### 2014-12-08 Toy Example/expected value/mu_parameterization.Rmd
initial commit

---
author: "Andee Kaplan"
date: "11/08/2014"
output:
  html_document:
    theme: readable
runtime: shiny
---

# $\mu$-Parametrization
In Random Graph Model theory, degeneracy occurs when the model places a disproportionate probability on only a few elements of the sample space, $\mathcal{X}$. Handcock, et al. define a model to be near degenerate if $\mu(\boldsymbol \theta)$, the mean parametrization on the model parameters, is close to the boundary of the convex hull of $\{t({\bf x}): {\bf x} \in \mathcal{X}\}$, where $t({\bf x})$ is the set of statistics in the neg-potential function $Q({\bf x})$ from the joint distribution of the node states.

This document will explore the behavior of the $\mu$-parametrized parameters for various dimensional models.

## Explorer

```{r, echo=FALSE, message=FALSE}

library(plyr)
library(dplyr)
library(reshape2)

stats <- function(H, V, type="negative") {
  if(type == "binary") t <- data.frame(c(0,1))[,rep(1, H + V)]
  else if(type == "negative") t <- data.frame(c(-1,1))[,rep(1, H + V)]
  else if(type == "experiment") t <- data.frame(c(1,3))[,rep(1, H + V)]
  else stop("Type must be 'binary', 'negative', or 'experiment'")

  names(t) <- c(paste0("h", 1:H), paste0("v", 1:V))

  #all possibilities of statistics in (H + V)-space
  t.grid <- expand.grid(t)

  for(i in 1:H){
    for(j in (H+1):(H+V)){
      t.grid <- cbind(t.grid, t.grid[,i]*t.grid[,j])
      names(t.grid)[ncol(t.grid)] <- paste0("theta", i, j - H)
    }
  }

  return(t.grid)
}

expected_value <- function(stats, theta, normalized = TRUE) {
  if(normalized) result <- t(as.matrix(stats)) %*% exp(as.matrix(stats) %*% as.matrix(theta)) %*% diag(1/apply(exp(as.matrix(stats) %*% as.matrix(theta)), 2, sum))
  else result <- t(as.matrix(stats)) %*% exp(as.matrix(stats) %*% as.matrix(theta))

  rownames(result) <- paste0("exp_", rownames(result))
  return(result)
}

reactiveSvg <- function(outputId) 
{
  div(id = outputId, class="d3graph")
}

shinyApp(
  ui = fluidPage(
      tags$head(
        includeScript("http://d3js.org/d3.v3.js"),
        includeCSS("www/css/surface3d.css"),
        includeCSS("www/css/linechart.css"),
        includeScript("www/js/surface3d.js"),
        includeScript("www/js/plots.js"),
        includeCSS("www/readable.min.css") 
      ),
      sidebarLayout(
        sidebarPanel(sliderInput("numHidden", label = "Number of hidden nodes:", 
                                 min = 1, max = 10, value = 1, step = 1),
                     sliderInput("numVisible", label = "Number of visible nodes:", 
                                 min = 1, max = 10, value = 1, step = 1),
                     radioButtons("encoding", label="Encoding:", 
                                  choices = list("Binary (0,1)" = "binary", "Negative (-1,1)" = "negative"),
                                  inline = FALSE),
                     selectizeInput("variableParams", "Parameters to vary:", 
                                    choices = NULL, multiple = TRUE,
                                    options = list(maxItems = 2))
        ),
        mainPanel(reactiveSvg(outputId = "surfacePlot"),
                  fluidRow(column(6, h4("Fixed Values"), uiOutput("fixedVals")),
                           column(6, h4("Variable Ranges"), uiOutput("variableVals"))
                  )
        )        
      )      
    ),
  server = function(input, output, session){
      stat_vals <- reactive({
        stats(H = input$numHidden, V = input$numVisible, type = input$encoding)
      })

      observe({
        vals <- stat_vals()
        updateSelectizeInput(session, inputId = "variableParams", choices = names(vals))
      })

      output$fixedVals <- renderUI({ 
        vals <- stat_vals()

        if(length(input$variableParams) > 0) {
          ## add fixed values
          lapply(names(vals)[!(names(vals) %in% input$variableParams)], function(name) {
            numericInput(paste0("fixedVal_", name), label = paste0(name, ":"), value=NULL)
          })    
        }
      })

      output$variableVals <- renderUI({
        vals <- stat_vals()

        if(length(input$variableParams) > 0) {
          ## add ranges for variable
          lapply(names(vals)[names(vals) %in% input$variableParams], function(name) {
            sliderInput(paste0("variableRange_", name), label = paste0(name, ":"), value=c(-10,10),
                        min=-20, max=20)
          })  
        }
      })

      output$surfacePlot <- reactive({ 
        vals <- stat_vals()

        if(length(input$variableParams) > 0) {
          fixed_vals <- data.frame(do.call("rbind", lapply(names(vals)[!(names(vals) %in% input$variableParams)], function(z) c(z, input[[paste0("fixedVal_", z)]]))))
          variable_ranges <- data.frame(do.call("rbind", lapply(names(vals)[names(vals) %in% input$variableParams], function(z) c(z, input[[paste0("variableRange_", z)]]))))

          if(ncol(fixed_vals) == 2) {
            names(fixed_vals) <- c("param","val")
            names(variable_ranges) <- c("param", "min", "max")

            if(sum(complete.cases(fixed_vals)) == nrow(fixed_vals) & nrow(variable_ranges) > 0) {
              ranges <- data.frame(apply(variable_ranges, 1, function(x) seq(as.numeric(x["min"]), as.numeric(x["max"]), length.out = 40)))
              ranges.grid <- expand.grid(ranges)

              ranges.grid <- ranges.grid %>%
                cbind(matrix(rep(as.numeric(as.character(fixed_vals[,"val"])), nrow(ranges.grid)), nrow = nrow(ranges.grid), byrow = TRUE))
              names(ranges.grid) <- c(as.character(variable_ranges$param), as.character(fixed_vals$param))

              surf_plot <- nrow(variable_ranges) == 2

              exp.grid0 <- t(expected_value(vals[,names(ranges.grid)], t(ranges.grid)))
              exp.grid <- cbind(exp.grid0, data.frame(ranges.grid[, variable_ranges$param])) 
              names(exp.grid) <- c(colnames(exp.grid0), as.character(variable_ranges$param))

              surf_values <- melt(exp.grid, id.vars=as.character(variable_ranges$param)) %>%
                dlply(.(variable), function(x){
                  dcast(x, formula = formula(paste0(variable_ranges$param[1], "~", variable_ranges$param[2])), value.var = "value") %>%
                    select(-1)
                }) %>%          
                llply(function(x) {
                  names(x) <- 0:(ncol(x) - 1) 
                  return(x)
                })

              surf_values <- llply(surf_values, function(x) x*10)

              line_values <- melt(exp.grid, id.vars=as.character(variable_ranges$param)) %>%
                dlply(.(variable), function(x){
                  x %>% select(-variable) %>%
                    mutate_("x" = as.character(variable_ranges$param))
                })

              if(surf_plot) return(list(values=surf_values, surf_plot=surf_plot))
              else return(list(values=line_values, surf_plot=surf_plot))
            }
          }
        }    
      })  
    },
  options = list(width="100%", height=600)
  )
```

### 2014-12-08 Toy Example/writeup/writeup.Rnw
initial commit

\documentclass{article}
\usepackage{subcaption, float}
\usepackage{tikz}
\usepackage{tikz-3dplot}
\usetikzlibrary{arrows,%
                shapes,positioning}
\usepackage{amsmath}
\usepackage[bibstyle=numeric,citestyle=authoryear,backend=bibtex,sorting=nyt]{biblatex}
\addbibresource{references/refs.bib}

\title{Degeneracy in a restricted Boltzmann machine}
\date{}

\begin{document}
\maketitle

\section{Toy examples}
To explore the idea of graph degeneracy in a restricted Boltzmann machine, we will consider three very small toy examples that are increasing in complexity. The first model consists of one visible node $v_1$ and one hidden node $h_1$, both binary. The second model consists of two visible nodes $\{v_1, v_2\}$ and one hidden $h_1$, all binary. The third model has three visible nodes $\{v_1, v_2, v_3\}$ and two hidden nodes $\{h_1, h_2\}$, all binary. A schematic of these three machines can be found in figure~\ref{fig:models}. Within the framework, we will learn the weights on each edge $\{\theta_{ij}| i=1,\dots,H, j=1,\dots, V\}$, where $V$ is the number of visible nodes and $H$ the number of hiddens, as well as the individual weights on the states of the nodes (on/off) $\{\theta_{hi}, \theta_{vj} | i=1,,H, j=1,\dots,V\}$.

\begin{figure}
   \begin{subfigure}[b]{0.32\textwidth}
        \centering
        \resizebox{0.12\linewidth}{0.12\textheight}{\input{figures/model1.tikz}}
   \end{subfigure}
   \begin{subfigure}[b]{0.32\textwidth}
        \centering
        \resizebox{\linewidth}{!}{\input{figures/model2.tikz}}
   \end{subfigure}
   \begin{subfigure}[b]{0.32\textwidth}
        \centering
        \resizebox{\linewidth}{!}{\input{figures/model3.tikz}}
   \end{subfigure}
   \caption{The three toy examples ranging from least complicated to most complicated. Gray nodes indicate hidden variables and white nodes are visibles.}
   \label{fig:models}
\end{figure}

Within each model, we will have $V\times H + V + H$ parameters to fit. In our three toy models, this correponds to table~\ref{tab:paramcount}. To begin exploring the behavior of the Boltzmann machine in relation to different values of parameters, we begin by specifying the joint distribution for the values of the nodes.

\begin{table}[ht!]
\centering
\begin{tabular}{ l | r r r }
  Model & H & V & Number of parameters \\
  \hline
  1 & 1 & 1 & 3 \\
  2 & 1 & 2 & 5 \\
  3 & 2 & 3 & 11 \\
\end{tabular}
\caption{The number of parameters needed to estimate in each toy model.}
\label{tab:paramcount}
\end{table}

\subsection{Joint distribution}
Let ${\bf x} = \{h_1, \dots, h_H, v_1,\dots,v_V\}$ represent the states of the five nodes in our example. Then we posit a parametric form for probabilities corresponding to the state of each node being a 1.

\begin{align}
\label{eqn:pmf}
f_{\boldsymbol \theta} ({\bf x}) = \frac{\exp\left(\sum\limits_{i = 1}^H \sum\limits_{j=1}^V \theta_{ij} h_i v_j + \sum\limits_{i = 1}^H\theta_{hi} h_i + \sum\limits_{j = 1}^V\theta_{vj} v_j\right)}{\sum\limits_{{\bf x} \in \{0,1\}^{H+V}}\exp\left(\sum\limits_{i = 1}^H \sum\limits_{j=1}^V \theta_{ij} h_i v_j + \sum\limits_{i = 1}^H\theta_{hi} h_i + \sum\limits_{j = 1}^V\theta_{vj} v_j\right)} 
\end{align}

Let the normalizing constant in equation~\ref{eqn:pmf} be denoted $\gamma(\boldsymbol \theta)$ \parencite[pp. 163--167]{Vardeman602}. We can also refer to $Q({\bf x}) = \sum\limits_{i = 1}^H\sum\limits_{j=1}^V \theta_{ij} h_i v_j + \sum\limits_{i = 1}^H\theta_{hi} h_i + \sum\limits_{j = 1}^V\theta_{vj} v_j$ as the neg-potential function in our model. In our very small examples, this normalizing constant can actually be computed, due to the fact that $\max\limits_{\text{model 1, 2 or 3}}|\{0,1\}^{V+H}| = \Sexpr{2^5}$. We will exploit this design in order to compare varying sets of values for $\boldsymbol \theta$ as they affect the probability of seeing any particular state of our graph.

\subsection{Degeneracy}
In Random Graph Model theory, degeneracy occurs when the model places a disproportionate probability on only a few elements of the sample space, $\mathcal{X}$. Handcock, et al. define a model to be near degenerate if $\mu(\boldsymbol \theta)$, the mean parametrization on the model parameters, is close to the boundary of the convex hull of $\{t({\bf x}): {\bf x} \in \mathcal{X}\}$, where $t({\bf x})$ is the set of statistics in the neg-potential function $Q({\bf x})$ from equation~\ref{eqn:pmf} \parencite{handcock2003assessing}.

There is a package in \texttt{R} \parencite{r} called \texttt{geometry} \parencite{r-geometry} that calculates a convex hull from a set of points. We will leverage this to create the convex hulls of $t({\bf x})$ in each of our toy models. For a finite set of points, the convex hull is a convex polytope for any number of dimensions, whose vertices are some of the points in the input set \parencite{Swart198517}. In higher dimensions, even if the vertices of a convex polytope are known, construction of its faces is a non-trivial task \parencite{Avis1997265}. Thus we will explore analytically creating our hulls as well.

<<source, echo=FALSE, message=FALSE, warning=FALSE, results='hide'>>=
source("helpers/volume_relationship.R")

#function for pretty function printing
insert_fun <- function(name) {
  read_chunk(lines = capture.output(dump(name, '')), labels = paste(name, 'source', sep = '-'))
}
@

<<insert-calc_hull, echo=FALSE>>=
insert_fun('calc_hull')
<<calc_hull-source, eval=FALSE>>=
@

\section{Model 1}
In our model 1, $t({\bf x}) \in \{0,1\}^{3}$ or $t({\bf x}) \in \{-1,1\}^{3}$ depending on how we encode on and off in the nodes, so the convex hull can be conceptualized geometrically in only 3 dimensions. A rendering of the hull with binary and (1,-1) encoding can be found from multiple perspectives in figure~\ref{fig:hull_1}.

<<models>>=
m1_bin <- calc_hull(1,1)
m1_neg <- calc_hull(1,1, "negative")
@

We include the bounding hulls of unrestricted space $\{0,1\}^{3}$ or $\{-1,1\}^{3}$ to emphasize how much the restriction in our statistics by including the cross product terms incurs.

\begin{figure}[H]
\begin{subfigure}[b]{\textwidth}
\begin{subfigure}[b]{0.5\textwidth}
        \centering
        \resizebox{\linewidth}{!}{
          \tdplotsetmaincoords{60}{-60}
          \begin{tikzpicture}[tdplot_main_coords]
          \draw[dotted,->,black] (0,0,0) -- (3,0,0) node[anchor=west]{$x$};
          \draw[dotted,->] (0,0,0) -- (0,3,0) node[anchor=north east]{$y$};
          \draw[dotted,->] (0,0,0) -- (0,0,3) node[anchor=east]{$z$};

          \draw[thin,color=lightgray] (0,0,0) -- (1,0,0) -- (1,1,0) -- (0,1,0) -- (0,0,0);
          \draw[thin,color=lightgray] (0,0,0) -- (0,0,1) -- (1,0,1) -- (1,0,0) -- (0,0,0);
          \draw[thin,color=lightgray] (0,0,0) -- (0,0,1) -- (0,1,1) -- (0,1,0) -- (0,0,0);
          \draw[thin,color=lightgray] (0,0,1) -- (1,0,1) -- (1,1,1) -- (0,1,1) -- (0,0,1);
          \draw[thin,color=lightgray] (0,1,0) -- (1,1,0) -- (1,1,1) -- (0,1,1) -- (0,1,0);
          \draw[thin,color=lightgray] (1,1,0) -- (1,1,1) -- (1,0,1) -- (1,0,0) -- (1,1,0);

          <<, results='asis', echo=FALSE, include=TRUE, message=FALSE>>=
          cat(sapply(1:nrow(m1_bin$c_hull$hull), function(x) paste0("\\draw", paste(paste0("(", apply(m1_bin$possible_t[cbind(m1_bin$c_hull$hull, m1_bin$c_hull$hull[,1])[x,],], 1, paste, collapse=","), ")"), collapse=" -- "), "; \n")))
          @
          \end{tikzpicture}
        }
        \end{subfigure}
   \begin{subfigure}[b]{0.5\textwidth}
        \centering
        \resizebox{\linewidth}{!}{
          \tdplotsetmaincoords{60}{120}
          \begin{tikzpicture}[tdplot_main_coords]
          \draw[dotted,->,black] (0,0,0) -- (3,0,0) node[anchor=north east]{$x$};
          \draw[dotted,->] (0,0,0) -- (0,3,0) node[anchor=north east]{$y$};
          \draw[dotted,->] (0,0,0) -- (0,0,3) node[anchor=east]{$z$};

          \draw[thin,color=lightgray] (0,0,0) -- (1,0,0) -- (1,1,0) -- (0,1,0) -- (0,0,0);
          \draw[thin,color=lightgray] (0,0,0) -- (0,0,1) -- (1,0,1) -- (1,0,0) -- (0,0,0);
          \draw[thin,color=lightgray] (0,0,0) -- (0,0,1) -- (0,1,1) -- (0,1,0) -- (0,0,0);
          \draw[thin,color=lightgray] (0,0,1) -- (1,0,1) -- (1,1,1) -- (0,1,1) -- (0,0,1);
          \draw[thin,color=lightgray] (0,1,0) -- (1,1,0) -- (1,1,1) -- (0,1,1) -- (0,1,0);
          \draw[thin,color=lightgray] (1,1,0) -- (1,1,1) -- (1,0,1) -- (1,0,0) -- (1,1,0);

          <<,results='asis', echo=FALSE, include=TRUE, message=FALSE>>=
          cat(sapply(1:nrow(m1_bin$c_hull$hull), function(x) paste0("\\draw", paste(paste0("(", apply(m1_bin$possible_t[cbind(m1_bin$c_hull$hull, m1_bin$c_hull$hull[,1])[x,],], 1, paste, collapse=","), ")"), collapse=" -- "), "; \n")))
          @
          \end{tikzpicture}
        }
   \end{subfigure}
\end{subfigure}
\begin{subfigure}[b]{\textwidth}
\begin{subfigure}[b]{0.5\textwidth}
        \centering
        \resizebox{\linewidth}{!}{
          \tdplotsetmaincoords{60}{30}
          \begin{tikzpicture}[tdplot_main_coords]
          \draw[dotted,->,black] (0,0,0) -- (3,0,0) node[anchor=west]{$x$};
          \draw[dotted,->] (0,0,0) -- (0,3,0) node[anchor=north east]{$y$};
          \draw[dotted,->] (0,0,0) -- (0,0,3) node[anchor=east]{$z$};

          \draw[thin,color=lightgray] (-1,-1,-1) -- (1,-1,-1) -- (1,1,-1) -- (-1,1,-1) -- (-1,-1,-1);
          \draw[thin,color=lightgray] (-1,-1,-1) -- (-1,-1,1) -- (1,-1,1) -- (1,-1,-1) -- (-1,-1,-1);
          \draw[thin,color=lightgray] (-1,-1,-1) -- (-1,-1,1) -- (-1,1,1) -- (-1,1,-1) -- (-1,-1,-1);
          \draw[thin,color=lightgray] (-1,-1,1) -- (1,-1,1) -- (1,1,1) -- (-1,1,1) -- (-1,-1,1);
          \draw[thin,color=lightgray] (-1,1,-1) -- (1,1,-1) -- (1,1,1) -- (-1,1,1) -- (-1,1,-1);
          \draw[thin,color=lightgray] (1,1,-1) -- (1,1,1) -- (1,-1,1) -- (1,-1,-1) -- (1,1,-1);

          <<, results='asis', echo=FALSE, include=TRUE, message=FALSE>>=
          cat(sapply(1:nrow(m1_neg$c_hull$hull), function(x) paste0("\\draw", paste(paste0("(", apply(m1_neg$possible_t[cbind(m1_neg$c_hull$hull, m1_neg$c_hull$hull[,1])[x,],], 1, paste, collapse=","), ")"), collapse=" -- "), "; \n")))
          @
          \end{tikzpicture}
        }
        \end{subfigure}
   \begin{subfigure}[b]{0.5\textwidth}
        \centering
        \resizebox{\linewidth}{!}{
          \tdplotsetmaincoords{60}{120}
          \begin{tikzpicture}[tdplot_main_coords]
          \draw[dotted,->,black] (0,0,0) -- (3,0,0) node[anchor=north east]{$x$};
          \draw[dotted,->] (0,0,0) -- (0,3,0) node[anchor=north east]{$y$};
          \draw[dotted,->] (0,0,0) -- (0,0,3) node[anchor=east]{$z$};

          \draw[thin,color=lightgray] (-1,-1,-1) -- (1,-1,-1) -- (1,1,-1) -- (-1,1,-1) -- (-1,-1,-1);
          \draw[thin,color=lightgray] (-1,-1,-1) -- (-1,-1,1) -- (1,-1,1) -- (1,-1,-1) -- (-1,-1,-1);
          \draw[thin,color=lightgray] (-1,-1,-1) -- (-1,-1,1) -- (-1,1,1) -- (-1,1,-1) -- (-1,-1,-1);
          \draw[thin,color=lightgray] (-1,-1,1) -- (1,-1,1) -- (1,1,1) -- (-1,1,1) -- (-1,-1,1);
          \draw[thin,color=lightgray] (-1,1,-1) -- (1,1,-1) -- (1,1,1) -- (-1,1,1) -- (-1,1,-1);
          \draw[thin,color=lightgray] (1,1,-1) -- (1,1,1) -- (1,-1,1) -- (1,-1,-1) -- (1,1,-1);

          <<,results='asis', echo=FALSE, include=TRUE, message=FALSE>>=
          cat(sapply(1:nrow(m1_neg$c_hull$hull), function(x) paste0("\\draw", paste(paste0("(", apply(m1_neg$possible_t[cbind(m1_neg$c_hull$hull, m1_neg$c_hull$hull[,1])[x,],], 1, paste, collapse=","), ")"), collapse=" -- "), "; \n")))
          @
          \end{tikzpicture}
        }
   \end{subfigure}
\end{subfigure}
\caption{The convex hull of our statistic space in three dimensions for model 1 from multiple perspectives. The top two figures show the binary encoding (1, 0) for the nodes enclosed by the unrestricted hull of $\{0,1\}^3$ space. The bottom two figures show the (1, -1) encoding for the nodes eclosed by the unrestricted hull of $\{-1,1\}^3$.}
\label{fig:hull_1}
\end{figure}

In terms of lost volume, the binary encoding loses \Sexpr{round((1-m1_bin$c_hull$vol)*100,2)}\% of the volume compared to unrestricted space and the negative encoding loses \Sexpr{round((1-m1_neg$c_hull$vol/(2^3))*100,2)}\% of the volume compared to unrestricted space. This notion of lost volume can be helpful to conceptualize how difficult will be for the mean parameterized vector $\mu(\boldsymbol \theta)$ to avoid the boundary, and thus avoid near degeneracy. In fact, if we look at the ratio of volume within a convex hull versus the volume of an unrestricted hull, we can see a relationship emerge as the number of nodes (and thus parameters) increases. In figure~\ref{fig:volume_plot}, it is evident that as the number of parameters increses, this ratio is decreasing at an increasing rate, meaning it gets more and more difficult to avoid the boundary of the convex hull and thus near degeneracy. Additionally, it appears that the $\{-1,1\}$ encoding suffers slightly less from this problem.

<<volume_plot, echo=FALSE, fig.cap="The relationship between volume within the convex hull of our statistics and the convex hull of unrestricted space for different configurations of nodes. We compare the two encodings for on and off in a node.", fig.width=6, fig.height=4, out.width='.8\\textwidth', fig.align='center'>>=
volume_plot
@

For model 1, we can compute the $\mu$-parametrization of the model parameters for the $\{-1,1\}$ encoding as 
\begin{align*}
\mu(\boldsymbol \theta) &= \text{E}_{\boldsymbol \theta}\left[ t(X) \right] \\
&= \sum\limits_{x \in \{-1,1\}^3} \left\{ t(x) \frac{\exp\left( \theta_{11} h_1 v_1 + \theta_{h1} h_1 + \theta_{v1} v_1\right)}{\sum\limits_{{\bf x} \in \{-1,1\}^{3}}\exp\left( \theta_{11} h_1 v_1 + \theta_{h1} h_1 + \theta_{v1} v_1\right)} \right\} \\
&= \left[\begin{matrix}
\frac{e^{-\theta_{h1} - \theta_{v1} + \theta_{11}} - e^{-\theta_{h1} + \theta_{v1} - \theta_{11}} - e^{\theta_{h1} - \theta_{v1} - \theta_{11}} + e^{\theta_{h1} + \theta_{v1} + \theta_{11}}}{e^{-\theta_{h1} - \theta_{v1} + \theta_{11}} + e^{-\theta_{h1} + \theta_{v1} - \theta_{11}} + e^{\theta_{h1} - \theta_{v1} - \theta_{11}} + e^{\theta_{h1} + \theta_{v1} + \theta_{11}}} \\
\frac{-e^{-\theta_{h1} - \theta_{v1} + \theta_{11}} - e^{-\theta_{h1} + \theta_{v1} - \theta_{11}} + e^{\theta_{h1} - \theta_{v1} - \theta_{11}} + e^{\theta_{h1} + \theta_{v1} + \theta_{11}}}{e^{-\theta_{h1} - \theta_{v1} + \theta_{11}} + e^{-\theta_{h1} + \theta_{v1} - \theta_{11}} + e^{\theta_{h1} - \theta_{v1} - \theta_{11}} + e^{\theta_{h1} + \theta_{v1} + \theta_{11}}} \\
\frac{-e^{-\theta_{h1} - \theta_{v1} + \theta_{11}} + e^{ -\theta_{h1} + \theta_{v1} - \theta_{11}} - e^{\theta_{h1} - \theta_{v1} - \theta_{11}} + e^{\theta_{h1} + \theta_{v1} + \theta_{11}}}{e^{-\theta_{h1} - \theta_{v1} + \theta_{11}} + e^{-\theta_{h1} + \theta_{v1} - \theta_{11}} + e^{\theta_{h1} - \theta_{v1} - \theta_{11}} + e^{\theta_{h1} + \theta_{v1} + \theta_{11}}} \\
\end{matrix} \right]
\end{align*}

%\section{Model 3}
%In our toy example, $t({\bf x}) \in \{0,1\}^{11}$ and below we attempt to compute the convex hull of this set in 11-space.

%Luckily the vertices of all our points lie on a set of zeros and ones only. To attempt to get some understanding of what our hull consists of, we present the 2D and 3D cases in figure~\ref{fig:hull}.

%With our toy example then, $\mathcal{C} = \text{ the convex hull of } \{t({\bf x}): {\bf x} \in \mathcal{X}\}$ is the 11-dimensional unit hypercube.

%We can compute the $\mu$-parametrization of the model parameters as $\mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta}\left[ t(X) \right]$.

\printbibliography
\end{document}

### 2014-12-08 Neural Nets - Hinton Course/Programming Assignments/assignment1/assignment.m

@@ -1,0 +1,14 @@
+load dataset1_ancient_octave.mat
+w1 = learn_perceptron(neg_examples_nobias,pos_examples_nobias,w_init,w_gen_feas)
+
+load dataset2_ancient_octave.mat
+w2 = learn_perceptron(neg_examples_nobias,pos_examples_nobias,w_init,w_gen_feas)
+
+
+load dataset3_ancient_octave.mat
+w3 = learn_perceptron(neg_examples_nobias,pos_examples_nobias,w_init,w_gen_feas)
+
+
+load dataset4_ancient_octave.mat
+w4 = learn_perceptron(neg_examples_nobias,pos_examples_nobias,w_init,w_gen_feas)
+

### 2014-12-08 Neural Nets - Hinton Course/Programming Assignments/assignment1/learn_perceptron.m

@@ -1,0 +1,162 @@
+%% Learns the weights of a perceptron and displays the results.
+function [w] = learn_perceptron(neg_examples_nobias,pos_examples_nobias,w_init,w_gen_feas)
+%% 
+% Learns the weights of a perceptron for a 2-dimensional dataset and plots
+% the perceptron at each iteration where an iteration is defined as one
+% full pass through the data. If a generously feasible weight vector
+% is provided then the visualization will also show the distance
+% of the learned weight vectors to the generously feasible weight vector.
+% Required Inputs:
+%   neg_examples_nobias - The num_neg_examples x 2 matrix for the examples with target 0.
+%       num_neg_examples is the number of examples for the negative class.
+%   pos_examples_nobias - The num_pos_examples x 2 matrix for the examples with target 1.
+%       num_pos_examples is the number of examples for the positive class.
+%   w_init - A 3-dimensional initial weight vector. The last element is the bias.
+%   w_gen_feas - A generously feasible weight vector.
+% Returns:
+%   w - The learned weight vector.
+%%
+
+%Bookkeeping
+num_neg_examples = size(neg_examples_nobias,1);
+num_pos_examples = size(pos_examples_nobias,1);
+num_err_history = [];
+w_dist_history = [];
+
+%Here we add a column of ones to the examples in order to allow us to learn
+%bias parameters.
+neg_examples = [neg_examples_nobias,ones(num_neg_examples,1)];
+pos_examples = [pos_examples_nobias,ones(num_pos_examples,1)];
+
+%If weight vectors have not been provided, initialize them appropriately.
+if (~exist('w_init','var') || isempty(w_init))
+    w = randn(3,1);
+else
+    w = w_init;
+end
+
+if (~exist('w_gen_feas','var'))
+    w_gen_feas = [];
+end
+
+%Find the data points that the perceptron has incorrectly classified
+%and record the number of errors it makes.
+iter = 0;
+[mistakes0, mistakes1] = eval_perceptron(neg_examples,pos_examples,w);
+num_errs = size(mistakes0,1) + size(mistakes1,1);
+num_err_history(end+1) = num_errs;
+fprintf('Number of errors in iteration %d:\t%d\n',iter,num_errs);
+fprintf(['weights:\t', mat2str(w), '\n']);
+plot_perceptron(neg_examples, pos_examples, mistakes0, mistakes1, num_err_history, w, w_dist_history);
[…]

### 2014-12-08 Neural Nets - Hinton Course/Programming Assignments/assignment2/fprop.m

@@ -1,0 +1,86 @@
+function [embedding_layer_state, hidden_layer_state, output_layer_state] = ...
+  fprop(input_batch, word_embedding_weights, embed_to_hid_weights,...
+  hid_to_output_weights, hid_bias, output_bias)
+% This method forward propagates through a neural network.
+% Inputs:
+%   input_batch: The input data as a matrix of size numwords X batchsize where,
+%     numwords is the number of words, batchsize is the number of data points.
+%     So, if input_batch(i, j) = k then the ith word in data point j is word
+%     index k of the vocabulary.
+%
+%   word_embedding_weights: Word embedding as a matrix of size
+%     vocab_size X numhid1, where vocab_size is the size of the vocabulary
+%     numhid1 is the dimensionality of the embedding space.
+%
+%   embed_to_hid_weights: Weights between the word embedding layer and hidden
+%     layer as a matrix of soze numhid1*numwords X numhid2, numhid2 is the
+%     number of hidden units.
+%
+%   hid_to_output_weights: Weights between the hidden layer and output softmax
+%               unit as a matrix of size numhid2 X vocab_size
+%
+%   hid_bias: Bias of the hidden layer as a matrix of size numhid2 X 1.
+%
+%   output_bias: Bias of the output layer as a matrix of size vocab_size X 1.
+%
+% Outputs:
+%   embedding_layer_state: State of units in the embedding layer as a matrix of
+%     size numhid1*numwords X batchsize
+%
+%   hidden_layer_state: State of units in the hidden layer as a matrix of size
+%     numhid2 X batchsize
+%
+%   output_layer_state: State of units in the output layer as a matrix of size
+%     vocab_size X batchsize
+%
+
+[numwords, batchsize] = size(input_batch);
+[vocab_size, numhid1] = size(word_embedding_weights);
+numhid2 = size(embed_to_hid_weights, 2);
+
+%% COMPUTE STATE OF WORD EMBEDDING LAYER.
+% Look up the inputs word indices in the word_embedding_weights matrix.
+embedding_layer_state = reshape(...
+  word_embedding_weights(reshape(input_batch, 1, []),:)',...
+  numhid1 * numwords, []);
+
+%% COMPUTE STATE OF HIDDEN LAYER.
+% Compute inputs to hidden units.
[…]

### 2014-12-08 Neural Nets - Hinton Course/Programming Assignments/assignment2/train.m

@@ -1,0 +1,240 @@
+% This function trains a neural network language model.
+function [model] = train(epochs)
+% Inputs:
+%   epochs: Number of epochs to run.
+% Output:
+%   model: A struct containing the learned weights and biases and vocabulary.
+
+if size(ver('Octave'),1)
+  OctaveMode = 1;
+  warning('error', 'Octave:broadcast');
+  start_time = time;
+else
+  OctaveMode = 0;
+  start_time = clock;
+end
+
+% SET HYPERPARAMETERS HERE.
+batchsize = 100;  % Mini-batch size.
+learning_rate = 0.1;  % Learning rate; default = 0.1.
+momentum = 0.9;  % Momentum; default = 0.9.
+numhid1 = 50;  % Dimensionality of embedding space; default = 50.
+numhid2 = 200;  % Number of units in hidden layer; default = 200.
+init_wt = 0.01;  % Standard deviation of the normal distribution
+                 % which is sampled to get the initial weights; default = 0.01
+
+% VARIABLES FOR TRACKING TRAINING PROGRESS.
+show_training_CE_after = 100;
+show_validation_CE_after = 1000;
+
+% LOAD DATA.
+[train_input, train_target, valid_input, valid_target, ...
+  test_input, test_target, vocab] = load_data(batchsize);
+[numwords, batchsize, numbatches] = size(train_input); 
+vocab_size = size(vocab, 2);
+
+% INITIALIZE WEIGHTS AND BIASES.
+word_embedding_weights = init_wt * randn(vocab_size, numhid1);
+embed_to_hid_weights = init_wt * randn(numwords * numhid1, numhid2);
+hid_to_output_weights = init_wt * randn(numhid2, vocab_size);
+hid_bias = zeros(numhid2, 1);
+output_bias = zeros(vocab_size, 1);
+
+word_embedding_weights_delta = zeros(vocab_size, numhid1);
+word_embedding_weights_gradient = zeros(vocab_size, numhid1);
+embed_to_hid_weights_delta = zeros(numwords * numhid1, numhid2);
+hid_to_output_weights_delta = zeros(numhid2, vocab_size);
+hid_bias_delta = zeros(numhid2, 1);
+output_bias_delta = zeros(vocab_size, 1);
+expansion_matrix = eye(vocab_size);
+count = 0;
+tiny = exp(-30);
+
+% TRAIN.
+for epoch = 1:epochs
+  fprintf(1, 'Epoch %d\n', epoch);
+  this_chunk_CE = 0;
[…]

### 2014-12-08 Toy Example/expected value/closest_points.R

@@ -1,0 +1,54 @@
+source("functions.R")
+library(scatterplot3d)
+library(plyr)
+library(dplyr)
+library(reshape2)
+library(ggplot2)
+
+###############################################
+## Generate data
+###############################################
+
+## Num nodes
+H <- 1
+V <- 1
+
+## Num to sample
+n <- 10000
+
+## Radius to sample
+r <- c(seq(0,1, by=0.05), seq(1.5, 10, by=0.5))
+
+stat <- stats(H, V, "negative")
+theta <- mlply(r, function(x) sample_sphere(space = ncol(stat), n, x))
+exp_vals <- llply(theta, function(x) expected_value(stats = stat, theta = x, normalized = TRUE))
+lines <- generate_hull_from_stats(stat)
+
+dist <- ldply(exp_vals, function(y) apply(y, 2, function(x){ min_distance_to_hull(x, lines) }))
+min_dist <- apply(dist[,-1], 1, min)
+min_point <- apply(dist[,-1], 1, which.min)
+
+## Data frame of points at each radius with minimum distance to a closest line
+min_points <- ldply(1:length(theta), function(i) theta[[i]][, min_point[i]])
+names(min_points) <- names(stat)
+min_points$r <- r
+min_points$min_dist <- min_dist
+
+###############################################
+## Plots
+###############################################
+
+## Plot the progression of "worst" parameter
+s <- scatterplot3d(min_points[, 1:3], pch=19)
+segs <- NULL
+for(i in 1:(nrow(min_points)-1)) {
+  start <- s$xyz.convert(min_points[i, 1:3])
+  end <- s$xyz.convert(min_points[i + 1, 1:3])
+  segs <- rbind(segs, c(start$x, start$y, end$x, end$y))
+}
+segments(segs[,1], segs[,2], segs[,3], segs[,4], lty=2)
+## Doesn't look like any sort of pattern
+
+## Plot of min_dist by radius
+qplot(r, min_dist, colour = min_dist < 1e-4, data=min_points) + geom_hline(aes(yintercept=1e-4))
+

### 2014-12-08 Toy Example/expected value/functions.R

@@ -1,0 +1,81 @@
+
+stats <- function(H, V, type="negative") {
+  if(type == "binary") t <- data.frame(c(0,1))[,rep(1, H + V)]
+  else if(type == "negative") t <- data.frame(c(-1,1))[,rep(1, H + V)]
+  else if(type == "experiment") t <- data.frame(c(1,3))[,rep(1, H + V)]
+  else stop("Type must be 'binary', 'negative', or 'experiment'")
+  
+  names(t) <- c(paste0("h", 1:H), paste0("v", 1:V))  
+  
+  #all possibilities of statistics in (H + V)-space
+  t.grid <- expand.grid(t)
+  
+  for(i in 1:H){
+    for(j in (H+1):(H+V)){
+      t.grid <- cbind(t.grid, t.grid[,i]*t.grid[,j])
+      names(t.grid)[ncol(t.grid)] <- paste0("theta", i, j - H)
+    }
+  }
+  
+  return(t.grid)
+}
+
+expected_value <- function(stats, theta, normalized = TRUE) {
+  if(normalized) result <- t(as.matrix(stats)) %*% exp(as.matrix(stats) %*% as.matrix(theta)) %*% diag(1/apply(exp(as.matrix(stats) %*% as.matrix(theta)), 2, sum))
+  else result <- t(as.matrix(stats)) %*% exp(as.matrix(stats) %*% as.matrix(theta))
+  
+  rownames(result) <- paste0("exp_", rownames(result))
+  return(result)
+}
+
+sample_sphere <- function(space, n, r = 1) {
+  return(data.frame(sapply(1:n, function(i){
+    vec <- rnorm(space)
+    r*vec/sqrt(sum(vec^2))
+  })))  
+}
+
+calc_hull <- function(stats) {
+  #calculate convex hull
+  require(geometry)
+  C <- convhulln(stats, options="FA")
+  return(C)
+}
+
+
+generate_hull_from_stats <- function(stats) {
+  ## stat is the df of possible statistics as generated from stats
+  lines <- NULL
+  for(i in 1:nrow(stats)){
+    for(j in 1:nrow(stats)){
+      if(i < j) {
+        lines <- c(lines, list(list(x1=stats[i,], x2=stats[j,])))
+      }
+    }
+  }
+  
+  return(lines)
+}
+
+
+min_distance_to_hull <- function(x0, lines) {
+  ## x0 is a single point (numeric vector)
[…]

### 2014-12-08 Toy Example/expected value/random_search.R

@@ -1,0 +1,89 @@
+source("functions.R")
+library(scatterplot3d)
+library(plyr)
+library(dplyr)
+library(reshape2)
+library(ggplot2)
+
+
+###############################################
+## Generate data
+###############################################
+
+## Num nodes
+H <- 1
+V <- 1
+
+## Num to sample
+n <- 10000
+
+## Radius to sample
+r <- c(seq(0,1, by=0.05), seq(1.5, 10, by=0.5))
+
+stat <- stats(H, V, "negative")
+theta <- mlply(r, function(x) sample_sphere(space = ncol(stat), n, x))
+exp_vals <- llply(theta, function(x) expected_value(stats = stat, theta = x, normalized = TRUE))
+lines <- generate_hull_from_stats(stat)
+
+dist <- ldply(exp_vals, function(y) apply(y, 2, function(x){ min_distance_to_hull(x, lines) }))
+min_dist <- apply(dist[,-1], 1, min)
+min_point <- apply(dist[,-1], 1, which.min)
+
+## Data frame of points at each radius with minimum distance to a closest line
+min_points <- ldply(1:length(theta), function(i) theta[[i]][, min_point[i]])
+names(min_points) <- names(stat)
+min_points$r <- r
+min_points$min_dist <- min_dist
+
+## Start with first radius that has a point within .01 of a line
+cut <- .01
+sigma <- .5
+n0 <- 500
+
+mu0 <- min_points[min_dist < cut,][1, 1:ncol(stat)] 
+sphere = list(center = c(0,0,0), radius = min(min_points[min_dist < cut, "r"]))
+
+
+local_search <- function(mu0, sigma, sphere, dim = 3, n = 500, cutoff = .01) {
+  store_mus <- rbind(mu0, c(0,0,0))
+  
+  search <- function(mu0, sigma, sphere, dim, n, cutoff) {
+    local_theta <- t(matrix(rnorm(n*dim, sd = sigma), nrow=dim) + mu0)
+    proj_theta <- t(apply(local_theta, 1, function(point) project_to_sphere(sphere=sphere, point=point)))
+    exp_vals <- expected_value(stats = stat, theta = t(proj_theta), normalized = TRUE)
+    dist <- apply(exp_vals, 2, function(x)  min_distance_to_hull(x, lines))
+    mu1 <- proj_theta[which.min(dist),]
[…]

### 2014-12-08 Toy Example/expected value/www/js/surface3d.js

@@ -1,0 +1,143 @@
+(function(){
+  var Surface=function(node){
+    var heightFunction,colorFunction,timer,timer,transformPrecalc=[];
+    var displayWidth=300, displayHeight=300, zoom=1;
+    var trans;
+    
+    var data = [$.map(node.datum(), function(value, index) {  return [value]; })];
+
+    this.setZoom=function(zoomLevel){
+      zoom=zoomLevel;
+      if(timer) clearTimeout(timer);
+      timer=setTimeout(renderSurface);
+    };
+    var getHeights=function(){
+      
+      var output=[];
+      var xlength=data.length;
+      var ylength=data[0].length;
+      for(var x=0;x<xlength;x++){
+        output.push(t=[]);
+        for(var y=0;y<ylength;y++){
+            var value=heightFunction(data[x][y],x,y);
+            t.push(value);
+        }
+      }
+      return output;
+    };
+    var transformPoint=function(point){
+      var x=transformPrecalc[0]*point[0]+transformPrecalc[1]*point[1]+transformPrecalc[2]*point[2];
+      var y=transformPrecalc[3]*point[0]+transformPrecalc[4]*point[1]+transformPrecalc[5]*point[2];
+      var z=transformPrecalc[6]*point[0]+transformPrecalc[7]*point[1]+transformPrecalc[8]*point[2];
+      return [x,y,z];
+    };
+    var getTransformedData=function(){
+      
+      if(!heightFunction) return [[]];
+      var t, output=[];
+      var heights=getHeights();
+
+      var xlength=data[0].length;
+      var ylength=data[0][0].length;
+      
+      for(var x=0;x<xlength;x++){
+        output.push(t=[]);
+        for(var y=0;y<ylength;y++){
+          t.push(transformPoint([(x-xlength/2)/(xlength*1.41)*displayWidth*zoom, heights[0][x][y]*zoom, (y-ylength/2)/(ylength*1.41)*displayWidth*zoom]));
+        }
+      }
+      
+      return output;
+    };
+    var renderSurface=function(){
+      var newdata=getTransformedData();
+      var xlength=newdata.length;
+      var ylength=newdata[0].length;
+      var d0=[];
+      var idx=0;
+      for(var x=0;x<xlength-1;x++){
+        for(var y=0;y<ylength-1;y++){
+          var depth=newdata[x][y][2]+newdata[x+1][y][2]+newdata[x+1][y+1][2]+newdata[x][y+1][2];
[…]

### 2014-12-08 Toy Example/writeup/helpers/volume_relationship.R

@@ -1,0 +1,60 @@
+library(plyr)
+library(dplyr)
+library(tidyr)
+library(ggplot2)
+
+calc_hull <- function(H, V, type="binary") {
+  #only possible values of t are 0 or 1
+  if(type == "binary") t <- data.frame(c(0,1))[,rep(1, H + V)]
+  else if(type == "negative") t <- data.frame(c(-1,1))[,rep(1, H + V)]
+  else if(type == "experiment") t <- data.frame(c(1,3))[,rep(1, H + V)]
+  else stop("Type must be 'binary', 'negative', or 'experiment'")
+  
+  names(t) <- c(paste0("h", 1:H), paste0("v", 1:V))  
+  
+  #all possibilities of statistics in (H + V)-space
+  t.grid <- expand.grid(t)
+
+  for(i in 1:H){
+    for(j in (H+1):(H+V)){
+      t.grid <- cbind(t.grid, t.grid[,i]*t.grid[,j])
+      names(t.grid)[ncol(t.grid)] <- paste0("theta", i, j - H)
+    }
+  }
+  
+  #calculate convex hull
+  library(geometry)
+  C <- convhulln(t.grid, options="FA")
+  return(list(possible_t=t.grid, c_hull=C))
+}
+
+test_cases <- expand.grid(data.frame(1:4)[,rep(1,2)]) %>% 
+  rename(H = X1.4, V = X1.4.1) %>%
+  filter(H <= V) %>% #remove those cases with less visibles than hiddens. Is this necessary?
+  mutate(n_param = H*V + H + V) %>%
+  mutate(max_facets = (2^(H+V))^(floor(n_param/2))) %>%
+  filter(n_param <= 11) #calc_hull can't handle any higher dimensions currently
+
+bin_res <- dlply(test_cases, .(n_param), function(x) calc_hull(x$V, x$H, "binary"), .progress="text")
+neg_res <- dlply(test_cases, .(n_param), function(x) calc_hull(x$V, x$H, "negative"), .progress="text")
+
+volume_plot <- ldply(bin_res, function(x) x$c_hull$vol) %>% 
+  mutate(frac_vol = V1/(1^n_param)) %>%
+  inner_join(ldply(neg_res, function(x) x$c_hull$vol) %>% 
+               mutate(frac_vol = V1/(2^n_param)),
+             by="n_param") %>%
[…]

### 2015-03-02 Toy Example/expected value/degen_magnitude.Rmd
example for meeting

---
title: "Degeneracy and parameter magnitude"
author: "Andee Kaplan"
date: "March 1, 2015"
output:
  html_document:
    theme: readable
runtime: shiny
---

```{r libraries, message=FALSE, warning=FALSE, echo=FALSE}
library(ggvis)
library(ggplot2)
library(plyr)
library(dplyr)
library(tidyr)
library(shiny)
```

```{r data, message=FALSE, warning=FALSE, echo=FALSE}
load("results.RData")

## params ----------------------------------
n <- 100
r_center <- seq(0.2, 3, by = 0.2)
r_exp <- .05

plot_data <- function(res) {

  plot.data <- data.frame()
  for(i in 1:nrow(res)) {
    tmp <- res$samp[[i]]
    H <- res[i,]$H
    V <- res[i,]$V

    tmp %>% 
      data.frame() %>% 
      rowwise() %>% 
      mutate_(ss_ratio = paste0("(", paste(paste0(names(.)[(H + V + 1):(H + V + H*V)], "^2"), collapse = " + "), ")/sum(", paste(paste0(names(.)[1:(H+V)],"^2"), collapse = " + "), ")")) %>%    
      ungroup() -> ratio

    inner_join(ratio, res$outside[[i]] %>% ungroup()) %>%
      select(ss_ratio, near_hull) %>%
      mutate(H = H, V = V, n_param = H + V + H*V) %>%
      rbind(plot.data) -> plot.data
  }
  return(plot.data)
}

plot.dat <- lapply(res, plot_data)
names(plot.dat) <- gsub("[.]", "_", r_center)

do.call(rbind, plot.dat) %>%
  mutate(name = rownames(do.call(rbind, plot.dat))) %>%
  separate(name, into = c("multiplier", "num"), "[.]") %>% 
  mutate(multiplier = as.numeric(gsub("_", ".", multiplier))) %>%
  mutate(n_param_f = factor(n_param)) %>%
  group_by(n_param_f, multiplier) %>%
  summarise(frac_degen = sum(near_hull)/n()) -> ggvis.dat
```

```{r plots, message=FALSE, warning=FALSE, echo=FALSE, out.height=300}
multiplier <- reactiveValues(current = 0.2)

on_click <- function(data, ...) {
  if(nrow(data) > 0) multiplier$current <- data$multiplier
}

ggvis.subset <- reactive({
  do.call(rbind, plot.dat) %>%
    mutate(name = rownames(do.call(rbind, plot.dat))) %>%
    mutate(n_param_f = factor(n_param)) %>%
    separate(name, into = c("multiplier", "num"), "[.]") %>% 
    mutate(multiplier = as.numeric(gsub("_", ".", multiplier))) %>%
    filter(multiplier == multiplier$current) 
})

ggvis.dat %>%
  ggvis(~multiplier, ~frac_degen) %>%
  layer_points(fill = ~n_param_f) %>% 
  layer_lines(stroke = ~n_param_f) %>%
  add_axis("x", title = "Radius Multiplier") %>%
  add_axis("y", title = "Fraction Degenerate") %>%
  add_legend("fill", title = "# Parameters") %>%
  add_legend("stroke", title = "# Parameters") %>%
  handle_click(on_click)

```

```{r plots2, message=FALSE, warning=FALSE, echo=FALSE, out.height=300}

ggvis.subset %>%
  group_by(multiplier, n_param_f, near_hull) %>%
  summarise(num = n()) %>%
  ggvis(~n_param_f, ~num) %>%
  layer_bars(fill = ~as.character(near_hull)) %>%
  add_axis("x", title = "# Parameters") %>%
  add_axis("y", title = "Count") %>%
  add_legend("fill", title = "Within .05 of Hull")

# ggvis.subset %>%
#   ggvis(~n_param_f, ~ss_ratio) %>%
#   layer_points(fill = ~as.character(near_hull), opacity := 0.4, size := 200) %>%
#   add_axis("x", title = "# Parameters") %>%
#   add_axis("y", title = "Ratio of SS of cross terms to main effects") %>%
#   add_legend("fill", title = "Within 0.05 of Hull")

renderPlot(
  ggvis.subset() %>%
    ggplot() + 
    geom_boxplot(aes(n_param_f, ss_ratio, colour = near_hull)) +
    xlab("# Parameters") +
    ylab("Ratio of sum of squares of cross terms to main effects") +
    theme_bw() +
    scale_colour_discrete("Within .05 of hull")
)
```

#Code Appendix
```{r eval = FALSE}
## load libraries ---------------------------
library(dplyr)
library(tidyr)
library(ggplot2)

## background functions -------------------------------------
stats <- function(H, V, type="negative") {
  if(type == "binary") t <- data.frame(c(0,1))[,rep(1, H + V)]
  else if(type == "negative") t <- data.frame(c(-1,1))[,rep(1, H + V)]
  else stop("Type must be 'binary', 'negative'")

  names(t) <- c(paste0("h", 1:H), paste0("v", 1:V))

  #all possibilities of statistics in (H + V)-space
  t.grid <- expand.grid(t)

  for(i in 1:H){
    for(j in (H+1):(H+V)){
      t.grid <- cbind(t.grid, t.grid[,i]*t.grid[,j])
      names(t.grid)[ncol(t.grid)] <- paste0("theta", i, j - H)
    }
  }

  return(data.matrix(t.grid))
}

expected_value <- function(theta, stats, normalized = TRUE) {
  result <- crossprod(t(crossprod(stats, exp(crossprod(t(stats), theta)))), diag(1/apply(exp(crossprod(t(stats), theta)), 2, sum), nrow=length(apply(exp(crossprod(t(stats), theta)), 2, sum))))
  rownames(result) <- paste0("exp_", rownames(result))
  return(result)
}

sample_sphere <- function(stat, n, r = 1) {
  require(dplyr)
  require(tidyr)

  samp <- function(space, r) {
    vec <- rnorm(space)
    paste0(r*vec/sqrt(sum(vec^2)), collapse=",")
  }

  res <- data.frame(samp = 1:n) %>%
    group_by(samp) %>%
    mutate(vals = samp(ncol(stat), r)) %>%
    separate(col = vals, sep = ",", into = colnames(stat)) %>%
    select(-samp) %>%
    data.matrix() %>%
    t()

  return(res)  
}

## main function --------------------------------------------
find_prop <- function(g_theta, stat, r_exp) {
  g_theta %>%
    data.frame() %>%
    group_by_(.dots = c(colnames(g_theta))) %>%
    do(samp = sample_points_outside(as.numeric(.), stat, r_exp)$in_hull) %>%
    group_by_(.dots = c(colnames(g_theta))) %>%
    do(as.data.frame(sum(.$samp[[1]]) > 0)) -> is_outside # count of not in hull greater than zero => there exists a point outside hull for this model

  is_outside %>% 
    rename_(near_hull = as.name(names(is_outside)[ncol(is_outside)])) %>%
    mutate(r = r_exp) -> inside_outside

  return(inside_outside)
}

## additional useful functions -------------------------------------
sample_points_outside <- function(g_theta, stats, r, n = 100) {
  #g_theta is the expected value mapping of a point theta in R^H+V+H*V
  x0 <- data.frame(samp = 1:n) %>%
    mutate(vals = samp_exp(length(g_theta), g_theta, r)) %>%
    separate(col = vals, sep = ",", into = colnames(stats)) %>%
    select(-samp)

  x0$in_hull <- !apply(x0, 1, in_hull, stats)

  return(x0)
}

samp_exp <- function(space, center, r) {
  vec <- rnorm(space)
  paste0(r*vec/sqrt(sum(vec^2)) + center, collapse=",")
}

in_hull <- function(point, hull_points) {
  require(lpSolveAPI)
  P <- data.matrix(point)
  A <- t(data.matrix(hull_points))

  lp_obj <- make.lp(nrow = 0, ncol = ncol(A))
  sapply(1:nrow(A), function(x) add.constraint(lp_obj, A[x,], type = "=", rhs = P[x]))
  add.constraint(lp_obj, rep(1, ncol(A)), type = "=", rhs = 1)
  sapply(1:ncol(A), function(x) add.constraint(lp_obj, 1, type = ">=", rhs = 0, indices = x))

  return(solve(lp_obj) == 0)
}

## params ----------------------------------
n <- 100
r_center <- seq(0.2, 3, by = 0.2)
#r_center <- 1
r_exp <- .05

## function for running simulation -------------------------------
get_res <- function(n, r_center, r_exp) {
  ## perform functions 
  test_cases <- expand.grid(data.frame(1:4)[,rep(1,2)]) %>% 
    rename(H = X1.4, V = X1.4.1) %>%
    filter(H <= V) %>% #remove those cases with less visibles than hiddens. Is this necessary?
    mutate(n_param = H*V + H + V) %>%
    mutate(max_facets = (2^(H+V))^(floor(n_param/2))) %>%
    mutate(n = n, r = r_center*(H + V)) %>%
    group_by(H, V, n_param, n, r) %>%
    do(stat = stats(.$H, .$V, "negative")) %>%
    ungroup() %>%
    group_by(H, V, n_param, n, r) %>%
    do(samp = sample_sphere(.$stat[[1]], .$n, .$r))

  test_cases_stat <- expand.grid(data.frame(1:4)[,rep(1,2)]) %>% 
    rename(H = X1.4, V = X1.4.1) %>%
    filter(H <= V) %>% #remove those cases with less visibles than hiddens. Is this necessary?
    mutate(n_param = H*V + H + V) %>%
    mutate(max_facets = (2^(H+V))^(floor(n_param/2))) %>%
    mutate(n = n, r = r_center*(H + V)) %>%
    group_by(H, V, n_param, n, r) %>%
    do(stat = stats(.$H, .$V, "negative"))

  test_cases <- inner_join(test_cases, test_cases_stat)

  test_cases %>%
    group_by(H, V, n_param, n, r) %>%
    do(g_theta = t(rbind(.$samp[[1]], expected_value(theta = .$samp[[1]], stats = data.matrix(.$stat[[1]]))))) -> cases

  cases <- inner_join(cases, test_cases)

  cases %>%
    group_by(H, V, r) %>%
    do(outside = find_prop(.$g_theta[[1]] %>% data.frame() %>% select(starts_with("exp")), .$stat[[1]], r_exp)) -> tmp

  res <- inner_join(tmp, cases %>% group_by(H, V, r, n_param) %>% do(samp = .$g_theta[[1]])) 
}

## run simulations -------------------------------------
res <- lapply(r_center, function(r) get_res(n, r, r_exp))
save(res, file = "results.RData")

```

### 2015-03-02 Toy Example/expected value/random_search.R

@@ -67,8 +67,8 @@ local_search <- function(mu0, sigma, sphere, dim = 3, n = 500, cutoff = .01) {
 
 mus <- lapply(1:100, function(i){
   cat("iteration = ", i, "\n")
-  local_search(mu0, sigma, sphere, cutoff=.005))
-} 
+  local_search(mu0, sigma, sphere, cutoff=.005)
+})
 
 
 ###############################################

### 2015-03-02 Toy Example/expected value/apps/degen_magnitude.Rmd
file structure for apps

--- 2015-03-02 Toy Example/expected value/degen_magnitude.Rmd
@@
   ggvis(~multiplier, ~frac_degen) %>%
-  layer_points(fill = ~n_param_f) %>% 
+  layer_points(fill = ~n_param_f, size.hover := 200) %>% 
   layer_lines(stroke = ~n_param_f) %>%

### 2015-03-09 Toy Example/expected value/apps/degen_magnitude.Rmd
added powers

@@
 ```{r data, message=FALSE, warning=FALSE, echo=FALSE}
-load("results.RData")
 
@@
 r_exp <- .05
+sliderInput("power", "Power of N:", min = 0, max = 1, value = 0, step = 0.1)
 
@@
 }
+plot.dat <- list()
+for(i in seq(0, 1, by = 0.1)) {
+  load(paste0("results_", i, ".RData")) 
+  plot_dat <- lapply(res, plot_data)
+  names(plot_dat) <- gsub("[.]", "_", r_center)
+  plot.dat[[i*10 + 1]] <- plot_dat
+}
 
-plot.dat <- lapply(res, plot_data)
-names(plot.dat) <- gsub("[.]", "_", r_center)
-
-do.call(rbind, plot.dat) %>%
-  mutate(name = rownames(do.call(rbind, plot.dat))) %>%
-  separate(name, into = c("multiplier", "num"), "[.]") %>% 
-  mutate(multiplier = as.numeric(gsub("_", ".", multiplier))) %>%
-  mutate(n_param_f = factor(n_param)) %>%
-  group_by(n_param_f, multiplier) %>%
-  summarise(frac_degen = sum(near_hull)/n()) -> ggvis.dat
 ```
@@
 
-ggvis.subset <- reactive({
-  do.call(rbind, plot.dat) %>%
-    mutate(name = rownames(do.call(rbind, plot.dat))) %>%
-    mutate(n_param_f = factor(n_param)) %>%
+ggvis.dat <- reactive({
+  if(is.null(input$power)) {
+    dat <- plot.dat[[1]]
+  } else
+    dat <- plot.dat[[input$power*10 + 1]]
+
+  do.call(rbind, dat) %>%
+    mutate(name = rownames(do.call(rbind, dat))) %>%
     separate(name, into = c("multiplier", "num"), "[.]") %>% 
     mutate(multiplier = as.numeric(gsub("_", ".", multiplier))) %>%
-    filter(multiplier == multiplier$current) 
+    mutate(n_param_f = factor(n_param)) %>%
+    group_by(n_param_f, multiplier) %>%
+    summarise(frac_degen = sum(near_hull)/n()) 
 })
@@
 ```{r plots2, message=FALSE, warning=FALSE, echo=FALSE, out.height=300}
+ggvis.subset <- reactive({
+  if(is.null(input$power)) {
+    dat <- plot.dat[[1]]
+  } else
+    dat <- plot.dat[[input$power*10 + 1]]
+
+  do.call(rbind, dat) %>%
+    mutate(name = rownames(do.call(rbind, dat))) %>%
+    mutate(n_param_f = factor(n_param)) %>%
+    separate(name, into = c("multiplier", "num"), "[.]") %>% 
+    mutate(multiplier = as.numeric(gsub("_", ".", multiplier))) %>%
+    filter(multiplier == multiplier$current) %>%
+    return()
+})
 
@@
 r_center <- seq(0.2, 3, by = 0.2)
-#r_center <- 1
 r_exp <- .05

### 2015-03-09 Toy Example/expected value/radius_percent_degen.R

@@ -57,16 +57,17 @@ n <- 100
 r_center <- seq(0.2, 3, by = 0.2)
 #r_center <- 1
 r_exp <- .05
+power <- seq(0, 1, by = .1)
 
 ## function for running simulation -------------------------------
-get_res <- function(n, r_center, r_exp) {
+get_res <- function(n, r_center, r_exp, power) {
   ## perform functions 
   test_cases <- expand.grid(data.frame(1:4)[,rep(1,2)]) %>% 
     rename(H = X1.4, V = X1.4.1) %>%
     filter(H <= V) %>% #remove those cases with less visibles than hiddens. Is this necessary?
     mutate(n_param = H*V + H + V) %>%
     mutate(max_facets = (2^(H+V))^(floor(n_param/2))) %>%
-    mutate(n = n, r = r_center*(H + V)) %>%
+    mutate(n = n, r = r_center*(H + V)^(power)) %>%
     group_by(H, V, n_param, n, r) %>%
     do(stat = stats(.$H, .$V, "negative")) %>%
     ungroup() %>%
@@ -78,7 +79,7 @@ get_res <- function(n, r_center, r_exp) {
     filter(H <= V) %>% #remove those cases with less visibles than hiddens. Is this necessary?
     mutate(n_param = H*V + H + V) %>%
     mutate(max_facets = (2^(H+V))^(floor(n_param/2))) %>%
-    mutate(n = n, r = r_center*(H + V)) %>%
+    mutate(n = n, r = r_center*(H + V)^(power)) %>%
     group_by(H, V, n_param, n, r) %>%
     do(stat = stats(.$H, .$V, "negative"))
   
@@ -98,9 +99,10 @@ get_res <- function(n, r_center, r_exp) {
 }
 
 ## run simulations -------------------------------------
-res <- lapply(r_center, function(r) get_res(n, r, r_exp))
-save(res, file = "results.RData")
-
+for(i in power) {
+  res <- lapply(r_center, function(r) get_res(n, r, r_exp, i))
+  save(res, file = paste0("results_", i, ".RData"))
+}
 
 ## analyze results -------------------------------------
 ## plot function ----------------------------------
@@ -129,15 +131,15 @@ plot_data <- function(res) {
 make_plot <- function(plot.data) { 
   plot.data %>%
     ggplot() + 
-    geom_boxplot(aes(as.factor(H + V), ss_ratio, colour = near_hull)) +
-    xlab("Number of Nodes") +
[…]

### 2015-09-24 Toy Example/4x4 Model/notes/mcmc_algorithm.Rmd
stuff is working

---
title: "Notes on 4x4 Model - MCMC Algorithm"
author: "Andee Kaplan"
date: "September 14, 2015"
output: 
  pdf_document:
    includes:
      in_header: header.tex
---

# Tikz test image

\begin{tikzpicture}[node distance = 2cm, auto]

  \tikzstyle{decision} = [diamond, draw, fill=blue!20, text width=4.5em, text badly centered, node distance=3cm, inner sep=0pt]
  \tikzstyle{block} = [rectangle, draw, fill=blue!20, text width=5em, text centered, rounded corners, minimum height=4em]
  \tikzstyle{line} = [draw, -latex']
  \tikzstyle{cloud} = [draw, ellipse, fill=red!20, node distance=3cm, minimum height=2em]

    % Place nodes
    \node [block] (init) {initialize model};
    \node [cloud, left of=init] (expert) {expert};
    \node [cloud, right of=init] (system) {system};
    \node [block, below of=init] (identify) {identify candidate models};
    \node [block, below of=identify] (evaluate) {evaluate candidate models};
    \node [block, left of=evaluate, node distance=3cm] (update) {update model};
    \node [decision, below of=evaluate] (decide) {is best candidate better?};
    \node [block, below of=decide, node distance=3cm] (stop) {stop};
    % Draw edges
    \path [line] (init) -- (identify);
    \path [line] (identify) -- (evaluate);
    \path [line] (evaluate) -- (decide);
    \path [line] (decide) -| node [near start] {yes} (update);
    \path [line] (update) |- (identify);
    \path [line] (decide) -- node {no}(stop);
    \path [line,dashed] (expert) -- (init);
    \path [line,dashed] (system) -- (init);
    \path [line,dashed] (system) |- (evaluate);
\end{tikzpicture}

### 2015-09-24 Toy Example/4x4 Model/choose_params.R

@@ -1,0 +1,69 @@
+#libraries -----------------------
+library(dplyr)
+library(tidyr)
+library(ggplot2)
+
+#function -----------------------
+#this function takes the results from the simulation study and rearranges for ease of use
+rearrange_data <- function(res) {
+  
+  data <- data.frame()
+  for(i in 1:nrow(res)) {
+    tmp <- res$samp[[i]]
+    
+    H <- res[i,]$H
+    V <- res[i,]$V
+    
+    if(H == 4 & V == 4) { #only looking at 4x4 model
+      tmp %>% 
+        data.frame() %>% 
+        rowwise() %>% 
+        mutate_(ss_ratio = paste0("(", paste(paste0(names(.)[(H + V + 1):(H + V + H*V)], "^2"), collapse = " + "), ")/sum(", paste(paste0(names(.)[1:(H+V)],"^2"), collapse = " + "), ")")) %>%    
+        ungroup() -> ratio
+      
+      data <- rbind(data, inner_join(ratio, res$outside[[i]] %>% ungroup()) %>%
+        mutate(H = H, V = V, n_param = H + V + H*V))
+    }
+  }
+  return(data)
+}
+
+#do it -------------------------
+#uses the function to rearrange the data
+n <- 100
+r_center <- seq(0.2, 3, by = 0.2)
+r_exp <- .05
+
+plot.dat <- list()
+for(i in seq(0, 1, by = 0.1)) {
+  load(paste0("../expected value/apps/results_", i, ".RData")) 
+  plot_dat <- lapply(res, rearrange_data)
+  names(plot_dat) <- gsub("[.]", "_", r_center)
+  
+  do.call(rbind, plot_dat) %>% 
+    mutate(name = rownames(do.call(rbind, plot_dat))) %>%
+    separate(name, into = c("multiplier", "num"), "[.]") %>% 
+    mutate(multiplier = as.numeric(gsub("_", ".", multiplier))) %>%
+    mutate(n_param_f = factor(n_param)) %>%
+    select(-starts_with("exp")) -> plot_dat
+  
+  plot.dat[[i*10 + 1]] <- plot_dat
+}
+
+names(plot.dat) <- gsub("[.]", "_", seq(0, 1, by = 0.1))
+do.call(rbind, plot.dat) %>%
+  mutate(name2 = rownames(do.call(rbind, plot.dat))) %>%
+  separate(name2, into = c("power", "num2"), "[.]") %>% 
[…]

### 2015-09-24 Toy Example/4x4 Model/choose_variance_params.R

@@ -1,0 +1,53 @@
+#libraries -----------------------
+library(dplyr)
+library(tidyr)
+library(ggplot2)
+
+#reshape data ---------------------
+load("../expected value/apps/results_split.RData")
+plot_data <- function(res) {
+  
+  plot.data <- data.frame()
+  for(i in 1:nrow(res)) {
+    tmp <- res$g_theta[[i]] %>% data.frame()
+    H <- res[i,]$H
+    V <- res[i,]$V
+    N <- res[i,]$N
+    r1 <- res[i,]$r1
+    r2 <- res[i,]$r2
+    C <- res[i,]$C
+    epsilon <- res[i,]$epsilon
+    
+    
+    tmp %>% 
+      rowwise() %>% 
+      mutate_(ss_ratio = paste0("(", paste(paste0(names(.)[(H + V + 1):(H + V + H*V)], "^2"), collapse = " + "), ")/sum(", paste(paste0(names(.)[1:(H+V)],"^2"), collapse = " + "), ")")) %>%    
+      ungroup() -> ratio
+    
+    
+    inner_join(ratio, res$outside[[i]] %>% ungroup()) %>%
+      select(ss_ratio, near_hull) %>%
+      mutate(H = H, V = V, n_param = H + V + H*V, N = H + V, N = N, r1 = r1, r2 = r2, C = C, epsilon = epsilon) %>%
+      rbind(plot.data) -> plot.data
+  }
+  return(plot.data)
+}
+plot_dat <- res %>% plot_data
+
+#choose C, C_prime --------------
+plot_dat %>%
+  filter(H == 4 & V == 4) %>%
+  group_by(H, V, C, epsilon) %>%
+  summarise(percent_degen = sum(near_hull)/n()) %>%
+  mutate(C_prime = C*((H + V)/(H*V))^(1+epsilon)) %>%
+  mutate(percent_degen = round(percent_degen, 2)) %>%
+  group_by(percent_degen) %>%
+  mutate(C_C_prime = C + C_prime) %>%
+  top_n(1, C_C_prime) %>%
+  arrange(percent_degen, C_prime, C) %>%
+  ungroup() %>%
+  top_n(1, C_C_prime) %>%
+  select(percent_degen, C, C_prime) %>%
+  data.frame -> variance_params
+
+save(variance_params, file = "written/variance_params.Rdata")

### 2015-09-24 Toy Example/4x4 Model/fit_gibbs.R

@@ -1,0 +1,45 @@
+#libraries --------------------
+library(dplyr)
+library(tidyr)
+library(ggplot2)
+source("sample_functs.R")
+
+#data and params ------------------------
+H <- 4
+V <- 4
+mc.iter <- 5000
+set.seed(102285) #reproducible seed
+
+load("written/sample_images.Rdata")
+load("written/variance_params.Rdata")
+load("written/sample.params.Rdata")
+
+params <- list(main_hidden = rnorm(H, mean = 0, sd = sqrt(variance_params$C)),
+               main_visible = rnorm(V, mean = 0, sd = sqrt(variance_params$C)),
+               interaction = matrix(rnorm(H*V, mean = 0, sd = sqrt(variance_params$C_prime)), nrow = H))
+
+N <- nrow(flat_images_good$visibles)
+const <- 1
+models_good <- sample_gibbs(visibles = flat_images_good$visibles, params0 = params, C = const*variance_params$C/N, C_prime = const*variance_params$C_prime/N, mc.iter = mc.iter)
+models_bad <- sample_gibbs(visibles = flat_images_good$visibles, params0 = params, C = const*variance_params$C/N, C_prime = const*variance_params$C/N, mc.iter = mc.iter)
+models_degen <- sample_gibbs(visibles = flat_images_degen$visibles, params0 = params, C = const*variance_params$C/N, C_prime = const*variance_params$C_prime/N, mc.iter = mc.iter)
+
+#burnin + thinning
+burnin_thinning <- function(models) {
+  idx <- which(1:mc.iter %% 5 == 0 & 1:mc.iter > mc.iter/2)
+  models$params$main_hidden <- models$params$main_hidden[idx, ]
+  models$params$main_visible <- models$params$main_visible[idx, ]
+  models$params$interaction <- models$params$interaction[idx, , ]
+  models$hiddens <- models$hiddens[idx, , ]
+  #models$post_pred <- models$post_pred[idx, , ]
+  models$distn <- models$distn[idx, , ]
+  
+  return(models)
+}
+
+models_good <- burnin_thinning(models_good)
+models_bad <- burnin_thinning(models_bad)
+models_degen <- burnin_thinning(models_degen)
+
+save(models_good, models_bad, models_degen, file = paste0("written/fitted_models_distn_", const, ".Rdata"))
+

### 2015-09-24 Toy Example/4x4 Model/generate_images.R

@@ -1,31 +1,33 @@
 #libraries --------------------
 library(dplyr)
 library(tidyr)
-source("sample_functs_4x4.R")
+source("sample_functs.R")
 
 #data and params ------------------------
 H <- 4
 V <- 4
 N <- 1
+mc.iter <- 10e4
+set.seed(102285)
 
 load("written/sample.params.Rdata")
 params_degen <- list(main_hidden = sample.params %>% ungroup() %>% filter(near_hull) %>% select(starts_with("h"), -H) %>% data.matrix(),
                      main_visible = sample.params %>% ungroup() %>% filter(near_hull) %>% select(starts_with("v"), -V) %>% data.matrix(),
-                     interaction = sample.params %>% ungroup() %>% filter(near_hull) %>% select(starts_with("theta")) %>% data.matrix() %>% matrix(4, byrow = TRUE))
+                     interaction = sample.params %>% ungroup() %>% filter(near_hull) %>% select(starts_with("theta")) %>% data.matrix() %>% matrix(4))
 params_good <- list(main_hidden = sample.params %>% ungroup() %>% filter(!near_hull) %>% select(starts_with("h"), -H) %>% data.matrix(),
                     main_visible = sample.params %>% ungroup() %>% filter(!near_hull) %>% select(starts_with("v"), -V) %>% data.matrix(),
-                    interaction = sample.params %>% ungroup() %>% filter(!near_hull) %>% select(starts_with("theta")) %>% data.matrix() %>% matrix(4, byrow = TRUE))
+                    interaction = sample.params %>% ungroup() %>% filter(!near_hull) %>% select(starts_with("theta")) %>% data.matrix() %>% matrix(4))
 
 hidden0 <- matrix(sample(c(-1,1), H*N, replace = TRUE), nrow = N)
 visible0 <- matrix(sample(c(-1,1), V*N, replace = TRUE), nrow = N)
 
-flat_images_degen <- sample_images(params = params_degen, hidden0 = hidden0, visible0 = visible0, mc.iter = 5e4)
-flat_images_good <- sample_images(params = params_good, hidden0 = hidden0, visible0 = visible0, mc.iter = 5e4)
+flat_images_degen <- sample_images(params = params_degen, hidden0 = hidden0, visible0 = visible0, mc.iter = mc.iter)
[…]

### 2015-09-24 Toy Example/4x4 Model/ml_functs.R

@@ -1,0 +1,114 @@
+#libraries --------------------
+library(dplyr)
+library(tidyr)
+
+loglik <- function(theta, parms) {
+  #theta is a H + V + H*V length vector containing main_visible, then main_hidden, then interactions
+  
+  #parms is a named list containing data and H and V
+  visibles <- parms$visibles
+  hiddens <- parms$hiddens
+  H <- parms$H
+  V <- parms$V
+  
+  #things for derivative, build data and possible values
+  N <- visibles %>% nrow()
+  data <- cbind(visibles, hiddens)
+  for(i in 1:V) {
+    for(j in 1:H) {
+      data <- cbind(data, visibles[i,]*hiddens[j,])
+    }
+  }
+  possibles <- stats(H, V)
+  
+  W <- exp(possibles %*% theta)
+  sum(theta*colSums(data)/N) - log(sum(W)) 
+}
+
+loglik_derivs <- function(theta, parms) {
+  #theta is a H + V + H*V length vector containing main_visible, then main_hidden, then interactions
+  
+  #parms is a named list containing data and H and V
+  visibles <- parms$visibles
+  hiddens <- parms$hiddens
+  H <- parms$H
+  V <- parms$V
+  
+  #things for derivative, build data and possible values
+  N <- visibles %>% nrow()
+  data <- cbind(visibles, hiddens)
+  for(i in 1:V) {
+    for(j in 1:H) {
+      data <- cbind(data, visibles[i,]*hiddens[j,])
+    }
+  }
+  possibles <- stats(H, V)
+  
+  
+  #vector of derivatives set to zero
+  W <- exp(possibles %*% theta)
+  colSums(data)/N - colSums(apply(possibles, 2, function(x) { W*x }))/sum(W)
+}
+
+loglik_hessian <- function(theta, parms) {
+  #theta is a H + V + H*V length vector containing main_visible, then main_hidden, then interactions
+  
+  #parms is a named list containing data and H and V
+  visibles <- parms$visibles
+  hiddens <- parms$hiddens
+  H <- parms$H
+  V <- parms$V
+  
+  #things for derivative, build data and possible values
+  N <- visibles %>% nrow()
+  data <- cbind(visibles, hiddens)
+  for(i in 1:V) {
+    for(j in 1:H) {
[…]

### 2015-09-24 Toy Example/4x4 Model/sample_functs.R

@@ -1,0 +1,165 @@
+sample_main_hidden <- function(hiddens, C) {
+  #hiddens is an NxH matrix of hidden node values
+  #C is the variance constant
+  res <- sapply(colSums(hiddens), function(mean) rnorm(1, mean*C, sqrt(C)))
+  names(res) <- paste0("H", 1:ncol(hiddens))
+  return(res)
+}
+
+sample_main_visible <- function(visibles, C) {
+  #visibles is an NxV matrix of visible node values
+  #C is the variance constant
+  res <- sapply(colSums(visibles), function(mean) rnorm(1, mean*C, sqrt(C)))
+  names(res) <- paste0("V", 1:ncol(visibles))
+  return(res)
+}
+
+sample_interaction <- function(hiddens, visibles, C) {
+  #hiddens is an NxH matrix of hidden node values
+  #visibles is an NxV matrix of visible node values
+  #C is the variance constant
+  H <- ncol(hiddens)
+  V <- ncol(visibles)
+  
+  stopifnot(nrow(hiddens) == nrow(visibles)) #something wrong with the data structure if these don't match
+  
+  res <- matrix(NA, nrow = H, ncol = V)
+  for(i in 1:V) {
+    for(j in 1:H) {
+      res[i, j] <- rnorm(1, sum(hiddens[, j]*visibles[, i])*C, sqrt(C))
+    }
+  }
+  
+  rownames(res) <- paste0("V", 1:V)
+  colnames(res) <- paste0("H", 1:H)
+  
+  return(res)
+}
+
+sample_hidden <- function(main_hidden, interaction, visibles) {
+  #main_hidden is a 1xH matrix of hidden main effects param values
+  #interaction is a HxV matrix of interaction effects param values
+  #visibles is an NxV matrix of visible node values
+  a <- exp(2*(matrix(main_hidden, nrow = nrow(visibles), ncol = length(main_hidden), byrow = TRUE) + crossprod(t(visibles), interaction)))
+  res <- apply(a, 1:2, function(b) rbinom(1, 1, b/(1 + b))*2 - 1)
+  
+  #a <- exp(2*(matrix(main_hidden, nrow = nrow(visibles), ncol = length(main_hidden), byrow = TRUE) + matrix(rowSums(interaction), nrow = nrow(visibles), ncol = nrow(interaction), byrow = TRUE)*visibles))
[…]

### 2015-09-24 Toy Example/4x4 Model/test_all_params.R

@@ -2,6 +2,7 @@
 library(dplyr)
 library(tidyr)
 library(ggplot2)
+source("sample_functs.R")
 
 #function -----------------------
 #this function takes the results from the simulation study and rearranges for ease of use
@@ -22,7 +23,7 @@ rearrange_data <- function(res) {
         ungroup() -> ratio
       
       data <- rbind(data, inner_join(ratio, res$outside[[i]] %>% ungroup()) %>%
-        mutate(H = H, V = V, n_param = H + V + H*V))
+                      mutate(H = H, V = V, n_param = H + V + H*V))
     }
   }
   return(data)
@@ -56,14 +57,45 @@ do.call(rbind, plot.dat) %>%
   separate(name2, into = c("power", "num2"), "[.]") %>% 
   mutate(power = as.numeric(gsub("_", ".", power))) -> plot.dat
 
-#sample --------------------------------
-#samples the parameters to get one set from the "degenerate" and one from "non-degenerate"
-
-set.seed(1000) #reproducible
-
+#plot --------------------------------
 plot.dat %>%
-  group_by(near_hull) %>%
-  sample_n(1) -> sample.params
+  group_by(h1, h2, h3, h4, v1, v2, v3, v4, theta11, theta12, theta13, theta14, theta21, theta22, theta23, theta24, theta31, theta32, theta33, theta34, theta41, theta42, theta43, theta44, near_hull) %>%
+  do(params = list(main_hidden = data.frame(.) %>% select(starts_with("h"), -H) %>% data.matrix(),
+                   main_visible = data.frame(.) %>% select(starts_with("v"), -V) %>% data.matrix(),
+                   interaction = data.frame(.) %>% select(starts_with("theta")) %>% data.matrix() %>% matrix(4))) %>%
+  ungroup() %>%
+  group_by(h1, h2, h3, h4, v1, v2, v3, v4, theta11, theta12, theta13, theta14, theta21, theta22, theta23, theta24, theta31, theta32, theta33, theta34, theta41, theta42, theta43, theta44, near_hull) %>%
+  do(prob = visible_distn(params = .$params[[1]])) %>%
+  rowwise() %>%
+  do(probs = data.frame(near_hull = .$near_hull, .$prob)) -> probs
 
-save(sample.params, file = "written/sample.params.Rdata")
[…]

### 2015-09-24 Toy Example/4x4 Model/test_fitted_model.R

@@ -1,0 +1,111 @@
+#libraries --------------------
+library(dplyr)
+library(tidyr)
+library(ggplot2)
+source("sample_functs.R")
+
+#data and params ------------------------
+H <- 4
+V <- 4
+N <- 1
+
+load("written/sample_images.Rdata")
+load("written/fitted_models_distn_1.Rdata")
+load("written/sample.params.Rdata")
+params_degen <- list(main_hidden = sample.params %>% ungroup() %>% filter(near_hull) %>% select(starts_with("h"), -H) %>% data.matrix(),
+                     main_visible = sample.params %>% ungroup() %>% filter(near_hull) %>% select(starts_with("v"), -V) %>% data.matrix(),
+                     interaction = sample.params %>% ungroup() %>% filter(near_hull) %>% select(starts_with("theta")) %>% data.matrix() %>% matrix(4))
+params_good <- list(main_hidden = sample.params %>% ungroup() %>% filter(!near_hull) %>% select(starts_with("h"), -H) %>% data.matrix(),
+                    main_visible = sample.params %>% ungroup() %>% filter(!near_hull) %>% select(starts_with("v"), -V) %>% data.matrix(),
+                    interaction = sample.params %>% ungroup() %>% filter(!near_hull) %>% select(starts_with("theta")) %>% data.matrix() %>% matrix(4))
+
+
+##plots -----------------------
+
+#computing actual distributions
+distn_good <- visible_distn(params = params_good)
+distn_degen <- visible_distn(params = params_degen)
+
+get_props <- function(model) {
+  props <- apply(model$post_pred, 1, function(x) {
+    x %>% 
+      data.frame %>%
+      rename(v1 = X1, v2 = X2, v3 = X3, v4 = X4) %>%
+      right_join(distn_good) %>% 
+      group_by(image_id) %>%
+      summarise(prop = n()/nrow(x))
+  })
+  
+  props %>% 
+    bind_rows(.id = "iteration") %>% 
+    mutate(image_id = as.numeric(image_id), iteration = as.numeric(iteration)) %>%
+    data.frame
+}
+
+props_good <- get_props(models_good)
+props_bad <- get_props(models_bad)
+props_degen <- get_props(models_degen)
+
+data_props_good <- flat_images_good$visibles %>% 
[…]

### 2015-09-24 Toy Example/4x4 Model/test_ml.R

@@ -1,0 +1,99 @@
+#libraries --------------------
+library(dplyr)
+library(tidyr)
+library(ggplot2)
+library(gridExtra)
+source("ml_functs.R")
+
+
+#data and params ------------------------
+H <- 4
+V <- 4
+load("written/fitted_ml.Rdata")
+load("written/sample_images.Rdata")
+load("written/sample.params.Rdata")
+
+parms <- list(visibles = flat_images_good$visibles, 
+              hiddens = flat_images_good$hiddens,
+              H = H, V = V)
+
+theta_good <- sample.params %>% filter(!near_hull) %>% ungroup() %>% select(starts_with("v"), starts_with("h"), starts_with("theta"), -H, -V) %>% data.matrix
+theta_ml <- cbind(theta, likelihood) %>% data.frame()
+names(theta_ml) <- c(colnames(theta_good), "likelihood")
+
+#plots--------------------------
+theta_ml %>%
+  mutate(iter = 1:nrow(theta)) %>%
+  gather(variable, value, -iter) %>%
+  ggplot() +
+  geom_point(aes(x = iter, y = value)) +
+  facet_wrap(~variable, scales = "free_y") 
+
+plot_two <- function(idx1, idx2, theta_true, theta_ml, parms) {
+
+  expand.grid(seq(min(theta_ml[, idx1]), max(theta_ml[, idx1]), length.out = 30), 
+              seq(min(theta_ml[, idx2]), max(theta_ml[, idx2]), length.out = 30)) %>% 
+    data.frame() -> grid
+  
+  names(grid) <- colnames(theta_true)[c(idx1, idx2)]
+  
+  matrix(theta_true, nrow = nrow(grid), ncol = length(theta_true), byrow = TRUE) %>% 
+    data.frame() -> theta_true_expand
+  names(theta_true_expand) <- colnames(theta_true)
+  
+  theta_true_expand[, colnames(theta_true)[c(idx1, idx2)]] <- grid
+  theta_true_expand$likelihood <- apply(theta_true_expand, 1, function(x) loglik(theta = x, parms = parms))
+  
+  theta_true_expand %>%
+    ggplot(aes_string(x = colnames(theta_true)[idx1], y = colnames(theta_true)[idx2], z = "likelihood")) +
[…]

### 2015-09-24 Toy Example/expected value/split_sample.R

@@ -178,6 +178,13 @@ data.frame(expand.grid(H = seq(1, 11, by = 2), V = seq(1, 11, by = 2))) %>%
                            p = p)) -> explore
 
 
+do.call(rbind, explore$plotting) %>% 
+  ggplot(aes(x = C, y = C_prime, z = pred_trans)) +
+  stat_contour(aes(colour = ..level..), size = 0.5) +
+  stat_contour(breaks = c(p), size = 2, colour = "black") +
+  facet_grid(H ~ V) +
+  theme_bw()
+
 data.frame(expand.grid(H = 4, V = 4)) %>%
   group_by(H, V) %>%
   do(finding = find_C_C_prime(p, .$H, .$V, Cs = seq(0, 3, by = .1), C_primes =  seq(0, 3, by = .1), degen.m2)) %>%

### 2015-09-24 Toy Example/gibbs sampler/sample_functs_4x4.R

@@ -1,121 +1,0 @@
-sample_main_hidden <- function(hiddens, C) {
-  #hiddens is an NxH matrix of hidden node values
-  #C is the variance constant
-  res <- sapply(colSums(hiddens), function(mean) rnorm(1, mean/(2*C), 1/(2*C^2)))
-  names(res) <- paste0("H", 1:ncol(hiddens))
-  return(res)
-}
-
-sample_main_visible <- function(visibles, C) {
-  #visibles is an NxV matrix of visible node values
-  #C is the variance constant
-  res <- sapply(colSums(visibles), function(mean) rnorm(1, mean/(2*C), 1/(2*C^2)))
-  names(res) <- paste0("V", 1:ncol(visibles))
-  return(res)
-}
-
-sample_interaction <- function(hiddens, visibles, C) {
-  #hiddens is an NxH matrix of hidden node values
-  #visibles is an NxV matrix of visible node values
-  #C is the variance constant
-  H <- ncol(hiddens)
-  V <- ncol(visibles)
-  
-  stopifnot(nrow(hiddens) == nrow(visibles)) #something wrong with the data structure if these don't match
-  res <- matrix(NA, nrow = H, ncol = V)
-  for(i in 1:H) {
-    for(j in 1:V) {
-      res[i, j] <- rnorm(1, sum(hiddens[, i]*hiddens[, j])/(2*C), 1/(2*C^2))
-    }
-  }
-  
-  colnames(res) <- paste0("V", 1:V)
-  rownames(res) <- paste0("H", 1:H)
-  
-  return(res)
-}
-
-sample_hidden <- function(main_hidden, interaction, visibles) {
-  #main_hidden is a 1xH matrix of hidden main effects param values
-  #interaction is a HxV matrix of interaction effects param values
-  #visibles is an NxV matrix of visible node values
-  
-  a <- exp(2*(sum(main_hidden) + rowSums(t(colSums(interaction) * t(visibles)))))
-  res <- t(sapply(a, function(b) { rbinom(length(main_hidden), 1, b/(b + 1))*2 - 1 })) #make 0,1 -> -1,1
-  colnames(res) <- paste0("H", 1:H)
-  return(res)
-}
-
-sample_visible <- function(main_visible, interaction, hiddens) {
-  #main_visible is a 1xH matrix of visible main effects param values
-  #interaction is a HxV matrix of interaction effects param values
[…]

### 2015-09-24 Toy Example/gibbs sampler/test_sample_images_4x4.R

@@ -1,62 +1,0 @@
-#libraries --------------------
-library(dplyr)
-library(tidyr)
-library(ggplot2)
-source("sample_functs_4x4.R")
-
-#data and params ------------------------
-H <- 4
-V <- 4
-N <- 1
-
-load("written/sample.params.Rdata")
-params_degen <- list(main_hidden = sample.params %>% ungroup() %>% filter(near_hull) %>% select(starts_with("h"), -H) %>% data.matrix(),
-                     main_visible = sample.params %>% ungroup() %>% filter(near_hull) %>% select(starts_with("v"), -V) %>% data.matrix(),
-                     interaction = sample.params %>% ungroup() %>% filter(near_hull) %>% select(starts_with("theta")) %>% data.matrix() %>% matrix(4, byrow = TRUE))
-params_good <- list(main_hidden = sample.params %>% ungroup() %>% filter(!near_hull) %>% select(starts_with("h"), -H) %>% data.matrix(),
-                    main_visible = sample.params %>% ungroup() %>% filter(!near_hull) %>% select(starts_with("v"), -V) %>% data.matrix(),
-                    interaction = sample.params %>% ungroup() %>% filter(!near_hull) %>% select(starts_with("theta")) %>% data.matrix() %>% matrix(4, byrow = TRUE))
-
-
-cdf_visible <- function(main_visible, interaction, hiddens, visibles) {
-  #main_visible is a 1xH matrix of visible main effects param values
-  #interaction is a HxV matrix of interaction effects param values
-  #hiddens is an NxH matrix of hidden node values
-  stopifnot(nrow(hiddens) == nrow(visibles))
-  
-  N <- nrow(hiddens)
-  res <- matrix(NA, nrow = N, ncol = ncol(visibles))
-  
-  
-  for(i in 1:N) {
-    a <- exp(2*(sum(main_visible) + rowSums(t(colSums(interaction) * t(matrix(hiddens[i, ], nrow = 1))))))
-    A <- runif(1)
-    f <- dbinom(0, 1, a/(a + 1))
-    res[i, ] <- A*f*(visibles[i, ] == -1) + (f + A*(1 - f))*(visibles[i, ] == 1)
-  }
-
-  return(res)
-}
-
-pit_test_degen <- cdf_visible(params_degen$main_visible, params_degen$interaction, flat_images_degen$hiddens, flat_images_degen$visibles)
[…]

### 2015-11-18 4x4 model/docs/ok_model.Rmd
reworking structure

---
title: "OK Restricted Boltzmann Machine"
author: "Andee Kaplan"
date: "October 28, 2015"
output: pdf_document
---

```{r, message=FALSE}
library(dplyr)
library(tidyr)
library(knitr)
source("../sample_functs.R")

load("../written/sample.params.Rdata")
load("../written/sample_images.Rdata")

sample.params %>%
  ungroup() %>%
  filter(!near_hull) %>%
  select(starts_with("v"), starts_with("h"), starts_with("theta"), -H, -V) %>%
  gather(parameter, value, everything()) %>%
  kable()

params_good <- list(main_hidden = sample.params %>% ungroup() %>% filter(!near_hull) %>% select(starts_with("h"), -H) %>% data.matrix(),
                    main_visible = sample.params %>% ungroup() %>% filter(!near_hull) %>% select(starts_with("v"), -V) %>% data.matrix(),
                    interaction = sample.params %>% ungroup() %>% filter(!near_hull) %>% select(starts_with("theta")) %>% data.matrix() %>% matrix(4))

visible_distn(params = params_good) %>%
  kable()

possibles <- stats(4, 4)
theta <- sample.params %>%
  ungroup() %>%
  filter(!near_hull) %>%
  select(starts_with("v"), starts_with("h"), starts_with("theta"), -H, -V) %>%
  data.matrix() %>% t

sum(exp(possibles %*% theta))

```

### 2015-11-18 degeneracy/apps/degen_magnitude.Rmd
reworking structure

--- 2015-03-09 Toy Example/expected value/apps/degen_magnitude.Rmd
@@
 library(ggplot2)
-library(plyr)
 library(dplyr)
@@
 sliderInput("power", "Power of N:", min = 0, max = 1, value = 0, step = 0.1)
+checkboxInput("split", "Split interaction power")
 
@@
 }
+
 plot.dat <- list()
@@
   plot.dat[[i*10 + 1]] <- plot_dat
+}
+
+plot.dat_split <- list()
+for(i in seq(0, 1, by = 0.1)) { #full data not compiled yet
+  load(paste0("results_split_", i, ".RData")) 
+  res <- lapply(res, function(x) x %>% mutate(r = r2)) #dumb naming convention
+  plot_dat_split <- lapply(res, plot_data)
+  names(plot_dat_split) <- gsub("[.]", "_", r_center)
+  plot.dat_split[[i*10 + 1]] <- plot_dat_split
 }
@@
 ggvis.dat <- reactive({
+
   if(is.null(input$power)) {
     dat <- plot.dat[[1]]
-  } else
+  } else {
+    if(input$split)
+      dat <- plot.dat_split[[input$power*10 + 1]]
+    else 
       dat <- plot.dat[[input$power*10 + 1]]
+  }
 
@@
     dat <- plot.dat[[1]]
-  } else
+  } else {
+    if(input$split)
+      dat <- plot.dat_split[[input$power*10 + 1]]
+    else 
       dat <- plot.dat[[input$power*10 + 1]]
+  }
 
@@
     mutate(multiplier = as.numeric(gsub("_", ".", multiplier))) %>%
-    filter(multiplier == multiplier$current) %>%
-    return()
+    filter(multiplier == multiplier$current) 
 })

### 2015-11-18 degeneracy/apps/mu_parameterization_3d.Rmd
reworking structure

---
author: "Andee Kaplan"
date: "11/18/2014"
output:
  html_document:
    theme: readable
runtime: shiny
---

# $\mu$ parameterization in 3 dimensions

For one hidden and one visible node, we sample points uniformly on a 3 dimensional sphere of radius `r` centered at `(0,0,0)` and view the resulting expected values in 3 dimensions.
```{r, echo=FALSE, message=FALSE}
source("functions.R")
library(scatterplot3d)
library(plyr)
library(dplyr)
library(reshape2)
library(ggvis)

## Num nodes
H <- 1
V <- 1

## Num to sample
n <- 500

## Radius to sample
r <- c(seq(0,1, by=0.05), seq(1.5, 10, by=0.5))

```

```{r, echo=FALSE, message=FALSE}

shinyApp(
  ui=
    fluidPage(
      tags$head(
        includeCSS("www/readable.min.css"),
        includeCSS("www/css/numeric_inputs.css")
      ),
      responsive = FALSE,
      fluidRow(style = "padding-bottom: 20px;",
        column(4, radioButtons("encoding", label="Encoding:", 
                                      choices = list("Negative (-1,1)" = "negative", "Binary (0,1)" = "binary"),
                                      inline = FALSE)),
        column(4, sliderInput("rSlider", label = "Radius:", min = 0, max = 10, value=0, step = 0.1, animate = list(interval=1000, loop=FALSE, playButton=icon("play"), pauseButton=icon("pause")), )),
        column(4, 
               column(6, span("Translate origin by (x, y, z): ")),
               column(2, numericInput("x", value=NA, label="")),
               column(2, numericInput("y", value=NA, label="")), 
               column(2, numericInput("z", value=NA, label=""))
               )
      ),
      fluidRow(
        column(6, plotOutput("exp_val_3d")),
        column(6, ggvisOutput("min_dist_hist"))
      )
    ),
  server=function(input, output, session){

    data <- reactive({
      x <- ifelse(is.na(input$x), 0, input$x)
      y <- ifelse(is.na(input$y), 0, input$y)
      z <- ifelse(is.na(input$z), 0, input$z)

      stat <- stats(H, V, input$encoding)
      theta <- sample_sphere(space = ncol(stat), n, input$rSlider)  + matrix(rep(c(x,y,z), 10), nrow = 3)
      exp_vals <- expected_value(stats = stat, theta = theta, normalized = TRUE)

      return(list(stat=stat, exp_vals=exp_vals))
    })

#     cdf_mu <- reactive({
#       dat <- data()
#       stat <- dat$stat
#       exp_vals <- dat$exp_vals
#       
#       cdf_mu <- melt(apply(apply(t(exp_vals), 2, abs), 2, sort)) %>%
#         group_by(Var2)
#       cdf_mu$Var1 <- cdf_mu$Var1/ncol(exp_vals)
#       
#       return(cdf_mu)
#     })

    min_dist_data <- reactive({
      dat <- data()
      stat <- dat$stat
      exp_vals <- dat$exp_vals

      lines <- generate_hull_from_stats(stat)
      dist <- apply(exp_vals, 2, function(x){ min_distance_to_hull(x, lines) })
      return(data.frame(dist = dist))      
    })

    output$exp_val_3d <- renderPlot({
      dat <- data()
      stat <- dat$stat
      exp_vals <- dat$exp_vals

      s <- scatterplot3d(stat, pch='.')

      segs <- NULL
      for(i in 1:nrow(stat)) {
        for(j in 1:nrow(stat)) {
          if(i < j) {
            start <- s$xyz.convert(stat[i,])
            end <- s$xyz.convert(stat[j,])
            segs <- rbind(segs, c(start$x, start$y, end$x, end$y))
          }
        }
      }

      segments(segs[,1], segs[,2], segs[,3], segs[,4], lty=2)
      points(s$xyz.convert(t(exp_vals)), col="blue", pch=19)

    })

#     cdf_mu %>%
#       ggvis(~value, ~Var1, stroke = ~Var2) %>%
#       layer_paths() %>%
#       scale_numeric("x", domain = c(0,1)) %>%
#       scale_numeric("y", domain = c(0,1)) %>%
#       add_axis("x", title = "Absolute Expected Value") %>%
#       add_axis("y", title = "Empirical CDF") %>%
#       add_legend("stroke", title = "Mu Parameterized") %>%
#       set_options(width=200, height=200, keep_aspect = TRUE, resizable = FALSE) %>%
#       bind_shiny(plot_id = "exp_val_ecdf")

    min_dist_data %>%
      ggvis(~dist) %>%
      layer_histograms() %>%
      scale_numeric("x", domain = c(0,1)) %>%
      add_axis("x", title = "Minimum distance to convex hull edge") %>%
      set_options(width=200, height=200, keep_aspect = TRUE, resizable = FALSE) %>%
      bind_shiny(plot_id = "min_dist_hist")
  },
  options = list(height = 800)
)

```

### 2016-01-14 writing/degeneracy/draft.Rmd
started writing

---
title: "Degeneracy in a restricted Boltzmann machine"
author: "Andee Kaplan, Stephen Vardeman, Daniel Nordman"
date: "January 8, 2016"
output: 
  pdf_document:
    fig_caption: yes
    includes:
      in_header: header.tex
    number_sections: yes
bibliography: ../references/refs.bib
---

```{r libraries_options, echo=FALSE, message=FALSE, warnings=FALSE}
library(knitr)
library(ggplot2)
library(dplyr)
library(tidyr)

opts_chunk$set(echo=FALSE, message=FALSE, warnings=FALSE)
theme_set(theme_bw(base_family="serif"))

source("../helpers/functions.R")
```

#Restricted Boltzmann Machines

A restricted Boltzmann machine (RBM) is a specific undirected graphical model specified for discrete random variables, binary being the most common. The architecture of an RBM is specified as having two layers, a hidden ($\mathcal{H}$) and a visible layer ($\mathcal{V}$), with no connections within a layer. An example of this structure is visualized in figure \ref{fig:rbm} with the hidden nodes indicated by gray circles and the visible nodes indicated by white circles.

\begin{figure}[ht]
  \centering
  \resizebox{\linewidth}{!}{\input{tikz/rbm.tikz}}
  \caption{An example restricted Boltzmann machine (RBM), which consists of two layers, a hidden ($\mathcal{H}$) and a visible layer ($\mathcal{V}$), with no connections within a layer. Hidden nodes indicated by gray circles and the visible nodes indicated by white circles.}
  \label{fig:rbm}
\end{figure}

A common use for the RBM is to create features for use in classification. For example, a binary image could be classified where the pixel values would be considered the visible variables $v_i$.

##Joint Distribution
Let ${\bf x} = \{h_1, \dots, h_H, v_1,\dots,v_V\}$ represent the states of the visible and hidden nodes in an RBM. Then we posit a parametric form for probabilities corresponding to the state of each node taking the value of $1$.

\begin{align}
\label{eqn:pmf}
f_{\boldsymbol \theta} ({\bf x}) = \frac{\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\sum\limits_{{\bf x} \in \{0,1\}^{H+V}}\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)} 
\end{align}

Let the normalizing constant in equation \ref{eqn:pmf} be denoted $\gamma(\boldsymbol \theta)$ [@Vardeman602, pp.163-167]. Additionally, we can refer to $Q({\bf x}) = \sum\limits_{i = 1}^H\sum\limits_{j=1}^V \theta_{ij} h_i v_j + \sum\limits_{i = 1}^H\theta_{hi} h_i + \sum\limits_{j = 1}^V\theta_{vj} v_j$ as the neg-potential function in our model. The RBM model is parameterized by $\boldsymbol \theta$, which contains two types of parameters, main effects and interaction effects. The main effects parameters ($\theta_{v_i}, \theta_{h_j}$) weight to the values of the visible and hidden nodes and the interaction effects parameters ($\theta_{ij}$) weight the values of the connections, or dependencies, between hidden and visible layers.

Due to the potential size of the model the normalizing constant $\gamma(\boldsymbol \theta)$ can be infeasible to calculate, making it impossible to estimate the model exactly. An approximation can be achieved using Gibbs sampling, which is a natural fit due to the restricted architecture of the RBM. Specifically, the conditional independence of each layer given the other allows for simulation of the main and hidden layers.

##Connection to Deep Learning

#Degeneracy in Random Graph Models
In Random Graph Model theory, \emph{degeneracy} occurs when there is a disproportionate amount of probability placed on only a few elements of the sample space, $\mathcal{X}$, by the model. @handcock2003assessing define a model to be near degenerate if $\mu(\boldsymbol \theta)$, the mean parametrization on the model parameters, is close to the boundary of the convex hull of $\{t({\bf x}): {\bf x} \in \mathcal{X}\}$, where $t({\bf x})$ is the set of statistics in the neg-potential function $Q({\bf x})$ from equation \ref{eqn:pmf}.

The mean parameterization on the model parameters, $\mu(\boldsymbol \theta)$, is defined as
\begin{align*}
\mu(\boldsymbol \theta) &= \text{E}_{\boldsymbol \theta}\left[ t(\boldsymbol X) \right] \\
&= \sum\limits_{x \in \{-1,1\}^{H + V}} \left\{ t(\boldsymbol x) f_{\boldsymbol \theta} ({\bf x}) \right\} \\
&= \sum\limits_{x \in \{-1,1\}^{H + V}} \left\{ t(\boldsymbol x)\dfrac{\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\sum\limits_{{\bf x} \in \{0,1\}^{H+V}}\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}\right\} \\
\end{align*}

##Small example
To illustrate the idea of graph degeneracy in a RBM, we will consider a the smallest possible toy example that consists of one visible node $v_1$ and one hidden node $h_1$, both binary. A schematic of this model can be found in figure \ref{fig:toymodel}.

\begin{figure}[ht]
  \centering
  \resizebox{1cm}{!}{\input{tikz/toymodel.tikz}}
  \caption{A small example restricted Boltzmann machine (RBM), which consists of two nodes, one hidden and one visible.}
  \label{fig:toymodel}
\end{figure}

For this small model, we are able to illustrate the boundary of the convex hull of $\{t({\bf x}): {\bf x} \in \mathcal{X}\}$ because we are in three dimensions. In figure \ref{fig:toyhull}, the convex hull of our statistic space in three dimensions for the toy RBM with one visible and one hidden node from multiple perspectives enclosed by an unrestricted hull of $\{0,1\}^3$ space is shown. In this small model, $\{t({\bf x})\}$ is defined as $\{t({\bf x})\} = \{v_1, h_1, v_1 h_1\}$ by definition and form of the neg-potential function $Q({\bf x})$. Note that the convex hull of $\{t({\bf x})\}$ does not fill the unrestricted hull because of the relationship between the elements of $\{t({\bf x})\}$.

```{r models}
m1_bin <- calc_hull(1, 1)
m1_neg <- calc_hull(1, 1, type = "negative")
```

\begin{figure}[ht]
  \begin{minipage}{0.45\textwidth}
  \resizebox{\linewidth}{!}{
    \tdplotsetmaincoords{60}{-60}
    \input{tikz/toyhull_top.tikz}

    ```{r, results='asis', echo=FALSE, include=TRUE, message=FALSE}
    cat(sapply(1:nrow(m1_bin$c_hull$hull), function(x) paste0("\\draw", paste(paste0("(", apply(m1_bin$possible_t[cbind(m1_bin$c_hull$hull, m1_bin$c_hull$hull[,1])[x,],], 1, paste, collapse=","), ")"), collapse=" -- "), "; \n")))
   ```
    \end{tikzpicture}
  }
  \end{minipage}
  \begin{minipage}{0.45\textwidth}
  \resizebox{\linewidth}{!}{
    \tdplotsetmaincoords{60}{120}
    \input{tikz/toyhull_top.tikz}

    ```{r, results='asis', echo=FALSE, include=TRUE, message=FALSE}
    cat(sapply(1:nrow(m1_bin$c_hull$hull), function(x) paste0("\\draw", paste(paste0("(", apply(m1_bin$possible_t[cbind(m1_bin$c_hull$hull, m1_bin$c_hull$hull[,1])[x,],], 1, paste, collapse=","), ")"), collapse=" -- "), "; \n")))
    ```
    \end{tikzpicture}
  }
  \end{minipage}
  \caption{The convex hull of our statistic space in three dimensions for the toy RBM with one visible and one hidden node from multiple perspectives enclosed by an unrestricted hull of $\{0,1\}^3$ space. Note that the convex hull of $\{t({\bf x})\}$ does not fill the unrestricted hull because of the relationship between the elements of $\{t({\bf x})\}$.}
\label{fig:toyhull}
\end{figure}

Additionally, we can compute the mean parametrization of the model parameters as 
\begin{align*}
\mu(\boldsymbol \theta) &= \text{E}_{\boldsymbol \theta}\left[ t(X) \right] \\
&= \sum\limits_{x \in \{0,1\}^2} \left\{ t(x) \frac{\exp\left( \theta_{11} h_1 v_1 + \theta_{h1} h_1 + \theta_{v1} v_1\right)}{\sum\limits_{{\bf x} \in \{0,1\}^{2}}\exp\left( \theta_{11} h_1 v_1 + \theta_{h1} h_1 + \theta_{v1} v_1\right)} \right\} \\
&= \sum\limits_{x \in \{0,1\}^2} \left\{ t(x) \frac{\exp\left( \theta_{11} h_1 v_1 + \theta_{h1} h_1 + \theta_{v1} v_1\right)}{\gamma(\boldsymbol \theta)} \right\} \\
&= \left[
\begin{matrix}
\frac{\exp\left(\theta_{v_1}\right) + \exp\left(\theta_{11} + \theta_{v_1} + \theta_{h_1}\right)}{\gamma(\boldsymbol \theta)} \\
\frac{\exp\left(\theta_{h_1}\right) + \exp\left(\theta_{11} + \theta_{v_1} + \theta_{h_1}\right)}{\gamma(\boldsymbol \theta)} \\
\frac{\exp\left(\theta_{11} + \theta_{v_1} + \theta_{h_1}\right)}{\gamma(\boldsymbol \theta)} \\
\end{matrix} \right]
\end{align*}
These three functions can be visualized over multiple values of $\boldsymbol \theta$, as seen in figure \ref{fig:toymodel}. By examining these functions, we can see that as values of $\boldsymbol \theta$ grow larger in magnitude, at least one mean-parameterization functions is approaching the values 0 or 1, indicating a potential to be near to the boundary of the convex hull of $\{t({\bf x}): {\bf x} \in \mathcal{X}\}$. Thus, for a very small example we can see the relationship between values of $\boldsymbol \theta$ and model degeneracy.

```{r toymodel, fig.cap="\\label{fig:toymodel}The three mean-parameterization functions of the model parameters for a RBM with one visible and one hidden node.", out.height = "4in", fig.align='center'}
possibles <- stats(1, 1, type = "binary")

expand.grid(theta_h1 = seq(-2, 2, length.out = 5),
            theta_v1 = seq(-2, 2, length.out = 5),
            theta_11 = seq(-2, 2, length.out = 11)) %>%
  data.frame() %>%
  group_by(theta_v1, theta_h1, theta_11) %>%
  do(data.frame(gamma = sum(exp(possibles %*% t(data.matrix(.)))),
                mu_v1 = possibles[,"v1"] %*% exp(possibles %*% t(data.matrix(.))),
                mu_h1 = possibles[,"h1"] %*% exp(possibles %*% t(data.matrix(.))),
                mu_11 = possibles[,"theta11"] %*% exp(possibles %*% t(data.matrix(.))))) %>%
  mutate_each(funs(./gamma), starts_with("mu")) %>%
  select(-gamma) %>%
  gather(mu_parameter, value, mu_v1, mu_h1, mu_11) %>%
  mutate(theta_h1 = factor(theta_h1, labels = paste0("theta[h1]==", unique(theta_h1))), 
         theta_v1 = factor(theta_v1, labels = paste0("theta[v1]==", unique(theta_v1)))) %>%
  ggplot() +
  geom_line(aes(theta_11, value, colour = mu_parameter)) +
  facet_grid(theta_h1 ~ theta_v1, labeller = label_parsed) +
  xlab(expression(theta[11])) +
  scale_colour_discrete(expression(mu(theta)), labels = c(expression(theta[v1]), 
                                                          expression(theta[h1]), 
                                                          expression(theta[11])))

```

##Variable Encoding
Multiple encodings of the binary variables are possible. For example, we could allow $\mathcal{H} \subset \{0,1\}^H, \mathcal{V} \subset \{0,1\}^V$, as in the previous sections or we could encode the state of the variables as $\mathcal{H} \subset \{-1,1\}^H, \mathcal{V} \subset \{-1,1\}^V$. This will result in $t({\bf x}) \in \{0,1\}^{H + V + HV}$ or $t({\bf x}) \in \{-1,1\}^{H + V + HV}$ depending on how we encode on and off states in the nodes. To explore the effect of this encoding on the potential near degeneracy of an RBM, we can explore the behavior of the ratio of the volume of the unrestricted convex hull to the restructed convex hull of $\{t({\bf x})\}$ to determine how much the restriction in our statistics by including the cross product terms incurs.

For the small two node example, the binary encoding loses `r round((1-m1_bin$c_hull$vol)*100,2)`% of the volume compared to unrestricted space and the negative encoding loses `r round((1-m1_neg$c_hull$vol/(2^3))*100,2)`% of the volume compared to unrestricted space. This notion of lost volume can be helpful to conceptualize how difficult it will be for the mean parameterized vector $\mu(\boldsymbol \theta)$ to avoid the boundary, and thus avoid near degeneracy. In fact, if we look at the ratio of volume within a convex hull versus the volume of an unrestricted hull, we can see a relationship emerge as the number of nodes (and thus parameters) increases. In figure \ref{fig:volume_plot}, it is evident that as the number of parameters increses, this ratio is decreasing at an increasing rate, meaning it gets more and more difficult to avoid the boundary of the convex hull and thus near degeneracy. Additionally, it appears that the $\{-1,1\}$ encoding suffers slightly less from this problem.

```{r volume_plot, echo=FALSE, fig.cap="\\label{fig:volume_plot}The relationship between volume within the convex hull of our statistics and the convex hull of unrestricted space for different configurations of nodes. We compare the two encodings for on and off in a node.", out.height = "3in", fig.align='center'}
test_cases <- expand.grid(data.frame(1:4)[,rep(1,2)]) %>% 
  rename(H = X1.4, V = X1.4.1) %>%
  filter(H <= V) %>% #remove those cases with less visibles than hiddens. Is this necessary?
  mutate(n_param = H*V + H + V) %>%
  mutate(max_facets = (2^(H+V))^(floor(n_param/2))) %>%
  filter(n_param <= 11) #calc_hull can't handle any higher dimensions currently

bin_res <- plyr::dlply(test_cases, plyr::.(n_param), function(x) calc_hull(x$V, x$H, "binary"))
neg_res <- plyr::dlply(test_cases, plyr::.(n_param), function(x) calc_hull(x$V, x$H, "negative"))

plyr::ldply(bin_res, function(x) x$c_hull$vol) %>% 
  mutate(frac_vol = V1/(1^n_param)) %>%
  inner_join(plyr::ldply(neg_res, function(x) x$c_hull$vol) %>% 
               mutate(frac_vol = V1/(2^n_param)),
             by="n_param") %>%
   rename(vol.bin = V1.x, vol.neg = V1.y, frac_vol.bin = frac_vol.x, frac_vol.neg = frac_vol.y) %>%
  gather(vars, value, -n_param) %>%
  separate(vars, c("type", "encoding"), "\\.") %>%
  spread(type, value) %>%
  ggplot() +
  geom_point(aes(x=n_param, y=frac_vol, colour=encoding)) +
  geom_line(aes(x=n_param, y=frac_vol, colour=encoding, group=encoding)) +
  ylab("Fraction of unrestricted volume") +
  xlab("Number of parameters") +
  scale_colour_discrete("Encoding", labels=c("Binary (1,0)", "Negative (1,-1)"))
```

Thus, moving forward we will restrict our examinations to the $\{-1, 1\}$-encoding of the binary variables in the RBM.

#Exploring Manageable Examples

#Discussion

### 2016-01-14 4x4 model/fit_single_adaptive_mh_within_gibbs_marginal.R

@@ -14,9 +14,13 @@ load("written/sample_images.Rdata")
 load("written/variance_params.Rdata")
 load("written/sample.params.Rdata")
 
-params <- list(main_hidden = rep(0, H),
-               main_visible = rep(0, V),
-               interaction = rep(0, H*V))
+# params <- list(main_hidden = rep(0, H),
+#                main_visible = rep(0, V),
+#                interaction = rep(0, H*V))
+
+params <- list(main_hidden = rnorm(H, 0, 10),
+               main_visible = rnorm(V, 0, 10),
+               interaction = rnorm(H*V, 0, 10))
 
 h <- .01
 s1 <- 1e-4
@@ -29,13 +33,13 @@ trunc_const <- 1
 
 models_good <- sample_single_adaptive_mh_within_gibbs(visibles = flat_images_good$visibles, params0 = params, C = variance_params$C, C_prime =  variance_params$C_prime, trunc_const = trunc_const, h = h, s1 = s1, s2 = s2, mc.iter = mc.iter, conditional_function = log_conditional_marginal)
 models_bad <- sample_single_adaptive_mh_within_gibbs(visibles = flat_images_good$visibles, params0 = params, C = variance_params$C, C_prime = variance_params$C, trunc_const = trunc_const, h = h, s1 = s1, s2 = s2, mc.iter = mc.iter, conditional_function = log_conditional_marginal)
-#models_degen <- sample_single_adaptive_mh_within_gibbs(visibles = flat_images_degen$visibles, params0 = params, C = variance_params$C, C_prime = variance_params$C_prime, trunc_const = trunc_const, h = h, s1 = s1, s2 = s2, mc.iter = mc.iter, conditional_function = log_conditional_marginal)
+models_degen <- sample_single_adaptive_mh_within_gibbs(visibles = flat_images_degen$visibles, params0 = params, C = variance_params$C, C_prime = variance_params$C_prime, trunc_const = trunc_const, h = h, s1 = s1, s2 = s2, mc.iter = mc.iter, conditional_function = log_conditional_marginal)
 
-save(models_good, models_bad, file = "written/fitted_models_adaptive_mh_full_trunc_distn_marginal.Rdata")
+save(models_good, models_bad, models_degen, file = "written/fitted_models_adaptive_mh_full_trunc_distn_marginal_3.Rdata")
 
 #burnin + thinning
[…]

### 2016-01-14 4x4 model/test_compare_marginal_inits.R

@@ -1,0 +1,133 @@
+#libraries --------------------
+library(dplyr)
+library(tidyr)
+library(ggplot2)
+source("sample_functs.R")
+
+#data and params ------------------------
+H <- 4
+V <- 4
+
+load("written/sample_images.Rdata")
+
+load("written/fitted_models_adaptive_mh_full_trunc_distn_marginal_1.Rdata")
+marginal_bad_1 <- models_bad
+marginal_good_1 <- models_good
+load("written/fitted_models_adaptive_mh_full_trunc_distn_marginal_2.Rdata")
+marginal_bad_2 <- models_bad
+marginal_good_2 <- models_good
+load("written/fitted_models_adaptive_mh_full_trunc_distn_marginal_3.Rdata")
+marginal_bad_3 <- models_bad
+marginal_good_3 <- models_good
+
+
+#rm unneccesary data
+rm(models_bad)
+rm(models_good)
+
+load("written/sample.params.Rdata")
+params <- sample.params %>% 
+  mutate(label = ifelse(near_hull, "degen", "good")) %>% 
+  ungroup() %>%
+  select(label, starts_with("v"), starts_with("h"), starts_with("theta"), -H, -V) %>%
+  gather(theta, value, -label)
+
+
+reshape_loglik <- function(model, visibles) {
+  if("theta" %in% names(model)) {
+    theta <- model$theta
+  } else {
+    interaction <- t(apply(model$params$interaction, 1, t))
+    theta <- cbind(model$params$main_visible, model$params$main_hidden, interaction)
+  }
+  #hiddens <- model$hiddens
+  N <- visibles %>% nrow()
+  H <- dim(model$hiddens)[3]
+  V <- ncol(visibles)
+  
+  loglik <- do.call(rbind, lapply(seq_along(theta[, 1]), function(m) {
+    
+    possibles <- stats(H, V)
+    data <- visibles %>% data.frame
+    names(data) <- paste0("v", 1:V)
+    
+    data <- data %>%
+      left_join(possibles %>% data.frame()) %>% data.frame
+    
+    N <- nrow(data)
+    require(dplyr)
+    N <- nrow(data)
+    data <- data %>% data.frame() 
+    names(data) <- colnames(possibles)
+    H <- data %>% select(starts_with("h")) %>% ncol
+    V <- data %>% select(starts_with("V")) %>% ncol
+    
+    data %>%
[…]

### 2016-01-14 4x4 model/test_compare_mcmc.R

@@ -11,18 +11,28 @@ V <- 4
 #load truncatednormal prior
 load("written/fitted_models_adaptive_mh_trunc_distn_shrunk_0.6667.Rdata")
 shrunk_bad <- models_bad
-shrunk_degen <- models_degen
 shrunk_good <- models_good
+
+load("written/fitted_models_adaptive_mh_trunc_distn_marginal_1.Rdata")
+marginal_bad_1 <- models_bad
+marginal_good_1 <- models_good
+
+
+load("written/fitted_models_adaptive_mh_trunc_distn_marginal_2.Rdata")
+marginal_bad_2 <- models_bad
+marginal_good_2 <- models_good
+
+load("written/fitted_models_adaptive_mh_trunc_distn_marginal_3.Rdata")
+marginal_bad_3 <- models_bad
+marginal_good_3 <- models_good
 
 #load Jing's prior
 load("written/fitted_models_distn_0.26.Rdata")
 jing_bad <- models_bad
-jing_degen <- models_degen
 jing_good <- models_good
 
 load("written/fitted_models_adaptive_mh_trunc_distn_shrunk_jing_match.Rdata")
 shrunk_jing_bad <- models_bad
-shrunk_jing_degen <- models_degen
 shrunk_jing_good <- models_good
 
 #rm unneccesary data
@@ -61,15 +71,21 @@ reshape_sample_distn <- function(model) {
 #burn in more on the truncated, there are more iterations than the Jing, which gets complicated with plotting
 shrunk_sample_good <- reshape_sample_distn(shrunk_good) %>% filter(iter > 500) %>% mutate(iter = 1:n()) 
 shrunk_sample_bad <- reshape_sample_distn(shrunk_bad) %>% filter(iter > 500) %>% mutate(iter = 1:n()) 
-shrunk_sample_degen <- reshape_sample_distn(shrunk_degen) %>% filter(iter > 500) %>% mutate(iter = 1:n()) 
+
+marginal_sample_good_1 <- reshape_sample_distn(marginal_good_1)
+marginal_sample_bad_1 <- reshape_sample_distn(marginal_bad_1)
+
+marginal_sample_good_2 <- reshape_sample_distn(marginal_good_2)
+marginal_sample_bad_2 <- reshape_sample_distn(marginal_bad_2)
+
+marginal_sample_good_3 <- reshape_sample_distn(marginal_good_3)
[…]

### 2016-01-14 4x4 model/test_convergence.R

@@ -2,7 +2,7 @@ library(dplyr)
 library(ggplot2)
 library(tidyr)
 
-load("written/fitted_models_adaptive_mh_full_trunc_distn_shrunk_jing_match.Rdata")
+load("written/fitted_models_adaptive_mh_full_trunc_distn_marginal.Rdata")
 load("written/sample.params.Rdata")
 theta_good <- sample.params %>% 
   ungroup() %>% 
@@ -23,7 +23,7 @@ theta_est %>%
   mutate(iter = 1:n()) %>% 
   gather(variable, value, -iter) %>% 
   ggplot() + 
-  #geom_abline(aes(intercept = true_value, slope = 0), data = theta_good %>% data.frame %>% gather(variable, true_value), colour = "red") +
+  geom_abline(aes(intercept = true_value, slope = 0), data = theta_good %>% data.frame %>% gather(variable, true_value), colour = "red") +
   geom_line(aes(iter, value)) + 
   facet_wrap(~variable, scales = "free")
 

### 2016-01-14 4x4 model/test_fitted_model.R

@@ -10,7 +10,7 @@ V <- 4
 
 load("written/sample_images.Rdata")
 # load("written/fitted_models_adaptive_mh_trunc_distn.Rdata")
-load("written/fitted_models_distn_0.24.Rdata")
+load("written/fitted_models_adaptive_mh_full_trunc_distn_marginal_2.Rdata")
 load("written/sample.params.Rdata")
 params_degen <- list(main_hidden = sample.params %>% ungroup() %>% filter(near_hull) %>% select(starts_with("h"), -H) %>% data.matrix(),
                      main_visible = sample.params %>% ungroup() %>% filter(near_hull) %>% select(starts_with("v"), -V) %>% data.matrix(),

### 2016-01-14 writing/helpers/functions.R

@@ -1,0 +1,45 @@
+##function for creating all possible values of the t(x)
+stats <- function(H, V, type="negative") {
+  if(type == "binary") t <- data.frame(c(0,1))[,rep(1, H + V)]
+  else if(type == "negative") t <- data.frame(c(-1,1))[,rep(1, H + V)]
+  else stop("Type must be 'binary', 'negative'")
+  
+  names(t) <- c(paste0("h", 1:H), paste0("v", 1:V))  
+  
+  #all possibilities of statistics in (H + V)-space
+  t.grid <- expand.grid(t)
+  
+  for(i in 1:V){
+    for(j in (V+1):(H+V)){
+      t.grid <- cbind(t.grid, t.grid[,i]*t.grid[,j])
+      names(t.grid)[ncol(t.grid)] <- paste0("theta", i, j - H)
+    }
+  }
+  
+  return(data.matrix(t.grid))
+}
+
+##function for calculating the convex hull of t(x)
+calc_hull <- function(H, V, type="binary") {
+  #only possible values of t are 0 or 1
+  if(type == "binary") t <- data.frame(c(0,1))[,rep(1, H + V)]
+  else if(type == "negative") t <- data.frame(c(-1,1))[,rep(1, H + V)]
+  else stop("Type must be 'binary', 'negative', or 'experiment'")
+  
+  names(t) <- c(paste0("h", 1:H), paste0("v", 1:V))  
+  
+  #all possibilities of statistics in (H + V)-space
+  t.grid <- expand.grid(t)
+  
+  for(i in 1:H){
+    for(j in (H+1):(H+V)){
+      t.grid <- cbind(t.grid, t.grid[,i]*t.grid[,j])
+      names(t.grid)[ncol(t.grid)] <- paste0("theta", i, j - H)
+    }
+  }
+  
+  #calculate convex hull
+  library(geometry)
+  C <- convhulln(t.grid, options="FA")
+  return(list(possible_t=t.grid, c_hull=C))
+}

### 2016-01-14 writing/helpers/volume_relationship.R

@@ -1,60 +1,0 @@
-library(plyr)
-library(dplyr)
-library(tidyr)
-library(ggplot2)
-
-calc_hull <- function(H, V, type="binary") {
-  #only possible values of t are 0 or 1
-  if(type == "binary") t <- data.frame(c(0,1))[,rep(1, H + V)]
-  else if(type == "negative") t <- data.frame(c(-1,1))[,rep(1, H + V)]
-  else if(type == "experiment") t <- data.frame(c(1,3))[,rep(1, H + V)]
-  else stop("Type must be 'binary', 'negative', or 'experiment'")
-  
-  names(t) <- c(paste0("h", 1:H), paste0("v", 1:V))  
-  
-  #all possibilities of statistics in (H + V)-space
-  t.grid <- expand.grid(t)
-
-  for(i in 1:H){
-    for(j in (H+1):(H+V)){
-      t.grid <- cbind(t.grid, t.grid[,i]*t.grid[,j])
-      names(t.grid)[ncol(t.grid)] <- paste0("theta", i, j - H)
-    }
-  }
-  
-  #calculate convex hull
-  library(geometry)
-  C <- convhulln(t.grid, options="FA")
-  return(list(possible_t=t.grid, c_hull=C))
-}
-
-test_cases <- expand.grid(data.frame(1:4)[,rep(1,2)]) %>% 
-  dplyr::rename(H = X1.4, V = X1.4.1) %>%
-  dplyr::filter(H <= V) %>% #remove those cases with less visibles than hiddens. Is this necessary?
-  dplyr::mutate(n_param = H*V + H + V) %>%
-  dplyr::mutate(max_facets = (2^(H+V))^(floor(n_param/2))) %>%
-  dplyr::filter(n_param <= 11) #calc_hull can't handle any higher dimensions currently
-
-bin_res <- dlply(test_cases, .(n_param), function(x) calc_hull(x$V, x$H, "binary"), .progress="text")
-neg_res <- dlply(test_cases, .(n_param), function(x) calc_hull(x$V, x$H, "negative"), .progress="text")
-
-volume_plot <- ldply(bin_res, function(x) x$c_hull$vol) %>% 
-  dplyr::mutate(frac_vol = V1/(1^n_param)) %>%
-  dplyr::inner_join(ldply(neg_res, function(x) x$c_hull$vol) %>% 
-               mutate(frac_vol = V1/(2^n_param)),
[…]

### 2016-01-29 writing/degeneracy/draft.Rmd
added grid sampling for degeneracy

@@
 ##Joint Distribution
-Let ${\bf x} = \{h_1, \dots, h_H, v_1,\dots,v_V\}$ represent the states of the visible and hidden nodes in an RBM.
+Let $\boldsymbol x = \{h_1, \dots, h_H, v_1,\dots,v_V\}$ represent the states of the visible and hidden nodes in an RBM.
 Then we posit a parametric form for probabilities corresponding to the state of each node taking the value of $1$.
@@
 \label{eqn:pmf}
-f_{\boldsymbol \theta} ({\bf x}) = \frac{\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\sum\limits_{{\bf x} \in \{0,1\}^{H+V}}\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}
+f_{\boldsymbol \theta} (\boldsymbol x) = \frac{\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}
 \end{align}
@@
 Let the normalizing constant in equation \ref{eqn:pmf} be denoted $\gamma(\boldsymbol \theta)$ [@Vardeman602, pp.163-167].
-Additionally, we can refer to $Q({\bf x}) = \sum\limits_{i = 1}^H\sum\limits_{j=1}^V \theta_{ij} h_i v_j + \sum\limits_{i = 1}^H\theta_{hi} h_i + \sum\limits_{j = 1}^V\theta_{vj} v_j$ as the neg-potential function in our model.
+Additionally, we can refer to $Q(\boldsymbol x) = \sum\limits_{i = 1}^H\sum\limits_{j=1}^V \theta_{ij} h_i v_j + \sum\limits_{i = 1}^H\theta_{hi} h_i + \sum\limits_{j = 1}^V\theta_{vj} v_j$ as the neg-potential function in our model.
 The RBM model is parameterized by $\boldsymbol \theta$, which contains two types of parameters, main effects and interaction effects.
@@
 ##Connection to Deep Learning
+RBMs have risen to prominence in recent years due to their connection to deep learning [see @hinton2006fast, @salakhutdinov2012efficient, or @srivastava2013modeling for examples].
+By stacking multiple layers of RBMs in a deep architecture, the method's proponents claim the ability to learn "internal representations that become increasingly complex, which is considered to be a promising way of solving object and speech recognition problems" [@salakhutdinov2009deep, pp.
+450].
+The stacking is achieved by treating a hidden layer of one RBM as the visible layer in a second RBM, and so on until the desired architecture is created.
 
@@
 In Random Graph Model theory, \emph{degeneracy} occurs when there is a disproportionate amount of probability placed on only a few elements of the sample space, $\mathcal{X}$, by the model.
-@handcock2003assessing define a model to be near degenerate if $\mu(\boldsymbol \theta)$, the mean parametrization on the model parameters, is close to the boundary of the convex hull of $\{t({\bf x}): {\bf x} \in \mathcal{X}\}$, where $t({\bf x})$ is the set of statistics in the neg-potential function $Q({\bf x})$ from equation \ref{eqn:pmf}.
+@handcock2003assessing define a model to be near degenerate if $\mu(\boldsymbol \theta)$, the mean parametrization on the model parameters, is close to the boundary of the convex hull of $\{t_k(\boldsymbol x): \boldsymbol x \in \mathcal{X}, k = 1, \dots, K\}$, where $t_k(\boldsymbol x), k = 1, \dots, K\}$ is the set of statistics in the neg-potential function $Q(\boldsymbol x)$ from equation \ref{eqn:pmf}.
 
@@
 \mu(\boldsymbol \theta) &= \text{E}_{\boldsymbol \theta}\left[ t(\boldsymbol X) \right] \\
-&= \sum\limits_{x \in \{-1,1\}^{H + V}} \left\{ t(\boldsymbol x) f_{\boldsymbol \theta} ({\bf x}) \right\} \\
-&= \sum\limits_{x \in \{-1,1\}^{H + V}} \left\{ t(\boldsymbol x)\dfrac{\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\sum\limits_{{\bf x} \in \{0,1\}^{H+V}}\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}\right\} \\
+&= \sum\limits_{x \in \mathcal{X}} \left\{ t(\boldsymbol x) f_{\boldsymbol \theta} (\boldsymbol x) \right\} \\
+&= \sum\limits_{x \in \mathcal{X}} \left\{ t(\boldsymbol x)\dfrac{\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}\right\} \\
 \end{align*}
+
+@schweinberger2011instability introduces a related concept, instability, which  is characterized
+by excessive sensitivity and near-degeneracy. He provides the following tool for detecting instability in the parameters of a discrete exponential family model, and thus near-degeneracy.
+
+\begin{defn}
+A sufficient statistic $t_k(\boldsymbol x_N)$ is \emph{stable} if there exist constants $C > 0$ and $N_C > 0$ such that
+\begin{align}
+\max\limits_{\boldsymbol x_N \in \mathcal{X}_N} [t_k(\boldsymbol x_N)] \le CN \text{ for all } N > N_C
+\end{align}
+and \emph{unstable} if for any $C > 0$, however large, there exists $N_C > 0$ such that
+\begin{align}
+\max\limits_{\boldsymbol x_N \in \mathcal{X}_N} [t_k(\boldsymbol x_N)] > CN \text{ for all } N > N_C
+\end{align}
+\end{defn}
+
+\begin{lemma}
+A $K$-parameter exponential family $\{P_{\theta} , \theta \in \Theta \}$ with natural parameters $\theta_1,\dots, \theta_K$ and $K - 1$ stable sufficient statistics, $t_1(\boldsymbol x_N),\dots, t_{K-1}(\boldsymbol x_N)$, as well as one unstable sufficient statistic, $t_K(\boldsymbol x_N)$, is degenerate with respect to $t_K(\boldsymbol x_N)$.
+\end{lemma}
 
@@
 
-For this small model, we are able to illustrate the boundary of the convex hull of $\{t({\bf x}): {\bf x} \in \mathcal{X}\}$ because we are in three dimensions.
+For this small model, we are able to illustrate the boundary of the convex hull of $\{t_k(\boldsymbol x): \boldsymbol x \in \mathcal{X}, k = 1, \dots, K\}$ because we are in three dimensions.
 In figure \ref{fig:toyhull}, the convex hull of our statistic space in three dimensions for the toy RBM with one visible and one hidden node from multiple perspectives enclosed by an unrestricted hull of $\{0,1\}^3$ space is shown.
-In this small model, $\{t({\bf x})\}$ is defined as $\{t({\bf x})\} = \{v_1, h_1, v_1 h_1\}$ by definition and form of the neg-potential function $Q({\bf x})$.
-Note that the convex hull of $\{t({\bf x})\}$ does not fill the unrestricted hull because of the relationship between the elements of $\{t({\bf x})\}$.
+In this small model, $\{t(\boldsymbol x)\}$ is defined as $\{t(\boldsymbol x)\} = \{v_1, h_1, v_1 h_1\}$ by definition and form of the neg-potential function $Q(\boldsymbol x)$.
+Note that the convex hull of $\{t(\boldsymbol x)\}$ does not fill the unrestricted hull because of the relationship between the elements of $\{t(\boldsymbol x)\}$.
 
@@
   \caption{The convex hull of our statistic space in three dimensions for the toy RBM with one visible and one hidden node from multiple perspectives enclosed by an unrestricted hull of $\{0,1\}^3$ space.
-Note that the convex hull of $\{t({\bf x})\}$ does not fill the unrestricted hull because of the relationship between the elements of $\{t({\bf x})\}$.}
+Note that the convex hull of $\{t(\boldsymbol x)\}$ does not fill the unrestricted hull because of the relationship between the elements of $\{t(\boldsymbol x)\}$.}
 \label{fig:toyhull}
@@
 \mu(\boldsymbol \theta) &= \text{E}_{\boldsymbol \theta}\left[ t(X) \right] \\
-&= \sum\limits_{x \in \{0,1\}^2} \left\{ t(x) \frac{\exp\left( \theta_{11} h_1 v_1 + \theta_{h1} h_1 + \theta_{v1} v_1\right)}{\sum\limits_{{\bf x} \in \{0,1\}^{2}}\exp\left( \theta_{11} h_1 v_1 + \theta_{h1} h_1 + \theta_{v1} v_1\right)} \right\} \\
-&= \sum\limits_{x \in \{0,1\}^2} \left\{ t(x) \frac{\exp\left( \theta_{11} h_1 v_1 + \theta_{h1} h_1 + \theta_{v1} v_1\right)}{\gamma(\boldsymbol \theta)} \right\} \\
+&= \sum\limits_{x \in \mathcal{X}} \left\{ t(x) \frac{\exp\left( \theta_{11} h_1 v_1 + \theta_{h1} h_1 + \theta_{v1} v_1\right)}{\sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left( \theta_{11} h_1 v_1 + \theta_{h1} h_1 + \theta_{v1} v_1\right)} \right\} \\
+&= \sum\limits_{x \in \mathcal{X}} \left\{ t(x) \frac{\exp\left( \theta_{11} h_1 v_1 + \theta_{h1} h_1 + \theta_{v1} v_1\right)}{\gamma(\boldsymbol \theta)} \right\} \\
 &= \left[
@@
 These three functions can be visualized over multiple values of $\boldsymbol \theta$, as seen in figure \ref{fig:toymodel}.
-By examining these functions, we can see that as values of $\boldsymbol \theta$ grow larger in magnitude, at least one mean-parameterization functions is approaching the values 0 or 1, indicating a potential to be near to the boundary of the convex hull of $\{t({\bf x}): {\bf x} \in \mathcal{X}\}$.
+By examining these functions, we can see that as values of $\boldsymbol \theta$ grow larger in magnitude, at least one mean-parameterization functions is approaching the values 0 or 1, indicating a potential to be near to the boundary of the convex hull of $\{t_k(\boldsymbol x): \boldsymbol x \in \mathcal{X}, k = 1, \dots, K\}$.
 Thus, for a very small example we can see the relationship between values of $\boldsymbol \theta$ and model degeneracy.
@@
 For example, we could allow $\mathcal{H} \subset \{0,1\}^H, \mathcal{V} \subset \{0,1\}^V$, as in the previous sections or we could encode the state of the variables as $\mathcal{H} \subset \{-1,1\}^H, \mathcal{V} \subset \{-1,1\}^V$.
-This will result in $t({\bf x}) \in \{0,1\}^{H + V + HV}$ or $t({\bf x}) \in \{-1,1\}^{H + V + HV}$ depending on how we encode on and off states in the nodes.
-To explore the effect of this encoding on the potential near degeneracy of an RBM, we can explore the behavior of the ratio of the volume of the unrestricted convex hull to the restructed convex hull of $\{t({\bf x})\}$ to determine how much the restriction in our statistics by including the cross product terms incurs.
+This will result in $t(\boldsymbol x) \in \{0,1\}^{H + V + HV}$ or $t(\boldsymbol x) \in \{-1,1\}^{H + V + HV}$ depending on how we encode on and off states in the nodes.
+To explore the effect of this encoding on the potential near degeneracy of an RBM, we can explore the behavior of the ratio of the volume of the unrestricted convex hull to the restructed convex hull of $\{t(\boldsymbol x)\}$ to determine how much the restriction in our statistics by including the cross product terms incurs.
 
@@
 #Exploring Manageable Examples
+```{r split_sample, cache=TRUE}
+#reshape data function
+plot_data <- function(res, grid = FALSE) {
+
+  plot.data <- data.frame()
+
+  if(!grid) {
+    for(i in 1:nrow(res)) {
+      tmp <- res$g_theta[[i]] %>% data.frame()
+      H <- res[i,]$H
+      V <- res[i,]$V
+      N <- res[i,]$N
+      r1 <- res[i,]$r1
+      r2 <- res[i,]$r2
+      C <- res[i,]$C
+      epsilon <- res[i,]$epsilon
+
+      tmp %>% 
+        rowwise() %>% 
+        mutate_(ss_interaction = paste0("(", paste(paste0(names(.)[(H + V + 1):(H + V + H*V)], "^2"), collapse = " + "), ")"),
+                ss_main = paste0("sum(", paste(paste0(names(.)[1:(H+V)],"^2"), collapse = " + "), ")")) %>%    
+        ungroup() -> ratio
+
+      inner_join(ratio, res$outside[[i]] %>% ungroup()) %>%
+        select(ss_interaction, ss_main, near_hull) %>%
+        mutate(H = H, V = V, n_param = H + V + H*V, N = H + V, N = N, r1 = r1, r2 = r2, C = C, epsilon = epsilon) %>%
+        rbind(plot.data) -> plot.data
+      }
+    } else {
+      for(i in 1:nrow(res)) {
+        tmp <- res$g_theta[[i]] %>% data.frame()
+        H <- res[i,]$H
+        V <- res[i,]$V
+        N <- res[i,]$N
+        r1 <- res[i,]$r1
+        r2 <- res[i,]$r2
+
+        tmp %>% 
+          rowwise() %>% 
+          mutate_(ss_interaction = paste0("(", paste(paste0(names(.)[(H + V + 1):(H + V + H*V)], "^2"), collapse = " + "), ")"),
+                ss_main = paste0("sum(", paste(paste0(names(.)[1:(H+V)],"^2"), collapse = " + "), ")")) %>%  
+          ungroup() -> ratio
+
+        inner_join(ratio, res$outside[[i]] %>% ungroup()) %>%
+          select(ss_interaction, ss_main, near_hull) %>%
+          mutate(H = H, V = V, n_param = H + V + H*V, N = H + V, N = N, r1 = r1, r2 = r2) %>%
+          rbind(plot.data) -> plot.data
+      }
+    }
+
+  return(plot.data)
+}
+
+load("../data/results_split.RData")
+plot_dat_split <- res %>% plot_data
+
+load("../data/results_grid.RData")
+plot_dat_grid <- res %>% plot_data(grid = TRUE)
+```
+
+```{r split_plots}
+plot_dat_grid %>%
+  group_by(r1, r2, n_param) %>%
+  summarise(frac_degen = sum(near_hull)/n(), count = n()) %>% 
+  ungroup() %>% 
+  ggplot() +
+  geom_tile(aes(x = r1, y = r2, fill = frac_degen)) +
+  geom_contour(aes(x = r1, y = r2, z = frac_degen), colour = "black", bins = 8) +
+  geom_contour(aes(x = r1, y = r2, z = frac_degen), colour = "black", breaks = .05, size = 1.5) +
+  geom_abline(aes(intercept = 0, slope = 1), alpha = .5, lty = 2) +
+  scale_fill_gradient(low = "yellow", high = "red") +
+#   geom_point(aes(r1, r2, alpha = frac_degen), 
+#              data = plot_dat_split %>%
+#                group_by(r1, r2, n_param) %>%
+#                summarise(frac_degen = sum(near_hull)/n(), count = n())) +
+  facet_grid(~n_param) +
+  theme(aspect.ratio = 1) +
+  theme(legend.position = "bottom")
+
+plot_dat_grid %>% 
+  mutate(ss_interaction = sqrt(ss_interaction)/(H*V), ss_main = sqrt(ss_main)/(H + V)) %>%
+  #filter(near_hull) %>%
+  ggplot() +
+  geom_density_2d(aes(ss_main, ss_interaction, colour = ..level..)) +
+  scale_colour_gradient(low = "yellow", high = "red") +
+  facet_grid(~n_param) +
+  theme(aspect.ratio = 1) +
+  theme(legend.position = "none")
+
+```
 
 #Discussion
+
+#References

### 2016-01-29 4x4 model/test_compare_mcmc.R

@@ -12,28 +12,34 @@ V <- 4
 load("written/fitted_models_adaptive_mh_trunc_distn_shrunk_0.6667.Rdata")
 shrunk_bad <- models_bad
 shrunk_good <- models_good
+shrunk_degen <- models_degen
 
 load("written/fitted_models_adaptive_mh_trunc_distn_marginal_1.Rdata")
 marginal_bad_1 <- models_bad
 marginal_good_1 <- models_good
+marginal_degen_1 <- models_degen
 
 
 load("written/fitted_models_adaptive_mh_trunc_distn_marginal_2.Rdata")
 marginal_bad_2 <- models_bad
 marginal_good_2 <- models_good
+marginal_degen_2 <- models_degen
 
 load("written/fitted_models_adaptive_mh_trunc_distn_marginal_3.Rdata")
 marginal_bad_3 <- models_bad
 marginal_good_3 <- models_good
+marginal_degen_3 <- models_degen
 
 #load Jing's prior
 load("written/fitted_models_distn_0.26.Rdata")
 jing_bad <- models_bad
 jing_good <- models_good
+jing_degen <- models_degen
 
 load("written/fitted_models_adaptive_mh_trunc_distn_shrunk_jing_match.Rdata")
 shrunk_jing_bad <- models_bad
 shrunk_jing_good <- models_good
+shrunk_jing_degen <- models_degen
 
 #rm unneccesary data
 rm(models_bad)
@@ -70,23 +76,28 @@ reshape_sample_distn <- function(model) {
 
 #burn in more on the truncated, there are more iterations than the Jing, which gets complicated with plotting
 shrunk_sample_good <- reshape_sample_distn(shrunk_good) %>% filter(iter > 500) %>% mutate(iter = 1:n()) 
-shrunk_sample_bad <- reshape_sample_distn(shrunk_bad) %>% filter(iter > 500) %>% mutate(iter = 1:n()) 
+shrunk_sample_bad <- reshape_sample_distn(shrunk_bad) %>% filter(iter > 500) %>% mutate(iter = 1:n())
+shrunk_sample_degen <- reshape_sample_distn(shrunk_degen) %>% filter(iter > 500) %>% mutate(iter = 1:n()) 
 
 marginal_sample_good_1 <- reshape_sample_distn(marginal_good_1)
 marginal_sample_bad_1 <- reshape_sample_distn(marginal_bad_1)
+marginal_sample_degen_1 <- reshape_sample_distn(marginal_degen_1)
 
[…]

### 2016-01-29 degeneracy/grid_sample.R

@@ -38,6 +38,11 @@ samp_exp <- function(space, center, r) {
   paste0(r*vec/sqrt(sum(vec^2)) + center, collapse=",")
 }
 
+sample_sphere_unif <- function(space, rep, center, r) {
+  vec <- matrix(rnorm(space*rep), nrow = rep)
+  r*vec/t(matrix(rep(sqrt(rowSums(vec^2)), each = space), nrow = space))
+}
+
 in_hull <- function(point, hull_points) {
   require(lpSolveAPI)
   P <- data.matrix(point)
@@ -53,35 +58,32 @@ in_hull <- function(point, hull_points) {
 
 #sample split radii ---------------------------------------------
 n <- 100
-C <- seq(0.2, 3, by = 0.2)
 r_exp <- .05
-epsilon <- seq(-1, 1, length.out = 11)
 
-expand.grid(H = 1:4, V = 1:4, C = C, epsilon = epsilon) %>%
+expand.grid(H = 1:4, V = 1:4, r1 = seq(0.001, 3, length.out = 20), r2 = seq(0.001, 3, length.out = 20)) %>%
   mutate(N = H + V) %>%
-  mutate(n = n, r1 = sqrt(C), r2 = sqrt(C*(N/(H*V))^(epsilon))) %>%
-  group_by(H, V, C, epsilon, r1, r2, N) %>% 
-  do(samp = cbind(matrix(rnorm(n*.$N, mean = 0, sd = .$r1/3), nrow = n), matrix(rnorm(n*.$H*.$V, mean = 0, sd = .$r2/3), nrow = n))) -> split_sample
+  group_by(H, V, r1, r2, N) %>% 
+  do(samp = cbind(sample_sphere_unif(.$N, n, 0, .$r1), sample_sphere_unif(.$H*.$V, n, 0, .$r2))) -> grid_sample
 
-expand.grid(H = 1:4, V = 1:4, C = C, epsilon = epsilon) %>%
+expand.grid(H = 1:4, V = 1:4, r1 = seq(0.001, 3, length.out = 20), r2 = seq(0.001, 3, length.out = 20)) %>%
   mutate(N = H + V) %>%
-  mutate(n = n, r1 = sqrt(C), r2 = sqrt(C*(N/(H*V))^(epsilon))) %>%
-  group_by(H, V, C, epsilon, r1, r2, N) %>% 
-  do(stat = stats(.$H, .$V, "negative")) -> split_sample_stat
+  group_by(H, V, r1, r2, N) %>% 
[…]

### 2016-01-29 degeneracy/kaiser_method.R

@@ -47,8 +47,8 @@ expand.grid(power = as.numeric(names(model.dat)), multiplier = as.numeric(names(
 
 frame %>%
   group_by(power, multiplier, n_param) %>%
-  mutate(frac_degen = sum(models[[1]][, "near_hull"])/nrow(models[[1]]), 
-         radius = (models[[1]][1, "H"] + models[[1]][1, "V"])^power,
+  mutate(frac_degen = sum(models[[1]]$near_hull)/nrow(models[[1]]), 
+         radius = (models[[1]]$H[1] + models[[1]]$V[1])^power,
          happy = frac_degen < .05) %>%
   group_by(n_param, happy) %>% 
   sample_n(20) -> models.samp
@@ -75,7 +75,7 @@ no_int %>%
        t()) %>%
   left_join(no_int %>%
               group_by(n_param) %>%
-              do(stats = stats(.$models[[1]][1, c("H")], .$models[[1]][1, c("V")]))) %>%
+              do(stats = stats(.$models[[1]]$H[1], .$models[[1]]$V[1]))) %>%
   group_by(n_param) %>%
   do(no_int = expected_value(theta = .$compare[[1]], stats = .$stats[[1]]) %>% t()) %>%
   left_join(no_int %>% 

### 2016-01-29 degeneracy/radius_percent_degen_analyze_res.R

@@ -52,6 +52,7 @@ for (i in 1:length(plots)) {
 
 names(plot.dat) <- gsub("[.]", "_", r_center)
 do.call(rbind, plot.dat) %>%
+  ungroup() %>%
   mutate(name = rownames(do.call(rbind, plot.dat))) %>%
   separate(name, into = c("multiplier", "num"), "[.]") %>% 
   mutate(multiplier = as.numeric(gsub("_", ".", multiplier))) %>%

### 2016-01-29 degeneracy/radius_percent_degen_max_no_degen_plots_split.R

@@ -65,12 +65,14 @@ plot.dat %>%
   summarise(frac_degen = sum(near_hull)/n()) %>%
   filter(frac_degen < .05) %>% 
   group_by(n_param) %>%
-  summarise(best_radius = max(true_r)) -> best_rad_nodegen
+  summarise(best_radius = max(true_r),
+            split_radius = max(split_r)) -> best_rad_nodegen
 
 best_rad_nodegen %>%
+  gather(radii, value, best_radius, split_radius) %>%
   ggplot() +
-  geom_point(aes(n_param, best_radius)) +
-  geom_smooth(aes(n_param, best_radius), method = "lm") +
+  geom_point(aes(n_param, value, colour = radii)) +
+  geom_smooth(aes(n_param, value, colour = radii), method = "lm") +
   theme_bw(base_family = "serif") +
   theme(legend.position = "bottom")
 
@@ -80,7 +82,7 @@ plot.dat %>%
   summarise(frac_degen = sum(near_hull)/n()) %>%
   ggplot() +
   geom_jitter(aes(true_r, frac_degen, colour = factor(power)), alpha = .5) +
-  facet_wrap(~n_param_f) +
+  facet_grid(power~n_param_f) +
   theme_bw(base_family = "serif") +
   theme(legend.position = "bottom")
 

### 2016-02-03 writing/degeneracy/draft.Rmd
more writing

@@
 
-The mean parameterization on the model parameters, $\mu(\boldsymbol \theta)$, is defined as
+The mean parameterization on the model parameters, $\mu(\boldsymbol \theta)$, for the RBM is defined as
 \begin{align*}
@@
 @schweinberger2011instability introduces a related concept, instability, which  is characterized
-by excessive sensitivity and near-degeneracy. He provides the following tool for detecting instability in the parameters of a discrete exponential family model, and thus near-degeneracy.
+by excessive sensitivity and near-degeneracy. He provides the following tool for detecting instability in the a discrete exponential family model, and thus near-degeneracy.
 
 \begin{defn}
-A sufficient statistic $t_k(\boldsymbol x_N)$ is \emph{stable} if there exist constants $C > 0$ and $N_C > 0$ such that
+A discrete exponential family distribution is unstable if for any $C > 0$, however large, there exists $N_C > 0$ such that 
 \begin{align}
-\max\limits_{\boldsymbol x_N \in \mathcal{X}_N} [t_k(\boldsymbol x_N)] \le CN \text{ for all } N > N_C
-\end{align}
-and \emph{unstable} if for any $C > 0$, however large, there exists $N_C > 0$ such that
-\begin{align}
-\max\limits_{\boldsymbol x_N \in \mathcal{X}_N} [t_k(\boldsymbol x_N)] > CN \text{ for all } N > N_C
+\max\limits_{\boldsymbol x_N \in \mathcal{X}_N}[Q(\boldsymbol x_N)] > CN \text{ for all } N > N_C \label{eq:unstable}
 \end{align}
@@
 \begin{lemma}
-A $K$-parameter exponential family $\{P_{\theta} , \theta \in \Theta \}$ with natural parameters $\theta_1,\dots, \theta_K$ and $K - 1$ stable sufficient statistics, $t_1(\boldsymbol x_N),\dots, t_{K-1}(\boldsymbol x_N)$, as well as one unstable sufficient statistic, $t_K(\boldsymbol x_N)$, is degenerate with respect to $t_K(\boldsymbol x_N)$.
+If a discrete exponential family distribution is unstable, then it is degenerate in the sense that for any $0 < \epsilon < 1$, $P_\theta(\boldsymbol X_N \in \mathcal{M}_{\epsilon, N}) \rightarrow 1$ as $N \rightarrow \infty$ where $\mathcal{M}_{\epsilon, N}$ is the subset of $\epsilon$-modes of the probability mass function.
 \end{lemma}
+
+In the context of the RBM, $N$ is the number of nodes in the model, $H + V$. So, the distribution is unstable if for any $C > 0$,
+
+\begin{align*}
+\max\limits_{\boldsymbol x_N \in \mathcal{X}_N}[Q(\boldsymbol x_N)] &= \max\limits_{\boldsymbol x \in \mathcal{X}}\left[ \sum\limits_{i = 1}^H\sum\limits_{j=1}^V \theta_{ij} h_i v_j + \sum\limits_{i = 1}^H\theta_{hi} h_i + \sum\limits_{j = 1}^V\theta_{vj} v_j \right] \\
+&= \max\limits_{h_j, v_i \in \{0,1\}}\left[ \sum\limits_{i = 1}^H\sum\limits_{j=1}^V \theta_{ij} h_i v_j + \sum\limits_{i = 1}^H\theta_{hi} h_i + \sum\limits_{j = 1}^V\theta_{vj} v_j \right] \\
+&> C(H + V)
+\end{align*}
+
+There are $H + V + H*V$ terms in the sum that comprises $Q(\boldsymbol x)$, which leads to a natural split of the quantity into main effects, those terms based solely on $h_i$ and $v_j$ and interaction effects, those terms based on the product of $h_i v_j$.
+There are $H+V$ main effects terms and $H*V$ interaction effects terms.
+For $H$ and $V$ large enough, the number of interaction terms is growing much faster than $H + V$, meaning no matter the $C$, we are able to find values of $\boldsymbol theta$ that can make the left side of equation \ref{eq:unstable} large compared to $C(H+V)$.
+This points to the interaction terms as potentially the more dangerous parameters for causing near-degeneracy in the RBM.
+
+```{r interaction size, include=FALSE}
+expand.grid(H = 1:10, V = 1:10) %>%
+  mutate(interaction_terms = H*V) %>%
+  ggplot() +
+  geom_contour(aes(H, V, z = interaction_terms/(H + V), colour = ..level..), bins = 20) +
+  geom_contour(aes(H, V, z = interaction_terms/(H + V)), colour = "black", breaks = 1) +
+  scale_colour_gradient("H*V/(H + V)", low = "yellow", high = "red")
+
+```
 
@@
   gather(mu_parameter, value, mu_v1, mu_h1, mu_11) %>%
+  ungroup() %>%
   mutate(theta_h1 = factor(theta_h1, labels = paste0("theta[h1]==", unique(theta_h1))), 
@@
 
-```{r split_plots}
+To explore the behavior of the RBM parameters $\boldsymbol \theta$ as it related to model degeneracy and near degeneracy, we example various models of small size.
+For $H, V \in \{1, \dots, 4\}$, we simulate `r dim(res$samp[[1]])[1]` values of $\boldsymbol \theta$ with magnitude, $||\boldsymbol \theta || = \sqrt{\sum_{i = 1, \dots, H + V + H*V} \theta_i^2}$, between `r min(plot_dat_grid$r1)` and `r max(plot_dat_grid$r1)`.
+For each set of parameters we then calculate the mean parameterization, as detailed in section \ref{degeneracy-in-random-graph-models} and can classify each model as near-degenerate or viable and look at the fraction of models that were near-degenerate for each combination of magnitude of $\boldsymbol \theta$ and model size.
+
+In this numerical experiment, we split $\boldsymbol \theta$ into $\boldsymbol \theta_{interaction}$ and $\boldsymbol \theta_{main}$, in reference to which sufficient statistics the parameters correspond to, and allow the two types of terms to have varying magnitudes.
+The results of this numerical study can be seen in figure \ref{fig:split_plots}.
+
+```{r split_plots, fig.cap="\\label{fig:split_plots}A numerical experiment looking at the fraction of models that were near-degenerate for each combination of magnitude of $\\boldsymbol \\theta$ and model size."}
 plot_dat_grid %>%
-  group_by(r1, r2, n_param) %>%
+  group_by(r1, r2, H, V) %>%
   summarise(frac_degen = sum(near_hull)/n(), count = n()) %>% 
   ungroup() %>% 
+  mutate(Hiddens = paste0("Hiddens: ", H), Visibles = paste0("Visibles: ", V)) %>%
   ggplot() +
@@
   geom_abline(aes(intercept = 0, slope = 1), alpha = .5, lty = 2) +
-  scale_fill_gradient(low = "yellow", high = "red") +
-#   geom_point(aes(r1, r2, alpha = frac_degen), 
+  scale_fill_gradient("Fraction Degenerate", low = "yellow", high = "red") +
+#   geom_point(aes(r1, r2), alpha = .1, 
 #              data = plot_dat_split %>%
+#                mutate(ss_interaction = sqrt(ss_interaction), ss_main = sqrt(ss_main)) %>%
+#                select(-r1, -r2) %>%
+#                rename(r1 = ss_main, r2 = ss_interaction) %>%
 #                group_by(r1, r2, n_param) %>%
-#                summarise(frac_degen = sum(near_hull)/n(), count = n())) +
-  facet_grid(~n_param) +
-  theme(aspect.ratio = 1) +
-  theme(legend.position = "bottom")
+#                summarise(frac_degen = sum(near_hull)/n(), count = n()) %>%
+#                filter(frac_degen > 0)) +
+  facet_grid(Visibles~Hiddens) +
+  xlab(expression(group("||", theta[main], "||"))) +
+  ylab(expression(group("||", theta[interaction], "||"))) +
+  theme(aspect.ratio = 1)
+  #theme(legend.position = "bottom")
+```
 
+```{r split_plots2, warning=FALSE}
 plot_dat_grid %>% 
-  mutate(ss_interaction = sqrt(ss_interaction)/(H*V), ss_main = sqrt(ss_main)/(H + V)) %>%
-  #filter(near_hull) %>%
+  mutate(dist = sqrt(r1^2 + r2^2)) %>%
+  group_by(dist, H, V) %>%
+  summarise(frac_degen = sum(near_hull)/n()) %>%
+  mutate(Hiddens = as.character(H), Visibles = paste0("Visibles: ", V)) %>%
   ggplot() +
-  geom_density_2d(aes(ss_main, ss_interaction, colour = ..level..)) +
-  scale_colour_gradient(low = "yellow", high = "red") +
-  facet_grid(~n_param) +
-  theme(aspect.ratio = 1) +
-  theme(legend.position = "none")
-
+  geom_point(aes(dist, frac_degen, colour = Hiddens)) +
+  geom_smooth(aes(dist, frac_degen, colour = Hiddens, group = Hiddens)) +
+  facet_grid(~Visibles) +
+  ylim(c(0,1)) +
+  xlab(expression(paste(group("||", theta, "||"), ", Distance from the origin"))) +
+  ylab("Fraction Degenerate")
 ```

### 2016-02-04 writing/degeneracy/draft.Rmd
some more writing before the meeting

@@
 To explore the behavior of the RBM parameters $\boldsymbol \theta$ as it related to model degeneracy and near degeneracy, we example various models of small size.
-For $H, V \in \{1, \dots, 4\}$, we simulate `r dim(res$samp[[1]])[1]` values of $\boldsymbol \theta$ with magnitude, $||\boldsymbol \theta || = \sqrt{\sum_{i = 1, \dots, H + V + H*V} \theta_i^2}$, between `r min(plot_dat_grid$r1)` and `r max(plot_dat_grid$r1)`.
+For $H, V \in \{1, \dots, 4\}$, we sample uniformly `r dim(res$samp[[1]])[1]` values of $\boldsymbol \theta$ with magnitude, $||\boldsymbol \theta || = \sqrt{\sum_{i = 1, \dots, H + V + H*V} \theta_i^2}$, between `r min(plot_dat_grid$r1)` and `r max(plot_dat_grid$r1)`.
 For each set of parameters we then calculate the mean parameterization, as detailed in section \ref{degeneracy-in-random-graph-models} and can classify each model as near-degenerate or viable and look at the fraction of models that were near-degenerate for each combination of magnitude of $\boldsymbol \theta$ and model size.
@@
 The results of this numerical study can be seen in figure \ref{fig:split_plots}.
+From this figure, it is evident that the fraction of near-degenerate models simulated at specified magnitudes of $\boldsymbol theta_{main}$ and $\boldsymbol theta_{interaction}$ is more closely related to the combined distance from the origin of these two vectors, than any individual group of parameter values.
 
-```{r split_plots, fig.cap="\\label{fig:split_plots}A numerical experiment looking at the fraction of models that were near-degenerate for each combination of magnitude of $\\boldsymbol \\theta$ and model size."}
+```{r split_plots, fig.cap="\\label{fig:split_plots}A numerical experiment looking at the fraction of models that were near-degenerate for each combination of magnitude of $\\boldsymbol \\theta$ and model size.
+The thick black line shows the level where the fraction of -near degenerate models is .05."}
 plot_dat_grid %>%
@@
   geom_abline(aes(intercept = 0, slope = 1), alpha = .5, lty = 2) +
-  scale_fill_gradient("Fraction Degenerate", low = "yellow", high = "red") +
+  scale_fill_gradient("Fraction Near-degenerate", low = "yellow", high = "red") +
 #   geom_point(aes(r1, r2), alpha = .1, 
@@
 
-```{r split_plots2, warning=FALSE}
+Figure \ref{fig:split_plots_dist} shows the fraction near-degenerate for each magnitude of $\boldsymbol \theta$ for each model.
+For each number of visibles in the model, the fraction near-degenerate becomes greater than zero at larger values of $||\boldsymbol \theta||$ as $H$ increases and the slope becomes steeper as $H$ increases also.
+This shows that as model size gets larger, the risk of degeneracy starts at a slightly larger magnitude of parameters, but very quickly increases until reaching close to 1.
+
+```{r split_plots2, warning=FALSE, fig.cap="\\label{fig:split_plots_dist}The fraction near-degenerate for each magnitude of $\\boldsymbol \\theta$ for each model.
+For each number of visibles in the model, the fraction near-degenerate becomes greater than zero at larger values of $||\\boldsymbol \\theta||$ as $H$ increases and the slope becomes steeper as $H$ increases also."}
 plot_dat_grid %>% 
@@
   xlab(expression(paste(group("||", theta, "||"), ", Distance from the origin"))) +
-  ylab("Fraction Degenerate")
+  ylab("Fraction Near-degenerate")
+```
+
+```{r, frac-degen, include=FALSE}
+plot_dat_grid %>%
+  mutate(dist = sqrt(r1^2 + r2^2)) %>%
+  group_by(dist, H, V) %>%
+  summarise(frac_degen = sum(near_hull)/n()) %>%
+  ungroup() %>%
+  group_by(H, V) -> tmp
+
+tmp %>%
+  filter(frac_degen > 0) %>%
+  mutate(min_degen_pos = min(dist)) %>%
+  select(H, V, min_degen_pos) %>% unique() %>%
+left_join(tmp %>%
+  filter(frac_degen > .6) %>%
+  mutate(min_degen_50 = min(dist)) %>%
+  select(H, V, min_degen_50) %>% unique()) %>%
+  mutate(min_degen_pos = min_degen_pos/(H + V + H*V),
+         min_degen_50 = min_degen_50/(H + V + H*V)) %>%
+  mutate(degen_ratio = min_degen_50/min_degen_pos) %>%
+  ggplot() +
+  geom_point(aes(V, degen_ratio, colour = factor(H))) +
+  geom_line(aes(V, degen_ratio, colour = factor(H), group = factor(H))) +
+  ylab("Ratio of Min Distance above 0% near-degenerate to 50% near-degenerate") +
+  xlab("Visibles") +
+  scale_colour_discrete("Hiddens")
+
 ```

### 2016-02-17 writing/degeneracy/draft.Rmd
started on poster

@@
 In Random Graph Model theory, \emph{degeneracy} occurs when there is a disproportionate amount of probability placed on only a few elements of the sample space, $\mathcal{X}$, by the model.
+We consider three related definitions of degeneracy as they relate to a graphical model.
+Section \ref{mean-parameterization} presents a definition from @handcock2003assessing based on the mean parameterization of the model parameters, section \ref{neg-potential} presents a definition from @schweinberger2011instability based on the neg-potential function, and section \ref{conditional-expectation} presents a definition from @kaiser2006markov that involves the difference between marginal and conditional expectations of the model.
+
+##Mean Parameterization
 @handcock2003assessing define a model to be near degenerate if $\mu(\boldsymbol \theta)$, the mean parametrization on the model parameters, is close to the boundary of the convex hull of $\{t_k(\boldsymbol x): \boldsymbol x \in \mathcal{X}, k = 1, \dots, K\}$, where $t_k(\boldsymbol x), k = 1, \dots, K\}$ is the set of statistics in the neg-potential function $Q(\boldsymbol x)$ from equation \ref{eqn:pmf}.
@@
 
+##Neg-potential
 @schweinberger2011instability introduces a related concept, instability, which  is characterized
@@
 
+##Conditional Expectation
+
 ```{r interaction size, include=FALSE}
@@
 
-##Small example
+#Tiny example
 To illustrate the idea of graph degeneracy in a RBM, we will consider a the smallest possible toy example that consists of one visible node $v_1$ and one hidden node $h_1$, both binary.
@@
 \end{figure}
+
+##Degeneracy three ways
 
@@
 ```{r split_plots2, warning=FALSE, fig.cap="\\label{fig:split_plots_dist}The fraction near-degenerate for each magnitude of $\\boldsymbol \\theta$ for each model.
-For each number of visibles in the model, the fraction near-degenerate becomes greater than zero at larger values of $||\\boldsymbol \\theta||$ as $H$ increases and the slope becomes steeper as $H$ increases also."}
+For each number of visibles in the model, the fraction near-degenerate becomes greater than zero at larger values of $||\\boldsymbol \\theta||$ as $H$ increases and the slope becomes steeper as $H$ increases also.", fig.height=3}
 plot_dat_grid %>% 

### 2016-02-18 presentations/coda2016/coda2016_poster.tex
added kaiser method on the grid and some writing

%%big poster size with big fonts
\documentclass[extrafontsizes, 30pt]{memoir}
\usepackage[paperheight=31in,paperwidth=47in,margin=1in,heightrounded,showframe]{geometry}

%%background from pptx
\usepackage{background}
\backgroundsetup{
scale=1,
angle=0,
opacity=1,  %% adjust
contents={\includegraphics[width=\paperwidth,height=\paperheight]{images/coda2016_poster_background.pdf}}
}

%%put words where I want
\usepackage[absolute, overlay]{textpos}
\setlength{\TPHorizModule}{1in}
\setlength{\TPVertModule}{\TPHorizModule}

%%tikz setup
\usepackage{tikz, subfig, amsthm}
\usepackage{tikz-3dplot}
\usetikzlibrary{arrows, shapes, positioning}

%%other options
\definecolor{isublue}{RGB}{58,117,196}
\definecolor{isured}{RGB}{206,17,38}
\counterwithout{figure}{chapter}
\usepackage{amsmath}
\setlength{\parindent}{0cm}

\begin{document}

%%-----------------------------------------------
%%
%%Middle big box
%%
\begin{textblock}{18.01}(14.53, 6.71)
{\large \bfseries Restricted Boltzmann machine (RBM)}

\begin{figure}[ht]
  \centering
  \resizebox{\linewidth}{!}{\input{images/rbm.tikz}}
  \caption{An example restricted Boltzmann machine (RBM), which consists of two layers, a hidden ($\mathcal{H}$) and a visible layer ($\mathcal{V}$), with no connections within a layer. Hidden nodes indicated by gray circles and the visible nodes indicated by white circles.}
  \label{fig:rbm}
\end{figure}

\begin{figure}[ht]
  \centering
  \input{images/visibles.tikz}
  \caption{Visibles diagram.}
  \label{fig:visibles}
\end{figure}

~\\[-1cm]
{\bfseries Joint distribution} \\[.25cm]
Let $\boldsymbol x = \{h_1, \dots, h_H, v_1,\dots,v_V\}$ represent the states of the visible and hidden nodes in an RBM. Then  the probability corresponding to the state of each node taking the value of $1$:
\begin{align}
\label{eqn:pmf}
f_{\boldsymbol \theta} (\boldsymbol x) = \dfrac{\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)} 
\end{align}
\end{textblock}

%%-----------------------------------------------
%%
%%Top left box
%%
\begin{textblock}{14.08}(.8, 5.71)
{\large \bfseries Deep learning}

\end{textblock}

%%-----------------------------------------------
%%
%%Middle left box
%%
\begin{textblock}{14.08}(.8, 11.83)
{\large \bfseries Degeneracy in random graph models}

\end{textblock}

%%-----------------------------------------------
%%
%%Bottom left box
%%
\begin{textblock}{14.08}(.8, 20.64)
{\large \bfseries Tiny example}

\end{textblock}

%%-----------------------------------------------
%%
%%Top right box
%%
\begin{textblock}{14.08}(33.5, 5.71)
{\large \bfseries Managable examples}
\end{textblock}

%%-----------------------------------------------
%%
%%Middle right box
%%
\begin{textblock}{14.08}(33.5, 12.91)
{\large \bfseries Fitting}

\end{textblock}

%%-----------------------------------------------
%%
%%Bottom right box
%%
\begin{textblock}{14.08}(33.5, 23.85)
{\large \bfseries Discussion}
\end{textblock}

%%-----------------------------------------------
%%
%%Bottom middle box
%%
\begin{textblock}{15.08}(16.17, 24.84)
{\large \bfseries References}
\end{textblock}

\end{document}

### 2016-02-18 writing/degeneracy/draft.Rmd
added kaiser method on the grid and some writing

@@
 We consider three related definitions of degeneracy as they relate to a graphical model.
-Section \ref{mean-parameterization} presents a definition from @handcock2003assessing based on the mean parameterization of the model parameters, section \ref{neg-potential} presents a definition from @schweinberger2011instability based on the neg-potential function, and section \ref{conditional-expectation} presents a definition from @kaiser2006markov that involves the difference between marginal and conditional expectations of the model.
+Section \ref{mean-parameterization} presents a definition from @handcock2003assessing based on the mean parameterization of the model parameters, section \ref{neg-potential} presents a definition from @schweinberger2011instability based on the neg-potential function, and section \ref{conditional-expectation} presents a definition from @kaiser2006markov that involves the difference between expectations of the model and expectations under independence.
 
@@
 
-##Conditional Expectation
-
 ```{r interaction size, include=FALSE}
@@
 ```
+
+##Expectation Differences
+The characterization put forth by @kaiser2006markov of degeneracy in a model concerns the difference between model expectations and expectations given independence.
+Using this concept of dependence in a model, it is possible to understand what conditions lead to near-degeneracy in a model versus those that guarantee non-degenerate models.
+In this method, a model is considered without near-degeneracy if its expected value is not very different from the same model with assumed independence.
+The quantities to compare in the RBM case are
+
+\begin{align*}
+\text{E}\left[\boldsymbol X\right] &= \sum\limits_{\boldsymbol x \in \mathcal{X}} \boldsymbol x f_{\boldsymbol \theta}(\boldsymbol x) \\
+&= \sum\limits_{\boldsymbol x \in \mathcal{X}} \boldsymbol x \frac{\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)} \\
+&= \frac{\sum\limits_{\boldsymbol x \in \mathcal{X}} x_k \exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}  \qquad k = 1, \dots, H + V
+\end{align*}
 
@@
 The results of this numerical study can be seen in figure \ref{fig:split_plots}.
-From this figure, it is evident that the fraction of near-degenerate models simulated at specified magnitudes of $\boldsymbol theta_{main}$ and $\boldsymbol theta_{interaction}$ is more closely related to the combined distance from the origin of these two vectors, than any individual group of parameter values.
+From this figure, it is evident that the fraction of near-degenerate models simulated at specified magnitudes of $\boldsymbol \theta_{main}$ and $\boldsymbol \theta_{interaction}$ is more closely related to the combined distance from the origin of these two vectors, than any individual group of parameter values.
 
@@
 ```{r split_plots2, warning=FALSE, fig.cap="\\label{fig:split_plots_dist}The fraction near-degenerate for each magnitude of $\\boldsymbol \\theta$ for each model.
-For each number of visibles in the model, the fraction near-degenerate becomes greater than zero at larger values of $||\\boldsymbol \\theta||$ as $H$ increases and the slope becomes steeper as $H$ increases also.", fig.height=3}
+For each number of visibles in the model, the fraction near-degenerate becomes greater than zero at larger values of $||\\boldsymbol \\theta||$ as $H$ increases and the slope becomes steeper as $H$ increases also.", fig.height=2}
 plot_dat_grid %>% 

### 2016-02-18 degeneracy/kaiser_method_grid.R

@@ -1,0 +1,34 @@
+#libraries -----------------------
+library(dplyr)
+library(tidyr)
+library(ggplot2)
+source("functions.R")
+
+#load sample params ---------------------------------------------
+n <- 100
+load("apps/results_grid.RData") 
+
+indep_params <- function(samp, H, V) {
+  samp[, (H + V + 1):ncol(samp)] <- 0
+  samp
+}
+
+res %>%
+  group_by(H, V, N, r1, r2) %>%
+  do(indep_exp = t(expected_value(t(indep_params(.$samp[[1]], .$H, .$V)), .$stat[[1]]))[, -((.$H + .$V + 1):(.$H + .$V + .$H*.$V))],
+     marg_exp = .$g_theta[[1]][, (.$H + .$V + .$H*.$V + 1):(ncol(.$g_theta[[1]])-.$H*.$V)]) -> exp_vals
+
+exp_vals %>%
+  group_by(H, V, N, r1, r2) %>%
+  do(data.frame(mean_abs_diff = mean(abs(.$indep_exp[[1]] - .$marg_exp[[1]])))) %>%
+  mutate(Hiddens = paste0("Hiddens: ", H), Visibles = paste0("Visibles: ", V)) %>%
+  ggplot() +
+  geom_tile(aes(x = r1, y = r2, fill = mean_abs_diff)) +
+  geom_contour(aes(x = r1, y = r2, z = mean_abs_diff), colour = "black", bins = 8) +
+  #geom_contour(aes(x = r1, y = r2, z = frac_degen), colour = "black", breaks = .05, size = 1.5) +
+  #geom_abline(aes(intercept = 0, slope = 1), alpha = .5, lty = 2) +
+  scale_fill_gradient("Mean absolute difference in expectation", low = "yellow", high = "red") +
+  facet_grid(Visibles~Hiddens) +
+  xlab(expression(group("||", theta[main], "||"))) +
+  ylab(expression(group("||", theta[interaction], "||"))) +
+  theme(aspect.ratio = 1)

### 2016-02-20 presentations/coda2016/coda2016_poster.tex
good progress on poster

@@
 \usepackage{tikz-3dplot}
-\usetikzlibrary{arrows, shapes, positioning}
+\usetikzlibrary{arrows, shapes, positioning, backgrounds}
 
@@
 \counterwithout{figure}{chapter}
+\counterwithout{table}{chapter}
 \usepackage{amsmath}
 \setlength{\parindent}{0cm}
+\def\vdotswide{\vbox{\baselineskip=20pt \lineskiplimit=0pt \kern6pt \hbox{.}\hbox{.}\hbox{.}}}
+
+%%big sums
+\usepackage{graphicx}
+\usepackage{calc}
+\newlength{\depthofsumsign}
+\setlength{\depthofsumsign}{\depthof{$\sum$}}
+\newcommand{\nsum}[1][1.4]{% only for \displaystyle
+    \mathop{%
+        \raisebox
+            {-#1\depthofsumsign+1\depthofsumsign}
+            {\scalebox
+                {#1}
+                {$\displaystyle\sum$}%
+            }
+    }
+}
+
+\newcommand{\Nsum}[1][1.8]{% only for \displaystyle
+    \mathop{%
+        \raisebox
+            {-#1\depthofsumsign+1\depthofsumsign}
+            {\scalebox
+                {#1}
+                {$\displaystyle\sum$}%
+            }
+    }
+}
+
+%%bibliography
+\usepackage[backend=bibtex]{biblatex}
+\addbibresource{refs.bib}
+\renewcommand*{\bibfont}{\tiny}
 
 \begin{document}
+%% For every picture that defines or uses external nodes, you'll have to
+%% apply the 'remember picture' style. To avoid some typing, we'll apply
+%% the style to all pictures.
+%%http://www.texample.net/tikz/examples/connecting-text-and-graphics/
+\tikzstyle{every picture}+=[remember picture]
+\tikzstyle{na} = [baseline=-.5ex]
 
@@
 %%
-\begin{textblock}{18.01}(14.53, 6.71)
+\begin{textblock}{16}(15.5, 6.71)
 {\large \bfseries Restricted Boltzmann machine (RBM)}
@@
   \caption{An example restricted Boltzmann machine (RBM), which consists of two layers, a hidden ($\mathcal{H}$) and a visible layer ($\mathcal{V}$), with no connections within a layer.
-Hidden nodes indicated by gray circles and the visible nodes indicated by white circles.}
+Hidden nodes indicated by gray circles and the visible nodes indicated by white circles \cite{friedman2001elements}.}
   \label{fig:rbm}
@@
 \label{eqn:pmf}
-f_{\boldsymbol \theta} (\boldsymbol x) = \dfrac{\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}
+f_{\boldsymbol \theta} (\boldsymbol x) = \frac{\exp\left(\nsum\limits_{i = 1}^V \nsum\limits_{j=1}^H \theta_{ij} v_i h_j + \nsum\limits_{i = 1}^V\theta_{v_i} v_i + \nsum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\Nsum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\nsum\limits_{i = 1}^V \nsum\limits_{j=1}^H \theta_{ij} v_i h_j + \nsum\limits_{i = 1}^V\theta_{v_i} v_i + \nsum\limits_{j = 1}^H\theta_{h_j} h_j\right)}
 \end{align}
+
 \end{textblock}
@@
 %%
-\begin{textblock}{14.08}(.8, 5.71)
-{\large \bfseries Deep learning}
+\begin{textblock}{14}(.65, 5.71)
+{\large \bfseries Deep learning} \\[.25cm]
+\begin{minipage}{.62\textwidth}
+Three layer deep Boltzmann machine, with visible-to-hidden and hidden-to-hidden connections but no within-layer connections.
+This can be considered as multiple single layer restricted Boltzmann machines with the lower stack hidden layer acting as the visible layer for the higher stacked model.
+Claimed ability to learn "internal representations that become increasingly complex" \cite{salakhutdinov2009deep}, used in classification problems.
+\end{minipage}
+\begin{minipage}{.38\textwidth}
+\begin{figure}
+  \centering
+  \input{images/deep_rbm.tikz}
+  \caption{Deep RBM example.}
+  \label{fig:deep_rbm}
+\end{figure}
+\end{minipage}
 
@@
 %%
-\begin{textblock}{14.08}(.8, 11.83)
-{\large \bfseries Degeneracy in random graph models}
-
+\begin{textblock}{14}(.65, 11.33)
+{\large \bfseries Degeneracy, instability, and uninterpretability... Oh my!}
+~\\[.25cm]
+\begin{table}
+\centering
+\begin{tabular}{| p{.48\linewidth} | p{.5\linewidth} |}
+\hline
+Characteristic & Detection \\
+\hline
+\hline
+{\bfseries Near-degeneracy.} Occurs when there is a disproportionate amount of probability placed on only a few elements of the sample space by the model \cite{handcock2003assessing}.
+& If the mean parametrization on the model parameters, $\mu(\boldsymbol \theta)$, is close to the boundary of the convex hull of the set of statistics in the neg-potential function $Q(\boldsymbol x)$.
+\\
+\hline
+{\bfseries Instability.} Small changes in natural parameters result in large changes of the pmf, excessive sensitivity \cite{schweinberger2011instability}.
+& If for any $C > 0$ there exists $N_C > 0$ such that
+$\max\limits_{\boldsymbol x_N \in \mathcal{X}_N}[Q(\boldsymbol x_N)] > CN$ for all $N > N_C$. Where $Q(\cdot)$ is the neg-potential function of the model. \\
+\hline
+{\bfseries Uninterpretability.} Due to the existence of dependence, marginal mean-structure no longer maintained \cite{kaiser2007statistical}.
+& If the magnitude of the difference between model expectations and expectations under independence, $\left|\text{E}(\boldsymbol X | \boldsymbol \theta) - \text{E}(\boldsymbol X | \emptyset ) \right|$, is large.\\
+\hline
+\end{tabular}
+\label{tab:degen}
+\caption{Table of ``bad" model characteristics.}
+\end{table}
 \end{textblock}
@@
 %%
-\begin{textblock}{14.08}(.8, 20.64)
-{\large \bfseries Tiny example}
+\begin{textblock}{14}(.65, 22.94)
+{\large \bfseries Variable encoding}
 
@@
 %%
-\begin{textblock}{14.08}(33.5, 5.71)
-{\large \bfseries Managable examples}
+\begin{textblock}{14}(32.4, 5.71)
+{\large \bfseries Managable examples} \\[.25cm]
 \end{textblock}
@@
 %%
-\begin{textblock}{14.08}(33.5, 12.91)
-{\large \bfseries Fitting}
+\begin{textblock}{14}(32.4, 12.91)
+{\large \bfseries Model fitting} \\[.25cm]
+Fitting is really hard to get right (Bayesian or ML).
+If you are able to fit this model properly, the result is the empirical distribution of images.
+You spent a lot of time just to get the empirical distribution back.
+If you do not have an instance of every possible image in your training set, certain parameter must $\rightarrow \infty$, making fitting more complicated (shrink)
+In reality it's practically impossible to have at least one of every possible image in your training set (large images, color scales, etc.)
 
@@
 %%
-\begin{textblock}{14.08}(33.5, 23.85)
-{\large \bfseries Discussion}
+\begin{textblock}{14}(32.4, 23.85)
+{\large \bfseries Discussion} \\[.25cm]
+\end{textblock}
+
+\begin{textblock}{14.54}(31.86, 24.6)
+These models very easily are degenerate, unstable, and uninterpretable. Meaning, the space of possible fitted parameter values that leads to a proper model is highly restricted.
+
+To further complicate things, the proper fitting these models is very intricate and easily leads to parameter values running off to $\infty$.
+
+If you're thinking of using RBMs, or stacked RBMs in a deep architecture, don't. Instead, fit a less flexible and more sensible model.
 \end{textblock}
@@
 %%
-\begin{textblock}{15.08}(16.17, 24.84)
+\begin{textblock}{14.5}(16.17, 24.84)
 {\large \bfseries References}
+\printbibliography[heading=none]
+
 \end{textblock}

### 2016-02-20 degeneracy/kaiser_method_grid.R

@@ -20,15 +20,29 @@ res %>%
 
 exp_vals %>%
   group_by(H, V, N, r1, r2) %>%
-  do(data.frame(mean_abs_diff = mean(abs(.$indep_exp[[1]] - .$marg_exp[[1]])))) %>%
+  do(data.frame(mean_abs_diff = apply(abs(.$indep_exp[[1]] - .$marg_exp[[1]]), 1, mean))) %>%
+  group_by(H, V, N, r1, r2) %>%
+  summarise(mean_abs_diff = mean(mean_abs_diff)) %>%
   mutate(Hiddens = paste0("Hiddens: ", H), Visibles = paste0("Visibles: ", V)) %>%
   ggplot() +
   geom_tile(aes(x = r1, y = r2, fill = mean_abs_diff)) +
   geom_contour(aes(x = r1, y = r2, z = mean_abs_diff), colour = "black", bins = 8) +
-  #geom_contour(aes(x = r1, y = r2, z = frac_degen), colour = "black", breaks = .05, size = 1.5) +
+  #geom_contour(aes(x = r1, y = r2, z = mean_abs_diff), colour = "black", breaks = .1, size = 1.5) +
   #geom_abline(aes(intercept = 0, slope = 1), alpha = .5, lty = 2) +
   scale_fill_gradient("Mean absolute difference in expectation", low = "yellow", high = "red") +
   facet_grid(Visibles~Hiddens) +
   xlab(expression(group("||", theta[main], "||"))) +
   ylab(expression(group("||", theta[interaction], "||"))) +
   theme(aspect.ratio = 1)
+
+exp_vals %>% 
+  mutate(dist = sqrt(r1^2 + r2^2)) %>%
+  group_by(dist, H, V) %>%
+  do(data.frame(mean_abs_diff = mean(abs(.$indep_exp[[1]] - .$marg_exp[[1]])))) %>%
+  mutate(Hiddens = as.character(H), Visibles = paste0("Visibles: ", V)) %>%
+  ggplot() +
+  geom_point(aes(dist, mean_abs_diff, colour = Hiddens)) + ylim(c(0, .0025))
+  #geom_smooth(aes(dist, mean_abs_diff, colour = Hiddens, group = Hiddens)) +
+  facet_grid(~Visibles) +
+  xlab(expression(paste(group("||", theta, "||"), ", Distance from the origin"))) +
+  ylab("Mean absolute difference in expectation")

### 2016-02-20 presentations/coda2016/coda2016_poster.tex
left side complete

@@
 \usepackage{tikz-3dplot}
-\usetikzlibrary{arrows, shapes, positioning, backgrounds}
+\usetikzlibrary{arrows, shapes, positioning, backgrounds, decorations.pathreplacing}
 
@@
 \definecolor{isured}{RGB}{206,17,38}
+\definecolor{isugreen}{RGB}{7,109,84}
 \counterwithout{figure}{chapter}
@@
   \caption{An example restricted Boltzmann machine (RBM), which consists of two layers, a hidden ($\mathcal{H}$) and a visible layer ($\mathcal{V}$), with no connections within a layer.
-Hidden nodes indicated by gray circles and the visible nodes indicated by white circles \cite{friedman2001elements}.}
+Hidden nodes indicated by white circles and the visible nodes indicated by blue circles \cite{friedman2001elements}.}
   \label{fig:rbm}
@@
 
+\vspace{1cm}
 \begin{figure}[ht]
@@
   \input{images/visibles.tikz}
-  \caption{Visibles diagram.}
+  \caption{Image classification using a RBM.
+On the left, each image pixel comprises a node in the visible layer, $\mathcal{V}$.
+On the right, the output of the RBM is used to create features which are then passed to a supervised learning algorithm.}
   \label{fig:visibles}
@@
 Let $\boldsymbol x = \{h_1, \dots, h_H, v_1,\dots,v_V\}$ represent the states of the visible and hidden nodes in an RBM.
-Then  the probability corresponding to the state of each node taking the value of $1$:
+Then the probability each node taking the the value corresponding to $\boldsymbol x$ is:
 \begin{align}
@@
 \begin{textblock}{14}(.65, 11.33)
-{\large \bfseries Degeneracy, instability, and uninterpretability... Oh my!}
-~\\[.25cm]
+{\large \bfseries Degeneracy, instability, and uninterpretability... Oh my!}\\[-1cm]
+
+The highly flexible nature of the RBM ($H + V + HV$ parameters) makes the following characteristics of model impropriety of particular concern. \\[-.5cm]
+
 \begin{table}
@@
 \label{tab:degen}
-\caption{Table of ``bad" model characteristics.}
+\caption{Table of ``improper model'' characteristics.}
 \end{table}
@@
 %%
-\begin{textblock}{14}(.65, 22.94)
-{\large \bfseries Variable encoding}
+\begin{textblock}{14}(.65, 21.5)
+{\large \bfseries Avoiding degeneracy} \\[-.75cm]
+
+For the $\{-1, 1 \}$ encoding of $\mathcal{V}$ and $\mathcal{H}$, the origin is the center of the parameter space.
+In particular, at $\boldsymbol \theta = \boldsymbol 0$, the RBM is equivalent to elements of $\boldsymbol X$ being distributed as iid Bernoulli$\left(\frac{1}{2}\right)$ r.v.s.
+$\Rightarrow$ No \emph{near-degeneracy}, \emph{instability}, or \emph{uninterpretability}!
+\end{textblock}
+\begin{textblock}{14.5}(.65, 23.9)
+\begin{minipage}{.42\textwidth}
+\begin{figure}[ht]
+  \begin{minipage}{0.49\textwidth}
+  \resizebox{\linewidth}{!}{
+    \tdplotsetmaincoords{60}{-60}
+    \input{images/toyhull_top.tikz}
+
+    \draw(0,1,0) -- (1,0,0) -- (0,0,0) -- (0,1,0); 
+     \draw(1,1,1) -- (1,0,0) -- (0,0,0) -- (1,1,1); 
+     \draw(1,1,1) -- (0,1,0) -- (0,0,0) -- (1,1,1); 
+     \draw(1,1,1) -- (0,1,0) -- (1,0,0) -- (1,1,1); 
+    \end{tikzpicture}
+  }
+  \end{minipage}
+  \begin{minipage}{0.49\textwidth}
+  \resizebox{\linewidth}{!}{
+    \tdplotsetmaincoords{60}{-60}
+    \input{images/toyhull_top_2.tikz}
+
+    \draw(-1,1,-1) -- (1,-1,-1) -- (-1,-1,-1) -- (-1,1,-1); 
+     \draw(1,1,1) -- (1,-1,-1) -- (-1,-1,-1) -- (1,1,1); 
+     \draw(1,1,1) -- (-1,1,-1) -- (-1,-1,-1) -- (1,1,1); 
+     \draw(1,1,1) -- (-1,1,-1) -- (1,-1,-1) -- (1,1,1); 
+    \end{tikzpicture}
+  }
+  \end{minipage}
+  \caption{The convex hulls of the statistic space in three dimensions for a toy RBM with $|\mathcal{V}| = |\mathcal{H}| = 1$ for $\{0,1\}$-encoding (left) and $\{-1,1\}$-encoding (right) enclosed by an unrestricted hull of 3-space.}
+\label{fig:toyhull}
+\end{figure}
+\end{minipage}
+\begin{minipage}{.02\textwidth}
+\hfill
+\end{minipage}
+\begin{minipage}{.55\textwidth}
+\begin{figure}
+\centering
+\includegraphics[scale=1]{images/frac_volume.pdf}
+\label{fig:frac_volume}
+\caption{Volume relationship for the convex hulls of statistics in $Q(\cdot)$ vs. unrestricted space.}
+\end{figure}
+\end{minipage}
 
@@
 \begin{textblock}{14}(32.4, 5.71)
-{\large \bfseries Managable examples} \\[.25cm]
+{\large \bfseries Manageable (a.k.a. small) examples} \\[.25cm]
+
+As the magnitude of $\boldsymbol \theta$ grows, so does the occurence of near-degeneracy, instability, and uninterpretability for RBMs of varying sizes.
+
 \end{textblock}
@@
 
-\begin{textblock}{14.54}(31.86, 24.6)
+\begin{textblock}{14.54}(31.86, 24.8)
 These models very easily are degenerate, unstable, and uninterpretable. Meaning, the space of possible fitted parameter values that leads to a proper model is highly restricted.
@@
 %%
-\begin{textblock}{14.5}(16.17, 24.84)
+\begin{textblock}{14.5}(16.17, 25.45)
 {\large \bfseries References}

### 2016-02-20 writing/degeneracy/draft.Rmd
left side complete

@@
       in_header: header.tex
+    keep_tex: yes
     number_sections: yes

### 2016-02-20 presentations/coda2016/images/ggplots.R

@@ -1,0 +1,111 @@
+library(knitr)
+library(ggplot2)
+library(dplyr)
+library(tidyr)
+
+opts_chunk$set(echo=FALSE, message=FALSE, warnings=FALSE)
+theme_set(theme_bw(base_family="serif"))
+
+source("../../degeneracy/functions.R")
+
+##variable encoding ----------------------
+test_cases <- expand.grid(data.frame(1:4)[,rep(1,2)]) %>% 
+  rename(H = X1.4, V = X1.4.1) %>%
+  filter(H <= V) %>% #remove those cases with less visibles than hiddens. Is this necessary?
+  mutate(n_param = H*V + H + V) %>%
+  mutate(max_facets = (2^(H+V))^(floor(n_param/2))) %>%
+  filter(n_param <= 11) #calc_hull can't handle any higher dimensions currently
+
+bin_res <- plyr::dlply(test_cases, plyr::.(n_param), function(x) calc_hull(x$V, x$H, "binary"))
+neg_res <- plyr::dlply(test_cases, plyr::.(n_param), function(x) calc_hull(x$V, x$H, "negative"))
+
+plyr::ldply(bin_res, function(x) x$c_hull$vol) %>% 
+  mutate(frac_vol = V1/(1^n_param)) %>%
+  inner_join(plyr::ldply(neg_res, function(x) x$c_hull$vol) %>% 
+               mutate(frac_vol = V1/(2^n_param)),
+             by="n_param") %>%
+  rename(vol.bin = V1.x, vol.neg = V1.y, frac_vol.bin = frac_vol.x, frac_vol.neg = frac_vol.y) %>%
+  gather(vars, value, -n_param) %>%
+  separate(vars, c("type", "encoding"), "\\.") %>%
+  spread(type, value) %>%
+  ggplot() +
+  geom_point(aes(x=n_param, y=frac_vol, colour=encoding)) +
+  geom_line(aes(x=n_param, y=frac_vol, colour=encoding, group=encoding)) +
+  ylab("Fraction of unrestricted volume") +
+  xlab("Number of parameters") +
+  scale_colour_discrete("Encoding", labels=c("Binary (1,0)", "Negative (1,-1)")) +           
+  theme(
+    plot.background = element_rect(fill = "transparent", colour = NA),
+    legend.background = element_rect(fill = "transparent", colour = NA)) -> p
+
[…]

### 2016-02-21 presentations/coda2016/coda2016_poster.tex
added manageable examples

@@
 \begin{textblock}{14}(32.4, 5.71)
-{\large \bfseries Manageable (a.k.a. small) examples} \\[.25cm]
-
-As the magnitude of $\boldsymbol \theta$ grows, so does the occurence of near-degeneracy, instability, and uninterpretability for RBMs of varying sizes.
-
+{\large \bfseries Manageable (a.k.a. small) examples}
+\begin{figure}
+\centering
+\includegraphics{images/degeneracy.pdf}
+\includegraphics{images/instability.pdf}
+\includegraphics{images/uninterpretability.pdf}
+\label{fig:three_ways}
+\caption{As the magnitude of $\boldsymbol \theta$ grows, so does the occurence of near-degeneracy, instability, and uninterpretability for RBMs of varying sizes.}
+\end{figure}
 \end{textblock}
@@
 %%
-\begin{textblock}{14}(32.4, 12.91)
+\begin{textblock}{14}(32.4, 14)
 {\large \bfseries Model fitting} \\[.25cm]

### 2016-02-21 presentations/coda2016/images/ggplots.R

@@ -46,7 +46,7 @@ ggsave("images/frac_volume.pdf",
        units = "in")
 
 #manageable examples ---------
-#reshape data function
+#reshape data functions
 plot_data <- function(res, grid = FALSE) {
   
   plot.data <- data.frame()
@@ -100,12 +100,118 @@ plot_data <- function(res, grid = FALSE) {
   
   return(plot.data)
 }
+indep_params <- function(samp, H, V) {
+  samp[, (H + V + 1):ncol(samp)] <- 0
+  samp
+}
+max_Q <- function(theta, stats) {
+  apply(crossprod(t(stats), theta), 2, max)
+}
 
+#grid data
 load("../../writing/data/results_grid.RData")
+
+#near-degeneracy
 plot_dat_grid <- res %>% plot_data(grid = TRUE)
 
+plot_dat_grid %>%
+  group_by(r1, r2, H, V) %>%
+  summarise(frac_degen = sum(near_hull)/n(), count = n()) %>% 
+  ungroup() %>% 
+  mutate(Hiddens = paste0("Hiddens: ", H), Visibles = paste0("Visibles: ", V)) -> convex_hull_summary
+
+#uninterpretability
 res %>%
   group_by(H, V, N, r1, r2) %>%
   do(indep_exp = t(expected_value(t(indep_params(.$samp[[1]], .$H, .$V)), .$stat[[1]]))[, -((.$H + .$V + 1):(.$H + .$V + .$H*.$V))],
      marg_exp = .$g_theta[[1]][, (.$H + .$V + .$H*.$V + 1):(ncol(.$g_theta[[1]])-.$H*.$V)]) -> exp_vals
 
+exp_vals %>%
+  group_by(H, V, N, r1, r2) %>%
+  do(data.frame(mean_abs_diff = apply(abs(.$indep_exp[[1]] - .$marg_exp[[1]]), 1, mean))) %>%
+  group_by(H, V, N, r1, r2) %>%
+  summarise(mean_abs_diff = mean(mean_abs_diff)) %>%
+  mutate(Hiddens = paste0("Hiddens: ", H), Visibles = paste0("Visibles: ", V)) -> exp_vals_summary
+
+#instability
+res %>%
+  group_by(H, V, N, r1, r2) %>%
+  do(data.frame(LHS = max_Q(t(.$samp[[1]]), .$stat[[1]])/(.$H + .$V))) %>%
[…]

### 2016-02-22 presentations/coda2016/coda2016_poster.tex
first draft complete

@@
 %%
-\begin{textblock}{14}(32.4, 14)
-{\large \bfseries Model fitting} \\[.25cm]
-Fitting is really hard to get right (Bayesian or ML).
-If you are able to fit this model properly, the result is the empirical distribution of images.
-You spent a lot of time just to get the empirical distribution back.
-If you do not have an instance of every possible image in your training set, certain parameter must $\rightarrow \infty$, making fitting more complicated (shrink)
-In reality it's practically impossible to have at least one of every possible image in your training set (large images, color scales, etc.)
+\begin{textblock}{14}(32.4, 13.8)
+{\large \bfseries Bayesian model fitting} \\
+{\bfseries Idea:} To avoid model impropriety, avoid parts of the parameter space that lead to \emph{near-degeneracy}, \emph{instability}, and \emph{uninterpretability} (i.e., shrink $\boldsymbol \theta$).
+\\[-.75cm]
 
+Simulated $n = 5,000$ images (4 pixels) from RBM model with 4 hiddens then fit using Bayesian methods, 
+\vspace{-.5cm}
+\begin{itemize}
+\setlength\itemsep{.05cm}
+\item {\bfseries Trick prior.} Cancel out the normalizing term, resulting full conditionals are normally distributed. {\bfseries Conclusion:} Scalable solution, but requires tuning.
+\vspace{-.75cm}
+$$\pi(\boldsymbol \theta) \propto \gamma(\boldsymbol \theta)^n \exp\left(-\frac{1}{2C_{1}}\boldsymbol \theta_{main}'\boldsymbol \theta_{main} -\frac{1}{2C_{2}}\boldsymbol \theta_{interaction}'\boldsymbol \theta_{interaction}\right), \vspace{-.75cm}
+$$ 
+where $\gamma(\boldsymbol \theta) = \Nsum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\nsum\limits_{i = 1}^V \nsum\limits_{j=1}^H \theta_{ij} v_i h_j + \nsum\limits_{i = 1}^V\theta_{v_i} v_i + \nsum\limits_{j = 1}^H\theta_{h_j} h_j\right)$ and $C_{2} < C_{1}$ \cite{li2014biclustering}.
+\item{\bfseries Truncated Normal prior.} Use two independent truncated spherical normal distributions as priors for $\theta_{main}$ and $\theta_{interaction}$ with $\sigma_{interaction} < \sigma_{main}$.
+Full conditional distributions are not conjugate, requires a geometric adaptive MH step \cite{zhou2014some} and calculation of likelihood normalizing constant.
+\\{\bfseries Conclusion:} Computationally intensive and convergence issues.
+\item{\bfseries Marginalized likelihood.} Marginalize out $\boldsymbol h$ in $f_{\boldsymbol \theta}(\boldsymbol x)$, and use the truncated Normal prior.
+{\bfseries Conclusion:} Least scalable, but removes need to gain MCMC convergence for $Hn$ sampled hidden nodes.
+\vspace{-.75cm}
+$$
+g_{\boldsymbol \theta}(\boldsymbol v) = \Nsum\limits_{\boldsymbol h \in \{-1,1\}^H} \exp\left(\nsum\limits_{i = 1}^V \nsum\limits_{j=1}^H \theta_{ij} v_i h_j + \nsum\limits_{i = 1}^V\theta_{v_i} v_i + \nsum\limits_{j = 1}^H\theta_{h_j} h_j\right).
+\vspace{-.75cm}
+$$
+\end{itemize}
+
+\end{textblock}
+
+\begin{textblock}{14.54}(31.86, 25.5)
+\begin{figure}
+\centering
+\includegraphics{images/image_prediction.pdf}
+\label{fig:image_pred}
+\vspace*{-15mm}
+\caption{Posterior probability of each possible 4-pixel image using priors above.}
+\end{figure}
+
+{\bfseries Big takeaway:} RBMs very easily are degenerate, unstable, and uninterpretable.
+To further complicate things, a rigorous fitting method for these models is not scalable and replicates the nonparametric solution (empirical distribution).
 \end{textblock}
@@
 %%
-\begin{textblock}{14}(32.4, 23.85)
-{\large \bfseries Discussion} \\[.25cm]
-\end{textblock}
+%%\begin{textblock}{14}(32.4, 23.85)
+%%{\large \bfseries Discussion} \\[.25cm]
+%%\end{textblock}
 
-\begin{textblock}{14.54}(31.86, 24.8)
-These models very easily are degenerate, unstable, and uninterpretable. Meaning, the space of possible fitted parameter values that leads to a proper model is highly restricted.
+%%\begin{textblock}{14.54}(31.86, 24.8)
+%%These models very easily are degenerate, unstable, and uninterpretable. Meaning, the space of possible fitted parameter values that leads to a proper model is highly restricted.
 
-To further complicate things, the proper fitting these models is very intricate and easily leads to parameter values running off to $\infty$.
+%%To further complicate things, the proper fitting these models is very intricate and easily leads to parameter values running off to $\infty$.
 
-If you're thinking of using RBMs, or stacked RBMs in a deep architecture, don't. Instead, fit a less flexible and more sensible model.
-\end{textblock}
+%%If you're thinking of using RBMs, or stacked RBMs in a deep architecture, don't. Instead, fit a less flexible and more sensible model.
+
+%%If you do not have an instance of every possible image in your training set, certain parameter must $\rightarrow \infty$, making fitting more complicated (shrink)
+
+%%In reality it's practically impossible to have at least one of every possible image in your training set (large images, color scales, etc.)
+%%\end{textblock}
 

### 2016-02-22 presentations/coda2016/images/degeneracy_plots.R

@@ -168,7 +168,7 @@ three_ways %>%
   geom_contour(aes(x = r1, y = r2, z = mean_abs_diff), colour = "black", bins = 8) +
   #geom_contour(aes(x = r1, y = r2, z = mean_abs_diff), colour = "black", breaks = .05, size = 1.5) +
   geom_abline(aes(intercept = 0, slope = 1), alpha = .5, lty = 2) +
-  scale_fill_gradient(expression(paste0("|", E(bold(X), "|", bold(theta)) - E(bold(X), "|", plain(independence)), "|")), low = "yellow", high = "red") +
+  scale_fill_gradient(expression(group("|", E(group("", bold(X), "|")*bold(theta)) - E(group("", bold(X), "|")*plain(independence)), "|")), low = "yellow", high = "red") +
   facet_grid(Visibles~Hiddens) +
   xlab(expression(group("||", theta[main], "||"))) +
   ylab(expression(group("||", theta[interaction], "||"))) +

### 2016-02-29 degeneracy/apps/degen_magnitude.Rmd
apps and restructure

@@
 title: "Degeneracy and parameter magnitude"
-author: "Andee Kaplan"
-date: "March 1, 2015"
 output:
@@
 library(shiny)
+source("functions.R")
 ```
@@
 
-plot_data <- function(res) {
-
-  plot.data <- data.frame()
-  for(i in 1:nrow(res)) {
-    tmp <- res$samp[[i]]
-    H <- res[i,]$H
-    V <- res[i,]$V
-
-    tmp %>% 
-      data.frame() %>% 
-      rowwise() %>% 
-      mutate_(ss_ratio = paste0("(", paste(paste0(names(.)[(H + V + 1):(H + V + H*V)], "^2"), collapse = " + "), ")/sum(", paste(paste0(names(.)[1:(H+V)],"^2"), collapse = " + "), ")")) %>%    
-      ungroup() -> ratio
-
-    inner_join(ratio, res$outside[[i]] %>% ungroup()) %>%
-      select(ss_ratio, near_hull) %>%
-      mutate(H = H, V = V, n_param = H + V + H*V) %>%
-      rbind(plot.data) -> plot.data
-  }
-  return(plot.data)
-}
-
-plot.dat <- list()
-for(i in seq(0, 1, by = 0.1)) {
-  load(paste0("results_", i, ".RData")) 
-  plot_dat <- lapply(res, plot_data)
-  names(plot_dat) <- gsub("[.]", "_", r_center)
-  plot.dat[[i*10 + 1]] <- plot_dat
-}
-
-plot.dat_split <- list()
-for(i in seq(0, 1, by = 0.1)) { #full data not compiled yet
-  load(paste0("results_split_", i, ".RData")) 
-  res <- lapply(res, function(x) x %>% mutate(r = r2)) #dumb naming convention
-  plot_dat_split <- lapply(res, plot_data)
-  names(plot_dat_split) <- gsub("[.]", "_", r_center)
-  plot.dat_split[[i*10 + 1]] <- plot_dat_split
-}
-
+load("plot_dat.RData")
 ```
@@
   add_legend("fill", title = "Within .05 of Hull")
-
-# ggvis.subset %>%
-#   ggvis(~n_param_f, ~ss_ratio) %>%
-#   layer_points(fill = ~as.character(near_hull), opacity := 0.4, size := 200) %>%
-#   add_axis("x", title = "# Parameters") %>%
-#   add_axis("y", title = "Ratio of SS of cross terms to main effects") %>%
-#   add_legend("fill", title = "Within 0.05 of Hull")
 

### 2016-02-29 presentations/coda2016/coda2016_poster.tex
apps and restructure

@@
 {\bfseries Near-degeneracy.} Occurs when there is a disproportionate amount of probability placed on only a few elements of the sample space by the model \cite{handcock2003assessing}.
-& If the mean parametrization on the model parameters, $\mu(\boldsymbol \theta)$, is close to the boundary of the convex hull of the set of statistics in the neg-potential function $Q(\boldsymbol x)$.
+& If random variables in the neg-potential function $Q(\cdot)$, having support set $\mathcal{S}$, have a collective mean $\mu(\boldsymbol \theta)$ close to the boundary of the convex hull of $\mathcal{S}$.
+\\
+%If the mean parametrization on the model parameters, $\mu(\boldsymbol \theta)$, is close to the boundary of the convex hull of the set of statistics in the neg-potential function $Q(\boldsymbol x)$.
 \\
 \hline
-{\bfseries Instability.} Small changes in natural parameters result in large changes of the pmf, excessive sensitivity \cite{schweinberger2011instability}.
+{\bfseries Instability.} Small changes in natural parameters result in large changes in probability masses, excessive sensitivity \cite{schweinberger2011instability}.
 & If for any $C > 0$ there exists $N_C > 0$ such that
-$\max\limits_{\boldsymbol x_N \in \mathcal{X}_N}[Q(\boldsymbol x_N)] > CN$ for all $N > N_C$. Where $Q(\cdot)$ is the neg-potential function of the model. \\
+$\max\limits_{\boldsymbol x_N \in \mathcal{X}_N}[Q(\boldsymbol x_N)] > CN$ for all $N > N_C$, where $Q(\cdot)$ is the neg-potential function of the model. \\
 \hline
 {\bfseries Uninterpretability.} Due to the existence of dependence, marginal mean-structure no longer maintained \cite{kaiser2007statistical}.
-& If the magnitude of the difference between model expectations and expectations under independence, $\left|\text{E}(\boldsymbol X | \boldsymbol \theta) - \text{E}(\boldsymbol X | \emptyset ) \right|$, is large.\\
+& If the magnitude of the difference between model expectations and expectations under independence (dependence parameters of zero), $\left|\text{E}(\boldsymbol X | \boldsymbol \theta) - \text{E}(\boldsymbol X | \emptyset ) \right|$, is large.\\
 \hline
@@
 \begin{textblock}{14}(.65, 21.5)
-{\large \bfseries Avoiding degeneracy} \\[-.75cm]
+{\large \bfseries Data coding to mitigate degeneracy} \\[-.75cm]
 
 For the $\{-1, 1 \}$ encoding of $\mathcal{V}$ and $\mathcal{H}$, the origin is the center of the parameter space.
-In particular, at $\boldsymbol \theta = \boldsymbol 0$, the RBM is equivalent to elements of $\boldsymbol X$ being distributed as iid Bernoulli$\left(\frac{1}{2}\right)$ r.v.s.
-$\Rightarrow$ No \emph{near-degeneracy}, \emph{instability}, or \emph{uninterpretability}!
+In particular, at $\boldsymbol \theta = \boldsymbol 0$, the RBM is equivalent to elements of $\boldsymbol X$ being distributed as iid Bernoulli$\left(\frac{1}{2}\right)$ $\Rightarrow$ No \emph{near-degeneracy}, \emph{instability}, or \emph{uninterpretability}!
 \end{textblock}
@@
 \label{fig:three_ways}
-\caption{As the magnitude of $\boldsymbol \theta$ grows, so does the occurence of near-degeneracy, instability, and uninterpretability for RBMs of varying sizes.}
+\caption{As the magnitude of $\boldsymbol \theta$ grows ($\boldsymbol \theta$ moves from $\boldsymbol 0$), so does the occurence of near-degeneracy, instability, and uninterpretability for RBMs of varying sizes.}
 \end{figure}
@@
 {\bfseries Big takeaway:} RBMs very easily are degenerate, unstable, and uninterpretable.
-To further complicate things, a rigorous fitting method for these models is not scalable and replicates the nonparametric solution (empirical distribution).
+As compounding issues, a rigorous fitting method for these models is not scalable and merely replicates the empirical data distribution, a.k.a optimal nonparametric solution.
+%only replicates the nonparametric solution (empirical distribution).
 \end{textblock}

### 2016-02-29 degeneracy/apps/functions.R

@@ -37,6 +37,7 @@ sample_sphere <- function(stat, n, r = 1) {
     group_by(samp) %>%
     mutate(vals = samp(ncol(stat), r)) %>%
     separate(col = vals, sep = ",", into = colnames(stat)) %>%
+    ungroup() %>%
     select(-samp) %>%
     data.matrix() %>%
     t()

### 2016-02-29 degeneracy/closest_points.R

@@ -51,4 +51,3 @@ segments(segs[,1], segs[,2], segs[,3], segs[,4], lty=2)
 
 ## Plot of min_dist by radius
 qplot(r, min_dist, colour = min_dist < 1e-4, data=min_points) + geom_hline(aes(yintercept=1e-4))
-

### 2016-02-29 degeneracy/kaiser_method.R

@@ -4,19 +4,19 @@ library(tidyr)
 source("functions.R")
 
 model_data <- function(res) {
-  
+
   model.data <- list()
   for(i in 1:nrow(res)) {
     tmp <- res$samp[[i]]
     H <- res[i,]$H
     V <- res[i,]$V
-    
-    tmp %>% 
-      data.frame() %>% 
-      rowwise() %>% 
-      mutate_(ss_ratio = paste0("(", paste(paste0(names(.)[(H + V + 1):(H + V + H*V)], "^2"), collapse = " + "), ")/sum(", paste(paste0(names(.)[1:(H+V)],"^2"), collapse = " + "), ")")) %>%    
+
+    tmp %>%
+      data.frame() %>%
+      rowwise() %>%
+      mutate_(ss_ratio = paste0("(", paste(paste0(names(.)[(H + V + 1):(H + V + H*V)], "^2"), collapse = " + "), ")/sum(", paste(paste0(names(.)[1:(H+V)],"^2"), collapse = " + "), ")")) %>%
       ungroup() -> ratio
-    
+
     inner_join(ratio, res$outside[[i]] %>% ungroup()) %>%
       mutate(H = H, V = V, n_param = H + V + H*V) -> model.data[[i]]
   }
@@ -32,10 +32,10 @@ set.seed(10-22-1985) #best day ever!
 
 model.dat <- list()
 for(i in seq(0, 1, by = 0.1)) {
-  load(paste0("apps/results_", i, ".RData")) 
+  load(paste0("written/results_", i, ".RData"))
   model_dat <- lapply(res, model_data)
   names(model_dat) <- r_center
-  
+
   model.dat[[i*10 + 1]] <- model_dat
 }
 names(model.dat) <- seq(0, 1, by = 0.1)
@@ -47,10 +47,10 @@ expand.grid(power = as.numeric(names(model.dat)), multiplier = as.numeric(names(
 
 frame %>%
   group_by(power, multiplier, n_param) %>%
-  mutate(frac_degen = sum(models[[1]]$near_hull)/nrow(models[[1]]), 
+  mutate(frac_degen = sum(models[[1]]$near_hull)/nrow(models[[1]]),
          radius = (models[[1]]$H[1] + models[[1]]$V[1])^power,
          happy = frac_degen < .05) %>%
-  group_by(n_param, happy) %>% 
+  group_by(n_param, happy) %>%
   sample_n(20) -> models.samp
 
 
@@ -68,7 +68,7 @@ no_int %>%
[…]

### 2016-02-29 degeneracy/kaiser_method_grid.R

@@ -6,7 +6,7 @@ source("functions.R")
 
 #load sample params ---------------------------------------------
 n <- 100
-load("apps/results_grid.RData") 
+load("written/results_grid.RData")
 
 indep_params <- function(samp, H, V) {
   samp[, (H + V + 1):ncol(samp)] <- 0
@@ -35,7 +35,7 @@ exp_vals %>%
   ylab(expression(group("||", theta[interaction], "||"))) +
   theme(aspect.ratio = 1)
 
-exp_vals %>% 
+exp_vals %>%
   mutate(dist = sqrt(r1^2 + r2^2)) %>%
   group_by(dist, H, V) %>%
   do(data.frame(mean_abs_diff = mean(abs(.$indep_exp[[1]] - .$marg_exp[[1]])))) %>%

### 2016-02-29 degeneracy/radius_percent_degen_max_no_degen_plots_split.R

@@ -3,25 +3,25 @@ library(dplyr)
 library(tidyr)
 
 plot_data <- function(res) {
-  
+
   plot.data <- data.frame()
   for(i in 1:nrow(res)) {
     tmp <- res$samp[[i]]
     H <- res[i,]$H
     V <- res[i,]$V
-    
-    tmp %>% 
-      data.frame() %>% 
+
+    tmp %>%
+      data.frame() %>%
       rowwise() %>%
-      mutate_(ss_ratio = paste0("(", paste(paste0(names(.)[(H + V + 1):(H + V + H*V)], "^2"), collapse = " + "), ")/sum(", paste(paste0(names(.)[1:(H+V)],"^2"), collapse = " + "), ")")) %>%    
+      mutate_(ss_ratio = paste0("(", paste(paste0(names(.)[(H + V + 1):(H + V + H*V)], "^2"), collapse = " + "), ")/sum(", paste(paste0(names(.)[1:(H+V)],"^2"), collapse = " + "), ")")) %>%
       ungroup() -> ratio
-    
+
     tmp %>%
       data.frame() %>%
       select(-starts_with("exp")) %>%
       rowwise() %>%
       mutate_(true_r = paste0("sqrt(", paste(paste0(names(.), "^2"), collapse = " + "), ")")) -> radius
-      
+
     inner_join(radius, ratio) %>%
     inner_join(res$outside[[i]] %>% ungroup()) %>%
       select(ss_ratio, near_hull, true_r) %>%
@@ -38,24 +38,24 @@ r_exp <- .05
 
 plot.dat <- list()
 for(i in seq(0, 1, by = 0.1)) {
-  load(paste0("apps/results_split_", i, ".RData")) 
+  load(paste0("written/results_split_", i, ".RData"))
   plot_dat <- lapply(res, plot_data)
   names(plot_dat) <- gsub("[.]", "_", r_center)
-  
+
   do.call(rbind, plot_dat) %>%
     ungroup() %>%
     mutate(name = rownames(do.call(rbind, plot_dat))) %>%
-    separate(name, into = c("multiplier", "num"), "[.]") %>% 
+    separate(name, into = c("multiplier", "num"), "[.]") %>%
     mutate(multiplier = as.numeric(gsub("_", ".", multiplier))) %>%
     mutate(n_param_f = factor(n_param)) -> plot_dat
-  
+
   plot.dat[[i*10 + 1]] <- plot_dat
 }
 
[…]

### 2016-02-29 degeneracy/radius_percent_degen_split.R

@@ -12,11 +12,11 @@ find_prop <- function(g_theta, stat, r_exp) {
     do(samp = sample_points_outside(as.numeric(.), stat, r_exp)$in_hull) %>%
     group_by_(.dots = c(colnames(g_theta))) %>%
     do(as.data.frame(sum(.$samp[[1]]) > 0)) -> is_outside # count of not in hull greater than zero => there exists a point outside hull for this model
-  
-  is_outside %>% 
+
+  is_outside %>%
     rename_(near_hull = as.name(names(is_outside)[ncol(is_outside)])) %>%
     mutate(r = r_exp) -> inside_outside
-  
+
   return(inside_outside)
 }
 
@@ -27,9 +27,9 @@ sample_points_outside <- function(g_theta, stats, r, n = 100) {
     mutate(vals = samp_exp(length(g_theta), g_theta, r)) %>%
     separate(col = vals, sep = ",", into = colnames(stats)) %>%
     select(-samp)
-  
+
   x0$in_hull <- !apply(x0, 1, in_hull, stats)
-  
+
   return(x0)
 }
 
@@ -42,24 +42,24 @@ in_hull <- function(point, hull_points) {
   require(lpSolveAPI)
   P <- data.matrix(point)
   A <- t(data.matrix(hull_points))
-  
+
   lp_obj <- make.lp(nrow = 0, ncol = ncol(A))
   sapply(1:nrow(A), function(x) add.constraint(lp_obj, A[x,], type = "=", rhs = P[x]))
   add.constraint(lp_obj, rep(1, ncol(A)), type = "=", rhs = 1)
   sapply(1:ncol(A), function(x) add.constraint(lp_obj, 1, type = ">=", rhs = 0, indices = x))
-  
+
   return(solve(lp_obj) == 0)
 }
 
 sample_sphere_main <- function(stat, n, r = 1) {
   require(dplyr)
   require(tidyr)
-  
+
   samp <- function(space, r) {
     vec <- rnorm(space)
     paste0(r*vec/sqrt(sum(vec^2)), collapse=",")
   }
-  
+
   res <- data.frame(samp = 1:n) %>%
     group_by(samp) %>%
     mutate(vals = samp(ncol(stat %>% data.frame() %>% select(-contains("theta"))), r)) %>%
@@ -67,19 +67,19 @@ sample_sphere_main <- function(stat, n, r = 1) {
     select(-samp) %>%
     data.matrix() %>%
     t()
-  
[…]

### 2016-02-29 degeneracy/split_sample.R

@@ -12,11 +12,11 @@ find_prop <- function(g_theta, stat, r_exp) {
     do(samp = sample_points_outside(as.numeric(.), stat, r_exp)$in_hull) %>%
     group_by_(.dots = c(colnames(g_theta))) %>%
     do(as.data.frame(sum(.$samp[[1]]) > 0)) -> is_outside # count of not in hull greater than zero => there exists a point outside hull for this model
-  
-  is_outside %>% 
+
+  is_outside %>%
     rename_(near_hull = as.name(names(is_outside)[ncol(is_outside)])) %>%
     mutate(r = r_exp) -> inside_outside
-  
+
   return(inside_outside)
 }
 
@@ -27,9 +27,9 @@ sample_points_outside <- function(g_theta, stats, r, n = 100) {
     mutate(vals = samp_exp(length(g_theta), g_theta, r)) %>%
     separate(col = vals, sep = ",", into = colnames(stats)) %>%
     select(-samp)
-  
+
   x0$in_hull <- !apply(x0, 1, in_hull, stats)
-  
+
   return(x0)
 }
 
@@ -42,12 +42,12 @@ in_hull <- function(point, hull_points) {
   require(lpSolveAPI)
   P <- data.matrix(point)
   A <- t(data.matrix(hull_points))
-  
+
   lp_obj <- make.lp(nrow = 0, ncol = ncol(A))
   sapply(1:nrow(A), function(x) add.constraint(lp_obj, A[x,], type = "=", rhs = P[x]))
   add.constraint(lp_obj, rep(1, ncol(A)), type = "=", rhs = 1)
   sapply(1:ncol(A), function(x) add.constraint(lp_obj, 1, type = ">=", rhs = 0, indices = x))
-  
+
   return(solve(lp_obj) == 0)
 }
 
@@ -60,13 +60,13 @@ epsilon <- seq(-1, 1, length.out = 11)
 expand.grid(H = 1:4, V = 1:4, C = C, epsilon = epsilon) %>%
   mutate(N = H + V) %>%
   mutate(n = n, r1 = sqrt(C), r2 = sqrt(C*(N/(H*V))^(epsilon))) %>%
-  group_by(H, V, C, epsilon, r1, r2, N) %>% 
+  group_by(H, V, C, epsilon, r1, r2, N) %>%
[…]

### 2016-04-20 writing/degeneracy/draft.Rmd
restructured the degeneracy section

--- 2016-02-18 writing/degeneracy/draft.Rmd
@@
 ---
-title: "Degeneracy in a restricted Boltzmann machine"
+title: "An exposition on the propriety of restricted Boltzmann machines"
 author: "Andee Kaplan, Stephen Vardeman, Daniel Nordman"
-date: "January 8, 2016"
+date: "May, 2016"
 output: 
@@
 
-#Degeneracy in Random Graph Models
-In Random Graph Model theory, \emph{degeneracy} occurs when there is a disproportionate amount of probability placed on only a few elements of the sample space, $\mathcal{X}$, by the model.
-We consider three related definitions of degeneracy as they relate to a graphical model.
-Section \ref{mean-parameterization} presents a definition from @handcock2003assessing based on the mean parameterization of the model parameters, section \ref{neg-potential} presents a definition from @schweinberger2011instability based on the neg-potential function, and section \ref{conditional-expectation} presents a definition from @kaiser2006markov that involves the difference between expectations of the model and expectations under independence.
+#Degeneracy, instability, and uninterpretability... Oh my!
+The highly flexible nature of the RBM ($H + V + H V$ parameters) makes three characteristics of model impropriety of particular concern, *degeneracy, instability, and uninterpretability*.
+In this section we define these characteristics, discuss how to detect them in the RBM, and point out their relationship to one another.
 
-##Mean Parameterization
+##Near-degeneracy
+In Random Graph Model theory, *degeneracy* occurs when there is a disproportionate amount of probability placed on only a few elements of the sample space, $\mathcal{X}$, by the model.
 @handcock2003assessing define a model to be near degenerate if $\mu(\boldsymbol \theta)$, the mean parametrization on the model parameters, is close to the boundary of the convex hull of $\{t_k(\boldsymbol x): \boldsymbol x \in \mathcal{X}, k = 1, \dots, K\}$, where $t_k(\boldsymbol x), k = 1, \dots, K\}$ is the set of statistics in the neg-potential function $Q(\boldsymbol x)$ from equation \ref{eqn:pmf}.
+In other words, if random variables in the neg-potential function $Q(\cdot)$, having support set $\mathcal{S}$, have a collective mean $\mu(\boldsymbol \theta)$ close to the boundary of the convex hull of $\mathcal{S}$ then the model is near-degenerate, as defined by @handcock2003assessing.
 
@@
 
-##Neg-potential
-@schweinberger2011instability introduces a related concept, instability, which  is characterized
-by excessive sensitivity and near-degeneracy. He provides the following tool for detecting instability in the a discrete exponential family model, and thus near-degeneracy.
+##Instability
+@schweinberger2011instability introduces a related concept, *instability*, which occurs when small changes in the natural parameters result in large changes in probability masses.
+This can be thought of as excessive sensitivity in the model.
+@schweinberger2011instability provides the following tool for detecting instability in the a discrete exponential family model.
 
@@
 
-\begin{lemma}
-If a discrete exponential family distribution is unstable, then it is degenerate in the sense that for any $0 < \epsilon < 1$, $P_\theta(\boldsymbol X_N \in \mathcal{M}_{\epsilon, N}) \rightarrow 1$ as $N \rightarrow \infty$ where $\mathcal{M}_{\epsilon, N}$ is the subset of $\epsilon$-modes of the probability mass function.
-\end{lemma}
+Conversely, a discrete exponential family distribution is stable if there exists constants $C > 0$ and $N_C > 0$ such that 
+\begin{align}
+\max\limits_{\boldsymbol x_N \in \mathcal{X}_N}[Q(\boldsymbol x_N)] \le CN \text{ for all } N > N_C. \label{eq:stable}
+\end{align}
 
-In the context of the RBM, $N$ is the number of nodes in the model, $H + V$. So, the distribution is unstable if for any $C > 0$,
+Within the context of the RBM, $N$ is the number of nodes in the model, $H + V$.
+To ensure the model is stable, we must restrict $\boldsymbol \theta$ such that there exists constants $C > 0$ and $N_C > 0$ that force equation \ref{eq:stable} to hold.
 
@@
 &= \max\limits_{h_j, v_i \in \{0,1\}}\left[ \sum\limits_{i = 1}^H\sum\limits_{j=1}^V \theta_{ij} h_i v_j + \sum\limits_{i = 1}^H\theta_{hi} h_i + \sum\limits_{j = 1}^V\theta_{vj} v_j \right] \\
-&> C(H + V)
+&\le C(H + V)
 \end{align*}
@@
 There are $H+V$ main effects terms and $H*V$ interaction effects terms.
-For $H$ and $V$ large enough, the number of interaction terms is growing much faster than $H + V$, meaning no matter the $C$, we are able to find values of $\boldsymbol theta$ that can make the left side of equation \ref{eq:unstable} large compared to $C(H+V)$.
-This points to the interaction terms as potentially the more dangerous parameters for causing near-degeneracy in the RBM.
+For $H$ and $V$ large enough, the number of interaction terms is growing much faster than $H + V$, meaning no matter the $C$, we can restrict the size of the left side of equation \ref{eq:stable} effectively by restricting the size of the interaction terms.
+This points to the interaction terms as potentially the more dangerous parameters for causing instability in the RBM simply because there are so many more of them.
 
@@
 
-##Expectation Differences
-The characterization put forth by @kaiser2006markov of degeneracy in a model concerns the difference between model expectations and expectations given independence.
-Using this concept of dependence in a model, it is possible to understand what conditions lead to near-degeneracy in a model versus those that guarantee non-degenerate models.
-In this method, a model is considered without near-degeneracy if its expected value is not very different from the same model with assumed independence.
+Instability in a model is in itself a measure of model impropriety, however there is also a relationship between instability and near-degeneracy.
+
+\begin{lemma}[Schweinberger (2011)]
+If a discrete exponential family distribution is unstable, then it is degenerate in the sense that for any $0 < \epsilon < 1$, $P_\theta(\boldsymbol X_N \in \mathcal{M}_{\epsilon, N}) \rightarrow 1$ as $N \rightarrow \infty$ where $\mathcal{M}_{\epsilon, N}$ is the subset of $\epsilon$-modes of the probability mass function.
+\end{lemma}
+
+##Uninterpretability
+@kaiser2007statistical defines a measure of model impropriety called *uninterpretability*, which is characterized as due to the existence of dependence, marginal mean-structure no longer being maintained in the model.
+This measure concerns the difference between model expectations, E$\left[\boldsymbol X | \boldsymbol \theta\right]$, and expectations given independence, E$\left[\boldsymbol X | \emptyset\right ]$.
+Using this concept of dependence in a model, it is possible to understand what conditions lead to uninterpretability in a model versus those that guarantee interpretable models.
+In this method, a model is considered without uninterpretability if its expected value is not very different from the same model with assumed independence.
 The quantities to compare in the RBM case are
@@
 \begin{align*}
-\text{E}\left[\boldsymbol X\right] &= \sum\limits_{\boldsymbol x \in \mathcal{X}} \boldsymbol x f_{\boldsymbol \theta}(\boldsymbol x) \\
+\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] &= \sum\limits_{\boldsymbol x \in \mathcal{X}} \boldsymbol x f_{\boldsymbol \theta}(\boldsymbol x) \\
 &= \sum\limits_{\boldsymbol x \in \mathcal{X}} \boldsymbol x \frac{\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)} \\
@@
 \end{align*}
+and
+\begin{align*}
+\text{E}\left[\boldsymbol X | \emptyset \right] &= \sum\limits_{\boldsymbol x \in \mathcal{X}} \boldsymbol x \frac{\exp\left(\sum\limits_{i = 1}^V \theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)} \\
+&= \frac{\sum\limits_{\boldsymbol x \in \mathcal{X}} x_k \exp\left(\sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}  \qquad k = 1, \dots, H + V
+\end{align*}
 
 #Tiny example
-To illustrate the idea of graph degeneracy in a RBM, we will consider a the smallest possible toy example that consists of one visible node $v_1$ and one hidden node $h_1$, both binary.
+To illustrate the ideas of graph near-degeneracy, instability, and uninterpretability in a RBM, we will consider a the smallest possible toy example that consists of one visible node $v_1$ and one hidden node $h_1$, both binary.
 A schematic of this model can be found in figure \ref{fig:toymodel}.
@@
 
-##Degeneracy three ways
+##Impropriety three ways
 

### 2016-05-11 writing/cut_text.Rmd
stupid git

---
title: "Cut text for later"
author: "Andee Kaplan"
date: "April 20, 2016"
output: html_document
---

We consider three related definitions of degeneracy as they relate to a graphical model. Section \ref{mean-parameterization} presents a definition from @handcock2003assessing based on the mean parameterization of the model parameters, section \ref{neg-potential} presents a definition from @schweinberger2011instability based on the neg-potential function, and section \ref{conditional-expectation} presents a definition from @kaiser2006markov that involves the difference between expectations of the model and expectations under independence.

which is characterized by excessive sensitivity and near-degeneracy

it is evident that the fraction of near-degenerate models simulated at specified magnitudes of $\boldsymbol \theta_{main}$ and $\boldsymbol \theta_{interaction}$ is more closely related to the combined distance from the origin of these two vectors, than any individual group of parameter values.

```{r}
three_ways %>% 
  mutate(dist = sqrt(r1^2 + r2^2)) %>% 
  group_by(N, dist) %>% 
  summarise(mean_max_q = mean(mean_max_q)) %>% 
  ggplot() + 
  geom_line(aes(N, mean_max_q, colour = dist, group = dist))

three_ways %>% 
  ggplot() + 
  geom_point(aes(mean_abs_diff, frac_degen), alpha = .3) +
  geom_smooth(aes(mean_abs_diff, frac_degen), method = "lm") +
  facet_grid(Hiddens ~ Visibles)

three_ways %>% group_by(Hiddens, Visibles) %>% summarise(corr = cor(mean_abs_diff, frac_degen))

```

### 2016-05-11 writing/draft.Rmd
stupid git

--- 2016-04-20 writing/degeneracy/draft.Rmd
@@
       in_header: header.tex
-    keep_tex: yes
     number_sections: yes
-bibliography: ../references/refs.bib
+bibliography: references/refs.bib
 ---
@@
 
-source("../helpers/functions.R")
+source("helpers/functions.R")
 ```
 
-#Restricted Boltzmann Machines
+#Restricted Boltzmann machines
 
@@
 
-##Joint Distribution
+##Joint distribution
 Let $\boldsymbol x = \{h_1, \dots, h_H, v_1,\dots,v_V\}$ represent the states of the visible and hidden nodes in an RBM.
@@
 \begin{defn}
-A discrete exponential family distribution is unstable if for any $C > 0$, however large, there exists $N_C > 0$ such that 
+A discrete exponential family distribution is \emph{unstable} if for any $C > 0$, however large, there exists $N_C > 0$ such that 
 \begin{align}
@@
 \end{align}
-\end{defn}
 
-Conversely, a discrete exponential family distribution is stable if there exists constants $C > 0$ and $N_C > 0$ such that 
+Conversely, a discrete exponential family distribution is \emph{stable} if there exists constants $C > 0$ and $N_C > 0$ such that 
 \begin{align}
@@
 \end{align}
+\end{defn}
 
@@
 
+If $|\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] - \text{E}\left[\boldsymbol X | \emptyset\right]$ is large then the RBM is uninterpetable.
+
 #Tiny example
@@
 
-For this small model, we are able to illustrate the boundary of the convex hull of $\{t_k(\boldsymbol x): \boldsymbol x \in \mathcal{X}, k = 1, \dots, K\}$ because we are in three dimensions.
+For this small model, we are able to investigate the symptoms of the three measure of model impropriety, near-degeneracy, instability uninterpretability.
+First, we are able to illustrate the shape of boundary of the convex hull of $\{t_k(\boldsymbol x): \boldsymbol x \in \mathcal{X}, k = 1, \dots, K\}$ and explore the behavior of the mean parameterization of the parameters because we are in three dimensions.
 In figure \ref{fig:toyhull}, the convex hull of our statistic space in three dimensions for the toy RBM with one visible and one hidden node from multiple perspectives enclosed by an unrestricted hull of $\{0,1\}^3$ space is shown.
@@
 
-    ```{r, results='asis', echo=FALSE, include=TRUE, message=FALSE}
+    ```{r hull, results='asis', echo=FALSE, include=TRUE, message=FALSE}
     cat(sapply(1:nrow(m1_bin$c_hull$hull), function(x) paste0("\\draw", paste(paste0("(", apply(m1_bin$possible_t[cbind(m1_bin$c_hull$hull, m1_bin$c_hull$hull[,1])[x,],], 1, paste, collapse=","), ")"), collapse=" -- "), "; \n")))
@@
 
-```{r toymodel, fig.cap="\\label{fig:toymodel}The three mean-parameterization functions of the model parameters for a RBM with one visible and one hidden node.", out.height = "4in", fig.align='center'}
+```{r near_dependency, fig.cap="\\label{fig:toymodel}The three mean-parameterization functions of the model parameters for a RBM with one visible and one hidden node.", fig.height = 4, fig.align='center', message=FALSE, cache=TRUE}
 possibles <- stats(1, 1, type = "binary")
 
-expand.grid(theta_h1 = seq(-2, 2, length.out = 5),
-            theta_v1 = seq(-2, 2, length.out = 5),
-            theta_11 = seq(-2, 2, length.out = 11)) %>%
+expand.grid(theta_h1 = seq(-5, 5, length.out = 5),
+            theta_v1 = seq(-5, 5, length.out = 5),
+            theta_11 = seq(-5, 5, length.out = 11)) %>%
   data.frame() %>%
@@
 
-##Variable Encoding
+Secondly, we can look at $\max\limits_{\boldsymbol x_N \in \mathcal{X}_N}[Q(\boldsymbol x_N)]$ for this tiny model.
+
+\begin{align*}
+\max\limits_{\boldsymbol x_N \in \mathcal{X}_N}[Q(\boldsymbol x_N)] &= \max\limits_{\boldsymbol x \in \mathcal{X}}\left[ \theta_{11} h_1 v_1 + \theta_{h_1}h_1 + \theta_{v_1} v_1 \right] \\
+&=\max\left\{ \theta_{11} + \theta_{h_1} + \theta_{v_1}, \theta_{h_1}, \theta_{v_1}, 0  \right\} \\
+\end{align*}
+
+Figure \ref{fig:instab} shows $\max\limits_{\boldsymbol x_N \in \mathcal{X}_N}[Q(\boldsymbol x_N)]/N$ for various values of $\boldsymbol \theta$.
+Of course, we are interested in how this quantity grows as $N = H+V$ grows for various parameter values.
+We can see that this quantity is increasing for larger positive values of $\boldsymbol \theta$, which may indicate instability as $|\boldsymbol \theta|$ is large, agreeing with the indications about near-degeneracy from looking at $\mu(\boldsymbol \theta)$.
+
+```{r instab, fig.cap="\\label{fig:instab}$\\max\\limits_{\\boldsymbol x_N \\in \\mathcal{X}_N}[Q(\\boldsymbol x_N)]/N$ for various values of $\\boldsymbol \\theta$ for the tiny example model.
+This quantity is increasing for larger positive values of $\\boldsymbol \\theta$.", fig.height = 4, fig.align='center'}
+expand.grid(theta_h1 = seq(-5, 5, length.out = 5),
+            theta_v1 = seq(-5, 5, length.out = 5),
+            theta_11 = seq(-5, 5, length.out = 31)) %>%
+  data.frame() %>%
+  group_by(theta_v1, theta_h1, theta_11) %>%
+  do(data.frame(q_max = max(possibles %*% t(data.matrix(.))))) %>%
+  ungroup() %>%
+  mutate(theta_h1 = factor(theta_h1, labels = paste0("theta[h1]==", unique(theta_h1))), 
+         theta_v1 = factor(theta_v1, labels = paste0("theta[v1]==", unique(theta_v1)))) %>%
+  ggplot() +
+  geom_line(aes(theta_11, q_max/2)) +
+  facet_grid(theta_h1 ~ theta_v1, labeller = label_parsed) +
+  xlab(expression(theta[11])) +
+  ylab(expression(frac(max(Q), H + V)))
+```
+
+Finally, we can look at the difference between model expectations, E$\left[\boldsymbol X | \boldsymbol \theta\right]$, and expectations given independence, E$\left[\boldsymbol X | \emptyset\right ]$ for the tiny toy model to understand the effect of uninterpretability on the RBM.
+
+\begin{align*}
+\left|\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] -   \text{E}\left[\boldsymbol X | \emptyset \right] \right|
+&= \left|\left[
+\begin{matrix}
+\frac{\exp\left(\theta_{v_1}\right) + \exp\left(\theta_{11} + \theta_{v_1} + \theta_{h_1}\right)}{\exp\left(\theta_{v_1}\right) + \exp\left(\theta_{h_1}\right) + \exp\left(\theta_{11} + \theta_{v_1} + \theta_{h_1}\right)} \\
+\frac{\exp\left(\theta_{h_1}\right) + \exp\left(\theta_{11} + \theta_{v_1} + \theta_{h_1}\right)}{\exp\left(\theta_{v_1}\right) + \exp\left(\theta_{h_1}\right) + \exp\left(\theta_{11} + \theta_{v_1} + \theta_{h_1}\right)}
+\end{matrix} \right] - \left[
+\begin{matrix}
+\frac{\exp\left(\theta_{v_1}\right) + \exp\left(\theta_{v_1} + \theta_{h_1}\right)}{\exp\left(\theta_{v_1}\right) + \exp\left(\theta_{h_1}\right) + \exp\left(\theta_{v_1} + \theta_{h_1}\right)} \\
+\frac{\exp\left(\theta_{h_1}\right) + \exp\left(+ \theta_{v_1} + \theta_{h_1}\right)}{\exp\left(\theta_{v_1}\right) + \exp\left(\theta_{h_1}\right) + \exp\left(+ \theta_{v_1} + \theta_{h_1}\right)} 
+\end{matrix} \right] \right|\\
+&= 
+\begin{matrix}
+\frac{\left|\exp\left(\theta_{11} + \theta_{v_1} + 2\theta_{h_1}\right) - \exp\left( \theta_{v_1} + 2\theta_{h_1}\right) \right| }{\left(\exp\left(\theta_{v_1}\right) + \exp\left(\theta_{h_1}\right) + \exp\left(\theta_{11} + \theta_{v_1} + \theta_{h_1}\right)\right)\left(\exp\left(\theta_{v_1}\right) + \exp\left(\theta_{h_1}\right) + \exp\left(+ \theta_{v_1} + \theta_{h_1}\right)\right)} \\
+\frac{\left|\exp\left(\theta_{11} + 2\theta_{v_1} + \theta_{h_1}\right) - \exp\left( 2\theta_{v_1} + \theta_{h_1}\right) \right| }{\left(\exp\left(\theta_{v_1}\right) + \exp\left(\theta_{h_1}\right) + \exp\left(\theta_{11} + \theta_{v_1} + \theta_{h_1}\right)\right)\left(\exp\left(\theta_{v_1}\right) + \exp\left(\theta_{h_1}\right) + \exp\left(+ \theta_{v_1} + \theta_{h_1}\right)\right)}
+\end{matrix}
+\end{align*}
+
+Again, we can inspect these functions over multiple values of $\boldsymbol \theta$ to look for a relationship between parameter values and uninterpretability for the toy RBM.
+
+```{r uninterp, fig.cap="\\label{fig:uninterp}The difference between model expectations, E$\\left[\\boldsymbol X | \\boldsymbol \\theta\\right]$, and expectations given independence, E$\\left[\\boldsymbol X | \\emptyset\\right ]$ for a RBM with one visible and one hidden node.", fig.height = 4, fig.align='center', message=FALSE, cache=TRUE}
+possibles <- stats(1, 1, type = "binary")
+
+expand.grid(theta_h1 = seq(-5, 5, length.out = 5),
+            theta_v1 = seq(-5, 5, length.out = 5),
+            theta_11 = seq(-5, 5, length.out = 11)) %>%
+  data.frame() %>%
+  group_by(theta_v1, theta_h1, theta_11) %>%
+  do(data.frame(gamma = sum(exp(possibles %*% t(data.matrix(.)))),
+                mu_v1_dep = possibles[,"v1"] %*% exp(possibles %*% t(data.matrix(.))),
+                mu_h1_dep = possibles[,"h1"] %*% exp(possibles %*% t(data.matrix(.))))) %>%
+  mutate_each(funs(./gamma), starts_with("mu")) %>%
+  select(-gamma) -> exp_dep
+
+expand.grid(theta_h1 = seq(-5, 5, length.out = 5),
+            theta_v1 = seq(-5, 5, length.out = 5),
+            theta_11 = 0) %>%
+  data.frame() %>%
+  group_by(theta_v1, theta_h1, theta_11) %>%
+  do(data.frame(gamma = sum(exp(possibles %*% t(data.matrix(.)))),
+                mu_v1_indep = possibles[,"v1"] %*% exp(possibles %*% t(data.matrix(.))),
+                mu_h1_indep = possibles[,"h1"] %*% exp(possibles %*% t(data.matrix(.))))) %>%
+  mutate_each(funs(./gamma), starts_with("mu")) %>% ungroup() %>%
+  select(-gamma, -theta_11) -> exp_indep
+
+exp_dep %>%
+  left_join(exp_indep) %>%
+  mutate(mu_v1 = abs(mu_v1_dep - mu_v1_indep),
+         mu_h1 = abs(mu_h1_dep - mu_h1_indep)) %>%
+  gather(mu_parameter, value, mu_v1, mu_h1) %>%
+  ungroup() %>%
+  mutate(theta_h1 = factor(theta_h1, labels = paste0("theta[h1]==", unique(theta_h1))), 
+         theta_v1 = factor(theta_v1, labels = paste0("theta[v1]==", unique(theta_v1)))) %>%
+  ggplot() +
+  geom_line(aes(theta_11, value, colour = mu_parameter)) +
+  facet_grid(theta_h1 ~ theta_v1, labeller = label_parsed) +
+  xlab(expression(theta[11])) + ylab(expression(group("|", E(group("", bold(X), "|")*bold(theta)) - E(group("", bold(X), "|")*plain(independence)), "|"))) +
+  scale_colour_discrete(expression(bold(X)), labels = c(expression(v[1]), 
+                                                          expression(h[1])))
+
+```
+
+Figure \ref{fig:uninterp} shows that the absolute difference between model expectations, E$\left[\boldsymbol X | \boldsymbol \theta\right]$, and expectations given independence, E$\left[\boldsymbol X | \emptyset\right ]$ for the RBM with one hidden and one visible node grows as the values of $\boldsymbol \theta$ aare farther from zero.
+This is a third indication that parameters of large magnetude could lead to model impropriety in an RBM using a model with only one hidden and one visible node.
+
+##Data coding to mitigate degeneracy
 Multiple encodings of the binary variables are possible.
@@
 
+In addition to the argument that the $\{-1,1\} data encoding mitigates some near-degeneracy it also has the benefit of providing a guaranteed non-degenerate model at $\boldsymbol \theta = \boldsymbol 0 \in \mathbb{R}^{H + V + H*V}$.
+The proof of this is included in appendix \ref{appendix:center}.
+
+**TODO: Add center of the universe proof**
+
 Thus, moving forward we will restrict our examinations to the $\{-1, 1\}$-encoding of the binary variables in the RBM.
 
-#Exploring Manageable Examples
-```{r split_sample, cache=TRUE}
-#reshape data function
+#Exploring manageable examples
+```{r degen-data, message=FALSE, warning=FALSE, cache=TRUE}
+#reshape data functions
 plot_data <- function(res, grid = FALSE) {
@@
 }
+indep_params <- function(samp, H, V) {
+  samp[, (H + V + 1):ncol(samp)] <- 0
+  samp
+}
+max_Q <- function(theta, stats) {
+  apply(crossprod(t(stats), theta), 2, max)
+}
 
-load("../data/results_split.RData")
-plot_dat_split <- res %>% plot_data
+#grid data
+load("data/results_grid.RData")
 
-load("../data/results_grid.RData")
+#near-degeneracy
 plot_dat_grid <- res %>% plot_data(grid = TRUE)
+
+plot_dat_grid %>%
+  group_by(r1, r2, H, V) %>%
+  summarise(frac_degen = sum(near_hull)/n(), count = n()) %>% 
+  ungroup() %>% 
+  mutate(Hiddens = paste0("Hiddens: ", H), Visibles = paste0("Visibles: ", V)) -> convex_hull_summary
+
+#uninterpretability
+res %>%
+  group_by(H, V, N, r1, r2) %>%
+  do(indep_exp = t(expected_value(t(indep_params(.$samp[[1]], .$H, .$V)), .$stat[[1]]))[, -((.$H + .$V + 1):(.$H + .$V + .$H*.$V))],
+     marg_exp = .$g_theta[[1]][, (.$H + .$V + .$H*.$V + 1):(ncol(.$g_theta[[1]])-.$H*.$V)]) -> exp_vals
+
+exp_vals %>%
+  group_by(H, V, N, r1, r2) %>%
+  do(data.frame(mean_abs_diff = apply(abs(.$indep_exp[[1]] - .$marg_exp[[1]]), 1, mean))) %>%
+  group_by(H, V, N, r1, r2) %>%
+  summarise(mean_abs_diff = mean(mean_abs_diff)) %>%
+  mutate(Hiddens = paste0("Hiddens: ", H), Visibles = paste0("Visibles: ", V)) -> exp_vals_summary
+
+#instability
+res %>%
+  group_by(H, V, N, r1, r2) %>%
+  do(data.frame(LHS = max_Q(t(.$samp[[1]]), .$stat[[1]])/(.$H + .$V))) %>%
+  summarise(mean_max_q = mean(LHS)) -> max_q_summary
+
+convex_hull_summary %>%
+  left_join(max_q_summary) %>%
+  left_join(exp_vals_summary) -> three_ways
+
 ```
@@
 For $H, V \in \{1, \dots, 4\}$, we sample uniformly `r dim(res$samp[[1]])[1]` values of $\boldsymbol \theta$ with magnitude, $||\boldsymbol \theta || = \sqrt{\sum_{i = 1, \dots, H + V + H*V} \theta_i^2}$, between `r min(plot_dat_grid$r1)` and `r max(plot_dat_grid$r1)`.
-For each set of parameters we then calculate the mean parameterization, as detailed in section \ref{degeneracy-in-random-graph-models} and can classify each model as near-degenerate or viable and look at the fraction of models that were near-degenerate for each combination of magnitude of $\boldsymbol \theta$ and model size.
+For each set of parameters we then calculate the three metrics of model impropriety introduced in section \ref{degeneracy-instability-and-uninterpretability-oh-my}, $\mu(\boldsymbol \theta)$, $\max\limits_{\boldsymbol x_N \in \mathcal{X}_N}[Q(\boldsymbol x_N)]/N$, and $\left|\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] -   \text{E}\left[\boldsymbol X | \emptyset \right] \right|$.
+In the case of near-degeneracy, we can go further and classify each model as near-degenerate or viable based on the distance of $\mu(\boldsymbol \theta)$ from the convex hull of $\{t_k(\boldsymbol x): \boldsymbol x \in \mathcal{X}, k = 1, \dots, K\}$ and look at the fraction of models that were near-degenerate for each combination of magnitude of $\boldsymbol \theta$ and model size.
 
 In this numerical experiment, we split $\boldsymbol \theta$ into $\boldsymbol \theta_{interaction}$ and $\boldsymbol \theta_{main}$, in reference to which sufficient statistics the parameters correspond to, and allow the two types of terms to have varying magnitudes.
-The results of this numerical study can be seen in figure \ref{fig:split_plots}.
-From this figure, it is evident that the fraction of near-degenerate models simulated at specified magnitudes of $\boldsymbol \theta_{main}$ and $\boldsymbol \theta_{interaction}$ is more closely related to the combined distance from the origin of these two vectors, than any individual group of parameter values.
+The results of this numerical study can be seen in figures \ref{fig:degen_plots}, \ref{fig:instab_plots}, and \ref{fig:uninterp_plots}.
+From these three figures, it s clear to see that all three indications of model impropriety show higher values the larger the magnitude of the parameters.
+Additionally, since there are $H*V$ interaction terms in $\boldsymbol \theta$ versus only $H + V$ main effects terms, for large models this dictates that there are many more interaction parameters in the model.
+And so, perhaps limiting the magnitude of the overall distance from the origin for $\boldsymbol \theta_{interaction}$ may help to limit the effects of model impropriety.
 
-```{r split_plots, fig.cap="\\label{fig:split_plots}A numerical experiment looking at the fraction of models that were near-degenerate for each combination of magnitude of $\\boldsymbol \\theta$ and model size.
-The thick black line shows the level where the fraction of -near degenerate models is .05."}
-plot_dat_grid %>%
-  group_by(r1, r2, H, V) %>%
-  summarise(frac_degen = sum(near_hull)/n(), count = n()) %>% 
-  ungroup() %>% 
-  mutate(Hiddens = paste0("Hiddens: ", H), Visibles = paste0("Visibles: ", V)) %>%
+```{r degen_plots, fig.cap="\\label{fig:degen_plots}A numerical experiment looking at the fraction of models that were near-degenerate for each combination of magnitude of $\\boldsymbol \\theta$ and model size.
+The thick black line shows the level where the fraction of -near degenerate models is .05.", fig.height=4}
+three_ways %>%
   ggplot() +
@@
   geom_abline(aes(intercept = 0, slope = 1), alpha = .5, lty = 2) +
-  scale_fill_gradient("Fraction Near-degenerate", low = "yellow", high = "red") +
-#   geom_point(aes(r1, r2), alpha = .1, 
-#              data = plot_dat_split %>%
-#                mutate(ss_interaction = sqrt(ss_interaction), ss_main = sqrt(ss_main)) %>%
-#                select(-r1, -r2) %>%
-#                rename(r1 = ss_main, r2 = ss_interaction) %>%
-#                group_by(r1, r2, n_param) %>%
-#                summarise(frac_degen = sum(near_hull)/n(), count = n()) %>%
-#                filter(frac_degen > 0)) +
+  scale_fill_gradient("Fraction near-degenerate", low = "yellow", high = "red") +
   facet_grid(Visibles~Hiddens) +
@@
   ylab(expression(group("||", theta[interaction], "||"))) +
-  theme(aspect.ratio = 1)
-  #theme(legend.position = "bottom")
+  theme(aspect.ratio = 1)  -> p.degen
+
+three_ways %>%
+  ggplot() +
+  geom_tile(aes(x = r1, y = r2, fill = mean_abs_diff)) +
+  geom_contour(aes(x = r1, y = r2, z = mean_abs_diff), colour = "black", bins = 8) +
+  #geom_contour(aes(x = r1, y = r2, z = mean_abs_diff), colour = "black", breaks = .05, size = 1.5) +
+  geom_abline(aes(intercept = 0, slope = 1), alpha = .5, lty = 2) +
+  scale_fill_gradient(expression(group("|", E(group("", bold(X), "|")*bold(theta)) - E(group("", bold(X), "|")*plain(independence)), "|")), low = "yellow", high = "red") +
+  facet_grid(Visibles~Hiddens) +
+  xlab(expression(group("||", theta[main], "||"))) +
+  ylab(expression(group("||", theta[interaction], "||"))) +
+  theme(aspect.ratio = 1) -> p.exp_diff
+
+three_ways %>%
+  ggplot() +
+  geom_tile(aes(x = r1, y = r2, fill = mean_max_q)) +
+  geom_contour(aes(x = r1, y = r2, z = mean_max_q), colour = "black", bins = 8) +
+  #geom_contour(aes(x = r1, y = r2, z = mean_max_q), colour = "black", breaks = .05, size = 1.5) +
+  geom_abline(aes(intercept = 0, slope = 1), alpha = .5, lty = 2) +
+  scale_fill_gradient(expression(frac(max(Q), H + V)), low = "yellow", high = "red") +
+  facet_grid(Visibles~Hiddens) +
+  xlab(expression(group("||", theta[main], "||"))) +
+  ylab(expression(group("||", theta[interaction], "||"))) +
+  theme(aspect.ratio = 1) -> p.max_q
+
+p.degen
+```
+
+```{r instab_plots, fig.cap="\\label{fig:instab_plots}A numerical experiment looking at $\\max\\limits_{\\boldsymbol x_N \\in \\mathcal{X}_N}[Q(\\boldsymbol x_N)]/N$ for each combination of magnitude of $\\boldsymbol \\theta$ and model size.
+As the magnitude of $\\boldsymbol \\theta$ grows, so does the value of this metric.", fig.height=4}
+p.max_q
+```
+
+```{r uninterp_plots, fig.cap="\\label{fig:split_plots}The difference between model expectations, E$\\left[\\boldsymbol X | \\boldsymbol \\theta\\right]$, and expectations given independence, E$\\left[\\boldsymbol X | \\emptyset\\right ]$ for each combination of magnitude of $\\boldsymbol \\theta$ and model size.
+Larger magnitudes of $\\boldsymbol \\theta$ correspond to larger differences, thus indicating higher uninterpretability.", fig.height=4}
+p.exp_diff
 ```
@@
 ```{r split_plots2, warning=FALSE, fig.cap="\\label{fig:split_plots_dist}The fraction near-degenerate for each magnitude of $\\boldsymbol \\theta$ for each model.
-For each number of visibles in the model, the fraction near-degenerate becomes greater than zero at larger values of $||\\boldsymbol \\theta||$ as $H$ increases and the slope becomes steeper as $H$ increases also.", fig.height=2}
+For each number of visibles in the model, the fraction near-degenerate becomes greater than zero at larger values of $||\\boldsymbol \\theta||$ as $H$ increases and the slope becomes steeper as $H$ increases also.", fig.height=5}
 plot_dat_grid %>% 
@@
   ggplot() +
-  geom_point(aes(dist, frac_degen, colour = Hiddens)) +
+  geom_point(aes(dist, frac_degen, colour = Hiddens), size = .5) +
   geom_smooth(aes(dist, frac_degen, colour = Hiddens, group = Hiddens)) +
-  facet_grid(~Visibles) +
+  facet_wrap(~Visibles) +
   ylim(c(0,1)) +
@@
 ```
+It is evidenced from these manageable examples that RBMs very easily are near-degenerate, unstable, and uninterpretable for large portions of the parameter space.
+The question becomes, can a rigorous fitting method be developed that avoids these issues?
 
-#Discussion
+#Bayesian model fitting
+To avoid model impropriety, we want to avoid parts of the parameter space that lead to near-degeneracy}, instability, and uninterpretability.
+Based on the evidence in section \ref{exploring-manageable-examples}, one idea is to shrink $\boldsymbol \theta$ by specifying priors that place low probability on high values of $\boldsymbol \theta$.
+
+Using this idea of shrinkage, we simulated $n = 5,000$ images with four pixels (four visible nodes) from an RBM model with 4 hiddens and fit the RBM using Bayesian methods, and three separate methods.
+
+1. *Trick prior.* Cancel out the normalizing term, resulting full conditionals are normally distributed. 
+    \begin{align*}
+    \pi(\boldsymbol \theta) \propto \gamma(\boldsymbol \theta)^n \exp\left(-\frac{1}{2C_{1}}\boldsymbol \theta_{main}'\boldsymbol \theta_{main} -\frac{1}{2C_{2}}\boldsymbol \theta_{interaction}'\boldsymbol \theta_{interaction}\right), \vspace{-.75cm}
+    \end{align*}
+
+    where
+
+    \begin{align*}
+    \gamma(\boldsymbol \theta) = \sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V\sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right) \text{ and } C_{2} < C_{1}
+    \end{align*}
+    [@li2014biclustering].
+2.
+*Truncated Normal prior.* Use two independent truncated spherical normal distributions as priors for $\theta_{main}$ and $\theta_{interaction}$ with $\sigma_{interaction} < \sigma_{main}$.
+Full conditional distributions are not conjugate, requires a geometric adaptive Metropolis Hastings step [@zhou2014some] and calculation of likelihood normalizing constant.
+3. *Marginalized likelihood* Marginalize out $\boldsymbol h$ in $f_{\boldsymbol \theta}(\boldsymbol x)$, and use the truncated Normal prior. 
+    \begin{align*}
+    g_{\boldsymbol \theta}(\boldsymbol v) = \sum\limits_{\boldsymbol h \in \{-1,1\}^H} \exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right).
+\vspace{-.75cm}
+    \end{align*}
+
+The three fitting methods are ordered by computational feasibility in a real data situation with the trick prior being the most possible due to conjugacy and the marginalized likelihood method the least feasible due to the marginalization and need for an adaptive Metropolis Hastings step.
+The trick prior, however, does require extensive tuning to choose $C_1$ and $C_2$, making it less appealing.
+
+Figure \ref{fig:fitting_plot} shows the posterior probability of each possible binary image with four pixels after fitting the model in the three ways detailed in this section.
+The black lines show the true probabilities of each image based on the parameters used to simulate the images.
+From these posterior predictive checks, it is evident that the marginalized likelihood method shows the closest fit to the actual data.
+However, this model requires a marginalization step, which is infeasible for a model of any real size.
+
+```{r fitting}
+#data and params ------------------------
+H <- 4
+V <- 4
+
+#marginalized likelihood
+load("data/fitted_models_adaptive_mh_trunc_distn_marginal_1.Rdata")
+marginal_bad_1 <- models_bad
+marginal_good_1 <- models_good
+marginal_degen_1 <- models_degen
+
+#load Jing's prior
+load("data/fitted_models_distn_0.26.Rdata")
+jing_bad <- models_bad
+jing_good <- models_good
+jing_degen <- models_degen
+
+#truncated normal
+load("data/fitted_models_adaptive_mh_trunc_distn_shrunk_jing_match.Rdata")
+shrunk_jing_bad <- models_bad
+shrunk_jing_good <- models_good
+shrunk_jing_degen <- models_degen
+
+#rm unneccesary data
+rm(models_bad)
+rm(models_degen)
+rm(models_good)
+
+load("data/sample.params.Rdata")
+params_degen <- list(main_hidden = sample.params %>% ungroup() %>% filter(near_hull) %>% select(starts_with("h"), -H) %>% data.matrix(),
+                     main_visible = sample.params %>% ungroup() %>% filter(near_hull) %>% select(starts_with("v"), -V) %>% data.matrix(),
+                     interaction = sample.params %>% ungroup() %>% filter(near_hull) %>% select(starts_with("theta")) %>% data.matrix() %>% matrix(4))
+params_good <- list(main_hidden = sample.params %>% ungroup() %>% filter(!near_hull) %>% select(starts_with("h"), -H) %>% data.matrix(),
+                    main_visible = sample.params %>% ungroup() %>% filter(!near_hull) %>% select(starts_with("v"), -V) %>% data.matrix(),
+                    interaction = sample.params %>% ungroup() %>% filter(!near_hull) %>% select(starts_with("theta")) %>% data.matrix() %>% matrix(4))
+
+load("data/sample_images.Rdata")
+
+#computing actual distributions ---------------------
+distn_good <- visible_distn(params = params_good)
+distn_degen <- visible_distn(params = params_degen)
+
+reshape_sample_distn <- function(model) {
+  sample_distn <- model$distn
+  dim(sample_distn) <- c(dim(sample_distn)[1]*dim(sample_distn)[2], dim(sample_distn)[3])
+  sample_distn %>% data.frame() -> sample_distn
+  names(sample_distn) <- names(distn_degen)
+
+  sample_distn %>%
+    group_by(image_id) %>%
+    mutate(iter = 1:n()) -> sample_distn
+
+  return(sample_distn)
+}
+
+marginal_sample_good_1 <- reshape_sample_distn(marginal_good_1)
+jing_sample_good <- reshape_sample_distn(jing_good)
+shrunk_jing_sample_good <- reshape_sample_distn(shrunk_jing_good) %>% filter(iter > 500) %>% mutate(iter = 1:n())
+
+#plots -------------------
+marginal_sample_good_1 %>% rename(marginal = prob) %>%
+  left_join(jing_sample_good %>% ungroup() %>% select(-image_id) %>% rename(trick = prob)) %>%
+  left_join(shrunk_jing_sample_good %>% ungroup() %>% select(-image_id) %>% rename(trunc = prob)) %>%
+  ungroup() %>%
+  mutate(method = "good", image_id = paste0("image_", image_id)) %>% 
+  #select_(.dots = paste0("-v", 1:V)) %>%
+  gather(prior, prob, trick, marginal, trunc) %>%
+  spread(image_id, prob) -> all_statistics
+
+all_statistics %>%
+  gather(statistic, value, -iter, -method, -prior, -starts_with("v")) %>%
+  filter(grepl("image", statistic) & !is.na(value)) %>%
+  separate(statistic, into = c("junk", "statistic")) %>%
+  mutate(statistic = factor(statistic, labels = paste("Image", 1:length(unique(statistic))))) %>%
+  left_join(distn_good %>% select(-image_id)) %>% 
+  ggplot() +
+  geom_line(aes(iter/100, value, colour = prior)) +
+  geom_abline(aes(slope = 0, intercept = prob)) +
+  facet_wrap(~statistic) +
+  ylim(c(0,.5)) +
+  xlab("MCMC Iteration (in 100s)") + ylab("Posterior Probability") +
+  theme(legend.position = "bottom") +
+  scale_colour_discrete("Method", labels = c("Marginalized likelihood", "Trick prior", "Truncated Normal prior")) -> p
+
+```
+
+```{r fitting_plot, fig.cap="\\label{fig:fitting_plot}Posterior probability of each possible 4-pixel image using three Bayesian fitting techniques, trick prior, truncated Normal prior, and marginalized likelihood.
+The black lines show the true probabilities of each image based on the parameters used to simulate the images.
+The marginalized likelihood method has the best fit for the data, however is also the most computationally intensive and least feasible with a real dataset.", fig.height=6}
+p
+```
+
+It should also be noted that each of these rigorous fitting methods seeks to merely replicate the empirical data distribution, which is the optimal nonparametric solution.
+In a data application with a large sample space it is unlikely that the training set will include one of each possible image, thus there will be possible images that are not realized in the training set, unlike this small example.
+
+**TODO: Add proof that likelihood goes to infinity.**  
+**TODO: Add marginalized Jing prior?**  
+**TODO: Add comparison to contrastive divergence**
 

### 2016-06-10 writing/draft.Rmd
updated grid to be more useful and wrote a little

@@
 author: "Andee Kaplan, Stephen Vardeman, Daniel Nordman"
-date: "May, 2016"
+date: "June, 2016"
 output: 
@@
 ```
+#Introduction
 
-#Restricted Boltzmann machines
+#Background
+
+##Restricted Boltzmann machines
 
@@
 
-##Joint distribution
+###Joint distribution
 Let $\boldsymbol x = \{h_1, \dots, h_H, v_1,\dots,v_V\}$ represent the states of the visible and hidden nodes in an RBM.
@@
 
-##Connection to Deep Learning
+###Connection to Deep Learning
 RBMs have risen to prominence in recent years due to their connection to deep learning [see @hinton2006fast, @salakhutdinov2012efficient, or @srivastava2013modeling for examples].
@@
 
-#Degeneracy, instability, and uninterpretability... Oh my!
+##Degeneracy, instability, and uninterpretability... Oh my!
 The highly flexible nature of the RBM ($H + V + H V$ parameters) makes three characteristics of model impropriety of particular concern, *degeneracy, instability, and uninterpretability*.
@@
 
-##Near-degeneracy
+###Near-degeneracy
 In Random Graph Model theory, *degeneracy* occurs when there is a disproportionate amount of probability placed on only a few elements of the sample space, $\mathcal{X}$, by the model.
@@
 
-##Instability
+###Instability
 @schweinberger2011instability introduces a related concept, *instability*, which occurs when small changes in the natural parameters result in large changes in probability masses.
@@
 
-##Uninterpretability
+###Uninterpretability
 @kaiser2007statistical defines a measure of model impropriety called *uninterpretability*, which is characterized as due to the existence of dependence, marginal mean-structure no longer being maintained in the model.
@@
 
-#Tiny example
+#Characterizations
+##Tiny example
 To illustrate the ideas of graph near-degeneracy, instability, and uninterpretability in a RBM, we will consider a the smallest possible toy example that consists of one visible node $v_1$ and one hidden node $h_1$, both binary.
@@
 
-##Impropriety three ways
+###Impropriety three ways
 
@@
 
-```{r instab, fig.cap="\\label{fig:instab}$\\max\\limits_{\\boldsymbol x_N \\in \\mathcal{X}_N}[Q(\\boldsymbol x_N)]/N$ for various values of $\\boldsymbol \\theta$ for the tiny example model.
+```{r instab, fig.cap="\\label{fig:instab}$(\\max\\limits_{\\boldsymbol x_N \\in \\mathcal{X}_N}[Q(\\boldsymbol x_N)] - \\min\\limits_{\\boldsymbol x_N \\in \\mathcal{X}_N}[Q(\\boldsymbol x_N)])/N$ for various values of $\\boldsymbol \\theta$ for the tiny example model.
 This quantity is increasing for larger positive values of $\\boldsymbol \\theta$.", fig.height = 4, fig.align='center'}
@@
   group_by(theta_v1, theta_h1, theta_11) %>%
-  do(data.frame(q_max = max(possibles %*% t(data.matrix(.))))) %>%
+  do(data.frame(q_max = max(possibles %*% t(data.matrix(.))) - min(possibles %*% t(data.matrix(.))))) %>%
   ungroup() %>%
@@
   xlab(expression(theta[11])) +
-  ylab(expression(frac(max(Q), H + V)))
+  ylab(expression(frac(max(Q) - min(Q), H + V)))
 ```
@@
 
-##Data coding to mitigate degeneracy
+###Data coding to mitigate degeneracy
 Multiple encodings of the binary variables are possible.
@@
 
-In addition to the argument that the $\{-1,1\} data encoding mitigates some near-degeneracy it also has the benefit of providing a guaranteed non-degenerate model at $\boldsymbol \theta = \boldsymbol 0 \in \mathbb{R}^{H + V + H*V}$.
+In addition to the argument that the $\{-1,1\}$ data encoding mitigates some near-degeneracy it also has the benefit of providing a guaranteed non-degenerate model at $\boldsymbol \theta = \boldsymbol 0 \in \mathbb{R}^{H + V + H*V}$.
 The proof of this is included in appendix \ref{appendix:center}.
@@
 
-#Exploring manageable examples
+##Exploring manageable examples
 ```{r degen-data, message=FALSE, warning=FALSE, cache=TRUE}
@@
 }
+min_Q <- function(theta, stats) {
+  apply(crossprod(t(stats), theta), 2, min)
+}
+min_max_Q <- function(theta, stats) {
+  require(dplyr)
+  tcrossprod(stats, theta) %>% data.frame() %>%
+    cbind(stats) %>%
+    group_by_(.dots = colnames(stats)[grepl("v", colnames(stats)) & !grepl("theta", colnames(stats))]) %>%
+    summarise_each(funs(max), contains("X")) %>%
+    ungroup() %>%
+    summarise_each(funs(min), contains("X")) %>%
+    select(contains("X")) %>%
+    data.matrix() %>%
+    t()
+}
 
@@
   group_by(H, V, N, r1, r2) %>%
-  do(data.frame(mean_abs_diff = apply(abs(.$indep_exp[[1]] - .$marg_exp[[1]]), 1, mean))) %>%
+  do(data.frame(max_abs_diff = apply(abs(.$indep_exp[[1]] - .$marg_exp[[1]]), 1, max))) %>%
   group_by(H, V, N, r1, r2) %>%
-  summarise(mean_abs_diff = mean(mean_abs_diff)) %>%
+  summarise(max_abs_diff = mean(max_abs_diff)) %>%
   mutate(Hiddens = paste0("Hiddens: ", H), Visibles = paste0("Visibles: ", V)) -> exp_vals_summary
@@
   group_by(H, V, N, r1, r2) %>%
-  do(data.frame(LHS = max_Q(t(.$samp[[1]]), .$stat[[1]])/(.$H + .$V))) %>%
-  summarise(mean_max_q = mean(LHS)) -> max_q_summary
+  do(data.frame(max = max_Q(t(.$samp[[1]]), .$stat[[1]]),
+                min = min_Q(t(.$samp[[1]]), .$stat[[1]]),
+                min_max = min_max_Q(.$samp[[1]], .$stat[[1]]))) %>%
+  ungroup() %>% 
+  group_by(H, V, N, r1, r2) %>%
+  mutate(LHS1 = (max - min)/N,
+         LHS2 = (max - min_max - H*log(2))/N) %>%
+  summarise_each(funs(mean), LHS1, LHS2) -> max_q_summary
 
@@
 three_ways %>%
+  mutate(r1 = round(r1/(H + V), 8), r2 = round(r2/(H * V), 8)) %>% 
   ggplot() +
@@
   facet_grid(Visibles~Hiddens) +
-  xlab(expression(group("||", theta[main], "||"))) +
-  ylab(expression(group("||", theta[interaction], "||"))) +
+  xlab(expression(paste(group("||", theta[main], "||"), "/(H + V)"))) +
+  ylab(expression(paste(group("||", theta[interaction], "||"), "/(H*V)"))) +
   theme(aspect.ratio = 1) -> p.degen
@@
 three_ways %>%
+  mutate(r1 = round(r1/(H + V), 8), r2 = round(r2/(H * V), 8)) %>% 
   ggplot() +
-  geom_tile(aes(x = r1, y = r2, fill = mean_abs_diff)) +
-  geom_contour(aes(x = r1, y = r2, z = mean_abs_diff), colour = "black", bins = 8) +
+  geom_tile(aes(x = r1, y = r2, fill = max_abs_diff)) +
+  geom_contour(aes(x = r1, y = r2, z = max_abs_diff), colour = "black", bins = 8) +
   #geom_contour(aes(x = r1, y = r2, z = mean_abs_diff), colour = "black", breaks = .05, size = 1.5) +
   geom_abline(aes(intercept = 0, slope = 1), alpha = .5, lty = 2) +
-  scale_fill_gradient(expression(group("|", E(group("", bold(X), "|")*bold(theta)) - E(group("", bold(X), "|")*plain(independence)), "|")), low = "yellow", high = "red") +
+  scale_fill_gradient(expression(group("|", E(group("", bold(X), "|")*bold(theta)) - E(group("", bold(X), "|")*plain(independence)), "|")), low = "yellow", high = "red", limits = c(0,2)) +
   facet_grid(Visibles~Hiddens) +
-  xlab(expression(group("||", theta[main], "||"))) +
-  ylab(expression(group("||", theta[interaction], "||"))) +
+  xlab(expression(paste(group("||", theta[main], "||"), "/(H + V)"))) +
+  ylab(expression(paste(group("||", theta[interaction], "||"), "/(H*V)"))) +
   theme(aspect.ratio = 1) -> p.exp_diff
@@
 three_ways %>%
+  mutate(r1 = round(r1/(H + V), 8), r2 = round(r2/(H * V), 8)) %>% 
   ggplot() +
-  geom_tile(aes(x = r1, y = r2, fill = mean_max_q)) +
-  geom_contour(aes(x = r1, y = r2, z = mean_max_q), colour = "black", bins = 8) +
+  geom_tile(aes(x = r1, y = r2, fill = LHS1)) +
+  geom_contour(aes(x = r1, y = r2, z = LHS1), colour = "black", bins = 8) +
   #geom_contour(aes(x = r1, y = r2, z = mean_max_q), colour = "black", breaks = .05, size = 1.5) +
   geom_abline(aes(intercept = 0, slope = 1), alpha = .5, lty = 2) +
-  scale_fill_gradient(expression(frac(max(Q), H + V)), low = "yellow", high = "red") +
+  scale_fill_gradient(expression(frac(max(Q) - min(Q), H + V)), low = "yellow", high = "red") +
   facet_grid(Visibles~Hiddens) +
-  xlab(expression(group("||", theta[main], "||"))) +
-  ylab(expression(group("||", theta[interaction], "||"))) +
+  xlab(expression(paste(group("||", theta[main], "||"), "/(H + V)"))) +
+  ylab(expression(paste(group("||", theta[interaction], "||"), "/(H*V)"))) +
   theme(aspect.ratio = 1) -> p.max_q
+
+three_ways %>%
+  mutate(r1 = round(r1/(H + V), 8), r2 = round(r2/(H * V), 8)) %>% 
+  ggplot() +
+  geom_tile(aes(x = r1, y = r2, fill = LHS2)) +
+  geom_contour(aes(x = r1, y = r2, z = LHS2), colour = "black", bins = 8) +
+  #geom_contour(aes(x = r1, y = r2, z = mean_max_q), colour = "black", breaks = .05, size = 1.5) +
+  geom_abline(aes(intercept = 0, slope = 1), alpha = .5, lty = 2) +
+  scale_fill_gradient(expression(frac(max(Q) - min[v](max[h](Q)) - Hlog(2), H + V)), low = "yellow", high = "red") +
+  facet_grid(Visibles~Hiddens) +
+  xlab(expression(paste(group("||", theta[main], "||"), "/(H + V)"))) +
+  ylab(expression(paste(group("||", theta[interaction], "||"), "/(H*V)"))) +
+  theme(aspect.ratio = 1) -> p.min_max_q
 
@@
 
-```{r instab_plots, fig.cap="\\label{fig:instab_plots}A numerical experiment looking at $\\max\\limits_{\\boldsymbol x_N \\in \\mathcal{X}_N}[Q(\\boldsymbol x_N)]/N$ for each combination of magnitude of $\\boldsymbol \\theta$ and model size.
+```{r instab_plots, fig.cap="\\label{fig:instab_plots}A numerical experiment looking at $(\\max\\limits_{\\boldsymbol x_N \\in \\mathcal{X}_N}[Q(\\boldsymbol x_N)] - \\min\\limits_{\\boldsymbol x_N \\in \\mathcal{X}_N}[Q(\\boldsymbol x_N)])/N$ for each combination of magnitude of $\\boldsymbol \\theta$ and model size.
 As the magnitude of $\\boldsymbol \\theta$ grows, so does the value of this metric.", fig.height=4}
 p.max_q
+```
+
+```{r instab_plots_2, fig.cap="\\label{fig:instab_plots}A numerical experiment looking at $(\\max\\limits_{\\boldsymbol x_N \\in \\mathcal{X}_N}[Q(\\boldsymbol x_N)] - \\min\\limits_{\\boldsymbol v} \\max\\limits_{\\boldsymbol h}[Q(\\boldsymbol x_N)] - H\\log(2))/N$ for each combination of magnitude of $\\boldsymbol \\theta$ and model size.
+As the magnitude of $\\boldsymbol \\theta$ grows, so does the value of this metric.", fig.height=4}
+p.min_max_q
 ```
@@
 plot_dat_grid %>% 
-  mutate(dist = sqrt(r1^2 + r2^2)) %>%
+  mutate(dist = sqrt(r1^2 + r2^2)/(H + V + H*V)) %>%
   group_by(dist, H, V) %>%
@@
   geom_point(aes(dist, frac_degen, colour = Hiddens), size = .5) +
-  geom_smooth(aes(dist, frac_degen, colour = Hiddens, group = Hiddens)) +
+  #geom_smooth(aes(dist, frac_degen, colour = Hiddens, group = Hiddens)) +
   facet_wrap(~Visibles) +
   ylim(c(0,1)) +
-  xlab(expression(paste(group("||", theta, "||"), ", Distance from the origin"))) +
+  xlab(expression(paste(group("||", theta, "||"), "/(H + V + H*V), Average distance from the origin"))) +
   ylab("Fraction Near-degenerate")
@@
 
+#Discussion
+
+#Appendix
+
 #References

### 2016-06-10 degeneracy/grid_sample.R

@@ -61,11 +61,13 @@ n <- 100
 r_exp <- .05
 
 expand.grid(H = 1:4, V = 1:4, r1 = seq(0.001, 3, length.out = 20), r2 = seq(0.001, 3, length.out = 20)) %>%
+  mutate(r1 = r1*(H + V), r2 = r2*(H*V)) %>%
   mutate(N = H + V) %>%
   group_by(H, V, r1, r2, N) %>%
   do(samp = cbind(sample_sphere_unif(.$N, n, 0, .$r1), sample_sphere_unif(.$H*.$V, n, 0, .$r2))) -> grid_sample
 
 expand.grid(H = 1:4, V = 1:4, r1 = seq(0.001, 3, length.out = 20), r2 = seq(0.001, 3, length.out = 20)) %>%
+  mutate(r1 = r1*(H + V), r2 = r2*(H*V)) %>%
   mutate(N = H + V) %>%
   group_by(H, V, r1, r2, N) %>%
   do(stat = stats(.$H, .$V, "negative")) -> grid_sample_stat

### 2016-06-22 writing/cut_text.Rmd
beef up instability

@@
 ```
+
+```{r instab_plots, fig.cap="\\label{fig:instab_plots}A numerical experiment looking at $(\\max\\limits_{\\boldsymbol x_N \\in \\mathcal{X}_N}[Q(\\boldsymbol x_N)] - \\min\\limits_{\\boldsymbol x_N \\in \\mathcal{X}_N}[Q(\\boldsymbol x_N)])/N$ for each combination of magnitude of $\\boldsymbol \\theta$ and model size.
+As the magnitude of $\\boldsymbol \\theta$ grows, so does the value of this metric.", fig.height=4}
+p.max_q
+```
+
+There are $H + V + H*V$ terms in the sum that comprises $Q(\boldsymbol x)$, which leads to a natural split of the quantity into main effects, those terms based solely on $h_i$ and $v_j$ and interaction effects, those terms based on the product of $h_i v_j$.
+There are $H+V$ main effects terms and $H*V$ interaction effects terms.
+For $H$ and $V$ large enough, the number of interaction terms is growing much faster than $H + V$, meaning no matter the $C$, we can restrict the size of the left side of equation \ref{eq:stable} effectively by restricting the size of the interaction terms.
+This points to the interaction terms as potentially the more dangerous parameters for causing instability in the RBM simply because there are so many more of them.
+
+```{r interaction size, include=FALSE}
+expand.grid(H = 1:10, V = 1:10) %>%
+  mutate(interaction_terms = H*V) %>%
+  ggplot() +
+  geom_contour(aes(H, V, z = interaction_terms/(H + V), colour = ..level..), bins = 20) +
+  geom_contour(aes(H, V, z = interaction_terms/(H + V)), colour = "black", breaks = 1) +
+  scale_colour_gradient("H*V/(H + V)", low = "yellow", high = "red")
+
+```

### 2016-06-22 writing/draft.Rmd
beef up instability

@@
 @schweinberger2011instability introduces a related concept, *instability*, which occurs when small changes in the natural parameters result in large changes in probability masses.
-This can be thought of as excessive sensitivity in the model.
+This can be characterized of as excessive sensitivity in the model, where a small change in $\boldsymbol x$ leads to a large change in the probability mass function.
 @schweinberger2011instability provides the following tool for detecting instability in the a discrete exponential family model.
@@
 \begin{align}
-\max\limits_{\boldsymbol x_N \in \mathcal{X}_N}[Q(\boldsymbol x_N)] > CN \text{ for all } N > N_C \label{eq:unstable}
+\max\limits_{\boldsymbol x_N \in \mathcal{X}_N}[Q(\boldsymbol x_N)] - \min\limits_{\boldsymbol x_N \in \mathcal{X}_N}[Q(\boldsymbol x_N)] > CN \text{ for all } N > N_C \label{eq:unstable}
 \end{align}
@@
 \begin{align}
-\max\limits_{\boldsymbol x_N \in \mathcal{X}_N}[Q(\boldsymbol x_N)] \le CN \text{ for all } N > N_C. \label{eq:stable}
+\max\limits_{\boldsymbol x_N \in \mathcal{X}_N}[Q(\boldsymbol x_N)] - \min\limits_{\boldsymbol x_N \in \mathcal{X}_N}[Q(\boldsymbol x_N)] \le CN \text{ for all } N > N_C. \label{eq:stable}
 \end{align}
@@
 
-Within the context of the RBM, $N$ is the number of nodes in the model, $H + V$.
-To ensure the model is stable, we must restrict $\boldsymbol \theta$ such that there exists constants $C > 0$ and $N_C > 0$ that force equation \ref{eq:stable} to hold.
+Within the context of the RBM, this definition concerns the joint distribution of both hidden and visible nodes, where $N$ is the number of nodes in the model, $H + V$.
+However, instability in terms of only the visible nodes is a more useful measure of model impropriety to be concerned with.
+To see this, consider the example of classifying images using an RBM.
+One would be interested in the maximum change in the likelihood when a single pixel is changed in the visibles as a proxy for instability in the model.
+Thus, we can instead concern ourselves with the notion of instability that deals with the maximum likelihood ratio of two images that differ in only one pixel.
 
 \begin{align*}
-\max\limits_{\boldsymbol x_N \in \mathcal{X}_N}[Q(\boldsymbol x_N)] &= \max\limits_{\boldsymbol x \in \mathcal{X}}\left[ \sum\limits_{i = 1}^H\sum\limits_{j=1}^V \theta_{ij} h_i v_j + \sum\limits_{i = 1}^H\theta_{hi} h_i + \sum\limits_{j = 1}^V\theta_{vj} v_j \right] \\
-&= \max\limits_{h_j, v_i \in \{0,1\}}\left[ \sum\limits_{i = 1}^H\sum\limits_{j=1}^V \theta_{ij} h_i v_j + \sum\limits_{i = 1}^H\theta_{hi} h_i + \sum\limits_{j = 1}^V\theta_{vj} v_j \right] \\
-&\le C(H + V)
+&\max\left\{\log\frac{P_{\boldsymbol \theta}(v_1, \dots, v_V)}{P_{\boldsymbol \theta}(v^*_1, \dots, v^*_V)}: (v_1, \dots, v_V) \text{ and } (v^*_1, \dots, v^*_V) \in \{-1, 1\}^V \text{ differ by one component} \right\} \\
+&\quad = \max\limits_{k \in 1, \dots, V}\left\{\log\frac{\sum\limits_{\boldsymbol h \in \{-1,1\}^H} \exp\left(
+\sum\limits_{i = 1}^V \theta_{v_i}v_i + \sum\limits_{j = 1}^H \theta_{h_j}h_j + \sum\limits_{i = 1}^V\sum\limits_{j = 1}^H \theta_{ij}v_i h_j\right)}{\sum\limits_{\boldsymbol h \in \{-1,1\}^H} \exp\left(
+\sum\limits_{i \not= k} \theta_{v_i}v_i + \sum\limits_{j = 1}^H \theta_{h_j}h_j + \sum\limits_{i \not=k}\sum\limits_{j = 1}^H \theta_{ij}v_i h_j + \theta_{v_k}v_k^* + v_k^*\sum\limits_{j = 1}^H \theta_{kj}h_j
+\right)}: v_k \not= v^*_k \right\} \\
+& \quad \ge \max\limits_{k \in 1, \dots, V}\left\{\log\frac{\max\limits_{\boldsymbol h \in \{-1,1\}^H}\exp\left(
+\sum\limits_{i = 1}^V \theta_{v_i}v_i + \sum\limits_{j = 1}^H \theta_{h_j}h_j + \sum\limits_{i = 1}^V\sum\limits_{j = 1}^H \theta_{ij}v_i h_j\right)}{2^H\max\limits_{\boldsymbol h \in \{-1,1\}^H}\exp\left(
+\sum\limits_{i \not= k} \theta_{v_i}v_i + \sum\limits_{j = 1}^H \theta_{h_j}h_j + \sum\limits_{i \not=k}\sum\limits_{j = 1}^H \theta_{ij}v_i h_j + \theta_{v_k}v_k^* + v_k^*\sum\limits_{j = 1}^H \theta_{kj}h_j
+\right)}: v_k \not= v^*_k \right\} \\
+&\quad = \max\limits_{k \in 1, \dots, V}\left\{\max\limits_{\boldsymbol h \in \{-1,1\}^H}\left\{
+\sum\limits_{i = 1}^V \theta_{v_i}v_i + \sum\limits_{j = 1}^H \theta_{h_j}h_j + \sum\limits_{i = 1}^V\sum\limits_{j = 1}^H \theta_{ij}v_i h_j\right\}\right. \\
+& \quad \quad \left. - \max\limits_{\boldsymbol h \in \{-1,1\}^H}
+\left\{\sum\limits_{i \not= k} \theta_{v_i}v_i + \sum\limits_{j = 1}^H \theta_{h_j}h_j + \sum\limits_{i \not=k}\sum\limits_{j = 1}^H \theta_{ij}v_i h_j + \theta_{v_k}v_k^* + v_k^*\sum\limits_{j = 1}^H \theta_{kj}h_j
+\right\} - H\log2: v_k \not= v^*_k \right\} \\
+&\quad \ge \max\limits_{\boldsymbol v \in \{-1,1\}^V}\max\limits_{\boldsymbol h \in \{-1,1\}^H}Q(\boldsymbol x) - \min\limits_{\boldsymbol v \in \{-1,1\}^V}\max\limits_{\boldsymbol h \in \{-1,1\}^H}Q(\boldsymbol x) - H\log 2
 \end{align*}
 
-There are $H + V + H*V$ terms in the sum that comprises $Q(\boldsymbol x)$, which leads to a natural split of the quantity into main effects, those terms based solely on $h_i$ and $v_j$ and interaction effects, those terms based on the product of $h_i v_j$.
-There are $H+V$ main effects terms and $H*V$ interaction effects terms.
-For $H$ and $V$ large enough, the number of interaction terms is growing much faster than $H + V$, meaning no matter the $C$, we can restrict the size of the left side of equation \ref{eq:stable} effectively by restricting the size of the interaction terms.
-This points to the interaction terms as potentially the more dangerous parameters for causing instability in the RBM simply because there are so many more of them.
-
-```{r interaction size, include=FALSE}
-expand.grid(H = 1:10, V = 1:10) %>%
-  mutate(interaction_terms = H*V) %>%
-  ggplot() +
-  geom_contour(aes(H, V, z = interaction_terms/(H + V), colour = ..level..), bins = 20) +
-  geom_contour(aes(H, V, z = interaction_terms/(H + V)), colour = "black", breaks = 1) +
-  scale_colour_gradient("H*V/(H + V)", low = "yellow", high = "red")
-
-```
+Let $R(\boldsymbol \theta) \equiv \max_{\boldsymbol v} \max_{\boldsymbol h}Q(\boldsymbol x) - \min_{\boldsymbol v}\max_{\boldsymbol h}Q(\boldsymbol x) - H\log 2$.
+Then, if $R(\boldsymbol \theta) > C*V$, this implies that the maximum log-likelihood ratio of two images that differ in only one pixel is greater than $C$.
+So, by examining $R(\boldsymbol\theta)/V$, we can determine a lower bound for the maximum log-likelihood ratio of two images that differ in only one pixel, which if large is consistent with the definition of instability detailed above.
 
@@
 
-expand.grid(theta_h1 = seq(-5, 5, length.out = 5),
-            theta_v1 = seq(-5, 5, length.out = 5),
-            theta_11 = seq(-5, 5, length.out = 11)) %>%
+expand.grid(theta_h1 = seq(-5, 5, length.out = 21),
+            theta_v1 = seq(-5, 5, length.out = 21),
+            theta_11 = seq(-5, 5, length.out = 5)) %>%
   data.frame() %>%
@@
   ungroup() %>%
-  mutate(theta_h1 = factor(theta_h1, labels = paste0("theta[h1]==", unique(theta_h1))), 
-         theta_v1 = factor(theta_v1, labels = paste0("theta[v1]==", unique(theta_v1)))) %>%
+  mutate(theta_11 = factor(theta_11, labels = paste0("theta[11]==", unique(theta_11))),
+         mu_parameter = factor(mu_parameter, labels = paste0(gsub("_", "(theta[", unique(mu_parameter)), "])"))) %>%
   ggplot() +
-  geom_line(aes(theta_11, value, colour = mu_parameter)) +
-  facet_grid(theta_h1 ~ theta_v1, labeller = label_parsed) +
-  xlab(expression(theta[11])) +
-  scale_colour_discrete(expression(mu(theta)), labels = c(expression(theta[v1]), 
-                                                          expression(theta[h1]), 
-                                                          expression(theta[11])))
-
+  geom_tile(aes(theta_v1, theta_h1, fill = value))  +
+  geom_contour(aes(theta_v1, theta_h1, z = value), colour = "black") +
+  facet_grid(mu_parameter ~ theta_11, labeller = label_parsed) +
+  xlab(expression(theta[v1])) +
+  ylab(expression(theta[h1])) +
+  scale_fill_gradient2("Mean-parameterization", low = "#af8dc3", high = "#7fbf7b", midpoint = .5, limits = c(0, 1)) +
+  theme(aspect.ratio = 1)
 ```
 
-Secondly, we can look at $\max\limits_{\boldsymbol x_N \in \mathcal{X}_N}[Q(\boldsymbol x_N)]$ for this tiny model.
+Secondly, we can look at $R(\boldsymbol \theta)$ for this tiny model, which gives us a lower bound for the loglikelihood ratio of the smallest possible change in $\boldsymbol x$, indicating instability.
 
 \begin{align*}
-\max\limits_{\boldsymbol x_N \in \mathcal{X}_N}[Q(\boldsymbol x_N)] &= \max\limits_{\boldsymbol x \in \mathcal{X}}\left[ \theta_{11} h_1 v_1 + \theta_{h_1}h_1 + \theta_{v_1} v_1 \right] \\
+R(\boldsymbol \theta) &= \max_{\boldsymbol v} \max_{\boldsymbol h}Q(\boldsymbol x) - \min_{\boldsymbol v}\max_{\boldsymbol h}Q(\boldsymbol x) - H\log 2\\
+&= \max_{\boldsymbol v} \max_{\boldsymbol h}\left[ \theta_{11} h_1 v_1 + \theta_{h_1}h_1 + \theta_{v_1} v_1 \right] -  \min_{\boldsymbol v}\max_{\boldsymbol h}\left[ \theta_{11} h_1 v_1 + \theta_{h_1}h_1 + \theta_{v_1} v_1 \right] -\log 2\\
 &=\max\left\{ \theta_{11} + \theta_{h_1} + \theta_{v_1}, \theta_{h_1}, \theta_{v_1}, 0  \right\} \\
+&\qquad - \min\left\{\theta_{v_1} + \max\left\{\theta_{11} + \theta_{h1}, -\theta_{11} - \theta_{h1}\right\}, -\theta_{v_1} + \max\left\{-\theta_{11} + \theta_{h1}, \theta_{11} - \theta_{h1}\right\}\right\} - \log2\\
 \end{align*}
 
-Figure \ref{fig:instab} shows $\max\limits_{\boldsymbol x_N \in \mathcal{X}_N}[Q(\boldsymbol x_N)]/N$ for various values of $\boldsymbol \theta$.
-Of course, we are interested in how this quantity grows as $N = H+V$ grows for various parameter values.
-We can see that this quantity is increasing for larger positive values of $\boldsymbol \theta$, which may indicate instability as $|\boldsymbol \theta|$ is large, agreeing with the indications about near-degeneracy from looking at $\mu(\boldsymbol \theta)$.
+Figure \ref{fig:instab} shows $R(\boldsymbol \theta)/V$ for various values of $\boldsymbol \theta$.
+We can see that this quantity is increasing for larger magnitudes of $\boldsymbol \theta$, especially for large values of $\theta_11$.
+This may indicate instability as $|\boldsymbol \theta|$ is large, agreeing with the indications about near-degeneracy from looking at $\mu(\boldsymbol \theta)$.
 
-```{r instab, fig.cap="\\label{fig:instab}$(\\max\\limits_{\\boldsymbol x_N \\in \\mathcal{X}_N}[Q(\\boldsymbol x_N)] - \\min\\limits_{\\boldsymbol x_N \\in \\mathcal{X}_N}[Q(\\boldsymbol x_N)])/N$ for various values of $\\boldsymbol \\theta$ for the tiny example model.
+```{r instab, fig.cap="\\label{fig:instab}$R(\\boldsymbol \\theta)/V$ for various values of $\\boldsymbol \\theta$ for the tiny example model.
 This quantity is increasing for larger positive values of $\\boldsymbol \\theta$.", fig.height = 4, fig.align='center'}
-expand.grid(theta_h1 = seq(-5, 5, length.out = 5),
-            theta_v1 = seq(-5, 5, length.out = 5),
-            theta_11 = seq(-5, 5, length.out = 31)) %>%
+expand.grid(theta_h1 = seq(-5, 5, length.out = 21),
+            theta_v1 = seq(-5, 5, length.out = 21),
+            theta_11 = seq(-5, 5, length.out = 5)) %>%
   data.frame() %>%
   group_by(theta_v1, theta_h1, theta_11) %>%
-  do(data.frame(q_max = max(possibles %*% t(data.matrix(.))) - min(possibles %*% t(data.matrix(.))))) %>%
+  do(data.frame(Q = cbind(possibles, possibles %*% t(data.matrix(.))))) %>%
+  group_by(theta_v1, theta_h1, theta_11, Q.v1) %>%
+  do(data.frame(maxq = max(.$Q.V4))) %>%
+  group_by(theta_v1, theta_h1, theta_11) %>%
+  do(data.frame(q_max = max(.$maxq) - min(.$maxq) - log(2))) %>%
   ungroup() %>%
-  mutate(theta_h1 = factor(theta_h1, labels = paste0("theta[h1]==", unique(theta_h1))), 
-         theta_v1 = factor(theta_v1, labels = paste0("theta[v1]==", unique(theta_v1)))) %>%
+  mutate(theta_11 = factor(theta_11, labels = paste0("theta[11]==", unique(theta_11)))) %>%
   ggplot() +
-  geom_line(aes(theta_11, q_max/2)) +
-  facet_grid(theta_h1 ~ theta_v1, labeller = label_parsed) +
-  xlab(expression(theta[11])) +
-  ylab(expression(frac(max(Q) - min(Q), H + V)))
+  geom_tile(aes(theta_v1, theta_h1, fill = q_max)) +
+  geom_contour(aes(theta_v1, theta_h1, z = q_max), colour = "black") +
+  facet_grid(. ~ theta_11, labeller = label_parsed) +
+  xlab(expression(theta[v1])) +
+  ylab(expression(theta[h1])) +
+  scale_fill_gradient(expression(frac(R(theta), V)), low = "yellow", high = "red") +
+  theme(aspect.ratio = 1)
 ```
@@
 
-```{r uninterp, fig.cap="\\label{fig:uninterp}The difference between model expectations, E$\\left[\\boldsymbol X | \\boldsymbol \\theta\\right]$, and expectations given independence, E$\\left[\\boldsymbol X | \\emptyset\\right ]$ for a RBM with one visible and one hidden node.", fig.height = 4, fig.align='center', message=FALSE, cache=TRUE}
-possibles <- stats(1, 1, type = "binary")
-
-expand.grid(theta_h1 = seq(-5, 5, length.out = 5),
-            theta_v1 = seq(-5, 5, length.out = 5),
-            theta_11 = seq(-5, 5, length.out = 11)) %>%
+```{r uninterp, fig.cap="\\label{fig:uninterp}The difference between model expectations, E$\\left[\\boldsymbol X | \\boldsymbol \\theta\\right]$, and expectations given independence, E$\\left[\\boldsymbol X | \\emptyset\\right ]$ for a RBM with one visible and one hidden node.", fig.height = 4, fig.align='center', message=FALSE, cache=TRUE, warning=FALSE}
+expand.grid(theta_h1 = seq(-5, 5, length.out = 21),
+            theta_v1 = seq(-5, 5, length.out = 21),
+            theta_11 = seq(-5, 5, length.out = 5)) %>%
   data.frame() %>%
@@
 
-expand.grid(theta_h1 = seq(-5, 5, length.out = 5),
-            theta_v1 = seq(-5, 5, length.out = 5),
+expand.grid(theta_h1 = seq(-5, 5, length.out = 21),
+            theta_v1 = seq(-5, 5, length.out = 21),
             theta_11 = 0) %>%
@@
   ungroup() %>%
-  mutate(theta_h1 = factor(theta_h1, labels = paste0("theta[h1]==", unique(theta_h1))), 
-         theta_v1 = factor(theta_v1, labels = paste0("theta[v1]==", unique(theta_v1)))) %>%
+  mutate(theta_11 = factor(theta_11, labels = paste0("theta[11]==", unique(theta_11))),
+         mu_parameter = factor(mu_parameter, labels = paste0(gsub("_", "(theta[", unique(mu_parameter)), "])"))) %>%
   ggplot() +
-  geom_line(aes(theta_11, value, colour = mu_parameter)) +
-  facet_grid(theta_h1 ~ theta_v1, labeller = label_parsed) +
-  xlab(expression(theta[11])) + ylab(expression(group("|", E(group("", bold(X), "|")*bold(theta)) - E(group("", bold(X), "|")*plain(independence)), "|"))) +
-  scale_colour_discrete(expression(bold(X)), labels = c(expression(v[1]), 
-                                                          expression(h[1])))
-
+  geom_tile(aes(theta_v1, theta_h1, fill = value))  +
+  geom_contour(aes(theta_v1, theta_h1, z = value), colour = "black") +
+  facet_grid(mu_parameter ~ theta_11, labeller = label_parsed) +
+  xlab(expression(theta[v1])) +
+  ylab(expression(theta[h1])) +
+  scale_fill_gradient(expression(group("|", E(group("", bold(X), "|")*bold(theta)) - E(group("", bold(X), "|")*plain(independence)), "|")), low = "yellow", high = "red", limits = c(0, 1)) +
+  theme(aspect.ratio = 1)
 ```
 
-Figure \ref{fig:uninterp} shows that the absolute difference between model expectations, E$\left[\boldsymbol X | \boldsymbol \theta\right]$, and expectations given independence, E$\left[\boldsymbol X | \emptyset\right ]$ for the RBM with one hidden and one visible node grows as the values of $\boldsymbol \theta$ aare farther from zero.
-This is a third indication that parameters of large magnetude could lead to model impropriety in an RBM using a model with only one hidden and one visible node.
+Figure \ref{fig:uninterp} shows that the absolute difference between model expectations, E$\left[\boldsymbol X | \boldsymbol \theta\right]$, and expectations given independence, E$\left[\boldsymbol X | \emptyset\right ]$ for the RBM with one hidden and one visible node grows as the values of $\boldsymbol \theta$ are farther from zero, especially for large magnitudes of $\theta_11$.
+This is a third indication that parameters of large magnitude could lead to model impropriety in an RBM using a model with only one hidden and one visible node.
 
@@
 
-```{r instab_plots, fig.cap="\\label{fig:instab_plots}A numerical experiment looking at $(\\max\\limits_{\\boldsymbol x_N \\in \\mathcal{X}_N}[Q(\\boldsymbol x_N)] - \\min\\limits_{\\boldsymbol x_N \\in \\mathcal{X}_N}[Q(\\boldsymbol x_N)])/N$ for each combination of magnitude of $\\boldsymbol \\theta$ and model size.
-As the magnitude of $\\boldsymbol \\theta$ grows, so does the value of this metric.", fig.height=4}
-p.max_q
-```
-
 ```{r instab_plots_2, fig.cap="\\label{fig:instab_plots}A numerical experiment looking at $(\\max\\limits_{\\boldsymbol x_N \\in \\mathcal{X}_N}[Q(\\boldsymbol x_N)] - \\min\\limits_{\\boldsymbol v} \\max\\limits_{\\boldsymbol h}[Q(\\boldsymbol x_N)] - H\\log(2))/N$ for each combination of magnitude of $\\boldsymbol \\theta$ and model size.
@@
 **TODO: Add proof that likelihood goes to infinity.**  
-**TODO: Add marginalized Jing prior?**  
 **TODO: Add comparison to contrastive divergence**
@@
 
+##Center of the universe
+
+##Incomplete cases
+
 #References

### 2016-06-22 writing/draft.Rmd
more updates

@@
 \end{matrix} \right] \right|\\
-&= 
+&= \left[
 \begin{matrix}
@@
 \frac{\left|\exp\left(\theta_{11} + 2\theta_{v_1} + \theta_{h_1}\right) - \exp\left( 2\theta_{v_1} + \theta_{h_1}\right) \right| }{\left(\exp\left(\theta_{v_1}\right) + \exp\left(\theta_{h_1}\right) + \exp\left(\theta_{11} + \theta_{v_1} + \theta_{h_1}\right)\right)\left(\exp\left(\theta_{v_1}\right) + \exp\left(\theta_{h_1}\right) + \exp\left(+ \theta_{v_1} + \theta_{h_1}\right)\right)}
-\end{matrix}
+\end{matrix}\right]
 \end{align*}
@@
   ylab(expression(theta[h1])) +
-  scale_fill_gradient(expression(group("|", E(group("", bold(X), "|")*bold(theta)) - E(group("", bold(X), "|")*plain(independence)), "|")), low = "yellow", high = "red", limits = c(0, 1)) +
+  scale_fill_gradient(expression(group("|", E(group("", bold(X), "|")*bold(theta)) - E(group("", bold(X), "|")*symbol("\306")), "|")), low = "yellow", high = "red", limits = c(0, 1)) +
   theme(aspect.ratio = 1)
@@
 To explore the behavior of the RBM parameters $\boldsymbol \theta$ as it related to model degeneracy and near degeneracy, we example various models of small size.
-For $H, V \in \{1, \dots, 4\}$, we sample uniformly `r dim(res$samp[[1]])[1]` values of $\boldsymbol \theta$ with magnitude, $||\boldsymbol \theta || = \sqrt{\sum_{i = 1, \dots, H + V + H*V} \theta_i^2}$, between `r min(plot_dat_grid$r1)` and `r max(plot_dat_grid$r1)`.
-For each set of parameters we then calculate the three metrics of model impropriety introduced in section \ref{degeneracy-instability-and-uninterpretability-oh-my}, $\mu(\boldsymbol \theta)$, $\max\limits_{\boldsymbol x_N \in \mathcal{X}_N}[Q(\boldsymbol x_N)]/N$, and $\left|\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] -   \text{E}\left[\boldsymbol X | \emptyset \right] \right|$.
+For $H, V \in \{1, \dots, 4\}$, we sample uniformly `r dim(res$samp[[1]])[1]` values of $\boldsymbol \theta$ with various magnitudes.
+For each set of parameters we then calculate the three metrics of model impropriety introduced in section \ref{degeneracy-instability-and-uninterpretability-oh-my}, $\mu(\boldsymbol \theta)$, $R(\boldsymbol x)/V$, and $\left|\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] -   \text{E}\left[\boldsymbol X | \emptyset \right] \right|$.
 In the case of near-degeneracy, we can go further and classify each model as near-degenerate or viable based on the distance of $\mu(\boldsymbol \theta)$ from the convex hull of $\{t_k(\boldsymbol x): \boldsymbol x \in \mathcal{X}, k = 1, \dots, K\}$ and look at the fraction of models that were near-degenerate for each combination of magnitude of $\boldsymbol \theta$ and model size.
 
-In this numerical experiment, we split $\boldsymbol \theta$ into $\boldsymbol \theta_{interaction}$ and $\boldsymbol \theta_{main}$, in reference to which sufficient statistics the parameters correspond to, and allow the two types of terms to have varying magnitudes.
+In this numerical experiment, we split $\boldsymbol \theta$ into $\boldsymbol \theta_{interaction}$ and $\boldsymbol \theta_{main}$, in reference to which sufficient statistics the parameters correspond to, and allow the two types of terms to have varying average magnitudes, $||\boldsymbol \theta_{main} || /(H+V)$ and $||\boldsymbol \theta_{interaction} || /(H*V)$.
+These magnitudes vary on a grid between `r min(plot_dat_grid$r1/(plot_dat_grid$H + plot_dat_grid$V))` and `r max(plot_dat_grid$r1/(plot_dat_grid$H + plot_dat_grid$V))`.
+By looking at the average magnitudes, we are able to consider the potential benefit of shrinking each parameter value $\theta_i$ towards zero in a Bayesian fitting technique.
 The results of this numerical study can be seen in figures \ref{fig:degen_plots}, \ref{fig:instab_plots}, and \ref{fig:uninterp_plots}.
@@
   geom_abline(aes(intercept = 0, slope = 1), alpha = .5, lty = 2) +
-  scale_fill_gradient(expression(group("|", E(group("", bold(X), "|")*bold(theta)) - E(group("", bold(X), "|")*plain(independence)), "|")), low = "yellow", high = "red", limits = c(0,2)) +
+  scale_fill_gradient(expression(group("|", E(group("", bold(X), "|")*bold(theta)) - E(group("", bold(X), "|")*symbol("\306")), "|")), low = "yellow", high = "red", limits = c(0,2)) +
   facet_grid(Visibles~Hiddens) +
@@
   geom_abline(aes(intercept = 0, slope = 1), alpha = .5, lty = 2) +
-  scale_fill_gradient(expression(frac(max(Q) - min[v](max[h](Q)) - Hlog(2), H + V)), low = "yellow", high = "red") +
+  scale_fill_gradient(expression(frac(R(theta), V)), low = "yellow", high = "red") +
   facet_grid(Visibles~Hiddens) +
@@
 
-```{r instab_plots_2, fig.cap="\\label{fig:instab_plots}A numerical experiment looking at $(\\max\\limits_{\\boldsymbol x_N \\in \\mathcal{X}_N}[Q(\\boldsymbol x_N)] - \\min\\limits_{\\boldsymbol v} \\max\\limits_{\\boldsymbol h}[Q(\\boldsymbol x_N)] - H\\log(2))/N$ for each combination of magnitude of $\\boldsymbol \\theta$ and model size.
-As the magnitude of $\\boldsymbol \\theta$ grows, so does the value of this metric.", fig.height=4}
+```{r instab_plots_2, fig.cap="\\label{fig:instab_plots}A numerical experiment looking at $R(\\boldsymbol \\theta)/V$ for each combination of magnitude of $\\boldsymbol \\theta$ and model size.
+As the magnitude of $\\boldsymbol \\theta$ grows, so does the value of this metric, indicating instability in the model.", fig.height=4}
 p.min_max_q
@@
 
-```{r uninterp_plots, fig.cap="\\label{fig:split_plots}The difference between model expectations, E$\\left[\\boldsymbol X | \\boldsymbol \\theta\\right]$, and expectations given independence, E$\\left[\\boldsymbol X | \\emptyset\\right ]$ for each combination of magnitude of $\\boldsymbol \\theta$ and model size.
+```{r uninterp_plots, fig.cap="\\label{fig:uninterp_plots}The difference between model expectations, E$\\left[\\boldsymbol X | \\boldsymbol \\theta\\right]$, and expectations given independence, E$\\left[\\boldsymbol X | \\emptyset\\right ]$ for each combination of magnitude of $\\boldsymbol \\theta$ and model size.
 Larger magnitudes of $\\boldsymbol \\theta$ correspond to larger differences, thus indicating higher uninterpretability.", fig.height=4}

### 2016-07-07 writing/draft.Rmd
revisiting 4x4 model with theta based on grid

@@
 marginal_good_1 <- models_good
-marginal_degen_1 <- models_degen
 
@@
 jing_good <- models_good
-jing_degen <- models_degen
 
@@
 shrunk_jing_good <- models_good
-shrunk_jing_degen <- models_degen
 

### 2016-07-07 4x4 model/choose_params.R

@@ -3,67 +3,28 @@ library(dplyr)
 library(tidyr)
 library(ggplot2)
 
-#function -----------------------
-#this function takes the results from the simulation study and rearranges for ease of use
-rearrange_data <- function(res) {
-  
-  data <- data.frame()
-  for(i in 1:nrow(res)) {
-    tmp <- res$samp[[i]]
-    
-    H <- res[i,]$H
-    V <- res[i,]$V
-    
-    if(H == 4 & V == 4) { #only looking at 4x4 model
-      tmp %>% 
-        data.frame() %>% 
-        rowwise() %>% 
-        mutate_(ss_ratio = paste0("(", paste(paste0(names(.)[(H + V + 1):(H + V + H*V)], "^2"), collapse = " + "), ")/sum(", paste(paste0(names(.)[1:(H+V)],"^2"), collapse = " + "), ")")) %>%    
-        ungroup() -> ratio
-      
-      data <- rbind(data, inner_join(ratio, res$outside[[i]] %>% ungroup()) %>%
-        mutate(H = H, V = V, n_param = H + V + H*V))
-    }
-  }
-  return(data)
-}
-
-#do it -------------------------
-#uses the function to rearrange the data
-n <- 100
-r_center <- seq(0.2, 3, by = 0.2)
-r_exp <- .05
-
-plot.dat <- list()
-for(i in seq(0, 1, by = 0.1)) {
-  load(paste0("../expected value/apps/results_", i, ".RData")) 
-  plot_dat <- lapply(res, rearrange_data)
-  names(plot_dat) <- gsub("[.]", "_", r_center)
-  
-  do.call(rbind, plot_dat) %>% 
-    mutate(name = rownames(do.call(rbind, plot_dat))) %>%
-    separate(name, into = c("multiplier", "num"), "[.]") %>% 
-    mutate(multiplier = as.numeric(gsub("_", ".", multiplier))) %>%
-    mutate(n_param_f = factor(n_param)) %>%
-    select(-starts_with("exp")) -> plot_dat
-  
-  plot.dat[[i*10 + 1]] <- plot_dat
-}
-
-names(plot.dat) <- gsub("[.]", "_", seq(0, 1, by = 0.1))
-do.call(rbind, plot.dat) %>%
-  mutate(name2 = rownames(do.call(rbind, plot.dat))) %>%
-  separate(name2, into = c("power", "num2"), "[.]") %>% 
-  mutate(power = as.numeric(gsub("_", ".", power))) -> plot.dat
[…]

### 2016-07-07 4x4 model/choose_variance_params.R

@@ -3,51 +3,21 @@ library(dplyr)
 library(tidyr)
 library(ggplot2)
 
-#reshape data ---------------------
-load("../expected value/apps/results_split.RData")
-plot_data <- function(res) {
-  
-  plot.data <- data.frame()
-  for(i in 1:nrow(res)) {
-    tmp <- res$g_theta[[i]] %>% data.frame()
-    H <- res[i,]$H
-    V <- res[i,]$V
-    N <- res[i,]$N
-    r1 <- res[i,]$r1
-    r2 <- res[i,]$r2
-    C <- res[i,]$C
-    epsilon <- res[i,]$epsilon
-    
-    
-    tmp %>% 
-      rowwise() %>% 
-      mutate_(ss_ratio = paste0("(", paste(paste0(names(.)[(H + V + 1):(H + V + H*V)], "^2"), collapse = " + "), ")/sum(", paste(paste0(names(.)[1:(H+V)],"^2"), collapse = " + "), ")")) %>%    
-      ungroup() -> ratio
-    
-    
-    inner_join(ratio, res$outside[[i]] %>% ungroup()) %>%
-      select(ss_ratio, near_hull) %>%
-      mutate(H = H, V = V, n_param = H + V + H*V, N = H + V, N = N, r1 = r1, r2 = r2, C = C, epsilon = epsilon) %>%
-      rbind(plot.data) -> plot.data
-  }
-  return(plot.data)
-}
-plot_dat <- res %>% plot_data
+#load data -------------------------
+load("../degeneracy/written/results_grid.RData")
 
-#choose C, C_prime --------------
-plot_dat %>%
+#find appropriate radii and correlate with variance constants
+res %>%
   filter(H == 4 & V == 4) %>%
-  group_by(H, V, C, epsilon) %>%
+  group_by(H, V, r1, r2) %>%
+  do(data.frame(.$outside)) %>%
+  group_by(H, V, r1, r2) %>%
   summarise(percent_degen = sum(near_hull)/n()) %>%
-  mutate(C_prime = C*((H + V)/(H*V))^(1+epsilon)) %>%
-  mutate(percent_degen = round(percent_degen, 2)) %>%
-  group_by(percent_degen) %>%
-  mutate(C_C_prime = C + C_prime) %>%
-  top_n(1, C_C_prime) %>%
-  arrange(percent_degen, C_prime, C) %>%
+  filter(percent_degen <= .05 & r1 > r2) %>%
[…]

### 2016-07-07 4x4 model/fit_gibbs.R

@@ -19,27 +19,31 @@ params <- list(main_hidden = rnorm(H, mean = 0, sd = sqrt(variance_params$C)),
                interaction = matrix(rnorm(H*V, mean = 0, sd = sqrt(variance_params$C_prime)), nrow = H))
 
 N <- nrow(flat_images_good$visibles)
-const <- 2
-models_good <- sample_gibbs(visibles = flat_images_good$visibles, params0 = params, C = const*variance_params$C/N, C_prime = const*variance_params$C_prime/N, mc.iter = mc.iter)
-models_bad <- sample_gibbs(visibles = flat_images_good$visibles, params0 = params, C = const*variance_params$C/N, C_prime = const*variance_params$C/N, mc.iter = mc.iter)
-models_degen <- sample_gibbs(visibles = flat_images_degen$visibles, params0 = params, C = const*variance_params$C/N, C_prime = const*variance_params$C_prime/N, mc.iter = mc.iter)
+consts <- 6:8 #tune constant 
 
-#burnin + thinning
-burnin_thinning <- function(models) {
-  idx <- which(1:mc.iter %% 5 == 0 & 1:mc.iter > mc.iter/2)
-  models$params$main_hidden <- models$params$main_hidden[idx, ]
-  models$params$main_visible <- models$params$main_visible[idx, ]
-  models$params$interaction <- models$params$interaction[idx, , ]
-  models$hiddens <- models$hiddens[idx, , ]
-  #models$post_pred <- models$post_pred[idx, , ]
-  models$distn <- models$distn[idx, , ]
+for(const in consts) {
+  models_good <- sample_gibbs(visibles = flat_images_good$visibles, params0 = params, C = const*variance_params$C/N, C_prime = const*variance_params$C_prime/N, mc.iter = mc.iter)
+  models_bad <- sample_gibbs(visibles = flat_images_good$visibles, params0 = params, C = const*variance_params$C/N, C_prime = const*variance_params$C/N, mc.iter = mc.iter)
+  #models_degen <- sample_gibbs(visibles = flat_images_degen$visibles, params0 = params, C = const*variance_params$C/N, C_prime = const*variance_params$C_prime/N, mc.iter = mc.iter)
   
-  return(models)
+  #burnin + thinning
+  burnin_thinning <- function(models) {
+    idx <- which(1:mc.iter %% 5 == 0 & 1:mc.iter > mc.iter/2)
+    models$params$main_hidden <- models$params$main_hidden[idx, ]
[…]

### 2016-07-07 4x4 model/fit_single_adaptive_mh_within_gibbs_marginal.R

@@ -33,9 +33,9 @@ trunc_const <- 1
 
 models_good <- sample_single_adaptive_mh_within_gibbs(visibles = flat_images_good$visibles, params0 = params, C = variance_params$C, C_prime =  variance_params$C_prime, trunc_const = trunc_const, h = h, s1 = s1, s2 = s2, mc.iter = mc.iter, conditional_function = log_conditional_marginal)
 models_bad <- sample_single_adaptive_mh_within_gibbs(visibles = flat_images_good$visibles, params0 = params, C = variance_params$C, C_prime = variance_params$C, trunc_const = trunc_const, h = h, s1 = s1, s2 = s2, mc.iter = mc.iter, conditional_function = log_conditional_marginal)
-models_degen <- sample_single_adaptive_mh_within_gibbs(visibles = flat_images_degen$visibles, params0 = params, C = variance_params$C, C_prime = variance_params$C_prime, trunc_const = trunc_const, h = h, s1 = s1, s2 = s2, mc.iter = mc.iter, conditional_function = log_conditional_marginal)
+#models_degen <- sample_single_adaptive_mh_within_gibbs(visibles = flat_images_degen$visibles, params0 = params, C = variance_params$C, C_prime = variance_params$C_prime, trunc_const = trunc_const, h = h, s1 = s1, s2 = s2, mc.iter = mc.iter, conditional_function = log_conditional_marginal)
 
-save(models_good, models_bad, models_degen, file = "written/fitted_models_adaptive_mh_full_trunc_distn_marginal_3.Rdata")
+save(models_good, models_bad, models_degen, file = "written/fitted_models_adaptive_mh_full_trunc_distn_marginal.Rdata")
 
 #burnin + thinning
 burnin_thinning <- function(models) {
@@ -51,6 +51,6 @@ burnin_thinning <- function(models) {
 
 models_good <- burnin_thinning(models_good)
 models_bad <- burnin_thinning(models_bad)
-models_degen <- burnin_thinning(models_degen)
+#models_degen <- burnin_thinning(models_degen)
 
-save(models_good, models_bad, models_degen, file = "written/fitted_models_adaptive_mh_trunc_distn_marginal_3.Rdata")
+save(models_good, models_bad, models_degen, file = "written/fitted_models_adaptive_mh_trunc_distn_marginal.Rdata")

### 2016-07-07 4x4 model/test_fitted_model.R

@@ -10,7 +10,7 @@ V <- 4
 
 load("written/sample_images.Rdata")
 # load("written/fitted_models_adaptive_mh_trunc_distn.Rdata")
-load("written/fitted_models_adaptive_mh_full_trunc_distn_marginal_2.Rdata")
+load("written/fitted_models_jing_7.Rdata")
 load("written/sample.params.Rdata")
 params_degen <- list(main_hidden = sample.params %>% ungroup() %>% filter(near_hull) %>% select(starts_with("h"), -H) %>% data.matrix(),
                      main_visible = sample.params %>% ungroup() %>% filter(near_hull) %>% select(starts_with("v"), -V) %>% data.matrix(),
@@ -85,7 +85,7 @@ reshape_sample_distn <- function(model) {
   sample_distn <- model$distn
   dim(sample_distn) <- c(dim(sample_distn)[1]*dim(sample_distn)[2], dim(sample_distn)[3])
   sample_distn %>% data.frame() -> sample_distn
-  names(sample_distn) <- names(distn_degen)
+  names(sample_distn) <- names(distn_good)
   
   sample_distn %>%
     group_by(image_id) %>%
@@ -155,7 +155,5 @@ data.frame(good = colSums(exp(possibles %*% t(models_good$theta))),
   gather(model, normalizer, -iter) %>%
   ggplot() +
   geom_line(aes(iter, normalizer, colour = model))
-
-  
 
 

### 2016-07-07 4x4 model/tune_mh_within_gibbs.R

@@ -8,7 +8,7 @@ source("sample_functs.R")
 H <- 4
 V <- 4
 set.seed(102285) #reproducible seed
-mc.iter <- 500
+mc.iter <- 200
 
 load("written/sample_images.Rdata")
 load("written/variance_params.Rdata")
@@ -19,7 +19,7 @@ params <- list(main_hidden = rnorm(H, mean = 0, sd = sqrt(variance_params$C)),
                main_visible = rnorm(V, mean = 0, sd = sqrt(variance_params$C)),
                interaction = matrix(rnorm(H*V, mean = 0, sd = sqrt(variance_params$C_prime)), nrow = H))
 
-# data.frame(expand.grid(tau_main = seq(.1, 1, by = .1), tau_interaction = seq(.1, 1, by = .1))) %>%
+# data.frame(expand.grid(tau_main = seq(.001, .02, by = .002), tau_interaction = seq(.001, .02, by = .002))) %>%
 #   group_by(tau_main, tau_interaction) %>%
 #   do(model = sample_mh_within_gibbs(visibles = flat_images_good$visibles, params0 = params, C = variance_params$C, C_prime = variance_params$C_prime, tau_main = .$tau_main, tau_interaction = .$tau_interaction, mc.iter = mc.iter)) -> models
 # 

### 2016-07-28 presentations/coda2016/coda2016_poster.tex
jsm presentation

@@
 opacity=1,  %% adjust
-contents={\includegraphics[width=\paperwidth,height=\paperheight]{images/coda2016_poster_background.pdf}}
+contents={\includegraphics[width=\paperwidth,height=\paperheight]{../images/coda2016_poster_background.pdf}}
 }
@@
   \centering
-  \resizebox{\linewidth}{!}{\input{images/rbm.tikz}}
+  \resizebox{\linewidth}{!}{\input{../images/rbm.tikz}}
   \caption{An example restricted Boltzmann machine (RBM), which consists of two layers, a hidden ($\mathcal{H}$) and a visible layer ($\mathcal{V}$), with no connections within a layer.
@@
   \centering
-  \input{images/visibles.tikz}
+  \input{../images/visibles.tikz}
   \caption{Image classification using a RBM.
@@
   \centering
-  \input{images/deep_rbm.tikz}
+  \input{../images/deep_rbm.tikz}
   \caption{Deep RBM example.}
@@
     \tdplotsetmaincoords{60}{-60}
-    \input{images/toyhull_top.tikz}
+    \input{../images/toyhull_top.tikz}
 
@@
     \tdplotsetmaincoords{60}{-60}
-    \input{images/toyhull_top_2.tikz}
+    \input{../images/toyhull_top_2.tikz}
 
@@
 \centering
-\includegraphics[scale=1]{images/frac_volume.pdf}
+\includegraphics[scale=1]{../images/frac_volume.pdf}
 \label{fig:frac_volume}
@@
 \centering
-\includegraphics{images/degeneracy.pdf}
-\includegraphics{images/instability.pdf}
-\includegraphics{images/uninterpretability.pdf}
+\includegraphics{../images/degeneracy.pdf}
+\includegraphics{../images/instability.pdf}
+\includegraphics{../images/uninterpretability.pdf}
 \label{fig:three_ways}
@@
 \centering
-\includegraphics{images/image_prediction.pdf}
+\includegraphics{../images/image_prediction.pdf}
 \label{fig:image_pred}

### 2016-07-28 presentations/jsm2016/index.Rmd
jsm presentation

---
title: "An Exposition on the Propriety of Restricted Boltzmann Machines"
author: http://bit.ly/jsm2016
ratio: 16x10
output:
  rmdshower::shower_presentation:
    self_contained: false
    theme: ribbon
    katex: true
    fig_caption: true
    css: style.css
bibliography: refs.bib
---

## An exposition on the propriety of restricted Boltzmann machines { .white }

<img src="images/title_background.jpg" class="cover" style="height: 100%" alt="https://www.flickr.com/photos/kevin53/13427959404/in/photolist-mszNvs-9MQYEj-4yexiR-ks7nip-6o3CdZ-ngtetn-canv9W-bmdAKd-4vidoh-6PLNA8-5gvtyL-3MDEAo-8SsP23-ks6CS4-oJ44cc-93LLtu-jYhtWc-e4nSK4-pJrmi4-62mmJ-dANTW2-8zyj8N-ks7nFP-cSDsyd-pmTfZt-9dL7vS-nyf63n-8q76xZ-dzE85u-63vAS1-djhyF4-pZKfE7-dr9oRc-8i1GXY-jmsNrz-4wC15t-4wBZMp-8XmJa8-co9uys-63nQSn-a5tzkd-fqbZm4-au3EME-mszUiu-sbVvKY-ajAP7X-9kUypA-o4stRL-J5n3or-7icjfY">

<p class="white">
Andee Kaplan, Daniel Nordman, and Stephen Vardeman<br/>
Iowa State University<br/>
July 31, 2016<br/>
JSM - Chicago, IL<br/>
</br></br></br></br>
<a href="http://bit.ly/jsm2016">http://bit.ly/jsm2016</a>
</p>

# Restricted Boltzmann machines

## What is this?

<div class="double">
<p class="double-flow">
Restricted Boltzmann machine (RBM), which consists of two layers - a hidden ($\mathcal{H}$) and a visible ($\mathcal{V}$).
</p>
<p class="double-flow">
<img src="images/rbm.pdf" class="one-col-image"></br>
</p>
</div>
<img src="images/visibles_image.pdf"></br>
Image classification using a RBM. Each image pixel comprises a node in the visible layer, $\mathcal{V}$. The output is used to create features which are then passed to a supervised learning algorithm [@friedman2001elements].

## Joint Distribution

Let $x = \{h_1, ..., h_H, v_1, ...,v_V\}$ represent the states of the visible and hidden nodes in an RBM. Then the probability each node taking the the value corresponding to $x$ is:

$$
f_{\theta} (x) = \frac{\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\sum\limits_{x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}
$$

## Deep learning

<div class="double">
<p class="double=flow">
"Deep Boltzmann machine" - multiple single layer restricted Boltzmann machines with the lower stack hidden layer acting as the visible layer for the higher stacked model
</p>
<p class="double-flow">
<img src="images/deep_rbm.pdf"></br>
</p>
</div><br/>

Claimed ability to learn "internal representations that become increasingly complex" [@salakhutdinov2009deep], used in classification problems.

# Degeneracy, instability, and uninterpretability. Oh my!

## Near-degeneracy

The highly flexible nature of the RBM ($H + V + HV$ parameters) makes three characteristics of model impropriety of particular concern.

Let $Q(\cdot)$ denotes the neg-potentional function of the model, having support set $\mathcal{S}$.

| Characteristic | Detection |
|:----------------------------------------------------------------|:-------------------------------------------------------|
|Disproportionate amount of probability placed on only a few elements of the sample space by the model [@handcock2003assessing] | If random variables in $Q(\cdot)$ have a collective mean $\mu(\theta)$ close to the boundary of the convex hull of $\mathcal{S}$. |

## Instability

Let $R(\theta) = \max_{v} \max_{h}Q(x) - \min_{v}\max_{h}Q(x) - H\log 2$.

| Characteristic | Detection |
|:----------------------------------------------------------------|:-------------------------------------------------------|
| Small changes in natural parameters result in large changes in probability masses, excessive sensitivity [@schweinberger2011instability]. | If $R(\theta)/V$ is large, then the the maximum log-likelihood ratio of two images that differ in only one pixel is large. |

## Uninterpretability

| Characteristic | Detection |
|:----------------------------------------------------------------|:-------------------------------------------------------|
| Due to the existence of dependence, marginal mean-structure no longer maintained [@kaiser2007statistical]. | If the magnitude of the difference between model expectations and expectations under independence (dependence parameters of zero), $\left\vert E( X \vert \theta) -E( X \vert \emptyset ) \right\vert$, is large. |

## Data coding to mitigate degeneracy

<div class="double">
<p class="double-flow">
<img src="images/toyhull_top.pdf">
<img src="images/toyhull_top_2.pdf">
</p>
<p class="double-flow">
Convex hulls of the statistic space for a toy RBM with $V = H = 1$ for $\{0,1\}$-encoding and $\{-1,1\}$-encoding enclosed by an unrestricted hull of 3-space.
</p>
</div>

- For the $\{-1, 1 \}$ encoding of $\mathcal{V}$ and $\mathcal{H}$, the origin is the center of the parameter space
- At $\theta = 0$, the RBM is equivalent to elements of $X$ being distributed as iid Bernoulli$\left(\frac{1}{2}\right)$ $\Rightarrow$ No *near-degeneracy*, *instability*, or *uninterpretability*!

## Manageable (a.k.a. small) examples

# Bayesian model fitting

# Thank you!

## References {.small}

### 2016-07-28 presentations/jsm2016/index.Rmd
update bitly

@@
 title: "An Exposition on the Propriety of Restricted Boltzmann Machines"
-author: http://bit.ly/jsm2016
+author: http://bit.ly/jsm2016-rbm
 ratio: 16x10
@@
 </br></br></br></br>
-<a href="http://bit.ly/jsm2016">http://bit.ly/jsm2016</a>
+<a href="http://bit.ly/jsm2016-rbm">http://bit.ly/jsm2016-rbm</a>
 </p>

### 2016-07-28 presentations/jsm2016/index.Rmd
change pdf to png

@@
 <p class="double-flow">
-<img src="images/rbm.pdf" class="one-col-image"></br>
+<img src="images/rbm.png" class="one-col-image"></br>
 </p>
 </div>
-<img src="images/visibles_image.pdf"></br>
+<img src="images/visibles_image.png"></br>
 Image classification using a RBM.
@@
 <p class="double-flow">
-<img src="images/deep_rbm.pdf"></br>
+<img src="images/deep_rbm.png"></br>
 </p>
@@
 <p class="double-flow">
-<img src="images/toyhull_top.pdf">
-<img src="images/toyhull_top_2.pdf">
+<img src="images/toyhull_top.png">
+<img src="images/toyhull_top_2.png">
 </p>

### 2016-07-28 presentations/jsm2016/index.Rmd
clearer images

@@
 <p class="double-flow">
-Restricted Boltzmann machine (RBM), which consists of two layers - a hidden ($\mathcal{H}$) and a visible ($\mathcal{V}$).
+Restricted Boltzmann machine (RBM) with two layers - hidden ($\mathcal{H}$) and visible ($\mathcal{V}$) [@friedman2001elements].
 </p>
 <p class="double-flow">
-<img src="images/rbm.png" class="one-col-image"></br>
+<img src="images/rbm.png" width="100%"></br>
 </p>
 </div>
-<img src="images/visibles_image.png"></br>
-Image classification using a RBM.
-Each image pixel comprises a node in the visible layer, $\mathcal{V}$.
-The output is used to create features which are then passed to a supervised learning algorithm [@friedman2001elements].
+<img src="images/visibles_image.png" width="100%"></br>
+Used for image classification. Each image pixel is a node in the visible layer. The output creates features, passed to supervised learning.
 
@@
 <p class="double-flow">
-<img src="images/deep_rbm.png"></br>
+<img src="images/deep_rbm.png" width="100%"></br>
 </p>
@@
 <p class="double-flow">
-<img src="images/toyhull_top.png">
-<img src="images/toyhull_top_2.png">
+<img src="images/toyhull_top.png" width="100%">
 </p>
 <p class="double-flow">
-Convex hulls of the statistic space for a toy RBM with $V = H = 1$ for $\{0,1\}$-encoding and $\{-1,1\}$-encoding enclosed by an unrestricted hull of 3-space.
+<img src="images/toyhull_top_2.png" width="70%">
 </p>
 </div>
+
+Convex hulls of the statistic space for a toy RBM with $V = H = 1$ for $\{0,1\}$ and $\{-1,1\}$-encoding enclosed by an unrestricted hull of 3-space.
+
+## Data coding to mitigate degeneracy (cont'd)
 

### 2016-07-28 presentations/jsm2016/index.Rmd
finished the first part

@@
 ---
+```{r libraries, echo=FALSE, message=FALSE, warning=FALSE}
+library(knitr)
+library(ggplot2)
+library(dplyr)
+library(tidyr)
+
+opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE)
+theme_set(theme_bw(base_family="serif"))
+
+source("../../writing/helpers/functions.R")
+```
 
@@
 
+```{r encoding-volume, echo = FALSE, fig.height=3}
+test_cases <- expand.grid(data.frame(1:4)[,rep(1,2)]) %>% 
+  rename(H = X1.4, V = X1.4.1) %>%
+  filter(H <= V) %>% #remove those cases with less visibles than hiddens. Is this necessary?
+  mutate(n_param = H*V + H + V) %>%
+  mutate(max_facets = (2^(H+V))^(floor(n_param/2))) %>%
+  filter(n_param <= 11) #calc_hull can't handle any higher dimensions currently
+
+bin_res <- plyr::dlply(test_cases, plyr::.(n_param), function(x) calc_hull(x$V, x$H, "binary"))
+neg_res <- plyr::dlply(test_cases, plyr::.(n_param), function(x) calc_hull(x$V, x$H, "negative"))
+
+plyr::ldply(bin_res, function(x) x$c_hull$vol) %>% 
+  mutate(frac_vol = V1/(1^n_param)) %>%
+  inner_join(plyr::ldply(neg_res, function(x) x$c_hull$vol) %>% 
+               mutate(frac_vol = V1/(2^n_param)),
+             by="n_param") %>%
+  rename(vol.bin = V1.x, vol.neg = V1.y, frac_vol.bin = frac_vol.x, frac_vol.neg = frac_vol.y) %>%
+  gather(vars, value, -n_param) %>%
+  separate(vars, c("type", "encoding"), "\\.") %>%
+  spread(type, value) %>%
+  ggplot() +
+  geom_point(aes(x=n_param, y=frac_vol, colour=encoding)) +
+  geom_line(aes(x=n_param, y=frac_vol, colour=encoding, group=encoding)) +
+  ylab("Fraction of unrestricted volume") +
+  xlab("Number of parameters") +
+  scale_colour_discrete("Encoding", labels=c("Binary (1,0)", "Negative (1,-1)")) +           
+  theme(
+    plot.background = element_rect(fill = "transparent", colour = NA),
+    legend.background = element_rect(fill = "transparent", colour = NA)) -> p
+
+p
+```
+
 ## Manageable (a.k.a. small) examples
 
+```{r three-ways, cache=TRUE, echo=FALSE}
+#reshape data functions
+plot_data <- function(res, grid = FALSE) {
+
+  plot.data <- data.frame()
+
+  if(!grid) {
+    for(i in 1:nrow(res)) {
+      tmp <- res$g_theta[[i]] %>% data.frame()
+      H <- res[i,]$H
+      V <- res[i,]$V
+      N <- res[i,]$N
+      r1 <- res[i,]$r1
+      r2 <- res[i,]$r2
+      C <- res[i,]$C
+      epsilon <- res[i,]$epsilon
+
+      tmp %>% 
+        rowwise() %>% 
+        mutate_(ss_interaction = paste0("(", paste(paste0(names(.)[(H + V + 1):(H + V + H*V)], "^2"), collapse = " + "), ")"),
+                ss_main = paste0("sum(", paste(paste0(names(.)[1:(H+V)],"^2"), collapse = " + "), ")")) %>%    
+        ungroup() -> ratio
+
+      inner_join(ratio, res$outside[[i]] %>% ungroup()) %>%
+        select(ss_interaction, ss_main, near_hull) %>%
+        mutate(H = H, V = V, n_param = H + V + H*V, N = H + V, N = N, r1 = r1, r2 = r2, C = C, epsilon = epsilon) %>%
+        rbind(plot.data) -> plot.data
+    }
+  } else {
+    for(i in 1:nrow(res)) {
+      tmp <- res$g_theta[[i]] %>% data.frame()
+      H <- res[i,]$H
+      V <- res[i,]$V
+      N <- res[i,]$N
+      r1 <- res[i,]$r1
+      r2 <- res[i,]$r2
+
+      tmp %>% 
+        rowwise() %>% 
+        mutate_(ss_interaction = paste0("(", paste(paste0(names(.)[(H + V + 1):(H + V + H*V)], "^2"), collapse = " + "), ")"),
+                ss_main = paste0("sum(", paste(paste0(names(.)[1:(H+V)],"^2"), collapse = " + "), ")")) %>%  
+        ungroup() -> ratio
+
+      inner_join(ratio, res$outside[[i]] %>% ungroup()) %>%
+        select(ss_interaction, ss_main, near_hull) %>%
+        mutate(H = H, V = V, n_param = H + V + H*V, N = H + V, N = N, r1 = r1, r2 = r2) %>%
+        rbind(plot.data) -> plot.data
+    }
+  }
+
+  return(plot.data)
+}
+indep_params <- function(samp, H, V) {
+  samp[, (H + V + 1):ncol(samp)] <- 0
+  samp
+}
+max_Q <- function(theta, stats) {
+  apply(crossprod(t(stats), theta), 2, max)
+}
+min_Q <- function(theta, stats) {
+  apply(crossprod(t(stats), theta), 2, min)
+}
+min_max_Q <- function(theta, stats) {
+  require(dplyr)
+  tcrossprod(stats, theta) %>% data.frame() %>%
+    cbind(stats) %>%
+    group_by_(.dots = colnames(stats)[grepl("v", colnames(stats)) & !grepl("theta", colnames(stats))]) %>%
+    summarise_each(funs(max), contains("X")) %>%
+    ungroup() %>%
+    summarise_each(funs(min), contains("X")) %>%
+    select(contains("X")) %>%
+    data.matrix() %>%
+    t()
+}
+
+#grid data
+load("../../writing/data/results_grid.RData")
+
+#near-degeneracy
+plot_dat_grid <- res %>% plot_data(grid = TRUE)
+
+plot_dat_grid %>%
+  group_by(r1, r2, H, V) %>%
+  summarise(frac_degen = sum(near_hull)/n(), count = n()) %>% 
+  ungroup() %>% 
+  mutate(Hiddens = paste0("H: ", H), Visibles = paste0("V: ", V)) -> convex_hull_summary
+
+#uninterpretability
+res %>%
+  group_by(H, V, N, r1, r2) %>%
+  do(indep_exp = t(expected_value(t(indep_params(.$samp[[1]], .$H, .$V)), .$stat[[1]]))[, -((.$H + .$V + 1):(.$H + .$V + .$H*.$V))],
+     marg_exp = .$g_theta[[1]][, (.$H + .$V + .$H*.$V + 1):(ncol(.$g_theta[[1]])-.$H*.$V)]) -> exp_vals
+
+exp_vals %>%
+  group_by(H, V, N, r1, r2) %>%
+  do(data.frame(max_abs_diff = apply(abs(.$indep_exp[[1]] - .$marg_exp[[1]]), 1, max))) %>%
+  group_by(H, V, N, r1, r2) %>%
+  summarise(max_abs_diff = mean(max_abs_diff)) %>%
+  mutate(Hiddens = paste0("H: ", H), Visibles = paste0("V: ", V)) -> exp_vals_summary
+
+#instability
+res %>%
+  group_by(H, V, N, r1, r2) %>%
+  do(data.frame(max = max_Q(t(.$samp[[1]]), .$stat[[1]]),
+                min = min_Q(t(.$samp[[1]]), .$stat[[1]]),
+                min_max = min_max_Q(.$samp[[1]], .$stat[[1]]))) %>%
+  ungroup() %>% 
+  group_by(H, V, N, r1, r2) %>%
+  mutate(LHS1 = (max - min)/N,
+         LHS2 = (max - min_max - H*log(2))/N) %>%
+  summarise_each(funs(mean), LHS1, LHS2) -> max_q_summary
+
+convex_hull_summary %>%
+  left_join(max_q_summary) %>%
+  left_join(exp_vals_summary) -> three_ways
+```
+```{r degen-plots, echo=FALSE}
+three_ways %>%
+  mutate(r1 = round(r1/(H + V), 8), r2 = round(r2/(H * V), 8)) %>% 
+  ggplot() +
+  geom_tile(aes(x = r1, y = r2, fill = frac_degen)) +
+  geom_contour(aes(x = r1, y = r2, z = frac_degen), colour = "black", bins = 8) +
+  geom_contour(aes(x = r1, y = r2, z = frac_degen), colour = "black", breaks = .05, size = 1.5) +
+  geom_abline(aes(intercept = 0, slope = 1), alpha = .5, lty = 2) +
+  scale_fill_gradient("Fraction near-degenerate", low = "yellow", high = "red") +
+  facet_grid(Visibles~Hiddens) +
+  xlab(expression(paste(group("||", theta[main], "||"), "/(H + V)"))) +
+  ylab(expression(paste(group("||", theta[interaction], "||"), "/(H*V)"))) +
+  theme(aspect.ratio = 1) +
+  theme(legend.position = "bottom") +
+  ggtitle("Near-degeneracy") -> p.degen
+
+three_ways %>%
+  mutate(r1 = round(r1/(H + V), 8), r2 = round(r2/(H * V), 8)) %>% 
+  ggplot() +
+  geom_tile(aes(x = r1, y = r2, fill = max_abs_diff)) +
+  geom_contour(aes(x = r1, y = r2, z = max_abs_diff), colour = "black", bins = 8) +
+  #geom_contour(aes(x = r1, y = r2, z = mean_abs_diff), colour = "black", breaks = .05, size = 1.5) +
+  geom_abline(aes(intercept = 0, slope = 1), alpha = .5, lty = 2) +
+  scale_fill_gradient(expression(group("|", E(group("", bold(X), "|")*bold(theta)) - E(group("", bold(X), "|")*symbol("\306")), "|")), low = "yellow", high = "red", limits = c(0,2)) +
+  facet_grid(Visibles~Hiddens) +
+  xlab(expression(paste(group("||", theta[main], "||"), "/(H + V)"))) +
+  ylab(expression(paste(group("||", theta[interaction], "||"), "/(H*V)"))) +
+  theme(legend.position = "bottom") +
+  ggtitle("Uninterpretability") +
+  theme(aspect.ratio = 1) -> p.exp_diff
+
+three_ways %>%
+  mutate(r1 = round(r1/(H + V), 8), r2 = round(r2/(H * V), 8)) %>% 
+  ggplot() +
+  geom_tile(aes(x = r1, y = r2, fill = LHS2)) +
+  geom_contour(aes(x = r1, y = r2, z = LHS2), colour = "black", bins = 8) +
+  #geom_contour(aes(x = r1, y = r2, z = mean_max_q), colour = "black", breaks = .05, size = 1.5) +
+  geom_abline(aes(intercept = 0, slope = 1), alpha = .5, lty = 2) +
+  scale_fill_gradient(expression(frac(R(theta), V)), low = "yellow", high = "red") +
+  facet_grid(Visibles~Hiddens) +
+  xlab(expression(paste(group("||", theta[main], "||"), "/(H + V)"))) +
+  ylab(expression(paste(group("||", theta[interaction], "||"), "/(H*V)"))) +
+  theme(aspect.ratio = 1) +
+  theme(legend.position = "bottom") +
+  ggtitle("Instability") -> p.min_max_q
+```
+
+```{r, fig.hold=TRUE, fig.width=3, out.width='33%'}
+p.degen
+p.min_max_q
+p.exp_diff
+```
+RBMs easily are *near-degenerate*, *unstable*, and *uninterpretable* for large portions of parameter space.
+
 # Bayesian model fitting
+
+## A tale of three priors
 

### 2016-07-28 presentations/jsm2016/index.Rmd
slides complete, but check densities

@@
 
-## A tale of three priors
+## A tale of three methods
+
+> 1. *Trick prior.* Cancel out the normalizing term, full conditionals normally dist'd. 
+   $$
+    \pi(\theta) \propto \gamma(\theta)^n \exp\left(-\frac{1}{2C_{1}} \theta_{main}' \theta_{main} -\frac{1}{2C_{2}} \theta_{int}' \theta_{int}\right),
+   $$
+    where $\gamma(\theta)$ is the normalizing term [@li2014biclustering].
+> 2.
+*Truncated Normal prior.* Use two independent truncated spherical normal distributions as priors for $\theta_{main}$ and $\theta_{int}$ with $\sigma_{int} < \sigma_{main}$.
+Not conjugate $\Rightarrow$ geometric adaptive MH step [@zhou2014some] and calculation of $\gamma(\theta)$.
+> 3. *Marginalized likelihood* Marginalize out $h$ in $f_{\theta}(x)$, use trunc. Normal prior. 
+    $$
+    g_{ \theta}(v) = \sum\limits_{\in \{-1,1\}^H} \exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)
+    $$
+
+## Posterior distributions of images
+
+```{r models, cache=TRUE}
+source("../../4x4 model/functs_sample.R")
+
+#data and params ------------------------
+H <- 4
+V <- 4
+
+#marginalized likelihood
+load("../../4x4 model/written/fitted_models_trunc_marginal_thin.Rdata")
+marginal_bad <- models_bad
+marginal_good <- models_good
+
+#load trick prior
+load("../../4x4 model/written/fitted_models_jing_5.8.Rdata")
+trick_bad <- models_bad
+trick_good <- models_good
+
+#truncated normal
+load("../../4x4 model/written/fitted_models_trunc_thin.Rdata")
+trunc_bad <- models_bad
+trunc_good <- models_good
+
+#rm unneccesary data
+rm(models_bad)
+rm(models_degen)
+rm(models_good)
+
+load("../../4x4 model/written/params_theta.Rdata")
+
+params_good <- list(main_hidden = sample.params %>% ungroup() %>% filter(!near_hull) %>% select(starts_with("h"), -H) %>% data.matrix(),
+                    main_visible = sample.params %>% ungroup() %>% filter(!near_hull) %>% select(starts_with("v"), -V) %>% data.matrix(),
+                    interaction = sample.params %>% ungroup() %>% filter(!near_hull) %>% select(starts_with("theta")) %>% data.matrix() %>% matrix(4))
+
+#computing actual distributions ---------------------
+distn_good <- visible_distn(params = params_good)
+
+reshape_sample_distn <- function(model) {
+  sample_distn <- model$distn
+  dim(sample_distn) <- c(dim(sample_distn)[1]*dim(sample_distn)[2], dim(sample_distn)[3])
+  sample_distn %>% data.frame() -> sample_distn
+  names(sample_distn) <- names(distn_good)
+
+  sample_distn %>%
+    group_by(image_id) %>%
+    mutate(iter = 1:n()) -> sample_distn
+
+  return(sample_distn)
+}
+
+marginal_sample_good <- reshape_sample_distn(marginal_good)
+trick_sample_good <- reshape_sample_distn(trick_good)
+trunc_sample_good <- reshape_sample_distn(trunc_good)
+
+marginal_sample_good %>% rename(marginal = prob) %>%
+  left_join(trick_sample_good %>% ungroup() %>% select(-image_id) %>% rename(trick = prob)) %>%
+  left_join(trunc_sample_good %>% ungroup() %>% select(-image_id) %>% rename(trunc = prob)) %>%
+  ungroup() %>%
+  mutate(image_id = paste0("image_", image_id)) %>% 
+  gather(method, prob, trick, marginal, trunc) %>%
+  spread(image_id, prob) -> all_statistics
+
+all_statistics %>%
+  gather(statistic, value, -iter, -method, -starts_with("v")) %>%
+  filter(grepl("image", statistic) & !is.na(value)) %>%
+  separate(statistic, into = c("junk", "statistic")) %>%
+  mutate(statistic = factor(statistic, labels = paste("Image", 1:length(unique(statistic))))) %>%
+  left_join(distn_good %>% select(-image_id)) %>% 
+  ggplot() +
+  geom_density(aes(value, colour = method, fill = method), alpha = .2) +
+  geom_vline(aes(xintercept = prob)) +
+  facet_wrap(~statistic, scales = "free_x") + 
+  ylab("Posterior Density") + xlab("Probability of Image") +
+  scale_colour_discrete("Method", labels = c("Marginalized likelihood", "Trick prior", "Truncated Normal prior")) +
+  scale_fill_discrete("Method", labels = c("Marginalized likelihood", "Trick prior", "Truncated Normal prior")) -> p.models
+```
+```{r models-plot, fig.width=8, fig.height=4}
+p.models
+```
+Each of these rigorous fitting methods seeks to merely replicate the empirical data distribution, which is the optimal nonparametric solution.
 

### 2016-07-28 4x4 model/test_fitted_model.R

@@ -9,9 +9,9 @@ H <- 4
 V <- 4
 
 load("written/sample_images.Rdata")
-load("written/fitted_models_trunc_marginal_thin.Rdata")
+# load("written/fitted_models_trunc_marginal_thin.Rdata")
 # load("written/fitted_models_trunc_thin.Rdata")
-# load("written/fitted_models_jing_4.Rdata")
+load("written/fitted_models_jing_5.8.Rdata")
 load("written/params_theta.Rdata")
 params_degen <- list(main_hidden = sample.params %>% ungroup() %>% filter(near_hull) %>% select(starts_with("h"), -H) %>% data.matrix(),
                      main_visible = sample.params %>% ungroup() %>% filter(near_hull) %>% select(starts_with("v"), -V) %>% data.matrix(),
@@ -135,8 +135,8 @@ sample_good %>% rename(good = prob) %>%
   geom_line(aes(iter, prob, colour = model)) +
   geom_abline(aes(intercept = true_prob, slope = 0)) +
   #geom_abline(aes(intercept = prop, slope = 0), data = data_props_good, colour = "red") +
-  facet_wrap(~image_id) +
-  ylim(c(0,1))
+  facet_wrap(~image_id, scales="free_y")
+  #ylim(c(0,1))
 
 # sample_degen %>% 
 #   left_join(distn_degen %>% rename(true_prob = prob) %>% select(-image_id)) %>%

### 2016-07-28 presentations/jsm2016/index.Rmd
scaled densities for slides

@@
   ggplot() +
-  geom_density(aes(value, colour = method, fill = method), alpha = .2) +
+  geom_density(aes(value, y=..scaled.., colour = method, fill = method), alpha = .2) +
   geom_vline(aes(xintercept = prob)) +
   facet_wrap(~statistic, scales = "free_x") + 
-  ylab("Posterior Density") + xlab("Probability of Image") +
+  ylab("Scaled Posterior Density") + xlab("Probability of Image") +
   scale_colour_discrete("Method", labels = c("Marginalized likelihood", "Trick prior", "Truncated Normal prior")) +

### 2016-07-28 4x4 model/test_fitted_model.R

@@ -135,7 +135,7 @@ sample_good %>% rename(good = prob) %>%
   geom_line(aes(iter, prob, colour = model)) +
   geom_abline(aes(intercept = true_prob, slope = 0)) +
   #geom_abline(aes(intercept = prop, slope = 0), data = data_props_good, colour = "red") +
-  facet_wrap(~image_id, scales="free_y")
+  facet_wrap(~image_id, scales="free_y") 
   #ylim(c(0,1))
 
 # sample_degen %>% 

### 2016-07-28 presentations/jsm2016/index.Rmd
added motivation and conclusion

--- 2016-07-28 presentations/jsm2016/index.Rmd
@@
 <p class="double-flow">
-Restricted Boltzmann machine (RBM) with two layers - hidden ($\mathcal{H}$) and visible ($\mathcal{V}$) [@friedman2001elements].
+Restricted Boltzmann machine (RBM) with two layers - hidden ($\mathcal{H}$) and visible ($\mathcal{V}$) [smolensky1986information].
 </p>
@@
 Claimed ability to learn "internal representations that become increasingly complex" [@salakhutdinov2009deep], used in classification problems.
+
+## Why do I care?
+
+Current heuristic fitting methods seem to work for classification. Beyond classification, RBMs are generative models:
+
+> To generate data from an RBM, we can start with a random state in one of the layers and then perform alternating Gibbs sampling.
+[@hinton2006fast]
+
+Can we fit a model that generates data that looks like data?
 
@@
 |
-
-## Data coding to mitigate degeneracy
-
-<div class="double">
-<p class="double-flow">
-<img src="images/toyhull_top.png" width="100%">
-</p>
-<p class="double-flow">
-<img src="images/toyhull_top_2.png" width="70%">
-</p>
-</div>
-
-Convex hulls of the statistic space for a toy RBM with $V = H = 1$ for $\{0,1\}$ and $\{-1,1\}$-encoding enclosed by an unrestricted hull of 3-space.
-
-## Data coding to mitigate degeneracy (cont'd)
-
-- For the $\{-1, 1 \}$ encoding of $\mathcal{V}$ and $\mathcal{H}$, the origin is the center of the parameter space
-- At $\theta = 0$, the RBM is equivalent to elements of $X$ being distributed as iid Bernoulli$\left(\frac{1}{2}\right)$ $\Rightarrow$ No *near-degeneracy*, *instability*, or *uninterpretability*!
-
-```{r encoding-volume, echo = FALSE, fig.height=3}
-test_cases <- expand.grid(data.frame(1:4)[,rep(1,2)]) %>% 
-  rename(H = X1.4, V = X1.4.1) %>%
-  filter(H <= V) %>% #remove those cases with less visibles than hiddens. Is this necessary?
-  mutate(n_param = H*V + H + V) %>%
-  mutate(max_facets = (2^(H+V))^(floor(n_param/2))) %>%
-  filter(n_param <= 11) #calc_hull can't handle any higher dimensions currently
-
-bin_res <- plyr::dlply(test_cases, plyr::.(n_param), function(x) calc_hull(x$V, x$H, "binary"))
-neg_res <- plyr::dlply(test_cases, plyr::.(n_param), function(x) calc_hull(x$V, x$H, "negative"))
-
-plyr::ldply(bin_res, function(x) x$c_hull$vol) %>% 
-  mutate(frac_vol = V1/(1^n_param)) %>%
-  inner_join(plyr::ldply(neg_res, function(x) x$c_hull$vol) %>% 
-               mutate(frac_vol = V1/(2^n_param)),
-             by="n_param") %>%
-  rename(vol.bin = V1.x, vol.neg = V1.y, frac_vol.bin = frac_vol.x, frac_vol.neg = frac_vol.y) %>%
-  gather(vars, value, -n_param) %>%
-  separate(vars, c("type", "encoding"), "\\.") %>%
-  spread(type, value) %>%
-  ggplot() +
-  geom_point(aes(x=n_param, y=frac_vol, colour=encoding)) +
-  geom_line(aes(x=n_param, y=frac_vol, colour=encoding, group=encoding)) +
-  ylab("Fraction of unrestricted volume") +
-  xlab("Number of parameters") +
-  scale_colour_discrete("Encoding", labels=c("Binary (1,0)", "Negative (1,-1)")) +           
-  theme(
-    plot.background = element_rect(fill = "transparent", colour = NA),
-    legend.background = element_rect(fill = "transparent", colour = NA)) -> p
-
-p
-```
 
@@
 
+## Data coding to mitigate degeneracy
+
+<div class="double">
+<p class="double-flow">
+<img src="images/toyhull_top.png" width="100%">
+</p>
+<p class="double-flow">
+<img src="images/toyhull_top_2.png" width="70%">
+</p>
+</div>
+
+Convex hulls of the statistic space for a toy RBM with $V = H = 1$ for $\{0,1\}$ and $\{-1,1\}$-encoding enclosed by an unrestricted hull of 3-space.
+
+## Data coding to mitigate degeneracy (cont'd)
+
+- For the $\{-1, 1 \}$ encoding of $\mathcal{V}$ and $\mathcal{H}$, the origin is the center of the parameter space
+- At $\theta = 0$, the RBM is equivalent to elements of $X$ being distributed as iid Bernoulli$\left(\frac{1}{2}\right)$ $\Rightarrow$ No *near-degeneracy*, *instability*, or *uninterpretability*!
+
+```{r encoding-volume, echo = FALSE, fig.height=3}
+test_cases <- expand.grid(data.frame(1:4)[,rep(1,2)]) %>% 
+  rename(H = X1.4, V = X1.4.1) %>%
+  filter(H <= V) %>% #remove those cases with less visibles than hiddens. Is this necessary?
+  mutate(n_param = H*V + H + V) %>%
+  mutate(max_facets = (2^(H+V))^(floor(n_param/2))) %>%
+  filter(n_param <= 11) #calc_hull can't handle any higher dimensions currently
+
+bin_res <- plyr::dlply(test_cases, plyr::.(n_param), function(x) calc_hull(x$V, x$H, "binary"))
+neg_res <- plyr::dlply(test_cases, plyr::.(n_param), function(x) calc_hull(x$V, x$H, "negative"))
+
+plyr::ldply(bin_res, function(x) x$c_hull$vol) %>% 
+  mutate(frac_vol = V1/(1^n_param)) %>%
+  inner_join(plyr::ldply(neg_res, function(x) x$c_hull$vol) %>% 
+               mutate(frac_vol = V1/(2^n_param)),
+             by="n_param") %>%
+  rename(vol.bin = V1.x, vol.neg = V1.y, frac_vol.bin = frac_vol.x, frac_vol.neg = frac_vol.y) %>%
+  gather(vars, value, -n_param) %>%
+  separate(vars, c("type", "encoding"), "\\.") %>%
+  spread(type, value) %>%
+  ggplot() +
+  geom_point(aes(x=n_param, y=frac_vol, colour=encoding)) +
+  geom_line(aes(x=n_param, y=frac_vol, colour=encoding, group=encoding)) +
+  ylab("Fraction of unrestricted volume") +
+  xlab("Number of parameters") +
+  scale_colour_discrete("Encoding", labels=c("Binary (1,0)", "Negative (1,-1)")) +           
+  theme(
+    plot.background = element_rect(fill = "transparent", colour = NA),
+    legend.background = element_rect(fill = "transparent", colour = NA)) -> p
+
+p
+```
+
 # Bayesian model fitting
@@
     $$
-    g_{ \theta}(v) = \sum\limits_{\in \{-1,1\}^H} \exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)
+    g_{ \theta}(v) = \sum\limits_{h \in \{-1,1\}^H} \exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)
     $$
@@
 ```{r models, cache=TRUE}
+
+##TODO: Add empirical line
+
 source("../../4x4 model/functs_sample.R")
@@
 ```
-Each of these rigorous fitting methods seeks to merely replicate the empirical data distribution, which is the optimal nonparametric solution.
+Each of these rigorous fitting methods seeks to merely replicate the empirical data distribution, which is the likelihood maximizer.
+
+## Wrapping up
+
+- While RMBs can be useful for classification, in the context of a statistical model as a representation of data, RBMs poor fit due to
+    1. *near-degeneracy*, 
+    2. *instability*, and
+    3. *uninterpretability*.
+- Rigorous fitting methodology is possible but slow, and ultimately results in the empirical data distribution.
 

### 2016-07-29 presentations/jsm2016/index.Rmd
typos

@@
 <p class="double-flow">
-Restricted Boltzmann machine (RBM) with two layers - hidden ($\mathcal{H}$) and visible ($\mathcal{V}$) [smolensky1986information].
+Restricted Boltzmann machine (RBM) with two layers - hidden ($\mathcal{H}$) and visible ($\mathcal{V}$) [@smolensky1986information].
 </p>
@@
 $$
-f_{\theta} (x) = \frac{\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\sum\limits_{x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}
+f_{\theta} (x) = \frac{\exp\left(Q(x)\right)}{\sum\limits_{x \in \mathcal{X}}\exp\left(Q(x)\right)}
 $$
+
+Where $Q(x) = \sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j$ denotes the neg-potentional function of the model, having support set $\mathcal{S}$.
 
@@
 The highly flexible nature of the RBM ($H + V + HV$ parameters) makes three characteristics of model impropriety of particular concern.
-
-Let $Q(\cdot)$ denotes the neg-potentional function of the model, having support set $\mathcal{S}$.
 
@@
 Not conjugate $\Rightarrow$ geometric adaptive MH step [@zhou2014some] and calculation of $\gamma(\theta)$.
-> 3. *Marginalized likelihood* Marginalize out $h$ in $f_{\theta}(x)$, use trunc. Normal prior. 
+> 3. *Marginalized likelihood.* Marginalize out $h$ in $f_{\theta}(x)$, use trunc. Normal prior. 
     $$
@@
 ```{r models, cache=TRUE}
-
-##TODO: Add empirical line
-
 source("../../4x4 model/functs_sample.R")
+load("../../4x4 model/written/sample_images.Rdata")
 
@@
 rm(models_bad)
-rm(models_degen)
 rm(models_good)
@@
 #computing actual distributions ---------------------
+flat_images_good$visibles %>%
+  data.frame() %>% 
+  rename(v1 = X1, v2 = X2, v3 = X3, v4 = X4) %>%
+  group_by(v1, v2, v3, v4) %>%
+  summarise(prob = n()/nrow(flat_images_good$visibles)) -> distn_emp
+
 distn_good <- visible_distn(params = params_good)
@@
   mutate(statistic = factor(statistic, labels = paste("Image", 1:length(unique(statistic))))) %>%
-  left_join(distn_good %>% select(-image_id)) %>% 
+  left_join(distn_emp %>% rename(prob_emp = prob) %>% right_join(distn_good %>% select(-image_id))) %>%
   ggplot() +
@@
   geom_vline(aes(xintercept = prob)) +
+  geom_vline(aes(xintercept = prob_emp), colour = "red") +
   facet_wrap(~statistic, scales = "free_x") + 
@@
 ```
-```{r models-plot, fig.width=8, fig.height=4}
+```{r models-plot, fig.width=8, fig.height=5.5}
 p.models
 ```
-Each of these rigorous fitting methods seeks to merely replicate the empirical data distribution, which is the likelihood maximizer.
 
@@
 
-- While RMBs can be useful for classification, in the context of a statistical model as a representation of data, RBMs poor fit due to
+- While RBMs can be useful for classification, in the context of a statistical model as a representation of data, RBMs poor fit due to
     1. *near-degeneracy*, 

### 2016-07-31 presentations/jsm2016/index.Rmd
sizing

@@
 author: http://bit.ly/jsm2016-rbm
-ratio: 16x10
+ratio: 4x3
 output:

### 2016-08-11 writing/draft.Rmd
slides complete, but check densities
update document to JSM content

@@
 author: "Andee Kaplan, Stephen Vardeman, Daniel Nordman"
-date: "June, 2016"
+date: "August, 2016"
 output: 
@@
 
+Thus, if a discrete exponential family model is unstable, then it is also degenerate.
+
 ###Uninterpretability
@@
 
-#Characterizations
+#Simulations
+
+We explore the relationship between magnitudes of $\boldsymbol \theta$ and the three measures of model impropriety, near-degeneracy, instability, and uninterpretability, for varying sized RBM models through simulated data.
+
 ##Tiny example
@@
 #Bayesian model fitting
-To avoid model impropriety, we want to avoid parts of the parameter space that lead to near-degeneracy}, instability, and uninterpretability.
-Based on the evidence in section \ref{exploring-manageable-examples}, one idea is to shrink $\boldsymbol \theta$ by specifying priors that place low probability on high values of $\boldsymbol \theta$.
+To avoid model impropriety, we want to avoid parts of the parameter space that lead to near-degeneracy, instability, and uninterpretability.
+Based on the evidence in section \ref{exploring-manageable-examples}, one idea is to shrink $\boldsymbol \theta$ by specifying priors that place low probability on high values of $\boldsymbol \theta$, specifically focusing on shrinking $\boldsymbol \theta_{interaction}$ more than $\boldsymbol \theta_{main}$.
 
@@
 
-1. *Trick prior.* Cancel out the normalizing term, resulting full conditionals are normally distributed. 
+1. *Trick prior.* Cancel out normalizing term, resulting full conditionals normally distributed. 
     \begin{align*}
@@
 The three fitting methods are ordered by computational feasibility in a real data situation with the trick prior being the most possible due to conjugacy and the marginalized likelihood method the least feasible due to the marginalization and need for an adaptive Metropolis Hastings step.
-The trick prior, however, does require extensive tuning to choose $C_1$ and $C_2$, making it less appealing.
+All three priors do require choosing the values of hyperparameters.
+In each case, we have chosed the values based on a rule of thumb that maintains the property of shrinking $\boldsymbol \theta_{interaction}$ more than $\boldsymbol \theta_{main}$.
+Additionally, the trick prior requires additional tuning to choose $C_1$ and $C_2$, making it less appealing.
+The values used for the hyperparameters are presented in table \ref{tab:hyperparam}.
+
+\begin{table}[ht]
+\centering
+\begin{tabular}{|l|c|c|}
+\hline 
+Method & Hyperparameter & Value \\ 
+\hline \hline
+\multirow{2}{*}{Trick Prior} & $C_1$ & $\frac{C}{N}\frac{1}{H + V}$ \\
+ & $C_2$ & $\frac{C}{N}\frac{1}{H*V}$ \\
+\hline
+\multirow{2}{*}{Truncated Normal} & $\sigma^2_{main}$ & $\frac{1}{H + V}$ \\
+ & $\sigma^2_{interaction}$ & $\frac{1}{H*V}$ \\
+\hline
+\multirow{2}{*}{Marginalized Likelihood} & $\sigma^2_{main}$ & $\frac{1}{H + V}$ \\
+ & $\sigma^2_{interaction}$ & $\frac{1}{H*V}$ \\
+\hline
+\end{tabular}
+\caption{The values used for the hyperparameters for all three fitting methods.
+A rule of thumb is imposed which decreases as the size of the model increases and also shrinks $\boldsymbol \theta_{interaction}$ more than $\boldsymbol \theta_{main}$.}
+\label{tab:hyperparam}
+\end{table}
 
 Figure \ref{fig:fitting_plot} shows the posterior probability of each possible binary image with four pixels after fitting the model in the three ways detailed in this section.
-The black lines show the true probabilities of each image based on the parameters used to simulate the images.
+The black vertical lines show the true probabilities of each image based on the parameters used to simulate the images while the red vertical lines show the empirical distribution.
 From these posterior predictive checks, it is evident that the marginalized likelihood method shows the closest fit to the actual data.
@@
 ```{r fitting}
+source("helpers/functions.R")
+load("data/sample_images.Rdata")
+
 #data and params ------------------------
@@
 #marginalized likelihood
-load("data/fitted_models_adaptive_mh_trunc_distn_marginal_1.Rdata")
-marginal_bad_1 <- models_bad
-marginal_good_1 <- models_good
+load("data/fitted_models_trunc_marginal_thin.Rdata")
+marginal_bad <- models_bad
+marginal_good <- models_good
 
-#load Jing's prior
-load("data/fitted_models_distn_0.26.Rdata")
-jing_bad <- models_bad
-jing_good <- models_good
+#load trick prior
+load("data/fitted_models_jing_5.8.Rdata")
+trick_bad <- models_bad
+trick_good <- models_good
 
 #truncated normal
-load("data/fitted_models_adaptive_mh_trunc_distn_shrunk_jing_match.Rdata")
-shrunk_jing_bad <- models_bad
-shrunk_jing_good <- models_good
+load("data/fitted_models_trunc_thin.Rdata")
+trunc_bad <- models_bad
+trunc_good <- models_good
 
@@
 rm(models_bad)
-rm(models_degen)
 rm(models_good)
 
-load("data/sample.params.Rdata")
-params_degen <- list(main_hidden = sample.params %>% ungroup() %>% filter(near_hull) %>% select(starts_with("h"), -H) %>% data.matrix(),
-                     main_visible = sample.params %>% ungroup() %>% filter(near_hull) %>% select(starts_with("v"), -V) %>% data.matrix(),
-                     interaction = sample.params %>% ungroup() %>% filter(near_hull) %>% select(starts_with("theta")) %>% data.matrix() %>% matrix(4))
+load("data/params_theta.Rdata")
+
 params_good <- list(main_hidden = sample.params %>% ungroup() %>% filter(!near_hull) %>% select(starts_with("h"), -H) %>% data.matrix(),
@@
 
-load("data/sample_images.Rdata")
+#computing actual distributions ---------------------
+flat_images_good$visibles %>%
+  data.frame() %>% 
+  rename(v1 = X1, v2 = X2, v3 = X3, v4 = X4) %>%
+  group_by(v1, v2, v3, v4) %>%
+  summarise(prob = n()/nrow(flat_images_good$visibles)) -> distn_emp
 
-#computing actual distributions ---------------------
 distn_good <- visible_distn(params = params_good)
-distn_degen <- visible_distn(params = params_degen)
 
@@
   sample_distn %>% data.frame() -> sample_distn
-  names(sample_distn) <- names(distn_degen)
+  names(sample_distn) <- names(distn_good)
 
@@
 
-marginal_sample_good_1 <- reshape_sample_distn(marginal_good_1)
-jing_sample_good <- reshape_sample_distn(jing_good)
-shrunk_jing_sample_good <- reshape_sample_distn(shrunk_jing_good) %>% filter(iter > 500) %>% mutate(iter = 1:n())
+marginal_sample_good <- reshape_sample_distn(marginal_good)
+trick_sample_good <- reshape_sample_distn(trick_good)
+trunc_sample_good <- reshape_sample_distn(trunc_good)
 
-#plots -------------------
-marginal_sample_good_1 %>% rename(marginal = prob) %>%
-  left_join(jing_sample_good %>% ungroup() %>% select(-image_id) %>% rename(trick = prob)) %>%
-  left_join(shrunk_jing_sample_good %>% ungroup() %>% select(-image_id) %>% rename(trunc = prob)) %>%
+marginal_sample_good %>% rename(marginal = prob) %>%
+  left_join(trick_sample_good %>% ungroup() %>% select(-image_id) %>% rename(trick = prob)) %>%
+  left_join(trunc_sample_good %>% ungroup() %>% select(-image_id) %>% rename(trunc = prob)) %>%
   ungroup() %>%
-  mutate(method = "good", image_id = paste0("image_", image_id)) %>% 
-  #select_(.dots = paste0("-v", 1:V)) %>%
-  gather(prior, prob, trick, marginal, trunc) %>%
+  mutate(image_id = paste0("image_", image_id)) %>% 
+  gather(method, prob, trick, marginal, trunc) %>%
   spread(image_id, prob) -> all_statistics
@@
 all_statistics %>%
-  gather(statistic, value, -iter, -method, -prior, -starts_with("v")) %>%
+  gather(statistic, value, -iter, -method, -starts_with("v")) %>%
   filter(grepl("image", statistic) & !is.na(value)) %>%
@@
   mutate(statistic = factor(statistic, labels = paste("Image", 1:length(unique(statistic))))) %>%
-  left_join(distn_good %>% select(-image_id)) %>% 
+  left_join(distn_emp %>% rename(prob_emp = prob) %>% right_join(distn_good %>% select(-image_id))) %>%
   ggplot() +
-  geom_line(aes(iter/100, value, colour = prior)) +
-  geom_abline(aes(slope = 0, intercept = prob)) +
-  facet_wrap(~statistic) +
-  ylim(c(0,.5)) +
-  xlab("MCMC Iteration (in 100s)") + ylab("Posterior Probability") +
-  theme(legend.position = "bottom") +
-  scale_colour_discrete("Method", labels = c("Marginalized likelihood", "Trick prior", "Truncated Normal prior")) -> p
-
+  geom_density(aes(value, y=..scaled.., colour = method, fill = method), alpha = .2) +
+  geom_vline(aes(xintercept = prob)) +
+  geom_vline(aes(xintercept = prob_emp), colour = "red") +
+  facet_wrap(~statistic, scales = "free_x") + 
+  ylab("Scaled Posterior Density") + xlab("Probability of Image") +
+  scale_colour_discrete("Method", labels = c("Marginalized likelihood", "Trick prior", "Truncated Normal prior")) +
+  scale_fill_discrete("Method", labels = c("Marginalized likelihood", "Trick prior", "Truncated Normal prior")) +
+  theme(legend.position = "bottom") -> p.models
 ```
@@
 ```{r fitting_plot, fig.cap="\\label{fig:fitting_plot}Posterior probability of each possible 4-pixel image using three Bayesian fitting techniques, trick prior, truncated Normal prior, and marginalized likelihood.
-The black lines show the true probabilities of each image based on the parameters used to simulate the images.
+The black vertical lines show the true probabilities of each image based on the parameters used to simulate the images while the red vertical lines show the empirical distribution.
 The marginalized likelihood method has the best fit for the data, however is also the most computationally intensive and least feasible with a real dataset.", fig.height=6}
-p
+p.models
 ```
@@
 **TODO: Add proof that likelihood goes to infinity.**
-**TODO: Add comparison to contrastive divergence**
 
 #Discussion
+
+**TODO: Add comparison to contrastive divergence**
 

### 2016-08-18 writing/draft.Rmd
added draft intro

@@
 title: "An exposition on the propriety of restricted Boltzmann machines"
-author: "Andee Kaplan, Stephen Vardeman, Daniel Nordman"
+author: "Andee Kaplan, Daniel Nordman, Stephen Vardeman"
 date: "August, 2016"
@@
 bibliography: references/refs.bib
+abstract: |
+  A restricted Boltzmann machine (RBM) is an undirected graphical model constructed for discrete or continuous random variables, specified as having two layers, a hidden and a visible layer with no conditional dependency within a layer.
+In recent years, RBMs have risen to prominence due to their connection to deep learning - by treating a hidden layer of one RBM as the visible layer in a second RBM a deep architecture can be created.
+The method is claimed to have the ability to learn very complex and rich structures in data, making these models attractive for supervised learning.
+However, the generative behavior of RBMs has largely been unexplored.
+In this paper, we discuss the relationship between parameter specification and the prevalence of degenerate models as well as an exposition of the difficulties in fitting such highly flexible models in the binary case of RBMs.
 ---
@@
 
+The data mining and machine learning communities have recently found a great interest in deep learning, specifically in stacked restricted Boltzmann machines (RBM).
+(TODO: deep learning, deep RBM references.) A RBM is an undirected graphical model constructed for discrete or continuous random variables, specified as having two layers, a hidden and a visible layer with no conditional dependency within a layer.
+These models are typically used with success in classification of images.
+(TODO: classification references.) However, the model properties are largely unexplored in the literature and the fitting methodology remains heuristic based and abstruse.
+In this paper, we attept to give a thorough understanding of the model and its behavior from the context of statistical model theory as well as explore the possibility of a rigorous fitting methodology.
+(TODO: statistical model definition) Through our study, we find the RBM model to be deficient in two fundamental ways.
+
+First, the models are unsatisfactory as a conceptualization of how data is generated.
+Specifically, these models often fail to generate data that resembles realistic data.
+Additionally, it is evidenced that RBMs easily exhibit instability in the parameter space.
+In practice, this is seen when a single pixel change in an image results in a wildly different classification (TODO: add ref).
+We investigate these phenomena in section \ref{explorations} through simulations of small, manageable examples.
+
+Secondly, the fitting of these models is problematic.
+As the size of these models grows, a computationally intractible situation is created quickly when fitting with either maximum likelihood or through Bayesian methods.
+However, even in models small enough to feasibly fit, there are unique challenges due to the overparameterization of RBMs.
+In section \ref{bayesian-model-fitting} we try three Bayesian modelling techniques, each with the goal of avoiding parts of the parameter space that yield improper models.
+As could be expected, with computational complexity comes an increase in accuracy, but at the cost of feasibility.
+Additionally, because these models are overparametrized, any honest fitting method will reproduce the empirical distribution.
+Thus, there is no smoothing achieved with a RBM that is fit using a rigourous method.
+We are therefore very skeptical that any model built using these structures, like a deep Boltzmann machine, can achieve prediction or inference in a principled way.
+
+This paper is structured as follows, section \ref{background} gives a definition of the RBM including the joint distribution and shows the connection to deep learning.
+Additionally, measures of model impropriety and methods of detecting them are defined.
+Section \ref{explorations} details our explorations into the model behavior and propriety (or lack thereof) of the RBM.
+We discuss three Bayesian fitting techniques with the goal of avoiding model impropriety in section \ref{bayesian-model-fitting} and wrap up with a discussion in section \ref{discussion}.
+Additionally, there are two accompanying appendices with auxillary proofs relating to the choice of data encoding and parameter behavior during fitting with incomplete training data in section \ref{appendix}.
+
 #Background
@@
 
-A restricted Boltzmann machine (RBM) is a specific undirected graphical model specified for discrete random variables, binary being the most common.
+A restricted Boltzmann machine (RBM) is a specific undirected graphical model specified for discrete or continuous random variables, binary being the most common.
+In this paper, we consider the binary case of an RBM only.
 The architecture of an RBM is specified as having two layers, a hidden ($\mathcal{H}$) and a visible layer ($\mathcal{V}$), with no connections within a layer.
@@
 
-#Simulations
+#Explorations
 

### 2016-08-22 writing/draft.Rmd
update intro and refs

@@
 ```
-#Introduction
+
+# Introduction
 
 The data mining and machine learning communities have recently found a great interest in deep learning, specifically in stacked restricted Boltzmann machines (RBM).
-(TODO: deep learning, deep RBM references.) A RBM is an undirected graphical model constructed for discrete or continuous random variables, specified as having two layers, a hidden and a visible layer with no conditional dependency within a layer.
+(**TODO: deep learning, deep RBM references.**) A RBM is an undirected graphical model constructed for discrete or continuous random variables, specified as having two layers, a hidden and a visible layer with no conditional dependency within a layer.
 These models are typically used with success in classification of images.
-(TODO: classification references.) However, the model properties are largely unexplored in the literature and the fitting methodology remains heuristic based and abstruse.
-In this paper, we attept to give a thorough understanding of the model and its behavior from the context of statistical model theory as well as explore the possibility of a rigorous fitting methodology.
-(TODO: statistical model definition) Through our study, we find the RBM model to be deficient in two fundamental ways.
+(**TODO: classification references.**) However, the model properties are largely unexplored in the literature and the fitting methodology remains heuristic based and abstruse.
+In this paper, we attempt to give a thorough understanding of the model and its behavior from the context of statistical model theory as well as explore the possibility of a rigorous fitting methodology.
+Through our study, we find the RBM model to be deficient in two fundamental ways.
 
 First, the models are unsatisfactory as a conceptualization of how data is generated.
-Specifically, these models often fail to generate data that resembles realistic data.
+According to Fisher, the aim of a statistical model is to represent data in a compact way [@fisher1922mathematical].
+Neyman and Box agree with this assessment and go further to state that a model should "provide an explanation of the mechanism underlying the observed phenomena" [@lehmann1990model, @box1967discrimination].
+In the case of RBMs, these models often fail to generate data that resembles realistic data.
+Thus, providing an unsatisfactory conceptualization of the data generation process.
 Additionally, it is evidenced that RBMs easily exhibit instability in the parameter space.
-In practice, this is seen when a single pixel change in an image results in a wildly different classification (TODO: add ref).
-We investigate these phenomena in section \ref{explorations} through simulations of small, manageable examples.
+In practice, this is seen when a single pixel change in an image results in a wildly different classification.
+These model impropriety measures has been documented in other deep architectures recently as well [@szegedy2013intriguing, @nguyen2014deep].
+We investigate these phenomena in RBMs in section \ref{explorations} through simulations of small, manageable examples.
 
 Secondly, the fitting of these models is problematic.
-As the size of these models grows, a computationally intractible situation is created quickly when fitting with either maximum likelihood or through Bayesian methods.
+As the size of these models grows, a computationally intractable situation is created quickly when fitting with either maximum likelihood or through Bayesian methods.
 However, even in models small enough to feasibly fit, there are unique challenges due to the overparameterization of RBMs.
@@
 Additionally, because these models are overparametrized, any honest fitting method will reproduce the empirical distribution.
-Thus, there is no smoothing achieved with a RBM that is fit using a rigourous method.
+Thus, there is no smoothing achieved with a RBM that is fit using a rigorous method.
 We are therefore very skeptical that any model built using these structures, like a deep Boltzmann machine, can achieve prediction or inference in a principled way.
@@
 We discuss three Bayesian fitting techniques with the goal of avoiding model impropriety in section \ref{bayesian-model-fitting} and wrap up with a discussion in section \ref{discussion}.
-Additionally, there are two accompanying appendices with auxillary proofs relating to the choice of data encoding and parameter behavior during fitting with incomplete training data in section \ref{appendix}.
+Additionally, there are two accompanying appendices with auxiliary proofs relating to the choice of data encoding and parameter behavior during fitting with incomplete training data in section \ref{appendix}.
 
-#Background
+# Background
 
-##Restricted Boltzmann machines
+## Restricted Boltzmann machines
 
@@
 
-###Joint distribution
+### Joint distribution
 Let $\boldsymbol x = \{h_1, \dots, h_H, v_1,\dots,v_V\}$ represent the states of the visible and hidden nodes in an RBM.
@@
 
-Due to the potential size of the model the normalizing constant $\gamma(\boldsymbol \theta)$ can be infeasible to calculate, making it impossible to estimate the model exactly.
+Due to the potential size of the model the normalizing constant $\gamma(\boldsymbol \theta)$ can be unfeasible to calculate, making it impossible to estimate the model exactly.
 An approximation can be achieved using Gibbs sampling, which is a natural fit due to the restricted architecture of the RBM.
@@
 
-###Connection to Deep Learning
+### Connection to Deep Learning
 RBMs have risen to prominence in recent years due to their connection to deep learning [see @hinton2006fast, @salakhutdinov2012efficient, or @srivastava2013modeling for examples].
@@
 
-##Degeneracy, instability, and uninterpretability... Oh my!
+## Degeneracy, instability, and uninterpretability... Oh my!
 The highly flexible nature of the RBM ($H + V + H V$ parameters) makes three characteristics of model impropriety of particular concern, *degeneracy, instability, and uninterpretability*.
@@
 
-###Near-degeneracy
+### Near-degeneracy
 In Random Graph Model theory, *degeneracy* occurs when there is a disproportionate amount of probability placed on only a few elements of the sample space, $\mathcal{X}$, by the model.
@@
 
-###Instability
+### Instability
 @schweinberger2011instability introduces a related concept, *instability*, which occurs when small changes in the natural parameters result in large changes in probability masses.
@@
 
-###Uninterpretability
+### Uninterpretability
 @kaiser2007statistical defines a measure of model impropriety called *uninterpretability*, which is characterized as due to the existence of dependence, marginal mean-structure no longer being maintained in the model.
@@
 
-#Explorations
+# Explorations
 
@@
 
-##Tiny example
+## Tiny example
 To illustrate the ideas of graph near-degeneracy, instability, and uninterpretability in a RBM, we will consider a the smallest possible toy example that consists of one visible node $v_1$ and one hidden node $h_1$, both binary.
@@
 
-###Impropriety three ways
+### Impropriety three ways
 
@@
 
-###Data coding to mitigate degeneracy
+### Data coding to mitigate degeneracy
 Multiple encodings of the binary variables are possible.
@@
 This will result in $t(\boldsymbol x) \in \{0,1\}^{H + V + HV}$ or $t(\boldsymbol x) \in \{-1,1\}^{H + V + HV}$ depending on how we encode on and off states in the nodes.
-To explore the effect of this encoding on the potential near degeneracy of an RBM, we can explore the behavior of the ratio of the volume of the unrestricted convex hull to the restructed convex hull of $\{t(\boldsymbol x)\}$ to determine how much the restriction in our statistics by including the cross product terms incurs.
+To explore the effect of this encoding on the potential near degeneracy of an RBM, we can explore the behavior of the ratio of the volume of the unrestricted convex hull to the restricted convex hull of $\{t(\boldsymbol x)\}$ to determine how much the restriction in our statistics by including the cross product terms incurs.
 
@@
 In fact, if we look at the ratio of volume within a convex hull versus the volume of an unrestricted hull, we can see a relationship emerge as the number of nodes (and thus parameters) increases.
-In figure \ref{fig:volume_plot}, it is evident that as the number of parameters increses, this ratio is decreasing at an increasing rate, meaning it gets more and more difficult to avoid the boundary of the convex hull and thus near degeneracy.
+In figure \ref{fig:volume_plot}, it is evident that as the number of parameters increases, this ratio is decreasing at an increasing rate, meaning it gets more and more difficult to avoid the boundary of the convex hull and thus near degeneracy.
 Additionally, it appears that the $\{-1,1\}$ encoding suffers slightly less from this problem.
@@
 
-##Exploring manageable examples
+## Exploring manageable examples
 ```{r degen-data, message=FALSE, warning=FALSE, cache=TRUE}
@@
 
-#Bayesian model fitting
+# Bayesian model fitting
 To avoid model impropriety, we want to avoid parts of the parameter space that lead to near-degeneracy, instability, and uninterpretability.
@@
 All three priors do require choosing the values of hyperparameters.
-In each case, we have chosed the values based on a rule of thumb that maintains the property of shrinking $\boldsymbol \theta_{interaction}$ more than $\boldsymbol \theta_{main}$.
+In each case, we have chosen the values based on a rule of thumb that maintains the property of shrinking $\boldsymbol \theta_{interaction}$ more than $\boldsymbol \theta_{main}$.
 Additionally, the trick prior requires additional tuning to choose $C_1$ and $C_2$, making it less appealing.
@@
 From these posterior predictive checks, it is evident that the marginalized likelihood method shows the closest fit to the actual data.
-However, this model requires a marginalization step, which is infeasible for a model of any real size.
+However, this model requires a marginalization step, which is unfeasible for a model of any real size.
 
@@
 
-#Discussion
+# Discussion
 

### 2016-09-03 writing/draft.Rmd
add Vardeman's edits for abstract and intro

@@
 date: "August, 2016"
+linestretch: 2 
 output: 
@@
 abstract: |
-  A restricted Boltzmann machine (RBM) is an undirected graphical model constructed for discrete or continuous random variables, specified as having two layers, a hidden and a visible layer with no conditional dependency within a layer.
-In recent years, RBMs have risen to prominence due to their connection to deep learning - by treating a hidden layer of one RBM as the visible layer in a second RBM a deep architecture can be created.
-The method is claimed to have the ability to learn very complex and rich structures in data, making these models attractive for supervised learning.
+  A restricted Boltzmann machine (RBM) is an undirected graphical model constructed for discrete or continuous random variables, with two layers, one hidden and one visible, with no conditional dependency within a layer.
+In recent years, RBMs have risen to prominence due to their connection to deep learning.
+By treating a hidden layer of one RBM as the visible layer in a second RBM a deep architecture can be created.
+RBMs are thought to have the ability to encode very complex and rich structures in data, making them attractive for supervised learning.
 However, the generative behavior of RBMs has largely been unexplored.
-In this paper, we discuss the relationship between parameter specification and the prevalence of degenerate models as well as an exposition of the difficulties in fitting such highly flexible models in the binary case of RBMs.
+In this paper, we discuss the relationship between parameter RBM specification and the prevalence of degenerate models as well as the difficulties that arise in fitting such highly flexible models in the binary case.
 ---
@@
 
-The data mining and machine learning communities have recently found a great interest in deep learning, specifically in stacked restricted Boltzmann machines (RBM).
-(**TODO: deep learning, deep RBM references.**) A RBM is an undirected graphical model constructed for discrete or continuous random variables, specified as having two layers, a hidden and a visible layer with no conditional dependency within a layer.
-These models are typically used with success in classification of images.
-(**TODO: classification references.**) However, the model properties are largely unexplored in the literature and the fitting methodology remains heuristic based and abstruse.
-In this paper, we attempt to give a thorough understanding of the model and its behavior from the context of statistical model theory as well as explore the possibility of a rigorous fitting methodology.
-Through our study, we find the RBM model to be deficient in two fundamental ways.
+The data mining and machine learning communities have recently found a great interest in deep learning, specifically in stacked restricted Boltzmann machines (RBMs).
+(**TODO: deep learning, deep RBM references.**) A RBM is an undirected probabilistic graphical model (for discrete or continuous random variables) with two layers, one hidden and one visible, with no conditional dependency within a layer (**TODO: RBM references.**).
+These models have reportedly been used with success in classification of images.
+(**TODO: classification references.**) However, the model properties are largely unexplored in the literature and the commonly cited fitting methodology remains heuristic-based and abstruse.
+In this paper, we attempt to provide steps toward a thorough understanding of the model and its behavior from the perspective of statistical model theoryand then explore the possibility of a rigorous fitting methodology.
+We find the RBM model class to be deficient in two fundamental ways.
 
@@
 Neyman and Box agree with this assessment and go further to state that a model should "provide an explanation of the mechanism underlying the observed phenomena" [@lehmann1990model, @box1967discrimination].
-In the case of RBMs, these models often fail to generate data that resembles realistic data.
-Thus, providing an unsatisfactory conceptualization of the data generation process.
-Additionally, it is evidenced that RBMs easily exhibit instability in the parameter space.
+RBMs often fail to generate data that resemble realistic data and thus an unsatisfactory conceptualization of the data generation process.
+Additionally, we find that RBMs easily exhibit a kind of instability in the parameter space.
 In practice, this is seen when a single pixel change in an image results in a wildly different classification.
-These model impropriety measures has been documented in other deep architectures recently as well [@szegedy2013intriguing, @nguyen2014deep].
+Such model impropriety issues have  been documented in other deep architectures recently [@szegedy2013intriguing, @nguyen2014deep].
 We investigate these phenomena in RBMs in section \ref{explorations} through simulations of small, manageable examples.
@@
 Secondly, the fitting of these models is problematic.
-As the size of these models grows, a computationally intractable situation is created quickly when fitting with either maximum likelihood or through Bayesian methods.
-However, even in models small enough to feasibly fit, there are unique challenges due to the overparameterization of RBMs.
-In section \ref{bayesian-model-fitting} we try three Bayesian modelling techniques, each with the goal of avoiding parts of the parameter space that yield improper models.
-As could be expected, with computational complexity comes an increase in accuracy, but at the cost of feasibility.
-Additionally, because these models are overparametrized, any honest fitting method will reproduce the empirical distribution.
-Thus, there is no smoothing achieved with a RBM that is fit using a rigorous method.
-We are therefore very skeptical that any model built using these structures, like a deep Boltzmann machine, can achieve prediction or inference in a principled way.
+As the size of these models grows, both maximum likelihood and Bayesian methods of fitting quickly being intractable.
+But even when fitting is feasible, there are unique challenges due to the overparameterization of RBMs.
+In section \ref{bayesian-model-fitting} we compare three Bayesian modelling techniques, each proposed with the goal of avoiding parts of the RBM parameter space that yield improper models.
+As could be expected, with computational complexity comes an increase in fitting accuracy, but at the cost of feasibility.
+Additionally, because these models are highly overparametrized, any fully principled fitting method will reproduce the empirical distribution of most realistic training sets.
+There is no "smoothing" achieved with a RBM fit using a rigorous method.
+We are therefore led to be skeptical that any model built using these structures, like a deep Boltzmann machine, can achieve prediction or inference in a principled way.
 
-This paper is structured as follows, section \ref{background} gives a definition of the RBM including the joint distribution and shows the connection to deep learning.
+This paper is structured as follows, section \ref{background} gives a definition of the RBM including the joint distribution of hidden and visible variables and shows the connection to deep learning.
 Additionally, measures of model impropriety and methods of detecting them are defined.
-Section \ref{explorations} details our explorations into the model behavior and propriety (or lack thereof) of the RBM.
+Section \ref{explorations} details our explorations into the model behavior and propriety (or lack thereof) of the RBM class.
 We discuss three Bayesian fitting techniques with the goal of avoiding model impropriety in section \ref{bayesian-model-fitting} and wrap up with a discussion in section \ref{discussion}.
@@
 ### Connection to Deep Learning
+
 RBMs have risen to prominence in recent years due to their connection to deep learning [see @hinton2006fast, @salakhutdinov2012efficient, or @srivastava2013modeling for examples].

### 2016-09-03 writing/draft.Rmd
joint and connection edits

@@
 
-A restricted Boltzmann machine (RBM) is a specific undirected graphical model specified for discrete or continuous random variables, binary being the most common.
+A restricted Boltzmann machine (RBM) is an undirected graphical model specified for discrete or continuous random variables, binary variables being most common.
 In this paper, we consider the binary case of an RBM only.
-The architecture of an RBM is specified as having two layers, a hidden ($\mathcal{H}$) and a visible layer ($\mathcal{V}$), with no connections within a layer.
-An example of this structure is visualized in figure \ref{fig:rbm} with the hidden nodes indicated by gray circles and the visible nodes indicated by white circles.
+A RBM architecture has two layers, hidden ($\mathcal{H}$) and visible ($\mathcal{V}$), with no connections within a layer.
+An example of this structure is in figure \ref{fig:rbm} with the hidden nodes indicated by gray circles and the visible nodes indicated by white circles.
 
@@
   \resizebox{\linewidth}{!}{\input{tikz/rbm.tikz}}
-  \caption{An example restricted Boltzmann machine (RBM), which consists of two layers, a hidden ($\mathcal{H}$) and a visible layer ($\mathcal{V}$), with no connections within a layer.
-Hidden nodes indicated by gray circles and the visible nodes indicated by white circles.}
+  \caption{An example restricted Boltzmann machine (RBM), consisting of two layers, one hidden ($\mathcal{H}$) and one visible ($\mathcal{V}$), with no connections within a layer.
+Hidden nodes are indicated by gray circles and the visible nodes indicated by white circles.}
   \label{fig:rbm}
@@
 
-A common use for the RBM is to create features for use in classification. For example, a binary image could be classified where the pixel values would be considered the visible variables $v_i$.
+A common use for RBMs is to create features for use in classification. For example, binary images could be classified where the pixel values would be treated as the visible variables $v_i$.
 
@@
 Let $\boldsymbol x = \{h_1, \dots, h_H, v_1,\dots,v_V\}$ represent the states of the visible and hidden nodes in an RBM.
-Then we posit a parametric form for probabilities corresponding to the state of each node taking the value of $1$.
+We will consider both $\boldsymbol x \in \{0,1\}^{H + V}$ and $\boldsymbol x \in \{-1, 1\}^{H + V}$, two binary $H + V$-dimensional spaces.
+We posit a parametric form for probabilities corresponding to a potential vector of states of each node taking the value of $1$.
 
@@
 
-Let the normalizing constant in equation \ref{eqn:pmf} be denoted $\gamma(\boldsymbol \theta)$ [@Vardeman602, pp.163-167].
-Additionally, we can refer to $Q(\boldsymbol x) = \sum\limits_{i = 1}^H\sum\limits_{j=1}^V \theta_{ij} h_i v_j + \sum\limits_{i = 1}^H\theta_{hi} h_i + \sum\limits_{j = 1}^V\theta_{vj} v_j$ as the neg-potential function in our model.
+Let the normalizing constant in equation (\ref{eqn:pmf}) be denoted $\gamma(\boldsymbol \theta)$.It is standard to refer to $Q(\boldsymbol x) = \sum\limits_{i = 1}^H\sum\limits_{j=1}^V \theta_{ij} h_i v_j + \sum\limits_{i = 1}^H\theta_{hi} h_i + \sum\limits_{j = 1}^V\theta_{vj} v_j$ as the neg-potential function in this model.
 The RBM model is parameterized by $\boldsymbol \theta$, which contains two types of parameters, main effects and interaction effects.
-The main effects parameters ($\theta_{v_i}, \theta_{h_j}$) weight to the values of the visible and hidden nodes and the interaction effects parameters ($\theta_{ij}$) weight the values of the connections, or dependencies, between hidden and visible layers.
+The main effects parameters ($\theta_{v_i}, \theta_{h_j}$) weight the values of the visible and hidden nodes and the interaction effects parameters ($\theta_{ij}$) weight the values of the connections, or dependencies, between hidden and visible layers.
 
-Due to the potential size of the model the normalizing constant $\gamma(\boldsymbol \theta)$ can be unfeasible to calculate, making it impossible to estimate the model exactly.
-An approximation can be achieved using Gibbs sampling, which is a natural fit due to the restricted architecture of the RBM.
-Specifically, the conditional independence of each layer given the other allows for simulation of the main and hidden layers.
+Due to the potential size of the model the normalizing constant $\gamma(\boldsymbol \theta)$ can be practically impossible to calculate, making it impossible to estimate the model parameter vector in an obvious fashion.
+A kind of Gibbs sampling can be tried, which is natural due to the restricted architecture of the RBM.
+Specifically, the conditional independence of nodes in each layer given those in the other allows for simulation of the main and hidden layers.
 
@@
 RBMs have risen to prominence in recent years due to their connection to deep learning [see @hinton2006fast, @salakhutdinov2012efficient, or @srivastava2013modeling for examples].
-By stacking multiple layers of RBMs in a deep architecture, the method's proponents claim the ability to learn "internal representations that become increasingly complex, which is considered to be a promising way of solving object and speech recognition problems" [@salakhutdinov2009deep, pp.
+By stacking multiple layers of RBMs in a deep architecture, proponents of the models claim the ability to learn "internal representations that become increasingly complex, which is considered to be a promising way of solving object and speech recognition problems" [@salakhutdinov2009deep, pp.
 450].
-The stacking is achieved by treating a hidden layer of one RBM as the visible layer in a second RBM, and so on until the desired architecture is created.
+The stacking is achieved by treating a hidden layer of one RBM as the visible layer in a second RBM, and so on until the desired multi-layer architecture is created.
 

### 2016-09-03 writing/draft.Rmd
attempt to make degeneracy section make some sense

@@
 \label{eqn:pmf}
-f_{\boldsymbol \theta} (\boldsymbol x) = \frac{\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}
+f_{\boldsymbol \theta} (\boldsymbol x) = \frac{\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\gamma(\boldsymbol \theta)}
 \end{align}
 
-Let the normalizing constant in equation (\ref{eqn:pmf}) be denoted $\gamma(\boldsymbol \theta)$.It is standard to refer to $Q(\boldsymbol x) = \sum\limits_{i = 1}^H\sum\limits_{j=1}^V \theta_{ij} h_i v_j + \sum\limits_{i = 1}^H\theta_{hi} h_i + \sum\limits_{j = 1}^V\theta_{vj} v_j$ as the neg-potential function in this model.
+Where the denominator $\gamma(\boldsymbol \theta)$ is the normalizing function that ensures the distribution sums to one, $\gamma(\boldsymbol \theta) = \sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)$.
+It is standard to refer to $Q(\boldsymbol x) = \sum\limits_{i = 1}^H\sum\limits_{j=1}^V \theta_{ij} h_i v_j + \sum\limits_{i = 1}^H\theta_{hi} h_i + \sum\limits_{j = 1}^V\theta_{vj} v_j$ as the neg-potential function in this model.
 The RBM model is parameterized by $\boldsymbol \theta$, which contains two types of parameters, main effects and interaction effects.
@@
 ## Degeneracy, instability, and uninterpretability... Oh my!
-The highly flexible nature of the RBM ($H + V + H V$ parameters) makes three characteristics of model impropriety of particular concern, *degeneracy, instability, and uninterpretability*.
-In this section we define these characteristics, discuss how to detect them in the RBM, and point out their relationship to one another.
+The highly flexible nature of a RBM (having as it does $H + V + H V$ parameters) makes three kinds of potential model impropriety of concern, *degeneracy, instability, and uninterpretability*.
+In this section we define these characteristics, consider how to detect them in a RBM, and point out their relationship to one another.
 
 ### Near-degeneracy
-In Random Graph Model theory, *degeneracy* occurs when there is a disproportionate amount of probability placed on only a few elements of the sample space, $\mathcal{X}$, by the model.
-@handcock2003assessing define a model to be near degenerate if $\mu(\boldsymbol \theta)$, the mean parametrization on the model parameters, is close to the boundary of the convex hull of $\{t_k(\boldsymbol x): \boldsymbol x \in \mathcal{X}, k = 1, \dots, K\}$, where $t_k(\boldsymbol x), k = 1, \dots, K\}$ is the set of statistics in the neg-potential function $Q(\boldsymbol x)$ from equation \ref{eqn:pmf}.
-In other words, if random variables in the neg-potential function $Q(\cdot)$, having support set $\mathcal{S}$, have a collective mean $\mu(\boldsymbol \theta)$ close to the boundary of the convex hull of $\mathcal{S}$ then the model is near-degenerate, as defined by @handcock2003assessing.
+In Random Graph Model theory, *model degeneracy* means there is a disproportionate amount of probability placed on only a few elements of the sample space, $\mathcal{X}$, by the model.
+In the scenario of random graph models, $\mathcal{X}$ denotes all possible graphs that can be constructed from a set of nodes.
+An exponentially parameterized random graph model has a distribution of the form
+\begin{align*}
+f_{\boldsymbol \theta} (\boldsymbol x) = \frac{\exp\left(\boldsymbol \theta^T t(\boldsymbol x)\right)}{\gamma(\boldsymbol \theta)}
+\end{align*}
+where $\boldsymbol \theta \in \Theta \subset \mathbb{R}^q$ is the model parameter, and $t: \mathcal{X} \rightarrow \mathbb{R}^q$ are statistics based on the adjacency matrix of a graph.
+Again, $\gamma(\boldsymbol \theta)$ is the normalizing function that guarantees the distibrution is valid, $\gamma(\boldsymbol \theta) = \sum_{\boldsymbol x \in \mathcal{X}} \exp\left(\boldsymbol \theta^T t(\boldsymbol x)\right)$.
+Let $C$ be the convex hull of $\{t(\boldsymbol x): \boldsymbol x \in \mathcal{X}\}$.
+@handcock2003assessing classify an exponentially parametrized random graph model as *near-degenerate* if $\mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta}t(X)$ is close to the boundary of the $C$.
 
-The mean parameterization on the model parameters, $\mu(\boldsymbol \theta)$, for the RBM is defined as
+The RBM class of models can be thought to exhibit an analogous occurence of *near-degeneracy* when random variables in the neg-potential function $Q(\cdot)$ have a mean vector $\mu(\boldsymbol \theta)$ close to the boundary of the convex hull of $\mathcal{S}$, where $t(\boldsymbol x) = \{v_1, \dots, v_V, h_1, \dots, h_H, v_1 h_1, \dots, v_V h_H \} \in \mathcal{S}$ and the mean parameterization on the model parameters, $\mu(\boldsymbol \theta)$ is
 \begin{align*}
@@
 From these posterior predictive checks, it is evident that the marginalized likelihood method shows the closest fit to the actual data.
-However, this model requires a marginalization step, which is unfeasible for a model of any real size.
+However, this model requires a marginalization step, which is infeasible for a model of any real size.
 

### 2016-09-03 writing/draft.Rmd
update uninterpetability

@@
 
-The RBM class of models can be thought to exhibit an analogous occurence of *near-degeneracy* when random variables in the neg-potential function $Q(\cdot)$ have a mean vector $\mu(\boldsymbol \theta)$ close to the boundary of the convex hull of $\mathcal{S}$, where $t(\boldsymbol x) = \{v_1, \dots, v_V, h_1, \dots, h_H, v_1 h_1, \dots, v_V h_H \} \in \mathcal{S}$ and the mean parameterization on the model parameters, $\mu(\boldsymbol \theta)$ is
+The RBM class of models can be thought to exhibit an analogous occurence of *near-degeneracy* when random variables in the neg-potential function $Q(\cdot)$, $t(\boldsymbol x)$, have a mean vector $\mu(\boldsymbol \theta)$ close to the boundary of the convex hull of $\mathcal{X}$, where $t(\boldsymbol x) = \{v_1, \dots, v_V, h_1, \dots, h_H, v_1 h_1, \dots, v_V h_H \} \in \mathcal{X}$ and the mean parameterization on the model parameters, $\mu(\boldsymbol \theta)$ is
 \begin{align*}
-\mu(\boldsymbol \theta) &= \text{E}_{\boldsymbol \theta}\left[ t(\boldsymbol X) \right] \\
+\mu(\boldsymbol \theta) &= \text{E}_{\boldsymbol \theta} t(\boldsymbol X)  \\
 &= \sum\limits_{x \in \mathcal{X}} \left\{ t(\boldsymbol x) f_{\boldsymbol \theta} (\boldsymbol x) \right\} \\
@@
 ### Uninterpretability
-@kaiser2007statistical defines a measure of model impropriety called *uninterpretability*, which is characterized as due to the existence of dependence, marginal mean-structure no longer being maintained in the model.
-This measure concerns the difference between model expectations, E$\left[\boldsymbol X | \boldsymbol \theta\right]$, and expectations given independence, E$\left[\boldsymbol X | \emptyset\right ]$.
-Using this concept of dependence in a model, it is possible to understand what conditions lead to uninterpretability in a model versus those that guarantee interpretable models.
-In this method, a model is considered without uninterpretability if its expected value is not very different from the same model with assumed independence.
+@kaiser2007statistical defines a measure of model impropriety he calls *uninterpretability*, which is characterized by marginal mean-structure not being maintained in the model.
+A measure of this is the magnitude of the difference between model expectations, E$\left[\boldsymbol X | \boldsymbol \theta\right]$, and expectations given independence, E$\left[\boldsymbol X | \boldsymbol \theta^* \right]$, where $\boldsymbol \theta^*$ is defined to equal $\boldsymbol \theta$ with all $\theta_{ij} = 0$ for $i = 1, \dots, V, j = 1, \dots, H$.
+Using this concept it is possible to investigate what conditions lead to uninterpretability in a model versus those that guarantee interpretable models.
+A model is considered without uninterpretability if its expected value is not very different from the same model with assumed independence.
 The quantities to compare in the RBM case are
@@
 \begin{align*}
-\text{E}\left[\boldsymbol X | \emptyset \right] &= \sum\limits_{\boldsymbol x \in \mathcal{X}} \boldsymbol x \frac{\exp\left(\sum\limits_{i = 1}^V \theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)} \\
+\text{E}\left[\boldsymbol X | \boldsymbol 
+theta^*\right] &= \sum\limits_{\boldsymbol x \in \mathcal{X}} \boldsymbol x \frac{\exp\left(\sum\limits_{i = 1}^V \theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)} \\
 &= \frac{\sum\limits_{\boldsymbol x \in \mathcal{X}} x_k \exp\left(\sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}  \qquad k = 1, \dots, H + V
@@
 
-If $|\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] - \text{E}\left[\boldsymbol X | \emptyset\right]$ is large then the RBM is uninterpetable.
+If $|\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] - \text{E}\left[\boldsymbol X | \emptyset\right]$ is large then the RBM with parameter vector $\boldsymbol \theta$ $is uninterpetable.
 

### 2016-09-03 writing/draft.Rmd
update beginning of exporations section with Vardeman's comments

@@
 
-We explore the relationship between magnitudes of $\boldsymbol \theta$ and the three measures of model impropriety, near-degeneracy, instability, and uninterpretability, for varying sized RBM models through simulated data.
+We explore the relationship between magnitudes of $\boldsymbol \theta$ and the three measures of model impropriety, near-degeneracy, instability, and uninterpretability, for RBM models of varying size through simulated data.
 
 ## Tiny example
-To illustrate the ideas of graph near-degeneracy, instability, and uninterpretability in a RBM, we will consider a the smallest possible toy example that consists of one visible node $v_1$ and one hidden node $h_1$, both binary.
+To illustrate the ideas of graph near-degeneracy, instability, and uninterpretability in a RBM, we consider first the smallest possible (toy) example that consists of one visible node $v_1$ and one hidden node $h_1$, both binary.
 A schematic of this model can be found in figure \ref{fig:toymodel}.
@@
   \resizebox{1cm}{!}{\input{tikz/toymodel.tikz}}
-  \caption{A small example restricted Boltzmann machine (RBM), which consists of two nodes, one hidden and one visible.}
+  \caption{A small example restricted Boltzmann machine (RBM), with two nodes, one hidden and one visible.}
   \label{fig:toymodel}
@@
 \end{align*}
-These three functions can be visualized over multiple values of $\boldsymbol \theta$, as seen in figure \ref{fig:toymodel}.
+These three functions can be visualized over multiple values of $\boldsymbol \theta$, as seen in figure \ref{fig:degen_toy}.
 By examining these functions, we can see that as values of $\boldsymbol \theta$ grow larger in magnitude, at least one mean-parameterization functions is approaching the values 0 or 1, indicating a potential to be near to the boundary of the convex hull of $\{t_k(\boldsymbol x): \boldsymbol x \in \mathcal{X}, k = 1, \dots, K\}$.
@@
 
-```{r near_dependency, fig.cap="\\label{fig:toymodel}The three mean-parameterization functions of the model parameters for a RBM with one visible and one hidden node.", fig.height = 4, fig.align='center', message=FALSE, cache=TRUE}
+```{r near_dependency, fig.cap="\\label{fig:degen_toy}The three mean-parameterization functions of the model parameters for a RBM with one visible and one hidden node.", fig.height = 4, fig.align='center', message=FALSE, cache=TRUE}
 possibles <- stats(1, 1, type = "binary")

### 2016-09-05 writing/draft.Rmd
started on instability section

@@
 (**TODO: classification references.**) However, the model properties are largely unexplored in the literature and the commonly cited fitting methodology remains heuristic-based and abstruse.
-In this paper, we attempt to provide steps toward a thorough understanding of the model and its behavior from the perspective of statistical model theoryand then explore the possibility of a rigorous fitting methodology.
+In this paper, we attempt to provide steps toward a thorough understanding of the model and its behavior from the perspective of statistical model theory and then explore the possibility of a rigorous fitting methodology.
 We find the RBM model class to be deficient in two fundamental ways.
@@
 ### Instability
-@schweinberger2011instability introduces a related concept, *instability*, which occurs when small changes in the natural parameters result in large changes in probability masses.
-This can be characterized of as excessive sensitivity in the model, where a small change in $\boldsymbol x$ leads to a large change in the probability mass function.
-@schweinberger2011instability provides the following tool for detecting instability in the a discrete exponential family model.
+
+@schweinberger2011instability introduces a related concept, *instability*, which means [**TODO: that small changes in the natural parameters result in large changes in probability masses**].
+This is characterized by excessive sensitivity in the model, where a small change in $\boldsymbol x$ leads to a large change in the probability mass function.
+@schweinberger2011instability provides the following formal definition for instability in a discrete exponential family model where [**TODO: notation goes here**].
 
@@
 
-Within the context of the RBM, this definition concerns the joint distribution of both hidden and visible nodes, where $N$ is the number of nodes in the model, $H + V$.
-However, instability in terms of only the visible nodes is a more useful measure of model impropriety to be concerned with.
-To see this, consider the example of classifying images using an RBM.
+For RBMs, an alaogous concept could concern the joint distribution of both hidden and visible nodes, where $N$ is the number of nodes in the model, $H + V$.
+However, instability in terms of only the visible nodes is a more useful measure of model impropriety.
+To see this, consider modeling images using an RBM.
 One would be interested in the maximum change in the likelihood when a single pixel is changed in the visibles as a proxy for instability in the model.
-Thus, we can instead concern ourselves with the notion of instability that deals with the maximum likelihood ratio of two images that differ in only one pixel.
+So we concern ourselves with a notion of instability that deals with the maximum likelihood ratio of two vectors of visibless that differ at only one coordinate.
 
@@
 Let $R(\boldsymbol \theta) \equiv \max_{\boldsymbol v} \max_{\boldsymbol h}Q(\boldsymbol x) - \min_{\boldsymbol v}\max_{\boldsymbol h}Q(\boldsymbol x) - H\log 2$.
-Then, if $R(\boldsymbol \theta) > C*V$, this implies that the maximum log-likelihood ratio of two images that differ in only one pixel is greater than $C$.
-So, by examining $R(\boldsymbol\theta)/V$, we can determine a lower bound for the maximum log-likelihood ratio of two images that differ in only one pixel, which if large is consistent with the definition of instability detailed above.
+Then, if $R(\boldsymbol \theta) > C*V$, this implies that the maximum log-likelihood ratio of two observations that differ in only one element is greater than $C$.
+So, by examining $R(\boldsymbol\theta)/V$, we can determine a lower bound for the maximum log-likelihood ratio of two sets of observables that differ in only one element, which if large is consistent with the concept of instability detailed above.
 
-Instability in a model is in itself a measure of model impropriety, however there is also a relationship between instability and near-degeneracy.
+Instability in a model is in itself a measure of model impropriety, however there is also a relationship between instability and near-degeneracy in a discrete exponental model.
 

### 2016-09-11 writing/draft.Rmd
first stab at instability edits

@@
 where $\boldsymbol \theta \in \Theta \subset \mathbb{R}^q$ is the model parameter, and $t: \mathcal{X} \rightarrow \mathbb{R}^q$ are statistics based on the adjacency matrix of a graph.
-Again, $\gamma(\boldsymbol \theta)$ is the normalizing function that guarantees the distibrution is valid, $\gamma(\boldsymbol \theta) = \sum_{\boldsymbol x \in \mathcal{X}} \exp\left(\boldsymbol \theta^T t(\boldsymbol x)\right)$.
+Again, $\gamma(\boldsymbol \theta)$ is the normalizing function that guarantees the distribution is valid, $\gamma(\boldsymbol \theta) = \sum_{\boldsymbol x \in \mathcal{X}} \exp\left(\boldsymbol \theta^T t(\boldsymbol x)\right)$.
 Let $C$ be the convex hull of $\{t(\boldsymbol x): \boldsymbol x \in \mathcal{X}\}$.
@@
 
-@schweinberger2011instability introduces a related concept, *instability*, which means [**TODO: that small changes in the natural parameters result in large changes in probability masses**].
-This is characterized by excessive sensitivity in the model, where a small change in $\boldsymbol x$ leads to a large change in the probability mass function.
-@schweinberger2011instability provides the following formal definition for instability in a discrete exponential family model where [**TODO: notation goes here**].
+@schweinberger2011instability introduces a related concept, *instability*.
+Instability is characterized by excessive sensitivity in the model, where a small change in $\boldsymbol x$ leads to a large change in the probability mass function.
+Consider discrete exponential families of distributions with probability mass functions of the form
+\begin{align*}
+f_{\boldsymbol \theta} (\boldsymbol x) = \frac{\exp\left(\eta^T(\boldsymbol \theta) t(\boldsymbol x)\right)}{\gamma(\boldsymbol \theta)}
+\end{align*}
+where, $\eta(\boldsymbol \theta)$ is a vector of natural parameters, and $t: \mathcal{X} \rightarrow \mathbb{R}^q$ is a vector of sufficient statistics based on $\boldsymbol x \in \mathcal{X}^N$ for $\mathcal{X}$ a discrete set of $M$ elements and $N$ degrees of freedom.
+In the random graph application, $\mathcal{X}^N$ may correspond to the number of relationships among $n$ nodes in a graph.
+In the instance of RBMs, $N$ refers to the number of hiddens and visible nodes included in the model, $N = H + V$.
+Again, $\gamma(\boldsymbol \theta)$ is the normalizing function that guarantees the distribution is valid, $\gamma(\boldsymbol \theta) = \sum_{\boldsymbol x \in \mathcal{X}^N} \exp\left(\eta^T(\boldsymbol \theta) t(\boldsymbol x)\right))$.
+
+@schweinberger2011instability provides the following formal definition for instability in a discrete exponential family model.
 
@@
 \begin{align}
-\max\limits_{\boldsymbol x_N \in \mathcal{X}_N}[Q(\boldsymbol x_N)] - \min\limits_{\boldsymbol x_N \in \mathcal{X}_N}[Q(\boldsymbol x_N)] > CN \text{ for all } N > N_C \label{eq:unstable}
+\max\limits_{\boldsymbol x_N \in \mathcal{X}_N}[\eta^T(\boldsymbol \theta) t(\boldsymbol x)\right)] - \min\limits_{\boldsymbol x_N \in \mathcal{X}_N}[\eta^T(\boldsymbol \theta) t(\boldsymbol x)\right)] > CN \text{ for all } N > N_C \label{eq:unstable}
 \end{align}
@@
 \begin{align}
-\max\limits_{\boldsymbol x_N \in \mathcal{X}_N}[Q(\boldsymbol x_N)] - \min\limits_{\boldsymbol x_N \in \mathcal{X}_N}[Q(\boldsymbol x_N)] \le CN \text{ for all } N > N_C. \label{eq:stable}
+\max\limits_{\boldsymbol x_N \in \mathcal{X}_N}[\eta^T(\boldsymbol \theta) t(\boldsymbol x)\right)] - \min\limits_{\boldsymbol x_N \in \mathcal{X}_N}[\eta^T(\boldsymbol \theta) t(\boldsymbol x)\right)] \le CN \text{ for all } N > N_C.
+\label{eq:stable}
 \end{align}
@@
 
-For RBMs, an alaogous concept could concern the joint distribution of both hidden and visible nodes, where $N$ is the number of nodes in the model, $H + V$.
+For RBMs, an analagous concept could concern the joint distribution of both hidden and visible nodes, where $N$ is the number of nodes in the model, $H + V$.
 However, instability in terms of only the visible nodes is a more useful measure of model impropriety.
@@
 
-Instability in a model is in itself a measure of model impropriety, however there is also a relationship between instability and near-degeneracy in a discrete exponental model.
+Instability in a discrete exponential model is in itself a measure of model impropriety, however there is also a relationship between instability and near-degeneracy in a discrete exponental model.
 
 \begin{lemma}[Schweinberger (2011)]
-If a discrete exponential family distribution is unstable, then it is degenerate in the sense that for any $0 < \epsilon < 1$, $P_\theta(\boldsymbol X_N \in \mathcal{M}_{\epsilon, N}) \rightarrow 1$ as $N \rightarrow \infty$ where $\mathcal{M}_{\epsilon, N}$ is the subset of $\epsilon$-modes of the probability mass function.
+If a discrete exponential family distribution is unstable, then it is degenerate in the sense that for any $0 < \epsilon < 1$, $P_\theta(\boldsymbol X \in \mathcal{M}_{\epsilon, N}) \rightarrow 1$ as $N \rightarrow \infty$ where $\mathcal{M}_{\epsilon, N} = \{boldsymbol x \in \mathcal{X}^N:  \eta^T(\boldsymbol \theta) t(\boldsymbol x)\right) > (1-\epsilon)\max\limits_{\boldsymbol x_N \in \mathcal{X}_N}[\eta^T(\boldsymbol \theta) t(\boldsymbol x)\right)]$ is the set of points in the sample space that yield arbitrarily high values of the unnormalized pmf.
 \end{lemma}
 
-Thus, if a discrete exponential family model is unstable, then it is also degenerate.
+Thus, if a discrete exponential family model is unstable, then it is also near-degenerate.
+Discrete RBMs (the focus of this paper) are curved exponential family models for particular choice of sufficient statistics ($\{h_j, v_i, v_i h_j: i = 1, \dots, V, j= 1, \dots, H\}$), meaning instability in the RBM implies near-degeneracy as well.
 

### 2016-09-12 writing/draft.Rmd
through toy model, blue edits

@@
 author: "Andee Kaplan, Daniel Nordman, Stephen Vardeman"
-date: "August, 2016"
+date: "September, 2016"
 linestretch: 2 
@@
 Where the denominator $\gamma(\boldsymbol \theta)$ is the normalizing function that ensures the distribution sums to one, $\gamma(\boldsymbol \theta) = \sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)$.
-It is standard to refer to $Q(\boldsymbol x) = \sum\limits_{i = 1}^H\sum\limits_{j=1}^V \theta_{ij} h_i v_j + \sum\limits_{i = 1}^H\theta_{hi} h_i + \sum\limits_{j = 1}^V\theta_{vj} v_j$ as the neg-potential function in this model.
+Let $\boldsymbol \tau = \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{X}\} = \{v_i, h_j, v_i h_j: i = 1, \dots, V, j = 1, \dots H\}$ be the vector of sufficient statistics in the model and $Q(\boldsymbol x) = \sum\limits_{i = 1}^H\sum\limits_{j=1}^V \theta_{ij} h_i v_j + \sum\limits_{i = 1}^H\theta_{hi} h_i + \sum\limits_{j = 1}^V\theta_{vj} v_j$, the neg-potential function.
 The RBM model is parameterized by $\boldsymbol \theta$, which contains two types of parameters, main effects and interaction effects.
@@
 
-The RBM class of models can be thought to exhibit an analogous occurence of *near-degeneracy* when random variables in the neg-potential function $Q(\cdot)$, $t(\boldsymbol x)$, have a mean vector $\mu(\boldsymbol \theta)$ close to the boundary of the convex hull of $\mathcal{X}$, where $t(\boldsymbol x) = \{v_1, \dots, v_V, h_1, \dots, h_H, v_1 h_1, \dots, v_V h_H \} \in \mathcal{X}$ and the mean parameterization on the model parameters, $\mu(\boldsymbol \theta)$ is
+The RBM class of models can be thought to exhibit an analogous occurrence of *near-degeneracy* when random variables in the neg-potential function $Q(\cdot)$, $t(\boldsymbol x)$, have a mean vector $\mu(\boldsymbol \theta)$ close to the boundary of the convex hull of $\mathcal{X}$, where $t(\boldsymbol x) = \{v_1, \dots, v_V, h_1, \dots, h_H, v_1 h_1, \dots, v_V h_H \} \in \mathcal{X}$ and the mean parameterization on the model parameters, $\mu(\boldsymbol \theta)$ is
 \begin{align*}
@@
 \end{align*}
-where, $\eta(\boldsymbol \theta)$ is a vector of natural parameters, and $t: \mathcal{X} \rightarrow \mathbb{R}^q$ is a vector of sufficient statistics based on $\boldsymbol x \in \mathcal{X}^N$ for $\mathcal{X}$ a discrete set of $M$ elements and $N$ degrees of freedom.
-In the random graph application, $\mathcal{X}^N$ may correspond to the number of relationships among $n$ nodes in a graph.
-In the instance of RBMs, $N$ refers to the number of hiddens and visible nodes included in the model, $N = H + V$.
-Again, $\gamma(\boldsymbol \theta)$ is the normalizing function that guarantees the distribution is valid, $\gamma(\boldsymbol \theta) = \sum_{\boldsymbol x \in \mathcal{X}^N} \exp\left(\eta^T(\boldsymbol \theta) t(\boldsymbol x)\right))$.
-
 @schweinberger2011instability provides the following formal definition for instability in a discrete exponential family model.
@@
 \begin{align}
-\max\limits_{\boldsymbol x_N \in \mathcal{X}_N}[\eta^T(\boldsymbol \theta) t(\boldsymbol x)\right)] - \min\limits_{\boldsymbol x_N \in \mathcal{X}_N}[\eta^T(\boldsymbol \theta) t(\boldsymbol x)\right)] > CN \text{ for all } N > N_C \label{eq:unstable}
+\max\limits_{\boldsymbol x_N \in \mathcal{X}_N}[\eta^T(\boldsymbol \theta) t(\boldsymbol x)] - \min\limits_{\boldsymbol x_N \in \mathcal{X}_N}[\eta^T(\boldsymbol \theta) t(\boldsymbol x)] > CN \text{ for all } N > N_C \label{eq:unstable}
 \end{align}
@@
 \begin{align}
-\max\limits_{\boldsymbol x_N \in \mathcal{X}_N}[\eta^T(\boldsymbol \theta) t(\boldsymbol x)\right)] - \min\limits_{\boldsymbol x_N \in \mathcal{X}_N}[\eta^T(\boldsymbol \theta) t(\boldsymbol x)\right)] \le CN \text{ for all } N > N_C.
+\max\limits_{\boldsymbol x_N \in \mathcal{X}_N}[\eta^T(\boldsymbol \theta) t(\boldsymbol x)] - \min\limits_{\boldsymbol x_N \in \mathcal{X}_N}[\eta^T(\boldsymbol \theta) t(\boldsymbol x)] \le CN \text{ for all } N > N_C.
 \label{eq:stable}
@@
 
-For RBMs, an analagous concept could concern the joint distribution of both hidden and visible nodes, where $N$ is the number of nodes in the model, $H + V$.
+For RBMs, an analogous concept could concern the joint distribution of both hidden and visible nodes, where $N$ is the number of nodes in the model, $H + V$.
 However, instability in terms of only the visible nodes is a more useful measure of model impropriety.
@@
 One would be interested in the maximum change in the likelihood when a single pixel is changed in the visibles as a proxy for instability in the model.
-So we concern ourselves with a notion of instability that deals with the maximum likelihood ratio of two vectors of visibless that differ at only one coordinate.
+So we concern ourselves with a notion of instability that deals with the maximum likelihood ratio of two vectors of visibles that differ at only one coordinate.
 
@@
 
-Instability in a discrete exponential model is in itself a measure of model impropriety, however there is also a relationship between instability and near-degeneracy in a discrete exponental model.
+Instability in a discrete exponential model is in itself a measure of model impropriety, however there is also a relationship between instability and near-degeneracy in a discrete exponential model.
 
 \begin{lemma}[Schweinberger (2011)]
-If a discrete exponential family distribution is unstable, then it is degenerate in the sense that for any $0 < \epsilon < 1$, $P_\theta(\boldsymbol X \in \mathcal{M}_{\epsilon, N}) \rightarrow 1$ as $N \rightarrow \infty$ where $\mathcal{M}_{\epsilon, N} = \{boldsymbol x \in \mathcal{X}^N:  \eta^T(\boldsymbol \theta) t(\boldsymbol x)\right) > (1-\epsilon)\max\limits_{\boldsymbol x_N \in \mathcal{X}_N}[\eta^T(\boldsymbol \theta) t(\boldsymbol x)\right)]$ is the set of points in the sample space that yield arbitrarily high values of the unnormalized pmf.
+If a discrete exponential family distribution is unstable, then it is degenerate in the sense that for any $0 < \epsilon < 1$, $P_\theta(\boldsymbol X \in \mathcal{M}_{\epsilon, N}) \rightarrow 1$ as $N \rightarrow \infty$ where $\mathcal{M}_{\epsilon, N} = \{\boldsymbol x \in \mathcal{X}^N:  \eta^T(\boldsymbol \theta) t(\boldsymbol x) > (1-\epsilon)\max\limits_{\boldsymbol x_N \in \mathcal{X}_N}[\eta^T(\boldsymbol \theta) t(\boldsymbol x)]\}$ is the set of points in the sample space that yield arbitrarily high values of the unnormalized pmf.
 \end{lemma}
@@
 Thus, if a discrete exponential family model is unstable, then it is also near-degenerate.
-Discrete RBMs (the focus of this paper) are curved exponential family models for particular choice of sufficient statistics ($\{h_j, v_i, v_i h_j: i = 1, \dots, V, j= 1, \dots, H\}$), meaning instability in the RBM implies near-degeneracy as well.
+Discrete RBMs (the focus of this paper) are curved exponential family models for particular choice of sufficient statistics ($\boldsymbol \tau = \{h_j, v_i, v_i h_j: i = 1, \dots, V, j= 1, \dots, H\}$), meaning instability in the RBM implies near-degeneracy as well.
 
@@
 ## Tiny example
-To illustrate the ideas of graph near-degeneracy, instability, and uninterpretability in a RBM, we consider first the smallest possible (toy) example that consists of one visible node $v_1$ and one hidden node $h_1$, both binary.
+To illustrate the ideas of model near-degeneracy, instability, and uninterpretability in a RBM, we consider first the smallest possible (toy) example that consists of one visible node $v_1$ and one hidden node $h_1$ that are both binary.
 A schematic of this model can be found in figure \ref{fig:toymodel}.
+Because it seems most common, we'll begin by employing $0/1$ encoding of binary variables (both $h_1$ and $v_1$ taking values in $\{0,1\}$).
+Eventually we'll note in section \ref{data-coding-to-mitigate-degeneracy} that $-1/1$ coding has substantial advantages.
 
@@
 
-For this small model, we are able to investigate the symptoms of the three measure of model impropriety, near-degeneracy, instability uninterpretability.
-First, we are able to illustrate the shape of boundary of the convex hull of $\{t_k(\boldsymbol x): \boldsymbol x \in \mathcal{X}, k = 1, \dots, K\}$ and explore the behavior of the mean parameterization of the parameters because we are in three dimensions.
-In figure \ref{fig:toyhull}, the convex hull of our statistic space in three dimensions for the toy RBM with one visible and one hidden node from multiple perspectives enclosed by an unrestricted hull of $\{0,1\}^3$ space is shown.
-In this small model, $\{t(\boldsymbol x)\}$ is defined as $\{t(\boldsymbol x)\} = \{v_1, h_1, v_1 h_1\}$ by definition and form of the neg-potential function $Q(\boldsymbol x)$.
-Note that the convex hull of $\{t(\boldsymbol x)\}$ does not fill the unrestricted hull because of the relationship between the elements of $\{t(\boldsymbol x)\}$.
+For this small model, we are able to investigate the symptoms of model impropriety (near-degeneracy, instability uninterpretability).
+First, because we are in three dimensions, we are able to explicitly illustrate the shape of boundary of the convex hull of $\boldsymbol tau$ and explore the behavior of the mean parameterization of the parameters.
+In figure \ref{fig:toyhull}, the convex hull of our "statistic space", $\boldsymbol \tau$, in three dimensions for this toy problem is shown from multiple perspectives enclosed by the unit cube, it convex hull of $\{0,1\}^3$ space.
+In this small model, note that the convex hull of $\boldsymbol tau$ does not fill the unrestricted hull because of the relationship between the elements of $\{(v_1, h_1, v_1 h_1 : (v_1, h_1) \in \{0, 1\}^2 \}$.
 
@@
   \end{minipage}
-  \caption{The convex hull of our statistic space in three dimensions for the toy RBM with one visible and one hidden node from multiple perspectives enclosed by an unrestricted hull of $\{0,1\}^3$ space.
-Note that the convex hull of $\{t(\boldsymbol x)\}$ does not fill the unrestricted hull because of the relationship between the elements of $\{t(\boldsymbol x)\}$.}
+  \caption{The convex hull of the "statistic space"" in three dimensions for the toy RBM with one visible and one hidden node.
+The convex hull of $\boldsymbol \tau = \{\boldsymbol t(\boldsymbol x) : \boldsymbol x \in \mathcal{X}\}$ does not fill the unit cube because of the relationship between the elements of $\boldsymbol \tau$.}
 \label{fig:toyhull}
@@
 
-Additionally, we can compute the mean parametrization of the model parameters as 
+Additionally, we can compute the mean vector for $\boldsymbol t(\boldsymbol x)$ as a function of the model parameters as 
 \begin{align*}
-\mu(\boldsymbol \theta) &= \text{E}_{\boldsymbol \theta}\left[ t(X) \right] \\
-&= \sum\limits_{x \in \mathcal{X}} \left\{ t(x) \frac{\exp\left( \theta_{11} h_1 v_1 + \theta_{h1} h_1 + \theta_{v1} v_1\right)}{\sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left( \theta_{11} h_1 v_1 + \theta_{h1} h_1 + \theta_{v1} v_1\right)} \right\} \\
-&= \sum\limits_{x \in \mathcal{X}} \left\{ t(x) \frac{\exp\left( \theta_{11} h_1 v_1 + \theta_{h1} h_1 + \theta_{v1} v_1\right)}{\gamma(\boldsymbol \theta)} \right\} \\
+\boldsymbol \mu(\boldsymbol \theta) &= \text{E}_{\boldsymbol \theta}\left[ \boldsymbol t(\boldsymbol X) \right] \\
+&= \sum\limits_{\boldsymbol x \in \mathcal{X}} \left\{ \boldsymbol t(\boldsymbol x) \frac{\exp\left( \theta_{11} h_1 v_1 + \theta_{h1} h_1 + \theta_{v1} v_1\right)}{\sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left( \theta_{11} h_1 v_1 + \theta_{h1} h_1 + \theta_{v1} v_1\right)} \right\} \\
+&= \sum\limits_{\boldsymbol x \in \mathcal{X}} \left\{ t(x) \frac{\exp\left( \theta_{11} h_1 v_1 + \theta_{h1} h_1 + \theta_{v1} v_1\right)}{\gamma(\boldsymbol \theta)} \right\} \\
 &= \left[
@@
 \end{align*}
-These three functions can be visualized over multiple values of $\boldsymbol \theta$, as seen in figure \ref{fig:degen_toy}.
-By examining these functions, we can see that as values of $\boldsymbol \theta$ grow larger in magnitude, at least one mean-parameterization functions is approaching the values 0 or 1, indicating a potential to be near to the boundary of the convex hull of $\{t_k(\boldsymbol x): \boldsymbol x \in \mathcal{X}, k = 1, \dots, K\}$.
+The three parametric functions giving $\boldsymbol \mu(\boldsymbol \theta)$ can be visualized in figure \ref{fig:degen_toy}.
+By examining these, we see that as values of coordinates of $\boldsymbol \theta$ grow larger in magnitude, at least one mean function for the entries of $\boldsymbol t(\boldsymbol x)$ is approaching a value 0 or 1, indicating a potential for $\text{E}_{\boldsymbol \theta} \boldsymbol t(\boldsymbol x)$ to be near to the boundary of the convex hull of $\boldsymbol \tau$.
 Thus, for a very small example we can see the relationship between values of $\boldsymbol \theta$ and model degeneracy.
 
-```{r near_dependency, fig.cap="\\label{fig:degen_toy}The three mean-parameterization functions of the model parameters for a RBM with one visible and one hidden node.", fig.height = 4, fig.align='center', message=FALSE, cache=TRUE}
+```{r near_dependency, fig.cap="\\label{fig:degen_toy}The three parametric mean functions of sufficient statistics for a RBM with one visible and one hidden node.", fig.height = 4, fig.align='center', message=FALSE, cache=TRUE}
 possibles <- stats(1, 1, type = "binary")
@@
   mutate(theta_11 = factor(theta_11, labels = paste0("theta[11]==", unique(theta_11))),
-         mu_parameter = factor(mu_parameter, labels = paste0(gsub("_", "(theta[", unique(mu_parameter)), "])"))) %>%
+         mu_parameter = ifelse(mu_parameter == "mu_v1", "paste(E[theta], v[1])", 
+                               ifelse(mu_parameter == "mu_h1", "paste(E[theta], h[1])", "paste(E[theta], v[1], h[1])"))) %>%
+  #factor(mu_parameter, labels = paste0(gsub("_", "(theta[", unique(mu_parameter)), "])"))) %>%
   ggplot() +
@@
   ylab(expression(theta[h1])) +
-  scale_fill_gradient2("Mean-parameterization", low = "#af8dc3", high = "#7fbf7b", midpoint = .5, limits = c(0, 1)) +
+  scale_fill_gradient2("Mean", low = "#af8dc3", high = "#7fbf7b", midpoint = .5, limits = c(0, 1)) +
   theme(aspect.ratio = 1)
@@
 
-Secondly, we can look at $R(\boldsymbol \theta)$ for this tiny model, which gives us a lower bound for the loglikelihood ratio of the smallest possible change in $\boldsymbol x$, indicating instability.
+Secondly, we can look at $R(\boldsymbol \theta)$ for this tiny model.
+Recall that gives us a lower bound for the loglikelihood ratio of the minimum possible change in $\boldsymbol x$, indicating instability.
 
@@
 Figure \ref{fig:instab} shows $R(\boldsymbol \theta)/V$ for various values of $\boldsymbol \theta$.
-We can see that this quantity is increasing for larger magnitudes of $\boldsymbol \theta$, especially for large values of $\theta_11$.
-This may indicate instability as $|\boldsymbol \theta|$ is large, agreeing with the indications about near-degeneracy from looking at $\mu(\boldsymbol \theta)$.
+We can see that this quantity is increasing for larger magnitudes of $\boldsymbol \theta$, especially for large values of $\theta_{11}$.
+This suggests instability as $|\boldsymbol \theta|$ becomes large, agreeing with the concerns about near-degeneracy produced by consideration of $\boldsymbol \mu(\boldsymbol \theta)$.
 
 ```{r instab, fig.cap="\\label{fig:instab}$R(\\boldsymbol \\theta)/V$ for various values of $\\boldsymbol \\theta$ for the tiny example model.
-This quantity is increasing for larger positive values of $\\boldsymbol \\theta$.", fig.height = 4, fig.align='center'}
+(Recall that here $V$ is the number of visible nodes and here is $1$.
+This quantity is increasing for larger magnitudes of $\\boldsymbol \\theta$.", fig.height = 4, fig.align='center'}
 expand.grid(theta_h1 = seq(-5, 5, length.out = 21),
@@
 
-Finally, we can look at the difference between model expectations, E$\left[\boldsymbol X | \boldsymbol \theta\right]$, and expectations given independence, E$\left[\boldsymbol X | \emptyset\right ]$ for the tiny toy model to understand the effect of uninterpretability on the RBM.
+Finally, we can look at the difference between model expectations, E$\left[\boldsymbol X | \boldsymbol \theta\right]$, and expectations given independence, E$\left[\boldsymbol X | \boldsymbol \theta^*\right ]$ for the tiny toy model to understand the effect of $\boldsymbol \theta$ on uninterpretability in the RBM.
 
 \begin{align*}
-\left|\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] -   \text{E}\left[\boldsymbol X | \emptyset \right] \right|
-&= \left|\left[
+\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] -   \text{E}\left[\boldsymbol X | \boldsymbol \theta^* \right] 
+&= \left[
 \begin{matrix}
@@
 \frac{\exp\left(\theta_{h_1}\right) + \exp\left(+ \theta_{v_1} + \theta_{h_1}\right)}{\exp\left(\theta_{v_1}\right) + \exp\left(\theta_{h_1}\right) + \exp\left(+ \theta_{v_1} + \theta_{h_1}\right)} 
-\end{matrix} \right] \right|\\
+\end{matrix} \right] \\
 &= \left[
@@
 
-Again, we can inspect these functions over multiple values of $\boldsymbol \theta$ to look for a relationship between parameter values and uninterpretability for the toy RBM.
+Again, we can inspect these coordinate functions of $\boldsymbol \theta$ to look for a relationship between parameter values and uninterpretability for the toy RBM.
 
-```{r uninterp, fig.cap="\\label{fig:uninterp}The difference between model expectations, E$\\left[\\boldsymbol X | \\boldsymbol \\theta\\right]$, and expectations given independence, E$\\left[\\boldsymbol X | \\emptyset\\right ]$ for a RBM with one visible and one hidden node.", fig.height = 4, fig.align='center', message=FALSE, cache=TRUE, warning=FALSE}
+```{r uninterp, fig.cap="\\label{fig:uninterp}The difference between coordinates of model expectations, E$\\left[\\boldsymbol X | \\boldsymbol \\theta\\right]$, and expectations given independence, E$\\left[\\boldsymbol X | \\boldsymbol \\theta^* \\right ]$ for a RBM with one visible and one hidden node.", fig.height = 4, fig.align='center', message=FALSE, cache=TRUE, warning=FALSE}
 expand.grid(theta_h1 = seq(-5, 5, length.out = 21),
@@
   mutate(theta_11 = factor(theta_11, labels = paste0("theta[11]==", unique(theta_11))),
-         mu_parameter = factor(mu_parameter, labels = paste0(gsub("_", "(theta[", unique(mu_parameter)), "])"))) %>%
+          mu_parameter = ifelse(mu_parameter == "mu_v1", "paste(E[theta], v[1])", "paste(E[theta], h[1])")) %>%
   ggplot() +
@@
   ylab(expression(theta[h1])) +
-  scale_fill_gradient(expression(group("|", E(group("", bold(X), "|")*bold(theta)) - E(group("", bold(X), "|")*symbol("\306")), "|")), low = "yellow", high = "red", limits = c(0, 1)) +
+  scale_fill_gradient("Component difference", low = "yellow", high = "red", limits = c(0, 1)) +
   theme(aspect.ratio = 1)
@@
 
-Figure \ref{fig:uninterp} shows that the absolute difference between model expectations, E$\left[\boldsymbol X | \boldsymbol \theta\right]$, and expectations given independence, E$\left[\boldsymbol X | \emptyset\right ]$ for the RBM with one hidden and one visible node grows as the values of $\boldsymbol \theta$ are farther from zero, especially for large magnitudes of $\theta_11$.
-This is a third indication that parameters of large magnitude could lead to model impropriety in an RBM using a model with only one hidden and one visible node.
+Figure \ref{fig:uninterp} shows that the absolute difference between coordinates of the vector of model expectations, E$\left[\boldsymbol X | \boldsymbol \theta\right]$, and corresponding expectations given independence, E$\left[\boldsymbol X | \boldsymbol \theta^*\right ]$ for the toy RBM grow as the values of $\boldsymbol \theta$ are farther from zero, especially for large magnitudes of $\theta_{11}$.
+This is a third indication that parameter vectors of large magnitude could lead to model impropriety in an RBM.
 
@@
 Multiple encodings of the binary variables are possible.
-For example, we could allow $\mathcal{H} \subset \{0,1\}^H, \mathcal{V} \subset \{0,1\}^V$, as in the previous sections or we could encode the state of the variables as $\mathcal{H} \subset \{-1,1\}^H, \mathcal{V} \subset \{-1,1\}^V$.
-This will result in $t(\boldsymbol x) \in \{0,1\}^{H + V + HV}$ or $t(\boldsymbol x) \in \{-1,1\}^{H + V + HV}$ depending on how we encode on and off states in the nodes.
-To explore the effect of this encoding on the potential near degeneracy of an RBM, we can explore the behavior of the ratio of the volume of the unrestricted convex hull to the restricted convex hull of $\{t(\boldsymbol x)\}$ to determine how much the restriction in our statistics by including the cross product terms incurs.
+For example, we could allow $\mathcal{H} \in \{0,1\}^H, \mathcal{V} \in \{0,1\}^V$, as in the previous sections or we could encode the state of the variables as $\mathcal{H} \in \{-1,1\}^H, \mathcal{V} \in \{-1,1\}^V$.
+This will result in $\boldsymbol t(\boldsymbol x) \in \{0,1\}^{H + V + HV}$ or $\boldsymbol t(\boldsymbol x) \in \{-1,1\}^{H + V + HV}$ depending on how we encode "on" and "off" states in the nodes.
+To explore the effect of this encoding on the potential for apparent near-degeneracy of an RBM, we can consider the ratio of the volume of the hypercube with corners at elements of $\boldsymbol \tau$ to the volume of the convex hull of $\boldsymbol \tau$.
 
-For the small two node example, the binary encoding loses `r round((1-m1_bin$c_hull$vol)*100,2)`% of the volume compared to unrestricted space and the negative encoding loses `r round((1-m1_neg$c_hull$vol/(2^3))*100,2)`% of the volume compared to unrestricted space.
-This notion of lost volume can be helpful to conceptualize how difficult it will be for the mean parameterized vector $\mu(\boldsymbol \theta)$ to avoid the boundary, and thus avoid near degeneracy.
-In fact, if we look at the ratio of volume within a convex hull versus the volume of an unrestricted hull, we can see a relationship emerge as the number of nodes (and thus parameters) increases.
+For the small two node example, the $0/1$ encoding loses `r round((1-m1_bin$c_hull$vol)*100,2)`% of the volume of $[0,1]^3$ and the $-1/1$ encoding loses `r round((1-m1_neg$c_hull$vol/(2^3))*100,2)`% of the volume of $[-1,1]^3$.
+While there is clearly a one-to-one mapping of models for these two possible codings, this notion of lost volume can be helpful to conceptualize how difficult it will be for the mean vector $\boldsymbol \mu(\boldsymbol \theta)$ to avoid the boundary, and thus avoid near-degeneracy.
+In fact, if we look at the ratio of volume within a convex hull defined by $\boldsymbol \tau$ and the corresponding hypercube, we can see a relationship emerge as the number of nodes (and thus parameters) increases.
 In figure \ref{fig:volume_plot}, it is evident that as the number of parameters increases, this ratio is decreasing at an increasing rate, meaning it gets more and more difficult to avoid the boundary of the convex hull and thus near degeneracy.
-Additionally, it appears that the $\{-1,1\}$ encoding suffers slightly less from this problem.
+Additionally, it appears that the $-1/1$ encoding suffers slightly less from this problem.
+This suggests that if intuition about models is to be formed in terms of volume fractions of vectors of "non-pathological" means for $\boldsymbol t(\boldsymbol x)$, $-1/1$ encoding is preferable to $0/1$ coding.
 
-```{r volume_plot, echo=FALSE, fig.cap="\\label{fig:volume_plot}The relationship between volume within the convex hull of our statistics and the convex hull of unrestricted space for different configurations of nodes.
-We compare the two encodings for on and off in a node.", out.height = "3in", fig.align='center'}
+```{r volume_plot, echo=FALSE, fig.cap="\\label{fig:volume_plot}The relationship between volume of the convex hull of possible values of the RBM sufficient statistics and the cube containing it for different configurations of nodes.", out.height = "3in", fig.align='center'}
 test_cases <- expand.grid(data.frame(1:4)[,rep(1,2)]) %>% 
@@
 
-In addition to the argument that the $\{-1,1\}$ data encoding mitigates some near-degeneracy it also has the benefit of providing a guaranteed non-degenerate model at $\boldsymbol \theta = \boldsymbol 0 \in \mathbb{R}^{H + V + H*V}$.
+In addition to the argument that the $-1/1$ data encoding mitigates some perceived prevalence of near-degeneracy it also has the benefit of providing a guaranteed non-degenerate model at $\boldsymbol \theta = \boldsymbol 0 \in \mathbb{R}^{H + V + H*V}$.
 The proof of this is included in appendix \ref{appendix:center}.
+And while from some computing perspectives $0/1$ coding might seem most natural, the $-1/1$ coding is far more convenient and interpretable from the point of view of statistical modeling, where it makes parameters simply interpreted in terms of symmetrically defined main effects and interactions.
+In light of all of these matters we will henceforth employ the $-1/1$ coding.
 
 **TODO: Add center of the universe proof**
-
-Thus, moving forward we will restrict our examinations to the $\{-1, 1\}$-encoding of the binary variables in the RBM.
 

### 2016-09-12 writing/draft.Rmd
worked through all edits from Vardeman

@@
   ylab(expression(theta[h1])) +
-  scale_fill_gradient("Component difference", low = "yellow", high = "red", limits = c(0, 1)) +
+  scale_fill_gradient("Component absolute difference", low = "yellow", high = "red", limits = c(0, 1)) +
   theme(aspect.ratio = 1)
@@
 In addition to the argument that the $-1/1$ data encoding mitigates some perceived prevalence of near-degeneracy it also has the benefit of providing a guaranteed non-degenerate model at $\boldsymbol \theta = \boldsymbol 0 \in \mathbb{R}^{H + V + H*V}$.
-The proof of this is included in appendix \ref{appendix:center}.
+The proof of this is included in appendix \ref{center-of-the-universe}.
 And while from some computing perspectives $0/1$ coding might seem most natural, the $-1/1$ coding is far more convenient and interpretable from the point of view of statistical modeling, where it makes parameters simply interpreted in terms of symmetrically defined main effects and interactions.
 In light of all of these matters we will henceforth employ the $-1/1$ coding.
-
-**TODO: Add center of the universe proof**
 
@@
   ungroup() %>% 
-  mutate(Hiddens = paste0("Hiddens: ", H), Visibles = paste0("Visibles: ", V)) -> convex_hull_summary
+  mutate(Hiddens = paste0("H = ", H), Visibles = paste0("V = ", V)) -> convex_hull_summary
 
@@
   summarise(max_abs_diff = mean(max_abs_diff)) %>%
-  mutate(Hiddens = paste0("Hiddens: ", H), Visibles = paste0("Visibles: ", V)) -> exp_vals_summary
+  mutate(Hiddens = paste0("H = ", H), Visibles = paste0("V = ", V)) -> exp_vals_summary
 
@@
 
-To explore the behavior of the RBM parameters $\boldsymbol \theta$ as it related to model degeneracy and near degeneracy, we example various models of small size.
-For $H, V \in \{1, \dots, 4\}$, we sample uniformly `r dim(res$samp[[1]])[1]` values of $\boldsymbol \theta$ with various magnitudes.
-For each set of parameters we then calculate the three metrics of model impropriety introduced in section \ref{degeneracy-instability-and-uninterpretability-oh-my}, $\mu(\boldsymbol \theta)$, $R(\boldsymbol x)/V$, and $\left|\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] -   \text{E}\left[\boldsymbol X | \emptyset \right] \right|$.
-In the case of near-degeneracy, we can go further and classify each model as near-degenerate or viable based on the distance of $\mu(\boldsymbol \theta)$ from the convex hull of $\{t_k(\boldsymbol x): \boldsymbol x \in \mathcal{X}, k = 1, \dots, K\}$ and look at the fraction of models that were near-degenerate for each combination of magnitude of $\boldsymbol \theta$ and model size.
+To explore the behavior of the RBM parameters $\boldsymbol \theta$ as it relates to near-degeneracy, instability, and uninterpretability.
+we consider models of small size.
+For $H, V \in \{1, \dots, 4\}$, we sample `r dim(res$samp[[1]])[1]` values of $\boldsymbol \theta$ with various magnitudes (details to follow).
+For each set of parameters we then calculate the three metrics of model impropriety introduced in section \ref{degeneracy-instability-and-uninterpretability-oh-my}, $\boldsymbol \mu(\boldsymbol \theta)$, $R(\boldsymbol x)/V$, and the coordinates of $\left|\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] -   \text{E}\left[\boldsymbol X | \boldsymbol \theta^* \right] \right|$.
+In the case of near-degeneracy, we can go further and classify each model as near-degenerate or "viable" based on the distance of $\boldsymbol \mu(\boldsymbol \theta)$ from the boundary of the convex hull of $\boldsymbol \tau$ and look at the fraction of models that are "near-degenerate", meaning they are within a small distance to the boundary of the convex hull.
 
-In this numerical experiment, we split $\boldsymbol \theta$ into $\boldsymbol \theta_{interaction}$ and $\boldsymbol \theta_{main}$, in reference to which sufficient statistics the parameters correspond to, and allow the two types of terms to have varying average magnitudes, $||\boldsymbol \theta_{main} || /(H+V)$ and $||\boldsymbol \theta_{interaction} || /(H*V)$.
-These magnitudes vary on a grid between `r min(plot_dat_grid$r1/(plot_dat_grid$H + plot_dat_grid$V))` and `r max(plot_dat_grid$r1/(plot_dat_grid$H + plot_dat_grid$V))`.
-By looking at the average magnitudes, we are able to consider the potential benefit of shrinking each parameter value $\theta_i$ towards zero in a Bayesian fitting technique.
-The results of this numerical study can be seen in figures \ref{fig:degen_plots}, \ref{fig:instab_plots}, and \ref{fig:uninterp_plots}.
-From these three figures, it s clear to see that all three indications of model impropriety show higher values the larger the magnitude of the parameters.
-Additionally, since there are $H*V$ interaction terms in $\boldsymbol \theta$ versus only $H + V$ main effects terms, for large models this dictates that there are many more interaction parameters in the model.
-And so, perhaps limiting the magnitude of the overall distance from the origin for $\boldsymbol \theta_{interaction}$ may help to limit the effects of model impropriety.
+In this numerical experiment, we split $\boldsymbol \theta$ into $\boldsymbol \theta_{interaction}$ and $\boldsymbol \theta_{main}$, in reference to which sufficient statistics the parameters correspond to (whether they multiply a $v_i$ or a $h_j$ or they multiply a $v_i h_j$), and allow the two types of terms to have varying average magnitudes, $||\boldsymbol \theta_{main} || /(H+V)$ and $||\boldsymbol \theta_{interaction} || /(H*V)$.
+These average magnitudes vary on a grid between `r min(plot_dat_grid$r1/(plot_dat_grid$H + plot_dat_grid$V))` and `r max(plot_dat_grid$r1/(plot_dat_grid$H + plot_dat_grid$V))` with `r length(unique(plot_dat_grid$r1/(plot_dat_grid$H + plot_dat_grid$V)))` breaks, yielding `r length(unique(plot_dat_grid$r1/(plot_dat_grid$H + plot_dat_grid$V)))^2` grid points.
+(By looking at the average magnitudes, we are able to later consider the potential benefit of shrinking each parameter value $\theta_i$ towards zero in a Bayesian fitting technique.) At each point in the grid, `r dim(res$samp[[1]])[1]` vectors are uniformly sampled on a sphere with radius corresponding to the first coordinate in the grid ($\boldsymbol \theta_{main}$) and `r dim(res$samp[[1]])[1]` vectors are uniformly sampled on a sphere with radius corresponding to the second coordinate in the grid ($\boldsymbol \theta_{main}$) via sums of squared and scaled iid Normal$(0, 1)$ variables.
+These vectors are then paired to create `r dim(res$samp[[1]])[1]` values of $\boldsymbol \theta$ with magnitudes at each point in the grid.
 
-```{r degen_plots, fig.cap="\\label{fig:degen_plots}A numerical experiment looking at the fraction of models that were near-degenerate for each combination of magnitude of $\\boldsymbol \\theta$ and model size.
-The thick black line shows the level where the fraction of -near degenerate models is .05.", fig.height=4}
+The results of this numerical study are summarized in figures \ref{fig:degen_plots}, \ref{fig:instab_plots}, and \ref{fig:uninterp_plots}.
+From these three figures, it s clear to see that all three measures of model impropriety show higher values the larger the magnitude of the parameter vectors.
+Additionally, since there are $H*V$ interaction terms in $\boldsymbol \theta$ versus only $H + V$ main effects terms, for large models there are many more interaction parameters than main effects in the model.
+And so, severely limiting the magnitude of the indivisual interactions may well help prevent model impropriety.
+
+```{r degen_plots, fig.cap="\\label{fig:degen_plots}Results from the numerical experiment, here looking at the fraction of models that were near-degenerate for each combination of magnitude of $\\boldsymbol \\theta$ and model size.
+Black lines show the contour levels for fraction of near-degenerace, while the thick black line shows the level where the fraction of near-degenerate models is .05.", fig.height=4}
 three_ways %>%
@@
   facet_grid(Visibles~Hiddens) +
-  xlab(expression(paste(group("||", theta[main], "||"), "/(H + V)"))) +
-  ylab(expression(paste(group("||", theta[interaction], "||"), "/(H*V)"))) +
+  xlab(expression(paste(paste("||", theta[main], "||", "/(H + V)")))) +
+  ylab(expression(paste(paste("||", theta[interaction], "||", "/(H*V)")))) +
   theme(aspect.ratio = 1) -> p.degen
@@
   geom_abline(aes(intercept = 0, slope = 1), alpha = .5, lty = 2) +
-  scale_fill_gradient(expression(group("|", E(group("", bold(X), "|")*bold(theta)) - E(group("", bold(X), "|")*symbol("\306")), "|")), low = "yellow", high = "red", limits = c(0,2)) +
+  scale_fill_gradient("Mean max absolute difference", low = "yellow", high = "red", limits = c(0,2)) +
   facet_grid(Visibles~Hiddens) +
-  xlab(expression(paste(group("||", theta[main], "||"), "/(H + V)"))) +
-  ylab(expression(paste(group("||", theta[interaction], "||"), "/(H*V)"))) +
+  xlab(expression(paste(paste("||", theta[main], "||", "/(H + V)")))) +
+  ylab(expression(paste(paste("||", theta[interaction], "||", "/(H*V)")))) +
   theme(aspect.ratio = 1) -> p.exp_diff
@@
   facet_grid(Visibles~Hiddens) +
-  xlab(expression(paste(group("||", theta[main], "||"), "/(H + V)"))) +
-  ylab(expression(paste(group("||", theta[interaction], "||"), "/(H*V)"))) +
+  xlab(expression(paste(paste("||", theta[main], "||", "/(H + V)")))) +
+  ylab(expression(paste(paste("||", theta[interaction], "||", "/(H*V)")))) +
   theme(aspect.ratio = 1) -> p.max_q
@@
   geom_abline(aes(intercept = 0, slope = 1), alpha = .5, lty = 2) +
-  scale_fill_gradient(expression(frac(R(theta), V)), low = "yellow", high = "red") +
+  scale_fill_gradient(expression(frac(bar(R(theta)), V)), low = "yellow", high = "red") +
   facet_grid(Visibles~Hiddens) +
-  xlab(expression(paste(group("||", theta[main], "||"), "/(H + V)"))) +
-  ylab(expression(paste(group("||", theta[interaction], "||"), "/(H*V)"))) +
+  xlab(expression(paste(paste("||", theta[main], "||", "/(H + V)")))) +
+  ylab(expression(paste(paste("||", theta[interaction], "||", "/(H*V)")))) +
   theme(aspect.ratio = 1) -> p.min_max_q
@@
 
-```{r instab_plots_2, fig.cap="\\label{fig:instab_plots}A numerical experiment looking at $R(\\boldsymbol \\theta)/V$ for each combination of magnitude of $\\boldsymbol \\theta$ and model size.
-As the magnitude of $\\boldsymbol \\theta$ grows, so does the value of this metric, indicating instability in the model.", fig.height=4}
+```{r instab_plots_2, fig.cap="\\label{fig:instab_plots}Results from the experiment, here looking at the sample mean value of $R(\\boldsymbol \\theta)/V$ at each grid point for each combination of magnitude of $\\boldsymbol \\theta$ and model size.
+As the magnitude of $\\boldsymbol \\theta$ grows, so does the value of this metric, indicating typical instability in the model.", fig.height=4}
 p.min_max_q
@@
 
-```{r uninterp_plots, fig.cap="\\label{fig:uninterp_plots}The difference between model expectations, E$\\left[\\boldsymbol X | \\boldsymbol \\theta\\right]$, and expectations given independence, E$\\left[\\boldsymbol X | \\emptyset\\right ]$ for each combination of magnitude of $\\boldsymbol \\theta$ and model size.
-Larger magnitudes of $\\boldsymbol \\theta$ correspond to larger differences, thus indicating higher uninterpretability.", fig.height=4}
+```{r uninterp_plots, fig.cap="\\label{fig:uninterp_plots}The sample mean of the maximum component of the absolute difference between the model expectation vector, E$\\left[\\boldsymbol X | \\boldsymbol \\theta\\right]$, and the expectation vector given independence, E$\\left[\\boldsymbol X | \\boldsymbol \\theta^* \\right ]$.
+Larger magnitudes of $\\boldsymbol \\theta$ correspond to larger differences, thus indicating reduced interpretability.", fig.height=4}
 p.exp_diff
@@
 
-Figure \ref{fig:split_plots_dist} shows the fraction near-degenerate for each magnitude of $\boldsymbol \theta$ for each model.
+Figure \ref{fig:split_plots_dist} shows the fraction near-degenerate for each magnitude of $\boldsymbol \theta$ for each model architecture.
 For each number of visibles in the model, the fraction near-degenerate becomes greater than zero at larger values of $||\boldsymbol \theta||$ as $H$ increases and the slope becomes steeper as $H$ increases also.
@@
 
-```{r split_plots2, warning=FALSE, fig.cap="\\label{fig:split_plots_dist}The fraction near-degenerate for each magnitude of $\\boldsymbol \\theta$ for each model.
+```{r split_plots2, warning=FALSE, fig.cap="\\label{fig:split_plots_dist}The fraction near-degenerate for each magnitude of $\\boldsymbol \\theta$.
 For each number of visibles in the model, the fraction near-degenerate becomes greater than zero at larger values of $||\\boldsymbol \\theta||$ as $H$ increases and the slope becomes steeper as $H$ increases also.", fig.height=5}
@@
   summarise(frac_degen = sum(near_hull)/n()) %>%
-  mutate(Hiddens = as.character(H), Visibles = paste0("Visibles: ", V)) %>%
+  mutate(Hiddens = as.character(H), Visibles = paste0("V = ", V)) %>%
   ggplot() +
@@
   ylim(c(0,1)) +
-  xlab(expression(paste(group("||", theta, "||"), "/(H + V + H*V), Average distance from the origin"))) +
-  ylab("Fraction Near-degenerate")
+  xlab(expression(paste("||", theta, "||", "/(H + V + H*V), Average distance from the origin"))) +
+  ylab("Fraction Near-degenerate") +
+  scale_colour_discrete("H")
 ```
@@
 ```
-It is evidenced from these manageable examples that RBMs very easily are near-degenerate, unstable, and uninterpretable for large portions of the parameter space.
-The question becomes, can a rigorous fitting method be developed that avoids these issues?
+These manageable examples indicate that RBMs are near-degenerate, unstable, and uninterpretable for large portions of the parameter space.
+This, however, is not the only potential problem to be faced when using these models.
+There is the matter of pricipled/rigorous fitting of RBM models.
 
-# Bayesian model fitting
-To avoid model impropriety, we want to avoid parts of the parameter space that lead to near-degeneracy, instability, and uninterpretability.
-Based on the evidence in section \ref{exploring-manageable-examples}, one idea is to shrink $\boldsymbol \theta$ by specifying priors that place low probability on high values of $\boldsymbol \theta$, specifically focusing on shrinking $\boldsymbol \theta_{interaction}$ more than $\boldsymbol \theta_{main}$.
+# Model Fitting
 
-Using this idea of shrinkage, we simulated $n = 5,000$ images with four pixels (four visible nodes) from an RBM model with 4 hiddens and fit the RBM using Bayesian methods, and three separate methods.
+Typically, fitting a RBM via maximum likelihood (ML) methods will be infeasible mainly due to the intractibility of the normalizing term $\gamma(\boldsymbol \theta)$ in a model of any realistic size.
+Ad hoc methods are used instead, which aim to avoid this problem by using stochastic ML that employ a small number of MCMC draws.
 
-1. *Trick prior.* Cancel out normalizing term, resulting full conditionals normally distributed. 
+However, computational concerns are not the only issues with fitting an RBM using ML.
+In addition, the RBM model, with the appropriate choice of parameters, has the potential to re-create any distribution for the data (i.e., reproduce any specification of cell probabilities for the -1/1 valued data outcomes).
+This hyper-flexibility dictates that ML fitting will merely replicate the empirical data distribution, which is the optimal nonparametric solution.
+In a data application with a large sample space it is unlikely that the training set will include one of each possible image, thus there will be possible images that are not realized in the training set, unlike this small example.
+The presence of empty cells in the empirical distribution drives coordinates of $\boldsymbol \theta$ to $\infty$ in order to make the corresponding cell probabilities arbitrarily close to zero.
+As we have already discussed in secion \ref{explorations}, large magnitudes of $\boldsymbol \theta$ lead to model impropriety in the RBM.
+A detailed discussion of this phenomenon are included in appendix \ref{incomplete-cases}.
+
+Here we consider what might be done in a principled manner, testing on a $V = H = 4$ case that already stretched the limits of what is computable - in particular we consider Bayes methods.
+
+## Bayesian model fitting
+To avoid model impropriety for a fitted RBM, we want to avoid parts of the parameter space $\mathbb{R}^{V + H + VH}$ that lead to near-degeneracy, instability, and uninterpretability.
+Motivated by the insights in section \ref{exploring-manageable-examples}, one idea is to shrink $\boldsymbol \theta$ toward $\boldsymbol 0$ by specifying priors that place low probability on large values of $||\boldsymbol \theta||$, specifically focusing on shrinking $\boldsymbol \theta_{interaction}$ more than $\boldsymbol \theta_{main}$.
+
+```{r param-table}
+load("data/params_theta.Rdata")
+
+params_good <- list(main_hidden = sample.params %>% ungroup() %>% filter(!near_hull) %>% select(starts_with("h"), -H) %>% data.matrix(),
+                    main_visible = sample.params %>% ungroup() %>% filter(!near_hull) %>% select(starts_with("v"), -V) %>% data.matrix(),
+                    interaction = sample.params %>% ungroup() %>% filter(!near_hull) %>% select(starts_with("theta")) %>% data.matrix() %>% matrix(4))
+
+sample.params %>% 
+  ungroup %>% 
+  filter(!near_hull) %>% select(starts_with("v"), starts_with("h"), starts_with("theta"), -H, -V) %>%
+  rename(`$\\theta_{v1}$` = v1,
+         `$\\theta_{v2}$` = v2,
+         `$\\theta_{v3}$` = v3,
+         `$\\theta_{v4}$` = v4,
+         `$\\theta_{h1}$` = h1,
+         `$\\theta_{h2}$` = h2,
+         `$\\theta_{h3}$` = h3,
+         `$\\theta_{h4}$` = h4,
+         `$\\theta_{11}$` = theta11,
+         `$\\theta_{12}$` = theta12,
+         `$\\theta_{13}$` = theta13,
+         `$\\theta_{14}$` = theta14,
+         `$\\theta_{21}$` = theta21,
+         `$\\theta_{22}$` = theta22,
+         `$\\theta_{23}$` = theta23,
+         `$\\theta_{24}$` = theta24,
+         `$\\theta_{31}$` = theta31,
+         `$\\theta_{32}$` = theta32,
+         `$\\theta_{33}$` = theta33,
+         `$\\theta_{34}$` = theta34,
+         `$\\theta_{41}$` = theta41,
+         `$\\theta_{42}$` = theta42,
+         `$\\theta_{43}$` = theta43,
+         `$\\theta_{44}$` = theta44) %>%
+  gather("Parameter", "Value") -> tbl
+
+cbind(tbl[1:8,], tbl[9:16,], tbl[17:24,]) %>%
+  kable(caption = "\\label{tab:theta}Parameters used to fit a test case with $V = H = 4$.
+This parameter vector was chosen as a sampled value of $\\boldsymbol \\theta$ that was not near the convex hull of the sufficient statistics for a grid point in figure \\ref{fig:degen_plots} with $< 5%$ near-degeneracy.")
+```
+
+We considered a test case with $V = H = 4$ and parameters given in table \ref{tab:theta}.
+This parameter vector was chosen as a sampled value of $\\boldsymbol \\theta$ that was not near the convex hull of the sufficient statistics for a grid point in figure \\ref{fig:degen_plots} with $< 5%$ near-degeneracy.
+We simulated $n = 5,000$ as a training set and fit the RBM using three Bayes methodologies.
+These involved the following:
+
+1.
+*A trick prior.* Here we cancel out normalizing term in the likelihood so that resulting full conditionals of $\boldsymbol \theta$ are multivariate Normal.
+The $h_j$ are carried allong in the MCMC sampling from the posterior as latent variables.
     \begin{align*}
@@
     \end{align*}
-    [@li2014biclustering].
+    This is the method of @li2014biclustering.
 2.
-*Truncated Normal prior.* Use two independent truncated spherical normal distributions as priors for $\theta_{main}$ and $\theta_{interaction}$ with $\sigma_{interaction} < \sigma_{main}$.
-Full conditional distributions are not conjugate, requires a geometric adaptive Metropolis Hastings step [@zhou2014some] and calculation of likelihood normalizing constant.
-3. *Marginalized likelihood* Marginalize out $\boldsymbol h$ in $f_{\boldsymbol \theta}(\boldsymbol x)$, and use the truncated Normal prior. 
+*A truncated Normal prior.* Here we use independent truncated spherical normal distributions as priors for $\boldsymbol \theta_{main}$ and $\boldsymbol \theta_{interaction}$ with $\sigma_{interaction} < \sigma_{main}$.
+Full conditional distributions are not conjugate, and simulaion from the posterior was accomplished using a geometric adaptive Metropolis Hastings step [@zhou2014some] and calculation of likelihood normalizing constant.
+(This computation is barely feasible for a problem of this size.) Here the $h_j$ are carried along in the MCMC implementation as latent variables.
+3. *A truncated Normal prior and marginalized likelihood.* Here we marginalize out $\boldsymbol h$ in $f_{\boldsymbol \theta}(\boldsymbol x)$, and use the truncated Normal prior. 
     \begin{align*}
@@
 
-The three fitting methods are ordered by computational feasibility in a real data situation with the trick prior being the most possible due to conjugacy and the marginalized likelihood method the least feasible due to the marginalization and need for an adaptive Metropolis Hastings step.
-All three priors do require choosing the values of hyperparameters.
-In each case, we have chosen the values based on a rule of thumb that maintains the property of shrinking $\boldsymbol \theta_{interaction}$ more than $\boldsymbol \theta_{main}$.
-Additionally, the trick prior requires additional tuning to choose $C_1$ and $C_2$, making it less appealing.
-The values used for the hyperparameters are presented in table \ref{tab:hyperparam}.
+The three fitting methods are ordered by computational feasibility in a real-data situation, with the trick prior being the most feasible due to conjugacy and the marginalized likelihood method the least feasible due to the marginalization and need for an adaptive Metropolis Hastings step.
+All three methods require choosing the values of hyperparameters.
+In each case, we have chosen the values based on a rule of thumb that shrinks $\boldsymbol \theta_{interaction}$ more than $\boldsymbol \theta_{main}$.
+Additionally, the trick prior requires additional tuning to choose $C_1$ and $C_2$, reducing its appeal.
+The values used for the hyperparameters in our simulation are presented in table \ref{tab:hyperparam}.
 
@@
 \caption{The values used for the hyperparameters for all three fitting methods.
-A rule of thumb is imposed which decreases as the size of the model increases and also shrinks $\boldsymbol \theta_{interaction}$ more than $\boldsymbol \theta_{main}$.}
+A rule of thumb is imposed which decreases prior variances for the model parameters as the size of the model increases and also shrinks $\boldsymbol \theta_{interaction}$ more than $\boldsymbol \theta_{main}$.}
 \label{tab:hyperparam}
@@
 
-Figure \ref{fig:fitting_plot} shows the posterior probability of each possible binary image with four pixels after fitting the model in the three ways detailed in this section.
-The black vertical lines show the true probabilities of each image based on the parameters used to simulate the images while the red vertical lines show the empirical distribution.
-From these posterior predictive checks, it is evident that the marginalized likelihood method shows the closest fit to the actual data.
-However, this model requires a marginalization step, which is infeasible for a model of any real size.
+Figure \ref{fig:fitting_plot} shows the posterior probability of each possible $\boldsymbol v \in \{-1,1\}^4$ after fitting the model in the three ways detailed in this section.
+The black vertical lines show the true probabilities of each image based on the parameters used to simulate the training set while the red vertical lines show the empirical distribution.
+From these posterior predictive checks, it is evident that the marginalized likelihood method produces the closest fit to the actual data.
+However, this method requires a marginalization step, which is infeasible for a model with $H$ of any real size.
 
@@
 rm(models_good)
-
-load("data/params_theta.Rdata")
-
-params_good <- list(main_hidden = sample.params %>% ungroup() %>% filter(!near_hull) %>% select(starts_with("h"), -H) %>% data.matrix(),
-                    main_visible = sample.params %>% ungroup() %>% filter(!near_hull) %>% select(starts_with("v"), -V) %>% data.matrix(),
-                    interaction = sample.params %>% ungroup() %>% filter(!near_hull) %>% select(starts_with("theta")) %>% data.matrix() %>% matrix(4))
 
@@
 
-It should also be noted that each of these rigorous fitting methods seeks to merely replicate the empirical data distribution, which is the optimal nonparametric solution.
-In a data application with a large sample space it is unlikely that the training set will include one of each possible image, thus there will be possible images that are not realized in the training set, unlike this small example.
-
-**TODO: Add proof that likelihood goes to infinity.**
-
 # Discussion
@@
 
+**TODO: Add**
+
 ##Incomplete cases
 
+**TODO: Add**
+
 #References

### 2016-09-12 writing/draft.Rmd
fix some latex table crap

@@
   kable(caption = "\\label{tab:theta}Parameters used to fit a test case with $V = H = 4$.
-This parameter vector was chosen as a sampled value of $\\boldsymbol \\theta$ that was not near the convex hull of the sufficient statistics for a grid point in figure \\ref{fig:degen_plots} with $< 5%$ near-degeneracy.")
+This parameter vector was chosen as a sampled value of $\\boldsymbol \\theta$ that was not near the convex hull of the sufficient statistics for a grid point in figure \\ref{fig:degen_plots} with $< 5$\\% near-degeneracy.")
 ```
@@
 We considered a test case with $V = H = 4$ and parameters given in table \ref{tab:theta}.
-This parameter vector was chosen as a sampled value of $\\boldsymbol \\theta$ that was not near the convex hull of the sufficient statistics for a grid point in figure \\ref{fig:degen_plots} with $< 5%$ near-degeneracy.
+This parameter vector was chosen as a sampled value of $\boldsymbol \theta$ that was not near the convex hull of the sufficient statistics for a grid point in figure \ref{fig:degen_plots} with $< 5$\% near-degeneracy.
 We simulated $n = 5,000$ as a training set and fit the RBM using three Bayes methodologies.
@@
 \centering
+\caption{The values used for the hyperparameters for all three fitting methods.
+A rule of thumb is imposed which decreases prior variances for the model parameters as the size of the model increases and also shrinks $\boldsymbol \theta_{interaction}$ more than $\boldsymbol \theta_{main}$.}
+\label{tab:hyperparam}
 \begin{tabular}{|l|c|c|}
@@
 \end{tabular}
-\caption{The values used for the hyperparameters for all three fitting methods.
-A rule of thumb is imposed which decreases prior variances for the model parameters as the size of the model increases and also shrinks $\boldsymbol \theta_{interaction}$ more than $\boldsymbol \theta_{main}$.}
-\label{tab:hyperparam}
 \end{table}

### 2016-09-17 writing/appendix.Rmd
add effective sample size test and nordman's proofs

For integers $m,n \geq 1$, consider random vectors $\bm{V}_m \equiv(V_1,\ldots,V_m)$ and $\bm{H}_n\equiv (H_1,\ldots,H_n)$, where the random variables $V_i$ and $H_j$ assume values in $\{1,-1\}$, $i=1,\ldots,m$, $j=1,\ldots,n$. We suppose the vector $(\bm{V}_m,\bm{H}_n)$ follows a restricted Boltzmann machine (RBM) distribution

\begin{eqnarray*}
&&P(V_1= v_1,\ldots,V_m= v_m, H_1= h_1,\ldots,H_n=h_n|\bm{\theta})   \\&\propto &\exp\left[ \sum_{i=1}^m v_i \theta_{v_j} + \sum_{j=1}^n h_j \theta_{h_j} + \sum_{i=1}^m  \sum_{j=1}^n v_i  h_j \theta_{ij} \right], \quad \{v_i\}_{i=1}^m,\{h_j\}_{j=1}^n \subset\{\pm 1\},
\end{eqnarray*}
with $m+n+mn$ real parameters $\theta_{v_1},\ldots,\theta_{v_m},\theta_{h_1},\ldots,\theta_{h_m}$ and
$\theta_{ij}$, $i=1,\ldots,m$, $j=1,\ldots,n$ collected in a parameter vector denoted as $\bm{\theta}\in \mathbb{R}^{m+n+mn}$.
Denote the distribution of $(\bm{V}_m,\bm{H}_n)$ as RBM$(m,n,\bm{\theta})$. In the following, for convenience, we say a symmetric random variable $X$ has a Bernoulli$(1/2)$ distribution if $P(X=1)=1/2=P(X=-1)$.

Denote the random vector $\bm{V}_m*\bm{H}_n = (V_1 H_1,\ldots,V_1 H_n, V_2 H_1,\ldots,V_2 H_n,\ldots, V_m H_1,\ldots,V_m H_n)$.
For the RBM$(m,n,\bm{\theta})$ model, we are interested in the function $\bm{g}: \mathbb{R}^{m+n+mn} \rightarrow   \mathcal{I} \subset \mathbb{R}^{m+n+mn}$ given by
\begin{eqnarray*}
\bm{g}(\bm{\theta}) &=& \E_{\bm{\theta}}(\bm{V}_m,\bm{H}_n,\bm{V}_m*\bm{H}_n)\\
& = & \sum_{\bm{v}_m \in \{\pm 1\}^m, \bm{h}_n \in \{\pm 1\}^n} (\bm{v}_m,\bm{h}_n,\bm{v}_m*\bm{h}_n)P(\bm{V}_m=\bm{v}_m,\bm{H}_n=\bm{h}_n|\bm{\theta})
\end{eqnarray*}
for $\bm{\theta}\in \mathbb{R}^{m+n+mn}$; above $\mathcal{I} \equiv \{\bm{g}(\bm{\theta}) : \bm{\theta}\in \mathbb{R}^{m+n+mn}\}$ denotes the image of the mapping.

We have the following properties:

1. The function $\bm{g}: \mathbb{R}^{m+n+mn} \rightarrow \mathcal{I}$ is one-to-one; see Theorem 1 below. Hence, these moments uniquely characterize the parameters in the RBM$(m,n,\bm{\theta})$ model. (Additionally, two RBM models have the same parameters if and only if the models have the same probability functions.)
1. The image $\mathcal{I}$ is a symmetric set (i.e., $\bm{g}(\bm{\theta})=-\bm{g}(-\bm{\theta})$) and must be a connected (not necessarily convex) region in $\mathbb{R}^{m+n+mn}$, without any voids in it (by the continuity of $\bm{g}(\bm{\theta})$).
1. The center of the image $\mathcal{I}$ (the zero vector in $\mathbb{R}^{m+n+mn}$) is given only by $\bm{\theta}=\bm{0}$ (the zero vector in the parameter space).  This corresponds to the case where all the random vectors $V_1,\ldots,V_m,H_1,\ldots,H_n$ are iid with a symmetric Bernoulli$(1/2)$ distribution; see Theorem 2.

\newtheorem{theoremnew4}{Theorem}
\begin{theoremnew4} Let $m,n\geq 1$ and suppose the random vector   $(\bm{V}_m,\bm{H}_n) = (V_1,\ldots,V_m, H_1,\ldots,H_n)$ follows a RBM$(m,n,\bm{\theta})$ distribution. Let $\bm{\theta}_1,\bm{\theta}_2 \in \mathbb{R}^{m+n+mn}$ and
let $P_i$ and $\E$ denote probability and expectation under $\bm{\theta}_i$, $i=1,2$. Then, the following are equivalent
\begin{enumerate}
\item $\E_1 V_i=\E_2 V_i$, $\E_1 H_j=\E_2 H_j$ and $\E_1 V_i H_j =\E_2 V_i H_j  $ for any $i=1,\ldots,m$, $j=1,\ldots,n$.
\item $\bm{\theta}_1 =\bm{\theta}_2$.
\item $P_1$ and  $P_2$ are the same distribution for  $(\bm{V}_m,\bm{H}_n)$.\\
\end{enumerate}
\end{theoremnew4}

**Proof of Theorem 1.** We first establish the equivalence of claims 2 and 3. If claim 2 holds, then
 claim 3 holds trivially. Suppose claim 3 holds and pick $\ell \in \{1,\ldots,m\}$. Let  $\theta_{v_1},\ldots,\theta_{v_m},\theta_{h_1},\ldots,\theta_{h_n}$, $\theta_{ij}$, $i=1,\ldots,m$, $j=1,\ldots,n$, denote the components of $\bm{\theta}_1$ and let $\tilde{\theta}_{v_1},\ldots,\tilde{\theta}_{v_m},\tilde{\theta}_{h_1},\ldots,\tilde{\theta}_{h_n}$, $\tilde{\theta}_{ij}$, $i=1,\ldots,m$, $j=1,\ldots,n$, denote the analogous components of $\bm{\theta}_2$. Then, by assumption,
\begin{eqnarray*}
&& \exp\left[ 2\theta_{v_\ell} +2 \sum_{j=1}^n  h_j \theta_{\ell j} \right]\\
&=&  \frac{P_1\left(\{V_\ell = 1\} \cap \bigcap_{i=1 \atop i \neq \ell}^m \{V_i=v_i\} \cap \bigcap_{j=1}^n \{H_j=h_j\}\right)} {P_1\left(\{V_\ell = -1\} \cap \bigcap_{i=1 \atop i \neq \ell}^m \{V_i=v_i\} \cap \bigcap_{j=1}^n \{H_j=h_j\}\right)}\\
&=& \frac{P_2\left(\{V_\ell = 1\} \cap \bigcap_{i=1 \atop i \neq \ell}^m \{V_i=v_i\} \cap \bigcap_{j=1}^n \{H_j=h_j\}\right)}
{P_2\left(\{V_\ell = -1\} \cap \bigcap_{i=1 \atop i \neq \ell}^m \{V_i=v_i\} \cap \bigcap_{j=1}^n \{H_j=h_j\}\right)}\\
&& \exp\left[ 2\tilde{\theta}_{v_\ell} +2 \sum_{j=1}^n  h_j \tilde{\theta}_{\ell j} \right]
\end{eqnarray*}
for any $\{v_i\}_{i=1,i\neq \ell}^m,\{h_j\}_{j=1}^n \subset\{\pm 1\}$. Equating exponents, we have $\theta_{v_\ell} - \tilde{\theta}_{v_\ell} = \sum_{j=1}^n  h_j (  \tilde{\theta}_{\ell j} -\theta{\ell j})$ for any $\{h_j\}_{j=1}^n \subset \{\pm 1\}$. Taking all $h_j=1$ or all $h_j=-1$ for $1 \leq j \leq n$ yields $\theta_{v_\ell} = \tilde{\theta}_{v_\ell}$. This result further implies $\tilde{\theta}_{\ell 1} -\theta{\ell 1} = -  \sum_{j=2}^n  h_j (  \tilde{\theta}_{\ell j} -\theta_{\ell j})$ for
any given $\{h_j\}_{j=2}^n \subset \{\pm 1\}$, so that $\theta{\ell 1}=\tilde{\theta}_{\ell 1}$ must hold. Iterating this argument, sequentially for each $j=2,\ldots,m$, shows $\theta{\ell j}=\tilde{\theta}_{\ell j}$ for any $1 \leq j \leq n$ in addition to $\theta_{v_\ell}=\tilde{\theta}_{v_\ell}$.  As $\ell\in\{1,\ldots,m\}$ was arbitrary, we have $\theta_{v_i}=\tilde{\theta}_{v_i}$ and $\theta{i j}=\tilde{\theta}_{i j}$ for any $1 \leq i \leq m$, $1 \leq j \leq n$.

Now pick $k \in \{1,\ldots,n\}$ so that, by the RBM$(m,n,\bm{\theta})$ probability structure as above, we have
 analogously that
$$
\theta_{h_k} + \sum_{i=1}^m  v_i \theta_{i k} =  \tilde{\theta}_{h_k} + \sum_{i=1}^m  v_i \tilde{\theta}_{i k}
$$
fsor any $\{v_i\}_{i=1}^m \subset \{\pm 1\}$, so that $\theta_{h_k}  =  \tilde{\theta}_{h_k}$. As $k \in \{1,\ldots,n\}$ was arbitrary, we have now established claim 2 from claim 3.

To establish the equivalence of claims 1 and 3 in Theorem 1 under the RBM$(m,n,\bm{\theta})$ model, we note that claim 3 easily implies claim 1. Hence, it suffices now to establish claim 3 from claim 1. This follows immediately from Lemma 1 below and the fact that the variables $V_1,\ldots,V_m$ are independent given $H_1,\ldots,H_n$ in the RBM$(m,n,\bm{\theta})$ model (and likewise $H_1,\ldots,H_n$ are independent given $V_1,\ldots,V_m$).

\begin{lemma} Let $m,n\geq 1$.  Let $P_1,P_2$ denote two probability distributions for a random vector $(\bm{V}_m,\bm{H}_n) = (V_1,\ldots,V_m, H_1,\ldots,H_n)$ supported on $\{a,b\}^{m+n}$, for some real constants $a\neq b \in\mathbb{R}$, such that $P_i( \bm{V}_m=\bm{v}_m, \bm{H}_n=\bm{h}_n)>0$, $i=1,2$, for any $\bm{v}_m \in \{a,b\}^{m}$, $\bm{h}_n \in \{a,b\}^{n}$. Suppose additionally that, under $P_1$ or $P_2$,  the variables $V_1,\ldots,V_m$ are independent given $\bm{H}_n=\bm{h}_n \in \{a,b\}^{n}$ and that the variables $H_1,\ldots,H_n$ are independent given $\bm{V}_m=\bm{v}_m \in \{a,b\}^{m}$.

Let $\E_i$ denote expectation under the distribution $P_i$, $i=1,2$. Then, for any $\bm{v}_m \in \{a,b\}^{m}, \bm{h}_n \in \{a,b\}^{n}$,
$$
P_1( \bm{V}_m=\bm{v}_m, \bm{H}_n=\bm{h}_n) = P_2( \bm{V}_m=\bm{v}_m, \bm{H}_n=\bm{h}_n), \qquad
$$
follows if $\E_1 V_i = \E_2 V_i$, $\E_1 H_j = \E_2 H_j$, $\E_1 V_i H_j = \E_2 V_j H_j$ holds for all $i\in\{1,\ldots,m\}$, $j\in \{1,\ldots,n\}$.
\end{lemma}

*Proof.* The case $m=n=1$ follows from Lemma 2. We now use an induction argument, assuming that the result of Lemma 1 holds for some given order of $(m,n)$ with $m,n\geq 1$ and show the the result continues to hold for $(m+1,n)$ or $(m,n+1)$. We shall treat the case $(m+1,n)$ (where the other case $(m,n+1)$ follows by symmetrical arguments).

Considering random variables $(\bm{V}_{m+1},\bm{H}_n) = (V_1,\ldots,V_m,V_{m+1}, H_1,\ldots,H_n)$, the induction hypothesis
applies to $(\bm{V}_{m},\bm{H}_n)$ where $\bm{V}_m =(V_1,\ldots,V_m)$. Hence, for any $\bm{v}_m \in \{a,b\}^{m}, \bm{h}_n \in \{a,b\}^{n}$, we have $P_1(\bm{V}_m=\bm{v}_m,\bm{H}_n=\bm{h}_n) = P_2(\bm{V}_m=\bm{v}_m,\bm{H}_n=\bm{h}_n)>0$, $P_1( \bm{H}_n=\bm{h}_n) = P_2(\bm{H}_n=\bm{h}_n)>0$ and consequently
\begin{equation}\label{eqn0}
P_1(\bm{V}_m=\bm{v}_m|\bm{H}_n=\bm{h}_n) =P_2(\bm{V}_m=\bm{v}_m |\bm{H}_n=\bm{h}_n).
\end{equation}
Likewise, by the induction hypothesis applied to $(\bm{V}_{m-},\bm{H}_n)$ for $\bm{V}_{m-} = (V_2,\ldots,V_{m+1})$, we have
\begin{eqnarray}\label{eqn1}
 1- P_1(V_{m+1} = b | \bm{H}_n=\bm{h}_n) &=& P_1(V_{m+1} = a |\bm{H}_n=\bm{h}_n)\\
\nonumber  &=&  P_2(V_{m+1} = a | \bm{H}_n=\bm{h}_n)\\
\nonumber  & = &1-P_2(V_{m+1} = b | \bm{H}_n=\bm{h}_n)
\end{eqnarray}
for any $\bm{h}_n \in \{a,b\}^{n}$. Now fix  $\bm{v}_m \in \{a,b\}^{m}, \bm{h}_n \in \{a,b\}^{n}$ and let $\bm{v}_{m+1}= (v_{m+1},\bm{v}_m)$ for $v_{m+1}\in\{a,b\}$.  Then, by the conditional independence assumption,
$$
P_i( \bm{V}_{m+1}=\bm{v}_{m+1}| \bm{H}_n=\bm{h}_n) = P_i( V_{m+1}=v_{m+1}| \bm{H}_n=\bm{h}_n) P_i( \bm{V}_{m}=v_{m}| \bm{H}_n=\bm{h}_n)
$$
for $i=1,2,$,  so that we have   $P_1( \bm{V}_{m+1}=\bm{v}_{m+1}| \bm{H}_n=\bm{h}_n) = P_2( \bm{V}_{m+1}=\bm{v}_{m+1}| \bm{H}_n=\bm{h}_n)$ by (\ref{eqn0})-(\ref{eqn1}). Consequently, it follows that
\begin{eqnarray*}
P_1( \bm{V}_{m+1}=\bm{v}_{m+1}, \bm{H}_n=\bm{h}_n) &=& P_1( \bm{V}_{m+1}=\bm{v}_{m+1}| \bm{H}_n=\bm{h}_n) P_1(\bm{H}_n=\bm{h}_n)\\
& =& P_2( \bm{V}_{m+1}=\bm{v}_{m+1}| \bm{H}_n=\bm{h}_n)P_2(\bm{H}_n=\bm{h}_n) \\&=&P_2( \bm{V}_{m+1}=\bm{v}_{m+1}, \bm{H}_n=\bm{h}_n)
\end{eqnarray*}
by $P_1(\bm{H}_n=\bm{h}_n)=P_2(\bm{H}_n=\bm{h}_n)$. As $\bm{v}_{m+1}\in \{a,b\}^{m+1},\bm{h}_n \in \{a,b\}^{n}$ were arbitrary, this completes the proof of Lemma 1. $\Box$

\begin{lemma} Suppose the discrete random vector $(X,Y)$ has support $\{a,b\}\times \{a,b\}$, for some $a\neq b \in \mathbb{R}$,
under two probability distributions $P_1$ and $P_2$. Let $\E_i$ denote expectation under the distribution $P_i$, $i=1,2$. Then, the following are equivalent
\begin{enumerate}
\item $\E_1 X = \E_2 X$, $\E_1 Y = \E_2 Y$, $ \E_1 X Y = \E_2 X Y$, where $\E_i$ denotes expectation under $P_i$, $i=1,2$.
\item $P_1(X=x,Y=y)=P_2(X=x,Y=y)$ for $x,y\in\{a,b\}$, i.e., $(X,Y)$ has the same distribution under $P_1$ and $P_2$.
\end{enumerate}
\end{lemma}

*Proof.* Denote the four probabilities by which the random vector $(X,Y)$ assume pairs $(b,b),(b,a),(a,b),(a,a)$, respectively, as $c_1,c_2,c_3,1-c_1-c_2-c_3$ under $P_1$ and $d_1,d_2,d_3,1-d_1-d_2-d_3$ under $P_2$.  The conditions $\E_1 X = \E_2 X$, $\E_1 Y = \E_2 Y$, $ \E_1 X Y = \E_2 X Y$ with $b\neq a$ imply that
$$
(c_1+c_2) = (d_1+d_2), \quad  (c_1+c_3) = (d_1+d_3) \quad (b-a) c_1 + a (2c_1+c_2+c_3) =  (b-a) d_1 + a (2d_1+d_2+d_3).
$$
As $(2c_1+c_2+c_3) = (2d_1+d_2+d_3) $ and $b\neq a$, we conclude that $c_1=d_1$, from which it follows that $c_2=d_2$ and $c_3=d_3$. Hence, claim 1 implies claim 2 in Lemma 2.  Claim 2 also trivially implies claim 1. $\Box$

\begin{theoremnew4} Let $m,n\geq 1$ and suppose the random vector   $(\bm{V}_m,\bm{H}_n) = (V_1,\ldots,V_m, H_1,\ldots,H_n)$ follows a RBM$(m,n,\bm{\theta})$ distribution.  Then, the following are equivalent
\begin{enumerate}
\item $\E V_i=0$, $\E H_j=0$ and $\E V_i H_j =0$ for any $i=1,\ldots,m$, $j=1,\ldots,n$.
\item $\bm{\theta} = \bm{0}\in \mathbb{R}^{m+n+mn}$.
\item $V_1,\ldots,V_m, H_1,\ldots,H_n$ are iid Bernoulli$(1/2)$ random variables.\\
\end{enumerate}
\end{theoremnew4}

**Proof of Theorem 2.**  We first establish the equivalence of claims 2 and 3. If claim 2 holds, then claim 3 follows easily from the resulting uniform cell probabilities: $P(V_1= v_1,\ldots,V_m= v_m, H_1= h_1,\ldots,H_n=h_n|\bm{\theta})=2^{-mn}$ for any $v_i,h_j\in\{\pm 1\}$, $i=1,\ldots,m$, $j=1,\ldots,n$.

Now suppose claim 3 holds. Pick $\ell \in \{1,\ldots,m\}$. Then, by the RBM$(m,n,\bm{\theta})$ probability structure and the iid Bernoulli assumption, we have the conditional probability
\begin{eqnarray*}
\frac{1}{2} &=& P\left(V_\ell = v_{\ell}\Big| \bigcap_{i=1 \atop i \neq \ell}^m \{V_i=v_i\} \cap \bigcap_{j=1}^n \{H_j=h_j\}\right) \\
&\equiv &\frac{\exp\left( v_\ell\theta_\ell +  v_\ell\sum_{j=1}^n  h_j \lambda_{\ell,j}\right)}{\exp\left(
\theta_\ell + \sum_{j=1}^n   h_j\lambda_{\ell,j} \right) +\exp\left(-\theta_\ell  -\sum_{j=1}^n  h_j\lambda_{\ell,j}\right) }
\end{eqnarray*}
for any subsets $\{v_i\}_{i=1}^m,\{h_j\}_{j=1}^n \subset\{\pm 1\}$. Equating conditional probabilities when $v_\ell=1$ or when $v_\ell=-1$, we have   $\theta_\ell = -  \sum_{j=1}^m  h_j \lambda_{\ell,j}$ for any given $\{h_j\}_{j=1}^n \subset \{\pm 1\}$. Taking all $h_j=1$ or all $h_j=-1$ for $1 \leq j \leq n$ yields $\theta_\ell=0$.  This result further implies $\lambda_{\ell,1} = - \sum_{j=2}^n  h_j \lambda_{\ell,j}$ for any given $\{h_j\}_{j=2}^n \subset \{\pm 1\}$, so that $\lambda_{\ell,1}=0$ must hold. Iterating this argument, sequentially for each $j=2,\ldots,m$, shows $\lambda_{\ell,j}=0$ for any $1 \leq j \leq n$ in addition to $\theta_\ell=0$. As $\ell\in\{1,\ldots,m\}$ was arbitrary, we have $\theta_i=0$ and $\lambda_{i,j}=0$ for any $1 \leq i \leq m$, $1 \leq j \leq n$.

Now pick $k \in \{1,\ldots,n\}$ so that, by the RBM$(m,n,\bm{\theta})$ probability structure and the iid assumption, we have the conditional probability
$$
 \frac{1}{2} = P\left(H_k = h_k \Big| \bigcap_{j=1 \atop j \neq k}^n \{H_j=h_j\}\right) \equiv \frac{\exp(h_k \theta_{h_k})}{\exp(\theta_{h_k}) + \exp(-\theta_{h_k})}
$$
for any $\{h_j\}_{j=1}^n \subset \{\pm 1\}$.  Considering $h_k=1$ or $h_k=-1$, we conclude that $\theta_{h_k} = -\theta_{h_k}$
or $\theta_{h_k}=0$.  As $k \in \{1,\ldots,n\}$ was arbitrary, we have now established claim 2 from claim 3.

To establish the equivalence of claims 1 and 2 in Theorem 2 under the RBM$(m,n,\bm{\theta})$ model, we note that claim 2 again implies claim 3, which then easily implies claim 1. By Theorem 1, if claim 1 holds then the only possibility for this is $\bm{\theta} = \bm{0} \in \mathbb{R}^{m+n+mn}$. $\Box$

### 2016-09-17 writing/draft.Rmd
add effective sample size test and nordman's proofs

@@
 We discuss three Bayesian fitting techniques with the goal of avoiding model impropriety in section \ref{bayesian-model-fitting} and wrap up with a discussion in section \ref{discussion}.
-Additionally, there are two accompanying appendices with auxiliary proofs relating to the choice of data encoding and parameter behavior during fitting with incomplete training data in section \ref{appendix}.
+Additionally, there is an accompanying appendix with auxiliary proofs relating to the choice of data encoding in section \ref{appendix}.
 
@@
 In addition to the argument that the $-1/1$ data encoding mitigates some perceived prevalence of near-degeneracy it also has the benefit of providing a guaranteed non-degenerate model at $\boldsymbol \theta = \boldsymbol 0 \in \mathbb{R}^{H + V + H*V}$.
-The proof of this is included in appendix \ref{center-of-the-universe}.
+The proof of this is included in appendix \ref{on-the-one-to-one-correspondence-between parameters-and moments-in-the-rbm-distribution}.
 And while from some computing perspectives $0/1$ coding might seem most natural, the $-1/1$ coding is far more convenient and interpretable from the point of view of statistical modeling, where it makes parameters simply interpreted in terms of symmetrically defined main effects and interactions.
@@
 In addition, the RBM model, with the appropriate choice of parameters, has the potential to re-create any distribution for the data (i.e., reproduce any specification of cell probabilities for the -1/1 valued data outcomes).
-This hyper-flexibility dictates that ML fitting will merely replicate the empirical data distribution, which is the optimal nonparametric solution.
+To see this, we provide a small example.
+
+For simplicity, consider a model with two visible variables $(V_1,V_2)$ and one hidden $H_1$.  In this case,
+there are four possible data realizations for $(V_1,V_2)$  given by $(\pm 1, \pm 1)$ and we may express the model probabilities
+as
+$$
+P(V_1=v_1,V_2=v_2| \theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21}) \propto  \exp\left( v_1  \theta_{v_1} + v_2 \theta_{v_2}\right) \sum_{h \in \{\pm 1\}}\exp\left( h[ \theta_{h_1}  + \theta_{11} v_1 + \theta_{21} v_2] \right),
+$$
+for $(v_1,v_2)\in \{-1, 1\}^2$, in terms of real-valued parameters $\theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21}$. Given any  specified cell probabilities, say
+\begin{equation}
+\label{eqn:0}
+0\leq p_{(-1,-1)},\;p_{(1,-1)}, \;p_{(-1,1)},\;p_{(1,1)},
+\end{equation} 
+for the outcomes $(\pm 1, \pm 1)$,  values for parameters $(\theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21})$ may be chosen to mimic such  cell probabilities with arbitrary closeness.
+In fact, when the cell probabilities (\ref{eqn:0}) are all strictly positive, parameters in the RBM model can be specified to reflect these probabilities exactly.
+And, when one or more of the cell probabilities (\ref{eqn:0}) are zero, the corresponding RBM  probabilities $P(V_1=v_1,V_2=_v2| \theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21})$ may never be identically zero (due to the exponential model form) but
+ parameters can be still selected to make the appropriate RBM cell probabilities arbitrarily small.
+
+To demonstrate such model approximations,  we  assume $p_{(-1,-1)}>0$ (without loss of generality) in the specified cell probabilities (\ref{eqn:0}) and replace parameters $\theta_{11},\theta_{21}$ with $\Delta_1 \equiv \theta_{11} +\theta_{21}$ and $\Delta_2 \equiv \theta_{11} -\theta_{21}$.
+We may then prescribe values of $\theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \Delta_1,\Delta_2$  so that the model probability ratio
+$$
+P(V_1=v_1,V_2=v_2|\theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \Delta_1,\Delta_2)/  P(V_1=-1,V_2=-1| \theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \Delta_1,\Delta_2)
+$$
+matches the corresponding  ratio $p_{(v_1,v_2)}/p_{(-1,-1)}$  over three values of $(v_1,v_2) = (1,-1),(-1,1),(1,1)$.
+
+For instance, assuming the cell probabilities from (\ref{eqn:0}) are all positive, these probabilities
+ can be exactly reproduced by choosing
+ \begin{eqnarray*}
+ \theta_{v_1} &=& \frac{1}{2} \log\left( \frac{p_{(1,-1)}}{p_{(-1,-1)}} \frac{\exp(\theta_{h_1}-\Delta_1) + \exp(-\theta_{h_1}+\Delta_1)}{\exp(\theta_{h_1}+\Delta_2) + \exp(-\theta_{h_1}-\Delta_2)}  \right),\\
+ \theta_{v_2}&= &\frac{1}{2} \log\left( \frac{p_{(-1,1)}}{p_{(-1,-1)}} \frac{\exp(\theta_{h_1}-\Delta_1) + \exp(-\theta_{h_1}+\Delta_1)}{\exp(\theta_{h_1}-\Delta_2) + \exp(-\theta_{h_1}+\Delta_2)}  \right)
+\end{eqnarray*}
+and selecting $\theta_{h_1}, \Delta_1,\Delta_2$ to solve
+\begin{equation}
+\label{eqn:1}
+\frac{p_{(1,1)}p_{(-1,-1)}}{p_{(-1,1)}p_{(1,-1)}} = \frac{\ell(|\theta_{h_1}|) + \ell(|\Delta_1|) }{\ell(|\theta_{h_1}|) + \ell(|\Delta_2|)},
+\end{equation}
+based on a monotonically increasing function $\ell(x)\equiv \exp(-2x)+\exp(2x)$, $x \geq 0$.
+If  $[p_{(1,1)}p_{(-1,-1)}]/ [p_{(-1,1)}p_{(1,-1)}] \geq 1$, one can pick any values for $\theta_{h_1}, \Delta_2\in \mathbb{R}$ and solve (\ref{eqn:1}) for $|\Delta_1|$; likewise, when $[p_{(1,1)}p_{(-1,-1)}]/ [p_{(-1,1)}p_{(1,-1)}] < 1$ in (\ref{eqn:1}), one may solve for $|\Delta_2|$ upon choosing any values for $\theta_{h_1}, \Delta_1\in \mathbb{R}$.
+
+Alternatively, if exactly one specified cell probability in (\ref{eqn:0}) is zero, say $p_{(1,1)}$ (without loss of generality), we can select parameters $\theta_{v_1},\theta_{v_2}$ as above based on a sequence $(\theta_{h_1}, \Delta_{1}, \Delta_{2}) \equiv (\theta_{h_1}^{(m)}, \Delta_{1}^{(m)}, \Delta_{2}^{(m)})$, $m\in\{1,2,\ldots,\}$ of the remaining parameter values such that $\lim_{m\to \infty}|\Delta_{1}^{(m)}| = \infty$ and $\lim_{m\to \infty}  (|\theta_{h_1}^{(m)}| + |\Delta_{2}^{(m)}|)/|\Delta_{1}^{(m)}|=0$ hold.
+This guarantees that the resulting RBM model matches the given cell probabilities (\ref{eqn:0}) in the limit:
+\begin{equation}
+\label{eqn:2}
+\lim_{m\to \infty}P(V_1=v_1,V_2=v_2| \theta_{v_1}, \theta_{v_2}, \theta_{h_1},\Delta_1,\Delta_2) = p_{(v_1,v_2)},\quad (v_1,v_2)\in\{(\pm 1,\pm 1)\}.
+\end{equation}  
+If exactly two specified probabilities in (\ref{eqn:0}) are zero, say  $p_{(1,1)}$ and $p_{(-1,1)}$ (without loss of generality), then a limit approximation as in (\ref{eqn:2}) follows by picking $\theta_{v_1}$ as above based on any choices of $(\theta_{h_1}, \Delta_1,\Delta_2)$ and choosing  a sequence of $\theta_{v_2}\equiv \theta_{v_2}^{(m)}$ values for which $\theta_{v_2}^{({m})} \rightarrow -\infty$.
+
+The previous discussion intends to indicate that the RBM model suffers from parameter identifiability issues that go beyond mere symmetries in the parametrization.
+Quite different parameter settings can induce the same essential RBM model.However, this  is not the most disastrous implication of the RBM parameterization.
+A far worse consequence is that, when fitting the RBM model by likelihood-based methods, we already know what the answer must be before we begin: namely, such  fitting will simply aim to reproduce the empirical distribution from the data.
+That is, based on a random sample of visible variables, the model for the cell probabilities that has the highest likelihood over  *all possible model classes* (i.e., RBM-based or not) is the empirical distribution, and the parametrization of the RBM model itself ensures that this empirical distribution can be arbitrarily well approximated.
+
+For illustration, continue the simple example from above with $n$ iid observations, each consisting of two realized visibles $(V_1,V_2)$.
+In which case, when the specified cell probabilities $p_{(-1,-1)},p_{(1,-1)}, p_{(-1,1)},p_{(1,1)}$ in (\ref{eqn:0}) are taken as the empirical cell frequencies from the sample, there is no better model based on maximum likelihood and the discussion above (c.f.
+(\ref{eqn:2})) shows that RBM model parameters can be chosen to re-create this empirical distribution to an arbitrarily close degree.
+Hence, RBM model fitting based on ML will often simply seek to reproduce the empirical distribution (i.e., as an extreme type of over-fitting).
+What's more, whenever this empirical distribution contains empty cells, fitting steps for the RBM model will necessarily aim to chase parameters that necessarily diverge in magnitude in order to zero out the corresponding RBM cell probabilities.
 In a data application with a large sample space it is unlikely that the training set will include one of each possible image, thus there will be possible images that are not realized in the training set, unlike this small example.
-The presence of empty cells in the empirical distribution drives coordinates of $\boldsymbol \theta$ to $\infty$ in order to make the corresponding cell probabilities arbitrarily close to zero.
 As we have already discussed in secion \ref{explorations}, large magnitudes of $\boldsymbol \theta$ lead to model impropriety in the RBM.
-A detailed discussion of this phenomenon are included in appendix \ref{incomplete-cases}.
 
@@
 
-#Appendix
+# Appendix
 
-##Center of the universe
+## On the one-to-one correspondence between parameters and moments in the RBM distribution
 
-**TODO: Add**
-
-##Incomplete cases
-
-**TODO: Add**
+```{r child='appendix.Rmd'}
+```
 

### 2016-09-17 4x4 model/test_mixing.R

@@ -42,7 +42,7 @@ marginal_sample <- reshape_sample_distn(marginal)
 trunc_sample <- reshape_sample_distn(trunc)
 
 
-
+## acf -------------------------------
 marginal_sample %>% 
   group_by(v1, v2, v3, v4, image_id) %>% 
   do(data.frame(acf = acf(.$prob, plot=FALSE)$acf,
@@ -63,9 +63,105 @@ marginal_acfs %>% ungroup() %>% select(-image_id) %>% rename(marginal = acf) %>%
   xlab("Lag") +
   ylab("ACF")
 
-
-
-
+## block bootstrap effective sample size ----------------
+###
+### MBB Statistic
+###
+mbb_mean <- function(data_star, data, b) {
+  m <- length(data_star)
+  n <- length(data)
+  N <- n - b + 1
   
+  blockmeans <- rep(0, N)
+  for(i in 1:N) {
+    blockmeans[i] <- mean(data[i:(b + i - 1)])
+  }
+  
+  sqrt(m)*(mean(data_star) - mean(blockmeans))
+}
 
+###
+### MBB Sample
+###
+mbb_sample <- function(data, b) {
+  n <- length(data)
+  k <- floor(n/b)
+  
+  start <- sample.int(n - b + 1, k, replace = TRUE)
+  data_star <- NULL
+  for(i in 1:k) {
+    data_star <- c(data_star, data[start[i]:(start[i] + b - 1)])
+  }
+  
+  return(data_star)
+}
+
+###
+### Block Bootstrap function
+###
+bb <- function(x, b, B, boot_stat, boot_sample) {
+  #ii. For blocksize b, generate B MBB versions of statistic
+  bb.stat <- array(NA, B)
+  for(i in seq_len(B)) {
+    bb.stat[i] <- boot_stat(boot_sample(x, b), x, b)
+  }
+  return(bb.stat)
+}
+
+marginal_sample %>% 
+  group_by(v1, v2, v3, v4, image_id) %>%
+  do(mbb = bb(.$prob, 2000^(1/3), 2000, mbb_mean, mbb_sample)) -> marginal_mbb
+
+trunc_sample %>% 
+  group_by(v1, v2, v3, v4, image_id) %>%
+  do(mbb = bb(.$prob, 5000^(1/3), 2000, mbb_mean, mbb_sample)) -> trunc_mbb
+
+epsilon <- 1/1000
+
+marginal_mbb %>%
+  group_by(v1, v2, v3, v4, image_id) %>%
[…]

### 2016-09-17 writing/appendix.Rmd
formatting

@@
 We suppose the vector $(\bm{V}_m,\bm{H}_n)$ follows a restricted Boltzmann machine (RBM) distribution
-
-\begin{eqnarray*}
-&&P(V_1= v_1,\ldots,V_m= v_m, H_1= h_1,\ldots,H_n=h_n|\bm{\theta})   \\&\propto &\exp\left[ \sum_{i=1}^m v_i \theta_{v_j} + \sum_{j=1}^n h_j \theta_{h_j} + \sum_{i=1}^m  \sum_{j=1}^n v_i  h_j \theta_{ij} \right], \quad \{v_i\}_{i=1}^m,\{h_j\}_{j=1}^n \subset\{\pm 1\},
-\end{eqnarray*}
+\begin{align*}
+&P(V_1= v_1,\ldots,V_m= v_m, H_1= h_1,\ldots,H_n=h_n|\bm{\theta}) \propto \exp\left[ \sum_{i=1}^m v_i \theta_{v_j} + \sum_{j=1}^n h_j \theta_{h_j} + \sum_{i=1}^m  \sum_{j=1}^n v_i  h_j \theta_{ij} \right], \\
+& \quad \{v_i\}_{i=1}^m,\{h_j\}_{j=1}^n \subset\{\pm 1\},
+\end{align*}
 with $m+n+mn$ real parameters $\theta_{v_1},\ldots,\theta_{v_m},\theta_{h_1},\ldots,\theta_{h_m}$ and
@@
 For the RBM$(m,n,\bm{\theta})$ model, we are interested in the function $\bm{g}: \mathbb{R}^{m+n+mn} \rightarrow   \mathcal{I} \subset \mathbb{R}^{m+n+mn}$ given by
-\begin{eqnarray*}
-\bm{g}(\bm{\theta}) &=& \E_{\bm{\theta}}(\bm{V}_m,\bm{H}_n,\bm{V}_m*\bm{H}_n)\\
-& = & \sum_{\bm{v}_m \in \{\pm 1\}^m, \bm{h}_n \in \{\pm 1\}^n} (\bm{v}_m,\bm{h}_n,\bm{v}_m*\bm{h}_n)P(\bm{V}_m=\bm{v}_m,\bm{H}_n=\bm{h}_n|\bm{\theta})
-\end{eqnarray*}
+\begin{align*}
+\bm{g}(\bm{\theta}) &= \E_{\bm{\theta}}(\bm{V}_m,\bm{H}_n,\bm{V}_m*\bm{H}_n)\\
+&= \sum_{\bm{v}_m \in \{\pm 1\}^m, \bm{h}_n \in \{\pm 1\}^n} (\bm{v}_m,\bm{h}_n,\bm{v}_m*\bm{h}_n)P(\bm{V}_m=\bm{v}_m,\bm{H}_n=\bm{h}_n|\bm{\theta})
+\end{align*}
 for $\bm{\theta}\in \mathbb{R}^{m+n+mn}$; above $\mathcal{I} \equiv \{\bm{g}(\bm{\theta}) : \bm{\theta}\in \mathbb{R}^{m+n+mn}\}$ denotes the image of the mapping.
@@
 Then, by assumption,
-\begin{eqnarray*}
-&& \exp\left[ 2\theta_{v_\ell} +2 \sum_{j=1}^n  h_j \theta_{\ell j} \right]\\
-&=&  \frac{P_1\left(\{V_\ell = 1\} \cap \bigcap_{i=1 \atop i \neq \ell}^m \{V_i=v_i\} \cap \bigcap_{j=1}^n \{H_j=h_j\}\right)} {P_1\left(\{V_\ell = -1\} \cap \bigcap_{i=1 \atop i \neq \ell}^m \{V_i=v_i\} \cap \bigcap_{j=1}^n \{H_j=h_j\}\right)}\\
-&=& \frac{P_2\left(\{V_\ell = 1\} \cap \bigcap_{i=1 \atop i \neq \ell}^m \{V_i=v_i\} \cap \bigcap_{j=1}^n \{H_j=h_j\}\right)}
-{P_2\left(\{V_\ell = -1\} \cap \bigcap_{i=1 \atop i \neq \ell}^m \{V_i=v_i\} \cap \bigcap_{j=1}^n \{H_j=h_j\}\right)}\\
-&& \exp\left[ 2\tilde{\theta}_{v_\ell} +2 \sum_{j=1}^n  h_j \tilde{\theta}_{\ell j} \right]
-\end{eqnarray*}
+\begin{align*}
+\exp\left[ 2\theta_{v_\ell} +2 \sum_{j=1}^n  h_j \theta_{\ell j} \right] &= \frac{P_1\left(\{V_\ell = 1\} \cap \bigcap_{i=1 \atop i \neq \ell}^m \{V_i=v_i\} \cap \bigcap_{j=1}^n \{H_j=h_j\}\right)} {P_1\left(\{V_\ell = -1\} \cap \bigcap_{i=1 \atop i \neq \ell}^m \{V_i=v_i\} \cap \bigcap_{j=1}^n \{H_j=h_j\}\right)}\\
+&= \frac{P_2\left(\{V_\ell = 1\} \cap \bigcap_{i=1 \atop i \neq \ell}^m \{V_i=v_i\} \cap \bigcap_{j=1}^n \{H_j=h_j\}\right)}
+{P_2\left(\{V_\ell = -1\} \cap \bigcap_{i=1 \atop i \neq \ell}^m \{V_i=v_i\} \cap \bigcap_{j=1}^n \{H_j=h_j\}\right)}\exp\left[ 2\tilde{\theta}_{v_\ell} +2 \sum_{j=1}^n  h_j \tilde{\theta}_{\ell j} \right]
+\end{align*}
 for any $\{v_i\}_{i=1,i\neq \ell}^m,\{h_j\}_{j=1}^n \subset\{\pm 1\}$.
@@
 
-Now pick $k \in \{1,\ldots,n\}$ so that, by the RBM$(m,n,\bm{\theta})$ probability structure as above, we have
- analogously that
+Now pick $k \in \{1,\ldots,n\}$ so that, by the RBM$(m,n,\bm{\theta})$ probability structure as above, we have analogously that
 $$
@@
 Hence, for any $\bm{v}_m \in \{a,b\}^{m}, \bm{h}_n \in \{a,b\}^{n}$, we have $P_1(\bm{V}_m=\bm{v}_m,\bm{H}_n=\bm{h}_n) = P_2(\bm{V}_m=\bm{v}_m,\bm{H}_n=\bm{h}_n)>0$, $P_1( \bm{H}_n=\bm{h}_n) = P_2(\bm{H}_n=\bm{h}_n)>0$ and consequently
-\begin{equation}\label{eqn0}
+\begin{align}\label{eqn0}
 P_1(\bm{V}_m=\bm{v}_m|\bm{H}_n=\bm{h}_n) =P_2(\bm{V}_m=\bm{v}_m |\bm{H}_n=\bm{h}_n).
-\end{equation}
+\end{align}
 Likewise, by the induction hypothesis applied to $(\bm{V}_{m-},\bm{H}_n)$ for $\bm{V}_{m-} = (V_2,\ldots,V_{m+1})$, we have
@@
 $$
-for $i=1,2,$,  so that we have   $P_1( \bm{V}_{m+1}=\bm{v}_{m+1}| \bm{H}_n=\bm{h}_n) = P_2( \bm{V}_{m+1}=\bm{v}_{m+1}| \bm{H}_n=\bm{h}_n)$ by (\ref{eqn0})-(\ref{eqn1}). Consequently, it follows that
+for $i=1,2$,  so that we have $P_1( \bm{V}_{m+1}=\bm{v}_{m+1}| \bm{H}_n=\bm{h}_n) = P_2( \bm{V}_{m+1}=\bm{v}_{m+1}| \bm{H}_n=\bm{h}_n)$ by (\ref{eqn0})-(\ref{eqn1}). Consequently, it follows that
 \begin{eqnarray*}
@@
 \begin{enumerate}
-\item $\E_1 X = \E_2 X$, $\E_1 Y = \E_2 Y$, $ \E_1 X Y = \E_2 X Y$, where $\E_i$ denotes expectation under $P_i$, $i=1,2$.
+\item $\E_1 X = \E_2 X$, $\E_1 Y = \E_2 Y$, $\E_1 X Y = \E_2 X Y$, where $\E_i$ denotes expectation under $P_i$, $i=1,2$.
 \item $P_1(X=x,Y=y)=P_2(X=x,Y=y)$ for $x,y\in\{a,b\}$, i.e., $(X,Y)$ has the same distribution under $P_1$ and $P_2$.
@@
 *Proof.* Denote the four probabilities by which the random vector $(X,Y)$ assume pairs $(b,b),(b,a),(a,b),(a,a)$, respectively, as $c_1,c_2,c_3,1-c_1-c_2-c_3$ under $P_1$ and $d_1,d_2,d_3,1-d_1-d_2-d_3$ under $P_2$.
-The conditions $\E_1 X = \E_2 X$, $\E_1 Y = \E_2 Y$, $ \E_1 X Y = \E_2 X Y$ with $b\neq a$ imply that
+The conditions $\E_1 X = \E_2 X$, $\E_1 Y = \E_2 Y$, $\E_1 X Y = \E_2 X Y$ with $b\neq a$ imply that
 $$
@@
 $$
-As $(2c_1+c_2+c_3) = (2d_1+d_2+d_3) $ and $b\neq a$, we conclude that $c_1=d_1$, from which it follows that $c_2=d_2$ and $c_3=d_3$.
+As $(2c_1+c_2+c_3) = (2d_1+d_2+d_3)$ and $b\neq a$, we conclude that $c_1=d_1$, from which it follows that $c_2=d_2$ and $c_3=d_3$.
 Hence, claim 1 implies claim 2 in Lemma 2.
@@
 \frac{1}{2} &=& P\left(V_\ell = v_{\ell}\Big| \bigcap_{i=1 \atop i \neq \ell}^m \{V_i=v_i\} \cap \bigcap_{j=1}^n \{H_j=h_j\}\right) \\
-&\equiv &\frac{\exp\left( v_\ell\theta_\ell +  v_\ell\sum_{j=1}^n  h_j \lambda_{\ell,j}\right)}{\exp\left(
-\theta_\ell + \sum_{j=1}^n   h_j\lambda_{\ell,j} \right) +\exp\left(-\theta_\ell  -\sum_{j=1}^n  h_j\lambda_{\ell,j}\right) }
+&\equiv &\frac{\exp\left( v_\ell\theta_\ell +  v_\ell\sum_{j=1}^n  h_j \theta_{\ell j}\right)}{\exp\left(
+\theta_\ell + \sum_{j=1}^n   h_j\theta_{\ell j} \right) +\exp\left(-\theta_\ell  -\sum_{j=1}^n  h_j\theta_{\ell j}\right) }
 \end{eqnarray*}
 for any subsets $\{v_i\}_{i=1}^m,\{h_j\}_{j=1}^n \subset\{\pm 1\}$.
-Equating conditional probabilities when $v_\ell=1$ or when $v_\ell=-1$, we have   $\theta_\ell = -  \sum_{j=1}^m  h_j \lambda_{\ell,j}$ for any given $\{h_j\}_{j=1}^n \subset \{\pm 1\}$.
+Equating conditional probabilities when $v_\ell=1$ or when $v_\ell=-1$, we have   $\theta_\ell = -  \sum_{j=1}^m  h_j \theta_{\ell j}$ for any given $\{h_j\}_{j=1}^n \subset \{\pm 1\}$.
 Taking all $h_j=1$ or all $h_j=-1$ for $1 \leq j \leq n$ yields $\theta_\ell=0$.
-This result further implies $\lambda_{\ell,1} = - \sum_{j=2}^n  h_j \lambda_{\ell,j}$ for any given $\{h_j\}_{j=2}^n \subset \{\pm 1\}$, so that $\lambda_{\ell,1}=0$ must hold.
-Iterating this argument, sequentially for each $j=2,\ldots,m$, shows $\lambda_{\ell,j}=0$ for any $1 \leq j \leq n$ in addition to $\theta_\ell=0$.
-As $\ell\in\{1,\ldots,m\}$ was arbitrary, we have $\theta_i=0$ and $\lambda_{i,j}=0$ for any $1 \leq i \leq m$, $1 \leq j \leq n$.
+This result further implies $\theta_{\ell 1} = - \sum_{j=2}^n  h_j \theta_{\ell j}$ for any given $\{h_j\}_{j=2}^n \subset \{\pm 1\}$, so that $\theta_{\ell 1}=0$ must hold.
+Iterating this argument, sequentially for each $j=2,\ldots,m$, shows $\theta_{\ell j}=0$ for any $1 \leq j \leq n$ in addition to $\theta_{v_\ell}=0$.
+As $\ell\in\{1,\ldots,m\}$ was arbitrary, we have $\theta_{v_i}=0$ and $\theta_{i j}=0$ for any $1 \leq i \leq m$, $1 \leq j \leq n$.
 

### 2016-09-17 writing/draft.Rmd
formatting

@@
 
-Figure \ref{fig:fitting_plot} shows the posterior probability of each possible $\boldsymbol v \in \{-1,1\}^4$ after fitting the model in the three ways detailed in this section.
-The black vertical lines show the true probabilities of each image based on the parameters used to simulate the training set while the red vertical lines show the empirical distribution.
-From these posterior predictive checks, it is evident that the marginalized likelihood method produces the closest fit to the actual data.
-However, this method requires a marginalization step, which is infeasible for a model with $H$ of any real size.
-
-```{r fitting}
+It should be noted that the truncated normal method (2) and the marginalized likelihood method (3) are drawing from the same stationary posterior distribution for images.
+The difference between these two methods is in how well the chains mix and how quickly they arrive at the target distribution.
+After a burn-in period of 50 iterations selected by inspecting the trace plots, we assess the issue of mixing in two ways.
+First, the autocorrelation functions (ACF) for each image are assessed at multiple lags for each method and plotted in figure \ref{fig:acf} with the truncated normal method in black and the marginalized likelihood method in red.
+As expected, ACF corresponding to the method that marginalizes out the hidden variables from the likelihood decreases to zero at a much faster rate, indicating better mixing for the chain.
+Secondly,
+
+```{r models}
 source("helpers/functions.R")
@@
 #marginalized likelihood
-load("data/fitted_models_trunc_marginal_thin.Rdata")
+load("data/fitted_models_trunc_marginal_full.Rdata")
 marginal_bad <- models_bad
@@
 #truncated normal
-load("data/fitted_models_trunc_thin.Rdata")
+load("data/fitted_models_trunc_full.Rdata")
 trunc_bad <- models_bad
@@
 trunc_sample_good <- reshape_sample_distn(trunc_good)
-
+```
+
+```{r acf, fig.cap="\\label{fig:acf}The autocorrelation functions (ACF) for each image assessed at multiple lags for each method with the truncated normal method in black and the marginalized likelihood method in red.
+As expected, ACF corresponding to the method that marginalizes out the hidden variables from the likelihood decreases to zero at a much faster rate, indicating better mixing for the chain."}
+marginal_sample_good %>% 
+  group_by(v1, v2, v3, v4, image_id) %>% 
+  do(data.frame(acf = acf(.$prob, plot=FALSE)$acf,
+     lag = acf(.$prob, plot=FALSE)$lag)) -> marginal_acfs
+
+trunc_sample_good %>% 
+  group_by(v1, v2, v3, v4, image_id) %>% 
+  do(data.frame(acf = acf(.$prob, plot=FALSE)$acf,
+                lag = acf(.$prob, plot=FALSE)$lag)) -> trunc_acfs
+
+marginal_acfs %>% ungroup() %>% select(-image_id) %>% rename(marginal = acf) %>%
+  left_join(trunc_acfs %>% rename(truncated = acf)) %>%
+  ggplot() +
+  geom_segment(aes(x=lag, xend=lag, y=0, yend=truncated), colour = "black") +
+  geom_segment(aes(x=lag, xend=lag, y=0, yend=marginal), colour = "red") +
+  facet_wrap(~image_id) +
+  xlab("Lag") +
+  ylab("ACF")
+
+```
+
+Figure \ref{fig:fitting_plot} shows the posterior probability of each possible $\boldsymbol v \in \{-1,1\}^4$ after fitting the model in the three ways detailed in this section.
+The black vertical lines show the true probabilities of each image based on the parameters used to simulate the training set while the red vertical lines show the empirical distribution.
+From these posterior predictive checks, it is evident that the marginalized likelihood method produces the closest fit to the actual data.
+However, this method requires a marginalization step, which is infeasible for a model with $H$ of any real size.
+
+```{r posterior-dsn}
 marginal_sample_good %>% rename(marginal = prob) %>%

### 2016-09-18 writing/draft.Rmd
added mixing issues

@@
 
-It should be noted that the truncated normal method (2) and the marginalized likelihood method (3) are drawing from the same stationary posterior distribution for images.
+It should be noted that the truncated Normal method (2) and the marginalized likelihood method (3) are drawing from the same stationary posterior distribution for images.
 The difference between these two methods is in how well the chains mix and how quickly they arrive at the target distribution.
 After a burn-in period of 50 iterations selected by inspecting the trace plots, we assess the issue of mixing in two ways.
-First, the autocorrelation functions (ACF) for each image are assessed at multiple lags for each method and plotted in figure \ref{fig:acf} with the truncated normal method in black and the marginalized likelihood method in red.
+First, the autocorrelation functions (ACF) for each image are assessed at multiple lags for each method and plotted in figure \ref{fig:acf} with the truncated Normal method in black and the marginalized likelihood method in red.
 As expected, ACF corresponding to the method that marginalizes out the hidden variables from the likelihood decreases to zero at a much faster rate, indicating better mixing for the chain.
-Secondly,
-
-```{r models}
+
+```{r models-load}
 source("helpers/functions.R")
@@
 
-```{r acf, fig.cap="\\label{fig:acf}The autocorrelation functions (ACF) for each image assessed at multiple lags for each method with the truncated normal method in black and the marginalized likelihood method in red.
+```{r acf, fig.cap="\\label{fig:acf}The autocorrelation functions (ACF) for each image assessed at multiple lags for each method with the truncated Normal method in black and the marginalized likelihood method in red.
 As expected, ACF corresponding to the method that marginalizes out the hidden variables from the likelihood decreases to zero at a much faster rate, indicating better mixing for the chain."}
 marginal_sample_good %>% 
+  filter(iter > 50) %>%
   group_by(v1, v2, v3, v4, image_id) %>% 
@@
 trunc_sample_good %>% 
+  filter(iter > 50) %>%
   group_by(v1, v2, v3, v4, image_id) %>% 
@@
 
-Figure \ref{fig:fitting_plot} shows the posterior probability of each possible $\boldsymbol v \in \{-1,1\}^4$ after fitting the model in the three ways detailed in this section.
+```{r M_eff, cache=TRUE}
+###
+### MBB Statistic
+###
+mbb_mean <- function(data_star, data, b) {
+  m <- length(data_star)
+  n <- length(data)
+  N <- n - b + 1
+
+  blockmeans <- rep(0, N)
+  for(i in 1:N) {
+    blockmeans[i] <- mean(data[i:(b + i - 1)])
+  }
+
+  sqrt(m)*(mean(data_star) - mean(blockmeans))
+}
+
+###
+### MBB Sample
+###
+mbb_sample <- function(data, b) {
+  n <- length(data)
+  k <- floor(n/b)
+
+  start <- sample.int(n - b + 1, k, replace = TRUE)
+  data_star <- NULL
+  for(i in 1:k) {
+    data_star <- c(data_star, data[start[i]:(start[i] + b - 1)])
+  }
+
+  return(data_star)
+}
+
+###
+### Block Bootstrap function
+###
+bb <- function(x, b, B, boot_stat, boot_sample) {
+  #ii. For blocksize b, generate B MBB versions of statistic
+  bb.stat <- array(NA, B)
+  for(i in seq_len(B)) {
+    bb.stat[i] <- boot_stat(boot_sample(x, b), x, b)
+  }
+  return(bb.stat)
+}
+
+marginal_sample_good %>% 
+  group_by(v1, v2, v3, v4, image_id) %>%
+  do(mbb = bb(.$prob, 2000^(1/3), 2000, mbb_mean, mbb_sample)) -> marginal_mbb
+
+trunc_sample_good %>% 
+  group_by(v1, v2, v3, v4, image_id) %>%
+  do(mbb = bb(.$prob, 5000^(1/3), 2000, mbb_mean, mbb_sample)) -> trunc_mbb
+
+marginal_mbb %>%
+  group_by(v1, v2, v3, v4, image_id) %>%
+  do(data.frame(C = var(.$mbb[[1]]))) %>%
+  left_join(marginal_sample_good %>% group_by(v1, v2, v3, v4) %>% summarise(sigma2 = var(prob))) %>%
+  mutate(M_eff = sigma2/C) %>%
+  ungroup() %>%
+  summarise(M_eff = min(M_eff)) %>% as.numeric() -> marginal_M_eff
+
+trunc_mbb %>%
+  group_by(v1, v2, v3, v4, image_id) %>%
+  do(data.frame(C = var(.$mbb[[1]]))) %>%
+  left_join(trunc_sample_good %>% group_by(v1, v2, v3, v4) %>% summarise(sigma2 = var(prob))) %>%
+  mutate(M_eff = sigma2/C) %>%
+  ungroup() %>%
+  summarise(M_eff = min(M_eff)) %>% as.numeric() -> trunc_M_eff
+```
+
+Secondly, we can assess the mixing of our chain using an idea of effective sample size.
+If the MCMC chain were truly iid draws from the target distribution, then the estimate of the posterior mean probability of image $i$, $\bar{p}^{(i)}$ would be distributed approximately Normal with mean the true posterior probability of image $i$, $p^{(i)}$, and variance $\sigma^2_i/M$, where $\sigma^2_i$ is the true posterior variance of image $i$ and $M$ is the length of the chain.
+However, with the presence of correlation in our chain, the aysmptotic variance of $\bar{p}^{(i)}$ is instead some $C_i/M$, where $C_i$ is some positive constant such that $C_i > \sigma^2_i$.
+We use a block bootstrap method with overlapping blocks to get an estimate for $C_i$, $\hat{C}_i$, and compare it to an estimate of $\sigma^2_i$ using the raw chain, $\hat{\sigma}^2_i$.
+Formally,
+
+$$
+M_{eff}^{(i)} = M\frac{\hat{\sigma}^2_i}{\hat{C}_i}.
+$$
+
+Using this metric, a chain of length $M = 1000$ for the marginalized likelihood method yields $\min_i M_{eff}^{(i)} = `r round(1000*marginal_M_eff, 2)`$, while the truncated Normal method only yields $\min_i M_{eff}^{(i)} = `r round(1000*trunc_M_eff, 2)`$.
+Thus, the truncated likelihood would require $`r round(marginal_M_eff/trunc_M_eff, 1)`$ times more iterations to be run of the MCMC chain in order to achieve the same amount of effective information about the posterior distribution.
+For this reason, in conjuction with the ACF results, the truncated Normal method is not an effective method for fitting the RBM.
+
+Figure \ref{fig:fitting_plot} shows the posterior probability of each possible $\boldsymbol v \in \{-1,1\}^4$ after fitting the RBM model in the two ways detailed in this section, excluding the truncated Normal method.
 The black vertical lines show the true probabilities of each image based on the parameters used to simulate the training set while the red vertical lines show the empirical distribution.
@@
 ```{r posterior-dsn}
-marginal_sample_good %>% rename(marginal = prob) %>%
+marginal_sample_good %>% rename(marginal = prob) %>% filter(iter > 50) %>%
   left_join(trick_sample_good %>% ungroup() %>% select(-image_id) %>% rename(trick = prob)) %>%
-  left_join(trunc_sample_good %>% ungroup() %>% select(-image_id) %>% rename(trunc = prob)) %>%
   ungroup() %>%
   mutate(image_id = paste0("image_", image_id)) %>% 
-  gather(method, prob, trick, marginal, trunc) %>%
+  gather(method, prob, trick, marginal) %>%
   spread(image_id, prob) -> all_statistics
@@
   ylab("Scaled Posterior Density") + xlab("Probability of Image") +
-  scale_colour_discrete("Method", labels = c("Marginalized likelihood", "Trick prior", "Truncated Normal prior")) +
-  scale_fill_discrete("Method", labels = c("Marginalized likelihood", "Trick prior", "Truncated Normal prior")) +
+  scale_colour_discrete("Method", labels = c("Marginalized likelihood", "Trick prior")) +
+  scale_fill_discrete("Method", labels = c("Marginalized likelihood", "Trick prior")) +
   theme(legend.position = "bottom") -> p.models
@@
 
-```{r fitting_plot, fig.cap="\\label{fig:fitting_plot}Posterior probability of each possible 4-pixel image using three Bayesian fitting techniques, trick prior, truncated Normal prior, and marginalized likelihood.
+```{r fitting_plot, fig.cap="\\label{fig:fitting_plot}Posterior probability of each possible 4-pixel image using two of the three Bayesian fitting techniques, trick prior and marginalized likelihood.
 The black vertical lines show the true probabilities of each image based on the parameters used to simulate the images while the red vertical lines show the empirical distribution.

### 2016-09-19 writing/draft.Rmd
updated to block means and add table

@@
 ```{r models-load}
-source("helpers/functions.R")
 load("data/sample_images.Rdata")
@@
 ###
-### MBB Statistic
+### Overlapping block means
 ###
-mbb_mean <- function(data_star, data, b) {
-  m <- length(data_star)
+overlap_mean <- function(data, b) {
   n <- length(data)
@@
 
-  sqrt(m)*(mean(data_star) - mean(blockmeans))
-}
-
-###
-### MBB Sample
-###
-mbb_sample <- function(data, b) {
-  n <- length(data)
-  k <- floor(n/b)
-
-  start <- sample.int(n - b + 1, k, replace = TRUE)
-  data_star <- NULL
-  for(i in 1:k) {
-    data_star <- c(data_star, data[start[i]:(start[i] + b - 1)])
-  }
-
-  return(data_star)
-}
-
-###
-### Block Bootstrap function
-###
-bb <- function(x, b, B, boot_stat, boot_sample) {
-  #ii. For blocksize b, generate B MBB versions of statistic
-  bb.stat <- array(NA, B)
-  for(i in seq_len(B)) {
-    bb.stat[i] <- boot_stat(boot_sample(x, b), x, b)
-  }
-  return(bb.stat)
+  blockmeans
 }
@@
   group_by(v1, v2, v3, v4, image_id) %>%
-  do(mbb = bb(.$prob, 2000^(1/3), 2000, mbb_mean, mbb_sample)) -> marginal_mbb
+  do(means = overlap_mean(.$prob, 2000^(1/3))) -> marginal_means
 
@@
   group_by(v1, v2, v3, v4, image_id) %>%
-  do(mbb = bb(.$prob, 5000^(1/3), 2000, mbb_mean, mbb_sample)) -> trunc_mbb
-
-marginal_mbb %>%
+  do(means = overlap_mean(.$prob, 5000^(1/3))) -> trunc_means
+
+marginal_means %>%
   group_by(v1, v2, v3, v4, image_id) %>%
-  do(data.frame(C = var(.$mbb[[1]]))) %>%
+  do(data.frame(C = 2000^(1/3)*var(.$means[[1]]))) %>%
   left_join(marginal_sample_good %>% group_by(v1, v2, v3, v4) %>% summarise(sigma2 = var(prob))) %>%
-  mutate(M_eff = sigma2/C) %>%
+  mutate(M_eff = sigma2/C,
+         model = "Marginal Likelihood") %>%
+  bind_rows(trunc_means %>%
+    group_by(v1, v2, v3, v4, image_id) %>%
+    do(data.frame(C = 5000^(1/3)*var(.$means[[1]]))) %>%
+    left_join(trunc_sample_good %>% group_by(v1, v2, v3, v4) %>% summarise(sigma2 = var(prob))) %>%
+    mutate(M_eff = sigma2/C,
+           model = "Truncated Normal Prior")) %>%
+  select(-C, -sigma2) %>%
+  spread(model, M_eff) %>%
   ungroup() %>%
-  summarise(M_eff = min(M_eff)) %>% as.numeric() -> marginal_M_eff
-
-trunc_mbb %>%
-  group_by(v1, v2, v3, v4, image_id) %>%
-  do(data.frame(C = var(.$mbb[[1]]))) %>%
-  left_join(trunc_sample_good %>% group_by(v1, v2, v3, v4) %>% summarise(sigma2 = var(prob))) %>%
-  mutate(M_eff = sigma2/C) %>%
-  ungroup() %>%
-  summarise(M_eff = min(M_eff)) %>% as.numeric() -> trunc_M_eff
+  select(-(v1:v4)) %>%
+  gather(Model, `$M_{eff}$`, -image_id) %>%
+  mutate(`$M_{eff}$` = 1000*`$M_{eff}$`) %>%
+  spread(image_id, `$M_{eff}$`) -> M_eff_table
+
+M_eff_ratio <- M_eff_table[1,-1]/M_eff_table[2,-1]
 ```
@@
 However, with the presence of correlation in our chain, the aysmptotic variance of $\bar{p}^{(i)}$ is instead some $C_i/M$, where $C_i$ is some positive constant such that $C_i > \sigma^2_i$.
-We use a block bootstrap method with overlapping blocks to get an estimate for $C_i$, $\hat{C}_i$, and compare it to an estimate of $\sigma^2_i$ using the raw chain, $\hat{\sigma}^2_i$.
+We can use an overlapping blockmeans approach to get a crude estimate for $C_i$, $\hat{C}_i = b \text{Var} \overline{p_i}$ where $b$ is the block length and $\overline{p_i}$ are the blockmeans for image $i$.
+We compare it to an estimate of $\sigma^2_i$ using the raw chain, $\hat{\sigma}^2_i$.
 Formally,
@@
 
-Using this metric, a chain of length $M = 1000$ for the marginalized likelihood method yields $\min_i M_{eff}^{(i)} = `r round(1000*marginal_M_eff, 2)`$, while the truncated Normal method only yields $\min_i M_{eff}^{(i)} = `r round(1000*trunc_M_eff, 2)`$.
-Thus, the truncated likelihood would require $`r round(marginal_M_eff/trunc_M_eff, 1)`$ times more iterations to be run of the MCMC chain in order to achieve the same amount of effective information about the posterior distribution.
+```{r m_eff_table}
+M_eff_table %>%
+  gather(Image, M_eff, -Model) %>%
+  spread(Model, M_eff) %>%
+  mutate(Image = as.numeric(Image)) %>%
+  arrange(Image) -> M_eff_tmp
+
+cbind(M_eff_tmp[1:8,], M_eff_tmp[9:16,]) %>%
+  kable(digits = 2, caption = paste0("\\label{tab:m_eff}The effective sample sizes for a chain of length $M = 1000$ of all $16$ images.
+The truncated likelihood would require at least $", round(min(M_eff_table[1, -1])/min(M_eff_table[2, -1]), 1), "$ times more iterations to be run of the MCMC chain in order to achieve the same amount of effective information about the posterior distribution."))
+```
+
+The effective sample sizes for a chain of length $M = 1000$ of all $16$ images are presented in table \ref{tab:m_eff}.
+These range from  $`r round(min(M_eff_table[1, -1]), 2)`$ to $`r round(max(M_eff_table[1, -1]), 2)`$ for the marginalized likelihood method, while the truncated Normal method only yields between $`r round(min(M_eff_table[2, -1]), 2)`$ and $`r round(max(M_eff_table[2, -1]), 2)`$ effective draws.
+Thus, the truncated likelihood would require at least $`r round(min(M_eff_table[1, -1])/min(M_eff_table[2, -1]), 1)`$ times more iterations to be run of the MCMC chain in order to achieve the same amount of effective information about the posterior distribution.
 For this reason, in conjuction with the ACF results, the truncated Normal method is not an effective method for fitting the RBM.

### 2016-09-28 writing/draft.Rmd
starting to change up instability section

@@
 
-@schweinberger2011instability introduces a related concept, *instability*.
-Instability is characterized by excessive sensitivity in the model, where a small change in $\boldsymbol x$ leads to a large change in the probability mass function.
-Consider discrete exponential families of distributions with probability mass functions of the form
-\begin{align*}
-f_{\boldsymbol \theta} (\boldsymbol x) = \frac{\exp\left(\eta^T(\boldsymbol \theta) t(\boldsymbol x)\right)}{\gamma(\boldsymbol \theta)}
-\end{align*}
-@schweinberger2011instability provides the following formal definition for instability in a discrete exponential family model.
-
-\begin{defn}
-A discrete exponential family distribution is \emph{unstable} if for any $C > 0$, however large, there exists $N_C > 0$ such that 
+Considering a class of exponential families of distributions, @schweinberger2011instability introduced a concept of model defficiency related to *instability*.
+Instability is characterized by excessive sensitivity in the model, where small changes in the components of potential data outcomes, $\boldsymbol x$, may lead to substantial changes in the probability mass function.
+To quantify "instability", it is useful to imagine how a data model (the RBM in particular) might be expanded to incorporate more and more observations.
+To increase the size of a RBM model to handle more "visible variables", it becomes necessary to expand the number of model parameters (and in the process of increasing the model size in visibles, one may also arbitrarily expand the number of hidden variables to be included).
+A relevant quantity to this end is a (scaled) extremal log-probability ratio given by
 \begin{align}
-\max\limits_{\boldsymbol x_N \in \mathcal{X}_N}[\eta^T(\boldsymbol \theta) t(\boldsymbol x)] - \min\limits_{\boldsymbol x_N \in \mathcal{X}_N}[\eta^T(\boldsymbol \theta) t(\boldsymbol x)] > CN \text{ for all } N > N_C \label{eq:unstable}
+\frac{1}{V} \log \left[\frac{\max\limits_{(v_1, \dots, v_V) \in \{-1,1\}^V}P_{\boldsymbol \theta}(v_1, \dots, v_V)}{\min\limits_{(v_1, \dots, v_V) \in \{-1,1\}^V}P_{\boldsymbol \theta}(v_1, \dots, v_V)}\right] \equiv \frac{1}{V} \text{ELPR}_V(\boldsymbol \theta) \label{eqn:elpr}
 \end{align}
-
-Conversely, a discrete exponential family distribution is \emph{stable} if there exists constants $C > 0$ and $N_C > 0$ such that 
-\begin{align}
-\max\limits_{\boldsymbol x_N \in \mathcal{X}_N}[\eta^T(\boldsymbol \theta) t(\boldsymbol x)] - \min\limits_{\boldsymbol x_N \in \mathcal{X}_N}[\eta^T(\boldsymbol \theta) t(\boldsymbol x)] \le CN \text{ for all } N > N_C.
-\label{eq:stable}
-\end{align}
+where $P_{\boldsymbol \theta}(v_1, \dots, v_V) \propto \sum_{\{h_j\}\in \{-1,1\}}\exp\left(\sum_{i = 1}^V \theta_{v_i} v_i + \sum_{j = 1}^H \theta_{h_j} h_j + \sum_{i = 1}^V \sum_{j = 1}^H \theta_{ij} v_i h_j\right)$ is the RBM probability of observing outcome $(v_1, \dots, v_V)$ for the visible variables for parameters $\boldsymbol \theta = (\theta_{v_1}, \dots, \theta_{HV})$.
+
+Essentially, in the process of developing a model to handle increasing numbers of visibles (i.e., $V \rightarrow \infty$), we want (\ref{eqn:elpr}) to stay bounded, meaning that the largest probability under this model should maintain a magnitude within the smallest probability allowed under the model.
+This leads to the following definition:
+
+\begin{defn}[unstable RBM]
+A RBM model formulation is \emph{unstable} if, as the number of visible variables increase ($V \rightarrow \infty$), it holds that 
+\begin{align*}
+\lim\limits_{V \rightarrow \infty} \frac{1}{V} \log \left[\frac{\max\limits_{(v_1, \dots, v_V) \in \{-1,1\}^V}P_{\boldsymbol \theta}(v_1, \dots, v_V)}{\min\limits_{(v_1, \dots, v_V) \in \{-1,1\}^V}P_{\boldsymbol \theta}(v_1, \dots, v_V)}\right] = \infty.
+\end{align*}
 \end{defn}
+In other words, given any $C > 0$, there exists an integer $V_C > 0$ so that $\frac{1}{V}\text{ELPR}_V(\boldsymbol \theta) > C$ for all $V \ge V_C$.
+
+Note that this definition of *unstable* is a generalization of that in @schweinberger2011instability allowing nonexponential family models and an increasing number of parameters (while this definition appears different, it does match that in @schweinberger2011instability for the special exponential models considered there).
+
+Unstable RBM models are undesirable for several reasons. One is that, as mentioned above, small changes in data can lead to overly-sensitive changes in probability. Consider, for example, 
+$$
+\Delta_V(\boldsymbol \theta) \equiv \max \left\{\log \frac{P_{\boldsymbol \theta}(v_1, \dots, v_V)}{P_{\boldsymbol \theta}(v_1^*, \dots, v_V^*)} : (v_1, \dots, v_V) \& (v_1^*, \dots, v_V^*) \in \{-1,1\}^V \text{ differ by exactly one component}\right\},
+$$
+the biggest log-probability ratio for a one-pixed (one component) change in data outcomes (visibles). We then have the following result.
 
@@
 So, by examining $R(\boldsymbol\theta)/V$, we can determine a lower bound for the maximum log-likelihood ratio of two sets of observables that differ in only one element, which if large is consistent with the concept of instability detailed above.
-
-Instability in a discrete exponential model is in itself a measure of model impropriety, however there is also a relationship between instability and near-degeneracy in a discrete exponential model.
-
-\begin{lemma}[Schweinberger (2011)]
-If a discrete exponential family distribution is unstable, then it is degenerate in the sense that for any $0 < \epsilon < 1$, $P_\theta(\boldsymbol X \in \mathcal{M}_{\epsilon, N}) \rightarrow 1$ as $N \rightarrow \infty$ where $\mathcal{M}_{\epsilon, N} = \{\boldsymbol x \in \mathcal{X}^N:  \eta^T(\boldsymbol \theta) t(\boldsymbol x) > (1-\epsilon)\max\limits_{\boldsymbol x_N \in \mathcal{X}_N}[\eta^T(\boldsymbol \theta) t(\boldsymbol x)]\}$ is the set of points in the sample space that yield arbitrarily high values of the unnormalized pmf.
-\end{lemma}
-
-Thus, if a discrete exponential family model is unstable, then it is also near-degenerate.
-Discrete RBMs (the focus of this paper) are curved exponential family models for particular choice of sufficient statistics ($\boldsymbol \tau = \{h_j, v_i, v_i h_j: i = 1, \dots, V, j= 1, \dots, H\}$), meaning instability in the RBM implies near-degeneracy as well.
 

### 2016-10-01 writing/appendix-instability.Rmd
body of instability rewrite complete

--- 2016-09-18 writing/draft.Rmd
@@
----
-title: "An exposition on the propriety of restricted Boltzmann machines"
-author: "Andee Kaplan, Daniel Nordman, Stephen Vardeman"
-date: "September, 2016"
-linestretch: 2 
-output: 
-  pdf_document:
-    fig_caption: yes
-    includes:
-      in_header: header.tex
-    number_sections: yes
-bibliography: references/refs.bib
-abstract: |
-  A restricted Boltzmann machine (RBM) is an undirected graphical model constructed for discrete or continuous random variables, with two layers, one hidden and one visible, with no conditional dependency within a layer.
-In recent years, RBMs have risen to prominence due to their connection to deep learning.
-By treating a hidden layer of one RBM as the visible layer in a second RBM a deep architecture can be created.
-RBMs are thought to have the ability to encode very complex and rich structures in data, making them attractive for supervised learning.
-However, the generative behavior of RBMs has largely been unexplored.
-In this paper, we discuss the relationship between parameter RBM specification and the prevalence of degenerate models as well as the difficulties that arise in fitting such highly flexible models in the binary case.
----
-
-```{r libraries_options, echo=FALSE, message=FALSE, warnings=FALSE}
-library(knitr)
-library(ggplot2)
-library(dplyr)
-library(tidyr)
-
-opts_chunk$set(echo=FALSE, message=FALSE, warnings=FALSE)
-theme_set(theme_bw(base_family="serif"))
-
-source("helpers/functions.R")
-```
-
-# Introduction
-
-The data mining and machine learning communities have recently found a great interest in deep learning, specifically in stacked restricted Boltzmann machines (RBMs).
-(**TODO: deep learning, deep RBM references.**) A RBM is an undirected probabilistic graphical model (for discrete or continuous random variables) with two layers, one hidden and one visible, with no conditional dependency within a layer (**TODO: RBM references.**).
-These models have reportedly been used with success in classification of images.
-(**TODO: classification references.**) However, the model properties are largely unexplored in the literature and the commonly cited fitting methodology remains heuristic-based and abstruse.
-In this paper, we attempt to provide steps toward a thorough understanding of the model and its behavior from the perspective of statistical model theory and then explore the possibility of a rigorous fitting methodology.
-We find the RBM model class to be deficient in two fundamental ways.
-
-First, the models are unsatisfactory as a conceptualization of how data is generated.
-According to Fisher, the aim of a statistical model is to represent data in a compact way [@fisher1922mathematical].
-Neyman and Box agree with this assessment and go further to state that a model should "provide an explanation of the mechanism underlying the observed phenomena" [@lehmann1990model, @box1967discrimination].
-RBMs often fail to generate data that resemble realistic data and thus an unsatisfactory conceptualization of the data generation process.
-Additionally, we find that RBMs easily exhibit a kind of instability in the parameter space.
-In practice, this is seen when a single pixel change in an image results in a wildly different classification.
-Such model impropriety issues have  been documented in other deep architectures recently [@szegedy2013intriguing, @nguyen2014deep].
-We investigate these phenomena in RBMs in section \ref{explorations} through simulations of small, manageable examples.
-
-Secondly, the fitting of these models is problematic.
-As the size of these models grows, both maximum likelihood and Bayesian methods of fitting quickly being intractable.
-But even when fitting is feasible, there are unique challenges due to the overparameterization of RBMs.
-In section \ref{bayesian-model-fitting} we compare three Bayesian modelling techniques, each proposed with the goal of avoiding parts of the RBM parameter space that yield improper models.
-As could be expected, with computational complexity comes an increase in fitting accuracy, but at the cost of feasibility.
-Additionally, because these models are highly overparametrized, any fully principled fitting method will reproduce the empirical distribution of most realistic training sets.
-There is no "smoothing" achieved with a RBM fit using a rigorous method.
-We are therefore led to be skeptical that any model built using these structures, like a deep Boltzmann machine, can achieve prediction or inference in a principled way.
-
-This paper is structured as follows, section \ref{background} gives a definition of the RBM including the joint distribution of hidden and visible variables and shows the connection to deep learning.
-Additionally, measures of model impropriety and methods of detecting them are defined.
-Section \ref{explorations} details our explorations into the model behavior and propriety (or lack thereof) of the RBM class.
-We discuss three Bayesian fitting techniques with the goal of avoiding model impropriety in section \ref{bayesian-model-fitting} and wrap up with a discussion in section \ref{discussion}.
-Additionally, there is an accompanying appendix with auxiliary proofs relating to the choice of data encoding in section \ref{appendix}.
-
-# Background
-
-## Restricted Boltzmann machines
-
-A restricted Boltzmann machine (RBM) is an undirected graphical model specified for discrete or continuous random variables, binary variables being most common.
-In this paper, we consider the binary case of an RBM only.
-A RBM architecture has two layers, hidden ($\mathcal{H}$) and visible ($\mathcal{V}$), with no connections within a layer.
-An example of this structure is in figure \ref{fig:rbm} with the hidden nodes indicated by gray circles and the visible nodes indicated by white circles.
-
-\begin{figure}[ht]
-  \centering
-  \resizebox{\linewidth}{!}{\input{tikz/rbm.tikz}}
-  \caption{An example restricted Boltzmann machine (RBM), consisting of two layers, one hidden ($\mathcal{H}$) and one visible ($\mathcal{V}$), with no connections within a layer.
-Hidden nodes are indicated by gray circles and the visible nodes indicated by white circles.}
-  \label{fig:rbm}
-\end{figure}
-
-A common use for RBMs is to create features for use in classification. For example, binary images could be classified where the pixel values would be treated as the visible variables $v_i$.
-
-### Joint distribution
-Let $\boldsymbol x = \{h_1, \dots, h_H, v_1,\dots,v_V\}$ represent the states of the visible and hidden nodes in an RBM.
-We will consider both $\boldsymbol x \in \{0,1\}^{H + V}$ and $\boldsymbol x \in \{-1, 1\}^{H + V}$, two binary $H + V$-dimensional spaces.
-We posit a parametric form for probabilities corresponding to a potential vector of states of each node taking the value of $1$.
-
-\begin{align}
-\label{eqn:pmf}
-f_{\boldsymbol \theta} (\boldsymbol x) = \frac{\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\gamma(\boldsymbol \theta)}
-\end{align}
-
-Where the denominator $\gamma(\boldsymbol \theta)$ is the normalizing function that ensures the distribution sums to one, $\gamma(\boldsymbol \theta) = \sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)$.
-Let $\boldsymbol \tau = \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{X}\} = \{v_i, h_j, v_i h_j: i = 1, \dots, V, j = 1, \dots H\}$ be the vector of sufficient statistics in the model and $Q(\boldsymbol x) = \sum\limits_{i = 1}^H\sum\limits_{j=1}^V \theta_{ij} h_i v_j + \sum\limits_{i = 1}^H\theta_{hi} h_i + \sum\limits_{j = 1}^V\theta_{vj} v_j$, the neg-potential function.
-The RBM model is parameterized by $\boldsymbol \theta$, which contains two types of parameters, main effects and interaction effects.
-The main effects parameters ($\theta_{v_i}, \theta_{h_j}$) weight the values of the visible and hidden nodes and the interaction effects parameters ($\theta_{ij}$) weight the values of the connections, or dependencies, between hidden and visible layers.
-
-Due to the potential size of the model the normalizing constant $\gamma(\boldsymbol \theta)$ can be practically impossible to calculate, making it impossible to estimate the model parameter vector in an obvious fashion.
-A kind of Gibbs sampling can be tried, which is natural due to the restricted architecture of the RBM.
-Specifically, the conditional independence of nodes in each layer given those in the other allows for simulation of the main and hidden layers.
-
-### Connection to Deep Learning
-
-RBMs have risen to prominence in recent years due to their connection to deep learning [see @hinton2006fast, @salakhutdinov2012efficient, or @srivastava2013modeling for examples].
-By stacking multiple layers of RBMs in a deep architecture, proponents of the models claim the ability to learn "internal representations that become increasingly complex, which is considered to be a promising way of solving object and speech recognition problems" [@salakhutdinov2009deep, pp.
-450].
-The stacking is achieved by treating a hidden layer of one RBM as the visible layer in a second RBM, and so on until the desired multi-layer architecture is created.
-
-## Degeneracy, instability, and uninterpretability... Oh my!
-The highly flexible nature of a RBM (having as it does $H + V + H V$ parameters) makes three kinds of potential model impropriety of concern, *degeneracy, instability, and uninterpretability*.
-In this section we define these characteristics, consider how to detect them in a RBM, and point out their relationship to one another.
-
-### Near-degeneracy
-In Random Graph Model theory, *model degeneracy* means there is a disproportionate amount of probability placed on only a few elements of the sample space, $\mathcal{X}$, by the model.
-In the scenario of random graph models, $\mathcal{X}$ denotes all possible graphs that can be constructed from a set of nodes.
-An exponentially parameterized random graph model has a distribution of the form
-\begin{align*}
-f_{\boldsymbol \theta} (\boldsymbol x) = \frac{\exp\left(\boldsymbol \theta^T t(\boldsymbol x)\right)}{\gamma(\boldsymbol \theta)}
-\end{align*}
-where $\boldsymbol \theta \in \Theta \subset \mathbb{R}^q$ is the model parameter, and $t: \mathcal{X} \rightarrow \mathbb{R}^q$ are statistics based on the adjacency matrix of a graph.
-Again, $\gamma(\boldsymbol \theta)$ is the normalizing function that guarantees the distribution is valid, $\gamma(\boldsymbol \theta) = \sum_{\boldsymbol x \in \mathcal{X}} \exp\left(\boldsymbol \theta^T t(\boldsymbol x)\right)$.
-Let $C$ be the convex hull of $\{t(\boldsymbol x): \boldsymbol x \in \mathcal{X}\}$.
-@handcock2003assessing classify an exponentially parametrized random graph model as *near-degenerate* if $\mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta}t(X)$ is close to the boundary of the $C$.
-
-The RBM class of models can be thought to exhibit an analogous occurrence of *near-degeneracy* when random variables in the neg-potential function $Q(\cdot)$, $t(\boldsymbol x)$, have a mean vector $\mu(\boldsymbol \theta)$ close to the boundary of the convex hull of $\mathcal{X}$, where $t(\boldsymbol x) = \{v_1, \dots, v_V, h_1, \dots, h_H, v_1 h_1, \dots, v_V h_H \} \in \mathcal{X}$ and the mean parameterization on the model parameters, $\mu(\boldsymbol \theta)$ is
-\begin{align*}
-\mu(\boldsymbol \theta) &= \text{E}_{\boldsymbol \theta} t(\boldsymbol X)  \\
-&= \sum\limits_{x \in \mathcal{X}} \left\{ t(\boldsymbol x) f_{\boldsymbol \theta} (\boldsymbol x) \right\} \\
-&= \sum\limits_{x \in \mathcal{X}} \left\{ t(\boldsymbol x)\dfrac{\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}\right\} \\
-\end{align*}
-
-### Instability
-
-@schweinberger2011instability introduces a related concept, *instability*.
-Instability is characterized by excessive sensitivity in the model, where a small change in $\boldsymbol x$ leads to a large change in the probability mass function.
-Consider discrete exponential families of distributions with probability mass functions of the form
-\begin{align*}
-f_{\boldsymbol \theta} (\boldsymbol x) = \frac{\exp\left(\eta^T(\boldsymbol \theta) t(\boldsymbol x)\right)}{\gamma(\boldsymbol \theta)}
-\end{align*}
-@schweinberger2011instability provides the following formal definition for instability in a discrete exponential family model.
-
-\begin{defn}
-A discrete exponential family distribution is \emph{unstable} if for any $C > 0$, however large, there exists $N_C > 0$ such that 
-\begin{align}
-\max\limits_{\boldsymbol x_N \in \mathcal{X}_N}[\eta^T(\boldsymbol \theta) t(\boldsymbol x)] - \min\limits_{\boldsymbol x_N \in \mathcal{X}_N}[\eta^T(\boldsymbol \theta) t(\boldsymbol x)] > CN \text{ for all } N > N_C \label{eq:unstable}
-\end{align}
-
-Conversely, a discrete exponential family distribution is \emph{stable} if there exists constants $C > 0$ and $N_C > 0$ such that 
-\begin{align}
-\max\limits_{\boldsymbol x_N \in \mathcal{X}_N}[\eta^T(\boldsymbol \theta) t(\boldsymbol x)] - \min\limits_{\boldsymbol x_N \in \mathcal{X}_N}[\eta^T(\boldsymbol \theta) t(\boldsymbol x)] \le CN \text{ for all } N > N_C.
-\label{eq:stable}
-\end{align}
-\end{defn}
-
-For RBMs, an analogous concept could concern the joint distribution of both hidden and visible nodes, where $N$ is the number of nodes in the model, $H + V$.
-However, instability in terms of only the visible nodes is a more useful measure of model impropriety.
-To see this, consider modeling images using an RBM.
-One would be interested in the maximum change in the likelihood when a single pixel is changed in the visibles as a proxy for instability in the model.
-So we concern ourselves with a notion of instability that deals with the maximum likelihood ratio of two vectors of visibles that differ at only one coordinate.
-
 \begin{align*}
@@
 \end{align*}
-
-Let $R(\boldsymbol \theta) \equiv \max_{\boldsymbol v} \max_{\boldsymbol h}Q(\boldsymbol x) - \min_{\boldsymbol v}\max_{\boldsymbol h}Q(\boldsymbol x) - H\log 2$.
-Then, if $R(\boldsymbol \theta) > C*V$, this implies that the maximum log-likelihood ratio of two observations that differ in only one element is greater than $C$.
-So, by examining $R(\boldsymbol\theta)/V$, we can determine a lower bound for the maximum log-likelihood ratio of two sets of observables that differ in only one element, which if large is consistent with the concept of instability detailed above.
-
-Instability in a discrete exponential model is in itself a measure of model impropriety, however there is also a relationship between instability and near-degeneracy in a discrete exponential model.
-
-\begin{lemma}[Schweinberger (2011)]
-If a discrete exponential family distribution is unstable, then it is degenerate in the sense that for any $0 < \epsilon < 1$, $P_\theta(\boldsymbol X \in \mathcal{M}_{\epsilon, N}) \rightarrow 1$ as $N \rightarrow \infty$ where $\mathcal{M}_{\epsilon, N} = \{\boldsymbol x \in \mathcal{X}^N:  \eta^T(\boldsymbol \theta) t(\boldsymbol x) > (1-\epsilon)\max\limits_{\boldsymbol x_N \in \mathcal{X}_N}[\eta^T(\boldsymbol \theta) t(\boldsymbol x)]\}$ is the set of points in the sample space that yield arbitrarily high values of the unnormalized pmf.
-\end{lemma}
-
-Thus, if a discrete exponential family model is unstable, then it is also near-degenerate.
-Discrete RBMs (the focus of this paper) are curved exponential family models for particular choice of sufficient statistics ($\boldsymbol \tau = \{h_j, v_i, v_i h_j: i = 1, \dots, V, j= 1, \dots, H\}$), meaning instability in the RBM implies near-degeneracy as well.
-
-### Uninterpretability
-@kaiser2007statistical defines a measure of model impropriety he calls *uninterpretability*, which is characterized by marginal mean-structure not being maintained in the model.
-A measure of this is the magnitude of the difference between model expectations, E$\left[\boldsymbol X | \boldsymbol \theta\right]$, and expectations given independence, E$\left[\boldsymbol X | \boldsymbol \theta^* \right]$, where $\boldsymbol \theta^*$ is defined to equal $\boldsymbol \theta$ with all $\theta_{ij} = 0$ for $i = 1, \dots, V, j = 1, \dots, H$.
-Using this concept it is possible to investigate what conditions lead to uninterpretability in a model versus those that guarantee interpretable models.
-A model is considered without uninterpretability if its expected value is not very different from the same model with assumed independence.
-The quantities to compare in the RBM case are
-
-\begin{align*}
-\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] &= \sum\limits_{\boldsymbol x \in \mathcal{X}} \boldsymbol x f_{\boldsymbol \theta}(\boldsymbol x) \\
-&= \sum\limits_{\boldsymbol x \in \mathcal{X}} \boldsymbol x \frac{\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)} \\
-&= \frac{\sum\limits_{\boldsymbol x \in \mathcal{X}} x_k \exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}  \qquad k = 1, \dots, H + V
-\end{align*}
-and
-\begin{align*}
-\text{E}\left[\boldsymbol X | \boldsymbol 
-theta^*\right] &= \sum\limits_{\boldsymbol x \in \mathcal{X}} \boldsymbol x \frac{\exp\left(\sum\limits_{i = 1}^V \theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)} \\
-&= \frac{\sum\limits_{\boldsymbol x \in \mathcal{X}} x_k \exp\left(\sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}  \qquad k = 1, \dots, H + V
-\end{align*}
-
-If $|\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] - \text{E}\left[\boldsymbol X | \emptyset\right]$ is large then the RBM with parameter vector $\boldsymbol \theta$ $is uninterpetable.
-
-# Explorations
-
-We explore the relationship between magnitudes of $\boldsymbol \theta$ and the three measures of model impropriety, near-degeneracy, instability, and uninterpretability, for RBM models of varying size through simulated data.
-
-## Tiny example
-To illustrate the ideas of model near-degeneracy, instability, and uninterpretability in a RBM, we consider first the smallest possible (toy) example that consists of one visible node $v_1$ and one hidden node $h_1$ that are both binary.
-A schematic of this model can be found in figure \ref{fig:toymodel}.
-Because it seems most common, we'll begin by employing $0/1$ encoding of binary variables (both $h_1$ and $v_1$ taking values in $\{0,1\}$).
-Eventually we'll note in section \ref{data-coding-to-mitigate-degeneracy} that $-1/1$ coding has substantial advantages.
-
-\begin{figure}[ht]
-  \centering
-  \resizebox{1cm}{!}{\input{tikz/toymodel.tikz}}
-  \caption{A small example restricted Boltzmann machine (RBM), with two nodes, one hidden and one visible.}
-  \label{fig:toymodel}
-\end{figure}
-
-### Impropriety three ways
-
-For this small model, we are able to investigate the symptoms of model impropriety (near-degeneracy, instability uninterpretability).
-First, because we are in three dimensions, we are able to explicitly illustrate the shape of boundary of the convex hull of $\boldsymbol tau$ and explore the behavior of the mean parameterization of the parameters.
-In figure \ref{fig:toyhull}, the convex hull of our "statistic space", $\boldsymbol \tau$, in three dimensions for this toy problem is shown from multiple perspectives enclosed by the unit cube, it convex hull of $\{0,1\}^3$ space.
-In this small model, note that the convex hull of $\boldsymbol tau$ does not fill the unrestricted hull because of the relationship between the elements of $\{(v_1, h_1, v_1 h_1 : (v_1, h_1) \in \{0, 1\}^2 \}$.
-
-```{r models}
-m1_bin <- calc_hull(1, 1)
-m1_neg <- calc_hull(1, 1, type = "negative")
-```
-
-\begin{figure}[ht]
-  \begin{minipage}{0.45\textwidth}
-  \resizebox{\linewidth}{!}{
-    \tdplotsetmaincoords{60}{-60}
-    \input{tikz/toyhull_top.tikz}
-
-    ```{r hull, results='asis', echo=FALSE, include=TRUE, message=FALSE}
-    cat(sapply(1:nrow(m1_bin$c_hull$hull), function(x) paste0("\\draw", paste(paste0("(", apply(m1_bin$possible_t[cbind(m1_bin$c_hull$hull, m1_bin$c_hull$hull[,1])[x,],], 1, paste, collapse=","), ")"), collapse=" -- "), "; \n")))
-   ```
-    \end{tikzpicture}
-  }
-  \end{minipage}
-  \begin{minipage}{0.45\textwidth}
-  \resizebox{\linewidth}{!}{
-    \tdplotsetmaincoords{60}{120}
-    \input{tikz/toyhull_top.tikz}
-
-    ```{r, results='asis', echo=FALSE, include=TRUE, message=FALSE}
-    cat(sapply(1:nrow(m1_bin$c_hull$hull), function(x) paste0("\\draw", paste(paste0("(", apply(m1_bin$possible_t[cbind(m1_bin$c_hull$hull, m1_bin$c_hull$hull[,1])[x,],], 1, paste, collapse=","), ")"), collapse=" -- "), "; \n")))
-    ```
-    \end{tikzpicture}
-  }
-  \end{minipage}
-  \caption{The convex hull of the "statistic space"" in three dimensions for the toy RBM with one visible and one hidden node.
-The convex hull of $\boldsymbol \tau = \{\boldsymbol t(\boldsymbol x) : \boldsymbol x \in \mathcal{X}\}$ does not fill the unit cube because of the relationship between the elements of $\boldsymbol \tau$.}
-\label{fig:toyhull}
-\end{figure}
-
-Additionally, we can compute the mean vector for $\boldsymbol t(\boldsymbol x)$ as a function of the model parameters as 
-\begin{align*}
-\boldsymbol \mu(\boldsymbol \theta) &= \text{E}_{\boldsymbol \theta}\left[ \boldsymbol t(\boldsymbol X) \right] \\
-&= \sum\limits_{\boldsymbol x \in \mathcal{X}} \left\{ \boldsymbol t(\boldsymbol x) \frac{\exp\left( \theta_{11} h_1 v_1 + \theta_{h1} h_1 + \theta_{v1} v_1\right)}{\sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left( \theta_{11} h_1 v_1 + \theta_{h1} h_1 + \theta_{v1} v_1\right)} \right\} \\
-&= \sum\limits_{\boldsymbol x \in \mathcal{X}} \left\{ t(x) \frac{\exp\left( \theta_{11} h_1 v_1 + \theta_{h1} h_1 + \theta_{v1} v_1\right)}{\gamma(\boldsymbol \theta)} \right\} \\
-&= \left[
-\begin{matrix}
-\frac{\exp\left(\theta_{v_1}\right) + \exp\left(\theta_{11} + \theta_{v_1} + \theta_{h_1}\right)}{\gamma(\boldsymbol \theta)} \\
-\frac{\exp\left(\theta_{h_1}\right) + \exp\left(\theta_{11} + \theta_{v_1} + \theta_{h_1}\right)}{\gamma(\boldsymbol \theta)} \\
-\frac{\exp\left(\theta_{11} + \theta_{v_1} + \theta_{h_1}\right)}{\gamma(\boldsymbol \theta)} \\
-\end{matrix} \right]
-\end{align*}
-The three parametric functions giving $\boldsymbol \mu(\boldsymbol \theta)$ can be visualized in figure \ref{fig:degen_toy}.
-By examining these, we see that as values of coordinates of $\boldsymbol \theta$ grow larger in magnitude, at least one mean function for the entries of $\boldsymbol t(\boldsymbol x)$ is approaching a value 0 or 1, indicating a potential for $\text{E}_{\boldsymbol \theta} \boldsymbol t(\boldsymbol x)$ to be near to the boundary of the convex hull of $\boldsymbol \tau$.
-Thus, for a very small example we can see the relationship between values of $\boldsymbol \theta$ and model degeneracy.
-
-```{r near_dependency, fig.cap="\\label{fig:degen_toy}The three parametric mean functions of sufficient statistics for a RBM with one visible and one hidden node.", fig.height = 4, fig.align='center', message=FALSE, cache=TRUE}
-possibles <- stats(1, 1, type = "binary")
-
-expand.grid(theta_h1 = seq(-5, 5, length.out = 21),
-            theta_v1 = seq(-5, 5, length.out = 21),
-            theta_11 = seq(-5, 5, length.out = 5)) %>%
-  data.frame() %>%
-  group_by(theta_v1, theta_h1, theta_11) %>%
-  do(data.frame(gamma = sum(exp(possibles %*% t(data.matrix(.)))),
-                mu_v1 = possibles[,"v1"] %*% exp(possibles %*% t(data.matrix(.))),
-                mu_h1 = possibles[,"h1"] %*% exp(possibles %*% t(data.matrix(.))),
-                mu_11 = possibles[,"theta11"] %*% exp(possibles %*% t(data.matrix(.))))) %>%
-  mutate_each(funs(./gamma), starts_with("mu")) %>%
-  select(-gamma) %>%
-  gather(mu_parameter, value, mu_v1, mu_h1, mu_11) %>%
-  ungroup() %>%
-  mutate(theta_11 = factor(theta_11, labels = paste0("theta[11]==", unique(theta_11))),
-         mu_parameter = ifelse(mu_parameter == "mu_v1", "paste(E[theta], v[1])", 
-                               ifelse(mu_parameter == "mu_h1", "paste(E[theta], h[1])", "paste(E[theta], v[1], h[1])"))) %>%
-  #factor(mu_parameter, labels = paste0(gsub("_", "(theta[", unique(mu_parameter)), "])"))) %>%
-  ggplot() +
-  geom_tile(aes(theta_v1, theta_h1, fill = value))  +
-  geom_contour(aes(theta_v1, theta_h1, z = value), colour = "black") +
-  facet_grid(mu_parameter ~ theta_11, labeller = label_parsed) +
-  xlab(expression(theta[v1])) +
-  ylab(expression(theta[h1])) +
-  scale_fill_gradient2("Mean", low = "#af8dc3", high = "#7fbf7b", midpoint = .5, limits = c(0, 1)) +
-  theme(aspect.ratio = 1)
-```
-
-Secondly, we can look at $R(\boldsymbol \theta)$ for this tiny model.
-Recall that gives us a lower bound for the loglikelihood ratio of the minimum possible change in $\boldsymbol x$, indicating instability.
-
-\begin{align*}
-R(\boldsymbol \theta) &= \max_{\boldsymbol v} \max_{\boldsymbol h}Q(\boldsymbol x) - \min_{\boldsymbol v}\max_{\boldsymbol h}Q(\boldsymbol x) - H\log 2\\
-&= \max_{\boldsymbol v} \max_{\boldsymbol h}\left[ \theta_{11} h_1 v_1 + \theta_{h_1}h_1 + \theta_{v_1} v_1 \right] -  \min_{\boldsymbol v}\max_{\boldsymbol h}\left[ \theta_{11} h_1 v_1 + \theta_{h_1}h_1 + \theta_{v_1} v_1 \right] -\log 2\\
-&=\max\left\{ \theta_{11} + \theta_{h_1} + \theta_{v_1}, \theta_{h_1}, \theta_{v_1}, 0  \right\} \\
-&\qquad - \min\left\{\theta_{v_1} + \max\left\{\theta_{11} + \theta_{h1}, -\theta_{11} - \theta_{h1}\right\}, -\theta_{v_1} + \max\left\{-\theta_{11} + \theta_{h1}, \theta_{11} - \theta_{h1}\right\}\right\} - \log2\\
-\end{align*}
-
-Figure \ref{fig:instab} shows $R(\boldsymbol \theta)/V$ for various values of $\boldsymbol \theta$.
-We can see that this quantity is increasing for larger magnitudes of $\boldsymbol \theta$, especially for large values of $\theta_{11}$.
-This suggests instability as $|\boldsymbol \theta|$ becomes large, agreeing with the concerns about near-degeneracy produced by consideration of $\boldsymbol \mu(\boldsymbol \theta)$.
-
-```{r instab, fig.cap="\\label{fig:instab}$R(\\boldsymbol \\theta)/V$ for various values of $\\boldsymbol \\theta$ for the tiny example model.
-(Recall that here $V$ is the number of visible nodes and here is $1$.
-This quantity is increasing for larger magnitudes of $\\boldsymbol \\theta$.", fig.height = 4, fig.align='center'}
-expand.grid(theta_h1 = seq(-5, 5, length.out = 21),
-            theta_v1 = seq(-5, 5, length.out = 21),
-            theta_11 = seq(-5, 5, length.out = 5)) %>%
-  data.frame() %>%
-  group_by(theta_v1, theta_h1, theta_11) %>%
-  do(data.frame(Q = cbind(possibles, possibles %*% t(data.matrix(.))))) %>%
-  group_by(theta_v1, theta_h1, theta_11, Q.v1) %>%
-  do(data.frame(maxq = max(.$Q.V4))) %>%
-  group_by(theta_v1, theta_h1, theta_11) %>%
-  do(data.frame(q_max = max(.$maxq) - min(.$maxq) - log(2))) %>%
-  ungroup() %>%
-  mutate(theta_11 = factor(theta_11, labels = paste0("theta[11]==", unique(theta_11)))) %>%
-  ggplot() +
-  geom_tile(aes(theta_v1, theta_h1, fill = q_max)) +
-  geom_contour(aes(theta_v1, theta_h1, z = q_max), colour = "black") +
-  facet_grid(. ~ theta_11, labeller = label_parsed) +
-  xlab(expression(theta[v1])) +
-  ylab(expression(theta[h1])) +
-  scale_fill_gradient(expression(frac(R(theta), V)), low = "yellow", high = "red") +
-  theme(aspect.ratio = 1)
-```
-
-Finally, we can look at the difference between model expectations, E$\left[\boldsymbol X | \boldsymbol \theta\right]$, and expectations given independence, E$\left[\boldsymbol X | \boldsymbol \theta^*\right ]$ for the tiny toy model to understand the effect of $\boldsymbol \theta$ on uninterpretability in the RBM.
-
-\begin{align*}
-\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] -   \text{E}\left[\boldsymbol X | \boldsymbol \theta^* \right] 
-&= \left[
-\begin{matrix}
-\frac{\exp\left(\theta_{v_1}\right) + \exp\left(\theta_{11} + \theta_{v_1} + \theta_{h_1}\right)}{\exp\left(\theta_{v_1}\right) + \exp\left(\theta_{h_1}\right) + \exp\left(\theta_{11} + \theta_{v_1} + \theta_{h_1}\right)} \\
-\frac{\exp\left(\theta_{h_1}\right) + \exp\left(\theta_{11} + \theta_{v_1} + \theta_{h_1}\right)}{\exp\left(\theta_{v_1}\right) + \exp\left(\theta_{h_1}\right) + \exp\left(\theta_{11} + \theta_{v_1} + \theta_{h_1}\right)}
-\end{matrix} \right] - \left[
-\begin{matrix}
-\frac{\exp\left(\theta_{v_1}\right) + \exp\left(\theta_{v_1} + \theta_{h_1}\right)}{\exp\left(\theta_{v_1}\right) + \exp\left(\theta_{h_1}\right) + \exp\left(\theta_{v_1} + \theta_{h_1}\right)} \\
-\frac{\exp\left(\theta_{h_1}\right) + \exp\left(+ \theta_{v_1} + \theta_{h_1}\right)}{\exp\left(\theta_{v_1}\right) + \exp\left(\theta_{h_1}\right) + \exp\left(+ \theta_{v_1} + \theta_{h_1}\right)} 
-\end{matrix} \right] \\
-&= \left[
-\begin{matrix}
-\frac{\left|\exp\left(\theta_{11} + \theta_{v_1} + 2\theta_{h_1}\right) - \exp\left( \theta_{v_1} + 2\theta_{h_1}\right) \right| }{\left(\exp\left(\theta_{v_1}\right) + \exp\left(\theta_{h_1}\right) + \exp\left(\theta_{11} + \theta_{v_1} + \theta_{h_1}\right)\right)\left(\exp\left(\theta_{v_1}\right) + \exp\left(\theta_{h_1}\right) + \exp\left(+ \theta_{v_1} + \theta_{h_1}\right)\right)} \\
-\frac{\left|\exp\left(\theta_{11} + 2\theta_{v_1} + \theta_{h_1}\right) - \exp\left( 2\theta_{v_1} + \theta_{h_1}\right) \right| }{\left(\exp\left(\theta_{v_1}\right) + \exp\left(\theta_{h_1}\right) + \exp\left(\theta_{11} + \theta_{v_1} + \theta_{h_1}\right)\right)\left(\exp\left(\theta_{v_1}\right) + \exp\left(\theta_{h_1}\right) + \exp\left(+ \theta_{v_1} + \theta_{h_1}\right)\right)}
-\end{matrix}\right]
-\end{align*}
-
-Again, we can inspect these coordinate functions of $\boldsymbol \theta$ to look for a relationship between parameter values and uninterpretability for the toy RBM.
-
-```{r uninterp, fig.cap="\\label{fig:uninterp}The difference between coordinates of model expectations, E$\\left[\\boldsymbol X | \\boldsymbol \\theta\\right]$, and expectations given independence, E$\\left[\\boldsymbol X | \\boldsymbol \\theta^* \\right ]$ for a RBM with one visible and one hidden node.", fig.height = 4, fig.align='center', message=FALSE, cache=TRUE, warning=FALSE}
-expand.grid(theta_h1 = seq(-5, 5, length.out = 21),
-            theta_v1 = seq(-5, 5, length.out = 21),
-            theta_11 = seq(-5, 5, length.out = 5)) %>%
-  data.frame() %>%
-  group_by(theta_v1, theta_h1, theta_11) %>%
-  do(data.frame(gamma = sum(exp(possibles %*% t(data.matrix(.)))),
-                mu_v1_dep = possibles[,"v1"] %*% exp(possibles %*% t(data.matrix(.))),
-                mu_h1_dep = possibles[,"h1"] %*% exp(possibles %*% t(data.matrix(.))))) %>%
-  mutate_each(funs(./gamma), starts_with("mu")) %>%
-  select(-gamma) -> exp_dep
-
-expand.grid(theta_h1 = seq(-5, 5, length.out = 21),
-            theta_v1 = seq(-5, 5, length.out = 21),
-            theta_11 = 0) %>%
-  data.frame() %>%
-  group_by(theta_v1, theta_h1, theta_11) %>%
-  do(data.frame(gamma = sum(exp(possibles %*% t(data.matrix(.)))),
-                mu_v1_indep = possibles[,"v1"] %*% exp(possibles %*% t(data.matrix(.))),
-                mu_h1_indep = possibles[,"h1"] %*% exp(possibles %*% t(data.matrix(.))))) %>%
-  mutate_each(funs(./gamma), starts_with("mu")) %>% ungroup() %>%
-  select(-gamma, -theta_11) -> exp_indep
-
-exp_dep %>%
-  left_join(exp_indep) %>%
-  mutate(mu_v1 = abs(mu_v1_dep - mu_v1_indep),
-         mu_h1 = abs(mu_h1_dep - mu_h1_indep)) %>%
-  gather(mu_parameter, value, mu_v1, mu_h1) %>%
-  ungroup() %>%
-  mutate(theta_11 = factor(theta_11, labels = paste0("theta[11]==", unique(theta_11))),
-          mu_parameter = ifelse(mu_parameter == "mu_v1", "paste(E[theta], v[1])", "paste(E[theta], h[1])")) %>%
-  ggplot() +
-  geom_tile(aes(theta_v1, theta_h1, fill = value))  +
-  geom_contour(aes(theta_v1, theta_h1, z = value), colour = "black") +
-  facet_grid(mu_parameter ~ theta_11, labeller = label_parsed) +
-  xlab(expression(theta[v1])) +
-  ylab(expression(theta[h1])) +
-  scale_fill_gradient("Component absolute difference", low = "yellow", high = "red", limits = c(0, 1)) +
-  theme(aspect.ratio = 1)
-```
-
-Figure \ref{fig:uninterp} shows that the absolute difference between coordinates of the vector of model expectations, E$\left[\boldsymbol X | \boldsymbol \theta\right]$, and corresponding expectations given independence, E$\left[\boldsymbol X | \boldsymbol \theta^*\right ]$ for the toy RBM grow as the values of $\boldsymbol \theta$ are farther from zero, especially for large magnitudes of $\theta_{11}$.
-This is a third indication that parameter vectors of large magnitude could lead to model impropriety in an RBM.
-
-### Data coding to mitigate degeneracy
-Multiple encodings of the binary variables are possible.
-For example, we could allow $\mathcal{H} \in \{0,1\}^H, \mathcal{V} \in \{0,1\}^V$, as in the previous sections or we could encode the state of the variables as $\mathcal{H} \in \{-1,1\}^H, \mathcal{V} \in \{-1,1\}^V$.
-This will result in $\boldsymbol t(\boldsymbol x) \in \{0,1\}^{H + V + HV}$ or $\boldsymbol t(\boldsymbol x) \in \{-1,1\}^{H + V + HV}$ depending on how we encode "on" and "off" states in the nodes.
-To explore the effect of this encoding on the potential for apparent near-degeneracy of an RBM, we can consider the ratio of the volume of the hypercube with corners at elements of $\boldsymbol \tau$ to the volume of the convex hull of $\boldsymbol \tau$.
-
-For the small two node example, the $0/1$ encoding loses `r round((1-m1_bin$c_hull$vol)*100,2)`% of the volume of $[0,1]^3$ and the $-1/1$ encoding loses `r round((1-m1_neg$c_hull$vol/(2^3))*100,2)`% of the volume of $[-1,1]^3$.
-While there is clearly a one-to-one mapping of models for these two possible codings, this notion of lost volume can be helpful to conceptualize how difficult it will be for the mean vector $\boldsymbol \mu(\boldsymbol \theta)$ to avoid the boundary, and thus avoid near-degeneracy.
-In fact, if we look at the ratio of volume within a convex hull defined by $\boldsymbol \tau$ and the corresponding hypercube, we can see a relationship emerge as the number of nodes (and thus parameters) increases.
-In figure \ref{fig:volume_plot}, it is evident that as the number of parameters increases, this ratio is decreasing at an increasing rate, meaning it gets more and more difficult to avoid the boundary of the convex hull and thus near degeneracy.
-Additionally, it appears that the $-1/1$ encoding suffers slightly less from this problem.
-This suggests that if intuition about models is to be formed in terms of volume fractions of vectors of "non-pathological" means for $\boldsymbol t(\boldsymbol x)$, $-1/1$ encoding is preferable to $0/1$ coding.
-
-```{r volume_plot, echo=FALSE, fig.cap="\\label{fig:volume_plot}The relationship between volume of the convex hull of possible values of the RBM sufficient statistics and the cube containing it for different configurations of nodes.", out.height = "3in", fig.align='center'}
-test_cases <- expand.grid(data.frame(1:4)[,rep(1,2)]) %>% 
-  rename(H = X1.4, V = X1.4.1) %>%
-  filter(H <= V) %>% #remove those cases with less visibles than hiddens. Is this necessary?
-  mutate(n_param = H*V + H + V) %>%
-  mutate(max_facets = (2^(H+V))^(floor(n_param/2))) %>%
-  filter(n_param <= 11) #calc_hull can't handle any higher dimensions currently
-
-bin_res <- plyr::dlply(test_cases, plyr::.(n_param), function(x) calc_hull(x$V, x$H, "binary"))
-neg_res <- plyr::dlply(test_cases, plyr::.(n_param), function(x) calc_hull(x$V, x$H, "negative"))
-
-plyr::ldply(bin_res, function(x) x$c_hull$vol) %>% 
-  mutate(frac_vol = V1/(1^n_param)) %>%
-  inner_join(plyr::ldply(neg_res, function(x) x$c_hull$vol) %>% 
-               mutate(frac_vol = V1/(2^n_param)),
-             by="n_param") %>%
-   rename(vol.bin = V1.x, vol.neg = V1.y, frac_vol.bin = frac_vol.x, frac_vol.neg = frac_vol.y) %>%
-  gather(vars, value, -n_param) %>%
-  separate(vars, c("type", "encoding"), "\\.") %>%
-  spread(type, value) %>%
-  ggplot() +
-  geom_point(aes(x=n_param, y=frac_vol, colour=encoding)) +
-  geom_line(aes(x=n_param, y=frac_vol, colour=encoding, group=encoding)) +
-  ylab("Fraction of unrestricted volume") +
-  xlab("Number of parameters") +
-  scale_colour_discrete("Encoding", labels=c("Binary (1,0)", "Negative (1,-1)"))
-```
-
-In addition to the argument that the $-1/1$ data encoding mitigates some perceived prevalence of near-degeneracy it also has the benefit of providing a guaranteed non-degenerate model at $\boldsymbol \theta = \boldsymbol 0 \in \mathbb{R}^{H + V + H*V}$.
-The proof of this is included in appendix \ref{on-the-one-to-one-correspondence-between parameters-and moments-in-the-rbm-distribution}.
-And while from some computing perspectives $0/1$ coding might seem most natural, the $-1/1$ coding is far more convenient and interpretable from the point of view of statistical modeling, where it makes parameters simply interpreted in terms of symmetrically defined main effects and interactions.
-In light of all of these matters we will henceforth employ the $-1/1$ coding.
-
-## Exploring manageable examples
-```{r degen-data, message=FALSE, warning=FALSE, cache=TRUE}
-#reshape data functions
-plot_data <- function(res, grid = FALSE) {
-
-  plot.data <- data.frame()
-
-  if(!grid) {
-    for(i in 1:nrow(res)) {
-      tmp <- res$g_theta[[i]] %>% data.frame()
-      H <- res[i,]$H
-      V <- res[i,]$V
-      N <- res[i,]$N
-      r1 <- res[i,]$r1
-      r2 <- res[i,]$r2
-      C <- res[i,]$C
-      epsilon <- res[i,]$epsilon
-
-      tmp %>% 
-        rowwise() %>% 
-        mutate_(ss_interaction = paste0("(", paste(paste0(names(.)[(H + V + 1):(H + V + H*V)], "^2"), collapse = " + "), ")"),
-                ss_main = paste0("sum(", paste(paste0(names(.)[1:(H+V)],"^2"), collapse = " + "), ")")) %>%    
-        ungroup() -> ratio
-
-      inner_join(ratio, res$outside[[i]] %>% ungroup()) %>%
-        select(ss_interaction, ss_main, near_hull) %>%
-        mutate(H = H, V = V, n_param = H + V + H*V, N = H + V, N = N, r1 = r1, r2 = r2, C = C, epsilon = epsilon) %>%
-        rbind(plot.data) -> plot.data
-    }
-  } else {
-    for(i in 1:nrow(res)) {
-      tmp <- res$g_theta[[i]] %>% data.frame()
-      H <- res[i,]$H
-      V <- res[i,]$V
-      N <- res[i,]$N
-      r1 <- res[i,]$r1
-      r2 <- res[i,]$r2
-
-      tmp %>% 
-        rowwise() %>% 
-        mutate_(ss_interaction = paste0("(", paste(paste0(names(.)[(H + V + 1):(H + V + H*V)], "^2"), collapse = " + "), ")"),
-                ss_main = paste0("sum(", paste(paste0(names(.)[1:(H+V)],"^2"), collapse = " + "), ")")) %>%  
-        ungroup() -> ratio
-
-      inner_join(ratio, res$outside[[i]] %>% ungroup()) %>%
-        select(ss_interaction, ss_main, near_hull) %>%
-        mutate(H = H, V = V, n_param = H + V + H*V, N = H + V, N = N, r1 = r1, r2 = r2) %>%
-        rbind(plot.data) -> plot.data
-    }
-  }
-
-  return(plot.data)
-}
-indep_params <- function(samp, H, V) {
-  samp[, (H + V + 1):ncol(samp)] <- 0
-  samp
-}
-max_Q <- function(theta, stats) {
-  apply(crossprod(t(stats), theta), 2, max)
-}
-min_Q <- function(theta, stats) {
-  apply(crossprod(t(stats), theta), 2, min)
-}
-min_max_Q <- function(theta, stats) {
-  require(dplyr)
-  tcrossprod(stats, theta) %>% data.frame() %>%
-    cbind(stats) %>%
-    group_by_(.dots = colnames(stats)[grepl("v", colnames(stats)) & !grepl("theta", colnames(stats))]) %>%
-    summarise_each(funs(max), contains("X")) %>%
-    ungroup() %>%
-    summarise_each(funs(min), contains("X")) %>%
-    select(contains("X")) %>%
-    data.matrix() %>%
-    t()
-}
-
-#grid data
-load("data/results_grid.RData")
-
-#near-degeneracy
-plot_dat_grid <- res %>% plot_data(grid = TRUE)
-
-plot_dat_grid %>%
-  group_by(r1, r2, H, V) %>%
-  summarise(frac_degen = sum(near_hull)/n(), count = n()) %>% 
-  ungroup() %>% 
-  mutate(Hiddens = paste0("H = ", H), Visibles = paste0("V = ", V)) -> convex_hull_summary
-
-#uninterpretability
-res %>%
-  group_by(H, V, N, r1, r2) %>%
-  do(indep_exp = t(expected_value(t(indep_params(.$samp[[1]], .$H, .$V)), .$stat[[1]]))[, -((.$H + .$V + 1):(.$H + .$V + .$H*.$V))],
-     marg_exp = .$g_theta[[1]][, (.$H + .$V + .$H*.$V + 1):(ncol(.$g_theta[[1]])-.$H*.$V)]) -> exp_vals
-
-exp_vals %>%
-  group_by(H, V, N, r1, r2) %>%
-  do(data.frame(max_abs_diff = apply(abs(.$indep_exp[[1]] - .$marg_exp[[1]]), 1, max))) %>%
-  group_by(H, V, N, r1, r2) %>%
-  summarise(max_abs_diff = mean(max_abs_diff)) %>%
-  mutate(Hiddens = paste0("H = ", H), Visibles = paste0("V = ", V)) -> exp_vals_summary
-
-#instability
-res %>%
-  group_by(H, V, N, r1, r2) %>%
-  do(data.frame(max = max_Q(t(.$samp[[1]]), .$stat[[1]]),
-                min = min_Q(t(.$samp[[1]]), .$stat[[1]]),
-                min_max = min_max_Q(.$samp[[1]], .$stat[[1]]))) %>%
-  ungroup() %>% 
-  group_by(H, V, N, r1, r2) %>%
-  mutate(LHS1 = (max - min)/N,
-         LHS2 = (max - min_max - H*log(2))/N) %>%
-  summarise_each(funs(mean), LHS1, LHS2) -> max_q_summary
-
-convex_hull_summary %>%
-  left_join(max_q_summary) %>%
-  left_join(exp_vals_summary) -> three_ways
-
-```
-
-To explore the behavior of the RBM parameters $\boldsymbol \theta$ as it relates to near-degeneracy, instability, and uninterpretability.
-we consider models of small size.
-For $H, V \in \{1, \dots, 4\}$, we sample `r dim(res$samp[[1]])[1]` values of $\boldsymbol \theta$ with various magnitudes (details to follow).
-For each set of parameters we then calculate the three metrics of model impropriety introduced in section \ref{degeneracy-instability-and-uninterpretability-oh-my}, $\boldsymbol \mu(\boldsymbol \theta)$, $R(\boldsymbol x)/V$, and the coordinates of $\left|\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] -   \text{E}\left[\boldsymbol X | \boldsymbol \theta^* \right] \right|$.
-In the case of near-degeneracy, we can go further and classify each model as near-degenerate or "viable" based on the distance of $\boldsymbol \mu(\boldsymbol \theta)$ from the boundary of the convex hull of $\boldsymbol \tau$ and look at the fraction of models that are "near-degenerate", meaning they are within a small distance to the boundary of the convex hull.
-
-In this numerical experiment, we split $\boldsymbol \theta$ into $\boldsymbol \theta_{interaction}$ and $\boldsymbol \theta_{main}$, in reference to which sufficient statistics the parameters correspond to (whether they multiply a $v_i$ or a $h_j$ or they multiply a $v_i h_j$), and allow the two types of terms to have varying average magnitudes, $||\boldsymbol \theta_{main} || /(H+V)$ and $||\boldsymbol \theta_{interaction} || /(H*V)$.
-These average magnitudes vary on a grid between `r min(plot_dat_grid$r1/(plot_dat_grid$H + plot_dat_grid$V))` and `r max(plot_dat_grid$r1/(plot_dat_grid$H + plot_dat_grid$V))` with `r length(unique(plot_dat_grid$r1/(plot_dat_grid$H + plot_dat_grid$V)))` breaks, yielding `r length(unique(plot_dat_grid$r1/(plot_dat_grid$H + plot_dat_grid$V)))^2` grid points.
-(By looking at the average magnitudes, we are able to later consider the potential benefit of shrinking each parameter value $\theta_i$ towards zero in a Bayesian fitting technique.) At each point in the grid, `r dim(res$samp[[1]])[1]` vectors are uniformly sampled on a sphere with radius corresponding to the first coordinate in the grid ($\boldsymbol \theta_{main}$) and `r dim(res$samp[[1]])[1]` vectors are uniformly sampled on a sphere with radius corresponding to the second coordinate in the grid ($\boldsymbol \theta_{main}$) via sums of squared and scaled iid Normal$(0, 1)$ variables.
-These vectors are then paired to create `r dim(res$samp[[1]])[1]` values of $\boldsymbol \theta$ with magnitudes at each point in the grid.
-
-The results of this numerical study are summarized in figures \ref{fig:degen_plots}, \ref{fig:instab_plots}, and \ref{fig:uninterp_plots}.
-From these three figures, it s clear to see that all three measures of model impropriety show higher values the larger the magnitude of the parameter vectors.
-Additionally, since there are $H*V$ interaction terms in $\boldsymbol \theta$ versus only $H + V$ main effects terms, for large models there are many more interaction parameters than main effects in the model.
-And so, severely limiting the magnitude of the indivisual interactions may well help prevent model impropriety.
-
-```{r degen_plots, fig.cap="\\label{fig:degen_plots}Results from the numerical experiment, here looking at the fraction of models that were near-degenerate for each combination of magnitude of $\\boldsymbol \\theta$ and model size.
-Black lines show the contour levels for fraction of near-degenerace, while the thick black line shows the level where the fraction of near-degenerate models is .05.", fig.height=4}
-three_ways %>%
-  mutate(r1 = round(r1/(H + V), 8), r2 = round(r2/(H * V), 8)) %>% 
-  ggplot() +
-  geom_tile(aes(x = r1, y = r2, fill = frac_degen)) +
-  geom_contour(aes(x = r1, y = r2, z = frac_degen), colour = "black", bins = 8) +
-  geom_contour(aes(x = r1, y = r2, z = frac_degen), colour = "black", breaks = .05, size = 1.5) +
-  geom_abline(aes(intercept = 0, slope = 1), alpha = .5, lty = 2) +
-  scale_fill_gradient("Fraction near-degenerate", low = "yellow", high = "red") +
-  facet_grid(Visibles~Hiddens) +
-  xlab(expression(paste(paste("||", theta[main], "||", "/(H + V)")))) +
-  ylab(expression(paste(paste("||", theta[interaction], "||", "/(H*V)")))) +
-  theme(aspect.ratio = 1) -> p.degen
-
-three_ways %>%
-  mutate(r1 = round(r1/(H + V), 8), r2 = round(r2/(H * V), 8)) %>% 
-  ggplot() +
-  geom_tile(aes(x = r1, y = r2, fill = max_abs_diff)) +
-  geom_contour(aes(x = r1, y = r2, z = max_abs_diff), colour = "black", bins = 8) +
-  #geom_contour(aes(x = r1, y = r2, z = mean_abs_diff), colour = "black", breaks = .05, size = 1.5) +
-  geom_abline(aes(intercept = 0, slope = 1), alpha = .5, lty = 2) +
-  scale_fill_gradient("Mean max absolute difference", low = "yellow", high = "red", limits = c(0,2)) +
-  facet_grid(Visibles~Hiddens) +
-  xlab(expression(paste(paste("||", theta[main], "||", "/(H + V)")))) +
-  ylab(expression(paste(paste("||", theta[interaction], "||", "/(H*V)")))) +
-  theme(aspect.ratio = 1) -> p.exp_diff
-
-three_ways %>%
-  mutate(r1 = round(r1/(H + V), 8), r2 = round(r2/(H * V), 8)) %>% 
-  ggplot() +
-  geom_tile(aes(x = r1, y = r2, fill = LHS1)) +
-  geom_contour(aes(x = r1, y = r2, z = LHS1), colour = "black", bins = 8) +
-  #geom_contour(aes(x = r1, y = r2, z = mean_max_q), colour = "black", breaks = .05, size = 1.5) +
-  geom_abline(aes(intercept = 0, slope = 1), alpha = .5, lty = 2) +
-  scale_fill_gradient(expression(frac(max(Q) - min(Q), H + V)), low = "yellow", high = "red") +
-  facet_grid(Visibles~Hiddens) +
-  xlab(expression(paste(paste("||", theta[main], "||", "/(H + V)")))) +
-  ylab(expression(paste(paste("||", theta[interaction], "||", "/(H*V)")))) +
-  theme(aspect.ratio = 1) -> p.max_q
-
-three_ways %>%
-  mutate(r1 = round(r1/(H + V), 8), r2 = round(r2/(H * V), 8)) %>% 
-  ggplot() +
-  geom_tile(aes(x = r1, y = r2, fill = LHS2)) +
-  geom_contour(aes(x = r1, y = r2, z = LHS2), colour = "black", bins = 8) +
-  #geom_contour(aes(x = r1, y = r2, z = mean_max_q), colour = "black", breaks = .05, size = 1.5) +
-  geom_abline(aes(intercept = 0, slope = 1), alpha = .5, lty = 2) +
-  scale_fill_gradient(expression(frac(bar(R(theta)), V)), low = "yellow", high = "red") +
-  facet_grid(Visibles~Hiddens) +
-  xlab(expression(paste(paste("||", theta[main], "||", "/(H + V)")))) +
-  ylab(expression(paste(paste("||", theta[interaction], "||", "/(H*V)")))) +
-  theme(aspect.ratio = 1) -> p.min_max_q
-
-p.degen
-```
-
-```{r instab_plots_2, fig.cap="\\label{fig:instab_plots}Results from the experiment, here looking at the sample mean value of $R(\\boldsymbol \\theta)/V$ at each grid point for each combination of magnitude of $\\boldsymbol \\theta$ and model size.
-As the magnitude of $\\boldsymbol \\theta$ grows, so does the value of this metric, indicating typical instability in the model.", fig.height=4}
-p.min_max_q
-```
-
-```{r uninterp_plots, fig.cap="\\label{fig:uninterp_plots}The sample mean of the maximum component of the absolute difference between the model expectation vector, E$\\left[\\boldsymbol X | \\boldsymbol \\theta\\right]$, and the expectation vector given independence, E$\\left[\\boldsymbol X | \\boldsymbol \\theta^* \\right ]$.
-Larger magnitudes of $\\boldsymbol \\theta$ correspond to larger differences, thus indicating reduced interpretability.", fig.height=4}
-p.exp_diff
-```
-
-Figure \ref{fig:split_plots_dist} shows the fraction near-degenerate for each magnitude of $\boldsymbol \theta$ for each model architecture.
-For each number of visibles in the model, the fraction near-degenerate becomes greater than zero at larger values of $||\boldsymbol \theta||$ as $H$ increases and the slope becomes steeper as $H$ increases also.
-This shows that as model size gets larger, the risk of degeneracy starts at a slightly larger magnitude of parameters, but very quickly increases until reaching close to 1.
-
-```{r split_plots2, warning=FALSE, fig.cap="\\label{fig:split_plots_dist}The fraction near-degenerate for each magnitude of $\\boldsymbol \\theta$.
-For each number of visibles in the model, the fraction near-degenerate becomes greater than zero at larger values of $||\\boldsymbol \\theta||$ as $H$ increases and the slope becomes steeper as $H$ increases also.", fig.height=5}
-plot_dat_grid %>% 
-  mutate(dist = sqrt(r1^2 + r2^2)/(H + V + H*V)) %>%
-  group_by(dist, H, V) %>%
-  summarise(frac_degen = sum(near_hull)/n()) %>%
-  mutate(Hiddens = as.character(H), Visibles = paste0("V = ", V)) %>%
-  ggplot() +
-  geom_point(aes(dist, frac_degen, colour = Hiddens), size = .5) +
-  #geom_smooth(aes(dist, frac_degen, colour = Hiddens, group = Hiddens)) +
-  facet_wrap(~Visibles) +
-  ylim(c(0,1)) +
-  xlab(expression(paste("||", theta, "||", "/(H + V + H*V), Average distance from the origin"))) +
-  ylab("Fraction Near-degenerate") +
-  scale_colour_discrete("H")
-```
-
-```{r, frac-degen, include=FALSE}
-plot_dat_grid %>%
-  mutate(dist = sqrt(r1^2 + r2^2)) %>%
-  group_by(dist, H, V) %>%
-  summarise(frac_degen = sum(near_hull)/n()) %>%
-  ungroup() %>%
-  group_by(H, V) -> tmp
-
-tmp %>%
-  filter(frac_degen > 0) %>%
-  mutate(min_degen_pos = min(dist)) %>%
-  select(H, V, min_degen_pos) %>% unique() %>%
-left_join(tmp %>%
-  filter(frac_degen > .6) %>%
-  mutate(min_degen_50 = min(dist)) %>%
-  select(H, V, min_degen_50) %>% unique()) %>%
-  mutate(min_degen_pos = min_degen_pos/(H + V + H*V),
-         min_degen_50 = min_degen_50/(H + V + H*V)) %>%
-  mutate(degen_ratio = min_degen_50/min_degen_pos) %>%
-  ggplot() +
-  geom_point(aes(V, degen_ratio, colour = factor(H))) +
-  geom_line(aes(V, degen_ratio, colour = factor(H), group = factor(H))) +
-  ylab("Ratio of Min Distance above 0% near-degenerate to 50% near-degenerate") +
-  xlab("Visibles") +
-  scale_colour_discrete("Hiddens")
-
-```
-These manageable examples indicate that RBMs are near-degenerate, unstable, and uninterpretable for large portions of the parameter space.
-This, however, is not the only potential problem to be faced when using these models.
-There is the matter of pricipled/rigorous fitting of RBM models.
-
-# Model Fitting
-
-Typically, fitting a RBM via maximum likelihood (ML) methods will be infeasible mainly due to the intractibility of the normalizing term $\gamma(\boldsymbol \theta)$ in a model of any realistic size.
-Ad hoc methods are used instead, which aim to avoid this problem by using stochastic ML that employ a small number of MCMC draws.
-
-However, computational concerns are not the only issues with fitting an RBM using ML.
-In addition, the RBM model, with the appropriate choice of parameters, has the potential to re-create any distribution for the data (i.e., reproduce any specification of cell probabilities for the -1/1 valued data outcomes).
-To see this, we provide a small example.
-
-For simplicity, consider a model with two visible variables $(V_1,V_2)$ and one hidden $H_1$.  In this case,
-there are four possible data realizations for $(V_1,V_2)$  given by $(\pm 1, \pm 1)$ and we may express the model probabilities
-as
-$$
-P(V_1=v_1,V_2=v_2| \theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21}) \propto  \exp\left( v_1  \theta_{v_1} + v_2 \theta_{v_2}\right) \sum_{h \in \{\pm 1\}}\exp\left( h[ \theta_{h_1}  + \theta_{11} v_1 + \theta_{21} v_2] \right),
-$$
-for $(v_1,v_2)\in \{-1, 1\}^2$, in terms of real-valued parameters $\theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21}$. Given any  specified cell probabilities, say
-\begin{equation}
-\label{eqn:0}
-0\leq p_{(-1,-1)},\;p_{(1,-1)}, \;p_{(-1,1)},\;p_{(1,1)},
-\end{equation} 
-for the outcomes $(\pm 1, \pm 1)$,  values for parameters $(\theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21})$ may be chosen to mimic such  cell probabilities with arbitrary closeness.
-In fact, when the cell probabilities (\ref{eqn:0}) are all strictly positive, parameters in the RBM model can be specified to reflect these probabilities exactly.
-And, when one or more of the cell probabilities (\ref{eqn:0}) are zero, the corresponding RBM  probabilities $P(V_1=v_1,V_2=_v2| \theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21})$ may never be identically zero (due to the exponential model form) but
- parameters can be still selected to make the appropriate RBM cell probabilities arbitrarily small.
-
-To demonstrate such model approximations,  we  assume $p_{(-1,-1)}>0$ (without loss of generality) in the specified cell probabilities (\ref{eqn:0}) and replace parameters $\theta_{11},\theta_{21}$ with $\Delta_1 \equiv \theta_{11} +\theta_{21}$ and $\Delta_2 \equiv \theta_{11} -\theta_{21}$.
-We may then prescribe values of $\theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \Delta_1,\Delta_2$  so that the model probability ratio
-$$
-P(V_1=v_1,V_2=v_2|\theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \Delta_1,\Delta_2)/  P(V_1=-1,V_2=-1| \theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \Delta_1,\Delta_2)
-$$
-matches the corresponding  ratio $p_{(v_1,v_2)}/p_{(-1,-1)}$  over three values of $(v_1,v_2) = (1,-1),(-1,1),(1,1)$.
-
-For instance, assuming the cell probabilities from (\ref{eqn:0}) are all positive, these probabilities
- can be exactly reproduced by choosing
- \begin{eqnarray*}
- \theta_{v_1} &=& \frac{1}{2} \log\left( \frac{p_{(1,-1)}}{p_{(-1,-1)}} \frac{\exp(\theta_{h_1}-\Delta_1) + \exp(-\theta_{h_1}+\Delta_1)}{\exp(\theta_{h_1}+\Delta_2) + \exp(-\theta_{h_1}-\Delta_2)}  \right),\\
- \theta_{v_2}&= &\frac{1}{2} \log\left( \frac{p_{(-1,1)}}{p_{(-1,-1)}} \frac{\exp(\theta_{h_1}-\Delta_1) + \exp(-\theta_{h_1}+\Delta_1)}{\exp(\theta_{h_1}-\Delta_2) + \exp(-\theta_{h_1}+\Delta_2)}  \right)
-\end{eqnarray*}
-and selecting $\theta_{h_1}, \Delta_1,\Delta_2$ to solve
-\begin{equation}
-\label{eqn:1}
-\frac{p_{(1,1)}p_{(-1,-1)}}{p_{(-1,1)}p_{(1,-1)}} = \frac{\ell(|\theta_{h_1}|) + \ell(|\Delta_1|) }{\ell(|\theta_{h_1}|) + \ell(|\Delta_2|)},
-\end{equation}
-based on a monotonically increasing function $\ell(x)\equiv \exp(-2x)+\exp(2x)$, $x \geq 0$.
-If  $[p_{(1,1)}p_{(-1,-1)}]/ [p_{(-1,1)}p_{(1,-1)}] \geq 1$, one can pick any values for $\theta_{h_1}, \Delta_2\in \mathbb{R}$ and solve (\ref{eqn:1}) for $|\Delta_1|$; likewise, when $[p_{(1,1)}p_{(-1,-1)}]/ [p_{(-1,1)}p_{(1,-1)}] < 1$ in (\ref{eqn:1}), one may solve for $|\Delta_2|$ upon choosing any values for $\theta_{h_1}, \Delta_1\in \mathbb{R}$.
-
-Alternatively, if exactly one specified cell probability in (\ref{eqn:0}) is zero, say $p_{(1,1)}$ (without loss of generality), we can select parameters $\theta_{v_1},\theta_{v_2}$ as above based on a sequence $(\theta_{h_1}, \Delta_{1}, \Delta_{2}) \equiv (\theta_{h_1}^{(m)}, \Delta_{1}^{(m)}, \Delta_{2}^{(m)})$, $m\in\{1,2,\ldots,\}$ of the remaining parameter values such that $\lim_{m\to \infty}|\Delta_{1}^{(m)}| = \infty$ and $\lim_{m\to \infty}  (|\theta_{h_1}^{(m)}| + |\Delta_{2}^{(m)}|)/|\Delta_{1}^{(m)}|=0$ hold.
-This guarantees that the resulting RBM model matches the given cell probabilities (\ref{eqn:0}) in the limit:
-\begin{equation}
-\label{eqn:2}
-\lim_{m\to \infty}P(V_1=v_1,V_2=v_2| \theta_{v_1}, \theta_{v_2}, \theta_{h_1},\Delta_1,\Delta_2) = p_{(v_1,v_2)},\quad (v_1,v_2)\in\{(\pm 1,\pm 1)\}.
-\end{equation}  
-If exactly two specified probabilities in (\ref{eqn:0}) are zero, say  $p_{(1,1)}$ and $p_{(-1,1)}$ (without loss of generality), then a limit approximation as in (\ref{eqn:2}) follows by picking $\theta_{v_1}$ as above based on any choices of $(\theta_{h_1}, \Delta_1,\Delta_2)$ and choosing  a sequence of $\theta_{v_2}\equiv \theta_{v_2}^{(m)}$ values for which $\theta_{v_2}^{({m})} \rightarrow -\infty$.
-
-The previous discussion intends to indicate that the RBM model suffers from parameter identifiability issues that go beyond mere symmetries in the parametrization.
-Quite different parameter settings can induce the same essential RBM model.However, this  is not the most disastrous implication of the RBM parameterization.
-A far worse consequence is that, when fitting the RBM model by likelihood-based methods, we already know what the answer must be before we begin: namely, such  fitting will simply aim to reproduce the empirical distribution from the data.
-That is, based on a random sample of visible variables, the model for the cell probabilities that has the highest likelihood over  *all possible model classes* (i.e., RBM-based or not) is the empirical distribution, and the parametrization of the RBM model itself ensures that this empirical distribution can be arbitrarily well approximated.
-
-For illustration, continue the simple example from above with $n$ iid observations, each consisting of two realized visibles $(V_1,V_2)$.
-In which case, when the specified cell probabilities $p_{(-1,-1)},p_{(1,-1)}, p_{(-1,1)},p_{(1,1)}$ in (\ref{eqn:0}) are taken as the empirical cell frequencies from the sample, there is no better model based on maximum likelihood and the discussion above (c.f.
-(\ref{eqn:2})) shows that RBM model parameters can be chosen to re-create this empirical distribution to an arbitrarily close degree.
-Hence, RBM model fitting based on ML will often simply seek to reproduce the empirical distribution (i.e., as an extreme type of over-fitting).
-What's more, whenever this empirical distribution contains empty cells, fitting steps for the RBM model will necessarily aim to chase parameters that necessarily diverge in magnitude in order to zero out the corresponding RBM cell probabilities.
-In a data application with a large sample space it is unlikely that the training set will include one of each possible image, thus there will be possible images that are not realized in the training set, unlike this small example.
-As we have already discussed in secion \ref{explorations}, large magnitudes of $\boldsymbol \theta$ lead to model impropriety in the RBM.
-
-Here we consider what might be done in a principled manner, testing on a $V = H = 4$ case that already stretched the limits of what is computable - in particular we consider Bayes methods.
-
-## Bayesian model fitting
-To avoid model impropriety for a fitted RBM, we want to avoid parts of the parameter space $\mathbb{R}^{V + H + VH}$ that lead to near-degeneracy, instability, and uninterpretability.
-Motivated by the insights in section \ref{exploring-manageable-examples}, one idea is to shrink $\boldsymbol \theta$ toward $\boldsymbol 0$ by specifying priors that place low probability on large values of $||\boldsymbol \theta||$, specifically focusing on shrinking $\boldsymbol \theta_{interaction}$ more than $\boldsymbol \theta_{main}$.
-
-```{r param-table}
-load("data/params_theta.Rdata")
-
-params_good <- list(main_hidden = sample.params %>% ungroup() %>% filter(!near_hull) %>% select(starts_with("h"), -H) %>% data.matrix(),
-                    main_visible = sample.params %>% ungroup() %>% filter(!near_hull) %>% select(starts_with("v"), -V) %>% data.matrix(),
-                    interaction = sample.params %>% ungroup() %>% filter(!near_hull) %>% select(starts_with("theta")) %>% data.matrix() %>% matrix(4))
-
-sample.params %>% 
-  ungroup %>% 
-  filter(!near_hull) %>% select(starts_with("v"), starts_with("h"), starts_with("theta"), -H, -V) %>%
-  rename(`$\\theta_{v1}$` = v1,
-         `$\\theta_{v2}$` = v2,
-         `$\\theta_{v3}$` = v3,
-         `$\\theta_{v4}$` = v4,
-         `$\\theta_{h1}$` = h1,
-         `$\\theta_{h2}$` = h2,
-         `$\\theta_{h3}$` = h3,
-         `$\\theta_{h4}$` = h4,
-         `$\\theta_{11}$` = theta11,
-         `$\\theta_{12}$` = theta12,
-         `$\\theta_{13}$` = theta13,
-         `$\\theta_{14}$` = theta14,
-         `$\\theta_{21}$` = theta21,
-         `$\\theta_{22}$` = theta22,
-         `$\\theta_{23}$` = theta23,
-         `$\\theta_{24}$` = theta24,
-         `$\\theta_{31}$` = theta31,
-         `$\\theta_{32}$` = theta32,
-         `$\\theta_{33}$` = theta33,
-         `$\\theta_{34}$` = theta34,
-         `$\\theta_{41}$` = theta41,
-         `$\\theta_{42}$` = theta42,
-         `$\\theta_{43}$` = theta43,
-         `$\\theta_{44}$` = theta44) %>%
-  gather("Parameter", "Value") -> tbl
-
-cbind(tbl[1:8,], tbl[9:16,], tbl[17:24,]) %>%
-  kable(caption = "\\label{tab:theta}Parameters used to fit a test case with $V = H = 4$.
-This parameter vector was chosen as a sampled value of $\\boldsymbol \\theta$ that was not near the convex hull of the sufficient statistics for a grid point in figure \\ref{fig:degen_plots} with $< 5$\\% near-degeneracy.")
-```
-
-We considered a test case with $V = H = 4$ and parameters given in table \ref{tab:theta}.
-This parameter vector was chosen as a sampled value of $\boldsymbol \theta$ that was not near the convex hull of the sufficient statistics for a grid point in figure \ref{fig:degen_plots} with $< 5$\% near-degeneracy.
-We simulated $n = 5,000$ as a training set and fit the RBM using three Bayes methodologies.
-These involved the following:
-
-1.
-*A trick prior.* Here we cancel out normalizing term in the likelihood so that resulting full conditionals of $\boldsymbol \theta$ are multivariate Normal.
-The $h_j$ are carried allong in the MCMC sampling from the posterior as latent variables.
-    \begin{align*}
-    \pi(\boldsymbol \theta) \propto \gamma(\boldsymbol \theta)^n \exp\left(-\frac{1}{2C_{1}}\boldsymbol \theta_{main}'\boldsymbol \theta_{main} -\frac{1}{2C_{2}}\boldsymbol \theta_{interaction}'\boldsymbol \theta_{interaction}\right), \vspace{-.75cm}
-    \end{align*}
-
-    where
-
-    \begin{align*}
-    \gamma(\boldsymbol \theta) = \sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V\sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right) \text{ and } C_{2} < C_{1}
-    \end{align*}
-    This is the method of @li2014biclustering.
-2.
-*A truncated Normal prior.* Here we use independent truncated spherical normal distributions as priors for $\boldsymbol \theta_{main}$ and $\boldsymbol \theta_{interaction}$ with $\sigma_{interaction} < \sigma_{main}$.
-Full conditional distributions are not conjugate, and simulaion from the posterior was accomplished using a geometric adaptive Metropolis Hastings step [@zhou2014some] and calculation of likelihood normalizing constant.
-(This computation is barely feasible for a problem of this size.) Here the $h_j$ are carried along in the MCMC implementation as latent variables.
-3. *A truncated Normal prior and marginalized likelihood.* Here we marginalize out $\boldsymbol h$ in $f_{\boldsymbol \theta}(\boldsymbol x)$, and use the truncated Normal prior. 
-    \begin{align*}
-    g_{\boldsymbol \theta}(\boldsymbol v) = \sum\limits_{\boldsymbol h \in \{-1,1\}^H} \exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right).
-\vspace{-.75cm}
-    \end{align*}
-
-The three fitting methods are ordered by computational feasibility in a real-data situation, with the trick prior being the most feasible due to conjugacy and the marginalized likelihood method the least feasible due to the marginalization and need for an adaptive Metropolis Hastings step.
-All three methods require choosing the values of hyperparameters.
-In each case, we have chosen the values based on a rule of thumb that shrinks $\boldsymbol \theta_{interaction}$ more than $\boldsymbol \theta_{main}$.
-Additionally, the trick prior requires additional tuning to choose $C_1$ and $C_2$, reducing its appeal.
-The values used for the hyperparameters in our simulation are presented in table \ref{tab:hyperparam}.
-
-\begin{table}[ht]
-\centering
-\caption{The values used for the hyperparameters for all three fitting methods.
-A rule of thumb is imposed which decreases prior variances for the model parameters as the size of the model increases and also shrinks $\boldsymbol \theta_{interaction}$ more than $\boldsymbol \theta_{main}$.}
-\label{tab:hyperparam}
-\begin{tabular}{|l|c|c|}
-\hline 
-Method & Hyperparameter & Value \\ 
-\hline \hline
-\multirow{2}{*}{Trick Prior} & $C_1$ & $\frac{C}{N}\frac{1}{H + V}$ \\
- & $C_2$ & $\frac{C}{N}\frac{1}{H*V}$ \\
-\hline
-\multirow{2}{*}{Truncated Normal} & $\sigma^2_{main}$ & $\frac{1}{H + V}$ \\
- & $\sigma^2_{interaction}$ & $\frac{1}{H*V}$ \\
-\hline
-\multirow{2}{*}{Marginalized Likelihood} & $\sigma^2_{main}$ & $\frac{1}{H + V}$ \\
- & $\sigma^2_{interaction}$ & $\frac{1}{H*V}$ \\
-\hline
-\end{tabular}
-\end{table}
-
-It should be noted that the truncated Normal method (2) and the marginalized likelihood method (3) are drawing from the same stationary posterior distribution for images.
-The difference between these two methods is in how well the chains mix and how quickly they arrive at the target distribution.
-After a burn-in period of 50 iterations selected by inspecting the trace plots, we assess the issue of mixing in two ways.
-First, the autocorrelation functions (ACF) for each image are assessed at multiple lags for each method and plotted in figure \ref{fig:acf} with the truncated Normal method in black and the marginalized likelihood method in red.
-As expected, ACF corresponding to the method that marginalizes out the hidden variables from the likelihood decreases to zero at a much faster rate, indicating better mixing for the chain.
-
-```{r models-load}
-source("helpers/functions.R")
-load("data/sample_images.Rdata")
-
-#data and params ------------------------
-H <- 4
-V <- 4
-
-#marginalized likelihood
-load("data/fitted_models_trunc_marginal_full.Rdata")
-marginal_bad <- models_bad
-marginal_good <- models_good
-
-#load trick prior
-load("data/fitted_models_jing_5.8.Rdata")
-trick_bad <- models_bad
-trick_good <- models_good
-
-#truncated normal
-load("data/fitted_models_trunc_full.Rdata")
-trunc_bad <- models_bad
-trunc_good <- models_good
-
-#rm unneccesary data
-rm(models_bad)
-rm(models_good)
-
-#computing actual distributions ---------------------
-flat_images_good$visibles %>%
-  data.frame() %>% 
-  rename(v1 = X1, v2 = X2, v3 = X3, v4 = X4) %>%
-  group_by(v1, v2, v3, v4) %>%
-  summarise(prob = n()/nrow(flat_images_good$visibles)) -> distn_emp
-
-distn_good <- visible_distn(params = params_good)
-
-reshape_sample_distn <- function(model) {
-  sample_distn <- model$distn
-  dim(sample_distn) <- c(dim(sample_distn)[1]*dim(sample_distn)[2], dim(sample_distn)[3])
-  sample_distn %>% data.frame() -> sample_distn
-  names(sample_distn) <- names(distn_good)
-
-  sample_distn %>%
-    group_by(image_id) %>%
-    mutate(iter = 1:n()) -> sample_distn
-
-  return(sample_distn)
-}
-
-marginal_sample_good <- reshape_sample_distn(marginal_good)
-trick_sample_good <- reshape_sample_distn(trick_good)
-trunc_sample_good <- reshape_sample_distn(trunc_good)
-```
-
-```{r acf, fig.cap="\\label{fig:acf}The autocorrelation functions (ACF) for each image assessed at multiple lags for each method with the truncated Normal method in black and the marginalized likelihood method in red.
-As expected, ACF corresponding to the method that marginalizes out the hidden variables from the likelihood decreases to zero at a much faster rate, indicating better mixing for the chain."}
-marginal_sample_good %>% 
-  filter(iter > 50) %>%
-  group_by(v1, v2, v3, v4, image_id) %>% 
-  do(data.frame(acf = acf(.$prob, plot=FALSE)$acf,
-     lag = acf(.$prob, plot=FALSE)$lag)) -> marginal_acfs
-
-trunc_sample_good %>% 
-  filter(iter > 50) %>%
-  group_by(v1, v2, v3, v4, image_id) %>% 
-  do(data.frame(acf = acf(.$prob, plot=FALSE)$acf,
-                lag = acf(.$prob, plot=FALSE)$lag)) -> trunc_acfs
-
-marginal_acfs %>% ungroup() %>% select(-image_id) %>% rename(marginal = acf) %>%
-  left_join(trunc_acfs %>% rename(truncated = acf)) %>%
-  ggplot() +
-  geom_segment(aes(x=lag, xend=lag, y=0, yend=truncated), colour = "black") +
-  geom_segment(aes(x=lag, xend=lag, y=0, yend=marginal), colour = "red") +
-  facet_wrap(~image_id) +
-  xlab("Lag") +
-  ylab("ACF")
-
-```
-
-```{r M_eff, cache=TRUE}
-###
-### MBB Statistic
-###
-mbb_mean <- function(data_star, data, b) {
-  m <- length(data_star)
-  n <- length(data)
-  N <- n - b + 1
-
-  blockmeans <- rep(0, N)
-  for(i in 1:N) {
-    blockmeans[i] <- mean(data[i:(b + i - 1)])
-  }
-
-  sqrt(m)*(mean(data_star) - mean(blockmeans))
-}
-
-###
-### MBB Sample
-###
-mbb_sample <- function(data, b) {
-  n <- length(data)
-  k <- floor(n/b)
-
-  start <- sample.int(n - b + 1, k, replace = TRUE)
-  data_star <- NULL
-  for(i in 1:k) {
-    data_star <- c(data_star, data[start[i]:(start[i] + b - 1)])
-  }
-
-  return(data_star)
-}
-
-###
-### Block Bootstrap function
-###
-bb <- function(x, b, B, boot_stat, boot_sample) {
-  #ii. For blocksize b, generate B MBB versions of statistic
-  bb.stat <- array(NA, B)
-  for(i in seq_len(B)) {
-    bb.stat[i] <- boot_stat(boot_sample(x, b), x, b)
-  }
-  return(bb.stat)
-}
-
-marginal_sample_good %>% 
-  group_by(v1, v2, v3, v4, image_id) %>%
-  do(mbb = bb(.$prob, 2000^(1/3), 2000, mbb_mean, mbb_sample)) -> marginal_mbb
-
-trunc_sample_good %>% 
-  group_by(v1, v2, v3, v4, image_id) %>%
-  do(mbb = bb(.$prob, 5000^(1/3), 2000, mbb_mean, mbb_sample)) -> trunc_mbb
-
-marginal_mbb %>%
-  group_by(v1, v2, v3, v4, image_id) %>%
-  do(data.frame(C = var(.$mbb[[1]]))) %>%
-  left_join(marginal_sample_good %>% group_by(v1, v2, v3, v4) %>% summarise(sigma2 = var(prob))) %>%
-  mutate(M_eff = sigma2/C) %>%
-  ungroup() %>%
-  summarise(M_eff = min(M_eff)) %>% as.numeric() -> marginal_M_eff
-
-trunc_mbb %>%
-  group_by(v1, v2, v3, v4, image_id) %>%
-  do(data.frame(C = var(.$mbb[[1]]))) %>%
-  left_join(trunc_sample_good %>% group_by(v1, v2, v3, v4) %>% summarise(sigma2 = var(prob))) %>%
-  mutate(M_eff = sigma2/C) %>%
-  ungroup() %>%
-  summarise(M_eff = min(M_eff)) %>% as.numeric() -> trunc_M_eff
-```
-
-Secondly, we can assess the mixing of our chain using an idea of effective sample size.
-If the MCMC chain were truly iid draws from the target distribution, then the estimate of the posterior mean probability of image $i$, $\bar{p}^{(i)}$ would be distributed approximately Normal with mean the true posterior probability of image $i$, $p^{(i)}$, and variance $\sigma^2_i/M$, where $\sigma^2_i$ is the true posterior variance of image $i$ and $M$ is the length of the chain.
-However, with the presence of correlation in our chain, the aysmptotic variance of $\bar{p}^{(i)}$ is instead some $C_i/M$, where $C_i$ is some positive constant such that $C_i > \sigma^2_i$.
-We use a block bootstrap method with overlapping blocks to get an estimate for $C_i$, $\hat{C}_i$, and compare it to an estimate of $\sigma^2_i$ using the raw chain, $\hat{\sigma}^2_i$.
-Formally,
-
-$$
-M_{eff}^{(i)} = M\frac{\hat{\sigma}^2_i}{\hat{C}_i}.
-$$
-
-Using this metric, a chain of length $M = 1000$ for the marginalized likelihood method yields $\min_i M_{eff}^{(i)} = `r round(1000*marginal_M_eff, 2)`$, while the truncated Normal method only yields $\min_i M_{eff}^{(i)} = `r round(1000*trunc_M_eff, 2)`$.
-Thus, the truncated likelihood would require $`r round(marginal_M_eff/trunc_M_eff, 1)`$ times more iterations to be run of the MCMC chain in order to achieve the same amount of effective information about the posterior distribution.
-For this reason, in conjuction with the ACF results, the truncated Normal method is not an effective method for fitting the RBM.
-
-Figure \ref{fig:fitting_plot} shows the posterior probability of each possible $\boldsymbol v \in \{-1,1\}^4$ after fitting the RBM model in the two ways detailed in this section, excluding the truncated Normal method.
-The black vertical lines show the true probabilities of each image based on the parameters used to simulate the training set while the red vertical lines show the empirical distribution.
-From these posterior predictive checks, it is evident that the marginalized likelihood method produces the closest fit to the actual data.
-However, this method requires a marginalization step, which is infeasible for a model with $H$ of any real size.
-
-```{r posterior-dsn}
-marginal_sample_good %>% rename(marginal = prob) %>% filter(iter > 50) %>%
-  left_join(trick_sample_good %>% ungroup() %>% select(-image_id) %>% rename(trick = prob)) %>%
-  ungroup() %>%
-  mutate(image_id = paste0("image_", image_id)) %>% 
-  gather(method, prob, trick, marginal) %>%
-  spread(image_id, prob) -> all_statistics
-
-all_statistics %>%
-  gather(statistic, value, -iter, -method, -starts_with("v")) %>%
-  filter(grepl("image", statistic) & !is.na(value)) %>%
-  separate(statistic, into = c("junk", "statistic")) %>%
-  mutate(statistic = factor(statistic, labels = paste("Image", 1:length(unique(statistic))))) %>%
-  left_join(distn_emp %>% rename(prob_emp = prob) %>% right_join(distn_good %>% select(-image_id))) %>%
-  ggplot() +
-  geom_density(aes(value, y=..scaled.., colour = method, fill = method), alpha = .2) +
-  geom_vline(aes(xintercept = prob)) +
-  geom_vline(aes(xintercept = prob_emp), colour = "red") +
-  facet_wrap(~statistic, scales = "free_x") + 
-  ylab("Scaled Posterior Density") + xlab("Probability of Image") +
-  scale_colour_discrete("Method", labels = c("Marginalized likelihood", "Trick prior")) +
-  scale_fill_discrete("Method", labels = c("Marginalized likelihood", "Trick prior")) +
-  theme(legend.position = "bottom") -> p.models
-```
-
-```{r fitting_plot, fig.cap="\\label{fig:fitting_plot}Posterior probability of each possible 4-pixel image using two of the three Bayesian fitting techniques, trick prior and marginalized likelihood.
-The black vertical lines show the true probabilities of each image based on the parameters used to simulate the images while the red vertical lines show the empirical distribution.
-The marginalized likelihood method has the best fit for the data, however is also the most computationally intensive and least feasible with a real dataset.", fig.height=6}
-p.models
-```
-
-# Discussion
-
-**TODO: Add comparison to contrastive divergence**
-
-# Appendix
-
-## On the one-to-one correspondence between parameters and moments in the RBM distribution
-
-```{r child='appendix.Rmd'}
-```
-
-#References

### 2016-10-01 writing/draft.Rmd
body of instability rewrite complete

@@
 
-For RBMs, an analogous concept could concern the joint distribution of both hidden and visible nodes, where $N$ is the number of nodes in the model, $H + V$.
-However, instability in terms of only the visible nodes is a more useful measure of model impropriety.
-To see this, consider modeling images using an RBM.
-One would be interested in the maximum change in the likelihood when a single pixel is changed in the visibles as a proxy for instability in the model.
-So we concern ourselves with a notion of instability that deals with the maximum likelihood ratio of two vectors of visibles that differ at only one coordinate.
-
-\begin{align*}
-&\max\left\{\log\frac{P_{\boldsymbol \theta}(v_1, \dots, v_V)}{P_{\boldsymbol \theta}(v^*_1, \dots, v^*_V)}: (v_1, \dots, v_V) \text{ and } (v^*_1, \dots, v^*_V) \in \{-1, 1\}^V \text{ differ by one component} \right\} \\
-&\quad = \max\limits_{k \in 1, \dots, V}\left\{\log\frac{\sum\limits_{\boldsymbol h \in \{-1,1\}^H} \exp\left(
-\sum\limits_{i = 1}^V \theta_{v_i}v_i + \sum\limits_{j = 1}^H \theta_{h_j}h_j + \sum\limits_{i = 1}^V\sum\limits_{j = 1}^H \theta_{ij}v_i h_j\right)}{\sum\limits_{\boldsymbol h \in \{-1,1\}^H} \exp\left(
-\sum\limits_{i \not= k} \theta_{v_i}v_i + \sum\limits_{j = 1}^H \theta_{h_j}h_j + \sum\limits_{i \not=k}\sum\limits_{j = 1}^H \theta_{ij}v_i h_j + \theta_{v_k}v_k^* + v_k^*\sum\limits_{j = 1}^H \theta_{kj}h_j
-\right)}: v_k \not= v^*_k \right\} \\
-& \quad \ge \max\limits_{k \in 1, \dots, V}\left\{\log\frac{\max\limits_{\boldsymbol h \in \{-1,1\}^H}\exp\left(
-\sum\limits_{i = 1}^V \theta_{v_i}v_i + \sum\limits_{j = 1}^H \theta_{h_j}h_j + \sum\limits_{i = 1}^V\sum\limits_{j = 1}^H \theta_{ij}v_i h_j\right)}{2^H\max\limits_{\boldsymbol h \in \{-1,1\}^H}\exp\left(
-\sum\limits_{i \not= k} \theta_{v_i}v_i + \sum\limits_{j = 1}^H \theta_{h_j}h_j + \sum\limits_{i \not=k}\sum\limits_{j = 1}^H \theta_{ij}v_i h_j + \theta_{v_k}v_k^* + v_k^*\sum\limits_{j = 1}^H \theta_{kj}h_j
-\right)}: v_k \not= v^*_k \right\} \\
-&\quad = \max\limits_{k \in 1, \dots, V}\left\{\max\limits_{\boldsymbol h \in \{-1,1\}^H}\left\{
-\sum\limits_{i = 1}^V \theta_{v_i}v_i + \sum\limits_{j = 1}^H \theta_{h_j}h_j + \sum\limits_{i = 1}^V\sum\limits_{j = 1}^H \theta_{ij}v_i h_j\right\}\right. \\
-& \quad \quad \left. - \max\limits_{\boldsymbol h \in \{-1,1\}^H}
-\left\{\sum\limits_{i \not= k} \theta_{v_i}v_i + \sum\limits_{j = 1}^H \theta_{h_j}h_j + \sum\limits_{i \not=k}\sum\limits_{j = 1}^H \theta_{ij}v_i h_j + \theta_{v_k}v_k^* + v_k^*\sum\limits_{j = 1}^H \theta_{kj}h_j
-\right\} - H\log2: v_k \not= v^*_k \right\} \\
-&\quad \ge \max\limits_{\boldsymbol v \in \{-1,1\}^V}\max\limits_{\boldsymbol h \in \{-1,1\}^H}Q(\boldsymbol x) - \min\limits_{\boldsymbol v \in \{-1,1\}^V}\max\limits_{\boldsymbol h \in \{-1,1\}^H}Q(\boldsymbol x) - H\log 2
-\end{align*}
-
-Let $R(\boldsymbol \theta) \equiv \max_{\boldsymbol v} \max_{\boldsymbol h}Q(\boldsymbol x) - \min_{\boldsymbol v}\max_{\boldsymbol h}Q(\boldsymbol x) - H\log 2$.
-Then, if $R(\boldsymbol \theta) > C*V$, this implies that the maximum log-likelihood ratio of two observations that differ in only one element is greater than $C$.
-So, by examining $R(\boldsymbol\theta)/V$, we can determine a lower bound for the maximum log-likelihood ratio of two sets of observables that differ in only one element, which if large is consistent with the concept of instability detailed above.
+\begin{prop}
+\label{prop:instab}
+If $\frac{1}{V}\text{ELPR}_V(\boldsymbol \theta) > C$, then $\Delta_V(\boldsymbol \theta) > C$.
+\end{prop}
+
+In other words, if the quantity (\ref{eqn:elpr}) is too large, then the RBM model will exhibit large probability shifts for very small changes in the data configuration (i.e.
+instability mentioned earlier).
+
+Additionally, unstable RBM models are connected to degenerate models (placing all probability on a small pieace of the sample space). Define a model set 
+$$
+M_{\epsilon, \boldsymbol \theta} \equiv \left\{(v_1, \dots, v_V) \in \{-1,1\}^V: \log P_{\boldsymbol \theta}(v_1, \dots, v_V) > (1-\epsilon)\max\limits_{(v_1^*, \dots, v_V^*)}P_{\boldsymbol \theta}(v_1^*, \dots, v_V^*) + \epsilon\min\limits_{(v_1^*, \dots, v_V^*)}P_{\boldsymbol \theta}(v_1^*, \dots, v_V^*) \right\}
+$$
+of possible outcomes, for a given $0 < \epsilon < 1$.
+
+\begin{prop}
+\label{prop:degen}
+For an unstable RBM model, and for any given $0 < \epsilon < 1$, $P_{\boldsymbol \theta}\left((v_1, \dots, v_V) \in M_{\epsilon, \boldsymbol \theta}\right) \rightarrow 1$ holds as $V \rightarrow \infty$.
+\end{prop}
+
+In other words, as a consequence of unstable models, all probability will stack up on mode sets or potentially those few outcomes with the highest probability.
+Proofs of both proposition \ref{prop:instab} and proposition \ref{prop:degen} are provided in appendix \ref{instability-proofs}.
+Propositions \ref{prop:instab}-\ref{prop:degen} have counterpart results in @schweinberger2011instability, but, unlike results there, we do not consider exponential family forms with a fixednumber of parameters (i.e.
+our results are broader for a more complicated model form for the visibles).
+Essentially, with our generalized definition of *unstable* models, one can establish instability results as in @schweinberger2011instability under assumptions that the sample space is finite and all model outcomes have a non-zero probability.
+
+To later numerically investigate unstable RBM models, it is useful to define a computationally tractable lower bound on (\ref{eqn:elpr}) given by
+$$
+\frac{1}{V}R(\boldsymbol \theta) \text{ for } R(\boldsymbol \theta) \equiv \max_{\boldsymbol v} \max_{\boldsymbol h}Q(\boldsymbol x) - \min_{\boldsymbol v}\max_{\boldsymbol h}Q(\boldsymbol x) - H\log 2.
+$$
+That is, it holds that, in (\ref{eqn:elpr}),
+\begin{align}
+\frac{1}{V} \text{ELPR}_V(\boldsymbol \theta) \equiv \frac{1}{V} \log \left[\frac{\max\limits_{(v_1, \dots, v_V) \in \{-1,1\}^V}P_{\boldsymbol \theta}(v_1, \dots, v_V)}{\min\limits_{(v_1, \dots, v_V) \in \{-1,1\}^V}P_{\boldsymbol \theta}(v_1, \dots, v_V)}\right] \ge \frac{1}{V}R(\boldsymbol \theta).
+\label{eqn:lowerbnd}
+\end{align}
+(A proof of (\ref{eqn:lowerbnd}) appears in appendix \ref{instability-proofs}.)
+
+The benefit of (\ref{eqn:lowerbnd}) is that $R(\boldsymbol \theta)/V$ is simple to compute (i.e.
+compared to (\ref{eqn:elpr}), $R(\boldsymbol \theta)$ requires no summation over hidden variables) and that propositions \ref{prop:instab}-\ref{prop:degen} would continue to hold upon replcing (\ref{eqn:elpr}) with $R(\boldsymbol \theta)/V$, due to (\ref{eqn:lowerbnd}).
+Hence, larger values of $R(\boldsymbol \theta)/V$ indicate unstable RBM models.
 
@@
 
-```{r child='appendix.Rmd'}
+```{r child='appendix-center.Rmd'}
+```
+
+## Instability proofs
+
+```{r child='appendix-instability.Rmd'}
 ```

### 2016-10-01 writing/appendix-center.Rmd
cleanup center of universe appendix

--- 2016-09-17 writing/appendix.Rmd
@@
 1.
-The function $\bm{g}: \mathbb{R}^{m+n+mn} \rightarrow \mathcal{I}$ is one-to-one; see Theorem 1 below.
+The function $\bm{g}: \mathbb{R}^{m+n+mn} \rightarrow \mathcal{I}$ is one-to-one; see Theorem \ref{thm:one} below.
 Hence, these moments uniquely characterize the parameters in the RBM$(m,n,\bm{\theta})$ model.
@@
 The center of the image $\mathcal{I}$ (the zero vector in $\mathbb{R}^{m+n+mn}$) is given only by $\bm{\theta}=\bm{0}$ (the zero vector in the parameter space).
-This corresponds to the case where all the random vectors $V_1,\ldots,V_m,H_1,\ldots,H_n$ are iid with a symmetric Bernoulli$(1/2)$ distribution; see Theorem 2.
+This corresponds to the case where all the random vectors $V_1,\ldots,V_m,H_1,\ldots,H_n$ are iid with a symmetric Bernoulli$(1/2)$ distribution; see Theorem \ref{thm:two}.
 
-\newtheorem{theoremnew4}{Theorem}
-\begin{theoremnew4} Let $m,n\geq 1$ and suppose the random vector   $(\bm{V}_m,\bm{H}_n) = (V_1,\ldots,V_m, H_1,\ldots,H_n)$ follows a RBM$(m,n,\bm{\theta})$ distribution.
+\begin{apptheorem} 
+\label{thm:one}
+Let $m,n\geq 1$ and suppose the random vector   $(\bm{V}_m,\bm{H}_n) = (V_1,\ldots,V_m, H_1,\ldots,H_n)$ follows a RBM$(m,n,\bm{\theta})$ distribution.
 Let $\bm{\theta}_1,\bm{\theta}_2 \in \mathbb{R}^{m+n+mn}$ and
@@
 \end{enumerate}
-\end{theoremnew4}
+\end{apptheorem}
 
-**Proof of Theorem 1.** We first establish the equivalence of claims 2 and 3. If claim 2 holds, then
+{\bf Proof of Theorem \ref{thm:one}.} We first establish the equivalence of claims 2 and 3. If claim 2 holds, then
  claim 3 holds trivially.
@@
 for any $\{v_i\}_{i=1,i\neq \ell}^m,\{h_j\}_{j=1}^n \subset\{\pm 1\}$.
-Equating exponents, we have $\theta_{v_\ell} - \tilde{\theta}_{v_\ell} = \sum_{j=1}^n  h_j (  \tilde{\theta}_{\ell j} -\theta{\ell j})$ for any $\{h_j\}_{j=1}^n \subset \{\pm 1\}$.
+Equating exponents, we have $\theta_{v_\ell} - \tilde{\theta}_{v_\ell} = \sum_{j=1}^n  h_j (  \tilde{\theta}_{\ell j} -\theta_{\ell j})$ for any $\{h_j\}_{j=1}^n \subset \{\pm 1\}$.
 Taking all $h_j=1$ or all $h_j=-1$ for $1 \leq j \leq n$ yields $\theta_{v_\ell} = \tilde{\theta}_{v_\ell}$.
 This result further implies $\tilde{\theta}_{\ell 1} -\theta{\ell 1} = -  \sum_{j=2}^n  h_j (  \tilde{\theta}_{\ell j} -\theta_{\ell j})$ for
-any given $\{h_j\}_{j=2}^n \subset \{\pm 1\}$, so that $\theta{\ell 1}=\tilde{\theta}_{\ell 1}$ must hold.
-Iterating this argument, sequentially for each $j=2,\ldots,m$, shows $\theta{\ell j}=\tilde{\theta}_{\ell j}$ for any $1 \leq j \leq n$ in addition to $\theta_{v_\ell}=\tilde{\theta}_{v_\ell}$.
+any given $\{h_j\}_{j=2}^n \subset \{\pm 1\}$, so that $\theta_{\ell 1}=\tilde{\theta}_{\ell 1}$ must hold.
+Iterating this argument, sequentially for each $j=2,\ldots,m$, shows $\theta_{\ell j}=\tilde{\theta}_{\ell j}$ for any $1 \leq j \leq n$ in addition to $\theta_{v_\ell}=\tilde{\theta}_{v_\ell}$.
 As $\ell\in\{1,\ldots,m\}$ was arbitrary, we have $\theta_{v_i}=\tilde{\theta}_{v_i}$ and $\theta{i j}=\tilde{\theta}_{i j}$ for any $1 \leq i \leq m$, $1 \leq j \leq n$.
@@
 
-To establish the equivalence of claims 1 and 3 in Theorem 1 under the RBM$(m,n,\bm{\theta})$ model, we note that claim 3 easily implies claim 1.
+To establish the equivalence of claims 1 and 3 in Theorem \ref{thm:one} under the RBM$(m,n,\bm{\theta})$ model, we note that claim 3 easily implies claim 1.
 Hence, it suffices now to establish claim 3 from claim 1.
-This follows immediately from Lemma 1 below and the fact that the variables $V_1,\ldots,V_m$ are independent given $H_1,\ldots,H_n$ in the RBM$(m,n,\bm{\theta})$ model (and likewise $H_1,\ldots,H_n$ are independent given $V_1,\ldots,V_m$).
+This follows immediately from Lemma \ref{lemma:one} below and the fact that the variables $V_1,\ldots,V_m$ are independent given $H_1,\ldots,H_n$ in the RBM$(m,n,\bm{\theta})$ model (and likewise $H_1,\ldots,H_n$ are independent given $V_1,\ldots,V_m$).
 
-\begin{lemma} Let $m,n\geq 1$.
+\begin{applemma} 
+\label{lemma:one}
+Let $m,n\geq 1$.
 Let $P_1,P_2$ denote two probability distributions for a random vector $(\bm{V}_m,\bm{H}_n) = (V_1,\ldots,V_m, H_1,\ldots,H_n)$ supported on $\{a,b\}^{m+n}$, for some real constants $a\neq b \in\mathbb{R}$, such that $P_i( \bm{V}_m=\bm{v}_m, \bm{H}_n=\bm{h}_n)>0$, $i=1,2$, for any $\bm{v}_m \in \{a,b\}^{m}$, $\bm{h}_n \in \{a,b\}^{n}$.
@@
 follows if $\E_1 V_i = \E_2 V_i$, $\E_1 H_j = \E_2 H_j$, $\E_1 V_i H_j = \E_2 V_j H_j$ holds for all $i\in\{1,\ldots,m\}$, $j\in \{1,\ldots,n\}$.
-\end{lemma}
+\end{applemma}
 
-*Proof.* The case $m=n=1$ follows from Lemma 2.
-We now use an induction argument, assuming that the result of Lemma 1 holds for some given order of $(m,n)$ with $m,n\geq 1$ and show the the result continues to hold for $(m+1,n)$ or $(m,n+1)$.
+{\bf Proof.} The case $m=n=1$ follows from Lemma \ref{lemma:two}.
+We now use an induction argument, assuming that the result of Lemma \ref{lemma:one} holds for some given order of $(m,n)$ with $m,n\geq 1$ and show the the result continues to hold for $(m+1,n)$ or $(m,n+1)$.
 We shall treat the case $(m+1,n)$ (where the other case $(m,n+1)$ follows by symmetrical arguments).
@@
 \end{eqnarray*}
-by $P_1(\bm{H}_n=\bm{h}_n)=P_2(\bm{H}_n=\bm{h}_n)$. As $\bm{v}_{m+1}\in \{a,b\}^{m+1},\bm{h}_n \in \{a,b\}^{n}$ were arbitrary, this completes the proof of Lemma 1. $\Box$
+by $P_1(\bm{H}_n=\bm{h}_n)=P_2(\bm{H}_n=\bm{h}_n)$. As $\bm{v}_{m+1}\in \{a,b\}^{m+1},\bm{h}_n \in \{a,b\}^{n}$ were arbitrary, this completes the proof of Lemma \ref{lemma:one}. $\Box$
 
-\begin{lemma} Suppose the discrete random vector $(X,Y)$ has support $\{a,b\}\times \{a,b\}$, for some $a\neq b \in \mathbb{R}$,
+\begin{applemma} 
+\label{lemma:two}
+Suppose the discrete random vector $(X,Y)$ has support $\{a,b\}\times \{a,b\}$, for some $a\neq b \in \mathbb{R}$,
 under two probability distributions $P_1$ and $P_2$. Let $\E_i$ denote expectation under the distribution $P_i$, $i=1,2$. Then, the following are equivalent
@@
 \end{enumerate}
-\end{lemma}
+\end{applemma}
 
-*Proof.* Denote the four probabilities by which the random vector $(X,Y)$ assume pairs $(b,b),(b,a),(a,b),(a,a)$, respectively, as $c_1,c_2,c_3,1-c_1-c_2-c_3$ under $P_1$ and $d_1,d_2,d_3,1-d_1-d_2-d_3$ under $P_2$.
+{\bf Proof.} Denote the four probabilities by which the random vector $(X,Y)$ assume pairs $(b,b),(b,a),(a,b),(a,a)$, respectively, as $c_1,c_2,c_3,1-c_1-c_2-c_3$ under $P_1$ and $d_1,d_2,d_3,1-d_1-d_2-d_3$ under $P_2$.
 The conditions $\E_1 X = \E_2 X$, $\E_1 Y = \E_2 Y$, $\E_1 X Y = \E_2 X Y$ with $b\neq a$ imply that
@@
 As $(2c_1+c_2+c_3) = (2d_1+d_2+d_3)$ and $b\neq a$, we conclude that $c_1=d_1$, from which it follows that $c_2=d_2$ and $c_3=d_3$.
-Hence, claim 1 implies claim 2 in Lemma 2.
+Hence, claim 1 implies claim 2 in Lemma \ref{lemma:two}.
 Claim 2 also trivially implies claim 1.
@@
 
-\begin{theoremnew4} Let $m,n\geq 1$ and suppose the random vector $(\bm{V}_m,\bm{H}_n) = (V_1,\ldots,V_m, H_1,\ldots,H_n)$ follows a RBM$(m,n,\bm{\theta})$ distribution.
-Then, the following are equivalent
+\begin{apptheorem} 
+\label{thm:two}
+Let $m,n\geq 1$ and suppose the random vector $(\bm{V}_m,\bm{H}_n) = (V_1,\ldots,V_m, H_1,\ldots,H_n)$ follows a RBM$(m,n,\bm{\theta})$ distribution.  Then, the following are equivalent
 \begin{enumerate}
@@
 \end{enumerate}
-\end{theoremnew4}
+\end{apptheorem}
 
-**Proof of Theorem 2.**  We first establish the equivalence of claims 2 and 3.
+{\bf Proof of Theorem \ref{thm:two}.}  We first establish the equivalence of claims 2 and 3.
 If claim 2 holds, then claim 3 follows easily from the resulting uniform cell probabilities: $P(V_1= v_1,\ldots,V_m= v_m, H_1= h_1,\ldots,H_n=h_n|\bm{\theta})=2^{-mn}$ for any $v_i,h_j\in\{\pm 1\}$, $i=1,\ldots,m$, $j=1,\ldots,n$.
@@
 
-To establish the equivalence of claims 1 and 2 in Theorem 2 under the RBM$(m,n,\bm{\theta})$ model, we note that claim 2 again implies claim 3, which then easily implies claim 1.
-By Theorem 1, if claim 1 holds then the only possibility for this is $\bm{\theta} = \bm{0} \in \mathbb{R}^{m+n+mn}$.
+To establish the equivalence of claims 1 and 2 in Theorem \ref{thm:two} under the RBM$(m,n,\bm{\theta})$ model, we note that claim 2 again implies claim 3, which then easily implies claim 1.
+By Theorem \ref{thm:one}, if claim 1 holds then the only possibility for this is $\bm{\theta} = \bm{0} \in \mathbb{R}^{m+n+mn}$.
 $\Box$

### 2016-10-01 writing/appendix-instability.Rmd
add proofs for propositions

@@
+{\bf Proof of Proposition \ref{prop:instab}.} We prove the contrapositive.
+Suppose that $\Delta_V(\boldsymbol \theta) \le C$ holds for some $C > 0$.
+Under the RBM model fo visibles, $P_{\boldsymbol \theta}(\boldsymbol v) > 0$ holds for each outcome $\boldsymbol v \in \{-1,1\}^V$.
+Let $\boldsymbol v_{min} \equiv \argmin\limits_{\boldsymbol v \in \{-1, 1\}^V}P_{\boldsymbol \theta}(\boldsymbol v)$ and $\boldsymbol v_{max} \equiv \argmax\limits_{\boldsymbol v \in \{-1, 1\}^V}P_{\boldsymbol \theta}(\boldsymbol v)$.
+Note there exists a sequence $\boldsymbol v_{min} \equiv \boldsymbol v_0, \boldsymbol v_1, \dots, \boldsymbol v_k \equiv \boldsymbol v_{max}$ in $\{-1,1\}^V$ of component-wise switches to move from $\boldsymbol v_{min}$ to $\boldsymbol v_{max}$ in the sample space (i.e.
+$\boldsymbol v_i, \boldsymbol v_{i + 1} \in \{-1,1\}^V$ differ by exactly $1$ component for $i = 0, \dots, k -1$) for some integer $k \in \{0, 1, \dots, V\}$.
+Then
+\begin{align*}
+\log\left[\frac{P_{\boldsymbol \theta}(\boldsymbol v_{max})}{P_{\boldsymbol \theta}(\boldsymbol v_{min})}\right] &= \left|\sum\limits_{i = 1}^k\log\left(\frac{P_{\boldsymbol \theta}(\boldsymbol v_i)}{P_{\boldsymbol \theta}(\boldsymbol v_{i-1})}\right)\right| \\
+& \le \sum\limits_{i = 1}^k\left|\log\left(\frac{P_{\boldsymbol \theta}(\boldsymbol v_i)}{P_{\boldsymbol \theta}(\boldsymbol v_{i-1})}\right)\right| \\
+& \le k \Delta_V(\boldsymbol \theta) \le VC
+\end{align*}
+using $k \le V$ and $\Delta_V(\boldsymbol \theta) \le C$.$\Box$
+
+{\bf Proof of Proposition \ref{prop:degen}.} Define $\boldsymbol v_{max}$ and $\boldsymbol v_{min}$ as in the proof of Proposition \ref{prop:instab}.
+Fix $0 < \epsilon < 1$ Then, $\boldsymbol x_{max} \in M_{\epsilon, \boldsymbol \theta}$, so $P_{\boldsymbol \theta}(M_{\epsilon, \boldsymbol \theta}) \ge P_{\boldsymbol \theta}(\boldsymbol v_{max})$.
+If $\boldsymbol v \in \{-1, 1\}^V \setminus M_{\epsilon, \boldsymbol \theta}$, then by definition $P_{\boldsymbol \theta}(\boldsymbol v) \le [P_{\boldsymbol \theta}(\boldsymbol v_{max})]^{1-\epsilon}[P_{\boldsymbol \theta}(\boldsymbol v_{min})]^{\epsilon}$ holds so that
+\begin{align*}
+1-P_{\boldsymbol \theta}(M_{\epsilon, \boldsymbol \theta}) &= P_{\boldsymbol \theta}(M_{\epsilon, \boldsymbol \theta}^C) \\
+& = \sum\limits_{\boldsymbol v \in \{-1, 1\}^V \setminus M_{\epsilon, \boldsymbol \theta}}P_{\boldsymbol \theta}(\boldsymbol v) \\
+& \le (2^V)[P_{\boldsymbol \theta}(\boldsymbol v_{max})]^{1-\epsilon}[P_{\boldsymbol \theta}(\boldsymbol v_{min})]^{\epsilon}
+\end{align*}
+Then,
+\begin{align*}
+\frac{1}{V}\log\left[\frac{P_{\boldsymbol \theta}(M_{\epsilon, \boldsymbol \theta})}{1-P_{\boldsymbol \theta}(M_{\epsilon, \boldsymbol \theta})}\right] & \ge \frac{1}{V} \log\left[\frac{P_{\boldsymbol\theta}(\boldsymbol v_{max})}{(2^V)[P_{\boldsymbol \theta}(\boldsymbol v_{max})]^{1-\epsilon}[P_{\boldsymbol \theta}(\boldsymbol v_{min})]^{\epsilon}}\right] \\
+&= \frac{\epsilon}{V} \log\left[\frac{P_{\boldsymbol \theta}(\boldsymbol v_{max})}{P_{\boldsymbol \theta}(\boldsymbol v_{min})}\right] - \log 2 \rightarrow \infty
+\end{align*}
+as $V \rightarrow \infty$ by the definition of an unstable RBM model.$\Box$
+
 \begin{align*}

### 2016-10-01 writing/appendix-instability.Rmd
fix references and alter last instability proof

@@
 
+{\bf Proof of (\ref{eqn:lowerbnd}).} With algebra, we have
+
 \begin{align*}
-&\max\left\{\log\frac{P_{\boldsymbol \theta}(v_1, \dots, v_V)}{P_{\boldsymbol \theta}(v^*_1, \dots, v^*_V)}: (v_1, \dots, v_V) \text{ and } (v^*_1, \dots, v^*_V) \in \{-1, 1\}^V \text{ differ by one component} \right\} \\
-&\quad = \max\limits_{k \in 1, \dots, V}\left\{\log\frac{\sum\limits_{\boldsymbol h \in \{-1,1\}^H} \exp\left(
-\sum\limits_{i = 1}^V \theta_{v_i}v_i + \sum\limits_{j = 1}^H \theta_{h_j}h_j + \sum\limits_{i = 1}^V\sum\limits_{j = 1}^H \theta_{ij}v_i h_j\right)}{\sum\limits_{\boldsymbol h \in \{-1,1\}^H} \exp\left(
-\sum\limits_{i \not= k} \theta_{v_i}v_i + \sum\limits_{j = 1}^H \theta_{h_j}h_j + \sum\limits_{i \not=k}\sum\limits_{j = 1}^H \theta_{ij}v_i h_j + \theta_{v_k}v_k^* + v_k^*\sum\limits_{j = 1}^H \theta_{kj}h_j
-\right)}: v_k \not= v^*_k \right\} \\
-& \quad \ge \max\limits_{k \in 1, \dots, V}\left\{\log\frac{\max\limits_{\boldsymbol h \in \{-1,1\}^H}\exp\left(
-\sum\limits_{i = 1}^V \theta_{v_i}v_i + \sum\limits_{j = 1}^H \theta_{h_j}h_j + \sum\limits_{i = 1}^V\sum\limits_{j = 1}^H \theta_{ij}v_i h_j\right)}{2^H\max\limits_{\boldsymbol h \in \{-1,1\}^H}\exp\left(
-\sum\limits_{i \not= k} \theta_{v_i}v_i + \sum\limits_{j = 1}^H \theta_{h_j}h_j + \sum\limits_{i \not=k}\sum\limits_{j = 1}^H \theta_{ij}v_i h_j + \theta_{v_k}v_k^* + v_k^*\sum\limits_{j = 1}^H \theta_{kj}h_j
-\right)}: v_k \not= v^*_k \right\} \\
-&\quad = \max\limits_{k \in 1, \dots, V}\left\{\max\limits_{\boldsymbol h \in \{-1,1\}^H}\left\{
-\sum\limits_{i = 1}^V \theta_{v_i}v_i + \sum\limits_{j = 1}^H \theta_{h_j}h_j + \sum\limits_{i = 1}^V\sum\limits_{j = 1}^H \theta_{ij}v_i h_j\right\}\right. \\
-& \quad \quad \left. - \max\limits_{\boldsymbol h \in \{-1,1\}^H}
-\left\{\sum\limits_{i \not= k} \theta_{v_i}v_i + \sum\limits_{j = 1}^H \theta_{h_j}h_j + \sum\limits_{i \not=k}\sum\limits_{j = 1}^H \theta_{ij}v_i h_j + \theta_{v_k}v_k^* + v_k^*\sum\limits_{j = 1}^H \theta_{kj}h_j
-\right\} - H\log2: v_k \not= v^*_k \right\} \\
-&\quad \ge \max\limits_{\boldsymbol v \in \{-1,1\}^V}\max\limits_{\boldsymbol h \in \{-1,1\}^H}Q(\boldsymbol x) - \min\limits_{\boldsymbol v \in \{-1,1\}^V}\max\limits_{\boldsymbol h \in \{-1,1\}^H}Q(\boldsymbol x) - H\log 2
+&\log\left[\frac{\max\limits_{(v_1, \dots, v_V) \in \{-1, 1\}^V}P_{\boldsymbol \theta}(v_1, \dots, v_V)}{\min\limits_{(v_1, \dots, v_V) \in \{-1, 1\}^V}P_{\boldsymbol \theta}(v_1, \dots, v_V)} \right] = \\
+& \qquad \log\left[\frac{\max\limits_{(v_1, \dots, v_V)}\sum\limits_{\boldsymbol h \in \{-1,1\}^H} \exp\left(\sum\limits_{i = 1}^V \theta_{v_i}v_i + \sum\limits_{j = 1}^H \theta_{h_j}h_j + \sum\limits_{i = 1}^V\sum\limits_{j = 1}^H \theta_{ij}v_i h_j\right)}
+{\min\limits_{(v_1, \dots, v_V)}\sum\limits_{\boldsymbol h \in \{-1,1\}^H} \exp\left(\sum\limits_{i = 1}^V \theta_{v_i}v_i + \sum\limits_{j = 1}^H \theta_{h_j}h_j + \sum\limits_{i = 1}^V\sum\limits_{j = 1}^H \theta_{ij}v_i h_j\right)}\right] \\
+& \qquad \ge \log\frac{\max\limits_{(v_1, \dots, v_V)}\max\limits_{\boldsymbol h \in \{-1,1\}^H}\exp\left(\sum\limits_{i = 1}^V \theta_{v_i}v_i + \sum\limits_{j = 1}^H \theta_{h_j}h_j + \sum\limits_{i = 1}^V\sum\limits_{j = 1}^H \theta_{ij}v_i h_j\right)}
+{\min\limits_{(v_1, \dots, v_V)}2^H\max\limits_{\boldsymbol h \in \{-1,1\}^H}\exp\left(\sum\limits_{i = 1}^V \theta_{v_i}v_i + \sum\limits_{j = 1}^H \theta_{h_j}h_j + \sum\limits_{i = 1}^V\sum\limits_{j = 1}^H \theta_{ij}v_i h_j\right)} \\
+& \qquad = \max\limits_{(v_1, \dots, v_V)}\max\limits_{\boldsymbol h \in \{-1,1\}^H}\left\{\sum\limits_{i = 1}^V \theta_{v_i}v_i + \sum\limits_{j = 1}^H \theta_{h_j}h_j + \sum\limits_{i = 1}^V\sum\limits_{j = 1}^H \theta_{ij}v_i h_j\right\} - \\
+& \qquad \qquad \min\limits_{(v_1, \dots, v_V)}\max\limits_{\boldsymbol h \in \{-1,1\}^H}\left\{\sum\limits_{i = 1}^V \theta_{v_i}v_i + \sum\limits_{j = 1}^H \theta_{h_j}h_j + \sum\limits_{i = 1}^V\sum\limits_{j = 1}^H \theta_{ij}v_i h_j\right\} -H\log2 \\
+& \qquad \equiv R(\boldsymbol \theta)
 \end{align*}
+using above the monotonicity of $\log(\cdot)$.

### 2016-10-01 writing/draft.Rmd
cleanup center of universe appendix
fix references and alter last instability proof

@@
 We discuss three Bayesian fitting techniques with the goal of avoiding model impropriety in section \ref{bayesian-model-fitting} and wrap up with a discussion in section \ref{discussion}.
-Additionally, there is an accompanying appendix with auxiliary proofs relating to the choice of data encoding in section \ref{appendix}.
+Additionally, there is an accompanying appendix with auxiliary proofs relating to the choice of data encoding and propositions on instability in section \ref{appendix}.
 
@@
 In addition to the argument that the $-1/1$ data encoding mitigates some perceived prevalence of near-degeneracy it also has the benefit of providing a guaranteed non-degenerate model at $\boldsymbol \theta = \boldsymbol 0 \in \mathbb{R}^{H + V + H*V}$.
-The proof of this is included in appendix \ref{on-the-one-to-one-correspondence-between parameters-and moments-in-the-rbm-distribution}.
+The proof of this is included in appendix \ref{center-of-universe}.
 And while from some computing perspectives $0/1$ coding might seem most natural, the $-1/1$ coding is far more convenient and interpretable from the point of view of statistical modeling, where it makes parameters simply interpreted in terms of symmetrically defined main effects and interactions.
@@
 
-# Appendix
-
-## On the one-to-one correspondence between parameters and moments in the RBM distribution
+\begin{appendix}
+\section{Appendices}
+\label{appendix}
+\newtheorem{apptheorem}{Theorem}[section]
+\newtheorem{applemma}{Lemma}[section]
+
+\subsection{On the one-to-one correspondence between parameters and moments in the RBM distribution}
+\label{center-of-universe}
 
@@
 
-## Instability proofs
+\subsection{Instability proofs}
+\label{instability-proofs}
 
@@
 
+\end{appendix}
+
 #References

### 2016-10-04 writing/appendix-instability.Rmd
small typos and start note

@@
 Suppose that $\Delta_V(\boldsymbol \theta) \le C$ holds for some $C > 0$.
-Under the RBM model fo visibles, $P_{\boldsymbol \theta}(\boldsymbol v) > 0$ holds for each outcome $\boldsymbol v \in \{-1,1\}^V$.
+Under the RBM model for visibles, $P_{\boldsymbol \theta}(\boldsymbol v) > 0$ holds for each outcome $\boldsymbol v \in \{-1,1\}^V$.
 Let $\boldsymbol v_{min} \equiv \argmin\limits_{\boldsymbol v \in \{-1, 1\}^V}P_{\boldsymbol \theta}(\boldsymbol v)$ and $\boldsymbol v_{max} \equiv \argmax\limits_{\boldsymbol v \in \{-1, 1\}^V}P_{\boldsymbol \theta}(\boldsymbol v)$.
 Note there exists a sequence $\boldsymbol v_{min} \equiv \boldsymbol v_0, \boldsymbol v_1, \dots, \boldsymbol v_k \equiv \boldsymbol v_{max}$ in $\{-1,1\}^V$ of component-wise switches to move from $\boldsymbol v_{min}$ to $\boldsymbol v_{max}$ in the sample space (i.e.
-$\boldsymbol v_i, \boldsymbol v_{i + 1} \in \{-1,1\}^V$ differ by exactly $1$ component for $i = 0, \dots, k -1$) for some integer $k \in \{0, 1, \dots, V\}$.
+$\boldsymbol v_i, \boldsymbol v_{i + 1} \in \{-1,1\}^V$ differ by exactly $1$ component for $i = 0, \dots, k$) for some integer $k \in \{0, 1, \dots, V\}$.
 Then
@@
 {\bf Proof of Proposition \ref{prop:degen}.} Define $\boldsymbol v_{max}$ and $\boldsymbol v_{min}$ as in the proof of Proposition \ref{prop:instab}.
-Fix $0 < \epsilon < 1$ Then, $\boldsymbol x_{max} \in M_{\epsilon, \boldsymbol \theta}$, so $P_{\boldsymbol \theta}(M_{\epsilon, \boldsymbol \theta}) \ge P_{\boldsymbol \theta}(\boldsymbol v_{max})$.
+Fix $0 < \epsilon < 1$ Then, $\boldsymbol v_{max} \in M_{\epsilon, \boldsymbol \theta}$, so $P_{\boldsymbol \theta}(M_{\epsilon, \boldsymbol \theta}) \ge P_{\boldsymbol \theta}(\boldsymbol v_{max})$.
 If $\boldsymbol v \in \{-1, 1\}^V \setminus M_{\epsilon, \boldsymbol \theta}$, then by definition $P_{\boldsymbol \theta}(\boldsymbol v) \le [P_{\boldsymbol \theta}(\boldsymbol v_{max})]^{1-\epsilon}[P_{\boldsymbol \theta}(\boldsymbol v_{min})]^{\epsilon}$ holds so that

### 2016-10-04 writing/draft.Rmd
warning not warnings
small typos and start note

@@
 
-opts_chunk$set(echo=FALSE, message=FALSE, warnings=FALSE)
+opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE)
 theme_set(theme_bw(base_family="serif"))
@@
 $$
-the biggest log-probability ratio for a one-pixed (one component) change in data outcomes (visibles). We then have the following result.
+the biggest log-probability ratio for a one-pixel (one component) change in data outcomes (visibles). We then have the following result.
 

### 2016-10-04 writing/note.Rmd
small typos and start note

---
title: "A note on the instability and degeneracy of deep learning models"
author: "Andee Kaplan, Daniel Nordman, Stephen Vardeman"
date: "October, 2016"
linestretch: 2
output: 
  pdf_document:
    fig_caption: yes
    includes:
      in_header: header.tex
    number_sections: yes
bibliography: references/refs.bib
abstract: |
---

# Discrete Expanding Finite space models and notions of instability

Consider $(X_1, \dots X_N)$, a set of discrete random variables with a finite sample space, $\mathcal{X}^N$. Let $P_{\boldsymbol \theta_N}(x_1, \dots, x_N)$ denote the probability of the data outcome $(x_1, \dots, x_N) \in \mathcal{X}^N$, which is a sequence of expanding models that potentially grow size as the dimension of the data become larger (i.e. $\boldsymbol \theta_N \in \mathbb{R}^{q(N)}$, where $q(N)$ is a function of $N$). Further, the model support of $P_{\boldsymbol \theta_N}(x_1, \dots, x_N)$ is equal to the sample space $\mathcal{X}^N$. We will call this class of models *Discrete Expanding Finite space*, or DEF models.

To quantify "instability" in a model with discrete outcomes, it is useful to imagine how a data model might be expanded to incorporate more and more observations. A relevant quantity to this end is a (scaled) extremal log-probability ratio given by 
\begin{align}
\frac{1}{N} \log \left[\frac{\max\limits_{(x_1, \dots, x_N) \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(x_1, \dots, x_N)}{\min\limits_{(x_1, \dots, x_N) \in \mathcal{X}^N}P_{\boldsymbol \theta}(x_1, \dots, x_N)}\right] \equiv \frac{1}{V} \text{ELPR}_N(\boldsymbol \theta_N) \label{eqn:elpr}
\end{align}

Essentially, in the process of developing a model to handle increasing numbers of variable (i.e., $N \rightarrow \infty$), we want (\ref{eqn:elpr}) to stay bounded, meaning that the largest probability under this model should maintain a magnitude within the smallest probability allowed under the model. This leads to the following definition:

\begin{defn}[unstable DEF]
A DEF model formulation is \emph{unstable} if, as the number of variables increase ($N \rightarrow \infty$), it holds that 
\begin{align*}
\lim\limits_{N \rightarrow \infty} \frac{1}{N} \log \left[\frac{\max\limits_{(x_1, \dots, x_N) \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(x_1, \dots, x_N)}{\min\limits_{(x_1, \dots, x_N) \in \mathcal{X}^N}P_{\boldsymbol \theta}(x_1, \dots, x_N)}\right] = \infty.
\end{align*}
\end{defn}
In other words, given any $C > 0$, there exists an integer $N_C > 0$ so that $\frac{1}{N}\text{ELPR}_N(\boldsymbol \theta_N) > C$ for all $N \ge N_C$.

Note that this definition of *unstable* is a generalization of that in @schweinberger2011instability allowing nonexponential family models and an increasing number of parameters (while this definition appears different, it does match that in @schweinberger2011instability for the special exponential models considered there, see Section \ref{discrete-exponential-family-models} for details).

Unstable DEF models are undesirable for several reasons. One is that small changes in data can lead to overly-sensitive changes in probability. Consider, for example, 
$$
\Delta_N(\boldsymbol \theta_N) \equiv \max \left\{\log \frac{P_{\boldsymbol \theta_N}(x_1, \dots, x_N)}{P_{\boldsymbol \theta_N}(x_1^*, \dots, x_N^*)} : (x_1, \dots, x_N) \& (x_1^*, \dots, x_N^*) \in \mathcal{X}^N \text{ differ by exactly one component}\right\},
$$
the biggest log-probability ratio for a one component change in data outcomes. We then have the following result.

\begin{prop}
\label{prop:instab}
If $\frac{1}{N}\text{ELPR}_N(\boldsymbol \theta_N) > C$, then $\Delta_N(\boldsymbol \theta_N) > C$.
\end{prop}

In other words, if the quantity (\ref{eqn:elpr}) is too large, then the DEF model will exhibit large probability shifts for very small changes in the data configuration (i.e. instability mentioned earlier).

**Proof of Proposition \ref{prop:instab}.** We prove the contrapositive. Suppose that $\Delta_N(\boldsymbol \theta_N) \le C$ holds for some $C > 0$. Under the DEF model, $P_{\boldsymbol \theta_V}(\boldsymbol x) > 0$ holds for each outcome $\boldsymbol x \in \mathcal{X}^N$. Let $\boldsymbol x_{min} \equiv \argmin\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)$ and $\boldsymbol x_{max} \equiv \argmax\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)$. Note there exists a sequence $\boldsymbol x_{min} \equiv \boldsymbol x_0, \boldsymbol x_1, \dots, \boldsymbol x_k \equiv \boldsymbol x_{max}$ in $\mathcal{X}^N$ of component-wise switches to move from $\boldsymbol x_{min}$ to $\boldsymbol x_{max}$ in the sample space (i.e. $\boldsymbol x_i, \boldsymbol x_{i + 1} \in \mathcal{X}^N$ differ by exactly $1$ component for $i = 0, \dots, k$) for some integer $k \in \{0, 1, \dots, N\}$. Then
\begin{align*}
\log\left[\frac{P_{\boldsymbol \theta_N}(\boldsymbol x_{max})}{P_{\boldsymbol \theta_N}(\boldsymbol x_{min})}\right] &= \left|\sum\limits_{i = 1}^k\log\left(\frac{P_{\boldsymbol \theta_N}(\boldsymbol x_i)}{P_{\boldsymbol \theta_N}(\boldsymbol x_{i-1})}\right)\right| \\
& \le \sum\limits_{i = 1}^k\left|\log\left(\frac{P_{\boldsymbol \theta_N}(\boldsymbol x_i)}{P_{\boldsymbol \theta}(\boldsymbol x_{i-1})}\right)\right| \\
& \le k \Delta_N(\boldsymbol \theta_N) \le NC
\end{align*}
using $k \le N$ and $\Delta_N(\boldsymbol \theta_N) \le C$.$\Box$

Additionally, unstable RBM models are connected to degenerate models, where *degeneracy* implies placing all probability on a small pieace of the sample space. Define a model set 
$$
M_{\epsilon, \boldsymbol \theta_N} \equiv \left\{(x_1, \dots, x_N) \in \mathcal{X}^N: \log P_{\boldsymbol \theta_N}(x_1, \dots, x_N) > (1-\epsilon)\max\limits_{(x_1^*, \dots, x_N^*)}P_{\boldsymbol \theta_N}(x_1^*, \dots, x_N^*) + \epsilon\min\limits_{(x_1^*, \dots, x_N^*)}P_{\boldsymbol \theta_N}(x_1^*, \dots, x_N^*) \right\}
$$
of possible outcomes, for a given $0 < \epsilon < 1$.

\begin{prop}
\label{prop:degen}
For an unstable DEF model, and for any given $0 < \epsilon < 1$, $P_{\boldsymbol \theta_N}\left((x_1, \dots, x_N) \in M_{\epsilon, \boldsymbol \theta_N}\right) \rightarrow 1$ holds as $N \rightarrow \infty$.
\end{prop}

In other words, as a consequence of unstable models, all probability will stack up on mode sets or potentially those few outcomes with the highest probability.

**Proof of Proposition \ref{prop:degen}.** Define $\boldsymbol x_{max}$ and $\boldsymbol x_{min}$ as in the proof of Proposition \ref{prop:instab}. Fix $0 < \epsilon < 1$ Then, $\boldsymbol x_{max} \in M_{\epsilon, \boldsymbol \theta_N}$, so $P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N}) \ge P_{\boldsymbol \theta_N}(\boldsymbol x_{max})$. If $\boldsymbol x \in \mathcal{X}^N \setminus M_{\epsilon, \boldsymbol \theta_N}$, then by definition $P_{\boldsymbol \theta_N}(\boldsymbol x) \le [P_{\boldsymbol \theta_N}(\boldsymbol x_{max})]^{1-\epsilon}[P_{\boldsymbol \theta_N}(\boldsymbol x_{min})]^{\epsilon}$ holds so that
\begin{align*}
1-P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N}) &= P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N}^C) \\
& = \sum\limits_{\boldsymbol x \in \mathcal{X}^N \setminus M_{\epsilon, \boldsymbol \theta_N}}P_{\boldsymbol \theta_N}(\boldsymbol x) \\
& \le (2^N)[P_{\boldsymbol \theta_N}(\boldsymbol x_{max})]^{1-\epsilon}[P_{\boldsymbol \theta_N}(\boldsymbol x_{min})]^{\epsilon}
\end{align*}
Then,
\begin{align*}
\frac{1}{N}\log\left[\frac{P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N})}{1-P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N})}\right] & \ge \frac{1}{N} \log\left[\frac{P_{\boldsymbol\theta_N}(\boldsymbol x_{max})}{(2^V)[P_{\boldsymbol \theta_N}(\boldsymbol x_{max})]^{1-\epsilon}[P_{\boldsymbol \theta_N}(\boldsymbol x_{min})]^{\epsilon}}\right] \\
&= \frac{\epsilon}{N} \log\left[\frac{P_{\boldsymbol \theta_N}(\boldsymbol x_{max})}{P_{\boldsymbol \theta_N}(\boldsymbol x_{min})}\right] - \log 2 \rightarrow \infty
\end{align*}
as $N \rightarrow \infty$ by the definition of an unstable DEF model.$\Box$

Note this somewhat general result can be applied to many model families. In Section \ref{examples-of-def-models}, we present two specific examples of DEF models, including models with deep architectures.

# Examples of DEF models

## Discrete exponential family models

## Restricted Boltzmann machines

## Deep learning

# Implications

# References

### 2016-10-06 writing/note.Rmd
restructure from Nordman

@@
 
-# Discrete Expanding Finite space models and notions of instability
+# Discrete Expanding Finite space models
 
@@
 
+Note many model families can fall under the umbrella of DEF models.
+In Sections \ref{discrete-exponential-family-models}-\ref{deep-learning}, we present three specific examples of DEF models, including models with deep architectures.
+
+## Discrete exponential family models
+
+## Restricted Boltzmann machines
+
+## Deep learning
+
+# Instability and degeneracy results
+
 To quantify "instability" in a model with discrete outcomes, it is useful to imagine how a data model might be expanded to incorporate more and more observations.
@@
 \begin{align}
-\frac{1}{N} \log \left[\frac{\max\limits_{(x_1, \dots, x_N) \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(x_1, \dots, x_N)}{\min\limits_{(x_1, \dots, x_N) \in \mathcal{X}^N}P_{\boldsymbol \theta}(x_1, \dots, x_N)}\right] \equiv \frac{1}{V} \text{ELPR}_N(\boldsymbol \theta_N) \label{eqn:elpr}
+\frac{1}{N} \log \left[\frac{\max\limits_{(x_1, \dots, x_N) \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(x_1, \dots, x_N)}{\min\limits_{(x_1, \dots, x_N) \in \mathcal{X}^N}P_{\boldsymbol \theta}(x_1, \dots, x_N)}\right] \equiv \frac{1}{N} \text{ELPR}_N(\boldsymbol \theta_N) \label{eqn:elpr}
 \end{align}
@@
 
-Note this somewhat general result can be applied to many model families.
-In Section \ref{examples-of-def-models}, we present two specific examples of DEF models, including models with deep architectures.
-
-# Examples of DEF models
-
-## Discrete exponential family models
-
-## Restricted Boltzmann machines
-
-## Deep learning
-
 # Implications

### 2016-10-08 writing/draft.Rmd
2 examples done

@@
 The data mining and machine learning communities have recently found a great interest in deep learning, specifically in stacked restricted Boltzmann machines (RBMs).
-(**TODO: deep learning, deep RBM references.**) A RBM is an undirected probabilistic graphical model (for discrete or continuous random variables) with two layers, one hidden and one visible, with no conditional dependency within a layer (**TODO: RBM references.**).
+(**TODO: deep learning, deep RBM references.**) A RBM is an undirected probabilistic graphical model (for discrete or continuous random variables) with two layers, one hidden and one visible, with no conditional dependency within a layer [@smolensky1986information].
 These models have reportedly been used with success in classification of images.

### 2016-10-08 writing/note.Rmd
2 examples done

@@
 date: "October, 2016"
-linestretch: 2
+linestretch: 1.5
 output: 
@@
 
+Let $X_1, \dots, X_N$ be discrete random variables with sample space $\mathcal{X}^N$, where $|\mathcal{X}| < \infty$.
+Consider discrete exponential family model with probability mass function of the form
+$$
+p_{\boldsymbol \lambda}(\boldsymbol x) = \exp\left[\boldsymbol\eta^T(\boldsymbol \lambda) \boldsymbol g(\boldsymbol x) - \psi(\boldsymbol \lambda)\right], \qquad \boldsymbol x \in \mathcal{X}^N.
+$$ 
+where $\boldsymbol \lambda \in \Lambda \subset \mathbb{R}^{q(N)}$, $\boldsymbol \eta : \mathbb{R}^q \mapsto \mathbb{R}^L$ is a vector of natural parameters, $\boldsymbol g : \mathcal{X}^N \mapsto \mathbb{R}^L$ is a vector of sufficient statistics,
+$$
+\psi(\boldsymbol \lambda) = \log \sum\limits_{\boldsymbol x \in \mathcal{X}^N}\exp\left[\boldsymbol \eta^T(\boldsymbol \lambda) \boldsymbol g(\boldsymbol x) \right], \qquad \boldsymbol \lambda \in \Lambda
+$$
+is the normalizing function, and $\Lambda = \{\boldsymbol \lambda \in \mathbb{R}^k : \psi(\boldsymbol \lambda) < \infty, k \le q(N) \}$ is the parameter space.
+
+By above definition, $X_1, \dots, X_N$ are discrete with a finite sample space.
+Let $\boldsymbol \theta = (\lambda_1, \dots, \lambda_k, 0, \dots, 0)$, where the first $k$ elements are equal to the $k$ elements of $\boldsymbol \lambda$ and for $i = k + 1, \dots, q(N)$, $\theta_i = 0$.
+Further, because $p_{\boldsymbol \lambda}(\boldsymbol x) > 0$ for all $\boldsymbol x \in \mathcal{X}^N$ by properties of the exponential function, the discrete exponential family models defined in this section are a special case of the DEF model.
+
 ## Restricted Boltzmann machines
+
+A restricted Boltzmann machine (RBM) is an undirected graphical model specified for discrete or continuous random variables, binary variables being most common [@smolensky1986information].
+Consider the binary case of an RBM only.
+A RBM architecture has two layers, hidden ($\mathcal{H}$) and visible ($\mathcal{V}$), with conditional independence within each layer.
+Let $\boldsymbol X = \{H_1, \dots, H_H, V_1,\dots, V_V\}$ be random variables corresponding to the hidden and visible variables in an RBM.
+We will consider the sample space $\mathcal{X} = \{-1,1\}$.
+The RBM is then has the joint probability mass function
+$$
+p_{\boldsymbol \theta} (\boldsymbol x) = \exp\left[ \boldsymbol \alpha^T \boldsymbol h + \boldsymbol \beta^T \boldsymbol v + \boldsymbol h^T \Gamma \boldsymbol v - \psi(\boldsymbol \theta)\right], \qquad \boldsymbol x = (\boldsymbol h, \boldsymbol v) \in \mathcal{X}^{H+V}
+$$
+where 
+$$
+\psi(\boldsymbol \theta) = \log \sum_{\boldsymbol x \in \mathcal{X}^{H + V}} \exp\left[ \boldsymbol \alpha^T \boldsymbol h + \boldsymbol \beta^T \boldsymbol v + \boldsymbol h^T \Gamma \boldsymbol v\right], \qquad \boldsymbol \theta \in \Theta
+$$ 
+is the normalizing function, $\boldsymbol \alpha \in \mathbb{R}^H$, $\boldsymbol \beta \in \mathbb{R}^V$, and $\Gamma$ is a matrix with dimension $H \times V$ are the parameters in the model.
+Let $\boldsymbol \theta = (\boldsymbol \alpha, \boldsymbol \beta,\Gamma) \in \Theta \subset \mathbb{R}^{H + V + H*V}$ denote the combined vector of parameters.
+
+The probability mass function for the visible random variables $V_1, \dots, V_V$ follows from this joint specification:
+$$
+p_{\boldsymbol \theta} (\boldsymbol v) = \sum\limits_{\boldsymbol h \in \mathcal{X}^H} p_{\boldsymbol \theta} (\boldsymbol x), \qquad \boldsymbol v \in \mathcal{X}^V
+$$
+
+By definition, $V_1, \dots, V_V$ are discrete with a finite sample space.
+The vector of model parameters $\boldsymbol \theta$, of size $q(V) \equiv V + H + H*V$, is expanding in size as a function of the dimension of $V_1, \dots, V_V$.
+Additionally, $p_{\boldsymbol \theta}(\boldsymbol v) > 0$ for all $\boldsymbol v \in \mathcal{X}^V$ by properties of the exponential function.
+Thus, the visible RBM model specification is an example of a DEF model.
 

### 2016-10-17 writing/note.Rmd
deep learning and implications

@@
 ## Deep learning
+
+Consider two models with "deep architechture" that contain multiple hidden (or latent) layers in addition to a visible layer of data, a deep Boltzmann machine [@salakhutdinov2009deep] and a deep belief network [@hinton2006fast].
+Let $M$ denote the number of hidden layers included in the model and $H_1, \dots, H_M$ the number of hidden variables within each hidden layer.
+Then let $\boldsymbol X = \{H^{(1)}_1, \dots, H^{(1)}_{H_1}, \dots, H^{(M)}_1, \dots, H^{(M)}_{H_M}, V_1,\dots, V_V\}$ be random variables corresponding to the hidden and visible variables in a deep probabilitic model.
+We will consider the sample space $\mathcal{X} = \{-1,1\}$.
+
+**Deep Boltzmann machine (DBM).** The DBM class of models maintains conditional independence within all layers in the model by stacking RBM models and maintaining the restriction of only allowing conditional dependence between layers.
+The joint probability mass function for a DBM is
+$$
+p_{\boldsymbol \theta} (\boldsymbol x) = \exp\left[ \sum\limits_{i = 1}^M\boldsymbol \alpha^{(i)T} \boldsymbol h^{(i)} + \boldsymbol \beta^T \boldsymbol v + \boldsymbol h^{(1)T} \Gamma^{(0)} \boldsymbol v + \sum\limits_{i = 1}^{M - 1} \boldsymbol h^{(i)T} \Gamma^{(i)} \boldsymbol h^{(i + 1)} - \psi(\boldsymbol \theta) \right],
+$$
+where $\boldsymbol x = (\boldsymbol h^{(1)}, \dots, \boldsymbol h^{(M)}, \boldsymbol v) \in \mathcal{X}^{H_1 + \cdots + H_M +V}$ and
+$$
+\psi(\boldsymbol \theta) = \log \sum_{\boldsymbol x \in \mathcal{X}^{H_1 + \cdots + H_M + V}} \exp\left[ \sum\limits_{i = 1}^M\boldsymbol \alpha^{(i)T} \boldsymbol h^{(i)} + \boldsymbol \beta^T \boldsymbol v + \boldsymbol h^{(1)T} \Gamma^{(0)} \boldsymbol v + \sum\limits_{i = 1}^{M - 1} \boldsymbol h^{(i)T} \Gamma^{(i)} \boldsymbol h^{(i + 1)}\right], \qquad \boldsymbol \theta \in \Theta
+$$ 
+is the normalizing function, $\boldsymbol \beta \in \mathbb{R}^V$, $\boldsymbol \alpha^{(i)} \in \mathbb{R}^{H_i}$ for $i = 1, \dots, M$, $\Gamma^{(0)}$ is a matrix with dimension $H_1 \times V$, and $\Gamma^{(i)}$ is a matrix with dimension $H_i \times H_{i + 1}$ for $i = 1, \dots, M-1$ are the parameters in the model.
+Let $\boldsymbol \theta = (\boldsymbol \alpha^{(1)}, \dots, \boldsymbol \alpha^{(M)}, \boldsymbol \beta,\Gamma^{(0)}, \dots, \Gamma^{(M - 1)}) \in \Theta \subset \mathbb{R}^{H_1 + \cdots + H_M + V + H_1*V + H_1 *H_{2} + \cdots + H_{M-1}*H_M}$ denote the combined vector of parameters.
+
+The probability mass function for the visible random variables $V_1, \dots, V_V$ follows from this joint specification:
+$$
+p_{\boldsymbol \theta} (\boldsymbol v) = \sum\limits_{(\boldsymbol h^{(1)}, \dots, \boldsymbol h^{(M)}) \in \mathcal{X}^{H_1 + \cdots + H_M}} p_{\boldsymbol \theta} (\boldsymbol x), \qquad \boldsymbol v \in \mathcal{X}^V
+$$
+
+By definition, $V_1, \dots, V_V$ are discrete with a finite sample space.
+The vector of model parameters $\boldsymbol \theta$, of size $q(V) \equiv H_1 + \cdots + H_M + V + H_1*V + H_1 *H_{2} + \cdots + H_{M-1}*H_M$, is expanding in size as a function of the dimension of $V_1, \dots, V_V$.
+Additionally, $p_{\boldsymbol \theta}(\boldsymbol v) > 0$ for all $\boldsymbol v \in \mathcal{X}^V$ by properties of the exponential function.
+Thus, the visible RBM model specification is an example of a DEF model.
+
+**Deep belief network (DBN).** A DBN is similar to a DBM in that there are multiple layers of latent random variables stacked in a deep architecture with no conditional dependence between layers.
+The difference between the two models is that all but the last stacked model are Bayesian networks, rather than RBMs.
+Thus for $V_1, \dots, V_V$ discrete with finite sample space and support equal to the sample space $\mathcal{X}^V$, a DBN is also a DEF model if the number of components in the parameter vector is dependent on the dimension of the data, $V$.
+Commonly, as in a logistic belief net [@neal1992connectionist], a "weight" parameter is placed on each interaction between visibles, $V_1, \dots, V_V$ and the first layer of latent variables, $H^{(1)}_1, \dots, H^{(1)}_{H_1}$, satisfying the definition of a DEF model.
 
@@
 
+For a large class of models that cover a broad range of applications (including "deep learning"), we have developed a formal definition and elucidated multiple consequences of instability.
+We have shown for DEF models that instability manifests through small changes in data leading to overly-sensitive changes in probability as well as the potential to place all probability on a small pieace of the sample space.
+Models that fall within the definition of a DEF model should be used with caution to ensure that the effects of instability are not experienced.
+
 # References

### 2016-10-19 writing/draft.Rmd
first part of edits from Nordman

@@
 ---
-title: "An exposition on the propriety of restricted Boltzmann machines"
+title: "On the propriety of restricted Boltzmann machines"
 author: "Andee Kaplan, Daniel Nordman, Stephen Vardeman"
@@
 In recent years, RBMs have risen to prominence due to their connection to deep learning.
-By treating a hidden layer of one RBM as the visible layer in a second RBM a deep architecture can be created.
-RBMs are thought to have the ability to encode very complex and rich structures in data, making them attractive for supervised learning.
+By treating a hidden layer of one RBM as the visible layer in a second RBM, a deep architecture can be created.
+RBMs are thereby thought to have the ability to encode very complex and rich structures in data, making them attractive for supervised learning.
 However, the generative behavior of RBMs has largely been unexplored.
-In this paper, we discuss the relationship between parameter RBM specification and the prevalence of degenerate models as well as the difficulties that arise in fitting such highly flexible models in the binary case.
+In this paper, we discuss the relationship between parameter RBM specification and the prevalence of degeneracy, instability and/or uninterpretability in such models in the binary case.
+We also describe the difficulties that arise in likelihood-based fitting such highly flexible models, especially as Gibbs sampling methods are often advocated for the RBM model structure.
 ---
@@
 These models have reportedly been used with success in classification of images.
-(**TODO: classification references.**) However, the model properties are largely unexplored in the literature and the commonly cited fitting methodology remains heuristic-based and abstruse.
+(**TODO: classification references.**) However, the model properties are largely unexplored in the literature and the commonly cited fitting methodology remains heuristic-based and abstruse (**TODO: citation**).
 In this paper, we attempt to provide steps toward a thorough understanding of the model and its behavior from the perspective of statistical model theory and then explore the possibility of a rigorous fitting methodology.
@@
 
-First, the models are unsatisfactory as a conceptualization of how data is generated.
+First, the models are unsatisfactory as a conceptualization of how data are generated.
 According to Fisher, the aim of a statistical model is to represent data in a compact way [@fisher1922mathematical].
 Neyman and Box agree with this assessment and go further to state that a model should "provide an explanation of the mechanism underlying the observed phenomena" [@lehmann1990model, @box1967discrimination].
-RBMs often fail to generate data that resemble realistic data and thus an unsatisfactory conceptualization of the data generation process.
-Additionally, we find that RBMs easily exhibit a kind of instability in the parameter space.
-In practice, this is seen when a single pixel change in an image results in a wildly different classification.
+RBMs can often generate data lacking realistic variability and thereby fail to provide a satisfactory conceptualization of the data generation process.
+In addition to such degeneracy, we find that RBMs easily exhibit types of instability in the parameter space.
+In practice, this may seen when a single pixel change in an image results in a wildly different classification.
 Such model impropriety issues have  been documented in other deep architectures recently [@szegedy2013intriguing, @nguyen2014deep].
-We investigate these phenomena in RBMs in section \ref{explorations} through simulations of small, manageable examples.
+We investigate these phenomena in RBMs in Section \ref{explorations} through simulations of small, manageable examples.
 
@@
 But even when fitting is feasible, there are unique challenges due to the overparameterization of RBMs.
-In section \ref{bayesian-model-fitting} we compare three Bayesian modelling techniques, each proposed with the goal of avoiding parts of the RBM parameter space that yield improper models.
-As could be expected, with computational complexity comes an increase in fitting accuracy, but at the cost of feasibility.
-Additionally, because these models are highly overparametrized, any fully principled fitting method will reproduce the empirical distribution of most realistic training sets.
-There is no "smoothing" achieved with a RBM fit using a rigorous method.
+The literature often suggests MCMC approximations to fit these models (exploiting conditional structure in the hidden and visible variables), but little is said about monitoring convergence [hinton2010practical, @hinton2006fast].
+Related to this, the MCMC algorithms involve updating potentially many latent variables (hiddens) per observation (visibles) which can critically influence convergence in MCMC-based likelihood methods.
+In Section \ref{bayesian-model-fitting}, we compare three Bayesian modelling techniques (which can be computationally more accessible than direct maximum likelihood), each proposed with the goal of avoiding parts of the RBM parameter space that yield improper models.
+As might be expected, with greater computational complexity comes an increase in fitting accuracy, but at the cost of feasibility.
+Additionally, because these models are highly overparametrized, any fully principled fitting method based on the likelihood will simply reproduce the empirical distribution of most realistic training sets.
+There is no "smoothing" achieved with a RBM fit using a rigorous likelihood-based method.
 We are therefore led to be skeptical that any model built using these structures, like a deep Boltzmann machine, can achieve prediction or inference in a principled way.
 
-This paper is structured as follows, section \ref{background} gives a definition of the RBM including the joint distribution of hidden and visible variables and shows the connection to deep learning.
+This paper is structured as follows, Section \ref{background} gives a definition of the RBM including the joint distribution of hidden and visible variables and shows the connection to deep learning.
 Additionally, measures of model impropriety and methods of detecting them are defined.
 Section \ref{explorations} details our explorations into the model behavior and propriety (or lack thereof) of the RBM class.
-We discuss three Bayesian fitting techniques with the goal of avoiding model impropriety in section \ref{bayesian-model-fitting} and wrap up with a discussion in section \ref{discussion}.
-Additionally, there is an accompanying appendix with auxiliary proofs relating to the choice of data encoding and propositions on instability in section \ref{appendix}.
+We discuss three Bayesian fitting techniques with the goal of avoiding model impropriety in Section \ref{bayesian-model-fitting} and conclude with a discussion in Section \ref{discussion}.
+Additionally, there is an accompanying appendix with auxiliary proofs relating to the choice of data encoding and propositions on instability.
 
@@
 A restricted Boltzmann machine (RBM) is an undirected graphical model specified for discrete or continuous random variables, binary variables being most common.
-In this paper, we consider the binary case of an RBM only.
+In this paper, we consider the binary case of an RBM for concreteness.
 A RBM architecture has two layers, hidden ($\mathcal{H}$) and visible ($\mathcal{V}$), with no connections within a layer.
-An example of this structure is in figure \ref{fig:rbm} with the hidden nodes indicated by gray circles and the visible nodes indicated by white circles.
+An example of this structure is in Figure \ref{fig:rbm} with the hidden nodes indicated by gray circles and the visible nodes indicated by white circles.
 
@@
 
-A common use for RBMs is to create features for use in classification. For example, binary images could be classified where the pixel values would be treated as the visible variables $v_i$.
+A common use for RBMs is to create features for use in classification.
+For example, binary images could be classified where the pixel values would be treated as the visible variables $v_i$ [@hinton2006fast].
 
@@
 Let $\boldsymbol x = \{h_1, \dots, h_H, v_1,\dots,v_V\}$ represent the states of the visible and hidden nodes in an RBM.
-We will consider both $\boldsymbol x \in \{0,1\}^{H + V}$ and $\boldsymbol x \in \{-1, 1\}^{H + V}$, two binary $H + V$-dimensional spaces.
-We posit a parametric form for probabilities corresponding to a potential vector of states of each node taking the value of $1$.
+We will consider both $\boldsymbol x \in \{0,1\}^{H + V}$ and $\boldsymbol x \in \{-1, 1\}^{H + V}$, as the potential encodings for a binary $H + V$-dimensional space, $\mathcal{X}$.
+We posit a parametric form for probabilities corresponding to a potential vector of states, $X = (H_1, \dots, H_H, V_1,\dots,V_V)$, for the nodes as
 
@@
 \label{eqn:pmf}
-f_{\boldsymbol \theta} (\boldsymbol x) = \frac{\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\gamma(\boldsymbol \theta)}
+f_{\boldsymbol \theta} (\boldsymbol x)\equiv P_{\boldsymbol \theta}(\boldsymbol X = \boldsymbol x) = \frac{\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\gamma(\boldsymbol \theta)}, \quad \boldsymbol x \in \mathcal{X}
 \end{align}
 
-Where the denominator $\gamma(\boldsymbol \theta)$ is the normalizing function that ensures the distribution sums to one, $\gamma(\boldsymbol \theta) = \sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)$.
-Let $\boldsymbol \tau = \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{X}\} = \{v_i, h_j, v_i h_j: i = 1, \dots, V, j = 1, \dots H\}$ be the vector of sufficient statistics in the model and $Q(\boldsymbol x) = \sum\limits_{i = 1}^H\sum\limits_{j=1}^V \theta_{ij} h_i v_j + \sum\limits_{i = 1}^H\theta_{hi} h_i + \sum\limits_{j = 1}^V\theta_{vj} v_j$, the neg-potential function.
-The RBM model is parameterized by $\boldsymbol \theta$, which contains two types of parameters, main effects and interaction effects.
+where $\boldsybol \theta \equiv (\theta_{11}, \dots, \theta_{1H}, \dots, \theta_{V1}, \dots, \theta_{VH}, \theta_{v_1}, \dots, \theta_{v_V}, \theta_{h_1}, \dots, \theta_{h_H}) \in \mathcc{R}^{V + H + V*H}$ denotes hte vector of model parameters and the denominator $\gamma(\boldsymbol \theta)$ is the normalizing function that ensures the distribution sums to one, $\gamma(\boldsymbol \theta) = \sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)$.
+Let $\boldsymbol \tau = \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{X}\} = \{v_i, h_j, v_i h_j: i = 1, \dots, V, j = 1, \dots H\}$ be the vector of sufficient statistics in the model and write $Q_{\boldsymbol \theta}(\boldsymbol x) = \sum\limits_{i = 1}^H\sum\limits_{j=1}^V \theta_{ij} h_i v_j + \sum\limits_{i = 1}^H\theta_{hi} h_i + \sum\limits_{j = 1}^V\theta_{vj} v_j$, the neg-potential function.
+Note that the RBM model is parameterized by $\boldsymbol \theta$containing two types of parameters, main effects and interaction effects.
 The main effects parameters ($\theta_{v_i}, \theta_{h_j}$) weight the values of the visible and hidden nodes and the interaction effects parameters ($\theta_{ij}$) weight the values of the connections, or dependencies, between hidden and visible layers.
 
-Due to the potential size of the model the normalizing constant $\gamma(\boldsymbol \theta)$ can be practically impossible to calculate, making it impossible to estimate the model parameter vector in an obvious fashion.
-A kind of Gibbs sampling can be tried, which is natural due to the restricted architecture of the RBM.
-Specifically, the conditional independence of nodes in each layer given those in the other allows for simulation of the main and hidden layers.
+Due to the potential size of the model, the normalizing constant $\gamma(\boldsymbol \theta)$ can be practically impossible to calculate, making it impossible to estimate the model parameter vector in an obvious fashion.
+A kind of Gibbs sampling can be tried, which is natural due to the restricted architecture of conditioal independence of the RBM (i.e.
+visibles given hiddens or vice versa).
+Specifically, the conditional independence of nodes in each layer given those in the other allows for simulation of the visible and hidden layers.
 
@@
 ## Degeneracy, instability, and uninterpretability... Oh my!
-The highly flexible nature of a RBM (having as it does $H + V + H V$ parameters) makes three kinds of potential model impropriety of concern, *degeneracy, instability, and uninterpretability*.
-In this section we define these characteristics, consider how to detect them in a RBM, and point out their relationship to one another.
+The highly flexible nature of a RBM (having as it does $H + V + H V$ parameters) creates three kinds of potential model impropriety of concern: *degeneracy, instability, and uninterpretability*.
+In this section we define these characteristics, consider how to detect them in a RBM, and point out their relationships to one another.
 

### 2016-10-19 writing/appendix-instability.Rmd
degeneracy and instability

@@
-{\bf Proof of Proposition \ref{prop:instab}.} We prove the contrapositive.
-Suppose that $\Delta_V(\boldsymbol \theta) \le C$ holds for some $C > 0$.
-Under the RBM model for visibles, $P_{\boldsymbol \theta}(\boldsymbol v) > 0$ holds for each outcome $\boldsymbol v \in \{-1,1\}^V$.
-Let $\boldsymbol v_{min} \equiv \argmin\limits_{\boldsymbol v \in \{-1, 1\}^V}P_{\boldsymbol \theta}(\boldsymbol v)$ and $\boldsymbol v_{max} \equiv \argmax\limits_{\boldsymbol v \in \{-1, 1\}^V}P_{\boldsymbol \theta}(\boldsymbol v)$.
-Note there exists a sequence $\boldsymbol v_{min} \equiv \boldsymbol v_0, \boldsymbol v_1, \dots, \boldsymbol v_k \equiv \boldsymbol v_{max}$ in $\{-1,1\}^V$ of component-wise switches to move from $\boldsymbol v_{min}$ to $\boldsymbol v_{max}$ in the sample space (i.e.
-$\boldsymbol v_i, \boldsymbol v_{i + 1} \in \{-1,1\}^V$ differ by exactly $1$ component for $i = 0, \dots, k$) for some integer $k \in \{0, 1, \dots, V\}$.
-Then
-\begin{align*}
-\log\left[\frac{P_{\boldsymbol \theta}(\boldsymbol v_{max})}{P_{\boldsymbol \theta}(\boldsymbol v_{min})}\right] &= \left|\sum\limits_{i = 1}^k\log\left(\frac{P_{\boldsymbol \theta}(\boldsymbol v_i)}{P_{\boldsymbol \theta}(\boldsymbol v_{i-1})}\right)\right| \\
-& \le \sum\limits_{i = 1}^k\left|\log\left(\frac{P_{\boldsymbol \theta}(\boldsymbol v_i)}{P_{\boldsymbol \theta}(\boldsymbol v_{i-1})}\right)\right| \\
-& \le k \Delta_V(\boldsymbol \theta) \le VC
-\end{align*}
-using $k \le V$ and $\Delta_V(\boldsymbol \theta) \le C$.$\Box$
-
-{\bf Proof of Proposition \ref{prop:degen}.} Define $\boldsymbol v_{max}$ and $\boldsymbol v_{min}$ as in the proof of Proposition \ref{prop:instab}.
-Fix $0 < \epsilon < 1$ Then, $\boldsymbol v_{max} \in M_{\epsilon, \boldsymbol \theta}$, so $P_{\boldsymbol \theta}(M_{\epsilon, \boldsymbol \theta}) \ge P_{\boldsymbol \theta}(\boldsymbol v_{max})$.
-If $\boldsymbol v \in \{-1, 1\}^V \setminus M_{\epsilon, \boldsymbol \theta}$, then by definition $P_{\boldsymbol \theta}(\boldsymbol v) \le [P_{\boldsymbol \theta}(\boldsymbol v_{max})]^{1-\epsilon}[P_{\boldsymbol \theta}(\boldsymbol v_{min})]^{\epsilon}$ holds so that
-\begin{align*}
-1-P_{\boldsymbol \theta}(M_{\epsilon, \boldsymbol \theta}) &= P_{\boldsymbol \theta}(M_{\epsilon, \boldsymbol \theta}^C) \\
-& = \sum\limits_{\boldsymbol v \in \{-1, 1\}^V \setminus M_{\epsilon, \boldsymbol \theta}}P_{\boldsymbol \theta}(\boldsymbol v) \\
-& \le (2^V)[P_{\boldsymbol \theta}(\boldsymbol v_{max})]^{1-\epsilon}[P_{\boldsymbol \theta}(\boldsymbol v_{min})]^{\epsilon}
-\end{align*}
-Then,
-\begin{align*}
-\frac{1}{V}\log\left[\frac{P_{\boldsymbol \theta}(M_{\epsilon, \boldsymbol \theta})}{1-P_{\boldsymbol \theta}(M_{\epsilon, \boldsymbol \theta})}\right] & \ge \frac{1}{V} \log\left[\frac{P_{\boldsymbol\theta}(\boldsymbol v_{max})}{(2^V)[P_{\boldsymbol \theta}(\boldsymbol v_{max})]^{1-\epsilon}[P_{\boldsymbol \theta}(\boldsymbol v_{min})]^{\epsilon}}\right] \\
-&= \frac{\epsilon}{V} \log\left[\frac{P_{\boldsymbol \theta}(\boldsymbol v_{max})}{P_{\boldsymbol \theta}(\boldsymbol v_{min})}\right] - \log 2 \rightarrow \infty
-\end{align*}
-as $V \rightarrow \infty$ by the definition of an unstable RBM model.$\Box$
-
 {\bf Proof of (\ref{eqn:lowerbnd}).} With algebra, we have

### 2016-10-19 writing/draft.Rmd
degeneracy and instability

@@
 ### Near-degeneracy
-In Random Graph Model theory, *model degeneracy* means there is a disproportionate amount of probability placed on only a few elements of the sample space, $\mathcal{X}$, by the model.
+In Random Graph Model theory, *model degeneracy* means there is a disproportionate amount of probability placed on only a few elements of the sample space, $\mathcal{X}$, by the model [@handcock2003assessing].
 In the scenario of random graph models, $\mathcal{X}$ denotes all possible graphs that can be constructed from a set of nodes.
@@
 where $\boldsymbol \theta \in \Theta \subset \mathbb{R}^q$ is the model parameter, and $t: \mathcal{X} \rightarrow \mathbb{R}^q$ are statistics based on the adjacency matrix of a graph.
-Again, $\gamma(\boldsymbol \theta)$ is the normalizing function that guarantees the distribution is valid, $\gamma(\boldsymbol \theta) = \sum_{\boldsymbol x \in \mathcal{X}} \exp\left(\boldsymbol \theta^T t(\boldsymbol x)\right)$.
-Let $C$ be the convex hull of $\{t(\boldsymbol x): \boldsymbol x \in \mathcal{X}\}$.
-@handcock2003assessing classify an exponentially parametrized random graph model as *near-degenerate* if $\mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta}t(X)$ is close to the boundary of the $C$.
-
-The RBM class of models can be thought to exhibit an analogous occurrence of *near-degeneracy* when random variables in the neg-potential function $Q(\cdot)$, $t(\boldsymbol x)$, have a mean vector $\mu(\boldsymbol \theta)$ close to the boundary of the convex hull of $\mathcal{X}$, where $t(\boldsymbol x) = \{v_1, \dots, v_V, h_1, \dots, h_H, v_1 h_1, \dots, v_V h_H \} \in \mathcal{X}$ and the mean parameterization on the model parameters, $\mu(\boldsymbol \theta)$ is
+Again, $\gamma(\boldsymbol \theta)$ is the normalizing function that guarantees the distribution is valid, e.g.
+$\gamma(\boldsymbol \theta) = \sum_{\boldsymbol x \in \mathcal{X}} \exp\left(\boldsymbol \theta^T t(\boldsymbol x)\right)$.
+Let $C$ denote the convex hull spanned by the potential outcomes of sufficient statistics, $\{t(\boldsymbol x): \boldsymbol x \in \mathcal{X}\}$.
+@handcock2003assessing classify an exponentially parametrized random graph model at $\boldsymbol \theta$ as *near-degenerate* if the mean value of sufficient statistics under $\boldsymbol \theta$, $\mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta}t(X)$, is close to the boundary of $C$.
+
+The RBM class of models can be thought to exhibit an analogous occurrence of *near-degeneracy* as follows.
+When random variables $t(\boldsymbol x) = \{v_1, \dots, v_V, h_1, \dots, h_H, v_1 h_1, \dots, v_V h_H \} \in \mathcal{X}$ in the neg-potential function $Q_{\boldsymbol \theta}(\cdot)$, have a mean vector $\mu(\boldsymbol \theta)$ close to the boundary of the convex hull of $\mathcal{X}$, then the RBM model is near-degenerate at $\boldsymbol \theta \in \mathbb{R}^{V + H + H*V}$.
+Here the mean of $t(\boldsymbol x)$ isgiven as
 \begin{align*}
@@
 Considering a class of exponential families of distributions, @schweinberger2011instability introduced a concept of model defficiency related to *instability*.
-Instability is characterized by excessive sensitivity in the model, where small changes in the components of potential data outcomes, $\boldsymbol x$, may lead to substantial changes in the probability mass function.
-To quantify "instability", it is useful to imagine how a data model (the RBM in particular) might be expanded to incorporate more and more observations.
-To increase the size of a RBM model to handle more "visible variables", it becomes necessary to expand the number of model parameters (and in the process of increasing the model size in visibles, one may also arbitrarily expand the number of hidden variables to be included).
-A relevant quantity to this end is a (scaled) extremal log-probability ratio given by
+Instability can be roughly characterized by excessive sensitivity in the model, where small changes in the components of potential data outcomes, $\boldsymbol x$, may lead to substantial changes in the probability mass function.
+To quantify "instability" more rigorously (particularly beyond the definition given by @schweinberger2011instability), it is useful to imagine how a data model (the RBM in particular) might be expanded to incorporate more and more observations.
+When increasing the size of a RBM model to handle more visible variables, it becomes necessary to expand the number of model parameters (and in this process one may also arbitrarily expand the number of hidden variables to be included).
+To this end, let $\boldsymbol \theta_{V} \equiv (\theta_{v_1}, \dots, \theta_{v_V}, \theta_{h_1}, \dots, \theta_{h_H}, \theta_{11}, \theta_{VH}), V \ge 1$, denote a sequence of RBM parameters indexed by the number $V$ of visibles $(V_1, \dots, V_V)$ and define a (scaled) extremal log-probability ratio of the RBM model at $\boldsymbol \theta_{V}$ $(V \ge 1)$ as
 \begin{align}
-\frac{1}{V} \log \left[\frac{\max\limits_{(v_1, \dots, v_V) \in \{-1,1\}^V}P_{\boldsymbol \theta}(v_1, \dots, v_V)}{\min\limits_{(v_1, \dots, v_V) \in \{-1,1\}^V}P_{\boldsymbol \theta}(v_1, \dots, v_V)}\right] \equiv \frac{1}{V} \text{ELPR}_V(\boldsymbol \theta) \label{eqn:elpr}
+\frac{1}{V} \log \left[\frac{\max\limits_{(v_1, \dots, v_V) \in \{-1,1\}^V}P_{\boldsymbol \theta_V}(v_1, \dots, v_V)}{\min\limits_{(v_1, \dots, v_V) \in \{-1,1\}^V}P_{\boldsymbol \theta_V}(v_1, \dots, v_V)}\right] \equiv \frac{1}{V} \text{ELPR}(\boldsymbol \theta_V) \label{eqn:elpr}
 \end{align}
-where $P_{\boldsymbol \theta}(v_1, \dots, v_V) \propto \sum_{\{h_j\}\in \{-1,1\}}\exp\left(\sum_{i = 1}^V \theta_{v_i} v_i + \sum_{j = 1}^H \theta_{h_j} h_j + \sum_{i = 1}^V \sum_{j = 1}^H \theta_{ij} v_i h_j\right)$ is the RBM probability of observing outcome $(v_1, \dots, v_V)$ for the visible variables for parameters $\boldsymbol \theta = (\theta_{v_1}, \dots, \theta_{HV})$.
-
-Essentially, in the process of developing a model to handle increasing numbers of visibles (i.e., $V \rightarrow \infty$), we want (\ref{eqn:elpr}) to stay bounded, meaning that the largest probability under this model should maintain a magnitude within the smallest probability allowed under the model.
+where $P_{\boldsymbol \theta_V}(v_1, \dots, v_V) \propto \sum_{\{h_j\}\in \{-1,1\}}\exp\left(\sum_{i = 1}^V \theta_{v_i} v_i + \sum_{j = 1}^H \theta_{h_j} h_j + \sum_{i = 1}^V \sum_{j = 1}^H \theta_{ij} v_i h_j\right)$ is the RBM probability of observing outcome $(v_1, \dots, v_V)$ for the visible variables $(V_1, \dots, V_V)$ under parameters $\boldsymbol \theta_V$.
+
+Essentially, in formulating a RBM model for a potentially large number of visibles (i.e., $V \rightarrow \infty$),the ratio (\ref{eqn:elpr}) should stay bounded, meaning that the largest probability under this model at parameter $\boldsymbol \theta_V$ should maintain a magnitude within the smallest probability allowed under the same model.
 This leads to the following definition:
@@
 \begin{defn}[unstable RBM]
-A RBM model formulation is \emph{unstable} if, as the number of visible variables increase ($V \rightarrow \infty$), it holds that 
+Let $\boldsymbol \theta_V \in \mathbb{R}^{V + H +  H*V}, v \ge 1$, be a sequence of RBM parameters where the number of hiddens, $H \equiv H(V) \ge 1$, can be an arbitrary function of the number $V$ of visibles.
+A RBM model formulation is \emph{Scheinberger-unstable} or \emph{S-unstable} if, as the number of visible variables increase ($V \rightarrow \infty$), it holds that
 \begin{align*}
-\lim\limits_{V \rightarrow \infty} \frac{1}{V} \log \left[\frac{\max\limits_{(v_1, \dots, v_V) \in \{-1,1\}^V}P_{\boldsymbol \theta}(v_1, \dots, v_V)}{\min\limits_{(v_1, \dots, v_V) \in \{-1,1\}^V}P_{\boldsymbol \theta}(v_1, \dots, v_V)}\right] = \infty.
+\lim\limits_{V \rightarrow \infty} \frac{1}{V} \text{ELPR}(\boldsymbol \theta_V) \equiv \lim\limits_{V \rightarrow \infty} \frac{1}{V} \log \left[\frac{\max\limits_{(v_1, \dots, v_V) \in \{-1,1\}^V}P_{\boldsymbol \theta}(v_1, \dots, v_V)}{\min\limits_{(v_1, \dots, v_V) \in \{-1,1\}^V}P_{\boldsymbol \theta}(v_1, \dots, v_V)}\right] = \infty.
 \end{align*}
 \end{defn}
-In other words, given any $C > 0$, there exists an integer $V_C > 0$ so that $\frac{1}{V}\text{ELPR}_V(\boldsymbol \theta) > C$ for all $V \ge V_C$.
-
-Note that this definition of *unstable* is a generalization of that in @schweinberger2011instability allowing nonexponential family models and an increasing number of parameters (while this definition appears different, it does match that in @schweinberger2011instability for the special exponential models considered there).
-
-Unstable RBM models are undesirable for several reasons. One is that, as mentioned above, small changes in data can lead to overly-sensitive changes in probability. Consider, for example, 
+In other words, given any $C > 0$, there exists an integer $V_C > 0$ so that $\frac{1}{V}\text{ELPR}(\boldsymbol \theta_V) > C$ for all $V \ge V_C$.
+
+Note that this definition of *S-unstable* is a generalization or re-interpretation of "unstable" used in @schweinberger2011instability allowing nonexponential family models for visibles $(v_1, \dots, v_V)$ and an increasing number of parameters.
+While this definition appears different, it matchs that of @schweinberger2011instability for the special exponential models considered there.
+
+S-Unstable RBM models are undesirable for several reasons. One is that, as mentioned above, small changes in data can lead to overly-sensitive changes in probability. Consider, for example, 
 $$
-\Delta_V(\boldsymbol \theta) \equiv \max \left\{\log \frac{P_{\boldsymbol \theta}(v_1, \dots, v_V)}{P_{\boldsymbol \theta}(v_1^*, \dots, v_V^*)} : (v_1, \dots, v_V) \& (v_1^*, \dots, v_V^*) \in \{-1,1\}^V \text{ differ by exactly one component}\right\},
+\Delta(\boldsymbol \theta_V) \equiv \max \left\{\log \frac{P_{\boldsymbol \theta_V}(v_1, \dots, v_V)}{P_{\boldsymbol \theta_V}(v_1^*, \dots, v_V^*)} : (v_1, \dots, v_V) \& (v_1^*, \dots, v_V^*) \in \{-1,1\}^V \text{ differ by exactly one component}\right\},
 $$
-the biggest log-probability ratio for a one-pixel (one component) change in data outcomes (visibles). We then have the following result.
+the biggest log-probability ratio for a one-pixel (one component) change in data outcomes (visibles) at a RBM parameter $\boldsymbol \theta_V$. We then have the following result.
 
@@
 \label{prop:instab}
+Let $C > 0$ and let $\text{ELPR}(\boldsymbol \theta_V)$ be as in (\ref{eqn:elpr}) for an integer $V \ge 1$.
 If $\frac{1}{V}\text{ELPR}_V(\boldsymbol \theta) > C$, then $\Delta_V(\boldsymbol \theta) > C$.
@@
 
-In other words, if the quantity (\ref{eqn:elpr}) is too large, then the RBM model will exhibit large probability shifts for very small changes in the data configuration (i.e.
-instability mentioned earlier).
-
-Additionally, unstable RBM models are connected to degenerate models (placing all probability on a small pieace of the sample space). Define a model set 
+In other words, if the probability ratio (\ref{eqn:elpr}) is too large, then the RBM model will exhibit large probability shifts for very small changes in the data configuration (i.e., instability).
+
+Additionally, S-unstable RBM models are connected to degenerate models  placing all probability on a small portion of the sample space. To see this, define a modal set 
 $$
-M_{\epsilon, \boldsymbol \theta} \equiv \left\{(v_1, \dots, v_V) \in \{-1,1\}^V: \log P_{\boldsymbol \theta}(v_1, \dots, v_V) > (1-\epsilon)\max\limits_{(v_1^*, \dots, v_V^*)}P_{\boldsymbol \theta}(v_1^*, \dots, v_V^*) + \epsilon\min\limits_{(v_1^*, \dots, v_V^*)}P_{\boldsymbol \theta}(v_1^*, \dots, v_V^*) \right\}
+M_{\epsilon, \boldsymbol \theta_V} \equiv \left\{(v_1, \dots, v_V) \in \{-1,1\}^V: \log P_{\boldsymbol \theta_V}(v_1, \dots, v_V) > (1-\epsilon)\max\limits_{(v_1^*, \dots, v_V^*)}P_{\boldsymbol \theta_V}(v_1^*, \dots, v_V^*) + \epsilon\min\limits_{(v_1^*, \dots, v_V^*)}P_{\boldsymbol \theta_V}(v_1^*, \dots, v_V^*) \right\}
 $$
-of possible outcomes, for a given $0 < \epsilon < 1$.
+of possible outcomes, for a given $0 < \epsilon < 1$, and along a sequence $\boldsymbol \theta_V, V \ge 1$, of RBM parameters.
 
@@
 \label{prop:degen}
-For an unstable RBM model, and for any given $0 < \epsilon < 1$, $P_{\boldsymbol \theta}\left((v_1, \dots, v_V) \in M_{\epsilon, \boldsymbol \theta}\right) \rightarrow 1$ holds as $V \rightarrow \infty$.
+For an S-unstable RBM model, and for any given $0 < \epsilon < 1$, $P_{\boldsymbol \theta_V}\left((v_1, \dots, v_V) \in M_{\epsilon, \boldsymbol \theta_V}\right) \rightarrow 1$ holds as $V \rightarrow \infty$.
 \end{prop}
 
-In other words, as a consequence of unstable models, all probability will stack up on mode sets or potentially those few outcomes with the highest probability.
-Proofs of both proposition \ref{prop:instab} and proposition \ref{prop:degen} are provided in appendix \ref{instability-proofs}.
-Propositions \ref{prop:instab}-\ref{prop:degen} have counterpart results in @schweinberger2011instability, but, unlike results there, we do not consider exponential family forms with a fixednumber of parameters (i.e.
-our results are broader for a more complicated model form for the visibles).
-Essentially, with our generalized definition of *unstable* models, one can establish instability results as in @schweinberger2011instability under assumptions that the sample space is finite and all model outcomes have a non-zero probability.
-
-To later numerically investigate unstable RBM models, it is useful to define a computationally tractable lower bound on (\ref{eqn:elpr}) given by
+In other words, as a consequence of S-unstable models, all probability in the RBM model formulation with a lerge number of visible variables will stack up on mode sets or simply those outcomes for visibles with with the highest probability.
+Proofs of Propositions \ref{prop:instab}-\ref{prop:degen} appear in @kaplan2016note.
+These findings also have counterparts to results in @schweinberger2011instability, but, unlike results there, we do not consider exponential family forms with a fixed number of parameters (i.e.
+our results are broader for a more complicated model form for the visible variables).
+
+To later numerically investigate S-unstable RBM models, it is useful to define a computationally tractable lower bound on (\ref{eqn:elpr}) given by
 $$
-\frac{1}{V}R(\boldsymbol \theta) \text{ for } R(\boldsymbol \theta) \equiv \max_{\boldsymbol v} \max_{\boldsymbol h}Q(\boldsymbol x) - \min_{\boldsymbol v}\max_{\boldsymbol h}Q(\boldsymbol x) - H\log 2.
+\frac{1}{V}R(\boldsymbol \theta_V) \text{ for } R(\boldsymbol \theta_V) \equiv \max_{\boldsymbol v} \max_{\boldsymbol h}Q_{\boldsymbol \theta_V}(\boldsymbol x) - \min_{\boldsymbol v}\max_{\boldsymbol h}Q_{\boldsymbol \theta_V}(\boldsymbol x) - H\log 2.
 $$
@@
 \begin{align}
-\frac{1}{V} \text{ELPR}_V(\boldsymbol \theta) \equiv \frac{1}{V} \log \left[\frac{\max\limits_{(v_1, \dots, v_V) \in \{-1,1\}^V}P_{\boldsymbol \theta}(v_1, \dots, v_V)}{\min\limits_{(v_1, \dots, v_V) \in \{-1,1\}^V}P_{\boldsymbol \theta}(v_1, \dots, v_V)}\right] \ge \frac{1}{V}R(\boldsymbol \theta).
-\label{eqn:lowerbnd}
+\frac{1}{V} \text{ELPR}(\boldsymbol \theta_V) \equiv \frac{1}{V} \log \left[\frac{\max\limits_{(v_1, \dots, v_V) \in \{-1,1\}^V}P_{\boldsymbol \theta}(v_1, \dots, v_V)}{\min\limits_{(v_1, \dots, v_V) \in \{-1,1\}^V}P_{\boldsymbol \theta}(v_1, \dots, v_V)}\right] \ge \frac{1}{V}R(\boldsymbol \theta_V); \label{eqn:lowerbnd}
 \end{align}
-(A proof of (\ref{eqn:lowerbnd}) appears in appendix \ref{instability-proofs}.)
-
-The benefit of (\ref{eqn:lowerbnd}) is that $R(\boldsymbol \theta)/V$ is simple to compute (i.e.
-compared to (\ref{eqn:elpr}), $R(\boldsymbol \theta)$ requires no summation over hidden variables) and that propositions \ref{prop:instab}-\ref{prop:degen} would continue to hold upon replcing (\ref{eqn:elpr}) with $R(\boldsymbol \theta)/V$, due to (\ref{eqn:lowerbnd}).
-Hence, larger values of $R(\boldsymbol \theta)/V$ indicate unstable RBM models.
+the appendix provides a proof of (\ref{eqn:lowerbnd}).
+
+The benefit of $R(\boldsymbol \theta_V)$ is that $R(\boldsymbol \theta_V)/V$ is simple to compute (i.e.
+requiring no summation over hidden variables, unlike (\ref{eqn:elpr})) and that Propositions \ref{prop:instab}-\ref{prop:degen} continue to hold upon replacing (\ref{eqn:elpr}) with $R(\boldsymbol \theta_V)/V$ due to (\ref{eqn:lowerbnd}).
+Hence, large values of $R(\boldsymbol \theta_V)/V$ indicate S-unstable RBM models.
 

### 2016-10-19 writing/draft.Rmd
through edits on uninterpretability

@@
 ### Uninterpretability
-@kaiser2007statistical defines a measure of model impropriety he calls *uninterpretability*, which is characterized by marginal mean-structure not being maintained in the model.
-A measure of this is the magnitude of the difference between model expectations, E$\left[\boldsymbol X | \boldsymbol \theta\right]$, and expectations given independence, E$\left[\boldsymbol X | \boldsymbol \theta^* \right]$, where $\boldsymbol \theta^*$ is defined to equal $\boldsymbol \theta$ with all $\theta_{ij} = 0$ for $i = 1, \dots, V, j = 1, \dots, H$.
-Using this concept it is possible to investigate what conditions lead to uninterpretability in a model versus those that guarantee interpretable models.
-A model is considered without uninterpretability if its expected value is not very different from the same model with assumed independence.
+@kaiser2007statistical defines a measure of model impropriety he calls *uninterpretability*, which is characterized by a marginal mean-structure failing to hold when dependence parameters in a model become too extreme.
+Note that, in a RBM, the parameters $\theta_{v_1}, \dots, \theta_{v_V}$ and $\theta_{h_1}, \dots, \theta_{h_H}$ are naturally associated with main effects of visible and hidden variables and can be interpreted as (logit functions of) means for variables $V_1, \dots, V_V, H_1, \dots, H_V$ in a model with no interaction parameters, $\theta_{ij} = 0, i = 1, \dots, V, j = 1, \dots, H$.
+Hence, the main effect parameters have a clear interpretation under an independence model ($\theta_{ij} = 0$) but this interpretation (as well as an understanding of the nture of dependence) can break down as interaction parameters increase in magnitude (e.g.
+$\max |\theta_{i,j}| \rightarrow \infty$) relative to the size of the main effects.
+To access which parameter values $\boldsymbol \theta$ may cause difficulties in interpretation, we use the difference $\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] - \text{E}\left[\boldsymbol X | \boldsymbol \theta^* \right]$ between model expectations, E$\left[\boldsymbol X | \boldsymbol \theta\right]$, at $\boldsymbol \theta$ and expectations given independence, E$\left[\boldsymbol X | \boldsymbol \theta^* \right]$, where $\boldsymbol \theta^*$ is defined to equal $\boldsymbol \theta$ with all $\theta_{ij} = 0$ for $i = 1, \dots, V, j = 1, \dots, H$.
+Using this, it is possible to investigate what parametric conditions lead to uninterpretability in a model versus those that guarantee interpretable models.
+Uninterpretability is avoided at a parametric specification $\boldsymbol \theta$ if the model expected value at $\boldsymbol \theta$ is not very different from the corresponding model expectation with assumed independence.
 The quantities to compare in the RBM case are
@@
 \text{E}\left[\boldsymbol X | \boldsymbol \theta\right] &= \sum\limits_{\boldsymbol x \in \mathcal{X}} \boldsymbol x f_{\boldsymbol \theta}(\boldsymbol x) \\
-&= \sum\limits_{\boldsymbol x \in \mathcal{X}} \boldsymbol x \frac{\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)} \\
-&= \frac{\sum\limits_{\boldsymbol x \in \mathcal{X}} x_k \exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}  \qquad k = 1, \dots, H + V
+&= \sum\limits_{\boldsymbol x \in \mathcal{X}} \boldsymbol x \frac{\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}
 \end{align*}
@@
 \text{E}\left[\boldsymbol X | \boldsymbol 
-theta^*\right] &= \sum\limits_{\boldsymbol x \in \mathcal{X}} \boldsymbol x \frac{\exp\left(\sum\limits_{i = 1}^V \theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)} \\
-&= \frac{\sum\limits_{\boldsymbol x \in \mathcal{X}} x_k \exp\left(\sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}  \qquad k = 1, \dots, H + V
+\theta^*\right] &= \sum\limits_{\boldsymbol x \in \mathcal{X}} \boldsymbol x \frac{\exp\left(\sum\limits_{i = 1}^V \theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)} \\
 \end{align*}
 
-If $|\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] - \text{E}\left[\boldsymbol X | \emptyset\right]$ is large then the RBM with parameter vector $\boldsymbol \theta$ $is uninterpetable.
+If $||\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] - \text{E}\left[\boldsymbol X | \boldsymbol \theta^*\right]||$ is large then the RBM model with parameter vector $\boldsymbol \theta$ is said to be uninterpetable.
 

### 2016-10-19 writing/appendix-instability.Rmd
typos and refs

@@
 \begin{align*}
-&\log\left[\frac{\max\limits_{(v_1, \dots, v_V) \in \{-1, 1\}^V}P_{\boldsymbol \theta}(v_1, \dots, v_V)}{\min\limits_{(v_1, \dots, v_V) \in \{-1, 1\}^V}P_{\boldsymbol \theta}(v_1, \dots, v_V)} \right] = \\
+&\log\left[\frac{\max\limits_{(v_1, \dots, v_V) \in \{-1, 1\}^V}P_{\boldsymbol \theta_V}(v_1, \dots, v_V)}{\min\limits_{(v_1, \dots, v_V) \in \{-1, 1\}^V}P_{\boldsymbol \theta_V}(v_1, \dots, v_V)} \right] = \\
 & \qquad \log\left[\frac{\max\limits_{(v_1, \dots, v_V)}\sum\limits_{\boldsymbol h \in \{-1,1\}^H} \exp\left(\sum\limits_{i = 1}^V \theta_{v_i}v_i + \sum\limits_{j = 1}^H \theta_{h_j}h_j + \sum\limits_{i = 1}^V\sum\limits_{j = 1}^H \theta_{ij}v_i h_j\right)}
@@
 & \qquad \qquad \min\limits_{(v_1, \dots, v_V)}\max\limits_{\boldsymbol h \in \{-1,1\}^H}\left\{\sum\limits_{i = 1}^V \theta_{v_i}v_i + \sum\limits_{j = 1}^H \theta_{h_j}h_j + \sum\limits_{i = 1}^V\sum\limits_{j = 1}^H \theta_{ij}v_i h_j\right\} -H\log2 \\
-& \qquad \equiv R(\boldsymbol \theta)
+& \qquad \equiv R(\boldsymbol \theta_V)
 \end{align*}

### 2016-10-19 writing/draft.Rmd
typos and refs

@@
 
-The data mining and machine learning communities have recently found a great interest in deep learning, specifically in stacked restricted Boltzmann machines (RBMs).
-(**TODO: deep learning, deep RBM references.**) A RBM is an undirected probabilistic graphical model (for discrete or continuous random variables) with two layers, one hidden and one visible, with no conditional dependency within a layer [@smolensky1986information].
-These models have reportedly been used with success in classification of images.
-(**TODO: classification references.**) However, the model properties are largely unexplored in the literature and the commonly cited fitting methodology remains heuristic-based and abstruse (**TODO: citation**).
+The data mining and machine learning communities have recently found a great interest in deep learning, specifically in stacked restricted Boltzmann machines (RBMs) [see @salakhutdinov2009deep, @salakhutdinov2012efficient, @srivastava2013modeling, @le2008representational for examples].
+A RBM is an undirected probabilistic graphical model (for discrete or continuous random variables) with two layers, one hidden and one visible, with no conditional dependency within a layer [@smolensky1986information].
+These models have reportedly been used with success in classification of images [@larochelle2008classification, @srivastava2012multimodal].
+However, the model properties are largely unexplored in the literature and the commonly cited fitting methodology remains heuristic-based and abstruse [@hinton2006fast].
 In this paper, we attempt to provide steps toward a thorough understanding of the model and its behavior from the perspective of statistical model theory and then explore the possibility of a rigorous fitting methodology.
@@
 
-where $\boldsybol \theta \equiv (\theta_{11}, \dots, \theta_{1H}, \dots, \theta_{V1}, \dots, \theta_{VH}, \theta_{v_1}, \dots, \theta_{v_V}, \theta_{h_1}, \dots, \theta_{h_H}) \in \mathcc{R}^{V + H + V*H}$ denotes hte vector of model parameters and the denominator $\gamma(\boldsymbol \theta)$ is the normalizing function that ensures the distribution sums to one, $\gamma(\boldsymbol \theta) = \sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)$.
+where $\boldsymbol \theta \equiv (\theta_{11}, \dots, \theta_{1H}, \dots, \theta_{V1}, \dots, \theta_{VH}, \theta_{v_1}, \dots, \theta_{v_V}, \theta_{h_1}, \dots, \theta_{h_H}) \in \mathbb{R}^{V + H + V*H}$ denotes hte vector of model parameters and the denominator $\gamma(\boldsymbol \theta)$ is the normalizing function that ensures the distribution sums to one, $\gamma(\boldsymbol \theta) = \sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)$.
 Let $\boldsymbol \tau = \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{X}\} = \{v_i, h_j, v_i h_j: i = 1, \dots, V, j = 1, \dots H\}$ be the vector of sufficient statistics in the model and write $Q_{\boldsymbol \theta}(\boldsymbol x) = \sum\limits_{i = 1}^H\sum\limits_{j=1}^V \theta_{ij} h_i v_j + \sum\limits_{i = 1}^H\theta_{hi} h_i + \sum\limits_{j = 1}^V\theta_{vj} v_j$, the neg-potential function.
@@
 
-**TODO: Add comparison to contrastive divergence**
-
 \begin{appendix}

### 2016-10-22 writing/draft.Rmd
2nd edits from Nordman

@@
 abstract: |
-  A restricted Boltzmann machine (RBM) is an undirected graphical model constructed for discrete or continuous random variables, with two layers, one hidden and one visible, with no conditional dependency within a layer.
+  A restricted Boltzmann machine (RBM) is an undirected graphical model constructed for discrete or continuous random variables, and two layers, one hidden and one visible, with no conditional dependency within a layer.
 In recent years, RBMs have risen to prominence due to their connection to deep learning.
@@
 However, the generative behavior of RBMs has largely been unexplored.
-In this paper, we discuss the relationship between parameter RBM specification and the prevalence of degeneracy, instability and/or uninterpretability in such models in the binary case.
-We also describe the difficulties that arise in likelihood-based fitting such highly flexible models, especially as Gibbs sampling methods are often advocated for the RBM model structure.
+In this paper, we discuss the relationship between RBM parameter specification in the binary case and the prevalence of undesirable model behaviors such as degeneracy, instability and/or uninterpretability.
+We also describe the difficulties that arise in likelihood-based fitting of such highly flexible models, especially as Gibbs sampling (quasi-Bayes) methods are often advocated for the RBM model structure.
 ---
@@
 The data mining and machine learning communities have recently found a great interest in deep learning, specifically in stacked restricted Boltzmann machines (RBMs) [see @salakhutdinov2009deep, @salakhutdinov2012efficient, @srivastava2013modeling, @le2008representational for examples].
-A RBM is an undirected probabilistic graphical model (for discrete or continuous random variables) with two layers, one hidden and one visible, with no conditional dependency within a layer [@smolensky1986information].
+A RBM is an undirected probabilistic graphical model (for discrete or continuous random variables) with two layers, one hidden and one visible, and with no conditional dependency within a layer [@smolensky1986information].
 These models have reportedly been used with success in classification of images [@larochelle2008classification, @srivastava2012multimodal].
 However, the model properties are largely unexplored in the literature and the commonly cited fitting methodology remains heuristic-based and abstruse [@hinton2006fast].
-In this paper, we attempt to provide steps toward a thorough understanding of the model and its behavior from the perspective of statistical model theory and then explore the possibility of a rigorous fitting methodology.
+In this paper, we attempt to provide steps toward a thorough understanding of the model and its behavior from the perspective of statistical theory and then explore the possibility of a rigorous fitting methodology.
 We find the RBM model class to be deficient in two fundamental ways.
 
-First, the models are unsatisfactory as a conceptualization of how data are generated.
+First, the models are often unsatisfactory as a conceptualization of how data are generated.
 According to Fisher, the aim of a statistical model is to represent data in a compact way [@fisher1922mathematical].
 Neyman and Box agree with this assessment and go further to state that a model should "provide an explanation of the mechanism underlying the observed phenomena" [@lehmann1990model, @box1967discrimination].
-RBMs can often generate data lacking realistic variability and thereby fail to provide a satisfactory conceptualization of the data generation process.
-In addition to such degeneracy, we find that RBMs easily exhibit types of instability in the parameter space.
+At issue, RBMs can often produce data lacking realistic variability and may thereby fail to provide a satisfactory conceptualization of the data generation process.
+In addition to such degeneracy, we find that RBMs can easily exhibit types of instability in the parameter space.
 In practice, this may seen when a single pixel change in an image results in a wildly different classification.
-Such model impropriety issues have  been documented in other deep architectures recently [@szegedy2013intriguing, @nguyen2014deep].
-We investigate these phenomena in RBMs in Section \ref{explorations} through simulations of small, manageable examples.
+Such model impropriety issues have recently been documented in RBMs [@li2014biclustering], as well as other deep architectures [@szegedy2013intriguing, @nguyen2014deep].
+We investigate and these phenomena for RBMs in Section \ref{explorations} through simulations of small, manageable examples.
 
@@
 But even when fitting is feasible, there are unique challenges due to the overparameterization of RBMs.
-The literature often suggests MCMC approximations to fit these models (exploiting conditional structure in the hidden and visible variables), but little is said about monitoring convergence [hinton2010practical, @hinton2006fast].
-Related to this, the MCMC algorithms involve updating potentially many latent variables (hiddens) per observation (visibles) which can critically influence convergence in MCMC-based likelihood methods.
-In Section \ref{bayesian-model-fitting}, we compare three Bayesian modelling techniques (which can be computationally more accessible than direct maximum likelihood), each proposed with the goal of avoiding parts of the RBM parameter space that yield improper models.
+The literature often suggests Markov chain Monet Carlo (MCMC) approximations to fit these models (e.g., Gibbs sampling to exploit conditional structure in the hidden and visible variables), but little is said about monitoring convergence [@hinton2010practical, @hinton2006fast].
+Related to this, these MCMC algorithms require updating potentially many latent variables (hiddens) per observation (visibles) which can critically influence convergence in MCMC-based likelihood methods.
+In Section \ref{bayesian-model-fitting}, we compare three Bayesian modeling techniques (involving MCMC approximations that are computationally more accessible than direct maximum likelihood), each proposed with the goal of avoiding parts of the RBM parameter space that yield improper models.
 As might be expected, with greater computational complexity comes an increase in fitting accuracy, but at the cost of feasibility.
-Additionally, because these models are highly overparametrized, any fully principled fitting method based on the likelihood will simply reproduce the empirical distribution of most realistic training sets.
-There is no "smoothing" achieved with a RBM fit using a rigorous likelihood-based method.
+Additionally, because RBM models are highly overparametrized, any fully principled fitting method based on the likelihood will simply reproduce the (discrete) empirical distribution of most realistic training sets.
+That is, there is no "smoothed distribution" achieved in fitting a RBM model with a rigorous likelihood-based method.
 We are therefore led to be skeptical that any model built using these structures, like a deep Boltzmann machine, can achieve prediction or inference in a principled way.
 
-This paper is structured as follows, Section \ref{background} gives a definition of the RBM including the joint distribution of hidden and visible variables and shows the connection to deep learning.
+This paper is structured as follows.
+Section \ref{background} formally definesthe RBM including the joint distribution of hidden and visible variables and explains the model's connection to deep learning.
 Additionally, measures of model impropriety and methods of detecting them are defined.
 Section \ref{explorations} details our explorations into the model behavior and propriety (or lack thereof) of the RBM class.
-We discuss three Bayesian fitting techniques with the goal of avoiding model impropriety in Section \ref{bayesian-model-fitting} and conclude with a discussion in Section \ref{discussion}.
-Additionally, there is an accompanying appendix with auxiliary proofs relating to the choice of data encoding and propositions on instability.
+We discuss three Bayesian fitting techniques with the goal of avoiding model impropriety in Section \ref{model-fitting} and conclude with a discussion in Section \ref{discussion}.
+Additionally, there is an accompanying appendix with proofs of main results.
 
@@
 A restricted Boltzmann machine (RBM) is an undirected graphical model specified for discrete or continuous random variables, binary variables being most common.
-In this paper, we consider the binary case of an RBM for concreteness.
-A RBM architecture has two layers, hidden ($\mathcal{H}$) and visible ($\mathcal{V}$), with no connections within a layer.
+In this paper, we consider the binary case of a RBM for concreteness.
+A RBM architecture has two layers, hidden ($\mathcal{H}$) and visible ($\mathcal{V}$), with no dependency connections within a layer.
 An example of this structure is in Figure \ref{fig:rbm} with the hidden nodes indicated by gray circles and the visible nodes indicated by white circles.
@@
 ### Joint distribution
-Let $\boldsymbol x = \{h_1, \dots, h_H, v_1,\dots,v_V\}$ represent the states of the visible and hidden nodes in an RBM.
+Let $\boldsymbol x = (h_1, \dots, h_H, v_1,\dots,v_V)$ represent the states of the visible and hidden nodes in an RBM for some integers $V, H \ge 1$.
 We will consider both $\boldsymbol x \in \{0,1\}^{H + V}$ and $\boldsymbol x \in \{-1, 1\}^{H + V}$, as the potential encodings for a binary $H + V$-dimensional space, $\mathcal{X}$.
@@
 
-where $\boldsymbol \theta \equiv (\theta_{11}, \dots, \theta_{1H}, \dots, \theta_{V1}, \dots, \theta_{VH}, \theta_{v_1}, \dots, \theta_{v_V}, \theta_{h_1}, \dots, \theta_{h_H}) \in \mathbb{R}^{V + H + V*H}$ denotes hte vector of model parameters and the denominator $\gamma(\boldsymbol \theta)$ is the normalizing function that ensures the distribution sums to one, $\gamma(\boldsymbol \theta) = \sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)$.
-Let $\boldsymbol \tau = \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{X}\} = \{v_i, h_j, v_i h_j: i = 1, \dots, V, j = 1, \dots H\}$ be the vector of sufficient statistics in the model and write $Q_{\boldsymbol \theta}(\boldsymbol x) = \sum\limits_{i = 1}^H\sum\limits_{j=1}^V \theta_{ij} h_i v_j + \sum\limits_{i = 1}^H\theta_{hi} h_i + \sum\limits_{j = 1}^V\theta_{vj} v_j$, the neg-potential function.
-Note that the RBM model is parameterized by $\boldsymbol \theta$containing two types of parameters, main effects and interaction effects.
-The main effects parameters ($\theta_{v_i}, \theta_{h_j}$) weight the values of the visible and hidden nodes and the interaction effects parameters ($\theta_{ij}$) weight the values of the connections, or dependencies, between hidden and visible layers.
+where $\boldsymbol \theta \equiv (\theta_{11}, \dots, \theta_{1H}, \dots, \theta_{V1}, \dots, \theta_{VH}, \theta_{v_1}, \dots, \theta_{v_V}, \theta_{h_1}, \dots, \theta_{h_H}) \in \mathbb{R}^{V + H + V*H}$ denotes the vector of model parameters and the denominator $\gamma(\boldsymbol \theta) = \sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)$ is the normalizing function to ensure the probabilities (\ref{eqn:pmf}) sums to one.
+For
+\begin{align}
+\boldsymbol t(\boldsymbol x) = (h_1, \dots, h_H, v_1, \dots, v_V, v_1 h_1, \dots, v_V h_H) \in \{0,1\}^{H + V + H*V}, \boldsymbol x = (h_1, \dots, h_H, v_1, \dots, v_V) \in \mathcal{X} = \{0,1\}^{V + H}, \label{eqn:t}
+\end{align}
+let $\boldsymbol \tau = \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{X}\}$ be the vector of random variables needed to compute probabilities (\ref{eqn:pmf}) in the model and write $Q_{\boldsymbol \theta}(\boldsymbol x) = \sum\limits_{i = 1}^H\sum\limits_{j=1}^V \theta_{ij} h_i v_j + \sum\limits_{i = 1}^H\theta_{hi} h_i + \sum\limits_{j = 1}^V\theta_{vj} v_j$ for the neg-potential function.
+Note that the RBM model is parameterized by $\boldsymbol \theta$ containing two types of parameters, main effects and interaction effects.
+The main effects parameters ($\{\theta_{v_i}, \theta_{h_j}\}_{\substack{i = 1, \dots, V,\\j = 1, \dots, H}}$) weight the values of the visible $v_i$ and hidden $h_j$ nodes in probabilities (\ref{eqn:pmf}) and the interaction effects parameters ($\theta_{ij}$) weight the values of the connections $v_i h_j$, or dependencies, between hidden and visible layers.
 
 Due to the potential size of the model, the normalizing constant $\gamma(\boldsymbol \theta)$ can be practically impossible to calculate, making it impossible to estimate the model parameter vector in an obvious fashion.
-A kind of Gibbs sampling can be tried, which is natural due to the restricted architecture of conditioal independence of the RBM (i.e.
+A kind of Gibbs sampling can be tried, which is natural due to conditional independence architecture within the RBM (i.e.
 visibles given hiddens or vice versa).
-Specifically, the conditional independence of nodes in each layer given those in the other allows for simulation of the visible and hidden layers.
+Specifically, the conditional independence of nodes in each layer, given those nodes in the other layer allows for stepwise simulation of both the visible and hidden layers.
 
@@
 In Random Graph Model theory, *model degeneracy* means there is a disproportionate amount of probability placed on only a few elements of the sample space, $\mathcal{X}$, by the model [@handcock2003assessing].
-In the scenario of random graph models, $\mathcal{X}$ denotes all possible graphs that can be constructed from a set of nodes.
-An exponentially parameterized random graph model has a distribution of the form
+For random graph models, $\mathcal{X}$ denotes all possible graphs that can be constructed from a set of nodes and an exponentially parameterized random graph model has a distribution of the form
 \begin{align*}
-f_{\boldsymbol \theta} (\boldsymbol x) = \frac{\exp\left(\boldsymbol \theta^T t(\boldsymbol x)\right)}{\gamma(\boldsymbol \theta)}
+f_{\boldsymbol \theta} (\boldsymbol x) = \frac{\exp\left(\boldsymbol \theta^T \boldsymbol t(\boldsymbol x)\right)}{\gamma(\boldsymbol \theta)}, \boldsymbol x \in \mathcal{X},
 \end{align*}
 where $\boldsymbol \theta \in \Theta \subset \mathbb{R}^q$ is the model parameter, and $t: \mathcal{X} \rightarrow \mathbb{R}^q$ are statistics based on the adjacency matrix of a graph.
-Again, $\gamma(\boldsymbol \theta)$ is the normalizing function that guarantees the distribution is valid, e.g.
-$\gamma(\boldsymbol \theta) = \sum_{\boldsymbol x \in \mathcal{X}} \exp\left(\boldsymbol \theta^T t(\boldsymbol x)\right)$.
-Let $C$ denote the convex hull spanned by the potential outcomes of sufficient statistics, $\{t(\boldsymbol x): \boldsymbol x \in \mathcal{X}\}$.
-@handcock2003assessing classify an exponentially parametrized random graph model at $\boldsymbol \theta$ as *near-degenerate* if the mean value of sufficient statistics under $\boldsymbol \theta$, $\mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta}t(X)$, is close to the boundary of $C$.
+Again, $\gamma(\boldsymbol \theta) = \sum_{\boldsymbol x \in \mathcal{X}} \exp\left(\boldsymbol \theta^T \boldsymbol t(\boldsymbol x)\right)$ is the normalizing function for a valid distribution.
+Let $C$ denote the convex hull spanned by the potential outcomes of sufficient statistics, $\{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{X}\}$.
+@handcock2003assessing classify an exponentially parametrized random graph model at $\boldsymbol \theta$ as *near-degenerate* if the mean value of sufficient statistics under $\boldsymbol \theta$, $\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta}t(X)$, is close to the boundary of $C$.
 
 The RBM class of models can be thought to exhibit an analogous occurrence of *near-degeneracy* as follows.
-When random variables $t(\boldsymbol x) = \{v_1, \dots, v_V, h_1, \dots, h_H, v_1 h_1, \dots, v_V h_H \} \in \mathcal{X}$ in the neg-potential function $Q_{\boldsymbol \theta}(\cdot)$, have a mean vector $\mu(\boldsymbol \theta)$ close to the boundary of the convex hull of $\mathcal{X}$, then the RBM model is near-degenerate at $\boldsymbol \theta \in \mathbb{R}^{V + H + H*V}$.
-Here the mean of $t(\boldsymbol x)$ isgiven as
-\begin{align*}
-\mu(\boldsymbol \theta) &= \text{E}_{\boldsymbol \theta} t(\boldsymbol X)  \\
-&= \sum\limits_{x \in \mathcal{X}} \left\{ t(\boldsymbol x) f_{\boldsymbol \theta} (\boldsymbol x) \right\} \\
-&= \sum\limits_{x \in \mathcal{X}} \left\{ t(\boldsymbol x)\dfrac{\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}\right\} \\
-\end{align*}
+When random variables $\boldsymbol t(\boldsymbol x) = (v_1, \dots, v_V, h_1, \dots, h_H, v_1 h_1, \dots, v_V h_H ) \in \boldsymbol \tau \equiv \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{X} = \{0,1\}^{V + H}\} \subset \{0,1\}^{V + H + V*H}$ in the neg-potential function $Q_{\boldsymbol \theta}(\cdot)$ have a mean vector $\boldsymbol \mu(\boldsymbol \theta)$ close to the boundary of the convex hull of $\boldsymbol \tau$, then the RBM model can be said to be near-degenerate at $\boldsymbol \theta \in \mathbb{R}^{V + H + H*V}$.
+Here the mean of $\boldsymbol t(\boldsymbol x)$ is given as
+$$
+\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta} \boldsymbol t(\boldsymbol X)  = \sum\limits_{x \in \mathcal{X}} \left\{ \boldsymbol t(\boldsymbol x) f_{\boldsymbol \theta} (\boldsymbol x) \right\} = \sum\limits_{x \in \mathcal{X}} \left\{ \boldsymbol t(\boldsymbol x)\dfrac{\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}\right\}.
+$$
 
@@
 Considering a class of exponential families of distributions, @schweinberger2011instability introduced a concept of model defficiency related to *instability*.
-Instability can be roughly characterized by excessive sensitivity in the model, where small changes in the components of potential data outcomes, $\boldsymbol x$, may lead to substantial changes in the probability mass function.
-To quantify "instability" more rigorously (particularly beyond the definition given by @schweinberger2011instability), it is useful to imagine how a data model (the RBM in particular) might be expanded to incorporate more and more observations.
+Instability can be roughly characterized by excessive sensitivity in the model, where small changes in the components of potential data outcomes, $\boldsymbol x$, may lead to substantial changes in the probability function $f_{\boldsymbol \theta}(\boldsymbol x)$.
+To quantify "instability" more rigorously (particularly beyond the definition given by @schweinberger2011instability), it is useful to imagine how the RBM model might be expanded to incorporate more and more observations.
 When increasing the size of a RBM model to handle more visible variables, it becomes necessary to expand the number of model parameters (and in this process one may also arbitrarily expand the number of hidden variables to be included).
-To this end, let $\boldsymbol \theta_{V} \equiv (\theta_{v_1}, \dots, \theta_{v_V}, \theta_{h_1}, \dots, \theta_{h_H}, \theta_{11}, \theta_{VH}), V \ge 1$, denote a sequence of RBM parameters indexed by the number $V$ of visibles $(V_1, \dots, V_V)$ and define a (scaled) extremal log-probability ratio of the RBM model at $\boldsymbol \theta_{V}$ $(V \ge 1)$ as
+To this end, let $\boldsymbol \theta_{V} \equiv (\theta_{v_1}, \dots, \theta_{v_V}, \theta_{h_1}, \dots, \theta_{h_H}, \theta_{11}, \dots, \theta_{VH}), V \ge 1$, denote a sequence of RBM parameters indexed by the number $V$ of visibles $(V_1, \dots, V_V)$ and define a (scaled) extremal log-probability ratio of the RBM model at $\boldsymbol \theta_{V}$ as
 \begin{align}
@@
 
-Essentially, in formulating a RBM model for a potentially large number of visibles (i.e., $V \rightarrow \infty$),the ratio (\ref{eqn:elpr}) should stay bounded, meaning that the largest probability under this model at parameter $\boldsymbol \theta_V$ should maintain a magnitude within the smallest probability allowed under the same model.
+In formulating a RBM model for a potentially large number of visibles (i.e., $V \rightarrow \infty$), we will see that the ratio (\ref{eqn:elpr}) needs to stay bounded for the RBM model to be stable.
 This leads to the following definition:
 
-\begin{defn}[unstable RBM]
+\begin{defn}[S-unstable RBM]
 Let $\boldsymbol \theta_V \in \mathbb{R}^{V + H +  H*V}, v \ge 1$, be a sequence of RBM parameters where the number of hiddens, $H \equiv H(V) \ge 1$, can be an arbitrary function of the number $V$ of visibles.
@@
 
-Note that this definition of *S-unstable* is a generalization or re-interpretation of "unstable" used in @schweinberger2011instability allowing nonexponential family models for visibles $(v_1, \dots, v_V)$ and an increasing number of parameters.
-While this definition appears different, it matchs that of @schweinberger2011instability for the special exponential models considered there.
-
-S-Unstable RBM models are undesirable for several reasons. One is that, as mentioned above, small changes in data can lead to overly-sensitive changes in probability. Consider, for example, 
+Note that this definition of *S-unstable* is a generalization or re-interpretation of "unstable" used in @schweinberger2011instability by allowing nonexponential family models for visibles $(v_1, \dots, v_V)$ and an increasing number of parameters.
+
+S-unstable RBM models are undesirable for several reasons. One is that, as mentioned above, small changes in data can lead to overly-sensitive changes in probability. Consider, for example, 
 $$
@@
 $$
-the biggest log-probability ratio for a one-pixel (one component) change in data outcomes (visibles) at a RBM parameter $\boldsymbol \theta_V$. We then have the following result.
+denoting the biggest log-probability ratio for a one-pixel (one component) change in data outcomes (visibles) at a RBM parameter $\boldsymbol \theta_V$. We then have the following result.
 
@@
 
-Additionally, S-unstable RBM models are connected to degenerate models  placing all probability on a small portion of the sample space. To see this, define a modal set 
+Additionally, S-unstable RBM models are connected to degenerate models  placing all probability on a small portion of the sample space.
+To see this, define a modal set of possible outcomes (i.e.
+set of highest probability outcomes) in RBM models with parameters $\boldsymbol \theta_V, V \ge 1$ as
 $$
@@
 $$
-of possible outcomes, for a given $0 < \epsilon < 1$, and along a sequence $\boldsymbol \theta_V, V \ge 1$, of RBM parameters.
+for a given $0 < \epsilon < 1$. Then S-unstable models are guaranteed to be degenerate, as the following result shows.
 
@@
 \label{prop:degen}
-For an S-unstable RBM model, and for any given $0 < \epsilon < 1$, $P_{\boldsymbol \theta_V}\left((v_1, \dots, v_V) \in M_{\epsilon, \boldsymbol \theta_V}\right) \rightarrow 1$ holds as $V \rightarrow \infty$.
+For an S-unstable RBM model and any given $0 < \epsilon < 1$, $P_{\boldsymbol \theta_V}\left((v_1, \dots, v_V) \in M_{\epsilon, \boldsymbol \theta_V}\right) \rightarrow 1$ holds as $V \rightarrow \infty$.
 \end{prop}
 
-In other words, as a consequence of S-unstable models, all probability in the RBM model formulation with a lerge number of visible variables will stack up on mode sets or simply those outcomes for visibles with with the highest probability.
+In other words, in S-unstable RBM models with a large number of visible variables, all probability in the RBM model formulation  will stack up on those outcomes for visibles with with the highest probability.
 Proofs of Propositions \ref{prop:instab}-\ref{prop:degen} appear in @kaplan2016note.
-These findings also have counterparts to results in @schweinberger2011instability, but, unlike results there, we do not consider exponential family forms with a fixed number of parameters (i.e.
-our results are broader for a more complicated model form for the visible variables).
-
+These findings also have counterparts to results in @schweinberger2011instability, but, unlike results there, we do not consider exponential family forms with a fixed number of parameters.
+
+\begin{remark}
+\label{remark:bnd}
 To later numerically investigate S-unstable RBM models, it is useful to define a computationally tractable lower bound on (\ref{eqn:elpr}) given by
-$$
+\begin{align}
 \frac{1}{V}R(\boldsymbol \theta_V) \text{ for } R(\boldsymbol \theta_V) \equiv \max_{\boldsymbol v} \max_{\boldsymbol h}Q_{\boldsymbol \theta_V}(\boldsymbol x) - \min_{\boldsymbol v}\max_{\boldsymbol h}Q_{\boldsymbol \theta_V}(\boldsymbol x) - H\log 2.
-$$
+\label{eqn:r}
+\end{align}
 That is, it holds that, in (\ref{eqn:elpr}),
 \begin{align}
-\frac{1}{V} \text{ELPR}(\boldsymbol \theta_V) \equiv \frac{1}{V} \log \left[\frac{\max\limits_{(v_1, \dots, v_V) \in \{-1,1\}^V}P_{\boldsymbol \theta}(v_1, \dots, v_V)}{\min\limits_{(v_1, \dots, v_V) \in \{-1,1\}^V}P_{\boldsymbol \theta}(v_1, \dots, v_V)}\right] \ge \frac{1}{V}R(\boldsymbol \theta_V); \label{eqn:lowerbnd}
+\frac{1}{V} \text{ELPR}(\boldsymbol \theta_V) \equiv \frac{1}{V} \log \left[\frac{\max\limits_{(v_1, \dots, v_V) \in \{-1,1\}^V}P_{\boldsymbol \theta}(v_1, \dots, v_V)}{\min\limits_{(v_1, \dots, v_V) \in \{-1,1\}^V}P_{\boldsymbol \theta}(v_1, \dots, v_V)}\right] \ge \frac{1}{V}R(\boldsymbol \theta_V).
+\label{eqn:lowerbnd}
 \end{align}
-the appendix provides a proof of (\ref{eqn:lowerbnd}).
-
+\end{remark}
+The appendix provides a proof of (\ref{eqn:lowerbnd}).
 The benefit of $R(\boldsymbol \theta_V)$ is that $R(\boldsymbol \theta_V)/V$ is simple to compute (i.e.
 requiring no summation over hidden variables, unlike (\ref{eqn:elpr})) and that Propositions \ref{prop:instab}-\ref{prop:degen} continue to hold upon replacing (\ref{eqn:elpr}) with $R(\boldsymbol \theta_V)/V$ due to (\ref{eqn:lowerbnd}).
-Hence, large values of $R(\boldsymbol \theta_V)/V$ indicate S-unstable RBM models.
+Hence, large values of $R(\boldsymbol \theta_V)/V$ also indicate S-unstable RBM models.
 
 ### Uninterpretability
-@kaiser2007statistical defines a measure of model impropriety he calls *uninterpretability*, which is characterized by a marginal mean-structure failing to hold when dependence parameters in a model become too extreme.
+For spatial Markov models, kaiser2007statistical defines a measure of model impropriety he calls *uninterpretability*, which is characterized by dependence parameters in a model being so extreme that marginal mean-structures fail to hold as anticipated by the model statement.
+We adapt this notion to RBM models as follows.
 Note that, in a RBM, the parameters $\theta_{v_1}, \dots, \theta_{v_V}$ and $\theta_{h_1}, \dots, \theta_{h_H}$ are naturally associated with main effects of visible and hidden variables and can be interpreted as (logit functions of) means for variables $V_1, \dots, V_V, H_1, \dots, H_V$ in a model with no interaction parameters, $\theta_{ij} = 0, i = 1, \dots, V, j = 1, \dots, H$.
-Hence, the main effect parameters have a clear interpretation under an independence model ($\theta_{ij} = 0$) but this interpretation (as well as an understanding of the nture of dependence) can break down as interaction parameters increase in magnitude (e.g.
+Hence, these main effect parameters have a clear interpretation under an independence model ($\theta_{ij} = 0$) but this interpretation (as well as an understanding of the nature of dependence) can break down as interaction parameters increase in magnitude (e.g.
 $\max |\theta_{i,j}| \rightarrow \infty$) relative to the size of the main effects.
-To access which parameter values $\boldsymbol \theta$ may cause difficulties in interpretation, we use the difference $\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] - \text{E}\left[\boldsymbol X | \boldsymbol \theta^* \right]$ between model expectations, E$\left[\boldsymbol X | \boldsymbol \theta\right]$, at $\boldsymbol \theta$ and expectations given independence, E$\left[\boldsymbol X | \boldsymbol \theta^* \right]$, where $\boldsymbol \theta^*$ is defined to equal $\boldsymbol \theta$ with all $\theta_{ij} = 0$ for $i = 1, \dots, V, j = 1, \dots, H$.
+In which case, the main effect parameters $\theta_{v_1}$ and $\theta_{h_j}$ are no longer interpretable as originally intended in the modes (statements of marginal means) and the dependence parameters are so large as to dominate the entire model probability structure (losing the model interpretation of dependence as local, conditional modifications of all overall marginal mean structure).
+To access which parameter values $\boldsymbol \theta$ may cause difficulties in interpretation, we use the difference $\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] - \text{E}\left[\boldsymbol X | \boldsymbol \theta^* \right]$ between two model expectations: E$\left[\boldsymbol X | \boldsymbol \theta\right]$ at $\boldsymbol \theta$ and expectations E$\left[\boldsymbol X | \boldsymbol \theta^* \right]$ given independence where $\boldsymbol \theta^*$ is matches $\boldsymbol \theta$ for all main effects but otherwise has $\theta_{ij} = 0$ for $i = 1, \dots, V, j = 1, \dots, H$.
+Hence, $\boldsymbol \theta^*$ and $\boldsymbol \theta$ have the same main effects but $\boldsymbol \theta^*$ has no dependence parameters.
+Uninterpretability is then avoided at a parametric specification $\boldsymbol \theta$ if the model expected value at $\boldsymbol \theta$ is not very different from the corresponding model expectation with assumed independence.
 Using this, it is possible to investigate what parametric conditions lead to uninterpretability in a model versus those that guarantee interpretable models.
-Uninterpretability is avoided at a parametric specification $\boldsymbol \theta$ if the model expected value at $\boldsymbol \theta$ is not very different from the corresponding model expectation with assumed independence.
+If $||\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] - \text{E}\left[\boldsymbol X | \boldsymbol \theta^*\right]||$ is large then the RBM model with parameter vector $\boldsymbol \theta$ is said to be uninterpetable.
 The quantities to compare in the RBM case are
 
-\begin{align*}
-\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] &= \sum\limits_{\boldsymbol x \in \mathcal{X}} \boldsymbol x f_{\boldsymbol \theta}(\boldsymbol x) \\
-&= \sum\limits_{\boldsymbol x \in \mathcal{X}} \boldsymbol x \frac{\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}
-\end{align*}
+$$
+\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] = \sum\limits_{\boldsymbol x \in \mathcal{X}} \boldsymbol x f_{\boldsymbol \theta}(\boldsymbol x) = \sum\limits_{\boldsymbol x \in \mathcal{X}} \boldsymbol x \frac{\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}
+$$
 and
@@
 
-If $||\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] - \text{E}\left[\boldsymbol X | \boldsymbol \theta^*\right]||$ is large then the RBM model with parameter vector $\boldsymbol \theta$ is said to be uninterpetable.
-
 # Explorations
 
-We explore the relationship between magnitudes of $\boldsymbol \theta$ and the three measures of model impropriety, near-degeneracy, instability, and uninterpretability, for RBM models of varying size through simulated data.
+We next explore and numerically explain the relationship between values of $\boldsymbol \theta$ and the three measures of model impropriety, near-degeneracy, instability, and uninterpretability, for RBM models of varying size.
 
@@
 A schematic of this model can be found in figure \ref{fig:toymodel}.
-Because it seems most common, we'll begin by employing $0/1$ encoding of binary variables (both $h_1$ and $v_1$ taking values in $\{0,1\}$).
-Eventually we'll note in section \ref{data-coding-to-mitigate-degeneracy} that $-1/1$ coding has substantial advantages.
+Because it seems most common, we shall begin by employing $0/1$ encoding of binary variables (both $h_1$ and $v_1$ taking values in $\{0,1\}$).
+Eventually we'll note in Slinkection \ref{data-coding-to-mitigate-degeneracy} that $-1/1$ coding has substantial advantages.
 
@@
 
-For this small model, we are able to investigate the symptoms of model impropriety (near-degeneracy, instability uninterpretability).
-First, because we are in three dimensions, we are able to explicitly illustrate the shape of boundary of the convex hull of $\boldsymbol tau$ and explore the behavior of the mean parameterization of the parameters.
-In figure \ref{fig:toyhull}, the convex hull of our "statistic space", $\boldsymbol \tau$, in three dimensions for this toy problem is shown from multiple perspectives enclosed by the unit cube, it convex hull of $\{0,1\}^3$ space.
-In this small model, note that the convex hull of $\boldsymbol tau$ does not fill the unrestricted hull because of the relationship between the elements of $\{(v_1, h_1, v_1 h_1 : (v_1, h_1) \in \{0, 1\}^2 \}$.
+For this small model, we are able to investigate the symptoms of model impropriety, beginning with near-degeneracy.
+To this end, recall that one characterization requires the convex hull of statistics $\boldsymbol \tau = \{\boldsymbol t(\boldsymbol x): \boldsymbol x = (v_1, h_1) \in \{0,1\}^2\} \equiv \{(v_1, h_1, v_1 h_1): v_1, h_1 \in \{0,1\}\}$ appearing in the RBM probabilities for this model.
+As this set is in three dimensions, we are able to explicitly illustrate the shape of boundary of the convex hull of $\boldsymbol \tau$ and explore the behavior of the mean parameterization of the parameters.
+Figure \ref{fig:toyhull} shows the convex hull of our "statistic space", $\boldsymbol \tau \subset \{0,1\}^3$, for this toy problem from multiple perspectives enclosed by the unit cube, representing the convex hull of $\{0,1\}^3$ space.
+In this small model, note that the convex hull of $\boldsymbol \tau$ does not fill the unrestricted hull of $\{0,1\}^3$ because of the relationship between the elements of $\boldsymbol \tau = \{(v_1, h_1, v_1 h_1 : v_1, h_1 \in \{0, 1\} \}$ (i.1.
+$v_1 h_1 = 1$ only if $v_1 = h_1 = 1$).
 
@@
 Additionally, we can compute the mean vector for $\boldsymbol t(\boldsymbol x)$ as a function of the model parameters as 
-\begin{align*}
-\boldsymbol \mu(\boldsymbol \theta) &= \text{E}_{\boldsymbol \theta}\left[ \boldsymbol t(\boldsymbol X) \right] \\
-&= \sum\limits_{\boldsymbol x \in \mathcal{X}} \left\{ \boldsymbol t(\boldsymbol x) \frac{\exp\left( \theta_{11} h_1 v_1 + \theta_{h1} h_1 + \theta_{v1} v_1\right)}{\sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left( \theta_{11} h_1 v_1 + \theta_{h1} h_1 + \theta_{v1} v_1\right)} \right\} \\
-&= \sum\limits_{\boldsymbol x \in \mathcal{X}} \left\{ t(x) \frac{\exp\left( \theta_{11} h_1 v_1 + \theta_{h1} h_1 + \theta_{v1} v_1\right)}{\gamma(\boldsymbol \theta)} \right\} \\
-&= \left[
+\begin{align}
+\boldsymbol \mu(\boldsymbol \theta) &= \text{E}_{\boldsymbol \theta}\left[ \boldsymbol t(\boldsymbol X) \right] =  \sum\limits_{\boldsymbol x = (v_1, h_1) \in \mathcal{X} = \{0,1\}^2} \left\{ t(x) \frac{\exp\left( \theta_{11} h_1 v_1 + \theta_{h1} h_1 + \theta_{v1} v_1\right)}{\gamma(\boldsymbol \theta)} \right\}
+= \left[
 \begin{matrix}
@@
 \end{matrix} \right]
-\end{align*}
-The three parametric functions giving $\boldsymbol \mu(\boldsymbol \theta)$ can be visualized in figure \ref{fig:degen_toy}.
-By examining these, we see that as values of coordinates of $\boldsymbol \theta$ grow larger in magnitude, at least one mean function for the entries of $\boldsymbol t(\boldsymbol x)$ is approaching a value 0 or 1, indicating a potential for $\text{E}_{\boldsymbol \theta} \boldsymbol t(\boldsymbol x)$ to be near to the boundary of the convex hull of $\boldsymbol \tau$.
+\end{align}
+where $\gamma(\boldsymbol \theta) = \sum\limits_{h_1 = 0}^1\sum\limits_{v_1 = 0}^1 \exp(\theta_{11} h_1 v_1 + \theta_{h_1}h_1 + \theta_{v_1}v_1)$.
+The three parametric functions giving $\boldsymbol \mu(\boldsymbol \theta)$ can be visualized in Figure \ref{fig:degen_toy}.
+In examining these, we see that as values of coordinates of $\boldsymbol \theta$ grow larger in magnitude, at least one mean function for the entries of $\boldsymbol t(\boldsymbol x)$ is approaching a value 0 or 1, indicating a potential for $\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta} \boldsymbol t(\boldsymbol x)$ to be near to the boundary of the convex hull of $\boldsymbol \tau$, as a sign of model near-degeneracy.
 Thus, for a very small example we can see the relationship between values of $\boldsymbol \theta$ and model degeneracy.
@@
 
-Secondly, we can look at $R(\boldsymbol \theta)$ for this tiny model.
-Recall that gives us a lower bound for the loglikelihood ratio of the minimum possible change in $\boldsymbol x$, indicating instability.
+Secondly, we can look at $R(\boldsymbol \theta)$ from (\ref{eqn:r}) for this tiny model in order to consider model instability as a function of RBM parameters.
+Recall that larger values of $R(\boldsymbol \theta)$ imply an extreme sensitivity of the model probabilities $f_{\boldsymbol \theta}(\boldsymbol x)$ to small changes in $\boldsymbol x$ (see Remark \ref{remark:bnd} and Proposition \ref{prop:instab}).
+The quantity $R(\boldsymbol \theta)$ is computed for this small RBM as
 
 \begin{align*}
-R(\boldsymbol \theta) &= \max_{\boldsymbol v} \max_{\boldsymbol h}Q(\boldsymbol x) - \min_{\boldsymbol v}\max_{\boldsymbol h}Q(\boldsymbol x) - H\log 2\\
+R(\boldsymbol \theta) &= \max_{\boldsymbol v} \max_{\boldsymbol h}Q_{\boldsymbol \theta}(\boldsymbol x) - \min_{\boldsymbol v}\max_{\boldsymbol h}Q_{\boldsymbol \theta}(\boldsymbol x) - H\log 2\\
 &= \max_{\boldsymbol v} \max_{\boldsymbol h}\left[ \theta_{11} h_1 v_1 + \theta_{h_1}h_1 + \theta_{v_1} v_1 \right] -  \min_{\boldsymbol v}\max_{\boldsymbol h}\left[ \theta_{11} h_1 v_1 + \theta_{h_1}h_1 + \theta_{v_1} v_1 \right] -\log 2\\
 &=\max\left\{ \theta_{11} + \theta_{h_1} + \theta_{v_1}, \theta_{h_1}, \theta_{v_1}, 0  \right\} \\
-&\qquad - \min\left\{\theta_{v_1} + \max\left\{\theta_{11} + \theta_{h1}, -\theta_{11} - \theta_{h1}\right\}, -\theta_{v_1} + \max\left\{-\theta_{11} + \theta_{h1}, \theta_{11} - \theta_{h1}\right\}\right\} - \log2\\
+&\qquad - \min\left\{\theta_{v_1} + \max\left\{\theta_{11} + \theta_{h1}, -\theta_{11} - \theta_{h1}\right\}, -\theta_{v_1} + \max\left\{-\theta_{11} + \theta_{h1}, \theta_{11} - \theta_{h1}\right\}\right\} - \log2.\\
 \end{align*}
 
-Figure \ref{fig:instab} shows $R(\boldsymbol \theta)/V$ for various values of $\boldsymbol \theta$.
-We can see that this quantity is increasing for larger magnitudes of $\boldsymbol \theta$, especially for large values of $\theta_{11}$.
-This suggests instability as $|\boldsymbol \theta|$ becomes large, agreeing with the concerns about near-degeneracy produced by consideration of $\boldsymbol \mu(\boldsymbol \theta)$.
+Figure \ref{fig:instab} shows $R(\boldsymbol \theta)/V$ for various values of $\boldsymbol \theta$ in this model with $V = 1$.
+We can see that this quantity is increasing for larger magnitudes of $\boldsymbol \theta$, especially for large values of the dependence/interaction parameter $\theta_{11}$.
+This suggests instability as $|\boldsymbol \theta|$ becomes large, agreeing also with the concerns about near-degeneracy produced by consideration of $\boldsymbol \mu(\boldsymbol \theta)$.
 
@@
 
-Finally, we can look at the difference between model expectations, E$\left[\boldsymbol X | \boldsymbol \theta\right]$, and expectations given independence, E$\left[\boldsymbol X | \boldsymbol \theta^*\right ]$ for the tiny toy model to understand the effect of $\boldsymbol \theta$ on uninterpretability in the RBM.
+Finally to consider the effect of $\boldsymbol \theta$ on potential model uninterpretability, we can look at the difference between model expectations, E$\left[\boldsymbol X | \boldsymbol \theta\right]$, and expectations given independence, E$\left[\boldsymbol X | \boldsymbol \theta^*\right ]$ for the tiny toy RBM model where $\boldsymbol X = (V_1, H_1, V_1 H_1)$; this is given by
 
@@
 \frac{\left|\exp\left(\theta_{11} + 2\theta_{v_1} + \theta_{h_1}\right) - \exp\left( 2\theta_{v_1} + \theta_{h_1}\right) \right| }{\left(\exp\left(\theta_{v_1}\right) + \exp\left(\theta_{h_1}\right) + \exp\left(\theta_{11} + \theta_{v_1} + \theta_{h_1}\right)\right)\left(\exp\left(\theta_{v_1}\right) + \exp\left(\theta_{h_1}\right) + \exp\left(+ \theta_{v_1} + \theta_{h_1}\right)\right)}
-\end{matrix}\right]
+\end{matrix}\right].
 \end{align*}
 
-Again, we can inspect these coordinate functions of $\boldsymbol \theta$ to look for a relationship between parameter values and uninterpretability for the toy RBM.
+Again, we can inspect these coordinate functions of $\boldsymbol \theta$ to look for a relationship between parameter values and large values of $||\text{E}[\boldsymbol X|\boldsymbol \theta] - \text{E}[\boldsymbol X| \boldsymbol \theta^*]||$ as a signal of uninterpretability for the toy RBM.
 
@@
 
-Figure \ref{fig:uninterp} shows that the absolute difference between coordinates of the vector of model expectations, E$\left[\boldsymbol X | \boldsymbol \theta\right]$, and corresponding expectations given independence, E$\left[\boldsymbol X | \boldsymbol \theta^*\right ]$ for the toy RBM grow as the values of $\boldsymbol \theta$ are farther from zero, especially for large magnitudes of $\theta_{11}$.
-This is a third indication that parameter vectors of large magnitude could lead to model impropriety in an RBM.
+Figure \ref{fig:uninterp} shows that the absolute difference between coordinates of the vector of model expectations, E$\left[\boldsymbol X | \boldsymbol \theta\right]$, and corresponding expectations E$\left[\boldsymbol X | \boldsymbol \theta^*\right ]$ given independence grow for the toy RBM as the values of $\boldsymbol \theta$ are farther from zero, especially for large magnitudes of $\theta_{11}$.
+This is a third indication supporting that parameter vectors of large magnitude could lead to model impropriety in a RBM.
 
@@
 Multiple encodings of the binary variables are possible.
-For example, we could allow $\mathcal{H} \in \{0,1\}^H, \mathcal{V} \in \{0,1\}^V$, as in the previous sections or we could encode the state of the variables as $\mathcal{H} \in \{-1,1\}^H, \mathcal{V} \in \{-1,1\}^V$.
-This will result in $\boldsymbol t(\boldsymbol x) \in \{0,1\}^{H + V + HV}$ or $\boldsymbol t(\boldsymbol x) \in \{-1,1\}^{H + V + HV}$ depending on how we encode "on" and "off" states in the nodes.
-To explore the effect of this encoding on the potential for apparent near-degeneracy of an RBM, we can consider the ratio of the volume of the hypercube with corners at elements of $\boldsymbol \tau$ to the volume of the convex hull of $\boldsymbol \tau$.
+For example, we could allow hiddens $(H_1, \dots, H_H) \in \{0,1\}^H$ and visibles $(V_1, \dots, V_V) \in \{0,1\}^V$, as in the previous sections or we could instead encode the state of the variables as $\{-1,1\}^H$ and $\{-1,1\}^V$.
+This will result in variables $\boldsymbol t(\boldsymbol X)$ from (\ref{eqn:t}) satisfying $\boldsymbol t(\boldsymbol x) \in \{0,1\}^{H + V + HV}$ or $\boldsymbol t(\boldsymbol x) \in \{-1,1\}^{H + V + HV}$ depending on how we encode "on" and "off" states in the nodes.
+To explore the effect of this encoding on the potential for apparent near-degeneracy of an RBM, we can consider the ratio of the volume of the hypercube with corners at elements of $\boldsymbol \tau$ to the volume of the convex hull of $\boldsymbol \tau$ under both possible encodings (i.e., compare convex hull of $\boldsymbol \tau$ to either $[0,1]^3$ or $[-1,1]^3$ under the respective encodings).
 
 For the small two node example, the $0/1$ encoding loses `r round((1-m1_bin$c_hull$vol)*100,2)`% of the volume of $[0,1]^3$ and the $-1/1$ encoding loses `r round((1-m1_neg$c_hull$vol/(2^3))*100,2)`% of the volume of $[-1,1]^3$.
-While there is clearly a one-to-one mapping of models for these two possible codings, this notion of lost volume can be helpful to conceptualize how difficult it will be for the mean vector $\boldsymbol \mu(\boldsymbol \theta)$ to avoid the boundary, and thus avoid near-degeneracy.
-In fact, if we look at the ratio of volume within a convex hull defined by $\boldsymbol \tau$ and the corresponding hypercube, we can see a relationship emerge as the number of nodes (and thus parameters) increases.
-In figure \ref{fig:volume_plot}, it is evident that as the number of parameters increases, this ratio is decreasing at an increasing rate, meaning it gets more and more difficult to avoid the boundary of the convex hull and thus near degeneracy.
+While there is clearly a one-to-one mapping of models for these two possible codings, this notion of lost volume can be helpful to conceptualize how difficult it will be for the mean vector $\boldsymbol \mu(\boldsymbol \theta)$ to avoid the boundary of $\boldsymbol \tau$, and thus avoid near-degeneracy.
+In fact, if we look at the ratio of volume within the convex hull defined by $\boldsymbol \tau$ and the corresponding hypercube, we can see a relationship emerge as the number of nodes (and thus parameters) increases.
+In Figure \ref{fig:volume_plot}, it is evident that as the number of parameters increases, this ratio is decreasing at an increasing rate, meaning it gets more and more difficult to avoid the boundary of the convex hull and thus near degeneracy.
 Additionally, it appears that the $-1/1$ encoding suffers slightly less from this problem.
-This suggests that if intuition about models is to be formed in terms of volume fractions of vectors of "non-pathological" means for $\boldsymbol t(\boldsymbol x)$, $-1/1$ encoding is preferable to $0/1$ coding.
+This suggests that if intuition about models is to be formed in terms of a data encoding that encourages of "non-pathological" means $\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta}(\boldsymbol t(\boldsymbol x))$ for $\boldsymbol t(\boldsymbol x)$, then the $-1/1$ encoding is preferable to $0/1$ coding.
 
@@
 
-In addition to the argument that the $-1/1$ data encoding mitigates some perceived prevalence of near-degeneracy it also has the benefit of providing a guaranteed non-degenerate model at $\boldsymbol \theta = \boldsymbol 0 \in \mathbb{R}^{H + V + H*V}$.
+In addition to the argument that the $-1/1$ data encoding mitigates some perceived prevalence of near-degeneracy it also has the benefit of providing a guaranteed non-degenerate model at $\boldsymbol \theta = \boldsymbol 0 \in \mathbb{R}^{H + V + H*V}$, where the zero vector that serves as the natural center of the parameter space and induces model properties for the RBM (i.e., at $\boldsymbol \theta = \boldsymbol 0$, all variables are independent and visible variables are independent and uniformly distributed on $\{-1,1\}^V$.
 The proof of this is included in appendix \ref{center-of-universe}.
-And while from some computing perspectives $0/1$ coding might seem most natural, the $-1/1$ coding is far more convenient and interpretable from the point of view of statistical modeling, where it makes parameters simply interpreted in terms of symmetrically defined main effects and interactions.
+Hence while from some computing perspectives $0/1$ coding might seem most natural, the $-1/1$ coding is far more convenient and interpretable from the point of view of statistical modeling, where it makes parameters simply interpreted in terms of symmetrically defined main effects and interactions.
 In light of all of these matters we will henceforth employ the $-1/1$ coding.
@@
 For $H, V \in \{1, \dots, 4\}$, we sample `r dim(res$samp[[1]])[1]` values of $\boldsymbol \theta$ with various magnitudes (details to follow).
-For each set of parameters we then calculate the three metrics of model impropriety introduced in section \ref{degeneracy-instability-and-uninterpretability-oh-my}, $\boldsymbol \mu(\boldsymbol \theta)$, $R(\boldsymbol x)/V$, and the coordinates of $\left|\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] -   \text{E}\left[\boldsymbol X | \boldsymbol \theta^* \right] \right|$.
+For each set of parameters we then calculate the three metrics of model impropriety introduced in Section \ref{degeneracy-instability-and-uninterpretability-oh-my}, $\boldsymbol \mu(\boldsymbol \theta)$, $R(\boldsymbol x)/V$, and the coordinates of $\left|\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] -   \text{E}\left[\boldsymbol X | \boldsymbol \theta^* \right] \right|$.
 In the case of near-degeneracy, we can go further and classify each model as near-degenerate or "viable" based on the distance of $\boldsymbol \mu(\boldsymbol \theta)$ from the boundary of the convex hull of $\boldsymbol \tau$ and look at the fraction of models that are "near-degenerate", meaning they are within a small distance to the boundary of the convex hull.
 
-In this numerical experiment, we split $\boldsymbol \theta$ into $\boldsymbol \theta_{interaction}$ and $\boldsymbol \theta_{main}$, in reference to which sufficient statistics the parameters correspond to (whether they multiply a $v_i$ or a $h_j$ or they multiply a $v_i h_j$), and allow the two types of terms to have varying average magnitudes, $||\boldsymbol \theta_{main} || /(H+V)$ and $||\boldsymbol \theta_{interaction} || /(H*V)$.
+In this numerical experiment, we split $\boldsymbol \theta$ into $\boldsymbol \theta_{interaction}$ and $\boldsymbol \theta_{main}$, in reference to which variables in the probability function the parameters correspond to (whether they multiply a $v_i$ or a $h_j$ or they multiply a $v_i h_j$), and allow the two types of terms to have varying average magnitudes, $||\boldsymbol \theta_{main} || /(H+V)$ and $||\boldsymbol \theta_{interaction} || /(H*V)$.
 These average magnitudes vary on a grid between `r min(plot_dat_grid$r1/(plot_dat_grid$H + plot_dat_grid$V))` and `r max(plot_dat_grid$r1/(plot_dat_grid$H + plot_dat_grid$V))` with `r length(unique(plot_dat_grid$r1/(plot_dat_grid$H + plot_dat_grid$V)))` breaks, yielding `r length(unique(plot_dat_grid$r1/(plot_dat_grid$H + plot_dat_grid$V)))^2` grid points.
@@
 
-The results of this numerical study are summarized in figures \ref{fig:degen_plots}, \ref{fig:instab_plots}, and \ref{fig:uninterp_plots}.
-From these three figures, it s clear to see that all three measures of model impropriety show higher values the larger the magnitude of the parameter vectors.
+The results of this numerical study are summarized in Figures \ref{fig:degen_plots}, \ref{fig:instab_plots}, and \ref{fig:uninterp_plots}.
+From these three figures, it is clear to see that all three measures of model impropriety show higher values for larger magnitudes of the parameter vectors.
 Additionally, since there are $H*V$ interaction terms in $\boldsymbol \theta$ versus only $H + V$ main effects terms, for large models there are many more interaction parameters than main effects in the model.
-And so, severely limiting the magnitude of the indivisual interactions may well help prevent model impropriety.
+And so, severely limiting the magnitude of the individual interactions may well help prevent model impropriety.
 
@@
 In fact, when the cell probabilities (\ref{eqn:0}) are all strictly positive, parameters in the RBM model can be specified to reflect these probabilities exactly.
-And, when one or more of the cell probabilities (\ref{eqn:0}) are zero, the corresponding RBM  probabilities $P(V_1=v_1,V_2=_v2| \theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21})$ may never be identically zero (due to the exponential model form) but
+And, when one or more of the cell probabilities (\ref{eqn:0}) are zero, the corresponding RBM  probabilities $P(V_1=v_1,V_2=_v2| \theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21})$ may never be identically zero (due to exponential terms in the model) but
  parameters can be still selected to make the appropriate RBM cell probabilities arbitrarily small.
@@
 The previous discussion intends to indicate that the RBM model suffers from parameter identifiability issues that go beyond mere symmetries in the parametrization.
-Quite different parameter settings can induce the same essential RBM model.However, this  is not the most disastrous implication of the RBM parameterization.
+Quite different parameter settings can induce the same essential RBM model.
+However, this  is not the most disastrous implication of the RBM parameterization.
 A far worse consequence is that, when fitting the RBM model by likelihood-based methods, we already know what the answer must be before we begin: namely, such  fitting will simply aim to reproduce the empirical distribution from the data.
@@
 To avoid model impropriety for a fitted RBM, we want to avoid parts of the parameter space $\mathbb{R}^{V + H + VH}$ that lead to near-degeneracy, instability, and uninterpretability.
-Motivated by the insights in section \ref{exploring-manageable-examples}, one idea is to shrink $\boldsymbol \theta$ toward $\boldsymbol 0$ by specifying priors that place low probability on large values of $||\boldsymbol \theta||$, specifically focusing on shrinking $\boldsymbol \theta_{interaction}$ more than $\boldsymbol \theta_{main}$.
+Motivated by the insights in Section \ref{exploring-manageable-examples}, one idea is to shrink $\boldsymbol \theta$ toward $\boldsymbol 0$ by specifying priors that place low probability on large values of $||\boldsymbol \theta||$, specifically focusing on shrinking $\boldsymbol \theta_{interaction}$ more than $\boldsymbol \theta_{main}$.
 
@@
 
-We considered a test case with $V = H = 4$ and parameters given in table \ref{tab:theta}.
+We considered a test case with $V = H = 4$ and parameters given in Table \ref{tab:theta}.
 This parameter vector was chosen as a sampled value of $\boldsymbol \theta$ that was not near the convex hull of the sufficient statistics for a grid point in figure \ref{fig:degen_plots} with $< 5$\% near-degeneracy.
@@
 *A truncated Normal prior.* Here we use independent truncated spherical normal distributions as priors for $\boldsymbol \theta_{main}$ and $\boldsymbol \theta_{interaction}$ with $\sigma_{interaction} < \sigma_{main}$.
-Full conditional distributions are not conjugate, and simulaion from the posterior was accomplished using a geometric adaptive Metropolis Hastings step [@zhou2014some] and calculation of likelihood normalizing constant.
+Full conditional distributions are not conjugate, and simulation from the posterior was accomplished using a geometric adaptive Metropolis Hastings step [@zhou2014some] and calculation of likelihood normalizing constant.
 (This computation is barely feasible for a problem of this size.) Here the $h_j$ are carried along in the MCMC implementation as latent variables.
-3. *A truncated Normal prior and marginalized likelihood.* Here we marginalize out $\boldsymbol h$ in $f_{\boldsymbol \theta}(\boldsymbol x)$, and use the truncated Normal prior. 
+3.
+*A truncated Normal prior and marginalized likelihood.* Here we marginalize out $\boldsymbol h$ in $f_{\boldsymbol \theta}(\boldsymbol x)$, and use the truncated Normal priors applied to the marginal probabilities for visible variables given by
     \begin{align*}
-    g_{\boldsymbol \theta}(\boldsymbol v) = \sum\limits_{\boldsymbol h \in \{-1,1\}^H} \exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right).
-\vspace{-.75cm}
+    g_{\boldsymbol \theta}(\boldsymbol v) \propto \sum\limits_{\boldsymbol h \in \{-1,1\}^H} \exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right), \boldsymbol v \in \{-1,1\}^V.
     \end{align*}
@@
 Additionally, the trick prior requires additional tuning to choose $C_1$ and $C_2$, reducing its appeal.
-The values used for the hyperparameters in our simulation are presented in table \ref{tab:hyperparam}.
+The values used for the hyperparameters in our simulation are presented in Table \ref{tab:hyperparam}.
 
@@
 Secondly, we can assess the mixing of our chain using an idea of effective sample size.
-If the MCMC chain were truly iid draws from the target distribution, then the estimate of the posterior mean probability of image $i$, $\bar{p}^{(i)}$ would be distributed approximately Normal with mean the true posterior probability of image $i$, $p^{(i)}$, and variance $\sigma^2_i/M$, where $\sigma^2_i$ is the true posterior variance of image $i$ and $M$ is the length of the chain.
-However, with the presence of correlation in our chain, the aysmptotic variance of $\bar{p}^{(i)}$ is instead some $C_i/M$, where $C_i$ is some positive constant such that $C_i > \sigma^2_i$.
-We can use an overlapping blockmeans approach to get a crude estimate for $C_i$, $\hat{C}_i = b \text{Var} \overline{p_i}$ where $b$ is the block length and $\overline{p_i}$ are the blockmeans for image $i$.
-We compare it to an estimate of $\sigma^2_i$ using the raw chain, $\hat{\sigma}^2_i$.
-Formally,
+If the MCMC chain were truly iid draws from the target distribution, then the estimate of the posterior mean probability of image $i$, $\bar{p}^{(i)}$ would be distributed approximately Normal with mean the true posterior marginal probability of image $i$, $p^{(i)}$, and variance $\sigma^2_i/M$, where $\sigma^2_i$ is the true posterior variance of image $i$ and $M$ is the length of the chain.
+However, with the presence of correlation in our chain, the aysmptotic variance of $\bar{p}^{(i)}$ is instead approximately some $C_i/M$, where $C_i$ is some positive constant such that $C_i > \sigma^2_i$.
+We can use an overlapping block-means approach [@gelman2011inference] to get a crude estimate for $C_i$, $\hat{C}_i = b S^2(\overline{p_i})$ where $b$ is the block length and $S^2(\overline{p_i})$ denotes the sample variance of the blockmeans $\overline{p}_i$ for image $i$.
+We compare it to an estimate of $\sigma^2_i$ using sample variance of the raw chain, $\hat{\sigma}^2_i$.
+Formally, we approximate the effective sample size of the length $M$ MCMC chain as
 
@@
 Thus, the truncated likelihood would require at least $`r round(min(M_eff_table[1, -1])/min(M_eff_table[2, -1]), 1)`$ times more iterations to be run of the MCMC chain in order to achieve the same amount of effective information about the posterior distribution.
-For this reason, in conjuction with the ACF results, the truncated Normal method is not an effective method for fitting the RBM.
+For this reason, in conjuction with the ACF results in Figure \ref{fig:acf}, the truncated Normal method is not an effective method for fitting the RBM.
 
@@
 From these posterior predictive checks, it is evident that the marginalized likelihood method produces the closest fit to the actual data.
-However, this method requires a marginalization step, which is infeasible for a model with $H$ of any real size.
+However, this method requires a marginalization step to obtain the probability function of visible observations alone, which is infeasible for a model with $H$ of any real size.
 

### 2016-10-22 writing/note.Rmd
2nd edits from Nordman

@@
 date: "October, 2016"
-linestretch: 1.5
+linestretch: 2
 output: 
@@
 
-Essentially, in the process of developing a model to handle increasing numbers of variable (i.e., $N \rightarrow \infty$), we want (\ref{eqn:elpr}) to stay bounded, meaning that the largest probability under this model should maintain a magnitude within the smallest probability allowed under the model.
+Essentially, in formulating a model for potentially increasing numbers of variables (i.e., $N \rightarrow \infty$), the ratio (\ref{eqn:elpr}) should bounded, meaning that the largest probability under this model at parameter $\boldsymbol \theta_N$ should maintain a magnitude within the smallest probability allowed under the same model.
 This leads to the following definition:
 
-\begin{defn}[unstable DEF]
-A DEF model formulation is \emph{unstable} if, as the number of variables increase ($N \rightarrow \infty$), it holds that 
+\begin{defn}[S-unstable DEF]
+Let $\boldsymbol \theta_N \in \mathbb{R}^{q(N)}$ be a sequence of DEF model parameters where the size of the model $q(N)$ is a function of the number of random variables $N$.
+A DEF model formulation is \emph{Schweinberger-unstable} or \emph{S-unstable} if, as the number of variables increase ($N \rightarrow \infty$), it holds that
 \begin{align*}
-\lim\limits_{N \rightarrow \infty} \frac{1}{N} \log \left[\frac{\max\limits_{(x_1, \dots, x_N) \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(x_1, \dots, x_N)}{\min\limits_{(x_1, \dots, x_N) \in \mathcal{X}^N}P_{\boldsymbol \theta}(x_1, \dots, x_N)}\right] = \infty.
+\lim\limits_{N \rightarrow \infty} \frac{1}{N} \text{ELPR}(\boldsymbol \theta_N) \equiv \lim\limits_{N \rightarrow \infty} \frac{1}{N} \log \left[\frac{\max\limits_{(x_1, \dots, x_N) \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(x_1, \dots, x_N)}{\min\limits_{(x_1, \dots, x_N) \in \mathcal{X}^N}P_{\boldsymbol \theta}(x_1, \dots, x_N)}\right] = \infty.
 \end{align*}
@@
 
-Note that this definition of *unstable* is a generalization of that in @schweinberger2011instability allowing nonexponential family models and an increasing number of parameters (while this definition appears different, it does match that in @schweinberger2011instability for the special exponential models considered there, see Section \ref{discrete-exponential-family-models} for details).
+Note that this definition of *S-unstable* is a generalization or reinterpretation of "unstable" used in @schweinberger2011instability by allowing nonexponential family models and an increasing number of parameters.
+While this definition appears different, it does match that in @schweinberger2011instability for the special exponential models considered there, see Section \ref{discrete-exponential-family-models} for details.
 
-Unstable DEF models are undesirable for several reasons. One is that small changes in data can lead to overly-sensitive changes in probability. Consider, for example, 
+S-unstable DEF models are undesirable for several reasons. One is that small changes in data can lead to overly-sensitive changes in probability. Consider, for example, 
 $$
-\Delta_N(\boldsymbol \theta_N) \equiv \max \left\{\log \frac{P_{\boldsymbol \theta_N}(x_1, \dots, x_N)}{P_{\boldsymbol \theta_N}(x_1^*, \dots, x_N^*)} : (x_1, \dots, x_N) \& (x_1^*, \dots, x_N^*) \in \mathcal{X}^N \text{ differ by exactly one component}\right\},
+\Delta(\boldsymbol \theta_N) \equiv \max \left\{\log \frac{P_{\boldsymbol \theta_N}(x_1, \dots, x_N)}{P_{\boldsymbol \theta_N}(x_1^*, \dots, x_N^*)} : (x_1, \dots, x_N) \& (x_1^*, \dots, x_N^*) \in \mathcal{X}^N \text{ differ by exactly one component}\right\},
 $$
-the biggest log-probability ratio for a one component change in data outcomes. We then have the following result.
+the biggest log-probability ratio for a one component change in data outcomes at a DEF parameter $\boldsymbol \theta_N$. We then have the following result.
 
@@
 \label{prop:instab}
+Let $C > 0$ and let $\text{ELPR}(\boldsymbol \theta_N)$ be as in (\ref{eqn:elpr}) for an integer $N \ge 1$.
 If $\frac{1}{N}\text{ELPR}_N(\boldsymbol \theta_N) > C$, then $\Delta_N(\boldsymbol \theta_N) > C$.
@@
 
-In other words, if the quantity (\ref{eqn:elpr}) is too large, then the DEF model will exhibit large probability shifts for very small changes in the data configuration (i.e.
-instability mentioned earlier).
+In other words, if the probability ratio (\ref{eqn:elpr}) is too large, then the DEF model will exhibit large probability shifts for very small changes in the data configuration (i.e., instability).
 
 **Proof of Proposition \ref{prop:instab}.** We prove the contrapositive.
-Suppose that $\Delta_N(\boldsymbol \theta_N) \le C$ holds for some $C > 0$.
-Under the DEF model, $P_{\boldsymbol \theta_V}(\boldsymbol x) > 0$ holds for each outcome $\boldsymbol x \in \mathcal{X}^N$.
+Suppose that $\Delta(\boldsymbol \theta_N) \le C$ holds for some $C > 0$.
+Under the DEF model, $P_{\boldsymbol \theta_N}(\boldsymbol x) > 0$ holds for each outcome $\boldsymbol x \in \mathcal{X}^N$.
 Let $\boldsymbol x_{min} \equiv \argmin\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)$ and $\boldsymbol x_{max} \equiv \argmax\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)$.
@@
 $\boldsymbol x_i, \boldsymbol x_{i + 1} \in \mathcal{X}^N$ differ by exactly $1$ component for $i = 0, \dots, k$) for some integer $k \in \{0, 1, \dots, N\}$.
-Then
+Then, if $k > 0$, it follows that
 \begin{align*}
@@
 \end{align*}
-using $k \le N$ and $\Delta_N(\boldsymbol \theta_N) \le C$.$\Box$
+using $k \le N$ and $\Delta(\boldsymbol \theta_N) \le C$. If $k = 0$, then $\boldsymbol x_{max} = \boldsymbol x_{min}$ and the same bound above holds.$\Box$
 
-Additionally, unstable RBM models are connected to degenerate models, where *degeneracy* implies placing all probability on a small pieace of the sample space. Define a model set 
+Additionally, S-unstable DEF models are connected to degenerate models, where *degeneracy* implies placing all probability on a small portion of the sample space. Define a modal set 
 $$
@@
 $$
-of possible outcomes, for a given $0 < \epsilon < 1$.
+of possible outcomes, for a given $0 < \epsilon < 1$ and along a sequence $\boldsymbol \theta_N$, $N \ge 1$, of DEF parameters.
 
@@
 
-In other words, as a consequence of unstable models, all probability will stack up on mode sets or potentially those few outcomes with the highest probability.
+In other words, as a consequence of S-unstable models, all probability in the DEF model formulation with a large number of random variables will stack up on mode sets or simply those few outcomes with the highest probability.
 
@@
 \begin{align*}
-1-P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N}) &= P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N}^C) \\
+1-P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N})
 & = \sum\limits_{\boldsymbol x \in \mathcal{X}^N \setminus M_{\epsilon, \boldsymbol \theta_N}}P_{\boldsymbol \theta_N}(\boldsymbol x) \\
@@
 \end{align*}
-Then,
+Then, it follows that
 \begin{align*}
@@
 \end{align*}
-as $N \rightarrow \infty$ by the definition of an unstable DEF model.$\Box$
+as $N \rightarrow \infty$ by the definition of an unstable DEF model.
+Consequently, $P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N}) \rightarrow 1$ as $N \rightarrow \infty$ as claimed.$\Box$
+
+\begin{remark}
+There is a further generalization to the notion of "unstable" used in Schweinberger (2011) that deals with replication.
+One can conceivably imagine replications of the data $\boldsymbol X_1, \dots, \boldsymbol X_n$, where $X_i \in \mathbb{R}^N, i = 1, \dots, n$ such that there are now $n*N$ random variables in the model.
+However, the definition of S-unstable and propositions \ref{prop:instab}-\ref{prop:degen} still hold if the replications are independent and identically distributed.
+The reason for this is that for $n$ observations, $\left(\max\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)\right)^n$ is the largest probability possible under the model and similarly $\left(\min\limits_{\boldsymbol x \in  \mathcal{X}}P_{\boldsymbol \theta_N}(\boldsymbol x)\right)^n$ is the smallest.
+Thus,
+\begin{align*}
+\frac{\text{extremal log-probability ratio}}{\text{\# random variables in the model}} &\equiv \frac{1}{n*N}\log \left[\frac{\left\{\max\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)\right\}^n}{\left\{\min\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta}(\boldsymbol x)\right\}^n}\right] \\
+&= \frac{1}{N} \log \left[\frac{\max\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)}{\min\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta}(\boldsymbol x)}\right]\\
+&= \frac{1}{N} \text{ELPR}(\boldsymbol \theta_N).
+\end{align*}
+Meaning, the definition of an S-unstable DEF model is invariant to replication.
+\end{remark}
 

### 2016-10-22 writing/draft.Rmd
discussion section

@@
 
+RBM models are an interesting class of undirected graphical models that have shown to be useful for supervised learning tasks.
+However, when viewed as a statistical model, RBMs very easily fall prey to instances of model impropriety such as near-degeneracy, S-instability, and uninterpretability.
+Additionally, these models are difficult to fit using a rigorous methodology due to size of the model space.
+
+In this paper, we have presented three honest Bayes MCMC-based methods for fitting RBMs.
+Common practice is to use MCMC to overcome fitting complexities, however because of the size of the space to be spanned ($H$ hidden variables for each visible observation) convergence and mixing of these methods will be slow.
+Marginalization of the latent variables in the model can improve mixing, but is numerically intractible due to repeated calculation of the normalizing constant.
+
+Ultimately, it is not clear RBM models are useful as generative models.
+Due to the extreme flexibility in these models, a rigorous likelihood-based method for fitting an RBM will merely return the (discrete) empirical distribution for visibles.
+Meaning, even if a rigorous likelihood-based fitting method is able to overcome the inherent difficulties of mixing and convergence in a large space, we know the answer before we even begin.
+For these reasons, we are thus skeptical of the generative claims made about RBMs.
+
 \begin{appendix}

### 2016-10-24 writing/draft.Rmd
notation on encoding cleared up

@@
 abstract: |
-  A restricted Boltzmann machine (RBM) is an undirected graphical model constructed for discrete or continuous random variables, and two layers, one hidden and one visible, with no conditional dependency within a layer.
+  A restricted Boltzmann machine (RBM) is an undirected graphical model constructed for discrete or continuous random variables, with two layers, one hidden and one visible, and no conditional dependency within a layer.
 In recent years, RBMs have risen to prominence due to their connection to deep learning.
@@
 RBMs are thereby thought to have the ability to encode very complex and rich structures in data, making them attractive for supervised learning.
-However, the generative behavior of RBMs has largely been unexplored.
+However, the generative behavior of RBMs has been largely unexplored.
 In this paper, we discuss the relationship between RBM parameter specification in the binary case and the prevalence of undesirable model behaviors such as degeneracy, instability and/or uninterpretability.
-We also describe the difficulties that arise in likelihood-based fitting of such highly flexible models, especially as Gibbs sampling (quasi-Bayes) methods are often advocated for the RBM model structure.
+We also describe the difficulties that arise in likelihood-based fitting and Bayes of such (highly flexible) models, especially as Gibbs sampling (quasi-Bayes) methods are often advocated for the RBM model structure.
 ---
@@
 
-The data mining and machine learning communities have recently found a great interest in deep learning, specifically in stacked restricted Boltzmann machines (RBMs) [see @salakhutdinov2009deep, @salakhutdinov2012efficient, @srivastava2013modeling, @le2008representational for examples].
-A RBM is an undirected probabilistic graphical model (for discrete or continuous random variables) with two layers, one hidden and one visible, and with no conditional dependency within a layer [@smolensky1986information].
+The data mining and machine learning communities have recently shown a great interest in deep learning, specifically in stacked restricted Boltzmann machines (RBMs) [see @salakhutdinov2009deep, @salakhutdinov2012efficient, @srivastava2013modeling, @le2008representational for examples].
+A RBM is an undirected probabilistic graphical model (for discrete or continuous random variables) with two layers, one hidden and one visible, with no conditional dependency within a layer [@smolensky1986information].
 These models have reportedly been used with success in classification of images [@larochelle2008classification, @srivastava2012multimodal].
 However, the model properties are largely unexplored in the literature and the commonly cited fitting methodology remains heuristic-based and abstruse [@hinton2006fast].
-In this paper, we attempt to provide steps toward a thorough understanding of the model and its behavior from the perspective of statistical theory and then explore the possibility of a rigorous fitting methodology.
+In this paper, we attempt to provide steps toward a thorough understanding of the model and its behavior from the perspective of statistical theory, and then explore the possibility of a rigorous fitting methodology.
 We find the RBM model class to be deficient in two fundamental ways.
@@
 Neyman and Box agree with this assessment and go further to state that a model should "provide an explanation of the mechanism underlying the observed phenomena" [@lehmann1990model, @box1967discrimination].
-At issue, RBMs can often produce data lacking realistic variability and may thereby fail to provide a satisfactory conceptualization of the data generation process.
-In addition to such degeneracy, we find that RBMs can easily exhibit types of instability in the parameter space.
-In practice, this may seen when a single pixel change in an image results in a wildly different classification.
+At issue, RBMs  often produce data lacking realistic variability and may thereby fail to provide a satisfactory conceptualization of the data generation process.
+In addition to such degeneracy, we find that RBMs can easily exhibit types of instability.
+In practice, this may seen when a single pixel change in an image results in a wildly different classification in an image classification problem.
 Such model impropriety issues have recently been documented in RBMs [@li2014biclustering], as well as other deep architectures [@szegedy2013intriguing, @nguyen2014deep].
@@
 Secondly, the fitting of these models is problematic.
-As the size of these models grows, both maximum likelihood and Bayesian methods of fitting quickly being intractable.
-But even when fitting is feasible, there are unique challenges due to the overparameterization of RBMs.
-The literature often suggests Markov chain Monet Carlo (MCMC) approximations to fit these models (e.g., Gibbs sampling to exploit conditional structure in the hidden and visible variables), but little is said about monitoring convergence [@hinton2010practical, @hinton2006fast].
+As the size of these models grows, both maximum likelihood and Bayesian methods of fitting quickly become intractable.
+The literature often suggests Markov chain Monte Carlo (MCMC) approximations to posteriors to fit these models (e.g., Gibbs sampling to exploit conditional structure in the hidden and visible variables), but little is said about acheiving convergence [@hinton2010practical, @hinton2006fast].
 Related to this, these MCMC algorithms require updating potentially many latent variables (hiddens) per observation (visibles) which can critically influence convergence in MCMC-based likelihood methods.
+
 In Section \ref{bayesian-model-fitting}, we compare three Bayesian modeling techniques (involving MCMC approximations that are computationally more accessible than direct maximum likelihood), each proposed with the goal of avoiding parts of the RBM parameter space that yield improper models.
-As might be expected, with greater computational complexity comes an increase in fitting accuracy, but at the cost of feasibility.
+As might be expected, with greater computational complexity comes an increase in fitting accuracy, but at the cost of practical feasibility.
+
 Additionally, because RBM models are highly overparametrized, any fully principled fitting method based on the likelihood will simply reproduce the (discrete) empirical distribution of most realistic training sets.
@@
 Section \ref{background} formally definesthe RBM including the joint distribution of hidden and visible variables and explains the model's connection to deep learning.
-Additionally, measures of model impropriety and methods of detecting them are defined.
+Additionally, measures of model impropriety and methods of detecting it are defined.
 Section \ref{explorations} details our explorations into the model behavior and propriety (or lack thereof) of the RBM class.
-We discuss three Bayesian fitting techniques with the goal of avoiding model impropriety in Section \ref{model-fitting} and conclude with a discussion in Section \ref{discussion}.
+We discuss three Bayesian fitting techniques intended to avoid fitted model impropriety in Section \ref{model-fitting} and conclude with a discussion in Section \ref{discussion}.
 Additionally, there is an accompanying appendix with proofs of main results.
@@
 Let $\boldsymbol x = (h_1, \dots, h_H, v_1,\dots,v_V)$ represent the states of the visible and hidden nodes in an RBM for some integers $V, H \ge 1$.
-We will consider both $\boldsymbol x \in \{0,1\}^{H + V}$ and $\boldsymbol x \in \{-1, 1\}^{H + V}$, as the potential encodings for a binary $H + V$-dimensional space, $\mathcal{X}$.
+Each single "binary" random variable, visible or hidden, will take its values in a common coding set $\mathcal{C}$, where we allow one of two possibilities for the coding set $\mathcal{C}=\{0,1\}$ or $\mathcal{C}=\{-1,1\}$ with "$1$" always indicating the "high" value of the variable.
+While $\mathcal{C}=\{0,1\}$ may be a natural or na{\"i}ve starting point, we show later that the coding $\mathcal{C}=\{-1,1\}$ induces better model properties for the RBM.
 We posit a parametric form for probabilities corresponding to a potential vector of states, $X = (H_1, \dots, H_H, V_1,\dots,V_V)$, for the nodes as
@@
 
-where $\boldsymbol \theta \equiv (\theta_{11}, \dots, \theta_{1H}, \dots, \theta_{V1}, \dots, \theta_{VH}, \theta_{v_1}, \dots, \theta_{v_V}, \theta_{h_1}, \dots, \theta_{h_H}) \in \mathbb{R}^{V + H + V*H}$ denotes the vector of model parameters and the denominator $\gamma(\boldsymbol \theta) = \sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)$ is the normalizing function to ensure the probabilities (\ref{eqn:pmf}) sums to one.
+where $\boldsymbol \theta \equiv (\theta_{11}, \dots, \theta_{1H}, \dots, \theta_{V1}, \dots, \theta_{VH}, \theta_{v_1}, \dots, \theta_{v_V}, \theta_{h_1}, \dots, \theta_{h_H}) \in \mathbb{R}^{V + H + V*H}$ denotes the vector of model parameters and the denominator $\gamma(\boldsymbol \theta) = \sum\limits_{\boldsymbol x \in \mathcal{C}^{H+V}}\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)$ is the normalizing function that ensures the probabilities (\ref{eqn:pmf}) sum to one.
 For
 \begin{align}
-\boldsymbol t(\boldsymbol x) = (h_1, \dots, h_H, v_1, \dots, v_V, v_1 h_1, \dots, v_V h_H) \in \{0,1\}^{H + V + H*V}, \boldsymbol x = (h_1, \dots, h_H, v_1, \dots, v_V) \in \mathcal{X} = \{0,1\}^{V + H}, \label{eqn:t}
+\boldsymbol t(\boldsymbol x) = (h_1, \dots, h_H, v_1, \dots, v_V, v_1 h_1, \dots, v_V h_H) \in \mathcal{C}^{H + V + H*V}, \boldsymbol x = (h_1, \dots, h_H, v_1, \dots, v_V) \in \mathcal{C}^{V + H}, \label{eqn:t}
 \end{align}
-let $\boldsymbol \tau = \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{X}\}$ be the vector of random variables needed to compute probabilities (\ref{eqn:pmf}) in the model and write $Q_{\boldsymbol \theta}(\boldsymbol x) = \sum\limits_{i = 1}^H\sum\limits_{j=1}^V \theta_{ij} h_i v_j + \sum\limits_{i = 1}^H\theta_{hi} h_i + \sum\limits_{j = 1}^V\theta_{vj} v_j$ for the neg-potential function.
+let $\mathcal{T} = \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{X}\}$ be the vector of variables needed to compute probabilities (\ref{eqn:pmf}) in the model and write $Q_{\boldsymbol \theta}(\boldsymbol x) = \sum\limits_{i = 1}^H\sum\limits_{j=1}^V \theta_{ij} h_i v_j + \sum\limits_{i = 1}^H\theta_{hi} h_i + \sum\limits_{j = 1}^V\theta_{vj} v_j$ for the neg-potential function.
 Note that the RBM model is parameterized by $\boldsymbol \theta$ containing two types of parameters, main effects and interaction effects.
@@
 Due to the potential size of the model, the normalizing constant $\gamma(\boldsymbol \theta)$ can be practically impossible to calculate, making it impossible to estimate the model parameter vector in an obvious fashion.
-A kind of Gibbs sampling can be tried, which is natural due to conditional independence architecture within the RBM (i.e.
+A kind of Gibbs sampling can be tried, which is natural due to conditional independence architecture of the RBM (i.e.
 visibles given hiddens or vice versa).
@@
 RBMs have risen to prominence in recent years due to their connection to deep learning [see @hinton2006fast, @salakhutdinov2012efficient, or @srivastava2013modeling for examples].
-By stacking multiple layers of RBMs in a deep architecture, proponents of the models claim the ability to learn "internal representations that become increasingly complex, which is considered to be a promising way of solving object and speech recognition problems" [@salakhutdinov2009deep, pp.
+By stacking multiple layers of RBMs in a deep architecture, proponents of the models claim to produce the ability to learn "internal representations that become increasingly complex, which is considered to be a promising way of solving object and speech recognition problems" [@salakhutdinov2009deep, pp.
 450].
@@
 ## Degeneracy, instability, and uninterpretability... Oh my!
-The highly flexible nature of a RBM (having as it does $H + V + H V$ parameters) creates three kinds of potential model impropriety of concern: *degeneracy, instability, and uninterpretability*.
-In this section we define these characteristics, consider how to detect them in a RBM, and point out their relationships to one another.
+The highly flexible nature of a RBM (having as it does $H + V + H V$ parameters) creates at least three kinds of potential model impropriety of concern that we will call *degeneracy, instability, and uninterpretability*.
+In this section we define these characteristics, consider how to detect them in a RBM, and point out relationships among them.
 
@@
 where $\boldsymbol \theta \in \Theta \subset \mathbb{R}^q$ is the model parameter, and $t: \mathcal{X} \rightarrow \mathbb{R}^q$ are statistics based on the adjacency matrix of a graph.
-Again, $\gamma(\boldsymbol \theta) = \sum_{\boldsymbol x \in \mathcal{X}} \exp\left(\boldsymbol \theta^T \boldsymbol t(\boldsymbol x)\right)$ is the normalizing function for a valid distribution.
+Again, $\gamma(\boldsymbol \theta) = \sum_{\boldsymbol x \in \mathcal{X}} \exp\left(\boldsymbol \theta^T \boldsymbol t(\boldsymbol x)\right)$ is the normalizing function.
 Let $C$ denote the convex hull spanned by the potential outcomes of sufficient statistics, $\{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{X}\}$.
-@handcock2003assessing classify an exponentially parametrized random graph model at $\boldsymbol \theta$ as *near-degenerate* if the mean value of sufficient statistics under $\boldsymbol \theta$, $\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta}t(X)$, is close to the boundary of $C$.
+@handcock2003assessing classify an exponentially parametrized random graph model at $\boldsymbol \theta$ as *near-degenerate* if the mean value of sufficient statistics under $\boldsymbol \theta$, $\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta}\boldsymbol t(X)$, is close to the boundary of $C$.
 
 The RBM class of models can be thought to exhibit an analogous occurrence of *near-degeneracy* as follows.
-When random variables $\boldsymbol t(\boldsymbol x) = (v_1, \dots, v_V, h_1, \dots, h_H, v_1 h_1, \dots, v_V h_H ) \in \boldsymbol \tau \equiv \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{X} = \{0,1\}^{V + H}\} \subset \{0,1\}^{V + H + V*H}$ in the neg-potential function $Q_{\boldsymbol \theta}(\cdot)$ have a mean vector $\boldsymbol \mu(\boldsymbol \theta)$ close to the boundary of the convex hull of $\boldsymbol \tau$, then the RBM model can be said to be near-degenerate at $\boldsymbol \theta \in \mathbb{R}^{V + H + H*V}$.
-Here the mean of $\boldsymbol t(\boldsymbol x)$ is given as
+When random variables $\boldsymbol t(\boldsymbol x) = (v_1, \dots, v_V, h_1, \dots, h_H, v_1 h_1, \dots, v_V h_H ) \in \mathcal{T} \equiv \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{C}^{H + V} \}$ in the neg-potential function $Q_{\boldsymbol \theta}(\cdot)$ have a mean vector $\boldsymbol \mu(\boldsymbol \theta)$ close to the boundary of the convex hull of $\mathcal{T}$, then the RBM model can be said to be near-degenerate at $\boldsymbol \theta \in \mathbb{R}^{V + H + H*V}$.
+Here the mean of $\boldsymbol t(\boldsymbol x)$ is
 $$
@@
 Considering a class of exponential families of distributions, @schweinberger2011instability introduced a concept of model defficiency related to *instability*.
-Instability can be roughly characterized by excessive sensitivity in the model, where small changes in the components of potential data outcomes, $\boldsymbol x$, may lead to substantial changes in the probability function $f_{\boldsymbol \theta}(\boldsymbol x)$.
-To quantify "instability" more rigorously (particularly beyond the definition given by @schweinberger2011instability), it is useful to imagine how the RBM model might be expanded to incorporate more and more observations.
-When increasing the size of a RBM model to handle more visible variables, it becomes necessary to expand the number of model parameters (and in this process one may also arbitrarily expand the number of hidden variables to be included).
-To this end, let $\boldsymbol \theta_{V} \equiv (\theta_{v_1}, \dots, \theta_{v_V}, \theta_{h_1}, \dots, \theta_{h_H}, \theta_{11}, \dots, \theta_{VH}), V \ge 1$, denote a sequence of RBM parameters indexed by the number $V$ of visibles $(V_1, \dots, V_V)$ and define a (scaled) extremal log-probability ratio of the RBM model at $\boldsymbol \theta_{V}$ as
+Instability can be roughly characterized as excessive sensitivity in the model, where small changes in the components of potential data outcomes, $\boldsymbol x$, may lead to substantial changes in the probability function $f_{\boldsymbol \theta}(\boldsymbol x)$.
+To quantify "instability" more rigorously (particularly beyond the definition given by @schweinberger2011instability) it is useful to consider how the RBM model might be expanded to incorporate more and more visibles When increasing the size of a RBM model, it becomes necessary to grow the number of model parameters (and in this process one may also arbitrarily expand the number of hidden variables used).
+To this end, let $\boldsymbol \theta_{V} \equiv (\theta_{v_1}, \dots, \theta_{v_V}, \theta_{h_1}, \dots, \theta_{h_H}, \theta_{11}, \dots, \theta_{VH}), V \ge 1$, denote an element of a sequence of RBM parameters indexed by the number $V$ of visibles $(V_1, \dots, V_V)$ and define a (scaled) extremal log-probability ratio of the RBM model at $\boldsymbol \theta_{V}$ as
 \begin{align}
-\frac{1}{V} \log \left[\frac{\max\limits_{(v_1, \dots, v_V) \in \{-1,1\}^V}P_{\boldsymbol \theta_V}(v_1, \dots, v_V)}{\min\limits_{(v_1, \dots, v_V) \in \{-1,1\}^V}P_{\boldsymbol \theta_V}(v_1, \dots, v_V)}\right] \equiv \frac{1}{V} \text{ELPR}(\boldsymbol \theta_V) \label{eqn:elpr}
+\frac{1}{V} \log \left[\frac{\max\limits_{(v_1, \dots, v_V) \in \mathcal{C}^V}P_{\boldsymbol \theta_V}(v_1, \dots, v_V)}{\min\limits_{(v_1, \dots, v_V) \in \mathcal{C}^V}P_{\boldsymbol \theta_V}(v_1, \dots, v_V)}\right] \equiv \frac{1}{V} \text{ELPR}(\boldsymbol \theta_V) \label{eqn:elpr}
 \end{align}
-where $P_{\boldsymbol \theta_V}(v_1, \dots, v_V) \propto \sum_{\{h_j\}\in \{-1,1\}}\exp\left(\sum_{i = 1}^V \theta_{v_i} v_i + \sum_{j = 1}^H \theta_{h_j} h_j + \sum_{i = 1}^V \sum_{j = 1}^H \theta_{ij} v_i h_j\right)$ is the RBM probability of observing outcome $(v_1, \dots, v_V)$ for the visible variables $(V_1, \dots, V_V)$ under parameters $\boldsymbol \theta_V$.
-
-In formulating a RBM model for a potentially large number of visibles (i.e., $V \rightarrow \infty$), we will see that the ratio (\ref{eqn:elpr}) needs to stay bounded for the RBM model to be stable.
-This leads to the following definition:
+where $P_{\boldsymbol \theta_V}(v_1, \dots, v_V) \propto \sum_{\{h_j\}\in \mathcal{C}}\exp\left(\sum_{i = 1}^V \theta_{v_i} v_i + \sum_{j = 1}^H \theta_{h_j} h_j + \sum_{i = 1}^V \sum_{j = 1}^H \theta_{ij} v_i h_j\right)$ is the RBM probability of observing outcome $(v_1, \dots, v_V)$ for the visible variables $(V_1, \dots, V_V)$ under parameters $\boldsymbol \theta_V$.
+
+In formulating a RBM model for a potentially large number of visibles (i.e., $V \rightarrow \infty$), we will say that the ratio (\ref{eqn:elpr}) needs to stay bounded for the sequence of RBM models to be stable.
+That is, we make the following convention
 
 \begin{defn}[S-unstable RBM]
-Let $\boldsymbol \theta_V \in \mathbb{R}^{V + H +  H*V}, v \ge 1$, be a sequence of RBM parameters where the number of hiddens, $H \equiv H(V) \ge 1$, can be an arbitrary function of the number $V$ of visibles.
-A RBM model formulation is \emph{Scheinberger-unstable} or \emph{S-unstable} if, as the number of visible variables increase ($V \rightarrow \infty$), it holds that
+Let $\boldsymbol \theta_V \in \mathbb{R}^{V + H +  H*V}, v \ge 1$, be an element of a sequence of RBM parameters where the number of hiddens, $H \equiv H(V) \ge 1$, can be an arbitrary function of the number $V$ of visibles.
+A RBM model formulation is \emph{Scheinberger-unstable} or \emph{S-unstable} if
 \begin{align*}
-\lim\limits_{V \rightarrow \infty} \frac{1}{V} \text{ELPR}(\boldsymbol \theta_V) \equiv \lim\limits_{V \rightarrow \infty} \frac{1}{V} \log \left[\frac{\max\limits_{(v_1, \dots, v_V) \in \{-1,1\}^V}P_{\boldsymbol \theta}(v_1, \dots, v_V)}{\min\limits_{(v_1, \dots, v_V) \in \{-1,1\}^V}P_{\boldsymbol \theta}(v_1, \dots, v_V)}\right] = \infty.
+\lim\limits_{V \rightarrow \infty} \frac{1}{V} \text{ELPR}(\boldsymbol \theta_V) \equiv \lim\limits_{V \rightarrow \infty} \frac{1}{V} \log \left[\frac{\max\limits_{(v_1, \dots, v_V) \in \mathcal{C}^V}P_{\boldsymbol \theta}(v_1, \dots, v_V)}{\min\limits_{(v_1, \dots, v_V) \in \mathcal{C}^V}P_{\boldsymbol \theta}(v_1, \dots, v_V)}\right] = \infty.
 \end{align*}
@@
 
-Note that this definition of *S-unstable* is a generalization or re-interpretation of "unstable" used in @schweinberger2011instability by allowing nonexponential family models for visibles $(v_1, \dots, v_V)$ and an increasing number of parameters.
-
-S-unstable RBM models are undesirable for several reasons. One is that, as mentioned above, small changes in data can lead to overly-sensitive changes in probability. Consider, for example, 
-$$
-\Delta(\boldsymbol \theta_V) \equiv \max \left\{\log \frac{P_{\boldsymbol \theta_V}(v_1, \dots, v_V)}{P_{\boldsymbol \theta_V}(v_1^*, \dots, v_V^*)} : (v_1, \dots, v_V) \& (v_1^*, \dots, v_V^*) \in \{-1,1\}^V \text{ differ by exactly one component}\right\},
+This definition of *S-unstable* is a generalization or re-interpretation of the "unstable" concept of @schweinberger2011instability in that here RBM models for visibles $(v_1, \dots, v_V)$ do not form an exponential family and the dimensionality of $\boldsymbol \theta_V$ is not fixed, but rather grows with $V$.
+
+S-unstable RBM model sequences are undesirable for several reasons. One is that, as mentioned above, small changes in data can lead to overly-sensitive changes in probability. Consider, for example, 
+$$
+\Delta(\boldsymbol \theta_V) \equiv \max \left\{\log \frac{P_{\boldsymbol \theta_V}(v_1, \dots, v_V)}{P_{\boldsymbol \theta_V}(v_1^*, \dots, v_V^*)} : (v_1, \dots, v_V) \& (v_1^*, \dots, v_V^*) \in \mathcal{C}^V \text{ differ by exactly one component}\right\},
 $$
@@
 
-In other words, if the probability ratio (\ref{eqn:elpr}) is too large, then the RBM model will exhibit large probability shifts for very small changes in the data configuration (i.e., instability).
-
-Additionally, S-unstable RBM models are connected to degenerate models  placing all probability on a small portion of the sample space.
+In other words, if the probability ratio (\ref{eqn:elpr}) is too large, then a RBM model sequence will exhibit large probability shifts for very small changes in data configurations (i.e., instability).
+
+Additionally, S-unstable RBM model sequences are connected to degenerate model sequences placing all probability on a small portion of their sample spaces.
 To see this, define a modal set of possible outcomes (i.e.
@@
 $$
-M_{\epsilon, \boldsymbol \theta_V} \equiv \left\{(v_1, \dots, v_V) \in \{-1,1\}^V: \log P_{\boldsymbol \theta_V}(v_1, \dots, v_V) > (1-\epsilon)\max\limits_{(v_1^*, \dots, v_V^*)}P_{\boldsymbol \theta_V}(v_1^*, \dots, v_V^*) + \epsilon\min\limits_{(v_1^*, \dots, v_V^*)}P_{\boldsymbol \theta_V}(v_1^*, \dots, v_V^*) \right\}
-$$
-for a given $0 < \epsilon < 1$. Then S-unstable models are guaranteed to be degenerate, as the following result shows.
+M_{\epsilon, \boldsymbol \theta_V} \equiv \left\{(v_1, \dots, v_V) \in \mathcal{C}^V: \log P_{\boldsymbol \theta_V}(v_1, \dots, v_V) > (1-\epsilon)\max\limits_{(v_1^*, \dots, v_V^*)}P_{\boldsymbol \theta_V}(v_1^*, \dots, v_V^*) + \epsilon\min\limits_{(v_1^*, \dots, v_V^*)}P_{\boldsymbol \theta_V}(v_1^*, \dots, v_V^*) \right\}
+$$
+for a given $0 < \epsilon < 1$. Then S-unstable model sequences are guaranteed to be degenerate, as the following result shows.
 
@@
 
-In other words, in S-unstable RBM models with a large number of visible variables, all probability in the RBM model formulation  will stack up on those outcomes for visibles with with the highest probability.
+In other words, in S-unstable RBM model sequences with a large number of visible variables, all probability in the RBM model formulation  will stack up on those outcomes for visibles with with the highest probability.
 Proofs of Propositions \ref{prop:instab}-\ref{prop:degen} appear in @kaplan2016note.
-These findings also have counterparts to results in @schweinberger2011instability, but, unlike results there, we do not consider exponential family forms with a fixed number of parameters.
+These findings also have counterparts to results in @schweinberger2011instability, but, unlike results there, we do not limit consideration to exponential family forms with a fixed number of parameters.
 
@@
 \label{remark:bnd}
-To later numerically investigate S-unstable RBM models, it is useful to define a computationally tractable lower bound on (\ref{eqn:elpr}) given by
+To later numerically investigate S-unstable RBM model sequences, it is useful to define a computationally tractable lower bound on (\ref{eqn:elpr}) given by
 \begin{align}
@@
 \end{align}
-That is, it holds that, in (\ref{eqn:elpr}),
+That is, in (\ref{eqn:elpr}),
 \begin{align}
-\frac{1}{V} \text{ELPR}(\boldsymbol \theta_V) \equiv \frac{1}{V} \log \left[\frac{\max\limits_{(v_1, \dots, v_V) \in \{-1,1\}^V}P_{\boldsymbol \theta}(v_1, \dots, v_V)}{\min\limits_{(v_1, \dots, v_V) \in \{-1,1\}^V}P_{\boldsymbol \theta}(v_1, \dots, v_V)}\right] \ge \frac{1}{V}R(\boldsymbol \theta_V).
+\frac{1}{V} \text{ELPR}(\boldsymbol \theta_V) \equiv \frac{1}{V} \log \left[\frac{\max\limits_{(v_1, \dots, v_V) \in \mathcal{C}^V}P_{\boldsymbol \theta}(v_1, \dots, v_V)}{\min\limits_{(v_1, \dots, v_V) \in \mathcal{C}^V}P_{\boldsymbol \theta}(v_1, \dots, v_V)}\right] \ge \frac{1}{V}R(\boldsymbol \theta_V).
 \label{eqn:lowerbnd}
@@
 The benefit of $R(\boldsymbol \theta_V)$ is that $R(\boldsymbol \theta_V)/V$ is simple to compute (i.e.
-requiring no summation over hidden variables, unlike (\ref{eqn:elpr})) and that Propositions \ref{prop:instab}-\ref{prop:degen} continue to hold upon replacing (\ref{eqn:elpr}) with $R(\boldsymbol \theta_V)/V$ due to (\ref{eqn:lowerbnd}).
-Hence, large values of $R(\boldsymbol \theta_V)/V$ also indicate S-unstable RBM models.
+unlike (\ref{eqn:elpr}), requiring no summation over hidden variables) and that Propositions \ref{prop:instab}-\ref{prop:degen} continue to hold upon replacing $\text{ELPR}(\boldsymbol \theta_V)/V$ with $R(\boldsymbol \theta_V)/V$ due to (\ref{eqn:lowerbnd}).
+Hence, large values of $R(\boldsymbol \theta_V)/V$ also indicate S-unstable RBM model sequences.
 
 ### Uninterpretability
-For spatial Markov models, kaiser2007statistical defines a measure of model impropriety he calls *uninterpretability*, which is characterized by dependence parameters in a model being so extreme that marginal mean-structures fail to hold as anticipated by the model statement.
+For spatial Markov models, @kaiser2007statistical defines a measure of model impropriety he calls *uninterpretability*, which is characterized by dependence parameters in a model being so extreme that marginal mean-structures fail to hold as anticipated by the model statement.
 We adapt this notion to RBM models as follows.
 Note that, in a RBM, the parameters $\theta_{v_1}, \dots, \theta_{v_V}$ and $\theta_{h_1}, \dots, \theta_{h_H}$ are naturally associated with main effects of visible and hidden variables and can be interpreted as (logit functions of) means for variables $V_1, \dots, V_V, H_1, \dots, H_V$ in a model with no interaction parameters, $\theta_{ij} = 0, i = 1, \dots, V, j = 1, \dots, H$.
-Hence, these main effect parameters have a clear interpretation under an independence model ($\theta_{ij} = 0$) but this interpretation (as well as an understanding of the nature of dependence) can break down as interaction parameters increase in magnitude (e.g.
-$\max |\theta_{i,j}| \rightarrow \infty$) relative to the size of the main effects.
-In which case, the main effect parameters $\theta_{v_1}$ and $\theta_{h_j}$ are no longer interpretable as originally intended in the modes (statements of marginal means) and the dependence parameters are so large as to dominate the entire model probability structure (losing the model interpretation of dependence as local, conditional modifications of all overall marginal mean structure).
-To access which parameter values $\boldsymbol \theta$ may cause difficulties in interpretation, we use the difference $\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] - \text{E}\left[\boldsymbol X | \boldsymbol \theta^* \right]$ between two model expectations: E$\left[\boldsymbol X | \boldsymbol \theta\right]$ at $\boldsymbol \theta$ and expectations E$\left[\boldsymbol X | \boldsymbol \theta^* \right]$ given independence where $\boldsymbol \theta^*$ is matches $\boldsymbol \theta$ for all main effects but otherwise has $\theta_{ij} = 0$ for $i = 1, \dots, V, j = 1, \dots, H$.
-Hence, $\boldsymbol \theta^*$ and $\boldsymbol \theta$ have the same main effects but $\boldsymbol \theta^*$ has no dependence parameters.
-Uninterpretability is then avoided at a parametric specification $\boldsymbol \theta$ if the model expected value at $\boldsymbol \theta$ is not very different from the corresponding model expectation with assumed independence.
+Hence, these main effect parameters have a clear interpretation under an independence model ($\theta_{ij} = 0$) but this interpretation (as well as an understanding of the nature of dependence) can break down as interaction parameters increase in magnitude relative to the size of the main effects.
+In such case, the main effect parameters $\theta_{v_1}$ and $\theta_{h_j}$ are no longer interpretable as originally intended in the modes (statements of marginal means) and the dependence parameters are so large as to dominate the entire model probability structure (destroying the model interpretation of dependence as local, conditional modifications of all overall marginal mean structure).
+To assess which parameter values $\boldsymbol \theta$ may cause difficulties in interpretation, we use the difference $\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] - \text{E}\left[\boldsymbol X | \boldsymbol \theta^* \right]$ between two model expectations: E$\left[\boldsymbol X | \boldsymbol \theta\right]$ at $\boldsymbol \theta$ and expectations E$\left[\boldsymbol X | \boldsymbol \theta^* \right]$ given independence where $\boldsymbol \theta^*$ matches $\boldsymbol \theta$ for all main effects but otherwise has $\theta_{ij} = 0$ for $i = 1, \dots, V, j = 1, \dots, H$.
+Hence, $\boldsymbol \theta^*$ and $\boldsymbol \theta$ have the same main effects but $\boldsymbol \theta^*$ has $0$ dependence parameters.
+Uninterpretability is then avoided at a parametric specification $\boldsymbol \theta$ if the model expected value at $\boldsymbol \theta$ is not very different from the corresponding model expectation under independence.
 Using this, it is possible to investigate what parametric conditions lead to uninterpretability in a model versus those that guarantee interpretable models.
-If $||\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] - \text{E}\left[\boldsymbol X | \boldsymbol \theta^*\right]||$ is large then the RBM model with parameter vector $\boldsymbol \theta$ is said to be uninterpetable.
+If $||\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] - \text{E}\left[\boldsymbol X | \boldsymbol \theta^*\right]||$ is large, then the RBM model with parameter vector $\boldsymbol \theta$ is said to be uninterpetable.
 The quantities to compare in the RBM case are
@@
 A schematic of this model can be found in figure \ref{fig:toymodel}.
-Because it seems most common, we shall begin by employing $0/1$ encoding of binary variables (both $h_1$ and $v_1$ taking values in $\{0,1\}$).
-Eventually we'll note in Slinkection \ref{data-coding-to-mitigate-degeneracy} that $-1/1$ coding has substantial advantages.
+Because it seems most common, we shall begin by employing $0/1$ encoding of binary variables (both $h_1$ and $v_1$ taking values in $\mathcal{C} = \{0,1\}$).
+Eventually we'll note in Section \ref{data-coding-to-mitigate-degeneracy} that $-1/1$ coding has substantial advantages.
 
@@
 For this small model, we are able to investigate the symptoms of model impropriety, beginning with near-degeneracy.
-To this end, recall that one characterization requires the convex hull of statistics $\boldsymbol \tau = \{\boldsymbol t(\boldsymbol x): \boldsymbol x = (v_1, h_1) \in \{0,1\}^2\} \equiv \{(v_1, h_1, v_1 h_1): v_1, h_1 \in \{0,1\}\}$ appearing in the RBM probabilities for this model.
-As this set is in three dimensions, we are able to explicitly illustrate the shape of boundary of the convex hull of $\boldsymbol \tau$ and explore the behavior of the mean parameterization of the parameters.
-Figure \ref{fig:toyhull} shows the convex hull of our "statistic space", $\boldsymbol \tau \subset \{0,1\}^3$, for this toy problem from multiple perspectives enclosed by the unit cube, representing the convex hull of $\{0,1\}^3$ space.
-In this small model, note that the convex hull of $\boldsymbol \tau$ does not fill the unrestricted hull of $\{0,1\}^3$ because of the relationship between the elements of $\boldsymbol \tau = \{(v_1, h_1, v_1 h_1 : v_1, h_1 \in \{0, 1\} \}$ (i.1.
+To this end, recall that one characterization requires the convex hull of statistics $\mathcal{T} = \{\boldsymbol t(\boldsymbol x): \boldsymbol x = (v_1, h_1) \in \{0,1\}^2\} \equiv \{(v_1, h_1, v_1 h_1): v_1, h_1 \in \{0,1\}\}$ appearing in the RBM probabilities for this model.
+As this set is in three dimensions, we are able to explicitly illustrate the shape of boundary of the convex hull of $\mathcal{T}$ and explore the behavior of the mean parameterization of the parameters.
+Figure \ref{fig:toyhull} shows the convex hull of our "statistic space", $\mathcal{T} \subset \{0,1\}^3$, for this toy problem from multiple perspectives enclosed by the unit cube, representing the convex hull of $\{0,1\}^3$ space.
+In this small model, note that the convex hull of $\mathcal{T}$ does not fill the unrestricted hull of $\{0,1\}^3$ because of the relationship between the elements of $\mathcal{T} = \{(v_1, h_1, v_1 h_1 : v_1, h_1 \in \{0, 1\} \}$ (i.1.
 $v_1 h_1 = 1$ only if $v_1 = h_1 = 1$).
@@
   \caption{The convex hull of the "statistic space"" in three dimensions for the toy RBM with one visible and one hidden node.
-The convex hull of $\boldsymbol \tau = \{\boldsymbol t(\boldsymbol x) : \boldsymbol x \in \mathcal{X}\}$ does not fill the unit cube because of the relationship between the elements of $\boldsymbol \tau$.}
+The convex hull of $\mathcal{T} = \{\boldsymbol t(\boldsymbol x) : \boldsymbol x \in \mathcal{X}\}$ does not fill the unit cube because of the relationship between the elements of $\mathcal{T}$.}
 \label{fig:toyhull}
@@
 The three parametric functions giving $\boldsymbol \mu(\boldsymbol \theta)$ can be visualized in Figure \ref{fig:degen_toy}.
-In examining these, we see that as values of coordinates of $\boldsymbol \theta$ grow larger in magnitude, at least one mean function for the entries of $\boldsymbol t(\boldsymbol x)$ is approaching a value 0 or 1, indicating a potential for $\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta} \boldsymbol t(\boldsymbol x)$ to be near to the boundary of the convex hull of $\boldsymbol \tau$, as a sign of model near-degeneracy.
+In examining these, we see that as values of coordinates of $\boldsymbol \theta$ grow larger in magnitude, at least one mean function for the entries of $\boldsymbol t(\boldsymbol x)$ is approaching a value 0 or 1, indicating a potential for $\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta} \boldsymbol t(\boldsymbol x)$ to be near to the boundary of the convex hull of $\mathcal{T}$, as a sign of model near-degeneracy.
 Thus, for a very small example we can see the relationship between values of $\boldsymbol \theta$ and model degeneracy.
@@
 This will result in variables $\boldsymbol t(\boldsymbol X)$ from (\ref{eqn:t}) satisfying $\boldsymbol t(\boldsymbol x) \in \{0,1\}^{H + V + HV}$ or $\boldsymbol t(\boldsymbol x) \in \{-1,1\}^{H + V + HV}$ depending on how we encode "on" and "off" states in the nodes.
-To explore the effect of this encoding on the potential for apparent near-degeneracy of an RBM, we can consider the ratio of the volume of the hypercube with corners at elements of $\boldsymbol \tau$ to the volume of the convex hull of $\boldsymbol \tau$ under both possible encodings (i.e., compare convex hull of $\boldsymbol \tau$ to either $[0,1]^3$ or $[-1,1]^3$ under the respective encodings).
+To explore the effect of this encoding on the potential for apparent near-degeneracy of an RBM, we can consider the ratio of the volume of the hypercube with corners at elements of $\mathcal{T}$ to the volume of the convex hull of $\mathcal{T}$ under both possible encodings (i.e., compare convex hull of $\mathcal{T}$ to either $[0,1]^3$ or $[-1,1]^3$ under the respective encodings).
 
 For the small two node example, the $0/1$ encoding loses `r round((1-m1_bin$c_hull$vol)*100,2)`% of the volume of $[0,1]^3$ and the $-1/1$ encoding loses `r round((1-m1_neg$c_hull$vol/(2^3))*100,2)`% of the volume of $[-1,1]^3$.
-While there is clearly a one-to-one mapping of models for these two possible codings, this notion of lost volume can be helpful to conceptualize how difficult it will be for the mean vector $\boldsymbol \mu(\boldsymbol \theta)$ to avoid the boundary of $\boldsymbol \tau$, and thus avoid near-degeneracy.
-In fact, if we look at the ratio of volume within the convex hull defined by $\boldsymbol \tau$ and the corresponding hypercube, we can see a relationship emerge as the number of nodes (and thus parameters) increases.
+While there is clearly a one-to-one mapping of models for these two possible codings, this notion of lost volume can be helpful to conceptualize how difficult it will be for the mean vector $\boldsymbol \mu(\boldsymbol \theta)$ to avoid the boundary of $\mathcal{T}$, and thus avoid near-degeneracy.
+In fact, if we look at the ratio of volume within the convex hull defined by $\mathcal{T}$ and the corresponding hypercube, we can see a relationship emerge as the number of nodes (and thus parameters) increases.
 In Figure \ref{fig:volume_plot}, it is evident that as the number of parameters increases, this ratio is decreasing at an increasing rate, meaning it gets more and more difficult to avoid the boundary of the convex hull and thus near degeneracy.
@@
 For each set of parameters we then calculate the three metrics of model impropriety introduced in Section \ref{degeneracy-instability-and-uninterpretability-oh-my}, $\boldsymbol \mu(\boldsymbol \theta)$, $R(\boldsymbol x)/V$, and the coordinates of $\left|\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] -   \text{E}\left[\boldsymbol X | \boldsymbol \theta^* \right] \right|$.
-In the case of near-degeneracy, we can go further and classify each model as near-degenerate or "viable" based on the distance of $\boldsymbol \mu(\boldsymbol \theta)$ from the boundary of the convex hull of $\boldsymbol \tau$ and look at the fraction of models that are "near-degenerate", meaning they are within a small distance to the boundary of the convex hull.
+In the case of near-degeneracy, we can go further and classify each model as near-degenerate or "viable" based on the distance of $\boldsymbol \mu(\boldsymbol \theta)$ from the boundary of the convex hull of $\mathcal{T}$ and look at the fraction of models that are "near-degenerate", meaning they are within a small distance to the boundary of the convex hull.
 

### 2016-10-24 writing/note.Rmd
note intro and abstract

@@
 abstract: |
+  We introduce a broad class of discrete space models that are defined in relation to how a probability model is built to incorporate more variables.
+We will call this class of models *Discrete Expanding Finite space*, or DEF models.
+They are defined as models in discrete finite space with a sequence of expanding model parameter vectors that potentially grow size as the dimension of the data become larger and model support equal the sample space.
+In this paper, we present results that define the notion of model instability as well as near-degeneracy within the framework of DEF models in addition to demonstrating example models of the DEF class.
 ---
 
-# Discrete Expanding Finite space models
+# Introduction
 
+In this paper, we introduce a broad class of discrete space models that are defined in relation to how a probability model is built to incorporate more variables.
 Consider $(X_1, \dots X_N)$, a set of discrete random variables with a finite sample space, $\mathcal{X}^N$.
@@
 
-Note many model families can fall under the umbrella of DEF models.
-In Sections \ref{discrete-exponential-family-models}-\ref{deep-learning}, we present three specific examples of DEF models, including models with deep architectures.
+This paper is structured as follows.
+First, in Section \ref{examples}, we present specific examples of DEF models and discuss how they fit in this framework (including deep learning models).
+Section \ref{instability-and-degeneracy-results} establishes some formal results regarding the model propriety of DEF models.
+Lastly, Section \ref{implications} emphasizes the implications of our model prepriety results.
+
+# Examples
+
+Note many model families can fall under the umbrella of DEF models. In this section, we present three specific examples of DEF models, including models with deep architectures.
 

### 2016-10-24 writing/note.Rmd
clean up

@@
 abstract: |
-  We introduce a broad class of discrete space models that are defined in relation to how a probability model is built to incorporate more variables.
-We will call this class of models *Discrete Expanding Finite space*, or DEF models.
-They are defined as models in discrete finite space with a sequence of expanding model parameter vectors that potentially grow size as the dimension of the data become larger and model support equal the sample space.
+  We introduce *Discrete Expanding Finite space*, or DEF models, a broad class of discrete space models that are defined in relation to how a probability model is built to incorporate more variables.
+These models are defined to be in discrete finite space with model support equal the sample space and with a sequence of expanding model parameter vectors that potentially grow size as the dimension of the data become larger.
 In this paper, we present results that define the notion of model instability as well as near-degeneracy within the framework of DEF models in addition to demonstrating example models of the DEF class.
@@
 First, in Section \ref{examples}, we present specific examples of DEF models and discuss how they fit in this framework (including deep learning models).
-Section \ref{instability-and-degeneracy-results} establishes some formal results regarding the model propriety of DEF models.
-Lastly, Section \ref{implications} emphasizes the implications of our model prepriety results.
+Section \ref{instability-and-degeneracy-results} establishes formal results regarding the model propriety of DEF models.
+Lastly, Section \ref{implications} emphasizes the implications of our model propriety results.
 

### 2016-10-31 writing/draft.Rmd
update some things after oral

@@
 ### Joint distribution
-Let $\boldsymbol x = (h_1, \dots, h_H, v_1,\dots,v_V)$ represent the states of the visible and hidden nodes in an RBM for some integers $V, H \ge 1$.
+Let $\boldsymbol x = (h_1, \dots, h_\nh, v_1,\dots,v_\nv)$ represent the states of the visible and hidden nodes in an RBM for some integers $\nv, \nh \ge 1$.
 Each single "binary" random variable, visible or hidden, will take its values in a common coding set $\mathcal{C}$, where we allow one of two possibilities for the coding set $\mathcal{C}=\{0,1\}$ or $\mathcal{C}=\{-1,1\}$ with "$1$" always indicating the "high" value of the variable.
 While $\mathcal{C}=\{0,1\}$ may be a natural or na{\"i}ve starting point, we show later that the coding $\mathcal{C}=\{-1,1\}$ induces better model properties for the RBM.
-We posit a parametric form for probabilities corresponding to a potential vector of states, $X = (H_1, \dots, H_H, V_1,\dots,V_V)$, for the nodes as
+We posit a parametric form for probabilities corresponding to a potential vector of states, $X = (H_1, \dots, H_\nh, V_1,\dots,V_\nv)$, for the nodes as
 
@@
 \label{eqn:pmf}
-f_{\boldsymbol \theta} (\boldsymbol x)\equiv P_{\boldsymbol \theta}(\boldsymbol X = \boldsymbol x) = \frac{\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\gamma(\boldsymbol \theta)}, \quad \boldsymbol x \in \mathcal{X}
+f_{\boldsymbol \theta} (\boldsymbol x)\equiv P_{\boldsymbol \theta}(\boldsymbol X = \boldsymbol x) = \frac{\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}{\gamma(\boldsymbol \theta)}, \quad \boldsymbol x \in \mathcal{C}^{\nh + \nv}
 \end{align}
 
-where $\boldsymbol \theta \equiv (\theta_{11}, \dots, \theta_{1H}, \dots, \theta_{V1}, \dots, \theta_{VH}, \theta_{v_1}, \dots, \theta_{v_V}, \theta_{h_1}, \dots, \theta_{h_H}) \in \mathbb{R}^{V + H + V*H}$ denotes the vector of model parameters and the denominator $\gamma(\boldsymbol \theta) = \sum\limits_{\boldsymbol x \in \mathcal{C}^{H+V}}\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)$ is the normalizing function that ensures the probabilities (\ref{eqn:pmf}) sum to one.
+where $\boldsymbol \theta \equiv (\theta_{11}, \dots, \theta_{1\nh}, \dots, \theta_{\nv 1}, \dots, \theta_{\nv \nh}, \theta_{v_1}, \dots, \theta_{v_\nv}, \theta_{h_1}, \dots, \theta_{h_\nh}) \in \mathbb{R}^{\nv + \nh + \nv*\nh}$ denotes the vector of model parameters and the denominator $\gamma(\boldsymbol \theta) = \sum\limits_{\boldsymbol x \in \mathcal{C}^{\nh+\nv}}\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)$ is the normalizing function that ensures the probabilities (\ref{eqn:pmf}) sum to one.
 For
 \begin{align}
-\boldsymbol t(\boldsymbol x) = (h_1, \dots, h_H, v_1, \dots, v_V, v_1 h_1, \dots, v_V h_H) \in \mathcal{C}^{H + V + H*V}, \boldsymbol x = (h_1, \dots, h_H, v_1, \dots, v_V) \in \mathcal{C}^{V + H}, \label{eqn:t}
+\boldsymbol t(\boldsymbol x) = (h_1, \dots, h_\nh, v_1, \dots, v_\nv, v_1 h_1, \dots, v_\nv h_\nh) \in \mathcal{C}^{\nh + \nv + \nh*\nv}, \boldsymbol x = (h_1, \dots, h_\nh, v_1, \dots, v_\nh) \in \mathcal{C}^{\nv + \nh}, \label{eqn:t}
 \end{align}
-let $\mathcal{T} = \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{X}\}$ be the vector of variables needed to compute probabilities (\ref{eqn:pmf}) in the model and write $Q_{\boldsymbol \theta}(\boldsymbol x) = \sum\limits_{i = 1}^H\sum\limits_{j=1}^V \theta_{ij} h_i v_j + \sum\limits_{i = 1}^H\theta_{hi} h_i + \sum\limits_{j = 1}^V\theta_{vj} v_j$ for the neg-potential function.
+let $\mathcal{T} = \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{C}^{\nv + \nh}\}$ be the vector of variables needed to compute probabilities (\ref{eqn:pmf}) in the model and write $Q_{\boldsymbol \theta}(\boldsymbol x) = \sum\limits_{i = 1}^\nh\sum\limits_{j=1}^\nv \theta_{ij} h_i v_j + \sum\limits_{i = 1}^\nh\theta_{hi} h_i + \sum\limits_{j = 1}^\nv\theta_{vj} v_j$ for the neg-potential function.
 Note that the RBM model is parameterized by $\boldsymbol \theta$ containing two types of parameters, main effects and interaction effects.
-The main effects parameters ($\{\theta_{v_i}, \theta_{h_j}\}_{\substack{i = 1, \dots, V,\\j = 1, \dots, H}}$) weight the values of the visible $v_i$ and hidden $h_j$ nodes in probabilities (\ref{eqn:pmf}) and the interaction effects parameters ($\theta_{ij}$) weight the values of the connections $v_i h_j$, or dependencies, between hidden and visible layers.
+The main effects parameters ($\{\theta_{v_i}, \theta_{h_j}\}_{\substack{i = 1, \dots, \nv,\\j = 1, \dots, \nh}}$) weight the values of the visible $v_i$ and hidden $h_j$ nodes in probabilities (\ref{eqn:pmf}) and the interaction effects parameters ($\theta_{ij}$) weight the values of the connections $v_i h_j$, or dependencies, between hidden and visible layers.
 
@@
 ## Degeneracy, instability, and uninterpretability... Oh my!
-The highly flexible nature of a RBM (having as it does $H + V + H V$ parameters) creates at least three kinds of potential model impropriety of concern that we will call *degeneracy, instability, and uninterpretability*.
+The highly flexible nature of a RBM (having as it does $\nh + \nv + \nh*\nv$ parameters) creates at least three kinds of potential model impropriety of concern that we will call *degeneracy, instability, and uninterpretability*.
 In this section we define these characteristics, consider how to detect them in a RBM, and point out relationships among them.
@@
 The RBM class of models can be thought to exhibit an analogous occurrence of *near-degeneracy* as follows.
-When random variables $\boldsymbol t(\boldsymbol x) = (v_1, \dots, v_V, h_1, \dots, h_H, v_1 h_1, \dots, v_V h_H ) \in \mathcal{T} \equiv \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{C}^{H + V} \}$ in the neg-potential function $Q_{\boldsymbol \theta}(\cdot)$ have a mean vector $\boldsymbol \mu(\boldsymbol \theta)$ close to the boundary of the convex hull of $\mathcal{T}$, then the RBM model can be said to be near-degenerate at $\boldsymbol \theta \in \mathbb{R}^{V + H + H*V}$.
+When random variables $\boldsymbol t(\boldsymbol x) = (v_1, \dots, v_\nv, h_1, \dots, h_\nh, v_1 h_1, \dots, v_V h_\nh ) \in \mathcal{T} \equiv \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{C}^{\nh + \nv} \}$ in the neg-potential function $Q_{\boldsymbol \theta}(\cdot)$ have a mean vector $\boldsymbol \mu(\boldsymbol \theta)$ close to the boundary of the convex hull of $\mathcal{T}$, then the RBM model can be said to be near-degenerate at $\boldsymbol \theta \in \mathbb{R}^{V + H + H*V}$.
 Here the mean of $\boldsymbol t(\boldsymbol x)$ is
@@
 }
+elpr <- function(theta, stats) {
+  require(dplyr)
+  exp(tcrossprod(stats, theta)) %>% data.frame() %>%
+    cbind(stats) %>%
+    group_by_(.dots = colnames(stats)[grepl("v", colnames(stats)) & !grepl("theta", colnames(stats))]) %>%
+    summarise_each(funs(sum), contains("X")) %>%
+    ungroup() -> marginalized
+
+  marginalized %>%
+    summarise_each(funs(max), contains("X")) %>%
+    select(contains("X")) %>%
+    data.matrix() -> max_marg
+
+  marginalized %>%
+    summarise_each(funs(min), contains("X")) %>%
+    select(contains("X")) %>%
+    data.matrix() -> min_marg
+
+  t(log(max_marg/min_marg))
+}
 
@@
   group_by(H, V, N, r1, r2) %>%
-  mutate(LHS1 = (max - min)/N,
-         LHS2 = (max - min_max - H*log(2))/N) %>%
+  mutate(LHS1 = (max - min)/V,
+         LHS2 = (max - min_max - H*log(2))/V) %>%
   summarise_each(funs(mean), LHS1, LHS2) -> max_q_summary
 
+res %>%
+  group_by(H, V, N, r1, r2) %>%
+  do(data.frame(elpr = elpr(.$samp[[1]], .$stat[[1]]))) %>%
+  ungroup() %>%
+  group_by(H, V, N, r1, r2) %>% 
+  summarise(mean_elpr = mean(elpr)) %>%
+  mutate(scaled_mean_elpr = mean_elpr/V) -> elpr_summary
+
 convex_hull_summary %>%
-  left_join(max_q_summary) %>%
+  left_join(elpr_summary) %>%
   left_join(exp_vals_summary) -> three_ways
@@
   ggplot() +
-  geom_tile(aes(x = r1, y = r2, fill = LHS2)) +
-  geom_contour(aes(x = r1, y = r2, z = LHS2), colour = "black", bins = 8) +
+  geom_tile(aes(x = r1, y = r2, fill = scaled_mean_elpr)) +
+  geom_contour(aes(x = r1, y = r2, z = scaled_mean_elpr), colour = "black", bins = 8) +
   #geom_contour(aes(x = r1, y = r2, z = mean_max_q), colour = "black", breaks = .05, size = 1.5) +
   geom_abline(aes(intercept = 0, slope = 1), alpha = .5, lty = 2) +
-  scale_fill_gradient(expression(frac(bar(R(theta)), V)), low = "yellow", high = "red") +
+  scale_fill_gradient(expression(frac(bar(ELPR(theta)), V)), low = "yellow", high = "red") +
   facet_grid(Visibles~Hiddens) +
@@
   ylab(expression(paste(paste("||", theta[interaction], "||", "/(H*V)")))) +
-  theme(aspect.ratio = 1) -> p.min_max_q
+  theme(aspect.ratio = 1) -> p.elpr
 
@@
 As the magnitude of $\\boldsymbol \\theta$ grows, so does the value of this metric, indicating typical instability in the model.", fig.height=4}
-p.min_max_q
+p.elpr
 ```

### 2016-10-31 degeneracy/grid_sample.R

@@ -80,10 +80,11 @@ grid_sample %>%
 
 grid <- inner_join(grid, grid_sample)
 
-grid %>%
+grid %>% 
   ungroup() %>%
   group_by(H, V, N, r1, r2) %>%
-  do(outside = find_prop(.$g_theta[[1]] %>% data.frame() %>% select(starts_with("exp")), .$stat[[1]], r_exp)) -> tmp
+  mutate(epsilon = (1-(1-2*r_exp)^(3/(H + V + H*V)))/2) %>%
+  do(outside = find_prop(.$g_theta[[1]] %>% data.frame() %>% select(starts_with("exp")), .$stat[[1]], .$epsilon)) -> tmp
 
 res <- inner_join(tmp, grid)
 save(res, file = "written/results_grid.RData")

### 2016-11-01 writing/draft.Rmd
Dr. Vardeman's edits

@@
 In this paper, we discuss the relationship between RBM parameter specification in the binary case and the prevalence of undesirable model behaviors such as degeneracy, instability and/or uninterpretability.
-We also describe the difficulties that arise in likelihood-based fitting and Bayes of such (highly flexible) models, especially as Gibbs sampling (quasi-Bayes) methods are often advocated for the RBM model structure.
+We also describe the difficulties that arise in likelihood-based and Bayes fitting  of such (highly flexible) models, especially as Gibbs sampling (quasi-Bayes) methods are often advocated for the RBM model structure.
 ---
@@
 The literature often suggests Markov chain Monte Carlo (MCMC) approximations to posteriors to fit these models (e.g., Gibbs sampling to exploit conditional structure in the hidden and visible variables), but little is said about acheiving convergence [@hinton2010practical, @hinton2006fast].
-Related to this, these MCMC algorithms require updating potentially many latent variables (hiddens) per observation (visibles) which can critically influence convergence in MCMC-based likelihood methods.
-
-In Section \ref{bayesian-model-fitting}, we compare three Bayesian modeling techniques (involving MCMC approximations that are computationally more accessible than direct maximum likelihood), each proposed with the goal of avoiding parts of the RBM parameter space that yield improper models.
+Related to this, these MCMC algorithms require updating potentially many latent variables (hiddens) which can critically influence convergence in MCMC-based likelihood methods.
+
+In Section \ref{bayesian-model-fitting}, we compare three Bayesian modeling techniques (involving MCMC approximations that are computationally more accessible than direct maximum likelihood), each proposed with the goal of avoiding parts of the RBM parameter space that yield unattractive models.
 As might be expected, with greater computational complexity comes an increase in fitting accuracy, but at the cost of practical feasibility.
@@
 That is, there is no "smoothed distribution" achieved in fitting a RBM model with a rigorous likelihood-based method.
-We are therefore led to be skeptical that any model built using these structures, like a deep Boltzmann machine, can achieve prediction or inference in a principled way.
+We are therefore led to be skeptical that any model built using these structures (like a deep Boltzmann machine) can achieve prediction or inference in a principled way.
 
 This paper is structured as follows.
-Section \ref{background} formally definesthe RBM including the joint distribution of hidden and visible variables and explains the model's connection to deep learning.
+Section \ref{background} formally defines the RBM including the joint distribution of hidden and visible variables and explains the model's connection to deep learning.
 Additionally, measures of model impropriety and methods of detecting it are defined.
@@
 Each single "binary" random variable, visible or hidden, will take its values in a common coding set $\mathcal{C}$, where we allow one of two possibilities for the coding set $\mathcal{C}=\{0,1\}$ or $\mathcal{C}=\{-1,1\}$ with "$1$" always indicating the "high" value of the variable.
-While $\mathcal{C}=\{0,1\}$ may be a natural or na{\"i}ve starting point, we show later that the coding $\mathcal{C}=\{-1,1\}$ induces better model properties for the RBM.
+While $\mathcal{C}=\{0,1\}$ may be a natural or naive starting point, we show later that the coding $\mathcal{C}=\{-1,1\}$ induces better model properties for the RBM.
 We posit a parametric form for probabilities corresponding to a potential vector of states, $X = (H_1, \dots, H_\nh, V_1,\dots,V_\nv)$, for the nodes as
@@
 \label{eqn:pmf}
-f_{\boldsymbol \theta} (\boldsymbol x)\equiv P_{\boldsymbol \theta}(\boldsymbol X = \boldsymbol x) = \frac{\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}{\gamma(\boldsymbol \theta)}, \quad \boldsymbol x \in \mathcal{C}^{\nh + \nv}
+f_{\boldsymbol \theta} (\boldsymbol x)\equiv P_{\boldsymbol \theta}(\boldsymbol X = \boldsymbol x) = \frac{\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}{\gamma(\boldsymbol \theta)}, \quad \boldsymbol x \in \mathcal{C}^{\nh + \nv}
 \end{align}
@@
 
-Due to the potential size of the model, the normalizing constant $\gamma(\boldsymbol \theta)$ can be practically impossible to calculate, making it impossible to estimate the model parameter vector in an obvious fashion.
-A kind of Gibbs sampling can be tried, which is natural due to conditional independence architecture of the RBM (i.e.
+Due to the potential size of the model, the normalizing constant $\gamma(\boldsymbol \theta)$ can be practically impossible to calculate, making simple estimation of the model parameter vector problematic.
+A kind of Gibbs sampling can be tried (which is natural due to conditional independence architecture of the RBM, i.e.
 visibles given hiddens or vice versa).
-Specifically, the conditional independence of nodes in each layer, given those nodes in the other layer allows for stepwise simulation of both the visible and hidden layers.
+Specifically, the conditional independence of nodes in each layer, given those nodes in the other layer allows for stepwise simulation of both hidden layers and model parameters possible.
 
@@
 ## Degeneracy, instability, and uninterpretability... Oh my!
-The highly flexible nature of a RBM (having as it does $\nh + \nv + \nh*\nv$ parameters) creates at least three kinds of potential model impropriety of concern that we will call *degeneracy, instability, and uninterpretability*.
+The highly flexible nature of a RBM (having as it does $\nh + \nv + \nh*\nv$ parameters) creates at least three kinds of potential model impropriety of concern that we will call *degeneracy*, *instability*, and *uninterpretability*.
 In this section we define these characteristics, consider how to detect them in a RBM, and point out relationships among them.
@@
 where $\boldsymbol \theta \in \Theta \subset \mathbb{R}^q$ is the model parameter, and $t: \mathcal{X} \rightarrow \mathbb{R}^q$ are statistics based on the adjacency matrix of a graph.
-Again, $\gamma(\boldsymbol \theta) = \sum_{\boldsymbol x \in \mathcal{X}} \exp\left(\boldsymbol \theta^T \boldsymbol t(\boldsymbol x)\right)$ is the normalizing function.
+Here, as earlier, $\gamma(\boldsymbol \theta) = \sum_{\boldsymbol x \in \mathcal{X}} \exp\left(\boldsymbol \theta^T \boldsymbol t(\boldsymbol x)\right)$ is the normalizing function.
 Let $C$ denote the convex hull spanned by the potential outcomes of sufficient statistics, $\{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{X}\}$.
-@handcock2003assessing classify an exponentially parametrized random graph model at $\boldsymbol \theta$ as *near-degenerate* if the mean value of sufficient statistics under $\boldsymbol \theta$, $\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta}\boldsymbol t(X)$, is close to the boundary of $C$.
-
-The RBM class of models can be thought to exhibit an analogous occurrence of *near-degeneracy* as follows.
-When random variables $\boldsymbol t(\boldsymbol x) = (v_1, \dots, v_\nv, h_1, \dots, h_\nh, v_1 h_1, \dots, v_V h_\nh ) \in \mathcal{T} \equiv \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{C}^{\nh + \nv} \}$ in the neg-potential function $Q_{\boldsymbol \theta}(\cdot)$ have a mean vector $\boldsymbol \mu(\boldsymbol \theta)$ close to the boundary of the convex hull of $\mathcal{T}$, then the RBM model can be said to be near-degenerate at $\boldsymbol \theta \in \mathbb{R}^{V + H + H*V}$.
+@handcock2003assessing classify an exponentially parametrized random graph model at $\boldsymbol \theta$ as *near-degenerate* if the mean value of sufficient statistics under $\boldsymbol \theta$, $\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta}\boldsymbol t( \boldsymbol X)$, is close to the boundary of $C$.
+
+The RBM class of models can be thought to exhibit an analogous form of *near-degeneracy* as follows.
+When random variables $\boldsymbol t(\boldsymbol x) = (v_1, \dots, v_\nv, h_1, \dots, h_\nh, v_1 h_1, \dots, v_V h_\nh ) \in \mathcal{T} \equiv \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{C}^{\nh + \nv} \}$ in the neg-potential function $Q_{\boldsymbol \theta}(\cdot)$ have a mean vector $\boldsymbol \mu(\boldsymbol \theta)$ close to the boundary of the convex hull of $\mathcal{T}$, then the RBM model can be said to be near-degenerate at $\boldsymbol \theta \in \mathbb{R}^{\nv + \nh + \nh*\nv}$.
 Here the mean of $\boldsymbol t(\boldsymbol x)$ is
 $$
-\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta} \boldsymbol t(\boldsymbol X)  = \sum\limits_{x \in \mathcal{X}} \left\{ \boldsymbol t(\boldsymbol x) f_{\boldsymbol \theta} (\boldsymbol x) \right\} = \sum\limits_{x \in \mathcal{X}} \left\{ \boldsymbol t(\boldsymbol x)\dfrac{\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}\right\}.
+\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta} \boldsymbol t(\boldsymbol X)  = \sum\limits_{x \in \mathcal{C}^{\nv + \nh}} \left\{ \boldsymbol t(\boldsymbol x) f_{\boldsymbol \theta} (\boldsymbol x) \right\} = \sum\limits_{x \in \mathcal{C}^{\nv + \nh}} \left\{ \boldsymbol t(\boldsymbol x)\dfrac{\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{C}^{\nh + \nv}}\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}\right\}.
 $$
@@
 Instability can be roughly characterized as excessive sensitivity in the model, where small changes in the components of potential data outcomes, $\boldsymbol x$, may lead to substantial changes in the probability function $f_{\boldsymbol \theta}(\boldsymbol x)$.
-To quantify "instability" more rigorously (particularly beyond the definition given by @schweinberger2011instability) it is useful to consider how the RBM model might be expanded to incorporate more and more visibles When increasing the size of a RBM model, it becomes necessary to grow the number of model parameters (and in this process one may also arbitrarily expand the number of hidden variables used).
-To this end, let $\boldsymbol \theta_{V} \equiv (\theta_{v_1}, \dots, \theta_{v_V}, \theta_{h_1}, \dots, \theta_{h_H}, \theta_{11}, \dots, \theta_{VH}), V \ge 1$, denote an element of a sequence of RBM parameters indexed by the number $V$ of visibles $(V_1, \dots, V_V)$ and define a (scaled) extremal log-probability ratio of the RBM model at $\boldsymbol \theta_{V}$ as
+To quantify "instability" more rigorously (particularly beyond the definition given by @schweinberger2011instability) it is useful to consider how the RBM model might be expanded to incorporate more and more visibles.
+When increasing the size of a RBM model, it becomes necessary to grow the number of model parameters (and in this process one may also arbitrarily expand the number of hidden variables used).
+To this end, let $\boldsymbol \theta_{\nv} \equiv (\theta_{v_1}, \dots, \theta_{v_\nv}, \theta_{h_1}, \dots, \theta_{h_\nh}, \theta_{11}, \dots, \theta_{\nv \nh}), \nv \ge 1$, denote an element of a sequence of RBM parameters indexed by the number $V$ of visibles $(V_1, \dots, V_V)$ and define a (scaled) extremal log-probability ratio of the RBM model at $\boldsymbol \theta_{V}$ as
 \begin{align}
-\frac{1}{V} \log \left[\frac{\max\limits_{(v_1, \dots, v_V) \in \mathcal{C}^V}P_{\boldsymbol \theta_V}(v_1, \dots, v_V)}{\min\limits_{(v_1, \dots, v_V) \in \mathcal{C}^V}P_{\boldsymbol \theta_V}(v_1, \dots, v_V)}\right] \equiv \frac{1}{V} \text{ELPR}(\boldsymbol \theta_V) \label{eqn:elpr}
+\frac{1}{\nv} \log \left[\frac{\max\limits_{(v_1, \dots, v_\nv) \in \mathcal{C}^\nv}P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}{\min\limits_{(v_1, \dots, v_\nv) \in \mathcal{C}^\nv}P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}\right] \equiv \frac{1}{\nv} \text{ELPR}(\boldsymbol \theta_\nv) \label{eqn:elpr}
 \end{align}
-where $P_{\boldsymbol \theta_V}(v_1, \dots, v_V) \propto \sum_{\{h_j\}\in \mathcal{C}}\exp\left(\sum_{i = 1}^V \theta_{v_i} v_i + \sum_{j = 1}^H \theta_{h_j} h_j + \sum_{i = 1}^V \sum_{j = 1}^H \theta_{ij} v_i h_j\right)$ is the RBM probability of observing outcome $(v_1, \dots, v_V)$ for the visible variables $(V_1, \dots, V_V)$ under parameters $\boldsymbol \theta_V$.
-
-In formulating a RBM model for a potentially large number of visibles (i.e., $V \rightarrow \infty$), we will say that the ratio (\ref{eqn:elpr}) needs to stay bounded for the sequence of RBM models to be stable.
+where $P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv) \propto \sum_{\{h_j\}\in \mathcal{C}}\exp\left(\sum_{i = 1}^\nv \theta_{v_i} v_i + \sum_{j = 1}^\nh \theta_{h_j} h_j + \sum_{i = 1}^\nv \sum_{j = 1}^\nh \theta_{ij} v_i h_j\right)$ is the RBM probability of observing outcome $(v_1, \dots, v_\nv)$ for the visible variables $(V_1, \dots, V_\nv)$ under parameters $\boldsymbol \theta_\nv$.
+
+In formulating a RBM model for a potentially large number of visibles (i.e., as $\nv \rightarrow \infty$), we will say that the ratio (\ref{eqn:elpr}) needs to stay bounded for the sequence of RBM models to be stable.
 That is, we make the following convention
@@
 \begin{defn}[S-unstable RBM]
-Let $\boldsymbol \theta_V \in \mathbb{R}^{V + H +  H*V}, v \ge 1$, be an element of a sequence of RBM parameters where the number of hiddens, $H \equiv H(V) \ge 1$, can be an arbitrary function of the number $V$ of visibles.
+Let $\boldsymbol \theta_\nv \in \mathbb{R}^{\nv + \nh +  \nh*\nv}, \nv \ge 1$, be an element of a sequence of RBM parameters where the number of hiddens, $\nh \equiv \nh(\nv) \ge 1$, can be an arbitrary function of the number $\nv$ of visibles.
 A RBM model formulation is \emph{Scheinberger-unstable} or \emph{S-unstable} if
 \begin{align*}
-\lim\limits_{V \rightarrow \infty} \frac{1}{V} \text{ELPR}(\boldsymbol \theta_V) \equiv \lim\limits_{V \rightarrow \infty} \frac{1}{V} \log \left[\frac{\max\limits_{(v_1, \dots, v_V) \in \mathcal{C}^V}P_{\boldsymbol \theta}(v_1, \dots, v_V)}{\min\limits_{(v_1, \dots, v_V) \in \mathcal{C}^V}P_{\boldsymbol \theta}(v_1, \dots, v_V)}\right] = \infty.
+\lim\limits_{\nv \rightarrow \infty} \frac{1}{\nv} \text{ELPR}(\boldsymbol \theta_\nv) \equiv \lim\limits_{\nv \rightarrow \infty} \frac{1}{\nv} \log \left[\frac{\max\limits_{(v_1, \dots, v_\nv) \in \mathcal{C}^\nv}P_{\boldsymbol \theta}(v_1, \dots, v_\nv)}{\min\limits_{(v_1, \dots, v_\nv) \in \mathcal{C}^\nv}P_{\boldsymbol \theta}(v_1, \dots, v_\nv)}\right] = \infty.
 \end{align*}
 \end{defn}
-In other words, given any $C > 0$, there exists an integer $V_C > 0$ so that $\frac{1}{V}\text{ELPR}(\boldsymbol \theta_V) > C$ for all $V \ge V_C$.
-
-This definition of *S-unstable* is a generalization or re-interpretation of the "unstable" concept of @schweinberger2011instability in that here RBM models for visibles $(v_1, \dots, v_V)$ do not form an exponential family and the dimensionality of $\boldsymbol \theta_V$ is not fixed, but rather grows with $V$.
+In other words, given any $C > 0$, there exists an integer $n_C > 0$ so that $\frac{1}{\nv}\text{ELPR}(\boldsymbol \theta_\nv) > C$ for all $\nv \ge n_C$.
+
+This definition of *S-unstable* is a generalization or re-interpretation of the "unstable" concept of @schweinberger2011instability in that here RBM models for visibles $(v_1, \dots, v_\nv)$ do not form an exponential family and the dimensionality of $\boldsymbol \theta_\nv$ is not fixed, but rather grows with $\nv$.
 
@@
 $$
-\Delta(\boldsymbol \theta_V) \equiv \max \left\{\log \frac{P_{\boldsymbol \theta_V}(v_1, \dots, v_V)}{P_{\boldsymbol \theta_V}(v_1^*, \dots, v_V^*)} : (v_1, \dots, v_V) \& (v_1^*, \dots, v_V^*) \in \mathcal{C}^V \text{ differ by exactly one component}\right\},
-$$
-denoting the biggest log-probability ratio for a one-pixel (one component) change in data outcomes (visibles) at a RBM parameter $\boldsymbol \theta_V$. We then have the following result.
+\Delta(\boldsymbol \theta_\nv) \equiv \max \left\{\log \frac{P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}{P_{\boldsymbol \theta_\nv}(v_1^*, \dots, v_\nv^*)} : (v_1, \dots, v_\nv) \& (v_1^*, \dots, v_\nv^*) \in \mathcal{C}^\nv \text{ differ by exactly one component}\right\},
+$$
+denoting the biggest log-probability ratio for a one-pixel (one component) change in data outcomes (visibles) at a RBM parameter $\boldsymbol \theta_\nv$. We then have the following result.
 
@@
 \label{prop:instab}
-Let $C > 0$ and let $\text{ELPR}(\boldsymbol \theta_V)$ be as in (\ref{eqn:elpr}) for an integer $V \ge 1$.
-If $\frac{1}{V}\text{ELPR}_V(\boldsymbol \theta) > C$, then $\Delta_V(\boldsymbol \theta) > C$.
+Let $C > 0$ and let $\text{ELPR}(\boldsymbol \theta_\nv)$ be as in (\ref{eqn:elpr}) for an integer $\nv \ge 1$.
+If $\frac{1}{\nv}\text{ELPR}(\boldsymbol \theta_\nv) > C$, then $\Delta(\boldsymbol \theta_\nv) > C$.
 \end{prop}
@@
 To see this, define a modal set of possible outcomes (i.e.
-set of highest probability outcomes) in RBM models with parameters $\boldsymbol \theta_V, V \ge 1$ as
-$$
-M_{\epsilon, \boldsymbol \theta_V} \equiv \left\{(v_1, \dots, v_V) \in \mathcal{C}^V: \log P_{\boldsymbol \theta_V}(v_1, \dots, v_V) > (1-\epsilon)\max\limits_{(v_1^*, \dots, v_V^*)}P_{\boldsymbol \theta_V}(v_1^*, \dots, v_V^*) + \epsilon\min\limits_{(v_1^*, \dots, v_V^*)}P_{\boldsymbol \theta_V}(v_1^*, \dots, v_V^*) \right\}
+set of highest probability outcomes) in RBM models with parameters $\boldsymbol \theta_\nv, \nv \ge 1$ as
+$$
+M_{\epsilon, \boldsymbol \theta_\nv} \equiv \left\{\boldsymbol v \in \mathcal{C}^\nv: \log P_{\boldsymbol \theta_\nv}(\boldsymbol v) > (1-\epsilon)\max\limits_{\boldsymbol v^*}P_{\boldsymbol \theta_\nv}(\boldsymbol v^*) + \epsilon\min\limits_{\boldsymbol v^*}P_{\boldsymbol \theta_\nv}(\boldsymbol v^*) \right\}
 $$
@@
 \label{prop:degen}
-For an S-unstable RBM model and any given $0 < \epsilon < 1$, $P_{\boldsymbol \theta_V}\left((v_1, \dots, v_V) \in M_{\epsilon, \boldsymbol \theta_V}\right) \rightarrow 1$ holds as $V \rightarrow \infty$.
+For an S-unstable RBM model sequence and any given $0 < \epsilon < 1$, $P_{\boldsymbol \theta_\nv}\left((v_1, \dots, v_\nv) \in M_{\epsilon, \boldsymbol \theta_\nv}\right) \rightarrow 1$ as $\nv \rightarrow \infty$.
 \end{prop}
@@
 
-\begin{remark}
-\label{remark:bnd}
-To later numerically investigate S-unstable RBM model sequences, it is useful to define a computationally tractable lower bound on (\ref{eqn:elpr}) given by
-\begin{align}
-\frac{1}{V}R(\boldsymbol \theta_V) \text{ for } R(\boldsymbol \theta_V) \equiv \max_{\boldsymbol v} \max_{\boldsymbol h}Q_{\boldsymbol \theta_V}(\boldsymbol x) - \min_{\boldsymbol v}\max_{\boldsymbol h}Q_{\boldsymbol \theta_V}(\boldsymbol x) - H\log 2.
-\label{eqn:r}
-\end{align}
-That is, in (\ref{eqn:elpr}),
-\begin{align}
-\frac{1}{V} \text{ELPR}(\boldsymbol \theta_V) \equiv \frac{1}{V} \log \left[\frac{\max\limits_{(v_1, \dots, v_V) \in \mathcal{C}^V}P_{\boldsymbol \theta}(v_1, \dots, v_V)}{\min\limits_{(v_1, \dots, v_V) \in \mathcal{C}^V}P_{\boldsymbol \theta}(v_1, \dots, v_V)}\right] \ge \frac{1}{V}R(\boldsymbol \theta_V).
-\label{eqn:lowerbnd}
-\end{align}
-\end{remark}
-The appendix provides a proof of (\ref{eqn:lowerbnd}).
-The benefit of $R(\boldsymbol \theta_V)$ is that $R(\boldsymbol \theta_V)/V$ is simple to compute (i.e.
-unlike (\ref{eqn:elpr}), requiring no summation over hidden variables) and that Propositions \ref{prop:instab}-\ref{prop:degen} continue to hold upon replacing $\text{ELPR}(\boldsymbol \theta_V)/V$ with $R(\boldsymbol \theta_V)/V$ due to (\ref{eqn:lowerbnd}).
-Hence, large values of $R(\boldsymbol \theta_V)/V$ also indicate S-unstable RBM model sequences.
-
 ### Uninterpretability
@@
 We adapt this notion to RBM models as follows.
-Note that, in a RBM, the parameters $\theta_{v_1}, \dots, \theta_{v_V}$ and $\theta_{h_1}, \dots, \theta_{h_H}$ are naturally associated with main effects of visible and hidden variables and can be interpreted as (logit functions of) means for variables $V_1, \dots, V_V, H_1, \dots, H_V$ in a model with no interaction parameters, $\theta_{ij} = 0, i = 1, \dots, V, j = 1, \dots, H$.
-Hence, these main effect parameters have a clear interpretation under an independence model ($\theta_{ij} = 0$) but this interpretation (as well as an understanding of the nature of dependence) can break down as interaction parameters increase in magnitude relative to the size of the main effects.
-In such case, the main effect parameters $\theta_{v_1}$ and $\theta_{h_j}$ are no longer interpretable as originally intended in the modes (statements of marginal means) and the dependence parameters are so large as to dominate the entire model probability structure (destroying the model interpretation of dependence as local, conditional modifications of all overall marginal mean structure).
-To assess which parameter values $\boldsymbol \theta$ may cause difficulties in interpretation, we use the difference $\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] - \text{E}\left[\boldsymbol X | \boldsymbol \theta^* \right]$ between two model expectations: E$\left[\boldsymbol X | \boldsymbol \theta\right]$ at $\boldsymbol \theta$ and expectations E$\left[\boldsymbol X | \boldsymbol \theta^* \right]$ given independence where $\boldsymbol \theta^*$ matches $\boldsymbol \theta$ for all main effects but otherwise has $\theta_{ij} = 0$ for $i = 1, \dots, V, j = 1, \dots, H$.
+Note that, in a RBM, the parameters $\theta_{v_1}, \dots, \theta_{v_\nv}$ and $\theta_{h_1}, \dots, \theta_{h_\nh}$ are naturally associated with main effects of visible and hidden variables and can be interpreted as (logit functions of) means for variables $V_1, \dots, V_\nv, H_1, \dots, H_\nh$ in a model with no interaction parameters, $\theta_{ij} = 0, i = 1, \dots, \nv, j = 1, \dots, \nh$.
+Hence, these main effect parameters have a clear interpretation under an independence model (one with $\theta_{ij} = 0$) but this interpretation (as well as an understanding of the nature of dependence) can break down as interaction parameters increase in magnitude relative to the size of the main effects.
+In such case, the main effect parameters $\theta_{v_1}$ and $\theta_{h_j}$ are no longer interpretable as originally intended in the modes (statements of marginal means) and the dependence parameters are so large as to dominate the entire model probability structure (destroying the model interpretation of dependence as local conditional modifications of all overall marginal mean structure).
+To assess which parameter values $\boldsymbol \theta$ may cause difficulties in interpretation, we use the difference $\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] - \text{E}\left[\boldsymbol X | \boldsymbol \theta^* \right]$ between two model expectations: E$\left[\boldsymbol X | \boldsymbol \theta\right]$ at $\boldsymbol \theta$ and expectations E$\left[\boldsymbol X | \boldsymbol \theta^* \right]$ given independence where $\boldsymbol \theta^*$ matches $\boldsymbol \theta$ for all main effects but otherwise has $\theta_{ij} = 0$ for $i = 1, \dots, \nv, j = 1, \dots, \nh$.
 Hence, $\boldsymbol \theta^*$ and $\boldsymbol \theta$ have the same main effects but $\boldsymbol \theta^*$ has $0$ dependence parameters.
@@
 $$
-\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] = \sum\limits_{\boldsymbol x \in \mathcal{X}} \boldsymbol x f_{\boldsymbol \theta}(\boldsymbol x) = \sum\limits_{\boldsymbol x \in \mathcal{X}} \boldsymbol x \frac{\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}
+\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] = \sum\limits_{\boldsymbol x \in \mathcal{C}^{\nv + \nh}} \boldsymbol x f_{\boldsymbol \theta}(\boldsymbol x) = \sum\limits_{\boldsymbol x \in \mathcal{C}^{\nv + \nh}} \boldsymbol x \frac{\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{C}^{\nv + \nh}}\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}
 $$
@@
 \text{E}\left[\boldsymbol X | \boldsymbol 
-\theta^*\right] &= \sum\limits_{\boldsymbol x \in \mathcal{X}} \boldsymbol x \frac{\exp\left(\sum\limits_{i = 1}^V \theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)} \\
+\theta^*\right] &= \sum\limits_{\boldsymbol x \in \mathcal{C}^{\nv + \nh}} \boldsymbol x \frac{\exp\left(\sum\limits_{i = 1}^\nv \theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{C}^{\nv + \nh}}\exp\left(\sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)} \\
 \end{align*}
@@
 Because it seems most common, we shall begin by employing $0/1$ encoding of binary variables (both $h_1$ and $v_1$ taking values in $\mathcal{C} = \{0,1\}$).
-Eventually we'll note in Section \ref{data-coding-to-mitigate-degeneracy} that $-1/1$ coding has substantial advantages.
+(Eventually we'll argue in Section \ref{data-coding-to-mitigate-apparent-degeneracy} that $-1/1$ coding has substantial advantages.)
 
@@
 For this small model, we are able to investigate the symptoms of model impropriety, beginning with near-degeneracy.
-To this end, recall that one characterization requires the convex hull of statistics $\mathcal{T} = \{\boldsymbol t(\boldsymbol x): \boldsymbol x = (v_1, h_1) \in \{0,1\}^2\} \equiv \{(v_1, h_1, v_1 h_1): v_1, h_1 \in \{0,1\}\}$ appearing in the RBM probabilities for this model.
+To this end, recall that one characterization requires the convex hull of statistics
+$$\mathcal{T} = \{\boldsymbol t(\boldsymbol x): \boldsymbol x = (v_1, h_1) \in \{0,1\}^2\} \equiv \{(v_1, h_1, v_1 h_1): v_1, h_1 \in \{0,1\}\}$$ 
+appearing in the RBM probabilities for this model.
 As this set is in three dimensions, we are able to explicitly illustrate the shape of boundary of the convex hull of $\mathcal{T}$ and explore the behavior of the mean parameterization of the parameters.
-Figure \ref{fig:toyhull} shows the convex hull of our "statistic space", $\mathcal{T} \subset \{0,1\}^3$, for this toy problem from multiple perspectives enclosed by the unit cube, representing the convex hull of $\{0,1\}^3$ space.
+Figure \ref{fig:toyhull} shows the convex hull of our "statistic space," $\mathcal{T} \subset \{0,1\}^3$, for this toy problem from two perspectives (enclosed by the unit cube, representing the convex hull of $\{0,1\}^3$ space).
 In this small model, note that the convex hull of $\mathcal{T}$ does not fill the unrestricted hull of $\{0,1\}^3$ because of the relationship between the elements of $\mathcal{T} = \{(v_1, h_1, v_1 h_1 : v_1, h_1 \in \{0, 1\} \}$ (i.1.
@@
   \caption{The convex hull of the "statistic space"" in three dimensions for the toy RBM with one visible and one hidden node.
-The convex hull of $\mathcal{T} = \{\boldsymbol t(\boldsymbol x) : \boldsymbol x \in \mathcal{X}\}$ does not fill the unit cube because of the relationship between the elements of $\mathcal{T}$.}
+The convex hull of $\mathcal{T} = \{\boldsymbol t(\boldsymbol x) : \boldsymbol x \in \mathcal{C}^{\nv + \nh}\}$ does not fill the unit cube because of the relationship between the elements of $\mathcal{T}$.}
 \label{fig:toyhull}
@@
 where $\gamma(\boldsymbol \theta) = \sum\limits_{h_1 = 0}^1\sum\limits_{v_1 = 0}^1 \exp(\theta_{11} h_1 v_1 + \theta_{h_1}h_1 + \theta_{v_1}v_1)$.
-The three parametric functions giving $\boldsymbol \mu(\boldsymbol \theta)$ can be visualized in Figure \ref{fig:degen_toy}.
-In examining these, we see that as values of coordinates of $\boldsymbol \theta$ grow larger in magnitude, at least one mean function for the entries of $\boldsymbol t(\boldsymbol x)$ is approaching a value 0 or 1, indicating a potential for $\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta} \boldsymbol t(\boldsymbol x)$ to be near to the boundary of the convex hull of $\mathcal{T}$, as a sign of model near-degeneracy.
+The three parametric coordinate functions of $\boldsymbol \mu(\boldsymbol \theta)$ can be visualized in Figure \ref{fig:degen_toy}.
+(Contour plots for three coordinate functions are shown in columns for various values of $\theta_{11}$.) In examining these, we see that as values of coordinates of $\boldsymbol \theta$ grow larger in magnitude, at least one mean function for the entries of $\boldsymbol t(\boldsymbol x)$ is approaching a value 0 or 1, forcing $\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta} \boldsymbol t(\boldsymbol x)$ to be near to the boundary of the convex hull of $\mathcal{T}$, as a sign of model near-degeneracy.
 Thus, for a very small example we can see the relationship between values of $\boldsymbol \theta$ and model degeneracy.
 
-```{r near_dependency, fig.cap="\\label{fig:degen_toy}The three parametric mean functions of sufficient statistics for a RBM with one visible and one hidden node.", fig.height = 4, fig.align='center', message=FALSE, cache=TRUE}
+```{r near_dependency, fig.cap="\\label{fig:degen_toy}Contour plots for the three parametric mean functions of sufficient statistics for a RBM with one visible and one hidden node.", fig.height = 4, fig.align='center', message=FALSE, cache=TRUE}
 possibles <- stats(1, 1, type = "binary")
@@
 
-Secondly, we can look at $R(\boldsymbol \theta)$ from (\ref{eqn:r}) for this tiny model in order to consider model instability as a function of RBM parameters.
-Recall that larger values of $R(\boldsymbol \theta)$ imply an extreme sensitivity of the model probabilities $f_{\boldsymbol \theta}(\boldsymbol x)$ to small changes in $\boldsymbol x$ (see Remark \ref{remark:bnd} and Proposition \ref{prop:instab}).
-The quantity $R(\boldsymbol \theta)$ is computed for this small RBM as
+Secondly, we can look at $\text{ELPR}(\boldsymbol \theta)$ from (\ref{eqn:elpr}) for this tiny model in order to consider model instability as a function of RBM parameters.
+Recall that large values of $\text{ELPR}(\boldsymbol \theta)$ are associated with an extreme sensitivity of the model probabilities $f_{\boldsymbol \theta}(\boldsymbol x)$ to small changes in $\boldsymbol x$ (see Proposition \ref{prop:instab}).
+The quantity $\text{ELPR}(\boldsymbol \theta)$ for this small RBM is
 
 \begin{align*}
-R(\boldsymbol \theta) &= \max_{\boldsymbol v} \max_{\boldsymbol h}Q_{\boldsymbol \theta}(\boldsymbol x) - \min_{\boldsymbol v}\max_{\boldsymbol h}Q_{\boldsymbol \theta}(\boldsymbol x) - H\log 2\\
-&= \max_{\boldsymbol v} \max_{\boldsymbol h}\left[ \theta_{11} h_1 v_1 + \theta_{h_1}h_1 + \theta_{v_1} v_1 \right] -  \min_{\boldsymbol v}\max_{\boldsymbol h}\left[ \theta_{11} h_1 v_1 + \theta_{h_1}h_1 + \theta_{v_1} v_1 \right] -\log 2\\
-&=\max\left\{ \theta_{11} + \theta_{h_1} + \theta_{v_1}, \theta_{h_1}, \theta_{v_1}, 0  \right\} \\
-&\qquad - \min\left\{\theta_{v_1} + \max\left\{\theta_{11} + \theta_{h1}, -\theta_{11} - \theta_{h1}\right\}, -\theta_{v_1} + \max\left\{-\theta_{11} + \theta_{h1}, \theta_{11} - \theta_{h1}\right\}\right\} - \log2.\\
+\text{ELPR}(\boldsymbol \theta) &= \log \left[\frac{\max\limits_{(v_1, \dots, v_\nv) \in \mathcal{C}^\nv}P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}{\min\limits_{(v_1, \dots, v_\nv) \in \mathcal{C}^\nv}P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}\right]\\
+&= \log \left[\frac{\max\limits_{v_1 \in \mathcal{C}}\sum\limits_{h_1 \in \mathcal{C}}\exp\left\{\theta_{11} h_1 v_1 + \theta_{h_1}h_1 + \theta_{v_1} v_1 \right\}}{\min\limits_{v_1 \in \mathcal{C}}\sum\limits_{h_1 \in \mathcal{C}}\exp\left\{\theta_{11} h_1 v_1 + \theta_{h_1}h_1 + \theta_{v_1} v_1 \right\}}\right].
+\\
 \end{align*}
 
-Figure \ref{fig:instab} shows $R(\boldsymbol \theta)/V$ for various values of $\boldsymbol \theta$ in this model with $V = 1$.
-We can see that this quantity is increasing for larger magnitudes of $\boldsymbol \theta$, especially for large values of the dependence/interaction parameter $\theta_{11}$.
+Figure \ref{fig:instab} shows contour plots of $\text{ELPR}(\boldsymbol \theta)/\nv$ for various values of $\boldsymbol \theta$ in this model with $\nv = 1$.
+We can see that this quantity is large for large magnitudes of $\boldsymbol \theta$, especially for large values of the dependence/interaction parameter $\theta_{11}$.
 This suggests instability as $|\boldsymbol \theta|$ becomes large, agreeing also with the concerns about near-degeneracy produced by consideration of $\boldsymbol \mu(\boldsymbol \theta)$.
 
-```{r instab, fig.cap="\\label{fig:instab}$R(\\boldsymbol \\theta)/V$ for various values of $\\boldsymbol \\theta$ for the tiny example model.
-(Recall that here $V$ is the number of visible nodes and here is $1$.
-This quantity is increasing for larger magnitudes of $\\boldsymbol \\theta$.", fig.height = 4, fig.align='center'}
+```{r instab, fig.cap="\\label{fig:instab}$\\text{ELPR}(\\boldsymbol \\theta)/\\nv$ for various values of $\\boldsymbol \\theta$ for the tiny example model.
+(Recall that here $\\nv$ is the number of visible nodes and here is $1$.
+This quantity is large for large magnitudes of $\\boldsymbol \\theta$.", fig.height = 4, fig.align='center'}
 expand.grid(theta_h1 = seq(-5, 5, length.out = 21),
@@
   group_by(theta_v1, theta_h1, theta_11, Q.v1) %>%
-  do(data.frame(maxq = max(.$Q.V4))) %>%
+  do(data.frame(marginal = sum(exp(.$Q.V4)))) %>%
   group_by(theta_v1, theta_h1, theta_11) %>%
-  do(data.frame(q_max = max(.$maxq) - min(.$maxq) - log(2))) %>%
+  do(data.frame(max_marg = max(.$marginal),
+                min_marg = min(.$marginal))) %>%
+  mutate(elpr = log(max_marg/min_marg)) %>%
   ungroup() %>%
@@
   ggplot() +
-  geom_tile(aes(theta_v1, theta_h1, fill = q_max)) +
-  geom_contour(aes(theta_v1, theta_h1, z = q_max), colour = "black") +
+  geom_tile(aes(theta_v1, theta_h1, fill = elpr)) +
+  geom_contour(aes(theta_v1, theta_h1, z = elpr), colour = "black") +
   facet_grid(. ~ theta_11, labeller = label_parsed) +
@@
   ylab(expression(theta[h1])) +
-  scale_fill_gradient(expression(frac(R(theta), V)), low = "yellow", high = "red") +
+  scale_fill_gradient(expression(frac(ELPR(theta), V)), low = "yellow", high = "red") +
   theme(aspect.ratio = 1)
@@
 
-Finally to consider the effect of $\boldsymbol \theta$ on potential model uninterpretability, we can look at the difference between model expectations, E$\left[\boldsymbol X | \boldsymbol \theta\right]$, and expectations given independence, E$\left[\boldsymbol X | \boldsymbol \theta^*\right ]$ for the tiny toy RBM model where $\boldsymbol X = (V_1, H_1, V_1 H_1)$; this is given by
+Finally to consider the effect of $\boldsymbol \theta$ on potential model uninterpretability, we can look at the difference between model expectations, E$\left[\boldsymbol X | \boldsymbol \theta\right]$, and expectations given independence, E$\left[\boldsymbol X | \boldsymbol \theta^*\right ]$ for the tiny toy RBM model where $\boldsymbol X = (V_1, H_1, V_1 H_1)$.
+This is given by
 
@@
 
-Again, we can inspect these coordinate functions of $\boldsymbol \theta$ to look for a relationship between parameter values and large values of $||\text{E}[\boldsymbol X|\boldsymbol \theta] - \text{E}[\boldsymbol X| \boldsymbol \theta^*]||$ as a signal of uninterpretability for the toy RBM.
+Again, we can inspect these coordinate functions of this vector difference to look for a relationship between parameter values and large values of $||\text{E}[\boldsymbol X|\boldsymbol \theta] - \text{E}[\boldsymbol X| \boldsymbol \theta^*]||$ as a signal of uninterpretability for the toy RBM.
 
@@
 
-Figure \ref{fig:uninterp} shows that the absolute difference between coordinates of the vector of model expectations, E$\left[\boldsymbol X | \boldsymbol \theta\right]$, and corresponding expectations E$\left[\boldsymbol X | \boldsymbol \theta^*\right ]$ given independence grow for the toy RBM as the values of $\boldsymbol \theta$ are farther from zero, especially for large magnitudes of $\theta_{11}$.
-This is a third indication supporting that parameter vectors of large magnitude could lead to model impropriety in a RBM.
-
-### Data coding to mitigate degeneracy
+Figure \ref{fig:uninterp} shows that the absolute difference between coordinates of the vector of model expectations, E$\left[\boldsymbol X | \boldsymbol \theta\right]$ and corresponding expectations E$\left[\boldsymbol X | \boldsymbol \theta^*\right ]$ given independence grow for the toy RBM as the values of $\boldsymbol \theta$ are farther from zero, especially for large magnitudes of $\theta_{11}$.
+This is a third indication supporting that parameter vectors of large magnitude lead to model impropriety in a RBM.
+
+### Data coding to mitigate apparent degeneracy
 Multiple encodings of the binary variables are possible.
-For example, we could allow hiddens $(H_1, \dots, H_H) \in \{0,1\}^H$ and visibles $(V_1, \dots, V_V) \in \{0,1\}^V$, as in the previous sections or we could instead encode the state of the variables as $\{-1,1\}^H$ and $\{-1,1\}^V$.
-This will result in variables $\boldsymbol t(\boldsymbol X)$ from (\ref{eqn:t}) satisfying $\boldsymbol t(\boldsymbol x) \in \{0,1\}^{H + V + HV}$ or $\boldsymbol t(\boldsymbol x) \in \{-1,1\}^{H + V + HV}$ depending on how we encode "on" and "off" states in the nodes.
+For example, we could allow hiddens $(H_1, \dots, H_\nh) \in \{0,1\}^\nh$ and visibles $(V_1, \dots, V_\nv) \in \{0,1\}^\nv$, as in the previous sections or we could instead encode the state of the variables as $\{-1,1\}^\nh$ and $\{-1,1\}^\nv$.
+This will result in variables $\boldsymbol t(\boldsymbol X)$ from (\ref{eqn:t}) satisfying $\boldsymbol t(\boldsymbol x) \in \{0,1\}^{\nh + \nv + \nh*\nv}$ or $\boldsymbol t(\boldsymbol x) \in \{-1,1\}^{\nh + \nv + \nh*\nv}$ depending on how we encode "on" and "off" states in the nodes.
 To explore the effect of this encoding on the potential for apparent near-degeneracy of an RBM, we can consider the ratio of the volume of the hypercube with corners at elements of $\mathcal{T}$ to the volume of the convex hull of $\mathcal{T}$ under both possible encodings (i.e., compare convex hull of $\mathcal{T}$ to either $[0,1]^3$ or $[-1,1]^3$ under the respective encodings).
@@
 For the small two node example, the $0/1$ encoding loses `r round((1-m1_bin$c_hull$vol)*100,2)`% of the volume of $[0,1]^3$ and the $-1/1$ encoding loses `r round((1-m1_neg$c_hull$vol/(2^3))*100,2)`% of the volume of $[-1,1]^3$.
-While there is clearly a one-to-one mapping of models for these two possible codings, this notion of lost volume can be helpful to conceptualize how difficult it will be for the mean vector $\boldsymbol \mu(\boldsymbol \theta)$ to avoid the boundary of $\mathcal{T}$, and thus avoid near-degeneracy.
+While there is clearly a one-to-one mapping of models for these two possible codings, this notion of lost volume can be helpful to conceptualize how difficult it will be for the mean vector $\boldsymbol \mu(\boldsymbol \theta)$ to avoid the boundary of $\mathcal{T}$, and thus avoid apparent near-degeneracy.
 In fact, if we look at the ratio of volume within the convex hull defined by $\mathcal{T}$ and the corresponding hypercube, we can see a relationship emerge as the number of nodes (and thus parameters) increases.
-In Figure \ref{fig:volume_plot}, it is evident that as the number of parameters increases, this ratio is decreasing at an increasing rate, meaning it gets more and more difficult to avoid the boundary of the convex hull and thus near degeneracy.
+In Figure \ref{fig:volume_plot}, it is evident that as the number of parameters increases, this ratio is decreasing at an increasing rate, meaning it gets more and more difficult to avoid the boundary of the convex hull and thus appearance of near degeneracy.
 Additionally, it appears that the $-1/1$ encoding suffers slightly less from this problem.
@@
 
-In addition to the argument that the $-1/1$ data encoding mitigates some perceived prevalence of near-degeneracy it also has the benefit of providing a guaranteed non-degenerate model at $\boldsymbol \theta = \boldsymbol 0 \in \mathbb{R}^{H + V + H*V}$, where the zero vector that serves as the natural center of the parameter space and induces model properties for the RBM (i.e., at $\boldsymbol \theta = \boldsymbol 0$, all variables are independent and visible variables are independent and uniformly distributed on $\{-1,1\}^V$.
+In addition to the argument that the $-1/1$ data encoding mitigates some perceived prevalence of near-degeneracy it also has the benefit of providing a guaranteed-to-be non-degenerate model at $\boldsymbol \theta = \boldsymbol 0 \in \mathbb{R}^{\nh + \nv + \nh*\nv}$, where the zero vector that serves as the natural center of the parameter space and induces the simplest possible model properties for the RBM (i.e., at $\boldsymbol \theta = \boldsymbol 0$, all variables are independent and visible variables are independent and uniformly distributed on $\{-1,1\}^\nv$).
 The proof of this is included in appendix \ref{center-of-universe}.
@@
   ungroup() %>% 
-  mutate(Hiddens = paste0("H = ", H), Visibles = paste0("V = ", V)) -> convex_hull_summary
+  mutate(Hiddens = paste0("paste(n[h], '= ", H, "')"), Visibles = paste0("paste(n[v], '= ", V, "')")) -> convex_hull_summary
 
@@
   summarise(max_abs_diff = mean(max_abs_diff)) %>%
-  mutate(Hiddens = paste0("H = ", H), Visibles = paste0("V = ", V)) -> exp_vals_summary
+  mutate(Hiddens = paste0("paste(n[h], '= ", H, "')"), Visibles = paste0("paste(n[v], '= ", V, "')")) -> exp_vals_summary
 
@@
 
-To explore the behavior of the RBM parameters $\boldsymbol \theta$ as it relates to near-degeneracy, instability, and uninterpretability.
-we consider models of small size.
-For $H, V \in \{1, \dots, 4\}$, we sample `r dim(res$samp[[1]])[1]` values of $\boldsymbol \theta$ with various magnitudes (details to follow).
-For each set of parameters we then calculate the three metrics of model impropriety introduced in Section \ref{degeneracy-instability-and-uninterpretability-oh-my}, $\boldsymbol \mu(\boldsymbol \theta)$, $R(\boldsymbol x)/V$, and the coordinates of $\left|\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] -   \text{E}\left[\boldsymbol X | \boldsymbol \theta^* \right] \right|$.
-In the case of near-degeneracy, we can go further and classify each model as near-degenerate or "viable" based on the distance of $\boldsymbol \mu(\boldsymbol \theta)$ from the boundary of the convex hull of $\mathcal{T}$ and look at the fraction of models that are "near-degenerate", meaning they are within a small distance to the boundary of the convex hull.
-
-In this numerical experiment, we split $\boldsymbol \theta$ into $\boldsymbol \theta_{interaction}$ and $\boldsymbol \theta_{main}$, in reference to which variables in the probability function the parameters correspond to (whether they multiply a $v_i$ or a $h_j$ or they multiply a $v_i h_j$), and allow the two types of terms to have varying average magnitudes, $||\boldsymbol \theta_{main} || /(H+V)$ and $||\boldsymbol \theta_{interaction} || /(H*V)$.
+To explore the effects of the RBM parameter vector $\boldsymbol \theta$ as it relates to near-degeneracy, instability, and uninterpretability, we consider models of small size.
+For $\nh, \nv \in \{1, \dots, 4\}$, we sample `r dim(res$samp[[1]])[1]` values of $\boldsymbol \theta$ with various magnitudes (details to follow).
+For each set of parameters we then calculate the three metrics of model impropriety introduced in Section \ref{degeneracy-instability-and-uninterpretability-oh-my}, $\boldsymbol \mu(\boldsymbol \theta)$, $\text{ELPR}(\boldsymbol \theta)/\nv$, and the coordinates of $\left|\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] -   \text{E}\left[\boldsymbol X | \boldsymbol \theta^* \right] \right|$.
+In the case of near-degeneracy, we can go further and classify each model as near-degenerate or "viable" based on the distance of $\boldsymbol \mu(\boldsymbol \theta)$ from the boundary of the convex hull of $\mathcal{T}$ and look at the fraction of models that are "near-degenerate," meaning they are within a small distance to the boundary of the convex hull.
+We define a "small" through a rough estimation to the volume of the hull for each model size.
+We pick $\epsilon_0 = 0.05$ for $\nh=\nv=1$ and then, for every other $\nh$ and $\nv$, set $m=\nh+\nv+\nv*\nh$ and pick $\epsilon$ so that $1-(1-2\epsilon_0)^3 = 1 - (1-2\epsilon)^m$.
+In this way, we roughly scale the volume of the small distance to the boundary of the convex hull to be equivalent in all dimensions.
+
+In this numerical experiment, we split $\boldsymbol \theta$ into $\boldsymbol \theta_{main}$ and $\boldsymbol \theta_{interaction}$, in reference to which variables in the probability function the parameters correspond to (whether they multiply a $v_i$ or a $h_j$ or they multiply a $v_i h_j$), and allow the two types of terms to have varying average magnitudes, $||\boldsymbol \theta_{main} || /(\nh+\nv)$ and $||\boldsymbol \theta_{interaction} || /(\nh*\nv)$.
 These average magnitudes vary on a grid between `r min(plot_dat_grid$r1/(plot_dat_grid$H + plot_dat_grid$V))` and `r max(plot_dat_grid$r1/(plot_dat_grid$H + plot_dat_grid$V))` with `r length(unique(plot_dat_grid$r1/(plot_dat_grid$H + plot_dat_grid$V)))` breaks, yielding `r length(unique(plot_dat_grid$r1/(plot_dat_grid$H + plot_dat_grid$V)))^2` grid points.
@@
 The results of this numerical study are summarized in Figures \ref{fig:degen_plots}, \ref{fig:instab_plots}, and \ref{fig:uninterp_plots}.
-From these three figures, it is clear to see that all three measures of model impropriety show higher values for larger magnitudes of the parameter vectors.
-Additionally, since there are $H*V$ interaction terms in $\boldsymbol \theta$ versus only $H + V$ main effects terms, for large models there are many more interaction parameters than main effects in the model.
+From these three figures, it is clear that all three measures of model impropriety show higher values for larger magnitudes of the parameter vectors.
+Additionally, since there are $\nh*\nv$ interaction terms in $\boldsymbol \theta$ versus only $\nh + \nv$ main effect terms, for large models there are many more interaction parameters than main effects in the model.
 And so, severely limiting the magnitude of the individual interactions may well help prevent model impropriety.
@@
   scale_fill_gradient("Fraction near-degenerate", low = "yellow", high = "red") +
-  facet_grid(Visibles~Hiddens) +
-  xlab(expression(paste(paste("||", theta[main], "||", "/(H + V)")))) +
-  ylab(expression(paste(paste("||", theta[interaction], "||", "/(H*V)")))) +
+  facet_grid(Visibles~Hiddens, labeller = label_parsed) +
+  xlab(expression(paste(paste("||", theta[main], "||", "/(", n[h]," + ", n[v], ")")))) +
+  ylab(expression(paste(paste("||", theta[interaction], "||", "/(", n[h], "*", n[v], ")")))) +
   theme(aspect.ratio = 1) -> p.degen
@@
   scale_fill_gradient("Mean max absolute difference", low = "yellow", high = "red", limits = c(0,2)) +
-  facet_grid(Visibles~Hiddens) +
-  xlab(expression(paste(paste("||", theta[main], "||", "/(H + V)")))) +
-  ylab(expression(paste(paste("||", theta[interaction], "||", "/(H*V)")))) +
+  facet_grid(Visibles~Hiddens, labeller = label_parsed) +
+  xlab(expression(paste(paste("||", theta[main], "||", "/(", n[h]," + ", n[v], ")")))) +
+  ylab(expression(paste(paste("||", theta[interaction], "||", "/(", n[h], "*", n[v], ")")))) +
   theme(aspect.ratio = 1) -> p.exp_diff
@@
   scale_fill_gradient(expression(frac(max(Q) - min(Q), H + V)), low = "yellow", high = "red") +
-  facet_grid(Visibles~Hiddens) +
-  xlab(expression(paste(paste("||", theta[main], "||", "/(H + V)")))) +
-  ylab(expression(paste(paste("||", theta[interaction], "||", "/(H*V)")))) +
+  facet_grid(Visibles~Hiddens, labeller = label_parsed) +
+  xlab(expression(paste(paste("||", theta[main], "||", "/(", n[h]," + ", n[v], ")")))) +
+  ylab(expression(paste(paste("||", theta[interaction], "||", "/(", n[h], "*", n[v], ")")))) +
   theme(aspect.ratio = 1) -> p.max_q
@@
   scale_fill_gradient(expression(frac(bar(ELPR(theta)), V)), low = "yellow", high = "red") +
-  facet_grid(Visibles~Hiddens) +
-  xlab(expression(paste(paste("||", theta[main], "||", "/(H + V)")))) +
-  ylab(expression(paste(paste("||", theta[interaction], "||", "/(H*V)")))) +
+  facet_grid(Visibles~Hiddens, labeller = label_parsed) +
+  xlab(expression(paste(paste("||", theta[main], "||", "/(", n[h]," + ", n[v], ")")))) +
+  ylab(expression(paste(paste("||", theta[interaction], "||", "/(", n[h], "*", n[v], ")")))) +
   theme(aspect.ratio = 1) -> p.elpr
@@
 
-```{r instab_plots_2, fig.cap="\\label{fig:instab_plots}Results from the experiment, here looking at the sample mean value of $R(\\boldsymbol \\theta)/V$ at each grid point for each combination of magnitude of $\\boldsymbol \\theta$ and model size.
+```{r instab_plots_2, fig.cap="\\label{fig:instab_plots}Results from the experiment, here looking at the sample mean value of $\\text{ELPR}(\\boldsymbol \\theta)/V$ at each grid point for each combination of magnitude of $\\boldsymbol \\theta$ and model size.
 As the magnitude of $\\boldsymbol \\theta$ grows, so does the value of this metric, indicating typical instability in the model.", fig.height=4}
@@
 
-Figure \ref{fig:split_plots_dist} shows the fraction near-degenerate for each magnitude of $\boldsymbol \theta$ for each model architecture.
-For each number of visibles in the model, the fraction near-degenerate becomes greater than zero at larger values of $||\boldsymbol \theta||$ as $H$ increases and the slope becomes steeper as $H$ increases also.
+Figure \ref{fig:split_plots_dist} shows the fraction of near-degenerate models for each magnitude of $\boldsymbol \theta$ for each model architecture.
+For each number of visibles in the model, the fraction near-degenerate becomes greater than zero at larger values of $||\boldsymbol \theta||$ as $\nh$ increases and the slope becomes steeper as $\nh$ increases also.
 This shows that as model size gets larger, the risk of degeneracy starts at a slightly larger magnitude of parameters, but very quickly increases until reaching close to 1.
 
-```{r split_plots2, warning=FALSE, fig.cap="\\label{fig:split_plots_dist}The fraction near-degenerate for each magnitude of $\\boldsymbol \\theta$.
-For each number of visibles in the model, the fraction near-degenerate becomes greater than zero at larger values of $||\\boldsymbol \\theta||$ as $H$ increases and the slope becomes steeper as $H$ increases also.", fig.height=5}
+```{r split_plots2, warning=FALSE, fig.cap="\\label{fig:split_plots_dist}The fraction of near-degenerate models for each magnitude of $\\boldsymbol \\theta$.
+For each number of visibles in the model, the fraction near-degenerate becomes greater than zero at larger values of $||\\boldsymbol \\theta||$ as $\\nh$ increases and the slope becomes steeper as $\\nh$ increases also.", fig.height=5}
 plot_dat_grid %>% 
@@
   summarise(frac_degen = sum(near_hull)/n()) %>%
-  mutate(Hiddens = as.character(H), Visibles = paste0("V = ", V)) %>%
+  mutate(Hiddens = as.character(H), Visibles = paste0("paste(n[v], '= ", V, "')")) %>%
   ggplot() +
@@
   #geom_smooth(aes(dist, frac_degen, colour = Hiddens, group = Hiddens)) +
-  facet_wrap(~Visibles) +
+  facet_wrap(~Visibles, labeller = label_parsed) +
   ylim(c(0,1)) +
-  xlab(expression(paste("||", theta, "||", "/(H + V + H*V), Average distance from the origin"))) +
+  xlab(expression(paste("||", theta, "||", "/(", n[h], " + ", n[v], " + ", n[h], "*", n[v], "), Average distance from the origin"))) +
   ylab("Fraction Near-degenerate") +
-  scale_colour_discrete("H")
+  scale_colour_discrete(expression(n[h]))
 ```
@@
 This, however, is not the only potential problem to be faced when using these models.
-There is the matter of pricipled/rigorous fitting of RBM models.
+There is the matter of principled/rigorous fitting of RBM models.
 
@@
 
-Typically, fitting a RBM via maximum likelihood (ML) methods will be infeasible mainly due to the intractibility of the normalizing term $\gamma(\boldsymbol \theta)$ in a model of any realistic size.
+Typically, fitting a RBM via maximum likelihood (ML) methods will be infeasible due mainly to the intractibility of the normalizing term $\gamma(\boldsymbol \theta)$ in a model of any realistic size.
 Ad hoc methods are used instead, which aim to avoid this problem by using stochastic ML that employ a small number of MCMC draws.
@@
 However, computational concerns are not the only issues with fitting an RBM using ML.
-In addition, the RBM model, with the appropriate choice of parameters, has the potential to re-create any distribution for the data (i.e., reproduce any specification of cell probabilities for the -1/1 valued data outcomes).
+In addition, the RBM model, with the appropriate choice of parameters, has the potential to re-create any distribution for the data (i.e., reproduce any specification of cell probabilities for the $-1/1$ valued data outcomes).
 To see this, we provide a small example.
@@
 
-The previous discussion intends to indicate that the RBM model suffers from parameter identifiability issues that go beyond mere symmetries in the parametrization.
+The previous discussion intends to indicate that the RBM model class suffers from parameter identifiability issues that go beyond mere symmetries in the parametrization.
 Quite different parameter settings can induce the same essential RBM model.
@@
 Hence, RBM model fitting based on ML will often simply seek to reproduce the empirical distribution (i.e., as an extreme type of over-fitting).
-What's more, whenever this empirical distribution contains empty cells, fitting steps for the RBM model will necessarily aim to chase parameters that necessarily diverge in magnitude in order to zero out the corresponding RBM cell probabilities.
-In a data application with a large sample space it is unlikely that the training set will include one of each possible image, thus there will be possible images that are not realized in the training set, unlike this small example.
+What's more, whenever this empirical distribution contains empty cells, fitting steps for the RBM model will necessarily aim to choose parameters that necessarily diverge in magnitude in order to zero out the corresponding RBM cell probabilities.
+In a data application with a large sample space, it is unlikely that the training set will include one of each possible outcome, unlike this small example.
 As we have already discussed in secion \ref{explorations}, large magnitudes of $\boldsymbol \theta$ lead to model impropriety in the RBM.
 
-Here we consider what might be done in a principled manner, testing on a $V = H = 4$ case that already stretched the limits of what is computable - in particular we consider Bayes methods.
+Here we consider what might be done in a principled manner, testing on a $\nv = \nh = 4$ case that already stretches the limits of what is computable - in particular we consider Bayes methods.
 
 ## Bayesian model fitting
-To avoid model impropriety for a fitted RBM, we want to avoid parts of the parameter space $\mathbb{R}^{V + H + VH}$ that lead to near-degeneracy, instability, and uninterpretability.
+To avoid model impropriety for a fitted RBM, we want to avoid parts of the parameter space $\mathbb{R}^{\nv + \nh + \nv*\nh}$ that lead to near-degeneracy, instability, and uninterpretability.
 Motivated by the insights in Section \ref{exploring-manageable-examples}, one idea is to shrink $\boldsymbol \theta$ toward $\boldsymbol 0$ by specifying priors that place low probability on large values of $||\boldsymbol \theta||$, specifically focusing on shrinking $\boldsymbol \theta_{interaction}$ more than $\boldsymbol \theta_{main}$.
@@
          `$\\theta_{44}$` = theta44) %>%
-  gather("Parameter", "Value") -> tbl
+  gather("Parameter", "Value") %>%
+  mutate(Value = paste0("$", Value, "$")) -> tbl
 
@@
   kable(caption = "\\label{tab:theta}Parameters used to fit a test case with $V = H = 4$.
-This parameter vector was chosen as a sampled value of $\\boldsymbol \\theta$ that was not near the convex hull of the sufficient statistics for a grid point in figure \\ref{fig:degen_plots} with $< 5$\\% near-degeneracy.")
-```
-
-We considered a test case with $V = H = 4$ and parameters given in Table \ref{tab:theta}.
+This parameter vector was chosen as a sampled value of $\\boldsymbol \\theta$ that was not near the convex hull of the sufficient statistics for a grid point in Figure \\ref{fig:degen_plots} with $< 5$\\% near-degeneracy.")
+```
+
+We considered a test case with $\nv = \nh = 4$ and parameters given in Table \ref{tab:theta}.
 This parameter vector was chosen as a sampled value of $\boldsymbol \theta$ that was not near the convex hull of the sufficient statistics for a grid point in figure \ref{fig:degen_plots} with $< 5$\% near-degeneracy.
-We simulated $n = 5,000$ as a training set and fit the RBM using three Bayes methodologies.
+We simulated $n = 5,000$ realizations as visibles as a training set and fit the RBM using three Bayes methodologies.
 These involved the following:
@@
 1.
-*A trick prior.* Here we cancel out normalizing term in the likelihood so that resulting full conditionals of $\boldsymbol \theta$ are multivariate Normal.
+*A "trick" prior.* Here we cancel out normalizing term in the likelihood so that resulting full conditionals of $\boldsymbol \theta$ are multivariate Normal.
 The $h_j$ are carried allong in the MCMC sampling from the posterior as latent variables.
@@
     \begin{align*}
-    \gamma(\boldsymbol \theta) = \sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V\sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right) \text{ and } C_{2} < C_{1}
+    \gamma(\boldsymbol \theta) = \sum\limits_{\boldsymbol x \in \mathcal{C}^{\nh + \nv}}\exp\left(\sum\limits_{i = 1}^\nv\sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right) \text{ and } C_{2} < C_{1}
     \end{align*}
@@
     \begin{align*}
-    g_{\boldsymbol \theta}(\boldsymbol v) \propto \sum\limits_{\boldsymbol h \in \{-1,1\}^H} \exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right), \boldsymbol v \in \{-1,1\}^V.
+    g_{\boldsymbol \theta}(\boldsymbol v) \propto \sum\limits_{\boldsymbol h \in \mathcal{C}^\nh} \exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right), \boldsymbol v \in \mathcal{C}^\nv.
     \end{align*}
 
-The three fitting methods are ordered by computational feasibility in a real-data situation, with the trick prior being the most feasible due to conjugacy and the marginalized likelihood method the least feasible due to the marginalization and need for an adaptive Metropolis Hastings step.
+The three fitting methods are ordered by computational feasibility in a real-data situation, with the trick prior being the most computationally feasible due to conjugacy and the marginalized likelihood method the least feasible due to the marginalization and need for an adaptive Metropolis Hastings step.
 All three methods require choosing the values of hyperparameters.
@@
 \hline \hline
-\multirow{2}{*}{Trick Prior} & $C_1$ & $\frac{C}{N}\frac{1}{H + V}$ \\
- & $C_2$ & $\frac{C}{N}\frac{1}{H*V}$ \\
+\multirow{2}{*}{Trick Prior} & $C_1$ & $\frac{C}{n}\frac{1}{\nh + \nv}$ \\
+ & $C_2$ & $\frac{C}{n}\frac{1}{\nh*\nv}$ \\
 \hline
-\multirow{2}{*}{Truncated Normal} & $\sigma^2_{main}$ & $\frac{1}{H + V}$ \\
- & $\sigma^2_{interaction}$ & $\frac{1}{H*V}$ \\
+\multirow{2}{*}{Truncated Normal} & $\sigma^2_{main}$ & $\frac{1}{\nh + \nv}$ \\
+ & $\sigma^2_{interaction}$ & $\frac{1}{\nh*\nv}$ \\
 \hline
-\multirow{2}{*}{Marginalized Likelihood} & $\sigma^2_{main}$ & $\frac{1}{H + V}$ \\
- & $\sigma^2_{interaction}$ & $\frac{1}{H*V}$ \\
+\multirow{2}{*}{Marginalized Likelihood} & $\sigma^2_{main}$ & $\frac{1}{\nh + \nv}$ \\
+ & $\sigma^2_{interaction}$ & $\frac{1}{\nh*\nv}$ \\
 \hline
@@
 
-It should be noted that the truncated Normal method (2) and the marginalized likelihood method (3) are drawing from the same stationary posterior distribution for images.
+It should be noted that the truncated Normal method (2) and the marginalized likelihood method (3) are drawing from the same stationary posterior distribution for vectors of visibles.
 The difference between these two methods is in how well the chains mix and how quickly they arrive at the target distribution.
 After a burn-in period of 50 iterations selected by inspecting the trace plots, we assess the issue of mixing in two ways.
-First, the autocorrelation functions (ACF) for each image are assessed at multiple lags for each method and plotted in figure \ref{fig:acf} with the truncated Normal method in black and the marginalized likelihood method in red.
+First, the autocorrelation functions (ACF) for each possible vector of visibles are assessed at multiple lags for each method and plotted in Figure \ref{fig:acf} with the truncated Normal method in black and the marginalized likelihood method in red.
 As expected, ACF corresponding to the method that marginalizes out the hidden variables from the likelihood decreases to zero at a much faster rate, indicating better mixing for the chain.
@@
 
-```{r acf, fig.cap="\\label{fig:acf}The autocorrelation functions (ACF) for each image assessed at multiple lags for each method with the truncated Normal method in black and the marginalized likelihood method in red.
+```{r acf, fig.cap="\\label{fig:acf}The autocorrelation functions (ACF) for all $2^4 - 16$ possibles vectors of visibles assessed at multiple lags for each method with the truncated Normal method in black and the marginalized likelihood method in red.
 As expected, ACF corresponding to the method that marginalizes out the hidden variables from the likelihood decreases to zero at a much faster rate, indicating better mixing for the chain."}
@@
 Secondly, we can assess the mixing of our chain using an idea of effective sample size.
-If the MCMC chain were truly iid draws from the target distribution, then the estimate of the posterior mean probability of image $i$, $\bar{p}^{(i)}$ would be distributed approximately Normal with mean the true posterior marginal probability of image $i$, $p^{(i)}$, and variance $\sigma^2_i/M$, where $\sigma^2_i$ is the true posterior variance of image $i$ and $M$ is the length of the chain.
+If the MCMC chain were truly iid draws from the target distribution, then the estimate of the posterior mean probability of visibles vector $i$, $\bar{p}^{(i)}$ would be distributed approximately Normal with mean the true posterior marginal probability of visibles vector $i$, $p^{(i)}$, and variance $\sigma^2_i/M$, where $\sigma^2_i$ is the true posterior variance of visibles vector $i$ and $M$ is the length of the chain.
 However, with the presence of correlation in our chain, the aysmptotic variance of $\bar{p}^{(i)}$ is instead approximately some $C_i/M$, where $C_i$ is some positive constant such that $C_i > \sigma^2_i$.
-We can use an overlapping block-means approach [@gelman2011inference] to get a crude estimate for $C_i$, $\hat{C}_i = b S^2(\overline{p_i})$ where $b$ is the block length and $S^2(\overline{p_i})$ denotes the sample variance of the blockmeans $\overline{p}_i$ for image $i$.
+We can use an overlapping block-means approach [@gelman2011inference] to get a crude estimate for $C_i$, $\hat{C}_i = b S^2(\overline{p_i})$ where $b$ is the block length and $S^2(\overline{p_i})$ denotes the sample variance of the blockmeans $\overline{p}_i$ for visibles vector $i$.
 We compare it to an estimate of $\sigma^2_i$ using sample variance of the raw chain, $\hat{\sigma}^2_i$.
@@
   kable(digits = 2, caption = paste0("\\label{tab:m_eff}The effective sample sizes for a chain of length $M = 1000$ of all $16$ images.
-The truncated likelihood would require at least $", round(min(M_eff_table[1, -1])/min(M_eff_table[2, -1]), 1), "$ times more iterations to be run of the MCMC chain in order to achieve the same amount of effective information about the posterior distribution."))
-```
-
-The effective sample sizes for a chain of length $M = 1000$ of all $16$ images are presented in table \ref{tab:m_eff}.
+The truncated likelihood would require at least $", round(min(M_eff_table[1, -1])/min(M_eff_table[2, -1]), 1), "$ times as many MCMC iterations to be run in order to achieve the same amount of effective information about the posterior distribution."))
+```
+
+The effective sample sizes for a chain of length $M = 1000$ of all $2^4 = 16$ possible vectors of visibles are presented in table \ref{tab:m_eff}.
 These range from  $`r round(min(M_eff_table[1, -1]), 2)`$ to $`r round(max(M_eff_table[1, -1]), 2)`$ for the marginalized likelihood method, while the truncated Normal method only yields between $`r round(min(M_eff_table[2, -1]), 2)`$ and $`r round(max(M_eff_table[2, -1]), 2)`$ effective draws.
 Thus, the truncated likelihood would require at least $`r round(min(M_eff_table[1, -1])/min(M_eff_table[2, -1]), 1)`$ times more iterations to be run of the MCMC chain in order to achieve the same amount of effective information about the posterior distribution.
-For this reason, in conjuction with the ACF results in Figure \ref{fig:acf}, the truncated Normal method is not an effective method for fitting the RBM.
+For this reason, consistent with the ACF results in Figure \ref{fig:acf}, the truncated Normal method does not seem to be an effective method for fitting the RBM.
 
@@
 From these posterior predictive checks, it is evident that the marginalized likelihood method produces the closest fit to the actual data.
-However, this method requires a marginalization step to obtain the probability function of visible observations alone, which is infeasible for a model with $H$ of any real size.
+However, this method requires a marginalization step to obtain the probability function of visible observations alone, which is infeasible for a model with $\nh$ of any real size.
 
@@
 RBM models are an interesting class of undirected graphical models that have shown to be useful for supervised learning tasks.
-However, when viewed as a statistical model, RBMs very easily fall prey to instances of model impropriety such as near-degeneracy, S-instability, and uninterpretability.
+However, when viewed as generative statistical models, RBMs very easily fall prey to instances of model impropriety such as near-degeneracy, S-instability, and uninterpretability.
 Additionally, these models are difficult to fit using a rigorous methodology due to size of the model space.
@@
 In this paper, we have presented three honest Bayes MCMC-based methods for fitting RBMs.
-Common practice is to use MCMC to overcome fitting complexities, however because of the size of the space to be spanned ($H$ hidden variables for each visible observation) convergence and mixing of these methods will be slow.
+Common practice is to use MCMC to overcome fitting complexities, however because of the size of the space to be spanned ($\nh$ hidden variables for each visible observation) convergence and mixing of these methods will be slow.
 Marginalization of the latent variables in the model can improve mixing, but is numerically intractible due to repeated calculation of the normalizing constant.
@@
 
-\subsection{Instability proofs}
-\label{instability-proofs}
-
-```{r child='appendix-instability.Rmd'}
-```
-
 \end{appendix}

### 2016-11-05 writing/appendix/appendix-center.Rmd
reorganize dir and try to soften language about empirical likelihood

--- 2016-10-01 writing/appendix-center.Rmd
@@
 
-1.
-The function $\bm{g}: \mathbb{R}^{m+n+mn} \rightarrow \mathcal{I}$ is one-to-one; see Theorem \ref{thm:one} below.
+\begin{itemize}
+\item The function $\bm{g}: \mathbb{R}^{m+n+mn} \rightarrow \mathcal{I}$ is one-to-one; see Theorem \ref{thm:one} below.
 Hence, these moments uniquely characterize the parameters in the RBM$(m,n,\bm{\theta})$ model.
-(Additionally, two RBM models have the same parameters if and only if the models have the same probability functions.)
-1.
-The image $\mathcal{I}$ is a symmetric set (i.e., $\bm{g}(\bm{\theta})=-\bm{g}(-\bm{\theta})$) and must be a connected (not necessarily convex) region in $\mathbb{R}^{m+n+mn}$, without any voids in it (by the continuity of $\bm{g}(\bm{\theta})$).
-1.
-The center of the image $\mathcal{I}$ (the zero vector in $\mathbb{R}^{m+n+mn}$) is given only by $\bm{\theta}=\bm{0}$ (the zero vector in the parameter space).
+\item The image $\mathcal{I}$ is a symmetric set (i.e., $\bm{g}(\bm{\theta})=-\bm{g}(-\bm{\theta})$) and must be a connected (not necessarily convex) region in $\mathbb{R}^{m+n+mn}$, without any voids in it (by the continuity of $\bm{g}(\bm{\theta})$).
+\item The center of the image $\mathcal{I}$ (the zero vector in $\mathbb{R}^{m+n+mn}$) is given only by $\bm{\theta}=\bm{0}$ (the zero vector in the parameter space).
 This corresponds to the case where all the random vectors $V_1,\ldots,V_m,H_1,\ldots,H_n$ are iid with a symmetric Bernoulli$(1/2)$ distribution; see Theorem \ref{thm:two}.
+\end{itemize}
 
@@
 Taking all $h_j=1$ or all $h_j=-1$ for $1 \leq j \leq n$ yields $\theta_{v_\ell} = \tilde{\theta}_{v_\ell}$.
-This result further implies $\tilde{\theta}_{\ell 1} -\theta{\ell 1} = -  \sum_{j=2}^n  h_j (  \tilde{\theta}_{\ell j} -\theta_{\ell j})$ for
+This result further implies $\tilde{\theta}_{\ell 1} -\theta_{\ell 1} = -  \sum_{j=2}^n  h_j (  \tilde{\theta}_{\ell j} -\theta_{\ell j})$ for
 any given $\{h_j\}_{j=2}^n \subset \{\pm 1\}$, so that $\theta_{\ell 1}=\tilde{\theta}_{\ell 1}$ must hold.
 Iterating this argument, sequentially for each $j=2,\ldots,m$, shows $\theta_{\ell j}=\tilde{\theta}_{\ell j}$ for any $1 \leq j \leq n$ in addition to $\theta_{v_\ell}=\tilde{\theta}_{v_\ell}$.
-As $\ell\in\{1,\ldots,m\}$ was arbitrary, we have $\theta_{v_i}=\tilde{\theta}_{v_i}$ and $\theta{i j}=\tilde{\theta}_{i j}$ for any $1 \leq i \leq m$, $1 \leq j \leq n$.
+As $\ell\in\{1,\ldots,m\}$ was arbitrary, we have $\theta_{v_i}=\tilde{\theta}_{v_i}$ and $\theta_{i j}=\tilde{\theta}_{i j}$ for any $1 \leq i \leq m$, $1 \leq j \leq n$.
 
@@
 $$
-fsor any $\{v_i\}_{i=1}^m \subset \{\pm 1\}$, so that $\theta_{h_k}  =  \tilde{\theta}_{h_k}$. As $k \in \{1,\ldots,n\}$ was arbitrary, we have now established claim 2 from claim 3.
+for any $\{v_i\}_{i=1}^m \subset \{\pm 1\}$, so that $\theta_{h_k}  =  \tilde{\theta}_{h_k}$. As $k \in \{1,\ldots,n\}$ was arbitrary, we have now established claim 2 from claim 3.
 
@@
 Hence, it suffices now to establish claim 3 from claim 1.
-This follows immediately from Lemma \ref{lemma:one} below and the fact that the variables $V_1,\ldots,V_m$ are independent given $H_1,\ldots,H_n$ in the RBM$(m,n,\bm{\theta})$ model (and likewise $H_1,\ldots,H_n$ are independent given $V_1,\ldots,V_m$).
+This follows immediately from Lemma \ref{lemma:one} below and the fact that the variables $V_1,\ldots,V_m$ are conditionally independent given $H_1,\ldots,H_n$ in the RBM$(m,n,\bm{\theta})$ model (and likewise $H_1,\ldots,H_n$ are independent given $V_1,\ldots,V_m$).
 
@@
 Let $P_1,P_2$ denote two probability distributions for a random vector $(\bm{V}_m,\bm{H}_n) = (V_1,\ldots,V_m, H_1,\ldots,H_n)$ supported on $\{a,b\}^{m+n}$, for some real constants $a\neq b \in\mathbb{R}$, such that $P_i( \bm{V}_m=\bm{v}_m, \bm{H}_n=\bm{h}_n)>0$, $i=1,2$, for any $\bm{v}_m \in \{a,b\}^{m}$, $\bm{h}_n \in \{a,b\}^{n}$.
-Suppose additionally that, under $P_1$ or $P_2$,  the variables $V_1,\ldots,V_m$ are independent given $\bm{H}_n=\bm{h}_n \in \{a,b\}^{n}$ and that the variables $H_1,\ldots,H_n$ are independent given $\bm{V}_m=\bm{v}_m \in \{a,b\}^{m}$.
+Suppose additionally that, under $P_1$ or $P_2$,  the variables $V_1,\ldots,V_m$ are conditionally independent given $\bm{H}_n=\bm{h}_n \in \{a,b\}^{n}$ and that the variables $H_1,\ldots,H_n$ are conditionally independent given $\bm{V}_m=\bm{v}_m \in \{a,b\}^{m}$.
 

### 2016-11-05 writing/draft.Rmd
reorganize dir and try to soften language about empirical likelihood

@@
 title: "On the propriety of restricted Boltzmann machines"
-author: "Andee Kaplan, Daniel Nordman, Stephen Vardeman"
-date: "September, 2016"
+author: "Andee Kaplan, Daniel Nordman, and Stephen Vardeman"
+date: "November, 2016"
 linestretch: 2 
@@
     includes:
-      in_header: header.tex
+      in_header: latex/header.tex
     number_sections: yes
@@
 
-Additionally, because RBM models are highly overparametrized, any fully principled fitting method based on the likelihood will simply reproduce the (discrete) empirical distribution of most realistic training sets.
-That is, there is no "smoothed distribution" achieved in fitting a RBM model with a rigorous likelihood-based method.
-We are therefore led to be skeptical that any model built using these structures (like a deep Boltzmann machine) can achieve prediction or inference in a principled way.
+For a RBM model with sufficient hidden variables, it has been shown that any distribution on the data can be approximated arbitrarily well [@le2008representational, @montufar2011refinements, and @montufar2011expressive].
+The empirical distribution is the best fitting model to the observed cell data and so the RBM will seek to achieve this when possible.
+Thus, any fully principled fitting method based on the likelihood of a RBM with enough hidden variables will simply reproduce the (discrete) empirical distribution of most realistic training sets.
+That is, there is no "smoothed distribution" achieved in fitting a RBM model of sufficient size with a rigorous likelihood-based method.
+We are therefore led to be skeptical that any model built using these structures (like a deep Boltzmann machine) can achieve prediction or inference in a principled way without intentially limiting the flexibility of the model.
 
@@
 However, computational concerns are not the only issues with fitting an RBM using ML.
-In addition, the RBM model, with the appropriate choice of parameters, has the potential to re-create any distribution for the data (i.e., reproduce any specification of cell probabilities for the $-1/1$ valued data outcomes).
+In addition, the RBM model, with the appropriate choice of parameters and number of hiddens, has the potential to re-create any distribution for the data (i.e., reproduce any specification of cell probabilities for the $-1/1$ valued data outcomes).
+@montufar2011refinements show that any distribution on $\{0, 1\}^{\nv}$ can be approximated arbitrarily well by a RBM with $2^{\nv}/2 - 1$ hidden units.
 To see this, we provide a small example.
@@
 In fact, when the cell probabilities (\ref{eqn:0}) are all strictly positive, parameters in the RBM model can be specified to reflect these probabilities exactly.
-And, when one or more of the cell probabilities (\ref{eqn:0}) are zero, the corresponding RBM  probabilities $P(V_1=v_1,V_2=_v2| \theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21})$ may never be identically zero (due to exponential terms in the model) but
- parameters can be still selected to make the appropriate RBM cell probabilities arbitrarily small.
+And, when one or more of the cell probabilities (\ref{eqn:0}) are zero, the corresponding RBM  probabilities $P(V_1=v_1,V_2=_v2| \theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21})$ may never be identically zero (due to exponential terms in the model) but parameters can be still selected to make the appropriate RBM cell probabilities arbitrarily small.
 
@@
 
-For instance, assuming the cell probabilities from (\ref{eqn:0}) are all positive, these probabilities
- can be exactly reproduced by choosing
+For instance, assuming the cell probabilities from (\ref{eqn:0}) are all positive, these probabilities can be exactly reproduced by choosing
  \begin{eqnarray*}
@@
 However, this is not the most disastrous implication of the RBM parameterization.
-A far worse consequence is that, when fitting the RBM model by likelihood-based methods, we already know what the answer must be before we begin: namely, such  fitting will simply aim to reproduce the empirical distribution from the data.
+A far worse consequence is that, when fitting the RBM model by likelihood-based methods, we already know what the answer must be before we begin: namely, such  fitting will simply aim to reproduce the empirical distribution from the data if sufficiently many hiddens are in the model.
 That is, based on a random sample of visible variables, the model for the cell probabilities that has the highest likelihood over  *all possible model classes* (i.e., RBM-based or not) is the empirical distribution, and the parametrization of the RBM model itself ensures that this empirical distribution can be arbitrarily well approximated.
@@
 
-```{r child='appendix-center.Rmd'}
+```{r child='appendix/appendix-center.Rmd'}
 ```

### 2016-11-11 writing/draft.Rmd
abstrand and intro edits

@@
 By treating a hidden layer of one RBM as the visible layer in a second RBM, a deep architecture can be created.
-RBMs are thereby thought to have the ability to encode very complex and rich structures in data, making them attractive for supervised learning.
-However, the generative behavior of RBMs has been largely unexplored.
-In this paper, we discuss the relationship between RBM parameter specification in the binary case and the prevalence of undesirable model behaviors such as degeneracy, instability and/or uninterpretability.
+RBMs are thought to thereby have the ability to encode very complex and rich structures in data, making them attractive for supervised learning.
+However, the generative behavior of RBMs is largely unexplored.
+In this paper, we discuss the relationship between RBM parameter specification in the binary case and the prevalence of undesirable model properties such as degeneracy, instability and uninterpretability.
 We also describe the difficulties that arise in likelihood-based and Bayes fitting  of such (highly flexible) models, especially as Gibbs sampling (quasi-Bayes) methods are often advocated for the RBM model structure.
@@
 
-The data mining and machine learning communities have recently shown a great interest in deep learning, specifically in stacked restricted Boltzmann machines (RBMs) [see @salakhutdinov2009deep, @salakhutdinov2012efficient, @srivastava2013modeling, @le2008representational for examples].
-A RBM is an undirected probabilistic graphical model (for discrete or continuous random variables) with two layers, one hidden and one visible, with no conditional dependency within a layer [@smolensky1986information].
+The data mining and machine learning communities have recently shown great interest in deep learning, specifically in stacked restricted Boltzmann machines (RBMs) [see @salakhutdinov2009deep, @salakhutdinov2012efficient, @srivastava2013modeling, @le2008representational for examples].
+A RBM is a probabilistic undirected graphical model (for discrete or continuous random variables) with two layers, one hidden and one visible, with no conditional dependency within a layer [@smolensky1986information].
 These models have reportedly been used with success in classification of images [@larochelle2008classification, @srivastava2012multimodal].
 However, the model properties are largely unexplored in the literature and the commonly cited fitting methodology remains heuristic-based and abstruse [@hinton2006fast].
-In this paper, we attempt to provide steps toward a thorough understanding of the model and its behavior from the perspective of statistical theory, and then explore the possibility of a rigorous fitting methodology.
+In this paper, we provide steps toward a thorough understanding of the model class and its properties from the perspective of statistical theory, and then explore the possibility of a rigorous fitting methodology.
 We find the RBM model class to be deficient in two fundamental ways.
 
-First, the models are often unsatisfactory as a conceptualization of how data are generated.
+First, the models are often unsatisfactory as conceptualizations of how data are generated.
 According to Fisher, the aim of a statistical model is to represent data in a compact way [@fisher1922mathematical].
 Neyman and Box agree with this assessment and go further to state that a model should "provide an explanation of the mechanism underlying the observed phenomena" [@lehmann1990model, @box1967discrimination].
-At issue, RBMs  often produce data lacking realistic variability and may thereby fail to provide a satisfactory conceptualization of the data generation process.
+At issue, RBMs  often produce data lacking realistic variability and may thereby fail to provide a satisfactory conceptualization of a data generation process.
 In addition to such degeneracy, we find that RBMs can easily exhibit types of instability.
@@
 Such model impropriety issues have recently been documented in RBMs [@li2014biclustering], as well as other deep architectures [@szegedy2013intriguing, @nguyen2014deep].
-We investigate and these phenomena for RBMs in Section \ref{explorations} through simulations of small, manageable examples.
+We investigate these phenomena for RBMs in Section \ref{explorations} through simulations of small, manageable examples.
 
@@
 
-In Section \ref{bayesian-model-fitting}, we compare three Bayesian modeling techniques (involving MCMC approximations that are computationally more accessible than direct maximum likelihood), each proposed with the goal of avoiding parts of the RBM parameter space that yield unattractive models.
+In Section \ref{bayesian-model-fitting}, we compare three Bayesian techniques (involving MCMC approximations that are computationally more accessible than direct maximum likelihood), each proposed with the goal of avoiding parts of a RBM parameter space that yield unattractive models.
 As might be expected, with greater computational complexity comes an increase in fitting accuracy, but at the cost of practical feasibility.
 
-For a RBM model with sufficient hidden variables, it has been shown that any distribution on the data can be approximated arbitrarily well [@le2008representational, @montufar2011refinements, and @montufar2011expressive].
-The empirical distribution is the best fitting model to the observed cell data and so the RBM will seek to achieve this when possible.
-Thus, any fully principled fitting method based on the likelihood of a RBM with enough hidden variables will simply reproduce the (discrete) empirical distribution of most realistic training sets.
-That is, there is no "smoothed distribution" achieved in fitting a RBM model of sufficient size with a rigorous likelihood-based method.
-We are therefore led to be skeptical that any model built using these structures (like a deep Boltzmann machine) can achieve prediction or inference in a principled way without intentially limiting the flexibility of the model.
+For a RBM model with enough hidden variables, it has been shown that any distribution for the visibles can be approximated arbitrarily well [@le2008representational, @montufar2011refinements, and @montufar2011expressive].
+The empirical distribution of a training set of vectors of visibles is the best fitting model for observed cell data and so principled RBM fitting will seek to produce this when possible.
+That is, any fully principled fitting method based on the likelihood for RBM with enough hidden variables will simply reproduce the (discrete) empirical distribution of a training set.
+Thus, there is no "smoothed distribution" achieved in fitting a RBM model of sufficient size with a rigorous likelihood-based method.
+We are therefore led to be skeptical that any model built using these structures (like a deep Boltzmann machine) can achieve prediction or inference in a principled way without intentially limiting the flexibility of the fitted model.
 
@@
 Section \ref{background} formally defines the RBM including the joint distribution of hidden and visible variables and explains the model's connection to deep learning.
-Additionally, measures of model impropriety and methods of detecting it are defined.
+Additionally, measures of model impropriety and methods of quantifying/detecting it are defined.
 Section \ref{explorations} details our explorations into the model behavior and propriety (or lack thereof) of the RBM class.
@@
 @handcock2003assessing classify an exponentially parametrized random graph model at $\boldsymbol \theta$ as *near-degenerate* if the mean value of sufficient statistics under $\boldsymbol \theta$, $\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta}\boldsymbol t( \boldsymbol X)$, is close to the boundary of $C$.
-
-The RBM class of models can be thought to exhibit an analogous form of *near-degeneracy* as follows.
-When random variables $\boldsymbol t(\boldsymbol x) = (v_1, \dots, v_\nv, h_1, \dots, h_\nh, v_1 h_1, \dots, v_V h_\nh ) \in \mathcal{T} \equiv \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{C}^{\nh + \nv} \}$ in the neg-potential function $Q_{\boldsymbol \theta}(\cdot)$ have a mean vector $\boldsymbol \mu(\boldsymbol \theta)$ close to the boundary of the convex hull of $\mathcal{T}$, then the RBM model can be said to be near-degenerate at $\boldsymbol \theta \in \mathbb{R}^{\nv + \nh + \nh*\nv}$.
+Conceptually, this makes sense because if a model is near-degenerate, then there are only a small number of elements of the sample space $\mathcal{X}$ that have positive probability, meaning the the expected value $\text{E}_{\boldsymbol \theta}\boldsymbol t( \boldsymbol X)$ is only averaging that same number of values of $t( \boldsymbol x)$.
+Thus, this average value will not be pulled away from the boundary of $C$.
+
+The RBM class of models can be thought to exhibit an analogous form of *near-degeneracy* when there is a disproportionate amount of probability placed on a small number of elements in the sample space, $\mathcal{C}^{\nv + \nh}$.
+Using the method of @handcock2003assessing, when random variables $\boldsymbol t(\boldsymbol x) = (v_1, \dots, v_\nv, h_1, \dots, h_\nh, v_1 h_1, \dots, v_V h_\nh ) \in \mathcal{T} \equiv \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{C}^{\nh + \nv} \}$ in the neg-potential function $Q_{\boldsymbol \theta}(\cdot)$ have a mean vector $\boldsymbol \mu(\boldsymbol \theta)$ close to the boundary of the convex hull of $\mathcal{T}$, then the RBM model can be said to exhibit near-degeneracy at $\boldsymbol \theta \in \mathbb{R}^{\nv + \nh + \nh*\nv}$.
 Here the mean of $\boldsymbol t(\boldsymbol x)$ is
@@
 
-Additionally, S-unstable RBM model sequences are connected to degenerate model sequences placing all probability on a small portion of their sample spaces.
+Additionally, S-unstable RBM model sequences are connected to near-degeneracy of section \ref{near-degeneracy}, in which model sequences place all probability on a small portion of their sample spaces.
 To see this, define a modal set of possible outcomes (i.e.
@@
 
+three_ways$Visibles <- factor(three_ways$Visibles, levels=rev(unique(three_ways$Visibles)))
+
 ```
@@
   theme(aspect.ratio = 1) -> p.exp_diff
-
-three_ways %>%
-  mutate(r1 = round(r1/(H + V), 8), r2 = round(r2/(H * V), 8)) %>% 
-  ggplot() +
-  geom_tile(aes(x = r1, y = r2, fill = LHS1)) +
-  geom_contour(aes(x = r1, y = r2, z = LHS1), colour = "black", bins = 8) +
-  #geom_contour(aes(x = r1, y = r2, z = mean_max_q), colour = "black", breaks = .05, size = 1.5) +
-  geom_abline(aes(intercept = 0, slope = 1), alpha = .5, lty = 2) +
-  scale_fill_gradient(expression(frac(max(Q) - min(Q), H + V)), low = "yellow", high = "red") +
-  facet_grid(Visibles~Hiddens, labeller = label_parsed) +
-  xlab(expression(paste(paste("||", theta[main], "||", "/(", n[h]," + ", n[v], ")")))) +
-  ylab(expression(paste(paste("||", theta[interaction], "||", "/(", n[h], "*", n[v], ")")))) +
-  theme(aspect.ratio = 1) -> p.max_q
 
@@
 
-For simplicity, consider a model with two visible variables $(V_1,V_2)$ and one hidden $H_1$.  In this case,
-there are four possible data realizations for $(V_1,V_2)$  given by $(\pm 1, \pm 1)$ and we may express the model probabilities
+For simplicity, consider a model with two visible variables $(V_1,V_2)$ and one hidden $H_1$.
+In this case, there are four possible data realizations for $(V_1,V_2)$  given by $(\pm 1, \pm 1)$ and we may express the model probabilities
 as
@@
 Motivated by the insights in Section \ref{exploring-manageable-examples}, one idea is to shrink $\boldsymbol \theta$ toward $\boldsymbol 0$ by specifying priors that place low probability on large values of $||\boldsymbol \theta||$, specifically focusing on shrinking $\boldsymbol \theta_{interaction}$ more than $\boldsymbol \theta_{main}$.
+This is similar to an idea advocated by @hinton2010practical called *weight decay*, in which a penalty is added to the interaction terms in the model, $\boldsymbol \theta_{interaction}$, shrinking their magnitudes.
 

### 2016-11-11 writing/draft.Rmd
section 2

@@
 
-A restricted Boltzmann machine (RBM) is an undirected graphical model specified for discrete or continuous random variables, binary variables being most common.
+A restricted Boltzmann machine (RBM) is an undirected graphical model specified for discrete or continuous random variables, binary variables being the case most commonly considered.
 In this paper, we consider the binary case of a RBM for concreteness.
@@
   \caption{An example restricted Boltzmann machine (RBM), consisting of two layers, one hidden ($\mathcal{H}$) and one visible ($\mathcal{V}$), with no connections within a layer.
-Hidden nodes are indicated by gray circles and the visible nodes indicated by white circles.}
+Hidden nodes are indicated by gray filled circles and the visible nodes indicated by unfilled circles.}
   \label{fig:rbm}
@@
 
-A common use for RBMs is to create features for use in classification.
-For example, binary images could be classified where the pixel values would be treated as the visible variables $v_i$ [@hinton2006fast].
+A common use for RBMs is to create features for use in classification. For example, binary images can be classified where the pixel values are treated as the visible variables $v_i$ [@hinton2006fast].
 
@@
 Let $\boldsymbol x = (h_1, \dots, h_\nh, v_1,\dots,v_\nv)$ represent the states of the visible and hidden nodes in an RBM for some integers $\nv, \nh \ge 1$.
-Each single "binary" random variable, visible or hidden, will take its values in a common coding set $\mathcal{C}$, where we allow one of two possibilities for the coding set $\mathcal{C}=\{0,1\}$ or $\mathcal{C}=\{-1,1\}$ with "$1$" always indicating the "high" value of the variable.
-While $\mathcal{C}=\{0,1\}$ may be a natural or naive starting point, we show later that the coding $\mathcal{C}=\{-1,1\}$ induces better model properties for the RBM.
-We posit a parametric form for probabilities corresponding to a potential vector of states, $X = (H_1, \dots, H_\nh, V_1,\dots,V_\nv)$, for the nodes as
+Each single binary random variable, visible or hidden, will take its values in a common coding set $\mathcal{C}$, where we allow one of two possibilities for the coding set, $\mathcal{C}=\{0,1\}$ or $\mathcal{C}=\{-1,1\}$, with "$1$" always indicating the "high" value of the variable.
+While $\mathcal{C}=\{0,1\}$ may be a natural or naive starting point, we argue later that the coding $\mathcal{C}=\{-1,1\}$ induces more interpretable model properties for the RBM.
+A standard parametric form for probabilities corresponding to a potential vector of states, $X = (H_1, \dots, H_\nh, V_1,\dots,V_\nv)$, for the nodes is
 
@@
 \begin{align}
-\boldsymbol t(\boldsymbol x) = (h_1, \dots, h_\nh, v_1, \dots, v_\nv, v_1 h_1, \dots, v_\nv h_\nh) \in \mathcal{C}^{\nh + \nv + \nh*\nv}, \boldsymbol x = (h_1, \dots, h_\nh, v_1, \dots, v_\nh) \in \mathcal{C}^{\nv + \nh}, \label{eqn:t}
+\boldsymbol x &= (h_1, \dots, h_\nh, v_1, \dots, v_\nh) \in \mathcal{C}^{\nv + \nh} \text{ and} \\
+\boldsymbol t(\boldsymbol x) &= (h_1, \dots, h_\nh, v_1, \dots, v_\nv, v_1 h_1, \dots, v_\nv h_\nh) \in \mathcal{C}^{\nh + \nv + \nh*\nv}, \label{eqn:t}
 \end{align}
-let $\mathcal{T} = \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{C}^{\nv + \nh}\}$ be the vector of variables needed to compute probabilities (\ref{eqn:pmf}) in the model and write $Q_{\boldsymbol \theta}(\boldsymbol x) = \sum\limits_{i = 1}^\nh\sum\limits_{j=1}^\nv \theta_{ij} h_i v_j + \sum\limits_{i = 1}^\nh\theta_{hi} h_i + \sum\limits_{j = 1}^\nv\theta_{vj} v_j$ for the neg-potential function.
-Note that the RBM model is parameterized by $\boldsymbol \theta$ containing two types of parameters, main effects and interaction effects.
+let $\mathcal{T} = \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{C}^{\nv + \nh}\} \subset \mathbb{R}^{\nv + \nh + \nv * \nh}$ be the set of possible values for the vector of variables needed to compute probabilities (\ref{eqn:pmf}) in the model and write $Q_{\boldsymbol \theta}(\boldsymbol x) = \sum\limits_{i = 1}^\nh\sum\limits_{j=1}^\nv \theta_{ij} h_i v_j + \sum\limits_{i = 1}^\nh\theta_{hi} h_i + \sum\limits_{j = 1}^\nv\theta_{vj} v_j$ for the "neg-potential" function.
+The RBM model is parameterized by $\boldsymbol \theta$ containing two types of parameters, main effects and interaction effects.
 The main effects parameters ($\{\theta_{v_i}, \theta_{h_j}\}_{\substack{i = 1, \dots, \nv,\\j = 1, \dots, \nh}}$) weight the values of the visible $v_i$ and hidden $h_j$ nodes in probabilities (\ref{eqn:pmf}) and the interaction effects parameters ($\theta_{ij}$) weight the values of the connections $v_i h_j$, or dependencies, between hidden and visible layers.
@@
 450].
-The stacking is achieved by treating a hidden layer of one RBM as the visible layer in a second RBM, and so on until the desired multi-layer architecture is created.
+The stacking is achieved by treating a hidden layer of one RBM as the visible layer in a second RBM, and so on, until the desired multi-layer architecture is created.
 
@@
 \end{align*}
-where $\boldsymbol \theta \in \Theta \subset \mathbb{R}^q$ is the model parameter, and $t: \mathcal{X} \rightarrow \mathbb{R}^q$ are statistics based on the adjacency matrix of a graph.
+where $\boldsymbol \theta \in \Theta \subset \mathbb{R}^q$ is the model parameter, and $\boldsymbol t: \mathcal{X} \rightarrow \mathbb{R}^q$ is a vector of statistics based on the adjacency matrix of a graph.
 Here, as earlier, $\gamma(\boldsymbol \theta) = \sum_{\boldsymbol x \in \mathcal{X}} \exp\left(\boldsymbol \theta^T \boldsymbol t(\boldsymbol x)\right)$ is the normalizing function.
-Let $C$ denote the convex hull spanned by the potential outcomes of sufficient statistics, $\{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{X}\}$.
-@handcock2003assessing classify an exponentially parametrized random graph model at $\boldsymbol \theta$ as *near-degenerate* if the mean value of sufficient statistics under $\boldsymbol \theta$, $\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta}\boldsymbol t( \boldsymbol X)$, is close to the boundary of $C$.
-Conceptually, this makes sense because if a model is near-degenerate, then there are only a small number of elements of the sample space $\mathcal{X}$ that have positive probability, meaning the the expected value $\text{E}_{\boldsymbol \theta}\boldsymbol t( \boldsymbol X)$ is only averaging that same number of values of $t( \boldsymbol x)$.
-Thus, this average value will not be pulled away from the boundary of $C$.
-
-The RBM class of models can be thought to exhibit an analogous form of *near-degeneracy* when there is a disproportionate amount of probability placed on a small number of elements in the sample space, $\mathcal{C}^{\nv + \nh}$.
-Using the method of @handcock2003assessing, when random variables $\boldsymbol t(\boldsymbol x) = (v_1, \dots, v_\nv, h_1, \dots, h_\nh, v_1 h_1, \dots, v_V h_\nh ) \in \mathcal{T} \equiv \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{C}^{\nh + \nv} \}$ in the neg-potential function $Q_{\boldsymbol \theta}(\cdot)$ have a mean vector $\boldsymbol \mu(\boldsymbol \theta)$ close to the boundary of the convex hull of $\mathcal{T}$, then the RBM model can be said to exhibit near-degeneracy at $\boldsymbol \theta \in \mathbb{R}^{\nv + \nh + \nh*\nv}$.
+Let $C$ denote the convex hull of the potential outcomes of sufficient statistics, $\{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{X}\}$.
+@handcock2003assessing classify an exponentially parametrized random graph model at $\boldsymbol \theta$ as *near-degenerate* if the mean value of the vector of sufficient statistics under $\boldsymbol \theta$, $\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta}\boldsymbol t( \boldsymbol X)$, is close to the boundary of $C$.
+This makes sense because if a model is near-degenerate, in the sense that only a small number of elements of the sample space $\mathcal{X}$ have positive probability, the expected value $\text{E}_{\boldsymbol \theta}\boldsymbol t( \boldsymbol X)$ is an average of that same small number of values of $t( \boldsymbol x)$ and can be expected to not be pulled deep into the interior of $C$.
+
+A RBM model can be thought to exhibit an analogous form of *near-degeneracy* when there is a disproportionate amount of probability placed on a small number of elements in the sample space of visibles and hiddens, $\mathcal{C}^{\nv + \nh}$.
+Using the idea of @handcock2003assessing, when the random vector $\boldsymbol t(\boldsymbol x) = (v_1, \dots, v_\nv, h_1, \dots, h_\nh, v_1 h_1, \dots, v_V h_\nh ) \in \mathcal{T} \equiv \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{C}^{\nh + \nv} \}$ in the neg-potential function $Q_{\boldsymbol \theta}(\cdot)$ has a mean vector $\boldsymbol \mu(\boldsymbol \theta) \in \mathbb{R}^{\nv + \nh + \nv * \nh}$ close to the boundary of the convex hull of $\mathcal{T}$, then the RBM model can be said to exhibit near-degeneracy at $\boldsymbol \theta \in \mathbb{R}^{\nv + \nh + \nh*\nv}$.
 Here the mean of $\boldsymbol t(\boldsymbol x)$ is
@@
 Considering a class of exponential families of distributions, @schweinberger2011instability introduced a concept of model defficiency related to *instability*.
-Instability can be roughly characterized as excessive sensitivity in the model, where small changes in the components of potential data outcomes, $\boldsymbol x$, may lead to substantial changes in the probability function $f_{\boldsymbol \theta}(\boldsymbol x)$.
-To quantify "instability" more rigorously (particularly beyond the definition given by @schweinberger2011instability) it is useful to consider how the RBM model might be expanded to incorporate more and more visibles.
-When increasing the size of a RBM model, it becomes necessary to grow the number of model parameters (and in this process one may also arbitrarily expand the number of hidden variables used).
-To this end, let $\boldsymbol \theta_{\nv} \equiv (\theta_{v_1}, \dots, \theta_{v_\nv}, \theta_{h_1}, \dots, \theta_{h_\nh}, \theta_{11}, \dots, \theta_{\nv \nh}), \nv \ge 1$, denote an element of a sequence of RBM parameters indexed by the number $V$ of visibles $(V_1, \dots, V_V)$ and define a (scaled) extremal log-probability ratio of the RBM model at $\boldsymbol \theta_{V}$ as
+Instability can be roughly thought of as excessive sensitivity in the model, where small changes in the components of potential data outcomes, $\boldsymbol x$, may lead to substantial changes in the probability function $f_{\boldsymbol \theta}(\boldsymbol x)$.
+To quantify "instability" more rigorously (particularly beyond the definition given by @schweinberger2011instability) it is useful to consider how RBM models might be expanded to incorporate more and more visibles.
+When increasing the size of RBM models, it becomes necessary to grow the number of model parameters (and in this process one may also arbitrarily expand the number of hidden variables used).
+To this end, let $\boldsymbol \theta_{\nv} \equiv (\theta_{v_1}, \dots, \theta_{v_\nv}, \theta_{h_1}, \dots, \theta_{h_\nh}, \theta_{11}, \dots, \theta_{\nv \nh}), \nv \ge 1$, denote an element of a sequence of RBM parameters indexed by the number $V$ of visibles $(V_1, \dots, V_V)$ and define a (scaled) extremal log-probability ratio of the RBM model at $\boldsymbol \theta_{\nv}$ as
 \begin{align}
@@
 \end{align}
-where $P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv) \propto \sum_{\{h_j\}\in \mathcal{C}}\exp\left(\sum_{i = 1}^\nv \theta_{v_i} v_i + \sum_{j = 1}^\nh \theta_{h_j} h_j + \sum_{i = 1}^\nv \sum_{j = 1}^\nh \theta_{ij} v_i h_j\right)$ is the RBM probability of observing outcome $(v_1, \dots, v_\nv)$ for the visible variables $(V_1, \dots, V_\nv)$ under parameters $\boldsymbol \theta_\nv$.
+where $P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv) \propto \sum_{\{h_j\}\in \mathcal{C}}\exp\left(\sum_{i = 1}^\nv \theta_{v_i} v_i + \sum_{j = 1}^\nh \theta_{h_j} h_j + \sum_{i = 1}^\nv \sum_{j = 1}^\nh \theta_{ij} v_i h_j\right)$ is the RBM probability of observing outcome $(v_1, \dots, v_\nv)$ for the visible variables $(V_1, \dots, V_\nv)$ under parameter vector $\boldsymbol \theta_\nv$.
 
 In formulating a RBM model for a potentially large number of visibles (i.e., as $\nv \rightarrow \infty$), we will say that the ratio (\ref{eqn:elpr}) needs to stay bounded for the sequence of RBM models to be stable.
-That is, we make the following convention
+That is, we make the following convention.
 
@@
 \begin{align*}
-\lim\limits_{\nv \rightarrow \infty} \frac{1}{\nv} \text{ELPR}(\boldsymbol \theta_\nv) \equiv \lim\limits_{\nv \rightarrow \infty} \frac{1}{\nv} \log \left[\frac{\max\limits_{(v_1, \dots, v_\nv) \in \mathcal{C}^\nv}P_{\boldsymbol \theta}(v_1, \dots, v_\nv)}{\min\limits_{(v_1, \dots, v_\nv) \in \mathcal{C}^\nv}P_{\boldsymbol \theta}(v_1, \dots, v_\nv)}\right] = \infty.
+\lim\limits_{\nv \rightarrow \infty} \frac{1}{\nv} \text{ELPR}(\boldsymbol \theta_\nv) = \infty.
 \end{align*}
@@
 $$
-denoting the biggest log-probability ratio for a one-pixel (one component) change in data outcomes (visibles) at a RBM parameter $\boldsymbol \theta_\nv$. We then have the following result.
+denoting the biggest log-probability ratio for a one component change in data outcomes (visibles) at a RBM parameter $\boldsymbol \theta_\nv$. We then have the following result.
 
@@
 
-In other words, if the probability ratio (\ref{eqn:elpr}) is too large, then a RBM model sequence will exhibit large probability shifts for very small changes in data configurations (i.e., instability).
-
-Additionally, S-unstable RBM model sequences are connected to near-degeneracy of section \ref{near-degeneracy}, in which model sequences place all probability on a small portion of their sample spaces.
+In other words, if the probability ratio (\ref{eqn:elpr}) is too large, then a RBM model sequence will exhibit large probability shifts for very small changes in data configurations (i.e., will exhibit instability).
+
+Additionally, S-unstable RBM model sequences are connected to the near-degeneracy of section \ref{near-degeneracy} (in which model sequences place all probability on a small portion of their sample spaces).
 To see this, define a modal set of possible outcomes (i.e.
@@
 ### Uninterpretability
-For spatial Markov models, @kaiser2007statistical defines a measure of model impropriety he calls *uninterpretability*, which is characterized by dependence parameters in a model being so extreme that marginal mean-structures fail to hold as anticipated by the model statement.
+For spatial Markov models, @kaiser2007statistical defines a measure of model impropriety he calls *uninterpretability*, which is characterized by dependence parameters in a model being so extreme that marginal mean-structures fail to hold as anticipated by a model statement.
 We adapt this notion to RBM models as follows.
-Note that, in a RBM, the parameters $\theta_{v_1}, \dots, \theta_{v_\nv}$ and $\theta_{h_1}, \dots, \theta_{h_\nh}$ are naturally associated with main effects of visible and hidden variables and can be interpreted as (logit functions of) means for variables $V_1, \dots, V_\nv, H_1, \dots, H_\nh$ in a model with no interaction parameters, $\theta_{ij} = 0, i = 1, \dots, \nv, j = 1, \dots, \nh$.
+Note that in a RBM, the parameters $\theta_{v_1}, \dots, \theta_{v_\nv}$ and $\theta_{h_1}, \dots, \theta_{h_\nh}$ are naturally associated with main effects of visible and hidden variables and can be interpreted as (logit functions of) means for variables $V_1, \dots, V_\nv, H_1, \dots, H_\nh$ in a model with no interaction parameters, $\theta_{ij} = 0, i = 1, \dots, \nv, j = 1, \dots, \nh$.
 Hence, these main effect parameters have a clear interpretation under an independence model (one with $\theta_{ij} = 0$) but this interpretation (as well as an understanding of the nature of dependence) can break down as interaction parameters increase in magnitude relative to the size of the main effects.
-In such case, the main effect parameters $\theta_{v_1}$ and $\theta_{h_j}$ are no longer interpretable as originally intended in the modes (statements of marginal means) and the dependence parameters are so large as to dominate the entire model probability structure (destroying the model interpretation of dependence as local conditional modifications of all overall marginal mean structure).
+In such cases, the main effect parameters $\theta_{v_1}$ and $\theta_{h_j}$ are no longer interpretable as originally intended in the models (statements of marginal means) and the dependence parameters are so large as to dominate the entire model probability structure (destroying the model interpretation of dependence as local conditional modifications of an overall marginal mean structure).
 To assess which parameter values $\boldsymbol \theta$ may cause difficulties in interpretation, we use the difference $\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] - \text{E}\left[\boldsymbol X | \boldsymbol \theta^* \right]$ between two model expectations: E$\left[\boldsymbol X | \boldsymbol \theta\right]$ at $\boldsymbol \theta$ and expectations E$\left[\boldsymbol X | \boldsymbol \theta^* \right]$ given independence where $\boldsymbol \theta^*$ matches $\boldsymbol \theta$ for all main effects but otherwise has $\theta_{ij} = 0$ for $i = 1, \dots, \nv, j = 1, \dots, \nh$.

### 2016-11-14 writing/appendix/appendix-center.Rmd
more vardeman edits

@@
 For integers $m,n \geq 1$, consider random vectors $\bm{V}_m \equiv(V_1,\ldots,V_m)$ and $\bm{H}_n\equiv (H_1,\ldots,H_n)$, where the random variables $V_i$ and $H_j$ assume values in $\{1,-1\}$, $i=1,\ldots,m$, $j=1,\ldots,n$.
-We suppose the vector $(\bm{V}_m,\bm{H}_n)$ follows a restricted Boltzmann machine (RBM) distribution
+We suppose the vector $(\bm{V}_m,\bm{H}_n)$ follows a restricted Boltzmann machine (RBM) distribution, i.e.
+has probability mass function
 \begin{align*}
@@
 
-Denote the random vector $\bm{V}_m*\bm{H}_n = (V_1 H_1,\ldots,V_1 H_n, V_2 H_1,\ldots,V_2 H_n,\ldots, V_m H_1,\ldots,V_m H_n)$.
-For the RBM$(m,n,\bm{\theta})$ model, we are interested in the function $\bm{g}: \mathbb{R}^{m+n+mn} \rightarrow   \mathcal{I} \subset \mathbb{R}^{m+n+mn}$ given by
+Use the notation $\bm{V}_m*\bm{H}_n = (V_1 H_1,\ldots,V_1 H_n, V_2 H_1,\ldots,V_2 H_n,\ldots, V_m H_1,\ldots,V_m H_n)$.
+For the RBM$(m,n,\bm{\theta})$ model, consider in the function $\bm{g}: \mathbb{R}^{m+n+mn} \rightarrow   \mathcal{I} \subset \mathbb{R}^{m+n+mn}$ given by
 \begin{align*}

### 2016-11-14 writing/draft.Rmd
more vardeman edits

@@
 ### Joint distribution
-Let $\boldsymbol x = (h_1, \dots, h_\nh, v_1,\dots,v_\nv)$ represent the states of the visible and hidden nodes in an RBM for some integers $\nv, \nh \ge 1$.
+Let $\boldsymbol x = (h_1, \dots, h_\nh, v_1,\dots,v_\nv)$ represent the states of the visible and hidden nodes in a RBM for some integers $\nv, \nh \ge 1$.
 Each single binary random variable, visible or hidden, will take its values in a common coding set $\mathcal{C}$, where we allow one of two possibilities for the coding set, $\mathcal{C}=\{0,1\}$ or $\mathcal{C}=\{-1,1\}$, with "$1$" always indicating the "high" value of the variable.
@@
 
-We next explore and numerically explain the relationship between values of $\boldsymbol \theta$ and the three measures of model impropriety, near-degeneracy, instability, and uninterpretability, for RBM models of varying size.
+We next explore and numerically explain the relationship between values of $\boldsymbol \theta$ and the three measures of model impropriety, near-degeneracy, instability, and uninterpretability, for RBM models of varying sizes.
 
@@
 For this small model, we are able to investigate the symptoms of model impropriety, beginning with near-degeneracy.
-To this end, recall that one characterization requires the convex hull of statistics
+To this end, recall that one characterization requires consideration of the convex hull of possible values of statistics $\boldsymbol t(\boldsymbol x)$,
 $$\mathcal{T} = \{\boldsymbol t(\boldsymbol x): \boldsymbol x = (v_1, h_1) \in \{0,1\}^2\} \equiv \{(v_1, h_1, v_1 h_1): v_1, h_1 \in \{0,1\}\}$$ 
 appearing in the RBM probabilities for this model.
-As this set is in three dimensions, we are able to explicitly illustrate the shape of boundary of the convex hull of $\mathcal{T}$ and explore the behavior of the mean parameterization of the parameters.
+As this set is in three dimensions, we are able to explicitly illustrate the shape of boundary of the convex hull of $\mathcal{T}$ and explore the behavior of the mean vector $\text{E}_{\boldsymbol \theta} \boldsymbol t(\boldsymbol x)$ as a function of the parameter vector $\boldsymbol \theta$.
 Figure \ref{fig:toyhull} shows the convex hull of our "statistic space," $\mathcal{T} \subset \{0,1\}^3$, for this toy problem from two perspectives (enclosed by the unit cube, representing the convex hull of $\{0,1\}^3$ space).
@@
   \end{minipage}
-  \caption{The convex hull of the "statistic space"" in three dimensions for the toy RBM with one visible and one hidden node.
+  \caption{The convex hull of the "statistic space" in three dimensions for the toy RBM with one visible and one hidden node.
 The convex hull of $\mathcal{T} = \{\boldsymbol t(\boldsymbol x) : \boldsymbol x \in \mathcal{C}^{\nv + \nh}\}$ does not fill the unit cube because of the relationship between the elements of $\mathcal{T}$.}
@@
 
-Additionally, we can compute the mean vector for $\boldsymbol t(\boldsymbol x)$ as a function of the model parameters as 
+We can compute the mean vector for $\boldsymbol t(\boldsymbol x)$ as a function of the model parameters as 
 \begin{align}
@@
 where $\gamma(\boldsymbol \theta) = \sum\limits_{h_1 = 0}^1\sum\limits_{v_1 = 0}^1 \exp(\theta_{11} h_1 v_1 + \theta_{h_1}h_1 + \theta_{v_1}v_1)$.
-The three parametric coordinate functions of $\boldsymbol \mu(\boldsymbol \theta)$ can be visualized in Figure \ref{fig:degen_toy}.
+The three parametric coordinate functions of $\boldsymbol \mu(\boldsymbol \theta)$ can be represented as in Figure \ref{fig:degen_toy}.
 (Contour plots for three coordinate functions are shown in columns for various values of $\theta_{11}$.) In examining these, we see that as values of coordinates of $\boldsymbol \theta$ grow larger in magnitude, at least one mean function for the entries of $\boldsymbol t(\boldsymbol x)$ is approaching a value 0 or 1, forcing $\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta} \boldsymbol t(\boldsymbol x)$ to be near to the boundary of the convex hull of $\mathcal{T}$, as a sign of model near-degeneracy.
@@
 
-Finally to consider the effect of $\boldsymbol \theta$ on potential model uninterpretability, we can look at the difference between model expectations, E$\left[\boldsymbol X | \boldsymbol \theta\right]$, and expectations given independence, E$\left[\boldsymbol X | \boldsymbol \theta^*\right ]$ for the tiny toy RBM model where $\boldsymbol X = (V_1, H_1, V_1 H_1)$.
-This is given by
+Finally to consider the effect of $\boldsymbol \theta$ on potential model uninterpretability, we can look at the absolute difference between model expectations, E$\left[\boldsymbol X | \boldsymbol \theta\right]$, and expectations given independence, E$\left[\boldsymbol X | \boldsymbol \theta^*\right ]$ for the tiny toy RBM model where $\boldsymbol X = (V_1, H_1, V_1 H_1)$.
+This difference is given by
 
@@
 \begin{matrix}
-\frac{\left|\exp\left(\theta_{11} + \theta_{v_1} + 2\theta_{h_1}\right) - \exp\left( \theta_{v_1} + 2\theta_{h_1}\right) \right| }{\left(\exp\left(\theta_{v_1}\right) + \exp\left(\theta_{h_1}\right) + \exp\left(\theta_{11} + \theta_{v_1} + \theta_{h_1}\right)\right)\left(\exp\left(\theta_{v_1}\right) + \exp\left(\theta_{h_1}\right) + \exp\left(+ \theta_{v_1} + \theta_{h_1}\right)\right)} \\
-\frac{\left|\exp\left(\theta_{11} + 2\theta_{v_1} + \theta_{h_1}\right) - \exp\left( 2\theta_{v_1} + \theta_{h_1}\right) \right| }{\left(\exp\left(\theta_{v_1}\right) + \exp\left(\theta_{h_1}\right) + \exp\left(\theta_{11} + \theta_{v_1} + \theta_{h_1}\right)\right)\left(\exp\left(\theta_{v_1}\right) + \exp\left(\theta_{h_1}\right) + \exp\left(+ \theta_{v_1} + \theta_{h_1}\right)\right)}
+\frac{\exp\left(\theta_{11} + \theta_{v_1} + 2\theta_{h_1}\right) - \exp\left( \theta_{v_1} + 2\theta_{h_1}\right) }{\left(\exp\left(\theta_{v_1}\right) + \exp\left(\theta_{h_1}\right) + \exp\left(\theta_{11} + \theta_{v_1} + \theta_{h_1}\right)\right)\left(\exp\left(\theta_{v_1}\right) + \exp\left(\theta_{h_1}\right) + \exp\left(+ \theta_{v_1} + \theta_{h_1}\right)\right)} \\
+\frac{\exp\left(\theta_{11} + 2\theta_{v_1} + \theta_{h_1}\right) - \exp\left( 2\theta_{v_1} + \theta_{h_1}\right)  }{\left(\exp\left(\theta_{v_1}\right) + \exp\left(\theta_{h_1}\right) + \exp\left(\theta_{11} + \theta_{v_1} + \theta_{h_1}\right)\right)\left(\exp\left(\theta_{v_1}\right) + \exp\left(\theta_{h_1}\right) + \exp\left(+ \theta_{v_1} + \theta_{h_1}\right)\right)}
 \end{matrix}\right].
@@
 
-```{r uninterp, fig.cap="\\label{fig:uninterp}The difference between coordinates of model expectations, E$\\left[\\boldsymbol X | \\boldsymbol \\theta\\right]$, and expectations given independence, E$\\left[\\boldsymbol X | \\boldsymbol \\theta^* \\right ]$ for a RBM with one visible and one hidden node.", fig.height = 4, fig.align='center', message=FALSE, cache=TRUE, warning=FALSE}
+```{r uninterp, fig.cap="\\label{fig:uninterp}The absolute difference between coordinates of model expectations, E$\\left[\\boldsymbol X | \\boldsymbol \\theta\\right]$, and expectations given independence, E$\\left[\\boldsymbol X | \\boldsymbol \\theta^* \\right ]$ for a RBM with one visible and one hidden node.", fig.height = 4, fig.align='center', message=FALSE, cache=TRUE, warning=FALSE}
 expand.grid(theta_h1 = seq(-5, 5, length.out = 21),
@@
 This will result in variables $\boldsymbol t(\boldsymbol X)$ from (\ref{eqn:t}) satisfying $\boldsymbol t(\boldsymbol x) \in \{0,1\}^{\nh + \nv + \nh*\nv}$ or $\boldsymbol t(\boldsymbol x) \in \{-1,1\}^{\nh + \nv + \nh*\nv}$ depending on how we encode "on" and "off" states in the nodes.
-To explore the effect of this encoding on the potential for apparent near-degeneracy of an RBM, we can consider the ratio of the volume of the hypercube with corners at elements of $\mathcal{T}$ to the volume of the convex hull of $\mathcal{T}$ under both possible encodings (i.e., compare convex hull of $\mathcal{T}$ to either $[0,1]^3$ or $[-1,1]^3$ under the respective encodings).
+To explore the effect of this encoding on the potential for apparent near-degeneracy of a RBM, we can consider the ratio of the volume of the hypercube with corners at elements of $\mathcal{T}$ to the volume of the convex hull of $\mathcal{T}$ under both possible encodings (i.e., compare convex hull of $\mathcal{T}$ to either $[0,1]^3$ or $[-1,1]^3$ under the respective encodings).
 
 For the small two node example, the $0/1$ encoding loses `r round((1-m1_bin$c_hull$vol)*100,2)`% of the volume of $[0,1]^3$ and the $-1/1$ encoding loses `r round((1-m1_neg$c_hull$vol/(2^3))*100,2)`% of the volume of $[-1,1]^3$.
-While there is clearly a one-to-one mapping of models for these two possible codings, this notion of lost volume can be helpful to conceptualize how difficult it will be for the mean vector $\boldsymbol \mu(\boldsymbol \theta)$ to avoid the boundary of $\mathcal{T}$, and thus avoid apparent near-degeneracy.
+While there is clearly a one-to-one mapping between models for these two possible codings, this notion of lost volume can be helpful to geometrically conceptualize how difficult it will be for the mean vector $\boldsymbol \mu(\boldsymbol \theta)$ to avoid the boundary of $\mathcal{T}$, and thus avoid apparent near-degeneracy.
 In fact, if we look at the ratio of volume within the convex hull defined by $\mathcal{T}$ and the corresponding hypercube, we can see a relationship emerge as the number of nodes (and thus parameters) increases.
@@
 In addition to the argument that the $-1/1$ data encoding mitigates some perceived prevalence of near-degeneracy it also has the benefit of providing a guaranteed-to-be non-degenerate model at $\boldsymbol \theta = \boldsymbol 0 \in \mathbb{R}^{\nh + \nv + \nh*\nv}$, where the zero vector that serves as the natural center of the parameter space and induces the simplest possible model properties for the RBM (i.e., at $\boldsymbol \theta = \boldsymbol 0$, all variables are independent and visible variables are independent and uniformly distributed on $\{-1,1\}^\nv$).
-The proof of this is included in appendix \ref{center-of-universe}.
+The proof of this and further exploration o the equivalence of the $\boldsymbol \theta$ parameterization of the RBM model class and parameterization by $\boldsymbol \mu(\boldsymbol \theta)$ is in appendix \ref{center-of-universe}.
 Hence while from some computing perspectives $0/1$ coding might seem most natural, the $-1/1$ coding is far more convenient and interpretable from the point of view of statistical modeling, where it makes parameters simply interpreted in terms of symmetrically defined main effects and interactions.
@@
 
-To explore the effects of the RBM parameter vector $\boldsymbol \theta$ as it relates to near-degeneracy, instability, and uninterpretability, we consider models of small size.
+To explore the impact of RBM parameter vector $\boldsymbol \theta$ magnitude on near-degeneracy, instability, and uninterpretability, we consider models of small size.
 For $\nh, \nv \in \{1, \dots, 4\}$, we sample `r dim(res$samp[[1]])[1]` values of $\boldsymbol \theta$ with various magnitudes (details to follow).
-For each set of parameters we then calculate the three metrics of model impropriety introduced in Section \ref{degeneracy-instability-and-uninterpretability-oh-my}, $\boldsymbol \mu(\boldsymbol \theta)$, $\text{ELPR}(\boldsymbol \theta)/\nv$, and the coordinates of $\left|\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] -   \text{E}\left[\boldsymbol X | \boldsymbol \theta^* \right] \right|$.
-In the case of near-degeneracy, we can go further and classify each model as near-degenerate or "viable" based on the distance of $\boldsymbol \mu(\boldsymbol \theta)$ from the boundary of the convex hull of $\mathcal{T}$ and look at the fraction of models that are "near-degenerate," meaning they are within a small distance to the boundary of the convex hull.
-We define a "small" through a rough estimation to the volume of the hull for each model size.
+For each set of parameters we then calculate metrics of model impropriety introduced in Section \ref{degeneracy-instability-and-uninterpretability-oh-my} based on $\boldsymbol \mu(\boldsymbol \theta)$, $\text{ELPR}(\boldsymbol \theta)/\nv$, and the coordinates of $\left|\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] -   \text{E}\left[\boldsymbol X | \boldsymbol \theta^* \right] \right|$.
+In the case of near-degeneracy, we classify each model as near-degenerate or "viable" based on the distance of $\boldsymbol \mu(\boldsymbol \theta)$ from the boundary of the convex hull of $\mathcal{T}$ and look at the fraction of models that are "near-degenerate," meaning they are within a small distance of the boundary of the convex hull.
+We define a "small" through a rough estimation of the volume of the hull for each model size.
 We pick $\epsilon_0 = 0.05$ for $\nh=\nv=1$ and then, for every other $\nh$ and $\nv$, set $m=\nh+\nv+\nv*\nh$ and pick $\epsilon$ so that $1-(1-2\epsilon_0)^3 = 1 - (1-2\epsilon)^m$.
-In this way, we roughly scale the volume of the small distance to the boundary of the convex hull to be equivalent in all dimensions.
-
-In this numerical experiment, we split $\boldsymbol \theta$ into $\boldsymbol \theta_{main}$ and $\boldsymbol \theta_{interaction}$, in reference to which variables in the probability function the parameters correspond to (whether they multiply a $v_i$ or a $h_j$ or they multiply a $v_i h_j$), and allow the two types of terms to have varying average magnitudes, $||\boldsymbol \theta_{main} || /(\nh+\nv)$ and $||\boldsymbol \theta_{interaction} || /(\nh*\nv)$.
+In this way, we roughly scale the volume of the "small distance" to the boundary of the convex hull to be equivalent across dimensions.
+
+In our numerical experiment, we split $\boldsymbol \theta$ into $\boldsymbol \theta_{main}$ and $\boldsymbol \theta_{interaction}$, in reference to which variables in the probability function the parameters correspond (whether they multiply a $v_i$ or a $h_j$ or they multiply a $v_i h_j$), and allow the two types of terms to have varying average magnitudes, $||\boldsymbol \theta_{main} || /(\nh+\nv)$ and $||\boldsymbol \theta_{interaction} || /(\nh*\nv)$.
 These average magnitudes vary on a grid between `r min(plot_dat_grid$r1/(plot_dat_grid$H + plot_dat_grid$V))` and `r max(plot_dat_grid$r1/(plot_dat_grid$H + plot_dat_grid$V))` with `r length(unique(plot_dat_grid$r1/(plot_dat_grid$H + plot_dat_grid$V)))` breaks, yielding `r length(unique(plot_dat_grid$r1/(plot_dat_grid$H + plot_dat_grid$V)))^2` grid points.
@@
 From these three figures, it is clear that all three measures of model impropriety show higher values for larger magnitudes of the parameter vectors.
-Additionally, since there are $\nh*\nv$ interaction terms in $\boldsymbol \theta$ versus only $\nh + \nv$ main effect terms, for large models there are many more interaction parameters than main effects in the model.
+Additionally, since there are $\nh*\nv$ interaction terms in $\boldsymbol \theta$ versus only $\nh + \nv$ main effect terms, for large models there are many more interaction parameters than main effects in the models.
 And so, severely limiting the magnitude of the individual interactions may well help prevent model impropriety.
@@
 ```{r degen_plots, fig.cap="\\label{fig:degen_plots}Results from the numerical experiment, here looking at the fraction of models that were near-degenerate for each combination of magnitude of $\\boldsymbol \\theta$ and model size.
-Black lines show the contour levels for fraction of near-degenerace, while the thick black line shows the level where the fraction of near-degenerate models is .05.", fig.height=4}
+Black lines show the contour levels for fraction of near-degeneracy, while the thick black line shows the level where the fraction of near-degenerate models is .05.", fig.height=4}
 three_ways %>%
@@
 
-However, computational concerns are not the only issues with fitting an RBM using ML.
-In addition, the RBM model, with the appropriate choice of parameters and number of hiddens, has the potential to re-create any distribution for the data (i.e., reproduce any specification of cell probabilities for the $-1/1$ valued data outcomes).
+However, computational concerns are not the only issues with fitting a RBM using ML.
+In addition, a RBM model, with the appropriate choice of parameters and number of hiddens, has the potential to re-create any distribution for the data (i.e., reproduce any specification of cell probabilities for the $-1/1$ valued data outcomes).
 @montufar2011refinements show that any distribution on $\{0, 1\}^{\nv}$ can be approximated arbitrarily well by a RBM with $2^{\nv}/2 - 1$ hidden units.
-To see this, we provide a small example.
+To illustrate, we provide a small example.
 
@@
 \end{equation} 
-for the outcomes $(\pm 1, \pm 1)$,  values for parameters $(\theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21})$ may be chosen to mimic such  cell probabilities with arbitrary closeness.
+for the outcomes $(\pm 1, \pm 1)$,  values for parameters $(\theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21})$ may be chosen to approximate such  cell probabilities with arbitrary closeness.
 In fact, when the cell probabilities (\ref{eqn:0}) are all strictly positive, parameters in the RBM model can be specified to reflect these probabilities exactly.
@@
 However, this is not the most disastrous implication of the RBM parameterization.
-A far worse consequence is that, when fitting the RBM model by likelihood-based methods, we already know what the answer must be before we begin: namely, such  fitting will simply aim to reproduce the empirical distribution from the data if sufficiently many hiddens are in the model.
-That is, based on a random sample of visible variables, the model for the cell probabilities that has the highest likelihood over  *all possible model classes* (i.e., RBM-based or not) is the empirical distribution, and the parametrization of the RBM model itself ensures that this empirical distribution can be arbitrarily well approximated.
+A far worse consequence is that, when fitting the RBM model by likelihood-based methods, we already know the nature of the answer before we begin: namely, such  fitting will simply aim to reproduce the empirical distribution from the training data if sufficiently many hiddens are in the model.
+That is, based on a random sample of vectors of visible variables, the model for the cell probabilities that has the highest likelihood over  *all possible model classes* (i.e., RBM-based or not) is the empirical distribution, and the parametrization of the RBM model itself ensures that this empirical distribution can be arbitrarily well-approximated.
 
@@
 Hence, RBM model fitting based on ML will often simply seek to reproduce the empirical distribution (i.e., as an extreme type of over-fitting).
-What's more, whenever this empirical distribution contains empty cells, fitting steps for the RBM model will necessarily aim to choose parameters that necessarily diverge in magnitude in order to zero out the corresponding RBM cell probabilities.
-In a data application with a large sample space, it is unlikely that the training set will include one of each possible outcome, unlike this small example.
-As we have already discussed in secion \ref{explorations}, large magnitudes of $\boldsymbol \theta$ lead to model impropriety in the RBM.
-
-Here we consider what might be done in a principled manner, testing on a $\nv = \nh = 4$ case that already stretches the limits of what is computable - in particular we consider Bayes methods.
+What's more, whenever this empirical distribution contains empty cells, fitting steps for the RBM model will necessarily aim to choose parameters that necessarily diverge to infinity in magnitude in order to zero-out the corresponding RBM cell probabilities.
+In data applications with a large sample space, it is unlikely that the training set will include at least one of each possible vector outcome (unlike this small example).
+As we have already discussed in Section \ref{explorations}, large magnitudes of $\boldsymbol \theta$ lead to model impropriety in the RBM.
+
+Here we consider what might be done in a principled manner to prevent both overfitting and model impropriety, testing on a $\nv = \nh = 4$ case that already stretches the limits of what is computable - in particular we consider Bayes methods.
 
@@
   kable(caption = "\\label{tab:theta}Parameters used to fit a test case with $V = H = 4$.
-This parameter vector was chosen as a sampled value of $\\boldsymbol \\theta$ that was not near the convex hull of the sufficient statistics for a grid point in Figure \\ref{fig:degen_plots} with $< 5$\\% near-degeneracy.")
+This parameter vector was chosen as a sampled value of $\\boldsymbol \\theta$ that was not near the convex hull of the sufficient statistics for a grid point in Figure \\ref{fig:degen_plots} with $< 5$\\% near-degeneracy.", digits = 10)
 ```
@@
 We considered a test case with $\nv = \nh = 4$ and parameters given in Table \ref{tab:theta}.
-This parameter vector was chosen as a sampled value of $\boldsymbol \theta$ that was not near the convex hull of the sufficient statistics for a grid point in figure \ref{fig:degen_plots} with $< 5$\% near-degeneracy.
+This parameter vector was chosen as a sampled value of $\boldsymbol \theta$ that was not near the convex hull of the sufficient statistics for a grid point in Figure \ref{fig:degen_plots} with $< 5$\% near-degeneracy.
 We simulated $n = 5,000$ realizations as visibles as a training set and fit the RBM using three Bayes methodologies.
@@
 Full conditional distributions are not conjugate, and simulation from the posterior was accomplished using a geometric adaptive Metropolis Hastings step [@zhou2014some] and calculation of likelihood normalizing constant.
-(This computation is barely feasible for a problem of this size.) Here the $h_j$ are carried along in the MCMC implementation as latent variables.
+(This computation is barely feasible for a problem of this size and would be infeasible for larger problems.) Here the $h_j$ are carried along in the MCMC implementation as latent variables.
 3.
@@
 After a burn-in period of 50 iterations selected by inspecting the trace plots, we assess the issue of mixing in two ways.
-First, the autocorrelation functions (ACF) for each possible vector of visibles are assessed at multiple lags for each method and plotted in Figure \ref{fig:acf} with the truncated Normal method in black and the marginalized likelihood method in red.
+First, the autocorrelation functions (ACF) for each possible vector of visibles are assessed and plotted in Figure \ref{fig:acf} with the truncated Normal method in black and the marginalized likelihood method in red.
 As expected, ACF corresponding to the method that marginalizes out the hidden variables from the likelihood decreases to zero at a much faster rate, indicating better mixing for the chain.
@@
 
-```{r acf, fig.cap="\\label{fig:acf}The autocorrelation functions (ACF) for all $2^4 - 16$ possibles vectors of visibles assessed at multiple lags for each method with the truncated Normal method in black and the marginalized likelihood method in red.
+```{r acf, fig.cap="\\label{fig:acf}The autocorrelation functions (ACF) for all $2^4 = 16$ possibles vectors of visibles assessed at multiple lags for each method with the truncated Normal method in black and the marginalized likelihood method in red.
 As expected, ACF corresponding to the method that marginalizes out the hidden variables from the likelihood decreases to zero at a much faster rate, indicating better mixing for the chain."}
@@
 
-Secondly, we can assess the mixing of our chain using an idea of effective sample size.
+Secondly, we can assess the mixing of our chains using an idea of effective sample size.
 If the MCMC chain were truly iid draws from the target distribution, then the estimate of the posterior mean probability of visibles vector $i$, $\bar{p}^{(i)}$ would be distributed approximately Normal with mean the true posterior marginal probability of visibles vector $i$, $p^{(i)}$, and variance $\sigma^2_i/M$, where $\sigma^2_i$ is the true posterior variance of visibles vector $i$ and $M$ is the length of the chain.
@@
   spread(Model, M_eff) %>%
-  mutate(Image = as.numeric(Image)) %>%
-  arrange(Image) -> M_eff_tmp
+  mutate(Vector = as.numeric(Image)) %>%
+  arrange(Vector) -> M_eff_tmp
 
 cbind(M_eff_tmp[1:8,], M_eff_tmp[9:16,]) %>%
-  kable(digits = 2, caption = paste0("\\label{tab:m_eff}The effective sample sizes for a chain of length $M = 1000$ of all $16$ images.
-The truncated likelihood would require at least $", round(min(M_eff_table[1, -1])/min(M_eff_table[2, -1]), 1), "$ times as many MCMC iterations to be run in order to achieve the same amount of effective information about the posterior distribution."))
-```
-
-The effective sample sizes for a chain of length $M = 1000$ of all $2^4 = 16$ possible vectors of visibles are presented in table \ref{tab:m_eff}.
+  kable(digits = 2, caption = paste0("\\label{tab:m_eff}The effective sample sizes for a chain of length $M = 1000$ of all $16$ possible vectors of visibles.
+The truncated likelihood would require at least $", round(min(M_eff_table[1, -1])/min(M_eff_table[2, -1]), 1), "$ times as many MCMC iterations to achieve the same amount of effective information about the posterior distribution."))
+```
+
+The effective sample sizes for a chain of length $M = 1000$ of all $2^4 = 16$ possible vectors of visibles are presented in Table \ref{tab:m_eff}.
 These range from  $`r round(min(M_eff_table[1, -1]), 2)`$ to $`r round(max(M_eff_table[1, -1]), 2)`$ for the marginalized likelihood method, while the truncated Normal method only yields between $`r round(min(M_eff_table[2, -1]), 2)`$ and $`r round(max(M_eff_table[2, -1]), 2)`$ effective draws.
 Thus, the truncated likelihood would require at least $`r round(min(M_eff_table[1, -1])/min(M_eff_table[2, -1]), 1)`$ times more iterations to be run of the MCMC chain in order to achieve the same amount of effective information about the posterior distribution.
-For this reason, consistent with the ACF results in Figure \ref{fig:acf}, the truncated Normal method does not seem to be an effective method for fitting the RBM.
+For this reason, consistent with the ACF results in Figure \ref{fig:acf}, the truncated Normal method does not seem to be an effective method for fitting the RBM of computing resources are at all limited.
 
 Figure \ref{fig:fitting_plot} shows the posterior probability of each possible $\boldsymbol v \in \{-1,1\}^4$ after fitting the RBM model in the two ways detailed in this section, excluding the truncated Normal method.
-The black vertical lines show the true probabilities of each image based on the parameters used to simulate the training set while the red vertical lines show the empirical distribution.
-From these posterior predictive checks, it is evident that the marginalized likelihood method produces the closest fit to the actual data.
+The black vertical lines show the true probabilities of each image based on the parameters used to simulate the training set while the red vertical lines show the empirical distribution for the training set of $5,000$ vectors.
+From these posterior predictive checks, it is evident that the marginalized likelihood method produces the best fit to the data.
 However, this method requires a marginalization step to obtain the probability function of visible observations alone, which is infeasible for a model with $\nh$ of any real size.
@@
   separate(statistic, into = c("junk", "statistic")) %>%
-  mutate(statistic = factor(statistic, labels = paste("Image", 1:length(unique(statistic))))) %>%
+  mutate(statistic = factor(statistic, labels = paste("Vector", 1:length(unique(statistic))))) %>%
   left_join(distn_emp %>% rename(prob_emp = prob) %>% right_join(distn_good %>% select(-image_id))) %>%
@@
   facet_wrap(~statistic, scales = "free_x") + 
-  ylab("Scaled Posterior Density") + xlab("Probability of Image") +
+  ylab("Scaled Posterior Density") + xlab("Probability of Vector of Visibles") +
   scale_colour_discrete("Method", labels = c("Marginalized likelihood", "Trick prior")) +
@@
 
-```{r fitting_plot, fig.cap="\\label{fig:fitting_plot}Posterior probability of each possible 4-pixel image using two of the three Bayesian fitting techniques, trick prior and marginalized likelihood.
-The black vertical lines show the true probabilities of each image based on the parameters used to simulate the images while the red vertical lines show the empirical distribution.
-The marginalized likelihood method has the best fit for the data, however is also the most computationally intensive and least feasible with a real dataset.", fig.height=6}
+```{r fitting_plot, fig.cap="\\label{fig:fitting_plot}Posterior probabilities of $16 = 2^4$ possible realizations of $4$ visibles using two of the three Bayesian fitting techniques, trick prior and marginalized likelihood.
+The black vertical lines show the true probabilities of each vector of visibles based on the parameters used to simulate the training data while the red vertical lines show the empirical distribution.
+The marginalized likelihood method produces the best fit for the data, however is also the most computationally intensive and least feasible with a real dataset.", fig.height=6}
 p.models
@@
 
-RBM models are an interesting class of undirected graphical models that have shown to be useful for supervised learning tasks.
-However, when viewed as generative statistical models, RBMs very easily fall prey to instances of model impropriety such as near-degeneracy, S-instability, and uninterpretability.
-Additionally, these models are difficult to fit using a rigorous methodology due to size of the model space.
+RBM models constitute an interesting class of undirected graphical models that are thought to be useful for supervised learning tasks.
+However, when viewed as generative statistical models, RBMs are prone to forms of model impropriety such as near-degeneracy, S-instability, and uninterpretability.
+Additionally, these models are difficult to fit using a rigorous methodology due dimension of the parameter space coupled with the size of the latent variable space.
 
 In this paper, we have presented three honest Bayes MCMC-based methods for fitting RBMs.
-Common practice is to use MCMC to overcome fitting complexities, however because of the size of the space to be spanned ($\nh$ hidden variables for each visible observation) convergence and mixing of these methods will be slow.
-Marginalization of the latent variables in the model can improve mixing, but is numerically intractible due to repeated calculation of the normalizing constant.
-
-Ultimately, it is not clear RBM models are useful as generative models.
-Due to the extreme flexibility in these models, a rigorous likelihood-based method for fitting an RBM will merely return the (discrete) empirical distribution for visibles.
-Meaning, even if a rigorous likelihood-based fitting method is able to overcome the inherent difficulties of mixing and convergence in a large space, we know the answer before we even begin.
-For these reasons, we are thus skeptical of the generative claims made about RBMs.
+Common practice is to use a kind of MCMC to overcome fitting complexities, however because of the size of the space to be filled with MCMC iterates convergence and mixing of these methods will be slow.
+Marginalization over the latent variables in the model can improve mixing, but is numerically intractible due to the necessity of repeated calculation of the normalizing constant.
+
+Ultimately, it is not clear that RBM models are useful as generative models.
+Due to the extreme flexibility in this model class, rigorous likelihood-based fitting for a RBM will typically merely return the (discrete) empirical distribution for visibles.
+Meaning, even if a rigorous likelihood-based fitting method is practically possible, we know what it will produce for fitted probabilities before we even begin.
+For these reasons, we are skeptical of the claims made about RBMs as generative.
 

### 2016-11-15 writing/draft.Rmd
rename methods

@@
     \end{align*}
-    This is the method of @li2014biclustering.
+    This is the method of @li2014biclustering. We will refer to this method as Bayes with Trick Prior and Latent Variables (BwTPLV).
 2.
@@
 (This computation is barely feasible for a problem of this size and would be infeasible for larger problems.) Here the $h_j$ are carried along in the MCMC implementation as latent variables.
+We will refer to this method as Bayes with Truncated Normal prior and Latent Variables (BwTNLV).
 3.
@@
     \end{align*}
-
-The three fitting methods are ordered by computational feasibility in a real-data situation, with the trick prior being the most computationally feasible due to conjugacy and the marginalized likelihood method the least feasible due to the marginalization and need for an adaptive Metropolis Hastings step.
+    We will refer to this method as Bayes with Truncated Normal prior and Marginalized Likelihood (BwTNML).
+
+The three fitting methods are ordered by computational feasibility in a real-data situation, with BwTPLV being the most computationally feasible due to conjugacy and BwTNML the least feasible due to the marginalization and need for an adaptive Metropolis Hastings step.
 All three methods require choosing the values of hyperparameters.
 In each case, we have chosen the values based on a rule of thumb that shrinks $\boldsymbol \theta_{interaction}$ more than $\boldsymbol \theta_{main}$.
-Additionally, the trick prior requires additional tuning to choose $C_1$ and $C_2$, reducing its appeal.
+Additionally, BwTPLV requires additional tuning to choose $C_1$ and $C_2$, reducing its appeal.
 The values used for the hyperparameters in our simulation are presented in Table \ref{tab:hyperparam}.
@@
 \hline \hline
-\multirow{2}{*}{Trick Prior} & $C_1$ & $\frac{C}{n}\frac{1}{\nh + \nv}$ \\
+\multirow{2}{*}{BwTPLV} & $C_1$ & $\frac{C}{n}\frac{1}{\nh + \nv}$ \\
  & $C_2$ & $\frac{C}{n}\frac{1}{\nh*\nv}$ \\
 \hline
-\multirow{2}{*}{Truncated Normal} & $\sigma^2_{main}$ & $\frac{1}{\nh + \nv}$ \\
+\multirow{2}{*}{BwTNLV} & $\sigma^2_{main}$ & $\frac{1}{\nh + \nv}$ \\
  & $\sigma^2_{interaction}$ & $\frac{1}{\nh*\nv}$ \\
 \hline
-\multirow{2}{*}{Marginalized Likelihood} & $\sigma^2_{main}$ & $\frac{1}{\nh + \nv}$ \\
+\multirow{2}{*}{BwTNML} & $\sigma^2_{main}$ & $\frac{1}{\nh + \nv}$ \\
  & $\sigma^2_{interaction}$ & $\frac{1}{\nh*\nv}$ \\
@@
 
-It should be noted that the truncated Normal method (2) and the marginalized likelihood method (3) are drawing from the same stationary posterior distribution for vectors of visibles.
+It should be noted that BwTNLV (2) and BwTNML (3) are drawing from the same stationary posterior distribution for vectors of visibles.
 The difference between these two methods is in how well the chains mix and how quickly they arrive at the target distribution.
 After a burn-in period of 50 iterations selected by inspecting the trace plots, we assess the issue of mixing in two ways.
-First, the autocorrelation functions (ACF) for each possible vector of visibles are assessed and plotted in Figure \ref{fig:acf} with the truncated Normal method in black and the marginalized likelihood method in red.
+First, the autocorrelation functions (ACF) for each possible vector of visibles are assessed and plotted in Figure \ref{fig:acf} with BwTNLV in black and BwTNML in red.
 As expected, ACF corresponding to the method that marginalizes out the hidden variables from the likelihood decreases to zero at a much faster rate, indicating better mixing for the chain.
@@
 
-```{r acf, fig.cap="\\label{fig:acf}The autocorrelation functions (ACF) for all $2^4 = 16$ possibles vectors of visibles assessed at multiple lags for each method with the truncated Normal method in black and the marginalized likelihood method in red.
+```{r acf, fig.cap="\\label{fig:acf}The autocorrelation functions (ACF) for all $2^4 = 16$ possibles vectors of visibles assessed at multiple lags for each method with BwTNLV in black and BwTNML in red.
 As expected, ACF corresponding to the method that marginalizes out the hidden variables from the likelihood decreases to zero at a much faster rate, indicating better mixing for the chain."}
@@
   kable(digits = 2, caption = paste0("\\label{tab:m_eff}The effective sample sizes for a chain of length $M = 1000$ of all $16$ possible vectors of visibles.
-The truncated likelihood would require at least $", round(min(M_eff_table[1, -1])/min(M_eff_table[2, -1]), 1), "$ times as many MCMC iterations to achieve the same amount of effective information about the posterior distribution."))
+BwTNLV would require at least $", round(min(M_eff_table[1, -1])/min(M_eff_table[2, -1]), 1), "$ times as many MCMC iterations to achieve the same amount of effective information about the posterior distribution."))
 ```
@@
 The effective sample sizes for a chain of length $M = 1000$ of all $2^4 = 16$ possible vectors of visibles are presented in Table \ref{tab:m_eff}.
-These range from  $`r round(min(M_eff_table[1, -1]), 2)`$ to $`r round(max(M_eff_table[1, -1]), 2)`$ for the marginalized likelihood method, while the truncated Normal method only yields between $`r round(min(M_eff_table[2, -1]), 2)`$ and $`r round(max(M_eff_table[2, -1]), 2)`$ effective draws.
-Thus, the truncated likelihood would require at least $`r round(min(M_eff_table[1, -1])/min(M_eff_table[2, -1]), 1)`$ times more iterations to be run of the MCMC chain in order to achieve the same amount of effective information about the posterior distribution.
-For this reason, consistent with the ACF results in Figure \ref{fig:acf}, the truncated Normal method does not seem to be an effective method for fitting the RBM of computing resources are at all limited.
-
-Figure \ref{fig:fitting_plot} shows the posterior probability of each possible $\boldsymbol v \in \{-1,1\}^4$ after fitting the RBM model in the two ways detailed in this section, excluding the truncated Normal method.
+These range from  $`r round(min(M_eff_table[1, -1]), 2)`$ to $`r round(max(M_eff_table[1, -1]), 2)`$ for BwTNML, while BwTNLV only yields between $`r round(min(M_eff_table[2, -1]), 2)`$ and $`r round(max(M_eff_table[2, -1]), 2)`$ effective draws.
+Thus, BwTNLV would require at least $`r round(min(M_eff_table[1, -1])/min(M_eff_table[2, -1]), 1)`$ times more iterations to be run of the MCMC chain in order to achieve the same amount of effective information about the posterior distribution.
+For this reason, consistent with the ACF results in Figure \ref{fig:acf}, BeTNLV does not seem to be an effective method for fitting the RBM of computing resources are at all limited.
+
+Figure \ref{fig:fitting_plot} shows the posterior probability of each possible $\boldsymbol v \in \{-1,1\}^4$ after fitting the RBM model in the two ways detailed in this section, excluding BwTNLV.
 The black vertical lines show the true probabilities of each image based on the parameters used to simulate the training set while the red vertical lines show the empirical distribution for the training set of $5,000$ vectors.
-From these posterior predictive checks, it is evident that the marginalized likelihood method produces the best fit to the data.
+From these posterior predictive checks, it is evident that BwTNML produces the best fit to the data.
 However, this method requires a marginalization step to obtain the probability function of visible observations alone, which is infeasible for a model with $\nh$ of any real size.
@@
   ylab("Scaled Posterior Density") + xlab("Probability of Vector of Visibles") +
-  scale_colour_discrete("Method", labels = c("Marginalized likelihood", "Trick prior")) +
-  scale_fill_discrete("Method", labels = c("Marginalized likelihood", "Trick prior")) +
+  scale_colour_discrete("Method", labels = c("BwTNML", "BwTPLV")) +
+  scale_fill_discrete("Method", labels = c("BwTNML", "BwTPLV")) +
   theme(legend.position = "bottom") -> p.models
@@
 
-```{r fitting_plot, fig.cap="\\label{fig:fitting_plot}Posterior probabilities of $16 = 2^4$ possible realizations of $4$ visibles using two of the three Bayesian fitting techniques, trick prior and marginalized likelihood.
+```{r fitting_plot, fig.cap="\\label{fig:fitting_plot}Posterior probabilities of $16 = 2^4$ possible realizations of $4$ visibles using two of the three Bayesian fitting techniques, BwTPLV and BwTNML.
 The black vertical lines show the true probabilities of each vector of visibles based on the parameters used to simulate the training data while the red vertical lines show the empirical distribution.
-The marginalized likelihood method produces the best fit for the data, however is also the most computationally intensive and least feasible with a real dataset.", fig.height=6}
+BwTNML produces the best fit for the data, however is also the most computationally intensive and least feasible with a real dataset.", fig.height=6}
 p.models

### 2016-11-17 writing/draft.Rmd
formatting

@@
 ```{r degen_plots, fig.cap="\\label{fig:degen_plots}Results from the numerical experiment, here looking at the fraction of models that were near-degenerate for each combination of magnitude of $\\boldsymbol \\theta$ and model size.
-Black lines show the contour levels for fraction of near-degeneracy, while the thick black line shows the level where the fraction of near-degenerate models is .05.", fig.height=4}
+Black lines show the contour levels for fraction of near-degeneracy, while the thick black line shows the level where the fraction of near-degenerate models is .05.", fig.height=3.5, fig.align="ht!"}
 three_ways %>%
@@
 ```{r instab_plots_2, fig.cap="\\label{fig:instab_plots}Results from the experiment, here looking at the sample mean value of $\\text{ELPR}(\\boldsymbol \\theta)/V$ at each grid point for each combination of magnitude of $\\boldsymbol \\theta$ and model size.
-As the magnitude of $\\boldsymbol \\theta$ grows, so does the value of this metric, indicating typical instability in the model.", fig.height=4}
+As the magnitude of $\\boldsymbol \\theta$ grows, so does the value of this metric, indicating typical instability in the model.", fig.height=3.5, fig.align="ht!"}
 p.elpr
@@
 ```{r uninterp_plots, fig.cap="\\label{fig:uninterp_plots}The sample mean of the maximum component of the absolute difference between the model expectation vector, E$\\left[\\boldsymbol X | \\boldsymbol \\theta\\right]$, and the expectation vector given independence, E$\\left[\\boldsymbol X | \\boldsymbol \\theta^* \\right ]$.
-Larger magnitudes of $\\boldsymbol \\theta$ correspond to larger differences, thus indicating reduced interpretability.", fig.height=4}
+Larger magnitudes of $\\boldsymbol \\theta$ correspond to larger differences, thus indicating reduced interpretability.", fig.height=3.5, fig.align="ht!"}
 p.exp_diff
@@
 ```{r split_plots2, warning=FALSE, fig.cap="\\label{fig:split_plots_dist}The fraction of near-degenerate models for each magnitude of $\\boldsymbol \\theta$.
-For each number of visibles in the model, the fraction near-degenerate becomes greater than zero at larger values of $||\\boldsymbol \\theta||$ as $\\nh$ increases and the slope becomes steeper as $\\nh$ increases also.", fig.height=5}
+For each number of visibles in the model, the fraction near-degenerate becomes greater than zero at larger values of $||\\boldsymbol \\theta||$ as $\\nh$ increases and the slope becomes steeper as $\\nh$ increases also.", fig.height=4, fig.align="ht!"}
 plot_dat_grid %>% 
@@
 
+old_pen <- options()$scipen
+options("scipen" = 100)
 sample.params %>% 
@@
   gather("Parameter", "Value") %>%
-  mutate(Value = paste0("$", Value, "$")) -> tbl
+  mutate(Value = paste0("$", round(Value, 10), "$")) -> tbl
+options("scipen" = old_pen)
 
@@
   kable(caption = "\\label{tab:theta}Parameters used to fit a test case with $V = H = 4$.
-This parameter vector was chosen as a sampled value of $\\boldsymbol \\theta$ that was not near the convex hull of the sufficient statistics for a grid point in Figure \\ref{fig:degen_plots} with $< 5$\\% near-degeneracy.", digits = 10)
+This parameter vector was chosen as a sampled value of $\\boldsymbol \\theta$ that was not near the convex hull of the sufficient statistics for a grid point in Figure \\ref{fig:degen_plots} with $< 5$\\% near-degeneracy.", align = rep(c("l", "r"), times = 3))
 ```
@@
   mutate(M_eff = sigma2/C,
-         model = "Marginal Likelihood") %>%
+         model = "BwTNML") %>%
   bind_rows(trunc_means %>%
@@
     mutate(M_eff = sigma2/C,
-           model = "Truncated Normal Prior")) %>%
+           model = "BwTNLV")) %>%
   select(-C, -sigma2) %>%
@@
   mutate(Vector = as.numeric(Image)) %>%
+  select(Vector, BwTNML, BwTNLV) %>%
   arrange(Vector) -> M_eff_tmp
@@
 Thus, BwTNLV would require at least $`r round(min(M_eff_table[1, -1])/min(M_eff_table[2, -1]), 1)`$ times more iterations to be run of the MCMC chain in order to achieve the same amount of effective information about the posterior distribution.
-For this reason, consistent with the ACF results in Figure \ref{fig:acf}, BeTNLV does not seem to be an effective method for fitting the RBM of computing resources are at all limited.
+For this reason, consistent with the ACF results in Figure \ref{fig:acf}, BwTNLV does not seem to be an effective method for fitting the RBM of computing resources are at all limited.
 
@@
   theme(legend.position = "bottom") -> p.models
+
+sx <- scale_x_continuous()
+sx$trans$breaks <- function(range) pretty(range, n = 3)
 ```
@@
 BwTNML produces the best fit for the data, however is also the most computationally intensive and least feasible with a real dataset.", fig.height=6}
-p.models
+p.models + sx
 ```

### 2016-11-17 writing/note.Rmd
nordman's note suggestions

--- 2016-10-22 writing/note.Rmd
@@
 title: "A note on the instability and degeneracy of deep learning models"
-author: "Andee Kaplan, Daniel Nordman, Stephen Vardeman"
+author: "Andee Kaplan, Daniel Nordman, and Stephen Vardeman"
 date: "October, 2016"
@@
 abstract: |
+  We introduce *Finitely Supported Finite Sequence*, or FSFS models, a broad class of discrete space models that are defined in relation to how a probability model is built to incorporate more variables.
+These models are defined to be in discrete finite space with model support equal the sample space and with a sequence of expanding model parameter vectors that potentially grow size as the dimension of the data become larger.
+In this paper, we present results that define the notion of model instability as well as near-degeneracy within the framework of FSFS models in addition to demonstrating example models of the FSFS class.
 ---
 
-# Discrete Expanding Finite space models
+# Introduction
 
-Consider $(X_1, \dots X_N)$, a set of discrete random variables with a finite sample space, $\mathcal{X}^N$.
-Let $P_{\boldsymbol \theta_N}(x_1, \dots, x_N)$ denote the probability of the data outcome $(x_1, \dots, x_N) \in \mathcal{X}^N$, which is a sequence of expanding models that potentially grow size as the dimension of the data become larger (i.e.
-$\boldsymbol \theta_N \in \mathbb{R}^{q(N)}$, where $q(N)$ is a function of $N$).
-Further, the model support of $P_{\boldsymbol \theta_N}(x_1, \dots, x_N)$ is equal to the sample space $\mathcal{X}^N$.
-We will call this class of models *Discrete Expanding Finite space*, or DEF models.
+A probability model exhibits instability if small changes in a data outcome result in large, perhaps unanticipated changes in probability.
+As data structures become larger and more complex, there is increasing interest in identifying instability in potential model specifications for such data (e.g., models for
+random graphs or networks).
+We consider the problem of  quantifying instability for general probability models defined on sequences of observations, where each sequence of a given length $N$ has a finite number of possible outcomes.
+As the number of available observations increases, a sequence of probability models results to accommodate the data of expanding dimension.
 
-Note many model families can fall under the umbrella of DEF models.
-In Sections \ref{discrete-exponential-family-models}-\ref{deep-learning}, we present three specific examples of DEF models, including models with deep architectures.
+Model instability is formally shown to occur when a certain log-probability ratio under such models grows faster than the length $n$ of the data sequence.
+In this case, a one component change in the data sequence can provably shift probability
+by orders of magnitude. Also, as the measure of instability becomes more extreme, the resulting probability models are shown  to slip into degeneracy, thereby placing all probability
+on arbitrarily small portions of the sample space.
+These results on instability apply to large classes of exponential models commonly used in random graphs, network analysis, and machine learning contexts.
+
+In this paper, we consider the behavior of probability models built to incorporate a sequence of discrete observations with an arbitrarily specified length $N$.
+Let $(X_1, \dots X_N)$ denote a set of sequence of discrete random variables with a finite sample space, $\mathcal{X}^N$.
+That is, $\mathcal{X}$ represents a finite set of potential outcomes for the variable $X_i$, and the data sequence $(X_1,\ldots,X_N)$ assumes values in the $N$-fold product space $\mathcal{X}^N$.
+For any $N$, let $P_{\theta_N}$ denote a probability model on $\mathcal{X}^N$, under which $P_{\boldsymbol \theta_N}(x_1, \dots, x_N) > 0$ is the probability of the data outcome $(x_1, \dots, x_N) \in \mathcal{X}^N$.
+We assume that the model support of $P_{\boldsymbol \theta_N}$ is equal to the sample space $\mathcal{X}^N$.
+This framework produces a series $P_{\theta_N}$ of probability models, indexed by a generic sequence of parameters $\theta_N$, to describe data of any given length $N \geq 1$.
+The size and structure of such parameters may be generally specified, and one may intuitvely consider Euclidean $\theta_N \in \mathbb{R}^{q(N)}$ for some arbitrary integer-valued function $q(\cdot) \geq 1$.
+We will call this class *Finitely Supported Finite Sequence (FSFS) models*.
+
+Section \ref{examples} provides several examples of FSFS models commonly used in graph/network analysis and machine learning (i.e., deep learning models) among other application areas.
+Section \ref{instability-results} establishes formal results regarding the model propriety of FSFS models.
+These results help to measure, and qualify the effects of, model instability that can arise when certain maximal probabilities in such models are too extreme relevant to the length $N$ of the data sample.
+**TODO: Somewhere here or in Section 3 we need a brief literature review connecting instablity here to previous work (e.g., Schweinberger) and related notions of degeneracy (e.g., Handcock)**.
+Lastly, Section \ref{implications} emphasizes the implications of our model propriety results.
+
+# Examples
+
+Note many model families can fall under the umbrella of FSFS models. In this section, we present three specific examples of FSFS models, including models with deep architectures.
 
@@
 Let $\boldsymbol \theta = (\lambda_1, \dots, \lambda_k, 0, \dots, 0)$, where the first $k$ elements are equal to the $k$ elements of $\boldsymbol \lambda$ and for $i = k + 1, \dots, q(N)$, $\theta_i = 0$.
-Further, because $p_{\boldsymbol \lambda}(\boldsymbol x) > 0$ for all $\boldsymbol x \in \mathcal{X}^N$ by properties of the exponential function, the discrete exponential family models defined in this section are a special case of the DEF model.
+Further, because $p_{\boldsymbol \lambda}(\boldsymbol x) > 0$ for all $\boldsymbol x \in \mathcal{X}^N$ by properties of the exponential function, the discrete exponential family models defined in this section are a special case of the FSFS model.
+**TODO: need to briefly add where such exponential models arise (e.g., spatial data, random graphs, network data) to indicate wide-spread applications**
 
@@
 Additionally, $p_{\boldsymbol \theta}(\boldsymbol v) > 0$ for all $\boldsymbol v \in \mathcal{X}^V$ by properties of the exponential function.
-Thus, the visible RBM model specification is an example of a DEF model.
+Thus, the visible RBM model specification is an example of a FSFS model.
+This example also indicates that models formed by marginalizing a base FSFS model (e.g., exponential family) is again a FSFS model class.
 
@@
 Additionally, $p_{\boldsymbol \theta}(\boldsymbol v) > 0$ for all $\boldsymbol v \in \mathcal{X}^V$ by properties of the exponential function.
-Thus, the visible RBM model specification is an example of a DEF model.
+Thus, the visible RBM model specification is an example of a FSFS model.
 
@@
 The difference between the two models is that all but the last stacked model are Bayesian networks, rather than RBMs.
-Thus for $V_1, \dots, V_V$ discrete with finite sample space and support equal to the sample space $\mathcal{X}^V$, a DBN is also a DEF model if the number of components in the parameter vector is dependent on the dimension of the data, $V$.
-Commonly, as in a logistic belief net [@neal1992connectionist], a "weight" parameter is placed on each interaction between visibles, $V_1, \dots, V_V$ and the first layer of latent variables, $H^{(1)}_1, \dots, H^{(1)}_{H_1}$, satisfying the definition of a DEF model.
+Thus for $V_1, \dots, V_V$ discrete with finite sample space and support equal to the sample space $\mathcal{X}^V$, a DBN is also a FSFS model if the number of components in the parameter vector is dependent on the dimension of the data, $V$.
+Commonly, as in a logistic belief net [@neal1992connectionist], a "weight" parameter is placed on each interaction between visibles, $V_1, \dots, V_V$ and the first layer of latent variables, $H^{(1)}_1, \dots, H^{(1)}_{H_1}$, satisfying the definition of a FSFS model.
 
-# Instability and degeneracy results
+# Instability results
 
-To quantify "instability" in a model with discrete outcomes, it is useful to imagine how a data model might be expanded to incorporate more and more observations.
+To measure "instability" in a FSFS model, it is useful to imagine the behavior of a data model sequence $P_{\theta_N}$ specified to incorporate more and more observations.
 A relevant quantity to this end is a (scaled) extremal log-probability ratio given by
 \begin{align}
-\frac{1}{N} \log \left[\frac{\max\limits_{(x_1, \dots, x_N) \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(x_1, \dots, x_N)}{\min\limits_{(x_1, \dots, x_N) \in \mathcal{X}^N}P_{\boldsymbol \theta}(x_1, \dots, x_N)}\right] \equiv \frac{1}{N} \text{ELPR}_N(\boldsymbol \theta_N) \label{eqn:elpr}
+\frac{1}{N} \log \left[\frac{\max\limits_{(x_1, \dots, x_N) \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(x_1, \dots, x_N)}{\min\limits_{(x_1, \dots, x_N) \in \mathcal{X}^N}P_{\boldsymbol \theta}(x_1, \dots, x_N)}\right] \equiv \frac{1}{N} \text{ELPR}_N(\boldsymbol \theta_N).
+\label{eqn:elpr}
 \end{align}
 
-Essentially, in formulating a model for potentially increasing numbers of variables (i.e., $N \rightarrow \infty$), the ratio (\ref{eqn:elpr}) should bounded, meaning that the largest probability under this model at parameter $\boldsymbol \theta_N$ should maintain a magnitude within the smallest probability allowed under the same model.
+The main idea is that, in formulating a FSFS model for potentially increasing numbers of variables (i.e., $N \rightarrow \infty$), the ratio (\ref{eqn:elpr}) should remain bounded, entailing that the largest probability possible under $P_{\theta_N}$ should maintain a magnitude within the smallest probability allowed under the same model.
+Specifically, the log of the fold change should grow at mostly linearly with the sample size $N$.
 This leads to the following definition:
 
-\begin{defn}[S-unstable DEF]
-Let $\boldsymbol \theta_N \in \mathbb{R}^{q(N)}$ be a sequence of DEF model parameters where the size of the model $q(N)$ is a function of the number of random variables $N$.
-A DEF model formulation is \emph{Schweinberger-unstable} or \emph{S-unstable} if, as the number of variables increase ($N \rightarrow \infty$), it holds that
+\begin{defn}[S-unstable FSFS]
+Let $\boldsymbol \theta_N \in \mathbb{R}^{q(N)}$ be a sequence of FSFS model parameters where the size of the model $q(N)$ is a function of the number of random variables $N$.
+A FSFS model formulation is \emph{Schweinberger-unstable} or \emph{S-unstable} if, as the number of variables increase ($N \rightarrow \infty$), it holds that
 \begin{align*}
@@
 
-S-unstable DEF models are undesirable for several reasons. One is that small changes in data can lead to overly-sensitive changes in probability. Consider, for example, 
+S-unstable FSFS models are undesirable for several reasons. One is that small changes in data can lead to overly-sensitive changes in probability. Consider, for example, 
 $$
@@
 $$
-the biggest log-probability ratio for a one component change in data outcomes at a DEF parameter $\boldsymbol \theta_N$. We then have the following result.
+the biggest log-probability ratio for a one component change in data outcomes at a FSFS parameter $\boldsymbol \theta_N$. We then have the following result.
 
@@
 \label{prop:instab}
-Let $C > 0$ and let $\text{ELPR}(\boldsymbol \theta_N)$ be as in (\ref{eqn:elpr}) for an integer $N \ge 1$.
-If $\frac{1}{N}\text{ELPR}_N(\boldsymbol \theta_N) > C$, then $\Delta_N(\boldsymbol \theta_N) > C$.
+Let $C > 0$ and let $\text{ELPR}(\boldsymbol \theta_N)$ be as in (\ref{eqn:elpr}) for an integer $N \ge 1$. If $$\frac{1}{N}\text{ELPR}_N(\boldsymbol \theta_N) > C,$$ 
+then 
+$$\Delta_N(\boldsymbol \theta_N) > C.$$
 \end{prop}
 
-In other words, if the probability ratio (\ref{eqn:elpr}) is too large, then the DEF model will exhibit large probability shifts for very small changes in the data configuration (i.e., instability).
+Again, if the probability ratio (\ref{eqn:elpr}) is too large, then the FSFS model will exhibit large probability shifts for very small changes in the data configuration, which exemplifies the intuitive notation of instability..
+
+Additionally, S-unstable FSFS models are connected to degenerate models, where *degeneracy* implies placing all probability on a small portion of the sample space.
+That is, instability in FSFS models, as measured by (\ref{eqn:elpr}), is a spectrum and Proposition \ref{prop:instab} indicates increasing levels of one undesirable type of model behavior as (\ref{eqn:elpr}) increases.
+Furthermore, as (\ref{eqn:elpr}) increases, FSFS models will slide into full degeneracy as Proposition \ref{prop:degen} next shows.
+Define a modal set
+$$
+M_{\epsilon, \boldsymbol \theta_N} \equiv \left\{\boldsymbol x \in \mathcal{X}^N: \log P_{\boldsymbol \theta_N}(\boldsymbol x) > (1-\epsilon)\max\limits_{\boldsymbol x^* \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x^*) + \epsilon\min\limits_{\boldsymbol x^* \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x^*) \right\}
+$$
+of possible outcomes, for a given $0 < \epsilon < 1$ and along a sequence $\boldsymbol \theta_N$, $N \ge 1$, of FSFS parameters.
+
+\begin{prop}
+\label{prop:degen}
+For an unstable FSFS model, and for any given $0 < \epsilon < 1$, $P_{\boldsymbol \theta_N}\left((x_1, \dots, x_N) \in M_{\epsilon, \boldsymbol \theta_N}\right) \rightarrow 1$ holds as $N \rightarrow \infty$.
+\end{prop}
+
+In other words, as a consequence of S-unstable models, all probability in the FSFS model formulation with a large number of random variables will stack up on mode sets or simply those few outcomes with the highest probability.
 
@@
 Suppose that $\Delta(\boldsymbol \theta_N) \le C$ holds for some $C > 0$.
-Under the DEF model, $P_{\boldsymbol \theta_N}(\boldsymbol x) > 0$ holds for each outcome $\boldsymbol x \in \mathcal{X}^N$.
+Under the FSFS model, $P_{\boldsymbol \theta_N}(\boldsymbol x) > 0$ holds for each outcome $\boldsymbol x \in \mathcal{X}^N$.
 Let $\boldsymbol x_{min} \equiv \argmin\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)$ and $\boldsymbol x_{max} \equiv \argmax\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)$.
@@
 using $k \le N$ and $\Delta(\boldsymbol \theta_N) \le C$. If $k = 0$, then $\boldsymbol x_{max} = \boldsymbol x_{min}$ and the same bound above holds.$\Box$
-
-Additionally, S-unstable DEF models are connected to degenerate models, where *degeneracy* implies placing all probability on a small portion of the sample space. Define a modal set 
-$$
-M_{\epsilon, \boldsymbol \theta_N} \equiv \left\{(x_1, \dots, x_N) \in \mathcal{X}^N: \log P_{\boldsymbol \theta_N}(x_1, \dots, x_N) > (1-\epsilon)\max\limits_{(x_1^*, \dots, x_N^*)}P_{\boldsymbol \theta_N}(x_1^*, \dots, x_N^*) + \epsilon\min\limits_{(x_1^*, \dots, x_N^*)}P_{\boldsymbol \theta_N}(x_1^*, \dots, x_N^*) \right\}
-$$
-of possible outcomes, for a given $0 < \epsilon < 1$ and along a sequence $\boldsymbol \theta_N$, $N \ge 1$, of DEF parameters.
-
-\begin{prop}
-\label{prop:degen}
-For an unstable DEF model, and for any given $0 < \epsilon < 1$, $P_{\boldsymbol \theta_N}\left((x_1, \dots, x_N) \in M_{\epsilon, \boldsymbol \theta_N}\right) \rightarrow 1$ holds as $N \rightarrow \infty$.
-\end{prop}
-
-In other words, as a consequence of S-unstable models, all probability in the DEF model formulation with a large number of random variables will stack up on mode sets or simply those few outcomes with the highest probability.
 
@@
 \end{align*}
-as $N \rightarrow \infty$ by the definition of an unstable DEF model.
+as $N \rightarrow \infty$ by the definition of an unstable FSFS model.
 Consequently, $P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N}) \rightarrow 1$ as $N \rightarrow \infty$ as claimed.$\Box$
@@
 \end{align*}
-Meaning, the definition of an S-unstable DEF model is invariant to replication.
+Meaning, the definition of an S-unstable FSFS model is invariant to replication.
 \end{remark}
@@
 For a large class of models that cover a broad range of applications (including "deep learning"), we have developed a formal definition and elucidated multiple consequences of instability.
-We have shown for DEF models that instability manifests through small changes in data leading to overly-sensitive changes in probability as well as the potential to place all probability on a small pieace of the sample space.
-Models that fall within the definition of a DEF model should be used with caution to ensure that the effects of instability are not experienced.
+We have shown for FSFS models that instability manifests through small changes in data leading to overly-sensitive changes in probability as well as the potential to place all probability on a small pieace of the sample space.
+Models that fall within the definition of a FSFS model should be used with caution to ensure that the effects of instability are not experienced.
 

### 2016-11-17 writing/note.Rmd
try to do nordman's ideas

@@
 abstract: |
-  We introduce *Finitely Supported Finite Sequence*, or FSFS models, a broad class of discrete space models that are defined in relation to how a probability model is built to incorporate more variables.
-These models are defined to be in discrete finite space with model support equal the sample space and with a sequence of expanding model parameter vectors that potentially grow size as the dimension of the data become larger.
-In this paper, we present results that define the notion of model instability as well as near-degeneracy within the framework of FSFS models in addition to demonstrating example models of the FSFS class.
+  A probability model exhibits instability if small changes in a data outcome result in large, perhaps unanticipated changes in probability.
+As data structures become larger and more complex, there is increasing interest in identifying instability in potential model specifications for such data (e.g., models for random graphs or networks).
+We consider the problem of quantifying instability for general probability models defined on sequences of observations, where each sequence of a given length $N$ has a finite number of possible outcomes.
+As the number of available observations increases, a sequence of probability models results to accommodate the data of expanding dimension.
+  Model instability is formally shown to occur when a certain log-probability ratio under such models grows faster than the length $N$ of the data sequence.
+In this case, a one component change in the data sequence can provably shift probability by orders of magnitude.
+Also, as the measure of instability becomes more extreme, the resulting probability models are shown  to slip into degeneracy, thereby placing all probability on arbitrarily small portions of the sample space.
+These results on instability apply to large classes of exponential models commonly used in random graphs, network analysis, and machine learning contexts.
 ---
@@
 A probability model exhibits instability if small changes in a data outcome result in large, perhaps unanticipated changes in probability.
-As data structures become larger and more complex, there is increasing interest in identifying instability in potential model specifications for such data (e.g., models for
-random graphs or networks).
+As data structures become larger and more complex, there is increasing interest in identifying instability in potential model specifications for such data (e.g., models for random graphs or networks).
+The concept of instability, introduced in the field statistical physics by @ruelle1999statistical, was extended to include a notion of detection and quantification by @schweinberger2011instability.
 We consider the problem of quantifying instability for general probability models defined on sequences of observations, where each sequence of a given length $N$ has a finite number of possible outcomes.
@@
 
-Model instability is formally shown to occur when a certain log-probability ratio under such models grows faster than the length $n$ of the data sequence.
-In this case, a one component change in the data sequence can provably shift probability
-by orders of magnitude. Also, as the measure of instability becomes more extreme, the resulting probability models are shown  to slip into degeneracy, thereby placing all probability
-on arbitrarily small portions of the sample space.
-These results on instability apply to large classes of exponential models commonly used in random graphs, network analysis, and machine learning contexts.
+Model instability is formally shown to occur when a certain log-probability ratio under such models grows faster than the length $N$ of the data sequence.
+In this case, a one component change in the data sequence can provably shift probability by orders of magnitude.
+Also, as the measure of instability becomes more extreme, the resulting probability models are shown to slip into degeneracy, thereby placing all probability on arbitrarily small portions of the sample space.
+This relationship between instability and degeneracy, as defined by @handcock2003assessing, was described by @schweinberger2011instability.
+We extend these results on instability to large classes of exponential models commonly used in random graphs, network analysis, and machine learning contexts.
 
@@
 These results help to measure, and qualify the effects of, model instability that can arise when certain maximal probabilities in such models are too extreme relevant to the length $N$ of the data sample.
-**TODO: Somewhere here or in Section 3 we need a brief literature review connecting instablity here to previous work (e.g., Schweinberger) and related notions of degeneracy (e.g., Handcock)**.
 Lastly, Section \ref{implications} emphasizes the implications of our model propriety results.
@@
 Further, because $p_{\boldsymbol \lambda}(\boldsymbol x) > 0$ for all $\boldsymbol x \in \mathcal{X}^N$ by properties of the exponential function, the discrete exponential family models defined in this section are a special case of the FSFS model.
-**TODO: need to briefly add where such exponential models arise (e.g., spatial data, random graphs, network data) to indicate wide-spread applications**
+Such exponential models arise in spatial data on a lattice [@besag1974spatial], network data [@wasserman1994social, @handcock2003assessing], and even in binomial sampling with $N$ iid Bernoulli random variables.
 

### 2016-11-18 writing/appendix/appendix-center.Rmd
Nordman's edits

@@
+---
+title: "Supplement to: On the propriety of restricted Boltzmann machines"
+author: "Andee Kaplan, Daniel Nordman, and Stephen Vardeman"
+date: "November, 2016"
+linestretch: 2 
+output: 
+  pdf_document:
+    fig_caption: yes
+    includes:
+      in_header: ../latex/header.tex
+    number_sections: yes
+---
+
+# On the one-to-one correspondence between parameters and moments in the RBM distribution
+
 For integers $m,n \geq 1$, consider random vectors $\bm{V}_m \equiv(V_1,\ldots,V_m)$ and $\bm{H}_n\equiv (H_1,\ldots,H_n)$, where the random variables $V_i$ and $H_j$ assume values in $\{1,-1\}$, $i=1,\ldots,m$, $j=1,\ldots,n$.
@@
 
-{\bf Proof of Theorem \ref{thm:one}.} We first establish the equivalence of claims 2 and 3. If claim 2 holds, then
+**Proof of Theorem \ref{thm:one}.** We first establish the equivalence of claims 2 and 3. If claim 2 holds, then
  claim 3 holds trivially.
@@
 
-{\bf Proof.} The case $m=n=1$ follows from Lemma \ref{lemma:two}.
+**Proof.** The case $m=n=1$ follows from Lemma \ref{lemma:two}.
 We now use an induction argument, assuming that the result of Lemma \ref{lemma:one} holds for some given order of $(m,n)$ with $m,n\geq 1$ and show the the result continues to hold for $(m+1,n)$ or $(m,n+1)$.
@@
 
-{\bf Proof.} Denote the four probabilities by which the random vector $(X,Y)$ assume pairs $(b,b),(b,a),(a,b),(a,a)$, respectively, as $c_1,c_2,c_3,1-c_1-c_2-c_3$ under $P_1$ and $d_1,d_2,d_3,1-d_1-d_2-d_3$ under $P_2$.
+**Proof.** Denote the four probabilities by which the random vector $(X,Y)$ assume pairs $(b,b),(b,a),(a,b),(a,a)$, respectively, as $c_1,c_2,c_3,1-c_1-c_2-c_3$ under $P_1$ and $d_1,d_2,d_3,1-d_1-d_2-d_3$ under $P_2$.
 The conditions $\E_1 X = \E_2 X$, $\E_1 Y = \E_2 Y$, $\E_1 X Y = \E_2 X Y$ with $b\neq a$ imply that
@@
 
-{\bf Proof of Theorem \ref{thm:two}.}  We first establish the equivalence of claims 2 and 3.
+**Proof of Theorem \ref{thm:two}.**  We first establish the equivalence of claims 2 and 3.
 If claim 2 holds, then claim 3 follows easily from the resulting uniform cell probabilities: $P(V_1= v_1,\ldots,V_m= v_m, H_1= h_1,\ldots,H_n=h_n|\bm{\theta})=2^{-mn}$ for any $v_i,h_j\in\{\pm 1\}$, $i=1,\ldots,m$, $j=1,\ldots,n$.

### 2016-11-18 writing/draft.Rmd
Nordman's edits

@@
 First, the models are often unsatisfactory as conceptualizations of how data are generated.
-According to Fisher, the aim of a statistical model is to represent data in a compact way [@fisher1922mathematical].
-Neyman and Box agree with this assessment and go further to state that a model should "provide an explanation of the mechanism underlying the observed phenomena" [@lehmann1990model, @box1967discrimination].
+Recalling [@fisher1922mathematical], the aim of a statistical model is to represent data in a compact way.
+Neyman and Box further state that a model should "provide an explanation of the mechanism underlying the observed phenomena" [@lehmann1990model, @box1967discrimination].
 At issue, RBMs  often produce data lacking realistic variability and may thereby fail to provide a satisfactory conceptualization of a data generation process.
@@
 As the size of these models grows, both maximum likelihood and Bayesian methods of fitting quickly become intractable.
-The literature often suggests Markov chain Monte Carlo (MCMC) approximations to posteriors to fit these models (e.g., Gibbs sampling to exploit conditional structure in the hidden and visible variables), but little is said about acheiving convergence [@hinton2010practical, @hinton2006fast].
+The literature often suggests Markov chain Monte Carlo (MCMC) tools for approximate maximization of likelihoods to fit these models (e.g., Gibbs sampling to exploit conditional structure in hidden and visible variables), but little is said about acheiving convergence [@hinton2010practical, @hinton2006fast].
 Related to this, these MCMC algorithms require updating potentially many latent variables (hiddens) which can critically influence convergence in MCMC-based likelihood methods.
 
-In Section \ref{bayesian-model-fitting}, we compare three Bayesian techniques (involving MCMC approximations that are computationally more accessible than direct maximum likelihood), each proposed with the goal of avoiding parts of a RBM parameter space that yield unattractive models.
+In Section \ref{bayesian-model-fitting}, we compare three Bayesian techniques involving MCMC approximations that are computationally more accessible than direct maximum likelihood, which also aim to avoid parts of a RBM parameter space that yield unattractive models.
 As might be expected, with greater computational complexity comes an increase in fitting accuracy, but at the cost of practical feasibility.
@@
 For a RBM model with enough hidden variables, it has been shown that any distribution for the visibles can be approximated arbitrarily well [@le2008representational, @montufar2011refinements, and @montufar2011expressive].
-The empirical distribution of a training set of vectors of visibles is the best fitting model for observed cell data and so principled RBM fitting will seek to produce this when possible.
-That is, any fully principled fitting method based on the likelihood for RBM with enough hidden variables will simply reproduce the (discrete) empirical distribution of a training set.
+However, the empirical distribution of a training set of vectors of visibles is the best fitting model for observed cell data.
+As a consequence, we find that any fully principled fitting method based on the likelihood for RBM with enough hidden variables will simply seek to reproduce the (discrete) empirical distribution of a training set.
 Thus, there is no "smoothed distribution" achieved in fitting a RBM model of sufficient size with a rigorous likelihood-based method.
@@
 Section \ref{explorations} details our explorations into the model behavior and propriety (or lack thereof) of the RBM class.
-We discuss three Bayesian fitting techniques intended to avoid fitted model impropriety in Section \ref{model-fitting} and conclude with a discussion in Section \ref{discussion}.
-Additionally, there is an accompanying appendix with proofs of main results.
+We discuss three Bayesian fitting techniques intended to avoid model impropriety in Section \ref{model-fitting} and conclude with a discussion in Section \ref{discussion}.
+On-line supplementary materials provide proofs for results on RBM parameterizations and data codings described in Section \ref{data-coding-to-mitigate-apparent-degeneracy}.
 
@@
 
-A restricted Boltzmann machine (RBM) is an undirected graphical model specified for discrete or continuous random variables, binary variables being the case most commonly considered.
-In this paper, we consider the binary case of a RBM for concreteness.
+A restricted Boltzmann machine (RBM) is an undirected graphical model specified for discrete or continuous random variables, binary variables being most commonly considered.
+In this paper, we consider the binary case for concreteness.
 A RBM architecture has two layers, hidden ($\mathcal{H}$) and visible ($\mathcal{V}$), with no dependency connections within a layer.
@@
 
-A common use for RBMs is to create features for use in classification. For example, binary images can be classified where the pixel values are treated as the visible variables $v_i$ [@hinton2006fast].
+A common use for RBMs is to create features for use in classification.
+For example, binary images can be classified through a process that treats the pixel values as the visible variables $v_i$ in a RBM model [@hinton2006fast].
 
@@
 Each single binary random variable, visible or hidden, will take its values in a common coding set $\mathcal{C}$, where we allow one of two possibilities for the coding set, $\mathcal{C}=\{0,1\}$ or $\mathcal{C}=\{-1,1\}$, with "$1$" always indicating the "high" value of the variable.
-While $\mathcal{C}=\{0,1\}$ may be a natural or naive starting point, we argue later that the coding $\mathcal{C}=\{-1,1\}$ induces more interpretable model properties for the RBM.
+While $\mathcal{C}=\{0,1\}$ may be a natural starting point, we argue later that the coding $\mathcal{C}=\{-1,1\}$ induces more interpretable model properties for the RBM (cf.
+Section \ref{explorations}).
 A standard parametric form for probabilities corresponding to a potential vector of states, $X = (H_1, \dots, H_\nh, V_1,\dots,V_\nv)$, for the nodes is
@@
 
-where $\boldsymbol \theta \equiv (\theta_{11}, \dots, \theta_{1\nh}, \dots, \theta_{\nv 1}, \dots, \theta_{\nv \nh}, \theta_{v_1}, \dots, \theta_{v_\nv}, \theta_{h_1}, \dots, \theta_{h_\nh}) \in \mathbb{R}^{\nv + \nh + \nv*\nh}$ denotes the vector of model parameters and the denominator $\gamma(\boldsymbol \theta) = \sum\limits_{\boldsymbol x \in \mathcal{C}^{\nh+\nv}}\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)$ is the normalizing function that ensures the probabilities (\ref{eqn:pmf}) sum to one.
+where $\boldsymbol \theta \equiv (\theta_{11}, \dots, \theta_{1\nh}, \dots, \theta_{\nv 1}, \dots, \theta_{\nv \nh}, \theta_{v_1}, \dots, \theta_{v_\nv}, \theta_{h_1}, \dots, \theta_{h_\nh}) \in \mathbb{R}^{\nv + \nh + \nv*\nh}$ denotes the vector of model parameters and the denominator $\gamma(\boldsymbol \theta) = \sum_{\boldsymbol x \in \mathcal{C}^{\nh+\nv}}\exp\left(\sum_{i = 1}^\nv \sum_{j=1}^\nh \theta_{ij} v_i h_j + \sum_{i = 1}^\nv\theta_{v_i} v_i + \sum_{j = 1}^\nh\theta_{h_j} h_j\right)$ is the normalizing function that ensures the probabilities (\ref{eqn:pmf}) sum to one.
 For
 \begin{align}
-\boldsymbol x &= (h_1, \dots, h_\nh, v_1, \dots, v_\nh) \in \mathcal{C}^{\nv + \nh} \text{ and} \\
+\boldsymbol x &= (h_1, \dots, h_\nh, v_1, \dots, v_\nh) \in \mathcal{C}^{\nv + \nh} \text{ and} \nonumber\\
 \boldsymbol t(\boldsymbol x) &= (h_1, \dots, h_\nh, v_1, \dots, v_\nv, v_1 h_1, \dots, v_\nv h_\nh) \in \mathcal{C}^{\nh + \nv + \nh*\nv}, \label{eqn:t}
@@
 Due to the potential size of the model, the normalizing constant $\gamma(\boldsymbol \theta)$ can be practically impossible to calculate, making simple estimation of the model parameter vector problematic.
-A kind of Gibbs sampling can be tried (which is natural due to conditional independence architecture of the RBM, i.e.
+A kind of Gibbs sampling can be tried (due to conditional architecture of the RBM, i.e.
 visibles given hiddens or vice versa).
-Specifically, the conditional independence of nodes in each layer, given those nodes in the other layer allows for stepwise simulation of both hidden layers and model parameters possible.
+Specifically, the conditional independence of nodes in each layer, given those nodes in the other layer, allows for stepwise simulation of both hidden layers and model parameters possible (e.g., contrastive divergence of @hinton2002training or Bayes methods in Section \ref{model-fitting}).
 
@@
 A RBM model can be thought to exhibit an analogous form of *near-degeneracy* when there is a disproportionate amount of probability placed on a small number of elements in the sample space of visibles and hiddens, $\mathcal{C}^{\nv + \nh}$.
-Using the idea of @handcock2003assessing, when the random vector $\boldsymbol t(\boldsymbol x) = (v_1, \dots, v_\nv, h_1, \dots, h_\nh, v_1 h_1, \dots, v_V h_\nh ) \in \mathcal{T} \equiv \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{C}^{\nh + \nv} \}$ in the neg-potential function $Q_{\boldsymbol \theta}(\cdot)$ has a mean vector $\boldsymbol \mu(\boldsymbol \theta) \in \mathbb{R}^{\nv + \nh + \nv * \nh}$ close to the boundary of the convex hull of $\mathcal{T}$, then the RBM model can be said to exhibit near-degeneracy at $\boldsymbol \theta \in \mathbb{R}^{\nv + \nh + \nh*\nv}$.
+Using the idea of @handcock2003assessing, when the random vector $\boldsymbol t(\boldsymbol x) = (v_1, \dots, v_\nv, h_1, \dots, h_\nh, v_1 h_1, \dots, v_V h_\nh ) \in \mathcal{T} \equiv \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{C}^{\nh + \nv} \}$ from (\ref{eqn:t}), appearing in the neg-potential function $Q_{\boldsymbol \theta}(\cdot)$, has a mean vector $\boldsymbol \mu(\boldsymbol \theta) \in \mathbb{R}^{\nv + \nh + \nv * \nh}$ close to the boundary of the convex hull of $\mathcal{T}$, then the RBM model can be said to exhibit near-degeneracy at $\boldsymbol \theta \in \mathbb{R}^{\nv + \nh + \nh*\nv}$.
 Here the mean of $\boldsymbol t(\boldsymbol x)$ is
@@
 When increasing the size of RBM models, it becomes necessary to grow the number of model parameters (and in this process one may also arbitrarily expand the number of hidden variables used).
-To this end, let $\boldsymbol \theta_{\nv} \equiv (\theta_{v_1}, \dots, \theta_{v_\nv}, \theta_{h_1}, \dots, \theta_{h_\nh}, \theta_{11}, \dots, \theta_{\nv \nh}), \nv \ge 1$, denote an element of a sequence of RBM parameters indexed by the number $V$ of visibles $(V_1, \dots, V_V)$ and define a (scaled) extremal log-probability ratio of the RBM model at $\boldsymbol \theta_{\nv}$ as
+To this end, let $\boldsymbol \theta_{\nv} \equiv (\theta_{v_1}, \dots, \theta_{v_\nv}, \theta_{h_1}, \dots, \theta_{h_\nh}, \theta_{11}, \dots, \theta_{\nv \nh}), \nv \ge 1$, denote an element of a sequence of RBM parameters indexed by the number $\nv$ of visibles $(V_1, \dots, V_\nv)$ and define a (scaled) extremal log-probability ratio of the RBM model at $\boldsymbol \theta_{\nv}$ as
 \begin{align}
@@
 
-Additionally, S-unstable RBM model sequences are connected to the near-degeneracy of section \ref{near-degeneracy} (in which model sequences place all probability on a small portion of their sample spaces).
-To see this, define a modal set of possible outcomes (i.e.
+Additionally, S-unstable RBM model sequences are connected to the near-degeneracy of Section \ref{near-degeneracy} (in which model sequences place all probability on a small portion of their sample spaces).
+To see this, define an arbitrary modal set of possible outcomes (i.e.
 set of highest probability outcomes) in RBM models with parameters $\boldsymbol \theta_\nv, \nv \ge 1$ as
@@
 
-In other words, in S-unstable RBM model sequences with a large number of visible variables, all probability in the RBM model formulation  will stack up on those outcomes for visibles with with the highest probability.
+In other words, S-unstable RBM model sequences are gwauranteed to stack up all probability on arbitrarily narrow outcomes for visibles.
 Proofs of Propositions \ref{prop:instab}-\ref{prop:degen} appear in @kaplan2016note.
@@
 Note that in a RBM, the parameters $\theta_{v_1}, \dots, \theta_{v_\nv}$ and $\theta_{h_1}, \dots, \theta_{h_\nh}$ are naturally associated with main effects of visible and hidden variables and can be interpreted as (logit functions of) means for variables $V_1, \dots, V_\nv, H_1, \dots, H_\nh$ in a model with no interaction parameters, $\theta_{ij} = 0, i = 1, \dots, \nv, j = 1, \dots, \nh$.
-Hence, these main effect parameters have a clear interpretation under an independence model (one with $\theta_{ij} = 0$) but this interpretation (as well as an understanding of the nature of dependence) can break down as interaction parameters increase in magnitude relative to the size of the main effects.
-In such cases, the main effect parameters $\theta_{v_1}$ and $\theta_{h_j}$ are no longer interpretable as originally intended in the models (statements of marginal means) and the dependence parameters are so large as to dominate the entire model probability structure (destroying the model interpretation of dependence as local conditional modifications of an overall marginal mean structure).
+That is, with no interaction parameters, we have from (\ref{eqn:pmf}) that
+$$
+P_\theta(V_i=1) \propto e^{\theta_{v_i}} \quad \& \quad P_\theta(H_j=1) \propto e^{\theta_{h_j}}, \quad  i=1,\ldots,\nv,j=1,\ldots,\nh
+$$
+so that, for example, $logit(P_\theta(V_i=1)=\theta_{v_i}$ (or $2 \theta_{v_i}$) under the coding $\mathcal{C}=\{0,1\}$ (or $\{-1,1\}$).
+Hence, these main effect parameters have a clear interpretation under an independence model (one with $\theta_{ij} = 0$) but this interpretation can break down as interaction parameters increase in magnitude relative to the size of the main effects.
+In such cases, the main effect parameters $\theta_{v_1}$ and $\theta_{h_j}$ are no longer interpretable as originally intended in the models (statements of marginal means) and the dependence parameters are so large as to dominate the entire model probability structure (also destroying the model interpretation of dependence as local conditional modifications of an overall marginal mean structure).
 To assess which parameter values $\boldsymbol \theta$ may cause difficulties in interpretation, we use the difference $\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] - \text{E}\left[\boldsymbol X | \boldsymbol \theta^* \right]$ between two model expectations: E$\left[\boldsymbol X | \boldsymbol \theta\right]$ at $\boldsymbol \theta$ and expectations E$\left[\boldsymbol X | \boldsymbol \theta^* \right]$ given independence where $\boldsymbol \theta^*$ matches $\boldsymbol \theta$ for all main effects but otherwise has $\theta_{ij} = 0$ for $i = 1, \dots, \nv, j = 1, \dots, \nh$.
@@
 To illustrate the ideas of model near-degeneracy, instability, and uninterpretability in a RBM, we consider first the smallest possible (toy) example that consists of one visible node $v_1$ and one hidden node $h_1$ that are both binary.
-A schematic of this model can be found in figure \ref{fig:toymodel}.
+A schematic of this model can be found in Figure \ref{fig:toymodel}.
 Because it seems most common, we shall begin by employing $0/1$ encoding of binary variables (both $h_1$ and $v_1$ taking values in $\mathcal{C} = \{0,1\}$).
-(Eventually we'll argue in Section \ref{data-coding-to-mitigate-apparent-degeneracy} that $-1/1$ coding has substantial advantages.)
+(Eventually we shall argue in Section \ref{data-coding-to-mitigate-apparent-degeneracy} that $-1/1$ coding has substantial advantages.)
 
@@
 For this small model, we are able to investigate the symptoms of model impropriety, beginning with near-degeneracy.
-To this end, recall that one characterization requires consideration of the convex hull of possible values of statistics $\boldsymbol t(\boldsymbol x)$,
+To this end, recall from Section \ref{near-degeneracy} that one characterization requires consideration of the convex hull of possible values of statistics $\boldsymbol t(\boldsymbol x)$,
 $$\mathcal{T} = \{\boldsymbol t(\boldsymbol x): \boldsymbol x = (v_1, h_1) \in \{0,1\}^2\} \equiv \{(v_1, h_1, v_1 h_1): v_1, h_1 \in \{0,1\}\}$$ 
 appearing in the RBM probabilities for this model.
-As this set is in three dimensions, we are able to explicitly illustrate the shape of boundary of the convex hull of $\mathcal{T}$ and explore the behavior of the mean vector $\text{E}_{\boldsymbol \theta} \boldsymbol t(\boldsymbol x)$ as a function of the parameter vector $\boldsymbol \theta$.
-Figure \ref{fig:toyhull} shows the convex hull of our "statistic space," $\mathcal{T} \subset \{0,1\}^3$, for this toy problem from two perspectives (enclosed by the unit cube, representing the convex hull of $\{0,1\}^3$ space).
-In this small model, note that the convex hull of $\mathcal{T}$ does not fill the unrestricted hull of $\{0,1\}^3$ because of the relationship between the elements of $\mathcal{T} = \{(v_1, h_1, v_1 h_1 : v_1, h_1 \in \{0, 1\} \}$ (i.1.
+As this set is in three dimensions, we are able to explicitly illustrate the shape of boundary of the convex hull of $\mathcal{T}$ and explore the behavior of the mean vector $\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta} \boldsymbol t(\boldsymbol x)$ as a function of the parameter vector $\boldsymbol \theta$.
+Figure \ref{fig:toyhull} shows the convex hull of our "statistic space," $\mathcal{T} \subset \{0,1\}^3$, for this toy problem from two perspectives (enclosed by the unit cube $[0,1]^3$, representing the convex hull of $\{0,1\}^3$ space).
+In this small model, note that the convex hull of $\mathcal{T}$ does not fill the unrestricted hull of $\{0,1\}^3$ because of the relationship between the elements of $\mathcal{T} = \{(v_1, h_1, v_1 h_1 : v_1, h_1 \in \{0, 1\} \}$ (i.e.
 $v_1 h_1 = 1$ only if $v_1 = h_1 = 1$).
@@
 \begin{align*}
-\text{ELPR}(\boldsymbol \theta) &= \log \left[\frac{\max\limits_{(v_1, \dots, v_\nv) \in \mathcal{C}^\nv}P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}{\min\limits_{(v_1, \dots, v_\nv) \in \mathcal{C}^\nv}P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}\right]\\
-&= \log \left[\frac{\max\limits_{v_1 \in \mathcal{C}}\sum\limits_{h_1 \in \mathcal{C}}\exp\left\{\theta_{11} h_1 v_1 + \theta_{h_1}h_1 + \theta_{v_1} v_1 \right\}}{\min\limits_{v_1 \in \mathcal{C}}\sum\limits_{h_1 \in \mathcal{C}}\exp\left\{\theta_{11} h_1 v_1 + \theta_{h_1}h_1 + \theta_{v_1} v_1 \right\}}\right].
+\text{ELPR}(\boldsymbol \theta) &= \log \left[\frac{\max\limits_{(v_1, \dots, v_\nv) \in \mathcal{C}^\nv}P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}{\min\limits_{(v_1, \dots, v_\nv) \in \mathcal{C}^\nv}P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}\right] = \log \left[\frac{\max\limits_{v_1 \in \mathcal{C}}\sum\limits_{h_1 \in \mathcal{C}}\exp\left\{\theta_{11} h_1 v_1 + \theta_{h_1}h_1 + \theta_{v_1} v_1 \right\}}{\min\limits_{v_1 \in \mathcal{C}}\sum\limits_{h_1 \in \mathcal{C}}\exp\left\{\theta_{11} h_1 v_1 + \theta_{h_1}h_1 + \theta_{v_1} v_1 \right\}}\right].
 \\
@@
 \begin{matrix}
-\frac{\exp\left(\theta_{v_1}\right) + \exp\left(\theta_{11} + \theta_{v_1} + \theta_{h_1}\right)}{\exp\left(\theta_{v_1}\right) + \exp\left(\theta_{h_1}\right) + \exp\left(\theta_{11} + \theta_{v_1} + \theta_{h_1}\right)} \\
-\frac{\exp\left(\theta_{h_1}\right) + \exp\left(\theta_{11} + \theta_{v_1} + \theta_{h_1}\right)}{\exp\left(\theta_{v_1}\right) + \exp\left(\theta_{h_1}\right) + \exp\left(\theta_{11} + \theta_{v_1} + \theta_{h_1}\right)}
-\end{matrix} \right] - \left[
-\begin{matrix}
-\frac{\exp\left(\theta_{v_1}\right) + \exp\left(\theta_{v_1} + \theta_{h_1}\right)}{\exp\left(\theta_{v_1}\right) + \exp\left(\theta_{h_1}\right) + \exp\left(\theta_{v_1} + \theta_{h_1}\right)} \\
-\frac{\exp\left(\theta_{h_1}\right) + \exp\left(+ \theta_{v_1} + \theta_{h_1}\right)}{\exp\left(\theta_{v_1}\right) + \exp\left(\theta_{h_1}\right) + \exp\left(+ \theta_{v_1} + \theta_{h_1}\right)} 
-\end{matrix} \right] \\
-&= \left[
-\begin{matrix}
 \frac{\exp\left(\theta_{11} + \theta_{v_1} + 2\theta_{h_1}\right) - \exp\left( \theta_{v_1} + 2\theta_{h_1}\right) }{\left(\exp\left(\theta_{v_1}\right) + \exp\left(\theta_{h_1}\right) + \exp\left(\theta_{11} + \theta_{v_1} + \theta_{h_1}\right)\right)\left(\exp\left(\theta_{v_1}\right) + \exp\left(\theta_{h_1}\right) + \exp\left(+ \theta_{v_1} + \theta_{h_1}\right)\right)} \\
@@
 In addition to the argument that the $-1/1$ data encoding mitigates some perceived prevalence of near-degeneracy it also has the benefit of providing a guaranteed-to-be non-degenerate model at $\boldsymbol \theta = \boldsymbol 0 \in \mathbb{R}^{\nh + \nv + \nh*\nv}$, where the zero vector that serves as the natural center of the parameter space and induces the simplest possible model properties for the RBM (i.e., at $\boldsymbol \theta = \boldsymbol 0$, all variables are independent and visible variables are independent and uniformly distributed on $\{-1,1\}^\nv$).
-The proof of this and further exploration o the equivalence of the $\boldsymbol \theta$ parameterization of the RBM model class and parameterization by $\boldsymbol \mu(\boldsymbol \theta)$ is in appendix \ref{center-of-universe}.
+The proof of this and further exploration of the equivalence of the $\boldsymbol \theta$ parameterization of the RBM model class and parameterization by $\boldsymbol \mu(\boldsymbol \theta)$ is in the supplementary materials.
 Hence while from some computing perspectives $0/1$ coding might seem most natural, the $-1/1$ coding is far more convenient and interpretable from the point of view of statistical modeling, where it makes parameters simply interpreted in terms of symmetrically defined main effects and interactions.
@@
 For each set of parameters we then calculate metrics of model impropriety introduced in Section \ref{degeneracy-instability-and-uninterpretability-oh-my} based on $\boldsymbol \mu(\boldsymbol \theta)$, $\text{ELPR}(\boldsymbol \theta)/\nv$, and the coordinates of $\left|\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] -   \text{E}\left[\boldsymbol X | \boldsymbol \theta^* \right] \right|$.
-In the case of near-degeneracy, we classify each model as near-degenerate or "viable" based on the distance of $\boldsymbol \mu(\boldsymbol \theta)$ from the boundary of the convex hull of $\mathcal{T}$ and look at the fraction of models that are "near-degenerate," meaning they are within a small distance of the boundary of the convex hull.
+In the case of near-degeneracy, we classify each model as near-degenerate or "viable" based on the distance of $\boldsymbol \mu(\boldsymbol \theta)$ from the boundary of the convex hull of $\mathcal{T}$ and look at the fraction of models that are "near-degenerate," meaning they are within a small distance $\epsilon > 0$ of the boundary of the convex hull.
 We define a "small" through a rough estimation of the volume of the hull for each model size.
 We pick $\epsilon_0 = 0.05$ for $\nh=\nv=1$ and then, for every other $\nh$ and $\nv$, set $m=\nh+\nv+\nv*\nh$ and pick $\epsilon$ so that $1-(1-2\epsilon_0)^3 = 1 - (1-2\epsilon)^m$.
-In this way, we roughly scale the volume of the "small distance" to the boundary of the convex hull to be equivalent across dimensions.
-
-In our numerical experiment, we split $\boldsymbol \theta$ into $\boldsymbol \theta_{main}$ and $\boldsymbol \theta_{interaction}$, in reference to which variables in the probability function the parameters correspond (whether they multiply a $v_i$ or a $h_j$ or they multiply a $v_i h_j$), and allow the two types of terms to have varying average magnitudes, $||\boldsymbol \theta_{main} || /(\nh+\nv)$ and $||\boldsymbol \theta_{interaction} || /(\nh*\nv)$.
+In this way, we roughly scale the volume of the "small distance" to the boundary of the convex hull to be equivalent across model dimensions.
+
+In our numerical experiment, we split $\boldsymbol \theta = (\boldsymbol \theta_{main}, \boldsymbol \theta_{interaction})$ into $\boldsymbol \theta_{main}$ and $\boldsymbol \theta_{interaction}$, in reference to which variables in the probability function the parameters correspond (whether they multiply a $v_i$ or a $h_j$ or they multiply a $v_i h_j$), and allow the two types of terms to have varying average magnitudes, $||\boldsymbol \theta_{main} || /(\nh+\nv)$ and $||\boldsymbol \theta_{interaction} || /(\nh*\nv)$.
 These average magnitudes vary on a grid between `r min(plot_dat_grid$r1/(plot_dat_grid$H + plot_dat_grid$V))` and `r max(plot_dat_grid$r1/(plot_dat_grid$H + plot_dat_grid$V))` with `r length(unique(plot_dat_grid$r1/(plot_dat_grid$H + plot_dat_grid$V)))` breaks, yielding `r length(unique(plot_dat_grid$r1/(plot_dat_grid$H + plot_dat_grid$V)))^2` grid points.
-(By looking at the average magnitudes, we are able to later consider the potential benefit of shrinking each parameter value $\theta_i$ towards zero in a Bayesian fitting technique.) At each point in the grid, `r dim(res$samp[[1]])[1]` vectors are uniformly sampled on a sphere with radius corresponding to the first coordinate in the grid ($\boldsymbol \theta_{main}$) and `r dim(res$samp[[1]])[1]` vectors are uniformly sampled on a sphere with radius corresponding to the second coordinate in the grid ($\boldsymbol \theta_{main}$) via sums of squared and scaled iid Normal$(0, 1)$ variables.
+(By looking at the average magnitudes, we are able to later consider the potential benefit of shrinking each parameter value $\theta_i$ towards zero in a Bayesian fitting technique.) At each point in the grid, `r dim(res$samp[[1]])[1]` vectors ($\boldsymbol \theta_{main}$) are uniformly sampled on a sphere with radius corresponding to the first coordinate in the grid and `r dim(res$samp[[1]])[1]` vectors ($\boldsymbol \theta_{interction}$) are uniformly sampled on a sphere with radius corresponding to the second coordinate in the grid via sums of squared and scaled iid Normal$(0, 1)$ variables.
 These vectors are then paired to create `r dim(res$samp[[1]])[1]` values of $\boldsymbol \theta$ with magnitudes at each point in the grid.
@@
 Figure \ref{fig:split_plots_dist} shows the fraction of near-degenerate models for each magnitude of $\boldsymbol \theta$ for each model architecture.
-For each number of visibles in the model, the fraction near-degenerate becomes greater than zero at larger values of $||\boldsymbol \theta||$ as $\nh$ increases and the slope becomes steeper as $\nh$ increases also.
+For each number $\nv$ of visibles in the model, as the number $\nh$ of hiddens increase, the fraction near-degenerate becomes diverges from zero at increasing rates for larger values of $||\boldsymbol \theta||$.
 This shows that as model size gets larger, the risk of degeneracy starts at a slightly larger magnitude of parameters, but very quickly increases until reaching close to 1.
@@
 ```{r split_plots2, warning=FALSE, fig.cap="\\label{fig:split_plots_dist}The fraction of near-degenerate models for each magnitude of $\\boldsymbol \\theta$.
-For each number of visibles in the model, the fraction near-degenerate becomes greater than zero at larger values of $||\\boldsymbol \\theta||$ as $\\nh$ increases and the slope becomes steeper as $\\nh$ increases also.", fig.height=4, fig.align="ht!"}
+For each number $\nv$ of visibles in the model, the fraction near-degenerate becomes greater than zero at larger values of $||\\boldsymbol \\theta||$ as $\\nh$ increases and the slope becomes steeper as $\\nh$ increases as well.", fig.height=4, fig.align="ht!"}
 plot_dat_grid %>% 
@@
 
-Typically, fitting a RBM via maximum likelihood (ML) methods will be infeasible due mainly to the intractibility of the normalizing term $\gamma(\boldsymbol \theta)$ in a model of any realistic size.
-Ad hoc methods are used instead, which aim to avoid this problem by using stochastic ML that employ a small number of MCMC draws.
+Typically, fitting a RBM via maximum likelihood (ML) methods will be infeasible due mainly to the intractibility of the normalizing term $\gamma(\boldsymbol \theta)$ in a model (\ref{eqn:pmf}) of any realistic size.
+Ad hoc methods are used instead, which aim to avoid this problem by using stochastic ML approximations that employ a small number of MCMC draws (i.e., contrastive divergence, [@hinton2002training]).
 
@@
 In addition, a RBM model, with the appropriate choice of parameters and number of hiddens, has the potential to re-create any distribution for the data (i.e., reproduce any specification of cell probabilities for the $-1/1$ valued data outcomes).
-@montufar2011refinements show that any distribution on $\{0, 1\}^{\nv}$ can be approximated arbitrarily well by a RBM with $2^{\nv}/2 - 1$ hidden units.
+For example, @montufar2011refinements show that any distribution on $\{0, 1\}^{\nv}$ can be approximated arbitrarily well by a RBM with $2^{\nv}/2 - 1$ hidden units.
 To illustrate, we provide a small example.
@@
 
-To demonstrate such model approximations,  we  assume $p_{(-1,-1)}>0$ (without loss of generality) in the specified cell probabilities (\ref{eqn:0}) and replace parameters $\theta_{11},\theta_{21}$ with $\Delta_1 \equiv \theta_{11} +\theta_{21}$ and $\Delta_2 \equiv \theta_{11} -\theta_{21}$.
+To demonstrate, =we  assume $p_{(-1,-1)}>0$ (without loss of generality) in the specified cell probabilities (\ref{eqn:0}) and replace parameters $\theta_{11},\theta_{21}$ with $\Delta_1 \equiv \theta_{11} +\theta_{21}$ and $\Delta_2 \equiv \theta_{11} -\theta_{21}$.
 We may then prescribe values of $\theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \Delta_1,\Delta_2$  so that the model probability ratio
@@
 The previous discussion intends to indicate that the RBM model class suffers from parameter identifiability issues that go beyond mere symmetries in the parametrization.
-Quite different parameter settings can induce the same essential RBM model.
+Not only it is possible to approximate any distribution on the visibles arbitrarily well (cf.
+@montufar2011refinements), quite different parameter settings can induce the same essential RBM model.
 However, this is not the most disastrous implication of the RBM parameterization.
@@
 (\ref{eqn:2})) shows that RBM model parameters can be chosen to re-create this empirical distribution to an arbitrarily close degree.
-Hence, RBM model fitting based on ML will often simply seek to reproduce the empirical distribution (i.e., as an extreme type of over-fitting).
+Hence, RBM model fitting based on ML will simply seek to reproduce the empirical distribution.
 What's more, whenever this empirical distribution contains empty cells, fitting steps for the RBM model will necessarily aim to choose parameters that necessarily diverge to infinity in magnitude in order to zero-out the corresponding RBM cell probabilities.
 In data applications with a large sample space, it is unlikely that the training set will include at least one of each possible vector outcome (unlike this small example).
-As we have already discussed in Section \ref{explorations}, large magnitudes of $\boldsymbol \theta$ lead to model impropriety in the RBM.
+This implies that some RBM model parameters must diverge to $+\infty$ to mimic the empirical distribution with empty cells and, as we have already discussed in Section \ref{explorations}, large magnitudes of $\boldsymbol \theta$ lead to model impropriety in the RBM.
 
@@
 To avoid model impropriety for a fitted RBM, we want to avoid parts of the parameter space $\mathbb{R}^{\nv + \nh + \nv*\nh}$ that lead to near-degeneracy, instability, and uninterpretability.
-Motivated by the insights in Section \ref{exploring-manageable-examples}, one idea is to shrink $\boldsymbol \theta$ toward $\boldsymbol 0$ by specifying priors that place low probability on large values of $||\boldsymbol \theta||$, specifically focusing on shrinking $\boldsymbol \theta_{interaction}$ more than $\boldsymbol \theta_{main}$.
+Motivated by the insights in Section \ref{exploring-manageable-examples}, one idea is to shrink $\boldsymbol \theta = (\boldsymbol \theta_{main}, \boldsymbol \theta_{interaction})$ toward $\boldsymbol 0$ by specifying priors that place low probability on large values of $||\boldsymbol \theta||$, specifically focusing on shrinking $\boldsymbol \theta_{interaction}$ more than $\boldsymbol \theta_{main}$.
 This is similar to an idea advocated by @hinton2010practical called *weight decay*, in which a penalty is added to the interaction terms in the model, $\boldsymbol \theta_{interaction}$, shrinking their magnitudes.
@@
 2.
-*A truncated Normal prior.* Here we use independent truncated spherical normal distributions as priors for $\boldsymbol \theta_{main}$ and $\boldsymbol \theta_{interaction}$ with $\sigma_{interaction} < \sigma_{main}$.
+*A truncated Normal prior.* Here we use independent spherical normal distributions as priors for $\boldsymbol \theta_{main}$ and $\boldsymbol \theta_{interaction}$ with $\sigma_{interaction} < \sigma_{main}$, truncated at $3\sigma_{main}$ and $3\sigma_{interaction}$, respectively.
 Full conditional distributions are not conjugate, and simulation from the posterior was accomplished using a geometric adaptive Metropolis Hastings step [@zhou2014some] and calculation of likelihood normalizing constant.
@@
 All three methods require choosing the values of hyperparameters.
-In each case, we have chosen the values based on a rule of thumb that shrinks $\boldsymbol \theta_{interaction}$ more than $\boldsymbol \theta_{main}$.
+In each case, we have chosen these values based on a rule of thumb that shrinks $\boldsymbol \theta_{interaction}$ more than $\boldsymbol \theta_{main}$.
 Additionally, BwTPLV requires additional tuning to choose $C_1$ and $C_2$, reducing its appeal.
@@
 It should be noted that BwTNLV (2) and BwTNML (3) are drawing from the same stationary posterior distribution for vectors of visibles.
-The difference between these two methods is in how well the chains mix and how quickly they arrive at the target distribution.
+The difference between these two methods is in how well the chains mix and how quickly they arrive at the target posterior distribution.
 After a burn-in period of 50 iterations selected by inspecting the trace plots, we assess the issue of mixing in two ways.
@@
 Secondly, we can assess the mixing of our chains using an idea of effective sample size.
-If the MCMC chain were truly iid draws from the target distribution, then the estimate of the posterior mean probability of visibles vector $i$, $\bar{p}^{(i)}$ would be distributed approximately Normal with mean the true posterior marginal probability of visibles vector $i$, $p^{(i)}$, and variance $\sigma^2_i/M$, where $\sigma^2_i$ is the true posterior variance of visibles vector $i$ and $M$ is the length of the chain.
+If the MCMC chain were truly iid draws from the target distribution, then for the parameter $p^{(i)}$ denoting the probability of the $i$th vector outcome for the four visibles $\mathbf{v}=(v_1,v_2,v_3,v_4)\in\{\pm 1\}^4$, $i=1,\ldots,16$ (i.e., computed from $\theta$ under (\ref{eqn:pmf})), its estimate as the average $\bar{p}^{(i)}$ of posterior sample versions would be distributed approximately Normal with mean given by the true posterior marginal probability of $p^{(i)}$, and variance $\sigma^2_i/M$, where $\sigma^2_i$ is the true posterior variance of $p^{(i)}$ and $M$ is the length of the chain.
 However, with the presence of correlation in our chain, the aysmptotic variance of $\bar{p}^{(i)}$ is instead approximately some $C_i/M$, where $C_i$ is some positive constant such that $C_i > \sigma^2_i$.
-We can use an overlapping block-means approach [@gelman2011inference] to get a crude estimate for $C_i$, $\hat{C}_i = b S^2(\overline{p_i})$ where $b$ is the block length and $S^2(\overline{p_i})$ denotes the sample variance of the blockmeans $\overline{p}_i$ for visibles vector $i$.
-We compare it to an estimate of $\sigma^2_i$ using sample variance of the raw chain, $\hat{\sigma}^2_i$.
+We can use an overlapping block-means approach [@gelman2011inference] to get a crude estimate for $C_i$ as $\hat{C}_i = bS_b^2$, where $S_b^2$  denotes the sample variance of overlapping block means $\{\bar{p}_j^{(i)}=\sum_{k=j}^{j+b-1} p_k^{(i)}/b\}_{j=1}^{M-b+1}$ of length $b$ computed from the posterior samples $\{p_k^{(i)}\}_{k=1}^M$.
+We compare it to an estimate of $\sigma^2_i$ using sample variance $\hat{sigma}^2_i$ of the raw chain, $\{p_k^{(i)}\}_{k=1}^{M}$.
 Formally, we approximate the effective sample size of the length $M$ MCMC chain as
@@
   spread(Model, M_eff) %>%
-  mutate(Vector = as.numeric(Image)) %>%
-  select(Vector, BwTNML, BwTNLV) %>%
-  arrange(Vector) -> M_eff_tmp
+  mutate(Outcome = as.numeric(Image)) %>%
+  select(Outcome, BwTNML, BwTNLV) %>%
+  arrange(Outcome) -> M_eff_tmp
 
 cbind(M_eff_tmp[1:8,], M_eff_tmp[9:16,]) %>%
-  kable(digits = 2, caption = paste0("\\label{tab:m_eff}The effective sample sizes for a chain of length $M = 1000$ of all $16$ possible vectors of visibles.
+  kable(digits = 2, caption = paste0("\\label{tab:m_eff}The effective sample sizes for a chain of length $M = 1000$ of all $16$ possible vector outcomes of visibles.
 BwTNLV would require at least $", round(min(M_eff_table[1, -1])/min(M_eff_table[2, -1]), 1), "$ times as many MCMC iterations to achieve the same amount of effective information about the posterior distribution."))
@@
 Thus, BwTNLV would require at least $`r round(min(M_eff_table[1, -1])/min(M_eff_table[2, -1]), 1)`$ times more iterations to be run of the MCMC chain in order to achieve the same amount of effective information about the posterior distribution.
-For this reason, consistent with the ACF results in Figure \ref{fig:acf}, BwTNLV does not seem to be an effective method for fitting the RBM of computing resources are at all limited.
+For this reason, consistent with the ACF results in Figure \ref{fig:acf}, BwTNLV does not seem to be an effective method for fitting the RBM if computing resources are at all limited.
 
@@
 In this paper, we have presented three honest Bayes MCMC-based methods for fitting RBMs.
-Common practice is to use a kind of MCMC to overcome fitting complexities, however because of the size of the space to be filled with MCMC iterates convergence and mixing of these methods will be slow.
+Common practice is to use a kind of MCMC to overcome fitting complexities.
+However because of the size of the space to be filled with MCMC iterates convergence and mixing of these methods will be slow.
 Marginalization over the latent variables in the model can improve mixing, but is numerically intractible due to the necessity of repeated calculation of the normalizing constant.
@@
 
-\begin{appendix}
-\section{Appendices}
-\label{appendix}
-\newtheorem{apptheorem}{Theorem}[section]
-\newtheorem{applemma}{Lemma}[section]
-
-\subsection{On the one-to-one correspondence between parameters and moments in the RBM distribution}
-\label{center-of-universe}
-
-```{r child='appendix/appendix-center.Rmd'}
-```
-
-\end{appendix}
-
 #References

### 2016-11-20 writing/appendix/appendix-center.Rmd
more edits

@@
 ---
-title: "Supplement to: On the propriety of restricted Boltzmann machines"
+title: "Supplementary Material to:  On the propriety of restricted Boltzmann machines"
 author: "Andee Kaplan, Daniel Nordman, and Stephen Vardeman"
@@
 ---
+
+Supplement material to follow consists of Section \ref{on-the-one-to-one-correspondence-between-parameters-and-moments-in-the-rbm-distribution}, providing proofs and discussion of results relating to the data coding in restricted Boltzmann machine models (RBMs).
 

### 2016-11-20 writing/draft.Rmd
more edits

@@
 where $\boldsymbol \theta \equiv (\theta_{11}, \dots, \theta_{1\nh}, \dots, \theta_{\nv 1}, \dots, \theta_{\nv \nh}, \theta_{v_1}, \dots, \theta_{v_\nv}, \theta_{h_1}, \dots, \theta_{h_\nh}) \in \mathbb{R}^{\nv + \nh + \nv*\nh}$ denotes the vector of model parameters and the denominator $\gamma(\boldsymbol \theta) = \sum_{\boldsymbol x \in \mathcal{C}^{\nh+\nv}}\exp\left(\sum_{i = 1}^\nv \sum_{j=1}^\nh \theta_{ij} v_i h_j + \sum_{i = 1}^\nv\theta_{v_i} v_i + \sum_{j = 1}^\nh\theta_{h_j} h_j\right)$ is the normalizing function that ensures the probabilities (\ref{eqn:pmf}) sum to one.
-For
+For $\boldsymbol x = (h_1, \dots, h_\nh, v_1, \dots, v_\nh) \in \mathcal{C}^{\nv + \nh}$ and
 \begin{align}
-\boldsymbol x &= (h_1, \dots, h_\nh, v_1, \dots, v_\nh) \in \mathcal{C}^{\nv + \nh} \text{ and} \nonumber\\
 \boldsymbol t(\boldsymbol x) &= (h_1, \dots, h_\nh, v_1, \dots, v_\nv, v_1 h_1, \dots, v_\nv h_\nh) \in \mathcal{C}^{\nh + \nv + \nh*\nv}, \label{eqn:t}
@@
 A RBM model can be thought to exhibit an analogous form of *near-degeneracy* when there is a disproportionate amount of probability placed on a small number of elements in the sample space of visibles and hiddens, $\mathcal{C}^{\nv + \nh}$.
-Using the idea of @handcock2003assessing, when the random vector $\boldsymbol t(\boldsymbol x) = (v_1, \dots, v_\nv, h_1, \dots, h_\nh, v_1 h_1, \dots, v_V h_\nh ) \in \mathcal{T} \equiv \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{C}^{\nh + \nv} \}$ from (\ref{eqn:t}), appearing in the neg-potential function $Q_{\boldsymbol \theta}(\cdot)$, has a mean vector $\boldsymbol \mu(\boldsymbol \theta) \in \mathbb{R}^{\nv + \nh + \nv * \nh}$ close to the boundary of the convex hull of $\mathcal{T}$, then the RBM model can be said to exhibit near-degeneracy at $\boldsymbol \theta \in \mathbb{R}^{\nv + \nh + \nh*\nv}$.
+Using the idea of @handcock2003assessing, when the random vector $\boldsymbol t(\boldsymbol x) = (v_1, \dots, v_\nv, h_1, \dots, h_\nh, v_1 h_1, \dots, v_V h_\nh ) \in \mathcal{T} \equiv \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{C}^{\nh + \nv} \}$ appearing in the neg-potential function $Q_{\boldsymbol \theta}(\cdot)$, has a mean vector $\boldsymbol \mu(\boldsymbol \theta) \in \mathbb{R}^{\nv + \nh + \nv * \nh}$ close to the boundary of the convex hull of $\mathcal{T}$, then the RBM model can be said to exhibit near-degeneracy at $\boldsymbol \theta \in \mathbb{R}^{\nv + \nh + \nh*\nv}$.
 Here the mean of $\boldsymbol t(\boldsymbol x)$ is
@@
 $$
-P_\theta(V_i=1) \propto e^{\theta_{v_i}} \quad \& \quad P_\theta(H_j=1) \propto e^{\theta_{h_j}}, \quad  i=1,\ldots,\nv,j=1,\ldots,\nh
-$$
-so that, for example, $logit(P_\theta(V_i=1)=\theta_{v_i}$ (or $2 \theta_{v_i}$) under the coding $\mathcal{C}=\{0,1\}$ (or $\{-1,1\}$).
+P_\theta(V_i=1) \propto e^{\theta_{v_i}} \quad \text{ and } \quad P_\theta(H_j=1) \propto e^{\theta_{h_j}}, \quad  i=1,\ldots,\nv,j=1,\ldots,\nh
+$$
+so that, for example, $logit(P_\theta(V_i=1))=\theta_{v_i}$ (or $2 \theta_{v_i}$) under the coding $\mathcal{C}=\{0,1\}$ (or $\{-1,1\}$).
 Hence, these main effect parameters have a clear interpretation under an independence model (one with $\theta_{ij} = 0$) but this interpretation can break down as interaction parameters increase in magnitude relative to the size of the main effects.
@@
 
-It should be noted that BwTNLV (2) and BwTNML (3) are drawing from the same stationary posterior distribution for vectors of visibles.
+It should be noted that BwTNLV (2.) and BwTNML (3.) are drawing from the same stationary posterior distribution for vectors of visibles.
 The difference between these two methods is in how well the chains mix and how quickly they arrive at the target posterior distribution.
@@
 Secondly, we can assess the mixing of our chains using an idea of effective sample size.
-If the MCMC chain were truly iid draws from the target distribution, then for the parameter $p^{(i)}$ denoting the probability of the $i$th vector outcome for the four visibles $\mathbf{v}=(v_1,v_2,v_3,v_4)\in\{\pm 1\}^4$, $i=1,\ldots,16$ (i.e., computed from $\theta$ under (\ref{eqn:pmf})), its estimate as the average $\bar{p}^{(i)}$ of posterior sample versions would be distributed approximately Normal with mean given by the true posterior marginal probability of $p^{(i)}$, and variance $\sigma^2_i/M$, where $\sigma^2_i$ is the true posterior variance of $p^{(i)}$ and $M$ is the length of the chain.
+If the MCMC chain were truly iid draws from the target distribution, then for the parameter $p^{(i)}$ denoting the probability of the $i$th vector outcome for the four visibles $\mathbf{v}=(v_1,v_2,v_3,v_4)\in\{\pm 1\}^4$, $i=1,\ldots,16$ (i.e., computed from $\boldsymbol \theta$ under (\ref{eqn:pmf})), its estimate as the average $\bar{p}^{(i)}$ of posterior sample versions would be distributed approximately Normal with mean given by the true posterior marginal probability of $p^{(i)}$, and variance $\sigma^2_i/M$, where $\sigma^2_i$ is the true posterior variance of $p^{(i)}$ and $M$ is the length of the chain.
 However, with the presence of correlation in our chain, the aysmptotic variance of $\bar{p}^{(i)}$ is instead approximately some $C_i/M$, where $C_i$ is some positive constant such that $C_i > \sigma^2_i$.
 We can use an overlapping block-means approach [@gelman2011inference] to get a crude estimate for $C_i$ as $\hat{C}_i = bS_b^2$, where $S_b^2$  denotes the sample variance of overlapping block means $\{\bar{p}_j^{(i)}=\sum_{k=j}^{j+b-1} p_k^{(i)}/b\}_{j=1}^{M-b+1}$ of length $b$ computed from the posterior samples $\{p_k^{(i)}\}_{k=1}^M$.
-We compare it to an estimate of $\sigma^2_i$ using sample variance $\hat{sigma}^2_i$ of the raw chain, $\{p_k^{(i)}\}_{k=1}^{M}$.
+We compare it to an estimate of $\sigma^2_i$ using sample variance $\hat{\sigma}^2_i$ of the raw chain, $\{p_k^{(i)}\}_{k=1}^{M}$.
 Formally, we approximate the effective sample size of the length $M$ MCMC chain as

### 2016-11-20 writing/note.Rmd
note edits intro and examples

@@
 author: "Andee Kaplan, Daniel Nordman, and Stephen Vardeman"
-date: "October, 2016"
+date: "November, 2016"
 linestretch: 2
@@
 abstract: |
-  A probability model exhibits instability if small changes in a data outcome result in large, perhaps unanticipated changes in probability.
-As data structures become larger and more complex, there is increasing interest in identifying instability in potential model specifications for such data (e.g., models for random graphs or networks).
+  A probability model exhibits instability if small changes in a data outcome result in large, and often unanticipated, changes in probability.
+For correlaated data structures found in several application areas, there is increasing interest in identifying instability in potential model specifications for such data.
 We consider the problem of quantifying instability for general probability models defined on sequences of observations, where each sequence of a given length $N$ has a finite number of possible outcomes.
@@
 Also, as the measure of instability becomes more extreme, the resulting probability models are shown  to slip into degeneracy, thereby placing all probability on arbitrarily small portions of the sample space.
-These results on instability apply to large classes of exponential models commonly used in random graphs, network analysis, and machine learning contexts.
+These results on instability apply to large classes of models commonly used in random graphs, network analysis, and machine learning contexts.
 ---
@@
 
-A probability model exhibits instability if small changes in a data outcome result in large, perhaps unanticipated changes in probability.
 As data structures become larger and more complex, there is increasing interest in identifying instability in potential model specifications for such data (e.g., models for random graphs or networks).
-The concept of instability, introduced in the field statistical physics by @ruelle1999statistical, was extended to include a notion of detection and quantification by @schweinberger2011instability.
 We consider the problem of quantifying instability for general probability models defined on sequences of observations, where each sequence of a given length $N$ has a finite number of possible outcomes.
@@
 Also, as the measure of instability becomes more extreme, the resulting probability models are shown to slip into degeneracy, thereby placing all probability on arbitrarily small portions of the sample space.
-This relationship between instability and degeneracy, as defined by @handcock2003assessing, was described by @schweinberger2011instability.
-We extend these results on instability to large classes of exponential models commonly used in random graphs, network analysis, and machine learning contexts.
 
-In this paper, we consider the behavior of probability models built to incorporate a sequence of discrete observations with an arbitrarily specified length $N$.
+We consider the behavior, and the potential impropriety, of probability models built to incorporate a sequence of discrete observations with an arbitrarily specified length $N$.
 Let $(X_1, \dots X_N)$ denote a set of sequence of discrete random variables with a finite sample space, $\mathcal{X}^N$.
 That is, $\mathcal{X}$ represents a finite set of potential outcomes for the variable $X_i$, and the data sequence $(X_1,\ldots,X_N)$ assumes values in the $N$-fold product space $\mathcal{X}^N$.
-For any $N$, let $P_{\theta_N}$ denote a probability model on $\mathcal{X}^N$, under which $P_{\boldsymbol \theta_N}(x_1, \dots, x_N) > 0$ is the probability of the data outcome $(x_1, \dots, x_N) \in \mathcal{X}^N$.
+For any $N$, let $P_{\boldsymbol \theta_N}$ denote a probability model on $\mathcal{X}^N$, under which $P_{\boldsymbol \theta_N}(x_1, \dots, x_N) > 0$ is the probability of the data outcome $(x_1, \dots, x_N) \in \mathcal{X}^N$.
 We assume that the model support of $P_{\boldsymbol \theta_N}$ is equal to the sample space $\mathcal{X}^N$.
-This framework produces a series $P_{\theta_N}$ of probability models, indexed by a generic sequence of parameters $\theta_N$, to describe data of any given length $N \geq 1$.
-The size and structure of such parameters may be generally specified, and one may intuitvely consider Euclidean $\theta_N \in \mathbb{R}^{q(N)}$ for some arbitrary integer-valued function $q(\cdot) \geq 1$.
+This framework produces a series $P_{\boldsymbol \theta_N}$ of probability models, indexed by a generic sequence of parameters $\boldsymbol \theta_N$, to describe data of any given length $N \geq 1$.
+The size and structure of such parameters may be generally specified, and one may intuitvely consider Euclidean $\boldsymbol \theta_N \in \mathbb{R}^{q(N)}$ for some arbitrary integer-valued function $q(\cdot) \geq 1$.
 We will call this class *Finitely Supported Finite Sequence (FSFS) models*.
 
-Section \ref{examples} provides several examples of FSFS models commonly used in graph/network analysis and machine learning (i.e., deep learning models) among other application areas.
-Section \ref{instability-results} establishes formal results regarding the model propriety of FSFS models.
-These results help to measure, and qualify the effects of, model instability that can arise when certain maximal probabilities in such models are too extreme relevant to the length $N$ of the data sample.
-Lastly, Section \ref{implications} emphasizes the implications of our model propriety results.
+Section \ref{examples} provides several examples of FSFS models commonly used in graph/network analysis and machine learning (i.e., deep learning models).
+Section \ref{instability-results} establishes formal results regarding the model propriety of FSFS models with regard to stability.
+A FSFS probability model intuitively exhibits instability if small changes in the components of a data outcome $(x_1,\ldots,x_N)$ may result in large in probability $P_{\theta_N}(x_1,\ldots,x_N)$.
+The concept of instability, introduced in the field statistical physics by @ruelle1999statistical, was extended to include a notion of detection and quantification for certain exponential
+family models by @schweinberger2011instability.
+For similar exponential models, particularly in connection to random graphs/networks, @handcock2003assessing considered (mean-based) characterizations for so-called model degeneracy, whereby a probability model places all mass on a small subset of the sample space and produces undesirably low variability in model outcomes.
+As described by @schweinberger2011instability, model instability and model degeneracy are related by viewing degeneracy as an extreme or limiting form of instability.
+However, the instability results of @schweinberger2011instability are only intended for a special parametric case of exponential family models.
+In considering the broad framework of FSFS models, the main results here develop a more general measure of model instability, which is defined to arise when certain maximal probabilities in FSFS models are too extreme relative to the length $N$ of the data sample.
+In this case, a one component change in the data sequence
+may provably shift probability by orders of magnitude, and FSFS models are rigorously shown to become degenerate as the measure of instability increases.
+Lastly, Section \ref{implications} emphasizes the implications of our model propriety results and proofs of the main results appear in the Appendix.
 
@@
 
-Note many model families can fall under the umbrella of FSFS models. In this section, we present three specific examples of FSFS models, including models with deep architectures.
+Many model families can fall under the umbrella of FSFS models. For illustration, this section presents three specific examples of FSFS models, including models with deep architectures.
 
@@
 
-Let $X_1, \dots, X_N$ be discrete random variables with sample space $\mathcal{X}^N$, where $|\mathcal{X}| < \infty$.
-Consider discrete exponential family model with probability mass function of the form
+For $X_1, \dots, X_N$ discrete random variables with sample space $\mathcal{X}^N$, $|\mathcal{X}| < \infty$, consider an exponential family model with probability mass function of the form
 $$
@@
 $$ 
-where $\boldsymbol \lambda \in \Lambda \subset \mathbb{R}^{q(N)}$, $\boldsymbol \eta : \mathbb{R}^q \mapsto \mathbb{R}^L$ is a vector of natural parameters, $\boldsymbol g : \mathcal{X}^N \mapsto \mathbb{R}^L$ is a vector of sufficient statistics,
+where $\boldsymbol \lambda \in \Lambda \subset \mathbb{R}^{k}$, $\boldsymbol \eta : \mathbb{R}^k \mapsto \mathbb{R}^L$ is a vector of natural parameters, $\boldsymbol g : \mathcal{X}^N \mapsto \mathbb{R}^L$ is a vector of sufficient statistics, positive integers $k$ and $L$ denote the dimensions of the parameter and natural parameters spaces, respectfully,
 $$
@@
 
-By above definition, $X_1, \dots, X_N$ are discrete with a finite sample space.
-Let $\boldsymbol \theta = (\lambda_1, \dots, \lambda_k, 0, \dots, 0)$, where the first $k$ elements are equal to the $k$ elements of $\boldsymbol \lambda$ and for $i = k + 1, \dots, q(N)$, $\theta_i = 0$.
-Further, because $p_{\boldsymbol \lambda}(\boldsymbol x) > 0$ for all $\boldsymbol x \in \mathcal{X}^N$ by properties of the exponential function, the discrete exponential family models defined in this section are a special case of the FSFS model.
+Setting $\boldsymbol \theta_N=\lambda_N$ to a subsequence of $\mathbb{R}^k$ and noting that $p_{\boldsymbol \lambda}(\boldsymbol x) > 0$ for all $\boldsymbol x \in \mathcal{X}^N$ by properties of the exponential function, the discrete exponential family models defined in this section are a special case of the FSFS model.
 Such exponential models arise in spatial data on a lattice [@besag1974spatial], network data [@wasserman1994social, @handcock2003assessing], and even in binomial sampling with $N$ iid Bernoulli random variables.
+@schweinberger2011instability considered instability in such exponential models (e.g., for random graphs) for fixed parameter sequences $\theta_N=\lambda\in\mathbb{R}^k$.
 
@@
 A restricted Boltzmann machine (RBM) is an undirected graphical model specified for discrete or continuous random variables, binary variables being most common [@smolensky1986information].
-Consider the binary case of an RBM only.
 A RBM architecture has two layers, hidden ($\mathcal{H}$) and visible ($\mathcal{V}$), with conditional independence within each layer.
-Let $\boldsymbol X = \{H_1, \dots, H_H, V_1,\dots, V_V\}$ be random variables corresponding to the hidden and visible variables in an RBM.
-We will consider the sample space $\mathcal{X} = \{-1,1\}$.
-The RBM is then has the joint probability mass function
+Let $\boldsymbol X=(X_1,\ldots,X_N)$ denote the $N$ random variables for visibles with support $\mathcal{X}^N$ and $\boldsymbol H=(H_1,\ldots,H_{N_H})$ denote the $N_H$ random variables for hiddens with support $\mathcal{X}^{N_H}$ where  $\mathcal{X} = \{-1,1\}$.
+The RBM for $\tilde{\boldsymbol X}=(\boldsymbol X,\boldsymbol H})$ then has the joint probability mass function
 $$
-p_{\boldsymbol \theta} (\boldsymbol x) = \exp\left[ \boldsymbol \alpha^T \boldsymbol h + \boldsymbol \beta^T \boldsymbol v + \boldsymbol h^T \Gamma \boldsymbol v - \psi(\boldsymbol \theta)\right], \qquad \boldsymbol x = (\boldsymbol h, \boldsymbol v) \in \mathcal{X}^{H+V}
+p_{\boldsymbol \theta} (\tilde{\boldsymbol x}) = \exp\left[ \boldsymbol \alpha^T \boldsymbol h + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^T \Gamma \boldsymbol x - \psi(\boldsymbol \theta)\right], \qquad \tilde{\boldsymbol x} = (\boldsymbol h, \boldsymbol x) \in \mathcal{X}^{N+N_H}
 $$
@@
 $$
-\psi(\boldsymbol \theta) = \log \sum_{\boldsymbol x \in \mathcal{X}^{H + V}} \exp\left[ \boldsymbol \alpha^T \boldsymbol h + \boldsymbol \beta^T \boldsymbol v + \boldsymbol h^T \Gamma \boldsymbol v\right], \qquad \boldsymbol \theta \in \Theta
+\psi(\boldsymbol \theta) = \log \sum_{\tilde{\boldsymbol x} \in \mathcal{X}^{N + N_H}} \exp\left[ \boldsymbol \alpha^T \boldsymbol h + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^T \Gamma \boldsymbol x\right], \qquad \boldsymbol \theta \in \Theta
 $$ 
-is the normalizing function, $\boldsymbol \alpha \in \mathbb{R}^H$, $\boldsymbol \beta \in \mathbb{R}^V$, and $\Gamma$ is a matrix with dimension $H \times V$ are the parameters in the model.
-Let $\boldsymbol \theta = (\boldsymbol \alpha, \boldsymbol \beta,\Gamma) \in \Theta \subset \mathbb{R}^{H + V + H*V}$ denote the combined vector of parameters.
+is the normalizing function, $\boldsymbol \alpha \in \mathbb{R}^{N_H}$, $\boldsymbol \beta \in \mathbb{R}^N$, and $\Gamma$ is a matrix with dimension $N_H \times N$ are the parameters in the model.
+Let $\boldsymbol \theta_N = (\boldsymbol \alpha, \boldsymbol \beta,\Gamma) \in \Theta \subset \mathbb{R}^{q(N)}$ with $q(N) = N + N_H + N*N_H$ denote the vector of parameters.
 
-The probability mass function for the visible random variables $V_1, \dots, V_V$ follows from this joint specification:
+The probability mass function for the visible variables $X_1, \dots, X_N$ follows from marginalizing this joint specification:
 $$
-p_{\boldsymbol \theta} (\boldsymbol v) = \sum\limits_{\boldsymbol h \in \mathcal{X}^H} p_{\boldsymbol \theta} (\boldsymbol x), \qquad \boldsymbol v \in \mathcal{X}^V
+p_{\boldsymbol \theta} (\boldsymbol x) = \sum\limits_{\boldsymbol h \in \mathcal{X}^{N_H} p_{\boldsymbol \theta} (\boldsymbol x, \boldsymbol h), \qquad \boldsymbol x \in \mathcal{X}^N
 $$
 
-By definition, $V_1, \dots, V_V$ are discrete with a finite sample space.
-The vector of model parameters $\boldsymbol \theta$, of size $q(V) \equiv V + H + H*V$, is expanding in size as a function of the dimension of $V_1, \dots, V_V$.
-Additionally, $p_{\boldsymbol \theta}(\boldsymbol v) > 0$ for all $\boldsymbol v \in \mathcal{X}^V$ by properties of the exponential function.
-Thus, the visible RBM model specification is an example of a FSFS model.
+Note that the vector of model parameters $\boldsymbol \theta_N$, of size $q(N)$, is expanding in size as a function of the visible variable dimension $N$, and one may choose the number $N_H$ of hidden variables as an arbitrary function of $N$ as well (which may also potentially increase with $N$).
+Additionally, $p_{\boldsymbol \theta}(\boldsymbol x) > 0$ for all $\boldsymbol x \in \mathcal{X}^N$ by properties of the exponential function so that the visible RBM model specification is an example of a FSFS model.
 This example also indicates that models formed by marginalizing a base FSFS model (e.g., exponential family) is again a FSFS model class.
@@
 Consider two models with "deep architechture" that contain multiple hidden (or latent) layers in addition to a visible layer of data, a deep Boltzmann machine [@salakhutdinov2009deep] and a deep belief network [@hinton2006fast].
-Let $M$ denote the number of hidden layers included in the model and $H_1, \dots, H_M$ the number of hidden variables within each hidden layer.
-Then let $\boldsymbol X = \{H^{(1)}_1, \dots, H^{(1)}_{H_1}, \dots, H^{(M)}_1, \dots, H^{(M)}_{H_M}, V_1,\dots, V_V\}$ be random variables corresponding to the hidden and visible variables in a deep probabilitic model.
+Let $M$ denote the number of hidden layers included in the model and $N_{H,1}, \dots, N_{H, M}$ the number of hidden variables within each hidden layer.
+Then let $\tilde{\boldsymbol X} = \{H^{(1)}_1, \dots, H^{(1)}_{N_{H,1}}, \dots, H^{(M)}_1, \dots, H^{(M)}_{N_{H,M}}, \boldsymbol X\}$ be random variables corresponding to the hidden and visible $\boldsymbol X =(X_1,\ldots,X_N)$ variables in a deep probabilitic model.
 We will consider the sample space $\mathcal{X} = \{-1,1\}$.
@@
 $$
-p_{\boldsymbol \theta} (\boldsymbol x) = \exp\left[ \sum\limits_{i = 1}^M\boldsymbol \alpha^{(i)T} \boldsymbol h^{(i)} + \boldsymbol \beta^T \boldsymbol v + \boldsymbol h^{(1)T} \Gamma^{(0)} \boldsymbol v + \sum\limits_{i = 1}^{M - 1} \boldsymbol h^{(i)T} \Gamma^{(i)} \boldsymbol h^{(i + 1)} - \psi(\boldsymbol \theta) \right],
+p_{\boldsymbol \theta} (\tilde{\boldsymbol x}) = \exp\left[ \sum\limits_{i = 1}^M\boldsymbol \alpha^{(i)T} \boldsymbol h^{(i)} + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^{(1)T} \Gamma^{(0)} \boldsymbol x + \sum\limits_{i = 1}^{M - 1} \boldsymbol h^{(i)T} \Gamma^{(i)} \boldsymbol h^{(i + 1)} - \psi(\boldsymbol \theta) \right],
 $$
-where $\boldsymbol x = (\boldsymbol h^{(1)}, \dots, \boldsymbol h^{(M)}, \boldsymbol v) \in \mathcal{X}^{H_1 + \cdots + H_M +V}$ and
+where $\tilde{\boldsymbol x} = (\boldsymbol h^{(1)}, \dots, \boldsymbol h^{(M)}, \boldsymbol x) \in \mathcal{X}^{N_{H,1} + \cdots + N_{H,M} +N}$ and
 $$
-\psi(\boldsymbol \theta) = \log \sum_{\boldsymbol x \in \mathcal{X}^{H_1 + \cdots + H_M + V}} \exp\left[ \sum\limits_{i = 1}^M\boldsymbol \alpha^{(i)T} \boldsymbol h^{(i)} + \boldsymbol \beta^T \boldsymbol v + \boldsymbol h^{(1)T} \Gamma^{(0)} \boldsymbol v + \sum\limits_{i = 1}^{M - 1} \boldsymbol h^{(i)T} \Gamma^{(i)} \boldsymbol h^{(i + 1)}\right], \qquad \boldsymbol \theta \in \Theta
+\psi(\boldsymbol \theta) = \log \sum\limits_{\tilde{\boldsymbol x} \in \mathcal{X}^{N_{H,1} + \cdots + N_{H,M} +N}} \exp\left[ \sum\limits_{i = 1}^M\boldsymbol \alpha^{(i)T} \boldsymbol h^{(i)} + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^{(1)T} \Gamma^{(0)} \boldsymbol x + \sum\limits_{i = 1}^{M - 1} \boldsymbol h^{(i)T} \Gamma^{(i)} \boldsymbol h^{(i + 1)}\right], \qquad \boldsymbol \theta \in \Theta
 $$ 
-is the normalizing function, $\boldsymbol \beta \in \mathbb{R}^V$, $\boldsymbol \alpha^{(i)} \in \mathbb{R}^{H_i}$ for $i = 1, \dots, M$, $\Gamma^{(0)}$ is a matrix with dimension $H_1 \times V$, and $\Gamma^{(i)}$ is a matrix with dimension $H_i \times H_{i + 1}$ for $i = 1, \dots, M-1$ are the parameters in the model.
-Let $\boldsymbol \theta = (\boldsymbol \alpha^{(1)}, \dots, \boldsymbol \alpha^{(M)}, \boldsymbol \beta,\Gamma^{(0)}, \dots, \Gamma^{(M - 1)}) \in \Theta \subset \mathbb{R}^{H_1 + \cdots + H_M + V + H_1*V + H_1 *H_{2} + \cdots + H_{M-1}*H_M}$ denote the combined vector of parameters.
+is the normalizing function, $\boldsymbol \beta \in \mathbb{R}^V$, $\boldsymbol \alpha^{(i)} \in \mathbb{R}^{N_{H,i}}$ for $i = 1, \dots, M$, $\Gamma^{(0)}$ is a matrix with dimension $N_{H,1} \times N$, and $\Gamma^{(i)}$ is a matrix with dimension $N_{H,i} \times N_{H,i+1}$ for $i = 1, \dots, M-1$ are the parameters in the model.
+Let $\boldsymbol \theta = (\boldsymbol \alpha^{(1)}, \dots, \boldsymbol \alpha^{(M)}, \boldsymbol \beta,\Gamma^{(0)}, \dots, \Gamma^{(M - 1)}) \in \Theta \subset \mathbb{R}^{q(N)}$ denote the combined vector of parameters with $q(N)= N_{H,1}+\cdots N_{M,H} + N + N_{H,1}*N+N_{H,2}*H_{H,1}+\cdots +N_{H,M}*H_{H,M-1}$.
 
-The probability mass function for the visible random variables $V_1, \dots, V_V$ follows from this joint specification:
+The probability mass function for the visible random variables $X_1, \dots, X_N$ follows from this joint specification:
 $$
-p_{\boldsymbol \theta} (\boldsymbol v) = \sum\limits_{(\boldsymbol h^{(1)}, \dots, \boldsymbol h^{(M)}) \in \mathcal{X}^{H_1 + \cdots + H_M}} p_{\boldsymbol \theta} (\boldsymbol x), \qquad \boldsymbol v \in \mathcal{X}^V
+p_{\boldsymbol \theta} (\boldsymbol x) = \sum\limits_{(\boldsymbol h^{(1)}, \dots, \boldsymbol h^{(M)}) \in \mathcal{X}^{N_{H,1} + \cdots + N_{H,M}}} p_{\boldsymbol \theta} (\tilde{\boldsymbol x}), \qquad \boldsymbol x \in \mathcal{X}^N
 $$
 
-By definition, $V_1, \dots, V_V$ are discrete with a finite sample space.
-The vector of model parameters $\boldsymbol \theta$, of size $q(V) \equiv H_1 + \cdots + H_M + V + H_1*V + H_1 *H_{2} + \cdots + H_{M-1}*H_M$, is expanding in size as a function of the dimension of $V_1, \dots, V_V$.
-Additionally, $p_{\boldsymbol \theta}(\boldsymbol v) > 0$ for all $\boldsymbol v \in \mathcal{X}^V$ by properties of the exponential function.
-Thus, the visible RBM model specification is an example of a FSFS model.
+Again similar to the RBM case, the visible DBM model specification is an example of a FSFS model.
 
@@
 The difference between the two models is that all but the last stacked model are Bayesian networks, rather than RBMs.
-Thus for $V_1, \dots, V_V$ discrete with finite sample space and support equal to the sample space $\mathcal{X}^V$, a DBN is also a FSFS model if the number of components in the parameter vector is dependent on the dimension of the data, $V$.
-Commonly, as in a logistic belief net [@neal1992connectionist], a "weight" parameter is placed on each interaction between visibles, $V_1, \dots, V_V$ and the first layer of latent variables, $H^{(1)}_1, \dots, H^{(1)}_{H_1}$, satisfying the definition of a FSFS model.
+Thus for visibles $X_1, \dots, X_N$ with support equal to the sample space $\mathcal{X}^V$, a DBN is also a FSFS model if the number $q(N)$ of components in the parameter vector is dependent on the dimension of the visibles.
+Commonly, as in a logistic belief net [@neal1992connectionist], a "weight" parameter is placed on each interaction between visibles, $X_1, \dots, X_N$ and the first layer of latent variables, $H^{(1)}_1, \dots, H^{(1)}_{N_{H,1}}$, satisfying the definition of a FSFS model.
 

### 2016-11-20 writing/appendix/appendix-note.Rmd
remaining edits

--- 2016-11-17 writing/note.Rmd
@@
----
-title: "A note on the instability and degeneracy of deep learning models"
-author: "Andee Kaplan, Daniel Nordman, and Stephen Vardeman"
-date: "October, 2016"
-linestretch: 2
-output: 
-  pdf_document:
-    fig_caption: yes
-    includes:
-      in_header: latex/header.tex
-    number_sections: yes
-bibliography: references/refs.bib
-abstract: |
-  A probability model exhibits instability if small changes in a data outcome result in large, perhaps unanticipated changes in probability.
-As data structures become larger and more complex, there is increasing interest in identifying instability in potential model specifications for such data (e.g., models for random graphs or networks).
-We consider the problem of quantifying instability for general probability models defined on sequences of observations, where each sequence of a given length $N$ has a finite number of possible outcomes.
-As the number of available observations increases, a sequence of probability models results to accommodate the data of expanding dimension.
-  Model instability is formally shown to occur when a certain log-probability ratio under such models grows faster than the length $N$ of the data sequence.
-In this case, a one component change in the data sequence can provably shift probability by orders of magnitude.
-Also, as the measure of instability becomes more extreme, the resulting probability models are shown  to slip into degeneracy, thereby placing all probability on arbitrarily small portions of the sample space.
-These results on instability apply to large classes of exponential models commonly used in random graphs, network analysis, and machine learning contexts.
----
-
-# Introduction
-
-A probability model exhibits instability if small changes in a data outcome result in large, perhaps unanticipated changes in probability.
-As data structures become larger and more complex, there is increasing interest in identifying instability in potential model specifications for such data (e.g., models for random graphs or networks).
-The concept of instability, introduced in the field statistical physics by @ruelle1999statistical, was extended to include a notion of detection and quantification by @schweinberger2011instability.
-We consider the problem of quantifying instability for general probability models defined on sequences of observations, where each sequence of a given length $N$ has a finite number of possible outcomes.
-As the number of available observations increases, a sequence of probability models results to accommodate the data of expanding dimension.
-
-Model instability is formally shown to occur when a certain log-probability ratio under such models grows faster than the length $N$ of the data sequence.
-In this case, a one component change in the data sequence can provably shift probability by orders of magnitude.
-Also, as the measure of instability becomes more extreme, the resulting probability models are shown to slip into degeneracy, thereby placing all probability on arbitrarily small portions of the sample space.
-This relationship between instability and degeneracy, as defined by @handcock2003assessing, was described by @schweinberger2011instability.
-We extend these results on instability to large classes of exponential models commonly used in random graphs, network analysis, and machine learning contexts.
-
-In this paper, we consider the behavior of probability models built to incorporate a sequence of discrete observations with an arbitrarily specified length $N$.
-Let $(X_1, \dots X_N)$ denote a set of sequence of discrete random variables with a finite sample space, $\mathcal{X}^N$.
-That is, $\mathcal{X}$ represents a finite set of potential outcomes for the variable $X_i$, and the data sequence $(X_1,\ldots,X_N)$ assumes values in the $N$-fold product space $\mathcal{X}^N$.
-For any $N$, let $P_{\theta_N}$ denote a probability model on $\mathcal{X}^N$, under which $P_{\boldsymbol \theta_N}(x_1, \dots, x_N) > 0$ is the probability of the data outcome $(x_1, \dots, x_N) \in \mathcal{X}^N$.
-We assume that the model support of $P_{\boldsymbol \theta_N}$ is equal to the sample space $\mathcal{X}^N$.
-This framework produces a series $P_{\theta_N}$ of probability models, indexed by a generic sequence of parameters $\theta_N$, to describe data of any given length $N \geq 1$.
-The size and structure of such parameters may be generally specified, and one may intuitvely consider Euclidean $\theta_N \in \mathbb{R}^{q(N)}$ for some arbitrary integer-valued function $q(\cdot) \geq 1$.
-We will call this class *Finitely Supported Finite Sequence (FSFS) models*.
-
-Section \ref{examples} provides several examples of FSFS models commonly used in graph/network analysis and machine learning (i.e., deep learning models) among other application areas.
-Section \ref{instability-results} establishes formal results regarding the model propriety of FSFS models.
-These results help to measure, and qualify the effects of, model instability that can arise when certain maximal probabilities in such models are too extreme relevant to the length $N$ of the data sample.
-Lastly, Section \ref{implications} emphasizes the implications of our model propriety results.
-
-# Examples
-
-Note many model families can fall under the umbrella of FSFS models. In this section, we present three specific examples of FSFS models, including models with deep architectures.
-
-## Discrete exponential family models
-
-Let $X_1, \dots, X_N$ be discrete random variables with sample space $\mathcal{X}^N$, where $|\mathcal{X}| < \infty$.
-Consider discrete exponential family model with probability mass function of the form
-$$
-p_{\boldsymbol \lambda}(\boldsymbol x) = \exp\left[\boldsymbol\eta^T(\boldsymbol \lambda) \boldsymbol g(\boldsymbol x) - \psi(\boldsymbol \lambda)\right], \qquad \boldsymbol x \in \mathcal{X}^N.
-$$ 
-where $\boldsymbol \lambda \in \Lambda \subset \mathbb{R}^{q(N)}$, $\boldsymbol \eta : \mathbb{R}^q \mapsto \mathbb{R}^L$ is a vector of natural parameters, $\boldsymbol g : \mathcal{X}^N \mapsto \mathbb{R}^L$ is a vector of sufficient statistics,
-$$
-\psi(\boldsymbol \lambda) = \log \sum\limits_{\boldsymbol x \in \mathcal{X}^N}\exp\left[\boldsymbol \eta^T(\boldsymbol \lambda) \boldsymbol g(\boldsymbol x) \right], \qquad \boldsymbol \lambda \in \Lambda
-$$
-is the normalizing function, and $\Lambda = \{\boldsymbol \lambda \in \mathbb{R}^k : \psi(\boldsymbol \lambda) < \infty, k \le q(N) \}$ is the parameter space.
-
-By above definition, $X_1, \dots, X_N$ are discrete with a finite sample space.
-Let $\boldsymbol \theta = (\lambda_1, \dots, \lambda_k, 0, \dots, 0)$, where the first $k$ elements are equal to the $k$ elements of $\boldsymbol \lambda$ and for $i = k + 1, \dots, q(N)$, $\theta_i = 0$.
-Further, because $p_{\boldsymbol \lambda}(\boldsymbol x) > 0$ for all $\boldsymbol x \in \mathcal{X}^N$ by properties of the exponential function, the discrete exponential family models defined in this section are a special case of the FSFS model.
-Such exponential models arise in spatial data on a lattice [@besag1974spatial], network data [@wasserman1994social, @handcock2003assessing], and even in binomial sampling with $N$ iid Bernoulli random variables.
-
-## Restricted Boltzmann machines
-
-A restricted Boltzmann machine (RBM) is an undirected graphical model specified for discrete or continuous random variables, binary variables being most common [@smolensky1986information].
-Consider the binary case of an RBM only.
-A RBM architecture has two layers, hidden ($\mathcal{H}$) and visible ($\mathcal{V}$), with conditional independence within each layer.
-Let $\boldsymbol X = \{H_1, \dots, H_H, V_1,\dots, V_V\}$ be random variables corresponding to the hidden and visible variables in an RBM.
-We will consider the sample space $\mathcal{X} = \{-1,1\}$.
-The RBM is then has the joint probability mass function
-$$
-p_{\boldsymbol \theta} (\boldsymbol x) = \exp\left[ \boldsymbol \alpha^T \boldsymbol h + \boldsymbol \beta^T \boldsymbol v + \boldsymbol h^T \Gamma \boldsymbol v - \psi(\boldsymbol \theta)\right], \qquad \boldsymbol x = (\boldsymbol h, \boldsymbol v) \in \mathcal{X}^{H+V}
-$$
-where 
-$$
-\psi(\boldsymbol \theta) = \log \sum_{\boldsymbol x \in \mathcal{X}^{H + V}} \exp\left[ \boldsymbol \alpha^T \boldsymbol h + \boldsymbol \beta^T \boldsymbol v + \boldsymbol h^T \Gamma \boldsymbol v\right], \qquad \boldsymbol \theta \in \Theta
-$$ 
-is the normalizing function, $\boldsymbol \alpha \in \mathbb{R}^H$, $\boldsymbol \beta \in \mathbb{R}^V$, and $\Gamma$ is a matrix with dimension $H \times V$ are the parameters in the model.
-Let $\boldsymbol \theta = (\boldsymbol \alpha, \boldsymbol \beta,\Gamma) \in \Theta \subset \mathbb{R}^{H + V + H*V}$ denote the combined vector of parameters.
-
-The probability mass function for the visible random variables $V_1, \dots, V_V$ follows from this joint specification:
-$$
-p_{\boldsymbol \theta} (\boldsymbol v) = \sum\limits_{\boldsymbol h \in \mathcal{X}^H} p_{\boldsymbol \theta} (\boldsymbol x), \qquad \boldsymbol v \in \mathcal{X}^V
-$$
-
-By definition, $V_1, \dots, V_V$ are discrete with a finite sample space.
-The vector of model parameters $\boldsymbol \theta$, of size $q(V) \equiv V + H + H*V$, is expanding in size as a function of the dimension of $V_1, \dots, V_V$.
-Additionally, $p_{\boldsymbol \theta}(\boldsymbol v) > 0$ for all $\boldsymbol v \in \mathcal{X}^V$ by properties of the exponential function.
-Thus, the visible RBM model specification is an example of a FSFS model.
-This example also indicates that models formed by marginalizing a base FSFS model (e.g., exponential family) is again a FSFS model class.
-
-## Deep learning
-
-Consider two models with "deep architechture" that contain multiple hidden (or latent) layers in addition to a visible layer of data, a deep Boltzmann machine [@salakhutdinov2009deep] and a deep belief network [@hinton2006fast].
-Let $M$ denote the number of hidden layers included in the model and $H_1, \dots, H_M$ the number of hidden variables within each hidden layer.
-Then let $\boldsymbol X = \{H^{(1)}_1, \dots, H^{(1)}_{H_1}, \dots, H^{(M)}_1, \dots, H^{(M)}_{H_M}, V_1,\dots, V_V\}$ be random variables corresponding to the hidden and visible variables in a deep probabilitic model.
-We will consider the sample space $\mathcal{X} = \{-1,1\}$.
-
-**Deep Boltzmann machine (DBM).** The DBM class of models maintains conditional independence within all layers in the model by stacking RBM models and maintaining the restriction of only allowing conditional dependence between layers.
-The joint probability mass function for a DBM is
-$$
-p_{\boldsymbol \theta} (\boldsymbol x) = \exp\left[ \sum\limits_{i = 1}^M\boldsymbol \alpha^{(i)T} \boldsymbol h^{(i)} + \boldsymbol \beta^T \boldsymbol v + \boldsymbol h^{(1)T} \Gamma^{(0)} \boldsymbol v + \sum\limits_{i = 1}^{M - 1} \boldsymbol h^{(i)T} \Gamma^{(i)} \boldsymbol h^{(i + 1)} - \psi(\boldsymbol \theta) \right],
-$$
-where $\boldsymbol x = (\boldsymbol h^{(1)}, \dots, \boldsymbol h^{(M)}, \boldsymbol v) \in \mathcal{X}^{H_1 + \cdots + H_M +V}$ and
-$$
-\psi(\boldsymbol \theta) = \log \sum_{\boldsymbol x \in \mathcal{X}^{H_1 + \cdots + H_M + V}} \exp\left[ \sum\limits_{i = 1}^M\boldsymbol \alpha^{(i)T} \boldsymbol h^{(i)} + \boldsymbol \beta^T \boldsymbol v + \boldsymbol h^{(1)T} \Gamma^{(0)} \boldsymbol v + \sum\limits_{i = 1}^{M - 1} \boldsymbol h^{(i)T} \Gamma^{(i)} \boldsymbol h^{(i + 1)}\right], \qquad \boldsymbol \theta \in \Theta
-$$ 
-is the normalizing function, $\boldsymbol \beta \in \mathbb{R}^V$, $\boldsymbol \alpha^{(i)} \in \mathbb{R}^{H_i}$ for $i = 1, \dots, M$, $\Gamma^{(0)}$ is a matrix with dimension $H_1 \times V$, and $\Gamma^{(i)}$ is a matrix with dimension $H_i \times H_{i + 1}$ for $i = 1, \dots, M-1$ are the parameters in the model.
-Let $\boldsymbol \theta = (\boldsymbol \alpha^{(1)}, \dots, \boldsymbol \alpha^{(M)}, \boldsymbol \beta,\Gamma^{(0)}, \dots, \Gamma^{(M - 1)}) \in \Theta \subset \mathbb{R}^{H_1 + \cdots + H_M + V + H_1*V + H_1 *H_{2} + \cdots + H_{M-1}*H_M}$ denote the combined vector of parameters.
-
-The probability mass function for the visible random variables $V_1, \dots, V_V$ follows from this joint specification:
-$$
-p_{\boldsymbol \theta} (\boldsymbol v) = \sum\limits_{(\boldsymbol h^{(1)}, \dots, \boldsymbol h^{(M)}) \in \mathcal{X}^{H_1 + \cdots + H_M}} p_{\boldsymbol \theta} (\boldsymbol x), \qquad \boldsymbol v \in \mathcal{X}^V
-$$
-
-By definition, $V_1, \dots, V_V$ are discrete with a finite sample space.
-The vector of model parameters $\boldsymbol \theta$, of size $q(V) \equiv H_1 + \cdots + H_M + V + H_1*V + H_1 *H_{2} + \cdots + H_{M-1}*H_M$, is expanding in size as a function of the dimension of $V_1, \dots, V_V$.
-Additionally, $p_{\boldsymbol \theta}(\boldsymbol v) > 0$ for all $\boldsymbol v \in \mathcal{X}^V$ by properties of the exponential function.
-Thus, the visible RBM model specification is an example of a FSFS model.
-
-**Deep belief network (DBN).** A DBN is similar to a DBM in that there are multiple layers of latent random variables stacked in a deep architecture with no conditional dependence between layers.
-The difference between the two models is that all but the last stacked model are Bayesian networks, rather than RBMs.
-Thus for $V_1, \dots, V_V$ discrete with finite sample space and support equal to the sample space $\mathcal{X}^V$, a DBN is also a FSFS model if the number of components in the parameter vector is dependent on the dimension of the data, $V$.
-Commonly, as in a logistic belief net [@neal1992connectionist], a "weight" parameter is placed on each interaction between visibles, $V_1, \dots, V_V$ and the first layer of latent variables, $H^{(1)}_1, \dots, H^{(1)}_{H_1}$, satisfying the definition of a FSFS model.
-
-# Instability results
-
-To measure "instability" in a FSFS model, it is useful to imagine the behavior of a data model sequence $P_{\theta_N}$ specified to incorporate more and more observations.
-A relevant quantity to this end is a (scaled) extremal log-probability ratio given by
-\begin{align}
-\frac{1}{N} \log \left[\frac{\max\limits_{(x_1, \dots, x_N) \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(x_1, \dots, x_N)}{\min\limits_{(x_1, \dots, x_N) \in \mathcal{X}^N}P_{\boldsymbol \theta}(x_1, \dots, x_N)}\right] \equiv \frac{1}{N} \text{ELPR}_N(\boldsymbol \theta_N).
-\label{eqn:elpr}
-\end{align}
-
-The main idea is that, in formulating a FSFS model for potentially increasing numbers of variables (i.e., $N \rightarrow \infty$), the ratio (\ref{eqn:elpr}) should remain bounded, entailing that the largest probability possible under $P_{\theta_N}$ should maintain a magnitude within the smallest probability allowed under the same model.
-Specifically, the log of the fold change should grow at mostly linearly with the sample size $N$.
-This leads to the following definition:
-
-\begin{defn}[S-unstable FSFS]
-Let $\boldsymbol \theta_N \in \mathbb{R}^{q(N)}$ be a sequence of FSFS model parameters where the size of the model $q(N)$ is a function of the number of random variables $N$.
-A FSFS model formulation is \emph{Schweinberger-unstable} or \emph{S-unstable} if, as the number of variables increase ($N \rightarrow \infty$), it holds that
-\begin{align*}
-\lim\limits_{N \rightarrow \infty} \frac{1}{N} \text{ELPR}(\boldsymbol \theta_N) \equiv \lim\limits_{N \rightarrow \infty} \frac{1}{N} \log \left[\frac{\max\limits_{(x_1, \dots, x_N) \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(x_1, \dots, x_N)}{\min\limits_{(x_1, \dots, x_N) \in \mathcal{X}^N}P_{\boldsymbol \theta}(x_1, \dots, x_N)}\right] = \infty.
-\end{align*}
-\end{defn}
-In other words, given any $C > 0$, there exists an integer $N_C > 0$ so that $\frac{1}{N}\text{ELPR}_N(\boldsymbol \theta_N) > C$ for all $N \ge N_C$.
-
-Note that this definition of *S-unstable* is a generalization or reinterpretation of "unstable" used in @schweinberger2011instability by allowing nonexponential family models and an increasing number of parameters.
-While this definition appears different, it does match that in @schweinberger2011instability for the special exponential models considered there, see Section \ref{discrete-exponential-family-models} for details.
-
-S-unstable FSFS models are undesirable for several reasons. One is that small changes in data can lead to overly-sensitive changes in probability. Consider, for example, 
-$$
-\Delta(\boldsymbol \theta_N) \equiv \max \left\{\log \frac{P_{\boldsymbol \theta_N}(x_1, \dots, x_N)}{P_{\boldsymbol \theta_N}(x_1^*, \dots, x_N^*)} : (x_1, \dots, x_N) \& (x_1^*, \dots, x_N^*) \in \mathcal{X}^N \text{ differ by exactly one component}\right\},
-$$
-the biggest log-probability ratio for a one component change in data outcomes at a FSFS parameter $\boldsymbol \theta_N$. We then have the following result.
-
-\begin{prop}
-\label{prop:instab}
-Let $C > 0$ and let $\text{ELPR}(\boldsymbol \theta_N)$ be as in (\ref{eqn:elpr}) for an integer $N \ge 1$. If $$\frac{1}{N}\text{ELPR}_N(\boldsymbol \theta_N) > C,$$ 
-then 
-$$\Delta_N(\boldsymbol \theta_N) > C.$$
-\end{prop}
-
-Again, if the probability ratio (\ref{eqn:elpr}) is too large, then the FSFS model will exhibit large probability shifts for very small changes in the data configuration, which exemplifies the intuitive notation of instability..
-
-Additionally, S-unstable FSFS models are connected to degenerate models, where *degeneracy* implies placing all probability on a small portion of the sample space.
-That is, instability in FSFS models, as measured by (\ref{eqn:elpr}), is a spectrum and Proposition \ref{prop:instab} indicates increasing levels of one undesirable type of model behavior as (\ref{eqn:elpr}) increases.
-Furthermore, as (\ref{eqn:elpr}) increases, FSFS models will slide into full degeneracy as Proposition \ref{prop:degen} next shows.
-Define a modal set
-$$
-M_{\epsilon, \boldsymbol \theta_N} \equiv \left\{\boldsymbol x \in \mathcal{X}^N: \log P_{\boldsymbol \theta_N}(\boldsymbol x) > (1-\epsilon)\max\limits_{\boldsymbol x^* \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x^*) + \epsilon\min\limits_{\boldsymbol x^* \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x^*) \right\}
-$$
-of possible outcomes, for a given $0 < \epsilon < 1$ and along a sequence $\boldsymbol \theta_N$, $N \ge 1$, of FSFS parameters.
-
-\begin{prop}
-\label{prop:degen}
-For an unstable FSFS model, and for any given $0 < \epsilon < 1$, $P_{\boldsymbol \theta_N}\left((x_1, \dots, x_N) \in M_{\epsilon, \boldsymbol \theta_N}\right) \rightarrow 1$ holds as $N \rightarrow \infty$.
-\end{prop}
-
-In other words, as a consequence of S-unstable models, all probability in the FSFS model formulation with a large number of random variables will stack up on mode sets or simply those few outcomes with the highest probability.
-
-**Proof of Proposition \ref{prop:instab}.** We prove the contrapositive.
+{\bf Proof of Proposition \ref{prop:instab}.} We prove the contrapositive.
 Suppose that $\Delta(\boldsymbol \theta_N) \le C$ holds for some $C > 0$.
@@
 \begin{align*}
-\log\left[\frac{P_{\boldsymbol \theta_N}(\boldsymbol x_{max})}{P_{\boldsymbol \theta_N}(\boldsymbol x_{min})}\right] &= \left|\sum\limits_{i = 1}^k\log\left(\frac{P_{\boldsymbol \theta_N}(\boldsymbol x_i)}{P_{\boldsymbol \theta_N}(\boldsymbol x_{i-1})}\right)\right| \\
-& \le \sum\limits_{i = 1}^k\left|\log\left(\frac{P_{\boldsymbol \theta_N}(\boldsymbol x_i)}{P_{\boldsymbol \theta}(\boldsymbol x_{i-1})}\right)\right| \\
-& \le k \Delta_N(\boldsymbol \theta_N) \le NC
+\log\left[\frac{P_{\boldsymbol \theta_N}(\boldsymbol x_{max})}{P_{\boldsymbol \theta_N}(\boldsymbol x_{min})}\right] &= \left|\sum\limits_{i = 1}^k\log\left(\frac{P_{\boldsymbol \theta_N}(\boldsymbol x_i)}{P_{\boldsymbol \theta_N}(\boldsymbol x_{i-1})}\right)\right| \le \sum\limits_{i = 1}^k\left|\log\left(\frac{P_{\boldsymbol \theta_N}(\boldsymbol x_i)}{P_{\boldsymbol \theta}(\boldsymbol x_{i-1})}\right)\right| \le k \Delta_N(\boldsymbol \theta_N) \le NC
 \end{align*}
@@
 
-**Proof of Proposition \ref{prop:degen}.** Define $\boldsymbol x_{max}$ and $\boldsymbol x_{min}$ as in the proof of Proposition \ref{prop:instab}.
+{\bf Proof of Proposition \ref{prop:degen}.} Define $\boldsymbol x_{max}$ and $\boldsymbol x_{min}$ as in the proof of Proposition \ref{prop:instab}.
 Fix $0 < \epsilon < 1$ Then, $\boldsymbol x_{max} \in M_{\epsilon, \boldsymbol \theta_N}$, so $P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N}) \ge P_{\boldsymbol \theta_N}(\boldsymbol x_{max})$.
@@
 Consequently, $P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N}) \rightarrow 1$ as $N \rightarrow \infty$ as claimed.$\Box$
-
-\begin{remark}
-There is a further generalization to the notion of "unstable" used in Schweinberger (2011) that deals with replication.
-One can conceivably imagine replications of the data $\boldsymbol X_1, \dots, \boldsymbol X_n$, where $X_i \in \mathbb{R}^N, i = 1, \dots, n$ such that there are now $n*N$ random variables in the model.
-However, the definition of S-unstable and propositions \ref{prop:instab}-\ref{prop:degen} still hold if the replications are independent and identically distributed.
-The reason for this is that for $n$ observations, $\left(\max\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)\right)^n$ is the largest probability possible under the model and similarly $\left(\min\limits_{\boldsymbol x \in \mathcal{X}}P_{\boldsymbol \theta_N}(\boldsymbol x)\right)^n$ is the smallest.
-Thus,
-\begin{align*}
-\frac{\text{extremal log-probability ratio}}{\text{\# random variables in the model}} &\equiv \frac{1}{n*N}\log \left[\frac{\left\{\max\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)\right\}^n}{\left\{\min\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta}(\boldsymbol x)\right\}^n}\right] \\
-&= \frac{1}{N} \log \left[\frac{\max\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)}{\min\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta}(\boldsymbol x)}\right]\\
-&= \frac{1}{N} \text{ELPR}(\boldsymbol \theta_N).
-\end{align*}
-Meaning, the definition of an S-unstable FSFS model is invariant to replication.
-\end{remark}
-
-# Implications
-
-For a large class of models that cover a broad range of applications (including "deep learning"), we have developed a formal definition and elucidated multiple consequences of instability.
-We have shown for FSFS models that instability manifests through small changes in data leading to overly-sensitive changes in probability as well as the potential to place all probability on a small pieace of the sample space.
-Models that fall within the definition of a FSFS model should be used with caution to ensure that the effects of instability are not experienced.
-
-# References

### 2016-11-20 writing/note.Rmd
remaining edits

@@
   pdf_document:
+    keep_tex: true
     fig_caption: yes
@@
 Let $\boldsymbol X=(X_1,\ldots,X_N)$ denote the $N$ random variables for visibles with support $\mathcal{X}^N$ and $\boldsymbol H=(H_1,\ldots,H_{N_H})$ denote the $N_H$ random variables for hiddens with support $\mathcal{X}^{N_H}$ where  $\mathcal{X} = \{-1,1\}$.
-The RBM for $\tilde{\boldsymbol X}=(\boldsymbol X,\boldsymbol H})$ then has the joint probability mass function
+The RBM for $\tilde{\boldsymbol X}=(\boldsymbol X,\boldsymbol H)$ then has the joint probability mass function
 $$
@@
 $$
-p_{\boldsymbol \theta} (\boldsymbol x) = \sum\limits_{\boldsymbol h \in \mathcal{X}^{N_H} p_{\boldsymbol \theta} (\boldsymbol x, \boldsymbol h), \qquad \boldsymbol x \in \mathcal{X}^N
+p_{\boldsymbol \theta} (\boldsymbol x) = \sum\limits_{\boldsymbol h \in \mathcal{X}^{N_H}} p_{\boldsymbol \theta} (\boldsymbol x, \boldsymbol h), \qquad \boldsymbol x \in \mathcal{X}^N
 $$
@@
 
-Note that this definition of *S-unstable* is a generalization or reinterpretation of "unstable" used in @schweinberger2011instability by allowing nonexponential family models and an increasing number of parameters.
-While this definition appears different, it does match that in @schweinberger2011instability for the special exponential models considered there, see Section \ref{discrete-exponential-family-models} for details.
+Note that this definition of *S-unstable* is a generalization or reinterpretation of "unstable" used in @schweinberger2011instability by allowing nonexponential family models (e.g.
+RBM and DBM models in Sections \ref{restricted-boltzmann-machines}-\ref{deep-learning}) and an increasing number of parameters.
+While this definition differs in form, it does match that in @schweinberger2011instability for the special exponential models (cf.
+Section \ref{discrete-exponential-family-models}) considered there.
 
@@
 $$
-the biggest log-probability ratio for a one component change in data outcomes at a FSFS parameter $\boldsymbol \theta_N$. We then have the following result.
+the biggest log-probability ratio for a one component change in data outcomes at a FSFS parameter $\boldsymbol \theta_N$. We then have the following non-asymptotic result.
 
@@
 
-Again, if the probability ratio (\ref{eqn:elpr}) is too large, then the FSFS model will exhibit large probability shifts for very small changes in the data configuration, which exemplifies the intuitive notation of instability..
+Again, if the probability ratio (\ref{eqn:elpr}) is too large, then the FSFS model will exhibit large changes in probability for very small differences in the data configuration, which exemplifies the intuitive notation of instability.
 
 Additionally, S-unstable FSFS models are connected to degenerate models, where *degeneracy* implies placing all probability on a small portion of the sample space.
-That is, instability in FSFS models, as measured by (\ref{eqn:elpr}), is a spectrum and Proposition \ref{prop:instab} indicates increasing levels of one undesirable type of model behavior as (\ref{eqn:elpr}) increases.
-Furthermore, as (\ref{eqn:elpr}) increases, FSFS models will slide into full degeneracy as Proposition \ref{prop:degen} next shows.
-Define a modal set
+For perspective, note that instability in FSFS models, as measured by (\ref{eqn:elpr}), is a spectrum and Proposition \ref{prop:instab} indicates increasing and undesirable sensitivity to model probabilities (i.e., in one component changes to outcomes) as (\ref{eqn:elpr}) increases.
+Furthermore, as the instability measure (\ref{eqn:elpr}) grows, FSFS models are guaranteed to slide into full degeneracy as Proposition \ref{prop:degen} next shows.
+Define a $\epsilon$-modal set
 $$
@@
 $$
-of possible outcomes, for a given $0 < \epsilon < 1$ and along a sequence $\boldsymbol \theta_N$, $N \ge 1$, of FSFS parameters.
+of possible outcomes, for a given $0 < \epsilon < 1$, and along a sequence $\boldsymbol \theta_N$, $N \ge 1$, of FSFS parameters.
 
@@
 
-In other words, as a consequence of S-unstable models, all probability in the FSFS model formulation with a large number of random variables will stack up on mode sets or simply those few outcomes with the highest probability.
-
-**Proof of Proposition \ref{prop:instab}.** We prove the contrapositive.
-Suppose that $\Delta(\boldsymbol \theta_N) \le C$ holds for some $C > 0$.
-Under the FSFS model, $P_{\boldsymbol \theta_N}(\boldsymbol x) > 0$ holds for each outcome $\boldsymbol x \in \mathcal{X}^N$.
-Let $\boldsymbol x_{min} \equiv \argmin\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)$ and $\boldsymbol x_{max} \equiv \argmax\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)$.
-Note there exists a sequence $\boldsymbol x_{min} \equiv \boldsymbol x_0, \boldsymbol x_1, \dots, \boldsymbol x_k \equiv \boldsymbol x_{max}$ in $\mathcal{X}^N$ of component-wise switches to move from $\boldsymbol x_{min}$ to $\boldsymbol x_{max}$ in the sample space (i.e.
-$\boldsymbol x_i, \boldsymbol x_{i + 1} \in \mathcal{X}^N$ differ by exactly $1$ component for $i = 0, \dots, k$) for some integer $k \in \{0, 1, \dots, N\}$.
-Then, if $k > 0$, it follows that
-\begin{align*}
-\log\left[\frac{P_{\boldsymbol \theta_N}(\boldsymbol x_{max})}{P_{\boldsymbol \theta_N}(\boldsymbol x_{min})}\right] &= \left|\sum\limits_{i = 1}^k\log\left(\frac{P_{\boldsymbol \theta_N}(\boldsymbol x_i)}{P_{\boldsymbol \theta_N}(\boldsymbol x_{i-1})}\right)\right| \\
-& \le \sum\limits_{i = 1}^k\left|\log\left(\frac{P_{\boldsymbol \theta_N}(\boldsymbol x_i)}{P_{\boldsymbol \theta}(\boldsymbol x_{i-1})}\right)\right| \\
-& \le k \Delta_N(\boldsymbol \theta_N) \le NC
-\end{align*}
-using $k \le N$ and $\Delta(\boldsymbol \theta_N) \le C$. If $k = 0$, then $\boldsymbol x_{max} = \boldsymbol x_{min}$ and the same bound above holds.$\Box$
-
-**Proof of Proposition \ref{prop:degen}.** Define $\boldsymbol x_{max}$ and $\boldsymbol x_{min}$ as in the proof of Proposition \ref{prop:instab}.
-Fix $0 < \epsilon < 1$ Then, $\boldsymbol x_{max} \in M_{\epsilon, \boldsymbol \theta_N}$, so $P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N}) \ge P_{\boldsymbol \theta_N}(\boldsymbol x_{max})$.
-If $\boldsymbol x \in \mathcal{X}^N \setminus M_{\epsilon, \boldsymbol \theta_N}$, then by definition $P_{\boldsymbol \theta_N}(\boldsymbol x) \le [P_{\boldsymbol \theta_N}(\boldsymbol x_{max})]^{1-\epsilon}[P_{\boldsymbol \theta_N}(\boldsymbol x_{min})]^{\epsilon}$ holds so that
-\begin{align*}
-1-P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N})
-& = \sum\limits_{\boldsymbol x \in \mathcal{X}^N \setminus M_{\epsilon, \boldsymbol \theta_N}}P_{\boldsymbol \theta_N}(\boldsymbol x) \\
-& \le (2^N)[P_{\boldsymbol \theta_N}(\boldsymbol x_{max})]^{1-\epsilon}[P_{\boldsymbol \theta_N}(\boldsymbol x_{min})]^{\epsilon}
-\end{align*}
-Then, it follows that
-\begin{align*}
-\frac{1}{N}\log\left[\frac{P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N})}{1-P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N})}\right] & \ge \frac{1}{N} \log\left[\frac{P_{\boldsymbol\theta_N}(\boldsymbol x_{max})}{(2^V)[P_{\boldsymbol \theta_N}(\boldsymbol x_{max})]^{1-\epsilon}[P_{\boldsymbol \theta_N}(\boldsymbol x_{min})]^{\epsilon}}\right] \\
-&= \frac{\epsilon}{N} \log\left[\frac{P_{\boldsymbol \theta_N}(\boldsymbol x_{max})}{P_{\boldsymbol \theta_N}(\boldsymbol x_{min})}\right] - \log 2 \rightarrow \infty
-\end{align*}
-as $N \rightarrow \infty$ by the definition of an unstable FSFS model.
-Consequently, $P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N}) \rightarrow 1$ as $N \rightarrow \infty$ as claimed.$\Box$
+In other words, in S-unstable FSFS models, all probability in the FSFS model formulation with a large number of random variables will mass on mode sets of arbitrarily small size, or simply those few outcomes with the highest probability.
 
@@
 There is a further generalization to the notion of "unstable" used in Schweinberger (2011) that deals with replication.
-One can conceivably imagine replications of the data $\boldsymbol X_1, \dots, \boldsymbol X_n$, where $X_i \in \mathbb{R}^N, i = 1, \dots, n$ such that there are now $n*N$ random variables in the model.
-However, the definition of S-unstable and propositions \ref{prop:instab}-\ref{prop:degen} still hold if the replications are independent and identically distributed.
-The reason for this is that for $n$ observations, $\left(\max\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)\right)^n$ is the largest probability possible under the model and similarly $\left(\min\limits_{\boldsymbol x \in \mathcal{X}}P_{\boldsymbol \theta_N}(\boldsymbol x)\right)^n$ is the smallest.
+One can conceivably imagine $n$ independent and identically distributed (iid) replications $\boldsymbol X_1, \dots, \boldsymbol X_n$, where each $\boldsymbol X_i=(X_{i,1},\ldots,X_{i,N}) \in \mathcal{X}^N$ follows a common FSFS model with probabilities $P_{\theta_N}(\boldsymbol x)>0$,  $\boldsymbol x\in\mathcal{X}^N$ and $|\mathcal{X}<\infty|$, for $i=1,\ldots,n$.
+This leads to a total of $n*N$ random variables in the joint FSFS model.
+However, the definition of S-unstable and Propositions \ref{prop:instab}-\ref{prop:degen} still hold for such iid replications.
+The reason for this is that $\left(\max\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)\right)^n$ is the largest probability possible under the joint model for the $n$ replications while $\left(\min\limits_{\boldsymbol x \in \mathcal{X}}P_{\boldsymbol \theta_N}(\boldsymbol x)\right)^n$ is the smallest probability.
 Thus,
@@
 &= \frac{1}{N} \log \left[\frac{\max\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)}{\min\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta}(\boldsymbol x)}\right]\\
-&= \frac{1}{N} \text{ELPR}(\boldsymbol \theta_N).
+&= \frac{1}{N} \text{ELPR}(\boldsymbol \theta_N),
 \end{align*}
-Meaning, the definition of an S-unstable FSFS model is invariant to replication.
+implying that the definition of an S-unstable FSFS model is invariant to the level $n$ of independent replication.
 \end{remark}
@@
 
+\begin{appendix}
+\section{Appendices}
+\label{appendix}
+\subsection{Proofs of main results}
+\label{proofs}
+
+```{r child='appendix/appendix-note.Rmd'}
+```
+\end{appendix}
+
 # References

### 2016-11-21 writing/appendix/appendix-center.Rmd
nordman edits and references cleanup

@@
 ---
-title: "Supplementary Material to:  On the propriety of restricted Boltzmann machines"
-author: "Andee Kaplan, Daniel Nordman, and Stephen Vardeman"
+title: "Supplementary material to:  On the propriety of restricted Boltzmann machines"
+author: | 
+  | Andee Kaplan, Daniel Nordman, and Stephen Vardeman
+  | Iowa State University
+  | ajkaplan@iastate.edu, dnordman@iastate.edu, vardeman@iastate.edu
 date: "November, 2016"

### 2016-11-21 writing/appendix/appendix-note.Rmd
nordman edits and references cleanup

@@
-{\bf Proof of Proposition \ref{prop:instab}.} We prove the contrapositive.
-Suppose that $\Delta(\boldsymbol \theta_N) \le C$ holds for some $C > 0$.
-Under the FSFS model, $P_{\boldsymbol \theta_N}(\boldsymbol x) > 0$ holds for each outcome $\boldsymbol x \in \mathcal{X}^N$.
+{\bf Proof of Proposition \ref{prop:instab}.} We prove the contrapositive, supposing that $\Delta(\boldsymbol \theta_N) \le C$ holds for some $C > 0$ and show $\text{ELPR}(\boldsymbol \theta_N) \leq C$.
 Let $\boldsymbol x_{min} \equiv \argmin\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)$ and $\boldsymbol x_{max} \equiv \argmax\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)$.
@@
 $\boldsymbol x_i, \boldsymbol x_{i + 1} \in \mathcal{X}^N$ differ by exactly $1$ component for $i = 0, \dots, k$) for some integer $k \in \{0, 1, \dots, N\}$.
+Under the FSFS model, recall $P_{\boldsymbol \theta_N}(\boldsymbol x) > 0$ holds so that $\log P_{\boldsymbol \theta_N}(\boldsymbol x)$ is well-defined for each outcome $\boldsymbol x \in \mathcal{X}^N$.
 Then, if $k > 0$, it follows that
-\begin{align*}
-\log\left[\frac{P_{\boldsymbol \theta_N}(\boldsymbol x_{max})}{P_{\boldsymbol \theta_N}(\boldsymbol x_{min})}\right] &= \left|\sum\limits_{i = 1}^k\log\left(\frac{P_{\boldsymbol \theta_N}(\boldsymbol x_i)}{P_{\boldsymbol \theta_N}(\boldsymbol x_{i-1})}\right)\right| \le \sum\limits_{i = 1}^k\left|\log\left(\frac{P_{\boldsymbol \theta_N}(\boldsymbol x_i)}{P_{\boldsymbol \theta}(\boldsymbol x_{i-1})}\right)\right| \le k \Delta_N(\boldsymbol \theta_N) \le NC
-\end{align*}
+$$
+N\cdot\text{ELPR}(\boldsymbol \theta_N) = \log\left[\frac{P_{\boldsymbol \theta_N}(\boldsymbol x_{max})}{P_{\boldsymbol \theta_N}(\boldsymbol x_{min})}\right] = \left|\sum\limits_{i = 1}^k\log\left(\frac{P_{\boldsymbol \theta_N}(\boldsymbol x_i)}{P_{\boldsymbol \theta_N}(\boldsymbol x_{i-1})}\right)\right| \le \sum\limits_{i = 1}^k\left|\log\left(\frac{P_{\boldsymbol \theta_N}(\boldsymbol x_i)}{P_{\boldsymbol \theta}(\boldsymbol x_{i-1})}\right)\right| \le k \Delta_N(\boldsymbol \theta_N) \le NC,
+$$
 using $k \le N$ and $\Delta(\boldsymbol \theta_N) \le C$. If $k = 0$, then $\boldsymbol x_{max} = \boldsymbol x_{min}$ and the same bound above holds.$\Box$
 
-{\bf Proof of Proposition \ref{prop:degen}.} Define $\boldsymbol x_{max}$ and $\boldsymbol x_{min}$ as in the proof of Proposition \ref{prop:instab}.
-Fix $0 < \epsilon < 1$ Then, $\boldsymbol x_{max} \in M_{\epsilon, \boldsymbol \theta_N}$, so $P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N}) \ge P_{\boldsymbol \theta_N}(\boldsymbol x_{max})$.
+{\bf Proof of Proposition \ref{prop:degen}.} Define $\boldsymbol x_{max}$ and $\boldsymbol x_{min} \in \mathcal{X}^N$ as in the proof of Proposition \ref{prop:instab} where $|\mathcal{X}|<\infty$ holds in the FSFS model.
+We may suppose $|\mathcal{X}|>1$ (i.e., $\mathcal{X}^N$ has more than one outcome) because otherwise the model is trivially degenerate for all $N \geq 1$.
+Fix $0 < \epsilon < 1$ Then, $\boldsymbol x_{max} \in M_{\epsilon, \boldsymbol \theta_N}$, so $P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N}) \ge P_{\boldsymbol \theta_N}(\boldsymbol x_{max}) > 0$.
 If $\boldsymbol x \in \mathcal{X}^N \setminus M_{\epsilon, \boldsymbol \theta_N}$, then by definition $P_{\boldsymbol \theta_N}(\boldsymbol x) \le [P_{\boldsymbol \theta_N}(\boldsymbol x_{max})]^{1-\epsilon}[P_{\boldsymbol \theta_N}(\boldsymbol x_{min})]^{\epsilon}$ holds so that
@@
 & = \sum\limits_{\boldsymbol x \in \mathcal{X}^N \setminus M_{\epsilon, \boldsymbol \theta_N}}P_{\boldsymbol \theta_N}(\boldsymbol x) \\
-& \le (2^N)[P_{\boldsymbol \theta_N}(\boldsymbol x_{max})]^{1-\epsilon}[P_{\boldsymbol \theta_N}(\boldsymbol x_{min})]^{\epsilon}
+& \le (|\mathcal{X}|^N)[P_{\boldsymbol \theta_N}(\boldsymbol x_{max})]^{1-\epsilon}[P_{\boldsymbol \theta_N}(\boldsymbol x_{min})]^{\epsilon}.
 \end{align*}
-Then, it follows that
+From the lower bound on $P_{\boldsymbol \theta_N}(\boldsymbol x_{max})$ and the upper bound on $1-P_{\boldsymbol \theta_N}(\boldsymbol x_{max})$, it follows that
 \begin{align*}
-\frac{1}{N}\log\left[\frac{P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N})}{1-P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N})}\right] & \ge \frac{1}{N} \log\left[\frac{P_{\boldsymbol\theta_N}(\boldsymbol x_{max})}{(2^V)[P_{\boldsymbol \theta_N}(\boldsymbol x_{max})]^{1-\epsilon}[P_{\boldsymbol \theta_N}(\boldsymbol x_{min})]^{\epsilon}}\right] \\
-&= \frac{\epsilon}{N} \log\left[\frac{P_{\boldsymbol \theta_N}(\boldsymbol x_{max})}{P_{\boldsymbol \theta_N}(\boldsymbol x_{min})}\right] - \log 2 \rightarrow \infty
+\frac{1}{N}\log\left[\frac{P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N})}{1-P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N})}\right] & \ge \frac{1}{N} \log\left[\frac{P_{\boldsymbol\theta_N}(\boldsymbol x_{max})}{(|\mathcal{X}|^N)[P_{\boldsymbol \theta_N}(\boldsymbol x_{max})]^{1-\epsilon}[P_{\boldsymbol \theta_N}(\boldsymbol x_{min})]^{\epsilon}}\right] \\
+&= \frac{\epsilon}{N} \log\left[\frac{P_{\boldsymbol \theta_N}(\boldsymbol x_{max})}{P_{\boldsymbol \theta_N}(\boldsymbol x_{min})}\right] - \log |\mathcal{X}| \rightarrow \infty
 \end{align*}
-as $N \rightarrow \infty$ by the definition of an unstable FSFS model.
+as $N \rightarrow \infty$ by the definition of an unstable FSFS model (c.f.
+Definition \ref{defn:instab}).
 Consequently, $P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N}) \rightarrow 1$ as $N \rightarrow \infty$ as claimed.$\Box$

### 2016-11-21 writing/draft.Rmd
nordman edits and references cleanup

@@
 title: "On the propriety of restricted Boltzmann machines"
-author: "Andee Kaplan, Daniel Nordman, and Stephen Vardeman"
+author: | 
+  | Andee Kaplan, Daniel Nordman, and Stephen Vardeman
+  | Iowa State University
+  | ajkaplan@iastate.edu, dnordman@iastate.edu, vardeman@iastate.edu
 date: "November, 2016"
@@
 
-The data mining and machine learning communities have recently shown great interest in deep learning, specifically in stacked restricted Boltzmann machines (RBMs) [see @salakhutdinov2009deep, @salakhutdinov2012efficient, @srivastava2013modeling, @le2008representational for examples].
+The data mining and machine learning communities have recently shown great interest in deep learning, specifically in stacked restricted Boltzmann machines (RBMs) [see @salakhutdinov2009deep; @salakhutdinov2012efficient; @srivastava2013modeling; @le2008representational for examples].
 A RBM is a probabilistic undirected graphical model (for discrete or continuous random variables) with two layers, one hidden and one visible, with no conditional dependency within a layer [@smolensky1986information].
-These models have reportedly been used with success in classification of images [@larochelle2008classification, @srivastava2012multimodal].
+These models have reportedly been used with success in classification of images [@larochelle2008classification; @srivastava2012multimodal].
 However, the model properties are largely unexplored in the literature and the commonly cited fitting methodology remains heuristic-based and abstruse [@hinton2006fast].
@@
 Recalling [@fisher1922mathematical], the aim of a statistical model is to represent data in a compact way.
-Neyman and Box further state that a model should "provide an explanation of the mechanism underlying the observed phenomena" [@lehmann1990model, @box1967discrimination].
+Neyman and Box further state that a model should "provide an explanation of the mechanism underlying the observed phenomena" [@lehmann1990model; @box1967discrimination].
 At issue, RBMs  often produce data lacking realistic variability and may thereby fail to provide a satisfactory conceptualization of a data generation process.
@@
 In practice, this may seen when a single pixel change in an image results in a wildly different classification in an image classification problem.
-Such model impropriety issues have recently been documented in RBMs [@li2014biclustering], as well as other deep architectures [@szegedy2013intriguing, @nguyen2014deep].
+Such model impropriety issues have recently been documented in RBMs [@li2014biclustering], as well as other deep architectures [@szegedy2013intriguing; @nguyen2014deep].
 We investigate these phenomena for RBMs in Section \ref{explorations} through simulations of small, manageable examples.
@@
 As the size of these models grows, both maximum likelihood and Bayesian methods of fitting quickly become intractable.
-The literature often suggests Markov chain Monte Carlo (MCMC) tools for approximate maximization of likelihoods to fit these models (e.g., Gibbs sampling to exploit conditional structure in hidden and visible variables), but little is said about acheiving convergence [@hinton2010practical, @hinton2006fast].
+The literature often suggests Markov chain Monte Carlo (MCMC) tools for approximate maximization of likelihoods to fit these models (e.g., Gibbs sampling to exploit conditional structure in hidden and visible variables), but little is said about acheiving convergence [@hinton2010practical; @hinton2006fast].
 Related to this, these MCMC algorithms require updating potentially many latent variables (hiddens) which can critically influence convergence in MCMC-based likelihood methods.
@@
 
-For a RBM model with enough hidden variables, it has been shown that any distribution for the visibles can be approximated arbitrarily well [@le2008representational, @montufar2011refinements, and @montufar2011expressive].
+For a RBM model with enough hidden variables, it has been shown that any distribution for the visibles can be approximated arbitrarily well [@le2008representational; @montufar2011refinements; and @montufar2011expressive].
 However, the empirical distribution of a training set of vectors of visibles is the best fitting model for observed cell data.
@@
 
-RBMs have risen to prominence in recent years due to their connection to deep learning [see @hinton2006fast, @salakhutdinov2012efficient, or @srivastava2013modeling for examples].
+RBMs have risen to prominence in recent years due to their connection to deep learning [see @hinton2006fast; @salakhutdinov2012efficient; @srivastava2013modeling for examples].
 By stacking multiple layers of RBMs in a deep architecture, proponents of the models claim to produce the ability to learn "internal representations that become increasingly complex, which is considered to be a promising way of solving object and speech recognition problems" [@salakhutdinov2009deep, pp.
@@
 The previous discussion intends to indicate that the RBM model class suffers from parameter identifiability issues that go beyond mere symmetries in the parametrization.
-Not only it is possible to approximate any distribution on the visibles arbitrarily well (cf.
-@montufar2011refinements), quite different parameter settings can induce the same essential RBM model.
+Not only it is possible to approximate any distribution on the visibles arbitrarily well [cf.
+@montufar2011refinements], quite different parameter settings can induce the same essential RBM model.
 However, this is not the most disastrous implication of the RBM parameterization.
@@
 For illustration, continue the simple example from above with $n$ iid observations, each consisting of two realized visibles $(V_1,V_2)$.
-In which case, when the specified cell probabilities $p_{(-1,-1)},p_{(1,-1)}, p_{(-1,1)},p_{(1,1)}$ in (\ref{eqn:0}) are taken as the empirical cell frequencies from the sample, there is no better model based on maximum likelihood and the discussion above (c.f.
+In which case, when the specified cell probabilities $p_{(-1,-1)},p_{(1,-1)}, p_{(-1,1)},p_{(1,1)}$ in (\ref{eqn:0}) are taken as the empirical cell frequencies from the sample, there is no better model based on maximum likelihood and the discussion above (cf.
 (\ref{eqn:2})) shows that RBM model parameters can be chosen to re-create this empirical distribution to an arbitrarily close degree.
@@
 \caption{The values used for the hyperparameters for all three fitting methods.
-A rule of thumb is imposed which decreases prior variances for the model parameters as the size of the model increases and also shrinks $\boldsymbol \theta_{interaction}$ more than $\boldsymbol \theta_{main}$.}
+A rule of thumb is imposed which decreases prior variances for the model parameters as the size of the model increases and also shrinks $\boldsymbol \theta_{interaction}$ more than $\boldsymbol \theta_{main}$.
+The common $C$ defining $C_1$ and $C_2$  in the BwTPLV method is chosen by tuning.}
 \label{tab:hyperparam}

### 2016-11-21 writing/note.Rmd
nordman edits and references cleanup

@@
 title: "A note on the instability and degeneracy of deep learning models"
-author: "Andee Kaplan, Daniel Nordman, and Stephen Vardeman"
+author: | 
+  | Andee Kaplan, Daniel Nordman, and Stephen Vardeman
+  | Iowa State University
+  | ajkaplan@iastate.edu, dnordman@iastate.edu, vardeman@iastate.edu
 date: "November, 2016"
@@
 
-As data structures become larger and more complex, there is increasing interest in identifying instability in potential model specifications for such data (e.g., models for random graphs or networks).
-We consider the problem of quantifying instability for general probability models defined on sequences of observations, where each sequence of a given length $N$ has a finite number of possible outcomes.
-As the number of available observations increases, a sequence of probability models results to accommodate the data of expanding dimension.
-
-Model instability is formally shown to occur when a certain log-probability ratio under such models grows faster than the length $N$ of the data sequence.
-In this case, a one component change in the data sequence can provably shift probability by orders of magnitude.
-Also, as the measure of instability becomes more extreme, the resulting probability models are shown to slip into degeneracy, thereby placing all probability on arbitrarily small portions of the sample space.
-
 We consider the behavior, and the potential impropriety, of probability models built to incorporate a sequence of discrete observations with an arbitrarily specified length $N$.
 Let $(X_1, \dots X_N)$ denote a set of sequence of discrete random variables with a finite sample space, $\mathcal{X}^N$.
-That is, $\mathcal{X}$ represents a finite set of potential outcomes for the variable $X_i$, and the data sequence $(X_1,\ldots,X_N)$ assumes values in the $N$-fold product space $\mathcal{X}^N$.
+That is, $\mathcal{X}$ with $|\mathcal{X}| < \infty$ represents a finite set of potential outcomes for the variable $X_i$, and the data sequence $(X_1,\ldots,X_N)$ assumes values in the $N$-fold product space $\mathcal{X}^N$.
 For any $N$, let $P_{\boldsymbol \theta_N}$ denote a probability model on $\mathcal{X}^N$, under which $P_{\boldsymbol \theta_N}(x_1, \dots, x_N) > 0$ is the probability of the data outcome $(x_1, \dots, x_N) \in \mathcal{X}^N$.
@@
 Section \ref{instability-results} establishes formal results regarding the model propriety of FSFS models with regard to stability.
-A FSFS probability model intuitively exhibits instability if small changes in the components of a data outcome $(x_1,\ldots,x_N)$ may result in large in probability $P_{\theta_N}(x_1,\ldots,x_N)$.
+A FSFS probability model intuitively exhibits instability if small changes in the components of a data outcome $(x_1,\ldots,x_N)$ may result in large changes in probability $P_{\theta_N}(x_1,\ldots,x_N)$.
 The concept of instability, introduced in the field statistical physics by @ruelle1999statistical, was extended to include a notion of detection and quantification for certain exponential
@@
 $$
-p_{\boldsymbol \lambda}(\boldsymbol x) = \exp\left[\boldsymbol\eta^T(\boldsymbol \lambda) \boldsymbol g(\boldsymbol x) - \psi(\boldsymbol \lambda)\right], \qquad \boldsymbol x \in \mathcal{X}^N.
+p_{\boldsymbol \lambda}(\boldsymbol x) = \exp\left[\boldsymbol\eta^T(\boldsymbol \lambda) \boldsymbol g(\boldsymbol x) - \psi(\boldsymbol \lambda)\right], \qquad \boldsymbol x \in \mathcal{X}^N,
 $$ 
@@
 $$
-\psi(\boldsymbol \lambda) = \log \sum\limits_{\boldsymbol x \in \mathcal{X}^N}\exp\left[\boldsymbol \eta^T(\boldsymbol \lambda) \boldsymbol g(\boldsymbol x) \right], \qquad \boldsymbol \lambda \in \Lambda
+\psi(\boldsymbol \lambda) = \log \sum\limits_{\boldsymbol x \in \mathcal{X}^N}\exp\left[\boldsymbol \eta^T(\boldsymbol \lambda) \boldsymbol g(\boldsymbol x) \right], \qquad \boldsymbol \lambda \in \Lambda,
 $$
@@
 
-Setting $\boldsymbol \theta_N=\lambda_N$ to a subsequence of $\mathbb{R}^k$ and noting that $p_{\boldsymbol \lambda}(\boldsymbol x) > 0$ for all $\boldsymbol x \in \mathcal{X}^N$ by properties of the exponential function, the discrete exponential family models defined in this section are a special case of the FSFS model.
-Such exponential models arise in spatial data on a lattice [@besag1974spatial], network data [@wasserman1994social, @handcock2003assessing], and even in binomial sampling with $N$ iid Bernoulli random variables.
-@schweinberger2011instability considered instability in such exponential models (e.g., for random graphs) for fixed parameter sequences $\theta_N=\lambda\in\mathbb{R}^k$.
+Defining $\boldsymbol \theta_N=\lambda_N$ to a subsequence of $\Lambda \subset \mathbb{R}^k$ and noting that $p_{\boldsymbol \lambda}(\boldsymbol x) > 0$ for all $\boldsymbol x \in \mathcal{X}^N$ by properties of the exponential function, the discrete exponential family models defined in this section are a special case of the FSFS model.
+Such exponential models arise with spatial data on a lattice [@besag1974spatial], network data [@wasserman1994social; @handcock2003assessing], and even binomial sampling with $N$ iid Bernoulli random variables.
+For these exponential models, the length of the parameter sequence $\boldsymbol \theta_N$ is the same for each $N$ (i.e., $\boldsymbol \theta_N$ lies in a parameter space of fixed Euclidean dimension $k$).
+This aspect may not be true for other types of FSFS models considered in Sections \ref{restricted-boltzmann-machines}-\ref{deep-learning}.
+We also note that @schweinberger2011instability considered instability in such exponential models (e.g., for random graphs) for but with fixed parameter sequences $\theta_N=\lambda\in\mathbb{R}^k$, $N \geq 1$, of non-varying length $k$.
 
@@
 
-A restricted Boltzmann machine (RBM) is an undirected graphical model specified for discrete or continuous random variables, binary variables being most common [@smolensky1986information].
+A restricted Boltzmann machine (RBM) is an undirected graphical model specified for discrete or continuous random variables, binary variables being most common [cf.
+@smolensky1986information].
 A RBM architecture has two layers, hidden ($\mathcal{H}$) and visible ($\mathcal{V}$), with conditional independence within each layer.
@@
 $$
-p_{\boldsymbol \theta} (\tilde{\boldsymbol x}) = \exp\left[ \boldsymbol \alpha^T \boldsymbol h + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^T \Gamma \boldsymbol x - \psi(\boldsymbol \theta)\right], \qquad \tilde{\boldsymbol x} = (\boldsymbol h, \boldsymbol x) \in \mathcal{X}^{N+N_H}
+p_{\boldsymbol \theta} (\tilde{\boldsymbol x}) = \exp\left[ \boldsymbol \alpha^T \boldsymbol h + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^T \Gamma \boldsymbol x - \psi(\boldsymbol \theta)\right], \qquad \tilde{\boldsymbol x} = (\boldsymbol h, \boldsymbol x) \in \mathcal{X}^{N+N_H},
 $$
@@
 $$
-\psi(\boldsymbol \theta) = \log \sum_{\tilde{\boldsymbol x} \in \mathcal{X}^{N + N_H}} \exp\left[ \boldsymbol \alpha^T \boldsymbol h + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^T \Gamma \boldsymbol x\right], \qquad \boldsymbol \theta \in \Theta
+\psi(\boldsymbol \theta) = \log \sum_{\tilde{\boldsymbol x} \in \mathcal{X}^{N + N_H}} \exp\left[ \boldsymbol \alpha^T \boldsymbol h + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^T \Gamma \boldsymbol x\right], \qquad \boldsymbol \theta \in \Theta,
 $$ 
@@
 $$
-p_{\boldsymbol \theta} (\boldsymbol x) = \sum\limits_{\boldsymbol h \in \mathcal{X}^{N_H}} p_{\boldsymbol \theta} (\boldsymbol x, \boldsymbol h), \qquad \boldsymbol x \in \mathcal{X}^N
+p_{\boldsymbol \theta} (\boldsymbol x) = \sum\limits_{\boldsymbol h \in \mathcal{X}^{N_H}} p_{\boldsymbol \theta} (\boldsymbol x, \boldsymbol h), \qquad \boldsymbol x \in \mathcal{X}^N.
 $$
 
-Note that the vector of model parameters $\boldsymbol \theta_N$, of size $q(N)$, is expanding in size as a function of the visible variable dimension $N$, and one may choose the number $N_H$ of hidden variables as an arbitrary function of $N$ as well (which may also potentially increase with $N$).
-Additionally, $p_{\boldsymbol \theta}(\boldsymbol x) > 0$ for all $\boldsymbol x \in \mathcal{X}^N$ by properties of the exponential function so that the visible RBM model specification is an example of a FSFS model.
-This example also indicates that models formed by marginalizing a base FSFS model (e.g., exponential family) is again a FSFS model class.
+Note that the vector of model parameters $\boldsymbol \theta_N$, of size $q(N)$, is expanding in size as a function of sample dimension $N$ to accomodate the dimension of visible variables $X_1, \dots, X_N$, and one may further choose the number $N_H$ of hidden variables to vary as an arbitrary function of $N$ as well; in particular, the number $N_H$ of hideen may also potentially increase with $N$.
+Additionally, as $|\mathcal{X}| = 2$ and $p_{\boldsymbol \theta}(\boldsymbol x) > 0$ for all $\boldsymbol x \in \mathcal{X}^N$ by properties of the exponential function, the visible RBM model specification is a FSFS model.
+This example also indicates that models formed by marginalizing a base FSFS model (e.g., a type of exponential family) is again a FSFS model class.
 
@@
 Consider two models with "deep architechture" that contain multiple hidden (or latent) layers in addition to a visible layer of data, a deep Boltzmann machine [@salakhutdinov2009deep] and a deep belief network [@hinton2006fast].
-Let $M$ denote the number of hidden layers included in the model and $N_{H,1}, \dots, N_{H, M}$ the number of hidden variables within each hidden layer.
-Then let $\tilde{\boldsymbol X} = \{H^{(1)}_1, \dots, H^{(1)}_{N_{H,1}}, \dots, H^{(M)}_1, \dots, H^{(M)}_{N_{H,M}}, \boldsymbol X\}$ be random variables corresponding to the hidden and visible $\boldsymbol X =(X_1,\ldots,X_N)$ variables in a deep probabilitic model.
-We will consider the sample space $\mathcal{X} = \{-1,1\}$.
+Let $M$ denote the number of hidden layers included in the model and $N_{(H,1)}, \dots, N_{((H,M))}$ the number of hidden variables within each hidden layer.
+Then let $\tilde{\boldsymbol X} = \{H^{(1)}_1, \dots, H^{(1)}_{N_{(H,1)}}, \dots, H^{(M)}_1, \dots, H^{(M)}_{N_{(H,M)}}, \boldsymbol X\}$ be random variables corresponding to the hidden variables $\{ H_{i}^{(j)} : i=1,\ldots, N_{(H,j)}, j=1,\ldots,M\}$ and visible variables $\boldsymbol X =(X_1,\ldots,X_N)$ in a deep probabilitic model.
+Each variable outcome will again lie in $\mathcal{X} = \{-1,1\}$.
 
@@
 $$
-where $\tilde{\boldsymbol x} = (\boldsymbol h^{(1)}, \dots, \boldsymbol h^{(M)}, \boldsymbol x) \in \mathcal{X}^{N_{H,1} + \cdots + N_{H,M} +N}$ and
+for $\tilde{\boldsymbol x} = (\boldsymbol h^{(1)}, \dots, \boldsymbol h^{(M)}, \boldsymbol x) \in \mathcal{X}^{N_{(H,1)} + \cdots + N_{(H,M)} +N}$ where
 $$
-\psi(\boldsymbol \theta) = \log \sum\limits_{\tilde{\boldsymbol x} \in \mathcal{X}^{N_{H,1} + \cdots + N_{H,M} +N}} \exp\left[ \sum\limits_{i = 1}^M\boldsymbol \alpha^{(i)T} \boldsymbol h^{(i)} + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^{(1)T} \Gamma^{(0)} \boldsymbol x + \sum\limits_{i = 1}^{M - 1} \boldsymbol h^{(i)T} \Gamma^{(i)} \boldsymbol h^{(i + 1)}\right], \qquad \boldsymbol \theta \in \Theta
+\psi(\boldsymbol \theta) = \log \sum\limits_{\tilde{\boldsymbol x} \in \mathcal{X}^{N_{(H,1)} + \cdots + N_{(H,M)} +N}} \exp\left[ \sum\limits_{i = 1}^M\boldsymbol \alpha^{(i)T} \boldsymbol h^{(i)} + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^{(1)T} \Gamma^{(0)} \boldsymbol x + \sum\limits_{i = 1}^{M - 1} \boldsymbol h^{(i)T} \Gamma^{(i)} \boldsymbol h^{(i + 1)}\right], \qquad \boldsymbol \theta \in \Theta
 $$ 
-is the normalizing function, $\boldsymbol \beta \in \mathbb{R}^V$, $\boldsymbol \alpha^{(i)} \in \mathbb{R}^{N_{H,i}}$ for $i = 1, \dots, M$, $\Gamma^{(0)}$ is a matrix with dimension $N_{H,1} \times N$, and $\Gamma^{(i)}$ is a matrix with dimension $N_{H,i} \times N_{H,i+1}$ for $i = 1, \dots, M-1$ are the parameters in the model.
-Let $\boldsymbol \theta = (\boldsymbol \alpha^{(1)}, \dots, \boldsymbol \alpha^{(M)}, \boldsymbol \beta,\Gamma^{(0)}, \dots, \Gamma^{(M - 1)}) \in \Theta \subset \mathbb{R}^{q(N)}$ denote the combined vector of parameters with $q(N)= N_{H,1}+\cdots N_{M,H} + N + N_{H,1}*N+N_{H,2}*H_{H,1}+\cdots +N_{H,M}*H_{H,M-1}$.
+is the normalizing functionand parameters in the model are given by $\boldsymbol \beta \in \mathbb{R}^V$, $\boldsymbol \alpha^{(i)} \in \mathbb{R}^{N_{(H,i)}}$ for $i = 1, \dots, M$, along with a matrix $\Gamma^{(0)}$ of dimension $N_{(H,1)} \times N$, and matrices $\Gamma^{(i)}$ of dimension $N_{(H,i)} \times N_{(H,i+1)}$ for $i = 1, \dots, M-1$.
+Let $\boldsymbol \theta = (\boldsymbol \alpha^{(1)}, \dots, \boldsymbol \alpha^{(M)}, \boldsymbol \beta,\Gamma^{(0)}, \dots, \Gamma^{(M - 1)}) \in \Theta \subset \mathbb{R}^{q(N)}$ denote the combined vector of parameters with total length $q(N)= N_{(H,1)}+\cdots N_{(H,M)} + N + N_{(H,1)}*N+N_{H,2}*H_{(H,1)}+\cdots +N_{(H,M)}*H_{(H,M)-1}$.
 
@@
 $$
-p_{\boldsymbol \theta} (\boldsymbol x) = \sum\limits_{(\boldsymbol h^{(1)}, \dots, \boldsymbol h^{(M)}) \in \mathcal{X}^{N_{H,1} + \cdots + N_{H,M}}} p_{\boldsymbol \theta} (\tilde{\boldsymbol x}), \qquad \boldsymbol x \in \mathcal{X}^N
+p_{\boldsymbol \theta} (\boldsymbol x) = \sum\limits_{(\boldsymbol h^{(1)}, \dots, \boldsymbol h^{(M)}) \in \mathcal{X}^{N_{(H,1)} + \cdots + N_{(H,M)}}} p_{\boldsymbol \theta} (\tilde{\boldsymbol x}), \qquad \boldsymbol x \in \mathcal{X}^N
 $$
@@
 
-**Deep belief network (DBN).** A DBN is similar to a DBM in that there are multiple layers of latent random variables stacked in a deep architecture with no conditional dependence between layers.
+**Deep belief network (DBN).** A DBN resembles a DBM in that there are multiple layers of latent random variables stacked in a deep architecture with no conditional dependence between layers.
 The difference between the two models is that all but the last stacked model are Bayesian networks, rather than RBMs.
-Thus for visibles $X_1, \dots, X_N$ with support equal to the sample space $\mathcal{X}^V$, a DBN is also a FSFS model if the number $q(N)$ of components in the parameter vector is dependent on the dimension of the visibles.
-Commonly, as in a logistic belief net [@neal1992connectionist], a "weight" parameter is placed on each interaction between visibles, $X_1, \dots, X_N$ and the first layer of latent variables, $H^{(1)}_1, \dots, H^{(1)}_{N_{H,1}}$, satisfying the definition of a FSFS model.
+Thus for visibles $X_1, \dots, X_N$ with support equal to the sample space $\mathcal{X}^N$, a DBN is also a FSFS model if the number $q(N)$ of components in the parameter vector is dependent on the dimension of the visibles.
+Commonly, as in a logistic belief net [@neal1992connectionist], a "weight" parameter is placed on each interaction between visibles, $X_1, \dots, X_N$ and the first layer of latent variables, $H^{(1)}_1, \dots, H^{(1)}_{N_{(H,1)}}$, satisfying the definition of a FSFS model.
 
@@
 
-To measure "instability" in a FSFS model, it is useful to imagine the behavior of a data model sequence $P_{\theta_N}$ specified to incorporate more and more observations.
-A relevant quantity to this end is a (scaled) extremal log-probability ratio given by
+To define or measure instability in a FSFS model, it is useful to imagine the behavior of a data model sequence $P_{\theta_N}$ specified to incorporate more and more observations to reach a sample length $N$.
+A relevant quantity to this end is a (scaled) extremal log-probability ratio (ELPR) given by
 \begin{align}
@@
 \begin{align*}
+\label{defn:instab}
 \lim\limits_{N \rightarrow \infty} \frac{1}{N} \text{ELPR}(\boldsymbol \theta_N) \equiv \lim\limits_{N \rightarrow \infty} \frac{1}{N} \log \left[\frac{\max\limits_{(x_1, \dots, x_N) \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(x_1, \dots, x_N)}{\min\limits_{(x_1, \dots, x_N) \in \mathcal{X}^N}P_{\boldsymbol \theta}(x_1, \dots, x_N)}\right] = \infty.
@@
 RBM and DBM models in Sections \ref{restricted-boltzmann-machines}-\ref{deep-learning}) and an increasing number of parameters.
-While this definition differs in form, it does match that in @schweinberger2011instability for the special exponential models (cf.
+While this definition differs in form and scope, it does match that in @schweinberger2011instability for the special exponential models (cf.
 Section \ref{discrete-exponential-family-models}) considered there.
@@
 Additionally, S-unstable FSFS models are connected to degenerate models, where *degeneracy* implies placing all probability on a small portion of the sample space.
-For perspective, note that instability in FSFS models, as measured by (\ref{eqn:elpr}), is a spectrum and Proposition \ref{prop:instab} indicates increasing and undesirable sensitivity to model probabilities (i.e., in one component changes to outcomes) as (\ref{eqn:elpr}) increases.
+For perspective, note that instability in FSFS models, as measured by ELPR in (\ref{eqn:elpr}), is a spectrum and Proposition \ref{prop:instab} indicates increasing and undesirable sensitivity of model probabilities (i.e., for one component changes in outcomes) as (\ref{eqn:elpr}) increases.
 Furthermore, as the instability measure (\ref{eqn:elpr}) grows, FSFS models are guaranteed to slide into full degeneracy as Proposition \ref{prop:degen} next shows.
@@
 $$
-of possible outcomes, for a given $0 < \epsilon < 1$, and along a sequence $\boldsymbol \theta_N$, $N \ge 1$, of FSFS parameters.
+of possible outcomes, for a given $0 < \epsilon < 1$, along a sequence $\boldsymbol \theta_N$, $N \ge 1$, of FSFS parameters.
 
@@
 \label{prop:degen}
-For an unstable FSFS model, and for any given $0 < \epsilon < 1$, $P_{\boldsymbol \theta_N}\left((x_1, \dots, x_N) \in M_{\epsilon, \boldsymbol \theta_N}\right) \rightarrow 1$ holds as $N \rightarrow \infty$.
+For an unstable FSFS model in Definition \ref{defn:instab}, and for any given $0 < \epsilon < 1$, 
+$$
+P_{\boldsymbol \theta_N}\left((x_1, \dots, x_N) \in M_{\epsilon, \boldsymbol \theta_N}\right) \rightarrow 1 \text{ as } N \rightarrow \infty.
+$$
 \end{prop}
 
-In other words, in S-unstable FSFS models, all probability in the FSFS model formulation with a large number of random variables will mass on mode sets of arbitrarily small size, or simply those few outcomes with the highest probability.
+In other words, in S-unstable FSFS models, all probability in the model formulation with a large number of random variables will mass on mode sets of arbitrarily small size or simply on those few outcomes with the highest probability.
 
 \begin{remark}
-There is a further generalization to the notion of "unstable" used in Schweinberger (2011) that deals with replication.
-One can conceivably imagine $n$ independent and identically distributed (iid) replications $\boldsymbol X_1, \dots, \boldsymbol X_n$, where each $\boldsymbol X_i=(X_{i,1},\ldots,X_{i,N}) \in \mathcal{X}^N$ follows a common FSFS model with probabilities $P_{\theta_N}(\boldsymbol x)>0$,  $\boldsymbol x\in\mathcal{X}^N$ and $|\mathcal{X}<\infty|$, for $i=1,\ldots,n$.
-This leads to a total of $n*N$ random variables in the joint FSFS model.
+There is a further generalization the notion of instability in Definition \ref{defn:instab} to address independent replications of data sequences.
+That is, one might consider data as $n$ independent and identically distributed replications $\boldsymbol X_1, \dots, \boldsymbol X_n$, where each $\boldsymbol X_i=(X_{i,1},\ldots,X_{i,N}) \in \mathcal{X}^N$ follows a common FSFS model with probabilities $P_{\theta_N}(\boldsymbol x)>0$,  $\boldsymbol x\in\mathcal{X}^N$ and $|\mathcal{X}|<\infty$, for $i=1,\ldots,n$.
+This leads to a total of $n*N$ random variables in the joint model.
 However, the definition of S-unstable and Propositions \ref{prop:instab}-\ref{prop:degen} still hold for such iid replications.
@@
 Thus,
-\begin{align*}
-\frac{\text{extremal log-probability ratio}}{\text{\# random variables in the model}} &\equiv \frac{1}{n*N}\log \left[\frac{\left\{\max\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)\right\}^n}{\left\{\min\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta}(\boldsymbol x)\right\}^n}\right] \\
-&= \frac{1}{N} \log \left[\frac{\max\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)}{\min\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta}(\boldsymbol x)}\right]\\
-&= \frac{1}{N} \text{ELPR}(\boldsymbol \theta_N),
-\end{align*}
+$$
+\frac{\text{extremal log-probability ratio}}{\text{\# random variables in the model}} \equiv \frac{1}{n*N}\log \left[\frac{\left(\max\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)\right)^n}{\left(\min\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta}(\boldsymbol x)\right)^n}\right] = \frac{1}{N} \log \left[\frac{\max\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)}{\min\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta}(\boldsymbol x)}\right] = \frac{1}{N} \text{ELPR}(\boldsymbol \theta_N),
+$$
 implying that the definition of an S-unstable FSFS model is invariant to the level $n$ of independent replication.
+Consequently, overall model instabilities may be characterized by those of one replication from the common FSFS model.
 \end{remark}
@@
 \begin{appendix}
-\section{Appendices}
+\section{Appendix: Proofs of main results}
 \label{appendix}
-\subsection{Proofs of main results}
 \label{proofs}

### 2016-11-23 writing/appendix/appendix-note.Rmd
vardeman's note edits

@@
 Note there exists a sequence $\boldsymbol x_{min} \equiv \boldsymbol x_0, \boldsymbol x_1, \dots, \boldsymbol x_k \equiv \boldsymbol x_{max}$ in $\mathcal{X}^N$ of component-wise switches to move from $\boldsymbol x_{min}$ to $\boldsymbol x_{max}$ in the sample space (i.e.
-$\boldsymbol x_i, \boldsymbol x_{i + 1} \in \mathcal{X}^N$ differ by exactly $1$ component for $i = 0, \dots, k$) for some integer $k \in \{0, 1, \dots, N\}$.
+$\boldsymbol x_i, \boldsymbol x_{i + 1} \in \mathcal{X}^N$ differ in exactly $1$ component for $i = 0, \dots, k$) for some integer $k \in \{0, 1, \dots, N\}$.
 Under the FSFS model, recall $P_{\boldsymbol \theta_N}(\boldsymbol x) > 0$ holds so that $\log P_{\boldsymbol \theta_N}(\boldsymbol x)$ is well-defined for each outcome $\boldsymbol x \in \mathcal{X}^N$.

### 2016-11-23 writing/note.Rmd
vardeman's note edits

@@
   A probability model exhibits instability if small changes in a data outcome result in large, and often unanticipated, changes in probability.
-For correlaated data structures found in several application areas, there is increasing interest in identifying instability in potential model specifications for such data.
-We consider the problem of quantifying instability for general probability models defined on sequences of observations, where each sequence of a given length $N$ has a finite number of possible outcomes.
-As the number of available observations increases, a sequence of probability models results to accommodate the data of expanding dimension.
-  Model instability is formally shown to occur when a certain log-probability ratio under such models grows faster than the length $N$ of the data sequence.
+For correlated data structures found in several application areas, there is increasing interest in predicting/identifying instability.
+We consider the problem of quantifying instability for general probability models defined on sequences of observations, where each sequence of length $N$ has a finite number of possible outcomes.
+(A sequence of probability models results indexed by $N$ accommodates data of expanding dimension.) Model instability is formally shown to occur when a certain log-probability ratio under such models grows faster than $N$.
 In this case, a one component change in the data sequence can provably shift probability by orders of magnitude.
-Also, as the measure of instability becomes more extreme, the resulting probability models are shown  to slip into degeneracy, thereby placing all probability on arbitrarily small portions of the sample space.
+Also, as a measure of instability becomes more extreme, the resulting probability models are shown  to tend to degeneracy, placing all their probability on arbitrarily small portions of the sample space.
 These results on instability apply to large classes of models commonly used in random graphs, network analysis, and machine learning contexts.
@@
 
-We consider the behavior, and the potential impropriety, of probability models built to incorporate a sequence of discrete observations with an arbitrarily specified length $N$.
-Let $(X_1, \dots X_N)$ denote a set of sequence of discrete random variables with a finite sample space, $\mathcal{X}^N$.
-That is, $\mathcal{X}$ with $|\mathcal{X}| < \infty$ represents a finite set of potential outcomes for the variable $X_i$, and the data sequence $(X_1,\ldots,X_N)$ assumes values in the $N$-fold product space $\mathcal{X}^N$.
-For any $N$, let $P_{\boldsymbol \theta_N}$ denote a probability model on $\mathcal{X}^N$, under which $P_{\boldsymbol \theta_N}(x_1, \dots, x_N) > 0$ is the probability of the data outcome $(x_1, \dots, x_N) \in \mathcal{X}^N$.
-We assume that the model support of $P_{\boldsymbol \theta_N}$ is equal to the sample space $\mathcal{X}^N$.
-This framework produces a series $P_{\boldsymbol \theta_N}$ of probability models, indexed by a generic sequence of parameters $\boldsymbol \theta_N$, to describe data of any given length $N \geq 1$.
-The size and structure of such parameters may be generally specified, and one may intuitvely consider Euclidean $\boldsymbol \theta_N \in \mathbb{R}^{q(N)}$ for some arbitrary integer-valued function $q(\cdot) \geq 1$.
-We will call this class *Finitely Supported Finite Sequence (FSFS) models*.
+We consider the behavior, and the potential impropriety, of probability models built to incorporate a sequence of discrete observations with length $N$.
+Let $(X_1, \dots, X_N)$ denote a set of sequence of discrete random variables with a finite sample space, $\mathcal{X}^N$.
+That is, $\mathcal{X}$ with $|\mathcal{X}| < \infty$ represents a finite set of potential outcomes for the variable $X_i$, and the data sequence $(X_1,\ldots,X_N)$ takes values in the $N$-fold product space $\mathcal{X}^N$.
+For each $N$, let $P_{\boldsymbol \theta_N}$ denote a probability model on $\mathcal{X}^N$, under which $P_{\boldsymbol \theta_N}(x_1, \dots, x_N) > 0$ is the probability of the data outcome $(x_1, \dots, x_N) \in \mathcal{X}^N$.
+We assume that the model support of $P_{\boldsymbol \theta_N}$ is the sample space $\mathcal{X}^N$.
+This framework produces a series $P_{\boldsymbol \theta_N}$ of probability models, indexed by a generic sequence of parameters $\boldsymbol \theta_N$, to describe data of each length $N \geq 1$.
+(The size and structure of such parameters are without restriction, and natural cases include those where $\boldsymbol \theta_N \in \mathbb{R}^{q(N)}$ for some arbitrary integer-valued function $q(\cdot) \geq 1$.) We will call this model class *Finitely Supported Finite Sequence (FSFS) models*.
 
 Section \ref{examples} provides several examples of FSFS models commonly used in graph/network analysis and machine learning (i.e., deep learning models).
-Section \ref{instability-results} establishes formal results regarding the model propriety of FSFS models with regard to stability.
-A FSFS probability model intuitively exhibits instability if small changes in the components of a data outcome $(x_1,\ldots,x_N)$ may result in large changes in probability $P_{\theta_N}(x_1,\ldots,x_N)$.
-The concept of instability, introduced in the field statistical physics by @ruelle1999statistical, was extended to include a notion of detection and quantification for certain exponential
+Section \ref{instability-results} establishes formal results regarding the propriety of FSFS models with regard to stability.
+A FSFS probability model sequence exhibits instability if small changes in the components of a data outcome $(x_1,\ldots,x_N)$ can result in large changes in probability $P_{\theta_N}(x_1,\ldots,x_N)$.
+The concept of instability, introduced in the field of statistical physics by @ruelle1999statistical, was extended to include a notion of detection and quantification for certain exponential
 family models by @schweinberger2011instability.
@@
 As described by @schweinberger2011instability, model instability and model degeneracy are related by viewing degeneracy as an extreme or limiting form of instability.
-However, the instability results of @schweinberger2011instability are only intended for a special parametric case of exponential family models.
-In considering the broad framework of FSFS models, the main results here develop a more general measure of model instability, which is defined to arise when certain maximal probabilities in FSFS models are too extreme relative to the length $N$ of the data sample.
-In this case, a one component change in the data sequence
-may provably shift probability by orders of magnitude, and FSFS models are rigorously shown to become degenerate as the measure of instability increases.
+The instability results of @schweinberger2011instability were developed for the case of discrete exponential family models.
+The main results here concern a general measure of model instability, appropriate accross the whole FSFS model class.
+This can be used to identify when certain maximal probabilities in FSFS models are too extreme relative to the length $N$.
+In this case, a one component change in the data sequence may provably shift probability by orders of magnitude, and FSFS models are rigorously shown to become degenerate as the measure of instability increases.
 Lastly, Section \ref{implications} emphasizes the implications of our model propriety results and proofs of the main results appear in the Appendix.
@@
 
-Many model families can fall under the umbrella of FSFS models. For illustration, this section presents three specific examples of FSFS models, including models with deep architectures.
+Many model families fall under the umbrella of FSFS models. For illustration, this section presents three specific examples of FSFS models, including models with deep architectures.
 
@@
 
-For $X_1, \dots, X_N$ discrete random variables with sample space $\mathcal{X}^N$, $|\mathcal{X}| < \infty$, consider an exponential family model with probability mass function of the form
+For $\boldsymbol X = X_1, \dots, X_N$ discrete random variables with sample space $\mathcal{X}^N$, $|\mathcal{X}| < \infty$, consider an exponential family model for $\boldsymbol X$ with probability mass function of the form
 $$
@@
 $$ 
-where $\boldsymbol \lambda \in \Lambda \subset \mathbb{R}^{k}$, $\boldsymbol \eta : \mathbb{R}^k \mapsto \mathbb{R}^L$ is a vector of natural parameters, $\boldsymbol g : \mathcal{X}^N \mapsto \mathbb{R}^L$ is a vector of sufficient statistics, positive integers $k$ and $L$ denote the dimensions of the parameter and natural parameters spaces, respectfully,
+for fixed positive integers $k$ and $L$ denoting the dimensions of the parameter and natural parameters spaces respectively, $\boldsymbol \lambda \in \Lambda \subset \mathbb{R}^{k}$, $\boldsymbol \eta : \mathbb{R}^k \mapsto \mathbb{R}^L$, $\boldsymbol g : \mathcal{X}^N \mapsto \mathbb{R}^L$ a vector of sufficient statistics,
 $$
@@
 $$
-is the normalizing function, and $\Lambda = \{\boldsymbol \lambda \in \mathbb{R}^k : \psi(\boldsymbol \lambda) < \infty, k \le q(N) \}$ is the parameter space.
+the normalizing function, and $\Lambda = \{\boldsymbol \lambda \in \mathbb{R}^k : \psi(\boldsymbol \lambda) < \infty, k \le q(N) \}$ is the parameter space.
 
-Defining $\boldsymbol \theta_N=\lambda_N$ to a subsequence of $\Lambda \subset \mathbb{R}^k$ and noting that $p_{\boldsymbol \lambda}(\boldsymbol x) > 0$ for all $\boldsymbol x \in \mathcal{X}^N$ by properties of the exponential function, the discrete exponential family models defined in this section are a special case of the FSFS model.
+Defining $\boldsymbol \theta_N=\lambda_N$ to a sequence of elements of $\Lambda \subset \mathbb{R}^k$ and noting that $p_{\boldsymbol \lambda}(\boldsymbol x) > 0$ for all $\boldsymbol x \in \mathcal{X}^N$, these discrete exponential family models are special cases of the FSFS models.
 Such exponential models arise with spatial data on a lattice [@besag1974spatial], network data [@wasserman1994social; @handcock2003assessing], and even binomial sampling with $N$ iid Bernoulli random variables.
-For these exponential models, the length of the parameter sequence $\boldsymbol \theta_N$ is the same for each $N$ (i.e., $\boldsymbol \theta_N$ lies in a parameter space of fixed Euclidean dimension $k$).
-This aspect may not be true for other types of FSFS models considered in Sections \ref{restricted-boltzmann-machines}-\ref{deep-learning}.
-We also note that @schweinberger2011instability considered instability in such exponential models (e.g., for random graphs) for but with fixed parameter sequences $\theta_N=\lambda\in\mathbb{R}^k$, $N \geq 1$, of non-varying length $k$.
+For these exponential models, the dimension of the parameter $\boldsymbol \theta_N$ is the same for each $N$ (i.e., $\boldsymbol \theta_N$ lies in a parameter space of fixed Euclidean dimension $k$).
+This need not be true for other types of FSFS models considered in Sections \ref{restricted-boltzmann-machines}-\ref{deep-learning}.
+@schweinberger2011instability considered instability in such exponential models (e.g., for random graphs) for sequences of fixed parameters $\theta_N=\lambda\in\mathbb{R}^k$, $N \geq 1$, of non-varying dimension $k$.
 
@@
 Let $\boldsymbol X=(X_1,\ldots,X_N)$ denote the $N$ random variables for visibles with support $\mathcal{X}^N$ and $\boldsymbol H=(H_1,\ldots,H_{N_H})$ denote the $N_H$ random variables for hiddens with support $\mathcal{X}^{N_H}$ where  $\mathcal{X} = \{-1,1\}$.
-The RBM for $\tilde{\boldsymbol X}=(\boldsymbol X,\boldsymbol H)$ then has the joint probability mass function
+For parameters $\boldsymbol \alpha \in \mathbb{R}^{N_H}$, $\boldsymbol \beta \in \mathbb{R}^N$, and $\Gamma$ is a matrix with dimension $N_H \times N$, the RBM for $\tilde{\boldsymbol X}=(\boldsymbol X,\boldsymbol H)$ then has the joint probability mass function
 $$
@@
 $$ 
-is the normalizing function, $\boldsymbol \alpha \in \mathbb{R}^{N_H}$, $\boldsymbol \beta \in \mathbb{R}^N$, and $\Gamma$ is a matrix with dimension $N_H \times N$ are the parameters in the model.
-Let $\boldsymbol \theta_N = (\boldsymbol \alpha, \boldsymbol \beta,\Gamma) \in \Theta \subset \mathbb{R}^{q(N)}$ with $q(N) = N + N_H + N*N_H$ denote the vector of parameters.
+is the normalizing function.
+Let $\boldsymbol \theta_N = (\boldsymbol \alpha, \boldsymbol \beta,\Gamma) \in \Theta \subset \mathbb{R}^{q(N)}$ with $q(N) = N + N_H + N*N_H$ denote the vector of parameters for the RBM.
 
@@
 
-Note that the vector of model parameters $\boldsymbol \theta_N$, of size $q(N)$, is expanding in size as a function of sample dimension $N$ to accomodate the dimension of visible variables $X_1, \dots, X_N$, and one may further choose the number $N_H$ of hidden variables to vary as an arbitrary function of $N$ as well; in particular, the number $N_H$ of hideen may also potentially increase with $N$.
-Additionally, as $|\mathcal{X}| = 2$ and $p_{\boldsymbol \theta}(\boldsymbol x) > 0$ for all $\boldsymbol x \in \mathcal{X}^N$ by properties of the exponential function, the visible RBM model specification is a FSFS model.
-This example also indicates that models formed by marginalizing a base FSFS model (e.g., a type of exponential family) is again a FSFS model class.
+Note that the vector of model parameters $\boldsymbol \theta_N$, of size $q(N)$, grows in size as a function of sample dimension $N$ to accomodate the dimension of visible variables $X_1, \dots, X_N$, and one may further choose the number $N_H$ of hidden variables to change with $N$ as well; in particular, the number $N_H$ of hidens may also potentially increase with $N$.
+Additionally, as $|\mathcal{X}| = 2$ and $p_{\boldsymbol \theta}(\boldsymbol x) > 0$ for all $\boldsymbol x \in \mathcal{X}^N$, the RBM model specification for visibles is a FSFS model.
+This example also indicates that models formed by marginalizing a base FSFS model (e.g., a type of exponential family model) is again a FSFS model class.
 
@@
 
-**Deep Boltzmann machine (DBM).** The DBM class of models maintains conditional independence within all layers in the model by stacking RBM models and maintaining the restriction of only allowing conditional dependence between layers.
+**Deep Boltzmann machine (DBM).** The DBM class of models maintains conditional independence within all layers in the model by stacking RBM models and maintaining the restriction of only allowing conditional dependence between neighboring layers.
 The joint probability mass function for a DBM is
@@
 $$ 
-is the normalizing functionand parameters in the model are given by $\boldsymbol \beta \in \mathbb{R}^V$, $\boldsymbol \alpha^{(i)} \in \mathbb{R}^{N_{(H,i)}}$ for $i = 1, \dots, M$, along with a matrix $\Gamma^{(0)}$ of dimension $N_{(H,1)} \times N$, and matrices $\Gamma^{(i)}$ of dimension $N_{(H,i)} \times N_{(H,i+1)}$ for $i = 1, \dots, M-1$.
+is the normalizing function and parameters in the model are $\boldsymbol \beta \in \mathbb{R}^V$, $\boldsymbol \alpha^{(i)} \in \mathbb{R}^{N_{(H,i)}}$ for $i = 1, \dots, M$, along with a matrix $\Gamma^{(0)}$ of dimension $N_{(H,1)} \times N$, and matrices $\Gamma^{(i)}$ of dimension $N_{(H,i)} \times N_{(H,i+1)}$ for $i = 1, \dots, M-1$.
 Let $\boldsymbol \theta = (\boldsymbol \alpha^{(1)}, \dots, \boldsymbol \alpha^{(M)}, \boldsymbol \beta,\Gamma^{(0)}, \dots, \Gamma^{(M - 1)}) \in \Theta \subset \mathbb{R}^{q(N)}$ denote the combined vector of parameters with total length $q(N)= N_{(H,1)}+\cdots N_{(H,M)} + N + N_{(H,1)}*N+N_{H,2}*H_{(H,1)}+\cdots +N_{(H,M)}*H_{(H,M)-1}$.
 
-The probability mass function for the visible random variables $X_1, \dots, X_N$ follows from this joint specification:
+The probability mass function for the visible random variables $X_1, \dots, X_N$ follows from this joint specification as
 $$
@@
 
-Again similar to the RBM case, the visible DBM model specification is an example of a FSFS model.
+Again like the RBM case, the visible DBM model specification is an example of a FSFS model.
 
 **Deep belief network (DBN).** A DBN resembles a DBM in that there are multiple layers of latent random variables stacked in a deep architecture with no conditional dependence between layers.
-The difference between the two models is that all but the last stacked model are Bayesian networks, rather than RBMs.
-Thus for visibles $X_1, \dots, X_N$ with support equal to the sample space $\mathcal{X}^N$, a DBN is also a FSFS model if the number $q(N)$ of components in the parameter vector is dependent on the dimension of the visibles.
-Commonly, as in a logistic belief net [@neal1992connectionist], a "weight" parameter is placed on each interaction between visibles, $X_1, \dots, X_N$ and the first layer of latent variables, $H^{(1)}_1, \dots, H^{(1)}_{N_{(H,1)}}$, satisfying the definition of a FSFS model.
+The difference between the DBM and DBN models is that all but the last stacked layer in a DBN are Bayesian networks [see @pearl985bayesian], rather than RBMs.
+Thus for visibles $X_1, \dots, X_N$ with support $\mathcal{X}^N$, a DBN is also a FSFS model if the number $q(N)$ of components in the parameter vector is dependent on the dimension of the visibles.
+Commonly, as in logistic belief nets [@neal1992connectionist], a "weight" parameter is placed on each interaction between visibles, $X_1, \dots, X_N$ and the first layer of latent variables, $H^{(1)}_1, \dots, H^{(1)}_{N_{(H,1)}}$, satisfying the definition of a FSFS model.
 
@@
 
-To define or measure instability in a FSFS model, it is useful to imagine the behavior of a data model sequence $P_{\theta_N}$ specified to incorporate more and more observations to reach a sample length $N$.
-A relevant quantity to this end is a (scaled) extremal log-probability ratio (ELPR) given by
+To define or measure instability in FSFS models, it is useful to consider the behavior of a data model sequence $P_{\theta_N}$.
+A relevant quantity to this end is a (scaled) extremal log-probability ratio (ELPR)
 \begin{align}
@@
 
-The main idea is that, in formulating a FSFS model for potentially increasing numbers of variables (i.e., $N \rightarrow \infty$), the ratio (\ref{eqn:elpr}) should remain bounded, entailing that the largest probability possible under $P_{\theta_N}$ should maintain a magnitude within the smallest probability allowed under the same model.
-Specifically, the log of the fold change should grow at mostly linearly with the sample size $N$.
-This leads to the following definition:
+The main idea is that, in formulating FSFS models for potentially increasing numbers of variables (i.e., for $N \rightarrow \infty$), the ratio (\ref{eqn:elpr}) should remain bounded, requiring that the largest probability possible under $P_{\theta_N}$ should maintain a fixed order of magnitude relative to the smallest probability allowed under the same model.
+Specifically, the log of the ratio should grow at mostly linearly with the sample size $N$.
+This leads to the following definition.
 
 \begin{defn}[S-unstable FSFS]
+\label{defn:instab}
 Let $\boldsymbol \theta_N \in \mathbb{R}^{q(N)}$ be a sequence of FSFS model parameters where the size of the model $q(N)$ is a function of the number of random variables $N$.
-A FSFS model formulation is \emph{Schweinberger-unstable} or \emph{S-unstable} if, as the number of variables increase ($N \rightarrow \infty$), it holds that
+A FSFS model formulation is \emph{Schweinberger-unstable} or \emph{S-unstable} if, as the number of variables increase ($N \rightarrow \infty$),
 \begin{align*}
-\label{defn:instab}
 \lim\limits_{N \rightarrow \infty} \frac{1}{N} \text{ELPR}(\boldsymbol \theta_N) \equiv \lim\limits_{N \rightarrow \infty} \frac{1}{N} \log \left[\frac{\max\limits_{(x_1, \dots, x_N) \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(x_1, \dots, x_N)}{\min\limits_{(x_1, \dots, x_N) \in \mathcal{X}^N}P_{\boldsymbol \theta}(x_1, \dots, x_N)}\right] = \infty.
@@
 
-Note that this definition of *S-unstable* is a generalization or reinterpretation of "unstable" used in @schweinberger2011instability by allowing nonexponential family models (e.g.
+This definition of *S-unstable* is a generalization or reinterpretation of "unstable" used in @schweinberger2011instability by allowing nonexponential family models (e.g.
 RBM and DBM models in Sections \ref{restricted-boltzmann-machines}-\ref{deep-learning}) and an increasing number of parameters.
-While this definition differs in form and scope, it does match that in @schweinberger2011instability for the special exponential models (cf.
+While this definition differs in form and scope from the original, it does match that in @schweinberger2011instability for the special case of exponential models (cf.
 Section \ref{discrete-exponential-family-models}) considered there.
 
-S-unstable FSFS models are undesirable for several reasons. One is that small changes in data can lead to overly-sensitive changes in probability. Consider, for example, 
+S-unstable FSFS model seqences are undesirable for several reasons. One is that small changes in data can lead to overly-sensitive changes in probability. Consider, for example, 
 $$
-\Delta(\boldsymbol \theta_N) \equiv \max \left\{\log \frac{P_{\boldsymbol \theta_N}(x_1, \dots, x_N)}{P_{\boldsymbol \theta_N}(x_1^*, \dots, x_N^*)} : (x_1, \dots, x_N) \& (x_1^*, \dots, x_N^*) \in \mathcal{X}^N \text{ differ by exactly one component}\right\},
+\Delta(\boldsymbol \theta_N) \equiv \max \left\{\log \frac{P_{\boldsymbol \theta_N}(x_1, \dots, x_N)}{P_{\boldsymbol \theta_N}(x_1^*, \dots, x_N^*)} : (x_1, \dots, x_N) \& (x_1^*, \dots, x_N^*) \in \mathcal{X}^N \text{ differ in exactly one component}\right\},
 $$
-the biggest log-probability ratio for a one component change in data outcomes at a FSFS parameter $\boldsymbol \theta_N$. We then have the following non-asymptotic result.
+the biggest log-probability ratio for a one component change in data outcomes at a FSFS parameter $\boldsymbol \theta_N$. We then have the following (non-asymptotic) result.
 
@@
 
-Additionally, S-unstable FSFS models are connected to degenerate models, where *degeneracy* implies placing all probability on a small portion of the sample space.
-For perspective, note that instability in FSFS models, as measured by ELPR in (\ref{eqn:elpr}), is a spectrum and Proposition \ref{prop:instab} indicates increasing and undesirable sensitivity of model probabilities (i.e., for one component changes in outcomes) as (\ref{eqn:elpr}) increases.
-Furthermore, as the instability measure (\ref{eqn:elpr}) grows, FSFS models are guaranteed to slide into full degeneracy as Proposition \ref{prop:degen} next shows.
+Additionally, S-unstable FSFS model sequences are connected to degenerate models, where model *degeneracy* implies placing all probability on a small portion of the sample space.
+For perspective, note that differing bounds on $1/N\cdot\text{ELPR}(\boldsymbol \theta_N$ in (\ref{eqn:elpr}) provide a spectrum of levels of "stability" and Proposition \ref{prop:instab} indicates increasing and undesirable sensitivity of model probabilities (i.e., for one component changes in outcomes) as (\ref{eqn:elpr}) increases.
+Furthermore, as the instability measure (\ref{eqn:elpr}) grows, FSFS model sequences are guaranteed to slide into full degeneracy as Proposition \ref{prop:degen} will show.
 Define a $\epsilon$-modal set
@@
 $$
-of possible outcomes, for a given $0 < \epsilon < 1$, along a sequence $\boldsymbol \theta_N$, $N \ge 1$, of FSFS parameters.
+of possible outcomes, for a given $0 < \epsilon < 1$.
 
@@
 
-In other words, in S-unstable FSFS models, all probability in the model formulation with a large number of random variables will mass on mode sets of arbitrarily small size or simply on those few outcomes with the highest probability.
+In other words, in S-unstable FSFS models, all probability in the model formulation with a large number of random variables will concentrate mass on mode sets of arbitrarily small size or simply on those few outcomes with the highest probability.
 
 \begin{remark}
-There is a further generalization the notion of instability in Definition \ref{defn:instab} to address independent replications of data sequences.
+There is a further generalization the notion of instability in Definition \ref{defn:instab} meant to address independent replications of data sequences.
 That is, one might consider data as $n$ independent and identically distributed replications $\boldsymbol X_1, \dots, \boldsymbol X_n$, where each $\boldsymbol X_i=(X_{i,1},\ldots,X_{i,N}) \in \mathcal{X}^N$ follows a common FSFS model with probabilities $P_{\theta_N}(\boldsymbol x)>0$,  $\boldsymbol x\in\mathcal{X}^N$ and $|\mathcal{X}|<\infty$, for $i=1,\ldots,n$.
@@
 However, the definition of S-unstable and Propositions \ref{prop:instab}-\ref{prop:degen} still hold for such iid replications.
-The reason for this is that $\left(\max\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)\right)^n$ is the largest probability possible under the joint model for the $n$ replications while $\left(\min\limits_{\boldsymbol x \in \mathcal{X}}P_{\boldsymbol \theta_N}(\boldsymbol x)\right)^n$ is the smallest probability.
+This is because $\left(\max\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)\right)^n$ is the largest probability possible under the joint model for the $n$ replications while $\left(\min\limits_{\boldsymbol x \in \mathcal{X}}P_{\boldsymbol \theta_N}(\boldsymbol x)\right)^n$ is the smallest probability.
 Thus,
@@
 $$
-implying that the definition of an S-unstable FSFS model is invariant to the level $n$ of independent replication.
-Consequently, overall model instabilities may be characterized by those of one replication from the common FSFS model.
+implying that the definition of an S-unstable FSFS model sequence is invariant to the level ($n$) of independent replication.
+Consequently, overall model instabilities may be characterized by those of one observation from the common FSFS model.
 \end{remark}
@@
 
-For a large class of models that cover a broad range of applications (including "deep learning"), we have developed a formal definition and elucidated multiple consequences of instability.
-We have shown for FSFS models that instability manifests through small changes in data leading to overly-sensitive changes in probability as well as the potential to place all probability on a small pieace of the sample space.
+For a large class of models that covers a broad range of applications (including "deep learning"), we have developed a formal definition and elucidated multiple consequences of instability.
+We have shown for FSFS models that instability manifests through small changes in data leading to overly-sensitive changes in probability as well as the potential to place all probability on a small piece of the sample space.
 Models that fall within the definition of a FSFS model should be used with caution to ensure that the effects of instability are not experienced.

### 2016-11-23 writing/draft.Rmd
last few vardeman edits

@@
 However, the generative behavior of RBMs is largely unexplored.
-In this paper, we discuss the relationship between RBM parameter specification in the binary case and the prevalence of undesirable model properties such as degeneracy, instability and uninterpretability.
+In this paper, we discuss the relationship between RBM parameter specification in the binary case and the tendency to undesirable model properties such as degeneracy, instability and uninterpretability.
 We also describe the difficulties that arise in likelihood-based and Bayes fitting of such (highly flexible) models, especially as Gibbs sampling (quasi-Bayes) methods are often advocated for the RBM model structure.
@@
 
-where $\boldsymbol \theta \equiv (\theta_{11}, \dots, \theta_{1\nh}, \dots, \theta_{\nv 1}, \dots, \theta_{\nv \nh}, \theta_{v_1}, \dots, \theta_{v_\nv}, \theta_{h_1}, \dots, \theta_{h_\nh}) \in \mathbb{R}^{\nv + \nh + \nv*\nh}$ denotes the vector of model parameters and the denominator $\gamma(\boldsymbol \theta) = \sum_{\boldsymbol x \in \mathcal{C}^{\nh+\nv}}\exp\left(\sum_{i = 1}^\nv \sum_{j=1}^\nh \theta_{ij} v_i h_j + \sum_{i = 1}^\nv\theta_{v_i} v_i + \sum_{j = 1}^\nh\theta_{h_j} h_j\right)$ is the normalizing function that ensures the probabilities (\ref{eqn:pmf}) sum to one.
-For $\boldsymbol x = (h_1, \dots, h_\nh, v_1, \dots, v_\nh) \in \mathcal{C}^{\nv + \nh}$ and
+where $\boldsymbol \theta \equiv (\theta_{11}, \dots, \theta_{1\nh}, \dots, \theta_{\nv 1}, \dots, \theta_{\nv \nh}, \theta_{v_1}, \dots, \theta_{v_\nv}, \theta_{h_1}, \dots, \theta_{h_\nh}) \in \mathbb{R}^{\nv + \nh + \nv*\nh}$ denotes the vector of model parameters and the denominator
+$$\gamma(\boldsymbol \theta) = \sum_{\boldsymbol x \in \mathcal{C}^{\nh+\nv}}\exp\left(\sum_{i = 1}^\nv \sum_{j=1}^\nh \theta_{ij} v_i h_j + \sum_{i = 1}^\nv\theta_{v_i} v_i + \sum_{j = 1}^\nh\theta_{h_j} h_j\right)
+$$ 
+is the normalizing function that ensures the probabilities (\ref{eqn:pmf}) sum to one. For $\boldsymbol x = (h_1, \dots, h_\nh, v_1, \dots, v_\nh) \in \mathcal{C}^{\nv + \nh}$ and 
 \begin{align}
@@
 The RBM model is parameterized by $\boldsymbol \theta$ containing two types of parameters, main effects and interaction effects.
-The main effects parameters ($\{\theta_{v_i}, \theta_{h_j}\}_{\substack{i = 1, \dots, \nv,\\j = 1, \dots, \nh}}$) weight the values of the visible $v_i$ and hidden $h_j$ nodes in probabilities (\ref{eqn:pmf}) and the interaction effects parameters ($\theta_{ij}$) weight the values of the connections $v_i h_j$, or dependencies, between hidden and visible layers.
+The main effects parameters ($\{\theta_{v_i}, \theta_{h_j}\}_{\substack{i = 1, \dots, \nv,\\j = 1, \dots, \nh}}$) weight the values of the visible $v_i$ and hidden $h_j$ nodes in probabilities (\ref{eqn:pmf}) and the interaction effect parameters ($\theta_{ij}$) weight the values of the connections $v_i h_j$, or dependencies, between hidden and visible layers.
 
@@
 visibles given hiddens or vice versa).
-Specifically, the conditional independence of nodes in each layer, given those nodes in the other layer, allows for stepwise simulation of both hidden layers and model parameters possible (e.g., contrastive divergence of @hinton2002training or Bayes methods in Section \ref{model-fitting}).
+Specifically, the conditional independence of nodes in each layer (given those nodes in the other layer) allows for stepwise simulation of both hidden layers and model parameters (e.g., contrastive divergence of @hinton2002training or Bayes methods in Section \ref{model-fitting}).
 
@@
 ## Degeneracy, instability, and uninterpretability... Oh my!
-The highly flexible nature of a RBM (having as it does $\nh + \nv + \nh*\nv$ parameters) creates at least three kinds of potential model impropriety of concern that we will call *degeneracy*, *instability*, and *uninterpretability*.
+The highly flexible nature of a RBM (having as it does $\nh + \nv + \nh*\nv$ parameters) creates at least three kinds of potential model impropriety that we will call *degeneracy*, *instability*, and *uninterpretability*.
 In this section we define these characteristics, consider how to detect them in a RBM, and point out relationships among them.
@@
 @handcock2003assessing classify an exponentially parametrized random graph model at $\boldsymbol \theta$ as *near-degenerate* if the mean value of the vector of sufficient statistics under $\boldsymbol \theta$, $\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta}\boldsymbol t( \boldsymbol X)$, is close to the boundary of $C$.
-This makes sense because if a model is near-degenerate, in the sense that only a small number of elements of the sample space $\mathcal{X}$ have positive probability, the expected value $\text{E}_{\boldsymbol \theta}\boldsymbol t( \boldsymbol X)$ is an average of that same small number of values of $t( \boldsymbol x)$ and can be expected to not be pulled deep into the interior of $C$.
+This makes sense because if a model is near-degenerate in the sense that only a small number of elements of the sample space $\mathcal{X}$ have positive probability, the expected value $\text{E}_{\boldsymbol \theta}\boldsymbol t( \boldsymbol X)$ is an average of that same small number of values of $t( \boldsymbol x)$ and can be expected to not be pulled deep into the interior of $C$.
 
@@
 
-Considering a class of exponential families of distributions, @schweinberger2011instability introduced a concept of model defficiency related to *instability*.
+Considering exponential families of distributions, @schweinberger2011instability introduced a concept of model defficiency related to *instability*.
 Instability can be roughly thought of as excessive sensitivity in the model, where small changes in the components of potential data outcomes, $\boldsymbol x$, may lead to substantial changes in the probability function $f_{\boldsymbol \theta}(\boldsymbol x)$.
@@
 $$
-\Delta(\boldsymbol \theta_\nv) \equiv \max \left\{\log \frac{P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}{P_{\boldsymbol \theta_\nv}(v_1^*, \dots, v_\nv^*)} : (v_1, \dots, v_\nv) \& (v_1^*, \dots, v_\nv^*) \in \mathcal{C}^\nv \text{ differ by exactly one component}\right\},
+\Delta(\boldsymbol \theta_\nv) \equiv \max \left\{\log \frac{P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}{P_{\boldsymbol \theta_\nv}(v_1^*, \dots, v_\nv^*)} : (v_1, \dots, v_\nv) \& (v_1^*, \dots, v_\nv^*) \in \mathcal{C}^\nv \text{ differ in exactly one component}\right\},
 $$
@@
 \label{prop:degen}
-For an S-unstable RBM model sequence and any given $0 < \epsilon < 1$, $P_{\boldsymbol \theta_\nv}\left((v_1, \dots, v_\nv) \in M_{\epsilon, \boldsymbol \theta_\nv}\right) \rightarrow 1$ as $\nv \rightarrow \infty$.
+For an S-unstable RBM model sequence and any $0 < \epsilon < 1$, $P_{\boldsymbol \theta_\nv}\left((v_1, \dots, v_\nv) \in M_{\epsilon, \boldsymbol \theta_\nv}\right) \rightarrow 1$ as $\nv \rightarrow \infty$.
 \end{prop}
 
-In other words, S-unstable RBM model sequences are gwauranteed to stack up all probability on arbitrarily narrow outcomes for visibles.
+In other words, S-unstable RBM model sequences are guaranteed to stack up all probability on an arbitrarily narrow set of outcomes for visibles.
 Proofs of Propositions \ref{prop:instab}-\ref{prop:degen} appear in @kaplan2016note.
-These findings also have counterparts to results in @schweinberger2011instability, but, unlike results there, we do not limit consideration to exponential family forms with a fixed number of parameters.
+These findings also have counterparts in results in @schweinberger2011instability, but unlike results there, we do not limit consideration to exponential family forms with a fixed number of parameters.
 
@@
 $$
-so that, for example, $logit(P_\theta(V_i=1))=\theta_{v_i}$ (or $2 \theta_{v_i}$) under the coding $\mathcal{C}=\{0,1\}$ (or $\{-1,1\}$).
+so that, for example, $\text{logit}(P_\theta(V_i=1))=\theta_{v_i}$ (or $2 \theta_{v_i}$) under the coding $\mathcal{C}=\{0,1\}$ (or $\{-1,1\}$).
 Hence, these main effect parameters have a clear interpretation under an independence model (one with $\theta_{ij} = 0$) but this interpretation can break down as interaction parameters increase in magnitude relative to the size of the main effects.
 In such cases, the main effect parameters $\theta_{v_1}$ and $\theta_{h_j}$ are no longer interpretable as originally intended in the models (statements of marginal means) and the dependence parameters are so large as to dominate the entire model probability structure (also destroying the model interpretation of dependence as local conditional modifications of an overall marginal mean structure).
-To assess which parameter values $\boldsymbol \theta$ may cause difficulties in interpretation, we use the difference $\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] - \text{E}\left[\boldsymbol X | \boldsymbol \theta^* \right]$ between two model expectations: E$\left[\boldsymbol X | \boldsymbol \theta\right]$ at $\boldsymbol \theta$ and expectations E$\left[\boldsymbol X | \boldsymbol \theta^* \right]$ given independence where $\boldsymbol \theta^*$ matches $\boldsymbol \theta$ for all main effects but otherwise has $\theta_{ij} = 0$ for $i = 1, \dots, \nv, j = 1, \dots, \nh$.
+To assess which parameter values $\boldsymbol \theta$ may cause difficulties in interpretation, we use the difference $\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] - \text{E}\left[\boldsymbol X | \boldsymbol \theta^* \right]$ between two model expectations: E$\left[\boldsymbol X | \boldsymbol \theta\right]$ at $\boldsymbol \theta$ and expectations E$\left[\boldsymbol X | \boldsymbol \theta^* \right]$ where $\boldsymbol \theta^*$ matches $\boldsymbol \theta$ for all main effects but otherwise has $\theta_{ij} = 0$ for $i = 1, \dots, \nv, j = 1, \dots, \nh$.
 Hence, $\boldsymbol \theta^*$ and $\boldsymbol \theta$ have the same main effects but $\boldsymbol \theta^*$ has $0$ dependence parameters.
@@
 
-We next explore and numerically explain the relationship between values of $\boldsymbol \theta$ and the three measures of model impropriety, near-degeneracy, instability, and uninterpretability, for RBM models of varying sizes.
+We next explore and numerically explain the relationship between values of $\boldsymbol \theta$ and the three notions of model impropriety, near-degeneracy, instability, and uninterpretability, for RBM models of varying sizes.
 
@@
 Figure \ref{fig:uninterp} shows that the absolute difference between coordinates of the vector of model expectations, E$\left[\boldsymbol X | \boldsymbol \theta\right]$ and corresponding expectations E$\left[\boldsymbol X | \boldsymbol \theta^*\right ]$ given independence grow for the toy RBM as the values of $\boldsymbol \theta$ are farther from zero, especially for large magnitudes of $\theta_{11}$.
-This is a third indication supporting that parameter vectors of large magnitude lead to model impropriety in a RBM.
+This is a third indication that parameter vectors of large magnitude lead to model impropriety in a RBM.
 
@@
 In addition to the argument that the $-1/1$ data encoding mitigates some perceived prevalence of near-degeneracy it also has the benefit of providing a guaranteed-to-be non-degenerate model at $\boldsymbol \theta = \boldsymbol 0 \in \mathbb{R}^{\nh + \nv + \nh*\nv}$, where the zero vector that serves as the natural center of the parameter space and induces the simplest possible model properties for the RBM (i.e., at $\boldsymbol \theta = \boldsymbol 0$, all variables are independent and visible variables are independent and uniformly distributed on $\{-1,1\}^\nv$).
-The proof of this and further exploration of the equivalence of the $\boldsymbol \theta$ parameterization of the RBM model class and parameterization by $\boldsymbol \mu(\boldsymbol \theta)$ is in the supplementary materials.
+The proof of this and further exploration of the equivalence of the $\boldsymbol \theta$ parameterization of the RBM model class and parameterization by $\boldsymbol \mu(\boldsymbol \theta)$ is in the on-line supplementary materials.
 Hence while from some computing perspectives $0/1$ coding might seem most natural, the $-1/1$ coding is far more convenient and interpretable from the point of view of statistical modeling, where it makes parameters simply interpreted in terms of symmetrically defined main effects and interactions.
@@
 For $\nh, \nv \in \{1, \dots, 4\}$, we sample `r dim(res$samp[[1]])[1]` values of $\boldsymbol \theta$ with various magnitudes (details to follow).
-For each set of parameters we then calculate metrics of model impropriety introduced in Section \ref{degeneracy-instability-and-uninterpretability-oh-my} based on $\boldsymbol \mu(\boldsymbol \theta)$, $\text{ELPR}(\boldsymbol \theta)/\nv$, and the coordinates of $\left|\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] -   \text{E}\left[\boldsymbol X | \boldsymbol \theta^* \right] \right|$.
+For each set of parameters we then calculate metrics of model impropriety introduced in Section \ref{degeneracy-instability-and-uninterpretability-oh-my} based on $\boldsymbol \mu(\boldsymbol \theta)$, $\text{ELPR}(\boldsymbol \theta)/\nv$, and the absolute coordinates of $\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] -   \text{E}\left[\boldsymbol X | \boldsymbol \theta^* \right]$.
 In the case of near-degeneracy, we classify each model as near-degenerate or "viable" based on the distance of $\boldsymbol \mu(\boldsymbol \theta)$ from the boundary of the convex hull of $\mathcal{T}$ and look at the fraction of models that are "near-degenerate," meaning they are within a small distance $\epsilon > 0$ of the boundary of the convex hull.
-We define a "small" through a rough estimation of the volume of the hull for each model size.
+We define "small" through a rough estimation of the volume of the hull for each model size.
 We pick $\epsilon_0 = 0.05$ for $\nh=\nv=1$ and then, for every other $\nh$ and $\nv$, set $m=\nh+\nv+\nv*\nh$ and pick $\epsilon$ so that $1-(1-2\epsilon_0)^3 = 1 - (1-2\epsilon)^m$.
@@
 These average magnitudes vary on a grid between `r min(plot_dat_grid$r1/(plot_dat_grid$H + plot_dat_grid$V))` and `r max(plot_dat_grid$r1/(plot_dat_grid$H + plot_dat_grid$V))` with `r length(unique(plot_dat_grid$r1/(plot_dat_grid$H + plot_dat_grid$V)))` breaks, yielding `r length(unique(plot_dat_grid$r1/(plot_dat_grid$H + plot_dat_grid$V)))^2` grid points.
-(By looking at the average magnitudes, we are able to later consider the potential benefit of shrinking each parameter value $\theta_i$ towards zero in a Bayesian fitting technique.) At each point in the grid, `r dim(res$samp[[1]])[1]` vectors ($\boldsymbol \theta_{main}$) are uniformly sampled on a sphere with radius corresponding to the first coordinate in the grid and `r dim(res$samp[[1]])[1]` vectors ($\boldsymbol \theta_{interction}$) are uniformly sampled on a sphere with radius corresponding to the second coordinate in the grid via sums of squared and scaled iid Normal$(0, 1)$ variables.
+(By looking at the average magnitudes, we are able to later consider the potential benefit of shrinking each parameter value $\theta_i$ towards zero in a Bayesian fitting technique.) At each point in the grid, `r dim(res$samp[[1]])[1]` vectors ($\boldsymbol \theta_{main}$) are sampled uniformly on a sphere with radius corresponding to the first coordinate in the grid and `r dim(res$samp[[1]])[1]` vectors ($\boldsymbol \theta_{interction}$) are sampled uniformly on a sphere with radius corresponding to the second coordinate in the grid via sums of squared and scaled iid Normal$(0, 1)$ variables.
 These vectors are then paired to create `r dim(res$samp[[1]])[1]` values of $\boldsymbol \theta$ with magnitudes at each point in the grid.
@@
 Figure \ref{fig:split_plots_dist} shows the fraction of near-degenerate models for each magnitude of $\boldsymbol \theta$ for each model architecture.
-For each number $\nv$ of visibles in the model, as the number $\nh$ of hiddens increase, the fraction near-degenerate becomes diverges from zero at increasing rates for larger values of $||\boldsymbol \theta||$.
+For each number $\nv$ of visibles in the model, as the number $\nh$ of hiddens increase, the fraction near-degenerate diverges from zero at increasing rates for larger values of $||\boldsymbol \theta||$.
 This shows that as model size gets larger, the risk of degeneracy starts at a slightly larger magnitude of parameters, but very quickly increases until reaching close to 1.
@@
 ```
-These manageable examples indicate that RBMs are near-degenerate, unstable, and uninterpretable for large portions of the parameter space.
+These manageable examples indicate that RBMs are near-degenerate, unstable, and uninterpretable for large portions of the parameter space with large $\|\boldsymbol \theta\|$.
 This, however, is not the only potential problem to be faced when using these models.
@@
 
-The previous discussion intends to indicate that the RBM model class suffers from parameter identifiability issues that go beyond mere symmetries in the parametrization.
+The previous discussion illustrates the fact that the RBM model class suffers from parameter identifiability issues that go beyond mere symmetries in the parametrization.
 Not only it is possible to approximate any distribution on the visibles arbitrarily well [cf.
-@montufar2011refinements], quite different parameter settings can induce the same essential RBM model.
+@montufar2011refinements], but quite different parameter settings can induce the same essential RBM model.
 However, this is not the most disastrous implication of the RBM parameterization.
 A far worse consequence is that, when fitting the RBM model by likelihood-based methods, we already know the nature of the answer before we begin: namely, such  fitting will simply aim to reproduce the empirical distribution from the training data if sufficiently many hiddens are in the model.
-That is, based on a random sample of vectors of visible variables, the model for the cell probabilities that has the highest likelihood over  *all possible model classes* (i.e., RBM-based or not) is the empirical distribution, and the parametrization of the RBM model itself ensures that this empirical distribution can be arbitrarily well-approximated.
+That is, based on a random sample of vectors of visible variables, the model for the cell probabilities that has the highest likelihood over  *all possible model classes* (i.e., RBM-based or not) is the empirical distribution, and the over-parametrization of the RBM model itself ensures that this empirical distribution can be arbitrarily well-approximated.
 
 For illustration, continue the simple example from above with $n$ iid observations, each consisting of two realized visibles $(V_1,V_2)$.
-In which case, when the specified cell probabilities $p_{(-1,-1)},p_{(1,-1)}, p_{(-1,1)},p_{(1,1)}$ in (\ref{eqn:0}) are taken as the empirical cell frequencies from the sample, there is no better model based on maximum likelihood and the discussion above (cf.
+In which case, when the specified cell probabilities $p_{(-1,-1)},p_{(1,-1)}, p_{(-1,1)},p_{(1,1)}$ in (\ref{eqn:0}) are taken as the empirical cell frequencies from the sample, there is no better model based on maximum likelihood, and the discussion above (cf.
 (\ref{eqn:2})) shows that RBM model parameters can be chosen to re-create this empirical distribution to an arbitrarily close degree.
@@
   gather("Parameter", "Value") %>%
-  mutate(Value = paste0("$", round(Value, 10), "$")) -> tbl
+  mutate(Value = paste0("$", round(Value, 7), "$")) -> tbl
 options("scipen" = old_pen)
@@
 This parameter vector was chosen as a sampled value of $\boldsymbol \theta$ that was not near the convex hull of the sufficient statistics for a grid point in Figure \ref{fig:degen_plots} with $< 5$\% near-degeneracy.
-We simulated $n = 5,000$ realizations as visibles as a training set and fit the RBM using three Bayes methodologies.
+We simulated $n = 5,000$ realizations of visibles as a training set and fit the RBM using three Bayes methodologies.
 These involved the following:
@@
 Secondly, we can assess the mixing of our chains using an idea of effective sample size.
-If the MCMC chain were truly iid draws from the target distribution, then for the parameter $p^{(i)}$ denoting the probability of the $i$th vector outcome for the four visibles $\mathbf{v}=(v_1,v_2,v_3,v_4)\in\{\pm 1\}^4$, $i=1,\ldots,16$ (i.e., computed from $\boldsymbol \theta$ under (\ref{eqn:pmf})), its estimate as the average $\bar{p}^{(i)}$ of posterior sample versions would be distributed approximately Normal with mean given by the true posterior marginal probability of $p^{(i)}$, and variance $\sigma^2_i/M$, where $\sigma^2_i$ is the true posterior variance of $p^{(i)}$ and $M$ is the length of the chain.
+If the MCMC chain were truly iid draws from the target distribution, then for the parameter $p^{(i)}$ denoting the probability of the $i$th vector outcome for the four visibles $\mathbf{v}=(v_1,v_2,v_3,v_4)\in\{\pm 1\}^4$, $i=1,\ldots,16$ (i.e., computed from $\boldsymbol \theta$ under (\ref{eqn:pmf})), its estimate as the average $\bar{p}^{(i)}$ of posterior sample versions would be approximately Normal with mean the true posterior marginal probability of $p^{(i)}$, and variance $\sigma^2_i/M$, where $\sigma^2_i$ is the true posterior variance of $p^{(i)}$ and $M$ is the length of the chain.
 However, with the presence of correlation in our chain, the aysmptotic variance of $\bar{p}^{(i)}$ is instead approximately some $C_i/M$, where $C_i$ is some positive constant such that $C_i > \sigma^2_i$.
@@
   mutate(Outcome = as.numeric(Image)) %>%
-  select(Outcome, BwTNML, BwTNLV) %>%
+  select(Outcome, BwTNLV, BwTNML) %>%
   arrange(Outcome) -> M_eff_tmp
@@
 
-Figure \ref{fig:fitting_plot} shows the posterior probability of each possible $\boldsymbol v \in \{-1,1\}^4$ after fitting the RBM model in the two ways detailed in this section, excluding BwTNLV.
-The black vertical lines show the true probabilities of each image based on the parameters used to simulate the training set while the red vertical lines show the empirical distribution for the training set of $5,000$ vectors.
+Figure \ref{fig:fitting_plot} shows the posterior probability of each possible $\boldsymbol v \in \{-1,1\}^4$ after fitting the RBM model in the two ways detailed in this section (excluding BwTNLV).
+The black vertical lines show the true probabilities of each image based on the parameters used to generate the training set while the red vertical lines show the empirical distribution for the training set of $5,000$ vectors.
 From these posterior predictive checks, it is evident that BwTNML produces the best fit to the data.
@@
 ```{r fitting_plot, fig.cap="\\label{fig:fitting_plot}Posterior probabilities of $16 = 2^4$ possible realizations of $4$ visibles using two of the three Bayesian fitting techniques, BwTPLV and BwTNML.
-The black vertical lines show the true probabilities of each vector of visibles based on the parameters used to simulate the training data while the red vertical lines show the empirical distribution.
+The black vertical lines show the true probabilities of each vector of visibles based on the parameters used to generate the training data while the red vertical lines show the empirical distribution.
 BwTNML produces the best fit for the data, however is also the most computationally intensive and least feasible with a real dataset.", fig.height=6}
@@
 However, when viewed as generative statistical models, RBMs are prone to forms of model impropriety such as near-degeneracy, S-instability, and uninterpretability.
-Additionally, these models are difficult to fit using a rigorous methodology due dimension of the parameter space coupled with the size of the latent variable space.
+Additionally, these models are difficult to fit using a rigorous methodology, due to the dimension of the parameter space coupled with the size of the latent variable space.
 
@@
 Common practice is to use a kind of MCMC to overcome fitting complexities.
-However because of the size of the space to be filled with MCMC iterates convergence and mixing of these methods will be slow.
+However because of the size of the space to be filled with MCMC iterates, convergence and mixing of these methods will be slow.
 Marginalization over the latent variables in the model can improve mixing, but is numerically intractible due to the necessity of repeated calculation of the normalizing constant.
@@
 Due to the extreme flexibility in this model class, rigorous likelihood-based fitting for a RBM will typically merely return the (discrete) empirical distribution for visibles.
-Meaning, even if a rigorous likelihood-based fitting method is practically possible, we know what it will produce for fitted probabilities before we even begin.
-For these reasons, we are skeptical of the claims made about RBMs as generative.
+Even if a rigorous likelihood-based fitting method is practically possible, we know what it will produce for fitted probabilities before we even begin.
+For these reasons, we are skeptical of the claims made about RBMs as generative tools.
 

### 2016-11-27 writing/draft.Rmd
sentences about applied instability and index

@@
 In other words, if the probability ratio (\ref{eqn:elpr}) is too large, then a RBM model sequence will exhibit large probability shifts for very small changes in data configurations (i.e., will exhibit instability).
+Recall the applied example of RBM models as a means to classify images.
+This instability result manifests itself as a one pixel change in an image (one component of the visible vector) resulting in a large shift in the probability, which in turn results in a vastly different classification of the image.
+Examples of this behavior have been presented in @szegedy2013intriguing for trained deep learning models, in which a one pixel change in a test image results in a wildly different classification.
 

### 2016-11-27 writing/note.Rmd
sentences about applied instability and index

@@
 $$
-p_{\boldsymbol \lambda}(\boldsymbol x) = \exp\left[\boldsymbol\eta^T(\boldsymbol \lambda) \boldsymbol g(\boldsymbol x) - \psi(\boldsymbol \lambda)\right], \qquad \boldsymbol x \in \mathcal{X}^N,
+p_{N, \boldsymbol \lambda}(\boldsymbol x) = \exp\left[\boldsymbol\eta^T(\boldsymbol \lambda) \boldsymbol g_N(\boldsymbol x) - \psi(\boldsymbol \lambda)\right], \qquad \boldsymbol x \in \mathcal{X}^N,
 $$ 
-for fixed positive integers $k$ and $L$ denoting the dimensions of the parameter and natural parameters spaces respectively, $\boldsymbol \lambda \in \Lambda \subset \mathbb{R}^{k}$, $\boldsymbol \eta : \mathbb{R}^k \mapsto \mathbb{R}^L$, $\boldsymbol g : \mathcal{X}^N \mapsto \mathbb{R}^L$ a vector of sufficient statistics,
+for fixed positive integers $k$ and $L$ denoting the dimensions of the parameter and natural parameters spaces respectively, $\boldsymbol \lambda \in \Lambda \subset \mathbb{R}^{k}$, $\boldsymbol \eta : \mathbb{R}^k \mapsto \mathbb{R}^L$, $\boldsymbol g_N : \mathcal{X}^N \mapsto \mathbb{R}^L$ a vector of sufficient statistics,
 $$
-\psi(\boldsymbol \lambda) = \log \sum\limits_{\boldsymbol x \in \mathcal{X}^N}\exp\left[\boldsymbol \eta^T(\boldsymbol \lambda) \boldsymbol g(\boldsymbol x) \right], \qquad \boldsymbol \lambda \in \Lambda,
+\psi(\boldsymbol \lambda) = \log \sum\limits_{\boldsymbol x \in \mathcal{X}^N}\exp\left[\boldsymbol \eta^T(\boldsymbol \lambda) \boldsymbol g_N(\boldsymbol x) \right], \qquad \boldsymbol \lambda \in \Lambda,
 $$
@@
 
-Defining $\boldsymbol \theta_N=\lambda_N$ to a sequence of elements of $\Lambda \subset \mathbb{R}^k$ and noting that $p_{\boldsymbol \lambda}(\boldsymbol x) > 0$ for all $\boldsymbol x \in \mathcal{X}^N$, these discrete exponential family models are special cases of the FSFS models.
+Defining $\boldsymbol \theta_N=\lambda$ to a sequence of elements of $\Lambda \subset \mathbb{R}^k$ and noting that $p_{N, \boldsymbol \lambda}(\boldsymbol x) > 0$ for all $\boldsymbol x \in \mathcal{X}^N$, these discrete exponential family models are special cases of the FSFS models.
 Such exponential models arise with spatial data on a lattice [@besag1974spatial], network data [@wasserman1994social; @handcock2003assessing], and even binomial sampling with $N$ iid Bernoulli random variables.

### 2016-11-29 writing/draft.Rmd
small updates from meeting

--- 2016-11-23 writing/draft.Rmd
@@
 @handcock2003assessing classify an exponentially parametrized random graph model at $\boldsymbol \theta$ as *near-degenerate* if the mean value of the vector of sufficient statistics under $\boldsymbol \theta$, $\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta}\boldsymbol t( \boldsymbol X)$, is close to the boundary of $C$.
-This makes sense because if a model is near-degenerate in the sense that only a small number of elements of the sample space $\mathcal{X}$ have positive probability, the expected value $\text{E}_{\boldsymbol \theta}\boldsymbol t( \boldsymbol X)$ is an average of that same small number of values of $t( \boldsymbol x)$ and can be expected to not be pulled deep into the interior of $C$.
+This makes sense because if a model is near-degenerate in the sense that only a small number of elements of the sample space $\mathcal{X}$ have positive probability, the expected value $\text{E}_{\boldsymbol \theta}\boldsymbol t( \boldsymbol X)$ is an average of that same small number of values of $t( \boldsymbol x)$ and can be expected to *not* be pulled deep into the interior of $C$.
 
@@
 In other words, if the probability ratio (\ref{eqn:elpr}) is too large, then a RBM model sequence will exhibit large probability shifts for very small changes in data configurations (i.e., will exhibit instability).
+Recall the applied example of RBM models as a means to classify images.
+This instability result manifests itself as a one pixel change in an image (one component of the visible vector) resulting in a large shift in the probability, which in turn could result in a vastly different classification of the image.
+Examples of this behavior have been presented in @szegedy2013intriguing for other deep learning models, in which a one pixel change in a test image results in a wildly different classification.
 
@@
 2.
-*A truncated Normal prior.* Here we use independent spherical normal distributions as priors for $\boldsymbol \theta_{main}$ and $\boldsymbol \theta_{interaction}$ with $\sigma_{interaction} < \sigma_{main}$, truncated at $3\sigma_{main}$ and $3\sigma_{interaction}$, respectively.
+*A truncated Normal prior.* Here we use independent spherical normal distributions as priors for $\boldsymbol \theta_{main}$ and $\boldsymbol \theta_{interaction}$ with $\sigma_{interaction} < \sigma_{main}$, *truncated* at $3\sigma_{main}$ and $3\sigma_{interaction}$, respectively.
 Full conditional distributions are not conjugate, and simulation from the posterior was accomplished using a geometric adaptive Metropolis Hastings step [@zhou2014some] and calculation of likelihood normalizing constant.

### 2016-11-29 writing/note.Rmd
small updates from meeting

@@
 $$
-p_{N, \boldsymbol \lambda}(\boldsymbol x) = \exp\left[\boldsymbol\eta^T(\boldsymbol \lambda) \boldsymbol g_N(\boldsymbol x) - \psi(\boldsymbol \lambda)\right], \qquad \boldsymbol x \in \mathcal{X}^N,
+p_{\boldsymbol \lambda}(\boldsymbol x) = \exp\left[\boldsymbol\eta^T(\boldsymbol \lambda) \boldsymbol g_N(\boldsymbol x) - \psi(\boldsymbol \lambda)\right], \qquad \boldsymbol x \in \mathcal{X}^N,
 $$ 
@@
 
-Defining $\boldsymbol \theta_N=\lambda$ to a sequence of elements of $\Lambda \subset \mathbb{R}^k$ and noting that $p_{N, \boldsymbol \lambda}(\boldsymbol x) > 0$ for all $\boldsymbol x \in \mathcal{X}^N$, these discrete exponential family models are special cases of the FSFS models.
+Defining $\boldsymbol \theta_N=\lambda$ to a sequence of elements of $\Lambda \subset \mathbb{R}^k$ and noting that $p_{\boldsymbol \lambda}(\boldsymbol x) > 0$ for all $\boldsymbol x \in \mathcal{X}^N$, these discrete exponential family models are special cases of the FSFS models.
 Such exponential models arise with spatial data on a lattice [@besag1974spatial], network data [@wasserman1994social; @handcock2003assessing], and even binomial sampling with $N$ iid Bernoulli random variables.
@@
 $$
-\frac{\text{extremal log-probability ratio}}{\text{\# random variables in the model}} \equiv \frac{1}{n*N}\log \left[\frac{\left(\max\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)\right)^n}{\left(\min\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta}(\boldsymbol x)\right)^n}\right] = \frac{1}{N} \log \left[\frac{\max\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)}{\min\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta}(\boldsymbol x)}\right] = \frac{1}{N} \text{ELPR}(\boldsymbol \theta_N),
+\frac{\text{extremal log-probability ratio}}{\text{\# random variables in the model}} \equiv \frac{1}{n*N}\log \left[\frac{\left(\max\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)\right)^n}{\left(\min\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)\right)^n}\right] = \frac{1}{N} \log \left[\frac{\max\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)}{\min\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)}\right] = \frac{1}{N} \text{ELPR}(\boldsymbol \theta_N),
 $$

### 2016-11-29 writing/appendix/appendix-note.Rmd
edits from dr. nordman on note

@@
 We may suppose $|\mathcal{X}|>1$ (i.e., $\mathcal{X}^N$ has more than one outcome) because otherwise the model is trivially degenerate for all $N \geq 1$.
-Fix $0 < \epsilon < 1$ Then, $\boldsymbol x_{max} \in M_{\epsilon, \boldsymbol \theta_N}$, so $P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N}) \ge P_{\boldsymbol \theta_N}(\boldsymbol x_{max}) > 0$.
+Fix $0 < \epsilon < 1$.
+Then, $\boldsymbol x_{max} \in M_{\epsilon, \boldsymbol \theta_N}$, so $P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N}) \ge P_{\boldsymbol \theta_N}(\boldsymbol x_{max}) > 0$.
 If $\boldsymbol x \in \mathcal{X}^N \setminus M_{\epsilon, \boldsymbol \theta_N}$, then by definition $P_{\boldsymbol \theta_N}(\boldsymbol x) \le [P_{\boldsymbol \theta_N}(\boldsymbol x_{max})]^{1-\epsilon}[P_{\boldsymbol \theta_N}(\boldsymbol x_{min})]^{\epsilon}$ holds so that
@@
 \end{align*}
-From the lower bound on $P_{\boldsymbol \theta_N}(\boldsymbol x_{max})$ and the upper bound on $1-P_{\boldsymbol \theta_N}(\boldsymbol x_{max})$, it follows that
+From the lower bound on $P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N})$ and the upper bound on $1-P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N})$, it follows that
 \begin{align*}

### 2016-11-29 writing/note.Rmd
edits from dr. nordman on note

@@
 We consider the problem of quantifying instability for general probability models defined on sequences of observations, where each sequence of length $N$ has a finite number of possible outcomes.
-(A sequence of probability models results indexed by $N$ accommodates data of expanding dimension.) Model instability is formally shown to occur when a certain log-probability ratio under such models grows faster than $N$.
+(A sequence of probability models results indexed by $N$ that accommodates data of expanding dimension.) Model instability is formally shown to occur when a certain log-probability ratio under such models grows faster than $N$.
 In this case, a one component change in the data sequence can provably shift probability by orders of magnitude.
@@
 We consider the behavior, and the potential impropriety, of probability models built to incorporate a sequence of discrete observations with length $N$.
-Let $(X_1, \dots, X_N)$ denote a set of sequence of discrete random variables with a finite sample space, $\mathcal{X}^N$.
+Let $(X_1, \dots, X_N)$ denote a set of discrete random variables with a finite sample space, $\mathcal{X}^N$.
 That is, $\mathcal{X}$ with $|\mathcal{X}| < \infty$ represents a finite set of potential outcomes for the variable $X_i$, and the data sequence $(X_1,\ldots,X_N)$ takes values in the $N$-fold product space $\mathcal{X}^N$.
@@
 
-For $\boldsymbol X = X_1, \dots, X_N$ discrete random variables with sample space $\mathcal{X}^N$, $|\mathcal{X}| < \infty$, consider an exponential family model for $\boldsymbol X$ with probability mass function of the form
+For $\boldsymbol X = (X_1, \dots, X_N)$ discrete random variables with sample space $\mathcal{X}^N$, $|\mathcal{X}| < \infty$, consider an exponential family model for $\boldsymbol X$ with probability mass function of the form
 $$
-p_{\boldsymbol \lambda}(\boldsymbol x) = \exp\left[\boldsymbol\eta^T(\boldsymbol \lambda) \boldsymbol g_N(\boldsymbol x) - \psi(\boldsymbol \lambda)\right], \qquad \boldsymbol x \in \mathcal{X}^N,
+p_{N, \boldsymbol \lambda}(\boldsymbol x) = \exp\left[\boldsymbol\eta^T(\boldsymbol \lambda) \boldsymbol g_N(\boldsymbol x) - \psi(\boldsymbol \lambda)\right], \qquad \boldsymbol x \in \mathcal{X}^N,
 $$ 
-for fixed positive integers $k$ and $L$ denoting the dimensions of the parameter and natural parameters spaces respectively, $\boldsymbol \lambda \in \Lambda \subset \mathbb{R}^{k}$, $\boldsymbol \eta : \mathbb{R}^k \mapsto \mathbb{R}^L$, $\boldsymbol g_N : \mathcal{X}^N \mapsto \mathbb{R}^L$ a vector of sufficient statistics,
+for fixed positive integers $k$ and $L$ denoting the dimensions of the parameter, $\boldsymbol \lambda \in \Lambda \subset \mathbb{R}^{k}$ and natural parameter $\boldsymbol \eta : \mathbb{R}^k \mapsto \mathbb{R}^L$ spaces, $\boldsymbol g_N : \mathcal{X}^N \mapsto \mathbb{R}^L$ a vector of sufficient statistics,
 $$
@@
 
-Defining $\boldsymbol \theta_N=\lambda$ to a sequence of elements of $\Lambda \subset \mathbb{R}^k$ and noting that $p_{\boldsymbol \lambda}(\boldsymbol x) > 0$ for all $\boldsymbol x \in \mathcal{X}^N$, these discrete exponential family models are special cases of the FSFS models.
+Defining $P_{\boldsymbol \theta_N}(\boldsymbol x)\equiv p_{N,\boldsymbol \lambda_N}(\boldsymbol x)$ with $\boldsmbol \theta}_N=\boldsymbol \lambda_N$ as $\boldsymbol \theta_N=\lambda$ to a sequence of elements of $\Lambda \subset \mathbb{R}^k$ and noting that $P_{\boldsymbol \theta_N}(\boldsymbol x) > 0$ for all $\boldsymbol x \in \mathcal{X}^N$, these discrete exponential family models are special cases of the FSFS models.
 Such exponential models arise with spatial data on a lattice [@besag1974spatial], network data [@wasserman1994social; @handcock2003assessing], and even binomial sampling with $N$ iid Bernoulli random variables.
-For these exponential models, the dimension of the parameter $\boldsymbol \theta_N$ is the same for each $N$ (i.e., $\boldsymbol \theta_N$ lies in a parameter space of fixed Euclidean dimension $k$).
+(Note for random graphs or networks with, say, $m$ nodes, one may wish to consider $N={m \choose 2}$ edges as binary (presence/absence) variables $X_i$; in this case, the length $N$ of data sequence may naturally increase as a function of $m$.) For these exponential models, the dimension of the parameter $\boldsymbol \theta_N$ is the same for each $N$ (i.e., $\boldsymbol \theta_N$ lies in a parameter space of fixed Euclidean dimension $k$).
 This need not be true for other types of FSFS models considered in Sections \ref{restricted-boltzmann-machines}-\ref{deep-learning}.
-@schweinberger2011instability considered instability in such exponential models (e.g., for random graphs) for sequences of fixed parameters $\theta_N=\lambda\in\mathbb{R}^k$, $N \geq 1$, of non-varying dimension $k$.
+@schweinberger2011instability considered instability in such exponential models (e.g., for random graphs) for sequences of fixed parameters $\boldsymbol \theta_N=\boldsymbol \lambda\in\mathbb{R}^k$, $N \geq 1$, of non-varying dimension $k$.
 
@@
 Let $\boldsymbol X=(X_1,\ldots,X_N)$ denote the $N$ random variables for visibles with support $\mathcal{X}^N$ and $\boldsymbol H=(H_1,\ldots,H_{N_H})$ denote the $N_H$ random variables for hiddens with support $\mathcal{X}^{N_H}$ where  $\mathcal{X} = \{-1,1\}$.
-For parameters $\boldsymbol \alpha \in \mathbb{R}^{N_H}$, $\boldsymbol \beta \in \mathbb{R}^N$, and $\Gamma$ is a matrix with dimension $N_H \times N$, the RBM for $\tilde{\boldsymbol X}=(\boldsymbol X,\boldsymbol H)$ then has the joint probability mass function
+For parameters $\boldsymbol \alpha \in \mathbb{R}^{N_H}$, $\boldsymbol \beta \in \mathbb{R}^N$, and $\Gamma$ as a matrix with dimension $N_H \times N$, the RBM for $\tilde{\boldsymbol X}=(\boldsymbol X,\boldsymbol H)$ then has the joint probability mass function
 $$
-p_{\boldsymbol \theta} (\tilde{\boldsymbol x}) = \exp\left[ \boldsymbol \alpha^T \boldsymbol h + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^T \Gamma \boldsymbol x - \psi(\boldsymbol \theta)\right], \qquad \tilde{\boldsymbol x} = (\boldsymbol h, \boldsymbol x) \in \mathcal{X}^{N+N_H},
+P_{\boldsymbol \theta} (\tilde{\boldsymbol x}) = \exp\left[ \boldsymbol \alpha^T \boldsymbol h + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^T \Gamma \boldsymbol x - \psi(\boldsymbol \theta)\right], \qquad \tilde{\boldsymbol x} = (\boldsymbol h, \boldsymbol x) \in \mathcal{X}^{N+N_H},
 $$
@@
 $$
-p_{\boldsymbol \theta} (\boldsymbol x) = \sum\limits_{\boldsymbol h \in \mathcal{X}^{N_H}} p_{\boldsymbol \theta} (\boldsymbol x, \boldsymbol h), \qquad \boldsymbol x \in \mathcal{X}^N.
+P_{\boldsymbol \theta} (\boldsymbol x) = \sum\limits_{\boldsymbol h \in \mathcal{X}^{N_H}} P_{\boldsymbol \theta} (\boldsymbol x, \boldsymbol h), \qquad \boldsymbol x \in \mathcal{X}^N.
 $$
 
-Note that the vector of model parameters $\boldsymbol \theta_N$, of size $q(N)$, grows in size as a function of sample dimension $N$ to accomodate the dimension of visible variables $X_1, \dots, X_N$, and one may further choose the number $N_H$ of hidden variables to change with $N$ as well; in particular, the number $N_H$ of hidens may also potentially increase with $N$.
+Note that the vector of model parameters $\boldsymbol \theta_N$, of size $q(N)$, grows in size as a function of sample dimension $N$ to accomodate the dimension of visible variables $X_1, \dots, X_N$, and one may further choose the number $N_H$ of hidden variables to change with $N$ as well; in particular, the number $N_H$ of hiddens may also potentially increase with $N$.
 Additionally, as $|\mathcal{X}| = 2$ and $p_{\boldsymbol \theta}(\boldsymbol x) > 0$ for all $\boldsymbol x \in \mathcal{X}^N$, the RBM model specification for visibles is a FSFS model.
@@
 $$
-p_{\boldsymbol \theta} (\tilde{\boldsymbol x}) = \exp\left[ \sum\limits_{i = 1}^M\boldsymbol \alpha^{(i)T} \boldsymbol h^{(i)} + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^{(1)T} \Gamma^{(0)} \boldsymbol x + \sum\limits_{i = 1}^{M - 1} \boldsymbol h^{(i)T} \Gamma^{(i)} \boldsymbol h^{(i + 1)} - \psi(\boldsymbol \theta) \right],
+P_{\boldsymbol \theta} (\tilde{\boldsymbol x}) = \exp\left[ \sum\limits_{i = 1}^M\boldsymbol \alpha^{(i)T} \boldsymbol h^{(i)} + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^{(1)T} \Gamma^{(0)} \boldsymbol x + \sum\limits_{i = 1}^{M - 1} \boldsymbol h^{(i)T} \Gamma^{(i)} \boldsymbol h^{(i + 1)} - \psi(\boldsymbol \theta) \right],
 $$
@@
 $$
-p_{\boldsymbol \theta} (\boldsymbol x) = \sum\limits_{(\boldsymbol h^{(1)}, \dots, \boldsymbol h^{(M)}) \in \mathcal{X}^{N_{(H,1)} + \cdots + N_{(H,M)}}} p_{\boldsymbol \theta} (\tilde{\boldsymbol x}), \qquad \boldsymbol x \in \mathcal{X}^N
+P_{\boldsymbol \theta} (\boldsymbol x) = \sum\limits_{(\boldsymbol h^{(1)}, \dots, \boldsymbol h^{(M)}) \in \mathcal{X}^{N_{(H,1)} + \cdots + N_{(H,M)}}} p_{\boldsymbol \theta} (\tilde{\boldsymbol x}), \qquad \boldsymbol x \in \mathcal{X}^N
 $$
@@
 Additionally, S-unstable FSFS model sequences are connected to degenerate models, where model *degeneracy* implies placing all probability on a small portion of the sample space.
-For perspective, note that differing bounds on $1/N\cdot\text{ELPR}(\boldsymbol \theta_N$ in (\ref{eqn:elpr}) provide a spectrum of levels of "stability" and Proposition \ref{prop:instab} indicates increasing and undesirable sensitivity of model probabilities (i.e., for one component changes in outcomes) as (\ref{eqn:elpr}) increases.
+For perspective, note that differing bounds on $1/N\cdot\text{ELPR}(\boldsymbol \theta_N)$ in (\ref{eqn:elpr}) provide a spectrum of levels of "stability" and Proposition \ref{prop:instab} indicates increasing and undesirable sensitivity of model probabilities (i.e., for one component changes in outcomes) as (\ref{eqn:elpr}) increases.
 Furthermore, as the instability measure (\ref{eqn:elpr}) grows, FSFS model sequences are guaranteed to slide into full degeneracy as Proposition \ref{prop:degen} will show.

### 2016-11-30 writing/draft.Rmd
edits from dr. nordman

@@
 First, the models are often unsatisfactory as conceptualizations of how data are generated.
-Recalling [@fisher1922mathematical], the aim of a statistical model is to represent data in a compact way.
+Recalling @fisher1922mathematical, the aim of a statistical model is to represent data in a compact way.
 Neyman and Box further state that a model should "provide an explanation of the mechanism underlying the observed phenomena" [@lehmann1990model; @box1967discrimination].
@@
 As the size of these models grows, both maximum likelihood and Bayesian methods of fitting quickly become intractable.
-The literature often suggests Markov chain Monte Carlo (MCMC) tools for approximate maximization of likelihoods to fit these models (e.g., Gibbs sampling to exploit conditional structure in hidden and visible variables), but little is said about acheiving convergence [@hinton2010practical; @hinton2006fast].
+The literature often suggests Markov chain Monte Carlo (MCMC) tools for approximate maximization of likelihoods to fit these models (e.g., Gibbs sampling to exploit conditional structure in hidden and visible variables), but little is said about achieving convergence [@hinton2010practical; @hinton2006fast].
 Related to this, these MCMC algorithms require updating potentially many latent variables (hiddens) which can critically influence convergence in MCMC-based likelihood methods.
@@
 As a consequence, we find that any fully principled fitting method based on the likelihood for RBM with enough hidden variables will simply seek to reproduce the (discrete) empirical distribution of a training set.
-Thus, there is no "smoothed distribution" achieved in fitting a RBM model of sufficient size with a rigorous likelihood-based method.
-We are therefore led to be skeptical that any model built using these structures (like a deep Boltzmann machine) can achieve prediction or inference in a principled way without intentially limiting the flexibility of the fitted model.
+Thus, there can be no "smoothed distribution" achieved in fitting a RBM model of sufficient size with a rigorous likelihood-based method.
+We are therefore led to be skeptical that any model built using these structures (like a deep Boltzmann machine) can achieve prediction or inference in a principled way without intentionally limiting the flexibility of the fitted model.
 
@@
 A kind of Gibbs sampling can be tried (due to conditional architecture of the RBM, i.e.
-visibles given hiddens or vice versa).
+visibles given hiddens or vice verse).
 Specifically, the conditional independence of nodes in each layer (given those nodes in the other layer) allows for stepwise simulation of both hidden layers and model parameters (e.g., contrastive divergence of @hinton2002training or Bayes methods in Section \ref{model-fitting}).
@@
 Here, as earlier, $\gamma(\boldsymbol \theta) = \sum_{\boldsymbol x \in \mathcal{X}} \exp\left(\boldsymbol \theta^T \boldsymbol t(\boldsymbol x)\right)$ is the normalizing function.
-Let $C$ denote the convex hull of the potential outcomes of sufficient statistics, $\{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{X}\}$.
+Let $C$ denote the convex hull of the potential outcomes of sufficient statistics, $\{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{X}\}$, under the model above.
 @handcock2003assessing classify an exponentially parametrized random graph model at $\boldsymbol \theta$ as *near-degenerate* if the mean value of the vector of sufficient statistics under $\boldsymbol \theta$, $\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta}\boldsymbol t( \boldsymbol X)$, is close to the boundary of $C$.
@@
 A RBM model can be thought to exhibit an analogous form of *near-degeneracy* when there is a disproportionate amount of probability placed on a small number of elements in the sample space of visibles and hiddens, $\mathcal{C}^{\nv + \nh}$.
-Using the idea of @handcock2003assessing, when the random vector $\boldsymbol t(\boldsymbol x) = (v_1, \dots, v_\nv, h_1, \dots, h_\nh, v_1 h_1, \dots, v_V h_\nh ) \in \mathcal{T} \equiv \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{C}^{\nh + \nv} \}$ appearing in the neg-potential function $Q_{\boldsymbol \theta}(\cdot)$, has a mean vector $\boldsymbol \mu(\boldsymbol \theta) \in \mathbb{R}^{\nv + \nh + \nv * \nh}$ close to the boundary of the convex hull of $\mathcal{T}$, then the RBM model can be said to exhibit near-degeneracy at $\boldsymbol \theta \in \mathbb{R}^{\nv + \nh + \nh*\nv}$.
+Using the idea of @handcock2003assessing, when the random vector $\boldsymbol t(\boldsymbol x) = (v_1, \dots, v_\nv, h_1, \dots, h_\nh, v_1 h_1, \dots, v_V h_\nh ) \in \mathcal{T} \equiv \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{C}^{\nh + \nv} \}$ from (\ref{eqn:t}), appearing in the neg-potential function $Q_{\boldsymbol \theta}(\cdot)$, has a mean vector $\boldsymbol \mu(\boldsymbol \theta) \in \mathbb{R}^{\nv + \nh + \nv * \nh}$ close to the boundary of the convex hull of $\mathcal{T}$, then the RBM model can be said to exhibit near-degeneracy at $\boldsymbol \theta \in \mathbb{R}^{\nv + \nh + \nh*\nv}$.
 Here the mean of $\boldsymbol t(\boldsymbol x)$ is
@@
 
-Considering exponential families of distributions, @schweinberger2011instability introduced a concept of model defficiency related to *instability*.
+Considering exponential families of distributions, @schweinberger2011instability introduced a concept of model deficiency related to *instability*.
 Instability can be roughly thought of as excessive sensitivity in the model, where small changes in the components of potential data outcomes, $\boldsymbol x$, may lead to substantial changes in the probability function $f_{\boldsymbol \theta}(\boldsymbol x)$.
@@
 Let $\boldsymbol \theta_\nv \in \mathbb{R}^{\nv + \nh +  \nh*\nv}, \nv \ge 1$, be an element of a sequence of RBM parameters where the number of hiddens, $\nh \equiv \nh(\nv) \ge 1$, can be an arbitrary function of the number $\nv$ of visibles.
-A RBM model formulation is \emph{Scheinberger-unstable} or \emph{S-unstable} if
+A RBM model formulation is \emph{Schweinberger-unstable} or \emph{S-unstable} if
 \begin{align*}
@@
 Recall the applied example of RBM models as a means to classify images.
-This instability result manifests itself as a one pixel change in an image (one component of the visible vector) resulting in a large shift in the probability, which in turn could result in a vastly different classification of the image.
+For data as pixels in an image, the instability result in Proposition \ref{prop:instab} manifests itself as a one pixel change in an image (one component of the visible vector) resulting in a large shift in the probability, which in turn could result in a vastly different classification of the image.
 Examples of this behavior have been presented in @szegedy2013intriguing for other deep learning models, in which a one pixel change in a test image results in a wildly different classification.
@@
 ```{r instab, fig.cap="\\label{fig:instab}$\\text{ELPR}(\\boldsymbol \\theta)/\\nv$ for various values of $\\boldsymbol \\theta$ for the tiny example model.
-(Recall that here $\\nv$ is the number of visible nodes and here is $1$.
+Recall here $\\nv$ is the number of visible nodes and here is $1$.
 This quantity is large for large magnitudes of $\\boldsymbol \\theta$.", fig.height = 4, fig.align='center'}
@@
 
-```{r uninterp, fig.cap="\\label{fig:uninterp}The absolute difference between coordinates of model expectations, E$\\left[\\boldsymbol X | \\boldsymbol \\theta\\right]$, and expectations given independence, E$\\left[\\boldsymbol X | \\boldsymbol \\theta^* \\right ]$ for a RBM with one visible and one hidden node.", fig.height = 4, fig.align='center', message=FALSE, cache=TRUE, warning=FALSE}
+```{r uninterp, fig.cap="\\label{fig:uninterp}The absolute difference between coordinates of model expectations, E$\\left[\\boldsymbol X | \\boldsymbol \\theta\\right]$, and expectations given independence, E$\\left[\\boldsymbol X | \\boldsymbol \\theta^* \\right ]$ for a RBM with one visible and one hidden node.
+As an indicator of uninterpretability, note that differences in expectations increase as the dependence parameter $\theta_{11}$ deviates from zero.", fig.height = 4, fig.align='center', message=FALSE, cache=TRUE, warning=FALSE}
 expand.grid(theta_h1 = seq(-5, 5, length.out = 21),
@@
 
-Figure \ref{fig:uninterp} shows that the absolute difference between coordinates of the vector of model expectations, E$\left[\boldsymbol X | \boldsymbol \theta\right]$ and corresponding expectations E$\left[\boldsymbol X | \boldsymbol \theta^*\right ]$ given independence grow for the toy RBM as the values of $\boldsymbol \theta$ are farther from zero, especially for large magnitudes of $\theta_{11}$.
+Figure \ref{fig:uninterp} shows that the absolute difference between coordinates of the vector of model expectations, E$\left[\boldsymbol X | \boldsymbol \theta\right]$ and corresponding expectations E$\left[\boldsymbol X | \boldsymbol \theta^*\right ]$ given independence grow for the toy RBM as the values of $\boldsymbol \theta$ are farther from zero, especially for large magnitudes of the dependence parameter $\theta_{11}$.
 This is a third indication that parameter vectors of large magnitude lead to model impropriety in a RBM.
@@
 
-In addition to the argument that the $-1/1$ data encoding mitigates some perceived prevalence of near-degeneracy it also has the benefit of providing a guaranteed-to-be non-degenerate model at $\boldsymbol \theta = \boldsymbol 0 \in \mathbb{R}^{\nh + \nv + \nh*\nv}$, where the zero vector that serves as the natural center of the parameter space and induces the simplest possible model properties for the RBM (i.e., at $\boldsymbol \theta = \boldsymbol 0$, all variables are independent and visible variables are independent and uniformly distributed on $\{-1,1\}^\nv$).
+In addition to the argument that the $-1/1$ data encoding mitigates some perceived prevalence of near-degeneracy it also has the benefit of providing a guaranteed-to-be non-degenerate model at $\boldsymbol \theta = \boldsymbol 0 \in \mathbb{R}^{\nh + \nv + \nh*\nv}$, where the zero vector then serves as the natural center of the parameter space and induces the simplest possible model properties for the RBM (i.e., at $\boldsymbol \theta = \boldsymbol 0$, all variables are independent and visible variables are independent and uniformly distributed on $\{-1,1\}^\nv$).
 The proof of this and further exploration of the equivalence of the $\boldsymbol \theta$ parameterization of the RBM model class and parameterization by $\boldsymbol \mu(\boldsymbol \theta)$ is in the on-line supplementary materials.
@@
 
-Typically, fitting a RBM via maximum likelihood (ML) methods will be infeasible due mainly to the intractibility of the normalizing term $\gamma(\boldsymbol \theta)$ in a model (\ref{eqn:pmf}) of any realistic size.
+Typically, fitting a RBM via maximum likelihood (ML) methods will be infeasible due mainly to the intractability of the normalizing term $\gamma(\boldsymbol \theta)$ in a model (\ref{eqn:pmf}) of any realistic size.
 Ad hoc methods are used instead, which aim to avoid this problem by using stochastic ML approximations that employ a small number of MCMC draws (i.e., contrastive divergence, [@hinton2002training]).
@@
 However, computational concerns are not the only issues with fitting a RBM using ML.
-In addition, a RBM model, with the appropriate choice of parameters and number of hiddens, has the potential to re-create any distribution for the data (i.e., reproduce any specification of cell probabilities for the $-1/1$ valued data outcomes).
+In addition, a RBM model, with the appropriate choice of parameters and number of hiddens, has the potential to re-create any distribution for the data (i.e., reproduce any specification of cell probabilities for the binary data outcomes).
 For example, @montufar2011refinements show that any distribution on $\{0, 1\}^{\nv}$ can be approximated arbitrarily well by a RBM with $2^{\nv}/2 - 1$ hidden units.
@@
 
-To demonstrate, =we  assume $p_{(-1,-1)}>0$ (without loss of generality) in the specified cell probabilities (\ref{eqn:0}) and replace parameters $\theta_{11},\theta_{21}$ with $\Delta_1 \equiv \theta_{11} +\theta_{21}$ and $\Delta_2 \equiv \theta_{11} -\theta_{21}$.
+To demonstrate, we  assume $p_{(-1,-1)}>0$ (without loss of generality) in the specified cell probabilities (\ref{eqn:0}) and replace parameters $\theta_{11},\theta_{21}$ with $\Delta_1 \equiv \theta_{11} +\theta_{21}$ and $\Delta_2 \equiv \theta_{11} -\theta_{21}$.
 We may then prescribe values of $\theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \Delta_1,\Delta_2$  so that the model probability ratio
@@
 *A "trick" prior.* Here we cancel out normalizing term in the likelihood so that resulting full conditionals of $\boldsymbol \theta$ are multivariate Normal.
-The $h_j$ are carried allong in the MCMC sampling from the posterior as latent variables.
+The $h_j$ are carried along in the MCMC sampling from the posterior as latent variables.
     \begin{align*}
@@
 2.
-*A truncated Normal prior.* Here we use independent spherical normal distributions as priors for $\boldsymbol \theta_{main}$ and $\boldsymbol \theta_{interaction}$ with $\sigma_{interaction} < \sigma_{main}$, *truncated* at $3\sigma_{main}$ and $3\sigma_{interaction}$, respectively.
+*A truncated Normal prior.* Here we use independent spherical normal distributions as priors for $\boldsymbol \theta_{main}$ and $\boldsymbol \theta_{interaction}$, with $\sigma_{interaction} < \sigma_{main}$, *truncated* at $3\sigma_{main}$ and $3\sigma_{interaction}$, respectively.
 Full conditional distributions are not conjugate, and simulation from the posterior was accomplished using a geometric adaptive Metropolis Hastings step [@zhou2014some] and calculation of likelihood normalizing constant.
@@
 After a burn-in period of 50 iterations selected by inspecting the trace plots, we assess the issue of mixing in two ways.
-First, the autocorrelation functions (ACF) for each possible vector of visibles are assessed and plotted in Figure \ref{fig:acf} with BwTNLV in black and BwTNML in red.
+First, the autocorrelation functions (ACF) from each posterior sample corresponding to a model probability for a visible vector outcome $\mathbf{v}=(v_1,v_2,v_3,v_4)\in\{\pm 1\}^4$ (i.e., computed from $\boldsymbol \theta$ under (\ref{eqn:pmf})) of visibles are assessed and plotted in Figure \ref{fig:acf} with BwTNLV in black and BwTNML in red.
 As expected, ACF corresponding to the method that marginalizes out the hidden variables from the likelihood decreases to zero at a much faster rate, indicating better mixing for the chain.
@@
 
-```{r acf, fig.cap="\\label{fig:acf}The autocorrelation functions (ACF) for all $2^4 = 16$ possibles vectors of visibles assessed at multiple lags for each method with BwTNLV in black and BwTNML in red.
+```{r acf, fig.cap="\\label{fig:acf}The autocorrelation functions (ACF) for the posterior probabilityies of all $2^4 = 16$ possible outcomes for the vector of four visibles assessed at multiple lags for each method with BwTNLV in black and BwTNML in red.
 As expected, ACF corresponding to the method that marginalizes out the hidden variables from the likelihood decreases to zero at a much faster rate, indicating better mixing for the chain."}
@@
   spread(image_id, `$M_{eff}$`) -> M_eff_table
-
-M_eff_ratio <- M_eff_table[1,-1]/M_eff_table[2,-1]
 ```
@@
 Secondly, we can assess the mixing of our chains using an idea of effective sample size.
-If the MCMC chain were truly iid draws from the target distribution, then for the parameter $p^{(i)}$ denoting the probability of the $i$th vector outcome for the four visibles $\mathbf{v}=(v_1,v_2,v_3,v_4)\in\{\pm 1\}^4$, $i=1,\ldots,16$ (i.e., computed from $\boldsymbol \theta$ under (\ref{eqn:pmf})), its estimate as the average $\bar{p}^{(i)}$ of posterior sample versions would be approximately Normal with mean the true posterior marginal probability of $p^{(i)}$, and variance $\sigma^2_i/M$, where $\sigma^2_i$ is the true posterior variance of $p^{(i)}$ and $M$ is the length of the chain.
-However, with the presence of correlation in our chain, the aysmptotic variance of $\bar{p}^{(i)}$ is instead approximately some $C_i/M$, where $C_i$ is some positive constant such that $C_i > \sigma^2_i$.
+If the MCMC chain were truly iid draws from the target distribution, then for the parameter $p^{(i)}$ denoting the probability of the $i$th vector outcome for the four visibles $\mathbf{v}=(v_1,v_2,v_3,v_4)\in\{\pm 1\}^4$, $i=1,\ldots,16$, its estimate as the average $\bar{p}^{(i)}$ of posterior sample versions would be approximately Normal with mean the true posterior marginal probability of $p^{(i)}$, and variance $\sigma^2_i/M$, where $\sigma^2_i$ is the true posterior variance of $p^{(i)}$ and $M$ is the length of the chain.
+However, with the presence of correlation in our chain, the asymptotic variance of $\bar{p}^{(i)}$ is instead approximately some $C_i/M$, where $C_i$ is some positive constant such that $C_i > \sigma^2_i$.
 We can use an overlapping block-means approach [@gelman2011inference] to get a crude estimate for $C_i$ as $\hat{C}_i = bS_b^2$, where $S_b^2$  denotes the sample variance of overlapping block means $\{\bar{p}_j^{(i)}=\sum_{k=j}^{j+b-1} p_k^{(i)}/b\}_{j=1}^{M-b+1}$ of length $b$ computed from the posterior samples $\{p_k^{(i)}\}_{k=1}^M$.
@@
 cbind(M_eff_tmp[1:8,], M_eff_tmp[9:16,]) %>%
-  kable(digits = 2, caption = paste0("\\label{tab:m_eff}The effective sample sizes for a chain of length $M = 1000$ of all $16$ possible vector outcomes of visibles.
-BwTNLV would require at least $", round(min(M_eff_table[1, -1])/min(M_eff_table[2, -1]), 1), "$ times as many MCMC iterations to achieve the same amount of effective information about the posterior distribution."))
-```
-
-The effective sample sizes for a chain of length $M = 1000$ of all $2^4 = 16$ possible vectors of visibles are presented in Table \ref{tab:m_eff}.
-These range from  $`r round(min(M_eff_table[1, -1]), 2)`$ to $`r round(max(M_eff_table[1, -1]), 2)`$ for BwTNML, while BwTNLV only yields between $`r round(min(M_eff_table[2, -1]), 2)`$ and $`r round(max(M_eff_table[2, -1]), 2)`$ effective draws.
-Thus, BwTNLV would require at least $`r round(min(M_eff_table[1, -1])/min(M_eff_table[2, -1]), 1)`$ times more iterations to be run of the MCMC chain in order to achieve the same amount of effective information about the posterior distribution.
+  kable(digits = 2, caption = paste0("\\label{tab:m_eff}The effective sample sizes for a chain of length $M = 1000$ regarding all $16$ probabilities for possible vector outcomes of visibles.
+BwTNLV would require at least $", round(min(M_eff_tmp[, "BwTNML"])/min(M_eff_tmp[, "BwTNLV"]), 1), "$ times as many MCMC iterations to achieve the same amount of effective information about the posterior distribution."))
+```
+
+The effective sample sizes for a chain of length $M = 1000$ for inference about each of the $2^4 = 16$ model probabilities are presented in Table \ref{tab:m_eff}.
+These range from  $`r round(min(M_eff_tmp[, "BwTNML"]), 2)`$ to $`r round(max(M_eff_tmp[, "BwTNML"]), 2)`$ for BwTNML, while BwTNLV only yields between $`r round(min(M_eff_tmp[, "BwTNLV"]), 2)`$ and $`r round(max(M_eff_tmp[, "BwTNLV"]), 2)`$ effective draws.
+Thus, BwTNLV would require at least $`r round(min(M_eff_tmp[, "BwTNML"])/min(M_eff_tmp[, "BwTNLV"]), 1)`$ times more iterations to be run of the MCMC chain in order to achieve the same amount of effective information about the posterior distribution.
 For this reason, consistent with the ACF results in Figure \ref{fig:acf}, BwTNLV does not seem to be an effective method for fitting the RBM if computing resources are at all limited.
@@
 However because of the size of the space to be filled with MCMC iterates, convergence and mixing of these methods will be slow.
-Marginalization over the latent variables in the model can improve mixing, but is numerically intractible due to the necessity of repeated calculation of the normalizing constant.
+Marginalization over the latent variables in the model can improve mixing, but is numerically intractable due to the necessity of repeated calculation of the normalizing constant.
 

### 2016-11-30 writing/note.Rmd
edits from dr. nordman

@@
 The instability results of @schweinberger2011instability were developed for the case of discrete exponential family models.
-The main results here concern a general measure of model instability, appropriate accross the whole FSFS model class.
+The main results here concern a general measure of model instability, appropriate across the whole FSFS model class.
 This can be used to identify when certain maximal probabilities in FSFS models are too extreme relative to the length $N$.
@@
 
-Defining $P_{\boldsymbol \theta_N}(\boldsymbol x)\equiv p_{N,\boldsymbol \lambda_N}(\boldsymbol x)$ with $\boldsmbol \theta}_N=\boldsymbol \lambda_N$ as $\boldsymbol \theta_N=\lambda$ to a sequence of elements of $\Lambda \subset \mathbb{R}^k$ and noting that $P_{\boldsymbol \theta_N}(\boldsymbol x) > 0$ for all $\boldsymbol x \in \mathcal{X}^N$, these discrete exponential family models are special cases of the FSFS models.
+Defining $P_{\boldsymbol \theta_N}(\boldsymbol x)\equiv p_{N,\boldsymbol \lambda_N}(\boldsymbol x)$ with $\boldsymbol \theta_N=\boldsymbol \lambda_N$ as $\boldsymbol \theta_N=\lambda$ to a sequence of elements of $\Lambda \subset \mathbb{R}^k$ and noting that $P_{\boldsymbol \theta_N}(\boldsymbol x) > 0$ for all $\boldsymbol x \in \mathcal{X}^N$, these discrete exponential family models are special cases of the FSFS models.
 Such exponential models arise with spatial data on a lattice [@besag1974spatial], network data [@wasserman1994social; @handcock2003assessing], and even binomial sampling with $N$ iid Bernoulli random variables.
@@
 
-Note that the vector of model parameters $\boldsymbol \theta_N$, of size $q(N)$, grows in size as a function of sample dimension $N$ to accomodate the dimension of visible variables $X_1, \dots, X_N$, and one may further choose the number $N_H$ of hidden variables to change with $N$ as well; in particular, the number $N_H$ of hiddens may also potentially increase with $N$.
+Note that the vector of model parameters $\boldsymbol \theta_N$, of size $q(N)$, grows in size as a function of sample dimension $N$ to accommodate the dimension of visible variables $X_1, \dots, X_N$, and one may further choose the number $N_H$ of hidden variables to change with $N$ as well; in particular, the number $N_H$ of hiddens may also potentially increase with $N$.
 Additionally, as $|\mathcal{X}| = 2$ and $p_{\boldsymbol \theta}(\boldsymbol x) > 0$ for all $\boldsymbol x \in \mathcal{X}^N$, the RBM model specification for visibles is a FSFS model.
@@
 
-Consider two models with "deep architechture" that contain multiple hidden (or latent) layers in addition to a visible layer of data, a deep Boltzmann machine [@salakhutdinov2009deep] and a deep belief network [@hinton2006fast].
+Consider two models with "deep architecture" that contain multiple hidden (or latent) layers in addition to a visible layer of data, a deep Boltzmann machine [@salakhutdinov2009deep] and a deep belief network [@hinton2006fast].
 Let $M$ denote the number of hidden layers included in the model and $N_{(H,1)}, \dots, N_{((H,M))}$ the number of hidden variables within each hidden layer.
-Then let $\tilde{\boldsymbol X} = \{H^{(1)}_1, \dots, H^{(1)}_{N_{(H,1)}}, \dots, H^{(M)}_1, \dots, H^{(M)}_{N_{(H,M)}}, \boldsymbol X\}$ be random variables corresponding to the hidden variables $\{ H_{i}^{(j)} : i=1,\ldots, N_{(H,j)}, j=1,\ldots,M\}$ and visible variables $\boldsymbol X =(X_1,\ldots,X_N)$ in a deep probabilitic model.
+Then let $\tilde{\boldsymbol X} = \{H^{(1)}_1, \dots, H^{(1)}_{N_{(H,1)}}, \dots, H^{(M)}_1, \dots, H^{(M)}_{N_{(H,M)}}, \boldsymbol X\}$ be random variables corresponding to the hidden variables $\{ H_{i}^{(j)} : i=1,\ldots, N_{(H,j)}, j=1,\ldots,M\}$ and visible variables $\boldsymbol X =(X_1,\ldots,X_N)$ in a deep probabilistic model.
 Each variable outcome will again lie in $\mathcal{X} = \{-1,1\}$.
@@
 
-This definition of *S-unstable* is a generalization or reinterpretation of "unstable" used in @schweinberger2011instability by allowing nonexponential family models (e.g.
+This definition of *S-unstable* is a generalization or reinterpretation of "unstable" used in @schweinberger2011instability by allowing non-exponential family models (e.g.
 RBM and DBM models in Sections \ref{restricted-boltzmann-machines}-\ref{deep-learning}) and an increasing number of parameters.
@@
 
-S-unstable FSFS model seqences are undesirable for several reasons. One is that small changes in data can lead to overly-sensitive changes in probability. Consider, for example, 
+S-unstable FSFS model sequences are undesirable for several reasons. One is that small changes in data can lead to overly-sensitive changes in probability. Consider, for example, 
 $$

### 2016-12-01 writing/appendix/appendix-note.Rmd
nordman and vardeman final edits

@@
-{\bf Proof of Proposition \ref{prop:instab}.} We prove the contrapositive, supposing that $\Delta(\boldsymbol \theta_N) \le C$ holds for some $C > 0$ and show $\text{ELPR}(\boldsymbol \theta_N) \leq C$.
+{\bf Proof of Proposition \ref{prop:instab}.} We prove the contrapositive, supposing that $\Delta(\boldsymbol \theta_N) \le C$ holds for some $C > 0$ and show $\text{ELPR}(\boldsymbol \theta_N) \leq NC$.
 Let $\boldsymbol x_{min} \equiv \argmin\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)$ and $\boldsymbol x_{max} \equiv \argmax\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)$.
@@
 $$
-N\cdot\text{ELPR}(\boldsymbol \theta_N) = \log\left[\frac{P_{\boldsymbol \theta_N}(\boldsymbol x_{max})}{P_{\boldsymbol \theta_N}(\boldsymbol x_{min})}\right] = \left|\sum\limits_{i = 1}^k\log\left(\frac{P_{\boldsymbol \theta_N}(\boldsymbol x_i)}{P_{\boldsymbol \theta_N}(\boldsymbol x_{i-1})}\right)\right| \le \sum\limits_{i = 1}^k\left|\log\left(\frac{P_{\boldsymbol \theta_N}(\boldsymbol x_i)}{P_{\boldsymbol \theta}(\boldsymbol x_{i-1})}\right)\right| \le k \Delta_N(\boldsymbol \theta_N) \le NC,
+\text{ELPR}(\boldsymbol \theta_N) = \log\left[\frac{P_{\boldsymbol \theta_N}(\boldsymbol x_{max})}{P_{\boldsymbol \theta_N}(\boldsymbol x_{min})}\right] = \left|\sum\limits_{i = 1}^k\log\left(\frac{P_{\boldsymbol \theta_N}(\boldsymbol x_i)}{P_{\boldsymbol \theta_N}(\boldsymbol x_{i-1})}\right)\right| \le \sum\limits_{i = 1}^k\left|\log\left(\frac{P_{\boldsymbol \theta_N}(\boldsymbol x_i)}{P_{\boldsymbol \theta}(\boldsymbol x_{i-1})}\right)\right| \le k \Delta_N(\boldsymbol \theta_N) \le NC,
 $$

### 2016-12-01 writing/draft.Rmd
nordman and vardeman final edits

@@
 However, the empirical distribution of a training set of vectors of visibles is the best fitting model for observed cell data.
-As a consequence, we find that any fully principled fitting method based on the likelihood for RBM with enough hidden variables will simply seek to reproduce the (discrete) empirical distribution of a training set.
+As a consequence, we find that any fully principled fitting method based on the likelihood for a RBM with enough hidden variables will simply seek to reproduce the (discrete) empirical distribution of a training set.
 Thus, there can be no "smoothed distribution" achieved in fitting a RBM model of sufficient size with a rigorous likelihood-based method.
-We are therefore led to be skeptical that any model built using these structures (like a deep Boltzmann machine) can achieve prediction or inference in a principled way without intentionally limiting the flexibility of the fitted model.
+We are therefore led to be skeptical that any model built using these structures (like a deep Boltzmann machine) can achieve useful prediction or inference in a principled way without intentionally limiting the flexibility of the fitted model.
 
@@
 \end{align}
-let $\mathcal{T} = \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{C}^{\nv + \nh}\} \subset \mathbb{R}^{\nv + \nh + \nv * \nh}$ be the set of possible values for the vector of variables needed to compute probabilities (\ref{eqn:pmf}) in the model and write $Q_{\boldsymbol \theta}(\boldsymbol x) = \sum\limits_{i = 1}^\nh\sum\limits_{j=1}^\nv \theta_{ij} h_i v_j + \sum\limits_{i = 1}^\nh\theta_{hi} h_i + \sum\limits_{j = 1}^\nv\theta_{vj} v_j$ for the "neg-potential" function.
+let $\mathcal{T} = \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{C}^{\nv + \nh}\} \subset \mathbb{R}^{\nv + \nh + \nv * \nh}$ be the set of possible values for the vector of variables needed to compute probabilities (\ref{eqn:pmf}) in the model, and write $Q_{\boldsymbol \theta}(\boldsymbol x) = \sum\limits_{i = 1}^\nh\sum\limits_{j=1}^\nv \theta_{ij} h_i v_j + \sum\limits_{i = 1}^\nh\theta_{hi} h_i + \sum\limits_{j = 1}^\nv\theta_{vj} v_j$ for the "neg-potential" function.
 The RBM model is parameterized by $\boldsymbol \theta$ containing two types of parameters, main effects and interaction effects.
@@
 Due to the potential size of the model, the normalizing constant $\gamma(\boldsymbol \theta)$ can be practically impossible to calculate, making simple estimation of the model parameter vector problematic.
-A kind of Gibbs sampling can be tried (due to conditional architecture of the RBM, i.e.
+A kind of Gibbs sampling can be tried (due to the conditional architecture of the RBM, i.e.
 visibles given hiddens or vice verse).
-Specifically, the conditional independence of nodes in each layer (given those nodes in the other layer) allows for stepwise simulation of both hidden layers and model parameters (e.g., contrastive divergence of @hinton2002training or Bayes methods in Section \ref{model-fitting}).
+Specifically, the conditional independence of nodes in each layer (given those nodes in the other layer) allows for stepwise simulation of both hidden layers and model parameters (e.g., see the contrastive divergence of @hinton2002training or Bayes methods in Section \ref{model-fitting}).
 
@@
 \end{defn}
-In other words, given any $C > 0$, there exists an integer $n_C > 0$ so that $\frac{1}{\nv}\text{ELPR}(\boldsymbol \theta_\nv) > C$ for all $\nv \ge n_C$.
+In other words, given any $c > 0$, there exists an integer $n_c > 0$ so that $\frac{1}{\nv}\text{ELPR}(\boldsymbol \theta_\nv) > c$ for all $\nv \ge n_c$.
 
@@
 \label{prop:instab}
-Let $C > 0$ and let $\text{ELPR}(\boldsymbol \theta_\nv)$ be as in (\ref{eqn:elpr}) for an integer $\nv \ge 1$.
-If $\frac{1}{\nv}\text{ELPR}(\boldsymbol \theta_\nv) > C$, then $\Delta(\boldsymbol \theta_\nv) > C$.
+Let $c > 0$ and let $\text{ELPR}(\boldsymbol \theta_\nv)$ be as in (\ref{eqn:elpr}) for an integer $\nv \ge 1$.
+If $\frac{1}{\nv}\text{ELPR}(\boldsymbol \theta_\nv) > c$, then $\Delta(\boldsymbol \theta_\nv) > c$.
 \end{prop}
@@
 In other words, S-unstable RBM model sequences are guaranteed to stack up all probability on an arbitrarily narrow set of outcomes for visibles.
-Proofs of Propositions \ref{prop:instab}-\ref{prop:degen} appear in @kaplan2016note.
+Proofs of Propositions \ref{prop:instab} and \ref{prop:degen} appear in @kaplan2016note.
 These findings also have counterparts in results in @schweinberger2011instability, but unlike results there, we do not limit consideration to exponential family forms with a fixed number of parameters.
@@
 Using this, it is possible to investigate what parametric conditions lead to uninterpretability in a model versus those that guarantee interpretable models.
-If $||\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] - \text{E}\left[\boldsymbol X | \boldsymbol \theta^*\right]||$ is large, then the RBM model with parameter vector $\boldsymbol \theta$ is said to be uninterpetable.
+If $\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] - \text{E}\left[\boldsymbol X | \boldsymbol \theta^*\right]$ is large, then the RBM model with parameter vector $\boldsymbol \theta$ is said to be uninterpetable.
 The quantities to compare in the RBM case are
@@
 As this set is in three dimensions, we are able to explicitly illustrate the shape of boundary of the convex hull of $\mathcal{T}$ and explore the behavior of the mean vector $\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta} \boldsymbol t(\boldsymbol x)$ as a function of the parameter vector $\boldsymbol \theta$.
-Figure \ref{fig:toyhull} shows the convex hull of our "statistic space," $\mathcal{T} \subset \{0,1\}^3$, for this toy problem from two perspectives (enclosed by the unit cube $[0,1]^3$, representing the convex hull of $\{0,1\}^3$ space).
+Figure \ref{fig:toyhull} shows the convex hull of our "statistic space," $\mathcal{T} \subset \{0,1\}^3$, for this toy problem from two perspectives (enclosed by the unit cube $[0,1]^3$, the convex hull of $\{0,1\}^3$).
 In this small model, note that the convex hull of $\mathcal{T}$ does not fill the unrestricted hull of $\{0,1\}^3$ because of the relationship between the elements of $\mathcal{T} = \{(v_1, h_1, v_1 h_1 : v_1, h_1 \in \{0, 1\} \}$ (i.e.
@@
 The three parametric coordinate functions of $\boldsymbol \mu(\boldsymbol \theta)$ can be represented as in Figure \ref{fig:degen_toy}.
-(Contour plots for three coordinate functions are shown in columns for various values of $\theta_{11}$.) In examining these, we see that as values of coordinates of $\boldsymbol \theta$ grow larger in magnitude, at least one mean function for the entries of $\boldsymbol t(\boldsymbol x)$ is approaching a value 0 or 1, forcing $\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta} \boldsymbol t(\boldsymbol x)$ to be near to the boundary of the convex hull of $\mathcal{T}$, as a sign of model near-degeneracy.
+(Contour plots for three coordinate functions are shown in columns for various values of $\theta_{11}$.) In examining these, we see that as coordinates of $\boldsymbol \theta$ grow larger in magnitude, at least one mean function for the entries of $\boldsymbol t(\boldsymbol x)$ approaches a value 0 or 1, forcing $\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta} \boldsymbol t(\boldsymbol x)$ to be near to the boundary of the convex hull of $\mathcal{T}$, as a sign of model near-degeneracy.
 Thus, for a very small example we can see the relationship between values of $\boldsymbol \theta$ and model degeneracy.
@@
 ```{r uninterp, fig.cap="\\label{fig:uninterp}The absolute difference between coordinates of model expectations, E$\\left[\\boldsymbol X | \\boldsymbol \\theta\\right]$, and expectations given independence, E$\\left[\\boldsymbol X | \\boldsymbol \\theta^* \\right ]$ for a RBM with one visible and one hidden node.
-As an indicator of uninterpretability, note that differences in expectations increase as the dependence parameter $\theta_{11}$ deviates from zero.", fig.height = 4, fig.align='center', message=FALSE, cache=TRUE, warning=FALSE}
+As an indicator of uninterpretability, note that differences in expectations increase as the dependence parameter $\\theta_{11}$ deviates from zero.", fig.height = 4, fig.align='center', message=FALSE, cache=TRUE, warning=FALSE}
 expand.grid(theta_h1 = seq(-5, 5, length.out = 21),
@@
 The proof of this and further exploration of the equivalence of the $\boldsymbol \theta$ parameterization of the RBM model class and parameterization by $\boldsymbol \mu(\boldsymbol \theta)$ is in the on-line supplementary materials.
-Hence while from some computing perspectives $0/1$ coding might seem most natural, the $-1/1$ coding is far more convenient and interpretable from the point of view of statistical modeling, where it makes parameters simply interpreted in terms of symmetrically defined main effects and interactions.
+Hence, while from some computing perspectives $0/1$ coding might seem most natural, the $-1/1$ coding is far more convenient and interpretable from the point of view of statistical modeling, where it makes parameters simply interpreted in terms of symmetrically defined main effects and interactions.
 In light of all of these matters we will henceforth employ the $-1/1$ coding.
@@
 In addition, a RBM model, with the appropriate choice of parameters and number of hiddens, has the potential to re-create any distribution for the data (i.e., reproduce any specification of cell probabilities for the binary data outcomes).
-For example, @montufar2011refinements show that any distribution on $\{0, 1\}^{\nv}$ can be approximated arbitrarily well by a RBM with $2^{\nv}/2 - 1$ hidden units.
-To illustrate, we provide a small example.
+For example, @montufar2011refinements show that any distribution on $\{0, 1\}^{\nv}$ can be approximated arbitrarily well by a RBM with $2^{\nv-1} - 1$ hidden units.
+We provide a small example that illustrates that in fact there can be many such approximations.
 
@@
   kable(caption = "\\label{tab:theta}Parameters used to fit a test case with $V = H = 4$.
-This parameter vector was chosen as a sampled value of $\\boldsymbol \\theta$ that was not near the convex hull of the sufficient statistics for a grid point in Figure \\ref{fig:degen_plots} with $< 5$\\% near-degeneracy.", align = rep(c("l", "r"), times = 3))
+This parameter vector was chosen as a sampled value of $\\boldsymbol \\theta$ that was not near the convex hull of the sufficient statistics for a grid point in Figure \\ref{fig:degen_plots} with $< 5$\\% near-degeneracy.", align = rep(c("r", "r"), times = 3))
 ```
@@
 We considered a test case with $\nv = \nh = 4$ and parameters given in Table \ref{tab:theta}.
-This parameter vector was chosen as a sampled value of $\boldsymbol \theta$ that was not near the convex hull of the sufficient statistics for a grid point in Figure \ref{fig:degen_plots} with $< 5$\% near-degeneracy.
+This parameter vector was chosen as a sampled value of $\boldsymbol \theta$ that was not near the convex hull of the space of values of sufficient statistics for a grid point in Figure \ref{fig:degen_plots} with $< 5$\% near-degeneracy.
 We simulated $n = 5,000$ realizations of visibles as a training set and fit the RBM using three Bayes methodologies.
@@
 These range from  $`r round(min(M_eff_tmp[, "BwTNML"]), 2)`$ to $`r round(max(M_eff_tmp[, "BwTNML"]), 2)`$ for BwTNML, while BwTNLV only yields between $`r round(min(M_eff_tmp[, "BwTNLV"]), 2)`$ and $`r round(max(M_eff_tmp[, "BwTNLV"]), 2)`$ effective draws.
-Thus, BwTNLV would require at least $`r round(min(M_eff_tmp[, "BwTNML"])/min(M_eff_tmp[, "BwTNLV"]), 1)`$ times more iterations to be run of the MCMC chain in order to achieve the same amount of effective information about the posterior distribution.
+Thus, BwTNLV would require at least $`r round(min(M_eff_tmp[, "BwTNML"])/min(M_eff_tmp[, "BwTNLV"]), 1)`$ times as many iterations to be run of the MCMC chain in order to achieve the same amount of effective information about the posterior distribution.
 For this reason, consistent with the ACF results in Figure \ref{fig:acf}, BwTNLV does not seem to be an effective method for fitting the RBM if computing resources are at all limited.

### 2016-12-01 writing/note.Rmd
nordman and vardeman final edits

@@
 (A sequence of probability models results indexed by $N$ that accommodates data of expanding dimension.) Model instability is formally shown to occur when a certain log-probability ratio under such models grows faster than $N$.
-In this case, a one component change in the data sequence can provably shift probability by orders of magnitude.
+In this case, a one component change in the data sequence can shift probability by orders of magnitude.
 Also, as a measure of instability becomes more extreme, the resulting probability models are shown  to tend to degeneracy, placing all their probability on arbitrarily small portions of the sample space.
@@
 This can be used to identify when certain maximal probabilities in FSFS models are too extreme relative to the length $N$.
-In this case, a one component change in the data sequence may provably shift probability by orders of magnitude, and FSFS models are rigorously shown to become degenerate as the measure of instability increases.
+In this case, a one component change in the data sequence may shift probability by orders of magnitude, and FSFS models are rigorously shown to become degenerate as the measure of instability increases.
 Lastly, Section \ref{implications} emphasizes the implications of our model propriety results and proofs of the main results appear in the Appendix.
@@
 
-Defining $P_{\boldsymbol \theta_N}(\boldsymbol x)\equiv p_{N,\boldsymbol \lambda_N}(\boldsymbol x)$ with $\boldsymbol \theta_N=\boldsymbol \lambda_N$ as $\boldsymbol \theta_N=\lambda$ to a sequence of elements of $\Lambda \subset \mathbb{R}^k$ and noting that $P_{\boldsymbol \theta_N}(\boldsymbol x) > 0$ for all $\boldsymbol x \in \mathcal{X}^N$, these discrete exponential family models are special cases of the FSFS models.
+Defining $P_{\boldsymbol \theta_N}(\boldsymbol x)\equiv p_{N,\boldsymbol \lambda_N}(\boldsymbol x)$ with $\boldsymbol \theta_N=\boldsymbol \lambda_N$ to be a sequence of elements of $\Lambda \subset \mathbb{R}^k$ and noting that $P_{\boldsymbol \theta_N}(\boldsymbol x) > 0$ for all $\boldsymbol x \in \mathcal{X}^N$, these discrete exponential family models are special cases of the FSFS models.
 Such exponential models arise with spatial data on a lattice [@besag1974spatial], network data [@wasserman1994social; @handcock2003assessing], and even binomial sampling with $N$ iid Bernoulli random variables.
-(Note for random graphs or networks with, say, $m$ nodes, one may wish to consider $N={m \choose 2}$ edges as binary (presence/absence) variables $X_i$; in this case, the length $N$ of data sequence may naturally increase as a function of $m$.) For these exponential models, the dimension of the parameter $\boldsymbol \theta_N$ is the same for each $N$ (i.e., $\boldsymbol \theta_N$ lies in a parameter space of fixed Euclidean dimension $k$).
+(Note that for random graphs or networks with, say, $m$ nodes, one may wish to consider $N={m \choose 2}$ edges as binary (presence/absence) variables $X_i$.
+In this case, the length $N$ of data sequence may naturally increase as a function of $m$.) For these exponential models, the dimension of the parameter $\boldsymbol \theta_N$ is the same for each $N$ as $\boldsymbol \theta_N$ lies in a parameter space of fixed Euclidean dimension $k$.
 This need not be true for other types of FSFS models considered in Sections \ref{restricted-boltzmann-machines}-\ref{deep-learning}.
@@
 $$
-P_{\boldsymbol \theta} (\tilde{\boldsymbol x}) = \exp\left[ \boldsymbol \alpha^T \boldsymbol h + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^T \Gamma \boldsymbol x - \psi(\boldsymbol \theta)\right], \qquad \tilde{\boldsymbol x} = (\boldsymbol h, \boldsymbol x) \in \mathcal{X}^{N+N_H},
+P_{\boldsymbol \theta_N} (\tilde{\boldsymbol x}) = \exp\left[ \boldsymbol \alpha^T \boldsymbol h + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^T \Gamma \boldsymbol x - \psi(\boldsymbol \theta_N)\right], \qquad \tilde{\boldsymbol x} = (\boldsymbol h, \boldsymbol x) \in \mathcal{X}^{N+N_H},
 $$
@@
 $$
-\psi(\boldsymbol \theta) = \log \sum_{\tilde{\boldsymbol x} \in \mathcal{X}^{N + N_H}} \exp\left[ \boldsymbol \alpha^T \boldsymbol h + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^T \Gamma \boldsymbol x\right], \qquad \boldsymbol \theta \in \Theta,
+\psi(\boldsymbol \theta_N) = \log \sum_{\tilde{\boldsymbol x} \in \mathcal{X}^{N + N_H}} \exp\left[ \boldsymbol \alpha^T \boldsymbol h + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^T \Gamma \boldsymbol x\right], \qquad \boldsymbol \theta_N \in \Theta_N,
 $$ 
 is the normalizing function.
-Let $\boldsymbol \theta_N = (\boldsymbol \alpha, \boldsymbol \beta,\Gamma) \in \Theta \subset \mathbb{R}^{q(N)}$ with $q(N) = N + N_H + N*N_H$ denote the vector of parameters for the RBM.
+Let $\boldsymbol \theta_N = (\boldsymbol \alpha, \boldsymbol \beta,\Gamma) \in \Theta_N \subset \mathbb{R}^{q(N)}$ with $q(N) = N + N_H + N*N_H$ denote the vector of parameters for the RBM.
 
@@
 $$
-P_{\boldsymbol \theta} (\boldsymbol x) = \sum\limits_{\boldsymbol h \in \mathcal{X}^{N_H}} P_{\boldsymbol \theta} (\boldsymbol x, \boldsymbol h), \qquad \boldsymbol x \in \mathcal{X}^N.
+P_{\boldsymbol \theta_N} (\boldsymbol x) = \sum\limits_{\boldsymbol h \in \mathcal{X}^{N_H}} P_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h), \qquad \boldsymbol x \in \mathcal{X}^N.
 $$
@@
 Note that the vector of model parameters $\boldsymbol \theta_N$, of size $q(N)$, grows in size as a function of sample dimension $N$ to accommodate the dimension of visible variables $X_1, \dots, X_N$, and one may further choose the number $N_H$ of hidden variables to change with $N$ as well; in particular, the number $N_H$ of hiddens may also potentially increase with $N$.
-Additionally, as $|\mathcal{X}| = 2$ and $p_{\boldsymbol \theta}(\boldsymbol x) > 0$ for all $\boldsymbol x \in \mathcal{X}^N$, the RBM model specification for visibles is a FSFS model.
+Additionally, as $|\mathcal{X}| = 2$ and $P_{\boldsymbol \theta_N}(\boldsymbol x) > 0$ for all $\boldsymbol x \in \mathcal{X}^N$, the RBM model specification for visibles is a FSFS model.
 This example also indicates that models formed by marginalizing a base FSFS model (e.g., a type of exponential family model) is again a FSFS model class.
@@
 $$
-P_{\boldsymbol \theta} (\tilde{\boldsymbol x}) = \exp\left[ \sum\limits_{i = 1}^M\boldsymbol \alpha^{(i)T} \boldsymbol h^{(i)} + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^{(1)T} \Gamma^{(0)} \boldsymbol x + \sum\limits_{i = 1}^{M - 1} \boldsymbol h^{(i)T} \Gamma^{(i)} \boldsymbol h^{(i + 1)} - \psi(\boldsymbol \theta) \right],
+P_{\boldsymbol \theta_N} (\tilde{\boldsymbol x}) = \exp\left[ \sum\limits_{i = 1}^M\boldsymbol \alpha^{(i)T} \boldsymbol h^{(i)} + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^{(1)T} \Gamma^{(0)} \boldsymbol x + \sum\limits_{i = 1}^{M - 1} \boldsymbol h^{(i)T} \Gamma^{(i)} \boldsymbol h^{(i + 1)} - \psi(\boldsymbol \theta_N) \right],
 $$
@@
 $$
-\psi(\boldsymbol \theta) = \log \sum\limits_{\tilde{\boldsymbol x} \in \mathcal{X}^{N_{(H,1)} + \cdots + N_{(H,M)} +N}} \exp\left[ \sum\limits_{i = 1}^M\boldsymbol \alpha^{(i)T} \boldsymbol h^{(i)} + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^{(1)T} \Gamma^{(0)} \boldsymbol x + \sum\limits_{i = 1}^{M - 1} \boldsymbol h^{(i)T} \Gamma^{(i)} \boldsymbol h^{(i + 1)}\right], \qquad \boldsymbol \theta \in \Theta
+\psi(\boldsymbol \theta_N) = \log \sum\limits_{\tilde{\boldsymbol x} \in \mathcal{X}^{N_{(H,1)} + \cdots + N_{(H,M)} +N}} \exp\left[ \sum\limits_{i = 1}^M\boldsymbol \alpha^{(i)T} \boldsymbol h^{(i)} + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^{(1)T} \Gamma^{(0)} \boldsymbol x + \sum\limits_{i = 1}^{M - 1} \boldsymbol h^{(i)T} \Gamma^{(i)} \boldsymbol h^{(i + 1)}\right], \qquad \boldsymbol \theta_N \in \Theta_N
 $$ 
 is the normalizing function and parameters in the model are $\boldsymbol \beta \in \mathbb{R}^V$, $\boldsymbol \alpha^{(i)} \in \mathbb{R}^{N_{(H,i)}}$ for $i = 1, \dots, M$, along with a matrix $\Gamma^{(0)}$ of dimension $N_{(H,1)} \times N$, and matrices $\Gamma^{(i)}$ of dimension $N_{(H,i)} \times N_{(H,i+1)}$ for $i = 1, \dots, M-1$.
-Let $\boldsymbol \theta = (\boldsymbol \alpha^{(1)}, \dots, \boldsymbol \alpha^{(M)}, \boldsymbol \beta,\Gamma^{(0)}, \dots, \Gamma^{(M - 1)}) \in \Theta \subset \mathbb{R}^{q(N)}$ denote the combined vector of parameters with total length $q(N)= N_{(H,1)}+\cdots N_{(H,M)} + N + N_{(H,1)}*N+N_{H,2}*H_{(H,1)}+\cdots +N_{(H,M)}*H_{(H,M)-1}$.
+Let $\boldsymbol \theta_N = (\boldsymbol \alpha^{(1)}, \dots, \boldsymbol \alpha^{(M)}, \boldsymbol \beta,\Gamma^{(0)}, \dots, \Gamma^{(M - 1)}) \in \Theta_N \subset \mathbb{R}^{q(N)}$ denote the combined vector of parameters with total length $q(N)= N_{(H,1)}+\cdots N_{(H,M)} + N + N_{(H,1)}*N+N_{H,2}*H_{(H,1)}+\cdots +N_{(H,M)}*H_{(H,M)-1}$.
 
@@
 $$
-P_{\boldsymbol \theta} (\boldsymbol x) = \sum\limits_{(\boldsymbol h^{(1)}, \dots, \boldsymbol h^{(M)}) \in \mathcal{X}^{N_{(H,1)} + \cdots + N_{(H,M)}}} p_{\boldsymbol \theta} (\tilde{\boldsymbol x}), \qquad \boldsymbol x \in \mathcal{X}^N
+P_{\boldsymbol \theta_N} (\boldsymbol x) = \sum\limits_{(\boldsymbol h^{(1)}, \dots, \boldsymbol h^{(M)}) \in \mathcal{X}^{N_{(H,1)} + \cdots + N_{(H,M)}}} P_{\boldsymbol \theta_N} (\tilde{\boldsymbol x}), \qquad \boldsymbol x \in \mathcal{X}^N
 $$
@@
 \begin{align}
-\frac{1}{N} \log \left[\frac{\max\limits_{(x_1, \dots, x_N) \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(x_1, \dots, x_N)}{\min\limits_{(x_1, \dots, x_N) \in \mathcal{X}^N}P_{\boldsymbol \theta}(x_1, \dots, x_N)}\right] \equiv \frac{1}{N} \text{ELPR}_N(\boldsymbol \theta_N).
+\frac{1}{N} \log \left[\frac{\max\limits_{(x_1, \dots, x_N) \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(x_1, \dots, x_N)}{\min\limits_{(x_1, \dots, x_N) \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(x_1, \dots, x_N)}\right] \equiv \frac{1}{N} \text{ELPR}_N(\boldsymbol \theta_N).
 \label{eqn:elpr}
@@
 \begin{align*}
-\lim\limits_{N \rightarrow \infty} \frac{1}{N} \text{ELPR}(\boldsymbol \theta_N) \equiv \lim\limits_{N \rightarrow \infty} \frac{1}{N} \log \left[\frac{\max\limits_{(x_1, \dots, x_N) \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(x_1, \dots, x_N)}{\min\limits_{(x_1, \dots, x_N) \in \mathcal{X}^N}P_{\boldsymbol \theta}(x_1, \dots, x_N)}\right] = \infty.
+\lim\limits_{N \rightarrow \infty} \frac{1}{N} \text{ELPR}(\boldsymbol \theta_N) \equiv \lim\limits_{N \rightarrow \infty} \frac{1}{N} \log \left[\frac{\max\limits_{(x_1, \dots, x_N) \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(x_1, \dots, x_N)}{\min\limits_{(x_1, \dots, x_N) \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(x_1, \dots, x_N)}\right] = \infty.
 \end{align*}
@@
 \label{prop:instab}
-Let $C > 0$ and let $\text{ELPR}(\boldsymbol \theta_N)$ be as in (\ref{eqn:elpr}) for an integer $N \ge 1$. If $$\frac{1}{N}\text{ELPR}_N(\boldsymbol \theta_N) > C,$$ 
+Let $\text{ELPR}(\boldsymbol \theta_N)$ be as in (\ref{eqn:elpr}) for an integer $N \ge 1$. For a given $C>0$, if $$\frac{1}{N}\text{ELPR}_N(\boldsymbol \theta_N) > C,$$ 
 then 
@@
 This is because $\left(\max\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)\right)^n$ is the largest probability possible under the joint model for the $n$ replications while $\left(\min\limits_{\boldsymbol x \in \mathcal{X}}P_{\boldsymbol \theta_N}(\boldsymbol x)\right)^n$ is the smallest probability.
-Thus,
+Thus, for the combined replications $\boldsymbol X_1, \ldots, \boldsymbol X_n$, the analog definition of the extremal log-probability becomes
 $$

### 2016-12-01 writing/draft.Rmd
forgot one more edit

@@
 After a burn-in period of 50 iterations selected by inspecting the trace plots, we assess the issue of mixing in two ways.
-First, the autocorrelation functions (ACF) from each posterior sample corresponding to a model probability for a visible vector outcome $\mathbf{v}=(v_1,v_2,v_3,v_4)\in\{\pm 1\}^4$ (i.e., computed from $\boldsymbol \theta$ under (\ref{eqn:pmf})) of visibles are assessed and plotted in Figure \ref{fig:acf} with BwTNLV in black and BwTNML in red.
+First, the autocorrelation functions (ACF) from each posterior sample corresponding to a model probability for a visible vector outcome $\mathbf{v}=(v_1,v_2,v_3,v_4)\in\{\pm 1\}^4$ (i.e., computed from $\boldsymbol \theta$ under (\ref{eqn:pmf})) are assessed and plotted in Figure \ref{fig:acf} with BwTNLV in black and BwTNML in red.
 As expected, ACF corresponding to the method that marginalizes out the hidden variables from the likelihood decreases to zero at a much faster rate, indicating better mixing for the chain.

### 2016-12-02 writing/draft.Rmd
date change

@@
   | ajkaplan@iastate.edu, dnordman@iastate.edu, vardeman@iastate.edu
-date: "November, 2016"
+date: "December, 2016"
 linestretch: 2 
@@
 
-Finally to consider the effect of $\boldsymbol \theta$ on potential model uninterpretability, we can look at the absolute difference between model expectations, E$\left[\boldsymbol X | \boldsymbol \theta\right]$, and expectations given independence, E$\left[\boldsymbol X | \boldsymbol \theta^*\right ]$ for the tiny toy RBM model where $\boldsymbol X = (V_1, H_1, V_1 H_1)$.
+Finally to consider the effect of $\boldsymbol \theta$ on potential model uninterpretability, we can look at the difference between model expectations, E$\left[\boldsymbol X | \boldsymbol \theta\right]$, and expectations given independence, E$\left[\boldsymbol X | \boldsymbol \theta^*\right ]$ for the tiny toy RBM model where $\boldsymbol X = (V_1, H_1, V_1 H_1)$.
 This difference is given by
@@
 
-Again, we can inspect these coordinate functions of this vector difference to look for a relationship between parameter values and large values of $||\text{E}[\boldsymbol X|\boldsymbol \theta] - \text{E}[\boldsymbol X| \boldsymbol \theta^*]||$ as a signal of uninterpretability for the toy RBM.
+Again, we can inspect these coordinate functions of this vector difference to look for a relationship between parameter values and large values of $\text{E}[\boldsymbol X|\boldsymbol \theta] - \text{E}[\boldsymbol X| \boldsymbol \theta^*]$ as a signal of uninterpretability for the toy RBM.
 

### 2016-12-02 writing/note.Rmd
date change

@@
   | ajkaplan@iastate.edu, dnordman@iastate.edu, vardeman@iastate.edu
-date: "November, 2016"
+date: "December, 2016"
 linestretch: 2

### 2016-12-02 code/degeneracy/apps/degen_instab_uninterp.Rmd
dumb restructure

---
output:
  html_document:
    theme: readable
resource_files:
- index.html
runtime: shiny
---

#Degeneracy, instability, and uninterpretability... Oh my!

```{r libraries, message=FALSE, warning=FALSE, echo=FALSE}
library(knitr)
library(ggplot2)
library(dplyr)
library(tidyr)
library(shiny)
library(gridExtra)

opts_chunk$set(echo=FALSE, message=FALSE, warnings=FALSE)
theme_set(theme_bw(base_family="serif"))

source("functions.R")
```

```{r}
load("grid_measures.RData")

top_plots <- list(p1 = three_ways %>%
    ggplot() +
    geom_tile(aes(x = r1, y = r2, fill = frac_degen)) +
    geom_contour(aes(x = r1, y = r2, z = frac_degen), colour = "black", bins = 8) +
    geom_contour(aes(x = r1, y = r2, z = frac_degen), colour = "black", breaks = .05, size = 1.5) +
    #geom_abline(aes(intercept = 0, slope = 1), alpha = .5, lty = 2) +
    scale_fill_gradient("Fraction near-degenerate", low = "yellow", high = "red") +
    facet_grid(Visibles~Hiddens) +
    xlab(expression(group("||", theta[main], "||"))) +
    ylab(expression(group("||", theta[interaction], "||"))) +
    theme(aspect.ratio = 1) +
    theme(legend.position = "bottom") +
    ggtitle("Near-degeneracy"),
    p2 =  three_ways %>%
    ggplot() +
    geom_tile(aes(x = r1, y = r2, fill = mean_abs_diff)) +
    geom_contour(aes(x = r1, y = r2, z = mean_abs_diff), colour = "black", bins = 8) +
    #geom_contour(aes(x = r1, y = r2, z = mean_abs_diff), colour = "black", breaks = .05, size = 1.5) +
    #geom_abline(aes(intercept = 0, slope = 1), alpha = .5, lty = 2) +
    scale_fill_gradient(expression(group("|", E(group("", bold(X), "|")*bold(theta)) - E(group("", bold(X), "|")*plain(independence)), "|")), low = "yellow", high = "red") +
    facet_grid(Visibles~Hiddens) +
    xlab(expression(group("||", theta[main], "||"))) +
    ylab(expression(group("||", theta[interaction], "||"))) +
    theme(aspect.ratio = 1) +
    theme(legend.position = "bottom") +
    ggtitle("Uninterpretability"),
    p3 = three_ways %>%
    ggplot() +
    geom_tile(aes(x = r1, y = r2, fill = mean_max_q)) +
    geom_contour(aes(x = r1, y = r2, z = mean_max_q), colour = "black", bins = 8) +
    #geom_contour(aes(x = r1, y = r2, z = mean_max_q), colour = "black", breaks = .05, size = 1.5) +
    #geom_abline(aes(intercept = 0, slope = 1), alpha = .5, lty = 2) +
    scale_fill_gradient(expression(frac(max(Q), H + V)), low = "yellow", high = "red") +
    facet_grid(Visibles~Hiddens) +
    xlab(expression(group("||", theta[main], "||"))) +
    ylab(expression(group("||", theta[interaction], "||"))) +
    theme(aspect.ratio = 1) +
    theme(legend.position = "bottom") +
    ggtitle("Instability")
)

inputPanel(
  selectizeInput("variable", "Select radius to fix:", 
                 choices = list("Main effects" = "r1", "Interaction effects" = "r2"), 
                 multiple = FALSE,
                 selected = "r1"),
  numericInput("slice", "Fix radius at:",
               min = min(three_ways$r1), 
               max = max(three_ways$r1), 
               value = min(three_ways$r1), step = .1)
)

renderPlot({

  if(input$variable == "r1") {
    p1 <- top_plots$p1 + geom_vline(aes(xintercept = input$slice))
    p2 <- top_plots$p2 + geom_vline(aes(xintercept = input$slice))
    p3 <- top_plots$p3 + geom_vline(aes(xintercept = input$slice))
  } else {
    p1 <- top_plots$p1 + geom_hline(aes(yintercept = input$slice))
    p2 <- top_plots$p2 + geom_hline(aes(yintercept = input$slice))
    p3 <- top_plots$p3 + geom_hline(aes(yintercept = input$slice))
  }

  grid.arrange(p1, p3, p2, nrow = 1)
})

filter_dat <- reactive({
    slice <- unique(three_ways[which.min(abs(three_ways[, input$variable] %>% data.matrix() - input$slice)), input$variable])[1] %>% 
    as.numeric()

    three_ways[three_ways[,input$variable] == slice, ]
})

renderPlot({
  axes <- c("r1", "r2")
  variable <- axes[axes != input$variable]
  label <- ifelse(variable == "r1", 
                  expression(group("||", theta[main], "||")), 
                  expression(group("||", theta[interaction], "||")))

  dat <- filter_dat()

   dat %>%
    ggplot() +
    geom_line(aes_string(variable, "frac_degen")) +
    facet_grid(Visibles~Hiddens) +
    xlab(label) +
    ylab("Fraction Near-degenerate") +
    theme(aspect.ratio = 1) +
    ggtitle("Near-degeneracy") -> p1

  dat %>%
    ggplot() +
    geom_line(aes_string(variable, "mean_abs_diff")) +
    facet_grid(Visibles~Hiddens) +
    xlab(label) +
    ylab(expression(group("|", E(group("", bold(X), "|")*bold(theta)) - E(group("", bold(X), "|")*plain(independence)), "|"))) +
    theme(aspect.ratio = 1) +
    ggtitle("Uninterpretability") -> p2

  dat %>%
    ggplot() +
    geom_line(aes_string(variable, "mean_max_q")) +
    facet_grid(Visibles~Hiddens) +
    xlab(label) +
    ylab(expression(frac(max(Q), H + V))) +
    theme(aspect.ratio = 1) +
    ggtitle("Instability") -> p3

  grid.arrange(p1, p3, p2, nrow = 1)
})

```

The highly flexible nature of the RBM ($H + V + HV$ parameters) makes the following characteristics of model impropriety of particular concern.

 - **Near-degeneracy.** Occurs when there is a disproportionate amount of probability placed on only a few elements of the sample space by the model. If random variables in the neg-potential function $Q(\cdot)$, having support set $\mathcal{S}$, have a collective mean $\mu(\boldsymbol \theta)$ close to the boundary of the convex hull of $\mathcal{S}$.
 - **Instability.** Small changes in natural parameters result in large changes in probability masses, excessive sensitivity. If for any $C > 0$ there exists $N_C > 0$ such that 
$\max\limits_{\boldsymbol x_N \in \mathcal{X}_N}[Q(\boldsymbol x_N)] > CN$ for all $N > N_C$, where $Q(\cdot)$ is the neg-potential function of the model.
 - **Uninterpretability.** Due to the existence of dependence, marginal mean-structure no longer maintained. If the magnitude of the difference between model expectations and expectations under independence (dependence parameters of zero), $\left|\text{E}(\boldsymbol X | \boldsymbol \theta) - \text{E}(\boldsymbol X | \emptyset ) \right|$, is large.

### 2016-12-02 code/degeneracy/apps/index.Rmd
dumb restructure

---
title: "Exploring the restricted Boltzmann machine (RBM)"
output: 
  html_document: 
    theme: readable
---
A selection of `Shiny` applications to accompany the paper "An exposition on the propriety of restricted Boltzmann machines" by <a href="http://andeekaplan.com">Andee Kaplan</a>, <a href="http://dnordman.public.iastate.edu/">Daniel Nordman</a>, and <a href="http://www.public.iastate.edu/~vardeman/">Stephen Vardeman</a>, all in the Department of Statistics at Iowa State University. All code is available on <a href="http://github.com/andeek/deep-learning">GitHub</a>.

<div class="row">
<div class="col-md-3">
<a href="mu_parameterization.Rmd">
<img src="images/icon_mu.png" class="img-responsive thumbnail">
</a>
</div>
<div class="col-md-9">
<h4>$\mu$-parameterization</h4>
<p>Explore the behavior of the $\mu$-parametrized parameters for various dimensional RBMs.</p>
</div>
</div>

<div class="row">
<div class="col-md-3">
<a href="mu_parameterization_3d.Rmd">
<img src="images/icon_mu_3d.png" class="img-responsive thumbnail">
</a>
</div>
<div class="col-md-9">
<h4>$\mu$-parameterization in 3 dimensions</h4>
<p>For one hidden and one visible node, sample points uniformly on a 3 dimensional sphere of radius `r` centered at `(0,0,0)` and view the resulting expected values in 3 dimensions.</p>
</div>
</div>

<div class="row">
<div class="col-md-3">
<a href="degen_magnitude.Rmd">
<img src="images/icon_perc_degen.png" class="img-responsive thumbnail">
</a>
</div>
<div class="col-md-9">
<h4>Degeneracy and parameter magnitude</h4>
<p>What is the relationship between parameter magnitude and near-degeneracy?</p>
</div>
</div>

<div class="row">
<div class="col-md-3">
<a href="degen_instab_uninterp.Rmd">
<img src="images/icon_degeneracy.png" class="img-responsive thumbnail">
</a>
</div>
<div class="col-md-9">
<h4>Degeneracy, instability, and uninterpretability... Oh my!</h4>
<p>The highly flexible nature of the RBM ($H + V + HV$ parameters) makes three characteristics of model impropriety of particular concern.</p>
</div>
</div>

### 2016-12-02 code/degeneracy/apps/mu_parameterization_3d.Rmd
dumb restructure

--- 2015-11-18 degeneracy/apps/mu_parameterization_3d.Rmd
@@
 ---
-author: "Andee Kaplan"
-date: "11/18/2014"
 output:
@@
 library(ggvis)
+library(shinythemes)
 
@@
     fluidPage(
-      tags$head(
-        includeCSS("www/readable.min.css"),
-        includeCSS("www/css/numeric_inputs.css")
-      ),
-      responsive = FALSE,
+      theme = shinytheme("readable"),
       fluidRow(style = "padding-bottom: 20px;",
-        column(4, radioButtons("encoding", label="Encoding:", 
+        column(2, radioButtons("encoding", label="Encoding:", 
                                       choices = list("Negative (-1,1)" = "negative", "Binary (0,1)" = "binary"),
                                       inline = FALSE)),
-        column(4, sliderInput("rSlider", label = "Radius:", min = 0, max = 10, value=0, step = 0.1, animate = list(interval=1000, loop=FALSE, playButton=icon("play"), pauseButton=icon("pause")), )),
-        column(4, 
-               column(6, span("Translate origin by (x, y, z): ")),
-               column(2, numericInput("x", value=NA, label="")),
-               column(2, numericInput("y", value=NA, label="")), 
-               column(2, numericInput("z", value=NA, label=""))
+        column(4, sliderInput("rSlider", label = "Radius:", min = 0, max = 10, value=0, step = 0.1, animate = list(interval=1000, loop=FALSE, playButton=icon("play"), pauseButton=icon("pause")))),
+        column(6, 
+               column(3, span("Translate origin by (x, y, z): ")),
+               column(3, numericInput("x", value=NA, label="")),
+               column(3, numericInput("y", value=NA, label="")), 
+               column(3, numericInput("z", value=NA, label=""))
                )
@@
       stat <- stats(H, V, input$encoding)
-      theta <- sample_sphere(space = ncol(stat), n, input$rSlider)  + matrix(rep(c(x,y,z), 10), nrow = 3)
+      theta <- sample_sphere(stat, n, input$rSlider) + matrix(rep(c(x,y,z), n), nrow = 3)
       exp_vals <- expected_value(stats = stat, theta = theta, normalized = TRUE)
@@
       lines <- generate_hull_from_stats(stat)
-      dist <- apply(exp_vals, 2, function(x){ min_distance_to_hull(x, lines) })
-      return(data.frame(dist = dist))      
+      dist <- do.call(rbind, apply(exp_vals, 2, function(x){ min_distance_to_hull(x, lines) }))
+      return(dist)      
     })
@@
           if(i < j) {
-            start <- s$xyz.convert(stat[i,])
-            end <- s$xyz.convert(stat[j,])
+            start <- s$xyz.convert(t(stat[i,]))
+            end <- s$xyz.convert(t(stat[j,]))
             segs <- rbind(segs, c(start$x, start$y, end$x, end$y))
@@
     min_dist_data %>%
-      ggvis(~dist) %>%
+      ggvis(~min_dist) %>%
       layer_histograms() %>%
       scale_numeric("x", domain = c(0,1)) %>%
-      add_axis("x", title = "Minimum distance to convex hull edge") %>%
-      set_options(width=200, height=200, keep_aspect = TRUE, resizable = FALSE) %>%
+      add_axis("x", title = "Minimum approximate distance to convex hull edge") %>%
+      set_options(width="auto", height=100, keep_aspect = TRUE, resizable = FALSE) %>%
       bind_shiny(plot_id = "min_dist_hist")
   },
-  options = list(height = 800)
+  options = list(height = 600)
 )

### 2016-12-02 code/4x4 model/fit_adaptive_mh_within_gibbs.R

@@ -1,0 +1,45 @@
+#libraries --------------------
+library(dplyr)
+library(tidyr)
+library(ggplot2)
+source("sample_functs.R")
+
+#data and params ------------------------
+H <- 4
+V <- 4
+mc.iter <- 5000
+set.seed(102285) #reproducible seed
+
+load("written/sample_images.Rdata")
+load("written/variance_params.Rdata")
+load("written/sample.params.Rdata")
+
+params <- list(main_hidden = rnorm(H, mean = 0, sd = sqrt(variance_params$C)),
+               main_visible = rnorm(V, mean = 0, sd = sqrt(variance_params$C)),
+               interaction = matrix(rnorm(H*V, mean = 0, sd = sqrt(variance_params$C_prime)), nrow = H))
+h <- .01
+s1 <- 10e-4
+s2 <- 10e4
+
+models_good <- sample_adaptive_mh_within_gibbs(visibles = flat_images_good$visibles, params0 = params, C = variance_params$C, C_prime = variance_params$C_prime, h = h, s1 = s1, s2 = s2, mc.iter = mc.iter)
+models_bad <- sample_adaptive_mh_within_gibbs(visibles = flat_images_good$visibles, params0 = params, C = variance_params$C, C_prime = variance_params$C, h = h, s1 = s1, s2 = s2, mc.iter = mc.iter)
+models_degen <- sample_adaptive_mh_within_gibbs(visibles = flat_images_degen$visibles, params0 = params, C = variance_params$C, C_prime = variance_params$C_prime, h = h, s1 = s1, s2 = s2, mc.iter = mc.iter)
+
+save(models_good, models_bad, models_degen, file = "written/fitted_models_adaptive_mh_full_distn.Rdata")
+
+#burnin + thinning
+burnin_thinning <- function(models) {
+  idx <- which(1:mc.iter %% 5 == 0 & 1:mc.iter > mc.iter/2)
+  models$theta <- models$theta[idx, ]
+  models$hiddens <- models$hiddens[idx, , ]
+  #models$post_pred <- models$post_pred[idx, , ]
+  models$distn <- models$distn[idx, , ]
+  
+  return(models)
+}
+
+models_good <- burnin_thinning(models_good)
+models_bad <- burnin_thinning(models_bad)
+models_degen <- burnin_thinning(models_degen)
+
+save(models_good, models_bad, models_degen, file = "written/fitted_models_adaptive_mh_distn.Rdata")

### 2016-12-02 code/4x4 model/fit_gibbs.R

@@ -1,0 +1,52 @@
+#libraries --------------------
+library(dplyr)
+library(tidyr)
+library(ggplot2)
+source("functs_sample.R")
+
+#data and params ------------------------
+H <- 4
+V <- 4
+mc.iter <- 5000
+set.seed(102285) #reproducible seed
+
+#rule of thumb for variance params
+C <- 1/(H + V)
+C_prime = 1/(H*V)
+
+load("written/sample_images.Rdata")
+load("written/params_theta.Rdata")
+
+params <- list(main_hidden = rnorm(H, mean = 0, sd = sqrt(C)),
+               main_visible = rnorm(V, mean = 0, sd = sqrt(C)),
+               interaction = matrix(rnorm(H*V, mean = 0, sd = sqrt(C_prime)), nrow = H))
+
+N <- nrow(flat_images_good$visibles)
+consts <- c(5.5, 5.6, 5.7) #tune constant 
+
+for(const in consts) {
+  models_good <- sample_gibbs(visibles = flat_images_good$visibles, params0 = params, C = const*C/N, C_prime = const*C_prime/N, mc.iter = mc.iter)
+  models_bad <- sample_gibbs(visibles = flat_images_good$visibles, params0 = params, C = const*C/N, C_prime = const*C/N, mc.iter = mc.iter)
+  #models_degen <- sample_gibbs(visibles = flat_images_degen$visibles, params0 = params, C = const*variance_params$C/N, C_prime = const*variance_params$C_prime/N, mc.iter = mc.iter)
+  
+  #burnin + thinning
+  burnin_thinning <- function(models) {
+    idx <- which(1:mc.iter %% 5 == 0 & 1:mc.iter > mc.iter/2)
+    models$params$main_hidden <- models$params$main_hidden[idx, ]
+    models$params$main_visible <- models$params$main_visible[idx, ]
+    models$params$interaction <- models$params$interaction[idx, , ]
+    models$hiddens <- models$hiddens[idx, , ]
+    #models$post_pred <- models$post_pred[idx, , ]
+    models$distn <- models$distn[idx, , ]
+    
+    return(models)
+  }
+  
+  models_good <- burnin_thinning(models_good)
+  models_bad <- burnin_thinning(models_bad)
+  #models_degen <- burnin_thinning(models_degen)
+  
+  save(models_good, models_bad, file = paste0("written/fitted_models_jing_", const, ".Rdata"))
+  
+  
+}

### 2016-12-02 code/4x4 model/fit_mh_within_gibbs.R

@@ -1,0 +1,42 @@
+#libraries --------------------
+library(dplyr)
+library(tidyr)
+library(ggplot2)
+source("sample_functs.R")
+
+#data and params ------------------------
+H <- 4
+V <- 4
+mc.iter <- 10000
+set.seed(102285) #reproducible seed
+
+load("written/sample_images.Rdata")
+load("written/variance_params.Rdata")
+load("written/sample.params.Rdata")
+
+params <- list(main_hidden = rnorm(H, mean = 0, sd = sqrt(variance_params$C)),
+               main_visible = rnorm(V, mean = 0, sd = sqrt(variance_params$C)),
+               interaction = matrix(rnorm(H*V, mean = 0, sd = sqrt(variance_params$C_prime)), nrow = H))
+tau1 <- 1
+tau2 <- .25 #tuning parameter
+
+models_good <- sample_mh_within_gibbs(visibles = flat_images_good$visibles, params0 = params, C = variance_params$C, C_prime = variance_params$C_prime, tau_main = tau1, tau_interaction = tau2, mc.iter = mc.iter)
+models_bad <- sample_mh_within_gibbs(visibles = flat_images_good$visibles, params0 = params, C = variance_params$C, C_prime = variance_params$C, tau_main = tau1, tau_interaction = tau2, mc.iter = mc.iter)
+models_degen <- sample_mh_within_gibbs(visibles = flat_images_degen$visibles, params0 = params, C = variance_params$C, C_prime = variance_params$C_prime, tau_main = tau1, tau_interaction = tau2, mc.iter = mc.iter)
+
+#burnin + thinning
+burnin_thinning <- function(models) {
+  idx <- which(1:mc.iter %% 5 == 0 & 1:mc.iter > mc.iter/2)
+  models$theta <- models$theta[idx, ]
+  models$hiddens <- models$hiddens[idx, , ]
+  #models$post_pred <- models$post_pred[idx, , ]
+  models$distn <- models$distn[idx, , ]
+  
+  return(models)
+}
+
+models_good <- burnin_thinning(models_good)
+models_bad <- burnin_thinning(models_bad)
+models_degen <- burnin_thinning(models_degen)
+
+save(models_good, models_bad, models_degen, file = "written/fitted_models_mh_distn.Rdata")

### 2016-12-02 code/4x4 model/fit_ml.R

@@ -1,0 +1,64 @@
+#libraries --------------------
+library(dplyr)
+library(tidyr)
+library(ggplot2)
+source("sample_functs.R")
+source("ml_functs.R")
+
+#data and params ------------------------
+H <- 4
+V <- 4
+load("written/sample_images.Rdata")
+load("written/sample.params.Rdata")
+set.seed(102285)
+
+
+#system of nonlinear eqns to solve -----------------
+parms <- list(visibles = flat_images_good$visibles, 
+              hiddens = flat_images_good$hiddens,
+              H = H, V = V)
+
+max.iter <- 10
+theta <- matrix(NA, nrow = max.iter + 1, ncol = H + V + H*V)
+likelihood <- matrix(NA, nrow = max.iter + 1, ncol = 1)
+
+#inits
+theta[1, ] <- sample.params %>% ungroup() %>% filter(!near_hull) %>% select(starts_with("v"), starts_with("h"), starts_with("theta"), -H, -V) %>% data.matrix()
+likelihood[1, ] <- loglik(theta = theta[1, ], parms = parms)
+
+for(i in 1:max.iter) {
+  cat(paste0(i, "\r"))
+  optim(par = theta[i, ],
+        fn = loglik,
+        #gr = loglik_derivs,
+        parms = parms,
+        method = "BFGS",
+        control = list(fnscale = -1, maxit = 1)) -> mle
+  
+  if(mle$convergence == 1) {
+    theta[i + 1, ] <- mle$par
+    likelihood[i + 1, ] <- mle$value
+  } else {
+    return
+  }
+}
+
+save(theta, likelihood, file = "written/fitted_ml_margin.Rdata")
+
+
+cbind(theta, likelihood) %>%
+  data.frame() %>%
+  mutate(iter = 1:(max.iter + 1)) %>%
+  rename(likelihood = X25) %>%
+  gather(variable, value, -iter) %>%
+  ggplot() +
+  geom_line(aes(x = iter, y = value)) +
+  facet_wrap(~variable, scales = "free_y") 
+
+
+
+
+
+
+
+

### 2016-12-02 code/4x4 model/fit_single_adaptive_mh_within_gibbs.R

@@ -1,0 +1,52 @@
+#libraries --------------------
+library(dplyr)
+library(tidyr)
+library(ggplot2)
+source("functs_sample.R")
+
+#data and params ------------------------
+H <- 4
+V <- 4
+mc.iter <- 5000
+set.seed(102285) #reproducible seed
+
+load("written/sample_images.Rdata")
+load("written/params_theta.Rdata")
+
+params <- list(main_hidden = rep(0, H),
+               main_visible = rep(0, V),
+               interaction = rep(0, H*V))
+
+h <- .01
+s1 <- 1e-4
+s2 <- 1e4
+trunc_const <- 1
+
+#rule of thumb for variance params
+C <- 1/(H + V)
+C_prime = 1/(H*V)
+
+#run models ------------
+models_good <- sample_single_adaptive_mh_within_gibbs(visibles = flat_images_good$visibles, params0 = params, C = C, C_prime = C_prime, trunc_const = trunc_const, h = h, s1 = s1, s2 = s2, mc.iter = mc.iter)
+models_bad <- sample_single_adaptive_mh_within_gibbs(visibles = flat_images_good$visibles, params0 = params, C = C, C_prime = C, trunc_const = trunc_const, h = h, s1 = s1, s2 = s2, mc.iter = mc.iter)
+#models_degen <- sample_single_adaptive_mh_within_gibbs(visibles = flat_images_degen$visibles, params0 = params, C = C, C_prime = C_prime, trunc_const = trunc_const, h = h, s1 = s1, s2 = s2, mc.iter = mc.iter)
+
+save(models_good, models_bad, file = "written/fitted_models_trunc_full.Rdata")
+
+#burnin + thinning --------------
+burnin_thinning <- function(models) {
+  idx <- which(1:mc.iter %% 5 == 0 & 1:mc.iter > mc.iter/2)
+  models$theta <- models$theta[idx, ]
+  models$hiddens <- models$hiddens[idx, , ]
+  #models$post_pred <- models$post_pred[idx, , ]
+  models$distn <- models$distn[idx, , ]
+  models$var <- models$var[idx, ]
+  
+  return(models)
+}
+
+models_good <- burnin_thinning(models_good)
+models_bad <- burnin_thinning(models_bad)
+#models_degen <- burnin_thinning(models_degen)
+
+save(models_good, models_bad, file = "written/fitted_models_trunc_thin.Rdata")

### 2016-12-02 code/4x4 model/fit_single_adaptive_mh_within_gibbs_marginal.R

@@ -1,0 +1,51 @@
+#libraries --------------------
+library(dplyr)
+library(tidyr)
+library(ggplot2)
+source("functs_sample.R")
+
+#data and params ------------------------
+H <- 4
+V <- 4
+mc.iter <- 2000
+set.seed(102285) #reproducible seed
+
+load("written/sample_images.Rdata")
+load("written/params_theta.Rdata")
+
+params <- list(main_hidden = rep(0, H),
+               main_visible = rep(0, V),
+               interaction = rep(0, H*V))
+
+h <- .01
+s1 <- 1e-4
+s2 <- 1e4
+trunc_const <- 1
+
+#rule of thumb for variance params
+C <- 1/(H + V)
+C_prime = 1/(H*V)
+
+models_good <- sample_single_adaptive_mh_within_gibbs(visibles = flat_images_good$visibles, params0 = params, C = C, C_prime =  C_prime, trunc_const = trunc_const, h = h, s1 = s1, s2 = s2, mc.iter = mc.iter, conditional_function = log_conditional_marginal)
+models_bad <- sample_single_adaptive_mh_within_gibbs(visibles = flat_images_good$visibles, params0 = params, C = C, C_prime = C, trunc_const = trunc_const, h = h, s1 = s1, s2 = s2, mc.iter = mc.iter, conditional_function = log_conditional_marginal)
+#models_degen <- sample_single_adaptive_mh_within_gibbs(visibles = flat_images_degen$visibles, params0 = params, C = variance_params$C, C_prime = variance_params$C_prime, trunc_const = trunc_const, h = h, s1 = s1, s2 = s2, mc.iter = mc.iter, conditional_function = log_conditional_marginal)
+
+save(models_good, models_bad, file = "written/fitted_models_trunc_marginal_full.Rdata")
+
+#burnin + thinning
+burnin_thinning <- function(models) {
+  idx <- which(1:mc.iter %% 2 == 0 & 1:mc.iter > mc.iter/2)
+  models$theta <- models$theta[idx, ]
+  models$hiddens <- models$hiddens[idx, , ]
+  #models$post_pred <- models$post_pred[idx, , ]
+  models$distn <- models$distn[idx, , ]
+  models$var <- models$var[idx, ]
+  
+  return(models)
+}
+
+models_good <- burnin_thinning(models_good)
+models_bad <- burnin_thinning(models_bad)
+#models_degen <- burnin_thinning(models_degen)
+
[…]

### 2016-12-02 code/4x4 model/functs_ml.R

@@ -1,0 +1,152 @@
+#libraries --------------------
+library(dplyr)
+library(tidyr)
+
+loglik_num <- function(theta, parms) {
+  #theta is a H + V + H*V length vector containing main_visible, then main_hidden, then interactions
+  
+  #parms is a named list containing data and H and V
+  visibles <- parms$visibles
+  hiddens <- parms$hiddens
+  H <- parms$H
+  V <- parms$V
+  
+  #things for derivative, build data and possible values
+  N <- visibles %>% nrow()
+  data <- cbind(visibles, hiddens)
+  for(i in 1:V) {
+    for(j in 1:H) {
+      data <- cbind(data, visibles[i,]*hiddens[j,])
+    }
+  }
+  
+  sum(theta*colSums(data)/N)
+}
+
+
+loglik <- function(theta, parms) {
+  #theta is a H + V + H*V length vector containing main_visible, then main_hidden, then interactions
+  
+  #parms is a named list containing data and H and V
+  visibles <- parms$visibles %>% data.frame()
+  H <- parms$H
+  V <- parms$V
+  names(visibles) <- paste0("v", 1:V)
+  
+  #build data and possible values
+  N <- visibles %>% nrow()
+  possibles <- stats(H, V) 
+  W <- exp(possibles %*% theta)
+  
+  # marginalize out the hiddens
+  data <- visibles %>%
+    left_join(possibles %>% data.frame(), by = paste0("v", 1:V))
+
+  data %>%
+    group_by_(.dots = paste0("h", 1:H)) %>%
+    do(parms = list(visibles = data.frame(.) %>% select(starts_with("v")) %>% data.matrix(),
+                    hiddens = data.frame(.) %>% select(starts_with("h")) %>% data.matrix(),
+                    H = H,
+                    V = V)) %>%
+    mutate(A = loglik_num(theta = theta, parms)) %>%
+    mutate(num = exp(A)) %>%
+    ungroup() %>%
+    summarise(numerator = sum(num)) %>%
+    as.numeric() -> numerator
+  
+    log(numerator) - log(sum(W))
+}
+
+#not working
+loglik_derivs <- function(theta, parms) {
+  #theta is a H + V + H*V length vector containing main_visible, then main_hidden, then interactions
+  
[…]

### 2016-12-02 code/4x4 model/functs_sample.R

@@ -1,0 +1,393 @@
+sample_main_hidden <- function(hiddens, C) {
+  #hiddens is an NxH matrix of hidden node values
+  #C is the variance constant
+  res <- sapply(colSums(hiddens), function(mean) rnorm(1, mean*C, sqrt(C)))
+  names(res) <- paste0("H", 1:ncol(hiddens))
+  return(res)
+}
+
+sample_main_visible <- function(visibles, C) {
+  #visibles is an NxV matrix of visible node values
+  #C is the variance constant
+  res <- sapply(colSums(visibles), function(mean) rnorm(1, mean*C, sqrt(C)))
+  names(res) <- paste0("V", 1:ncol(visibles))
+  return(res)
+}
+
+sample_interaction <- function(hiddens, visibles, C) {
+  #hiddens is an NxH matrix of hidden node values
+  #visibles is an NxV matrix of visible node values
+  #C is the variance constant
+  H <- ncol(hiddens)
+  V <- ncol(visibles)
+  
+  stopifnot(nrow(hiddens) == nrow(visibles)) #something wrong with the data structure if these don't match
+  
+  res <- matrix(NA, nrow = H, ncol = V)
+  for(i in 1:V) {
+    for(j in 1:H) {
+      res[i, j] <- rnorm(1, sum(hiddens[, j]*visibles[, i])*C, sqrt(C))
+    }
+  }
+  
+  rownames(res) <- paste0("V", 1:V)
+  colnames(res) <- paste0("H", 1:H)
+  
+  return(res)
+}
+
+sample_main_mh <- function(theta, hiddens, visibles, tau, C) {
+  #theta is an H + V + H*V vector of parameter values
+  #hiddens is an NxH matrix of hidden node values
+  #visibles is an NxV matrix of visible node values
+  #tau is a variance parameter 
+  V <- ncol(visibles)
+  H <- ncol(hiddens)
+  
+  theta_star_main <- MASS::mvrnorm(n = 1, mu = theta[1:(V + H)], Sigma = diag(tau, nrow = V + H, ncol = H + V))
+  theta_star <- c(theta_star_main, theta[(V + H + 1):(V + H + V*H)])
+  u <- runif(1)
+  
+  N <- visibles %>% nrow()
+  data <- cbind(visibles, hiddens)
+  for(i in 1:V) {
+    for(j in 1:H) {
+      data <- cbind(data, visibles[i,]*hiddens[j,])
+    }
+  }
+  possibles <- stats(H, V)
+  W <- exp(possibles %*% theta)
[…]

### 2016-12-02 code/4x4 model/test_all_fitted_models.R

@@ -1,0 +1,129 @@
+#libraries --------------------
+library(dplyr)
+library(tidyr)
+library(ggplot2)
+source("sample_functs.R")
+
+#data and params ------------------------
+H <- 4
+V <- 4
+
+load("written/sample_images.Rdata")
+load("written/sample.params.Rdata")
+params_degen <- list(main_hidden = sample.params %>% ungroup() %>% filter(near_hull) %>% select(starts_with("h"), -H) %>% data.matrix(),
+                     main_visible = sample.params %>% ungroup() %>% filter(near_hull) %>% select(starts_with("v"), -V) %>% data.matrix(),
+                     interaction = sample.params %>% ungroup() %>% filter(near_hull) %>% select(starts_with("theta")) %>% data.matrix() %>% matrix(4))
+params_good <- list(main_hidden = sample.params %>% ungroup() %>% filter(!near_hull) %>% select(starts_with("h"), -H) %>% data.matrix(),
+                    main_visible = sample.params %>% ungroup() %>% filter(!near_hull) %>% select(starts_with("v"), -V) %>% data.matrix(),
+                    interaction = sample.params %>% ungroup() %>% filter(!near_hull) %>% select(starts_with("theta")) %>% data.matrix() %>% matrix(4))
+
+#computing actual distributions
+distn_good <- visible_distn(params = params_good)
+distn_degen <- visible_distn(params = params_degen)
+
+
+reshape_sample_distn <- function(model) {
+  sample_distn <- model$distn
+  dim(sample_distn) <- c(dim(sample_distn)[1]*dim(sample_distn)[2], dim(sample_distn)[3])
+  sample_distn %>% data.frame() -> sample_distn
+  names(sample_distn) <- names(distn_degen)
+  
+  sample_distn %>%
+    group_by(image_id) %>%
+    mutate(iter = 1:n()) -> sample_distn
+  
+  return(sample_distn)
+}
+
+## all models
+model_files <- paste0("written/", list.files("written/")[grep("fitted_models_distn_", list.files("written/"))])
+
+load_model <- function(file) {
+  load(file)
+  sample_good <- reshape_sample_distn(models_good)
+  sample_bad <- reshape_sample_distn(models_bad)
+  sample_degen <- reshape_sample_distn(models_degen)
[…]

### 2016-12-02 code/4x4 model/test_all_params.R

@@ -1,0 +1,101 @@
+#libraries -----------------------
+library(dplyr)
+library(tidyr)
+library(ggplot2)
+source("sample_functs.R")
+
+#function -----------------------
+#this function takes the results from the simulation study and rearranges for ease of use
+rearrange_data <- function(res) {
+  
+  data <- data.frame()
+  for(i in 1:nrow(res)) {
+    tmp <- res$samp[[i]]
+    
+    H <- res[i,]$H
+    V <- res[i,]$V
+    
+    if(H == 4 & V == 4) { #only looking at 4x4 model
+      tmp %>% 
+        data.frame() %>% 
+        rowwise() %>% 
+        mutate_(ss_ratio = paste0("(", paste(paste0(names(.)[(H + V + 1):(H + V + H*V)], "^2"), collapse = " + "), ")/sum(", paste(paste0(names(.)[1:(H+V)],"^2"), collapse = " + "), ")")) %>%    
+        ungroup() -> ratio
+      
+      data <- rbind(data, inner_join(ratio, res$outside[[i]] %>% ungroup()) %>%
+                      mutate(H = H, V = V, n_param = H + V + H*V))
+    }
+  }
+  return(data)
+}
+
+#do it -------------------------
+#uses the function to rearrange the data
+n <- 100
+r_center <- seq(0.2, 3, by = 0.2)
+r_exp <- .05
+
+plot.dat <- list()
+for(i in seq(0, 1, by = 0.1)) {
+  load(paste0("../expected value/apps/results_", i, ".RData")) 
+  plot_dat <- lapply(res, rearrange_data)
+  names(plot_dat) <- gsub("[.]", "_", r_center)
+  
+  do.call(rbind, plot_dat) %>% 
+    mutate(name = rownames(do.call(rbind, plot_dat))) %>%
+    separate(name, into = c("multiplier", "num"), "[.]") %>% 
+    mutate(multiplier = as.numeric(gsub("_", ".", multiplier))) %>%
+    mutate(n_param_f = factor(n_param)) %>%
+    select(-starts_with("exp")) -> plot_dat
+  
+  plot.dat[[i*10 + 1]] <- plot_dat
+}
+
+names(plot.dat) <- gsub("[.]", "_", seq(0, 1, by = 0.1))
+do.call(rbind, plot.dat) %>%
+  mutate(name2 = rownames(do.call(rbind, plot.dat))) %>%
[…]

### 2016-12-02 code/4x4 model/test_compare_mcmc.R

@@ -1,0 +1,421 @@
+#libraries --------------------
+library(dplyr)
+library(tidyr)
+library(ggplot2)
+source("sample_functs.R")
+
+#data and params ------------------------
+H <- 4
+V <- 4
+
+#load truncatednormal prior
+load("written/fitted_models_adaptive_mh_trunc_distn_shrunk_0.6667.Rdata")
+shrunk_bad <- models_bad
+shrunk_good <- models_good
+shrunk_degen <- models_degen
+
+load("written/fitted_models_adaptive_mh_trunc_distn_marginal_1.Rdata")
+marginal_bad_1 <- models_bad
+marginal_good_1 <- models_good
+marginal_degen_1 <- models_degen
+
+
+load("written/fitted_models_adaptive_mh_trunc_distn_marginal_2.Rdata")
+marginal_bad_2 <- models_bad
+marginal_good_2 <- models_good
+marginal_degen_2 <- models_degen
+
+load("written/fitted_models_adaptive_mh_trunc_distn_marginal_3.Rdata")
+marginal_bad_3 <- models_bad
+marginal_good_3 <- models_good
+marginal_degen_3 <- models_degen
+
+#load Jing's prior
+load("written/fitted_models_distn_0.26.Rdata")
+jing_bad <- models_bad
+jing_good <- models_good
+jing_degen <- models_degen
+
+load("written/fitted_models_adaptive_mh_trunc_distn_shrunk_jing_match.Rdata")
+shrunk_jing_bad <- models_bad
+shrunk_jing_good <- models_good
+shrunk_jing_degen <- models_degen
+
+#rm unneccesary data
+rm(models_bad)
+rm(models_degen)
+rm(models_good)
+
+load("written/sample.params.Rdata")
+params_degen <- list(main_hidden = sample.params %>% ungroup() %>% filter(near_hull) %>% select(starts_with("h"), -H) %>% data.matrix(),
+                     main_visible = sample.params %>% ungroup() %>% filter(near_hull) %>% select(starts_with("v"), -V) %>% data.matrix(),
+                     interaction = sample.params %>% ungroup() %>% filter(near_hull) %>% select(starts_with("theta")) %>% data.matrix() %>% matrix(4))
+params_good <- list(main_hidden = sample.params %>% ungroup() %>% filter(!near_hull) %>% select(starts_with("h"), -H) %>% data.matrix(),
[…]

### 2016-12-02 code/4x4 model/test_convergence.R

@@ -1,0 +1,56 @@
+library(dplyr)
+library(ggplot2)
+library(tidyr)
+
+load("written/fitted_models_trunc_marginal_full.Rdata")
+load("written/params_theta.Rdata")
+theta_good <- sample.params %>% 
+  ungroup() %>% 
+  filter(!near_hull) %>% 
+  select(starts_with("v"), starts_with("h"), starts_with("theta"), -H, -V) %>% 
+  data.matrix()
+
+
+if("params" %in% names(models_good)) {
+  theta_est <- data.frame(models_good$params$main_visible, models_good$params$main_hidden, models_good$params$interaction)
+} else if("theta" %in% names(models_good)) {
+  theta_est <- models_good$theta %>% data.frame()
+}
+names(theta_est) <- colnames(theta_good)
+
+
+theta_est %>%
+  mutate(iter = 1:n()) %>% 
+  gather(variable, value, -iter) %>% 
+  ggplot() + 
+  geom_abline(aes(intercept = true_value, slope = 0), data = theta_good %>% data.frame %>% gather(variable, true_value), colour = "red") +
+  geom_line(aes(iter, value)) + 
+  facet_wrap(~variable, scales = "free")
+
+theta_est %>%
+  gather(variable, value) %>% 
+  ggplot() + 
+  geom_histogram(aes(value)) + 
+  facet_wrap(~variable, scales = "free")
+
+var <- models_good$var %>% data.frame
+names(var) <- colnames(theta_good)
+
+var %>%
+  mutate(iter = 1:n()) %>% 
+  gather(variable, value, -iter) %>% 
+  ggplot() + 
+  #geom_abline(aes(intercept = true_value, slope = 0), data = theta_good %>% data.frame %>% gather(variable, true_value), colour = "red") +
+  geom_line(aes(iter, value)) + 
+  facet_wrap(~variable, scales = "free")
+
+
+theta_est %>%
+  mutate(iter = 1:n()) %>% 
+  mutate_(ss_main = paste(paste0(names(theta_est)[1:8], "^2"), collapse=" + "),
+          ss_inter = paste(paste0(names(theta_est)[9:24], "^2"), collapse=" + ")) %>%
+  gather(type, ss, ss_main, ss_inter) %>%
+  ggplot() +
+  geom_line(aes(iter, ss, colour = type))
+
+

### 2016-12-02 code/4x4 model/test_fitted_model.R

@@ -1,0 +1,160 @@
+#libraries --------------------
+library(dplyr)
+library(tidyr)
+library(ggplot2)
+source("functs_sample.R")
+
+#data and params ------------------------
+H <- 4
+V <- 4
+
+load("written/sample_images.Rdata")
+# load("written/fitted_models_trunc_marginal_thin.Rdata")
+# load("written/fitted_models_trunc_thin.Rdata")
+load("written/fitted_models_jing_5.8.Rdata")
+load("written/params_theta.Rdata")
+params_degen <- list(main_hidden = sample.params %>% ungroup() %>% filter(near_hull) %>% select(starts_with("h"), -H) %>% data.matrix(),
+                     main_visible = sample.params %>% ungroup() %>% filter(near_hull) %>% select(starts_with("v"), -V) %>% data.matrix(),
+                     interaction = sample.params %>% ungroup() %>% filter(near_hull) %>% select(starts_with("theta")) %>% data.matrix() %>% matrix(4))
+params_good <- list(main_hidden = sample.params %>% ungroup() %>% filter(!near_hull) %>% select(starts_with("h"), -H) %>% data.matrix(),
+                    main_visible = sample.params %>% ungroup() %>% filter(!near_hull) %>% select(starts_with("v"), -V) %>% data.matrix(),
+                    interaction = sample.params %>% ungroup() %>% filter(!near_hull) %>% select(starts_with("theta")) %>% data.matrix() %>% matrix(4))
+
+
+##plots -----------------------
+
+#computing actual distributions
+distn_good <- visible_distn(params = params_good)
+distn_degen <- visible_distn(params = params_degen)
+
+# get_props <- function(model) {
+#   props <- apply(model$post_pred, 1, function(x) {
+#     x %>% 
+#       data.frame %>%
+#       rename(v1 = X1, v2 = X2, v3 = X3, v4 = X4) %>%
+#       right_join(distn_good) %>% 
+#       group_by(image_id) %>%
+#       summarise(prop = n()/nrow(x))
+#   })
+#   
+#   props %>% 
+#     bind_rows(.id = "iteration") %>% 
+#     mutate(image_id = as.numeric(image_id), iteration = as.numeric(iteration)) %>%
+#     data.frame
+# }
+# 
[…]

### 2016-12-02 code/4x4 model/test_mixing.R

@@ -1,0 +1,165 @@
+library(dplyr)
+library(tidyr)
+
+load("written/sample_images.Rdata")
+
+#data and params ------------------------
+H <- 4
+V <- 4
+
+load("written/fitted_models_trunc_marginal_full.Rdata")
+marginal <- models_good
+
+load("written/fitted_models_trunc_full.Rdata")
+trunc <- models_good
+
+rm(models_good)
+rm(models_bad)
+
+source("functs_sample.R")
+load("written/params_theta.Rdata")
+
+params_good <- list(main_hidden = sample.params %>% ungroup() %>% filter(!near_hull) %>% select(starts_with("h"), -H) %>% data.matrix(),
+                    main_visible = sample.params %>% ungroup() %>% filter(!near_hull) %>% select(starts_with("v"), -V) %>% data.matrix(),
+                    interaction = sample.params %>% ungroup() %>% filter(!near_hull) %>% select(starts_with("theta")) %>% data.matrix() %>% matrix(4))
+distn_good <- visible_distn(params = params_good)
+
+
+reshape_sample_distn <- function(model) {
+  sample_distn <- model$distn
+  dim(sample_distn) <- c(dim(sample_distn)[1]*dim(sample_distn)[2], dim(sample_distn)[3])
+  sample_distn %>% data.frame() -> sample_distn
+  names(sample_distn) <- names(distn_good)
+  
+  sample_distn %>%
+    group_by(image_id) %>%
+    mutate(iter = 1:n()) -> sample_distn
+  
+  return(sample_distn)
+}
+
+marginal_sample <- reshape_sample_distn(marginal)
+trunc_sample <- reshape_sample_distn(trunc)
+
+
+## acf -------------------------------
+marginal_sample %>% 
+  group_by(v1, v2, v3, v4, image_id) %>% 
+  do(data.frame(acf = acf(.$prob, plot=FALSE)$acf,
+     lag = acf(.$prob, plot=FALSE)$lag)) -> marginal_acfs
+
+trunc_sample %>% 
+  group_by(v1, v2, v3, v4, image_id) %>% 
+  do(data.frame(acf = acf(.$prob, plot=FALSE)$acf,
+                lag = acf(.$prob, plot=FALSE)$lag)) -> trunc_acfs
+
[…]

### 2016-12-02 code/4x4 model/test_ml_2.R

@@ -1,0 +1,55 @@
+#libraries --------------------
+library(dplyr)
+library(tidyr)
+library(ggplot2)
+source("sample_functs.R")
+
+#data and params ------------------------
+load("written/sample.params.Rdata")
+params_good <- list(main_hidden = sample.params %>% ungroup() %>% filter(!near_hull) %>% select(starts_with("h"), -H) %>% data.matrix(),
+                    main_visible = sample.params %>% ungroup() %>% filter(!near_hull) %>% select(starts_with("v"), -V) %>% data.matrix(),
+                    interaction = sample.params %>% ungroup() %>% filter(!near_hull) %>% select(starts_with("theta")) %>% data.matrix() %>% matrix(4))
+
+distn_good <- visible_distn(params = params_good)
+possibles <- stats(4, 4)
+
+load("written/fitted_ml_margin_zeros.Rdata")
+theta.df <- theta %>% data.frame()
+colnames(theta.df) <- colnames(possibles)
+
+theta.df %>%
+  rowwise() %>%
+  do(params = list(main_hidden = data.frame(.) %>% select(starts_with("h")) %>% data.matrix(),
+                    main_visible = data.frame(.) %>% select(starts_with("v")) %>% data.matrix(),
+                    interaction = data.frame(.) %>% select(starts_with("theta")) %>% data.matrix() %>% matrix(4))
+  ) %>%
+  ungroup() %>%
+  mutate(iter = 1:n()) %>%
+  group_by(iter) %>%
+  do(data.frame(visible_distn(.$params[[1]]))) -> distn_ml
+ 
+distn_good %>% rename(true = prob) %>%
+  left_join(distn_ml %>% rename(est = prob) %>% ungroup %>% select(-image_id)) %>% 
+  #gather(variable, value, true, est) %>%
+  ggplot() +
+  geom_line(aes(iter, est), colour = 'blue') +
+  geom_line(aes(iter, true)) +
+  facet_wrap(~image_id)
+
+#likelihood-----
+len <- 20
+
+expand.grid(V1 = seq(-10, 10, length.out = len), V2 = seq(-10, 10, length.out = len)) %>%
+  cbind(matrix(rep(rnorm(H + V + H*V - 2, 0, 10), len*len), nrow = len*len, byrow = TRUE)) %>%
+  data.frame() %>% 
+  group_by(V1, V2) %>%
[…]

### 2016-12-02 code/4x4 model/test_sample_images.R

@@ -1,0 +1,78 @@
+#libraries --------------------
+library(dplyr)
+library(tidyr)
+library(ggplot2)
+source("sample_functs.R")
+
+#data and params ------------------------
+H <- 4
+V <- 4
+N <- 1
+
+load("written/sample_images.Rdata")
+load("written/sample.params.Rdata")
+params_degen <- list(main_hidden = sample.params %>% ungroup() %>% filter(near_hull) %>% select(starts_with("h"), -H) %>% data.matrix(),
+                     main_visible = sample.params %>% ungroup() %>% filter(near_hull) %>% select(starts_with("v"), -V) %>% data.matrix(),
+                     interaction = sample.params %>% ungroup() %>% filter(near_hull) %>% select(starts_with("theta")) %>% data.matrix() %>% matrix(4))
+params_good <- list(main_hidden = sample.params %>% ungroup() %>% filter(!near_hull) %>% select(starts_with("h"), -H) %>% data.matrix(),
+                    main_visible = sample.params %>% ungroup() %>% filter(!near_hull) %>% select(starts_with("v"), -V) %>% data.matrix(),
+                    interaction = sample.params %>% ungroup() %>% filter(!near_hull) %>% select(starts_with("theta")) %>% data.matrix() %>% matrix(4))
+
+
+cdf_visible <- function(main_visible, interaction, hiddens, visibles) {
+  #main_visible is a 1xH matrix of visible main effects param values
+  #interaction is a HxV matrix of interaction effects param values
+  #hiddens is an NxH matrix of hidden node values
+  stopifnot(nrow(hiddens) == nrow(visibles))
+  
+  N <- nrow(hiddens)
+
+  a <- exp(2*(matrix(main_visible, nrow = nrow(hiddens), ncol = length(main_visible), byrow = TRUE) + tcrossprod(hiddens, interaction)))
+  f <- apply(a, 1:2, function(b) dbinom(1, 1, b/(1 + b)))
+  A <- matrix(runif(N), nrow = N, ncol = ncol(visibles))
+  
+  res <- A*f*(visibles == -1) + (f + A*(1 - f))*(visibles == 1)
+
+  return(res)
+}
+
+pit_test_degen <- cdf_visible(params_degen$main_visible, params_degen$interaction, flat_images_degen$hiddens, flat_images_degen$visibles)
[…]

### 2016-12-02 code/4x4 model/tune_gibbs.R

@@ -1,0 +1,67 @@
+#libraries --------------------
+library(dplyr)
+library(tidyr)
+library(ggplot2)
+source("functs_sample.R")
+
+#params ------------------------
+H <- 4
+V <- 4
+
+load("written/sample_images.Rdata")
+load("written/params_theta.Rdata")
+params_degen <- list(main_hidden = sample.params %>% ungroup() %>% filter(near_hull) %>% select(starts_with("h"), -H) %>% data.matrix(),
+                     main_visible = sample.params %>% ungroup() %>% filter(near_hull) %>% select(starts_with("v"), -V) %>% data.matrix(),
+                     interaction = sample.params %>% ungroup() %>% filter(near_hull) %>% select(starts_with("theta")) %>% data.matrix() %>% matrix(4))
+params_good <- list(main_hidden = sample.params %>% ungroup() %>% filter(!near_hull) %>% select(starts_with("h"), -H) %>% data.matrix(),
+                    main_visible = sample.params %>% ungroup() %>% filter(!near_hull) %>% select(starts_with("v"), -V) %>% data.matrix(),
+                    interaction = sample.params %>% ungroup() %>% filter(!near_hull) %>% select(starts_with("theta")) %>% data.matrix() %>% matrix(4))
+
+#computing actual distributions --------------
+distn_good <- visible_distn(params = params_good)
+distn_degen <- visible_distn(params = params_degen)
+
+#data ----------------------
+consts <- c(0.1, 0.2, 0.3, 0.4, 0.5, 1, 2, 3, 4, 5, 5.5, 5.6, 5.7, 5.8, 5.9, 6, 6.1, 6.2, 7, 8)
+reshape_sample_distn <- function(model) {
+  sample_distn <- model$distn
+  dim(sample_distn) <- c(dim(sample_distn)[1]*dim(sample_distn)[2], dim(sample_distn)[3])
+  sample_distn %>% data.frame() -> sample_distn
+  names(sample_distn) <- names(distn_good)
+  
+  sample_distn %>%
+    group_by(image_id) %>%
+    mutate(iter = 1:n()) -> sample_distn
+  
+  return(sample_distn)
+}
+
+res <- data.frame()
+for(const in consts) {
+  load(paste0("written/fitted_models_jing_", const, ".Rdata"))
+  sample_good <- reshape_sample_distn(models_good)
+  sample_bad <- reshape_sample_distn(models_bad)
+  
[…]

### 2016-12-02 code/degeneracy/apps/functions.R

@@ -1,0 +1,132 @@
+stats <- function(H, V, type="negative") {
+  if(type == "binary") t <- data.frame(c(0,1))[,rep(1, H + V)]
+  else if(type == "negative") t <- data.frame(c(-1,1))[,rep(1, H + V)]
+  else stop("Type must be 'binary', 'negative'")
+  
+  names(t) <- c(paste0("h", 1:H), paste0("v", 1:V))  
+  
+  #all possibilities of statistics in (H + V)-space
+  t.grid <- expand.grid(t)
+  
+  for(i in 1:H){
+    for(j in (H+1):(H+V)){
+      t.grid <- cbind(t.grid, t.grid[,i]*t.grid[,j])
+      names(t.grid)[ncol(t.grid)] <- paste0("theta", i, j - H)
+    }
+  }
+  
+  return(data.matrix(t.grid))
+}
+
+expected_value <- function(theta, stats, normalized = TRUE) {
+  result <- crossprod(t(crossprod(stats, exp(crossprod(t(stats), theta)))), diag(1/apply(exp(crossprod(t(stats), theta)), 2, sum), nrow=length(apply(exp(crossprod(t(stats), theta)), 2, sum))))
+  rownames(result) <- paste0("exp_", rownames(result))
+  return(result)
+}
+
+sample_sphere <- function(stat, n, r = 1) {
+  require(dplyr)
+  require(tidyr)
+  
+  samp <- function(space, r) {
+    vec <- rnorm(space)
+    paste0(r*vec/sqrt(sum(vec^2)), collapse=",")
+  }
+  
+  res <- data.frame(samp = 1:n) %>%
+    group_by(samp) %>%
+    mutate(vals = samp(ncol(stat), r)) %>%
+    separate(col = vals, sep = ",", into = colnames(stat)) %>%
+    ungroup() %>%
+    select(-samp) %>%
+    data.matrix() %>%
+    t()
+    
+  
+  return(res)  
+}
+
+calc_hull <- function(stats, options="FA") {
+  #calculate convex hull
+  require(geometry)
+  C <- convhulln(stats, options=options)
+  return(C)
+}
+
+generate_hull_from_stats <- function(stats) {
+  ## stat is the df of possible statistics as generated from stats
+  lines <- NULL
+  for(i in 1:nrow(stats)){
+    for(j in 1:nrow(stats)){
+      if(i < j) {
[…]

### 2016-12-02 code/degeneracy/apps/www/js/plots.js

@@ -1,0 +1,183 @@
+var outputBinding = new Shiny.OutputBinding();
+
+$.extend(outputBinding, {
+  find: function(scope) {
+    return $(scope).find('.d3graph');
+  },
+  renderValue: function(el, data) {  
+    wrapper(el, data);
+  },
+});
+Shiny.outputBindings.register(outputBinding);
+
+
+function wrapper(el, data) {  
+  var w = 500,
+      h = 200,
+      yaw = 0.5,
+      pitch = 0.5, 
+      drag = false;
+  
+  var plot_dat = null,
+      surf_plot = null;
+
+  if(data != null) {
+    plot_dat = data.values;
+    surf_plot = data.surf_plot;
+  }
+
+  
+  d3.select(el).selectAll(".plotHolder").remove();
+
+  if(surf_plot) {
+  var selection = d3.select(el).selectAll("svg")
+      .data(d3.keys(plot_dat));
+  
+  var enterSelection = selection.enter()
+    .append("div")
+    .attr("class", "plotHolder");
+      
+  var h3 = enterSelection
+    .append("h5")
+    .text(function(d) {return d;});
+    
+  var svg = enterSelection  
+    .append("svg")
+    .attr("width", w)
+    .attr("height", h);
+    
+  selection.exit().remove();
+  
+  var group = svg.append("g");
+  
+  var md = group
+    .data(d3.values(plot_dat));
+
+    var surf = [] 
+    md.each(function(d) {  
+      surf.push(
+        d3.select(this)
+          .surface3D(w, h)
+          .surfaceHeight(function(d){ 
+              return d;
+          })
+          .surfaceColor(function(d){
+              var c = d3.hsl((10*d + 100), 0.6, 0.5).rgb();
+              return "rgb("+parseInt(c.r)+","+parseInt(c.g)+","+parseInt(c.b)+")";
+          })
+        );
+    });
+  
+    d3.select(el).on("mousedown",function() {
+      drag = [d3.mouse(this), yaw, pitch];
+    })
+    .on("mouseup",function() {
+      drag = false;
+    })
+    .on("mousemove",function() {
+      if(drag){            
+        var mouse = d3.mouse(this);
+        yaw = drag[1] - (mouse[0] - drag[0][0])/50;
+        pitch = drag[2] + (mouse[1] - drag[0][1])/50;
+        pitch = Math.max(-Math.PI/2, Math.min(Math.PI/2, pitch));
+  
+        for(i = 0; i < surf.length; i++) {
[…]

### 2016-12-02 code/degeneracy/functions.R

@@ -1,0 +1,131 @@
+stats <- function(H, V, type="negative") {
+  if(type == "binary") t <- data.frame(c(0,1))[,rep(1, H + V)]
+  else if(type == "negative") t <- data.frame(c(-1,1))[,rep(1, H + V)]
+  else stop("Type must be 'binary', 'negative'")
+  
+  names(t) <- c(paste0("h", 1:H), paste0("v", 1:V))  
+  
+  #all possibilities of statistics in (H + V)-space
+  t.grid <- expand.grid(t)
+  
+  for(i in 1:H){
+    for(j in (H+1):(H+V)){
+      t.grid <- cbind(t.grid, t.grid[,i]*t.grid[,j])
+      names(t.grid)[ncol(t.grid)] <- paste0("theta", i, j - H)
+    }
+  }
+  
+  return(data.matrix(t.grid))
+}
+
+expected_value <- function(theta, stats, normalized = TRUE) {
+  result <- crossprod(t(crossprod(stats, exp(crossprod(t(stats), theta)))), diag(1/apply(exp(crossprod(t(stats), theta)), 2, sum), nrow=length(apply(exp(crossprod(t(stats), theta)), 2, sum))))
+  rownames(result) <- paste0("exp_", rownames(result))
+  return(result)
+}
+
+sample_sphere <- function(stat, n, r = 1) {
+  require(dplyr)
+  require(tidyr)
+  
+  samp <- function(space, r) {
+    vec <- rnorm(space)
+    paste0(r*vec/sqrt(sum(vec^2)), collapse=",")
+  }
+  
+  res <- data.frame(samp = 1:n) %>%
+    group_by(samp) %>%
+    mutate(vals = samp(ncol(stat), r)) %>%
+    separate(col = vals, sep = ",", into = colnames(stat)) %>%
+    select(-samp) %>%
+    data.matrix() %>%
+    t()
+    
+  
+  return(res)  
+}
+
+calc_hull <- function(stats, options="FA") {
+  #calculate convex hull
+  require(geometry)
+  C <- convhulln(stats, options=options)
+  return(C)
+}
+
+generate_hull_from_stats <- function(stats) {
+  ## stat is the df of possible statistics as generated from stats
+  lines <- NULL
+  for(i in 1:nrow(stats)){
+    for(j in 1:nrow(stats)){
+      if(i < j) {
+        lines <- c(lines, list(list(x1=stats[i,], x2=stats[j,])))
[…]

### 2016-12-02 code/degeneracy/grid_sample.R

@@ -1,0 +1,90 @@
+#libraries -----------------------
+library(dplyr)
+library(tidyr)
+library(ggplot2)
+source("functions.R")
+
+## main function --------------------------------------------
+find_prop <- function(g_theta, stat, r_exp) {
+  g_theta %>%
+    data.frame() %>%
+    group_by_(.dots = c(colnames(g_theta))) %>%
+    do(samp = sample_points_outside(as.numeric(.), stat, r_exp)$in_hull) %>%
+    group_by_(.dots = c(colnames(g_theta))) %>%
+    do(as.data.frame(sum(.$samp[[1]]) > 0)) -> is_outside # count of not in hull greater than zero => there exists a point outside hull for this model
+
+  is_outside %>%
+    rename_(near_hull = as.name(names(is_outside)[ncol(is_outside)])) %>%
+    mutate(r = r_exp) -> inside_outside
+
+  return(inside_outside)
+}
+
+## additional useful functions -------------------------------------
+sample_points_outside <- function(g_theta, stats, r, n = 100) {
+  #g_theta is the expected value mapping of a point theta in R^H+V+H*V
+  x0 <- data.frame(samp = 1:n) %>%
+    mutate(vals = samp_exp(length(g_theta), g_theta, r)) %>%
+    separate(col = vals, sep = ",", into = colnames(stats)) %>%
+    select(-samp)
+
+  x0$in_hull <- !apply(x0, 1, in_hull, stats)
+
+  return(x0)
+}
+
+samp_exp <- function(space, center, r) {
+  vec <- rnorm(space)
+  paste0(r*vec/sqrt(sum(vec^2)) + center, collapse=",")
+}
+
+sample_sphere_unif <- function(space, rep, center, r) {
+  vec <- matrix(rnorm(space*rep), nrow = rep)
+  r*vec/t(matrix(rep(sqrt(rowSums(vec^2)), each = space), nrow = space))
+}
+
+in_hull <- function(point, hull_points) {
+  require(lpSolveAPI)
+  P <- data.matrix(point)
+  A <- t(data.matrix(hull_points))
+
+  lp_obj <- make.lp(nrow = 0, ncol = ncol(A))
[…]

### 2016-12-02 code/degeneracy/kaiser_method.R

@@ -1,0 +1,106 @@
+library(ggplot2)
+library(dplyr)
+library(tidyr)
+source("functions.R")
+
+model_data <- function(res) {
+
+  model.data <- list()
+  for(i in 1:nrow(res)) {
+    tmp <- res$samp[[i]]
+    H <- res[i,]$H
+    V <- res[i,]$V
+
+    tmp %>%
+      data.frame() %>%
+      rowwise() %>%
+      mutate_(ss_ratio = paste0("(", paste(paste0(names(.)[(H + V + 1):(H + V + H*V)], "^2"), collapse = " + "), ")/sum(", paste(paste0(names(.)[1:(H+V)],"^2"), collapse = " + "), ")")) %>%
+      ungroup() -> ratio
+
+    inner_join(ratio, res$outside[[i]] %>% ungroup()) %>%
+      mutate(H = H, V = V, n_param = H + V + H*V) -> model.data[[i]]
+  }
+  names(model.data) <- res$n_param
+  return(model.data)
+}
+
+n <- 100
+r_center <- seq(0.2, 3, by = 0.2)
+r_exp <- .05
+
+set.seed(10-22-1985) #best day ever!
+
+model.dat <- list()
+for(i in seq(0, 1, by = 0.1)) {
+  load(paste0("written/results_", i, ".RData"))
+  model_dat <- lapply(res, model_data)
+  names(model_dat) <- r_center
+
+  model.dat[[i*10 + 1]] <- model_dat
+}
+names(model.dat) <- seq(0, 1, by = 0.1)
+
+expand.grid(power = as.numeric(names(model.dat)), multiplier = as.numeric(names(model.dat[[1]])), n_param = as.numeric(names(model.dat[[1]][[1]]))) %>%
+  cbind(expand.grid(i = 1:length(model.dat), j = 1:length(model.dat[[1]]), k = 1:length(model.dat[[1]][[1]]))) %>%
+  group_by(power, multiplier, n_param) %>%
+  do(models = model.dat[[.$i]][[.$j]][[.$k]]) -> frame
+
+frame %>%
+  group_by(power, multiplier, n_param) %>%
+  mutate(frac_degen = sum(models[[1]]$near_hull)/nrow(models[[1]]),
+         radius = (models[[1]]$H[1] + models[[1]]$V[1])^power,
+         happy = frac_degen < .05) %>%
+  group_by(n_param, happy) %>%
+  sample_n(20) -> models.samp
+
+
[…]

### 2016-12-02 code/degeneracy/kaiser_method_grid.R

@@ -1,0 +1,48 @@
+#libraries -----------------------
+library(dplyr)
+library(tidyr)
+library(ggplot2)
+source("functions.R")
+
+#load sample params ---------------------------------------------
+n <- 100
+load("written/results_grid.RData")
+
+indep_params <- function(samp, H, V) {
+  samp[, (H + V + 1):ncol(samp)] <- 0
+  samp
+}
+
+res %>%
+  group_by(H, V, N, r1, r2) %>%
+  do(indep_exp = t(expected_value(t(indep_params(.$samp[[1]], .$H, .$V)), .$stat[[1]]))[, -((.$H + .$V + 1):(.$H + .$V + .$H*.$V))],
+     marg_exp = .$g_theta[[1]][, (.$H + .$V + .$H*.$V + 1):(ncol(.$g_theta[[1]])-.$H*.$V)]) -> exp_vals
+
+exp_vals %>%
+  group_by(H, V, N, r1, r2) %>%
+  do(data.frame(mean_abs_diff = apply(abs(.$indep_exp[[1]] - .$marg_exp[[1]]), 1, mean))) %>%
+  group_by(H, V, N, r1, r2) %>%
+  summarise(mean_abs_diff = mean(mean_abs_diff)) %>%
+  mutate(Hiddens = paste0("Hiddens: ", H), Visibles = paste0("Visibles: ", V)) %>%
+  ggplot() +
+  geom_tile(aes(x = r1, y = r2, fill = mean_abs_diff)) +
+  geom_contour(aes(x = r1, y = r2, z = mean_abs_diff), colour = "black", bins = 8) +
+  #geom_contour(aes(x = r1, y = r2, z = mean_abs_diff), colour = "black", breaks = .1, size = 1.5) +
+  #geom_abline(aes(intercept = 0, slope = 1), alpha = .5, lty = 2) +
+  scale_fill_gradient("Mean absolute difference in expectation", low = "yellow", high = "red") +
+  facet_grid(Visibles~Hiddens) +
+  xlab(expression(group("||", theta[main], "||"))) +
+  ylab(expression(group("||", theta[interaction], "||"))) +
+  theme(aspect.ratio = 1)
+
+exp_vals %>%
+  mutate(dist = sqrt(r1^2 + r2^2)) %>%
+  group_by(dist, H, V) %>%
+  do(data.frame(mean_abs_diff = mean(abs(.$indep_exp[[1]] - .$marg_exp[[1]])))) %>%
[…]

### 2016-12-02 code/degeneracy/min_radius.R

@@ -1,0 +1,84 @@
+source("functions.R")
+library(dplyr)
+library(reshape2)
+library(ggplot2)
+
+## constants
+n <- 500 #num points to sample on the sphere
+cutoff <- .01
+step <- 1
+backstep <- .01
+
+test_cases <- expand.grid(data.frame(1:4)[,rep(1,2)]) %>% 
+  rename(H = X1.4, V = X1.4.1) %>%
+  filter(H <= V) %>% #remove those cases with less visibles than hiddens. Is this necessary?
+  mutate(n_param = H*V + H + V) %>%
+  mutate(max_facets = (2^(H+V))^(floor(n_param/2))) %>%
+  filter(n_param <= 11) %>% #can't handle any higher dimensions currently
+  rowwise() %>%
+  mutate(min_radius(H, V, n=n, cutoff=cutoff, backstep=backstep, step=step))
+
+min_radius <- function(H, V, type='negative', n, cutoff, backstep, step) {
+  stat <- stats(H, V, type)
+  
+  h <- calc_hull(stat, options="Qt")
+  triangles <- h %>%
+    data.frame() %>%
+    rowwise() %>%
+    do(data.frame(stat[unlist(.),])) %>%
+    ungroup() %>%
+    transmute_(point=paste0("paste(",paste(colnames(stat), collapse=","),", sep=',')")) %>%
+    mutate(triangle=rep(1:(n()/3), each = 3)) %>%
+    mutate(point_num=paste0("point", rep(1:3, (n()/3)))) %>%
+    dcast(triangle ~ point_num, value.var = "point")
+  
+  
+  ## inits
+  r0 <- 0
+  theta0 <- sample_sphere(stat, n, r0)
+  exp_vals0 <- expected_value(stats = stat, theta = theta0, normalized = TRUE)
+  d0 <- exp_vals0 %>% t() %>%
+    data.frame() %>%
+    rowwise() %>% 
+    transmute_(P=paste0("paste(",paste(rownames(exp_vals0), collapse=","),", sep=',')")) %>%
+    merge(triangles) %>%    
+    select(-triangle) %>%
+    ungroup() %>%
+    apply(1, function(x) distance_to_simplex(P=x["P"], simplex=as.list(x[c("point1", "point2", "point3")]))) %>%
+    min()
+ 
+
+
+  success <- FALSE
+  while(!success){
+
+    r1 <- r0 + step
[…]

### 2016-12-02 code/degeneracy/min_radius_lp.R

@@ -1,0 +1,81 @@
+source("functions.R")
+library(dplyr)
+library(tidyr)
+library(lpSolveAPI)
+
+## constants
+n <- 500 #num points to sample on the sphere
+cutoff <- .05
+step <- .5
+backstep <- .01
+v_step <- .01
+
+test_cases <- expand.grid(data.frame(1:4)[,rep(1,2)]) %>% 
+  rename(H = X1.4, V = X1.4.1) %>%
+  filter(H <= V) %>% #remove those cases with less visibles than hiddens. Is this necessary?
+  mutate(n_param = H*V + H + V) %>%
+  #filter(n_param <= 11) %>%
+  rowwise() %>%
+  mutate(min_rad = min_radius_lp(H, V, n=n, cutoff=cutoff, backstep=backstep, step=step, v_step=v_step)) %>%
+  mutate(volume = pi^(n_param/2)/gamma(n_param/2 + 1)*min_rad^n_param)
+
+save(test_cases, file="min_radius_lp.Rdata")
+
+
+min_radius_lp <- function(H, V, type = "negative", n, cutoff, backstep, step, v_step) {
+  stat <- stats(H, V, type)
+  r0 <- 0
+  #theta0 <- sample_sphere(stat, n, r0)
+  #exp_vals0 <- expected_value(stats = stat, theta = theta0, normalized = TRUE)
+  d0 <- 1
+  
+  success <- FALSE
+  while(!success){  
+    r1 <- r0 + step
+    theta1 <- sample_sphere(stat, n, r1)
+    exp_vals1 <- expected_value(stats = stat, theta = theta1, normalized = TRUE)
+      
+    outside <- FALSE
+    v <- max(round(1/max(apply(exp_vals1, 2, function(x) sqrt(sum(x^2)))) - .1, 2), 1)
+    while(!outside) {
+      num_outside <- sum(apply(exp_vals1*v, 2, function(y) !in_hull(stat, y)))
+      if(num_outside > 0) {
+        outside <- TRUE      
+      } else {
+        v <- v + v_step
+      }      
+      cat(v, " ")
+    }
+    
+    exp_vals1_v <- exp_vals1*v   
+    d1 <- min(sapply(1:nrow(t(exp_vals1)), function(x) dist(matrix(c(exp_vals1[,x], exp_vals1_v[,x]), nrow=2, byrow = TRUE))))
+    
[…]

### 2016-12-02 code/degeneracy/min_radius_nordman.R

@@ -1,0 +1,95 @@
+source("functions.R")
+library(dplyr)
+library(tidyr)
+
+n <- 100
+r_center <- 2
+
+test_cases <- expand.grid(data.frame(1:4)[,rep(1,2)]) %>% 
+  rename(H = X1.4, V = X1.4.1) %>%
+  filter(H <= V) %>% #remove those cases with less visibles than hiddens. Is this necessary?
+  mutate(n_param = H*V + H + V) %>%
+  mutate(max_facets = (2^(H+V))^(floor(n_param/2))) %>%
+  mutate(n = n, r = r_center) %>%
+  group_by(H, V, n_param, n, r) %>%
+  do(stat = stats(.$H, .$V, "negative")) %>%
+  ungroup() %>%
+  group_by(H, V, n_param, n, r) %>%
+  do(g_theta = t(expected_value(theta = sample_sphere(.$stat[[1]], .$n, .$r), stats = data.matrix(.$stat[[1]]))))
+
+test_cases_stat <- expand.grid(data.frame(1:4)[,rep(1,2)]) %>% 
+  rename(H = X1.4, V = X1.4.1) %>%
+  filter(H <= V) %>% #remove those cases with less visibles than hiddens. Is this necessary?
+  mutate(n_param = H*V + H + V) %>%
+  mutate(max_facets = (2^(H+V))^(floor(n_param/2))) %>%
+  mutate(n = n, r = r_center) %>%
+  group_by(H, V, n_param, n, r) %>%
+  do(stat = stats(.$H, .$V, "negative"))
+
+cases <- inner_join(test_cases, test_cases_stat)
+
+cases %>%
+  group_by(H, V) %>%
+  do(min_dist = find_min_dist(.$g_theta[[1]], .$stat[[1]], .05, .005, .$H + .$V)) %>%
+  mutate(min_dist = as.numeric(min_dist))-> dists
+
+find_min_dist <- function(g_theta, stat, step, backstep, n_param) {
+  r_exp_0 <- 0
+  num_0 <- 0
+  
+  success <- FALSE
+  while(!success) {
+    r_exp_1 <- r_exp_0 + step
+    
+    for(i in 1:nrow(g_theta)) {
+      g_theta[i,] %>%
+        sample_points_outside(stat, r_exp_1) %>%
+        filter(!in_hull) %>%
+        dim() -> num_1
+      
+      if(num_1[1] > 0) break
+    }
+    
+
[…]

### 2016-12-02 code/degeneracy/radius_percent_degen.R

@@ -1,0 +1,105 @@
+## load libraries ---------------------------
+
+source("functions.R")
+library(dplyr)
+library(tidyr)
+library(ggplot2)
+
+## main function --------------------------------------------
+find_prop <- function(g_theta, stat, r_exp) {
+  g_theta %>%
+    data.frame() %>%
+    group_by_(.dots = c(colnames(g_theta))) %>%
+    do(samp = sample_points_outside(as.numeric(.), stat, r_exp)$in_hull) %>%
+    group_by_(.dots = c(colnames(g_theta))) %>%
+    do(as.data.frame(sum(.$samp[[1]]) > 0)) -> is_outside # count of not in hull greater than zero => there exists a point outside hull for this model
+  
+  is_outside %>% 
+    rename_(near_hull = as.name(names(is_outside)[ncol(is_outside)])) %>%
+    mutate(r = r_exp) -> inside_outside
+  
+  return(inside_outside)
+}
+
+## additional useful functions -------------------------------------
+sample_points_outside <- function(g_theta, stats, r, n = 100) {
+  #g_theta is the expected value mapping of a point theta in R^H+V+H*V
+  x0 <- data.frame(samp = 1:n) %>%
+    mutate(vals = samp_exp(length(g_theta), g_theta, r)) %>%
+    separate(col = vals, sep = ",", into = colnames(stats)) %>%
+    select(-samp)
+  
+  x0$in_hull <- !apply(x0, 1, in_hull, stats)
+  
+  return(x0)
+}
+
+samp_exp <- function(space, center, r) {
+  vec <- rnorm(space)
+  paste0(r*vec/sqrt(sum(vec^2)) + center, collapse=",")
+}
+
+in_hull <- function(point, hull_points) {
+  require(lpSolveAPI)
+  P <- data.matrix(point)
+  A <- t(data.matrix(hull_points))
+  
+  lp_obj <- make.lp(nrow = 0, ncol = ncol(A))
+  sapply(1:nrow(A), function(x) add.constraint(lp_obj, A[x,], type = "=", rhs = P[x]))
+  add.constraint(lp_obj, rep(1, ncol(A)), type = "=", rhs = 1)
+  sapply(1:ncol(A), function(x) add.constraint(lp_obj, 1, type = ">=", rhs = 0, indices = x))
+  
+  return(solve(lp_obj) == 0)
+}
+
[…]

### 2016-12-02 code/degeneracy/radius_percent_degen_analyze_res.R

@@ -1,0 +1,70 @@
+
+## analyze results -------------------------------------
+## plot function ----------------------------------
+plot_data <- function(res) {
+  
+  plot.data <- data.frame()
+  for(i in 1:nrow(res)) {
+    tmp <- res$samp[[i]]
+    H <- res[i,]$H
+    V <- res[i,]$V
+    
+    tmp %>% 
+      data.frame() %>% 
+      rowwise() %>% 
+      mutate_(ss_ratio = paste0("(", paste(paste0(names(.)[(H + V + 1):(H + V + H*V)], "^2"), collapse = " + "), ")/sum(", paste(paste0(names(.)[1:(H+V)],"^2"), collapse = " + "), ")")) %>%    
+      ungroup() -> ratio
+    
+    inner_join(ratio, res$outside[[i]] %>% ungroup()) %>%
+      select(ss_ratio, near_hull) %>%
+      mutate(H = H, V = V, n_param = H + V + H*V) %>%
+      rbind(plot.data) -> plot.data
+  }
+  return(plot.data)
+}
+
+make_plot <- function(plot.data) { 
+  plot.data %>%
+    ggplot() + 
+    geom_boxplot(aes(as.factor(n_param), ss_ratio, colour = near_hull)) +
+    xlab("Number of params") +
+    ylab("Ratio of sum of squares of cross terms to main effects") +
+    scale_colour_discrete("Within .05 of hull") -> plot1
+  
+  plot.data %>% 
+    ggplot() + 
+    geom_bar(aes(as.factor(n_param), y = ..count.., fill = near_hull), position = "fill") +
+    xlab("Number of params") +
+    scale_fill_discrete("Within .05 of hull") -> plot2
+  
+  return(list(p1 = plot1, p2 = plot2))
+}
+
+plot.dat <- lapply(res, plot_data)
+plots <- lapply(plot.dat, make_plot)
+
+
+library(gridExtra)
+for (i in 1:length(plots)) {
+  grid.arrange(plots[[i]][[1]], plots[[i]][[2]], main = paste0("Radius multiplier: ", r_center[i]))
+}
+
+
+names(plot.dat) <- gsub("[.]", "_", r_center)
+do.call(rbind, plot.dat) %>%
+  ungroup() %>%
+  mutate(name = rownames(do.call(rbind, plot.dat))) %>%
+  separate(name, into = c("multiplier", "num"), "[.]") %>% 
[…]

### 2016-12-02 code/degeneracy/radius_percent_degen_max_no_degen_plots.R

@@ -1,0 +1,88 @@
+library(ggplot2)
+library(dplyr)
+library(tidyr)
+
+plot_data <- function(res) {
+
+  plot.data <- data.frame()
+  for(i in 1:nrow(res)) {
+    tmp <- res$samp[[i]]
+    H <- res[i,]$H
+    V <- res[i,]$V
+
+    tmp %>%
+      data.frame() %>%
+      rowwise() %>%
+      mutate_(ss_ratio = paste0("(", paste(paste0(names(.)[(H + V + 1):(H + V + H*V)], "^2"), collapse = " + "), ")/sum(", paste(paste0(names(.)[1:(H+V)],"^2"), collapse = " + "), ")")) %>%
+      ungroup() -> ratio
+
+    inner_join(ratio, res$outside[[i]] %>% ungroup()) %>%
+      select(ss_ratio, near_hull) %>%
+      mutate(H = H, V = V, n_param = H + V + H*V) %>%
+      rbind(plot.data) -> plot.data
+  }
+  return(plot.data)
+}
+
+n <- 100
+r_center <- seq(0.2, 3, by = 0.2)
+r_exp <- .05
+
+
+plot.dat <- list()
+for(i in seq(0, 1, by = 0.1)) {
+  load(paste0("written/results_", i, ".RData"))
+  plot_dat <- lapply(res, plot_data)
+  names(plot_dat) <- gsub("[.]", "_", r_center)
+
+  do.call(rbind, plot_dat) %>%
+    mutate(name = rownames(do.call(rbind, plot_dat))) %>%
+    separate(name, into = c("multiplier", "num"), "[.]") %>%
+    mutate(multiplier = as.numeric(gsub("_", ".", multiplier))) %>%
+    mutate(n_param_f = factor(n_param)) -> plot_dat
+
+  plot.dat[[i*10 + 1]] <- plot_dat
+}
+
+names(plot.dat) <- gsub("[.]", "_", seq(0, 1, by = 0.1))
+do.call(rbind, plot.dat) %>%
+  mutate(name2 = rownames(do.call(rbind, plot.dat))) %>%
+  separate(name2, into = c("power", "num2"), "[.]") %>%
+  mutate(power = as.numeric(gsub("_", ".", power))) -> plot.dat
+
+
+plot.dat %>%
+  mutate(radius = multiplier*(H + V)^power) %>%
+  group_by(radius, n_param) %>%
+  summarise(frac_degen = sum(near_hull)/n()) %>%
[…]

### 2016-12-02 code/degeneracy/radius_percent_degen_max_no_degen_plots_split.R

@@ -1,0 +1,95 @@
+library(ggplot2)
+library(dplyr)
+library(tidyr)
+
+plot_data <- function(res) {
+
+  plot.data <- data.frame()
+  for(i in 1:nrow(res)) {
+    tmp <- res$samp[[i]]
+    H <- res[i,]$H
+    V <- res[i,]$V
+
+    tmp %>%
+      data.frame() %>%
+      rowwise() %>%
+      mutate_(ss_ratio = paste0("(", paste(paste0(names(.)[(H + V + 1):(H + V + H*V)], "^2"), collapse = " + "), ")/sum(", paste(paste0(names(.)[1:(H+V)],"^2"), collapse = " + "), ")")) %>%
+      ungroup() -> ratio
+
+    tmp %>%
+      data.frame() %>%
+      select(-starts_with("exp")) %>%
+      rowwise() %>%
+      mutate_(true_r = paste0("sqrt(", paste(paste0(names(.), "^2"), collapse = " + "), ")")) -> radius
+
+    inner_join(radius, ratio) %>%
+    inner_join(res$outside[[i]] %>% ungroup()) %>%
+      select(ss_ratio, near_hull, true_r) %>%
+      mutate(H = H, V = V, n_param = H + V + H*V) %>%
+      rbind(plot.data) -> plot.data
+  }
+  return(plot.data)
+}
+
+n <- 100
+r_center <- seq(0.2, 3, by = 0.2)
+r_exp <- .05
+
+
+plot.dat <- list()
+for(i in seq(0, 1, by = 0.1)) {
+  load(paste0("written/results_split_", i, ".RData"))
+  plot_dat <- lapply(res, plot_data)
+  names(plot_dat) <- gsub("[.]", "_", r_center)
+
+  do.call(rbind, plot_dat) %>%
+    ungroup() %>%
+    mutate(name = rownames(do.call(rbind, plot_dat))) %>%
+    separate(name, into = c("multiplier", "num"), "[.]") %>%
+    mutate(multiplier = as.numeric(gsub("_", ".", multiplier))) %>%
+    mutate(n_param_f = factor(n_param)) -> plot_dat
+
+  plot.dat[[i*10 + 1]] <- plot_dat
+}
+
+names(plot.dat) <- gsub("[.]", "_", seq(0, 1, by = 0.1))
+do.call(rbind, plot.dat) %>%
+  mutate(name2 = rownames(do.call(rbind, plot.dat))) %>%
+  separate(name2, into = c("power", "num2"), "[.]") %>%
[…]

### 2016-12-02 code/degeneracy/radius_percent_degen_split.R

@@ -1,0 +1,151 @@
+## load libraries --------------------------- 
+source("functions.R")
+library(dplyr)
+library(tidyr)
+library(ggplot2)
+
+## main function --------------------------------------------
+find_prop <- function(g_theta, stat, r_exp) {
+  g_theta %>%
+    data.frame() %>%
+    group_by_(.dots = c(colnames(g_theta))) %>%
+    do(samp = sample_points_outside(as.numeric(.), stat, r_exp)$in_hull) %>%
+    group_by_(.dots = c(colnames(g_theta))) %>%
+    do(as.data.frame(sum(.$samp[[1]]) > 0)) -> is_outside # count of not in hull greater than zero => there exists a point outside hull for this model
+
+  is_outside %>%
+    rename_(near_hull = as.name(names(is_outside)[ncol(is_outside)])) %>%
+    mutate(r = r_exp) -> inside_outside
+
+  return(inside_outside)
+}
+
+## additional useful functions -------------------------------------
+sample_points_outside <- function(g_theta, stats, r, n = 100) {
+  #g_theta is the expected value mapping of a point theta in R^H+V+H*V
+  x0 <- data.frame(samp = 1:n) %>%
+    mutate(vals = samp_exp(length(g_theta), g_theta, r)) %>%
+    separate(col = vals, sep = ",", into = colnames(stats)) %>%
+    select(-samp)
+
+  x0$in_hull <- !apply(x0, 1, in_hull, stats)
+
+  return(x0)
+}
+
+samp_exp <- function(space, center, r) {
+  vec <- rnorm(space)
+  paste0(r*vec/sqrt(sum(vec^2)) + center, collapse=",")
+}
+
+in_hull <- function(point, hull_points) {
+  require(lpSolveAPI)
+  P <- data.matrix(point)
+  A <- t(data.matrix(hull_points))
+
+  lp_obj <- make.lp(nrow = 0, ncol = ncol(A))
+  sapply(1:nrow(A), function(x) add.constraint(lp_obj, A[x,], type = "=", rhs = P[x]))
+  add.constraint(lp_obj, rep(1, ncol(A)), type = "=", rhs = 1)
+  sapply(1:ncol(A), function(x) add.constraint(lp_obj, 1, type = ">=", rhs = 0, indices = x))
+
+  return(solve(lp_obj) == 0)
+}
+
[…]

### 2016-12-02 code/degeneracy/random_search.R

@@ -1,0 +1,89 @@
+source("functions.R")
+library(scatterplot3d)
+library(plyr)
+library(dplyr)
+library(reshape2)
+library(ggplot2)
+
+
+###############################################
+## Generate data
+###############################################
+
+## Num nodes
+H <- 1
+V <- 1
+
+## Num to sample
+n <- 10000
+
+## Radius to sample
+r <- c(seq(0,1, by=0.05), seq(1.5, 10, by=0.5))
+
+stat <- stats(H, V, "negative")
+theta <- mlply(r, function(x) sample_sphere(space = ncol(stat), n, x))
+exp_vals <- llply(theta, function(x) expected_value(stats = stat, theta = x, normalized = TRUE))
+lines <- generate_hull_from_stats(stat)
+
+dist <- ldply(exp_vals, function(y) apply(y, 2, function(x){ min_distance_to_hull(x, lines) }))
+min_dist <- apply(dist[,-1], 1, min)
+min_point <- apply(dist[,-1], 1, which.min)
+
+## Data frame of points at each radius with minimum distance to a closest line
+min_points <- ldply(1:length(theta), function(i) theta[[i]][, min_point[i]])
+names(min_points) <- names(stat)
+min_points$r <- r
+min_points$min_dist <- min_dist
+
+## Start with first radius that has a point within .01 of a line
+cut <- .01
+sigma <- .5
+n0 <- 500
+
+mu0 <- min_points[min_dist < cut,][1, 1:ncol(stat)] 
+sphere = list(center = c(0,0,0), radius = min(min_points[min_dist < cut, "r"]))
+
+
+local_search <- function(mu0, sigma, sphere, dim = 3, n = 500, cutoff = .01) {
+  store_mus <- rbind(mu0, c(0,0,0))
+  
+  search <- function(mu0, sigma, sphere, dim, n, cutoff) {
+    local_theta <- t(matrix(rnorm(n*dim, sd = sigma), nrow=dim) + mu0)
+    proj_theta <- t(apply(local_theta, 1, function(point) project_to_sphere(sphere=sphere, point=point)))
+    exp_vals <- expected_value(stats = stat, theta = t(proj_theta), normalized = TRUE)
+    dist <- apply(exp_vals, 2, function(x)  min_distance_to_hull(x, lines))
+    mu1 <- proj_theta[which.min(dist),]
[…]

### 2016-12-02 writing/draft.Rmd
add formatting

@@
 title: "On the propriety of restricted Boltzmann machines"
-author: | 
-  | Andee Kaplan, Daniel Nordman, and Stephen Vardeman
-  | Iowa State University
-  | ajkaplan@iastate.edu, dnordman@iastate.edu, vardeman@iastate.edu
-date: "December, 2016"
-linestretch: 2 
+author: 
+  - name: Andee Kaplan
+    university: Iowa State University
+    department: Department of Statistics
+  - name: Daniel Nordman
+    university: Iowa State University
+    department: Department of Statistics
+  - name: Stephen Vardeman
+    university: Iowa State University
+    department: Department of Statistics
 output: 
@@
     number_sections: yes
+    template: latex/technometrics_article.tex
+fontsize: 12pt
 bibliography: references/refs.bib
@@
 We also describe the difficulties that arise in likelihood-based and Bayes fitting of such (highly flexible) models, especially as Gibbs sampling (quasi-Bayes) methods are often advocated for the RBM model structure.
+keywords: Degeneracy, Instability, Classification, Deep Learning, Graphical Models
+blind: false
 ---
@@
 Here the mean of $\boldsymbol t(\boldsymbol x)$ is
-$$
-\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta} \boldsymbol t(\boldsymbol X)  = \sum\limits_{x \in \mathcal{C}^{\nv + \nh}} \left\{ \boldsymbol t(\boldsymbol x) f_{\boldsymbol \theta} (\boldsymbol x) \right\} = \sum\limits_{x \in \mathcal{C}^{\nv + \nh}} \left\{ \boldsymbol t(\boldsymbol x)\dfrac{\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{C}^{\nh + \nv}}\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}\right\}.
-$$
+\begin{align*}
+\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta} \boldsymbol t(\boldsymbol X)  &= \sum\limits_{x \in \mathcal{C}^{\nv + \nh}} \left\{ \boldsymbol t(\boldsymbol x) f_{\boldsymbol \theta} (\boldsymbol x) \right\} \\
+&= \sum\limits_{x \in \mathcal{C}^{\nv + \nh}} \left\{ \boldsymbol t(\boldsymbol x)\dfrac{\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{C}^{\nh + \nv}}\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}\right\}.
+\end{align*}
 
@@
 $$
-\Delta(\boldsymbol \theta_\nv) \equiv \max \left\{\log \frac{P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}{P_{\boldsymbol \theta_\nv}(v_1^*, \dots, v_\nv^*)} : (v_1, \dots, v_\nv) \& (v_1^*, \dots, v_\nv^*) \in \mathcal{C}^\nv \text{ differ in exactly one component}\right\},
+\Delta(\boldsymbol \theta_\nv) \equiv \max \left\{\log \frac{P_{\boldsymbol \theta_\nv}(\boldsymbol v)}{P_{\boldsymbol \theta_\nv}(\boldsymbol v^*)} : \boldsymbol v \text{ }\& \text{ } \boldsymbol v^* \in \mathcal{C}^\nv \text{ differ in exactly one component}\right\},
 $$
@@
 We can compute the mean vector for $\boldsymbol t(\boldsymbol x)$ as a function of the model parameters as 
-\begin{align}
-\boldsymbol \mu(\boldsymbol \theta) &= \text{E}_{\boldsymbol \theta}\left[ \boldsymbol t(\boldsymbol X) \right] =  \sum\limits_{\boldsymbol x = (v_1, h_1) \in \mathcal{X} = \{0,1\}^2} \left\{ t(x) \frac{\exp\left( \theta_{11} h_1 v_1 + \theta_{h1} h_1 + \theta_{v1} v_1\right)}{\gamma(\boldsymbol \theta)} \right\}
-= \left[
+\begin{align*}
+\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta}\left[ \boldsymbol t(\boldsymbol X) \right] =  \sum\limits_{\boldsymbol x = (v_1, h_1) \in \{0,1\}^2} \left\{ t(x) \frac{\exp\left( \theta_{11} h_1 v_1 + \theta_{h1} h_1 + \theta_{v1} v_1\right)}{\gamma(\boldsymbol \theta)} \right\} = \left[
 \begin{matrix}
@@
 \end{matrix} \right]
-\end{align}
+\end{align*}
 where $\gamma(\boldsymbol \theta) = \sum\limits_{h_1 = 0}^1\sum\limits_{v_1 = 0}^1 \exp(\theta_{11} h_1 v_1 + \theta_{h_1}h_1 + \theta_{v_1}v_1)$.
@@
   geom_segment(aes(x=lag, xend=lag, y=0, yend=marginal), colour = "red") +
+  geom_point(aes(x=lag, y=marginal), colour = "red", size = 0.5) +
   facet_wrap(~image_id) +

### 2016-12-02 writing/note.Rmd
add formatting

@@
 title: "A note on the instability and degeneracy of deep learning models"
-author: | 
-  | Andee Kaplan, Daniel Nordman, and Stephen Vardeman
-  | Iowa State University
-  | ajkaplan@iastate.edu, dnordman@iastate.edu, vardeman@iastate.edu
-date: "December, 2016"
-linestretch: 2
+author: 
+  - name: Andee Kaplan
+    affiliation: Iowa State University
+    department: Department of Statistics
+  - name: Daniel Nordman
+    affiliation: Iowa State University
+    department: Department of Statistics
+  - name: Stephen Vardeman
+    affiliation: Iowa State University
+    department: Department of Statistics
 output: 
@@
 These results on instability apply to large classes of models commonly used in random graphs, network analysis, and machine learning contexts.
+keywords: Degeneracy, Instability, Deep Learning, Graphical Models
+blind: false
 ---
@@
 $$
-\psi(\boldsymbol \theta_N) = \log \sum\limits_{\tilde{\boldsymbol x} \in \mathcal{X}^{N_{(H,1)} + \cdots + N_{(H,M)} +N}} \exp\left[ \sum\limits_{i = 1}^M\boldsymbol \alpha^{(i)T} \boldsymbol h^{(i)} + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^{(1)T} \Gamma^{(0)} \boldsymbol x + \sum\limits_{i = 1}^{M - 1} \boldsymbol h^{(i)T} \Gamma^{(i)} \boldsymbol h^{(i + 1)}\right], \qquad \boldsymbol \theta_N \in \Theta_N
+\psi(\boldsymbol \theta_N) = \log \sum\limits_{\tilde{\boldsymbol x} \in \mathcal{X}^{N_{(H,1)} + \cdots + N_{(H,M)} +N}} \exp\left[ \sum\limits_{i = 1}^M\boldsymbol \alpha^{(i)T} \boldsymbol h^{(i)} + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^{(1)T} \Gamma^{(0)} \boldsymbol x + \sum\limits_{i = 1}^{M - 1} \boldsymbol h^{(i)T} \Gamma^{(i)} \boldsymbol h^{(i + 1)}\right],
 $$ 
-is the normalizing function and parameters in the model are $\boldsymbol \beta \in \mathbb{R}^V$, $\boldsymbol \alpha^{(i)} \in \mathbb{R}^{N_{(H,i)}}$ for $i = 1, \dots, M$, along with a matrix $\Gamma^{(0)}$ of dimension $N_{(H,1)} \times N$, and matrices $\Gamma^{(i)}$ of dimension $N_{(H,i)} \times N_{(H,i+1)}$ for $i = 1, \dots, M-1$.
+for $\boldsymbol \theta_N \in \Theta_N$ is the normalizing function and parameters in the model are $\boldsymbol \beta \in \mathbb{R}^V$, $\boldsymbol \alpha^{(i)} \in \mathbb{R}^{N_{(H,i)}}$ for $i = 1, \dots, M$, along with a matrix $\Gamma^{(0)}$ of dimension $N_{(H,1)} \times N$, and matrices $\Gamma^{(i)}$ of dimension $N_{(H,i)} \times N_{(H,i+1)}$ for $i = 1, \dots, M-1$.
 Let $\boldsymbol \theta_N = (\boldsymbol \alpha^{(1)}, \dots, \boldsymbol \alpha^{(M)}, \boldsymbol \beta,\Gamma^{(0)}, \dots, \Gamma^{(M - 1)}) \in \Theta_N \subset \mathbb{R}^{q(N)}$ denote the combined vector of parameters with total length $q(N)= N_{(H,1)}+\cdots N_{(H,M)} + N + N_{(H,1)}*N+N_{H,2}*H_{(H,1)}+\cdots +N_{(H,M)}*H_{(H,M)-1}$.
@@
 $$
-\Delta(\boldsymbol \theta_N) \equiv \max \left\{\log \frac{P_{\boldsymbol \theta_N}(x_1, \dots, x_N)}{P_{\boldsymbol \theta_N}(x_1^*, \dots, x_N^*)} : (x_1, \dots, x_N) \& (x_1^*, \dots, x_N^*) \in \mathcal{X}^N \text{ differ in exactly one component}\right\},
+\Delta(\boldsymbol \theta_N) \equiv \max \left\{\log \frac{P_{\boldsymbol \theta_N}(\boldsymbol x)}{P_{\boldsymbol \theta_N}(\boldsymbol x^*)} : \boldsymbol x \text{ }\& \text{ } \boldsymbol x^* \in \mathcal{X}^N \text{ differ in exactly one component}\right\},
 $$
@@
 Thus, for the combined replications $\boldsymbol X_1, \ldots, \boldsymbol X_n$, the analog definition of the extremal log-probability becomes
-$$
-\frac{\text{extremal log-probability ratio}}{\text{\# random variables in the model}} \equiv \frac{1}{n*N}\log \left[\frac{\left(\max\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)\right)^n}{\left(\min\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)\right)^n}\right] = \frac{1}{N} \log \left[\frac{\max\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)}{\min\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)}\right] = \frac{1}{N} \text{ELPR}(\boldsymbol \theta_N),
-$$
+\begin{align*}
+\frac{\text{extremal log-probability ratio}}{\text{\# random variables in the model}} &\equiv \frac{1}{n*N}\log \left[\frac{\left(\max\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)\right)^n}{\left(\min\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)\right)^n}\right] = \frac{1}{N} \log \left[\frac{\max\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)}{\min\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)}\right] \\
+&= \frac{1}{N} \text{ELPR}(\boldsymbol \theta_N),
+\end{align*}
 implying that the definition of an S-unstable FSFS model sequence is invariant to the level ($n$) of independent replication.

### 2016-12-03 writing/draft.Rmd
update vardeman's affiliation

@@
     university: Iowa State University
-    department: Department of Statistics
+    department: |
+      | Department of Statistics and Industrial Manufacturing 
+      | Systems Engineering Department
 output: 

### 2016-12-03 writing/note.Rmd
update vardeman's affiliation

@@
   - name: Andee Kaplan
-    affiliation: Iowa State University
+    university: Iowa State University
     department: Department of Statistics
   - name: Daniel Nordman
-    affiliation: Iowa State University
+    university: Iowa State University
     department: Department of Statistics
   - name: Stephen Vardeman
-    affiliation: Iowa State University
-    department: Department of Statistics
+    university: Iowa State University
+    department: |
+      | Department of Statistics and Industrial Manufacturing 
+      | Systems Engineering Department
 output: 
   pdf_document:
-    keep_tex: true
+    keep_tex: false
     fig_caption: yes

### 2016-12-04 writing/appendix/appendix-center.Rmd
Submission

--- 2016-11-20 writing/appendix/appendix-center.Rmd
@@
 ---
-title: "Supplementary Material to:  On the propriety of restricted Boltzmann machines"
-author: "Andee Kaplan, Daniel Nordman, and Stephen Vardeman"
-date: "November, 2016"
-linestretch: 2 
+title: "Supplementary material to:  On the propriety of restricted Boltzmann machines"
+author: 
+  - name: Andee Kaplan
+    university: Iowa State University
+    department: Department of Statistics
+  - name: Daniel Nordman
+    university: Iowa State University
+    department: Department of Statistics
+  - name: Stephen Vardeman
+    university: Iowa State University
+    department: |
+      | Department of Statistics and Industrial Manufacturing 
+      | Systems Engineering Department
 output: 
@@
     number_sections: yes
+    template: ../latex/technometrics_article.tex
+fontsize: 12pt
+blind: true
 ---

### 2016-12-04 writing/draft.Rmd
Submission

--- 2016-12-02 writing/draft.Rmd
@@
     university: Iowa State University
-    department: Department of Statistics
+    department: |
+      | Department of Statistics and Department of Industrial and
+      | Manufacturing Systems Engineering
 output: 
   pdf_document:
+    keep_tex: true
     fig_caption: yes

### 2016-12-11 presentations/amherst/slides.Rmd
slides start

---
title: "Model matters with restricted Boltzmann machines"
shorttitle: "Model matters with RBMs"
author: Andee Kaplan, Daniel Nordman, and Stephen Vardeman
shortname: Kaplan, et al.
institute: ajkaplan@iastate.edu
date: "December 20, 2016"
output: 
  beamer_presentation:
    template: latex/beamer.tex
    includes:
      in_header: latex/front-matter.tex
theme: CambridgeUS
bibliography: latex/refs.bib
fig_caption: true
nocite: |
    @salakhutdinov2009deep, @kaiser2007statistical
---

----

```{r libraries, echo=FALSE, message=FALSE, warning=FALSE}
library(knitr)
library(ggplot2)
library(dplyr)
library(tidyr)

opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE)
theme_set(theme_bw())

source("helpers/functions.R")
```

# What is this?

A Restricted Boltzman Machine (RBM) is an undirected probabilistic graphical model (for discrete or continuous random variables) with two layers, one hidden and one visible, with conditional independence within a layer [@smolensky1986information].

\begin{figure}
\includegraphics[width=.6\linewidth]{images/rbm.png}
\caption{An example RBM, which consists of two layers. Hidden nodes are indicated by white circles and the visible nodes are indicated by blue circles}
\end{figure}

# How is it used?

Typically used for image classification. Each image pixel is a node in the visible layer. The output creates features, which are passed to a supervised learning algorithm.

\begin{figure}
\includegraphics[width=\linewidth]{images/visibles_image.png}
\caption{Image classification using a RBM. On the left, each image pixel comprises a node in the visible layer, $\mathcal{V}$. On the right, the output of the RBM is used to create features which are then passed to a supervised learning algorithm.}
\end{figure}

# Joint distribution
Let $\boldsymbol x = (h_1, \dots, h_{\nh}, v_1,\dots,v_{\nv})$ represent the states of the visible and hidden nodes in an RBM. Each single "binary" random variable, visible or hidden, will take its values in a common coding set $\mathcal{C}$, with two possibilities for the coding set $\mathcal{C}=\{0,1\}$ or $\mathcal{C}=\{-1,1\}$. A parametric form for probabilities corresponding to a potential vector of states of each node

\begin{align} 
\label{eqn:pmf} 
f_{\boldsymbol \theta} (\boldsymbol x) = \frac{\exp\left(\sum\limits_{i = 1}^{\nv} \sum\limits_{j=1}^{\nh} \theta_{ij} v_i h_j + \sum\limits_{i = 1}^{\nv}\theta_{v_i} v_i + \sum\limits_{j = 1}^{\nh}\theta_{h_j} h_j\right)}{\gamma(\boldsymbol \theta)} \end{align}

where

$$\gamma(\boldsymbol \theta) = \sum\limits_{\boldsymbol x \in \mathcal{C}^{\nh + \nv}}\exp\left(\sum\limits_{i = 1}^{\nv} \sum\limits_{j=1}^{\nh} \theta_{ij} v_i h_j + \sum\limits_{i = 1}^{\nv}\theta_{v_i} v_i + \sum\limits_{j = 1}^{\nh}\theta_{h_j} h_j\right)$$.

# Deep learning

\begin{columns}[T] % align columns
\begin{column}{.48\textwidth}
\begin{itemize}
\item By stacking layers of RBMs in a deep architecture, proponents of the models claim the ability to learn "internal representations that become increasingly complex, which is considered to be a promising way of solving object and speech recognition problems" (Salakhutdinov and Hinton 2009, pp. 450).
\item Stacking is achieved by treating a hidden layer of one RBM as the visible layer in a second RBM, etc.
\end{itemize}
\end{column}
\hfill
\begin{column}{.48\textwidth}
\begin{figure}
\includegraphics[width=\linewidth]{images/deep_rbm.png}
\caption{Three layer deep Boltzmann machine, with visible-to-hidden and hidden-to-hidden connections but no within-layer connections.}
\end{figure}
\end{column}
\end{columns}

# Why do I care?

The model properties are largely unexplored in the literature and the commonly cited fitting methodology remains heuristic-based and abstruse [@hinton2006fast].

We want to

1. provide steps toward a thorough understanding of the model class and its properties from the perspective of statistical theory, and 
2. explore the possibility of a rigorous fitting methodology.

# Degeneracy, instability, and uninterpretability. Oh my!

The highly flexible nature of a RBM (having as it does $\nh + \nv + \nh*\nv$ parameters) makes at least three kinds of potential model impropriety of concern, *degeneracy*, *instability*, and *uninterpretability*.

> A model should "provide an explanation of the mechanism underlying the observed phenomena" [@lehmann1990model; @box1967discrimination].

RBMs often

- fail to generate data with realistic variability and thus an unsatisfactory conceptualization of the data generation process. 
- exhibit instability in the parameter space.

Such model impropriety issues have been documented in in RBMs [@li2014biclustering], as well as other deep architectures [@szegedy2013intriguing; @nguyen2014deep].

# Near-degeneracy

\begin{definition}[Model Degeneracy]
There is a disproportionate amount of probability placed on only a few elements of the sample space, $\mathcal{C}^{\nh + \nv}$, by the model.
\end{definition}

RBM models exhibit *near-degeneracy* when random variables in the neg-potential function 
$$Q_{\boldsymbol \theta}(\boldsymbol x) = \sum\limits_{i = 1}^{\nv} \sum\limits_{j=1}^{\nh} \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j,
$$ 
have a mean vector $\boldsymbol \mu(\boldsymbol \theta)$ close to the boundary of the convex hull of $\mathcal{T} = \{\boldsymbol t(x): x \in \mathcal{C}^{\nh + \nv}\}$ [@handcock2003assessing], where $\boldsymbol t(\boldsymbol x) = \{v_1, \dots, v_{\nv}, h_1, \dots, h_{\nh}, v_1 h_1, \dots, v_{\nv} h_{\nh} \}$ and the mean parameterization on the model parameters, $\boldsymbol \mu(\boldsymbol \theta)$ is

---

\begin{align*}
\boldsymbol \mu(\boldsymbol \theta) &= \text{E}_{\boldsymbol \theta} \boldsymbol t(\boldsymbol X)  \\
&= \sum\limits_{\boldsymbol x \in \mathcal{C}^{\nh + \nv}} \left\{ \boldsymbol t(\boldsymbol x)\dfrac{\exp\left(\sum\limits_{i = 1}^{\nv} \sum\limits_{j=1}^{\nh} \theta_{ij} v_i h_j + \sum\limits_{i = 1}^{\nv}\theta_{v_i} v_i + \sum\limits_{j = 1}^{\nh}\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{C}^{\nh + \nv}}\exp\left(\sum\limits_{i = 1}^{\nv} \sum\limits_{j=1}^{\nh} \theta_{ij} v_i h_j + \sum\limits_{i = 1}^{\nv}\theta_{v_i} v_i + \sum\limits_{j = 1}^{\nh}\theta_{h_j} h_j\right)}\right\}. \\
\end{align*}

# Instability

\begin{definition}[Instability]
Characterized by excessive sensitivity in the model, where small changes in the components of data outcomes, $\boldsymbol x$, lead to substantial changes in probability.
\end{definition}

@schweinberger2011instability introduced a concept of model deficiency related to *instability* considering only a class of exponential families of distributions.

To quantify *instability* in the RBM, it is useful to consider how a data model might be expanded to incorporate more visibles. For this, it becomes necessary to grow the number of model parameters in a sequence $\boldsymbol \theta_{\nv} \in \mathbb{R}^{\nv + \nh + \nv*\nh}, \nv \ge 1$ (one may also arbitrarily expand the number of hidden variables used).

# Unstable RBMs

\begin{definition}[S-unstable RBM]
A RBM model formulation is \emph{S-unstable} if
\begin{align*}
\lim\limits_{\nv \rightarrow \infty} \frac{1}{\nv} \text{ELPR}(\boldsymbol \theta_{\nv}) = \infty.
\end{align*}
where 
\begin{align}
\label{eqn:elpr}
\text{ELPR}(\boldsymbol \theta_{\nv}) = \log \left[\frac{\max\limits_{(v_1, \dots, v_{\nv}) \in \mathcal{C}^\nv}P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}{\min\limits_{(v_1, \dots, v_\nv) \in \mathcal{C}^\nv}P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}\right]
\end{align}
\end{definition}

S-unstable RBM model sequences are undesirable for several reasons. Again, one is that small changes in data can lead to overly-sensitive changes in probability.

# One-pixel change

Consider, for example, the biggest log-probability ratio for a one-pixel (one component) change in data outcomes (visibles).

$$
\Delta(\boldsymbol \theta_\nv) \equiv \max \left\{\log \frac{P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}{P_{\boldsymbol \theta_\nv}(v_1^*, \dots, v_\nv^*)} \right\},
$$

where $(v_1, \dots, v_\nv) \& (v_1^*, \dots, v_\nv^*) \in \mathcal{C}^\nv$ differ by exactly one component

\begin{result}
\label{prop:instab}
Let $C > 0$ and $\nv \ge 1$. If $\frac{1}{\nv}\text{ELPR}(\boldsymbol \theta_\nv) > C$, then $\Delta(\boldsymbol \theta_\nv) > C$.
\end{result}

If the quantity (\ref{eqn:elpr}) is too large, then a RBM model sequence will exhibit large probability shifts for very small changes in the data configuration.

# Tie to degeneracy

Unstable RBM model sequences are connected to degenerate model sequences (placing all probability on a small portion of their sample spaces). Define a modal set

\small
\begin{align*}
M_{\epsilon, \boldsymbol \theta_\nv} &\equiv \left\{\boldsymbol v \in \mathcal{C}^\nv: \log P_{\boldsymbol \theta_\nv}(\boldsymbol v)  > (1-\epsilon)\max\limits_{\boldsymbol v^*}P_{\boldsymbol \theta_\nv}(\boldsymbol v^*) + \epsilon\min\limits_{\boldsymbol v^*}P_{\boldsymbol \theta_\nv}(\boldsymbol v^*) \right\}
\end{align*}

\normalsize
of possible outcomes, for a given $0 < \epsilon < 1$.

\begin{result}
\label{prop:degen}
For an S-unstable RBM model, and for any given $0 < \epsilon < 1$, $P_{\boldsymbol \theta_\nv}\left((v_1, \dots, v_\nv) \in M_{\epsilon, \boldsymbol \theta_\nv}\right) \rightarrow 1$ holds as $\nv \rightarrow \infty$.
\end{result}

As a consequence of unstable models, all probability will stack up on mode sets or potentially those few outcomes with the highest probability. Proofs of results \ref{prop:instab}-\ref{prop:degen} are provided later.

# Uninterpretability

\begin{definition}[Uninterpretability]
Characterized by marginal mean-structure (controlled by main effect parameters $\theta_{v_i}, \theta_{h_j}$) not being maintained in the model due to dependence (interaction parameters $\theta_{ij}$) (Kaiser 2007).
\end{definition}

A measure of this is the magnitude of the difference between model expectations, E$\left[\boldsymbol X | \boldsymbol \theta\right]$, and expectations given independence, E$\left[\boldsymbol X | \boldsymbol \theta^* \right]$, where $\boldsymbol \theta^*$ is defined to equal $\boldsymbol \theta$ with all $\theta_{ij} = 0$ for $i = 1, \dots, \nv, j = 1, \dots, \nh$.

Using this, it is possible to investigate what parameters lead to uninterpretability in a model versus those that guarantee interpretability.

If $|\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] - \text{E}\left[\boldsymbol X | \boldsymbol \theta^*\right]|$ is large then the RBM with parameter vector $\boldsymbol \theta$ is *uninterpetable*.

# RBM quantities to compare

\begin{align*}
\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] &= \sum\limits_{\boldsymbol x \in \mathcal{C}^{\nh + \nv}} \boldsymbol x \frac{\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{C}^{\nh + \nv}}\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)} \\
~\\
\text{E}\left[\boldsymbol X | \boldsymbol \theta^*\right] &= \sum\limits_{\boldsymbol x \in \mathcal{C}^{\nh + \nv}} \boldsymbol x \frac{\exp\left(\sum\limits_{i = 1}^\nv \theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{C}^{\nh + V=\nv}}\exp\left(\sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)} 
\end{align*}

# Data coding to mitigate degeneracy

\begin{figure}
\includegraphics[width=.55\linewidth]{images/toyhull_top.png}
\includegraphics[width=.35\linewidth]{images/toyhull_top_2.png}
\caption{The convex hull of the "statistic space" $\mathcal{T} = \{(v_1, h_1, v_1 h_1): v_1, h_1 \in \mathcal{C}\}$ in three dimensions for the toy RBM with one visible and one hidden node for $\mathcal{C} = \{0,1\}$ (left) and $\mathcal{C} = \{-1,1\}$ (right) data encoding. The convex hull of $\mathcal{T} \subset \mathcal{C}^3$ does not fill the unit cube $[0,1]^3$ (left), but does better with $[-1,1]^3$ (right).}
\end{figure}

# The center of the universe

- For the $\mathcal{C} = \{-1, 1 \}$ encoding of hiddens $(H_1, \dots, H_\nh)$ and visibles $(V_1, \dots, V_\nv)$, the origin is the center of the parameter space.
- At $\boldsymbol \theta = \boldsymbol 0$, RBM is equivalent to elements of $X$ being distributed as iid Bernoulli$\left(\frac{1}{2}\right)$ $\Rightarrow$ No *near-degeneracy*, *instability*, or *uninterpretability*!

```{r encoding-volume, echo = FALSE, fig.height=3, fig.cap="\\label{fig:volume_plot}Relationship between volume of the convex hull of possible values of the RBM sufficient statistics and the cube containing it for different size models."}
test_cases <- expand.grid(data.frame(1:4)[,rep(1,2)]) %>% 
  rename(H = X1.4, V = X1.4.1) %>%
  filter(H <= V) %>% #remove those cases with less visibles than hiddens. Is this necessary?
  mutate(n_param = H*V + H + V) %>%
  mutate(max_facets = (2^(H+V))^(floor(n_param/2))) %>%
  filter(n_param <= 11) #calc_hull can't handle any higher dimensions currently

bin_res <- plyr::dlply(test_cases, plyr::.(n_param), function(x) calc_hull(x$V, x$H, "binary"))
neg_res <- plyr::dlply(test_cases, plyr::.(n_param), function(x) calc_hull(x$V, x$H, "negative"))

plyr::ldply(bin_res, function(x) x$c_hull$vol) %>% 
  mutate(frac_vol = V1/(1^n_param)) %>%
  inner_join(plyr::ldply(neg_res, function(x) x$c_hull$vol) %>% 
               mutate(frac_vol = V1/(2^n_param)),
             by="n_param") %>%
  rename(vol.bin = V1.x, vol.neg = V1.y, frac_vol.bin = frac_vol.x, frac_vol.neg = frac_vol.y) %>%
  gather(vars, value, -n_param) %>%
  separate(vars, c("type", "encoding"), "\\.") %>%
  spread(type, value) %>%
  ggplot() +
  geom_point(aes(x=n_param, y=frac_vol, colour=encoding)) +
  geom_line(aes(x=n_param, y=frac_vol, colour=encoding, group=encoding)) +
  ylab("Fraction of unrestricted volume") +
  xlab("Number of parameters") +
  scale_colour_discrete("Encoding", labels=c("Binary (1,0)", "Negative (1,-1)")) +           
  theme(
    plot.background = element_rect(fill = "transparent", colour = NA),
    legend.background = element_rect(fill = "transparent", colour = NA)) -> p

p
```

# Manageable (a.k.a. small) examples

```{r degen-data, message=FALSE, warning=FALSE, cache=TRUE}
#reshape data functions
plot_data <- function(res, grid = FALSE) {

  plot.data <- data.frame()

  if(!grid) {
    for(i in 1:nrow(res)) {
      tmp <- res$g_theta[[i]] %>% data.frame()
      H <- res[i,]$H
      V <- res[i,]$V
      N <- res[i,]$N
      r1 <- res[i,]$r1
      r2 <- res[i,]$r2
      C <- res[i,]$C
      epsilon <- res[i,]$epsilon

      tmp %>% 
        rowwise() %>% 
        mutate_(ss_interaction = paste0("(", paste(paste0(names(.)[(H + V + 1):(H + V + H*V)], "^2"), collapse = " + "), ")"),
                ss_main = paste0("sum(", paste(paste0(names(.)[1:(H+V)],"^2"), collapse = " + "), ")")) %>%    
        ungroup() -> ratio

      inner_join(ratio, res$outside[[i]] %>% ungroup()) %>%
        select(ss_interaction, ss_main, near_hull) %>%
        mutate(H = H, V = V, n_param = H + V + H*V, N = H + V, N = N, r1 = r1, r2 = r2, C = C, epsilon = epsilon) %>%
        rbind(plot.data) -> plot.data
    }
  } else {
    for(i in 1:nrow(res)) {
      tmp <- res$g_theta[[i]] %>% data.frame()
      H <- res[i,]$H
      V <- res[i,]$V
      N <- res[i,]$N
      r1 <- res[i,]$r1
      r2 <- res[i,]$r2

      tmp %>% 
        rowwise() %>% 
        mutate_(ss_interaction = paste0("(", paste(paste0(names(.)[(H + V + 1):(H + V + H*V)], "^2"), collapse = " + "), ")"),
                ss_main = paste0("sum(", paste(paste0(names(.)[1:(H+V)],"^2"), collapse = " + "), ")")) %>%  
        ungroup() -> ratio

      inner_join(ratio, res$outside[[i]] %>% ungroup()) %>%
        select(ss_interaction, ss_main, near_hull) %>%
        mutate(H = H, V = V, n_param = H + V + H*V, N = H + V, N = N, r1 = r1, r2 = r2) %>%
        rbind(plot.data) -> plot.data
    }
  }

  return(plot.data)
}
indep_params <- function(samp, H, V) {
  samp[, (H + V + 1):ncol(samp)] <- 0
  samp
}
max_Q <- function(theta, stats) {
  apply(crossprod(t(stats), theta), 2, max)
}
min_Q <- function(theta, stats) {
  apply(crossprod(t(stats), theta), 2, min)
}
min_max_Q <- function(theta, stats) {
  require(dplyr)
  tcrossprod(stats, theta) %>% data.frame() %>%
    cbind(stats) %>%
    group_by_(.dots = colnames(stats)[grepl("v", colnames(stats)) & !grepl("theta", colnames(stats))]) %>%
    summarise_each(funs(max), contains("X")) %>%
    ungroup() %>%
    summarise_each(funs(min), contains("X")) %>%
    select(contains("X")) %>%
    data.matrix() %>%
    t()
}
elpr <- function(theta, stats) {
  require(dplyr)
  exp(tcrossprod(stats, theta)) %>% data.frame() %>%
    cbind(stats) %>%
    group_by_(.dots = colnames(stats)[grepl("v", colnames(stats)) & !grepl("theta", colnames(stats))]) %>%
    summarise_each(funs(sum), contains("X")) %>%
    ungroup() -> marginalized

  marginalized %>%
    summarise_each(funs(max), contains("X")) %>%
    select(contains("X")) %>%
    data.matrix() -> max_marg

  marginalized %>%
    summarise_each(funs(min), contains("X")) %>%
    select(contains("X")) %>%
    data.matrix() -> min_marg

  t(log(max_marg/min_marg))
}

#grid data
load("data/results_grid.RData")

#near-degeneracy
plot_dat_grid <- res %>% plot_data(grid = TRUE)

plot_dat_grid %>%
  group_by(r1, r2, H, V) %>%
  summarise(frac_degen = sum(near_hull)/n(), count = n()) %>% 
  ungroup() %>% 
  mutate(Hiddens = paste0("paste(n[h], '= ", H, "')"), Visibles = paste0("paste(n[v], '= ", V, "')")) -> convex_hull_summary

#uninterpretability
res %>%
  group_by(H, V, N, r1, r2) %>%
  do(indep_exp = t(expected_value(t(indep_params(.$samp[[1]], .$H, .$V)), .$stat[[1]]))[, -((.$H + .$V + 1):(.$H + .$V + .$H*.$V))],
     marg_exp = .$g_theta[[1]][, (.$H + .$V + .$H*.$V + 1):(ncol(.$g_theta[[1]])-.$H*.$V)]) -> exp_vals

exp_vals %>%
  group_by(H, V, N, r1, r2) %>%
  do(data.frame(max_abs_diff = apply(abs(.$indep_exp[[1]] - .$marg_exp[[1]]), 1, max))) %>%
  group_by(H, V, N, r1, r2) %>%
  summarise(max_abs_diff = mean(max_abs_diff)) %>%
  mutate(Hiddens = paste0("paste(n[h], '= ", H, "')"), Visibles = paste0("paste(n[v], '= ", V, "')")) -> exp_vals_summary

#instability
res %>%
  group_by(H, V, N, r1, r2) %>%
  do(data.frame(max = max_Q(t(.$samp[[1]]), .$stat[[1]]),
                min = min_Q(t(.$samp[[1]]), .$stat[[1]]),
                min_max = min_max_Q(.$samp[[1]], .$stat[[1]]))) %>%
  ungroup() %>% 
  group_by(H, V, N, r1, r2) %>%
  mutate(LHS1 = (max - min)/V,
         LHS2 = (max - min_max - H*log(2))/V) %>%
  summarise_each(funs(mean), LHS1, LHS2) -> max_q_summary

res %>%
  group_by(H, V, N, r1, r2) %>%
  do(data.frame(elpr = elpr(.$samp[[1]], .$stat[[1]]))) %>%
  ungroup() %>%
  group_by(H, V, N, r1, r2) %>% 
  summarise(mean_elpr = mean(elpr)) %>%
  mutate(scaled_mean_elpr = mean_elpr/V) -> elpr_summary

convex_hull_summary %>%
  left_join(elpr_summary) %>%
  left_join(exp_vals_summary) -> three_ways

three_ways$Visibles <- factor(three_ways$Visibles, levels=rev(unique(three_ways$Visibles)))

```

- To explore the behavior of the RBM parameters $\boldsymbol \theta$ as it relates to near-degeneracy, instability, and uninterpretability, we consider models of small size.  For $\nh, \nv \in \{1, \dots, 4\}$, we sample `r dim(res$samp[[1]])[1]` values of $\boldsymbol \theta$.
    1. Split $\boldsymbol \theta$ into $\boldsymbol \theta_{interaction}$ and $\boldsymbol \theta_{main}$, in reference to which sufficient statistics the parameters correspond to.
    2. Allow the two types of terms to have varying average magnitudes, $||\boldsymbol \theta_{main} || /(\nh+\nv)$ and $||\boldsymbol \theta_{interaction} || /(\nh*\nv)$.
    3. Average magnitudes vary on a grid between `r min(plot_dat_grid$r1/(plot_dat_grid$H + plot_dat_grid$V))` and `r max(plot_dat_grid$r1/(plot_dat_grid$H + plot_dat_grid$V))` with `r length(unique(plot_dat_grid$r1/(plot_dat_grid$H + plot_dat_grid$V)))` breaks, yielding `r length(unique(plot_dat_grid$r1/(plot_dat_grid$H + plot_dat_grid$V)))^2` grid points.
- Calculate the three metrics of model impropriety, $\boldsymbol \mu(\boldsymbol \theta)$, $\text{ELPR}(\boldsymbol x)/\nv$, and the coordinates of $\left|\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] -   \text{E}\left[\boldsymbol X | \boldsymbol \theta^* \right] \right|$. 
- In the case of *near-degeneracy*, we can go further and classify each model as near-degenerate or "viable" based on the distance of $\boldsymbol \mu(\boldsymbol \theta)$ from the boundary of the convex hull of $\mathcal{T}$

# Simulation results
```{r degen_plots}
three_ways %>%
  mutate(r1 = round(r1/(H + V), 8), r2 = round(r2/(H * V), 8)) %>% 
  ggplot() +
  geom_tile(aes(x = r1, y = r2, fill = frac_degen)) +
  geom_contour(aes(x = r1, y = r2, z = frac_degen), colour = "black", bins = 8) +
  geom_contour(aes(x = r1, y = r2, z = frac_degen), colour = "black", breaks = .05, size = 1.5) +
  geom_abline(aes(intercept = 0, slope = 1), alpha = .5, lty = 2) +
  scale_fill_gradient("Fraction near-\ndegenerate", low = "yellow", high = "red") +
  facet_grid(Visibles~Hiddens, labeller = label_parsed) +
  xlab(expression(paste(paste("||", theta[main], "||", "/(", n[h]," + ", n[v], ")")))) +
  ylab(expression(paste(paste("||", theta[interaction], "||", "/(", n[h], "*", n[v], ")")))) +
  theme(aspect.ratio = 1, legend.position = "bottom") -> p.degen

three_ways %>%
  mutate(r1 = round(r1/(H + V), 8), r2 = round(r2/(H * V), 8)) %>% 
  ggplot() +
  geom_tile(aes(x = r1, y = r2, fill = max_abs_diff)) +
  geom_contour(aes(x = r1, y = r2, z = max_abs_diff), colour = "black", bins = 8) +
  #geom_contour(aes(x = r1, y = r2, z = mean_abs_diff), colour = "black", breaks = .05, size = 1.5) +
  geom_abline(aes(intercept = 0, slope = 1), alpha = .5, lty = 2) +
  scale_fill_gradient("Mean max \nabs. difference", low = "yellow", high = "red", limits = c(0,2)) +
  facet_grid(Visibles~Hiddens, labeller = label_parsed) +
  xlab(expression(paste(paste("||", theta[main], "||", "/(", n[h]," + ", n[v], ")")))) +
  ylab(expression(paste(paste("||", theta[interaction], "||", "/(", n[h], "*", n[v], ")")))) +
  theme(aspect.ratio = 1, legend.position = "bottom") -> p.exp_diff

three_ways %>%
  mutate(r1 = round(r1/(H + V), 8), r2 = round(r2/(H * V), 8)) %>% 
  ggplot() +
  geom_tile(aes(x = r1, y = r2, fill = scaled_mean_elpr)) +
  geom_contour(aes(x = r1, y = r2, z = scaled_mean_elpr), colour = "black", bins = 8) +
  #geom_contour(aes(x = r1, y = r2, z = mean_max_q), colour = "black", breaks = .05, size = 1.5) +
  geom_abline(aes(intercept = 0, slope = 1), alpha = .5, lty = 2) +
  scale_fill_gradient(expression(frac(bar(ELPR(theta)), V)), low = "yellow", high = "red") +
  facet_grid(Visibles~Hiddens, labeller = label_parsed) +
  xlab(expression(paste(paste("||", theta[main], "||", "/(", n[h]," + ", n[v], ")")))) +
  ylab(expression(paste(paste("||", theta[interaction], "||", "/(", n[h], "*", n[v], ")")))) +
  theme(aspect.ratio = 1, legend.position = "bottom") -> p.elpr
```

```{r, fig.show="hold", fig.width=3, out.width='.32\\linewidth', fig.cap="\\label{fig:degen_plots}Results from the numerical experiment, here looking at the fraction of models that were near-degenerate (left), the sample mean value of $\\text{ELPR}(\\boldsymbol \\theta)/\\nv$ (middle), and the sample mean of the maximum component of the absolute difference between the model expectation vector, E$\\left[\\boldsymbol X | \\boldsymbol \\theta\\right]$, and the expectation vector given independence, E$\\left[\\boldsymbol X | \\boldsymbol \\theta^* \\right ]$ (right)."}
p.degen
p.elpr
p.exp_diff
```

# Model fitting

- Typically, fitting a RBM via maximum likelihood (ML) methods will be infeasible mainly due to the intractibility of the normalizing term $\gamma(\boldsymbol \theta)$ in a model of any realistic size
    - Ad hoc methods are used instead, which aim to avoid this problem by using stochastic ML that employ a small number of MCMC draws. 
- Computational concerns are not the only issues with fitting an RBM using ML, the RBM model, has the potential to re-create any distribution for the data.
    - Based on a random sample of visible variables, the model for the cell probabilities that has the highest likelihood over  *all possible model classes* is the empirical distribution, and the parametrization of the RBM model itself ensures that this empirical distribution can be arbitrarily well approximated.
    - Whenever the empirical distribution contains empty cells, fitting steps for the RBM model will aim to chase parameters that necessarily diverge in magnitude in order to zero out the corresponding RBM cell probabilities.

# Bayesian methods

- We consider what might be done in a principled manner, testing on a $\nv = \nh = 4$ case that already stretched the limits of what is computable - in particular we consider Bayes methods. 
- To avoid model impropriety for a fitted RBM, we want to avoid parts of the parameter space $\mathbb{R}^{\nv + \nh + \nv*\nh}$ that lead to *near-degeneracy*, *instability*, and *uninterpretability*.
    - Shrink $\boldsymbol \theta$ toward $\boldsymbol 0$ by specifying priors that place low probability on large values of $||\boldsymbol \theta||$, shrinking $\boldsymbol \theta_{interaction}$ more than $\boldsymbol \theta_{main}$.
  - We considered a test case with $\nv = \nh = 4$ and parameters given in in appendix. This parameter vector was chosen as a sampled value of $\boldsymbol \theta$ that was not near the convex hull of the sufficient statistics for a grid point in figure \ref{fig:degen_plots} with $< 5$\% near-degeneracy. We simulated $n = 5,000$ as a training set and fit the RBM using three Bayes methodologies.

# Fitting methodologies

1. *A "trick" prior (BwTPLV).* Cancel out normalizing term in the likelihood, resulting full conditionals of $\boldsymbol \theta$ are multivariate Normal, $h_j$ are carried along as latent variables.
    \begin{align*}
    \pi(\boldsymbol \theta) \propto \gamma(\boldsymbol \theta)^n \exp\left(-\frac{1}{2C_{1}}\boldsymbol \theta_{main}'\boldsymbol \theta_{main} -\frac{1}{2C_{2}}\boldsymbol \theta_{interaction}'\boldsymbol \theta_{interaction}\right), \vspace{-.75cm}
    \end{align*}
    where $C_{2} < C_{1}$.
    This is the method of @li2014biclustering. 
2. *A truncated Normal prior (BwTNLV).* Independent spherical normal distributions as priors for $\boldsymbol \theta_{main}$ and $\boldsymbol \theta_{interaction}$, with $\sigma_{interaction} < \sigma_{main}$, *truncated* at $3\sigma_{main}$ and $3\sigma_{interaction}$, respectively. Simulation from the posterior using a geometric adaptive MH step [@zhou2014some], $h_j$ are carried along in the MCMC implementation as latent variables. 
3. *A truncated Normal prior and marginalized likelihood (BwTNML).* Marginalize out $\boldsymbol h$ in $f_{\boldsymbol \theta}(\boldsymbol x)$, and use the truncated Normal priors applied to the marginal probabilities for visible variables.

# Hyperparameters

\begin{table}[ht]
\centering
\caption{The values used for the hyperparameters for all three fitting methods. A rule of thumb is imposed which decreases prior variances for the model parameters as the size of the model increases and also shrinks $\boldsymbol \theta_{interaction}$ more than $\boldsymbol \theta_{main}$. The common $C$ defining $C_1$ and $C_2$  in the BwTPLV method is chosen by tuning.}
\label{tab:hyperparam}
\begin{tabular}{|l|c|c|}
\hline 
Method & Hyperparameter & Value \\ 
\hline \hline
\multirow{2}{*}{BwTPLV} & $C_1$ & $\frac{C}{n}\frac{1}{\nh + \nv}$ \\
 & $C_2$ & $\frac{C}{n}\frac{1}{\nh*\nv}$ \\
\hline
\multirow{2}{*}{BwTNLV} & $\sigma^2_{main}$ & $\frac{1}{\nh + \nv}$ \\
 & $\sigma^2_{interaction}$ & $\frac{1}{\nh*\nv}$ \\
\hline
\multirow{2}{*}{BwTNML} & $\sigma^2_{main}$ & $\frac{1}{\nh + \nv}$ \\
 & $\sigma^2_{interaction}$ & $\frac{1}{\nh*\nv}$ \\
\hline
\end{tabular}
\end{table}

# Mixing

The truncated Normal method (2) and the marginalized likelihood method (3) are drawing from the same stationary posterior distribution for images.

```{r models-load}
load("data/sample_images.Rdata")
load("data/params_theta.Rdata")

params_good <- list(main_hidden = sample.params %>% ungroup() %>% filter(!near_hull) %>% select(starts_with("h"), -H) %>% data.matrix(),
                    main_visible = sample.params %>% ungroup() %>% filter(!near_hull) %>% select(starts_with("v"), -V) %>% data.matrix(),
                    interaction = sample.params %>% ungroup() %>% filter(!near_hull) %>% select(starts_with("theta")) %>% data.matrix() %>% matrix(4))

#data and params ------------------------
H <- 4
V <- 4

#marginalized likelihood
load("data/fitted_models_trunc_marginal_full.Rdata")
marginal_bad <- models_bad
marginal_good <- models_good

#load trick prior
load("data/fitted_models_jing_5.8.Rdata")
trick_bad <- models_bad
trick_good <- models_good

#truncated normal
load("data/fitted_models_trunc_full.Rdata")
trunc_bad <- models_bad
trunc_good <- models_good

#rm unneccesary data
rm(models_bad)
rm(models_good)

#computing actual distributions ---------------------
flat_images_good$visibles %>%
  data.frame() %>% 
  rename(v1 = X1, v2 = X2, v3 = X3, v4 = X4) %>%
  group_by(v1, v2, v3, v4) %>%
  summarise(prob = n()/nrow(flat_images_good$visibles)) -> distn_emp

distn_good <- visible_distn(params = params_good)

reshape_sample_distn <- function(model) {
  sample_distn <- model$distn
  dim(sample_distn) <- c(dim(sample_distn)[1]*dim(sample_distn)[2], dim(sample_distn)[3])
  sample_distn %>% data.frame() -> sample_distn
  names(sample_distn) <- names(distn_good)

  sample_distn %>%
    group_by(image_id) %>%
    mutate(iter = 1:n()) -> sample_distn

  return(sample_distn)
}

marginal_sample_good <- reshape_sample_distn(marginal_good)
trick_sample_good <- reshape_sample_distn(trick_good)
trunc_sample_good <- reshape_sample_distn(trunc_good)
```

```{r acf, fig.cap="\\label{fig:acf}The autocorrelation functions (ACF) for the posterior probabilityies of all $2^4 = 16$ possible outcomes for the vector of $4$ visibles assessed at multiple lags for each method with BwTNLV in black and BwTNML in red.", fig.height=4.5, fig.width=8, out.width=".7\\linewidth"}
marginal_sample_good %>% 
  filter(iter > 50) %>%
  group_by(v1, v2, v3, v4, image_id) %>% 
  do(data.frame(acf = acf(.$prob, plot=FALSE)$acf,
     lag = acf(.$prob, plot=FALSE)$lag)) -> marginal_acfs

trunc_sample_good %>% 
  filter(iter > 50) %>%
  group_by(v1, v2, v3, v4, image_id) %>% 
  do(data.frame(acf = acf(.$prob, plot=FALSE)$acf,
                lag = acf(.$prob, plot=FALSE)$lag)) -> trunc_acfs

marginal_acfs %>% ungroup() %>% select(-image_id) %>% rename(marginal = acf) %>%
  left_join(trunc_acfs %>% rename(truncated = acf)) %>%
  ggplot() +
  geom_segment(aes(x=lag, xend=lag, y=0, yend=truncated), colour = "black") +
  geom_segment(aes(x=lag, xend=lag, y=0, yend=marginal), colour = "red") +
  facet_wrap(~image_id) +
  xlab("Lag") +
  ylab("ACF")

```

```{r M_eff, cache=TRUE}
###
### Overlapping block means
###
overlap_mean <- function(data, b) {
  n <- length(data)
  N <- n - b + 1

  blockmeans <- rep(0, N)
  for(i in 1:N) {
    blockmeans[i] <- mean(data[i:(b + i - 1)])
  }

  blockmeans
}

marginal_sample_good %>% 
  group_by(v1, v2, v3, v4, image_id) %>%
  do(means = overlap_mean(.$prob, 2000^(1/3))) -> marginal_means

trunc_sample_good %>% 
  group_by(v1, v2, v3, v4, image_id) %>%
  do(means = overlap_mean(.$prob, 5000^(1/3))) -> trunc_means

marginal_means %>%
  group_by(v1, v2, v3, v4, image_id) %>%
  do(data.frame(C = 2000^(1/3)*var(.$means[[1]]))) %>%
  left_join(marginal_sample_good %>% group_by(v1, v2, v3, v4) %>% summarise(sigma2 = var(prob))) %>%
  mutate(M_eff = sigma2/C,
         model = "BwTNML") %>%
  bind_rows(trunc_means %>%
    group_by(v1, v2, v3, v4, image_id) %>%
    do(data.frame(C = 5000^(1/3)*var(.$means[[1]]))) %>%
    left_join(trunc_sample_good %>% group_by(v1, v2, v3, v4) %>% summarise(sigma2 = var(prob))) %>%
    mutate(M_eff = sigma2/C,
           model = "BwTNLV")) %>%
  select(-C, -sigma2) %>%
  spread(model, M_eff) %>%
  ungroup() %>%
  select(-(v1:v4)) %>%
  gather(Model, `$M_{eff}$`, -image_id) %>%
  mutate(`$M_{eff}$` = 1000*`$M_{eff}$`) %>%
  spread(image_id, `$M_{eff}$`) -> M_eff_table
```

# Effective sample size

We can use an overlapping blockmeans approach to get a crude estimate for the aysmptotic variance of the probability of each image and compare it to an estimate of the asymptotic variance assuming IID draws from the target distribution.

\tiny

```{r m_eff_table}
M_eff_table %>%
  gather(Image, M_eff, -Model) %>%
  spread(Model, M_eff) %>%
  mutate(Outcome = as.numeric(Image)) %>%
  select(Outcome, BwTNLV, BwTNML) %>%
  arrange(Outcome) -> M_eff_tmp

cbind(M_eff_tmp[1:8,], M_eff_tmp[9:16,]) %>%
  kable(digits = 2, caption = paste0("\\label{tab:m_eff}The effective sample sizes for a chain of length $M = 1000$ regarding all $16$ probabilities for possible vector outcomes of visibles. BwTNLV would require at least $", round(min(M_eff_tmp[, "BwTNML"])/min(M_eff_tmp[, "BwTNLV"]), 1), "$ times as many MCMC iterations to achieve the same amount of effective information about the posterior distribution."))
```

# Posterior distributions of images
```{r posterior-dsn}
marginal_sample_good %>% rename(marginal = prob) %>% filter(iter > 50) %>%
  left_join(trick_sample_good %>% ungroup() %>% select(-image_id) %>% rename(trick = prob)) %>%
  ungroup() %>%
  mutate(image_id = paste0("image_", image_id)) %>% 
  gather(method, prob, trick, marginal) %>%
  spread(image_id, prob) -> all_statistics

all_statistics %>%
  gather(statistic, value, -iter, -method, -starts_with("v")) %>%
  filter(grepl("image", statistic) & !is.na(value)) %>%
  separate(statistic, into = c("junk", "statistic")) %>%
  mutate(statistic = factor(statistic, labels = paste("Image", 1:length(unique(statistic))))) %>%
  left_join(distn_emp %>% rename(prob_emp = prob) %>% right_join(distn_good %>% select(-image_id))) %>%
  ggplot() +
  geom_density(aes(value, y=..scaled.., colour = method, fill = method), alpha = .2) +
  geom_vline(aes(xintercept = prob)) +
  geom_vline(aes(xintercept = prob_emp), colour = "red") +
  facet_wrap(~statistic, scales = "free_x") + 
  ylab("Scaled Posterior Density") + xlab("Probability of Image") +
  scale_colour_discrete("Method", labels = c("Marginalized likelihood", "Trick prior")) +
  scale_fill_discrete("Method", labels = c("Marginalized likelihood", "Trick prior")) +
  theme(legend.position = "bottom") -> p.models
```

```{r fitting_plot, fig.cap="\\label{fig:fitting_plot}Posterior probabilities of $16 = 2^4$ possible realizations of $4$ visibles using two of the three Bayesian fitting techniques, BwTPLV and BwTNML. Black lines show true probabilities of each vector of visibles based on the parameters used to generate the training data while red lines show the empirical distribution.", fig.height=4.9}
p.models
```

# Wrapping up

- While RBMs are thought to be useful for classification, in the context of generative statistical models, RBMs are a poor fit due to *near-degeneracy*, *S-instability*, and w*uninterpretability*.
- Rigorous fitting methodology is difficult due to the dimension of the parameter space coupled with the size of the latent variable space.
- For a RBM model with enough hidden variables, any distribution for the visibles can be approximated arbitrarily well [@le2008representational; @montufar2011refinements; and @montufar2011expressive]. 
    - The empirical distribution of a training set of vectors of visibles is the best fitting model for observed cell data. 
    - There can be no "smoothed distribution" achieved in fitting a RBM model of sufficient size with a rigorous likelihood-based method.

We are skeptical that any model built using these structures (like a deep Boltzmann machine) can achieve useful prediction or inference in a principled way without limiting the flexibility of the fitted model.

# Future work

---- 
\begin{center}
\Huge
Appendices
\end{center}

# Appendix: Proof of Result \ref{prop:instab}
\scriptsize

We prove the contrapositive. Suppose that $\Delta(\boldsymbol \theta_\nv) \le C$ holds for some $C > 0$. Under the RBM model for visibles, $P_{\boldsymbol \theta_\nv}(\boldsymbol v) > 0$ holds for each outcome $\boldsymbol v \in \mathcal{C}^\nv$. Let $\boldsymbol v_{min} \equiv \argmin\limits_{\boldsymbol v \in \mathcal{C}^\nv}P_{\boldsymbol \theta_\nv}(\boldsymbol v)$ and $\boldsymbol v_{max} \equiv \argmax\limits_{\boldsymbol v \in \mathcal{C}^V\nv}P_{\boldsymbol \theta_\nv}(\boldsymbol v)$. Note there exists a sequence $\boldsymbol v_{min} \equiv \boldsymbol v_0, \boldsymbol v_1, \dots, \boldsymbol v_k \equiv \boldsymbol v_{max}$ in $\mathcal{C}^\nv$ of component-wise switches to move from $\boldsymbol v_{min}$ to $\boldsymbol v_{max}$ in the sample space (i.e. $\boldsymbol v_i, \boldsymbol v_{i + 1} \in \mathcal{C}^\nv$ differ by exactly $1$ component for $i = 0, \dots, k$) for some integer $k \in \{0, 1, \dots, \nv\}$. Then
\begin{align*}
\log\left[\frac{P_{\boldsymbol \theta_\nv}(\boldsymbol v_{max})}{P_{\boldsymbol \theta_\nv}(\boldsymbol v_{min})}\right] &= \left|\sum\limits_{i = 1}^k\log\left(\frac{P_{\boldsymbol \theta_\nv}(\boldsymbol v_i)}{P_{\boldsymbol \theta_\nv}(\boldsymbol v_{i-1})}\right)\right| \\
& \le \sum\limits_{i = 1}^k\left|\log\left(\frac{P_{\boldsymbol \theta_\nv}(\boldsymbol v_i)}{P_{\boldsymbol \theta_\nv}(\boldsymbol v_{i-1})}\right)\right| \\
& \le k \Delta(\boldsymbol \theta_\nv) \le \nv C
\end{align*}
using $k \le \nv$ and $\Delta(\boldsymbol \theta_\nv) \le C$.$\Box$

# Appendix: Proof of Result \ref{prop:degen}

\scriptsize

Define $\boldsymbol v_{max}$ and $\boldsymbol v_{min}$ as in the proof of Proposition \ref{prop:instab}. Fix $0 < \epsilon < 1$ Then, $\boldsymbol v_{max} \in M_{\epsilon, \boldsymbol \theta_\nv}$, so $P_{\boldsymbol \theta_\nv}(M_{\epsilon, \boldsymbol \theta}) \ge P_{\boldsymbol \theta_\nv}(\boldsymbol v_{max})$. If $\boldsymbol v \in \mathcal{C}^\nv \setminus M_{\epsilon, \boldsymbol \theta_\nv}$, then by definition $P_{\boldsymbol \theta_\nv}(\boldsymbol v) \le [P_{\boldsymbol \theta_\nv}(\boldsymbol v_{max})]^{1-\epsilon}[P_{\boldsymbol \theta_\nv}(\boldsymbol v_{min})]^{\epsilon}$ holds so that
\begin{align*}
1-P_{\boldsymbol \theta_\nv}(M_{\epsilon, \boldsymbol \theta_\nv}) &= P_{\boldsymbol \theta_\nv}(M_{\epsilon, \boldsymbol \theta_\nv}^C) \\
& = \sum\limits_{\boldsymbol v \in \mathcal{C}^\nv \setminus M_{\epsilon, \boldsymbol \theta_\nv}}P_{\boldsymbol \theta_\nv}(\boldsymbol v) \\
& \le (2^\nv)[P_{\boldsymbol \theta_\nv}(\boldsymbol v_{max})]^{1-\epsilon}[P_{\boldsymbol \theta_\nv}(\boldsymbol v_{min})]^{\epsilon}
\end{align*}
Then,
\begin{align*}
\frac{1}{\nv}\log\left[\frac{P_{\boldsymbol \theta_\nv}(M_{\epsilon, \boldsymbol \theta_\nv})}{1-P_{\boldsymbol \theta_\nv}(M_{\epsilon, \boldsymbol \theta_\nv})}\right] & \ge \frac{1}{\nv} \log\left[\frac{P_{\boldsymbol\theta_\nv}(\boldsymbol v_{max})}{(2^\nv)[P_{\boldsymbol \theta_\nv}(\boldsymbol v_{max})]^{1-\epsilon}[P_{\boldsymbol \theta_\nv}(\boldsymbol v_{min})]^{\epsilon}}\right] \\
&= \frac{\epsilon}{\nv} \log\left[\frac{P_{\boldsymbol \theta_\nv}(\boldsymbol v_{max})}{P_{\boldsymbol \theta_\nv}(\boldsymbol v_{min})}\right] - \log 2 \rightarrow \infty
\end{align*}
as $\nv \rightarrow \infty$ by the definition of an unstable RBM model.$\Box$

# Appendix: Parameters used

\scriptsize

```{r param-table}
sample.params %>% 
  ungroup %>% 
  filter(!near_hull) %>% select(starts_with("v"), starts_with("h"), starts_with("theta"), -H, -V) %>%
  rename(`$\\theta_{v1}$` = v1,
         `$\\theta_{v2}$` = v2,
         `$\\theta_{v3}$` = v3,
         `$\\theta_{v4}$` = v4,
         `$\\theta_{h1}$` = h1,
         `$\\theta_{h2}$` = h2,
         `$\\theta_{h3}$` = h3,
         `$\\theta_{h4}$` = h4,
         `$\\theta_{11}$` = theta11,
         `$\\theta_{12}$` = theta12,
         `$\\theta_{13}$` = theta13,
         `$\\theta_{14}$` = theta14,
         `$\\theta_{21}$` = theta21,
         `$\\theta_{22}$` = theta22,
         `$\\theta_{23}$` = theta23,
         `$\\theta_{24}$` = theta24,
         `$\\theta_{31}$` = theta31,
         `$\\theta_{32}$` = theta32,
         `$\\theta_{33}$` = theta33,
         `$\\theta_{34}$` = theta34,
         `$\\theta_{41}$` = theta41,
         `$\\theta_{42}$` = theta42,
         `$\\theta_{43}$` = theta43,
         `$\\theta_{44}$` = theta44) %>%
  gather("Parameter", "Value") -> tbl

cbind(tbl[1:8,], tbl[9:16,], tbl[17:24,]) %>%
  kable(caption = "\\label{tab:theta}Parameters used to fit a test case with $\\nv = \\nh = 4$. This parameter vector was chosen as a sampled value of $\\boldsymbol \\theta$ that was not near the convex hull of the sufficient statistics for a grid point in figure \\ref{fig:degen_plots} with $< 5$\\% near-degeneracy.")
```

# References {.allowframebreaks}
\tiny

### 2016-12-11 writing/appendix/appendix-center.Rmd
slides start

--- 2016-11-20 writing/appendix/appendix-center.Rmd
@@
     number_sections: yes
+    template: ../latex/technometrics_article.tex
+fontsize: 12pt
+blind: false
 ---

### 2016-12-11 presentations/amherst/slides.Rmd
remove slide

@@
 $$ 
-have a mean vector $\boldsymbol \mu(\boldsymbol \theta)$ close to the boundary of the convex hull of $\mathcal{T} = \{\boldsymbol t(x): x \in \mathcal{C}^{\nh + \nv}\}$ [@handcock2003assessing], where $\boldsymbol t(\boldsymbol x) = \{v_1, \dots, v_{\nv}, h_1, \dots, h_{\nh}, v_1 h_1, \dots, v_{\nv} h_{\nh} \}$ and the mean parameterization on the model parameters, $\boldsymbol \mu(\boldsymbol \theta)$ is
-
----
-
-\begin{align*}
-\boldsymbol \mu(\boldsymbol \theta) &= \text{E}_{\boldsymbol \theta} \boldsymbol t(\boldsymbol X)  \\
-&= \sum\limits_{\boldsymbol x \in \mathcal{C}^{\nh + \nv}} \left\{ \boldsymbol t(\boldsymbol x)\dfrac{\exp\left(\sum\limits_{i = 1}^{\nv} \sum\limits_{j=1}^{\nh} \theta_{ij} v_i h_j + \sum\limits_{i = 1}^{\nv}\theta_{v_i} v_i + \sum\limits_{j = 1}^{\nh}\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{C}^{\nh + \nv}}\exp\left(\sum\limits_{i = 1}^{\nv} \sum\limits_{j=1}^{\nh} \theta_{ij} v_i h_j + \sum\limits_{i = 1}^{\nv}\theta_{v_i} v_i + \sum\limits_{j = 1}^{\nh}\theta_{h_j} h_j\right)}\right\}.
-\\
-\end{align*}
+have a mean vector $\boldsymbol \mu(\boldsymbol \theta)$ close to the boundary of the convex hull of $\mathcal{T} = \{\boldsymbol t(x): x \in \mathcal{C}^{\nh + \nv}\}$ [@handcock2003assessing], where $\boldsymbol t(\boldsymbol x) = \{v_1, \dots, v_{\nv}, h_1, \dots, h_{\nh}, v_1 h_1, \dots, v_{\nv} h_{\nh} \}$ and the mean parameterization on the model parameters, $\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta} \boldsymbol t(\boldsymbol X)$.
 

### 2016-12-11 presentations/amherst/slides.Rmd
add contct info

@@
 shortname: Kaplan, et al.
-institute: ajkaplan@iastate.edu
-date: "December 20, 2016"
+institute: |
+    | Iowa State University
+    | ajkaplan@iastate.edu
+shortinstitute: ajkaplan@iastate.edu
+date: |
+  | December 20, 2016
+  | 
+  |
+  | Slides available at <http://bitly.com/kaplan-amherst>
+shortdate: "December 20, 2016"
 output: 
@@
 
-- While RBMs are thought to be useful for classification, in the context of generative statistical models, RBMs are a poor fit due to *near-degeneracy*, *S-instability*, and w*uninterpretability*.
+- While RBMs are thought to be useful for classification, in the context of generative statistical models, RBMs are a poor fit due to *near-degeneracy*, *S-instability*, and *uninterpretability*.
 - Rigorous fitting methodology is difficult due to the dimension of the parameter space coupled with the size of the latent variable space.
@@
 # Future work
+
+# Thank you
+
+* Slides -- <http://bitly.com/kaplan-amherst>
+
+* Contact
+    * Email -- <ajkaplan@iastate.edu>
+    * Twitter -- <http://twitter.com/andeekaplan>
+    * GitHub -- <http://github.com/andeek>
 

### 2016-12-11 presentations/amherst/slides.Rmd
fix link

@@
   |
-  | Slides available at <http://bitly.com/kaplan-amherst>
+  | Slides available at <http://bit.ly/kaplan-umass>
 shortdate: "December 20, 2016"
@@
 
-* Slides -- <http://bitly.com/kaplan-amherst>
+* Slides -- <http://bit.ly/kaplan-umass>
 

### 2016-12-16 presentations/amherst/slides.Rmd
edits to slides, starting to cut

@@
 
-A Restricted Boltzman Machine (RBM) is an undirected probabilistic graphical model (for discrete or continuous random variables) with two layers, one hidden and one visible, with conditional independence within a layer [@smolensky1986information].
+- A restricted Boltzman machine (RBM) is an undirected probabilistic graphical model with
+    1. Discrete or continuous random variables 
+    2. two layers - one hidden and one visible
+    3. conditional independence within a layer [@smolensky1986information]
 
@@
 
-Typically used for image classification. Each image pixel is a node in the visible layer. The output creates features, which are passed to a supervised learning algorithm.
+- Supervised learning, specifically image classification
 
 \begin{figure}
-\includegraphics[width=\linewidth]{images/visibles_image.png}
+\includegraphics[width=\linewidth]{images/visibles_image.pdf}
 \caption{Image classification using a RBM.
@@
 # Joint distribution
-Let $\boldsymbol x = (h_1, \dots, h_{\nh}, v_1,\dots,v_{\nv})$ represent the states of the visible and hidden nodes in an RBM.
-Each single "binary" random variable, visible or hidden, will take its values in a common coding set $\mathcal{C}$, with two possibilities for the coding set $\mathcal{C}=\{0,1\}$ or $\mathcal{C}=\{-1,1\}$.
-A parametric form for probabilities corresponding to a potential vector of states of each node
+
+- $\boldsymbol x = (h_1, \dots, h_{\nh}, v_1,\dots,v_{\nv})$ represents the states of the visible and hidden nodes in a RBM 
+- Each single "binary" random variable, visible or hidden, will take its values in a common coding set $\mathcal{C}$
+    - Two possibilities for the coding set $\mathcal{C}=\{0,1\}$ or $\mathcal{C}=\{-1,1\}$. 
+- A parametric form for probabilities
 
@@
 
-$$\gamma(\boldsymbol \theta) = \sum\limits_{\boldsymbol x \in \mathcal{C}^{\nh + \nv}}\exp\left(\sum\limits_{i = 1}^{\nv} \sum\limits_{j=1}^{\nh} \theta_{ij} v_i h_j + \sum\limits_{i = 1}^{\nv}\theta_{v_i} v_i + \sum\limits_{j = 1}^{\nh}\theta_{h_j} h_j\right)$$.
+    $$\gamma(\boldsymbol \theta) = \sum\limits_{\boldsymbol x \in \mathcal{C}^{\nh + \nv}}\exp\left(\sum\limits_{i = 1}^{\nv} \sum\limits_{j=1}^{\nh} \theta_{ij} v_i h_j + \sum\limits_{i = 1}^{\nv}\theta_{v_i} v_i + \sum\limits_{j = 1}^{\nh}\theta_{h_j} h_j\right)$$
 
@@
 \begin{itemize}
-\item By stacking layers of RBMs in a deep architecture, proponents of the models claim the ability to learn "internal representations that become increasingly complex, which is considered to be a promising way of solving object and speech recognition problems" (Salakhutdinov and Hinton 2009, pp.
+\item Stacking layers of RBMs in a deep architecture
+\item Proponents claim the ability to learn "internal representations that become increasingly complex, which is considered to be a promising way of solving object and speech recognition problems" (Salakhutdinov and Hinton 2009, pp.
 450).
-\item Stacking is achieved by treating a hidden layer of one RBM as the visible layer in a second RBM, etc.
+\item Treating a hidden layer of one RBM as the visible layer in a second RBM, etc.
 \end{itemize}
@@
 
-The model properties are largely unexplored in the literature and the commonly cited fitting methodology remains heuristic-based and abstruse [@hinton2006fast].
+1. The model properties are largely unexplored in the literature and 
+2. The commonly cited fitting methodology remains heuristic-based and abstruse [@hinton2006fast]
 
@@
 
-1. provide steps toward a thorough understanding of the model class and its properties from the perspective of statistical theory, and 
-2. explore the possibility of a rigorous fitting methodology.
+1. Provide steps toward a thorough understanding of the model class and its properties from the perspective of statistical theory, and 
+2. Explore the possibility of a rigorous fitting methodology
 
@@
 
-The highly flexible nature of a RBM (having as it does $\nh + \nv + \nh*\nv$ parameters) makes at least three kinds of potential model impropriety of concern, *degeneracy*, *instability*, and *uninterpretability*.
+The highly flexible nature of a RBM ($\nh + \nv + \nh*\nv$ parameters) makes at least three kinds of potential model impropriety of concern
+
+1. *degeneracy*
+2. *instability*, and 
+3. *uninterpretability*
 
@@
 
-- fail to generate data with realistic variability and thus an unsatisfactory conceptualization of the data generation process. 
-- exhibit instability in the parameter space.
-
-Such model impropriety issues have been documented in in RBMs [@li2014biclustering], as well as other deep architectures [@szegedy2013intriguing; @nguyen2014deep].
+- fail to generate data with realistic variability and thus an unsatisfactory conceptualization of the data generation process [@li2014biclustering] 
+- exhibit instability in the parameter space [@szegedy2013intriguing; @nguyen2014deep]
 
@@
     - Ad hoc methods are used instead, which aim to avoid this problem by using stochastic ML that employ a small number of MCMC draws. 
-- Computational concerns are not the only issues with fitting an RBM using ML, the RBM model, has the potential to re-create any distribution for the data.
+- Computational concerns are not the only issues with fitting a RBM using ML, the RBM model, has the potential to re-create any distribution for the data.
     - Based on a random sample of visible variables, the model for the cell probabilities that has the highest likelihood over  *all possible model classes* is the empirical distribution, and the parametrization of the RBM model itself ensures that this empirical distribution can be arbitrarily well approximated.

### 2016-12-16 presentations/amherst/slides.Rmd
some more cutting

@@
 
-- A restricted Boltzman machine (RBM) is an undirected probabilistic graphical model with
+A restricted Boltzman machine (RBM) is an undirected probabilistic graphical model with
+
 1. Discrete or continuous random variables 
@@
 
-RBM models exhibit *near-degeneracy* when random variables in the neg-potential function 
+RBM models exhibit *near-degeneracy* when random variables in 
 $$Q_{\boldsymbol \theta}(\boldsymbol x) = \sum\limits_{i = 1}^{\nv} \sum\limits_{j=1}^{\nh} \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j,
 $$ 
-have a mean vector $\boldsymbol \mu(\boldsymbol \theta)$ close to the boundary of the convex hull of $\mathcal{T} = \{\boldsymbol t(x): x \in \mathcal{C}^{\nh + \nv}\}$ [@handcock2003assessing], where $\boldsymbol t(\boldsymbol x) = \{v_1, \dots, v_{\nv}, h_1, \dots, h_{\nh}, v_1 h_1, \dots, v_{\nv} h_{\nh} \}$ and the mean parameterization on the model parameters, $\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta} \boldsymbol t(\boldsymbol X)$.
+have a mean vector $\boldsymbol \mu(\boldsymbol \theta)$ close to the boundary of the convex hull of $\mathcal{T} = \{\boldsymbol t(x): x \in \mathcal{C}^{\nh + \nv}\}$ [@handcock2003assessing], where $$\boldsymbol t(\boldsymbol x) = \{v_1, \dots, v_{\nv}, h_1, \dots, h_{\nh}, v_1 h_1, \dots, v_{\nv} h_{\nh} \}$$ and $$\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta} \boldsymbol t(\boldsymbol X)$$.
 
@@
 
-@schweinberger2011instability introduced a concept of model deficiency related to *instability* considering only a class of exponential families of distributions.
-
-To quantify *instability* in the RBM, it is useful to consider how a data model might be expanded to incorporate more visibles.
-For this, it becomes necessary to grow the number of model parameters in a sequence $\boldsymbol \theta_{\nv} \in \mathbb{R}^{\nv + \nh + \nv*\nh}, \nv \ge 1$ (one may also arbitrarily expand the number of hidden variables used).
+- Concept of model deficiency related to *instability* for a class of exponential families of distributions [@schweinberger2011instability]
+- For the RBM, consider how a data model might be expanded to incorporate more visibles
+    - Necessary to grow the number of model parameters in a sequence $\boldsymbol \theta_{\nv} \in \mathbb{R}^{\nv + \nh + \nv*\nh}, \nv \ge 1$ 
+    - May also arbitrarily expand the number of hidden variables used.
 
@@
 
-S-unstable RBM model sequences are undesirable for several reasons. Again, one is that small changes in data can lead to overly-sensitive changes in probability.
+S-unstable RBM model sequences are undesirable for several reasons - small changes in data can lead to overly-sensitive changes in probability.
 
@@
 
-Consider, for example, the biggest log-probability ratio for a one-pixel (one component) change in data outcomes (visibles).
+Consider the biggest log-probability ratio for a one-pixel (one component) change in data outcomes (visibles)
 
@@
 \label{prop:instab}
-Let $C > 0$ and $\nv \ge 1$. If $\frac{1}{\nv}\text{ELPR}(\boldsymbol \theta_\nv) > C$, then $\Delta(\boldsymbol \theta_\nv) > C$.
+Let $c > 0$ and let $\text{ELPR}(\boldsymbol \theta_\nv)$ be as in (\ref{eqn:elpr}) for an integer $\nv \ge 1$.
+If $\frac{1}{\nv}\text{ELPR}(\boldsymbol \theta_\nv) > c$, then $\Delta(\boldsymbol \theta_\nv) > c$.
 \end{result}
@@
 
-Unstable RBM model sequences are connected to degenerate model sequences (placing all probability on a small portion of their sample spaces). Define a modal set
+Define an arbitrary modal set of possible outcomes (i.e. set of highest probability outcomes) for a given $0 < \epsilon < 1$ as
 
@@
 \begin{align*}
-M_{\epsilon, \boldsymbol \theta_\nv} &\equiv \left\{\boldsymbol v \in \mathcal{C}^\nv: \log P_{\boldsymbol \theta_\nv}(\boldsymbol v)  > (1-\epsilon)\max\limits_{\boldsymbol v^*}P_{\boldsymbol \theta_\nv}(\boldsymbol v^*) + \epsilon\min\limits_{\boldsymbol v^*}P_{\boldsymbol \theta_\nv}(\boldsymbol v^*) \right\}
+M_{\epsilon, \boldsymbol \theta_\nv} \equiv \left\{\boldsymbol v \in \mathcal{C}^\nv: \log P_{\boldsymbol \theta_\nv}(\boldsymbol v) > (1-\epsilon)\max\limits_{\boldsymbol v^*}P_{\boldsymbol \theta_\nv}(\boldsymbol v^*) + \epsilon\min\limits_{\boldsymbol v^*}P_{\boldsymbol \theta_\nv}(\boldsymbol v^*) \right\}
 \end{align*}
-
 \normalsize
-of possible outcomes, for a given $0 < \epsilon < 1$.
 
@@
 
-As a consequence of unstable models, all probability will stack up on mode sets or potentially those few outcomes with the highest probability.
-Proofs of results \ref{prop:instab}-\ref{prop:degen} are provided later.
+- All probability will stack up on mode sets or potentially those few outcomes with the highest probability 
+- Proofs of results \ref{prop:instab}-\ref{prop:degen} can be found in [@kaplan2016note]
 
@@
 
-A measure of this is the magnitude of the difference between model expectations, E$\left[\boldsymbol X | \boldsymbol \theta\right]$, and expectations given independence, E$\left[\boldsymbol X | \boldsymbol \theta^* \right]$, where $\boldsymbol \theta^*$ is defined to equal $\boldsymbol \theta$ with all $\theta_{ij} = 0$ for $i = 1, \dots, \nv, j = 1, \dots, \nh$.
-
-Using this, it is possible to investigate what parameters lead to uninterpretability in a model versus those that guarantee interpretability.
-
-If $|\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] - \text{E}\left[\boldsymbol X | \boldsymbol \theta^*\right]|$ is large then the RBM with parameter vector $\boldsymbol \theta$ is *uninterpetable*.
+- Measure: the magnitude of the difference between
+    1. Model expectations, E$\left[\boldsymbol X | \boldsymbol \theta\right]$, and 
+    2.
+Expectations given independence, E$\left[\boldsymbol X | \boldsymbol \theta^* \right]$, where $\boldsymbol \theta^*$ matches $\boldsymbol \theta$ for all main effects but otherwise has $\theta_{ij} = 0$ for $i = 1, \dots, \nv, j = 1, \dots, \nh$
+- If $|\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] - \text{E}\left[\boldsymbol X | \boldsymbol \theta^*\right]|$ is large then the RBM with parameter vector $\boldsymbol \theta$ is *uninterpetable*
 
@@
 \includegraphics[width=.35\linewidth]{images/toyhull_top_2.png}
-\caption{The convex hull of the "statistic space" $\mathcal{T} = \{(v_1, h_1, v_1 h_1): v_1, h_1 \in \mathcal{C}\}$ in three dimensions for the toy RBM with one visible and one hidden node for $\mathcal{C} = \{0,1\}$ (left) and $\mathcal{C} = \{-1,1\}$ (right) data encoding.
-The convex hull of $\mathcal{T} \subset \mathcal{C}^3$ does not fill the unit cube $[0,1]^3$ (left), but does better with $[-1,1]^3$ (right).}
+\caption{The convex hull of the "statistic space" $\mathcal{T} = \{(v_1, h_1, v_1 h_1): v_1, h_1 \in \mathcal{C}\}$ in three dimensions for the toy RBM with one visible and one hidden node for $\mathcal{C} = \{0,1\}$ (left) and $\mathcal{C} = \{-1,1\}$ (right) data encoding.}
 \end{figure}
+
+The convex hull of $\mathcal{T} \subset \mathcal{C}^3$ does not fill the unit cube $[0,1]^3$ (left), but does better with $[-1,1]^3$ (right).
 
@@
 
-- To explore the behavior of the RBM parameters $\boldsymbol \theta$ as it relates to near-degeneracy, instability, and uninterpretability, we consider models of small size.
-For $\nh, \nv \in \{1, \dots, 4\}$, we sample `r dim(res$samp[[1]])[1]` values of $\boldsymbol \theta$.
-    1. Split $\boldsymbol \theta$ into $\boldsymbol \theta_{interaction}$ and $\boldsymbol \theta_{main}$, in reference to which sufficient statistics the parameters correspond to.
-    2. Allow the two types of terms to have varying average magnitudes, $||\boldsymbol \theta_{main} || /(\nh+\nv)$ and $||\boldsymbol \theta_{interaction} || /(\nh*\nv)$.
+- To explore the behavior of the RBM parameters $\boldsymbol \theta$ as it relates to *near-degeneracy*, *instability*, and *uninterpretability*, consider models of small size  
+- For $\nh, \nv \in \{1, \dots, 4\}$, sample `r dim(res$samp[[1]])[1]` values of $\boldsymbol \theta$
+    1. Split $\boldsymbol \theta$ into $\boldsymbol \theta_{interaction}$ and $\boldsymbol \theta_{main}$, in reference to which sufficient statistics the parameters correspond to
+    2. Allow the two types of terms to have varying average magnitudes, $||\boldsymbol \theta_{main} || /(\nh+\nv)$ and $||\boldsymbol \theta_{interaction} || /(\nh*\nv)$
     3.
-Average magnitudes vary on a grid between `r min(plot_dat_grid$r1/(plot_dat_grid$H + plot_dat_grid$V))` and `r max(plot_dat_grid$r1/(plot_dat_grid$H + plot_dat_grid$V))` with `r length(unique(plot_dat_grid$r1/(plot_dat_grid$H + plot_dat_grid$V)))` breaks, yielding `r length(unique(plot_dat_grid$r1/(plot_dat_grid$H + plot_dat_grid$V)))^2` grid points.
-- Calculate the three metrics of model impropriety, $\boldsymbol \mu(\boldsymbol \theta)$, $\text{ELPR}(\boldsymbol x)/\nv$, and the coordinates of $\left|\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] -   \text{E}\left[\boldsymbol X | \boldsymbol \theta^* \right] \right|$.
-- In the case of *near-degeneracy*, we can go further and classify each model as near-degenerate or "viable" based on the distance of $\boldsymbol \mu(\boldsymbol \theta)$ from the boundary of the convex hull of $\mathcal{T}$
+Average magnitudes vary on a grid between `r min(plot_dat_grid$r1/(plot_dat_grid$H + plot_dat_grid$V))` and `r max(plot_dat_grid$r1/(plot_dat_grid$H + plot_dat_grid$V))` with `r length(unique(plot_dat_grid$r1/(plot_dat_grid$H + plot_dat_grid$V)))` breaks, yielding `r length(unique(plot_dat_grid$r1/(plot_dat_grid$H + plot_dat_grid$V)))^2` grid points
+- Calculate metrics of model impropriety, $\boldsymbol \mu(\boldsymbol \theta)$, $\text{ELPR}(\boldsymbol \theta)/\nv$, and the coordinates of $\left|\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] - \text{E}\left[\boldsymbol X | \boldsymbol \theta^* \right] \right|$.
+- In the case of *near-degeneracy*, classify each model as near-degenerate or "viable" based on the distance of $\boldsymbol \mu(\boldsymbol \theta)$ from the boundary of the convex hull of $\mathcal{T}$
 
@@
 
-```{r, fig.show="hold", fig.width=3, out.width='.32\\linewidth', fig.cap="\\label{fig:degen_plots}Results from the numerical experiment, here looking at the fraction of models that were near-degenerate (left), the sample mean value of $\\text{ELPR}(\\boldsymbol \\theta)/\\nv$ (middle), and the sample mean of the maximum component of the absolute difference between the model expectation vector, E$\\left[\\boldsymbol X | \\boldsymbol \\theta\\right]$, and the expectation vector given independence, E$\\left[\\boldsymbol X | \\boldsymbol \\theta^* \\right ]$ (right)."}
+```{r, fig.show="hold", fig.width=3, out.width='.32\\linewidth', fig.cap="\\label{fig:degen_plots}The fraction of models that were near-degenerate (left), the sample mean value of $\\text{ELPR}(\\boldsymbol \\theta)/\\nv$ (middle), and the sample mean of the maximum component of the absolute difference between the model expectation vector, E$\\left[\\boldsymbol X | \\boldsymbol \\theta\\right]$, and the expectation vector given independence, E$\\left[\\boldsymbol X | \\boldsymbol \\theta^* \\right ]$ (right)."}
 p.degen
@@
 
-- Typically, fitting a RBM via maximum likelihood (ML) methods will be infeasible mainly due to the intractibility of the normalizing term $\gamma(\boldsymbol \theta)$ in a model of any realistic size
-    - Ad hoc methods are used instead, which aim to avoid this problem by using stochastic ML that employ a small number of MCMC draws. 
-- Computational concerns are not the only issues with fitting a RBM using ML, the RBM model, has the potential to re-create any distribution for the data.
-    - Based on a random sample of visible variables, the model for the cell probabilities that has the highest likelihood over  *all possible model classes* is the empirical distribution, and the parametrization of the RBM model itself ensures that this empirical distribution can be arbitrarily well approximated.
-    - Whenever the empirical distribution contains empty cells, fitting steps for the RBM model will aim to chase parameters that necessarily diverge in magnitude in order to zero out the corresponding RBM cell probabilities.
+1. Computational concerns
+    - Fitting a RBM via maximum likelihood (ML) methods infeasible due to the intractibility of the normalizing term $\gamma(\boldsymbol \theta)$
+        - Ad hoc methods are used, aim to avoid this problem by using stochastic ML 
+        - employ a small number of MCMC draws to approximate $\gamma(\boldsymbol \theta)$ 
+2. With $\nh$ large enough 
+    - Potential to re-create any distribution for the data [@le2008representational; @montufar2011refinements; and @montufar2011expressive]
+        - The model for the cell probabilities that has the highest likelihood over  *all possible model classes* is the empirical distribution
+        - The RBM model ensures that this empirical distribution can be arbitrarily well approximated
+    - When empirical distribution contains empty cells, fitting will chase parameters to $\infty$ in order to zero out the corresponding RBM cell probabilities
 
@@
 
-- We consider what might be done in a principled manner, testing on a $\nv = \nh = 4$ case that already stretched the limits of what is computable - in particular we consider Bayes methods. 
-- To avoid model impropriety for a fitted RBM, we want to avoid parts of the parameter space $\mathbb{R}^{\nv + \nh + \nv*\nh}$ that lead to *near-degeneracy*, *instability*, and *uninterpretability*.
-    - Shrink $\boldsymbol \theta$ toward $\boldsymbol 0$ by specifying priors that place low probability on large values of $||\boldsymbol \theta||$, shrinking $\boldsymbol \theta_{interaction}$ more than $\boldsymbol \theta_{main}$.
-  - We considered a test case with $\nv = \nh = 4$ and parameters given in in appendix.
-This parameter vector was chosen as a sampled value of $\boldsymbol \theta$ that was not near the convex hull of the sufficient statistics for a grid point in figure \ref{fig:degen_plots} with $< 5$\% near-degeneracy.
-We simulated $n = 5,000$ as a training set and fit the RBM using three Bayes methodologies.
+- Consider what might be done in a principled manner, testing on $\nv = \nh = 4$ 
+- To avoid model impropriety, we want to avoid parts of the parameter space $\mathbb{R}^{\nv + \nh + \nv*\nh}$ that lead to *near-degeneracy*, *instability*, and *uninterpretability*.
+    - Shrink $\boldsymbol \theta$ toward $\boldsymbol 0$ 
+        1. Specify priors that place low probability on large values of $||\boldsymbol \theta||$
+        2. Shrink $\boldsymbol \theta_{interaction}$ more than $\boldsymbol \theta_{main}$
+- Consider a test case with $\nv = \nh = 4$ and parameters given in in appendix 
+    - $\boldsymbol \theta$ chosen as a sampled value from a grid point in figure \ref{fig:degen_plots} with $< 5$\% near-degeneracy (not near the convex hull of the sufficient statistics) 
+    - simulate $n = 5,000$ as a training set and fit the RBM using three Bayes methodologies
 
@@
 
-1. *A "trick" prior (BwTPLV).* Cancel out normalizing term in the likelihood, resulting full conditionals of $\boldsymbol \theta$ are multivariate Normal, $h_j$ are carried along as latent variables.
+1. *A "trick" prior (BwTPLV).* Cancel out normalizing term in the likelihood, resulting full conditionals of $\boldsymbol \theta$ are multivariate Normal, $h_j$ are carried along as latent variables
     \begin{align*}
@@
     \end{align*}
-    where $C_{2} < C_{1}$.
-    This is the method of @li2014biclustering. 
+    where $C_{2} < C_{1}$ [@li2014biclustering] 
 2.
 *A truncated Normal prior (BwTNLV).* Independent spherical normal distributions as priors for $\boldsymbol \theta_{main}$ and $\boldsymbol \theta_{interaction}$, with $\sigma_{interaction} < \sigma_{main}$, *truncated* at $3\sigma_{main}$ and $3\sigma_{interaction}$, respectively.
-Simulation from the posterior using a geometric adaptive MH step [@zhou2014some], $h_j$ are carried along in the MCMC implementation as latent variables.
+Simulation from the posterior using a geometric adaptive MH step [@zhou2014some], $h_j$ are carried along in the MCMC implementation as latent variables
 3.
-*A truncated Normal prior and marginalized likelihood (BwTNML).* Marginalize out $\boldsymbol h$ in $f_{\boldsymbol \theta}(\boldsymbol x)$, and use the truncated Normal priors applied to the marginal probabilities for visible variables.
+*A truncated Normal prior and marginalized likelihood (BwTNML).* Marginalize out $\boldsymbol h$ in $f_{\boldsymbol \theta}(\boldsymbol x)$, and use the truncated Normal priors applied to the marginal probabilities for visible variables
 
@@
 
-# Appendix: Proof of Result \ref{prop:instab}
-\scriptsize
-
-We prove the contrapositive.
-Suppose that $\Delta(\boldsymbol \theta_\nv) \le C$ holds for some $C > 0$.
-Under the RBM model for visibles, $P_{\boldsymbol \theta_\nv}(\boldsymbol v) > 0$ holds for each outcome $\boldsymbol v \in \mathcal{C}^\nv$.
-Let $\boldsymbol v_{min} \equiv \argmin\limits_{\boldsymbol v \in \mathcal{C}^\nv}P_{\boldsymbol \theta_\nv}(\boldsymbol v)$ and $\boldsymbol v_{max} \equiv \argmax\limits_{\boldsymbol v \in \mathcal{C}^V\nv}P_{\boldsymbol \theta_\nv}(\boldsymbol v)$.
-Note there exists a sequence $\boldsymbol v_{min} \equiv \boldsymbol v_0, \boldsymbol v_1, \dots, \boldsymbol v_k \equiv \boldsymbol v_{max}$ in $\mathcal{C}^\nv$ of component-wise switches to move from $\boldsymbol v_{min}$ to $\boldsymbol v_{max}$ in the sample space (i.e.
-$\boldsymbol v_i, \boldsymbol v_{i + 1} \in \mathcal{C}^\nv$ differ by exactly $1$ component for $i = 0, \dots, k$) for some integer $k \in \{0, 1, \dots, \nv\}$.
-Then
-\begin{align*}
-\log\left[\frac{P_{\boldsymbol \theta_\nv}(\boldsymbol v_{max})}{P_{\boldsymbol \theta_\nv}(\boldsymbol v_{min})}\right] &= \left|\sum\limits_{i = 1}^k\log\left(\frac{P_{\boldsymbol \theta_\nv}(\boldsymbol v_i)}{P_{\boldsymbol \theta_\nv}(\boldsymbol v_{i-1})}\right)\right| \\
-& \le \sum\limits_{i = 1}^k\left|\log\left(\frac{P_{\boldsymbol \theta_\nv}(\boldsymbol v_i)}{P_{\boldsymbol \theta_\nv}(\boldsymbol v_{i-1})}\right)\right| \\
-& \le k \Delta(\boldsymbol \theta_\nv) \le \nv C
-\end{align*}
-using $k \le \nv$ and $\Delta(\boldsymbol \theta_\nv) \le C$.$\Box$
-
-# Appendix: Proof of Result \ref{prop:degen}
-
-\scriptsize
-
-Define $\boldsymbol v_{max}$ and $\boldsymbol v_{min}$ as in the proof of Proposition \ref{prop:instab}.
-Fix $0 < \epsilon < 1$ Then, $\boldsymbol v_{max} \in M_{\epsilon, \boldsymbol \theta_\nv}$, so $P_{\boldsymbol \theta_\nv}(M_{\epsilon, \boldsymbol \theta}) \ge P_{\boldsymbol \theta_\nv}(\boldsymbol v_{max})$.
-If $\boldsymbol v \in \mathcal{C}^\nv \setminus M_{\epsilon, \boldsymbol \theta_\nv}$, then by definition $P_{\boldsymbol \theta_\nv}(\boldsymbol v) \le [P_{\boldsymbol \theta_\nv}(\boldsymbol v_{max})]^{1-\epsilon}[P_{\boldsymbol \theta_\nv}(\boldsymbol v_{min})]^{\epsilon}$ holds so that
-\begin{align*}
-1-P_{\boldsymbol \theta_\nv}(M_{\epsilon, \boldsymbol \theta_\nv}) &= P_{\boldsymbol \theta_\nv}(M_{\epsilon, \boldsymbol \theta_\nv}^C) \\
-& = \sum\limits_{\boldsymbol v \in \mathcal{C}^\nv \setminus M_{\epsilon, \boldsymbol \theta_\nv}}P_{\boldsymbol \theta_\nv}(\boldsymbol v) \\
-& \le (2^\nv)[P_{\boldsymbol \theta_\nv}(\boldsymbol v_{max})]^{1-\epsilon}[P_{\boldsymbol \theta_\nv}(\boldsymbol v_{min})]^{\epsilon}
-\end{align*}
-Then,
-\begin{align*}
-\frac{1}{\nv}\log\left[\frac{P_{\boldsymbol \theta_\nv}(M_{\epsilon, \boldsymbol \theta_\nv})}{1-P_{\boldsymbol \theta_\nv}(M_{\epsilon, \boldsymbol \theta_\nv})}\right] & \ge \frac{1}{\nv} \log\left[\frac{P_{\boldsymbol\theta_\nv}(\boldsymbol v_{max})}{(2^\nv)[P_{\boldsymbol \theta_\nv}(\boldsymbol v_{max})]^{1-\epsilon}[P_{\boldsymbol \theta_\nv}(\boldsymbol v_{min})]^{\epsilon}}\right] \\
-&= \frac{\epsilon}{\nv} \log\left[\frac{P_{\boldsymbol \theta_\nv}(\boldsymbol v_{max})}{P_{\boldsymbol \theta_\nv}(\boldsymbol v_{min})}\right] - \log 2 \rightarrow \infty
-\end{align*}
-as $\nv \rightarrow \infty$ by the definition of an unstable RBM model.$\Box$
-
 # Appendix: Parameters used

### 2016-12-16 presentations/amherst/slides.Rmd
cuts and future work

@@
 
-1. *A "trick" prior (BwTPLV).* Cancel out normalizing term in the likelihood, resulting full conditionals of $\boldsymbol \theta$ are multivariate Normal, $h_j$ are carried along as latent variables
+1. *A "trick" prior (BwTPLV)* 
+    - Cancel out normalizing term in the likelihood
+    - Resulting full conditionals of $\boldsymbol \theta$ are multivariate Normal
+    - $h_j$ are carried along as latent variables
     \begin{align*}
@@
     where $C_{2} < C_{1}$ [@li2014biclustering]
-2.
-*A truncated Normal prior (BwTNLV).* Independent spherical normal distributions as priors for $\boldsymbol \theta_{main}$ and $\boldsymbol \theta_{interaction}$, with $\sigma_{interaction} < \sigma_{main}$, *truncated* at $3\sigma_{main}$ and $3\sigma_{interaction}$, respectively.
-Simulation from the posterior using a geometric adaptive MH step [@zhou2014some], $h_j$ are carried along in the MCMC implementation as latent variables
-3.
-*A truncated Normal prior and marginalized likelihood (BwTNML).* Marginalize out $\boldsymbol h$ in $f_{\boldsymbol \theta}(\boldsymbol x)$, and use the truncated Normal priors applied to the marginal probabilities for visible variables
+
+# Fitting methodologies (cont'd)
+
+2. *A truncated Normal prior (BwTNLV)* 
+    - Independent spherical normal distributions as priors for $\boldsymbol \theta_{main}$ and $\boldsymbol \theta_{interaction}$
+        - $\sigma_{interaction} < \sigma_{main}$ 
+        - *truncated* at $3\sigma_{main}$ and $3\sigma_{interaction}$, respectively
+    - Simulation from the posterior using a geometric adaptive MH step [@zhou2014some]
+    - $h_j$ are carried along in the MCMC implementation as latent variables 
+3. *A truncated Normal prior and marginalized likelihood (BwTNML)* 
+    - Marginalize out $\boldsymbol h$ in $f_{\boldsymbol \theta}(\boldsymbol x)$
+    - Use the truncated Normal priors applied to the marginal probabilities for visible variables
 
@@
   geom_segment(aes(x=lag, xend=lag, y=0, yend=marginal), colour = "red") +
+  geom_point(aes(x=lag, y=marginal), colour = "red", size=.5) +
   facet_wrap(~image_id) +
@@
 
-We can use an overlapping blockmeans approach to get a crude estimate for the aysmptotic variance of the probability of each image and compare it to an estimate of the asymptotic variance assuming IID draws from the target distribution.
+- Overlapping blockmeans approach [@gelman2011inference]
+    - Crude estimate for the aysmptotic variance of the probability of each image 
+    - Compare it to an estimate of the asymptotic variance assuming IID draws from the target distribution
 
@@
 
-- While RBMs are thought to be useful for classification, in the context of generative statistical models, RBMs are a poor fit due to *near-degeneracy*, *S-instability*, and *uninterpretability*.
-- Rigorous fitting methodology is difficult due to the dimension of the parameter space coupled with the size of the latent variable space.
-- For a RBM model with enough hidden variables, any distribution for the visibles can be approximated arbitrarily well [@le2008representational; @montufar2011refinements; and @montufar2011expressive].
-    - The empirical distribution of a training set of vectors of visibles is the best fitting model for observed cell data. 
-    - There can be no "smoothed distribution" achieved in fitting a RBM model of sufficient size with a rigorous likelihood-based method.
+- RBMs are thought to be useful for classification, but in the context of generative statistical models, poor fit due to *near-degeneracy*, *S-instability*, and *uninterpretability*
+- Rigorous fitting methodology is difficult due to the dimension of the parameter space & size of the latent variable space
+- For a RBM model with enough hidden variables, any distribution for the visibles can be approximated arbitrarily well [@le2008representational; @montufar2011refinements; and @montufar2011expressive] 
+    - The empirical distribution of a training set is the best fitting model for observed cell data 
+    - There can be no "smoothed distribution" achieved in a RBM model of sufficient size with a rigorous likelihood-based method
 
-We are skeptical that any model built using these structures (like a deep Boltzmann machine) can achieve useful prediction or inference in a principled way without limiting the flexibility of the fitted model.
+Skeptical that any model built using RBMs (i.e. deep Boltzmann machine) can achieve useful **prediction** or **inference** in a principled way without limiting the flexibility of the fitted model
 
 # Future work
+
+1. Generalization of instability results [ongoing, see @kaplan2016note]
+2. Image classification
+    - Ensemble methods (super learners) using AdaBoost [@freund1995desicion] 
+    - Decision theoretic based approach to approximating the likelihood ratio test for classification
+3. Markov chain Monte Carlo methods for data with Markovian dependence
+    - Spatial data
+    - Network data
 

### 2016-12-16 presentations/amherst/slides.Rmd
edits from dr. nordman

@@
 shorttitle: "Model matters with RBMs"
-author: Andee Kaplan, Daniel Nordman, and Stephen Vardeman
+author: Andee Kaplan
 shortname: Kaplan, et al.
@@
   |
+  | Slides available at <http://bit.ly/kaplan-umass>
   | 
-  | Slides available at <http://bit.ly/kaplan-umass>
+  | \footnotesize Joint work with D. Nordman and S. Vardeman
 shortdate: "December 20, 2016"
@@
 
-1. Discrete or continuous random variables 
-2. two layers - one hidden and one visible
+2. two layers of random variables - one hidden and one visible
 3. conditional independence within a layer [@smolensky1986information]
@@
 \includegraphics[width=.6\linewidth]{images/rbm.png}
-\caption{An example RBM, which consists of two layers. Hidden nodes are indicated by white circles and the visible nodes are indicated by blue circles}
+\caption{Hidden nodes are indicated by white circles and the visible nodes are indicated by blue circles.}
 \end{figure}
@@
 \includegraphics[width=\linewidth]{images/visibles_image.pdf}
-\caption{Image classification using a RBM.
-On the left, each image pixel comprises a node in the visible layer, $\mathcal{V}$.
-On the right, the output of the RBM is used to create features which are then passed to a supervised learning algorithm.}
+\caption{Image classification using a RBM: each image pixel comprises a node in the visible layer, $\mathcal{V}$ and the output of the RBM is used to create features passed to a supervised learning algorithm.}
 \end{figure}
@@
 
-- $\boldsymbol x = (h_1, \dots, h_{\nh}, v_1,\dots,v_{\nv})$ represents the states of the visible and hidden nodes in a RBM 
-- Each single "binary" random variable, visible or hidden, will take its values in a common coding set $\mathcal{C}$
-    - Two possibilities for the coding set $\mathcal{C}=\{0,1\}$ or $\mathcal{C}=\{-1,1\}$. 
+- $\boldsymbol x = (h_1, \dots, h_{\nh}, v_1,\dots,v_{\nv})$ represents visible and hidden nodes in a RBM 
+- Each single "binary" random variable, visible $v_i$ or hidden $h_j$, takes values in a common coding set 
+    - $\mathcal{C}=\{0,1\}$ or $\mathcal{C}=\{-1,1\}$. 
 - A parametric form for probabilities
 
-    \begin{align} 
+    \begin{align*} 
     \label{eqn:pmf} 
-    f_{\boldsymbol \theta} (\boldsymbol x) = \frac{\exp\left(\sum\limits_{i = 1}^{\nv} \sum\limits_{j=1}^{\nh} \theta_{ij} v_i h_j + \sum\limits_{i = 1}^{\nv}\theta_{v_i} v_i + \sum\limits_{j = 1}^{\nh}\theta_{h_j} h_j\right)}{\gamma(\boldsymbol \theta)} \end{align}
+    f_{\boldsymbol \theta} (\boldsymbol x) = \frac{\exp\left(\sum\limits_{i = 1}^{\nv} \sum\limits_{j=1}^{\nh} \theta_{ij} v_i h_j + \sum\limits_{i = 1}^{\nv}\theta_{v_i} v_i + \sum\limits_{j = 1}^{\nh}\theta_{h_j} h_j\right)}{\gamma(\boldsymbol \theta)} \end{align*}
 
@@
 450).
-\item Treating a hidden layer of one RBM as the visible layer in a second RBM, etc.
 \end{itemize}
@@
 
-1. The model properties are largely unexplored in the literature and 
+1. The model properties are largely unexplored in the literature
 2. The commonly cited fitting methodology remains heuristic-based and abstruse [@hinton2006fast]
@@
 
-1. Provide steps toward a thorough understanding of the model class and its properties from the perspective of statistical theory, and 
+1. Provide steps toward understanding properties of the model class from the perspective of statistical theory
 2. Explore the possibility of a rigorous fitting methodology
@@
 
-> A model should "provide an explanation of the mechanism underlying the observed phenomena" [@lehmann1990model; @box1967discrimination].
+> A model should "provide an explanation of the mechanism underlying the observed phenomena" [@box1967discrimination].
 
@@
 \begin{definition}[Model Degeneracy]
-There is a disproportionate amount of probability placed on only a few elements of the sample space, $\mathcal{C}^{\nh + \nv}$, by the model.
+A disproportionate amount of probability is placed on only a few elements of the sample space, $\mathcal{C}^{\nh + \nv}$, by the model.
 \end{definition}
@@
 - Concept of model deficiency related to *instability* for a class of exponential families of distributions [@schweinberger2011instability]
-- For the RBM, consider how a data model might be expanded to incorporate more visibles
-    - Necessary to grow the number of model parameters in a sequence $\boldsymbol \theta_{\nv} \in \mathbb{R}^{\nv + \nh + \nv*\nh}, \nv \ge 1$ 
+- For the RBM, consider how model incorporates more visibles
+    - Model parameters in a longer sequence $\boldsymbol \theta_{\nv} \in \mathbb{R}^{\nv + \nh + \nv*\nh}, \nv \ge 1$ 
     - May also arbitrarily expand the number of hidden variables used.
@@
 
-S-unstable RBM model sequences are undesirable for several reasons - small changes in data can lead to overly-sensitive changes in probability.
+S-unstable RBM models are undesirable for several reasons - small changes in data outcomes can lead to overly-sensitive changes in probability.
 
@@
 \label{prop:instab}
-Let $c > 0$ and let $\text{ELPR}(\boldsymbol \theta_\nv)$ be as in (\ref{eqn:elpr}) for an integer $\nv \ge 1$.
-If $\frac{1}{\nv}\text{ELPR}(\boldsymbol \theta_\nv) > c$, then $\Delta(\boldsymbol \theta_\nv) > c$.
+Let $c > 0$ and fix an integer $\nv \ge 1$. If $\frac{1}{\nv}\text{ELPR}(\boldsymbol \theta_\nv) > c$, then $\Delta(\boldsymbol \theta_\nv) > c$.
 \end{result}
 
-If the quantity (\ref{eqn:elpr}) is too large, then a RBM model sequence will exhibit large probability shifts for very small changes in the data configuration.
+If the $\nv^{-1}ELPR(\boldsymbol\theta_{\nv})$ is too large, then a RBM model will exhibit large probability shifts for very small changes in the data configuration.
 
@@
 - All probability will stack up on mode sets or potentially those few outcomes with the highest probability 
-- Proofs of results \ref{prop:instab}-\ref{prop:degen} can be found in [@kaplan2016note]
+- Proofs found in [@kaplan2016note]
 
@@
 
-- Measure: the magnitude of the difference between
-    1. Model expectations, E$\left[\boldsymbol X | \boldsymbol \theta\right]$, and 
-    2.
-Expectations given independence, E$\left[\boldsymbol X | \boldsymbol \theta^* \right]$, where $\boldsymbol \theta^*$ matches $\boldsymbol \theta$ for all main effects but otherwise has $\theta_{ij} = 0$ for $i = 1, \dots, \nv, j = 1, \dots, \nh$
+- Model expectations, E$\left[\boldsymbol X | \boldsymbol \theta\right]$
+- Expectations given independence, E$\left[\boldsymbol X | \boldsymbol \theta^* \right]$, where $\boldsymbol \theta^*$ matches $\boldsymbol \theta$ for all main effects but otherwise has $\theta_{ij} = 0$ for $i = 1, \dots, \nv, j = 1, \dots, \nh$
 - If $|\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] - \text{E}\left[\boldsymbol X | \boldsymbol \theta^*\right]|$ is large then the RBM with parameter vector $\boldsymbol \theta$ is *uninterpetable*
@@
 \includegraphics[width=.35\linewidth]{images/toyhull_top_2.png}
-\caption{The convex hull of the "statistic space" $\mathcal{T} = \{(v_1, h_1, v_1 h_1): v_1, h_1 \in \mathcal{C}\}$ in three dimensions for the toy RBM with one visible and one hidden node for $\mathcal{C} = \{0,1\}$ (left) and $\mathcal{C} = \{-1,1\}$ (right) data encoding.}
+\caption{The convex hull of the "statistic space" $\mathcal{T} = \{(v_1, h_1, v_1 h_1): v_1, h_1 \in \mathcal{C}\}$ in three dimensions for a toy RBM with one visible and one hidden node for $\mathcal{C} = \{0,1\}$ (left) and $\mathcal{C} = \{-1,1\}$ (right) data encoding.}
 \end{figure}
@@
 
-- To explore the behavior of the RBM parameters $\boldsymbol \theta$ as it relates to *near-degeneracy*, *instability*, and *uninterpretability*, consider models of small size  
+- To explore the effects of RBM parameters $\boldsymbol \theta$ on *near-degeneracy*, *instability*, and *uninterpretability*, consider models of small size  
 - For $\nh, \nv \in \{1, \dots, 4\}$, sample `r dim(res$samp[[1]])[1]` values of $\boldsymbol \theta$
-    1. Split $\boldsymbol \theta$ into $\boldsymbol \theta_{interaction}$ and $\boldsymbol \theta_{main}$, in reference to which sufficient statistics the parameters correspond to
+    1. Split $\boldsymbol \theta$ into $\boldsymbol \theta_{interaction}$ and $\boldsymbol \theta_{main}$
     2. Allow the two types of terms to have varying average magnitudes, $||\boldsymbol \theta_{main} || /(\nh+\nv)$ and $||\boldsymbol \theta_{interaction} || /(\nh*\nv)$
@@
 
-1. Computational concerns
-    - Fitting a RBM via maximum likelihood (ML) methods infeasible due to the intractibility of the normalizing term $\gamma(\boldsymbol \theta)$
-        - Ad hoc methods are used, aim to avoid this problem by using stochastic ML 
-        - employ a small number of MCMC draws to approximate $\gamma(\boldsymbol \theta)$ 
-2. With $\nh$ large enough 
+1. Computational concerns: Fitting a RBM via maximum likelihood (ML) methods infeasible due to the intractibility of the normalizing term $\gamma(\boldsymbol \theta)$
+    - Ad hoc methods used to avoid this problem with stochastic ML 
+    - Employ a small number of MCMC draws to approximate $\gamma(\boldsymbol \theta)$ 
+2. Model parameterization concerns: With enough hiddens, 
     - Potential to re-create any distribution for the data [@le2008representational; @montufar2011refinements; and @montufar2011expressive]
@@
         - The RBM model ensures that this empirical distribution can be arbitrarily well approximated
-    - When empirical distribution contains empty cells, fitting will chase parameters to $\infty$ in order to zero out the corresponding RBM cell probabilities
+    - When empirical distribution contains empty cells, ML will chase parameters to $\infty$ in order to zero out corresponding RBM cell probabilities
 
@@
 
-- Consider what might be done in a principled manner, testing on $\nv = \nh = 4$ 
-- To avoid model impropriety, we want to avoid parts of the parameter space $\mathbb{R}^{\nv + \nh + \nv*\nh}$ that lead to *near-degeneracy*, *instability*, and *uninterpretability*.
+- Consider what might be done in a principled manner, small test
+- To avoid model impropriety, avoid parts of the parameter space $\mathbb{R}^{\nv + \nh + \nv*\nh}$ leading to *near-degeneracy*, *instability*, and *uninterpretability*.
     - Shrink $\boldsymbol \theta$ toward $\boldsymbol 0$ 
@@
         2. Shrink $\boldsymbol \theta_{interaction}$ more than $\boldsymbol \theta_{main}$
-- Consider a test case with $\nv = \nh = 4$ and parameters given in in appendix 
+- Consider a test case with $\nv = \nh = 4$ 
     - $\boldsymbol \theta$ chosen as a sampled value from a grid point in figure \ref{fig:degen_plots} with $< 5$\% near-degeneracy (not near the convex hull of the sufficient statistics) 
@@
     - Marginalize out $\boldsymbol h$ in $f_{\boldsymbol \theta}(\boldsymbol x)$
-    - Use the truncated Normal priors applied to the marginal probabilities for visible variables
+    - Use the truncated Normal priors applied to the marginal probabilities for visible variables (recall visibles are the observed data, hiddens are not)
 
@@
 
-- RBMs are thought to be useful for classification, but in the context of generative statistical models, poor fit due to *near-degeneracy*, *S-instability*, and *uninterpretability*
+- RBMs used for classification, but are concerning as statistical models due to *near-degeneracy*, *S-instability*, and *uninterpretability*
 - Rigorous fitting methodology is difficult due to the dimension of the parameter space & size of the latent variable space
-- For a RBM model with enough hidden variables, any distribution for the visibles can be approximated arbitrarily well [@le2008representational; @montufar2011refinements; and @montufar2011expressive] 
-    - The empirical distribution of a training set is the best fitting model for observed cell data 
+- Fitting a RBM model is also questionable as any distribution for the visibles can be approximated arbitrarily well
+    - The empirical distribution of visibles is the best fitting model for observed cell data 
     - There can be no "smoothed distribution" achieved in a RBM model of sufficient size with a rigorous likelihood-based method
@@
 
-1. Generalization of instability results [ongoing, see @kaplan2016note]
+1. Generalization of instability results for other network models [ongoing, see @kaplan2016note]
 2. Image classification
@@
     * GitHub -- <http://github.com/andeek>
-
----- 
-\begin{center}
-\Huge
-Appendices
-\end{center}
 

### 2016-12-18 presentations/amherst/slides.Rmd
small typos

@@
     - Model parameters in a longer sequence $\boldsymbol \theta_{\nv} \in \mathbb{R}^{\nv + \nh + \nv*\nh}, \nv \ge 1$ 
-    - May also arbitrarily expand the number of hidden variables used.
+    - May also arbitrarily expand the number of hidden variables used
 
@@
 
-Consider the biggest log-probability ratio for a one-pixel (one component) change in data outcomes (visibles)
+Consider the biggest log-probability ratio for a one-pixel (one component) change in visibles (data outcomes)
 
@@
   geom_abline(aes(intercept = 0, slope = 1), alpha = .5, lty = 2) +
-  scale_fill_gradient(expression(frac(bar(ELPR(theta)), V)), low = "yellow", high = "red") +
+  scale_fill_gradient(expression(frac(bar(ELPR(theta)), n[v])), low = "yellow", high = "red") +
   facet_grid(Visibles~Hiddens, labeller = label_parsed) +
@@
 
-The truncated Normal method (2) and the marginalized likelihood method (3) are drawing from the same stationary posterior distribution for images.
+The BwTNLV (2) and the BwTNML method (3) are drawing from the same stationary posterior distribution for images.
 

### 2017-01-05 presentations/asu/slides.Rmd
add asu slides

--- 2016-12-18 presentations/amherst/slides.Rmd
@@
 date: |
-  | December 20, 2016
+  | January 11, 2017
   |
-  | Slides available at <http://bit.ly/kaplan-umass>
+  | Slides available at <http://bit.ly/kaplan-asu>
   | 
   | \footnotesize Joint work with D. Nordman and S. Vardeman
-shortdate: "December 20, 2016"
+shortdate: "January 11, 2017"
 output: 
@@
 
-* Slides -- <http://bit.ly/kaplan-umass>
+* Slides -- <http://bit.ly/kaplan-asu>
 

### 2017-01-16 presentations/uva/slides.Rmd
add uva slides

--- 2017-01-05 presentations/asu/slides.Rmd
@@
 date: |
-  | January 11, 2017
+  | January 19, 2017
   |
-  | Slides available at <http://bit.ly/kaplan-asu>
+  | Slides available at <http://bit.ly/kaplan-uva>
   | 
   | \footnotesize Joint work with D. Nordman and S. Vardeman
-shortdate: "January 11, 2017"
+shortdate: "January 19, 2017"
 output: 
@@
 
-```{r acf, fig.cap="\\label{fig:acf}The autocorrelation functions (ACF) for the posterior probabilityies of all $2^4 = 16$ possible outcomes for the vector of $4$ visibles assessed at multiple lags for each method with BwTNLV in black and BwTNML in red.", fig.height=4.5, fig.width=8, out.width=".7\\linewidth"}
+```{r acf, fig.cap="\\label{fig:acf}The autocorrelation functions (ACF) for the posterior probabilities of all $2^4 = 16$ possible outcomes for the vector of $4$ visibles assessed at multiple lags for each method with BwTNLV in black and BwTNML in red.", fig.height=4.5, fig.width=8, out.width=".7\\linewidth"}
 marginal_sample_good %>% 
@@
   ylab("Scaled Posterior Density") + xlab("Probability of Image") +
-  scale_colour_discrete("Method", labels = c("Marginalized likelihood", "Trick prior")) +
-  scale_fill_discrete("Method", labels = c("Marginalized likelihood", "Trick prior")) +
+  scale_colour_discrete("Method", labels = c("BwTNML", "BwTPLV")) +
+  scale_fill_discrete("Method", labels = c("BwTNML", "BwTPLV")) +
   theme(legend.position = "bottom") -> p.models
@@
 
-* Slides -- <http://bit.ly/kaplan-asu>
+* Slides -- <http://bit.ly/kaplan-uva>
 

### 2017-01-16 presentations/uva/images/plot_visibles.R

@@ -1,0 +1,14 @@
+library(ElemStatLearn)
+library(ggplot2)
+
+image <- cbind(expand.grid(x = 1:16, y = -1:-16), zip = zip.test[50,-1])
+ggplot(image) +
+  geom_tile(aes(x = x, y = y, fill = zip), colour = "#0066cc") +
+  scale_fill_gradient(low = "white", high = "black") +
+  theme(axis.line=element_blank(),axis.text.x=element_blank(),
+        axis.text.y=element_blank(),axis.ticks=element_blank(),
+        axis.title.x=element_blank(),
+        axis.title.y=element_blank(),legend.position="none",
+        panel.background=element_blank(),panel.border=element_blank(),panel.grid.major=element_blank(),
+        panel.grid.minor=element_blank(),plot.background=element_blank(),
+        aspect.ratio = 1) 

### 2017-01-18 presentations/uva/slides.Rmd
cleanup language surrounding instability

@@
 - fail to generate data with realistic variability and thus an unsatisfactory conceptualization of the data generation process [@li2014biclustering] 
-- exhibit instability in the parameter space [@szegedy2013intriguing; @nguyen2014deep]
+- exhibit model instability (over-sensitivity) [@szegedy2013intriguing; @nguyen2014deep]
 

### 2017-01-21 presentations/rice/slides.Rmd
i'm dumb

--- 2017-01-18 presentations/uva/slides.Rmd
@@
 date: |
-  | January 19, 2017
+  | January 23, 2017
   |
-  | Slides available at <http://bit.ly/kaplan-uva>
+  | Slides available at <http://bit.ly/kaplan-rice>
   | 
@@
 
-* Slides -- <http://bit.ly/kaplan-uva>
+* Slides -- <http://bit.ly/kaplan-rice>
 

### 2017-01-21 presentations/uva/slides.Rmd
i'm dumb

@@
   | \footnotesize Joint work with D. Nordman and S. Vardeman
-shortdate: "January 19, 2017"
+shortdate: "January 23, 2017"
 output: 

### 2017-01-25 presentations/samsi/slides.Rmd
samsi

--- 2017-01-18 presentations/uva/slides.Rmd
@@
 date: |
-  | January 19, 2017
+  | January 30, 2017
   |
-  | Slides available at <http://bit.ly/kaplan-uva>
+  | Slides available at <http://bit.ly/kaplan-samsi>
   | 
   | \footnotesize Joint work with D. Nordman and S. Vardeman
-shortdate: "January 19, 2017"
+shortdate: "January 30, 2017"
 output: 
@@
 
-* Slides -- <http://bit.ly/kaplan-uva>
+* Slides -- <http://bit.ly/kaplan-samsi>
 

### 2017-01-31 presentations/ufl/slides.Rmd
ufl and uiuc slides

--- 2017-01-18 presentations/uva/slides.Rmd
@@
 date: |
-  | January 19, 2017
+  | February 2, 2017
   |
-  | Slides available at <http://bit.ly/kaplan-uva>
+  | Slides available at <http://bit.ly/kaplan-ufl>
   | 
   | \footnotesize Joint work with D. Nordman and S. Vardeman
-shortdate: "January 19, 2017"
+shortdate: "February 2, 2017"
 output: 
@@
 ~\\
-\text{E}\left[\boldsymbol X | \boldsymbol \theta^*\right] &= \sum\limits_{\boldsymbol x \in \mathcal{C}^{\nh + \nv}} \boldsymbol x \frac{\exp\left(\sum\limits_{i = 1}^\nv \theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{C}^{\nh + V=\nv}}\exp\left(\sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}
+\text{E}\left[\boldsymbol X | \boldsymbol \theta^*\right] &= \sum\limits_{\boldsymbol x \in \mathcal{C}^{\nh + \nv}} \boldsymbol x \frac{\exp\left(\sum\limits_{i = 1}^\nv \theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{C}^{\nh + \nv}}\exp\left(\sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}
 \end{align*}
@@
 
-* Slides -- <http://bit.ly/kaplan-uva>
+* Slides -- <http://bit.ly/kaplan-ufl>
 

### 2017-01-31 presentations/uiuc/slides.Rmd
ufl and uiuc slides

--- 2017-01-31 presentations/ufl/slides.Rmd
@@
 date: |
-  | February 2, 2017
+  | February 7, 2017
   |
-  | Slides available at <http://bit.ly/kaplan-ufl>
+  | Slides available at <http://bit.ly/kaplan-uiuc>
   | 
   | \footnotesize Joint work with D. Nordman and S. Vardeman
-shortdate: "February 2, 2017"
+shortdate: "February 7, 2017"
 output: 
@@
 
-* Slides -- <http://bit.ly/kaplan-ufl>
+* Slides -- <http://bit.ly/kaplan-uiuc>
 

### 2017-01-31 presentations/ufl/images/degeneracy_plots.R

@@ -1,0 +1,221 @@
+library(knitr)
+library(ggplot2)
+library(dplyr)
+library(tidyr)
+
+opts_chunk$set(echo=FALSE, message=FALSE, warnings=FALSE)
+theme_set(theme_bw(base_family="serif"))
+
+source("../../degeneracy/functions.R")
+
+##variable encoding ----------------------
+test_cases <- expand.grid(data.frame(1:4)[,rep(1,2)]) %>% 
+  rename(H = X1.4, V = X1.4.1) %>%
+  filter(H <= V) %>% #remove those cases with less visibles than hiddens. Is this necessary?
+  mutate(n_param = H*V + H + V) %>%
+  mutate(max_facets = (2^(H+V))^(floor(n_param/2))) %>%
+  filter(n_param <= 11) #calc_hull can't handle any higher dimensions currently
+
+bin_res <- plyr::dlply(test_cases, plyr::.(n_param), function(x) calc_hull(x$V, x$H, "binary"))
+neg_res <- plyr::dlply(test_cases, plyr::.(n_param), function(x) calc_hull(x$V, x$H, "negative"))
+
+plyr::ldply(bin_res, function(x) x$c_hull$vol) %>% 
+  mutate(frac_vol = V1/(1^n_param)) %>%
+  inner_join(plyr::ldply(neg_res, function(x) x$c_hull$vol) %>% 
+               mutate(frac_vol = V1/(2^n_param)),
+             by="n_param") %>%
+  rename(vol.bin = V1.x, vol.neg = V1.y, frac_vol.bin = frac_vol.x, frac_vol.neg = frac_vol.y) %>%
+  gather(vars, value, -n_param) %>%
+  separate(vars, c("type", "encoding"), "\\.") %>%
+  spread(type, value) %>%
+  ggplot() +
+  geom_point(aes(x=n_param, y=frac_vol, colour=encoding)) +
+  geom_line(aes(x=n_param, y=frac_vol, colour=encoding, group=encoding)) +
+  ylab("Fraction of unrestricted volume") +
+  xlab("Number of parameters") +
+  scale_colour_discrete("Encoding", labels=c("Binary (1,0)", "Negative (1,-1)")) +           
+  theme(
+    plot.background = element_rect(fill = "transparent", colour = NA),
+    legend.background = element_rect(fill = "transparent", colour = NA)) -> p
+
+ggsave("images/frac_volume.pdf",
+       plot = p,
[…]

### 2017-01-31 presentations/uiuc/images/fitting_plots.R

@@ -1,0 +1,99 @@
+#libraries --------------------
+library(dplyr)
+library(tidyr)
+library(ggplot2)
+source("../../4x4 model/sample_functs.R")
+
+#data and params ------------------------
+H <- 4
+V <- 4
+
+#marginalized likelihood
+load("../../4x4 model/written/fitted_models_adaptive_mh_trunc_distn_marginal_1.Rdata")
+marginal_bad_1 <- models_bad
+marginal_good_1 <- models_good
+marginal_degen_1 <- models_degen
+
+#load Jing's prior
+load("../../4x4 model/written/fitted_models_distn_0.26.Rdata")
+jing_bad <- models_bad
+jing_good <- models_good
+jing_degen <- models_degen
+
+#truncated normal
+load("../../4x4 model/written/fitted_models_adaptive_mh_trunc_distn_shrunk_jing_match.Rdata")
+shrunk_jing_bad <- models_bad
+shrunk_jing_good <- models_good
+shrunk_jing_degen <- models_degen
+
+#rm unneccesary data
+rm(models_bad)
+rm(models_degen)
+rm(models_good)
+
+load("../../4x4 model/written/sample.params.Rdata")
+params_degen <- list(main_hidden = sample.params %>% ungroup() %>% filter(near_hull) %>% select(starts_with("h"), -H) %>% data.matrix(),
+                     main_visible = sample.params %>% ungroup() %>% filter(near_hull) %>% select(starts_with("v"), -V) %>% data.matrix(),
+                     interaction = sample.params %>% ungroup() %>% filter(near_hull) %>% select(starts_with("theta")) %>% data.matrix() %>% matrix(4))
+params_good <- list(main_hidden = sample.params %>% ungroup() %>% filter(!near_hull) %>% select(starts_with("h"), -H) %>% data.matrix(),
+                    main_visible = sample.params %>% ungroup() %>% filter(!near_hull) %>% select(starts_with("v"), -V) %>% data.matrix(),
+                    interaction = sample.params %>% ungroup() %>% filter(!near_hull) %>% select(starts_with("theta")) %>% data.matrix() %>% matrix(4))
+
+
+load("../../4x4 model/written/sample_images.Rdata")
+
+#computing actual distributions ---------------------
+distn_good <- visible_distn(params = params_good)
+distn_degen <- visible_distn(params = params_degen)
+
[…]

### 2017-02-13 presentations/msu/slides.Rmd
add MSU

--- 2017-01-31 presentations/uiuc/slides.Rmd
@@
   |
-  | Slides available at <http://bit.ly/kaplan-uiuc>
+  | Slides available at <http://bit.ly/kaplan-msu>
   | 
@@
 
-2. two layers of random variables - one hidden and one visible
-3. conditional independence within a layer [@smolensky1986information]
+1. two layers of random variables - one hidden and one visible
+2. conditional independence within a layer [@smolensky1986information]
 
@@
 
-* Slides -- <http://bit.ly/kaplan-uiuc>
+* Slides -- <http://bit.ly/kaplan-msu>
 

### 2017-02-13 presentations/msu/slides.Rmd
date fix

@@
 date: |
-  | February 7, 2017
+  | February 14, 2017
   |
@@
   | \footnotesize Joint work with D. Nordman and S. Vardeman
-shortdate: "February 7, 2017"
+shortdate: "February 14, 2017"
 output: 

### 2017-03-25 presentations/amherst/slides.Rmd
fix stupid structure

@@
   beamer_presentation:
-    template: latex/beamer.tex
+    template: ../resources/latex/beamer.tex
     includes:
-      in_header: latex/front-matter.tex
+      in_header: ../resources/latex/front-matter.tex
 theme: CambridgeUS
-bibliography: latex/refs.bib
+bibliography: ../resources/latex/refs.bib
 fig_caption: true
@@
 
-source("helpers/functions.R")
+source("../resources/helpers/functions.R")
 ```
@@
 \begin{figure}
-\includegraphics[width=.6\linewidth]{images/rbm.png}
+\includegraphics[width=.6\linewidth]{../resources/images/rbm.png}
 \caption{Hidden nodes are indicated by white circles and the visible nodes are indicated by blue circles.}
@@
 \begin{figure}
-\includegraphics[width=\linewidth]{images/visibles_image.pdf}
+\includegraphics[width=\linewidth]{../resources/images/visibles_image.pdf}
 \caption{Image classification using a RBM: each image pixel comprises a node in the visible layer, $\mathcal{V}$ and the output of the RBM is used to create features passed to a supervised learning algorithm.}
@@
 \begin{figure}
-\includegraphics[width=\linewidth]{images/deep_rbm.png}
+\includegraphics[width=\linewidth]{../resources/images/deep_rbm.png}
 \caption{Three layer deep Boltzmann machine, with visible-to-hidden and hidden-to-hidden connections but no within-layer connections.}
@@
 \begin{figure}
-\includegraphics[width=.55\linewidth]{images/toyhull_top.png}
-\includegraphics[width=.35\linewidth]{images/toyhull_top_2.png}
+\includegraphics[width=.55\linewidth]{../resources/images/toyhull_top.png}
+\includegraphics[width=.35\linewidth]{../resources/images/toyhull_top_2.png}
 \caption{The convex hull of the "statistic space" $\mathcal{T} = \{(v_1, h_1, v_1 h_1): v_1, h_1 \in \mathcal{C}\}$ in three dimensions for a toy RBM with one visible and one hidden node for $\mathcal{C} = \{0,1\}$ (left) and $\mathcal{C} = \{-1,1\}$ (right) data encoding.}
@@
 #grid data
-load("data/results_grid.RData")
+load("../resources/data/results_grid.RData")
 
@@
 ```{r models-load}
-load("data/sample_images.Rdata")
-load("data/params_theta.Rdata")
+load("../resources/data/sample_images.Rdata")
+load("../resources/data/params_theta.Rdata")
 
@@
 #marginalized likelihood
-load("data/fitted_models_trunc_marginal_full.Rdata")
+load("../resources/data/fitted_models_trunc_marginal_full.Rdata")
 marginal_bad <- models_bad
@@
 #load trick prior
-load("data/fitted_models_jing_5.8.Rdata")
+load("../resources/data/fitted_models_jing_5.8.Rdata")
 trick_bad <- models_bad
@@
 #truncated normal
-load("data/fitted_models_trunc_full.Rdata")
+load("../resources/data/fitted_models_trunc_full.Rdata")
 trunc_bad <- models_bad

### 2017-03-25 presentations/asu/helpers/functions.R

@@ -1,70 +1,0 @@
-##function for creating all possible values of the t(x)
-stats <- function(H, V, type="negative") {
-  if(type == "binary") t <- data.frame(c(0,1))[,rep(1, H + V)]
-  else if(type == "negative") t <- data.frame(c(-1,1))[,rep(1, H + V)]
-  else stop("Type must be 'binary', 'negative'")
-  
-  names(t) <- c(paste0("v", 1:H), paste0("h", 1:V))  
-  
-  #all possibilities of statistics in (H + V)-space
-  t.grid <- expand.grid(t)
-  
-  for(i in 1:V){
-    for(j in (V+1):(H+V)){
-      t.grid <- cbind(t.grid, t.grid[,i]*t.grid[,j])
-      names(t.grid)[ncol(t.grid)] <- paste0("theta", i, j - V)
-    }
-  }
-  
-  return(data.matrix(t.grid))
-}
-
-
-##function for calculating the convex hull of t(x)
-calc_hull <- function(H, V, type="binary") {
-  #only possible values of t are 0 or 1
-  if(type == "binary") t <- data.frame(c(0,1))[,rep(1, H + V)]
-  else if(type == "negative") t <- data.frame(c(-1,1))[,rep(1, H + V)]
-  else stop("Type must be 'binary', 'negative', or 'experiment'")
-  
-  names(t) <- c(paste0("h", 1:H), paste0("v", 1:V))  
-  
-  #all possibilities of statistics in (H + V)-space
-  t.grid <- expand.grid(t)
-  
-  for(i in 1:H){
-    for(j in (H+1):(H+V)){
-      t.grid <- cbind(t.grid, t.grid[,i]*t.grid[,j])
-      names(t.grid)[ncol(t.grid)] <- paste0("theta", i, j - H)
-    }
-  }
-  
-  #calculate convex hull
-  library(geometry)
-  C <- convhulln(t.grid, options="FA")
-  return(list(possible_t=t.grid, c_hull=C))
-}
-
-expected_value <- function(theta, stats, normalized = TRUE) {
-  result <- crossprod(t(crossprod(stats, exp(crossprod(t(stats), theta)))), diag(1/apply(exp(crossprod(t(stats), theta)), 2, sum), nrow=length(apply(exp(crossprod(t(stats), theta)), 2, sum))))
-  rownames(result) <- paste0("exp_", rownames(result))
-  return(result)
-}
-
-visible_distn <- function(params) {
[…]

### 2017-03-25 presentations/msu/images/plot_visibles.R

@@ -1,14 +1,0 @@
-library(ElemStatLearn)
-library(ggplot2)
-
-image <- cbind(expand.grid(x = 1:16, y = -1:-16), zip = zip.test[50,-1])
-ggplot(image) +
-  geom_tile(aes(x = x, y = y, fill = zip), colour = "#0066cc") +
-  scale_fill_gradient(low = "white", high = "black") +
-  theme(axis.line=element_blank(),axis.text.x=element_blank(),
-        axis.text.y=element_blank(),axis.ticks=element_blank(),
-        axis.title.x=element_blank(),
-        axis.title.y=element_blank(),legend.position="none",
-        panel.background=element_blank(),panel.border=element_blank(),panel.grid.major=element_blank(),
-        panel.grid.minor=element_blank(),plot.background=element_blank(),
-        aspect.ratio = 1) 

### 2017-03-25 presentations/uva/images/fitting_plots.R

@@ -1,99 +1,0 @@
-#libraries --------------------
-library(dplyr)
-library(tidyr)
-library(ggplot2)
-source("../../4x4 model/sample_functs.R")
-
-#data and params ------------------------
-H <- 4
-V <- 4
-
-#marginalized likelihood
-load("../../4x4 model/written/fitted_models_adaptive_mh_trunc_distn_marginal_1.Rdata")
-marginal_bad_1 <- models_bad
-marginal_good_1 <- models_good
-marginal_degen_1 <- models_degen
-
-#load Jing's prior
-load("../../4x4 model/written/fitted_models_distn_0.26.Rdata")
-jing_bad <- models_bad
-jing_good <- models_good
-jing_degen <- models_degen
-
-#truncated normal
-load("../../4x4 model/written/fitted_models_adaptive_mh_trunc_distn_shrunk_jing_match.Rdata")
-shrunk_jing_bad <- models_bad
-shrunk_jing_good <- models_good
-shrunk_jing_degen <- models_degen
-
-#rm unneccesary data
-rm(models_bad)
-rm(models_degen)
-rm(models_good)
-
-load("../../4x4 model/written/sample.params.Rdata")
-params_degen <- list(main_hidden = sample.params %>% ungroup() %>% filter(near_hull) %>% select(starts_with("h"), -H) %>% data.matrix(),
-                     main_visible = sample.params %>% ungroup() %>% filter(near_hull) %>% select(starts_with("v"), -V) %>% data.matrix(),
-                     interaction = sample.params %>% ungroup() %>% filter(near_hull) %>% select(starts_with("theta")) %>% data.matrix() %>% matrix(4))
-params_good <- list(main_hidden = sample.params %>% ungroup() %>% filter(!near_hull) %>% select(starts_with("h"), -H) %>% data.matrix(),
-                    main_visible = sample.params %>% ungroup() %>% filter(!near_hull) %>% select(starts_with("v"), -V) %>% data.matrix(),
-                    interaction = sample.params %>% ungroup() %>% filter(!near_hull) %>% select(starts_with("theta")) %>% data.matrix() %>% matrix(4))
-
-
-load("../../4x4 model/written/sample_images.Rdata")
-
-#computing actual distributions ---------------------
-distn_good <- visible_distn(params = params_good)
-distn_degen <- visible_distn(params = params_degen)
-
[…]

### 2017-04-17 presentations/coms474/slides.Rmd
guest lecture

--- 2017-03-25 presentations/msu/slides.Rmd
@@
 date: |
-  | February 14, 2017
+  | April 17, 2017
   |
-  | Slides available at <http://bit.ly/kaplan-msu>
+  | Slides available at <http://bit.ly/kaplan-coms474>
   | 
   | \footnotesize Joint work with D. Nordman and S. Vardeman
-shortdate: "February 14, 2017"
+shortdate: "April 17, 2017"
 output: 
@@
 
-# RBM quantities to compare
-
-\begin{align*}
-\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] &= \sum\limits_{\boldsymbol x \in \mathcal{C}^{\nh + \nv}} \boldsymbol x \frac{\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{C}^{\nh + \nv}}\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)} \\
-~\\
-\text{E}\left[\boldsymbol X | \boldsymbol \theta^*\right] &= \sum\limits_{\boldsymbol x \in \mathcal{C}^{\nh + \nv}} \boldsymbol x \frac{\exp\left(\sum\limits_{i = 1}^\nv \theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{C}^{\nh + \nv}}\exp\left(\sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}
-\end{align*}
-
-# Data coding to mitigate degeneracy
-
-\begin{figure}
-\includegraphics[width=.55\linewidth]{../resources/images/toyhull_top.png}
-\includegraphics[width=.35\linewidth]{../resources/images/toyhull_top_2.png}
-\caption{The convex hull of the "statistic space" $\mathcal{T} = \{(v_1, h_1, v_1 h_1): v_1, h_1 \in \mathcal{C}\}$ in three dimensions for a toy RBM with one visible and one hidden node for $\mathcal{C} = \{0,1\}$ (left) and $\mathcal{C} = \{-1,1\}$ (right) data encoding.}
-\end{figure}
-
-The convex hull of $\mathcal{T} \subset \mathcal{C}^3$ does not fill the unit cube $[0,1]^3$ (left), but does better with $[-1,1]^3$ (right).
-
 # The center of the universe
@@
 - For the $\mathcal{C} = \{-1, 1 \}$ encoding of hiddens $(H_1, \dots, H_\nh)$ and visibles $(V_1, \dots, V_\nv)$, the origin is the center of the parameter space.
+\vspace{.6in}
 - At $\boldsymbol \theta = \boldsymbol 0$, RBM is equivalent to elements of $X$ being distributed as iid Bernoulli$\left(\frac{1}{2}\right)$ $\Rightarrow$ No *near-degeneracy*, *instability*, or *uninterpretability*!
-
-```{r encoding-volume, echo = FALSE, fig.height=3, fig.cap="\\label{fig:volume_plot}Relationship between volume of the convex hull of possible values of the RBM sufficient statistics and the cube containing it for different size models."}
-test_cases <- expand.grid(data.frame(1:4)[,rep(1,2)]) %>% 
-  rename(H = X1.4, V = X1.4.1) %>%
-  filter(H <= V) %>% #remove those cases with less visibles than hiddens. Is this necessary?
-  mutate(n_param = H*V + H + V) %>%
-  mutate(max_facets = (2^(H+V))^(floor(n_param/2))) %>%
-  filter(n_param <= 11) #calc_hull can't handle any higher dimensions currently
-
-bin_res <- plyr::dlply(test_cases, plyr::.(n_param), function(x) calc_hull(x$V, x$H, "binary"))
-neg_res <- plyr::dlply(test_cases, plyr::.(n_param), function(x) calc_hull(x$V, x$H, "negative"))
-
-plyr::ldply(bin_res, function(x) x$c_hull$vol) %>% 
-  mutate(frac_vol = V1/(1^n_param)) %>%
-  inner_join(plyr::ldply(neg_res, function(x) x$c_hull$vol) %>% 
-               mutate(frac_vol = V1/(2^n_param)),
-             by="n_param") %>%
-  rename(vol.bin = V1.x, vol.neg = V1.y, frac_vol.bin = frac_vol.x, frac_vol.neg = frac_vol.y) %>%
-  gather(vars, value, -n_param) %>%
-  separate(vars, c("type", "encoding"), "\\.") %>%
-  spread(type, value) %>%
-  ggplot() +
-  geom_point(aes(x=n_param, y=frac_vol, colour=encoding)) +
-  geom_line(aes(x=n_param, y=frac_vol, colour=encoding, group=encoding)) +
-  ylab("Fraction of unrestricted volume") +
-  xlab("Number of parameters") +
-  scale_colour_discrete("Encoding", labels=c("Binary (1,0)", "Negative (1,-1)")) +           
-  theme(
-    plot.background = element_rect(fill = "transparent", colour = NA),
-    legend.background = element_rect(fill = "transparent", colour = NA)) -> p
-
-p
-```
 
@@
 
-# Hyperparameters
-
-\begin{table}[ht]
-\centering
-\caption{The values used for the hyperparameters for all three fitting methods.
-A rule of thumb is imposed which decreases prior variances for the model parameters as the size of the model increases and also shrinks $\boldsymbol \theta_{interaction}$ more than $\boldsymbol \theta_{main}$.
-The common $C$ defining $C_1$ and $C_2$  in the BwTPLV method is chosen by tuning.}
-\label{tab:hyperparam}
-\begin{tabular}{|l|c|c|}
-\hline 
-Method & Hyperparameter & Value \\ 
-\hline \hline
-\multirow{2}{*}{BwTPLV} & $C_1$ & $\frac{C}{n}\frac{1}{\nh + \nv}$ \\
- & $C_2$ & $\frac{C}{n}\frac{1}{\nh*\nv}$ \\
-\hline
-\multirow{2}{*}{BwTNLV} & $\sigma^2_{main}$ & $\frac{1}{\nh + \nv}$ \\
- & $\sigma^2_{interaction}$ & $\frac{1}{\nh*\nv}$ \\
-\hline
-\multirow{2}{*}{BwTNML} & $\sigma^2_{main}$ & $\frac{1}{\nh + \nv}$ \\
- & $\sigma^2_{interaction}$ & $\frac{1}{\nh*\nv}$ \\
-\hline
-\end{tabular}
-\end{table}
-
 # Mixing
@@
 
-# Future work
-
-1. Generalization of instability results for other network models [ongoing, see @kaplan2016note]
-2. Image classification
-    - Ensemble methods (super learners) using AdaBoost [@freund1995desicion] 
-    - Decision theoretic based approach to approximating the likelihood ratio test for classification
-3. Markov chain Monte Carlo methods for data with Markovian dependence
-    - Spatial data
-    - Network data
-
 # Thank you
 
-* Slides -- <http://bit.ly/kaplan-msu>
+* Slides -- <http://bit.ly/kaplan-coms474>
 
@@
 
+# Appendix: Hyperparameters
+
+\begin{table}[ht]
+\centering
+\caption{The values used for the hyperparameters for all three fitting methods.
+A rule of thumb is imposed which decreases prior variances for the model parameters as the size of the model increases and also shrinks $\boldsymbol \theta_{interaction}$ more than $\boldsymbol \theta_{main}$.
+The common $C$ defining $C_1$ and $C_2$  in the BwTPLV method is chosen by tuning.}
+\label{tab:hyperparam}
+\begin{tabular}{|l|c|c|}
+\hline 
+Method & Hyperparameter & Value \\ 
+\hline \hline
+\multirow{2}{*}{BwTPLV} & $C_1$ & $\frac{C}{n}\frac{1}{\nh + \nv}$ \\
+ & $C_2$ & $\frac{C}{n}\frac{1}{\nh*\nv}$ \\
+\hline
+\multirow{2}{*}{BwTNLV} & $\sigma^2_{main}$ & $\frac{1}{\nh + \nv}$ \\
+ & $\sigma^2_{interaction}$ & $\frac{1}{\nh*\nv}$ \\
+\hline
+\multirow{2}{*}{BwTNML} & $\sigma^2_{main}$ & $\frac{1}{\nh + \nv}$ \\
+ & $\sigma^2_{interaction}$ & $\frac{1}{\nh*\nv}$ \\
+\hline
+\end{tabular}
+\end{table}
+
 # References {.allowframebreaks}

### 2017-05-10 writing/tech_submission/appendix/appendix-note.Rmd
forgot to add old files

--- 2016-12-01 writing/appendix/appendix-note.Rmd
@@
 Then, if $k > 0$, it follows that
-$$
-\text{ELPR}(\boldsymbol \theta_N) = \log\left[\frac{P_{\boldsymbol \theta_N}(\boldsymbol x_{max})}{P_{\boldsymbol \theta_N}(\boldsymbol x_{min})}\right] = \left|\sum\limits_{i = 1}^k\log\left(\frac{P_{\boldsymbol \theta_N}(\boldsymbol x_i)}{P_{\boldsymbol \theta_N}(\boldsymbol x_{i-1})}\right)\right| \le \sum\limits_{i = 1}^k\left|\log\left(\frac{P_{\boldsymbol \theta_N}(\boldsymbol x_i)}{P_{\boldsymbol \theta}(\boldsymbol x_{i-1})}\right)\right| \le k \Delta_N(\boldsymbol \theta_N) \le NC,
-$$
+\begin{align*}
+\text{ELPR}(\boldsymbol \theta_N) = \log\left[\frac{P_{\boldsymbol \theta_N}(\boldsymbol x_{max})}{P_{\boldsymbol \theta_N}(\boldsymbol x_{min})}\right] &= \left|\sum\limits_{i = 1}^k\log\left(\frac{P_{\boldsymbol \theta_N}(\boldsymbol x_i)}{P_{\boldsymbol \theta_N}(\boldsymbol x_{i-1})}\right)\right| \\
+&\le \sum\limits_{i = 1}^k\left|\log\left(\frac{P_{\boldsymbol \theta_N}(\boldsymbol x_i)}{P_{\boldsymbol \theta}(\boldsymbol x_{i-1})}\right)\right| \le k \Delta_N(\boldsymbol \theta_N) \le NC,
+\end{align*}
 using $k \le N$ and $\Delta(\boldsymbol \theta_N) \le C$. If $k = 0$, then $\boldsymbol x_{max} = \boldsymbol x_{min}$ and the same bound above holds.$\Box$

### 2017-05-10 writing/tech_submission/note.Rmd
forgot to add old files

--- 2016-12-02 writing/note.Rmd
@@
   - name: Andee Kaplan
-    affiliation: Iowa State University
+    university: Iowa State University
     department: Department of Statistics
   - name: Daniel Nordman
-    affiliation: Iowa State University
+    university: Iowa State University
     department: Department of Statistics
   - name: Stephen Vardeman
-    affiliation: Iowa State University
-    department: Department of Statistics
+    university: Iowa State University
+    department: |
+      | Department of Statistics and Department of Industrial and
+      | Manufacturing Systems Engineering
 output: 

### 2017-05-21 writing/draft.Rmd
some edits

--- 2016-12-02 writing/draft.Rmd
@@
 ---
-title: "On the propriety of restricted Boltzmann machines"
+title: "Properties and Bayes fitting of restricted Boltzmann machines"
 author: 
@@
 However, the generative behavior of RBMs is largely unexplored.
-In this paper, we discuss the relationship between RBM parameter specification in the binary case and the tendency to undesirable model properties such as degeneracy, instability and uninterpretability.
+In this paper, we discuss the relationship between RBM parameter specification in the binary case and the tendency to exhibit model properties such as degeneracy, instability and uninterpretability.
 We also describe the difficulties that arise in likelihood-based and Bayes fitting of such (highly flexible) models, especially as Gibbs sampling (quasi-Bayes) methods are often advocated for the RBM model structure.
@@
 These models have reportedly been used with success in classification of images [@larochelle2008classification; @srivastava2012multimodal].
-However, the model properties are largely unexplored in the literature and the commonly cited fitting methodology remains heuristic-based and abstruse [@hinton2006fast].
-In this paper, we provide steps toward a thorough understanding of the model class and its properties from the perspective of statistical theory, and then explore the possibility of a rigorous fitting methodology.
-We find the RBM model class to be deficient in two fundamental ways.
+However, the model properties are largely unexplored in the literature and the commonly cited fitting methodology remains heuristic-based and relies on rough approximation [@hinton2006fast].
+In this paper, we provide steps toward a thorough understanding of the model class and its properties from the perspective of statistical model theory, and then explore the possibility of a rigorous fitting methodology.
+We find the RBM model class to be concerning in two fundamental ways.
 
@@
 At issue, RBMs  often produce data lacking realistic variability and may thereby fail to provide a satisfactory conceptualization of a data generation process.
+For example, when generating data from a degenerate RBM, only a small number of output possibilities have probability greater than zero, and thus a sample of images will all be copies of the same one image (or small number of images).
 In addition to such degeneracy, we find that RBMs can easily exhibit types of instability.
-In practice, this may seen when a single pixel change in an image results in a wildly different classification in an image classification problem.
-Such model impropriety issues have recently been documented in RBMs [@li2014biclustering], as well as other deep architectures [@szegedy2013intriguing; @nguyen2014deep].
-We investigate these phenomena for RBMs in Section \ref{explorations} through simulations of small, manageable examples.
-
-Secondly, the fitting of these models is problematic.
+In practice, this may be seen when a single pixel change in an image results in a wildly different classification in an image classification problem.
+Such model properties have recently been documented in RBMs [@li2014biclustering], as well as other deep architectures [@szegedy2013intriguing; @nguyen2014deep].
+We investigate the presence of these properties for the RBM class of models in Section \ref{explorations-of-model-properties-through-simulation} through simulations of small, manageable examples.
+
+In addition to model properties, we also explore the issues with fitting these models.
+The fitting can be problematic for two reasons, the first being computational and the second concerning flexibility.
 As the size of these models grows, both maximum likelihood and Bayesian methods of fitting quickly become intractable.
@@
 As might be expected, with greater computational complexity comes an increase in fitting accuracy, but at the cost of practical feasibility.
+While the computational concerns are inconvenient, the flexibility issues are more alarming.
 
@@
 Thus, there can be no "smoothed distribution" achieved in fitting a RBM model of sufficient size with a rigorous likelihood-based method.
-We are therefore led to be skeptical that any model built using these structures (like a deep Boltzmann machine) can achieve useful prediction or inference in a principled way without intentionally limiting the flexibility of the fitted model.
+We are therefore led to be skeptical that models that involve these structures (like a deep Boltzmann machine) can achieve useful prediction or inference in a principled way without intentionally limiting the flexibility of the fitted model.
+This concern with overfitting is addressed via specification of a Bayesian prior in Section \ref{bayesian-model-fitting}.
 
@@
 Additionally, measures of model impropriety and methods of quantifying/detecting it are defined.
-Section \ref{explorations} details our explorations into the model behavior and propriety (or lack thereof) of the RBM class.
+Section \ref{explorations-of-model-properties-through-simulation} details our explorations into the model behavior and propriety (or lack thereof) of the RBM class.
 We discuss three Bayesian fitting techniques intended to avoid model impropriety in Section \ref{model-fitting} and conclude with a discussion in Section \ref{discussion}.
-On-line supplementary materials provide proofs for results on RBM parameterizations and data codings described in Section \ref{data-coding-to-mitigate-apparent-degeneracy}.
+On-line supplementary materials provide proofs for results on RBM parameterizations and data codings described in Section \ref{data-encoding}.
 
@@
 Each single binary random variable, visible or hidden, will take its values in a common coding set $\mathcal{C}$, where we allow one of two possibilities for the coding set, $\mathcal{C}=\{0,1\}$ or $\mathcal{C}=\{-1,1\}$, with "$1$" always indicating the "high" value of the variable.
-While $\mathcal{C}=\{0,1\}$ may be a natural starting point, we argue later that the coding $\mathcal{C}=\{-1,1\}$ induces more interpretable model properties for the RBM (cf.
-Section \ref{explorations}).
+While $\mathcal{C}=\{0,1\}$ may be a natural starting point, we argue in Section \ref{explorations-of-model-properties-through-simulation} that the coding $\mathcal{C}=\{-1,1\}$ induces more interpretable model properties for the RBM.
 A standard parametric form for probabilities corresponding to a potential vector of states, $X = (H_1, \dots, H_\nh, V_1,\dots,V_\nv)$, for the nodes is
@@
 
-# Explorations
+# Explorations of model properties through simulation
 
@@
 Because it seems most common, we shall begin by employing $0/1$ encoding of binary variables (both $h_1$ and $v_1$ taking values in $\mathcal{C} = \{0,1\}$).
-(Eventually we shall argue in Section \ref{data-coding-to-mitigate-apparent-degeneracy} that $-1/1$ coding has substantial advantages.)
+(Eventually we shall argue in Section \ref{data-encoding} that $-1/1$ coding has substantial advantages.)
 
@@
 
-### Data coding to mitigate apparent degeneracy
+### Data encoding
 Multiple encodings of the binary variables are possible.
@@
 This will result in variables $\boldsymbol t(\boldsymbol X)$ from (\ref{eqn:t}) satisfying $\boldsymbol t(\boldsymbol x) \in \{0,1\}^{\nh + \nv + \nh*\nv}$ or $\boldsymbol t(\boldsymbol x) \in \{-1,1\}^{\nh + \nv + \nh*\nv}$ depending on how we encode "on" and "off" states in the nodes.
-To explore the effect of this encoding on the potential for apparent near-degeneracy of a RBM, we can consider the ratio of the volume of the hypercube with corners at elements of $\mathcal{T}$ to the volume of the convex hull of $\mathcal{T}$ under both possible encodings (i.e., compare convex hull of $\mathcal{T}$ to either $[0,1]^3$ or $[-1,1]^3$ under the respective encodings).
-
-For the small two node example, the $0/1$ encoding loses `r round((1-m1_bin$c_hull$vol)*100,2)`% of the volume of $[0,1]^3$ and the $-1/1$ encoding loses `r round((1-m1_neg$c_hull$vol/(2^3))*100,2)`% of the volume of $[-1,1]^3$.
-While there is clearly a one-to-one mapping between models for these two possible codings, this notion of lost volume can be helpful to geometrically conceptualize how difficult it will be for the mean vector $\boldsymbol \mu(\boldsymbol \theta)$ to avoid the boundary of $\mathcal{T}$, and thus avoid apparent near-degeneracy.
-In fact, if we look at the ratio of volume within the convex hull defined by $\mathcal{T}$ and the corresponding hypercube, we can see a relationship emerge as the number of nodes (and thus parameters) increases.
-In Figure \ref{fig:volume_plot}, it is evident that as the number of parameters increases, this ratio is decreasing at an increasing rate, meaning it gets more and more difficult to avoid the boundary of the convex hull and thus appearance of near degeneracy.
-Additionally, it appears that the $-1/1$ encoding suffers slightly less from this problem.
-This suggests that if intuition about models is to be formed in terms of a data encoding that encourages of "non-pathological" means $\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta}(\boldsymbol t(\boldsymbol x))$ for $\boldsymbol t(\boldsymbol x)$, then the $-1/1$ encoding is preferable to $0/1$ coding.
-
-```{r volume_plot, echo=FALSE, fig.cap="\\label{fig:volume_plot}The relationship between volume of the convex hull of possible values of the RBM sufficient statistics and the cube containing it for different configurations of nodes.", out.height = "3in", fig.align='center'}
-test_cases <- expand.grid(data.frame(1:4)[,rep(1,2)]) %>% 
-  rename(H = X1.4, V = X1.4.1) %>%
-  filter(H <= V) %>% #remove those cases with less visibles than hiddens. Is this necessary?
-  mutate(n_param = H*V + H + V) %>%
-  mutate(max_facets = (2^(H+V))^(floor(n_param/2))) %>%
-  filter(n_param <= 11) #calc_hull can't handle any higher dimensions currently
-
-bin_res <- plyr::dlply(test_cases, plyr::.(n_param), function(x) calc_hull(x$V, x$H, "binary"))
-neg_res <- plyr::dlply(test_cases, plyr::.(n_param), function(x) calc_hull(x$V, x$H, "negative"))
-
-plyr::ldply(bin_res, function(x) x$c_hull$vol) %>% 
-  mutate(frac_vol = V1/(1^n_param)) %>%
-  inner_join(plyr::ldply(neg_res, function(x) x$c_hull$vol) %>% 
-               mutate(frac_vol = V1/(2^n_param)),
-             by="n_param") %>%
-   rename(vol.bin = V1.x, vol.neg = V1.y, frac_vol.bin = frac_vol.x, frac_vol.neg = frac_vol.y) %>%
-  gather(vars, value, -n_param) %>%
-  separate(vars, c("type", "encoding"), "\\.") %>%
-  spread(type, value) %>%
-  ggplot() +
-  geom_point(aes(x=n_param, y=frac_vol, colour=encoding)) +
-  geom_line(aes(x=n_param, y=frac_vol, colour=encoding, group=encoding)) +
-  ylab("Fraction of unrestricted volume") +
-  xlab("Number of parameters") +
-  scale_colour_discrete("Encoding", labels=c("Binary (1,0)", "Negative (1,-1)"))
-```
-
-In addition to the argument that the $-1/1$ data encoding mitigates some perceived prevalence of near-degeneracy it also has the benefit of providing a guaranteed-to-be non-degenerate model at $\boldsymbol \theta = \boldsymbol 0 \in \mathbb{R}^{\nh + \nv + \nh*\nv}$, where the zero vector then serves as the natural center of the parameter space and induces the simplest possible model properties for the RBM (i.e., at $\boldsymbol \theta = \boldsymbol 0$, all variables are independent and visible variables are independent and uniformly distributed on $\{-1,1\}^\nv$).
+
+The $-1/1$ data encoding has the benefit of providing a guaranteed-to-be non-degenerate model at $\boldsymbol \theta = \boldsymbol 0 \in \mathbb{R}^{\nh + \nv + \nh*\nv}$, where the zero vector then serves as the natural center of the parameter space and induces the simplest possible model properties for the RBM (i.e., at $\boldsymbol \theta = \boldsymbol 0$, all variables are independent and visible variables are independent and uniformly distributed on $\{-1,1\}^\nv$).
 The proof of this and further exploration of the equivalence of the $\boldsymbol \theta$ parameterization of the RBM model class and parameterization by $\boldsymbol \mu(\boldsymbol \theta)$ is in the on-line supplementary materials.
@@
 In fact, when the cell probabilities (\ref{eqn:0}) are all strictly positive, parameters in the RBM model can be specified to reflect these probabilities exactly.
-And, when one or more of the cell probabilities (\ref{eqn:0}) are zero, the corresponding RBM  probabilities $P(V_1=v_1,V_2=_v2| \theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21})$ may never be identically zero (due to exponential terms in the model) but parameters can be still selected to make the appropriate RBM cell probabilities arbitrarily small.
+And, when one or more of the cell probabilities (\ref{eqn:0}) are zero, the corresponding RBM  probabilities $P(V_1=v_1,V_2= v_2| \theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21})$ may never be identically zero (due to exponential terms in the model) but parameters can be still selected to make the appropriate RBM cell probabilities arbitrarily small.
 
@@
 In data applications with a large sample space, it is unlikely that the training set will include at least one of each possible vector outcome (unlike this small example).
-This implies that some RBM model parameters must diverge to $+\infty$ to mimic the empirical distribution with empty cells and, as we have already discussed in Section \ref{explorations}, large magnitudes of $\boldsymbol \theta$ lead to model impropriety in the RBM.
+This implies that some RBM model parameters must diverge to $+\infty$ to mimic the empirical distribution with empty cells and, as we have already discussed in Section \ref{explorations-of-model-properties-through-simulation}, large magnitudes of $\boldsymbol \theta$ lead to model impropriety in the RBM.
 
@@
 1.
-*A "trick" prior.* Here we cancel out normalizing term in the likelihood so that resulting full conditionals of $\boldsymbol \theta$ are multivariate Normal.
+*A "trick" prior.* Here we cancel out normalizing term in the likelihood so that resulting full conditionals of $\boldsymbol \theta$ are multivariate normal.
 The $h_j$ are carried along in the MCMC sampling from the posterior as latent variables.
@@
 Marginalization over the latent variables in the model can improve mixing, but is numerically intractable due to the necessity of repeated calculation of the normalizing constant.
+Due to the extreme flexibility in this model class, rigorous likelihood-based fitting for a RBM will typically merely return the (discrete) empirical distribution for visibles, meaning any practitioner should be aware of this and employ some form of regularization or penalization in the model.
 
 Ultimately, it is not clear that RBM models are useful as generative models.
-Due to the extreme flexibility in this model class, rigorous likelihood-based fitting for a RBM will typically merely return the (discrete) empirical distribution for visibles.
-Even if a rigorous likelihood-based fitting method is practically possible, we know what it will produce for fitted probabilities before we even begin.
+Without the generative behavior of the RBM model, the ability to quantify uncertainty in the estimated model parameters becomes impossible and it is no longer useful as a predictive distribution.
+In the case of classification this may not be the ultimate goal, however when S-instability is present, small (imperceptable) differences in the data can lead to greatly different probabilities and thus greatly different classifications.
 For these reasons, we are skeptical of the claims made about RBMs as generative tools.

### 2017-05-24 writing/appendix/appendix-center.Rmd
add simulation example to intro

--- 2016-11-21 writing/appendix/appendix-center.Rmd
@@
 title: "Supplementary material to:  On the propriety of restricted Boltzmann machines"
-author: | 
-  | Andee Kaplan, Daniel Nordman, and Stephen Vardeman
-  | Iowa State University
-  | ajkaplan@iastate.edu, dnordman@iastate.edu, vardeman@iastate.edu
-date: "November, 2016"
-linestretch: 2 
+author: 
+  - name: Andee Kaplan
+    university: Iowa State University
+    department: Department of Statistics
+  - name: Daniel Nordman
+    university: Iowa State University
+    department: Department of Statistics
+  - name: Stephen Vardeman
+    university: Iowa State University
+    department: |
+      | Department of Statistics and Department of Industrial and
+      | Manufacturing Systems Engineering
 output: 
   pdf_document:
+    keep_tex: true
     fig_caption: yes
@@
     number_sections: yes
+    template: ../latex/jasa_article.tex
+blind: false
 ---

### 2017-05-24 writing/draft.Rmd
add simulation example to intro

@@
     number_sections: yes
-    template: latex/technometrics_article.tex
-fontsize: 12pt
+    template: latex/jasa_article.tex
 bibliography: references/refs.bib
@@
 
+set.seed(1022)
+
 opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE)
@@
 # Introduction
+
+```{r sample-models, fig.show='hide', fig.height=2}
+load("data/params_theta.Rdata")
+params_good <- list(main_hidden = sample.params %>% ungroup() %>% filter(!near_hull) %>% select(starts_with("h"), -H) %>% data.matrix(),
+                    main_visible = sample.params %>% ungroup() %>% filter(!near_hull) %>% select(starts_with("v"), -V) %>% data.matrix(),
+                    interaction = sample.params %>% ungroup() %>% filter(!near_hull) %>% select(starts_with("theta")) %>% data.matrix() %>% matrix(4))
+params_bad <- list(main_hidden = sample.params %>% ungroup() %>% filter(near_hull) %>% select(starts_with("h"), -H) %>% data.matrix(),
+                    main_visible = sample.params %>% ungroup() %>% filter(near_hull) %>% select(starts_with("v"), -V) %>% data.matrix(),
+                    interaction = sample.params %>% ungroup() %>% filter(near_hull) %>% select(starts_with("theta")) %>% data.matrix() %>% matrix(4))
+
+distn_good <- visible_distn(params_good)
+distn_bad <- visible_distn(params_bad)
+
+sample_good <- sample_n(distn_good, size = 10, replace = TRUE, weight = prob)
+sample_bad <- sample_n(distn_bad, size = 10, replace = TRUE, weight = prob)
+
+sample_good %>%
+  mutate(image = 1:n()) %>%
+  gather(pixel, value, -prob, -image_id, -image) %>%
+  mutate(value = factor(value)) %>%
+  mutate(x = ifelse(pixel %in% c("v1", "v3"), 1, 2),
+         y = ifelse(pixel %in% c("v1", "v2"), 1, 2)) %>%
+  ggplot() +
+  geom_tile(aes(x = x, y = y, fill = value)) +
+  facet_wrap(~image, nrow = 2) +
+  scale_fill_manual(values = c("white", "black")) +
+  theme(axis.line=element_blank(), axis.text.x=element_blank(),
+        axis.text.y=element_blank(), axis.ticks=element_blank(),
+        axis.title.x=element_blank(),
+        axis.title.y=element_blank(), legend.position="none",
+        panel.background=element_blank(), panel.grid.major=element_blank(),
+        panel.grid.minor=element_blank(), plot.background=element_blank(),
+        panel.border = element_rect(colour = "black", fill=NA, size=1),
+        strip.background = element_blank(),
+        strip.text.x = element_blank()) +
+  ggtitle("(b)")
+
+sample_bad %>%
+  mutate(image = 1:n()) %>%
+  gather(pixel, value, -prob, -image_id, -image) %>%
+  mutate(value = factor(value)) %>%
+  mutate(x = ifelse(pixel %in% c("v1", "v3"), 1, 2),
+         y = ifelse(pixel %in% c("v1", "v2"), 1, 2)) %>%
+  ggplot() +
+  geom_tile(aes(x = x, y = y, fill = value)) +
+  facet_wrap(~image, nrow = 2) +
+  scale_fill_manual(values = c("white", "black")) +
+  theme(axis.line=element_blank(), axis.text.x=element_blank(),
+        axis.text.y=element_blank(), axis.ticks=element_blank(),
+        axis.title.x=element_blank(),
+        axis.title.y=element_blank(), legend.position="none",
+        panel.background=element_blank(), panel.grid.major=element_blank(),
+        panel.grid.minor=element_blank(), plot.background=element_blank(),
+        panel.border = element_rect(colour = "black", fill=NA, size=1),
+        strip.background = element_blank(),
+        strip.text.x = element_blank()) +
+  ggtitle("(a)")
+```
 
@@
 For example, when generating data from a degenerate RBM, only a small number of output possibilities have probability greater than zero, and thus a sample of images will all be copies of the same one image (or small number of images).
+An example of ten 4-pixel images simulated from a degenerate model is compared to ten 4-pixel images simulated from a non-degenerate model in Figure \ref{fig:sample-models}.
+The degenerate model places almost all probability on one outcome, causing the image to be generated repeatedly, whereas the non-degenerate model shows more realistic variation.
+\begin{figure}
+\includegraphics{draft_files/figure-latex/sample-models-2.pdf} \\[.1in]
+\includegraphics{draft_files/figure-latex/sample-models-1.pdf}
+\caption{Ten 4-pixel images simulated from a degenerate model (a) compared to ten 4-pixel images simulated from a non-degenerate model (b).
+The degenerate model places almost all probability on one outcome, causing the image to be generated repeatedly, whereas the non-degenerate model shows more realistic variation.}
+\label{fig:sample-models}
+\end{figure}
+
 In addition to such degeneracy, we find that RBMs can easily exhibit types of instability.
@@
 However, the empirical distribution of a training set of vectors of visibles is the best fitting model for observed cell data.
-As a consequence, we find that any fully principled fitting method based on the likelihood for a RBM with enough hidden variables will simply seek to reproduce the (discrete) empirical distribution of a training set.
+As a consequence, we find that any fully principled fitting method based on the likelihood for a RBM with enough hidden variables will seek to reproduce the (discrete) empirical distribution of a training set.
 Thus, there can be no "smoothed distribution" achieved in fitting a RBM model of sufficient size with a rigorous likelihood-based method.
@@
 ```{r param-table}
-load("data/params_theta.Rdata")
-
-params_good <- list(main_hidden = sample.params %>% ungroup() %>% filter(!near_hull) %>% select(starts_with("h"), -H) %>% data.matrix(),
-                    main_visible = sample.params %>% ungroup() %>% filter(!near_hull) %>% select(starts_with("v"), -V) %>% data.matrix(),
-                    interaction = sample.params %>% ungroup() %>% filter(!near_hull) %>% select(starts_with("theta")) %>% data.matrix() %>% matrix(4))
-
 old_pen <- options()$scipen
@@
 Without the generative behavior of the RBM model, the ability to quantify uncertainty in the estimated model parameters becomes impossible and it is no longer useful as a predictive distribution.
-In the case of classification this may not be the ultimate goal, however when S-instability is present, small (imperceptable) differences in the data can lead to greatly different probabilities and thus greatly different classifications.
+In the case of classification this may not be the ultimate goal, however when S-instability is present, small (imperceptible) differences in the data can lead to greatly different probabilities and thus greatly different classifications.
 For these reasons, we are skeptical of the claims made about RBMs as generative tools.

### 2017-05-31 writing/note.Rmd
format note

--- 2016-12-02 writing/note.Rmd
@@
     number_sections: yes
-    template: latex/technometrics_article.tex
+    template: latex/jasa_article.tex
 fontsize: 12pt

### 2017-06-13 writing/draft.Rmd
small tweeks

@@
 library(tidyr)
+library(xtable)
 
@@
 
-```{r param-table}
+```{r param-table, results='asis'}
 old_pen <- options()$scipen
@@
 cbind(tbl[1:8,], tbl[9:16,], tbl[17:24,]) %>%
-  kable(caption = "\\label{tab:theta}Parameters used to fit a test case with $V = H = 4$.
-This parameter vector was chosen as a sampled value of $\\boldsymbol \\theta$ that was not near the convex hull of the sufficient statistics for a grid point in Figure \\ref{fig:degen_plots} with $< 5$\\% near-degeneracy.", align = rep(c("r", "r"), times = 3))
+  xtable(label = "tab:theta", 
+         caption = "Parameters used to fit a test case with $V = H = 4$.
+This parameter vector was chosen as a sampled value of $\\boldsymbol \\theta$ that was not near the convex hull of the sufficient statistics for a grid point in Figure \\ref{fig:degen-plots} with $< 5$\\% near-degeneracy.", align = c("l", rep(c("r", "r"), times = 3))) %>%
+  print(sanitize.text.function=function(x){x}, include.rownames = FALSE, comment = FALSE)
+
 ```
@@
 \centering
-\caption{The values used for the hyperparameters for all three fitting methods.
-A rule of thumb is imposed which decreases prior variances for the model parameters as the size of the model increases and also shrinks $\boldsymbol \theta_{interaction}$ more than $\boldsymbol \theta_{main}$.
-The common $C$ defining $C_1$ and $C_2$  in the BwTPLV method is chosen by tuning.}
-\label{tab:hyperparam}
 \begin{tabular}{|l|c|c|}
@@
 \end{tabular}
+\caption{The values used for the hyperparameters for all three fitting methods.
+A rule of thumb is imposed which decreases prior variances for the model parameters as the size of the model increases and also shrinks $\boldsymbol \theta_{interaction}$ more than $\boldsymbol \theta_{main}$.
+The common $C$ defining $C_1$ and $C_2$  in the BwTPLV method is chosen by tuning.}
+\label{tab:hyperparam}
 \end{table}
@@
 
-```{r m_eff_table}
+```{r m_eff_table, results='asis'}
 M_eff_table %>%
@@
   select(Outcome, BwTNLV, BwTNML) %>%
-  arrange(Outcome) -> M_eff_tmp
+  arrange(Outcome) %>%
+  mutate(Outcome = as.character(Outcome)) -> M_eff_tmp
 
 cbind(M_eff_tmp[1:8,], M_eff_tmp[9:16,]) %>%
-  kable(digits = 2, caption = paste0("\\label{tab:m_eff}The effective sample sizes for a chain of length $M = 1000$ regarding all $16$ probabilities for possible vector outcomes of visibles.
-BwTNLV would require at least $", round(min(M_eff_tmp[, "BwTNML"])/min(M_eff_tmp[, "BwTNLV"]), 1), "$ times as many MCMC iterations to achieve the same amount of effective information about the posterior distribution."))
-```
-
-The effective sample sizes for a chain of length $M = 1000$ for inference about each of the $2^4 = 16$ model probabilities are presented in Table \ref{tab:m_eff}.
+  xtable(label = "tab:m-eff", 
+         digits = 2, caption = paste0("The effective sample sizes for a chain of length $M = 1000$ regarding all $16$ probabilities for possible vector outcomes of visibles.
+BwTNLV would require at least $", round(min(M_eff_tmp[, "BwTNML"])/min(M_eff_tmp[, "BwTNLV"]), 1), "$ times as many MCMC iterations to achieve the same amount of effective information about the posterior distribution.")) %>%
+  print(include.rownames = FALSE, comment = FALSE)
+```
+
+The effective sample sizes for a chain of length $M = 1000$ for inference about each of the $2^4 = 16$ model probabilities are presented in Table \ref{tab:m-eff}.
 These range from  $`r round(min(M_eff_tmp[, "BwTNML"]), 2)`$ to $`r round(max(M_eff_tmp[, "BwTNML"]), 2)`$ for BwTNML, while BwTNLV only yields between $`r round(min(M_eff_tmp[, "BwTNLV"]), 2)`$ and $`r round(max(M_eff_tmp[, "BwTNLV"]), 2)`$ effective draws.

### 2017-06-13 writing/draft.Rmd
stupid cross-ref

@@
          caption = "Parameters used to fit a test case with $V = H = 4$.
-This parameter vector was chosen as a sampled value of $\\boldsymbol \\theta$ that was not near the convex hull of the sufficient statistics for a grid point in Figure \\ref{fig:degen-plots} with $< 5$\\% near-degeneracy.", align = c("l", rep(c("r", "r"), times = 3))) %>%
+This parameter vector was chosen as a sampled value of $\\boldsymbol \\theta$ that was not near the convex hull of the sufficient statistics for a grid point in Figure \\ref{fig:degen_plots} with $< 5$\\% near-degeneracy.", align = c("l", rep(c("r", "r"), times = 3))) %>%
   print(sanitize.text.function=function(x){x}, include.rownames = FALSE, comment = FALSE)

### 2017-07-08 writing/paper-note/appendices/appendix-note.Rmd
paper to bookdown

--- 2017-05-10 writing/tech_submission/appendix/appendix-note.Rmd
@@
-{\bf Proof of Proposition \ref{prop:instab}.} We prove the contrapositive, supposing that $\Delta(\boldsymbol \theta_N) \le C$ holds for some $C > 0$ and show $\text{ELPR}(\boldsymbol \theta_N) \leq NC$.
+\clearpage
+
+# (APPENDIX) Appendix {-}
+
+# Proofs of instability results {#appendix-instab}
+
+**Proof of Proposition \@ref(prp:instab-elpr).** We prove the contrapositive, supposing that $\Delta(\boldsymbol \theta_N) \le C$ holds for some $C > 0$ and show $\text{ELPR}(\boldsymbol \theta_N) \leq NC$.
 Let $\boldsymbol x_{min} \equiv \argmin\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)$ and $\boldsymbol x_{max} \equiv \argmax\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)$.
@@
 \end{align*}
-using $k \le N$ and $\Delta(\boldsymbol \theta_N) \le C$. If $k = 0$, then $\boldsymbol x_{max} = \boldsymbol x_{min}$ and the same bound above holds.$\Box$
+using $k \le N$ and $\Delta(\boldsymbol \theta_N) \le C$. If $k = 0$, then $\boldsymbol x_{max} = \boldsymbol x_{min}$ and the same bound above holds.\hfill $\Box$
 
-{\bf Proof of Proposition \ref{prop:degen}.} Define $\boldsymbol x_{max}$ and $\boldsymbol x_{min} \in \mathcal{X}^N$ as in the proof of Proposition \ref{prop:instab} where $|\mathcal{X}|<\infty$ holds in the FSFS model.
+**Proof of Proposition \@ref(prp:degenFSFS).** where $|\mathcal{X}|<\infty$ holds in the FSFS model.
 We may suppose $|\mathcal{X}|>1$ (i.e., $\mathcal{X}^N$ has more than one outcome) because otherwise the model is trivially degenerate for all $N \geq 1$.
@@
 \end{align*}
-as $N \rightarrow \infty$ by the definition of an unstable FSFS model (c.f.
-Definition \ref{defn:instab}).
-Consequently, $P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N}) \rightarrow 1$ as $N \rightarrow \infty$ as claimed.$\Box$
+as $N \rightarrow \infty$ by the definition of an unstable FSFS model (cf.
+Definition \@ref(def:instabFSFS)).
+Consequently, $P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N}) \rightarrow 1$ as $N \rightarrow \infty$ as claimed.
+\hfill $\Box$
+
+\clearpage
+
+# References {-}

### 2017-07-08 writing/paper-note/chapters/instab.Rmd
paper to bookdown

--- 2016-12-02 writing/note.Rmd
@@
----
-title: "A note on the instability and degeneracy of deep learning models"
-author: 
-  - name: Andee Kaplan
-    affiliation: Iowa State University
-    department: Department of Statistics
-  - name: Daniel Nordman
-    affiliation: Iowa State University
-    department: Department of Statistics
-  - name: Stephen Vardeman
-    affiliation: Iowa State University
-    department: Department of Statistics
-output: 
-  pdf_document:
-    keep_tex: true
-    fig_caption: yes
-    includes:
-      in_header: latex/header.tex
-    number_sections: yes
-    template: latex/technometrics_article.tex
-fontsize: 12pt
-bibliography: references/refs.bib
-abstract: |
-  A probability model exhibits instability if small changes in a data outcome result in large, and often unanticipated, changes in probability.
-For correlated data structures found in several application areas, there is increasing interest in predicting/identifying instability.
-We consider the problem of quantifying instability for general probability models defined on sequences of observations, where each sequence of length $N$ has a finite number of possible outcomes.
-(A sequence of probability models results indexed by $N$ that accommodates data of expanding dimension.) Model instability is formally shown to occur when a certain log-probability ratio under such models grows faster than $N$.
-In this case, a one component change in the data sequence can shift probability by orders of magnitude.
-Also, as a measure of instability becomes more extreme, the resulting probability models are shown  to tend to degeneracy, placing all their probability on arbitrarily small portions of the sample space.
-These results on instability apply to large classes of models commonly used in random graphs, network analysis, and machine learning contexts.
-keywords: Degeneracy, Instability, Deep Learning, Graphical Models
-blind: false
----
-
 # Introduction
@@
 Let $(X_1, \dots, X_N)$ denote a set of discrete random variables with a finite sample space, $\mathcal{X}^N$.
-That is, $\mathcal{X}$ with $|\mathcal{X}| < \infty$ represents a finite set of potential outcomes for the variable $X_i$, and the data sequence $(X_1,\ldots,X_N)$ takes values in the $N$-fold product space $\mathcal{X}^N$.
+That is, $\mathcal{X}$ with $|\mathcal{X}| < \infty$ represents a finite set of potential outcomes for each single variable $X_i$, and the data sequence $(X_1,\ldots,X_N)$ takes values in the $N$-fold product space $\mathcal{X}^N$.
 For each $N$, let $P_{\boldsymbol \theta_N}$ denote a probability model on $\mathcal{X}^N$, under which $P_{\boldsymbol \theta_N}(x_1, \dots, x_N) > 0$ is the probability of the data outcome $(x_1, \dots, x_N) \in \mathcal{X}^N$.
-We assume that the model support of $P_{\boldsymbol \theta_N}$ is the sample space $\mathcal{X}^N$.
+In this, we assume that the model support of $P_{\boldsymbol \theta_N}$ is the sample space $\mathcal{X}^N$.
 This framework produces a series $P_{\boldsymbol \theta_N}$ of probability models, indexed by a generic sequence of parameters $\boldsymbol \theta_N$, to describe data of each length $N \geq 1$.
-(The size and structure of such parameters are without restriction, and natural cases include those where $\boldsymbol \theta_N \in \mathbb{R}^{q(N)}$ for some arbitrary integer-valued function $q(\cdot) \geq 1$.) We will call this model class *Finitely Supported Finite Sequence (FSFS) models*.
+The size and structure of such parameters are without restriction, and natural cases include those where $\boldsymbol \theta_N \in \mathbb{R}^{q(N)}$ for some arbitrary integer-valued function $q(\cdot) \geq 1$.
+We will refer to this model class as *Finitely Supported Finite Sequence (FSFS) models*.
 
-Section \ref{examples} provides several examples of FSFS models commonly used in graph/network analysis and machine learning (i.e., deep learning models).
-Section \ref{instability-results} establishes formal results regarding the propriety of FSFS models with regard to stability.
+Section \@ref(examples) provides several examples of FSFS models commonly used in graph/network analysis and machine learning (i.e., deep learning models).
+Section \@ref(instability-results) establishes formal results regarding the propriety of FSFS models with regard to stability.
 A FSFS probability model sequence exhibits instability if small changes in the components of a data outcome $(x_1,\ldots,x_N)$ can result in large changes in probability $P_{\theta_N}(x_1,\ldots,x_N)$.
-The concept of instability, introduced in the field of statistical physics by @ruelle1999statistical, was extended to include a notion of detection and quantification for certain exponential
-family models by @schweinberger2011instability.
+The concept of instability, introduced in the field of statistical physics by @ruelle1999statistical, was extended to include a notion of detection and quantification for certain exponential family models by @schweinberger2011instability.
 For similar exponential models, particularly in connection to random graphs/networks, @handcock2003assessing considered (mean-based) characterizations for so-called model degeneracy, whereby a probability model places all mass on a small subset of the sample space and produces undesirably low variability in model outcomes.
@@
 The main results here concern a general measure of model instability, appropriate across the whole FSFS model class.
-This can be used to identify when certain maximal probabilities in FSFS models are too extreme relative to the length $N$.
+This can be used to identify when certain maximal probabilities in FSFS models are too extreme relative to the length $N$ and may thereby induce a potentially undesirable probability structure.
 In this case, a one component change in the data sequence may shift probability by orders of magnitude, and FSFS models are rigorously shown to become degenerate as the measure of instability increases.
-Lastly, Section \ref{implications} emphasizes the implications of our model propriety results and proofs of the main results appear in the Appendix.
+Lastly, Section \@ref(implications) emphasizes the implications of our model propriety results and proofs of the main results appear in Appendix \@ref(appendix-instab).
 
@@
 
-For $\boldsymbol X = (X_1, \dots, X_N)$ discrete random variables with sample space $\mathcal{X}^N$, $|\mathcal{X}| < \infty$, consider an exponential family model for $\boldsymbol X$ with probability mass function of the form
+For discrete random variables $\boldsymbol X = (X_1, \dots, X_N)$ with sample space $\mathcal{X}^N$, $|\mathcal{X}| < \infty$, consider an exponential family model for $\boldsymbol X$ with probability mass function of the form
 $$
@@
 $$ 
-for fixed positive integers $k$ and $L$ denoting the dimensions of the parameter, $\boldsymbol \lambda \in \Lambda \subset \mathbb{R}^{k}$ and natural parameter $\boldsymbol \eta : \mathbb{R}^k \mapsto \mathbb{R}^L$ spaces, $\boldsymbol g_N : \mathcal{X}^N \mapsto \mathbb{R}^L$ a vector of sufficient statistics,
+depending on parameter vector $\boldsymbol \lambda \in \Lambda \subset \mathbb{R}^{k}$ and natural parameter function $\boldsymbol \eta : \mathbb{R}^k \mapsto \mathbb{R}^L$ with fixed positive integers $k$ and $L$ denoting their dimensions.
+Above, $\boldsymbol g_N : \mathcal{X}^N \mapsto \mathbb{R}^L$ is a vector of sufficient statistics, while
 $$
@@
 $$
-the normalizing function, and $\Lambda = \{\boldsymbol \lambda \in \mathbb{R}^k : \psi(\boldsymbol \lambda) < \infty, k \le q(N) \}$ is the parameter space.
+denotes the normalizing function, and $\Lambda = \{\boldsymbol \lambda \in \mathbb{R}^k : \psi(\boldsymbol \lambda) < \infty, k \le q(N) \}$ is the parameter space.
 
 Defining $P_{\boldsymbol \theta_N}(\boldsymbol x)\equiv p_{N,\boldsymbol \lambda_N}(\boldsymbol x)$ with $\boldsymbol \theta_N=\boldsymbol \lambda_N$ to be a sequence of elements of $\Lambda \subset \mathbb{R}^k$ and noting that $P_{\boldsymbol \theta_N}(\boldsymbol x) > 0$ for all $\boldsymbol x \in \mathcal{X}^N$, these discrete exponential family models are special cases of the FSFS models.
-Such exponential models arise with spatial data on a lattice [@besag1974spatial], network data [@wasserman1994social; @handcock2003assessing], and even binomial sampling with $N$ iid Bernoulli random variables.
+Such exponential models arise with spatial data on a lattice [@besag1974spatial], network data [@wasserman1994social; @handcock2003assessing], and even standard independence models for discrete data, such as with $N$ iid Bernoulli random variables.
 (Note that for random graphs or networks with, say, $m$ nodes, one may wish to consider $N={m \choose 2}$ edges as binary (presence/absence) variables $X_i$.
-In this case, the length $N$ of data sequence may naturally increase as a function of $m$.) For these exponential models, the dimension of the parameter $\boldsymbol \theta_N$ is the same for each $N$ as $\boldsymbol \theta_N$ lies in a parameter space of fixed Euclidean dimension $k$.
-This need not be true for other types of FSFS models considered in Sections \ref{restricted-boltzmann-machines}-\ref{deep-learning}.
-@schweinberger2011instability considered instability in such exponential models (e.g., for random graphs) for sequences of fixed parameters $\boldsymbol \theta_N=\boldsymbol \lambda\in\mathbb{R}^k$, $N \geq 1$, of non-varying dimension $k$.
+In this case, the length $N$ of data sequence may naturally increase as a function of $m$.) For these exponential models, the dimension of the parameter $\boldsymbol \theta_N$ remains constant over each $N$, as $\boldsymbol \theta_N$ lies in a parameter space of fixed Euclidean dimension $k$.
+This need not be true for other types of FSFS models considered in Sections \@ref(restricted-boltzmann-machines)-\@ref(deep-learning).
+@schweinberger2011instability considered instability in such exponential models (e.g., for random graphs) for sequences of fixed parameters $\boldsymbol \theta_N=\boldsymbol \lambda\in\mathbb{R}^k$, $N \geq 1$.
 
@@
 
-A restricted Boltzmann machine (RBM) is an undirected graphical model specified for discrete or continuous random variables, binary variables being most common [cf.
+A restricted Boltzmann machine (RBM) is an undirected graphical model specified for discrete or continuous random variables, with binary variables being most common [cf.
 @smolensky1986information].
@@
 Let $\boldsymbol X=(X_1,\ldots,X_N)$ denote the $N$ random variables for visibles with support $\mathcal{X}^N$ and $\boldsymbol H=(H_1,\ldots,H_{N_H})$ denote the $N_H$ random variables for hiddens with support $\mathcal{X}^{N_H}$ where  $\mathcal{X} = \{-1,1\}$.
-For parameters $\boldsymbol \alpha \in \mathbb{R}^{N_H}$, $\boldsymbol \beta \in \mathbb{R}^N$, and $\Gamma$ as a matrix with dimension $N_H \times N$, the RBM for $\tilde{\boldsymbol X}=(\boldsymbol X,\boldsymbol H)$ then has the joint probability mass function
+For parameters $\boldsymbol \alpha \in \mathbb{R}^{N_H}$, $\boldsymbol \beta \in \mathbb{R}^N$, and $\Gamma$ as a matrix with dimension $N_H \times N$, the RBM model for $\tilde{\boldsymbol X}=(\boldsymbol X,\boldsymbol H)$ then has the joint probability mass function
 $$
@@
 Let $\boldsymbol \theta_N = (\boldsymbol \alpha, \boldsymbol \beta,\Gamma) \in \Theta_N \subset \mathbb{R}^{q(N)}$ with $q(N) = N + N_H + N*N_H$ denote the vector of parameters for the RBM.
-
 The probability mass function for the visible variables $X_1, \dots, X_N$ follows from marginalizing this joint specification:
@@
 $$
-
-Note that the vector of model parameters $\boldsymbol \theta_N$, of size $q(N)$, grows in size as a function of sample dimension $N$ to accommodate the dimension of visible variables $X_1, \dots, X_N$, and one may further choose the number $N_H$ of hidden variables to change with $N$ as well; in particular, the number $N_H$ of hiddens may also potentially increase with $N$.
+Note that the vector of model parameters $\boldsymbol \theta_N$, of size $q(N)$, grows in size as a function of sample dimension $N$ to accommodate the dimension of visible variables $X_1, \dots, X_N$, and one may further choose the number $N_H$ of hidden variables to change with $N$ as well.
+In particular, the number $N_H$ of hiddens may also potentially and arbitrarily increase with $N$.
 Additionally, as $|\mathcal{X}| = 2$ and $P_{\boldsymbol \theta_N}(\boldsymbol x) > 0$ for all $\boldsymbol x \in \mathcal{X}^N$, the RBM model specification for visibles is a FSFS model.
@@
 
-Consider two models with "deep architecture" that contain multiple hidden (or latent) layers in addition to a visible layer of data, a deep Boltzmann machine [@salakhutdinov2009deep] and a deep belief network [@hinton2006fast].
-Let $M$ denote the number of hidden layers included in the model and $N_{(H,1)}, \dots, N_{((H,M))}$ the number of hidden variables within each hidden layer.
-Then let $\tilde{\boldsymbol X} = \{H^{(1)}_1, \dots, H^{(1)}_{N_{(H,1)}}, \dots, H^{(M)}_1, \dots, H^{(M)}_{N_{(H,M)}}, \boldsymbol X\}$ be random variables corresponding to the hidden variables $\{ H_{i}^{(j)} : i=1,\ldots, N_{(H,j)}, j=1,\ldots,M\}$ and visible variables $\boldsymbol X =(X_1,\ldots,X_N)$ in a deep probabilistic model.
+Consider two models with "deep architecture" that contain multiple hidden (or latent) layers in addition to a visible layer of data, namely a deep Boltzmann machine [@salakhutdinov2009deep] and a deep belief network [@hinton2006fast].
+Let $M$ denote the number of hidden layers included in the model and let $N_{(H,1)}, \dots, N_{(H,M)}$ denote the numbers of hidden variables within each hidden layer.
+Then the random vector $\tilde{\boldsymbol X} = \{H^{(1)}_1, \dots, H^{(1)}_{N_{(H,1)}}, \dots, H^{(M)}_1, \dots, H^{(M)}_{N_{(H,M)}}, \boldsymbol X\}$ collects both the hidden variables $\{ H_{i}^{(j)} : i=1,\ldots, N_{(H,j)}, j=1,\ldots,M\}$ and visible variables $\boldsymbol X =(X_1,\ldots,X_N)$ in a deep probabilistic model.
 Each variable outcome will again lie in $\mathcal{X} = \{-1,1\}$.
 
-**Deep Boltzmann machine (DBM).** The DBM class of models maintains conditional independence within all layers in the model by stacking RBM models and maintaining the restriction of only allowing conditional dependence between neighboring layers.
+**Deep Boltzmann machine (DBM).** The DBM class of models maintains conditional independence within all layers in the model by stacking RBM models and only allowing conditional dependence between neighboring layers.
 The joint probability mass function for a DBM is
@@
 $$
-
-Again like the RBM case, the visible DBM model specification is an example of a FSFS model.
+Again like the RBM case, the DBM model specification examples a FSFS model.
 
@@
 Thus for visibles $X_1, \dots, X_N$ with support $\mathcal{X}^N$, a DBN is also a FSFS model if the number $q(N)$ of components in the parameter vector is dependent on the dimension of the visibles.
-Commonly, as in logistic belief nets [@neal1992connectionist], a "weight" parameter is placed on each interaction between visibles, $X_1, \dots, X_N$ and the first layer of latent variables, $H^{(1)}_1, \dots, H^{(1)}_{N_{(H,1)}}$, satisfying the definition of a FSFS model.
+Commonly, as in logistic belief nets [@neal1992connectionist], a "weight" parameter is placed on each interaction between visibles, $X_1, \dots, X_N$, and the first layer of latent variables, $H^{(1)}_1, \dots, H^{(1)}_{N_{(H,1)}}$, in the definition of a FSFS model.
 
@@
 \frac{1}{N} \log \left[\frac{\max\limits_{(x_1, \dots, x_N) \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(x_1, \dots, x_N)}{\min\limits_{(x_1, \dots, x_N) \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(x_1, \dots, x_N)}\right] \equiv \frac{1}{N} \text{ELPR}_N(\boldsymbol \theta_N).
-\label{eqn:elpr}
+(\#eq:elpr)
 \end{align}
 
-The main idea is that, in formulating FSFS models for potentially increasing numbers of variables (i.e., for $N \rightarrow \infty$), the ratio (\ref{eqn:elpr}) should remain bounded, requiring that the largest probability possible under $P_{\theta_N}$ should maintain a fixed order of magnitude relative to the smallest probability allowed under the same model.
+The main idea is that, in formulating FSFS models for potentially increasing numbers of variables (i.e., for $N \rightarrow \infty$), the ratio (\@ref(eq:elpr)) should remain bounded to better ensure model stability, requiring that the largest probability possible under $P_{\theta_N}$ maintain a fixed order of magnitude relative to the smallest probability allowed under the same model.
 Specifically, the log of the ratio should grow at mostly linearly with the sample size $N$.
 This leads to the following definition.
-
-\begin{defn}[S-unstable FSFS]
-\label{defn:instab}
+```{definition, name = "S-unstable FSFS", label="instabFSFS", echo=TRUE}
 Let $\boldsymbol \theta_N \in \mathbb{R}^{q(N)}$ be a sequence of FSFS model parameters where the size of the model $q(N)$ is a function of the number of random variables $N$.
@@
 \end{align*}
-\end{defn}
-In other words, given any $C > 0$, there exists an integer $N_C > 0$ so that $\frac{1}{N}\text{ELPR}_N(\boldsymbol \theta_N) > C$ for all $N \ge N_C$.
+```
+In other words, a model is S-unstable if, given any $C > 0$, there exists an integer $N_C > 0$ so that $\frac{1}{N}\text{ELPR}_N(\boldsymbol \theta_N) > C$ for all $N \ge N_C$.
+A FSFS model formulation may be termed S-stable if it fails to be S-unstable.
 
-This definition of *S-unstable* is a generalization or reinterpretation of "unstable" used in @schweinberger2011instability by allowing non-exponential family models (e.g.
-RBM and DBM models in Sections \ref{restricted-boltzmann-machines}-\ref{deep-learning}) and an increasing number of parameters.
+This definition of S-unstable is a generalization or reinterpretation of "unstable" used in @schweinberger2011instability by allowing non-exponential family models (e.g.
+RBM and DBM models in Sections \@ref(restricted-boltzmann-machines)-\@ref(deep-learning)) and an increasing number of parameters.
 While this definition differs in form and scope from the original, it does match that in @schweinberger2011instability for the special case of exponential models (cf.
-Section \ref{discrete-exponential-family-models}) considered there.
+Section \@ref(discrete-exponential-family-models)) considered there.
 
-S-unstable FSFS model sequences are undesirable for several reasons. One is that small changes in data can lead to overly-sensitive changes in probability. Consider, for example, 
+S-unstable FSFS model sequences may be undesirable for several reasons.
+One is that small changes in data can lead to overly-sensitive changes in probability.
+Consider, for example, the quantity given by
 $$
@@
 $$
-the biggest log-probability ratio for a one component change in data outcomes at a FSFS parameter $\boldsymbol \theta_N$. We then have the following (non-asymptotic) result.
+which represents the biggest log-probability ratio for a one component change in data outcomes at a FSFS parameter $\boldsymbol \theta_N$. We then have the following (non-asymptotic) result.
 
-\begin{prop}
-\label{prop:instab}
-Let $\text{ELPR}(\boldsymbol \theta_N)$ be as in (\ref{eqn:elpr}) for an integer $N \ge 1$. For a given $C>0$, if $$\frac{1}{N}\text{ELPR}_N(\boldsymbol \theta_N) > C,$$ 
+```{proposition, label="instab-elpr", echo=TRUE}
+Let $\text{ELPR}(\boldsymbol \theta_N)$ be as in (\@ref(eq:elpr)) for an integer $N \ge 1$. For a given $C>0$, if $$\frac{1}{N}\text{ELPR}_N(\boldsymbol \theta_N) > C,$$ 
 then 
 $$\Delta_N(\boldsymbol \theta_N) > C.$$
-\end{prop}
+```
 
-Again, if the probability ratio (\ref{eqn:elpr}) is too large, then the FSFS model will exhibit large changes in probability for very small differences in the data configuration, which exemplifies the intuitive notation of instability.
+Again, if the probability ratio (\@ref(eq:elpr)) is too large, then the FSFS model will exhibit large changes in probability for very small differences in the data configuration, which exemplifies the intuitive notation of instability.
 
-Additionally, S-unstable FSFS model sequences are connected to degenerate models, where model *degeneracy* implies placing all probability on a small portion of the sample space.
-For perspective, note that differing bounds on $1/N\cdot\text{ELPR}(\boldsymbol \theta_N)$ in (\ref{eqn:elpr}) provide a spectrum of levels of "stability" and Proposition \ref{prop:instab} indicates increasing and undesirable sensitivity of model probabilities (i.e., for one component changes in outcomes) as (\ref{eqn:elpr}) increases.
-Furthermore, as the instability measure (\ref{eqn:elpr}) grows, FSFS model sequences are guaranteed to slide into full degeneracy as Proposition \ref{prop:degen} will show.
+Additionally, S-unstable FSFS model sequences are connected to degenerate models, where model *degeneracy* typically entails placing all probability on a small portion of the sample space.
+For perspective, note that differing sizes of $1/N\cdot\text{ELPR}(\boldsymbol \theta_N)$ in (\@ref(eq:elpr)) may induce a spectrum of levels of "stability" and Proposition \@ref(prp:instab-elpr) indicates increasing sensitivity of model probabilities (i.e., for one component changes in outcomes) as (\@ref(eq:elpr)) increases.
+Furthermore, as the instability measure (\@ref(eq:elpr)) grows, FSFS model sequences are guaranteed to slide into full degeneracy as Proposition \@ref(prp:degenFSFS) shows.
 Define a $\epsilon$-modal set
@@
 
-\begin{prop}
-\label{prop:degen}
-For an unstable FSFS model in Definition \ref{defn:instab}, and for any given $0 < \epsilon < 1$, 
+```{proposition, label="degenFSFS", echo=TRUE}
+For an unstable FSFS model in Definition \@ref(def:instabFSFS), and for any given $0 < \epsilon < 1$, 
 $$
@@
 $$
-\end{prop}
+```
 
-In other words, in S-unstable FSFS models, all probability in the model formulation with a large number of random variables will concentrate mass on mode sets of arbitrarily small size or simply on those few outcomes with the highest probability.
+In other words, in S-unstable FSFS models, all probability in the model formulation with a large number of random variables will concentrate mass on an $\epsilon$-modal set, where $\epsilon$ can be made arbitrarily small.
+The associated mode set could potentially be quite small, in which case Proposition \ref{prp:degenFSFS} would suggest that the unstable model asymptotically stacks all probability on a few outcomes.
 
-\begin{remark}
-There is a further generalization the notion of instability in Definition \ref{defn:instab} meant to address independent replications of data sequences.
+```{remark, echo = TRUE}
+There is a further generalization the notion of instability in Definition \@ref(def:instabFSFS) meant to address independent replications of data sequences.
 That is, one might consider data as $n$ independent and identically distributed replications $\boldsymbol X_1, \dots, \boldsymbol X_n$, where each $\boldsymbol X_i=(X_{i,1},\ldots,X_{i,N}) \in \mathcal{X}^N$ follows a common FSFS model with probabilities $P_{\theta_N}(\boldsymbol x)>0$,  $\boldsymbol x\in\mathcal{X}^N$ and $|\mathcal{X}|<\infty$, for $i=1,\ldots,n$.
 This leads to a total of $n*N$ random variables in the joint model.
-However, the definition of S-unstable and Propositions \ref{prop:instab}-\ref{prop:degen} still hold for such iid replications.
+However, the definition of S-unstable and Propositions \@ref(prp:instab-elpr)-\@ref(prp:degenFSFS) still hold for such iid replications.
 This is because $\left(\max\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)\right)^n$ is the largest probability possible under the joint model for the $n$ replications while $\left(\min\limits_{\boldsymbol x \in \mathcal{X}}P_{\boldsymbol \theta_N}(\boldsymbol x)\right)^n$ is the smallest probability.
@@
 Consequently, overall model instabilities may be characterized by those of one observation from the common FSFS model.
-\end{remark}
+```
 
@@
 
-For a large class of models that covers a broad range of applications (including "deep learning"), we have developed a formal definition and elucidated multiple consequences of instability.
-We have shown for FSFS models that instability manifests through small changes in data leading to overly-sensitive changes in probability as well as the potential to place all probability on a small piece of the sample space.
-Models that fall within the definition of a FSFS model should be used with caution to ensure that the effects of instability are not experienced.
-
-\begin{appendix}
-\section{Appendix: Proofs of main results}
-\label{appendix}
-\label{proofs}
-
-```{r child='appendix/appendix-note.Rmd'}
-```
-\end{appendix}
-
-# References
+For a large class of models that covers a broad range of applications (including "deep learning"), we have developed a formal definition of instability in model probability structure and elucidated multiple consequences of instability.
+We have shown for FSFS models that instability manifests through small changes in data leading to potentially large changes in probability as well as the potential to place all probability on certain modal subsections of the sample space, which could be potentially small.
+The FSFS model class is quite broad and, particularly in developing FSFS models for large data sets, some caution should be used in parameter specification to control effects of model instability.

### 2017-07-08 writing/paper-note/docs/note.tex
paper to bookdown

\documentclass[12pt]{article}
\usepackage{amsmath,amssymb}
\usepackage{graphicx,psfrag,epsf}
\usepackage{enumerate}
\usepackage{natbib}
\usepackage{url} % not crucial - just used below for the URL

\usepackage{longtable,booktabs}
\usepackage[unicode=true]{hyperref}
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\usepackage{tikz, subfig, amsthm, multirow, float}
\usepackage{tikz-3dplot}
\usetikzlibrary{arrows, shapes, positioning}
\usepackage{bm}

%%INCLUDED-AK
\usepackage{amsthm}
\newtheorem{theorem}{Theorem}
\newtheorem{proposition}{Proposition}
\newtheorem{lemma}{Lemma}

\theoremstyle{definition}
\newtheorem{definition}{Definition}
\newtheorem{assumption}{Assumption}
\newtheorem{remark}{Remark}

\DeclareMathOperator*{\argmin}{arg\,min}
\DeclareMathOperator*{\argmax}{arg\,max}

\newcommand{\nv}{{n_{\scriptscriptstyle V}}}
\newcommand{\nh}{{n_{\scriptscriptstyle H}}}
\newcommand{\E}{E}

%\pdfminorversion=4
% NOTE: To produce blinded version, replace "0" with "1" below.
\newcommand{\blind}{0}

% DON'T change margins - should be 1 inch all around.
\addtolength{\oddsidemargin}{-.5in}%
\addtolength{\evensidemargin}{-.5in}%
\addtolength{\textwidth}{1in}%
\addtolength{\textheight}{-.3in}%
\addtolength{\topmargin}{-.8in}%

\let\BeginKnitrBlock\begin \let\EndKnitrBlock\end
\begin{document}

\def\spacingset#1{\renewcommand{\baselinestretch}%
{#1}\small\normalsize} \spacingset{1}

\title{\bf A note on the instability and degeneracy of deep learning models}
\author{Andee Kaplan \\ Department of Statistics, Iowa State University \\  and \\ Daniel Nordman \\ Department of Statistics, Iowa State University \\  and \\ Stephen Vardeman \\ Department of Statistics and Department of Industrial and\\
Manufacturing Systems Engineering, Iowa State University \\ }

\date{}
\maketitle
\begin{abstract}
A probability model exhibits instability if small changes in a data
outcome result in large, and often unanticipated, changes in
probability. For correlated data structures found in several application
areas, there is increasing interest in predicting/identifying such
sensitivity in model probability structure. We consider the problem of
quantifying instability for general probability models defined on
sequences of observations, where each sequence of length \(N\) has a
finite number of possible outcomes. A sequence of probability models
results, indexed by \(N\), that accommodates data of expanding
dimension. Model instability is formally shown to occur when a certain
log-probability ratio under such models grows faster than \(N\). In this
case, a one component change in the data sequence can shift probability
by orders of magnitude. Also, as instability becomes more extreme, the
resulting probability models are shown to tend to degeneracy, placing
all their probability on potentially small portions of the sample space.
These results on instability apply to large classes of models commonly
used in random graphs, network analysis, and machine learning contexts.
\end{abstract}
\noindent%
{\it Keywords:}  Degeneracy, Instability, Classification, Deep Learning, Graphical Models
\vfill

\newpage
\spacingset{1.45}

\section{Introduction}\label{introduction}

We consider the behavior, and the potential impropriety, of probability
models built to incorporate a sequence of discrete observations with
length \(N\). Let \((X_1, \dots, X_N)\) denote a set of discrete random
variables with a finite sample space, \(\mathcal{X}^N\). That is,
\(\mathcal{X}\) with \(|\mathcal{X}| < \infty\) represents a finite set
of potential outcomes for each single variable \(X_i\), and the data
sequence \((X_1,\ldots,X_N)\) takes values in the \(N\)-fold product
space \(\mathcal{X}^N\). For each \(N\), let
\(P_{\boldsymbol \theta_N}\) denote a probability model on
\(\mathcal{X}^N\), under which
\(P_{\boldsymbol \theta_N}(x_1, \dots, x_N) > 0\) is the probability of
the data outcome \((x_1, \dots, x_N) \in \mathcal{X}^N\). In this, we
assume that the model support of \(P_{\boldsymbol \theta_N}\) is the
sample space \(\mathcal{X}^N\). This framework produces a series
\(P_{\boldsymbol \theta_N}\) of probability models, indexed by a generic
sequence of parameters \(\boldsymbol \theta_N\), to describe data of
each length \(N \geq 1\). The size and structure of such parameters are
without restriction, and natural cases include those where
\(\boldsymbol \theta_N \in \mathbb{R}^{q(N)}\) for some arbitrary
integer-valued function \(q(\cdot) \geq 1\). We will refer to this model
class as \emph{Finitely Supported Finite Sequence (FSFS) models}.

Section \ref{examples} provides several examples of FSFS models commonly
used in graph/network analysis and machine learning (i.e., deep learning
models). Section \ref{instability-results} establishes formal results
regarding the propriety of FSFS models with regard to stability. A FSFS
probability model sequence exhibits instability if small changes in the
components of a data outcome \((x_1,\ldots,x_N)\) can result in large
changes in probability \(P_{\theta_N}(x_1,\ldots,x_N)\). The concept of
instability, introduced in the field of statistical physics by Ruelle
(\protect\hyperlink{ref-ruelle1999statistical}{1999}), was extended to
include a notion of detection and quantification for certain exponential
family models by Schweinberger
(\protect\hyperlink{ref-schweinberger2011instability}{2011}). For
similar exponential models, particularly in connection to random
graphs/networks, Handcock
(\protect\hyperlink{ref-handcock2003assessing}{2003}) considered
(mean-based) characterizations for so-called model degeneracy, whereby a
probability model places all mass on a small subset of the sample space
and produces undesirably low variability in model outcomes. As described
by Schweinberger
(\protect\hyperlink{ref-schweinberger2011instability}{2011}), model
instability and model degeneracy are related by viewing degeneracy as an
extreme or limiting form of instability. The instability results of
Schweinberger
(\protect\hyperlink{ref-schweinberger2011instability}{2011}) were
developed for the case of discrete exponential family models. The main
results here concern a general measure of model instability, appropriate
across the whole FSFS model class. This can be used to identify when
certain maximal probabilities in FSFS models are too extreme relative to
the length \(N\) and may thereby induce a potentially undesirable
probability structure. In this case, a one component change in the data
sequence may shift probability by orders of magnitude, and FSFS models
are rigorously shown to become degenerate as the measure of instability
increases. Lastly, Section \ref{implications} emphasizes the
implications of our model propriety results and proofs of the main
results appear in Appendix \ref{appendix-instab}.

\section{Examples}\label{examples}

Many model families fall under the umbrella of FSFS models. For
illustration, this section presents three specific examples of FSFS
models, including models with deep architectures.

\subsection{Discrete exponential family
models}\label{discrete-exponential-family-models}

For discrete random variables \(\boldsymbol X = (X_1, \dots, X_N)\) with
sample space \(\mathcal{X}^N\), \(|\mathcal{X}| < \infty\), consider an
exponential family model for \(\boldsymbol X\) with probability mass
function of the form \[
p_{N, \boldsymbol \lambda}(\boldsymbol x) = \exp\left[\boldsymbol\eta^T(\boldsymbol \lambda) \boldsymbol g_N(\boldsymbol x) - \psi(\boldsymbol \lambda)\right], \qquad \boldsymbol x \in \mathcal{X}^N,
\] depending on parameter vector
\(\boldsymbol \lambda \in \Lambda \subset \mathbb{R}^{k}\) and natural
parameter function
\(\boldsymbol \eta : \mathbb{R}^k \mapsto \mathbb{R}^L\) with fixed
positive integers \(k\) and \(L\) denoting their dimensions. Above,
\(\boldsymbol g_N : \mathcal{X}^N \mapsto \mathbb{R}^L\) is a vector of
sufficient statistics, while \[
\psi(\boldsymbol \lambda) = \log \sum\limits_{\boldsymbol x \in \mathcal{X}^N}\exp\left[\boldsymbol \eta^T(\boldsymbol \lambda) \boldsymbol g_N(\boldsymbol x) \right], \qquad \boldsymbol \lambda \in \Lambda,
\] denotes the normalizing function, and
\(\Lambda = \{\boldsymbol \lambda \in \mathbb{R}^k : \psi(\boldsymbol \lambda) < \infty, k \le q(N) \}\)
is the parameter space.

Defining
\(P_{\boldsymbol \theta_N}(\boldsymbol x)\equiv p_{N,\boldsymbol \lambda_N}(\boldsymbol x)\)
with \(\boldsymbol \theta_N=\boldsymbol \lambda_N\) to be a sequence of
elements of \(\Lambda \subset \mathbb{R}^k\) and noting that
\(P_{\boldsymbol \theta_N}(\boldsymbol x) > 0\) for all
\(\boldsymbol x \in \mathcal{X}^N\), these discrete exponential family
models are special cases of the FSFS models. Such exponential models
arise with spatial data on a lattice (Besag
\protect\hyperlink{ref-besag1974spatial}{1974}), network data (Wasserman
and Faust \protect\hyperlink{ref-wasserman1994social}{1994}; Handcock
\protect\hyperlink{ref-handcock2003assessing}{2003}), and even standard
independence models for discrete data, such as with \(N\) iid Bernoulli
random variables. (Note that for random graphs or networks with, say,
\(m\) nodes, one may wish to consider \(N={m \choose 2}\) edges as
binary (presence/absence) variables \(X_i\). In this case, the length
\(N\) of data sequence may naturally increase as a function of \(m\).)
For these exponential models, the dimension of the parameter
\(\boldsymbol \theta_N\) remains constant over each \(N\), as
\(\boldsymbol \theta_N\) lies in a parameter space of fixed Euclidean
dimension \(k\). This need not be true for other types of FSFS models
considered in Sections
\ref{restricted-boltzmann-machines}-\ref{deep-learning}. Schweinberger
(\protect\hyperlink{ref-schweinberger2011instability}{2011}) considered
instability in such exponential models (e.g., for random graphs) for
sequences of fixed parameters
\(\boldsymbol \theta_N=\boldsymbol \lambda\in\mathbb{R}^k\),
\(N \geq 1\).

\subsection{Restricted Boltzmann
machines}\label{restricted-boltzmann-machines}

A restricted Boltzmann machine (RBM) is an undirected graphical model
specified for discrete or continuous random variables, with binary
variables being most common (cf. Smolensky
\protect\hyperlink{ref-smolensky1986information}{1986}). A RBM
architecture has two layers, hidden (\(\mathcal{H}\)) and visible
(\(\mathcal{V}\)), with conditional independence within each layer. Let
\(\boldsymbol X=(X_1,\ldots,X_N)\) denote the \(N\) random variables for
visibles with support \(\mathcal{X}^N\) and
\(\boldsymbol H=(H_1,\ldots,H_{N_H})\) denote the \(N_H\) random
variables for hiddens with support \(\mathcal{X}^{N_H}\) where
\(\mathcal{X} = \{-1,1\}\). For parameters
\(\boldsymbol \alpha \in \mathbb{R}^{N_H}\),
\(\boldsymbol \beta \in \mathbb{R}^N\), and \(\Gamma\) as a matrix with
dimension \(N_H \times N\), the RBM model for
\(\tilde{\boldsymbol X}=(\boldsymbol X,\boldsymbol H)\) then has the
joint probability mass function \[
P_{\boldsymbol \theta_N} (\tilde{\boldsymbol x}) = \exp\left[ \boldsymbol \alpha^T \boldsymbol h + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^T \Gamma \boldsymbol x - \psi(\boldsymbol \theta_N)\right], \qquad \tilde{\boldsymbol x} = (\boldsymbol h, \boldsymbol x) \in \mathcal{X}^{N+N_H},
\] where \[
\psi(\boldsymbol \theta_N) = \log \sum_{\tilde{\boldsymbol x} \in \mathcal{X}^{N + N_H}} \exp\left[ \boldsymbol \alpha^T \boldsymbol h + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^T \Gamma \boldsymbol x\right], \qquad \boldsymbol \theta_N \in \Theta_N,
\] is the normalizing function. Let
\(\boldsymbol \theta_N = (\boldsymbol \alpha, \boldsymbol \beta,\Gamma) \in \Theta_N \subset \mathbb{R}^{q(N)}\)
with \(q(N) = N + N_H + N*N_H\) denote the vector of parameters for the
RBM. The probability mass function for the visible variables
\(X_1, \dots, X_N\) follows from marginalizing this joint specification:
\[
P_{\boldsymbol \theta_N} (\boldsymbol x) = \sum\limits_{\boldsymbol h \in \mathcal{X}^{N_H}} P_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h), \qquad \boldsymbol x \in \mathcal{X}^N.
\] Note that the vector of model parameters \(\boldsymbol \theta_N\), of
size \(q(N)\), grows in size as a function of sample dimension \(N\) to
accommodate the dimension of visible variables \(X_1, \dots, X_N\), and
one may further choose the number \(N_H\) of hidden variables to change
with \(N\) as well. In particular, the number \(N_H\) of hiddens may
also potentially and arbitrarily increase with \(N\). Additionally, as
\(|\mathcal{X}| = 2\) and
\(P_{\boldsymbol \theta_N}(\boldsymbol x) > 0\) for all
\(\boldsymbol x \in \mathcal{X}^N\), the RBM model specification for
visibles is a FSFS model. This example also indicates that models formed
by marginalizing a base FSFS model (e.g., a type of exponential family
model) is again a FSFS model class.

\subsection{Deep learning}\label{deep-learning}

Consider two models with ``deep architecture'' that contain multiple
hidden (or latent) layers in addition to a visible layer of data, namely
a deep Boltzmann machine (Salakhutdinov and Hinton
\protect\hyperlink{ref-salakhutdinov2009deep}{2009}) and a deep belief
network (Hinton, Osindero, and Teh
\protect\hyperlink{ref-hinton2006fast}{2006}). Let \(M\) denote the
number of hidden layers included in the model and let
\(N_{(H,1)}, \dots, N_{(H,M)}\) denote the numbers of hidden variables
within each hidden layer. Then the random vector
\(\tilde{\boldsymbol X} = \{H^{(1)}_1, \dots, H^{(1)}_{N_{(H,1)}}, \dots, H^{(M)}_1, \dots, H^{(M)}_{N_{(H,M)}}, \boldsymbol X\}\)
collects both the hidden variables
\(\{ H_{i}^{(j)} : i=1,\ldots, N_{(H,j)}, j=1,\ldots,M\}\) and visible
variables \(\boldsymbol X =(X_1,\ldots,X_N)\) in a deep probabilistic
model. Each variable outcome will again lie in
\(\mathcal{X} = \{-1,1\}\).

\textbf{Deep Boltzmann machine (DBM).} The DBM class of models maintains
conditional independence within all layers in the model by stacking RBM
models and only allowing conditional dependence between neighboring
layers. The joint probability mass function for a DBM is \[
P_{\boldsymbol \theta_N} (\tilde{\boldsymbol x}) = \exp\left[ \sum\limits_{i = 1}^M\boldsymbol \alpha^{(i)T} \boldsymbol h^{(i)} + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^{(1)T} \Gamma^{(0)} \boldsymbol x + \sum\limits_{i = 1}^{M - 1} \boldsymbol h^{(i)T} \Gamma^{(i)} \boldsymbol h^{(i + 1)} - \psi(\boldsymbol \theta_N) \right], 
\] for
\(\tilde{\boldsymbol x} = (\boldsymbol h^{(1)}, \dots, \boldsymbol h^{(M)}, \boldsymbol x) \in \mathcal{X}^{N_{(H,1)} + \cdots + N_{(H,M)} +N}\)
where \[
\psi(\boldsymbol \theta_N) = \log \sum\limits_{\tilde{\boldsymbol x} \in \mathcal{X}^{N_{(H,1)} + \cdots + N_{(H,M)} +N}} \exp\left[ \sum\limits_{i = 1}^M\boldsymbol \alpha^{(i)T} \boldsymbol h^{(i)} + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^{(1)T} \Gamma^{(0)} \boldsymbol x + \sum\limits_{i = 1}^{M - 1} \boldsymbol h^{(i)T} \Gamma^{(i)} \boldsymbol h^{(i + 1)}\right],
\] for \(\boldsymbol \theta_N \in \Theta_N\) is the normalizing function
and parameters in the model are \(\boldsymbol \beta \in \mathbb{R}^V\),
\(\boldsymbol \alpha^{(i)} \in \mathbb{R}^{N_{(H,i)}}\) for
\(i = 1, \dots, M\), along with a matrix \(\Gamma^{(0)}\) of dimension
\(N_{(H,1)} \times N\), and matrices \(\Gamma^{(i)}\) of dimension
\(N_{(H,i)} \times N_{(H,i+1)}\) for \(i = 1, \dots, M-1\). Let
\(\boldsymbol \theta_N = (\boldsymbol \alpha^{(1)}, \dots, \boldsymbol \alpha^{(M)}, \boldsymbol \beta,\Gamma^{(0)}, \dots, \Gamma^{(M - 1)}) \in \Theta_N \subset \mathbb{R}^{q(N)}\)
denote the combined vector of parameters with total length
\(q(N)= N_{(H,1)}+\cdots N_{(H,M)} + N + N_{(H,1)}*N+N_{H,2}*H_{(H,1)}+\cdots +N_{(H,M)}*H_{(H,M)-1}\).

The probability mass function for the visible random variables
\(X_1, \dots, X_N\) follows from this joint specification as \[
P_{\boldsymbol \theta_N} (\boldsymbol x) = \sum\limits_{(\boldsymbol h^{(1)}, \dots, \boldsymbol h^{(M)}) \in \mathcal{X}^{N_{(H,1)} + \cdots + N_{(H,M)}}} P_{\boldsymbol \theta_N} (\tilde{\boldsymbol x}), \qquad \boldsymbol x \in \mathcal{X}^N
\] Again like the RBM case, the DBM model specification examples a FSFS
model.

\textbf{Deep belief network (DBN).} A DBN resembles a DBM in that there
are multiple layers of latent random variables stacked in a deep
architecture with no conditional dependence between layers. The
difference between the DBM and DBN models is that all but the last
stacked layer in a DBN are Bayesian networks (see Pearl
\protect\hyperlink{ref-pearl985bayesian}{1985}), rather than RBMs. Thus
for visibles \(X_1, \dots, X_N\) with support \(\mathcal{X}^N\), a DBN
is also a FSFS model if the number \(q(N)\) of components in the
parameter vector is dependent on the dimension of the visibles.
Commonly, as in logistic belief nets (Neal
\protect\hyperlink{ref-neal1992connectionist}{1992}), a ``weight''
parameter is placed on each interaction between visibles,
\(X_1, \dots, X_N\), and the first layer of latent variables,
\(H^{(1)}_1, \dots, H^{(1)}_{N_{(H,1)}}\), in the definition of a FSFS
model.

\section{Instability results}\label{instability-results}

To define or measure instability in FSFS models, it is useful to
consider the behavior of a data model sequence \(P_{\theta_N}\). A
relevant quantity to this end is a (scaled) extremal log-probability
ratio (ELPR)
\begin{align}
\frac{1}{N} \log \left[\frac{\max\limits_{(x_1, \dots, x_N) \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(x_1, \dots, x_N)}{\min\limits_{(x_1, \dots, x_N) \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(x_1, \dots, x_N)}\right] \equiv \frac{1}{N} \text{ELPR}_N(\boldsymbol \theta_N). \label{eq:elpr}
\end{align}
The main idea is that, in formulating FSFS models for potentially
increasing numbers of variables (i.e., for \(N \rightarrow \infty\)),
the ratio (\eqref{eq:elpr}) should remain bounded to better ensure model
stability, requiring that the largest probability possible under
\(P_{\theta_N}\) maintain a fixed order of magnitude relative to the
smallest probability allowed under the same model. Specifically, the log
of the ratio should grow at mostly linearly with the sample size \(N\).
This leads to the following definition.
\BeginKnitrBlock{definition}[S-unstable FSFS]
\protect\hypertarget{def:instabFSFS}{}{\label{def:instabFSFS}
\iffalse (S-unstable FSFS) \fi{} }Let
\(\boldsymbol \theta_N \in \mathbb{R}^{q(N)}\) be a sequence of FSFS
model parameters where the size of the model \(q(N)\) is a function of
the number of random variables \(N\). A FSFS model formulation is
\emph{Schweinberger-unstable} or \emph{S-unstable} if, as the number of
variables increase (\(N \rightarrow \infty\)),
\begin{align*}
\lim\limits_{N \rightarrow \infty} \frac{1}{N} \text{ELPR}(\boldsymbol \theta_N) \equiv \lim\limits_{N \rightarrow \infty} \frac{1}{N} \log \left[\frac{\max\limits_{(x_1, \dots, x_N) \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(x_1, \dots, x_N)}{\min\limits_{(x_1, \dots, x_N) \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(x_1, \dots, x_N)}\right] = \infty.
\end{align*}
\EndKnitrBlock{definition} In other words, a model is S-unstable if,
given any \(C > 0\), there exists an integer \(N_C > 0\) so that
\(\frac{1}{N}\text{ELPR}_N(\boldsymbol \theta_N) > C\) for all
\(N \ge N_C\). A FSFS model formulation may be termed S-stable if it
fails to be S-unstable.

This definition of S-unstable is a generalization or reinterpretation of
``unstable'' used in Schweinberger
(\protect\hyperlink{ref-schweinberger2011instability}{2011}) by allowing
non-exponential family models (e.g.~RBM and DBM models in Sections
\ref{restricted-boltzmann-machines}-\ref{deep-learning}) and an
increasing number of parameters. While this definition differs in form
and scope from the original, it does match that in Schweinberger
(\protect\hyperlink{ref-schweinberger2011instability}{2011}) for the
special case of exponential models (cf.~Section
\ref{discrete-exponential-family-models}) considered there.

S-unstable FSFS model sequences may be undesirable for several reasons.
One is that small changes in data can lead to overly-sensitive changes
in probability. Consider, for example, the quantity given by \[
\Delta(\boldsymbol \theta_N) \equiv \max \left\{\log \frac{P_{\boldsymbol \theta_N}(\boldsymbol x)}{P_{\boldsymbol \theta_N}(\boldsymbol x^*)} : \boldsymbol x \text{ }\& \text{ } \boldsymbol x^* \in \mathcal{X}^N \text{ differ in exactly one component}\right\},
\] which represents the biggest log-probability ratio for a one
component change in data outcomes at a FSFS parameter
\(\boldsymbol \theta_N\). We then have the following (non-asymptotic)
result.

\BeginKnitrBlock{proposition}
\protect\hypertarget{prp:instab-elpr}{}{\label{prp:instab-elpr}}Let
\(\text{ELPR}(\boldsymbol \theta_N)\) be as in (\eqref{eq:elpr}) for an
integer \(N \ge 1\). For a given \(C>0\), if
\[\frac{1}{N}\text{ELPR}_N(\boldsymbol \theta_N) > C,\] then
\[\Delta_N(\boldsymbol \theta_N) > C.\]
\EndKnitrBlock{proposition}

Again, if the probability ratio (\eqref{eq:elpr}) is too large, then the
FSFS model will exhibit large changes in probability for very small
differences in the data configuration, which exemplifies the intuitive
notation of instability.

Additionally, S-unstable FSFS model sequences are connected to
degenerate models, where model \emph{degeneracy} typically entails
placing all probability on a small portion of the sample space. For
perspective, note that differing sizes of
\(1/N\cdot\text{ELPR}(\boldsymbol \theta_N)\) in (\eqref{eq:elpr}) may
induce a spectrum of levels of ``stability'' and Proposition
\ref{prp:instab-elpr} indicates increasing sensitivity of model
probabilities (i.e., for one component changes in outcomes) as
(\eqref{eq:elpr}) increases. Furthermore, as the instability measure
(\eqref{eq:elpr}) grows, FSFS model sequences are guaranteed to slide into
full degeneracy as Proposition \ref{prp:degenFSFS} shows. Define a
\(\epsilon\)-modal set \[
M_{\epsilon, \boldsymbol \theta_N} \equiv \left\{\boldsymbol x \in \mathcal{X}^N: \log P_{\boldsymbol \theta_N}(\boldsymbol x) > (1-\epsilon)\max\limits_{\boldsymbol x^* \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x^*) + \epsilon\min\limits_{\boldsymbol x^* \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x^*) \right\}
\] of possible outcomes, for a given \(0 < \epsilon < 1\).

\BeginKnitrBlock{proposition}
\protect\hypertarget{prp:degenFSFS}{}{\label{prp:degenFSFS}}For an unstable
FSFS model in Definition \ref{def:instabFSFS}, and for any given
\(0 < \epsilon < 1\), \[
P_{\boldsymbol \theta_N}\left((x_1, \dots, x_N) \in M_{\epsilon, \boldsymbol \theta_N}\right) \rightarrow 1 \text{ as } N \rightarrow \infty.
\]
\EndKnitrBlock{proposition}

In other words, in S-unstable FSFS models, all probability in the model
formulation with a large number of random variables will concentrate
mass on an \(\epsilon\)-modal set, where \(\epsilon\) can be made
arbitrarily small. The associated mode set could potentially be quite
small, in which case Proposition \ref{prp:degenFSFS} would suggest that
the unstable model asymptotically stacks all probability on a few
outcomes.

\BeginKnitrBlock{remark}
\iffalse <span class="remark"><em>Remark. \fi{} There is a further
generalization the notion of instability in Definition
\ref{def:instabFSFS} meant to address independent replications of data
sequences. That is, one might consider data as \(n\) independent and
identically distributed replications
\(\boldsymbol X_1, \dots, \boldsymbol X_n\), where each
\(\boldsymbol X_i=(X_{i,1},\ldots,X_{i,N}) \in \mathcal{X}^N\) follows a
common FSFS model with probabilities \(P_{\theta_N}(\boldsymbol x)>0\),
\(\boldsymbol x\in\mathcal{X}^N\) and \(|\mathcal{X}|<\infty\), for
\(i=1,\ldots,n\). This leads to a total of \(n*N\) random variables in
the joint model. However, the definition of S-unstable and Propositions
\ref{prp:instab-elpr}-\ref{prp:degenFSFS} still hold for such iid
replications. This is because
\(\left(\max\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)\right)^n\)
is the largest probability possible under the joint model for the \(n\)
replications while
\(\left(\min\limits_{\boldsymbol x \in \mathcal{X}}P_{\boldsymbol \theta_N}(\boldsymbol x)\right)^n\)
is the smallest probability. Thus, for the combined replications
\(\boldsymbol X_1, \ldots, \boldsymbol X_n\), the analog definition of
the extremal log-probability becomes
\begin{align*}
\frac{\text{extremal log-probability ratio}}{\text{\# random variables in the model}} &\equiv \frac{1}{n*N}\log \left[\frac{\left(\max\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)\right)^n}{\left(\min\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)\right)^n}\right] = \frac{1}{N} \log \left[\frac{\max\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)}{\min\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)}\right] \\
&= \frac{1}{N} \text{ELPR}(\boldsymbol \theta_N),
\end{align*}
implying that the definition of an S-unstable FSFS model sequence is
invariant to the level (\(n\)) of independent replication. Consequently,
overall model instabilities may be characterized by those of one
observation from the common FSFS model.
\EndKnitrBlock{remark}

\section{Implications}\label{implications}

For a large class of models that covers a broad range of applications
(including ``deep learning''), we have developed a formal definition of
instability in model probability structure and elucidated multiple
consequences of instability. We have shown for FSFS models that
instability manifests through small changes in data leading to
potentially large changes in probability as well as the potential to
place all probability on certain modal subsections of the sample space,
which could be potentially small. The FSFS model class is quite broad
and, particularly in developing FSFS models for large data sets, some
caution should be used in parameter specification to control effects of
model instability.

\clearpage

\appendix

\section{Proofs of instability results}\label{appendix-instab}

\textbf{Proof of Proposition \ref{prp:instab-elpr}.} We prove the
contrapositive, supposing that \(\Delta(\boldsymbol \theta_N) \le C\)
holds for some \(C > 0\) and show
\(\text{ELPR}(\boldsymbol \theta_N) \leq NC\). Let
\(\boldsymbol x_{min} \equiv \argmin\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)\)
and
\(\boldsymbol x_{max} \equiv \argmax\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)\).
Note there exists a sequence
\(\boldsymbol x_{min} \equiv \boldsymbol x_0, \boldsymbol x_1, \dots, \boldsymbol x_k \equiv \boldsymbol x_{max}\)
in \(\mathcal{X}^N\) of component-wise switches to move from
\(\boldsymbol x_{min}\) to \(\boldsymbol x_{max}\) in the sample space
(i.e. \(\boldsymbol x_i, \boldsymbol x_{i + 1} \in \mathcal{X}^N\)
differ in exactly \(1\) component for \(i = 0, \dots, k\)) for some
integer \(k \in \{0, 1, \dots, N\}\). Under the FSFS model, recall
\(P_{\boldsymbol \theta_N}(\boldsymbol x) > 0\) holds so that
\(\log P_{\boldsymbol \theta_N}(\boldsymbol x)\) is well-defined for
each outcome \(\boldsymbol x \in \mathcal{X}^N\). Then, if \(k > 0\), it
follows that
\begin{align*}
\text{ELPR}(\boldsymbol \theta_N) = \log\left[\frac{P_{\boldsymbol \theta_N}(\boldsymbol x_{max})}{P_{\boldsymbol \theta_N}(\boldsymbol x_{min})}\right] &= \left|\sum\limits_{i = 1}^k\log\left(\frac{P_{\boldsymbol \theta_N}(\boldsymbol x_i)}{P_{\boldsymbol \theta_N}(\boldsymbol x_{i-1})}\right)\right| \\
&\le \sum\limits_{i = 1}^k\left|\log\left(\frac{P_{\boldsymbol \theta_N}(\boldsymbol x_i)}{P_{\boldsymbol \theta}(\boldsymbol x_{i-1})}\right)\right| \le k \Delta_N(\boldsymbol \theta_N) \le NC,
\end{align*}
using \(k \le N\) and \(\Delta(\boldsymbol \theta_N) \le C\). If
\(k = 0\), then \(\boldsymbol x_{max} = \boldsymbol x_{min}\) and the
same bound above holds.\hfill \(\Box\)

\textbf{Proof of Proposition \ref{prp:degenFSFS}.} where
\(|\mathcal{X}|<\infty\) holds in the FSFS model. We may suppose
\(|\mathcal{X}|>1\) (i.e., \(\mathcal{X}^N\) has more than one outcome)
because otherwise the model is trivially degenerate for all
\(N \geq 1\). Fix \(0 < \epsilon < 1\). Then,
\(\boldsymbol x_{max} \in M_{\epsilon, \boldsymbol \theta_N}\), so
\(P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N}) \ge P_{\boldsymbol \theta_N}(\boldsymbol x_{max}) > 0\).
If
\(\boldsymbol x \in \mathcal{X}^N \setminus M_{\epsilon, \boldsymbol \theta_N}\),
then by definition
\(P_{\boldsymbol \theta_N}(\boldsymbol x) \le [P_{\boldsymbol \theta_N}(\boldsymbol x_{max})]^{1-\epsilon}[P_{\boldsymbol \theta_N}(\boldsymbol x_{min})]^{\epsilon}\)
holds so that
\begin{align*}
1-P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N})
& = \sum\limits_{\boldsymbol x \in \mathcal{X}^N \setminus M_{\epsilon, \boldsymbol \theta_N}}P_{\boldsymbol \theta_N}(\boldsymbol x) \\
& \le (|\mathcal{X}|^N)[P_{\boldsymbol \theta_N}(\boldsymbol x_{max})]^{1-\epsilon}[P_{\boldsymbol \theta_N}(\boldsymbol x_{min})]^{\epsilon}.
\end{align*}
From the lower bound on
\(P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N})\) and the
upper bound on
\(1-P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N})\), it
follows that
\begin{align*}
\frac{1}{N}\log\left[\frac{P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N})}{1-P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N})}\right] & \ge \frac{1}{N} \log\left[\frac{P_{\boldsymbol\theta_N}(\boldsymbol x_{max})}{(|\mathcal{X}|^N)[P_{\boldsymbol \theta_N}(\boldsymbol x_{max})]^{1-\epsilon}[P_{\boldsymbol \theta_N}(\boldsymbol x_{min})]^{\epsilon}}\right] \\
&= \frac{\epsilon}{N} \log\left[\frac{P_{\boldsymbol \theta_N}(\boldsymbol x_{max})}{P_{\boldsymbol \theta_N}(\boldsymbol x_{min})}\right] - \log |\mathcal{X}| \rightarrow \infty
\end{align*}
as \(N \rightarrow \infty\) by the definition of an unstable FSFS model
(cf.~Definition \ref{def:instabFSFS}). Consequently,
\(P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N}) \rightarrow 1\)
as \(N \rightarrow \infty\) as claimed. \hfill \(\Box\)

\clearpage

\section*{References}\label{references}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\hypertarget{ref-besag1974spatial}{}
Besag, Julian. 1974. ``Spatial Interaction and the Statistical Analysis
of Lattice Systems.'' \emph{Journal of the Royal Statistical Society.
Series B (Methodological)}. JSTOR, 192--236.

\hypertarget{ref-handcock2003assessing}{}
Handcock, Mark S. 2003. ``Assessing Degeneracy in Statistical Models of
Social Networks.'' Center for Statistics; the Social Sciences,
University of Washington. \url{http://www.csss.washington.edu/}.

\hypertarget{ref-hinton2006fast}{}
Hinton, Geoffrey E, Simon Osindero, and Yee-Whye Teh. 2006. ``A Fast
Learning Algorithm for Deep Belief Nets.'' \emph{Neural Computation} 18
(7). MIT Press: 1527--54.

\hypertarget{ref-neal1992connectionist}{}
Neal, Radford M. 1992. ``Connectionist Learning of Belief Networks.''
\emph{Artificial Intelligence} 56 (1). Elsevier: 71--113.

\hypertarget{ref-pearl985bayesian}{}
Pearl, Judea. 1985. ``Bayesian Networks: A Model of Self-Activated
Memory for Evidential Reasoning.'' UCLA Computer Science Department.

\hypertarget{ref-ruelle1999statistical}{}
Ruelle, D. 1999. \emph{Statistical Mechanics: Rigorous Results}. London:
Imperial College Press.

\hypertarget{ref-salakhutdinov2009deep}{}
Salakhutdinov, Ruslan, and Geoffrey E Hinton. 2009. ``Deep Boltzmann
Machines.'' In \emph{International Conference on Artificial Intelligence
and Statistics}, 448--55. AI \& Statistics.

\hypertarget{ref-schweinberger2011instability}{}
Schweinberger, Michael. 2011. ``Instability, Sensitivity, and Degeneracy
of Discrete Exponential Families.'' \emph{Journal of the American
Statistical Association} 106 (496). Taylor \& Francis: 1361--70.

\hypertarget{ref-smolensky1986information}{}
Smolensky, Paul. 1986. ``Information Processing in Dynamical Systems:
Foundations of Harmony Theory.'' DTIC Document.

\hypertarget{ref-wasserman1994social}{}
Wasserman, Stanley, and Katherine Faust. 1994. \emph{Social Network
Analysis: Methods and Applications}. Vol. 8. Cambridge: Cambridge
University Press.

\end{document}

### 2017-07-08 writing/paper-note/index.Rmd
paper to bookdown

---
title: "A note on the instability and degeneracy of deep learning models"
author: 
  - name: Andee Kaplan
    university: Iowa State University
    department: Department of Statistics
  - name: Daniel Nordman
    university: Iowa State University
    department: Department of Statistics
  - name: Stephen Vardeman
    university: Iowa State University
    department: |
      | Department of Statistics and Department of Industrial and
      | Manufacturing Systems Engineering
site: bookdown::bookdown_site
output: 
  bookdown::pdf_book:
    includes:
      in_header: ../resources/latex/header.tex
    toc: no
bibliography: [../resources/latex/refs_rbm.bib]
github-repo: andeek/rbm
abstract: |
  A probability model exhibits instability if small changes in a data outcome result in large, and often unanticipated, changes in probability. For correlated data structures found in several application areas, there is increasing interest in predicting/identifying such sensitivity in model probability structure. We consider the problem of quantifying instability for general probability models defined on sequences of observations, where each sequence of length $N$ has a finite number of possible outcomes. A sequence of probability models results, indexed by $N$, that accommodates data of expanding dimension. Model instability is formally shown to occur when a certain log-probability ratio under such models grows faster than $N$. In this case, a one component change in the data sequence can shift probability by orders of magnitude. Also, as instability becomes more extreme, the resulting probability models are shown to tend to degeneracy, placing all their probability on potentially small portions of the sample space. These results on instability apply to large classes of models commonly used in random graphs, network analysis, and machine learning contexts.
keywords: Degeneracy, Instability, Classification, Deep Learning, Graphical Models
blind: false
---

```{r setup, include=FALSE}
# https://github.com/rstudio/bookdown/pull/374
fix_envs <- function(x) {
  beg_reg <- '^\\s*\\\\begin\\{.*\\}'
  end_reg <- '^\\s*\\\\end\\{.*\\}'
  i3 = if (length(i1 <- grep(beg_reg, x))) (i1 - 1)[grepl("^\\s*$", x[i1 - 1])]
  i3 = c(i3, if (length(i2 <- grep(end_reg, x))) (i2 + 1)[grepl("^\\s*$", x[i2 + 1])])
  if (length(i3)) x = x[-i3]
  x
}

options(
  dplyr.print_min = 6, dplyr.print_max = 6, width = 70,
  digits = 3,
  bookdown.post.latex = fix_envs
)
knitr::opts_chunk$set(
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  cache = TRUE,
  echo = FALSE
)

ggplot2::theme_set(ggplot2::theme_bw(base_family="serif"))

library(knitr)
library(ggplot2)
library(dplyr)
library(tidyr)
library(scales)
library(mvtnorm)
library(rootSolve)
library(agridat)
library(xtable)
```

### 2017-07-08 writing/paper-rbm/appendices/appendix-center.Rmd
paper to bookdown

--- 2017-05-24 writing/appendix/appendix-center.Rmd
@@
----
-title: "Supplementary material to:  On the propriety of restricted Boltzmann machines"
-author: 
-  - name: Andee Kaplan
-    university: Iowa State University
-    department: Department of Statistics
-  - name: Daniel Nordman
-    university: Iowa State University
-    department: Department of Statistics
-  - name: Stephen Vardeman
-    university: Iowa State University
-    department: |
-      | Department of Statistics and Department of Industrial and
-      | Manufacturing Systems Engineering
-output: 
-  pdf_document:
-    keep_tex: true
-    fig_caption: yes
-    includes:
-      in_header: ../latex/header.tex
-    number_sections: yes
-    template: ../latex/jasa_article.tex
-blind: false
----
+\clearpage
 
-Supplement material to follow consists of Section \ref{on-the-one-to-one-correspondence-between-parameters-and-moments-in-the-rbm-distribution}, providing proofs and discussion of results relating to the data coding in restricted Boltzmann machine models (RBMs).
+# (APPENDIX) Appendix {-}
 
-# On the one-to-one correspondence between parameters and moments in the RBM distribution
+# On the one-to-one correspondence between parameters and moments in the RBM distribution {#appendix-rbm}
 
@@
 Denote the distribution of $(\bm{V}_m,\bm{H}_n)$ as RBM$(m,n,\bm{\theta})$.
-In the following, for convenience, we say a symmetric random variable $X$ has a Bernoulli$(1/2)$ distribution if $P(X=1)=1/2=P(X=-1)$.
+In the following, for convenience, we say a symmetric random variable $X$ has a Bernoulli$(1/2)$ distribution if $P(X = 1) = 1/2 = P(X = -1)$.
 
@@
 
-\begin{itemize}
-\item The function $\bm{g}: \mathbb{R}^{m+n+mn} \rightarrow \mathcal{I}$ is one-to-one; see Theorem \ref{thm:one} below.
+- The function $\bm{g}: \mathbb{R}^{m+n+mn} \rightarrow \mathcal{I}$ is one-to-one; see Theorem \@ref(thm:one) below.
 Hence, these moments uniquely characterize the parameters in the RBM$(m,n,\bm{\theta})$ model.
-\item The image $\mathcal{I}$ is a symmetric set (i.e., $\bm{g}(\bm{\theta})=-\bm{g}(-\bm{\theta})$) and must be a connected (not necessarily convex) region in $\mathbb{R}^{m+n+mn}$, without any voids in it (by the continuity of $\bm{g}(\bm{\theta})$).
-\item The center of the image $\mathcal{I}$ (the zero vector in $\mathbb{R}^{m+n+mn}$) is given only by $\bm{\theta}=\bm{0}$ (the zero vector in the parameter space).
-This corresponds to the case where all the random vectors $V_1,\ldots,V_m,H_1,\ldots,H_n$ are iid with a symmetric Bernoulli$(1/2)$ distribution; see Theorem \ref{thm:two}.
-\end{itemize}
+- The image $\mathcal{I}$ is a symmetric set (i.e., $\bm{g}(\bm{\theta})=-\bm{g}(-\bm{\theta})$) and must be a connected (not necessarily convex) region in $\mathbb{R}^{m+n+mn}$, without any voids in it (by the continuity of $\bm{g}(\bm{\theta})$).
+- The center of the image $\mathcal{I}$ (the zero vector in $\mathbb{R}^{m+n+mn}$) is given only by $\bm{\theta}=\bm{0}$ (the zero vector in the parameter space).
+This corresponds to the case where all the random vectors $V_1,\ldots,V_m,H_1,\ldots,H_n$ are iid with a symmetric Bernoulli$(1/2)$ distribution; see Theorem \@ref(thm:two).
 
-\begin{apptheorem} 
-\label{thm:one}
+```{theorem, label="one", echo=TRUE}
 Let $m,n\geq 1$ and suppose the random vector $(\bm{V}_m,\bm{H}_n) = (V_1,\ldots,V_m, H_1,\ldots,H_n)$ follows a RBM$(m,n,\bm{\theta})$ distribution.
@@
 let $P_i$ and $\E$ denote probability and expectation under $\bm{\theta}_i$, $i=1,2$. Then, the following are equivalent
-\begin{enumerate}
-\item $\E_1 V_i=\E_2 V_i$, $\E_1 H_j=\E_2 H_j$ and $\E_1 V_i H_j =\E_2 V_i H_j  $ for any $i=1,\ldots,m$, $j=1,\ldots,n$.
-\item $\bm{\theta}_1 =\bm{\theta}_2$.
-\item $P_1$ and  $P_2$ are the same distribution for  $(\bm{V}_m,\bm{H}_n)$.\\
-\end{enumerate}
-\end{apptheorem}
 
-**Proof of Theorem \ref{thm:one}.** We first establish the equivalence of claims 2 and 3. If claim 2 holds, then
- claim 3 holds trivially.
+1. $\E_1 V_i=\E_2 V_i$, $\E_1 H_j=\E_2 H_j$ and $\E_1 V_i H_j =\E_2 V_i H_j$ for any $i=1,\ldots,m$, $j=1,\ldots,n$.
+2. $\bm{\theta}_1 =\bm{\theta}_2$.
+3. $P_1$ and  $P_2$ are the same distribution for  $(\bm{V}_m,\bm{H}_n)$.\\
+
+```
+
+```{proof, echo=TRUE}
+We first establish the equivalence of claims 2 and 3.
+If claim 2 holds, then claim 3 holds trivially.
 Suppose claim 3 holds and pick $\ell \in \{1,\ldots,m\}$.
@@
 
-To establish the equivalence of claims 1 and 3 in Theorem \ref{thm:one} under the RBM$(m,n,\bm{\theta})$ model, we note that claim 3 easily implies claim 1.
+To establish the equivalence of claims 1 and 3 in Theorem \@ref(thm:one) under the RBM$(m,n,\bm{\theta})$ model, we note that claim 3 easily implies claim 1.
 Hence, it suffices now to establish claim 3 from claim 1.
-This follows immediately from Lemma \ref{lemma:one} below and the fact that the variables $V_1,\ldots,V_m$ are conditionally independent given $H_1,\ldots,H_n$ in the RBM$(m,n,\bm{\theta})$ model (and likewise $H_1,\ldots,H_n$ are independent given $V_1,\ldots,V_m$).
+This follows immediately from Lemma \@ref(lem:first) below and the fact that the variables $V_1,\ldots,V_m$ are conditionally independent given $H_1,\ldots,H_n$ in the RBM$(m,n,\bm{\theta})$ model (and likewise $H_1,\ldots,H_n$ are independent given $V_1,\ldots,V_m$).
+```
 
-\begin{applemma} 
-\label{lemma:one}
+```{lemma, label="first", echo=TRUE}
 Let $m,n\geq 1$.
@@
 follows if $\E_1 V_i = \E_2 V_i$, $\E_1 H_j = \E_2 H_j$, $\E_1 V_i H_j = \E_2 V_j H_j$ holds for all $i\in\{1,\ldots,m\}$, $j\in \{1,\ldots,n\}$.
-\end{applemma}
+```
 
-**Proof.** The case $m=n=1$ follows from Lemma \ref{lemma:two}.
-We now use an induction argument, assuming that the result of Lemma \ref{lemma:one} holds for some given order of $(m,n)$ with $m,n\geq 1$ and show the the result continues to hold for $(m+1,n)$ or $(m,n+1)$.
+```{proof, echo=TRUE}
+The case $m=n=1$ follows from Lemma \@ref(lem:second).
+We now use an induction argument, assuming that the result of Lemma \@ref(lem:first) holds for some given order of $(m,n)$ with $m,n\geq 1$ and show the the result continues to hold for $(m+1,n)$ or $(m,n+1)$.
 We shall treat the case $(m+1,n)$ (where the other case $(m,n+1)$ follows by symmetrical arguments).
 
-Considering random variables $(\bm{V}_{m+1},\bm{H}_n) = (V_1,\ldots,V_m,V_{m+1}, H_1,\ldots,H_n)$, the induction hypothesis
-applies to $(\bm{V}_{m},\bm{H}_n)$ where $\bm{V}_m =(V_1,\ldots,V_m)$.
+Considering random variables $(\bm{V}_{m+1},\bm{H}_n) = (V_1,\ldots,V_m,V_{m+1}, H_1,\ldots,H_n)$, the induction hypothesis applies to $(\bm{V}_{m},\bm{H}_n)$ where $\bm{V}_m =(V_1,\ldots,V_m)$.
 Hence, for any $\bm{v}_m \in \{a,b\}^{m}, \bm{h}_n \in \{a,b\}^{n}$, we have $P_1(\bm{V}_m=\bm{v}_m,\bm{H}_n=\bm{h}_n) = P_2(\bm{V}_m=\bm{v}_m,\bm{H}_n=\bm{h}_n)>0$, $P_1( \bm{H}_n=\bm{h}_n) = P_2(\bm{H}_n=\bm{h}_n)>0$ and consequently
-\begin{align}\label{eqn0}
+\begin{align}(\#eq:0)
 P_1(\bm{V}_m=\bm{v}_m|\bm{H}_n=\bm{h}_n) =P_2(\bm{V}_m=\bm{v}_m |\bm{H}_n=\bm{h}_n).
@@
 Likewise, by the induction hypothesis applied to $(\bm{V}_{m-},\bm{H}_n)$ for $\bm{V}_{m-} = (V_2,\ldots,V_{m+1})$, we have
-\begin{eqnarray}\label{eqn1}
+\begin{eqnarray}(\#eq:0)
  1- P_1(V_{m+1} = b | \bm{H}_n=\bm{h}_n) &=& P_1(V_{m+1} = a |\bm{H}_n=\bm{h}_n)\\
@@
 $$
-for $i=1,2$,  so that we have $P_1( \bm{V}_{m+1}=\bm{v}_{m+1}| \bm{H}_n=\bm{h}_n) = P_2( \bm{V}_{m+1}=\bm{v}_{m+1}| \bm{H}_n=\bm{h}_n)$ by (\ref{eqn0})-(\ref{eqn1}). Consequently, it follows that
+for $i=1,2$,  so that we have $P_1( \bm{V}_{m+1}=\bm{v}_{m+1}| \bm{H}_n=\bm{h}_n) = P_2( \bm{V}_{m+1}=\bm{v}_{m+1}| \bm{H}_n=\bm{h}_n)$ by (\@ref{eq:0})-(\@ref(eq:1)). Consequently, it follows that
 \begin{eqnarray*}
@@
 \end{eqnarray*}
-by $P_1(\bm{H}_n=\bm{h}_n)=P_2(\bm{H}_n=\bm{h}_n)$. As $\bm{v}_{m+1}\in \{a,b\}^{m+1},\bm{h}_n \in \{a,b\}^{n}$ were arbitrary, this completes the proof of Lemma \ref{lemma:one}. $\Box$
+by $P_1(\bm{H}_n=\bm{h}_n)=P_2(\bm{H}_n=\bm{h}_n)$. As $\bm{v}_{m+1}\in \{a,b\}^{m+1},\bm{h}_n \in \{a,b\}^{n}$ were arbitrary, this completes the proof of Lemma \@ref(lem:first). 
+```
 
-\begin{applemma} 
-\label{lemma:two}
+```{lemma, label="second", echo=TRUE}
 Suppose the discrete random vector $(X,Y)$ has support $\{a,b\}\times \{a,b\}$, for some $a\neq b \in \mathbb{R}$,
 under two probability distributions $P_1$ and $P_2$. Let $\E_i$ denote expectation under the distribution $P_i$, $i=1,2$. Then, the following are equivalent
-\begin{enumerate}
-\item $\E_1 X = \E_2 X$, $\E_1 Y = \E_2 Y$, $\E_1 X Y = \E_2 X Y$, where $\E_i$ denotes expectation under $P_i$, $i=1,2$.
-\item $P_1(X=x,Y=y)=P_2(X=x,Y=y)$ for $x,y\in\{a,b\}$, i.e., $(X,Y)$ has the same distribution under $P_1$ and $P_2$.
-\end{enumerate}
-\end{applemma}
 
-**Proof.** Denote the four probabilities by which the random vector $(X,Y)$ assume pairs $(b,b),(b,a),(a,b),(a,a)$, respectively, as $c_1,c_2,c_3,1-c_1-c_2-c_3$ under $P_1$ and $d_1,d_2,d_3,1-d_1-d_2-d_3$ under $P_2$.
+1. $\E_1 X = \E_2 X$, $\E_1 Y = \E_2 Y$, $\E_1 X Y = \E_2 X Y$, where $\E_i$ denotes expectation under $P_i$, $i=1,2$.
+2. $P_1(X=x,Y=y)=P_2(X=x,Y=y)$ for $x,y\in\{a,b\}$, i.e., $(X,Y)$ has the same distribution under $P_1$ and $P_2$.
+
+```
+
+```{proof, echo=TRUE}
+Denote the four probabilities by which the random vector $(X,Y)$ assume pairs $(b,b),(b,a),(a,b),(a,a)$, respectively, as $c_1,c_2,c_3,1-c_1-c_2-c_3$ under $P_1$ and $d_1,d_2,d_3,1-d_1-d_2-d_3$ under $P_2$.
 The conditions $\E_1 X = \E_2 X$, $\E_1 Y = \E_2 Y$, $\E_1 X Y = \E_2 X Y$ with $b\neq a$ imply that
@@
 As $(2c_1+c_2+c_3) = (2d_1+d_2+d_3)$ and $b\neq a$, we conclude that $c_1=d_1$, from which it follows that $c_2=d_2$ and $c_3=d_3$.
-Hence, claim 1 implies claim 2 in Lemma \ref{lemma:two}.
+Hence, claim 1 implies claim 2 in Lemma \@ref(lem:second).
 Claim 2 also trivially implies claim 1.
-$\Box$
+```
 
-\begin{apptheorem} 
-\label{thm:two}
+```{theorem, label="two", echo=TRUE}
 Let $m,n\geq 1$ and suppose the random vector $(\bm{V}_m,\bm{H}_n) = (V_1,\ldots,V_m, H_1,\ldots,H_n)$ follows a RBM$(m,n,\bm{\theta})$ distribution.  Then, the following are equivalent
-\begin{enumerate}
-\item $\E V_i=0$, $\E H_j=0$ and $\E V_i H_j =0$ for any $i=1,\ldots,m$, $j=1,\ldots,n$.
-\item $\bm{\theta} = \bm{0}\in \mathbb{R}^{m+n+mn}$.
-\item $V_1,\ldots,V_m, H_1,\ldots,H_n$ are iid Bernoulli$(1/2)$ random variables.\\
-\end{enumerate}
-\end{apptheorem}
 
-**Proof of Theorem \ref{thm:two}.**  We first establish the equivalence of claims 2 and 3.
+1. $\E V_i=0$, $\E H_j=0$ and $\E V_i H_j =0$ for any $i=1,\ldots,m$, $j=1,\ldots,n$.
+2. $\bm{\theta} = \bm{0}\in \mathbb{R}^{m+n+mn}$.
+3. $V_1,\ldots,V_m, H_1,\ldots,H_n$ are iid Bernoulli$(1/2)$ random variables.
+
+```
+
+```{proof, echo=TRUE}
+We first establish the equivalence of claims 2 and 3.
 If claim 2 holds, then claim 3 follows easily from the resulting uniform cell probabilities: $P(V_1= v_1,\ldots,V_m= v_m, H_1= h_1,\ldots,H_n=h_n|\bm{\theta})=2^{-mn}$ for any $v_i,h_j\in\{\pm 1\}$, $i=1,\ldots,m$, $j=1,\ldots,n$.
@@
 
-To establish the equivalence of claims 1 and 2 in Theorem \ref{thm:two} under the RBM$(m,n,\bm{\theta})$ model, we note that claim 2 again implies claim 3, which then easily implies claim 1.
-By Theorem \ref{thm:one}, if claim 1 holds then the only possibility for this is $\bm{\theta} = \bm{0} \in \mathbb{R}^{m+n+mn}$.
-$\Box$
+To establish the equivalence of claims 1 and 2 in Theorem \@ref(thm:two) under the RBM$(m,n,\bm{\theta})$ model, we note that claim 2 again implies claim 3, which then easily implies claim 1.
+By Theorem \@ref(thm:one), if claim 1 holds then the only possibility for this is $\bm{\theta} = \bm{0} \in \mathbb{R}^{m+n+mn}$.
+```
+
+# Flexibility of the RBM model {#appendix-rbm2}
+
+In an RBM model with enough hidden variables, parameter values may be chosen to match any given cell probabilities with arbitrary closeness.
+Additionally, when one or more of the cell probabilities (\@ref(eq:0)) are zero, the corresponding RBM probabilities may never be identically zero (due to exponential terms in the model) but parameters can be still selected to make the appropriate RBM cell probabilities arbitrarily small.
+We show this for a model with two visible variables $(V_1,V_2)$ and one hidden $H_1$.
+
+To demonstrate, we assume $p_{(-1,-1)}>0$ (without loss of generality) in the specified cell probabilities (\@ref(eq:0)) and replace parameters $\theta_{11},\theta_{21}$ with $\Delta_1 \equiv \theta_{11} +\theta_{21}$ and $\Delta_2 \equiv \theta_{11} -\theta_{21}$.
+We may then prescribe values of $\theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \Delta_1,\Delta_2$  so that the model probability ratio
+$$
+P(V_1=v_1,V_2=v_2|\theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \Delta_1,\Delta_2)/  P(V_1=-1,V_2=-1| \theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \Delta_1,\Delta_2)
+$$
+matches the corresponding ratio $p_{(v_1,v_2)}/p_{(-1,-1)}$ over three values of $(v_1,v_2) = (1,-1),(-1,1),(1,1)$.
+
+For instance, assuming the cell probabilities from (\@ref(eq:0)) are all positive, these probabilities can be exactly reproduced by choosing
+\begin{eqnarray*}
+\theta_{v_1} &=& \frac{1}{2} \log\left( \frac{p_{(1,-1)}}{p_{(-1,-1)}} \frac{\exp(\theta_{h_1}-\Delta_1) + \exp(-\theta_{h_1}+\Delta_1)}{\exp(\theta_{h_1}+\Delta_2) + \exp(-\theta_{h_1}-\Delta_2)}  \right),\\
+\theta_{v_2}&= &\frac{1}{2} \log\left( \frac{p_{(-1,1)}}{p_{(-1,-1)}} \frac{\exp(\theta_{h_1}-\Delta_1) + \exp(-\theta_{h_1}+\Delta_1)}{\exp(\theta_{h_1}-\Delta_2) + \exp(-\theta_{h_1}+\Delta_2)}  \right)
+\end{eqnarray*}
+and selecting $\theta_{h_1}, \Delta_1,\Delta_2$ to solve
+\begin{equation}
+(\#eq:1)
+\frac{p_{(1,1)}p_{(-1,-1)}}{p_{(-1,1)}p_{(1,-1)}} = \frac{\ell(|\theta_{h_1}|) + \ell(|\Delta_1|) }{\ell(|\theta_{h_1}|) + \ell(|\Delta_2|)},
+\end{equation}
+based on a monotonically increasing function $\ell(x)\equiv \exp(-2x)+\exp(2x)$, $x \geq 0$.
+If  $[p_{(1,1)}p_{(-1,-1)}]/ [p_{(-1,1)}p_{(1,-1)}] \geq 1$, one can pick any values for $\theta_{h_1}, \Delta_2\in \mathbb{R}$ and solve (\@ref(eq:1)) for $|\Delta_1|$; likewise, when $[p_{(1,1)}p_{(-1,-1)}]/ [p_{(-1,1)}p_{(1,-1)}] < 1$ in (\@ref(eq:1))}), one may solve for $|\Delta_2|$ upon choosing any values for $\theta_{h_1}, \Delta_1\in \mathbb{R}$.
+
+Alternatively, if exactly one specified cell probability in (\@ref(eq:0)) is zero, say $p_{(1,1)}$ (without loss of generality), we can select parameters $\theta_{v_1},\theta_{v_2}$ as above based on a sequence $(\theta_{h_1}, \Delta_{1}, \Delta_{2}) \equiv (\theta_{h_1}^{(m)}, \Delta_{1}^{(m)}, \Delta_{2}^{(m)})$, $m\in\{1,2,\ldots,\}$ of the remaining parameter values such that $\lim_{m\to \infty}|\Delta_{1}^{(m)}| = \infty$ and $\lim_{m\to \infty}  (|\theta_{h_1}^{(m)}| + |\Delta_{2}^{(m)}|)/|\Delta_{1}^{(m)}|=0$ hold.
+This guarantees that the resulting RBM model matches the given cell probabilities (\@ref(eq:0)) in the limit:
+\begin{equation}
+(\#eq:2)
+\lim_{m\to \infty}P(V_1=v_1,V_2=v_2| \theta_{v_1}, \theta_{v_2}, \theta_{h_1},\Delta_1,\Delta_2) = p_{(v_1,v_2)},\quad (v_1,v_2)\in\{(\pm 1,\pm 1)\}.
+\end{equation}  
+If exactly two specified probabilities in (\@ref(eq:0)) are zero, say  $p_{(1,1)}$ and $p_{(-1,1)}$ (without loss of generality), then a limit approximation as in (\@ref(eq:2)) follows by picking $\theta_{v_1}$ as above based on any choices of $(\theta_{h_1}, \Delta_1,\Delta_2)$ and choosing  a sequence of $\theta_{v_2}\equiv \theta_{v_2}^{(m)}$ values for which $\theta_{v_2}^{({m})} \rightarrow -\infty$.
+
+# References {-}

### 2017-07-08 writing/paper-rbm/chapters/rbm.Rmd
paper to bookdown

--- 2017-06-13 writing/draft.Rmd
@@
----
-title: "Properties and Bayes fitting of restricted Boltzmann machines"
-author: 
-  - name: Andee Kaplan
-    university: Iowa State University
-    department: Department of Statistics
-  - name: Daniel Nordman
-    university: Iowa State University
-    department: Department of Statistics
-  - name: Stephen Vardeman
-    university: Iowa State University
-    department: |
-      | Department of Statistics and Department of Industrial and
-      | Manufacturing Systems Engineering
-output: 
-  pdf_document:
-    keep_tex: true
-    fig_caption: yes
-    includes:
-      in_header: latex/header.tex
-    number_sections: yes
-    template: latex/jasa_article.tex
-bibliography: references/refs.bib
-abstract: |
-  A restricted Boltzmann machine (RBM) is an undirected graphical model constructed for discrete or continuous random variables, with two layers, one hidden and one visible, and no conditional dependency within a layer.
-In recent years, RBMs have risen to prominence due to their connection to deep learning.
-By treating a hidden layer of one RBM as the visible layer in a second RBM, a deep architecture can be created.
-RBMs are thought to thereby have the ability to encode very complex and rich structures in data, making them attractive for supervised learning.
-However, the generative behavior of RBMs is largely unexplored.
-In this paper, we discuss the relationship between RBM parameter specification in the binary case and the tendency to exhibit model properties such as degeneracy, instability and uninterpretability.
-We also describe the difficulties that arise in likelihood-based and Bayes fitting of such (highly flexible) models, especially as Gibbs sampling (quasi-Bayes) methods are often advocated for the RBM model structure.
-keywords: Degeneracy, Instability, Classification, Deep Learning, Graphical Models
-blind: false
----
-
-```{r libraries_options, echo=FALSE, message=FALSE, warnings=FALSE}
-library(knitr)
-library(ggplot2)
-library(dplyr)
-library(tidyr)
-library(xtable)
-
-set.seed(1022)
-
-opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE)
-theme_set(theme_bw(base_family="serif"))
-
-source("helpers/functions.R")
+```{r helpers, echo=FALSE, message=FALSE, warnings=FALSE}
+source("../resources/code/functions.R")
 ```
@@
 
-```{r sample-models, fig.show='hide', fig.height=2}
-load("data/params_theta.Rdata")
+The data mining and machine learning communities have recently shown great interest in deep learning, specifically in stacked restricted Boltzmann machines (RBMs) [see @salakhutdinov2009deep; @salakhutdinov2012efficient; @srivastava2013modeling; @le2008representational for examples].
+A RBM is a probabilistic undirected graphical model (for discrete or continuous random variables) with two layers, one hidden and one visible, with no conditional dependency within a layer [@smolensky1986information].
+These models have reportedly been used with success in classification of images [@larochelle2008classification; @srivastava2012multimodal].
+However, the model properties are largely unexplored in the literature and the commonly cited fitting methodology remains heuristic-based and relies on rough approximation [@hinton2006fast].
+In this paper, we provide steps toward a fuller understanding of the model class and its properties from the perspective of statistical model theory, and we then explore the possibility of a rigorous fitting methodology.
+We find the RBM model class to be concerning in two fundamental ways.
+
+First, the models can be unsatisfactory as conceptualizations of how data are generated.
+That is, recalling @fisher1922mathematical, the aim of a statistical model is to represent data in a compact way.
+Neyman and Box further state that a model should "provide an explanation of the mechanism underlying the observed phenomena" [@lehmann1990model; @box1967discrimination].
+At issue, simulation from RBMs can often produce data lacking realistic variability so that such models may thereby fail to satisfactorily reflect a true data generation process.
+Such behavior relates to model degeneracy (or near degeneracy), which is a statistical concern in that most data processes being modeled are realistically not degenerate in their spectrum of potential outcomes.
+For example, when sampling data from a nearly degenerate RBM, only a small number of output possibilities have probability greater than zero, and thus a sample of images will all be copies of the same one image (or small number of images).
+An example of ten 4-pixel images simulated from a nearly degenerate RBM model is compared to ten 4-pixel images simulated from a non-degenerate RBM model in Figure \@ref(fig:sample-models).
+The degenerate model places almost all probability on one outcome, causing the image to be generated repeatedly, whereas the non-degenerate model shows more potentially realistic variation.
+\par
+
+```{r sample-models, fig.show='hold', fig.height=2, fig.cap='Ten 4-pixel images simulated from a degenerate model (a) compared to ten 4-pixel images simulated from a non-degenerate model (b).
+The degenerate model places almost all probability on one outcome, causing the image to be generated repeatedly, whereas the non-degenerate model shows more realistic variation.', out.width='100%'}
+load("../resources/data/params_theta.Rdata")
 params_good <- list(main_hidden = sample.params %>% ungroup() %>% filter(!near_hull) %>% select(starts_with("h"), -H) %>% data.matrix(),
@@
 
-The data mining and machine learning communities have recently shown great interest in deep learning, specifically in stacked restricted Boltzmann machines (RBMs) [see @salakhutdinov2009deep; @salakhutdinov2012efficient; @srivastava2013modeling; @le2008representational for examples].
-A RBM is a probabilistic undirected graphical model (for discrete or continuous random variables) with two layers, one hidden and one visible, with no conditional dependency within a layer [@smolensky1986information].
-These models have reportedly been used with success in classification of images [@larochelle2008classification; @srivastava2012multimodal].
-However, the model properties are largely unexplored in the literature and the commonly cited fitting methodology remains heuristic-based and relies on rough approximation [@hinton2006fast].
-In this paper, we provide steps toward a thorough understanding of the model class and its properties from the perspective of statistical model theory, and then explore the possibility of a rigorous fitting methodology.
-We find the RBM model class to be concerning in two fundamental ways.
-
-First, the models are often unsatisfactory as conceptualizations of how data are generated.
-Recalling @fisher1922mathematical, the aim of a statistical model is to represent data in a compact way.
-Neyman and Box further state that a model should "provide an explanation of the mechanism underlying the observed phenomena" [@lehmann1990model; @box1967discrimination].
-At issue, RBMs  often produce data lacking realistic variability and may thereby fail to provide a satisfactory conceptualization of a data generation process.
-For example, when generating data from a degenerate RBM, only a small number of output possibilities have probability greater than zero, and thus a sample of images will all be copies of the same one image (or small number of images).
-An example of ten 4-pixel images simulated from a degenerate model is compared to ten 4-pixel images simulated from a non-degenerate model in Figure \ref{fig:sample-models}.
-The degenerate model places almost all probability on one outcome, causing the image to be generated repeatedly, whereas the non-degenerate model shows more realistic variation.
-\begin{figure}
-\includegraphics{draft_files/figure-latex/sample-models-2.pdf} \\[.1in]
-\includegraphics{draft_files/figure-latex/sample-models-1.pdf}
-\caption{Ten 4-pixel images simulated from a degenerate model (a) compared to ten 4-pixel images simulated from a non-degenerate model (b).
-The degenerate model places almost all probability on one outcome, causing the image to be generated repeatedly, whereas the non-degenerate model shows more realistic variation.}
-\label{fig:sample-models}
-\end{figure}
-
-In addition to such degeneracy, we find that RBMs can easily exhibit types of instability.
+In addition to such degeneracy, we find that RBMs can easily exhibit types of instability, related to how sensitive the model probability structure can be to small changes in data outcomes.
 In practice, this may be seen when a single pixel change in an image results in a wildly different classification in an image classification problem.
-Such model properties have recently been documented in RBMs [@li2014biclustering], as well as other deep architectures [@szegedy2013intriguing; @nguyen2014deep].
-We investigate the presence of these properties for the RBM class of models in Section \ref{explorations-of-model-properties-through-simulation} through simulations of small, manageable examples.
-
-In addition to model properties, we also explore the issues with fitting these models.
+Occurrences of such behavior have recently been documented in RBMs [@li2014biclustering], as well as other deep architectures [@szegedy2013intriguing; @nguyen2014deep].
+We describe potential issues of model instability, degeneracy and uninterpretability for the RBM class, which are related properties, and examine the presence of these in Section \@ref(explorations-of-model-properties-through-simulation) through simulations of small, manageable examples.
+
+In addition to model properties, we also explore the quality of estimation in fitting these models.
 The fitting can be problematic for two reasons, the first being computational and the second concerning flexibility.
 As the size of these models grows, both maximum likelihood and Bayesian methods of fitting quickly become intractable.
-The literature often suggests Markov chain Monte Carlo (MCMC) tools for approximate maximization of likelihoods to fit these models (e.g., Gibbs sampling to exploit conditional structure in hidden and visible variables), but little is said about achieving convergence [@hinton2010practical; @hinton2006fast].
+The literature often suggests Markov chain Monte Carlo (MCMC) tools for approximate maximization of likelihoods to fit these models (e.g., Gibbs sampling to exploit conditional structure in hidden and visible variables), but little is said about the attributes of realized estimates [@hinton2010practical; @hinton2006fast].
 Related to this, these MCMC algorithms require updating potentially many latent variables (hiddens) which can critically influence convergence in MCMC-based likelihood methods.
-
-In Section \ref{bayesian-model-fitting}, we compare three Bayesian techniques involving MCMC approximations that are computationally more accessible than direct maximum likelihood, which also aim to avoid parts of a RBM parameter space that yield unattractive models.
+Applying basic statistical principles in fitting RBM models of tractable sizes, we compare three fully Bayesian techniques involving MCMC, which are computationally more accessible than direct maximum likelihood and also aim to avoid parts of a RBM parameter space that yield unattractive models.
 As might be expected, with greater computational complexity comes an increase in fitting accuracy, but at the cost of practical feasibility.
-While the computational concerns are inconvenient, the flexibility issues are more alarming.
-
+
+While the computational concerns are inconvenient in fitting, issues due to model flexibility are potentially more concerning.
 For a RBM model with enough hidden variables, it has been shown that any distribution for the visibles can be approximated arbitrarily well [@le2008representational; @montufar2011refinements; and @montufar2011expressive].
-However, the empirical distribution of a training set of vectors of visibles is the best fitting model for observed cell data.
+However, for cell data, the empirical distribution of an observed training set of visible variables provides a highest likelihood benchmark, before any parametric model class is even introduced and applied to obtain a refinement of model fit.
 As a consequence, we find that any fully principled fitting method based on the likelihood for a RBM with enough hidden variables will seek to reproduce the (discrete) empirical distribution of a training set.
-Thus, there can be no "smoothed distribution" achieved in fitting a RBM model of sufficient size with a rigorous likelihood-based method.
+This aspect can be undesirable, and perhaps even unexpected compared to most modeling scenarios, in that no "smoothed distribution" may result when fitting a RBM model of sufficient size with a rigorous likelihood-based method.
 We are therefore led to be skeptical that models that involve these structures (like a deep Boltzmann machine) can achieve useful prediction or inference in a principled way without intentionally limiting the flexibility of the fitted model.
-This concern with overfitting is addressed via specification of a Bayesian prior in Section \ref{bayesian-model-fitting}.
+
+The concepts of weight-decay (penalization) and sparsity (regularization) are suggested in the literature as practical remedies for both over-fitting and poor mixing in the Markov chain during fitting [@hinton2010practical; @tieleman2008training; @cho2012tikhonov].
+Both $L_1$ and $L_2$ type penalties are mentioned to achieve weight-decay however it is unclear the benefits of these particular forms in different situations.
+Additionally, it is not clear the degree to which regularization and penalization are used by practitioners because they are not (by default) a part of the accepted fitted methodology [@hinton2002training; @carreira2005contrastive].
+In this paper, we attempt to address the concerns with overfitting and poor mixing via specification of a Bayesian prior in Section \@ref(bayesian-model-fitting).
 
 This paper is structured as follows.
-Section \ref{background} formally defines the RBM including the joint distribution of hidden and visible variables and explains the model's connection to deep learning.
+Section \@ref(background) formally defines the RBM including the joint distribution of hidden and visible variables and explains the model's connection to deep learning.
 Additionally, measures of model impropriety and methods of quantifying/detecting it are defined.
-Section \ref{explorations-of-model-properties-through-simulation} details our explorations into the model behavior and propriety (or lack thereof) of the RBM class.
-We discuss three Bayesian fitting techniques intended to avoid model impropriety in Section \ref{model-fitting} and conclude with a discussion in Section \ref{discussion}.
-On-line supplementary materials provide proofs for results on RBM parameterizations and data codings described in Section \ref{data-encoding}.
+Section \@ref(explorations-of-model-properties-through-simulation) details our explorations into the model behavior and potential propriety issues with the RBM class.
+We examine three Bayesian fitting techniques intended to avoid model impropriety in Section \@ref(model-fitting) and conclude with a discussion in Section \@ref(discussion).
+Appendix \@ref(appendix-rbm) provides proofs for results on RBM parameterizations and data codings described in Section \@ref(data-encoding).
+
+While applications of the RBM have claimed some successes in classification problems, it is unclear if the model class allows one to go beyond fitting to other statistical matters of importance in using the models, such as quantification of the uncertainty of estimation and the formulation of predictive distributions.
+These are closely tied to the probability properties of the RBM model for explaining data generation and it is hoped that the current exposition contributes to a better understanding in this direction.
 
@@
 A RBM architecture has two layers, hidden ($\mathcal{H}$) and visible ($\mathcal{V}$), with no dependency connections within a layer.
-An example of this structure is in Figure \ref{fig:rbm} with the hidden nodes indicated by gray circles and the visible nodes indicated by white circles.
-
-\begin{figure}[ht]
+An example of this structure is in Figure \@ref(fig:rbm) with the hidden nodes indicated by gray circles and the visible nodes indicated by white circles.
+\begin{figure}
   \centering
-  \resizebox{\linewidth}{!}{\input{tikz/rbm.tikz}}
+  \resizebox{\linewidth}{!}{\input{../resources/images/tikz/rbm.tikz}}
   \caption{An example restricted Boltzmann machine (RBM), consisting of two layers, one hidden ($\mathcal{H}$) and one visible ($\mathcal{V}$), with no connections within a layer.
@@
 \end{figure}
-
 A common use for RBMs is to create features for use in classification.
@@
 ### Joint distribution
+
 Let $\boldsymbol x = (h_1, \dots, h_\nh, v_1,\dots,v_\nv)$ represent the states of the visible and hidden nodes in a RBM for some integers $\nv, \nh \ge 1$.
 Each single binary random variable, visible or hidden, will take its values in a common coding set $\mathcal{C}$, where we allow one of two possibilities for the coding set, $\mathcal{C}=\{0,1\}$ or $\mathcal{C}=\{-1,1\}$, with "$1$" always indicating the "high" value of the variable.
-While $\mathcal{C}=\{0,1\}$ may be a natural starting point, we argue in Section \ref{explorations-of-model-properties-through-simulation} that the coding $\mathcal{C}=\{-1,1\}$ induces more interpretable model properties for the RBM.
+While $\mathcal{C}=\{0,1\}$ may be a natural starting point, we argue in Section \@ref(explorations-of-model-properties-through-simulation) that the coding $\mathcal{C}=\{-1,1\}$ induces more interpretable model properties for the RBM.
 A standard parametric form for probabilities corresponding to a potential vector of states, $X = (H_1, \dots, H_\nh, V_1,\dots,V_\nv)$, for the nodes is
-
 \begin{align}
-\label{eqn:pmf}
+(\#eq:pmf)
 f_{\boldsymbol \theta} (\boldsymbol x)\equiv P_{\boldsymbol \theta}(\boldsymbol X = \boldsymbol x) = \frac{\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}{\gamma(\boldsymbol \theta)}, \quad \boldsymbol x \in \mathcal{C}^{\nh + \nv}
 \end{align}
-
 where $\boldsymbol \theta \equiv (\theta_{11}, \dots, \theta_{1\nh}, \dots, \theta_{\nv 1}, \dots, \theta_{\nv \nh}, \theta_{v_1}, \dots, \theta_{v_\nv}, \theta_{h_1}, \dots, \theta_{h_\nh}) \in \mathbb{R}^{\nv + \nh + \nv*\nh}$ denotes the vector of model parameters and the denominator
-$$\gamma(\boldsymbol \theta) = \sum_{\boldsymbol x \in \mathcal{C}^{\nh+\nv}}\exp\left(\sum_{i = 1}^\nv \sum_{j=1}^\nh \theta_{ij} v_i h_j + \sum_{i = 1}^\nv\theta_{v_i} v_i + \sum_{j = 1}^\nh\theta_{h_j} h_j\right)
-$$ 
-is the normalizing function that ensures the probabilities (\ref{eqn:pmf}) sum to one. For $\boldsymbol x = (h_1, \dots, h_\nh, v_1, \dots, v_\nh) \in \mathcal{C}^{\nv + \nh}$ and 
+$$
+\gamma(\boldsymbol \theta) = \sum_{\boldsymbol x \in \mathcal{C}^{\nh+\nv}}\exp\left(\sum_{i = 1}^\nv \sum_{j=1}^\nh \theta_{ij} v_i h_j + \sum_{i = 1}^\nv\theta_{v_i} v_i + \sum_{j = 1}^\nh\theta_{h_j} h_j\right)
+$$ 
+is the normalizing function that ensures the probabilities (\@ref(eq:pmf)) sum to one. For $\boldsymbol x = (h_1, \dots, h_\nh, v_1, \dots, v_\nh) \in \mathcal{C}^{\nv + \nh}$ and 
 \begin{align}
-\boldsymbol t(\boldsymbol x) &= (h_1, \dots, h_\nh, v_1, \dots, v_\nv, v_1 h_1, \dots, v_\nv h_\nh) \in \mathcal{C}^{\nh + \nv + \nh*\nv}, \label{eqn:t}
+\boldsymbol t(\boldsymbol x) &= (h_1, \dots, h_\nh, v_1, \dots, v_\nv, v_1 h_1, \dots, v_\nv h_\nh) \in \mathcal{C}^{\nh + \nv + \nh*\nv}, (\#eq:t)
 \end{align}
-let $\mathcal{T} = \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{C}^{\nv + \nh}\} \subset \mathbb{R}^{\nv + \nh + \nv * \nh}$ be the set of possible values for the vector of variables needed to compute probabilities (\ref{eqn:pmf}) in the model, and write $Q_{\boldsymbol \theta}(\boldsymbol x) = \sum\limits_{i = 1}^\nh\sum\limits_{j=1}^\nv \theta_{ij} h_i v_j + \sum\limits_{i = 1}^\nh\theta_{hi} h_i + \sum\limits_{j = 1}^\nv\theta_{vj} v_j$ for the "neg-potential" function.
+let $\mathcal{T} = \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{C}^{\nv + \nh}\} \subset \mathbb{R}^{\nv + \nh + \nv * \nh}$ be the set of possible values for the vector of variables needed to compute probabilities (\@ref(eq:pmf)) in the model, and write $Q_{\boldsymbol \theta}(\boldsymbol x) = \sum\limits_{i = 1}^\nh\sum\limits_{j=1}^\nv \theta_{ij} h_i v_j + \sum\limits_{i = 1}^\nh\theta_{hi} h_i + \sum\limits_{j = 1}^\nv\theta_{vj} v_j$ for the "neg-potential" function.
 The RBM model is parameterized by $\boldsymbol \theta$ containing two types of parameters, main effects and interaction effects.
-The main effects parameters ($\{\theta_{v_i}, \theta_{h_j}\}_{\substack{i = 1, \dots, \nv,\\j = 1, \dots, \nh}}$) weight the values of the visible $v_i$ and hidden $h_j$ nodes in probabilities (\ref{eqn:pmf}) and the interaction effect parameters ($\theta_{ij}$) weight the values of the connections $v_i h_j$, or dependencies, between hidden and visible layers.
+The main effects parameters ($\{\theta_{v_i}, \theta_{h_j}\}_{\substack{i = 1, \dots, \nv,\\j = 1, \dots, \nh}}$) weight the values of the visible $v_i$ and hidden $h_j$ nodes in probabilities (\@ref(eq:pmf)) and the interaction effect parameters ($\theta_{ij}$) weight the values of the connections $v_i h_j$, or dependencies, between hidden and visible layers.
 
 Due to the potential size of the model, the normalizing constant $\gamma(\boldsymbol \theta)$ can be practically impossible to calculate, making simple estimation of the model parameter vector problematic.
-A kind of Gibbs sampling can be tried (due to the conditional architecture of the RBM, i.e.
+In model fitting, a kind of Gibbs sampling can be tried due to the conditional architecture of the RBM (i.e.
 visibles given hiddens or vice verse).
-Specifically, the conditional independence of nodes in each layer (given those nodes in the other layer) allows for stepwise simulation of both hidden layers and model parameters (e.g., see the contrastive divergence of @hinton2002training or Bayes methods in Section \ref{model-fitting}).
+Specifically, the conditional independence of nodes in each layer (given those nodes in the other layer) allows for stepwise simulation of both hidden layers and model parameters (e.g., see the contrastive divergence of @hinton2002training or Bayes methods in Section \@ref(model-fitting)).
 
@@
 
-## Degeneracy, instability, and uninterpretability... Oh my!
-The highly flexible nature of a RBM (having as it does $\nh + \nv + \nh*\nv$ parameters) creates at least three kinds of potential model impropriety that we will call *degeneracy*, *instability*, and *uninterpretability*.
-In this section we define these characteristics, consider how to detect them in a RBM, and point out relationships among them.
+## Degeneracy, instability, and uninterpretability
+The highly flexible nature of a RBM (having as it does $\nh + \nv + \nh*\nv$ parameters) creates at least three kinds of potential issues in model impropriety that we will call *degeneracy*, *instability*, and *uninterpretability*.
+In this section we define these characteristics, consider how to quantify these in a RBM, and point out relationships among them.
 
 ### Near-degeneracy
+
 In Random Graph Model theory, *model degeneracy* means there is a disproportionate amount of probability placed on only a few elements of the sample space, $\mathcal{X}$, by the model [@handcock2003assessing].
@@
 Let $C$ denote the convex hull of the potential outcomes of sufficient statistics, $\{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{X}\}$, under the model above.
-@handcock2003assessing classify an exponentially parametrized random graph model at $\boldsymbol \theta$ as *near-degenerate* if the mean value of the vector of sufficient statistics under $\boldsymbol \theta$, $\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta}\boldsymbol t( \boldsymbol X)$, is close to the boundary of $C$.
-This makes sense because if a model is near-degenerate in the sense that only a small number of elements of the sample space $\mathcal{X}$ have positive probability, the expected value $\text{E}_{\boldsymbol \theta}\boldsymbol t( \boldsymbol X)$ is an average of that same small number of values of $t( \boldsymbol x)$ and can be expected to *not* be pulled deep into the interior of $C$.
+@handcock2003assessing classifies an exponentially parametrized random graph model at $\boldsymbol \theta$ as *near-degenerate* if the mean value of the vector of sufficient statistics under $\boldsymbol \theta$, $\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta}\boldsymbol t( \boldsymbol X)$, is close to the boundary of $C$.
+Intuitively, if a model is near-degenerate in the sense that only a small number of elements of the sample space $\mathcal{X}$ have positive probability, the expected value $\text{E}_{\boldsymbol \theta}\boldsymbol t( \boldsymbol X)$ is an average of that same small number of values of $t( \boldsymbol x)$ (defining the boundary of the hull $C$) and can be expected to *not* be pulled deep into the interior of $C$.
 
 A RBM model can be thought to exhibit an analogous form of *near-degeneracy* when there is a disproportionate amount of probability placed on a small number of elements in the sample space of visibles and hiddens, $\mathcal{C}^{\nv + \nh}$.
-Using the idea of @handcock2003assessing, when the random vector $\boldsymbol t(\boldsymbol x) = (v_1, \dots, v_\nv, h_1, \dots, h_\nh, v_1 h_1, \dots, v_V h_\nh ) \in \mathcal{T} \equiv \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{C}^{\nh + \nv} \}$ from (\ref{eqn:t}), appearing in the neg-potential function $Q_{\boldsymbol \theta}(\cdot)$, has a mean vector $\boldsymbol \mu(\boldsymbol \theta) \in \mathbb{R}^{\nv + \nh + \nv * \nh}$ close to the boundary of the convex hull of $\mathcal{T}$, then the RBM model can be said to exhibit near-degeneracy at $\boldsymbol \theta \in \mathbb{R}^{\nv + \nh + \nh*\nv}$.
+Using the idea of @handcock2003assessing, when the random vector $\boldsymbol t(\boldsymbol x) = (v_1, \dots, v_\nv, h_1, \dots, h_\nh, v_1 h_1, \dots, v_V h_\nh ) \in \mathcal{T} \equiv \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{C}^{\nh + \nv} \}$ from (\@ref(eq:t)), appearing in the neg-potential function $Q_{\boldsymbol \theta}(\cdot)$, has a mean vector $\boldsymbol \mu(\boldsymbol \theta) \in \mathbb{R}^{\nv + \nh + \nv * \nh}$ close to the boundary of the convex hull of $\mathcal{T}$, then the RBM model can be said to exhibit near-degeneracy at $\boldsymbol \theta \in \mathbb{R}^{\nv + \nh + \nh*\nv}$.
 Here the mean of $\boldsymbol t(\boldsymbol x)$ is
@@
 Instability can be roughly thought of as excessive sensitivity in the model, where small changes in the components of potential data outcomes, $\boldsymbol x$, may lead to substantial changes in the probability function $f_{\boldsymbol \theta}(\boldsymbol x)$.
+Furthermore, model instability can be viewed on a spectrum of potential sensitivity in probability structure, with model degeneracy included as a extreme or limiting case of instability.
 To quantify "instability" more rigorously (particularly beyond the definition given by @schweinberger2011instability) it is useful to consider how RBM models might be expanded to incorporate more and more visibles.
@@
 \begin{align}
-\frac{1}{\nv} \log \left[\frac{\max\limits_{(v_1, \dots, v_\nv) \in \mathcal{C}^\nv}P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}{\min\limits_{(v_1, \dots, v_\nv) \in \mathcal{C}^\nv}P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}\right] \equiv \frac{1}{\nv} \text{ELPR}(\boldsymbol \theta_\nv) \label{eqn:elpr}
+\frac{1}{\nv} \log \left[\frac{\max\limits_{(v_1, \dots, v_\nv) \in \mathcal{C}^\nv}P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}{\min\limits_{(v_1, \dots, v_\nv) \in \mathcal{C}^\nv}P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}\right] \equiv \frac{1}{\nv} \text{ELPR}(\boldsymbol \theta_\nv) (\#eq:elpr)
 \end{align}
-where $P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv) \propto \sum_{\{h_j\}\in \mathcal{C}}\exp\left(\sum_{i = 1}^\nv \theta_{v_i} v_i + \sum_{j = 1}^\nh \theta_{h_j} h_j + \sum_{i = 1}^\nv \sum_{j = 1}^\nh \theta_{ij} v_i h_j\right)$ is the RBM probability of observing outcome $(v_1, \dots, v_\nv)$ for the visible variables $(V_1, \dots, V_\nv)$ under parameter vector $\boldsymbol \theta_\nv$.
-
-In formulating a RBM model for a potentially large number of visibles (i.e., as $\nv \rightarrow \infty$), we will say that the ratio (\ref{eqn:elpr}) needs to stay bounded for the sequence of RBM models to be stable.
+where $P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv) \propto \sum_{(h_1, \dots, h_\nh) \in \mathcal{C}^\nh}\exp\left(\sum_{i = 1}^\nv \theta_{v_i} v_i + \sum_{j = 1}^\nh \theta_{h_j} h_j + \sum_{i = 1}^\nv \sum_{j = 1}^\nh \theta_{ij} v_i h_j\right)$ is the RBM probability of observing outcome $(v_1, \dots, v_\nv)$ for the visible variables $(V_1, \dots, V_\nv)$ under parameter vector $\boldsymbol \theta_\nv$, after marginalization of hidden variables.
+
+In formulating a RBM model for a potentially large number of visibles (i.e., as $\nv \rightarrow \infty$), we will say that the ratio (\@ref(eq:elpr)) needs to stay bounded for the sequence of RBM models to be stable.
 That is, we make the following convention.
-
-\begin{defn}[S-unstable RBM]
+```{definition, name = "S-unstable RBM", echo=TRUE}
 Let $\boldsymbol \theta_\nv \in \mathbb{R}^{\nv + \nh +  \nh*\nv}, \nv \ge 1$, be an element of a sequence of RBM parameters where the number of hiddens, $\nh \equiv \nh(\nv) \ge 1$, can be an arbitrary function of the number $\nv$ of visibles.
@@
 \end{align*}
-\end{defn}
-In other words, given any $c > 0$, there exists an integer $n_c > 0$ so that $\frac{1}{\nv}\text{ELPR}(\boldsymbol \theta_\nv) > c$ for all $\nv \ge n_c$.
-
+```
+In other words, the RBM model sequence is unstable if, given any $c > 0$, there exists an integer $n_c > 0$ so that $\frac{1}{\nv}\text{ELPR}(\boldsymbol \theta_\nv) > c$ for all $\nv \ge n_c$.
 This definition of *S-unstable* is a generalization or re-interpretation of the "unstable" concept of @schweinberger2011instability in that here RBM models for visibles $(v_1, \dots, v_\nv)$ do not form an exponential family and the dimensionality of $\boldsymbol \theta_\nv$ is not fixed, but rather grows with $\nv$.
 
-S-unstable RBM model sequences are undesirable for several reasons. One is that, as mentioned above, small changes in data can lead to overly-sensitive changes in probability. Consider, for example, 
+S-unstable RBM model sequences are undesirable for several reasons.
+One is that, as mentioned above, small changes in data can lead to overly-sensitive changes in probability under the data model.
+Consider, for example,
 $$
@@
 
-\begin{prop}
-\label{prop:instab}
-Let $c > 0$ and let $\text{ELPR}(\boldsymbol \theta_\nv)$ be as in (\ref{eqn:elpr}) for an integer $\nv \ge 1$.
+```{proposition, label = "instab1", echo=TRUE}
+Let $c > 0$ and let $\text{ELPR}(\boldsymbol \theta_\nv)$ be as in (\@ref(eq:elpr)) for an integer $\nv \ge 1$.
 If $\frac{1}{\nv}\text{ELPR}(\boldsymbol \theta_\nv) > c$, then $\Delta(\boldsymbol \theta_\nv) > c$.
-\end{prop}
-
-In other words, if the probability ratio (\ref{eqn:elpr}) is too large, then a RBM model sequence will exhibit large probability shifts for very small changes in data configurations (i.e., will exhibit instability).
+```
+
+In other words, if the probability ratio (\@ref(eq:elpr)) is too large, then a RBM model sequence will exhibit large probability shifts for very small changes in data configurations (i.e., will exhibit instability).
+Such instability can be a concern in a model for the similar reasons to degeneracy: as outcomes vary over the sample space, the geography of probabilities is extremely rugged, with deep pits following sharp mountains.
 Recall the applied example of RBM models as a means to classify images.
-For data as pixels in an image, the instability result in Proposition \ref{prop:instab} manifests itself as a one pixel change in an image (one component of the visible vector) resulting in a large shift in the probability, which in turn could result in a vastly different classification of the image.
+For data as pixels in an image, the instability result in Proposition \@ref(prp:instab1) manifests itself as a one pixel change in an image (one component of the visible vector) resulting in a large shift in the probability, which in turn could result in a vastly different classification of the image.
 Examples of this behavior have been presented in @szegedy2013intriguing for other deep learning models, in which a one pixel change in a test image results in a wildly different classification.
 
-Additionally, S-unstable RBM model sequences are connected to the near-degeneracy of Section \ref{near-degeneracy} (in which model sequences place all probability on a small portion of their sample spaces).
+Additionally, S-unstable RBM model sequences may be formally connected to the near-degeneracy of Section \@ref(near-degeneracy) (in which model sequences place all probability on a small portion of their sample spaces).
 To see this, define an arbitrary modal set of possible outcomes (i.e.
@@
 
-\begin{prop}
-\label{prop:degen}
-For an S-unstable RBM model sequence and any $0 < \epsilon < 1$, $P_{\boldsymbol \theta_\nv}\left((v_1, \dots, v_\nv) \in M_{\epsilon, \boldsymbol \theta_\nv}\right) \rightarrow 1$ as $\nv \rightarrow \infty$.
-\end{prop}
-
-In other words, S-unstable RBM model sequences are guaranteed to stack up all probability on an arbitrarily narrow set of outcomes for visibles.
-Proofs of Propositions \ref{prop:instab} and \ref{prop:degen} appear in @kaplan2016note.
+```{proposition, label = "degen", echo=TRUE}
+For an S-unstable RBM model sequence and any $0 < \epsilon < 1$, 
+$$
+P_{\boldsymbol \theta_\nv}\left((v_1, \dots, v_\nv) \in M_{\epsilon, \boldsymbol \theta_\nv}\right) \rightarrow 1 \text{ as } \nv \rightarrow \infty.
+$$
+```
+
+In other words, S-unstable RBM model sequences are guaranteed to stack up all probability on a specific set of outcomes for visibles, which could potentially be arbitrarily narrow.
+Proofs of Propositions \@ref(prp:instab1) and \@ref(prp:degen) appear in @kaplan2017note.
 These findings also have counterparts in results in @schweinberger2011instability, but unlike results there, we do not limit consideration to exponential family forms with a fixed number of parameters.
@@
 ### Uninterpretability
+
 For spatial Markov models, @kaiser2007statistical defines a measure of model impropriety he calls *uninterpretability*, which is characterized by dependence parameters in a model being so extreme that marginal mean-structures fail to hold as anticipated by a model statement.
@@
 Note that in a RBM, the parameters $\theta_{v_1}, \dots, \theta_{v_\nv}$ and $\theta_{h_1}, \dots, \theta_{h_\nh}$ are naturally associated with main effects of visible and hidden variables and can be interpreted as (logit functions of) means for variables $V_1, \dots, V_\nv, H_1, \dots, H_\nh$ in a model with no interaction parameters, $\theta_{ij} = 0, i = 1, \dots, \nv, j = 1, \dots, \nh$.
-That is, with no interaction parameters, we have from (\ref{eqn:pmf}) that
+That is, with no interaction parameters, we have from (\@ref(eq:pmf)) that
 $$
@@
 Hence, these main effect parameters have a clear interpretation under an independence model (one with $\theta_{ij} = 0$) but this interpretation can break down as interaction parameters increase in magnitude relative to the size of the main effects.
-In such cases, the main effect parameters $\theta_{v_1}$ and $\theta_{h_j}$ are no longer interpretable as originally intended in the models (statements of marginal means) and the dependence parameters are so large as to dominate the entire model probability structure (also destroying the model interpretation of dependence as local conditional modifications of an overall marginal mean structure).
+In such cases, the main effect parameters $\theta_{v_1}$ and $\theta_{h_j}$ are no longer interpretable in the models (statements of marginal means) and the dependence parameters are so large as to dominate the entire model probability structure (also destroying the model interpretation of dependence as local conditional modifications of an overall marginal mean structure).
+Whether or not parameter interpretation is itself a goal in a given application of RBM models, this concept of interpretation can provide an additional device for examining other aspects of model propriety related to instability and degeneracy.
+As explained by @kaiser2007statistical, models with interpretable dependence parameters typically correspond to non-degenerate models, while degradation in interpretability is often associated with model drift into degeneracy.
 To assess which parameter values $\boldsymbol \theta$ may cause difficulties in interpretation, we use the difference $\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] - \text{E}\left[\boldsymbol X | \boldsymbol \theta^* \right]$ between two model expectations: E$\left[\boldsymbol X | \boldsymbol \theta\right]$ at $\boldsymbol \theta$ and expectations E$\left[\boldsymbol X | \boldsymbol \theta^* \right]$ where $\boldsymbol \theta^*$ matches $\boldsymbol \theta$ for all main effects but otherwise has $\theta_{ij} = 0$ for $i = 1, \dots, \nv, j = 1, \dots, \nh$.
@@
 The quantities to compare in the RBM case are
-
 $$
@@
 ## Tiny example
+
 To illustrate the ideas of model near-degeneracy, instability, and uninterpretability in a RBM, we consider first the smallest possible (toy) example that consists of one visible node $v_1$ and one hidden node $h_1$ that are both binary.
-A schematic of this model can be found in Figure \ref{fig:toymodel}.
+A schematic of this model can be found in Figure \@ref(fig:toymodel).
 Because it seems most common, we shall begin by employing $0/1$ encoding of binary variables (both $h_1$ and $v_1$ taking values in $\mathcal{C} = \{0,1\}$).
-(Eventually we shall argue in Section \ref{data-encoding} that $-1/1$ coding has substantial advantages.)
+(Eventually we shall argue in Section \@ref(data-encoding) that $-1/1$ coding has advantages.)
 
@@
   \centering
-  \resizebox{1cm}{!}{\input{tikz/toymodel.tikz}}
+  \resizebox{1cm}{!}{\input{../resources/images/tikz/toymodel.tikz}}
   \caption{A small example restricted Boltzmann machine (RBM), with two nodes, one hidden and one visible.}
@@
 For this small model, we are able to investigate the symptoms of model impropriety, beginning with near-degeneracy.
-To this end, recall from Section \ref{near-degeneracy} that one characterization requires consideration of the convex hull of possible values of statistics $\boldsymbol t(\boldsymbol x)$,
-$$\mathcal{T} = \{\boldsymbol t(\boldsymbol x): \boldsymbol x = (v_1, h_1) \in \{0,1\}^2\} \equiv \{(v_1, h_1, v_1 h_1): v_1, h_1 \in \{0,1\}\}$$ 
+To this end, recall from Section \@ref(near-degeneracy) that one characterization requires consideration of the convex hull of possible values of statistics $\boldsymbol t(\boldsymbol x)$,
+$$
+\mathcal{T} = \{\boldsymbol t(\boldsymbol x): \boldsymbol x = (v_1, h_1) \in \{0,1\}^2\} \equiv \{(v_1, h_1, v_1 h_1): v_1, h_1 \in \{0,1\}\}
+$$ 
 appearing in the RBM probabilities for this model.
 As this set is in three dimensions, we are able to explicitly illustrate the shape of boundary of the convex hull of $\mathcal{T}$ and explore the behavior of the mean vector $\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta} \boldsymbol t(\boldsymbol x)$ as a function of the parameter vector $\boldsymbol \theta$.
-Figure \ref{fig:toyhull} shows the convex hull of our "statistic space," $\mathcal{T} \subset \{0,1\}^3$, for this toy problem from two perspectives (enclosed by the unit cube $[0,1]^3$, the convex hull of $\{0,1\}^3$).
+Figure \@ref(fig:toyhull) shows the convex hull of our "statistic space," $\mathcal{T} \subset \{0,1\}^3$, for this toy problem from two perspectives (enclosed by the unit cube $[0,1]^3$, the convex hull of $\{0,1\}^3$).
 In this small model, note that the convex hull of $\mathcal{T}$ does not fill the unrestricted hull of $\{0,1\}^3$ because of the relationship between the elements of $\mathcal{T} = \{(v_1, h_1, v_1 h_1 : v_1, h_1 \in \{0, 1\} \}$ (i.e.
 $v_1 h_1 = 1$ only if $v_1 = h_1 = 1$).
+\par
 
@@
     \tdplotsetmaincoords{60}{-60}
-    \input{tikz/toyhull_top.tikz}
+    \input{../resources/images/tikz/toyhull_top.tikz}
 
@@
     \tdplotsetmaincoords{60}{120}
-    \input{tikz/toyhull_top.tikz}
+    \input{../resources/images/tikz/toyhull_top.tikz}
 
@@
   \end{minipage}
-  \caption{The convex hull of the "statistic space" in three dimensions for the toy RBM with one visible and one hidden node.
-The convex hull of $\mathcal{T} = \{\boldsymbol t(\boldsymbol x) : \boldsymbol x \in \mathcal{C}^{\nv + \nh}\}$ does not fill the unit cube because of the relationship between the elements of $\mathcal{T}$.}
+  \caption{Two perspectives of the convex hull of the ``statistic space" in three dimensions for the toy RBM with one visible and one hidden node.}
 \label{fig:toyhull}
@@
 where $\gamma(\boldsymbol \theta) = \sum\limits_{h_1 = 0}^1\sum\limits_{v_1 = 0}^1 \exp(\theta_{11} h_1 v_1 + \theta_{h_1}h_1 + \theta_{v_1}v_1)$.
-The three parametric coordinate functions of $\boldsymbol \mu(\boldsymbol \theta)$ can be represented as in Figure \ref{fig:degen_toy}.
-(Contour plots for three coordinate functions are shown in columns for various values of $\theta_{11}$.) In examining these, we see that as coordinates of $\boldsymbol \theta$ grow larger in magnitude, at least one mean function for the entries of $\boldsymbol t(\boldsymbol x)$ approaches a value 0 or 1, forcing $\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta} \boldsymbol t(\boldsymbol x)$ to be near to the boundary of the convex hull of $\mathcal{T}$, as a sign of model near-degeneracy.
+The three parametric coordinate functions of $\boldsymbol \mu(\boldsymbol \theta)$ can be represented as in Figure \@ref(fig:degen-toy).
+(Contour plots for three coordinate functions are shown in columns for various values of $\theta_{11}$, which can be interpreted here as an absolute log-odds ratio as the visible changes between $0$ and $1$.) In examining these, we see that as coordinates of $\boldsymbol \theta$ grow larger in magnitude, at least one mean function for the entries of $\boldsymbol t(\boldsymbol x)$ approaches a value 0 or 1, forcing $\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta} \boldsymbol t(\boldsymbol x)$ to be near to the boundary of the convex hull of $\mathcal{T}$, as a sign of model near-degeneracy.
 Thus, for a very small example we can see the relationship between values of $\boldsymbol \theta$ and model degeneracy.
-
-```{r near_dependency, fig.cap="\\label{fig:degen_toy}Contour plots for the three parametric mean functions of sufficient statistics for a RBM with one visible and one hidden node.", fig.height = 4, fig.align='center', message=FALSE, cache=TRUE}
+\par
+
+```{r degen-toy, fig.cap="Contour plots for the three parametric mean functions of sufficient statistics for a RBM with one visible and one hidden node.", fig.height = 4, fig.align='center', message=FALSE, cache=TRUE}
 possibles <- stats(1, 1, type = "binary")
@@
                                ifelse(mu_parameter == "mu_h1", "paste(E[theta], h[1])", "paste(E[theta], v[1], h[1])"))) %>%
-  #factor(mu_parameter, labels = paste0(gsub("_", "(theta[", unique(mu_parameter)), "])"))) %>%
   ggplot() +
@@
 
-Secondly, we can look at $\text{ELPR}(\boldsymbol \theta)$ from (\ref{eqn:elpr}) for this tiny model in order to consider model instability as a function of RBM parameters.
-Recall that large values of $\text{ELPR}(\boldsymbol \theta)$ are associated with an extreme sensitivity of the model probabilities $f_{\boldsymbol \theta}(\boldsymbol x)$ to small changes in $\boldsymbol x$ (see Proposition \ref{prop:instab}).
+Secondly, we can look at $\text{ELPR}(\boldsymbol \theta)$ from (\@ref(eq:elpr)) for this tiny model in order to consider model instability as a function of RBM parameters.
+Recall that large values of $\text{ELPR}(\boldsymbol \theta)$ are associated with an extreme sensitivity of the model probabilities $f_{\boldsymbol \theta}(\boldsymbol x)$ to small changes in $\boldsymbol x$ (see Proposition \@ref(prp:instab1)).
 The quantity $\text{ELPR}(\boldsymbol \theta)$ for this small RBM is
-
 \begin{align*}
@@
 
-Figure \ref{fig:instab} shows contour plots of $\text{ELPR}(\boldsymbol \theta)/\nv$ for various values of $\boldsymbol \theta$ in this model with $\nv = 1$.
+Figure \@ref(fig:instab) shows contour plots of $\text{ELPR}(\boldsymbol \theta)/\nv$ for various values of $\boldsymbol \theta$ in this model with $\nv = 1$.
 We can see that this quantity is large for large magnitudes of $\boldsymbol \theta$, especially for large values of the dependence/interaction parameter $\theta_{11}$.
 This suggests instability as $|\boldsymbol \theta|$ becomes large, agreeing also with the concerns about near-degeneracy produced by consideration of $\boldsymbol \mu(\boldsymbol \theta)$.
-
-```{r instab, fig.cap="\\label{fig:instab}$\\text{ELPR}(\\boldsymbol \\theta)/\\nv$ for various values of $\\boldsymbol \\theta$ for the tiny example model.
+\par
+
+```{r instab, fig.cap="$\\text{ELPR}(\\boldsymbol \\theta)/\\nv$ for various values of $\\boldsymbol \\theta$ for the tiny example model.
 Recall here $\\nv$ is the number of visible nodes and here is $1$.
@@
 This difference is given by
-
 \begin{align*}
@@
 Again, we can inspect these coordinate functions of this vector difference to look for a relationship between parameter values and large values of $\text{E}[\boldsymbol X|\boldsymbol \theta] - \text{E}[\boldsymbol X| \boldsymbol \theta^*]$ as a signal of uninterpretability for the toy RBM.
-
-```{r uninterp, fig.cap="\\label{fig:uninterp}The absolute difference between coordinates of model expectations, E$\\left[\\boldsymbol X | \\boldsymbol \\theta\\right]$, and expectations given independence, E$\\left[\\boldsymbol X | \\boldsymbol \\theta^* \\right ]$ for a RBM with one visible and one hidden node.
+\par
+
+```{r uninterp, fig.cap="The absolute difference between coordinates of model expectations, E$\\left[\\boldsymbol X | \\boldsymbol \\theta\\right]$, and expectations given independence, E$\\left[\\boldsymbol X | \\boldsymbol \\theta^* \\right ]$ for a RBM with one visible and one hidden node.
 As an indicator of uninterpretability, note that differences in expectations increase as the dependence parameter $\\theta_{11}$ deviates from zero.", fig.height = 4, fig.align='center', message=FALSE, cache=TRUE, warning=FALSE}
@@
 
-Figure \ref{fig:uninterp} shows that the absolute difference between coordinates of the vector of model expectations, E$\left[\boldsymbol X | \boldsymbol \theta\right]$ and corresponding expectations E$\left[\boldsymbol X | \boldsymbol \theta^*\right ]$ given independence grow for the toy RBM as the values of $\boldsymbol \theta$ are farther from zero, especially for large magnitudes of the dependence parameter $\theta_{11}$.
+Figure \@ref(fig:uninterp) shows that the absolute difference between coordinates of the vector of model expectations, E$\left[\boldsymbol X | \boldsymbol \theta\right]$ and corresponding expectations E$\left[\boldsymbol X | \boldsymbol \theta^*\right ]$ given independence grow for the toy RBM as the values of $\boldsymbol \theta$ are farther from zero, especially for large magnitudes of the dependence parameter $\theta_{11}$.
 This is a third indication that parameter vectors of large magnitude lead to model impropriety in a RBM.
@@
 ### Data encoding
+
 Multiple encodings of the binary variables are possible.
 For example, we could allow hiddens $(H_1, \dots, H_\nh) \in \{0,1\}^\nh$ and visibles $(V_1, \dots, V_\nv) \in \{0,1\}^\nv$, as in the previous sections or we could instead encode the state of the variables as $\{-1,1\}^\nh$ and $\{-1,1\}^\nv$.
-This will result in variables $\boldsymbol t(\boldsymbol X)$ from (\ref{eqn:t}) satisfying $\boldsymbol t(\boldsymbol x) \in \{0,1\}^{\nh + \nv + \nh*\nv}$ or $\boldsymbol t(\boldsymbol x) \in \{-1,1\}^{\nh + \nv + \nh*\nv}$ depending on how we encode "on" and "off" states in the nodes.
-
-The $-1/1$ data encoding has the benefit of providing a guaranteed-to-be non-degenerate model at $\boldsymbol \theta = \boldsymbol 0 \in \mathbb{R}^{\nh + \nv + \nh*\nv}$, where the zero vector then serves as the natural center of the parameter space and induces the simplest possible model properties for the RBM (i.e., at $\boldsymbol \theta = \boldsymbol 0$, all variables are independent and visible variables are independent and uniformly distributed on $\{-1,1\}^\nv$).
+This will result in variables $\boldsymbol t(\boldsymbol X)$ from (\@ref(eq:t)) satisfying $\boldsymbol t(\boldsymbol x) \in \{0,1\}^{\nh + \nv + \nh*\nv}$ or $\boldsymbol t(\boldsymbol x) \in \{-1,1\}^{\nh + \nv + \nh*\nv}$ depending on how we encode "on" and "off" states in the nodes.
+
+The $-1/1$ data encoding has the benefit of providing a guaranteed-to-be non-degenerate model at $\boldsymbol \theta = \boldsymbol 0 \in \mathbb{R}^{\nh + \nv + \nh*\nv}$, where the zero vector then serves as the natural center of the parameter space and induces the simplest possible model properties for the RBM (i.e., at $\boldsymbol \theta = \boldsymbol 0$, all variables are independent uniformly distributed on $\{-1,1\}^\nv$).
 The proof of this and further exploration of the equivalence of the $\boldsymbol \theta$ parameterization of the RBM model class and parameterization by $\boldsymbol \mu(\boldsymbol \theta)$ is in the on-line supplementary materials.
 Hence, while from some computing perspectives $0/1$ coding might seem most natural, the $-1/1$ coding is far more convenient and interpretable from the point of view of statistical modeling, where it makes parameters simply interpreted in terms of symmetrically defined main effects and interactions.
+Under the data encoding $-1/1$, the parameter space centered at 0 is also helpful for framing parameter configurations that are undesirably large (i.e., these are naturally parameters that have moved too far away from $\bm{0}$ where the RBM model is anchored to be trivially describable and completely problem-free).
 In light of all of these matters we will henceforth employ the $-1/1$ coding.
@@
 ## Exploring manageable examples
+
 ```{r degen-data, message=FALSE, warning=FALSE, cache=TRUE}
@@
 three_ways$Visibles <- factor(three_ways$Visibles, levels=rev(unique(three_ways$Visibles)))
-
 ```
@@
 For $\nh, \nv \in \{1, \dots, 4\}$, we sample `r dim(res$samp[[1]])[1]` values of $\boldsymbol \theta$ with various magnitudes (details to follow).
-For each set of parameters we then calculate metrics of model impropriety introduced in Section \ref{degeneracy-instability-and-uninterpretability-oh-my} based on $\boldsymbol \mu(\boldsymbol \theta)$, $\text{ELPR}(\boldsymbol \theta)/\nv$, and the absolute coordinates of $\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] -   \text{E}\left[\boldsymbol X | \boldsymbol \theta^* \right]$.
-In the case of near-degeneracy, we classify each model as near-degenerate or "viable" based on the distance of $\boldsymbol \mu(\boldsymbol \theta)$ from the boundary of the convex hull of $\mathcal{T}$ and look at the fraction of models that are "near-degenerate," meaning they are within a small distance $\epsilon > 0$ of the boundary of the convex hull.
+For each set of parameters we then calculate metrics of model impropriety introduced in Section \@ref(degeneracy-instability-and-uninterpretability) based on $\boldsymbol \mu(\boldsymbol \theta)$, $\text{ELPR}(\boldsymbol \theta)/\nv$, and the absolute coordinates of $\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] -   \text{E}\left[\boldsymbol X | \boldsymbol \theta^* \right]$.
+In the case of near-degeneracy, we classify each model as near-degenerate based on the distance of $\boldsymbol \mu(\boldsymbol \theta)$ from the boundary of the convex hull of $\mathcal{T}$ and look at the fraction of models that are "near-degenerate," meaning they are within a small distance $\epsilon > 0$ of the boundary of the convex hull.
 We define "small" through a rough estimation of the volume of the hull for each model size.
@@
 
-The results of this numerical study are summarized in Figures \ref{fig:degen_plots}, \ref{fig:instab_plots}, and \ref{fig:uninterp_plots}.
-From these three figures, it is clear that all three measures of model impropriety show higher values for larger magnitudes of the parameter vectors.
+The results of this numerical study are summarized in Figures \@ref(fig:degen-plots), \@ref(fig:instab-plots), and \@ref(fig:uninterp-plots).
+From these three figures, it is clear that all three measures of model impropriety show higher values for larger magnitudes of the parameter vectors, supporting the RBM model properties developed in Section \@ref(background).
+As a compounding issue, these figures show that, as model grow in size, it becomes easier for more parameter configurations to push RBM models into near-degeneracy, instability and uninterpretability.
 Additionally, since there are $\nh*\nv$ interaction terms in $\boldsymbol \theta$ versus only $\nh + \nv$ main effect terms, for large models there are many more interaction parameters than main effects in the models.
 And so, severely limiting the magnitude of the individual interactions may well help prevent model impropriety.
-
-```{r degen_plots, fig.cap="\\label{fig:degen_plots}Results from the numerical experiment, here looking at the fraction of models that were near-degenerate for each combination of magnitude of $\\boldsymbol \\theta$ and model size.
+\par
+
+```{r degen-plots, fig.cap="Results from the numerical experiment, here looking at the fraction of models that were near-degenerate for each combination of magnitude of $\\boldsymbol \\theta$ and model size, where $\\boldsymbol \\theta = (\\boldsymbol \\theta_{main}, \\theta_{interaction})$ is split into main and interaction parameters.
 Black lines show the contour levels for fraction of near-degeneracy, while the thick black line shows the level where the fraction of near-degenerate models is .05.", fig.height=3.5, fig.align="ht!"}
@@
   geom_contour(aes(x = r1, y = r2, z = max_abs_diff), colour = "black", bins = 8) +
-  #geom_contour(aes(x = r1, y = r2, z = mean_abs_diff), colour = "black", breaks = .05, size = 1.5) +
   geom_abline(aes(intercept = 0, slope = 1), alpha = .5, lty = 2) +
@@
   geom_contour(aes(x = r1, y = r2, z = scaled_mean_elpr), colour = "black", bins = 8) +
-  #geom_contour(aes(x = r1, y = r2, z = mean_max_q), colour = "black", breaks = .05, size = 1.5) +
   geom_abline(aes(intercept = 0, slope = 1), alpha = .5, lty = 2) +
@@
 
-```{r instab_plots_2, fig.cap="\\label{fig:instab_plots}Results from the experiment, here looking at the sample mean value of $\\text{ELPR}(\\boldsymbol \\theta)/V$ at each grid point for each combination of magnitude of $\\boldsymbol \\theta$ and model size.
+```{r instab-plots, fig.cap="The sample mean value of $\\text{ELPR}(\\boldsymbol \\theta)/V$ at each grid point for each combination of magnitude of $\\boldsymbol \\theta$ and model size.
 As the magnitude of $\\boldsymbol \\theta$ grows, so does the value of this metric, indicating typical instability in the model.", fig.height=3.5, fig.align="ht!"}
@@
 
-```{r uninterp_plots, fig.cap="\\label{fig:uninterp_plots}The sample mean of the maximum component of the absolute difference between the model expectation vector, E$\\left[\\boldsymbol X | \\boldsymbol \\theta\\right]$, and the expectation vector given independence, E$\\left[\\boldsymbol X | \\boldsymbol \\theta^* \\right ]$.
+```{r uninterp-plots, fig.cap="The sample mean of the maximum component of the absolute difference between the model expectation vector, E$\\left[\\boldsymbol X | \\boldsymbol \\theta\\right]$, and the expectation vector given independence, E$\\left[\\boldsymbol X | \\boldsymbol \\theta^* \\right ]$.
 Larger magnitudes of $\\boldsymbol \\theta$ correspond to larger differences, thus indicating reduced interpretability.", fig.height=3.5, fig.align="ht!"}
@@
 
-Figure \ref{fig:split_plots_dist} shows the fraction of near-degenerate models for each magnitude of $\boldsymbol \theta$ for each model architecture.
+Figure \@ref(fig:split-plots-dist) shows the fraction of near-degenerate models for each magnitude of $\boldsymbol \theta$ for each model architecture.
 For each number $\nv$ of visibles in the model, as the number $\nh$ of hiddens increase, the fraction near-degenerate diverges from zero at increasing rates for larger values of $||\boldsymbol \theta||$.
-This shows that as model size gets larger, the risk of degeneracy starts at a slightly larger magnitude of parameters, but very quickly increases until reaching close to 1.
-
-```{r split_plots2, warning=FALSE, fig.cap="\\label{fig:split_plots_dist}The fraction of near-degenerate models for each magnitude of $\\boldsymbol \\theta$.
-For each number $\nv$ of visibles in the model, the fraction near-degenerate becomes greater than zero at larger values of $||\\boldsymbol \\theta||$ as $\\nh$ increases and the slope becomes steeper as $\\nh$ increases as well.", fig.height=4, fig.align="ht!"}
+This shows that, as model size gets larger, the risk of degeneracy starts at a slightly larger magnitude of parameters, but very quickly increases until reaching close to 1.
+\par
+
+```{r split-plots-dist, warning=FALSE, fig.cap="The fraction of near-degenerate models for each magnitude of $\\boldsymbol \\theta$.
+For each number $\\nv$ of visibles in the model, the fraction near-degenerate moves away from zero at larger values of $||\\boldsymbol \\theta||$ as the number $\\nh$ of hidden variables increases and the slope becomes steeper as $\\nh$ increases as well.", fig.height=4, fig.align="ht!"}
 plot_dat_grid %>% 
@@
   geom_point(aes(dist, frac_degen, colour = Hiddens), size = .5) +
-  #geom_smooth(aes(dist, frac_degen, colour = Hiddens, group = Hiddens)) +
   facet_wrap(~Visibles, labeller = label_parsed) +
@@
 These manageable examples indicate that RBMs are near-degenerate, unstable, and uninterpretable for large portions of the parameter space with large $\|\boldsymbol \theta\|$.
-This, however, is not the only potential problem to be faced when using these models.
-There is the matter of principled/rigorous fitting of RBM models.
+These problematic aspects require serious consideration when using RBM models, on top of the additional matter of principled/rigorous fitting of RBM models.
 
@@
 
-Typically, fitting a RBM via maximum likelihood (ML) methods will be infeasible due mainly to the intractability of the normalizing term $\gamma(\boldsymbol \theta)$ in a model (\ref{eqn:pmf}) of any realistic size.
+Typically, fitting a RBM via maximum likelihood (ML) methods will be infeasible due mainly to the intractability of the normalizing term $\gamma(\boldsymbol \theta)$ in a model (\@ref(eq:pmf)) of any realistic size.
 Ad hoc methods are used instead, which aim to avoid this problem by using stochastic ML approximations that employ a small number of MCMC draws (i.e., contrastive divergence, [@hinton2002training]).
@@
 We provide a small example that illustrates that in fact there can be many such approximations.
-
-For simplicity, consider a model with two visible variables $(V_1,V_2)$ and one hidden $H_1$.
-In this case, there are four possible data realizations for $(V_1,V_2)$  given by $(\pm 1, \pm 1)$ and we may express the model probabilities
+For simplicity, consider a model with two visible variables $(V_1,V_2)$ and one hidden $H_1$ so that there are four possible data realizations for $(V_1,V_2)$  given by $(\pm 1, \pm 1)$ and we may express the model probabilities
 as
@@
 $$
-for $(v_1,v_2)\in \{-1, 1\}^2$, in terms of real-valued parameters $\theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21}$. Given any  specified cell probabilities, say
+for $(v_1,v_2)\in \{-1, 1\}^2$, in terms of real-valued parameters $\theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21}$. Given some specified cell probabilities, say
 \begin{equation}
-\label{eqn:0}
+(\#eq:0)
 0\leq p_{(-1,-1)},\;p_{(1,-1)}, \;p_{(-1,1)},\;p_{(1,1)},
 \end{equation} 
-for the outcomes $(\pm 1, \pm 1)$,  values for parameters $(\theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21})$ may be chosen to approximate such  cell probabilities with arbitrary closeness.
-In fact, when the cell probabilities (\ref{eqn:0}) are all strictly positive, parameters in the RBM model can be specified to reflect these probabilities exactly.
-And, when one or more of the cell probabilities (\ref{eqn:0}) are zero, the corresponding RBM  probabilities $P(V_1=v_1,V_2= v_2| \theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21})$ may never be identically zero (due to exponential terms in the model) but parameters can be still selected to make the appropriate RBM cell probabilities arbitrarily small.
-
-To demonstrate, we  assume $p_{(-1,-1)}>0$ (without loss of generality) in the specified cell probabilities (\ref{eqn:0}) and replace parameters $\theta_{11},\theta_{21}$ with $\Delta_1 \equiv \theta_{11} +\theta_{21}$ and $\Delta_2 \equiv \theta_{11} -\theta_{21}$.
-We may then prescribe values of $\theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \Delta_1,\Delta_2$  so that the model probability ratio
-$$
-P(V_1=v_1,V_2=v_2|\theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \Delta_1,\Delta_2)/  P(V_1=-1,V_2=-1| \theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \Delta_1,\Delta_2)
-$$
-matches the corresponding ratio $p_{(v_1,v_2)}/p_{(-1,-1)}$ over three values of $(v_1,v_2) = (1,-1),(-1,1),(1,1)$.
-
-For instance, assuming the cell probabilities from (\ref{eqn:0}) are all positive, these probabilities can be exactly reproduced by choosing
- \begin{eqnarray*}
- \theta_{v_1} &=& \frac{1}{2} \log\left( \frac{p_{(1,-1)}}{p_{(-1,-1)}} \frac{\exp(\theta_{h_1}-\Delta_1) + \exp(-\theta_{h_1}+\Delta_1)}{\exp(\theta_{h_1}+\Delta_2) + \exp(-\theta_{h_1}-\Delta_2)}  \right),\\
- \theta_{v_2}&= &\frac{1}{2} \log\left( \frac{p_{(-1,1)}}{p_{(-1,-1)}} \frac{\exp(\theta_{h_1}-\Delta_1) + \exp(-\theta_{h_1}+\Delta_1)}{\exp(\theta_{h_1}-\Delta_2) + \exp(-\theta_{h_1}+\Delta_2)}  \right)
-\end{eqnarray*}
-and selecting $\theta_{h_1}, \Delta_1,\Delta_2$ to solve
-\begin{equation}
-\label{eqn:1}
-\frac{p_{(1,1)}p_{(-1,-1)}}{p_{(-1,1)}p_{(1,-1)}} = \frac{\ell(|\theta_{h_1}|) + \ell(|\Delta_1|) }{\ell(|\theta_{h_1}|) + \ell(|\Delta_2|)},
-\end{equation}
-based on a monotonically increasing function $\ell(x)\equiv \exp(-2x)+\exp(2x)$, $x \geq 0$.
-If  $[p_{(1,1)}p_{(-1,-1)}]/ [p_{(-1,1)}p_{(1,-1)}] \geq 1$, one can pick any values for $\theta_{h_1}, \Delta_2\in \mathbb{R}$ and solve (\ref{eqn:1}) for $|\Delta_1|$; likewise, when $[p_{(1,1)}p_{(-1,-1)}]/ [p_{(-1,1)}p_{(1,-1)}] < 1$ in (\ref{eqn:1}), one may solve for $|\Delta_2|$ upon choosing any values for $\theta_{h_1}, \Delta_1\in \mathbb{R}$.
-
-Alternatively, if exactly one specified cell probability in (\ref{eqn:0}) is zero, say $p_{(1,1)}$ (without loss of generality), we can select parameters $\theta_{v_1},\theta_{v_2}$ as above based on a sequence $(\theta_{h_1}, \Delta_{1}, \Delta_{2}) \equiv (\theta_{h_1}^{(m)}, \Delta_{1}^{(m)}, \Delta_{2}^{(m)})$, $m\in\{1,2,\ldots,\}$ of the remaining parameter values such that $\lim_{m\to \infty}|\Delta_{1}^{(m)}| = \infty$ and $\lim_{m\to \infty}  (|\theta_{h_1}^{(m)}| + |\Delta_{2}^{(m)}|)/|\Delta_{1}^{(m)}|=0$ hold.
-This guarantees that the resulting RBM model matches the given cell probabilities (\ref{eqn:0}) in the limit:
-\begin{equation}
-\label{eqn:2}
-\lim_{m\to \infty}P(V_1=v_1,V_2=v_2| \theta_{v_1}, \theta_{v_2}, \theta_{h_1},\Delta_1,\Delta_2) = p_{(v_1,v_2)},\quad (v_1,v_2)\in\{(\pm 1,\pm 1)\}.
-\end{equation}  
-If exactly two specified probabilities in (\ref{eqn:0}) are zero, say  $p_{(1,1)}$ and $p_{(-1,1)}$ (without loss of generality), then a limit approximation as in (\ref{eqn:2}) follows by picking $\theta_{v_1}$ as above based on any choices of $(\theta_{h_1}, \Delta_1,\Delta_2)$ and choosing  a sequence of $\theta_{v_2}\equiv \theta_{v_2}^{(m)}$ values for which $\theta_{v_2}^{({m})} \rightarrow -\infty$.
-
-The previous discussion illustrates the fact that the RBM model class suffers from parameter identifiability issues that go beyond mere symmetries in the parametrization.
-Not only it is possible to approximate any distribution on the visibles arbitrarily well [cf.
-@montufar2011refinements], but quite different parameter settings can induce the same essential RBM model.
-However, this is not the most disastrous implication of the RBM parameterization.
-A far worse consequence is that, when fitting the RBM model by likelihood-based methods, we already know the nature of the answer before we begin: namely, such  fitting will simply aim to reproduce the empirical distribution from the training data if sufficiently many hiddens are in the model.
-That is, based on a random sample of vectors of visible variables, the model for the cell probabilities that has the highest likelihood over  *all possible model classes* (i.e., RBM-based or not) is the empirical distribution, and the over-parametrization of the RBM model itself ensures that this empirical distribution can be arbitrarily well-approximated.
-
-For illustration, continue the simple example from above with $n$ iid observations, each consisting of two realized visibles $(V_1,V_2)$.
-In which case, when the specified cell probabilities $p_{(-1,-1)},p_{(1,-1)}, p_{(-1,1)},p_{(1,1)}$ in (\ref{eqn:0}) are taken as the empirical cell frequencies from the sample, there is no better model based on maximum likelihood, and the discussion above (cf.
-(\ref{eqn:2})) shows that RBM model parameters can be chosen to re-create this empirical distribution to an arbitrarily close degree.
-Hence, RBM model fitting based on ML will simply seek to reproduce the empirical distribution.
-What's more, whenever this empirical distribution contains empty cells, fitting steps for the RBM model will necessarily aim to choose parameters that necessarily diverge to infinity in magnitude in order to zero-out the corresponding RBM cell probabilities.
+for the outcomes $(\pm 1, \pm 1)$, the appendix provides technical details to show parameter values $(\theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21})$ may be chosen to match any given cell probabilities with arbitrary closeness.
+In fact, when the cell probabilities (\@ref(eq:0)) are all strictly positive, parameters in the RBM model can be specified to reflect these probabilities exactly.
+And, when one or more of the cell probabilities (\@ref(eq:0)) are zero, the corresponding RBM probabilities $P(V_1=v_1,V_2=_v2| \theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21})$ may never be identically zero (due to exponential terms in the model) but parameters can be still selected to make the appropriate RBM cell probabilities arbitrarily small.
+Furthermore, as development in the appendix shows, only it is possible to approximate any distribution on the visibles arbitrarily well [cf.
+@montufar2011refinements], but dramatically different parameter settings can induce the same RBM model (beyond mere symmetries in parameterization).
+A further, and somewhat odd consequence of the RBM parameterization is then that, when fitting the RBM model by likelihood-based methods, we already know the nature of the answer before we begin: namely, such  fitting will simply reproduce the empirical distribution of the training data if sufficiently many hiddens are in the model and there can be no model refinements or smoothing from the RBM.
+That is, based on a random sample of vectors of visible variables, the model for the cell probabilities with the highest likelihood over  *all possible model classes* (i.e., RBM-based or not) is the empirical distribution, and the over-parameterization of the RBM model itself ensures that this empirical distribution can be arbitrarily well-approximated.
+For illustration, considering the simple example from above with $n$ iid observations, each consisting of two realized visibles $(V_1,V_2)$ the empirical cell frequencies from the sample are the highest likelihood choices for the cell probabilities $p_{(-1,-1)},p_{(1,-1)}, p_{(-1,1)},p_{(1,1)}$ in (\@ref(eq:0)) before any parametric model is introduced for improved analysis, and the discussion above indicates that RBM model parameters can be chosen to re-create this empirical distribution to an arbitrarily close degree.
+Not only does RBM model fitting based on ML seek to reproduce the empirical distribution, whenever this empirical distribution contains empty cells, fitting steps for the RBM model will further aim to choose parameters that necessarily diverge to infinity in magnitude in order to zero-out the corresponding RBM cell probabilities.
 In data applications with a large sample space, it is unlikely that the training set will include at least one of each possible vector outcome (unlike this small example).
-This implies that some RBM model parameters must diverge to $+\infty$ to mimic the empirical distribution with empty cells and, as we have already discussed in Section \ref{explorations-of-model-properties-through-simulation}, large magnitudes of $\boldsymbol \theta$ lead to model impropriety in the RBM.
+This implies that some RBM model parameters must diverge to $+\infty$ to mimic the empirical distribution with empty cells and, as we have already discussed in Section \@ref(explorations-of-model-properties-through-simulation), large magnitudes of $\boldsymbol \theta$ lead to model impropriety in the RBM.
 
@@
 ## Bayesian model fitting
-To avoid model impropriety for a fitted RBM, we want to avoid parts of the parameter space $\mathbb{R}^{\nv + \nh + \nv*\nh}$ that lead to near-degeneracy, instability, and uninterpretability.
-Motivated by the insights in Section \ref{exploring-manageable-examples}, one idea is to shrink $\boldsymbol \theta = (\boldsymbol \theta_{main}, \boldsymbol \theta_{interaction})$ toward $\boldsymbol 0$ by specifying priors that place low probability on large values of $||\boldsymbol \theta||$, specifically focusing on shrinking $\boldsymbol \theta_{interaction}$ more than $\boldsymbol \theta_{main}$.
+
+To avoid model impropriety for a fitted RBM, we wish to avoid parts of the parameter space $\mathbb{R}^{\nv + \nh + \nv*\nh}$ that lead to near-degeneracy, instability, and uninterpretability.
+Motivated by the insights in Section \@ref(exploring-manageable-examples), one idea is to shrink $\boldsymbol \theta = (\boldsymbol \theta_{main}, \boldsymbol \theta_{interaction})$ toward $\boldsymbol 0$ by specifying priors that place low probability on large values of $||\boldsymbol \theta||$, specifically focusing on shrinking $\boldsymbol \theta_{interaction}$ more than $\boldsymbol \theta_{main}$.
 This is similar to an idea advocated by @hinton2010practical called *weight decay*, in which a penalty is added to the interaction terms in the model, $\boldsymbol \theta_{interaction}$, shrinking their magnitudes.
-
-```{r param-table, results='asis'}
+\par
+
+```{r theta, results='asis'}
+load("../resources/data/params_theta.Rdata")
+
+params_good <- list(main_hidden = sample.params %>% ungroup() %>% filter(!near_hull) %>% select(starts_with("h"), -H) %>% data.matrix(),
+                    main_visible = sample.params %>% ungroup() %>% filter(!near_hull) %>% select(starts_with("v"), -V) %>% data.matrix(),
+                    interaction = sample.params %>% ungroup() %>% filter(!near_hull) %>% select(starts_with("theta")) %>% data.matrix() %>% matrix(4))
+
 old_pen <- options()$scipen
@@
          caption = "Parameters used to fit a test case with $V = H = 4$.
-This parameter vector was chosen as a sampled value of $\\boldsymbol \\theta$ that was not near the convex hull of the sufficient statistics for a grid point in Figure \\ref{fig:degen_plots} with $< 5$\\% near-degeneracy.", align = c("l", rep(c("r", "r"), times = 3))) %>%
+This parameter vector was chosen as a sampled value of $\\boldsymbol \\theta$ that was not near the convex hull of the sufficient statistics for a grid point in Figure \\ref{fig:degen-plots} with $< 5$\\% near-degeneracy.", align = c("l", rep(c("r", "r"), times = 3))) %>%
   print(sanitize.text.function=function(x){x}, include.rownames = FALSE, comment = FALSE)
-
-```
-
-We considered a test case with $\nv = \nh = 4$ and parameters given in Table \ref{tab:theta}.
-This parameter vector was chosen as a sampled value of $\boldsymbol \theta$ that was not near the convex hull of the space of values of sufficient statistics for a grid point in Figure \ref{fig:degen_plots} with $< 5$\% near-degeneracy.
+```
+
+We considered a test case with $\nv = \nh = 4$ and parameters given in Table \@ref(tab:theta).
+This parameter vector was chosen as a sampled value of $\boldsymbol \theta$ at which the resulting RBM model would not be clearly degenerate.
 We simulated $n = 5,000$ realizations of visibles as a training set and fit the RBM using three Bayes methodologies.
-These involved the following:
+These involved the following set-ups with choice of prior distribution $\pi(\boldsymbol \theta)$ for parameters $\boldsymbol \theta$.
 
 1.
-*A "trick" prior.* Here we cancel out normalizing term in the likelihood so that resulting full conditionals of $\boldsymbol \theta$ are multivariate normal.
-The $h_j$ are carried along in the MCMC sampling from the posterior as latent variables.
+*A "trick" prior.* Here we cancel out normalizing term in the likelihood (from $\gamma(\boldsymbol \theta)$ in (\@ref(eq:pmf))) so that resulting full conditionals of $\boldsymbol \theta$ are multivariate Normal.
+Namely, this involves a prior of the form
     \begin{align*}
@@
     \end{align*}
-
     where 
-
     \begin{align*}
-    \gamma(\boldsymbol \theta) = \sum\limits_{\boldsymbol x \in \mathcal{C}^{\nh + \nv}}\exp\left(\sum\limits_{i = 1}^\nv\sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right) \text{ and } C_{2} < C_{1}
+    \gamma(\boldsymbol \theta) = \sum\limits_{\boldsymbol x \in \mathcal{C}^{\nh + \nv}}\exp\left(\sum\limits_{i = 1}^\nv\sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)
     \end{align*}
-    This is the method of @li2014biclustering. We will refer to this method as Bayes with Trick Prior and Latent Variables (BwTPLV).
+    for hyperparameter choices $0< C_2 < C_1$.
+The unknown hidden variables $h_j$ are also directly treated as latent variables and are sampled in each MCMC iterative draw from the posterior distribution.
+This is the method of @li2014biclustering.
+We will refer to this method as Bayes with Trick Prior and Latent Variables (BwTPLV).
 2.
-*A truncated Normal prior.* Here we use independent spherical normal distributions as priors for $\boldsymbol \theta_{main}$ and $\boldsymbol \theta_{interaction}$, with $\sigma_{interaction} < \sigma_{main}$, *truncated* at $3\sigma_{main}$ and $3\sigma_{interaction}$, respectively.
+*A truncated Normal prior.* Here we use independent spherical normal distributions as priors for $\boldsymbol \theta_{main}$ and $\boldsymbol \theta_{interaction}$, which are *truncated* at $3\sigma_{main}$ and $3\sigma_{interaction}$, respectively, based on standard deviation hyperparameters $0<\sigma_{interaction}<\sigma_{main}$.
 Full conditional distributions are not conjugate, and simulation from the posterior was accomplished using a geometric adaptive Metropolis Hastings step [@zhou2014some] and calculation of likelihood normalizing constant.
-(This computation is barely feasible for a problem of this size and would be infeasible for larger problems.) Here the $h_j$ are carried along in the MCMC implementation as latent variables.
+(This computation is barely feasible for a problem of this size and would be infeasible for larger problems.) Here the hidden variables $h_j$ are again carried along in the MCMC implementation as latent variables.
 We will refer to this method as Bayes with Truncated Normal prior and Latent Variables (BwTNLV).
 3.
-*A truncated Normal prior and marginalized likelihood.* Here we marginalize out $\boldsymbol h$ in $f_{\boldsymbol \theta}(\boldsymbol x)$, and use the truncated Normal priors applied to the marginal probabilities for visible variables given by
+*A truncated Normal prior and marginalized likelihood.* Here we marginalize out the hidden variables $\boldsymbol h = (h_1, \dots, h_\nh)$ in $f_{\boldsymbol \theta}(\boldsymbol x)$, and use the truncated Normal priors applied to the marginal probabilities for visible variables given by
     \begin{align*}
@@
 
-The three fitting methods are ordered by computational feasibility in a real-data situation, with BwTPLV being the most computationally feasible due to conjugacy and BwTNML the least feasible due to the marginalization and need for an adaptive Metropolis Hastings step.
+The three fitting methods are ordered above according to  computational feasibility in a real-data situation, with BwTPLV being the most computationally feasible due to conjugacy and BwTNML the least feasible due to the marginalization and need for an adaptive Metropolis Hastings step.
 All three methods require choosing the values of hyperparameters.
 In each case, we have chosen these values based on a rule of thumb that shrinks $\boldsymbol \theta_{interaction}$ more than $\boldsymbol \theta_{main}$.
-Additionally, BwTPLV requires additional tuning to choose $C_1$ and $C_2$, reducing its appeal.
-The values used for the hyperparameters in our simulation are presented in Table \ref{tab:hyperparam}.
+Additionally, BwTPLV requires additional tuning (i.e.
+a tuning parameter $C > 0$ in Table \@ref(tab:hyperparam)) to choose $C_1$ and $C_2$, reducing its appeal.
+The forms used for the hyperparameters in our simulation are presented in Table \@ref(tab:hyperparam).
 
@@
 
-It should be noted that BwTNLV (2.) and BwTNML (3.) are drawing from the same stationary posterior distribution for vectors of visibles.
-The difference between these two methods is in how well the chains mix and how quickly they arrive at the target posterior distribution.
+It should be noted that, due to the common prior distributions for $\boldsymbol \theta$, both BwTNLV (method 2 above) and BwTNML (method 3) are drawing from the same stationary posterior distribution for vectors of visibles.
+A fundamental difference between these two methods lies in how well these two chains mix and how quickly they arrive at the target posterior distribution.
 After a burn-in period of 50 iterations selected by inspecting the trace plots, we assess the issue of mixing in two ways.
-First, the autocorrelation functions (ACF) from each posterior sample corresponding to a model probability for a visible vector outcome $\mathbf{v}=(v_1,v_2,v_3,v_4)\in\{\pm 1\}^4$ (i.e., computed from $\boldsymbol \theta$ under (\ref{eqn:pmf})) are assessed and plotted in Figure \ref{fig:acf} with BwTNLV in black and BwTNML in red.
-As expected, ACF corresponding to the method that marginalizes out the hidden variables from the likelihood decreases to zero at a much faster rate, indicating better mixing for the chain.
+First, the autocorrelation functions (ACF) from each posterior sample corresponding to a model probability for a visible vector outcome $\mathbf{v}=(v_1,v_2,v_3,v_4)\in\{\pm 1\}^4$ (i.e., computed from $\boldsymbol \theta$ under (\@ref(eq:pmf))) are determined and plotted in Figure \@ref(fig:acf) with BwTNLV in black and BwTNML in red.
+As expected, ACF corresponding to the method (BwTNML) that marginalizes out the hidden variables from the likelihood decreases to zero at a much faster rate, indicating better mixing for the chain.
+\par
 
 ```{r models-load}
-load("data/sample_images.Rdata")
+load("../resources/data/sample_images.Rdata")
 
@@
 
-```{r acf, fig.cap="\\label{fig:acf}The autocorrelation functions (ACF) for the posterior probabilityies of all $2^4 = 16$ possible outcomes for the vector of four visibles assessed at multiple lags for each method with BwTNLV in black and BwTNML in red.
+```{r acf, fig.cap="The autocorrelation functions (ACF) for the posterior probabilityies of all $2^4 = 16$ possible outcomes for the vector of four visibles assessed at multiple lags for each method with BwTNLV in black and BwTNML in red.
 As expected, ACF corresponding to the method that marginalizes out the hidden variables from the likelihood decreases to zero at a much faster rate, indicating better mixing for the chain."}
@@
 
-```{r M_eff, cache=TRUE}
+```{r m-eff-values, cache=TRUE}
 ###
@@
 
-Secondly, we can assess the mixing of our chains using an idea of effective sample size.
-If the MCMC chain were truly iid draws from the target distribution, then for the parameter $p^{(i)}$ denoting the probability of the $i$th vector outcome for the four visibles $\mathbf{v}=(v_1,v_2,v_3,v_4)\in\{\pm 1\}^4$, $i=1,\ldots,16$, its estimate as the average $\bar{p}^{(i)}$ of posterior sample versions would be approximately Normal with mean the true posterior marginal probability of $p^{(i)}$, and variance $\sigma^2_i/M$, where $\sigma^2_i$ is the true posterior variance of $p^{(i)}$ and $M$ is the length of the chain.
+Secondly, we can assess the mixing of the BwTNLV/BwTNML chains using the notion of effective sample size.
+If the MCMC chain were truly iid draws from the target distribution, then for the parameter $p^{(i)}$ denoting the probability of the $i$th vector outcome for the four visibles $\mathbf{v}=(v_1,v_2,v_3,v_4)\in\{\pm 1\}^4$, $i=1,\ldots,16$, its estimate as the average $\bar{p}^{(i)}$ of posterior sample versions would be approximately Normal with mean given by the posterior marginal mean of $p^{(i)}$, and variance given by $\sigma^2_i/M$, where $\sigma^2_i$ is the true posterior variance of $p^{(i)}$ and $M$ is the length of the chain.
 However, with the presence of correlation in our chain, the asymptotic variance of $\bar{p}^{(i)}$ is instead approximately some $C_i/M$, where $C_i$ is some positive constant such that $C_i > \sigma^2_i$.
@@
 Formally, we approximate the effective sample size of the length $M$ MCMC chain as
-
 $$
@@
 
-```{r m_eff_table, results='asis'}
+```{r m-eff, results='asis'}
 M_eff_table %>%
@@
 
-The effective sample sizes for a chain of length $M = 1000$ for inference about each of the $2^4 = 16$ model probabilities are presented in Table \ref{tab:m-eff}.
+For both BwTNLV and BwTNML methods, effective sample sizes for a chain of length $M = 1000$ for inference about each of the $2^4 = 16$ model probabilities are presented in Table \@ref(tab:m-eff).
 These range from  $`r round(min(M_eff_tmp[, "BwTNML"]), 2)`$ to $`r round(max(M_eff_tmp[, "BwTNML"]), 2)`$ for BwTNML, while BwTNLV only yields between $`r round(min(M_eff_tmp[, "BwTNLV"]), 2)`$ and $`r round(max(M_eff_tmp[, "BwTNLV"]), 2)`$ effective draws.
 Thus, BwTNLV would require at least $`r round(min(M_eff_tmp[, "BwTNML"])/min(M_eff_tmp[, "BwTNLV"]), 1)`$ times as many iterations to be run of the MCMC chain in order to achieve the same amount of effective information about the posterior distribution.
-For this reason, consistent with the ACF results in Figure \ref{fig:acf}, BwTNLV does not seem to be an effective method for fitting the RBM if computing resources are at all limited.
-
-Figure \ref{fig:fitting_plot} shows the posterior probability of each possible $\boldsymbol v \in \{-1,1\}^4$ after fitting the RBM model in the two ways detailed in this section (excluding BwTNLV).
+For this reason, consistent with the ACF results in Figure \@ref(fig:acf), BwTNLV does not seem to be an effective method for fitting the RBM, though computing resources can hinder use of the alternative BwTNML involving marginalization of hidden variables.
+
+Figure \@ref(fig:fitting-plot) shows the posterior probability of each possible $\boldsymbol v \in \{-1,1\}^4$ after fitting the RBM model according to method 1 (BwTPLV using trick prior) and method 3 (BwTNML) (excluding method 2 (BwTNLV) that seeks the same posterior as method 3) .
 The black vertical lines show the true probabilities of each image based on the parameters used to generate the training set while the red vertical lines show the empirical distribution for the training set of $5,000$ vectors.
 From these posterior predictive checks, it is evident that BwTNML produces the best fit to the data.
+Furthermore, along with the discussion of Section \@ref(model-fitting), Figure \@ref(fig:fitting-plot) also shows that it can be undesirable to seek to perfectly re-create an empirical distribution in fitting RBM models (i.e., true model probabilities may differ substantially).
+The priors in the BwTNML method constrain the RBM model fit to avoid replication of the empirical distribution and better estimate the underlying true data generating probabilities.
 However, this method requires a marginalization step to obtain the probability function of visible observations alone, which is infeasible for a model with $\nh$ of any real size.
+\par
 
@@
 
-```{r fitting_plot, fig.cap="\\label{fig:fitting_plot}Posterior probabilities of $16 = 2^4$ possible realizations of $4$ visibles using two of the three Bayesian fitting techniques, BwTPLV and BwTNML.
+```{r fitting-plot, fig.cap="Posterior probabilities of $16 = 2^4$ possible realizations of $4$ visibles using two of the three Bayesian fitting techniques, BwTPLV and BwTNML.
 The black vertical lines show the true probabilities of each vector of visibles based on the parameters used to generate the training data while the red vertical lines show the empirical distribution.
@@
 
-In this paper, we have presented three honest Bayes MCMC-based methods for fitting RBMs.
+In this paper, we have presented three fully Bayes-principled MCMC-based methods for fitting RBMs.
 Common practice is to use a kind of MCMC to overcome fitting complexities.
 However because of the size of the space to be filled with MCMC iterates, convergence and mixing of these methods will be slow.
-Marginalization over the latent variables in the model can improve mixing, but is numerically intractable due to the necessity of repeated calculation of the normalizing constant.
-Due to the extreme flexibility in this model class, rigorous likelihood-based fitting for a RBM will typically merely return the (discrete) empirical distribution for visibles, meaning any practitioner should be aware of this and employ some form of regularization or penalization in the model.
+Marginalization over the latent variables in the model is shown to improve mixing, but is numerically intractable due to the necessity of repeated calculation of the normalizing constant.
+Due to the extreme flexibility in this model class, rigorous likelihood-based fitting for a RBM will typically aim to merely merely return the (discrete) empirical distribution for visibles, meaning any practitioner should be aware of this and employ some form of regularization or penalization in the model.
 
 Ultimately, it is not clear that RBM models are useful as generative models.
-Without the generative behavior of the RBM model, the ability to quantify uncertainty in the estimated model parameters becomes impossible and it is no longer useful as a predictive distribution.
-In the case of classification this may not be the ultimate goal, however when S-instability is present, small (imperceptible) differences in the data can lead to greatly different probabilities and thus greatly different classifications.
-For these reasons, we are skeptical of the claims made about RBMs as generative tools.
-
-#References
+Furthermore, without the appropriate generative behavior in the RBM model, the ability to quantify uncertainty in the estimated model parameters becomes impossible and the model loses useful application to prediction problems (e.g., realistic predictive distributions).
+In the case of classification, predictive distributions not be the ultimate goal, but when S-instability is present, small (imperceptible) differences in the data may lead to greatly different probabilities and thus greatly different classifications.
+For these reasons, we are skeptical about RBMs as data generative tools.

### 2017-07-08 writing/paper-rbm/docs/paper.tex
paper to bookdown

\documentclass[12pt]{article}
\usepackage{amsmath,amssymb}
\usepackage{graphicx,psfrag,epsf}
\usepackage{enumerate}
\usepackage{natbib}
\usepackage{url} % not crucial - just used below for the URL

\usepackage{longtable,booktabs}
\usepackage[unicode=true]{hyperref}
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\usepackage{tikz, subfig, amsthm, multirow, float}
\usepackage{tikz-3dplot}
\usetikzlibrary{arrows, shapes, positioning}
\usepackage{bm}

%%INCLUDED-AK
\usepackage{amsthm}
\newtheorem{theorem}{Theorem}
\newtheorem{proposition}{Proposition}
\newtheorem{lemma}{Lemma}

\theoremstyle{definition}
\newtheorem{definition}{Definition}
\newtheorem{assumption}{Assumption}
\newtheorem{remark}{Remark}

\DeclareMathOperator*{\argmin}{arg\,min}
\DeclareMathOperator*{\argmax}{arg\,max}

\newcommand{\nv}{{n_{\scriptscriptstyle V}}}
\newcommand{\nh}{{n_{\scriptscriptstyle H}}}
\newcommand{\E}{E}

%\pdfminorversion=4
% NOTE: To produce blinded version, replace "0" with "1" below.
\newcommand{\blind}{0}

% DON'T change margins - should be 1 inch all around.
\addtolength{\oddsidemargin}{-.5in}%
\addtolength{\evensidemargin}{-.5in}%
\addtolength{\textwidth}{1in}%
\addtolength{\textheight}{-.3in}%
\addtolength{\topmargin}{-.8in}%

\let\BeginKnitrBlock\begin \let\EndKnitrBlock\end
\begin{document}

\def\spacingset#1{\renewcommand{\baselinestretch}%
{#1}\small\normalsize} \spacingset{1}

\title{\bf Properties and Bayes fitting of restricted Boltzmann machines}
\author{Andee Kaplan \\ Department of Statistics, Iowa State University \\  and \\ Daniel Nordman \\ Department of Statistics, Iowa State University \\  and \\ Stephen Vardeman \\ Department of Statistics and Department of Industrial and\\
Manufacturing Systems Engineering, Iowa State University \\ }

\date{}
\maketitle
\begin{abstract}
A restricted Boltzmann machine (RBM) is an undirected graphical model
constructed for discrete or continuous random variables, with two
layers, one hidden and one visible, and no conditional dependency within
a layer. In recent years, RBMs have risen to prominence due to their
connection to deep learning. By treating a hidden layer of one RBM as
the visible layer in a second RBM, a deep architecture can be created.
RBMs are thought to thereby have the ability to encode very complex and
rich structures in data, making them attractive for supervised learning.
However, the generative behavior of RBMs is largely unexplored. In this
paper, we discuss the relationship between RBM parameter specification
in the binary case and model properties such as degeneracy, instability
and uninterpretability. We also describe the difficulties that arise in
likelihood-based and Bayes fitting of such (highly flexible) models,
especially as Gibbs sampling (quasi-Bayes) methods are often advocated
for the RBM model structure.
\end{abstract}
\noindent%
{\it Keywords:}  Degeneracy, Instability, Classification, Deep Learning, Graphical Models
\vfill

\newpage
\spacingset{1.45}

\section{Introduction}\label{introduction}

The data mining and machine learning communities have recently shown
great interest in deep learning, specifically in stacked restricted
Boltzmann machines (RBMs) (see R. Salakhutdinov and Hinton
\protect\hyperlink{ref-salakhutdinov2009deep}{2009}; R. Salakhutdinov
and Hinton \protect\hyperlink{ref-salakhutdinov2012efficient}{2012};
Srivastava, Salakhutdinov, and Hinton
\protect\hyperlink{ref-srivastava2013modeling}{2013}; Le Roux and Bengio
\protect\hyperlink{ref-le2008representational}{2008} for examples). A
RBM is a probabilistic undirected graphical model (for discrete or
continuous random variables) with two layers, one hidden and one
visible, with no conditional dependency within a layer (Smolensky
\protect\hyperlink{ref-smolensky1986information}{1986}). These models
have reportedly been used with success in classification of images
(Larochelle and Bengio
\protect\hyperlink{ref-larochelle2008classification}{2008}; Srivastava
and Salakhutdinov
\protect\hyperlink{ref-srivastava2012multimodal}{2012}). However, the
model properties are largely unexplored in the literature and the
commonly cited fitting methodology remains heuristic-based and relies on
rough approximation (Hinton, Osindero, and Teh
\protect\hyperlink{ref-hinton2006fast}{2006}). In this paper, we provide
steps toward a fuller understanding of the model class and its
properties from the perspective of statistical model theory, and we then
explore the possibility of a rigorous fitting methodology. We find the
RBM model class to be concerning in two fundamental ways.

First, the models can be unsatisfactory as conceptualizations of how
data are generated. That is, recalling Fisher
(\protect\hyperlink{ref-fisher1922mathematical}{1922}), the aim of a
statistical model is to represent data in a compact way. Neyman and Box
further state that a model should ``provide an explanation of the
mechanism underlying the observed phenomena'' (Lehmann
\protect\hyperlink{ref-lehmann1990model}{1990}; G. E. P. Box
\protect\hyperlink{ref-box1967discrimination}{1967}). At issue,
simulation from RBMs can often produce data lacking realistic
variability so that such models may thereby fail to satisfactorily
reflect a true data generation process. Such behavior relates to model
degeneracy (or near degeneracy), which is a statistical concern in that
most data processes being modeled are realistically not degenerate in
their spectrum of potential outcomes. For example, when sampling data
from a nearly degenerate RBM, only a small number of output
possibilities have probability greater than zero, and thus a sample of
images will all be copies of the same one image (or small number of
images). An example of ten 4-pixel images simulated from a nearly
degenerate RBM model is compared to ten 4-pixel images simulated from a
non-degenerate RBM model in Figure \ref{fig:sample-models}. The
degenerate model places almost all probability on one outcome, causing
the image to be generated repeatedly, whereas the non-degenerate model
shows more potentially realistic variation. \par
\begin{figure}
\includegraphics[width=1\linewidth]{paper_files/figure-latex/sample-models-1} \includegraphics[width=1\linewidth]{paper_files/figure-latex/sample-models-2} \caption{Ten 4-pixel images simulated from a degenerate model (a) compared to ten 4-pixel images simulated from a non-degenerate model (b). The degenerate model places almost all probability on one outcome, causing the image to be generated repeatedly, whereas the non-degenerate model shows more realistic variation.}\label{fig:sample-models}
\end{figure}
In addition to such degeneracy, we find that RBMs can easily exhibit
types of instability, related to how sensitive the model probability
structure can be to small changes in data outcomes. In practice, this
may be seen when a single pixel change in an image results in a wildly
different classification in an image classification problem. Occurrences
of such behavior have recently been documented in RBMs (Li
\protect\hyperlink{ref-li2014biclustering}{2014}), as well as other deep
architectures (Szegedy et al.
\protect\hyperlink{ref-szegedy2013intriguing}{2013}; Nguyen, Yosinski,
and Clune \protect\hyperlink{ref-nguyen2014deep}{2014}). We describe
potential issues of model instability, degeneracy and uninterpretability
for the RBM class, which are related properties, and examine the
presence of these in Section
\ref{explorations-of-model-properties-through-simulation} through
simulations of small, manageable examples.

In addition to model properties, we also explore the quality of
estimation in fitting these models. The fitting can be problematic for
two reasons, the first being computational and the second concerning
flexibility. As the size of these models grows, both maximum likelihood
and Bayesian methods of fitting quickly become intractable. The
literature often suggests Markov chain Monte Carlo (MCMC) tools for
approximate maximization of likelihoods to fit these models (e.g., Gibbs
sampling to exploit conditional structure in hidden and visible
variables), but little is said about the attributes of realized
estimates (Hinton \protect\hyperlink{ref-hinton2010practical}{2010};
Hinton, Osindero, and Teh \protect\hyperlink{ref-hinton2006fast}{2006}).
Related to this, these MCMC algorithms require updating potentially many
latent variables (hiddens) which can critically influence convergence in
MCMC-based likelihood methods. Applying basic statistical principles in
fitting RBM models of tractable sizes, we compare three fully Bayesian
techniques involving MCMC, which are computationally more accessible
than direct maximum likelihood and also aim to avoid parts of a RBM
parameter space that yield unattractive models. As might be expected,
with greater computational complexity comes an increase in fitting
accuracy, but at the cost of practical feasibility.

While the computational concerns are inconvenient in fitting, issues due
to model flexibility are potentially more concerning. For a RBM model
with enough hidden variables, it has been shown that any distribution
for the visibles can be approximated arbitrarily well (Le Roux and
Bengio \protect\hyperlink{ref-le2008representational}{2008}; Montufar
and Ay \protect\hyperlink{ref-montufar2011refinements}{2011}; and
Montúfar, Rauh, and Ay
\protect\hyperlink{ref-montufar2011expressive}{2011}). However, for cell
data, the empirical distribution of an observed training set of visible
variables provides a highest likelihood benchmark, before any parametric
model class is even introduced and applied to obtain a refinement of
model fit. As a consequence, we find that any fully principled fitting
method based on the likelihood for a RBM with enough hidden variables
will seek to reproduce the (discrete) empirical distribution of a
training set. This aspect can be undesirable, and perhaps even
unexpected compared to most modeling scenarios, in that no ``smoothed
distribution'' may result when fitting a RBM model of sufficient size
with a rigorous likelihood-based method. We are therefore led to be
skeptical that models that involve these structures (like a deep
Boltzmann machine) can achieve useful prediction or inference in a
principled way without intentionally limiting the flexibility of the
fitted model.

The concepts of weight-decay (penalization) and sparsity
(regularization) are suggested in the literature as practical remedies
for both over-fitting and poor mixing in the Markov chain during fitting
(Hinton \protect\hyperlink{ref-hinton2010practical}{2010}; Tieleman
\protect\hyperlink{ref-tieleman2008training}{2008}; Cho, Ilin, and Raiko
\protect\hyperlink{ref-cho2012tikhonov}{2012}). Both \(L_1\) and \(L_2\)
type penalties are mentioned to achieve weight-decay however it is
unclear the benefits of these particular forms in different situations.
Additionally, it is not clear the degree to which regularization and
penalization are used by practitioners because they are not (by default)
a part of the accepted fitted methodology (Hinton
\protect\hyperlink{ref-hinton2002training}{2002}; Carreira-Perpinan and
Hinton \protect\hyperlink{ref-carreira2005contrastive}{2005}). In this
paper, we attempt to address the concerns with overfitting and poor
mixing via specification of a Bayesian prior in Section
\ref{bayesian-model-fitting}.

This paper is structured as follows. Section \ref{background} formally
defines the RBM including the joint distribution of hidden and visible
variables and explains the model's connection to deep learning.
Additionally, measures of model impropriety and methods of
quantifying/detecting it are defined. Section
\ref{explorations-of-model-properties-through-simulation} details our
explorations into the model behavior and potential propriety issues with
the RBM class. We examine three Bayesian fitting techniques intended to
avoid model impropriety in Section \ref{model-fitting} and conclude with
a discussion in Section \ref{discussion}. Appendix \ref{appendix-rbm}
provides proofs for results on RBM parameterizations and data codings
described in Section \ref{data-encoding}.

While applications of the RBM have claimed some successes in
classification problems, it is unclear if the model class allows one to
go beyond fitting to other statistical matters of importance in using
the models, such as quantification of the uncertainty of estimation and
the formulation of predictive distributions. These are closely tied to
the probability properties of the RBM model for explaining data
generation and it is hoped that the current exposition contributes to a
better understanding in this direction.

\section{Background}\label{background}

\subsection{Restricted Boltzmann
machines}\label{restricted-boltzmann-machines}

A restricted Boltzmann machine (RBM) is an undirected graphical model
specified for discrete or continuous random variables, binary variables
being most commonly considered. In this paper, we consider the binary
case for concreteness. A RBM architecture has two layers, hidden
(\(\mathcal{H}\)) and visible (\(\mathcal{V}\)), with no dependency
connections within a layer. An example of this structure is in Figure
\ref{fig:rbm} with the hidden nodes indicated by gray circles and the
visible nodes indicated by white circles.
\begin{figure}
  \centering
  \resizebox{\linewidth}{!}{\input{../resources/images/tikz/rbm.tikz}}
  \caption{An example restricted Boltzmann machine (RBM), consisting of two layers, one hidden ($\mathcal{H}$) and one visible ($\mathcal{V}$), with no connections within a layer. Hidden nodes are indicated by gray filled circles and the visible nodes indicated by unfilled circles.}
  \label{fig:rbm}
\end{figure}
A common use for RBMs is to create features for use in classification.
For example, binary images can be classified through a process that
treats the pixel values as the visible variables \(v_i\) in a RBM model
(Hinton, Osindero, and Teh
\protect\hyperlink{ref-hinton2006fast}{2006}).

\subsubsection{Joint distribution}\label{joint-distribution}

Let \(\boldsymbol x = (h_1, \dots, h_\nh, v_1,\dots,v_\nv)\) represent
the states of the visible and hidden nodes in a RBM for some integers
\(\nv, \nh \ge 1\). Each single binary random variable, visible or
hidden, will take its values in a common coding set \(\mathcal{C}\),
where we allow one of two possibilities for the coding set,
\(\mathcal{C}=\{0,1\}\) or \(\mathcal{C}=\{-1,1\}\), with ``\(1\)''
always indicating the ``high'' value of the variable. While
\(\mathcal{C}=\{0,1\}\) may be a natural starting point, we argue in
Section \ref{explorations-of-model-properties-through-simulation} that
the coding \(\mathcal{C}=\{-1,1\}\) induces more interpretable model
properties for the RBM. A standard parametric form for probabilities
corresponding to a potential vector of states,
\(X = (H_1, \dots, H_\nh, V_1,\dots,V_\nv)\), for the nodes is
\begin{align}
\label{eq:pmf}
f_{\boldsymbol \theta} (\boldsymbol x)\equiv P_{\boldsymbol \theta}(\boldsymbol X = \boldsymbol x) = \frac{\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}{\gamma(\boldsymbol \theta)}, \quad \boldsymbol x \in \mathcal{C}^{\nh + \nv} 
\end{align}
where
\(\boldsymbol \theta \equiv (\theta_{11}, \dots, \theta_{1\nh}, \dots, \theta_{\nv 1}, \dots, \theta_{\nv \nh}, \theta_{v_1}, \dots, \theta_{v_\nv}, \theta_{h_1}, \dots, \theta_{h_\nh}) \in \mathbb{R}^{\nv + \nh + \nv*\nh}\)
denotes the vector of model parameters and the denominator \[
\gamma(\boldsymbol \theta) = \sum_{\boldsymbol x \in \mathcal{C}^{\nh+\nv}}\exp\left(\sum_{i = 1}^\nv \sum_{j=1}^\nh \theta_{ij} v_i h_j + \sum_{i = 1}^\nv\theta_{v_i} v_i + \sum_{j = 1}^\nh\theta_{h_j} h_j\right)
\] is the normalizing function that ensures the probabilities
(\eqref{eq:pmf}) sum to one. For
\(\boldsymbol x = (h_1, \dots, h_\nh, v_1, \dots, v_\nh) \in \mathcal{C}^{\nv + \nh}\)
and
\begin{align}
\boldsymbol t(\boldsymbol x) &= (h_1, \dots, h_\nh, v_1, \dots, v_\nv, v_1 h_1, \dots, v_\nv h_\nh) \in \mathcal{C}^{\nh + \nv + \nh*\nv}, \label{eq:t}
\end{align}
let
\(\mathcal{T} = \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{C}^{\nv + \nh}\} \subset \mathbb{R}^{\nv + \nh + \nv * \nh}\)
be the set of possible values for the vector of variables needed to
compute probabilities (\eqref{eq:pmf}) in the model, and write
\(Q_{\boldsymbol \theta}(\boldsymbol x) = \sum\limits_{i = 1}^\nh\sum\limits_{j=1}^\nv \theta_{ij} h_i v_j + \sum\limits_{i = 1}^\nh\theta_{hi} h_i + \sum\limits_{j = 1}^\nv\theta_{vj} v_j\)
for the ``neg-potential'' function. The RBM model is parameterized by
\(\boldsymbol \theta\) containing two types of parameters, main effects
and interaction effects. The main effects parameters
(\(\{\theta_{v_i}, \theta_{h_j}\}_{\substack{i = 1, \dots, \nv,\\j = 1, \dots, \nh}}\))
weight the values of the visible \(v_i\) and hidden \(h_j\) nodes in
probabilities (\eqref{eq:pmf}) and the interaction effect parameters
(\(\theta_{ij}\)) weight the values of the connections \(v_i h_j\), or
dependencies, between hidden and visible layers.

Due to the potential size of the model, the normalizing constant
\(\gamma(\boldsymbol \theta)\) can be practically impossible to
calculate, making simple estimation of the model parameter vector
problematic. In model fitting, a kind of Gibbs sampling can be tried due
to the conditional architecture of the RBM (i.e.~visibles given hiddens
or vice verse). Specifically, the conditional independence of nodes in
each layer (given those nodes in the other layer) allows for stepwise
simulation of both hidden layers and model parameters (e.g., see the
contrastive divergence of Hinton
(\protect\hyperlink{ref-hinton2002training}{2002}) or Bayes methods in
Section \ref{model-fitting}).

\subsubsection{Connection to Deep
Learning}\label{connection-to-deep-learning}

RBMs have risen to prominence in recent years due to their connection to
deep learning (see Hinton, Osindero, and Teh
\protect\hyperlink{ref-hinton2006fast}{2006}; R. Salakhutdinov and
Hinton \protect\hyperlink{ref-salakhutdinov2012efficient}{2012};
Srivastava, Salakhutdinov, and Hinton
\protect\hyperlink{ref-srivastava2013modeling}{2013} for examples). By
stacking multiple layers of RBMs in a deep architecture, proponents of
the models claim to produce the ability to learn ``internal
representations that become increasingly complex, which is considered to
be a promising way of solving object and speech recognition problems''
(R. Salakhutdinov and Hinton
\protect\hyperlink{ref-salakhutdinov2009deep}{2009}, 450). The stacking
is achieved by treating a hidden layer of one RBM as the visible layer
in a second RBM, and so on, until the desired multi-layer architecture
is created.

\subsection{Degeneracy, instability, and
uninterpretability}\label{degeneracy-instability-and-uninterpretability}

The highly flexible nature of a RBM (having as it does
\(\nh + \nv + \nh*\nv\) parameters) creates at least three kinds of
potential issues in model impropriety that we will call
\emph{degeneracy}, \emph{instability}, and \emph{uninterpretability}. In
this section we define these characteristics, consider how to quantify
these in a RBM, and point out relationships among them.

\subsubsection{Near-degeneracy}\label{near-degeneracy}

In Random Graph Model theory, \emph{model degeneracy} means there is a
disproportionate amount of probability placed on only a few elements of
the sample space, \(\mathcal{X}\), by the model (Handcock
\protect\hyperlink{ref-handcock2003assessing}{2003}). For random graph
models, \(\mathcal{X}\) denotes all possible graphs that can be
constructed from a set of nodes and an exponentially parameterized
random graph model has a distribution of the form
\begin{align*}
f_{\boldsymbol \theta} (\boldsymbol x) = \frac{\exp\left(\boldsymbol \theta^T \boldsymbol t(\boldsymbol x)\right)}{\gamma(\boldsymbol \theta)}, \boldsymbol x \in \mathcal{X},
\end{align*}
where \(\boldsymbol \theta \in \Theta \subset \mathbb{R}^q\) is the
model parameter, and
\(\boldsymbol t: \mathcal{X} \rightarrow \mathbb{R}^q\) is a vector of
statistics based on the adjacency matrix of a graph. Here, as earlier,
\(\gamma(\boldsymbol \theta) = \sum_{\boldsymbol x \in \mathcal{X}} \exp\left(\boldsymbol \theta^T \boldsymbol t(\boldsymbol x)\right)\)
is the normalizing function. Let \(C\) denote the convex hull of the
potential outcomes of sufficient statistics,
\(\{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{X}\}\),
under the model above. Handcock
(\protect\hyperlink{ref-handcock2003assessing}{2003}) classifies an
exponentially parametrized random graph model at \(\boldsymbol \theta\)
as \emph{near-degenerate} if the mean value of the vector of sufficient
statistics under \(\boldsymbol \theta\),
\(\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta}\boldsymbol t( \boldsymbol X)\),
is close to the boundary of \(C\). Intuitively, if a model is
near-degenerate in the sense that only a small number of elements of the
sample space \(\mathcal{X}\) have positive probability, the expected
value \(\text{E}_{\boldsymbol \theta}\boldsymbol t( \boldsymbol X)\) is
an average of that same small number of values of \(t( \boldsymbol x)\)
(defining the boundary of the hull \(C\)) and can be expected to
\emph{not} be pulled deep into the interior of \(C\).

A RBM model can be thought to exhibit an analogous form of
\emph{near-degeneracy} when there is a disproportionate amount of
probability placed on a small number of elements in the sample space of
visibles and hiddens, \(\mathcal{C}^{\nv + \nh}\). Using the idea of
Handcock (\protect\hyperlink{ref-handcock2003assessing}{2003}), when the
random vector
\(\boldsymbol t(\boldsymbol x) = (v_1, \dots, v_\nv, h_1, \dots, h_\nh, v_1 h_1, \dots, v_V h_\nh ) \in \mathcal{T} \equiv \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{C}^{\nh + \nv} \}\)
from (\eqref{eq:t}), appearing in the neg-potential function
\(Q_{\boldsymbol \theta}(\cdot)\), has a mean vector
\(\boldsymbol \mu(\boldsymbol \theta) \in \mathbb{R}^{\nv + \nh + \nv * \nh}\)
close to the boundary of the convex hull of \(\mathcal{T}\), then the
RBM model can be said to exhibit near-degeneracy at
\(\boldsymbol \theta \in \mathbb{R}^{\nv + \nh + \nh*\nv}\). Here the
mean of \(\boldsymbol t(\boldsymbol x)\) is
\begin{align*}
\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta} \boldsymbol t(\boldsymbol X)  &= \sum\limits_{x \in \mathcal{C}^{\nv + \nh}} \left\{ \boldsymbol t(\boldsymbol x) f_{\boldsymbol \theta} (\boldsymbol x) \right\} \\
&= \sum\limits_{x \in \mathcal{C}^{\nv + \nh}} \left\{ \boldsymbol t(\boldsymbol x)\dfrac{\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{C}^{\nh + \nv}}\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}\right\}.
\end{align*}
\subsubsection{Instability}\label{instability}

Considering exponential families of distributions, Schweinberger
(\protect\hyperlink{ref-schweinberger2011instability}{2011}) introduced
a concept of model deficiency related to \emph{instability}. Instability
can be roughly thought of as excessive sensitivity in the model, where
small changes in the components of potential data outcomes,
\(\boldsymbol x\), may lead to substantial changes in the probability
function \(f_{\boldsymbol \theta}(\boldsymbol x)\). Furthermore, model
instability can be viewed on a spectrum of potential sensitivity in
probability structure, with model degeneracy included as a extreme or
limiting case of instability. To quantify ``instability'' more
rigorously (particularly beyond the definition given by Schweinberger
(\protect\hyperlink{ref-schweinberger2011instability}{2011})) it is
useful to consider how RBM models might be expanded to incorporate more
and more visibles. When increasing the size of RBM models, it becomes
necessary to grow the number of model parameters (and in this process
one may also arbitrarily expand the number of hidden variables used). To
this end, let
\(\boldsymbol \theta_{\nv} \equiv (\theta_{v_1}, \dots, \theta_{v_\nv}, \theta_{h_1}, \dots, \theta_{h_\nh}, \theta_{11}, \dots, \theta_{\nv \nh}), \nv \ge 1\),
denote an element of a sequence of RBM parameters indexed by the number
\(\nv\) of visibles \((V_1, \dots, V_\nv)\) and define a (scaled)
extremal log-probability ratio of the RBM model at
\(\boldsymbol \theta_{\nv}\) as
\begin{align}
\frac{1}{\nv} \log \left[\frac{\max\limits_{(v_1, \dots, v_\nv) \in \mathcal{C}^\nv}P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}{\min\limits_{(v_1, \dots, v_\nv) \in \mathcal{C}^\nv}P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}\right] \equiv \frac{1}{\nv} \text{ELPR}(\boldsymbol \theta_\nv) \label{eq:elpr}
\end{align}
where
\(P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv) \propto \sum_{(h_1, \dots, h_\nh) \in \mathcal{C}^\nh}\exp\left(\sum_{i = 1}^\nv \theta_{v_i} v_i + \sum_{j = 1}^\nh \theta_{h_j} h_j + \sum_{i = 1}^\nv \sum_{j = 1}^\nh \theta_{ij} v_i h_j\right)\)
is the RBM probability of observing outcome \((v_1, \dots, v_\nv)\) for
the visible variables \((V_1, \dots, V_\nv)\) under parameter vector
\(\boldsymbol \theta_\nv\), after marginalization of hidden variables.

In formulating a RBM model for a potentially large number of visibles
(i.e., as \(\nv \rightarrow \infty\)), we will say that the ratio
(\eqref{eq:elpr}) needs to stay bounded for the sequence of RBM models to
be stable. That is, we make the following convention.
\BeginKnitrBlock{definition}[S-unstable RBM]
\protect\hypertarget{def:unnamed-chunk-1}{}{\label{def:unnamed-chunk-1}
\iffalse (S-unstable RBM) \fi{} }Let
\(\boldsymbol \theta_\nv \in \mathbb{R}^{\nv + \nh + \nh*\nv}, \nv \ge 1\),
be an element of a sequence of RBM parameters where the number of
hiddens, \(\nh \equiv \nh(\nv) \ge 1\), can be an arbitrary function of
the number \(\nv\) of visibles. A RBM model formulation is
\emph{Schweinberger-unstable} or \emph{S-unstable} if
\begin{align*}
\lim\limits_{\nv \rightarrow \infty} \frac{1}{\nv} \text{ELPR}(\boldsymbol \theta_\nv) = \infty.
\end{align*}
\EndKnitrBlock{definition} In other words, the RBM model sequence is
unstable if, given any \(c > 0\), there exists an integer \(n_c > 0\) so
that \(\frac{1}{\nv}\text{ELPR}(\boldsymbol \theta_\nv) > c\) for all
\(\nv \ge n_c\). This definition of \emph{S-unstable} is a
generalization or re-interpretation of the ``unstable'' concept of
Schweinberger
(\protect\hyperlink{ref-schweinberger2011instability}{2011}) in that
here RBM models for visibles \((v_1, \dots, v_\nv)\) do not form an
exponential family and the dimensionality of \(\boldsymbol \theta_\nv\)
is not fixed, but rather grows with \(\nv\).

S-unstable RBM model sequences are undesirable for several reasons. One
is that, as mentioned above, small changes in data can lead to
overly-sensitive changes in probability under the data model. Consider,
for example, \[
\Delta(\boldsymbol \theta_\nv) \equiv \max \left\{\log \frac{P_{\boldsymbol \theta_\nv}(\boldsymbol v)}{P_{\boldsymbol \theta_\nv}(\boldsymbol v^*)} : \boldsymbol v \text{ }\& \text{ } \boldsymbol v^* \in \mathcal{C}^\nv \text{ differ in exactly one component}\right\},
\] denoting the biggest log-probability ratio for a one component change
in data outcomes (visibles) at a RBM parameter
\(\boldsymbol \theta_\nv\). We then have the following result.

\BeginKnitrBlock{proposition}
\protect\hypertarget{prp:instab1}{}{\label{prp:instab1}}Let \(c > 0\) and
let \(\text{ELPR}(\boldsymbol \theta_\nv)\) be as in (\eqref{eq:elpr}) for
an integer \(\nv \ge 1\). If
\(\frac{1}{\nv}\text{ELPR}(\boldsymbol \theta_\nv) > c\), then
\(\Delta(\boldsymbol \theta_\nv) > c\).
\EndKnitrBlock{proposition}

In other words, if the probability ratio (\eqref{eq:elpr}) is too large,
then a RBM model sequence will exhibit large probability shifts for very
small changes in data configurations (i.e., will exhibit instability).
Such instability can be a concern in a model for the similar reasons to
degeneracy: as outcomes vary over the sample space, the geography of
probabilities is extremely rugged, with deep pits following sharp
mountains. Recall the applied example of RBM models as a means to
classify images. For data as pixels in an image, the instability result
in Proposition \ref{prp:instab1} manifests itself as a one pixel change
in an image (one component of the visible vector) resulting in a large
shift in the probability, which in turn could result in a vastly
different classification of the image. Examples of this behavior have
been presented in Szegedy et al.
(\protect\hyperlink{ref-szegedy2013intriguing}{2013}) for other deep
learning models, in which a one pixel change in a test image results in
a wildly different classification.

Additionally, S-unstable RBM model sequences may be formally connected
to the near-degeneracy of Section \ref{near-degeneracy} (in which model
sequences place all probability on a small portion of their sample
spaces). To see this, define an arbitrary modal set of possible outcomes
(i.e.~set of highest probability outcomes) in RBM models with parameters
\(\boldsymbol \theta_\nv, \nv \ge 1\) as \[
M_{\epsilon, \boldsymbol \theta_\nv} \equiv \left\{\boldsymbol v \in \mathcal{C}^\nv: \log P_{\boldsymbol \theta_\nv}(\boldsymbol v) > (1-\epsilon)\max\limits_{\boldsymbol v^*}P_{\boldsymbol \theta_\nv}(\boldsymbol v^*) + \epsilon\min\limits_{\boldsymbol v^*}P_{\boldsymbol \theta_\nv}(\boldsymbol v^*) \right\}
\] for a given \(0 < \epsilon < 1\). Then S-unstable model sequences are
guaranteed to be degenerate, as the following result shows.

\BeginKnitrBlock{proposition}
\protect\hypertarget{prp:degen}{}{\label{prp:degen}}For an S-unstable RBM
model sequence and any \(0 < \epsilon < 1\), \[
P_{\boldsymbol \theta_\nv}\left((v_1, \dots, v_\nv) \in M_{\epsilon, \boldsymbol \theta_\nv}\right) \rightarrow 1 \text{ as } \nv \rightarrow \infty.
\]
\EndKnitrBlock{proposition}

In other words, S-unstable RBM model sequences are guaranteed to stack
up all probability on a specific set of outcomes for visibles, which
could potentially be arbitrarily narrow. Proofs of Propositions
\ref{prp:instab1} and \ref{prp:degen} appear in Kaplan, Nordman, and
Vardeman (\protect\hyperlink{ref-kaplan2017note}{2017}). These findings
also have counterparts in results in Schweinberger
(\protect\hyperlink{ref-schweinberger2011instability}{2011}), but unlike
results there, we do not limit consideration to exponential family forms
with a fixed number of parameters.

\subsubsection{Uninterpretability}\label{uninterpretability}

For spatial Markov models, Kaiser
(\protect\hyperlink{ref-kaiser2007statistical}{2007}) defines a measure
of model impropriety he calls \emph{uninterpretability}, which is
characterized by dependence parameters in a model being so extreme that
marginal mean-structures fail to hold as anticipated by a model
statement. We adapt this notion to RBM models as follows. Note that in a
RBM, the parameters \(\theta_{v_1}, \dots, \theta_{v_\nv}\) and
\(\theta_{h_1}, \dots, \theta_{h_\nh}\) are naturally associated with
main effects of visible and hidden variables and can be interpreted as
(logit functions of) means for variables
\(V_1, \dots, V_\nv, H_1, \dots, H_\nh\) in a model with no interaction
parameters, \(\theta_{ij} = 0, i = 1, \dots, \nv, j = 1, \dots, \nh\).
That is, with no interaction parameters, we have from (\eqref{eq:pmf})
that \[
P_\theta(V_i=1) \propto e^{\theta_{v_i}} \quad \text{ and } \quad P_\theta(H_j=1) \propto e^{\theta_{h_j}}, \quad  i=1,\ldots,\nv,j=1,\ldots,\nh
\] so that, for example, \(\text{logit}(P_\theta(V_i=1))=\theta_{v_i}\)
(or \(2 \theta_{v_i}\)) under the coding \(\mathcal{C}=\{0,1\}\) (or
\(\{-1,1\}\)). Hence, these main effect parameters have a clear
interpretation under an independence model (one with
\(\theta_{ij} = 0\)) but this interpretation can break down as
interaction parameters increase in magnitude relative to the size of the
main effects. In such cases, the main effect parameters \(\theta_{v_1}\)
and \(\theta_{h_j}\) are no longer interpretable in the models
(statements of marginal means) and the dependence parameters are so
large as to dominate the entire model probability structure (also
destroying the model interpretation of dependence as local conditional
modifications of an overall marginal mean structure). Whether or not
parameter interpretation is itself a goal in a given application of RBM
models, this concept of interpretation can provide an additional device
for examining other aspects of model propriety related to instability
and degeneracy. As explained by Kaiser
(\protect\hyperlink{ref-kaiser2007statistical}{2007}), models with
interpretable dependence parameters typically correspond to
non-degenerate models, while degradation in interpretability is often
associated with model drift into degeneracy. To assess which parameter
values \(\boldsymbol \theta\) may cause difficulties in interpretation,
we use the difference
\(\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] - \text{E}\left[\boldsymbol X | \boldsymbol \theta^* \right]\)
between two model expectations:
E\(\left[\boldsymbol X | \boldsymbol \theta\right]\) at
\(\boldsymbol \theta\) and expectations
E\(\left[\boldsymbol X | \boldsymbol \theta^* \right]\) where
\(\boldsymbol \theta^*\) matches \(\boldsymbol \theta\) for all main
effects but otherwise has \(\theta_{ij} = 0\) for
\(i = 1, \dots, \nv, j = 1, \dots, \nh\). Hence,
\(\boldsymbol \theta^*\) and \(\boldsymbol \theta\) have the same main
effects but \(\boldsymbol \theta^*\) has \(0\) dependence parameters.
Uninterpretability is then avoided at a parametric specification
\(\boldsymbol \theta\) if the model expected value at
\(\boldsymbol \theta\) is not very different from the corresponding
model expectation under independence. Using this, it is possible to
investigate what parametric conditions lead to uninterpretability in a
model versus those that guarantee interpretable models. If
\(\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] - \text{E}\left[\boldsymbol X | \boldsymbol \theta^*\right]\)
is large, then the RBM model with parameter vector
\(\boldsymbol \theta\) is said to be uninterpetable. The quantities to
compare in the RBM case are \[
\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] = \sum\limits_{\boldsymbol x \in \mathcal{C}^{\nv + \nh}} \boldsymbol x f_{\boldsymbol \theta}(\boldsymbol x) = \sum\limits_{\boldsymbol x \in \mathcal{C}^{\nv + \nh}} \boldsymbol x \frac{\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{C}^{\nv + \nh}}\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}
\] and
\begin{align*}
\text{E}\left[\boldsymbol X | \boldsymbol 
\theta^*\right] &= \sum\limits_{\boldsymbol x \in \mathcal{C}^{\nv + \nh}} \boldsymbol x \frac{\exp\left(\sum\limits_{i = 1}^\nv \theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{C}^{\nv + \nh}}\exp\left(\sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)} \\
\end{align*}
\section{Explorations of model properties through
simulation}\label{explorations-of-model-properties-through-simulation}

We next explore and numerically explain the relationship between values
of \(\boldsymbol \theta\) and the three notions of model impropriety,
near-degeneracy, instability, and uninterpretability, for RBM models of
varying sizes.

\subsection{Tiny example}\label{tiny-example}

To illustrate the ideas of model near-degeneracy, instability, and
uninterpretability in a RBM, we consider first the smallest possible
(toy) example that consists of one visible node \(v_1\) and one hidden
node \(h_1\) that are both binary. A schematic of this model can be
found in Figure \ref{fig:toymodel}. Because it seems most common, we
shall begin by employing \(0/1\) encoding of binary variables (both
\(h_1\) and \(v_1\) taking values in \(\mathcal{C} = \{0,1\}\)).
(Eventually we shall argue in Section \ref{data-encoding} that \(-1/1\)
coding has advantages.)
\begin{figure}[ht]
  \centering
  \resizebox{1cm}{!}{\input{../resources/images/tikz/toymodel.tikz}}
  \caption{A small example restricted Boltzmann machine (RBM), with two nodes, one hidden and one visible.}
  \label{fig:toymodel}
\end{figure}
\subsubsection{Impropriety three ways}\label{impropriety-three-ways}

For this small model, we are able to investigate the symptoms of model
impropriety, beginning with near-degeneracy. To this end, recall from
Section \ref{near-degeneracy} that one characterization requires
consideration of the convex hull of possible values of statistics
\(\boldsymbol t(\boldsymbol x)\), \[
\mathcal{T} = \{\boldsymbol t(\boldsymbol x): \boldsymbol x = (v_1, h_1) \in \{0,1\}^2\} \equiv \{(v_1, h_1, v_1 h_1): v_1, h_1 \in \{0,1\}\}
\] appearing in the RBM probabilities for this model. As this set is in
three dimensions, we are able to explicitly illustrate the shape of
boundary of the convex hull of \(\mathcal{T}\) and explore the behavior
of the mean vector
\(\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta} \boldsymbol t(\boldsymbol x)\)
as a function of the parameter vector \(\boldsymbol \theta\). Figure
\ref{fig:toyhull} shows the convex hull of our ``statistic space,''
\(\mathcal{T} \subset \{0,1\}^3\), for this toy problem from two
perspectives (enclosed by the unit cube \([0,1]^3\), the convex hull of
\(\{0,1\}^3\)). In this small model, note that the convex hull of
\(\mathcal{T}\) does not fill the unrestricted hull of \(\{0,1\}^3\)
because of the relationship between the elements of
\(\mathcal{T} = \{(v_1, h_1, v_1 h_1 : v_1, h_1 \in \{0, 1\} \}\) (i.e.
\(v_1 h_1 = 1\) only if \(v_1 = h_1 = 1\)). \par
\begin{figure}[ht]
  \begin{minipage}{0.45\textwidth}
  \resizebox{\linewidth}{!}{
    \tdplotsetmaincoords{60}{-60}
    \input{../resources/images/tikz/toyhull_top.tikz}

    \draw(0,1,0) -- (1,0,0) -- (0,0,0) -- (0,1,0); 
     \draw(1,1,1) -- (1,0,0) -- (0,0,0) -- (1,1,1); 
     \draw(1,1,1) -- (0,1,0) -- (0,0,0) -- (1,1,1); 
     \draw(1,1,1) -- (0,1,0) -- (1,0,0) -- (1,1,1); 
    \end{tikzpicture}
  }
  \end{minipage}
  \begin{minipage}{0.45\textwidth}
  \resizebox{\linewidth}{!}{
    \tdplotsetmaincoords{60}{120}
    \input{../resources/images/tikz/toyhull_top.tikz}

    \draw(0,1,0) -- (1,0,0) -- (0,0,0) -- (0,1,0); 
     \draw(1,1,1) -- (1,0,0) -- (0,0,0) -- (1,1,1); 
     \draw(1,1,1) -- (0,1,0) -- (0,0,0) -- (1,1,1); 
     \draw(1,1,1) -- (0,1,0) -- (1,0,0) -- (1,1,1); 
    \end{tikzpicture}
  }
  \end{minipage}
  \caption{Two perspectives of the convex hull of the ``statistic space" in three dimensions for the toy RBM with one visible and one hidden node.}
\label{fig:toyhull}
\end{figure}
We can compute the mean vector for \(\boldsymbol t(\boldsymbol x)\) as a
function of the model parameters as
\begin{align*}
\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta}\left[ \boldsymbol t(\boldsymbol X) \right] =  \sum\limits_{\boldsymbol x = (v_1, h_1) \in \{0,1\}^2} \left\{ t(x) \frac{\exp\left( \theta_{11} h_1 v_1 + \theta_{h1} h_1 + \theta_{v1} v_1\right)}{\gamma(\boldsymbol \theta)} \right\} = \left[
\begin{matrix}
\frac{\exp\left(\theta_{v_1}\right) + \exp\left(\theta_{11} + \theta_{v_1} + \theta_{h_1}\right)}{\gamma(\boldsymbol \theta)} \\
\frac{\exp\left(\theta_{h_1}\right) + \exp\left(\theta_{11} + \theta_{v_1} + \theta_{h_1}\right)}{\gamma(\boldsymbol \theta)} \\
\frac{\exp\left(\theta_{11} + \theta_{v_1} + \theta_{h_1}\right)}{\gamma(\boldsymbol \theta)} \\
\end{matrix} \right]
\end{align*}
where
\(\gamma(\boldsymbol \theta) = \sum\limits_{h_1 = 0}^1\sum\limits_{v_1 = 0}^1 \exp(\theta_{11} h_1 v_1 + \theta_{h_1}h_1 + \theta_{v_1}v_1)\).
The three parametric coordinate functions of
\(\boldsymbol \mu(\boldsymbol \theta)\) can be represented as in Figure
\ref{fig:degen-toy}. (Contour plots for three coordinate functions are
shown in columns for various values of \(\theta_{11}\), which can be
interpreted here as an absolute log-odds ratio as the visible changes
between \(0\) and \(1\).) In examining these, we see that as coordinates
of \(\boldsymbol \theta\) grow larger in magnitude, at least one mean
function for the entries of \(\boldsymbol t(\boldsymbol x)\) approaches
a value 0 or 1, forcing
\(\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta} \boldsymbol t(\boldsymbol x)\)
to be near to the boundary of the convex hull of \(\mathcal{T}\), as a
sign of model near-degeneracy. Thus, for a very small example we can see
the relationship between values of \(\boldsymbol \theta\) and model
degeneracy. \par
\begin{figure}

{\centering \includegraphics{paper_files/figure-latex/degen-toy-1}

}

\caption{Contour plots for the three parametric mean functions of sufficient statistics for a RBM with one visible and one hidden node.}\label{fig:degen-toy}
\end{figure}
Secondly, we can look at \(\text{ELPR}(\boldsymbol \theta)\) from
(\eqref{eq:elpr}) for this tiny model in order to consider model
instability as a function of RBM parameters. Recall that large values of
\(\text{ELPR}(\boldsymbol \theta)\) are associated with an extreme
sensitivity of the model probabilities
\(f_{\boldsymbol \theta}(\boldsymbol x)\) to small changes in
\(\boldsymbol x\) (see Proposition \ref{prp:instab1}). The quantity
\(\text{ELPR}(\boldsymbol \theta)\) for this small RBM is
\begin{align*}
\text{ELPR}(\boldsymbol \theta) &= \log \left[\frac{\max\limits_{(v_1, \dots, v_\nv) \in \mathcal{C}^\nv}P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}{\min\limits_{(v_1, \dots, v_\nv) \in \mathcal{C}^\nv}P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}\right] = \log \left[\frac{\max\limits_{v_1 \in \mathcal{C}}\sum\limits_{h_1 \in \mathcal{C}}\exp\left\{\theta_{11} h_1 v_1 + \theta_{h_1}h_1 + \theta_{v_1} v_1 \right\}}{\min\limits_{v_1 \in \mathcal{C}}\sum\limits_{h_1 \in \mathcal{C}}\exp\left\{\theta_{11} h_1 v_1 + \theta_{h_1}h_1 + \theta_{v_1} v_1 \right\}}\right]. \\
\end{align*}
Figure \ref{fig:instab} shows contour plots of
\(\text{ELPR}(\boldsymbol \theta)/\nv\) for various values of
\(\boldsymbol \theta\) in this model with \(\nv = 1\). We can see that
this quantity is large for large magnitudes of \(\boldsymbol \theta\),
especially for large values of the dependence/interaction parameter
\(\theta_{11}\). This suggests instability as \(|\boldsymbol \theta|\)
becomes large, agreeing also with the concerns about near-degeneracy
produced by consideration of \(\boldsymbol \mu(\boldsymbol \theta)\).
\par
\begin{figure}

{\centering \includegraphics{paper_files/figure-latex/instab-1}

}

\caption{$\text{ELPR}(\boldsymbol \theta)/\nv$ for various values of $\boldsymbol \theta$ for the tiny example model. Recall here $\nv$ is the number of visible nodes and here is $1$. This quantity is large for large magnitudes of $\boldsymbol \theta$.}\label{fig:instab}
\end{figure}
Finally to consider the effect of \(\boldsymbol \theta\) on potential
model uninterpretability, we can look at the difference between model
expectations, E\(\left[\boldsymbol X | \boldsymbol \theta\right]\), and
expectations given independence,
E\(\left[\boldsymbol X | \boldsymbol \theta^*\right ]\) for the tiny toy
RBM model where \(\boldsymbol X = (V_1, H_1, V_1 H_1)\). This difference
is given by
\begin{align*}
\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] -   \text{E}\left[\boldsymbol X | \boldsymbol \theta^* \right] 
&= \left[
\begin{matrix}
\frac{\exp\left(\theta_{11} + \theta_{v_1} + 2\theta_{h_1}\right) - \exp\left( \theta_{v_1} + 2\theta_{h_1}\right) }{\left(\exp\left(\theta_{v_1}\right) + \exp\left(\theta_{h_1}\right) + \exp\left(\theta_{11} + \theta_{v_1} + \theta_{h_1}\right)\right)\left(\exp\left(\theta_{v_1}\right) + \exp\left(\theta_{h_1}\right) + \exp\left(+ \theta_{v_1} + \theta_{h_1}\right)\right)} \\
\frac{\exp\left(\theta_{11} + 2\theta_{v_1} + \theta_{h_1}\right) - \exp\left( 2\theta_{v_1} + \theta_{h_1}\right)  }{\left(\exp\left(\theta_{v_1}\right) + \exp\left(\theta_{h_1}\right) + \exp\left(\theta_{11} + \theta_{v_1} + \theta_{h_1}\right)\right)\left(\exp\left(\theta_{v_1}\right) + \exp\left(\theta_{h_1}\right) + \exp\left(+ \theta_{v_1} + \theta_{h_1}\right)\right)}
\end{matrix}\right].
\end{align*}
Again, we can inspect these coordinate functions of this vector
difference to look for a relationship between parameter values and large
values of
\(\text{E}[\boldsymbol X|\boldsymbol \theta] - \text{E}[\boldsymbol X| \boldsymbol \theta^*]\)
as a signal of uninterpretability for the toy RBM. \par
\begin{figure}

{\centering \includegraphics{paper_files/figure-latex/uninterp-1}

}

\caption{The absolute difference between coordinates of model expectations, E$\left[\boldsymbol X | \boldsymbol \theta\right]$, and expectations given independence, E$\left[\boldsymbol X | \boldsymbol \theta^* \right ]$ for a RBM with one visible and one hidden node. As an indicator of uninterpretability, note that differences in expectations increase as the dependence parameter $\theta_{11}$ deviates from zero.}\label{fig:uninterp}
\end{figure}
Figure \ref{fig:uninterp} shows that the absolute difference between
coordinates of the vector of model expectations,
E\(\left[\boldsymbol X | \boldsymbol \theta\right]\) and corresponding
expectations E\(\left[\boldsymbol X | \boldsymbol \theta^*\right ]\)
given independence grow for the toy RBM as the values of
\(\boldsymbol \theta\) are farther from zero, especially for large
magnitudes of the dependence parameter \(\theta_{11}\). This is a third
indication that parameter vectors of large magnitude lead to model
impropriety in a RBM.

\subsubsection{Data encoding}\label{data-encoding}

Multiple encodings of the binary variables are possible. For example, we
could allow hiddens \((H_1, \dots, H_\nh) \in \{0,1\}^\nh\) and visibles
\((V_1, \dots, V_\nv) \in \{0,1\}^\nv\), as in the previous sections or
we could instead encode the state of the variables as \(\{-1,1\}^\nh\)
and \(\{-1,1\}^\nv\). This will result in variables
\(\boldsymbol t(\boldsymbol X)\) from (\eqref{eq:t}) satisfying
\(\boldsymbol t(\boldsymbol x) \in \{0,1\}^{\nh + \nv + \nh*\nv}\) or
\(\boldsymbol t(\boldsymbol x) \in \{-1,1\}^{\nh + \nv + \nh*\nv}\)
depending on how we encode ``on'' and ``off'' states in the nodes.

The \(-1/1\) data encoding has the benefit of providing a
guaranteed-to-be non-degenerate model at
\(\boldsymbol \theta = \boldsymbol 0 \in \mathbb{R}^{\nh + \nv + \nh*\nv}\),
where the zero vector then serves as the natural center of the parameter
space and induces the simplest possible model properties for the RBM
(i.e., at \(\boldsymbol \theta = \boldsymbol 0\), all variables are
independent uniformly distributed on \(\{-1,1\}^\nv\)). The proof of
this and further exploration of the equivalence of the
\(\boldsymbol \theta\) parameterization of the RBM model class and
parameterization by \(\boldsymbol \mu(\boldsymbol \theta)\) is in the
on-line supplementary materials. Hence, while from some computing
perspectives \(0/1\) coding might seem most natural, the \(-1/1\) coding
is far more convenient and interpretable from the point of view of
statistical modeling, where it makes parameters simply interpreted in
terms of symmetrically defined main effects and interactions. Under the
data encoding \(-1/1\), the parameter space centered at 0 is also
helpful for framing parameter configurations that are undesirably large
(i.e., these are naturally parameters that have moved too far away from
\(\bm{0}\) where the RBM model is anchored to be trivially describable
and completely problem-free). In light of all of these matters we will
henceforth employ the \(-1/1\) coding.

\subsection{Exploring manageable
examples}\label{exploring-manageable-examples}

To explore the impact of RBM parameter vector \(\boldsymbol \theta\)
magnitude on near-degeneracy, instability, and uninterpretability, we
consider models of small size. For \(\nh, \nv \in \{1, \dots, 4\}\), we
sample 100 values of \(\boldsymbol \theta\) with various magnitudes
(details to follow). For each set of parameters we then calculate
metrics of model impropriety introduced in Section
\ref{degeneracy-instability-and-uninterpretability} based on
\(\boldsymbol \mu(\boldsymbol \theta)\),
\(\text{ELPR}(\boldsymbol \theta)/\nv\), and the absolute coordinates of
\(\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] - \text{E}\left[\boldsymbol X | \boldsymbol \theta^* \right]\).
In the case of near-degeneracy, we classify each model as
near-degenerate based on the distance of
\(\boldsymbol \mu(\boldsymbol \theta)\) from the boundary of the convex
hull of \(\mathcal{T}\) and look at the fraction of models that are
``near-degenerate,'' meaning they are within a small distance
\(\epsilon > 0\) of the boundary of the convex hull. We define ``small''
through a rough estimation of the volume of the hull for each model
size. We pick \(\epsilon_0 = 0.05\) for \(\nh=\nv=1\) and then, for
every other \(\nh\) and \(\nv\), set \(m=\nh+\nv+\nv*\nh\) and pick
\(\epsilon\) so that \(1-(1-2\epsilon_0)^3 = 1 - (1-2\epsilon)^m\). In
this way, we roughly scale the volume of the ``small distance'' to the
boundary of the convex hull to be equivalent across model dimensions.

In our numerical experiment, we split
\(\boldsymbol \theta = (\boldsymbol \theta_{main}, \boldsymbol \theta_{interaction})\)
into \(\boldsymbol \theta_{main}\) and
\(\boldsymbol \theta_{interaction}\), in reference to which variables in
the probability function the parameters correspond (whether they
multiply a \(v_i\) or a \(h_j\) or they multiply a \(v_i h_j\)), and
allow the two types of terms to have varying average magnitudes,
\(||\boldsymbol \theta_{main} || /(\nh+\nv)\) and
\(||\boldsymbol \theta_{interaction} || /(\nh*\nv)\). These average
magnitudes vary on a grid between 0.001 and 3 with 24 breaks, yielding
576 grid points. (By looking at the average magnitudes, we are able to
later consider the potential benefit of shrinking each parameter value
\(\theta_i\) towards zero in a Bayesian fitting technique.) At each
point in the grid, 100 vectors (\(\boldsymbol \theta_{main}\)) are
sampled uniformly on a sphere with radius corresponding to the first
coordinate in the grid and 100 vectors
(\(\boldsymbol \theta_{interction}\)) are sampled uniformly on a sphere
with radius corresponding to the second coordinate in the grid via sums
of squared and scaled iid Normal\((0, 1)\) variables. These vectors are
then paired to create 100 values of \(\boldsymbol \theta\) with
magnitudes at each point in the grid.

The results of this numerical study are summarized in Figures
\ref{fig:degen-plots}, \ref{fig:instab-plots}, and
\ref{fig:uninterp-plots}. From these three figures, it is clear that all
three measures of model impropriety show higher values for larger
magnitudes of the parameter vectors, supporting the RBM model properties
developed in Section \ref{background}. As a compounding issue, these
figures show that, as model grow in size, it becomes easier for more
parameter configurations to push RBM models into near-degeneracy,
instability and uninterpretability. Additionally, since there are
\(\nh*\nv\) interaction terms in \(\boldsymbol \theta\) versus only
\(\nh + \nv\) main effect terms, for large models there are many more
interaction parameters than main effects in the models. And so, severely
limiting the magnitude of the individual interactions may well help
prevent model impropriety. \par
\begin{figure}
\includegraphics{paper_files/figure-latex/degen-plots-1} \caption{Results from the numerical experiment, here looking at the fraction of models that were near-degenerate for each combination of magnitude of $\boldsymbol \theta$ and model size, where $\boldsymbol \theta = (\boldsymbol \theta_{main}, \theta_{interaction})$ is split into main and interaction parameters. Black lines show the contour levels for fraction of near-degeneracy, while the thick black line shows the level where the fraction of near-degenerate models is .05.}\label{fig:degen-plots}
\end{figure}
\begin{figure}
\includegraphics{paper_files/figure-latex/instab-plots-1} \caption{The sample mean value of $\text{ELPR}(\boldsymbol \theta)/V$ at each grid point for each combination of magnitude of $\boldsymbol \theta$ and model size. As the magnitude of $\boldsymbol \theta$ grows, so does the value of this metric, indicating typical instability in the model.}\label{fig:instab-plots}
\end{figure}
\begin{figure}
\includegraphics{paper_files/figure-latex/uninterp-plots-1} \caption{The sample mean of the maximum component of the absolute difference between the model expectation vector, E$\left[\boldsymbol X | \boldsymbol \theta\right]$, and the expectation vector given independence, E$\left[\boldsymbol X | \boldsymbol \theta^* \right ]$. Larger magnitudes of $\boldsymbol \theta$ correspond to larger differences, thus indicating reduced interpretability.}\label{fig:uninterp-plots}
\end{figure}
Figure \ref{fig:split-plots-dist} shows the fraction of near-degenerate
models for each magnitude of \(\boldsymbol \theta\) for each model
architecture. For each number \(\nv\) of visibles in the model, as the
number \(\nh\) of hiddens increase, the fraction near-degenerate
diverges from zero at increasing rates for larger values of
\(||\boldsymbol \theta||\). This shows that, as model size gets larger,
the risk of degeneracy starts at a slightly larger magnitude of
parameters, but very quickly increases until reaching close to 1. \par
\begin{figure}
\includegraphics{paper_files/figure-latex/split-plots-dist-1} \caption{The fraction of near-degenerate models for each magnitude of $\boldsymbol \theta$. For each number $\nv$ of visibles in the model, the fraction near-degenerate moves away from zero at larger values of $||\boldsymbol \theta||$ as the number $\nh$ of hidden variables increases and the slope becomes steeper as $\nh$ increases as well.}\label{fig:split-plots-dist}
\end{figure}
These manageable examples indicate that RBMs are near-degenerate,
unstable, and uninterpretable for large portions of the parameter space
with large \(\|\boldsymbol \theta\|\). These problematic aspects require
serious consideration when using RBM models, on top of the additional
matter of principled/rigorous fitting of RBM models.

\section{Model Fitting}\label{model-fitting}

Typically, fitting a RBM via maximum likelihood (ML) methods will be
infeasible due mainly to the intractability of the normalizing term
\(\gamma(\boldsymbol \theta)\) in a model (\eqref{eq:pmf}) of any
realistic size. Ad hoc methods are used instead, which aim to avoid this
problem by using stochastic ML approximations that employ a small number
of MCMC draws (i.e., contrastive divergence, (Hinton
\protect\hyperlink{ref-hinton2002training}{2002})).

However, computational concerns are not the only issues with fitting a
RBM using ML. In addition, a RBM model, with the appropriate choice of
parameters and number of hiddens, has the potential to re-create any
distribution for the data (i.e., reproduce any specification of cell
probabilities for the binary data outcomes). For example, Montufar and
Ay (\protect\hyperlink{ref-montufar2011refinements}{2011}) show that any
distribution on \(\{0, 1\}^{\nv}\) can be approximated arbitrarily well
by a RBM with \(2^{\nv-1} - 1\) hidden units. We provide a small example
that illustrates that in fact there can be many such approximations. For
simplicity, consider a model with two visible variables \((V_1,V_2)\)
and one hidden \(H_1\) so that there are four possible data realizations
for \((V_1,V_2)\) given by \((\pm 1, \pm 1)\) and we may express the
model probabilities as \[
P(V_1=v_1,V_2=v_2| \theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21}) \propto  \exp\left( v_1  \theta_{v_1} + v_2 \theta_{v_2}\right) \sum_{h \in \{\pm 1\}}\exp\left( h[ \theta_{h_1}  + \theta_{11} v_1 + \theta_{21} v_2] \right),
\] for \((v_1,v_2)\in \{-1, 1\}^2\), in terms of real-valued parameters
\(\theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21}\).
Given some specified cell probabilities, say
\begin{equation}
\label{eq:0}
0\leq p_{(-1,-1)},\;p_{(1,-1)}, \;p_{(-1,1)},\;p_{(1,1)},
\end{equation}
for the outcomes \((\pm 1, \pm 1)\), the appendix provides technical
details to show parameter values
\((\theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21})\)
may be chosen to match any given cell probabilities with arbitrary
closeness. In fact, when the cell probabilities (\eqref{eq:0}) are all
strictly positive, parameters in the RBM model can be specified to
reflect these probabilities exactly. And, when one or more of the cell
probabilities (\eqref{eq:0}) are zero, the corresponding RBM probabilities
\(P(V_1=v_1,V_2=_v2| \theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21})\)
may never be identically zero (due to exponential terms in the model)
but parameters can be still selected to make the appropriate RBM cell
probabilities arbitrarily small. Furthermore, as development in the
appendix shows, only it is possible to approximate any distribution on
the visibles arbitrarily well (cf. Montufar and Ay
\protect\hyperlink{ref-montufar2011refinements}{2011}), but dramatically
different parameter settings can induce the same RBM model (beyond mere
symmetries in parameterization). A further, and somewhat odd consequence
of the RBM parameterization is then that, when fitting the RBM model by
likelihood-based methods, we already know the nature of the answer
before we begin: namely, such fitting will simply reproduce the
empirical distribution of the training data if sufficiently many hiddens
are in the model and there can be no model refinements or smoothing from
the RBM. That is, based on a random sample of vectors of visible
variables, the model for the cell probabilities with the highest
likelihood over \emph{all possible model classes} (i.e., RBM-based or
not) is the empirical distribution, and the over-parameterization of the
RBM model itself ensures that this empirical distribution can be
arbitrarily well-approximated. For illustration, considering the simple
example from above with \(n\) iid observations, each consisting of two
realized visibles \((V_1,V_2)\) the empirical cell frequencies from the
sample are the highest likelihood choices for the cell probabilities
\(p_{(-1,-1)},p_{(1,-1)}, p_{(-1,1)},p_{(1,1)}\) in (\eqref{eq:0}) before
any parametric model is introduced for improved analysis, and the
discussion above indicates that RBM model parameters can be chosen to
re-create this empirical distribution to an arbitrarily close degree.
Not only does RBM model fitting based on ML seek to reproduce the
empirical distribution, whenever this empirical distribution contains
empty cells, fitting steps for the RBM model will further aim to choose
parameters that necessarily diverge to infinity in magnitude in order to
zero-out the corresponding RBM cell probabilities. In data applications
with a large sample space, it is unlikely that the training set will
include at least one of each possible vector outcome (unlike this small
example). This implies that some RBM model parameters must diverge to
\(+\infty\) to mimic the empirical distribution with empty cells and, as
we have already discussed in Section
\ref{explorations-of-model-properties-through-simulation}, large
magnitudes of \(\boldsymbol \theta\) lead to model impropriety in the
RBM.

Here we consider what might be done in a principled manner to prevent
both overfitting and model impropriety, testing on a \(\nv = \nh = 4\)
case that already stretches the limits of what is computable - in
particular we consider Bayes methods.

\subsection{Bayesian model fitting}\label{bayesian-model-fitting}

To avoid model impropriety for a fitted RBM, we wish to avoid parts of
the parameter space \(\mathbb{R}^{\nv + \nh + \nv*\nh}\) that lead to
near-degeneracy, instability, and uninterpretability. Motivated by the
insights in Section \ref{exploring-manageable-examples}, one idea is to
shrink
\(\boldsymbol \theta = (\boldsymbol \theta_{main}, \boldsymbol \theta_{interaction})\)
toward \(\boldsymbol 0\) by specifying priors that place low probability
on large values of \(||\boldsymbol \theta||\), specifically focusing on
shrinking \(\boldsymbol \theta_{interaction}\) more than
\(\boldsymbol \theta_{main}\). This is similar to an idea advocated by
Hinton (\protect\hyperlink{ref-hinton2010practical}{2010}) called
\emph{weight decay}, in which a penalty is added to the interaction
terms in the model, \(\boldsymbol \theta_{interaction}\), shrinking
their magnitudes. \par 
\begin{table}[ht]
\centering
\begin{tabular}{rrrrrr}
  \hline
Parameter & Value & Parameter & Value & Parameter & Value \\ 
  \hline
$\theta_{v1}$ & $-1.104376$ & $\theta_{11}$ & $-0.0006334$ & $\theta_{31}$ & $-0.0038301$ \\ 
  $\theta_{v2}$ & $-0.2630044$ & $\theta_{12}$ & $-0.0021401$ & $\theta_{32}$ & $0.0032237$ \\ 
  $\theta_{v3}$ & $0.3411915$ & $\theta_{13}$ & $0.0047799$ & $\theta_{33}$ & $0.0020681$ \\ 
  $\theta_{v4}$ & $-0.2583769$ & $\theta_{14}$ & $0.0025282$ & $\theta_{34}$ & $0.0041429$ \\ 
  $\theta_{h1}$ & $-0.1939302$ & $\theta_{21}$ & $0.0012975$ & $\theta_{41}$ & $0.0089533$ \\ 
  $\theta_{h2}$ & $-0.0572858$ & $\theta_{22}$ & $0.0000253$ & $\theta_{42}$ & $-0.0042403$ \\ 
  $\theta_{h3}$ & $-0.2101802$ & $\theta_{23}$ & $-0.0004352$ & $\theta_{43}$ & $-0.000048$ \\ 
  $\theta_{h4}$ & $0.2402456$ & $\theta_{24}$ & $-0.0086621$ & $\theta_{44}$ & $0.0004767$ \\ 
   \hline
\end{tabular}
\caption{Parameters used to fit a test case with $V = H = 4$. This parameter vector was chosen as a sampled value of $\boldsymbol \theta$ that was not near the convex hull of the sufficient statistics for a grid point in Figure \ref{fig:degen-plots} with $< 5$\% near-degeneracy.} 
\label{tab:theta}
\end{table}
We considered a test case with \(\nv = \nh = 4\) and parameters given in
Table \ref{tab:theta}. This parameter vector was chosen as a sampled
value of \(\boldsymbol \theta\) at which the resulting RBM model would
not be clearly degenerate. We simulated \(n = 5,000\) realizations of
visibles as a training set and fit the RBM using three Bayes
methodologies. These involved the following set-ups with choice of prior
distribution \(\pi(\boldsymbol \theta)\) for parameters
\(\boldsymbol \theta\).
\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  \emph{A ``trick'' prior.} Here we cancel out normalizing term in the
  likelihood (from \(\gamma(\boldsymbol \theta)\) in (\eqref{eq:pmf})) so
  that resulting full conditionals of \(\boldsymbol \theta\) are
  multivariate Normal. Namely, this involves a prior of the form
  \begin{align*}
  \pi(\boldsymbol \theta) \propto \gamma(\boldsymbol \theta)^n \exp\left(-\frac{1}{2C_{1}}\boldsymbol \theta_{main}'\boldsymbol \theta_{main} -\frac{1}{2C_{2}}\boldsymbol \theta_{interaction}'\boldsymbol \theta_{interaction}\right), \vspace{-.75cm}
  \end{align*}
  where
  \begin{align*}
  \gamma(\boldsymbol \theta) = \sum\limits_{\boldsymbol x \in \mathcal{C}^{\nh + \nv}}\exp\left(\sum\limits_{i = 1}^\nv\sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right) 
  \end{align*}
  for hyperparameter choices \(0< C_2 < C_1\). The unknown hidden
  variables \(h_j\) are also directly treated as latent variables and
  are sampled in each MCMC iterative draw from the posterior
  distribution. This is the method of Li
  (\protect\hyperlink{ref-li2014biclustering}{2014}). We will refer to
  this method as Bayes with Trick Prior and Latent Variables (BwTPLV).
\item
  \emph{A truncated Normal prior.} Here we use independent spherical
  normal distributions as priors for \(\boldsymbol \theta_{main}\) and
  \(\boldsymbol \theta_{interaction}\), which are \emph{truncated} at
  \(3\sigma_{main}\) and \(3\sigma_{interaction}\), respectively, based
  on standard deviation hyperparameters
  \(0<\sigma_{interaction}<\sigma_{main}\). Full conditional
  distributions are not conjugate, and simulation from the posterior was
  accomplished using a geometric adaptive Metropolis Hastings step (Zhou
  \protect\hyperlink{ref-zhou2014some}{2014}) and calculation of
  likelihood normalizing constant. (This computation is barely feasible
  for a problem of this size and would be infeasible for larger
  problems.) Here the hidden variables \(h_j\) are again carried along
  in the MCMC implementation as latent variables. We will refer to this
  method as Bayes with Truncated Normal prior and Latent Variables
  (BwTNLV).
\item
  \emph{A truncated Normal prior and marginalized likelihood.} Here we
  marginalize out the hidden variables
  \(\boldsymbol h = (h_1, \dots, h_\nh)\) in
  \(f_{\boldsymbol \theta}(\boldsymbol x)\), and use the truncated
  Normal priors applied to the marginal probabilities for visible
  variables given by
  \begin{align*}
  g_{\boldsymbol \theta}(\boldsymbol v) \propto \sum\limits_{\boldsymbol h \in \mathcal{C}^\nh} \exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right), \boldsymbol v \in \mathcal{C}^\nv.
  \end{align*}
  We will refer to this method as Bayes with Truncated Normal prior and
  Marginalized Likelihood (BwTNML).
\end{enumerate}
The three fitting methods are ordered above according to computational
feasibility in a real-data situation, with BwTPLV being the most
computationally feasible due to conjugacy and BwTNML the least feasible
due to the marginalization and need for an adaptive Metropolis Hastings
step. All three methods require choosing the values of hyperparameters.
In each case, we have chosen these values based on a rule of thumb that
shrinks \(\boldsymbol \theta_{interaction}\) more than
\(\boldsymbol \theta_{main}\). Additionally, BwTPLV requires additional
tuning (i.e.~a tuning parameter \(C > 0\) in Table \ref{tab:hyperparam})
to choose \(C_1\) and \(C_2\), reducing its appeal. The forms used for
the hyperparameters in our simulation are presented in Table
\ref{tab:hyperparam}.
\begin{table}[ht]
\centering
\begin{tabular}{|l|c|c|}
\hline 
Method & Hyperparameter & Value \\ 
\hline \hline
\multirow{2}{*}{BwTPLV} & $C_1$ & $\frac{C}{n}\frac{1}{\nh + \nv}$ \\
 & $C_2$ & $\frac{C}{n}\frac{1}{\nh*\nv}$ \\
\hline
\multirow{2}{*}{BwTNLV} & $\sigma^2_{main}$ & $\frac{1}{\nh + \nv}$ \\
 & $\sigma^2_{interaction}$ & $\frac{1}{\nh*\nv}$ \\
\hline
\multirow{2}{*}{BwTNML} & $\sigma^2_{main}$ & $\frac{1}{\nh + \nv}$ \\
 & $\sigma^2_{interaction}$ & $\frac{1}{\nh*\nv}$ \\
\hline
\end{tabular}
\caption{The values used for the hyperparameters for all three fitting methods. A rule of thumb is imposed which decreases prior variances for the model parameters as the size of the model increases and also shrinks $\boldsymbol \theta_{interaction}$ more than $\boldsymbol \theta_{main}$. The common $C$ defining $C_1$ and $C_2$  in the BwTPLV method is chosen by tuning.}
\label{tab:hyperparam}
\end{table}
It should be noted that, due to the common prior distributions for
\(\boldsymbol \theta\), both BwTNLV (method 2 above) and BwTNML (method
3) are drawing from the same stationary posterior distribution for
vectors of visibles. A fundamental difference between these two methods
lies in how well these two chains mix and how quickly they arrive at the
target posterior distribution. After a burn-in period of 50 iterations
selected by inspecting the trace plots, we assess the issue of mixing in
two ways. First, the autocorrelation functions (ACF) from each posterior
sample corresponding to a model probability for a visible vector outcome
\(\mathbf{v}=(v_1,v_2,v_3,v_4)\in\{\pm 1\}^4\) (i.e., computed from
\(\boldsymbol \theta\) under (\eqref{eq:pmf})) are determined and plotted
in Figure \ref{fig:acf} with BwTNLV in black and BwTNML in red. As
expected, ACF corresponding to the method (BwTNML) that marginalizes out
the hidden variables from the likelihood decreases to zero at a much
faster rate, indicating better mixing for the chain. \par
\begin{figure}
\centering
\includegraphics{paper_files/figure-latex/acf-1.pdf}
\caption{\label{fig:acf}The autocorrelation functions (ACF) for the
posterior probabilityies of all \(2^4 = 16\) possible outcomes for the
vector of four visibles assessed at multiple lags for each method with
BwTNLV in black and BwTNML in red. As expected, ACF corresponding to the
method that marginalizes out the hidden variables from the likelihood
decreases to zero at a much faster rate, indicating better mixing for
the chain.}
\end{figure}
Secondly, we can assess the mixing of the BwTNLV/BwTNML chains using the
notion of effective sample size. If the MCMC chain were truly iid draws
from the target distribution, then for the parameter \(p^{(i)}\)
denoting the probability of the \(i\)th vector outcome for the four
visibles \(\mathbf{v}=(v_1,v_2,v_3,v_4)\in\{\pm 1\}^4\),
\(i=1,\ldots,16\), its estimate as the average \(\bar{p}^{(i)}\) of
posterior sample versions would be approximately Normal with mean given
by the posterior marginal mean of \(p^{(i)}\), and variance given by
\(\sigma^2_i/M\), where \(\sigma^2_i\) is the true posterior variance of
\(p^{(i)}\) and \(M\) is the length of the chain. However, with the
presence of correlation in our chain, the asymptotic variance of
\(\bar{p}^{(i)}\) is instead approximately some \(C_i/M\), where \(C_i\)
is some positive constant such that \(C_i > \sigma^2_i\). We can use an
overlapping block-means approach (Gelman, Shirley, and others
\protect\hyperlink{ref-gelman2011inference}{2011}) to get a crude
estimate for \(C_i\) as \(\hat{C}_i = bS_b^2\), where \(S_b^2\) denotes
the sample variance of overlapping block means
\(\{\bar{p}_j^{(i)}=\sum_{k=j}^{j+b-1} p_k^{(i)}/b\}_{j=1}^{M-b+1}\) of
length \(b\) computed from the posterior samples
\(\{p_k^{(i)}\}_{k=1}^M\). We compare it to an estimate of
\(\sigma^2_i\) using sample variance \(\hat{\sigma}^2_i\) of the raw
chain, \(\{p_k^{(i)}\}_{k=1}^{M}\). Formally, we approximate the
effective sample size of the length \(M\) MCMC chain as \[
M_{eff}^{(i)} = M\frac{\hat{\sigma}^2_i}{\hat{C}_i}.
\]
\begin{table}[ht]
\centering
\begin{tabular}{lrrlrr}
  \hline
Outcome & BwTNLV & BwTNML & Outcome & BwTNLV & BwTNML \\ 
  \hline
1 & 73.00 & 509.43 & 9 & 83.47 & 394.90 \\ 
  2 & 65.05 & 472.51 & 10 & 95.39 & 327.35 \\ 
  3 & 87.10 & 1229.39 & 11 & 70.74 & 356.56 \\ 
  4 & 72.64 & 577.73 & 12 & 81.40 & 338.30 \\ 
  5 & 71.67 & 452.01 & 13 & 105.98 & 373.59 \\ 
  6 & 66.49 & 389.78 & 14 & 132.61 & 306.91 \\ 
  7 & 84.30 & 660.37 & 15 & 82.15 & 365.30 \\ 
  8 & 75.46 & 515.09 & 16 & 98.05 & 304.57 \\ 
   \hline
\end{tabular}
\caption{The effective sample sizes for a chain of length $M = 1000$ regarding all $16$ probabilities for possible vector outcomes of visibles. BwTNLV would require at least $4.7$ times as many MCMC iterations to achieve the same amount of effective information about the posterior distribution.} 
\label{tab:m-eff}
\end{table}
For both BwTNLV and BwTNML methods, effective sample sizes for a chain
of length \(M = 1000\) for inference about each of the \(2^4 = 16\)
model probabilities are presented in Table \ref{tab:m-eff}. These range
from \(304.57\) to \(1229.39\) for BwTNML, while BwTNLV only yields
between \(65.05\) and \(132.61\) effective draws. Thus, BwTNLV would
require at least \(4.7\) times as many iterations to be run of the MCMC
chain in order to achieve the same amount of effective information about
the posterior distribution. For this reason, consistent with the ACF
results in Figure \ref{fig:acf}, BwTNLV does not seem to be an effective
method for fitting the RBM, though computing resources can hinder use of
the alternative BwTNML involving marginalization of hidden variables.

Figure \ref{fig:fitting-plot} shows the posterior probability of each
possible \(\boldsymbol v \in \{-1,1\}^4\) after fitting the RBM model
according to method 1 (BwTPLV using trick prior) and method 3 (BwTNML)
(excluding method 2 (BwTNLV) that seeks the same posterior as method 3)
. The black vertical lines show the true probabilities of each image
based on the parameters used to generate the training set while the red
vertical lines show the empirical distribution for the training set of
\(5,000\) vectors. From these posterior predictive checks, it is evident
that BwTNML produces the best fit to the data. Furthermore, along with
the discussion of Section \ref{model-fitting}, Figure
\ref{fig:fitting-plot} also shows that it can be undesirable to seek to
perfectly re-create an empirical distribution in fitting RBM models
(i.e., true model probabilities may differ substantially). The priors in
the BwTNML method constrain the RBM model fit to avoid replication of
the empirical distribution and better estimate the underlying true data
generating probabilities. However, this method requires a
marginalization step to obtain the probability function of visible
observations alone, which is infeasible for a model with \(\nh\) of any
real size. \par
\begin{figure}
\centering
\includegraphics{paper_files/figure-latex/fitting-plot-1.pdf}
\caption{\label{fig:fitting-plot}Posterior probabilities of \(16 = 2^4\)
possible realizations of \(4\) visibles using two of the three Bayesian
fitting techniques, BwTPLV and BwTNML. The black vertical lines show the
true probabilities of each vector of visibles based on the parameters
used to generate the training data while the red vertical lines show the
empirical distribution. BwTNML produces the best fit for the data,
however is also the most computationally intensive and least feasible
with a real dataset.}
\end{figure}
\section{Discussion}\label{discussion}

RBM models constitute an interesting class of undirected graphical
models that are thought to be useful for supervised learning tasks.
However, when viewed as generative statistical models, RBMs are prone to
forms of model impropriety such as near-degeneracy, S-instability, and
uninterpretability. Additionally, these models are difficult to fit
using a rigorous methodology, due to the dimension of the parameter
space coupled with the size of the latent variable space.

In this paper, we have presented three fully Bayes-principled MCMC-based
methods for fitting RBMs. Common practice is to use a kind of MCMC to
overcome fitting complexities. However because of the size of the space
to be filled with MCMC iterates, convergence and mixing of these methods
will be slow. Marginalization over the latent variables in the model is
shown to improve mixing, but is numerically intractable due to the
necessity of repeated calculation of the normalizing constant. Due to
the extreme flexibility in this model class, rigorous likelihood-based
fitting for a RBM will typically aim to merely merely return the
(discrete) empirical distribution for visibles, meaning any practitioner
should be aware of this and employ some form of regularization or
penalization in the model.

Ultimately, it is not clear that RBM models are useful as generative
models. Furthermore, without the appropriate generative behavior in the
RBM model, the ability to quantify uncertainty in the estimated model
parameters becomes impossible and the model loses useful application to
prediction problems (e.g., realistic predictive distributions). In the
case of classification, predictive distributions not be the ultimate
goal, but when S-instability is present, small (imperceptible)
differences in the data may lead to greatly different probabilities and
thus greatly different classifications. For these reasons, we are
skeptical about RBMs as data generative tools.

\clearpage

\appendix

\section{On the one-to-one correspondence between parameters and moments
in the RBM distribution}\label{appendix-rbm}

For integers \(m,n \geq 1\), consider random vectors
\(\bm{V}_m \equiv(V_1,\ldots,V_m)\) and
\(\bm{H}_n\equiv (H_1,\ldots,H_n)\), where the random variables \(V_i\)
and \(H_j\) assume values in \(\{1,-1\}\), \(i=1,\ldots,m\),
\(j=1,\ldots,n\). We suppose the vector \((\bm{V}_m,\bm{H}_n)\) follows
a restricted Boltzmann machine (RBM) distribution, i.e.~has probability
mass function
\begin{align*}
&P(V_1= v_1,\ldots,V_m= v_m, H_1= h_1,\ldots,H_n=h_n|\bm{\theta}) \propto \exp\left[ \sum_{i=1}^m v_i \theta_{v_j} + \sum_{j=1}^n h_j \theta_{h_j} + \sum_{i=1}^m  \sum_{j=1}^n v_i  h_j \theta_{ij} \right], \\
& \quad \{v_i\}_{i=1}^m,\{h_j\}_{j=1}^n \subset\{\pm 1\},
\end{align*}
with \(m+n+mn\) real parameters
\(\theta_{v_1},\ldots,\theta_{v_m},\theta_{h_1},\ldots,\theta_{h_m}\)
and \(\theta_{ij}\), \(i=1,\ldots,m\), \(j=1,\ldots,n\) collected in a
parameter vector denoted as \(\bm{\theta}\in \mathbb{R}^{m+n+mn}\).
Denote the distribution of \((\bm{V}_m,\bm{H}_n)\) as
RBM\((m,n,\bm{\theta})\). In the following, for convenience, we say a
symmetric random variable \(X\) has a Bernoulli\((1/2)\) distribution if
\(P(X = 1) = 1/2 = P(X = -1)\).

Use the notation
\(\bm{V}_m*\bm{H}_n = (V_1 H_1,\ldots,V_1 H_n, V_2 H_1,\ldots,V_2 H_n,\ldots, V_m H_1,\ldots,V_m H_n)\).
For the RBM\((m,n,\bm{\theta})\) model, consider in the function
\(\bm{g}: \mathbb{R}^{m+n+mn} \rightarrow \mathcal{I} \subset \mathbb{R}^{m+n+mn}\)
given by
\begin{align*}
\bm{g}(\bm{\theta}) &= \E_{\bm{\theta}}(\bm{V}_m,\bm{H}_n,\bm{V}_m*\bm{H}_n)\\
&= \sum_{\bm{v}_m \in \{\pm 1\}^m, \bm{h}_n \in \{\pm 1\}^n} (\bm{v}_m,\bm{h}_n,\bm{v}_m*\bm{h}_n)P(\bm{V}_m=\bm{v}_m,\bm{H}_n=\bm{h}_n|\bm{\theta})
\end{align*}
for \(\bm{\theta}\in \mathbb{R}^{m+n+mn}\); above
\(\mathcal{I} \equiv \{\bm{g}(\bm{\theta}) : \bm{\theta}\in \mathbb{R}^{m+n+mn}\}\)
denotes the image of the mapping.

We have the following properties:
\begin{itemize}
\tightlist
\item
  The function \(\bm{g}: \mathbb{R}^{m+n+mn} \rightarrow \mathcal{I}\)
  is one-to-one; see Theorem \ref{thm:one} below. Hence, these moments
  uniquely characterize the parameters in the RBM\((m,n,\bm{\theta})\)
  model.
\item
  The image \(\mathcal{I}\) is a symmetric set (i.e.,
  \(\bm{g}(\bm{\theta})=-\bm{g}(-\bm{\theta})\)) and must be a connected
  (not necessarily convex) region in \(\mathbb{R}^{m+n+mn}\), without
  any voids in it (by the continuity of \(\bm{g}(\bm{\theta})\)).
\item
  The center of the image \(\mathcal{I}\) (the zero vector in
  \(\mathbb{R}^{m+n+mn}\)) is given only by \(\bm{\theta}=\bm{0}\) (the
  zero vector in the parameter space). This corresponds to the case
  where all the random vectors \(V_1,\ldots,V_m,H_1,\ldots,H_n\) are iid
  with a symmetric Bernoulli\((1/2)\) distribution; see Theorem
  \ref{thm:two}.
\end{itemize}
\BeginKnitrBlock{theorem}
\protect\hypertarget{thm:one}{}{\label{thm:one}}Let \(m,n\geq 1\) and
suppose the random vector
\((\bm{V}_m,\bm{H}_n) = (V_1,\ldots,V_m, H_1,\ldots,H_n)\) follows a
RBM\((m,n,\bm{\theta})\) distribution. Let
\(\bm{\theta}_1,\bm{\theta}_2 \in \mathbb{R}^{m+n+mn}\) and let \(P_i\)
and \(\E\) denote probability and expectation under \(\bm{\theta}_i\),
\(i=1,2\). Then, the following are equivalent
\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  \(\E_1 V_i=\E_2 V_i\), \(\E_1 H_j=\E_2 H_j\) and
  \(\E_1 V_i H_j =\E_2 V_i H_j\) for any \(i=1,\ldots,m\),
  \(j=1,\ldots,n\).
\item
  \(\bm{\theta}_1 =\bm{\theta}_2\).
\item
  \(P_1\) and \(P_2\) are the same distribution for
  \((\bm{V}_m,\bm{H}_n)\).\textbackslash{}
\end{enumerate}
\EndKnitrBlock{theorem}

\BeginKnitrBlock{proof}
\iffalse <span class="proof"><em>Proof. \fi{} We first establish the
equivalence of claims 2 and 3. If claim 2 holds, then claim 3 holds
trivially. Suppose claim 3 holds and pick \(\ell \in \{1,\ldots,m\}\).
Let
\(\theta_{v_1},\ldots,\theta_{v_m},\theta_{h_1},\ldots,\theta_{h_n}\),
\(\theta_{ij}\), \(i=1,\ldots,m\), \(j=1,\ldots,n\), denote the
components of \(\bm{\theta}_1\) and let
\(\tilde{\theta}_{v_1},\ldots,\tilde{\theta}_{v_m},\tilde{\theta}_{h_1},\ldots,\tilde{\theta}_{h_n}\),
\(\tilde{\theta}_{ij}\), \(i=1,\ldots,m\), \(j=1,\ldots,n\), denote the
analogous components of \(\bm{\theta}_2\). Then, by assumption,
\begin{align*}
\exp\left[ 2\theta_{v_\ell} +2 \sum_{j=1}^n  h_j \theta_{\ell j} \right] &= \frac{P_1\left(\{V_\ell = 1\} \cap \bigcap_{i=1 \atop i \neq \ell}^m \{V_i=v_i\} \cap \bigcap_{j=1}^n \{H_j=h_j\}\right)} {P_1\left(\{V_\ell = -1\} \cap \bigcap_{i=1 \atop i \neq \ell}^m \{V_i=v_i\} \cap \bigcap_{j=1}^n \{H_j=h_j\}\right)}\\
&= \frac{P_2\left(\{V_\ell = 1\} \cap \bigcap_{i=1 \atop i \neq \ell}^m \{V_i=v_i\} \cap \bigcap_{j=1}^n \{H_j=h_j\}\right)}
{P_2\left(\{V_\ell = -1\} \cap \bigcap_{i=1 \atop i \neq \ell}^m \{V_i=v_i\} \cap \bigcap_{j=1}^n \{H_j=h_j\}\right)}\exp\left[ 2\tilde{\theta}_{v_\ell} +2 \sum_{j=1}^n  h_j \tilde{\theta}_{\ell j} \right]
\end{align*}
for any \(\{v_i\}_{i=1,i\neq \ell}^m,\{h_j\}_{j=1}^n \subset\{\pm 1\}\).
Equating exponents, we have
\(\theta_{v_\ell} - \tilde{\theta}_{v_\ell} = \sum_{j=1}^n h_j ( \tilde{\theta}_{\ell j} -\theta_{\ell j})\)
for any \(\{h_j\}_{j=1}^n \subset \{\pm 1\}\). Taking all \(h_j=1\) or
all \(h_j=-1\) for \(1 \leq j \leq n\) yields
\(\theta_{v_\ell} = \tilde{\theta}_{v_\ell}\). This result further
implies
\(\tilde{\theta}_{\ell 1} -\theta_{\ell 1} = - \sum_{j=2}^n h_j ( \tilde{\theta}_{\ell j} -\theta_{\ell j})\)
for any given \(\{h_j\}_{j=2}^n \subset \{\pm 1\}\), so that
\(\theta_{\ell 1}=\tilde{\theta}_{\ell 1}\) must hold. Iterating this
argument, sequentially for each \(j=2,\ldots,m\), shows
\(\theta_{\ell j}=\tilde{\theta}_{\ell j}\) for any \(1 \leq j \leq n\)
in addition to \(\theta_{v_\ell}=\tilde{\theta}_{v_\ell}\). As
\(\ell\in\{1,\ldots,m\}\) was arbitrary, we have
\(\theta_{v_i}=\tilde{\theta}_{v_i}\) and
\(\theta_{i j}=\tilde{\theta}_{i j}\) for any \(1 \leq i \leq m\),
\(1 \leq j \leq n\).

Now pick \(k \in \{1,\ldots,n\}\) so that, by the
RBM\((m,n,\bm{\theta})\) probability structure as above, we have
analogously that \[
\theta_{h_k} + \sum_{i=1}^m  v_i \theta_{i k} =  \tilde{\theta}_{h_k} + \sum_{i=1}^m  v_i \tilde{\theta}_{i k}
\] for any \(\{v_i\}_{i=1}^m \subset \{\pm 1\}\), so that
\(\theta_{h_k} = \tilde{\theta}_{h_k}\). As \(k \in \{1,\ldots,n\}\) was
arbitrary, we have now established claim 2 from claim 3.

To establish the equivalence of claims 1 and 3 in Theorem \ref{thm:one}
under the RBM\((m,n,\bm{\theta})\) model, we note that claim 3 easily
implies claim 1. Hence, it suffices now to establish claim 3 from claim
1. This follows immediately from Lemma \ref{lem:first} below and the
fact that the variables \(V_1,\ldots,V_m\) are conditionally independent
given \(H_1,\ldots,H_n\) in the RBM\((m,n,\bm{\theta})\) model (and
likewise \(H_1,\ldots,H_n\) are independent given \(V_1,\ldots,V_m\)).
\EndKnitrBlock{proof}

\BeginKnitrBlock{lemma}
\protect\hypertarget{lem:first}{}{\label{lem:first}}Let \(m,n\geq 1\). Let
\(P_1,P_2\) denote two probability distributions for a random vector
\((\bm{V}_m,\bm{H}_n) = (V_1,\ldots,V_m, H_1,\ldots,H_n)\) supported on
\(\{a,b\}^{m+n}\), for some real constants \(a\neq b \in\mathbb{R}\),
such that \(P_i( \bm{V}_m=\bm{v}_m, \bm{H}_n=\bm{h}_n)>0\), \(i=1,2\),
for any \(\bm{v}_m \in \{a,b\}^{m}\), \(\bm{h}_n \in \{a,b\}^{n}\).
Suppose additionally that, under \(P_1\) or \(P_2\), the variables
\(V_1,\ldots,V_m\) are conditionally independent given
\(\bm{H}_n=\bm{h}_n \in \{a,b\}^{n}\) and that the variables
\(H_1,\ldots,H_n\) are conditionally independent given
\(\bm{V}_m=\bm{v}_m \in \{a,b\}^{m}\).

Let \(\E_i\) denote expectation under the distribution \(P_i\),
\(i=1,2\). Then, for any
\(\bm{v}_m \in \{a,b\}^{m}, \bm{h}_n \in \{a,b\}^{n}\), \[
P_1( \bm{V}_m=\bm{v}_m, \bm{H}_n=\bm{h}_n) = P_2( \bm{V}_m=\bm{v}_m, \bm{H}_n=\bm{h}_n), \qquad
\] follows if \(\E_1 V_i = \E_2 V_i\), \(\E_1 H_j = \E_2 H_j\),
\(\E_1 V_i H_j = \E_2 V_j H_j\) holds for all \(i\in\{1,\ldots,m\}\),
\(j\in \{1,\ldots,n\}\).
\EndKnitrBlock{lemma}

\BeginKnitrBlock{proof}
\iffalse <span class="proof"><em>Proof. \fi{} The case \(m=n=1\) follows
from Lemma \ref{lem:second}. We now use an induction argument, assuming
that the result of Lemma \ref{lem:first} holds for some given order of
\((m,n)\) with \(m,n\geq 1\) and show the the result continues to hold
for \((m+1,n)\) or \((m,n+1)\). We shall treat the case \((m+1,n)\)
(where the other case \((m,n+1)\) follows by symmetrical arguments).

Considering random variables
\((\bm{V}_{m+1},\bm{H}_n) = (V_1,\ldots,V_m,V_{m+1}, H_1,\ldots,H_n)\),
the induction hypothesis applies to \((\bm{V}_{m},\bm{H}_n)\) where
\(\bm{V}_m =(V_1,\ldots,V_m)\). Hence, for any
\(\bm{v}_m \in \{a,b\}^{m}, \bm{h}_n \in \{a,b\}^{n}\), we have
\(P_1(\bm{V}_m=\bm{v}_m,\bm{H}_n=\bm{h}_n) = P_2(\bm{V}_m=\bm{v}_m,\bm{H}_n=\bm{h}_n)>0\),
\(P_1( \bm{H}_n=\bm{h}_n) = P_2(\bm{H}_n=\bm{h}_n)>0\) and consequently
\begin{align}\label{eq:0}
P_1(\bm{V}_m=\bm{v}_m|\bm{H}_n=\bm{h}_n) =P_2(\bm{V}_m=\bm{v}_m |\bm{H}_n=\bm{h}_n).
\end{align}
Likewise, by the induction hypothesis applied to
\((\bm{V}_{m-},\bm{H}_n)\) for \(\bm{V}_{m-} = (V_2,\ldots,V_{m+1})\),
we have
\begin{eqnarray}\label{eq:0}
 1- P_1(V_{m+1} = b | \bm{H}_n=\bm{h}_n) &=& P_1(V_{m+1} = a |\bm{H}_n=\bm{h}_n)\\
\nonumber  &=&  P_2(V_{m+1} = a | \bm{H}_n=\bm{h}_n)\\
\nonumber  & = &1-P_2(V_{m+1} = b | \bm{H}_n=\bm{h}_n)
\end{eqnarray}
for any \(\bm{h}_n \in \{a,b\}^{n}\). Now fix
\(\bm{v}_m \in \{a,b\}^{m}, \bm{h}_n \in \{a,b\}^{n}\) and let
\(\bm{v}_{m+1}= (v_{m+1},\bm{v}_m)\) for \(v_{m+1}\in\{a,b\}\). Then, by
the conditional independence assumption, \[
P_i( \bm{V}_{m+1}=\bm{v}_{m+1}| \bm{H}_n=\bm{h}_n) = P_i( V_{m+1}=v_{m+1}| \bm{H}_n=\bm{h}_n) P_i( \bm{V}_{m}=v_{m}| \bm{H}_n=\bm{h}_n)
\] for \(i=1,2\), so that we have
\(P_1( \bm{V}_{m+1}=\bm{v}_{m+1}| \bm{H}_n=\bm{h}_n) = P_2( \bm{V}_{m+1}=\bm{v}_{m+1}| \bm{H}_n=\bm{h}_n)\)
by (@ref\{eq:0\})-(\eqref{eq:1}). Consequently, it follows that
\begin{eqnarray*}
P_1( \bm{V}_{m+1}=\bm{v}_{m+1}, \bm{H}_n=\bm{h}_n) &=& P_1( \bm{V}_{m+1}=\bm{v}_{m+1}| \bm{H}_n=\bm{h}_n) P_1(\bm{H}_n=\bm{h}_n)\\
& =& P_2( \bm{V}_{m+1}=\bm{v}_{m+1}| \bm{H}_n=\bm{h}_n)P_2(\bm{H}_n=\bm{h}_n) \\&=&P_2( \bm{V}_{m+1}=\bm{v}_{m+1}, \bm{H}_n=\bm{h}_n)
\end{eqnarray*}
by \(P_1(\bm{H}_n=\bm{h}_n)=P_2(\bm{H}_n=\bm{h}_n)\). As
\(\bm{v}_{m+1}\in \{a,b\}^{m+1},\bm{h}_n \in \{a,b\}^{n}\) were
arbitrary, this completes the proof of Lemma \ref{lem:first}.
\EndKnitrBlock{proof}

\BeginKnitrBlock{lemma}
\protect\hypertarget{lem:second}{}{\label{lem:second}}Suppose the discrete
random vector \((X,Y)\) has support \(\{a,b\}\times \{a,b\}\), for some
\(a\neq b \in \mathbb{R}\), under two probability distributions \(P_1\)
and \(P_2\). Let \(\E_i\) denote expectation under the distribution
\(P_i\), \(i=1,2\). Then, the following are equivalent
\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  \(\E_1 X = \E_2 X\), \(\E_1 Y = \E_2 Y\), \(\E_1 X Y = \E_2 X Y\),
  where \(\E_i\) denotes expectation under \(P_i\), \(i=1,2\).
\item
  \(P_1(X=x,Y=y)=P_2(X=x,Y=y)\) for \(x,y\in\{a,b\}\), i.e., \((X,Y)\)
  has the same distribution under \(P_1\) and \(P_2\).
\end{enumerate}
\EndKnitrBlock{lemma}

\BeginKnitrBlock{proof}
\iffalse <span class="proof"><em>Proof. \fi{} Denote the four
probabilities by which the random vector \((X,Y)\) assume pairs
\((b,b),(b,a),(a,b),(a,a)\), respectively, as
\(c_1,c_2,c_3,1-c_1-c_2-c_3\) under \(P_1\) and
\(d_1,d_2,d_3,1-d_1-d_2-d_3\) under \(P_2\). The conditions
\(\E_1 X = \E_2 X\), \(\E_1 Y = \E_2 Y\), \(\E_1 X Y = \E_2 X Y\) with
\(b\neq a\) imply that \[
(c_1+c_2) = (d_1+d_2), \quad  (c_1+c_3) = (d_1+d_3) \quad (b-a) c_1 + a (2c_1+c_2+c_3) =  (b-a) d_1 + a (2d_1+d_2+d_3).
\] As \((2c_1+c_2+c_3) = (2d_1+d_2+d_3)\) and \(b\neq a\), we conclude
that \(c_1=d_1\), from which it follows that \(c_2=d_2\) and
\(c_3=d_3\). Hence, claim 1 implies claim 2 in Lemma \ref{lem:second}.
Claim 2 also trivially implies claim 1.
\EndKnitrBlock{proof}

\BeginKnitrBlock{theorem}
\protect\hypertarget{thm:two}{}{\label{thm:two}}Let \(m,n\geq 1\) and
suppose the random vector
\((\bm{V}_m,\bm{H}_n) = (V_1,\ldots,V_m, H_1,\ldots,H_n)\) follows a
RBM\((m,n,\bm{\theta})\) distribution. Then, the following are
equivalent
\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  \(\E V_i=0\), \(\E H_j=0\) and \(\E V_i H_j =0\) for any
  \(i=1,\ldots,m\), \(j=1,\ldots,n\).
\item
  \(\bm{\theta} = \bm{0}\in \mathbb{R}^{m+n+mn}\).
\item
  \(V_1,\ldots,V_m, H_1,\ldots,H_n\) are iid Bernoulli\((1/2)\) random
  variables.
\end{enumerate}
\EndKnitrBlock{theorem}

\BeginKnitrBlock{proof}
\iffalse <span class="proof"><em>Proof. \fi{} We first establish the
equivalence of claims 2 and 3. If claim 2 holds, then claim 3 follows
easily from the resulting uniform cell probabilities:
\(P(V_1= v_1,\ldots,V_m= v_m, H_1= h_1,\ldots,H_n=h_n|\bm{\theta})=2^{-mn}\)
for any \(v_i,h_j\in\{\pm 1\}\), \(i=1,\ldots,m\), \(j=1,\ldots,n\).

Now suppose claim 3 holds. Pick \(\ell \in \{1,\ldots,m\}\). Then, by
the RBM\((m,n,\bm{\theta})\) probability structure and the iid Bernoulli
assumption, we have the conditional probability
\begin{eqnarray*}
\frac{1}{2} &=& P\left(V_\ell = v_{\ell}\Big| \bigcap_{i=1 \atop i \neq \ell}^m \{V_i=v_i\} \cap \bigcap_{j=1}^n \{H_j=h_j\}\right) \\
&\equiv &\frac{\exp\left( v_\ell\theta_\ell +  v_\ell\sum_{j=1}^n  h_j \theta_{\ell j}\right)}{\exp\left(
\theta_\ell + \sum_{j=1}^n   h_j\theta_{\ell j} \right) +\exp\left(-\theta_\ell  -\sum_{j=1}^n  h_j\theta_{\ell j}\right) }
\end{eqnarray*}
for any subsets \(\{v_i\}_{i=1}^m,\{h_j\}_{j=1}^n \subset\{\pm 1\}\).
Equating conditional probabilities when \(v_\ell=1\) or when
\(v_\ell=-1\), we have
\(\theta_\ell = - \sum_{j=1}^m h_j \theta_{\ell j}\) for any given
\(\{h_j\}_{j=1}^n \subset \{\pm 1\}\). Taking all \(h_j=1\) or all
\(h_j=-1\) for \(1 \leq j \leq n\) yields \(\theta_\ell=0\). This result
further implies \(\theta_{\ell 1} = - \sum_{j=2}^n h_j \theta_{\ell j}\)
for any given \(\{h_j\}_{j=2}^n \subset \{\pm 1\}\), so that
\(\theta_{\ell 1}=0\) must hold. Iterating this argument, sequentially
for each \(j=2,\ldots,m\), shows \(\theta_{\ell j}=0\) for any
\(1 \leq j \leq n\) in addition to \(\theta_{v_\ell}=0\). As
\(\ell\in\{1,\ldots,m\}\) was arbitrary, we have \(\theta_{v_i}=0\) and
\(\theta_{i j}=0\) for any \(1 \leq i \leq m\), \(1 \leq j \leq n\).

Now pick \(k \in \{1,\ldots,n\}\) so that, by the
RBM\((m,n,\bm{\theta})\) probability structure and the iid assumption,
we have the conditional probability \[
 \frac{1}{2} = P\left(H_k = h_k \Big| \bigcap_{j=1 \atop j \neq k}^n \{H_j=h_j\}\right) \equiv \frac{\exp(h_k \theta_{h_k})}{\exp(\theta_{h_k}) + \exp(-\theta_{h_k})}
\] for any \(\{h_j\}_{j=1}^n \subset \{\pm 1\}\). Considering \(h_k=1\)
or \(h_k=-1\), we conclude that \(\theta_{h_k} = -\theta_{h_k}\) or
\(\theta_{h_k}=0\). As \(k \in \{1,\ldots,n\}\) was arbitrary, we have
now established claim 2 from claim 3.

To establish the equivalence of claims 1 and 2 in Theorem \ref{thm:two}
under the RBM\((m,n,\bm{\theta})\) model, we note that claim 2 again
implies claim 3, which then easily implies claim 1. By Theorem
\ref{thm:one}, if claim 1 holds then the only possibility for this is
\(\bm{\theta} = \bm{0} \in \mathbb{R}^{m+n+mn}\).
\EndKnitrBlock{proof}

\section{Flexibility of the RBM model}\label{appendix-rbm2}

In an RBM model with enough hidden variables, parameter values may be
chosen to match any given cell probabilities with arbitrary closeness.
Additionally, when one or more of the cell probabilities (\eqref{eq:0})
are zero, the corresponding RBM probabilities may never be identically
zero (due to exponential terms in the model) but parameters can be still
selected to make the appropriate RBM cell probabilities arbitrarily
small. We show this for a model with two visible variables \((V_1,V_2)\)
and one hidden \(H_1\).

To demonstrate, we assume \(p_{(-1,-1)}>0\) (without loss of generality)
in the specified cell probabilities (\eqref{eq:0}) and replace parameters
\(\theta_{11},\theta_{21}\) with
\(\Delta_1 \equiv \theta_{11} +\theta_{21}\) and
\(\Delta_2 \equiv \theta_{11} -\theta_{21}\). We may then prescribe
values of
\(\theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \Delta_1,\Delta_2\) so that
the model probability ratio \[
P(V_1=v_1,V_2=v_2|\theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \Delta_1,\Delta_2)/  P(V_1=-1,V_2=-1| \theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \Delta_1,\Delta_2)
\] matches the corresponding ratio \(p_{(v_1,v_2)}/p_{(-1,-1)}\) over
three values of \((v_1,v_2) = (1,-1),(-1,1),(1,1)\).

For instance, assuming the cell probabilities from (\eqref{eq:0}) are all
positive, these probabilities can be exactly reproduced by choosing
\begin{eqnarray*}
\theta_{v_1} &=& \frac{1}{2} \log\left( \frac{p_{(1,-1)}}{p_{(-1,-1)}} \frac{\exp(\theta_{h_1}-\Delta_1) + \exp(-\theta_{h_1}+\Delta_1)}{\exp(\theta_{h_1}+\Delta_2) + \exp(-\theta_{h_1}-\Delta_2)}  \right),\\ 
\theta_{v_2}&= &\frac{1}{2} \log\left( \frac{p_{(-1,1)}}{p_{(-1,-1)}} \frac{\exp(\theta_{h_1}-\Delta_1) + \exp(-\theta_{h_1}+\Delta_1)}{\exp(\theta_{h_1}-\Delta_2) + \exp(-\theta_{h_1}+\Delta_2)}  \right)
\end{eqnarray*}
and selecting \(\theta_{h_1}, \Delta_1,\Delta_2\) to solve
\begin{equation}
\label{eq:1}
\frac{p_{(1,1)}p_{(-1,-1)}}{p_{(-1,1)}p_{(1,-1)}} = \frac{\ell(|\theta_{h_1}|) + \ell(|\Delta_1|) }{\ell(|\theta_{h_1}|) + \ell(|\Delta_2|)},
\end{equation}
based on a monotonically increasing function
\(\ell(x)\equiv \exp(-2x)+\exp(2x)\), \(x \geq 0\). If
\([p_{(1,1)}p_{(-1,-1)}]/ [p_{(-1,1)}p_{(1,-1)}] \geq 1\), one can pick
any values for \(\theta_{h_1}, \Delta_2\in \mathbb{R}\) and solve
(\eqref{eq:1}) for \(|\Delta_1|\); likewise, when
\([p_{(1,1)}p_{(-1,-1)}]/ [p_{(-1,1)}p_{(1,-1)}] < 1\) in
(\eqref{eq:1})\}), one may solve for \(|\Delta_2|\) upon choosing any
values for \(\theta_{h_1}, \Delta_1\in \mathbb{R}\).

Alternatively, if exactly one specified cell probability in (\eqref{eq:0})
is zero, say \(p_{(1,1)}\) (without loss of generality), we can select
parameters \(\theta_{v_1},\theta_{v_2}\) as above based on a sequence
\((\theta_{h_1}, \Delta_{1}, \Delta_{2}) \equiv (\theta_{h_1}^{(m)}, \Delta_{1}^{(m)}, \Delta_{2}^{(m)})\),
\(m\in\{1,2,\ldots,\}\) of the remaining parameter values such that
\(\lim_{m\to \infty}|\Delta_{1}^{(m)}| = \infty\) and
\(\lim_{m\to \infty} (|\theta_{h_1}^{(m)}| + |\Delta_{2}^{(m)}|)/|\Delta_{1}^{(m)}|=0\)
hold. This guarantees that the resulting RBM model matches the given
cell probabilities (\eqref{eq:0}) in the limit:
\begin{equation}
\label{eq:2}
\lim_{m\to \infty}P(V_1=v_1,V_2=v_2| \theta_{v_1}, \theta_{v_2}, \theta_{h_1},\Delta_1,\Delta_2) = p_{(v_1,v_2)},\quad (v_1,v_2)\in\{(\pm 1,\pm 1)\}.
\end{equation}
If exactly two specified probabilities in (\eqref{eq:0}) are zero, say
\(p_{(1,1)}\) and \(p_{(-1,1)}\) (without loss of generality), then a
limit approximation as in (\eqref{eq:2}) follows by picking
\(\theta_{v_1}\) as above based on any choices of
\((\theta_{h_1}, \Delta_1,\Delta_2)\) and choosing a sequence of
\(\theta_{v_2}\equiv \theta_{v_2}^{(m)}\) values for which
\(\theta_{v_2}^{({m})} \rightarrow -\infty\).

\section*{References}\label{references}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\hypertarget{ref-carreira2005contrastive}{}
Carreira-Perpinan, Miguel A, and Geoffrey E Hinton. 2005. ``On
Contrastive Divergence Learning.'' In \emph{Aistats}, 10:33--40.

\hypertarget{ref-cho2012tikhonov}{}
Cho, KyungHyun, Alexander Ilin, and Tapani Raiko. 2012. ``Tikhonov-Type
Regularization for Restricted Boltzmann Machines.'' \emph{Artificial
Neural Networks and Machine Learning--ICANN 2012}. Springer, 81--88.

\hypertarget{ref-fisher1922mathematical}{}
Fisher, R. A. 1922. ``On the Mathematical Foundations of Theoretical
Statistics.'' \emph{Philosophical Transactions of the Royal Society of
London A: Mathematical, Physical and Engineering Sciences} 222
(594-604). The Royal Society: 309--68.

\hypertarget{ref-box1967discrimination}{}
G. E. P. Box, W. J. Hill. 1967. ``Discrimination Among Mechanistic
Models.'' \emph{Technometrics} 9 (1). {[}Taylor \& Francis, Ltd.,
American Statistical Association, American Society for Quality{]}:
57--71.

\hypertarget{ref-gelman2011inference}{}
Gelman, Andrew, Kenneth Shirley, and others. 2011. ``Inference from
Simulations and Monitoring Convergence.'' \emph{Handbook of Markov Chain
Monte Carlo}, 163--74.

\hypertarget{ref-handcock2003assessing}{}
Handcock, Mark S. 2003. ``Assessing Degeneracy in Statistical Models of
Social Networks.'' Center for Statistics; the Social Sciences,
University of Washington. \url{http://www.csss.washington.edu/}.

\hypertarget{ref-hinton2010practical}{}
Hinton, Geoffrey. 2010. ``A Practical Guide to Training Restricted
Boltzmann Machines.'' \emph{Momentum} 9 (1): 926.

\hypertarget{ref-hinton2002training}{}
Hinton, Geoffrey E. 2002. ``Training Products of Experts by Minimizing
Contrastive Divergence.'' \emph{Neural Computation} 14 (8). MIT Press:
1771--1800.

\hypertarget{ref-hinton2006fast}{}
Hinton, Geoffrey E, Simon Osindero, and Yee-Whye Teh. 2006. ``A Fast
Learning Algorithm for Deep Belief Nets.'' \emph{Neural Computation} 18
(7). MIT Press: 1527--54.

\hypertarget{ref-kaiser2007statistical}{}
Kaiser, Mark S. 2007. ``Statistical Dependence in Markov Random Field
Models.'' \emph{Statistics Preprints} Paper 57. Digital Repository @
Iowa State University.
\url{http://lib.dr.iastate.edu/stat_las_preprints/57/}.

\hypertarget{ref-kaplan2017note}{}
Kaplan, Andee, Daniel Nordman, and Stephen Vardeman. 2017. ``A Note on
the Instability and Degeneracy of Deep Learning Models.'' \emph{In
Preparation}.

\hypertarget{ref-larochelle2008classification}{}
Larochelle, Hugo, and Yoshua Bengio. 2008. ``Classification Using
Discriminative Restricted Boltzmann Machines.'' In \emph{Proceedings of
the 25th International Conference on Machine Learning}, 536--43. ACM.

\hypertarget{ref-le2008representational}{}
Le Roux, Nicolas, and Yoshua Bengio. 2008. ``Representational Power of
Restricted Boltzmann Machines and Deep Belief Networks.'' \emph{Neural
Computation} 20 (6). MIT Press: 1631--49.

\hypertarget{ref-lehmann1990model}{}
Lehmann, E. L. 1990. ``Model Specification: The Views of Fisher and
Neyman, and Later Developments.'' \emph{Statistical Science} 5 (2).
Institute of Mathematical Statistics: 160--68.

\hypertarget{ref-li2014biclustering}{}
Li, Jing. 2014. ``Biclustering Methods and a Bayesian Approach to
Fitting Boltzmann Machines in Statistical Learning.'' PhD thesis, Iowa
State University; Graduate Theses; Dissertations.
\url{http://lib.dr.iastate.edu/etd/14173/}.

\hypertarget{ref-montufar2011refinements}{}
Montufar, Guido, and Nihat Ay. 2011. ``Refinements of Universal
Approximation Results for Deep Belief Networks and Restricted Boltzmann
Machines.'' \emph{Neural Computation} 23 (5). MIT Press: 1306--19.

\hypertarget{ref-montufar2011expressive}{}
Montúfar, Guido F, Johannes Rauh, and Nihat Ay. 2011. ``Expressive Power
and Approximation Errors of Restricted Boltzmann Machines.'' In
\emph{Advances in Neural Information Processing Systems}, 415--23. NIPS.

\hypertarget{ref-nguyen2014deep}{}
Nguyen, Anh Mai, Jason Yosinski, and Jeff Clune. 2014. ``Deep Neural
Networks Are Easily Fooled: High Confidence Predictions for
Unrecognizable Images.'' \emph{arXiv Preprint arXiv:1412.1897}.
\url{http://arxiv.org/abs/1412.1897}.

\hypertarget{ref-salakhutdinov2012efficient}{}
Salakhutdinov, Ruslan, and Geoffrey Hinton. 2012. ``An Efficient
Learning Procedure for Deep Boltzmann Machines.'' \emph{Neural
Computation} 24 (8). MIT Press: 1967--2006.

\hypertarget{ref-salakhutdinov2009deep}{}
Salakhutdinov, Ruslan, and Geoffrey E Hinton. 2009. ``Deep Boltzmann
Machines.'' In \emph{International Conference on Artificial Intelligence
and Statistics}, 448--55. AI \& Statistics.

\hypertarget{ref-schweinberger2011instability}{}
Schweinberger, Michael. 2011. ``Instability, Sensitivity, and Degeneracy
of Discrete Exponential Families.'' \emph{Journal of the American
Statistical Association} 106 (496). Taylor \& Francis: 1361--70.

\hypertarget{ref-smolensky1986information}{}
Smolensky, Paul. 1986. ``Information Processing in Dynamical Systems:
Foundations of Harmony Theory.'' DTIC Document.

\hypertarget{ref-srivastava2012multimodal}{}
Srivastava, Nitish, and Ruslan R Salakhutdinov. 2012. ``Multimodal
Learning with Deep Boltzmann Machines.'' In \emph{Advances in Neural
Information Processing Systems}, 2222--30. NIPS.

\hypertarget{ref-srivastava2013modeling}{}
Srivastava, Nitish, Ruslan R Salakhutdinov, and Geoffrey E Hinton. 2013.
``Modeling Documents with Deep Boltzmann Machines.'' \emph{arXiv
Preprint arXiv:1309.6865}. \url{http://arxiv.org/abs/1309.6865}.

\hypertarget{ref-szegedy2013intriguing}{}
Szegedy, Christian, Wojciech Zaremba, Ilya Sutskever, Joan Bruna,
Dumitru Erhan, Ian J. Goodfellow, and Rob Fergus. 2013. ``Intriguing
Properties of Neural Networks.'' \emph{arXiv Preprint arXiv:1312.6199}.
\url{http://arxiv.org/abs/1312.6199}.

\hypertarget{ref-tieleman2008training}{}
Tieleman, Tijmen. 2008. ``Training Restricted Boltzmann Machines Using
Approximations to the Likelihood Gradient.'' In \emph{Proceedings of the
25th International Conference on Machine Learning}, 1064--71. ACM.

\hypertarget{ref-zhou2014some}{}
Zhou, Wen. 2014. ``Some Bayesian and Multivariate Analysis Methods in
Statistical Machine Learning and Applications.'' PhD thesis, Iowa State
University; Graduate Theses; Dissertations.
\url{http://lib.dr.iastate.edu/etd/13816/}.

\end{document}

### 2017-07-08 writing/paper-rbm/index.Rmd
paper to bookdown

---
title: "Properties and Bayes fitting of restricted Boltzmann machines"
author: 
  - name: Andee Kaplan
    university: Iowa State University
    department: Department of Statistics
  - name: Daniel Nordman
    university: Iowa State University
    department: Department of Statistics
  - name: Stephen Vardeman
    university: Iowa State University
    department: |
      | Department of Statistics and Department of Industrial and
      | Manufacturing Systems Engineering
site: bookdown::bookdown_site
output: 
  bookdown::pdf_book:
    includes:
      in_header: ../resources/latex/header.tex
    toc: no
bibliography: [../resources/latex/refs_rbm.bib]
github-repo: andeek/rbm
abstract: |
   A restricted Boltzmann machine (RBM) is an undirected graphical model constructed for discrete or continuous random variables, with two layers, one hidden and one visible, and no conditional dependency within a layer. In recent years, RBMs have risen to prominence due to their connection to deep learning. By treating a hidden layer of one RBM as the visible layer in a second RBM, a deep architecture can be created. RBMs are thought to thereby have the ability to encode very complex and rich structures in data, making them attractive for supervised learning. However, the generative behavior of RBMs is largely unexplored. In this paper, we discuss the relationship between RBM parameter specification in the binary case and model properties such as degeneracy, instability and uninterpretability. We also describe the difficulties that arise in likelihood-based and Bayes fitting of such (highly flexible) models, especially as Gibbs sampling (quasi-Bayes) methods are often advocated for the RBM model structure.
keywords: Degeneracy, Instability, Classification, Deep Learning, Graphical Models
blind: false
---

```{r setup, include=FALSE}
# https://github.com/rstudio/bookdown/pull/374
fix_envs <- function(x) {
  beg_reg <- '^\\s*\\\\begin\\{.*\\}'
  end_reg <- '^\\s*\\\\end\\{.*\\}'
  i3 = if (length(i1 <- grep(beg_reg, x))) (i1 - 1)[grepl("^\\s*$", x[i1 - 1])]
  i3 = c(i3, if (length(i2 <- grep(end_reg, x))) (i2 + 1)[grepl("^\\s*$", x[i2 + 1])])
  if (length(i3)) x = x[-i3]
  x
}

options(
  dplyr.print_min = 6, dplyr.print_max = 6, width = 70,
  digits = 3,
  bookdown.post.latex = fix_envs
)
knitr::opts_chunk$set(
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  cache = TRUE,
  echo = FALSE
)

ggplot2::theme_set(ggplot2::theme_bw(base_family="serif"))

library(knitr)
library(ggplot2)
library(dplyr)
library(tidyr)
library(scales)
library(mvtnorm)
library(rootSolve)
library(agridat)
library(xtable)
```

### 2017-07-09 writing/paper-rbm/chapters/rbm.Rmd
small edits

@@
 
-The concepts of weight-decay (penalization) and sparsity (regularization) are suggested in the literature as practical remedies for both over-fitting and poor mixing in the Markov chain during fitting [@hinton2010practical; @tieleman2008training; @cho2012tikhonov].
-Both $L_1$ and $L_2$ type penalties are mentioned to achieve weight-decay however it is unclear the benefits of these particular forms in different situations.
-Additionally, it is not clear the degree to which regularization and penalization are used by practitioners because they are not (by default) a part of the accepted fitted methodology [@hinton2002training; @carreira2005contrastive].
-In this paper, we attempt to address the concerns with overfitting and poor mixing via specification of a Bayesian prior in Section \@ref(bayesian-model-fitting).
+Notions of weight-decay (penalization) and sparsity (regularization) have been suggested in the RBM literature as practical remedies for both over-fitting and poor mixing in the Markov chain during fitting [@hinton2010practical; @tieleman2008training; @cho2012tikhonov].
+Both $L_1$ and $L_2$ type penalties are mentioned to achieve weight-decay, though the benefits of these particular forms are unknown in different situations.
+However,  the degree to which regularization and penalization are used by practitioners is not clear because these concepts are not (by default) a part of the accepted fitted methodology [@hinton2002training; @carreira2005contrastive].
+In this paper, we attempt to address the concerns with overfitting and poor mixing in an alternative, and perhaps more transparent manner, via specification of a Bayesian prior in Section \@ref(bayesian-model-fitting).
 

### 2017-07-09 writing/paper-rbm/docs/paper.tex
small edits

@@
 
-The concepts of weight-decay (penalization) and sparsity
-(regularization) are suggested in the literature as practical remedies
-for both over-fitting and poor mixing in the Markov chain during fitting
-(Hinton \protect\hyperlink{ref-hinton2010practical}{2010}; Tieleman
+Notions of weight-decay (penalization) and sparsity (regularization)
+have been suggested in the RBM literature as practical remedies for both
+over-fitting and poor mixing in the Markov chain during fitting (Hinton
+\protect\hyperlink{ref-hinton2010practical}{2010}; Tieleman
 \protect\hyperlink{ref-tieleman2008training}{2008}; Cho, Ilin, and Raiko
 \protect\hyperlink{ref-cho2012tikhonov}{2012}). Both \(L_1\) and \(L_2\)
-type penalties are mentioned to achieve weight-decay however it is
-unclear the benefits of these particular forms in different situations.
-Additionally, it is not clear the degree to which regularization and
-penalization are used by practitioners because they are not (by default)
-a part of the accepted fitted methodology (Hinton
+type penalties are mentioned to achieve weight-decay, though the
+benefits of these particular forms are unknown in different situations.
+However, the degree to which regularization and penalization are used by
+practitioners is not clear because these concepts are not (by default) a
+part of the accepted fitted methodology (Hinton
 \protect\hyperlink{ref-hinton2002training}{2002}; Carreira-Perpinan and
@@
 paper, we attempt to address the concerns with overfitting and poor
-mixing via specification of a Bayesian prior in Section
+mixing in an alternative, and perhaps more transparent manner, via
+specification of a Bayesian prior in Section
 \ref{bayesian-model-fitting}.

### 2017-07-11 writing/paper-note/chapters/instab.Rmd
edits from Dr. Vardeman

@@
 Let $\boldsymbol \theta_N = (\boldsymbol \alpha, \boldsymbol \beta,\Gamma) \in \Theta_N \subset \mathbb{R}^{q(N)}$ with $q(N) = N + N_H + N*N_H$ denote the vector of parameters for the RBM.
-The probability mass function for the visible variables $X_1, \dots, X_N$ follows from marginalizing this joint specification:
+The probability mass function for the visible variables $X_1, \dots, X_N$ follows from marginalizing the joint specification
 $$
@@
 $$
-Again like the RBM case, the DBM model specification examples a FSFS model.
+Again like the RBM case, the DBM model specification is an example of a FSFS model.
 
@@
 
-The main idea is that, in formulating FSFS models for potentially increasing numbers of variables (i.e., for $N \rightarrow \infty$), the ratio (\@ref(eq:elpr)) should remain bounded to better ensure model stability, requiring that the largest probability possible under $P_{\theta_N}$ maintain a fixed order of magnitude relative to the smallest probability allowed under the same model.
+The main idea is that, in formulating FSFS models for potentially increasing numbers of variables (i.e., for $N \rightarrow \infty$), the ratio (\@ref(eq:elpr)) should remain bounded to ensure model stability, requiring that the largest probability possible under $P_{\theta_N}$ maintain a fixed order of magnitude relative to the smallest probability allowed under the same model.
 Specifically, the log of the ratio should grow at mostly linearly with the sample size $N$.
@@
 
-S-unstable FSFS model sequences may be undesirable for several reasons.
-One is that small changes in data can lead to overly-sensitive changes in probability.
-Consider, for example, the quantity given by
+S-unstable FSFS model sequences may be undesirable for several reasons. One is that small changes in data can lead to very large changes in probability. Consider, for example, the quantity given by
 $$
@@
 For a large class of models that covers a broad range of applications (including "deep learning"), we have developed a formal definition of instability in model probability structure and elucidated multiple consequences of instability.
-We have shown for FSFS models that instability manifests through small changes in data leading to potentially large changes in probability as well as the potential to place all probability on certain modal subsections of the sample space, which could be potentially small.
+We have shown for FSFS models that instability manifests through small changes in data leading to potentially large changes in probability as well as the potential to place all probability on certain modal subsections of the sample space, which potentially could be small.
 The FSFS model class is quite broad and, particularly in developing FSFS models for large data sets, some caution should be used in parameter specification to control effects of model instability.

### 2017-07-11 writing/paper-note/docs/note.tex
edits from Dr. Vardeman

@@
 RBM. The probability mass function for the visible variables
-\(X_1, \dots, X_N\) follows from marginalizing this joint specification:
+\(X_1, \dots, X_N\) follows from marginalizing the joint specification
 \[
@@
 P_{\boldsymbol \theta_N} (\boldsymbol x) = \sum\limits_{(\boldsymbol h^{(1)}, \dots, \boldsymbol h^{(M)}) \in \mathcal{X}^{N_{(H,1)} + \cdots + N_{(H,M)}}} P_{\boldsymbol \theta_N} (\tilde{\boldsymbol x}), \qquad \boldsymbol x \in \mathcal{X}^N
-\] Again like the RBM case, the DBM model specification examples a FSFS
-model.
+\] Again like the RBM case, the DBM model specification is an example of
+a FSFS model.
 
@@
 increasing numbers of variables (i.e., for \(N \rightarrow \infty\)),
-the ratio (\eqref{eq:elpr}) should remain bounded to better ensure model
+the ratio (\eqref{eq:elpr}) should remain bounded to ensure model
 stability, requiring that the largest probability possible under
@@
 S-unstable FSFS model sequences may be undesirable for several reasons.
-One is that small changes in data can lead to overly-sensitive changes
-in probability. Consider, for example, the quantity given by \[
+One is that small changes in data can lead to very large changes in
+probability. Consider, for example, the quantity given by \[
 \Delta(\boldsymbol \theta_N) \equiv \max \left\{\log \frac{P_{\boldsymbol \theta_N}(\boldsymbol x)}{P_{\boldsymbol \theta_N}(\boldsymbol x^*)} : \boldsymbol x \text{ }\& \text{ } \boldsymbol x^* \in \mathcal{X}^N \text{ differ in exactly one component}\right\},
@@
 place all probability on certain modal subsections of the sample space,
-which could be potentially small. The FSFS model class is quite broad
+which potentially could be small. The FSFS model class is quite broad
 and, particularly in developing FSFS models for large data sets, some

### 2017-07-11 writing/paper-rbm/chapters/rbm.Rmd
edits from Dr. Vardeman

@@
 For a RBM model with enough hidden variables, it has been shown that any distribution for the visibles can be approximated arbitrarily well [@le2008representational; @montufar2011refinements; and @montufar2011expressive].
-However, for cell data, the empirical distribution of an observed training set of visible variables provides a highest likelihood benchmark, before any parametric model class is even introduced and applied to obtain a refinement of model fit.
+However, for binary data, the empirical distribution of an observed training set of visible variables provides a highest likelihood benchmark, before any parametric model class is even introduced and applied to obtain a refinement of model fit.
 As a consequence, we find that any fully principled fitting method based on the likelihood for a RBM with enough hidden variables will seek to reproduce the (discrete) empirical distribution of a training set.
 This aspect can be undesirable, and perhaps even unexpected compared to most modeling scenarios, in that no "smoothed distribution" may result when fitting a RBM model of sufficient size with a rigorous likelihood-based method.
-We are therefore led to be skeptical that models that involve these structures (like a deep Boltzmann machine) can achieve useful prediction or inference in a principled way without intentionally limiting the flexibility of the fitted model.
+We are therefore led to be skeptical that models that involve these structures (like deep Boltzmann machines) can achieve useful prediction or inference in a principled way without intentionally limiting the flexibility of the fitted model.
 
@@
 Both $L_1$ and $L_2$ type penalties are mentioned to achieve weight-decay, though the benefits of these particular forms are unknown in different situations.
-However,  the degree to which regularization and penalization are used by practitioners is not clear because these concepts are not (by default) a part of the accepted fitted methodology [@hinton2002training; @carreira2005contrastive].
+However, the degree to which regularization and penalization are used by practitioners is not clear because these concepts are not (by default) a part of the accepted fitting methodology [@hinton2002training; @carreira2005contrastive].
 In this paper, we attempt to address the concerns with overfitting and poor mixing in an alternative, and perhaps more transparent manner, via specification of a Bayesian prior in Section \@ref(bayesian-model-fitting).
@@
 Section \@ref(explorations-of-model-properties-through-simulation) details our explorations into the model behavior and potential propriety issues with the RBM class.
-We examine three Bayesian fitting techniques intended to avoid model impropriety in Section \@ref(model-fitting) and conclude with a discussion in Section \@ref(discussion).
+We examine three Bayesian fitting techniques intended to avoid model impropriety issues raised in Section \@ref(model-fitting) and conclude with a discussion in Section \@ref(discussion).
 Appendix \@ref(appendix-rbm) provides proofs for results on RBM parameterizations and data codings described in Section \@ref(data-encoding).
@@
 While applications of the RBM have claimed some successes in classification problems, it is unclear if the model class allows one to go beyond fitting to other statistical matters of importance in using the models, such as quantification of the uncertainty of estimation and the formulation of predictive distributions.
-These are closely tied to the probability properties of the RBM model for explaining data generation and it is hoped that the current exposition contributes to a better understanding in this direction.
+These are closely tied to the probability properties of the RBM model for explaining data generation and the current exposition intends to contribute to a better understanding in this direction.
 
@@
 RBMs have risen to prominence in recent years due to their connection to deep learning [see @hinton2006fast; @salakhutdinov2012efficient; @srivastava2013modeling for examples].
-By stacking multiple layers of RBMs in a deep architecture, proponents of the models claim to produce the ability to learn "internal representations that become increasingly complex, which is considered to be a promising way of solving object and speech recognition problems" [@salakhutdinov2009deep, pp.
+By stacking multiple layers of RBMs in a deep architecture, proponents of the model are claimed to produce the ability to learn "internal representations that become increasingly complex, which is considered to be a promising way of solving object and speech recognition problems" [@salakhutdinov2009deep, pp.
 450].
@@
 ## Degeneracy, instability, and uninterpretability
-The highly flexible nature of a RBM (having as it does $\nh + \nv + \nh*\nv$ parameters) creates at least three kinds of potential issues in model impropriety that we will call *degeneracy*, *instability*, and *uninterpretability*.
-In this section we define these characteristics, consider how to quantify these in a RBM, and point out relationships among them.
+The highly flexible nature of a RBM (having as it does $\nh + \nv + \nh*\nv$ parameters) creates at least three kinds of potential issues of model impropriety that we will call *degeneracy*, *instability*, and *uninterpretability*.
+In this section we define these, consider how to quantify them in a RBM, and point out relationships among them.
 
@@
 In Random Graph Model theory, *model degeneracy* means there is a disproportionate amount of probability placed on only a few elements of the sample space, $\mathcal{X}$, by the model [@handcock2003assessing].
-For random graph models, $\mathcal{X}$ denotes all possible graphs that can be constructed from a set of nodes and an exponentially parameterized random graph model has a distribution of the form
+For random graph models, $\mathcal{X}$ denotes all possible graphs that can be constructed from a set of nodes and an exponentially parameterized random graph model has a distribution with a probability mass function of the form
 \begin{align*}
@@
 Considering exponential families of distributions, @schweinberger2011instability introduced a concept of model deficiency related to *instability*.
-Instability can be roughly thought of as excessive sensitivity in the model, where small changes in the components of potential data outcomes, $\boldsymbol x$, may lead to substantial changes in the probability function $f_{\boldsymbol \theta}(\boldsymbol x)$.
+Instability can be roughly thought of as excessive sensitivity in the model, where small changes in the components of potential data outcomes, $\boldsymbol x$, may lead to substantial changes in the probability mass function $f_{\boldsymbol \theta}(\boldsymbol x)$.
 Furthermore, model instability can be viewed on a spectrum of potential sensitivity in probability structure, with model degeneracy included as a extreme or limiting case of instability.
@@
 
-In formulating a RBM model for a potentially large number of visibles (i.e., as $\nv \rightarrow \infty$), we will say that the ratio (\@ref(eq:elpr)) needs to stay bounded for the sequence of RBM models to be stable.
+In formulating a RBM model for a potentially large number of visibles (i.e., as $\nv \rightarrow \infty$), we will say that the ratio (\@ref(eq:elpr)) needs to stay bounded for a sequence of RBM models to be stable.
 That is, we make the following convention.
@@
 ```
-In other words, the RBM model sequence is unstable if, given any $c > 0$, there exists an integer $n_c > 0$ so that $\frac{1}{\nv}\text{ELPR}(\boldsymbol \theta_\nv) > c$ for all $\nv \ge n_c$.
+In other words, a RBM model sequence is unstable if, given any $c > 0$, there exists an integer $n_c > 0$ so that $\frac{1}{\nv}\text{ELPR}(\boldsymbol \theta_\nv) > c$ for all $\nv \ge n_c$.
 This definition of *S-unstable* is a generalization or re-interpretation of the "unstable" concept of @schweinberger2011instability in that here RBM models for visibles $(v_1, \dots, v_\nv)$ do not form an exponential family and the dimensionality of $\boldsymbol \theta_\nv$ is not fixed, but rather grows with $\nv$.
@@
 In other words, if the probability ratio (\@ref(eq:elpr)) is too large, then a RBM model sequence will exhibit large probability shifts for very small changes in data configurations (i.e., will exhibit instability).
-Such instability can be a concern in a model for the similar reasons to degeneracy: as outcomes vary over the sample space, the geography of probabilities is extremely rugged, with deep pits following sharp mountains.
+Such instability can be a concern in a model for reasons similar to those related to degeneracy: as outcomes vary over the sample space, the geography of probabilities is extremely rugged, with deep pits following sharp mountains.
 Recall the applied example of RBM models as a means to classify images.
@@
 
-We next explore and numerically explain the relationship between values of $\boldsymbol \theta$ and the three notions of model impropriety, near-degeneracy, instability, and uninterpretability, for RBM models of varying sizes.
+We next explore and numerically explain the relationship between values of $\boldsymbol \theta$ and the three notions of model impropriety (near-degeneracy, instability, and uninterpretability), for RBM models of varying sizes.
 
@@
 From these three figures, it is clear that all three measures of model impropriety show higher values for larger magnitudes of the parameter vectors, supporting the RBM model properties developed in Section \@ref(background).
-As a compounding issue, these figures show that, as model grow in size, it becomes easier for more parameter configurations to push RBM models into near-degeneracy, instability and uninterpretability.
+As a compounding issue, these figures show that, as models grow in size, it becomes easier for more parameter configurations to push RBM models into near-degeneracy, instability and uninterpretability.
 Additionally, since there are $\nh*\nv$ interaction terms in $\boldsymbol \theta$ versus only $\nh + \nv$ main effect terms, for large models there are many more interaction parameters than main effects in the models.
@@
   geom_abline(aes(intercept = 0, slope = 1), alpha = .5, lty = 2) +
-  scale_fill_gradient(expression(frac(bar(ELPR(theta)), V)), low = "yellow", high = "red") +
+  scale_fill_gradient(expression(paste("Mean ", frac(ELPR(theta), V))), low = "yellow", high = "red") +
   facet_grid(Visibles~Hiddens, labeller = label_parsed) +
@@
 ```
-These manageable examples indicate that RBMs are near-degenerate, unstable, and uninterpretable for large portions of the parameter space with large $\|\boldsymbol \theta\|$.
+These manageable examples indicate that RBMs are near-degenerate, unstable, and uninterpretable for large portions of the parameter space having large $\|\boldsymbol \theta\|$.
 These problematic aspects require serious consideration when using RBM models, on top of the additional matter of principled/rigorous fitting of RBM models.
@@
 Typically, fitting a RBM via maximum likelihood (ML) methods will be infeasible due mainly to the intractability of the normalizing term $\gamma(\boldsymbol \theta)$ in a model (\@ref(eq:pmf)) of any realistic size.
-Ad hoc methods are used instead, which aim to avoid this problem by using stochastic ML approximations that employ a small number of MCMC draws (i.e., contrastive divergence, [@hinton2002training]).
+Ad hoc methods are often recommended instead, which aim to avoid this problem by using stochastic ML approximations that employ a small number of MCMC draws (i.e., contrastive divergence, [@hinton2002training]).
 
@@
 And, when one or more of the cell probabilities (\@ref(eq:0)) are zero, the corresponding RBM probabilities $P(V_1=v_1,V_2=_v2| \theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21})$ may never be identically zero (due to exponential terms in the model) but parameters can be still selected to make the appropriate RBM cell probabilities arbitrarily small.
-Furthermore, as development in the appendix shows, only it is possible to approximate any distribution on the visibles arbitrarily well [cf.
+Furthermore, as development in the appendix shows, not only is it possible to approximate any distribution on the visibles arbitrarily well [cf.
 @montufar2011refinements], but dramatically different parameter settings can induce the same RBM model (beyond mere symmetries in parameterization).
@@
 These range from  $`r round(min(M_eff_tmp[, "BwTNML"]), 2)`$ to $`r round(max(M_eff_tmp[, "BwTNML"]), 2)`$ for BwTNML, while BwTNLV only yields between $`r round(min(M_eff_tmp[, "BwTNLV"]), 2)`$ and $`r round(max(M_eff_tmp[, "BwTNLV"]), 2)`$ effective draws.
-Thus, BwTNLV would require at least $`r round(min(M_eff_tmp[, "BwTNML"])/min(M_eff_tmp[, "BwTNLV"]), 1)`$ times as many iterations to be run of the MCMC chain in order to achieve the same amount of effective information about the posterior distribution.
+Thus, BwTNLV would require at least $`r round(min(M_eff_tmp[, "BwTNML"])/min(M_eff_tmp[, "BwTNLV"]), 1)`$ times as many iterations of the MCMC chain to be run in order to achieve the same amount of effective information about the posterior distribution.
 For this reason, consistent with the ACF results in Figure \@ref(fig:acf), BwTNLV does not seem to be an effective method for fitting the RBM, though computing resources can hinder use of the alternative BwTNML involving marginalization of hidden variables.
@@
 Ultimately, it is not clear that RBM models are useful as generative models.
-Furthermore, without the appropriate generative behavior in the RBM model, the ability to quantify uncertainty in the estimated model parameters becomes impossible and the model loses useful application to prediction problems (e.g., realistic predictive distributions).
-In the case of classification, predictive distributions not be the ultimate goal, but when S-instability is present, small (imperceptible) differences in the data may lead to greatly different probabilities and thus greatly different classifications.
+Furthermore, without the appropriate generative behavior in the RBM model, quantifying uncertainty in the estimated model parameters becomes impossible and the model loses useful application to prediction problems (e.g., realistic predictive distributions).
+In the case of classification, predictive distributions may not be the ultimate goal, but when S-instability is present, small (imperceptible) differences in the data may lead to greatly different probabilities and thus greatly different classifications.
 For these reasons, we are skeptical about RBMs as data generative tools.

### 2017-07-11 writing/paper-rbm/docs/paper.tex
edits from Dr. Vardeman

@@
 Montúfar, Rauh, and Ay
-\protect\hyperlink{ref-montufar2011expressive}{2011}). However, for cell
-data, the empirical distribution of an observed training set of visible
-variables provides a highest likelihood benchmark, before any parametric
-model class is even introduced and applied to obtain a refinement of
-model fit. As a consequence, we find that any fully principled fitting
-method based on the likelihood for a RBM with enough hidden variables
-will seek to reproduce the (discrete) empirical distribution of a
-training set. This aspect can be undesirable, and perhaps even
-unexpected compared to most modeling scenarios, in that no ``smoothed
-distribution'' may result when fitting a RBM model of sufficient size
-with a rigorous likelihood-based method. We are therefore led to be
-skeptical that models that involve these structures (like a deep
-Boltzmann machine) can achieve useful prediction or inference in a
-principled way without intentionally limiting the flexibility of the
-fitted model.
+\protect\hyperlink{ref-montufar2011expressive}{2011}). However, for
+binary data, the empirical distribution of an observed training set of
+visible variables provides a highest likelihood benchmark, before any
+parametric model class is even introduced and applied to obtain a
+refinement of model fit. As a consequence, we find that any fully
+principled fitting method based on the likelihood for a RBM with enough
+hidden variables will seek to reproduce the (discrete) empirical
+distribution of a training set. This aspect can be undesirable, and
+perhaps even unexpected compared to most modeling scenarios, in that no
+``smoothed distribution'' may result when fitting a RBM model of
+sufficient size with a rigorous likelihood-based method. We are
+therefore led to be skeptical that models that involve these structures
+(like deep Boltzmann machines) can achieve useful prediction or
+inference in a principled way without intentionally limiting the
+flexibility of the fitted model.
 
@@
 practitioners is not clear because these concepts are not (by default) a
-part of the accepted fitted methodology (Hinton
+part of the accepted fitting methodology (Hinton
 \protect\hyperlink{ref-hinton2002training}{2002}; Carreira-Perpinan and
@@
 the RBM class. We examine three Bayesian fitting techniques intended to
-avoid model impropriety in Section \ref{model-fitting} and conclude with
-a discussion in Section \ref{discussion}. Appendix \ref{appendix-rbm}
-provides proofs for results on RBM parameterizations and data codings
-described in Section \ref{data-encoding}.
+avoid model impropriety issues raised in Section \ref{model-fitting} and
+conclude with a discussion in Section \ref{discussion}. Appendix
+\ref{appendix-rbm} provides proofs for results on RBM parameterizations
+and data codings described in Section \ref{data-encoding}.
 
@@
 the probability properties of the RBM model for explaining data
-generation and it is hoped that the current exposition contributes to a
-better understanding in this direction.
+generation and the current exposition intends to contribute to a better
+understanding in this direction.
 
@@
 stacking multiple layers of RBMs in a deep architecture, proponents of
-the models claim to produce the ability to learn ``internal
+the model are claimed to produce the ability to learn ``internal
 representations that become increasingly complex, which is considered to
@@
 \(\nh + \nv + \nh*\nv\) parameters) creates at least three kinds of
-potential issues in model impropriety that we will call
+potential issues of model impropriety that we will call
 \emph{degeneracy}, \emph{instability}, and \emph{uninterpretability}. In
-this section we define these characteristics, consider how to quantify
-these in a RBM, and point out relationships among them.
+this section we define these, consider how to quantify them in a RBM,
+and point out relationships among them.
 
@@
 constructed from a set of nodes and an exponentially parameterized
-random graph model has a distribution of the form
+random graph model has a distribution with a probability mass function
+of the form
 \begin{align*}
@@
 \(\boldsymbol x\), may lead to substantial changes in the probability
-function \(f_{\boldsymbol \theta}(\boldsymbol x)\). Furthermore, model
-instability can be viewed on a spectrum of potential sensitivity in
-probability structure, with model degeneracy included as a extreme or
+mass function \(f_{\boldsymbol \theta}(\boldsymbol x)\). Furthermore,
+model instability can be viewed on a spectrum of potential sensitivity
+in probability structure, with model degeneracy included as a extreme or
 limiting case of instability. To quantify ``instability'' more
@@
 (i.e., as \(\nv \rightarrow \infty\)), we will say that the ratio
-(\eqref{eq:elpr}) needs to stay bounded for the sequence of RBM models to
-be stable. That is, we make the following convention.
+(\eqref{eq:elpr}) needs to stay bounded for a sequence of RBM models to be
+stable. That is, we make the following convention.
 \BeginKnitrBlock{definition}[S-unstable RBM]
@@
 \end{align*}
-\EndKnitrBlock{definition} In other words, the RBM model sequence is
+\EndKnitrBlock{definition} In other words, a RBM model sequence is
 unstable if, given any \(c > 0\), there exists an integer \(n_c > 0\) so
@@
 small changes in data configurations (i.e., will exhibit instability).
-Such instability can be a concern in a model for the similar reasons to
-degeneracy: as outcomes vary over the sample space, the geography of
-probabilities is extremely rugged, with deep pits following sharp
-mountains. Recall the applied example of RBM models as a means to
+Such instability can be a concern in a model for reasons similar to
+those related to degeneracy: as outcomes vary over the sample space, the
+geography of probabilities is extremely rugged, with deep pits following
+sharp mountains. Recall the applied example of RBM models as a means to
 classify images. For data as pixels in an image, the instability result
@@
 We next explore and numerically explain the relationship between values
-of \(\boldsymbol \theta\) and the three notions of model impropriety,
-near-degeneracy, instability, and uninterpretability, for RBM models of
-varying sizes.
+of \(\boldsymbol \theta\) and the three notions of model impropriety
+(near-degeneracy, instability, and uninterpretability), for RBM models
+of varying sizes.
 
@@
 developed in Section \ref{background}. As a compounding issue, these
-figures show that, as model grow in size, it becomes easier for more
+figures show that, as models grow in size, it becomes easier for more
 parameter configurations to push RBM models into near-degeneracy,
@@
 unstable, and uninterpretable for large portions of the parameter space
-with large \(\|\boldsymbol \theta\|\). These problematic aspects require
-serious consideration when using RBM models, on top of the additional
-matter of principled/rigorous fitting of RBM models.
+having large \(\|\boldsymbol \theta\|\). These problematic aspects
+require serious consideration when using RBM models, on top of the
+additional matter of principled/rigorous fitting of RBM models.
 
@@
 \(\gamma(\boldsymbol \theta)\) in a model (\eqref{eq:pmf}) of any
-realistic size. Ad hoc methods are used instead, which aim to avoid this
-problem by using stochastic ML approximations that employ a small number
-of MCMC draws (i.e., contrastive divergence, (Hinton
+realistic size. Ad hoc methods are often recommended instead, which aim
+to avoid this problem by using stochastic ML approximations that employ
+a small number of MCMC draws (i.e., contrastive divergence, (Hinton
 \protect\hyperlink{ref-hinton2002training}{2002})).
@@
 probabilities arbitrarily small. Furthermore, as development in the
-appendix shows, only it is possible to approximate any distribution on
-the visibles arbitrarily well (cf. Montufar and Ay
+appendix shows, not only is it possible to approximate any distribution
+on the visibles arbitrarily well (cf. Montufar and Ay
 \protect\hyperlink{ref-montufar2011refinements}{2011}), but dramatically
@@
 between \(65.05\) and \(132.61\) effective draws. Thus, BwTNLV would
-require at least \(4.7\) times as many iterations to be run of the MCMC
-chain in order to achieve the same amount of effective information about
-the posterior distribution. For this reason, consistent with the ACF
-results in Figure \ref{fig:acf}, BwTNLV does not seem to be an effective
-method for fitting the RBM, though computing resources can hinder use of
-the alternative BwTNML involving marginalization of hidden variables.
+require at least \(4.7\) times as many iterations of the MCMC chain to
+be run in order to achieve the same amount of effective information
+about the posterior distribution. For this reason, consistent with the
+ACF results in Figure \ref{fig:acf}, BwTNLV does not seem to be an
+effective method for fitting the RBM, though computing resources can
+hinder use of the alternative BwTNML involving marginalization of hidden
+variables.
 
@@
 models. Furthermore, without the appropriate generative behavior in the
-RBM model, the ability to quantify uncertainty in the estimated model
-parameters becomes impossible and the model loses useful application to
-prediction problems (e.g., realistic predictive distributions). In the
-case of classification, predictive distributions not be the ultimate
-goal, but when S-instability is present, small (imperceptible)
-differences in the data may lead to greatly different probabilities and
-thus greatly different classifications. For these reasons, we are
-skeptical about RBMs as data generative tools.
+RBM model, quantifying uncertainty in the estimated model parameters
+becomes impossible and the model loses useful application to prediction
+problems (e.g., realistic predictive distributions). In the case of
+classification, predictive distributions may not be the ultimate goal,
+but when S-instability is present, small (imperceptible) differences in
+the data may lead to greatly different probabilities and thus greatly
+different classifications. For these reasons, we are skeptical about
+RBMs as data generative tools.
 

### 2017-09-27 writing/paper-note/appendices/appendix-note.Rmd
some changes in the note from nordman

@@
 
-**Proof of Proposition \@ref(prp:degenFSFS).** where $|\mathcal{X}|<\infty$ holds in the FSFS model.
-We may suppose $|\mathcal{X}|>1$ (i.e., $\mathcal{X}^N$ has more than one outcome) because otherwise the model is trivially degenerate for all $N \geq 1$.
+**Proof of Proposition \@ref(prp:degenFSFS).** Where $|\mathcal{X}|<\infty$ holds in the FSFS model, we may suppose $|\mathcal{X}|>1$ (i.e., $\mathcal{X}^N$ has more than one outcome) because otherwise the model is trivially degenerate for all $N \geq 1$.
 Fix $0 < \epsilon < 1$.
@@
 \end{align*}
-as $N \rightarrow \infty$ by the definition of an unstable FSFS model (cf.
+as $N \rightarrow \infty$ by the definition of an S-unstable FSFS model (cf.
 Definition \@ref(def:instabFSFS)).

### 2017-09-27 writing/paper-note/chapters/instab.Rmd
some changes in the note from nordman

@@
 
-We consider the behavior, and the potential impropriety, of probability models built to incorporate a sequence of discrete observations with length $N$.
-Let $(X_1, \dots, X_N)$ denote a set of discrete random variables with a finite sample space, $\mathcal{X}^N$.
-That is, $\mathcal{X}$ with $|\mathcal{X}| < \infty$ represents a finite set of potential outcomes for each single variable $X_i$, and the data sequence $(X_1,\ldots,X_N)$ takes values in the $N$-fold product space $\mathcal{X}^N$.
+We consider the behavior, and the potential impropriety, sequences of discrete probability models built to incorporate observations of increasing sample size $N$.
+Interest is in identifying  instability in such models, which is roughly characterized by probabilities with extreme sensitivity to small changes in data configuration.
+The concept of instability was introduced in the field of statistical physics (i.e., point processes) by @ruelle1999statistical and then further extended by Schweinberger @schweinberger2011instability for a family of exponential models.
+At issue, models exhibiting instability are typically undesirable as these tend to provide poor representations of data or data-generation.
+As an example, such models can include near-degenerate distributions that assign all essential probability mass to only a subset of an overall sample space.
+The latter issue in connection to degeneracy has been recognized as concern in that dominant model outcomes  may not resemble observed data [cf.
+@handcock2003assessing].
+As a compounding issue, model instability often has direct negative impacts for statistical inference and computations based on likelihood functions.
+Namely, volatilities in probability structure can potentially hamper the numerical evaluations required for maximum likelihood estimation as well as other model-based simulations via Markov Chain Monte Carlo (MCMC).
+These reasons motivate our general study of instability for a broad class of probability models, described next.
+
+In the model framework, let $\boldsymbol X_N = (X_1, \dots, X_N)$ denote a collection of discrete random variables with a finite sample space, $\mathcal{X}^N$, represented as some $N$-fold Cartesian  product.
+That is, $\mathcal{X}$ with $|\mathcal{X}| < \infty$ denotes the set of potential outcomes for each single variable $X_i$, so that the product space $\mathcal{X}^N$ corresponds to values for the variables $\boldsymbol X_N=(X_1,\ldots,X_N)$.
 For each $N$, let $P_{\boldsymbol \theta_N}$ denote a probability model on $\mathcal{X}^N$, under which $P_{\boldsymbol \theta_N}(x_1, \dots, x_N) > 0$ is the probability of the data outcome $(x_1, \dots, x_N) \in \mathcal{X}^N$.
 In this, we assume that the model support of $P_{\boldsymbol \theta_N}$ is the sample space $\mathcal{X}^N$.
-This framework produces a series $P_{\boldsymbol \theta_N}$ of probability models, indexed by a generic sequence of parameters $\boldsymbol \theta_N$, to describe data of each length $N \geq 1$.
-The size and structure of such parameters are without restriction, and natural cases include those where $\boldsymbol \theta_N \in \mathbb{R}^{q(N)}$ for some arbitrary integer-valued function $q(\cdot) \geq 1$.
-We will refer to this model class as *Finitely Supported Finite Sequence (FSFS) models*.
+This framework produces probability models $P_{\boldsymbol \theta_N}$, indexed by a generic sequence of parameters $\boldsymbol \theta_N$, to describe data $\boldsymbol X_N$ of any given sample size $N \geq 1$.
+For simplicity, we will refer to this distributional class as *Finite Outcome Everywhere Supported (FOES)* models in the following.
+The dimension and structure of such parameters are generic, without restriction, though natural cases will be seen to include those where $\boldsymbol \theta_N \in \mathbb{R}^{q(N)}$ for some arbitrary integer-valued function $q(\cdot) \geq 1$.
 
-Section \@ref(examples) provides several examples of FSFS models commonly used in graph/network analysis and machine learning (i.e., deep learning models).
-Section \@ref(instability-results) establishes formal results regarding the propriety of FSFS models with regard to stability.
-A FSFS probability model sequence exhibits instability if small changes in the components of a data outcome $(x_1,\ldots,x_N)$ can result in large changes in probability $P_{\theta_N}(x_1,\ldots,x_N)$.
-The concept of instability, introduced in the field of statistical physics by @ruelle1999statistical, was extended to include a notion of detection and quantification for certain exponential family models by @schweinberger2011instability.
-For similar exponential models, particularly in connection to random graphs/networks, @handcock2003assessing considered (mean-based) characterizations for so-called model degeneracy, whereby a probability model places all mass on a small subset of the sample space and produces undesirably low variability in model outcomes.
-As described by @schweinberger2011instability, model instability and model degeneracy are related by viewing degeneracy as an extreme or limiting form of instability.
-The instability results of @schweinberger2011instability were developed for the case of discrete exponential family models.
-The main results here concern a general measure of model instability, appropriate across the whole FSFS model class.
-This can be used to identify when certain maximal probabilities in FSFS models are too extreme relative to the length $N$ and may thereby induce a potentially undesirable probability structure.
-In this case, a one component change in the data sequence may shift probability by orders of magnitude, and FSFS models are rigorously shown to become degenerate as the measure of instability increases.
-Lastly, Section \@ref(implications) emphasizes the implications of our model propriety results and proofs of the main results appear in Appendix \@ref(appendix-instab).
+Section \@ref(examples) provides some examples of FOES models encountered in graph/network analysis and machine learning (i.e., deep learning models).
+These are used as references for later illustrations.
+Section \@ref(instability-results) then  establishes several formal results for  FOES models with regard to instability.
+@schweinberger2011instability originally developed instability results framed for, or specific to, a certain class of discrete exponential models.
+For similar exponential models with random networks, @handcock2003assessing studied model degeneracy, where a probability model places near complete mass on  modes and may thereby narrow the feasible model outcomes.
+As findings here and from @schweinberger2011instability suggest, model instability and degeneracy may also be related by viewing degeneracy as an extreme, or limiting form, of instability.
+Our main results establish a broad characterization of model instability, appropriate across the whole FOES model class, that incorporates results of @schweinberger2011instability as a special case.
+We prescribe a general and simple condition  for identifying instability in a FOES model sequence, which  quantifies whether certain maximal probabilities in a FOES model  are too extreme relative to the sample size $N$.
+When fulfilled, the probability structure of a FOES model is shown to exhibit extreme sensitivity, with probability assignments possessing extreme peaks and troughs across nearly identical outcomes.
+As the measure of model instability increases, probabilities from an unstable FOES model additionally increase in volatility and provably slide into degeneracy.
+Section \@ref(implications) then emphasizes the implications of such model instability, showing that such impropriety can be expected to numerically hinder maximum likelihood estimation and   MCMC-based simulations.
+As one potential remedy, suggestions are given or constraining model parameterizations and avoid the most problematic regions of the parameter space.
+Proofs of the main results appear in Appendix \@ref(appendix-instab).
 
@@
 
-Many model families fall under the umbrella of FSFS models. For illustration, this section presents three specific examples of FSFS models, including models with deep architectures.
+Many model families fall under the umbrella of FOES models. For illustration, this section presents three specific examples of FOES models, including models with deep architectures.
 
@@
 
-For discrete random variables $\boldsymbol X = (X_1, \dots, X_N)$ with sample space $\mathcal{X}^N$, $|\mathcal{X}| < \infty$, consider an exponential family model for $\boldsymbol X$ with probability mass function of the form
-$$
-p_{N, \boldsymbol \lambda}(\boldsymbol x) = \exp\left[\boldsymbol\eta^T(\boldsymbol \lambda) \boldsymbol g_N(\boldsymbol x) - \psi(\boldsymbol \lambda)\right], \qquad \boldsymbol x \in \mathcal{X}^N,
-$$ 
-depending on parameter vector $\boldsymbol \lambda \in \Lambda \subset \mathbb{R}^{k}$ and natural parameter function $\boldsymbol \eta : \mathbb{R}^k \mapsto \mathbb{R}^L$ with fixed positive integers $k$ and $L$ denoting their dimensions.
+For random variables $\boldsymbol X \equiv\boldsymbol X_N= (X_1, \dots, X_N)$ with sample space $\mathcal{X}^N$, $|\mathcal{X}| < \infty$, consider an exponential family model for $\boldsymbol X$ with probability mass function given by
+\begin{equation}
+(\#eq:expo)
+p_{N, \boldsymbol \theta}(\boldsymbol x) = \exp\left[\boldsymbol\eta^T(\boldsymbol \theta) \boldsymbol g_N(\boldsymbol x) - \psi(\boldsymbol \theta)\right], \qquad \boldsymbol x \in \mathcal{X}^N,
+\end{equation}
+depending on parameter vector $\boldsymbol \theta \in \Theta_N \subset \mathbb{R}^{k}$ and natural parameter function $\boldsymbol \eta : \mathbb{R}^k \mapsto \mathbb{R}^L$ with fixed positive integers $k$ and $L$ denoting their dimensions.
 Above, $\boldsymbol g_N : \mathcal{X}^N \mapsto \mathbb{R}^L$ is a vector of sufficient statistics, while
 $$
-\psi(\boldsymbol \lambda) = \log \sum\limits_{\boldsymbol x \in \mathcal{X}^N}\exp\left[\boldsymbol \eta^T(\boldsymbol \lambda) \boldsymbol g_N(\boldsymbol x) \right], \qquad \boldsymbol \lambda \in \Lambda,
+\psi(\boldsymbol \theta) = \log \sum\limits_{\boldsymbol x \in \mathcal{X}^N}\exp\left[\boldsymbol \eta^T(\boldsymbol \theta) \boldsymbol g_N(\boldsymbol x) \right], \qquad \boldsymbol \theta \in \Theta_N\equiv \{\boldsymbol \theta \in \mathbb{R}^k : \psi(\boldsymbol \theta) < \infty \},
 $$
-denotes the normalizing function, and $\Lambda = \{\boldsymbol \lambda \in \mathbb{R}^k : \psi(\boldsymbol \lambda) < \infty, k \le q(N) \}$ is the parameter space.
+denotes the normalizing function with parameter space $\Theta_N$.
+The natural parameter function $\eta (\boldsymbol \theta)$ has a linear form (i.e., $\eta (\boldsymbol \theta)= \bm{A} \boldsymbol \theta$ for a given $L \times k$ matrix $\bm{A}$) in many common model formulaitons, though may also be nonlinear (e.g., curved exponential families).
+In the linear case, $\eta (\boldsymbol \theta) = \boldsymbol \theta$ may be generally assumed in the exponential parameterization with a minor modification to the definition of sufficient statistics $\boldsymbol g_N(\boldsymbol x)$.
 
-Defining $P_{\boldsymbol \theta_N}(\boldsymbol x)\equiv p_{N,\boldsymbol \lambda_N}(\boldsymbol x)$ with $\boldsymbol \theta_N=\boldsymbol \lambda_N$ to be a sequence of elements of $\Lambda \subset \mathbb{R}^k$ and noting that $P_{\boldsymbol \theta_N}(\boldsymbol x) > 0$ for all $\boldsymbol x \in \mathcal{X}^N$, these discrete exponential family models are special cases of the FSFS models.
-Such exponential models arise with spatial data on a lattice [@besag1974spatial], network data [@wasserman1994social; @handcock2003assessing], and even standard independence models for discrete data, such as with $N$ iid Bernoulli random variables.
-(Note that for random graphs or networks with, say, $m$ nodes, one may wish to consider $N={m \choose 2}$ edges as binary (presence/absence) variables $X_i$.
-In this case, the length $N$ of data sequence may naturally increase as a function of $m$.) For these exponential models, the dimension of the parameter $\boldsymbol \theta_N$ remains constant over each $N$, as $\boldsymbol \theta_N$ lies in a parameter space of fixed Euclidean dimension $k$.
-This need not be true for other types of FSFS models considered in Sections \@ref(restricted-boltzmann-machines)-\@ref(deep-learning).
-@schweinberger2011instability considered instability in such exponential models (e.g., for random graphs) for sequences of fixed parameters $\boldsymbol \theta_N=\boldsymbol \lambda\in\mathbb{R}^k$, $N \geq 1$.
+Such discrete exponential family models are special cases of the FOES models, as seen by  defining $P_{\boldsymbol \theta_N}(\boldsymbol x)\equiv p_{N,\boldsymbol \theta_N}(\boldsymbol x)> 0$,  $\boldsymbol x \in \mathcal{X}^N$, based on \@ref(eq:expo) and a parameter sequence $\boldsymbol \theta_N \in \Theta_N \subset \mathbb{R}^k$.
+For example, if observations $\boldsymbol X = (X_1,\ldots,X_N)$ correspond to $N$ independent and identically distributed Bernoulli random variables, each indicating a binary $0$-$1$ outcome,  the resulting probabilities have exponential form \@ref(eq:expo)  given by
+\begin{equation}
+(\#eq:mod1)
+P_{\boldsymbol \theta_N}(\boldsymbol x) \propto
+ \exp\left[\boldsymbol \theta_N \sum_{i=1}^N x_i\right], \qquad \boldsymbol x=(x_1,\ldots,x_N) \in\{0,1\}^N, 
+ \end{equation} 
+ with sufficient statistic $\boldsymbol g_N(\boldsymbol x)\equiv \sum_{i=1}^N  x_i$ and "log odds ratio" parameter $\boldsymbol \theta_N \equiv  \log[ P_{\boldsymbol \theta_N}(X_i=1)/P_{\boldsymbol \theta_N}(X_i=0) ] \in \mathbb{R}$.
+More generally, supposing $\boldsymbol X =(X_1,\ldots,X_N)$ represent $N$ independent trials, each assuming an outcome $\{1,\ldots,k\}$ among $k$ possibilities (e.g., a die roll), a multinomial distribution is given by
+\begin{equation}
+(\#eq:mod11)
+P_{\boldsymbol \theta_N}(\boldsymbol x) \propto  \exp\left[  \boldsymbol \theta_{N}^T g_N(\boldsymbol x)   \right] =
+\exp\left[ \sum_{j=1}^k {\theta_{j,N}} \sum_{i=1}^N \mathbb{I}(x_i=j) \right], \qquad \boldsymbol x  \in\{1,\ldots,k\}^N, 
+\end{equation}
+with sufficient statistic $\boldsymbol g_N(\boldsymbol x)$ involving a count $\sum_{i=1}^N \mathbb{I}(x_i=j)$ for each outcome $j \in \{1,\ldots,k\}$, where $\mathbb{I}(\cdot)$ denotes the indicator function, and parameters $\boldsymbol \theta_N=(\theta_{1,N},\ldots,\theta_{k,N})\in\mathbb{R}^k$ defining log-probability ratios $\theta_{i,N}-\theta_{j,N} =\log [P_{\boldsymbol \theta_N}(X_1=i)/P_{\boldsymbol \theta_N}(X_1=j)]$.
+In addition to such standard models for discrete independent data,
+exponential models of FOES type commonly arise with dependent spatial data [@besag1974spatial] and network/relational data [@wasserman1994social; @handcock2003assessing].
+For a random graph  or network  with, say, $n$ nodes,  consider $N={n \choose 2}$ random edges where the $i$th edge is associated with a pair of nodes $s_i \equiv \{v_i,u_i\}$  and a binary variable $X_i\in\{0,1\}$ indicating   presence/absence of an edge among the node pair $s_i$, $i=1,\ldots,N$.
+Here the length $N$ of the edge variable sequence $\boldsymbol X = (X_1,\ldots,X_N)$ increases as a function of node number $n$ and corresponding exponential models often incorporate graph topographical features derived from $\boldsymbol X$.
+As an example, consider a   graph model of exponential/FOES form prescribed by
+\begin{equation}
+(\#eq:mod2)
+P_{\boldsymbol \theta_N}(\boldsymbol x) \propto
+ \exp\left[\sum_{j=1}^3 \theta_{j,N} g_{j,N}(\boldsymbol x)\right], \quad\qquad \boldsymbol x=(x_1,\ldots,x_N)  \in\{0,1\}^N,
+\end{equation}
+$$
+  g_{1,N}(\boldsymbol x) \equiv \sum_{i=1}^N  x_i, \qquad\quad g_{2,N}(\boldsymbol x) \equiv \sum_{1 \leq i<j \leq N,\atop s_i \cap s_j \neq \emptyset}\!\!\!   x_i x_j, \qquad
+  g_{3,N}(\boldsymbol x) \equiv \sum_{1 \leq i<j<\ell \leq N, \atop s_i \cap s_j \neq \emptyset,s_i \cap s_\ell \neq \emptyset, \\ s_j \cap s_\ell \neq \emptyset } \!\!\!\!\!\!\!\!
+\!\!\!\!\!\!\!\!
+x_i x_j x_\ell,
+$$
+involving the numbers of edges, 2-stars and triangles among an outcome $\boldsymbol x$ given by $g_{1,N}(\boldsymbol x)$, $g_{2,N}(\boldsymbol x)$ and $g_{3,N}(\boldsymbol x)$, respectively, along with $k=3$ real parameters $\boldsymbol \theta_N \equiv (\theta_{1,N},\theta_{2,N},\theta_{3,N})$.
+For this network model \@ref(eq:mod2) in particular, as well as for more general models of form \@ref(eq:expo),  @schweinberger2011instability considered instability in such exponential models with sequences of fixed parameters $\boldsymbol \theta_N = (\theta_1,\ldots,\theta_k)\in\mathbb{R}^k$, $N \geq 1$, of fixed dimension $k$.
+
+For model sequences $P_{\boldsymbol \theta_N}(\boldsymbol x)\equiv p_{N,\boldsymbol \theta_N}(\boldsymbol x)$ of the exponential type \@ref(eq:expo), such as those in \@ref(eq:mod1)-\@ref(eq:mod2), note that the dimension $k$ of the parameter $\boldsymbol \theta_N\in\Theta \subset \mathbb{R}^k$ necessarily remains the same for all sample sizes $N \geq 1$ as the form of the natural parameter function $\eta(\cdot)$ in \@ref(eq:expo) and the number of sufficient statistics $\boldsymbol g_{N}(\boldsymbol x)$ do not depend on $N$.
+Consequently, $\boldsymbol \theta_N$ lies in a parameter space of fixed Euclidean dimension $k$.
+However, this aspect need not be true for other types of FOES models considered in Sections \@ref(restricted-boltzmann-machines) - \@ref(deep-learning), where instead the numbers of parameters and sufficient statistics commonly increase with the sample size $N$.
 
@@
 A RBM architecture has two layers, hidden ($\mathcal{H}$) and visible ($\mathcal{V}$), with conditional independence within each layer.
-Let $\boldsymbol X=(X_1,\ldots,X_N)$ denote the $N$ random variables for visibles with support $\mathcal{X}^N$ and $\boldsymbol H=(H_1,\ldots,H_{N_H})$ denote the $N_H$ random variables for hiddens with support $\mathcal{X}^{N_H}$ where  $\mathcal{X} = \{-1,1\}$.
-For parameters $\boldsymbol \alpha \in \mathbb{R}^{N_H}$, $\boldsymbol \beta \in \mathbb{R}^N$, and $\Gamma$ as a matrix with dimension $N_H \times N$, the RBM model for $\tilde{\boldsymbol X}=(\boldsymbol X,\boldsymbol H)$ then has the joint probability mass function
+Let $\boldsymbol X = (X_1,\ldots,X_N)$ denote the $N$ random variables for visibles with support $\mathcal{X}^N$ and $\boldsymbol H = (H_1,\ldots,H_{N_\mathcal{H}})$ denote the $N_\mathcal{H}$ random variables for hiddens with support $\mathcal{X}^{N_\mathcal{H}}$ where $\mathcal{X} = \{-1,1\}$.
+For parameters $\boldsymbol \theta_N^{\mathcal{H}} \in \mathbb{R}^{N_\mathcal{H}}$, $\boldsymbol \theta_N^{\mathcal{V}}\in \mathbb{R}^N$, and $\boldsymbol \theta_N^{\mathcal{HV}}$ as a real matrix with dimension $N_\mathcal{H} \times N$, the RBM model for $\tilde{\boldsymbol X}=(\boldsymbol X,\boldsymbol H)$  has the joint probability mass function
+\begin{equation}
+(\#eq:RBM1)
+\tilde{P}_{\boldsymbol \theta_N} (\tilde{\boldsymbol x}) = \exp\left[ (\boldsymbol \theta_N^{\mathcal{H}})^T \boldsymbol h + \boldsymbol (\boldsymbol \theta_N^{\mathcal{V}})^T \boldsymbol x + \boldsymbol h^T  \boldsymbol\theta_N^{\mathcal{HV}} \boldsymbol x - \psi(\boldsymbol \theta_N)\right], \quad \tilde{\boldsymbol x} = (\boldsymbol x, \boldsymbol h) \in \{\pm 1\}^{N+N_\mathcal{H}}
+\end{equation}
+with normalizing function
 $$
-P_{\boldsymbol \theta_N} (\tilde{\boldsymbol x}) = \exp\left[ \boldsymbol \alpha^T \boldsymbol h + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^T \Gamma \boldsymbol x - \psi(\boldsymbol \theta_N)\right], \qquad \tilde{\boldsymbol x} = (\boldsymbol h, \boldsymbol x) \in \mathcal{X}^{N+N_H},
+\psi(\boldsymbol \theta_N) = \log \sum_{\tilde{\boldsymbol x} \in \{\pm 1\}^{N+N_H} } \exp\left[ (\boldsymbol \theta_N^{\mathcal{H}})^T \boldsymbol h + \boldsymbol (\boldsymbol \theta_N^{\mathcal{V}})^T \boldsymbol x + \boldsymbol h^T  \boldsymbol\theta_N^{\mathcal{HV}} \boldsymbol x\right].
 $$
-where 
-$$
-\psi(\boldsymbol \theta_N) = \log \sum_{\tilde{\boldsymbol x} \in \mathcal{X}^{N + N_H}} \exp\left[ \boldsymbol \alpha^T \boldsymbol h + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^T \Gamma \boldsymbol x\right], \qquad \boldsymbol \theta_N \in \Theta_N,
-$$ 
-is the normalizing function.
-Let $\boldsymbol \theta_N = (\boldsymbol \alpha, \boldsymbol \beta,\Gamma) \in \Theta_N \subset \mathbb{R}^{q(N)}$ with $q(N) = N + N_H + N*N_H$ denote the vector of parameters for the RBM.
-The probability mass function for the visible variables $X_1, \dots, X_N$ follows from marginalizing the joint specification
-$$
-P_{\boldsymbol \theta_N} (\boldsymbol x) = \sum\limits_{\boldsymbol h \in \mathcal{X}^{N_H}} P_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h), \qquad \boldsymbol x \in \mathcal{X}^N.
-$$
-Note that the vector of model parameters $\boldsymbol \theta_N$, of size $q(N)$, grows in size as a function of sample dimension $N$ to accommodate the dimension of visible variables $X_1, \dots, X_N$, and one may further choose the number $N_H$ of hidden variables to change with $N$ as well.
-In particular, the number $N_H$ of hiddens may also potentially and arbitrarily increase with $N$.
-Additionally, as $|\mathcal{X}| = 2$ and $P_{\boldsymbol \theta_N}(\boldsymbol x) > 0$ for all $\boldsymbol x \in \mathcal{X}^N$, the RBM model specification for visibles is a FSFS model.
-This example also indicates that models formed by marginalizing a base FSFS model (e.g., a type of exponential family model) is again a FSFS model class.
+Let $\boldsymbol \theta_N = (\boldsymbol \theta_N^{\mathcal{H}}, \boldsymbol \theta_N^{\mathcal{V}}, \boldsymbol\theta_N^{\mathcal{HV}} ) \in \Theta_N \equiv \mathbb{R}^{q(N)}$, with $q(N) = N + N_\mathcal{H} + N*N_\mathcal{H}$, denote the parameter vector for the RBM, as indexed by the number $N$ of visible random variables (which may differ from the actual lengths of these parameter vectors).
+The probability mass function for the visible variables $\boldsymbol X = (X_1, \dots, X_N)$ follows from marginalizing the joint specification to yield
+\begin{equation}
+(\#eq:RBM2)
+P_{\boldsymbol \theta_N} (\boldsymbol x) = \sum\limits_{\boldsymbol h \in \{\pm 1\}^{N_{\mathcal{H}}}} \tilde{P}_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h), \qquad \boldsymbol x \in \{\pm 1\}^{N}\equiv \mathcal{X}^N.
+\end{equation}
+Here the baseline model \@ref(eq:RBM1) for hidden/visible variables is a linear exponential one in sufficient statistics $(\tilde{\boldsymbol X}, \boldsymbol X^T\boldsymbol H)$ using $\tilde{\boldsymbol X}=(\boldsymbol X,\boldsymbol H)$ from   \@ref(eq:RBM1), but the form differs from the previous exponential models in \@ref(eq:expo) in that the lengths of parameters $\boldsymbol \theta_N$ and statistics  $(\tilde{\boldsymbol X}, \boldsymbol X^T\boldsymbol H)$ increase to incorporate more visible variables.
+That is, in contrast to \@ref(eq:expo), the natural parameter function involved in the RBM model \@ref(eq:RBM1), as the identity mapping of the parameters $\boldsymbol \theta_N\in\mathbb{R}^{q(N)}$, naturally grows in  dimension  $q(N)\to \infty$ to accommodate visible variables $X_1, \dots, X_N$ of increasing sample size $N\to \infty$.
+Additionally, one may further arbitrarily choose the number $N_\mathcal{H}$ of hidden variables $\boldsymbol H$ in the joint RMB model \@ref(eq:RBM1) to define a marginal model \@ref(eq:RBM2) for the $N$ visible variables $\boldsymbol X$, and the number $N_\mathcal{H}$ of hiddens may also potentially   increase with $N$.
+Because $|\mathcal{X}| = 2$ and $P_{\boldsymbol \theta_N}(\boldsymbol x) > 0$ for all $\boldsymbol x \in \mathcal{X}^N$, the RBM  specification \@ref(eq:RBM2) for visibles $\boldsymbol$ corresponds to a FOES model, while the joint distribution \@ref(eq:RBM1) for $(\boldsymbol X, \boldsymbol H)$ is also a FOES model.
+As this example also indicates, any model formed by marginalizing a base FOES model class, such as the RBM joint specification \@ref(eq:RBM1), is again a FOES model.
 
@@
 
-Consider two models with "deep architecture" that contain multiple hidden (or latent) layers in addition to a visible layer of data, namely a deep Boltzmann machine [@salakhutdinov2009deep] and a deep belief network [@hinton2006fast].
-Let $M$ denote the number of hidden layers included in the model and let $N_{(H,1)}, \dots, N_{(H,M)}$ denote the numbers of hidden variables within each hidden layer.
-Then the random vector $\tilde{\boldsymbol X} = \{H^{(1)}_1, \dots, H^{(1)}_{N_{(H,1)}}, \dots, H^{(M)}_1, \dots, H^{(M)}_{N_{(H,M)}}, \boldsymbol X\}$ collects both the hidden variables $\{ H_{i}^{(j)} : i=1,\ldots, N_{(H,j)}, j=1,\ldots,M\}$ and visible variables $\boldsymbol X =(X_1,\ldots,X_N)$ in a deep probabilistic model.
+Consider two models with "deep architecture" that contain multiple hidden (or latent) layers in addition to a visible layer of data, namely a deep Boltzmann machine [@salakhutdinov2009deep] and a deep belief network @hinton2006fast].
+Let $M$ denote the
+number of hidden layers included in the model and let $N_{(H,1)}, \dots, N_{(H,M)}$ denote the numbers of hidden variables within each hidden layer. Then the random vector
+$\tilde{\boldsymbol X} = \{H^{(1)}_1, \dots, H^{(1)}_{N_{(H,1)}}, \dots, H^{(M)}_1, \dots, H^{(M)}_{N_{(H,M)}}, \boldsymbol X\}$ collects both the hidden variables
+$\{ H_{i}^{(j)} : i=1,\ldots, N_{(H,j)}, j=1,\ldots,M\}$ and visible variables $\boldsymbol X =(X_1,\ldots,X_N)$ in a deep probabilistic model.
 Each variable outcome will again lie in $\mathcal{X} = \{-1,1\}$.
@@
 $$
-P_{\boldsymbol \theta_N} (\tilde{\boldsymbol x}) = \exp\left[ \sum\limits_{i = 1}^M\boldsymbol \alpha^{(i)T} \boldsymbol h^{(i)} + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^{(1)T} \Gamma^{(0)} \boldsymbol x + \sum\limits_{i = 1}^{M - 1} \boldsymbol h^{(i)T} \Gamma^{(i)} \boldsymbol h^{(i + 1)} - \psi(\boldsymbol \theta_N) \right],
+\tilde{P}_{\boldsymbol \theta_N} ( \tilde{\boldsymbol x} ) = \exp\left[ \sum\limits_{i = 1}^M\boldsymbol \alpha^{(i)T} \boldsymbol h^{(i)} + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^{(1)T} \Gamma^{(0)} \boldsymbol x + \sum\limits_{i = 1}^{M - 1} \boldsymbol h^{(i)T} \Gamma^{(i)} \boldsymbol h^{(i + 1)} - \psi(\boldsymbol \theta_N) \right],
 $$ 
@@
 $$ 
-for $\boldsymbol \theta_N \in \Theta_N$ is the normalizing function and parameters in the model are $\boldsymbol \beta \in \mathbb{R}^V$, $\boldsymbol \alpha^{(i)} \in \mathbb{R}^{N_{(H,i)}}$ for $i = 1, \dots, M$, along with a matrix $\Gamma^{(0)}$ of dimension $N_{(H,1)} \times N$, and matrices $\Gamma^{(i)}$ of dimension $N_{(H,i)} \times N_{(H,i+1)}$ for $i = 1, \dots, M-1$.
-Let $\boldsymbol \theta_N = (\boldsymbol \alpha^{(1)}, \dots, \boldsymbol \alpha^{(M)}, \boldsymbol \beta,\Gamma^{(0)}, \dots, \Gamma^{(M - 1)}) \in \Theta_N \subset \mathbb{R}^{q(N)}$ denote the combined vector of parameters with total length $q(N)= N_{(H,1)}+\cdots N_{(H,M)} + N + N_{(H,1)}*N+N_{H,2}*H_{(H,1)}+\cdots +N_{(H,M)}*H_{(H,M)-1}$.
-
+is the normalizing function for $\boldsymbol \theta_N = (\boldsymbol \alpha^{(1)}, \dots, \boldsymbol \alpha^{(M)}, \boldsymbol \beta,\Gamma^{(0)}, \dots, \Gamma^{(M - 1)}) \in \Theta_N \subset \mathbb{R}^{q(N)}$, consisting of model parameters $\boldsymbol \beta \in \mathbb{R}^N$, $\boldsymbol \alpha^{(i)} \in \mathbb{R}^{N_{(H,i)}}$, $i = 1, \dots, M$, along with a matrix $\Gamma^{(0)}$ of dimension $N_{(H,1)} \times N$, and matrices $\Gamma^{(i)}$ of dimension
+$N_{(H,i)} \times N_{(H,i+1)}$ for $i = 1, \dots, M-1$.
+The combined parameter vector $\boldsymbol \theta_N$ has total length $q(N)= N_{(H,1)}+\cdots N_{(H,M)} + N + N_{(H,1)}*N+N_{H,2}*H_{(H,1)}+\cdots +N_{(H,M)}*H_{(H,M)-1}$.
 The probability mass function for the visible random variables $X_1, \dots, X_N$ follows from this joint specification as
 $$
-P_{\boldsymbol \theta_N} (\boldsymbol x) = \sum\limits_{(\boldsymbol h^{(1)}, \dots, \boldsymbol h^{(M)}) \in \mathcal{X}^{N_{(H,1)} + \cdots + N_{(H,M)}}} P_{\boldsymbol \theta_N} (\tilde{\boldsymbol x}), \qquad \boldsymbol x \in \mathcal{X}^N
+P_{\boldsymbol \theta_N} (\boldsymbol x) = \sum\limits_{(\boldsymbol h^{(1)}, \dots, \boldsymbol h^{(M)}) \in \mathcal{X}^{N_{(H,1)} + \cdots + N_{(H,M)}}} \tilde{P}_{\boldsymbol \theta_N} (\boldsymbol h^{(1)}, \dots, \boldsymbol h^{(M)}, \boldsymbol x) , \qquad \boldsymbol x \in \mathcal{X}^N.
 $$ 
-Again like the RBM case, the DBM model specification is an example of a FSFS model.
+Again like the RBM case, the DBM model specification is an example of a FOES model.
 
@@
 The difference between the DBM and DBN models is that all but the last stacked layer in a DBN are Bayesian networks [see @pearl985bayesian], rather than RBMs.
-Thus for visibles $X_1, \dots, X_N$ with support $\mathcal{X}^N$, a DBN is also a FSFS model if the number $q(N)$ of components in the parameter vector is dependent on the dimension of the visibles.
-Commonly, as in logistic belief nets [@neal1992connectionist], a "weight" parameter is placed on each interaction between visibles, $X_1, \dots, X_N$, and the first layer of latent variables, $H^{(1)}_1, \dots, H^{(1)}_{N_{(H,1)}}$, in the definition of a FSFS model.
+Thus for visibles $X_1, \dots, X_N$ with support $\mathcal{X}^N$, a DBN is also a FOES model if the number $q(N)$ of components in the parameter vector is dependent on the dimension of the visibles.
+Commonly, as in logistic belief nets [@neal1992connectionist], a "weight" parameter is placed on each interaction between visibles, $X_1, \dots, X_N$, and the first layer of latent variables, $H^{(1)}_1, \dots, H^{(1)}_{N_{(H,1)}}$, in the definition of a FOES model.
 
-# Instability results
+# Main results on model instability {#instability-results}
 
-To define or measure instability in FSFS models, it is useful to consider the behavior of a data model sequence $P_{\theta_N}$.
-A relevant quantity to this end is a (scaled) extremal log-probability ratio (ELPR)
+## A criterion for instability {#criterion}
+
+To define a measure of instability in FOES models, it is useful to consider the behavior of  data  models $P_{\theta_N}$, again supported on a set $\mathcal{X}^N$ of outcomes for $\boldsymbol X\equiv \boldsymbol X_N =(X_1,\ldots,X_N)$,  in connection to the sample size $N$.
+A relevant quantity to this end is a log-ratio of extremal probabilities (LREP), defined as
 \begin{align}
-\frac{1}{N} \log \left[\frac{\max\limits_{(x_1, \dots, x_N) \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(x_1, \dots, x_N)}{\min\limits_{(x_1, \dots, x_N) \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(x_1, \dots, x_N)}\right] \equiv \frac{1}{N} \text{ELPR}_N(\boldsymbol \theta_N).
 (\#eq:elpr)
+ \REP (\boldsymbol \theta_N)  =  \log \left[\frac{\max\limits_{  \boldsymbol x\in \mathcal{X}^N}P_{\boldsymbol \theta_N}( \boldsymbol x)}{\min\limits_{ \boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}( \boldsymbol x)}\right],
 \end{align}
+based on maximum and minimal model probabilities.
+In what follows, the main idea is that instability, and other negative model features, can be associated with a FOES model formulation for $N$ random variables where the log-ratio (\ref{eq:elpr}) is overly large relative to the sample size $N$.
+That is, a sequence of FOES probability models $P_{\boldsymbol \theta_N}$ results in specifying the distribution of observations $\boldsymbol X=(X_1,\ldots,X_N)$ for each sample size $N \geq 1$ and instability will generally occur among these models whenever the corresponding LREP \@ref(eq:elpr) grows faster than $N$.
+This leads to the following definition.
 
-The main idea is that, in formulating FSFS models for potentially increasing numbers of variables (i.e., for $N \rightarrow \infty$), the ratio (\@ref(eq:elpr)) should remain bounded to ensure model stability, requiring that the largest probability possible under $P_{\theta_N}$ maintain a fixed order of magnitude relative to the smallest probability allowed under the same model.
-Specifically, the log of the ratio should grow at mostly linearly with the sample size $N$.
-This leads to the following definition.
-```{definition, name = "S-unstable FSFS", label="instabFSFS", echo=TRUE}
-Let $\boldsymbol \theta_N \in \mathbb{R}^{q(N)}$ be a sequence of FSFS model parameters where the size of the model $q(N)$ is a function of the number of random variables $N$.
-A FSFS model formulation is \emph{Schweinberger-unstable} or \emph{S-unstable} if, as the number of variables increase ($N \rightarrow \infty$),
+```{definition, name="S-unstable FOES model", label="instabFSFS", echo=TRUE}
+A FOES model formulation for $\boldsymbol X_N=(X_1,\ldots,X_N)$ is *Schweinberger-unstable* or *S-unstable* if
+\begin{equation}
+(\#eq:Sun)
+\lim \limits_{N \rightarrow \infty} \frac{1}{N} \REP(\boldsymbol \theta_N) \equiv \lim \limits_{N \rightarrow \infty} \frac{1}{N}\log \left[\frac{\max\limits_{  \boldsymbol x\in \mathcal{X}^N}P_{\boldsymbol \theta_N}( \boldsymbol x)}{\min\limits_{ \boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}( \boldsymbol x)}\right] = \infty
+\end{equation}
+ as the number of variables increases ($N \rightarrow \infty$).
+```
+In other words, a model is S-unstable if $\REP(\boldsymbol \theta_N)/N$ is an unbounded sequence of sample size  $N$; namely, given any $C > 0$, there exists an integer $N_C > 0$ so that $\REP(\boldsymbol \theta_N)/N > C$ holds for all $N \ge N_C$.
+A FOES model formulation may be termed S-stable if it fails to be S-unstable, i.e., if $\sup_{N \geq 1}\REP(\boldsymbol \theta_N)/N $ is bounded.
+
+This definition of S-unstable is a generalization or reinterpretation of "unstable" used in @schweinberger2011instability by allowing possibility non-exponential family models (e.g., RBM and DBM models in Sections \@ref(restricted-boltzmann-machines) - \@ref(deep-learning) as well as a potentially increasing number $d(N)$ of parameters through the parameter sequence $\boldsymbol \theta_N\in \mathbb{R}^{d(N)}$.
+While this definition differs in form and scope from the original, it does match that in @schweinberger2011instability for the special case of exponential models (cf.
+Section \@ref(discrete-exponential-family-models) considered there.
+Section \ref@(illustrations) provides several examples  of unstable models as well as causes for model instability, where the latter may often be traced to issues in model form (i.e., data functions) and/or parameterization.
+We next describe several potentially undesirable features associated with unstable models.
+
+```{remark, echo=TRUE}
+In the definition \@ref(eq:Sun) of S-instability, we note that the numerical measure $\REP(\boldsymbol \theta_N)/N$ of model instability is invariant to *independent  replications* of data.
+That is, let $M \geq 1$ denote a possible number of replications and consider data $\boldsymbol Y_{N,M} \equiv (\boldsymbol X^{(1)}_N, \dots, \boldsymbol X^{(M)}_N)$ formed by $\{ \boldsymbol X^{(j)}_N\}_{j=1}^M$ as $M$  iid replications of a random vector $\boldsymbol X_N=(X_1,\ldots,X_N)$, where the latter follows a FOES model with probabilities \(P_{ \boldsymbol \theta_N}(\boldsymbol x)>0\), $\boldsymbol x\in\mathcal{X}^N$.
+This leads to a joint model, say $P_{ \boldsymbol \theta_N}(\boldsymbol y)$, $\boldsymbol y\in\mathcal{X}^{NM}$, for $\boldsymbol Y_{N,M}$ consisting of $N*M$ random variables in total.
+Then, the LREP for $\boldsymbol Y_{N,M}$, scaled by associated size, is given by
 \begin{align*}
-\lim\limits_{N \rightarrow \infty} \frac{1}{N} \text{ELPR}(\boldsymbol \theta_N) \equiv \lim\limits_{N \rightarrow \infty} \frac{1}{N} \log \left[\frac{\max\limits_{(x_1, \dots, x_N) \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(x_1, \dots, x_N)}{\min\limits_{(x_1, \dots, x_N) \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(x_1, \dots, x_N)}\right] = \infty.
+\frac{1}{NM}\REP_{\boldsymbol Y_{N,M}}( \boldsymbol\theta_N ) &\equiv \frac{1}{NM}\log\left[\frac{\max_{\boldsymbol y \in \mathcal{X}^{NM}}P_{ \boldsymbol \theta_N}(\boldsymbol y)  }{\min_{\boldsymbol y \in \mathcal{X}^{NM}}P_{ \boldsymbol \theta_N}(\boldsymbol y)} \right] \\
+&= \frac{1}{NM}\log\left[\frac{\max_{\boldsymbol x \in \mathcal{X}^{N}}P_{ \boldsymbol \theta_N}(\boldsymbol x)  }{\min_{\boldsymbol x \in \mathcal{X}^{N}}P_{ \boldsymbol \theta_N}(\boldsymbol x)} \right]^M \equiv \frac{1}{N}\REP_{\boldsymbol X_{N}}( \boldsymbol\theta_N ),
 \end{align*}
-```
-In other words, a model is S-unstable if, given any $C > 0$, there exists an integer $N_C > 0$ so that $\frac{1}{N}\text{ELPR}_N(\boldsymbol \theta_N) > C$ for all $N \ge N_C$.
-A FSFS model formulation may be termed S-stable if it fails to be S-unstable.
-
-This definition of S-unstable is a generalization or reinterpretation of "unstable" used in @schweinberger2011instability by allowing non-exponential family models (e.g.
-RBM and DBM models in Sections \@ref(restricted-boltzmann-machines)-\@ref(deep-learning)) and an increasing number of parameters.
-While this definition differs in form and scope from the original, it does match that in @schweinberger2011instability for the special case of exponential models (cf.
-Section \@ref(discrete-exponential-family-models)) considered there.
-
-S-unstable FSFS model sequences may be undesirable for several reasons. One is that small changes in data can lead to very large changes in probability. Consider, for example, the quantity given by
+where $\REP_{\boldsymbol X_{N}}( \boldsymbol\theta_N )  \equiv \REP( \boldsymbol\theta_N )$ denotes the log-ratio of extremal probabilities for $\boldsymbol X_N$ defined from \@ref(eq:elpr).
+That is, due to iid properties, the sample-size corrected  LREP for $\boldsymbol Y_{N,M}$ equals the analog, $\REP(\boldsymbol\theta_N )/N$, from the underlying common data model for $\boldsymbol X_N$ alone, regardless of the level $M \geq 1$ of independent replication.
+Consequently, the definition of an S-unstable  model  is  unaffected by independent replication and all instability properties  may be characterized by those of one observation  from the common FOES model.
+For computational purposes, this aspect also implies that if the original data $\boldsymbol X_N=(X_1,\ldots,X_N)$ in a FOES model consist of $N$ iid random variables, then the size-scaled log-ratio \@ref(eq:elpr)  may be calculated as
 $$
-\Delta(\boldsymbol \theta_N) \equiv \max \left\{\log \frac{P_{\boldsymbol \theta_N}(\boldsymbol x)}{P_{\boldsymbol \theta_N}(\boldsymbol x^*)} : \boldsymbol x \text{ }\& \text{ } \boldsymbol x^* \in \mathcal{X}^N \text{ differ in exactly one component}\right\},
+\frac{1}{N}\REP( \boldsymbol\theta_N ) \equiv \frac{1}{N}\REP_{\boldsymbol X_{N}}( \boldsymbol\theta_N ) = \log\left[ \frac{\max_{ x \in \mathcal{X}}P_{\boldsymbol\theta_N}(X_1=x)} {\min_{ x \in \mathcal{X}}P_{\boldsymbol\theta_N}(X_1=x)} \right]
 $$
-which represents the biggest log-probability ratio for a one component change in data outcomes at a FSFS parameter $\boldsymbol \theta_N$. We then have the following (non-asymptotic) result.
-
-```{proposition, label="instab-elpr", echo=TRUE}
-Let $\text{ELPR}(\boldsymbol \theta_N)$ be as in (\@ref(eq:elpr)) for an integer $N \ge 1$. For a given $C>0$, if $$\frac{1}{N}\text{ELPR}_N(\boldsymbol \theta_N) > C,$$ 
-then 
-$$\Delta_N(\boldsymbol \theta_N) > C.$$
+based on the extremal probabilities of just one random variable $X_1$.
 ```
 
-Again, if the probability ratio (\@ref(eq:elpr)) is too large, then the FSFS model will exhibit large changes in probability for very small differences in the data configuration, which exemplifies the intuitive notation of instability.
+## Characterizations and consequences of instability
 
-Additionally, S-unstable FSFS model sequences are connected to degenerate models, where model *degeneracy* typically entails placing all probability on a small portion of the sample space.
-For perspective, note that differing sizes of $1/N\cdot\text{ELPR}(\boldsymbol \theta_N)$ in (\@ref(eq:elpr)) may induce a spectrum of levels of "stability" and Proposition \@ref(prp:instab-elpr) indicates increasing sensitivity of model probabilities (i.e., for one component changes in outcomes) as (\@ref(eq:elpr)) increases.
-Furthermore, as the instability measure (\@ref(eq:elpr)) grows, FSFS model sequences are guaranteed to slide into full degeneracy as Proposition \@ref(prp:degenFSFS) shows.
-Define a $\epsilon$-modal set
+As a basic characteristic, S-unstable FOES model sequences  have extremely sensitive probability structures.
+One aspect is that small changes in data configuration can   lead to very large changes in probability.
+Consider, for example, the quantity given by
 $$
-M_{\epsilon, \boldsymbol \theta_N} \equiv \left\{\boldsymbol x \in \mathcal{X}^N: \log P_{\boldsymbol \theta_N}(\boldsymbol x) > (1-\epsilon)\max\limits_{\boldsymbol x^* \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x^*) + \epsilon\min\limits_{\boldsymbol x^* \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x^*) \right\}
+\DN(\boldsymbol \theta_N) \equiv \max \left\{\log \frac{P_{\boldsymbol \theta_N}(\boldsymbol x)}{P_{\boldsymbol \theta_N}(\boldsymbol x^*)} : \boldsymbol x \text{ }\& \text{ } \boldsymbol x^* \in \mathcal{X}^N \text{ differ in exactly one component}\right\},
 $$
-of possible outcomes, for a given $0 < \epsilon < 1$.
+which represents the biggest log-probability ratio for a one component change in data outcomes in a FOES model with parameter $\boldsymbol \theta_N$.
+We then have the following result prescribing the behavior of $\DN(\boldsymbol \theta_N)$ for unstable models.
 
-```{proposition, label="degenFSFS", echo=TRUE}
-For an unstable FSFS model in Definition \@ref(def:instabFSFS), and for any given $0 < \epsilon < 1$, 
+```{theorem, label="instab-elpr", echo=TRUE}
+Let $P_{\boldsymbol \theta_N}$, with support $\mathcal{X}^N$, $N\geq 1$, be a sequence of FOES models.
+
+(i) For any integer $N \geq 1$ and any given $C>0$, if $\REP(\boldsymbol \theta_N)/N > C$ in (\ref{eq:elpr}), then
+    $$ \DN(\boldsymbol \theta_N) > C,$$
+    or probabilities from a one component change in some outcome have log-ratio exceeding $C$.
+(ii) Suppose the FOES model sequence is S-unstable.
+Then, for all large $N$ and given any arbitrary $C>0$, there exist outcomes $\boldsymbol x,\boldsymbol x^*\in\mathcal{X}^N$, differing by one component, such that
+    $$ \frac{P_{\boldsymbol \theta_N}(\boldsymbol x)}{P_{\boldsymbol \theta_N}(\boldsymbol x^*)} > \exp[N C]. $$
+```
+
+Theorem \@ref(instab-elpr)(i) is a  non-asymptotic result, which connects to the definition of instability in a FOES model through a log-ratio of extreme model probabilities \@ref(eq:elpr) being too large relative to the associated sample size $N$.
+If so, Theorem \ref@(instab-elpr)(i) guarantees the FOES model must also exhibit correspondingly large changes in probability for very small differences among some data configurations, a property that intuitively captures a notion of instability.
+Furthermore, and perhaps more seriously under Theorem \@ref(instab-elpr)(ii),
+S-unstable models can never have universally bounded changes in probability among single component variations in data configurations.
+While not all one component changes in data may produce massive changes in probability, unstable models must have some such data outcomes with this property.
+As a consequence, unstable probability structure may exhibit extreme sensitivity through large peaks and troughs over the sample space.
+
+Additionally, S-unstable FOES model sequences are also connected to degenerate models, where {\it degeneracy} involves assigning essentially all probability to modes within the sample space, which could potentially represent a  small subset among the totality of outcomes.
+For perspective, note that differing sizes of the scaled log-ratio $\REP(\boldsymbol \theta_N)/N$ from \@ref(eq:elpr) induce a spectrum of levels of instability/stability and Theorem \@ref(instab-elpr) indicates increasing sensitivity of model probabilities as \@ref(eq:elpr) increases.
+Furthermore, as the instability measure grows and the log-ratio $\REP(\boldsymbol \theta_N)/N$ diverges, as in the definition \@ref(eq:Sun) of S-unstable models, then a FOES model sequence  will become degenerate.
+Theorem \@ref(degenFOES) provides a formal statement of such degeneracy due to instability.
+For a given $0 < \epsilon < 1$, define a $\epsilon$-modal set of   outcomes as
+\begin{equation}
+(\#eq:mode)
+\mathcal{M}_{\epsilon, \boldsymbol \theta_N} \equiv \left\{\boldsymbol x \in \mathcal{X}^N: \log P_{\boldsymbol \theta_N}(\boldsymbol x) > (1-\epsilon)\max\limits_{\boldsymbol y \in \mathcal{X}^N} \log  P_{\boldsymbol \theta_N}(\boldsymbol y) + \epsilon\min\limits_{\boldsymbol y \in \mathcal{X}^N} \log P_{\boldsymbol \theta_N}(\boldsymbol y) \right\}.
+\end{equation}
+
+```{theorem, label="degenFOES", echo=TRUE}
+For any arbitrarily small $0 < \epsilon < 1$, an S-unstable FOES model sequence $P_{\boldsymbol \theta_N}$, $N \geq 1$, for $\boldsymbol X_N=(X_1, \dots, X_N)$ satisfies
 $$
-P_{\boldsymbol \theta_N}\left((x_1, \dots, x_N) \in M_{\epsilon, \boldsymbol \theta_N}\right) \rightarrow 1 \text{ as } N \rightarrow \infty.
+P_{\boldsymbol \theta_N}\left( \boldsymbol X_N\in \mathcal{M}_{\epsilon, \boldsymbol \theta_N}\right) \rightarrow 1 \text{ as } N \rightarrow \infty.
 $$
@@
 
-In other words, in S-unstable FSFS models, all probability in the model formulation with a large number of random variables will concentrate mass on an $\epsilon$-modal set, where $\epsilon$ can be made arbitrarily small.
-The associated mode set could potentially be quite small, in which case Proposition \ref{prp:degenFSFS} would suggest that the unstable model asymptotically stacks all probability on a few outcomes.
+In other words, as the sample size grows in S-unstable FOES models, all probability   tends to concentrate mass on an $\epsilon$-modal set, where $\epsilon$ can be made arbitrarily small.
+Intuitively, the occurrence of such degeneracy can be explained
+by a type of "reverse" pigeonhole principle for unstable FOES models: if all outcomes should receive positive probability but the maximal probability far exceeds the minimal one in the model, then little probability remains for distribution among remaining model outcomes (i.e., if nearly all available pigeons are stuffed into one hole, the remaining pigeonholes must have few occupants).
+Degeneracy in unstable models can pose dangers in data modeling as well, particularly when a mode set represents a narrow collection of outcomes among those realistically possible for adequately describing data.
+In which case, model outcomes may fail to look like data of interest.
 
-```{remark, echo = TRUE}
-There is a further generalization the notion of instability in Definition \@ref(def:instabFSFS) meant to address independent replications of data sequences.
-That is, one might consider data as $n$ independent and identically distributed replications $\boldsymbol X_1, \dots, \boldsymbol X_n$, where each $\boldsymbol X_i=(X_{i,1},\ldots,X_{i,N}) \in \mathcal{X}^N$ follows a common FSFS model with probabilities $P_{\theta_N}(\boldsymbol x)>0$,  $\boldsymbol x\in\mathcal{X}^N$ and $|\mathcal{X}|<\infty$, for $i=1,\ldots,n$.
-This leads to a total of $n*N$ random variables in the joint model.
-However, the definition of S-unstable and Propositions \@ref(prp:instab-elpr)-\@ref(prp:degenFSFS) still hold for such iid replications.
-This is because $\left(\max\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)\right)^n$ is the largest probability possible under the joint model for the $n$ replications while $\left(\min\limits_{\boldsymbol x \in \mathcal{X}}P_{\boldsymbol \theta_N}(\boldsymbol x)\right)^n$ is the smallest probability.
-Thus, for the combined replications $\boldsymbol X_1, \ldots, \boldsymbol X_n$, the analog definition of the extremal log-probability becomes
-\begin{align*}
-\frac{\text{extremal log-probability ratio}}{\text{\# random variables in the model}} &\equiv \frac{1}{n*N}\log \left[\frac{\left(\max\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)\right)^n}{\left(\min\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)\right)^n}\right] = \frac{1}{N} \log \left[\frac{\max\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)}{\min\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)}\right] \\
-&= \frac{1}{N} \text{ELPR}(\boldsymbol \theta_N),
-\end{align*}
-implying that the definition of an S-unstable FSFS model sequence is invariant to the level ($n$) of independent replication.
-Consequently, overall model instabilities may be characterized by those of one observation from the common FSFS model.
+Connected to degeneracy, unstable models may also exhibit additional kinds of  extreme and undesirable sensitivity in probabilities if model parameters $\boldsymbol \theta_N$ can further be "dialed" between positive and negative values.
+That is, some FOES models naturally involve parameter spaces covering a positive-negative spectrum of parameter possibilities, where the signs of parameters provide a standard device for increasing or decreasing probabilities of outcomes in the model formulation.
+In fact, for many models, the switch of a parameter sign serves to produce reciprocal probabilities, as outlined in the following model assumption about parameter sign reversal (PSR).
+
+```{mcpsr, name="Reciprocal Probabilities from Parameter Sign Reversal", echo=TRUE}
+Let $P_{\boldsymbol \theta_N}$, with support $\mathcal{X}^N$, $N\geq 1$, represent a sequence of FOES models.
+For each $N \geq 1$ and any outcome $\boldsymbol x \in \mathcal{X}^N$, suppose it holds that
+$$
+P_{\boldsymbol \theta_N}(\boldsymbol x)  \cdot P_{-\boldsymbol \theta_N}(\boldsymbol x) \;=\;   \max\limits_{\boldsymbol y \in \mathcal{X}^N}P_{ \boldsymbol \theta_N}(\boldsymbol y)\cdot \min\limits_{\boldsymbol y \in \mathcal{X}^N}P_{-\boldsymbol \theta_N}(\boldsymbol y),
+$$
+where $\max_{\boldsymbol y \in \mathcal{X}^N}P_{ \boldsymbol \theta_N}(\boldsymbol y)$ and $\min_{\boldsymbol y \in \mathcal{X}^N}P_{-\boldsymbol \theta_N}(\boldsymbol y)$ denote the maximum and minimum probabilities under parameters $\boldsymbol \theta_N$   and $-\boldsymbol \theta_N$, respectively.
 ```
+
+The above model condition incorporates many standard parameterizations and follows, for instance, whenever $P_{\boldsymbol \theta_N}(\boldsymbol x)/P_{\boldsymbol \theta_N}(\boldsymbol y)  = [P_{-\boldsymbol \theta_N}(\boldsymbol x)/P_{-\boldsymbol \theta_N}(\boldsymbol y)]^{-1}$ holds for outcomes $\boldsymbol x, \boldsymbol y \in\mathcal{X}^N$ in a FOES model.
+For instance, this latter condition is fulfilled for all linear exponential families from Section \@ref(discrete-exponential-family-models) (e.g., \@ref(eq:mod1)-\@ref(eq:mod2)) as well as all network models from Sections \@ref(restricted-boltzmann-machines)-\@ref(deep-learning) (e.g., \@ref(eq:RBM1)-\@ref(eq:RBM2)).
+When parameters can be tuned in sign with effects prescribed in the model condition PSR, unstable FOES models will exhibit further probability sensitivities, as outlined in the following extension of Theorem \@ref(degenFOES}).
+
+```{corollary, label="sign", echo=TRUE}
+Let $P_{\boldsymbol \theta_N}$, with support $\mathcal{X}^N$, $N\geq 1$, be a sequence of FOES models satisfying model condition PSR.
+If the models $P_{\boldsymbol \theta_N}$  are additionally S-unstable, then
+
+(i) the models  $P_{-\boldsymbol \theta_N}$  defined by $-\boldsymbol \theta_N$  are also S-unstable;
+(ii) and for the complement $\mathcal{M}_{\epsilon, \boldsymbol \theta_N}^c \equiv \mathcal{X}^N \setminus \mathcal{M}_{\epsilon, \boldsymbol \theta_N}$ of any mode-set $\mathcal{M}_{\epsilon, \boldsymbol \theta_N} $ under $\boldsymbol \theta_N$ from (\ref{eqn:mode}), with $0<\epsilon<1$, it holds under $- \boldsymbol \theta_N$ that
+    $$ P_{-\boldsymbol \theta_N}( \boldsymbol X_N \in \mathcal{M}_{\epsilon, \boldsymbol \theta_N}^c) \rightarrow 1\qquad \mbox{as $N\to \infty$},$$
+    while, by Theorem \@ref(degenFOES), $P_{\boldsymbol \theta_N}( \boldsymbol X_N \in \mathcal{M}_{\epsilon, \boldsymbol \theta_N} ) \rightarrow 1$ holds for $ \boldsymbol X_N = (X_1,\ldots,X_N)$ under $  \boldsymbol \theta_N$.
+```
+
+For unstable models, Corollary \@ref(sign) shows that shifts in parameters around zero (i.e., from $\boldsymbol \theta_N$ to $-\boldsymbol \theta_N$) can induce extreme changes in probability among subsets of the sample space, as another manifestation  of instability and hyper-sensitivity in probability structure.
+For one-parameter exponential families, involving a fixed real-valued linear parameter  $\boldsymbol \theta_N = \theta \in \mathbb{R}$ and sufficient statistic $\boldsymbol g_N(\boldsymbol x)\in \mathbb{R}$ in \@ref(eq:expo), @schweinberger2011instability [Theorem 3] proved a result similar in spirit, though based on a characterization there in terms of maximum $U_N \equiv \max_{\boldsymbol x\in\mathcal{X}^N}g_N(\boldsymbol x)$ and minimal $L_N \equiv \min_{\boldsymbol x\in\mathcal{X}^N}g_N(\boldsymbol x)$ values of the sufficient statistic.
+For this case in particular,  mode sets have specific, and essentially complementary, forms over positive and negative parameters, namely, $\mathcal{M}_{\epsilon, \boldsymbol \theta_N} = \{\boldsymbol x \in\mathcal{X}^N: g_N(\boldsymbol x)  >  (1-\epsilon) U_N + \epsilon L_N \}$ and $\mathcal{M}_{\epsilon, -\boldsymbol \theta_N} = \{\boldsymbol x \in\mathcal{X}^N: g_N(\boldsymbol x)  <  \epsilon U_N + (1-\epsilon) L_N \}$ for any $\boldsymbol \theta_N>0$, and
+ @schweinberger2011instability [Theorem 3] showed each mode set collects all mass, under positive and negative parameters, respectively, with unstable models of this exponential type.
+However, for all unstable FOES models, Corollary \@ref(sign) generalizes the same principle that unstable models can push all probability to different, and in fact disjoint, parts of the sample space, depending on how parameters fall with respect to zero.
+This feature can numerically complicate likelihood manipulations, such as maximization or MCMC-based Bayes posterior sampling, as further discussed in Section \@ref(statistical-consequences).
 

### 2017-09-27 writing/paper-note/docs/note.tex
some changes in the note from nordman

@@
 increasing numbers of variables (i.e., for \(N \rightarrow \infty\)),
-the ratio (\eqref{eq:elpr}) should remain bounded to ensure model
-stability, requiring that the largest probability possible under
-\(P_{\theta_N}\) maintain a fixed order of magnitude relative to the
-smallest probability allowed under the same model. Specifically, the log
-of the ratio should grow at mostly linearly with the sample size \(N\).
-This leads to the following definition.
+the ratio \eqref{eq:elpr} should remain bounded to ensure model stability,
+requiring that the largest probability possible under \(P_{\theta_N}\)
+maintain a fixed order of magnitude relative to the smallest probability
+allowed under the same model. Specifically, the log of the ratio should
+grow at mostly linearly with the sample size \(N\). This leads to the
+following definition.
 \BeginKnitrBlock{definition}[S-unstable FSFS]
@@
 (\protect\hyperlink{ref-schweinberger2011instability}{2011}) by allowing
-non-exponential family models (e.g.~RBM and DBM models in Sections
+non-exponential family models (e.g., RBM and DBM models in Sections
 \ref{restricted-boltzmann-machines}-\ref{deep-learning}) and an
@@
 \protect\hypertarget{prp:instab-elpr}{}{\label{prp:instab-elpr}}Let
-\(\text{ELPR}(\boldsymbol \theta_N)\) be as in (\eqref{eq:elpr}) for an
+\(\text{ELPR}(\boldsymbol \theta_N)\) be as in \eqref{eq:elpr} for an
 integer \(N \ge 1\). For a given \(C>0\), if
@@
 
-Again, if the probability ratio (\eqref{eq:elpr}) is too large, then the
+Again, if the probability ratio \eqref{eq:elpr} is too large, then the
 FSFS model will exhibit large changes in probability for very small
@@
 perspective, note that differing sizes of
-\(1/N\cdot\text{ELPR}(\boldsymbol \theta_N)\) in (\eqref{eq:elpr}) may
+\(1/N\cdot\text{ELPR}(\boldsymbol \theta_N)\) in \eqref{eq:elpr} may
 induce a spectrum of levels of ``stability'' and Proposition
@@
 probabilities (i.e., for one component changes in outcomes) as
-(\eqref{eq:elpr}) increases. Furthermore, as the instability measure
-(\eqref{eq:elpr}) grows, FSFS model sequences are guaranteed to slide into
+\eqref{eq:elpr} increases. Furthermore, as the instability measure
+\eqref{eq:elpr} grows, FSFS model sequences are guaranteed to slide into
 full degeneracy as Proposition \ref{prp:degenFSFS} shows. Define a
@@
 \(0 < \epsilon < 1\), \[
-P_{\boldsymbol \theta_N}\left((x_1, \dots, x_N) \in M_{\epsilon, \boldsymbol \theta_N}\right) \rightarrow 1 \text{ as } N \rightarrow \infty.
+P_{\boldsymbol \theta_N}\left((X_1, \dots, X_N) \in M_{\epsilon, \boldsymbol \theta_N}\right) \rightarrow 1 \text{ as } N \rightarrow \infty.
 \]
@@
 
-\textbf{Proof of Proposition \ref{prp:degenFSFS}.} where
-\(|\mathcal{X}|<\infty\) holds in the FSFS model. We may suppose
+\textbf{Proof of Proposition \ref{prp:degenFSFS}.} Where
+\(|\mathcal{X}|<\infty\) holds in the FSFS model, we may suppose
 \(|\mathcal{X}|>1\) (i.e., \(\mathcal{X}^N\) has more than one outcome)
@@
 \end{align*}
-as \(N \rightarrow \infty\) by the definition of an unstable FSFS model
-(cf.~Definition \ref{def:instabFSFS}). Consequently,
+as \(N \rightarrow \infty\) by the definition of an S-unstable FSFS
+model (cf.~Definition \ref{def:instabFSFS}). Consequently,
 \(P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N}) \rightarrow 1\)

### 2017-09-30 writing/paper-note/chapters/instab.Rmd
what a mess

@@
 based on maximum and minimal model probabilities.
-In what follows, the main idea is that instability, and other negative model features, can be associated with a FOES model formulation for $N$ random variables where the log-ratio (\ref{eq:elpr}) is overly large relative to the sample size $N$.
+In what follows, the main idea is that instability, and other negative model features, can be associated with a FOES model formulation for $N$ random variables where the log-ratio \@ref(eq:elpr) is overly large relative to the sample size $N$.
 That is, a sequence of FOES probability models $P_{\boldsymbol \theta_N}$ results in specifying the distribution of observations $\boldsymbol X=(X_1,\ldots,X_N)$ for each sample size $N \geq 1$ and instability will generally occur among these models whenever the corresponding LREP \@ref(eq:elpr) grows faster than $N$.
@@
 In other words, a model is S-unstable if $\REP(\boldsymbol \theta_N)/N$ is an unbounded sequence of sample size $N$; namely, given any $C > 0$, there exists an integer $N_C > 0$ so that $\REP(\boldsymbol \theta_N)/N > C$ holds for all $N \ge N_C$.
-A FOES model formulation may be termed S-stable if it fails to be S-unstable, i.e., if $\sup_{N \geq 1}\REP(\boldsymbol \theta_N)/N $ is bounded.
+A FOES model formulation may be termed S-stable if it fails to be S-unstable, i.e., if $\sup_{N \geq 1}\REP(\boldsymbol \theta_N)/N$ is bounded.
 
@@
 Section \@ref(discrete-exponential-family-models) considered there.
-Section \ref@(illustrations) provides several examples  of unstable models as well as causes for model instability, where the latter may often be traced to issues in model form (i.e., data functions) and/or parameterization.
+Section \@ref(illustrations) provides several examples  of unstable models as well as causes for model instability, where the latter may often be traced to issues in model form (i.e., data functions) and/or parameterization.
 We next describe several potentially undesirable features associated with unstable models.
 
-```{remark, echo=TRUE}
+```{remark, label="rmk1", echo=TRUE}
 In the definition \@ref(eq:Sun) of S-instability, we note that the numerical measure $\REP(\boldsymbol \theta_N)/N$ of model instability is invariant to *independent  replications* of data.
@@
 Let $P_{\boldsymbol \theta_N}$, with support $\mathcal{X}^N$, $N\geq 1$, be a sequence of FOES models.
-
-(i) For any integer $N \geq 1$ and any given $C>0$, if $\REP(\boldsymbol \theta_N)/N > C$ in (\ref{eq:elpr}), then
+\begin{enumerate}[(i)]
+\item For any integer $N \geq 1$ and any given $C>0$, if $\REP(\boldsymbol \theta_N)/N > C$ in (\ref{eq:elpr}, then
     $$ \DN(\boldsymbol \theta_N) > C,$$
     or probabilities from a one component change in some outcome have log-ratio exceeding $C$.
-(ii) Suppose the FOES model sequence is S-unstable.
+\item Suppose the FOES model sequence is S-unstable.
 Then, for all large $N$ and given any arbitrary $C>0$, there exist outcomes $\boldsymbol x,\boldsymbol x^*\in\mathcal{X}^N$, differing by one component, such that
-    $$ \frac{P_{\boldsymbol \theta_N}(\boldsymbol x)}{P_{\boldsymbol \theta_N}(\boldsymbol x^*)} > \exp[N C]. $$
+    $$
+    \frac{P_{\boldsymbol \theta_N}(\boldsymbol x)}{P_{\boldsymbol \theta_N}(\boldsymbol x^*)} > \exp[N C].
+    $$
+\end{enumerate}
 ```
 
-Theorem \@ref(instab-elpr)(i) is a  non-asymptotic result, which connects to the definition of instability in a FOES model through a log-ratio of extreme model probabilities \@ref(eq:elpr) being too large relative to the associated sample size $N$.
-If so, Theorem \ref@(instab-elpr)(i) guarantees the FOES model must also exhibit correspondingly large changes in probability for very small differences among some data configurations, a property that intuitively captures a notion of instability.
-Furthermore, and perhaps more seriously under Theorem \@ref(instab-elpr)(ii),
+Theorem \@ref(thm:instab-elpr)(i) is a non-asymptotic result, which connects to the definition of instability in a FOES model through a log-ratio of extreme model probabilities \@ref(eq:elpr) being too large relative to the associated sample size $N$.
+If so, Theorem \@ref(thm:instab-elpr)(i) guarantees the FOES model must also exhibit correspondingly large changes in probability for very small differences among some data configurations, a property that intuitively captures a notion of instability.
+Furthermore, and perhaps more seriously under Theorem \@ref(thm:instab-elpr)(ii),
 S-unstable models can never have universally bounded changes in probability among single component variations in data configurations.
@@
 
-Additionally, S-unstable FOES model sequences are also connected to degenerate models, where {\it degeneracy} involves assigning essentially all probability to modes within the sample space, which could potentially represent a  small subset among the totality of outcomes.
-For perspective, note that differing sizes of the scaled log-ratio $\REP(\boldsymbol \theta_N)/N$ from \@ref(eq:elpr) induce a spectrum of levels of instability/stability and Theorem \@ref(instab-elpr) indicates increasing sensitivity of model probabilities as \@ref(eq:elpr) increases.
+Additionally, S-unstable FOES model sequences are also connected to degenerate models, where *degeneracy* involves assigning essentially all probability to modes within the sample space, which could potentially represent a  small subset among the totality of outcomes.
+For perspective, note that differing sizes of the scaled log-ratio $\REP(\boldsymbol \theta_N)/N$ from \@ref(eq:elpr) induce a spectrum of levels of instability/stability and Theorem \@ref(thm:instab-elpr) indicates increasing sensitivity of model probabilities as \@ref(eq:elpr) increases.
 Furthermore, as the instability measure grows and the log-ratio $\REP(\boldsymbol \theta_N)/N$ diverges, as in the definition \@ref(eq:Sun) of S-unstable models, then a FOES model sequence  will become degenerate.
-Theorem \@ref(degenFOES) provides a formal statement of such degeneracy due to instability.
+Theorem \@ref(thm:degenFOES) provides a formal statement of such degeneracy due to instability.
 For a given $0 < \epsilon < 1$, define a $\epsilon$-modal set of   outcomes as
@@
 
-```{mcpsr, name="Reciprocal Probabilities from Parameter Sign Reversal", echo=TRUE}
-Let $P_{\boldsymbol \theta_N}$, with support $\mathcal{X}^N$, $N\geq 1$, represent a sequence of FOES models.
+**Model Condition PSR** *(Reciprocal Probabilities from Parameter Sign Reversal)*: Let $P_{\boldsymbol \theta_N}$, with support $\mathcal{X}^N$, $N\geq 1$, represent a sequence of FOES models.
 For each $N \geq 1$ and any outcome $\boldsymbol x \in \mathcal{X}^N$, suppose it holds that
@@
 where $\max_{\boldsymbol y \in \mathcal{X}^N}P_{ \boldsymbol \theta_N}(\boldsymbol y)$ and $\min_{\boldsymbol y \in \mathcal{X}^N}P_{-\boldsymbol \theta_N}(\boldsymbol y)$ denote the maximum and minimum probabilities under parameters $\boldsymbol \theta_N$ and $-\boldsymbol \theta_N$, respectively.
-```
 
@@
 For instance, this latter condition is fulfilled for all linear exponential families from Section \@ref(discrete-exponential-family-models) (e.g., \@ref(eq:mod1)-\@ref(eq:mod2)) as well as all network models from Sections \@ref(restricted-boltzmann-machines)-\@ref(deep-learning) (e.g., \@ref(eq:RBM1)-\@ref(eq:RBM2)).
-When parameters can be tuned in sign with effects prescribed in the model condition PSR, unstable FOES models will exhibit further probability sensitivities, as outlined in the following extension of Theorem \@ref(degenFOES}).
+When parameters can be tuned in sign with effects prescribed in the model condition PSR, unstable FOES models will exhibit further probability sensitivities, as outlined in the following extension of Theorem \@ref(thm:degenFOES}).
 
@@
 If the models $P_{\boldsymbol \theta_N}$  are additionally S-unstable, then
-
-(i) the models  $P_{-\boldsymbol \theta_N}$  defined by $-\boldsymbol \theta_N$  are also S-unstable;
-(ii) and for the complement $\mathcal{M}_{\epsilon, \boldsymbol \theta_N}^c \equiv \mathcal{X}^N \setminus \mathcal{M}_{\epsilon, \boldsymbol \theta_N}$ of any mode-set $\mathcal{M}_{\epsilon, \boldsymbol \theta_N} $ under $\boldsymbol \theta_N$ from (\ref{eqn:mode}), with $0<\epsilon<1$, it holds under $- \boldsymbol \theta_N$ that
-    $$ P_{-\boldsymbol \theta_N}( \boldsymbol X_N \in \mathcal{M}_{\epsilon, \boldsymbol \theta_N}^c) \rightarrow 1\qquad \mbox{as $N\to \infty$},$$
-    while, by Theorem \@ref(degenFOES), $P_{\boldsymbol \theta_N}( \boldsymbol X_N \in \mathcal{M}_{\epsilon, \boldsymbol \theta_N} ) \rightarrow 1$ holds for $ \boldsymbol X_N = (X_1,\ldots,X_N)$ under $  \boldsymbol \theta_N$.
+\begin{enumerate}[(i)]
+\item the models  $P_{-\boldsymbol \theta_N}$  defined by $-\boldsymbol \theta_N$  are also S-unstable;
+\item and for the complement $\mathcal{M}_{\epsilon, \boldsymbol \theta_N}^c \equiv \mathcal{X}^N \setminus \mathcal{M}_{\epsilon, \boldsymbol \theta_N}$ of any mode-set $\mathcal{M}_{\epsilon, \boldsymbol \theta_N}$ under $\boldsymbol \theta_N$ from (\ref{eq:mode}), with $0<\epsilon<1$, it holds under $- \boldsymbol \theta_N$ that
+    $$ 
+    P_{-\boldsymbol \theta_N}( \boldsymbol X_N \in \mathcal{M}_{\epsilon, \boldsymbol \theta_N}^c) \rightarrow 1\qquad \mbox{as $N\to \infty$},
+    $$
+    while, by Theorem \ref{thm:degenFOES}, $P_{\boldsymbol \theta_N}( \boldsymbol X_N \in \mathcal{M}_{\epsilon, \boldsymbol \theta_N} ) \rightarrow 1$ holds for $\boldsymbol X_N = (X_1,\ldots,X_N)$ under $\boldsymbol \theta_N$.
+\end{enumerate}  
 ```
 
-For unstable models, Corollary \@ref(sign) shows that shifts in parameters around zero (i.e., from $\boldsymbol \theta_N$ to $-\boldsymbol \theta_N$) can induce extreme changes in probability among subsets of the sample space, as another manifestation  of instability and hyper-sensitivity in probability structure.
+For unstable models, Corollary \@ref(cor:sign) shows that shifts in parameters around zero (i.e., from $\boldsymbol \theta_N$ to $-\boldsymbol \theta_N$) can induce extreme changes in probability among subsets of the sample space, as another manifestation  of instability and hyper-sensitivity in probability structure.
 For one-parameter exponential families, involving a fixed real-valued linear parameter  $\boldsymbol \theta_N = \theta \in \mathbb{R}$ and sufficient statistic $\boldsymbol g_N(\boldsymbol x)\in \mathbb{R}$ in \@ref(eq:expo), @schweinberger2011instability [Theorem 3] proved a result similar in spirit, though based on a characterization there in terms of maximum $U_N \equiv \max_{\boldsymbol x\in\mathcal{X}^N}g_N(\boldsymbol x)$ and minimal $L_N \equiv \min_{\boldsymbol x\in\mathcal{X}^N}g_N(\boldsymbol x)$ values of the sufficient statistic.
@@
  @schweinberger2011instability [Theorem 3] showed each mode set collects all mass, under positive and negative parameters, respectively, with unstable models of this exponential type.
-However, for all unstable FOES models, Corollary \@ref(sign) generalizes the same principle that unstable models can push all probability to different, and in fact disjoint, parts of the sample space, depending on how parameters fall with respect to zero.
-This feature can numerically complicate likelihood manipulations, such as maximization or MCMC-based Bayes posterior sampling, as further discussed in Section \@ref(statistical-consequences).
+However, for all unstable FOES models, Corollary \@ref(cor:sign) generalizes the same principle that unstable models can push all probability to different, and in fact disjoint, parts of the sample space, depending on how parameters fall with respect to zero.
+This feature can numerically complicate likelihood manipulations, such as maximization or MCMC-based Bayes posterior sampling, as further discussed in Section \@ref(implications).
 
-# Implications
+```{remark, name="need to briefly finish", echo=TRUE}
+extensions of sensitivity...
+basically, model condition PSR and Corollary \ref{cor:sign} can be extended to cases where parameter components $\boldsymbol \theta_N =(\boldsymbol \theta_{1,N}, \boldsymbol \theta_{N,2})$ are not all changed in sign ($-\boldsymbol \theta_N$) but, more generally, are instead altered to another parameter configuration $\boldsymbol \theta_{N}^A = ( \boldsymbol \theta_{1,N}^A, \boldsymbol \theta_{2,N}^A )$ involving a switch in sign only among some model parameters $\boldsymbol \theta_{N,2}^A=- \boldsymbol \theta_{N,2}$ with remaining parameters $\boldsymbol \theta_{1,N}^A$ being arbitrarily chosen.
+If a change sign occurs among parameters ($\pm \boldsymbol \theta_{N,2}$) which dominate the probability structure of the model, then an unstable model under $\boldsymbol \theta_N$ will then imply that many more unstable models exist  over a broad spectrum of possibilities for variations $\boldsymbol \theta_{N}^A$ of $\boldsymbol \theta_N$.
+```
 
-For a large class of models that covers a broad range of applications (including "deep learning"), we have developed a formal definition of instability in model probability structure and elucidated multiple consequences of instability.
-We have shown for FSFS models that instability manifests through small changes in data leading to potentially large changes in probability as well as the potential to place all probability on certain modal subsections of the sample space, which potentially could be small.
-The FSFS model class is quite broad and, particularly in developing FSFS models for large data sets, some caution should be used in parameter specification to control effects of model instability.
+# Illustrations
+
+Model instability can depend intricately on how functions of parameters and data $\boldsymbol X_N=(X_1,\ldots,X_n)$ are combined in the formulation of the model probabilities, though some general causes may be identified.
+As one issue, a broad parameter space (or wide interpretation of this space) may admit some parameters as technically valid that have an undue and often undesirable impact on the model structure for a prescribed data size $N$.
+In this case, both the size and dimension of model parameters can be problematic and induce instability.
+In combination to this last point, further causes of instability may also be traced to the magnitude of statistics in the model.
+Potentially massive, and thereby unstable, statistics were the primary focus of instability studies of @schweinberger2011instability for certain discrete exponential models having parameters/statistics of fixed dimension.
+However, as shown in the following, bounded statistics may still lead to instability if the parameter dimension is high.
+We next provides some examples to illustrate instability in FOES models, which also suggest  some potential strategies for preventing unstable models.
+
+## Equi-probability models
+
+As a baseline for comparisons, consider a simplistic model for $\boldsymbol X_N=(X_1,\ldots,X_N)$ with uniform probabilities over the sample space, say $P_{\boldsymbol \theta_N}(\boldsymbol x)= |\mathcal{X}|^{-N}$, $\boldsymbol x \in \mathcal{X}^N$, where each random variable has $|\mathcal{X}| \geq 1$ outcomes.
+In contrast to instability, model probabilities here are completely insensitive to changes in data outcomes across the sample space, and the associated log-ratio of extreme probabilities \@ref(eq:elpr) is
+$$
+\frac{1}{N} \REP(\boldsymbol \theta_N)=0\quad \;(\mbox{uniform probability model}),
+$$ 
+which is as small as possible. In fact, a LREP value of zero can only occur for a FOES model having uniform probabilities, and such equi-probability models are always S-stable.
+
+## One-parameter exponential models {#one-param-exp}
+
+A fundamental model considered in the instability work of @schweinberger2011instability involves a one-parameter exponential model corresponding to \@ref(eq:expo) with a real-valued parameter, say $\boldsymbol \theta_N =  \eta(\boldsymbol \theta_N)\in \mathbb{R}$, and sufficient statistic $\boldsymbol g_N(\boldsymbol x)\in \mathbb{R}$.
+For such models,  upon scaling by sample size $N$, the log-ratio of extreme probabilities in \@ref(eq:elpr) for assessing instability becomes
+\begin{equation}
+(\#eq:UL)
+\frac{1}{N}\REP(\boldsymbol \theta_N ) \equiv   |\boldsymbol \theta_N| \frac{(U_N-L_N)}{N} \;\quad \mbox{(one-parameter exponential model)},
+\end{equation}
+where $U_N \equiv \max_{\boldsymbol x\in\mathcal{X}^N}g_N(\boldsymbol x)$  and $L_N \equiv \min_{\boldsymbol x\in\mathcal{X}^N}g_N(\boldsymbol x)$ denote the maximal and minimal values of the single sufficient statistic.
+In this case, an S-unstable model results, by definition \@ref(eq:Sun), whenever $\lim_{N\to \infty} |\boldsymbol \theta_N| (U_N-L_N)/N= \infty$ holds or, in other words, if the combined  magnitudes of parameter $|\boldsymbol \theta_N|$ and maximal difference $U_N-L_N$ in statistic values are overwhelmingly large relative to the sample size $N$.
+If we further assume that $\boldsymbol \theta_N =\theta\in\mathbb{R}\setminus \{0\}$ is a fixed (non-zero) parameter for all $N \geq 1$, as considered in @schweinberger2011instability, then an S-unstable model results solely if the sufficient statistic admits a value $U_N-L_N$ too large relative to number $N$ of observations, i.e., if $(U_N-L_N)/N\to \infty$ as $N\to \infty$.
+The latter aspect reflects the definition of @schweinberger2011instability, for this setting, that a real-valued *statistic* $g_N(\boldsymbol x)$ may be classified as *unstable* when  $\lim_{N\to \infty}|(U_N-L_N)/N=\infty$ holds and as *stable* otherwise (e.g., if $\sup_{N \geq 1}(U_N-L_N)/N<\infty$).
+
+For illustration, consider the iid Bernoulli model \@ref(eq:mod1) for $\boldsymbol X_N=(X_1,\ldots,X_N)$ with log-odds ratio parameter $\boldsymbol \theta_N = \log[ P_{\boldsymbol \theta_N}(X_1=1)/ P_{\boldsymbol \theta_N}(X_1=0)]\in\mathbb{R}$.
+Remark \@ref(rmk1) (Section \@ref(criterion)) then gives the model instability measure \@ref(eq:Sun) directly as
+$$
+\frac{1}{N}\REP(\boldsymbol \theta_N ) = |\boldsymbol \theta_N|\quad\; \mbox{(iid Bernoulli model)},
+$$
+so that an unstable (or stable) model results for a divergent (or bounded) parameter sequence  $|\boldsymbol \theta_N|$.
+The above instability expression for the Bernoulli model follows as well from  the $N$-scaled LREP value \@ref(eq:UL) for a one-parameter exponential distribution, using that the sufficient statistic involved $g_N(\boldsymbol x)= \sum_{i=1}^N x_i$, $\boldsymbol x =(x_1,\ldots,x_n)\in\{0,1\}^N$, has maximum and mimimum values $U_N=N$ and $L_N=0$.
+In this case, @schweinberger2011instability has noted that the statistic is stable (i.e., bounded $(U_N-L_N)/N=1$) and the Bernoulli model is as well when, in particular, $\boldsymbol \theta_N=\bold \theta \in\mathbb{R}$ is fixed for $N \geq 1$.
+
+Alternatively, considering a random graph with $N={n \choose 2}$ edges among $n$ nodes, the exponential graph model from \@ref(eq:mod2), when based purely on the number of $g_{2,N}(\boldsymbol x)$ of 2-stars or solely the number $g_{3,N}(\boldsymbol x)$  of triangles, $\boldsymbol x\in\{0,1\}^N$, has an measure of instability from \@ref(eq:Sun) as
+$$
+\frac{1}{N}\REP(\boldsymbol \theta_N )  = \left\{ \begin{array}{lcl} |\boldsymbol \theta_N| (n-2) && \mbox{(2-star graph model)}\\
+|\boldsymbol \theta_N|(n-2)/3 &&\mbox{(triangle graph model)},\end{array}\right.
+$$
+by using the (one-parameter exponential) LREP formula \@ref(eq:UL) with statistic maximums $U_N= N(n-2)$ for 2-stars or $U_N= N(n-2)/3$ for triangles  and with minimums $L_N=0$ in both cases.
+Because the variable number $N\to \infty$ as the node number $n\to \infty$, both counts of 2-stars and triangles are unstable statistics in the sense of @schweinberger2011instability (i.e., $\lim_{N\to \infty} (U_N-L_N)/N=\infty$).
+Furthermore, both types of graph models are always S-unstable for all possible of parameter sequences $\boldsymbol \theta_N \in\mathbb{R}$ that are bounded away from zero (i.e.,  $\lim_{N\to \infty}\REP(\boldsymbol \theta_N )/N=\infty$ then  holds, including the fixed parameter case $\boldsymbol \theta_N=\theta\in\mathbb{R}\setminus \{0\}$  from @schweinberger2011instability).
+
+## Fixed-dimensional linear exponential models {#fixed-dim-exp}
+
+As a generalization of the one-parameter exponential case, we next consider linear exponential families \@ref(eq:expo) with $k$ parameters $\boldsymbol \theta_N = (\theta_{1,N},\ldots,\theta_{k,N})^\prime$ and $k$ sufficient statistics $\boldsymbol g_N(\boldsymbol x) = (g_{1,N}(\boldsymbol x),\ldots, g_{k,N}(\boldsymbol x))^\prime$.
+Here the dimension $k$ of model parameters/statistics is fixed, and we next prescribe a condition helpful to avoiding instability in such models.
+For this, define  $U_{i,N}=\max_{\boldsymbol x \in\mathcal{X}^N} g_{i,N}(\boldsymbol x)$ and $L_{i,N}=\min_{\boldsymbol x \in\mathcal{X}^N} g_{i,N}(\boldsymbol x)$ as the maximal and minimal values of the $i$th statistic, $i=1,\ldots,k$, based on observations $\boldsymbol X_N=(X_1,\ldots,X_N)$.
+
+```{proposition, label="prop1", echo=TRUE}
+Let $P_{\boldsymbol \theta_N}$, $N \geq 1$,  denote linear exponential models (\ref{eq:expo}) with   parameters $\boldsymbol \theta_N = (\theta_{1,N},\ldots,\theta_{k,N})^\prime \in \mathbb{R}^k$ and statistics  $\boldsymbol g_N(\boldsymbol x) = (g_{1,N}(\boldsymbol x),\ldots, g_{k,N}(\boldsymbol x))^\prime \in \mathbb{R}^k$, for fixed $k \geq 1$.
+Then, the models $P_{\boldsymbol \theta_N}$ are {\rm S-stable}   if
+\begin{equation}
+(\#eq:prop1)
+\sup_{N \geq 1}\frac{1}{N} \max_{1 \leq i \leq k }|\theta_{i,N}|(U_{i,N}-L_{i,N})<\infty
+\end{equation}
+holds, i.e.,  if $\max_{1 \leq i \leq k } |\theta_{i,N}|(U_{i,N}-L_{i,N})/N$ is bounded sequence of sample size $N$.
+```
+
+```{remark, echo=TRUE}
+In the one-parameter exponential case $k=1$, recall the exponential model is stable/unstable depending on whether $\REP(\boldsymbol \theta_N)/N = |\theta_{1,N}|(U_{1,N}-L_{1,N})/N \equiv |\boldsymbol \theta_{N}|(U_{N}-L_{N})/N$ in \@ref(eq:UL) is  convergent/divergent.
+Hence, for $k=1$, the condition \@ref(eq:prop1) of Proposition \@ref(prp:prop1) captures the same notion of S-stability based on \@ref(eq:UL).
+```
+
+Proposition \@ref(prp:prop1) provides a sufficient condition for the stability of linear exponential models with fixed parameter dimension $k\geq 1$, whereby an S-stable model is guaranteed if the compounded magnitude of each combination of parameter $\theta_{i,N}$ and sufficient statistic value $(U_{i,N}-L_{i,N})$  is bounded by the sample size $N$, $i=1,\ldots,k$.
+This  supports the findings of @schweinberger2011instability, who showed degeneracy follows in such models under one type of violation of the condition \@ref(eq:prop1) in Proposition \@ref(prp:prop1) (namely, involving $k>1$ non-zero parameters  with $k-1$ statistics being $O(N)$ bounded while one statistic diverges in maximal size faster than the number $N$ of observations).
+To further illustrate the result in Proposition \@ref(prp:prop1), consider the multinomial distribution \@ref(eq:mod11) for $\boldsymbol X_N=(X_1,\ldots,X_N)$ having $k\geq 2$ categories $\{1,\ldots,k\}$ and $k$ parameters $\boldsymbol \theta_N = (\theta_{1,N},\ldots,\theta_{k,N})^\prime$.
+The variables are iid under this model so that Remark \@ref(rmk1) (Section \@ref(criterion) yields the corresponding $N$-scaled  log-ratio of extreme probabilities \@ref(eq:elpr) as
+\begin{align*}
+  \frac{1}{N}\REP(\boldsymbol \theta_N) &= \frac{\max_{1 \leq i \leq k} P_{\boldsymbol \theta_N}(X_1=i)}{\min_{1 \leq i \leq k} P_{\boldsymbol \theta_N}(X_1=i)}\\
+  &= \max_{1 \leq i \leq k} \theta_{i,N} - \min_{1 \leq i \leq k} \theta_{i,N} \qquad \mbox{(iid multinomial model)}.
+\end{align*}
+Hence, a multinomial model sequence is unstable (or stable) depending on whether (or not) the maximal parameter difference $\max_{1 \leq i \leq k} \theta_{i,N} - \min_{1 \leq i \leq k} \theta_{i,N}$ diverges.
+Furthermore, using that each of the $k$ sufficient (count) statistics from the multinomial model \@ref(eq:mod11) satisfies $(U_{i,N}-L_{i,N})/N=1$, we  see that \@ref(eq:prop1) of Proposition \@ref(prp:prop1) becomes purely a parameter condition, $\sup_{N \geq 1}\max_{1 \leq i \leq k } |\theta_{i,N}| <\infty$, for ensuring that $\REP(\boldsymbol \theta_N)/N =\max_{1 \leq i \leq k} \theta_{i,N} - \min_{1 \leq i \leq k} \theta_{i,N}$ is bounded and stability follows for the multinomial distribution.
+Additionally, a stable multinomial sequence (i.e., bounded $\REP(\boldsymbol \theta_N)/N$) turns out to be nearly equivalent to \@ref(eq:prop1)  (e.g., these are the same if the smallest  parameter $\min_{1 \leq i \leq k } |\theta_{i,N}|$ remains bounded).
+
+When the condition \@ref(eq:prop1) of Proposition \@ref(prp:prop1) is violated, this aspect suggests a potentially unstable model that may be investigated more closely.
+For example, consider the exponential graph model from \@ref(eq:mod2) involving counts of edges, 2-stars and triangles with fixed parameters $\boldsymbol \theta_N = (\theta_{1},\theta_2,\theta_3)^\prime \in \mathbb{R}^3$ for $N\geq 1$.
+If either the 2-star parameter $\theta_2 \neq 0$ or triangle parameter $\theta_3 \neq 0$ is non-zero, then $\max_{1 \leq i \leq 3 } |\theta_{i}|(U_{i,N}-L_{i,N})/N \propto (n-2)\to \infty$ holds in \@ref(eq:prop1) by  $(U_{2,N}-L_{2,N})/N = 3 (U_{3,N}-L_{3,N})/N=(n-2)$ for 2-star and triangle statistics ($i=2,3$), so that Proposition \@ref(prp:prop1) hints that an unstable model may result when $|\theta_2| + |\theta_3| \neq 0$.
+Relatedly, a result from @schweinberger2011instability [Result 3] states that this model is unstable for all fixed parameters excluding cases $\theta_2 =\theta_3=0$ or $\theta_2 = - \theta_3/3$.
+However, more is true in line with the instability suggested by Proposition \@ref(prp:prop1) whenever $|\theta_2| + |\theta_3| \neq 0$ (i.e., excluding $\theta_2 =\theta_3=0$).
+
+To see this, consider an even number $n>2$ of nodes and let $\boldsymbol x_0$ denote the data outcome in $\mathcal{X}^N \equiv \{0,1\}^N$ with all $N = {n \choose 2}$ edges being zero, let $\boldsymbol x_1$ denote the outcome with all edges being 1, and let $\boldsymbol x_2$ denote the edge configuration from dividing the nodes into two equal groups, with no edges within a group and all edges between the groups (so that no triangles exist in $\boldsymbol x_2$).
+Then, the $N$-scaled  log-ratio \@ref(eq:elpr) for the exponential graph model \@ref(eq:mod2) can, by definition, be bounded below by
+\begin{align*}
+\frac{1}{N}\REP_N(\boldsymbol \theta_N) &\geq \max_{i=1,2}\frac{1}{N}
+\left| \log\left[ \frac{P_{\boldsymbol \theta_N}(\boldsymbol x_i)}{P_{\boldsymbol \theta_N}(\boldsymbol x_0)}\right] \right| \\
+&= (n-2) \max\left\{ \left| \theta_2 + \frac{\theta_3}{3}+\frac{\theta_1}{n-2} \right|, \frac{n}{4(n-1)} \left| \theta_2 + \frac{8\theta_1}{n-2} \right| \right\};
+\end{align*}
+a similar expression also holds for an odd node number $n>2$.
+Consequently, for all fixed parameters excluding $\theta_2=\theta_3=0$, $\lim_{N\to \infty}\REP_N(\boldsymbol \theta_N)/N=\infty$ then follows and the graph model with 2-stars and triangles is S-unstable, as suggested by the breach of Proposition \@ref(prp:prop1) for this model when $|\theta_2|+|\theta_3|\neq 0$.
+That is, instability holds even under $\theta_2 = - \theta_3/3$ case potentially allowed by Schweinberger's results.
+
+## Latent variable models of increasing parameter dimension
+
+We next consider instability of discrete data models based on exponential formulations involving hidden, or latent, variables, such as those described in Sections \@ref(restricted-boltzmann-machines)-\@ref(deep-learning) for networks.
+We will focus on restricted Bolzmann machine (RBM) models (Section \@ref(restricted-boltzmann-machines), having one layer of latent variables for simplicity, though the same instability concepts may be extended to other deep learning models (Section \@ref(deep-learning).
+For $N$ visible variables $\boldsymbol X \equiv \boldsymbol X_N = (X_1,\ldots,X_n)$ as data, each observation $X_i\in\{\pm 1\}$ being binary, the RBM-based model \@ref(eq:RBM2) for  $\boldsymbol X$ is again of FOES-type, though not an exponential model.
+However, the distribution of visible variables is induced by an underlying joint exponential model \@ref(eq:RBM1) for both visible and latent variables $(\boldsymbol X, \boldsymbol H)$, where $\boldsymbol H=(H_1,\ldots,H_{N_{\mathcal{H}}})$ denotes a vector of $N_{\mathcal{H}}$ hidden variables (similarly binary).
+The joint model is of linear exponential form involving $d(N)\equiv N + N_{\mathcal{H}} + N* N_{\mathcal{H}}$ sufficient statistics given by $(\boldsymbol X, \boldsymbol H, \boldsymbol X^T\boldsymbol H)$ and parameters $\boldsymbol \theta_N = (\boldsymbol \theta_N^{\mathcal{V}},\boldsymbol \theta_N^{\mathcal{H}}, \boldsymbol \theta_N^{\mathcal{VH}}  ) \in\mathbb{R}^{d(N)}$ corresponding to the $N$ visible variables $\boldsymbol X$ (i.e., $\boldsymbol \theta_N^{\mathcal{V}}\in\mathbb{R}^N$), the $N_{\mathcal{H}}$ hidden variables $\boldsymbol H$ (i.e., $\boldsymbol \theta_N^{\mathcal{H}}\in\mathbb{R}^{N_{\mathcal{H}}}$), and the $N *N_{\mathcal{H}}$ cross-product variables $\boldsymbol X^T\boldsymbol H$ (i.e., $\boldsymbol \theta_N^{\mathcal{VH}}\in\mathbb{R}^{N N_{\mathcal{H}}}$).
+However, unlike some previous exponential models considered in Sections \@ref(one-param-exp)-\@ref(fixed-dim-exp) (cf.
+Proposition \@ref(prp:prop1), note that the RBM formulation always associates parameters with *bounded* statistics (i.e., the components of $(\boldsymbol X, \boldsymbol H, \boldsymbol X^T\boldsymbol H)$) so that  model instability cannot arise here due to the magnitude of sufficient statistics exceeding the sample size $N$.
+Instead, RBM instability  may be linked to solely to parameter configuration and the fact that the number $d(N) \geq N$ of parameters necessarily increases with the number $N$ of observations $\boldsymbol X$, in contrast to previous exponential cases of fixed parameter dimension.
+
+To highlight the instability issues for the RBM model, consider a simple  model for $N$ visibles $\boldsymbol X$ with no hidden variables ($N_{\mathcal{H}}=0$), for which model statements \@ref(eq:RBM1)-\@ref(eq:RBM2) coincide.
+An independence model then results for variables in $\boldsymbol X$, which has $d(N)=N$ parameters $\boldsymbol \theta_N^{\mathcal{V}} = (\theta_{1,N}^{\mathcal{V}}, \ldots, \theta_{N,N}^{\mathcal{V}}) \in \mathbb{R}^{N}$, and the measure of model instability becomes
+$$
+\frac{1}{N}\REP(\boldsymbol \theta_N) = \frac{2}{N} \sum_{i=1}^N|\theta_{i,N}^{\mathcal{V}}| \quad \mbox{(RBM model, no hiddens)}.
+$$
+Hence, this model sequence for $\boldsymbol X$ will be  S-unstable model if the aggregation of absolute parameters grows faster than the number $N$ of parameters/visible variables.
+Consequently, even for a simplest RBM model involving independence, preventing instability requires careful choice of parameters, particularly with regard to how a parameter configuration differs from zero.
+For more general RBM models, the number $N_{\mathcal{H}}$ of hidden variables $\boldsymbol H$ can also be chosen arbitrarily (i.e., as some function $N_{\mathcal{H}}\equiv N_{N,\mathcal{H}}$ of $N$), which can substantially inflate the number $d(N)$ of model parameters and further impact model instability through accumulated parameters.
+To better understand the effects of instability in the RBM structure,  Proposition \@ref(prp:prop2) next frames the general behavior of extreme probabilities in the joint RBM model \@ref(eq:RBM1) for  $(\boldsymbol X, \boldsymbol H)$  and the implied RBM data model \@ref(eq:RBM2) for  $\boldsymbol X$ alone.
+Specifically, critical measures of instability may be closely connected in both models through tight bounds on their respective LREP values \@ref(eq:elpr).
+As a result, Proposition \@ref(prp:prop2) shows how an unstable distribution for observations $\boldsymbol X$ may be traced to sources of instability in the original joint distribution for $(\boldsymbol X,\boldsymbol H)$.
+This also suggests a device for avoiding instability, as provided next.
+
+To state the result, let $\REP_{\boldsymbol X}(\boldsymbol \theta_N) \equiv \REP(\boldsymbol \theta_N)$ denote the LREP value \@ref(eq:elpr) from the marginal distribution $P_{\boldsymbol \theta_N}$ of visibles $\boldsymbol X$ in \@ref(eq:RBM2) and write the LREP for the joint distribution $\tilde{P}_{\boldsymbol \theta_N}$ of $(\boldsymbol X, \boldsymbol H)$ from \@ref(eq:RBM1) as
+\begin{align*}
+\REP_{(\boldsymbol X, \boldsymbol H)}(\boldsymbol \theta_N) &=\log\left[  \frac{\max_{ (\boldsymbol x, \boldsymbol h) \in \{\pm 1\}^{N+N_{\mathcal{H}}}}
+\tilde{P}_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)}{\min_{ (\boldsymbol x, \boldsymbol h) \in \{\pm 1\}^{N+N_{\mathcal{H}}}`}
+\tilde{P}_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)} \right] \qquad \mbox{(joint RBM model)}    \\    
+&=   \left(\max_{ (\boldsymbol x, \boldsymbol h) \in \{\pm 1\}^{N+N_{\mathcal{H}}}} f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)-\min_{ (\boldsymbol x, \boldsymbol h) \in \{\pm 1\}^{N+N_{\mathcal{H}}}} f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)\right),
+\end{align*}
+written as a function 
+\begin{equation}
+(\#eq:f)
+f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)  \equiv  \sum_{i=1}^N x_i  \theta_{i,N}^{\mathcal{V}} +  \sum_{j=1}^{N_\mathcal{H}} h_j \theta_{j,N}^{\mathcal{H}} + \sum_{i=1}^N \sum_{j=1}^{N_\mathcal{H}} x_i h_j  \theta_{ij,N}^{\mathcal{VH}}
+\end{equation}
+of outcomes $\boldsymbol x =(x_1,\ldots,x_N) \in\{\pm1\}^{N}$, $\boldsymbol h =(h_1,\ldots,h_{N_{\mathcal{H}}})  \in\{\pm1\}^{N_{\mathcal{H}}}$ and parameters $\boldsymbol \theta_N \equiv (\boldsymbol \theta_N^{\mathcal{V}},\boldsymbol \theta_N^{\mathcal{H}}, \boldsymbol \theta_N^{\mathcal{VH}})$, with $\theta_{i,N}^{\mathcal{V}}$, $\theta_{j,N}^{\mathcal{H}}$ and $\theta_{ij,N}^{\mathcal{VH}}$ denoting respective parameter components, $1 \leq i \leq N$, $1 \leq j \leq N_{\mathcal{H}}$.
+Due to the marginalization steps in defining the distribution \@ref(eq:RBM2) of $\boldsymbol X$, note that $\REP_{\boldsymbol X}(\boldsymbol \theta_N)$ has no immediate analytical expression similar to that of  $\REP_{(\boldsymbol X, \boldsymbol H)}(\boldsymbol \theta_N)$.
+For clarity, recall also that S-instability \@ref(eq:Sun) in each model type refers to a respective divergence (i.e., $\lim_{N\to \infty} \REP_{(\boldsymbol X, \boldsymbol H)}(\boldsymbol \theta_N) /(N+N_{\mathcal{H}})=\infty$, $\lim_{N\to \infty} \REP(\boldsymbol \theta_N) /N=\infty$)
+ upon scaling by the corresponding number of variables in a distribution.
+In the following, let $|\boldsymbol y|_1 = \sum_{i=1}^d |y_i|$ denote the L-1 norm of a generic vector $\boldsymbol y =(y_1,\ldots,y_d)$, $d \geq 1$.
+```{proposition, label="prop2", echo=TRUE}
+Let $P_{\boldsymbol \theta_N}$ denote a RBM-based  model \@ref(eq:RBM2) for $N\geq 1$ visible variables $\boldsymbol X \equiv \boldsymbol X_N$ derived from $\tilde{P}_{\boldsymbol \theta_N}$ as the joint RBM distribution \@ref(eq:RBM1) of $(\boldsymbol X, \boldsymbol H)$ involving some number $N_{\mathcal{H}} \equiv  N_{N,\mathcal{H}}\geq 0$ of hidden variables $\boldsymbol H \equiv \boldsymbol H_N$ and parameters $\boldsymbol \theta_N \equiv (\boldsymbol \theta_N^{\mathcal{V}},\boldsymbol \theta_N^{\mathcal{H}}, \boldsymbol \theta_N^{\mathcal{VH}}) \in\mathbb{R}^{N}\times \mathbb{R}^{N_{\mathcal{H}}} \times  \mathbb{R}^{N N_{\mathcal{H}}}$.
+Then,
+\begin{enumerate}[(i)]
+\item the instability measure $\REP_{(\boldsymbol X)}(\boldsymbol \theta_N)$ for the marginal model $P_{\boldsymbol \theta_N}$ of  $\boldsymbol X$ satisfies
+    $$
+    \left| \REP_{(\boldsymbol X)}(\boldsymbol \theta_N)  - \elt\right| \leq    N_{\mathcal{H}}  \log 2
+    $$
+    for
+    $$
+    \elt  \equiv   \max_{ \boldsymbol x} \max_{ \boldsymbol h  } f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)-\min_{ \boldsymbol x } \max_{ \boldsymbol h }f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)
+    $$
+    based on $f_{\boldsymbol \theta_N}$ from \@ref(eq:f) with components $\boldsymbol x \in \{\pm 1\}^{N}, \boldsymbol h \in \{\pm 1\}^{N_{\mathcal{H}}}$.
+\item The instability measure $\REP_{(\boldsymbol X, \boldsymbol H)}(\boldsymbol \theta_N)\equiv \left(\max_{ \boldsymbol x} \max_{ \boldsymbol h  } f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)-\min_{ \boldsymbol x } \min_{ \boldsymbol h }f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)\right)$ for the joint model $\tilde{P}_{\boldsymbol \theta_N}$ of $(\boldsymbol X, \boldsymbol H)$ satisfies
+    \begin{align*}
+    2\Gam +  2|\boldsymbol \theta_N^{\mathcal{H}} |_{1} \; \geq \; \REP_{(\boldsymbol X, \boldsymbol H)}(\boldsymbol \theta_N)    & \geq 
+    2\max\big\{\Gam,   |\boldsymbol \theta_N^{\mathcal{H}} |_{1}\big\} \\
+    &\geq  2\Gam\\
+    &\geq  \elt\\
+    &\geq   \max\big\{  \Gamc, \, \Gam  - 2|\boldsymbol \theta_N^{\mathcal{H}} |_{1}  \big\}
+    \end{align*}
+    for
+    $$
+    \Gam \equiv \max_{ \boldsymbol h} k_{\boldsymbol \theta_N} (\boldsymbol h)
+    \geq  |\boldsymbol \theta_N^{\mathcal{V}} |_{1},\qquad  k_{\boldsymbol \theta_N} (\boldsymbol h)  \equiv \sum_{i=1}^{N }\left| \theta_{i,N}^{\mathcal{V}}   + \sum_{j=1}^{N_{\mathcal{H}}} h_j \theta_{ij,N}^{\mathcal{VH}} \right|,
+    $$ 
+    and $\Gamc \equiv \min_{ \boldsymbol h} k_{\boldsymbol \theta_N} (\boldsymbol h)$ based on a function  $k_{\boldsymbol \theta_N} (\boldsymbol h)$ of hidden variable outcomes $\boldsymbol h = (h_1,\ldots,h_{N_{\mathcal{H}}})$ and visible-related parameters $\boldsymbol \theta_N^{\mathcal{V}}$ and $\boldsymbol \theta_N^{\mathcal{VH}}$.
+\item Assuming $\sup_{N \geq 1} N_{\mathcal{H}}/N<\infty$ additionally, then the following properties 1.-7. hold:
+\begin{enumerate}[1.] \itemsep 0cm
+    \item an S-unstable visible model $P_{\boldsymbol \theta_N}$ is equivalent to the condition $\lim_{N\to \infty}  \elt/N  =\infty$; further, $P_{\boldsymbol \theta_N}$ is stable when $\elt/N$, $N \geq 1$, is bounded.
+    \item an S-unstable joint model $P_{\boldsymbol \theta_N}$ is equivalent to the condition $\lim_{N\to \infty} \max\{|\boldsymbol \theta_N^{\mathcal{H}}|_1, \Gam\}/N =\infty$; further,  $\tilde{P}_{\boldsymbol \theta_N}$ is stable when $[|\boldsymbol \theta_N^{\mathcal{H}}|_1+ \Gam]/N$, $N \geq 1$, is bounded.
+    \item if the visible model $P_{\boldsymbol \theta_N}$   is S-unstable, then  the joint model $\tilde{P}_{\boldsymbol \theta_N}$ is also S-unstable.
+    \item when $\lim_{N\to \infty}  (|\boldsymbol \theta_N^{\mathcal{V}}|_1-2|\boldsymbol \theta_N^{\mathcal{H}}|_1)/N =\infty$, both $P_{\boldsymbol \theta_N}$ and $\tilde{P}_{\boldsymbol \theta_N}$ are necessarily S-unstable.
+    \item when $\lim_{N\to \infty}|\boldsymbol \theta_N^{\mathcal{H}}|_1/N =\infty$,  the joint model $\tilde{P}_{\boldsymbol \theta_N}$ is necessarily S-unstable.
+    \item when $\sup_{N \geq 1} |\boldsymbol \theta_N^{\mathcal{H}}|_1 /N<\infty$, the visible model $P_{\boldsymbol \theta_N}$ being S-stable or S-unstable is equivalent to the joint model  $\tilde{P}_{\boldsymbol \theta_N}$ being stable or unstable.
+    \item  an S-stable visible model $P_{\boldsymbol \theta_N}$ results if
+        $$
+        |\boldsymbol \theta_N^{\mathcal{V}}|_1+ |\boldsymbol \theta_N^{\mathcal{VH}} |_1  \leq CN,\quad N \geq 1,
+        $$
+        for some $C>0$, while an S-stable joint model $\tilde{P}_{\boldsymbol \theta_N}$  results if
+        $$
+        |\boldsymbol \theta_N|_1 \equiv  |\boldsymbol \theta_N^{\mathcal{V}}|_1+|\boldsymbol \theta_N^{\mathcal{H}}|_1 +|\boldsymbol \theta_N^{\mathcal{VH}} |_1  \leq  C N ,\quad N \geq 1.
+        $$
+\end{enumerate}
+\end{enumerate}
+```
+
+```{remark, echo=TRUE}
+The condition $\sup_{N \geq 1} N_{\mathcal{H}}/N<\infty$ in Proposition \@ref(prp:prop2)(iii) is often mild in practice (i.e., the number $N_{\mathcal{H}}$ of hidden variables is typically not excessively larger than the number $N$ of visible observations).
+This allows instability results for both marginal and joint RBM models to be more readily stated together, as the numbers $N$ and $N+N_{\mathcal{H}}$ of variables in these models become asymptotically equivalent.
+```
+
+In Proposition \@ref(prp:prop2)(iii), the relationships between RBM models with regard to instability, and the effects of different parameter types, follow from the bounds on model instability measures in Proposition \@ref(prp:prop2)(i)-(ii).
+Generally speaking, all instability in the marginal RBM model for the data $\boldsymbol X$ can be attributed to an excessively large model quantity $\elt$, which predominantly follows when main $\boldsymbol \theta_N^{\mathcal{V}}$ and interaction $\boldsymbol \theta_N^{\mathcal{VH}}$ parameters related to visible variables are too large in magnitude (e.g., upon accumulation in terms such as $|\boldsymbol \theta_N^{\mathcal{V}}|_1$, $\Gam$ or $\Gamc$).
+For example, for any bounded sequence $|\boldsymbol \theta_N^{\mathcal{H}}|/N$ of hidden parameters, if main visible parameters $\boldsymbol \theta_N^{\mathcal{V}}$ are too extreme ($|\boldsymbol \theta_N^{\mathcal{V}}|_1/N\to \infty$), this aspect  will guarantee instability in the visible model ($\elt/N\to \infty$).
+In fact, the instability measure $\elt \equiv \max_{ \boldsymbol x} \max_{ \boldsymbol h  } f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)-\min_{ \boldsymbol x } \max_{ \boldsymbol h }f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)$ for marginal/visible model represents a clearly smaller portion of the  instability measure $\REP_{(\boldsymbol X, \boldsymbol H)}(\boldsymbol \theta_N)\equiv \max_{ \boldsymbol x} \max_{ \boldsymbol h  } f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)-\min_{ \boldsymbol x } \min_{ \boldsymbol h } f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)$ in the joint RBM model, implying that an unstable marginal model (i.e., due to  $\boldsymbol \theta_N^{\mathcal{V}}$, $\boldsymbol \theta_N^{\mathcal{VH}}$) must always translate to an unstable joint model and that further potential causes of instability exist for the joint model, often due to the size $|\boldsymbol \theta_N^{\mathcal{H}}|_1$.
+For example, while the joint RBM model for $(\boldsymbol X,\boldsymbol H)$ must always be unstable due to a diverging combination of visible and/or interaction parameters ($|\boldsymbol \theta_N^{\mathcal{V}}|_1/N\to \infty$ or $\Gam/N\to \infty$) (Proposition \@ref(prp:prop2)(iii.2)), instability for the joint model can also result when the main hidden parameters $\boldsymbol \theta_N^{\mathcal{H}}$ become too large relative to sample size ($|\boldsymbol \theta_N^{\mathcal{H}}|_1/N\to \infty$ in Proposition \@ref(prp:prop2)(iii.5)).
+However, under Proposition \@ref(prp:prop2), the main hidden parameters $\boldsymbol \theta_N^{\mathcal{H}}$  do not necessarily entail a source of instability for the marginal visible model.
+To explain this distinction, consider a joint model where all parameters related to visibles are zero, $\boldsymbol \theta_N^{\mathcal{V}}= \boldsymbol \theta_N^{\mathcal{VH}}=\boldsymbol 0$, but the hidden-related parameters diverge in sum $|\boldsymbol \theta_N^{\mathcal{H}}|_1/N\to \infty$.
+Here, the explosive behavior among parameters $\boldsymbol \theta_N^{\mathcal{H}}$ induces instability in the joint model for $(\boldsymbol X, \boldsymbol H)$ but the marginal model for $\boldsymbol X$, however, has a perfectly stable (and in fact uniform) distribution in this case.
+When the hidden parameters are bounded relative to the sample size ($\sup_{N\geq 1} |\boldsymbol \theta_N^{\mathcal{H}}|_1/N<\infty$), then all instability in both the joint and marginal RBM models can be directly linked to excessively large visible $\boldsymbol \theta_N^{\mathcal{V}}$ and/or interaction parameters $\boldsymbol \theta_N^{\mathcal{VH}}$ so that features of stability/instability must be the same across both models (Proposition \@ref(prp:prop2)(iii.6)).
+Hence, to prevent instability in the joint model, the combined magnitudes of all parameters $\theta_N$ must be controlled (cf.
+Proposition \@ref(prp:prop2)(iii.7)), while a stable visible data model technically results in constraining only the sizes of visible-related parameters $\boldsymbol \theta_N^{\mathcal{V}}$, $\boldsymbol \theta_N^{\mathcal{VH}}$.
+Nevertheless, because the joint model is often employed in practice for purposes of simulation and simulation-based inference, it is still reasonable to consider parameter choices for ensuring a stable joint model (and, consequently, a stable visible model as well).
+
+In more complicated network models involving further or deeper hidden layers, the same issues and causes of instability similarly exist, but are compounded by a greater number of model parameters.
+S-unstable joint models will similarly follow if the combined sizes of all parameters are too great relative to the total number of variables, while instability in the data model for visible variables will depend only on the  main or interaction parameters directly related to visibles and how their accumulated magnitude compares to the observation sample size $N$.
+
+# Statistical consequences of instability {#implications}
+Due to their extreme sensitivity in probability structure, S-instability in FOES models often translates to numerical complications, and in fact obstructions, in both simulation and statistical inference based on likelihoods.
+
+## Implications for simulation
+
+Suppose one aims to apply MCMC to simulate data $\boldsymbol X=(X_1,\ldots,X_n)$ from a FOES model $P_{\boldsymbol \theta_N}(\boldsymbol x)$, $\boldsymbol x \in\mathcal{X}^N$ whereby a chain is constructed  with $P_{\boldsymbol \theta_N}$ as the stationary distribution.
+For an unstable FOES model, one component changes in outcomes may produce radically different probabilities which then creates numerical barriers to MCMC.
+For example, consider implementation of a Gibbs sampler from the full conditional distributions $P_{\boldsymbol \theta_N} (X_i = x| \boldsymbol x_{-i})$, $x\in\mathcal{X}$, of each variable $X_i$ based on values $\boldsymbol x_{-i}\in  \mathcal{X}^{N-1}$ for the remaining variables, say $\boldsymbol X_{-i}$, in $\boldsymbol X$.
+If a single change in $X_i$ from one value $x_1$ to another $x_2$ may produce two outcomes $\boldsymbol x^{(1)}$ and $\boldsymbol x^{(2)}$ for $\boldsymbol X$ with vastly different probabilities under the joint distribution $P_{\boldsymbol \theta_N}$, then the Gibbs sampler can have  extreme log-ratios in its transition probabilities
+$$
+\left| \frac{P_{\boldsymbol \theta_N} (X_i =x_1 | \boldsymbol x^{(1)}_{-i})}{P_{\boldsymbol \theta_N} (X_i = x_2| \boldsymbol x^{(2)}_{-i}) } \right|=\left| \frac{P_{\boldsymbol \theta_N} ( \boldsymbol x^{(1)} }{P_{\boldsymbol \theta_N} (  \boldsymbol x^{(2)} ) } \right|
+$$
+using that conditional probabilities are proportional to the joint probabilities and the fact that unstable models can have  unbounded log-probability ratios in one-component changes (Theorem \@ref(thm:instab-elpr).
+This can hinder the ability of chain to effectively explore the sample space of the observations $\boldsymbol X$, as the chain may then mix poorly by moving rapidly to, and slowly away from, sections of the sample space surface.
+Because the numerical problem is unbounded ratios of probabilities under the joint model, similar complications can also arise for Metropolis-Hastings algorithms for MCMC.
+Hence, while an unstable FOES model $P_{\boldsymbol \theta_N}$ may be valid and  technically open to simulation by MCMC, the aspect of instability can ruin  numerical applications of MCMC.
+
+## Implications for maximum likelihood inference
+
+The rugged or volatile probability structure in unstable models can also hamper efforts to maximize likelihood functions for statistical inference.
+Essentially, when a FOES model is unstable along a parameter sequence $\boldsymbol \theta_N$, the same model can consequently unstable along parameters $-\boldsymbol \theta_N$ opposite from the origin (model condition PSR and Corollary \@ref(cor:sign)).
+This leads to potential hyper-sensitivity in the likelihood function, which can be pulled in extreme and opposite directions depending on how a parameter lies with respect to zero and translates into numerical complications in maximizing the objective function.
+
+For a FOES model $P_{\boldsymbol \theta_N}(\boldsymbol x)$, $\boldsymbol x\in\mathcal{X}^N$, for $N$ observations $\boldsymbol X =(X_1,\ldots,X_n)$, define the likelihood function $\mathcal{L}(\boldsymbol \theta_N | \boldsymbol x) \equiv N^{-1}\log P_{\boldsymbol \theta_N }(\boldsymbol x)$ at $\boldsymbol \theta_N$  and the minimum probability ratio counterpart  $\mathcal{L}(\boldsymbol \theta_N | \boldsymbol x)  - N^{-1} \log \min_{\boldsymbol \theta_N}= N^{-1}\log[P_{\boldsymbol \theta_N }(\boldsymbol x) /\min_{\boldsymbol \theta_N}]$ where $\min_{\boldsymbol \theta_N} \equiv \min_{\boldsymbol y\in\mathcal{X}^N} P_{\boldsymbol \theta_N }(\boldsymbol y)]$.
+Also define $\max_{\boldsymbol \theta_N} \equiv \max_{\boldsymbol y\in\mathcal{X}^N} P_{\boldsymbol \theta_N }(\boldsymbol y)]$.
+
+```{theorem, label="something", echo=TRUE}
+Let $P_{\boldsymbol \theta_N}$, $N \geq 1$, denote an S-unstable FOES model sequence (i.e., $\lim_{N\to \infty}\REP(\boldsymbol \theta_N)/N=\infty$) which additionally satisfies model condition SPR.
+Then,
+\begin{align*}
+\log\left[ \frac{P_{\boldsymbol \theta_N }(\boldsymbol x) }{\min_{\boldsymbol \theta_N}}\right] +  \log\left[ \frac{P_{-\boldsymbol \theta_N }(\boldsymbol x) }{\min_{-\boldsymbol \theta_N}}\right]  &=    \log\left[ \frac{\max_{\boldsymbol \theta_N}}{\min_{\boldsymbol \theta_N}}\right]  \equiv  \REP(\boldsymbol \theta_N)\\
+& = \log\left[ \frac{\max_{-\boldsymbol \theta_N} }{\min_{-\boldsymbol \theta_N}}\right] \equiv \REP(-\boldsymbol \theta_N)
+\end{align*}
+for any $\boldsymbol x\in\mathcal{X}^N$ and, as $N\to \infty$, the following convergence
+$$
+\frac{1}{\REP(\boldsymbol \theta_N)} \log\left[ \frac{P_{\boldsymbol \theta_N }(\boldsymbol X) }{\min_{\boldsymbol \theta_N}}\right]  \rightarrow 1 \qquad \frac{1}{\REP(-\boldsymbol \theta_N)}\log\left[ \frac{P_{-\boldsymbol \theta_N }(\boldsymbol X) }{\min_{-\boldsymbol \theta_N}}\right]  \rightarrow 0
+$$
+holds in probability $P_{\boldsymbol \theta_N}$, as well as in expectation $E_{\boldsymbol \theta_N}$, under $\boldsymbol \theta_N$; the limits of convergence are reversed under $-\boldsymbol \theta_N$,
+$$
+\frac{1}{\REP(\boldsymbol \theta_N)} \log\left[ \frac{P_{\boldsymbol \theta_N }(\boldsymbol x) }{\min_{\boldsymbol \theta_N}}\right]  \rightarrow 0 \qquad \frac{1}{\REP(-\boldsymbol \theta_N)}\log\left[ \frac{P_{-\boldsymbol \theta_N }(\boldsymbol x) }{\min_{-\boldsymbol \theta_N}}\right]  \rightarrow 1
+$$
+```
+This needs to be connected to results in Schweinberger who considered the one-parameter linear exponential case showing that, for positive $\boldsymbol \theta_N>0$, 
+$$
+\frac{1}{\REP(\boldsymbol \theta_N)} \log\left[ \frac{P_{\boldsymbol \theta_N }(\boldsymbol x) }{\min_{\boldsymbol \theta_N}}\right] = \frac{\boldsymbol g(\boldsymbol x) - L_N}{U_N-L_N} 
+$$
+holds and converges to $1$ in expectation $E_{\boldsymbol \theta_N}$, while
+$$   
+\frac{1}{\REP(-\boldsymbol \theta_N)} \log\left[ \frac{P_{-\boldsymbol \theta_N }(\boldsymbol x) }{\min_{-\boldsymbol \theta_N}}\right] = \frac{U_N -\boldsymbol g(\boldsymbol x)}{U_N-L_N} 
+$$
+holds converges to $1$ in expectation $E_{-\boldsymbol \theta_N}$.
+This result is a special case of Theorem \@ref(thm:something) and the LREP expansion \@ref(eq:UL) for the one parameter exponential model case.
+
+The idea, more generally, is that probability in an unstable model is pushed to opposite regions of the sample space due to degeneracy and opposite signed parameters.
+This can cause numerical complications in likelihood maximization.
+The likelihood function knows what to do with an observed outcome $\boldsymbol x \equiv \argmax \mathcal{L}(\boldsymbol \theta_N)$ which is also the outcome with minimal probability under $P_{-\boldsymbol \theta_N}$.
+But unless the observed outcome $\boldsymbol x$ is at an extreme in the probability structures of   $P_{\boldsymbol \theta_N}, P_{-\boldsymbol \theta_N}$ (which are related) the likelihood function will be flat and optimization steps may shift around zero.
+
+## Implications for Bayes inference
+
+Considering that changes in sign of parameters can cause all probability to be placed on disjoint sections of the sample space (Corollary \@ref(cor:sign)), then this will create numerical difficulties in Bayes inference based on MCMC.
+We can write the MH transition density to indicate this.
+
+# Concluding Remarks {#conclusions}
+
+For a large class of models that covers a broad range of applications (including "deep learning''), we have developed a formal definition of instability in model probability structure and elucidated multiple consequences of instability.
+We have shown for FOES models that instability manifests through small changes in data leading to potentially large changes in probability as well as the potential to place all probability on certain modal subsections of the sample space, which potentially could be small.
+The FOES model class is quite broad and, particularly in developing FOES models for large data sets, some caution should be used in parameter specification to control effects of
+model instability.

### 2017-09-30 writing/paper-note/docs/note.tex
what a mess

@@
 
-\usepackage{tikz, subfig, amsthm, multirow, float}
+\usepackage{tikz, subfig, amsthm, multirow, float, enumerate}
 \usepackage{tikz-3dplot}
@@
 \newtheorem{lemma}{Lemma}
+\newtheorem{corollary}{Corollary}
 
@@
 
+\newcommand{\REP}{\mathrm{LREP}}
+\newcommand{\DN}{\Delta_N}
+
+\newcommand{\ma}{\mathrm{max}_{\boldsymbol \theta_N}}
+\newcommand{\mi}{\mathrm{min}_{\boldsymbol \theta_N}}
+
+\newcommand{\maa}{\mathrm{max}_{-\boldsymbol \theta_N}}
+\newcommand{\mii}{\mathrm{min}_{-\boldsymbol \theta_N}}
+
 \newcommand{\nv}{{n_{\scriptscriptstyle V}}}
@@
 \newcommand{\E}{E}
+
+\newcommand{\elt}{A_{N}(\boldsymbol \theta_N) }
+\newcommand{\Gam}{B_{N}(\boldsymbol \theta_N) }
+
+\newcommand{\Gamc}{C_{N}(\boldsymbol \theta_N) }
+\newcommand{\Gamt}{\Gamma_{N,2}(\boldsymbol \theta_N) }
 
@@
 
-We consider the behavior, and the potential impropriety, of probability
-models built to incorporate a sequence of discrete observations with
-length \(N\). Let \((X_1, \dots, X_N)\) denote a set of discrete random
-variables with a finite sample space, \(\mathcal{X}^N\). That is,
-\(\mathcal{X}\) with \(|\mathcal{X}| < \infty\) represents a finite set
-of potential outcomes for each single variable \(X_i\), and the data
-sequence \((X_1,\ldots,X_N)\) takes values in the \(N\)-fold product
-space \(\mathcal{X}^N\). For each \(N\), let
+We consider the behavior, and the potential impropriety, sequences of
+discrete probability models built to incorporate observations of
+increasing sample size \(N\). Interest is in identifying instability in
+such models, which is roughly characterized by probabilities with
+extreme sensitivity to small changes in data configuration. The concept
+of instability was introduced in the field of statistical physics (i.e.,
+point processes) by Ruelle
+(\protect\hyperlink{ref-ruelle1999statistical}{1999}) and then further
+extended by Schweinberger Schweinberger
+(\protect\hyperlink{ref-schweinberger2011instability}{2011}) for a
+family of exponential models. At issue, models exhibiting instability
+are typically undesirable as these tend to provide poor representations
+of data or data-generation. As an example, such models can include
+near-degenerate distributions that assign all essential probability mass
+to only a subset of an overall sample space. The latter issue in
+connection to degeneracy has been recognized as concern in that dominant
+model outcomes may not resemble observed data (cf. Handcock
+\protect\hyperlink{ref-handcock2003assessing}{2003}). As a compounding
+issue, model instability often has direct negative impacts for
+statistical inference and computations based on likelihood functions.
+Namely, volatilities in probability structure can potentially hamper the
+numerical evaluations required for maximum likelihood estimation as well
+as other model-based simulations via Markov Chain Monte Carlo (MCMC).
+These reasons motivate our general study of instability for a broad
+class of probability models, described next.
+
+In the model framework, let \(\boldsymbol X_N = (X_1, \dots, X_N)\)
+denote a collection of discrete random variables with a finite sample
+space, \(\mathcal{X}^N\), represented as some \(N\)-fold Cartesian
+product. That is, \(\mathcal{X}\) with \(|\mathcal{X}| < \infty\)
+denotes the set of potential outcomes for each single variable \(X_i\),
+so that the product space \(\mathcal{X}^N\) corresponds to values for
+the variables \(\boldsymbol X_N=(X_1,\ldots,X_N)\). For each \(N\), let
 \(P_{\boldsymbol \theta_N}\) denote a probability model on
@@
 assume that the model support of \(P_{\boldsymbol \theta_N}\) is the
-sample space \(\mathcal{X}^N\). This framework produces a series
-\(P_{\boldsymbol \theta_N}\) of probability models, indexed by a generic
-sequence of parameters \(\boldsymbol \theta_N\), to describe data of
-each length \(N \geq 1\). The size and structure of such parameters are
-without restriction, and natural cases include those where
+sample space \(\mathcal{X}^N\). This framework produces probability
+models \(P_{\boldsymbol \theta_N}\), indexed by a generic sequence of
+parameters \(\boldsymbol \theta_N\), to describe data
+\(\boldsymbol X_N\) of any given sample size \(N \geq 1\). For
+simplicity, we will refer to this distributional class as \emph{Finite
+Outcome Everywhere Supported (FOES)} models in the following. The
+dimension and structure of such parameters are generic, without
+restriction, though natural cases will be seen to include those where
 \(\boldsymbol \theta_N \in \mathbb{R}^{q(N)}\) for some arbitrary
-integer-valued function \(q(\cdot) \geq 1\). We will refer to this model
-class as \emph{Finitely Supported Finite Sequence (FSFS) models}.
+integer-valued function \(q(\cdot) \geq 1\).
 
-Section \ref{examples} provides several examples of FSFS models commonly
-used in graph/network analysis and machine learning (i.e., deep learning
-models). Section \ref{instability-results} establishes formal results
-regarding the propriety of FSFS models with regard to stability. A FSFS
-probability model sequence exhibits instability if small changes in the
-components of a data outcome \((x_1,\ldots,x_N)\) can result in large
-changes in probability \(P_{\theta_N}(x_1,\ldots,x_N)\). The concept of
-instability, introduced in the field of statistical physics by Ruelle
-(\protect\hyperlink{ref-ruelle1999statistical}{1999}), was extended to
-include a notion of detection and quantification for certain exponential
-family models by Schweinberger
-(\protect\hyperlink{ref-schweinberger2011instability}{2011}). For
-similar exponential models, particularly in connection to random
-graphs/networks, Handcock
-(\protect\hyperlink{ref-handcock2003assessing}{2003}) considered
-(mean-based) characterizations for so-called model degeneracy, whereby a
-probability model places all mass on a small subset of the sample space
-and produces undesirably low variability in model outcomes. As described
-by Schweinberger
-(\protect\hyperlink{ref-schweinberger2011instability}{2011}), model
-instability and model degeneracy are related by viewing degeneracy as an
-extreme or limiting form of instability. The instability results of
-Schweinberger
-(\protect\hyperlink{ref-schweinberger2011instability}{2011}) were
-developed for the case of discrete exponential family models. The main
-results here concern a general measure of model instability, appropriate
-across the whole FSFS model class. This can be used to identify when
-certain maximal probabilities in FSFS models are too extreme relative to
-the length \(N\) and may thereby induce a potentially undesirable
-probability structure. In this case, a one component change in the data
-sequence may shift probability by orders of magnitude, and FSFS models
-are rigorously shown to become degenerate as the measure of instability
-increases. Lastly, Section \ref{implications} emphasizes the
-implications of our model propriety results and proofs of the main
-results appear in Appendix \ref{appendix-instab}.
+Section \ref{examples} provides some examples of FOES models encountered
+in graph/network analysis and machine learning (i.e., deep learning
+models). These are used as references for later illustrations. Section
+\ref{instability-results} then establishes several formal results for
+FOES models with regard to instability. Schweinberger
+(\protect\hyperlink{ref-schweinberger2011instability}{2011}) originally
+developed instability results framed for, or specific to, a certain
+class of discrete exponential models. For similar exponential models
+with random networks, Handcock
+(\protect\hyperlink{ref-handcock2003assessing}{2003}) studied model
+degeneracy, where a probability model places near complete mass on modes
+and may thereby narrow the feasible model outcomes. As findings here and
+from Schweinberger
+(\protect\hyperlink{ref-schweinberger2011instability}{2011}) suggest,
+model instability and degeneracy may also be related by viewing
+degeneracy as an extreme, or limiting form, of instability. Our main
+results establish a broad characterization of model instability,
+appropriate across the whole FOES model class, that incorporates results
+of Schweinberger
+(\protect\hyperlink{ref-schweinberger2011instability}{2011}) as a
+special case. We prescribe a general and simple condition for
+identifying instability in a FOES model sequence, which quantifies
+whether certain maximal probabilities in a FOES model are too extreme
+relative to the sample size \(N\). When fulfilled, the probability
+structure of a FOES model is shown to exhibit extreme sensitivity, with
+probability assignments possessing extreme peaks and troughs across
+nearly identical outcomes. As the measure of model instability
+increases, probabilities from an unstable FOES model additionally
+increase in volatility and provably slide into degeneracy. Section
+\ref{implications} then emphasizes the implications of such model
+instability, showing that such impropriety can be expected to
+numerically hinder maximum likelihood estimation and MCMC-based
+simulations. As one potential remedy, suggestions are given or
+constraining model parameterizations and avoid the most problematic
+regions of the parameter space. Proofs of the main results appear in
+Appendix \ref{appendix-instab}.
 
@@
 
-Many model families fall under the umbrella of FSFS models. For
-illustration, this section presents three specific examples of FSFS
+Many model families fall under the umbrella of FOES models. For
+illustration, this section presents three specific examples of FOES
 models, including models with deep architectures.
@@
 
-For discrete random variables \(\boldsymbol X = (X_1, \dots, X_N)\) with
-sample space \(\mathcal{X}^N\), \(|\mathcal{X}| < \infty\), consider an
+For random variables
+\(\boldsymbol X \equiv\boldsymbol X_N= (X_1, \dots, X_N)\) with sample
+space \(\mathcal{X}^N\), \(|\mathcal{X}| < \infty\), consider an
 exponential family model for \(\boldsymbol X\) with probability mass
-function of the form \[
-p_{N, \boldsymbol \lambda}(\boldsymbol x) = \exp\left[\boldsymbol\eta^T(\boldsymbol \lambda) \boldsymbol g_N(\boldsymbol x) - \psi(\boldsymbol \lambda)\right], \qquad \boldsymbol x \in \mathcal{X}^N,
-\] depending on parameter vector
-\(\boldsymbol \lambda \in \Lambda \subset \mathbb{R}^{k}\) and natural
+function given by
+\begin{equation}
+\label{eq:expo}
+p_{N, \boldsymbol \theta}(\boldsymbol x) = \exp\left[\boldsymbol\eta^T(\boldsymbol \theta) \boldsymbol g_N(\boldsymbol x) - \psi(\boldsymbol \theta)\right], \qquad \boldsymbol x \in \mathcal{X}^N,
+\end{equation}
+depending on parameter vector
+\(\boldsymbol \theta \in \Theta_N \subset \mathbb{R}^{k}\) and natural
 parameter function
@@
 sufficient statistics, while \[
-\psi(\boldsymbol \lambda) = \log \sum\limits_{\boldsymbol x \in \mathcal{X}^N}\exp\left[\boldsymbol \eta^T(\boldsymbol \lambda) \boldsymbol g_N(\boldsymbol x) \right], \qquad \boldsymbol \lambda \in \Lambda,
-\] denotes the normalizing function, and
-\(\Lambda = \{\boldsymbol \lambda \in \mathbb{R}^k : \psi(\boldsymbol \lambda) < \infty, k \le q(N) \}\)
-is the parameter space.
+\psi(\boldsymbol \theta) = \log \sum\limits_{\boldsymbol x \in \mathcal{X}^N}\exp\left[\boldsymbol \eta^T(\boldsymbol \theta) \boldsymbol g_N(\boldsymbol x) \right], \qquad \boldsymbol \theta \in \Theta_N\equiv \{\boldsymbol \theta \in \mathbb{R}^k : \psi(\boldsymbol \theta) < \infty \},
+\] denotes the normalizing function with parameter space \(\Theta_N\).
+The natural parameter function \(\eta (\boldsymbol \theta)\) has a
+linear form (i.e.,
+\(\eta (\boldsymbol \theta)= \bm{A} \boldsymbol \theta\) for a given
+\(L \times k\) matrix \(\bm{A}\)) in many common model formulaitons,
+though may also be nonlinear (e.g., curved exponential families). In the
+linear case, \(\eta (\boldsymbol \theta) = \boldsymbol \theta\) may be
+generally assumed in the exponential parameterization with a minor
+modification to the definition of sufficient statistics
+\(\boldsymbol g_N(\boldsymbol x)\).
 
-Defining
-\(P_{\boldsymbol \theta_N}(\boldsymbol x)\equiv p_{N,\boldsymbol \lambda_N}(\boldsymbol x)\)
-with \(\boldsymbol \theta_N=\boldsymbol \lambda_N\) to be a sequence of
-elements of \(\Lambda \subset \mathbb{R}^k\) and noting that
-\(P_{\boldsymbol \theta_N}(\boldsymbol x) > 0\) for all
-\(\boldsymbol x \in \mathcal{X}^N\), these discrete exponential family
-models are special cases of the FSFS models. Such exponential models
-arise with spatial data on a lattice (Besag
-\protect\hyperlink{ref-besag1974spatial}{1974}), network data (Wasserman
-and Faust \protect\hyperlink{ref-wasserman1994social}{1994}; Handcock
-\protect\hyperlink{ref-handcock2003assessing}{2003}), and even standard
-independence models for discrete data, such as with \(N\) iid Bernoulli
-random variables. (Note that for random graphs or networks with, say,
-\(m\) nodes, one may wish to consider \(N={m \choose 2}\) edges as
-binary (presence/absence) variables \(X_i\). In this case, the length
-\(N\) of data sequence may naturally increase as a function of \(m\).)
-For these exponential models, the dimension of the parameter
-\(\boldsymbol \theta_N\) remains constant over each \(N\), as
-\(\boldsymbol \theta_N\) lies in a parameter space of fixed Euclidean
-dimension \(k\). This need not be true for other types of FSFS models
-considered in Sections
-\ref{restricted-boltzmann-machines}-\ref{deep-learning}. Schweinberger
+Such discrete exponential family models are special cases of the FOES
+models, as seen by defining
+\(P_{\boldsymbol \theta_N}(\boldsymbol x)\equiv p_{N,\boldsymbol \theta_N}(\boldsymbol x)> 0\),
+\(\boldsymbol x \in \mathcal{X}^N\), based on \eqref{eq:expo} and a
+parameter sequence
+\(\boldsymbol \theta_N \in \Theta_N \subset \mathbb{R}^k\). For example,
+if observations \(\boldsymbol X = (X_1,\ldots,X_N)\) correspond to \(N\)
+independent and identically distributed Bernoulli random variables, each
+indicating a binary \(0\)-\(1\) outcome, the resulting probabilities
+have exponential form \eqref{eq:expo} given by
+\begin{equation}
+\label{eq:mod1}
+P_{\boldsymbol \theta_N}(\boldsymbol x) \propto
+ \exp\left[\boldsymbol \theta_N \sum_{i=1}^N x_i\right], \qquad \boldsymbol x=(x_1,\ldots,x_N) \in\{0,1\}^N, 
+ \end{equation}
+with sufficient statistic
+\(\boldsymbol g_N(\boldsymbol x)\equiv \sum_{i=1}^N x_i\) and ``log odds
+ratio'' parameter
+\(\boldsymbol \theta_N \equiv \log[ P_{\boldsymbol \theta_N}(X_i=1)/P_{\boldsymbol \theta_N}(X_i=0) ] \in \mathbb{R}\).
+More generally, supposing \(\boldsymbol X =(X_1,\ldots,X_N)\) represent
+\(N\) independent trials, each assuming an outcome \(\{1,\ldots,k\}\)
+among \(k\) possibilities (e.g., a die roll), a multinomial distribution
+is given by
+\begin{equation}
+\label{eq:mod11}
+P_{\boldsymbol \theta_N}(\boldsymbol x) \propto  \exp\left[  \boldsymbol \theta_{N}^T g_N(\boldsymbol x)   \right] =
+\exp\left[ \sum_{j=1}^k {\theta_{j,N}} \sum_{i=1}^N \mathbb{I}(x_i=j) \right], \qquad \boldsymbol x  \in\{1,\ldots,k\}^N, 
+\end{equation}
+with sufficient statistic \(\boldsymbol g_N(\boldsymbol x)\) involving a
+count \(\sum_{i=1}^N \mathbb{I}(x_i=j)\) for each outcome
+\(j \in \{1,\ldots,k\}\), where \(\mathbb{I}(\cdot)\) denotes the
+indicator function, and parameters
+\(\boldsymbol \theta_N=(\theta_{1,N},\ldots,\theta_{k,N})\in\mathbb{R}^k\)
+defining log-probability ratios
+\(\theta_{i,N}-\theta_{j,N} =\log [P_{\boldsymbol \theta_N}(X_1=i)/P_{\boldsymbol \theta_N}(X_1=j)]\).
+In addition to such standard models for discrete independent data,
+exponential models of FOES type commonly arise with dependent spatial
+data (Besag \protect\hyperlink{ref-besag1974spatial}{1974}) and
+network/relational data (Wasserman and Faust
+\protect\hyperlink{ref-wasserman1994social}{1994}; Handcock
+\protect\hyperlink{ref-handcock2003assessing}{2003}). For a random graph
+or network with, say, \(n\) nodes, consider \(N={n \choose 2}\) random
+edges where the \(i\)th edge is associated with a pair of nodes
+\(s_i \equiv \{v_i,u_i\}\) and a binary variable \(X_i\in\{0,1\}\)
+indicating presence/absence of an edge among the node pair \(s_i\),
+\(i=1,\ldots,N\). Here the length \(N\) of the edge variable sequence
+\(\boldsymbol X = (X_1,\ldots,X_N)\) increases as a function of node
+number \(n\) and corresponding exponential models often incorporate
+graph topographical features derived from \(\boldsymbol X\). As an
+example, consider a graph model of exponential/FOES form prescribed by
+\begin{equation}
+\label{eq:mod2}
+P_{\boldsymbol \theta_N}(\boldsymbol x) \propto
+ \exp\left[\sum_{j=1}^3 \theta_{j,N} g_{j,N}(\boldsymbol x)\right], \quad\qquad \boldsymbol x=(x_1,\ldots,x_N)  \in\{0,1\}^N,
+\end{equation}
+\[
+  g_{1,N}(\boldsymbol x) \equiv \sum_{i=1}^N  x_i, \qquad\quad g_{2,N}(\boldsymbol x) \equiv \sum_{1 \leq i<j \leq N,\atop s_i \cap s_j \neq \emptyset}\!\!\!   x_i x_j, \qquad
+  g_{3,N}(\boldsymbol x) \equiv \sum_{1 \leq i<j<\ell \leq N, \atop s_i \cap s_j \neq \emptyset,s_i \cap s_\ell \neq \emptyset, \\ s_j \cap s_\ell \neq \emptyset } \!\!\!\!\!\!\!\!
+\!\!\!\!\!\!\!\!
+x_i x_j x_\ell,
+\] involving the numbers of edges, 2-stars and triangles among an
+outcome \(\boldsymbol x\) given by \(g_{1,N}(\boldsymbol x)\),
+\(g_{2,N}(\boldsymbol x)\) and \(g_{3,N}(\boldsymbol x)\), respectively,
+along with \(k=3\) real parameters
+\(\boldsymbol \theta_N \equiv (\theta_{1,N},\theta_{2,N},\theta_{3,N})\).
+For this network model \eqref{eq:mod2} in particular, as well as for more
+general models of form \eqref{eq:expo}, Schweinberger
 (\protect\hyperlink{ref-schweinberger2011instability}{2011}) considered
-instability in such exponential models (e.g., for random graphs) for
-sequences of fixed parameters
-\(\boldsymbol \theta_N=\boldsymbol \lambda\in\mathbb{R}^k\),
-\(N \geq 1\).
+instability in such exponential models with sequences of fixed
+parameters
+\(\boldsymbol \theta_N = (\theta_1,\ldots,\theta_k)\in\mathbb{R}^k\),
+\(N \geq 1\), of fixed dimension \(k\).
+
+For model sequences
+\(P_{\boldsymbol \theta_N}(\boldsymbol x)\equiv p_{N,\boldsymbol \theta_N}(\boldsymbol x)\)
+of the exponential type \eqref{eq:expo}, such as those in
+\eqref{eq:mod1}-\eqref{eq:mod2}, note that the dimension \(k\) of the
+parameter \(\boldsymbol \theta_N\in\Theta \subset \mathbb{R}^k\)
+necessarily remains the same for all sample sizes \(N \geq 1\) as the
+form of the natural parameter function \(\eta(\cdot)\) in \eqref{eq:expo}
+and the number of sufficient statistics
+\(\boldsymbol g_{N}(\boldsymbol x)\) do not depend on \(N\).
+Consequently, \(\boldsymbol \theta_N\) lies in a parameter space of
+fixed Euclidean dimension \(k\). However, this aspect need not be true
+for other types of FOES models considered in Sections
+\ref{restricted-boltzmann-machines} - \ref{deep-learning}, where instead
+the numbers of parameters and sufficient statistics commonly increase
+with the sample size \(N\).
 
@@
 (\(\mathcal{V}\)), with conditional independence within each layer. Let
-\(\boldsymbol X=(X_1,\ldots,X_N)\) denote the \(N\) random variables for
-visibles with support \(\mathcal{X}^N\) and
-\(\boldsymbol H=(H_1,\ldots,H_{N_H})\) denote the \(N_H\) random
-variables for hiddens with support \(\mathcal{X}^{N_H}\) where
-\(\mathcal{X} = \{-1,1\}\). For parameters
-\(\boldsymbol \alpha \in \mathbb{R}^{N_H}\),
-\(\boldsymbol \beta \in \mathbb{R}^N\), and \(\Gamma\) as a matrix with
-dimension \(N_H \times N\), the RBM model for
-\(\tilde{\boldsymbol X}=(\boldsymbol X,\boldsymbol H)\) then has the
-joint probability mass function \[
-P_{\boldsymbol \theta_N} (\tilde{\boldsymbol x}) = \exp\left[ \boldsymbol \alpha^T \boldsymbol h + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^T \Gamma \boldsymbol x - \psi(\boldsymbol \theta_N)\right], \qquad \tilde{\boldsymbol x} = (\boldsymbol h, \boldsymbol x) \in \mathcal{X}^{N+N_H},
-\] where \[
-\psi(\boldsymbol \theta_N) = \log \sum_{\tilde{\boldsymbol x} \in \mathcal{X}^{N + N_H}} \exp\left[ \boldsymbol \alpha^T \boldsymbol h + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^T \Gamma \boldsymbol x\right], \qquad \boldsymbol \theta_N \in \Theta_N,
-\] is the normalizing function. Let
-\(\boldsymbol \theta_N = (\boldsymbol \alpha, \boldsymbol \beta,\Gamma) \in \Theta_N \subset \mathbb{R}^{q(N)}\)
-with \(q(N) = N + N_H + N*N_H\) denote the vector of parameters for the
-RBM. The probability mass function for the visible variables
-\(X_1, \dots, X_N\) follows from marginalizing the joint specification
-\[
-P_{\boldsymbol \theta_N} (\boldsymbol x) = \sum\limits_{\boldsymbol h \in \mathcal{X}^{N_H}} P_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h), \qquad \boldsymbol x \in \mathcal{X}^N.
-\] Note that the vector of model parameters \(\boldsymbol \theta_N\), of
-size \(q(N)\), grows in size as a function of sample dimension \(N\) to
-accommodate the dimension of visible variables \(X_1, \dots, X_N\), and
-one may further choose the number \(N_H\) of hidden variables to change
-with \(N\) as well. In particular, the number \(N_H\) of hiddens may
-also potentially and arbitrarily increase with \(N\). Additionally, as
-\(|\mathcal{X}| = 2\) and
+\(\boldsymbol X = (X_1,\ldots,X_N)\) denote the \(N\) random variables
+for visibles with support \(\mathcal{X}^N\) and
+\(\boldsymbol H = (H_1,\ldots,H_{N_\mathcal{H}})\) denote the
+\(N_\mathcal{H}\) random variables for hiddens with support
+\(\mathcal{X}^{N_\mathcal{H}}\) where \(\mathcal{X} = \{-1,1\}\). For
+parameters
+\(\boldsymbol \theta_N^{\mathcal{H}} \in \mathbb{R}^{N_\mathcal{H}}\),
+\(\boldsymbol \theta_N^{\mathcal{V}}\in \mathbb{R}^N\), and
+\(\boldsymbol \theta_N^{\mathcal{HV}}\) as a real matrix with dimension
+\(N_\mathcal{H} \times N\), the RBM model for
+\(\tilde{\boldsymbol X}=(\boldsymbol X,\boldsymbol H)\) has the joint
+probability mass function
+\begin{equation}
+\label{eq:RBM1}
+\tilde{P}_{\boldsymbol \theta_N} (\tilde{\boldsymbol x}) = \exp\left[ (\boldsymbol \theta_N^{\mathcal{H}})^T \boldsymbol h + \boldsymbol (\boldsymbol \theta_N^{\mathcal{V}})^T \boldsymbol x + \boldsymbol h^T  \boldsymbol\theta_N^{\mathcal{HV}} \boldsymbol x - \psi(\boldsymbol \theta_N)\right], \quad \tilde{\boldsymbol x} = (\boldsymbol x, \boldsymbol h) \in \{\pm 1\}^{N+N_\mathcal{H}}
+\end{equation}
+with normalizing function \[
+\psi(\boldsymbol \theta_N) = \log \sum_{\tilde{\boldsymbol x} \in \{\pm 1\}^{N+N_H} } \exp\left[ (\boldsymbol \theta_N^{\mathcal{H}})^T \boldsymbol h + \boldsymbol (\boldsymbol \theta_N^{\mathcal{V}})^T \boldsymbol x + \boldsymbol h^T  \boldsymbol\theta_N^{\mathcal{HV}} \boldsymbol x\right].
+\] Let
+\(\boldsymbol \theta_N = (\boldsymbol \theta_N^{\mathcal{H}}, \boldsymbol \theta_N^{\mathcal{V}}, \boldsymbol\theta_N^{\mathcal{HV}} ) \in \Theta_N \equiv \mathbb{R}^{q(N)}\),
+with \(q(N) = N + N_\mathcal{H} + N*N_\mathcal{H}\), denote the
+parameter vector for the RBM, as indexed by the number \(N\) of visible
+random variables (which may differ from the actual lengths of these
+parameter vectors). The probability mass function for the visible
+variables \(\boldsymbol X = (X_1, \dots, X_N)\) follows from
+marginalizing the joint specification to yield
+\begin{equation}
+\label{eq:RBM2}
+P_{\boldsymbol \theta_N} (\boldsymbol x) = \sum\limits_{\boldsymbol h \in \{\pm 1\}^{N_{\mathcal{H}}}} \tilde{P}_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h), \qquad \boldsymbol x \in \{\pm 1\}^{N}\equiv \mathcal{X}^N.
+\end{equation}
+Here the baseline model \eqref{eq:RBM1} for hidden/visible variables is a
+linear exponential one in sufficient statistics
+\((\tilde{\boldsymbol X}, \boldsymbol X^T\boldsymbol H)\) using
+\(\tilde{\boldsymbol X}=(\boldsymbol X,\boldsymbol H)\) from
+\eqref{eq:RBM1}, but the form differs from the previous exponential models
+in \eqref{eq:expo} in that the lengths of parameters
+\(\boldsymbol \theta_N\) and statistics
+\((\tilde{\boldsymbol X}, \boldsymbol X^T\boldsymbol H)\) increase to
+incorporate more visible variables. That is, in contrast to
+\eqref{eq:expo}, the natural parameter function involved in the RBM model
+\eqref{eq:RBM1}, as the identity mapping of the parameters
+\(\boldsymbol \theta_N\in\mathbb{R}^{q(N)}\), naturally grows in
+dimension \(q(N)\to \infty\) to accommodate visible variables
+\(X_1, \dots, X_N\) of increasing sample size \(N\to \infty\).
+Additionally, one may further arbitrarily choose the number
+\(N_\mathcal{H}\) of hidden variables \(\boldsymbol H\) in the joint RMB
+model \eqref{eq:RBM1} to define a marginal model \eqref{eq:RBM2} for the
+\(N\) visible variables \(\boldsymbol X\), and the number
+\(N_\mathcal{H}\) of hiddens may also potentially increase with \(N\).
+Because \(|\mathcal{X}| = 2\) and
 \(P_{\boldsymbol \theta_N}(\boldsymbol x) > 0\) for all
-\(\boldsymbol x \in \mathcal{X}^N\), the RBM model specification for
-visibles is a FSFS model. This example also indicates that models formed
-by marginalizing a base FSFS model (e.g., a type of exponential family
-model) is again a FSFS model class.
+\(\boldsymbol x \in \mathcal{X}^N\), the RBM specification \eqref{eq:RBM2}
+for visibles \(\boldsymbol\) corresponds to a FOES model, while the
+joint distribution \eqref{eq:RBM1} for \((\boldsymbol X, \boldsymbol H)\)
+is also a FOES model. As this example also indicates, any model formed
+by marginalizing a base FOES model class, such as the RBM joint
+specification \eqref{eq:RBM1}, is again a FOES model.
 
@@
 \protect\hyperlink{ref-salakhutdinov2009deep}{2009}) and a deep belief
-network (Hinton, Osindero, and Teh
-\protect\hyperlink{ref-hinton2006fast}{2006}). Let \(M\) denote the
+network Hinton, Osindero, and Teh
+(\protect\hyperlink{ref-hinton2006fast}{2006}){]}. Let \(M\) denote the
 number of hidden layers included in the model and let
@@
 layers. The joint probability mass function for a DBM is \[
-P_{\boldsymbol \theta_N} (\tilde{\boldsymbol x}) = \exp\left[ \sum\limits_{i = 1}^M\boldsymbol \alpha^{(i)T} \boldsymbol h^{(i)} + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^{(1)T} \Gamma^{(0)} \boldsymbol x + \sum\limits_{i = 1}^{M - 1} \boldsymbol h^{(i)T} \Gamma^{(i)} \boldsymbol h^{(i + 1)} - \psi(\boldsymbol \theta_N) \right],
+\tilde{P}_{\boldsymbol \theta_N} ( \tilde{\boldsymbol x} ) = \exp\left[ \sum\limits_{i = 1}^M\boldsymbol \alpha^{(i)T} \boldsymbol h^{(i)} + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^{(1)T} \Gamma^{(0)} \boldsymbol x + \sum\limits_{i = 1}^{M - 1} \boldsymbol h^{(i)T} \Gamma^{(i)} \boldsymbol h^{(i + 1)} - \psi(\boldsymbol \theta_N) \right],
 \] for
@@
 \psi(\boldsymbol \theta_N) = \log \sum\limits_{\tilde{\boldsymbol x} \in \mathcal{X}^{N_{(H,1)} + \cdots + N_{(H,M)} +N}} \exp\left[ \sum\limits_{i = 1}^M\boldsymbol \alpha^{(i)T} \boldsymbol h^{(i)} + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^{(1)T} \Gamma^{(0)} \boldsymbol x + \sum\limits_{i = 1}^{M - 1} \boldsymbol h^{(i)T} \Gamma^{(i)} \boldsymbol h^{(i + 1)}\right],
-\] for \(\boldsymbol \theta_N \in \Theta_N\) is the normalizing function
-and parameters in the model are \(\boldsymbol \beta \in \mathbb{R}^V\),
-\(\boldsymbol \alpha^{(i)} \in \mathbb{R}^{N_{(H,i)}}\) for
+\] is the normalizing function for
+\(\boldsymbol \theta_N = (\boldsymbol \alpha^{(1)}, \dots, \boldsymbol \alpha^{(M)}, \boldsymbol \beta,\Gamma^{(0)}, \dots, \Gamma^{(M - 1)}) \in \Theta_N \subset \mathbb{R}^{q(N)}\),
+consisting of model parameters \(\boldsymbol \beta \in \mathbb{R}^N\),
+\(\boldsymbol \alpha^{(i)} \in \mathbb{R}^{N_{(H,i)}}\),
 \(i = 1, \dots, M\), along with a matrix \(\Gamma^{(0)}\) of dimension
 \(N_{(H,1)} \times N\), and matrices \(\Gamma^{(i)}\) of dimension
-\(N_{(H,i)} \times N_{(H,i+1)}\) for \(i = 1, \dots, M-1\). Let
-\(\boldsymbol \theta_N = (\boldsymbol \alpha^{(1)}, \dots, \boldsymbol \alpha^{(M)}, \boldsymbol \beta,\Gamma^{(0)}, \dots, \Gamma^{(M - 1)}) \in \Theta_N \subset \mathbb{R}^{q(N)}\)
-denote the combined vector of parameters with total length
+\(N_{(H,i)} \times N_{(H,i+1)}\) for \(i = 1, \dots, M-1\). The combined
+parameter vector \(\boldsymbol \theta_N\) has total length
 \(q(N)= N_{(H,1)}+\cdots N_{(H,M)} + N + N_{(H,1)}*N+N_{H,2}*H_{(H,1)}+\cdots +N_{(H,M)}*H_{(H,M)-1}\).
-
 The probability mass function for the visible random variables
 \(X_1, \dots, X_N\) follows from this joint specification as \[
-P_{\boldsymbol \theta_N} (\boldsymbol x) = \sum\limits_{(\boldsymbol h^{(1)}, \dots, \boldsymbol h^{(M)}) \in \mathcal{X}^{N_{(H,1)} + \cdots + N_{(H,M)}}} P_{\boldsymbol \theta_N} (\tilde{\boldsymbol x}), \qquad \boldsymbol x \in \mathcal{X}^N
+P_{\boldsymbol \theta_N} (\boldsymbol x) = \sum\limits_{(\boldsymbol h^{(1)}, \dots, \boldsymbol h^{(M)}) \in \mathcal{X}^{N_{(H,1)} + \cdots + N_{(H,M)}}} \tilde{P}_{\boldsymbol \theta_N} (\boldsymbol h^{(1)}, \dots, \boldsymbol h^{(M)}, \boldsymbol x) , \qquad \boldsymbol x \in \mathcal{X}^N.
 \] Again like the RBM case, the DBM model specification is an example of
-a FSFS model.
+a FOES model.
 
@@
 for visibles \(X_1, \dots, X_N\) with support \(\mathcal{X}^N\), a DBN
-is also a FSFS model if the number \(q(N)\) of components in the
+is also a FOES model if the number \(q(N)\) of components in the
 parameter vector is dependent on the dimension of the visibles.
@@
 \(X_1, \dots, X_N\), and the first layer of latent variables,
-\(H^{(1)}_1, \dots, H^{(1)}_{N_{(H,1)}}\), in the definition of a FSFS
+\(H^{(1)}_1, \dots, H^{(1)}_{N_{(H,1)}}\), in the definition of a FOES
 model.
 
-\section{Instability results}\label{instability-results}
+\section{Main results on model instability}\label{instability-results}
 
-To define or measure instability in FSFS models, it is useful to
-consider the behavior of a data model sequence \(P_{\theta_N}\). A
-relevant quantity to this end is a (scaled) extremal log-probability
-ratio (ELPR)
+\subsection{A criterion for instability}\label{criterion}
+
+To define a measure of instability in FOES models, it is useful to
+consider the behavior of data models \(P_{\theta_N}\), again supported
+on a set \(\mathcal{X}^N\) of outcomes for
+\(\boldsymbol X\equiv \boldsymbol X_N =(X_1,\ldots,X_N)\), in connection
+to the sample size \(N\). A relevant quantity to this end is a log-ratio
+of extremal probabilities (LREP), defined as
 \begin{align}
-\frac{1}{N} \log \left[\frac{\max\limits_{(x_1, \dots, x_N) \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(x_1, \dots, x_N)}{\min\limits_{(x_1, \dots, x_N) \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(x_1, \dots, x_N)}\right] \equiv \frac{1}{N} \text{ELPR}_N(\boldsymbol \theta_N).
 \label{eq:elpr}
+ \REP (\boldsymbol \theta_N)  =  \log \left[\frac{\max\limits_{  \boldsymbol x\in \mathcal{X}^N}P_{\boldsymbol \theta_N}( \boldsymbol x)}{\min\limits_{ \boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}( \boldsymbol x)}\right],
 \end{align}
-The main idea is that, in formulating FSFS models for potentially
-increasing numbers of variables (i.e., for \(N \rightarrow \infty\)),
-the ratio \eqref{eq:elpr} should remain bounded to ensure model stability,
-requiring that the largest probability possible under \(P_{\theta_N}\)
-maintain a fixed order of magnitude relative to the smallest probability
-allowed under the same model. Specifically, the log of the ratio should
-grow at mostly linearly with the sample size \(N\). This leads to the
-following definition.
-\BeginKnitrBlock{definition}[S-unstable FSFS]
+based on maximum and minimal model probabilities. In what follows, the
+main idea is that instability, and other negative model features, can be
+associated with a FOES model formulation for \(N\) random variables
+where the log-ratio \eqref{eq:elpr} is overly large relative to the sample
+size \(N\). That is, a sequence of FOES probability models
+\(P_{\boldsymbol \theta_N}\) results in specifying the distribution of
+observations \(\boldsymbol X=(X_1,\ldots,X_N)\) for each sample size
+\(N \geq 1\) and instability will generally occur among these models
+whenever the corresponding LREP \eqref{eq:elpr} grows faster than \(N\).
+This leads to the following definition.
+
+\BeginKnitrBlock{definition}[S-unstable FOES model]
 \protect\hypertarget{def:instabFSFS}{}{\label{def:instabFSFS}
-\iffalse (S-unstable FSFS) \fi{} }Let
-\(\boldsymbol \theta_N \in \mathbb{R}^{q(N)}\) be a sequence of FSFS
-model parameters where the size of the model \(q(N)\) is a function of
-the number of random variables \(N\). A FSFS model formulation is
-\emph{Schweinberger-unstable} or \emph{S-unstable} if, as the number of
-variables increase (\(N \rightarrow \infty\)),
-\begin{align*}
-\lim\limits_{N \rightarrow \infty} \frac{1}{N} \text{ELPR}(\boldsymbol \theta_N) \equiv \lim\limits_{N \rightarrow \infty} \frac{1}{N} \log \left[\frac{\max\limits_{(x_1, \dots, x_N) \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(x_1, \dots, x_N)}{\min\limits_{(x_1, \dots, x_N) \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(x_1, \dots, x_N)}\right] = \infty.
-\end{align*}
-\EndKnitrBlock{definition} In other words, a model is S-unstable if,
-given any \(C > 0\), there exists an integer \(N_C > 0\) so that
-\(\frac{1}{N}\text{ELPR}_N(\boldsymbol \theta_N) > C\) for all
-\(N \ge N_C\). A FSFS model formulation may be termed S-stable if it
-fails to be S-unstable.
+\iffalse (S-unstable FOES model) \fi{} }A FOES model formulation for
+\(\boldsymbol X_N=(X_1,\ldots,X_N)\) is \emph{Schweinberger-unstable} or
+\emph{S-unstable} if
+\begin{equation}
+\label{eq:Sun}
+\lim \limits_{N \rightarrow \infty} \frac{1}{N} \REP(\boldsymbol \theta_N) \equiv \lim \limits_{N \rightarrow \infty} \frac{1}{N}\log \left[\frac{\max\limits_{  \boldsymbol x\in \mathcal{X}^N}P_{\boldsymbol \theta_N}( \boldsymbol x)}{\min\limits_{ \boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}( \boldsymbol x)}\right] = \infty
+\end{equation}
+as the number of variables increases (\(N \rightarrow \infty\)).
+\EndKnitrBlock{definition} In other words, a model is S-unstable if
+\(\REP(\boldsymbol \theta_N)/N\) is an unbounded sequence of sample size
+\(N\); namely, given any \(C > 0\), there exists an integer \(N_C > 0\)
+so that \(\REP(\boldsymbol \theta_N)/N > C\) holds for all
+\(N \ge N_C\). A FOES model formulation may be termed S-stable if it
+fails to be S-unstable, i.e., if
+\(\sup_{N \geq 1}\REP(\boldsymbol \theta_N)/N\) is bounded.
 
@@
 (\protect\hyperlink{ref-schweinberger2011instability}{2011}) by allowing
-non-exponential family models (e.g., RBM and DBM models in Sections
-\ref{restricted-boltzmann-machines}-\ref{deep-learning}) and an
-increasing number of parameters. While this definition differs in form
-and scope from the original, it does match that in Schweinberger
+possibility non-exponential family models (e.g., RBM and DBM models in
+Sections \ref{restricted-boltzmann-machines} - \ref{deep-learning} as
+well as a potentially increasing number \(d(N)\) of parameters through
+the parameter sequence \(\boldsymbol \theta_N\in \mathbb{R}^{d(N)}\).
+While this definition differs in form and scope from the original, it
+does match that in Schweinberger
 (\protect\hyperlink{ref-schweinberger2011instability}{2011}) for the
 special case of exponential models (cf.~Section
-\ref{discrete-exponential-family-models}) considered there.
+\ref{discrete-exponential-family-models} considered there. Section
+\ref{illustrations} provides several examples of unstable models as well
+as causes for model instability, where the latter may often be traced to
+issues in model form (i.e., data functions) and/or parameterization. We
+next describe several potentially undesirable features associated with
+unstable models.
 
-S-unstable FSFS model sequences may be undesirable for several reasons.
-One is that small changes in data can lead to very large changes in
+\BeginKnitrBlock{remark}
+\iffalse <span class="remark"><em>Remark. \fi{} In the definition
+\eqref{eq:Sun} of S-instability, we note that the numerical measure
+\(\REP(\boldsymbol \theta_N)/N\) of model instability is invariant to
+\emph{independent replications} of data. That is, let \(M \geq 1\)
+denote a possible number of replications and consider data
+\(\boldsymbol Y_{N,M} \equiv (\boldsymbol X^{(1)}_N, \dots, \boldsymbol X^{(M)}_N)\)
+formed by \(\{ \boldsymbol X^{(j)}_N\}_{j=1}^M\) as \(M\) iid
+replications of a random vector \(\boldsymbol X_N=(X_1,\ldots,X_N)\),
+where the latter follows a FOES model with probabilities
+\(P_{ \boldsymbol \theta_N}(\boldsymbol x)>0\),
+\(\boldsymbol x\in\mathcal{X}^N\). This leads to a joint model, say
+\(P_{ \boldsymbol \theta_N}(\boldsymbol y)\),
+\(\boldsymbol y\in\mathcal{X}^{NM}\), for \(\boldsymbol Y_{N,M}\)
+consisting of \(N*M\) random variables in total. Then, the LREP for
+\(\boldsymbol Y_{N,M}\), scaled by associated size, is given by
+\begin{align*}
+\frac{1}{NM}\REP_{\boldsymbol Y_{N,M}}( \boldsymbol\theta_N ) &\equiv \frac{1}{NM}\log\left[\frac{\max_{\boldsymbol y \in \mathcal{X}^{NM}}P_{ \boldsymbol \theta_N}(\boldsymbol y)  }{\min_{\boldsymbol y \in \mathcal{X}^{NM}}P_{ \boldsymbol \theta_N}(\boldsymbol y)} \right] \\
+&= \frac{1}{NM}\log\left[\frac{\max_{\boldsymbol x \in \mathcal{X}^{N}}P_{ \boldsymbol \theta_N}(\boldsymbol x)  }{\min_{\boldsymbol x \in \mathcal{X}^{N}}P_{ \boldsymbol \theta_N}(\boldsymbol x)} \right]^M \equiv \frac{1}{N}\REP_{\boldsymbol X_{N}}( \boldsymbol\theta_N ),
+\end{align*}
+where
+\(\REP_{\boldsymbol X_{N}}( \boldsymbol\theta_N ) \equiv \REP( \boldsymbol\theta_N )\)
+denotes the log-ratio of extremal probabilities for \(\boldsymbol X_N\)
+defined from \eqref{eq:elpr}. That is, due to iid properties, the
+sample-size corrected LREP for \(\boldsymbol Y_{N,M}\) equals the
+analog, \(\REP(\boldsymbol\theta_N )/N\), from the underlying common
+data model for \(\boldsymbol X_N\) alone, regardless of the level
+\(M \geq 1\) of independent replication. Consequently, the definition of
+an S-unstable model is unaffected by independent replication and all
+instability properties may be characterized by those of one observation
+from the common FOES model. For computational purposes, this aspect also
+implies that if the original data \(\boldsymbol X_N=(X_1,\ldots,X_N)\)
+in a FOES model consist of \(N\) iid random variables, then the
+size-scaled log-ratio \eqref{eq:elpr} may be calculated as \[
+\frac{1}{N}\REP( \boldsymbol\theta_N ) \equiv \frac{1}{N}\REP_{\boldsymbol X_{N}}( \boldsymbol\theta_N ) = \log\left[ \frac{\max_{ x \in \mathcal{X}}P_{\boldsymbol\theta_N}(X_1=x)} {\min_{ x \in \mathcal{X}}P_{\boldsymbol\theta_N}(X_1=x)} \right]
+\] based on the extremal probabilities of just one random variable
+\(X_1\).
+\EndKnitrBlock{remark}
+
+\subsection{Characterizations and consequences of
+instability}\label{characterizations-and-consequences-of-instability}
+
+As a basic characteristic, S-unstable FOES model sequences have
+extremely sensitive probability structures. One aspect is that small
+changes in data configuration can lead to very large changes in
 probability. Consider, for example, the quantity given by \[
-\Delta(\boldsymbol \theta_N) \equiv \max \left\{\log \frac{P_{\boldsymbol \theta_N}(\boldsymbol x)}{P_{\boldsymbol \theta_N}(\boldsymbol x^*)} : \boldsymbol x \text{ }\& \text{ } \boldsymbol x^* \in \mathcal{X}^N \text{ differ in exactly one component}\right\},
+\DN(\boldsymbol \theta_N) \equiv \max \left\{\log \frac{P_{\boldsymbol \theta_N}(\boldsymbol x)}{P_{\boldsymbol \theta_N}(\boldsymbol x^*)} : \boldsymbol x \text{ }\& \text{ } \boldsymbol x^* \in \mathcal{X}^N \text{ differ in exactly one component}\right\},
 \] which represents the biggest log-probability ratio for a one
-component change in data outcomes at a FSFS parameter
-\(\boldsymbol \theta_N\). We then have the following (non-asymptotic)
-result.
+component change in data outcomes in a FOES model with parameter
+\(\boldsymbol \theta_N\). We then have the following result prescribing
+the behavior of \(\DN(\boldsymbol \theta_N)\) for unstable models.
+
+\BeginKnitrBlock{theorem}
+\protect\hypertarget{thm:instab-elpr}{}{\label{thm:instab-elpr}}Let
+\(P_{\boldsymbol \theta_N}\), with support \(\mathcal{X}^N\),
+\(N\geq 1\), be a sequence of FOES models.
+\begin{enumerate}[(i)]
+\item For any integer $N \geq 1$ and any given $C>0$, if $\REP(\boldsymbol \theta_N)/N > C$ in (\ref{eq:elpr}, then
+    $$ \DN(\boldsymbol \theta_N) > C,$$
+    or probabilities from a one component change in some outcome have log-ratio exceeding $C$.
+\item Suppose the FOES model sequence is S-unstable.
+Then, for all large $N$ and given any arbitrary $C>0$, there exist outcomes $\boldsymbol x,\boldsymbol x^*\in\mathcal{X}^N$, differing by one component, such that
+    $$
+    \frac{P_{\boldsymbol \theta_N}(\boldsymbol x)}{P_{\boldsymbol \theta_N}(\boldsymbol x^*)} > \exp[N C].
+    $$
+\end{enumerate}
+\EndKnitrBlock{theorem}
+
+Theorem \ref{thm:instab-elpr}(i) is a non-asymptotic result, which
+connects to the definition of instability in a FOES model through a
+log-ratio of extreme model probabilities \eqref{eq:elpr} being too large
+relative to the associated sample size \(N\). If so, Theorem
+\ref{thm:instab-elpr}(i) guarantees the FOES model must also exhibit
+correspondingly large changes in probability for very small differences
+among some data configurations, a property that intuitively captures a
+notion of instability. Furthermore, and perhaps more seriously under
+Theorem \ref{thm:instab-elpr}(ii), S-unstable models can never have
+universally bounded changes in probability among single component
+variations in data configurations. While not all one component changes
+in data may produce massive changes in probability, unstable models must
+have some such data outcomes with this property. As a consequence,
+unstable probability structure may exhibit extreme sensitivity through
+large peaks and troughs over the sample space.
+
+Additionally, S-unstable FOES model sequences are also connected to
+degenerate models, where \emph{degeneracy} involves assigning
+essentially all probability to modes within the sample space, which
+could potentially represent a small subset among the totality of
+outcomes. For perspective, note that differing sizes of the scaled
+log-ratio \(\REP(\boldsymbol \theta_N)/N\) from \eqref{eq:elpr} induce a
+spectrum of levels of instability/stability and Theorem
+\ref{thm:instab-elpr} indicates increasing sensitivity of model
+probabilities as \eqref{eq:elpr} increases. Furthermore, as the
+instability measure grows and the log-ratio
+\(\REP(\boldsymbol \theta_N)/N\) diverges, as in the definition
+\eqref{eq:Sun} of S-unstable models, then a FOES model sequence will
+become degenerate. Theorem \ref{thm:degenFOES} provides a formal
+statement of such degeneracy due to instability. For a given
+\(0 < \epsilon < 1\), define a \(\epsilon\)-modal set of outcomes as
+\begin{equation}
+\label{eq:mode}
+\mathcal{M}_{\epsilon, \boldsymbol \theta_N} \equiv \left\{\boldsymbol x \in \mathcal{X}^N: \log P_{\boldsymbol \theta_N}(\boldsymbol x) > (1-\epsilon)\max\limits_{\boldsymbol y \in \mathcal{X}^N} \log  P_{\boldsymbol \theta_N}(\boldsymbol y) + \epsilon\min\limits_{\boldsymbol y \in \mathcal{X}^N} \log P_{\boldsymbol \theta_N}(\boldsymbol y) \right\}.
+\end{equation}
+\BeginKnitrBlock{theorem}
+\protect\hypertarget{thm:degenFOES}{}{\label{thm:degenFOES}}For any
+arbitrarily small \(0 < \epsilon < 1\), an S-unstable FOES model
+sequence \(P_{\boldsymbol \theta_N}\), \(N \geq 1\), for
+\(\boldsymbol X_N=(X_1, \dots, X_N)\) satisfies \[
+P_{\boldsymbol \theta_N}\left( \boldsymbol X_N\in \mathcal{M}_{\epsilon, \boldsymbol \theta_N}\right) \rightarrow 1 \text{ as } N \rightarrow \infty.
+\]
+\EndKnitrBlock{theorem}
+
+In other words, as the sample size grows in S-unstable FOES models, all
+probability tends to concentrate mass on an \(\epsilon\)-modal set,
+where \(\epsilon\) can be made arbitrarily small. Intuitively, the
+occurrence of such degeneracy can be explained by a type of ``reverse''
+pigeonhole principle for unstable FOES models: if all outcomes should
+receive positive probability but the maximal probability far exceeds the
+minimal one in the model, then little probability remains for
+distribution among remaining model outcomes (i.e., if nearly all
+available pigeons are stuffed into one hole, the remaining pigeonholes
+must have few occupants). Degeneracy in unstable models can pose dangers
+in data modeling as well, particularly when a mode set represents a
+narrow collection of outcomes among those realistically possible for
+adequately describing data. In which case, model outcomes may fail to
+look like data of interest.
+
+Connected to degeneracy, unstable models may also exhibit additional
+kinds of extreme and undesirable sensitivity in probabilities if model
+parameters \(\boldsymbol \theta_N\) can further be ``dialed'' between
+positive and negative values. That is, some FOES models naturally
+involve parameter spaces covering a positive-negative spectrum of
+parameter possibilities, where the signs of parameters provide a
+standard device for increasing or decreasing probabilities of outcomes
+in the model formulation. In fact, for many models, the switch of a
+parameter sign serves to produce reciprocal probabilities, as outlined
+in the following model assumption about parameter sign reversal (PSR).
+
+\textbf{Model Condition PSR} \emph{(Reciprocal Probabilities from
+Parameter Sign Reversal)}: Let \(P_{\boldsymbol \theta_N}\), with
+support \(\mathcal{X}^N\), \(N\geq 1\), represent a sequence of FOES
+models. For each \(N \geq 1\) and any outcome
+\(\boldsymbol x \in \mathcal{X}^N\), suppose it holds that \[
+P_{\boldsymbol \theta_N}(\boldsymbol x)  \cdot P_{-\boldsymbol \theta_N}(\boldsymbol x) \;=\;   \max\limits_{\boldsymbol y \in \mathcal{X}^N}P_{ \boldsymbol \theta_N}(\boldsymbol y)\cdot \min\limits_{\boldsymbol y \in \mathcal{X}^N}P_{-\boldsymbol \theta_N}(\boldsymbol y),
+\] where
+\(\max_{\boldsymbol y \in \mathcal{X}^N}P_{ \boldsymbol \theta_N}(\boldsymbol y)\)
+and
+\(\min_{\boldsymbol y \in \mathcal{X}^N}P_{-\boldsymbol \theta_N}(\boldsymbol y)\)
+denote the maximum and minimum probabilities under parameters
+\(\boldsymbol \theta_N\) and \(-\boldsymbol \theta_N\), respectively.
+
+The above model condition incorporates many standard parameterizations
+and follows, for instance, whenever
+\(P_{\boldsymbol \theta_N}(\boldsymbol x)/P_{\boldsymbol \theta_N}(\boldsymbol y) = [P_{-\boldsymbol \theta_N}(\boldsymbol x)/P_{-\boldsymbol \theta_N}(\boldsymbol y)]^{-1}\)
+holds for outcomes \(\boldsymbol x, \boldsymbol y \in\mathcal{X}^N\) in
+a FOES model. For instance, this latter condition is fulfilled for all
+linear exponential families from Section
+\ref{discrete-exponential-family-models} (e.g.,
+\eqref{eq:mod1}-\eqref{eq:mod2}) as well as all network models from Sections
+\ref{restricted-boltzmann-machines}-\ref{deep-learning} (e.g.,
+\eqref{eq:RBM1}-\eqref{eq:RBM2}). When parameters can be tuned in sign with
+effects prescribed in the model condition PSR, unstable FOES models will
+exhibit further probability sensitivities, as outlined in the following
+extension of Theorem @ref(thm:degenFOES\}).
+
+\BeginKnitrBlock{corollary}
+\protect\hypertarget{cor:sign}{}{\label{cor:sign}}Let
+\(P_{\boldsymbol \theta_N}\), with support \(\mathcal{X}^N\),
+\(N\geq 1\), be a sequence of FOES models satisfying model condition
+PSR. If the models \(P_{\boldsymbol \theta_N}\) are additionally
+S-unstable, then
+\begin{enumerate}[(i)]
+\item the models  $P_{-\boldsymbol \theta_N}$  defined by $-\boldsymbol \theta_N$  are also S-unstable;
+\item and for the complement $\mathcal{M}_{\epsilon, \boldsymbol \theta_N}^c \equiv \mathcal{X}^N \setminus \mathcal{M}_{\epsilon, \boldsymbol \theta_N}$ of any mode-set $\mathcal{M}_{\epsilon, \boldsymbol \theta_N}$ under $\boldsymbol \theta_N$ from (\ref{eq:mode}), with $0<\epsilon<1$, it holds under $- \boldsymbol \theta_N$ that
+    $$ 
+    P_{-\boldsymbol \theta_N}( \boldsymbol X_N \in \mathcal{M}_{\epsilon, \boldsymbol \theta_N}^c) \rightarrow 1\qquad \mbox{as $N\to \infty$},
+    $$
+    while, by Theorem \ref{thm:degenFOES}, $P_{\boldsymbol \theta_N}( \boldsymbol X_N \in \mathcal{M}_{\epsilon, \boldsymbol \theta_N} ) \rightarrow 1$ holds for $\boldsymbol X_N = (X_1,\ldots,X_N)$ under $\boldsymbol \theta_N$.
+\end{enumerate}
+\EndKnitrBlock{corollary}
+
+For unstable models, Corollary \ref{cor:sign} shows that shifts in
+parameters around zero (i.e., from \(\boldsymbol \theta_N\) to
+\(-\boldsymbol \theta_N\)) can induce extreme changes in probability
+among subsets of the sample space, as another manifestation of
+instability and hyper-sensitivity in probability structure. For
+one-parameter exponential families, involving a fixed real-valued linear
+parameter \(\boldsymbol \theta_N = \theta \in \mathbb{R}\) and
+sufficient statistic \(\boldsymbol g_N(\boldsymbol x)\in \mathbb{R}\) in
+\eqref{eq:expo}, Schweinberger
+(\protect\hyperlink{ref-schweinberger2011instability}{2011} Theorem 3)
+proved a result similar in spirit, though based on a characterization
+there in terms of maximum
+\(U_N \equiv \max_{\boldsymbol x\in\mathcal{X}^N}g_N(\boldsymbol x)\)
+and minimal
+\(L_N \equiv \min_{\boldsymbol x\in\mathcal{X}^N}g_N(\boldsymbol x)\)
+values of the sufficient statistic. For this case in particular, mode
+sets have specific, and essentially complementary, forms over positive
+and negative parameters, namely,
+\(\mathcal{M}_{\epsilon, \boldsymbol \theta_N} = \{\boldsymbol x \in\mathcal{X}^N: g_N(\boldsymbol x) > (1-\epsilon) U_N + \epsilon L_N \}\)
+and
+\(\mathcal{M}_{\epsilon, -\boldsymbol \theta_N} = \{\boldsymbol x \in\mathcal{X}^N: g_N(\boldsymbol x) < \epsilon U_N + (1-\epsilon) L_N \}\)
+for any \(\boldsymbol \theta_N>0\), and Schweinberger
+(\protect\hyperlink{ref-schweinberger2011instability}{2011} Theorem 3)
+showed each mode set collects all mass, under positive and negative
+parameters, respectively, with unstable models of this exponential type.
+However, for all unstable FOES models, Corollary \ref{cor:sign}
+generalizes the same principle that unstable models can push all
+probability to different, and in fact disjoint, parts of the sample
+space, depending on how parameters fall with respect to zero. This
+feature can numerically complicate likelihood manipulations, such as
+maximization or MCMC-based Bayes posterior sampling, as further
+discussed in Section \ref{implications}.
+
+\BeginKnitrBlock{remark}[need to briefly finish]
+\iffalse <span class="remark"><em>Remark (need to briefly finish). \fi{}
+extensions of sensitivity\ldots{} basically, model condition PSR and
+Corollary \ref{cor:sign} can be extended to cases where parameter
+components
+\(\boldsymbol \theta_N =(\boldsymbol \theta_{1,N}, \boldsymbol \theta_{N,2})\)
+are not all changed in sign (\(-\boldsymbol \theta_N\)) but, more
+generally, are instead altered to another parameter configuration
+\(\boldsymbol \theta_{N}^A = ( \boldsymbol \theta_{1,N}^A, \boldsymbol \theta_{2,N}^A )\)
+involving a switch in sign only among some model parameters
+\(\boldsymbol \theta_{N,2}^A=- \boldsymbol \theta_{N,2}\) with remaining
+parameters \(\boldsymbol \theta_{1,N}^A\) being arbitrarily chosen. If a
+change sign occurs among parameters (\(\pm \boldsymbol \theta_{N,2}\))
+which dominate the probability structure of the model, then an unstable
+model under \(\boldsymbol \theta_N\) will then imply that many more
+unstable models exist over a broad spectrum of possibilities for
+variations \(\boldsymbol \theta_{N}^A\) of \(\boldsymbol \theta_N\).
+\EndKnitrBlock{remark}
+
+\section{Illustrations}\label{illustrations}
+
+Model instability can depend intricately on how functions of parameters
+and data \(\boldsymbol X_N=(X_1,\ldots,X_n)\) are combined in the
+formulation of the model probabilities, though some general causes may
+be identified. As one issue, a broad parameter space (or wide
+interpretation of this space) may admit some parameters as technically
+valid that have an undue and often undesirable impact on the model
+structure for a prescribed data size \(N\). In this case, both the size
+and dimension of model parameters can be problematic and induce
+instability. In combination to this last point, further causes of
+instability may also be traced to the magnitude of statistics in the
+model. Potentially massive, and thereby unstable, statistics were the
+primary focus of instability studies of Schweinberger
+(\protect\hyperlink{ref-schweinberger2011instability}{2011}) for certain
+discrete exponential models having parameters/statistics of fixed
+dimension. However, as shown in the following, bounded statistics may
+still lead to instability if the parameter dimension is high. We next
+provides some examples to illustrate instability in FOES models, which
+also suggest some potential strategies for preventing unstable models.
+
+\subsection{Equi-probability models}\label{equi-probability-models}
+
+As a baseline for comparisons, consider a simplistic model for
+\(\boldsymbol X_N=(X_1,\ldots,X_N)\) with uniform probabilities over the
+sample space, say
+\(P_{\boldsymbol \theta_N}(\boldsymbol x)= |\mathcal{X}|^{-N}\),
+\(\boldsymbol x \in \mathcal{X}^N\), where each random variable has
+\(|\mathcal{X}| \geq 1\) outcomes. In contrast to instability, model
+probabilities here are completely insensitive to changes in data
+outcomes across the sample space, and the associated log-ratio of
+extreme probabilities \eqref{eq:elpr} is \[
+\frac{1}{N} \REP(\boldsymbol \theta_N)=0\quad \;(\mbox{uniform probability model}),
+\] which is as small as possible. In fact, a LREP value of zero can only
+occur for a FOES model having uniform probabilities, and such
+equi-probability models are always S-stable.
+
+\subsection{One-parameter exponential models}\label{one-param-exp}
+
+A fundamental model considered in the instability work of Schweinberger
+(\protect\hyperlink{ref-schweinberger2011instability}{2011}) involves a
+one-parameter exponential model corresponding to \eqref{eq:expo} with a
+real-valued parameter, say
+\(\boldsymbol \theta_N = \eta(\boldsymbol \theta_N)\in \mathbb{R}\), and
+sufficient statistic \(\boldsymbol g_N(\boldsymbol x)\in \mathbb{R}\).
+For such models, upon scaling by sample size \(N\), the log-ratio of
+extreme probabilities in \eqref{eq:elpr} for assessing instability becomes
+\begin{equation}
+\label{eq:UL}
+\frac{1}{N}\REP(\boldsymbol \theta_N ) \equiv   |\boldsymbol \theta_N| \frac{(U_N-L_N)}{N} \;\quad \mbox{(one-parameter exponential model)},
+\end{equation}
+where
+\(U_N \equiv \max_{\boldsymbol x\in\mathcal{X}^N}g_N(\boldsymbol x)\)
+and
+\(L_N \equiv \min_{\boldsymbol x\in\mathcal{X}^N}g_N(\boldsymbol x)\)
+denote the maximal and minimal values of the single sufficient
+statistic. In this case, an S-unstable model results, by definition
+\eqref{eq:Sun}, whenever
+\(\lim_{N\to \infty} |\boldsymbol \theta_N| (U_N-L_N)/N= \infty\) holds
+or, in other words, if the combined magnitudes of parameter
+\(|\boldsymbol \theta_N|\) and maximal difference \(U_N-L_N\) in
+statistic values are overwhelmingly large relative to the sample size
+\(N\). If we further assume that
+\(\boldsymbol \theta_N =\theta\in\mathbb{R}\setminus \{0\}\) is a fixed
+(non-zero) parameter for all \(N \geq 1\), as considered in
+Schweinberger
+(\protect\hyperlink{ref-schweinberger2011instability}{2011}), then an
+S-unstable model results solely if the sufficient statistic admits a
+value \(U_N-L_N\) too large relative to number \(N\) of observations,
+i.e., if \((U_N-L_N)/N\to \infty\) as \(N\to \infty\). The latter aspect
+reflects the definition of Schweinberger
+(\protect\hyperlink{ref-schweinberger2011instability}{2011}), for this
+setting, that a real-valued \emph{statistic} \(g_N(\boldsymbol x)\) may
+be classified as \emph{unstable} when
+\(\lim_{N\to \infty}|(U_N-L_N)/N=\infty\) holds and as \emph{stable}
+otherwise (e.g., if \(\sup_{N \geq 1}(U_N-L_N)/N<\infty\)).
+
+For illustration, consider the iid Bernoulli model \eqref{eq:mod1} for
+\(\boldsymbol X_N=(X_1,\ldots,X_N)\) with log-odds ratio parameter
+\(\boldsymbol \theta_N = \log[ P_{\boldsymbol \theta_N}(X_1=1)/ P_{\boldsymbol \theta_N}(X_1=0)]\in\mathbb{R}\).
+Remark \ref{rmk1} (Section \ref{criterion}) then gives the model
+instability measure \eqref{eq:Sun} directly as \[
+\frac{1}{N}\REP(\boldsymbol \theta_N ) = |\boldsymbol \theta_N|\quad\; \mbox{(iid Bernoulli model)},
+\] so that an unstable (or stable) model results for a divergent (or
+bounded) parameter sequence \(|\boldsymbol \theta_N|\). The above
+instability expression for the Bernoulli model follows as well from the
+\(N\)-scaled LREP value \eqref{eq:UL} for a one-parameter exponential
+distribution, using that the sufficient statistic involved
+\(g_N(\boldsymbol x)= \sum_{i=1}^N x_i\),
+\(\boldsymbol x =(x_1,\ldots,x_n)\in\{0,1\}^N\), has maximum and mimimum
+values \(U_N=N\) and \(L_N=0\). In this case, Schweinberger
+(\protect\hyperlink{ref-schweinberger2011instability}{2011}) has noted
+that the statistic is stable (i.e., bounded \((U_N-L_N)/N=1\)) and the
+Bernoulli model is as well when, in particular,
+\(\boldsymbol \theta_N=\bold \theta \in\mathbb{R}\) is fixed for
+\(N \geq 1\).
+
+Alternatively, considering a random graph with \(N={n \choose 2}\) edges
+among \(n\) nodes, the exponential graph model from \eqref{eq:mod2}, when
+based purely on the number of \(g_{2,N}(\boldsymbol x)\) of 2-stars or
+solely the number \(g_{3,N}(\boldsymbol x)\) of triangles,
+\(\boldsymbol x\in\{0,1\}^N\), has an measure of instability from
+\eqref{eq:Sun} as \[
+\frac{1}{N}\REP(\boldsymbol \theta_N )  = \left\{ \begin{array}{lcl} |\boldsymbol \theta_N| (n-2) && \mbox{(2-star graph model)}\\
+|\boldsymbol \theta_N|(n-2)/3 &&\mbox{(triangle graph model)},\end{array}\right.
+\] by using the (one-parameter exponential) LREP formula \eqref{eq:UL}
+with statistic maximums \(U_N= N(n-2)\) for 2-stars or \(U_N= N(n-2)/3\)
+for triangles and with minimums \(L_N=0\) in both cases. Because the
+variable number \(N\to \infty\) as the node number \(n\to \infty\), both
+counts of 2-stars and triangles are unstable statistics in the sense of
+Schweinberger
+(\protect\hyperlink{ref-schweinberger2011instability}{2011}) (i.e.,
+\(\lim_{N\to \infty} (U_N-L_N)/N=\infty\)). Furthermore, both types of
+graph models are always S-unstable for all possible of parameter
+sequences \(\boldsymbol \theta_N \in\mathbb{R}\) that are bounded away
+from zero (i.e.,
+\(\lim_{N\to \infty}\REP(\boldsymbol \theta_N )/N=\infty\) then holds,
+including the fixed parameter case
+\(\boldsymbol \theta_N=\theta\in\mathbb{R}\setminus \{0\}\) from
+Schweinberger
+(\protect\hyperlink{ref-schweinberger2011instability}{2011})).
+
+\subsection{Fixed-dimensional linear exponential
+models}\label{fixed-dim-exp}
+
+As a generalization of the one-parameter exponential case, we next
+consider linear exponential families \eqref{eq:expo} with \(k\) parameters
+\(\boldsymbol \theta_N = (\theta_{1,N},\ldots,\theta_{k,N})^\prime\) and
+\(k\) sufficient statistics
+\(\boldsymbol g_N(\boldsymbol x) = (g_{1,N}(\boldsymbol x),\ldots, g_{k,N}(\boldsymbol x))^\prime\).
+Here the dimension \(k\) of model parameters/statistics is fixed, and we
+next prescribe a condition helpful to avoiding instability in such
+models. For this, define
+\(U_{i,N}=\max_{\boldsymbol x \in\mathcal{X}^N} g_{i,N}(\boldsymbol x)\)
+and
+\(L_{i,N}=\min_{\boldsymbol x \in\mathcal{X}^N} g_{i,N}(\boldsymbol x)\)
+as the maximal and minimal values of the \(i\)th statistic,
+\(i=1,\ldots,k\), based on observations
+\(\boldsymbol X_N=(X_1,\ldots,X_N)\).
 
 \BeginKnitrBlock{proposition}
-\protect\hypertarget{prp:instab-elpr}{}{\label{prp:instab-elpr}}Let
-\(\text{ELPR}(\boldsymbol \theta_N)\) be as in \eqref{eq:elpr} for an
-integer \(N \ge 1\). For a given \(C>0\), if
-\[\frac{1}{N}\text{ELPR}_N(\boldsymbol \theta_N) > C,\] then
-\[\Delta_N(\boldsymbol \theta_N) > C.\]
+\protect\hypertarget{prp:prop1}{}{\label{prp:prop1}}Let
+\(P_{\boldsymbol \theta_N}\), \(N \geq 1\), denote linear exponential
+models (\ref{eq:expo}) with parameters
+\(\boldsymbol \theta_N = (\theta_{1,N},\ldots,\theta_{k,N})^\prime \in \mathbb{R}^k\)
+and statistics
+\(\boldsymbol g_N(\boldsymbol x) = (g_{1,N}(\boldsymbol x),\ldots, g_{k,N}(\boldsymbol x))^\prime \in \mathbb{R}^k\),
+for fixed \(k \geq 1\). Then, the models \(P_{\boldsymbol \theta_N}\)
+are \{\rm S-stable\} if
+\begin{equation}
+\label{eq:prop1}
+\sup_{N \geq 1}\frac{1}{N} \max_{1 \leq i \leq k }|\theta_{i,N}|(U_{i,N}-L_{i,N})<\infty
+\end{equation}
+holds, i.e., if
+\(\max_{1 \leq i \leq k } |\theta_{i,N}|(U_{i,N}-L_{i,N})/N\) is bounded
+sequence of sample size \(N\).
 \EndKnitrBlock{proposition}
 
-Again, if the probability ratio \eqref{eq:elpr} is too large, then the
-FSFS model will exhibit large changes in probability for very small
-differences in the data configuration, which exemplifies the intuitive
-notation of instability.
+\BeginKnitrBlock{remark}
+\iffalse <span class="remark"><em>Remark. \fi{} In the one-parameter
+exponential case \(k=1\), recall the exponential model is
+stable/unstable depending on whether
+\(\REP(\boldsymbol \theta_N)/N = |\theta_{1,N}|(U_{1,N}-L_{1,N})/N \equiv |\boldsymbol \theta_{N}|(U_{N}-L_{N})/N\)
+in \eqref{eq:UL} is convergent/divergent. Hence, for \(k=1\), the
+condition \eqref{eq:prop1} of Proposition \ref{prp:prop1} captures the
+same notion of S-stability based on \eqref{eq:UL}.
+\EndKnitrBlock{remark}
 
-Additionally, S-unstable FSFS model sequences are connected to
-degenerate models, where model \emph{degeneracy} typically entails
-placing all probability on a small portion of the sample space. For
-perspective, note that differing sizes of
-\(1/N\cdot\text{ELPR}(\boldsymbol \theta_N)\) in \eqref{eq:elpr} may
-induce a spectrum of levels of ``stability'' and Proposition
-\ref{prp:instab-elpr} indicates increasing sensitivity of model
-probabilities (i.e., for one component changes in outcomes) as
-\eqref{eq:elpr} increases. Furthermore, as the instability measure
-\eqref{eq:elpr} grows, FSFS model sequences are guaranteed to slide into
-full degeneracy as Proposition \ref{prp:degenFSFS} shows. Define a
-\(\epsilon\)-modal set \[
-M_{\epsilon, \boldsymbol \theta_N} \equiv \left\{\boldsymbol x \in \mathcal{X}^N: \log P_{\boldsymbol \theta_N}(\boldsymbol x) > (1-\epsilon)\max\limits_{\boldsymbol x^* \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x^*) + \epsilon\min\limits_{\boldsymbol x^* \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x^*) \right\}
-\] of possible outcomes, for a given \(0 < \epsilon < 1\).
+Proposition \ref{prp:prop1} provides a sufficient condition for the
+stability of linear exponential models with fixed parameter dimension
+\(k\geq 1\), whereby an S-stable model is guaranteed if the compounded
+magnitude of each combination of parameter \(\theta_{i,N}\) and
+sufficient statistic value \((U_{i,N}-L_{i,N})\) is bounded by the
+sample size \(N\), \(i=1,\ldots,k\). This supports the findings of
+Schweinberger
+(\protect\hyperlink{ref-schweinberger2011instability}{2011}), who showed
+degeneracy follows in such models under one type of violation of the
+condition \eqref{eq:prop1} in Proposition \ref{prp:prop1} (namely,
+involving \(k>1\) non-zero parameters with \(k-1\) statistics being
+\(O(N)\) bounded while one statistic diverges in maximal size faster
+than the number \(N\) of observations). To further illustrate the result
+in Proposition \ref{prp:prop1}, consider the multinomial distribution
+\eqref{eq:mod11} for \(\boldsymbol X_N=(X_1,\ldots,X_N)\) having
+\(k\geq 2\) categories \(\{1,\ldots,k\}\) and \(k\) parameters
+\(\boldsymbol \theta_N = (\theta_{1,N},\ldots,\theta_{k,N})^\prime\).
+The variables are iid under this model so that Remark \ref{rmk1}
+(Section \ref{criterion} yields the corresponding \(N\)-scaled log-ratio
+of extreme probabilities \eqref{eq:elpr} as
+\begin{align*}
+  \frac{1}{N}\REP(\boldsymbol \theta_N) &= \frac{\max_{1 \leq i \leq k} P_{\boldsymbol \theta_N}(X_1=i)}{\min_{1 \leq i \leq k} P_{\boldsymbol \theta_N}(X_1=i)}\\
+  &= \max_{1 \leq i \leq k} \theta_{i,N} - \min_{1 \leq i \leq k} \theta_{i,N} \qquad \mbox{(iid multinomial model)}.
+\end{align*}
+Hence, a multinomial model sequence is unstable (or stable) depending on
+whether (or not) the maximal parameter difference
+\(\max_{1 \leq i \leq k} \theta_{i,N} - \min_{1 \leq i \leq k} \theta_{i,N}\)
+diverges. Furthermore, using that each of the \(k\) sufficient (count)
+statistics from the multinomial model \eqref{eq:mod11} satisfies
+\((U_{i,N}-L_{i,N})/N=1\), we see that \eqref{eq:prop1} of Proposition
+\ref{prp:prop1} becomes purely a parameter condition,
+\(\sup_{N \geq 1}\max_{1 \leq i \leq k } |\theta_{i,N}| <\infty\), for
+ensuring that
+\(\REP(\boldsymbol \theta_N)/N =\max_{1 \leq i \leq k} \theta_{i,N} - \min_{1 \leq i \leq k} \theta_{i,N}\)
+is bounded and stability follows for the multinomial distribution.
+Additionally, a stable multinomial sequence (i.e., bounded
+\(\REP(\boldsymbol \theta_N)/N\)) turns out to be nearly equivalent to
+\eqref{eq:prop1} (e.g., these are the same if the smallest parameter
+\(\min_{1 \leq i \leq k } |\theta_{i,N}|\) remains bounded).
 
-\BeginKnitrBlock{proposition}
-\protect\hypertarget{prp:degenFSFS}{}{\label{prp:degenFSFS}}For an unstable
-FSFS model in Definition \ref{def:instabFSFS}, and for any given
-\(0 < \epsilon < 1\), \[
-P_{\boldsymbol \theta_N}\left((X_1, \dots, X_N) \in M_{\epsilon, \boldsymbol \theta_N}\right) \rightarrow 1 \text{ as } N \rightarrow \infty.
-\]
+When the condition \eqref{eq:prop1} of Proposition \ref{prp:prop1} is
+violated, this aspect suggests a potentially unstable model that may be
+investigated more closely. For example, consider the exponential graph
+model from \eqref{eq:mod2} involving counts of edges, 2-stars and
+triangles with fixed parameters
+\(\boldsymbol \theta_N = (\theta_{1},\theta_2,\theta_3)^\prime \in \mathbb{R}^3\)
+for \(N\geq 1\). If either the 2-star parameter \(\theta_2 \neq 0\) or
+triangle parameter \(\theta_3 \neq 0\) is non-zero, then
+\(\max_{1 \leq i \leq 3 } |\theta_{i}|(U_{i,N}-L_{i,N})/N \propto (n-2)\to \infty\)
+holds in \eqref{eq:prop1} by
+\((U_{2,N}-L_{2,N})/N = 3 (U_{3,N}-L_{3,N})/N=(n-2)\) for 2-star and
+triangle statistics (\(i=2,3\)), so that Proposition \ref{prp:prop1}
+hints that an unstable model may result when
+\(|\theta_2| + |\theta_3| \neq 0\). Relatedly, a result from
+Schweinberger
+(\protect\hyperlink{ref-schweinberger2011instability}{2011} Result 3)
+states that this model is unstable for all fixed parameters excluding
+cases \(\theta_2 =\theta_3=0\) or \(\theta_2 = - \theta_3/3\). However,
+more is true in line with the instability suggested by Proposition
+\ref{prp:prop1} whenever \(|\theta_2| + |\theta_3| \neq 0\) (i.e.,
+excluding \(\theta_2 =\theta_3=0\)).
+
+To see this, consider an even number \(n>2\) of nodes and let
+\(\boldsymbol x_0\) denote the data outcome in
+\(\mathcal{X}^N \equiv \{0,1\}^N\) with all \(N = {n \choose 2}\) edges
+being zero, let \(\boldsymbol x_1\) denote the outcome with all edges
+being 1, and let \(\boldsymbol x_2\) denote the edge configuration from
+dividing the nodes into two equal groups, with no edges within a group
+and all edges between the groups (so that no triangles exist in
+\(\boldsymbol x_2\)). Then, the \(N\)-scaled log-ratio \eqref{eq:elpr} for
+the exponential graph model \eqref{eq:mod2} can, by definition, be bounded
+below by
+\begin{align*}
+\frac{1}{N}\REP_N(\boldsymbol \theta_N) &\geq \max_{i=1,2}\frac{1}{N}
+\left| \log\left[ \frac{P_{\boldsymbol \theta_N}(\boldsymbol x_i)}{P_{\boldsymbol \theta_N}(\boldsymbol x_0)}\right] \right| \\
+&= (n-2) \max\left\{ \left| \theta_2 + \frac{\theta_3}{3}+\frac{\theta_1}{n-2} \right|, \frac{n}{4(n-1)} \left| \theta_2 + \frac{8\theta_1}{n-2} \right| \right\};
+\end{align*}
+a similar expression also holds for an odd node number \(n>2\).
+Consequently, for all fixed parameters excluding
+\(\theta_2=\theta_3=0\),
+\(\lim_{N\to \infty}\REP_N(\boldsymbol \theta_N)/N=\infty\) then follows
+and the graph model with 2-stars and triangles is S-unstable, as
+suggested by the breach of Proposition \ref{prp:prop1} for this model
+when \(|\theta_2|+|\theta_3|\neq 0\). That is, instability holds even
+under \(\theta_2 = - \theta_3/3\) case potentially allowed by
+Schweinberger's results.
+
+\subsection{Latent variable models of increasing parameter
+dimension}\label{latent-variable-models-of-increasing-parameter-dimension}
+
+We next consider instability of discrete data models based on
+exponential formulations involving hidden, or latent, variables, such as
+those described in Sections
+\ref{restricted-boltzmann-machines}-\ref{deep-learning} for networks. We
+will focus on restricted Bolzmann machine (RBM) models (Section
+\ref{restricted-boltzmann-machines}, having one layer of latent
+variables for simplicity, though the same instability concepts may be
+extended to other deep learning models (Section \ref{deep-learning}. For
+\(N\) visible variables
+\(\boldsymbol X \equiv \boldsymbol X_N = (X_1,\ldots,X_n)\) as data,
+each observation \(X_i\in\{\pm 1\}\) being binary, the RBM-based model
+\eqref{eq:RBM2} for \(\boldsymbol X\) is again of FOES-type, though not an
+exponential model. However, the distribution of visible variables is
+induced by an underlying joint exponential model \eqref{eq:RBM1} for both
+visible and latent variables \((\boldsymbol X, \boldsymbol H)\), where
+\(\boldsymbol H=(H_1,\ldots,H_{N_{\mathcal{H}}})\) denotes a vector of
+\(N_{\mathcal{H}}\) hidden variables (similarly binary). The joint model
+is of linear exponential form involving
+\(d(N)\equiv N + N_{\mathcal{H}} + N* N_{\mathcal{H}}\) sufficient
+statistics given by
+\((\boldsymbol X, \boldsymbol H, \boldsymbol X^T\boldsymbol H)\) and
+parameters
+\(\boldsymbol \theta_N = (\boldsymbol \theta_N^{\mathcal{V}},\boldsymbol \theta_N^{\mathcal{H}}, \boldsymbol \theta_N^{\mathcal{VH}} ) \in\mathbb{R}^{d(N)}\)
+corresponding to the \(N\) visible variables \(\boldsymbol X\) (i.e.,
+\(\boldsymbol \theta_N^{\mathcal{V}}\in\mathbb{R}^N\)), the
+\(N_{\mathcal{H}}\) hidden variables \(\boldsymbol H\) (i.e.,
+\(\boldsymbol \theta_N^{\mathcal{H}}\in\mathbb{R}^{N_{\mathcal{H}}}\)),
+and the \(N *N_{\mathcal{H}}\) cross-product variables
+\(\boldsymbol X^T\boldsymbol H\) (i.e.,
+\(\boldsymbol \theta_N^{\mathcal{VH}}\in\mathbb{R}^{N N_{\mathcal{H}}}\)).
+However, unlike some previous exponential models considered in Sections
+\ref{one-param-exp}-\ref{fixed-dim-exp} (cf.~Proposition
+\ref{prp:prop1}, note that the RBM formulation always associates
+parameters with \emph{bounded} statistics (i.e., the components of
+\((\boldsymbol X, \boldsymbol H, \boldsymbol X^T\boldsymbol H)\)) so
+that model instability cannot arise here due to the magnitude of
+sufficient statistics exceeding the sample size \(N\). Instead, RBM
+instability may be linked to solely to parameter configuration and the
+fact that the number \(d(N) \geq N\) of parameters necessarily increases
+with the number \(N\) of observations \(\boldsymbol X\), in contrast to
+previous exponential cases of fixed parameter dimension.
+
+To highlight the instability issues for the RBM model, consider a simple
+model for \(N\) visibles \(\boldsymbol X\) with no hidden variables
+(\(N_{\mathcal{H}}=0\)), for which model statements
+\eqref{eq:RBM1}-\eqref{eq:RBM2} coincide. An independence model then results
+for variables in \(\boldsymbol X\), which has \(d(N)=N\) parameters
+\(\boldsymbol \theta_N^{\mathcal{V}} = (\theta_{1,N}^{\mathcal{V}}, \ldots, \theta_{N,N}^{\mathcal{V}}) \in \mathbb{R}^{N}\),
+and the measure of model instability becomes \[
+\frac{1}{N}\REP(\boldsymbol \theta_N) = \frac{2}{N} \sum_{i=1}^N|\theta_{i,N}^{\mathcal{V}}| \quad \mbox{(RBM model, no hiddens)}.
+\] Hence, this model sequence for \(\boldsymbol X\) will be S-unstable
+model if the aggregation of absolute parameters grows faster than the
+number \(N\) of parameters/visible variables. Consequently, even for a
+simplest RBM model involving independence, preventing instability
+requires careful choice of parameters, particularly with regard to how a
+parameter configuration differs from zero. For more general RBM models,
+the number \(N_{\mathcal{H}}\) of hidden variables \(\boldsymbol H\) can
+also be chosen arbitrarily (i.e., as some function
+\(N_{\mathcal{H}}\equiv N_{N,\mathcal{H}}\) of \(N\)), which can
+substantially inflate the number \(d(N)\) of model parameters and
+further impact model instability through accumulated parameters. To
+better understand the effects of instability in the RBM structure,
+Proposition \ref{prp:prop2} next frames the general behavior of extreme
+probabilities in the joint RBM model \eqref{eq:RBM1} for
+\((\boldsymbol X, \boldsymbol H)\) and the implied RBM data model
+\eqref{eq:RBM2} for \(\boldsymbol X\) alone. Specifically, critical
+measures of instability may be closely connected in both models through
+tight bounds on their respective LREP values \eqref{eq:elpr}. As a result,
+Proposition \ref{prp:prop2} shows how an unstable distribution for
+observations \(\boldsymbol X\) may be traced to sources of instability
+in the original joint distribution for
+\((\boldsymbol X,\boldsymbol H)\). This also suggests a device for
+avoiding instability, as provided next.
+
+To state the result, let
+\(\REP_{\boldsymbol X}(\boldsymbol \theta_N) \equiv \REP(\boldsymbol \theta_N)\)
+denote the LREP value \eqref{eq:elpr} from the marginal distribution
+\(P_{\boldsymbol \theta_N}\) of visibles \(\boldsymbol X\) in
+\eqref{eq:RBM2} and write the LREP for the joint distribution
+\(\tilde{P}_{\boldsymbol \theta_N}\) of
+\((\boldsymbol X, \boldsymbol H)\) from \eqref{eq:RBM1} as
+\begin{align*}
+\REP_{(\boldsymbol X, \boldsymbol H)}(\boldsymbol \theta_N) &=\log\left[  \frac{\max_{ (\boldsymbol x, \boldsymbol h) \in \{\pm 1\}^{N+N_{\mathcal{H}}}}
+\tilde{P}_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)}{\min_{ (\boldsymbol x, \boldsymbol h) \in \{\pm 1\}^{N+N_{\mathcal{H}}}`}
+\tilde{P}_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)} \right] \qquad \mbox{(joint RBM model)}    \\    
+&=   \left(\max_{ (\boldsymbol x, \boldsymbol h) \in \{\pm 1\}^{N+N_{\mathcal{H}}}} f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)-\min_{ (\boldsymbol x, \boldsymbol h) \in \{\pm 1\}^{N+N_{\mathcal{H}}}} f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)\right),
+\end{align*}
+written as a function
+\begin{equation}
+\label{eq:f}
+f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)  \equiv  \sum_{i=1}^N x_i  \theta_{i,N}^{\mathcal{V}} +  \sum_{j=1}^{N_\mathcal{H}} h_j \theta_{j,N}^{\mathcal{H}} + \sum_{i=1}^N \sum_{j=1}^{N_\mathcal{H}} x_i h_j  \theta_{ij,N}^{\mathcal{VH}}
+\end{equation}
+of outcomes \(\boldsymbol x =(x_1,\ldots,x_N) \in\{\pm1\}^{N}\),
+\(\boldsymbol h =(h_1,\ldots,h_{N_{\mathcal{H}}}) \in\{\pm1\}^{N_{\mathcal{H}}}\)
+and parameters
+\(\boldsymbol \theta_N \equiv (\boldsymbol \theta_N^{\mathcal{V}},\boldsymbol \theta_N^{\mathcal{H}}, \boldsymbol \theta_N^{\mathcal{VH}})\),
+with \(\theta_{i,N}^{\mathcal{V}}\), \(\theta_{j,N}^{\mathcal{H}}\) and
+\(\theta_{ij,N}^{\mathcal{VH}}\) denoting respective parameter
+components, \(1 \leq i \leq N\), \(1 \leq j \leq N_{\mathcal{H}}\). Due
+to the marginalization steps in defining the distribution \eqref{eq:RBM2}
+of \(\boldsymbol X\), note that
+\(\REP_{\boldsymbol X}(\boldsymbol \theta_N)\) has no immediate
+analytical expression similar to that of
+\(\REP_{(\boldsymbol X, \boldsymbol H)}(\boldsymbol \theta_N)\). For
+clarity, recall also that S-instability \eqref{eq:Sun} in each model type
+refers to a respective divergence (i.e.,
+\(\lim_{N\to \infty} \REP_{(\boldsymbol X, \boldsymbol H)}(\boldsymbol \theta_N) /(N+N_{\mathcal{H}})=\infty\),
+\(\lim_{N\to \infty} \REP(\boldsymbol \theta_N) /N=\infty\)) upon
+scaling by the corresponding number of variables in a distribution. In
+the following, let \(|\boldsymbol y|_1 = \sum_{i=1}^d |y_i|\) denote the
+L-1 norm of a generic vector \(\boldsymbol y =(y_1,\ldots,y_d)\),
+\(d \geq 1\). \BeginKnitrBlock{proposition}
+
+\protect\hypertarget{prp:prop2}{}{\label{prp:prop2}}Let
+\(P_{\boldsymbol \theta_N}\) denote a RBM-based model \eqref{eq:RBM2} for
+\(N\geq 1\) visible variables \(\boldsymbol X \equiv \boldsymbol X_N\)
+derived from \(\tilde{P}_{\boldsymbol \theta_N}\) as the joint RBM
+distribution \eqref{eq:RBM1} of \((\boldsymbol X, \boldsymbol H)\)
+involving some number \(N_{\mathcal{H}} \equiv N_{N,\mathcal{H}}\geq 0\)
+of hidden variables \(\boldsymbol H \equiv \boldsymbol H_N\) and
+parameters
+\(\boldsymbol \theta_N \equiv (\boldsymbol \theta_N^{\mathcal{V}},\boldsymbol \theta_N^{\mathcal{H}}, \boldsymbol \theta_N^{\mathcal{VH}}) \in\mathbb{R}^{N}\times \mathbb{R}^{N_{\mathcal{H}}} \times \mathbb{R}^{N N_{\mathcal{H}}}\).
+Then,
+\begin{enumerate}[(i)]
+\item the instability measure $\REP_{(\boldsymbol X)}(\boldsymbol \theta_N)$ for the marginal model $P_{\boldsymbol \theta_N}$ of  $\boldsymbol X$ satisfies
+    $$
+    \left| \REP_{(\boldsymbol X)}(\boldsymbol \theta_N)  - \elt\right| \leq    N_{\mathcal{H}}  \log 2
+    $$
+    for
+    $$
+    \elt  \equiv   \max_{ \boldsymbol x} \max_{ \boldsymbol h  } f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)-\min_{ \boldsymbol x } \max_{ \boldsymbol h }f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)
+    $$
+    based on $f_{\boldsymbol \theta_N}$ from \\eqref{eq:f} with components $\boldsymbol x \in \{\pm 1\}^{N}, \boldsymbol h \in \{\pm 1\}^{N_{\mathcal{H}}}$.
+\item The instability measure $\REP_{(\boldsymbol X, \boldsymbol H)}(\boldsymbol \theta_N)\equiv \left(\max_{ \boldsymbol x} \max_{ \boldsymbol h  } f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)-\min_{ \boldsymbol x } \min_{ \boldsymbol h }f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)\right)$ for the joint model $\tilde{P}_{\boldsymbol \theta_N}$ of $(\boldsymbol X, \boldsymbol H)$ satisfies
+    \begin{align*}
+    2\Gam +  2|\boldsymbol \theta_N^{\mathcal{H}} |_{1} \; \geq \; \REP_{(\boldsymbol X, \boldsymbol H)}(\boldsymbol \theta_N)    & \geq 
+    2\max\big\{\Gam,   |\boldsymbol \theta_N^{\mathcal{H}} |_{1}\big\} \\
+    &\geq  2\Gam\\
+    &\geq  \elt\\
+    &\geq   \max\big\{  \Gamc, \, \Gam  - 2|\boldsymbol \theta_N^{\mathcal{H}} |_{1}  \big\}
+    \end{align*}
+    for
+    $$
+    \Gam \equiv \max_{ \boldsymbol h} k_{\boldsymbol \theta_N} (\boldsymbol h)
+    \geq  |\boldsymbol \theta_N^{\mathcal{V}} |_{1},\qquad  k_{\boldsymbol \theta_N} (\boldsymbol h)  \equiv \sum_{i=1}^{N }\left| \theta_{i,N}^{\mathcal{V}}   + \sum_{j=1}^{N_{\mathcal{H}}} h_j \theta_{ij,N}^{\mathcal{VH}} \right|,
+    $$ 
+    and $\Gamc \equiv \min_{ \boldsymbol h} k_{\boldsymbol \theta_N} (\boldsymbol h)$ based on a function  $k_{\boldsymbol \theta_N} (\boldsymbol h)$ of hidden variable outcomes $\boldsymbol h = (h_1,\ldots,h_{N_{\mathcal{H}}})$ and visible-related parameters $\boldsymbol \theta_N^{\mathcal{V}}$ and $\boldsymbol \theta_N^{\mathcal{VH}}$.
+\item Assuming $\sup_{N \geq 1} N_{\mathcal{H}}/N<\infty$ additionally, then the following properties 1.-7. hold:
+\begin{enumerate}[1.] \itemsep 0cm
+    \item an S-unstable visible model $P_{\boldsymbol \theta_N}$ is equivalent to the condition $\lim_{N\to \infty}  \elt/N  =\infty$; further, $P_{\boldsymbol \theta_N}$ is stable when $\elt/N$, $N \geq 1$, is bounded.
+    \item an S-unstable joint model $P_{\boldsymbol \theta_N}$ is equivalent to the condition $\lim_{N\to \infty} \max\{|\boldsymbol \theta_N^{\mathcal{H}}|_1, \Gam\}/N =\infty$; further,  $\tilde{P}_{\boldsymbol \theta_N}$ is stable when $[|\boldsymbol \theta_N^{\mathcal{H}}|_1+ \Gam]/N$, $N \geq 1$, is bounded.
+    \item if the visible model $P_{\boldsymbol \theta_N}$   is S-unstable, then  the joint model $\tilde{P}_{\boldsymbol \theta_N}$ is also S-unstable.
+    \item when $\lim_{N\to \infty}  (|\boldsymbol \theta_N^{\mathcal{V}}|_1-2|\boldsymbol \theta_N^{\mathcal{H}}|_1)/N =\infty$, both $P_{\boldsymbol \theta_N}$ and $\tilde{P}_{\boldsymbol \theta_N}$ are necessarily S-unstable.
+    \item when $\lim_{N\to \infty}|\boldsymbol \theta_N^{\mathcal{H}}|_1/N =\infty$,  the joint model $\tilde{P}_{\boldsymbol \theta_N}$ is necessarily S-unstable.
+    \item when $\sup_{N \geq 1} |\boldsymbol \theta_N^{\mathcal{H}}|_1 /N<\infty$, the visible model $P_{\boldsymbol \theta_N}$ being S-stable or S-unstable is equivalent to the joint model  $\tilde{P}_{\boldsymbol \theta_N}$ being stable or unstable.
+    \item  an S-stable visible model $P_{\boldsymbol \theta_N}$ results if
+        $$
+        |\boldsymbol \theta_N^{\mathcal{V}}|_1+ |\boldsymbol \theta_N^{\mathcal{VH}} |_1  \leq CN,\quad N \geq 1,
+        $$
+        for some $C>0$, while an S-stable joint model $\tilde{P}_{\boldsymbol \theta_N}$  results if
+        $$
+        |\boldsymbol \theta_N|_1 \equiv  |\boldsymbol \theta_N^{\mathcal{V}}|_1+|\boldsymbol \theta_N^{\mathcal{H}}|_1 +|\boldsymbol \theta_N^{\mathcal{VH}} |_1  \leq  C N ,\quad N \geq 1.
+        $$
+\end{enumerate}
+\end{enumerate}
 \EndKnitrBlock{proposition}
 
-In other words, in S-unstable FSFS models, all probability in the model
-formulation with a large number of random variables will concentrate
-mass on an \(\epsilon\)-modal set, where \(\epsilon\) can be made
-arbitrarily small. The associated mode set could potentially be quite
-small, in which case Proposition \ref{prp:degenFSFS} would suggest that
-the unstable model asymptotically stacks all probability on a few
-outcomes.
-
 \BeginKnitrBlock{remark}
-\iffalse <span class="remark"><em>Remark. \fi{} There is a further
-generalization the notion of instability in Definition
-\ref{def:instabFSFS} meant to address independent replications of data
-sequences. That is, one might consider data as \(n\) independent and
-identically distributed replications
-\(\boldsymbol X_1, \dots, \boldsymbol X_n\), where each
-\(\boldsymbol X_i=(X_{i,1},\ldots,X_{i,N}) \in \mathcal{X}^N\) follows a
-common FSFS model with probabilities \(P_{\theta_N}(\boldsymbol x)>0\),
-\(\boldsymbol x\in\mathcal{X}^N\) and \(|\mathcal{X}|<\infty\), for
-\(i=1,\ldots,n\). This leads to a total of \(n*N\) random variables in
-the joint model. However, the definition of S-unstable and Propositions
-\ref{prp:instab-elpr}-\ref{prp:degenFSFS} still hold for such iid
-replications. This is because
-\(\left(\max\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)\right)^n\)
-is the largest probability possible under the joint model for the \(n\)
-replications while
-\(\left(\min\limits_{\boldsymbol x \in \mathcal{X}}P_{\boldsymbol \theta_N}(\boldsymbol x)\right)^n\)
-is the smallest probability. Thus, for the combined replications
-\(\boldsymbol X_1, \ldots, \boldsymbol X_n\), the analog definition of
-the extremal log-probability becomes
-\begin{align*}
-\frac{\text{extremal log-probability ratio}}{\text{\# random variables in the model}} &\equiv \frac{1}{n*N}\log \left[\frac{\left(\max\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)\right)^n}{\left(\min\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)\right)^n}\right] = \frac{1}{N} \log \left[\frac{\max\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)}{\min\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)}\right] \\
-&= \frac{1}{N} \text{ELPR}(\boldsymbol \theta_N),
-\end{align*}
-implying that the definition of an S-unstable FSFS model sequence is
-invariant to the level (\(n\)) of independent replication. Consequently,
-overall model instabilities may be characterized by those of one
-observation from the common FSFS model.
+\iffalse <span class="remark"><em>Remark. \fi{} The condition
+\(\sup_{N \geq 1} N_{\mathcal{H}}/N<\infty\) in Proposition
+\ref{prp:prop2}(iii) is often mild in practice (i.e., the number
+\(N_{\mathcal{H}}\) of hidden variables is typically not excessively
+larger than the number \(N\) of visible observations). This allows
+instability results for both marginal and joint RBM models to be more
+readily stated together, as the numbers \(N\) and \(N+N_{\mathcal{H}}\)
+of variables in these models become asymptotically equivalent.
 \EndKnitrBlock{remark}
 
-\section{Implications}\label{implications}
+In Proposition \ref{prp:prop2}(iii), the relationships between RBM
+models with regard to instability, and the effects of different
+parameter types, follow from the bounds on model instability measures in
+Proposition \ref{prp:prop2}(i)-(ii). Generally speaking, all instability
+in the marginal RBM model for the data \(\boldsymbol X\) can be
+attributed to an excessively large model quantity \(\elt\), which
+predominantly follows when main \(\boldsymbol \theta_N^{\mathcal{V}}\)
+and interaction \(\boldsymbol \theta_N^{\mathcal{VH}}\) parameters
+related to visible variables are too large in magnitude (e.g., upon
+accumulation in terms such as
+\(|\boldsymbol \theta_N^{\mathcal{V}}|_1\), \(\Gam\) or \(\Gamc\)). For
+example, for any bounded sequence
+\(|\boldsymbol \theta_N^{\mathcal{H}}|/N\) of hidden parameters, if main
+visible parameters \(\boldsymbol \theta_N^{\mathcal{V}}\) are too
+extreme (\(|\boldsymbol \theta_N^{\mathcal{V}}|_1/N\to \infty\)), this
+aspect will guarantee instability in the visible model
+(\(\elt/N\to \infty\)). In fact, the instability measure
+\(\elt \equiv \max_{ \boldsymbol x} \max_{ \boldsymbol h } f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)-\min_{ \boldsymbol x } \max_{ \boldsymbol h }f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)\)
+for marginal/visible model represents a clearly smaller portion of the
+instability measure
+\(\REP_{(\boldsymbol X, \boldsymbol H)}(\boldsymbol \theta_N)\equiv \max_{ \boldsymbol x} \max_{ \boldsymbol h } f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)-\min_{ \boldsymbol x } \min_{ \boldsymbol h } f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)\)
+in the joint RBM model, implying that an unstable marginal model (i.e.,
+due to \(\boldsymbol \theta_N^{\mathcal{V}}\),
+\(\boldsymbol \theta_N^{\mathcal{VH}}\)) must always translate to an
+unstable joint model and that further potential causes of instability
+exist for the joint model, often due to the size
+\(|\boldsymbol \theta_N^{\mathcal{H}}|_1\). For example, while the joint
+RBM model for \((\boldsymbol X,\boldsymbol H)\) must always be unstable
+due to a diverging combination of visible and/or interaction parameters
+(\(|\boldsymbol \theta_N^{\mathcal{V}}|_1/N\to \infty\) or
+\(\Gam/N\to \infty\)) (Proposition \ref{prp:prop2}(iii.2)), instability
+for the joint model can also result when the main hidden parameters
+\(\boldsymbol \theta_N^{\mathcal{H}}\) become too large relative to
+sample size (\(|\boldsymbol \theta_N^{\mathcal{H}}|_1/N\to \infty\) in
+Proposition \ref{prp:prop2}(iii.5)). However, under Proposition
+\ref{prp:prop2}, the main hidden parameters
+\(\boldsymbol \theta_N^{\mathcal{H}}\) do not necessarily entail a
+source of instability for the marginal visible model. To explain this
+distinction, consider a joint model where all parameters related to
+visibles are zero,
+\(\boldsymbol \theta_N^{\mathcal{V}}= \boldsymbol \theta_N^{\mathcal{VH}}=\boldsymbol 0\),
+but the hidden-related parameters diverge in sum
+\(|\boldsymbol \theta_N^{\mathcal{H}}|_1/N\to \infty\). Here, the
+explosive behavior among parameters
+\(\boldsymbol \theta_N^{\mathcal{H}}\) induces instability in the joint
+model for \((\boldsymbol X, \boldsymbol H)\) but the marginal model for
+\(\boldsymbol X\), however, has a perfectly stable (and in fact uniform)
+distribution in this case. When the hidden parameters are bounded
+relative to the sample size
+(\(\sup_{N\geq 1} |\boldsymbol \theta_N^{\mathcal{H}}|_1/N<\infty\)),
+then all instability in both the joint and marginal RBM models can be
+directly linked to excessively large visible
+\(\boldsymbol \theta_N^{\mathcal{V}}\) and/or interaction parameters
+\(\boldsymbol \theta_N^{\mathcal{VH}}\) so that features of
+stability/instability must be the same across both models (Proposition
+\ref{prp:prop2}(iii.6)). Hence, to prevent instability in the joint
+model, the combined magnitudes of all parameters \(\theta_N\) must be
+controlled (cf.~Proposition \ref{prp:prop2}(iii.7)), while a stable
+visible data model technically results in constraining only the sizes of
+visible-related parameters \(\boldsymbol \theta_N^{\mathcal{V}}\),
+\(\boldsymbol \theta_N^{\mathcal{VH}}\). Nevertheless, because the joint
+model is often employed in practice for purposes of simulation and
+simulation-based inference, it is still reasonable to consider parameter
+choices for ensuring a stable joint model (and, consequently, a stable
+visible model as well).
+
+In more complicated network models involving further or deeper hidden
+layers, the same issues and causes of instability similarly exist, but
+are compounded by a greater number of model parameters. S-unstable joint
+models will similarly follow if the combined sizes of all parameters are
+too great relative to the total number of variables, while instability
+in the data model for visible variables will depend only on the main or
+interaction parameters directly related to visibles and how their
+accumulated magnitude compares to the observation sample size \(N\).
+
+\section{Statistical consequences of instability}\label{implications}
+
+Due to their extreme sensitivity in probability structure, S-instability
+in FOES models often translates to numerical complications, and in fact
+obstructions, in both simulation and statistical inference based on
+likelihoods.
+
+\subsection{Implications for
+simulation}\label{implications-for-simulation}
+
+Suppose one aims to apply MCMC to simulate data
+\(\boldsymbol X=(X_1,\ldots,X_n)\) from a FOES model
+\(P_{\boldsymbol \theta_N}(\boldsymbol x)\),
+\(\boldsymbol x \in\mathcal{X}^N\) whereby a chain is constructed with
+\(P_{\boldsymbol \theta_N}\) as the stationary distribution. For an
+unstable FOES model, one component changes in outcomes may produce
+radically different probabilities which then creates numerical barriers
+to MCMC. For example, consider implementation of a Gibbs sampler from
+the full conditional distributions
+\(P_{\boldsymbol \theta_N} (X_i = x| \boldsymbol x_{-i})\),
+\(x\in\mathcal{X}\), of each variable \(X_i\) based on values
+\(\boldsymbol x_{-i}\in \mathcal{X}^{N-1}\) for the remaining variables,
+say \(\boldsymbol X_{-i}\), in \(\boldsymbol X\). If a single change in
+\(X_i\) from one value \(x_1\) to another \(x_2\) may produce two
+outcomes \(\boldsymbol x^{(1)}\) and \(\boldsymbol x^{(2)}\) for
+\(\boldsymbol X\) with vastly different probabilities under the joint
+distribution \(P_{\boldsymbol \theta_N}\), then the Gibbs sampler can
+have extreme log-ratios in its transition probabilities \[
+\left| \frac{P_{\boldsymbol \theta_N} (X_i =x_1 | \boldsymbol x^{(1)}_{-i})}{P_{\boldsymbol \theta_N} (X_i = x_2| \boldsymbol x^{(2)}_{-i}) } \right|=\left| \frac{P_{\boldsymbol \theta_N} ( \boldsymbol x^{(1)} }{P_{\boldsymbol \theta_N} (  \boldsymbol x^{(2)} ) } \right|
+\] using that conditional probabilities are proportional to the joint
+probabilities and the fact that unstable models can have unbounded
+log-probability ratios in one-component changes (Theorem
+\ref{thm:instab-elpr}. This can hinder the ability of chain to
+effectively explore the sample space of the observations
+\(\boldsymbol X\), as the chain may then mix poorly by moving rapidly
+to, and slowly away from, sections of the sample space surface. Because
+the numerical problem is unbounded ratios of probabilities under the
+joint model, similar complications can also arise for
+Metropolis-Hastings algorithms for MCMC. Hence, while an unstable FOES
+model \(P_{\boldsymbol \theta_N}\) may be valid and technically open to
+simulation by MCMC, the aspect of instability can ruin numerical
+applications of MCMC.
+
+\subsection{Implications for maximum likelihood
+inference}\label{implications-for-maximum-likelihood-inference}
+
+The rugged or volatile probability structure in unstable models can also
+hamper efforts to maximize likelihood functions for statistical
+inference. Essentially, when a FOES model is unstable along a parameter
+sequence \(\boldsymbol \theta_N\), the same model can consequently
+unstable along parameters \(-\boldsymbol \theta_N\) opposite from the
+origin (model condition PSR and Corollary \ref{cor:sign}). This leads to
+potential hyper-sensitivity in the likelihood function, which can be
+pulled in extreme and opposite directions depending on how a parameter
+lies with respect to zero and translates into numerical complications in
+maximizing the objective function.
+
+For a FOES model \(P_{\boldsymbol \theta_N}(\boldsymbol x)\),
+\(\boldsymbol x\in\mathcal{X}^N\), for \(N\) observations
+\(\boldsymbol X =(X_1,\ldots,X_n)\), define the likelihood function
+\(\mathcal{L}(\boldsymbol \theta_N | \boldsymbol x) \equiv N^{-1}\log P_{\boldsymbol \theta_N }(\boldsymbol x)\)
+at \(\boldsymbol \theta_N\) and the minimum probability ratio
+counterpart
+\(\mathcal{L}(\boldsymbol \theta_N | \boldsymbol x) - N^{-1} \log \min_{\boldsymbol \theta_N}= N^{-1}\log[P_{\boldsymbol \theta_N }(\boldsymbol x) /\min_{\boldsymbol \theta_N}]\)
+where
+\(\min_{\boldsymbol \theta_N} \equiv \min_{\boldsymbol y\in\mathcal{X}^N} P_{\boldsymbol \theta_N }(\boldsymbol y)]\).
+Also define
+\(\max_{\boldsymbol \theta_N} \equiv \max_{\boldsymbol y\in\mathcal{X}^N} P_{\boldsymbol \theta_N }(\boldsymbol y)]\).
+
+\BeginKnitrBlock{theorem}
+\protect\hypertarget{thm:something}{}{\label{thm:something}}Let
+\(P_{\boldsymbol \theta_N}\), \(N \geq 1\), denote an S-unstable FOES
+model sequence (i.e.,
+\(\lim_{N\to \infty}\REP(\boldsymbol \theta_N)/N=\infty\)) which
+additionally satisfies model condition SPR. Then,
+\begin{align*}
+\log\left[ \frac{P_{\boldsymbol \theta_N }(\boldsymbol x) }{\min_{\boldsymbol \theta_N}}\right] +  \log\left[ \frac{P_{-\boldsymbol \theta_N }(\boldsymbol x) }{\min_{-\boldsymbol \theta_N}}\right]  &=    \log\left[ \frac{\max_{\boldsymbol \theta_N}}{\min_{\boldsymbol \theta_N}}\right]  \equiv  \REP(\boldsymbol \theta_N)\\
+& = \log\left[ \frac{\max_{-\boldsymbol \theta_N} }{\min_{-\boldsymbol \theta_N}}\right] \equiv \REP(-\boldsymbol \theta_N)
+\end{align*}
+for any \(\boldsymbol x\in\mathcal{X}^N\) and, as \(N\to \infty\), the
+following convergence \[
+\frac{1}{\REP(\boldsymbol \theta_N)} \log\left[ \frac{P_{\boldsymbol \theta_N }(\boldsymbol X) }{\min_{\boldsymbol \theta_N}}\right]  \rightarrow 1 \qquad \frac{1}{\REP(-\boldsymbol \theta_N)}\log\left[ \frac{P_{-\boldsymbol \theta_N }(\boldsymbol X) }{\min_{-\boldsymbol \theta_N}}\right]  \rightarrow 0
+\] holds in probability \(P_{\boldsymbol \theta_N}\), as well as in
+expectation \(E_{\boldsymbol \theta_N}\), under
+\(\boldsymbol \theta_N\); the limits of convergence are reversed under
+\(-\boldsymbol \theta_N\), \[
+\frac{1}{\REP(\boldsymbol \theta_N)} \log\left[ \frac{P_{\boldsymbol \theta_N }(\boldsymbol x) }{\min_{\boldsymbol \theta_N}}\right]  \rightarrow 0 \qquad \frac{1}{\REP(-\boldsymbol \theta_N)}\log\left[ \frac{P_{-\boldsymbol \theta_N }(\boldsymbol x) }{\min_{-\boldsymbol \theta_N}}\right]  \rightarrow 1
+\]
+\EndKnitrBlock{theorem} This needs to be connected to results in
+Schweinberger who considered the one-parameter linear exponential case
+showing that, for positive \(\boldsymbol \theta_N>0\), \[
+\frac{1}{\REP(\boldsymbol \theta_N)} \log\left[ \frac{P_{\boldsymbol \theta_N }(\boldsymbol x) }{\min_{\boldsymbol \theta_N}}\right] = \frac{\boldsymbol g(\boldsymbol x) - L_N}{U_N-L_N} 
+\] holds and converges to \(1\) in expectation
+\(E_{\boldsymbol \theta_N}\), while \[   
+\frac{1}{\REP(-\boldsymbol \theta_N)} \log\left[ \frac{P_{-\boldsymbol \theta_N }(\boldsymbol x) }{\min_{-\boldsymbol \theta_N}}\right] = \frac{U_N -\boldsymbol g(\boldsymbol x)}{U_N-L_N} 
+\] holds converges to \(1\) in expectation
+\(E_{-\boldsymbol \theta_N}\). This result is a special case of Theorem
+\ref{thm:something} and the LREP expansion \eqref{eq:UL} for the one
+parameter exponential model case.
+
+The idea, more generally, is that probability in an unstable model is
+pushed to opposite regions of the sample space due to degeneracy and
+opposite signed parameters. This can cause numerical complications in
+likelihood maximization. The likelihood function knows what to do with
+an observed outcome
+\(\boldsymbol x \equiv \argmax \mathcal{L}(\boldsymbol \theta_N)\) which
+is also the outcome with minimal probability under
+\(P_{-\boldsymbol \theta_N}\). But unless the observed outcome
+\(\boldsymbol x\) is at an extreme in the probability structures of
+\(P_{\boldsymbol \theta_N}, P_{-\boldsymbol \theta_N}\) (which are
+related) the likelihood function will be flat and optimization steps may
+shift around zero.
+
+\subsection{Implications for Bayes
+inference}\label{implications-for-bayes-inference}
+
+Considering that changes in sign of parameters can cause all probability
+to be placed on disjoint sections of the sample space (Corollary
+\ref{cor:sign}), then this will create numerical difficulties in Bayes
+inference based on MCMC. We can write the MH transition density to
+indicate this.
+
+\section{Concluding Remarks}\label{conclusions}
 
@@
 instability in model probability structure and elucidated multiple
-consequences of instability. We have shown for FSFS models that
+consequences of instability. We have shown for FOES models that
 instability manifests through small changes in data leading to
@@
 place all probability on certain modal subsections of the sample space,
-which potentially could be small. The FSFS model class is quite broad
-and, particularly in developing FSFS models for large data sets, some
+which potentially could be small. The FOES model class is quite broad
+and, particularly in developing FOES models for large data sets, some
 caution should be used in parameter specification to control effects of

### 2017-09-30 writing/paper-note/chapters/instab.Rmd
remarks don't link

@@
 For illustration, consider the iid Bernoulli model \@ref(eq:mod1) for $\boldsymbol X_N=(X_1,\ldots,X_N)$ with log-odds ratio parameter $\boldsymbol \theta_N = \log[ P_{\boldsymbol \theta_N}(X_1=1)/ P_{\boldsymbol \theta_N}(X_1=0)]\in\mathbb{R}$.
-Remark \@ref(rmk1) (Section \@ref(criterion)) then gives the model instability measure \@ref(eq:Sun) directly as
+Remark 1 (Section \@ref(criterion)) then gives the model instability measure \@ref(eq:Sun) directly as
 $$
@@
 To further illustrate the result in Proposition \@ref(prp:prop1), consider the multinomial distribution \@ref(eq:mod11) for $\boldsymbol X_N=(X_1,\ldots,X_N)$ having $k\geq 2$ categories $\{1,\ldots,k\}$ and $k$ parameters $\boldsymbol \theta_N = (\theta_{1,N},\ldots,\theta_{k,N})^\prime$.
-The variables are iid under this model so that Remark \@ref(rmk1) (Section \@ref(criterion) yields the corresponding $N$-scaled  log-ratio of extreme probabilities \@ref(eq:elpr) as
+The variables are iid under this model so that Remark 1 (Section \@ref(criterion) yields the corresponding $N$-scaled  log-ratio of extreme probabilities \@ref(eq:elpr) as
 \begin{align*}

### 2017-09-30 writing/paper-note/docs/note.tex
remarks don't link

@@
 \(\boldsymbol \theta_N = \log[ P_{\boldsymbol \theta_N}(X_1=1)/ P_{\boldsymbol \theta_N}(X_1=0)]\in\mathbb{R}\).
-Remark \ref{rmk1} (Section \ref{criterion}) then gives the model
-instability measure \eqref{eq:Sun} directly as \[
+Remark 1 (Section \ref{criterion}) then gives the model instability
+measure \eqref{eq:Sun} directly as \[
 \frac{1}{N}\REP(\boldsymbol \theta_N ) = |\boldsymbol \theta_N|\quad\; \mbox{(iid Bernoulli model)},
@@
 \(\boldsymbol \theta_N = (\theta_{1,N},\ldots,\theta_{k,N})^\prime\).
-The variables are iid under this model so that Remark \ref{rmk1}
-(Section \ref{criterion} yields the corresponding \(N\)-scaled log-ratio
-of extreme probabilities \eqref{eq:elpr} as
+The variables are iid under this model so that Remark 1 (Section
+\ref{criterion} yields the corresponding \(N\)-scaled log-ratio of
+extreme probabilities \eqref{eq:elpr} as
 \begin{align*}

### 2017-09-30 writing/paper-note/docs/note.tex
backto tech

@@
-\documentclass[12pt]{article}
-\usepackage{amsmath,amssymb}
-\usepackage{graphicx,psfrag,epsf}
-\usepackage{enumerate}
-\usepackage{natbib}
-\usepackage{url} % not crucial - just used below for the URL
-
-\usepackage{longtable,booktabs}
-\usepackage[unicode=true]{hyperref}
+\documentclass[]{article}
+%\usepackage{nips15submit_e,times}
+\usepackage{url}
+
 \providecommand{\tightlist}{%
   \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
+
+\usepackage{lmodern}
+\usepackage{amssymb,amsmath}
+\usepackage{ifxetex,ifluatex}
+\usepackage{fixltx2e} % provides \textsubscript
+\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
+  \usepackage[T1]{fontenc}
+  \usepackage[utf8]{inputenc}
+\else % if luatex or xelatex
+  \ifxetex
+    \usepackage{mathspec}
+    \usepackage{xltxtra,xunicode}
+  \else
+    \usepackage{fontspec}
+  \fi
+  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
+  \newcommand{\euro}{€}
+\fi
+% use upquote if available, for straight quotes in verbatim environments
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
+% use microtype if available
+\IfFileExists{microtype.sty}{%
+\usepackage{microtype}
+\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
+}{}
+\usepackage{longtable,booktabs}
+\usepackage{graphicx}
+\makeatletter
+\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
+\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
+\makeatother
+% Scale images if necessary, so that they will not overflow the page
+% margins by default, and it is still possible to overwrite the defaults
+% using explicit options in \includegraphics[width, height, ...]{}
+\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
+\ifxetex
+  \usepackage[setpagesize=false, % page size defined by xetex
+              unicode=false, % unicode breaks when used with xetex
+              xetex]{hyperref}
+\else
+  \usepackage[unicode=true]{hyperref}
+\fi
+\hypersetup{breaklinks=true,
+            bookmarks=true,
+            pdfauthor={},
+            pdftitle={A note on the instability and degeneracy of deep learning models},
+            colorlinks=true,
+            citecolor=blue,
+            urlcolor=blue,
+            linkcolor=magenta,
+            pdfborder={0 0 0}}
+\urlstyle{same}  % don't use monospace font for urls
+\setlength{\parindent}{0pt}
+\setlength{\parskip}{6pt plus 2pt minus 1pt}
+\setlength{\emergencystretch}{3em}  % prevent overfull lines
+\setcounter{secnumdepth}{5}
+
+% Tech stuff starts here
+\usepackage{psfrag,epsf}
+%\pdfminorversion=4
 
@@
 \newcommand{\Gamt}{\Gamma_{N,2}(\boldsymbol \theta_N) }
-
-%\pdfminorversion=4
-% NOTE: To produce blinded version, replace "0" with "1" below.
-\newcommand{\blind}{0}
+\bigskip
 
@@
 \addtolength{\textwidth}{1in}%
-\addtolength{\textheight}{-.3in}%
+\addtolength{\textheight}{1.3in}%
 \addtolength{\topmargin}{-.8in}%
@@
 
-\date{}
 \maketitle
@@
 
+\hfill {\tiny technometrics tex template (do not remove)}
 \newpage
-\spacingset{1.45}
+\spacingset{1.45} % DON'T change the spacing!
 

### 2017-11-02 writing/paper-note/appendices/appendix-note.Rmd
forgot to push last time

--- 2017-07-08 writing/paper-note/appendices/appendix-note.Rmd
@@
 
-**Proof of Proposition \@ref(prp:instab-elpr).** We prove the contrapositive, supposing that $\Delta(\boldsymbol \theta_N) \le C$ holds for some $C > 0$ and show $\text{ELPR}(\boldsymbol \theta_N) \leq NC$.
-Let $\boldsymbol x_{min} \equiv \argmin\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)$ and $\boldsymbol x_{max} \equiv \argmax\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)$.
+**Proof of Proposition \@ref(thm:instab-elpr).** For part (i), we prove the contrapositive, supposing that $\DN(\boldsymbol \theta_N) \le C$ holds for some $C > 0$ and show $\REP(\boldsymbol \theta_N) \leq NC$.
+Let $\boldsymbol x_{min} \equiv \argmin\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)$
+and $\boldsymbol x_{max} \equiv \argmax\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)$.
 Note there exists a sequence $\boldsymbol x_{min} \equiv \boldsymbol x_0, \boldsymbol x_1, \dots, \boldsymbol x_k \equiv \boldsymbol x_{max}$ in $\mathcal{X}^N$ of component-wise switches to move from $\boldsymbol x_{min}$ to $\boldsymbol x_{max}$ in the sample space (i.e.
-$\boldsymbol x_i, \boldsymbol x_{i + 1} \in \mathcal{X}^N$ differ in exactly $1$ component for $i = 0, \dots, k$) for some integer $k \in \{0, 1, \dots, N\}$.
-Under the FSFS model, recall $P_{\boldsymbol \theta_N}(\boldsymbol x) > 0$ holds so that $\log P_{\boldsymbol \theta_N}(\boldsymbol x)$ is well-defined for each outcome $\boldsymbol x \in \mathcal{X}^N$.
+$\boldsymbol x_i, \boldsymbol x_{i + 1} \in \mathcal{X}^N$ differ in exactly $1$ component, $i = 0, \dots, k$) for some integer $k \in \{0, 1, \dots, N\}$.
+Under the FOES model, recall
+$P_{\boldsymbol \theta_N}(\boldsymbol x) > 0$ holds so that $\log P_{\boldsymbol \theta_N}(\boldsymbol x)$ is well-defined for each outcome $\boldsymbol x \in \mathcal{X}^N$.
 Then, if $k > 0$, it follows that
 \begin{align*}
-\text{ELPR}(\boldsymbol \theta_N) = \log\left[\frac{P_{\boldsymbol \theta_N}(\boldsymbol x_{max})}{P_{\boldsymbol \theta_N}(\boldsymbol x_{min})}\right] &= \left|\sum\limits_{i = 1}^k\log\left(\frac{P_{\boldsymbol \theta_N}(\boldsymbol x_i)}{P_{\boldsymbol \theta_N}(\boldsymbol x_{i-1})}\right)\right| \\
+\REP(\boldsymbol \theta_N) = \log\left[\frac{P_{\boldsymbol \theta_N}(\boldsymbol x_{max})}{P_{\boldsymbol \theta_N}(\boldsymbol x_{min})}\right] &= \left|\sum\limits_{i = 1}^k\log\left(\frac{P_{\boldsymbol \theta_N}(\boldsymbol x_i)}{P_{\boldsymbol \theta_N}(\boldsymbol x_{i-1})}\right)\right| \\
 &\le \sum\limits_{i = 1}^k\left|\log\left(\frac{P_{\boldsymbol \theta_N}(\boldsymbol x_i)}{P_{\boldsymbol \theta}(\boldsymbol x_{i-1})}\right)\right| \le k \Delta_N(\boldsymbol \theta_N) \le NC,
 \end{align*}
-using $k \le N$ and $\Delta(\boldsymbol \theta_N) \le C$. If $k = 0$, then $\boldsymbol x_{max} = \boldsymbol x_{min}$ and the same bound above holds.\hfill $\Box$
+using $k \le N$ and $\Delta(\boldsymbol \theta_N) \le C$.
+If $k = 0$, then $\boldsymbol x_{max} = \boldsymbol x_{min}$ and the same bound above holds.
+This establishes part (i).
+To show part (ii), note the definition of S-instability (i.e., $\lim_{N\to \infty}\REP(\boldsymbol \theta_N)/N= \infty$) combined with part (i) implies that $\lim_{N\to \infty}\DN(\boldsymbol \theta_N)=\infty$.
+\hfill $\Box$
 
-**Proof of Proposition \@ref(prp:degenFSFS).** where $|\mathcal{X}|<\infty$ holds in the FSFS model.
-We may suppose $|\mathcal{X}|>1$ (i.e., $\mathcal{X}^N$ has more than one outcome) because otherwise the model is trivially degenerate for all $N \geq 1$.
-Fix $0 < \epsilon < 1$.
+**Proof of Proposition \@ref(thm:degenFOES).** As $|\mathcal{X}|<\infty$ holds in the FOES model, we may suppose
+$|\mathcal{X}|>1$; otherwise, $\mathcal{X}^N$ has one outcome and the model is  trivially degenerate for all $N \geq 1$.
+Fix $0 < \epsilon < 1$ and write $\boldsymbol x_{min} \equiv \argmin\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)$ and $\boldsymbol x_{max} \equiv \argmax\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)$.
 Then, $\boldsymbol x_{max} \in M_{\epsilon, \boldsymbol \theta_N}$, so $P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N}) \ge P_{\boldsymbol \theta_N}(\boldsymbol x_{max}) > 0$.
-If $\boldsymbol x \in \mathcal{X}^N \setminus M_{\epsilon, \boldsymbol \theta_N}$, then by definition $P_{\boldsymbol \theta_N}(\boldsymbol x) \le [P_{\boldsymbol \theta_N}(\boldsymbol x_{max})]^{1-\epsilon}[P_{\boldsymbol \theta_N}(\boldsymbol x_{min})]^{\epsilon}$ holds so that
-\begin{align*}
+If $\boldsymbol x \in \mathcal{X}^N \setminus M_{\epsilon, \boldsymbol \theta_N}$, then by definition
+$P_{\boldsymbol \theta_N}(\boldsymbol x) \le [P_{\boldsymbol \theta_N}(\boldsymbol x_{max})]^{1-\epsilon}[P_{\boldsymbol \theta_N}(\boldsymbol x_{min})]^{\epsilon}$ holds so that
+$$
 1-P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N})
-& = \sum\limits_{\boldsymbol x \in \mathcal{X}^N \setminus M_{\epsilon, \boldsymbol \theta_N}}P_{\boldsymbol \theta_N}(\boldsymbol x) \\
-& \le (|\mathcal{X}|^N)[P_{\boldsymbol \theta_N}(\boldsymbol x_{max})]^{1-\epsilon}[P_{\boldsymbol \theta_N}(\boldsymbol x_{min})]^{\epsilon}.
-\end{align*}
-From the lower bound on $P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N})$ and the upper bound on $1-P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N})$, it follows that
+ = \sum\limits_{\boldsymbol x \in \mathcal{X}^N \setminus M_{\epsilon, \boldsymbol \theta_N}}P_{\boldsymbol \theta_N}(\boldsymbol x)
+  \le (|\mathcal{X}|^N)[P_{\boldsymbol \theta_N}(\boldsymbol x_{max})]^{1-\epsilon}[P_{\boldsymbol \theta_N}(\boldsymbol x_{min})]^{\epsilon}.
+$$
+From the lower bound on $P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N})$ and the
+upper bound on $1-P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N})$, it follows that
 \begin{align*}
@@
 \end{align*}
-as $N \rightarrow \infty$ by the definition of an unstable FSFS model (cf.
-Definition \@ref(def:instabFSFS)).
-Consequently, $P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N}) \rightarrow 1$ as $N \rightarrow \infty$ as claimed.
+as $N \rightarrow \infty$ by the definition of an S-unstable FOES model \@ref(eq:Sun). Consequently,
+$P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N}) \rightarrow 1$ as $N \rightarrow \infty$ as claimed. \hfill $\Box$
+
+**Proof of Corollary \@ref(cor:sign).** The model condition PSR implies that
+\begin{equation}
+(\#eq:R)
+\frac{\max_{\boldsymbol y \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol y)  }{\min_{\boldsymbol y \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol y) } = \frac{\max_{\boldsymbol y \in \mathcal{X}^N}P_{-\boldsymbol \theta_N}(\boldsymbol y) }{\min_{\boldsymbol y \in \mathcal{X}^N}P_{-\boldsymbol \theta_N}(\boldsymbol y) }
+\end{equation}
+so that the log-ratio $\REP(\boldsymbol \theta_N)=\REP(-\boldsymbol \theta_N)$ is the same for both $\boldsymbol \theta_N$ and $-\boldsymbol \theta_N$ in \@ref(eq:elpr).
+Now part (i) of Corollary \@ref(cor:sign) follows from $\REP(\boldsymbol \theta_N)/N=\REP(-\boldsymbol \theta_N)/N\to \infty$ as $N\to \infty$ in \@ref(eq:Sun).
+To show part (ii), fix $0 < \epsilon < 1$ and consider a $\epsilon$-mode set $\mathcal{M}_{\epsilon, \boldsymbol \theta_N}$ under $\boldsymbol \theta_N$ from \@ref(eq:mode).
+If $\boldsymbol x \in \mathcal{M}_{\epsilon, \boldsymbol \theta_N}^c \equiv \mathcal{X}^N \setminus \mathcal{M}_{\epsilon, \boldsymbol \theta_N}$, then, by definition,
+$$
+\frac{P_{\boldsymbol \theta_N}(\boldsymbol x)}{\min_{\boldsymbol y \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol y)} \leq \left[\frac{\max_{\boldsymbol y \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol y)  }{\min_{\boldsymbol y \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol y) } \right]^{1-\epsilon}
+$$
+holds, which is equivalent to
+$$
+\frac{\max_{\boldsymbol y \in \mathcal{X}^N}P_{-\boldsymbol \theta_N}(\boldsymbol y)}{P_{-\boldsymbol \theta_N}(\boldsymbol x)} \leq
+\left[\frac{\max_{\boldsymbol y \in \mathcal{X}^N}P_{-\boldsymbol \theta_N}(\boldsymbol y)  }{\min_{\boldsymbol y \in \mathcal{X}^N}P_{-\boldsymbol \theta_N}(\boldsymbol y) } \right]^{1-\epsilon}
+$$
+by model condition PSR and \@ref(eq:R). The latter is in turn equivalent to
+\begin{equation}
+(\#eq:R2)
+\log P_{-\boldsymbol \theta_N}(\boldsymbol x) \geq \epsilon \max\limits_{\boldsymbol y \in \mathcal{X}^N} \log  P_{-\boldsymbol \theta_N}(\boldsymbol y) + (1-\epsilon)\min\limits_{\boldsymbol y \in \mathcal{X}^N} \log P_{-\boldsymbol \theta_N}(\boldsymbol y),
+\end{equation}
+so that $\boldsymbol x \in \mathcal{M}_{\epsilon, \boldsymbol \theta_N}^c$ if and only if \@ref(eq:R2) holds.
+Next consider the $(1-\epsilon)$-mode set $\mathcal{M}_{1-\epsilon, -\boldsymbol \theta_N}$ under $-\boldsymbol \theta_N$ from \@ref(eq:mode).
+If $\boldsymbol x \in\mathcal{M}_{1-\epsilon, -\boldsymbol \theta_N}$, then by definition $\boldsymbol x$ fulfills \@ref(eq:R2) and so $\boldsymbol x \in \mathcal{M}_{\epsilon, \boldsymbol \theta_N}^c$, showing that $\mathcal{M}_{1-\epsilon, -\boldsymbol \theta_N} \subset \mathcal{M}_{\epsilon, \boldsymbol \theta_N}^c$.
+By this and the fact that that Theorem \@ref(thm:degenFOES) and Corollary \@ref(cor:sign)(i) entail that $P_{-\boldsymbol \theta_N}(\boldsymbol X_N \in \mathcal{M}_{1-\epsilon, -\boldsymbol \theta_N})\to 1$ as $N\to \infty$ (i.e., $P_{-\boldsymbol \theta_N}$ is S-unstable), we have
+$$
+1  = \lim_{N\to \infty} P_{-\boldsymbol \theta_N}(\boldsymbol X_N \in \mathcal{M}_{1-\epsilon, -\boldsymbol \theta_N}) \leq \lim_{N\to \infty} P_{-\boldsymbol \theta_N}(\boldsymbol X_N \in \mathcal{M}_{\epsilon, \boldsymbol \theta_N}^c ) \leq 1,
+$$
+proving Corollary \@ref(cor:sign)(ii) \hfill $\Box$
+
+**Proof of Proposition \@ref(prp:prop1).** For any two outcomes $\boldsymbol x_1, \boldsymbol x_2\in\mathcal{X}^N$, the log-ratio of probabilities from the linear exponential model \@ref(eq:expo) with $k$ parameters/statistics satisfies
+$$
+\left|\log \left[ \frac{P_{\boldsymbol \theta_N}(\boldsymbol x_1)}{P_{\boldsymbol \theta_N}(\boldsymbol x_2)}  \right] \right| =
+\left|  \sum_{i=1}^k \theta_{i,N} [g_{i,k}(\boldsymbol x_1) - g_{i,k}(\boldsymbol x_2) ] \right|  \leq  \sum_{i=1}^k | \theta_{i,N}| (U_{i,N}-L_{i,N});
+$$
+consequently, $\REP(\boldsymbol \theta_N )  \leq  \sum_{i=1}^k | \theta_{i,N}| (U_{i,N}-L_{i,N})$ holds in \@ref(eq:elpr) and model stability in Proposition \@ref(prp:prop1) follows from \@ref(eq:Sun).
+\hfill $\Box$
+
+**Proof of Proposition \@ref(prp:prop2).**
+Writing $\boldsymbol x=(x_1,\ldots,x_N)$ and $\boldsymbol h = (h_1,\ldots,h_{N_{\mathcal{H}}})$ with all components $x_i,h_j\in\{\pm 1\}$, probabilities in the joint RBM model \@ref(eq:RBM1) can be written as $\tilde{P}_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h) = c(\boldsymbol \theta_N)\exp[ f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)]$ in terms of the function $f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)$ from \@ref(eq:f) and the normalizing constant $c(\boldsymbol \theta_N)= \exp [-\psi(\boldsymbol \theta_N)]$ from \@ref(eq:RBM1).
+Let $\boldsymbol x_M, \boldsymbol x_m\in\{\pm 1\}^N$ be such that $P_{\boldsymbol \theta_N} (\boldsymbol x_M) = \max_{\boldsymbol x}P_{\boldsymbol \theta_N} (\boldsymbol x)$ and $P_{\boldsymbol \theta_N} (\boldsymbol x_m) = \min_{\boldsymbol x}P_{\boldsymbol \theta_N} (\boldsymbol x)$ under the marginal RBM model $P_{\boldsymbol \theta_N} (\boldsymbol x) = c(\boldsymbol \theta_N)\sum_{\boldsymbol h \in\{\pm 1\}^{\mathcal{N}_H}} \tilde{P}_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)= c(\boldsymbol \theta_N)\sum_{\boldsymbol h \in\{\pm 1\}^{\mathcal{N}_H}} \exp[ f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)]$ from \@ref(eq:RBM2).
+Also, $\boldsymbol x_0,x_1\in\{\pm 1\}^N$ be such that $\max_{\boldsymbol h}f_{\boldsymbol \theta_N} (\boldsymbol x_0, \boldsymbol h)=\max_{\boldsymbol x}\max_{\boldsymbol h}f_{\boldsymbol \theta_N} (\boldsymbol x , \boldsymbol h)$ and $\max_{\boldsymbol h}f_{\boldsymbol \theta_N} (\boldsymbol x_1, \boldsymbol h)=\min_{\boldsymbol x}\max_{\boldsymbol h}f_{\boldsymbol \theta_N} (\boldsymbol x , \boldsymbol h)$.
+Then, Proposition \@ref(prp:prop2)(i) follows from $\REP_{(\boldsymbol X)}(\boldsymbol \theta_N) = \log[P_{\boldsymbol \theta_N} (\boldsymbol x_M) /P_{\boldsymbol \theta_N} (\boldsymbol x_m) ]$ and the lower/upper bounds on $P_{\boldsymbol \theta_N} (\boldsymbol x_M)$ and $P_{\boldsymbol \theta_N} (\boldsymbol x_m)$ as
+$$
+c(\boldsymbol \theta_N) \exp[\max_{\boldsymbol h}f_{\boldsymbol \theta_N} (\boldsymbol x_0 , \boldsymbol h)]
+\leq P_{\boldsymbol \theta_N} (\boldsymbol x_0) \leq  P_{\boldsymbol \theta_N} (\boldsymbol x_M) \leq  2^{N_{\mathcal{H}}} c(\boldsymbol \theta_N) \exp[\max_{\boldsymbol x}\max_{\boldsymbol h}f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)]
+$$
+and
+\begin{align*}
+c(\boldsymbol \theta_N) \exp[\min_{\boldsymbol x}\max_{\boldsymbol h}f_{\boldsymbol \theta_N} (\boldsymbol x , \boldsymbol h)] &\leq c(\boldsymbol \theta_N) \exp[ \max_{\boldsymbol h}f_{\boldsymbol \theta_N} (\boldsymbol x_m , \boldsymbol h)] \\
+&\leq P_{\boldsymbol \theta_N} (\boldsymbol x_m) \\
+&\leq P_{\boldsymbol \theta_N} (\boldsymbol x_1) \\ 
+&\leq 2^{N_{\mathcal{H}}} c(\boldsymbol \theta_N) \exp[ \max_{\boldsymbol h}f_{\boldsymbol \theta_N} (\boldsymbol x_1 , \boldsymbol h)]\\&=&2^{N_{\mathcal{H}}} c(\boldsymbol \theta_N) \exp[\min_{\boldsymbol x}\max_{\boldsymbol h}f_{\boldsymbol \theta_N} (\boldsymbol x , \boldsymbol h)]
+\end{align*}
+
+To prove Proposition \@ref(prp:prop2), we next expand the function $f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)$ from \@ref(eq:f) as
+$$
+f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)= \sum_{j=1}^{N_\mathcal{H}} h_j \theta_{j,N}^{\mathcal{H}}  +   \sum_{i=1}^N \left( \theta_{i,N}^{\mathcal{V}} + \sum_{j=1}^{N_\mathcal{H}}  h_j  \theta_{ij,N}^{\mathcal{VH}}\right)x_i=\sum_{i=1}^N x_i  \theta_{i,N}^{\mathcal{V}} +  \sum_{j=1}^{N_\mathcal{H}} \left( \theta_{j,N}^{\mathcal{H}}  + \sum_{i=1}^N  x_i  \theta_{ij,N}^{\mathcal{VH}}\right)h_j.
+$$
+By this and the fact that $x_i,h_j\in\{\pm 1\}$, we then have
+\begin{align}
+\nonumber \max_{\boldsymbol x}  f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h) =  \sum_{j=1}^{N_\mathcal{H}} h_j \theta_{j,N}^{\mathcal{H}}  +   a_{\boldsymbol \theta_N, \mathcal{H}} (\boldsymbol h), & \min_{\boldsymbol x}  f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h) =  \sum_{j=1}^{N_\mathcal{H}} h_j \theta_{j,N}^{\mathcal{H}}  - a_{\boldsymbol \theta_N, \mathcal{H}} (\boldsymbol h),\\
+\nonumber \max_{\boldsymbol h}  f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h) = \sum_{i=1}^{N}x_i \theta_{i,N}^{\mathcal{V}} + b_{\boldsymbol \theta_N, \mathcal{V}} (\boldsymbol x), & \min_{\boldsymbol h}  f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h) = \sum_{i=1}^{N}x_i \theta_{i,N}^{\mathcal{V}}  -   b_{\boldsymbol \theta_N, \mathcal{V}} (\boldsymbol x), \\
+(\#eq:max)
+a_{\boldsymbol \theta_N, \mathcal{H}} (\boldsymbol h) \equiv \sum_{i=1}^N \left| \theta_{i,N}^{\mathcal{V}} + \sum_{j=1}^{N_\mathcal{H}}  h_j  \theta_{ij,N}^{\mathcal{VH}}\right|, &  b_{\boldsymbol \theta_N, \mathcal{V}} (\boldsymbol x) \equiv \sum_{j=1}^{N_\mathcal{H}} \left| \theta_{j,N}^{\mathcal{H}}  + \sum_{i=1}^N  x_i  \theta_{ij,N}^{\mathcal{VH}}\right|,
+\end{align}
+where  $\boldsymbol h^T \boldsymbol \theta_N^{\mathcal{H}}=\sum_{j=1}^{N_\mathcal{H}} h_j \theta_{j,N}^{\mathcal{H}}$, $\boldsymbol x^T \boldsymbol \theta_N^{\mathcal{V}}=  \sum_{i=1}^{N}x_i \theta_{i,N}^{\mathcal{V}}$ and $\Gam\equiv \max_{\boldsymbol h} a_{\boldsymbol \theta_N, \mathcal{H}} (\boldsymbol h)$.
+From this, it follows that
+\begin{align*}
+\REP_{(\boldsymbol X, \boldsymbol H)}(\boldsymbol \theta_N) &=  \max_{\boldsymbol h}\max_{\boldsymbol x}f_{\boldsymbol \theta_N} (\boldsymbol x , \boldsymbol h) -  \min_{\boldsymbol h}\min_{\boldsymbol x}f_{\boldsymbol \theta_N} (\boldsymbol x , \boldsymbol h)\\
+&= \max_{\boldsymbol h_1 }  \max_{\boldsymbol h_2 }\left[ (\boldsymbol h_1 - \boldsymbol h_2)^T \boldsymbol \theta_N^{\mathcal{H}}  +   a_{\boldsymbol \theta_N, \mathcal{H}} (\boldsymbol h_1)  + a_{\boldsymbol \theta_N, \mathcal{H}} (\boldsymbol h_2)\right],
+\end{align*}
+which leads to the upper bound $\REP_{(\boldsymbol X, \boldsymbol H)}(\boldsymbol \theta_N) \leq 2 \Gam + 2 |\boldsymbol \theta_N^{\mathcal{H}} |_1$.
+Then, taking $\boldsymbol h_1=\boldsymbol h_2$ (i.e., before maximization) gives $\REP_{(\boldsymbol X, \boldsymbol H)}(\boldsymbol \theta_N) \geq 2\Gam$ and taking $\boldsymbol h_1=-\boldsymbol h_2$, such that $\boldsymbol h_1^T \boldsymbol \theta_N^{\mathcal{H}} = |\theta_N^{\mathcal{H}}|_1$, gives $\REP_{(\boldsymbol X, \boldsymbol H)}(\boldsymbol \theta_N) \geq 2|\theta_N^{\mathcal{H}}|_1$; this yields the lower bound $\REP_{(\boldsymbol X, \boldsymbol H)}(\boldsymbol \theta_N)\geq 2\max\{\Gam, |\boldsymbol \theta_N^{\mathcal{H}} |_1\}$.
+
+We next consider $\elt$ and, by \@ref(eq:max), write
+\begin{align*}
+\elt &=  \max_{\boldsymbol h}\max_{\boldsymbol x}f_{\boldsymbol \theta_N} (\boldsymbol x , \boldsymbol h) -  \max_{\boldsymbol h}\min_{\boldsymbol x}f_{\boldsymbol \theta_N} (\boldsymbol x , \boldsymbol h)\\
+&= \max_{\boldsymbol h_1} \min_{\boldsymbol h_2 }\left[ (\boldsymbol h_1 - \boldsymbol h_2)^T \boldsymbol \theta_N^{\mathcal{H}} + a_{\boldsymbol \theta_N, \mathcal{H}} (\boldsymbol h_1)  + a_{\boldsymbol \theta_N, \mathcal{H}} (\boldsymbol h_2)\right].
+\end{align*}
+Taking $\boldsymbol h_1=\boldsymbol h_2$ and maximizing over both $\boldsymbol h_1,\boldsymbol h_2$ produces the upper bound $\elt \leq 2\Gam$.
+Then, using $(\boldsymbol h_1 - \boldsymbol h_2)^T \boldsymbol \theta_N^{\mathcal{H}} + a_{\boldsymbol \theta_N, \mathcal{H}} (\boldsymbol h_2) \geq - 2|\boldsymbol \theta_N^{\mathcal{H}} |_1$ and maximizing over $\boldsymbol h_1$ gives $\elt \geq  \Gam- 2|\boldsymbol \theta_N^{\mathcal{H}} |_1$, while setting $\boldsymbol h_1=\boldsymbol h_2^*$ for $\boldsymbol h_2^*$ such that $-(\boldsymbol h_2^*) ^T \boldsymbol \theta_N^{\mathcal{H}}  + a_{\boldsymbol \theta_N, \mathcal{H}} (\boldsymbol h_2^*) = \min_{\boldsymbol h_2} [-\boldsymbol h_2^T \boldsymbol \theta_N^{\mathcal{H}}  + a_{\boldsymbol \theta_N, \mathcal{H}} (\boldsymbol h_2)]$ gives $\elt \geq 2 a_{\boldsymbol \theta_N, \mathcal{H}} (\boldsymbol h_2^*) \geq \Gamc \equiv \min_{\boldsymbol h}  a_{\boldsymbol \theta_N, \mathcal{H}} (\boldsymbol h)$.
+Finally, note that for any $\boldsymbol h$, the triangle inequality gives
+\begin{align*}
+\Gam  \equiv \max_{\boldsymbol h_1}  a_{\boldsymbol \theta_N, \mathcal{H}} (\boldsymbol h_1) &\geq
+[a_{\boldsymbol \theta_N, \mathcal{H}} (\boldsymbol h)  + a_{\boldsymbol \theta_N, \mathcal{H}} (-\boldsymbol h)]/2 \\
+&=
+2^{-1}\sum_{i=1}^{N } \left(\left| \theta_{i,N}^{\mathcal{V}}   + \sum_{j=1}^{N_{\mathcal{H}}} h_j \theta_{ij,N}^{\mathcal{VH}} \right|
++ \left| \theta_{i,N}^{\mathcal{V}}   - \sum_{j=1}^{N_{\mathcal{H}}} h_j \theta_{ij,N}^{\mathcal{VH}} \right|\right)\\
+& \geq \sum_{i=1}^{N }  \left| \theta_{i,N}^{\mathcal{V}} \right| \equiv |\boldsymbol \theta_{N}^{\mathcal{V}}|_1.
+\end{align*}
 \hfill $\Box$

### 2017-11-02 writing/paper-note/chapters/instab.Rmd
forgot to push last time

@@
 
-We consider the behavior, and the potential impropriety, sequences of discrete probability models built to incorporate observations of increasing sample size $N$.
+We consider the behavior, and the potential impropriety, of sequences of discrete probability models built to incorporate observations of increasing sample size $N$.
 Interest is in identifying  instability in such models, which is roughly characterized by probabilities with extreme sensitivity to small changes in data configuration.
@@
 As an example, such models can include near-degenerate distributions that assign all essential probability mass to only a subset of an overall sample space.
-The latter issue in connection to degeneracy has been recognized as concern in that dominant model outcomes  may not resemble observed data [cf.
+The latter issue in connection to degeneracy has been recognized as a concern in that dominant model outcomes  may not resemble observed data [cf.
 @handcock2003assessing].
@@
 Section \@ref(instability-results) then establishes several formal results for  FOES models with regard to instability.
-@schweinberger2011instability originally developed instability results framed for, or specific to, a certain class of discrete exponential models.
+@schweinberger2011instability originally developed instability results specific to a certain class of discrete exponential models.
 For similar exponential models with random networks, @handcock2003assessing studied model degeneracy, where a probability model places near complete mass on modes and may thereby narrow the feasible model outcomes.
@@
 Section \@ref(implications) then emphasizes the implications of such model instability, showing that such impropriety can be expected to numerically hinder maximum likelihood estimation and   MCMC-based simulations.
-As one potential remedy, suggestions are given or constraining model parameterizations and avoid the most problematic regions of the parameter space.
+As one potential remedy, suggestions are given or constraining model parameterizations to avoid the most problematic regions of the parameter space.
 Proofs of the main results appear in Appendix \@ref(appendix-instab).
@@
 
-## Discrete exponential family models
+## Discrete Exponential Family Models
 
@@
 with sufficient statistic $\boldsymbol g_N(\boldsymbol x)$ involving a count $\sum_{i=1}^N \mathbb{I}(x_i=j)$ for each outcome $j \in \{1,\ldots,k\}$, where $\mathbb{I}(\cdot)$ denotes the indicator function, and parameters $\boldsymbol \theta_N=(\theta_{1,N},\ldots,\theta_{k,N})\in\mathbb{R}^k$ defining log-probability ratios $\theta_{i,N}-\theta_{j,N} =\log [P_{\boldsymbol \theta_N}(X_1=i)/P_{\boldsymbol \theta_N}(X_1=j)]$.
-In addition to such standard models for discrete independent data,
-exponential models of FOES type commonly arise with dependent spatial data [@besag1974spatial] and network/relational data [@wasserman1994social; @handcock2003assessing].
+In addition to such standard models for discrete independent data, exponential models of FOES type commonly arise with dependent spatial data [@besag1974spatial] and network/relational data [@wasserman1994social; @handcock2003assessing].
 For a random graph  or network  with, say, $n$ nodes,  consider $N={n \choose 2}$ random edges where the $i$th edge is associated with a pair of nodes $s_i \equiv \{v_i,u_i\}$  and a binary variable $X_i\in\{0,1\}$ indicating   presence/absence of an edge among the node pair $s_i$, $i=1,\ldots,N$.
@@
 
-## Restricted Boltzmann machines
+## Restricted Boltzmann Machines
 
@@
 That is, in contrast to \@ref(eq:expo), the natural parameter function involved in the RBM model \@ref(eq:RBM1), as the identity mapping of the parameters $\boldsymbol \theta_N\in\mathbb{R}^{q(N)}$, naturally grows in  dimension  $q(N)\to \infty$ to accommodate visible variables $X_1, \dots, X_N$ of increasing sample size $N\to \infty$.
-Additionally, one may further arbitrarily choose the number $N_\mathcal{H}$ of hidden variables $\boldsymbol H$ in the joint RMB model \@ref(eq:RBM1) to define a marginal model \@ref(eq:RBM2) for the $N$ visible variables $\boldsymbol X$, and the number $N_\mathcal{H}$ of hiddens may also potentially   increase with $N$.
+Additionally, one may further arbitrarily choose the number $N_\mathcal{H}$ of hidden variables $\boldsymbol H$ in the joint RBM model \@ref(eq:RBM1) to define a marginal model \@ref(eq:RBM2) for the $N$ visible variables $\boldsymbol X$, and the number $N_\mathcal{H}$ of hiddens may also potentially increase with $N$.
 Because $|\mathcal{X}| = 2$ and $P_{\boldsymbol \theta_N}(\boldsymbol x) > 0$ for all $\boldsymbol x \in \mathcal{X}^N$, the RBM  specification \@ref(eq:RBM2) for visibles $\boldsymbol$ corresponds to a FOES model, while the joint distribution \@ref(eq:RBM1) for $(\boldsymbol X, \boldsymbol H)$ is also a FOES model.
@@
 
-## Deep learning
+## Deep Learning
+
+<!-- http://search.library.duke.edu/search?id=DUKE002216318 -->
 
 Consider two models with "deep architecture" that contain multiple hidden (or latent) layers in addition to a visible layer of data, namely a deep Boltzmann machine [@salakhutdinov2009deep] and a deep belief network @hinton2006fast].
-Let $M$ denote the
-number of hidden layers included in the model and let $N_{(H,1)}, \dots, N_{(H,M)}$ denote the numbers of hidden variables within each hidden layer. Then the random vector
-$\tilde{\boldsymbol X} = \{H^{(1)}_1, \dots, H^{(1)}_{N_{(H,1)}}, \dots, H^{(M)}_1, \dots, H^{(M)}_{N_{(H,M)}}, \boldsymbol X\}$ collects both the hidden variables
+Let $M$ denote the number of hidden layers included in the model and let $N_{(H,1)}, \dots, N_{(H,M)}$ denote the numbers of hidden variables within each hidden layer.
+Then the random vector $\tilde{\boldsymbol X} = \{H^{(1)}_1, \dots, H^{(1)}_{N_{(H,1)}}, \dots, H^{(M)}_1, \dots, H^{(M)}_{N_{(H,M)}}, \boldsymbol X\}$ collects both the hidden variables
 $\{ H_{i}^{(j)} : i=1,\ldots, N_{(H,j)}, j=1,\ldots,M\}$ and visible variables $\boldsymbol X =(X_1,\ldots,X_N)$ in a deep probabilistic model.
@@
 The difference between the DBM and DBN models is that all but the last stacked layer in a DBN are Bayesian networks [see @pearl985bayesian], rather than RBMs.
-Thus for visibles $X_1, \dots, X_N$ with support $\mathcal{X}^N$, a DBN is also a FOES model if the number $q(N)$ of components in the parameter vector is dependent on the dimension of the visibles.
-Commonly, as in logistic belief nets [@neal1992connectionist], a "weight" parameter is placed on each interaction between visibles, $X_1, \dots, X_N$, and the first layer of latent variables, $H^{(1)}_1, \dots, H^{(1)}_{N_{(H,1)}}$, in the definition of a FOES model.
+A Bayesian network is a probabilistic graphical model that defines conditional dependence to be directed, rather than undirected (as with the RBM).
+Thus for visibles $X_1, \dots, X_N$ with support $\mathcal{X}^N, \mid \mathcal{X} \mid < \infty$, a DBN is also a FOES model with $q(N)$ the length of parameter vector is dependent on the dimension of the visibles because $P_{\boldsymbol \theta_N}(\boldsymbol x)>0$ for all $\boldsymbol x \in\mathcal{X}^N$.
+Commonly, as in logistic belief nets [@neal1992connectionist], a "weight" parameter is placed on each interaction between visibles, $X_1, \dots, X_N$, and the first layer of latent variables, $H^{(1)}_1, \dots, H^{(1)}_{N_{(H,1)}}$, satisfying the definition of a FOES model.
 
-# Main results on model instability {#instability-results}
+# Main Results on Model Instability {#instability-results}
 
-## A criterion for instability {#criterion}
+We now present a formal definition for instability of FOES models as well as a simple condition for identifying instability in a FOES model sequence.
+
+## A Criterion for Instability {#criterion}
 
@@
 based on maximum and minimal model probabilities.
-In what follows, the main idea is that instability, and other negative model features, can be associated with a FOES model formulation for $N$ random variables where the log-ratio \@ref(eq:elpr) is overly large relative to the sample size $N$.
+In what follows, the main idea is that instability, and other negative model features, can be associated with a FOES model formulation for $N$ random variables where the LREP \@ref(eq:elpr) is overly large relative to the sample size $N$.
 That is, a sequence of FOES probability models $P_{\boldsymbol \theta_N}$ results in specifying the distribution of observations $\boldsymbol X=(X_1,\ldots,X_N)$ for each sample size $N \geq 1$ and instability will generally occur among these models whenever the corresponding LREP \@ref(eq:elpr) grows faster than $N$.
@@
 
-This definition of S-unstable is a generalization or reinterpretation of "unstable" used in @schweinberger2011instability by allowing possibility non-exponential family models (e.g., RBM and DBM models in Sections \@ref(restricted-boltzmann-machines) - \@ref(deep-learning) as well as a potentially increasing number $d(N)$ of parameters through the parameter sequence $\boldsymbol \theta_N\in \mathbb{R}^{d(N)}$.
+This definition of S-unstable is a generalization or reinterpretation of "unstable" used in @schweinberger2011instability by allowing possibly non-exponential family models (e.g., RBM and DBM models in Sections \@ref(restricted-boltzmann-machines)-\@ref(deep-learning) as well as a potentially increasing number $q(N)$ of parameters through the parameter sequence $\boldsymbol \theta_N\in \mathbb{R}^{q(N)}$.
 While this definition differs in form and scope from the original, it does match that in @schweinberger2011instability for the special case of exponential models (cf.
@@
 Section \@ref(illustrations) provides several examples  of unstable models as well as causes for model instability, where the latter may often be traced to issues in model form (i.e., data functions) and/or parameterization.
-We next describe several potentially undesirable features associated with unstable models.
+We next describe several potentially undesirable features associated with S-unstable FOES models.
 
@@
 
-## Characterizations and consequences of instability
+## Characterizations and Consequences of Instability
 
@@
 which represents the biggest log-probability ratio for a one component change in data outcomes in a FOES model with parameter $\boldsymbol \theta_N$.
-We then have the following result prescribing the behavior of $\DN(\boldsymbol \theta_N)$ for unstable models.
+We then have the following result prescribing the behavior of $\DN(\boldsymbol \theta_N)$ for S-unstable FOES models.
 
@@
 While not all one component changes in data may produce massive changes in probability, unstable models must have some such data outcomes with this property.
-As a consequence, unstable probability structure may exhibit extreme sensitivity through large peaks and troughs over the sample space.
+As a consequence, unstable probability structures may exhibit extreme sensitivity through large peaks and troughs over the sample space.
 
@@
 Furthermore, as the instability measure grows and the log-ratio $\REP(\boldsymbol \theta_N)/N$ diverges, as in the definition \@ref(eq:Sun) of S-unstable models, then a FOES model sequence  will become degenerate.
-Theorem \@ref(thm:degenFOES) provides a formal statement of such degeneracy due to instability.
+Theorem \@ref(thm:degenFOES) provides a formal statement of such degeneracy due to S-instability.
 For a given $0 < \epsilon < 1$, define a $\epsilon$-modal set of outcomes as
@@
 In other words, as the sample size grows in S-unstable FOES models, all probability   tends to concentrate mass on an $\epsilon$-modal set, where $\epsilon$ can be made arbitrarily small.
-Intuitively, the occurrence of such degeneracy can be explained
-by a type of "reverse" pigeonhole principle for unstable FOES models: if all outcomes should receive positive probability but the maximal probability far exceeds the minimal one in the model, then little probability remains for distribution among remaining model outcomes (i.e., if nearly all available pigeons are stuffed into one hole, the remaining pigeonholes must have few occupants).
+Intuitively, the occurrence of such degeneracy can be explained by a type of "reverse" pigeonhole principle for unstable FOES models: if all outcomes should receive positive probability but the maximal probability far exceeds the minimal one in the model, then little probability remains for distribution among remaining model outcomes (i.e., if nearly all available pigeons are stuffed into one hole, the remaining pigeonholes must have few occupants).
 Degeneracy in unstable models can pose dangers in data modeling as well, particularly when a mode set represents a narrow collection of outcomes among those realistically possible for adequately describing data.
@@
 
-Connected to degeneracy, unstable models may also exhibit additional kinds of  extreme and undesirable sensitivity in probabilities if model parameters $\boldsymbol \theta_N$ can further be "dialed" between positive and negative values.
+Connected to degeneracy, S-unstable FOES models may also exhibit additional kinds of  extreme and undesirable sensitivity in probabilities if model parameters $\boldsymbol \theta_N$ can further be "dialed" between positive and negative values.
 That is, some FOES models naturally involve parameter spaces covering a positive-negative spectrum of parameter possibilities, where the signs of parameters provide a standard device for increasing or decreasing probabilities of outcomes in the model formulation.
@@
 
-```{remark, name="need to briefly finish", echo=TRUE}
-extensions of sensitivity...
-basically, model condition PSR and Corollary \ref{cor:sign} can be extended to cases where parameter components $\boldsymbol \theta_N =(\boldsymbol \theta_{1,N}, \boldsymbol \theta_{N,2})$ are not all changed in sign ($-\boldsymbol \theta_N$) but, more generally, are instead altered to another parameter configuration $\boldsymbol \theta_{N}^A = ( \boldsymbol \theta_{1,N}^A, \boldsymbol \theta_{2,N}^A )$ involving a switch in sign only among some model parameters $\boldsymbol \theta_{N,2}^A=- \boldsymbol \theta_{N,2}$ with remaining parameters $\boldsymbol \theta_{1,N}^A$ being arbitrarily chosen.
-If a change sign occurs among parameters ($\pm \boldsymbol \theta_{N,2}$) which dominate the probability structure of the model, then an unstable model under $\boldsymbol \theta_N$ will then imply that many more unstable models exist  over a broad spectrum of possibilities for variations $\boldsymbol \theta_{N}^A$ of $\boldsymbol \theta_N$.
+```{remark, echo=TRUE}
+Model condition PSR and Corollary \ref{cor:sign} can be extended to cases where parameter components $\boldsymbol \theta_N =(\boldsymbol \theta_{1,N}, \boldsymbol \theta_{N,2})$ are not all changed in sign ($-\boldsymbol \theta_N$) but, more generally, are instead altered to another parameter configuration $\boldsymbol \theta_{N}^A = (\boldsymbol \theta_{N,1}^A, \boldsymbol \theta_{N,2}^A)$ involving a switch in sign only among some model parameters $\boldsymbol \theta_{N,2}^A = -\boldsymbol \theta_{N,2}$ with remaining parameters $\boldsymbol \theta_{1,N}^A$ being arbitrarily chosen.
+If a sign change occurs among parameters ($\pm \boldsymbol \theta_{N,2}$) which dominate the probability structure of the model, then an unstable model under $\boldsymbol \theta_N$ will then imply that many more unstable models exist over a broad spectrum of possibilities for variations $\boldsymbol \theta_{N}^A$ of $\boldsymbol \theta_N$.
 ```
@@
 However, as shown in the following, bounded statistics may still lead to instability if the parameter dimension is high.
-We next provides some examples to illustrate instability in FOES models, which also suggest  some potential strategies for preventing unstable models.
+We next provides some examples to illustrate S-instability in FOES models, which also suggest some potential strategies for preventing unstable models.
 
-## Equi-probability models
+## Equi-probability Models
 
@@
 
-## One-parameter exponential models {#one-param-exp}
+## One-parameter Exponential Models {#one-param-exp}
 
@@
 
-## Fixed-dimensional linear exponential models {#fixed-dim-exp}
+## Fixed-dimensional Linear Exponential Models {#fixed-dim-exp}
 
@@
 Consequently, for all fixed parameters excluding $\theta_2=\theta_3=0$, $\lim_{N\to \infty}\REP_N(\boldsymbol \theta_N)/N=\infty$ then follows and the graph model with 2-stars and triangles is S-unstable, as suggested by the breach of Proposition \@ref(prp:prop1) for this model when $|\theta_2|+|\theta_3|\neq 0$.
-That is, instability holds even under $\theta_2 = - \theta_3/3$ case potentially allowed by Schweinberger's results.
+That is, instability holds even under $\theta_2 = - \theta_3/3$ case potentially allowed by Schweinberger's [-@schweinberger2011instability] results.
 
-## Latent variable models of increasing parameter dimension
+## Latent Variable Models of Increasing Parameter Dimension
 
-We next consider instability of discrete data models based on exponential formulations involving hidden, or latent, variables, such as those described in Sections \@ref(restricted-boltzmann-machines)-\@ref(deep-learning) for networks.
+We next consider instability of discrete data models based on exponential formulations involving hidden, or latent, variables, such as those probabilistic graphical models described in Sections \@ref(restricted-boltzmann-machines)-\@ref(deep-learning).
 We will focus on restricted Bolzmann machine (RBM) models (Section \@ref(restricted-boltzmann-machines), having one layer of latent variables for simplicity, though the same instability concepts may be extended to other deep learning models (Section \@ref(deep-learning).
 For $N$ visible variables $\boldsymbol X \equiv \boldsymbol X_N = (X_1,\ldots,X_n)$ as data, each observation $X_i\in\{\pm 1\}$ being binary, the RBM-based model \@ref(eq:RBM2) for  $\boldsymbol X$ is again of FOES-type, though not an exponential model.
-However, the distribution of visible variables is induced by an underlying joint exponential model \@ref(eq:RBM1) for both visible and latent variables $(\boldsymbol X, \boldsymbol H)$, where $\boldsymbol H=(H_1,\ldots,H_{N_{\mathcal{H}}})$ denotes a vector of $N_{\mathcal{H}}$ hidden variables (similarly binary).
-The joint model is of linear exponential form involving $d(N)\equiv N + N_{\mathcal{H}} + N* N_{\mathcal{H}}$ sufficient statistics given by $(\boldsymbol X, \boldsymbol H, \boldsymbol X^T\boldsymbol H)$ and parameters $\boldsymbol \theta_N = (\boldsymbol \theta_N^{\mathcal{V}},\boldsymbol \theta_N^{\mathcal{H}}, \boldsymbol \theta_N^{\mathcal{VH}}  ) \in\mathbb{R}^{d(N)}$ corresponding to the $N$ visible variables $\boldsymbol X$ (i.e., $\boldsymbol \theta_N^{\mathcal{V}}\in\mathbb{R}^N$), the $N_{\mathcal{H}}$ hidden variables $\boldsymbol H$ (i.e., $\boldsymbol \theta_N^{\mathcal{H}}\in\mathbb{R}^{N_{\mathcal{H}}}$), and the $N *N_{\mathcal{H}}$ cross-product variables $\boldsymbol X^T\boldsymbol H$ (i.e., $\boldsymbol \theta_N^{\mathcal{VH}}\in\mathbb{R}^{N N_{\mathcal{H}}}$).
+However, the distribution of visible variables is induced by an underlying joint exponential model \@ref(eq:RBM1) for both visible and latent variables $(\boldsymbol X, \boldsymbol H)$, where $\boldsymbol H = (H_1,\ldots,H_{N_{\mathcal{H}}})$ denotes a vector of $N_{\mathcal{H}}$ hidden variables (similarly binary).
+The joint model is of linear exponential form involving $q(N)\equiv N + N_{\mathcal{H}} + N*N_{\mathcal{H}}$ sufficient statistics given by $(\boldsymbol X, \boldsymbol H, \boldsymbol X^T\boldsymbol H)$ and parameters $\boldsymbol \theta_N = (\boldsymbol \theta_N^{\mathcal{V}},\boldsymbol \theta_N^{\mathcal{H}}, \boldsymbol \theta_N^{\mathcal{VH}}  ) \in\mathbb{R}^{q(N)}$ corresponding to the $N$ visible variables $\boldsymbol X$ (i.e., $\boldsymbol \theta_N^{\mathcal{V}}\in\mathbb{R}^N$), the $N_{\mathcal{H}}$ hidden variables $\boldsymbol H$ (i.e., $\boldsymbol \theta_N^{\mathcal{H}}\in\mathbb{R}^{N_{\mathcal{H}}}$), and the $N *N_{\mathcal{H}}$ cross-product variables $\boldsymbol X^T\boldsymbol H$ (i.e., $\boldsymbol \theta_N^{\mathcal{VH}}\in\mathbb{R}^{N*N_{\mathcal{H}}}$).
 However, unlike some previous exponential models considered in Sections \@ref(one-param-exp)-\@ref(fixed-dim-exp) (cf.
 Proposition \@ref(prp:prop1), note that the RBM formulation always associates parameters with *bounded* statistics (i.e., the components of $(\boldsymbol X, \boldsymbol H, \boldsymbol X^T\boldsymbol H)$) so that  model instability cannot arise here due to the magnitude of sufficient statistics exceeding the sample size $N$.
-Instead, RBM instability  may be linked to solely to parameter configuration and the fact that the number $d(N) \geq N$ of parameters necessarily increases with the number $N$ of observations $\boldsymbol X$, in contrast to previous exponential cases of fixed parameter dimension.
+Instead, RBM instability  may be linked solely to parameter configuration and the fact that the number $q(N) \geq N$ of parameters necessarily increases with the number $N$ of observations $\boldsymbol X$, in contrast to previous exponential cases of fixed parameter dimension.
 
 To highlight the instability issues for the RBM model, consider a simple  model for $N$ visibles $\boldsymbol X$ with no hidden variables ($N_{\mathcal{H}}=0$), for which model statements \@ref(eq:RBM1)-\@ref(eq:RBM2) coincide.
-An independence model then results for variables in $\boldsymbol X$, which has $d(N)=N$ parameters $\boldsymbol \theta_N^{\mathcal{V}} = (\theta_{1,N}^{\mathcal{V}}, \ldots, \theta_{N,N}^{\mathcal{V}}) \in \mathbb{R}^{N}$, and the measure of model instability becomes
+An independence model then results for variables in $\boldsymbol X$, which has $q(N)=N$ parameters $\boldsymbol \theta_N^{\mathcal{V}} = (\theta_{1,N}^{\mathcal{V}}, \ldots, \theta_{N,N}^{\mathcal{V}}) \in \mathbb{R}^{N}$, and the measure of model instability becomes
 $$
@@
 Consequently, even for a simplest RBM model involving independence, preventing instability requires careful choice of parameters, particularly with regard to how a parameter configuration differs from zero.
-For more general RBM models, the number $N_{\mathcal{H}}$ of hidden variables $\boldsymbol H$ can also be chosen arbitrarily (i.e., as some function $N_{\mathcal{H}}\equiv N_{N,\mathcal{H}}$ of $N$), which can substantially inflate the number $d(N)$ of model parameters and further impact model instability through accumulated parameters.
+For more general RBM models, the number $N_{\mathcal{H}}$ of hidden variables $\boldsymbol H$ can also be chosen arbitrarily (i.e., as some function $N_{\mathcal{H}}\equiv N_{N,\mathcal{H}}$ of $N$), which can substantially inflate the number $q(N)$ of model parameters and further impact model instability through accumulated parameters.
 To better understand the effects of instability in the RBM structure,  Proposition \@ref(prp:prop2) next frames the general behavior of extreme probabilities in the joint RBM model \@ref(eq:RBM1) for  $(\boldsymbol X, \boldsymbol H)$  and the implied RBM data model \@ref(eq:RBM2) for  $\boldsymbol X$ alone.
@@
 Due to the marginalization steps in defining the distribution \@ref(eq:RBM2) of $\boldsymbol X$, note that $\REP_{\boldsymbol X}(\boldsymbol \theta_N)$ has no immediate analytical expression similar to that of  $\REP_{(\boldsymbol X, \boldsymbol H)}(\boldsymbol \theta_N)$.
-For clarity, recall also that S-instability \@ref(eq:Sun) in each model type refers to a respective divergence (i.e., $\lim_{N\to \infty} \REP_{(\boldsymbol X, \boldsymbol H)}(\boldsymbol \theta_N) /(N+N_{\mathcal{H}})=\infty$, $\lim_{N\to \infty} \REP(\boldsymbol \theta_N) /N=\infty$)
- upon scaling by the corresponding number of variables in a distribution.
+For clarity, recall also that S-instability \@ref(eq:Sun) in each model type refers to a respective divergence (i.e., $\lim_{N\to \infty} \REP_{(\boldsymbol X, \boldsymbol H)}(\boldsymbol \theta_N) /(N+N_{\mathcal{H}})=\infty$, $\lim_{N\to \infty} \REP(\boldsymbol \theta_N) /N=\infty$) upon scaling by the corresponding number of variables in a distribution.
 In the following, let $|\boldsymbol y|_1 = \sum_{i=1}^d |y_i|$ denote the L-1 norm of a generic vector $\boldsymbol y =(y_1,\ldots,y_d)$, $d \geq 1$.
 ```{proposition, label="prop2", echo=TRUE}
-Let $P_{\boldsymbol \theta_N}$ denote a RBM-based  model \@ref(eq:RBM2) for $N\geq 1$ visible variables $\boldsymbol X \equiv \boldsymbol X_N$ derived from $\tilde{P}_{\boldsymbol \theta_N}$ as the joint RBM distribution \@ref(eq:RBM1) of $(\boldsymbol X, \boldsymbol H)$ involving some number $N_{\mathcal{H}} \equiv  N_{N,\mathcal{H}}\geq 0$ of hidden variables $\boldsymbol H \equiv \boldsymbol H_N$ and parameters $\boldsymbol \theta_N \equiv (\boldsymbol \theta_N^{\mathcal{V}},\boldsymbol \theta_N^{\mathcal{H}}, \boldsymbol \theta_N^{\mathcal{VH}}) \in\mathbb{R}^{N}\times \mathbb{R}^{N_{\mathcal{H}}} \times  \mathbb{R}^{N N_{\mathcal{H}}}$.
+Let $P_{\boldsymbol \theta_N}$ denote a RBM-based data model \@ref(eq:RBM2) for $N\geq 1$ visible variables $\boldsymbol X \equiv \boldsymbol X_N$ derived from $\tilde{P}_{\boldsymbol \theta_N}$ as the joint RBM distribution \@ref(eq:RBM1) of $(\boldsymbol X, \boldsymbol H)$ involving some number $N_{\mathcal{H}} \equiv  N_{N,\mathcal{H}}\geq 0$ of hidden variables $\boldsymbol H \equiv \boldsymbol H_N$ and parameters $\boldsymbol \theta_N \equiv (\boldsymbol \theta_N^{\mathcal{V}},\boldsymbol \theta_N^{\mathcal{H}}, \boldsymbol \theta_N^{\mathcal{VH}}) \in\mathbb{R}^{N}\times \mathbb{R}^{N_{\mathcal{H}}} \times  \mathbb{R}^{N*N_{\mathcal{H}}}$.
 Then,
 \begin{enumerate}[(i)]
-\item the instability measure $\REP_{(\boldsymbol X)}(\boldsymbol \theta_N)$ for the marginal model $P_{\boldsymbol \theta_N}$ of  $\boldsymbol X$ satisfies
+\item the instability measure $\REP(\boldsymbol \theta_N)$ for the marginal model $P_{\boldsymbol \theta_N}$ of  $\boldsymbol X$ satisfies
     $$
-    \left| \REP_{(\boldsymbol X)}(\boldsymbol \theta_N)  - \elt\right| \leq    N_{\mathcal{H}}  \log 2
+    \left| \REP(\boldsymbol \theta_N)  - \elt\right| \leq    N_{\mathcal{H}}  \log 2
     $$
@@
 
-In more complicated network models involving further or deeper hidden layers, the same issues and causes of instability similarly exist, but are compounded by a greater number of model parameters.
+In more complicated graphical models involving further or deeper hidden layers, the same issues and causes of instability similarly exist, but are compounded by a greater number of model parameters.
 S-unstable joint models will similarly follow if the combined sizes of all parameters are too great relative to the total number of variables, while instability in the data model for visible variables will depend only on the  main or interaction parameters directly related to visibles and how their accumulated magnitude compares to the observation sample size $N$.
 
-# Statistical consequences of instability {#implications}
+# Statistical Consequences of Instability {#implications}
+
 Due to their extreme sensitivity in probability structure, S-instability in FOES models often translates to numerical complications, and in fact obstructions, in both simulation and statistical inference based on likelihoods.
 
-## Implications for simulation
+## Implications for Simulation {#mcmc}
 
@@
 $$
-\left| \frac{P_{\boldsymbol \theta_N} (X_i =x_1 | \boldsymbol x^{(1)}_{-i})}{P_{\boldsymbol \theta_N} (X_i = x_2| \boldsymbol x^{(2)}_{-i}) } \right|=\left| \frac{P_{\boldsymbol \theta_N} ( \boldsymbol x^{(1)} }{P_{\boldsymbol \theta_N} (  \boldsymbol x^{(2)} ) } \right|
+\left| \frac{P_{\boldsymbol \theta_N} (X_i =x_1 | \boldsymbol x^{(1)}_{-i})}{P_{\boldsymbol \theta_N} (X_i = x_2| \boldsymbol x^{(2)}_{-i}) } \right|=\left| \frac{P_{\boldsymbol \theta_N} ( \boldsymbol x^{(1)} )}{P_{\boldsymbol \theta_N} (  \boldsymbol x^{(2)} ) } \right|
 $$
-using that conditional probabilities are proportional to the joint probabilities and the fact that unstable models can have  unbounded log-probability ratios in one-component changes (Theorem \@ref(thm:instab-elpr).
+using that conditional probabilities are proportional to the joint probabilities and the fact that unstable models can have  unbounded log-probability ratios in one-component changes (Theorem \@ref(thm:instab-elpr)).
 This can hinder the ability of chain to effectively explore the sample space of the observations $\boldsymbol X$, as the chain may then mix poorly by moving rapidly to, and slowly away from, sections of the sample space surface.
-Because the numerical problem is unbounded ratios of probabilities under the joint model, similar complications can also arise for Metropolis-Hastings algorithms for MCMC.
+This mixing problem is due to the unstable stationary distribution (unbounded ratios of probabilities under the joint model), rather than in the particulars of the MCMC algorithm, similar complications can also arise for Metropolis-Hastings algorithms for MCMC.
 Hence, while an unstable FOES model $P_{\boldsymbol \theta_N}$ may be valid and technically open to simulation by MCMC, the aspect of instability can ruin numerical applications of MCMC.
 
-## Implications for maximum likelihood inference
+## Implications for Maximum Likelihood Inference
 
@@
 
-For a FOES model $P_{\boldsymbol \theta_N}(\boldsymbol x)$, $\boldsymbol x\in\mathcal{X}^N$, for $N$ observations $\boldsymbol X =(X_1,\ldots,X_n)$, define the likelihood function $\mathcal{L}(\boldsymbol \theta_N | \boldsymbol x) \equiv N^{-1}\log P_{\boldsymbol \theta_N }(\boldsymbol x)$ at $\boldsymbol \theta_N$  and the minimum probability ratio counterpart  $\mathcal{L}(\boldsymbol \theta_N | \boldsymbol x)  - N^{-1} \log \min_{\boldsymbol \theta_N}= N^{-1}\log[P_{\boldsymbol \theta_N }(\boldsymbol x) /\min_{\boldsymbol \theta_N}]$ where $\min_{\boldsymbol \theta_N} \equiv \min_{\boldsymbol y\in\mathcal{X}^N} P_{\boldsymbol \theta_N }(\boldsymbol y)]$.
-Also define $\max_{\boldsymbol \theta_N} \equiv \max_{\boldsymbol y\in\mathcal{X}^N} P_{\boldsymbol \theta_N }(\boldsymbol y)]$.
+For a FOES model $P_{\boldsymbol \theta_N}(\boldsymbol x)$, $\boldsymbol x\in\mathcal{X}^N$, for $N$ observations $\boldsymbol X =(X_1,\ldots,X_N)$, define the likelihood function $\mathcal{L}(\boldsymbol \theta_N | \boldsymbol x) \equiv N^{-1}\log P_{\boldsymbol \theta_N }(\boldsymbol x)$ at $\boldsymbol \theta_N$  and the minimum probability ratio counterpart  $\mathcal{L}(\boldsymbol \theta_N | \boldsymbol x)  - N^{-1} \log \min_{\boldsymbol \theta_N}= N^{-1}\log[P_{\boldsymbol \theta_N }(\boldsymbol x) /\min_{\boldsymbol \theta_N}]$ where $\min_{\boldsymbol \theta_N} \equiv \min_{\boldsymbol y\in\mathcal{X}^N} P_{\boldsymbol \theta_N }(\boldsymbol y)$.
+Also define $\max_{\boldsymbol \theta_N} \equiv \max_{\boldsymbol y\in\mathcal{X}^N} P_{\boldsymbol \theta_N }(\boldsymbol y)$.
 
 ```{theorem, label="something", echo=TRUE}
-Let $P_{\boldsymbol \theta_N}$, $N \geq 1$, denote an S-unstable FOES model sequence (i.e., $\lim_{N\to \infty}\REP(\boldsymbol \theta_N)/N=\infty$) which additionally satisfies model condition SPR.
+Let $P_{\boldsymbol \theta_N}$, $N \geq 1$, denote an S-unstable FOES model sequence (i.e., $\lim_{N\to \infty}\REP(\boldsymbol \theta_N)/N=\infty$) which additionally satisfies model condition PSR.
 Then,
@@
 ```
-This needs to be connected to results in Schweinberger who considered the one-parameter linear exponential case showing that, for positive $\boldsymbol \theta_N>0$, 
+@schweinberger2011instability [Corollary 1] considered the one-parameter linear exponential case showing that, for positive $\boldsymbol \theta_N>0$, 
 $$
@@
 holds converges to $1$ in expectation $E_{-\boldsymbol \theta_N}$.
-This result is a special case of Theorem \@ref(thm:something) and the LREP expansion \@ref(eq:UL) for the one parameter exponential model case.
+This result is a special case of Theorem \@ref(thm:something) and the LREP expansion \@ref(eq:UL) for the one parameter exponential model case and is related to a result in @handcock2003assessing.
 
@@
 This can cause numerical complications in likelihood maximization.
-The likelihood function knows what to do with an observed outcome $\boldsymbol x \equiv \argmax \mathcal{L}(\boldsymbol \theta_N)$ which is also the outcome with minimal probability under $P_{-\boldsymbol \theta_N}$.
-But unless the observed outcome $\boldsymbol x$ is at an extreme in the probability structures of   $P_{\boldsymbol \theta_N}, P_{-\boldsymbol \theta_N}$ (which are related) the likelihood function will be flat and optimization steps may shift around zero.
+If an observed outcome $\boldsymbol x \equiv \argmax \mathcal{L}(\boldsymbol \theta_N)$ which is also the outcome with minimal probability under $P_{-\boldsymbol \theta_N}$, then the likelihood function will have a clear peak and maximization straightforward.
+However, when the observed outcome $\boldsymbol x$ is at an extreme in the probability structures of $P_{\boldsymbol \theta_N}, P_{-\boldsymbol \theta_N}$ (which are related) the likelihood function will be flat and optimization steps may shift around zero.
 
-## Implications for Bayes inference
+## Implications for Bayes Inference
 
 Considering that changes in sign of parameters can cause all probability to be placed on disjoint sections of the sample space (Corollary \@ref(cor:sign)), then this will create numerical difficulties in Bayes inference based on MCMC.
-We can write the MH transition density to indicate this.
+Given observations $\boldsymbol X = (X_1,\ldots,X_N)$, switching $\boldsymbol theta_N$ from positive to negative or vice versa could dramatically change probabilities (model condition PSR and Corollary \@ref(cor:sign)).
+This would translate to numerical challenges similar to those presented in Section \@ref(mcmc), i.e., inability of the chain to effectively explore the sample space because of unbounded ratios of probabilities.
+We can write the MH acceptance probability to indicate this,
+$$
+\alpha\left(\boldsymbol x^{(1)} \mid \boldsymbol x^{(2)}\right)= \min\left\{1, 
+\frac{q(\boldsymbol x^{(2)} \mid \boldsymbol x^{(1)})}{q(\boldsymbol x^{(1)} \mid \boldsymbol x^{(2)})}\frac{P_{\boldsymbol \theta_N} ( \boldsymbol x^{(1)} )}{P_{\boldsymbol \theta_N} (  \boldsymbol x^{(2)} ) } \right\}
+$$
+for some proposal distribution $q(\cdot)$.
 
@@
 
-For a large class of models that covers a broad range of applications (including "deep learning''), we have developed a formal definition of instability in model probability structure and elucidated multiple consequences of instability.
+For a large class of models that covers a broad range of applications (including "deep learning"), we have developed a formal definition of instability in model probability structure and elucidated multiple consequences of instability.
 We have shown for FOES models that instability manifests through small changes in data leading to potentially large changes in probability as well as the potential to place all probability on certain modal subsections of the sample space, which potentially could be small.
-The FOES model class is quite broad and, particularly in developing FOES models for large data sets, some caution should be used in parameter specification to control effects of
-model instability.
+The FOES model class is quite broad and, particularly in developing FOES models for large data sets, some caution should be used in parameter specification to control effects of model instability due to the consequences of instability in both simulation and likelihood inference.

### 2017-11-02 writing/paper-note/index.Rmd
some changes in the note from nordman
forgot to push last time

@@
 ---
-title: "A note on the instability and degeneracy of deep learning models"
+title: "A Note on the Instability and Degeneracy of Deep Learning Models"
 author: 

### 2017-11-02 writing/paper-note/docs/note.tex
nordman's changes

--- 2017-09-30 writing/paper-note/docs/note.tex
@@
 \usepackage{amsmath,amssymb}
-\usepackage{graphicx,psfrag,epsf}
+%\usepackage{graphicx,psfrag,epsf}
+\usepackage{graphicx}
 \usepackage{enumerate}
@@
 \usepackage{url} % not crucial - just used below for the URL
-
+\setlength{\textheight}{9in}
 \usepackage{longtable,booktabs}
@@
 
-\usepackage{tikz, subfig, amsthm, multirow, float, enumerate}
-\usepackage{tikz-3dplot}
+\usepackage{tikz,  amsthm, multirow, float}
+%\usepackage{tikz-3dplot}
 \usetikzlibrary{arrows, shapes, positioning}
@@
 
+\usepackage{color}
+\newcommand{\red}[1]{{\color{red} #1}}
+\newcommand{\blue}[1]{{\color{blue} #1}}
 %%INCLUDED-AK
@@
 \newtheorem{lemma}{Lemma}
-\newtheorem{corollary}{Corollary}
 
@@
 \newcommand{\Gamt}{\Gamma_{N,2}(\boldsymbol \theta_N) }
-
 %\pdfminorversion=4
@@
 
-We consider the behavior, and the potential impropriety, sequences of
-discrete probability models built to incorporate observations of
-increasing sample size \(N\). Interest is in identifying instability in
-such models, which is roughly characterized by probabilities with
-extreme sensitivity to small changes in data configuration. The concept
-of instability was introduced in the field of statistical physics (i.e.,
-point processes) by Ruelle
-(\protect\hyperlink{ref-ruelle1999statistical}{1999}) and then further
-extended by Schweinberger Schweinberger
-(\protect\hyperlink{ref-schweinberger2011instability}{2011}) for a
-family of exponential models. At issue, models exhibiting instability
-are typically undesirable as these tend to provide poor representations
-of data or data-generation. As an example, such models can include
-near-degenerate distributions that assign all essential probability mass
-to only a subset of an overall sample space. The latter issue in
-connection to degeneracy has been recognized as concern in that dominant
-model outcomes may not resemble observed data (cf. Handcock
-\protect\hyperlink{ref-handcock2003assessing}{2003}). As a compounding
-issue, model instability often has direct negative impacts for
-statistical inference and computations based on likelihood functions.
-Namely, volatilities in probability structure can potentially hamper the
-numerical evaluations required for maximum likelihood estimation as well
-as other model-based simulations via Markov Chain Monte Carlo (MCMC).
-These reasons motivate our general study of instability for a broad
-class of probability models, described next.
-
-In the model framework, let \(\boldsymbol X_N = (X_1, \dots, X_N)\)
-denote a collection of discrete random variables with a finite sample
-space, \(\mathcal{X}^N\), represented as some \(N\)-fold Cartesian
-product. That is, \(\mathcal{X}\) with \(|\mathcal{X}| < \infty\)
-denotes the set of potential outcomes for each single variable \(X_i\),
-so that the product space \(\mathcal{X}^N\) corresponds to values for
-the variables \(\boldsymbol X_N=(X_1,\ldots,X_N)\). For each \(N\), let
-\(P_{\boldsymbol \theta_N}\) denote a probability model on
-\(\mathcal{X}^N\), under which
-\(P_{\boldsymbol \theta_N}(x_1, \dots, x_N) > 0\) is the probability of
-the data outcome \((x_1, \dots, x_N) \in \mathcal{X}^N\). In this, we
-assume that the model support of \(P_{\boldsymbol \theta_N}\) is the
-sample space \(\mathcal{X}^N\). This framework produces probability
-models \(P_{\boldsymbol \theta_N}\), indexed by a generic sequence of
-parameters \(\boldsymbol \theta_N\), to describe data
-\(\boldsymbol X_N\) of any given sample size \(N \geq 1\). For
-simplicity, we will refer to this distributional class as \emph{Finite
-Outcome Everywhere Supported (FOES)} models in the following. The
-dimension and structure of such parameters are generic, without
-restriction, though natural cases will be seen to include those where
-\(\boldsymbol \theta_N \in \mathbb{R}^{q(N)}\) for some arbitrary
-integer-valued function \(q(\cdot) \geq 1\).
-
-Section \ref{examples} provides some examples of FOES models encountered
-in graph/network analysis and machine learning (i.e., deep learning
-models). These are used as references for later illustrations. Section
-\ref{instability-results} then establishes several formal results for
-FOES models with regard to instability. Schweinberger
-(\protect\hyperlink{ref-schweinberger2011instability}{2011}) originally
-developed instability results framed for, or specific to, a certain
-class of discrete exponential models. For similar exponential models
-with random networks, Handcock
-(\protect\hyperlink{ref-handcock2003assessing}{2003}) studied model
-degeneracy, where a probability model places near complete mass on modes
-and may thereby narrow the feasible model outcomes. As findings here and
-from Schweinberger
-(\protect\hyperlink{ref-schweinberger2011instability}{2011}) suggest,
-model instability and degeneracy may also be related by viewing
-degeneracy as an extreme, or limiting form, of instability. Our main
-results establish a broad characterization of model instability,
-appropriate across the whole FOES model class, that incorporates results
-of Schweinberger
-(\protect\hyperlink{ref-schweinberger2011instability}{2011}) as a
-special case. We prescribe a general and simple condition for
-identifying instability in a FOES model sequence, which quantifies
-whether certain maximal probabilities in a FOES model are too extreme
-relative to the sample size \(N\). When fulfilled, the probability
-structure of a FOES model is shown to exhibit extreme sensitivity, with
-probability assignments possessing extreme peaks and troughs across
-nearly identical outcomes. As the measure of model instability
-increases, probabilities from an unstable FOES model additionally
-increase in volatility and provably slide into degeneracy. Section
-\ref{implications} then emphasizes the implications of such model
-instability, showing that such impropriety can be expected to
-numerically hinder maximum likelihood estimation and MCMC-based
-simulations. As one potential remedy, suggestions are given or
-constraining model parameterizations and avoid the most problematic
-regions of the parameter space. Proofs of the main results appear in
-Appendix \ref{appendix-instab}.
+We consider the behavior, and the potential impropriety, sequences of discrete probability models built to incorporate observations of increasing sample size $N$.
+Interest is in identifying  instability in such models, which is roughly characterized by probabilities with extreme sensitivity to small changes in data configuration.
+The concept of instability was introduced in the field of statistical physics (i.e., point processes) by Ruelle (\protect\hyperlink{ref-ruelle1999statistical}{1999}) and then further extended by Schweinberger
+(\protect\hyperlink{ref-schweinberger2011instability}{2011})   for a family of exponential  models.
+At issue, models exhibiting instability are typically undesirable  as these tend to provide poor representations of data or data-generation.
+ As an example, such models  can include near-degenerate distributions that assign all essential probability mass to only a subset of
+ an overall sample space.
+The latter issue in connection to degeneracy has been recognized as concern  in that dominant model outcomes  may not resemble   observed data  (cf.~Handcock, \hyperlink{ref-handcock2003assessing}{2003}).
+As a compounding issue, model instability often has direct negative impacts for statistical inference and computations based on likelihood functions.
+Namely, volatilities  in probability structure  can potentially
+  hamper the numerical evaluations required for maximum likelihood estimation as well as other model-based simulations via Markov Chain Monte Carlo (MCMC).
+  These reasons motivate our general study of instability for a broad class of probability models, described next.
+
+    In the model framework,
+ let $  \boldsymbol X_N =(X_1, \dots, X_N)$ denote a collection of discrete random variables with a finite sample space, $\mathcal{X}^N$, represented as some $N$-fold Cartesian  product.
+That is, $\mathcal{X}$ with $|\mathcal{X}| < \infty$ denotes the set of potential outcomes for each single variable $X_i$, so that the product space $\mathcal{X}^N$ corresponds to values for the variables  $\boldsymbol X_N=(X_1,\ldots,X_N)$.
+For each $N$, let $P_{\boldsymbol \theta_N}$ denote a probability model on $\mathcal{X}^N$, under which $P_{\boldsymbol \theta_N}(x_1, \dots, x_N) > 0$ is the probability of the data outcome $(x_1, \dots, x_N) \in \mathcal{X}^N$.
+In this, we assume that the model support of $P_{\boldsymbol \theta_N}$ is the sample space $\mathcal{X}^N$.
+This framework produces probability models $P_{\boldsymbol \theta_N}$, indexed by a generic sequence of parameters $\boldsymbol \theta_N$, to describe data $\boldsymbol X_N$ of any given sample size $N \geq 1$.
+For simplicity, we will refer to this distributional class as {\it Finite Outcome Everywhere Supported} (FOES) models in the following.
+The dimension and structure of such parameters are generic, without restriction, though natural cases will be seen to include those where $\boldsymbol \theta_N \in \mathbb{R}^{q(N)}$ for some arbitrary integer-valued function $q(\cdot) \geq 1$.
+
+Section \ref{examples}   provides some examples of FOES models encountered in graph/network analysis and machine learning (i.e., deep learning models).
+These are used as references for later illustrations.
+Section \ref{instability-results} then  establishes several formal results for  FOES models with regard to instability.
+Schweinberger
+(\protect\hyperlink{ref-schweinberger2011instability}{2011}) originally developed instability results framed for, or specific to, a certain class of discrete exponential  models.
+For similar exponential models with random networks,
+Handcock (\protect\hyperlink{ref-handcock2003assessing}{2003})   studied model degeneracy, where a probability model places near complete mass on  modes and may thereby narrow the feasible
+  model outcomes.    As findings here and from  Schweinberger
+(\protect\hyperlink{ref-schweinberger2011instability}{2011}) suggest, model instability and degeneracy may also be related by viewing degeneracy as an extreme, or limiting form, of instability.
+Our main results establish a broad characterization of model instability, appropriate across the whole FOES model class,
+ that incorporates results of
+ Schweinberger
+(\protect\hyperlink{ref-schweinberger2011instability}{2011}) as a special case.
+ We prescribe a general and simple condition  for identifying instability in a FOES model sequence,   which  quantifies whether certain maximal probabilities in a FOES model  are too extreme relative to the sample size $N$.
+When fulfilled, the probability structure of a FOES model is shown to exhibit extreme sensitivity, with probability assignments possessing extreme  peaks and troughs across nearly identical outcomes.
+As the measure of model instability increases,
+probabilities from an unstable FOES model  additionally increase in volatility and provably slide into  degeneracy.
+Section \ref{implications} then emphasizes the implications of such model instability, showing that such impropriety
+can be expected to numerically hinder maximum likelihood estimation and   MCMC-based simulations.  As one potential remedy, suggestions are given
+ for constraining model parameterizations and avoid the most problematic regions of the parameter space. Proofs of the main results appear in Appendix \ref{appendix-instab}.
 
@@
 
-Many model families fall under the umbrella of FOES models. For
-illustration, this section presents three specific examples of FOES
-models, including models with deep architectures.
-
-\subsection{Discrete exponential family
-models}\label{discrete-exponential-family-models}
-
-For random variables
-\(\boldsymbol X \equiv\boldsymbol X_N= (X_1, \dots, X_N)\) with sample
-space \(\mathcal{X}^N\), \(|\mathcal{X}| < \infty\), consider an
-exponential family model for \(\boldsymbol X\) with probability mass
-function given by
+Many model families fall under the umbrella of FOES models. For illustration, this section presents three specific examples of FOES models, including models with deep architectures.
+
+\subsection{Discrete exponential family models}
+\label{discrete-exponential-family-models}
+
+For  random variables $\boldsymbol X \equiv\boldsymbol X_N= (X_1, \dots, X_N)$ with sample space $\mathcal{X}^N$, $|\mathcal{X}| < \infty$, consider an exponential family model for $\boldsymbol X$ with probability mass function given by
 \begin{equation}
-\label{eq:expo}
+\label{eqn:expo}
 p_{N, \boldsymbol \theta}(\boldsymbol x) = \exp\left[\boldsymbol\eta^T(\boldsymbol \theta) \boldsymbol g_N(\boldsymbol x) - \psi(\boldsymbol \theta)\right], \qquad \boldsymbol x \in \mathcal{X}^N,
 \end{equation}
-depending on parameter vector
-\(\boldsymbol \theta \in \Theta_N \subset \mathbb{R}^{k}\) and natural
-parameter function
-\(\boldsymbol \eta : \mathbb{R}^k \mapsto \mathbb{R}^L\) with fixed
-positive integers \(k\) and \(L\) denoting their dimensions. Above,
-\(\boldsymbol g_N : \mathcal{X}^N \mapsto \mathbb{R}^L\) is a vector of
-sufficient statistics, while \[
+depending on parameter vector $\boldsymbol \theta \in \Theta_N \subset \mathbb{R}^{k}$ and natural parameter function $\boldsymbol \eta : \mathbb{R}^k \mapsto \mathbb{R}^L$ with fixed positive integers $k$ and $L$ denoting their dimensions.
+Above, $\boldsymbol g_N : \mathcal{X}^N \mapsto \mathbb{R}^L$ is a vector of sufficient statistics, while
+$$
 \psi(\boldsymbol \theta) = \log \sum\limits_{\boldsymbol x \in \mathcal{X}^N}\exp\left[\boldsymbol \eta^T(\boldsymbol \theta) \boldsymbol g_N(\boldsymbol x) \right], \qquad \boldsymbol \theta \in \Theta_N\equiv \{\boldsymbol \theta \in \mathbb{R}^k : \psi(\boldsymbol \theta) < \infty \},
-\] denotes the normalizing function with parameter space \(\Theta_N\).
-The natural parameter function \(\eta (\boldsymbol \theta)\) has a
-linear form (i.e.,
-\(\eta (\boldsymbol \theta)= \bm{A} \boldsymbol \theta\) for a given
-\(L \times k\) matrix \(\bm{A}\)) in many common model formulaitons,
-though may also be nonlinear (e.g., curved exponential families). In the
-linear case, \(\eta (\boldsymbol \theta) = \boldsymbol \theta\) may be
-generally assumed in the exponential parameterization with a minor
-modification to the definition of sufficient statistics
-\(\boldsymbol g_N(\boldsymbol x)\).
-
-Such discrete exponential family models are special cases of the FOES
-models, as seen by defining
-\(P_{\boldsymbol \theta_N}(\boldsymbol x)\equiv p_{N,\boldsymbol \theta_N}(\boldsymbol x)> 0\),
-\(\boldsymbol x \in \mathcal{X}^N\), based on \eqref{eq:expo} and a
-parameter sequence
-\(\boldsymbol \theta_N \in \Theta_N \subset \mathbb{R}^k\). For example,
-if observations \(\boldsymbol X = (X_1,\ldots,X_N)\) correspond to \(N\)
-independent and identically distributed Bernoulli random variables, each
-indicating a binary \(0\)-\(1\) outcome, the resulting probabilities
-have exponential form \eqref{eq:expo} given by
+$$
+denotes the normalizing function with   parameter space $\Theta_N$.
+The natural parameter function $\eta (\boldsymbol \theta) $  has a linear form (i.e., $\eta (\boldsymbol \theta)= \bm{A} \boldsymbol \theta$ for a given $L \times k$ matrix $\bm{A}$) in many common model formulaitons, though may also be nonlinear (e.g., curved exponential families).
+In the linear case, $\eta (\boldsymbol \theta) =\boldsymbol \theta$ may be generally assumed in the exponential parameterization with a minor modification to the definition of sufficient statistics $\boldsymbol g_N(\boldsymbol x)$.
+
+Such discrete exponential family models are special cases of the FOES models, as seen by  defining $P_{\boldsymbol \theta_N}(\boldsymbol x)\equiv p_{N,\boldsymbol \theta_N}(\boldsymbol x)> 0$,  $\boldsymbol x \in \mathcal{X}^N$, based on (\ref{eqn:expo}) and a parameter sequence $\boldsymbol \theta_N \in \Theta_N \subset \mathbb{R}^k$.
+For example, if observations $\boldsymbol X=(X_1,\ldots,X_N)$ correspond to $N$ independent and identically distributed Bernoulli random variables, each indicating a binary $0$-$1$ outcome,  the resulting probabilities have exponential form (\ref{eqn:expo})  given by
 \begin{equation}
-\label{eq:mod1}
-P_{\boldsymbol \theta_N}(\boldsymbol x) \propto
- \exp\left[\boldsymbol \theta_N \sum_{i=1}^N x_i\right], \qquad \boldsymbol x=(x_1,\ldots,x_N) \in\{0,1\}^N, 
- \end{equation}
-with sufficient statistic
-\(\boldsymbol g_N(\boldsymbol x)\equiv \sum_{i=1}^N x_i\) and ``log odds
-ratio'' parameter
-\(\boldsymbol \theta_N \equiv \log[ P_{\boldsymbol \theta_N}(X_i=1)/P_{\boldsymbol \theta_N}(X_i=0) ] \in \mathbb{R}\).
-More generally, supposing \(\boldsymbol X =(X_1,\ldots,X_N)\) represent
-\(N\) independent trials, each assuming an outcome \(\{1,\ldots,k\}\)
-among \(k\) possibilities (e.g., a die roll), a multinomial distribution
-is given by
+\label{eqn:mod1} P_{\boldsymbol \theta_N}(\boldsymbol x) \propto
+ \exp\left[\boldsymbol \theta_N \sum_{i=1}^N x_i\right], \qquad \boldsymbol x=(x_1,\ldots,x_N) \in\{0,1\}^N, \end{equation} with sufficient statistic $\boldsymbol g_N(\boldsymbol x)\equiv \sum_{i=1}^N  x_i$ and ``log odds ratio" parameter $\boldsymbol \theta_N \equiv  \log[ P_{\boldsymbol \theta_N}(X_i=1)/P_{\boldsymbol \theta_N}(X_i=0) ] \in \mathbb{R}$.
+More generally,
+ supposing $\boldsymbol X =(X_1,\ldots,X_N)$ represent  $N$
+independent trials, each assuming an outcome $\{1,\ldots,k\}$ among $k$ possibilities (e.g., a die roll), a multinomial distribution is given by
    \begin{equation}
-\label{eq:mod11}
-P_{\boldsymbol \theta_N}(\boldsymbol x) \propto  \exp\left[  \boldsymbol \theta_{N}^T g_N(\boldsymbol x)   \right] =
-\exp\left[ \sum_{j=1}^k {\theta_{j,N}} \sum_{i=1}^N \mathbb{I}(x_i=j) \right], \qquad \boldsymbol x  \in\{1,\ldots,k\}^N, 
-\end{equation}
-with sufficient statistic \(\boldsymbol g_N(\boldsymbol x)\) involving a
-count \(\sum_{i=1}^N \mathbb{I}(x_i=j)\) for each outcome
-\(j \in \{1,\ldots,k\}\), where \(\mathbb{I}(\cdot)\) denotes the
-indicator function, and parameters
-\(\boldsymbol \theta_N=(\theta_{1,N},\ldots,\theta_{k,N})\in\mathbb{R}^k\)
-defining log-probability ratios
-\(\theta_{i,N}-\theta_{j,N} =\log [P_{\boldsymbol \theta_N}(X_1=i)/P_{\boldsymbol \theta_N}(X_1=j)]\).
-In addition to such standard models for discrete independent data,
-exponential models of FOES type commonly arise with dependent spatial
-data (Besag \protect\hyperlink{ref-besag1974spatial}{1974}) and
-network/relational data (Wasserman and Faust
-\protect\hyperlink{ref-wasserman1994social}{1994}; Handcock
-\protect\hyperlink{ref-handcock2003assessing}{2003}). For a random graph
-or network with, say, \(n\) nodes, consider \(N={n \choose 2}\) random
-edges where the \(i\)th edge is associated with a pair of nodes
-\(s_i \equiv \{v_i,u_i\}\) and a binary variable \(X_i\in\{0,1\}\)
-indicating presence/absence of an edge among the node pair \(s_i\),
-\(i=1,\ldots,N\). Here the length \(N\) of the edge variable sequence
-\(\boldsymbol X = (X_1,\ldots,X_N)\) increases as a function of node
-number \(n\) and corresponding exponential models often incorporate
-graph topographical features derived from \(\boldsymbol X\). As an
-example, consider a graph model of exponential/FOES form prescribed by
+   \label{eqn:mod11} P_{\boldsymbol \theta_N}(\boldsymbol x) \propto  \exp\left[  \boldsymbol \theta_{N}^T g_N(\boldsymbol x)   \right]=
+  \exp\left[ \sum_{j=1}^k {\theta_{j,N}} \sum_{i=1}^N \mathbb{I}(x_i=j) \right], \qquad \boldsymbol x  \in\{1,\ldots,k\}^N, \end{equation}
+with sufficient statistic $\boldsymbol g_N(\boldsymbol x)$ involving a count $\sum_{i=1}^N \mathbb{I}(x_i=j)$  for each outcome $j\in\{1,\ldots,k\}$, where $\mathbb{I}(\cdot)$ denotes the indicator function,   and parameters $\boldsymbol \theta_N=(\theta_{1,N},\ldots,\theta_{k,N})\in\mathbb{R}^k$ defining log-probability ratios
+$\theta_{i,N}-\theta_{j,N} =\log [P_{\boldsymbol \theta_N}(X_1=i)/P_{\boldsymbol \theta_N}(X_1=j)]$.  In addition to such standard models for discrete independent data,
+exponential models of FOES type commonly arise with
+ dependent spatial data   (Besag
+\protect\hyperlink{ref-besag1974spatial}{1974}) and network/relational data (Wasserman
+and Faust \protect\hyperlink{ref-wasserman1994social}{1994}; Handcock
+\protect\hyperlink{ref-handcock2003assessing}{2003}). For a random graph  or network  with, say, $n$ nodes,  consider $N={n \choose 2}$ random edges
+ where the $i$th edge is associated with a pair of nodes $s_i \equiv \{v_i,u_i\}$  and a binary variable $X_i\in\{0,1\}$  indicating   presence/absence of an edge among the   node pair $s_i$, $i=1,\ldots,N$.
+Here the length $N$ of the edge variable sequence $\boldsymbol X =(X_1,\ldots,X_N)$ increases as a function of node number $n$ and corresponding
+ exponential models often incorporate graph topographical features derived from $\boldsymbol X$. As an example,  consider a   graph model of exponential/FOES form prescribed by
  \begin{equation}
-\label{eq:mod2}
-P_{\boldsymbol \theta_N}(\boldsymbol x) \propto
- \exp\left[\sum_{j=1}^3 \theta_{j,N} g_{j,N}(\boldsymbol x)\right], \quad\qquad \boldsymbol x=(x_1,\ldots,x_N)  \in\{0,1\}^N,
-\end{equation}
+\label{eqn:mod2} P_{\boldsymbol \theta_N}(\boldsymbol x) \propto
+ \exp\left[\sum_{j=1}^3 \theta_{j,N} g_{j,N}(\boldsymbol x)\right], \quad\qquad \boldsymbol x=(x_1,\ldots,x_N)  \in\{0,1\}^N,\end{equation}
   \[
@@
 x_i x_j x_\ell,
-\] involving the numbers of edges, 2-stars and triangles among an
-outcome \(\boldsymbol x\) given by \(g_{1,N}(\boldsymbol x)\),
-\(g_{2,N}(\boldsymbol x)\) and \(g_{3,N}(\boldsymbol x)\), respectively,
-along with \(k=3\) real parameters
-\(\boldsymbol \theta_N \equiv (\theta_{1,N},\theta_{2,N},\theta_{3,N})\).
-For this network model \eqref{eq:mod2} in particular, as well as for more
-general models of form \eqref{eq:expo}, Schweinberger
-(\protect\hyperlink{ref-schweinberger2011instability}{2011}) considered
-instability in such exponential models with sequences of fixed
-parameters
-\(\boldsymbol \theta_N = (\theta_1,\ldots,\theta_k)\in\mathbb{R}^k\),
-\(N \geq 1\), of fixed dimension \(k\).
-
-For model sequences
-\(P_{\boldsymbol \theta_N}(\boldsymbol x)\equiv p_{N,\boldsymbol \theta_N}(\boldsymbol x)\)
-of the exponential type \eqref{eq:expo}, such as those in
-\eqref{eq:mod1}-\eqref{eq:mod2}, note that the dimension \(k\) of the
-parameter \(\boldsymbol \theta_N\in\Theta \subset \mathbb{R}^k\)
-necessarily remains the same for all sample sizes \(N \geq 1\) as the
-form of the natural parameter function \(\eta(\cdot)\) in \eqref{eq:expo}
-and the number of sufficient statistics
-\(\boldsymbol g_{N}(\boldsymbol x)\) do not depend on \(N\).
-Consequently, \(\boldsymbol \theta_N\) lies in a parameter space of
-fixed Euclidean dimension \(k\). However, this aspect need not be true
-for other types of FOES models considered in Sections
-\ref{restricted-boltzmann-machines} - \ref{deep-learning}, where instead
-the numbers of parameters and sufficient statistics commonly increase
-with the sample size \(N\).
+  \]
+ involving the numbers of edges, 2-stars and triangles among an outcome $\boldsymbol x$ given by   $ g_{1,N}(\boldsymbol x)$,   $ g_{2,N}(\boldsymbol x)$  and  $ g_{3,N}(\boldsymbol x)$, respectively, along   with $k=3$ real parameters $\boldsymbol \theta_N \equiv (\theta_{1,N},\theta_{2,N},\theta_{3,N})$.
+For this network model (\ref{eqn:mod2}) in particular, as well as for more general models of form (\ref{eqn:expo}),  Schweinberger
+(\protect\hyperlink{ref-schweinberger2011instability}{2011}) considered instability in such exponential models   with sequences of fixed parameters $\boldsymbol \theta_N= (\theta_1,\ldots,\theta_k)\in\mathbb{R}^k$, $N \geq 1$, of fixed dimension $k$.
+
+  For model sequences $P_{\boldsymbol \theta_N}(\boldsymbol x)\equiv p_{N,\boldsymbol \theta_N}(\boldsymbol x)$
+  of the exponential type (\ref{eqn:expo}), such as those in (\ref{eqn:mod1})-(\ref{eqn:mod2}),
+   note that the dimension $k$ of the parameter $\boldsymbol \theta_N\in\Theta \subset \mathbb{R}^k$  necessarily remains the same for all sample sizes $N \geq 1$ as the form of the natural parameter function $\eta(\cdot)$ in (\ref{eqn:expo}) and the number of sufficient statistics $\boldsymbol g_{N}(\boldsymbol x)$ do not depend on $N$.
+Consequently,
+   $\boldsymbol \theta_N$ lies in a parameter space of fixed Euclidean dimension $k$.
+However, this aspect need not be true for other types of FOES models considered in Sections \ref{restricted-boltzmann-machines}-\ref{deep-learning}, where instead the numbers of parameters and sufficient statistics commonly increase with the sample size $N$.
+
+%\subsection{Discrete exponential family
+%models}\label{discrete-exponential-family-models}
 
@@
 
-A restricted Boltzmann machine (RBM) is an undirected graphical model
-specified for discrete or continuous random variables, with binary
-variables being most common (cf. Smolensky
-\protect\hyperlink{ref-smolensky1986information}{1986}). A RBM
-architecture has two layers, hidden (\(\mathcal{H}\)) and visible
-(\(\mathcal{V}\)), with conditional independence within each layer. Let
-\(\boldsymbol X = (X_1,\ldots,X_N)\) denote the \(N\) random variables
-for visibles with support \(\mathcal{X}^N\) and
-\(\boldsymbol H = (H_1,\ldots,H_{N_\mathcal{H}})\) denote the
-\(N_\mathcal{H}\) random variables for hiddens with support
-\(\mathcal{X}^{N_\mathcal{H}}\) where \(\mathcal{X} = \{-1,1\}\). For
-parameters
-\(\boldsymbol \theta_N^{\mathcal{H}} \in \mathbb{R}^{N_\mathcal{H}}\),
-\(\boldsymbol \theta_N^{\mathcal{V}}\in \mathbb{R}^N\), and
-\(\boldsymbol \theta_N^{\mathcal{HV}}\) as a real matrix with dimension
-\(N_\mathcal{H} \times N\), the RBM model for
-\(\tilde{\boldsymbol X}=(\boldsymbol X,\boldsymbol H)\) has the joint
-probability mass function
+A restricted Boltzmann machine (RBM) is an undirected graphical model specified for discrete or continuous random variables, with binary variables being most common (cf.~Smolensky
+\protect\hyperlink{ref-smolensky1986information}{1986}).
+A RBM architecture has two layers, hidden ($\mathcal{H}$) and visible ($\mathcal{V}$), with conditional independence within each layer.
+Let $\boldsymbol X=(X_1,\ldots,X_N)$ denote the $N$ random variables for visibles with support $\mathcal{X}^N$ and $\boldsymbol H=(H_1,\ldots,H_{N_\mathcal{H}})$ denote the $N_\mathcal{H}$ random variables for hiddens with support $\mathcal{X}^{N_\mathcal{H}}$ where  $\mathcal{X} = \{-1,1\}$.
+For parameters $\boldsymbol \theta_N^{\mathcal{H}} \in \mathbb{R}^{N_\mathcal{H}}$, $\boldsymbol \theta_N^{\mathcal{V}}\in \mathbb{R}^N$, and $\boldsymbol \theta_N^{\mathcal{HV}}$ as a real matrix with dimension $N_\mathcal{H} \times N$, the RBM model for $\tilde{\boldsymbol X}=(\boldsymbol X,\boldsymbol H)$  has the joint probability mass function
 \begin{equation}
-\label{eq:RBM1}
+\label{eqn:RBM1}
 \tilde{P}_{\boldsymbol \theta_N} (\tilde{\boldsymbol x}) = \exp\left[ (\boldsymbol \theta_N^{\mathcal{H}})^T \boldsymbol h + \boldsymbol (\boldsymbol \theta_N^{\mathcal{V}})^T \boldsymbol x + \boldsymbol h^T  \boldsymbol\theta_N^{\mathcal{HV}} \boldsymbol x - \psi(\boldsymbol \theta_N)\right], \quad \tilde{\boldsymbol x} = (\boldsymbol x, \boldsymbol h) \in \{\pm 1\}^{N+N_\mathcal{H}}
 \end{equation}
-with normalizing function \[
+with normalizing function
+$$
 \psi(\boldsymbol \theta_N) = \log \sum_{\tilde{\boldsymbol x} \in \{\pm 1\}^{N+N_H} } \exp\left[ (\boldsymbol \theta_N^{\mathcal{H}})^T \boldsymbol h + \boldsymbol (\boldsymbol \theta_N^{\mathcal{V}})^T \boldsymbol x + \boldsymbol h^T  \boldsymbol\theta_N^{\mathcal{HV}} \boldsymbol x\right].
-\] Let
-\(\boldsymbol \theta_N = (\boldsymbol \theta_N^{\mathcal{H}}, \boldsymbol \theta_N^{\mathcal{V}}, \boldsymbol\theta_N^{\mathcal{HV}} ) \in \Theta_N \equiv \mathbb{R}^{q(N)}\),
-with \(q(N) = N + N_\mathcal{H} + N*N_\mathcal{H}\), denote the
-parameter vector for the RBM, as indexed by the number \(N\) of visible
-random variables (which may differ from the actual lengths of these
-parameter vectors). The probability mass function for the visible
-variables \(\boldsymbol X = (X_1, \dots, X_N)\) follows from
-marginalizing the joint specification to yield
-\begin{equation}
-\label{eq:RBM2}
+$$
+ Let $\boldsymbol \theta_N = (\boldsymbol \theta_N^{\mathcal{H}}, \boldsymbol \theta_N^{\mathcal{V}}, \boldsymbol\theta_N^{\mathcal{HV}} ) \in \Theta_N \equiv \mathbb{R}^{q(N)}$, with $q(N) = N + N_\mathcal{H} + N*N_\mathcal{H}$, denote the parameter vector  for the RBM,  as indexed by the number $N$
+of visible random variables (which may differ from the actual lengths of these parameter vectors).
+The probability mass function for the visible variables $\boldsymbol X =(X_1, \dots, X_N)$ follows from marginalizing the joint specification to yield
+\begin{equation}\label{eqn:RBM2}
 P_{\boldsymbol \theta_N} (\boldsymbol x) = \sum\limits_{\boldsymbol h \in \{\pm 1\}^{N_{\mathcal{H}}}} \tilde{P}_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h), \qquad \boldsymbol x \in \{\pm 1\}^{N}\equiv \mathcal{X}^N.
 \end{equation}
-Here the baseline model \eqref{eq:RBM1} for hidden/visible variables is a
-linear exponential one in sufficient statistics
-\((\tilde{\boldsymbol X}, \boldsymbol X^T\boldsymbol H)\) using
-\(\tilde{\boldsymbol X}=(\boldsymbol X,\boldsymbol H)\) from
-\eqref{eq:RBM1}, but the form differs from the previous exponential models
-in \eqref{eq:expo} in that the lengths of parameters
-\(\boldsymbol \theta_N\) and statistics
-\((\tilde{\boldsymbol X}, \boldsymbol X^T\boldsymbol H)\) increase to
-incorporate more visible variables. That is, in contrast to
-\eqref{eq:expo}, the natural parameter function involved in the RBM model
-\eqref{eq:RBM1}, as the identity mapping of the parameters
-\(\boldsymbol \theta_N\in\mathbb{R}^{q(N)}\), naturally grows in
-dimension \(q(N)\to \infty\) to accommodate visible variables
-\(X_1, \dots, X_N\) of increasing sample size \(N\to \infty\).
-Additionally, one may further arbitrarily choose the number
-\(N_\mathcal{H}\) of hidden variables \(\boldsymbol H\) in the joint RMB
-model \eqref{eq:RBM1} to define a marginal model \eqref{eq:RBM2} for the
-\(N\) visible variables \(\boldsymbol X\), and the number
-\(N_\mathcal{H}\) of hiddens may also potentially increase with \(N\).
-Because \(|\mathcal{X}| = 2\) and
-\(P_{\boldsymbol \theta_N}(\boldsymbol x) > 0\) for all
-\(\boldsymbol x \in \mathcal{X}^N\), the RBM specification \eqref{eq:RBM2}
-for visibles \(\boldsymbol\) corresponds to a FOES model, while the
-joint distribution \eqref{eq:RBM1} for \((\boldsymbol X, \boldsymbol H)\)
-is also a FOES model. As this example also indicates, any model formed
-by marginalizing a base FOES model class, such as the RBM joint
-specification \eqref{eq:RBM1}, is again a FOES model.
+Here the baseline model (\ref{eqn:RBM1}) for hidden/visible variables is a linear exponential one in sufficient statistics  $(\tilde{\boldsymbol X},  \boldsymbol X^T\boldsymbol H)$ using $\tilde{\boldsymbol X}=(\boldsymbol X,\boldsymbol H)$ from   (\ref{eqn:RBM1}), but the form differs from the previous exponential models
+in (\ref{eqn:expo}) in that the lengths of parameters $\boldsymbol \theta_N$ and   statistics  $(\tilde{\boldsymbol X},  \boldsymbol X^T\boldsymbol H)$ increase to incorporate more visible variables.
+That is, in contrast to (\ref{eqn:expo}), the natural parameter function involved in the RBM model (\ref{eqn:RBM1}), as the identity mapping of the parameters $\boldsymbol \theta_N\in\mathbb{R}^{q(N)}$,    naturally grows in  dimension  $q(N)\to \infty$   to accommodate visible variables $X_1, \dots, X_N$ of increasing sample size $N\to \infty$.
+Additionally, one may further arbitrarily choose the number $N_\mathcal{H}$ of hidden variables $\boldsymbol H$ in the joint RMB model (\ref{eqn:RBM1}) to define a marginal model (\ref{eqn:RBM2}) for the $N$ visible variables $\boldsymbol X$, and the number $N_\mathcal{H}$ of hiddens may also potentially   increase with $N$.
+Because   $|\mathcal{X}| = 2$ and $P_{\boldsymbol \theta_N}(\boldsymbol x) > 0$ for all $\boldsymbol x \in \mathcal{X}^N$, the RBM  specification (\ref{eqn:RBM2}) for visibles $\boldsymbol$ corresponds to a FOES model, while the joint distribution (\ref{eqn:RBM1}) for $(\boldsymbol X, \boldsymbol H)$ is also a FOES model.
+As this example also indicates,  any model formed by marginalizing a base FOES model class, such as the RBM joint specification (\ref{eqn:RBM1}), is again a FOES model.
 
@@
 \protect\hyperlink{ref-salakhutdinov2009deep}{2009}) and a deep belief
-network Hinton, Osindero, and Teh
-(\protect\hyperlink{ref-hinton2006fast}{2006}){]}. Let \(M\) denote the
+network (Hinton, Osindero, and Teh
+\protect\hyperlink{ref-hinton2006fast}{2006}). Let \(M\) denote the
 number of hidden layers included in the model and let
@@
 model. Each variable outcome will again lie in
-\(\mathcal{X} = \{-1,1\}\).
-
-\textbf{Deep Boltzmann machine (DBM).} The DBM class of models maintains
+\(\mathcal{X} = \{-1,1\}\).\\
+
+\noindent  \textbf{Deep Boltzmann machine (DBM).} The DBM class of models maintains
 conditional independence within all layers in the model by stacking RBM
@@
 \psi(\boldsymbol \theta_N) = \log \sum\limits_{\tilde{\boldsymbol x} \in \mathcal{X}^{N_{(H,1)} + \cdots + N_{(H,M)} +N}} \exp\left[ \sum\limits_{i = 1}^M\boldsymbol \alpha^{(i)T} \boldsymbol h^{(i)} + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^{(1)T} \Gamma^{(0)} \boldsymbol x + \sum\limits_{i = 1}^{M - 1} \boldsymbol h^{(i)T} \Gamma^{(i)} \boldsymbol h^{(i + 1)}\right],
-\] is the normalizing function for
-\(\boldsymbol \theta_N = (\boldsymbol \alpha^{(1)}, \dots, \boldsymbol \alpha^{(M)}, \boldsymbol \beta,\Gamma^{(0)}, \dots, \Gamma^{(M - 1)}) \in \Theta_N \subset \mathbb{R}^{q(N)}\),
-consisting of model parameters \(\boldsymbol \beta \in \mathbb{R}^N\),
+\] is the normalizing function for $\boldsymbol \theta_N = (\boldsymbol \alpha^{(1)}, \dots, \boldsymbol \alpha^{(M)}, \boldsymbol \beta,\Gamma^{(0)}, \dots, \Gamma^{(M - 1)}) \in \Theta_N \subset \mathbb{R}^{q(N)} $, consisting of
+model parameters $\boldsymbol \beta \in \mathbb{R}^N$,
 \(\boldsymbol \alpha^{(i)} \in \mathbb{R}^{N_{(H,i)}}\),
@@
 \(N_{(H,1)} \times N\), and matrices \(\Gamma^{(i)}\) of dimension
-\(N_{(H,i)} \times N_{(H,i+1)}\) for \(i = 1, \dots, M-1\). The combined
-parameter vector \(\boldsymbol \theta_N\) has total length
+\(N_{(H,i)} \times N_{(H,i+1)}\) for \(i = 1, \dots, M-1\). The combined   parameter vector $\boldsymbol \theta_N$ has total length
 \(q(N)= N_{(H,1)}+\cdots N_{(H,M)} + N + N_{(H,1)}*N+N_{H,2}*H_{(H,1)}+\cdots +N_{(H,M)}*H_{(H,M)-1}\).
@@
 \(X_1, \dots, X_N\) follows from this joint specification as \[
-P_{\boldsymbol \theta_N} (\boldsymbol x) = \sum\limits_{(\boldsymbol h^{(1)}, \dots, \boldsymbol h^{(M)}) \in \mathcal{X}^{N_{(H,1)} + \cdots + N_{(H,M)}}} \tilde{P}_{\boldsymbol \theta_N} (\boldsymbol h^{(1)}, \dots, \boldsymbol h^{(M)}, \boldsymbol x) , \qquad \boldsymbol x \in \mathcal{X}^N.
+P_{\boldsymbol \theta_N} (\boldsymbol x) = \sum\limits_{(\boldsymbol h^{(1)}, \dots, \boldsymbol h^{(M)}) \in \mathcal{X}^{N_{(H,1)} + \cdots + N_{(H,M)}}} \tilde{P}_{\boldsymbol \theta_N} (\boldsymbol h^{(1)}, \dots, \boldsymbol h^{(M)}, \boldsymbol x) , \qquad \boldsymbol x \in \mathcal{X}^N
 \] Again like the RBM case, the DBM model specification is an example of
-a FOES model.
-
-\textbf{Deep belief network (DBN).} A DBN resembles a DBM in that there
+a FOES model.\\
+
+\noindent \textbf{Deep belief network (DBN).} A DBN resembles a DBM in that there
 are multiple layers of latent random variables stacked in a deep
@@
 
-\subsection{A criterion for instability}\label{criterion}
-
-To define a measure of instability in FOES models, it is useful to
-consider the behavior of data models \(P_{\theta_N}\), again supported
-on a set \(\mathcal{X}^N\) of outcomes for
-\(\boldsymbol X\equiv \boldsymbol X_N =(X_1,\ldots,X_N)\), in connection
-to the sample size \(N\). A relevant quantity to this end is a log-ratio
-of extremal probabilities (LREP), defined as
+\subsection{A criterion for instability}
+\label{criterion}
+To define a measure of instability in FOES models, it is useful to consider the behavior of  data  models $P_{\theta_N}$, again supported on a set $\mathcal{X}^N$ of outcomes for $\boldsymbol X\equiv \boldsymbol X_N =(X_1,\ldots,X_N)$,  in connection to the sample size $N$.
+A relevant quantity to this end is a log-ratio of extremal probabilities (LREP), defined as
 \begin{align}
@@
 \end{align}
-based on maximum and minimal model probabilities. In what follows, the
-main idea is that instability, and other negative model features, can be
-associated with a FOES model formulation for \(N\) random variables
-where the log-ratio \eqref{eq:elpr} is overly large relative to the sample
-size \(N\). That is, a sequence of FOES probability models
-\(P_{\boldsymbol \theta_N}\) results in specifying the distribution of
-observations \(\boldsymbol X=(X_1,\ldots,X_N)\) for each sample size
-\(N \geq 1\) and instability will generally occur among these models
-whenever the corresponding LREP \eqref{eq:elpr} grows faster than \(N\).
-This leads to the following definition.
-
-\BeginKnitrBlock{definition}[S-unstable FOES model]
-\protect\hypertarget{def:instabFSFS}{}{\label{def:instabFSFS}
-\iffalse (S-unstable FOES model) \fi{} }A FOES model formulation for
-\(\boldsymbol X_N=(X_1,\ldots,X_N)\) is \emph{Schweinberger-unstable} or
-\emph{S-unstable} if
+based on maximum and minimal model probabilities.
+In what follows, the main idea is that instability, and other negative model features, can be associated with a FOES model formulation
+for $N$ random variables where the log-ratio (\ref{eq:elpr}) is overly large relative to the sample size $N$.  That is,  a   sequence of FOES probability models $P_{\boldsymbol \theta_N}$  results
+in specifying the distribution of observations $\boldsymbol X=(X_1,\ldots,X_N)$ for each sample size $N \geq 1$
+ and instability will generally occur among these models whenever   the corresponding LREP (\ref{eq:elpr}) grows faster than $N$.  This leads to the following definition.
+
+  \vspace{.25cm}
+
+\noindent \textbf{Definition} {\it(S-unstable FOES model)}:
+A FOES model formulation for $\boldsymbol X_N=(X_1,\ldots,X_N)$ is \emph{Schweinberger-unstable} or \emph{S-unstable} if
 \begin{equation}
-\label{eq:Sun}
+\label{eqn:Sun}
 \lim \limits_{N \rightarrow \infty} \frac{1}{N} \REP(\boldsymbol \theta_N) \equiv \lim \limits_{N \rightarrow \infty} \frac{1}{N}\log \left[\frac{\max\limits_{  \boldsymbol x\in \mathcal{X}^N}P_{\boldsymbol \theta_N}( \boldsymbol x)}{\min\limits_{ \boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}( \boldsymbol x)}\right] = \infty
 \end{equation}
-as the number of variables increases (\(N \rightarrow \infty\)).
-\EndKnitrBlock{definition} In other words, a model is S-unstable if
-\(\REP(\boldsymbol \theta_N)/N\) is an unbounded sequence of sample size
-\(N\); namely, given any \(C > 0\), there exists an integer \(N_C > 0\)
-so that \(\REP(\boldsymbol \theta_N)/N > C\) holds for all
-\(N \ge N_C\). A FOES model formulation may be termed S-stable if it
-fails to be S-unstable, i.e., if
-\(\sup_{N \geq 1}\REP(\boldsymbol \theta_N)/N\) is bounded.
-
-This definition of S-unstable is a generalization or reinterpretation of
-``unstable'' used in Schweinberger
-(\protect\hyperlink{ref-schweinberger2011instability}{2011}) by allowing
-possibility non-exponential family models (e.g., RBM and DBM models in
-Sections \ref{restricted-boltzmann-machines} - \ref{deep-learning} as
-well as a potentially increasing number \(d(N)\) of parameters through
-the parameter sequence \(\boldsymbol \theta_N\in \mathbb{R}^{d(N)}\).
-While this definition differs in form and scope from the original, it
-does match that in Schweinberger
-(\protect\hyperlink{ref-schweinberger2011instability}{2011}) for the
-special case of exponential models (cf.~Section
-\ref{discrete-exponential-family-models} considered there. Section
-\ref{illustrations} provides several examples of unstable models as well
-as causes for model instability, where the latter may often be traced to
-issues in model form (i.e., data functions) and/or parameterization. We
-next describe several potentially undesirable features associated with
-unstable models.
-
-\BeginKnitrBlock{remark}
-\iffalse <span class="remark"><em>Remark. \fi{} In the definition
-\eqref{eq:Sun} of S-instability, we note that the numerical measure
-\(\REP(\boldsymbol \theta_N)/N\) of model instability is invariant to
-\emph{independent replications} of data. That is, let \(M \geq 1\)
-denote a possible number of replications and consider data
-\(\boldsymbol Y_{N,M} \equiv (\boldsymbol X^{(1)}_N, \dots, \boldsymbol X^{(M)}_N)\)
-formed by \(\{ \boldsymbol X^{(j)}_N\}_{j=1}^M\) as \(M\) iid
-replications of a random vector \(\boldsymbol X_N=(X_1,\ldots,X_N)\),
-where the latter follows a FOES model with probabilities
-\(P_{ \boldsymbol \theta_N}(\boldsymbol x)>0\),
-\(\boldsymbol x\in\mathcal{X}^N\). This leads to a joint model, say
-\(P_{ \boldsymbol \theta_N}(\boldsymbol y)\),
-\(\boldsymbol y\in\mathcal{X}^{NM}\), for \(\boldsymbol Y_{N,M}\)
-consisting of \(N*M\) random variables in total. Then, the LREP for
-\(\boldsymbol Y_{N,M}\), scaled by associated size, is given by
-\begin{align*}
-\frac{1}{NM}\REP_{\boldsymbol Y_{N,M}}( \boldsymbol\theta_N ) &\equiv \frac{1}{NM}\log\left[\frac{\max_{\boldsymbol y \in \mathcal{X}^{NM}}P_{ \boldsymbol \theta_N}(\boldsymbol y)  }{\min_{\boldsymbol y \in \mathcal{X}^{NM}}P_{ \boldsymbol \theta_N}(\boldsymbol y)} \right] \\
-&= \frac{1}{NM}\log\left[\frac{\max_{\boldsymbol x \in \mathcal{X}^{N}}P_{ \boldsymbol \theta_N}(\boldsymbol x)  }{\min_{\boldsymbol x \in \mathcal{X}^{N}}P_{ \boldsymbol \theta_N}(\boldsymbol x)} \right]^M \equiv \frac{1}{N}\REP_{\boldsymbol X_{N}}( \boldsymbol\theta_N ),
-\end{align*}
-where
-\(\REP_{\boldsymbol X_{N}}( \boldsymbol\theta_N ) \equiv \REP( \boldsymbol\theta_N )\)
-denotes the log-ratio of extremal probabilities for \(\boldsymbol X_N\)
-defined from \eqref{eq:elpr}. That is, due to iid properties, the
-sample-size corrected LREP for \(\boldsymbol Y_{N,M}\) equals the
-analog, \(\REP(\boldsymbol\theta_N )/N\), from the underlying common
-data model for \(\boldsymbol X_N\) alone, regardless of the level
-\(M \geq 1\) of independent replication. Consequently, the definition of
-an S-unstable model is unaffected by independent replication and all
-instability properties may be characterized by those of one observation
-from the common FOES model. For computational purposes, this aspect also
-implies that if the original data \(\boldsymbol X_N=(X_1,\ldots,X_N)\)
-in a FOES model consist of \(N\) iid random variables, then the
-size-scaled log-ratio \eqref{eq:elpr} may be calculated as \[
+ as the number of variables increases ($N \rightarrow \infty$).\\[.15cm]
+In other words, a model is S-unstable if $\REP(\boldsymbol \theta_N)/N$ is an unbounded sequence of sample size  $N$; namely, given any $C > 0$, there exists an integer $N_C > 0$ so that $\REP(\boldsymbol \theta_N)/N > C$ holds for all $N \ge N_C$.
+A FOES model formulation may be termed S-stable if it fails to be S-unstable, i.e., if $\sup_{N \geq 1}\REP(\boldsymbol \theta_N)/N $ is bounded.
+
+This definition of S-unstable is a generalization or reinterpretation of ``unstable" used in  Schweinberger
+(\protect\hyperlink{ref-schweinberger2011instability}{2011}) by allowing possibility non-exponential family models (e.g., RBM and DBM models in Sections \ref{restricted-boltzmann-machines}-\ref{deep-learning}) as well as a potentially increasing number $d(N)$ of parameters through the parameter sequence $\boldsymbol \theta_N\in \mathbb{R}^{d(N)}$.
+While this definition differs in form and scope from the original, it does match that in Schweinberger
+(\protect\hyperlink{ref-schweinberger2011instability}{2011}) for the special case of exponential models (cf.~Section~\ref{discrete-exponential-family-models}) considered there.
+Section~4 provides several examples  of unstable models as well as causes for model instability, where the latter may often be traced to issues in model form (i.e., data functions) and/or parameterization.
+We next describe several potentially undesirable features associated with unstable models.\\
+
+\noindent \textbf{Remark 1:}   In the definition (\ref{eqn:Sun}) of S-instability, we note that the numerical measure $\REP(\boldsymbol \theta_N)/N$ of model instability is invariant to {\it independent  replications} of data.
+ That is, let $M \geq 1$ denote a possible number of replications and
+ consider data \(  \boldsymbol Y_{N,M} \equiv (\boldsymbol X^{(1)}_N, \dots, \boldsymbol X^{(M)}_N)\)
+formed by $\{ \boldsymbol X^{(j)}_N\}_{j=1}^M$ as
+  \(M\)  iid replications of a random vector $\boldsymbol X_N=(X_1,\ldots,X_N)$, where the latter follows a
+  FOES model with probabilities \(P_{ \boldsymbol \theta_N}(\boldsymbol x)>0\), $\boldsymbol x\in\mathcal{X}^N$.
+   This leads
+to a joint model, say $P_{ \boldsymbol \theta_N}(\boldsymbol y)$, $\boldsymbol y\in\mathcal{X}^{NM}$, for $\boldsymbol Y_{N,M}$ consisting of  \(N*M\) random variables in total.
+Then, the  LREP for $\boldsymbol Y_{N,M}$, scaled by associated size, is given by
+\begin{eqnarray*}
+\frac{1}{NM}\REP_{\boldsymbol Y_{N,M}}( \boldsymbol\theta_N ) &\equiv &\frac{1}{NM}\log\left[\frac{\max_{\boldsymbol y \in \mathcal{X}^{NM}}P_{ \boldsymbol \theta_N}(\boldsymbol y)  }{\min_{\boldsymbol y \in \mathcal{X}^{NM}}P_{ \boldsymbol \theta_N}(\boldsymbol y)} \right]\\&=&\frac{1}{NM}\log\left[\frac{\max_{\boldsymbol x \in \mathcal{X}^{N}}P_{ \boldsymbol \theta_N}(\boldsymbol x)  }{\min_{\boldsymbol x \in \mathcal{X}^{N}}P_{ \boldsymbol \theta_N}(\boldsymbol x)} \right]^M \equiv\frac{1}{N}\REP_{\boldsymbol X_{N}}( \boldsymbol\theta_N ),
+\end{eqnarray*}
+where $\REP_{\boldsymbol X_{N}}( \boldsymbol\theta_N )  \equiv \REP( \boldsymbol\theta_N ) $ denotes the log-ratio of extremal probabilities for $\boldsymbol X_N$ defined from (\ref{eq:elpr}).
+That is, due to iid properties,
+the sample-size corrected  LREP for   $\boldsymbol Y_{N,M}$ equals the analog, $\REP( \boldsymbol\theta_N )/N$, from the underlying common data model for $\boldsymbol X_N$ alone, regardless of the level $M \geq 1$ of independent replication.
+ Consequently,   the definition of an S-unstable  model  is  unaffected by independent replication and all   instability properties  may be characterized by those of one
+observation  from the common FOES model.  For  computational purposes, this aspect also implies that if the original data $\boldsymbol X_N=(X_1,\ldots,X_N)$ in a FOES model
+consist of $N$ iid random variables, then the size-scaled log-ratio (\ref{eq:elpr})  may be calculated as
+\[
 \frac{1}{N}\REP( \boldsymbol\theta_N ) \equiv \frac{1}{N}\REP_{\boldsymbol X_{N}}( \boldsymbol\theta_N ) = \log\left[ \frac{\max_{ x \in \mathcal{X}}P_{\boldsymbol\theta_N}(X_1=x)} {\min_{ x \in \mathcal{X}}P_{\boldsymbol\theta_N}(X_1=x)} \right]
-\] based on the extremal probabilities of just one random variable
-\(X_1\).
-\EndKnitrBlock{remark}
-
-\subsection{Characterizations and consequences of
-instability}\label{characterizations-and-consequences-of-instability}
-
-As a basic characteristic, S-unstable FOES model sequences have
-extremely sensitive probability structures. One aspect is that small
-changes in data configuration can lead to very large changes in
-probability. Consider, for example, the quantity given by \[
+\]
+based on the extremal probabilities of just one random variable $X_1$.
+
+\subsection{Characterizations and Consequences of Instability}
+
+  As a basic characteristic, S-unstable FOES model sequences  have extremely sensitive probability structures.
+One aspect is that small changes in data configuration can   lead to very large changes in probability.
+Consider, for example, the quantity given by
+$$
 \DN(\boldsymbol \theta_N) \equiv \max \left\{\log \frac{P_{\boldsymbol \theta_N}(\boldsymbol x)}{P_{\boldsymbol \theta_N}(\boldsymbol x^*)} : \boldsymbol x \text{ }\& \text{ } \boldsymbol x^* \in \mathcal{X}^N \text{ differ in exactly one component}\right\},
-\] which represents the biggest log-probability ratio for a one
-component change in data outcomes in a FOES model with parameter
-\(\boldsymbol \theta_N\). We then have the following result prescribing
-the behavior of \(\DN(\boldsymbol \theta_N)\) for unstable models.
-
-\BeginKnitrBlock{theorem}
-\protect\hypertarget{thm:instab-elpr}{}{\label{thm:instab-elpr}}Let
-\(P_{\boldsymbol \theta_N}\), with support \(\mathcal{X}^N\),
-\(N\geq 1\), be a sequence of FOES models.
-\begin{enumerate}[(i)]
-\item For any integer $N \geq 1$ and any given $C>0$, if $\REP(\boldsymbol \theta_N)/N > C$ in (\ref{eq:elpr}, then
-    $$ \DN(\boldsymbol \theta_N) > C,$$
-    or probabilities from a one component change in some outcome have log-ratio exceeding $C$.
-\item Suppose the FOES model sequence is S-unstable.
+$$
+which represents the biggest log-probability ratio for a one component change in data outcomes in a FOES model with parameter $\boldsymbol \theta_N$.
+We then have the following result prescribing the behavior of $\DN(\boldsymbol \theta_N)$ for   unstable models.
+
+\newtheorem{theorem1}{Theorem}
+
+\begin{theorem1}
+\label{instab-elpr}  {\it Let $P_{\boldsymbol \theta_N}$, with support $\mathcal{X}^N$, $N\geq 1$, be a sequence of FOES models. \\
+(i)   For any integer $N \geq 1$ and  any given $C>0$, if $\REP(\boldsymbol \theta_N)/N > C$ in (\ref{eq:elpr}), then
+\[\DN(\boldsymbol \theta_N) > C,\]
+or probabilities from a one component change in some outcome have    log-ratio  exceeding $C$.\\
+(ii) Suppose the FOES model sequence is S-unstable.
 Then,  for all large $N$ and given any arbitrary $C>0$,   there exist outcomes $\boldsymbol x,\boldsymbol x^*\in\mathcal{X}^N$, differing by one component, such that
-    $$
+\[
  \frac{P_{\boldsymbol \theta_N}(\boldsymbol x)}{P_{\boldsymbol \theta_N}(\boldsymbol x^*)} > \exp[N C].
-    $$
-\end{enumerate}
-\EndKnitrBlock{theorem}
-
-Theorem \ref{thm:instab-elpr}(i) is a non-asymptotic result, which
-connects to the definition of instability in a FOES model through a
-log-ratio of extreme model probabilities \eqref{eq:elpr} being too large
-relative to the associated sample size \(N\). If so, Theorem
-\ref{thm:instab-elpr}(i) guarantees the FOES model must also exhibit
-correspondingly large changes in probability for very small differences
-among some data configurations, a property that intuitively captures a
-notion of instability. Furthermore, and perhaps more seriously under
-Theorem \ref{thm:instab-elpr}(ii), S-unstable models can never have
-universally bounded changes in probability among single component
-variations in data configurations. While not all one component changes
-in data may produce massive changes in probability, unstable models must
-have some such data outcomes with this property. As a consequence,
-unstable probability structure may exhibit extreme sensitivity through
-large peaks and troughs over the sample space.
-
-Additionally, S-unstable FOES model sequences are also connected to
-degenerate models, where \emph{degeneracy} involves assigning
-essentially all probability to modes within the sample space, which
-could potentially represent a small subset among the totality of
-outcomes. For perspective, note that differing sizes of the scaled
-log-ratio \(\REP(\boldsymbol \theta_N)/N\) from \eqref{eq:elpr} induce a
-spectrum of levels of instability/stability and Theorem
-\ref{thm:instab-elpr} indicates increasing sensitivity of model
-probabilities as \eqref{eq:elpr} increases. Furthermore, as the
-instability measure grows and the log-ratio
-\(\REP(\boldsymbol \theta_N)/N\) diverges, as in the definition
-\eqref{eq:Sun} of S-unstable models, then a FOES model sequence will
-become degenerate. Theorem \ref{thm:degenFOES} provides a formal
-statement of such degeneracy due to instability. For a given
-\(0 < \epsilon < 1\), define a \(\epsilon\)-modal set of outcomes as
+\]}
+\end{theorem1}
+
+Theorem~\ref{instab-elpr}(i) is a  non-asymptotic  result,  which connects to the definition of instability in a FOES model through a log-ratio of extreme model probabilities (\ref{eq:elpr}) being too large relative to the associated sample size $N$.
+If so,  Theorem~\ref{instab-elpr}(i) guarantees the FOES model must also exhibit correspondingly large changes in probability for very small differences among some data configurations, a property that intuitively captures a notion of instability.
+Furthermore, and perhaps more seriously under
+Theorem~\ref{instab-elpr}(ii),
+S-unstable models can never have universally bounded changes in probability among single component variations in data configurations.
+While not all one component changes in data may produce massive changes in probability, unstable models must have some such data outcomes with this property.
+As a consequence, unstable probability structure may exhibit extreme sensitivity through large peaks and troughs over the sample space.
+
+Additionally, S-unstable FOES model sequences are also connected to degenerate models, where {\it degeneracy} involves assigning essentially all probability   to modes within the sample space, which could potentially represent a  small subset among the totality of outcomes.
+For perspective, note that differing sizes of the scaled log-ratio $\REP(\boldsymbol \theta_N)/N$ from (\ref{eq:elpr}) induce a spectrum of levels of instability/stability and Theorem~\ref{instab-elpr} indicates increasing sensitivity of model probabilities as (\ref{eq:elpr}) increases.
+Furthermore, as the instability measure grows and the log-ratio $\REP(\boldsymbol \theta_N)/N$ diverges, as in the definition (\ref{eqn:Sun}) of S-unstable models, then a FOES model sequence  will become degenerate.
+ Theorem~\ref{degenFOES} provides a formal statement of such degeneracy due to instability. For a given $0 < \epsilon < 1$, define a $\epsilon$-modal set of   outcomes as
 \begin{equation}
-\label{eq:mode}
+\label{eqn:mode}
 \mathcal{M}_{\epsilon, \boldsymbol \theta_N} \equiv \left\{\boldsymbol x \in \mathcal{X}^N: \log P_{\boldsymbol \theta_N}(\boldsymbol x) > (1-\epsilon)\max\limits_{\boldsymbol y \in \mathcal{X}^N} \log  P_{\boldsymbol \theta_N}(\boldsymbol y) + \epsilon\min\limits_{\boldsymbol y \in \mathcal{X}^N} \log P_{\boldsymbol \theta_N}(\boldsymbol y) \right\}.
 \end{equation}
-\BeginKnitrBlock{theorem}
-\protect\hypertarget{thm:degenFOES}{}{\label{thm:degenFOES}}For any
-arbitrarily small \(0 < \epsilon < 1\), an S-unstable FOES model
-sequence \(P_{\boldsymbol \theta_N}\), \(N \geq 1\), for
-\(\boldsymbol X_N=(X_1, \dots, X_N)\) satisfies \[
+
+\begin{theorem1}
+\label{degenFOES} {\it For any arbitrarily small  $0 < \epsilon < 1$,  an S-unstable FOES model sequence $P_{\boldsymbol \theta_N}$, $N \geq 1$, for $\boldsymbol X_N=(X_1, \dots, X_N)$ satisfies
+$$
 P_{\boldsymbol \theta_N}\left( \boldsymbol X_N\in \mathcal{M}_{\epsilon, \boldsymbol \theta_N}\right) \rightarrow 1 \text{ as } N \rightarrow \infty.
-\]
-\EndKnitrBlock{theorem}
-
-In other words, as the sample size grows in S-unstable FOES models, all
-probability tends to concentrate mass on an \(\epsilon\)-modal set,
-where \(\epsilon\) can be made arbitrarily small. Intuitively, the
-occurrence of such degeneracy can be explained by a type of ``reverse''
-pigeonhole principle for unstable FOES models: if all outcomes should
-receive positive probability but the maximal probability far exceeds the
-minimal one in the model, then little probability remains for
-distribution among remaining model outcomes (i.e., if nearly all
-available pigeons are stuffed into one hole, the remaining pigeonholes
-must have few occupants). Degeneracy in unstable models can pose dangers
-in data modeling as well, particularly when a mode set represents a
-narrow collection of outcomes among those realistically possible for
-adequately describing data. In which case, model outcomes may fail to
-look like data of interest.
-
-Connected to degeneracy, unstable models may also exhibit additional
-kinds of extreme and undesirable sensitivity in probabilities if model
-parameters \(\boldsymbol \theta_N\) can further be ``dialed'' between
-positive and negative values. That is, some FOES models naturally
-involve parameter spaces covering a positive-negative spectrum of
-parameter possibilities, where the signs of parameters provide a
-standard device for increasing or decreasing probabilities of outcomes
-in the model formulation. In fact, for many models, the switch of a
-parameter sign serves to produce reciprocal probabilities, as outlined
-in the following model assumption about parameter sign reversal (PSR).
-
-\textbf{Model Condition PSR} \emph{(Reciprocal Probabilities from
-Parameter Sign Reversal)}: Let \(P_{\boldsymbol \theta_N}\), with
-support \(\mathcal{X}^N\), \(N\geq 1\), represent a sequence of FOES
-models. For each \(N \geq 1\) and any outcome
-\(\boldsymbol x \in \mathcal{X}^N\), suppose it holds that \[
+$$}
+\end{theorem1}
+
+In other words, as the sample size grows in S-unstable FOES models, all probability   tends to concentrate mass on an $\epsilon$-modal set, where $\epsilon$ can be made arbitrarily small.
+Intuitively, the occurrence of such degeneracy can be explained
+by a type of ``reverse" pigeonhole principle for unstable FOES models: if all outcomes should receive positive probability but the maximal  probability far exceeds the minimal one in the model, then little probability remains for distribution among remaining model   outcomes (i.e., if nearly all available pigeons are stuffed into one hole, the remaining  pigeonholes must have few occupants).
+Degeneracy in unstable models can pose dangers in data modeling as well, particularly when a mode set  represents a  narrow collection of outcomes among those  realistically  possible for adequately describing data.
+In which case, model outcomes may fail to look like data of interest.
+
+Connected to degeneracy, unstable models may also exhibit additional kinds of  extreme and undesirable sensitivity in probabilities if model parameters $\boldsymbol \theta_N$ can further be ``dialed"  between positive and negative values.
+That is, some FOES  models naturally involve parameter spaces covering a positive-negative spectrum of parameter possibilities, where the signs of parameters provide a standard device for increasing or decreasing probabilities of outcomes in the model formulation.
+In fact, for many models, the switch of a parameter sign serves to produce reciprocal probabilities, as outlined in the following model assumption about parameter sign reversal (PSR).\\
+
+\noindent \textbf{Model Condition PSR} {\it (Reciprocal Probabilities from Parameter Sign Reversal)}: Let $P_{\boldsymbol \theta_N}$, with support $\mathcal{X}^N$, $N\geq 1$, represent a sequence of FOES models.
+For each $N \geq 1$ and any outcome
+$\boldsymbol x \in \mathcal{X}^N$, suppose it holds that
+\[
   P_{\boldsymbol \theta_N}(\boldsymbol x)  \cdot P_{-\boldsymbol \theta_N}(\boldsymbol x) \;=\;   \max\limits_{\boldsymbol y \in \mathcal{X}^N}P_{ \boldsymbol \theta_N}(\boldsymbol y)\cdot \min\limits_{\boldsymbol y \in \mathcal{X}^N}P_{-\boldsymbol \theta_N}(\boldsymbol y),
-\] where
-\(\max_{\boldsymbol y \in \mathcal{X}^N}P_{ \boldsymbol \theta_N}(\boldsymbol y)\)
-and
-\(\min_{\boldsymbol y \in \mathcal{X}^N}P_{-\boldsymbol \theta_N}(\boldsymbol y)\)
-denote the maximum and minimum probabilities under parameters
-\(\boldsymbol \theta_N\) and \(-\boldsymbol \theta_N\), respectively.
-
-The above model condition incorporates many standard parameterizations
-and follows, for instance, whenever
-\(P_{\boldsymbol \theta_N}(\boldsymbol x)/P_{\boldsymbol \theta_N}(\boldsymbol y) = [P_{-\boldsymbol \theta_N}(\boldsymbol x)/P_{-\boldsymbol \theta_N}(\boldsymbol y)]^{-1}\)
-holds for outcomes \(\boldsymbol x, \boldsymbol y \in\mathcal{X}^N\) in
-a FOES model. For instance, this latter condition is fulfilled for all
-linear exponential families from Section
-\ref{discrete-exponential-family-models} (e.g.,
-\eqref{eq:mod1}-\eqref{eq:mod2}) as well as all network models from Sections
-\ref{restricted-boltzmann-machines}-\ref{deep-learning} (e.g.,
-\eqref{eq:RBM1}-\eqref{eq:RBM2}). When parameters can be tuned in sign with
-effects prescribed in the model condition PSR, unstable FOES models will
-exhibit further probability sensitivities, as outlined in the following
-extension of Theorem @ref(thm:degenFOES\}).
-
-\BeginKnitrBlock{corollary}
-\protect\hypertarget{cor:sign}{}{\label{cor:sign}}Let
-\(P_{\boldsymbol \theta_N}\), with support \(\mathcal{X}^N\),
-\(N\geq 1\), be a sequence of FOES models satisfying model condition
-PSR. If the models \(P_{\boldsymbol \theta_N}\) are additionally
-S-unstable, then
-\begin{enumerate}[(i)]
-\item the models  $P_{-\boldsymbol \theta_N}$  defined by $-\boldsymbol \theta_N$  are also S-unstable;
-\item and for the complement $\mathcal{M}_{\epsilon, \boldsymbol \theta_N}^c \equiv \mathcal{X}^N \setminus \mathcal{M}_{\epsilon, \boldsymbol \theta_N}$ of any mode-set $\mathcal{M}_{\epsilon, \boldsymbol \theta_N}$ under $\boldsymbol \theta_N$ from (\ref{eq:mode}), with $0<\epsilon<1$, it holds under $- \boldsymbol \theta_N$ that
-    $$ 
+\]
+where $\max_{\boldsymbol y \in \mathcal{X}^N}P_{ \boldsymbol \theta_N}(\boldsymbol y)$ and $\min_{\boldsymbol y \in \mathcal{X}^N}P_{-\boldsymbol \theta_N}(\boldsymbol y)$ denote the   maximum and minimum probabilities under parameters $\boldsymbol \theta_N$   and
+ $-\boldsymbol \theta_N$, respectively.\\
+
+ The above model condition incorporates many standard parameterizations and follows, for instance, whenever $ P_{\boldsymbol \theta_N}(\boldsymbol x)/P_{\boldsymbol \theta_N}(\boldsymbol y)  = [P_{-\boldsymbol \theta_N}(\boldsymbol x)/P_{-\boldsymbol \theta_N}(\boldsymbol y)]^{-1}$ holds for outcomes $\boldsymbol x,\boldsymbol y\in\mathcal{X}^N$ in a FOES model.
+For instance, this latter condition is fulfilled for all
+ linear exponential families from Section~\ref{discrete-exponential-family-models} (e.g., (\ref{eqn:mod1})-(\ref{eqn:mod2})) as well as all network models from Sections~\ref{restricted-boltzmann-machines}-\ref{deep-learning} (e.g., (\ref{eqn:RBM1})-(\ref{eqn:RBM2})).
+When parameters can be tuned in sign with effects prescribed in the model condition PSR, unstable FOES models will exhibit further probability sensitivities, as outlined in the
+ following extension of Theorem~\ref{degenFOES}.
+
+\newtheorem{corollary1}{Corollary}
+
+  \begin{corollary1}
+  \label{sign}
+  {\it Let $P_{\boldsymbol \theta_N}$, with support $\mathcal{X}^N$, $N\geq 1$, be a sequence of FOES models satisfying model condition PSR.
+If the models $P_{\boldsymbol \theta_N}$  are additionally S-unstable, then \\
+  (i) the models  $P_{-\boldsymbol \theta_N}$  defined by $-\boldsymbol \theta_N$  are also S-unstable;\\
+  (ii)  and for the complement $\mathcal{M}_{\epsilon, \boldsymbol \theta_N}^c \equiv \mathcal{X}^N \setminus \mathcal{M}_{\epsilon, \boldsymbol \theta_N}$ of any mode-set $\mathcal{M}_{\epsilon, \boldsymbol \theta_N} $ under $\boldsymbol \theta_N$ from (\ref{eqn:mode}), with   $0<\epsilon<1$, it holds under $- \boldsymbol \theta_N$ that
+  \[
    P_{-\boldsymbol \theta_N}( \boldsymbol X_N \in \mathcal{M}_{\epsilon, \boldsymbol \theta_N}^c) \rightarrow 1\qquad \mbox{as $N\to \infty$},
-    $$
-    while, by Theorem \ref{thm:degenFOES}, $P_{\boldsymbol \theta_N}( \boldsymbol X_N \in \mathcal{M}_{\epsilon, \boldsymbol \theta_N} ) \rightarrow 1$ holds for $\boldsymbol X_N = (X_1,\ldots,X_N)$ under $\boldsymbol \theta_N$.
-\end{enumerate}
-\EndKnitrBlock{corollary}
-
-For unstable models, Corollary \ref{cor:sign} shows that shifts in
-parameters around zero (i.e., from \(\boldsymbol \theta_N\) to
-\(-\boldsymbol \theta_N\)) can induce extreme changes in probability
-among subsets of the sample space, as another manifestation of
-instability and hyper-sensitivity in probability structure. For
-one-parameter exponential families, involving a fixed real-valued linear
-parameter \(\boldsymbol \theta_N = \theta \in \mathbb{R}\) and
-sufficient statistic \(\boldsymbol g_N(\boldsymbol x)\in \mathbb{R}\) in
-\eqref{eq:expo}, Schweinberger
-(\protect\hyperlink{ref-schweinberger2011instability}{2011} Theorem 3)
-proved a result similar in spirit, though based on a characterization
-there in terms of maximum
-\(U_N \equiv \max_{\boldsymbol x\in\mathcal{X}^N}g_N(\boldsymbol x)\)
-and minimal
-\(L_N \equiv \min_{\boldsymbol x\in\mathcal{X}^N}g_N(\boldsymbol x)\)
-values of the sufficient statistic. For this case in particular, mode
-sets have specific, and essentially complementary, forms over positive
-and negative parameters, namely,
-\(\mathcal{M}_{\epsilon, \boldsymbol \theta_N} = \{\boldsymbol x \in\mathcal{X}^N: g_N(\boldsymbol x) > (1-\epsilon) U_N + \epsilon L_N \}\)
-and
-\(\mathcal{M}_{\epsilon, -\boldsymbol \theta_N} = \{\boldsymbol x \in\mathcal{X}^N: g_N(\boldsymbol x) < \epsilon U_N + (1-\epsilon) L_N \}\)
-for any \(\boldsymbol \theta_N>0\), and Schweinberger
-(\protect\hyperlink{ref-schweinberger2011instability}{2011} Theorem 3)
-showed each mode set collects all mass, under positive and negative
-parameters, respectively, with unstable models of this exponential type.
-However, for all unstable FOES models, Corollary \ref{cor:sign}
-generalizes the same principle that unstable models can push all
-probability to different, and in fact disjoint, parts of the sample
-space, depending on how parameters fall with respect to zero. This
-feature can numerically complicate likelihood manipulations, such as
-maximization or MCMC-based Bayes posterior sampling, as further
-discussed in Section \ref{implications}.
-
-\BeginKnitrBlock{remark}[need to briefly finish]
-\iffalse <span class="remark"><em>Remark (need to briefly finish). \fi{}
-extensions of sensitivity\ldots{} basically, model condition PSR and
-Corollary \ref{cor:sign} can be extended to cases where parameter
-components
-\(\boldsymbol \theta_N =(\boldsymbol \theta_{1,N}, \boldsymbol \theta_{N,2})\)
-are not all changed in sign (\(-\boldsymbol \theta_N\)) but, more
-generally, are instead altered to another parameter configuration
-\(\boldsymbol \theta_{N}^A = ( \boldsymbol \theta_{1,N}^A, \boldsymbol \theta_{2,N}^A )\)
-involving a switch in sign only among some model parameters
-\(\boldsymbol \theta_{N,2}^A=- \boldsymbol \theta_{N,2}\) with remaining
-parameters \(\boldsymbol \theta_{1,N}^A\) being arbitrarily chosen. If a
-change sign occurs among parameters (\(\pm \boldsymbol \theta_{N,2}\))
-which dominate the probability structure of the model, then an unstable
-model under \(\boldsymbol \theta_N\) will then imply that many more
-unstable models exist over a broad spectrum of possibilities for
-variations \(\boldsymbol \theta_{N}^A\) of \(\boldsymbol \theta_N\).
-\EndKnitrBlock{remark}
-
-\section{Illustrations}\label{illustrations}
-
-Model instability can depend intricately on how functions of parameters
-and data \(\boldsymbol X_N=(X_1,\ldots,X_n)\) are combined in the
-formulation of the model probabilities, though some general causes may
-be identified. As one issue, a broad parameter space (or wide
-interpretation of this space) may admit some parameters as technically
-valid that have an undue and often undesirable impact on the model
-structure for a prescribed data size \(N\). In this case, both the size
-and dimension of model parameters can be problematic and induce
-instability. In combination to this last point, further causes of
-instability may also be traced to the magnitude of statistics in the
-model. Potentially massive, and thereby unstable, statistics were the
-primary focus of instability studies of Schweinberger
-(\protect\hyperlink{ref-schweinberger2011instability}{2011}) for certain
-discrete exponential models having parameters/statistics of fixed
-dimension. However, as shown in the following, bounded statistics may
-still lead to instability if the parameter dimension is high. We next
-provides some examples to illustrate instability in FOES models, which
-also suggest some potential strategies for preventing unstable models.
-
-\subsection{Equi-probability models}\label{equi-probability-models}
-
-As a baseline for comparisons, consider a simplistic model for
-\(\boldsymbol X_N=(X_1,\ldots,X_N)\) with uniform probabilities over the
-sample space, say
-\(P_{\boldsymbol \theta_N}(\boldsymbol x)= |\mathcal{X}|^{-N}\),
-\(\boldsymbol x \in \mathcal{X}^N\), where each random variable has
-\(|\mathcal{X}| \geq 1\) outcomes. In contrast to instability, model
-probabilities here are completely insensitive to changes in data
-outcomes across the sample space, and the associated log-ratio of
-extreme probabilities \eqref{eq:elpr} is \[
+  \]
+   while,  by Theorem~\ref{degenFOES},  $P_{\boldsymbol \theta_N}( \boldsymbol X_N \in \mathcal{M}_{\epsilon, \boldsymbol \theta_N} ) \rightarrow 1$ holds for
+  $ \boldsymbol X_N =(X_1,\ldots,X_N)$ under $  \boldsymbol \theta_N$.}
+  \end{corollary1}
+
+ For unstable models, Corollary~\ref{sign} shows that shifts in parameters around zero (i.e., from $\boldsymbol \theta_N$ to $-\boldsymbol \theta_N$) can induce extreme changes in probability among subsets of the sample space, as another manifestation  of instability
+ and hyper-sensitivity in probability structure.
+For one-parameter exponential families, involving a fixed real-valued linear parameter  $\boldsymbol \theta_N = \theta \in \mathbb{R}$ and sufficient statistic $\boldsymbol g_N(\boldsymbol x)\in \mathbb{R}$ in (\ref{eqn:expo}),  Schweinberger~(\protect\hyperlink{ref-schweinberger2011instability}{2011}, Theorem~3) proved a result similar in spirit, though based on a characterization there in terms of maximum $U_N \equiv \max_{\boldsymbol x\in\mathcal{X}^N}g_N(\boldsymbol x)$  and minimal $L_N \equiv \min_{\boldsymbol x\in\mathcal{X}^N}g_N(\boldsymbol x)$ values of the sufficient statistic.
+For this case in particular,  mode sets have specific, and essentially complementary, forms  over positive and negative parameters, namely, $\mathcal{M}_{\epsilon, \boldsymbol \theta_N} = \{\boldsymbol x \in\mathcal{X}^N: g_N(\boldsymbol x)  >  (1-\epsilon) U_N + \epsilon L_N \}$ and $\mathcal{M}_{\epsilon, -\boldsymbol \theta_N} = \{\boldsymbol x \in\mathcal{X}^N: g_N(\boldsymbol x)  <  \epsilon U_N + (1-\epsilon) L_N \}$ for any $\boldsymbol \theta_N>0$, and
+ Schweinberger~(\protect\hyperlink{ref-schweinberger2011instability}{2011}, Theorem~3)  showed each mode set collects all mass, under positive and negative parameters, respectively, with unstable models of this exponential type.
+However, for all unstable FOES models, Corollary~\ref{sign} generalizes the same principle
+ that unstable models can push all probability to different, and in fact disjoint, parts of the sample space, depending on how parameters fall with respect to zero.
+This feature can numerically complicate likelihood manipulations, such as maximization or MCMC-based Bayes posterior sampling, as further discussed in Section~5.\\
+
+%############ begin blue marker
+\blue{
+
+\noindent \textbf{Remark 2:}  Under the model condition PSR, Corollary~\ref{sign} can also be extended to
+cases where parameter components $\boldsymbol \theta_N =(\boldsymbol \theta_{1,N}, \boldsymbol \theta_{N,2})$ (say) are not all changed in sign (e.g., $-\boldsymbol \theta_N$) but, more generally, are instead  altered to
+another parameter configuration $\boldsymbol \theta_{N}^A = ( \boldsymbol \theta_{1,N}^A, \boldsymbol \theta_{2,N}^A )$
+ involving a switch in sign only among some dominating model parameters $\boldsymbol \theta_{N,2}^A=- \boldsymbol \theta_{N,2}$ with remaining parameters $\boldsymbol \theta_{1,N}^A$ being arbitrarily chosen.
+If a change  sign occurs among  parameters ($\pm \boldsymbol \theta_{N,2}$) which dominate the probability structure of the model, then the results of Corollary~1  can still hold  with $\boldsymbol \theta_{N}^A$
+  replacing $-\boldsymbol \theta_N$; as an example of one sufficient condition, if $
+  \lim_{N\to \infty} \max_{\boldsymbol x \in \mathcal{X}^N} |G_N(\boldsymbol x, -\boldsymbol \theta_{N}) - G_N(\boldsymbol x, \boldsymbol \theta_{N}^A)|=0$ holds  in addition to Corollary~1 assumptions, where
+  \[
+  G_N(\boldsymbol x, \boldsymbol \theta) =   \frac{\log P_{\boldsymbol \theta}(\boldsymbol x) - \min_{\boldsymbol y \in\mathcal{X}^N}\log P_{\boldsymbol \theta}(\boldsymbol y) }{\max_{\boldsymbol y \in\mathcal{X}^N}\log P_{\boldsymbol \theta}(\boldsymbol y)  - \min_{\boldsymbol y \in\mathcal{X}^N}\log P_{\boldsymbol \theta}(\boldsymbol y) }, \quad \boldsymbol x \in\mathcal{X}^N,
+  \]
+ represents a standardized form of $\boldsymbol \theta$-model probabilities, then the results of Corollary~1 apply to 
+  $\boldsymbol \theta_{N}^A$ in addition to $-\boldsymbol \theta_N$.
+  As a consequence, an unstable model under $\boldsymbol \theta_N$ can then imply that many more unstable models exist  over a broader spectrum of possibilities for variations $\boldsymbol \theta_{N}^A$ of $\boldsymbol \theta_N$, which involves some amount of sign change among components of $\boldsymbol \theta_N$.
+
+%###################### end
+}
+\section{Illustrations}
+  Model instability can  depend intricately on how functions of parameters and data $\boldsymbol X_N=(X_1,\ldots,X_n)$ are combined in the formulation of the model probabilities, though some general causes may be identified.
+As one issue, a broad parameter space (or wide interpretation of this space) may admit some parameters as technically valid that have an undue and often undesirable  impact   on  the model structure for a prescribed data size $N$.
+In this case, both the size and dimension of model parameters can be problematic and induce instability.
+In combination to this last point, further causes of instability may also be traced to  the magnitude of statistics in the model.
+Potentially massive, and thereby unstable,   statistics were the primary focus
+  of instability studies of Schweinberger
+(\protect\hyperlink{ref-schweinberger2011instability}{2011}) for certain discrete exponential models having parameters/statistics of fixed dimension.
+However, as shown in the following, bounded statistics may still lead to instability if the  parameter dimension is high.
+   We next provides some examples to illustrate instability in FOES models, which also suggest  some potential strategies for preventing unstable models.
+
+\subsection{Equi-probability Models}  As a baseline for comparisons, consider a simplistic model for $\boldsymbol X_N=(X_1,\ldots,X_N)$ with uniform probabilities over the sample space, say
+$P_{\boldsymbol \theta_N}(\boldsymbol x)= |\mathcal{X}|^{-N}$, $\boldsymbol x \in \mathcal{X}^N$, where each random variable has $|\mathcal{X}| \geq 1$ outcomes.
+In contrast to instability, model probabilities here are completely insensitive to changes in data outcomes across the sample space, and the associated log-ratio of extreme probabilities (\ref{eq:elpr}) is
+\[
 \frac{1}{N} \REP(\boldsymbol \theta_N)=0\quad \;(\mbox{uniform probability model}),
-\] which is as small as possible. In fact, a LREP value of zero can only
-occur for a FOES model having uniform probabilities, and such
-equi-probability models are always S-stable.
-
-\subsection{One-parameter exponential models}\label{one-param-exp}
-
-A fundamental model considered in the instability work of Schweinberger
-(\protect\hyperlink{ref-schweinberger2011instability}{2011}) involves a
-one-parameter exponential model corresponding to \eqref{eq:expo} with a
-real-valued parameter, say
-\(\boldsymbol \theta_N = \eta(\boldsymbol \theta_N)\in \mathbb{R}\), and
-sufficient statistic \(\boldsymbol g_N(\boldsymbol x)\in \mathbb{R}\).
-For such models, upon scaling by sample size \(N\), the log-ratio of
-extreme probabilities in \eqref{eq:elpr} for assessing instability becomes
+\] which is as small as possible.  In fact, a LREP value of zero can only occur for a FOES model  having uniform probabilities, and such equi-probability models are always S-stable.
+
+\subsection{One-parameter Exponential Models}  A fundamental model considered in the instability work of Schweinberger~(\protect\hyperlink{ref-schweinberger2011instability}{2011}) involves a one-parameter exponential model corresponding to (\ref{eqn:expo}) with a real-valued parameter, say $\boldsymbol \theta_N =  \eta(\boldsymbol \theta_N)\in \mathbb{R}$, and sufficient statistic $\boldsymbol g_N(\boldsymbol x)\in \mathbb{R}$.
+For such models,  upon scaling by sample size $N$, the log-ratio of extreme probabilities in (\ref{eq:elpr}) for assessing instability   becomes
  \begin{equation}
@@
  \end{equation}
-where
-\(U_N \equiv \max_{\boldsymbol x\in\mathcal{X}^N}g_N(\boldsymbol x)\)
-and
-\(L_N \equiv \min_{\boldsymbol x\in\mathcal{X}^N}g_N(\boldsymbol x)\)
-denote the maximal and minimal values of the single sufficient
-statistic. In this case, an S-unstable model results, by definition
-\eqref{eq:Sun}, whenever
-\(\lim_{N\to \infty} |\boldsymbol \theta_N| (U_N-L_N)/N= \infty\) holds
-or, in other words, if the combined magnitudes of parameter
-\(|\boldsymbol \theta_N|\) and maximal difference \(U_N-L_N\) in
-statistic values are overwhelmingly large relative to the sample size
-\(N\). If we further assume that
-\(\boldsymbol \theta_N =\theta\in\mathbb{R}\setminus \{0\}\) is a fixed
-(non-zero) parameter for all \(N \geq 1\), as considered in
-Schweinberger
-(\protect\hyperlink{ref-schweinberger2011instability}{2011}), then an
-S-unstable model results solely if the sufficient statistic admits a
-value \(U_N-L_N\) too large relative to number \(N\) of observations,
-i.e., if \((U_N-L_N)/N\to \infty\) as \(N\to \infty\). The latter aspect
-reflects the definition of Schweinberger
-(\protect\hyperlink{ref-schweinberger2011instability}{2011}), for this
-setting, that a real-valued \emph{statistic} \(g_N(\boldsymbol x)\) may
-be classified as \emph{unstable} when
-\(\lim_{N\to \infty}|(U_N-L_N)/N=\infty\) holds and as \emph{stable}
-otherwise (e.g., if \(\sup_{N \geq 1}(U_N-L_N)/N<\infty\)).
-
-For illustration, consider the iid Bernoulli model \eqref{eq:mod1} for
-\(\boldsymbol X_N=(X_1,\ldots,X_N)\) with log-odds ratio parameter
-\(\boldsymbol \theta_N = \log[ P_{\boldsymbol \theta_N}(X_1=1)/ P_{\boldsymbol \theta_N}(X_1=0)]\in\mathbb{R}\).
-Remark 1 (Section \ref{criterion}) then gives the model instability
-measure \eqref{eq:Sun} directly as \[
+  where $U_N \equiv \max_{\boldsymbol x\in\mathcal{X}^N}g_N(\boldsymbol x)$  and $L_N \equiv \min_{\boldsymbol x\in\mathcal{X}^N}g_N(\boldsymbol x)$ denote the maximal and minimal values of the single sufficient statistic.
+    In this case, an S-unstable model results, by definition (\ref{eqn:Sun}), whenever
+  $\lim_{N\to \infty} |\boldsymbol \theta_N| (U_N-L_N)/N= \infty$ holds or, in order words, if the combined  magnitudes of parameter $|\boldsymbol \theta_N|$ and maximal difference $U_N-L_N$ in statistic values are overwhelmingly large relative to the sample size $N$.
+  If we further assume that
+   $\boldsymbol \theta_N =\theta\in\mathbb{R}\setminus \{0\}$ is a fixed (non-zero) parameter for all $N \geq 1$,  as considered in Schweinberger~(\protect\hyperlink{ref-schweinberger2011instability}{2011}), then an S-unstable model  results solely if the sufficient statistic admits a value   $ U_N-L_N $ too large relative to number $N$ of observations, i.e.,  if  $(U_N-L_N)/N\to \infty$ as $N\to \infty$.
+The latter aspect reflects the definition of Schweinberger~(\protect\hyperlink{ref-schweinberger2011instability}{2011}), for this setting, that a real-valued {\it statistic} $g_N(\boldsymbol x)$ may be classified as  {\it unstable} when  $\lim_{N\to \infty}|(U_N-L_N)/N=\infty$ holds and as {\it stable} otherwise (e.g., if $\sup_{N \geq 1}(U_N-L_N)/N<\infty$).
+
+    For illustration, consider the iid Bernoulli model (\ref{eqn:mod1}) for $\boldsymbol X_N=(X_1,\ldots,X_N)$ with log-odds ratio parameter $\boldsymbol \theta_N = \log[ P_{\boldsymbol \theta_N}(X_1=1)/ P_{\boldsymbol \theta_N}(X_1=0)]\in\mathbb{R}$.
+Remark 1 (Section~\ref{criterion}) then gives the model instability measure (\ref{eqn:Sun}) directly as
+      \[
      \frac{1}{N}\REP(\boldsymbol \theta_N ) = |\boldsymbol \theta_N|\quad\; \mbox{(iid Bernoulli model)},
-\] so that an unstable (or stable) model results for a divergent (or
-bounded) parameter sequence \(|\boldsymbol \theta_N|\). The above
-instability expression for the Bernoulli model follows as well from the
-\(N\)-scaled LREP value \eqref{eq:UL} for a one-parameter exponential
-distribution, using that the sufficient statistic involved
-\(g_N(\boldsymbol x)= \sum_{i=1}^N x_i\),
-\(\boldsymbol x =(x_1,\ldots,x_n)\in\{0,1\}^N\), has maximum and mimimum
-values \(U_N=N\) and \(L_N=0\). In this case, Schweinberger
-(\protect\hyperlink{ref-schweinberger2011instability}{2011}) has noted
-that the statistic is stable (i.e., bounded \((U_N-L_N)/N=1\)) and the
-Bernoulli model is as well when, in particular,
-\(\boldsymbol \theta_N=\bold \theta \in\mathbb{R}\) is fixed for
-\(N \geq 1\).
-
-Alternatively, considering a random graph with \(N={n \choose 2}\) edges
-among \(n\) nodes, the exponential graph model from \eqref{eq:mod2}, when
-based purely on the number of \(g_{2,N}(\boldsymbol x)\) of 2-stars or
-solely the number \(g_{3,N}(\boldsymbol x)\) of triangles,
-\(\boldsymbol x\in\{0,1\}^N\), has an measure of instability from
-\eqref{eq:Sun} as \[
+   \]
+    so that an unstable (or stable) model results for a divergent (or bounded) parameter sequence  $|\boldsymbol \theta_N|$.
+The above   instability expression  for the Bernoulli model  follows as well from  the $N$-scaled LREP value (\ref{eq:UL}) for a one-parameter exponential distribution, using that the sufficient statistic involved $g_N(\boldsymbol x)= \sum_{i=1}^N x_i$, $\boldsymbol x =(x_1,\ldots,x_n)\in\{0,1\}^N$, has maximum and mimimum values $U_N=N$ and $L_N=0$.
+In this case, Schweinberger~(\protect\hyperlink{ref-schweinberger2011instability}{2011}) has noted that the statistic is stable (i.e., bounded $(U_N-L_N)/N=1$) and the Bernoulli model is as well when, in particular, $\boldsymbol \theta_N=\bold \theta \in\mathbb{R}$ is fixed for $N \geq 1$.
+       Alternatively, considering a random graph with $N={n \choose 2}$ edges among $n$ nodes, the exponential graph model from (\ref{eqn:mod2}), when based purely on the number of $g_{2,N}(\boldsymbol x)$ of 2-stars  or solely the number $g_{3,N}(\boldsymbol x)$  of triangles,  $\boldsymbol x\in\{0,1\}^N$, has an measure of instability from (\ref{eqn:Sun}) as
+   \[
      \frac{1}{N}\REP(\boldsymbol \theta_N )  = \left\{ \begin{array}{lcl} |\boldsymbol \theta_N| (n-2) && \mbox{(2-star graph model)}\\
        |\boldsymbol \theta_N|(n-2)/3 &&\mbox{(triangle graph model)},\end{array}\right.
-\] by using the (one-parameter exponential) LREP formula \eqref{eq:UL}
-with statistic maximums \(U_N= N(n-2)\) for 2-stars or \(U_N= N(n-2)/3\)
-for triangles and with minimums \(L_N=0\) in both cases. Because the
-variable number \(N\to \infty\) as the node number \(n\to \infty\), both
-counts of 2-stars and triangles are unstable statistics in the sense of
-Schweinberger
-(\protect\hyperlink{ref-schweinberger2011instability}{2011}) (i.e.,
-\(\lim_{N\to \infty} (U_N-L_N)/N=\infty\)). Furthermore, both types of
-graph models are always S-unstable for all possible of parameter
-sequences \(\boldsymbol \theta_N \in\mathbb{R}\) that are bounded away
-from zero (i.e.,
-\(\lim_{N\to \infty}\REP(\boldsymbol \theta_N )/N=\infty\) then holds,
-including the fixed parameter case
-\(\boldsymbol \theta_N=\theta\in\mathbb{R}\setminus \{0\}\) from
-Schweinberger
-(\protect\hyperlink{ref-schweinberger2011instability}{2011})).
-
-\subsection{Fixed-dimensional linear exponential
-models}\label{fixed-dim-exp}
-
-As a generalization of the one-parameter exponential case, we next
-consider linear exponential families \eqref{eq:expo} with \(k\) parameters
-\(\boldsymbol \theta_N = (\theta_{1,N},\ldots,\theta_{k,N})^\prime\) and
-\(k\) sufficient statistics
-\(\boldsymbol g_N(\boldsymbol x) = (g_{1,N}(\boldsymbol x),\ldots, g_{k,N}(\boldsymbol x))^\prime\).
-Here the dimension \(k\) of model parameters/statistics is fixed, and we
-next prescribe a condition helpful to avoiding instability in such
-models. For this, define
-\(U_{i,N}=\max_{\boldsymbol x \in\mathcal{X}^N} g_{i,N}(\boldsymbol x)\)
-and
-\(L_{i,N}=\min_{\boldsymbol x \in\mathcal{X}^N} g_{i,N}(\boldsymbol x)\)
-as the maximal and minimal values of the \(i\)th statistic,
-\(i=1,\ldots,k\), based on observations
-\(\boldsymbol X_N=(X_1,\ldots,X_N)\).
-
-\BeginKnitrBlock{proposition}
-\protect\hypertarget{prp:prop1}{}{\label{prp:prop1}}Let
-\(P_{\boldsymbol \theta_N}\), \(N \geq 1\), denote linear exponential
-models (\ref{eq:expo}) with parameters
-\(\boldsymbol \theta_N = (\theta_{1,N},\ldots,\theta_{k,N})^\prime \in \mathbb{R}^k\)
-and statistics
-\(\boldsymbol g_N(\boldsymbol x) = (g_{1,N}(\boldsymbol x),\ldots, g_{k,N}(\boldsymbol x))^\prime \in \mathbb{R}^k\),
-for fixed \(k \geq 1\). Then, the models \(P_{\boldsymbol \theta_N}\)
-are \{\rm S-stable\} if
+     \]
+  by using the (one-parameter exponential) LREP formula (\ref{eq:UL}) with statistic maximums $U_N= N(n-2)$ for 2-stars or $U_N= N(n-2)/3$ for triangles  and with minimums $L_N=0$ in both cases.
+Because the variable number $N\to \infty$ as the node number $n\to \infty$, both counts of 2-stars and triangles are unstable statistics in the sense of Schweinberger~(\protect\hyperlink{ref-schweinberger2011instability}{2011}) (i.e., $\lim_{N\to \infty} (U_N-L_N)/N=\infty$).
+Furthermore, both types of graph models are always S-unstable for all possible of parameter sequences
+   $ \boldsymbol \theta_N \in\mathbb{R}$ that are bounded away from zero (i.e.,  $\lim_{N\to \infty}\REP(\boldsymbol \theta_N )/N=\infty$ then  holds, including the fixed parameter case $\boldsymbol \theta_N=\theta\in\mathbb{R}\setminus \{0\}$  from Schweinberger~(\protect\hyperlink{ref-schweinberger2011instability}{2011})).
+
+ \subsection{Fixed-Dimensional Linear Exponential Models}
+  As a generalization of the one-parameter exponential case, we next consider  linear exponential families  (\ref{eqn:expo}) with $k$
+   parameters $\boldsymbol \theta_N = (\theta_{1,N},\ldots,\theta_{k,N})^\prime$   and $k$ sufficient statistics
+   $\boldsymbol g_N(\boldsymbol x) = (g_{1,N}(\boldsymbol x),\ldots, g_{k,N}(\boldsymbol x))^\prime$.
+Here the dimension $k$ of model  parameters/statistics is fixed, and we next prescribe a condition helpful to avoiding instability in such models.
+For this, define  $U_{i,N}=\max_{\boldsymbol x \in\mathcal{X}^N} g_{i,N}(\boldsymbol x)$ and $L_{i,N}=\min_{\boldsymbol x \in\mathcal{X}^N} g_{i,N}(\boldsymbol x)$ as the maximal and minimal values of the $i$th statistic, $i=1,\ldots,k$, based on observations $\boldsymbol X_N=(X_1,\ldots,X_N)$.
+\newtheorem{proposition1}{Proposition}
+
+  \begin{proposition1}
+  \label{prop1} {\it Let $P_{\boldsymbol \theta_N}$, $N \geq 1$,  denote   linear exponential models   (\ref{eqn:expo})
+  with   parameters $\boldsymbol \theta_N = (\theta_{1,N},\ldots,\theta_{k,N})^\prime \in \mathbb{R}^k$ and statistics  $\boldsymbol g_N(\boldsymbol x) = (g_{1,N}(\boldsymbol x),\ldots, g_{k,N}(\boldsymbol x))^\prime \in \mathbb{R}^k$, for fixed $k \geq 1$.
+Then, the models
+  $P_{\boldsymbol \theta_N}$ are {\rm S-stable}   if
   \begin{equation}
-\label{eq:prop1}
-\sup_{N \geq 1}\frac{1}{N} \max_{1 \leq i \leq k }|\theta_{i,N}|(U_{i,N}-L_{i,N})<\infty
+  \label{eqn:prop1}
+  \sup_{N \geq 1}\frac{1}{N} \max_{1 \leq i \leq k } |\theta_{i,N}|(U_{i,N}-L_{i,N})<\infty
   \end{equation}
-holds, i.e., if
-\(\max_{1 \leq i \leq k } |\theta_{i,N}|(U_{i,N}-L_{i,N})/N\) is bounded
-sequence of sample size \(N\).
-\EndKnitrBlock{proposition}
-
-\BeginKnitrBlock{remark}
-\iffalse <span class="remark"><em>Remark. \fi{} In the one-parameter
-exponential case \(k=1\), recall the exponential model is
-stable/unstable depending on whether
-\(\REP(\boldsymbol \theta_N)/N = |\theta_{1,N}|(U_{1,N}-L_{1,N})/N \equiv |\boldsymbol \theta_{N}|(U_{N}-L_{N})/N\)
-in \eqref{eq:UL} is convergent/divergent. Hence, for \(k=1\), the
-condition \eqref{eq:prop1} of Proposition \ref{prp:prop1} captures the
-same notion of S-stability based on \eqref{eq:UL}.
-\EndKnitrBlock{remark}
-
-Proposition \ref{prp:prop1} provides a sufficient condition for the
-stability of linear exponential models with fixed parameter dimension
-\(k\geq 1\), whereby an S-stable model is guaranteed if the compounded
-magnitude of each combination of parameter \(\theta_{i,N}\) and
-sufficient statistic value \((U_{i,N}-L_{i,N})\) is bounded by the
-sample size \(N\), \(i=1,\ldots,k\). This supports the findings of
-Schweinberger
-(\protect\hyperlink{ref-schweinberger2011instability}{2011}), who showed
-degeneracy follows in such models under one type of violation of the
-condition \eqref{eq:prop1} in Proposition \ref{prp:prop1} (namely,
-involving \(k>1\) non-zero parameters with \(k-1\) statistics being
-\(O(N)\) bounded while one statistic diverges in maximal size faster
-than the number \(N\) of observations). To further illustrate the result
-in Proposition \ref{prp:prop1}, consider the multinomial distribution
-\eqref{eq:mod11} for \(\boldsymbol X_N=(X_1,\ldots,X_N)\) having
-\(k\geq 2\) categories \(\{1,\ldots,k\}\) and \(k\) parameters
-\(\boldsymbol \theta_N = (\theta_{1,N},\ldots,\theta_{k,N})^\prime\).
-The variables are iid under this model so that Remark 1 (Section
-\ref{criterion} yields the corresponding \(N\)-scaled log-ratio of
-extreme probabilities \eqref{eq:elpr} as
-\begin{align*}
-  \frac{1}{N}\REP(\boldsymbol \theta_N) &= \frac{\max_{1 \leq i \leq k} P_{\boldsymbol \theta_N}(X_1=i)}{\min_{1 \leq i \leq k} P_{\boldsymbol \theta_N}(X_1=i)}\\
-  &= \max_{1 \leq i \leq k} \theta_{i,N} - \min_{1 \leq i \leq k} \theta_{i,N} \qquad \mbox{(iid multinomial model)}.
-\end{align*}
-Hence, a multinomial model sequence is unstable (or stable) depending on
-whether (or not) the maximal parameter difference
-\(\max_{1 \leq i \leq k} \theta_{i,N} - \min_{1 \leq i \leq k} \theta_{i,N}\)
-diverges. Furthermore, using that each of the \(k\) sufficient (count)
-statistics from the multinomial model \eqref{eq:mod11} satisfies
-\((U_{i,N}-L_{i,N})/N=1\), we see that \eqref{eq:prop1} of Proposition
-\ref{prp:prop1} becomes purely a parameter condition,
-\(\sup_{N \geq 1}\max_{1 \leq i \leq k } |\theta_{i,N}| <\infty\), for
+  holds, i.e.,    if $\max_{1 \leq i \leq k } |\theta_{i,N}|(U_{i,N}-L_{i,N})/N $ is bounded sequence of sample size $N$.}
+\end{proposition1}
+\noindent\textbf{Remark 3:}  In the one-parameter exponential case $k=1$, recall the exponential model is stable/unstable depending on whether $\REP(\boldsymbol \theta_N)/N = |\theta_{1,N}|(U_{1,N}-L_{1,N})/N \equiv |\boldsymbol \theta_{N}|(U_{N}-L_{N})/N$ in (\ref{eq:UL}) is  convergent/divergent.
+Hence, for $k=1$, the condition (\ref{eqn:prop1}) of Proposition~\ref{prop1} captures the same notion of S-stability based on (\ref{eq:UL}).
+\\
+
+Proposition~\ref{prop1} provides a sufficient condition for the stability of linear exponential models with fixed parameter dimension $k\geq 1$,
+whereby an S-stable model is guaranteed if the compounded magnitude of each combination of parameter $\theta_{i,N}$ and sufficient statistic value $(U_{i,N}-L_{i,N})$  is bounded by the sample size $N$, $i=1,\ldots,k$.
+This  supports the findings of
+Schweinberger~(\protect\hyperlink{ref-schweinberger2011instability}{2011}), who showed degeneracy follows in such models under one type of violation of the condition   (\ref{eqn:prop1}) in Proposition~\ref{prop1} (namely, involving $k>1$ non-zero parameters  with $k-1$ statistics being $O(N)$ bounded  while one statistic  diverges in maximal size faster than the number  $N$ of observations).
+To further illustrate the result in Proposition~\ref{prop1}, consider  the multinomial distribution (\ref{eqn:mod11}) for $\boldsymbol X_N=(X_1,\ldots,X_N)$ having $k\geq 2$ categories $\{1,\ldots,k\}$ and $k$ parameters $\boldsymbol \theta_N = (\theta_{1,N},\ldots,\theta_{k,N})^\prime$.
+The variables are iid under this model so that Remark~1 (Section~\ref{criterion}) yields the corresponding $N$-scaled  log-ratio of extreme probabilities (\ref{eq:elpr}) as
+\begin{eqnarray*}
+  \frac{1}{N}\REP(\boldsymbol \theta_N) &=& \frac{\max_{1 \leq i \leq k} P_{\boldsymbol \theta_N}(X_1=i)}{\min_{1 \leq i \leq k} P_{\boldsymbol \theta_N}(X_1=i)}\\& =& \max_{1 \leq i \leq k} \theta_{i,N} - \min_{1 \leq i \leq k} \theta_{i,N} \qquad \mbox{(iid multinomial model)}.
+\end{eqnarray*}
+Hence, a multinomial model sequence is unstable (or stable) depending on whether (or not) the maximal parameter difference  $\max_{1 \leq i \leq k} \theta_{i,N} - \min_{1 \leq i \leq k} \theta_{i,N}$ diverges.
+Furthermore, using that each of the $k$ sufficient (count) statistics from the multinomial model  (\ref{eqn:mod11}) satisfies $(U_{i,N}-L_{i,N})/N=1$,  we  see that  (\ref{eqn:prop1}) of Proposition~\ref{prop1}  becomes purely a parameter condition,
+$\sup_{N \geq 1}\max_{1 \leq i \leq k } |\theta_{i,N}| <\infty$, for
   ensuring that
-\(\REP(\boldsymbol \theta_N)/N =\max_{1 \leq i \leq k} \theta_{i,N} - \min_{1 \leq i \leq k} \theta_{i,N}\)
-is bounded and stability follows for the multinomial distribution.
-Additionally, a stable multinomial sequence (i.e., bounded
-\(\REP(\boldsymbol \theta_N)/N\)) turns out to be nearly equivalent to
-\eqref{eq:prop1} (e.g., these are the same if the smallest parameter
-\(\min_{1 \leq i \leq k } |\theta_{i,N}|\) remains bounded).
-
-When the condition \eqref{eq:prop1} of Proposition \ref{prp:prop1} is
-violated, this aspect suggests a potentially unstable model that may be
-investigated more closely. For example, consider the exponential graph
-model from \eqref{eq:mod2} involving counts of edges, 2-stars and
-triangles with fixed parameters
-\(\boldsymbol \theta_N = (\theta_{1},\theta_2,\theta_3)^\prime \in \mathbb{R}^3\)
-for \(N\geq 1\). If either the 2-star parameter \(\theta_2 \neq 0\) or
-triangle parameter \(\theta_3 \neq 0\) is non-zero, then
-\(\max_{1 \leq i \leq 3 } |\theta_{i}|(U_{i,N}-L_{i,N})/N \propto (n-2)\to \infty\)
-holds in \eqref{eq:prop1} by
-\((U_{2,N}-L_{2,N})/N = 3 (U_{3,N}-L_{3,N})/N=(n-2)\) for 2-star and
-triangle statistics (\(i=2,3\)), so that Proposition \ref{prp:prop1}
-hints that an unstable model may result when
-\(|\theta_2| + |\theta_3| \neq 0\). Relatedly, a result from
-Schweinberger
-(\protect\hyperlink{ref-schweinberger2011instability}{2011} Result 3)
-states that this model is unstable for all fixed parameters excluding
-cases \(\theta_2 =\theta_3=0\) or \(\theta_2 = - \theta_3/3\). However,
-more is true in line with the instability suggested by Proposition
-\ref{prp:prop1} whenever \(|\theta_2| + |\theta_3| \neq 0\) (i.e.,
-excluding \(\theta_2 =\theta_3=0\)).
-
-To see this, consider an even number \(n>2\) of nodes and let
-\(\boldsymbol x_0\) denote the data outcome in
-\(\mathcal{X}^N \equiv \{0,1\}^N\) with all \(N = {n \choose 2}\) edges
-being zero, let \(\boldsymbol x_1\) denote the outcome with all edges
-being 1, and let \(\boldsymbol x_2\) denote the edge configuration from
-dividing the nodes into two equal groups, with no edges within a group
-and all edges between the groups (so that no triangles exist in
-\(\boldsymbol x_2\)). Then, the \(N\)-scaled log-ratio \eqref{eq:elpr} for
-the exponential graph model \eqref{eq:mod2} can, by definition, be bounded
-below by
-\begin{align*}
-\frac{1}{N}\REP_N(\boldsymbol \theta_N) &\geq \max_{i=1,2}\frac{1}{N}
+$\REP(\boldsymbol \theta_N)/N =\max_{1 \leq i \leq k} \theta_{i,N} - \min_{1 \leq i \leq k} \theta_{i,N}$ is bounded
+and  stability follows
+for the multinomial distribution.
+Additionally, a stable multinomial sequence (i.e., bounded $\REP(\boldsymbol \theta_N)/N$) turns out to be nearly equivalent to (\ref{eqn:prop1})  (e.g., these are the same if the smallest  parameter   $\min_{1 \leq i \leq k } |\theta_{i,N}| $ remains bounded).
+
+ When the condition (\ref{eqn:prop1}) of Proposition~\ref{prop1} is violated, this aspect suggests a potentially unstable model that may be investigated more closely.
+ For example,  consider the exponential graph model from (\ref{eqn:mod2}) involving counts of edges, 2-stars and triangles
+    with fixed parameters $\boldsymbol \theta_N = (\theta_{1},\theta_2,\theta_3)^\prime \in \mathbb{R}^3$ for $N\geq 1$.
+    If either the 2-star parameter $\theta_2 \neq 0$ or triangle parameter $\theta_3 \neq 0$ is non-zero, then   $ \max_{1 \leq i \leq 3 } |\theta_{i}|(U_{i,N}-L_{i,N})/N \propto (n-2)\to \infty$ holds in (\ref{eqn:prop1}) by  $(U_{2,N}-L_{2,N})/N = 3 (U_{3,N}-L_{3,N})/N=(n-2)$
+     for 2-star and triangle statistics ($i=2,3$),
+      so that Proposition~\ref{prop1} hints that an unstable model  may result when $|\theta_2| + |\theta_3| \neq 0$.
+   Relatedly, a result from Schweinberger~(\protect\hyperlink{ref-schweinberger2011instability}{2011}, Result 3) states that this model is unstable
+    for all fixed parameters excluding cases $\theta_2 =\theta_3=0$ or $\theta_2 = - \theta_3/3$.
+However, more is true in line with the instability suggested by Proposition~\ref{prop1} whenever $|\theta_2| + |\theta_3| \neq 0$ (i.e., excluding $\theta_2 =\theta_3=0$).
+      To see this, consider an even number $n>2$ of nodes and let $ \boldsymbol x_0 $ denote the data outcome in $\mathcal{X}^N \equiv \{0,1\}^N$ with all $N = {n \choose 2}$ edges being zero,
+       let $ \boldsymbol x_1 $ denote the outcome with all edges being 1, and let
+      $ \boldsymbol x_2$  denote  the edge configuration from dividing the nodes into two equal groups, with no edges within a group and all edges between the groups (so that no triangles exist in $ \boldsymbol x_2$).
+Then, the $N$-scaled  log-ratio (\ref{eq:elpr}) for the exponential graph model   (\ref{eqn:mod2}) can, by definition, be bounded below by
+    \begin{eqnarray*}
+     \frac{1}{N}\REP_N(\boldsymbol \theta_N) & \geq& \max_{i=1,2}\frac{1}{N}
      \left| \log\left[ \frac{P_{\boldsymbol \theta_N}(\boldsymbol x_i)}{P_{\boldsymbol \theta_N}(\boldsymbol x_0)}\right] \right| \\
-&= (n-2) \max\left\{ \left| \theta_2 + \frac{\theta_3}{3}+\frac{\theta_1}{n-2} \right|, \frac{n}{4(n-1)} \left| \theta_2 + \frac{8\theta_1}{n-2} \right| \right\};
-\end{align*}
-a similar expression also holds for an odd node number \(n>2\).
-Consequently, for all fixed parameters excluding
-\(\theta_2=\theta_3=0\),
-\(\lim_{N\to \infty}\REP_N(\boldsymbol \theta_N)/N=\infty\) then follows
-and the graph model with 2-stars and triangles is S-unstable, as
-suggested by the breach of Proposition \ref{prp:prop1} for this model
-when \(|\theta_2|+|\theta_3|\neq 0\). That is, instability holds even
-under \(\theta_2 = - \theta_3/3\) case potentially allowed by
-Schweinberger's results.
-
-\subsection{Latent variable models of increasing parameter
-dimension}\label{latent-variable-models-of-increasing-parameter-dimension}
-
-We next consider instability of discrete data models based on
-exponential formulations involving hidden, or latent, variables, such as
-those described in Sections
-\ref{restricted-boltzmann-machines}-\ref{deep-learning} for networks. We
-will focus on restricted Bolzmann machine (RBM) models (Section
-\ref{restricted-boltzmann-machines}, having one layer of latent
-variables for simplicity, though the same instability concepts may be
-extended to other deep learning models (Section \ref{deep-learning}. For
-\(N\) visible variables
-\(\boldsymbol X \equiv \boldsymbol X_N = (X_1,\ldots,X_n)\) as data,
-each observation \(X_i\in\{\pm 1\}\) being binary, the RBM-based model
-\eqref{eq:RBM2} for \(\boldsymbol X\) is again of FOES-type, though not an
-exponential model. However, the distribution of visible variables is
-induced by an underlying joint exponential model \eqref{eq:RBM1} for both
-visible and latent variables \((\boldsymbol X, \boldsymbol H)\), where
-\(\boldsymbol H=(H_1,\ldots,H_{N_{\mathcal{H}}})\) denotes a vector of
-\(N_{\mathcal{H}}\) hidden variables (similarly binary). The joint model
-is of linear exponential form involving
-\(d(N)\equiv N + N_{\mathcal{H}} + N* N_{\mathcal{H}}\) sufficient
-statistics given by
-\((\boldsymbol X, \boldsymbol H, \boldsymbol X^T\boldsymbol H)\) and
-parameters
-\(\boldsymbol \theta_N = (\boldsymbol \theta_N^{\mathcal{V}},\boldsymbol \theta_N^{\mathcal{H}}, \boldsymbol \theta_N^{\mathcal{VH}} ) \in\mathbb{R}^{d(N)}\)
-corresponding to the \(N\) visible variables \(\boldsymbol X\) (i.e.,
-\(\boldsymbol \theta_N^{\mathcal{V}}\in\mathbb{R}^N\)), the
-\(N_{\mathcal{H}}\) hidden variables \(\boldsymbol H\) (i.e.,
-\(\boldsymbol \theta_N^{\mathcal{H}}\in\mathbb{R}^{N_{\mathcal{H}}}\)),
-and the \(N *N_{\mathcal{H}}\) cross-product variables
-\(\boldsymbol X^T\boldsymbol H\) (i.e.,
-\(\boldsymbol \theta_N^{\mathcal{VH}}\in\mathbb{R}^{N N_{\mathcal{H}}}\)).
-However, unlike some previous exponential models considered in Sections
-\ref{one-param-exp}-\ref{fixed-dim-exp} (cf.~Proposition
-\ref{prp:prop1}, note that the RBM formulation always associates
-parameters with \emph{bounded} statistics (i.e., the components of
-\((\boldsymbol X, \boldsymbol H, \boldsymbol X^T\boldsymbol H)\)) so
-that model instability cannot arise here due to the magnitude of
-sufficient statistics exceeding the sample size \(N\). Instead, RBM
-instability may be linked to solely to parameter configuration and the
-fact that the number \(d(N) \geq N\) of parameters necessarily increases
-with the number \(N\) of observations \(\boldsymbol X\), in contrast to
-previous exponential cases of fixed parameter dimension.
-
-To highlight the instability issues for the RBM model, consider a simple
-model for \(N\) visibles \(\boldsymbol X\) with no hidden variables
-(\(N_{\mathcal{H}}=0\)), for which model statements
-\eqref{eq:RBM1}-\eqref{eq:RBM2} coincide. An independence model then results
-for variables in \(\boldsymbol X\), which has \(d(N)=N\) parameters
-\(\boldsymbol \theta_N^{\mathcal{V}} = (\theta_{1,N}^{\mathcal{V}}, \ldots, \theta_{N,N}^{\mathcal{V}}) \in \mathbb{R}^{N}\),
-and the measure of model instability becomes \[
+      &=& (n-2) \max\left\{ \left| \theta_2 + \frac{\theta_3}{3}+\frac{\theta_1}{n-2} \right|, \frac{n}{4(n-1)} \left| \theta_2 + \frac{8\theta_1}{n-2} \right| \right\};
+     \end{eqnarray*}
+a similar expression also holds for an odd node number $n>2$.
+    Consequently,  for all fixed parameters excluding $\theta_2=\theta_3=0$,   $\lim_{N\to \infty}\REP_N(\boldsymbol \theta_N)/N=\infty$ then follows and the graph model with 2-stars and triangles is S-unstable, as suggested by the breach of Proposition~\ref{prop1} for this model
+     when $|\theta_2|+|\theta_3|\neq 0$.  That is, instability holds even under  $\theta_2 = - \theta_3/3$ case potentially allowed by Schweinberger's results.
+
+   \subsection{Latent Variable Models of Increasing Parameter Dimension}  We next consider instability of discrete data models  based on exponential  formulations involving hidden, or latent, variables, such as those described in Sections~\ref{restricted-boltzmann-machines}-\ref{deep-learning} for networks.
+We will focus on restricted Bolzmann machine (RBM) models
+(Section~\ref{restricted-boltzmann-machines}), having one layer of latent variables for simplicity,  though the same  instability concepts may be extended to other deep learning models  (Section~\ref{deep-learning}).
+For $N$ visible variables $\boldsymbol X \equiv \boldsymbol X_N = (X_1,\ldots,X_n)$ as data,  each observation $X_i\in\{\pm 1\}$ being binary,
+the RBM-based model (\ref{eqn:RBM2}) for  $\boldsymbol X$ is again of FOES-type, though not an exponential model.  However, the distribution of
+visible variables is induced by an underlying joint exponential model  (\ref{eqn:RBM1})  for both visible and latent variables $(\boldsymbol X, \boldsymbol H)$, where $\boldsymbol H=(H_1,\ldots,H_{N_{\mathcal{H}}})$ denotes a vector of $N_{\mathcal{H}}$ hidden variables (similarly binary).
+ The joint model is of linear exponential form involving $d(N)\equiv N + N_{\mathcal{H}} + N* N_{\mathcal{H}}$ sufficient statistics given by $(\boldsymbol X, \boldsymbol H, \boldsymbol X^T\boldsymbol H)$ and parameters $\boldsymbol \theta_N = (\boldsymbol \theta_N^{\mathcal{V}},\boldsymbol \theta_N^{\mathcal{H}}, \boldsymbol \theta_N^{\mathcal{VH}}  ) \in\mathbb{R}^{d(N)}$ corresponding to
+ the $N$ visible variables $\boldsymbol X$ (i.e., $\boldsymbol \theta_N^{\mathcal{V}}\in\mathbb{R}^N$), the $N_{\mathcal{H}}$ hidden variables $\boldsymbol H$ (i.e., $\boldsymbol \theta_N^{\mathcal{H}}\in\mathbb{R}^{N_{\mathcal{H}}}$), and the $N *N_{\mathcal{H}}$ cross-product variables $\boldsymbol X^T\boldsymbol H$ (i.e., $\boldsymbol \theta_N^{\mathcal{VH}}\in\mathbb{R}^{N N_{\mathcal{H}}}$).
+However, unlike some previous exponential models considered in Sections~4.2-4.3 (cf.~Proposition~\ref{prop1}), note that the RBM formulation always associates parameters with {\it bounded} statistics  (i.e., the components of $(\boldsymbol X, \boldsymbol H, \boldsymbol X^T\boldsymbol H)$) so that  model instability cannot arise here
+ due to the magnitude of sufficient statistics exceeding the sample size $N$.
+Instead, RBM instability  may be linked to solely to parameter configuration and the fact that the number $d(N) \geq N$ of parameters necessarily increases with the number $N$ of observations $\boldsymbol X$,
+ in contrast to previous exponential cases of fixed parameter dimension.
+
+  To highlight the instability issues for the RBM model,
+ consider a simple  model for $N$ visibles $\boldsymbol X$ with no hidden variables ($N_{\mathcal{H}}=0$), for which model statements
+ (\ref{eqn:RBM1})-(\ref{eqn:RBM2})   coincide.  An independence model then results for variables in $\boldsymbol X$, which has
+   $d(N)=N$ parameters $\boldsymbol \theta_N^{\mathcal{V}} = (\theta_{1,N}^{\mathcal{V}},\ldots,\theta_{N,N}^{\mathcal{V}})\in\mathbb{R}^{N}$, and the measure of model instability becomes
+ \[
  \frac{1}{N}\REP(\boldsymbol \theta_N) = \frac{2}{N} \sum_{i=1}^N|\theta_{i,N}^{\mathcal{V}}| \quad \mbox{(RBM model, no hiddens)}.
-\] Hence, this model sequence for \(\boldsymbol X\) will be S-unstable
-model if the aggregation of absolute parameters grows faster than the
-number \(N\) of parameters/visible variables. Consequently, even for a
-simplest RBM model involving independence, preventing instability
-requires careful choice of parameters, particularly with regard to how a
-parameter configuration differs from zero. For more general RBM models,
-the number \(N_{\mathcal{H}}\) of hidden variables \(\boldsymbol H\) can
-also be chosen arbitrarily (i.e., as some function
-\(N_{\mathcal{H}}\equiv N_{N,\mathcal{H}}\) of \(N\)), which can
-substantially inflate the number \(d(N)\) of model parameters and
-further impact model instability through accumulated parameters. To
-better understand the effects of instability in the RBM structure,
-Proposition \ref{prp:prop2} next frames the general behavior of extreme
-probabilities in the joint RBM model \eqref{eq:RBM1} for
-\((\boldsymbol X, \boldsymbol H)\) and the implied RBM data model
-\eqref{eq:RBM2} for \(\boldsymbol X\) alone. Specifically, critical
-measures of instability may be closely connected in both models through
-tight bounds on their respective LREP values \eqref{eq:elpr}. As a result,
-Proposition \ref{prp:prop2} shows how an unstable distribution for
-observations \(\boldsymbol X\) may be traced to sources of instability
-in the original joint distribution for
-\((\boldsymbol X,\boldsymbol H)\). This also suggests a device for
-avoiding instability, as provided next.
-
-To state the result, let
-\(\REP_{\boldsymbol X}(\boldsymbol \theta_N) \equiv \REP(\boldsymbol \theta_N)\)
-denote the LREP value \eqref{eq:elpr} from the marginal distribution
-\(P_{\boldsymbol \theta_N}\) of visibles \(\boldsymbol X\) in
-\eqref{eq:RBM2} and write the LREP for the joint distribution
-\(\tilde{P}_{\boldsymbol \theta_N}\) of
-\((\boldsymbol X, \boldsymbol H)\) from \eqref{eq:RBM1} as
-\begin{align*}
-\REP_{(\boldsymbol X, \boldsymbol H)}(\boldsymbol \theta_N) &=\log\left[  \frac{\max_{ (\boldsymbol x, \boldsymbol h) \in \{\pm 1\}^{N+N_{\mathcal{H}}}}
+ \]
+ Hence, this model sequence for $\boldsymbol X$ will be  S-unstable model if the aggregation of absolute parameters grows faster than the number $N$ of parameters/visible variables.
+Consequently, even for a simplest RBM model involving independence,  preventing instability requires careful choice of parameters, particularly with regard to how a parameter configuration differs from zero.
+For more general RBM models, the number $N_{\mathcal{H}}$ of hidden variables $\boldsymbol H$ can also be chosen arbitrarily (i.e., as some function $N_{\mathcal{H}}\equiv N_{N,\mathcal{H}}$ of $N$), which can substantially inflate the number $d(N)$ of model parameters  and further impact model instability through accumulated parameters.
+  To better understand the effects of instability in the RBM structure,  Proposition~\ref{prop2} next frames the general behavior of
+    extreme probabilities in the joint RBM model (\ref{eqn:RBM1}) for  $(\boldsymbol X, \boldsymbol H)$  and the implied RBM data  model (\ref{eqn:RBM2}) for  $\boldsymbol X$ alone.  Specifically,
+    critical measures of instability may be closely connected in both models through   tight bounds on their respective LREP  values (\ref{eq:elpr}).
+As a result,   Proposition~\ref{prop2} shows how an unstable distribution for observations $\boldsymbol X$  may be traced to  sources of  instability in the original joint distribution for $(\boldsymbol X,\boldsymbol H)$.
+   This also suggests a device for avoiding instability, as provided next.
+
+   To state the result,
+     let   $\REP_{\boldsymbol X}(\boldsymbol \theta_N) \equiv \REP(\boldsymbol \theta_N) $ denote the LREP value (\ref{eq:elpr}) from the marginal distribution $P_{\boldsymbol \theta_N}$  of visibles $\boldsymbol X$ in (\ref{eqn:RBM2}) and write the LREP  for the joint distribution
+   $\tilde{P}_{\boldsymbol \theta_N}$ of  $(\boldsymbol X, \boldsymbol H)$ from   (\ref{eqn:RBM1}) as
+         \begin{eqnarray*}
+         \REP_{(\boldsymbol X, \boldsymbol H)}(\boldsymbol \theta_N) &=&\log\left[  \frac{\max_{ (\boldsymbol x, \boldsymbol h) \in \{\pm 1\}^{N+N_{\mathcal{H}}}}
          \tilde{P}_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)}{\min_{ (\boldsymbol x, \boldsymbol h) \in \{\pm 1\}^{N+N_{\mathcal{H}}}`}
-\tilde{P}_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)} \right] \qquad \mbox{(joint RBM model)}    \\    
-&=   \left(\max_{ (\boldsymbol x, \boldsymbol h) \in \{\pm 1\}^{N+N_{\mathcal{H}}}} f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)-\min_{ (\boldsymbol x, \boldsymbol h) \in \{\pm 1\}^{N+N_{\mathcal{H}}}} f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)\right),
-\end{align*}
-written as a function
-\begin{equation}
-\label{eq:f}
-f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)  \equiv  \sum_{i=1}^N x_i  \theta_{i,N}^{\mathcal{V}} +  \sum_{j=1}^{N_\mathcal{H}} h_j \theta_{j,N}^{\mathcal{H}} + \sum_{i=1}^N \sum_{j=1}^{N_\mathcal{H}} x_i h_j  \theta_{ij,N}^{\mathcal{VH}}
-\end{equation}
-of outcomes \(\boldsymbol x =(x_1,\ldots,x_N) \in\{\pm1\}^{N}\),
-\(\boldsymbol h =(h_1,\ldots,h_{N_{\mathcal{H}}}) \in\{\pm1\}^{N_{\mathcal{H}}}\)
-and parameters
-\(\boldsymbol \theta_N \equiv (\boldsymbol \theta_N^{\mathcal{V}},\boldsymbol \theta_N^{\mathcal{H}}, \boldsymbol \theta_N^{\mathcal{VH}})\),
-with \(\theta_{i,N}^{\mathcal{V}}\), \(\theta_{j,N}^{\mathcal{H}}\) and
-\(\theta_{ij,N}^{\mathcal{VH}}\) denoting respective parameter
-components, \(1 \leq i \leq N\), \(1 \leq j \leq N_{\mathcal{H}}\). Due
-to the marginalization steps in defining the distribution \eqref{eq:RBM2}
-of \(\boldsymbol X\), note that
-\(\REP_{\boldsymbol X}(\boldsymbol \theta_N)\) has no immediate
-analytical expression similar to that of
-\(\REP_{(\boldsymbol X, \boldsymbol H)}(\boldsymbol \theta_N)\). For
-clarity, recall also that S-instability \eqref{eq:Sun} in each model type
-refers to a respective divergence (i.e.,
-\(\lim_{N\to \infty} \REP_{(\boldsymbol X, \boldsymbol H)}(\boldsymbol \theta_N) /(N+N_{\mathcal{H}})=\infty\),
-\(\lim_{N\to \infty} \REP(\boldsymbol \theta_N) /N=\infty\)) upon
-scaling by the corresponding number of variables in a distribution. In
-the following, let \(|\boldsymbol y|_1 = \sum_{i=1}^d |y_i|\) denote the
-L-1 norm of a generic vector \(\boldsymbol y =(y_1,\ldots,y_d)\),
-\(d \geq 1\). \BeginKnitrBlock{proposition}
-
-\protect\hypertarget{prp:prop2}{}{\label{prp:prop2}}Let
-\(P_{\boldsymbol \theta_N}\) denote a RBM-based model \eqref{eq:RBM2} for
-\(N\geq 1\) visible variables \(\boldsymbol X \equiv \boldsymbol X_N\)
-derived from \(\tilde{P}_{\boldsymbol \theta_N}\) as the joint RBM
-distribution \eqref{eq:RBM1} of \((\boldsymbol X, \boldsymbol H)\)
-involving some number \(N_{\mathcal{H}} \equiv N_{N,\mathcal{H}}\geq 0\)
-of hidden variables \(\boldsymbol H \equiv \boldsymbol H_N\) and
-parameters
-\(\boldsymbol \theta_N \equiv (\boldsymbol \theta_N^{\mathcal{V}},\boldsymbol \theta_N^{\mathcal{H}}, \boldsymbol \theta_N^{\mathcal{VH}}) \in\mathbb{R}^{N}\times \mathbb{R}^{N_{\mathcal{H}}} \times \mathbb{R}^{N N_{\mathcal{H}}}\).
-Then,
-\begin{enumerate}[(i)]
-\item the instability measure $\REP_{(\boldsymbol X)}(\boldsymbol \theta_N)$ for the marginal model $P_{\boldsymbol \theta_N}$ of  $\boldsymbol X$ satisfies
-    $$
+         \tilde{P}_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)} \right] \qquad \mbox{(joint RBM model)}    \\    &= &  \left(\max_{ (\boldsymbol x, \boldsymbol h) \in \{\pm 1\}^{N+N_{\mathcal{H}}}} f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)-\min_{ (\boldsymbol x, \boldsymbol h) \in \{\pm 1\}^{N+N_{\mathcal{H}}}} f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)\right),
+          \end{eqnarray*}
+ written as a function \begin{equation}
+ \label{eqn:f}
+           f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)  \equiv  \sum_{i=1}^N x_i  \theta_{i,N}^{\mathcal{V}} +  \sum_{j=1}^{N_\mathcal{H}} h_j \theta_{j,N}^{\mathcal{H}}  + \sum_{i=1}^N \sum_{j=1}^{N_\mathcal{H}} x_i h_j  \theta_{ij,N}^{\mathcal{VH}}  \end{equation}
+  of outcomes $\boldsymbol x =(x_1,\ldots,x_N) \in\{\pm1\}^{N}$, $\boldsymbol h =(h_1,\ldots,h_{N_{\mathcal{H}}})  \in\{\pm1\}^{N_{\mathcal{H}}}$ and parameters $\boldsymbol \theta_N \equiv (\boldsymbol \theta_N^{\mathcal{V}},\boldsymbol \theta_N^{\mathcal{H}}, \boldsymbol \theta_N^{\mathcal{VH}}  )$, with
+  $\theta_{i,N}^{\mathcal{V}}$, $\theta_{j,N}^{\mathcal{H}}$ and $\theta_{ij,N}^{\mathcal{VH}}$ denoting respective parameter components,  $1 \leq i \leq N$, $1 \leq j \leq N_{\mathcal{H}}$.
+Due to the marginalization steps in defining the distribution (\ref{eqn:RBM2}) of $\boldsymbol X$, note that $\REP_{\boldsymbol X}(\boldsymbol \theta_N)$ has no immediate analytical expression similar to that of  $\REP_{(\boldsymbol X, \boldsymbol H)}(\boldsymbol \theta_N)$.
+For clarity, recall also that S-instability (\ref{eqn:Sun}) in each model type
+refers to a respective divergence (i.e., $\lim_{N\to \infty} \REP_{(\boldsymbol X, \boldsymbol H)}(\boldsymbol \theta_N) /(N+N_{\mathcal{H}})=\infty$, $\lim_{N\to \infty} \REP(\boldsymbol \theta_N) /N=\infty$)
+ upon scaling by the corresponding number of variables in a distribution.
+In the following, let $|\boldsymbol y|_1 = \sum_{i=1}^d |y_i|$ denote the L-1 norm of a generic vector $\boldsymbol y =(y_1,\ldots,y_d)$, $d \geq 1$.
+ \begin{proposition1}
+  \label{prop2} {\it Let $P_{\boldsymbol \theta_N}$ denote a RBM-based  model  (\ref{eqn:RBM2}) for $N\geq 1$ visible variables
+  $\boldsymbol X \equiv \boldsymbol X_N$ derived from $\tilde{P}_{\boldsymbol \theta_N}$ as the joint RBM distribution   (\ref{eqn:RBM1}) of $(\boldsymbol X,\boldsymbol H)$ involving some number  $N_{\mathcal{H}}\equiv  N_{N,\mathcal{H}}\geq 0$  of hidden variables $\boldsymbol H \equiv \boldsymbol H_N$ and parameters $\boldsymbol \theta_N \equiv (\boldsymbol \theta_N^{\mathcal{V}},\boldsymbol \theta_N^{\mathcal{H}}, \boldsymbol \theta_N^{\mathcal{VH}}  ) \in\mathbb{R}^{N}\times \mathbb{R}^{N_{\mathcal{H}}} \times  \mathbb{R}^{N N_{\mathcal{H}}}$.
+Then, \\
+  (i)   the instability measure $\REP_{(\boldsymbol X)}(\boldsymbol \theta_N)$ for the marginal model $P_{\boldsymbol \theta_N}$ of  $\boldsymbol X$ satisfies
+\[
          \left|   \REP_{(\boldsymbol X)}(\boldsymbol \theta_N)  - \elt\right| \leq    N_{\mathcal{H}}  \log 2
-    $$
+\]
 for
-    $$
+\[
 \elt  \equiv   \max_{ \boldsymbol x} \max_{ \boldsymbol h  } f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)-\min_{ \boldsymbol x } \max_{ \boldsymbol h }f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)
-    $$
-    based on $f_{\boldsymbol \theta_N}$ from \\eqref{eq:f} with components $\boldsymbol x \in \{\pm 1\}^{N}, \boldsymbol h \in \{\pm 1\}^{N_{\mathcal{H}}}$.
-\item The instability measure $\REP_{(\boldsymbol X, \boldsymbol H)}(\boldsymbol \theta_N)\equiv \left(\max_{ \boldsymbol x} \max_{ \boldsymbol h  } f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)-\min_{ \boldsymbol x } \min_{ \boldsymbol h }f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)\right)$ for the joint model $\tilde{P}_{\boldsymbol \theta_N}$ of $(\boldsymbol X, \boldsymbol H)$ satisfies
-    \begin{align*}
-    2\Gam +  2|\boldsymbol \theta_N^{\mathcal{H}} |_{1} \; \geq \; \REP_{(\boldsymbol X, \boldsymbol H)}(\boldsymbol \theta_N)    & \geq 
+\]
+based on $f_{\boldsymbol \theta_N}$ from (\ref{eqn:f}) with components
+ $\boldsymbol x \in \{\pm 1\}^{N}, \boldsymbol h \in \{\pm 1\}^{N_{\mathcal{H}}}$.  \\
+ (ii)  The instability measure $\REP_{(\boldsymbol X, \boldsymbol H)}(\boldsymbol \theta_N)\equiv \left(\max_{ \boldsymbol x} \max_{ \boldsymbol h  } f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)-\min_{ \boldsymbol x } \min_{ \boldsymbol h }f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)\right) $ for the joint model $\tilde{P}_{\boldsymbol \theta_N}$ of $(\boldsymbol X, \boldsymbol H)$ satisfies
+    \begin{eqnarray*}
+    2\Gam +  2|\boldsymbol \theta_N^{\mathcal{H}} |_{1} \; \geq \; \REP_{(\boldsymbol X, \boldsymbol H)}(\boldsymbol \theta_N)    & \geq &
      2\max\big\{\Gam,   |\boldsymbol \theta_N^{\mathcal{H}} |_{1}\big\} \\
-    &\geq  2\Gam\\
-    &\geq  \elt\\
-    &\geq   \max\big\{  \Gamc, \, \Gam  - 2|\boldsymbol \theta_N^{\mathcal{H}} |_{1}  \big\}
-    \end{align*}
+     &\geq & 2\Gam\\
+     &\geq & \elt\\
+       &\geq &  \max\big\{  \Gamc, \, \Gam  - 2|\boldsymbol \theta_N^{\mathcal{H}} |_{1}  \big\}
+\end{eqnarray*}
 for
-    $$
+\[
 \Gam \equiv \max_{ \boldsymbol h} k_{\boldsymbol \theta_N} (\boldsymbol h)
-    \geq  |\boldsymbol \theta_N^{\mathcal{V}} |_{1},\qquad  k_{\boldsymbol \theta_N} (\boldsymbol h)  \equiv \sum_{i=1}^{N }\left| \theta_{i,N}^{\mathcal{V}}   + \sum_{j=1}^{N_{\mathcal{H}}} h_j \theta_{ij,N}^{\mathcal{VH}} \right|,
-    $$ 
-    and $\Gamc \equiv \min_{ \boldsymbol h} k_{\boldsymbol \theta_N} (\boldsymbol h)$ based on a function  $k_{\boldsymbol \theta_N} (\boldsymbol h)$ of hidden variable outcomes $\boldsymbol h = (h_1,\ldots,h_{N_{\mathcal{H}}})$ and visible-related parameters $\boldsymbol \theta_N^{\mathcal{V}}$ and $\boldsymbol \theta_N^{\mathcal{VH}}$.
-\item Assuming $\sup_{N \geq 1} N_{\mathcal{H}}/N<\infty$ additionally, then the following properties 1.-7. hold:
-\begin{enumerate}[1.] \itemsep 0cm
-    \item an S-unstable visible model $P_{\boldsymbol \theta_N}$ is equivalent to the condition $\lim_{N\to \infty}  \elt/N  =\infty$; further, $P_{\boldsymbol \theta_N}$ is stable when $\elt/N$, $N \geq 1$, is bounded.
-    \item an S-unstable joint model $P_{\boldsymbol \theta_N}$ is equivalent to the condition $\lim_{N\to \infty} \max\{|\boldsymbol \theta_N^{\mathcal{H}}|_1, \Gam\}/N =\infty$; further,  $\tilde{P}_{\boldsymbol \theta_N}$ is stable when $[|\boldsymbol \theta_N^{\mathcal{H}}|_1+ \Gam]/N$, $N \geq 1$, is bounded.
+ \geq  |\boldsymbol \theta_N^{\mathcal{V}} |_{1},\qquad  k_{\boldsymbol \theta_N} (\boldsymbol h)  \equiv \sum_{i=1}^{N }\left| \theta_{i,N}^{\mathcal{V}}   + \sum_{j=1}^{N_{\mathcal{H}}} h_j \theta_{ij,N}^{\mathcal{VH}} \right|,\] and $\Gamc \equiv \min_{ \boldsymbol h} k_{\boldsymbol \theta_N} (\boldsymbol h)$    based on a function  $ k_{\boldsymbol \theta_N} (\boldsymbol h) $ of hidden variable outcomes $\boldsymbol h = (h_1,\ldots,h_{N_{\mathcal{H}}})$ and visible-related parameters $\boldsymbol \theta_N^{\mathcal{V}}$ and $\boldsymbol \theta_N^{\mathcal{VH}}$.\\
+\noindent (iii) Assuming   $\sup_{N \geq 1} N_{\mathcal{H}}/N<\infty$ additionally, then the following properties 1.-7. hold:
+\begin{enumerate} \itemsep 0cm
+  \item an S-unstable visible model $P_{\boldsymbol \theta_N}$ is equivalent to the condition $\lim_{N\to \infty}  \elt/N  =\infty$; further,  $P_{\boldsymbol \theta_N}$ is stable when
+$\elt/N$, $N \geq 1$, is bounded.
+\item an S-unstable joint model $P_{\boldsymbol \theta_N}$ is equivalent to the condition $\lim_{N\to \infty} \max\{|\boldsymbol \theta_N^{\mathcal{H}}|_1, \Gam\}/N =\infty$; further,  $\tilde{P}_{\boldsymbol \theta_N}$ is stable when
+$[|\boldsymbol \theta_N^{\mathcal{H}}|_1+ \Gam]/N$, $N \geq 1$, is bounded.
 \item if the visible model $P_{\boldsymbol \theta_N}$   is S-unstable, then  the joint model $\tilde{P}_{\boldsymbol \theta_N}$ is also S-unstable.
-    \item when $\lim_{N\to \infty}  (|\boldsymbol \theta_N^{\mathcal{V}}|_1-2|\boldsymbol \theta_N^{\mathcal{H}}|_1)/N =\infty$, both $P_{\boldsymbol \theta_N}$ and $\tilde{P}_{\boldsymbol \theta_N}$ are necessarily S-unstable.
-    \item when $\lim_{N\to \infty}|\boldsymbol \theta_N^{\mathcal{H}}|_1/N =\infty$,  the joint model $\tilde{P}_{\boldsymbol \theta_N}$ is necessarily S-unstable.
-    \item when $\sup_{N \geq 1} |\boldsymbol \theta_N^{\mathcal{H}}|_1 /N<\infty$, the visible model $P_{\boldsymbol \theta_N}$ being S-stable or S-unstable is equivalent to the joint model  $\tilde{P}_{\boldsymbol \theta_N}$ being stable or unstable.
+\item when $\lim_{N\to \infty}  (|\boldsymbol \theta_N^{\mathcal{V}}|_1-2|\boldsymbol \theta_N^{\mathcal{H}}|_1)/N =\infty$,   both
+$P_{\boldsymbol \theta_N}$ and $\tilde{P}_{\boldsymbol \theta_N}$ are necessarily S-unstable.
+\item when $\lim_{N\to \infty}  |\boldsymbol \theta_N^{\mathcal{H}}|_1/N =\infty$,  the joint model $\tilde{P}_{\boldsymbol \theta_N}$ is necessarily S-unstable.
+\item when $\sup_{N \geq 1} |\boldsymbol \theta_N^{\mathcal{H}}|_1 /N<\infty$,
+ the visible model $P_{\boldsymbol \theta_N}$  being S-stable or S-unstable is equivalent to the joint model  $\tilde{P}_{\boldsymbol \theta_N}$ being stable or unstable.
 \item  an S-stable visible model $P_{\boldsymbol \theta_N}$  results if
-        $$
+\[
          |\boldsymbol \theta_N^{\mathcal{V}}|_1+ |\boldsymbol \theta_N^{\mathcal{VH}} |_1  \leq CN,\quad N \geq 1,
-        $$
+ \]
 for some $C>0$, while an S-stable joint model $\tilde{P}_{\boldsymbol \theta_N}$  results if
-        $$
+ \[
         |\boldsymbol \theta_N|_1 \equiv  |\boldsymbol \theta_N^{\mathcal{V}}|_1+|\boldsymbol \theta_N^{\mathcal{H}}|_1 +|\boldsymbol \theta_N^{\mathcal{VH}} |_1  \leq  C N ,\quad N \geq 1.
-        $$
-\end{enumerate}
-\end{enumerate}
-\EndKnitrBlock{proposition}
-
-\BeginKnitrBlock{remark}
-\iffalse <span class="remark"><em>Remark. \fi{} The condition
-\(\sup_{N \geq 1} N_{\mathcal{H}}/N<\infty\) in Proposition
-\ref{prp:prop2}(iii) is often mild in practice (i.e., the number
-\(N_{\mathcal{H}}\) of hidden variables is typically not excessively
-larger than the number \(N\) of visible observations). This allows
-instability results for both marginal and joint RBM models to be more
-readily stated together, as the numbers \(N\) and \(N+N_{\mathcal{H}}\)
-of variables in these models become asymptotically equivalent.
-\EndKnitrBlock{remark}
-
-In Proposition \ref{prp:prop2}(iii), the relationships between RBM
-models with regard to instability, and the effects of different
-parameter types, follow from the bounds on model instability measures in
-Proposition \ref{prp:prop2}(i)-(ii). Generally speaking, all instability
-in the marginal RBM model for the data \(\boldsymbol X\) can be
-attributed to an excessively large model quantity \(\elt\), which
-predominantly follows when main \(\boldsymbol \theta_N^{\mathcal{V}}\)
-and interaction \(\boldsymbol \theta_N^{\mathcal{VH}}\) parameters
-related to visible variables are too large in magnitude (e.g., upon
-accumulation in terms such as
-\(|\boldsymbol \theta_N^{\mathcal{V}}|_1\), \(\Gam\) or \(\Gamc\)). For
-example, for any bounded sequence
-\(|\boldsymbol \theta_N^{\mathcal{H}}|/N\) of hidden parameters, if main
-visible parameters \(\boldsymbol \theta_N^{\mathcal{V}}\) are too
-extreme (\(|\boldsymbol \theta_N^{\mathcal{V}}|_1/N\to \infty\)), this
-aspect will guarantee instability in the visible model
-(\(\elt/N\to \infty\)). In fact, the instability measure
-\(\elt \equiv \max_{ \boldsymbol x} \max_{ \boldsymbol h } f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)-\min_{ \boldsymbol x } \max_{ \boldsymbol h }f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)\)
-for marginal/visible model represents a clearly smaller portion of the
-instability measure
-\(\REP_{(\boldsymbol X, \boldsymbol H)}(\boldsymbol \theta_N)\equiv \max_{ \boldsymbol x} \max_{ \boldsymbol h } f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)-\min_{ \boldsymbol x } \min_{ \boldsymbol h } f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)\)
-in the joint RBM model, implying that an unstable marginal model (i.e.,
-due to \(\boldsymbol \theta_N^{\mathcal{V}}\),
-\(\boldsymbol \theta_N^{\mathcal{VH}}\)) must always translate to an
-unstable joint model and that further potential causes of instability
-exist for the joint model, often due to the size
-\(|\boldsymbol \theta_N^{\mathcal{H}}|_1\). For example, while the joint
-RBM model for \((\boldsymbol X,\boldsymbol H)\) must always be unstable
-due to a diverging combination of visible and/or interaction parameters
-(\(|\boldsymbol \theta_N^{\mathcal{V}}|_1/N\to \infty\) or
-\(\Gam/N\to \infty\)) (Proposition \ref{prp:prop2}(iii.2)), instability
-for the joint model can also result when the main hidden parameters
-\(\boldsymbol \theta_N^{\mathcal{H}}\) become too large relative to
-sample size (\(|\boldsymbol \theta_N^{\mathcal{H}}|_1/N\to \infty\) in
-Proposition \ref{prp:prop2}(iii.5)). However, under Proposition
-\ref{prp:prop2}, the main hidden parameters
-\(\boldsymbol \theta_N^{\mathcal{H}}\) do not necessarily entail a
-source of instability for the marginal visible model. To explain this
-distinction, consider a joint model where all parameters related to
-visibles are zero,
-\(\boldsymbol \theta_N^{\mathcal{V}}= \boldsymbol \theta_N^{\mathcal{VH}}=\boldsymbol 0\),
-but the hidden-related parameters diverge in sum
-\(|\boldsymbol \theta_N^{\mathcal{H}}|_1/N\to \infty\). Here, the
-explosive behavior among parameters
-\(\boldsymbol \theta_N^{\mathcal{H}}\) induces instability in the joint
-model for \((\boldsymbol X, \boldsymbol H)\) but the marginal model for
-\(\boldsymbol X\), however, has a perfectly stable (and in fact uniform)
-distribution in this case. When the hidden parameters are bounded
-relative to the sample size
-(\(\sup_{N\geq 1} |\boldsymbol \theta_N^{\mathcal{H}}|_1/N<\infty\)),
-then all instability in both the joint and marginal RBM models can be
-directly linked to excessively large visible
-\(\boldsymbol \theta_N^{\mathcal{V}}\) and/or interaction parameters
-\(\boldsymbol \theta_N^{\mathcal{VH}}\) so that features of
-stability/instability must be the same across both models (Proposition
-\ref{prp:prop2}(iii.6)). Hence, to prevent instability in the joint
-model, the combined magnitudes of all parameters \(\theta_N\) must be
-controlled (cf.~Proposition \ref{prp:prop2}(iii.7)), while a stable
-visible data model technically results in constraining only the sizes of
-visible-related parameters \(\boldsymbol \theta_N^{\mathcal{V}}\),
-\(\boldsymbol \theta_N^{\mathcal{VH}}\). Nevertheless, because the joint
-model is often employed in practice for purposes of simulation and
-simulation-based inference, it is still reasonable to consider parameter
-choices for ensuring a stable joint model (and, consequently, a stable
-visible model as well).
-
-In more complicated network models involving further or deeper hidden
-layers, the same issues and causes of instability similarly exist, but
-are compounded by a greater number of model parameters. S-unstable joint
-models will similarly follow if the combined sizes of all parameters are
-too great relative to the total number of variables, while instability
-in the data model for visible variables will depend only on the main or
-interaction parameters directly related to visibles and how their
-accumulated magnitude compares to the observation sample size \(N\).
-
-\section{Statistical consequences of instability}\label{implications}
+ \]
+\end{enumerate}}
+\end{proposition1}
+
+\noindent \textbf{Remark 4:} The condition   $\sup_{N \geq 1} N_{\mathcal{H}}/N<\infty$ in Proposition~\ref{prop2}(iii) is often mild in practice (i.e., the number $N_{\mathcal{H}}$ of hidden variables is typically not excessively larger than the number $N$ of visible observations).
+This allows instability results for both marginal and joint RBM models to be more readily stated together,  as the numbers $N$ and $N+N_{\mathcal{H}}$ of variables in these models become asymptotically equivalent.
+\\
+
+In Proposition~\ref{prop2}(iii), the relationships between RBM models with regard to instability, and the effects of different parameter types, follow from the bounds
+on model instability measures in Proposition~\ref{prop2}(i)-(ii).
+Generally speaking, all instability in the marginal RBM model for the data $\boldsymbol X$ can be attributed to an excessively large model quantity $\elt  $, which predominantly follows when main $\boldsymbol \theta_N^{\mathcal{V}}$   and interaction $\boldsymbol \theta_N^{\mathcal{VH}}$  parameters related to visible variables are too large in magnitude (e.g., upon accumulation in terms such as $|\boldsymbol \theta_N^{\mathcal{V}}|_1$, $\Gam$ or $\Gamc$).
+For example, for any bounded sequence $|\boldsymbol \theta_N^{\mathcal{H}}|/N$ of hidden parameters, if main visible parameters $\boldsymbol \theta_N^{\mathcal{V}}$ are too extreme ($|\boldsymbol \theta_N^{\mathcal{V}}|_1/N\to \infty$), this aspect  will guarantee instability in the visible model ($\elt/N\to \infty$).
+In fact, the instability measure $\elt  \equiv   \max_{ \boldsymbol x} \max_{ \boldsymbol h  } f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)-\min_{ \boldsymbol x } \max_{ \boldsymbol h }f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)$ for marginal/visible model represents a clearly smaller
+   portion of the  instability measure
+ $\REP_{(\boldsymbol X, \boldsymbol H)}(\boldsymbol \theta_N)\equiv \max_{ \boldsymbol x} \max_{ \boldsymbol h  } f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)-\min_{ \boldsymbol x } \min_{ \boldsymbol h }f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h) $ in the joint RBM model, implying that an unstable marginal model (i.e., due to  $\boldsymbol \theta_N^{\mathcal{V}}$, $\boldsymbol \theta_N^{\mathcal{VH}}$) must always translate to an unstable joint model and that further potential causes of instability exist for the joint model, often due to the size $|\boldsymbol \theta_N^{\mathcal{H}}|_1$.
+For example,  while the joint RBM model for    $(\boldsymbol X,\boldsymbol H)$ must always be unstable
+  due to a diverging combination of visible and/or interaction parameters ($|\boldsymbol \theta_N^{\mathcal{V}}|_1/N\to \infty$ or $\Gam/N\to \infty$) (Proposition~\ref{prop2}(iii.2)),  instability for the joint model can also result when the main hidden parameters $\boldsymbol \theta_N^{\mathcal{H}}$ become  too large relative to sample size ($|\boldsymbol \theta_N^{\mathcal{H}}|_1/N\to \infty$ im Proposition~\ref{prop2}(iii.5)).
+However, under Proposition~\ref{prop2}, the main hidden parameters $\boldsymbol \theta_N^{\mathcal{H}}$  do not necessarily entail a source of instability for the marginal visible model.
+To explain this distinction, consider a joint model where all parameters related to visibles are zero, $\boldsymbol \theta_N^{\mathcal{V}}= \boldsymbol \theta_N^{\mathcal{VH}}=\boldsymbol 0$, but the hidden-related parameters diverge in sum $|\boldsymbol \theta_N^{\mathcal{H}}|_1/N\to \infty$.
+Here, the explosive behavior among parameters $\boldsymbol \theta_N^{\mathcal{H}}$ induces instability in the joint model for $(\boldsymbol X, \boldsymbol H)$ but the marginal model for $\boldsymbol X$, however, has a perfectly stable (and in fact uniform) distribution in this case.
+When the hidden parameters are bounded relative to the sample size ($\sup_{N\geq 1} |\boldsymbol \theta_N^{\mathcal{H}}|_1/N<\infty$), then all instability in both the joint and marginal RBM models can be directly linked  to
+    excessively large visible $\boldsymbol \theta_N^{\mathcal{V}}$ and/or interaction parameters $\boldsymbol \theta_N^{\mathcal{VH}}$ so that features of stability/instability must be the same across both models  (Proposition~\ref{prop2}(iii.6)).
+    Hence, to prevent instability in the joint model, the combined magnitudes of all parameters $\theta_N$ must be controlled (cf.~Proposition~\ref{prop2}(iii.7)),  while  a stable visible data model technically results in constraining only the sizes of visible-related parameters $\boldsymbol \theta_N^{\mathcal{V}}$, $\boldsymbol \theta_N^{\mathcal{VH}}$.
+Nevertheless, because the joint model is often employed in practice for purposes of simulation and simulation-based inference, it is still reasonable to consider parameter choices for ensuring a stable joint model (and, consequently, a stable visible model as well).
+
+  In more complicated network models involving further or deeper hidden layers, the same issues and causes of instability similarly exist, but are compounded by a greater number of model parameters.
+S-unstable joint models will similarly follow if the combined sizes of all parameters are too great relative to the total number of variables, while instability in the data model for visible variables will depend only on
+  the  main or interaction parameters  directly related to visibles and how their accumulated magnitude compares to the observation sample size $N$.
+
+%%#################
+%################# start of blue marker
+\blue{
+
+\section{Statistical Consequences of Instability}\label{implications}
 
 Due to their extreme sensitivity in probability structure, S-instability
-in FOES models often translates to numerical complications, and in fact
+in FOES models may often translate to numerical complications, and in fact
 obstructions, in both simulation and statistical inference based on
-likelihoods.
-
-\subsection{Implications for
-simulation}\label{implications-for-simulation}
+likelihoods.  We describe these aspects in Sections 5.1-5.3 with regard to data simulation, maximum likelihood estimation and
+Bayes inference, respectively.
+
+\subsection{Implications for Data Simulation}
 
@@
 unstable FOES model, one component changes in outcomes may produce
-radically different probabilities which then creates numerical barriers
+radically different probabilities, which may then impose numerical barriers
 to MCMC. For example, consider implementation of a Gibbs sampler from
@@
 distribution \(P_{\boldsymbol \theta_N}\), then the Gibbs sampler can
-have extreme log-ratios in its transition probabilities \[
-\left| \frac{P_{\boldsymbol \theta_N} (X_i =x_1 | \boldsymbol x^{(1)}_{-i})}{P_{\boldsymbol \theta_N} (X_i = x_2| \boldsymbol x^{(2)}_{-i}) } \right|=\left| \frac{P_{\boldsymbol \theta_N} ( \boldsymbol x^{(1)} }{P_{\boldsymbol \theta_N} (  \boldsymbol x^{(2)} ) } \right|
-\] using that conditional probabilities are proportional to the joint
-probabilities and the fact that unstable models can have unbounded
+have extreme log-ratios in its transition probabilities, \[
+\log \left| \frac{P_{\boldsymbol \theta_N} (X_i =x_1 | \boldsymbol x^{(1)}_{-i})}{P_{\boldsymbol \theta_N} (X_i = x_2| \boldsymbol x^{(2)}_{-i}) } \right|=\log \left| \frac{P_{\boldsymbol \theta_N} ( \boldsymbol x^{(1)} )}{P_{\boldsymbol \theta_N} (  \boldsymbol x^{(2)} ) } \right|,
+\]  as conditional probabilities are proportional to joint
+probabilities that, with unstable models, can have unbounded
 log-probability ratios in one-component changes (Theorem
-\ref{thm:instab-elpr}. This can hinder the ability of chain to
+\ref{thm:instab-elpr}). This can hinder the ability of chain to
 effectively explore the sample space of the observations
-\(\boldsymbol X\), as the chain may then mix poorly by moving rapidly
-to, and slowly away from, sections of the sample space surface. Because
-the numerical problem is unbounded ratios of probabilities under the
-joint model, similar complications can also arise for
-Metropolis-Hastings algorithms for MCMC. Hence, while an unstable FOES
-model \(P_{\boldsymbol \theta_N}\) may be valid and technically open to
-simulation by MCMC, the aspect of instability can ruin numerical
-applications of MCMC.
-
-\subsection{Implications for maximum likelihood
-inference}\label{implications-for-maximum-likelihood-inference}
-
-The rugged or volatile probability structure in unstable models can also
-hamper efforts to maximize likelihood functions for statistical
-inference. Essentially, when a FOES model is unstable along a parameter
-sequence \(\boldsymbol \theta_N\), the same model can consequently
-unstable along parameters \(-\boldsymbol \theta_N\) opposite from the
-origin (model condition PSR and Corollary \ref{cor:sign}). This leads to
-potential hyper-sensitivity in the likelihood function, which can be
-pulled in extreme and opposite directions depending on how a parameter
-lies with respect to zero and translates into numerical complications in
+\(\boldsymbol X\), as the chain may mix poorly by moving rapidly
+to, and slowly away from, sections of the sample space.
+In this case, for example, the Markov chain may become entrapped within a mode
+of the probability function, with rare chance  of escaping to adequately mimic
+the occupation frequencies in the overall sample space.
+If modes of the unstable model are not unique, then important outcomes may be missed
+without multiple chains or unpractically enormous numbers of    MCMC samples.
+This
+mixing problem is due to the unstable stationary distribution (unbounded
+ratios of probabilities under the joint model), rather than in any
+particulars of the MCMC algorithm, and similar complications can also arise
+for Metropolis-Hastings algorithms for MCMC. Hence, while an unstable
+FOES model \(P_{\boldsymbol \theta_N}\) may be valid and technically
+open to simulation by MCMC, the aspect of instability
+can render applications of MCMC as numerically infeasible for simulation purposes.
+This result is in line with conclusions of Handcock~(2002) and Schweinberger~(???) for other exponential models.
+
+\subsection{Implications for Maximum Likelihood Inference}
+
+Volatility  in the probability structure of an unstable model can also
+hamper efforts to maximize likelihood functions in statistical
+inference. When a FOES model is unstable along a parameter
+sequence \(\boldsymbol \theta_N\), the same model can further be
+unstable along parameters \(-\boldsymbol \theta_N\) in an opposite direction from the
+origin  (model condition PSR and Corollary \ref{cor:sign}).
+   This can translate into
+potential  sensitivity of the likelihood function around zero,  and lead to  numerical complications in
 maximizing the objective function.
-
-For a FOES model \(P_{\boldsymbol \theta_N}(\boldsymbol x)\),
-\(\boldsymbol x\in\mathcal{X}^N\), for \(N\) observations
-\(\boldsymbol X =(X_1,\ldots,X_n)\), define the likelihood function
-\(\mathcal{L}(\boldsymbol \theta_N | \boldsymbol x) \equiv N^{-1}\log P_{\boldsymbol \theta_N }(\boldsymbol x)\)
-at \(\boldsymbol \theta_N\) and the minimum probability ratio
-counterpart
-\(\mathcal{L}(\boldsymbol \theta_N | \boldsymbol x) - N^{-1} \log \min_{\boldsymbol \theta_N}= N^{-1}\log[P_{\boldsymbol \theta_N }(\boldsymbol x) /\min_{\boldsymbol \theta_N}]\)
-where
-\(\min_{\boldsymbol \theta_N} \equiv \min_{\boldsymbol y\in\mathcal{X}^N} P_{\boldsymbol \theta_N }(\boldsymbol y)]\).
-Also define
-\(\max_{\boldsymbol \theta_N} \equiv \max_{\boldsymbol y\in\mathcal{X}^N} P_{\boldsymbol \theta_N }(\boldsymbol y)]\).
-
-\BeginKnitrBlock{theorem}
-\protect\hypertarget{thm:something}{}{\label{thm:something}}Let
-\(P_{\boldsymbol \theta_N}\), \(N \geq 1\), denote an S-unstable FOES
-model sequence (i.e.,
-\(\lim_{N\to \infty}\REP(\boldsymbol \theta_N)/N=\infty\)) which
-additionally satisfies model condition SPR. Then,
-\begin{align*}
-\log\left[ \frac{P_{\boldsymbol \theta_N }(\boldsymbol x) }{\min_{\boldsymbol \theta_N}}\right] +  \log\left[ \frac{P_{-\boldsymbol \theta_N }(\boldsymbol x) }{\min_{-\boldsymbol \theta_N}}\right]  &=    \log\left[ \frac{\max_{\boldsymbol \theta_N}}{\min_{\boldsymbol \theta_N}}\right]  \equiv  \REP(\boldsymbol \theta_N)\\
-& = \log\left[ \frac{\max_{-\boldsymbol \theta_N} }{\min_{-\boldsymbol \theta_N}}\right] \equiv \REP(-\boldsymbol \theta_N)
-\end{align*}
-for any \(\boldsymbol x\in\mathcal{X}^N\) and, as \(N\to \infty\), the
-following convergence \[
-\frac{1}{\REP(\boldsymbol \theta_N)} \log\left[ \frac{P_{\boldsymbol \theta_N }(\boldsymbol X) }{\min_{\boldsymbol \theta_N}}\right]  \rightarrow 1 \qquad \frac{1}{\REP(-\boldsymbol \theta_N)}\log\left[ \frac{P_{-\boldsymbol \theta_N }(\boldsymbol X) }{\min_{-\boldsymbol \theta_N}}\right]  \rightarrow 0
-\] holds in probability \(P_{\boldsymbol \theta_N}\), as well as in
-expectation \(E_{\boldsymbol \theta_N}\), under
-\(\boldsymbol \theta_N\); the limits of convergence are reversed under
-\(-\boldsymbol \theta_N\), \[
-\frac{1}{\REP(\boldsymbol \theta_N)} \log\left[ \frac{P_{\boldsymbol \theta_N }(\boldsymbol x) }{\min_{\boldsymbol \theta_N}}\right]  \rightarrow 0 \qquad \frac{1}{\REP(-\boldsymbol \theta_N)}\log\left[ \frac{P_{-\boldsymbol \theta_N }(\boldsymbol x) }{\min_{-\boldsymbol \theta_N}}\right]  \rightarrow 1
-\]
-\EndKnitrBlock{theorem} This needs to be connected to results in
-Schweinberger who considered the one-parameter linear exponential case
-showing that, for positive \(\boldsymbol \theta_N>0\), \[
-\frac{1}{\REP(\boldsymbol \theta_N)} \log\left[ \frac{P_{\boldsymbol \theta_N }(\boldsymbol x) }{\min_{\boldsymbol \theta_N}}\right] = \frac{\boldsymbol g(\boldsymbol x) - L_N}{U_N-L_N} 
-\] holds and converges to \(1\) in expectation
-\(E_{\boldsymbol \theta_N}\), while \[   
-\frac{1}{\REP(-\boldsymbol \theta_N)} \log\left[ \frac{P_{-\boldsymbol \theta_N }(\boldsymbol x) }{\min_{-\boldsymbol \theta_N}}\right] = \frac{U_N -\boldsymbol g(\boldsymbol x)}{U_N-L_N} 
-\] holds converges to \(1\) in expectation
-\(E_{-\boldsymbol \theta_N}\). This result is a special case of Theorem
-\ref{thm:something} and the LREP expansion \eqref{eq:UL} for the one
-parameter exponential model case.
-
-The idea, more generally, is that probability in an unstable model is
-pushed to opposite regions of the sample space due to degeneracy and
-opposite signed parameters. This can cause numerical complications in
-likelihood maximization. The likelihood function knows what to do with
-an observed outcome
-\(\boldsymbol x \equiv \argmax \mathcal{L}(\boldsymbol \theta_N)\) which
-is also the outcome with minimal probability under
-\(P_{-\boldsymbol \theta_N}\). But unless the observed outcome
-\(\boldsymbol x\) is at an extreme in the probability structures of
-\(P_{\boldsymbol \theta_N}, P_{-\boldsymbol \theta_N}\) (which are
-related) the likelihood function will be flat and optimization steps may
-shift around zero.
-
-\subsection{Implications for Bayes
-inference}\label{implications-for-bayes-inference}
-
-Considering that changes in sign of parameters can cause all probability
-to be placed on disjoint sections of the sample space (Corollary
-\ref{cor:sign}), then this will create numerical difficulties in Bayes
-inference based on MCMC. We can write the MH transition density to
-indicate this.
+We next provide a discussion of this issue in a way that builds upon and extends  related findings by Schweinberger(???), who largely focused on the case of one-parameter exponential models.
+
+    With many probability models, the modes and anti-modes in the probability structure under one parameter \( \boldsymbol \theta_N\) are reversed in  role  when the parameter sign   changes \(-\boldsymbol \theta_N\).
+Because unstable models tend to degeneracy,
+the opposite signed parameters  further push unstable models to assign nearly all probability to extremely opposite data configurations, given by modes/anti-modes.
+This is made concrete in Theorem?,  relating the degeneracy from unstable models to the expected behavior of log-likelihood functions.
+
+\begin{theorem1}
+\label{thm:something}  {\it Let $P_{\boldsymbol \theta_N}$, $N \geq 1$, denote an S-unstable FOES model sequence, which additionally satisfies model condition SPR.
+Let $\boldsymbol{x}_{\max, \boldsymbol \theta_N},\boldsymbol{x}_{\min, \boldsymbol \theta_N}\in\mathcal{X}^N$ denote, respectively, a mode and anti-mode of the model $P_{\boldsymbol \theta_N}(\boldsymbol x)$, $\boldsymbol x\in\mathcal{X}^N$, for $N$ observations $\boldsymbol X =(X_1,\ldots,X_N)$,
+whereby $P_{\boldsymbol \theta_N }(\boldsymbol{x}_{\max, \boldsymbol \theta_N}) = \max_{\boldsymbol y\in\mathcal{X}^N} P_{\boldsymbol \theta_N }(\boldsymbol y)$ and $P_{\boldsymbol \theta_N }(\boldsymbol{x}_{\min, \boldsymbol \theta_N}) = \min_{\boldsymbol y\in\mathcal{X}^N} P_{\boldsymbol \theta_N }(\boldsymbol y)$.
+
+Then, letting $\stackrel{\mbox{$p, \mathrm{E}$}}{\longrightarrow}$ denote  convergence in  probability  and expectation,
+ as $N\to \infty$,
+ \[
+    \frac{1}{\REP(\boldsymbol \theta_N)} \log\left[ \frac{P_{\boldsymbol \theta_N }(\boldsymbol X) }{ \displaystyle{\min_{\boldsymbol y\in\mathcal{X}^N}} P_{\boldsymbol \theta_N }(\boldsymbol y)}\right] = \frac{ \log P_{\boldsymbol \theta_N } ( \boldsymbol X)  - \log P_{\boldsymbol \theta_N }(\boldsymbol{x}_{\min, \boldsymbol \theta_N}) }
+    { \log P_{\boldsymbol \theta_N } (\boldsymbol{x}_{\max, \boldsymbol \theta_N})  - \log P_{\boldsymbol \theta_N }(\boldsymbol{x}_{\min, \boldsymbol \theta_N}) }
+ \stackrel{\mbox{$p, \mathrm{E}$}}{\longrightarrow}
+ \; 1
+\]
+  under $\boldsymbol\theta_N$ while
+\[
+ \frac{1}{\REP(-\boldsymbol \theta_N)} \log\left[ \frac{P_{-\boldsymbol \theta_N }(\boldsymbol X) }{  \displaystyle{\min_{\boldsymbol y\in\mathcal{X}^N}} P_{-\boldsymbol \theta_N }(\boldsymbol y)} \right]
+ =    \frac{ \log P_{-\boldsymbol \theta_N } ( \boldsymbol X) - \log P_{ \boldsymbol \theta_N }(\boldsymbol{x}_{\max, \boldsymbol \theta_N})    }
+     { \log P_{-\boldsymbol \theta_N } (\boldsymbol{x}_{\min, \boldsymbol \theta_N})  - \log P_{-\boldsymbol \theta_N }(\boldsymbol{x}_{\max, \boldsymbol \theta_N}) }
+ \stackrel{\mbox{$p,\mathrm{E}$}}{\longrightarrow}
+ \; 1,
+\]
+ under $-\boldsymbol\theta_N$, where
+ \[
+ \REP( \boldsymbol \theta_N) \equiv \log \frac{P_{\boldsymbol \theta_N } (\boldsymbol{x}_{\max, \boldsymbol \theta_N})}{P_{\boldsymbol \theta_N }(\boldsymbol{x}_{\min, \boldsymbol \theta_N})} = \log  \frac{P_{-\boldsymbol \theta_N } (\boldsymbol{x}_{\min, \boldsymbol \theta_N}) }{ P_{-\boldsymbol \theta_N }(\boldsymbol{x}_{\max, \boldsymbol \theta_N})}=   \REP( -\boldsymbol \theta_N), \quad N \geq 1.
+  \]
+ }
+\end{theorem1}
+ Theorem~3 entails
+log-likelihood functions  based on unstable models  are both inversely related and degenerate at opposited signed parameters $ \boldsymbol \theta_N$ or $-\boldsymbol \theta_N$, so that  likelihoods are highest  at different extremes in data configuration (e.g., $\boldsymbol{x}_{\max, \boldsymbol \theta_N}$ under
+$\boldsymbol \theta$-probabilities or  $\boldsymbol{x}_{\min, \boldsymbol \theta_N}$ under
+$-\boldsymbol \theta$-probabilities).
+If the observed outcome $\boldsymbol x$ for data $\boldsymbol X$ is not a mode/anti-mode, then probabilities for the outcome may be small under both  parameters $ \boldsymbol \theta_N$ and $-\boldsymbol \theta_N$, in which case
+  associated optimization steps may then shift around zero and struggle to converge.
+In many model formulations, the zero parameter $ \boldsymbol \theta_N=\boldsymbol 0$ is a ``safe" position among parameters, representing a  guaranteed
+   stable model (having uniform probabilities among outcomes), which  can also  tether  a broad parameter search attempted  among     unstable models.     Handcock
+(\protect\hyperlink{ref-handcock2003assessing}{2003}) describes similar results for degenerate exponential models, and  Theorem~3  also supports an important finding of
+   Schweinberger
+(\protect\hyperlink{ref-schweinberger2011instability}{2011} Corollary 1)
+for one-parameter linear exponential models (??).
+In the latter case, the likelihood score function  at $\boldsymbol \theta_N$ is the expected value  $\mu(\boldsymbol \theta_N)\equiv \mathrm{E}_{\boldsymbol \theta_N} g(\boldsymbol X)$  of the sufficient statistic $g(\cdot)$,   and optimization involves
+solving $\mu(\cdot)=
+ g(\boldsymbol x)$ for an observed outcome $\boldsymbol x$. For unstable  models in this exponential class,
+  Schweinberger
+(\protect\hyperlink{ref-schweinberger2011instability}{2011} Corollary 1)  shows
+that
+\[
+\lim_{N\to \infty}\frac{\mu(\boldsymbol \theta_N) -L_n}{U_n-L_N}= \left\{ \begin{array}{lcl}
+1 && \mbox{for $\boldsymbol \theta_N>0$},\\
+0 && \mbox{for $\boldsymbol \theta_N<0$}, \end{array}\right.
+\]
+  where again $U_N$ and $L_N$ denote the maximum and minimum values of the statistic $g(\boldsymbol x)$, $\boldsymbol x\in\mathcal{X}^N$
+       As described by Schweinberger
+(\protect\hyperlink{ref-schweinberger2011instability}{2011}), the implication for maximum likelihood estimation   is that, unless an observed outcome $\boldsymbol x$ falls at an extreme $U_N, L_N$ (i.e., modes/anti-modes),
+optimization steps in the parameter space can   iterate in  relatively small increments around zero and  fail to  converge.   For   unstable   one-parameter exponential models,
+the maximum likelihood results of Schweinberger
+(\protect\hyperlink{ref-schweinberger2011instability}{2011})   turn  out to be a special case of Theorem~\ref{thm:something}  and the LREP expansion \eqref{eq:UL} in this setting; namely,    for an unstable   model with $\boldsymbol \theta_N>0$,
+   \[
+\frac{1}{\REP(\boldsymbol \theta_N)} \log\left[ \frac{P_{\boldsymbol \theta_N }(\boldsymbol X) }{\displaystyle{\min_{\boldsymbol y\in\mathcal{X}^N}} P_{\boldsymbol \theta_N }(\boldsymbol y)}\right] = \frac{\boldsymbol g(\boldsymbol X) - L_N}{U_N-L_N}\;\;\stackrel{\mbox{$p, \mathrm{E}$}}{\longrightarrow}\;\; 1
+\] holds as $N\to \infty$ by Theorem~\ref{thm:something}, while under   $-\boldsymbol \theta_N<0$
+\[
+\frac{1}{\REP(-\boldsymbol \theta_N)} \log\left[ \frac{P_{-\boldsymbol \theta_N }(\boldsymbol X) }{\displaystyle{\min_{\boldsymbol y\in\mathcal{X}^N}} P_{-\boldsymbol \theta_N }(\boldsymbol y)}\right] = \frac{U_N -\boldsymbol g(\boldsymbol X)}{U_N-L_N} = 1 - \frac{\boldsymbol g(\boldsymbol X) - L_N}{U_N-L_N} \;\;\stackrel{\mbox{$p, \mathrm{E}$}}{\longrightarrow}\;\; 1.
+\]
+Again, when all probability in  unstable models  may be
+  pushed to opposite extremes in the sample space, due to a combination of degeneracy and
+parameter sign,  numerical complications in
+likelihood maximization may occur.
+
+\subsection{Implications for Bayes Inference}
+The potential numerical difficulties with maximum likelihood with unstable models, as described in the previous section, 
+can naturally carry over to Bayes inference.
+Considering that the degeneracy issues related to unstable models can cause likelihoods can be flat (e.g., near zero) for many parameters under a given data outcome and that
+   sign changes in  parameters can shift tremendous probability to extreme and opposite outcomes in the sample space   (e.g., Corollary
+\ref{cor:sign}, Theorem~\ref{thm:something}), then numerical complications 
+may arise with  Bayes
+inference in sampling a posterior parameter space   based on MCMC.   The potential challenges in chain mixing  are similar to those presented in
+Section \ref{mcmc}, though in chain movements through the parameter space as opposed to the sample space in data generation.
+That is, in the Bayes setting for sampling a posterior distribution for $\boldsymbol \theta_N$,  a chain may  unstable to effectively explore
+the parameter space due partly to extreme and potentially unbounded probability ratios from parameter sign changes, which represents a parameter space analog to how one component changes in the sample space may impact data simulation with unstable models.
+For example,
+if $\pi(\cdot)$ denotes a prior density for $\boldsymbol \theta_N$ and  \(q(\cdot | \cdot)\)  denotes a proposal distribution 
+for use in a Metropologis-Hastings (MH) sampler, then 
+ MH acceptance probability becomes \[
+\alpha\left(\boldsymbol \theta^{(1)} \mid \boldsymbol \theta^{(2)}\right)= \min\left\{1,
+\frac{q(\boldsymbol \theta_N^{(2)} \mid \boldsymbol \theta_N^{(1)})}{q(\boldsymbol \theta_N^{(1)} \mid \boldsymbol \theta_N^{(2)})}
+\frac{P_{\boldsymbol \theta_N^{(1)}} ( \boldsymbol x ) \pi(\boldsymbol \theta_N^{(1)}) }{P_{\boldsymbol \theta^{(2)}_N} (  \boldsymbol x) \pi(\boldsymbol \theta_N^{(2)}) } \right\},
+\] 
+which indicates how parameter sensitivity in the likelihood $P_{\boldsymbol \theta_N} (  \boldsymbol x)$ may complicate sampling of the posterior $P_{\boldsymbol \theta_N} (  \boldsymbol x) \pi(\boldsymbol \theta_N)$    (i.e., moving from $\boldsymbol \theta^{(1)}$ to $\boldsymbol \theta^{(2)}$ in the parameter space).
+Furthermore, the potential for model unstability and the size of the parameter space can also become greater with the introduction of latent variables to existing data variables,
+as involved in some model formulations described in Section~(??).
+As latent variables are often sampled with parameters in a Bayes MCMC approach, this aspect may further compound numerical problems in chain mixing.
 
 \section{Concluding Remarks}\label{conclusions}
-
 For a large class of models that covers a broad range of applications
@@
 place all probability on certain modal subsections of the sample space,
-which potentially could be small. The FOES model class is quite broad
-and, particularly in developing FOES models for large data sets, some
+which potentially could be small.  Such instability is often due to a complex interaction between
+the model statistics used  (i.e., how numerous and large these may become) and the number and magnitudes of
+parameters  in the model formulation.  For many FOES models, the possibility exists, at least in principle, to constraint parameters in a way 
+balances their potential contributions  against those of model statistics in order to prevent   probability  instabilities.
+  The FOES model class is quite broad
+and, in developing such models for large data sets, some
 caution should be used in parameter specification to control effects of
@@
 
+%####################
+%#################### end of blue marker
+}
+
 \clearpage
@@
 
-\textbf{Proof of Proposition \ref{prp:instab-elpr}.} We prove the
-contrapositive, supposing that \(\Delta(\boldsymbol \theta_N) \le C\)
+\textbf{Proof of Theorem~\ref{instab-elpr}.} For part~(i), we prove the
+contrapositive, supposing that \(\DN(\boldsymbol \theta_N) \le C\)
 holds for some \(C > 0\) and show
-\(\text{ELPR}(\boldsymbol \theta_N) \leq NC\). Let
+\(\REP(\boldsymbol \theta_N) \leq NC\). Let
 \(\boldsymbol x_{min} \equiv \argmin\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)\)
@@
 (i.e. \(\boldsymbol x_i, \boldsymbol x_{i + 1} \in \mathcal{X}^N\)
-differ in exactly \(1\) component for \(i = 0, \dots, k\)) for some
-integer \(k \in \{0, 1, \dots, N\}\). Under the FSFS model, recall
+differ in exactly \(1\) component, \(i = 0, \dots, k\)) for some
+integer \(k \in \{0, 1, \dots, N\}\). Under the FOES model, recall
 \(P_{\boldsymbol \theta_N}(\boldsymbol x) > 0\) holds so that
@@
 \begin{align*}
-\text{ELPR}(\boldsymbol \theta_N) = \log\left[\frac{P_{\boldsymbol \theta_N}(\boldsymbol x_{max})}{P_{\boldsymbol \theta_N}(\boldsymbol x_{min})}\right] &= \left|\sum\limits_{i = 1}^k\log\left(\frac{P_{\boldsymbol \theta_N}(\boldsymbol x_i)}{P_{\boldsymbol \theta_N}(\boldsymbol x_{i-1})}\right)\right| \\
+\REP(\boldsymbol \theta_N) = \log\left[\frac{P_{\boldsymbol \theta_N}(\boldsymbol x_{max})}{P_{\boldsymbol \theta_N}(\boldsymbol x_{min})}\right] &= \left|\sum\limits_{i = 1}^k\log\left(\frac{P_{\boldsymbol \theta_N}(\boldsymbol x_i)}{P_{\boldsymbol \theta_N}(\boldsymbol x_{i-1})}\right)\right| \\
 &\le \sum\limits_{i = 1}^k\left|\log\left(\frac{P_{\boldsymbol \theta_N}(\boldsymbol x_i)}{P_{\boldsymbol \theta}(\boldsymbol x_{i-1})}\right)\right| \le k \Delta_N(\boldsymbol \theta_N) \le NC,
@@
 \(k = 0\), then \(\boldsymbol x_{max} = \boldsymbol x_{min}\) and the
-same bound above holds.\hfill \(\Box\)
-
-\textbf{Proof of Proposition \ref{prp:degenFSFS}.} Where
-\(|\mathcal{X}|<\infty\) holds in the FSFS model, we may suppose
-\(|\mathcal{X}|>1\) (i.e., \(\mathcal{X}^N\) has more than one outcome)
-because otherwise the model is trivially degenerate for all
-\(N \geq 1\). Fix \(0 < \epsilon < 1\). Then,
+same bound above holds.
+This establishes part~(i).
+To show part~(ii), note   the definition of S-instability (i.e., $\lim_{N\to \infty}\REP(\boldsymbol \theta_N)/N= \infty$) combined with part~(i) implies that $\lim_{N\to \infty}\DN(\boldsymbol \theta_N)=\infty$.
+\hfill \(\Box\)
+
+\noindent\textbf{Proof of Theorem~\ref{degenFOES}.} As
+\(|\mathcal{X}|<\infty\) holds in the FOES model, we may suppose
+\(|\mathcal{X}|>1\);
+otherwise,   \(\mathcal{X}^N\) has  one outcome and the model is  trivially degenerate for all
+\(N \geq 1\). Fix \(0 < \epsilon < 1\) and write
+\(\boldsymbol x_{min} \equiv \argmin\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)\)
+and
+\(\boldsymbol x_{max} \equiv \argmax\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)\). Then,
 \(\boldsymbol x_{max} \in M_{\epsilon, \boldsymbol \theta_N}\), so
@@
 holds so that
-\begin{align*}
+\[
 1-P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N})
-& = \sum\limits_{\boldsymbol x \in \mathcal{X}^N \setminus M_{\epsilon, \boldsymbol \theta_N}}P_{\boldsymbol \theta_N}(\boldsymbol x) \\
-& \le (|\mathcal{X}|^N)[P_{\boldsymbol \theta_N}(\boldsymbol x_{max})]^{1-\epsilon}[P_{\boldsymbol \theta_N}(\boldsymbol x_{min})]^{\epsilon}.
-\end{align*}
+ = \sum\limits_{\boldsymbol x \in \mathcal{X}^N \setminus M_{\epsilon, \boldsymbol \theta_N}}P_{\boldsymbol \theta_N}(\boldsymbol x)
+  \le (|\mathcal{X}|^N)[P_{\boldsymbol \theta_N}(\boldsymbol x_{max})]^{1-\epsilon}[P_{\boldsymbol \theta_N}(\boldsymbol x_{min})]^{\epsilon}.
+\]
 From the lower bound on
@@
 \end{align*}
-as \(N \rightarrow \infty\) by the definition of an S-unstable FSFS
-model (cf.~Definition \ref{def:instabFSFS}). Consequently,
+as \(N \rightarrow \infty\) by the definition of an S-unstable FOES
+model (\ref{eqn:Sun}). Consequently,
 \(P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N}) \rightarrow 1\)
-as \(N \rightarrow \infty\) as claimed. \hfill \(\Box\)
+as \(N \rightarrow \infty\) as claimed. \hfill \(\Box\)\\
+
+\noindent\textbf{Proof of Corollary~\ref{sign}.}\\
+The model condition PSR implies that
+\begin{equation}
+\label{eqn:R}
+  \frac{\max_{\boldsymbol y \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol y)  }{\min_{\boldsymbol y \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol y) } =   \frac{\max_{\boldsymbol y \in \mathcal{X}^N}P_{-\boldsymbol \theta_N}(\boldsymbol y)  }{\min_{\boldsymbol y \in \mathcal{X}^N}P_{-\boldsymbol \theta_N}(\boldsymbol y) }
+\end{equation}
+so that the log-ratio $\REP(\boldsymbol \theta_N)=\REP(-\boldsymbol \theta_N)$ is the same for both $\boldsymbol \theta_N$ and $-\boldsymbol \theta_N$ in (\ref{eq:elpr}).
+Now part~(i) of  Corollary~\ref{sign} follows from $\REP(\boldsymbol \theta_N)/N=\REP(-\boldsymbol \theta_N)/N\to \infty$ as $N\to \infty$ in (\ref{eqn:Sun}).
+To show part~(ii), fix $0 < \epsilon < 1$ and consider a $\epsilon$-mode set
+$\mathcal{M}_{\epsilon, \boldsymbol \theta_N}$ under $\boldsymbol \theta_N$ from (\ref{eqn:mode}).
+If $\boldsymbol x \in \mathcal{M}_{\epsilon, \boldsymbol \theta_N}^c \equiv \mathcal{X}^N \setminus \mathcal{M}_{\epsilon, \boldsymbol \theta_N}$, then, by definition,
+\[
+   \frac{P_{\boldsymbol \theta_N}(\boldsymbol x)}{\min_{\boldsymbol y \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol y)} \leq \left[\frac{\max_{\boldsymbol y \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol y)  }{\min_{\boldsymbol y \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol y) } \right]^{1-\epsilon}
+\]
+holds, which is equivalent to
+\[
+      \frac{\max_{\boldsymbol y \in \mathcal{X}^N}P_{-\boldsymbol \theta_N}(\boldsymbol y)}{P_{-\boldsymbol \theta_N}(\boldsymbol x)} \leq
+        \left[\frac{\max_{\boldsymbol y \in \mathcal{X}^N}P_{-\boldsymbol \theta_N}(\boldsymbol y)  }{\min_{\boldsymbol y \in \mathcal{X}^N}P_{-\boldsymbol \theta_N}(\boldsymbol y) } \right]^{1-\epsilon}
+\]
+by model condition PSR and (\ref{eqn:R}).  The latter is in turn equivalent to
+\begin{equation}
+\label{eqn:R2}\log P_{-\boldsymbol \theta_N}(\boldsymbol x) \geq    \epsilon \max\limits_{\boldsymbol y \in \mathcal{X}^N} \log  P_{-\boldsymbol \theta_N}(\boldsymbol y) + (1-\epsilon)\min\limits_{\boldsymbol y \in \mathcal{X}^N} \log P_{-\boldsymbol \theta_N}(\boldsymbol y),
+\end{equation}
+so that $\boldsymbol x \in \mathcal{M}_{\epsilon, \boldsymbol \theta_N}^c $ if and only if (\ref{eqn:R2}) holds.
+Next consider the $(1-\epsilon)$-mode set $\mathcal{M}_{1-\epsilon, -\boldsymbol \theta_N}$
+  under $-\boldsymbol \theta_N$ from (\ref{eqn:mode}).
+If  $\boldsymbol x \in\mathcal{M}_{1-\epsilon, -\boldsymbol \theta_N}$, then by definition $\boldsymbol x$ fulfilles (\ref{eqn:R2}) and so $\boldsymbol x \in \mathcal{M}_{\epsilon, \boldsymbol \theta_N}^c $, showing that $\mathcal{M}_{1-\epsilon, -\boldsymbol \theta_N} \subset \mathcal{M}_{\epsilon, \boldsymbol \theta_N}^c $.
+   By this and the fact that that Theorem~\ref{degenFOES} and Corollary~\ref{sign}(i) entail that
+  $P_{-\boldsymbol \theta_N}(\boldsymbol X_N \in \mathcal{M}_{1-\epsilon, -\boldsymbol \theta_N})\to 1$ as $N\to \infty$ (i.e., $P_{-\boldsymbol \theta_N}$ is S-unstable), we have
+  \[
+    1  = \lim_{N\to \infty} P_{-\boldsymbol \theta_N}(\boldsymbol X_N \in \mathcal{M}_{1-\epsilon, -\boldsymbol \theta_N}) \leq \lim_{N\to \infty} P_{-\boldsymbol \theta_N}(\boldsymbol X_N \in \mathcal{M}_{\epsilon, \boldsymbol \theta_N}^c ) \leq 1,
+  \]
+   proving  Corollary~\ref{sign}(ii)\hfill \(\Box\)\\
+
+\noindent\textbf{Proof of Proposition~\ref{prop1}.}\\
+For any two outcomes $\boldsymbol x_1, \boldsymbol x_2\in\mathcal{X}^N$, the log-ratio of probabilities from the linear exponential model (\ref{eqn:expo}) with $k$ parameters/statistics satisfies
+\[
+\left|\log \left[ \frac{P_{\boldsymbol \theta_N}(\boldsymbol x_1)}{P_{\boldsymbol \theta_N}(\boldsymbol x_2)}  \right] \right| =
+ \left|  \sum_{i=1}^k \theta_{i,N} [g_{i,k}(\boldsymbol x_1) - g_{i,k}(\boldsymbol x_2) ] \right|  \leq  \sum_{i=1}^k | \theta_{i,N}| (U_{i,N}-L_{i,N});
+\]
+consequently,   $\REP(\boldsymbol \theta_N )  \leq  \sum_{i=1}^k | \theta_{i,N}| (U_{i,N}-L_{i,N})$ holds in (\ref{eq:elpr}) and model stability in Proposition~\ref{prop1}
+follows from (\ref{eqn:Sun}).\hfill \(\Box\)\\
+
+\noindent\textbf{Proof of Proposition~\ref{prop2}.}\\
+Writing $\boldsymbol x=(x_1,\ldots,x_N)$ and $\boldsymbol h =(h_1,\ldots,h_{N_{\mathcal{H}}})$ with all components $x_i,h_j\in\{\pm 1\}$,
+probabilities in the joint RBM model (\ref{eqn:RBM1}) can be written as $\tilde{P}_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h) =
+c(\boldsymbol \theta_N)\exp[ f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)]$ in terms of the function
+$f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)$  from (\ref{eqn:f}) and the normalizing constant $ c(\boldsymbol \theta_N)= \exp [-
+\psi(\boldsymbol \theta_N)]$ from (\ref{eqn:RBM1}).  Let $\boldsymbol x_M, \boldsymbol x_m\in\{\pm 1\}^N$ be such that
+$P_{\boldsymbol \theta_N} (\boldsymbol x_M) = \max_{\boldsymbol x}P_{\boldsymbol \theta_N} (\boldsymbol x)$ and
+$P_{\boldsymbol \theta_N} (\boldsymbol x_m) = \min_{\boldsymbol x}P_{\boldsymbol \theta_N} (\boldsymbol x)$ under the marginal RBM model $P_{\boldsymbol \theta_N} (\boldsymbol x) =c(\boldsymbol \theta_N)\sum_{\boldsymbol h \in\{\pm 1\}^{\mathcal{N}_H}} \tilde{P}_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)= c(\boldsymbol \theta_N)\sum_{\boldsymbol h \in\{\pm 1\}^{\mathcal{N}_H}} \exp[ f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)]$ from (\ref{eqn:RBM2}).
+Also, $\boldsymbol x_0,x_1\in\{\pm 1\}^N$ be such that $\max_{\boldsymbol h}f_{\boldsymbol \theta_N} (\boldsymbol x_0, \boldsymbol h)=\max_{\boldsymbol x}\max_{\boldsymbol h}f_{\boldsymbol \theta_N} (\boldsymbol x , \boldsymbol h)$
+and $\max_{\boldsymbol h}f_{\boldsymbol \theta_N} (\boldsymbol x_1, \boldsymbol h)=\min_{\boldsymbol x}\max_{\boldsymbol h}f_{\boldsymbol \theta_N} (\boldsymbol x , \boldsymbol h)$.
+ Then, Proposition~\ref{prop2}(i) follows from $\REP_{(\boldsymbol X)}(\boldsymbol \theta_N) = \log[P_{\boldsymbol \theta_N} (\boldsymbol x_M) /P_{\boldsymbol \theta_N} (\boldsymbol x_m) ]$ and the lower/upper bounds on $P_{\boldsymbol \theta_N} (\boldsymbol x_M) $ and $P_{\boldsymbol \theta_N} (\boldsymbol x_m)$ as
+\[
+  c(\boldsymbol \theta_N) \exp[\max_{\boldsymbol h}f_{\boldsymbol \theta_N} (\boldsymbol x_0 , \boldsymbol h)]
+  \leq P_{\boldsymbol \theta_N} (\boldsymbol x_0) \leq  P_{\boldsymbol \theta_N} (\boldsymbol x_M) \leq  2^{N_{\mathcal{H}}} c(\boldsymbol \theta_N) \exp[\max_{\boldsymbol x}\max_{\boldsymbol h}f_{\boldsymbol \theta_N} (\boldsymbol x , \boldsymbol h)]
+\]
+ and
+ \begin{eqnarray*}
+  c(\boldsymbol \theta_N) \exp[\min_{\boldsymbol x}\max_{\boldsymbol h}f_{\boldsymbol \theta_N} (\boldsymbol x , \boldsymbol h)]
+  &\leq & c(\boldsymbol \theta_N) \exp[ \max_{\boldsymbol h}f_{\boldsymbol \theta_N} (\boldsymbol x_m , \boldsymbol h)]
+ \\& \leq& P_{\boldsymbol \theta_N} (\boldsymbol x_m) \\&\leq& P_{\boldsymbol \theta_N} (\boldsymbol x_1)\\ &\leq &  2^{N_{\mathcal{H}}} c(\boldsymbol \theta_N) \exp[ \max_{\boldsymbol h}f_{\boldsymbol \theta_N} (\boldsymbol x_1 , \boldsymbol h)]\\&=&2^{N_{\mathcal{H}}} c(\boldsymbol \theta_N) \exp[\min_{\boldsymbol x}\max_{\boldsymbol h}f_{\boldsymbol \theta_N} (\boldsymbol x , \boldsymbol h)]
+\end{eqnarray*}
+
+ To prove Proposition~\ref{prop2},
+ we next expand the function $f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)$  from (\ref{eqn:f}) as
+ \[
+f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)= \sum_{j=1}^{N_\mathcal{H}} h_j \theta_{j,N}^{\mathcal{H}}  +   \sum_{i=1}^N \left( \theta_{i,N}^{\mathcal{V}} + \sum_{j=1}^{N_\mathcal{H}}  h_j  \theta_{ij,N}^{\mathcal{VH}}\right)x_i=\sum_{i=1}^N x_i  \theta_{i,N}^{\mathcal{V}} +  \sum_{j=1}^{N_\mathcal{H}} \left( \theta_{j,N}^{\mathcal{H}}  + \sum_{i=1}^N  x_i  \theta_{ij,N}^{\mathcal{VH}}\right)h_j.\]
+By this and the fact that $x_i,h_j\in\{\pm 1\}$, we then have
+\begin{eqnarray}
+\nonumber \max_{\boldsymbol x}  f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h) =  \sum_{j=1}^{N_\mathcal{H}} h_j \theta_{j,N}^{\mathcal{H}}  +   a_{\boldsymbol \theta_N, \mathcal{H}} (\boldsymbol h),   &&
+ \min_{\boldsymbol x}  f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h) =  \sum_{j=1}^{N_\mathcal{H}} h_j \theta_{j,N}^{\mathcal{H}}  -   a_{\boldsymbol \theta_N, \mathcal{H}} (\boldsymbol h),\\\nonumber
+  \max_{\boldsymbol h}  f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h) =   \sum_{i=1}^{N}x_i \theta_{i,N}^{\mathcal{V}}  +   b_{\boldsymbol \theta_N, \mathcal{V}} (\boldsymbol x),  &&
+ \min_{\boldsymbol h}  f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h) =  \sum_{i=1}^{N}x_i \theta_{i,N}^{\mathcal{V}}  -   b_{\boldsymbol \theta_N, \mathcal{V}} (\boldsymbol x),\\ \label{eqn:max}
+  a_{\boldsymbol \theta_N, \mathcal{H}} (\boldsymbol h) \equiv \sum_{i=1}^N \left| \theta_{i,N}^{\mathcal{V}} + \sum_{j=1}^{N_\mathcal{H}}  h_j  \theta_{ij,N}^{\mathcal{VH}}\right|, &&  b_{\boldsymbol \theta_N, \mathcal{V}} (\boldsymbol x) \equiv \sum_{j=1}^{N_\mathcal{H}} \left| \theta_{j,N}^{\mathcal{H}}  + \sum_{i=1}^N  x_i  \theta_{ij,N}^{\mathcal{VH}}\right|,
+\end{eqnarray}
+where  $\boldsymbol h^T \boldsymbol \theta_N^{\mathcal{H}}=\sum_{j=1}^{N_\mathcal{H}} h_j \theta_{j,N}^{\mathcal{H}}$, $\boldsymbol x^T \boldsymbol \theta_N^{\mathcal{V}}=  \sum_{i=1}^{N}x_i \theta_{i,N}^{\mathcal{V}}$ and $\Gam\equiv \max_{\boldsymbol h} a_{\boldsymbol \theta_N, \mathcal{H}} (\boldsymbol h) $.
+From this, it follows that
+\begin{eqnarray*}
+ \REP_{(\boldsymbol X, \boldsymbol H)}(\boldsymbol \theta_N) &=&  \max_{\boldsymbol h}\max_{\boldsymbol x}f_{\boldsymbol \theta_N} (\boldsymbol x , \boldsymbol h) -  \min_{\boldsymbol h}\min_{\boldsymbol x}f_{\boldsymbol \theta_N} (\boldsymbol x , \boldsymbol h)\\&=&
+   \max_{\boldsymbol h_1 }  \max_{\boldsymbol h_2 }\left[ (\boldsymbol h_1 - \boldsymbol h_2)^T \boldsymbol \theta_N^{\mathcal{H}}  +   a_{\boldsymbol \theta_N, \mathcal{H}} (\boldsymbol h_1)  + a_{\boldsymbol \theta_N, \mathcal{H}} (\boldsymbol h_2)\right],
+\end{eqnarray*}
+which leads to the upper bound $ \REP_{(\boldsymbol X, \boldsymbol H)}(\boldsymbol \theta_N) \leq 2 \Gam + 2 |\boldsymbol \theta_N^{\mathcal{H}} |_1$.
+Then, taking $\boldsymbol h_1=\boldsymbol h_2$ (i.e., before maximization) gives $\REP_{(\boldsymbol X, \boldsymbol H)}(\boldsymbol \theta_N)
+\geq 2\Gam$ and taking $\boldsymbol h_1=-\boldsymbol h_2$, such that  $\boldsymbol h_1^T \boldsymbol \theta_N^{\mathcal{H}} = |\theta_N^{\mathcal{H}}|_1$, gives $\REP_{(\boldsymbol X, \boldsymbol H)}(\boldsymbol \theta_N)
+\geq 2|\theta_N^{\mathcal{H}}|_1 $; this yields the lower bound $\REP_{(\boldsymbol X, \boldsymbol H)}(\boldsymbol \theta_N)\geq 2\max\{\Gam,  |\boldsymbol \theta_N^{\mathcal{H}} |_1\}$.
+
+We next consider $\elt$ and, by (\ref{eqn:max}), write
+\begin{eqnarray*}
+\elt &=&  \max_{\boldsymbol h}\max_{\boldsymbol x}f_{\boldsymbol \theta_N} (\boldsymbol x , \boldsymbol h) -  \max_{\boldsymbol h}\min_{\boldsymbol x}f_{\boldsymbol \theta_N} (\boldsymbol x , \boldsymbol h)\\&=&
+   \max_{\boldsymbol h_1 }  \min_{\boldsymbol h_2 }\left[ (\boldsymbol h_1 - \boldsymbol h_2)^T \boldsymbol \theta_N^{\mathcal{H}}  +   a_{\boldsymbol \theta_N, \mathcal{H}} (\boldsymbol h_1)  + a_{\boldsymbol \theta_N, \mathcal{H}} (\boldsymbol h_2)\right].
+\end{eqnarray*}
+Taking $\boldsymbol h_1=\boldsymbol h_2$  and maximizing over both $\boldsymbol h_1,\boldsymbol h_2$ produces the upper bound
+$\elt \leq 2\Gam$.
+Then, using $ (\boldsymbol h_1 - \boldsymbol h_2)^T \boldsymbol \theta_N^{\mathcal{H}}+ a_{\boldsymbol \theta_N, \mathcal{H}} (\boldsymbol h_2) \geq - 2|\boldsymbol \theta_N^{\mathcal{H}} |_1$ and maximizing over   $\boldsymbol h_1$ gives
+$\elt \geq  \Gam- 2|\boldsymbol \theta_N^{\mathcal{H}} |_1$, while setting $\boldsymbol h_1=\boldsymbol h_2^*$
+for $\boldsymbol h_2^*$ such that  $
+  - (\boldsymbol h_2^*) ^T \boldsymbol \theta_N^{\mathcal{H}}  +     a_{\boldsymbol \theta_N, \mathcal{H}} (\boldsymbol h_2^*) = \min_{\boldsymbol h_2} [ -  \boldsymbol h_2  ^T \boldsymbol \theta_N^{\mathcal{H}}  +     a_{\boldsymbol \theta_N, \mathcal{H}} (\boldsymbol h_2)]$ gives
+  $\elt \geq 2   a_{\boldsymbol \theta_N, \mathcal{H}} (\boldsymbol h_2^*) \geq \Gamc \equiv \min_{\boldsymbol h}  a_{\boldsymbol \theta_N, \mathcal{H}} (\boldsymbol h)  $.
+Finally, note that for any $\boldsymbol h$, the triangle inequality gives
+  \begin{eqnarray*}
+  \Gam  \equiv \max_{\boldsymbol h_1}  a_{\boldsymbol \theta_N, \mathcal{H}} (\boldsymbol h_1) &\geq&
+  [a_{\boldsymbol \theta_N, \mathcal{H}} (\boldsymbol h)  + a_{\boldsymbol \theta_N, \mathcal{H}} (-\boldsymbol h)]/2\\
+  &=&
+    2^{-1}\sum_{i=1}^{N } \left(\left| \theta_{i,N}^{\mathcal{V}}   + \sum_{j=1}^{N_{\mathcal{H}}} h_j \theta_{ij,N}^{\mathcal{VH}} \right|
+   + \left| \theta_{i,N}^{\mathcal{V}}   - \sum_{j=1}^{N_{\mathcal{H}}} h_j \theta_{ij,N}^{\mathcal{VH}} \right|\right)\\
+   & \geq &\sum_{i=1}^{N }  \left| \theta_{i,N}^{\mathcal{V}} \right| \equiv |\boldsymbol \theta_{N}^{\mathcal{V}}|_1. \Box\\\end{eqnarray*}
+
+%###################
+\blue{
+
+\noindent\textbf{Proof of Theorem~\ref{thm:something}.}  Let $L_{\boldsymbol \theta_{N}}(\boldsymbol X) = \log[ P_{\boldsymbol \theta_{N}}(\boldsymbol X)/ \min_{\boldsymbol y \in \mathcal{X}^N} P_{\boldsymbol \theta_{N}}(\boldsymbol y) ]/\REP(\boldsymbol \theta_{N})$, where again    $\boldsymbol X=(X_1, \dots, X_N)$ and $\REP(\boldsymbol \theta_{N})= \log[\max_{\boldsymbol y \in \mathcal{X}^N} P_{\boldsymbol \theta_{N}}(\boldsymbol y)/ \min_{\boldsymbol y \in \mathcal{X}^N} P_{\boldsymbol \theta_{N}}(\boldsymbol y) ]$.
+As $L_{\boldsymbol \theta_{N}}(\boldsymbol X)\in[0,1]$, convergence of   $L_{\boldsymbol \theta_{N}}(\boldsymbol X)$ to 1 in probability
+under $P_{\boldsymbol \theta_{N}}$ is equivalent to convergence to 1
+  in expectation under $P_{\boldsymbol \theta_{N}}$ (i.e., convergence in expectation implies probabilistic convergence by Markov's inequality while probabilistic convergence implies  convergence in expectation by uniform integrability/boundedness).
+
+  For $\epsilon \in(0,1)$, let $\mathcal{M}_{\epsilon, \boldsymbol \theta_N}$ denote a modal set as in (\ref{eqn:mode}).  By Theorem~\ref{degenFOES},
+$P_{\boldsymbol \theta_N}\left( \boldsymbol X\in \mathcal{M}_{\epsilon, \boldsymbol \theta_N}\right) \rightarrow 1$ holds as $N \rightarrow \infty$ and, by definition of (\ref{eqn:mode}), $\boldsymbol X\in \mathcal{M}_{\epsilon, \boldsymbol \theta_N}$ follows if and only if
+$1-L_{\boldsymbol \theta_{N}}(\boldsymbol X)<\epsilon$.  Hence, $L_{\boldsymbol \theta_{N}}(\boldsymbol X)
+\stackrel{\mbox{$p,\mathrm{E}$}}{\longrightarrow} 1$ holds under $\boldsymbol \theta_{N} $ in Theorem~\ref{thm:something}.  The convergence
+ $L_{-\boldsymbol \theta_{N}}(\boldsymbol X)
+\stackrel{\mbox{$p,\mathrm{E}$}}{\longrightarrow} 1$  under $-\boldsymbol \theta_{N} $  likewise follows from Corollary~\ref{sign}. \hfill $\Box$
+
+%###############
+}
 
 \clearpage
-
 \section*{References}\label{references}

### 2017-11-02 writing/paper-note/docs/note.tex
Revert "nordman's changes"
This reverts commit 2972c56368bdf5d74b94c5b34c0788d109a74bad.

--- 2017-09-30 writing/paper-note/docs/note.tex
@@
             pdfauthor={},
-            pdftitle={A note on the instability and degeneracy of deep learning models},
+            pdftitle={A Note on the Instability and Degeneracy of Deep Learning Models},
             colorlinks=true,
@@
 
-\title{\bf A note on the instability and degeneracy of deep learning models}
+\title{\bf A Note on the Instability and Degeneracy of Deep Learning Models}
 \author{Andee Kaplan \\ Department of Statistics, Iowa State University \\  and \\ Daniel Nordman \\ Department of Statistics, Iowa State University \\  and \\ Stephen Vardeman \\ Department of Statistics and Department of Industrial and\\
@@
 
-We consider the behavior, and the potential impropriety, sequences of
+We consider the behavior, and the potential impropriety, of sequences of
 discrete probability models built to incorporate observations of
@@
 to only a subset of an overall sample space. The latter issue in
-connection to degeneracy has been recognized as concern in that dominant
-model outcomes may not resemble observed data (cf. Handcock
+connection to degeneracy has been recognized as a concern in that
+dominant model outcomes may not resemble observed data (cf. Handcock
 \protect\hyperlink{ref-handcock2003assessing}{2003}). As a compounding
@@
 (\protect\hyperlink{ref-schweinberger2011instability}{2011}) originally
-developed instability results framed for, or specific to, a certain
-class of discrete exponential models. For similar exponential models
-with random networks, Handcock
-(\protect\hyperlink{ref-handcock2003assessing}{2003}) studied model
-degeneracy, where a probability model places near complete mass on modes
-and may thereby narrow the feasible model outcomes. As findings here and
-from Schweinberger
+developed instability results specific to a certain class of discrete
+exponential models. For similar exponential models with random networks,
+Handcock (\protect\hyperlink{ref-handcock2003assessing}{2003}) studied
+model degeneracy, where a probability model places near complete mass on
+modes and may thereby narrow the feasible model outcomes. As findings
+here and from Schweinberger
 (\protect\hyperlink{ref-schweinberger2011instability}{2011}) suggest,
@@
 simulations. As one potential remedy, suggestions are given or
-constraining model parameterizations and avoid the most problematic
+constraining model parameterizations to avoid the most problematic
 regions of the parameter space. Proofs of the main results appear in
@@
 
-\subsection{Discrete exponential family
-models}\label{discrete-exponential-family-models}
+\subsection{Discrete Exponential Family
+Models}\label{discrete-exponential-family-models}
 
@@
 \subsection{Restricted Boltzmann
-machines}\label{restricted-boltzmann-machines}
+Machines}\label{restricted-boltzmann-machines}
 
@@
 Additionally, one may further arbitrarily choose the number
-\(N_\mathcal{H}\) of hidden variables \(\boldsymbol H\) in the joint RMB
+\(N_\mathcal{H}\) of hidden variables \(\boldsymbol H\) in the joint RBM
 model \eqref{eq:RBM1} to define a marginal model \eqref{eq:RBM2} for the
@@
 
-\subsection{Deep learning}\label{deep-learning}
+\subsection{Deep Learning}\label{deep-learning}
 
@@
 stacked layer in a DBN are Bayesian networks (see Pearl
-\protect\hyperlink{ref-pearl985bayesian}{1985}), rather than RBMs. Thus
-for visibles \(X_1, \dots, X_N\) with support \(\mathcal{X}^N\), a DBN
-is also a FOES model if the number \(q(N)\) of components in the
-parameter vector is dependent on the dimension of the visibles.
-Commonly, as in logistic belief nets (Neal
-\protect\hyperlink{ref-neal1992connectionist}{1992}), a ``weight''
+\protect\hyperlink{ref-pearl985bayesian}{1985}), rather than RBMs. A
+Bayesian network is a probabilistic graphical model that defines
+conditional dependence to be directed, rather than undirected (as with
+the RBM). Thus for visibles \(X_1, \dots, X_N\) with support
+\(\mathcal{X}^N, \mid \mathcal{X} \mid < \infty\), a DBN is also a FOES
+model with \(q(N)\) the length of parameter vector is dependent on the
+dimension of the visibles because
+\(P_{\boldsymbol \theta_N}(\boldsymbol x)>0\) for all
+\(\boldsymbol x \in\mathcal{X}^N\). Commonly, as in logistic belief nets
+(Neal \protect\hyperlink{ref-neal1992connectionist}{1992}), a ``weight''
 parameter is placed on each interaction between visibles,
 \(X_1, \dots, X_N\), and the first layer of latent variables,
-\(H^{(1)}_1, \dots, H^{(1)}_{N_{(H,1)}}\), in the definition of a FOES
-model.
-
-\section{Main results on model instability}\label{instability-results}
-
-\subsection{A criterion for instability}\label{criterion}
+\(H^{(1)}_1, \dots, H^{(1)}_{N_{(H,1)}}\), satisfying the definition of
+a FOES model.
+
+\section{Main Results on Model Instability}\label{instability-results}
+
+We now present a formal definition for instability of FOES models as
+well as a simple condition for identifying instability in a FOES model
+sequence.
+
+\subsection{A Criterion for Instability}\label{criterion}
 
@@
 associated with a FOES model formulation for \(N\) random variables
-where the log-ratio \eqref{eq:elpr} is overly large relative to the sample
-size \(N\). That is, a sequence of FOES probability models
+where the LREP \eqref{eq:elpr} is overly large relative to the sample size
+\(N\). That is, a sequence of FOES probability models
 \(P_{\boldsymbol \theta_N}\) results in specifying the distribution of
@@
 (\protect\hyperlink{ref-schweinberger2011instability}{2011}) by allowing
-possibility non-exponential family models (e.g., RBM and DBM models in
-Sections \ref{restricted-boltzmann-machines} - \ref{deep-learning} as
-well as a potentially increasing number \(d(N)\) of parameters through
-the parameter sequence \(\boldsymbol \theta_N\in \mathbb{R}^{d(N)}\).
-While this definition differs in form and scope from the original, it
-does match that in Schweinberger
+possibly non-exponential family models (e.g., RBM and DBM models in
+Sections \ref{restricted-boltzmann-machines}-\ref{deep-learning} as well
+as a potentially increasing number \(q(N)\) of parameters through the
+parameter sequence \(\boldsymbol \theta_N\in \mathbb{R}^{q(N)}\). While
+this definition differs in form and scope from the original, it does
+match that in Schweinberger
 (\protect\hyperlink{ref-schweinberger2011instability}{2011}) for the
@@
 next describe several potentially undesirable features associated with
-unstable models.
+S-unstable FOES models.
 
@@
 
-\subsection{Characterizations and consequences of
-instability}\label{characterizations-and-consequences-of-instability}
+\subsection{Characterizations and Consequences of
+Instability}\label{characterizations-and-consequences-of-instability}
 
@@
 \(\boldsymbol \theta_N\). We then have the following result prescribing
-the behavior of \(\DN(\boldsymbol \theta_N)\) for unstable models.
+the behavior of \(\DN(\boldsymbol \theta_N)\) for S-unstable FOES
+models.
 
@@
 have some such data outcomes with this property. As a consequence,
-unstable probability structure may exhibit extreme sensitivity through
+unstable probability structures may exhibit extreme sensitivity through
 large peaks and troughs over the sample space.
@@
 become degenerate. Theorem \ref{thm:degenFOES} provides a formal
-statement of such degeneracy due to instability. For a given
+statement of such degeneracy due to S-instability. For a given
 \(0 < \epsilon < 1\), define a \(\epsilon\)-modal set of outcomes as
@@
 
-Connected to degeneracy, unstable models may also exhibit additional
-kinds of extreme and undesirable sensitivity in probabilities if model
-parameters \(\boldsymbol \theta_N\) can further be ``dialed'' between
-positive and negative values. That is, some FOES models naturally
-involve parameter spaces covering a positive-negative spectrum of
-parameter possibilities, where the signs of parameters provide a
+Connected to degeneracy, S-unstable FOES models may also exhibit
+additional kinds of extreme and undesirable sensitivity in probabilities
+if model parameters \(\boldsymbol \theta_N\) can further be ``dialed''
+between positive and negative values. That is, some FOES models
+naturally involve parameter spaces covering a positive-negative spectrum
+of parameter possibilities, where the signs of parameters provide a
 standard device for increasing or decreasing probabilities of outcomes
@@
 
-\BeginKnitrBlock{remark}[need to briefly finish]
-\iffalse <span class="remark"><em>Remark (need to briefly finish). \fi{}
-extensions of sensitivity\ldots{} basically, model condition PSR and
+\BeginKnitrBlock{remark}
+\iffalse <span class="remark"><em>Remark. \fi{} Model condition PSR and
 Corollary \ref{cor:sign} can be extended to cases where parameter
@@
 generally, are instead altered to another parameter configuration
-\(\boldsymbol \theta_{N}^A = ( \boldsymbol \theta_{1,N}^A, \boldsymbol \theta_{2,N}^A )\)
+\(\boldsymbol \theta_{N}^A = (\boldsymbol \theta_{N,1}^A, \boldsymbol \theta_{N,2}^A)\)
 involving a switch in sign only among some model parameters
-\(\boldsymbol \theta_{N,2}^A=- \boldsymbol \theta_{N,2}\) with remaining
-parameters \(\boldsymbol \theta_{1,N}^A\) being arbitrarily chosen. If a
-change sign occurs among parameters (\(\pm \boldsymbol \theta_{N,2}\))
-which dominate the probability structure of the model, then an unstable
-model under \(\boldsymbol \theta_N\) will then imply that many more
-unstable models exist over a broad spectrum of possibilities for
-variations \(\boldsymbol \theta_{N}^A\) of \(\boldsymbol \theta_N\).
+\(\boldsymbol \theta_{N,2}^A = -\boldsymbol \theta_{N,2}\) with
+remaining parameters \(\boldsymbol \theta_{1,N}^A\) being arbitrarily
+chosen. If a sign change occurs among parameters
+(\(\pm \boldsymbol \theta_{N,2}\)) which dominate the probability
+structure of the model, then an unstable model under
+\(\boldsymbol \theta_N\) will then imply that many more unstable models
+exist over a broad spectrum of possibilities for variations
+\(\boldsymbol \theta_{N}^A\) of \(\boldsymbol \theta_N\).
 \EndKnitrBlock{remark}
@@
 still lead to instability if the parameter dimension is high. We next
-provides some examples to illustrate instability in FOES models, which
+provides some examples to illustrate S-instability in FOES models, which
 also suggest some potential strategies for preventing unstable models.
 
-\subsection{Equi-probability models}\label{equi-probability-models}
+\subsection{Equi-probability Models}\label{equi-probability-models}
 
@@
 
-\subsection{One-parameter exponential models}\label{one-param-exp}
+\subsection{One-parameter Exponential Models}\label{one-param-exp}
 
@@
 
-\subsection{Fixed-dimensional linear exponential
-models}\label{fixed-dim-exp}
+\subsection{Fixed-dimensional Linear Exponential
+Models}\label{fixed-dim-exp}
 
@@
 under \(\theta_2 = - \theta_3/3\) case potentially allowed by
-Schweinberger's results.
-
-\subsection{Latent variable models of increasing parameter
-dimension}\label{latent-variable-models-of-increasing-parameter-dimension}
+Schweinberger's
+(\protect\hyperlink{ref-schweinberger2011instability}{2011}) results.
+
+\subsection{Latent Variable Models of Increasing Parameter
+Dimension}\label{latent-variable-models-of-increasing-parameter-dimension}
 
@@
 exponential formulations involving hidden, or latent, variables, such as
-those described in Sections
-\ref{restricted-boltzmann-machines}-\ref{deep-learning} for networks. We
-will focus on restricted Bolzmann machine (RBM) models (Section
+those probabilistic graphical models described in Sections
+\ref{restricted-boltzmann-machines}-\ref{deep-learning}. We will focus
+on restricted Bolzmann machine (RBM) models (Section
 \ref{restricted-boltzmann-machines}, having one layer of latent
@@
 visible and latent variables \((\boldsymbol X, \boldsymbol H)\), where
-\(\boldsymbol H=(H_1,\ldots,H_{N_{\mathcal{H}}})\) denotes a vector of
+\(\boldsymbol H = (H_1,\ldots,H_{N_{\mathcal{H}}})\) denotes a vector of
 \(N_{\mathcal{H}}\) hidden variables (similarly binary). The joint model
 is of linear exponential form involving
-\(d(N)\equiv N + N_{\mathcal{H}} + N* N_{\mathcal{H}}\) sufficient
+\(q(N)\equiv N + N_{\mathcal{H}} + N*N_{\mathcal{H}}\) sufficient
 statistics given by
@@
 parameters
-\(\boldsymbol \theta_N = (\boldsymbol \theta_N^{\mathcal{V}},\boldsymbol \theta_N^{\mathcal{H}}, \boldsymbol \theta_N^{\mathcal{VH}} ) \in\mathbb{R}^{d(N)}\)
+\(\boldsymbol \theta_N = (\boldsymbol \theta_N^{\mathcal{V}},\boldsymbol \theta_N^{\mathcal{H}}, \boldsymbol \theta_N^{\mathcal{VH}} ) \in\mathbb{R}^{q(N)}\)
 corresponding to the \(N\) visible variables \(\boldsymbol X\) (i.e.,
@@
 \(\boldsymbol X^T\boldsymbol H\) (i.e.,
-\(\boldsymbol \theta_N^{\mathcal{VH}}\in\mathbb{R}^{N N_{\mathcal{H}}}\)).
+\(\boldsymbol \theta_N^{\mathcal{VH}}\in\mathbb{R}^{N*N_{\mathcal{H}}}\)).
 However, unlike some previous exponential models considered in Sections
@@
 sufficient statistics exceeding the sample size \(N\). Instead, RBM
-instability may be linked to solely to parameter configuration and the
-fact that the number \(d(N) \geq N\) of parameters necessarily increases
-with the number \(N\) of observations \(\boldsymbol X\), in contrast to
+instability may be linked solely to parameter configuration and the fact
+that the number \(q(N) \geq N\) of parameters necessarily increases with
+the number \(N\) of observations \(\boldsymbol X\), in contrast to
 previous exponential cases of fixed parameter dimension.
@@
 \eqref{eq:RBM1}-\eqref{eq:RBM2} coincide. An independence model then results
-for variables in \(\boldsymbol X\), which has \(d(N)=N\) parameters
+for variables in \(\boldsymbol X\), which has \(q(N)=N\) parameters
 \(\boldsymbol \theta_N^{\mathcal{V}} = (\theta_{1,N}^{\mathcal{V}}, \ldots, \theta_{N,N}^{\mathcal{V}}) \in \mathbb{R}^{N}\),
@@
 \(N_{\mathcal{H}}\equiv N_{N,\mathcal{H}}\) of \(N\)), which can
-substantially inflate the number \(d(N)\) of model parameters and
+substantially inflate the number \(q(N)\) of model parameters and
 further impact model instability through accumulated parameters. To
@@
 \protect\hypertarget{prp:prop2}{}{\label{prp:prop2}}Let
-\(P_{\boldsymbol \theta_N}\) denote a RBM-based model \eqref{eq:RBM2} for
-\(N\geq 1\) visible variables \(\boldsymbol X \equiv \boldsymbol X_N\)
-derived from \(\tilde{P}_{\boldsymbol \theta_N}\) as the joint RBM
-distribution \eqref{eq:RBM1} of \((\boldsymbol X, \boldsymbol H)\)
-involving some number \(N_{\mathcal{H}} \equiv N_{N,\mathcal{H}}\geq 0\)
-of hidden variables \(\boldsymbol H \equiv \boldsymbol H_N\) and
-parameters
-\(\boldsymbol \theta_N \equiv (\boldsymbol \theta_N^{\mathcal{V}},\boldsymbol \theta_N^{\mathcal{H}}, \boldsymbol \theta_N^{\mathcal{VH}}) \in\mathbb{R}^{N}\times \mathbb{R}^{N_{\mathcal{H}}} \times \mathbb{R}^{N N_{\mathcal{H}}}\).
+\(P_{\boldsymbol \theta_N}\) denote a RBM-based data model \eqref{eq:RBM2}
+for \(N\geq 1\) visible variables
+\(\boldsymbol X \equiv \boldsymbol X_N\) derived from
+\(\tilde{P}_{\boldsymbol \theta_N}\) as the joint RBM distribution
+\eqref{eq:RBM1} of \((\boldsymbol X, \boldsymbol H)\) involving some
+number \(N_{\mathcal{H}} \equiv N_{N,\mathcal{H}}\geq 0\) of hidden
+variables \(\boldsymbol H \equiv \boldsymbol H_N\) and parameters
+\(\boldsymbol \theta_N \equiv (\boldsymbol \theta_N^{\mathcal{V}},\boldsymbol \theta_N^{\mathcal{H}}, \boldsymbol \theta_N^{\mathcal{VH}}) \in\mathbb{R}^{N}\times \mathbb{R}^{N_{\mathcal{H}}} \times \mathbb{R}^{N*N_{\mathcal{H}}}\).
 Then,
 \begin{enumerate}[(i)]
-\item the instability measure $\REP_{(\boldsymbol X)}(\boldsymbol \theta_N)$ for the marginal model $P_{\boldsymbol \theta_N}$ of  $\boldsymbol X$ satisfies
+\item the instability measure $\REP(\boldsymbol \theta_N)$ for the marginal model $P_{\boldsymbol \theta_N}$ of  $\boldsymbol X$ satisfies
     $$
-    \left| \REP_{(\boldsymbol X)}(\boldsymbol \theta_N)  - \elt\right| \leq    N_{\mathcal{H}}  \log 2
+    \left| \REP(\boldsymbol \theta_N)  - \elt\right| \leq    N_{\mathcal{H}}  \log 2
     $$
@@
 
-In more complicated network models involving further or deeper hidden
+In more complicated graphical models involving further or deeper hidden
 layers, the same issues and causes of instability similarly exist, but
@@
 
-\section{Statistical consequences of instability}\label{implications}
+\section{Statistical Consequences of Instability}\label{implications}
 
@@
 
-\subsection{Implications for
-simulation}\label{implications-for-simulation}
+\subsection{Implications for Simulation}\label{mcmc}
 
@@
 have extreme log-ratios in its transition probabilities \[
-\left| \frac{P_{\boldsymbol \theta_N} (X_i =x_1 | \boldsymbol x^{(1)}_{-i})}{P_{\boldsymbol \theta_N} (X_i = x_2| \boldsymbol x^{(2)}_{-i}) } \right|=\left| \frac{P_{\boldsymbol \theta_N} ( \boldsymbol x^{(1)} }{P_{\boldsymbol \theta_N} (  \boldsymbol x^{(2)} ) } \right|
+\left| \frac{P_{\boldsymbol \theta_N} (X_i =x_1 | \boldsymbol x^{(1)}_{-i})}{P_{\boldsymbol \theta_N} (X_i = x_2| \boldsymbol x^{(2)}_{-i}) } \right|=\left| \frac{P_{\boldsymbol \theta_N} ( \boldsymbol x^{(1)} )}{P_{\boldsymbol \theta_N} (  \boldsymbol x^{(2)} ) } \right|
 \] using that conditional probabilities are proportional to the joint
@@
 log-probability ratios in one-component changes (Theorem
-\ref{thm:instab-elpr}. This can hinder the ability of chain to
+\ref{thm:instab-elpr}). This can hinder the ability of chain to
 effectively explore the sample space of the observations
 \(\boldsymbol X\), as the chain may then mix poorly by moving rapidly
-to, and slowly away from, sections of the sample space surface. Because
-the numerical problem is unbounded ratios of probabilities under the
-joint model, similar complications can also arise for
-Metropolis-Hastings algorithms for MCMC. Hence, while an unstable FOES
-model \(P_{\boldsymbol \theta_N}\) may be valid and technically open to
-simulation by MCMC, the aspect of instability can ruin numerical
+to, and slowly away from, sections of the sample space surface. This
+mixing problem is due to the unstable stationary distribution (unbounded
+ratios of probabilities under the joint model), rather than in the
+particulars of the MCMC algorithm, similar complications can also arise
+for Metropolis-Hastings algorithms for MCMC. Hence, while an unstable
+FOES model \(P_{\boldsymbol \theta_N}\) may be valid and technically
+open to simulation by MCMC, the aspect of instability can ruin numerical
 applications of MCMC.
 
-\subsection{Implications for maximum likelihood
-inference}\label{implications-for-maximum-likelihood-inference}
+\subsection{Implications for Maximum Likelihood
+Inference}\label{implications-for-maximum-likelihood-inference}
 
@@
 \(\boldsymbol x\in\mathcal{X}^N\), for \(N\) observations
-\(\boldsymbol X =(X_1,\ldots,X_n)\), define the likelihood function
+\(\boldsymbol X =(X_1,\ldots,X_N)\), define the likelihood function
 \(\mathcal{L}(\boldsymbol \theta_N | \boldsymbol x) \equiv N^{-1}\log P_{\boldsymbol \theta_N }(\boldsymbol x)\)
@@
 where
-\(\min_{\boldsymbol \theta_N} \equiv \min_{\boldsymbol y\in\mathcal{X}^N} P_{\boldsymbol \theta_N }(\boldsymbol y)]\).
+\(\min_{\boldsymbol \theta_N} \equiv \min_{\boldsymbol y\in\mathcal{X}^N} P_{\boldsymbol \theta_N }(\boldsymbol y)\).
 Also define
-\(\max_{\boldsymbol \theta_N} \equiv \max_{\boldsymbol y\in\mathcal{X}^N} P_{\boldsymbol \theta_N }(\boldsymbol y)]\).
+\(\max_{\boldsymbol \theta_N} \equiv \max_{\boldsymbol y\in\mathcal{X}^N} P_{\boldsymbol \theta_N }(\boldsymbol y)\).
 
@@
 \(\lim_{N\to \infty}\REP(\boldsymbol \theta_N)/N=\infty\)) which
-additionally satisfies model condition SPR. Then,
+additionally satisfies model condition PSR. Then,
 \begin{align*}
@@
 \]
-\EndKnitrBlock{theorem} This needs to be connected to results in
-Schweinberger who considered the one-parameter linear exponential case
-showing that, for positive \(\boldsymbol \theta_N>0\), \[
+\EndKnitrBlock{theorem} Schweinberger
+(\protect\hyperlink{ref-schweinberger2011instability}{2011} Corollary 1)
+considered the one-parameter linear exponential case showing that, for
+positive \(\boldsymbol \theta_N>0\), \[
 \frac{1}{\REP(\boldsymbol \theta_N)} \log\left[ \frac{P_{\boldsymbol \theta_N }(\boldsymbol x) }{\min_{\boldsymbol \theta_N}}\right] = \frac{\boldsymbol g(\boldsymbol x) - L_N}{U_N-L_N} 
@@
 \ref{thm:something} and the LREP expansion \eqref{eq:UL} for the one
-parameter exponential model case.
+parameter exponential model case and is related to a result in Handcock
+(\protect\hyperlink{ref-handcock2003assessing}{2003}).
 
@@
 opposite signed parameters. This can cause numerical complications in
-likelihood maximization. The likelihood function knows what to do with
-an observed outcome
+likelihood maximization. If an observed outcome
 \(\boldsymbol x \equiv \argmax \mathcal{L}(\boldsymbol \theta_N)\) which
 is also the outcome with minimal probability under
-\(P_{-\boldsymbol \theta_N}\). But unless the observed outcome
-\(\boldsymbol x\) is at an extreme in the probability structures of
-\(P_{\boldsymbol \theta_N}, P_{-\boldsymbol \theta_N}\) (which are
+\(P_{-\boldsymbol \theta_N}\), then the likelihood function will have a
+clear peak and maximization straightforward. However, when the observed
+outcome \(\boldsymbol x\) is at an extreme in the probability structures
+of \(P_{\boldsymbol \theta_N}, P_{-\boldsymbol \theta_N}\) (which are
 related) the likelihood function will be flat and optimization steps may
@@
 \subsection{Implications for Bayes
-inference}\label{implications-for-bayes-inference}
+Inference}\label{implications-for-bayes-inference}
 
@@
 \ref{cor:sign}), then this will create numerical difficulties in Bayes
-inference based on MCMC. We can write the MH transition density to
-indicate this.
+inference based on MCMC. Given observations
+\(\boldsymbol X = (X_1,\ldots,X_N)\), switching \(\boldsymbol theta_N\)
+from positive to negative or vice versa could dramatically change
+probabilities (model condition PSR and Corollary \ref{cor:sign}). This
+would translate to numerical challenges similar to those presented in
+Section \ref{mcmc}, i.e., inability of the chain to effectively explore
+the sample space because of unbounded ratios of probabilities. We can
+write the MH acceptance probability to indicate this, \[
+\alpha\left(\boldsymbol x^{(1)} \mid \boldsymbol x^{(2)}\right)= \min\left\{1, 
+\frac{q(\boldsymbol x^{(2)} \mid \boldsymbol x^{(1)})}{q(\boldsymbol x^{(1)} \mid \boldsymbol x^{(2)})}\frac{P_{\boldsymbol \theta_N} ( \boldsymbol x^{(1)} )}{P_{\boldsymbol \theta_N} (  \boldsymbol x^{(2)} ) } \right\}
+\] for some proposal distribution \(q(\cdot)\).
 
@@
 caution should be used in parameter specification to control effects of
-model instability.
+model instability due to the consequences of instability in both
+simulation and likelihood inference.
 
@@
 
-\textbf{Proof of Proposition \ref{prp:instab-elpr}.} We prove the
-contrapositive, supposing that \(\Delta(\boldsymbol \theta_N) \le C\)
-holds for some \(C > 0\) and show
-\(\text{ELPR}(\boldsymbol \theta_N) \leq NC\). Let
+\textbf{Proof of Proposition \ref{thm:instab-elpr}.} For part (i), we
+prove the contrapositive, supposing that
+\(\DN(\boldsymbol \theta_N) \le C\) holds for some \(C > 0\) and show
+\(\REP(\boldsymbol \theta_N) \leq NC\). Let
 \(\boldsymbol x_{min} \equiv \argmin\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)\)
@@
 (i.e. \(\boldsymbol x_i, \boldsymbol x_{i + 1} \in \mathcal{X}^N\)
-differ in exactly \(1\) component for \(i = 0, \dots, k\)) for some
-integer \(k \in \{0, 1, \dots, N\}\). Under the FSFS model, recall
+differ in exactly \(1\) component, \(i = 0, \dots, k\)) for some integer
+\(k \in \{0, 1, \dots, N\}\). Under the FOES model, recall
 \(P_{\boldsymbol \theta_N}(\boldsymbol x) > 0\) holds so that
@@
 \(k = 0\), then \(\boldsymbol x_{max} = \boldsymbol x_{min}\) and the
-same bound above holds.\hfill \(\Box\)
-
-\textbf{Proof of Proposition \ref{prp:degenFSFS}.} Where
-\(|\mathcal{X}|<\infty\) holds in the FSFS model, we may suppose
-\(|\mathcal{X}|>1\) (i.e., \(\mathcal{X}^N\) has more than one outcome)
-because otherwise the model is trivially degenerate for all
-\(N \geq 1\). Fix \(0 < \epsilon < 1\). Then,
-\(\boldsymbol x_{max} \in M_{\epsilon, \boldsymbol \theta_N}\), so
+same bound above holds. This establishes part (i). To show part (ii),
+note the definition of S-instability (i.e.,
+\(\lim_{N\to \infty}\REP(\boldsymbol \theta_N)/N= \infty\)) combined
+with part (i) implies that
+\(\lim_{N\to \infty}\DN(\boldsymbol \theta_N)=\infty\). \hfill \(\Box\)
+
+\textbf{Proof of Proposition \ref{thm:degenFOES}.} As
+\(|\mathcal{X}|<\infty\) holds in the FOES model, we may suppose
+\(|\mathcal{X}|>1\); otherwise, \(\mathcal{X}^N\) has one outcome and
+the model is trivially degenerate for all \(N \geq 1\). Fix
+\(0 < \epsilon < 1\) and write
+\(\boldsymbol x_{min} \equiv \argmin\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)\)
+and
+\(\boldsymbol x_{max} \equiv \argmax\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)\).
+Then, \(\boldsymbol x_{max} \in M_{\epsilon, \boldsymbol \theta_N}\), so
 \(P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N}) \ge P_{\boldsymbol \theta_N}(\boldsymbol x_{max}) > 0\).
@@
 \(P_{\boldsymbol \theta_N}(\boldsymbol x) \le [P_{\boldsymbol \theta_N}(\boldsymbol x_{max})]^{1-\epsilon}[P_{\boldsymbol \theta_N}(\boldsymbol x_{min})]^{\epsilon}\)
-holds so that
-\begin{align*}
+holds so that \[
 1-P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N})
-& = \sum\limits_{\boldsymbol x \in \mathcal{X}^N \setminus M_{\epsilon, \boldsymbol \theta_N}}P_{\boldsymbol \theta_N}(\boldsymbol x) \\
-& \le (|\mathcal{X}|^N)[P_{\boldsymbol \theta_N}(\boldsymbol x_{max})]^{1-\epsilon}[P_{\boldsymbol \theta_N}(\boldsymbol x_{min})]^{\epsilon}.
-\end{align*}
-From the lower bound on
+ = \sum\limits_{\boldsymbol x \in \mathcal{X}^N \setminus M_{\epsilon, \boldsymbol \theta_N}}P_{\boldsymbol \theta_N}(\boldsymbol x)
+  \le (|\mathcal{X}|^N)[P_{\boldsymbol \theta_N}(\boldsymbol x_{max})]^{1-\epsilon}[P_{\boldsymbol \theta_N}(\boldsymbol x_{min})]^{\epsilon}.
+\] From the lower bound on
 \(P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N})\) and the
@@
 \end{align*}
-as \(N \rightarrow \infty\) by the definition of an S-unstable FSFS
-model (cf.~Definition \ref{def:instabFSFS}). Consequently,
+as \(N \rightarrow \infty\) by the definition of an S-unstable FOES
+model \eqref{eq:Sun}. Consequently,
 \(P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N}) \rightarrow 1\)
 as \(N \rightarrow \infty\) as claimed. \hfill \(\Box\)
+
+\textbf{Proof of Corollary \ref{cor:sign}.} The model condition PSR
+implies that
+\begin{equation}
+\label{eq:R}
+\frac{\max_{\boldsymbol y \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol y)  }{\min_{\boldsymbol y \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol y) } = \frac{\max_{\boldsymbol y \in \mathcal{X}^N}P_{-\boldsymbol \theta_N}(\boldsymbol y) }{\min_{\boldsymbol y \in \mathcal{X}^N}P_{-\boldsymbol \theta_N}(\boldsymbol y) }
+\end{equation}
+so that the log-ratio
+\(\REP(\boldsymbol \theta_N)=\REP(-\boldsymbol \theta_N)\) is the same
+for both \(\boldsymbol \theta_N\) and \(-\boldsymbol \theta_N\) in
+\eqref{eq:elpr}. Now part (i) of Corollary \ref{cor:sign} follows from
+\(\REP(\boldsymbol \theta_N)/N=\REP(-\boldsymbol \theta_N)/N\to \infty\)
+as \(N\to \infty\) in \eqref{eq:Sun}. To show part (ii), fix
+\(0 < \epsilon < 1\) and consider a \(\epsilon\)-mode set
+\(\mathcal{M}_{\epsilon, \boldsymbol \theta_N}\) under
+\(\boldsymbol \theta_N\) from \eqref{eq:mode}. If
+\(\boldsymbol x \in \mathcal{M}_{\epsilon, \boldsymbol \theta_N}^c \equiv \mathcal{X}^N \setminus \mathcal{M}_{\epsilon, \boldsymbol \theta_N}\),
+then, by definition, \[
+\frac{P_{\boldsymbol \theta_N}(\boldsymbol x)}{\min_{\boldsymbol y \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol y)} \leq \left[\frac{\max_{\boldsymbol y \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol y)  }{\min_{\boldsymbol y \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol y) } \right]^{1-\epsilon}
+\] holds, which is equivalent to \[
+\frac{\max_{\boldsymbol y \in \mathcal{X}^N}P_{-\boldsymbol \theta_N}(\boldsymbol y)}{P_{-\boldsymbol \theta_N}(\boldsymbol x)} \leq
+\left[\frac{\max_{\boldsymbol y \in \mathcal{X}^N}P_{-\boldsymbol \theta_N}(\boldsymbol y)  }{\min_{\boldsymbol y \in \mathcal{X}^N}P_{-\boldsymbol \theta_N}(\boldsymbol y) } \right]^{1-\epsilon}
+\] by model condition PSR and \eqref{eq:R}. The latter is in turn
+equivalent to
+\begin{equation}
+\label{eq:R2}
+\log P_{-\boldsymbol \theta_N}(\boldsymbol x) \geq \epsilon \max\limits_{\boldsymbol y \in \mathcal{X}^N} \log  P_{-\boldsymbol \theta_N}(\boldsymbol y) + (1-\epsilon)\min\limits_{\boldsymbol y \in \mathcal{X}^N} \log P_{-\boldsymbol \theta_N}(\boldsymbol y),
+\end{equation}
+so that
+\(\boldsymbol x \in \mathcal{M}_{\epsilon, \boldsymbol \theta_N}^c\) if
+and only if \eqref{eq:R2} holds. Next consider the \((1-\epsilon)\)-mode
+set \(\mathcal{M}_{1-\epsilon, -\boldsymbol \theta_N}\) under
+\(-\boldsymbol \theta_N\) from \eqref{eq:mode}. If
+\(\boldsymbol x \in\mathcal{M}_{1-\epsilon, -\boldsymbol \theta_N}\),
+then by definition \(\boldsymbol x\) fulfills \eqref{eq:R2} and so
+\(\boldsymbol x \in \mathcal{M}_{\epsilon, \boldsymbol \theta_N}^c\),
+showing that
+\(\mathcal{M}_{1-\epsilon, -\boldsymbol \theta_N} \subset \mathcal{M}_{\epsilon, \boldsymbol \theta_N}^c\).
+By this and the fact that that Theorem \ref{thm:degenFOES} and Corollary
+\ref{cor:sign}(i) entail that
+\(P_{-\boldsymbol \theta_N}(\boldsymbol X_N \in \mathcal{M}_{1-\epsilon, -\boldsymbol \theta_N})\to 1\)
+as \(N\to \infty\) (i.e., \(P_{-\boldsymbol \theta_N}\) is S-unstable),
+we have \[
+1  = \lim_{N\to \infty} P_{-\boldsymbol \theta_N}(\boldsymbol X_N \in \mathcal{M}_{1-\epsilon, -\boldsymbol \theta_N}) \leq \lim_{N\to \infty} P_{-\boldsymbol \theta_N}(\boldsymbol X_N \in \mathcal{M}_{\epsilon, \boldsymbol \theta_N}^c ) \leq 1,
+\] proving Corollary \ref{cor:sign}(ii) \hfill \(\Box\)
+
+\textbf{Proof of Proposition \ref{prp:prop1}.} For any two outcomes
+\(\boldsymbol x_1, \boldsymbol x_2\in\mathcal{X}^N\), the log-ratio of
+probabilities from the linear exponential model \eqref{eq:expo} with \(k\)
+parameters/statistics satisfies \[
+\left|\log \left[ \frac{P_{\boldsymbol \theta_N}(\boldsymbol x_1)}{P_{\boldsymbol \theta_N}(\boldsymbol x_2)}  \right] \right| =
+\left|  \sum_{i=1}^k \theta_{i,N} [g_{i,k}(\boldsymbol x_1) - g_{i,k}(\boldsymbol x_2) ] \right|  \leq  \sum_{i=1}^k | \theta_{i,N}| (U_{i,N}-L_{i,N});
+\] consequently,
+\(\REP(\boldsymbol \theta_N ) \leq \sum_{i=1}^k | \theta_{i,N}| (U_{i,N}-L_{i,N})\)
+holds in \eqref{eq:elpr} and model stability in Proposition
+\ref{prp:prop1} follows from \eqref{eq:Sun}. \hfill \(\Box\)
+
+\textbf{Proof of Proposition \ref{prp:prop2}.} Writing
+\(\boldsymbol x=(x_1,\ldots,x_N)\) and
+\(\boldsymbol h = (h_1,\ldots,h_{N_{\mathcal{H}}})\) with all components
+\(x_i,h_j\in\{\pm 1\}\), probabilities in the joint RBM model
+\eqref{eq:RBM1} can be written as
+\(\tilde{P}_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h) = c(\boldsymbol \theta_N)\exp[ f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)]\)
+in terms of the function
+\(f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)\) from
+\eqref{eq:f} and the normalizing constant
+\(c(\boldsymbol \theta_N)= \exp [-\psi(\boldsymbol \theta_N)]\) from
+\eqref{eq:RBM1}. Let \(\boldsymbol x_M, \boldsymbol x_m\in\{\pm 1\}^N\) be
+such that
+\(P_{\boldsymbol \theta_N} (\boldsymbol x_M) = \max_{\boldsymbol x}P_{\boldsymbol \theta_N} (\boldsymbol x)\)
+and
+\(P_{\boldsymbol \theta_N} (\boldsymbol x_m) = \min_{\boldsymbol x}P_{\boldsymbol \theta_N} (\boldsymbol x)\)
+under the marginal RBM model
+\(P_{\boldsymbol \theta_N} (\boldsymbol x) = c(\boldsymbol \theta_N)\sum_{\boldsymbol h \in\{\pm 1\}^{\mathcal{N}_H}} \tilde{P}_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)= c(\boldsymbol \theta_N)\sum_{\boldsymbol h \in\{\pm 1\}^{\mathcal{N}_H}} \exp[ f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)]\)
+from \eqref{eq:RBM2}. Also, \(\boldsymbol x_0,x_1\in\{\pm 1\}^N\) be such
+that
+\(\max_{\boldsymbol h}f_{\boldsymbol \theta_N} (\boldsymbol x_0, \boldsymbol h)=\max_{\boldsymbol x}\max_{\boldsymbol h}f_{\boldsymbol \theta_N} (\boldsymbol x , \boldsymbol h)\)
+and
+\(\max_{\boldsymbol h}f_{\boldsymbol \theta_N} (\boldsymbol x_1, \boldsymbol h)=\min_{\boldsymbol x}\max_{\boldsymbol h}f_{\boldsymbol \theta_N} (\boldsymbol x , \boldsymbol h)\).
+Then, Proposition \ref{prp:prop2}(i) follows from
+\(\REP_{(\boldsymbol X)}(\boldsymbol \theta_N) = \log[P_{\boldsymbol \theta_N} (\boldsymbol x_M) /P_{\boldsymbol \theta_N} (\boldsymbol x_m) ]\)
+and the lower/upper bounds on
+\(P_{\boldsymbol \theta_N} (\boldsymbol x_M)\) and
+\(P_{\boldsymbol \theta_N} (\boldsymbol x_m)\) as \[
+c(\boldsymbol \theta_N) \exp[\max_{\boldsymbol h}f_{\boldsymbol \theta_N} (\boldsymbol x_0 , \boldsymbol h)]
+\leq P_{\boldsymbol \theta_N} (\boldsymbol x_0) \leq  P_{\boldsymbol \theta_N} (\boldsymbol x_M) \leq  2^{N_{\mathcal{H}}} c(\boldsymbol \theta_N) \exp[\max_{\boldsymbol x}\max_{\boldsymbol h}f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)]
+\] and
+\begin{align*}
+c(\boldsymbol \theta_N) \exp[\min_{\boldsymbol x}\max_{\boldsymbol h}f_{\boldsymbol \theta_N} (\boldsymbol x , \boldsymbol h)] &\leq c(\boldsymbol \theta_N) \exp[ \max_{\boldsymbol h}f_{\boldsymbol \theta_N} (\boldsymbol x_m , \boldsymbol h)] \\
+&\leq P_{\boldsymbol \theta_N} (\boldsymbol x_m) \\
+&\leq P_{\boldsymbol \theta_N} (\boldsymbol x_1) \\ 
+&\leq 2^{N_{\mathcal{H}}} c(\boldsymbol \theta_N) \exp[ \max_{\boldsymbol h}f_{\boldsymbol \theta_N} (\boldsymbol x_1 , \boldsymbol h)]\\&=&2^{N_{\mathcal{H}}} c(\boldsymbol \theta_N) \exp[\min_{\boldsymbol x}\max_{\boldsymbol h}f_{\boldsymbol \theta_N} (\boldsymbol x , \boldsymbol h)]
+\end{align*}
+To prove Proposition \ref{prp:prop2}, we next expand the function
+\(f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)\) from
+\eqref{eq:f} as \[
+f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)= \sum_{j=1}^{N_\mathcal{H}} h_j \theta_{j,N}^{\mathcal{H}}  +   \sum_{i=1}^N \left( \theta_{i,N}^{\mathcal{V}} + \sum_{j=1}^{N_\mathcal{H}}  h_j  \theta_{ij,N}^{\mathcal{VH}}\right)x_i=\sum_{i=1}^N x_i  \theta_{i,N}^{\mathcal{V}} +  \sum_{j=1}^{N_\mathcal{H}} \left( \theta_{j,N}^{\mathcal{H}}  + \sum_{i=1}^N  x_i  \theta_{ij,N}^{\mathcal{VH}}\right)h_j.
+\] By this and the fact that \(x_i,h_j\in\{\pm 1\}\), we then have
+\begin{align}
+\nonumber \max_{\boldsymbol x}  f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h) =  \sum_{j=1}^{N_\mathcal{H}} h_j \theta_{j,N}^{\mathcal{H}}  +   a_{\boldsymbol \theta_N, \mathcal{H}} (\boldsymbol h), & \min_{\boldsymbol x}  f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h) =  \sum_{j=1}^{N_\mathcal{H}} h_j \theta_{j,N}^{\mathcal{H}}  - a_{\boldsymbol \theta_N, \mathcal{H}} (\boldsymbol h),\\
+\nonumber \max_{\boldsymbol h}  f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h) = \sum_{i=1}^{N}x_i \theta_{i,N}^{\mathcal{V}} + b_{\boldsymbol \theta_N, \mathcal{V}} (\boldsymbol x), & \min_{\boldsymbol h}  f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h) = \sum_{i=1}^{N}x_i \theta_{i,N}^{\mathcal{V}}  -   b_{\boldsymbol \theta_N, \mathcal{V}} (\boldsymbol x), \\
+\label{eq:max}
+a_{\boldsymbol \theta_N, \mathcal{H}} (\boldsymbol h) \equiv \sum_{i=1}^N \left| \theta_{i,N}^{\mathcal{V}} + \sum_{j=1}^{N_\mathcal{H}}  h_j  \theta_{ij,N}^{\mathcal{VH}}\right|, &  b_{\boldsymbol \theta_N, \mathcal{V}} (\boldsymbol x) \equiv \sum_{j=1}^{N_\mathcal{H}} \left| \theta_{j,N}^{\mathcal{H}}  + \sum_{i=1}^N  x_i  \theta_{ij,N}^{\mathcal{VH}}\right|,
+\end{align}
+where
+\(\boldsymbol h^T \boldsymbol \theta_N^{\mathcal{H}}=\sum_{j=1}^{N_\mathcal{H}} h_j \theta_{j,N}^{\mathcal{H}}\),
+\(\boldsymbol x^T \boldsymbol \theta_N^{\mathcal{V}}= \sum_{i=1}^{N}x_i \theta_{i,N}^{\mathcal{V}}\)
+and
+\(\Gam\equiv \max_{\boldsymbol h} a_{\boldsymbol \theta_N, \mathcal{H}} (\boldsymbol h)\).
+From this, it follows that
+\begin{align*}
+\REP_{(\boldsymbol X, \boldsymbol H)}(\boldsymbol \theta_N) &=  \max_{\boldsymbol h}\max_{\boldsymbol x}f_{\boldsymbol \theta_N} (\boldsymbol x , \boldsymbol h) -  \min_{\boldsymbol h}\min_{\boldsymbol x}f_{\boldsymbol \theta_N} (\boldsymbol x , \boldsymbol h)\\
+&= \max_{\boldsymbol h_1 }  \max_{\boldsymbol h_2 }\left[ (\boldsymbol h_1 - \boldsymbol h_2)^T \boldsymbol \theta_N^{\mathcal{H}}  +   a_{\boldsymbol \theta_N, \mathcal{H}} (\boldsymbol h_1)  + a_{\boldsymbol \theta_N, \mathcal{H}} (\boldsymbol h_2)\right],
+\end{align*}
+which leads to the upper bound
+\(\REP_{(\boldsymbol X, \boldsymbol H)}(\boldsymbol \theta_N) \leq 2 \Gam + 2 |\boldsymbol \theta_N^{\mathcal{H}} |_1\).
+Then, taking \(\boldsymbol h_1=\boldsymbol h_2\) (i.e., before
+maximization) gives
+\(\REP_{(\boldsymbol X, \boldsymbol H)}(\boldsymbol \theta_N) \geq 2\Gam\)
+and taking \(\boldsymbol h_1=-\boldsymbol h_2\), such that
+\(\boldsymbol h_1^T \boldsymbol \theta_N^{\mathcal{H}} = |\theta_N^{\mathcal{H}}|_1\),
+gives
+\(\REP_{(\boldsymbol X, \boldsymbol H)}(\boldsymbol \theta_N) \geq 2|\theta_N^{\mathcal{H}}|_1\);
+this yields the lower bound
+\(\REP_{(\boldsymbol X, \boldsymbol H)}(\boldsymbol \theta_N)\geq 2\max\{\Gam, |\boldsymbol \theta_N^{\mathcal{H}} |_1\}\).
+
+We next consider \(\elt\) and, by \eqref{eq:max}, write
+\begin{align*}
+\elt &=  \max_{\boldsymbol h}\max_{\boldsymbol x}f_{\boldsymbol \theta_N} (\boldsymbol x , \boldsymbol h) -  \max_{\boldsymbol h}\min_{\boldsymbol x}f_{\boldsymbol \theta_N} (\boldsymbol x , \boldsymbol h)\\
+&= \max_{\boldsymbol h_1} \min_{\boldsymbol h_2 }\left[ (\boldsymbol h_1 - \boldsymbol h_2)^T \boldsymbol \theta_N^{\mathcal{H}} + a_{\boldsymbol \theta_N, \mathcal{H}} (\boldsymbol h_1)  + a_{\boldsymbol \theta_N, \mathcal{H}} (\boldsymbol h_2)\right].
+\end{align*}
+Taking \(\boldsymbol h_1=\boldsymbol h_2\) and maximizing over both
+\(\boldsymbol h_1,\boldsymbol h_2\) produces the upper bound
+\(\elt \leq 2\Gam\). Then, using
+\((\boldsymbol h_1 - \boldsymbol h_2)^T \boldsymbol \theta_N^{\mathcal{H}} + a_{\boldsymbol \theta_N, \mathcal{H}} (\boldsymbol h_2) \geq - 2|\boldsymbol \theta_N^{\mathcal{H}} |_1\)
+and maximizing over \(\boldsymbol h_1\) gives
+\(\elt \geq \Gam- 2|\boldsymbol \theta_N^{\mathcal{H}} |_1\), while
+setting \(\boldsymbol h_1=\boldsymbol h_2^*\) for \(\boldsymbol h_2^*\)
+such that
+\(-(\boldsymbol h_2^*) ^T \boldsymbol \theta_N^{\mathcal{H}} + a_{\boldsymbol \theta_N, \mathcal{H}} (\boldsymbol h_2^*) = \min_{\boldsymbol h_2} [-\boldsymbol h_2^T \boldsymbol \theta_N^{\mathcal{H}} + a_{\boldsymbol \theta_N, \mathcal{H}} (\boldsymbol h_2)]\)
+gives
+\(\elt \geq 2 a_{\boldsymbol \theta_N, \mathcal{H}} (\boldsymbol h_2^*) \geq \Gamc \equiv \min_{\boldsymbol h} a_{\boldsymbol \theta_N, \mathcal{H}} (\boldsymbol h)\).
+Finally, note that for any \(\boldsymbol h\), the triangle inequality
+gives
+\begin{align*}
+\Gam  \equiv \max_{\boldsymbol h_1}  a_{\boldsymbol \theta_N, \mathcal{H}} (\boldsymbol h_1) &\geq
+[a_{\boldsymbol \theta_N, \mathcal{H}} (\boldsymbol h)  + a_{\boldsymbol \theta_N, \mathcal{H}} (-\boldsymbol h)]/2 \\
+&=
+2^{-1}\sum_{i=1}^{N } \left(\left| \theta_{i,N}^{\mathcal{V}}   + \sum_{j=1}^{N_{\mathcal{H}}} h_j \theta_{ij,N}^{\mathcal{VH}} \right|
++ \left| \theta_{i,N}^{\mathcal{V}}   - \sum_{j=1}^{N_{\mathcal{H}}} h_j \theta_{ij,N}^{\mathcal{VH}} \right|\right)\\
+& \geq \sum_{i=1}^{N }  \left| \theta_{i,N}^{\mathcal{V}} \right| \equiv |\boldsymbol \theta_{N}^{\mathcal{V}}|_1.
+\end{align*}
+\hfill \(\Box\)
 

### 2017-11-02 writing/paper-note/appendices/appendix-note.Rmd
dr. nordman's changes

@@
 
+**Proof of Theorem \@ref(thm:something).**  Let $L_{\boldsymbol \theta_{N}}(\boldsymbol X) = \log[ P_{\boldsymbol \theta_{N}}(\boldsymbol X)/ \min_{\boldsymbol y \in \mathcal{X}^N} P_{\boldsymbol \theta_{N}}(\boldsymbol y) ]/\REP(\boldsymbol \theta_{N})$, where again $\boldsymbol X=(X_1, \dots, X_N)$ and $\REP(\boldsymbol \theta_{N})= \log[\max_{\boldsymbol y \in \mathcal{X}^N} P_{\boldsymbol \theta_{N}}(\boldsymbol y)/ \min_{\boldsymbol y \in \mathcal{X}^N} P_{\boldsymbol \theta_{N}}(\boldsymbol y) ]$.
+As $L_{\boldsymbol \theta_{N}}(\boldsymbol X)\in[0,1]$, convergence of $L_{\boldsymbol \theta_{N}}(\boldsymbol X)$ to 1 in probability under $P_{\boldsymbol \theta_{N}}$ is equivalent to convergence to $1$ in expectation under $P_{\boldsymbol \theta_{N}}$ (i.e., convergence in expectation implies probabilistic convergence by Markov's inequality while probabilistic convergence implies convergence in expectation by uniform integrability/boundedness).
+
+For $\epsilon \in(0,1)$, let $\mathcal{M}_{\epsilon, \boldsymbol \theta_N}$ denote a modal set as in \@ref(eq:mode).
+By Theorem \@ref(thm:degenFOES), $P_{\boldsymbol \theta_N}\left( \boldsymbol X\in \mathcal{M}_{\epsilon, \boldsymbol \theta_N}\right) \rightarrow 1$ holds as $N \rightarrow \infty$ and, by definition of \@ref(eq:mode), $\boldsymbol X\in \mathcal{M}_{\epsilon, \boldsymbol \theta_N}$ follows if and only if
+$1-L_{\boldsymbol \theta_{N}}(\boldsymbol X)<\epsilon$.
+Hence, $L_{\boldsymbol \theta_{N}}(\boldsymbol X) \stackrel{p,\mathrm{E}}{\longrightarrow} 1$ holds under $\boldsymbol \theta_{N}$ in Theorem \@ref(thm:something).
+The convergence $L_{-\boldsymbol \theta_{N}}(\boldsymbol X) \stackrel{p,\mathrm{E}}{\longrightarrow} 1$  under $-\boldsymbol \theta_{N}$ likewise follows from Corollary \@ref(cor:sign).
+\hfill $\Box$
+
 \clearpage

### 2017-11-02 writing/paper-note/chapters/instab.Rmd
dr. nordman's changes

@@
 Interest is in identifying  instability in such models, which is roughly characterized by probabilities with extreme sensitivity to small changes in data configuration.
-The concept of instability was introduced in the field of statistical physics (i.e., point processes) by @ruelle1999statistical and then further extended by Schweinberger @schweinberger2011instability for a family of exponential models.
+The concept of instability was introduced in the field of statistical physics (i.e., point processes) by @ruelle1999statistical and then further extended by @schweinberger2011instability for a family of exponential models.
 At issue, models exhibiting instability are typically undesirable as these tend to provide poor representations of data or data-generation.
@@
 denotes the normalizing function with parameter space $\Theta_N$.
-The natural parameter function $\eta (\boldsymbol \theta)$ has a linear form (i.e., $\eta (\boldsymbol \theta)= \bm{A} \boldsymbol \theta$ for a given $L \times k$ matrix $\bm{A}$) in many common model formulaitons, though may also be nonlinear (e.g., curved exponential families).
+The natural parameter function $\eta (\boldsymbol \theta)$ has a linear form (i.e., $\eta (\boldsymbol \theta)= \bm{A} \boldsymbol \theta$ for a given $L \times k$ matrix $\bm{A}$) in many common model formulations, though may also be nonlinear (e.g., curved exponential families).
 In the linear case, $\eta (\boldsymbol \theta) = \boldsymbol \theta$ may be generally assumed in the exponential parameterization with a minor modification to the definition of sufficient statistics $\boldsymbol g_N(\boldsymbol x)$.
@@
 Consequently, $\boldsymbol \theta_N$ lies in a parameter space of fixed Euclidean dimension $k$.
-However, this aspect need not be true for other types of FOES models considered in Sections \@ref(restricted-boltzmann-machines) - \@ref(deep-learning), where instead the numbers of parameters and sufficient statistics commonly increase with the sample size $N$.
+However, this aspect need not be true for other types of FOES models considered in Sections \@ref(rbm) - \@ref(deep-learning), where instead the numbers of parameters and sufficient statistics commonly increase with the sample size $N$.
 
-## Restricted Boltzmann Machines
+## Restricted Boltzmann Machines {#rbm}
 
@@
 
-This definition of S-unstable is a generalization or reinterpretation of "unstable" used in @schweinberger2011instability by allowing possibly non-exponential family models (e.g., RBM and DBM models in Sections \@ref(restricted-boltzmann-machines)-\@ref(deep-learning) as well as a potentially increasing number $q(N)$ of parameters through the parameter sequence $\boldsymbol \theta_N\in \mathbb{R}^{q(N)}$.
+This definition of S-unstable is a generalization or reinterpretation of "unstable" used in @schweinberger2011instability by allowing possibly non-exponential family models (e.g., RBM and DBM models in Sections \@ref(rbm)-\@ref(deep-learning) as well as a potentially increasing number $q(N)$ of parameters through the parameter sequence $\boldsymbol \theta_N\in \mathbb{R}^{q(N)}$.
 While this definition differs in form and scope from the original, it does match that in @schweinberger2011instability for the special case of exponential models (cf.
@@
 In the definition \@ref(eq:Sun) of S-instability, we note that the numerical measure $\REP(\boldsymbol \theta_N)/N$ of model instability is invariant to *independent  replications* of data.
-That is, let $M \geq 1$ denote a possible number of replications and consider data $\boldsymbol Y_{N,M} \equiv (\boldsymbol X^{(1)}_N, \dots, \boldsymbol X^{(M)}_N)$ formed by $\{ \boldsymbol X^{(j)}_N\}_{j=1}^M$ as $M$  iid replications of a random vector $\boldsymbol X_N=(X_1,\ldots,X_N)$, where the latter follows a FOES model with probabilities \(P_{ \boldsymbol \theta_N}(\boldsymbol x)>0\), $\boldsymbol x\in\mathcal{X}^N$.
+That is, let $M \geq 1$ denote a possible number of replications and consider data $\boldsymbol Y_{N,M} \equiv (\boldsymbol X^{(1)}_N, \dots, \boldsymbol X^{(M)}_N)$ formed by $\{ \boldsymbol X^{(j)}_N\}_{j=1}^M$ as $M$  iid replications of a random vector $\boldsymbol X_N=(X_1,\ldots,X_N)$, where the latter follows a FOES model with probabilities $P_{ \boldsymbol \theta_N}(\boldsymbol x)>0$, $\boldsymbol x\in\mathcal{X}^N$.
 This leads to a joint model, say $P_{ \boldsymbol \theta_N}(\boldsymbol y)$, $\boldsymbol y\in\mathcal{X}^{NM}$, for $\boldsymbol Y_{N,M}$ consisting of $N*M$ random variables in total.
@@
 The above model condition incorporates many standard parameterizations and follows, for instance, whenever $P_{\boldsymbol \theta_N}(\boldsymbol x)/P_{\boldsymbol \theta_N}(\boldsymbol y)  = [P_{-\boldsymbol \theta_N}(\boldsymbol x)/P_{-\boldsymbol \theta_N}(\boldsymbol y)]^{-1}$ holds for outcomes $\boldsymbol x, \boldsymbol y \in\mathcal{X}^N$ in a FOES model.
-For instance, this latter condition is fulfilled for all linear exponential families from Section \@ref(discrete-exponential-family-models) (e.g., \@ref(eq:mod1)-\@ref(eq:mod2)) as well as all network models from Sections \@ref(restricted-boltzmann-machines)-\@ref(deep-learning) (e.g., \@ref(eq:RBM1)-\@ref(eq:RBM2)).
+For instance, this latter condition is fulfilled for all linear exponential families from Section \@ref(discrete-exponential-family-models) (e.g., \@ref(eq:mod1)-\@ref(eq:mod2)) as well as all network models from Sections \@ref(rbm)-\@ref(deep-learning) (e.g., \@ref(eq:RBM1)-\@ref(eq:RBM2)).
 When parameters can be tuned in sign with effects prescribed in the model condition PSR, unstable FOES models will exhibit further probability sensitivities, as outlined in the following extension of Theorem \@ref(thm:degenFOES}).
@@
 \item the models  $P_{-\boldsymbol \theta_N}$  defined by $-\boldsymbol \theta_N$  are also S-unstable;
-\item and for the complement $\mathcal{M}_{\epsilon, \boldsymbol \theta_N}^c \equiv \mathcal{X}^N \setminus \mathcal{M}_{\epsilon, \boldsymbol \theta_N}$ of any mode-set $\mathcal{M}_{\epsilon, \boldsymbol \theta_N}$ under $\boldsymbol \theta_N$ from (\ref{eq:mode}), with $0<\epsilon<1$, it holds under $- \boldsymbol \theta_N$ that
+\item and for the complement $\mathcal{M}_{\epsilon, \boldsymbol \theta_N}^c \equiv \mathcal{X}^N \setminus \mathcal{M}_{\epsilon, \boldsymbol \theta_N}$ of any mode-set $\mathcal{M}_{\epsilon, \boldsymbol \theta_N}$ under $\boldsymbol \theta_N$ from (\ref{eq:mode}), with $0<\epsilon<1$, it holds under $-\boldsymbol \theta_N$ that
     $$ 
-    P_{-\boldsymbol \theta_N}( \boldsymbol X_N \in \mathcal{M}_{\epsilon, \boldsymbol \theta_N}^c) \rightarrow 1\qquad \mbox{as $N\to \infty$},
+    P_{-\boldsymbol \theta_N}( \boldsymbol X_N \in \mathcal{M}_{\epsilon, \boldsymbol \theta_N}^c) \rightarrow 1\qquad \text{as } N\to \infty,
     $$
@@
 ```{remark, echo=TRUE}
-Model condition PSR and Corollary \ref{cor:sign} can be extended to cases where parameter components $\boldsymbol \theta_N =(\boldsymbol \theta_{1,N}, \boldsymbol \theta_{N,2})$ are not all changed in sign ($-\boldsymbol \theta_N$) but, more generally, are instead altered to another parameter configuration $\boldsymbol \theta_{N}^A = (\boldsymbol \theta_{N,1}^A, \boldsymbol \theta_{N,2}^A)$ involving a switch in sign only among some model parameters $\boldsymbol \theta_{N,2}^A = -\boldsymbol \theta_{N,2}$ with remaining parameters $\boldsymbol \theta_{1,N}^A$ being arbitrarily chosen.
-If a sign change occurs among parameters ($\pm \boldsymbol \theta_{N,2}$) which dominate the probability structure of the model, then an unstable model under $\boldsymbol \theta_N$ will then imply that many more unstable models exist over a broad spectrum of possibilities for variations $\boldsymbol \theta_{N}^A$ of $\boldsymbol \theta_N$.
+Under the model condition PSR, Corollary \ref{cor:sign} can also be extended to
+cases where parameter components $\boldsymbol \theta_N =(\boldsymbol \theta_{1,N}, \boldsymbol \theta_{N,2})$ (say) are not all changed in sign (e.g., $-\boldsymbol \theta_N$) but, more generally, are instead  altered to another parameter configuration $\boldsymbol \theta_{N}^A = ( \boldsymbol \theta_{1,N}^A, \boldsymbol \theta_{2,N}^A )$ involving a switch in sign only among some dominating model parameters $\boldsymbol \theta_{N,2}^A=- \boldsymbol \theta_{N,2}$ with remaining parameters $\boldsymbol \theta_{1,N}^A$ being arbitrarily chosen.
+If a change sign occurs among parameters ($\pm \boldsymbol \theta_{N,2}$) which dominate the probability structure of the model, then the results of Corollary  \ref{cor:sign} can still hold  with $\boldsymbol \theta_{N}^A$ replacing $-\boldsymbol \theta_N$; as an example of one sufficient condition, if $\lim_{N\to \infty} \max_{\boldsymbol x \in \mathcal{X}^N} |G_N(\boldsymbol x, -\boldsymbol \theta_{N}) - G_N(\boldsymbol x, \boldsymbol \theta_{N}^A)|=0$ holds in addition to Corollary~1 assumptions, where
+$$
+G_N(\boldsymbol x, \boldsymbol \theta) =   \frac{\log P_{\boldsymbol \theta}(\boldsymbol x) - \min_{\boldsymbol y \in\mathcal{X}^N}\log P_{\boldsymbol \theta}(\boldsymbol y) }{\max_{\boldsymbol y \in\mathcal{X}^N}\log P_{\boldsymbol \theta}(\boldsymbol y)  - \min_{\boldsymbol y \in\mathcal{X}^N}\log P_{\boldsymbol \theta}(\boldsymbol y) }, \quad \boldsymbol x \in\mathcal{X}^N,
+$$
+represents a standardized form of $\boldsymbol \theta$-model probabilities, then the results of Corollary \ref{cor:sign} apply to $\boldsymbol \theta_{N}^A$ in addition to $-\boldsymbol \theta_N$.
+As a consequence, an unstable model under $\boldsymbol \theta_N$ can then imply that many more unstable models exist  over a broader spectrum of possibilities for variations $\boldsymbol \theta_{N}^A$ of $\boldsymbol \theta_N$, which involves some amount of sign change among components of $\boldsymbol \theta_N$.
 ```
@@
 $$
-\frac{1}{N} \REP(\boldsymbol \theta_N)=0\quad \;(\mbox{uniform probability model}),
+\frac{1}{N} \REP(\boldsymbol \theta_N)=0\quad \;(\text{uniform probability model}),
 $$ 
@@
 (\#eq:UL)
-\frac{1}{N}\REP(\boldsymbol \theta_N ) \equiv   |\boldsymbol \theta_N| \frac{(U_N-L_N)}{N} \;\quad \mbox{(one-parameter exponential model)},
+\frac{1}{N}\REP(\boldsymbol \theta_N ) \equiv   |\boldsymbol \theta_N| \frac{(U_N-L_N)}{N} \;\quad \text{(one-parameter exponential model)},
 \end{equation}
@@
 $$
-\frac{1}{N}\REP(\boldsymbol \theta_N ) = |\boldsymbol \theta_N|\quad\; \mbox{(iid Bernoulli model)},
+\frac{1}{N}\REP(\boldsymbol \theta_N ) = |\boldsymbol \theta_N|\quad\; \text{(iid Bernoulli model)},
 $$
 so that an unstable (or stable) model results for a divergent (or bounded) parameter sequence  $|\boldsymbol \theta_N|$.
-The above instability expression for the Bernoulli model follows as well from  the $N$-scaled LREP value \@ref(eq:UL) for a one-parameter exponential distribution, using that the sufficient statistic involved $g_N(\boldsymbol x)= \sum_{i=1}^N x_i$, $\boldsymbol x =(x_1,\ldots,x_n)\in\{0,1\}^N$, has maximum and mimimum values $U_N=N$ and $L_N=0$.
+The above instability expression for the Bernoulli model follows as well from  the $N$-scaled LREP value \@ref(eq:UL) for a one-parameter exponential distribution, using that the sufficient statistic involved $g_N(\boldsymbol x)= \sum_{i=1}^N x_i$, $\boldsymbol x =(x_1,\ldots,x_n)\in\{0,1\}^N$, has maximum and minimum values $U_N=N$ and $L_N=0$.
 In this case, @schweinberger2011instability has noted that the statistic is stable (i.e., bounded $(U_N-L_N)/N=1$) and the Bernoulli model is as well when, in particular, $\boldsymbol \theta_N=\bold \theta \in\mathbb{R}$ is fixed for $N \geq 1$.
@@
 $$
-\frac{1}{N}\REP(\boldsymbol \theta_N )  = \left\{ \begin{array}{lcl} |\boldsymbol \theta_N| (n-2) && \mbox{(2-star graph model)}\\
-|\boldsymbol \theta_N|(n-2)/3 &&\mbox{(triangle graph model)},\end{array}\right.
+\frac{1}{N}\REP(\boldsymbol \theta_N )  = \left\{ \begin{array}{lcl} |\boldsymbol \theta_N| (n-2) && \text{(2-star graph model)}\\
+|\boldsymbol \theta_N|(n-2)/3 &&\text{(triangle graph model)},\end{array}\right.
 $$
@@
   \frac{1}{N}\REP(\boldsymbol \theta_N) &= \frac{\max_{1 \leq i \leq k} P_{\boldsymbol \theta_N}(X_1=i)}{\min_{1 \leq i \leq k} P_{\boldsymbol \theta_N}(X_1=i)}\\
-  &= \max_{1 \leq i \leq k} \theta_{i,N} - \min_{1 \leq i \leq k} \theta_{i,N} \qquad \mbox{(iid multinomial model)}.
+  &= \max_{1 \leq i \leq k} \theta_{i,N} - \min_{1 \leq i \leq k} \theta_{i,N} \qquad \text{(iid multinomial model)}.
 \end{align*}
@@
 
-We next consider instability of discrete data models based on exponential formulations involving hidden, or latent, variables, such as those probabilistic graphical models described in Sections \@ref(restricted-boltzmann-machines)-\@ref(deep-learning).
-We will focus on restricted Bolzmann machine (RBM) models (Section \@ref(restricted-boltzmann-machines), having one layer of latent variables for simplicity, though the same instability concepts may be extended to other deep learning models (Section \@ref(deep-learning).
+We next consider instability of discrete data models based on exponential formulations involving hidden, or latent, variables, such as those probabilistic graphical models described in Sections \@ref(rbm)-\@ref(deep-learning).
+We will focus on restricted Boltzmann machine (RBM) models (Section \@ref(rbm), having one layer of latent variables for simplicity, though the same instability concepts may be extended to other deep learning models (Section \@ref(deep-learning).
 For $N$ visible variables $\boldsymbol X \equiv \boldsymbol X_N = (X_1,\ldots,X_n)$ as data, each observation $X_i\in\{\pm 1\}$ being binary, the RBM-based model \@ref(eq:RBM2) for  $\boldsymbol X$ is again of FOES-type, though not an exponential model.
@@
 $$
-\frac{1}{N}\REP(\boldsymbol \theta_N) = \frac{2}{N} \sum_{i=1}^N|\theta_{i,N}^{\mathcal{V}}| \quad \mbox{(RBM model, no hiddens)}.
+\frac{1}{N}\REP(\boldsymbol \theta_N) = \frac{2}{N} \sum_{i=1}^N|\theta_{i,N}^{\mathcal{V}}| \quad \text{(RBM model, no hiddens)}.
 $$
@@
 \tilde{P}_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)}{\min_{ (\boldsymbol x, \boldsymbol h) \in \{\pm 1\}^{N+N_{\mathcal{H}}}`}
-\tilde{P}_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)} \right] \qquad \mbox{(joint RBM model)}    \\    
+\tilde{P}_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)} \right] \qquad \text{(joint RBM model)}    \\    
 &= \left(\max_{ (\boldsymbol x, \boldsymbol h) \in \{\pm 1\}^{N+N_{\mathcal{H}}}} f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)-\min_{ (\boldsymbol x, \boldsymbol h) \in \{\pm 1\}^{N+N_{\mathcal{H}}}} f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)\right),
@@
 
-Due to their extreme sensitivity in probability structure, S-instability in FOES models often translates to numerical complications, and in fact obstructions, in both simulation and statistical inference based on likelihoods.
+Due to their extreme sensitivity in probability structure, S-instability in FOES models may often translate to numerical complications, and in fact obstructions, in both simulation and statistical inference based on likelihoods.
+We describe these aspects in Sections \@ref(mcmc)-\@ref(bayes) with regard to data simulation, maximum likelihood estimation and Bayes inference, respectively.
 
@@
 Suppose one aims to apply MCMC to simulate data $\boldsymbol X=(X_1,\ldots,X_n)$ from a FOES model $P_{\boldsymbol \theta_N}(\boldsymbol x)$, $\boldsymbol x \in\mathcal{X}^N$ whereby a chain is constructed with $P_{\boldsymbol \theta_N}$ as the stationary distribution.
-For an unstable FOES model, one component changes in outcomes may produce radically different probabilities which then creates numerical barriers to MCMC.
+For an unstable FOES model, one component changes in outcomes may produce
+radically different probabilities, which may then impose numerical barriers to MCMC.
 For example, consider implementation of a Gibbs sampler from the full conditional distributions $P_{\boldsymbol \theta_N} (X_i = x| \boldsymbol x_{-i})$, $x\in\mathcal{X}$, of each variable $X_i$ based on values $\boldsymbol x_{-i}\in \mathcal{X}^{N-1}$ for the remaining variables, say $\boldsymbol X_{-i}$, in $\boldsymbol X$.
-If a single change in $X_i$ from one value $x_1$ to another $x_2$ may produce two outcomes $\boldsymbol x^{(1)}$ and $\boldsymbol x^{(2)}$ for $\boldsymbol X$ with vastly different probabilities under the joint distribution $P_{\boldsymbol \theta_N}$, then the Gibbs sampler can have  extreme log-ratios in its transition probabilities
+If a single change in $X_i$ from one value $x_1$ to another $x_2$ may produce two outcomes $\boldsymbol x^{(1)}$ and $\boldsymbol x^{(2)}$ for $\boldsymbol X$ with vastly different probabilities under the joint distribution $P_{\boldsymbol \theta_N}$, then the Gibbs sampler can have extreme log-ratios in its transition probabilities,
 $$
-\left| \frac{P_{\boldsymbol \theta_N} (X_i =x_1 | \boldsymbol x^{(1)}_{-i})}{P_{\boldsymbol \theta_N} (X_i = x_2| \boldsymbol x^{(2)}_{-i}) } \right|=\left| \frac{P_{\boldsymbol \theta_N} ( \boldsymbol x^{(1)} )}{P_{\boldsymbol \theta_N} (  \boldsymbol x^{(2)} ) } \right|
+\log \left| \frac{P_{\boldsymbol \theta_N} (X_i =x_1 | \boldsymbol x^{(1)}_{-i})}{P_{\boldsymbol \theta_N} (X_i = x_2| \boldsymbol x^{(2)}_{-i}) } \right|=\log \left| \frac{P_{\boldsymbol \theta_N} ( \boldsymbol x^{(1)} )}{P_{\boldsymbol \theta_N} (  \boldsymbol x^{(2)} ) } \right|,
 $$  
-using that conditional probabilities are proportional to the joint probabilities and the fact that unstable models can have  unbounded log-probability ratios in one-component changes (Theorem \@ref(thm:instab-elpr)).
-This can hinder the ability of chain to effectively explore the sample space of the observations $\boldsymbol X$, as the chain may then mix poorly by moving rapidly to, and slowly away from, sections of the sample space surface.
-This mixing problem is due to the unstable stationary distribution (unbounded ratios of probabilities under the joint model), rather than in the particulars of the MCMC algorithm, similar complications can also arise for Metropolis-Hastings algorithms for MCMC.
-Hence, while an unstable FOES model $P_{\boldsymbol \theta_N}$ may be valid and technically open to simulation by MCMC, the aspect of instability can ruin numerical applications of MCMC.
+as conditional probabilities are proportional to joint probabilities that, with unstable models, can have unbounded log-probability ratios in one-component changes (Theorem
+\@ref(thm:instab-elpr)).
+This can hinder the ability of chain to effectively explore the sample space of the observations $\boldsymbol X$, as the chain may mix poorly by moving rapidly to, and slowly away from, sections of the sample space.
+In this case, for example, the Markov chain may become entrapped within a mode of the probability function, with rare chance  of escaping to adequately mimic the occupation frequencies in the overall sample space.
+If modes of the unstable model are not unique, then important outcomes may be missed without multiple chains or impractically enormous numbers of MCMC samples.
+This mixing problem is due to the unstable stationary distribution (unbounded ratios of probabilities under the joint model), rather than in any particulars of the MCMC algorithm, and similar complications can also arise for Metropolis-Hastings algorithms for MCMC.
+Hence, while an unstable FOES model $P_{\boldsymbol \theta_N}$ may be valid and technically open to simulation by MCMC, the aspect of instability can render applications of MCMC as numerically infeasible for simulation purposes.
+This result is in line with conclusions of @handcock2003assessing and @schweinberger2011instability for other exponential models.
 
@@
 
-The rugged or volatile probability structure in unstable models can also hamper efforts to maximize likelihood functions for statistical inference.
-Essentially, when a FOES model is unstable along a parameter sequence $\boldsymbol \theta_N$, the same model can consequently unstable along parameters $-\boldsymbol \theta_N$ opposite from the origin (model condition PSR and Corollary \@ref(cor:sign)).
-This leads to potential hyper-sensitivity in the likelihood function, which can be pulled in extreme and opposite directions depending on how a parameter lies with respect to zero and translates into numerical complications in maximizing the objective function.
+Volatility in the probability structure of an unstable model can also hamper efforts to maximize likelihood functions in statistical inference.
+When a FOES model is unstable along a parameter sequence \(\boldsymbol \theta_N\), the same model can further be unstable along parameters \(-\boldsymbol \theta_N\) in an opposite direction from the
+origin (model condition PSR and Corollary \@ref(cor:sign)). This can translate into
+potential sensitivity of the likelihood function around zero, and lead to numerical complications in maximizing the objective function.
+We next provide a discussion of this issue in a way that builds upon and extends related findings by @@schweinberger2011instability, who largely focused on the case of one-parameter exponential models.
 
-For a FOES model $P_{\boldsymbol \theta_N}(\boldsymbol x)$, $\boldsymbol x\in\mathcal{X}^N$, for $N$ observations $\boldsymbol X =(X_1,\ldots,X_N)$, define the likelihood function $\mathcal{L}(\boldsymbol \theta_N | \boldsymbol x) \equiv N^{-1}\log P_{\boldsymbol \theta_N }(\boldsymbol x)$ at $\boldsymbol \theta_N$  and the minimum probability ratio counterpart  $\mathcal{L}(\boldsymbol \theta_N | \boldsymbol x)  - N^{-1} \log \min_{\boldsymbol \theta_N}= N^{-1}\log[P_{\boldsymbol \theta_N }(\boldsymbol x) /\min_{\boldsymbol \theta_N}]$ where $\min_{\boldsymbol \theta_N} \equiv \min_{\boldsymbol y\in\mathcal{X}^N} P_{\boldsymbol \theta_N }(\boldsymbol y)$.
-Also define $\max_{\boldsymbol \theta_N} \equiv \max_{\boldsymbol y\in\mathcal{X}^N} P_{\boldsymbol \theta_N }(\boldsymbol y)$.
+With many probability models, the modes and anti-modes in the probability structure under one parameter $\boldsymbol \theta_N$ are reversed in role when the parameter sign changes \(-\boldsymbol \theta_N\).
+Because unstable models tend to degeneracy, the opposite signed parameters further push unstable models to assign nearly all probability to extremely opposite data configurations, given by modes/anti-modes.
+This is made concrete in Theorem \@ref(thm:something), relating the degeneracy from unstable models to the expected behavior of log-likelihood functions.
 
 ```{theorem, label="something", echo=TRUE}
-Let $P_{\boldsymbol \theta_N}$, $N \geq 1$, denote an S-unstable FOES model sequence (i.e., $\lim_{N\to \infty}\REP(\boldsymbol \theta_N)/N=\infty$) which additionally satisfies model condition PSR.
-Then,
-\begin{align*}
-\log\left[ \frac{P_{\boldsymbol \theta_N }(\boldsymbol x) }{\min_{\boldsymbol \theta_N}}\right] +  \log\left[ \frac{P_{-\boldsymbol \theta_N }(\boldsymbol x) }{\min_{-\boldsymbol \theta_N}}\right]  &=    \log\left[ \frac{\max_{\boldsymbol \theta_N}}{\min_{\boldsymbol \theta_N}}\right]  \equiv  \REP(\boldsymbol \theta_N)\\
-& = \log\left[ \frac{\max_{-\boldsymbol \theta_N} }{\min_{-\boldsymbol \theta_N}}\right] \equiv \REP(-\boldsymbol \theta_N)
-\end{align*}
-for any $\boldsymbol x\in\mathcal{X}^N$ and, as $N\to \infty$, the following convergence
+Let $P_{\boldsymbol \theta_N}$, $N \geq 1$, denote an S-unstable FOES model sequence, which additionally satisfies model condition PSR.
+Let $\boldsymbol{x}_{\max, \boldsymbol \theta_N},\boldsymbol{x}_{\min, \boldsymbol \theta_N}\in\mathcal{X}^N$ denote, respectively, a mode and anti-mode of the model $P_{\boldsymbol \theta_N}(\boldsymbol x)$, $\boldsymbol x\in\mathcal{X}^N$, for $N$ observations $\boldsymbol X = (X_1,\ldots,X_N)$, whereby $P_{\boldsymbol \theta_N }(\boldsymbol{x}_{\max, \boldsymbol \theta_N}) = \max_{\boldsymbol y\in\mathcal{X}^N} P_{\boldsymbol \theta_N }(\boldsymbol y)$ and $P_{\boldsymbol \theta_N }(\boldsymbol{x}_{\min, \boldsymbol \theta_N}) = \min_{\boldsymbol y\in\mathcal{X}^N} P_{\boldsymbol \theta_N }(\boldsymbol y)$.
+
+Then, letting $\stackrel{p, \mathrm{E}}{\longrightarrow}$ denote convergence in probability  and expectation, as $N\to \infty$,
 $$
-\frac{1}{\REP(\boldsymbol \theta_N)} \log\left[ \frac{P_{\boldsymbol \theta_N }(\boldsymbol X) }{\min_{\boldsymbol \theta_N}}\right]  \rightarrow 1 \qquad \frac{1}{\REP(-\boldsymbol \theta_N)}\log\left[ \frac{P_{-\boldsymbol \theta_N }(\boldsymbol X) }{\min_{-\boldsymbol \theta_N}}\right]  \rightarrow 0
+\frac{1}{\REP(\boldsymbol \theta_N)} \log\left[ \frac{P_{\boldsymbol \theta_N }(\boldsymbol X) }{ \displaystyle{\min_{\boldsymbol y\in\mathcal{X}^N}} P_{\boldsymbol \theta_N }(\boldsymbol y)}\right] = \frac{ \log P_{\boldsymbol \theta_N } ( \boldsymbol X)  - \log P_{\boldsymbol \theta_N }(\boldsymbol{x}_{\min, \boldsymbol \theta_N}) }
+{ \log P_{\boldsymbol \theta_N } (\boldsymbol{x}_{\max, \boldsymbol \theta_N})  - \log P_{\boldsymbol \theta_N }(\boldsymbol{x}_{\min, \boldsymbol \theta_N}) }
+\stackrel{p, \mathrm{E}}{\longrightarrow}
+\; 1
 $$
-holds in probability $P_{\boldsymbol \theta_N}$, as well as in expectation $E_{\boldsymbol \theta_N}$, under $\boldsymbol \theta_N$; the limits of convergence are reversed under $-\boldsymbol \theta_N$,
+under $\boldsymbol\theta_N$ while
 $$
-\frac{1}{\REP(\boldsymbol \theta_N)} \log\left[ \frac{P_{\boldsymbol \theta_N }(\boldsymbol x) }{\min_{\boldsymbol \theta_N}}\right]  \rightarrow 0 \qquad \frac{1}{\REP(-\boldsymbol \theta_N)}\log\left[ \frac{P_{-\boldsymbol \theta_N }(\boldsymbol x) }{\min_{-\boldsymbol \theta_N}}\right]  \rightarrow 1
+\frac{1}{\REP(-\boldsymbol \theta_N)} \log\left[ \frac{P_{-\boldsymbol \theta_N }(\boldsymbol X) }{  \displaystyle{\min_{\boldsymbol y\in\mathcal{X}^N}} P_{-\boldsymbol \theta_N }(\boldsymbol y)} \right]
+=    \frac{ \log P_{-\boldsymbol \theta_N } ( \boldsymbol X) - \log P_{ \boldsymbol \theta_N }(\boldsymbol{x}_{\max, \boldsymbol \theta_N})    }
+{ \log P_{-\boldsymbol \theta_N } (\boldsymbol{x}_{\min, \boldsymbol \theta_N})  - \log P_{-\boldsymbol \theta_N }(\boldsymbol{x}_{\max, \boldsymbol \theta_N}) }
+\stackrel{p,\mathrm{E}}{\longrightarrow}
+\; 1,
+$$
+under $-\boldsymbol\theta_N$, where
+$$
+\REP( \boldsymbol \theta_N) \equiv \log \frac{P_{\boldsymbol \theta_N } (\boldsymbol{x}_{\max, \boldsymbol \theta_N})}{P_{\boldsymbol \theta_N }(\boldsymbol{x}_{\min, \boldsymbol \theta_N})} = \log  \frac{P_{-\boldsymbol \theta_N } (\boldsymbol{x}_{\min, \boldsymbol \theta_N}) }{ P_{-\boldsymbol \theta_N }(\boldsymbol{x}_{\max, \boldsymbol \theta_N})}=   \REP( -\boldsymbol \theta_N), \quad N \geq 1.
 $$
 ```
-@schweinberger2011instability [Corollary 1] considered the one-parameter linear exponential case showing that, for positive $\boldsymbol \theta_N>0$, 
+
+Theorem \@ref(thm:something) entails log-likelihood functions based on unstable models are both inversely related and degenerate at opposited signed parameters $\boldsymbol \theta_N$ or $-\boldsymbol \theta_N$, so that likelihoods are highest at different extremes in data configuration (e.g., $\boldsymbol{x}_{\max, \boldsymbol \theta_N}$ under $\boldsymbol \theta$-probabilities or $\boldsymbol{x}_{\min, \boldsymbol \theta_N}$ under $-\boldsymbol \theta$-probabilities).
+If the observed outcome $\boldsymbol x$ for data $\boldsymbol X$ is not a mode/anti-mode, then probabilities for the outcome may be small under both parameters $\boldsymbol \theta_N$ and $-\boldsymbol \theta_N$, in which case associated optimization steps may then shift around zero and struggle to converge.
+In many model formulations, the zero parameter $\boldsymbol \theta_N=\boldsymbol 0$ is a "safe" position among parameters, representing a  guaranteed stable model (having uniform probabilities among outcomes), which can also  tether a broad parameter search attempted among unstable models.
+@handcock2003assessing describes similar results for degenerate exponential models, and Theorem \@ref(thm:something) also supports an important finding of @schweinberger2011instability [Corollary 1] for one-parameter linear exponential models \@ref(eq:expo).
+In the latter case, the likelihood score function at $\boldsymbol \theta_N$ is the expected value $\mu(\boldsymbol \theta_N)\equiv \mathrm{E}_{\boldsymbol \theta_N} g(\boldsymbol X)$  of the sufficient statistic $g(\cdot)$, and optimization involves solving $\mu(\cdot)=g(\boldsymbol x)$ for an observed outcome $\boldsymbol x$.
+For unstable  models in this exponential class, @schweinberger2011instability [Corollary 1] shows
+that
 $$
-\frac{1}{\REP(\boldsymbol \theta_N)} \log\left[ \frac{P_{\boldsymbol \theta_N }(\boldsymbol x) }{\min_{\boldsymbol \theta_N}}\right] = \frac{\boldsymbol g(\boldsymbol x) - L_N}{U_N-L_N} 
+\lim_{N\to \infty}\frac{\mu(\boldsymbol \theta_N) -L_n}{U_n-L_N}= \left\{ \begin{array}{lcl}
+1 && \text{for } \boldsymbol \theta_N>0,\\
+0 && \text{for } \boldsymbol \theta_N<0, \end{array}\right.
 $$
-holds and converges to $1$ in expectation $E_{\boldsymbol \theta_N}$, while
+where again $U_N$ and $L_N$ denote the maximum and minimum values of the statistic $g(\boldsymbol x)$, $\boldsymbol x\in\mathcal{X}^N$.
+As described by @schweinberger2011instability, the implication for maximum likelihood estimation is that, unless an observed outcome $\boldsymbol x$ falls at an extreme $U_N, L_N$ (i.e., modes/anti-modes), optimization steps in the parameter space can iterate in relatively small increments around zero and fail to converge.
+For unstable one-parameter exponential models, the maximum likelihood results of @schweinberger2011instability turn out to be a special case of Theorem \@ref(thm:something) and the LREP expansion \@ref(eq:UL) in this setting; namely, for an unstable model with $\boldsymbol \theta_N>0$,
 $$
-\frac{1}{\REP(-\boldsymbol \theta_N)} \log\left[ \frac{P_{-\boldsymbol \theta_N }(\boldsymbol x) }{\min_{-\boldsymbol \theta_N}}\right] = \frac{U_N -\boldsymbol g(\boldsymbol x)}{U_N-L_N} 
+\frac{1}{\REP(\boldsymbol \theta_N)} \log\left[ \frac{P_{\boldsymbol \theta_N }(\boldsymbol X) }{\displaystyle{\min_{\boldsymbol y\in\mathcal{X}^N}} P_{\boldsymbol \theta_N }(\boldsymbol y)}\right] = \frac{\boldsymbol g(\boldsymbol X) - L_N}{U_N-L_N}\;\;\stackrel{p, \mathrm{E}}{\longrightarrow}\;\; 1
 $$ 
-holds converges to $1$ in expectation $E_{-\boldsymbol \theta_N}$.
-This result is a special case of Theorem \@ref(thm:something) and the LREP expansion \@ref(eq:UL) for the one parameter exponential model case and is related to a result in @handcock2003assessing.
+holds as $N\to \infty$ by Theorem \@ref(thm:something), while under $-\boldsymbol \theta_N<0$
+$$
+\frac{1}{\REP(-\boldsymbol \theta_N)} \log\left[ \frac{P_{-\boldsymbol \theta_N }(\boldsymbol X) }{\displaystyle{\min_{\boldsymbol y\in\mathcal{X}^N}} P_{-\boldsymbol \theta_N }(\boldsymbol y)}\right] = \frac{U_N -\boldsymbol g(\boldsymbol X)}{U_N-L_N} = 1 - \frac{\boldsymbol g(\boldsymbol X) - L_N}{U_N-L_N} \;\;\stackrel{p, \mathrm{E}}{\longrightarrow}\;\; 1.
+$$
+Again, when all probability in  unstable models  may be pushed to opposite extremes in the sample space, due to a combination of degeneracy and parameter sign, numerical complications in likelihood maximization may occur.
 
-The idea, more generally, is that probability in an unstable model is pushed to opposite regions of the sample space due to degeneracy and opposite signed parameters.
-This can cause numerical complications in likelihood maximization.
-If an observed outcome $\boldsymbol x \equiv \argmax \mathcal{L}(\boldsymbol \theta_N)$ which is also the outcome with minimal probability under $P_{-\boldsymbol \theta_N}$, then the likelihood function will have a clear peak and maximization straightforward.
-However, when the observed outcome $\boldsymbol x$ is at an extreme in the probability structures of $P_{\boldsymbol \theta_N}, P_{-\boldsymbol \theta_N}$ (which are related) the likelihood function will be flat and optimization steps may shift around zero.
+## Implications for Bayes Inference {#bayes}
 
-## Implications for Bayes Inference
-
-Considering that changes in sign of parameters can cause all probability to be placed on disjoint sections of the sample space (Corollary \@ref(cor:sign)), then this will create numerical difficulties in Bayes inference based on MCMC.
-Given observations $\boldsymbol X = (X_1,\ldots,X_N)$, switching $\boldsymbol theta_N$ from positive to negative or vice versa could dramatically change probabilities (model condition PSR and Corollary \@ref(cor:sign)).
-This would translate to numerical challenges similar to those presented in Section \@ref(mcmc), i.e., inability of the chain to effectively explore the sample space because of unbounded ratios of probabilities.
-We can write the MH acceptance probability to indicate this,
+The potential numerical difficulties with maximum likelihood with unstable models, as described in the previous section, can naturally carry over to Bayes inference.
+Considering that the degeneracy issues related to unstable models can cause likelihoods can be flat (e.g., near zero) for many parameters under a given data outcome and that sign changes in parameters can shift tremendous probability to extreme and opposite outcomes in the sample space (e.g., Corollary \@ref(cor:sign), Theorem \@ref(thm:something)), then numerical complications may arise with Bayes inference in sampling a posterior parameter space based on MCMC.
+The potential challenges in chain mixing are similar to those presented in Section \@ref(mcmc), though in chain movements through the parameter space as opposed to the sample space in data generation.
+That is, in the Bayes setting for sampling a posterior distribution for $\boldsymbol \theta_N$,  a chain may  unstable to effectively explore the parameter space due partly to extreme and potentially unbounded probability ratios from parameter sign changes, which represents a parameter space analog to how one component changes in the sample space may impact data simulation with unstable models.
+For example, if $\pi(\cdot)$ denotes a prior density for $\boldsymbol \theta_N$ and $q(\cdot | \cdot)$  denotes a proposal distribution  for use in a Metropolis-Hastings (MH) sampler, then MH acceptance probability becomes
 $$
-\alpha\left(\boldsymbol x^{(1)} \mid \boldsymbol x^{(2)}\right)= \min\left\{1, 
-\frac{q(\boldsymbol x^{(2)} \mid \boldsymbol x^{(1)})}{q(\boldsymbol x^{(1)} \mid \boldsymbol x^{(2)})}\frac{P_{\boldsymbol \theta_N} ( \boldsymbol x^{(1)} )}{P_{\boldsymbol \theta_N} (  \boldsymbol x^{(2)} ) } \right\}
+\alpha\left(\boldsymbol \theta^{(1)} \mid \boldsymbol \theta^{(2)}\right)= \min\left\{1,
+\frac{q(\boldsymbol \theta_N^{(2)} \mid \boldsymbol \theta_N^{(1)})}{q(\boldsymbol \theta_N^{(1)} \mid \boldsymbol \theta_N^{(2)})}
+\frac{P_{\boldsymbol \theta_N^{(1)}} ( \boldsymbol x ) \pi(\boldsymbol \theta_N^{(1)}) }{P_{\boldsymbol \theta^{(2)}_N} (  \boldsymbol x) \pi(\boldsymbol \theta_N^{(2)}) } \right\},
 $$
-for some proposal distribution $q(\cdot)$.
+which indicates how parameter sensitivity in the likelihood $P_{\boldsymbol \theta_N} (  \boldsymbol x)$ may complicate sampling of the posterior $P_{\boldsymbol \theta_N} (  \boldsymbol x) \pi(\boldsymbol \theta_N)$ (i.e., moving from $\boldsymbol \theta^{(1)}$ to $\boldsymbol \theta^{(2)}$ in the parameter space).
+Furthermore, the potential for model instability and the size of the parameter space can also become greater with the introduction of latent variables to existing data variables, as involved in some model formulations described in Sections \@ref(rbm)-\@ref(deep-learning).
+As latent variables are often sampled with parameters in a Bayes MCMC approach, this aspect may further compound numerical problems in chain mixing.
 
@@
 We have shown for FOES models that instability manifests through small changes in data leading to potentially large changes in probability as well as the potential to place all probability on certain modal subsections of the sample space, which potentially could be small.
-The FOES model class is quite broad and, particularly in developing FOES models for large data sets, some caution should be used in parameter specification to control effects of model instability due to the consequences of instability in both simulation and likelihood inference.
+Such instability is often due to a complex interaction between the model statistics used  (i.e., how numerous and large these may become) and the number and magnitudes of
+parameters in the model formulation.
+For many FOES models, the possibility exists, at least in principle, to constraint parameters in a way balances their potential contributions against those of model statistics in order to prevent probability instabilities.
+The FOES model class is quite broad and, in developing such models for large data sets, some caution should be used in parameter specification to control effects of model instability.

### 2017-11-02 writing/paper-note/docs/note_blinded.tex
dr. nordman's changes

--- 2017-11-02 writing/paper-note/docs/note.tex
@@
             pdfauthor={},
-            pdftitle={A Note on the Instability and Degeneracy of Deep Learning Models},
+            pdftitle={A note on the instability and degeneracy of deep learning models},
             colorlinks=true,
@@
 
-\title{\bf A Note on the Instability and Degeneracy of Deep Learning Models}
-\author{Andee Kaplan \\ Department of Statistics, Iowa State University \\  and \\ Daniel Nordman \\ Department of Statistics, Iowa State University \\  and \\ Stephen Vardeman \\ Department of Statistics and Department of Industrial and\\
-Manufacturing Systems Engineering, Iowa State University \\ }
+\bigskip
+\bigskip
+\bigskip
+%\begin{center}
+  %{\LARGE\bf Title}
+  \title{\bf A note on the instability and degeneracy of deep learning models}
+%\end{center}
+\medskip
 
@@
 (\protect\hyperlink{ref-ruelle1999statistical}{1999}) and then further
-extended by Schweinberger Schweinberger
+extended by Schweinberger
 (\protect\hyperlink{ref-schweinberger2011instability}{2011}) for a
@@
 \(\eta (\boldsymbol \theta)= \bm{A} \boldsymbol \theta\) for a given
-\(L \times k\) matrix \(\bm{A}\)) in many common model formulaitons,
+\(L \times k\) matrix \(\bm{A}\)) in many common model formulations,
 though may also be nonlinear (e.g., curved exponential families). In the
@@
 fixed Euclidean dimension \(k\). However, this aspect need not be true
-for other types of FOES models considered in Sections
-\ref{restricted-boltzmann-machines} - \ref{deep-learning}, where instead
-the numbers of parameters and sufficient statistics commonly increase
-with the sample size \(N\).
-
-\subsection{Restricted Boltzmann
-Machines}\label{restricted-boltzmann-machines}
+for other types of FOES models considered in Sections \ref{rbm} -
+\ref{deep-learning}, where instead the numbers of parameters and
+sufficient statistics commonly increase with the sample size \(N\).
+
+\subsection{Restricted Boltzmann Machines}\label{rbm}
 
@@
 possibly non-exponential family models (e.g., RBM and DBM models in
-Sections \ref{restricted-boltzmann-machines}-\ref{deep-learning} as well
-as a potentially increasing number \(q(N)\) of parameters through the
-parameter sequence \(\boldsymbol \theta_N\in \mathbb{R}^{q(N)}\). While
-this definition differs in form and scope from the original, it does
-match that in Schweinberger
+Sections \ref{rbm}-\ref{deep-learning} as well as a potentially
+increasing number \(q(N)\) of parameters through the parameter sequence
+\(\boldsymbol \theta_N\in \mathbb{R}^{q(N)}\). While this definition
+differs in form and scope from the original, it does match that in
+Schweinberger
 (\protect\hyperlink{ref-schweinberger2011instability}{2011}) for the
@@
 \BeginKnitrBlock{remark}
-\iffalse <span class="remark"><em>Remark. \fi{} In the definition
-\eqref{eq:Sun} of S-instability, we note that the numerical measure
+\iffalse{} {Remark. } \fi{}In the definition \eqref{eq:Sun} of
+S-instability, we note that the numerical measure
 \(\REP(\boldsymbol \theta_N)/N\) of model instability is invariant to
@@
 \eqref{eq:mod1}-\eqref{eq:mod2}) as well as all network models from Sections
-\ref{restricted-boltzmann-machines}-\ref{deep-learning} (e.g.,
-\eqref{eq:RBM1}-\eqref{eq:RBM2}). When parameters can be tuned in sign with
-effects prescribed in the model condition PSR, unstable FOES models will
-exhibit further probability sensitivities, as outlined in the following
-extension of Theorem @ref(thm:degenFOES\}).
+\ref{rbm}-\ref{deep-learning} (e.g., \eqref{eq:RBM1}-\eqref{eq:RBM2}). When
+parameters can be tuned in sign with effects prescribed in the model
+condition PSR, unstable FOES models will exhibit further probability
+sensitivities, as outlined in the following extension of Theorem
+@ref(thm:degenFOES\}).
 
 \BeginKnitrBlock{corollary}
-\protect\hypertarget{cor:sign}{}{\label{cor:sign}}Let
+\protect\hypertarget{cor:sign}{}{\label{cor:sign} }Let
 \(P_{\boldsymbol \theta_N}\), with support \(\mathcal{X}^N\),
@@
 \BeginKnitrBlock{remark}
-\iffalse <span class="remark"><em>Remark. \fi{} Model condition PSR and
-Corollary \ref{cor:sign} can be extended to cases where parameter
-components
+\iffalse{} {Remark. } \fi{}Under the model condition PSR, Corollary
+\ref{cor:sign} can also be extended to cases where parameter components
 \(\boldsymbol \theta_N =(\boldsymbol \theta_{1,N}, \boldsymbol \theta_{N,2})\)
-are not all changed in sign (\(-\boldsymbol \theta_N\)) but, more
-generally, are instead altered to another parameter configuration
-\(\boldsymbol \theta_{N}^A = (\boldsymbol \theta_{N,1}^A, \boldsymbol \theta_{N,2}^A)\)
-involving a switch in sign only among some model parameters
-\(\boldsymbol \theta_{N,2}^A = -\boldsymbol \theta_{N,2}\) with
-remaining parameters \(\boldsymbol \theta_{1,N}^A\) being arbitrarily
-chosen. If a sign change occurs among parameters
-(\(\pm \boldsymbol \theta_{N,2}\)) which dominate the probability
-structure of the model, then an unstable model under
-\(\boldsymbol \theta_N\) will then imply that many more unstable models
-exist over a broad spectrum of possibilities for variations
-\(\boldsymbol \theta_{N}^A\) of \(\boldsymbol \theta_N\).
+(say) are not all changed in sign (e.g., \(-\boldsymbol \theta_N\)) but,
+more generally, are instead altered to another parameter configuration
+\(\boldsymbol \theta_{N}^A = ( \boldsymbol \theta_{1,N}^A, \boldsymbol \theta_{2,N}^A )\)
+involving a switch in sign only among some dominating model parameters
+\(\boldsymbol \theta_{N,2}^A=- \boldsymbol \theta_{N,2}\) with remaining
+parameters \(\boldsymbol \theta_{1,N}^A\) being arbitrarily chosen. If a
+change sign occurs among parameters (\(\pm \boldsymbol \theta_{N,2}\))
+which dominate the probability structure of the model, then the results
+of Corollary \ref{cor:sign} can still hold with
+\(\boldsymbol \theta_{N}^A\) replacing \(-\boldsymbol \theta_N\); as an
+example of one sufficient condition, if
+\(\lim_{N\to \infty} \max_{\boldsymbol x \in \mathcal{X}^N} |G_N(\boldsymbol x, -\boldsymbol \theta_{N}) - G_N(\boldsymbol x, \boldsymbol \theta_{N}^A)|=0\)
+holds in addition to Corollary\textasciitilde{}1 assumptions, where \[
+G_N(\boldsymbol x, \boldsymbol \theta) =   \frac{\log P_{\boldsymbol \theta}(\boldsymbol x) - \min_{\boldsymbol y \in\mathcal{X}^N}\log P_{\boldsymbol \theta}(\boldsymbol y) }{\max_{\boldsymbol y \in\mathcal{X}^N}\log P_{\boldsymbol \theta}(\boldsymbol y)  - \min_{\boldsymbol y \in\mathcal{X}^N}\log P_{\boldsymbol \theta}(\boldsymbol y) }, \quad \boldsymbol x \in\mathcal{X}^N,
+\] represents a standardized form of \(\boldsymbol \theta\)-model
+probabilities, then the results of Corollary \ref{cor:sign} apply to
+\(\boldsymbol \theta_{N}^A\) in addition to \(-\boldsymbol \theta_N\).
+As a consequence, an unstable model under \(\boldsymbol \theta_N\) can
+then imply that many more unstable models exist over a broader spectrum
+of possibilities for variations \(\boldsymbol \theta_{N}^A\) of
+\(\boldsymbol \theta_N\), which involves some amount of sign change
+among components of \(\boldsymbol \theta_N\).
 \EndKnitrBlock{remark}
@@
 extreme probabilities \eqref{eq:elpr} is \[
-\frac{1}{N} \REP(\boldsymbol \theta_N)=0\quad \;(\mbox{uniform probability model}),
+\frac{1}{N} \REP(\boldsymbol \theta_N)=0\quad \;(\text{uniform probability model}),
 \] which is as small as possible. In fact, a LREP value of zero can only
@@
 \label{eq:UL}
-\frac{1}{N}\REP(\boldsymbol \theta_N ) \equiv   |\boldsymbol \theta_N| \frac{(U_N-L_N)}{N} \;\quad \mbox{(one-parameter exponential model)},
+\frac{1}{N}\REP(\boldsymbol \theta_N ) \equiv   |\boldsymbol \theta_N| \frac{(U_N-L_N)}{N} \;\quad \text{(one-parameter exponential model)},
 \end{equation}
@@
 measure \eqref{eq:Sun} directly as \[
-\frac{1}{N}\REP(\boldsymbol \theta_N ) = |\boldsymbol \theta_N|\quad\; \mbox{(iid Bernoulli model)},
+\frac{1}{N}\REP(\boldsymbol \theta_N ) = |\boldsymbol \theta_N|\quad\; \text{(iid Bernoulli model)},
 \] so that an unstable (or stable) model results for a divergent (or
@@
 \(g_N(\boldsymbol x)= \sum_{i=1}^N x_i\),
-\(\boldsymbol x =(x_1,\ldots,x_n)\in\{0,1\}^N\), has maximum and mimimum
+\(\boldsymbol x =(x_1,\ldots,x_n)\in\{0,1\}^N\), has maximum and minimum
 values \(U_N=N\) and \(L_N=0\). In this case, Schweinberger
@@
 \eqref{eq:Sun} as \[
-\frac{1}{N}\REP(\boldsymbol \theta_N )  = \left\{ \begin{array}{lcl} |\boldsymbol \theta_N| (n-2) && \mbox{(2-star graph model)}\\
-|\boldsymbol \theta_N|(n-2)/3 &&\mbox{(triangle graph model)},\end{array}\right.
+\frac{1}{N}\REP(\boldsymbol \theta_N )  = \left\{ \begin{array}{lcl} |\boldsymbol \theta_N| (n-2) && \text{(2-star graph model)}\\
+|\boldsymbol \theta_N|(n-2)/3 &&\text{(triangle graph model)},\end{array}\right.
 \] by using the (one-parameter exponential) LREP formula \eqref{eq:UL}
@@
 those probabilistic graphical models described in Sections
-\ref{restricted-boltzmann-machines}-\ref{deep-learning}. We will focus
-on restricted Bolzmann machine (RBM) models (Section
-\ref{restricted-boltzmann-machines}, having one layer of latent
+\ref{rbm}-\ref{deep-learning}. We will focus on restricted Boltzmann
+machine (RBM) models (Section \ref{rbm}, having one layer of latent
 variables for simplicity, though the same instability concepts may be
@@
 and the measure of model instability becomes \[
-\frac{1}{N}\REP(\boldsymbol \theta_N) = \frac{2}{N} \sum_{i=1}^N|\theta_{i,N}^{\mathcal{V}}| \quad \mbox{(RBM model, no hiddens)}.
+\frac{1}{N}\REP(\boldsymbol \theta_N) = \frac{2}{N} \sum_{i=1}^N|\theta_{i,N}^{\mathcal{V}}| \quad \text{(RBM model, no hiddens)}.
 \] Hence, this model sequence for \(\boldsymbol X\) will be S-unstable
@@
 Due to their extreme sensitivity in probability structure, S-instability
-in FOES models often translates to numerical complications, and in fact
-obstructions, in both simulation and statistical inference based on
-likelihoods.
+in FOES models may often translate to numerical complications, and in
+fact obstructions, in both simulation and statistical inference based on
+likelihoods. We describe these aspects in Sections
+\ref{mcmc}-\ref{bayes} with regard to data simulation, maximum
+likelihood estimation and Bayes inference, respectively.
 
@@
 unstable FOES model, one component changes in outcomes may produce
-radically different probabilities which then creates numerical barriers
-to MCMC. For example, consider implementation of a Gibbs sampler from
-the full conditional distributions
+radically different probabilities, which may then impose numerical
+barriers to MCMC. For example, consider implementation of a Gibbs
+sampler from the full conditional distributions
 \(P_{\boldsymbol \theta_N} (X_i = x| \boldsymbol x_{-i})\),
@@
 distribution \(P_{\boldsymbol \theta_N}\), then the Gibbs sampler can
-have extreme log-ratios in its transition probabilities \[
-\left| \frac{P_{\boldsymbol \theta_N} (X_i =x_1 | \boldsymbol x^{(1)}_{-i})}{P_{\boldsymbol \theta_N} (X_i = x_2| \boldsymbol x^{(2)}_{-i}) } \right|=\left| \frac{P_{\boldsymbol \theta_N} ( \boldsymbol x^{(1)} )}{P_{\boldsymbol \theta_N} (  \boldsymbol x^{(2)} ) } \right|
-\] using that conditional probabilities are proportional to the joint
-probabilities and the fact that unstable models can have unbounded
-log-probability ratios in one-component changes (Theorem
-\ref{thm:instab-elpr}). This can hinder the ability of chain to
-effectively explore the sample space of the observations
-\(\boldsymbol X\), as the chain may then mix poorly by moving rapidly
-to, and slowly away from, sections of the sample space surface. This
-mixing problem is due to the unstable stationary distribution (unbounded
-ratios of probabilities under the joint model), rather than in the
-particulars of the MCMC algorithm, similar complications can also arise
-for Metropolis-Hastings algorithms for MCMC. Hence, while an unstable
-FOES model \(P_{\boldsymbol \theta_N}\) may be valid and technically
-open to simulation by MCMC, the aspect of instability can ruin numerical
-applications of MCMC.
+have extreme log-ratios in its transition probabilities, \[
+\log \left| \frac{P_{\boldsymbol \theta_N} (X_i =x_1 | \boldsymbol x^{(1)}_{-i})}{P_{\boldsymbol \theta_N} (X_i = x_2| \boldsymbol x^{(2)}_{-i}) } \right|=\log \left| \frac{P_{\boldsymbol \theta_N} ( \boldsymbol x^{(1)} )}{P_{\boldsymbol \theta_N} (  \boldsymbol x^{(2)} ) } \right|,
+\]\\
+as conditional probabilities are proportional to joint probabilities
+that, with unstable models, can have unbounded log-probability ratios in
+one-component changes (Theorem \ref{thm:instab-elpr}). This can hinder
+the ability of chain to effectively explore the sample space of the
+observations \(\boldsymbol X\), as the chain may mix poorly by moving
+rapidly to, and slowly away from, sections of the sample space. In this
+case, for example, the Markov chain may become entrapped within a mode
+of the probability function, with rare chance of escaping to adequately
+mimic the occupation frequencies in the overall sample space. If modes
+of the unstable model are not unique, then important outcomes may be
+missed without multiple chains or impractically enormous numbers of MCMC
+samples. This mixing problem is due to the unstable stationary
+distribution (unbounded ratios of probabilities under the joint model),
+rather than in any particulars of the MCMC algorithm, and similar
+complications can also arise for Metropolis-Hastings algorithms for
+MCMC. Hence, while an unstable FOES model \(P_{\boldsymbol \theta_N}\)
+may be valid and technically open to simulation by MCMC, the aspect of
+instability can render applications of MCMC as numerically infeasible
+for simulation purposes. This result is in line with conclusions of
+Handcock (\protect\hyperlink{ref-handcock2003assessing}{2003}) and
+Schweinberger
+(\protect\hyperlink{ref-schweinberger2011instability}{2011}) for other
+exponential models.
 
@@
 
-The rugged or volatile probability structure in unstable models can also
-hamper efforts to maximize likelihood functions for statistical
-inference. Essentially, when a FOES model is unstable along a parameter
-sequence \(\boldsymbol \theta_N\), the same model can consequently
-unstable along parameters \(-\boldsymbol \theta_N\) opposite from the
-origin (model condition PSR and Corollary \ref{cor:sign}). This leads to
-potential hyper-sensitivity in the likelihood function, which can be
-pulled in extreme and opposite directions depending on how a parameter
-lies with respect to zero and translates into numerical complications in
-maximizing the objective function.
-
-For a FOES model \(P_{\boldsymbol \theta_N}(\boldsymbol x)\),
+Volatility in the probability structure of an unstable model can also
+hamper efforts to maximize likelihood functions in statistical
+inference. When a FOES model is unstable along a parameter sequence
+\(\boldsymbol \theta_N\), the same model can further be unstable along
+parameters \(-\boldsymbol \theta_N\) in an opposite direction from the
+origin (model condition PSR and Corollary \ref{cor:sign}). This can
+translate into potential sensitivity of the likelihood function around
+zero, and lead to numerical complications in maximizing the objective
+function. We next provide a discussion of this issue in a way that
+builds upon and extends related findings by @Schweinberger
+(\protect\hyperlink{ref-schweinberger2011instability}{2011}), who
+largely focused on the case of one-parameter exponential models.
+
+With many probability models, the modes and anti-modes in the
+probability structure under one parameter \(\boldsymbol \theta_N\) are
+reversed in role when the parameter sign changes
+\(-\boldsymbol \theta_N\). Because unstable models tend to degeneracy,
+the opposite signed parameters further push unstable models to assign
+nearly all probability to extremely opposite data configurations, given
+by modes/anti-modes. This is made concrete in Theorem
+\ref{thm:something}, relating the degeneracy from unstable models to the
+expected behavior of log-likelihood functions.
+
+\BeginKnitrBlock{theorem}
+\protect\hypertarget{thm:something}{}{\label{thm:something} }Let
+\(P_{\boldsymbol \theta_N}\), \(N \geq 1\), denote an S-unstable FOES
+model sequence, which additionally satisfies model condition PSR. Let
+\(\boldsymbol{x}_{\max, \boldsymbol \theta_N},\boldsymbol{x}_{\min, \boldsymbol \theta_N}\in\mathcal{X}^N\)
+denote, respectively, a mode and anti-mode of the model
+\(P_{\boldsymbol \theta_N}(\boldsymbol x)\),
 \(\boldsymbol x\in\mathcal{X}^N\), for \(N\) observations
-\(\boldsymbol X =(X_1,\ldots,X_N)\), define the likelihood function
-\(\mathcal{L}(\boldsymbol \theta_N | \boldsymbol x) \equiv N^{-1}\log P_{\boldsymbol \theta_N }(\boldsymbol x)\)
-at \(\boldsymbol \theta_N\) and the minimum probability ratio
-counterpart
-\(\mathcal{L}(\boldsymbol \theta_N | \boldsymbol x) - N^{-1} \log \min_{\boldsymbol \theta_N}= N^{-1}\log[P_{\boldsymbol \theta_N }(\boldsymbol x) /\min_{\boldsymbol \theta_N}]\)
-where
-\(\min_{\boldsymbol \theta_N} \equiv \min_{\boldsymbol y\in\mathcal{X}^N} P_{\boldsymbol \theta_N }(\boldsymbol y)\).
-Also define
-\(\max_{\boldsymbol \theta_N} \equiv \max_{\boldsymbol y\in\mathcal{X}^N} P_{\boldsymbol \theta_N }(\boldsymbol y)\).
-
-\BeginKnitrBlock{theorem}
-\protect\hypertarget{thm:something}{}{\label{thm:something}}Let
-\(P_{\boldsymbol \theta_N}\), \(N \geq 1\), denote an S-unstable FOES
-model sequence (i.e.,
-\(\lim_{N\to \infty}\REP(\boldsymbol \theta_N)/N=\infty\)) which
-additionally satisfies model condition PSR. Then,
-\begin{align*}
-\log\left[ \frac{P_{\boldsymbol \theta_N }(\boldsymbol x) }{\min_{\boldsymbol \theta_N}}\right] +  \log\left[ \frac{P_{-\boldsymbol \theta_N }(\boldsymbol x) }{\min_{-\boldsymbol \theta_N}}\right]  &=    \log\left[ \frac{\max_{\boldsymbol \theta_N}}{\min_{\boldsymbol \theta_N}}\right]  \equiv  \REP(\boldsymbol \theta_N)\\
-& = \log\left[ \frac{\max_{-\boldsymbol \theta_N} }{\min_{-\boldsymbol \theta_N}}\right] \equiv \REP(-\boldsymbol \theta_N)
-\end{align*}
-for any \(\boldsymbol x\in\mathcal{X}^N\) and, as \(N\to \infty\), the
-following convergence \[
-\frac{1}{\REP(\boldsymbol \theta_N)} \log\left[ \frac{P_{\boldsymbol \theta_N }(\boldsymbol X) }{\min_{\boldsymbol \theta_N}}\right]  \rightarrow 1 \qquad \frac{1}{\REP(-\boldsymbol \theta_N)}\log\left[ \frac{P_{-\boldsymbol \theta_N }(\boldsymbol X) }{\min_{-\boldsymbol \theta_N}}\right]  \rightarrow 0
-\] holds in probability \(P_{\boldsymbol \theta_N}\), as well as in
-expectation \(E_{\boldsymbol \theta_N}\), under
-\(\boldsymbol \theta_N\); the limits of convergence are reversed under
-\(-\boldsymbol \theta_N\), \[
-\frac{1}{\REP(\boldsymbol \theta_N)} \log\left[ \frac{P_{\boldsymbol \theta_N }(\boldsymbol x) }{\min_{\boldsymbol \theta_N}}\right]  \rightarrow 0 \qquad \frac{1}{\REP(-\boldsymbol \theta_N)}\log\left[ \frac{P_{-\boldsymbol \theta_N }(\boldsymbol x) }{\min_{-\boldsymbol \theta_N}}\right]  \rightarrow 1
+\(\boldsymbol X = (X_1,\ldots,X_N)\), whereby
+\(P_{\boldsymbol \theta_N }(\boldsymbol{x}_{\max, \boldsymbol \theta_N}) = \max_{\boldsymbol y\in\mathcal{X}^N} P_{\boldsymbol \theta_N }(\boldsymbol y)\)
+and
+\(P_{\boldsymbol \theta_N }(\boldsymbol{x}_{\min, \boldsymbol \theta_N}) = \min_{\boldsymbol y\in\mathcal{X}^N} P_{\boldsymbol \theta_N }(\boldsymbol y)\).
+
+Then, letting \(\stackrel{p, \mathrm{E}}{\longrightarrow}\) denote
+convergence in probability and expectation, as \(N\to \infty\), \[
+\frac{1}{\REP(\boldsymbol \theta_N)} \log\left[ \frac{P_{\boldsymbol \theta_N }(\boldsymbol X) }{ \displaystyle{\min_{\boldsymbol y\in\mathcal{X}^N}} P_{\boldsymbol \theta_N }(\boldsymbol y)}\right] = \frac{ \log P_{\boldsymbol \theta_N } ( \boldsymbol X)  - \log P_{\boldsymbol \theta_N }(\boldsymbol{x}_{\min, \boldsymbol \theta_N}) }
+{ \log P_{\boldsymbol \theta_N } (\boldsymbol{x}_{\max, \boldsymbol \theta_N})  - \log P_{\boldsymbol \theta_N }(\boldsymbol{x}_{\min, \boldsymbol \theta_N}) }
+\stackrel{p, \mathrm{E}}{\longrightarrow}
+\; 1
+\] under \(\boldsymbol\theta_N\) while \[
+\frac{1}{\REP(-\boldsymbol \theta_N)} \log\left[ \frac{P_{-\boldsymbol \theta_N }(\boldsymbol X) }{  \displaystyle{\min_{\boldsymbol y\in\mathcal{X}^N}} P_{-\boldsymbol \theta_N }(\boldsymbol y)} \right]
+=    \frac{ \log P_{-\boldsymbol \theta_N } ( \boldsymbol X) - \log P_{ \boldsymbol \theta_N }(\boldsymbol{x}_{\max, \boldsymbol \theta_N})    }
+{ \log P_{-\boldsymbol \theta_N } (\boldsymbol{x}_{\min, \boldsymbol \theta_N})  - \log P_{-\boldsymbol \theta_N }(\boldsymbol{x}_{\max, \boldsymbol \theta_N}) }
+\stackrel{p,\mathrm{E}}{\longrightarrow}
+\; 1,
+\] under \(-\boldsymbol\theta_N\), where \[
+\REP( \boldsymbol \theta_N) \equiv \log \frac{P_{\boldsymbol \theta_N } (\boldsymbol{x}_{\max, \boldsymbol \theta_N})}{P_{\boldsymbol \theta_N }(\boldsymbol{x}_{\min, \boldsymbol \theta_N})} = \log  \frac{P_{-\boldsymbol \theta_N } (\boldsymbol{x}_{\min, \boldsymbol \theta_N}) }{ P_{-\boldsymbol \theta_N }(\boldsymbol{x}_{\max, \boldsymbol \theta_N})}=   \REP( -\boldsymbol \theta_N), \quad N \geq 1.
 \]
-\EndKnitrBlock{theorem} Schweinberger
+\EndKnitrBlock{theorem}
+
+Theorem \ref{thm:something} entails log-likelihood functions based on
+unstable models are both inversely related and degenerate at opposited
+signed parameters \(\boldsymbol \theta_N\) or \(-\boldsymbol \theta_N\),
+so that likelihoods are highest at different extremes in data
+configuration (e.g., \(\boldsymbol{x}_{\max, \boldsymbol \theta_N}\)
+under \(\boldsymbol \theta\)-probabilities or
+\(\boldsymbol{x}_{\min, \boldsymbol \theta_N}\) under
+\(-\boldsymbol \theta\)-probabilities). If the observed outcome
+\(\boldsymbol x\) for data \(\boldsymbol X\) is not a mode/anti-mode,
+then probabilities for the outcome may be small under both parameters
+\(\boldsymbol \theta_N\) and \(-\boldsymbol \theta_N\), in which case
+associated optimization steps may then shift around zero and struggle to
+converge. In many model formulations, the zero parameter
+\(\boldsymbol \theta_N=\boldsymbol 0\) is a ``safe'' position among
+parameters, representing a guaranteed stable model (having uniform
+probabilities among outcomes), which can also tether a broad parameter
+search attempted among unstable models. Handcock
+(\protect\hyperlink{ref-handcock2003assessing}{2003}) describes similar
+results for degenerate exponential models, and Theorem
+\ref{thm:something} also supports an important finding of Schweinberger
 (\protect\hyperlink{ref-schweinberger2011instability}{2011} Corollary 1)
-considered the one-parameter linear exponential case showing that, for
-positive \(\boldsymbol \theta_N>0\), \[
-\frac{1}{\REP(\boldsymbol \theta_N)} \log\left[ \frac{P_{\boldsymbol \theta_N }(\boldsymbol x) }{\min_{\boldsymbol \theta_N}}\right] = \frac{\boldsymbol g(\boldsymbol x) - L_N}{U_N-L_N} 
-\] holds and converges to \(1\) in expectation
-\(E_{\boldsymbol \theta_N}\), while \[   
-\frac{1}{\REP(-\boldsymbol \theta_N)} \log\left[ \frac{P_{-\boldsymbol \theta_N }(\boldsymbol x) }{\min_{-\boldsymbol \theta_N}}\right] = \frac{U_N -\boldsymbol g(\boldsymbol x)}{U_N-L_N} 
-\] holds converges to \(1\) in expectation
-\(E_{-\boldsymbol \theta_N}\). This result is a special case of Theorem
-\ref{thm:something} and the LREP expansion \eqref{eq:UL} for the one
-parameter exponential model case and is related to a result in Handcock
-(\protect\hyperlink{ref-handcock2003assessing}{2003}).
-
-The idea, more generally, is that probability in an unstable model is
-pushed to opposite regions of the sample space due to degeneracy and
-opposite signed parameters. This can cause numerical complications in
-likelihood maximization. If an observed outcome
-\(\boldsymbol x \equiv \argmax \mathcal{L}(\boldsymbol \theta_N)\) which
-is also the outcome with minimal probability under
-\(P_{-\boldsymbol \theta_N}\), then the likelihood function will have a
-clear peak and maximization straightforward. However, when the observed
-outcome \(\boldsymbol x\) is at an extreme in the probability structures
-of \(P_{\boldsymbol \theta_N}, P_{-\boldsymbol \theta_N}\) (which are
-related) the likelihood function will be flat and optimization steps may
-shift around zero.
-
-\subsection{Implications for Bayes
-Inference}\label{implications-for-bayes-inference}
-
-Considering that changes in sign of parameters can cause all probability
-to be placed on disjoint sections of the sample space (Corollary
-\ref{cor:sign}), then this will create numerical difficulties in Bayes
-inference based on MCMC. Given observations
-\(\boldsymbol X = (X_1,\ldots,X_N)\), switching \(\boldsymbol theta_N\)
-from positive to negative or vice versa could dramatically change
-probabilities (model condition PSR and Corollary \ref{cor:sign}). This
-would translate to numerical challenges similar to those presented in
-Section \ref{mcmc}, i.e., inability of the chain to effectively explore
-the sample space because of unbounded ratios of probabilities. We can
-write the MH acceptance probability to indicate this, \[
-\alpha\left(\boldsymbol x^{(1)} \mid \boldsymbol x^{(2)}\right)= \min\left\{1, 
-\frac{q(\boldsymbol x^{(2)} \mid \boldsymbol x^{(1)})}{q(\boldsymbol x^{(1)} \mid \boldsymbol x^{(2)})}\frac{P_{\boldsymbol \theta_N} ( \boldsymbol x^{(1)} )}{P_{\boldsymbol \theta_N} (  \boldsymbol x^{(2)} ) } \right\}
-\] for some proposal distribution \(q(\cdot)\).
+for one-parameter linear exponential models \eqref{eq:expo}. In the latter
+case, the likelihood score function at \(\boldsymbol \theta_N\) is the
+expected value
+\(\mu(\boldsymbol \theta_N)\equiv \mathrm{E}_{\boldsymbol \theta_N} g(\boldsymbol X)\)
+of the sufficient statistic \(g(\cdot)\), and optimization involves
+solving \(\mu(\cdot)=g(\boldsymbol x)\) for an observed outcome
+\(\boldsymbol x\). For unstable models in this exponential class,
+Schweinberger
+(\protect\hyperlink{ref-schweinberger2011instability}{2011} Corollary 1)
+shows that \[
+\lim_{N\to \infty}\frac{\mu(\boldsymbol \theta_N) -L_n}{U_n-L_N}= \left\{ \begin{array}{lcl}
+1 && \text{for } \boldsymbol \theta_N>0,\\
+0 && \text{for } \boldsymbol \theta_N<0, \end{array}\right.
+\] where again \(U_N\) and \(L_N\) denote the maximum and minimum values
+of the statistic \(g(\boldsymbol x)\),
+\(\boldsymbol x\in\mathcal{X}^N\). As described by Schweinberger
+(\protect\hyperlink{ref-schweinberger2011instability}{2011}), the
+implication for maximum likelihood estimation is that, unless an
+observed outcome \(\boldsymbol x\) falls at an extreme \(U_N, L_N\)
+(i.e., modes/anti-modes), optimization steps in the parameter space can
+iterate in relatively small increments around zero and fail to converge.
+For unstable one-parameter exponential models, the maximum likelihood
+results of Schweinberger
+(\protect\hyperlink{ref-schweinberger2011instability}{2011}) turn out to
+be a special case of Theorem \ref{thm:something} and the LREP expansion
+\eqref{eq:UL} in this setting; namely, for an unstable model with
+\(\boldsymbol \theta_N>0\), \[
+\frac{1}{\REP(\boldsymbol \theta_N)} \log\left[ \frac{P_{\boldsymbol \theta_N }(\boldsymbol X) }{\displaystyle{\min_{\boldsymbol y\in\mathcal{X}^N}} P_{\boldsymbol \theta_N }(\boldsymbol y)}\right] = \frac{\boldsymbol g(\boldsymbol X) - L_N}{U_N-L_N}\;\;\stackrel{p, \mathrm{E}}{\longrightarrow}\;\; 1
+\] holds as \(N\to \infty\) by Theorem \ref{thm:something}, while under
+\(-\boldsymbol \theta_N<0\) \[
+\frac{1}{\REP(-\boldsymbol \theta_N)} \log\left[ \frac{P_{-\boldsymbol \theta_N }(\boldsymbol X) }{\displaystyle{\min_{\boldsymbol y\in\mathcal{X}^N}} P_{-\boldsymbol \theta_N }(\boldsymbol y)}\right] = \frac{U_N -\boldsymbol g(\boldsymbol X)}{U_N-L_N} = 1 - \frac{\boldsymbol g(\boldsymbol X) - L_N}{U_N-L_N} \;\;\stackrel{p, \mathrm{E}}{\longrightarrow}\;\; 1.
+\] Again, when all probability in unstable models may be pushed to
+opposite extremes in the sample space, due to a combination of
+degeneracy and parameter sign, numerical complications in likelihood
+maximization may occur.
+
+\subsection{Implications for Bayes Inference}\label{bayes}
+
+The potential numerical difficulties with maximum likelihood with
+unstable models, as described in the previous section, can naturally
+carry over to Bayes inference. Considering that the degeneracy issues
+related to unstable models can cause likelihoods can be flat (e.g., near
+zero) for many parameters under a given data outcome and that sign
+changes in parameters can shift tremendous probability to extreme and
+opposite outcomes in the sample space (e.g., Corollary \ref{cor:sign},
+Theorem \ref{thm:something}), then numerical complications may arise
+with Bayes inference in sampling a posterior parameter space based on
+MCMC. The potential challenges in chain mixing are similar to those
+presented in Section \ref{mcmc}, though in chain movements through the
+parameter space as opposed to the sample space in data generation. That
+is, in the Bayes setting for sampling a posterior distribution for
+\(\boldsymbol \theta_N\), a chain may unstable to effectively explore
+the parameter space due partly to extreme and potentially unbounded
+probability ratios from parameter sign changes, which represents a
+parameter space analog to how one component changes in the sample space
+may impact data simulation with unstable models. For example, if
+\(\pi(\cdot)\) denotes a prior density for \(\boldsymbol \theta_N\) and
+\(q(\cdot | \cdot)\) denotes a proposal distribution for use in a
+Metropolis-Hastings (MH) sampler, then MH acceptance probability becomes
+\[
+\alpha\left(\boldsymbol \theta^{(1)} \mid \boldsymbol \theta^{(2)}\right)= \min\left\{1,
+\frac{q(\boldsymbol \theta_N^{(2)} \mid \boldsymbol \theta_N^{(1)})}{q(\boldsymbol \theta_N^{(1)} \mid \boldsymbol \theta_N^{(2)})}
+\frac{P_{\boldsymbol \theta_N^{(1)}} ( \boldsymbol x ) \pi(\boldsymbol \theta_N^{(1)}) }{P_{\boldsymbol \theta^{(2)}_N} (  \boldsymbol x) \pi(\boldsymbol \theta_N^{(2)}) } \right\},
+\] which indicates how parameter sensitivity in the likelihood
+\(P_{\boldsymbol \theta_N} ( \boldsymbol x)\) may complicate sampling of
+the posterior
+\(P_{\boldsymbol \theta_N} ( \boldsymbol x) \pi(\boldsymbol \theta_N)\)
+(i.e., moving from \(\boldsymbol \theta^{(1)}\) to
+\(\boldsymbol \theta^{(2)}\) in the parameter space). Furthermore, the
+potential for model instability and the size of the parameter space can
+also become greater with the introduction of latent variables to
+existing data variables, as involved in some model formulations
+described in Sections \ref{rbm}-\ref{deep-learning}. As latent variables
+are often sampled with parameters in a Bayes MCMC approach, this aspect
+may further compound numerical problems in chain mixing.
 
@@
 place all probability on certain modal subsections of the sample space,
-which potentially could be small. The FOES model class is quite broad
-and, particularly in developing FOES models for large data sets, some
-caution should be used in parameter specification to control effects of
-model instability due to the consequences of instability in both
-simulation and likelihood inference.
+which potentially could be small. Such instability is often due to a
+complex interaction between the model statistics used (i.e., how
+numerous and large these may become) and the number and magnitudes of
+parameters in the model formulation. For many FOES models, the
+possibility exists, at least in principle, to constraint parameters in a
+way balances their potential contributions against those of model
+statistics in order to prevent probability instabilities. The FOES model
+class is quite broad and, in developing such models for large data sets,
+some caution should be used in parameter specification to control
+effects of model instability.
 
@@
 
+\textbf{Proof of Theorem \ref{thm:something}.} Let
+\(L_{\boldsymbol \theta_{N}}(\boldsymbol X) = \log[ P_{\boldsymbol \theta_{N}}(\boldsymbol X)/ \min_{\boldsymbol y \in \mathcal{X}^N} P_{\boldsymbol \theta_{N}}(\boldsymbol y) ]/\REP(\boldsymbol \theta_{N})\),
+where again \(\boldsymbol X=(X_1, \dots, X_N)\) and
+\(\REP(\boldsymbol \theta_{N})= \log[\max_{\boldsymbol y \in \mathcal{X}^N} P_{\boldsymbol \theta_{N}}(\boldsymbol y)/ \min_{\boldsymbol y \in \mathcal{X}^N} P_{\boldsymbol \theta_{N}}(\boldsymbol y) ]\).
+As \(L_{\boldsymbol \theta_{N}}(\boldsymbol X)\in[0,1]\), convergence of
+\(L_{\boldsymbol \theta_{N}}(\boldsymbol X)\) to 1 in probability under
+\(P_{\boldsymbol \theta_{N}}\) is equivalent to convergence to \(1\) in
+expectation under \(P_{\boldsymbol \theta_{N}}\) (i.e., convergence in
+expectation implies probabilistic convergence by Markov's inequality
+while probabilistic convergence implies convergence in expectation by
+uniform integrability/boundedness).
+
+For \(\epsilon \in(0,1)\), let
+\(\mathcal{M}_{\epsilon, \boldsymbol \theta_N}\) denote a modal set as
+in \eqref{eq:mode}. By Theorem \ref{thm:degenFOES},
+\(P_{\boldsymbol \theta_N}\left( \boldsymbol X\in \mathcal{M}_{\epsilon, \boldsymbol \theta_N}\right) \rightarrow 1\)
+holds as \(N \rightarrow \infty\) and, by definition of \eqref{eq:mode},
+\(\boldsymbol X\in \mathcal{M}_{\epsilon, \boldsymbol \theta_N}\)
+follows if and only if
+\(1-L_{\boldsymbol \theta_{N}}(\boldsymbol X)<\epsilon\). Hence,
+\(L_{\boldsymbol \theta_{N}}(\boldsymbol X) \stackrel{p,\mathrm{E}}{\longrightarrow} 1\)
+holds under \(\boldsymbol \theta_{N}\) in Theorem \ref{thm:something}.
+The convergence
+\(L_{-\boldsymbol \theta_{N}}(\boldsymbol X) \stackrel{p,\mathrm{E}}{\longrightarrow} 1\)
+under \(-\boldsymbol \theta_{N}\) likewise follows from Corollary
+\ref{cor:sign}. \hfill \(\Box\)
+
 \clearpage

### 2017-11-06 writing/paper-note/chapters/instab.Rmd
ready?

@@
 For instance, this latter condition is fulfilled for all linear exponential families from Section \@ref(discrete-exponential-family-models) (e.g., \@ref(eq:mod1)-\@ref(eq:mod2)) as well as all network models from Sections \@ref(rbm)-\@ref(deep-learning) (e.g., \@ref(eq:RBM1)-\@ref(eq:RBM2)).
-When parameters can be tuned in sign with effects prescribed in the model condition PSR, unstable FOES models will exhibit further probability sensitivities, as outlined in the following extension of Theorem \@ref(thm:degenFOES}).
+When parameters can be tuned in sign with effects prescribed in the model condition PSR, unstable FOES models will exhibit further probability sensitivities, as outlined in the following extension of Theorem \@ref(thm:degenFOES).
 
@@
 cases where parameter components $\boldsymbol \theta_N =(\boldsymbol \theta_{1,N}, \boldsymbol \theta_{N,2})$ (say) are not all changed in sign (e.g., $-\boldsymbol \theta_N$) but, more generally, are instead  altered to another parameter configuration $\boldsymbol \theta_{N}^A = ( \boldsymbol \theta_{1,N}^A, \boldsymbol \theta_{2,N}^A )$ involving a switch in sign only among some dominating model parameters $\boldsymbol \theta_{N,2}^A=- \boldsymbol \theta_{N,2}$ with remaining parameters $\boldsymbol \theta_{1,N}^A$ being arbitrarily chosen.
-If a change sign occurs among parameters ($\pm \boldsymbol \theta_{N,2}$) which dominate the probability structure of the model, then the results of Corollary  \ref{cor:sign} can still hold  with $\boldsymbol \theta_{N}^A$ replacing $-\boldsymbol \theta_N$; as an example of one sufficient condition, if $\lim_{N\to \infty} \max_{\boldsymbol x \in \mathcal{X}^N} |G_N(\boldsymbol x, -\boldsymbol \theta_{N}) - G_N(\boldsymbol x, \boldsymbol \theta_{N}^A)|=0$ holds in addition to Corollary~1 assumptions, where
+If a change sign occurs among parameters ($\pm \boldsymbol \theta_{N,2}$) which dominate the probability structure of the model, then the results of Corollary  \ref{cor:sign} can still hold  with $\boldsymbol \theta_{N}^A$ replacing $-\boldsymbol \theta_N$; as an example of one sufficient condition, if $\lim_{N\to \infty} \max_{\boldsymbol x \in \mathcal{X}^N} |G_N(\boldsymbol x, -\boldsymbol \theta_{N}) - G_N(\boldsymbol x, \boldsymbol \theta_{N}^A)|=0$ holds in addition to Corollary \@ref(cor:sign) assumptions, where
 $$
@@
     $$
-    based on $f_{\boldsymbol \theta_N}$ from \@ref(eq:f) with components $\boldsymbol x \in \{\pm 1\}^{N}, \boldsymbol h \in \{\pm 1\}^{N_{\mathcal{H}}}$.
+    based on $f_{\boldsymbol \theta_N}$ from (\ref{eq:f}) with components $\boldsymbol x \in \{\pm 1\}^{N}, \boldsymbol h \in \{\pm 1\}^{N_{\mathcal{H}}}$.
 \item The instability measure $\REP_{(\boldsymbol X, \boldsymbol H)}(\boldsymbol \theta_N)\equiv \left(\max_{ \boldsymbol x} \max_{ \boldsymbol h  } f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)-\min_{ \boldsymbol x } \min_{ \boldsymbol h }f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)\right)$ for the joint model $\tilde{P}_{\boldsymbol \theta_N}$ of $(\boldsymbol X, \boldsymbol H)$ satisfies
@@
 Nevertheless, because the joint model is often employed in practice for purposes of simulation and simulation-based inference, it is still reasonable to consider parameter choices for ensuring a stable joint model (and, consequently, a stable visible model as well).
+Further evidence of this is seen in the following numerical example.
+\par
+
+```{r rbm-simul, echo=FALSE, warning=FALSE, message=FALSE, cache=TRUE, cache.lazy = FALSE}
+#libraries -----------------------
+source("../resources/code/functions.R")
+
+#functions ----------------------
+sample_sphere_unif <- function(space, rep, center, r) {
+  vec <- matrix(rnorm(space*rep), nrow = rep)
+  r*vec/t(matrix(rep(sqrt(rowSums(vec^2)), each = space), nrow = space))
+}
+
+#sample thetas --------------
+n <- 100
+V <- 9
+H <- 5
+
+expand.grid(H = H, V = V, r1 = seq(0.001, 3, length.out = 20), r2 = seq(0.001, 3, length.out = 20)) %>%
+  mutate(r1 = r1*(H + V), r2 = r2*(H*V)) %>%
+  mutate(N = H + V) %>%
+  group_by(H, V, r1, r2, N) %>%
+  do(samp = cbind(sample_sphere_unif(.$N, n, 0, .$r1), sample_sphere_unif(.$H*.$V, n, 0, .$r2))) -> grid_sample
+
+expand.grid(H = H, V = V, r1 = seq(0.001, 3, length.out = 20), r2 = seq(0.001, 3, length.out = 20)) %>%
+  mutate(r1 = r1*(H + V), r2 = r2*(H*V)) %>%
+  mutate(N = H + V) %>%
+  group_by(H, V, r1, r2, N) %>%
+  do(stat = stats(.$H, .$V, "negative")) -> grid_sample_stat
+
+grid_sample <- inner_join(grid_sample, grid_sample_stat)
+
+grid_sample %>%
+  group_by(H, V, N, r1, r2) %>%
+  do(g_theta = t(rbind(t(.$samp[[1]]), expected_value(theta = t(.$samp[[1]]), stats = data.matrix(.$stat[[1]]))))) -> grid
+
+res <- inner_join(grid, grid_sample)
+
+#reshape data functions ----------------
+elpr <- function(theta, stats) {
+  require(dplyr)
+  exp(tcrossprod(stats, theta)) %>% data.frame() %>%
+    cbind(stats) %>%
+    group_by_(.dots = colnames(stats)[grepl("v", colnames(stats)) & !grepl("theta", colnames(stats))]) %>%
+    summarise_each(funs(sum), contains("X")) %>%
+    ungroup() -> marginalized
+
+  marginalized %>%
+    summarise_each(funs(max), contains("X")) %>%
+    select(contains("X")) %>%
+    data.matrix() -> max_marg
+
+  marginalized %>%
+    summarise_each(funs(min), contains("X")) %>%
+    select(contains("X")) %>%
+    data.matrix() -> min_marg
+
+  t(log(max_marg/min_marg))
+}
+
+#instability
+res %>%
+  group_by(H, V, N, r1, r2) %>%
+  do(data.frame(elpr = elpr(.$samp[[1]], .$stat[[1]]))) %>%
+  ungroup() %>%
+  group_by(H, V, N, r1, r2) %>% 
+  summarise(mean_elpr = mean(elpr)) %>%
+  mutate(scaled_mean_elpr = mean_elpr/V) %>%
+  ungroup() -> elpr_summary
+
+res %>%
+  group_by(H, V, N, r1, r2) %>%
+  do(data.frame(max_prob = apply(.$samp[[1]], 1, function(row) {
+    params <- list(main_visible = row[1:V] %>% data.matrix(),
+                   main_hidden = row[(V + 1):(V + H)] %>% data.matrix(),
+                   interaction = row[(V + H + 1):length(row)] %>% data.matrix() %>% matrix(H))
+
+    distn <- visible_distn(params) %>% mutate(out_vals = pmap(lst_(paste0("v", 1:V)), c)) %>% select(-(starts_with("v")))
+
+    compare <- expand.grid(out1 = distn$image_id, out2 = distn$image_id) %>%
+      left_join(distn, by = c("out1" = "image_id")) %>%
+      left_join(distn, by = c("out2" = "image_id")) %>%
+      mutate(test = map2_dbl(out_vals.x, out_vals.y, function(a, b) sum(a != b))) %>%
+      filter(test == 1) %>%
+      mutate(delta = log(prob.x/prob.y))
+
+    max(compare$delta)
+    })
+    )) %>%
+  ungroup() %>%
+  group_by(H, V, N, r1, r2) %>% 
+  summarise(mean_max_prob = mean(max_prob)) %>%
+  ungroup() -> mean_max_prob_diff_summary
+```
+
+```{r rbm-plots, fig.show='hold', fig.cap="The sample mean value of $\\text{ELPR}(\\boldsymbol \\theta)/N_{\\mathcal{V}}$ (left) and $\\DN(\\boldsymbol \\theta)$ at each grid point for each combination of magnitude of $\\boldsymbol \\theta$.
+As the magnitude of $\\boldsymbol \\theta$ grows, so does the value of these metrics, indicating typical instability in the model.", fig.height=3.5, fig.align="ht!"}
+elpr_summary %>%
+  mutate(r1 = round(r1/(H + V), 8), r2 = round(r2/(H * V), 8)) %>% 
+  ggplot() +
+  geom_tile(aes(x = r1, y = r2, fill = scaled_mean_elpr)) +
+  geom_contour(aes(x = r1, y = r2, z = scaled_mean_elpr), colour = "black", bins = 8) +
+  geom_abline(aes(intercept = 0, slope = 1), alpha = .5, lty = 2) +
+  scale_fill_gradient(expression(paste("Mean ", frac(ELPR(theta), N[italic(V)]))), low = "yellow", high = "red") +
+  xlab(expression(paste(paste("||", theta[main], "||", "/(", N[italic(H)]," + ", N[italic(V)], ")")))) +
+  ylab(expression(paste(paste("||", theta[interaction], "||", "/(", N[italic(H)], "*", N[italic(V)], ")")))) +
+  theme(aspect.ratio = 1, legend.position = "bottom")
+
+# fix labels
+mean_max_prob_diff_summary %>%
+  mutate(r1 = round(r1/(H + V), 8), r2 = round(r2/(H * V), 8)) %>% 
+  ggplot() +
+  geom_tile(aes(x = r1, y = r2, fill = mean_max_prob)) +
+  geom_contour(aes(x = r1, y = r2, z = mean_max_prob), colour = "black", bins = 8) +
+  geom_abline(aes(intercept = 0, slope = 1), alpha = .5, lty = 2) +
+  scale_fill_gradient(expression(paste("Mean ", Delta[N](theta))), low = "yellow", high = "red") +
+  xlab(expression(paste(paste("||", theta[main], "||", "/(", N[italic(H)]," + ", N[italic(V)], ")")))) +
+  ylab(expression(paste(paste("||", theta[interaction], "||", "/(", N[italic(H)], "*", N[italic(V)], ")")))) +
+  theme(aspect.ratio = 1, legend.position = "bottom") 
+```
+
+In our numerical experiment, we allow the two types of terms (main effects terms corresponding to visible and hidden parameters $\boldsymbol \theta_{main} = (\boldsymbol \theta_N^{\mathcal{V}}, \boldsymbol \theta_N^{\mathcal{H}})$ and interaction parameters $\boldsymbol \theta_N^{\mathcal{VH}}$) to have varying average magnitudes, $||\boldsymbol \theta_{main} || /(N_{\mathcal{H}}+N_{\mathcal{V}})$ and $||\boldsymbol \theta_{interaction} || /(N_{\mathcal{H}}*N_{\mathcal{V}})$ for a RBM with $N_\mathcal{V} = `r V`$ visibles and $N_\mathcal{H} = `r H`$ hiddens.
+These average magnitudes vary on a grid between `r min(elpr_summary$r1/(elpr_summary$H + elpr_summary$V))` and `r max(elpr_summary$r1/(elpr_summary$H + elpr_summary$V))` with `r length(unique(elpr_summary$r1/(elpr_summary$H + elpr_summary$V)))` breaks, yielding `r length(unique(elpr_summary$r1/(elpr_summary$H + elpr_summary$V)))^2` grid points.
+At each point in the grid, `r n` vectors ($\boldsymbol \theta_{main}$) are sampled uniformly on a sphere with radius corresponding to the first coordinate in the grid and `r n` vectors ($\boldsymbol \theta_{interaction}$) are sampled uniformly on a sphere with radius corresponding to the second coordinate in the grid via sums of squared and scaled iid Normal$(0, 1)$ variables.
+These vectors are then paired to create `r n` values of $\boldsymbol \theta_N$ with magnitudes at each point in the grid.
+The values $\REP(\boldsymbol \theta_N)/N_{\mathcal{V}}$ and $\DN(\boldsymbol \theta_N)$ are then calculated for each $\boldsymbol \theta_N$ and then summarized for each point in the grid using the sample mean.
+The results of this numerical study are shown in Figure \@ref(fig:rbm-plots).
+From these two plots, it is clear that for larger magnitudes of the parameter vectors, there is evidence of S-instability in that the log-ratio of extremal probabilities scaled by $N_{\mathcal{V}}$ and the the biggest log-probability ratio for a one component change in data outcomes are both increasing away from $\boldsymbol \theta_N = \boldsymbol 0$, further supporting \@ref(prp:prop2)(iii.2 and iii.5).
 
@@
 potential sensitivity of the likelihood function around zero, and lead to numerical complications in maximizing the objective function.
-We next provide a discussion of this issue in a way that builds upon and extends related findings by @@schweinberger2011instability, who largely focused on the case of one-parameter exponential models.
+We next provide a discussion of this issue in a way that builds upon and extends related findings by @schweinberger2011instability, who largely focused on the case of one-parameter exponential models.
 

### 2017-11-06 writing/paper-note/docs/note.tex
ready?

@@
 
-\title{\bf A Note on the Instability and Degeneracy of Deep Learning Models}
+\title{\bf A note on the instability and degeneracy of deep learning models}
 \author{Andee Kaplan \\ Department of Statistics, Iowa State University \\  and \\ Daniel Nordman \\ Department of Statistics, Iowa State University \\  and \\ Stephen Vardeman \\ Department of Statistics and Department of Industrial and\\
@@
 outcome result in large, and often unanticipated, changes in
-probability. For correlated data structures found in several application
-areas, there is increasing interest in predicting/identifying such
-sensitivity in model probability structure. We consider the problem of
-quantifying instability for general probability models defined on
-sequences of observations, where each sequence of length \(N\) has a
-finite number of possible outcomes. A sequence of probability models
-results, indexed by \(N\), that accommodates data of expanding
-dimension. Model instability is formally shown to occur when a certain
-log-probability ratio under such models grows faster than \(N\). In this
-case, a one component change in the data sequence can shift probability
-by orders of magnitude. Also, as instability becomes more extreme, the
-resulting probability models are shown to tend to degeneracy, placing
-all their probability on potentially small portions of the sample space.
-These results on instability apply to large classes of models commonly
-used in random graphs, network analysis, and machine learning contexts.
+probability. This instability is a property of the probability model,
+rather than the fitted parameter vector. For correlated data structures
+found in several application areas, there is increasing interest in
+predicting/identifying such sensitivity in model probability structure.
+We consider the problem of quantifying instability for general
+probability models defined on sequences of observations, where each
+sequence of length \(N\) has a finite number of possible values. A
+sequence of probability models results, indexed by \(N\), that
+accommodates data of expanding dimension. Model instability is formally
+shown to occur when a certain log-probability ratio under such models
+grows faster than \(N\). In this case, a one component change in the
+data sequence can shift probability by orders of magnitude. Also, as
+instability becomes more extreme, the resulting probability models are
+shown to tend to degeneracy, placing all their probability on
+potentially small portions of the sample space. These results on
+instability apply to large classes of models commonly used in random
+graphs, network analysis, and machine learning contexts.
 \end{abstract}
@@
 \eqref{eq:mod1}-\eqref{eq:mod2}) as well as all network models from Sections
-\ref{restricted-boltzmann-machines}-\ref{deep-learning} (e.g.,
-\eqref{eq:RBM1}-\eqref{eq:RBM2}). When parameters can be tuned in sign with
-effects prescribed in the model condition PSR, unstable FOES models will
-exhibit further probability sensitivities, as outlined in the following
-extension of Theorem @ref(thm:degenFOES\}).
+\ref{rbm}-\ref{deep-learning} (e.g., \eqref{eq:RBM1}-\eqref{eq:RBM2}). When
+parameters can be tuned in sign with effects prescribed in the model
+condition PSR, unstable FOES models will exhibit further probability
+sensitivities, as outlined in the following extension of Theorem
+\ref{thm:degenFOES}.
 
 \BeginKnitrBlock{corollary}
-\protect\hypertarget{cor:sign}{}{\label{cor:sign}}Let
+\protect\hypertarget{cor:sign}{}{\label{cor:sign} }Let
 \(P_{\boldsymbol \theta_N}\), with support \(\mathcal{X}^N\),
@@
 \BeginKnitrBlock{remark}
-\iffalse <span class="remark"><em>Remark. \fi{} Model condition PSR and
-Corollary \ref{cor:sign} can be extended to cases where parameter
-components
+\iffalse{} {Remark. } \fi{}Under the model condition PSR, Corollary
+\ref{cor:sign} can also be extended to cases where parameter components
 \(\boldsymbol \theta_N =(\boldsymbol \theta_{1,N}, \boldsymbol \theta_{N,2})\)
-are not all changed in sign (\(-\boldsymbol \theta_N\)) but, more
-generally, are instead altered to another parameter configuration
-\(\boldsymbol \theta_{N}^A = (\boldsymbol \theta_{N,1}^A, \boldsymbol \theta_{N,2}^A)\)
-involving a switch in sign only among some model parameters
-\(\boldsymbol \theta_{N,2}^A = -\boldsymbol \theta_{N,2}\) with
-remaining parameters \(\boldsymbol \theta_{1,N}^A\) being arbitrarily
-chosen. If a sign change occurs among parameters
-(\(\pm \boldsymbol \theta_{N,2}\)) which dominate the probability
-structure of the model, then an unstable model under
-\(\boldsymbol \theta_N\) will then imply that many more unstable models
-exist over a broad spectrum of possibilities for variations
-\(\boldsymbol \theta_{N}^A\) of \(\boldsymbol \theta_N\).
+(say) are not all changed in sign (e.g., \(-\boldsymbol \theta_N\)) but,
+more generally, are instead altered to another parameter configuration
+\(\boldsymbol \theta_{N}^A = ( \boldsymbol \theta_{1,N}^A, \boldsymbol \theta_{2,N}^A )\)
+involving a switch in sign only among some dominating model parameters
+\(\boldsymbol \theta_{N,2}^A=- \boldsymbol \theta_{N,2}\) with remaining
+parameters \(\boldsymbol \theta_{1,N}^A\) being arbitrarily chosen. If a
+change sign occurs among parameters (\(\pm \boldsymbol \theta_{N,2}\))
+which dominate the probability structure of the model, then the results
+of Corollary \ref{cor:sign} can still hold with
+\(\boldsymbol \theta_{N}^A\) replacing \(-\boldsymbol \theta_N\); as an
+example of one sufficient condition, if
+\(\lim_{N\to \infty} \max_{\boldsymbol x \in \mathcal{X}^N} |G_N(\boldsymbol x, -\boldsymbol \theta_{N}) - G_N(\boldsymbol x, \boldsymbol \theta_{N}^A)|=0\)
+holds in addition to Corollary \ref{cor:sign} assumptions, where \[
+G_N(\boldsymbol x, \boldsymbol \theta) =   \frac{\log P_{\boldsymbol \theta}(\boldsymbol x) - \min_{\boldsymbol y \in\mathcal{X}^N}\log P_{\boldsymbol \theta}(\boldsymbol y) }{\max_{\boldsymbol y \in\mathcal{X}^N}\log P_{\boldsymbol \theta}(\boldsymbol y)  - \min_{\boldsymbol y \in\mathcal{X}^N}\log P_{\boldsymbol \theta}(\boldsymbol y) }, \quad \boldsymbol x \in\mathcal{X}^N,
+\] represents a standardized form of \(\boldsymbol \theta\)-model
+probabilities, then the results of Corollary \ref{cor:sign} apply to
+\(\boldsymbol \theta_{N}^A\) in addition to \(-\boldsymbol \theta_N\).
+As a consequence, an unstable model under \(\boldsymbol \theta_N\) can
+then imply that many more unstable models exist over a broader spectrum
+of possibilities for variations \(\boldsymbol \theta_{N}^A\) of
+\(\boldsymbol \theta_N\), which involves some amount of sign change
+among components of \(\boldsymbol \theta_N\).
 \EndKnitrBlock{remark}
@@
 
-\protect\hypertarget{prp:prop2}{}{\label{prp:prop2}}Let
+\protect\hypertarget{prp:prop2}{}{\label{prp:prop2} }Let
 \(P_{\boldsymbol \theta_N}\) denote a RBM-based data model \eqref{eq:RBM2}
@@
     $$
-    based on $f_{\boldsymbol \theta_N}$ from \\eqref{eq:f} with components $\boldsymbol x \in \{\pm 1\}^{N}, \boldsymbol h \in \{\pm 1\}^{N_{\mathcal{H}}}$.
+    based on $f_{\boldsymbol \theta_N}$ from (\ref{eq:f}) with components $\boldsymbol x \in \{\pm 1\}^{N}, \boldsymbol h \in \{\pm 1\}^{N_{\mathcal{H}}}$.
 \item The instability measure $\REP_{(\boldsymbol X, \boldsymbol H)}(\boldsymbol \theta_N)\equiv \left(\max_{ \boldsymbol x} \max_{ \boldsymbol h  } f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)-\min_{ \boldsymbol x } \min_{ \boldsymbol h }f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)\right)$ for the joint model $\tilde{P}_{\boldsymbol \theta_N}$ of $(\boldsymbol X, \boldsymbol H)$ satisfies
@@
 choices for ensuring a stable joint model (and, consequently, a stable
-visible model as well).
+visible model as well). Further evidence of this is seen in the
+following numerical example. \par
+\begin{figure}
+\includegraphics{note_files/figure-latex/rbm-plots-1} \includegraphics{note_files/figure-latex/rbm-plots-2} \caption{The sample mean value of $\text{ELPR}(\boldsymbol \theta)/N_{\mathcal{V}}$ (left) and $\DN(\boldsymbol \theta)$ at each grid point for each combination of magnitude of $\boldsymbol \theta$.
+As the magnitude of $\boldsymbol \theta$ grows, so does the value of these metrics, indicating typical instability in the model.}\label{fig:rbm-plots}
+\end{figure}
+In our numerical experiment, we allow the two types of terms (main
+effects terms corresponding to visible and hidden parameters
+\(\boldsymbol \theta_{main} = (\boldsymbol \theta_N^{\mathcal{V}}, \boldsymbol \theta_N^{\mathcal{H}})\)
+and interaction parameters \(\boldsymbol \theta_N^{\mathcal{VH}}\)) to
+have varying average magnitudes,
+\(||\boldsymbol \theta_{main} || /(N_{\mathcal{H}}+N_{\mathcal{V}})\)
+and
+\(||\boldsymbol \theta_{interaction} || /(N_{\mathcal{H}}*N_{\mathcal{V}})\)
+for a RBM with \(N_\mathcal{V} = 9\) visibles and \(N_\mathcal{H} = 5\)
+hiddens. These average magnitudes vary on a grid between 0.001 and 3
+with 20 breaks, yielding 400 grid points. At each point in the grid, 100
+vectors (\(\boldsymbol \theta_{main}\)) are sampled uniformly on a
+sphere with radius corresponding to the first coordinate in the grid and
+100 vectors (\(\boldsymbol \theta_{interaction}\)) are sampled uniformly
+on a sphere with radius corresponding to the second coordinate in the
+grid via sums of squared and scaled iid Normal\((0, 1)\) variables.
+These vectors are then paired to create 100 values of
+\(\boldsymbol \theta_N\) with magnitudes at each point in the grid. The
+values \(\REP(\boldsymbol \theta_N)/N_{\mathcal{V}}\) and
+\(\DN(\boldsymbol \theta_N)\) are then calculated for each
+\(\boldsymbol \theta_N\) and then summarized for each point in the grid
+using the sample mean. The results of this numerical study are shown in
+Figure \ref{fig:rbm-plots}. From these two plots, it is clear that for
+larger magnitudes of the parameter vectors, there is evidence of
+S-instability in that the log-ratio of extremal probabilities scaled by
+\(N_{\mathcal{V}}\) and the the biggest log-probability ratio for a one
+component change in data outcomes are both increasing away from
+\(\boldsymbol \theta_N = \boldsymbol 0\), further supporting
+\ref{prp:prop2}(iii.2 and iii.5).
 
@@
 
-The rugged or volatile probability structure in unstable models can also
-hamper efforts to maximize likelihood functions for statistical
-inference. Essentially, when a FOES model is unstable along a parameter
-sequence \(\boldsymbol \theta_N\), the same model can consequently
-unstable along parameters \(-\boldsymbol \theta_N\) opposite from the
-origin (model condition PSR and Corollary \ref{cor:sign}). This leads to
-potential hyper-sensitivity in the likelihood function, which can be
-pulled in extreme and opposite directions depending on how a parameter
-lies with respect to zero and translates into numerical complications in
-maximizing the objective function.
-
-For a FOES model \(P_{\boldsymbol \theta_N}(\boldsymbol x)\),
+Volatility in the probability structure of an unstable model can also
+hamper efforts to maximize likelihood functions in statistical
+inference. When a FOES model is unstable along a parameter sequence
+\(\boldsymbol \theta_N\), the same model can further be unstable along
+parameters \(-\boldsymbol \theta_N\) in an opposite direction from the
+origin (model condition PSR and Corollary \ref{cor:sign}). This can
+translate into potential sensitivity of the likelihood function around
+zero, and lead to numerical complications in maximizing the objective
+function. We next provide a discussion of this issue in a way that
+builds upon and extends related findings by Schweinberger
+(\protect\hyperlink{ref-schweinberger2011instability}{2011}), who
+largely focused on the case of one-parameter exponential models.
+
+With many probability models, the modes and anti-modes in the
+probability structure under one parameter \(\boldsymbol \theta_N\) are
+reversed in role when the parameter sign changes
+\(-\boldsymbol \theta_N\). Because unstable models tend to degeneracy,
+the opposite signed parameters further push unstable models to assign
+nearly all probability to extremely opposite data configurations, given
+by modes/anti-modes. This is made concrete in Theorem
+\ref{thm:something}, relating the degeneracy from unstable models to the
+expected behavior of log-likelihood functions.
+
+\BeginKnitrBlock{theorem}
+\protect\hypertarget{thm:something}{}{\label{thm:something} }Let
+\(P_{\boldsymbol \theta_N}\), \(N \geq 1\), denote an S-unstable FOES
+model sequence, which additionally satisfies model condition PSR. Let
+\(\boldsymbol{x}_{\max, \boldsymbol \theta_N},\boldsymbol{x}_{\min, \boldsymbol \theta_N}\in\mathcal{X}^N\)
+denote, respectively, a mode and anti-mode of the model
+\(P_{\boldsymbol \theta_N}(\boldsymbol x)\),
 \(\boldsymbol x\in\mathcal{X}^N\), for \(N\) observations
-\(\boldsymbol X =(X_1,\ldots,X_N)\), define the likelihood function
-\(\mathcal{L}(\boldsymbol \theta_N | \boldsymbol x) \equiv N^{-1}\log P_{\boldsymbol \theta_N }(\boldsymbol x)\)
-at \(\boldsymbol \theta_N\) and the minimum probability ratio
-counterpart
-\(\mathcal{L}(\boldsymbol \theta_N | \boldsymbol x) - N^{-1} \log \min_{\boldsymbol \theta_N}= N^{-1}\log[P_{\boldsymbol \theta_N }(\boldsymbol x) /\min_{\boldsymbol \theta_N}]\)
-where
-\(\min_{\boldsymbol \theta_N} \equiv \min_{\boldsymbol y\in\mathcal{X}^N} P_{\boldsymbol \theta_N }(\boldsymbol y)\).
-Also define
-\(\max_{\boldsymbol \theta_N} \equiv \max_{\boldsymbol y\in\mathcal{X}^N} P_{\boldsymbol \theta_N }(\boldsymbol y)\).
-
-\BeginKnitrBlock{theorem}
-\protect\hypertarget{thm:something}{}{\label{thm:something}}Let
-\(P_{\boldsymbol \theta_N}\), \(N \geq 1\), denote an S-unstable FOES
-model sequence (i.e.,
-\(\lim_{N\to \infty}\REP(\boldsymbol \theta_N)/N=\infty\)) which
-additionally satisfies model condition PSR. Then,
-\begin{align*}
-\log\left[ \frac{P_{\boldsymbol \theta_N }(\boldsymbol x) }{\min_{\boldsymbol \theta_N}}\right] +  \log\left[ \frac{P_{-\boldsymbol \theta_N }(\boldsymbol x) }{\min_{-\boldsymbol \theta_N}}\right]  &=    \log\left[ \frac{\max_{\boldsymbol \theta_N}}{\min_{\boldsymbol \theta_N}}\right]  \equiv  \REP(\boldsymbol \theta_N)\\
-& = \log\left[ \frac{\max_{-\boldsymbol \theta_N} }{\min_{-\boldsymbol \theta_N}}\right] \equiv \REP(-\boldsymbol \theta_N)
-\end{align*}
-for any \(\boldsymbol x\in\mathcal{X}^N\) and, as \(N\to \infty\), the
-following convergence \[
-\frac{1}{\REP(\boldsymbol \theta_N)} \log\left[ \frac{P_{\boldsymbol \theta_N }(\boldsymbol X) }{\min_{\boldsymbol \theta_N}}\right]  \rightarrow 1 \qquad \frac{1}{\REP(-\boldsymbol \theta_N)}\log\left[ \frac{P_{-\boldsymbol \theta_N }(\boldsymbol X) }{\min_{-\boldsymbol \theta_N}}\right]  \rightarrow 0
-\] holds in probability \(P_{\boldsymbol \theta_N}\), as well as in
-expectation \(E_{\boldsymbol \theta_N}\), under
-\(\boldsymbol \theta_N\); the limits of convergence are reversed under
-\(-\boldsymbol \theta_N\), \[
-\frac{1}{\REP(\boldsymbol \theta_N)} \log\left[ \frac{P_{\boldsymbol \theta_N }(\boldsymbol x) }{\min_{\boldsymbol \theta_N}}\right]  \rightarrow 0 \qquad \frac{1}{\REP(-\boldsymbol \theta_N)}\log\left[ \frac{P_{-\boldsymbol \theta_N }(\boldsymbol x) }{\min_{-\boldsymbol \theta_N}}\right]  \rightarrow 1
+\(\boldsymbol X = (X_1,\ldots,X_N)\), whereby
+\(P_{\boldsymbol \theta_N }(\boldsymbol{x}_{\max, \boldsymbol \theta_N}) = \max_{\boldsymbol y\in\mathcal{X}^N} P_{\boldsymbol \theta_N }(\boldsymbol y)\)
+and
+\(P_{\boldsymbol \theta_N }(\boldsymbol{x}_{\min, \boldsymbol \theta_N}) = \min_{\boldsymbol y\in\mathcal{X}^N} P_{\boldsymbol \theta_N }(\boldsymbol y)\).
+
+Then, letting \(\stackrel{p, \mathrm{E}}{\longrightarrow}\) denote
+convergence in probability and expectation, as \(N\to \infty\), \[
+\frac{1}{\REP(\boldsymbol \theta_N)} \log\left[ \frac{P_{\boldsymbol \theta_N }(\boldsymbol X) }{ \displaystyle{\min_{\boldsymbol y\in\mathcal{X}^N}} P_{\boldsymbol \theta_N }(\boldsymbol y)}\right] = \frac{ \log P_{\boldsymbol \theta_N } ( \boldsymbol X)  - \log P_{\boldsymbol \theta_N }(\boldsymbol{x}_{\min, \boldsymbol \theta_N}) }
+{ \log P_{\boldsymbol \theta_N } (\boldsymbol{x}_{\max, \boldsymbol \theta_N})  - \log P_{\boldsymbol \theta_N }(\boldsymbol{x}_{\min, \boldsymbol \theta_N}) }
+\stackrel{p, \mathrm{E}}{\longrightarrow}
+\; 1
+\] under \(\boldsymbol\theta_N\) while \[
+\frac{1}{\REP(-\boldsymbol \theta_N)} \log\left[ \frac{P_{-\boldsymbol \theta_N }(\boldsymbol X) }{  \displaystyle{\min_{\boldsymbol y\in\mathcal{X}^N}} P_{-\boldsymbol \theta_N }(\boldsymbol y)} \right]
+=    \frac{ \log P_{-\boldsymbol \theta_N } ( \boldsymbol X) - \log P_{ \boldsymbol \theta_N }(\boldsymbol{x}_{\max, \boldsymbol \theta_N})    }
+{ \log P_{-\boldsymbol \theta_N } (\boldsymbol{x}_{\min, \boldsymbol \theta_N})  - \log P_{-\boldsymbol \theta_N }(\boldsymbol{x}_{\max, \boldsymbol \theta_N}) }
+\stackrel{p,\mathrm{E}}{\longrightarrow}
+\; 1,
+\] under \(-\boldsymbol\theta_N\), where \[
+\REP( \boldsymbol \theta_N) \equiv \log \frac{P_{\boldsymbol \theta_N } (\boldsymbol{x}_{\max, \boldsymbol \theta_N})}{P_{\boldsymbol \theta_N }(\boldsymbol{x}_{\min, \boldsymbol \theta_N})} = \log  \frac{P_{-\boldsymbol \theta_N } (\boldsymbol{x}_{\min, \boldsymbol \theta_N}) }{ P_{-\boldsymbol \theta_N }(\boldsymbol{x}_{\max, \boldsymbol \theta_N})}=   \REP( -\boldsymbol \theta_N), \quad N \geq 1.
 \]
-\EndKnitrBlock{theorem} Schweinberger
+\EndKnitrBlock{theorem}
+
+Theorem \ref{thm:something} entails log-likelihood functions based on
+unstable models are both inversely related and degenerate at opposited
+signed parameters \(\boldsymbol \theta_N\) or \(-\boldsymbol \theta_N\),
+so that likelihoods are highest at different extremes in data
+configuration (e.g., \(\boldsymbol{x}_{\max, \boldsymbol \theta_N}\)
+under \(\boldsymbol \theta\)-probabilities or
+\(\boldsymbol{x}_{\min, \boldsymbol \theta_N}\) under
+\(-\boldsymbol \theta\)-probabilities). If the observed outcome
+\(\boldsymbol x\) for data \(\boldsymbol X\) is not a mode/anti-mode,
+then probabilities for the outcome may be small under both parameters
+\(\boldsymbol \theta_N\) and \(-\boldsymbol \theta_N\), in which case
+associated optimization steps may then shift around zero and struggle to
+converge. In many model formulations, the zero parameter
+\(\boldsymbol \theta_N=\boldsymbol 0\) is a ``safe'' position among
+parameters, representing a guaranteed stable model (having uniform
+probabilities among outcomes), which can also tether a broad parameter
+search attempted among unstable models. Handcock
+(\protect\hyperlink{ref-handcock2003assessing}{2003}) describes similar
+results for degenerate exponential models, and Theorem
+\ref{thm:something} also supports an important finding of Schweinberger
 (\protect\hyperlink{ref-schweinberger2011instability}{2011} Corollary 1)
-considered the one-parameter linear exponential case showing that, for
-positive \(\boldsymbol \theta_N>0\), \[
-\frac{1}{\REP(\boldsymbol \theta_N)} \log\left[ \frac{P_{\boldsymbol \theta_N }(\boldsymbol x) }{\min_{\boldsymbol \theta_N}}\right] = \frac{\boldsymbol g(\boldsymbol x) - L_N}{U_N-L_N} 
-\] holds and converges to \(1\) in expectation
-\(E_{\boldsymbol \theta_N}\), while \[   
-\frac{1}{\REP(-\boldsymbol \theta_N)} \log\left[ \frac{P_{-\boldsymbol \theta_N }(\boldsymbol x) }{\min_{-\boldsymbol \theta_N}}\right] = \frac{U_N -\boldsymbol g(\boldsymbol x)}{U_N-L_N} 
-\] holds converges to \(1\) in expectation
-\(E_{-\boldsymbol \theta_N}\). This result is a special case of Theorem
-\ref{thm:something} and the LREP expansion \eqref{eq:UL} for the one
-parameter exponential model case and is related to a result in Handcock
-(\protect\hyperlink{ref-handcock2003assessing}{2003}).
-
-The idea, more generally, is that probability in an unstable model is
-pushed to opposite regions of the sample space due to degeneracy and
-opposite signed parameters. This can cause numerical complications in
-likelihood maximization. If an observed outcome
-\(\boldsymbol x \equiv \argmax \mathcal{L}(\boldsymbol \theta_N)\) which
-is also the outcome with minimal probability under
-\(P_{-\boldsymbol \theta_N}\), then the likelihood function will have a
-clear peak and maximization straightforward. However, when the observed
-outcome \(\boldsymbol x\) is at an extreme in the probability structures
-of \(P_{\boldsymbol \theta_N}, P_{-\boldsymbol \theta_N}\) (which are
-related) the likelihood function will be flat and optimization steps may
-shift around zero.
-
-\subsection{Implications for Bayes
-Inference}\label{implications-for-bayes-inference}
-
-Considering that changes in sign of parameters can cause all probability
-to be placed on disjoint sections of the sample space (Corollary
-\ref{cor:sign}), then this will create numerical difficulties in Bayes
-inference based on MCMC. Given observations
-\(\boldsymbol X = (X_1,\ldots,X_N)\), switching \(\boldsymbol theta_N\)
-from positive to negative or vice versa could dramatically change
-probabilities (model condition PSR and Corollary \ref{cor:sign}). This
-would translate to numerical challenges similar to those presented in
-Section \ref{mcmc}, i.e., inability of the chain to effectively explore
-the sample space because of unbounded ratios of probabilities. We can
-write the MH acceptance probability to indicate this, \[
-\alpha\left(\boldsymbol x^{(1)} \mid \boldsymbol x^{(2)}\right)= \min\left\{1, 
-\frac{q(\boldsymbol x^{(2)} \mid \boldsymbol x^{(1)})}{q(\boldsymbol x^{(1)} \mid \boldsymbol x^{(2)})}\frac{P_{\boldsymbol \theta_N} ( \boldsymbol x^{(1)} )}{P_{\boldsymbol \theta_N} (  \boldsymbol x^{(2)} ) } \right\}
-\] for some proposal distribution \(q(\cdot)\).
+for one-parameter linear exponential models \eqref{eq:expo}. In the latter
+case, the likelihood score function at \(\boldsymbol \theta_N\) is the
+expected value
+\(\mu(\boldsymbol \theta_N)\equiv \mathrm{E}_{\boldsymbol \theta_N} g(\boldsymbol X)\)
+of the sufficient statistic \(g(\cdot)\), and optimization involves
+solving \(\mu(\cdot)=g(\boldsymbol x)\) for an observed outcome
+\(\boldsymbol x\). For unstable models in this exponential class,
+Schweinberger
+(\protect\hyperlink{ref-schweinberger2011instability}{2011} Corollary 1)
+shows that \[
+\lim_{N\to \infty}\frac{\mu(\boldsymbol \theta_N) -L_n}{U_n-L_N}= \left\{ \begin{array}{lcl}
+1 && \text{for } \boldsymbol \theta_N>0,\\
+0 && \text{for } \boldsymbol \theta_N<0, \end{array}\right.
+\] where again \(U_N\) and \(L_N\) denote the maximum and minimum values
+of the statistic \(g(\boldsymbol x)\),
+\(\boldsymbol x\in\mathcal{X}^N\). As described by Schweinberger
+(\protect\hyperlink{ref-schweinberger2011instability}{2011}), the
+implication for maximum likelihood estimation is that, unless an
+observed outcome \(\boldsymbol x\) falls at an extreme \(U_N, L_N\)
+(i.e., modes/anti-modes), optimization steps in the parameter space can
+iterate in relatively small increments around zero and fail to converge.
+For unstable one-parameter exponential models, the maximum likelihood
+results of Schweinberger
+(\protect\hyperlink{ref-schweinberger2011instability}{2011}) turn out to
+be a special case of Theorem \ref{thm:something} and the LREP expansion
+\eqref{eq:UL} in this setting; namely, for an unstable model with
+\(\boldsymbol \theta_N>0\), \[
+\frac{1}{\REP(\boldsymbol \theta_N)} \log\left[ \frac{P_{\boldsymbol \theta_N }(\boldsymbol X) }{\displaystyle{\min_{\boldsymbol y\in\mathcal{X}^N}} P_{\boldsymbol \theta_N }(\boldsymbol y)}\right] = \frac{\boldsymbol g(\boldsymbol X) - L_N}{U_N-L_N}\;\;\stackrel{p, \mathrm{E}}{\longrightarrow}\;\; 1
+\] holds as \(N\to \infty\) by Theorem \ref{thm:something}, while under
+\(-\boldsymbol \theta_N<0\) \[
+\frac{1}{\REP(-\boldsymbol \theta_N)} \log\left[ \frac{P_{-\boldsymbol \theta_N }(\boldsymbol X) }{\displaystyle{\min_{\boldsymbol y\in\mathcal{X}^N}} P_{-\boldsymbol \theta_N }(\boldsymbol y)}\right] = \frac{U_N -\boldsymbol g(\boldsymbol X)}{U_N-L_N} = 1 - \frac{\boldsymbol g(\boldsymbol X) - L_N}{U_N-L_N} \;\;\stackrel{p, \mathrm{E}}{\longrightarrow}\;\; 1.
+\] Again, when all probability in unstable models may be pushed to
+opposite extremes in the sample space, due to a combination of
+degeneracy and parameter sign, numerical complications in likelihood
+maximization may occur.
+
+\subsection{Implications for Bayes Inference}\label{bayes}
+
+The potential numerical difficulties with maximum likelihood with
+unstable models, as described in the previous section, can naturally
+carry over to Bayes inference. Considering that the degeneracy issues
+related to unstable models can cause likelihoods can be flat (e.g., near
+zero) for many parameters under a given data outcome and that sign
+changes in parameters can shift tremendous probability to extreme and
+opposite outcomes in the sample space (e.g., Corollary \ref{cor:sign},
+Theorem \ref{thm:something}), then numerical complications may arise
+with Bayes inference in sampling a posterior parameter space based on
+MCMC. The potential challenges in chain mixing are similar to those
+presented in Section \ref{mcmc}, though in chain movements through the
+parameter space as opposed to the sample space in data generation. That
+is, in the Bayes setting for sampling a posterior distribution for
+\(\boldsymbol \theta_N\), a chain may unstable to effectively explore
+the parameter space due partly to extreme and potentially unbounded
+probability ratios from parameter sign changes, which represents a
+parameter space analog to how one component changes in the sample space
+may impact data simulation with unstable models. For example, if
+\(\pi(\cdot)\) denotes a prior density for \(\boldsymbol \theta_N\) and
+\(q(\cdot | \cdot)\) denotes a proposal distribution for use in a
+Metropolis-Hastings (MH) sampler, then MH acceptance probability becomes
+\[
+\alpha\left(\boldsymbol \theta^{(1)} \mid \boldsymbol \theta^{(2)}\right)= \min\left\{1,
+\frac{q(\boldsymbol \theta_N^{(2)} \mid \boldsymbol \theta_N^{(1)})}{q(\boldsymbol \theta_N^{(1)} \mid \boldsymbol \theta_N^{(2)})}
+\frac{P_{\boldsymbol \theta_N^{(1)}} ( \boldsymbol x ) \pi(\boldsymbol \theta_N^{(1)}) }{P_{\boldsymbol \theta^{(2)}_N} (  \boldsymbol x) \pi(\boldsymbol \theta_N^{(2)}) } \right\},
+\] which indicates how parameter sensitivity in the likelihood
+\(P_{\boldsymbol \theta_N} ( \boldsymbol x)\) may complicate sampling of
+the posterior
+\(P_{\boldsymbol \theta_N} ( \boldsymbol x) \pi(\boldsymbol \theta_N)\)
+(i.e., moving from \(\boldsymbol \theta^{(1)}\) to
+\(\boldsymbol \theta^{(2)}\) in the parameter space). Furthermore, the
+potential for model instability and the size of the parameter space can
+also become greater with the introduction of latent variables to
+existing data variables, as involved in some model formulations
+described in Sections \ref{rbm}-\ref{deep-learning}. As latent variables
+are often sampled with parameters in a Bayes MCMC approach, this aspect
+may further compound numerical problems in chain mixing.
 

### 2017-11-06 writing/paper-note/docs/note_blinded.tex
ready?

@@
 sensitivities, as outlined in the following extension of Theorem
-@ref(thm:degenFOES\}).
+\ref{thm:degenFOES}.
 
@@
 \(\lim_{N\to \infty} \max_{\boldsymbol x \in \mathcal{X}^N} |G_N(\boldsymbol x, -\boldsymbol \theta_{N}) - G_N(\boldsymbol x, \boldsymbol \theta_{N}^A)|=0\)
-holds in addition to Corollary\textasciitilde{}1 assumptions, where \[
+holds in addition to Corollary \ref{cor:sign} assumptions, where \[
 G_N(\boldsymbol x, \boldsymbol \theta) =   \frac{\log P_{\boldsymbol \theta}(\boldsymbol x) - \min_{\boldsymbol y \in\mathcal{X}^N}\log P_{\boldsymbol \theta}(\boldsymbol y) }{\max_{\boldsymbol y \in\mathcal{X}^N}\log P_{\boldsymbol \theta}(\boldsymbol y)  - \min_{\boldsymbol y \in\mathcal{X}^N}\log P_{\boldsymbol \theta}(\boldsymbol y) }, \quad \boldsymbol x \in\mathcal{X}^N,
@@
 function. We next provide a discussion of this issue in a way that
-builds upon and extends related findings by @Schweinberger
+builds upon and extends related findings by Schweinberger
 (\protect\hyperlink{ref-schweinberger2011instability}{2011}), who

### 2017-11-06 writing/paper-note/index.Rmd
dr. nordman's changes
ready?

--- 2017-07-08 writing/paper-note/index.Rmd
@@
   A probability model exhibits instability if small changes in a data outcome result in large, and often unanticipated, changes in probability.
+This instability is a property of the probability model, rather than the fitted parameter vector.
 For correlated data structures found in several application areas, there is increasing interest in predicting/identifying such sensitivity in model probability structure.
-We consider the problem of quantifying instability for general probability models defined on sequences of observations, where each sequence of length $N$ has a finite number of possible outcomes.
+We consider the problem of quantifying instability for general probability models defined on sequences of observations, where each sequence of length $N$ has a finite number of possible values.
 A sequence of probability models results, indexed by $N$, that accommodates data of expanding dimension.
@@
 library(tidyr)
+library(purrr)
 library(scales)
-library(mvtnorm)
-library(rootSolve)
-library(agridat)
 library(xtable)

### 2017-11-06 writing/resources/code/functions.R

@@ -1,13 +1,13 @@
 ##function for creating all possible values of the t(x)
 stats <- function(H, V, type="negative") {
-  if(type == "binary") t <- data.frame(c(0,1))[,rep(1, H + V)]
-  else if(type == "negative") t <- data.frame(c(-1,1))[,rep(1, H + V)]
+  if(type == "binary") ti <- data.frame(c(0,1))[,rep(1, H + V)]
+  else if(type == "negative") ti <- data.frame(c(-1,1))[,rep(1, H + V)]
   else stop("Type must be 'binary', 'negative'")
   
-  names(t) <- c(paste0("v", 1:H), paste0("h", 1:V))  
+  names(ti) <- c(paste0("v", 1:V), paste0("h", 1:H))  
   
   #all possibilities of statistics in (H + V)-space
-  t.grid <- expand.grid(t)
+  t.grid <- expand.grid(ti)
   
   for(i in 1:V){
     for(j in (V+1):(H+V)){

### 2017-11-07 writing/paper-note/chapters/instab.Rmd
Vardeman's edits, submitted to tech

@@
 At issue, models exhibiting instability are typically undesirable as these tend to provide poor representations of data or data-generation.
-As an example, such models can include near-degenerate distributions that assign all essential probability mass to only a subset of an overall sample space.
+As an example, such models can include near-degenerate distributions that assign essentially all probability mass to only a subset of an overall sample space.
 The latter issue in connection to degeneracy has been recognized as a concern in that dominant model outcomes may not resemble observed data [cf.
@@
 We prescribe a general and simple condition for identifying instability in a FOES model sequence, which  quantifies whether certain maximal probabilities in a FOES model  are too extreme relative to the sample size $N$.
-When fulfilled, the probability structure of a FOES model is shown to exhibit extreme sensitivity, with probability assignments possessing extreme peaks and troughs across nearly identical outcomes.
+When these conditions are met, the probability structure of a FOES model is shown to exhibit extreme sensitivity, with probability assignments possessing extreme peaks and troughs across nearly identical outcomes.
 As the measure of model instability increases, probabilities from an unstable FOES model additionally increase in volatility and provably slide into degeneracy.
 Section \@ref(implications) then emphasizes the implications of such model instability, showing that such impropriety can be expected to numerically hinder maximum likelihood estimation and MCMC-based simulations.
-As one potential remedy, suggestions are given or constraining model parameterizations to avoid the most problematic regions of the parameter space.
+As one potential remedy, suggestions are given for constraining model parameterizations to avoid the most problematic regions of the parameter space.
 Proofs of the main results appear in Appendix \@ref(appendix-instab).
@@
 $$
-which represents the biggest log-probability ratio for a one component change in data outcomes in a FOES model with parameter $\boldsymbol \theta_N$.
+which represents the biggest log-probability ratio for a one-component change in data outcomes in a FOES model with parameter $\boldsymbol \theta_N$.
 We then have the following result prescribing the behavior of $\DN(\boldsymbol \theta_N)$ for S-unstable FOES models.
@@
 \begin{enumerate}[(i)]
-\item For any integer $N \geq 1$ and any given $C>0$, if $\REP(\boldsymbol \theta_N)/N > C$ in (\ref{eq:elpr}, then
+\item For any integer $N \geq 1$ and any given $C>0$, if $\REP(\boldsymbol \theta_N)/N > C$ in (\ref{eq:elpr}), then
     $$ \DN(\boldsymbol \theta_N) > C,$$
-    or probabilities from a one component change in some outcome have log-ratio exceeding $C$.
+    or probabilities from a one-component change in some outcome have log-ratio exceeding $C$.
 \item Suppose the FOES model sequence is S-unstable.
@@
 S-unstable models can never have universally bounded changes in probability among single component variations in data configurations.
-While not all one component changes in data may produce massive changes in probability, unstable models must have some such data outcomes with this property.
+While not all one-component changes in data may produce massive changes in probability, unstable models must have some such data outcomes with this property.
 As a consequence, unstable probability structures may exhibit extreme sensitivity through large peaks and troughs over the sample space.
@@
 Let $P_{\boldsymbol \theta_N}$, $N \geq 1$,  denote linear exponential models (\ref{eq:expo}) with   parameters $\boldsymbol \theta_N = (\theta_{1,N},\ldots,\theta_{k,N})^\prime \in \mathbb{R}^k$ and statistics  $\boldsymbol g_N(\boldsymbol x) = (g_{1,N}(\boldsymbol x),\ldots, g_{k,N}(\boldsymbol x))^\prime \in \mathbb{R}^k$, for fixed $k \geq 1$.
-Then, the models $P_{\boldsymbol \theta_N}$ are {\rm S-stable}   if
+Then, the models $P_{\boldsymbol \theta_N}$ are S-stable if
 \begin{equation}
@@
 For clarity, recall also that S-instability \@ref(eq:Sun) in each model type refers to a respective divergence (i.e., $\lim_{N\to \infty} \REP_{(\boldsymbol X, \boldsymbol H)}(\boldsymbol \theta_N) /(N+N_{\mathcal{H}})=\infty$, $\lim_{N\to \infty} \REP(\boldsymbol \theta_N) /N=\infty$) upon scaling by the corresponding number of variables in a distribution.
-In the following, let $|\boldsymbol y|_1 = \sum_{i=1}^d |y_i|$ denote the L-1 norm of a generic vector $\boldsymbol y =(y_1,\ldots,y_d)$, $d \geq 1$.
+In the following, let $|\boldsymbol y|_1 = \sum_{i=1}^d |y_i|$ denote the L1 norm of a generic vector $\boldsymbol y =(y_1,\ldots,y_d)$, $d \geq 1$.
 ```{proposition, label="prop2", echo=TRUE}
@@
   geom_abline(aes(intercept = 0, slope = 1), alpha = .5, lty = 2) +
-  scale_fill_gradient(expression(paste("Mean ", frac(ELPR(theta), N[italic(V)]))), low = "yellow", high = "red") +
+  scale_fill_gradient(expression(paste("Mean ", frac(ELPR(theta), N[italic(V)]))), low = "white", high = "grey20") +
   xlab(expression(paste(paste("||", theta[main], "||", "/(", N[italic(H)]," + ", N[italic(V)], ")")))) +
@@
   geom_abline(aes(intercept = 0, slope = 1), alpha = .5, lty = 2) +
-  scale_fill_gradient(expression(paste("Mean ", Delta[N](theta))), low = "yellow", high = "red") +
+  scale_fill_gradient(expression(paste("Mean ", Delta[N](theta))), low = "white", high = "grey20") +
   xlab(expression(paste(paste("||", theta[main], "||", "/(", N[italic(H)]," + ", N[italic(V)], ")")))) +
@@
 In our numerical experiment, we allow the two types of terms (main effects terms corresponding to visible and hidden parameters $\boldsymbol \theta_{main} = (\boldsymbol \theta_N^{\mathcal{V}}, \boldsymbol \theta_N^{\mathcal{H}})$ and interaction parameters $\boldsymbol \theta_N^{\mathcal{VH}}$) to have varying average magnitudes, $||\boldsymbol \theta_{main} || /(N_{\mathcal{H}}+N_{\mathcal{V}})$ and $||\boldsymbol \theta_{interaction} || /(N_{\mathcal{H}}*N_{\mathcal{V}})$ for a RBM with $N_\mathcal{V} = `r V`$ visibles and $N_\mathcal{H} = `r H`$ hiddens.
-These average magnitudes vary on a grid between `r min(elpr_summary$r1/(elpr_summary$H + elpr_summary$V))` and `r max(elpr_summary$r1/(elpr_summary$H + elpr_summary$V))` with `r length(unique(elpr_summary$r1/(elpr_summary$H + elpr_summary$V)))` breaks, yielding `r length(unique(elpr_summary$r1/(elpr_summary$H + elpr_summary$V)))^2` grid points.
-At each point in the grid, `r n` vectors ($\boldsymbol \theta_{main}$) are sampled uniformly on a sphere with radius corresponding to the first coordinate in the grid and `r n` vectors ($\boldsymbol \theta_{interaction}$) are sampled uniformly on a sphere with radius corresponding to the second coordinate in the grid via sums of squared and scaled iid Normal$(0, 1)$ variables.
-These vectors are then paired to create `r n` values of $\boldsymbol \theta_N$ with magnitudes at each point in the grid.
+These average magnitudes vary on a grid between $`r min(elpr_summary$r1/(elpr_summary$H + elpr_summary$V))`$ and $`r max(elpr_summary$r1/(elpr_summary$H + elpr_summary$V))`$ with $`r length(unique(elpr_summary$r1/(elpr_summary$H + elpr_summary$V)))`$ breaks, yielding $`r length(unique(elpr_summary$r1/(elpr_summary$H + elpr_summary$V)))^2`$ grid points.
+At each point in the grid, $`r n`$ vectors ($\boldsymbol \theta_{main}$) are sampled uniformly on a sphere with radius corresponding to the first coordinate in the grid and $`r n`$ vectors ($\boldsymbol \theta_{interaction}$) are sampled uniformly on a sphere with radius corresponding to the second coordinate in the grid via sums of squared and scaled iid Normal$(0, 1)$ variables.
+These vectors are then paired to create $`r n`$ values of $\boldsymbol \theta_N$ with magnitudes at each point in the grid.
 The values $\REP(\boldsymbol \theta_N)/N_{\mathcal{V}}$ and $\DN(\boldsymbol \theta_N)$ are then calculated for each $\boldsymbol \theta_N$ and then summarized for each point in the grid using the sample mean.
 The results of this numerical study are shown in Figure \@ref(fig:rbm-plots).
-From these two plots, it is clear that for larger magnitudes of the parameter vectors, there is evidence of S-instability in that the log-ratio of extremal probabilities scaled by $N_{\mathcal{V}}$ and the the biggest log-probability ratio for a one component change in data outcomes are both increasing away from $\boldsymbol \theta_N = \boldsymbol 0$, further supporting \@ref(prp:prop2)(iii.2 and iii.5).
+From these two plots, it is clear that for larger magnitudes of the parameter vectors, there is evidence of S-instability in that the log-ratio of extremal probabilities scaled by $N_{\mathcal{V}}$ and the the biggest log-probability ratio for a one-component change in data outcomes are both increasing away from $\boldsymbol \theta_N = \boldsymbol 0$, further supporting \@ref(prp:prop2)(iii.2 and iii.5).
 
@@
 
-Due to their extreme sensitivity in probability structure, S-instability in FOES models may often translate to numerical complications, and in fact obstructions, in both simulation and statistical inference based on likelihoods.
+Due to their induced sensitivity in probability structure, S-instability in FOES models may often translate to numerical complications, and in fact obstructions, in both simulation and statistical inference based on likelihoods.
 We describe these aspects in Sections \@ref(mcmc)-\@ref(bayes) with regard to data simulation, maximum likelihood estimation and Bayes inference, respectively.
@@
 Suppose one aims to apply MCMC to simulate data $\boldsymbol X=(X_1,\ldots,X_n)$ from a FOES model $P_{\boldsymbol \theta_N}(\boldsymbol x)$, $\boldsymbol x \in\mathcal{X}^N$ whereby a chain is constructed with $P_{\boldsymbol \theta_N}$ as the stationary distribution.
-For an unstable FOES model, one component changes in outcomes may produce
+For an unstable FOES model, one-component changes in outcomes may produce
 radically different probabilities, which may then impose numerical barriers to MCMC.
@@
 \@ref(thm:instab-elpr)).
-This can hinder the ability of chain to effectively explore the sample space of the observations $\boldsymbol X$, as the chain may mix poorly by moving rapidly to, and slowly away from, sections of the sample space.
+This can hinder the ability of a chain to effectively explore the sample space of the observations $\boldsymbol X$, as the chain may mix poorly by moving rapidly to, and slowly away from, sections of the sample space.
 In this case, for example, the Markov chain may become entrapped within a mode of the probability function, with rare chance  of escaping to adequately mimic the occupation frequencies in the overall sample space.
@@
 If the observed outcome $\boldsymbol x$ for data $\boldsymbol X$ is not a mode/anti-mode, then probabilities for the outcome may be small under both parameters $\boldsymbol \theta_N$ and $-\boldsymbol \theta_N$, in which case associated optimization steps may then shift around zero and struggle to converge.
+
 In many model formulations, the zero parameter $\boldsymbol \theta_N=\boldsymbol 0$ is a "safe" position among parameters, representing a  guaranteed stable model (having uniform probabilities among outcomes), which can also  tether a broad parameter search attempted among unstable models.
@@
 The potential challenges in chain mixing are similar to those presented in Section \@ref(mcmc), though in chain movements through the parameter space as opposed to the sample space in data generation.
-That is, in the Bayes setting for sampling a posterior distribution for $\boldsymbol \theta_N$,  a chain may  unstable to effectively explore the parameter space due partly to extreme and potentially unbounded probability ratios from parameter sign changes, which represents a parameter space analog to how one component changes in the sample space may impact data simulation with unstable models.
+That is, in the Bayes setting for sampling a posterior distribution for $\boldsymbol \theta_N$,  a chain may  unstable to effectively explore the parameter space due partly to extreme and potentially unbounded probability ratios from parameter sign changes, which represents a parameter space analog to how one-component changes in the sample space may impact data simulation with unstable models.
 For example, if $\pi(\cdot)$ denotes a prior density for $\boldsymbol \theta_N$ and $q(\cdot | \cdot)$  denotes a proposal distribution  for use in a Metropolis-Hastings (MH) sampler, then MH acceptance probability becomes

### 2017-11-07 writing/paper-note/docs/note.tex
Vardeman's edits, submitted to tech

@@
             pdfauthor={},
-            pdftitle={A note on the instability and degeneracy of deep learning models},
+            pdftitle={On the instability and degeneracy of deep learning models},
             colorlinks=true,
@@
 
-\title{\bf A note on the instability and degeneracy of deep learning models}
+\title{\bf On the instability and degeneracy of deep learning models}
 \author{Andee Kaplan \\ Department of Statistics, Iowa State University \\  and \\ Daniel Nordman \\ Department of Statistics, Iowa State University \\  and \\ Stephen Vardeman \\ Department of Statistics and Department of Industrial and\\
@@
 of data or data-generation. As an example, such models can include
-near-degenerate distributions that assign all essential probability mass
-to only a subset of an overall sample space. The latter issue in
+near-degenerate distributions that assign essentially all probability
+mass to only a subset of an overall sample space. The latter issue in
 connection to degeneracy has been recognized as a concern in that
@@
 whether certain maximal probabilities in a FOES model are too extreme
-relative to the sample size \(N\). When fulfilled, the probability
-structure of a FOES model is shown to exhibit extreme sensitivity, with
-probability assignments possessing extreme peaks and troughs across
-nearly identical outcomes. As the measure of model instability
-increases, probabilities from an unstable FOES model additionally
-increase in volatility and provably slide into degeneracy. Section
-\ref{implications} then emphasizes the implications of such model
-instability, showing that such impropriety can be expected to
+relative to the sample size \(N\). When these conditions are met, the
+probability structure of a FOES model is shown to exhibit extreme
+sensitivity, with probability assignments possessing extreme peaks and
+troughs across nearly identical outcomes. As the measure of model
+instability increases, probabilities from an unstable FOES model
+additionally increase in volatility and provably slide into degeneracy.
+Section \ref{implications} then emphasizes the implications of such
+model instability, showing that such impropriety can be expected to
 numerically hinder maximum likelihood estimation and MCMC-based
-simulations. As one potential remedy, suggestions are given or
+simulations. As one potential remedy, suggestions are given for
 constraining model parameterizations to avoid the most problematic
@@
 \DN(\boldsymbol \theta_N) \equiv \max \left\{\log \frac{P_{\boldsymbol \theta_N}(\boldsymbol x)}{P_{\boldsymbol \theta_N}(\boldsymbol x^*)} : \boldsymbol x \text{ }\& \text{ } \boldsymbol x^* \in \mathcal{X}^N \text{ differ in exactly one component}\right\},
-\] which represents the biggest log-probability ratio for a one
-component change in data outcomes in a FOES model with parameter
+\] which represents the biggest log-probability ratio for a
+one-component change in data outcomes in a FOES model with parameter
 \(\boldsymbol \theta_N\). We then have the following result prescribing
@@
 \BeginKnitrBlock{theorem}
-\protect\hypertarget{thm:instab-elpr}{}{\label{thm:instab-elpr}}Let
+\protect\hypertarget{thm:instab-elpr}{}{\label{thm:instab-elpr} }Let
 \(P_{\boldsymbol \theta_N}\), with support \(\mathcal{X}^N\),
@@
 universally bounded changes in probability among single component
-variations in data configurations. While not all one component changes
+variations in data configurations. While not all one-component changes
 in data may produce massive changes in probability, unstable models must
@@
 \BeginKnitrBlock{theorem}
-\protect\hypertarget{thm:degenFOES}{}{\label{thm:degenFOES}}For any
+\protect\hypertarget{thm:degenFOES}{}{\label{thm:degenFOES} }For any
 arbitrarily small \(0 < \epsilon < 1\), an S-unstable FOES model
@@
 \BeginKnitrBlock{proposition}
-\protect\hypertarget{prp:prop1}{}{\label{prp:prop1}}Let
+\protect\hypertarget{prp:prop1}{}{\label{prp:prop1} }Let
 \(P_{\boldsymbol \theta_N}\), \(N \geq 1\), denote linear exponential
@@
 for fixed \(k \geq 1\). Then, the models \(P_{\boldsymbol \theta_N}\)
-are \{\rm S-stable\} if
+are S-stable if
 \begin{equation}
@@
 \BeginKnitrBlock{remark}
-\iffalse <span class="remark"><em>Remark. \fi{} In the one-parameter
-exponential case \(k=1\), recall the exponential model is
-stable/unstable depending on whether
+\iffalse{} {Remark. } \fi{}In the one-parameter exponential case
+\(k=1\), recall the exponential model is stable/unstable depending on
+whether
 \(\REP(\boldsymbol \theta_N)/N = |\theta_{1,N}|(U_{1,N}-L_{1,N})/N \equiv |\boldsymbol \theta_{N}|(U_{N}-L_{N})/N\)
@@
 the following, let \(|\boldsymbol y|_1 = \sum_{i=1}^d |y_i|\) denote the
-L-1 norm of a generic vector \(\boldsymbol y =(y_1,\ldots,y_d)\),
+L1 norm of a generic vector \(\boldsymbol y =(y_1,\ldots,y_d)\),
 \(d \geq 1\). \BeginKnitrBlock{proposition}
@@
 \BeginKnitrBlock{remark}
-\iffalse <span class="remark"><em>Remark. \fi{} The condition
+\iffalse{} {Remark. } \fi{}The condition
 \(\sup_{N \geq 1} N_{\mathcal{H}}/N<\infty\) in Proposition
@@
 for a RBM with \(N_\mathcal{V} = 9\) visibles and \(N_\mathcal{H} = 5\)
-hiddens. These average magnitudes vary on a grid between 0.001 and 3
-with 20 breaks, yielding 400 grid points. At each point in the grid, 100
-vectors (\(\boldsymbol \theta_{main}\)) are sampled uniformly on a
-sphere with radius corresponding to the first coordinate in the grid and
-100 vectors (\(\boldsymbol \theta_{interaction}\)) are sampled uniformly
-on a sphere with radius corresponding to the second coordinate in the
-grid via sums of squared and scaled iid Normal\((0, 1)\) variables.
-These vectors are then paired to create 100 values of
-\(\boldsymbol \theta_N\) with magnitudes at each point in the grid. The
-values \(\REP(\boldsymbol \theta_N)/N_{\mathcal{V}}\) and
-\(\DN(\boldsymbol \theta_N)\) are then calculated for each
+hiddens. These average magnitudes vary on a grid between \(0.001\) and
+\(3\) with \(20\) breaks, yielding \(400\) grid points. At each point in
+the grid, \(100\) vectors (\(\boldsymbol \theta_{main}\)) are sampled
+uniformly on a sphere with radius corresponding to the first coordinate
+in the grid and \(100\) vectors (\(\boldsymbol \theta_{interaction}\))
+are sampled uniformly on a sphere with radius corresponding to the
+second coordinate in the grid via sums of squared and scaled iid
+Normal\((0, 1)\) variables. These vectors are then paired to create
+\(100\) values of \(\boldsymbol \theta_N\) with magnitudes at each point
+in the grid. The values \(\REP(\boldsymbol \theta_N)/N_{\mathcal{V}}\)
+and \(\DN(\boldsymbol \theta_N)\) are then calculated for each
 \(\boldsymbol \theta_N\) and then summarized for each point in the grid
@@
 S-instability in that the log-ratio of extremal probabilities scaled by
-\(N_{\mathcal{V}}\) and the the biggest log-probability ratio for a one
-component change in data outcomes are both increasing away from
+\(N_{\mathcal{V}}\) and the the biggest log-probability ratio for a
+one-component change in data outcomes are both increasing away from
 \(\boldsymbol \theta_N = \boldsymbol 0\), further supporting
@@
 
-Due to their extreme sensitivity in probability structure, S-instability
+Due to their induced sensitivity in probability structure, S-instability
 in FOES models may often translate to numerical complications, and in
@@
 \(P_{\boldsymbol \theta_N}\) as the stationary distribution. For an
-unstable FOES model, one component changes in outcomes may produce
+unstable FOES model, one-component changes in outcomes may produce
 radically different probabilities, which may then impose numerical
@@
 one-component changes (Theorem \ref{thm:instab-elpr}). This can hinder
-the ability of chain to effectively explore the sample space of the
+the ability of a chain to effectively explore the sample space of the
 observations \(\boldsymbol X\), as the chain may mix poorly by moving
@@
 associated optimization steps may then shift around zero and struggle to
-converge. In many model formulations, the zero parameter
+converge.
+
+In many model formulations, the zero parameter
 \(\boldsymbol \theta_N=\boldsymbol 0\) is a ``safe'' position among
@@
 probability ratios from parameter sign changes, which represents a
-parameter space analog to how one component changes in the sample space
+parameter space analog to how one-component changes in the sample space
 may impact data simulation with unstable models. For example, if

### 2017-11-07 writing/paper-note/docs/note_blinded.tex
Vardeman's edits, submitted to tech

@@
   %{\LARGE\bf Title}
-  \title{\bf A note on the instability and degeneracy of deep learning models}
+  \title{\bf On the instability and degeneracy of deep learning models}
 %\end{center}

### 2017-11-07 writing/paper-note/index.Rmd
Vardeman's edits, submitted to tech

@@
 ---
-title: "A note on the instability and degeneracy of deep learning models"
+title: "On the instability and degeneracy of deep learning models"
 author: 

### 2017-11-27 writing/paper-note/index.Rmd
getting ready for resubmission

@@
   - name: Andee Kaplan
-    university: Iowa State University
-    department: Department of Statistics
+    university: Duke University
+    department: Department of Statistical Science
   - name: Daniel Nordman

### 2017-11-27 writing/paper-rbm/chapters/rbm.Rmd
getting ready for resubmission

@@
 Occurrences of such behavior have recently been documented in RBMs [@li2014biclustering], as well as other deep architectures [@szegedy2013intriguing; @nguyen2014deep].
-We describe potential issues of model instability, degeneracy and uninterpretability for the RBM class, which are related properties, and examine the presence of these in Section \@ref(explorations-of-model-properties-through-simulation) through simulations of small, manageable examples.
+We describe potential issues of model instability, degeneracy and uninterpretability for the RBM class (which are related properties) and examine the presence of these in Section \@ref(explorations-of-model-properties-through-simulation) through simulations of small, manageable examples.
 
@@
 However, the degree to which regularization and penalization are used by practitioners is not clear because these concepts are not (by default) a part of the accepted fitting methodology [@hinton2002training; @carreira2005contrastive].
-In this paper, we attempt to address the concerns with overfitting and poor mixing in an alternative, and perhaps more transparent manner, via specification of a Bayesian prior in Section \@ref(bayesian-model-fitting).
+In this paper, we attempt to address the concerns with overfitting and poor mixing in an alternative (and perhaps more transparent) manner, via specification of a Bayesian prior in Section \@ref(bayesian-model-fitting).
 
@@
 We examine three Bayesian fitting techniques intended to avoid model impropriety issues raised in Section \@ref(model-fitting) and conclude with a discussion in Section \@ref(discussion).
-Appendix \@ref(appendix-rbm) provides proofs for results on RBM parameterizations and data codings described in Section \@ref(data-encoding).
-
-While applications of the RBM have claimed some successes in classification problems, it is unclear if the model class allows one to go beyond fitting to other statistical matters of importance in using the models, such as quantification of the uncertainty of estimation and the formulation of predictive distributions.
+Supplementary online material provides proofs for results on RBM parameterizations and data codings described in Section \@ref(data-encoding).
+
+While applications of the RBM have been claimed to produce some successes in classification problems, it is unclear if the model class allows one to go beyond fitting to other statistical matters of importance in using the models, such as quantification of the uncertainty of estimation and the formulation of predictive distributions.
 These are closely tied to the probability properties of the RBM model for explaining data generation and the current exposition intends to contribute to a better understanding in this direction.
@@
 When increasing the size of RBM models, it becomes necessary to grow the number of model parameters (and in this process one may also arbitrarily expand the number of hidden variables used).
-To this end, let $\boldsymbol \theta_{\nv} \equiv (\theta_{v_1}, \dots, \theta_{v_\nv}, \theta_{h_1}, \dots, \theta_{h_\nh}, \theta_{11}, \dots, \theta_{\nv \nh}), \nv \ge 1$, denote an element of a sequence of RBM parameters indexed by the number $\nv$ of visibles $(V_1, \dots, V_\nv)$ and define a (scaled) extremal log-probability ratio of the RBM model at $\boldsymbol \theta_{\nv}$ as
+To this end, let $\boldsymbol \theta_{\nv} \equiv (\theta_{v_1}, \dots, \theta_{v_\nv}, \theta_{h_1}, \dots, \theta_{h_\nh}, \theta_{11}, \dots, \theta_{\nv \nh}), \nv \ge 1$, denote an element of a sequence of RBM parameters indexed by the number $\nv$ of visibles $(V_1, \dots, V_\nv)$ and define a log-ratio of extremal
+probabilities (LREP) of the RBM model at $\boldsymbol \theta_{\nv}$ as
 \begin{align}
-\frac{1}{\nv} \log \left[\frac{\max\limits_{(v_1, \dots, v_\nv) \in \mathcal{C}^\nv}P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}{\min\limits_{(v_1, \dots, v_\nv) \in \mathcal{C}^\nv}P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}\right] \equiv \frac{1}{\nv} \text{ELPR}(\boldsymbol \theta_\nv) (\#eq:elpr)
+\frac{1}{\nv} \log \left[\frac{\max\limits_{(v_1, \dots, v_\nv) \in \mathcal{C}^\nv}P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}{\min\limits_{(v_1, \dots, v_\nv) \in \mathcal{C}^\nv}P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}\right] \equiv \frac{1}{\nv} \text{LREP}(\boldsymbol \theta_\nv) (\#eq:elpr)
 \end{align}
@@
 \begin{align*}
-\lim\limits_{\nv \rightarrow \infty} \frac{1}{\nv} \text{ELPR}(\boldsymbol \theta_\nv) = \infty.
+\lim\limits_{\nv \rightarrow \infty} \frac{1}{\nv} \text{LREP}(\boldsymbol \theta_\nv) = \infty.
 \end{align*}
 ```
-In other words, a RBM model sequence is unstable if, given any $c > 0$, there exists an integer $n_c > 0$ so that $\frac{1}{\nv}\text{ELPR}(\boldsymbol \theta_\nv) > c$ for all $\nv \ge n_c$.
+In other words, a RBM model sequence is unstable if, given any $c > 0$, there exists an integer $n_c > 0$ so that $\frac{1}{\nv}\text{LREP}(\boldsymbol \theta_\nv) > c$ for all $\nv \ge n_c$.
 This definition of *S-unstable* is a generalization or re-interpretation of the "unstable" concept of @schweinberger2011instability in that here RBM models for visibles $(v_1, \dots, v_\nv)$ do not form an exponential family and the dimensionality of $\boldsymbol \theta_\nv$ is not fixed, but rather grows with $\nv$.
@@
 ```{proposition, label = "instab1", echo=TRUE}
-Let $c > 0$ and let $\text{ELPR}(\boldsymbol \theta_\nv)$ be as in (\@ref(eq:elpr)) for an integer $\nv \ge 1$.
-If $\frac{1}{\nv}\text{ELPR}(\boldsymbol \theta_\nv) > c$, then $\Delta(\boldsymbol \theta_\nv) > c$.
+Let $c > 0$ and let $\text{LREP}(\boldsymbol \theta_\nv)$ be as in (\@ref(eq:elpr)) for an integer $\nv \ge 1$.
+If $\frac{1}{\nv}\text{LREP}(\boldsymbol \theta_\nv) > c$, then $\Delta(\boldsymbol \theta_\nv) > c$.
 ```
@@
 $$
-M_{\epsilon, \boldsymbol \theta_\nv} \equiv \left\{\boldsymbol v \in \mathcal{C}^\nv: \log P_{\boldsymbol \theta_\nv}(\boldsymbol v) > (1-\epsilon)\max\limits_{\boldsymbol v^*}P_{\boldsymbol \theta_\nv}(\boldsymbol v^*) + \epsilon\min\limits_{\boldsymbol v^*}P_{\boldsymbol \theta_\nv}(\boldsymbol v^*) \right\}
+M_{\epsilon, \boldsymbol \theta_\nv} \equiv \left\{\boldsymbol v \in \mathcal{C}^\nv: \log P_{\boldsymbol \theta_\nv}(\boldsymbol v) > (1-\epsilon)\max\log\limits_{\boldsymbol v^*}P_{\boldsymbol \theta_\nv}(\boldsymbol v^*) + \epsilon\min\log\limits_{\boldsymbol v^*}P_{\boldsymbol \theta_\nv}(\boldsymbol v^*) \right\}
 $$
@@
 In other words, S-unstable RBM model sequences are guaranteed to stack up all probability on a specific set of outcomes for visibles, which could potentially be arbitrarily narrow.
-Proofs of Propositions \@ref(prp:instab1) and \@ref(prp:degen) appear in @kaplan2017note.
+Proofs of Propositions \@ref(prp:instab1) and \@ref(prp:degen) appear in @kaplan2016propriety.
 These findings also have counterparts in results in @schweinberger2011instability, but unlike results there, we do not limit consideration to exponential family forms with a fixed number of parameters.
@@
 
-Secondly, we can look at $\text{ELPR}(\boldsymbol \theta)$ from (\@ref(eq:elpr)) for this tiny model in order to consider model instability as a function of RBM parameters.
-Recall that large values of $\text{ELPR}(\boldsymbol \theta)$ are associated with an extreme sensitivity of the model probabilities $f_{\boldsymbol \theta}(\boldsymbol x)$ to small changes in $\boldsymbol x$ (see Proposition \@ref(prp:instab1)).
-The quantity $\text{ELPR}(\boldsymbol \theta)$ for this small RBM is
+Secondly, we can look at $\text{LREP}(\boldsymbol \theta)$ from (\@ref(eq:elpr)) for this tiny model in order to consider model instability as a function of RBM parameters.
+Recall that large values of $\text{LREP}(\boldsymbol \theta)$ are associated with an extreme sensitivity of the model probabilities $f_{\boldsymbol \theta}(\boldsymbol x)$ to small changes in $\boldsymbol x$ (see Proposition \@ref(prp:instab1)).
+The quantity $\text{LREP}(\boldsymbol \theta)$ for this small RBM is
 \begin{align*}
-\text{ELPR}(\boldsymbol \theta) &= \log \left[\frac{\max\limits_{(v_1, \dots, v_\nv) \in \mathcal{C}^\nv}P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}{\min\limits_{(v_1, \dots, v_\nv) \in \mathcal{C}^\nv}P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}\right] = \log \left[\frac{\max\limits_{v_1 \in \mathcal{C}}\sum\limits_{h_1 \in \mathcal{C}}\exp\left\{\theta_{11} h_1 v_1 + \theta_{h_1}h_1 + \theta_{v_1} v_1 \right\}}{\min\limits_{v_1 \in \mathcal{C}}\sum\limits_{h_1 \in \mathcal{C}}\exp\left\{\theta_{11} h_1 v_1 + \theta_{h_1}h_1 + \theta_{v_1} v_1 \right\}}\right].
+\text{LREP}(\boldsymbol \theta) &= \log \left[\frac{\max\limits_{(v_1, \dots, v_\nv) \in \mathcal{C}^\nv}P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}{\min\limits_{(v_1, \dots, v_\nv) \in \mathcal{C}^\nv}P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}\right] = \log \left[\frac{\max\limits_{v_1 \in \mathcal{C}}\sum\limits_{h_1 \in \mathcal{C}}\exp\left\{\theta_{11} h_1 v_1 + \theta_{h_1}h_1 + \theta_{v_1} v_1 \right\}}{\min\limits_{v_1 \in \mathcal{C}}\sum\limits_{h_1 \in \mathcal{C}}\exp\left\{\theta_{11} h_1 v_1 + \theta_{h_1}h_1 + \theta_{v_1} v_1 \right\}}\right].
 \\
@@
 
-Figure \@ref(fig:instab) shows contour plots of $\text{ELPR}(\boldsymbol \theta)/\nv$ for various values of $\boldsymbol \theta$ in this model with $\nv = 1$.
+Figure \@ref(fig:instab) shows contour plots of $\text{LREP}(\boldsymbol \theta)/\nv$ for various values of $\boldsymbol \theta$ in this model with $\nv = 1$.
 We can see that this quantity is large for large magnitudes of $\boldsymbol \theta$, especially for large values of the dependence/interaction parameter $\theta_{11}$.
@@
 
-```{r instab, fig.cap="$\\text{ELPR}(\\boldsymbol \\theta)/\\nv$ for various values of $\\boldsymbol \\theta$ for the tiny example model.
+```{r instab, fig.cap="$\\text{LREP}(\\boldsymbol \\theta)/\\nv$ for various values of $\\boldsymbol \\theta$ for the tiny example model.
 Recall here $\\nv$ is the number of visible nodes and here is $1$.
@@
   ylab(expression(theta[h1])) +
-  scale_fill_gradient(expression(frac(ELPR(theta), V)), low = "yellow", high = "red") +
+  scale_fill_gradient(expression(frac(LREP(theta), V)), low = "yellow", high = "red") +
   theme(aspect.ratio = 1)
@@
 For $\nh, \nv \in \{1, \dots, 4\}$, we sample `r dim(res$samp[[1]])[1]` values of $\boldsymbol \theta$ with various magnitudes (details to follow).
-For each set of parameters we then calculate metrics of model impropriety introduced in Section \@ref(degeneracy-instability-and-uninterpretability) based on $\boldsymbol \mu(\boldsymbol \theta)$, $\text{ELPR}(\boldsymbol \theta)/\nv$, and the absolute coordinates of $\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] -   \text{E}\left[\boldsymbol X | \boldsymbol \theta^* \right]$.
+For each set of parameters we then calculate metrics of model impropriety introduced in Section \@ref(degeneracy-instability-and-uninterpretability) based on $\boldsymbol \mu(\boldsymbol \theta)$, $\text{LREP}(\boldsymbol \theta)/\nv$, and the absolute coordinates of $\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] -   \text{E}\left[\boldsymbol X | \boldsymbol \theta^* \right]$.
 In the case of near-degeneracy, we classify each model as near-degenerate based on the distance of $\boldsymbol \mu(\boldsymbol \theta)$ from the boundary of the convex hull of $\mathcal{T}$ and look at the fraction of models that are "near-degenerate," meaning they are within a small distance $\epsilon > 0$ of the boundary of the convex hull.
@@
   geom_abline(aes(intercept = 0, slope = 1), alpha = .5, lty = 2) +
-  scale_fill_gradient(expression(paste("Mean ", frac(ELPR(theta), V))), low = "yellow", high = "red") +
+  scale_fill_gradient(expression(paste("Mean ", frac(LREP(theta), V))), low = "yellow", high = "red") +
   facet_grid(Visibles~Hiddens, labeller = label_parsed) +
@@
 
-```{r instab-plots, fig.cap="The sample mean value of $\\text{ELPR}(\\boldsymbol \\theta)/V$ at each grid point for each combination of magnitude of $\\boldsymbol \\theta$ and model size.
+```{r instab-plots, fig.cap="The sample mean value of $\\text{LREP}(\\boldsymbol \\theta)/V$ at each grid point for each combination of magnitude of $\\boldsymbol \\theta$ and model size.
 As the magnitude of $\\boldsymbol \\theta$ grows, so does the value of this metric, indicating typical instability in the model.", fig.height=3.5, fig.align="ht!"}
@@
 \end{equation} 
-for the outcomes $(\pm 1, \pm 1)$, the appendix provides technical details to show parameter values $(\theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21})$ may be chosen to match any given cell probabilities with arbitrary closeness.
+for the outcomes $(\pm 1, \pm 1)$, the online appendix provides technical details to show parameter values $(\theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21})$ may be chosen to match any given cell probabilities with arbitrary closeness.
 In fact, when the cell probabilities (\@ref(eq:0)) are all strictly positive, parameters in the RBM model can be specified to reflect these probabilities exactly.
 And, when one or more of the cell probabilities (\@ref(eq:0)) are zero, the corresponding RBM probabilities $P(V_1=v_1,V_2=_v2| \theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21})$ may never be identically zero (due to exponential terms in the model) but parameters can be still selected to make the appropriate RBM cell probabilities arbitrarily small.
-Furthermore, as development in the appendix shows, not only is it possible to approximate any distribution on the visibles arbitrarily well [cf.
+Furthermore, as development in the online appendix shows, not only is it possible to approximate any distribution on the visibles arbitrarily well [cf.
 @montufar2011refinements], but dramatically different parameter settings can induce the same RBM model (beyond mere symmetries in parameterization).
@@
 For these reasons, we are skeptical about RBMs as data generative tools.
+
+# References {-}

### 2017-11-27 writing/paper-rbm/index.Rmd
getting ready for resubmission

@@
 ---
-title: "Properties and Bayes fitting of restricted Boltzmann machines"
+title: "Properties and Bayesian fitting of restricted Boltzmann machines"
 author: 
@@
 keywords: Degeneracy, Instability, Classification, Deep Learning, Graphical Models
-blind: false
+blind: true
 ---

### 2017-11-30 writing/paper-rbm/docs/paper.tex
submission 2

@@
 
-\title{\bf Properties and Bayes fitting of restricted Boltzmann machines}
-\author{Andee Kaplan \\ Department of Statistics, Iowa State University \\  and \\ Daniel Nordman \\ Department of Statistics, Iowa State University \\  and \\ Stephen Vardeman \\ Department of Statistics and Department of Industrial and\\
+\title{\bf Properties and Bayesian fitting of restricted Boltzmann machines}
+\author{Andee Kaplan \\ Department of Statistical Science, Duke University \\  and \\ Daniel Nordman \\ Department of Statistics, Iowa State University \\  and \\ Stephen Vardeman \\ Department of Statistics and Department of Industrial and\\
 Manufacturing Systems Engineering, Iowa State University \\ }
@@
 
-\section{Introduction}\label{introduction}
+\hypertarget{introduction}{%
+\section{Introduction}\label{introduction}}
 
@@
 Boltzmann machines (RBMs) (see R. Salakhutdinov and Hinton
-\protect\hyperlink{ref-salakhutdinov2009deep}{2009}; R. Salakhutdinov
-and Hinton \protect\hyperlink{ref-salakhutdinov2012efficient}{2012};
-Srivastava, Salakhutdinov, and Hinton
+\protect\hyperlink{ref-salakhutdinov2009deep}{2009},
+\protect\hyperlink{ref-salakhutdinov2012efficient}{2012}; Srivastava,
+Salakhutdinov, and Hinton
 \protect\hyperlink{ref-srivastava2013modeling}{2013}; Le Roux and Bengio
@@
 the image to be generated repeatedly, whereas the non-degenerate model
-shows more potentially realistic variation. \par
+shows more potentially realistic variation.
+
+\par
 \begin{figure}
@@
 potential issues of model instability, degeneracy and uninterpretability
-for the RBM class, which are related properties, and examine the
+for the RBM class (which are related properties) and examine the
 presence of these in Section
@@
 paper, we attempt to address the concerns with overfitting and poor
-mixing in an alternative, and perhaps more transparent manner, via
+mixing in an alternative (and perhaps more transparent) manner, via
 specification of a Bayesian prior in Section
@@
 avoid model impropriety issues raised in Section \ref{model-fitting} and
-conclude with a discussion in Section \ref{discussion}. Appendix
-\ref{appendix-rbm} provides proofs for results on RBM parameterizations
-and data codings described in Section \ref{data-encoding}.
-
-While applications of the RBM have claimed some successes in
-classification problems, it is unclear if the model class allows one to
-go beyond fitting to other statistical matters of importance in using
-the models, such as quantification of the uncertainty of estimation and
-the formulation of predictive distributions. These are closely tied to
-the probability properties of the RBM model for explaining data
-generation and the current exposition intends to contribute to a better
-understanding in this direction.
-
-\section{Background}\label{background}
-
+conclude with a discussion in Section \ref{discussion}. Supplementary
+online material provides proofs for results on RBM parameterizations and
+data codings described in Section \ref{data-encoding}.
+
+While applications of the RBM have been claimed to produce some
+successes in classification problems, it is unclear if the model class
+allows one to go beyond fitting to other statistical matters of
+importance in using the models, such as quantification of the
+uncertainty of estimation and the formulation of predictive
+distributions. These are closely tied to the probability properties of
+the RBM model for explaining data generation and the current exposition
+intends to contribute to a better understanding in this direction.
+
+\hypertarget{background}{%
+\section{Background}\label{background}}
+
+\hypertarget{restricted-boltzmann-machines}{%
 \subsection{Restricted Boltzmann
-machines}\label{restricted-boltzmann-machines}
+machines}\label{restricted-boltzmann-machines}}
 
@@
 
-\subsubsection{Joint distribution}\label{joint-distribution}
+\hypertarget{joint-distribution}{%
+\subsubsection{Joint distribution}\label{joint-distribution}}
 
@@
 f_{\boldsymbol \theta} (\boldsymbol x)\equiv P_{\boldsymbol \theta}(\boldsymbol X = \boldsymbol x) = \frac{\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}{\gamma(\boldsymbol \theta)}, \quad \boldsymbol x \in \mathcal{C}^{\nh + \nv}
-\end{align}
-where
+\end{align} where
 \(\boldsymbol \theta \equiv (\theta_{11}, \dots, \theta_{1\nh}, \dots, \theta_{\nv 1}, \dots, \theta_{\nv \nh}, \theta_{v_1}, \dots, \theta_{v_\nv}, \theta_{h_1}, \dots, \theta_{h_\nh}) \in \mathbb{R}^{\nv + \nh + \nv*\nh}\)
@@
 \(\boldsymbol x = (h_1, \dots, h_\nh, v_1, \dots, v_\nh) \in \mathcal{C}^{\nv + \nh}\)
-and
-\begin{align}
+and \begin{align}
 \boldsymbol t(\boldsymbol x) &= (h_1, \dots, h_\nh, v_1, \dots, v_\nv, v_1 h_1, \dots, v_\nv h_\nh) \in \mathcal{C}^{\nh + \nv + \nh*\nv}, \label{eq:t}
-\end{align}
-let
+\end{align} let
 \(\mathcal{T} = \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{C}^{\nv + \nh}\} \subset \mathbb{R}^{\nv + \nh + \nv * \nh}\)
@@
 
+\hypertarget{connection-to-deep-learning}{%
 \subsubsection{Connection to Deep
-Learning}\label{connection-to-deep-learning}
+Learning}\label{connection-to-deep-learning}}
 
@@
 
+\hypertarget{degeneracy-instability-and-uninterpretability}{%
 \subsection{Degeneracy, instability, and
-uninterpretability}\label{degeneracy-instability-and-uninterpretability}
+uninterpretability}\label{degeneracy-instability-and-uninterpretability}}
 
@@
 
-\subsubsection{Near-degeneracy}\label{near-degeneracy}
+\hypertarget{near-degeneracy}{%
+\subsubsection{Near-degeneracy}\label{near-degeneracy}}
 
@@
 random graph model has a distribution with a probability mass function
-of the form
-\begin{align*}
+of the form \begin{align*}
 f_{\boldsymbol \theta} (\boldsymbol x) = \frac{\exp\left(\boldsymbol \theta^T \boldsymbol t(\boldsymbol x)\right)}{\gamma(\boldsymbol \theta)}, \boldsymbol x \in \mathcal{X},
-\end{align*}
-where \(\boldsymbol \theta \in \Theta \subset \mathbb{R}^q\) is the
-model parameter, and
-\(\boldsymbol t: \mathcal{X} \rightarrow \mathbb{R}^q\) is a vector of
-statistics based on the adjacency matrix of a graph. Here, as earlier,
+\end{align*} where
+\(\boldsymbol \theta \in \Theta \subset \mathbb{R}^q\) is the model
+parameter, and \(\boldsymbol t: \mathcal{X} \rightarrow \mathbb{R}^q\)
+is a vector of statistics based on the adjacency matrix of a graph.
+Here, as earlier,
 \(\gamma(\boldsymbol \theta) = \sum_{\boldsymbol x \in \mathcal{X}} \exp\left(\boldsymbol \theta^T \boldsymbol t(\boldsymbol x)\right)\)
@@
 \(\boldsymbol \theta \in \mathbb{R}^{\nv + \nh + \nh*\nv}\). Here the
-mean of \(\boldsymbol t(\boldsymbol x)\) is
-\begin{align*}
+mean of \(\boldsymbol t(\boldsymbol x)\) is \begin{align*}
 \boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta} \boldsymbol t(\boldsymbol X)  &= \sum\limits_{x \in \mathcal{C}^{\nv + \nh}} \left\{ \boldsymbol t(\boldsymbol x) f_{\boldsymbol \theta} (\boldsymbol x) \right\} \\
@@
 \end{align*}
-\subsubsection{Instability}\label{instability}
+\hypertarget{instability}{%
+\subsubsection{Instability}\label{instability}}
 
@@
 denote an element of a sequence of RBM parameters indexed by the number
-\(\nv\) of visibles \((V_1, \dots, V_\nv)\) and define a (scaled)
-extremal log-probability ratio of the RBM model at
-\(\boldsymbol \theta_{\nv}\) as
-\begin{align}
-\frac{1}{\nv} \log \left[\frac{\max\limits_{(v_1, \dots, v_\nv) \in \mathcal{C}^\nv}P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}{\min\limits_{(v_1, \dots, v_\nv) \in \mathcal{C}^\nv}P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}\right] \equiv \frac{1}{\nv} \text{ELPR}(\boldsymbol \theta_\nv) \label{eq:elpr}
-\end{align}
-where
+\(\nv\) of visibles \((V_1, \dots, V_\nv)\) and define a log-ratio of
+extremal probabilities (LREP) of the RBM model at
+\(\boldsymbol \theta_{\nv}\) as \begin{align}
+\frac{1}{\nv} \log \left[\frac{\max\limits_{(v_1, \dots, v_\nv) \in \mathcal{C}^\nv}P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}{\min\limits_{(v_1, \dots, v_\nv) \in \mathcal{C}^\nv}P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}\right] \equiv \frac{1}{\nv} \text{LREP}(\boldsymbol \theta_\nv) \label{eq:elpr}
+\end{align} where
 \(P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv) \propto \sum_{(h_1, \dots, h_\nh) \in \mathcal{C}^\nh}\exp\left(\sum_{i = 1}^\nv \theta_{v_i} v_i + \sum_{j = 1}^\nh \theta_{h_j} h_j + \sum_{i = 1}^\nv \sum_{j = 1}^\nh \theta_{ij} v_i h_j\right)\)
@@
 the number \(\nv\) of visibles. A RBM model formulation is
-\emph{Schweinberger-unstable} or \emph{S-unstable} if
-\begin{align*}
-\lim\limits_{\nv \rightarrow \infty} \frac{1}{\nv} \text{ELPR}(\boldsymbol \theta_\nv) = \infty.
+\emph{Schweinberger-unstable} or \emph{S-unstable} if \begin{align*}
+\lim\limits_{\nv \rightarrow \infty} \frac{1}{\nv} \text{LREP}(\boldsymbol \theta_\nv) = \infty.
 \end{align*}
-\EndKnitrBlock{definition} In other words, a RBM model sequence is
-unstable if, given any \(c > 0\), there exists an integer \(n_c > 0\) so
-that \(\frac{1}{\nv}\text{ELPR}(\boldsymbol \theta_\nv) > c\) for all
+\EndKnitrBlock{definition}
+
+In other words, a RBM model sequence is unstable if, given any
+\(c > 0\), there exists an integer \(n_c > 0\) so that
+\(\frac{1}{\nv}\text{LREP}(\boldsymbol \theta_\nv) > c\) for all
 \(\nv \ge n_c\). This definition of \emph{S-unstable} is a
@@
 \BeginKnitrBlock{proposition}
-\protect\hypertarget{prp:instab1}{}{\label{prp:instab1}}Let \(c > 0\) and
-let \(\text{ELPR}(\boldsymbol \theta_\nv)\) be as in (\eqref{eq:elpr}) for
+\protect\hypertarget{prp:instab1}{}{\label{prp:instab1} }Let \(c > 0\) and
+let \(\text{LREP}(\boldsymbol \theta_\nv)\) be as in (\eqref{eq:elpr}) for
 an integer \(\nv \ge 1\). If
-\(\frac{1}{\nv}\text{ELPR}(\boldsymbol \theta_\nv) > c\), then
+\(\frac{1}{\nv}\text{LREP}(\boldsymbol \theta_\nv) > c\), then
 \(\Delta(\boldsymbol \theta_\nv) > c\).
@@
 \(\boldsymbol \theta_\nv, \nv \ge 1\) as \[
-M_{\epsilon, \boldsymbol \theta_\nv} \equiv \left\{\boldsymbol v \in \mathcal{C}^\nv: \log P_{\boldsymbol \theta_\nv}(\boldsymbol v) > (1-\epsilon)\max\limits_{\boldsymbol v^*}P_{\boldsymbol \theta_\nv}(\boldsymbol v^*) + \epsilon\min\limits_{\boldsymbol v^*}P_{\boldsymbol \theta_\nv}(\boldsymbol v^*) \right\}
+M_{\epsilon, \boldsymbol \theta_\nv} \equiv \left\{\boldsymbol v \in \mathcal{C}^\nv: \log P_{\boldsymbol \theta_\nv}(\boldsymbol v) > (1-\epsilon)\max\log\limits_{\boldsymbol v^*}P_{\boldsymbol \theta_\nv}(\boldsymbol v^*) + \epsilon\min\log\limits_{\boldsymbol v^*}P_{\boldsymbol \theta_\nv}(\boldsymbol v^*) \right\}
 \] for a given \(0 < \epsilon < 1\). Then S-unstable model sequences are
@@
 could potentially be arbitrarily narrow. Proofs of Propositions
-\ref{prp:instab1} and \ref{prp:degen} appear in Kaplan, Nordman, and
-Vardeman (\protect\hyperlink{ref-kaplan2017note}{2017}). These findings
-also have counterparts in results in Schweinberger
+\ref{prp:instab1} and \ref{prp:degen} follow from more general results
+in Kaplan, Nordman, and Vardeman
+(\protect\hyperlink{ref-kaplan2016propriety}{2016}). These findings also
+have counterparts in results in Schweinberger
 (\protect\hyperlink{ref-schweinberger2011instability}{2011}), but unlike
@@
 
-\subsubsection{Uninterpretability}\label{uninterpretability}
+\hypertarget{uninterpretability}{%
+\subsubsection{Uninterpretability}\label{uninterpretability}}
 
@@
 \text{E}\left[\boldsymbol X | \boldsymbol \theta\right] = \sum\limits_{\boldsymbol x \in \mathcal{C}^{\nv + \nh}} \boldsymbol x f_{\boldsymbol \theta}(\boldsymbol x) = \sum\limits_{\boldsymbol x \in \mathcal{C}^{\nv + \nh}} \boldsymbol x \frac{\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{C}^{\nv + \nh}}\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}
-\] and
-\begin{align*}
+\] and \begin{align*}
 \text{E}\left[\boldsymbol X | \boldsymbol 
@@
 \end{align*}
+\hypertarget{explorations-of-model-properties-through-simulation}{%
 \section{Explorations of model properties through
-simulation}\label{explorations-of-model-properties-through-simulation}
+simulation}\label{explorations-of-model-properties-through-simulation}}
 
@@
 
-\subsection{Tiny example}\label{tiny-example}
+\hypertarget{tiny-example}{%
+\subsection{Tiny example}\label{tiny-example}}
 
@@
 \end{figure}
-\subsubsection{Impropriety three ways}\label{impropriety-three-ways}
+\hypertarget{impropriety-three-ways}{%
+\subsubsection{Impropriety three ways}\label{impropriety-three-ways}}
 
@@
 \(\mathcal{T} = \{(v_1, h_1, v_1 h_1 : v_1, h_1 \in \{0, 1\} \}\) (i.e.
-\(v_1 h_1 = 1\) only if \(v_1 = h_1 = 1\)). \par
+\(v_1 h_1 = 1\) only if \(v_1 = h_1 = 1\)).
+
+\par
 \begin{figure}[ht]
@@
 We can compute the mean vector for \(\boldsymbol t(\boldsymbol x)\) as a
-function of the model parameters as
-\begin{align*}
+function of the model parameters as \begin{align*}
 \boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta}\left[ \boldsymbol t(\boldsymbol X) \right] =  \sum\limits_{\boldsymbol x = (v_1, h_1) \in \{0,1\}^2} \left\{ t(x) \frac{\exp\left( \theta_{11} h_1 v_1 + \theta_{h1} h_1 + \theta_{v1} v_1\right)}{\gamma(\boldsymbol \theta)} \right\} = \left[
@@
 \end{matrix} \right]
-\end{align*}
-where
+\end{align*} where
 \(\gamma(\boldsymbol \theta) = \sum\limits_{h_1 = 0}^1\sum\limits_{v_1 = 0}^1 \exp(\theta_{11} h_1 v_1 + \theta_{h_1}h_1 + \theta_{v_1}v_1)\).
@@
 the relationship between values of \(\boldsymbol \theta\) and model
-degeneracy. \par
+degeneracy.
+
+\par
 \begin{figure}
@@
 \end{figure}
-Secondly, we can look at \(\text{ELPR}(\boldsymbol \theta)\) from
+Secondly, we can look at \(\text{LREP}(\boldsymbol \theta)\) from
 (\eqref{eq:elpr}) for this tiny model in order to consider model
 instability as a function of RBM parameters. Recall that large values of
-\(\text{ELPR}(\boldsymbol \theta)\) are associated with an extreme
+\(\text{LREP}(\boldsymbol \theta)\) are associated with an extreme
 sensitivity of the model probabilities
@@
 \(\boldsymbol x\) (see Proposition \ref{prp:instab1}). The quantity
-\(\text{ELPR}(\boldsymbol \theta)\) for this small RBM is
-\begin{align*}
-\text{ELPR}(\boldsymbol \theta) &= \log \left[\frac{\max\limits_{(v_1, \dots, v_\nv) \in \mathcal{C}^\nv}P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}{\min\limits_{(v_1, \dots, v_\nv) \in \mathcal{C}^\nv}P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}\right] = \log \left[\frac{\max\limits_{v_1 \in \mathcal{C}}\sum\limits_{h_1 \in \mathcal{C}}\exp\left\{\theta_{11} h_1 v_1 + \theta_{h_1}h_1 + \theta_{v_1} v_1 \right\}}{\min\limits_{v_1 \in \mathcal{C}}\sum\limits_{h_1 \in \mathcal{C}}\exp\left\{\theta_{11} h_1 v_1 + \theta_{h_1}h_1 + \theta_{v_1} v_1 \right\}}\right].
+\(\text{LREP}(\boldsymbol \theta)\) for this small RBM is \begin{align*}
+\text{LREP}(\boldsymbol \theta) &= \log \left[\frac{\max\limits_{(v_1, \dots, v_\nv) \in \mathcal{C}^\nv}P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}{\min\limits_{(v_1, \dots, v_\nv) \in \mathcal{C}^\nv}P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}\right] = \log \left[\frac{\max\limits_{v_1 \in \mathcal{C}}\sum\limits_{h_1 \in \mathcal{C}}\exp\left\{\theta_{11} h_1 v_1 + \theta_{h_1}h_1 + \theta_{v_1} v_1 \right\}}{\min\limits_{v_1 \in \mathcal{C}}\sum\limits_{h_1 \in \mathcal{C}}\exp\left\{\theta_{11} h_1 v_1 + \theta_{h_1}h_1 + \theta_{v_1} v_1 \right\}}\right].
 \\
@@
 Figure \ref{fig:instab} shows contour plots of
-\(\text{ELPR}(\boldsymbol \theta)/\nv\) for various values of
+\(\text{LREP}(\boldsymbol \theta)/\nv\) for various values of
 \(\boldsymbol \theta\) in this model with \(\nv = 1\). We can see that
@@
 produced by consideration of \(\boldsymbol \mu(\boldsymbol \theta)\).
+
 \par
@@
 
-\caption{$\text{ELPR}(\boldsymbol \theta)/\nv$ for various values of $\boldsymbol \theta$ for the tiny example model.
+\caption{$\text{LREP}(\boldsymbol \theta)/\nv$ for various values of $\boldsymbol \theta$ for the tiny example model.
 Recall here $\nv$ is the number of visible nodes and here is $1$.
@@
 RBM model where \(\boldsymbol X = (V_1, H_1, V_1 H_1)\). This difference
-is given by
-\begin{align*}
+is given by \begin{align*}
 \text{E}\left[\boldsymbol X | \boldsymbol \theta\right] -   \text{E}\left[\boldsymbol X | \boldsymbol \theta^* \right] 
@@
 \(\text{E}[\boldsymbol X|\boldsymbol \theta] - \text{E}[\boldsymbol X| \boldsymbol \theta^*]\)
-as a signal of uninterpretability for the toy RBM. \par
+as a signal of uninterpretability for the toy RBM.
+
+\par
 \begin{figure}
@@
 
-\subsubsection{Data encoding}\label{data-encoding}
+\hypertarget{data-encoding}{%
+\subsubsection{Data encoding}\label{data-encoding}}
 
@@
 
+\hypertarget{exploring-manageable-examples}{%
 \subsection{Exploring manageable
-examples}\label{exploring-manageable-examples}
+examples}\label{exploring-manageable-examples}}
 
@@
 \(\boldsymbol \mu(\boldsymbol \theta)\),
-\(\text{ELPR}(\boldsymbol \theta)/\nv\), and the absolute coordinates of
+\(\text{LREP}(\boldsymbol \theta)/\nv\), and the absolute coordinates of
 \(\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] - \text{E}\left[\boldsymbol X | \boldsymbol \theta^* \right]\).
@@
 limiting the magnitude of the individual interactions may well help
-prevent model impropriety. \par
+prevent model impropriety.
+
+\par
 \begin{figure}
@@
 \begin{figure}
-\includegraphics{paper_files/figure-latex/instab-plots-1} \caption{The sample mean value of $\text{ELPR}(\boldsymbol \theta)/V$ at each grid point for each combination of magnitude of $\boldsymbol \theta$ and model size.
+\includegraphics{paper_files/figure-latex/instab-plots-1} \caption{The sample mean value of $\text{LREP}(\boldsymbol \theta)/V$ at each grid point for each combination of magnitude of $\boldsymbol \theta$ and model size.
 As the magnitude of $\boldsymbol \theta$ grows, so does the value of this metric, indicating typical instability in the model.}\label{fig:instab-plots}
@@
 the risk of degeneracy starts at a slightly larger magnitude of
-parameters, but very quickly increases until reaching close to 1. \par
+parameters, but very quickly increases until reaching close to 1.
+
+\par
 \begin{figure}
@@
 
-\section{Model Fitting}\label{model-fitting}
+\hypertarget{model-fitting}{%
+\section{Model Fitting}\label{model-fitting}}
 
@@
 \(\theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21}\).
-Given some specified cell probabilities, say
-\begin{equation}
+Given some specified cell probabilities, say \begin{equation}
 \label{eq:0}
 0\leq p_{(-1,-1)},\;p_{(1,-1)}, \;p_{(-1,1)},\;p_{(1,1)},
-\end{equation}
-for the outcomes \((\pm 1, \pm 1)\), the appendix provides technical
-details to show parameter values
+\end{equation} for the outcomes \((\pm 1, \pm 1)\), the online appendix
+provides technical details to show parameter values
 \((\theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21})\)
@@
 probabilities arbitrarily small. Furthermore, as development in the
-appendix shows, not only is it possible to approximate any distribution
-on the visibles arbitrarily well (cf. Montufar and Ay
+online appendix shows, not only is it possible to approximate any
+distribution on the visibles arbitrarily well (cf. Montufar and Ay
 \protect\hyperlink{ref-montufar2011refinements}{2011}), but dramatically
@@
 
-\subsection{Bayesian model fitting}\label{bayesian-model-fitting}
+\hypertarget{bayesian-model-fitting}{%
+\subsection{Bayesian model fitting}\label{bayesian-model-fitting}}
 
@@
 terms in the model, \(\boldsymbol \theta_{interaction}\), shrinking
-their magnitudes. \par 
+their magnitudes.
+
+\par 
 \begin{table}[ht]
@@
    \pi(\boldsymbol \theta) \propto \gamma(\boldsymbol \theta)^n \exp\left(-\frac{1}{2C_{1}}\boldsymbol \theta_{main}'\boldsymbol \theta_{main} -\frac{1}{2C_{2}}\boldsymbol \theta_{interaction}'\boldsymbol \theta_{interaction}\right), \vspace{-.75cm}
-  \end{align*}
-  where
-  \begin{align*}
+   \end{align*} where \begin{align*}
    \gamma(\boldsymbol \theta) = \sum\limits_{\boldsymbol x \in \mathcal{C}^{\nh + \nv}}\exp\left(\sum\limits_{i = 1}^\nv\sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)
-  \end{align*}
-  for hyperparameter choices \(0< C_2 < C_1\). The unknown hidden
-  variables \(h_j\) are also directly treated as latent variables and
-  are sampled in each MCMC iterative draw from the posterior
+   \end{align*} for hyperparameter choices \(0< C_2 < C_1\). The unknown
+  hidden variables \(h_j\) are also directly treated as latent variables
+  and are sampled in each MCMC iterative draw from the posterior
   distribution. This is the method of Li
@@
   Normal priors applied to the marginal probabilities for visible
-  variables given by
-  \begin{align*}
+  variables given by \begin{align*}
    g_{\boldsymbol \theta}(\boldsymbol v) \propto \sum\limits_{\boldsymbol h \in \mathcal{C}^\nh} \exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right), \boldsymbol v \in \mathcal{C}^\nv.
-  \end{align*}
-  We will refer to this method as Bayes with Truncated Normal prior and
-  Marginalized Likelihood (BwTNML).
+   \end{align*} We will refer to this method as Bayes with Truncated
+  Normal prior and Marginalized Likelihood (BwTNML).
 \end{enumerate}
@@
 the hidden variables from the likelihood decreases to zero at a much
-faster rate, indicating better mixing for the chain. \par
+faster rate, indicating better mixing for the chain.
+
+\par
 \begin{figure}
@@
 observations alone, which is infeasible for a model with \(\nh\) of any
-real size. \par
+real size.
+
+\par
 \begin{figure}
@@
 \end{figure}
-\section{Discussion}\label{discussion}
+\hypertarget{discussion}{%
+\section{Discussion}\label{discussion}}
 
@@
 
-\clearpage
-
-\appendix
-
-\section{On the one-to-one correspondence between parameters and moments
-in the RBM distribution}\label{appendix-rbm}
-
-For integers \(m,n \geq 1\), consider random vectors
-\(\bm{V}_m \equiv(V_1,\ldots,V_m)\) and
-\(\bm{H}_n\equiv (H_1,\ldots,H_n)\), where the random variables \(V_i\)
-and \(H_j\) assume values in \(\{1,-1\}\), \(i=1,\ldots,m\),
-\(j=1,\ldots,n\). We suppose the vector \((\bm{V}_m,\bm{H}_n)\) follows
-a restricted Boltzmann machine (RBM) distribution, i.e.~has probability
-mass function
-\begin{align*}
-&P(V_1= v_1,\ldots,V_m= v_m, H_1= h_1,\ldots,H_n=h_n|\bm{\theta}) \propto \exp\left[ \sum_{i=1}^m v_i \theta_{v_j} + \sum_{j=1}^n h_j \theta_{h_j} + \sum_{i=1}^m  \sum_{j=1}^n v_i  h_j \theta_{ij} \right], \\
-& \quad \{v_i\}_{i=1}^m,\{h_j\}_{j=1}^n \subset\{\pm 1\},
-\end{align*}
-with \(m+n+mn\) real parameters
-\(\theta_{v_1},\ldots,\theta_{v_m},\theta_{h_1},\ldots,\theta_{h_m}\)
-and \(\theta_{ij}\), \(i=1,\ldots,m\), \(j=1,\ldots,n\) collected in a
-parameter vector denoted as \(\bm{\theta}\in \mathbb{R}^{m+n+mn}\).
-Denote the distribution of \((\bm{V}_m,\bm{H}_n)\) as
-RBM\((m,n,\bm{\theta})\). In the following, for convenience, we say a
-symmetric random variable \(X\) has a Bernoulli\((1/2)\) distribution if
-\(P(X = 1) = 1/2 = P(X = -1)\).
-
-Use the notation
-\(\bm{V}_m*\bm{H}_n = (V_1 H_1,\ldots,V_1 H_n, V_2 H_1,\ldots,V_2 H_n,\ldots, V_m H_1,\ldots,V_m H_n)\).
-For the RBM\((m,n,\bm{\theta})\) model, consider in the function
-\(\bm{g}: \mathbb{R}^{m+n+mn} \rightarrow \mathcal{I} \subset \mathbb{R}^{m+n+mn}\)
-given by
-\begin{align*}
-\bm{g}(\bm{\theta}) &= \E_{\bm{\theta}}(\bm{V}_m,\bm{H}_n,\bm{V}_m*\bm{H}_n)\\
-&= \sum_{\bm{v}_m \in \{\pm 1\}^m, \bm{h}_n \in \{\pm 1\}^n} (\bm{v}_m,\bm{h}_n,\bm{v}_m*\bm{h}_n)P(\bm{V}_m=\bm{v}_m,\bm{H}_n=\bm{h}_n|\bm{\theta})
-\end{align*}
-for \(\bm{\theta}\in \mathbb{R}^{m+n+mn}\); above
-\(\mathcal{I} \equiv \{\bm{g}(\bm{\theta}) : \bm{\theta}\in \mathbb{R}^{m+n+mn}\}\)
-denotes the image of the mapping.
-
-We have the following properties:
-\begin{itemize}
-\tightlist
-\item
-  The function \(\bm{g}: \mathbb{R}^{m+n+mn} \rightarrow \mathcal{I}\)
-  is one-to-one; see Theorem \ref{thm:one} below. Hence, these moments
-  uniquely characterize the parameters in the RBM\((m,n,\bm{\theta})\)
-  model.
-\item
-  The image \(\mathcal{I}\) is a symmetric set (i.e.,
-  \(\bm{g}(\bm{\theta})=-\bm{g}(-\bm{\theta})\)) and must be a connected
-  (not necessarily convex) region in \(\mathbb{R}^{m+n+mn}\), without
-  any voids in it (by the continuity of \(\bm{g}(\bm{\theta})\)).
-\item
-  The center of the image \(\mathcal{I}\) (the zero vector in
-  \(\mathbb{R}^{m+n+mn}\)) is given only by \(\bm{\theta}=\bm{0}\) (the
-  zero vector in the parameter space). This corresponds to the case
-  where all the random vectors \(V_1,\ldots,V_m,H_1,\ldots,H_n\) are iid
-  with a symmetric Bernoulli\((1/2)\) distribution; see Theorem
-  \ref{thm:two}.
-\end{itemize}
-\BeginKnitrBlock{theorem}
-\protect\hypertarget{thm:one}{}{\label{thm:one}}Let \(m,n\geq 1\) and
-suppose the random vector
-\((\bm{V}_m,\bm{H}_n) = (V_1,\ldots,V_m, H_1,\ldots,H_n)\) follows a
-RBM\((m,n,\bm{\theta})\) distribution. Let
-\(\bm{\theta}_1,\bm{\theta}_2 \in \mathbb{R}^{m+n+mn}\) and let \(P_i\)
-and \(\E\) denote probability and expectation under \(\bm{\theta}_i\),
-\(i=1,2\). Then, the following are equivalent
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  \(\E_1 V_i=\E_2 V_i\), \(\E_1 H_j=\E_2 H_j\) and
-  \(\E_1 V_i H_j =\E_2 V_i H_j\) for any \(i=1,\ldots,m\),
-  \(j=1,\ldots,n\).
-\item
-  \(\bm{\theta}_1 =\bm{\theta}_2\).
-\item
-  \(P_1\) and \(P_2\) are the same distribution for
-  \((\bm{V}_m,\bm{H}_n)\).\textbackslash{}
-\end{enumerate}
-\EndKnitrBlock{theorem}
-
-\BeginKnitrBlock{proof}
-\iffalse <span class="proof"><em>Proof. \fi{} We first establish the
-equivalence of claims 2 and 3. If claim 2 holds, then claim 3 holds
-trivially. Suppose claim 3 holds and pick \(\ell \in \{1,\ldots,m\}\).
-Let
-\(\theta_{v_1},\ldots,\theta_{v_m},\theta_{h_1},\ldots,\theta_{h_n}\),
-\(\theta_{ij}\), \(i=1,\ldots,m\), \(j=1,\ldots,n\), denote the
-components of \(\bm{\theta}_1\) and let
-\(\tilde{\theta}_{v_1},\ldots,\tilde{\theta}_{v_m},\tilde{\theta}_{h_1},\ldots,\tilde{\theta}_{h_n}\),
-\(\tilde{\theta}_{ij}\), \(i=1,\ldots,m\), \(j=1,\ldots,n\), denote the
-analogous components of \(\bm{\theta}_2\). Then, by assumption,
-\begin{align*}
-\exp\left[ 2\theta_{v_\ell} +2 \sum_{j=1}^n  h_j \theta_{\ell j} \right] &= \frac{P_1\left(\{V_\ell = 1\} \cap \bigcap_{i=1 \atop i \neq \ell}^m \{V_i=v_i\} \cap \bigcap_{j=1}^n \{H_j=h_j\}\right)} {P_1\left(\{V_\ell = -1\} \cap \bigcap_{i=1 \atop i \neq \ell}^m \{V_i=v_i\} \cap \bigcap_{j=1}^n \{H_j=h_j\}\right)}\\
-&= \frac{P_2\left(\{V_\ell = 1\} \cap \bigcap_{i=1 \atop i \neq \ell}^m \{V_i=v_i\} \cap \bigcap_{j=1}^n \{H_j=h_j\}\right)}
-{P_2\left(\{V_\ell = -1\} \cap \bigcap_{i=1 \atop i \neq \ell}^m \{V_i=v_i\} \cap \bigcap_{j=1}^n \{H_j=h_j\}\right)}\exp\left[ 2\tilde{\theta}_{v_\ell} +2 \sum_{j=1}^n  h_j \tilde{\theta}_{\ell j} \right]
-\end{align*}
-for any \(\{v_i\}_{i=1,i\neq \ell}^m,\{h_j\}_{j=1}^n \subset\{\pm 1\}\).
-Equating exponents, we have
-\(\theta_{v_\ell} - \tilde{\theta}_{v_\ell} = \sum_{j=1}^n h_j ( \tilde{\theta}_{\ell j} -\theta_{\ell j})\)
-for any \(\{h_j\}_{j=1}^n \subset \{\pm 1\}\). Taking all \(h_j=1\) or
-all \(h_j=-1\) for \(1 \leq j \leq n\) yields
-\(\theta_{v_\ell} = \tilde{\theta}_{v_\ell}\). This result further
-implies
-\(\tilde{\theta}_{\ell 1} -\theta_{\ell 1} = - \sum_{j=2}^n h_j ( \tilde{\theta}_{\ell j} -\theta_{\ell j})\)
-for any given \(\{h_j\}_{j=2}^n \subset \{\pm 1\}\), so that
-\(\theta_{\ell 1}=\tilde{\theta}_{\ell 1}\) must hold. Iterating this
-argument, sequentially for each \(j=2,\ldots,m\), shows
-\(\theta_{\ell j}=\tilde{\theta}_{\ell j}\) for any \(1 \leq j \leq n\)
-in addition to \(\theta_{v_\ell}=\tilde{\theta}_{v_\ell}\). As
-\(\ell\in\{1,\ldots,m\}\) was arbitrary, we have
-\(\theta_{v_i}=\tilde{\theta}_{v_i}\) and
-\(\theta_{i j}=\tilde{\theta}_{i j}\) for any \(1 \leq i \leq m\),
-\(1 \leq j \leq n\).
-
-Now pick \(k \in \{1,\ldots,n\}\) so that, by the
-RBM\((m,n,\bm{\theta})\) probability structure as above, we have
-analogously that \[
-\theta_{h_k} + \sum_{i=1}^m  v_i \theta_{i k} =  \tilde{\theta}_{h_k} + \sum_{i=1}^m  v_i \tilde{\theta}_{i k}
-\] for any \(\{v_i\}_{i=1}^m \subset \{\pm 1\}\), so that
-\(\theta_{h_k} = \tilde{\theta}_{h_k}\). As \(k \in \{1,\ldots,n\}\) was
-arbitrary, we have now established claim 2 from claim 3.
-
-To establish the equivalence of claims 1 and 3 in Theorem \ref{thm:one}
-under the RBM\((m,n,\bm{\theta})\) model, we note that claim 3 easily
-implies claim 1. Hence, it suffices now to establish claim 3 from claim
-1. This follows immediately from Lemma \ref{lem:first} below and the
-fact that the variables \(V_1,\ldots,V_m\) are conditionally independent
-given \(H_1,\ldots,H_n\) in the RBM\((m,n,\bm{\theta})\) model (and
-likewise \(H_1,\ldots,H_n\) are independent given \(V_1,\ldots,V_m\)).
-\EndKnitrBlock{proof}
-
-\BeginKnitrBlock{lemma}
-\protect\hypertarget{lem:first}{}{\label{lem:first}}Let \(m,n\geq 1\). Let
-\(P_1,P_2\) denote two probability distributions for a random vector
-\((\bm{V}_m,\bm{H}_n) = (V_1,\ldots,V_m, H_1,\ldots,H_n)\) supported on
-\(\{a,b\}^{m+n}\), for some real constants \(a\neq b \in\mathbb{R}\),
-such that \(P_i( \bm{V}_m=\bm{v}_m, \bm{H}_n=\bm{h}_n)>0\), \(i=1,2\),
-for any \(\bm{v}_m \in \{a,b\}^{m}\), \(\bm{h}_n \in \{a,b\}^{n}\).
-Suppose additionally that, under \(P_1\) or \(P_2\), the variables
-\(V_1,\ldots,V_m\) are conditionally independent given
-\(\bm{H}_n=\bm{h}_n \in \{a,b\}^{n}\) and that the variables
-\(H_1,\ldots,H_n\) are conditionally independent given
-\(\bm{V}_m=\bm{v}_m \in \{a,b\}^{m}\).
-
-Let \(\E_i\) denote expectation under the distribution \(P_i\),
-\(i=1,2\). Then, for any
-\(\bm{v}_m \in \{a,b\}^{m}, \bm{h}_n \in \{a,b\}^{n}\), \[
-P_1( \bm{V}_m=\bm{v}_m, \bm{H}_n=\bm{h}_n) = P_2( \bm{V}_m=\bm{v}_m, \bm{H}_n=\bm{h}_n), \qquad
-\] follows if \(\E_1 V_i = \E_2 V_i\), \(\E_1 H_j = \E_2 H_j\),
-\(\E_1 V_i H_j = \E_2 V_j H_j\) holds for all \(i\in\{1,\ldots,m\}\),
-\(j\in \{1,\ldots,n\}\).
-\EndKnitrBlock{lemma}
-
-\BeginKnitrBlock{proof}
-\iffalse <span class="proof"><em>Proof. \fi{} The case \(m=n=1\) follows
-from Lemma \ref{lem:second}. We now use an induction argument, assuming
-that the result of Lemma \ref{lem:first} holds for some given order of
-\((m,n)\) with \(m,n\geq 1\) and show the the result continues to hold
-for \((m+1,n)\) or \((m,n+1)\). We shall treat the case \((m+1,n)\)
-(where the other case \((m,n+1)\) follows by symmetrical arguments).
-
-Considering random variables
-\((\bm{V}_{m+1},\bm{H}_n) = (V_1,\ldots,V_m,V_{m+1}, H_1,\ldots,H_n)\),
-the induction hypothesis applies to \((\bm{V}_{m},\bm{H}_n)\) where
-\(\bm{V}_m =(V_1,\ldots,V_m)\). Hence, for any
-\(\bm{v}_m \in \{a,b\}^{m}, \bm{h}_n \in \{a,b\}^{n}\), we have
-\(P_1(\bm{V}_m=\bm{v}_m,\bm{H}_n=\bm{h}_n) = P_2(\bm{V}_m=\bm{v}_m,\bm{H}_n=\bm{h}_n)>0\),
-\(P_1( \bm{H}_n=\bm{h}_n) = P_2(\bm{H}_n=\bm{h}_n)>0\) and consequently
-\begin{align}\label{eq:0}
-P_1(\bm{V}_m=\bm{v}_m|\bm{H}_n=\bm{h}_n) =P_2(\bm{V}_m=\bm{v}_m |\bm{H}_n=\bm{h}_n).
-\end{align}
-Likewise, by the induction hypothesis applied to
-\((\bm{V}_{m-},\bm{H}_n)\) for \(\bm{V}_{m-} = (V_2,\ldots,V_{m+1})\),
-we have
-\begin{eqnarray}\label{eq:0}
- 1- P_1(V_{m+1} = b | \bm{H}_n=\bm{h}_n) &=& P_1(V_{m+1} = a |\bm{H}_n=\bm{h}_n)\\
-\nonumber  &=&  P_2(V_{m+1} = a | \bm{H}_n=\bm{h}_n)\\
-\nonumber  & = &1-P_2(V_{m+1} = b | \bm{H}_n=\bm{h}_n)
-\end{eqnarray}
-for any \(\bm{h}_n \in \{a,b\}^{n}\). Now fix
-\(\bm{v}_m \in \{a,b\}^{m}, \bm{h}_n \in \{a,b\}^{n}\) and let
-\(\bm{v}_{m+1}= (v_{m+1},\bm{v}_m)\) for \(v_{m+1}\in\{a,b\}\). Then, by
-the conditional independence assumption, \[
-P_i( \bm{V}_{m+1}=\bm{v}_{m+1}| \bm{H}_n=\bm{h}_n) = P_i( V_{m+1}=v_{m+1}| \bm{H}_n=\bm{h}_n) P_i( \bm{V}_{m}=v_{m}| \bm{H}_n=\bm{h}_n)
-\] for \(i=1,2\), so that we have
-\(P_1( \bm{V}_{m+1}=\bm{v}_{m+1}| \bm{H}_n=\bm{h}_n) = P_2( \bm{V}_{m+1}=\bm{v}_{m+1}| \bm{H}_n=\bm{h}_n)\)
-by (@ref\{eq:0\})-(\eqref{eq:1}). Consequently, it follows that
-\begin{eqnarray*}
-P_1( \bm{V}_{m+1}=\bm{v}_{m+1}, \bm{H}_n=\bm{h}_n) &=& P_1( \bm{V}_{m+1}=\bm{v}_{m+1}| \bm{H}_n=\bm{h}_n) P_1(\bm{H}_n=\bm{h}_n)\\
-& =& P_2( \bm{V}_{m+1}=\bm{v}_{m+1}| \bm{H}_n=\bm{h}_n)P_2(\bm{H}_n=\bm{h}_n) \\&=&P_2( \bm{V}_{m+1}=\bm{v}_{m+1}, \bm{H}_n=\bm{h}_n)
-\end{eqnarray*}
-by \(P_1(\bm{H}_n=\bm{h}_n)=P_2(\bm{H}_n=\bm{h}_n)\). As
-\(\bm{v}_{m+1}\in \{a,b\}^{m+1},\bm{h}_n \in \{a,b\}^{n}\) were
-arbitrary, this completes the proof of Lemma \ref{lem:first}.
-\EndKnitrBlock{proof}
-
-\BeginKnitrBlock{lemma}
-\protect\hypertarget{lem:second}{}{\label{lem:second}}Suppose the discrete
-random vector \((X,Y)\) has support \(\{a,b\}\times \{a,b\}\), for some
-\(a\neq b \in \mathbb{R}\), under two probability distributions \(P_1\)
-and \(P_2\). Let \(\E_i\) denote expectation under the distribution
-\(P_i\), \(i=1,2\). Then, the following are equivalent
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  \(\E_1 X = \E_2 X\), \(\E_1 Y = \E_2 Y\), \(\E_1 X Y = \E_2 X Y\),
-  where \(\E_i\) denotes expectation under \(P_i\), \(i=1,2\).
-\item
-  \(P_1(X=x,Y=y)=P_2(X=x,Y=y)\) for \(x,y\in\{a,b\}\), i.e., \((X,Y)\)
-  has the same distribution under \(P_1\) and \(P_2\).
-\end{enumerate}
-\EndKnitrBlock{lemma}
-
-\BeginKnitrBlock{proof}
-\iffalse <span class="proof"><em>Proof. \fi{} Denote the four
-probabilities by which the random vector \((X,Y)\) assume pairs
-\((b,b),(b,a),(a,b),(a,a)\), respectively, as
-\(c_1,c_2,c_3,1-c_1-c_2-c_3\) under \(P_1\) and
-\(d_1,d_2,d_3,1-d_1-d_2-d_3\) under \(P_2\). The conditions
-\(\E_1 X = \E_2 X\), \(\E_1 Y = \E_2 Y\), \(\E_1 X Y = \E_2 X Y\) with
-\(b\neq a\) imply that \[
-(c_1+c_2) = (d_1+d_2), \quad  (c_1+c_3) = (d_1+d_3) \quad (b-a) c_1 + a (2c_1+c_2+c_3) =  (b-a) d_1 + a (2d_1+d_2+d_3).
-\] As \((2c_1+c_2+c_3) = (2d_1+d_2+d_3)\) and \(b\neq a\), we conclude
-that \(c_1=d_1\), from which it follows that \(c_2=d_2\) and
-\(c_3=d_3\). Hence, claim 1 implies claim 2 in Lemma \ref{lem:second}.
-Claim 2 also trivially implies claim 1.
-\EndKnitrBlock{proof}
-
-\BeginKnitrBlock{theorem}
-\protect\hypertarget{thm:two}{}{\label{thm:two}}Let \(m,n\geq 1\) and
-suppose the random vector
-\((\bm{V}_m,\bm{H}_n) = (V_1,\ldots,V_m, H_1,\ldots,H_n)\) follows a
-RBM\((m,n,\bm{\theta})\) distribution. Then, the following are
-equivalent
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  \(\E V_i=0\), \(\E H_j=0\) and \(\E V_i H_j =0\) for any
-  \(i=1,\ldots,m\), \(j=1,\ldots,n\).
-\item
-  \(\bm{\theta} = \bm{0}\in \mathbb{R}^{m+n+mn}\).
-\item
-  \(V_1,\ldots,V_m, H_1,\ldots,H_n\) are iid Bernoulli\((1/2)\) random
-  variables.
-\end{enumerate}
-\EndKnitrBlock{theorem}
-
-\BeginKnitrBlock{proof}
-\iffalse <span class="proof"><em>Proof. \fi{} We first establish the
-equivalence of claims 2 and 3. If claim 2 holds, then claim 3 follows
-easily from the resulting uniform cell probabilities:
-\(P(V_1= v_1,\ldots,V_m= v_m, H_1= h_1,\ldots,H_n=h_n|\bm{\theta})=2^{-mn}\)
-for any \(v_i,h_j\in\{\pm 1\}\), \(i=1,\ldots,m\), \(j=1,\ldots,n\).
-
-Now suppose claim 3 holds. Pick \(\ell \in \{1,\ldots,m\}\). Then, by
-the RBM\((m,n,\bm{\theta})\) probability structure and the iid Bernoulli
-assumption, we have the conditional probability
-\begin{eqnarray*}
-\frac{1}{2} &=& P\left(V_\ell = v_{\ell}\Big| \bigcap_{i=1 \atop i \neq \ell}^m \{V_i=v_i\} \cap \bigcap_{j=1}^n \{H_j=h_j\}\right) \\
-&\equiv &\frac{\exp\left( v_\ell\theta_\ell +  v_\ell\sum_{j=1}^n  h_j \theta_{\ell j}\right)}{\exp\left(
-\theta_\ell + \sum_{j=1}^n   h_j\theta_{\ell j} \right) +\exp\left(-\theta_\ell  -\sum_{j=1}^n  h_j\theta_{\ell j}\right) }
-\end{eqnarray*}
-for any subsets \(\{v_i\}_{i=1}^m,\{h_j\}_{j=1}^n \subset\{\pm 1\}\).
-Equating conditional probabilities when \(v_\ell=1\) or when
-\(v_\ell=-1\), we have
-\(\theta_\ell = - \sum_{j=1}^m h_j \theta_{\ell j}\) for any given
-\(\{h_j\}_{j=1}^n \subset \{\pm 1\}\). Taking all \(h_j=1\) or all
-\(h_j=-1\) for \(1 \leq j \leq n\) yields \(\theta_\ell=0\). This result
-further implies \(\theta_{\ell 1} = - \sum_{j=2}^n h_j \theta_{\ell j}\)
-for any given \(\{h_j\}_{j=2}^n \subset \{\pm 1\}\), so that
-\(\theta_{\ell 1}=0\) must hold. Iterating this argument, sequentially
-for each \(j=2,\ldots,m\), shows \(\theta_{\ell j}=0\) for any
-\(1 \leq j \leq n\) in addition to \(\theta_{v_\ell}=0\). As
-\(\ell\in\{1,\ldots,m\}\) was arbitrary, we have \(\theta_{v_i}=0\) and
-\(\theta_{i j}=0\) for any \(1 \leq i \leq m\), \(1 \leq j \leq n\).
-
-Now pick \(k \in \{1,\ldots,n\}\) so that, by the
-RBM\((m,n,\bm{\theta})\) probability structure and the iid assumption,
-we have the conditional probability \[
- \frac{1}{2} = P\left(H_k = h_k \Big| \bigcap_{j=1 \atop j \neq k}^n \{H_j=h_j\}\right) \equiv \frac{\exp(h_k \theta_{h_k})}{\exp(\theta_{h_k}) + \exp(-\theta_{h_k})}
-\] for any \(\{h_j\}_{j=1}^n \subset \{\pm 1\}\). Considering \(h_k=1\)
-or \(h_k=-1\), we conclude that \(\theta_{h_k} = -\theta_{h_k}\) or
-\(\theta_{h_k}=0\). As \(k \in \{1,\ldots,n\}\) was arbitrary, we have
-now established claim 2 from claim 3.
-
-To establish the equivalence of claims 1 and 2 in Theorem \ref{thm:two}
-under the RBM\((m,n,\bm{\theta})\) model, we note that claim 2 again
-implies claim 3, which then easily implies claim 1. By Theorem
-\ref{thm:one}, if claim 1 holds then the only possibility for this is
-\(\bm{\theta} = \bm{0} \in \mathbb{R}^{m+n+mn}\).
-\EndKnitrBlock{proof}
-
-\section{Flexibility of the RBM model}\label{appendix-rbm2}
-
-In an RBM model with enough hidden variables, parameter values may be
-chosen to match any given cell probabilities with arbitrary closeness.
-Additionally, when one or more of the cell probabilities (\eqref{eq:0})
-are zero, the corresponding RBM probabilities may never be identically
-zero (due to exponential terms in the model) but parameters can be still
-selected to make the appropriate RBM cell probabilities arbitrarily
-small. We show this for a model with two visible variables \((V_1,V_2)\)
-and one hidden \(H_1\).
-
-To demonstrate, we assume \(p_{(-1,-1)}>0\) (without loss of generality)
-in the specified cell probabilities (\eqref{eq:0}) and replace parameters
-\(\theta_{11},\theta_{21}\) with
-\(\Delta_1 \equiv \theta_{11} +\theta_{21}\) and
-\(\Delta_2 \equiv \theta_{11} -\theta_{21}\). We may then prescribe
-values of
-\(\theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \Delta_1,\Delta_2\) so that
-the model probability ratio \[
-P(V_1=v_1,V_2=v_2|\theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \Delta_1,\Delta_2)/  P(V_1=-1,V_2=-1| \theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \Delta_1,\Delta_2)
-\] matches the corresponding ratio \(p_{(v_1,v_2)}/p_{(-1,-1)}\) over
-three values of \((v_1,v_2) = (1,-1),(-1,1),(1,1)\).
-
-For instance, assuming the cell probabilities from (\eqref{eq:0}) are all
-positive, these probabilities can be exactly reproduced by choosing
-\begin{eqnarray*}
-\theta_{v_1} &=& \frac{1}{2} \log\left( \frac{p_{(1,-1)}}{p_{(-1,-1)}} \frac{\exp(\theta_{h_1}-\Delta_1) + \exp(-\theta_{h_1}+\Delta_1)}{\exp(\theta_{h_1}+\Delta_2) + \exp(-\theta_{h_1}-\Delta_2)}  \right),\\
-\theta_{v_2}&= &\frac{1}{2} \log\left( \frac{p_{(-1,1)}}{p_{(-1,-1)}} \frac{\exp(\theta_{h_1}-\Delta_1) + \exp(-\theta_{h_1}+\Delta_1)}{\exp(\theta_{h_1}-\Delta_2) + \exp(-\theta_{h_1}+\Delta_2)}  \right)
-\end{eqnarray*}
-and selecting \(\theta_{h_1}, \Delta_1,\Delta_2\) to solve
-\begin{equation}
-\label{eq:1}
-\frac{p_{(1,1)}p_{(-1,-1)}}{p_{(-1,1)}p_{(1,-1)}} = \frac{\ell(|\theta_{h_1}|) + \ell(|\Delta_1|) }{\ell(|\theta_{h_1}|) + \ell(|\Delta_2|)},
-\end{equation}
-based on a monotonically increasing function
-\(\ell(x)\equiv \exp(-2x)+\exp(2x)\), \(x \geq 0\). If
-\([p_{(1,1)}p_{(-1,-1)}]/ [p_{(-1,1)}p_{(1,-1)}] \geq 1\), one can pick
-any values for \(\theta_{h_1}, \Delta_2\in \mathbb{R}\) and solve
-(\eqref{eq:1}) for \(|\Delta_1|\); likewise, when
-\([p_{(1,1)}p_{(-1,-1)}]/ [p_{(-1,1)}p_{(1,-1)}] < 1\) in
-(\eqref{eq:1})\}), one may solve for \(|\Delta_2|\) upon choosing any
-values for \(\theta_{h_1}, \Delta_1\in \mathbb{R}\).
-
-Alternatively, if exactly one specified cell probability in (\eqref{eq:0})
-is zero, say \(p_{(1,1)}\) (without loss of generality), we can select
-parameters \(\theta_{v_1},\theta_{v_2}\) as above based on a sequence
-\((\theta_{h_1}, \Delta_{1}, \Delta_{2}) \equiv (\theta_{h_1}^{(m)}, \Delta_{1}^{(m)}, \Delta_{2}^{(m)})\),
-\(m\in\{1,2,\ldots,\}\) of the remaining parameter values such that
-\(\lim_{m\to \infty}|\Delta_{1}^{(m)}| = \infty\) and
-\(\lim_{m\to \infty} (|\theta_{h_1}^{(m)}| + |\Delta_{2}^{(m)}|)/|\Delta_{1}^{(m)}|=0\)
-hold. This guarantees that the resulting RBM model matches the given
-cell probabilities (\eqref{eq:0}) in the limit:
-\begin{equation}
-\label{eq:2}
-\lim_{m\to \infty}P(V_1=v_1,V_2=v_2| \theta_{v_1}, \theta_{v_2}, \theta_{h_1},\Delta_1,\Delta_2) = p_{(v_1,v_2)},\quad (v_1,v_2)\in\{(\pm 1,\pm 1)\}.
-\end{equation}
-If exactly two specified probabilities in (\eqref{eq:0}) are zero, say
-\(p_{(1,1)}\) and \(p_{(-1,1)}\) (without loss of generality), then a
-limit approximation as in (\eqref{eq:2}) follows by picking
-\(\theta_{v_1}\) as above based on any choices of
-\((\theta_{h_1}, \Delta_1,\Delta_2)\) and choosing a sequence of
-\(\theta_{v_2}\equiv \theta_{v_2}^{(m)}\) values for which
-\(\theta_{v_2}^{({m})} \rightarrow -\infty\).
-
-\section*{References}\label{references}
+\hypertarget{references}{%
+\section*{References}\label{references}}
 \addcontentsline{toc}{section}{References}
@@
 \hypertarget{refs}{}
-\hypertarget{ref-carreira2005contrastive}{}
+\leavevmode\hypertarget{ref-carreira2005contrastive}{}%
 Carreira-Perpinan, Miguel A, and Geoffrey E Hinton. 2005. ``On
@@
 
-\hypertarget{ref-cho2012tikhonov}{}
+\leavevmode\hypertarget{ref-cho2012tikhonov}{}%
 Cho, KyungHyun, Alexander Ilin, and Tapani Raiko. 2012. ``Tikhonov-Type
@@
 
-\hypertarget{ref-fisher1922mathematical}{}
+\leavevmode\hypertarget{ref-fisher1922mathematical}{}%
 Fisher, R. A. 1922. ``On the Mathematical Foundations of Theoretical
@@
 London A: Mathematical, Physical and Engineering Sciences} 222
-(594-604). The Royal Society: 309--68.
-
-\hypertarget{ref-box1967discrimination}{}
+(594-604). The Royal Society:309--68.
+
+\leavevmode\hypertarget{ref-box1967discrimination}{}%
 G. E. P. Box, W. J. Hill. 1967. ``Discrimination Among Mechanistic
 Models.'' \emph{Technometrics} 9 (1). {[}Taylor \& Francis, Ltd.,
-American Statistical Association, American Society for Quality{]}:
-57--71.
-
-\hypertarget{ref-gelman2011inference}{}
+American Statistical Association, American Society for
+Quality{]}:57--71.
+
+\leavevmode\hypertarget{ref-gelman2011inference}{}%
 Gelman, Andrew, Kenneth Shirley, and others. 2011. ``Inference from
@@
 
-\hypertarget{ref-handcock2003assessing}{}
+\leavevmode\hypertarget{ref-handcock2003assessing}{}%
 Handcock, Mark S. 2003. ``Assessing Degeneracy in Statistical Models of
@@
 
-\hypertarget{ref-hinton2010practical}{}
+\leavevmode\hypertarget{ref-hinton2010practical}{}%
 Hinton, Geoffrey. 2010. ``A Practical Guide to Training Restricted
-Boltzmann Machines.'' \emph{Momentum} 9 (1): 926.
-
-\hypertarget{ref-hinton2002training}{}
+Boltzmann Machines.'' \emph{Momentum} 9 (1):926.
+
+\leavevmode\hypertarget{ref-hinton2002training}{}%
 Hinton, Geoffrey E. 2002. ``Training Products of Experts by Minimizing
-Contrastive Divergence.'' \emph{Neural Computation} 14 (8). MIT Press:
-1771--1800.
-
-\hypertarget{ref-hinton2006fast}{}
+Contrastive Divergence.'' \emph{Neural Computation} 14 (8). MIT
+Press:1771--1800.
+
+\leavevmode\hypertarget{ref-hinton2006fast}{}%
 Hinton, Geoffrey E, Simon Osindero, and Yee-Whye Teh. 2006. ``A Fast
 Learning Algorithm for Deep Belief Nets.'' \emph{Neural Computation} 18
-(7). MIT Press: 1527--54.
-
-\hypertarget{ref-kaiser2007statistical}{}
+(7). MIT Press:1527--54.
+
+\leavevmode\hypertarget{ref-kaiser2007statistical}{}%
 Kaiser, Mark S. 2007. ``Statistical Dependence in Markov Random Field
@@
 
-\hypertarget{ref-kaplan2017note}{}
-Kaplan, Andee, Daniel Nordman, and Stephen Vardeman. 2017. ``A Note on
-the Instability and Degeneracy of Deep Learning Models.'' \emph{In
-Preparation}.
-
-\hypertarget{ref-larochelle2008classification}{}
+\leavevmode\hypertarget{ref-kaplan2016propriety}{}%
+Kaplan, Andee, Daniel Nordman, and Stephen Vardeman. 2016. ``On the
+Propriety of Restricted Boltzmann Machines.'' \emph{arXiv Preprint
+arXiv:1612.01158}.
+
+\leavevmode\hypertarget{ref-larochelle2008classification}{}%
 Larochelle, Hugo, and Yoshua Bengio. 2008. ``Classification Using
@@
 
-\hypertarget{ref-le2008representational}{}
+\leavevmode\hypertarget{ref-le2008representational}{}%
 Le Roux, Nicolas, and Yoshua Bengio. 2008. ``Representational Power of
 Restricted Boltzmann Machines and Deep Belief Networks.'' \emph{Neural
-Computation} 20 (6). MIT Press: 1631--49.
-
-\hypertarget{ref-lehmann1990model}{}
+Computation} 20 (6). MIT Press:1631--49.
+
+\leavevmode\hypertarget{ref-lehmann1990model}{}%
 Lehmann, E. L. 1990. ``Model Specification: The Views of Fisher and
 Neyman, and Later Developments.'' \emph{Statistical Science} 5 (2).
-Institute of Mathematical Statistics: 160--68.
-
-\hypertarget{ref-li2014biclustering}{}
+Institute of Mathematical Statistics:160--68.
+
+\leavevmode\hypertarget{ref-li2014biclustering}{}%
 Li, Jing. 2014. ``Biclustering Methods and a Bayesian Approach to
@@
 
-\hypertarget{ref-montufar2011refinements}{}
+\leavevmode\hypertarget{ref-montufar2011refinements}{}%
 Montufar, Guido, and Nihat Ay. 2011. ``Refinements of Universal
 Approximation Results for Deep Belief Networks and Restricted Boltzmann
-Machines.'' \emph{Neural Computation} 23 (5). MIT Press: 1306--19.
-
-\hypertarget{ref-montufar2011expressive}{}
+Machines.'' \emph{Neural Computation} 23 (5). MIT Press:1306--19.
+
+\leavevmode\hypertarget{ref-montufar2011expressive}{}%
 Montúfar, Guido F, Johannes Rauh, and Nihat Ay. 2011. ``Expressive Power
@@
 
-\hypertarget{ref-nguyen2014deep}{}
+\leavevmode\hypertarget{ref-nguyen2014deep}{}%
 Nguyen, Anh Mai, Jason Yosinski, and Jeff Clune. 2014. ``Deep Neural
@@
 
-\hypertarget{ref-salakhutdinov2012efficient}{}
+\leavevmode\hypertarget{ref-salakhutdinov2012efficient}{}%
 Salakhutdinov, Ruslan, and Geoffrey Hinton. 2012. ``An Efficient
 Learning Procedure for Deep Boltzmann Machines.'' \emph{Neural
-Computation} 24 (8). MIT Press: 1967--2006.
-
-\hypertarget{ref-salakhutdinov2009deep}{}
+Computation} 24 (8). MIT Press:1967--2006.
+
+\leavevmode\hypertarget{ref-salakhutdinov2009deep}{}%
 Salakhutdinov, Ruslan, and Geoffrey E Hinton. 2009. ``Deep Boltzmann
@@
 
-\hypertarget{ref-schweinberger2011instability}{}
+\leavevmode\hypertarget{ref-schweinberger2011instability}{}%
 Schweinberger, Michael. 2011. ``Instability, Sensitivity, and Degeneracy
 of Discrete Exponential Families.'' \emph{Journal of the American
-Statistical Association} 106 (496). Taylor \& Francis: 1361--70.
-
-\hypertarget{ref-smolensky1986information}{}
+Statistical Association} 106 (496). Taylor \& Francis:1361--70.
+
+\leavevmode\hypertarget{ref-smolensky1986information}{}%
 Smolensky, Paul. 1986. ``Information Processing in Dynamical Systems:
@@
 
-\hypertarget{ref-srivastava2012multimodal}{}
+\leavevmode\hypertarget{ref-srivastava2012multimodal}{}%
 Srivastava, Nitish, and Ruslan R Salakhutdinov. 2012. ``Multimodal
@@
 
-\hypertarget{ref-srivastava2013modeling}{}
+\leavevmode\hypertarget{ref-srivastava2013modeling}{}%
 Srivastava, Nitish, Ruslan R Salakhutdinov, and Geoffrey E Hinton. 2013.
@@
 
-\hypertarget{ref-szegedy2013intriguing}{}
+\leavevmode\hypertarget{ref-szegedy2013intriguing}{}%
 Szegedy, Christian, Wojciech Zaremba, Ilya Sutskever, Joan Bruna,
@@
 
-\hypertarget{ref-tieleman2008training}{}
+\leavevmode\hypertarget{ref-tieleman2008training}{}%
 Tieleman, Tijmen. 2008. ``Training Restricted Boltzmann Machines Using
@@
 
-\hypertarget{ref-zhou2014some}{}
+\leavevmode\hypertarget{ref-zhou2014some}{}%
 Zhou, Wen. 2014. ``Some Bayesian and Multivariate Analysis Methods in

### 2017-11-30 writing/paper-rbm/docs/paper_blinded.tex
submission 2

--- 2017-11-30 writing/paper-rbm/docs/paper.tex
@@
 
+\bigskip
+\bigskip
+\bigskip
+%\begin{center}
+  %{\LARGE\bf Title}
   \title{\bf Properties and Bayesian fitting of restricted Boltzmann machines}
-\author{Andee Kaplan \\ Department of Statistical Science, Duke University \\  and \\ Daniel Nordman \\ Department of Statistics, Iowa State University \\  and \\ Stephen Vardeman \\ Department of Statistics and Department of Industrial and\\
-Manufacturing Systems Engineering, Iowa State University \\ }
+%\end{center}
+\medskip
 
@@
 \ref{prp:instab1} and \ref{prp:degen} follow from more general results
-in Kaplan, Nordman, and Vardeman
-(\protect\hyperlink{ref-kaplan2016propriety}{2016}). These findings also
-have counterparts in results in Schweinberger
+in XXX (\protect\hyperlink{ref-kaplan2016propriety}{2016}). These
+findings also have counterparts in results in Schweinberger
 (\protect\hyperlink{ref-schweinberger2011instability}{2011}), but unlike
@@
 
-\leavevmode\hypertarget{ref-kaplan2016propriety}{}%
-Kaplan, Andee, Daniel Nordman, and Stephen Vardeman. 2016. ``On the
-Propriety of Restricted Boltzmann Machines.'' \emph{arXiv Preprint
-arXiv:1612.01158}.
-
 \leavevmode\hypertarget{ref-larochelle2008classification}{}%
@@
 
+\leavevmode\hypertarget{ref-kaplan2016propriety}{}%
+XXX, XXX. 2016. ``On the Propriety of Restricted Boltzmann Machines.''
+
 \leavevmode\hypertarget{ref-zhou2014some}{}%

### 2017-11-30 writing/paper-rbm/index.Rmd
submission 2

--- 2017-07-08 writing/paper-rbm/index.Rmd
@@
 ---
-title: "Properties and Bayes fitting of restricted Boltzmann machines"
+title: "Properties and Bayesian fitting of restricted Boltzmann machines"
 author: 
   - name: Andee Kaplan
-    university: Iowa State University
-    department: Department of Statistics
+    university: Duke University
+    department: Department of Statistical Science
   - name: Daniel Nordman

### 2017-11-30 writing/paper-rbm/letters/cover_letter.Rmd
submission 2

---
author: 
 - Andee Kaplan
 - Postdoctoral Associate
 - Department of Statistical Sciences 
 - Duke University
date: "`r format(Sys.time(), '%B %d, %Y')`"
opening: Dear Drs. Liu and Zhao,
closing: Best regards,
address: 
 - Regina Liu and Hongyu Zhao
 - Co-Editors-Elect
 - "*Journal of the American Statistical Association - Theory & Methods*"
return-address:
 - Andee Kaplan
 - Department of Statistical Sciences
 - Duke University
 - P.O. Box 90251
 - Durham, NC 27708-0251
output:
  pdf_document:
    template: template-letter.tex
geometry: margin=1in
---

I am writing to submit an article entitled "Properties and Bayesian fitting of restricted Boltzmann machines" for consideration for publication in the Journal of the American Statistical Association - Theory & Methods.  In this submission, we discuss the relationship between parameter specification and unattractive model properties---degeneracy, instability and uninterpretability---in restricted Boltzmann machines and describe related difficulties that can arise in likelihood-based and Bayes fitting of such models.

This paper references results from another manuscript, an invited revision of which is presently under review by Technometrics. We have provided a blinded copy of this second manuscript as a supplement for referees.

Thank you for your consideration.

### 2017-12-06 writing/paper-rbm/chapters/rbm.Rmd
submission 2
fix blinded

@@
 In other words, S-unstable RBM model sequences are guaranteed to stack up all probability on a specific set of outcomes for visibles, which could potentially be arbitrarily narrow.
-Proofs of Propositions \@ref(prp:instab1) and \@ref(prp:degen) appear in @kaplan2016propriety.
-These findings also have counterparts in results in @schweinberger2011instability, but unlike results there, we do not limit consideration to exponential family forms with a fixed number of parameters.
+Proofs of Propositions \@ref(prp:instab1) and \@ref(prp:degen) follow from more general results in @kaplan2016propriety.
+These findings also have counterparts in results in @schweinberger2011instability, but unlike results there, are not limited in consideration to exponential family forms with a fixed number of parameters.
 

### 2017-12-06 writing/paper-rbm/docs/paper.tex
fix blinded

@@
 in Kaplan, Nordman, and Vardeman
-(\protect\hyperlink{ref-kaplan2016propriety}{2016}). These findings also
+(\protect\hyperlink{ref-kaplan2016propriety}{2017}). These findings also
 have counterparts in results in Schweinberger
 (\protect\hyperlink{ref-schweinberger2011instability}{2011}), but unlike
-results there, we do not limit consideration to exponential family forms
-with a fixed number of parameters.
+results there, are not limited in consideration to exponential family
+forms with a fixed number of parameters.
 
@@
 \leavevmode\hypertarget{ref-kaplan2016propriety}{}%
-Kaplan, Andee, Daniel Nordman, and Stephen Vardeman. 2016. ``On the
-Propriety of Restricted Boltzmann Machines.'' \emph{arXiv Preprint
-arXiv:1612.01158}.
+Kaplan, Andee, Daniel Nordman, and Stephen Vardeman. 2017. ``On the
+Propriety of Restricted Boltzmann Machines.'' \emph{Unpublished}.
 

### 2017-12-06 writing/paper-rbm/docs/paper_blinded.tex
fix blinded

@@
 \ref{prp:instab1} and \ref{prp:degen} follow from more general results
-in XXX (\protect\hyperlink{ref-kaplan2016propriety}{2016}). These
-findings also have counterparts in results in Schweinberger
+in Kaplan, Nordman, and Vardeman
+(\protect\hyperlink{ref-kaplan2016propriety}{2017}). These findings also
+have counterparts in results in Schweinberger
 (\protect\hyperlink{ref-schweinberger2011instability}{2011}), but unlike
-results there, we do not limit consideration to exponential family forms
-with a fixed number of parameters.
+results there, are not limited in consideration to exponential family
+forms with a fixed number of parameters.
 
@@
 
+\leavevmode\hypertarget{ref-kaplan2016propriety}{}%
+Kaplan, Andee, Daniel Nordman, and Stephen Vardeman. 2017. ``On the
+Propriety of Restricted Boltzmann Machines.'' \emph{Unpublished}.
+
 \leavevmode\hypertarget{ref-larochelle2008classification}{}%
@@
 
-\leavevmode\hypertarget{ref-kaplan2016propriety}{}%
-XXX, XXX. 2016. ``On the Propriety of Restricted Boltzmann Machines.''
-
 \leavevmode\hypertarget{ref-zhou2014some}{}%

### 2018-02-22 presentations/fif2018/poster.tex
fif poster

--- 2016-02-29 presentations/coda2016/coda2016_poster.tex
@@
 opacity=1,  %% adjust
-contents={\includegraphics[width=\paperwidth,height=\paperheight]{images/coda2016_poster_background.pdf}}
+contents={\includegraphics[width=\paperwidth,height=\paperheight]{images/poster_background.pdf}}
 }
@@
 Then the probability each node taking the the value corresponding to $\boldsymbol x$ is:
+\small
 \begin{align}
+f_{\boldsymbol \theta} (\boldsymbol x) = \frac{\exp\left(\nsum\limits_{i = 1}^V \nsum\limits_{j=1}^H \theta_{ij} v_i h_j + \nsum\limits_{i = 1}^V\theta_{v_i} v_i + \nsum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\gamma(\boldsymbol \theta)}, \quad \gamma(\boldsymbol \theta) = \Nsum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\nsum\limits_{i = 1}^V \nsum\limits_{j=1}^H \theta_{ij} v_i h_j + \nsum\limits_{i = 1}^V\theta_{v_i} v_i + \nsum\limits_{j = 1}^H\theta_{h_j} h_j\right)
 \label{eqn:pmf}
-f_{\boldsymbol \theta} (\boldsymbol x) = \frac{\exp\left(\nsum\limits_{i = 1}^V \nsum\limits_{j=1}^H \theta_{ij} v_i h_j + \nsum\limits_{i = 1}^V\theta_{v_i} v_i + \nsum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\Nsum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\nsum\limits_{i = 1}^V \nsum\limits_{j=1}^H \theta_{ij} v_i h_j + \nsum\limits_{i = 1}^V\theta_{v_i} v_i + \nsum\limits_{j = 1}^H\theta_{h_j} h_j\right)}
 \end{align}
@@
 \hline
-{\bfseries Instability.} Small changes in natural parameters result in large changes in probability masses, excessive sensitivity \cite{schweinberger2011instability}.
-& If for any $C > 0$ there exists $N_C > 0$ such that
-$\max\limits_{\boldsymbol x_N \in \mathcal{X}_N}[Q(\boldsymbol x_N)] > CN$ for all $N > N_C$, where $Q(\cdot)$ is the neg-potential function of the model. \\
+{\bfseries Instability.} Small shifts in data result in large changes in probability masses, excessive sensitivity \cite{schweinberger2011instability}. & 
+If, for any $c > 0$, there exists $n_c > 0$ so that $\frac{1}{V}\text{LREP}(\boldsymbol \theta) > c$ for all $V \ge n_c$, where $\text{LREP}(\boldsymbol \theta) = \log \left[\frac{\max\limits_{(v_1, \dots, v_V) \in \mathcal{X}^V}P_{\boldsymbol \theta}(v_1, \dots, v_V)}{\min\limits_{(v_1, \dots, v_V) \in \mathcal{X}^V}P_{\boldsymbol \theta}(v_1, \dots, v_V)}\right]$ \\
+%If for any $C > 0$ there exists $N_C > 0$ such that  $\max\limits_{\boldsymbol x_N \in \mathcal{X}_N}[Q(\boldsymbol x_N)] > CN$ for all $N > N_C$, where $Q(\cdot)$ is the neg-potential function of the model.
+\\
 \hline
@@
 \begin{textblock}{14.5}(.65, 23.9)
-\begin{minipage}{.42\textwidth}
+%\begin{minipage}{.42\textwidth}
 \begin{figure}[ht]
@@
   \begin{minipage}{0.49\textwidth}
-  \resizebox{\linewidth}{!}{
+  \resizebox{.8\linewidth}{!}{
     \tdplotsetmaincoords{60}{-60}
@@
 \end{figure}
-\end{minipage}
-\begin{minipage}{.02\textwidth}
-\hfill
-\end{minipage}
-\begin{minipage}{.55\textwidth}
-\begin{figure}
-\centering
-\includegraphics[scale=1]{images/frac_volume.pdf}
-\label{fig:frac_volume}
-\caption{Volume relationship for the convex hulls of statistics in $Q(\cdot)$ vs. unrestricted space.}
-\end{figure}
-\end{minipage}
+%\end{minipage}
+%\begin{minipage}{.02\textwidth}
+%\hfill
+%\end{minipage}
+%\begin{minipage}{.55\textwidth}
+%\begin{figure}
+%\centering
+%\includegraphics[scale=1]{images/frac_volume.pdf}
+%\label{fig:frac_volume}
+%\caption{Volume relationship for the convex hulls of statistics in $Q(\cdot)$ vs. unrestricted space.}
+%\end{figure}
+%\end{minipage}
 
@@
 
-Simulated $n = 5,000$ images (4 pixels) from RBM model with 4 hiddens then fit using Bayesian methods, 
+Simulated $n = 5,000$ images ($V = 4, H = 4$) then fit using Bayesian methods, 
 \vspace{-.5cm}
@@
 \setlength\itemsep{.05cm}
-\item {\bfseries Trick prior.} Cancel out the normalizing term, resulting full conditionals are normally distributed. {\bfseries Conclusion:} Scalable solution, but requires tuning.
+\item {\bfseries Trick prior (BwTPLV).} Cancel out the normalizing term, full conditionals are normal. {\bfseries Conclusion:} Scalable solution, but requires tuning.
 \vspace{-.75cm}
-$$\pi(\boldsymbol \theta) \propto \gamma(\boldsymbol \theta)^n \exp\left(-\frac{1}{2C_{1}}\boldsymbol \theta_{main}'\boldsymbol \theta_{main} -\frac{1}{2C_{2}}\boldsymbol \theta_{interaction}'\boldsymbol \theta_{interaction}\right), \vspace{-.75cm}
-$$ 
-where $\gamma(\boldsymbol \theta) = \Nsum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\nsum\limits_{i = 1}^V \nsum\limits_{j=1}^H \theta_{ij} v_i h_j + \nsum\limits_{i = 1}^V\theta_{v_i} v_i + \nsum\limits_{j = 1}^H\theta_{h_j} h_j\right)$ and $C_{2} < C_{1}$ \cite{li2014biclustering}.
-\item{\bfseries Truncated Normal prior.} Use two independent truncated spherical normal distributions as priors for $\theta_{main}$ and $\theta_{interaction}$ with $\sigma_{interaction} < \sigma_{main}$.
-Full conditional distributions are not conjugate, requires a geometric adaptive MH step \cite{zhou2014some} and calculation of likelihood normalizing constant.
+%$$\pi(\boldsymbol \theta) \propto \gamma(\boldsymbol \theta)^n \exp\left(-\frac{1}{2C_{1}}\boldsymbol \theta_{main}'\boldsymbol \theta_{main} -\frac{1}{2C_{2}}\boldsymbol \theta_{interaction}'\boldsymbol \theta_{interaction}\right) \text{\cite{li2014biclustering}}.
+\vspace{-.75cm}
+%$$ 
+\item{\bfseries Truncated Normal prior (BwTNLV).} Use two independent truncated spherical normal distributions as priors for $\theta_{main}$ and $\theta_{interaction}$ with $\sigma_{interaction} < \sigma_{main}$.
+Requires a geometric adaptive MH step \cite{zhou2014some} and calculation of likelihood normalizing constant.
 \\{\bfseries Conclusion:} Computationally intensive and convergence issues.
-\item{\bfseries Marginalized likelihood.} Marginalize out $\boldsymbol h$ in $f_{\boldsymbol \theta}(\boldsymbol x)$, and use the truncated Normal prior.
+\item{\bfseries Marginalized likelihood (BwTNML).} Marginalize out $\boldsymbol h$ in $f_{\boldsymbol \theta}(\boldsymbol x)$, and use the truncated Normal prior.
 {\bfseries Conclusion:} Least scalable, but removes need to gain MCMC convergence for $Hn$ sampled hidden nodes.
-\vspace{-.75cm}
-$$
-g_{\boldsymbol \theta}(\boldsymbol v) = \Nsum\limits_{\boldsymbol h \in \{-1,1\}^H} \exp\left(\nsum\limits_{i = 1}^V \nsum\limits_{j=1}^H \theta_{ij} v_i h_j + \nsum\limits_{i = 1}^V\theta_{v_i} v_i + \nsum\limits_{j = 1}^H\theta_{h_j} h_j\right).
-\vspace{-.75cm}
-$$
 \end{itemize}
+
+\begin{figure}
+\centering
+\includegraphics{images/image_acf.pdf}
+\label{fig:image_pred}
+\vspace*{-15mm}
+\caption{ACF for the posterior probabilities of each possible 4-pixel image with BwTNLV in black and BwTNML in red.}
+\end{figure}
 
@@
 {\bfseries Big takeaway:} RBMs very easily are degenerate, unstable, and uninterpretable.
-As compounding issues, a rigorous fitting method for these models is not scalable and merely replicates the empirical data distribution, a.k.a optimal nonparametric solution.
+As compounding issues, a rigorous fitting method for these models is not scalable and merely approaches the empirical data distribution, a.k.a optimal nonparametric solution.
 %only replicates the nonparametric solution (empirical distribution).

### 2018-02-22 presentations/fif2018/images/degeneracy_plots.R

@@ -1,0 +1,237 @@
+library(knitr)
+library(ggplot2)
+library(dplyr)
+library(tidyr)
+
+opts_chunk$set(echo=FALSE, message=FALSE, warnings=FALSE)
+theme_set(theme_bw(base_family="serif"))
+
+source("../../../writing/resources/code/functions.R")
+
+#manageable examples ---------
+#reshape data functions
+plot_data <- function(res, grid = FALSE) {
+  
+  plot.data <- data.frame()
+  
+  if(!grid) {
+    for(i in 1:nrow(res)) {
+      tmp <- res$g_theta[[i]] %>% data.frame()
+      H <- res[i,]$H
+      V <- res[i,]$V
+      N <- res[i,]$N
+      r1 <- res[i,]$r1
+      r2 <- res[i,]$r2
+      C <- res[i,]$C
+      epsilon <- res[i,]$epsilon
+      
+      
+      tmp %>% 
+        rowwise() %>% 
+        mutate_(ss_interaction = paste0("(", paste(paste0(names(.)[(H + V + 1):(H + V + H*V)], "^2"), collapse = " + "), ")"),
+                ss_main = paste0("sum(", paste(paste0(names(.)[1:(H+V)],"^2"), collapse = " + "), ")")) %>%    
+        ungroup() -> ratio
+      
+      
+      inner_join(ratio, res$outside[[i]] %>% ungroup()) %>%
+        select(ss_interaction, ss_main, near_hull) %>%
+        mutate(H = H, V = V, n_param = H + V + H*V, N = H + V, N = N, r1 = r1, r2 = r2, C = C, epsilon = epsilon) %>%
+        rbind(plot.data) -> plot.data
+    }
+  } else {
+    for(i in 1:nrow(res)) {
+      tmp <- res$g_theta[[i]] %>% data.frame()
+      H <- res[i,]$H
+      V <- res[i,]$V
+      N <- res[i,]$N
+      r1 <- res[i,]$r1
+      r2 <- res[i,]$r2
+      
+      tmp %>% 
+        rowwise() %>% 
+        mutate_(ss_interaction = paste0("(", paste(paste0(names(.)[(H + V + 1):(H + V + H*V)], "^2"), collapse = " + "), ")"),
+                ss_main = paste0("sum(", paste(paste0(names(.)[1:(H+V)],"^2"), collapse = " + "), ")")) %>%  
+        ungroup() -> ratio
+      
+      
[…]

### 2018-02-22 presentations/fif2018/images/fitting_plots.R

@@ -1,0 +1,136 @@
+#libraries --------------------
+library(dplyr)
+library(tidyr)
+library(ggplot2)
+source("../../../writing/resources/code/functions.R")
+
+load("../../../writing/resources/data/sample_images.Rdata")
+
+#data and params ------------------------
+H <- 4
+V <- 4
+
+load("../../../writing/resources/data/params_theta.Rdata")
+
+params_good <- list(main_hidden = sample.params %>% ungroup() %>% filter(!near_hull) %>% select(starts_with("h"), -H) %>% data.matrix(),
+                    main_visible = sample.params %>% ungroup() %>% filter(!near_hull) %>% select(starts_with("v"), -V) %>% data.matrix(),
+                    interaction = sample.params %>% ungroup() %>% filter(!near_hull) %>% select(starts_with("theta")) %>% data.matrix() %>% matrix(4))
+
+
+#marginalized likelihood
+load("../../../writing/resources/data/fitted_models_trunc_marginal_full.Rdata")
+marginal_bad <- models_bad
+marginal_good <- models_good
+
+#load trick prior
+load("../../../writing/resources/data/fitted_models_jing_5.8.Rdata")
+trick_bad <- models_bad
+trick_good <- models_good
+
+#truncated normal
+load("../../../writing/resources/data/fitted_models_trunc_full.Rdata")
+trunc_bad <- models_bad
+trunc_good <- models_good
+
+#rm unneccesary data
+rm(models_bad)
+rm(models_good)
+
+#computing actual distributions ---------------------
+flat_images_good$visibles %>%
+  data.frame() %>% 
+  rename(v1 = X1, v2 = X2, v3 = X3, v4 = X4) %>%
+  group_by(v1, v2, v3, v4) %>%
+  summarise(prob = n()/nrow(flat_images_good$visibles)) -> distn_emp
+
+distn_good <- visible_distn(params = params_good)
+
+reshape_sample_distn <- function(model) {
+  sample_distn <- model$distn
+  dim(sample_distn) <- c(dim(sample_distn)[1]*dim(sample_distn)[2], dim(sample_distn)[3])
+  sample_distn %>% data.frame() -> sample_distn
+  names(sample_distn) <- names(distn_good)
+  
+  sample_distn %>%
+    group_by(image_id) %>%
+    mutate(iter = 1:n()) -> sample_distn
+  
[…]

### 2018-04-11 writing/paper-note/docs/note.tex
getting ready for another submission

@@
-\documentclass[]{article}
-%\usepackage{nips15submit_e,times}
+\documentclass[numbib]{imamat}
+
+\usepackage{amsmath}
+\usepackage{amsthm}
+\usepackage{cite}
+\usepackage{hyperref}
+\usepackage{graphics}
+\usepackage{algorithmic}
+\usepackage{subfig}
 \usepackage{url}
-
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-
-\usepackage{lmodern}
-\usepackage{amssymb,amsmath}
-\usepackage{ifxetex,ifluatex}
-\usepackage{fixltx2e} % provides \textsubscript
-\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
-  \usepackage[T1]{fontenc}
-  \usepackage[utf8]{inputenc}
-\else % if luatex or xelatex
-  \ifxetex
-    \usepackage{mathspec}
-    \usepackage{xltxtra,xunicode}
-  \else
-    \usepackage{fontspec}
-  \fi
-  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
-  \newcommand{\euro}{€}
-\fi
-% use upquote if available, for straight quotes in verbatim environments
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-% use microtype if available
-\IfFileExists{microtype.sty}{%
-\usepackage{microtype}
-\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
-}{}
-\usepackage{longtable,booktabs}
-\usepackage{graphicx}
-\makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
-\makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
-\ifxetex
-  \usepackage[setpagesize=false, % page size defined by xetex
-              unicode=false, % unicode breaks when used with xetex
-              xetex]{hyperref}
-\else
-  \usepackage[unicode=true]{hyperref}
-\fi
-\hypersetup{breaklinks=true,
-            bookmarks=true,
-            pdfauthor={},
-            pdftitle={On the instability and degeneracy of deep learning models},
-            colorlinks=true,
-            citecolor=blue,
-            urlcolor=blue,
-            linkcolor=magenta,
-            pdfborder={0 0 0}}
-\urlstyle{same}  % don't use monospace font for urls
-\setlength{\parindent}{0pt}
-\setlength{\parskip}{6pt plus 2pt minus 1pt}
-\setlength{\emergencystretch}{3em}  % prevent overfull lines
-\setcounter{secnumdepth}{5}
-
-% Tech stuff starts here
-\usepackage{psfrag,epsf}
-%\pdfminorversion=4
 
@@
 %%INCLUDED-AK
+%%DELETE HERE
 \usepackage{amsthm}
-\newtheorem{theorem}{Theorem}
-\newtheorem{proposition}{Proposition}
-\newtheorem{lemma}{Lemma}
-\newtheorem{corollary}{Corollary}
-
+\newtheoremstyle{theorem}{6pt}{6pt}{\rm}{}{\sffamily}{ }{ }{}
+\theoremstyle{theorem}
+\newtheorem{theorem}{\sc Theorem}[section]
+
+\newtheoremstyle{lemma}{6pt}{6pt}{\rm}{}{\sffamily}{ }{ }{}
+\theoremstyle{lemma}
+\newtheorem{lemma}{\sc Lemma}[section]
+
+\newtheorem{acknowledgement}[theorem]{Acknowledgement}
+\newtheorem{algorithm}[theorem]{Algorithm}
+\newtheorem{axiom}[theorem]{Axiom}
+\newtheorem{case}[theorem]{Case}
+\newtheorem{claim}[theorem]{Claim}
+\newtheorem{conclusion}[theorem]{Conclusion}
+\newtheorem{condition}[theorem]{Condition}
+\newtheorem{conjecture}[theorem]{Conjecture}
+\newtheorem{criterion}[theorem]{Criterion}
+
+\newtheoremstyle{example}{6pt}{6pt}{\rm}{}{\sffamily}{ }{ }{}
+\theoremstyle{example}
+\newtheorem{example}[theorem]{\sc Example}
+
+\newtheoremstyle{corollary}{6pt}{6pt}{\rm}{}{\sffamily}{ }{ }{}
+\theoremstyle{corollary}
+\newtheorem{corollary}{\sc Corollary}[section]
+
+\newtheoremstyle{definition}{6pt}{6pt}{\rm}{}{\sffamily}{ }{ }{}
 \theoremstyle{definition}
-\newtheorem{definition}{Definition}
-\newtheorem{assumption}{Assumption}
-\newtheorem{remark}{Remark}
+\newtheorem{definition}[theorem]{\sc Definition}
+
+\newtheorem{exercise}[theorem]{Exercise}
+
+\newtheorem{notation}[theorem]{Notation}
+\newtheorem{problem}[theorem]{Problem}
+\newtheorem{proposition}[theorem]{\sc Proposition}
+
+\newtheoremstyle{remark}{6pt}{6pt}{\rm}{}{\sffamily}{ }{ }{}
+\theoremstyle{remark}
+\newtheorem{remark}{\sc Remark}[section]
+\newtheorem{solution}[theorem]{Solution}
+\newtheorem{summary}[theorem]{Summary}
+
+\newtheoremstyle{approximation}{6pt}{6pt}{\rm}{}{\sffamily}{ }{ }{}
+\theoremstyle{approximation}
+\newtheorem{approximation}{\sc Approximation}
+
+\newtheoremstyle{scheme}{6pt}{6pt}{\rm}{}{\sffamily}{ }{ }{}
+\theoremstyle{scheme}
+\newtheorem{scheme}{\sc Scheme}
+
+% \newtheorem{proposition}{Proposition}
+% \newtheorem{lemma}{Lemma}
+% \newtheorem{corollary}{Corollary}
+
+% \theoremstyle{definition}
+% \newtheorem{definition}{Definition}
+% \newtheorem{assumption}{Assumption}
+% \newtheorem{remark}{Remark}
+%% STOP DELETING
 
@@
 \newcommand{\Gamt}{\Gamma_{N,2}(\boldsymbol \theta_N) }
-\bigskip
-
-% DON'T change margins - should be 1 inch all around.
-\addtolength{\oddsidemargin}{-.5in}%
-\addtolength{\evensidemargin}{-.5in}%
-\addtolength{\textwidth}{1in}%
-\addtolength{\textheight}{1.3in}%
-\addtolength{\topmargin}{-.8in}%
+
+%imaiai starts here
+%hacks
+%https://tex.stackexchange.com/questions/371955/hyperref-error-argument-of-hysetreflink-has-an-extra/384969
+\AtBeginDocument{
+  \label{FirstPage2}
+  \def\fpage{\pageref{FirstPage2}}
+}
 
@@
 
-\def\spacingset#1{\renewcommand{\baselinestretch}%
-{#1}\small\normalsize} \spacingset{1}
-
-\title{\bf On the instability and degeneracy of deep learning models}
-\author{Andee Kaplan \\ Department of Statistics, Iowa State University \\  and \\ Daniel Nordman \\ Department of Statistics, Iowa State University \\  and \\ Stephen Vardeman \\ Department of Statistics and Department of Industrial and\\
-Manufacturing Systems Engineering, Iowa State University \\ }
+\title{On the instability and degeneracy of deep learning models}
+\shorttitle{Instability of deep learning models}
+\shortauthorlist{Kaplan, Nordman, and Vardeman}
+
+\author{{\sc Andee Kaplan}\(^*\), \\[2pt] Department of Statistical Science, Duke University, P.O. Box 90251,
+Durham, NC 27708-0251, USA \\ \(^*\){\email{Corresponding author: \href{mailto:andrea.kaplan@duke.edu}{\nolinkurl{andrea.kaplan@duke.edu}}}}
+\\[6pt] {\sc Daniel Nordman} \\[2pt] Department of Statistics, Iowa State University, 2438 Osborn Dr., Ames,
+IA 50011-1090, USA \\ {\href{mailto:dnordman@iastate.edu}{\nolinkurl{dnordman@iastate.edu}}}  
+\\[6pt] {\sc Stephen Vardeman} \\[2pt] Departments of Statistics and Industrial and Manufacturing Systems
+Engineering, Iowa State University, 2438 Osborn Dr., Ames, IA
+50011-1090, USA \\ {\href{mailto:vardeman@iastate.edu}{\nolinkurl{vardeman@iastate.edu}}}  
+}
 
 \maketitle
+
 \begin{abstract}
-A probability model exhibits instability if small changes in a data
+{A probability model exhibits instability if small changes in a data
 outcome result in large, and often unanticipated, changes in
@@
 instability apply to large classes of models commonly used in random
-graphs, network analysis, and machine learning contexts.
+graphs, network analysis, and machine learning contexts.}
+{Degeneracy, Instability, Classification, Deep Learning, Graphical Models}
 \end{abstract}
-\noindent%
-{\it Keywords:}  Degeneracy, Instability, Classification, Deep Learning, Graphical Models
-\vfill
-
-\hfill {\tiny technometrics tex template (do not remove)}
-\newpage
-\spacingset{1.45} % DON'T change the spacing!
-
-\section{Introduction}\label{introduction}
+% end imaiai stuff
+
+\hypertarget{introduction}{%
+\section{Introduction}\label{introduction}}
 
@@
 of instability was introduced in the field of statistical physics (i.e.,
-point processes) by Ruelle
-(\protect\hyperlink{ref-ruelle1999statistical}{1999}) and then further
-extended by Schweinberger
-(\protect\hyperlink{ref-schweinberger2011instability}{2011}) for a
-family of exponential models. At issue, models exhibiting instability
-are typically undesirable as these tend to provide poor representations
-of data or data-generation. As an example, such models can include
+point processes) by \citet{ruelle1999statistical} and then further
+extended by \citet{schweinberger2011instability} for a family of
+exponential models. At issue, models exhibiting instability are
+typically undesirable as these tend to provide poor representations of
+data or data-generation. As an example, such models can include
 near-degenerate distributions that assign essentially all probability
@@
 connection to degeneracy has been recognized as a concern in that
-dominant model outcomes may not resemble observed data (cf. Handcock
-\protect\hyperlink{ref-handcock2003assessing}{2003}). As a compounding
-issue, model instability often has direct negative impacts for
-statistical inference and computations based on likelihood functions.
-Namely, volatilities in probability structure can potentially hamper the
-numerical evaluations required for maximum likelihood estimation as well
-as other model-based simulations via Markov Chain Monte Carlo (MCMC).
-These reasons motivate our general study of instability for a broad
-class of probability models, described next.
+dominant model outcomes may not resemble observed data
+\citep[cf.][]{handcock2003assessing}. As a compounding issue, model
+instability often has direct negative impacts for statistical inference
+and computations based on likelihood functions. Namely, volatilities in
+probability structure can potentially hamper the numerical evaluations
+required for maximum likelihood estimation as well as other model-based
+simulations via Markov Chain Monte Carlo (MCMC). These reasons motivate
+our general study of instability for a broad class of probability
+models, described next.
 
@@
 \ref{instability-results} then establishes several formal results for
-FOES models with regard to instability. Schweinberger
-(\protect\hyperlink{ref-schweinberger2011instability}{2011}) originally
-developed instability results specific to a certain class of discrete
-exponential models. For similar exponential models with random networks,
-Handcock (\protect\hyperlink{ref-handcock2003assessing}{2003}) studied
-model degeneracy, where a probability model places near complete mass on
-modes and may thereby narrow the feasible model outcomes. As findings
-here and from Schweinberger
-(\protect\hyperlink{ref-schweinberger2011instability}{2011}) suggest,
-model instability and degeneracy may also be related by viewing
-degeneracy as an extreme, or limiting form, of instability. Our main
-results establish a broad characterization of model instability,
-appropriate across the whole FOES model class, that incorporates results
-of Schweinberger
-(\protect\hyperlink{ref-schweinberger2011instability}{2011}) as a
-special case. We prescribe a general and simple condition for
-identifying instability in a FOES model sequence, which quantifies
-whether certain maximal probabilities in a FOES model are too extreme
-relative to the sample size \(N\). When these conditions are met, the
-probability structure of a FOES model is shown to exhibit extreme
-sensitivity, with probability assignments possessing extreme peaks and
-troughs across nearly identical outcomes. As the measure of model
-instability increases, probabilities from an unstable FOES model
-additionally increase in volatility and provably slide into degeneracy.
-Section \ref{implications} then emphasizes the implications of such
-model instability, showing that such impropriety can be expected to
-numerically hinder maximum likelihood estimation and MCMC-based
+FOES models with regard to instability.
+\citet{schweinberger2011instability} originally developed instability
+results specific to a certain class of discrete exponential models. For
+similar exponential models with random networks,
+\citet{handcock2003assessing} studied model degeneracy, where a
+probability model places near complete mass on modes and may thereby
+narrow the feasible model outcomes. As findings here and from
+\citet{schweinberger2011instability} suggest, model instability and
+degeneracy may also be related by viewing degeneracy as an extreme, or
+limiting form, of instability. Our main results establish a broad
+characterization of model instability, appropriate across the whole FOES
+model class, that incorporates results of
+\citet{schweinberger2011instability} as a special case. We prescribe a
+general and simple condition for identifying instability in a FOES model
+sequence, which quantifies whether certain maximal probabilities in a
+FOES model are too extreme relative to the sample size \(N\). When these
+conditions are met, the probability structure of a FOES model is shown
+to exhibit extreme sensitivity, with probability assignments possessing
+extreme peaks and troughs across nearly identical outcomes. As the
+measure of model instability increases, probabilities from an unstable
+FOES model additionally increase in volatility and provably slide into
+degeneracy. Section \ref{implications} then emphasizes the implications
+of such model instability, showing that such impropriety can be expected
+to numerically hinder maximum likelihood estimation and MCMC-based
 simulations. As one potential remedy, suggestions are given for
@@
 
-\section{Examples}\label{examples}
+\hypertarget{examples}{%
+\section{Examples}\label{examples}}
 
@@
 
+\hypertarget{discrete-exponential-family-models}{%
 \subsection{Discrete Exponential Family
-Models}\label{discrete-exponential-family-models}
+Models}\label{discrete-exponential-family-models}}
 
@@
 exponential family model for \(\boldsymbol X\) with probability mass
-function given by
-\begin{equation}
+function given by \begin{equation}
 \label{eq:expo}
 p_{N, \boldsymbol \theta}(\boldsymbol x) = \exp\left[\boldsymbol\eta^T(\boldsymbol \theta) \boldsymbol g_N(\boldsymbol x) - \psi(\boldsymbol \theta)\right], \qquad \boldsymbol x \in \mathcal{X}^N,
-\end{equation}
-depending on parameter vector
+\end{equation} depending on parameter vector
 \(\boldsymbol \theta \in \Theta_N \subset \mathbb{R}^{k}\) and natural
@@
 indicating a binary \(0\)-\(1\) outcome, the resulting probabilities
-have exponential form \eqref{eq:expo} given by
-\begin{equation}
+have exponential form \eqref{eq:expo} given by \begin{equation}
 \label{eq:mod1}
@@
  \exp\left[\boldsymbol \theta_N \sum_{i=1}^N x_i\right], \qquad \boldsymbol x=(x_1,\ldots,x_N) \in\{0,1\}^N, 
- \end{equation}
-with sufficient statistic
+ \end{equation} with sufficient statistic
 \(\boldsymbol g_N(\boldsymbol x)\equiv \sum_{i=1}^N x_i\) and ``log odds
@@
 among \(k\) possibilities (e.g., a die roll), a multinomial distribution
-is given by
-\begin{equation}
+is given by \begin{equation}
 \label{eq:mod11}
@@
 \exp\left[ \sum_{j=1}^k {\theta_{j,N}} \sum_{i=1}^N \mathbb{I}(x_i=j) \right], \qquad \boldsymbol x  \in\{1,\ldots,k\}^N, 
-\end{equation}
-with sufficient statistic \(\boldsymbol g_N(\boldsymbol x)\) involving a
-count \(\sum_{i=1}^N \mathbb{I}(x_i=j)\) for each outcome
+\end{equation} with sufficient statistic
+\(\boldsymbol g_N(\boldsymbol x)\) involving a count
+\(\sum_{i=1}^N \mathbb{I}(x_i=j)\) for each outcome
 \(j \in \{1,\ldots,k\}\), where \(\mathbb{I}(\cdot)\) denotes the
@@
 exponential models of FOES type commonly arise with dependent spatial
-data (Besag \protect\hyperlink{ref-besag1974spatial}{1974}) and
-network/relational data (Wasserman and Faust
-\protect\hyperlink{ref-wasserman1994social}{1994}; Handcock
-\protect\hyperlink{ref-handcock2003assessing}{2003}). For a random graph
+data \citep{besag1974spatial} and network/relational data
+\citep{wasserman1994social, handcock2003assessing}. For a random graph
 or network with, say, \(n\) nodes, consider \(N={n \choose 2}\) random
@@
  \exp\left[\sum_{j=1}^3 \theta_{j,N} g_{j,N}(\boldsymbol x)\right], \quad\qquad \boldsymbol x=(x_1,\ldots,x_N)  \in\{0,1\}^N,
-\end{equation}
-\[
+\end{equation} \[
   g_{1,N}(\boldsymbol x) \equiv \sum_{i=1}^N  x_i, \qquad\quad g_{2,N}(\boldsymbol x) \equiv \sum_{1 \leq i<j \leq N,\atop s_i \cap s_j \neq \emptyset}\!\!\!   x_i x_j, \qquad
@@
 For this network model \eqref{eq:mod2} in particular, as well as for more
-general models of form \eqref{eq:expo}, Schweinberger
-(\protect\hyperlink{ref-schweinberger2011instability}{2011}) considered
-instability in such exponential models with sequences of fixed
-parameters
+general models of form \eqref{eq:expo},
+\citet{schweinberger2011instability} considered instability in such
+exponential models with sequences of fixed parameters
 \(\boldsymbol \theta_N = (\theta_1,\ldots,\theta_k)\in\mathbb{R}^k\),
@@
 
-\subsection{Restricted Boltzmann Machines}\label{rbm}
+\hypertarget{rbm}{%
+\subsection{Restricted Boltzmann Machines}\label{rbm}}
 
@@
 specified for discrete or continuous random variables, with binary
-variables being most common (cf. Smolensky
-\protect\hyperlink{ref-smolensky1986information}{1986}). A RBM
-architecture has two layers, hidden (\(\mathcal{H}\)) and visible
+variables being most common \citep[cf.][]{smolensky1986information}. A
+RBM architecture has two layers, hidden (\(\mathcal{H}\)) and visible
 (\(\mathcal{V}\)), with conditional independence within each layer. Let
@@
 \(\tilde{\boldsymbol X}=(\boldsymbol X,\boldsymbol H)\) has the joint
-probability mass function
-\begin{equation}
+probability mass function \begin{equation}
 \label{eq:RBM1}
 \tilde{P}_{\boldsymbol \theta_N} (\tilde{\boldsymbol x}) = \exp\left[ (\boldsymbol \theta_N^{\mathcal{H}})^T \boldsymbol h + \boldsymbol (\boldsymbol \theta_N^{\mathcal{V}})^T \boldsymbol x + \boldsymbol h^T  \boldsymbol\theta_N^{\mathcal{HV}} \boldsymbol x - \psi(\boldsymbol \theta_N)\right], \quad \tilde{\boldsymbol x} = (\boldsymbol x, \boldsymbol h) \in \{\pm 1\}^{N+N_\mathcal{H}}
-\end{equation}
-with normalizing function \[
+\end{equation} with normalizing function \[
 \psi(\boldsymbol \theta_N) = \log \sum_{\tilde{\boldsymbol x} \in \{\pm 1\}^{N+N_H} } \exp\left[ (\boldsymbol \theta_N^{\mathcal{H}})^T \boldsymbol h + \boldsymbol (\boldsymbol \theta_N^{\mathcal{V}})^T \boldsymbol x + \boldsymbol h^T  \boldsymbol\theta_N^{\mathcal{HV}} \boldsymbol x\right].
@@
 variables \(\boldsymbol X = (X_1, \dots, X_N)\) follows from
-marginalizing the joint specification to yield
-\begin{equation}
+marginalizing the joint specification to yield \begin{equation}
 \label{eq:RBM2}
 P_{\boldsymbol \theta_N} (\boldsymbol x) = \sum\limits_{\boldsymbol h \in \{\pm 1\}^{N_{\mathcal{H}}}} \tilde{P}_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h), \qquad \boldsymbol x \in \{\pm 1\}^{N}\equiv \mathcal{X}^N.
-\end{equation}
-Here the baseline model \eqref{eq:RBM1} for hidden/visible variables is a
-linear exponential one in sufficient statistics
+\end{equation} Here the baseline model \eqref{eq:RBM1} for hidden/visible
+variables is a linear exponential one in sufficient statistics
 \((\tilde{\boldsymbol X}, \boldsymbol X^T\boldsymbol H)\) using
@@
 
-\subsection{Deep Learning}\label{deep-learning}
+\hypertarget{deep-learning}{%
+\subsection{Deep Learning}\label{deep-learning}}
 
@@
 hidden (or latent) layers in addition to a visible layer of data, namely
-a deep Boltzmann machine (Salakhutdinov and Hinton
-\protect\hyperlink{ref-salakhutdinov2009deep}{2009}) and a deep belief
-network Hinton, Osindero, and Teh
-(\protect\hyperlink{ref-hinton2006fast}{2006}){]}. Let \(M\) denote the
-number of hidden layers included in the model and let
-\(N_{(H,1)}, \dots, N_{(H,M)}\) denote the numbers of hidden variables
-within each hidden layer. Then the random vector
+a deep Boltzmann machine \citep{salakhutdinov2009deep} and a deep belief
+network \citet{hinton2006fast}{]}. Let \(M\) denote the number of hidden
+layers included in the model and let \(N_{(H,1)}, \dots, N_{(H,M)}\)
+denote the numbers of hidden variables within each hidden layer. Then
+the random vector
 \(\tilde{\boldsymbol X} = \{H^{(1)}_1, \dots, H^{(1)}_{N_{(H,1)}}, \dots, H^{(M)}_1, \dots, H^{(M)}_{N_{(H,M)}}, \boldsymbol X\}\)
@@
 difference between the DBM and DBN models is that all but the last
-stacked layer in a DBN are Bayesian networks (see Pearl
-\protect\hyperlink{ref-pearl985bayesian}{1985}), rather than RBMs. A
-Bayesian network is a probabilistic graphical model that defines
-conditional dependence to be directed, rather than undirected (as with
-the RBM). Thus for visibles \(X_1, \dots, X_N\) with support
+stacked layer in a DBN are Bayesian networks
+\citep[see][]{pearl985bayesian}, rather than RBMs. A Bayesian network is
+a probabilistic graphical model that defines conditional dependence to
+be directed, rather than undirected (as with the RBM). Thus for visibles
+\(X_1, \dots, X_N\) with support
 \(\mathcal{X}^N, \mid \mathcal{X} \mid < \infty\), a DBN is also a FOES
@@
 \(\boldsymbol x \in\mathcal{X}^N\). Commonly, as in logistic belief nets
-(Neal \protect\hyperlink{ref-neal1992connectionist}{1992}), a ``weight''
-parameter is placed on each interaction between visibles,
-\(X_1, \dots, X_N\), and the first layer of latent variables,
-\(H^{(1)}_1, \dots, H^{(1)}_{N_{(H,1)}}\), satisfying the definition of
-a FOES model.
-
-\section{Main Results on Model Instability}\label{instability-results}
+\citep{neal1992connectionist}, a ``weight'' parameter is placed on each
+interaction between visibles, \(X_1, \dots, X_N\), and the first layer
+of latent variables, \(H^{(1)}_1, \dots, H^{(1)}_{N_{(H,1)}}\),
+satisfying the definition of a FOES model.
+
+\hypertarget{instability-results}{%
+\section{Main Results on Model Instability}\label{instability-results}}
 
@@
 
-\subsection{A Criterion for Instability}\label{criterion}
+\hypertarget{criterion}{%
+\subsection{A Criterion for Instability}\label{criterion}}
 
@@
 to the sample size \(N\). A relevant quantity to this end is a log-ratio
-of extremal probabilities (LREP), defined as
-\begin{align}
+of extremal probabilities (LREP), defined as \begin{align}
 \label{eq:elpr}
  \REP (\boldsymbol \theta_N)  =  \log \left[\frac{\max\limits_{  \boldsymbol x\in \mathcal{X}^N}P_{\boldsymbol \theta_N}( \boldsymbol x)}{\min\limits_{ \boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}( \boldsymbol x)}\right],
-\end{align}
-based on maximum and minimal model probabilities. In what follows, the
-main idea is that instability, and other negative model features, can be
-associated with a FOES model formulation for \(N\) random variables
-where the LREP \eqref{eq:elpr} is overly large relative to the sample size
-\(N\). That is, a sequence of FOES probability models
+\end{align} based on maximum and minimal model probabilities. In what
+follows, the main idea is that instability, and other negative model
+features, can be associated with a FOES model formulation for \(N\)
+random variables where the LREP \eqref{eq:elpr} is overly large relative
+to the sample size \(N\). That is, a sequence of FOES probability models
 \(P_{\boldsymbol \theta_N}\) results in specifying the distribution of
@@
 \(\boldsymbol X_N=(X_1,\ldots,X_N)\) is \emph{Schweinberger-unstable} or
-\emph{S-unstable} if
-\begin{equation}
+\emph{S-unstable} if \begin{equation}
 \label{eq:Sun}
 \lim \limits_{N \rightarrow \infty} \frac{1}{N} \REP(\boldsymbol \theta_N) \equiv \lim \limits_{N \rightarrow \infty} \frac{1}{N}\log \left[\frac{\max\limits_{  \boldsymbol x\in \mathcal{X}^N}P_{\boldsymbol \theta_N}( \boldsymbol x)}{\min\limits_{ \boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}( \boldsymbol x)}\right] = \infty
-\end{equation}
-as the number of variables increases (\(N \rightarrow \infty\)).
-\EndKnitrBlock{definition} In other words, a model is S-unstable if
+\end{equation} as the number of variables increases
+(\(N \rightarrow \infty\)).
+\EndKnitrBlock{definition}
+
+In other words, a model is S-unstable if
 \(\REP(\boldsymbol \theta_N)/N\) is an unbounded sequence of sample size
@@
 This definition of S-unstable is a generalization or reinterpretation of
-``unstable'' used in Schweinberger
-(\protect\hyperlink{ref-schweinberger2011instability}{2011}) by allowing
+``unstable'' used in \citet{schweinberger2011instability} by allowing
 possibly non-exponential family models (e.g., RBM and DBM models in
@@
 differs in form and scope from the original, it does match that in
-Schweinberger
-(\protect\hyperlink{ref-schweinberger2011instability}{2011}) for the
-special case of exponential models (cf.~Section
-\ref{discrete-exponential-family-models} considered there. Section
-\ref{illustrations} provides several examples of unstable models as well
-as causes for model instability, where the latter may often be traced to
-issues in model form (i.e., data functions) and/or parameterization. We
-next describe several potentially undesirable features associated with
-S-unstable FOES models.
+\citet{schweinberger2011instability} for the special case of exponential
+models (cf.~Section \ref{discrete-exponential-family-models} considered
+there. Section \ref{illustrations} provides several examples of unstable
+models as well as causes for model instability, where the latter may
+often be traced to issues in model form (i.e., data functions) and/or
+parameterization. We next describe several potentially undesirable
+features associated with S-unstable FOES models.
 
@@
 &= \frac{1}{NM}\log\left[\frac{\max_{\boldsymbol x \in \mathcal{X}^{N}}P_{ \boldsymbol \theta_N}(\boldsymbol x)  }{\min_{\boldsymbol x \in \mathcal{X}^{N}}P_{ \boldsymbol \theta_N}(\boldsymbol x)} \right]^M \equiv \frac{1}{N}\REP_{\boldsymbol X_{N}}( \boldsymbol\theta_N ),
-\end{align*}
-where
+\end{align*} where
 \(\REP_{\boldsymbol X_{N}}( \boldsymbol\theta_N ) \equiv \REP( \boldsymbol\theta_N )\)
@@
 
+\hypertarget{characterizations-and-consequences-of-instability}{%
 \subsection{Characterizations and Consequences of
-Instability}\label{characterizations-and-consequences-of-instability}
+Instability}\label{characterizations-and-consequences-of-instability}}
 
@@
 sufficient statistic \(\boldsymbol g_N(\boldsymbol x)\in \mathbb{R}\) in
-\eqref{eq:expo}, Schweinberger
-(\protect\hyperlink{ref-schweinberger2011instability}{2011} Theorem 3)
-proved a result similar in spirit, though based on a characterization
-there in terms of maximum
+\eqref{eq:expo}, \citet[Theorem 3]{schweinberger2011instability} proved a
+result similar in spirit, though based on a characterization there in
+terms of maximum
 \(U_N \equiv \max_{\boldsymbol x\in\mathcal{X}^N}g_N(\boldsymbol x)\)
@@
 \(\mathcal{M}_{\epsilon, -\boldsymbol \theta_N} = \{\boldsymbol x \in\mathcal{X}^N: g_N(\boldsymbol x) < \epsilon U_N + (1-\epsilon) L_N \}\)
-for any \(\boldsymbol \theta_N>0\), and Schweinberger
-(\protect\hyperlink{ref-schweinberger2011instability}{2011} Theorem 3)
-showed each mode set collects all mass, under positive and negative
-parameters, respectively, with unstable models of this exponential type.
-However, for all unstable FOES models, Corollary \ref{cor:sign}
-generalizes the same principle that unstable models can push all
-probability to different, and in fact disjoint, parts of the sample
-space, depending on how parameters fall with respect to zero. This
-feature can numerically complicate likelihood manipulations, such as
-maximization or MCMC-based Bayes posterior sampling, as further
-discussed in Section \ref{implications}.
+for any \(\boldsymbol \theta_N>0\), and \citet[Theorem
+3]{schweinberger2011instability} showed each mode set collects all mass,
+under positive and negative parameters, respectively, with unstable
+models of this exponential type. However, for all unstable FOES models,
+Corollary \ref{cor:sign} generalizes the same principle that unstable
+models can push all probability to different, and in fact disjoint,
+parts of the sample space, depending on how parameters fall with respect
+to zero. This feature can numerically complicate likelihood
+manipulations, such as maximization or MCMC-based Bayes posterior
+sampling, as further discussed in Section \ref{implications}.
 
@@
 
-\section{Illustrations}\label{illustrations}
+\hypertarget{illustrations}{%
+\section{Illustrations}\label{illustrations}}
 
@@
 model. Potentially massive, and thereby unstable, statistics were the
-primary focus of instability studies of Schweinberger
-(\protect\hyperlink{ref-schweinberger2011instability}{2011}) for certain
-discrete exponential models having parameters/statistics of fixed
-dimension. However, as shown in the following, bounded statistics may
-still lead to instability if the parameter dimension is high. We next
-provides some examples to illustrate S-instability in FOES models, which
-also suggest some potential strategies for preventing unstable models.
-
-\subsection{Equi-probability Models}\label{equi-probability-models}
+primary focus of instability studies of
+\citet{schweinberger2011instability} for certain discrete exponential
+models having parameters/statistics of fixed dimension. However, as
+shown in the following, bounded statistics may still lead to instability
+if the parameter dimension is high. We next provides some examples to
+illustrate S-instability in FOES models, which also suggest some
+potential strategies for preventing unstable models.
+
+\hypertarget{equi-probability-models}{%
+\subsection{Equi-probability Models}\label{equi-probability-models}}
 
@@
 
-\subsection{One-parameter Exponential Models}\label{one-param-exp}
-
-A fundamental model considered in the instability work of Schweinberger
-(\protect\hyperlink{ref-schweinberger2011instability}{2011}) involves a
-one-parameter exponential model corresponding to \eqref{eq:expo} with a
-real-valued parameter, say
+\hypertarget{one-param-exp}{%
+\subsection{One-parameter Exponential Models}\label{one-param-exp}}
+
+A fundamental model considered in the instability work of
+\citet{schweinberger2011instability} involves a one-parameter
+exponential model corresponding to \eqref{eq:expo} with a real-valued
+parameter, say
 \(\boldsymbol \theta_N = \eta(\boldsymbol \theta_N)\in \mathbb{R}\), and
@@
 \frac{1}{N}\REP(\boldsymbol \theta_N ) \equiv   |\boldsymbol \theta_N| \frac{(U_N-L_N)}{N} \;\quad \text{(one-parameter exponential model)},
-\end{equation}
-where
+\end{equation} where
 \(U_N \equiv \max_{\boldsymbol x\in\mathcal{X}^N}g_N(\boldsymbol x)\)
@@
 (non-zero) parameter for all \(N \geq 1\), as considered in
-Schweinberger
-(\protect\hyperlink{ref-schweinberger2011instability}{2011}), then an
-S-unstable model results solely if the sufficient statistic admits a
-value \(U_N-L_N\) too large relative to number \(N\) of observations,
-i.e., if \((U_N-L_N)/N\to \infty\) as \(N\to \infty\). The latter aspect
-reflects the definition of Schweinberger
-(\protect\hyperlink{ref-schweinberger2011instability}{2011}), for this
+\citet{schweinberger2011instability}, then an S-unstable model results
+solely if the sufficient statistic admits a value \(U_N-L_N\) too large
+relative to number \(N\) of observations, i.e., if
+\((U_N-L_N)/N\to \infty\) as \(N\to \infty\). The latter aspect reflects
+the definition of \citet{schweinberger2011instability}, for this
 setting, that a real-valued \emph{statistic} \(g_N(\boldsymbol x)\) may
@@
 \(\boldsymbol x =(x_1,\ldots,x_n)\in\{0,1\}^N\), has maximum and minimum
-values \(U_N=N\) and \(L_N=0\). In this case, Schweinberger
-(\protect\hyperlink{ref-schweinberger2011instability}{2011}) has noted
-that the statistic is stable (i.e., bounded \((U_N-L_N)/N=1\)) and the
-Bernoulli model is as well when, in particular,
+values \(U_N=N\) and \(L_N=0\). In this case,
+\citet{schweinberger2011instability} has noted that the statistic is
+stable (i.e., bounded \((U_N-L_N)/N=1\)) and the Bernoulli model is as
+well when, in particular,
 \(\boldsymbol \theta_N=\bold \theta \in\mathbb{R}\) is fixed for
@@
 counts of 2-stars and triangles are unstable statistics in the sense of
-Schweinberger
-(\protect\hyperlink{ref-schweinberger2011instability}{2011}) (i.e.,
+\citet{schweinberger2011instability} (i.e.,
 \(\lim_{N\to \infty} (U_N-L_N)/N=\infty\)). Furthermore, both types of
@@
 \(\boldsymbol \theta_N=\theta\in\mathbb{R}\setminus \{0\}\) from
-Schweinberger
-(\protect\hyperlink{ref-schweinberger2011instability}{2011})).
-
+\citet{schweinberger2011instability}).
+
+\hypertarget{fixed-dim-exp}{%
 \subsection{Fixed-dimensional Linear Exponential
-Models}\label{fixed-dim-exp}
+Models}\label{fixed-dim-exp}}
 
@@
 for fixed \(k \geq 1\). Then, the models \(P_{\boldsymbol \theta_N}\)
-are S-stable if
-\begin{equation}
+are S-stable if \begin{equation}
 \label{eq:prop1}
 \sup_{N \geq 1}\frac{1}{N} \max_{1 \leq i \leq k }|\theta_{i,N}|(U_{i,N}-L_{i,N})<\infty
-\end{equation}
-holds, i.e., if
+\end{equation} holds, i.e., if
 \(\max_{1 \leq i \leq k } |\theta_{i,N}|(U_{i,N}-L_{i,N})/N\) is bounded
@@
 sample size \(N\), \(i=1,\ldots,k\). This supports the findings of
-Schweinberger
-(\protect\hyperlink{ref-schweinberger2011instability}{2011}), who showed
-degeneracy follows in such models under one type of violation of the
-condition \eqref{eq:prop1} in Proposition \ref{prp:prop1} (namely,
-involving \(k>1\) non-zero parameters with \(k-1\) statistics being
-\(O(N)\) bounded while one statistic diverges in maximal size faster
-than the number \(N\) of observations). To further illustrate the result
-in Proposition \ref{prp:prop1}, consider the multinomial distribution
-\eqref{eq:mod11} for \(\boldsymbol X_N=(X_1,\ldots,X_N)\) having
-\(k\geq 2\) categories \(\{1,\ldots,k\}\) and \(k\) parameters
+\citet{schweinberger2011instability}, who showed degeneracy follows in
+such models under one type of violation of the condition \eqref{eq:prop1}
+in Proposition \ref{prp:prop1} (namely, involving \(k>1\) non-zero
+parameters with \(k-1\) statistics being \(O(N)\) bounded while one
+statistic diverges in maximal size faster than the number \(N\) of
+observations). To further illustrate the result in Proposition
+\ref{prp:prop1}, consider the multinomial distribution \eqref{eq:mod11}
+for \(\boldsymbol X_N=(X_1,\ldots,X_N)\) having \(k\geq 2\) categories
+\(\{1,\ldots,k\}\) and \(k\) parameters
 \(\boldsymbol \theta_N = (\theta_{1,N},\ldots,\theta_{k,N})^\prime\).
@@
 \ref{criterion} yields the corresponding \(N\)-scaled log-ratio of
-extreme probabilities \eqref{eq:elpr} as
-\begin{align*}
+extreme probabilities \eqref{eq:elpr} as \begin{align*}
   \frac{1}{N}\REP(\boldsymbol \theta_N) &= \frac{\max_{1 \leq i \leq k} P_{\boldsymbol \theta_N}(X_1=i)}{\min_{1 \leq i \leq k} P_{\boldsymbol \theta_N}(X_1=i)}\\
   &= \max_{1 \leq i \leq k} \theta_{i,N} - \min_{1 \leq i \leq k} \theta_{i,N} \qquad \text{(iid multinomial model)}.
-\end{align*}
-Hence, a multinomial model sequence is unstable (or stable) depending on
-whether (or not) the maximal parameter difference
+\end{align*} Hence, a multinomial model sequence is unstable (or stable)
+depending on whether (or not) the maximal parameter difference
 \(\max_{1 \leq i \leq k} \theta_{i,N} - \min_{1 \leq i \leq k} \theta_{i,N}\)
@@
 \(|\theta_2| + |\theta_3| \neq 0\). Relatedly, a result from
-Schweinberger
-(\protect\hyperlink{ref-schweinberger2011instability}{2011} Result 3)
-states that this model is unstable for all fixed parameters excluding
-cases \(\theta_2 =\theta_3=0\) or \(\theta_2 = - \theta_3/3\). However,
-more is true in line with the instability suggested by Proposition
+\citet[Result 3]{schweinberger2011instability} states that this model is
+unstable for all fixed parameters excluding cases
+\(\theta_2 =\theta_3=0\) or \(\theta_2 = - \theta_3/3\). However, more
+is true in line with the instability suggested by Proposition
 \ref{prp:prop1} whenever \(|\theta_2| + |\theta_3| \neq 0\) (i.e.,
@@
 the exponential graph model \eqref{eq:mod2} can, by definition, be bounded
-below by
-\begin{align*}
+below by \begin{align*}
 \frac{1}{N}\REP_N(\boldsymbol \theta_N) &\geq \max_{i=1,2}\frac{1}{N}
@@
 &= (n-2) \max\left\{ \left| \theta_2 + \frac{\theta_3}{3}+\frac{\theta_1}{n-2} \right|, \frac{n}{4(n-1)} \left| \theta_2 + \frac{8\theta_1}{n-2} \right| \right\};
-\end{align*}
-a similar expression also holds for an odd node number \(n>2\).
-Consequently, for all fixed parameters excluding
+\end{align*} a similar expression also holds for an odd node number
+\(n>2\). Consequently, for all fixed parameters excluding
 \(\theta_2=\theta_3=0\),
@@
 under \(\theta_2 = - \theta_3/3\) case potentially allowed by
-Schweinberger's
-(\protect\hyperlink{ref-schweinberger2011instability}{2011}) results.
-
+Schweinberger's \citeyearpar{schweinberger2011instability} results.
+
+\hypertarget{latent-variable-models-of-increasing-parameter-dimension}{%
 \subsection{Latent Variable Models of Increasing Parameter
-Dimension}\label{latent-variable-models-of-increasing-parameter-dimension}
+Dimension}\label{latent-variable-models-of-increasing-parameter-dimension}}
 
@@
 \(\tilde{P}_{\boldsymbol \theta_N}\) of
-\((\boldsymbol X, \boldsymbol H)\) from \eqref{eq:RBM1} as
-\begin{align*}
+\((\boldsymbol X, \boldsymbol H)\) from \eqref{eq:RBM1} as \begin{align*}
 \REP_{(\boldsymbol X, \boldsymbol H)}(\boldsymbol \theta_N) &=\log\left[  \frac{\max_{ (\boldsymbol x, \boldsymbol h) \in \{\pm 1\}^{N+N_{\mathcal{H}}}}
@@
 &= \left(\max_{ (\boldsymbol x, \boldsymbol h) \in \{\pm 1\}^{N+N_{\mathcal{H}}}} f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)-\min_{ (\boldsymbol x, \boldsymbol h) \in \{\pm 1\}^{N+N_{\mathcal{H}}}} f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)\right),
-\end{align*}
-written as a function
-\begin{equation}
+\end{align*} written as a function \begin{equation}
 \label{eq:f}
 f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)  \equiv  \sum_{i=1}^N x_i  \theta_{i,N}^{\mathcal{V}} +  \sum_{j=1}^{N_\mathcal{H}} h_j \theta_{j,N}^{\mathcal{H}} + \sum_{i=1}^N \sum_{j=1}^{N_\mathcal{H}} x_i h_j  \theta_{ij,N}^{\mathcal{VH}}
-\end{equation}
-of outcomes \(\boldsymbol x =(x_1,\ldots,x_N) \in\{\pm1\}^{N}\),
+\end{equation} of outcomes
+\(\boldsymbol x =(x_1,\ldots,x_N) \in\{\pm1\}^{N}\),
 \(\boldsymbol h =(h_1,\ldots,h_{N_{\mathcal{H}}}) \in\{\pm1\}^{N_{\mathcal{H}}}\)
@@
 visible model as well). Further evidence of this is seen in the
-following numerical example. \par
+following numerical example.
+
+\par
 \begin{figure}
@@
 
-\section{Statistical Consequences of Instability}\label{implications}
+\hypertarget{implications}{%
+\section{Statistical Consequences of Instability}\label{implications}}
 
@@
 
-\subsection{Implications for Simulation}\label{mcmc}
+\hypertarget{mcmc}{%
+\subsection{Implications for Simulation}\label{mcmc}}
 
@@
 for simulation purposes. This result is in line with conclusions of
-Handcock (\protect\hyperlink{ref-handcock2003assessing}{2003}) and
-Schweinberger
-(\protect\hyperlink{ref-schweinberger2011instability}{2011}) for other
-exponential models.
-
+\citet{handcock2003assessing} and \citet{schweinberger2011instability}
+for other exponential models.
+
+\hypertarget{implications-for-maximum-likelihood-inference}{%
 \subsection{Implications for Maximum Likelihood
-Inference}\label{implications-for-maximum-likelihood-inference}
+Inference}\label{implications-for-maximum-likelihood-inference}}
 
@@
 function. We next provide a discussion of this issue in a way that
-builds upon and extends related findings by Schweinberger
-(\protect\hyperlink{ref-schweinberger2011instability}{2011}), who
-largely focused on the case of one-parameter exponential models.
+builds upon and extends related findings by
+\citet{schweinberger2011instability}, who largely focused on the case of
+one-parameter exponential models.
 
@@
 probabilities among outcomes), which can also tether a broad parameter
-search attempted among unstable models. Handcock
-(\protect\hyperlink{ref-handcock2003assessing}{2003}) describes similar
-results for degenerate exponential models, and Theorem
-\ref{thm:something} also supports an important finding of Schweinberger
-(\protect\hyperlink{ref-schweinberger2011instability}{2011} Corollary 1)
-for one-parameter linear exponential models \eqref{eq:expo}. In the latter
-case, the likelihood score function at \(\boldsymbol \theta_N\) is the
-expected value
+search attempted among unstable models. \citet{handcock2003assessing}
+describes similar results for degenerate exponential models, and Theorem
+\ref{thm:something} also supports an important finding of
+\citet[Corollary 1]{schweinberger2011instability} for one-parameter
+linear exponential models \eqref{eq:expo}. In the latter case, the
+likelihood score function at \(\boldsymbol \theta_N\) is the expected
+value
 \(\mu(\boldsymbol \theta_N)\equiv \mathrm{E}_{\boldsymbol \theta_N} g(\boldsymbol X)\)
@@
 \(\boldsymbol x\). For unstable models in this exponential class,
-Schweinberger
-(\protect\hyperlink{ref-schweinberger2011instability}{2011} Corollary 1)
-shows that \[
+\citet[Corollary 1]{schweinberger2011instability} shows that \[
 \lim_{N\to \infty}\frac{\mu(\boldsymbol \theta_N) -L_n}{U_n-L_N}= \left\{ \begin{array}{lcl}
@@
 of the statistic \(g(\boldsymbol x)\),
-\(\boldsymbol x\in\mathcal{X}^N\). As described by Schweinberger
-(\protect\hyperlink{ref-schweinberger2011instability}{2011}), the
-implication for maximum likelihood estimation is that, unless an
-observed outcome \(\boldsymbol x\) falls at an extreme \(U_N, L_N\)
-(i.e., modes/anti-modes), optimization steps in the parameter space can
-iterate in relatively small increments around zero and fail to converge.
-For unstable one-parameter exponential models, the maximum likelihood
-results of Schweinberger
-(\protect\hyperlink{ref-schweinberger2011instability}{2011}) turn out to
-be a special case of Theorem \ref{thm:something} and the LREP expansion
-\eqref{eq:UL} in this setting; namely, for an unstable model with
+\(\boldsymbol x\in\mathcal{X}^N\). As described by
+\citet{schweinberger2011instability}, the implication for maximum
+likelihood estimation is that, unless an observed outcome
+\(\boldsymbol x\) falls at an extreme \(U_N, L_N\) (i.e.,
+modes/anti-modes), optimization steps in the parameter space can iterate
+in relatively small increments around zero and fail to converge. For
+unstable one-parameter exponential models, the maximum likelihood
+results of \citet{schweinberger2011instability} turn out to be a special
+case of Theorem \ref{thm:something} and the LREP expansion \eqref{eq:UL}
+in this setting; namely, for an unstable model with
 \(\boldsymbol \theta_N>0\), \[
@@
 
-\subsection{Implications for Bayes Inference}\label{bayes}
+\hypertarget{bayes}{%
+\subsection{Implications for Bayes Inference}\label{bayes}}
 
@@
 
-\section{Concluding Remarks}\label{conclusions}
+\hypertarget{conclusions}{%
+\section{Concluding Remarks}\label{conclusions}}
 
@@
 
-\appendix
-
-\section{Proofs of instability results}\label{appendix-instab}
+\hypertarget{appendix-appendix}{%
+\appendix}
+
+\hypertarget{appendix-instab}{%
+\section{Proofs of instability results}\label{appendix-instab}}
 
@@
 each outcome \(\boldsymbol x \in \mathcal{X}^N\). Then, if \(k > 0\), it
-follows that
-\begin{align*}
+follows that \begin{align*}
 \REP(\boldsymbol \theta_N) = \log\left[\frac{P_{\boldsymbol \theta_N}(\boldsymbol x_{max})}{P_{\boldsymbol \theta_N}(\boldsymbol x_{min})}\right] &= \left|\sum\limits_{i = 1}^k\log\left(\frac{P_{\boldsymbol \theta_N}(\boldsymbol x_i)}{P_{\boldsymbol \theta_N}(\boldsymbol x_{i-1})}\right)\right| \\
 &\le \sum\limits_{i = 1}^k\left|\log\left(\frac{P_{\boldsymbol \theta_N}(\boldsymbol x_i)}{P_{\boldsymbol \theta}(\boldsymbol x_{i-1})}\right)\right| \le k \Delta_N(\boldsymbol \theta_N) \le NC,
-\end{align*}
-using \(k \le N\) and \(\Delta(\boldsymbol \theta_N) \le C\). If
-\(k = 0\), then \(\boldsymbol x_{max} = \boldsymbol x_{min}\) and the
-same bound above holds. This establishes part (i). To show part (ii),
-note the definition of S-instability (i.e.,
+\end{align*} using \(k \le N\) and
+\(\Delta(\boldsymbol \theta_N) \le C\). If \(k = 0\), then
+\(\boldsymbol x_{max} = \boldsymbol x_{min}\) and the same bound above
+holds. This establishes part (i). To show part (ii), note the definition
+of S-instability (i.e.,
 \(\lim_{N\to \infty}\REP(\boldsymbol \theta_N)/N= \infty\)) combined
@@
 \(1-P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N})\), it
-follows that
-\begin{align*}
+follows that \begin{align*}
 \frac{1}{N}\log\left[\frac{P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N})}{1-P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N})}\right] & \ge \frac{1}{N} \log\left[\frac{P_{\boldsymbol\theta_N}(\boldsymbol x_{max})}{(|\mathcal{X}|^N)[P_{\boldsymbol \theta_N}(\boldsymbol x_{max})]^{1-\epsilon}[P_{\boldsymbol \theta_N}(\boldsymbol x_{min})]^{\epsilon}}\right] \\
 &= \frac{\epsilon}{N} \log\left[\frac{P_{\boldsymbol \theta_N}(\boldsymbol x_{max})}{P_{\boldsymbol \theta_N}(\boldsymbol x_{min})}\right] - \log |\mathcal{X}| \rightarrow \infty
-\end{align*}
-as \(N \rightarrow \infty\) by the definition of an S-unstable FOES
-model \eqref{eq:Sun}. Consequently,
+\end{align*} as \(N \rightarrow \infty\) by the definition of an
+S-unstable FOES model \eqref{eq:Sun}. Consequently,
 \(P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N}) \rightarrow 1\)
@@
 \textbf{Proof of Corollary \ref{cor:sign}.} The model condition PSR
-implies that
-\begin{equation}
+implies that \begin{equation}
 \label{eq:R}
 \frac{\max_{\boldsymbol y \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol y)  }{\min_{\boldsymbol y \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol y) } = \frac{\max_{\boldsymbol y \in \mathcal{X}^N}P_{-\boldsymbol \theta_N}(\boldsymbol y) }{\min_{\boldsymbol y \in \mathcal{X}^N}P_{-\boldsymbol \theta_N}(\boldsymbol y) }
-\end{equation}
-so that the log-ratio
+\end{equation} so that the log-ratio
 \(\REP(\boldsymbol \theta_N)=\REP(-\boldsymbol \theta_N)\) is the same
@@
 \] by model condition PSR and \eqref{eq:R}. The latter is in turn
-equivalent to
-\begin{equation}
+equivalent to \begin{equation}
 \label{eq:R2}
 \log P_{-\boldsymbol \theta_N}(\boldsymbol x) \geq \epsilon \max\limits_{\boldsymbol y \in \mathcal{X}^N} \log  P_{-\boldsymbol \theta_N}(\boldsymbol y) + (1-\epsilon)\min\limits_{\boldsymbol y \in \mathcal{X}^N} \log P_{-\boldsymbol \theta_N}(\boldsymbol y),
-\end{equation}
-so that
+\end{equation} so that
 \(\boldsymbol x \in \mathcal{M}_{\epsilon, \boldsymbol \theta_N}^c\) if
@@
 \leq P_{\boldsymbol \theta_N} (\boldsymbol x_0) \leq  P_{\boldsymbol \theta_N} (\boldsymbol x_M) \leq  2^{N_{\mathcal{H}}} c(\boldsymbol \theta_N) \exp[\max_{\boldsymbol x}\max_{\boldsymbol h}f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)]
-\] and
-\begin{align*}
+\] and \begin{align*}
 c(\boldsymbol \theta_N) \exp[\min_{\boldsymbol x}\max_{\boldsymbol h}f_{\boldsymbol \theta_N} (\boldsymbol x , \boldsymbol h)] &\leq c(\boldsymbol \theta_N) \exp[ \max_{\boldsymbol h}f_{\boldsymbol \theta_N} (\boldsymbol x_m , \boldsymbol h)] \\
@@
 &\leq P_{\boldsymbol \theta_N} (\boldsymbol x_1) \\ 
-&\leq 2^{N_{\mathcal{H}}} c(\boldsymbol \theta_N) \exp[ \max_{\boldsymbol h}f_{\boldsymbol \theta_N} (\boldsymbol x_1 , \boldsymbol h)]\\&=&2^{N_{\mathcal{H}}} c(\boldsymbol \theta_N) \exp[\min_{\boldsymbol x}\max_{\boldsymbol h}f_{\boldsymbol \theta_N} (\boldsymbol x , \boldsymbol h)]
+&\leq 2^{N_{\mathcal{H}}} c(\boldsymbol \theta_N) \exp[ \max_{\boldsymbol h}f_{\boldsymbol \theta_N} (\boldsymbol x_1 , \boldsymbol h)]\\
+&=2^{N_{\mathcal{H}}} c(\boldsymbol \theta_N) \exp[\min_{\boldsymbol x}\max_{\boldsymbol h}f_{\boldsymbol \theta_N} (\boldsymbol x , \boldsymbol h)]
 \end{align*}
@@
 a_{\boldsymbol \theta_N, \mathcal{H}} (\boldsymbol h) \equiv \sum_{i=1}^N \left| \theta_{i,N}^{\mathcal{V}} + \sum_{j=1}^{N_\mathcal{H}}  h_j  \theta_{ij,N}^{\mathcal{VH}}\right|, &  b_{\boldsymbol \theta_N, \mathcal{V}} (\boldsymbol x) \equiv \sum_{j=1}^{N_\mathcal{H}} \left| \theta_{j,N}^{\mathcal{H}}  + \sum_{i=1}^N  x_i  \theta_{ij,N}^{\mathcal{VH}}\right|,
-\end{align}
-where
+\end{align} where
 \(\boldsymbol h^T \boldsymbol \theta_N^{\mathcal{H}}=\sum_{j=1}^{N_\mathcal{H}} h_j \theta_{j,N}^{\mathcal{H}}\),
@@
 \(\Gam\equiv \max_{\boldsymbol h} a_{\boldsymbol \theta_N, \mathcal{H}} (\boldsymbol h)\).
-From this, it follows that
-\begin{align*}
+From this, it follows that \begin{align*}
 \REP_{(\boldsymbol X, \boldsymbol H)}(\boldsymbol \theta_N) &=  \max_{\boldsymbol h}\max_{\boldsymbol x}f_{\boldsymbol \theta_N} (\boldsymbol x , \boldsymbol h) -  \min_{\boldsymbol h}\min_{\boldsymbol x}f_{\boldsymbol \theta_N} (\boldsymbol x , \boldsymbol h)\\
 &= \max_{\boldsymbol h_1 }  \max_{\boldsymbol h_2 }\left[ (\boldsymbol h_1 - \boldsymbol h_2)^T \boldsymbol \theta_N^{\mathcal{H}}  +   a_{\boldsymbol \theta_N, \mathcal{H}} (\boldsymbol h_1)  + a_{\boldsymbol \theta_N, \mathcal{H}} (\boldsymbol h_2)\right],
-\end{align*}
-which leads to the upper bound
+\end{align*} which leads to the upper bound
 \(\REP_{(\boldsymbol X, \boldsymbol H)}(\boldsymbol \theta_N) \leq 2 \Gam + 2 |\boldsymbol \theta_N^{\mathcal{H}} |_1\).
@@
 
-We next consider \(\elt\) and, by \eqref{eq:max}, write
-\begin{align*}
+We next consider \(\elt\) and, by \eqref{eq:max}, write \begin{align*}
 \elt &=  \max_{\boldsymbol h}\max_{\boldsymbol x}f_{\boldsymbol \theta_N} (\boldsymbol x , \boldsymbol h) -  \max_{\boldsymbol h}\min_{\boldsymbol x}f_{\boldsymbol \theta_N} (\boldsymbol x , \boldsymbol h)\\
 &= \max_{\boldsymbol h_1} \min_{\boldsymbol h_2 }\left[ (\boldsymbol h_1 - \boldsymbol h_2)^T \boldsymbol \theta_N^{\mathcal{H}} + a_{\boldsymbol \theta_N, \mathcal{H}} (\boldsymbol h_1)  + a_{\boldsymbol \theta_N, \mathcal{H}} (\boldsymbol h_2)\right].
-\end{align*}
-Taking \(\boldsymbol h_1=\boldsymbol h_2\) and maximizing over both
-\(\boldsymbol h_1,\boldsymbol h_2\) produces the upper bound
+\end{align*} Taking \(\boldsymbol h_1=\boldsymbol h_2\) and maximizing
+over both \(\boldsymbol h_1,\boldsymbol h_2\) produces the upper bound
 \(\elt \leq 2\Gam\). Then, using
@@
 Finally, note that for any \(\boldsymbol h\), the triangle inequality
-gives
-\begin{align*}
+gives \begin{align*}
 \Gam  \equiv \max_{\boldsymbol h_1}  a_{\boldsymbol \theta_N, \mathcal{H}} (\boldsymbol h_1) &\geq
@@
 & \geq \sum_{i=1}^{N }  \left| \theta_{i,N}^{\mathcal{V}} \right| \equiv |\boldsymbol \theta_{N}^{\mathcal{V}}|_1.
-\end{align*}
-\hfill \(\Box\)
-
+\end{align*} \hfill \(\Box\)
 \textbf{Proof of Theorem \ref{thm:something}.} Let
@@
 
-\section*{References}\label{references}
-\addcontentsline{toc}{section}{References}
-
-\hypertarget{refs}{}
-\hypertarget{ref-besag1974spatial}{}
-Besag, Julian. 1974. ``Spatial Interaction and the Statistical Analysis
-of Lattice Systems.'' \emph{Journal of the Royal Statistical Society.
-Series B (Methodological)}. JSTOR, 192--236.
-
-\hypertarget{ref-handcock2003assessing}{}
-Handcock, Mark S. 2003. ``Assessing Degeneracy in Statistical Models of
-Social Networks.'' Center for Statistics; the Social Sciences,
-University of Washington. \url{http://www.csss.washington.edu/}.
-
-\hypertarget{ref-hinton2006fast}{}
-Hinton, Geoffrey E, Simon Osindero, and Yee-Whye Teh. 2006. ``A Fast
-Learning Algorithm for Deep Belief Nets.'' \emph{Neural Computation} 18
-(7). MIT Press: 1527--54.
-
-\hypertarget{ref-neal1992connectionist}{}
-Neal, Radford M. 1992. ``Connectionist Learning of Belief Networks.''
-\emph{Artificial Intelligence} 56 (1). Elsevier: 71--113.
-
-\hypertarget{ref-pearl985bayesian}{}
-Pearl, Judea. 1985. ``Bayesian Networks: A Model of Self-Activated
-Memory for Evidential Reasoning.'' UCLA Computer Science Department.
-
-\hypertarget{ref-ruelle1999statistical}{}
-Ruelle, D. 1999. \emph{Statistical Mechanics: Rigorous Results}. London:
-Imperial College Press.
-
-\hypertarget{ref-salakhutdinov2009deep}{}
-Salakhutdinov, Ruslan, and Geoffrey E Hinton. 2009. ``Deep Boltzmann
-Machines.'' In \emph{International Conference on Artificial Intelligence
-and Statistics}, 448--55. AI \& Statistics.
-
-\hypertarget{ref-schweinberger2011instability}{}
-Schweinberger, Michael. 2011. ``Instability, Sensitivity, and Degeneracy
-of Discrete Exponential Families.'' \emph{Journal of the American
-Statistical Association} 106 (496). Taylor \& Francis: 1361--70.
-
-\hypertarget{ref-smolensky1986information}{}
-Smolensky, Paul. 1986. ``Information Processing in Dynamical Systems:
-Foundations of Harmony Theory.'' DTIC Document.
-
-\hypertarget{ref-wasserman1994social}{}
-Wasserman, Stanley, and Katherine Faust. 1994. \emph{Social Network
-Analysis: Methods and Applications}. Vol. 8. Cambridge: Cambridge
-University Press.
+\bibliographystyle{imamat}
+\bibliography{../resources/latex/rbm.bib}
 

### 2018-04-11 writing/paper-note/index.Rmd
getting ready for another submission

--- 2017-11-07 writing/paper-note/index.Rmd
@@
 title: "On the instability and degeneracy of deep learning models"
+shorttitle: "Instability of deep learning models"
+shortauthor: "Kaplan, Nordman, and Vardeman"
 author: 
   - name: Andee Kaplan
-    university: Iowa State University
-    department: Department of Statistics
+    corresponding: true
+    address: Department of Statistical Science, Duke University, P.O. Box 90251, Durham, NC 27708-0251, USA
+    email: andrea.kaplan@duke.edu
   - name: Daniel Nordman
-    university: Iowa State University
-    department: Department of Statistics
+    address: Department of Statistics, Iowa State University, 2438 Osborn Dr., Ames, IA 50011-1090, USA
+    email: dnordman@iastate.edu
   - name: Stephen Vardeman
-    university: Iowa State University
-    department: |
-      | Department of Statistics and Department of Industrial and
-      | Manufacturing Systems Engineering
+    address: Departments of Statistics and Industrial and Manufacturing Systems Engineering, Iowa State University, 2438 Osborn Dr., Ames, IA 50011-1090, USA
+    email: vardeman@iastate.edu
 site: bookdown::bookdown_site
@@
     includes:
-      in_header: ../resources/latex/header.tex
+      in_header: ../resources/latex/header2.tex
     toc: no
-bibliography: [../resources/latex/refs_rbm.bib]
+bibliography: "../resources/latex/rbm.bib"
 github-repo: andeek/rbm
@@
 keywords: Degeneracy, Instability, Classification, Deep Learning, Graphical Models
-blind: false
 ---

### 2018-05-08 presentations/msmlc/slides.Rmd
msmlc

--- 2017-03-25 presentations/msu/slides.Rmd
@@
 date: |
-  | February 14, 2017
+  | May 9, 2017
   |
-  | Slides available at <http://bit.ly/kaplan-msu>
+  | Slides available at <http://bit.ly/kaplan-msmlc>
   | 
   | \footnotesize Joint work with D. Nordman and S. Vardeman
-shortdate: "February 14, 2017"
+shortdate: "May 9, 2017"
 output: 
@@
 
+\vspace{-.1in}
 \begin{figure}
@@
 \includegraphics[width=.35\linewidth]{../resources/images/toyhull_top_2.png}
-\caption{The convex hull of the "statistic space" $\mathcal{T} = \{(v_1, h_1, v_1 h_1): v_1, h_1 \in \mathcal{C}\}$ in three dimensions for a toy RBM with one visible and one hidden node for $\mathcal{C} = \{0,1\}$ (left) and $\mathcal{C} = \{-1,1\}$ (right) data encoding.}
+\caption{Convex hull of the "statistic space" $\mathcal{T} = \{(v_1, h_1, v_1 h_1): v_1, h_1 \in \mathcal{C}\}$ for a toy RBM with one visible and one hidden node for $\mathcal{C} = \{0,1\}$ (left) and $\mathcal{C} = \{-1,1\}$ (right) data encoding.}
 \end{figure}
+\vspace{-.15in}
+- For the $\mathcal{C} = \{-1, 1 \}$ encoding of hiddens $(H_1, \dots, H_\nh)$ and visibles $(V_1, \dots, V_\nv)$, the origin is the center of the parameter space.
+- At $\boldsymbol \theta = \boldsymbol 0$, RBM is equivalent to elements of $X$ being distributed as iid Bernoulli$\left(\frac{1}{2}\right)$ $\Rightarrow$ No *near-degeneracy*, *instability*, or *uninterpretability*!
 
+<!--
 The convex hull of $\mathcal{T} \subset \mathcal{C}^3$ does not fill the unit cube $[0,1]^3$ (left), but does better with $[-1,1]^3$ (right).
@@
 # The center of the universe
-
-- For the $\mathcal{C} = \{-1, 1 \}$ encoding of hiddens $(H_1, \dots, H_\nh)$ and visibles $(V_1, \dots, V_\nv)$, the origin is the center of the parameter space.
-- At $\boldsymbol \theta = \boldsymbol 0$, RBM is equivalent to elements of $X$ being distributed as iid Bernoulli$\left(\frac{1}{2}\right)$ $\Rightarrow$ No *near-degeneracy*, *instability*, or *uninterpretability*!
 
@@
 ```
+-->
 
@@
 
-* Slides -- <http://bit.ly/kaplan-msu>
+* Slides -- <http://bit.ly/kaplan-msmlc>
 

### 2018-05-08 presentations/msmlc/slides.Rmd
small typos

@@
 institute: |
-    | Iowa State University
-    | ajkaplan@iastate.edu
-shortinstitute: ajkaplan@iastate.edu
+    | Duke University
+    | andrea.kaplan@duke.edu
+shortinstitute: andrea.kaplan@duke.edu
 date: |
-  | May 9, 2017
+  | May 9, 2018
   |
@@
   | \footnotesize Joint work with D. Nordman and S. Vardeman
-shortdate: "May 9, 2017"
+shortdate: "May 9, 2018"
 output: 
@@
 \begin{align*}
-\lim\limits_{\nv \rightarrow \infty} \frac{1}{\nv} \text{ELPR}(\boldsymbol \theta_{\nv}) = \infty.
+\lim\limits_{\nv \rightarrow \infty} \frac{1}{\nv} \text{LREP}(\boldsymbol \theta_{\nv}) = \infty.
 \end{align*}
@@
 \label{eqn:elpr}
-\text{ELPR}(\boldsymbol \theta_{\nv}) = \log \left[\frac{\max\limits_{(v_1, \dots, v_{\nv}) \in \mathcal{C}^\nv}P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}{\min\limits_{(v_1, \dots, v_\nv) \in \mathcal{C}^\nv}P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}\right]
+\text{LREP}(\boldsymbol \theta_{\nv}) = \log \left[\frac{\max\limits_{(v_1, \dots, v_{\nv}) \in \mathcal{C}^\nv}P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}{\min\limits_{(v_1, \dots, v_\nv) \in \mathcal{C}^\nv}P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}\right]
 \end{align}
@@
 \label{prop:instab}
-Let $c > 0$ and fix an integer $\nv \ge 1$. If $\frac{1}{\nv}\text{ELPR}(\boldsymbol \theta_\nv) > c$, then $\Delta(\boldsymbol \theta_\nv) > c$.
+Let $c > 0$ and fix an integer $\nv \ge 1$. If $\frac{1}{\nv}\text{LREP}(\boldsymbol \theta_\nv) > c$, then $\Delta(\boldsymbol \theta_\nv) > c$.
 \end{result}
 
-If the $\nv^{-1}ELPR(\boldsymbol\theta_{\nv})$ is too large, then a RBM model will exhibit large probability shifts for very small changes in the data configuration.
+If the $\nv^{-1}LREP(\boldsymbol\theta_{\nv})$ is too large, then a RBM model will exhibit large probability shifts for very small changes in the data configuration.
 
@@
 Average magnitudes vary on a grid between `r min(plot_dat_grid$r1/(plot_dat_grid$H + plot_dat_grid$V))` and `r max(plot_dat_grid$r1/(plot_dat_grid$H + plot_dat_grid$V))` with `r length(unique(plot_dat_grid$r1/(plot_dat_grid$H + plot_dat_grid$V)))` breaks, yielding `r length(unique(plot_dat_grid$r1/(plot_dat_grid$H + plot_dat_grid$V)))^2` grid points
-- Calculate metrics of model impropriety, $\boldsymbol \mu(\boldsymbol \theta)$, $\text{ELPR}(\boldsymbol \theta)/\nv$, and the coordinates of $\left|\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] - \text{E}\left[\boldsymbol X | \boldsymbol \theta^* \right] \right|$.
+- Calculate metrics of model impropriety, $\boldsymbol \mu(\boldsymbol \theta)$, $\text{LREP}(\boldsymbol \theta)/\nv$, and the coordinates of $\left|\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] - \text{E}\left[\boldsymbol X | \boldsymbol \theta^* \right] \right|$.
 - In the case of *near-degeneracy*, classify each model as near-degenerate or "viable" based on the distance of $\boldsymbol \mu(\boldsymbol \theta)$ from the boundary of the convex hull of $\mathcal{T}$
@@
   geom_abline(aes(intercept = 0, slope = 1), alpha = .5, lty = 2) +
-  scale_fill_gradient(expression(frac(bar(ELPR(theta)), n[v])), low = "yellow", high = "red") +
+  scale_fill_gradient(expression(frac(bar(LREP(theta)), n[v])), low = "yellow", high = "red") +
   facet_grid(Visibles~Hiddens, labeller = label_parsed) +
@@
 
-```{r, fig.show="hold", fig.width=3, out.width='.32\\linewidth', fig.cap="\\label{fig:degen_plots}The fraction of models that were near-degenerate (left), the sample mean value of $\\text{ELPR}(\\boldsymbol \\theta)/\\nv$ (middle), and the sample mean of the maximum component of the absolute difference between the model expectation vector, E$\\left[\\boldsymbol X | \\boldsymbol \\theta\\right]$, and the expectation vector given independence, E$\\left[\\boldsymbol X | \\boldsymbol \\theta^* \\right ]$ (right)."}
+```{r, fig.show="hold", fig.width=3, out.width='.32\\linewidth', fig.cap="\\label{fig:degen_plots}The fraction of models that were near-degenerate (left), the sample mean value of $\\text{LREP}(\\boldsymbol \\theta)/\\nv$ (middle), and the sample mean of the maximum component of the absolute difference between the model expectation vector, E$\\left[\\boldsymbol X | \\boldsymbol \\theta\\right]$, and the expectation vector given independence, E$\\left[\\boldsymbol X | \\boldsymbol \\theta^* \\right ]$ (right)."}
 p.degen
@@
 1. Generalization of instability results for other network models [ongoing, see @kaplan2016note]
+\vfill
 2. Image classification
@@
     - Decision theoretic based approach to approximating the likelihood ratio test for classification
-3. Markov chain Monte Carlo methods for data with Markovian dependence
-    - Spatial data
-    - Network data
+\vfill
 
@@
 * Contact
-    * Email -- <ajkaplan@iastate.edu>
+    * Email -- <andrea.kaplan@duke.edu>
     * Twitter -- <http://twitter.com/andeekaplan>

### 2018-09-01 writing/paper-rbm/chapters/rbm.Rmd
trying out some things

@@
 However, the model properties are largely unexplored in the literature and the commonly cited fitting methodology remains heuristic-based and relies on rough approximation [@hinton2006fast].
-In this paper, we provide steps toward a fuller understanding of the model class and its properties from the perspective of statistical model theory, and we then explore the possibility of a rigorous fitting methodology.
+Additionally, the current fitting methodology does not easily allow for interval estimation of parameters in the model, and instead only provides for point estimation.
+In this paper, we provide steps toward a fuller understanding of the model class and its properties from the perspective of statistical model theory, and we then explore the possibility of a rigorous fitting methodology with natural uncertainty quantification.
 We find the RBM model class to be concerning in two fundamental ways.
@@
 
-In addition to model properties, we also explore the quality of estimation in fitting these models.
-The fitting can be problematic for two reasons, the first being computational and the second concerning flexibility.
+Separate of the exploration of model properties, we also investigate the quality of estimation in fitting these models and propose an estimation procedure that will allow for uncertainty quantification rather than point estimates alone.
+The fitting of RBMs can be problematic for two reasons, the first being computational and the second concerning flexibility.
 As the size of these models grows, both maximum likelihood and Bayesian methods of fitting quickly become intractable.
-The literature often suggests Markov chain Monte Carlo (MCMC) tools for approximate maximization of likelihoods to fit these models (e.g., Gibbs sampling to exploit conditional structure in hidden and visible variables), but little is said about the attributes of realized estimates [@hinton2010practical; @hinton2006fast].
-Related to this, these MCMC algorithms require updating potentially many latent variables (hiddens) which can critically influence convergence in MCMC-based likelihood methods.
+The literature often suggests Markov chain Monte Carlo (MCMC) tools for approximate maximization of likelihoods to fit these models (e.g., Gibbs sampling to exploit conditional structure in hidden and visible variables), but little is said about the attributes of realized estimates [@hinton2010practical; @hinton2006fast], including the absence of interval estimates.
+
+Furthermore, these MCMC algorithms require updating potentially many latent variables (hiddens) which can critically influence convergence in MCMC-based likelihood methods.
 Applying basic statistical principles in fitting RBM models of tractable sizes, we compare three fully Bayesian techniques involving MCMC, which are computationally more accessible than direct maximum likelihood and also aim to avoid parts of a RBM parameter space that yield unattractive models.
@@
 Section \@ref(explorations-of-model-properties-through-simulation) details our explorations into the model behavior and potential propriety issues with the RBM class.
-We examine three Bayesian fitting techniques intended to avoid model impropriety issues raised in Section \@ref(model-fitting) and conclude with a discussion in Section \@ref(discussion).
+This follows with the call to action in @sculley2018winner to employ small toy examples to advance empirical understanding of deep learning models and strive for a "higher level of empirical rigor in the field." We examine three Bayesian fitting techniques intended to avoid model impropriety issues raised in Section \@ref(model-fitting) and conclude with a discussion in Section \@ref(discussion).
 Supplementary online material provides proofs for results on RBM parameterizations and data codings described in Section \@ref(data-encoding).
@@
 $$ 
-is the normalizing function that ensures the probabilities (\@ref(eq:pmf)) sum to one. For $\boldsymbol x = (h_1, \dots, h_\nh, v_1, \dots, v_\nh) \in \mathcal{C}^{\nv + \nh}$ and 
+is the normalizing function that ensures the probabilities \@ref(eq:pmf) sum to one. For $\boldsymbol x = (h_1, \dots, h_\nh, v_1, \dots, v_\nh) \in \mathcal{C}^{\nv + \nh}$ and 
 \begin{align}
@@
 \end{align}
-let $\mathcal{T} = \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{C}^{\nv + \nh}\} \subset \mathbb{R}^{\nv + \nh + \nv * \nh}$ be the set of possible values for the vector of variables needed to compute probabilities (\@ref(eq:pmf)) in the model, and write $Q_{\boldsymbol \theta}(\boldsymbol x) = \sum\limits_{i = 1}^\nh\sum\limits_{j=1}^\nv \theta_{ij} h_i v_j + \sum\limits_{i = 1}^\nh\theta_{hi} h_i + \sum\limits_{j = 1}^\nv\theta_{vj} v_j$ for the "neg-potential" function.
+let $\mathcal{T} = \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{C}^{\nv + \nh}\} \subset \mathbb{R}^{\nv + \nh + \nv * \nh}$ be the set of possible values for the vector of variables needed to compute probabilities \@ref(eq:pmf) in the model, and write $Q_{\boldsymbol \theta}(\boldsymbol x) = \sum\limits_{i = 1}^\nh\sum\limits_{j=1}^\nv \theta_{ij} h_i v_j + \sum\limits_{i = 1}^\nh\theta_{hi} h_i + \sum\limits_{j = 1}^\nv\theta_{vj} v_j$ for the "neg-potential" function.
 The RBM model is parameterized by $\boldsymbol \theta$ containing two types of parameters, main effects and interaction effects.
-The main effects parameters ($\{\theta_{v_i}, \theta_{h_j}\}_{\substack{i = 1, \dots, \nv,\\j = 1, \dots, \nh}}$) weight the values of the visible $v_i$ and hidden $h_j$ nodes in probabilities (\@ref(eq:pmf)) and the interaction effect parameters ($\theta_{ij}$) weight the values of the connections $v_i h_j$, or dependencies, between hidden and visible layers.
+The main effects parameters ($\{\theta_{v_i}, \theta_{h_j}\}_{\substack{i = 1, \dots, \nv,\\j = 1, \dots, \nh}}$) weight the values of the visible $v_i$ and hidden $h_j$ nodes in probabilities \@ref(eq:pmf) and the interaction effect parameters ($\theta_{ij}$) weight the values of the connections $v_i h_j$, or dependencies, between hidden and visible layers.
 
@@
 Let $C$ denote the convex hull of the potential outcomes of sufficient statistics, $\{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{X}\}$, under the model above.
-@handcock2003assessing classifies an exponentially parametrized random graph model at $\boldsymbol \theta$ as *near-degenerate* if the mean value of the vector of sufficient statistics under $\boldsymbol \theta$, $\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta}\boldsymbol t( \boldsymbol X)$, is close to the boundary of $C$.
+@handcock2003assessing classifies an exponentially parameterized random graph model at $\boldsymbol \theta$ as *near-degenerate* if the mean value of the vector of sufficient statistics under $\boldsymbol \theta$, $\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta}\boldsymbol t( \boldsymbol X)$, is close to the boundary of $C$.
 Intuitively, if a model is near-degenerate in the sense that only a small number of elements of the sample space $\mathcal{X}$ have positive probability, the expected value $\text{E}_{\boldsymbol \theta}\boldsymbol t( \boldsymbol X)$ is an average of that same small number of values of $t( \boldsymbol x)$ (defining the boundary of the hull $C$) and can be expected to *not* be pulled deep into the interior of $C$.
@@
 A RBM model can be thought to exhibit an analogous form of *near-degeneracy* when there is a disproportionate amount of probability placed on a small number of elements in the sample space of visibles and hiddens, $\mathcal{C}^{\nv + \nh}$.
-Using the idea of @handcock2003assessing, when the random vector $\boldsymbol t(\boldsymbol x) = (v_1, \dots, v_\nv, h_1, \dots, h_\nh, v_1 h_1, \dots, v_V h_\nh ) \in \mathcal{T} \equiv \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{C}^{\nh + \nv} \}$ from (\@ref(eq:t)), appearing in the neg-potential function $Q_{\boldsymbol \theta}(\cdot)$, has a mean vector $\boldsymbol \mu(\boldsymbol \theta) \in \mathbb{R}^{\nv + \nh + \nv * \nh}$ close to the boundary of the convex hull of $\mathcal{T}$, then the RBM model can be said to exhibit near-degeneracy at $\boldsymbol \theta \in \mathbb{R}^{\nv + \nh + \nh*\nv}$.
+Using the idea of @handcock2003assessing, when the random vector $\boldsymbol t(\boldsymbol x) = (v_1, \dots, v_\nv, h_1, \dots, h_\nh, v_1 h_1, \dots, v_V h_\nh ) \in \mathcal{T} \equiv \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{C}^{\nh + \nv} \}$ from \@ref(eq:t), appearing in the neg-potential function $Q_{\boldsymbol \theta}(\cdot)$, has a mean vector $\boldsymbol \mu(\boldsymbol \theta) \in \mathbb{R}^{\nv + \nh + \nv * \nh}$ close to the boundary of the convex hull of $\mathcal{T}$, then the RBM model can be said to exhibit near-degeneracy at $\boldsymbol \theta \in \mathbb{R}^{\nv + \nh + \nh*\nv}$.
 Here the mean of $\boldsymbol t(\boldsymbol x)$ is
@@
 
-In formulating a RBM model for a potentially large number of visibles (i.e., as $\nv \rightarrow \infty$), we will say that the ratio (\@ref(eq:elpr)) needs to stay bounded for a sequence of RBM models to be stable.
+In formulating a RBM model for a potentially large number of visibles (i.e., as $\nv \rightarrow \infty$), we will say that the ratio \@ref(eq:elpr) needs to stay bounded for a sequence of RBM models to be stable.
 That is, we make the following convention.
@@
 ```{proposition, label = "instab1", echo=TRUE}
-Let $c > 0$ and let $\text{LREP}(\boldsymbol \theta_\nv)$ be as in (\@ref(eq:elpr)) for an integer $\nv \ge 1$.
+Let $c > 0$ and let $\text{LREP}(\boldsymbol \theta_\nv)$ be as in \@ref(eq:elpr) for an integer $\nv \ge 1$.
 If $\frac{1}{\nv}\text{LREP}(\boldsymbol \theta_\nv) > c$, then $\Delta(\boldsymbol \theta_\nv) > c$.
@@
 
-In other words, if the probability ratio (\@ref(eq:elpr)) is too large, then a RBM model sequence will exhibit large probability shifts for very small changes in data configurations (i.e., will exhibit instability).
+In other words, if the probability ratio \@ref(eq:elpr) is too large, then a RBM model sequence will exhibit large probability shifts for very small changes in data configurations (i.e., will exhibit instability).
 Such instability can be a concern in a model for reasons similar to those related to degeneracy: as outcomes vary over the sample space, the geography of probabilities is extremely rugged, with deep pits following sharp mountains.
@@
 Note that in a RBM, the parameters $\theta_{v_1}, \dots, \theta_{v_\nv}$ and $\theta_{h_1}, \dots, \theta_{h_\nh}$ are naturally associated with main effects of visible and hidden variables and can be interpreted as (logit functions of) means for variables $V_1, \dots, V_\nv, H_1, \dots, H_\nh$ in a model with no interaction parameters, $\theta_{ij} = 0, i = 1, \dots, \nv, j = 1, \dots, \nh$.
-That is, with no interaction parameters, we have from (\@ref(eq:pmf)) that
+That is, with no interaction parameters, we have from \@ref(eq:pmf) that
 $$
@@
 
-Secondly, we can look at $\text{LREP}(\boldsymbol \theta)$ from (\@ref(eq:elpr)) for this tiny model in order to consider model instability as a function of RBM parameters.
+Secondly, we can look at $\text{LREP}(\boldsymbol \theta)$ from \@ref(eq:elpr) for this tiny model in order to consider model instability as a function of RBM parameters.
 Recall that large values of $\text{LREP}(\boldsymbol \theta)$ are associated with an extreme sensitivity of the model probabilities $f_{\boldsymbol \theta}(\boldsymbol x)$ to small changes in $\boldsymbol x$ (see Proposition \@ref(prp:instab1)).
@@
 For example, we could allow hiddens $(H_1, \dots, H_\nh) \in \{0,1\}^\nh$ and visibles $(V_1, \dots, V_\nv) \in \{0,1\}^\nv$, as in the previous sections or we could instead encode the state of the variables as $\{-1,1\}^\nh$ and $\{-1,1\}^\nv$.
-This will result in variables $\boldsymbol t(\boldsymbol X)$ from (\@ref(eq:t)) satisfying $\boldsymbol t(\boldsymbol x) \in \{0,1\}^{\nh + \nv + \nh*\nv}$ or $\boldsymbol t(\boldsymbol x) \in \{-1,1\}^{\nh + \nv + \nh*\nv}$ depending on how we encode "on" and "off" states in the nodes.
+This will result in variables $\boldsymbol t(\boldsymbol X)$ from \@ref(eq:t) satisfying $\boldsymbol t(\boldsymbol x) \in \{0,1\}^{\nh + \nv + \nh*\nv}$ or $\boldsymbol t(\boldsymbol x) \in \{-1,1\}^{\nh + \nv + \nh*\nv}$ depending on how we encode "on" and "off" states in the nodes.
 
@@
 
-Typically, fitting a RBM via maximum likelihood (ML) methods will be infeasible due mainly to the intractability of the normalizing term $\gamma(\boldsymbol \theta)$ in a model (\@ref(eq:pmf)) of any realistic size.
+Typically, fitting a RBM via maximum likelihood (ML) methods will be unfeasible due mainly to the intractability of the normalizing term $\gamma(\boldsymbol \theta)$ in a model \@ref(eq:pmf) of any realistic size.
 Ad hoc methods are often recommended instead, which aim to avoid this problem by using stochastic ML approximations that employ a small number of MCMC draws (i.e., contrastive divergence, [@hinton2002training]).
@@
 for the outcomes $(\pm 1, \pm 1)$, the online appendix provides technical details to show parameter values $(\theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21})$ may be chosen to match any given cell probabilities with arbitrary closeness.
-In fact, when the cell probabilities (\@ref(eq:0)) are all strictly positive, parameters in the RBM model can be specified to reflect these probabilities exactly.
-And, when one or more of the cell probabilities (\@ref(eq:0)) are zero, the corresponding RBM probabilities $P(V_1=v_1,V_2=_v2| \theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21})$ may never be identically zero (due to exponential terms in the model) but parameters can be still selected to make the appropriate RBM cell probabilities arbitrarily small.
+In fact, when the cell probabilities \@ref(eq:0) are all strictly positive, parameters in the RBM model can be specified to reflect these probabilities exactly.
+And, when one or more of the cell probabilities \@ref(eq:0) are zero, the corresponding RBM probabilities $P(V_1=v_1,V_2=_v2| \theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21})$ may never be identically zero (due to exponential terms in the model) but parameters can be still selected to make the appropriate RBM cell probabilities arbitrarily small.
 Furthermore, as development in the online appendix shows, not only is it possible to approximate any distribution on the visibles arbitrarily well [cf.
@@
 That is, based on a random sample of vectors of visible variables, the model for the cell probabilities with the highest likelihood over  *all possible model classes* (i.e., RBM-based or not) is the empirical distribution, and the over-parameterization of the RBM model itself ensures that this empirical distribution can be arbitrarily well-approximated.
-For illustration, considering the simple example from above with $n$ iid observations, each consisting of two realized visibles $(V_1,V_2)$ the empirical cell frequencies from the sample are the highest likelihood choices for the cell probabilities $p_{(-1,-1)},p_{(1,-1)}, p_{(-1,1)},p_{(1,1)}$ in (\@ref(eq:0)) before any parametric model is introduced for improved analysis, and the discussion above indicates that RBM model parameters can be chosen to re-create this empirical distribution to an arbitrarily close degree.
+For illustration, considering the simple example from above with $n$ iid observations, each consisting of two realized visibles $(V_1,V_2)$ the empirical cell frequencies from the sample are the highest likelihood choices for the cell probabilities $p_{(-1,-1)},p_{(1,-1)}, p_{(-1,1)},p_{(1,1)}$ in \@ref(eq:0) before any parametric model is introduced for improved analysis, and the discussion above indicates that RBM model parameters can be chosen to re-create this empirical distribution to an arbitrarily close degree.
 Not only does RBM model fitting based on ML seek to reproduce the empirical distribution, whenever this empirical distribution contains empty cells, fitting steps for the RBM model will further aim to choose parameters that necessarily diverge to infinity in magnitude in order to zero-out the corresponding RBM cell probabilities.
@@
 Here we consider what might be done in a principled manner to prevent both overfitting and model impropriety, testing on a $\nv = \nh = 4$ case that already stretches the limits of what is computable - in particular we consider Bayes methods.
+We employ a Bayesian framework because of the ability to obtain uncertainty quantification with little to no extra effort via posterior credible intervals.
 
@@
 1.
-*A "trick" prior.* Here we cancel out normalizing term in the likelihood (from $\gamma(\boldsymbol \theta)$ in (\@ref(eq:pmf))) so that resulting full conditionals of $\boldsymbol \theta$ are multivariate Normal.
+*A "trick" prior.* Here we cancel out normalizing term in the likelihood (from $\gamma(\boldsymbol \theta)$ in \@ref(eq:pmf)) so that resulting full conditionals of $\boldsymbol \theta$ are multivariate Normal.
 Namely, this involves a prior of the form
@@
 Full conditional distributions are not conjugate, and simulation from the posterior was accomplished using a geometric adaptive Metropolis Hastings step [@zhou2014some] and calculation of likelihood normalizing constant.
-(This computation is barely feasible for a problem of this size and would be infeasible for larger problems.) Here the hidden variables $h_j$ are again carried along in the MCMC implementation as latent variables.
+(This computation is barely feasible for a problem of this size and would be unfeasible for larger problems.) Here the hidden variables $h_j$ are again carried along in the MCMC implementation as latent variables.
 We will refer to this method as Bayes with Truncated Normal prior and Latent Variables (BwTNLV).
@@
 After a burn-in period of 50 iterations selected by inspecting the trace plots, we assess the issue of mixing in two ways.
-First, the autocorrelation functions (ACF) from each posterior sample corresponding to a model probability for a visible vector outcome $\mathbf{v}=(v_1,v_2,v_3,v_4)\in\{\pm 1\}^4$ (i.e., computed from $\boldsymbol \theta$ under (\@ref(eq:pmf))) are determined and plotted in Figure \@ref(fig:acf) with BwTNLV in black and BwTNML in red.
+First, the autocorrelation functions (ACF) from each posterior sample corresponding to a model probability for a visible vector outcome $\mathbf{v}=(v_1,v_2,v_3,v_4)\in\{\pm 1\}^4$ (i.e., computed from $\boldsymbol \theta$ under \@ref(eq:pmf) are determined and plotted in Figure \@ref(fig:acf) with BwTNLV in black and BwTNML in red.
 As expected, ACF corresponding to the method (BwTNML) that marginalizes out the hidden variables from the likelihood decreases to zero at a much faster rate, indicating better mixing for the chain.
@@
 The priors in the BwTNML method constrain the RBM model fit to avoid replication of the empirical distribution and better estimate the underlying true data generating probabilities.
-However, this method requires a marginalization step to obtain the probability function of visible observations alone, which is infeasible for a model with $\nh$ of any real size.
+However, this method requires a marginalization step to obtain the probability function of visible observations alone, which is unfeasible for a model with $\nh$ of any real size.
 \par
@@
 
+An alternative possibility to using MCMC to approximate the posterior for Bayesian inference in a RBM model is to use variational methods for approximation of the posterior [@salimans2015markov].
+The advantage to using variational methods would be computational scalability, but the disadvantages are potentially numerous.
+Namely, the lack of theoretical guarantees of convergence to a global optimum [@blei2017variational] and the difficulty of assessing how closely approximated the posterior is [@pmlr-v80-yao18a] are of concern.
+These two issues show potential difficulties in using variational Bayes methods for interval estimation at present.
+
 # Discussion

### 2018-09-01 writing/paper-rbm/docs/paper.tex
trying out some things

@@
-\documentclass[12pt]{article}
-\usepackage{amsmath,amssymb}
-\usepackage{graphicx,psfrag,epsf}
-\usepackage{enumerate}
-\usepackage{natbib}
-\usepackage{url} % not crucial - just used below for the URL
-
-\usepackage{longtable,booktabs}
-\usepackage[unicode=true]{hyperref}
+\documentclass[]{article}
+%\usepackage{nips15submit_e,times}
+\usepackage{url}
+
 \providecommand{\tightlist}{%
   \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
+
+\usepackage{lmodern}
+\usepackage{amssymb,amsmath}
+\usepackage{ifxetex,ifluatex}
+\usepackage{fixltx2e} % provides \textsubscript
+\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
+  \usepackage[T1]{fontenc}
+  \usepackage[utf8]{inputenc}
+\else % if luatex or xelatex
+  \ifxetex
+    \usepackage{mathspec}
+    \usepackage{xltxtra,xunicode}
+  \else
+    \usepackage{fontspec}
+  \fi
+  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
+  \newcommand{\euro}{€}
+\fi
+% use upquote if available, for straight quotes in verbatim environments
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
+% use microtype if available
+\IfFileExists{microtype.sty}{%
+\usepackage{microtype}
+\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
+}{}
+\usepackage{longtable,booktabs}
+\usepackage{graphicx}
+\makeatletter
+\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
+\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
+\makeatother
+% Scale images if necessary, so that they will not overflow the page
+% margins by default, and it is still possible to overwrite the defaults
+% using explicit options in \includegraphics[width, height, ...]{}
+\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
+\ifxetex
+  \usepackage[setpagesize=false, % page size defined by xetex
+              unicode=false, % unicode breaks when used with xetex
+              xetex]{hyperref}
+\else
+  \usepackage[unicode=true]{hyperref}
+\fi
+\hypersetup{breaklinks=true,
+            bookmarks=true,
+            pdfauthor={},
+            pdftitle={Properties and Bayesian fitting of restricted Boltzmann machines},
+            colorlinks=true,
+            citecolor=blue,
+            urlcolor=blue,
+            linkcolor=magenta,
+            pdfborder={0 0 0}}
+\urlstyle{same}  % don't use monospace font for urls
+\setlength{\parindent}{0pt}
+\setlength{\parskip}{6pt plus 2pt minus 1pt}
+\setlength{\emergencystretch}{3em}  % prevent overfull lines
+\setcounter{secnumdepth}{5}
+
+% Tech stuff starts here
+\usepackage{psfrag,epsf}
+%\pdfminorversion=4
 
@@
 rich structures in data, making them attractive for supervised learning.
-However, the generative behavior of RBMs is largely unexplored. In this
-paper, we discuss the relationship between RBM parameter specification
-in the binary case and model properties such as degeneracy, instability
-and uninterpretability. We also describe the difficulties that arise in
-likelihood-based and Bayes fitting of such (highly flexible) models,
-especially as Gibbs sampling (quasi-Bayes) methods are often advocated
-for the RBM model structure.
+However, the generative behavior of RBMs is largely unexplored and
+typical fitting methodology does not easily allow for uncertainty
+quantification in addition to point estimates. In this paper, we discuss
+the relationship between RBM parameter specification in the binary case
+and model properties such as degeneracy, instability and
+uninterpretability. We also describe the difficulties that arise in
+likelihood-based and discuss potential Bayes fitting of such (highly
+flexible) models, especially as Gibbs sampling (quasi-Bayes) methods are
+often advocated for the RBM model structure.
 \end{abstract}
@@
 great interest in deep learning, specifically in stacked restricted
-Boltzmann machines (RBMs) (see R. Salakhutdinov and Hinton
+Boltzmann machines (RBMs) (see Salakhutdinov and Hinton
 \protect\hyperlink{ref-salakhutdinov2009deep}{2009},
@@
 rough approximation (Hinton, Osindero, and Teh
-\protect\hyperlink{ref-hinton2006fast}{2006}). In this paper, we provide
-steps toward a fuller understanding of the model class and its
-properties from the perspective of statistical model theory, and we then
-explore the possibility of a rigorous fitting methodology. We find the
-RBM model class to be concerning in two fundamental ways.
+\protect\hyperlink{ref-hinton2006fast}{2006}). Additionally, the current
+fitting methodology does not easily allow for interval estimation of
+parameters in the model, and instead only provides for point estimation.
+In this paper, we provide steps toward a fuller understanding of the
+model class and its properties from the perspective of statistical model
+theory, and we then explore the possibility of a rigorous fitting
+methodology with natural uncertainty quantification. We find the RBM
+model class to be concerning in two fundamental ways.
 
@@
 
-In addition to model properties, we also explore the quality of
-estimation in fitting these models. The fitting can be problematic for
-two reasons, the first being computational and the second concerning
+Separate of the exploration of model properties, we also investigate the
+quality of estimation in fitting these models and propose an estimation
+procedure that will allow for uncertainty quantification rather than
+point estimates alone. The fitting of RBMs can be problematic for two
+reasons, the first being computational and the second concerning
 flexibility. As the size of these models grows, both maximum likelihood
@@
 estimates (Hinton \protect\hyperlink{ref-hinton2010practical}{2010};
-Hinton, Osindero, and Teh \protect\hyperlink{ref-hinton2006fast}{2006}).
-Related to this, these MCMC algorithms require updating potentially many
+Hinton, Osindero, and Teh \protect\hyperlink{ref-hinton2006fast}{2006}),
+including the absence of interval estimates.
+
+Furthermore, these MCMC algorithms require updating potentially many
 latent variables (hiddens) which can critically influence convergence in
@@
 explorations into the model behavior and potential propriety issues with
-the RBM class. We examine three Bayesian fitting techniques intended to
-avoid model impropriety issues raised in Section \ref{model-fitting} and
-conclude with a discussion in Section \ref{discussion}. Supplementary
-online material provides proofs for results on RBM parameterizations and
-data codings described in Section \ref{data-encoding}.
+the RBM class. This follows with the call to action in Sculley et al.
+(\protect\hyperlink{ref-sculley2018winner}{2018}) to employ small toy
+examples to advance empirical understanding of deep learning models and
+strive for a ``higher level of empirical rigor in the field.'' We
+examine three Bayesian fitting techniques intended to avoid model
+impropriety issues raised in Section \ref{model-fitting} and conclude
+with a discussion in Section \ref{discussion}. Supplementary online
+material provides proofs for results on RBM parameterizations and data
+codings described in Section \ref{data-encoding}.
 
@@
 \] is the normalizing function that ensures the probabilities
-(\eqref{eq:pmf}) sum to one. For
+\eqref{eq:pmf} sum to one. For
 \(\boldsymbol x = (h_1, \dots, h_\nh, v_1, \dots, v_\nh) \in \mathcal{C}^{\nv + \nh}\)
@@
 be the set of possible values for the vector of variables needed to
-compute probabilities (\eqref{eq:pmf}) in the model, and write
+compute probabilities \eqref{eq:pmf} in the model, and write
 \(Q_{\boldsymbol \theta}(\boldsymbol x) = \sum\limits_{i = 1}^\nh\sum\limits_{j=1}^\nv \theta_{ij} h_i v_j + \sum\limits_{i = 1}^\nh\theta_{hi} h_i + \sum\limits_{j = 1}^\nv\theta_{vj} v_j\)
@@
 weight the values of the visible \(v_i\) and hidden \(h_j\) nodes in
-probabilities (\eqref{eq:pmf}) and the interaction effect parameters
+probabilities \eqref{eq:pmf} and the interaction effect parameters
 (\(\theta_{ij}\)) weight the values of the connections \(v_i h_j\), or
@@
 deep learning (see Hinton, Osindero, and Teh
-\protect\hyperlink{ref-hinton2006fast}{2006}; R. Salakhutdinov and
-Hinton \protect\hyperlink{ref-salakhutdinov2012efficient}{2012};
-Srivastava, Salakhutdinov, and Hinton
+\protect\hyperlink{ref-hinton2006fast}{2006}; Salakhutdinov and Hinton
+\protect\hyperlink{ref-salakhutdinov2012efficient}{2012}; Srivastava,
+Salakhutdinov, and Hinton
 \protect\hyperlink{ref-srivastava2013modeling}{2013} for examples). By
@@
 be a promising way of solving object and speech recognition problems''
-(R. Salakhutdinov and Hinton
+(Salakhutdinov and Hinton
 \protect\hyperlink{ref-salakhutdinov2009deep}{2009}, 450). The stacking
@@
 (\protect\hyperlink{ref-handcock2003assessing}{2003}) classifies an
-exponentially parametrized random graph model at \(\boldsymbol \theta\)
+exponentially parameterized random graph model at \(\boldsymbol \theta\)
 as \emph{near-degenerate} if the mean value of the vector of sufficient
@@
 \(\boldsymbol t(\boldsymbol x) = (v_1, \dots, v_\nv, h_1, \dots, h_\nh, v_1 h_1, \dots, v_V h_\nh ) \in \mathcal{T} \equiv \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{C}^{\nh + \nv} \}\)
-from (\eqref{eq:t}), appearing in the neg-potential function
+from \eqref{eq:t}, appearing in the neg-potential function
 \(Q_{\boldsymbol \theta}(\cdot)\), has a mean vector
@@
 (i.e., as \(\nv \rightarrow \infty\)), we will say that the ratio
-(\eqref{eq:elpr}) needs to stay bounded for a sequence of RBM models to be
+\eqref{eq:elpr} needs to stay bounded for a sequence of RBM models to be
 stable. That is, we make the following convention.
@@
 \protect\hypertarget{prp:instab1}{}{\label{prp:instab1} }Let \(c > 0\) and
-let \(\text{LREP}(\boldsymbol \theta_\nv)\) be as in (\eqref{eq:elpr}) for
+let \(\text{LREP}(\boldsymbol \theta_\nv)\) be as in \eqref{eq:elpr} for
 an integer \(\nv \ge 1\). If
@@
 
-In other words, if the probability ratio (\eqref{eq:elpr}) is too large,
+In other words, if the probability ratio \eqref{eq:elpr} is too large,
 then a RBM model sequence will exhibit large probability shifts for very
@@
 parameters, \(\theta_{ij} = 0, i = 1, \dots, \nv, j = 1, \dots, \nh\).
-That is, with no interaction parameters, we have from (\eqref{eq:pmf})
-that \[
+That is, with no interaction parameters, we have from \eqref{eq:pmf} that
+\[
 P_\theta(V_i=1) \propto e^{\theta_{v_i}} \quad \text{ and } \quad P_\theta(H_j=1) \propto e^{\theta_{h_j}}, \quad  i=1,\ldots,\nv,j=1,\ldots,\nh
@@
 Secondly, we can look at \(\text{LREP}(\boldsymbol \theta)\) from
-(\eqref{eq:elpr}) for this tiny model in order to consider model
-instability as a function of RBM parameters. Recall that large values of
+\eqref{eq:elpr} for this tiny model in order to consider model instability
+as a function of RBM parameters. Recall that large values of
 \(\text{LREP}(\boldsymbol \theta)\) are associated with an extreme
@@
 and \(\{-1,1\}^\nv\). This will result in variables
-\(\boldsymbol t(\boldsymbol X)\) from (\eqref{eq:t}) satisfying
+\(\boldsymbol t(\boldsymbol X)\) from \eqref{eq:t} satisfying
 \(\boldsymbol t(\boldsymbol x) \in \{0,1\}^{\nh + \nv + \nh*\nv}\) or
@@
 Typically, fitting a RBM via maximum likelihood (ML) methods will be
-infeasible due mainly to the intractability of the normalizing term
-\(\gamma(\boldsymbol \theta)\) in a model (\eqref{eq:pmf}) of any
-realistic size. Ad hoc methods are often recommended instead, which aim
-to avoid this problem by using stochastic ML approximations that employ
-a small number of MCMC draws (i.e., contrastive divergence, (Hinton
+unfeasible due mainly to the intractability of the normalizing term
+\(\gamma(\boldsymbol \theta)\) in a model \eqref{eq:pmf} of any realistic
+size. Ad hoc methods are often recommended instead, which aim to avoid
+this problem by using stochastic ML approximations that employ a small
+number of MCMC draws (i.e., contrastive divergence, (Hinton
 \protect\hyperlink{ref-hinton2002training}{2002})).
@@
 may be chosen to match any given cell probabilities with arbitrary
-closeness. In fact, when the cell probabilities (\eqref{eq:0}) are all
+closeness. In fact, when the cell probabilities \eqref{eq:0} are all
 strictly positive, parameters in the RBM model can be specified to
 reflect these probabilities exactly. And, when one or more of the cell
-probabilities (\eqref{eq:0}) are zero, the corresponding RBM probabilities
+probabilities \eqref{eq:0} are zero, the corresponding RBM probabilities
 \(P(V_1=v_1,V_2=_v2| \theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21})\)
@@
 sample are the highest likelihood choices for the cell probabilities
-\(p_{(-1,-1)},p_{(1,-1)}, p_{(-1,1)},p_{(1,1)}\) in (\eqref{eq:0}) before
+\(p_{(-1,-1)},p_{(1,-1)}, p_{(-1,1)},p_{(1,1)}\) in \eqref{eq:0} before
 any parametric model is introduced for improved analysis, and the
@@
 case that already stretches the limits of what is computable - in
-particular we consider Bayes methods.
+particular we consider Bayes methods. We employ a Bayesian framework
+because of the ability to obtain uncertainty quantification with little
+to no extra effort via posterior credible intervals.
 
@@
   \emph{A ``trick'' prior.} Here we cancel out normalizing term in the
-  likelihood (from \(\gamma(\boldsymbol \theta)\) in (\eqref{eq:pmf})) so
+  likelihood (from \(\gamma(\boldsymbol \theta)\) in \eqref{eq:pmf}) so
   that resulting full conditionals of \(\boldsymbol \theta\) are
@@
   likelihood normalizing constant. (This computation is barely feasible
-  for a problem of this size and would be infeasible for larger
+  for a problem of this size and would be unfeasible for larger
   problems.) Here the hidden variables \(h_j\) are again carried along
@@
 \(\mathbf{v}=(v_1,v_2,v_3,v_4)\in\{\pm 1\}^4\) (i.e., computed from
-\(\boldsymbol \theta\) under (\eqref{eq:pmf})) are determined and plotted
-in Figure \ref{fig:acf} with BwTNLV in black and BwTNML in red. As
+\(\boldsymbol \theta\) under \eqref{eq:pmf} are determined and plotted in
+Figure \ref{fig:acf} with BwTNLV in black and BwTNML in red. As
 expected, ACF corresponding to the method (BwTNML) that marginalizes out
@@
 marginalization step to obtain the probability function of visible
-observations alone, which is infeasible for a model with \(\nh\) of any
+observations alone, which is unfeasible for a model with \(\nh\) of any
 real size.
@@
 \end{figure}
+An alternative possibility to using MCMC to approximate the posterior
+for Bayesian inference in a RBM model is to use variational methods for
+approximation of the posterior (Salimans, Kingma, and Welling
+\protect\hyperlink{ref-salimans2015markov}{2015}). The advantage to
+using variational methods would be computational scalability, but the
+disadvantages are potentially numerous. Namely, the lack of theoretical
+guarantees of convergence to a global optimum (Blei, Kucukelbir, and
+McAuliffe \protect\hyperlink{ref-blei2017variational}{2017}) and the
+difficulty of assessing how closely approximated the posterior is (Yao
+et al. \protect\hyperlink{ref-pmlr-v80-yao18a}{2018}) are of concern.
+These two issues show potential difficulties in using variational Bayes
+methods for interval estimation at present.
+
 \hypertarget{discussion}{%
@@
 \hypertarget{refs}{}
+\leavevmode\hypertarget{ref-blei2017variational}{}%
+Blei, David M, Alp Kucukelbir, and Jon D McAuliffe. 2017. ``Variational
+Inference: A Review for Statisticians.'' \emph{Journal of the American
+Statistical Association} 112 (518). Taylor \& Francis: 859--77.
+
 \leavevmode\hypertarget{ref-carreira2005contrastive}{}%
@@
 London A: Mathematical, Physical and Engineering Sciences} 222
-(594-604). The Royal Society:309--68.
-
-\leavevmode\hypertarget{ref-box1967discrimination}{}%
-G. E. P. Box, W. J. Hill. 1967. ``Discrimination Among Mechanistic
-Models.'' \emph{Technometrics} 9 (1). {[}Taylor \& Francis, Ltd.,
-American Statistical Association, American Society for
-Quality{]}:57--71.
+(594-604). The Royal Society: 309--68.
 
@@
 
+\leavevmode\hypertarget{ref-box1967discrimination}{}%
+G. E. P. Box, W. J. Hill. 1967. ``Discrimination Among Mechanistic
+Models.'' \emph{Technometrics} 9 (1). {[}Taylor \& Francis, Ltd.,
+American Statistical Association, American Society for Quality{]}:
+57--71.
+
 \leavevmode\hypertarget{ref-handcock2003assessing}{}%
@@
 Hinton, Geoffrey. 2010. ``A Practical Guide to Training Restricted
-Boltzmann Machines.'' \emph{Momentum} 9 (1):926.
+Boltzmann Machines.'' \emph{Momentum} 9 (1): 926.
 
@@
 Hinton, Geoffrey E. 2002. ``Training Products of Experts by Minimizing
-Contrastive Divergence.'' \emph{Neural Computation} 14 (8). MIT
-Press:1771--1800.
+Contrastive Divergence.'' \emph{Neural Computation} 14 (8). MIT Press:
+1771--1800.
 
@@
 Learning Algorithm for Deep Belief Nets.'' \emph{Neural Computation} 18
-(7). MIT Press:1527--54.
+(7). MIT Press: 1527--54.
 
@@
 
+\leavevmode\hypertarget{ref-lehmann1990model}{}%
+Lehmann, E. L. 1990. ``Model Specification: The Views of Fisher and
+Neyman, and Later Developments.'' \emph{Statistical Science} 5 (2).
+Institute of Mathematical Statistics: 160--68.
+
 \leavevmode\hypertarget{ref-le2008representational}{}%
@@
 Restricted Boltzmann Machines and Deep Belief Networks.'' \emph{Neural
-Computation} 20 (6). MIT Press:1631--49.
-
-\leavevmode\hypertarget{ref-lehmann1990model}{}%
-Lehmann, E. L. 1990. ``Model Specification: The Views of Fisher and
-Neyman, and Later Developments.'' \emph{Statistical Science} 5 (2).
-Institute of Mathematical Statistics:160--68.
+Computation} 20 (6). MIT Press: 1631--49.
 
@@
 Approximation Results for Deep Belief Networks and Restricted Boltzmann
-Machines.'' \emph{Neural Computation} 23 (5). MIT Press:1306--19.
+Machines.'' \emph{Neural Computation} 23 (5). MIT Press: 1306--19.
 
@@
 Learning Procedure for Deep Boltzmann Machines.'' \emph{Neural
-Computation} 24 (8). MIT Press:1967--2006.
+Computation} 24 (8). MIT Press: 1967--2006.
 
@@
 
+\leavevmode\hypertarget{ref-salimans2015markov}{}%
+Salimans, Tim, Diederik Kingma, and Max Welling. 2015. ``Markov Chain
+Monte Carlo and Variational Inference: Bridging the Gap.'' In
+\emph{International Conference on Machine Learning}, 1218--26.
+
 \leavevmode\hypertarget{ref-schweinberger2011instability}{}%
@@
 of Discrete Exponential Families.'' \emph{Journal of the American
-Statistical Association} 106 (496). Taylor \& Francis:1361--70.
+Statistical Association} 106 (496). Taylor \& Francis: 1361--70.
+
+\leavevmode\hypertarget{ref-sculley2018winner}{}%
+Sculley, D, Jasper Snoek, Alex Wiltschko, and Ali Rahimi. 2018.
+``Winner's Curse? On Pace, Progress, and Empirical Rigor.''
 
@@
 
+\leavevmode\hypertarget{ref-pmlr-v80-yao18a}{}%
+Yao, Yuling, Aki Vehtari, Daniel Simpson, and Andrew Gelman. 2018.
+``Yes, but Did It Work?: Evaluating Variational Inference.'' In
+\emph{Proceedings of the 35th International Conference on Machine
+Learning}, edited by Jennifer Dy and Andreas Krause, 80:5581--90.
+Proceedings of Machine Learning Research. Stockholmsmässan, Stockholm
+Sweden: PMLR. \url{http://proceedings.mlr.press/v80/yao18a.html}.
+
 \leavevmode\hypertarget{ref-zhou2014some}{}%

### 2018-09-01 writing/paper-rbm/index.Rmd
fix blinded
trying out some things

@@
 RBMs are thought to thereby have the ability to encode very complex and rich structures in data, making them attractive for supervised learning.
-However, the generative behavior of RBMs is largely unexplored.
+However, the generative behavior of RBMs is largely unexplored and typical fitting methodology does not easily allow for uncertainty quantification in addition to point estimates.
 In this paper, we discuss the relationship between RBM parameter specification in the binary case and model properties such as degeneracy, instability and uninterpretability.
-We also describe the difficulties that arise in likelihood-based and Bayes fitting of such (highly flexible) models, especially as Gibbs sampling (quasi-Bayes) methods are often advocated for the RBM model structure.
+We also describe the difficulties that arise in likelihood-based and discuss potential Bayes fitting of such (highly flexible) models, especially as Gibbs sampling (quasi-Bayes) methods are often advocated for the RBM model structure.
 keywords: Degeneracy, Instability, Classification, Deep Learning, Graphical Models

### 2018-09-02 writing/paper-rbm/chapters/rbm.Rmd
intro wording tweaks

@@
 Such behavior relates to model degeneracy (or near degeneracy), which is a statistical concern in that most data processes being modeled are realistically not degenerate in their spectrum of potential outcomes.
-For example, when sampling data from a nearly degenerate RBM, only a small number of output possibilities have probability greater than zero, and thus a sample of images will all be copies of the same one image (or small number of images).
+For example, when sampling data from a nearly degenerate RBM, only a small number of output possibilities receives nearly all probability, and thus a sample of images will all be copies of the same one image (or small number of images).
 An example of ten 4-pixel images simulated from a nearly degenerate RBM model is compared to ten 4-pixel images simulated from a non-degenerate RBM model in Figure \@ref(fig:sample-models).
@@
 
-Separate of the exploration of model properties, we also investigate the quality of estimation in fitting these models and propose an estimation procedure that will allow for uncertainty quantification rather than point estimates alone.
+Separate of the exploration of model properties, we also investigate the quality of estimation in fitting these models and examine an estimation procedure that allows for uncertainty quantification rather than point estimates alone.
 The fitting of RBMs can be problematic for two reasons, the first being computational and the second concerning flexibility.
@@
 
-While the computational concerns are inconvenient in fitting, issues due to model flexibility are potentially more concerning.
+As a compounding factor to the computational challenges, issues in model fitting due to model flexibility are potentially more concerning.
 For a RBM model with enough hidden variables, it has been shown that any distribution for the visibles can be approximated arbitrarily well [@le2008representational; @montufar2011refinements; and @montufar2011expressive].
@@
 Both $L_1$ and $L_2$ type penalties are mentioned to achieve weight-decay, though the benefits of these particular forms are unknown in different situations.
-However, the degree to which regularization and penalization are used by practitioners is not clear because these concepts are not (by default) a part of the accepted fitting methodology [@hinton2002training; @carreira2005contrastive].
+However, the degree to which regularization and penalization are used by practitioners is not clear because these concepts are not (by default) a part of the overall model-specification-and-fitting methodology [@hinton2002training; @carreira2005contrastive].
 In this paper, we attempt to address the concerns with overfitting and poor mixing in an alternative (and perhaps more transparent) manner, via specification of a Bayesian prior in Section \@ref(bayesian-model-fitting).
@@
 Section \@ref(explorations-of-model-properties-through-simulation) details our explorations into the model behavior and potential propriety issues with the RBM class.
-This follows with the call to action in @sculley2018winner to employ small toy examples to advance empirical understanding of deep learning models and strive for a "higher level of empirical rigor in the field." We examine three Bayesian fitting techniques intended to avoid model impropriety issues raised in Section \@ref(model-fitting) and conclude with a discussion in Section \@ref(discussion).
+This aligns with the call to action in @sculley2018winner to employ small toy examples to advance empirical understanding of deep learning models and strive for a "higher level of empirical rigor in the field." We examine three Bayesian fitting techniques intended to avoid model impropriety issues raised in Section \@ref(model-fitting) and conclude with a discussion in Section \@ref(discussion).
 Supplementary online material provides proofs for results on RBM parameterizations and data codings described in Section \@ref(data-encoding).

### 2018-09-02 writing/paper-rbm/docs/paper.tex
intro wording tweaks

@@
 and model properties such as degeneracy, instability and
-uninterpretability. We also describe the difficulties that arise in
-likelihood-based and discuss potential Bayes fitting of such (highly
-flexible) models, especially as Gibbs sampling (quasi-Bayes) methods are
-often advocated for the RBM model structure.
+uninterpretability. We also describe the associated difficulties that
+can arise with likelihood-based inference and further discuss the
+potential Bayes fitting of such (highly flexible) models, especially as
+Gibbs sampling (quasi-Bayes) methods are often advocated for the RBM
+model structure.
 \end{abstract}
@@
 from a nearly degenerate RBM, only a small number of output
-possibilities have probability greater than zero, and thus a sample of
+possibilities receives nearly all probability, and thus a sample of
 images will all be copies of the same one image (or small number of
@@
 Separate of the exploration of model properties, we also investigate the
-quality of estimation in fitting these models and propose an estimation
-procedure that will allow for uncertainty quantification rather than
-point estimates alone. The fitting of RBMs can be problematic for two
-reasons, the first being computational and the second concerning
-flexibility. As the size of these models grows, both maximum likelihood
-and Bayesian methods of fitting quickly become intractable. The
-literature often suggests Markov chain Monte Carlo (MCMC) tools for
-approximate maximization of likelihoods to fit these models (e.g., Gibbs
-sampling to exploit conditional structure in hidden and visible
-variables), but little is said about the attributes of realized
-estimates (Hinton \protect\hyperlink{ref-hinton2010practical}{2010};
-Hinton, Osindero, and Teh \protect\hyperlink{ref-hinton2006fast}{2006}),
-including the absence of interval estimates.
+quality of estimation in fitting these models and examine an estimation
+procedure that allows for uncertainty quantification rather than point
+estimates alone. The fitting of RBMs can be problematic for two reasons,
+the first being computational and the second concerning flexibility. As
+the size of these models grows, both maximum likelihood and Bayesian
+methods of fitting quickly become intractable. The literature often
+suggests Markov chain Monte Carlo (MCMC) tools for approximate
+maximization of likelihoods to fit these models (e.g., Gibbs sampling to
+exploit conditional structure in hidden and visible variables), but
+little is said about the attributes of realized estimates (Hinton
+\protect\hyperlink{ref-hinton2010practical}{2010}; Hinton, Osindero, and
+Teh \protect\hyperlink{ref-hinton2006fast}{2006}), including the absence
+of interval estimates.
 
@@
 
-While the computational concerns are inconvenient in fitting, issues due
-to model flexibility are potentially more concerning. For a RBM model
-with enough hidden variables, it has been shown that any distribution
-for the visibles can be approximated arbitrarily well (Le Roux and
-Bengio \protect\hyperlink{ref-le2008representational}{2008}; Montufar
-and Ay \protect\hyperlink{ref-montufar2011refinements}{2011}; and
-Montúfar, Rauh, and Ay
+As a compounding factor to the computational challenges, issues in model
+fitting due to model flexibility are potentially more concerning. For a
+RBM model with enough hidden variables, it has been shown that any
+distribution for the visibles can be approximated arbitrarily well (Le
+Roux and Bengio \protect\hyperlink{ref-le2008representational}{2008};
+Montufar and Ay \protect\hyperlink{ref-montufar2011refinements}{2011};
+and Montúfar, Rauh, and Ay
 \protect\hyperlink{ref-montufar2011expressive}{2011}). However, for
@@
 practitioners is not clear because these concepts are not (by default) a
-part of the accepted fitting methodology (Hinton
+part of the overall model-specification-and-fitting methodology (Hinton
 \protect\hyperlink{ref-hinton2002training}{2002}; Carreira-Perpinan and
@@
 explorations into the model behavior and potential propriety issues with
-the RBM class. This follows with the call to action in Sculley et al.
+the RBM class. This aligns with the call to action in Sculley et al.
 (\protect\hyperlink{ref-sculley2018winner}{2018}) to employ small toy

### 2018-09-02 writing/paper-rbm/index.Rmd
intro wording tweaks

@@
 In this paper, we discuss the relationship between RBM parameter specification in the binary case and model properties such as degeneracy, instability and uninterpretability.
-We also describe the difficulties that arise in likelihood-based and discuss potential Bayes fitting of such (highly flexible) models, especially as Gibbs sampling (quasi-Bayes) methods are often advocated for the RBM model structure.
+We also describe the associated difficulties that can arise with likelihood-based inference and further discuss the potential Bayes fitting of such (highly flexible) models, especially as Gibbs sampling (quasi-Bayes) methods are often advocated for the RBM model structure.
 keywords: Degeneracy, Instability, Classification, Deep Learning, Graphical Models

### 2018-09-04 writing/paper-rbm/appendices/appendix.Rmd
more wording changes

--- 2017-07-08 writing/paper-rbm/appendices/appendix-center.Rmd
@@
-\clearpage
-
-# (APPENDIX) Appendix {-}
+---
+title: Supplement to "Properties and Bayesian fitting of restricted Boltzmann machines"
+output: 
+  bookdown::pdf_book:
+    includes:
+      in_header: ../../resources/latex/header.tex
+    toc: no
+---
 
@@
 
-# References {-}
+# Small example illustrating approximation of an arbitrary binary distribution
+
+For simplicity, consider a model with two visible variables $(V_1,V_2)$ and one hidden $H_1$ so that there are four possible data realizations for $(V_1,V_2)$  given by $(\pm 1, \pm 1)$ and we may express the model probabilities
+as
+$$
+P(V_1=v_1,V_2=v_2| \theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21}) \propto  \exp\left( v_1  \theta_{v_1} + v_2 \theta_{v_2}\right) \sum_{h \in \{\pm 1\}}\exp\left( h[ \theta_{h_1}  + \theta_{11} v_1 + \theta_{21} v_2] \right),
+$$
+for $(v_1,v_2)\in \{-1, 1\}^2$, in terms of real-valued parameters $\theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21}$. Given some specified cell probabilities, say
+\begin{equation}
+(\#eq:0)
+0\leq p_{(-1,-1)},\;p_{(1,-1)}, \;p_{(-1,1)},\;p_{(1,1)},
+\end{equation} 
+for the outcomes $(\pm 1, \pm 1)$, the online appendix provides technical details to show parameter values $(\theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21})$ may be chosen to match any given cell probabilities with arbitrary closeness.
+In fact, when the cell probabilities \@ref(eq:0) are all strictly positive, parameters in the RBM model can be specified to reflect these probabilities exactly.
+And, when one or more of the cell probabilities \@ref(eq:0) are zero, the corresponding RBM probabilities $P(V_1=v_1,V_2=_v2| \theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21})$ may never be identically zero (due to exponential terms in the model) but parameters can be still selected to make the appropriate RBM cell probabilities arbitrarily small.
+
+For illustration, considering the simple example from above with $n$ iid observations, each consisting of two realized visibles $(V_1,V_2)$ the empirical cell frequencies from the sample are the highest likelihood choices for the cell probabilities $p_{(-1,-1)},p_{(1,-1)}, p_{(-1,1)},p_{(1,1)}$ in \@ref(eq:0) before any parametric model is introduced for improved analysis, and the discussion above indicates that RBM model parameters can be chosen to re-create this empirical distribution to an arbitrarily close degree.

### 2018-09-04 writing/paper-rbm/chapters/rbm.Rmd
more wording changes

--- 2018-09-01 writing/paper-rbm/chapters/rbm.Rmd
@@
 Neyman and Box further state that a model should "provide an explanation of the mechanism underlying the observed phenomena" [@lehmann1990model; @box1967discrimination].
-At issue, simulation from RBMs can often produce data lacking realistic variability so that such models may thereby fail to satisfactorily reflect a true data generation process.
-Such behavior relates to model degeneracy (or near degeneracy), which is a statistical concern in that most data processes being modeled are realistically not degenerate in their spectrum of potential outcomes.
-For example, when sampling data from a nearly degenerate RBM, only a small number of output possibilities have probability greater than zero, and thus a sample of images will all be copies of the same one image (or small number of images).
+At issue, simulation from RBMs can often produce data lacking realistic variability so that such models may thereby fail to satisfactorily reflect an observed data generation process.
+Such behavior relates to model degeneracy (or near degeneracy), which is a statistical concern in that many data processes of interest are realistically not degenerate in their spectrum of potential outcomes.
+For example, when sampling data from a nearly degenerate RBM used to model an imaging process, only a small set of output possibilities receives nearly all probability, and thus a sample of images will all be copies of the same one image (or small number of images).
 An example of ten 4-pixel images simulated from a nearly degenerate RBM model is compared to ten 4-pixel images simulated from a non-degenerate RBM model in Figure \@ref(fig:sample-models).
@@
 
-In addition to such degeneracy, we find that RBMs can easily exhibit types of instability, related to how sensitive the model probability structure can be to small changes in data outcomes.
+In addition to such degeneracy, we find that RBMs can easily exhibit types of instability, related to how sensitive the model probabilities can be to small changes in data outcomes.
 In practice, this may be seen when a single pixel change in an image results in a wildly different classification in an image classification problem.
 Occurrences of such behavior have recently been documented in RBMs [@li2014biclustering], as well as other deep architectures [@szegedy2013intriguing; @nguyen2014deep].
-We describe potential issues of model instability, degeneracy and uninterpretability for the RBM class (which are related properties) and examine the presence of these in Section \@ref(explorations-of-model-properties-through-simulation) through simulations of small, manageable examples.
-
-Separate of the exploration of model properties, we also investigate the quality of estimation in fitting these models and propose an estimation procedure that will allow for uncertainty quantification rather than point estimates alone.
+We describe (related) potential issues of model instability, degeneracy and uninterpretability for the RBM class (which are related properties) and examine the presence of these in Section \@ref(explorations-of-model-properties-through-simulation) through simulations of small, manageable examples.
+
+Separate from the exploration of model properties, we also investigate the quality of estimation possible in fitting these models and examine an estimation procedure that allows for uncertainty quantification rather than point estimates alone.
 The fitting of RBMs can be problematic for two reasons, the first being computational and the second concerning flexibility.
@@
 
-Furthermore, these MCMC algorithms require updating potentially many latent variables (hiddens) which can critically influence convergence in MCMC-based likelihood methods.
-Applying basic statistical principles in fitting RBM models of tractable sizes, we compare three fully Bayesian techniques involving MCMC, which are computationally more accessible than direct maximum likelihood and also aim to avoid parts of a RBM parameter space that yield unattractive models.
+Furthermore, these MCMC algorithms involve updating potentially many latent variables (hiddens) which can critically influence convergence in MCMC-based likelihood methods.
+Applying basic statistical principles in fitting RBM models of tractable size, we compare three fully Bayesian techniques involving MCMC, which are computationally more accessible than direct maximum likelihood and also aim to avoid parts of a RBM parameter space that yield unattractive models.
 As might be expected, with greater computational complexity comes an increase in fitting accuracy, but at the cost of practical feasibility.
 
-While the computational concerns are inconvenient in fitting, issues due to model flexibility are potentially more concerning.
-For a RBM model with enough hidden variables, it has been shown that any distribution for the visibles can be approximated arbitrarily well [@le2008representational; @montufar2011refinements; and @montufar2011expressive].
+As a factor compounding the computational challenges, issues in model fitting traceable to model flexibility are potentially more concerning.
+For a RBM model with enough hidden variables, any distribution for the visibles can be approximated arbitrarily well [@le2008representational; @montufar2011refinements; and @montufar2011expressive].
 However, for binary data, the empirical distribution of an observed training set of visible variables provides a highest likelihood benchmark, before any parametric model class is even introduced and applied to obtain a refinement of model fit.
@@
 Both $L_1$ and $L_2$ type penalties are mentioned to achieve weight-decay, though the benefits of these particular forms are unknown in different situations.
-However, the degree to which regularization and penalization are used by practitioners is not clear because these concepts are not (by default) a part of the accepted fitting methodology [@hinton2002training; @carreira2005contrastive].
+The degree to which regularization and penalization are used by practitioners is not clear because these concepts are not (by default) a part of the standard overall model-specification-and-fitting methodology [@hinton2002training; @carreira2005contrastive].
 In this paper, we attempt to address the concerns with overfitting and poor mixing in an alternative (and perhaps more transparent) manner, via specification of a Bayesian prior in Section \@ref(bayesian-model-fitting).
@@
 Section \@ref(explorations-of-model-properties-through-simulation) details our explorations into the model behavior and potential propriety issues with the RBM class.
-This follows with the call to action in @sculley2018winner to employ small toy examples to advance empirical understanding of deep learning models and strive for a "higher level of empirical rigor in the field." We examine three Bayesian fitting techniques intended to avoid model impropriety issues raised in Section \@ref(model-fitting) and conclude with a discussion in Section \@ref(discussion).
+This work aligns with the call to action by Google scientists in @sculley2018winner to employ small toy examples to advance empirical understanding of deep learning models and strive for a "higher level of empirical rigor in the field." We examine three Bayesian fitting techniques intended to avoid model impropriety issues raised in Section \@ref(model-fitting) and conclude with a discussion in Section \@ref(discussion).
 Supplementary online material provides proofs for results on RBM parameterizations and data codings described in Section \@ref(data-encoding).
@@
 While applications of the RBM have been claimed to produce some successes in classification problems, it is unclear if the model class allows one to go beyond fitting to other statistical matters of importance in using the models, such as quantification of the uncertainty of estimation and the formulation of predictive distributions.
-These are closely tied to the probability properties of the RBM model for explaining data generation and the current exposition intends to contribute to a better understanding in this direction.
+These are closely tied to the probability properties of the RBM model for explaining data generation and our exposition intends to contribute to a better understanding in this direction.
 
@@
 Let $\boldsymbol x = (h_1, \dots, h_\nh, v_1,\dots,v_\nv)$ represent the states of the visible and hidden nodes in a RBM for some integers $\nv, \nh \ge 1$.
-Each single binary random variable, visible or hidden, will take its values in a common coding set $\mathcal{C}$, where we allow one of two possibilities for the coding set, $\mathcal{C}=\{0,1\}$ or $\mathcal{C}=\{-1,1\}$, with "$1$" always indicating the "high" value of the variable.
+Each single binary random variable, visible or hidden, will take its values in a common coding set $\mathcal{C}$, where we allow (one of) two possibilities for the coding set, $\mathcal{C}=\{0,1\}$ or $\mathcal{C}=\{-1,1\}$, with "$1$" always indicating the "high" value of the variable.
 While $\mathcal{C}=\{0,1\}$ may be a natural starting point, we argue in Section \@ref(explorations-of-model-properties-through-simulation) that the coding $\mathcal{C}=\{-1,1\}$ induces more interpretable model properties for the RBM.
@@
 RBMs have risen to prominence in recent years due to their connection to deep learning [see @hinton2006fast; @salakhutdinov2012efficient; @srivastava2013modeling for examples].
-By stacking multiple layers of RBMs in a deep architecture, proponents of the model are claimed to produce the ability to learn "internal representations that become increasingly complex, which is considered to be a promising way of solving object and speech recognition problems" [@salakhutdinov2009deep, pp.
+By stacking multiple layers of RBMs in a deep architecture, proponents of the methodology claim to produce the ability to learn "internal representations that become increasingly complex, which is considered to be a promising way of solving object and speech recognition problems" [@salakhutdinov2009deep, pp.
 450].
@@
 where $\boldsymbol \theta \in \Theta \subset \mathbb{R}^q$ is the model parameter, and $\boldsymbol t: \mathcal{X} \rightarrow \mathbb{R}^q$ is a vector of statistics based on the adjacency matrix of a graph.
-Here, as earlier, $\gamma(\boldsymbol \theta) = \sum_{\boldsymbol x \in \mathcal{X}} \exp\left(\boldsymbol \theta^T \boldsymbol t(\boldsymbol x)\right)$ is the normalizing function.
+Here, as earlier for RBMs, $\gamma(\boldsymbol \theta) = \sum_{\boldsymbol x \in \mathcal{X}} \exp\left(\boldsymbol \theta^T \boldsymbol t(\boldsymbol x)\right)$ is the normalizing function.
 Let $C$ denote the convex hull of the potential outcomes of sufficient statistics, $\{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{X}\}$, under the model above.
@@
 A RBM model can be thought to exhibit an analogous form of *near-degeneracy* when there is a disproportionate amount of probability placed on a small number of elements in the sample space of visibles and hiddens, $\mathcal{C}^{\nv + \nh}$.
-Using the idea of @handcock2003assessing, when the random vector $\boldsymbol t(\boldsymbol x) = (v_1, \dots, v_\nv, h_1, \dots, h_\nh, v_1 h_1, \dots, v_V h_\nh ) \in \mathcal{T} \equiv \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{C}^{\nh + \nv} \}$ from \@ref(eq:t), appearing in the neg-potential function $Q_{\boldsymbol \theta}(\cdot)$, has a mean vector $\boldsymbol \mu(\boldsymbol \theta) \in \mathbb{R}^{\nv + \nh + \nv * \nh}$ close to the boundary of the convex hull of $\mathcal{T}$, then the RBM model can be said to exhibit near-degeneracy at $\boldsymbol \theta \in \mathbb{R}^{\nv + \nh + \nh*\nv}$.
+Using the idea of @handcock2003assessing, when the random vector $\boldsymbol t(\boldsymbol x) = (v_1, \dots, v_\nv, h_1, \dots, h_\nh, v_1 h_1, \dots, v_V h_\nh ) \in \mathcal{T} \equiv \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{C}^{\nh + \nv} \}$ from \@ref(eq:t), appearing in the neg-potential function $Q_{\boldsymbol \theta}(\cdot)$, has a mean vector $\boldsymbol \mu(\boldsymbol \theta) \in \mathbb{R}^{\nv + \nh + \nv * \nh}$ close to the boundary of the convex hull of $\mathcal{T}$, and the RBM model can be said to exhibit near-degeneracy at $\boldsymbol \theta \in \mathbb{R}^{\nv + \nh + \nh*\nv}$.
 Here the mean of $\boldsymbol t(\boldsymbol x)$ is
@@
 $$
-denoting the biggest log-probability ratio for a one component change in data outcomes (visibles) at a RBM parameter $\boldsymbol \theta_\nv$. We then have the following result.
+denoting the biggest log-probability ratio for a one-component change in data outcomes (visibles) at a RBM parameter $\boldsymbol \theta_\nv$. We then have the following result.
 
@@
 For data as pixels in an image, the instability result in Proposition \@ref(prp:instab1) manifests itself as a one pixel change in an image (one component of the visible vector) resulting in a large shift in the probability, which in turn could result in a vastly different classification of the image.
-Examples of this behavior have been presented in @szegedy2013intriguing for other deep learning models, in which a one pixel change in a test image results in a wildly different classification.
+Examples of this kind of behavior have been presented in @szegedy2013intriguing for deep learning models, in which a one pixel change in a test image results in a wildly different classification.
 
@@
 
-For spatial Markov models, @kaiser2007statistical defines a measure of model impropriety he calls *uninterpretability*, which is characterized by dependence parameters in a model being so extreme that marginal mean-structures fail to hold as anticipated by a model statement.
-We adapt this notion to RBM models as follows.
+For spatial Markov models, @kaiser2007statistical defines a measure of model impropriety he calls *uninterpretability*, which is characterized by dependence parameters in a model being so extreme that marginal mean-structures fail to hold as anticipated by consideration of a model statement.
+We adapt this notion to RBM models.
 Note that in a RBM, the parameters $\theta_{v_1}, \dots, \theta_{v_\nv}$ and $\theta_{h_1}, \dots, \theta_{h_\nh}$ are naturally associated with main effects of visible and hidden variables and can be interpreted as (logit functions of) means for variables $V_1, \dots, V_\nv, H_1, \dots, H_\nh$ in a model with no interaction parameters, $\theta_{ij} = 0, i = 1, \dots, \nv, j = 1, \dots, \nh$.
@@
 Hence, these main effect parameters have a clear interpretation under an independence model (one with $\theta_{ij} = 0$) but this interpretation can break down as interaction parameters increase in magnitude relative to the size of the main effects.
-In such cases, the main effect parameters $\theta_{v_1}$ and $\theta_{h_j}$ are no longer interpretable in the models (statements of marginal means) and the dependence parameters are so large as to dominate the entire model probability structure (also destroying the model interpretation of dependence as local conditional modifications of an overall marginal mean structure).
+In such cases, the main effect parameters $\theta_{v_1}$ and $\theta_{h_j}$ are no longer interpretable in the models (statements of marginal means) and the dependence parameters are so large as to dominate the entire model probability structure (also destroying simple interpretation of dependence parameters $\theta_{ij}$, $j=1,\ldots,\nh$ as local conditional modifications of an overall marginal mean structure $\theta_{v_i}$, as appearing for example in the (logit) conditional probability $\mathrm{logit P_\theta(V_i=1|H_1,\ldots,H_{\nh})}= \theta_{v_i} + \sum_{j=1}^{\nh} \theta_{ij} H_j$).
 Whether or not parameter interpretation is itself a goal in a given application of RBM models, this concept of interpretation can provide an additional device for examining other aspects of model propriety related to instability and degeneracy.
@@
 To assess which parameter values $\boldsymbol \theta$ may cause difficulties in interpretation, we use the difference $\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] - \text{E}\left[\boldsymbol X | \boldsymbol \theta^* \right]$ between two model expectations: E$\left[\boldsymbol X | \boldsymbol \theta\right]$ at $\boldsymbol \theta$ and expectations E$\left[\boldsymbol X | \boldsymbol \theta^* \right]$ where $\boldsymbol \theta^*$ matches $\boldsymbol \theta$ for all main effects but otherwise has $\theta_{ij} = 0$ for $i = 1, \dots, \nv, j = 1, \dots, \nh$.
-Hence, $\boldsymbol \theta^*$ and $\boldsymbol \theta$ have the same main effects but $\boldsymbol \theta^*$ has $0$ dependence parameters.
-Uninterpretability is then avoided at a parametric specification $\boldsymbol \theta$ if the model expected value at $\boldsymbol \theta$ is not very different from the corresponding model expectation under independence.
+(Hence, $\boldsymbol \theta^*$ and $\boldsymbol \theta$ have the same main effects but $\boldsymbol \theta^*$ has $0$ dependence parameters.) Uninterpretability is then avoided at a parametric specification $\boldsymbol \theta$ if the model expected value at $\boldsymbol \theta$ is not very different from the corresponding model expectation under independence.
 Using this, it is possible to investigate what parametric conditions lead to uninterpretability in a model versus those that guarantee interpretable models.
@@
 (Contour plots for three coordinate functions are shown in columns for various values of $\theta_{11}$, which can be interpreted here as an absolute log-odds ratio as the visible changes between $0$ and $1$.) In examining these, we see that as coordinates of $\boldsymbol \theta$ grow larger in magnitude, at least one mean function for the entries of $\boldsymbol t(\boldsymbol x)$ approaches a value 0 or 1, forcing $\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta} \boldsymbol t(\boldsymbol x)$ to be near to the boundary of the convex hull of $\mathcal{T}$, as a sign of model near-degeneracy.
-Thus, for a very small example we can see the relationship between values of $\boldsymbol \theta$ and model degeneracy.
+Thus, for a very small example we can see the relationship between values of $\boldsymbol \theta$ moving (sometimes only slightly) from zero and model degeneracy.
 \par
@@
 ```{r instab, fig.cap="$\\text{LREP}(\\boldsymbol \\theta)/\\nv$ for various values of $\\boldsymbol \\theta$ for the tiny example model.
-Recall here $\\nv$ is the number of visible nodes and here is $1$.
-This quantity is large for large magnitudes of $\\boldsymbol \\theta$.", fig.height = 4, fig.align='center'}
+(Recall here $\\nv$ is the number of visible nodes and here is $1$.) This quantity is large for large magnitudes of $\\boldsymbol \\theta$.", fig.height = 4, fig.align='center'}
 expand.grid(theta_h1 = seq(-5, 5, length.out = 21),
@@
 Figure \@ref(fig:uninterp) shows that the absolute difference between coordinates of the vector of model expectations, E$\left[\boldsymbol X | \boldsymbol \theta\right]$ and corresponding expectations E$\left[\boldsymbol X | \boldsymbol \theta^*\right ]$ given independence grow for the toy RBM as the values of $\boldsymbol \theta$ are farther from zero, especially for large magnitudes of the dependence parameter $\theta_{11}$.
-This is a third indication that parameter vectors of large magnitude lead to model impropriety in a RBM.
+This is a third indication that parameter vectors of large magnitude can readily lead to model impropriety in a RBM.
 
@@
 In our numerical experiment, we split $\boldsymbol \theta = (\boldsymbol \theta_{main}, \boldsymbol \theta_{interaction})$ into $\boldsymbol \theta_{main}$ and $\boldsymbol \theta_{interaction}$, in reference to which variables in the probability function the parameters correspond (whether they multiply a $v_i$ or a $h_j$ or they multiply a $v_i h_j$), and allow the two types of terms to have varying average magnitudes, $||\boldsymbol \theta_{main} || /(\nh+\nv)$ and $||\boldsymbol \theta_{interaction} || /(\nh*\nv)$.
-These average magnitudes vary on a grid between `r min(plot_dat_grid$r1/(plot_dat_grid$H + plot_dat_grid$V))` and `r max(plot_dat_grid$r1/(plot_dat_grid$H + plot_dat_grid$V))` with `r length(unique(plot_dat_grid$r1/(plot_dat_grid$H + plot_dat_grid$V)))` breaks, yielding `r length(unique(plot_dat_grid$r1/(plot_dat_grid$H + plot_dat_grid$V)))^2` grid points.
-(By looking at the average magnitudes, we are able to later consider the potential benefit of shrinking each parameter value $\theta_i$ towards zero in a Bayesian fitting technique.) At each point in the grid, `r dim(res$samp[[1]])[1]` vectors ($\boldsymbol \theta_{main}$) are sampled uniformly on a sphere with radius corresponding to the first coordinate in the grid and `r dim(res$samp[[1]])[1]` vectors ($\boldsymbol \theta_{interction}$) are sampled uniformly on a sphere with radius corresponding to the second coordinate in the grid via sums of squared and scaled iid Normal$(0, 1)$ variables.
-These vectors are then paired to create `r dim(res$samp[[1]])[1]` values of $\boldsymbol \theta$ with magnitudes at each point in the grid.
+These average magnitudes vary on a $1$-dimensional grid between `r min(plot_dat_grid$r1/(plot_dat_grid$H + plot_dat_grid$V))` and `r max(plot_dat_grid$r1/(plot_dat_grid$H + plot_dat_grid$V))` with `r length(unique(plot_dat_grid$r1/(plot_dat_grid$H + plot_dat_grid$V)))` breaks, yielding `r length(unique(plot_dat_grid$r1/(plot_dat_grid$H + plot_dat_grid$V)))^2` $2$-dimensional grid points.
+(By looking at the average magnitudes, we are able to later consider the potential benefit of shrinking each parameter value $\theta_i$ towards zero in a Bayesian fitting technique.) At each point in the grid, `r dim(res$samp[[1]])[1]` vectors ($\boldsymbol \theta_{main}$) are sampled uniformly on a sphere with radius corresponding to the first coordinate of the point and `r dim(res$samp[[1]])[1]` vectors ($\boldsymbol \theta_{interction}$) are sampled uniformly on a sphere with radius corresponding to the second coordinate of the point via sums of squared and scaled iid Normal$(0, 1)$ variables.
+These vectors are then paired to create `r dim(res$samp[[1]])[1]` values of $\boldsymbol \theta$ at each point in the grid.
 
@@
 
-```{r degen-plots, fig.cap="Results from the numerical experiment, here looking at the fraction of models that were near-degenerate for each combination of magnitude of $\\boldsymbol \\theta$ and model size, where $\\boldsymbol \\theta = (\\boldsymbol \\theta_{main}, \\theta_{interaction})$ is split into main and interaction parameters.
+```{r degen-plots, fig.cap="Results from the numerical experiment, here looking at the fraction of models that were near-degenerate for each combination of magnitudes of the points $\\boldsymbol \\theta$ and model size, where $\\boldsymbol \\theta = (\\boldsymbol \\theta_{main}, \\theta_{interaction})$ is split into main and interaction parameters.
 Black lines show the contour levels for fraction of near-degeneracy, while the thick black line shows the level where the fraction of near-degenerate models is .05.", fig.height=3.5, fig.align="ht!"}
@@
 
-```{r instab-plots, fig.cap="The sample mean value of $\\text{LREP}(\\boldsymbol \\theta)/V$ at each grid point for each combination of magnitude of $\\boldsymbol \\theta$ and model size.
+```{r instab-plots, fig.cap="The sample mean value of $\\text{LREP}(\\boldsymbol \\theta)/V$ at each grid point for each combination of magnitudes of the points of $\\boldsymbol \\theta$ and model size.
 As the magnitude of $\\boldsymbol \\theta$ grows, so does the value of this metric, indicating typical instability in the model.", fig.height=3.5, fig.align="ht!"}
@@
 For example, @montufar2011refinements show that any distribution on $\{0, 1\}^{\nv}$ can be approximated arbitrarily well by a RBM with $2^{\nv-1} - 1$ hidden units.
-We provide a small example that illustrates that in fact there can be many such approximations.
-For simplicity, consider a model with two visible variables $(V_1,V_2)$ and one hidden $H_1$ so that there are four possible data realizations for $(V_1,V_2)$  given by $(\pm 1, \pm 1)$ and we may express the model probabilities
-as
-$$
-P(V_1=v_1,V_2=v_2| \theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21}) \propto  \exp\left( v_1  \theta_{v_1} + v_2 \theta_{v_2}\right) \sum_{h \in \{\pm 1\}}\exp\left( h[ \theta_{h_1}  + \theta_{11} v_1 + \theta_{21} v_2] \right),
-$$
-for $(v_1,v_2)\in \{-1, 1\}^2$, in terms of real-valued parameters $\theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21}$. Given some specified cell probabilities, say
-\begin{equation}
-(\#eq:0)
-0\leq p_{(-1,-1)},\;p_{(1,-1)}, \;p_{(-1,1)},\;p_{(1,1)},
-\end{equation} 
-for the outcomes $(\pm 1, \pm 1)$, the online appendix provides technical details to show parameter values $(\theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21})$ may be chosen to match any given cell probabilities with arbitrary closeness.
-In fact, when the cell probabilities \@ref(eq:0) are all strictly positive, parameters in the RBM model can be specified to reflect these probabilities exactly.
-And, when one or more of the cell probabilities \@ref(eq:0) are zero, the corresponding RBM probabilities $P(V_1=v_1,V_2=_v2| \theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21})$ may never be identically zero (due to exponential terms in the model) but parameters can be still selected to make the appropriate RBM cell probabilities arbitrarily small.
+We provide a small example in the on-line supplementary materials that illustrates such approximations.
+
 Furthermore, as development in the online appendix shows, not only is it possible to approximate any distribution on the visibles arbitrarily well [cf.
 @montufar2011refinements], but dramatically different parameter settings can induce the same RBM model (beyond mere symmetries in parameterization).
-A further, and somewhat odd consequence of the RBM parameterization is then that, when fitting the RBM model by likelihood-based methods, we already know the nature of the answer before we begin: namely, such  fitting will simply reproduce the empirical distribution of the training data if sufficiently many hiddens are in the model and there can be no model refinements or smoothing from the RBM.
+A further (and somewhat odd) consequence of the RBM parameterization is then that when fitting the RBM model by likelihood-based methods, we may already know the nature of the answer before we begin: namely, such  fitting will simply reproduce the empirical distribution of the training data if sufficiently many hiddens are in the model.
+There then can be no model refinements or smoothing from the RBM.
 That is, based on a random sample of vectors of visible variables, the model for the cell probabilities with the highest likelihood over  *all possible model classes* (i.e., RBM-based or not) is the empirical distribution, and the over-parameterization of the RBM model itself ensures that this empirical distribution can be arbitrarily well-approximated.
-For illustration, considering the simple example from above with $n$ iid observations, each consisting of two realized visibles $(V_1,V_2)$ the empirical cell frequencies from the sample are the highest likelihood choices for the cell probabilities $p_{(-1,-1)},p_{(1,-1)}, p_{(-1,1)},p_{(1,1)}$ in \@ref(eq:0) before any parametric model is introduced for improved analysis, and the discussion above indicates that RBM model parameters can be chosen to re-create this empirical distribution to an arbitrarily close degree.
 Not only does RBM model fitting based on ML seek to reproduce the empirical distribution, whenever this empirical distribution contains empty cells, fitting steps for the RBM model will further aim to choose parameters that necessarily diverge to infinity in magnitude in order to zero-out the corresponding RBM cell probabilities.
@@
 
-Here we consider what might be done in a principled manner to prevent both overfitting and model impropriety, testing on a $\nv = \nh = 4$ case that already stretches the limits of what is computable - in particular we consider Bayes methods.
-We employ a Bayesian framework because of the ability to obtain uncertainty quantification with little to no extra effort via posterior credible intervals.
+Here we consider what might be done in a principled manner to prevent both overfitting and model impropriety, testing on a $\nv = \nh = 4$ case that already stretches the limits of what is computable --- in particular we consider Bayes methods.
+We employ a Bayesian framework because of the ability to obtain uncertainty quantification with little to no extra effort, via posterior credible intervals.
 
@@
 We simulated $n = 5,000$ realizations of visibles as a training set and fit the RBM using three Bayes methodologies.
-These involved the following set-ups with choice of prior distribution $\pi(\boldsymbol \theta)$ for parameters $\boldsymbol \theta$.
+These involved the following set-ups for choice of prior distribution $\pi(\boldsymbol \theta)$ for parameters $\boldsymbol \theta$.
 
@@
 
+\par
 It should be noted that, due to the common prior distributions for $\boldsymbol \theta$, both BwTNLV (method 2 above) and BwTNML (method 3) are drawing from the same stationary posterior distribution for vectors of visibles.
 A fundamental difference between these two methods lies in how well these two chains mix and how quickly they arrive at the target posterior distribution.
-After a burn-in period of 50 iterations selected by inspecting the trace plots, we assess the issue of mixing in two ways.
+After a burn-in of 50 iterations selected by inspecting the trace plots, we assess the issue of mixing in two ways.
 First, the autocorrelation functions (ACF) from each posterior sample corresponding to a model probability for a visible vector outcome $\mathbf{v}=(v_1,v_2,v_3,v_4)\in\{\pm 1\}^4$ (i.e., computed from $\boldsymbol \theta$ under \@ref(eq:pmf) are determined and plotted in Figure \@ref(fig:acf) with BwTNLV in black and BwTNML in red.
@@
 RBM models constitute an interesting class of undirected graphical models that are thought to be useful for supervised learning tasks.
-However, when viewed as generative statistical models, RBMs are prone to forms of model impropriety such as near-degeneracy, S-instability, and uninterpretability.
-Additionally, these models are difficult to fit using a rigorous methodology, due to the dimension of the parameter space coupled with the size of the latent variable space.
-
-In this paper, we have presented three fully Bayes-principled MCMC-based methods for fitting RBMs.
+However, when viewed as generative statistical models, RBMs cn be susceptible to forms of model impropriety such as near-degeneracy, S-instability, and uninterpretability.
+Additionally, these models are difficult to fit using a rigorous methodology tht cn incorporate estimtion uncertainty, due to the dimension of the parameter space coupled with the size of the latent variable space.
+
+In this paper, we have presented three fully Bayes-principled MCMC-based methods for fitting RBMs, which provide posterior distributions for quantifying uncertainty in estimating model parameters and probabilities (e.g., Figure \@ref(fig:fitting-plot)).
 Common practice is to use a kind of MCMC to overcome fitting complexities.
@@
 Marginalization over the latent variables in the model is shown to improve mixing, but is numerically intractable due to the necessity of repeated calculation of the normalizing constant.
-Due to the extreme flexibility in this model class, rigorous likelihood-based fitting for a RBM will typically aim to merely merely return the (discrete) empirical distribution for visibles, meaning any practitioner should be aware of this and employ some form of regularization or penalization in the model.
-
-Ultimately, it is not clear that RBM models are useful as generative models.
-Furthermore, without the appropriate generative behavior in the RBM model, quantifying uncertainty in the estimated model parameters becomes impossible and the model loses useful application to prediction problems (e.g., realistic predictive distributions).
+Due to the extreme flexibility in this model class, rigorous likelihood-based fitting for a RBM can typically seek to merely merely return the (discrete) empirical distribution for visibles, and practitioners should be aware of this and employ some form of regularization or penalization in the fitting.
+
+Ultimately, it is not unquestionably clear that RBM models are useful as generative models.
+As a concern, without appropriate generative behavior in a RBM model, the uncertainty in estimated model parameters becomes impossible to realistically assess and the model can further lose useful application to prediction problems (e.g., realistic predictive distributions).
 In the case of classification, predictive distributions may not be the ultimate goal, but when S-instability is present, small (imperceptible) differences in the data may lead to greatly different probabilities and thus greatly different classifications.
-For these reasons, we are skeptical about RBMs as data generative tools.
+For these reasons, we are skeptical about RBMs as probabilistic conceptualizations for data generation.
 

### 2018-09-04 writing/paper-rbm/docs/paper.tex
more wording changes

@@
 variability so that such models may thereby fail to satisfactorily
-reflect a true data generation process. Such behavior relates to model
-degeneracy (or near degeneracy), which is a statistical concern in that
-most data processes being modeled are realistically not degenerate in
+reflect an observed data generation process. Such behavior relates to
+model degeneracy (or near degeneracy), which is a statistical concern in
+that many data processes of interest are realistically not degenerate in
 their spectrum of potential outcomes. For example, when sampling data
-from a nearly degenerate RBM, only a small number of output
-possibilities receives nearly all probability, and thus a sample of
-images will all be copies of the same one image (or small number of
-images). An example of ten 4-pixel images simulated from a nearly
-degenerate RBM model is compared to ten 4-pixel images simulated from a
-non-degenerate RBM model in Figure \ref{fig:sample-models}. The
-degenerate model places almost all probability on one outcome, causing
-the image to be generated repeatedly, whereas the non-degenerate model
-shows more potentially realistic variation.
+from a nearly degenerate RBM used to model an imaging process, only a
+small set of output possibilities receives nearly all probability, and
+thus a sample of images will all be copies of the same one image (or
+small number of images). An example of ten 4-pixel images simulated from
+a nearly degenerate RBM model is compared to ten 4-pixel images
+simulated from a non-degenerate RBM model in Figure
+\ref{fig:sample-models}. The degenerate model places almost all
+probability on one outcome, causing the image to be generated
+repeatedly, whereas the non-degenerate model shows more potentially
+realistic variation.
 
@@
 In addition to such degeneracy, we find that RBMs can easily exhibit
-types of instability, related to how sensitive the model probability
-structure can be to small changes in data outcomes. In practice, this
-may be seen when a single pixel change in an image results in a wildly
-different classification in an image classification problem. Occurrences
-of such behavior have recently been documented in RBMs (Li
+types of instability, related to how sensitive the model probabilities
+can be to small changes in data outcomes. In practice, this may be seen
+when a single pixel change in an image results in a wildly different
+classification in an image classification problem. Occurrences of such
+behavior have recently been documented in RBMs (Li
 \protect\hyperlink{ref-li2014biclustering}{2014}), as well as other deep
@@
 and Clune \protect\hyperlink{ref-nguyen2014deep}{2014}). We describe
-potential issues of model instability, degeneracy and uninterpretability
-for the RBM class (which are related properties) and examine the
-presence of these in Section
+(related) potential issues of model instability, degeneracy and
+uninterpretability for the RBM class (which are related properties) and
+examine the presence of these in Section
 \ref{explorations-of-model-properties-through-simulation} through
@@
 
-Separate of the exploration of model properties, we also investigate the
-quality of estimation in fitting these models and examine an estimation
-procedure that allows for uncertainty quantification rather than point
-estimates alone. The fitting of RBMs can be problematic for two reasons,
-the first being computational and the second concerning flexibility. As
-the size of these models grows, both maximum likelihood and Bayesian
-methods of fitting quickly become intractable. The literature often
-suggests Markov chain Monte Carlo (MCMC) tools for approximate
-maximization of likelihoods to fit these models (e.g., Gibbs sampling to
-exploit conditional structure in hidden and visible variables), but
-little is said about the attributes of realized estimates (Hinton
+Separate from the exploration of model properties, we also investigate
+the quality of estimation possible in fitting these models and examine
+an estimation procedure that allows for uncertainty quantification
+rather than point estimates alone. The fitting of RBMs can be
+problematic for two reasons, the first being computational and the
+second concerning flexibility. As the size of these models grows, both
+maximum likelihood and Bayesian methods of fitting quickly become
+intractable. The literature often suggests Markov chain Monte Carlo
+(MCMC) tools for approximate maximization of likelihoods to fit these
+models (e.g., Gibbs sampling to exploit conditional structure in hidden
+and visible variables), but little is said about the attributes of
+realized estimates (Hinton
 \protect\hyperlink{ref-hinton2010practical}{2010}; Hinton, Osindero, and
@@
 
-Furthermore, these MCMC algorithms require updating potentially many
+Furthermore, these MCMC algorithms involve updating potentially many
 latent variables (hiddens) which can critically influence convergence in
 MCMC-based likelihood methods. Applying basic statistical principles in
-fitting RBM models of tractable sizes, we compare three fully Bayesian
+fitting RBM models of tractable size, we compare three fully Bayesian
 techniques involving MCMC, which are computationally more accessible
@@
 
-As a compounding factor to the computational challenges, issues in model
-fitting due to model flexibility are potentially more concerning. For a
-RBM model with enough hidden variables, it has been shown that any
-distribution for the visibles can be approximated arbitrarily well (Le
-Roux and Bengio \protect\hyperlink{ref-le2008representational}{2008};
-Montufar and Ay \protect\hyperlink{ref-montufar2011refinements}{2011};
-and Montúfar, Rauh, and Ay
-\protect\hyperlink{ref-montufar2011expressive}{2011}). However, for
-binary data, the empirical distribution of an observed training set of
-visible variables provides a highest likelihood benchmark, before any
-parametric model class is even introduced and applied to obtain a
-refinement of model fit. As a consequence, we find that any fully
-principled fitting method based on the likelihood for a RBM with enough
-hidden variables will seek to reproduce the (discrete) empirical
-distribution of a training set. This aspect can be undesirable, and
-perhaps even unexpected compared to most modeling scenarios, in that no
-``smoothed distribution'' may result when fitting a RBM model of
-sufficient size with a rigorous likelihood-based method. We are
-therefore led to be skeptical that models that involve these structures
-(like deep Boltzmann machines) can achieve useful prediction or
-inference in a principled way without intentionally limiting the
+As a factor compounding the computational challenges, issues in model
+fitting traceable to model flexibility are potentially more concerning.
+For a RBM model with enough hidden variables, any distribution for the
+visibles can be approximated arbitrarily well (Le Roux and Bengio
+\protect\hyperlink{ref-le2008representational}{2008}; Montufar and Ay
+\protect\hyperlink{ref-montufar2011refinements}{2011}; and Montúfar,
+Rauh, and Ay \protect\hyperlink{ref-montufar2011expressive}{2011}).
+However, for binary data, the empirical distribution of an observed
+training set of visible variables provides a highest likelihood
+benchmark, before any parametric model class is even introduced and
+applied to obtain a refinement of model fit. As a consequence, we find
+that any fully principled fitting method based on the likelihood for a
+RBM with enough hidden variables will seek to reproduce the (discrete)
+empirical distribution of a training set. This aspect can be
+undesirable, and perhaps even unexpected compared to most modeling
+scenarios, in that no ``smoothed distribution'' may result when fitting
+a RBM model of sufficient size with a rigorous likelihood-based method.
+We are therefore led to be skeptical that models that involve these
+structures (like deep Boltzmann machines) can achieve useful prediction
+or inference in a principled way without intentionally limiting the
 flexibility of the fitted model.
@@
 benefits of these particular forms are unknown in different situations.
-However, the degree to which regularization and penalization are used by
+The degree to which regularization and penalization are used by
 practitioners is not clear because these concepts are not (by default) a
-part of the overall model-specification-and-fitting methodology (Hinton
-\protect\hyperlink{ref-hinton2002training}{2002}; Carreira-Perpinan and
-Hinton \protect\hyperlink{ref-carreira2005contrastive}{2005}). In this
-paper, we attempt to address the concerns with overfitting and poor
-mixing in an alternative (and perhaps more transparent) manner, via
-specification of a Bayesian prior in Section
-\ref{bayesian-model-fitting}.
+part of the standard overall model-specification-and-fitting methodology
+(Hinton \protect\hyperlink{ref-hinton2002training}{2002};
+Carreira-Perpinan and Hinton
+\protect\hyperlink{ref-carreira2005contrastive}{2005}). In this paper,
+we attempt to address the concerns with overfitting and poor mixing in
+an alternative (and perhaps more transparent) manner, via specification
+of a Bayesian prior in Section \ref{bayesian-model-fitting}.
 
@@
 explorations into the model behavior and potential propriety issues with
-the RBM class. This aligns with the call to action in Sculley et al.
+the RBM class. This work aligns with the call to action by Google
+scientists in Sculley et al.
 (\protect\hyperlink{ref-sculley2018winner}{2018}) to employ small toy
@@
 distributions. These are closely tied to the probability properties of
-the RBM model for explaining data generation and the current exposition
-intends to contribute to a better understanding in this direction.
+the RBM model for explaining data generation and our exposition intends
+to contribute to a better understanding in this direction.
 
@@
 hidden, will take its values in a common coding set \(\mathcal{C}\),
-where we allow one of two possibilities for the coding set,
+where we allow (one of) two possibilities for the coding set,
 \(\mathcal{C}=\{0,1\}\) or \(\mathcal{C}=\{-1,1\}\), with ``\(1\)''
@@
 stacking multiple layers of RBMs in a deep architecture, proponents of
-the model are claimed to produce the ability to learn ``internal
+the methodology claim to produce the ability to learn ``internal
 representations that become increasingly complex, which is considered to
@@
 is a vector of statistics based on the adjacency matrix of a graph.
-Here, as earlier,
+Here, as earlier for RBMs,
 \(\gamma(\boldsymbol \theta) = \sum_{\boldsymbol x \in \mathcal{X}} \exp\left(\boldsymbol \theta^T \boldsymbol t(\boldsymbol x)\right)\)
@@
 \(\boldsymbol \mu(\boldsymbol \theta) \in \mathbb{R}^{\nv + \nh + \nv * \nh}\)
-close to the boundary of the convex hull of \(\mathcal{T}\), then the
-RBM model can be said to exhibit near-degeneracy at
+close to the boundary of the convex hull of \(\mathcal{T}\), and the RBM
+model can be said to exhibit near-degeneracy at
 \(\boldsymbol \theta \in \mathbb{R}^{\nv + \nh + \nh*\nv}\). Here the
@@
 \Delta(\boldsymbol \theta_\nv) \equiv \max \left\{\log \frac{P_{\boldsymbol \theta_\nv}(\boldsymbol v)}{P_{\boldsymbol \theta_\nv}(\boldsymbol v^*)} : \boldsymbol v \text{ }\& \text{ } \boldsymbol v^* \in \mathcal{C}^\nv \text{ differ in exactly one component}\right\},
-\] denoting the biggest log-probability ratio for a one component change
+\] denoting the biggest log-probability ratio for a one-component change
 in data outcomes (visibles) at a RBM parameter
@@
 shift in the probability, which in turn could result in a vastly
-different classification of the image. Examples of this behavior have
-been presented in Szegedy et al.
-(\protect\hyperlink{ref-szegedy2013intriguing}{2013}) for other deep
-learning models, in which a one pixel change in a test image results in
-a wildly different classification.
+different classification of the image. Examples of this kind of behavior
+have been presented in Szegedy et al.
+(\protect\hyperlink{ref-szegedy2013intriguing}{2013}) for deep learning
+models, in which a one pixel change in a test image results in a wildly
+different classification.
 
@@
 characterized by dependence parameters in a model being so extreme that
-marginal mean-structures fail to hold as anticipated by a model
-statement. We adapt this notion to RBM models as follows. Note that in a
+marginal mean-structures fail to hold as anticipated by consideration of
+a model statement. We adapt this notion to RBM models. Note that in a
 RBM, the parameters \(\theta_{v_1}, \dots, \theta_{v_\nv}\) and
@@
 large as to dominate the entire model probability structure (also
-destroying the model interpretation of dependence as local conditional
-modifications of an overall marginal mean structure). Whether or not
-parameter interpretation is itself a goal in a given application of RBM
-models, this concept of interpretation can provide an additional device
-for examining other aspects of model propriety related to instability
-and degeneracy. As explained by Kaiser
+destroying simple interpretation of dependence parameters
+\(\theta_{ij}\), \(j=1,\ldots,\nh\) as local conditional modifications
+of an overall marginal mean structure \(\theta_{v_i}\), as appearing for
+example in the (logit) conditional probability
+\(\mathrm{logit P_\theta(V_i=1|H_1,\ldots,H_{\nh})}= \theta_{v_i} + \sum_{j=1}^{\nh} \theta_{ij} H_j\)).
+Whether or not parameter interpretation is itself a goal in a given
+application of RBM models, this concept of interpretation can provide an
+additional device for examining other aspects of model propriety related
+to instability and degeneracy. As explained by Kaiser
 (\protect\hyperlink{ref-kaiser2007statistical}{2007}), models with
@@
 effects but otherwise has \(\theta_{ij} = 0\) for
-\(i = 1, \dots, \nv, j = 1, \dots, \nh\). Hence,
+\(i = 1, \dots, \nv, j = 1, \dots, \nh\). (Hence,
 \(\boldsymbol \theta^*\) and \(\boldsymbol \theta\) have the same main
-effects but \(\boldsymbol \theta^*\) has \(0\) dependence parameters.
+effects but \(\boldsymbol \theta^*\) has \(0\) dependence parameters.)
 Uninterpretability is then avoided at a parametric specification
@@
 sign of model near-degeneracy. Thus, for a very small example we can see
-the relationship between values of \(\boldsymbol \theta\) and model
-degeneracy.
+the relationship between values of \(\boldsymbol \theta\) moving
+(sometimes only slightly) from zero and model degeneracy.
 
@@
 \caption{$\text{LREP}(\boldsymbol \theta)/\nv$ for various values of $\boldsymbol \theta$ for the tiny example model.
-Recall here $\nv$ is the number of visible nodes and here is $1$.
-This quantity is large for large magnitudes of $\boldsymbol \theta$.}\label{fig:instab}
+(Recall here $\nv$ is the number of visible nodes and here is $1$.) This quantity is large for large magnitudes of $\boldsymbol \theta$.}\label{fig:instab}
 \end{figure}
@@
 magnitudes of the dependence parameter \(\theta_{11}\). This is a third
-indication that parameter vectors of large magnitude lead to model
-impropriety in a RBM.
+indication that parameter vectors of large magnitude can readily lead to
+model impropriety in a RBM.
 
@@
 \(||\boldsymbol \theta_{interaction} || /(\nh*\nv)\). These average
-magnitudes vary on a grid between 0.001 and 3 with 24 breaks, yielding
-576 grid points. (By looking at the average magnitudes, we are able to
-later consider the potential benefit of shrinking each parameter value
-\(\theta_i\) towards zero in a Bayesian fitting technique.) At each
-point in the grid, 100 vectors (\(\boldsymbol \theta_{main}\)) are
-sampled uniformly on a sphere with radius corresponding to the first
-coordinate in the grid and 100 vectors
-(\(\boldsymbol \theta_{interction}\)) are sampled uniformly on a sphere
-with radius corresponding to the second coordinate in the grid via sums
-of squared and scaled iid Normal\((0, 1)\) variables. These vectors are
-then paired to create 100 values of \(\boldsymbol \theta\) with
-magnitudes at each point in the grid.
+magnitudes vary on a \(1\)-dimensional grid between 0.001 and 3 with 24
+breaks, yielding 576 \(2\)-dimensional grid points. (By looking at the
+average magnitudes, we are able to later consider the potential benefit
+of shrinking each parameter value \(\theta_i\) towards zero in a
+Bayesian fitting technique.) At each point in the grid, 100 vectors
+(\(\boldsymbol \theta_{main}\)) are sampled uniformly on a sphere with
+radius corresponding to the first coordinate of the point and 100
+vectors (\(\boldsymbol \theta_{interction}\)) are sampled uniformly on a
+sphere with radius corresponding to the second coordinate of the point
+via sums of squared and scaled iid Normal\((0, 1)\) variables. These
+vectors are then paired to create 100 values of \(\boldsymbol \theta\)
+at each point in the grid.
 
@@
 \begin{figure}
-\includegraphics{paper_files/figure-latex/degen-plots-1} \caption{Results from the numerical experiment, here looking at the fraction of models that were near-degenerate for each combination of magnitude of $\boldsymbol \theta$ and model size, where $\boldsymbol \theta = (\boldsymbol \theta_{main}, \theta_{interaction})$ is split into main and interaction parameters.
+\includegraphics{paper_files/figure-latex/degen-plots-1} \caption{Results from the numerical experiment, here looking at the fraction of models that were near-degenerate for each combination of magnitudes of the points $\boldsymbol \theta$ and model size, where $\boldsymbol \theta = (\boldsymbol \theta_{main}, \theta_{interaction})$ is split into main and interaction parameters.
 Black lines show the contour levels for fraction of near-degeneracy, while the thick black line shows the level where the fraction of near-degenerate models is .05.}\label{fig:degen-plots}
@@
 \begin{figure}
-\includegraphics{paper_files/figure-latex/instab-plots-1} \caption{The sample mean value of $\text{LREP}(\boldsymbol \theta)/V$ at each grid point for each combination of magnitude of $\boldsymbol \theta$ and model size.
+\includegraphics{paper_files/figure-latex/instab-plots-1} \caption{The sample mean value of $\text{LREP}(\boldsymbol \theta)/V$ at each grid point for each combination of magnitudes of the points of $\boldsymbol \theta$ and model size.
 As the magnitude of $\boldsymbol \theta$ grows, so does the value of this metric, indicating typical instability in the model.}\label{fig:instab-plots}
@@
 by a RBM with \(2^{\nv-1} - 1\) hidden units. We provide a small example
-that illustrates that in fact there can be many such approximations. For
-simplicity, consider a model with two visible variables \((V_1,V_2)\)
-and one hidden \(H_1\) so that there are four possible data realizations
-for \((V_1,V_2)\) given by \((\pm 1, \pm 1)\) and we may express the
-model probabilities as \[
-P(V_1=v_1,V_2=v_2| \theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21}) \propto  \exp\left( v_1  \theta_{v_1} + v_2 \theta_{v_2}\right) \sum_{h \in \{\pm 1\}}\exp\left( h[ \theta_{h_1}  + \theta_{11} v_1 + \theta_{21} v_2] \right),
-\] for \((v_1,v_2)\in \{-1, 1\}^2\), in terms of real-valued parameters
-\(\theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21}\).
-Given some specified cell probabilities, say \begin{equation}
-\label{eq:0}
-0\leq p_{(-1,-1)},\;p_{(1,-1)}, \;p_{(-1,1)},\;p_{(1,1)},
-\end{equation} for the outcomes \((\pm 1, \pm 1)\), the online appendix
-provides technical details to show parameter values
-\((\theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21})\)
-may be chosen to match any given cell probabilities with arbitrary
-closeness. In fact, when the cell probabilities \eqref{eq:0} are all
-strictly positive, parameters in the RBM model can be specified to
-reflect these probabilities exactly. And, when one or more of the cell
-probabilities \eqref{eq:0} are zero, the corresponding RBM probabilities
-\(P(V_1=v_1,V_2=_v2| \theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21})\)
-may never be identically zero (due to exponential terms in the model)
-but parameters can be still selected to make the appropriate RBM cell
-probabilities arbitrarily small. Furthermore, as development in the
-online appendix shows, not only is it possible to approximate any
-distribution on the visibles arbitrarily well (cf. Montufar and Ay
+in the on-line supplementary materials that illustrates such
+approximations.
+
+Furthermore, as development in the online appendix shows, not only is it
+possible to approximate any distribution on the visibles arbitrarily
+well (cf. Montufar and Ay
 \protect\hyperlink{ref-montufar2011refinements}{2011}), but dramatically
 different parameter settings can induce the same RBM model (beyond mere
-symmetries in parameterization). A further, and somewhat odd consequence
-of the RBM parameterization is then that, when fitting the RBM model by
-likelihood-based methods, we already know the nature of the answer
-before we begin: namely, such fitting will simply reproduce the
-empirical distribution of the training data if sufficiently many hiddens
-are in the model and there can be no model refinements or smoothing from
-the RBM. That is, based on a random sample of vectors of visible
-variables, the model for the cell probabilities with the highest
+symmetries in parameterization). A further (and somewhat odd)
+consequence of the RBM parameterization is then that when fitting the
+RBM model by likelihood-based methods, we may already know the nature of
+the answer before we begin: namely, such fitting will simply reproduce
+the empirical distribution of the training data if sufficiently many
+hiddens are in the model. There then can be no model refinements or
+smoothing from the RBM. That is, based on a random sample of vectors of
+visible variables, the model for the cell probabilities with the highest
 likelihood over \emph{all possible model classes} (i.e., RBM-based or
@@
 RBM model itself ensures that this empirical distribution can be
-arbitrarily well-approximated. For illustration, considering the simple
-example from above with \(n\) iid observations, each consisting of two
-realized visibles \((V_1,V_2)\) the empirical cell frequencies from the
-sample are the highest likelihood choices for the cell probabilities
-\(p_{(-1,-1)},p_{(1,-1)}, p_{(-1,1)},p_{(1,1)}\) in \eqref{eq:0} before
-any parametric model is introduced for improved analysis, and the
-discussion above indicates that RBM model parameters can be chosen to
-re-create this empirical distribution to an arbitrarily close degree.
-Not only does RBM model fitting based on ML seek to reproduce the
-empirical distribution, whenever this empirical distribution contains
-empty cells, fitting steps for the RBM model will further aim to choose
-parameters that necessarily diverge to infinity in magnitude in order to
-zero-out the corresponding RBM cell probabilities. In data applications
-with a large sample space, it is unlikely that the training set will
-include at least one of each possible vector outcome (unlike this small
-example). This implies that some RBM model parameters must diverge to
-\(+\infty\) to mimic the empirical distribution with empty cells and, as
-we have already discussed in Section
+arbitrarily well-approximated. Not only does RBM model fitting based on
+ML seek to reproduce the empirical distribution, whenever this empirical
+distribution contains empty cells, fitting steps for the RBM model will
+further aim to choose parameters that necessarily diverge to infinity in
+magnitude in order to zero-out the corresponding RBM cell probabilities.
+In data applications with a large sample space, it is unlikely that the
+training set will include at least one of each possible vector outcome
+(unlike this small example). This implies that some RBM model parameters
+must diverge to \(+\infty\) to mimic the empirical distribution with
+empty cells and, as we have already discussed in Section
 \ref{explorations-of-model-properties-through-simulation}, large
@@
 both overfitting and model impropriety, testing on a \(\nv = \nh = 4\)
-case that already stretches the limits of what is computable - in
+case that already stretches the limits of what is computable --- in
 particular we consider Bayes methods. We employ a Bayesian framework
 because of the ability to obtain uncertainty quantification with little
-to no extra effort via posterior credible intervals.
+to no extra effort, via posterior credible intervals.
 
@@
 visibles as a training set and fit the RBM using three Bayes
-methodologies. These involved the following set-ups with choice of prior
+methodologies. These involved the following set-ups for choice of prior
 distribution \(\pi(\boldsymbol \theta)\) for parameters
@@
 \end{table}
+\par
+
 It should be noted that, due to the common prior distributions for
@@
 lies in how well these two chains mix and how quickly they arrive at the
-target posterior distribution. After a burn-in period of 50 iterations
-selected by inspecting the trace plots, we assess the issue of mixing in
-two ways. First, the autocorrelation functions (ACF) from each posterior
+target posterior distribution. After a burn-in of 50 iterations selected
+by inspecting the trace plots, we assess the issue of mixing in two
+ways. First, the autocorrelation functions (ACF) from each posterior
 sample corresponding to a model probability for a visible vector outcome
@@
 models that are thought to be useful for supervised learning tasks.
-However, when viewed as generative statistical models, RBMs are prone to
-forms of model impropriety such as near-degeneracy, S-instability, and
-uninterpretability. Additionally, these models are difficult to fit
-using a rigorous methodology, due to the dimension of the parameter
-space coupled with the size of the latent variable space.
+However, when viewed as generative statistical models, RBMs cn be
+susceptible to forms of model impropriety such as near-degeneracy,
+S-instability, and uninterpretability. Additionally, these models are
+difficult to fit using a rigorous methodology tht cn incorporate
+estimtion uncertainty, due to the dimension of the parameter space
+coupled with the size of the latent variable space.
 
 In this paper, we have presented three fully Bayes-principled MCMC-based
-methods for fitting RBMs. Common practice is to use a kind of MCMC to
-overcome fitting complexities. However because of the size of the space
-to be filled with MCMC iterates, convergence and mixing of these methods
-will be slow. Marginalization over the latent variables in the model is
-shown to improve mixing, but is numerically intractable due to the
-necessity of repeated calculation of the normalizing constant. Due to
-the extreme flexibility in this model class, rigorous likelihood-based
-fitting for a RBM will typically aim to merely merely return the
-(discrete) empirical distribution for visibles, meaning any practitioner
-should be aware of this and employ some form of regularization or
-penalization in the model.
-
-Ultimately, it is not clear that RBM models are useful as generative
-models. Furthermore, without the appropriate generative behavior in the
-RBM model, quantifying uncertainty in the estimated model parameters
-becomes impossible and the model loses useful application to prediction
-problems (e.g., realistic predictive distributions). In the case of
-classification, predictive distributions may not be the ultimate goal,
-but when S-instability is present, small (imperceptible) differences in
-the data may lead to greatly different probabilities and thus greatly
-different classifications. For these reasons, we are skeptical about
-RBMs as data generative tools.
+methods for fitting RBMs, which provide posterior distributions for
+quantifying uncertainty in estimating model parameters and probabilities
+(e.g., Figure \ref{fig:fitting-plot}). Common practice is to use a kind
+of MCMC to overcome fitting complexities. However because of the size of
+the space to be filled with MCMC iterates, convergence and mixing of
+these methods will be slow. Marginalization over the latent variables in
+the model is shown to improve mixing, but is numerically intractable due
+to the necessity of repeated calculation of the normalizing constant.
+Due to the extreme flexibility in this model class, rigorous
+likelihood-based fitting for a RBM can typically seek to merely merely
+return the (discrete) empirical distribution for visibles, and
+practitioners should be aware of this and employ some form of
+regularization or penalization in the fitting.
+
+Ultimately, it is not unquestionably clear that RBM models are useful as
+generative models. As a concern, without appropriate generative behavior
+in a RBM model, the uncertainty in estimated model parameters becomes
+impossible to realistically assess and the model can further lose useful
+application to prediction problems (e.g., realistic predictive
+distributions). In the case of classification, predictive distributions
+may not be the ultimate goal, but when S-instability is present, small
+(imperceptible) differences in the data may lead to greatly different
+probabilities and thus greatly different classifications. For these
+reasons, we are skeptical about RBMs as probabilistic conceptualizations
+for data generation.
 

### 2018-09-05 writing/paper-rbm/docs/paper.tex
fix citation

@@
 \leavevmode\hypertarget{ref-sculley2018winner}{}%
-Sculley, D, Jasper Snoek, Alex Wiltschko, and Ali Rahimi. 2018.
-``Winner's Curse? On Pace, Progress, and Empirical Rigor.''
+Sculley, D., Jasper Snoek, Alex Wiltschko, and Ali Rahimi. 2018.
+``Winner's Curse? On Pace, Progress, and Empirical Rigor.'' In \emph{6th
+International Conference on Learning Representations Workshop
+Submission}. \url{https://openreview.net/forum?id=rJWF0Fywf}.
 

### 2018-09-14 writing/paper-rbm/chapters/rbm.Rmd
try to clear up the discussion

@@
 
-In addition to such degeneracy, we find that RBMs can easily exhibit types of instability, related to how sensitive the model probabilities can be to small changes in data outcomes.
+In addition to such degeneracy, we find that RBMs can easily exhibit types of instability, related to how sensitive the model probabilities can be to small difference in data outcomes.
 In practice, this may be seen when a single pixel change in an image results in a wildly different classification in an image classification problem.
@@
 RBM models constitute an interesting class of undirected graphical models that are thought to be useful for supervised learning tasks.
-However, when viewed as generative statistical models, RBMs cn be susceptible to forms of model impropriety such as near-degeneracy, S-instability, and uninterpretability.
-Additionally, these models are difficult to fit using a rigorous methodology tht cn incorporate estimtion uncertainty, due to the dimension of the parameter space coupled with the size of the latent variable space.
-
-In this paper, we have presented three fully Bayes-principled MCMC-based methods for fitting RBMs, which provide posterior distributions for quantifying uncertainty in estimating model parameters and probabilities (e.g., Figure \@ref(fig:fitting-plot)).
-Common practice is to use a kind of MCMC to overcome fitting complexities.
-However because of the size of the space to be filled with MCMC iterates, convergence and mixing of these methods will be slow.
+However, when viewed as generative statistical models, RBMs can be susceptible to forms of model impropriety such as near-degeneracy, S-instability, and uninterpretability.
+These model instability problems relate to how useful the model may be for representing realistic data generation mechanisms.
+In this paper, we have provided a theoretical framework for discussing the model properties of RBMs, as well as provided empirical evidence that these model problems can arise away from the origin of the parameter space.
+
+Additionally, these models are difficult to fit using a rigorous methodology that can incorporate estimation uncertainty, due to the dimension of the parameter space coupled with the size of the latent variable space.
+We have presented three fully Bayes-principled MCMC-based methods for fitting RBMs, which provide posterior distributions for quantifying uncertainty in estimating model parameters and probabilities (e.g., Figure \@ref(fig:fitting-plot)) while aiming to avoid areas of the parameter space that will potentially lead to model impropriety.
+We have shown small examples where this fitting methodology does work, but becomes quickly intractable in large data cases.
+
+The current common practice is to use a kind of MCMC to overcome fitting complexities, and due to the extreme flexibility in this model class, rigorous likelihood-based fitting for a RBM can typically seek to merely merely return the (discrete) empirical distribution for visibles.
+Current fitting methodology for RBMs does not provide uncertainty quantification for the fitted parameters, and it is unclear that they seeks to address the issues in model impropriety detailed in this paper.
+Practitioners should be aware of this and employ some form of regularization or penalization in the fitting.
+
+<!-- because of the size of the space to be filled with MCMC iterates, convergence and mixing of these methods will be slow.
 Marginalization over the latent variables in the model is shown to improve mixing, but is numerically intractable due to the necessity of repeated calculation of the normalizing constant.
-Due to the extreme flexibility in this model class, rigorous likelihood-based fitting for a RBM can typically seek to merely merely return the (discrete) empirical distribution for visibles, and practitioners should be aware of this and employ some form of regularization or penalization in the fitting.
+-->
 

### 2018-09-14 writing/paper-rbm/docs/paper.tex
try to clear up the discussion

@@
 types of instability, related to how sensitive the model probabilities
-can be to small changes in data outcomes. In practice, this may be seen
-when a single pixel change in an image results in a wildly different
-classification in an image classification problem. Occurrences of such
-behavior have recently been documented in RBMs (Li
+can be to small difference in data outcomes. In practice, this may be
+seen when a single pixel change in an image results in a wildly
+different classification in an image classification problem. Occurrences
+of such behavior have recently been documented in RBMs (Li
 \protect\hyperlink{ref-li2014biclustering}{2014}), as well as other deep
@@
 models that are thought to be useful for supervised learning tasks.
-However, when viewed as generative statistical models, RBMs cn be
+However, when viewed as generative statistical models, RBMs can be
 susceptible to forms of model impropriety such as near-degeneracy,
-S-instability, and uninterpretability. Additionally, these models are
-difficult to fit using a rigorous methodology tht cn incorporate
-estimtion uncertainty, due to the dimension of the parameter space
-coupled with the size of the latent variable space.
-
-In this paper, we have presented three fully Bayes-principled MCMC-based
-methods for fitting RBMs, which provide posterior distributions for
-quantifying uncertainty in estimating model parameters and probabilities
-(e.g., Figure \ref{fig:fitting-plot}). Common practice is to use a kind
-of MCMC to overcome fitting complexities. However because of the size of
-the space to be filled with MCMC iterates, convergence and mixing of
-these methods will be slow. Marginalization over the latent variables in
-the model is shown to improve mixing, but is numerically intractable due
-to the necessity of repeated calculation of the normalizing constant.
-Due to the extreme flexibility in this model class, rigorous
-likelihood-based fitting for a RBM can typically seek to merely merely
-return the (discrete) empirical distribution for visibles, and
-practitioners should be aware of this and employ some form of
+S-instability, and uninterpretability. These model instability problems
+relate to how useful the model may be for representing realistic data
+generation mechanisms. In this paper, we have provided a theoretical
+framework for discussing the model properties of RBMs, as well as
+provided empirical evidence that these model problems can arise away
+from the origin of the parameter space.
+
+Additionally, these models are difficult to fit using a rigorous
+methodology that can incorporate estimation uncertainty, due to the
+dimension of the parameter space coupled with the size of the latent
+variable space. We have presented three fully Bayes-principled
+MCMC-based methods for fitting RBMs, which provide posterior
+distributions for quantifying uncertainty in estimating model parameters
+and probabilities (e.g., Figure \ref{fig:fitting-plot}) while aiming to
+avoid areas of the parameter space that will potentially lead to model
+impropriety. We have shown small examples where this fitting methodology
+does work, but becomes quickly intractable in large data cases.
+
+The current common practice is to use a kind of MCMC to overcome fitting
+complexities, and due to the extreme flexibility in this model class,
+rigorous likelihood-based fitting for a RBM can typically seek to merely
+merely return the (discrete) empirical distribution for visibles.
+Current fitting methodology for RBMs does not provide uncertainty
+quantification for the fitted parameters, and it is unclear that they
+seeks to address the issues in model impropriety detailed in this paper.
+Practitioners should be aware of this and employ some form of
 regularization or penalization in the fitting.

### 2018-09-18 writing/paper-rbm/paper.Rmd
include new files

--- 2018-09-14 writing/paper-rbm/chapters/rbm.Rmd
@@
+---
+title: "Properties and Bayesian fitting of restricted Boltzmann machines"
+author: 
+  - name: Andee Kaplan
+    shortname: Kaplan A.,
+    numaddress: 1
+    corresponding: true
+  - name: Daniel Nordman
+    shortname: D. Nordman, and
+    corresponding: false
+    numaddress: 2
+  - name: Stephen Vardeman
+    shortname: S. Vardeman
+    numaddress: "2,3"
+    corresponding: false
+shortauthor: "KAPLAN \\textsc{et al}"
+address:
+  - numaddress: 1
+    orgdiv: Department of Statistical Science
+    orgname: Duke University
+    orgstate: North Carolina
+    orgcountry: USA
+  - numaddress: 2
+    orgdiv: Department of Statistics
+    orgname: Iowa State University
+    orgstate: Iowa
+    orgcountry: USA
+  - numaddress: 3
+    orgdiv: Department of Industrial and Manufacturing Systems Engineering
+    orgname: Iowa State University
+    orgstate: Iowa
+    orgcountry: USA
+corresponding: true
+corresponding_name: Andee Kaplan
+corresponding_address: Department of Statistical Science, Duke University, P.O. Box 90251, Durham, NC 27708-0251
+corresponding_email: andrea.kaplan@duke.edu
+output: 
+  rmarkdown::pdf_document:
+    keep_tex: true
+    template: ../resources/latex/wileynjd-ams_article.tex
+    includes:
+      in_header: ../resources/latex/header.tex
+    toc: no
+bibliography: [../resources/latex/refs_rbm.bib]
+abstract: |
+   A restricted Boltzmann machine (RBM) is an undirected graphical model constructed for discrete or continuous random variables, with two layers, one hidden and one visible, and no conditional dependency within a layer.
+In recent years, RBMs have risen to prominence due to their connection to deep learning.
+By treating a hidden layer of one RBM as the visible layer in a second RBM, a deep architecture can be created.
+RBMs are thought to thereby have the ability to encode very complex and rich structures in data, making them attractive for supervised learning.
+However, the generative behavior of RBMs is largely unexplored and typical fitting methodology does not easily allow for uncertainty quantification in addition to point estimates.
+In this paper, we discuss the relationship between RBM parameter specification in the binary case and model properties such as degeneracy, instability and uninterpretability.
+We also describe the associated difficulties that can arise with likelihood-based inference and further discuss the potential Bayes fitting of such (highly flexible) models, especially as Gibbs sampling (quasi-Bayes) methods are often advocated for the RBM model structure.
+keywords: Degeneracy, Instability, Classification, Deep Learning, Graphical Models
+---
+
+```{r setup, include=FALSE}
+# https://github.com/rstudio/bookdown/pull/374
+fix_envs <- function(x) {
+  beg_reg <- '^\\s*\\\\begin\\{.*\\}'
+  end_reg <- '^\\s*\\\\end\\{.*\\}'
+  i3 = if (length(i1 <- grep(beg_reg, x))) (i1 - 1)[grepl("^\\s*$", x[i1 - 1])]
+  i3 = c(i3, if (length(i2 <- grep(end_reg, x))) (i2 + 1)[grepl("^\\s*$", x[i2 + 1])])
+  if (length(i3)) x = x[-i3]
+  x
+}
+
+options(
+  dplyr.print_min = 6, dplyr.print_max = 6, width = 70,
+  digits = 3,
+  bookdown.post.latex = fix_envs
+)
+knitr::opts_chunk$set(
+  comment = "#>",
+  warning = FALSE,
+  message = FALSE,
+  cache = TRUE,
+  echo = FALSE
+)
+
+ggplot2::theme_set(ggplot2::theme_bw(base_family="serif"))
+
+library(knitr)
+library(ggplot2)
+library(dplyr)
+library(tidyr)
+library(scales)
+library(mvtnorm)
+library(rootSolve)
+library(agridat)
+library(xtable)
+```
+
 ```{r helpers, echo=FALSE, message=FALSE, warnings=FALSE}
@@
 For example, when sampling data from a nearly degenerate RBM used to model an imaging process, only a small set of output possibilities receives nearly all probability, and thus a sample of images will all be copies of the same one image (or small number of images).
-An example of ten 4-pixel images simulated from a nearly degenerate RBM model is compared to ten 4-pixel images simulated from a non-degenerate RBM model in Figure \@ref(fig:sample-models).
+An example of ten 4-pixel images simulated from a nearly degenerate RBM model is compared to ten 4-pixel images simulated from a non-degenerate RBM model in Figure \ref{fig:sample-models}.
 The degenerate model places almost all probability on one outcome, causing the image to be generated repeatedly, whereas the non-degenerate model shows more potentially realistic variation.
@@
 
-```{r sample-models, fig.show='hold', fig.height=2, fig.cap='Ten 4-pixel images simulated from a degenerate model (a) compared to ten 4-pixel images simulated from a non-degenerate model (b).
+```{r sample-models, fig.show='hold', fig.height=2, fig.cap='\\label{fig:sample-models}Ten 4-pixel images simulated from a degenerate model (a) compared to ten 4-pixel images simulated from a non-degenerate model (b).
 The degenerate model places almost all probability on one outcome, causing the image to be generated repeatedly, whereas the non-degenerate model shows more realistic variation.', out.width='100%'}
@@
 Occurrences of such behavior have recently been documented in RBMs [@li2014biclustering], as well as other deep architectures [@szegedy2013intriguing; @nguyen2014deep].
-We describe (related) potential issues of model instability, degeneracy and uninterpretability for the RBM class (which are related properties) and examine the presence of these in Section \@ref(explorations-of-model-properties-through-simulation) through simulations of small, manageable examples.
+We describe (related) potential issues of model instability, degeneracy and uninterpretability for the RBM class (which are related properties) and examine the presence of these in Section \ref{explorations-of-model-properties-through-simulation} through simulations of small, manageable examples.
 
@@
 The degree to which regularization and penalization are used by practitioners is not clear because these concepts are not (by default) a part of the standard overall model-specification-and-fitting methodology [@hinton2002training; @carreira2005contrastive].
-In this paper, we attempt to address the concerns with overfitting and poor mixing in an alternative (and perhaps more transparent) manner, via specification of a Bayesian prior in Section \@ref(bayesian-model-fitting).
+In this paper, we attempt to address the concerns with overfitting and poor mixing in an alternative (and perhaps more transparent) manner, via specification of a Bayesian prior in Section \ref{bayesian-model-fitting}.
 
 This paper is structured as follows.
-Section \@ref(background) formally defines the RBM including the joint distribution of hidden and visible variables and explains the model's connection to deep learning.
+Section \ref{background} formally defines the RBM including the joint distribution of hidden and visible variables and explains the model's connection to deep learning.
 Additionally, measures of model impropriety and methods of quantifying/detecting it are defined.
-Section \@ref(explorations-of-model-properties-through-simulation) details our explorations into the model behavior and potential propriety issues with the RBM class.
-This work aligns with the call to action by Google scientists in @sculley2018winner to employ small toy examples to advance empirical understanding of deep learning models and strive for a "higher level of empirical rigor in the field." We examine three Bayesian fitting techniques intended to avoid model impropriety issues raised in Section \@ref(model-fitting) and conclude with a discussion in Section \@ref(discussion).
-Supplementary online material provides proofs for results on RBM parameterizations and data codings described in Section \@ref(data-encoding).
+Section \ref{explorations-of-model-properties-through-simulation} details our explorations into the model behavior and potential propriety issues with the RBM class.
+This work aligns with the call to action by Google scientists in @sculley2018winner to employ small toy examples to advance empirical understanding of deep learning models and strive for a "higher level of empirical rigor in the field." We examine three Bayesian fitting techniques intended to avoid model impropriety issues raised in Section \ref{model-fitting} and conclude with a discussion in Section \ref{discussion}.
+Supplementary online material provides proofs for results on RBM parameterizations and data codings described in Section \ref{data-encoding}.
 
@@
 A RBM architecture has two layers, hidden ($\mathcal{H}$) and visible ($\mathcal{V}$), with no dependency connections within a layer.
-An example of this structure is in Figure \@ref(fig:rbm) with the hidden nodes indicated by gray circles and the visible nodes indicated by white circles.
+An example of this structure is in Figure \ref{fig:rbm} with the hidden nodes indicated by gray circles and the visible nodes indicated by white circles.
 \begin{figure}
@@
 Each single binary random variable, visible or hidden, will take its values in a common coding set $\mathcal{C}$, where we allow (one of) two possibilities for the coding set, $\mathcal{C}=\{0,1\}$ or $\mathcal{C}=\{-1,1\}$, with "$1$" always indicating the "high" value of the variable.
-While $\mathcal{C}=\{0,1\}$ may be a natural starting point, we argue in Section \@ref(explorations-of-model-properties-through-simulation) that the coding $\mathcal{C}=\{-1,1\}$ induces more interpretable model properties for the RBM.
+While $\mathcal{C}=\{0,1\}$ may be a natural starting point, we argue in Section \ref{explorations-of-model-properties-through-simulation} that the coding $\mathcal{C}=\{-1,1\}$ induces more interpretable model properties for the RBM.
 A standard parametric form for probabilities corresponding to a potential vector of states, $X = (H_1, \dots, H_\nh, V_1,\dots,V_\nv)$, for the nodes is
 \begin{align}
-(\#eq:pmf)
+\label{eq:pmf}
 f_{\boldsymbol \theta} (\boldsymbol x)\equiv P_{\boldsymbol \theta}(\boldsymbol X = \boldsymbol x) = \frac{\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}{\gamma(\boldsymbol \theta)}, \quad \boldsymbol x \in \mathcal{C}^{\nh + \nv}
@@
 $$ 
-is the normalizing function that ensures the probabilities \@ref(eq:pmf) sum to one. For $\boldsymbol x = (h_1, \dots, h_\nh, v_1, \dots, v_\nh) \in \mathcal{C}^{\nv + \nh}$ and 
+is the normalizing function that ensures the probabilities \eqref{eq:pmf} sum to one. For $\boldsymbol x = (h_1, \dots, h_\nh, v_1, \dots, v_\nh) \in \mathcal{C}^{\nv + \nh}$ and 
 \begin{align}
-\boldsymbol t(\boldsymbol x) &= (h_1, \dots, h_\nh, v_1, \dots, v_\nv, v_1 h_1, \dots, v_\nv h_\nh) \in \mathcal{C}^{\nh + \nv + \nh*\nv}, (\#eq:t)
+\boldsymbol t(\boldsymbol x) &= (h_1, \dots, h_\nh, v_1, \dots, v_\nv, v_1 h_1, \dots, v_\nv h_\nh) \in \mathcal{C}^{\nh + \nv + \nh*\nv}, \label{eq:t}
 \end{align}
-let $\mathcal{T} = \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{C}^{\nv + \nh}\} \subset \mathbb{R}^{\nv + \nh + \nv * \nh}$ be the set of possible values for the vector of variables needed to compute probabilities \@ref(eq:pmf) in the model, and write $Q_{\boldsymbol \theta}(\boldsymbol x) = \sum\limits_{i = 1}^\nh\sum\limits_{j=1}^\nv \theta_{ij} h_i v_j + \sum\limits_{i = 1}^\nh\theta_{hi} h_i + \sum\limits_{j = 1}^\nv\theta_{vj} v_j$ for the "neg-potential" function.
+let $\mathcal{T} = \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{C}^{\nv + \nh}\} \subset \mathbb{R}^{\nv + \nh + \nv * \nh}$ be the set of possible values for the vector of variables needed to compute probabilities \eqref{eq:pmf} in the model, and write $Q_{\boldsymbol \theta}(\boldsymbol x) = \sum\limits_{i = 1}^\nh\sum\limits_{j=1}^\nv \theta_{ij} h_i v_j + \sum\limits_{i = 1}^\nh\theta_{hi} h_i + \sum\limits_{j = 1}^\nv\theta_{vj} v_j$ for the "neg-potential" function.
 The RBM model is parameterized by $\boldsymbol \theta$ containing two types of parameters, main effects and interaction effects.
-The main effects parameters ($\{\theta_{v_i}, \theta_{h_j}\}_{\substack{i = 1, \dots, \nv,\\j = 1, \dots, \nh}}$) weight the values of the visible $v_i$ and hidden $h_j$ nodes in probabilities \@ref(eq:pmf) and the interaction effect parameters ($\theta_{ij}$) weight the values of the connections $v_i h_j$, or dependencies, between hidden and visible layers.
+The main effects parameters ($\{\theta_{v_i}, \theta_{h_j}\}_{\substack{i = 1, \dots, \nv,\\j = 1, \dots, \nh}}$) weight the values of the visible $v_i$ and hidden $h_j$ nodes in probabilities \eqref{eq:pmf} and the interaction effect parameters ($\theta_{ij}$) weight the values of the connections $v_i h_j$, or dependencies, between hidden and visible layers.
 
@@
 visibles given hiddens or vice verse).
-Specifically, the conditional independence of nodes in each layer (given those nodes in the other layer) allows for stepwise simulation of both hidden layers and model parameters (e.g., see the contrastive divergence of @hinton2002training or Bayes methods in Section \@ref(model-fitting)).
+Specifically, the conditional independence of nodes in each layer (given those nodes in the other layer) allows for stepwise simulation of both hidden layers and model parameters (e.g., see the contrastive divergence of @hinton2002training or Bayes methods in Section \ref{model-fitting}).
 
@@
 A RBM model can be thought to exhibit an analogous form of *near-degeneracy* when there is a disproportionate amount of probability placed on a small number of elements in the sample space of visibles and hiddens, $\mathcal{C}^{\nv + \nh}$.
-Using the idea of @handcock2003assessing, when the random vector $\boldsymbol t(\boldsymbol x) = (v_1, \dots, v_\nv, h_1, \dots, h_\nh, v_1 h_1, \dots, v_V h_\nh ) \in \mathcal{T} \equiv \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{C}^{\nh + \nv} \}$ from \@ref(eq:t), appearing in the neg-potential function $Q_{\boldsymbol \theta}(\cdot)$, has a mean vector $\boldsymbol \mu(\boldsymbol \theta) \in \mathbb{R}^{\nv + \nh + \nv * \nh}$ close to the boundary of the convex hull of $\mathcal{T}$, and the RBM model can be said to exhibit near-degeneracy at $\boldsymbol \theta \in \mathbb{R}^{\nv + \nh + \nh*\nv}$.
+Using the idea of @handcock2003assessing, when the random vector $\boldsymbol t(\boldsymbol x) = (v_1, \dots, v_\nv, h_1, \dots, h_\nh, v_1 h_1, \dots, v_V h_\nh ) \in \mathcal{T} \equiv \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{C}^{\nh + \nv} \}$ from \eqref{eq:t}, appearing in the neg-potential function $Q_{\boldsymbol \theta}(\cdot)$, has a mean vector $\boldsymbol \mu(\boldsymbol \theta) \in \mathbb{R}^{\nv + \nh + \nv * \nh}$ close to the boundary of the convex hull of $\mathcal{T}$, and the RBM model can be said to exhibit near-degeneracy at $\boldsymbol \theta \in \mathbb{R}^{\nv + \nh + \nh*\nv}$.
 Here the mean of $\boldsymbol t(\boldsymbol x)$ is
@@
 \begin{align}
-\frac{1}{\nv} \log \left[\frac{\max\limits_{(v_1, \dots, v_\nv) \in \mathcal{C}^\nv}P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}{\min\limits_{(v_1, \dots, v_\nv) \in \mathcal{C}^\nv}P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}\right] \equiv \frac{1}{\nv} \text{LREP}(\boldsymbol \theta_\nv) (\#eq:elpr)
+\frac{1}{\nv} \log \left[\frac{\max\limits_{(v_1, \dots, v_\nv) \in \mathcal{C}^\nv}P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}{\min\limits_{(v_1, \dots, v_\nv) \in \mathcal{C}^\nv}P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}\right] \equiv \frac{1}{\nv} \text{LREP}(\boldsymbol \theta_\nv) \label{eq:elpr}
 \end{align}
@@
 
-In formulating a RBM model for a potentially large number of visibles (i.e., as $\nv \rightarrow \infty$), we will say that the ratio \@ref(eq:elpr) needs to stay bounded for a sequence of RBM models to be stable.
+In formulating a RBM model for a potentially large number of visibles (i.e., as $\nv \rightarrow \infty$), we will say that the ratio \eqref{eq:elpr} needs to stay bounded for a sequence of RBM models to be stable.
 That is, we make the following convention.
-```{definition, name = "S-unstable RBM", echo=TRUE}
+
+\begin{definition}[S-unstable RBM]
 Let $\boldsymbol \theta_\nv \in \mathbb{R}^{\nv + \nh +  \nh*\nv}, \nv \ge 1$, be an element of a sequence of RBM parameters where the number of hiddens, $\nh \equiv \nh(\nv) \ge 1$, can be an arbitrary function of the number $\nv$ of visibles.
@@
 \end{align*}
-```
+\end{definition}
+
 In other words, a RBM model sequence is unstable if, given any $c > 0$, there exists an integer $n_c > 0$ so that $\frac{1}{\nv}\text{LREP}(\boldsymbol \theta_\nv) > c$ for all $\nv \ge n_c$.
@@
 
-```{proposition, label = "instab1", echo=TRUE}
-Let $c > 0$ and let $\text{LREP}(\boldsymbol \theta_\nv)$ be as in \@ref(eq:elpr) for an integer $\nv \ge 1$.
+\begin{proposition}
+\label{prp:instab1}
+Let $c > 0$ and let $\text{LREP}(\boldsymbol \theta_\nv)$ be as in \eqref{eq:elpr} for an integer $\nv \ge 1$.
 If $\frac{1}{\nv}\text{LREP}(\boldsymbol \theta_\nv) > c$, then $\Delta(\boldsymbol \theta_\nv) > c$.
-```
-
-In other words, if the probability ratio \@ref(eq:elpr) is too large, then a RBM model sequence will exhibit large probability shifts for very small changes in data configurations (i.e., will exhibit instability).
+\end{proposition}
+
+In other words, if the probability ratio \eqref{eq:elpr} is too large, then a RBM model sequence will exhibit large probability shifts for very small changes in data configurations (i.e., will exhibit instability).
 Such instability can be a concern in a model for reasons similar to those related to degeneracy: as outcomes vary over the sample space, the geography of probabilities is extremely rugged, with deep pits following sharp mountains.
 Recall the applied example of RBM models as a means to classify images.
-For data as pixels in an image, the instability result in Proposition \@ref(prp:instab1) manifests itself as a one pixel change in an image (one component of the visible vector) resulting in a large shift in the probability, which in turn could result in a vastly different classification of the image.
+For data as pixels in an image, the instability result in Proposition \ref{prp:instab1} manifests itself as a one pixel change in an image (one component of the visible vector) resulting in a large shift in the probability, which in turn could result in a vastly different classification of the image.
 Examples of this kind of behavior have been presented in @szegedy2013intriguing for deep learning models, in which a one pixel change in a test image results in a wildly different classification.
 
-Additionally, S-unstable RBM model sequences may be formally connected to the near-degeneracy of Section \@ref(near-degeneracy) (in which model sequences place all probability on a small portion of their sample spaces).
+Additionally, S-unstable RBM model sequences may be formally connected to the near-degeneracy of Section \ref{near-degeneracy} (in which model sequences place all probability on a small portion of their sample spaces).
 To see this, define an arbitrary modal set of possible outcomes (i.e.
@@
 
-```{proposition, label = "degen", echo=TRUE}
+\begin{proposition}
+\label{prp:degen}
 For an S-unstable RBM model sequence and any $0 < \epsilon < 1$, 
@@
 $$
-```
+\end{proposition}
 
 In other words, S-unstable RBM model sequences are guaranteed to stack up all probability on a specific set of outcomes for visibles, which could potentially be arbitrarily narrow.
-Proofs of Propositions \@ref(prp:instab1) and \@ref(prp:degen) follow from more general results in @kaplan2016propriety.
+Proofs of Propositions \ref{prp:instab1} and \ref{prp:degen} follow from more general results in @kaplan2016propriety.
 These findings also have counterparts in results in @schweinberger2011instability, but unlike results there, are not limited in consideration to exponential family forms with a fixed number of parameters.
@@
 Note that in a RBM, the parameters $\theta_{v_1}, \dots, \theta_{v_\nv}$ and $\theta_{h_1}, \dots, \theta_{h_\nh}$ are naturally associated with main effects of visible and hidden variables and can be interpreted as (logit functions of) means for variables $V_1, \dots, V_\nv, H_1, \dots, H_\nh$ in a model with no interaction parameters, $\theta_{ij} = 0, i = 1, \dots, \nv, j = 1, \dots, \nh$.
-That is, with no interaction parameters, we have from \@ref(eq:pmf) that
+That is, with no interaction parameters, we have from \eqref{eq:pmf} that
 $$
@@
 To illustrate the ideas of model near-degeneracy, instability, and uninterpretability in a RBM, we consider first the smallest possible (toy) example that consists of one visible node $v_1$ and one hidden node $h_1$ that are both binary.
-A schematic of this model can be found in Figure \@ref(fig:toymodel).
+A schematic of this model can be found in Figure \ref{fig:toymodel}.
 Because it seems most common, we shall begin by employing $0/1$ encoding of binary variables (both $h_1$ and $v_1$ taking values in $\mathcal{C} = \{0,1\}$).
-(Eventually we shall argue in Section \@ref(data-encoding) that $-1/1$ coding has advantages.)
+(Eventually we shall argue in Section \ref{data-encoding} that $-1/1$ coding has advantages.)
 
@@
 For this small model, we are able to investigate the symptoms of model impropriety, beginning with near-degeneracy.
-To this end, recall from Section \@ref(near-degeneracy) that one characterization requires consideration of the convex hull of possible values of statistics $\boldsymbol t(\boldsymbol x)$,
+To this end, recall from Section \ref{near-degeneracy} that one characterization requires consideration of the convex hull of possible values of statistics $\boldsymbol t(\boldsymbol x)$,
 $$
@@
 As this set is in three dimensions, we are able to explicitly illustrate the shape of boundary of the convex hull of $\mathcal{T}$ and explore the behavior of the mean vector $\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta} \boldsymbol t(\boldsymbol x)$ as a function of the parameter vector $\boldsymbol \theta$.
-Figure \@ref(fig:toyhull) shows the convex hull of our "statistic space," $\mathcal{T} \subset \{0,1\}^3$, for this toy problem from two perspectives (enclosed by the unit cube $[0,1]^3$, the convex hull of $\{0,1\}^3$).
+Figure \ref{fig:toyhull} shows the convex hull of our "statistic space," $\mathcal{T} \subset \{0,1\}^3$, for this toy problem from two perspectives (enclosed by the unit cube $[0,1]^3$, the convex hull of $\{0,1\}^3$).
 In this small model, note that the convex hull of $\mathcal{T}$ does not fill the unrestricted hull of $\{0,1\}^3$ because of the relationship between the elements of $\mathcal{T} = \{(v_1, h_1, v_1 h_1 : v_1, h_1 \in \{0, 1\} \}$ (i.e.
@@
 where $\gamma(\boldsymbol \theta) = \sum\limits_{h_1 = 0}^1\sum\limits_{v_1 = 0}^1 \exp(\theta_{11} h_1 v_1 + \theta_{h_1}h_1 + \theta_{v_1}v_1)$.
-The three parametric coordinate functions of $\boldsymbol \mu(\boldsymbol \theta)$ can be represented as in Figure \@ref(fig:degen-toy).
+The three parametric coordinate functions of $\boldsymbol \mu(\boldsymbol \theta)$ can be represented as in Figure \ref{fig:degen-toy}.
 (Contour plots for three coordinate functions are shown in columns for various values of $\theta_{11}$, which can be interpreted here as an absolute log-odds ratio as the visible changes between $0$ and $1$.) In examining these, we see that as coordinates of $\boldsymbol \theta$ grow larger in magnitude, at least one mean function for the entries of $\boldsymbol t(\boldsymbol x)$ approaches a value 0 or 1, forcing $\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta} \boldsymbol t(\boldsymbol x)$ to be near to the boundary of the convex hull of $\mathcal{T}$, as a sign of model near-degeneracy.
@@
 
-```{r degen-toy, fig.cap="Contour plots for the three parametric mean functions of sufficient statistics for a RBM with one visible and one hidden node.", fig.height = 4, fig.align='center', message=FALSE, cache=TRUE}
+```{r degen-toy, fig.cap="\\label{fig:degen-toy}Contour plots for the three parametric mean functions of sufficient statistics for a RBM with one visible and one hidden node.", fig.height = 4, fig.align='center', message=FALSE, cache=TRUE}
 possibles <- stats(1, 1, type = "binary")
@@
 
-Secondly, we can look at $\text{LREP}(\boldsymbol \theta)$ from \@ref(eq:elpr) for this tiny model in order to consider model instability as a function of RBM parameters.
-Recall that large values of $\text{LREP}(\boldsymbol \theta)$ are associated with an extreme sensitivity of the model probabilities $f_{\boldsymbol \theta}(\boldsymbol x)$ to small changes in $\boldsymbol x$ (see Proposition \@ref(prp:instab1)).
+Secondly, we can look at $\text{LREP}(\boldsymbol \theta)$ from \eqref{eq:elpr} for this tiny model in order to consider model instability as a function of RBM parameters.
+Recall that large values of $\text{LREP}(\boldsymbol \theta)$ are associated with an extreme sensitivity of the model probabilities $f_{\boldsymbol \theta}(\boldsymbol x)$ to small changes in $\boldsymbol x$ (see Proposition \ref{prp:instab1}).
 The quantity $\text{LREP}(\boldsymbol \theta)$ for this small RBM is
@@
 
-Figure \@ref(fig:instab) shows contour plots of $\text{LREP}(\boldsymbol \theta)/\nv$ for various values of $\boldsymbol \theta$ in this model with $\nv = 1$.
+Figure \ref{fig:instab} shows contour plots of $\text{LREP}(\boldsymbol \theta)/\nv$ for various values of $\boldsymbol \theta$ in this model with $\nv = 1$.
 We can see that this quantity is large for large magnitudes of $\boldsymbol \theta$, especially for large values of the dependence/interaction parameter $\theta_{11}$.
@@
 
-```{r instab, fig.cap="$\\text{LREP}(\\boldsymbol \\theta)/\\nv$ for various values of $\\boldsymbol \\theta$ for the tiny example model.
+```{r instab, fig.cap="$\\label{fig:instab}\\text{LREP}(\\boldsymbol \\theta)/\\nv$ for various values of $\\boldsymbol \\theta$ for the tiny example model.
 (Recall here $\\nv$ is the number of visible nodes and here is $1$.) This quantity is large for large magnitudes of $\\boldsymbol \\theta$.", fig.height = 4, fig.align='center'}
@@
   ylab(expression(theta[h1])) +
-  scale_fill_gradient(expression(frac(LREP(theta), V)), low = "yellow", high = "red") +
+  scale_fill_gradient(expression(frac(LREP(theta), n[V])), low = "yellow", high = "red") +
   theme(aspect.ratio = 1)
@@
 
-```{r uninterp, fig.cap="The absolute difference between coordinates of model expectations, E$\\left[\\boldsymbol X | \\boldsymbol \\theta\\right]$, and expectations given independence, E$\\left[\\boldsymbol X | \\boldsymbol \\theta^* \\right ]$ for a RBM with one visible and one hidden node.
+```{r uninterp, fig.cap="\\label{fig:uninterp}The absolute difference between coordinates of model expectations, E$\\left[\\boldsymbol X | \\boldsymbol \\theta\\right]$, and expectations given independence, E$\\left[\\boldsymbol X | \\boldsymbol \\theta^* \\right ]$ for a RBM with one visible and one hidden node.
 As an indicator of uninterpretability, note that differences in expectations increase as the dependence parameter $\\theta_{11}$ deviates from zero.", fig.height = 4, fig.align='center', message=FALSE, cache=TRUE, warning=FALSE}
@@
 
-Figure \@ref(fig:uninterp) shows that the absolute difference between coordinates of the vector of model expectations, E$\left[\boldsymbol X | \boldsymbol \theta\right]$ and corresponding expectations E$\left[\boldsymbol X | \boldsymbol \theta^*\right ]$ given independence grow for the toy RBM as the values of $\boldsymbol \theta$ are farther from zero, especially for large magnitudes of the dependence parameter $\theta_{11}$.
+Figure \ref{fig:uninterp} shows that the absolute difference between coordinates of the vector of model expectations, E$\left[\boldsymbol X | \boldsymbol \theta\right]$ and corresponding expectations E$\left[\boldsymbol X | \boldsymbol \theta^*\right ]$ given independence grow for the toy RBM as the values of $\boldsymbol \theta$ are farther from zero, especially for large magnitudes of the dependence parameter $\theta_{11}$.
 This is a third indication that parameter vectors of large magnitude can readily lead to model impropriety in a RBM.
@@
 For example, we could allow hiddens $(H_1, \dots, H_\nh) \in \{0,1\}^\nh$ and visibles $(V_1, \dots, V_\nv) \in \{0,1\}^\nv$, as in the previous sections or we could instead encode the state of the variables as $\{-1,1\}^\nh$ and $\{-1,1\}^\nv$.
-This will result in variables $\boldsymbol t(\boldsymbol X)$ from \@ref(eq:t) satisfying $\boldsymbol t(\boldsymbol x) \in \{0,1\}^{\nh + \nv + \nh*\nv}$ or $\boldsymbol t(\boldsymbol x) \in \{-1,1\}^{\nh + \nv + \nh*\nv}$ depending on how we encode "on" and "off" states in the nodes.
+This will result in variables $\boldsymbol t(\boldsymbol X)$ from \eqref{eq:t} satisfying $\boldsymbol t(\boldsymbol x) \in \{0,1\}^{\nh + \nv + \nh*\nv}$ or $\boldsymbol t(\boldsymbol x) \in \{-1,1\}^{\nh + \nv + \nh*\nv}$ depending on how we encode "on" and "off" states in the nodes.
 
@@
   ungroup() %>% 
-  mutate(Hiddens = paste0("paste(n[h], '= ", H, "')"), Visibles = paste0("paste(n[v], '= ", V, "')")) -> convex_hull_summary
+  mutate(Hiddens = paste0("paste(n[H], '= ", H, "')"), Visibles = paste0("paste(n[V], '= ", V, "')")) -> convex_hull_summary
 
@@
   summarise(max_abs_diff = mean(max_abs_diff)) %>%
-  mutate(Hiddens = paste0("paste(n[h], '= ", H, "')"), Visibles = paste0("paste(n[v], '= ", V, "')")) -> exp_vals_summary
+  mutate(Hiddens = paste0("paste(n[H], '= ", H, "')"), Visibles = paste0("paste(n[V], '= ", V, "')")) -> exp_vals_summary
 
@@
 For $\nh, \nv \in \{1, \dots, 4\}$, we sample `r dim(res$samp[[1]])[1]` values of $\boldsymbol \theta$ with various magnitudes (details to follow).
-For each set of parameters we then calculate metrics of model impropriety introduced in Section \@ref(degeneracy-instability-and-uninterpretability) based on $\boldsymbol \mu(\boldsymbol \theta)$, $\text{LREP}(\boldsymbol \theta)/\nv$, and the absolute coordinates of $\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] -   \text{E}\left[\boldsymbol X | \boldsymbol \theta^* \right]$.
+For each set of parameters we then calculate metrics of model impropriety introduced in Section \ref{degeneracy-instability-and-uninterpretability} based on $\boldsymbol \mu(\boldsymbol \theta)$, $\text{LREP}(\boldsymbol \theta)/\nv$, and the absolute coordinates of $\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] -   \text{E}\left[\boldsymbol X | \boldsymbol \theta^* \right]$.
 In the case of near-degeneracy, we classify each model as near-degenerate based on the distance of $\boldsymbol \mu(\boldsymbol \theta)$ from the boundary of the convex hull of $\mathcal{T}$ and look at the fraction of models that are "near-degenerate," meaning they are within a small distance $\epsilon > 0$ of the boundary of the convex hull.
@@
 
-The results of this numerical study are summarized in Figures \@ref(fig:degen-plots), \@ref(fig:instab-plots), and \@ref(fig:uninterp-plots).
-From these three figures, it is clear that all three measures of model impropriety show higher values for larger magnitudes of the parameter vectors, supporting the RBM model properties developed in Section \@ref(background).
+The results of this numerical study are summarized in Figures \ref{fig:degen-plots}, \ref{fig:instab-plots}, and \ref{fig:uninterp-plots}.
+From these three figures, it is clear that all three measures of model impropriety show higher values for larger magnitudes of the parameter vectors, supporting the RBM model properties developed in Section \ref{background}.
 As a compounding issue, these figures show that, as models grow in size, it becomes easier for more parameter configurations to push RBM models into near-degeneracy, instability and uninterpretability.
@@
 
-```{r degen-plots, fig.cap="Results from the numerical experiment, here looking at the fraction of models that were near-degenerate for each combination of magnitudes of the points $\\boldsymbol \\theta$ and model size, where $\\boldsymbol \\theta = (\\boldsymbol \\theta_{main}, \\theta_{interaction})$ is split into main and interaction parameters.
-Black lines show the contour levels for fraction of near-degeneracy, while the thick black line shows the level where the fraction of near-degenerate models is .05.", fig.height=3.5, fig.align="ht!"}
+```{r degen-plots, fig.cap="\\label{fig:degen-plots}Results from the numerical experiment, here looking at the fraction of models that were near-degenerate for each combination of magnitudes of the points $\\boldsymbol \\theta$ and model size, where $\\boldsymbol \\theta = (\\boldsymbol \\theta_{main}, \\theta_{interaction})$ is split into main and interaction parameters.
+Black lines show the contour levels for fraction of near-degeneracy, while the thick black line shows the level where the fraction of near-degenerate models is .05.", fig.height=3.5, fig.pos="ht!"}
 three_ways %>%
@@
   facet_grid(Visibles~Hiddens, labeller = label_parsed) +
-  xlab(expression(paste(paste("||", theta[main], "||", "/(", n[h]," + ", n[v], ")")))) +
-  ylab(expression(paste(paste("||", theta[interaction], "||", "/(", n[h], "*", n[v], ")")))) +
+  xlab(expression(paste(paste("||", theta[main], "||", "/(", n[H]," + ", n[V], ")")))) +
+  ylab(expression(paste(paste("||", theta[interaction], "||", "/(", n[H], "*", n[V], ")")))) +
   theme(aspect.ratio = 1) -> p.degen
@@
   facet_grid(Visibles~Hiddens, labeller = label_parsed) +
-  xlab(expression(paste(paste("||", theta[main], "||", "/(", n[h]," + ", n[v], ")")))) +
-  ylab(expression(paste(paste("||", theta[interaction], "||", "/(", n[h], "*", n[v], ")")))) +
+  xlab(expression(paste(paste("||", theta[main], "||", "/(", n[H]," + ", n[V], ")")))) +
+  ylab(expression(paste(paste("||", theta[interaction], "||", "/(", n[H], "*", n[V], ")")))) +
   theme(aspect.ratio = 1) -> p.exp_diff
@@
   geom_abline(aes(intercept = 0, slope = 1), alpha = .5, lty = 2) +
-  scale_fill_gradient(expression(paste("Mean ", frac(LREP(theta), V))), low = "yellow", high = "red") +
+  scale_fill_gradient(expression(paste("Mean ", frac(LREP(theta), n[V]))), low = "yellow", high = "red") +
   facet_grid(Visibles~Hiddens, labeller = label_parsed) +
-  xlab(expression(paste(paste("||", theta[main], "||", "/(", n[h]," + ", n[v], ")")))) +
-  ylab(expression(paste(paste("||", theta[interaction], "||", "/(", n[h], "*", n[v], ")")))) +
+  xlab(expression(paste(paste("||", theta[main], "||", "/(", n[H]," + ", n[V], ")")))) +
+  ylab(expression(paste(paste("||", theta[interaction], "||", "/(", n[H], "*", n[V], ")")))) +
   theme(aspect.ratio = 1) -> p.elpr
@@
 
-```{r instab-plots, fig.cap="The sample mean value of $\\text{LREP}(\\boldsymbol \\theta)/V$ at each grid point for each combination of magnitudes of the points of $\\boldsymbol \\theta$ and model size.
-As the magnitude of $\\boldsymbol \\theta$ grows, so does the value of this metric, indicating typical instability in the model.", fig.height=3.5, fig.align="ht!"}
+```{r instab-plots, fig.cap="\\label{fig:instab-plots}The sample mean value of $\\text{LREP}(\\boldsymbol \\theta)/\\nv$ at each grid point for each combination of magnitudes of the points of $\\boldsymbol \\theta$ and model size.
+As the magnitude of $\\boldsymbol \\theta$ grows, so does the value of this metric, indicating typical instability in the model.", fig.height=3.5, fig.pos="ht!"}
 p.elpr
@@
 
-```{r uninterp-plots, fig.cap="The sample mean of the maximum component of the absolute difference between the model expectation vector, E$\\left[\\boldsymbol X | \\boldsymbol \\theta\\right]$, and the expectation vector given independence, E$\\left[\\boldsymbol X | \\boldsymbol \\theta^* \\right ]$.
-Larger magnitudes of $\\boldsymbol \\theta$ correspond to larger differences, thus indicating reduced interpretability.", fig.height=3.5, fig.align="ht!"}
+```{r uninterp-plots, fig.cap="\\label{fig:uninterp-plots}The sample mean of the maximum component of the absolute difference between the model expectation vector, E$\\left[\\boldsymbol X | \\boldsymbol \\theta\\right]$, and the expectation vector given independence, E$\\left[\\boldsymbol X | \\boldsymbol \\theta^* \\right ]$.
+Larger magnitudes of $\\boldsymbol \\theta$ correspond to larger differences, thus indicating reduced interpretability.", fig.height=3.5, fig.pos="ht!"}
 p.exp_diff
@@
 
-Figure \@ref(fig:split-plots-dist) shows the fraction of near-degenerate models for each magnitude of $\boldsymbol \theta$ for each model architecture.
+Figure \ref{fig:split-plots-dist} shows the fraction of near-degenerate models for each magnitude of $\boldsymbol \theta$ for each model architecture.
 For each number $\nv$ of visibles in the model, as the number $\nh$ of hiddens increase, the fraction near-degenerate diverges from zero at increasing rates for larger values of $||\boldsymbol \theta||$.
@@
 
-```{r split-plots-dist, warning=FALSE, fig.cap="The fraction of near-degenerate models for each magnitude of $\\boldsymbol \\theta$.
-For each number $\\nv$ of visibles in the model, the fraction near-degenerate moves away from zero at larger values of $||\\boldsymbol \\theta||$ as the number $\\nh$ of hidden variables increases and the slope becomes steeper as $\\nh$ increases as well.", fig.height=4, fig.align="ht!"}
+```{r split-plots-dist, warning=FALSE, fig.cap="\\label{fig:split-plots-dist}The fraction of near-degenerate models for each magnitude of $\\boldsymbol \\theta$.
+For each number $\\nv$ of visibles in the model, the fraction near-degenerate moves away from zero at larger values of $||\\boldsymbol \\theta||$ as the number $\\nh$ of hidden variables increases and the slope becomes steeper as $\\nh$ increases as well.", fig.height=4, fig.pos="ht!"}
 plot_dat_grid %>% 
@@
   summarise(frac_degen = sum(near_hull)/n()) %>%
-  mutate(Hiddens = as.character(H), Visibles = paste0("paste(n[v], '= ", V, "')")) %>%
+  mutate(Hiddens = as.character(H), Visibles = paste0("paste(n[V], '= ", V, "')")) %>%
   ggplot() +
@@
   ylim(c(0,1)) +
-  xlab(expression(paste("||", theta, "||", "/(", n[h], " + ", n[v], " + ", n[h], "*", n[v], "), Average distance from the origin"))) +
+  xlab(expression(paste("||", theta, "||", "/(", n[H], " + ", n[V], " + ", n[H], "*", n[V], "), Average distance from the origin"))) +
   ylab("Fraction Near-degenerate") +
-  scale_colour_discrete(expression(n[h]))
+  scale_colour_discrete(expression(n[H]))
 ```
@@
 
-Typically, fitting a RBM via maximum likelihood (ML) methods will be unfeasible due mainly to the intractability of the normalizing term $\gamma(\boldsymbol \theta)$ in a model \@ref(eq:pmf) of any realistic size.
+Typically, fitting a RBM via maximum likelihood (ML) methods will be unfeasible due mainly to the intractability of the normalizing term $\gamma(\boldsymbol \theta)$ in a model \eqref{eq:pmf} of any realistic size.
 Ad hoc methods are often recommended instead, which aim to avoid this problem by using stochastic ML approximations that employ a small number of MCMC draws (i.e., contrastive divergence, [@hinton2002training]).
@@
 In data applications with a large sample space, it is unlikely that the training set will include at least one of each possible vector outcome (unlike this small example).
-This implies that some RBM model parameters must diverge to $+\infty$ to mimic the empirical distribution with empty cells and, as we have already discussed in Section \@ref(explorations-of-model-properties-through-simulation), large magnitudes of $\boldsymbol \theta$ lead to model impropriety in the RBM.
+This implies that some RBM model parameters must diverge to $+\infty$ to mimic the empirical distribution with empty cells and, as we have already discussed in Section \ref{explorations-of-model-properties-through-simulation}, large magnitudes of $\boldsymbol \theta$ lead to model impropriety in the RBM.
 
@@
 To avoid model impropriety for a fitted RBM, we wish to avoid parts of the parameter space $\mathbb{R}^{\nv + \nh + \nv*\nh}$ that lead to near-degeneracy, instability, and uninterpretability.
-Motivated by the insights in Section \@ref(exploring-manageable-examples), one idea is to shrink $\boldsymbol \theta = (\boldsymbol \theta_{main}, \boldsymbol \theta_{interaction})$ toward $\boldsymbol 0$ by specifying priors that place low probability on large values of $||\boldsymbol \theta||$, specifically focusing on shrinking $\boldsymbol \theta_{interaction}$ more than $\boldsymbol \theta_{main}$.
+Motivated by the insights in Section \ref{exploring-manageable-examples}, one idea is to shrink $\boldsymbol \theta = (\boldsymbol \theta_{main}, \boldsymbol \theta_{interaction})$ toward $\boldsymbol 0$ by specifying priors that place low probability on large values of $||\boldsymbol \theta||$, specifically focusing on shrinking $\boldsymbol \theta_{interaction}$ more than $\boldsymbol \theta_{main}$.
 This is similar to an idea advocated by @hinton2010practical called *weight decay*, in which a penalty is added to the interaction terms in the model, $\boldsymbol \theta_{interaction}$, shrinking their magnitudes.
@@
 
-We considered a test case with $\nv = \nh = 4$ and parameters given in Table \@ref(tab:theta).
+We considered a test case with $\nv = \nh = 4$ and parameters given in Table \ref{tab:theta}.
 This parameter vector was chosen as a sampled value of $\boldsymbol \theta$ at which the resulting RBM model would not be clearly degenerate.
@@
 1.
-*A "trick" prior.* Here we cancel out normalizing term in the likelihood (from $\gamma(\boldsymbol \theta)$ in \@ref(eq:pmf)) so that resulting full conditionals of $\boldsymbol \theta$ are multivariate Normal.
+*A "trick" prior.* Here we cancel out normalizing term in the likelihood (from $\gamma(\boldsymbol \theta)$ in \eqref{eq:pmf}) so that resulting full conditionals of $\boldsymbol \theta$ are multivariate Normal.
 Namely, this involves a prior of the form
@@
 Additionally, BwTPLV requires additional tuning (i.e.
-a tuning parameter $C > 0$ in Table \@ref(tab:hyperparam)) to choose $C_1$ and $C_2$, reducing its appeal.
-The forms used for the hyperparameters in our simulation are presented in Table \@ref(tab:hyperparam).
+a tuning parameter $C > 0$ in Table \ref{tab:hyperparam}) to choose $C_1$ and $C_2$, reducing its appeal.
+The forms used for the hyperparameters in our simulation are presented in Table \ref{tab:hyperparam}.
 
@@
 After a burn-in of 50 iterations selected by inspecting the trace plots, we assess the issue of mixing in two ways.
-First, the autocorrelation functions (ACF) from each posterior sample corresponding to a model probability for a visible vector outcome $\mathbf{v}=(v_1,v_2,v_3,v_4)\in\{\pm 1\}^4$ (i.e., computed from $\boldsymbol \theta$ under \@ref(eq:pmf) are determined and plotted in Figure \@ref(fig:acf) with BwTNLV in black and BwTNML in red.
+First, the autocorrelation functions (ACF) from each posterior sample corresponding to a model probability for a visible vector outcome $\mathbf{v}=(v_1,v_2,v_3,v_4)\in\{\pm 1\}^4$ (i.e., computed from $\boldsymbol \theta$ under \eqref{eq:pmf} are determined and plotted in Figure \ref{fig:acf} with BwTNLV in black and BwTNML in red.
 As expected, ACF corresponding to the method (BwTNML) that marginalizes out the hidden variables from the likelihood decreases to zero at a much faster rate, indicating better mixing for the chain.
@@
 
-```{r acf, fig.cap="The autocorrelation functions (ACF) for the posterior probabilityies of all $2^4 = 16$ possible outcomes for the vector of four visibles assessed at multiple lags for each method with BwTNLV in black and BwTNML in red.
+```{r acf, fig.cap="\\label{fig:acf}The autocorrelation functions (ACF) for the posterior probabilityies of all $2^4 = 16$ possible outcomes for the vector of four visibles assessed at multiple lags for each method with BwTNLV in black and BwTNML in red.
 As expected, ACF corresponding to the method that marginalizes out the hidden variables from the likelihood decreases to zero at a much faster rate, indicating better mixing for the chain."}
@@
 
-For both BwTNLV and BwTNML methods, effective sample sizes for a chain of length $M = 1000$ for inference about each of the $2^4 = 16$ model probabilities are presented in Table \@ref(tab:m-eff).
+For both BwTNLV and BwTNML methods, effective sample sizes for a chain of length $M = 1000$ for inference about each of the $2^4 = 16$ model probabilities are presented in Table \ref{tab:m-eff}.
 These range from  $`r round(min(M_eff_tmp[, "BwTNML"]), 2)`$ to $`r round(max(M_eff_tmp[, "BwTNML"]), 2)`$ for BwTNML, while BwTNLV only yields between $`r round(min(M_eff_tmp[, "BwTNLV"]), 2)`$ and $`r round(max(M_eff_tmp[, "BwTNLV"]), 2)`$ effective draws.
 Thus, BwTNLV would require at least $`r round(min(M_eff_tmp[, "BwTNML"])/min(M_eff_tmp[, "BwTNLV"]), 1)`$ times as many iterations of the MCMC chain to be run in order to achieve the same amount of effective information about the posterior distribution.
-For this reason, consistent with the ACF results in Figure \@ref(fig:acf), BwTNLV does not seem to be an effective method for fitting the RBM, though computing resources can hinder use of the alternative BwTNML involving marginalization of hidden variables.
-
-Figure \@ref(fig:fitting-plot) shows the posterior probability of each possible $\boldsymbol v \in \{-1,1\}^4$ after fitting the RBM model according to method 1 (BwTPLV using trick prior) and method 3 (BwTNML) (excluding method 2 (BwTNLV) that seeks the same posterior as method 3) .
+For this reason, consistent with the ACF results in Figure \ref{fig:acf}, BwTNLV does not seem to be an effective method for fitting the RBM, though computing resources can hinder use of the alternative BwTNML involving marginalization of hidden variables.
+
+Figure \ref{fig:fitting-plot} shows the posterior probability of each possible $\boldsymbol v \in \{-1,1\}^4$ after fitting the RBM model according to method 1 (BwTPLV using trick prior) and method 3 (BwTNML) (excluding method 2 (BwTNLV) that seeks the same posterior as method 3) .
 The black vertical lines show the true probabilities of each image based on the parameters used to generate the training set while the red vertical lines show the empirical distribution for the training set of $5,000$ vectors.
 From these posterior predictive checks, it is evident that BwTNML produces the best fit to the data.
-Furthermore, along with the discussion of Section \@ref(model-fitting), Figure \@ref(fig:fitting-plot) also shows that it can be undesirable to seek to perfectly re-create an empirical distribution in fitting RBM models (i.e., true model probabilities may differ substantially).
+Furthermore, along with the discussion of Section \ref{model-fitting}, Figure \ref{fig:fitting-plot} also shows that it can be undesirable to seek to perfectly re-create an empirical distribution in fitting RBM models (i.e., true model probabilities may differ substantially).
 The priors in the BwTNML method constrain the RBM model fit to avoid replication of the empirical distribution and better estimate the underlying true data generating probabilities.
@@
 
-```{r fitting-plot, fig.cap="Posterior probabilities of $16 = 2^4$ possible realizations of $4$ visibles using two of the three Bayesian fitting techniques, BwTPLV and BwTNML.
+```{r fitting-plot, fig.cap="\\label{fig:fitting-plot}Posterior probabilities of $16 = 2^4$ possible realizations of $4$ visibles using two of the three Bayesian fitting techniques, BwTPLV and BwTNML.
 The black vertical lines show the true probabilities of each vector of visibles based on the parameters used to generate the training data while the red vertical lines show the empirical distribution.
@@
 Additionally, these models are difficult to fit using a rigorous methodology that can incorporate estimation uncertainty, due to the dimension of the parameter space coupled with the size of the latent variable space.
-We have presented three fully Bayes-principled MCMC-based methods for fitting RBMs, which provide posterior distributions for quantifying uncertainty in estimating model parameters and probabilities (e.g., Figure \@ref(fig:fitting-plot)) while aiming to avoid areas of the parameter space that will potentially lead to model impropriety.
+We have presented three fully Bayes-principled MCMC-based methods for fitting RBMs, which provide posterior distributions for quantifying uncertainty in estimating model parameters and probabilities (e.g., Figure \ref{fig:fitting-plot}) while aiming to avoid areas of the parameter space that will potentially lead to model impropriety.
 We have shown small examples where this fitting methodology does work, but becomes quickly intractable in large data cases.

### 2018-09-18 writing/paper-rbm/paper.tex
include new files

--- 2018-09-14 writing/paper-rbm/docs/paper.tex
@@
-\documentclass[]{article}
-%\usepackage{nips15submit_e,times}
+\documentclass[AMS,STIX1COL]{WileyNJD-v2}
+
+\articletype{Research Article}
+
+\received{17 September 2018}
+
+\raggedbottom
+
+\usepackage{amsmath}
+% \usepackage{amsthm}
+% \usepackage{cite}
+\usepackage{hyperref}
+\usepackage{graphics}
+\usepackage{subfig}
 \usepackage{url}
-
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-
-\usepackage{lmodern}
-\usepackage{amssymb,amsmath}
-\usepackage{ifxetex,ifluatex}
-\usepackage{fixltx2e} % provides \textsubscript
-\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
-  \usepackage[T1]{fontenc}
-  \usepackage[utf8]{inputenc}
-\else % if luatex or xelatex
-  \ifxetex
-    \usepackage{mathspec}
-    \usepackage{xltxtra,xunicode}
-  \else
-    \usepackage{fontspec}
-  \fi
-  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
-  \newcommand{\euro}{€}
-\fi
-% use upquote if available, for straight quotes in verbatim environments
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-% use microtype if available
-\IfFileExists{microtype.sty}{%
-\usepackage{microtype}
-\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
-}{}
-\usepackage{longtable,booktabs}
-\usepackage{graphicx}
-\makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
-\makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
-\ifxetex
-  \usepackage[setpagesize=false, % page size defined by xetex
-              unicode=false, % unicode breaks when used with xetex
-              xetex]{hyperref}
-\else
-  \usepackage[unicode=true]{hyperref}
-\fi
-\hypersetup{breaklinks=true,
-            bookmarks=true,
-            pdfauthor={},
-            pdftitle={Properties and Bayesian fitting of restricted Boltzmann machines},
-            colorlinks=true,
-            citecolor=blue,
-            urlcolor=blue,
-            linkcolor=magenta,
-            pdfborder={0 0 0}}
-\urlstyle{same}  % don't use monospace font for urls
-\setlength{\parindent}{0pt}
-\setlength{\parskip}{6pt plus 2pt minus 1pt}
-\setlength{\emergencystretch}{3em}  % prevent overfull lines
-\setcounter{secnumdepth}{5}
-
-% Tech stuff starts here
-\usepackage{psfrag,epsf}
-%\pdfminorversion=4
 
@@
 %%INCLUDED-AK
-\usepackage{amsthm}
-\newtheorem{theorem}{Theorem}
-\newtheorem{proposition}{Proposition}
-\newtheorem{lemma}{Lemma}
-\newtheorem{corollary}{Corollary}
-
-\theoremstyle{definition}
-\newtheorem{definition}{Definition}
-\newtheorem{assumption}{Assumption}
-\newtheorem{remark}{Remark}
+% \usepackage{amsthm}
+% \newtheorem{theorem}{Theorem}
+% \newtheorem{proposition}{Proposition}
+% \newtheorem{lemma}{Lemma}
+% \newtheorem{corollary}{Corollary}
+% 
+% \theoremstyle{definition}
+% \newtheorem{definition}{Definition}
+% \newtheorem{assumption}{Assumption}
+% \newtheorem{remark}{Remark}
 
@@
 \newcommand{\Gamt}{\Gamma_{N,2}(\boldsymbol \theta_N) }
-\bigskip
-
-% DON'T change margins - should be 1 inch all around.
-\addtolength{\oddsidemargin}{-.5in}%
-\addtolength{\evensidemargin}{-.5in}%
-\addtolength{\textwidth}{1in}%
-\addtolength{\textheight}{1.3in}%
-\addtolength{\topmargin}{-.8in}%
-
-\let\BeginKnitrBlock\begin \let\EndKnitrBlock\end
+
+%% rmarkdown being obstinant?
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
+
+%hacks
+%https://tex.stackexchange.com/questions/371955/hyperref-error-argument-of-hysetreflink-has-an-extra/384969
+\AtBeginDocument{
+  \label{FirstPage2}
+  \def\fpage{\pageref{FirstPage2}}
+}
+
 \begin{document}
 
-\def\spacingset#1{\renewcommand{\baselinestretch}%
-{#1}\small\normalsize} \spacingset{1}
-
-\title{\bf Properties and Bayesian fitting of restricted Boltzmann machines}
-\author{Andee Kaplan \\ Department of Statistical Science, Duke University \\  and \\ Daniel Nordman \\ Department of Statistics, Iowa State University \\  and \\ Stephen Vardeman \\ Department of Statistics and Department of Industrial and\\
-Manufacturing Systems Engineering, Iowa State University \\ }
-
-\maketitle
-\begin{abstract}
+\title{Properties and Bayesian fitting of restricted Boltzmann machines}
+\authormark{KAPLAN \textsc{et al}}
+
+\author[1]{Andee Kaplan*}
+\author[2]{Daniel Nordman}
+\author[2,3]{Stephen Vardeman}
+
+\address[1]{
+\orgdiv{Department of Statistical Science}, \orgname{Duke University}, \orgaddress{\state{North Carolina}, \country{USA}}}
+\address[2]{
+\orgdiv{Department of Statistics}, \orgname{Iowa State University}, \orgaddress{\state{Iowa}, \country{USA}}}
+\address[3]{
+\orgdiv{Department of Industrial and Manufacturing Systems Engineering}, \orgname{Iowa State University}, \orgaddress{\state{Iowa}, \country{USA}}}
+
+\corres{*Andee Kaplan, Department of Statistical Science, Duke University, P.O. Box 90251,
+Durham, NC 27708-0251. \email{\href{mailto:andrea.kaplan@duke.edu}{\nolinkurl{andrea.kaplan@duke.edu}}}}
+
+\abstract[Summary]{
 A restricted Boltzmann machine (RBM) is an undirected graphical model
@@
 model structure.
-\end{abstract}
-\noindent%
-{\it Keywords:}  Degeneracy, Instability, Classification, Deep Learning, Graphical Models
-\vfill
-
-\hfill {\tiny technometrics tex template (do not remove)}
-\newpage
-\spacingset{1.45} % DON'T change the spacing!
+}
+\keywords{Degeneracy, Instability, Classification, Deep Learning, Graphical Models}
+
+\jnlcitation{\cname{
+\author{Kaplan A.,} 
+\author{D. Nordman, and} 
+\author{S. Vardeman} 
+ (\cyear{2018+}), 
+\ctitle{Properties and Bayesian fitting of restricted Boltzmann machines}, \cjournal{Stat Anal DataMin: The ASA Data Sci Journal}, \cvol{2018+;00:00-00}.}}
+
+\maketitle
 
@@
 great interest in deep learning, specifically in stacked restricted
-Boltzmann machines (RBMs) (see Salakhutdinov and Hinton
-\protect\hyperlink{ref-salakhutdinov2009deep}{2009},
-\protect\hyperlink{ref-salakhutdinov2012efficient}{2012}; Srivastava,
-Salakhutdinov, and Hinton
-\protect\hyperlink{ref-srivastava2013modeling}{2013}; Le Roux and Bengio
-\protect\hyperlink{ref-le2008representational}{2008} for examples). A
-RBM is a probabilistic undirected graphical model (for discrete or
-continuous random variables) with two layers, one hidden and one
-visible, with no conditional dependency within a layer (Smolensky
-\protect\hyperlink{ref-smolensky1986information}{1986}). These models
-have reportedly been used with success in classification of images
-(Larochelle and Bengio
-\protect\hyperlink{ref-larochelle2008classification}{2008}; Srivastava
-and Salakhutdinov
-\protect\hyperlink{ref-srivastava2012multimodal}{2012}). However, the
-model properties are largely unexplored in the literature and the
-commonly cited fitting methodology remains heuristic-based and relies on
-rough approximation (Hinton, Osindero, and Teh
-\protect\hyperlink{ref-hinton2006fast}{2006}). Additionally, the current
-fitting methodology does not easily allow for interval estimation of
-parameters in the model, and instead only provides for point estimation.
-In this paper, we provide steps toward a fuller understanding of the
-model class and its properties from the perspective of statistical model
-theory, and we then explore the possibility of a rigorous fitting
-methodology with natural uncertainty quantification. We find the RBM
-model class to be concerning in two fundamental ways.
+Boltzmann machines (RBMs) (see Salakhutdinov and Hinton 2009, 2012;
+Srivastava, Salakhutdinov, and Hinton 2013; Le Roux and Bengio 2008 for
+examples). A RBM is a probabilistic undirected graphical model (for
+discrete or continuous random variables) with two layers, one hidden and
+one visible, with no conditional dependency within a layer (Smolensky
+1986). These models have reportedly been used with success in
+classification of images (Larochelle and Bengio 2008; Srivastava and
+Salakhutdinov 2012). However, the model properties are largely
+unexplored in the literature and the commonly cited fitting methodology
+remains heuristic-based and relies on rough approximation (Hinton,
+Osindero, and Teh 2006). Additionally, the current fitting methodology
+does not easily allow for interval estimation of parameters in the
+model, and instead only provides for point estimation. In this paper, we
+provide steps toward a fuller understanding of the model class and its
+properties from the perspective of statistical model theory, and we then
+explore the possibility of a rigorous fitting methodology with natural
+uncertainty quantification. We find the RBM model class to be concerning
+in two fundamental ways.
 
 First, the models can be unsatisfactory as conceptualizations of how
-data are generated. That is, recalling Fisher
-(\protect\hyperlink{ref-fisher1922mathematical}{1922}), the aim of a
+data are generated. That is, recalling Fisher (1922), the aim of a
 statistical model is to represent data in a compact way. Neyman and Box
 further state that a model should ``provide an explanation of the
-mechanism underlying the observed phenomena'' (Lehmann
-\protect\hyperlink{ref-lehmann1990model}{1990}; G. E. P. Box
-\protect\hyperlink{ref-box1967discrimination}{1967}). At issue,
-simulation from RBMs can often produce data lacking realistic
-variability so that such models may thereby fail to satisfactorily
-reflect an observed data generation process. Such behavior relates to
-model degeneracy (or near degeneracy), which is a statistical concern in
-that many data processes of interest are realistically not degenerate in
-their spectrum of potential outcomes. For example, when sampling data
-from a nearly degenerate RBM used to model an imaging process, only a
-small set of output possibilities receives nearly all probability, and
-thus a sample of images will all be copies of the same one image (or
-small number of images). An example of ten 4-pixel images simulated from
-a nearly degenerate RBM model is compared to ten 4-pixel images
-simulated from a non-degenerate RBM model in Figure
-\ref{fig:sample-models}. The degenerate model places almost all
-probability on one outcome, causing the image to be generated
+mechanism underlying the observed phenomena'' (Lehmann 1990; G. E. P.
+Box 1967). At issue, simulation from RBMs can often produce data lacking
+realistic variability so that such models may thereby fail to
+satisfactorily reflect an observed data generation process. Such
+behavior relates to model degeneracy (or near degeneracy), which is a
+statistical concern in that many data processes of interest are
+realistically not degenerate in their spectrum of potential outcomes.
+For example, when sampling data from a nearly degenerate RBM used to
+model an imaging process, only a small set of output possibilities
+receives nearly all probability, and thus a sample of images will all be
+copies of the same one image (or small number of images). An example of
+ten 4-pixel images simulated from a nearly degenerate RBM model is
+compared to ten 4-pixel images simulated from a non-degenerate RBM model
+in Figure \ref{fig:sample-models}. The degenerate model places almost
+all probability on one outcome, causing the image to be generated
 repeatedly, whereas the non-degenerate model shows more potentially
@@
 \par
+
 \begin{figure}
-\includegraphics[width=1\linewidth]{paper_files/figure-latex/sample-models-1} \includegraphics[width=1\linewidth]{paper_files/figure-latex/sample-models-2} \caption{Ten 4-pixel images simulated from a degenerate model (a) compared to ten 4-pixel images simulated from a non-degenerate model (b).
+\includegraphics[width=1\linewidth]{paper_files/figure-latex/sample-models-1} \includegraphics[width=1\linewidth]{paper_files/figure-latex/sample-models-2} \caption{\label{fig:sample-models}Ten 4-pixel images simulated from a degenerate model (a) compared to ten 4-pixel images simulated from a non-degenerate model (b).
 The degenerate model places almost all probability on one outcome, causing the image to be generated repeatedly, whereas the non-degenerate model shows more realistic variation.}\label{fig:sample-models}
 \end{figure}
+
 In addition to such degeneracy, we find that RBMs can easily exhibit
@@
 different classification in an image classification problem. Occurrences
-of such behavior have recently been documented in RBMs (Li
-\protect\hyperlink{ref-li2014biclustering}{2014}), as well as other deep
-architectures (Szegedy et al.
-\protect\hyperlink{ref-szegedy2013intriguing}{2013}; Nguyen, Yosinski,
-and Clune \protect\hyperlink{ref-nguyen2014deep}{2014}). We describe
-(related) potential issues of model instability, degeneracy and
-uninterpretability for the RBM class (which are related properties) and
-examine the presence of these in Section
+of such behavior have recently been documented in RBMs (Li 2014), as
+well as other deep architectures (Szegedy et al. 2013; Nguyen, Yosinski,
+and Clune 2014). We describe (related) potential issues of model
+instability, degeneracy and uninterpretability for the RBM class (which
+are related properties) and examine the presence of these in Section
 \ref{explorations-of-model-properties-through-simulation} through
@@
 and visible variables), but little is said about the attributes of
-realized estimates (Hinton
-\protect\hyperlink{ref-hinton2010practical}{2010}; Hinton, Osindero, and
-Teh \protect\hyperlink{ref-hinton2006fast}{2006}), including the absence
-of interval estimates.
+realized estimates (Hinton 2010; Hinton, Osindero, and Teh 2006),
+including the absence of interval estimates.
 
@@
 For a RBM model with enough hidden variables, any distribution for the
-visibles can be approximated arbitrarily well (Le Roux and Bengio
-\protect\hyperlink{ref-le2008representational}{2008}; Montufar and Ay
-\protect\hyperlink{ref-montufar2011refinements}{2011}; and Montúfar,
-Rauh, and Ay \protect\hyperlink{ref-montufar2011expressive}{2011}).
-However, for binary data, the empirical distribution of an observed
-training set of visible variables provides a highest likelihood
-benchmark, before any parametric model class is even introduced and
-applied to obtain a refinement of model fit. As a consequence, we find
-that any fully principled fitting method based on the likelihood for a
-RBM with enough hidden variables will seek to reproduce the (discrete)
-empirical distribution of a training set. This aspect can be
-undesirable, and perhaps even unexpected compared to most modeling
-scenarios, in that no ``smoothed distribution'' may result when fitting
-a RBM model of sufficient size with a rigorous likelihood-based method.
-We are therefore led to be skeptical that models that involve these
-structures (like deep Boltzmann machines) can achieve useful prediction
-or inference in a principled way without intentionally limiting the
+visibles can be approximated arbitrarily well (Le Roux and Bengio 2008;
+Montufar and Ay 2011; and Montúfar, Rauh, and Ay 2011). However, for
+binary data, the empirical distribution of an observed training set of
+visible variables provides a highest likelihood benchmark, before any
+parametric model class is even introduced and applied to obtain a
+refinement of model fit. As a consequence, we find that any fully
+principled fitting method based on the likelihood for a RBM with enough
+hidden variables will seek to reproduce the (discrete) empirical
+distribution of a training set. This aspect can be undesirable, and
+perhaps even unexpected compared to most modeling scenarios, in that no
+``smoothed distribution'' may result when fitting a RBM model of
+sufficient size with a rigorous likelihood-based method. We are
+therefore led to be skeptical that models that involve these structures
+(like deep Boltzmann machines) can achieve useful prediction or
+inference in a principled way without intentionally limiting the
 flexibility of the fitted model.
@@
 over-fitting and poor mixing in the Markov chain during fitting (Hinton
-\protect\hyperlink{ref-hinton2010practical}{2010}; Tieleman
-\protect\hyperlink{ref-tieleman2008training}{2008}; Cho, Ilin, and Raiko
-\protect\hyperlink{ref-cho2012tikhonov}{2012}). Both \(L_1\) and \(L_2\)
-type penalties are mentioned to achieve weight-decay, though the
+2010; Tieleman 2008; Cho, Ilin, and Raiko 2012). Both \(L_1\) and
+\(L_2\) type penalties are mentioned to achieve weight-decay, though the
 benefits of these particular forms are unknown in different situations.
@@
 part of the standard overall model-specification-and-fitting methodology
-(Hinton \protect\hyperlink{ref-hinton2002training}{2002};
-Carreira-Perpinan and Hinton
-\protect\hyperlink{ref-carreira2005contrastive}{2005}). In this paper,
-we attempt to address the concerns with overfitting and poor mixing in
-an alternative (and perhaps more transparent) manner, via specification
-of a Bayesian prior in Section \ref{bayesian-model-fitting}.
+(Hinton 2002; Carreira-Perpinan and Hinton 2005). In this paper, we
+attempt to address the concerns with overfitting and poor mixing in an
+alternative (and perhaps more transparent) manner, via specification of
+a Bayesian prior in Section \ref{bayesian-model-fitting}.
 
@@
 the RBM class. This work aligns with the call to action by Google
-scientists in Sculley et al.
-(\protect\hyperlink{ref-sculley2018winner}{2018}) to employ small toy
-examples to advance empirical understanding of deep learning models and
-strive for a ``higher level of empirical rigor in the field.'' We
-examine three Bayesian fitting techniques intended to avoid model
-impropriety issues raised in Section \ref{model-fitting} and conclude
-with a discussion in Section \ref{discussion}. Supplementary online
-material provides proofs for results on RBM parameterizations and data
-codings described in Section \ref{data-encoding}.
+scientists in Sculley et al. (2018) to employ small toy examples to
+advance empirical understanding of deep learning models and strive for a
+``higher level of empirical rigor in the field.'' We examine three
+Bayesian fitting techniques intended to avoid model impropriety issues
+raised in Section \ref{model-fitting} and conclude with a discussion in
+Section \ref{discussion}. Supplementary online material provides proofs
+for results on RBM parameterizations and data codings described in
+Section \ref{data-encoding}.
 
@@
 visible nodes indicated by white circles.
+
 \begin{figure}
@@
 \end{figure}
+
 A common use for RBMs is to create features for use in classification.
@@
 treats the pixel values as the visible variables \(v_i\) in a RBM model
-(Hinton, Osindero, and Teh
-\protect\hyperlink{ref-hinton2006fast}{2006}).
+(Hinton, Osindero, and Teh 2006).
 
@@
 simulation of both hidden layers and model parameters (e.g., see the
-contrastive divergence of Hinton
-(\protect\hyperlink{ref-hinton2002training}{2002}) or Bayes methods in
-Section \ref{model-fitting}).
+contrastive divergence of Hinton (2002) or Bayes methods in Section
+\ref{model-fitting}).
 
@@
 RBMs have risen to prominence in recent years due to their connection to
-deep learning (see Hinton, Osindero, and Teh
-\protect\hyperlink{ref-hinton2006fast}{2006}; Salakhutdinov and Hinton
-\protect\hyperlink{ref-salakhutdinov2012efficient}{2012}; Srivastava,
-Salakhutdinov, and Hinton
-\protect\hyperlink{ref-srivastava2013modeling}{2013} for examples). By
-stacking multiple layers of RBMs in a deep architecture, proponents of
-the methodology claim to produce the ability to learn ``internal
+deep learning (see Hinton, Osindero, and Teh 2006; Salakhutdinov and
+Hinton 2012; Srivastava, Salakhutdinov, and Hinton 2013 for examples).
+By stacking multiple layers of RBMs in a deep architecture, proponents
+of the methodology claim to produce the ability to learn ``internal
 representations that become increasingly complex, which is considered to
 be a promising way of solving object and speech recognition problems''
-(Salakhutdinov and Hinton
-\protect\hyperlink{ref-salakhutdinov2009deep}{2009}, 450). The stacking
-is achieved by treating a hidden layer of one RBM as the visible layer
-in a second RBM, and so on, until the desired multi-layer architecture
-is created.
+(Salakhutdinov and Hinton 2009, 450). The stacking is achieved by
+treating a hidden layer of one RBM as the visible layer in a second RBM,
+and so on, until the desired multi-layer architecture is created.
 
@@
 disproportionate amount of probability placed on only a few elements of
-the sample space, \(\mathcal{X}\), by the model (Handcock
-\protect\hyperlink{ref-handcock2003assessing}{2003}). For random graph
-models, \(\mathcal{X}\) denotes all possible graphs that can be
-constructed from a set of nodes and an exponentially parameterized
-random graph model has a distribution with a probability mass function
-of the form \begin{align*}
+the sample space, \(\mathcal{X}\), by the model (Handcock 2003). For
+random graph models, \(\mathcal{X}\) denotes all possible graphs that
+can be constructed from a set of nodes and an exponentially
+parameterized random graph model has a distribution with a probability
+mass function of the form \begin{align*}
 f_{\boldsymbol \theta} (\boldsymbol x) = \frac{\exp\left(\boldsymbol \theta^T \boldsymbol t(\boldsymbol x)\right)}{\gamma(\boldsymbol \theta)}, \boldsymbol x \in \mathcal{X},
@@
 \(\{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{X}\}\),
-under the model above. Handcock
-(\protect\hyperlink{ref-handcock2003assessing}{2003}) classifies an
-exponentially parameterized random graph model at \(\boldsymbol \theta\)
-as \emph{near-degenerate} if the mean value of the vector of sufficient
+under the model above. Handcock (2003) classifies an exponentially
+parameterized random graph model at \(\boldsymbol \theta\) as
+\emph{near-degenerate} if the mean value of the vector of sufficient
 statistics under \(\boldsymbol \theta\),
@@
 visibles and hiddens, \(\mathcal{C}^{\nv + \nh}\). Using the idea of
-Handcock (\protect\hyperlink{ref-handcock2003assessing}{2003}), when the
-random vector
+Handcock (2003), when the random vector
 \(\boldsymbol t(\boldsymbol x) = (v_1, \dots, v_\nv, h_1, \dots, h_\nh, v_1 h_1, \dots, v_V h_\nh ) \in \mathcal{T} \equiv \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{C}^{\nh + \nv} \}\)
@@
 \end{align*}
+
 \hypertarget{instability}{%
@@
 
-Considering exponential families of distributions, Schweinberger
-(\protect\hyperlink{ref-schweinberger2011instability}{2011}) introduced
-a concept of model deficiency related to \emph{instability}. Instability
-can be roughly thought of as excessive sensitivity in the model, where
-small changes in the components of potential data outcomes,
+Considering exponential families of distributions, Schweinberger (2011)
+introduced a concept of model deficiency related to \emph{instability}.
+Instability can be roughly thought of as excessive sensitivity in the
+model, where small changes in the components of potential data outcomes,
 \(\boldsymbol x\), may lead to substantial changes in the probability
@@
 rigorously (particularly beyond the definition given by Schweinberger
-(\protect\hyperlink{ref-schweinberger2011instability}{2011})) it is
-useful to consider how RBM models might be expanded to incorporate more
-and more visibles. When increasing the size of RBM models, it becomes
-necessary to grow the number of model parameters (and in this process
-one may also arbitrarily expand the number of hidden variables used). To
-this end, let
+(2011)) it is useful to consider how RBM models might be expanded to
+incorporate more and more visibles. When increasing the size of RBM
+models, it becomes necessary to grow the number of model parameters (and
+in this process one may also arbitrarily expand the number of hidden
+variables used). To this end, let
 \(\boldsymbol \theta_{\nv} \equiv (\theta_{v_1}, \dots, \theta_{v_\nv}, \theta_{h_1}, \dots, \theta_{h_\nh}, \theta_{11}, \dots, \theta_{\nv \nh}), \nv \ge 1\),
@@
 stable. That is, we make the following convention.
-\BeginKnitrBlock{definition}[S-unstable RBM]
-\protect\hypertarget{def:unnamed-chunk-1}{}{\label{def:unnamed-chunk-1}
-\iffalse (S-unstable RBM) \fi{} }Let
-\(\boldsymbol \theta_\nv \in \mathbb{R}^{\nv + \nh + \nh*\nv}, \nv \ge 1\),
-be an element of a sequence of RBM parameters where the number of
-hiddens, \(\nh \equiv \nh(\nv) \ge 1\), can be an arbitrary function of
-the number \(\nv\) of visibles. A RBM model formulation is
-\emph{Schweinberger-unstable} or \emph{S-unstable} if \begin{align*}
+
+\begin{definition}[S-unstable RBM]
+Let $\boldsymbol \theta_\nv \in \mathbb{R}^{\nv + \nh +  \nh*\nv}, \nv \ge 1$, be an element of a sequence of RBM parameters where the number of hiddens, $\nh \equiv \nh(\nv) \ge 1$, can be an arbitrary function of the number $\nv$ of visibles.
+A RBM model formulation is \emph{Schweinberger-unstable} or \emph{S-unstable} if
+\begin{align*}
 \lim\limits_{\nv \rightarrow \infty} \frac{1}{\nv} \text{LREP}(\boldsymbol \theta_\nv) = \infty.
 \end{align*}
-\EndKnitrBlock{definition}
+\end{definition}
 
@@
 generalization or re-interpretation of the ``unstable'' concept of
-Schweinberger
-(\protect\hyperlink{ref-schweinberger2011instability}{2011}) in that
-here RBM models for visibles \((v_1, \dots, v_\nv)\) do not form an
-exponential family and the dimensionality of \(\boldsymbol \theta_\nv\)
-is not fixed, but rather grows with \(\nv\).
+Schweinberger (2011) in that here RBM models for visibles
+\((v_1, \dots, v_\nv)\) do not form an exponential family and the
+dimensionality of \(\boldsymbol \theta_\nv\) is not fixed, but rather
+grows with \(\nv\).
 
@@
 
-\BeginKnitrBlock{proposition}
-\protect\hypertarget{prp:instab1}{}{\label{prp:instab1} }Let \(c > 0\) and
-let \(\text{LREP}(\boldsymbol \theta_\nv)\) be as in \eqref{eq:elpr} for
-an integer \(\nv \ge 1\). If
-\(\frac{1}{\nv}\text{LREP}(\boldsymbol \theta_\nv) > c\), then
-\(\Delta(\boldsymbol \theta_\nv) > c\).
-\EndKnitrBlock{proposition}
+\begin{proposition}
+\label{prp:instab1}
+Let $c > 0$ and let $\text{LREP}(\boldsymbol \theta_\nv)$ be as in \eqref{eq:elpr} for an integer $\nv \ge 1$.
+If $\frac{1}{\nv}\text{LREP}(\boldsymbol \theta_\nv) > c$, then $\Delta(\boldsymbol \theta_\nv) > c$.
+\end{proposition}
 
@@
 different classification of the image. Examples of this kind of behavior
-have been presented in Szegedy et al.
-(\protect\hyperlink{ref-szegedy2013intriguing}{2013}) for deep learning
-models, in which a one pixel change in a test image results in a wildly
+have been presented in Szegedy et al. (2013) for deep learning models,
+in which a one pixel change in a test image results in a wildly
 different classification.
@@
 
-\BeginKnitrBlock{proposition}
-\protect\hypertarget{prp:degen}{}{\label{prp:degen}}For an S-unstable RBM
-model sequence and any \(0 < \epsilon < 1\), \[
+\begin{proposition}
+\label{prp:degen}
+For an S-unstable RBM model sequence and any $0 < \epsilon < 1$, 
+$$
 P_{\boldsymbol \theta_\nv}\left((v_1, \dots, v_\nv) \in M_{\epsilon, \boldsymbol \theta_\nv}\right) \rightarrow 1 \text{ as } \nv \rightarrow \infty.
-\]
-\EndKnitrBlock{proposition}
+$$
+\end{proposition}
 
@@
 \ref{prp:instab1} and \ref{prp:degen} follow from more general results
-in Kaplan, Nordman, and Vardeman
-(\protect\hyperlink{ref-kaplan2016propriety}{2017}). These findings also
-have counterparts in results in Schweinberger
-(\protect\hyperlink{ref-schweinberger2011instability}{2011}), but unlike
-results there, are not limited in consideration to exponential family
-forms with a fixed number of parameters.
+in Kaplan, Nordman, and Vardeman (2017). These findings also have
+counterparts in results in Schweinberger (2011), but unlike results
+there, are not limited in consideration to exponential family forms with
+a fixed number of parameters.
 
@@
 
-For spatial Markov models, Kaiser
-(\protect\hyperlink{ref-kaiser2007statistical}{2007}) defines a measure
-of model impropriety he calls \emph{uninterpretability}, which is
-characterized by dependence parameters in a model being so extreme that
-marginal mean-structures fail to hold as anticipated by consideration of
-a model statement. We adapt this notion to RBM models. Note that in a
-RBM, the parameters \(\theta_{v_1}, \dots, \theta_{v_\nv}\) and
+For spatial Markov models, Kaiser (2007) defines a measure of model
+impropriety he calls \emph{uninterpretability}, which is characterized
+by dependence parameters in a model being so extreme that marginal
+mean-structures fail to hold as anticipated by consideration of a model
+statement. We adapt this notion to RBM models. Note that in a RBM, the
+parameters \(\theta_{v_1}, \dots, \theta_{v_\nv}\) and
 \(\theta_{h_1}, \dots, \theta_{h_\nh}\) are naturally associated with
@@
 parameters, \(\theta_{ij} = 0, i = 1, \dots, \nv, j = 1, \dots, \nh\).
-That is, with no interaction parameters, we have from \eqref{eq:pmf} that
-\[
+That is, with no interaction parameters, we have from \eqref{eq:pmf}
+that \[
 P_\theta(V_i=1) \propto e^{\theta_{v_i}} \quad \text{ and } \quad P_\theta(H_j=1) \propto e^{\theta_{h_j}}, \quad  i=1,\ldots,\nv,j=1,\ldots,\nh
@@
 additional device for examining other aspects of model propriety related
-to instability and degeneracy. As explained by Kaiser
-(\protect\hyperlink{ref-kaiser2007statistical}{2007}), models with
-interpretable dependence parameters typically correspond to
+to instability and degeneracy. As explained by Kaiser (2007), models
+with interpretable dependence parameters typically correspond to
 non-degenerate models, while degradation in interpretability is often
@@
 \end{align*}
+
 \hypertarget{explorations-of-model-properties-through-simulation}{%
@@
 coding has advantages.)
+
 \begin{figure}[ht]
@@
 \end{figure}
+
 \hypertarget{impropriety-three-ways}{%
@@
 \par
+
 \begin{figure}[ht]
@@
 \end{figure}
+
 We can compute the mean vector for \(\boldsymbol t(\boldsymbol x)\) as a
@@
 \par
+
 \begin{figure}
@@
 
-\caption{Contour plots for the three parametric mean functions of sufficient statistics for a RBM with one visible and one hidden node.}\label{fig:degen-toy}
+\caption{\label{fig:degen-toy}Contour plots for the three parametric mean functions of sufficient statistics for a RBM with one visible and one hidden node.}\label{fig:degen-toy}
 \end{figure}
+
 Secondly, we can look at \(\text{LREP}(\boldsymbol \theta)\) from
-\eqref{eq:elpr} for this tiny model in order to consider model instability
-as a function of RBM parameters. Recall that large values of
+\eqref{eq:elpr} for this tiny model in order to consider model
+instability as a function of RBM parameters. Recall that large values of
 \(\text{LREP}(\boldsymbol \theta)\) are associated with an extreme
@@
 \end{align*}
+
 Figure \ref{fig:instab} shows contour plots of
@@
 
-\caption{$\text{LREP}(\boldsymbol \theta)/\nv$ for various values of $\boldsymbol \theta$ for the tiny example model.
+\caption{$\label{fig:instab}\text{LREP}(\boldsymbol \theta)/\nv$ for various values of $\boldsymbol \theta$ for the tiny example model.
 (Recall here $\nv$ is the number of visible nodes and here is $1$.) This quantity is large for large magnitudes of $\boldsymbol \theta$.}\label{fig:instab}
 \end{figure}
+
 Finally to consider the effect of \(\boldsymbol \theta\) on potential
@@
 \end{align*}
+
 Again, we can inspect these coordinate functions of this vector
@@
 
-\caption{The absolute difference between coordinates of model expectations, E$\left[\boldsymbol X | \boldsymbol \theta\right]$, and expectations given independence, E$\left[\boldsymbol X | \boldsymbol \theta^* \right ]$ for a RBM with one visible and one hidden node.
+\caption{\label{fig:uninterp}The absolute difference between coordinates of model expectations, E$\left[\boldsymbol X | \boldsymbol \theta\right]$, and expectations given independence, E$\left[\boldsymbol X | \boldsymbol \theta^* \right ]$ for a RBM with one visible and one hidden node.
 As an indicator of uninterpretability, note that differences in expectations increase as the dependence parameter $\theta_{11}$ deviates from zero.}\label{fig:uninterp}
 \end{figure}
+
 Figure \ref{fig:uninterp} shows that the absolute difference between
@@
 \par
+
 \begin{figure}
-\includegraphics{paper_files/figure-latex/degen-plots-1} \caption{Results from the numerical experiment, here looking at the fraction of models that were near-degenerate for each combination of magnitudes of the points $\boldsymbol \theta$ and model size, where $\boldsymbol \theta = (\boldsymbol \theta_{main}, \theta_{interaction})$ is split into main and interaction parameters.
-Black lines show the contour levels for fraction of near-degeneracy, while the thick black line shows the level where the fraction of near-degenerate models is .05.}\label{fig:degen-plots}
+\centering
+\includegraphics{paper_files/figure-latex/degen-plots-1.pdf}
+\caption{\label{fig:degen-plots}Results from the numerical experiment,
+here looking at the fraction of models that were near-degenerate for
+each combination of magnitudes of the points \(\boldsymbol \theta\) and
+model size, where
+\(\boldsymbol \theta = (\boldsymbol \theta_{main}, \theta_{interaction})\)
+is split into main and interaction parameters. Black lines show the
+contour levels for fraction of near-degeneracy, while the thick black
+line shows the level where the fraction of near-degenerate models is
+.05.}
 \end{figure}
+
 \begin{figure}
-\includegraphics{paper_files/figure-latex/instab-plots-1} \caption{The sample mean value of $\text{LREP}(\boldsymbol \theta)/V$ at each grid point for each combination of magnitudes of the points of $\boldsymbol \theta$ and model size.
-As the magnitude of $\boldsymbol \theta$ grows, so does the value of this metric, indicating typical instability in the model.}\label{fig:instab-plots}
+\centering
+\includegraphics{paper_files/figure-latex/instab-plots-1.pdf}
+\caption{\label{fig:instab-plots}The sample mean value of
+\(\text{LREP}(\boldsymbol \theta)/\nv\) at each grid point for each
+combination of magnitudes of the points of \(\boldsymbol \theta\) and
+model size. As the magnitude of \(\boldsymbol \theta\) grows, so does
+the value of this metric, indicating typical instability in the model.}
 \end{figure}
+
 \begin{figure}
-\includegraphics{paper_files/figure-latex/uninterp-plots-1} \caption{The sample mean of the maximum component of the absolute difference between the model expectation vector, E$\left[\boldsymbol X | \boldsymbol \theta\right]$, and the expectation vector given independence, E$\left[\boldsymbol X | \boldsymbol \theta^* \right ]$.
-Larger magnitudes of $\boldsymbol \theta$ correspond to larger differences, thus indicating reduced interpretability.}\label{fig:uninterp-plots}
+\centering
+\includegraphics{paper_files/figure-latex/uninterp-plots-1.pdf}
+\caption{\label{fig:uninterp-plots}The sample mean of the maximum
+component of the absolute difference between the model expectation
+vector, E\(\left[\boldsymbol X | \boldsymbol \theta\right]\), and the
+expectation vector given independence,
+E\(\left[\boldsymbol X | \boldsymbol \theta^* \right ]\). Larger
+magnitudes of \(\boldsymbol \theta\) correspond to larger differences,
+thus indicating reduced interpretability.}
 \end{figure}
+
 Figure \ref{fig:split-plots-dist} shows the fraction of near-degenerate
@@
 \par
+
 \begin{figure}
-\includegraphics{paper_files/figure-latex/split-plots-dist-1} \caption{The fraction of near-degenerate models for each magnitude of $\boldsymbol \theta$.
-For each number $\nv$ of visibles in the model, the fraction near-degenerate moves away from zero at larger values of $||\boldsymbol \theta||$ as the number $\nh$ of hidden variables increases and the slope becomes steeper as $\nh$ increases as well.}\label{fig:split-plots-dist}
+\centering
+\includegraphics{paper_files/figure-latex/split-plots-dist-1.pdf}
+\caption{\label{fig:split-plots-dist}The fraction of near-degenerate
+models for each magnitude of \(\boldsymbol \theta\). For each number
+\(\nv\) of visibles in the model, the fraction near-degenerate moves
+away from zero at larger values of \(||\boldsymbol \theta||\) as the
+number \(\nh\) of hidden variables increases and the slope becomes
+steeper as \(\nh\) increases as well.}
 \end{figure}
+
 These manageable examples indicate that RBMs are near-degenerate,
@@
 unfeasible due mainly to the intractability of the normalizing term
-\(\gamma(\boldsymbol \theta)\) in a model \eqref{eq:pmf} of any realistic
-size. Ad hoc methods are often recommended instead, which aim to avoid
-this problem by using stochastic ML approximations that employ a small
-number of MCMC draws (i.e., contrastive divergence, (Hinton
-\protect\hyperlink{ref-hinton2002training}{2002})).
+\(\gamma(\boldsymbol \theta)\) in a model \eqref{eq:pmf} of any
+realistic size. Ad hoc methods are often recommended instead, which aim
+to avoid this problem by using stochastic ML approximations that employ
+a small number of MCMC draws (i.e., contrastive divergence, (Hinton
+2002)).
 
@@
 probabilities for the binary data outcomes). For example, Montufar and
-Ay (\protect\hyperlink{ref-montufar2011refinements}{2011}) show that any
-distribution on \(\{0, 1\}^{\nv}\) can be approximated arbitrarily well
-by a RBM with \(2^{\nv-1} - 1\) hidden units. We provide a small example
-in the on-line supplementary materials that illustrates such
-approximations.
+Ay (2011) show that any distribution on \(\{0, 1\}^{\nv}\) can be
+approximated arbitrarily well by a RBM with \(2^{\nv-1} - 1\) hidden
+units. We provide a small example in the on-line supplementary materials
+that illustrates such approximations.
 
@@
 possible to approximate any distribution on the visibles arbitrarily
-well (cf. Montufar and Ay
-\protect\hyperlink{ref-montufar2011refinements}{2011}), but dramatically
-different parameter settings can induce the same RBM model (beyond mere
-symmetries in parameterization). A further (and somewhat odd)
-consequence of the RBM parameterization is then that when fitting the
-RBM model by likelihood-based methods, we may already know the nature of
-the answer before we begin: namely, such fitting will simply reproduce
-the empirical distribution of the training data if sufficiently many
-hiddens are in the model. There then can be no model refinements or
-smoothing from the RBM. That is, based on a random sample of vectors of
-visible variables, the model for the cell probabilities with the highest
+well (cf. Montufar and Ay 2011), but dramatically different parameter
+settings can induce the same RBM model (beyond mere symmetries in
+parameterization). A further (and somewhat odd) consequence of the RBM
+parameterization is then that when fitting the RBM model by
+likelihood-based methods, we may already know the nature of the answer
+before we begin: namely, such fitting will simply reproduce the
+empirical distribution of the training data if sufficiently many hiddens
+are in the model. There then can be no model refinements or smoothing
+from the RBM. That is, based on a random sample of vectors of visible
+variables, the model for the cell probabilities with the highest
 likelihood over \emph{all possible model classes} (i.e., RBM-based or
@@
 \(\boldsymbol \theta_{main}\). This is similar to an idea advocated by
-Hinton (\protect\hyperlink{ref-hinton2010practical}{2010}) called
-\emph{weight decay}, in which a penalty is added to the interaction
-terms in the model, \(\boldsymbol \theta_{interaction}\), shrinking
-their magnitudes.
+Hinton (2010) called \emph{weight decay}, in which a penalty is added to
+the interaction terms in the model,
+\(\boldsymbol \theta_{interaction}\), shrinking their magnitudes.
 
 \par
+
 \begin{table}[ht]
@@
 \end{table}
+
 We considered a test case with \(\nv = \nh = 4\) and parameters given in
@@
 \(\boldsymbol \theta\).
+
 \begin{enumerate}
@@
   and are sampled in each MCMC iterative draw from the posterior
-  distribution. This is the method of Li
-  (\protect\hyperlink{ref-li2014biclustering}{2014}). We will refer to
-  this method as Bayes with Trick Prior and Latent Variables (BwTPLV).
+  distribution. This is the method of Li (2014). We will refer to this
+  method as Bayes with Trick Prior and Latent Variables (BwTPLV).
 \item
@@
   accomplished using a geometric adaptive Metropolis Hastings step (Zhou
-  \protect\hyperlink{ref-zhou2014some}{2014}) and calculation of
-  likelihood normalizing constant. (This computation is barely feasible
-  for a problem of this size and would be unfeasible for larger
-  problems.) Here the hidden variables \(h_j\) are again carried along
-  in the MCMC implementation as latent variables. We will refer to this
-  method as Bayes with Truncated Normal prior and Latent Variables
-  (BwTNLV).
+  2014) and calculation of likelihood normalizing constant. (This
+  computation is barely feasible for a problem of this size and would be
+  unfeasible for larger problems.) Here the hidden variables \(h_j\) are
+  again carried along in the MCMC implementation as latent variables. We
+  will refer to this method as Bayes with Truncated Normal prior and
+  Latent Variables (BwTNLV).
 \item
@@
 \end{enumerate}
+
 The three fitting methods are ordered above according to computational
@@
 \ref{tab:hyperparam}.
+
 \begin{table}[ht]
@@
 \end{table}
+
 \par
@@
 \(\mathbf{v}=(v_1,v_2,v_3,v_4)\in\{\pm 1\}^4\) (i.e., computed from
-\(\boldsymbol \theta\) under \eqref{eq:pmf} are determined and plotted in
-Figure \ref{fig:acf} with BwTNLV in black and BwTNML in red. As
+\(\boldsymbol \theta\) under \eqref{eq:pmf} are determined and plotted
+in Figure \ref{fig:acf} with BwTNLV in black and BwTNML in red. As
 expected, ACF corresponding to the method (BwTNML) that marginalizes out
@@
 \end{figure}
+
 Secondly, we can assess the mixing of the BwTNLV/BwTNML chains using the
@@
 is some positive constant such that \(C_i > \sigma^2_i\). We can use an
-overlapping block-means approach (Gelman, Shirley, and others
-\protect\hyperlink{ref-gelman2011inference}{2011}) to get a crude
-estimate for \(C_i\) as \(\hat{C}_i = bS_b^2\), where \(S_b^2\) denotes
-the sample variance of overlapping block means
+overlapping block-means approach (Gelman, Shirley, and others 2011) to
+get a crude estimate for \(C_i\) as \(\hat{C}_i = bS_b^2\), where
+\(S_b^2\) denotes the sample variance of overlapping block means
 \(\{\bar{p}_j^{(i)}=\sum_{k=j}^{j+b-1} p_k^{(i)}/b\}_{j=1}^{M-b+1}\) of
@@
 \]
+
 \begin{table}[ht]
@@
 \end{table}
+
 For both BwTNLV and BwTNML methods, effective sample sizes for a chain
@@
 \end{figure}
+
 An alternative possibility to using MCMC to approximate the posterior
 for Bayesian inference in a RBM model is to use variational methods for
-approximation of the posterior (Salimans, Kingma, and Welling
-\protect\hyperlink{ref-salimans2015markov}{2015}). The advantage to
-using variational methods would be computational scalability, but the
-disadvantages are potentially numerous. Namely, the lack of theoretical
-guarantees of convergence to a global optimum (Blei, Kucukelbir, and
-McAuliffe \protect\hyperlink{ref-blei2017variational}{2017}) and the
-difficulty of assessing how closely approximated the posterior is (Yao
-et al. \protect\hyperlink{ref-pmlr-v80-yao18a}{2018}) are of concern.
+approximation of the posterior (Salimans, Kingma, and Welling 2015). The
+advantage to using variational methods would be computational
+scalability, but the disadvantages are potentially numerous. Namely, the
+lack of theoretical guarantees of convergence to a global optimum (Blei,
+Kucukelbir, and McAuliffe 2017) and the difficulty of assessing how
+closely approximated the posterior is (Yao et al. 2018) are of concern.
 These two issues show potential difficulties in using variational Bayes
@@
 
+%\bibliography{../resources/latex/refs\_rbm.bib}
+
 \end{document}

### 2018-09-18 writing/paper-rbm/paper.Rmd
ready for submission

@@
   cache = TRUE,
-  echo = FALSE
+  echo = FALSE,
+  dev = 'postscript'
 )
 
-ggplot2::theme_set(ggplot2::theme_bw(base_family="serif"))
+ggplot2::theme_set(ggplot2::theme_bw())
 

### 2018-09-18 writing/paper-rbm/letters/cover_letter.Rmd
ready for submission

@@
 date: "`r format(Sys.time(), '%B %d, %Y')`"
-opening: Dear Drs. Liu and Zhao,
+opening: Dear Dr. Clarke,
 closing: Best regards,
 address: 
- - Regina Liu and Hongyu Zhao
- - Co-Editors-Elect
- - "*Journal of the American Statistical Association - Theory & Methods*"
+ - Dr. Bertrand Clarke
+ - Editor
+ - "Statistical Analysis and Data Mining"
 return-address:
@@
 
-I am writing to submit an article entitled "Properties and Bayesian fitting of restricted Boltzmann machines" for consideration for publication in the Journal of the American Statistical Association - Theory & Methods.
+My name is Andee Kaplan, and I am a postdoctoral fellow at Duke University in the Department of Statistical Science.
+I am writing on behalf of my co-authors, Dr.
+Daniel Nordman and Dr.
+Stephen Vardeman, to submit our article entitled "Properties and Bayesian fitting of restricted Boltzmann machines" for consideration for publication in Statistical Analysis and Data Mining: the ASA Data Mining Journal.
+
 In this submission, we discuss the relationship between parameter specification and unattractive model properties---degeneracy, instability and uninterpretability---in restricted Boltzmann machines and describe related difficulties that can arise in likelihood-based and Bayes fitting of such models.
+Our findings are that these models can experience certain types of model instability problems over the parameter space, which relates to how useful the model may be for representing realistic data-generation.
+In addition to the concepts and theories of instability, we also present empirical evidence that if parameter values stray too far from the origin of the parameter space, then these model problems can arise.
+We also investigate a principled Bayes fitting method with the aim to a) potentially quantify the uncertainty in estimation and b) potentially avoid problem areas of the parameter space by constraining parameters appropriately.
+Our methods show promise, but are numerically intractable in large data cases.
+In our estimation, it is unclear that current practice with these models is addressing our presented concerns.
 
-This paper references results from another manuscript, an invited revision of which is presently under review by Technometrics.
-We have provided a blinded copy of this second manuscript as a supplement for referees.
+In light of this, we do not expect this message to be well-received by proponents of the current methodology, and respectfully request that the review of the paper be carried out with this in mind and involve reviewers that have no personal stake in the existing practice.
 
-Thank you for your consideration.
+This paper references results from another manuscript, a paper which is presently under review by Information and Inference: A Journal of the IMA.
+We have provided a copy of this second manuscript as a supplement for referees.
+
+Thank you for your time and consideration, and do not hesitate to contact me if you have any questions regarding our paper.

### 2018-09-18 writing/paper-rbm/paper.tex
ready for submission

@@
 \begin{figure}
-\includegraphics[width=1\linewidth]{paper_files/figure-latex/sample-models-1} \includegraphics[width=1\linewidth]{paper_files/figure-latex/sample-models-2} \caption{\label{fig:sample-models}Ten 4-pixel images simulated from a degenerate model (a) compared to ten 4-pixel images simulated from a non-degenerate model (b).
+\includegraphics[width=1\linewidth]{sample-models-1} \includegraphics[width=1\linewidth]{sample-models-2} \caption{\label{fig:sample-models}Ten 4-pixel images simulated from a degenerate model (a) compared to ten 4-pixel images simulated from a non-degenerate model (b).
 The degenerate model places almost all probability on one outcome, causing the image to be generated repeatedly, whereas the non-degenerate model shows more realistic variation.}\label{fig:sample-models}
@@
   \centering
-  \resizebox{\linewidth}{!}{\input{../resources/images/tikz/rbm.tikz}}
+  \resizebox{\linewidth}{!}{\input{rbm.tikz}}
   \caption{An example restricted Boltzmann machine (RBM), consisting of two layers, one hidden ($\mathcal{H}$) and one visible ($\mathcal{V}$), with no connections within a layer.
@@
   \centering
-  \resizebox{1cm}{!}{\input{../resources/images/tikz/toymodel.tikz}}
+  \resizebox{1cm}{!}{\input{toymodel.tikz}}
   \caption{A small example restricted Boltzmann machine (RBM), with two nodes, one hidden and one visible.}
@@
     \tdplotsetmaincoords{60}{-60}
-    \input{../resources/images/tikz/toyhull_top.tikz}
+    \input{toyhull_top.tikz}
 
@@
     \tdplotsetmaincoords{60}{120}
-    \input{../resources/images/tikz/toyhull_top.tikz}
+    \input{toyhull_top.tikz}
 
@@
 
-{\centering \includegraphics{paper_files/figure-latex/degen-toy-1}
+{\centering \includegraphics{degen-toy-1}
 
@@
 
-{\centering \includegraphics{paper_files/figure-latex/instab-1}
+{\centering \includegraphics{instab-1}
 
@@
 
-{\centering \includegraphics{paper_files/figure-latex/uninterp-1}
+{\centering \includegraphics{uninterp-1}
 
@@
 \centering
-\includegraphics{paper_files/figure-latex/degen-plots-1.pdf}
+\includegraphics{degen-plots-1.eps}
 \caption{\label{fig:degen-plots}Results from the numerical experiment,
@@
 \centering
-\includegraphics{paper_files/figure-latex/instab-plots-1.pdf}
+\includegraphics{instab-plots-1.eps}
 \caption{\label{fig:instab-plots}The sample mean value of
@@
 \centering
-\includegraphics{paper_files/figure-latex/uninterp-plots-1.pdf}
+\includegraphics{uninterp-plots-1.eps}
 \caption{\label{fig:uninterp-plots}The sample mean of the maximum
@@
 \centering
-\includegraphics{paper_files/figure-latex/split-plots-dist-1.pdf}
+\includegraphics{split-plots-dist-1.eps}
 \caption{\label{fig:split-plots-dist}The fraction of near-degenerate
@@
 \centering
-\includegraphics{paper_files/figure-latex/acf-1.pdf}
+\includegraphics{acf-1.eps}
 \caption{\label{fig:acf}The autocorrelation functions (ACF) for the
@@
 \centering
-\includegraphics{paper_files/figure-latex/fitting-plot-1.pdf}
+\includegraphics{fitting-plot-1.eps}
 \caption{\label{fig:fitting-plot}Posterior probabilities of \(16 = 2^4\)

### 2018-09-27 writing/paper-rbm/paper.Rmd
grammar check

@@
 By treating a hidden layer of one RBM as the visible layer in a second RBM, a deep architecture can be created.
-RBMs are thought to thereby have the ability to encode very complex and rich structures in data, making them attractive for supervised learning.
-However, the generative behavior of RBMs is largely unexplored and typical fitting methodology does not easily allow for uncertainty quantification in addition to point estimates.
+RBMs thereby are thought to have the ability to encode very complex and rich structures in data, making them attractive for supervised learning.
+However, the generative behavior of RBMs largely is unexplored and typical fitting methodology does not easily allow for uncertainty quantification in addition to point estimates.
 In this paper, we discuss the relationship between RBM parameter specification in the binary case and model properties such as degeneracy, instability and uninterpretability.
-We also describe the associated difficulties that can arise with likelihood-based inference and further discuss the potential Bayes fitting of such (highly flexible) models, especially as Gibbs sampling (quasi-Bayes) methods are often advocated for the RBM model structure.
+We also describe the associated difficulties that can arise with likelihood-based inference and further discuss the potential Bayes fitting of such (highly flexible) models, especially as Gibbs sampling (quasi-Bayes) methods often are advocated for the RBM model structure.
 keywords: Degeneracy, Instability, Classification, Deep Learning, Graphical Models
@@
 
-The data mining and machine learning communities have recently shown great interest in deep learning, specifically in stacked restricted Boltzmann machines (RBMs) [see @salakhutdinov2009deep; @salakhutdinov2012efficient; @srivastava2013modeling; @le2008representational for examples].
-A RBM is a probabilistic undirected graphical model (for discrete or continuous random variables) with two layers, one hidden and one visible, with no conditional dependency within a layer [@smolensky1986information].
-These models have reportedly been used with success in classification of images [@larochelle2008classification; @srivastava2012multimodal].
+The data mining and machine learning communities recently have shown great interest in deep learning, specifically in stacked restricted Boltzmann machines (RBMs) [see @salakhutdinov2009deep; @salakhutdinov2012efficient; @srivastava2013modeling; @le2008representational for examples].
+An RBM is a probabilistic undirected graphical model (for discrete or continuous random variables) with two layers, one hidden and one visible, with no conditional dependency within a layer [@smolensky1986information].
+These models reportedly have been used with success in classification of images [@larochelle2008classification; @srivastava2012multimodal].
 However, the model properties are largely unexplored in the literature and the commonly cited fitting methodology remains heuristic-based and relies on rough approximation [@hinton2006fast].
@@
 Neyman and Box further state that a model should "provide an explanation of the mechanism underlying the observed phenomena" [@lehmann1990model; @box1967discrimination].
-At issue, simulation from RBMs can often produce data lacking realistic variability so that such models may thereby fail to satisfactorily reflect an observed data generation process.
-Such behavior relates to model degeneracy (or near degeneracy), which is a statistical concern in that many data processes of interest are realistically not degenerate in their spectrum of potential outcomes.
+At issue, simulation from RBMs often can produce data lacking realistic variability so that such models may thereby fail to satisfactorily reflect an observed data generation process.
+Such behavior relates to model degeneracy (or near degeneracy), which is a statistical concern in that, realistically, many data processes of interest are not degenerate in their spectrum of potential outcomes.
 For example, when sampling data from a nearly degenerate RBM used to model an imaging process, only a small set of output possibilities receives nearly all probability, and thus a sample of images will all be copies of the same one image (or small number of images).
@@
 
-In addition to such degeneracy, we find that RBMs can easily exhibit types of instability, related to how sensitive the model probabilities can be to small difference in data outcomes.
+In addition to such degeneracy, we find that RBMs easily can exhibit types of instability, related to how sensitive the model probabilities can be to small difference in data outcomes.
 In practice, this may be seen when a single pixel change in an image results in a wildly different classification in an image classification problem.
-Occurrences of such behavior have recently been documented in RBMs [@li2014biclustering], as well as other deep architectures [@szegedy2013intriguing; @nguyen2014deep].
+Occurrences of such behavior recently have been documented in RBMs [@li2014biclustering], as well as other deep architectures [@szegedy2013intriguing; @nguyen2014deep].
 We describe (related) potential issues of model instability, degeneracy and uninterpretability for the RBM class (which are related properties) and examine the presence of these in Section \ref{explorations-of-model-properties-through-simulation} through simulations of small, manageable examples.
@@
 
-Furthermore, these MCMC algorithms involve updating potentially many latent variables (hiddens) which can critically influence convergence in MCMC-based likelihood methods.
-Applying basic statistical principles in fitting RBM models of tractable size, we compare three fully Bayesian techniques involving MCMC, which are computationally more accessible than direct maximum likelihood and also aim to avoid parts of a RBM parameter space that yield unattractive models.
+Furthermore, these MCMC algorithms involve updating potentially many latent variables (hiddens), which can critically influence convergence in MCMC-based likelihood methods.
+Applying basic statistical principles in fitting RBM models of tractable size, we compare three fully Bayesian techniques involving MCMC, which are computationally more accessible than direct maximum likelihood and also aim to avoid parts of an RBM parameter space that yield unattractive models.
 As might be expected, with greater computational complexity comes an increase in fitting accuracy, but at the cost of practical feasibility.
 
-As a factor compounding the computational challenges, issues in model fitting traceable to model flexibility are potentially more concerning.
-For a RBM model with enough hidden variables, any distribution for the visibles can be approximated arbitrarily well [@le2008representational; @montufar2011refinements; and @montufar2011expressive].
-However, for binary data, the empirical distribution of an observed training set of visible variables provides a highest likelihood benchmark, before any parametric model class is even introduced and applied to obtain a refinement of model fit.
-As a consequence, we find that any fully principled fitting method based on the likelihood for a RBM with enough hidden variables will seek to reproduce the (discrete) empirical distribution of a training set.
-This aspect can be undesirable, and perhaps even unexpected compared to most modeling scenarios, in that no "smoothed distribution" may result when fitting a RBM model of sufficient size with a rigorous likelihood-based method.
-We are therefore led to be skeptical that models that involve these structures (like deep Boltzmann machines) can achieve useful prediction or inference in a principled way without intentionally limiting the flexibility of the fitted model.
+As a factor compounding the computational challenges, issues in model fitting traceable to model flexibility potentially are more concerning.
+For an RBM model with enough hidden variables, any distribution for the visibles can be approximated arbitrarily well [@le2008representational; @montufar2011refinements; and @montufar2011expressive].
+However, for binary data, the empirical distribution of an observed training set of visible variables provides a highest likelihood benchmark, before any parametric model class even is introduced and applied to obtain a refinement of model fit.
+As a consequence, we find that any fully principled fitting method based on the likelihood for an RBM with enough hidden variables will seek to reproduce the (discrete) empirical distribution of a training set.
+This aspect can be undesirable, and perhaps even unexpected compared to most modeling scenarios, in that no "smoothed distribution" may result when fitting an RBM model of sufficient size with a rigorous likelihood-based method.
+We therefore are led to be skeptical that models that involve these structures (like deep Boltzmann machines) can achieve useful prediction or inference in a principled way without intentionally limiting the flexibility of the fitted model.
 
@@
 While applications of the RBM have been claimed to produce some successes in classification problems, it is unclear if the model class allows one to go beyond fitting to other statistical matters of importance in using the models, such as quantification of the uncertainty of estimation and the formulation of predictive distributions.
-These are closely tied to the probability properties of the RBM model for explaining data generation and our exposition intends to contribute to a better understanding in this direction.
+These are tied closely to the probability properties of the RBM model for explaining data generation, and our exposition intends to contribute to a better understanding in this direction.
 
@@
 
-A restricted Boltzmann machine (RBM) is an undirected graphical model specified for discrete or continuous random variables, binary variables being most commonly considered.
+A restricted Boltzmann machine (RBM) is an undirected graphical model specified for discrete or continuous random variables, binary variables most commonly being considered.
 In this paper, we consider the binary case for concreteness.
-A RBM architecture has two layers, hidden ($\mathcal{H}$) and visible ($\mathcal{V}$), with no dependency connections within a layer.
+An RBM architecture has two layers, hidden ($\mathcal{H}$) and visible ($\mathcal{V}$), with no dependency connections within a layer.
 An example of this structure is in Figure \ref{fig:rbm} with the hidden nodes indicated by gray circles and the visible nodes indicated by white circles.
@@
 A common use for RBMs is to create features for use in classification.
-For example, binary images can be classified through a process that treats the pixel values as the visible variables $v_i$ in a RBM model [@hinton2006fast].
+For example, binary images can be classified through a process that treats the pixel values as the visible variables $v_i$ in an RBM model [@hinton2006fast].
 
@@
 
-Let $\boldsymbol x = (h_1, \dots, h_\nh, v_1,\dots,v_\nv)$ represent the states of the visible and hidden nodes in a RBM for some integers $\nv, \nh \ge 1$.
+Let $\boldsymbol x = (h_1, \dots, h_\nh, v_1,\dots,v_\nv)$ represent the states of the visible and hidden nodes in an RBM for some integers $\nv, \nh \ge 1$.
 Each single binary random variable, visible or hidden, will take its values in a common coding set $\mathcal{C}$, where we allow (one of) two possibilities for the coding set, $\mathcal{C}=\{0,1\}$ or $\mathcal{C}=\{-1,1\}$, with "$1$" always indicating the "high" value of the variable.
@@
 The RBM model is parameterized by $\boldsymbol \theta$ containing two types of parameters, main effects and interaction effects.
-The main effects parameters ($\{\theta_{v_i}, \theta_{h_j}\}_{\substack{i = 1, \dots, \nv,\\j = 1, \dots, \nh}}$) weight the values of the visible $v_i$ and hidden $h_j$ nodes in probabilities \eqref{eq:pmf} and the interaction effect parameters ($\theta_{ij}$) weight the values of the connections $v_i h_j$, or dependencies, between hidden and visible layers.
+The main effects parameters ($\{\theta_{v_i}, \theta_{h_j}\}_{\substack{i = 1, \dots, \nv,\\j = 1, \dots, \nh}}$) weight the values of the visible $v_i$ and hidden $h_j$ nodes in probabilities \eqref{eq:pmf}, and the interaction effect parameters ($\theta_{ij}$) weight the values of the connections $v_i h_j$, or dependencies, between hidden and visible layers.
 
@@
 ## Degeneracy, instability, and uninterpretability
-The highly flexible nature of a RBM (having as it does $\nh + \nv + \nh*\nv$ parameters) creates at least three kinds of potential issues of model impropriety that we will call *degeneracy*, *instability*, and *uninterpretability*.
-In this section we define these, consider how to quantify them in a RBM, and point out relationships among them.
+The highly flexible nature of an RBM (having as it does $\nh + \nv + \nh*\nv$ parameters) creates at least three kinds of potential issues of model impropriety that we will call *degeneracy*, *instability*, and *uninterpretability*.
+In this section we define these, consider how to quantify them in an RBM, and point out relationships among them.
 
@@
 In Random Graph Model theory, *model degeneracy* means there is a disproportionate amount of probability placed on only a few elements of the sample space, $\mathcal{X}$, by the model [@handcock2003assessing].
-For random graph models, $\mathcal{X}$ denotes all possible graphs that can be constructed from a set of nodes and an exponentially parameterized random graph model has a distribution with a probability mass function of the form
+For random graph models, $\mathcal{X}$ denotes all possible graphs that can be constructed from a set of nodes, and an exponentially parameterized random graph model has a distribution with a probability mass function of the form
 \begin{align*}
@@
 @handcock2003assessing classifies an exponentially parameterized random graph model at $\boldsymbol \theta$ as *near-degenerate* if the mean value of the vector of sufficient statistics under $\boldsymbol \theta$, $\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta}\boldsymbol t( \boldsymbol X)$, is close to the boundary of $C$.
-Intuitively, if a model is near-degenerate in the sense that only a small number of elements of the sample space $\mathcal{X}$ have positive probability, the expected value $\text{E}_{\boldsymbol \theta}\boldsymbol t( \boldsymbol X)$ is an average of that same small number of values of $t( \boldsymbol x)$ (defining the boundary of the hull $C$) and can be expected to *not* be pulled deep into the interior of $C$.
-
-A RBM model can be thought to exhibit an analogous form of *near-degeneracy* when there is a disproportionate amount of probability placed on a small number of elements in the sample space of visibles and hiddens, $\mathcal{C}^{\nv + \nh}$.
+Intuitively, if a model is near-degenerate in the sense that only a small number of elements of the sample space $\mathcal{X}$ have positive probability, the expected value $\text{E}_{\boldsymbol \theta}\boldsymbol t( \boldsymbol X)$ is an average of that same small number of values of $t( \boldsymbol x)$ (defining the boundary of the hull $C$) and can be expected to *not* be pulled deeply into the interior of $C$.
+
+An RBM model can be thought to exhibit an analogous form of *near-degeneracy* when there is a disproportionate amount of probability placed on a small number of elements in the sample space of visibles and hiddens, $\mathcal{C}^{\nv + \nh}$.
 Using the idea of @handcock2003assessing, when the random vector $\boldsymbol t(\boldsymbol x) = (v_1, \dots, v_\nv, h_1, \dots, h_\nh, v_1 h_1, \dots, v_V h_\nh ) \in \mathcal{T} \equiv \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{C}^{\nh + \nv} \}$ from \eqref{eq:t}, appearing in the neg-potential function $Q_{\boldsymbol \theta}(\cdot)$, has a mean vector $\boldsymbol \mu(\boldsymbol \theta) \in \mathbb{R}^{\nv + \nh + \nv * \nh}$ close to the boundary of the convex hull of $\mathcal{T}$, and the RBM model can be said to exhibit near-degeneracy at $\boldsymbol \theta \in \mathbb{R}^{\nv + \nh + \nh*\nv}$.
@@
 Considering exponential families of distributions, @schweinberger2011instability introduced a concept of model deficiency related to *instability*.
-Instability can be roughly thought of as excessive sensitivity in the model, where small changes in the components of potential data outcomes, $\boldsymbol x$, may lead to substantial changes in the probability mass function $f_{\boldsymbol \theta}(\boldsymbol x)$.
-Furthermore, model instability can be viewed on a spectrum of potential sensitivity in probability structure, with model degeneracy included as a extreme or limiting case of instability.
+Instability roughly can be thought of as excessive sensitivity in the model, where small changes in the components of potential data outcomes, $\boldsymbol x$, may lead to substantial changes in the probability mass function $f_{\boldsymbol \theta}(\boldsymbol x)$.
+Furthermore, model instability can be viewed on a spectrum of potential sensitivity in probability structure, with model degeneracy included as an extreme or limiting case of instability.
 To quantify "instability" more rigorously (particularly beyond the definition given by @schweinberger2011instability) it is useful to consider how RBM models might be expanded to incorporate more and more visibles.
-When increasing the size of RBM models, it becomes necessary to grow the number of model parameters (and in this process one may also arbitrarily expand the number of hidden variables used).
+When increasing the size of RBM models, it becomes necessary to grow the number of model parameters (and in this process one also may expand arbitrarily the number of hidden variables used).
 To this end, let $\boldsymbol \theta_{\nv} \equiv (\theta_{v_1}, \dots, \theta_{v_\nv}, \theta_{h_1}, \dots, \theta_{h_\nh}, \theta_{11}, \dots, \theta_{\nv \nh}), \nv \ge 1$, denote an element of a sequence of RBM parameters indexed by the number $\nv$ of visibles $(V_1, \dots, V_\nv)$ and define a log-ratio of extremal
@@
 
-In formulating a RBM model for a potentially large number of visibles (i.e., as $\nv \rightarrow \infty$), we will say that the ratio \eqref{eq:elpr} needs to stay bounded for a sequence of RBM models to be stable.
+In formulating an RBM model for a potentially large number of visibles (i.e., as $\nv \rightarrow \infty$), we will say that the ratio \eqref{eq:elpr} needs to stay bounded for a sequence of RBM models to be stable.
 That is, we make the following convention.
@@
 Let $\boldsymbol \theta_\nv \in \mathbb{R}^{\nv + \nh +  \nh*\nv}, \nv \ge 1$, be an element of a sequence of RBM parameters where the number of hiddens, $\nh \equiv \nh(\nv) \ge 1$, can be an arbitrary function of the number $\nv$ of visibles.
-A RBM model formulation is \emph{Schweinberger-unstable} or \emph{S-unstable} if
+An RBM model formulation is \emph{Schweinberger-unstable} or \emph{S-unstable} if
 \begin{align*}
@@
 
-In other words, a RBM model sequence is unstable if, given any $c > 0$, there exists an integer $n_c > 0$ so that $\frac{1}{\nv}\text{LREP}(\boldsymbol \theta_\nv) > c$ for all $\nv \ge n_c$.
-This definition of *S-unstable* is a generalization or re-interpretation of the "unstable" concept of @schweinberger2011instability in that here RBM models for visibles $(v_1, \dots, v_\nv)$ do not form an exponential family and the dimensionality of $\boldsymbol \theta_\nv$ is not fixed, but rather grows with $\nv$.
+In other words, an RBM model sequence is unstable if, given any $c > 0$, there exists an integer $n_c > 0$ so that $\frac{1}{\nv}\text{LREP}(\boldsymbol \theta_\nv) > c$ for all $\nv \ge n_c$.
+This definition of *S-unstable* is a generalization or re-interpretation of the "unstable" concept of @schweinberger2011instability in that here RBM models for visibles $(v_1, \dots, v_\nv)$ do not form an exponential family, and the dimensionality of $\boldsymbol \theta_\nv$ is not fixed but rather grows with $\nv$.
 
@@
 $$
-denoting the biggest log-probability ratio for a one-component change in data outcomes (visibles) at a RBM parameter $\boldsymbol \theta_\nv$. We then have the following result.
+denoting the largest log-probability ratio for a one-component change in data outcomes (visibles) at an RBM parameter $\boldsymbol \theta_\nv$. We then have the following result.
 
@@
 
-In other words, if the probability ratio \eqref{eq:elpr} is too large, then a RBM model sequence will exhibit large probability shifts for very small changes in data configurations (i.e., will exhibit instability).
+In other words, if the probability ratio \eqref{eq:elpr} is too large, then an RBM model sequence will exhibit large probability shifts for very small changes in data configurations (i.e., will exhibit instability).
 Such instability can be a concern in a model for reasons similar to those related to degeneracy: as outcomes vary over the sample space, the geography of probabilities is extremely rugged, with deep pits following sharp mountains.
 Recall the applied example of RBM models as a means to classify images.
-For data as pixels in an image, the instability result in Proposition \ref{prp:instab1} manifests itself as a one pixel change in an image (one component of the visible vector) resulting in a large shift in the probability, which in turn could result in a vastly different classification of the image.
+For data as pixels in an image, the instability result in Proposition \ref{prp:instab1} manifests itself as a one pixel change in an image (one component of the visible vector), resulting in a large shift in the probability, which in turn could result in a vastly different classification of the image.
 Examples of this kind of behavior have been presented in @szegedy2013intriguing for deep learning models, in which a one pixel change in a test image results in a wildly different classification.
 
-Additionally, S-unstable RBM model sequences may be formally connected to the near-degeneracy of Section \ref{near-degeneracy} (in which model sequences place all probability on a small portion of their sample spaces).
+Additionally, S-unstable RBM model sequences may be connected formally to the near-degeneracy of Section \ref{near-degeneracy} (in which model sequences place all probability on a small portion of their sample spaces).
 To see this, define an arbitrary modal set of possible outcomes (i.e.
@@
 $$
-for a given $0 < \epsilon < 1$. Then S-unstable model sequences are guaranteed to be degenerate, as the following result shows.
+for a given $0 < \epsilon < 1$. Then, S-unstable model sequences are guaranteed to be degenerate, as the following result shows.
 
@@
 
-In other words, S-unstable RBM model sequences are guaranteed to stack up all probability on a specific set of outcomes for visibles, which could potentially be arbitrarily narrow.
+In other words, S-unstable RBM model sequences are guaranteed to stack up all probability on a specific set of outcomes for visibles, which potentially could be arbitrarily narrow.
 Proofs of Propositions \ref{prp:instab1} and \ref{prp:degen} follow from more general results in @kaplan2016propriety.
@@
 We adapt this notion to RBM models.
-Note that in a RBM, the parameters $\theta_{v_1}, \dots, \theta_{v_\nv}$ and $\theta_{h_1}, \dots, \theta_{h_\nh}$ are naturally associated with main effects of visible and hidden variables and can be interpreted as (logit functions of) means for variables $V_1, \dots, V_\nv, H_1, \dots, H_\nh$ in a model with no interaction parameters, $\theta_{ij} = 0, i = 1, \dots, \nv, j = 1, \dots, \nh$.
+Note that in an RBM, the parameters $\theta_{v_1}, \dots, \theta_{v_\nv}$ and $\theta_{h_1}, \dots, \theta_{h_\nh}$ naturally are associated with main effects of visible and hidden variables and can be interpreted as (logit functions of) means for variables $V_1, \dots, V_\nv, H_1, \dots, H_\nh$ in a model with no interaction parameters, $\theta_{ij} = 0, i = 1, \dots, \nv, j = 1, \dots, \nh$.
 That is, with no interaction parameters, we have from \eqref{eq:pmf} that
@@
 Hence, these main effect parameters have a clear interpretation under an independence model (one with $\theta_{ij} = 0$) but this interpretation can break down as interaction parameters increase in magnitude relative to the size of the main effects.
-In such cases, the main effect parameters $\theta_{v_1}$ and $\theta_{h_j}$ are no longer interpretable in the models (statements of marginal means) and the dependence parameters are so large as to dominate the entire model probability structure (also destroying simple interpretation of dependence parameters $\theta_{ij}$, $j=1,\ldots,\nh$ as local conditional modifications of an overall marginal mean structure $\theta_{v_i}$, as appearing for example in the (logit) conditional probability $\mathrm{logit P_\theta(V_i=1|H_1,\ldots,H_{\nh})}= \theta_{v_i} + \sum_{j=1}^{\nh} \theta_{ij} H_j$).
+In such cases, the main effect parameters $\theta_{v_1}$ and $\theta_{h_j}$ no longer are interpretable in the models (statements of marginal means) and the dependence parameters are so large as to dominate the entire model probability structure (also destroying simple interpretation of dependence parameters $\theta_{ij}$, $j=1,\ldots,\nh$ as local conditional modifications of an overall marginal mean structure $\theta_{v_i}$, as appearing for example in the (logit) conditional probability $\mathrm{logit P_\theta(V_i=1|H_1,\ldots,H_{\nh})}= \theta_{v_i} + \sum_{j=1}^{\nh} \theta_{ij} H_j$).
 Whether or not parameter interpretation is itself a goal in a given application of RBM models, this concept of interpretation can provide an additional device for examining other aspects of model propriety related to instability and degeneracy.
-As explained by @kaiser2007statistical, models with interpretable dependence parameters typically correspond to non-degenerate models, while degradation in interpretability is often associated with model drift into degeneracy.
+As explained by @kaiser2007statistical, models with interpretable dependence parameters typically correspond to non-degenerate models, while degradation in interpretability often is associated with model drift into degeneracy.
 To assess which parameter values $\boldsymbol \theta$ may cause difficulties in interpretation, we use the difference $\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] - \text{E}\left[\boldsymbol X | \boldsymbol \theta^* \right]$ between two model expectations: E$\left[\boldsymbol X | \boldsymbol \theta\right]$ at $\boldsymbol \theta$ and expectations E$\left[\boldsymbol X | \boldsymbol \theta^* \right]$ where $\boldsymbol \theta^*$ matches $\boldsymbol \theta$ for all main effects but otherwise has $\theta_{ij} = 0$ for $i = 1, \dots, \nv, j = 1, \dots, \nh$.
-(Hence, $\boldsymbol \theta^*$ and $\boldsymbol \theta$ have the same main effects but $\boldsymbol \theta^*$ has $0$ dependence parameters.) Uninterpretability is then avoided at a parametric specification $\boldsymbol \theta$ if the model expected value at $\boldsymbol \theta$ is not very different from the corresponding model expectation under independence.
+(Hence, $\boldsymbol \theta^*$ and $\boldsymbol \theta$ have the same main effects but $\boldsymbol \theta^*$ has $0$ dependence parameters.) Uninterpretability then is avoided at a parametric specification $\boldsymbol \theta$ if the model expected value at $\boldsymbol \theta$ is not very different from the corresponding model expectation under independence.
 Using this, it is possible to investigate what parametric conditions lead to uninterpretability in a model versus those that guarantee interpretable models.
@@
 
-We next explore and numerically explain the relationship between values of $\boldsymbol \theta$ and the three notions of model impropriety (near-degeneracy, instability, and uninterpretability), for RBM models of varying sizes.
+We next explore and numerically explain the relationship between values of $\boldsymbol \theta$ and the three notions of model impropriety (near-degeneracy, instability, and uninterpretability) for RBM models of varying sizes.
 
@@
 
-To illustrate the ideas of model near-degeneracy, instability, and uninterpretability in a RBM, we consider first the smallest possible (toy) example that consists of one visible node $v_1$ and one hidden node $h_1$ that are both binary.
+To illustrate the ideas of model near-degeneracy, instability, and uninterpretability in an RBM, we consider first the smallest possible (toy) example that consists of one visible node $v_1$ and one hidden node $h_1$ that are both binary.
 A schematic of this model can be found in Figure \ref{fig:toymodel}.
@@
 appearing in the RBM probabilities for this model.
-As this set is in three dimensions, we are able to explicitly illustrate the shape of boundary of the convex hull of $\mathcal{T}$ and explore the behavior of the mean vector $\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta} \boldsymbol t(\boldsymbol x)$ as a function of the parameter vector $\boldsymbol \theta$.
+As this set is in three dimensions, we are able to illustrate explicitly the shape of boundary of the convex hull of $\mathcal{T}$ and explore the behavior of the mean vector $\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta} \boldsymbol t(\boldsymbol x)$ as a function of the parameter vector $\boldsymbol \theta$.
 Figure \ref{fig:toyhull} shows the convex hull of our "statistic space," $\mathcal{T} \subset \{0,1\}^3$, for this toy problem from two perspectives (enclosed by the unit cube $[0,1]^3$, the convex hull of $\{0,1\}^3$).
@@
 
-```{r degen-toy, fig.cap="\\label{fig:degen-toy}Contour plots for the three parametric mean functions of sufficient statistics for a RBM with one visible and one hidden node.", fig.height = 4, fig.align='center', message=FALSE, cache=TRUE}
+```{r degen-toy, fig.cap="\\label{fig:degen-toy}Contour plots for the three parametric mean functions of sufficient statistics for an RBM with one visible and one hidden node.", fig.height = 4, fig.align='center', message=FALSE, cache=TRUE}
 possibles <- stats(1, 1, type = "binary")
@@
 
-Finally to consider the effect of $\boldsymbol \theta$ on potential model uninterpretability, we can look at the difference between model expectations, E$\left[\boldsymbol X | \boldsymbol \theta\right]$, and expectations given independence, E$\left[\boldsymbol X | \boldsymbol \theta^*\right ]$ for the tiny toy RBM model where $\boldsymbol X = (V_1, H_1, V_1 H_1)$.
+Finally, to consider the effect of $\boldsymbol \theta$ on potential model uninterpretability, we can look at the difference between E$\left[\boldsymbol X | \boldsymbol \theta\right]$ model expectations and E$\left[\boldsymbol X | \boldsymbol \theta^*\right ]$ expectations given independence for the tiny toy RBM model where $\boldsymbol X = (V_1, H_1, V_1 H_1)$.
 This difference is given by
@@
 
-```{r uninterp, fig.cap="\\label{fig:uninterp}The absolute difference between coordinates of model expectations, E$\\left[\\boldsymbol X | \\boldsymbol \\theta\\right]$, and expectations given independence, E$\\left[\\boldsymbol X | \\boldsymbol \\theta^* \\right ]$ for a RBM with one visible and one hidden node.
-As an indicator of uninterpretability, note that differences in expectations increase as the dependence parameter $\\theta_{11}$ deviates from zero.", fig.height = 4, fig.align='center', message=FALSE, cache=TRUE, warning=FALSE}
+```{r uninterp, fig.cap="\\label{fig:uninterp}The absolute difference between coordinates of E$\\left[\\boldsymbol X | \\boldsymbol \\theta\\right]$ model expectations and E$\\left[\\boldsymbol X | \\boldsymbol \\theta^* \\right ]$ expectations given independence for an RBM with one visible and one hidden node.
+As an indicator of uninterpretability, note that differences in expectations increase as$\\theta_{11}$, the dependence parameter, deviates from zero.", fig.height = 4, fig.align='center', message=FALSE, cache=TRUE, warning=FALSE}
 expand.grid(theta_h1 = seq(-5, 5, length.out = 21),
@@
 
-Figure \ref{fig:uninterp} shows that the absolute difference between coordinates of the vector of model expectations, E$\left[\boldsymbol X | \boldsymbol \theta\right]$ and corresponding expectations E$\left[\boldsymbol X | \boldsymbol \theta^*\right ]$ given independence grow for the toy RBM as the values of $\boldsymbol \theta$ are farther from zero, especially for large magnitudes of the dependence parameter $\theta_{11}$.
-This is a third indication that parameter vectors of large magnitude can readily lead to model impropriety in a RBM.
+Figure \ref{fig:uninterp} shows that the absolute difference between coordinates of the vector of E$\left[\boldsymbol X | \boldsymbol \theta\right]$ model expectations and E$\left[\boldsymbol X | \boldsymbol \theta^*\right ]$ corresponding expectations given independence grow for the toy RBM as the values of $\boldsymbol \theta$ are farther from zero, especially for large magnitudes of $\theta_{11}$, the dependence parameter.
+This is a third indication that parameter vectors of large magnitude readily can lead to model impropriety in an RBM.
 
@@
 For example, we could allow hiddens $(H_1, \dots, H_\nh) \in \{0,1\}^\nh$ and visibles $(V_1, \dots, V_\nv) \in \{0,1\}^\nv$, as in the previous sections or we could instead encode the state of the variables as $\{-1,1\}^\nh$ and $\{-1,1\}^\nv$.
-This will result in variables $\boldsymbol t(\boldsymbol X)$ from \eqref{eq:t} satisfying $\boldsymbol t(\boldsymbol x) \in \{0,1\}^{\nh + \nv + \nh*\nv}$ or $\boldsymbol t(\boldsymbol x) \in \{-1,1\}^{\nh + \nv + \nh*\nv}$ depending on how we encode "on" and "off" states in the nodes.
+This will result in variables $\boldsymbol t(\boldsymbol X)$ from \eqref{eq:t} satisfying $\boldsymbol t(\boldsymbol x) \in \{0,1\}^{\nh + \nv + \nh*\nv}$ or $\boldsymbol t(\boldsymbol x) \in \{-1,1\}^{\nh + \nv + \nh*\nv}$, depending on how we encode "on" and "off" states in the nodes.
 
@@
 These average magnitudes vary on a $1$-dimensional grid between `r min(plot_dat_grid$r1/(plot_dat_grid$H + plot_dat_grid$V))` and `r max(plot_dat_grid$r1/(plot_dat_grid$H + plot_dat_grid$V))` with `r length(unique(plot_dat_grid$r1/(plot_dat_grid$H + plot_dat_grid$V)))` breaks, yielding `r length(unique(plot_dat_grid$r1/(plot_dat_grid$H + plot_dat_grid$V)))^2` $2$-dimensional grid points.
-(By looking at the average magnitudes, we are able to later consider the potential benefit of shrinking each parameter value $\theta_i$ towards zero in a Bayesian fitting technique.) At each point in the grid, `r dim(res$samp[[1]])[1]` vectors ($\boldsymbol \theta_{main}$) are sampled uniformly on a sphere with radius corresponding to the first coordinate of the point and `r dim(res$samp[[1]])[1]` vectors ($\boldsymbol \theta_{interction}$) are sampled uniformly on a sphere with radius corresponding to the second coordinate of the point via sums of squared and scaled iid Normal$(0, 1)$ variables.
-These vectors are then paired to create `r dim(res$samp[[1]])[1]` values of $\boldsymbol \theta$ at each point in the grid.
+(By looking at the average magnitudes, we later are able to consider the potential benefit of shrinking each parameter value $\theta_i$ towards zero in a Bayesian fitting technique.) At each point in the grid, `r dim(res$samp[[1]])[1]` vectors ($\boldsymbol \theta_{main}$) are sampled uniformly on a sphere with radius corresponding to the first coordinate of the point, and `r dim(res$samp[[1]])[1]` vectors ($\boldsymbol \theta_{interction}$) are sampled uniformly on a sphere with radius corresponding to the second coordinate of the point via sums of squared and scaled iid Normal$(0, 1)$ variables.
+These vectors then are paired to create `r dim(res$samp[[1]])[1]` values of $\boldsymbol \theta$ at each point in the grid.
 
@@
 ```{r degen-plots, fig.cap="\\label{fig:degen-plots}Results from the numerical experiment, here looking at the fraction of models that were near-degenerate for each combination of magnitudes of the points $\\boldsymbol \\theta$ and model size, where $\\boldsymbol \\theta = (\\boldsymbol \\theta_{main}, \\theta_{interaction})$ is split into main and interaction parameters.
-Black lines show the contour levels for fraction of near-degeneracy, while the thick black line shows the level where the fraction of near-degenerate models is .05.", fig.height=3.5, fig.pos="ht!"}
+Black lines show the contour levels for fraction of near-degeneracy, while the thick black line shows the level where the fraction of near-degenerate models is $0.05$.", fig.height=3.5, fig.pos="ht!"}
 three_ways %>%
@@
 ```{r split-plots-dist, warning=FALSE, fig.cap="\\label{fig:split-plots-dist}The fraction of near-degenerate models for each magnitude of $\\boldsymbol \\theta$.
-For each number $\\nv$ of visibles in the model, the fraction near-degenerate moves away from zero at larger values of $||\\boldsymbol \\theta||$ as the number $\\nh$ of hidden variables increases and the slope becomes steeper as $\\nh$ increases as well.", fig.height=4, fig.pos="ht!"}
+For each number $\\nv$ of visibles in the model, the fraction near-degenerate moves away from zero at larger values of $||\\boldsymbol \\theta||$ as the number $\\nh$ of hidden variables increases.
+Additionally, the slope becomes steeper as $\\nh$ increases as well.", fig.height=4, fig.pos="ht!"}
 plot_dat_grid %>% 
@@
 
-Typically, fitting a RBM via maximum likelihood (ML) methods will be unfeasible due mainly to the intractability of the normalizing term $\gamma(\boldsymbol \theta)$ in a model \eqref{eq:pmf} of any realistic size.
-Ad hoc methods are often recommended instead, which aim to avoid this problem by using stochastic ML approximations that employ a small number of MCMC draws (i.e., contrastive divergence, [@hinton2002training]).
-
-However, computational concerns are not the only issues with fitting a RBM using ML.
-In addition, a RBM model, with the appropriate choice of parameters and number of hiddens, has the potential to re-create any distribution for the data (i.e., reproduce any specification of cell probabilities for the binary data outcomes).
-For example, @montufar2011refinements show that any distribution on $\{0, 1\}^{\nv}$ can be approximated arbitrarily well by a RBM with $2^{\nv-1} - 1$ hidden units.
-We provide a small example in the on-line supplementary materials that illustrates such approximations.
-
+Typically, fitting an RBM via maximum likelihood (ML) methods will be unfeasible due mainly to the intractability of the normalizing term $\gamma(\boldsymbol \theta)$ in a model \eqref{eq:pmf} of any realistic size.
+Ad hoc methods often are recommended instead, which aim to avoid this problem by using stochastic ML approximations that employ a small number of MCMC draws (i.e., contrastive divergence, [@hinton2002training]).
+
+However, computational concerns are not the only issues with fitting an RBM using ML.
+In addition, an RBM model, with the appropriate choice of parameters and number of hiddens, has the potential to re-create any distribution for the data (i.e., reproduce any specification of cell probabilities for the binary data outcomes).
+For example, @montufar2011refinements show that any distribution on $\{0, 1\}^{\nv}$ can be approximated arbitrarily well by an RBM with $2^{\nv-1} - 1$ hidden units.
+We provide a small example that illustrates such approximations.
+
+For simplicity, consider a model with two visible variables $(V_1,V_2)$ and one hidden $H_1$ so that there are four possible data realizations for $(V_1,V_2)$  given by $(\pm 1, \pm 1)$, and we may express the model probabilities as
+$$
+P(V_1=v_1,V_2=v_2| \theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21}) \propto  \exp\left( v_1  \theta_{v_1} + v_2 \theta_{v_2}\right) \sum_{h \in \{\pm 1\}}\exp\left( h[ \theta_{h_1}  + \theta_{11} v_1 + \theta_{21} v_2] \right),
+$$
+for $(v_1,v_2)\in \{-1, 1\}^2$, in terms of real-valued parameters $\theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21}$. Given some specified cell probabilities, say
+\begin{equation}
+\label{eq:0}
+0\leq p_{(-1,-1)},\;p_{(1,-1)}, \;p_{(-1,1)},\;p_{(1,1)},
+\end{equation} 
+for the outcomes $(\pm 1, \pm 1)$, the online appendix provides technical details to show parameter values $(\theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21})$ may be chosen to match any given cell probabilities with arbitrary closeness.
+In fact, when the cell probabilities \eqref{eq:0} are all strictly positive, parameters in the RBM model can be specified to reflect these probabilities exactly.
+And, when one or more of the cell probabilities \eqref{eq:0} are zero, the corresponding RBM probabilities $P(V_1=v_1,V_2=_v2| \theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21})$ may never be identically zero (due to exponential terms in the model) but parameters still can be selected to make the appropriate RBM cell probabilities arbitrarily small.
 Furthermore, as development in the online appendix shows, not only is it possible to approximate any distribution on the visibles arbitrarily well [cf.
 @montufar2011refinements], but dramatically different parameter settings can induce the same RBM model (beyond mere symmetries in parameterization).
-A further (and somewhat odd) consequence of the RBM parameterization is then that when fitting the RBM model by likelihood-based methods, we may already know the nature of the answer before we begin: namely, such  fitting will simply reproduce the empirical distribution of the training data if sufficiently many hiddens are in the model.
+A further (and somewhat odd) consequence of the RBM parameterization is then that when fitting the RBM model by likelihood-based methods, we already may know the nature of the answer before we begin: namely, such fitting simply will reproduce the empirical distribution of the training data if sufficiently many hiddens are in the model.
 There then can be no model refinements or smoothing from the RBM.
 That is, based on a random sample of vectors of visible variables, the model for the cell probabilities with the highest likelihood over  *all possible model classes* (i.e., RBM-based or not) is the empirical distribution, and the over-parameterization of the RBM model itself ensures that this empirical distribution can be arbitrarily well-approximated.
-Not only does RBM model fitting based on ML seek to reproduce the empirical distribution, whenever this empirical distribution contains empty cells, fitting steps for the RBM model will further aim to choose parameters that necessarily diverge to infinity in magnitude in order to zero-out the corresponding RBM cell probabilities.
+
+For illustration, considering the simple example from above with $n$ iid observations, each consisting of two realized visibles $(V_1,V_2)$ the empirical cell frequencies from the sample are the highest likelihood choices for the cell probabilities $p_{(-1,-1)},p_{(1,-1)}, p_{(-1,1)},p_{(1,1)}$ in \eqref{eq:0} before any parametric model is introduced for improved analysis, and the discussion above indicates that RBM model parameters can be chosen to re-create this empirical distribution to an arbitrarily close degree.
+Not only does RBM model fitting based on ML seek to reproduce the empirical distribution, whenever this empirical distribution contains empty cells, fitting steps for the RBM model further will aim to choose parameters that necessarily diverge to infinity in magnitude in order to zero-out the corresponding RBM cell probabilities.
 In data applications with a large sample space, it is unlikely that the training set will include at least one of each possible vector outcome (unlike this small example).
@@
 
-Here we consider what might be done in a principled manner to prevent both overfitting and model impropriety, testing on a $\nv = \nh = 4$ case that already stretches the limits of what is computable --- in particular we consider Bayes methods.
+Here we consider what might be done in a principled manner to prevent both overfitting and model impropriety, testing on an $\nv = \nh = 4$ case that already stretches the limits of what is computable --- in particular we consider Bayes methods.
 We employ a Bayesian framework because of the ability to obtain uncertainty quantification with little to no extra effort, via posterior credible intervals.
@@
 1.
-*A "trick" prior.* Here we cancel out normalizing term in the likelihood (from $\gamma(\boldsymbol \theta)$ in \eqref{eq:pmf}) so that resulting full conditionals of $\boldsymbol \theta$ are multivariate Normal.
+*A "trick" prior.* Here we cancel out the normalizing term in the likelihood (from $\gamma(\boldsymbol \theta)$ in \eqref{eq:pmf}) so that resulting full conditionals of $\boldsymbol \theta$ are multivariate Normal.
 Namely, this involves a prior of the form
@@
     for hyperparameter choices $0< C_2 < C_1$.
-The unknown hidden variables $h_j$ are also directly treated as latent variables and are sampled in each MCMC iterative draw from the posterior distribution.
+The unknown hidden variables $h_j$ also are treated directly as latent variables and are sampled in each MCMC iterative draw from the posterior distribution.
 This is the method of @li2014biclustering.
@@
 Full conditional distributions are not conjugate, and simulation from the posterior was accomplished using a geometric adaptive Metropolis Hastings step [@zhou2014some] and calculation of likelihood normalizing constant.
-(This computation is barely feasible for a problem of this size and would be unfeasible for larger problems.) Here the hidden variables $h_j$ are again carried along in the MCMC implementation as latent variables.
+(This computation barely is feasible for a problem of this size and would be unfeasible for larger problems.) Here the hidden variables $h_j$ again are carried along in the MCMC implementation as latent variables.
 We will refer to this method as Bayes with Truncated Normal prior and Latent Variables (BwTNLV).
@@
 
-```{r acf, fig.cap="\\label{fig:acf}The autocorrelation functions (ACF) for the posterior probabilityies of all $2^4 = 16$ possible outcomes for the vector of four visibles assessed at multiple lags for each method with BwTNLV in black and BwTNML in red.
+```{r acf, fig.cap="\\label{fig:acf}The autocorrelation functions (ACF) for the posterior probabilities of all $2^4 = 16$ possible outcomes for the vector of four visibles assessed at multiple lags for each method with BwTNLV in black and BwTNML in red.
 As expected, ACF corresponding to the method that marginalizes out the hidden variables from the likelihood decreases to zero at a much faster rate, indicating better mixing for the chain."}
@@
 From these posterior predictive checks, it is evident that BwTNML produces the best fit to the data.
-Furthermore, along with the discussion of Section \ref{model-fitting}, Figure \ref{fig:fitting-plot} also shows that it can be undesirable to seek to perfectly re-create an empirical distribution in fitting RBM models (i.e., true model probabilities may differ substantially).
+Furthermore, along with the discussion of Section \ref{model-fitting}, Figure \ref{fig:fitting-plot} also shows that it can be undesirable to seek to re-create an empirical distribution perfectly in fitting RBM models (i.e., true model probabilities may differ substantially).
 The priors in the BwTNML method constrain the RBM model fit to avoid replication of the empirical distribution and better estimate the underlying true data generating probabilities.
@@
 ```{r fitting-plot, fig.cap="\\label{fig:fitting-plot}Posterior probabilities of $16 = 2^4$ possible realizations of $4$ visibles using two of the three Bayesian fitting techniques, BwTPLV and BwTNML.
-The black vertical lines show the true probabilities of each vector of visibles based on the parameters used to generate the training data while the red vertical lines show the empirical distribution.
+The black vertical lines show the true probabilities of each vector of visibles based on the parameters used to generate the training data, while the red vertical lines show the empirical distribution.
 BwTNML produces the best fit for the data, however is also the most computationally intensive and least feasible with a real dataset.", fig.height=6}
@@
 
-An alternative possibility to using MCMC to approximate the posterior for Bayesian inference in a RBM model is to use variational methods for approximation of the posterior [@salimans2015markov].
+An alternative possibility to using MCMC to approximate the posterior for Bayesian inference in an RBM model is to use variational methods for approximation of the posterior [@salimans2015markov].
 The advantage to using variational methods would be computational scalability, but the disadvantages are potentially numerous.
@@
 Additionally, these models are difficult to fit using a rigorous methodology that can incorporate estimation uncertainty, due to the dimension of the parameter space coupled with the size of the latent variable space.
-We have presented three fully Bayes-principled MCMC-based methods for fitting RBMs, which provide posterior distributions for quantifying uncertainty in estimating model parameters and probabilities (e.g., Figure \ref{fig:fitting-plot}) while aiming to avoid areas of the parameter space that will potentially lead to model impropriety.
+We have presented three fully Bayes-principled MCMC-based methods for fitting RBMs, which provide posterior distributions for quantifying uncertainty in estimating model parameters and probabilities (e.g., Figure \ref{fig:fitting-plot}) while aiming to avoid areas of the parameter space that potentially will lead to model impropriety.
 We have shown small examples where this fitting methodology does work, but becomes quickly intractable in large data cases.
 
-The current common practice is to use a kind of MCMC to overcome fitting complexities, and due to the extreme flexibility in this model class, rigorous likelihood-based fitting for a RBM can typically seek to merely merely return the (discrete) empirical distribution for visibles.
-Current fitting methodology for RBMs does not provide uncertainty quantification for the fitted parameters, and it is unclear that they seeks to address the issues in model impropriety detailed in this paper.
-Practitioners should be aware of this and employ some form of regularization or penalization in the fitting.
+The current common practice is to use a kind of MCMC to overcome fitting complexities, and due to the extreme flexibility in this model class, rigorous likelihood-based fitting for an RBM typically can merely seek to return the (discrete) empirical distribution for visibles.
+Current fitting methodology for RBMs does not provide uncertainty quantification for the fitted parameters, and it is unclear whether it seeks to address the issues in model impropriety detailed in this paper.
+Practitioners should be aware of this extreme flexibility and employ some form of regularization or penalization in the fitting.
 
@@
 Ultimately, it is not unquestionably clear that RBM models are useful as generative models.
-As a concern, without appropriate generative behavior in a RBM model, the uncertainty in estimated model parameters becomes impossible to realistically assess and the model can further lose useful application to prediction problems (e.g., realistic predictive distributions).
+As a concern, without appropriate generative behavior in an RBM model, the uncertainty in estimated model parameters becomes impossible to assess realistically and further, the model can lose useful application to prediction problems (e.g., realistic predictive distributions).
 In the case of classification, predictive distributions may not be the ultimate goal, but when S-instability is present, small (imperceptible) differences in the data may lead to greatly different probabilities and thus greatly different classifications.

### 2018-09-27 writing/paper-rbm/paper.tex
grammar check

--- 2018-09-18 writing/paper-rbm/paper.tex
@@
 the visible layer in a second RBM, a deep architecture can be created.
-RBMs are thought to thereby have the ability to encode very complex and
+RBMs thereby are thought to have the ability to encode very complex and
 rich structures in data, making them attractive for supervised learning.
-However, the generative behavior of RBMs is largely unexplored and
+However, the generative behavior of RBMs largely is unexplored and
 typical fitting methodology does not easily allow for uncertainty
@@
 potential Bayes fitting of such (highly flexible) models, especially as
-Gibbs sampling (quasi-Bayes) methods are often advocated for the RBM
+Gibbs sampling (quasi-Bayes) methods often are advocated for the RBM
 model structure.
@@
 
-The data mining and machine learning communities have recently shown
+The data mining and machine learning communities recently have shown
 great interest in deep learning, specifically in stacked restricted
@@
 Srivastava, Salakhutdinov, and Hinton 2013; Le Roux and Bengio 2008 for
-examples). A RBM is a probabilistic undirected graphical model (for
+examples). An RBM is a probabilistic undirected graphical model (for
 discrete or continuous random variables) with two layers, one hidden and
 one visible, with no conditional dependency within a layer (Smolensky
-1986). These models have reportedly been used with success in
+1986). These models reportedly have been used with success in
 classification of images (Larochelle and Bengio 2008; Srivastava and
@@
 mechanism underlying the observed phenomena'' (Lehmann 1990; G. E. P.
-Box 1967). At issue, simulation from RBMs can often produce data lacking
+Box 1967). At issue, simulation from RBMs often can produce data lacking
 realistic variability so that such models may thereby fail to
@@
 behavior relates to model degeneracy (or near degeneracy), which is a
-statistical concern in that many data processes of interest are
-realistically not degenerate in their spectrum of potential outcomes.
-For example, when sampling data from a nearly degenerate RBM used to
-model an imaging process, only a small set of output possibilities
-receives nearly all probability, and thus a sample of images will all be
-copies of the same one image (or small number of images). An example of
-ten 4-pixel images simulated from a nearly degenerate RBM model is
-compared to ten 4-pixel images simulated from a non-degenerate RBM model
-in Figure \ref{fig:sample-models}. The degenerate model places almost
-all probability on one outcome, causing the image to be generated
+statistical concern in that, realistically, many data processes of
+interest are not degenerate in their spectrum of potential outcomes. For
+example, when sampling data from a nearly degenerate RBM used to model
+an imaging process, only a small set of output possibilities receives
+nearly all probability, and thus a sample of images will all be copies
+of the same one image (or small number of images). An example of ten
+4-pixel images simulated from a nearly degenerate RBM model is compared
+to ten 4-pixel images simulated from a non-degenerate RBM model in
+Figure \ref{fig:sample-models}. The degenerate model places almost all
+probability on one outcome, causing the image to be generated
 repeatedly, whereas the non-degenerate model shows more potentially
@@
 
-In addition to such degeneracy, we find that RBMs can easily exhibit
+In addition to such degeneracy, we find that RBMs easily can exhibit
 types of instability, related to how sensitive the model probabilities
@@
 different classification in an image classification problem. Occurrences
-of such behavior have recently been documented in RBMs (Li 2014), as
+of such behavior recently have been documented in RBMs (Li 2014), as
 well as other deep architectures (Szegedy et al. 2013; Nguyen, Yosinski,
@@
 Furthermore, these MCMC algorithms involve updating potentially many
-latent variables (hiddens) which can critically influence convergence in
-MCMC-based likelihood methods. Applying basic statistical principles in
-fitting RBM models of tractable size, we compare three fully Bayesian
+latent variables (hiddens), which can critically influence convergence
+in MCMC-based likelihood methods. Applying basic statistical principles
+in fitting RBM models of tractable size, we compare three fully Bayesian
 techniques involving MCMC, which are computationally more accessible
-than direct maximum likelihood and also aim to avoid parts of a RBM
+than direct maximum likelihood and also aim to avoid parts of an RBM
 parameter space that yield unattractive models. As might be expected,
@@
 As a factor compounding the computational challenges, issues in model
-fitting traceable to model flexibility are potentially more concerning.
-For a RBM model with enough hidden variables, any distribution for the
+fitting traceable to model flexibility potentially are more concerning.
+For an RBM model with enough hidden variables, any distribution for the
 visibles can be approximated arbitrarily well (Le Roux and Bengio 2008;
@@
 visible variables provides a highest likelihood benchmark, before any
-parametric model class is even introduced and applied to obtain a
+parametric model class even is introduced and applied to obtain a
 refinement of model fit. As a consequence, we find that any fully
-principled fitting method based on the likelihood for a RBM with enough
+principled fitting method based on the likelihood for an RBM with enough
 hidden variables will seek to reproduce the (discrete) empirical
@@
 perhaps even unexpected compared to most modeling scenarios, in that no
-``smoothed distribution'' may result when fitting a RBM model of
-sufficient size with a rigorous likelihood-based method. We are
-therefore led to be skeptical that models that involve these structures
-(like deep Boltzmann machines) can achieve useful prediction or
-inference in a principled way without intentionally limiting the
-flexibility of the fitted model.
+``smoothed distribution'' may result when fitting an RBM model of
+sufficient size with a rigorous likelihood-based method. We therefore
+are led to be skeptical that models that involve these structures (like
+deep Boltzmann machines) can achieve useful prediction or inference in a
+principled way without intentionally limiting the flexibility of the
+fitted model.
 
@@
 uncertainty of estimation and the formulation of predictive
-distributions. These are closely tied to the probability properties of
-the RBM model for explaining data generation and our exposition intends
+distributions. These are tied closely to the probability properties of
+the RBM model for explaining data generation, and our exposition intends
 to contribute to a better understanding in this direction.
@@
 specified for discrete or continuous random variables, binary variables
-being most commonly considered. In this paper, we consider the binary
-case for concreteness. A RBM architecture has two layers, hidden
+most commonly being considered. In this paper, we consider the binary
+case for concreteness. An RBM architecture has two layers, hidden
 (\(\mathcal{H}\)) and visible (\(\mathcal{V}\)), with no dependency
@@
 For example, binary images can be classified through a process that
-treats the pixel values as the visible variables \(v_i\) in a RBM model
+treats the pixel values as the visible variables \(v_i\) in an RBM model
 (Hinton, Osindero, and Teh 2006).
@@
 Let \(\boldsymbol x = (h_1, \dots, h_\nh, v_1,\dots,v_\nv)\) represent
-the states of the visible and hidden nodes in a RBM for some integers
+the states of the visible and hidden nodes in an RBM for some integers
 \(\nv, \nh \ge 1\). Each single binary random variable, visible or
@@
 weight the values of the visible \(v_i\) and hidden \(h_j\) nodes in
-probabilities \eqref{eq:pmf} and the interaction effect parameters
+probabilities \eqref{eq:pmf}, and the interaction effect parameters
 (\(\theta_{ij}\)) weight the values of the connections \(v_i h_j\), or
@@
 
-The highly flexible nature of a RBM (having as it does
+The highly flexible nature of an RBM (having as it does
 \(\nh + \nv + \nh*\nv\) parameters) creates at least three kinds of
@@
 \emph{degeneracy}, \emph{instability}, and \emph{uninterpretability}. In
-this section we define these, consider how to quantify them in a RBM,
+this section we define these, consider how to quantify them in an RBM,
 and point out relationships among them.
@@
 random graph models, \(\mathcal{X}\) denotes all possible graphs that
-can be constructed from a set of nodes and an exponentially
+can be constructed from a set of nodes, and an exponentially
 parameterized random graph model has a distribution with a probability
@@
 (defining the boundary of the hull \(C\)) and can be expected to
-\emph{not} be pulled deep into the interior of \(C\).
-
-A RBM model can be thought to exhibit an analogous form of
+\emph{not} be pulled deeply into the interior of \(C\).
+
+An RBM model can be thought to exhibit an analogous form of
 \emph{near-degeneracy} when there is a disproportionate amount of
@@
 introduced a concept of model deficiency related to \emph{instability}.
-Instability can be roughly thought of as excessive sensitivity in the
+Instability roughly can be thought of as excessive sensitivity in the
 model, where small changes in the components of potential data outcomes,
@@
 model instability can be viewed on a spectrum of potential sensitivity
-in probability structure, with model degeneracy included as a extreme or
-limiting case of instability. To quantify ``instability'' more
+in probability structure, with model degeneracy included as an extreme
+or limiting case of instability. To quantify ``instability'' more
 rigorously (particularly beyond the definition given by Schweinberger
@@
 models, it becomes necessary to grow the number of model parameters (and
-in this process one may also arbitrarily expand the number of hidden
+in this process one also may expand arbitrarily the number of hidden
 variables used). To this end, let
@@
 
-In formulating a RBM model for a potentially large number of visibles
+In formulating an RBM model for a potentially large number of visibles
 (i.e., as \(\nv \rightarrow \infty\)), we will say that the ratio
@@
 
-In other words, a RBM model sequence is unstable if, given any
+In other words, an RBM model sequence is unstable if, given any
 \(c > 0\), there exists an integer \(n_c > 0\) so that
@@
 Schweinberger (2011) in that here RBM models for visibles
-\((v_1, \dots, v_\nv)\) do not form an exponential family and the
-dimensionality of \(\boldsymbol \theta_\nv\) is not fixed, but rather
+\((v_1, \dots, v_\nv)\) do not form an exponential family, and the
+dimensionality of \(\boldsymbol \theta_\nv\) is not fixed but rather
 grows with \(\nv\).
@@
 \Delta(\boldsymbol \theta_\nv) \equiv \max \left\{\log \frac{P_{\boldsymbol \theta_\nv}(\boldsymbol v)}{P_{\boldsymbol \theta_\nv}(\boldsymbol v^*)} : \boldsymbol v \text{ }\& \text{ } \boldsymbol v^* \in \mathcal{C}^\nv \text{ differ in exactly one component}\right\},
-\] denoting the biggest log-probability ratio for a one-component change
-in data outcomes (visibles) at a RBM parameter
+\] denoting the largest log-probability ratio for a one-component change
+in data outcomes (visibles) at an RBM parameter
 \(\boldsymbol \theta_\nv\). We then have the following result.
@@
 In other words, if the probability ratio \eqref{eq:elpr} is too large,
-then a RBM model sequence will exhibit large probability shifts for very
-small changes in data configurations (i.e., will exhibit instability).
-Such instability can be a concern in a model for reasons similar to
-those related to degeneracy: as outcomes vary over the sample space, the
-geography of probabilities is extremely rugged, with deep pits following
-sharp mountains. Recall the applied example of RBM models as a means to
-classify images. For data as pixels in an image, the instability result
-in Proposition \ref{prp:instab1} manifests itself as a one pixel change
-in an image (one component of the visible vector) resulting in a large
-shift in the probability, which in turn could result in a vastly
-different classification of the image. Examples of this kind of behavior
-have been presented in Szegedy et al. (2013) for deep learning models,
-in which a one pixel change in a test image results in a wildly
-different classification.
-
-Additionally, S-unstable RBM model sequences may be formally connected
+then an RBM model sequence will exhibit large probability shifts for
+very small changes in data configurations (i.e., will exhibit
+instability). Such instability can be a concern in a model for reasons
+similar to those related to degeneracy: as outcomes vary over the sample
+space, the geography of probabilities is extremely rugged, with deep
+pits following sharp mountains. Recall the applied example of RBM models
+as a means to classify images. For data as pixels in an image, the
+instability result in Proposition \ref{prp:instab1} manifests itself as
+a one pixel change in an image (one component of the visible vector),
+resulting in a large shift in the probability, which in turn could
+result in a vastly different classification of the image. Examples of
+this kind of behavior have been presented in Szegedy et al. (2013) for
+deep learning models, in which a one pixel change in a test image
+results in a wildly different classification.
+
+Additionally, S-unstable RBM model sequences may be connected formally
 to the near-degeneracy of Section \ref{near-degeneracy} (in which model
@@
 M_{\epsilon, \boldsymbol \theta_\nv} \equiv \left\{\boldsymbol v \in \mathcal{C}^\nv: \log P_{\boldsymbol \theta_\nv}(\boldsymbol v) > (1-\epsilon)\max\log\limits_{\boldsymbol v^*}P_{\boldsymbol \theta_\nv}(\boldsymbol v^*) + \epsilon\min\log\limits_{\boldsymbol v^*}P_{\boldsymbol \theta_\nv}(\boldsymbol v^*) \right\}
-\] for a given \(0 < \epsilon < 1\). Then S-unstable model sequences are
-guaranteed to be degenerate, as the following result shows.
+\] for a given \(0 < \epsilon < 1\). Then, S-unstable model sequences
+are guaranteed to be degenerate, as the following result shows.
 
@@
 up all probability on a specific set of outcomes for visibles, which
-could potentially be arbitrarily narrow. Proofs of Propositions
+potentially could be arbitrarily narrow. Proofs of Propositions
 \ref{prp:instab1} and \ref{prp:degen} follow from more general results
@@
 mean-structures fail to hold as anticipated by consideration of a model
-statement. We adapt this notion to RBM models. Note that in a RBM, the
+statement. We adapt this notion to RBM models. Note that in an RBM, the
 parameters \(\theta_{v_1}, \dots, \theta_{v_\nv}\) and
-\(\theta_{h_1}, \dots, \theta_{h_\nh}\) are naturally associated with
+\(\theta_{h_1}, \dots, \theta_{h_\nh}\) naturally are associated with
 main effects of visible and hidden variables and can be interpreted as
@@
 main effects. In such cases, the main effect parameters \(\theta_{v_1}\)
-and \(\theta_{h_j}\) are no longer interpretable in the models
+and \(\theta_{h_j}\) no longer are interpretable in the models
 (statements of marginal means) and the dependence parameters are so
@@
 with interpretable dependence parameters typically correspond to
-non-degenerate models, while degradation in interpretability is often
+non-degenerate models, while degradation in interpretability often is
 associated with model drift into degeneracy. To assess which parameter
@@
 effects but \(\boldsymbol \theta^*\) has \(0\) dependence parameters.)
-Uninterpretability is then avoided at a parametric specification
+Uninterpretability then is avoided at a parametric specification
 \(\boldsymbol \theta\) if the model expected value at
@@
 of \(\boldsymbol \theta\) and the three notions of model impropriety
-(near-degeneracy, instability, and uninterpretability), for RBM models
-of varying sizes.
+(near-degeneracy, instability, and uninterpretability) for RBM models of
+varying sizes.
 
@@
 To illustrate the ideas of model near-degeneracy, instability, and
-uninterpretability in a RBM, we consider first the smallest possible
+uninterpretability in an RBM, we consider first the smallest possible
 (toy) example that consists of one visible node \(v_1\) and one hidden
@@
 \] appearing in the RBM probabilities for this model. As this set is in
-three dimensions, we are able to explicitly illustrate the shape of
+three dimensions, we are able to illustrate explicitly the shape of
 boundary of the convex hull of \(\mathcal{T}\) and explore the behavior
@@
 
-\caption{\label{fig:degen-toy}Contour plots for the three parametric mean functions of sufficient statistics for a RBM with one visible and one hidden node.}\label{fig:degen-toy}
+\caption{\label{fig:degen-toy}Contour plots for the three parametric mean functions of sufficient statistics for an RBM with one visible and one hidden node.}\label{fig:degen-toy}
 \end{figure}
@@
 
-Finally to consider the effect of \(\boldsymbol \theta\) on potential
-model uninterpretability, we can look at the difference between model
-expectations, E\(\left[\boldsymbol X | \boldsymbol \theta\right]\), and
-expectations given independence,
-E\(\left[\boldsymbol X | \boldsymbol \theta^*\right ]\) for the tiny toy
-RBM model where \(\boldsymbol X = (V_1, H_1, V_1 H_1)\). This difference
-is given by \begin{align*}
+Finally, to consider the effect of \(\boldsymbol \theta\) on potential
+model uninterpretability, we can look at the difference between
+E\(\left[\boldsymbol X | \boldsymbol \theta\right]\) model expectations
+and E\(\left[\boldsymbol X | \boldsymbol \theta^*\right ]\) expectations
+given independence for the tiny toy RBM model where
+\(\boldsymbol X = (V_1, H_1, V_1 H_1)\). This difference is given by
+\begin{align*}
 \text{E}\left[\boldsymbol X | \boldsymbol \theta\right] -   \text{E}\left[\boldsymbol X | \boldsymbol \theta^* \right] 
@@
 
-\caption{\label{fig:uninterp}The absolute difference between coordinates of model expectations, E$\left[\boldsymbol X | \boldsymbol \theta\right]$, and expectations given independence, E$\left[\boldsymbol X | \boldsymbol \theta^* \right ]$ for a RBM with one visible and one hidden node.
-As an indicator of uninterpretability, note that differences in expectations increase as the dependence parameter $\theta_{11}$ deviates from zero.}\label{fig:uninterp}
+\caption{\label{fig:uninterp}The absolute difference between coordinates of E$\left[\boldsymbol X | \boldsymbol \theta\right]$ model expectations and E$\left[\boldsymbol X | \boldsymbol \theta^* \right ]$ expectations given independence for an RBM with one visible and one hidden node.
+As an indicator of uninterpretability, note that differences in expectations increase as$\theta_{11}$, the dependence parameter, deviates from zero.}\label{fig:uninterp}
 \end{figure}
@@
 Figure \ref{fig:uninterp} shows that the absolute difference between
-coordinates of the vector of model expectations,
-E\(\left[\boldsymbol X | \boldsymbol \theta\right]\) and corresponding
-expectations E\(\left[\boldsymbol X | \boldsymbol \theta^*\right ]\)
-given independence grow for the toy RBM as the values of
-\(\boldsymbol \theta\) are farther from zero, especially for large
-magnitudes of the dependence parameter \(\theta_{11}\). This is a third
-indication that parameter vectors of large magnitude can readily lead to
-model impropriety in a RBM.
+coordinates of the vector of
+E\(\left[\boldsymbol X | \boldsymbol \theta\right]\) model expectations
+and E\(\left[\boldsymbol X | \boldsymbol \theta^*\right ]\)
+corresponding expectations given independence grow for the toy RBM as
+the values of \(\boldsymbol \theta\) are farther from zero, especially
+for large magnitudes of \(\theta_{11}\), the dependence parameter. This
+is a third indication that parameter vectors of large magnitude readily
+can lead to model impropriety in an RBM.
 
@@
 \(\boldsymbol t(\boldsymbol x) \in \{0,1\}^{\nh + \nv + \nh*\nv}\) or
-\(\boldsymbol t(\boldsymbol x) \in \{-1,1\}^{\nh + \nv + \nh*\nv}\)
+\(\boldsymbol t(\boldsymbol x) \in \{-1,1\}^{\nh + \nv + \nh*\nv}\),
 depending on how we encode ``on'' and ``off'' states in the nodes.
@@
 breaks, yielding 576 \(2\)-dimensional grid points. (By looking at the
-average magnitudes, we are able to later consider the potential benefit
+average magnitudes, we later are able to consider the potential benefit
 of shrinking each parameter value \(\theta_i\) towards zero in a
@@
 (\(\boldsymbol \theta_{main}\)) are sampled uniformly on a sphere with
-radius corresponding to the first coordinate of the point and 100
+radius corresponding to the first coordinate of the point, and 100
 vectors (\(\boldsymbol \theta_{interction}\)) are sampled uniformly on a
@@
 via sums of squared and scaled iid Normal\((0, 1)\) variables. These
-vectors are then paired to create 100 values of \(\boldsymbol \theta\)
+vectors then are paired to create 100 values of \(\boldsymbol \theta\)
 at each point in the grid.
@@
 \centering
-\includegraphics{paper_files/figure-latex/degen-plots-1.pdf}
+\includegraphics{paper_files/figure-latex/degen-plots-1.eps}
 \caption{\label{fig:degen-plots}Results from the numerical experiment,
@@
 line shows the level where the fraction of near-degenerate models is
-.05.}
+\(0.05\).}
 \end{figure}
@@
 \centering
-\includegraphics{paper_files/figure-latex/instab-plots-1.pdf}
+\includegraphics{paper_files/figure-latex/instab-plots-1.eps}
 \caption{\label{fig:instab-plots}The sample mean value of
@@
 \centering
-\includegraphics{paper_files/figure-latex/uninterp-plots-1.pdf}
+\includegraphics{paper_files/figure-latex/uninterp-plots-1.eps}
 \caption{\label{fig:uninterp-plots}The sample mean of the maximum
@@
 \centering
-\includegraphics{paper_files/figure-latex/split-plots-dist-1.pdf}
+\includegraphics{paper_files/figure-latex/split-plots-dist-1.eps}
 \caption{\label{fig:split-plots-dist}The fraction of near-degenerate
@@
 away from zero at larger values of \(||\boldsymbol \theta||\) as the
-number \(\nh\) of hidden variables increases and the slope becomes
-steeper as \(\nh\) increases as well.}
+number \(\nh\) of hidden variables increases. Additionally, the slope
+becomes steeper as \(\nh\) increases as well.}
 \end{figure}
@@
 
-Typically, fitting a RBM via maximum likelihood (ML) methods will be
+Typically, fitting an RBM via maximum likelihood (ML) methods will be
 unfeasible due mainly to the intractability of the normalizing term
 \(\gamma(\boldsymbol \theta)\) in a model \eqref{eq:pmf} of any
-realistic size. Ad hoc methods are often recommended instead, which aim
+realistic size. Ad hoc methods often are recommended instead, which aim
 to avoid this problem by using stochastic ML approximations that employ
@@
 
-However, computational concerns are not the only issues with fitting a
-RBM using ML. In addition, a RBM model, with the appropriate choice of
+However, computational concerns are not the only issues with fitting an
+RBM using ML. In addition, an RBM model, with the appropriate choice of
 parameters and number of hiddens, has the potential to re-create any
@@
 Ay (2011) show that any distribution on \(\{0, 1\}^{\nv}\) can be
-approximated arbitrarily well by a RBM with \(2^{\nv-1} - 1\) hidden
-units. We provide a small example in the on-line supplementary materials
-that illustrates such approximations.
-
-Furthermore, as development in the online appendix shows, not only is it
-possible to approximate any distribution on the visibles arbitrarily
-well (cf. Montufar and Ay 2011), but dramatically different parameter
-settings can induce the same RBM model (beyond mere symmetries in
-parameterization). A further (and somewhat odd) consequence of the RBM
-parameterization is then that when fitting the RBM model by
-likelihood-based methods, we may already know the nature of the answer
-before we begin: namely, such fitting will simply reproduce the
-empirical distribution of the training data if sufficiently many hiddens
-are in the model. There then can be no model refinements or smoothing
-from the RBM. That is, based on a random sample of vectors of visible
-variables, the model for the cell probabilities with the highest
-likelihood over \emph{all possible model classes} (i.e., RBM-based or
-not) is the empirical distribution, and the over-parameterization of the
-RBM model itself ensures that this empirical distribution can be
-arbitrarily well-approximated. Not only does RBM model fitting based on
-ML seek to reproduce the empirical distribution, whenever this empirical
-distribution contains empty cells, fitting steps for the RBM model will
-further aim to choose parameters that necessarily diverge to infinity in
-magnitude in order to zero-out the corresponding RBM cell probabilities.
-In data applications with a large sample space, it is unlikely that the
-training set will include at least one of each possible vector outcome
-(unlike this small example). This implies that some RBM model parameters
-must diverge to \(+\infty\) to mimic the empirical distribution with
-empty cells and, as we have already discussed in Section
+approximated arbitrarily well by an RBM with \(2^{\nv-1} - 1\) hidden
+units. We provide a small example that illustrates such approximations.
+
+For simplicity, consider a model with two visible variables
+\((V_1,V_2)\) and one hidden \(H_1\) so that there are four possible
+data realizations for \((V_1,V_2)\) given by \((\pm 1, \pm 1)\), and we
+may express the model probabilities as \[
+P(V_1=v_1,V_2=v_2| \theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21}) \propto  \exp\left( v_1  \theta_{v_1} + v_2 \theta_{v_2}\right) \sum_{h \in \{\pm 1\}}\exp\left( h[ \theta_{h_1}  + \theta_{11} v_1 + \theta_{21} v_2] \right),
+\] for \((v_1,v_2)\in \{-1, 1\}^2\), in terms of real-valued parameters
+\(\theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21}\).
+Given some specified cell probabilities, say \begin{equation}
+\label{eq:0}
+0\leq p_{(-1,-1)},\;p_{(1,-1)}, \;p_{(-1,1)},\;p_{(1,1)},
+\end{equation} for the outcomes \((\pm 1, \pm 1)\), the online appendix
+provides technical details to show parameter values
+\((\theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21})\)
+may be chosen to match any given cell probabilities with arbitrary
+closeness. In fact, when the cell probabilities \eqref{eq:0} are all
+strictly positive, parameters in the RBM model can be specified to
+reflect these probabilities exactly. And, when one or more of the cell
+probabilities \eqref{eq:0} are zero, the corresponding RBM probabilities
+\(P(V_1=v_1,V_2=_v2| \theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21})\)
+may never be identically zero (due to exponential terms in the model)
+but parameters still can be selected to make the appropriate RBM cell
+probabilities arbitrarily small. Furthermore, as development in the
+online appendix shows, not only is it possible to approximate any
+distribution on the visibles arbitrarily well (cf. Montufar and Ay
+2011), but dramatically different parameter settings can induce the same
+RBM model (beyond mere symmetries in parameterization). A further (and
+somewhat odd) consequence of the RBM parameterization is then that when
+fitting the RBM model by likelihood-based methods, we already may know
+the nature of the answer before we begin: namely, such fitting simply
+will reproduce the empirical distribution of the training data if
+sufficiently many hiddens are in the model. There then can be no model
+refinements or smoothing from the RBM. That is, based on a random sample
+of vectors of visible variables, the model for the cell probabilities
+with the highest likelihood over \emph{all possible model classes}
+(i.e., RBM-based or not) is the empirical distribution, and the
+over-parameterization of the RBM model itself ensures that this
+empirical distribution can be arbitrarily well-approximated.
+
+For illustration, considering the simple example from above with \(n\)
+iid observations, each consisting of two realized visibles \((V_1,V_2)\)
+the empirical cell frequencies from the sample are the highest
+likelihood choices for the cell probabilities
+\(p_{(-1,-1)},p_{(1,-1)}, p_{(-1,1)},p_{(1,1)}\) in \eqref{eq:0} before
+any parametric model is introduced for improved analysis, and the
+discussion above indicates that RBM model parameters can be chosen to
+re-create this empirical distribution to an arbitrarily close degree.
+Not only does RBM model fitting based on ML seek to reproduce the
+empirical distribution, whenever this empirical distribution contains
+empty cells, fitting steps for the RBM model further will aim to choose
+parameters that necessarily diverge to infinity in magnitude in order to
+zero-out the corresponding RBM cell probabilities. In data applications
+with a large sample space, it is unlikely that the training set will
+include at least one of each possible vector outcome (unlike this small
+example). This implies that some RBM model parameters must diverge to
+\(+\infty\) to mimic the empirical distribution with empty cells and, as
+we have already discussed in Section
 \ref{explorations-of-model-properties-through-simulation}, large
@@
 Here we consider what might be done in a principled manner to prevent
-both overfitting and model impropriety, testing on a \(\nv = \nh = 4\)
+both overfitting and model impropriety, testing on an \(\nv = \nh = 4\)
 case that already stretches the limits of what is computable --- in
@@
 \item
-  \emph{A ``trick'' prior.} Here we cancel out normalizing term in the
-  likelihood (from \(\gamma(\boldsymbol \theta)\) in \eqref{eq:pmf}) so
-  that resulting full conditionals of \(\boldsymbol \theta\) are
+  \emph{A ``trick'' prior.} Here we cancel out the normalizing term in
+  the likelihood (from \(\gamma(\boldsymbol \theta)\) in \eqref{eq:pmf})
+  so that resulting full conditionals of \(\boldsymbol \theta\) are
   multivariate Normal. Namely, this involves a prior of the form
@@
    \end{align*} for hyperparameter choices \(0< C_2 < C_1\). The unknown
-  hidden variables \(h_j\) are also directly treated as latent variables
+  hidden variables \(h_j\) also are treated directly as latent variables
   and are sampled in each MCMC iterative draw from the posterior
@@
   2014) and calculation of likelihood normalizing constant. (This
-  computation is barely feasible for a problem of this size and would be
-  unfeasible for larger problems.) Here the hidden variables \(h_j\) are
-  again carried along in the MCMC implementation as latent variables. We
-  will refer to this method as Bayes with Truncated Normal prior and
-  Latent Variables (BwTNLV).
+  computation barely is feasible for a problem of this size and would be
+  unfeasible for larger problems.) Here the hidden variables \(h_j\)
+  again are carried along in the MCMC implementation as latent
+  variables. We will refer to this method as Bayes with Truncated Normal
+  prior and Latent Variables (BwTNLV).
 \item
@@
 \centering
-\includegraphics{paper_files/figure-latex/acf-1.pdf}
+\includegraphics{paper_files/figure-latex/acf-1.eps}
 \caption{\label{fig:acf}The autocorrelation functions (ACF) for the
-posterior probabilityies of all \(2^4 = 16\) possible outcomes for the
+posterior probabilities of all \(2^4 = 16\) possible outcomes for the
 vector of four visibles assessed at multiple lags for each method with
@@
 \ref{fig:fitting-plot} also shows that it can be undesirable to seek to
-perfectly re-create an empirical distribution in fitting RBM models
+re-create an empirical distribution perfectly in fitting RBM models
 (i.e., true model probabilities may differ substantially). The priors in
@@
 \centering
-\includegraphics{paper_files/figure-latex/fitting-plot-1.pdf}
+\includegraphics{paper_files/figure-latex/fitting-plot-1.eps}
 \caption{\label{fig:fitting-plot}Posterior probabilities of \(16 = 2^4\)
@@
 true probabilities of each vector of visibles based on the parameters
-used to generate the training data while the red vertical lines show the
-empirical distribution. BwTNML produces the best fit for the data,
+used to generate the training data, while the red vertical lines show
+the empirical distribution. BwTNML produces the best fit for the data,
 however is also the most computationally intensive and least feasible
@@
 An alternative possibility to using MCMC to approximate the posterior
-for Bayesian inference in a RBM model is to use variational methods for
+for Bayesian inference in an RBM model is to use variational methods for
 approximation of the posterior (Salimans, Kingma, and Welling 2015). The
@@
 and probabilities (e.g., Figure \ref{fig:fitting-plot}) while aiming to
-avoid areas of the parameter space that will potentially lead to model
+avoid areas of the parameter space that potentially will lead to model
 impropriety. We have shown small examples where this fitting methodology
@@
 complexities, and due to the extreme flexibility in this model class,
-rigorous likelihood-based fitting for a RBM can typically seek to merely
-merely return the (discrete) empirical distribution for visibles.
-Current fitting methodology for RBMs does not provide uncertainty
-quantification for the fitted parameters, and it is unclear that they
-seeks to address the issues in model impropriety detailed in this paper.
-Practitioners should be aware of this and employ some form of
+rigorous likelihood-based fitting for an RBM typically can merely seek
+to return the (discrete) empirical distribution for visibles. Current
+fitting methodology for RBMs does not provide uncertainty quantification
+for the fitted parameters, and it is unclear whether it seeks to address
+the issues in model impropriety detailed in this paper. Practitioners
+should be aware of this extreme flexibility and employ some form of
 regularization or penalization in the fitting.
@@
 generative models. As a concern, without appropriate generative behavior
-in a RBM model, the uncertainty in estimated model parameters becomes
-impossible to realistically assess and the model can further lose useful
-application to prediction problems (e.g., realistic predictive
+in an RBM model, the uncertainty in estimated model parameters becomes
+impossible to assess realistically and further, the model can lose
+useful application to prediction problems (e.g., realistic predictive
 distributions). In the case of classification, predictive distributions

### 2018-09-30 writing/paper-rbm/appendices/appendix.Rmd
revisions

--- 2017-07-08 writing/paper-rbm/appendices/appendix-center.Rmd
@@
-\clearpage
+---
+title: Supplement to "Properties and Bayesian fitting of restricted Boltzmann machines"
+output: 
+  bookdown::pdf_book:
+    includes:
+      in_header: ../../resources/latex/header.tex
+    toc: no
+---
 
-# (APPENDIX) Appendix {-}
-
-# On the one-to-one correspondence between parameters and moments in the RBM distribution {#appendix-rbm}
+# Some characterizations of RBM distributions on all nodes through parameter and moment conditions {#appendix-rbm}
 
@@
 In the following, for convenience, we say a symmetric random variable $X$ has a Bernoulli$(1/2)$ distribution if $P(X = 1) = 1/2 = P(X = -1)$.
+
+```{proposition, label="propone", echo=TRUE}
+Let $\nv, \nh \geq 1$ and suppose the random vector $X =(V_1,...,V_{m},H_1,...,H_{n})$ assumes values in $\{-1,1\}^{m+n}$ and follows the RBM$(m, n, \boldsymbol \theta)$ distribution with parameter $\boldsymbol \theta\in \mathbb{R}^{m+n+m*n}$.
+
+Let $\bm{\theta}_1,\bm{\theta}_2 \in \mathbb{R}^{m+n+m*n}$ and
+let $P_i$ and $\E_i$ denote probability and expectation under $\bm{\theta}_i$, $i=1,2$. Then, the following are equivalent
+
+1. $\E_1 V_i=\E_2 V_i$, $\E_1 H_j=\E_2 H_j$ and $\E_1 V_i H_j =\E_2 V_i H_j$ for any $i=1,\ldots,m$, $j=1,\ldots,n$.
+2. $\bm{\theta}_1 =\bm{\theta}_2$.
+3. $P_1$ and  $P_2$ are the same distribution for $\boldsymbol X$.
+
+Additionally, the following three statements are equivalent
+
+1. $\E V_i=0$, $\E H_j=0$ and $\E V_i H_j =0$ for any $i=1,\ldots,m$, $j=1,\ldots,n$.
+2. $\bm{\theta} = \bm{0}\in \mathbb{R}^{m+n+m*n}$.
+3. $V_1,\ldots,V_m, H_1,\ldots,H_n$ are independent random variables assuming $-1/1$ with equal probability $1/2$.
+
+```
 
@@
 Let $\bm{\theta}_1,\bm{\theta}_2 \in \mathbb{R}^{m+n+mn}$ and
-let $P_i$ and $\E$ denote probability and expectation under $\bm{\theta}_i$, $i=1,2$. Then, the following are equivalent
+let $P_i$ and $\E_i$ denote probability and expectation under $\bm{\theta}_i$, $i=1,2$. Then, the following are equivalent
 
-1. $\E_1 V_i=\E_2 V_i$, $\E_1 H_j=\E_2 H_j$ and $\E_1 V_i H_j =\E_2 V_i H_j$ for any $i=1,\ldots,m$, $j=1,\ldots,n$.
+1. $\E_1 V_i=\E_2 V_i$, $\E_1 H_j=\E_2 H_j$ and $\E_1 V_i H_j =\E_2 V_i H_j$ for all $i=1,\ldots,m$, $j=1,\ldots,n$.
 2. $\bm{\theta}_1 =\bm{\theta}_2$.
-3. $P_1$ and  $P_2$ are the same distribution for  $(\bm{V}_m,\bm{H}_n)$.\\
+3. $P_1$ and  $P_2$ are the same distribution for  $(\bm{V}_m,\bm{H}_n)$.
 
@@
 If exactly two specified probabilities in (\@ref(eq:0)) are zero, say  $p_{(1,1)}$ and $p_{(-1,1)}$ (without loss of generality), then a limit approximation as in (\@ref(eq:2)) follows by picking $\theta_{v_1}$ as above based on any choices of $(\theta_{h_1}, \Delta_1,\Delta_2)$ and choosing  a sequence of $\theta_{v_2}\equiv \theta_{v_2}^{(m)}$ values for which $\theta_{v_2}^{({m})} \rightarrow -\infty$.
-
-# References {-}

### 2018-09-30 writing/paper-rbm/paper.Rmd
revisions

@@
 $$
-\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] = \sum\limits_{\boldsymbol x \in \mathcal{C}^{\nv + \nh}} \boldsymbol x f_{\boldsymbol \theta}(\boldsymbol x) = \sum\limits_{\boldsymbol x \in \mathcal{C}^{\nv + \nh}} \boldsymbol x \frac{\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{C}^{\nv + \nh}}\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}
+\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] = \sum\limits_{\boldsymbol x \in \mathcal{C}^{\nv + \nh}} \boldsymbol x \frac{\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{C}^{\nv + \nh}}\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}
 $$
@@
 ```{r uninterp, fig.cap="\\label{fig:uninterp}The absolute difference between coordinates of E$\\left[\\boldsymbol X | \\boldsymbol \\theta\\right]$ model expectations and E$\\left[\\boldsymbol X | \\boldsymbol \\theta^* \\right ]$ expectations given independence for an RBM with one visible and one hidden node.
-As an indicator of uninterpretability, note that differences in expectations increase as$\\theta_{11}$, the dependence parameter, deviates from zero.", fig.height = 4, fig.align='center', message=FALSE, cache=TRUE, warning=FALSE}
+As an indicator of uninterpretability, note that differences in expectations increase as $\\theta_{11}$, the dependence parameter, deviates from zero.", fig.height = 4, fig.align='center', message=FALSE, cache=TRUE, warning=FALSE}
 expand.grid(theta_h1 = seq(-5, 5, length.out = 21),
@@
 The $-1/1$ data encoding has the benefit of providing a guaranteed-to-be non-degenerate model at $\boldsymbol \theta = \boldsymbol 0 \in \mathbb{R}^{\nh + \nv + \nh*\nv}$, where the zero vector then serves as the natural center of the parameter space and induces the simplest possible model properties for the RBM (i.e., at $\boldsymbol \theta = \boldsymbol 0$, all variables are independent uniformly distributed on $\{-1,1\}^\nv$).
-The proof of this and further exploration of the equivalence of the $\boldsymbol \theta$ parameterization of the RBM model class and parameterization by $\boldsymbol \mu(\boldsymbol \theta)$ is in the on-line supplementary materials.
+The proof of this and further relations between the $-1/1$ data encoding and the $\boldsymbol \theta$ parameterization of the RBM model class are given in the on-line supplementary materials.
+Based on this material, Proposition \ref{prp:three} also briefly summarizes how the $-1/1$ data encoding induces some attractive theoretical properties in characterizing RBM models.
+
+\begin{proposition}
+\label{prp:three}
+Let $\nv, \nh \geq 1$ and suppose the random vector $X =(V_1,...,V_{\nv},H_1,...,H_{\nh})$ assumes values in $\{-1,1\}^{\nv+\nh}$ and follows the RBM$(\nv, \nh, \boldsymbol \theta)$ distribution \eqref{eq:pmf} with parameter $\boldsymbol \theta\in \mathbb{R}^{\nv+\nh+\nv*\nh}$.
+Let $\bm{\theta}_1,\bm{\theta}_2 \in \mathbb{R}^{\nv+\nh+\nv*\nh}$ and let $P_i$ and $\E_i$ denote the full joint probability and expectation under $\bm{\theta}_i$, $i=1,2$.
+Then, the following are equivalent
+\begin{enumerate}
+\setlength{\itemsep}{0pt}
+\setlength{\parskip}{0pt}
+\setlength{\parsep}{0pt}
+\item $\E_1 V_i=\E_2 V_i$, $\E_1 H_j=\E_2 H_j$ and $\E_1 V_i H_j =\E_2 V_i H_j$ for all $i=1,\ldots, \nv$, $j=1,\ldots,\nh$.
+\item $\bm{\theta}_1 =\bm{\theta}_2$.
+\item $P_1$ and  $P_2$ are the same distribution for $\boldsymbol X$.
+\end{enumerate}
+Additionally, the following three statements are equivalent
+\begin{enumerate}
+\setlength{\itemsep}{0pt}
+\setlength{\parskip}{0pt}
+\setlength{\parsep}{0pt}
+\item $\E V_i=0$, $\E H_j=0$ and $\E V_i H_j =0$ for all $i=1,\ldots,\nv$, $j=1,\ldots,\nh$.
+\item $\bm{\theta} = \bm{0}\in \mathbb{R}^{\nv+\nh+\nv*\nh}$.
+\item $V_1,\ldots,V_\nv, H_1,\ldots,H_\nh$ are independent random variables assuming $-1/1$ with equal probability $1/2$.
+\end{enumerate}
+\end{proposition}
+
 Hence, while from some computing perspectives $0/1$ coding might seem most natural, the $-1/1$ coding is far more convenient and interpretable from the point of view of statistical modeling, where it makes parameters simply interpreted in terms of symmetrically defined main effects and interactions.
@@
 
-For simplicity, consider a model with two visible variables $(V_1,V_2)$ and one hidden $H_1$ so that there are four possible data realizations for $(V_1,V_2)$  given by $(\pm 1, \pm 1)$, and we may express the model probabilities as
+For simplicity, consider a model with two visible variables $(V_1,V_2)$ and one hidden $H_1$ so that there are four possible data realizations for $(V_1,V_2)$  given by $(\pm 1, \pm 1)$, and we may express the marginalized model probabilities as
 $$
@@
 \end{equation} 
-for the outcomes $(\pm 1, \pm 1)$, the online appendix provides technical details to show parameter values $(\theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21})$ may be chosen to match any given cell probabilities with arbitrary closeness.
+for the outcomes $(\pm 1, \pm 1)$, the online appendix provides technical details to show parameter values $(\theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11}, \theta_{21})$ may be chosen to match any given cell probabilities with arbitrary closeness.
 In fact, when the cell probabilities \eqref{eq:0} are all strictly positive, parameters in the RBM model can be specified to reflect these probabilities exactly.
-And, when one or more of the cell probabilities \eqref{eq:0} are zero, the corresponding RBM probabilities $P(V_1=v_1,V_2=_v2| \theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21})$ may never be identically zero (due to exponential terms in the model) but parameters still can be selected to make the appropriate RBM cell probabilities arbitrarily small.
+And, when one or more of the cell probabilities \eqref{eq:0} are zero, the corresponding marginalized RBM probabilities $P(V_1=v_1,V_2=_v2| \theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21})$ may never be identically zero (due to exponential terms in the model) but parameters still can be selected to make the appropriate RBM cell probabilities arbitrarily small.
 Furthermore, as development in the online appendix shows, not only is it possible to approximate any distribution on the visibles arbitrarily well [cf.
@@
 For illustration, considering the simple example from above with $n$ iid observations, each consisting of two realized visibles $(V_1,V_2)$ the empirical cell frequencies from the sample are the highest likelihood choices for the cell probabilities $p_{(-1,-1)},p_{(1,-1)}, p_{(-1,1)},p_{(1,1)}$ in \eqref{eq:0} before any parametric model is introduced for improved analysis, and the discussion above indicates that RBM model parameters can be chosen to re-create this empirical distribution to an arbitrarily close degree.
-Not only does RBM model fitting based on ML seek to reproduce the empirical distribution, whenever this empirical distribution contains empty cells, fitting steps for the RBM model further will aim to choose parameters that necessarily diverge to infinity in magnitude in order to zero-out the corresponding RBM cell probabilities.
+Not only does RBM model fitting based on ML seek to reproduce the empirical distribution, whenever this empirical distribution contains empty cells, fitting steps for the RBM model further will aim to choose parameters that necessarily diverge to infinity in magnitude in order to zero-out the corresponding marginalized RBM cell probabilities.
 In data applications with a large sample space, it is unlikely that the training set will include at least one of each possible vector outcome (unlike this small example).
@@
 
-Ultimately, it is not unquestionably clear that RBM models are useful as generative models.
+<!-- Ultimately, it is not unquestionably clear that RBM models are useful as generative models.
 As a concern, without appropriate generative behavior in an RBM model, the uncertainty in estimated model parameters becomes impossible to assess realistically and further, the model can lose useful application to prediction problems (e.g., realistic predictive distributions).
-In the case of classification, predictive distributions may not be the ultimate goal, but when S-instability is present, small (imperceptible) differences in the data may lead to greatly different probabilities and thus greatly different classifications.
+In the case of classification, predictive distributions may not be the ultimate goal.
+While interpretability may not always be of primary importance, we find that the issues of stability and degeneracy go hand in hand with interpretability, and these should not be ignored.
+In fact, when S-instability is present, small (imperceptible) differences in the data may lead to greatly different probabilities and thus greatly different classifications.
 For these reasons, we are skeptical about RBMs as probabilistic conceptualizations for data generation.
+-->
+
+While interpretability of parameters may not always be of primary importance in application of an RBM model (e.g.
+in some applications to prediction/classification), we find that the issues of stability and degeneracy go hand in hand with those of interpretability, and the former should not be ignored.
+When S-instability is present, small (imperceptible) differences in training data and corresponding fitted models may lead to greatly different probabilities (and thus substantially different classifications).
+Ultimately, it is not clear that RBM models are typically useful as generative models for observables.
+Without appropriate generative behavior in an RBM model, uncertainty in estimated model parameters becomes impossible to realistically assess.
+Further, the model cannot support other inferences important in many prediction problems (e.g., realistic predictive distributions).
+For these reasons, we are skeptical about RBMs as widely useful probabilistic constructs, at least in their present incarnation.
 

### 2018-09-30 writing/paper-rbm/paper.tex
revisions

@@
 compare in the RBM case are \[
-\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] = \sum\limits_{\boldsymbol x \in \mathcal{C}^{\nv + \nh}} \boldsymbol x f_{\boldsymbol \theta}(\boldsymbol x) = \sum\limits_{\boldsymbol x \in \mathcal{C}^{\nv + \nh}} \boldsymbol x \frac{\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{C}^{\nv + \nh}}\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}
+\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] = \sum\limits_{\boldsymbol x \in \mathcal{C}^{\nv + \nh}} \boldsymbol x \frac{\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{C}^{\nv + \nh}}\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}
 \] and \begin{align*}
@@
 \caption{\label{fig:uninterp}The absolute difference between coordinates of E$\left[\boldsymbol X | \boldsymbol \theta\right]$ model expectations and E$\left[\boldsymbol X | \boldsymbol \theta^* \right ]$ expectations given independence for an RBM with one visible and one hidden node.
-As an indicator of uninterpretability, note that differences in expectations increase as$\theta_{11}$, the dependence parameter, deviates from zero.}\label{fig:uninterp}
+As an indicator of uninterpretability, note that differences in expectations increase as $\theta_{11}$, the dependence parameter, deviates from zero.}\label{fig:uninterp}
 \end{figure}
@@
 independent uniformly distributed on \(\{-1,1\}^\nv\)). The proof of
-this and further exploration of the equivalence of the
-\(\boldsymbol \theta\) parameterization of the RBM model class and
-parameterization by \(\boldsymbol \mu(\boldsymbol \theta)\) is in the
-on-line supplementary materials. Hence, while from some computing
-perspectives \(0/1\) coding might seem most natural, the \(-1/1\) coding
-is far more convenient and interpretable from the point of view of
-statistical modeling, where it makes parameters simply interpreted in
-terms of symmetrically defined main effects and interactions. Under the
-data encoding \(-1/1\), the parameter space centered at 0 is also
-helpful for framing parameter configurations that are undesirably large
-(i.e., these are naturally parameters that have moved too far away from
-\(\bm{0}\) where the RBM model is anchored to be trivially describable
-and completely problem-free). In light of all of these matters we will
-henceforth employ the \(-1/1\) coding.
+this and further relations between the \(-1/1\) data encoding and the
+\(\boldsymbol \theta\) parameterization of the RBM model class are given
+in the on-line supplementary materials. Based on this material,
+Proposition \ref{prp:three} also briefly summarizes how the \(-1/1\)
+data encoding induces some attractive theoretical properties in
+characterizing RBM models.
+
+\begin{proposition}
+\label{prp:three}
+Let $\nv, \nh \geq 1$ and suppose the random vector $X =(V_1,...,V_{\nv},H_1,...,H_{\nh})$ assumes values in $\{-1,1\}^{\nv+\nh}$ and follows the RBM$(\nv, \nh, \boldsymbol \theta)$ distribution \eqref{eq:pmf} with parameter $\boldsymbol \theta\in \mathbb{R}^{\nv+\nh+\nv*\nh}$.
+Let $\bm{\theta}_1,\bm{\theta}_2 \in \mathbb{R}^{\nv+\nh+\nv*\nh}$ and let $P_i$ and $\E_i$ denote the full joint probability and expectation under $\bm{\theta}_i$, $i=1,2$.
+Then, the following are equivalent
+\begin{enumerate}
+\setlength{\itemsep}{0pt}
+\setlength{\parskip}{0pt}
+\setlength{\parsep}{0pt}
+\item $\E_1 V_i=\E_2 V_i$, $\E_1 H_j=\E_2 H_j$ and $\E_1 V_i H_j =\E_2 V_i H_j$ for all $i=1,\ldots, \nv$, $j=1,\ldots,\nh$.
+\item $\bm{\theta}_1 =\bm{\theta}_2$.
+\item $P_1$ and  $P_2$ are the same distribution for $\boldsymbol X$.
+\end{enumerate}
+Additionally, the following three statements are equivalent
+\begin{enumerate}
+\setlength{\itemsep}{0pt}
+\setlength{\parskip}{0pt}
+\setlength{\parsep}{0pt}
+\item $\E V_i=0$, $\E H_j=0$ and $\E V_i H_j =0$ for all $i=1,\ldots,\nv$, $j=1,\ldots,\nh$.
+\item $\bm{\theta} = \bm{0}\in \mathbb{R}^{\nv+\nh+\nv*\nh}$.
+\item $V_1,\ldots,V_\nv, H_1,\ldots,H_\nh$ are independent random variables assuming $-1/1$ with equal probability $1/2$.
+\end{enumerate}
+\end{proposition}
+
+Hence, while from some computing perspectives \(0/1\) coding might seem
+most natural, the \(-1/1\) coding is far more convenient and
+interpretable from the point of view of statistical modeling, where it
+makes parameters simply interpreted in terms of symmetrically defined
+main effects and interactions. Under the data encoding \(-1/1\), the
+parameter space centered at 0 is also helpful for framing parameter
+configurations that are undesirably large (i.e., these are naturally
+parameters that have moved too far away from \(\bm{0}\) where the RBM
+model is anchored to be trivially describable and completely
+problem-free). In light of all of these matters we will henceforth
+employ the \(-1/1\) coding.
 
@@
 data realizations for \((V_1,V_2)\) given by \((\pm 1, \pm 1)\), and we
-may express the model probabilities as \[
+may express the marginalized model probabilities as \[
 P(V_1=v_1,V_2=v_2| \theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21}) \propto  \exp\left( v_1  \theta_{v_1} + v_2 \theta_{v_2}\right) \sum_{h \in \{\pm 1\}}\exp\left( h[ \theta_{h_1}  + \theta_{11} v_1 + \theta_{21} v_2] \right),
@@
 provides technical details to show parameter values
-\((\theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21})\)
+\((\theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11}, \theta_{21})\)
 may be chosen to match any given cell probabilities with arbitrary
@@
 reflect these probabilities exactly. And, when one or more of the cell
-probabilities \eqref{eq:0} are zero, the corresponding RBM probabilities
+probabilities \eqref{eq:0} are zero, the corresponding marginalized RBM
+probabilities
 \(P(V_1=v_1,V_2=_v2| \theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21})\)
@@
 parameters that necessarily diverge to infinity in magnitude in order to
-zero-out the corresponding RBM cell probabilities. In data applications
-with a large sample space, it is unlikely that the training set will
-include at least one of each possible vector outcome (unlike this small
-example). This implies that some RBM model parameters must diverge to
-\(+\infty\) to mimic the empirical distribution with empty cells and, as
-we have already discussed in Section
+zero-out the corresponding marginalized RBM cell probabilities. In data
+applications with a large sample space, it is unlikely that the training
+set will include at least one of each possible vector outcome (unlike
+this small example). This implies that some RBM model parameters must
+diverge to \(+\infty\) to mimic the empirical distribution with empty
+cells and, as we have already discussed in Section
 \ref{explorations-of-model-properties-through-simulation}, large
@@
 
-Ultimately, it is not unquestionably clear that RBM models are useful as
-generative models. As a concern, without appropriate generative behavior
-in an RBM model, the uncertainty in estimated model parameters becomes
-impossible to assess realistically and further, the model can lose
-useful application to prediction problems (e.g., realistic predictive
-distributions). In the case of classification, predictive distributions
-may not be the ultimate goal, but when S-instability is present, small
-(imperceptible) differences in the data may lead to greatly different
-probabilities and thus greatly different classifications. For these
-reasons, we are skeptical about RBMs as probabilistic conceptualizations
-for data generation.
+While interpretability of parameters may not always be of primary
+importance in application of an RBM model (e.g.~in some applications to
+prediction/classification), we find that the issues of stability and
+degeneracy go hand in hand with those of interpretability, and the
+former should not be ignored. When S-instability is present, small
+(imperceptible) differences in training data and corresponding fitted
+models may lead to greatly different probabilities (and thus
+substantially different classifications). Ultimately, it is not clear
+that RBM models are typically useful as generative models for
+observables. Without appropriate generative behavior in an RBM model,
+uncertainty in estimated model parameters becomes impossible to
+realistically assess. Further, the model cannot support other inferences
+important in many prediction problems (e.g., realistic predictive
+distributions). For these reasons, we are skeptical about RBMs as widely
+useful probabilistic constructs, at least in their present incarnation.
 
@@
 Kaplan, Andee, Daniel Nordman, and Stephen Vardeman. 2017. ``On the
-Propriety of Restricted Boltzmann Machines.'' \emph{Unpublished}.
+Instability and Degeneracy of Deep Learning Models.'' \emph{arXiv
+Preprint arXiv:1612.01159}.
 

### 2019-06-11 writing/paper-note/chapters/instab.Rmd
getting ready for another submission
started edits

@@
 That is, $\mathcal{X}$ with $|\mathcal{X}| < \infty$ denotes the set of potential outcomes for each single variable $X_i$, so that the product space $\mathcal{X}^N$ corresponds to values for the variables $\boldsymbol X_N=(X_1,\ldots,X_N)$.
-For each $N$, let $P_{\boldsymbol \theta_N}$ denote a probability model on $\mathcal{X}^N$, under which $P_{\boldsymbol \theta_N}(x_1, \dots, x_N) > 0$ is the probability of the data outcome $(x_1, \dots, x_N) \in \mathcal{X}^N$.
-In this, we assume that the model support of $P_{\boldsymbol \theta_N}$ is the sample space $\mathcal{X}^N$.
-This framework produces probability models $P_{\boldsymbol \theta_N}$, indexed by a generic sequence of parameters $\boldsymbol \theta_N$, to describe data $\boldsymbol X_N$ of any given sample size $N \geq 1$.
+For each $N$, let $P_{\ak{\thetaN}}$ denote a probability model on $\mathcal{X}^N$, under which $P_{\ak{\thetaN}}(x_1, \dots, x_N) > 0$ is the probability of the data outcome $(x_1, \dots, x_N) \in \mathcal{X}^N$.
+In this, we assume that the model support of $P_{\ak{\thetaN}}$ is the sample space $\mathcal{X}^N$.
+This framework produces probability models $P_{\ak{\thetaN}}$, indexed by a \ak{defining} sequence of parameters $\ak{\thetaN}$, to describe data $\boldsymbol X_N$ of any given sample size $N \geq 1$.
 For simplicity, we will refer to this distributional class as *Finite Outcome Everywhere Supported (FOES)* models in the following.
-The dimension and structure of such parameters are generic, without restriction, though natural cases will be seen to include those where $\boldsymbol \theta_N \in \mathbb{R}^{q(N)}$ for some arbitrary integer-valued function $q(\cdot) \geq 1$.
+The dimension and structure of such parameters are generic, without restriction, though natural cases will be seen to include those where $\ak{\thetaN} \in \mathbb{R}^{q(N)}$ for some arbitrary integer-valued function $q(\cdot) \geq 1$.
 
@@
 
-Many model families fall under the umbrella of FOES models. For illustration, this section presents three specific examples of FOES models, including models with deep architectures.
+Many model families \ak{lead to} FOES models.
+\ak{(For clarity, we use ``model family" to refer to generic class or functional form of parametric distributions, while "model" often refers here to a member of the class for an instantiation or selection of parameters.
+This distinction is not so important for the examples presented in Sections \ref{discrete-exponential-family-models}-\ref{deep-learning}, but becomes more so in Section \ref{instability-results} where model instability depends on both form and a given parameter configuration, rather than just model class form.)} For illustration, this section presents three specific examples of FOES model \ak{classes}, including models with deep architectures.
 
@@
 
-For random variables $\boldsymbol X \equiv\boldsymbol X_N= (X_1, \dots, X_N)$ with sample space $\mathcal{X}^N$, $|\mathcal{X}| < \infty$, consider an exponential family model for $\boldsymbol X$ with probability mass function given by
+For random variables $\boldsymbol X \equiv\boldsymbol X_N= (X_1, \dots, X_N)$ with sample space $\mathcal{X}^N$, $|\mathcal{X}| < \infty$, consider an exponential \ak{model family} for $\boldsymbol X$ with probability mass function given by
 \begin{equation}
@@
 \end{equation}
-depending on parameter vector $\boldsymbol \theta \in \Theta_N \subset \mathbb{R}^{k}$ and natural parameter function $\boldsymbol \eta : \mathbb{R}^k \mapsto \mathbb{R}^L$ with fixed positive integers $k$ and $L$ denoting their dimensions.
+depending on parameter vector $\boldsymbol \theta \in \Theta_{\ak{\thetaidx}} \subset \mathbb{R}^{k}$ and natural parameter function $\boldsymbol \eta : \mathbb{R}^k \mapsto \mathbb{R}^L$ with fixed positive integers $k$ and $L$ denoting their dimensions.
 Above, $\boldsymbol g_N : \mathcal{X}^N \mapsto \mathbb{R}^L$ is a vector of sufficient statistics, while
 $$
-\psi(\boldsymbol \theta) = \log \sum\limits_{\boldsymbol x \in \mathcal{X}^N}\exp\left[\boldsymbol \eta^T(\boldsymbol \theta) \boldsymbol g_N(\boldsymbol x) \right], \qquad \boldsymbol \theta \in \Theta_N\equiv \{\boldsymbol \theta \in \mathbb{R}^k : \psi(\boldsymbol \theta) < \infty \},
+\psi(\boldsymbol \theta) = \log \sum\limits_{\boldsymbol x \in \mathcal{X}^N}\exp\left[\boldsymbol \eta^T(\boldsymbol \theta) \boldsymbol g_N(\boldsymbol x) \right], \qquad \boldsymbol \theta \in \Theta_\ak{\thetaidx}\equiv \{\boldsymbol \theta \in \mathbb{R}^k : \psi(\boldsymbol \theta) < \infty \},
 $$
-denotes the normalizing function with parameter space $\Theta_N$.
+denotes the normalizing function with parameter space $\Theta_{\ak{\thetaidx}}$.
 The natural parameter function $\eta (\boldsymbol \theta)$ has a linear form (i.e., $\eta (\boldsymbol \theta)= \bm{A} \boldsymbol \theta$ for a given $L \times k$ matrix $\bm{A}$) in many common model formulations, though may also be nonlinear (e.g., curved exponential families).
@@
 
-Such discrete exponential family models are special cases of the FOES models, as seen by  defining $P_{\boldsymbol \theta_N}(\boldsymbol x)\equiv p_{N,\boldsymbol \theta_N}(\boldsymbol x)> 0$,  $\boldsymbol x \in \mathcal{X}^N$, based on \@ref(eq:expo) and a parameter sequence $\boldsymbol \theta_N \in \Theta_N \subset \mathbb{R}^k$.
+Such discrete \ak{exponential models} are special cases of the FOES models, as seen by  defining $P_{\ak{\thetaN}}(\boldsymbol x)\equiv p_{N, \ak{\thetaN}}(\boldsymbol x)> 0$,  $\boldsymbol x \in \mathcal{X}^N$, based on \@ref(eq:expo) and a parameter sequence $\ak{\thetaN} \in \Theta_{\ak{\thetaidx}} \subset \mathbb{R}^k$.
 For example, if observations $\boldsymbol X = (X_1,\ldots,X_N)$ correspond to $N$ independent and identically distributed Bernoulli random variables, each indicating a binary $0$-$1$ outcome,  the resulting probabilities have exponential form \@ref(eq:expo)  given by
@@
 (\#eq:mod1)
-P_{\boldsymbol \theta_N}(\boldsymbol x) \propto
- \exp\left[\boldsymbol \theta_N \sum_{i=1}^N x_i\right], \qquad \boldsymbol x=(x_1,\ldots,x_N) \in\{0,1\}^N, 
+P_{\ak{\thetaN}}(\boldsymbol x) \propto
+ \exp\left[\ak{\thetaN} \sum_{i=1}^N x_i\right], \qquad \boldsymbol x=(x_1,\ldots,x_N) \in\{0,1\}^N, 
  \end{equation} 
- with sufficient statistic $\boldsymbol g_N(\boldsymbol x)\equiv \sum_{i=1}^N  x_i$ and "log odds ratio" parameter $\boldsymbol \theta_N \equiv  \log[ P_{\boldsymbol \theta_N}(X_i=1)/P_{\boldsymbol \theta_N}(X_i=0) ] \in \mathbb{R}$.
+ with sufficient statistic $\boldsymbol g_N(\boldsymbol x)\equiv \sum_{i=1}^N  x_i$ and "log odds ratio" parameter $\ak{\thetaN} \equiv  \log[ P_{\ak{\thetaN}}(X_i=1)/P_{\ak{\thetaN}}(X_i=0) ] \in \mathbb{R}$.
 More generally, supposing $\boldsymbol X =(X_1,\ldots,X_N)$ represent $N$ independent trials, each assuming an outcome $\{1,\ldots,k\}$ among $k$ possibilities (e.g., a die roll), a multinomial distribution is given by
@@
 (\#eq:mod11)
-P_{\boldsymbol \theta_N}(\boldsymbol x) \propto  \exp\left[  \boldsymbol \theta_{N}^T g_N(\boldsymbol x)   \right] =
-\exp\left[ \sum_{j=1}^k {\theta_{j,N}} \sum_{i=1}^N \mathbb{I}(x_i=j) \right], \qquad \boldsymbol x  \in\{1,\ldots,k\}^N, 
+P_{\ak{\thetaN}}(\boldsymbol x) \propto  \exp\left[  \ak{\thetaN}^T g_N(\boldsymbol x)   \right] =
+\exp\left[ \sum_{j=1}^k {\theta_{j,\ak{\thetaidx}}} \sum_{i=1}^N \mathbb{I}(x_i=j) \right], \qquad \boldsymbol x  \in\{1,\ldots,k\}^N, 
 \end{equation}
-with sufficient statistic $\boldsymbol g_N(\boldsymbol x)$ involving a count $\sum_{i=1}^N \mathbb{I}(x_i=j)$ for each outcome $j \in \{1,\ldots,k\}$, where $\mathbb{I}(\cdot)$ denotes the indicator function, and parameters $\boldsymbol \theta_N=(\theta_{1,N},\ldots,\theta_{k,N})\in\mathbb{R}^k$ defining log-probability ratios $\theta_{i,N}-\theta_{j,N} =\log [P_{\boldsymbol \theta_N}(X_1=i)/P_{\boldsymbol \theta_N}(X_1=j)]$.
+with sufficient statistic $\boldsymbol g_N(\boldsymbol x)$ involving a count $\sum_{i=1}^N \mathbb{I}(x_i=j)$ for each outcome $j \in \{1,\ldots,k\}$, where $\mathbb{I}(\cdot)$ denotes the indicator function, and parameters $\ak{\thetaN}=(\theta_{1,\ak{\thetaidx}},\ldots,\theta_{k,\ak{\thetaidx}})\in\mathbb{R}^k$ defining log-probability ratios $\theta_{i,\ak{\thetaidx}}-\theta_{j,\ak{\thetaidx}} =\log [P_{\ak{\thetaN}}(X_1=i)/P_{\ak{\thetaN}}(X_1=j)]$.
 In addition to such standard models for discrete independent data, exponential models of FOES type commonly arise with dependent spatial data [@besag1974spatial] and network/relational data [@wasserman1994social; @handcock2003assessing].
@@
 (\#eq:mod2)
-P_{\boldsymbol \theta_N}(\boldsymbol x) \propto
- \exp\left[\sum_{j=1}^3 \theta_{j,N} g_{j,N}(\boldsymbol x)\right], \quad\qquad \boldsymbol x=(x_1,\ldots,x_N)  \in\{0,1\}^N,
+P_{\ak{\thetaN}}(\boldsymbol x) \propto
+ \exp\left[\sum_{j=1}^3 \theta_{j,\ak{\thetaidx}} g_{j,N}(\boldsymbol x)\right], \quad\qquad \boldsymbol x=(x_1,\ldots,x_N)  \in\{0,1\}^N,
 \end{equation}
@@
 $$
-involving the numbers of edges, 2-stars and triangles among an outcome $\boldsymbol x$ given by $g_{1,N}(\boldsymbol x)$, $g_{2,N}(\boldsymbol x)$ and $g_{3,N}(\boldsymbol x)$, respectively, along with $k=3$ real parameters $\boldsymbol \theta_N \equiv (\theta_{1,N},\theta_{2,N},\theta_{3,N})$.
-For this network model \@ref(eq:mod2) in particular, as well as for more general models of form \@ref(eq:expo),  @schweinberger2011instability considered instability in such exponential models with sequences of fixed parameters $\boldsymbol \theta_N = (\theta_1,\ldots,\theta_k)\in\mathbb{R}^k$, $N \geq 1$, of fixed dimension $k$.
+involving the numbers of edges, 2-stars and triangles among an outcome $\boldsymbol x$ given by $g_{1,N}(\boldsymbol x)$, $g_{2,N}(\boldsymbol x)$ and $g_{3,N}(\boldsymbol x)$, respectively, along with $k=3$ real parameters $\ak{\thetaN} \equiv (\theta_{1,\ak{\thetaidx}},\theta_{2,\ak{\thetaidx}},\theta_{3,\ak{\thetaidx}})$.
+For this network model \@ref(eq:mod2) in particular, as well as for more general models of form \@ref(eq:expo),  @schweinberger2011instability considered instability in such exponential models with sequences of fixed parameters $\ak{\thetaN} = (\theta_1,\ldots,\theta_k)\in\mathbb{R}^k$, $N \geq 1$, of fixed dimension $k$.
 
-For model sequences $P_{\boldsymbol \theta_N}(\boldsymbol x)\equiv p_{N,\boldsymbol \theta_N}(\boldsymbol x)$ of the exponential type \@ref(eq:expo), such as those in \@ref(eq:mod1)-\@ref(eq:mod2), note that the dimension $k$ of the parameter $\boldsymbol \theta_N\in\Theta \subset \mathbb{R}^k$ necessarily remains the same for all sample sizes $N \geq 1$ as the form of the natural parameter function $\eta(\cdot)$ in \@ref(eq:expo) and the number of sufficient statistics $\boldsymbol g_{N}(\boldsymbol x)$ do not depend on $N$.
-Consequently, $\boldsymbol \theta_N$ lies in a parameter space of fixed Euclidean dimension $k$.
+For model sequences $P_{\ak{\thetaN}}(\boldsymbol x)\equiv p_{N,\ak{\thetaN}}(\boldsymbol x)$ of the exponential type \@ref(eq:expo), such as those in \@ref(eq:mod1)-\@ref(eq:mod2), note that the dimension $k$ of the parameter $\ak{\thetaN}\in\Theta \subset \mathbb{R}^k$ necessarily remains the same for all sample sizes $N \geq 1$ as the form of the natural parameter function $\eta(\cdot)$ in \@ref(eq:expo) and the number of sufficient statistics $\boldsymbol g_{N}(\boldsymbol x)$ do not depend on $N$.
+Consequently, $\ak{\thetaN}$ lies in a parameter space of fixed Euclidean dimension $k$.
 However, this aspect need not be true for other types of FOES models considered in Sections \@ref(rbm) - \@ref(deep-learning), where instead the numbers of parameters and sufficient statistics commonly increase with the sample size $N$.
@@
 Let $\boldsymbol X = (X_1,\ldots,X_N)$ denote the $N$ random variables for visibles with support $\mathcal{X}^N$ and $\boldsymbol H = (H_1,\ldots,H_{N_\mathcal{H}})$ denote the $N_\mathcal{H}$ random variables for hiddens with support $\mathcal{X}^{N_\mathcal{H}}$ where $\mathcal{X} = \{-1,1\}$.
-For parameters $\boldsymbol \theta_N^{\mathcal{H}} \in \mathbb{R}^{N_\mathcal{H}}$, $\boldsymbol \theta_N^{\mathcal{V}}\in \mathbb{R}^N$, and $\boldsymbol \theta_N^{\mathcal{HV}}$ as a real matrix with dimension $N_\mathcal{H} \times N$, the RBM model for $\tilde{\boldsymbol X}=(\boldsymbol X,\boldsymbol H)$  has the joint probability mass function
+For \ak{specified} parameters $\ak{\thetaN}^{\mathcal{H}} \in \mathbb{R}^{N_\mathcal{H}}$, $\ak{\thetaN}^{\mathcal{V}}\in \mathbb{R}^N$, and $\ak{\thetaN}^{\mathcal{HV}}$ as a real matrix with dimension $N_\mathcal{H} \times N$, the RBM model for $\tilde{\boldsymbol X}=(\boldsymbol X,\boldsymbol H)$  has the joint probability mass function
 \begin{equation}
 (\#eq:RBM1)
-\tilde{P}_{\boldsymbol \theta_N} (\tilde{\boldsymbol x}) = \exp\left[ (\boldsymbol \theta_N^{\mathcal{H}})^T \boldsymbol h + \boldsymbol (\boldsymbol \theta_N^{\mathcal{V}})^T \boldsymbol x + \boldsymbol h^T  \boldsymbol\theta_N^{\mathcal{HV}} \boldsymbol x - \psi(\boldsymbol \theta_N)\right], \quad \tilde{\boldsymbol x} = (\boldsymbol x, \boldsymbol h) \in \{\pm 1\}^{N+N_\mathcal{H}}
+\tilde{P}_{\ak{\thetaN}} (\tilde{\boldsymbol x}) = \exp\left[ (\ak{\thetaN}^{\mathcal{H}})^T \boldsymbol h + \boldsymbol (\ak{\thetaN}^{\mathcal{V}})^T \boldsymbol x + \boldsymbol h^T  \ak{\thetaN}^{\mathcal{HV}} \boldsymbol x - \psi(\ak{\thetaN})\right], \quad \tilde{\boldsymbol x} = (\boldsymbol x, \boldsymbol h) \in \{\pm 1\}^{N+N_\mathcal{H}}
 \end{equation}
@@
 $$
-\psi(\boldsymbol \theta_N) = \log \sum_{\tilde{\boldsymbol x} \in \{\pm 1\}^{N+N_H} } \exp\left[ (\boldsymbol \theta_N^{\mathcal{H}})^T \boldsymbol h + \boldsymbol (\boldsymbol \theta_N^{\mathcal{V}})^T \boldsymbol x + \boldsymbol h^T  \boldsymbol\theta_N^{\mathcal{HV}} \boldsymbol x\right].
+\psi(\ak{\thetaN}) = \log \sum_{\tilde{\boldsymbol x} \in \{\pm 1\}^{N+N_H} } \exp\left[ (\ak{\thetaN}^{\mathcal{H}})^T \boldsymbol h + \boldsymbol (\ak{\thetaN}^{\mathcal{V}})^T \boldsymbol x + \boldsymbol h^T  \ak{\thetaN}^{\mathcal{HV}} \boldsymbol x\right].
 $$
-Let $\boldsymbol \theta_N = (\boldsymbol \theta_N^{\mathcal{H}}, \boldsymbol \theta_N^{\mathcal{V}}, \boldsymbol\theta_N^{\mathcal{HV}} ) \in \Theta_N \equiv \mathbb{R}^{q(N)}$, with $q(N) = N + N_\mathcal{H} + N*N_\mathcal{H}$, denote the parameter vector for the RBM, as indexed by the number $N$ of visible random variables (which may differ from the actual lengths of these parameter vectors).
+Let $\ak{\thetaN} = (\ak{\thetaN}^{\mathcal{H}}, \ak{\thetaN}^{\mathcal{V}}, \ak{\thetaN}^{\mathcal{HV}} ) \in \Theta_\ak{\thetaidx} \equiv \mathbb{R}^{q(N)}$, with $q(N) = N + N_\mathcal{H} + N*N_\mathcal{H}$, denote \ak{a given} parameter vector for the RBM, as indexed by the number $N$ of visible random variables (which may differ from the actual lengths of these parameter vectors).
 The probability mass function for the visible variables $\boldsymbol X = (X_1, \dots, X_N)$ follows from marginalizing the joint specification to yield
@@
 (\#eq:RBM2)
-P_{\boldsymbol \theta_N} (\boldsymbol x) = \sum\limits_{\boldsymbol h \in \{\pm 1\}^{N_{\mathcal{H}}}} \tilde{P}_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h), \qquad \boldsymbol x \in \{\pm 1\}^{N}\equiv \mathcal{X}^N.
+P_{\ak{\thetaN}} (\boldsymbol x) = \sum\limits_{\boldsymbol h \in \{\pm 1\}^{N_{\mathcal{H}}}} \tilde{P}_{\ak{\thetaN}} (\boldsymbol x, \boldsymbol h), \qquad \boldsymbol x \in \{\pm 1\}^{N}\equiv \mathcal{X}^N.
 \end{equation}
-Here the baseline model \@ref(eq:RBM1) for hidden/visible variables is a linear exponential one in sufficient statistics $(\tilde{\boldsymbol X}, \boldsymbol X^T\boldsymbol H)$ using $\tilde{\boldsymbol X}=(\boldsymbol X,\boldsymbol H)$ from   \@ref(eq:RBM1), but the form differs from the previous exponential models in \@ref(eq:expo) in that the lengths of parameters $\boldsymbol \theta_N$ and statistics  $(\tilde{\boldsymbol X}, \boldsymbol X^T\boldsymbol H)$ increase to incorporate more visible variables.
-That is, in contrast to \@ref(eq:expo), the natural parameter function involved in the RBM model \@ref(eq:RBM1), as the identity mapping of the parameters $\boldsymbol \theta_N\in\mathbb{R}^{q(N)}$, naturally grows in  dimension  $q(N)\to \infty$ to accommodate visible variables $X_1, \dots, X_N$ of increasing sample size $N\to \infty$.
+Here the baseline model \@ref(eq:RBM1) for hidden/visible variables is a linear exponential one in sufficient statistics $(\tilde{\boldsymbol X}, \boldsymbol X^T\boldsymbol H)$ using $\tilde{\boldsymbol X}=(\boldsymbol X,\boldsymbol H)$ from   \@ref(eq:RBM1), but the form differs from the previous exponential models in \@ref(eq:expo) in that the lengths of parameters $\ak{\thetaN}$ and statistics  $(\tilde{\boldsymbol X}, \boldsymbol X^T\boldsymbol H)$ increase to incorporate more visible variables.
+That is, in contrast to \@ref(eq:expo), the natural parameter function involved in the RBM model \@ref(eq:RBM1), as the identity mapping of the parameters $\ak{\thetaN}\in\mathbb{R}^{q(N)}$, naturally grows in  dimension  $q(N)\to \infty$ to accommodate visible variables $X_1, \dots, X_N$ of increasing sample size $N\to \infty$.
 Additionally, one may further arbitrarily choose the number $N_\mathcal{H}$ of hidden variables $\boldsymbol H$ in the joint RBM model \@ref(eq:RBM1) to define a marginal model \@ref(eq:RBM2) for the $N$ visible variables $\boldsymbol X$, and the number $N_\mathcal{H}$ of hiddens may also potentially increase with $N$.
-Because $|\mathcal{X}| = 2$ and $P_{\boldsymbol \theta_N}(\boldsymbol x) > 0$ for all $\boldsymbol x \in \mathcal{X}^N$, the RBM  specification \@ref(eq:RBM2) for visibles $\boldsymbol$ corresponds to a FOES model, while the joint distribution \@ref(eq:RBM1) for $(\boldsymbol X, \boldsymbol H)$ is also a FOES model.
+Because $|\mathcal{X}| = 2$ and $P_{\ak{\thetaN}}(\boldsymbol x) > 0$ for all $\boldsymbol x \in \mathcal{X}^N$, the RBM  specification \@ref(eq:RBM2) for visibles $\boldsymbol$ corresponds to a FOES model, while the joint distribution \@ref(eq:RBM1) for $(\boldsymbol X, \boldsymbol H)$ is also a FOES model.
 As this example also indicates, any model formed by marginalizing a base FOES model class, such as the RBM joint specification \@ref(eq:RBM1), is again a FOES model.
@@
 
-Consider two models with "deep architecture" that contain multiple hidden (or latent) layers in addition to a visible layer of data, namely a deep Boltzmann machine [@salakhutdinov2009deep] and a deep belief network @hinton2006fast].
+Consider two models with "deep architecture" that contain multiple hidden (or latent) layers in addition to a visible layer of data, namely a deep Boltzmann machine [@salakhutdinov2009deep] and a deep belief network [@hinton2006fast].
 Let $M$ denote the number of hidden layers included in the model and let $N_{(H,1)}, \dots, N_{(H,M)}$ denote the numbers of hidden variables within each hidden layer.
-Then the random vector $\tilde{\boldsymbol X} = \{H^{(1)}_1, \dots, H^{(1)}_{N_{(H,1)}}, \dots, H^{(M)}_1, \dots, H^{(M)}_{N_{(H,M)}}, \boldsymbol X\}$ collects both the hidden variables
-$\{ H_{i}^{(j)} : i=1,\ldots, N_{(H,j)}, j=1,\ldots,M\}$ and visible variables $\boldsymbol X =(X_1,\ldots,X_N)$ in a deep probabilistic model.
+Then the random vector $\tilde{\boldsymbol X} = \{H^{(1)}_1, \dots, H^{(1)}_{N_{(H,1)}}, \dots, H^{(M)}_1, \dots, H^{(M)}_{N_{(H,M)}}, \boldsymbol X\}$ collects both the hidden variables $\{ H_{i}^{(j)} : i=1,\ldots, N_{(H,j)}, j=1,\ldots,M\}$ and visible variables $\boldsymbol X =(X_1,\ldots,X_N)$ in a deep probabilistic model.
 Each variable outcome will again lie in $\mathcal{X} = \{-1,1\}$.
@@
 $$
-\tilde{P}_{\boldsymbol \theta_N} ( \tilde{\boldsymbol x} ) = \exp\left[ \sum\limits_{i = 1}^M\boldsymbol \alpha^{(i)T} \boldsymbol h^{(i)} + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^{(1)T} \Gamma^{(0)} \boldsymbol x + \sum\limits_{i = 1}^{M - 1} \boldsymbol h^{(i)T} \Gamma^{(i)} \boldsymbol h^{(i + 1)} - \psi(\boldsymbol \theta_N) \right],
+\tilde{P}_{\ak{\thetaN}} ( \tilde{\boldsymbol x} ) = \exp\left[ \sum\limits_{i = 1}^M\boldsymbol \alpha^{(i)T} \boldsymbol h^{(i)} + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^{(1)T} \Gamma^{(0)} \boldsymbol x + \sum\limits_{i = 1}^{M - 1} \boldsymbol h^{(i)T} \Gamma^{(i)} \boldsymbol h^{(i + 1)} - \psi(\ak{\thetaN}) \right],
 $$ 
@@
 $$
-\psi(\boldsymbol \theta_N) = \log \sum\limits_{\tilde{\boldsymbol x} \in \mathcal{X}^{N_{(H,1)} + \cdots + N_{(H,M)} +N}} \exp\left[ \sum\limits_{i = 1}^M\boldsymbol \alpha^{(i)T} \boldsymbol h^{(i)} + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^{(1)T} \Gamma^{(0)} \boldsymbol x + \sum\limits_{i = 1}^{M - 1} \boldsymbol h^{(i)T} \Gamma^{(i)} \boldsymbol h^{(i + 1)}\right],
+\psi(\ak{\thetaN}) = \log \sum\limits_{\tilde{\boldsymbol x} \in \mathcal{X}^{N_{(H,1)} + \cdots + N_{(H,M)} +N}} \exp\left[ \sum\limits_{i = 1}^M\boldsymbol \alpha^{(i)T} \boldsymbol h^{(i)} + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^{(1)T} \Gamma^{(0)} \boldsymbol x + \sum\limits_{i = 1}^{M - 1} \boldsymbol h^{(i)T} \Gamma^{(i)} \boldsymbol h^{(i + 1)}\right],
 $$ 
-is the normalizing function for $\boldsymbol \theta_N = (\boldsymbol \alpha^{(1)}, \dots, \boldsymbol \alpha^{(M)}, \boldsymbol \beta,\Gamma^{(0)}, \dots, \Gamma^{(M - 1)}) \in \Theta_N \subset \mathbb{R}^{q(N)}$, consisting of model parameters $\boldsymbol \beta \in \mathbb{R}^N$, $\boldsymbol \alpha^{(i)} \in \mathbb{R}^{N_{(H,i)}}$, $i = 1, \dots, M$, along with a matrix $\Gamma^{(0)}$ of dimension $N_{(H,1)} \times N$, and matrices $\Gamma^{(i)}$ of dimension
+is the normalizing function for $\ak{\thetaN} = (\boldsymbol \alpha^{(1)}, \dots, \boldsymbol \alpha^{(M)}, \boldsymbol \beta,\Gamma^{(0)}, \dots, \Gamma^{(M - 1)}) \in \Theta_\ak{\thetaidx} \subset \mathbb{R}^{q(N)}$, consisting of model parameters $\boldsymbol \beta \in \mathbb{R}^N$, $\boldsymbol \alpha^{(i)} \in \mathbb{R}^{N_{(H,i)}}$, $i = 1, \dots, M$, along with a matrix $\Gamma^{(0)}$ of dimension $N_{(H,1)} \times N$, and matrices $\Gamma^{(i)}$ of dimension
 $N_{(H,i)} \times N_{(H,i+1)}$ for $i = 1, \dots, M-1$.
-The combined parameter vector $\boldsymbol \theta_N$ has total length $q(N)= N_{(H,1)}+\cdots N_{(H,M)} + N + N_{(H,1)}*N+N_{H,2}*H_{(H,1)}+\cdots +N_{(H,M)}*H_{(H,M)-1}$.
+The combined parameter vector $\ak{\thetaN}$ has total length $q(N)= N_{(H,1)}+\cdots N_{(H,M)} + N + N_{(H,1)}*N+N_{H,2}*H_{(H,1)}+\cdots +N_{(H,M)}*H_{(H,M)-1}$.
 The probability mass function for the visible random variables $X_1, \dots, X_N$ follows from this joint specification as
 $$
-P_{\boldsymbol \theta_N} (\boldsymbol x) = \sum\limits_{(\boldsymbol h^{(1)}, \dots, \boldsymbol h^{(M)}) \in \mathcal{X}^{N_{(H,1)} + \cdots + N_{(H,M)}}} \tilde{P}_{\boldsymbol \theta_N} (\boldsymbol h^{(1)}, \dots, \boldsymbol h^{(M)}, \boldsymbol x) , \qquad \boldsymbol x \in \mathcal{X}^N.
+P_{\ak{\thetaN}} (\boldsymbol x) = \sum\limits_{(\boldsymbol h^{(1)}, \dots, \boldsymbol h^{(M)}) \in \mathcal{X}^{N_{(H,1)} + \cdots + N_{(H,M)}}} \tilde{P}_{\ak{\thetaN}} (\boldsymbol h^{(1)}, \dots, \boldsymbol h^{(M)}, \boldsymbol x) , \qquad \boldsymbol x \in \mathcal{X}^N.
 $$ 
@@
 **Deep belief network (DBN).** A DBN resembles a DBM in that there are multiple layers of latent random variables stacked in a deep architecture with no conditional dependence between layers.
-The difference between the DBM and DBN models is that all but the last stacked layer in a DBN are Bayesian networks [see @pearl985bayesian], rather than RBMs.
+The difference between the DBM and DBN models is that all but the last stacked layer \ak{(layer $0$, which contains the visible random variables $X_1, \dots, X_N$)} in a DBN are Bayesian networks [see @pearl985bayesian], rather than RBMs.
 A Bayesian network is a probabilistic graphical model that defines conditional dependence to be directed, rather than undirected (as with the RBM).
-Thus for visibles $X_1, \dots, X_N$ with support $\mathcal{X}^N, \mid \mathcal{X} \mid < \infty$, a DBN is also a FOES model with $q(N)$ the length of parameter vector is dependent on the dimension of the visibles because $P_{\boldsymbol \theta_N}(\boldsymbol x)>0$ for all $\boldsymbol x \in\mathcal{X}^N$.
+Thus for visibles $X_1, \dots, X_N$ with support $\mathcal{X}^N, \mid \mathcal{X} \mid < \infty$, a DBN is also a FOES model with $q(N)$ the length of parameter vector is dependent on the dimension of the visibles because $P_{\ak{\thetaN}}(\boldsymbol x)>0$ for all $\boldsymbol x \in\mathcal{X}^N$.
 Commonly, as in logistic belief nets [@neal1992connectionist], a "weight" parameter is placed on each interaction between visibles, $X_1, \dots, X_N$, and the first layer of latent variables, $H^{(1)}_1, \dots, H^{(1)}_{N_{(H,1)}}$, satisfying the definition of a FOES model.
@@
 
-To define a measure of instability in FOES models, it is useful to consider the behavior of data  models $P_{\theta_N}$, again supported on a set $\mathcal{X}^N$ of outcomes for $\boldsymbol X\equiv \boldsymbol X_N =(X_1,\ldots,X_N)$,  in connection to the sample size $N$.
+To define a measure of instability in FOES models, it is useful to consider the behavior of data  models $P_{\ak{\thetaN}}$, again supported on a set $\mathcal{X}^N$ of outcomes for $\boldsymbol X\equiv \boldsymbol X_N =(X_1,\ldots,X_N)$ \ak{and for a prescribed configuration of parameters $\ak{\thetaN}$,  in connection to the sample size $N$.
 A relevant quantity to this end is a log-ratio of extremal probabilities (LREP), defined as
@@
 (\#eq:elpr)
- \REP (\boldsymbol \theta_N)  =  \log \left[\frac{\max\limits_{  \boldsymbol x\in \mathcal{X}^N}P_{\boldsymbol \theta_N}( \boldsymbol x)}{\min\limits_{ \boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}( \boldsymbol x)}\right],
+ \REP (\ak{\thetaN})  =  \log \left[\frac{\max\limits_{  \boldsymbol x\in \mathcal{X}^N}P_{\ak{\thetaN}}( \boldsymbol x)}{\min\limits_{ \boldsymbol x \in \mathcal{X}^N}P_{\ak{\thetaN}}( \boldsymbol x)}\right],
 \end{align}
@@
 In what follows, the main idea is that instability, and other negative model features, can be associated with a FOES model formulation for $N$ random variables where the LREP \@ref(eq:elpr) is overly large relative to the sample size $N$.
-That is, a sequence of FOES probability models $P_{\boldsymbol \theta_N}$ results in specifying the distribution of observations $\boldsymbol X=(X_1,\ldots,X_N)$ for each sample size $N \geq 1$ and instability will generally occur among these models whenever the corresponding LREP \@ref(eq:elpr) grows faster than $N$.
+That is, a sequence of FOES probability models $P_{\ak{\thetaN}}$ results in specifying \ak{a} distribution of observations $\boldsymbol X=(X_1,\ldots,X_N)$ for each sample size $N \geq 1$ \ak{(dependent on the prescribed values for $\ak{\thetaN}$, $N \geq 1$)} and instability will generally occur among these models whenever the corresponding LREP \@ref(eq:elpr) grows faster than $N$.
 This leads to the following definition.
@@
 ```{definition, name="S-unstable FOES model", label="instabFSFS", echo=TRUE}
-A FOES model formulation for $\boldsymbol X_N=(X_1,\ldots,X_N)$ is *Schweinberger-unstable* or *S-unstable* if
+A FOES model formulation \ak{$P_{\ak{\thetaN}}$} for $\boldsymbol X_N=(X_1,\ldots,X_N)$, \ak{based on parameter specification $\ak{\thetaN}$, $N \geq 1$,} is *Schweinberger-unstable* or *S-unstable* if
 \begin{equation}
 (\#eq:Sun)
-\lim \limits_{N \rightarrow \infty} \frac{1}{N} \REP(\boldsymbol \theta_N) \equiv \lim \limits_{N \rightarrow \infty} \frac{1}{N}\log \left[\frac{\max\limits_{  \boldsymbol x\in \mathcal{X}^N}P_{\boldsymbol \theta_N}( \boldsymbol x)}{\min\limits_{ \boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}( \boldsymbol x)}\right] = \infty
+\lim \limits_{N \rightarrow \infty} \frac{1}{N} \REP(\ak{\thetaN}) \equiv \lim \limits_{N \rightarrow \infty} \frac{1}{N}\log \left[\frac{\max\limits_{  \boldsymbol x\in \mathcal{X}^N}P_{\ak{\thetaN}}( \boldsymbol x)}{\min\limits_{ \boldsymbol x \in \mathcal{X}^N}P_{\ak{\thetaN}}( \boldsymbol x)}\right] = \infty
 \end{equation}
@@
 ```
-In other words, a model is S-unstable if $\REP(\boldsymbol \theta_N)/N$ is an unbounded sequence of sample size $N$; namely, given any $C > 0$, there exists an integer $N_C > 0$ so that $\REP(\boldsymbol \theta_N)/N > C$ holds for all $N \ge N_C$.
-A FOES model formulation may be termed S-stable if it fails to be S-unstable, i.e., if $\sup_{N \geq 1}\REP(\boldsymbol \theta_N)/N$ is bounded.
+In other words, a model is S-unstable if $\REP(\ak{\thetaN})/N$ is an unbounded sequence of sample size $N$; namely, given any $C > 0$, there exists an integer $N_C > 0$ so that $\REP(\ak{\thetaN})/N > C$ holds for all $N \ge N_C$.
+A FOES model formulation may be termed S-stable if it fails to be S-unstable, i.e., if $\sup_{N \geq 1}\REP(\ak{\thetaN})/N$ is bounded.
 
-This definition of S-unstable is a generalization or reinterpretation of "unstable" used in @schweinberger2011instability by allowing possibly non-exponential family models (e.g., RBM and DBM models in Sections \@ref(rbm)-\@ref(deep-learning) as well as a potentially increasing number $q(N)$ of parameters through the parameter sequence $\boldsymbol \theta_N\in \mathbb{R}^{q(N)}$.
+This definition of S-unstable is a generalization or reinterpretation of "unstable" used in @schweinberger2011instability by allowing possibly non-exponential family models (e.g., RBM and DBM models in Sections \@ref(rbm)-\@ref(deep-learning) as well as a potentially increasing number $q(N)$ of parameters through the parameter sequence $\ak{\thetaN}\in \mathbb{R}^{q(N)}$.
 While this definition differs in form and scope from the original, it does match that in @schweinberger2011instability for the special case of exponential models (cf.
-Section \@ref(discrete-exponential-family-models) considered there.
-Section \@ref(illustrations) provides several examples  of unstable models as well as causes for model instability, where the latter may often be traced to issues in model form (i.e., data functions) and/or parameterization.
+Section \@ref(discrete-exponential-family-models) considered there \ak{which translates to $\ak{\thetaN}=\boldsymbol{\theta} \in \mathbb{R}^k$, $N \geq 1$, in our model notation).
+Note the definition (\ref{eq:Sun}) of model instability depends intricately on the prescribed parameter sequence $\ak{\thetaN}$ or the particular parameter values indexing the distribution $P_{\ak{\thetaN}}$.
+This also agrees with the notion of unstable distributions from \citet{schweinberger2011instability}, as well as other characterizations of distributional stability in networks (cf.
+\citet{handcock2003assessing}), that model instability is crucially tied to the parameter configuration (i.e., where $\ak{\thetaN}$ lies in the parameter space) in addition to distributional form.
+Hence, a given model family or form may may result in an unstable/stable model, depending on the parameters $\ak{\thetaN}$ chosen.} Section \@ref(illustrations) provides several examples  of unstable models as well as causes for model instability, where the latter may often be traced to issues in model form (i.e., data functions) and/or \ak{type of} parameterization.
 We next describe several potentially undesirable features associated with S-unstable FOES models.
@@
 ```{remark, label="rmk1", echo=TRUE}
-In the definition \@ref(eq:Sun) of S-instability, we note that the numerical measure $\REP(\boldsymbol \theta_N)/N$ of model instability is invariant to *independent  replications* of data.
-That is, let $M \geq 1$ denote a possible number of replications and consider data $\boldsymbol Y_{N,M} \equiv (\boldsymbol X^{(1)}_N, \dots, \boldsymbol X^{(M)}_N)$ formed by $\{ \boldsymbol X^{(j)}_N\}_{j=1}^M$ as $M$  iid replications of a random vector $\boldsymbol X_N=(X_1,\ldots,X_N)$, where the latter follows a FOES model with probabilities $P_{ \boldsymbol \theta_N}(\boldsymbol x)>0$, $\boldsymbol x\in\mathcal{X}^N$.
-This leads to a joint model, say $P_{ \boldsymbol \theta_N}(\boldsymbol y)$, $\boldsymbol y\in\mathcal{X}^{NM}$, for $\boldsymbol Y_{N,M}$ consisting of $N*M$ random variables in total.
+In the definition \@ref(eq:Sun) of S-instability, we note that the numerical measure $\REP(\ak{\thetaN})/N$ of model instability is invariant to *independent  replications* of data.
+That is, let $M \geq 1$ denote a possible number of replications and consider data $\boldsymbol Y_{N,M} \equiv (\boldsymbol X^{(1)}_N, \dots, \boldsymbol X^{(M)}_N)$ formed by $\{ \boldsymbol X^{(j)}_N\}_{j=1}^M$ as $M$  iid replications of a random vector $\boldsymbol X_N=(X_1,\ldots,X_N)$, where the latter follows a FOES model with probabilities $P_{\ak{\thetaN}}(\boldsymbol x)>0$, $\boldsymbol x\in\mathcal{X}^N$.
+This leads to a joint model, say $P_{\ak{\thetaN}}(\boldsymbol y)$, $\boldsymbol y\in\mathcal{X}^{NM}$, for $\boldsymbol Y_{N,M}$ consisting of $N*M$ random variables in total.
 Then, the LREP for $\boldsymbol Y_{N,M}$, scaled by associated size, is given by
 \begin{align*}
-\frac{1}{NM}\REP_{\boldsymbol Y_{N,M}}( \boldsymbol\theta_N ) &\equiv \frac{1}{NM}\log\left[\frac{\max_{\boldsymbol y \in \mathcal{X}^{NM}}P_{ \boldsymbol \theta_N}(\boldsymbol y)  }{\min_{\boldsymbol y \in \mathcal{X}^{NM}}P_{ \boldsymbol \theta_N}(\boldsymbol y)} \right] \\
-&= \frac{1}{NM}\log\left[\frac{\max_{\boldsymbol x \in \mathcal{X}^{N}}P_{ \boldsymbol \theta_N}(\boldsymbol x)  }{\min_{\boldsymbol x \in \mathcal{X}^{N}}P_{ \boldsymbol \theta_N}(\boldsymbol x)} \right]^M \equiv \frac{1}{N}\REP_{\boldsymbol X_{N}}( \boldsymbol\theta_N ),
+\frac{1}{NM}\REP_{\boldsymbol Y_{N,M}}(\ak{\thetaN} ) &\equiv \frac{1}{NM}\log\left[\frac{\max_{\boldsymbol y \in \mathcal{X}^{NM}}P_{\ak{\thetaN}}(\boldsymbol y)  }{\min_{\boldsymbol y \in \mathcal{X}^{NM}}P_{\ak{\thetaN}}(\boldsymbol y)} \right] \\
+&= \frac{1}{NM}\log\left[\frac{\max_{\boldsymbol x \in \mathcal{X}^{N}}P_{\ak{\thetaN}}(\boldsymbol x)  }{\min_{\boldsymbol x \in \mathcal{X}^{N}}P_{\ak{\thetaN}}(\boldsymbol x)} \right]^M \equiv \frac{1}{N}\REP_{\boldsymbol X_{N}}(\ak{\thetaN} ),
 \end{align*}
-where $\REP_{\boldsymbol X_{N}}( \boldsymbol\theta_N )  \equiv \REP( \boldsymbol\theta_N )$ denotes the log-ratio of extremal probabilities for $\boldsymbol X_N$ defined from \@ref(eq:elpr).
-That is, due to iid properties, the sample-size corrected  LREP for $\boldsymbol Y_{N,M}$ equals the analog, $\REP(\boldsymbol\theta_N )/N$, from the underlying common data model for $\boldsymbol X_N$ alone, regardless of the level $M \geq 1$ of independent replication.
+where $\REP_{\boldsymbol X_{N}}(\ak{\thetaN} )  \equiv \REP(\ak{\thetaN} )$ denotes the log-ratio of extremal probabilities for $\boldsymbol X_N$ defined from \@ref(eq:elpr).
+That is, due to iid properties, the sample-size corrected  LREP for $\boldsymbol Y_{N,M}$ equals the analog, $\REP(\ak{\thetaN} )/N$, from the underlying common data model for $\boldsymbol X_N$ alone, regardless of the level $M \geq 1$ of independent replication.
 Consequently, the definition of an S-unstable  model  is  unaffected by independent replication and all instability properties  may be characterized by those of one observation  from the common FOES model.
@@
 $$
-\frac{1}{N}\REP( \boldsymbol\theta_N ) \equiv \frac{1}{N}\REP_{\boldsymbol X_{N}}( \boldsymbol\theta_N ) = \log\left[ \frac{\max_{ x \in \mathcal{X}}P_{\boldsymbol\theta_N}(X_1=x)} {\min_{ x \in \mathcal{X}}P_{\boldsymbol\theta_N}(X_1=x)} \right]
+\frac{1}{N}\REP( \ak{\thetaN} ) \equiv \frac{1}{N}\REP_{\boldsymbol X_{N}}( \ak{\thetaN} ) = \log\left[ \frac{\max_{ x \in \mathcal{X}}P_{\ak{\thetaN}}(X_1=x)} {\min_{ x \in \mathcal{X}}P_{\ak{\thetaN}}(X_1=x)} \right]
 $$
@@
 $$
-\DN(\boldsymbol \theta_N) \equiv \max \left\{\log \frac{P_{\boldsymbol \theta_N}(\boldsymbol x)}{P_{\boldsymbol \theta_N}(\boldsymbol x^*)} : \boldsymbol x \text{ }\& \text{ } \boldsymbol x^* \in \mathcal{X}^N \text{ differ in exactly one component}\right\},
+\DN(\ak{\thetaN}) \equiv \max \left\{\log \frac{P_{\ak{\thetaN}}(\boldsymbol x)}{P_{\ak{\thetaN}}(\boldsymbol x^*)} : \boldsymbol x \text{ }\& \text{ } \boldsymbol x^* \in \mathcal{X}^N \text{ differ in exactly one component}\right\},
 $$
-which represents the biggest log-probability ratio for a one-component change in data outcomes in a FOES model with parameter $\boldsymbol \theta_N$.
-We then have the following result prescribing the behavior of $\DN(\boldsymbol \theta_N)$ for S-unstable FOES models.
+which represents the biggest log-probability ratio for a one-component change \ak{(i.e., the data vector differs in exactly one component value)} in data outcomes in a FOES model with parameter $\ak{\thetaN}$.
+\ak{This is the smallest difference possible between two data vectors, making it a good notion of a small distance in the input space.} We then have the following result prescribing the behavior of $\DN(\ak{\thetaN})$ for S-unstable FOES models.
 
 ```{theorem, label="instab-elpr", echo=TRUE}
-Let $P_{\boldsymbol \theta_N}$, with support $\mathcal{X}^N$, $N\geq 1$, be a sequence of FOES models.
+Let $P_{\ak{\thetaN}}$, with support $\mathcal{X}^N$, $N\geq 1$, be a sequence of FOES models.
 \begin{enumerate}[(i)]
-\item For any integer $N \geq 1$ and any given $C>0$, if $\REP(\boldsymbol \theta_N)/N > C$ in (\ref{eq:elpr}), then
-    $$ \DN(\boldsymbol \theta_N) > C,$$
+\item For any integer $N \geq 1$ and any given $C>0$, if $\REP(\ak{\thetaN})/N > C$ in (\ref{eq:elpr}), then
+    $$ \DN(\ak{\thetaN}) > C,$$
     or probabilities from a one-component change in some outcome have log-ratio exceeding $C$.
 \item Suppose the FOES model sequence is S-unstable.
-Then, for all large $N$ and given any arbitrary $C>0$, there exist outcomes $\boldsymbol x,\boldsymbol x^*\in\mathcal{X}^N$, differing by one component, such that
+Then, \ak{for any $C>0$ there exists $N'$ such that for all $N>N'$}, there exist outcomes $\boldsymbol x,\boldsymbol x^*\in\mathcal{X}^N$, differing by one component, such that
     $$
-    \frac{P_{\boldsymbol \theta_N}(\boldsymbol x)}{P_{\boldsymbol \theta_N}(\boldsymbol x^*)} > \exp[N C].
+    \frac{P_{\ak{\thetaN}}(\boldsymbol x)}{P_{\ak{\thetaN}}(\boldsymbol x^*)} > \exp[N C].
     $$
@@
 Additionally, S-unstable FOES model sequences are also connected to degenerate models, where *degeneracy* involves assigning essentially all probability to modes within the sample space, which could potentially represent a  small subset among the totality of outcomes.
-For perspective, note that differing sizes of the scaled log-ratio $\REP(\boldsymbol \theta_N)/N$ from \@ref(eq:elpr) induce a spectrum of levels of instability/stability and Theorem \@ref(thm:instab-elpr) indicates increasing sensitivity of model probabilities as \@ref(eq:elpr) increases.
-Furthermore, as the instability measure grows and the log-ratio $\REP(\boldsymbol \theta_N)/N$ diverges, as in the definition \@ref(eq:Sun) of S-unstable models, then a FOES model sequence  will become degenerate.
+For perspective, note that differing sizes of the scaled log-ratio $\REP(\ak{\thetaN})/N$ from \@ref(eq:elpr) induce a spectrum of levels of instability/stability and Theorem \@ref(thm:instab-elpr) indicates increasing sensitivity of model probabilities as \@ref(eq:elpr) increases.
+Furthermore, as the instability measure grows and the log-ratio $\REP(\ak{\thetaN})/N$ diverges, as in the definition \@ref(eq:Sun) of S-unstable models, then a FOES model sequence  will become degenerate.
 Theorem \@ref(thm:degenFOES) provides a formal statement of such degeneracy due to S-instability.
@@
 (\#eq:mode)
-\mathcal{M}_{\epsilon, \boldsymbol \theta_N} \equiv \left\{\boldsymbol x \in \mathcal{X}^N: \log P_{\boldsymbol \theta_N}(\boldsymbol x) > (1-\epsilon)\max\limits_{\boldsymbol y \in \mathcal{X}^N} \log  P_{\boldsymbol \theta_N}(\boldsymbol y) + \epsilon\min\limits_{\boldsymbol y \in \mathcal{X}^N} \log P_{\boldsymbol \theta_N}(\boldsymbol y) \right\}.
+\mathcal{M}_{\epsilon, \ak{\thetaN}} \equiv \left\{\boldsymbol x \in \mathcal{X}^N: \log P_{\ak{\thetaN}}(\boldsymbol x) > (1-\epsilon)\max\limits_{\boldsymbol y \in \mathcal{X}^N} \log  P_{\ak{\thetaN}}(\boldsymbol y) + \epsilon\min\limits_{\boldsymbol y \in \mathcal{X}^N} \log P_{\ak{\thetaN}}(\boldsymbol y) \right\}.
 \end{equation}
@@
 ```{theorem, label="degenFOES", echo=TRUE}
-For any arbitrarily small $0 < \epsilon < 1$, an S-unstable FOES model sequence $P_{\boldsymbol \theta_N}$, $N \geq 1$, for $\boldsymbol X_N=(X_1, \dots, X_N)$ satisfies
+For any arbitrarily small $0 < \epsilon < 1$, an S-unstable FOES model sequence $P_{\ak{\thetaN}}$, $N \geq 1$, for $\boldsymbol X_N=(X_1, \dots, X_N)$ satisfies
 $$
-P_{\boldsymbol \theta_N}\left( \boldsymbol X_N\in \mathcal{M}_{\epsilon, \boldsymbol \theta_N}\right) \rightarrow 1 \text{ as } N \rightarrow \infty.
+P_{\ak{\thetaN}}\left( \boldsymbol X_N\in \mathcal{M}_{\epsilon, \ak{\thetaN}}\right) \rightarrow 1 \text{ as } N \rightarrow \infty.
 $$
@@
 
-Connected to degeneracy, S-unstable FOES models may also exhibit additional kinds of  extreme and undesirable sensitivity in probabilities if model parameters $\boldsymbol \theta_N$ can further be "dialed" between positive and negative values.
+Connected to degeneracy, S-unstable FOES models may also exhibit additional kinds of  extreme and undesirable sensitivity in probabilities if model parameters $\ak{\thetaN}$ can further be "dialed" between positive and negative values.
 That is, some FOES models naturally involve parameter spaces covering a positive-negative spectrum of parameter possibilities, where the signs of parameters provide a standard device for increasing or decreasing probabilities of outcomes in the model formulation.
@@
 
-**Model Condition PSR** *(Reciprocal Probabilities from Parameter Sign Reversal)*: Let $P_{\boldsymbol \theta_N}$, with support $\mathcal{X}^N$, $N\geq 1$, represent a sequence of FOES models.
+**Model Condition PSR** *(Reciprocal Probabilities from Parameter Sign Reversal)*: Let $P_{\ak{\thetaN}}$, with support $\mathcal{X}^N$, $N\geq 1$, represent a sequence of FOES models.
 For each $N \geq 1$ and any outcome $\boldsymbol x \in \mathcal{X}^N$, suppose it holds that
 $$
-P_{\boldsymbol \theta_N}(\boldsymbol x)  \cdot P_{-\boldsymbol \theta_N}(\boldsymbol x) \;=\;   \max\limits_{\boldsymbol y \in \mathcal{X}^N}P_{ \boldsymbol \theta_N}(\boldsymbol y)\cdot \min\limits_{\boldsymbol y \in \mathcal{X}^N}P_{-\boldsymbol \theta_N}(\boldsymbol y),
+P_{\ak{\thetaN}}(\boldsymbol x)  \cdot P_{-\ak{\thetaN}}(\boldsymbol x) \;=\;   \max\limits_{\boldsymbol y \in \mathcal{X}^N}P_{ \ak{\thetaN}}(\boldsymbol y)\cdot \min\limits_{\boldsymbol y \in \mathcal{X}^N}P_{-\ak{\thetaN}}(\boldsymbol y),
 $$
-where $\max_{\boldsymbol y \in \mathcal{X}^N}P_{ \boldsymbol \theta_N}(\boldsymbol y)$ and $\min_{\boldsymbol y \in \mathcal{X}^N}P_{-\boldsymbol \theta_N}(\boldsymbol y)$ denote the maximum and minimum probabilities under parameters $\boldsymbol \theta_N$ and $-\boldsymbol \theta_N$, respectively.
+where $\max_{\boldsymbol y \in \mathcal{X}^N}P_{\ak{\thetaN}}(\boldsymbol y)$ and $\min_{\boldsymbol y \in \mathcal{X}^N}P_{-\ak{\thetaN}}(\boldsymbol y)$ denote the maximum and minimum probabilities under parameters $\ak{\thetaN}$ and $-\ak{\thetaN}$, respectively.
 
-The above model condition incorporates many standard parameterizations and follows, for instance, whenever $P_{\boldsymbol \theta_N}(\boldsymbol x)/P_{\boldsymbol \theta_N}(\boldsymbol y)  = [P_{-\boldsymbol \theta_N}(\boldsymbol x)/P_{-\boldsymbol \theta_N}(\boldsymbol y)]^{-1}$ holds for outcomes $\boldsymbol x, \boldsymbol y \in\mathcal{X}^N$ in a FOES model.
+The above model condition incorporates many standard parameterizations and follows, for instance, whenever $P_{\ak{\thetaN}}(\boldsymbol x)/P_{\ak{\thetaN}}(\boldsymbol y)  = [P_{-\ak{\thetaN}}(\boldsymbol x)/P_{-\ak{\thetaN}}(\boldsymbol y)]^{-1}$ holds for outcomes $\boldsymbol x, \boldsymbol y \in\mathcal{X}^N$ in a FOES model.
 For instance, this latter condition is fulfilled for all linear exponential families from Section \@ref(discrete-exponential-family-models) (e.g., \@ref(eq:mod1)-\@ref(eq:mod2)) as well as all network models from Sections \@ref(rbm)-\@ref(deep-learning) (e.g., \@ref(eq:RBM1)-\@ref(eq:RBM2)).
@@
 ```{corollary, label="sign", echo=TRUE}
-Let $P_{\boldsymbol \theta_N}$, with support $\mathcal{X}^N$, $N\geq 1$, be a sequence of FOES models satisfying model condition PSR.
-If the models $P_{\boldsymbol \theta_N}$  are additionally S-unstable, then
+Let $P_{\ak{\thetaN}}$, with support $\mathcal{X}^N$, $N\geq 1$, be a sequence of FOES models satisfying model condition PSR. If the models $P_{\ak{\thetaN}}$  are additionally S-unstable, then
 \begin{enumerate}[(i)]
-\item the models  $P_{-\boldsymbol \theta_N}$  defined by $-\boldsymbol \theta_N$  are also S-unstable;
-\item and for the complement $\mathcal{M}_{\epsilon, \boldsymbol \theta_N}^c \equiv \mathcal{X}^N \setminus \mathcal{M}_{\epsilon, \boldsymbol \theta_N}$ of any mode-set $\mathcal{M}_{\epsilon, \boldsymbol \theta_N}$ under $\boldsymbol \theta_N$ from (\ref{eq:mode}), with $0<\epsilon<1$, it holds under $-\boldsymbol \theta_N$ that
+\item the models  $P_{-\ak{\thetaN}}$  defined by $-\ak{\thetaN}$  are also S-unstable;
+\item and for the complement $\mathcal{M}_{\epsilon, \ak{\thetaN}}^c \equiv \mathcal{X}^N \setminus \mathcal{M}_{\epsilon, \ak{\thetaN}}$ of any mode-set $\mathcal{M}_{\epsilon, \ak{\thetaN}}$ under $\ak{\thetaN}$ from (\ref{eq:mode}), with $0<\epsilon<1$, it holds under $-\ak{\thetaN}$ that
     $$ 
-    P_{-\boldsymbol \theta_N}( \boldsymbol X_N \in \mathcal{M}_{\epsilon, \boldsymbol \theta_N}^c) \rightarrow 1\qquad \text{as } N\to \infty,
+    P_{-\ak{\thetaN}}( \boldsymbol X_N \in \mathcal{M}_{\epsilon, \ak{\thetaN}}^c) \rightarrow 1\qquad \text{as } N\to \infty,
     $$
-    while, by Theorem \ref{thm:degenFOES}, $P_{\boldsymbol \theta_N}( \boldsymbol X_N \in \mathcal{M}_{\epsilon, \boldsymbol \theta_N} ) \rightarrow 1$ holds for $\boldsymbol X_N = (X_1,\ldots,X_N)$ under $\boldsymbol \theta_N$.
+    while, by Theorem \ref{thm:degenFOES}, $P_{\ak{\thetaN}}( \boldsymbol X_N \in \mathcal{M}_{\epsilon, \ak{\thetaN}} ) \rightarrow 1$ holds for $\boldsymbol X_N = (X_1,\ldots,X_N)$ under $\ak{\thetaN}$.
 \end{enumerate}  
@@
 
-For unstable models, Corollary \@ref(cor:sign) shows that shifts in parameters around zero (i.e., from $\boldsymbol \theta_N$ to $-\boldsymbol \theta_N$) can induce extreme changes in probability among subsets of the sample space, as another manifestation  of instability and hyper-sensitivity in probability structure.
-For one-parameter exponential families, involving a fixed real-valued linear parameter  $\boldsymbol \theta_N = \theta \in \mathbb{R}$ and sufficient statistic $\boldsymbol g_N(\boldsymbol x)\in \mathbb{R}$ in \@ref(eq:expo), @schweinberger2011instability [Theorem 3] proved a result similar in spirit, though based on a characterization there in terms of maximum $U_N \equiv \max_{\boldsymbol x\in\mathcal{X}^N}g_N(\boldsymbol x)$ and minimal $L_N \equiv \min_{\boldsymbol x\in\mathcal{X}^N}g_N(\boldsymbol x)$ values of the sufficient statistic.
-For this case in particular,  mode sets have specific, and essentially complementary, forms over positive and negative parameters, namely, $\mathcal{M}_{\epsilon, \boldsymbol \theta_N} = \{\boldsymbol x \in\mathcal{X}^N: g_N(\boldsymbol x)  >  (1-\epsilon) U_N + \epsilon L_N \}$ and $\mathcal{M}_{\epsilon, -\boldsymbol \theta_N} = \{\boldsymbol x \in\mathcal{X}^N: g_N(\boldsymbol x)  <  \epsilon U_N + (1-\epsilon) L_N \}$ for any $\boldsymbol \theta_N>0$, and
+For unstable models, Corollary \@ref(cor:sign) shows that shifts in parameters around zero (i.e., from $\ak{\thetaN}$ to $-\ak{\thetaN}$) can induce extreme changes in probability among subsets of the sample space, as another manifestation  of instability and hyper-sensitivity in probability structure.
+For one-parameter exponential families, involving a fixed real-valued linear parameter  $\ak{\thetaN} = \theta \in \mathbb{R}$ and sufficient statistic $\boldsymbol g_N(\boldsymbol x)\in \mathbb{R}$ in \@ref(eq:expo), @schweinberger2011instability [Theorem 3] proved a result similar in spirit, though based on a characterization there in terms of maximum $U_N \equiv \max_{\boldsymbol x\in\mathcal{X}^N}g_N(\boldsymbol x)$ and minimal $L_N \equiv \min_{\boldsymbol x\in\mathcal{X}^N}g_N(\boldsymbol x)$ values of the sufficient statistic.
+For this case in particular,  mode sets have specific, and essentially complementary, forms over positive and negative parameters, namely, $\mathcal{M}_{\epsilon, \ak{\thetaN}} = \{\boldsymbol x \in\mathcal{X}^N: g_N(\boldsymbol x)  >  (1-\epsilon) U_N + \epsilon L_N \}$ and $\mathcal{M}_{\epsilon, -\ak{\thetaN}} = \{\boldsymbol x \in\mathcal{X}^N: g_N(\boldsymbol x)  <  \epsilon U_N + (1-\epsilon) L_N \}$ for any $\ak{\thetaN}>0$, and
  @schweinberger2011instability [Theorem 3] showed each mode set collects all mass, under positive and negative parameters, respectively, with unstable models of this exponential type.
@@
 Under the model condition PSR, Corollary \ref{cor:sign} can also be extended to
-cases where parameter components $\boldsymbol \theta_N =(\boldsymbol \theta_{1,N}, \boldsymbol \theta_{N,2})$ (say) are not all changed in sign (e.g., $-\boldsymbol \theta_N$) but, more generally, are instead  altered to another parameter configuration $\boldsymbol \theta_{N}^A = ( \boldsymbol \theta_{1,N}^A, \boldsymbol \theta_{2,N}^A )$ involving a switch in sign only among some dominating model parameters $\boldsymbol \theta_{N,2}^A=- \boldsymbol \theta_{N,2}$ with remaining parameters $\boldsymbol \theta_{1,N}^A$ being arbitrarily chosen.
-If a change sign occurs among parameters ($\pm \boldsymbol \theta_{N,2}$) which dominate the probability structure of the model, then the results of Corollary  \ref{cor:sign} can still hold  with $\boldsymbol \theta_{N}^A$ replacing $-\boldsymbol \theta_N$; as an example of one sufficient condition, if $\lim_{N\to \infty} \max_{\boldsymbol x \in \mathcal{X}^N} |G_N(\boldsymbol x, -\boldsymbol \theta_{N}) - G_N(\boldsymbol x, \boldsymbol \theta_{N}^A)|=0$ holds in addition to Corollary \@ref(cor:sign) assumptions, where
+cases where parameter components $\ak{\thetaN} =(\boldsymbol \theta_{1,\ak{\thetaidx}}, \boldsymbol \theta_{2, \ak{\thetaidx}})$ (say) are not all changed in sign (e.g., $-\ak{\thetaN}$) but, more generally, are instead  altered to another parameter configuration $\ak{\thetaN}^A = ( \boldsymbol \theta_{1,\ak{\thetaidx}}^A, \boldsymbol \theta_{2,\ak{\thetaidx}}^A )$ involving a switch in sign only among some dominating model parameters $\boldsymbol \theta_{2,\ak{\thetaidx}}^A=- \boldsymbol \theta_{2,\ak{\thetaidx}}$ with remaining parameters $\boldsymbol \theta_{1,\ak{\thetaidx}}^A$ being arbitrarily chosen.
+If a change sign occurs among parameters ($\pm \boldsymbol \theta_{2,\ak{\thetaidx}}$) which dominate the probability structure of the model, then the results of Corollary  \ref{cor:sign} can still hold  with $\ak{\thetaN}^A$ replacing $-\ak{\thetaN}$; as an example of one sufficient condition, if $\lim_{N\to \infty} \max_{\boldsymbol x \in \mathcal{X}^N} |G_N(\boldsymbol x, -\ak{\thetaN}) - G_N(\boldsymbol x, \ak{\thetaN}^A)|=0$ holds in addition to Corollary \@ref(cor:sign) assumptions, where
 $$
@@
 $$
-represents a standardized form of $\boldsymbol \theta$-model probabilities, then the results of Corollary \ref{cor:sign} apply to $\boldsymbol \theta_{N}^A$ in addition to $-\boldsymbol \theta_N$.
-As a consequence, an unstable model under $\boldsymbol \theta_N$ can then imply that many more unstable models exist  over a broader spectrum of possibilities for variations $\boldsymbol \theta_{N}^A$ of $\boldsymbol \theta_N$, which involves some amount of sign change among components of $\boldsymbol \theta_N$.
+represents a standardized form of $\boldsymbol \theta$-model probabilities, then the results of Corollary \ref{cor:sign} apply to $\ak{\thetaN}^A$ in addition to $-\ak{\thetaN}$.
+As a consequence, an unstable model under $\ak{\thetaN}$ can then imply that many more unstable models exist  over a broader spectrum of possibilities for variations $\ak{\thetaN}^A$ of $\ak{\thetaN}$, which involves some amount of sign change among components of $\ak{\thetaN}$.
 ```
@@
 However, as shown in the following, bounded statistics may still lead to instability if the parameter dimension is high.
-We next provides some examples to illustrate S-instability in FOES models, which also suggest some potential strategies for preventing unstable models.
+We next \ak{provide} some examples to illustrate S-instability in FOES models, which also suggest some potential strategies for preventing unstable models.
 
@@
 
-As a baseline for comparisons, consider a simplistic model for $\boldsymbol X_N=(X_1,\ldots,X_N)$ with uniform probabilities over the sample space, say $P_{\boldsymbol \theta_N}(\boldsymbol x)= |\mathcal{X}|^{-N}$, $\boldsymbol x \in \mathcal{X}^N$, where each random variable has $|\mathcal{X}| \geq 1$ outcomes.
+As a baseline for comparisons, consider a simplistic model for $\boldsymbol X_N=(X_1,\ldots,X_N)$ with uniform probabilities over the sample space, say $P_{\ak{\thetaN}}(\boldsymbol x)= |\mathcal{X}|^{-N}$, $\boldsymbol x \in \mathcal{X}^N$, where each random variable has $|\mathcal{X}| \geq 1$ outcomes.
 In contrast to instability, model probabilities here are completely insensitive to changes in data outcomes across the sample space, and the associated log-ratio of extreme probabilities \@ref(eq:elpr) is
 $$
-\frac{1}{N} \REP(\boldsymbol \theta_N)=0\quad \;(\text{uniform probability model}),
+\frac{1}{N} \REP(\ak{\thetaN})=0\quad \;(\text{uniform probability model}),
 $$ 
@@
 
-A fundamental model considered in the instability work of @schweinberger2011instability involves a one-parameter exponential model corresponding to \@ref(eq:expo) with a real-valued parameter, say $\boldsymbol \theta_N =  \eta(\boldsymbol \theta_N)\in \mathbb{R}$, and sufficient statistic $\boldsymbol g_N(\boldsymbol x)\in \mathbb{R}$.
+A fundamental model considered in the instability work of @schweinberger2011instability involves a one-parameter exponential model corresponding to \@ref(eq:expo) with a real-valued parameter, say $\ak{\thetaN} =  \eta(\ak{\thetaN})\in \mathbb{R}$, and sufficient statistic $\boldsymbol g_N(\boldsymbol x)\in \mathbb{R}$.
 For such models, upon scaling by sample size $N$, the log-ratio of extreme probabilities in \@ref(eq:elpr) for assessing instability becomes
@@
 (\#eq:UL)
-\frac{1}{N}\REP(\boldsymbol \theta_N ) \equiv   |\boldsymbol \theta_N| \frac{(U_N-L_N)}{N} \;\quad \text{(one-parameter exponential model)},
+\frac{1}{N}\REP(\ak{\thetaN} ) \equiv   |\ak{\thetaN}| \frac{(U_N-L_N)}{N} \;\quad \text{(one-parameter exponential model)},
 \end{equation}
 where $U_N \equiv \max_{\boldsymbol x\in\mathcal{X}^N}g_N(\boldsymbol x)$  and $L_N \equiv \min_{\boldsymbol x\in\mathcal{X}^N}g_N(\boldsymbol x)$ denote the maximal and minimal values of the single sufficient statistic.
-In this case, an S-unstable model results, by definition \@ref(eq:Sun), whenever $\lim_{N\to \infty} |\boldsymbol \theta_N| (U_N-L_N)/N= \infty$ holds or, in other words, if the combined  magnitudes of parameter $|\boldsymbol \theta_N|$ and maximal difference $U_N-L_N$ in statistic values are overwhelmingly large relative to the sample size $N$.
-If we further assume that $\boldsymbol \theta_N =\theta\in\mathbb{R}\setminus \{0\}$ is a fixed (non-zero) parameter for all $N \geq 1$, as considered in @schweinberger2011instability, then an S-unstable model results solely if the sufficient statistic admits a value $U_N-L_N$ too large relative to number $N$ of observations, i.e., if $(U_N-L_N)/N\to \infty$ as $N\to \infty$.
+In this case, an S-unstable model results, by definition \@ref(eq:Sun), whenever $\lim_{N\to \infty} |\ak{\thetaN}| (U_N-L_N)/N= \infty$ holds or, in other words, if the combined  magnitudes of parameter $|\ak{\thetaN}|$ and maximal difference $U_N-L_N$ in statistic values are overwhelmingly large relative to the sample size $N$.
+If we further assume that $\ak{\thetaN} =\theta\in\mathbb{R}\setminus \{0\}$ is a fixed (non-zero) parameter for all $N \geq 1$, as considered in @schweinberger2011instability, then an S-unstable model results solely if the sufficient statistic admits a value $U_N-L_N$ too large relative to number $N$ of observations, i.e., if $(U_N-L_N)/N\to \infty$ as $N\to \infty$.
 The latter aspect reflects the definition of @schweinberger2011instability, for this setting, that a real-valued *statistic* $g_N(\boldsymbol x)$ may be classified as *unstable* when  $\lim_{N\to \infty}|(U_N-L_N)/N=\infty$ holds and as *stable* otherwise (e.g., if $\sup_{N \geq 1}(U_N-L_N)/N<\infty$).
 
-For illustration, consider the iid Bernoulli model \@ref(eq:mod1) for $\boldsymbol X_N=(X_1,\ldots,X_N)$ with log-odds ratio parameter $\boldsymbol \theta_N = \log[ P_{\boldsymbol \theta_N}(X_1=1)/ P_{\boldsymbol \theta_N}(X_1=0)]\in\mathbb{R}$.
-Remark 1 (Section \@ref(criterion)) then gives the model instability measure \@ref(eq:Sun) directly as
+For illustration, consider the iid Bernoulli model \@ref(eq:mod1) for $\boldsymbol X_N=(X_1,\ldots,X_N)$ with log-odds ratio parameter $\ak{\thetaN} = \log[ P_{\ak{\thetaN}}(X_1=1)/ P_{\ak{\thetaN}}(X_1=0)]\in\mathbb{R}$.
+\ak{Remark 3.1} (Section \@ref(criterion)) then gives the model instability measure \@ref(eq:Sun) directly as
 $$
-\frac{1}{N}\REP(\boldsymbol \theta_N ) = |\boldsymbol \theta_N|\quad\; \text{(iid Bernoulli model)},
+\frac{1}{N}\REP(\ak{\thetaN} ) = |\ak{\thetaN}|\quad\; \text{(iid Bernoulli model)},
 $$
-so that an unstable (or stable) model results for a divergent (or bounded) parameter sequence  $|\boldsymbol \theta_N|$.
+so that an unstable (or stable) model results for a divergent (or bounded) parameter sequence  $|\ak{\thetaN}|$.
 The above instability expression for the Bernoulli model follows as well from  the $N$-scaled LREP value \@ref(eq:UL) for a one-parameter exponential distribution, using that the sufficient statistic involved $g_N(\boldsymbol x)= \sum_{i=1}^N x_i$, $\boldsymbol x =(x_1,\ldots,x_n)\in\{0,1\}^N$, has maximum and minimum values $U_N=N$ and $L_N=0$.
-In this case, @schweinberger2011instability has noted that the statistic is stable (i.e., bounded $(U_N-L_N)/N=1$) and the Bernoulli model is as well when, in particular, $\boldsymbol \theta_N=\bold \theta \in\mathbb{R}$ is fixed for $N \geq 1$.
+In this case, @schweinberger2011instability has noted that the statistic is stable (i.e., bounded $(U_N-L_N)/N=1$) and the Bernoulli model is as well when, in particular, $\ak{\thetaN}=\theta \in\mathbb{R}$ is fixed for $N \geq 1$.
 
-Alternatively, considering a random graph with $N={n \choose 2}$ edges among $n$ nodes, the exponential graph model from \@ref(eq:mod2), when based purely on the number of $g_{2,N}(\boldsymbol x)$ of 2-stars or solely the number $g_{3,N}(\boldsymbol x)$  of triangles, $\boldsymbol x\in\{0,1\}^N$, has an measure of instability from \@ref(eq:Sun) as
+Alternatively, considering a random graph with $N={n \choose 2}$ edges among $n$ nodes, the exponential graph model from \@ref(eq:mod2), when based purely on the number of $g_{2,N}(\boldsymbol x)$ of 2-stars or solely the number $g_{3,N}(\boldsymbol x)$  of triangles, $\boldsymbol x\in\{0,1\}^N$, has a measure of instability from \@ref(eq:Sun) as
 $$
-\frac{1}{N}\REP(\boldsymbol \theta_N )  = \left\{ \begin{array}{lcl} |\boldsymbol \theta_N| (n-2) && \text{(2-star graph model)}\\
-|\boldsymbol \theta_N|(n-2)/3 &&\text{(triangle graph model)},\end{array}\right.
+\frac{1}{N}\REP(\ak{\thetaN} )  = \left\{ \begin{array}{lcl} |\ak{\thetaN}| (n-2) && \text{(2-star graph model)}\\
+|\ak{\thetaN}|(n-2)/3 &&\text{(triangle graph model)},\end{array}\right.
 $$
@@
 Because the variable number $N\to \infty$ as the node number $n\to \infty$, both counts of 2-stars and triangles are unstable statistics in the sense of @schweinberger2011instability (i.e., $\lim_{N\to \infty} (U_N-L_N)/N=\infty$).
-Furthermore, both types of graph models are always S-unstable for all possible of parameter sequences $\boldsymbol \theta_N \in\mathbb{R}$ that are bounded away from zero (i.e.,  $\lim_{N\to \infty}\REP(\boldsymbol \theta_N )/N=\infty$ then  holds, including the fixed parameter case $\boldsymbol \theta_N=\theta\in\mathbb{R}\setminus \{0\}$  from @schweinberger2011instability).
+Furthermore, both types of graph models are always S-unstable for all possible of parameter sequences $\ak{\thetaN} \in\mathbb{R}$ that are bounded away from zero (i.e.,  $\lim_{N\to \infty}\REP(\ak{\thetaN} )/N=\infty$ then  holds, including the fixed parameter case $ \ak{\thetaN}=\theta\in\mathbb{R}\setminus \{0\}$  from @schweinberger2011instability).
 
@@
 
-As a generalization of the one-parameter exponential case, we next consider linear exponential families \@ref(eq:expo) with $k$ parameters $\boldsymbol \theta_N = (\theta_{1,N},\ldots,\theta_{k,N})^\prime$ and $k$ sufficient statistics $\boldsymbol g_N(\boldsymbol x) = (g_{1,N}(\boldsymbol x),\ldots, g_{k,N}(\boldsymbol x))^\prime$.
+As a generalization of the one-parameter exponential case, we next consider linear exponential families \@ref(eq:expo) with $k$ parameters $\ak{\thetaN} = (\theta_{1,\ak{\thetaidx}},\ldots,\theta_{k,\ak{\thetaidx}})^\prime$ and $k$ sufficient statistics $\boldsymbol g_N(\boldsymbol x) = (g_{1,N}(\boldsymbol x),\ldots, g_{k,N}(\boldsymbol x))^\prime$.
 Here the dimension $k$ of model parameters/statistics is fixed, and we next prescribe a condition helpful to avoiding instability in such models.
@@
 ```{proposition, label="prop1", echo=TRUE}
-Let $P_{\boldsymbol \theta_N}$, $N \geq 1$,  denote linear exponential models (\ref{eq:expo}) with   parameters $\boldsymbol \theta_N = (\theta_{1,N},\ldots,\theta_{k,N})^\prime \in \mathbb{R}^k$ and statistics  $\boldsymbol g_N(\boldsymbol x) = (g_{1,N}(\boldsymbol x),\ldots, g_{k,N}(\boldsymbol x))^\prime \in \mathbb{R}^k$, for fixed $k \geq 1$.
-Then, the models $P_{\boldsymbol \theta_N}$ are S-stable if
+Let $P_{\ak{\thetaN}}$, $N \geq 1$,  denote linear exponential models (\ref{eq:expo}) with   parameters $\ak{\thetaN} = (\theta_{1,\ak{\thetaidx}},\ldots,\theta_{k,\ak{\thetaidx}})^\prime \in \mathbb{R}^k$ and statistics  $\boldsymbol g_N(\boldsymbol x) = (g_{1,N}(\boldsymbol x),\ldots, g_{k,N}(\boldsymbol x))^\prime \in \mathbb{R}^k$, for fixed $k \geq 1$.
+Then, the models $P_{\ak{\thetaN}}$ are S-stable if
 \begin{equation}
 (\#eq:prop1)
-\sup_{N \geq 1}\frac{1}{N} \max_{1 \leq i \leq k }|\theta_{i,N}|(U_{i,N}-L_{i,N})<\infty
+\sup_{N \geq 1}\frac{1}{N} \max_{1 \leq i \leq k }|\theta_{i,\ak{\thetaidx}}|(U_{i,N}-L_{i,N})<\infty
 \end{equation}
-holds, i.e.,  if $\max_{1 \leq i \leq k } |\theta_{i,N}|(U_{i,N}-L_{i,N})/N$ is bounded sequence of sample size $N$.
+holds, i.e.,  if $\max_{1 \leq i \leq k } |\theta_{i,\ak{\thetaidx}}|(U_{i,N}-L_{i,N})/N$ is bounded sequence of sample size $N$.
 ```
@@
 ```{remark, echo=TRUE}
-In the one-parameter exponential case $k=1$, recall the exponential model is stable/unstable depending on whether $\REP(\boldsymbol \theta_N)/N = |\theta_{1,N}|(U_{1,N}-L_{1,N})/N \equiv |\boldsymbol \theta_{N}|(U_{N}-L_{N})/N$ in \@ref(eq:UL) is  convergent/divergent.
+In the one-parameter exponential case $k=1$, recall the exponential model is stable/unstable depending on whether $\REP(\ak{\thetaN})/N = |\theta_{1,\ak{\thetaidx}}|(U_{1,N}-L_{1,N})/N \equiv | \ak{\thetaN}|(U_{N}-L_{N})/N$ in \@ref(eq:UL) is  convergent/divergent.
 Hence, for $k=1$, the condition \@ref(eq:prop1) of Proposition \@ref(prp:prop1) captures the same notion of S-stability based on \@ref(eq:UL).
@@
 
-Proposition \@ref(prp:prop1) provides a sufficient condition for the stability of linear exponential models with fixed parameter dimension $k\geq 1$, whereby an S-stable model is guaranteed if the compounded magnitude of each combination of parameter $\theta_{i,N}$ and sufficient statistic value $(U_{i,N}-L_{i,N})$  is bounded by the sample size $N$, $i=1,\ldots,k$.
+Proposition \@ref(prp:prop1) provides a sufficient condition for the stability of linear exponential models with fixed parameter dimension $k\geq 1$, whereby an S-stable model is guaranteed if the \ak{product of magnitudes} of each combination of parameter $\theta_{i,\ak{\thetaidx}}$ and sufficient statistic value $(U_{i,N}-L_{i,N})$ is bounded by the sample size $N$, $i=1,\ldots,k$.
 This supports the findings of @schweinberger2011instability, who showed degeneracy follows in such models under one type of violation of the condition \@ref(eq:prop1) in Proposition \@ref(prp:prop1) (namely, involving $k>1$ non-zero parameters  with $k-1$ statistics being $O(N)$ bounded while one statistic diverges in maximal size faster than the number $N$ of observations).
-To further illustrate the result in Proposition \@ref(prp:prop1), consider the multinomial distribution \@ref(eq:mod11) for $\boldsymbol X_N=(X_1,\ldots,X_N)$ having $k\geq 2$ categories $\{1,\ldots,k\}$ and $k$ parameters $\boldsymbol \theta_N = (\theta_{1,N},\ldots,\theta_{k,N})^\prime$.
-The variables are iid under this model so that Remark 1 (Section \@ref(criterion) yields the corresponding $N$-scaled  log-ratio of extreme probabilities \@ref(eq:elpr) as
+To further illustrate the result in Proposition \@ref(prp:prop1), consider the multinomial distribution \@ref(eq:mod11) for $\boldsymbol X_N=(X_1,\ldots,X_N)$ having $k\geq 2$ categories $\{1,\ldots,k\}$ and $k$ parameters $\ak{\thetaN} = (\theta_{1,\ak{\thetaidx}},\ldots,\theta_{k,\ak{\thetaidx}})^\prime$.
+The variables are iid under this model so that \ak{Remark 3.1} (Section \@ref(criterion) yields the corresponding $N$-scaled  log-ratio of extreme probabilities \@ref(eq:elpr) as
 \begin{align*}
-  \frac{1}{N}\REP(\boldsymbol \theta_N) &= \frac{\max_{1 \leq i \leq k} P_{\boldsymbol \theta_N}(X_1=i)}{\min_{1 \leq i \leq k} P_{\boldsymbol \theta_N}(X_1=i)}\\
-  &= \max_{1 \leq i \leq k} \theta_{i,N} - \min_{1 \leq i \leq k} \theta_{i,N} \qquad \text{(iid multinomial model)}.
+  \frac{1}{N}\REP(\ak{\thetaN}) &= \frac{\max_{1 \leq i \leq k} P_{\ak{\thetaN}}(X_1=i)}{\min_{1 \leq i \leq k} P_{\ak{\thetaN}}(X_1=i)}\\
+  &= \max_{1 \leq i \leq k} \theta_{i,\ak{\thetaidx}} - \min_{1 \leq i \leq k} \theta_{i,\ak{\thetaidx}} \qquad \text{(iid multinomial model)}.
 \end{align*}
-Hence, a multinomial model sequence is unstable (or stable) depending on whether (or not) the maximal parameter difference $\max_{1 \leq i \leq k} \theta_{i,N} - \min_{1 \leq i \leq k} \theta_{i,N}$ diverges.
-Furthermore, using that each of the $k$ sufficient (count) statistics from the multinomial model \@ref(eq:mod11) satisfies $(U_{i,N}-L_{i,N})/N=1$, we  see that \@ref(eq:prop1) of Proposition \@ref(prp:prop1) becomes purely a parameter condition, $\sup_{N \geq 1}\max_{1 \leq i \leq k } |\theta_{i,N}| <\infty$, for ensuring that $\REP(\boldsymbol \theta_N)/N =\max_{1 \leq i \leq k} \theta_{i,N} - \min_{1 \leq i \leq k} \theta_{i,N}$ is bounded and stability follows for the multinomial distribution.
-Additionally, a stable multinomial sequence (i.e., bounded $\REP(\boldsymbol \theta_N)/N$) turns out to be nearly equivalent to \@ref(eq:prop1)  (e.g., these are the same if the smallest  parameter $\min_{1 \leq i \leq k } |\theta_{i,N}|$ remains bounded).
+Hence, a multinomial model sequence is unstable (or stable) depending on whether (or not) the maximal parameter difference $\max_{1 \leq i \leq k} \theta_{i,\ak{\thetaidx}} - \min_{1 \leq i \leq k} \theta_{i,\ak{\thetaidx}}$ diverges.
+Furthermore, using that each of the $k$ sufficient (count) statistics from the multinomial model \@ref(eq:mod11) satisfies $(U_{i,N}-L_{i,N})/N=1$, we  see that \@ref(eq:prop1) of Proposition \@ref(prp:prop1) becomes purely a parameter condition, $\sup_{N \geq 1}\max_{1 \leq i \leq k } |\theta_{i,\ak{\thetaidx}}| <\infty$, for ensuring that $\REP(\ak{\thetaN})/N =\max_{1 \leq i \leq k} \theta_{i,\ak{\thetaidx}} - \min_{1 \leq i \leq k} \theta_{i,\ak{\thetaidx}}$ is bounded and stability follows for the multinomial distribution.
+Additionally, a stable multinomial sequence (i.e., bounded $\REP(\ak{\thetaN})/N$) turns out to be nearly equivalent to \@ref(eq:prop1)  (e.g., these are the same if the smallest  parameter $\min_{1 \leq i \leq k } |\theta_{i,\ak{\thetaidx}}|$ remains bounded).
 
 When the condition \@ref(eq:prop1) of Proposition \@ref(prp:prop1) is violated, this aspect suggests a potentially unstable model that may be investigated more closely.
-For example, consider the exponential graph model from \@ref(eq:mod2) involving counts of edges, 2-stars and triangles with fixed parameters $\boldsymbol \theta_N = (\theta_{1},\theta_2,\theta_3)^\prime \in \mathbb{R}^3$ for $N\geq 1$.
+For example, consider the exponential graph model from \@ref(eq:mod2) involving counts of edges, 2-stars and triangles with fixed parameters $\thet_N = (\theta_{1},\theta_2,\theta_3)^\prime \in \mathbb{R}^3$ for $N\geq 1$.
 If either the 2-star parameter $\theta_2 \neq 0$ or triangle parameter $\theta_3 \neq 0$ is non-zero, then $\max_{1 \leq i \leq 3 } |\theta_{i}|(U_{i,N}-L_{i,N})/N \propto (n-2)\to \infty$ holds in \@ref(eq:prop1) by  $(U_{2,N}-L_{2,N})/N = 3 (U_{3,N}-L_{3,N})/N=(n-2)$ for 2-star and triangle statistics ($i=2,3$), so that Proposition \@ref(prp:prop1) hints that an unstable model may result when $|\theta_2| + |\theta_3| \neq 0$.
-Relatedly, a result from @schweinberger2011instability [Result 3] states that this model is unstable for all fixed parameters excluding cases $\theta_2 =\theta_3=0$ or $\theta_2 = - \theta_3/3$.
-However, more is true in line with the instability suggested by Proposition \@ref(prp:prop1) whenever $|\theta_2| + |\theta_3| \neq 0$ (i.e., excluding $\theta_2 =\theta_3=0$).
+Relatedly, a result from @schweinberger2011instability [Result 3] states that this model is unstable for all fixed parameters excluding cases $\theta_2 =\theta_3=0$ or $\theta_2 = - \theta_3/3$\ak{, more in line} with the instability suggested by Proposition \@ref(prp:prop1) whenever $|\theta_2| + |\theta_3| \neq 0$ (i.e., excluding $\theta_2 =\theta_3=0$).
 
@@
 \begin{align*}
-\frac{1}{N}\REP_N(\boldsymbol \theta_N) &\geq \max_{i=1,2}\frac{1}{N}
-\left| \log\left[ \frac{P_{\boldsymbol \theta_N}(\boldsymbol x_i)}{P_{\boldsymbol \theta_N}(\boldsymbol x_0)}\right] \right| \\
+\frac{1}{N}\REP_N(\ak{\thetaN}) &\geq \max_{i=1,2}\frac{1}{N}
+\left| \log\left[ \frac{P_{\ak{\thetaN}}(\boldsymbol x_i)}{P_{\ak{\thetaN}}(\boldsymbol x_0)}\right] \right| \\
 &= (n-2) \max\left\{ \left| \theta_2 + \frac{\theta_3}{3}+\frac{\theta_1}{n-2} \right|, \frac{n}{4(n-1)} \left| \theta_2 + \frac{8\theta_1}{n-2} \right| \right\};
@@
 a similar expression also holds for an odd node number $n>2$.
-Consequently, for all fixed parameters excluding $\theta_2=\theta_3=0$, $\lim_{N\to \infty}\REP_N(\boldsymbol \theta_N)/N=\infty$ then follows and the graph model with 2-stars and triangles is S-unstable, as suggested by the breach of Proposition \@ref(prp:prop1) for this model when $|\theta_2|+|\theta_3|\neq 0$.
-That is, instability holds even under $\theta_2 = - \theta_3/3$ case potentially allowed by Schweinberger's [-@schweinberger2011instability] results.
+Consequently, for all fixed parameters excluding $\theta_2=\theta_3=0$, $\lim_{N\to \infty}\REP_N(\ak{\thetaN})/N=\infty$ then follows and the graph model with 2-stars and triangles is S-unstable, as suggested by the breach of Proposition \@ref(prp:prop1) for this model when $|\theta_2|+|\theta_3|\neq 0$.
+That is, instability holds even under \ak{the} $\theta_2 = - \theta_3/3$ case potentially allowed by Schweinberger's [-@schweinberger2011instability] results.
 
@@
 However, the distribution of visible variables is induced by an underlying joint exponential model \@ref(eq:RBM1) for both visible and latent variables $(\boldsymbol X, \boldsymbol H)$, where $\boldsymbol H = (H_1,\ldots,H_{N_{\mathcal{H}}})$ denotes a vector of $N_{\mathcal{H}}$ hidden variables (similarly binary).
-The joint model is of linear exponential form involving $q(N)\equiv N + N_{\mathcal{H}} + N*N_{\mathcal{H}}$ sufficient statistics given by $(\boldsymbol X, \boldsymbol H, \boldsymbol X^T\boldsymbol H)$ and parameters $\boldsymbol \theta_N = (\boldsymbol \theta_N^{\mathcal{V}},\boldsymbol \theta_N^{\mathcal{H}}, \boldsymbol \theta_N^{\mathcal{VH}}  ) \in\mathbb{R}^{q(N)}$ corresponding to the $N$ visible variables $\boldsymbol X$ (i.e., $\boldsymbol \theta_N^{\mathcal{V}}\in\mathbb{R}^N$), the $N_{\mathcal{H}}$ hidden variables $\boldsymbol H$ (i.e., $\boldsymbol \theta_N^{\mathcal{H}}\in\mathbb{R}^{N_{\mathcal{H}}}$), and the $N *N_{\mathcal{H}}$ cross-product variables $\boldsymbol X^T\boldsymbol H$ (i.e., $\boldsymbol \theta_N^{\mathcal{VH}}\in\mathbb{R}^{N*N_{\mathcal{H}}}$).
+The joint model is of linear exponential form involving $q(N)\equiv N + N_{\mathcal{H}} + N*N_{\mathcal{H}}$ sufficient statistics given by $(\boldsymbol X, \boldsymbol H, \boldsymbol X^T\boldsymbol H)$ and parameters $\ak{\thetaN} = (\ak{\thetaN}^{\mathcal{V}}, \ak{\thetaN}^{\mathcal{H}}, \ak{\thetaN}^{\mathcal{VH}}  ) \in\mathbb{R}^{q(N)}$ corresponding to the $N$ visible variables $\boldsymbol X$ (i.e., $\ak{\thetaN}^{\mathcal{V}}\in\mathbb{R}^N$), the $N_{\mathcal{H}}$ hidden variables $\boldsymbol H$ (i.e., $\ak{\thetaN}^{\mathcal{H}}\in\mathbb{R}^{N_{\mathcal{H}}}$), and the $N *N_{\mathcal{H}}$ cross-product variables $\boldsymbol X^T\boldsymbol H$ (i.e., $\ak{\thetaN}^{\mathcal{VH}}\in\mathbb{R}^{N*N_{\mathcal{H}}}$).
 However, unlike some previous exponential models considered in Sections \@ref(one-param-exp)-\@ref(fixed-dim-exp) (cf.
@@
 To highlight the instability issues for the RBM model, consider a simple  model for $N$ visibles $\boldsymbol X$ with no hidden variables ($N_{\mathcal{H}}=0$), for which model statements \@ref(eq:RBM1)-\@ref(eq:RBM2) coincide.
-An independence model then results for variables in $\boldsymbol X$, which has $q(N)=N$ parameters $\boldsymbol \theta_N^{\mathcal{V}} = (\theta_{1,N}^{\mathcal{V}}, \ldots, \theta_{N,N}^{\mathcal{V}}) \in \mathbb{R}^{N}$, and the measure of model instability becomes
+An independence model then results for variables in $\boldsymbol X$, which has $q(N)=N$ parameters $\ak{\thetaN}^{\mathcal{V}} = (\theta_{1,\ak{\thetaidx}}^{\mathcal{V}}, \ldots, \theta_{N,\ak{\thetaidx}}^{\mathcal{V}}) \in \mathbb{R}^{N}$, and the measure of model instability becomes
 $$
-\frac{1}{N}\REP(\boldsymbol \theta_N) = \frac{2}{N} \sum_{i=1}^N|\theta_{i,N}^{\mathcal{V}}| \quad \text{(RBM model, no hiddens)}.
+\frac{1}{N}\REP(\ak{\thetaN}) = \frac{2}{N} \sum_{i=1}^N|\theta_{i,\ak{\thetaidx}}^{\mathcal{V}}| \quad \text{(RBM model, no hiddens)}.
 $$
 Hence, this model sequence for $\boldsymbol X$ will be  S-unstable model if the aggregation of absolute parameters grows faster than the number $N$ of parameters/visible variables.
-Consequently, even for a simplest RBM model involving independence, preventing instability requires careful choice of parameters, particularly with regard to how a parameter configuration differs from zero.
+Consequently, even for \ak{the} simplest RBM model involving independence, preventing instability requires careful choice of parameters, particularly with regard to how a parameter configuration differs from zero.
 For more general RBM models, the number $N_{\mathcal{H}}$ of hidden variables $\boldsymbol H$ can also be chosen arbitrarily (i.e., as some function $N_{\mathcal{H}}\equiv N_{N,\mathcal{H}}$ of $N$), which can substantially inflate the number $q(N)$ of model parameters and further impact model instability through accumulated parameters.
@@
 
-To state the result, let $\REP_{\boldsymbol X}(\boldsymbol \theta_N) \equiv \REP(\boldsymbol \theta_N)$ denote the LREP value \@ref(eq:elpr) from the marginal distribution $P_{\boldsymbol \theta_N}$ of visibles $\boldsymbol X$ in \@ref(eq:RBM2) and write the LREP for the joint distribution $\tilde{P}_{\boldsymbol \theta_N}$ of $(\boldsymbol X, \boldsymbol H)$ from \@ref(eq:RBM1) as
+To state the result, let $\REP_{\boldsymbol X}(\ak{\thetaN}) \equiv \REP(\ak{\thetaN})$ denote the LREP value \@ref(eq:elpr) from the marginal distribution $P_{\ak{\thetaN}}$ of visibles $\boldsymbol X$ in \@ref(eq:RBM2) and write the LREP for the joint distribution $\tilde{P}_{\ak{\thetaN}}$ of $(\boldsymbol X, \boldsymbol H)$ from \@ref(eq:RBM1) as
 \begin{align*}
-\REP_{(\boldsymbol X, \boldsymbol H)}(\boldsymbol \theta_N) &=\log\left[  \frac{\max_{ (\boldsymbol x, \boldsymbol h) \in \{\pm 1\}^{N+N_{\mathcal{H}}}}
-\tilde{P}_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)}{\min_{ (\boldsymbol x, \boldsymbol h) \in \{\pm 1\}^{N+N_{\mathcal{H}}}`}
-\tilde{P}_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)} \right] \qquad \text{(joint RBM model)}    \\    
-&= \left(\max_{ (\boldsymbol x, \boldsymbol h) \in \{\pm 1\}^{N+N_{\mathcal{H}}}} f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)-\min_{ (\boldsymbol x, \boldsymbol h) \in \{\pm 1\}^{N+N_{\mathcal{H}}}} f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)\right),
+\REP_{(\boldsymbol X, \boldsymbol H)}(\ak{\thetaN}) &=\log\left[  \frac{\max_{ (\boldsymbol x, \boldsymbol h) \in \{\pm 1\}^{N+N_{\mathcal{H}}}}
+\tilde{P}_{\ak{\thetaN}} (\boldsymbol x, \boldsymbol h)}{\min_{ (\boldsymbol x, \boldsymbol h) \in \{\pm 1\}^{N+N_{\mathcal{H}}}`}
+\tilde{P}_{\ak{\thetaN}} (\boldsymbol x, \boldsymbol h)} \right] \qquad \text{(joint RBM model)}    \\    
+&= \left(\max_{ (\boldsymbol x, \boldsymbol h) \in \{\pm 1\}^{N+N_{\mathcal{H}}}} f_{\ak{\thetaN}} (\boldsymbol x, \boldsymbol h)-\min_{ (\boldsymbol x, \boldsymbol h) \in \{\pm 1\}^{N+N_{\mathcal{H}}}} f_{\ak{\thetaN}} (\boldsymbol x, \boldsymbol h)\right),
 \end{align*}
@@
 (\#eq:f)
-f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)  \equiv  \sum_{i=1}^N x_i  \theta_{i,N}^{\mathcal{V}} +  \sum_{j=1}^{N_\mathcal{H}} h_j \theta_{j,N}^{\mathcal{H}} + \sum_{i=1}^N \sum_{j=1}^{N_\mathcal{H}} x_i h_j  \theta_{ij,N}^{\mathcal{VH}}
+f_{\ak{\thetaN}} (\boldsymbol x, \boldsymbol h)  \equiv  \sum_{i=1}^N x_i  \theta_{i,\ak{\thetaidx}}^{\mathcal{V}} +  \sum_{j=1}^{N_\mathcal{H}} h_j \theta_{j,\ak{\thetaidx}}^{\mathcal{H}} + \sum_{i=1}^N \sum_{j=1}^{N_\mathcal{H}} x_i h_j  \theta_{ij,\ak{\thetaidx}}^{\mathcal{VH}}
 \end{equation}
-of outcomes $\boldsymbol x =(x_1,\ldots,x_N) \in\{\pm1\}^{N}$, $\boldsymbol h =(h_1,\ldots,h_{N_{\mathcal{H}}})  \in\{\pm1\}^{N_{\mathcal{H}}}$ and parameters $\boldsymbol \theta_N \equiv (\boldsymbol \theta_N^{\mathcal{V}},\boldsymbol \theta_N^{\mathcal{H}}, \boldsymbol \theta_N^{\mathcal{VH}})$, with $\theta_{i,N}^{\mathcal{V}}$, $\theta_{j,N}^{\mathcal{H}}$ and $\theta_{ij,N}^{\mathcal{VH}}$ denoting respective parameter components, $1 \leq i \leq N$, $1 \leq j \leq N_{\mathcal{H}}$.
-Due to the marginalization steps in defining the distribution \@ref(eq:RBM2) of $\boldsymbol X$, note that $\REP_{\boldsymbol X}(\boldsymbol \theta_N)$ has no immediate analytical expression similar to that of  $\REP_{(\boldsymbol X, \boldsymbol H)}(\boldsymbol \theta_N)$.
-For clarity, recall also that S-instability \@ref(eq:Sun) in each model type refers to a respective divergence (i.e., $\lim_{N\to \infty} \REP_{(\boldsymbol X, \boldsymbol H)}(\boldsymbol \theta_N) /(N+N_{\mathcal{H}})=\infty$, $\lim_{N\to \infty} \REP(\boldsymbol \theta_N) /N=\infty$) upon scaling by the corresponding number of variables in a distribution.
+of outcomes $\boldsymbol x =(x_1,\ldots,x_N) \in\{\pm1\}^{N}$, $\boldsymbol h =(h_1,\ldots,h_{N_{\mathcal{H}}})  \in\{\pm1\}^{N_{\mathcal{H}}}$ and parameters $\ak{\thetaN} \equiv (\ak{\thetaN}^{\mathcal{V}}, \ak{\thetaN}^{\mathcal{H}}, \ak{\thetaN}^{\mathcal{VH}})$, with $\theta_{i,\ak{\thetaidx}}^{\mathcal{V}}$, $\theta_{j,\ak{\thetaidx}}^{\mathcal{H}}$ and $\theta_{ij,\ak{\thetaidx}}^{\mathcal{VH}}$ denoting respective parameter components, $1 \leq i \leq N$, $1 \leq j \leq N_{\mathcal{H}}$.
+Due to the marginalization steps in defining the distribution \@ref(eq:RBM2) of $\boldsymbol X$, note that $\REP_{\boldsymbol X}(\ak{\thetaN})$ has no immediate analytical expression similar to that of  $\REP_{(\boldsymbol X, \boldsymbol H)}(\ak{\thetaN})$.
+For clarity, recall also that S-instability \@ref(eq:Sun) in each model type refers to a respective divergence (i.e., $\lim_{N\to \infty} \REP_{(\boldsymbol X, \boldsymbol H)}(\ak{\thetaN}) /(N+N_{\mathcal{H}})=\infty$, $\lim_{N\to \infty} \REP(\ak{\thetaN}) /N=\infty$) upon scaling by the corresponding number of variables in a distribution.
 In the following, let $|\boldsymbol y|_1 = \sum_{i=1}^d |y_i|$ denote the L1 norm of a generic vector $\boldsymbol y =(y_1,\ldots,y_d)$, $d \geq 1$.
 ```{proposition, label="prop2", echo=TRUE}
-Let $P_{\boldsymbol \theta_N}$ denote a RBM-based data model \@ref(eq:RBM2) for $N\geq 1$ visible variables $\boldsymbol X \equiv \boldsymbol X_N$ derived from $\tilde{P}_{\boldsymbol \theta_N}$ as the joint RBM distribution \@ref(eq:RBM1) of $(\boldsymbol X, \boldsymbol H)$ involving some number $N_{\mathcal{H}} \equiv  N_{N,\mathcal{H}}\geq 0$ of hidden variables $\boldsymbol H \equiv \boldsymbol H_N$ and parameters $\boldsymbol \theta_N \equiv (\boldsymbol \theta_N^{\mathcal{V}},\boldsymbol \theta_N^{\mathcal{H}}, \boldsymbol \theta_N^{\mathcal{VH}}) \in\mathbb{R}^{N}\times \mathbb{R}^{N_{\mathcal{H}}} \times  \mathbb{R}^{N*N_{\mathcal{H}}}$.
+Let $P_{\ak{\thetaN}}$ denote a RBM-based data model \@ref(eq:RBM2) for $N\geq 1$ visible variables $\boldsymbol X \equiv \boldsymbol X_N$ derived from $\tilde{P}_{\ak{\thetaN}}$ as the joint RBM distribution \@ref(eq:RBM1) of $(\boldsymbol X, \boldsymbol H)$ involving some number $N_{\mathcal{H}} \equiv  N_{N,\mathcal{H}}\geq 0$ of hidden variables $\boldsymbol H \equiv \boldsymbol H_N$ and parameters $\ak{\thetaN} \equiv (\ak{\thetaN}^{\mathcal{V}},\ak{\thetaN}^{\mathcal{H}}, \ak{\thetaN}^{\mathcal{VH}}) \in\mathbb{R}^{N}\times \mathbb{R}^{N_{\mathcal{H}}} \times  \mathbb{R}^{N*N_{\mathcal{H}}}$.
 Then,
 \begin{enumerate}[(i)]
-\item the instability measure $\REP(\boldsymbol \theta_N)$ for the marginal model $P_{\boldsymbol \theta_N}$ of  $\boldsymbol X$ satisfies
+\item the instability measure $\REP(\ak{\thetaN})$ for the marginal model $P_{\ak{\thetaN}}$ of  $\boldsymbol X$ satisfies
     $$
-    \left| \REP(\boldsymbol \theta_N)  - \elt\right| \leq    N_{\mathcal{H}}  \log 2
+    \left| \REP(\ak{\thetaN})  - \elt\right| \leq    N_{\mathcal{H}}  \log 2
     $$
@@
     $$
-    \elt  \equiv   \max_{ \boldsymbol x} \max_{ \boldsymbol h  } f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)-\min_{ \boldsymbol x } \max_{ \boldsymbol h }f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)
+    \elt  \equiv   \max_{ \boldsymbol x} \max_{ \boldsymbol h  } f_{\ak{\thetaN}} (\boldsymbol x, \boldsymbol h)-\min_{ \boldsymbol x } \max_{ \boldsymbol h }f_{\ak{\thetaN}} (\boldsymbol x, \boldsymbol h)
     $$
-    based on $f_{\boldsymbol \theta_N}$ from (\ref{eq:f}) with components $\boldsymbol x \in \{\pm 1\}^{N}, \boldsymbol h \in \{\pm 1\}^{N_{\mathcal{H}}}$.
-\item The instability measure $\REP_{(\boldsymbol X, \boldsymbol H)}(\boldsymbol \theta_N)\equiv \left(\max_{ \boldsymbol x} \max_{ \boldsymbol h  } f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)-\min_{ \boldsymbol x } \min_{ \boldsymbol h }f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)\right)$ for the joint model $\tilde{P}_{\boldsymbol \theta_N}$ of $(\boldsymbol X, \boldsymbol H)$ satisfies
+    based on $f_{\ak{\thetaN}}$ from (\ref{eq:f}) with components $\boldsymbol x \in \{\pm 1\}^{N}, \boldsymbol h \in \{\pm 1\}^{N_{\mathcal{H}}}$.
+\item The instability measure $\REP_{(\boldsymbol X, \boldsymbol H)}(\ak{\thetaN})\equiv \left(\max_{ \boldsymbol x} \max_{ \boldsymbol h  } f_{\ak{\thetaN}} (\boldsymbol x, \boldsymbol h)-\min_{ \boldsymbol x } \min_{ \boldsymbol h }f_{\ak{\thetaN}} (\boldsymbol x, \boldsymbol h)\right)$ for the joint model $\tilde{P}_{\ak{\thetaN}}$ of $(\boldsymbol X, \boldsymbol H)$ satisfies
     \begin{align*}
-    2\Gam +  2|\boldsymbol \theta_N^{\mathcal{H}} |_{1} \; \geq \; \REP_{(\boldsymbol X, \boldsymbol H)}(\boldsymbol \theta_N)    & \geq 
-    2\max\big\{\Gam,   |\boldsymbol \theta_N^{\mathcal{H}} |_{1}\big\} \\
+    2\Gam +  2|\ak{\thetaN}^{\mathcal{H}} |_{1} \; \geq \; \REP_{(\boldsymbol X, \boldsymbol H)}(\ak{\thetaN})    & \geq 
+    2\max\big\{\Gam,   |\ak{\thetaN}^{\mathcal{H}} |_{1}\big\} \\
     &\geq  2\Gam\\
     &\geq  \elt\\
-    &\geq   \max\big\{  \Gamc, \, \Gam  - 2|\boldsymbol \theta_N^{\mathcal{H}} |_{1}  \big\}
+    &\geq   \max\big\{  \Gamc, \, \Gam  - 2|\ak{\thetaN}^{\mathcal{H}} |_{1}  \big\}
     \end{align*}
@@
     $$
-    \Gam \equiv \max_{ \boldsymbol h} k_{\boldsymbol \theta_N} (\boldsymbol h)
-    \geq  |\boldsymbol \theta_N^{\mathcal{V}} |_{1},\qquad  k_{\boldsymbol \theta_N} (\boldsymbol h)  \equiv \sum_{i=1}^{N }\left| \theta_{i,N}^{\mathcal{V}}   + \sum_{j=1}^{N_{\mathcal{H}}} h_j \theta_{ij,N}^{\mathcal{VH}} \right|,
+    \Gam \equiv \max_{ \boldsymbol h} k_{\ak{\thetaN}} (\boldsymbol h)
+    \geq  |\ak{\thetaN}^{\mathcal{V}} |_{1},\qquad  k_{\ak{\thetaN}} (\boldsymbol h)  \equiv \sum_{i=1}^{N }\left| \theta_{i,\ak{\thetaidx}}^{\mathcal{V}}   + \sum_{j=1}^{N_{\mathcal{H}}} h_j \theta_{ij,\ak{\thetaidx}}^{\mathcal{VH}} \right|,
     $$ 
-    and $\Gamc \equiv \min_{ \boldsymbol h} k_{\boldsymbol \theta_N} (\boldsymbol h)$ based on a function  $k_{\boldsymbol \theta_N} (\boldsymbol h)$ of hidden variable outcomes $\boldsymbol h = (h_1,\ldots,h_{N_{\mathcal{H}}})$ and visible-related parameters $\boldsymbol \theta_N^{\mathcal{V}}$ and $\boldsymbol \theta_N^{\mathcal{VH}}$.
+    and $\Gamc \equiv \min_{ \boldsymbol h} k_{\ak{\thetaN}} (\boldsymbol h)$ based on a function  $k_{\ak{\thetaN}} (\boldsymbol h)$ of hidden variable outcomes $\boldsymbol h = (h_1,\ldots,h_{N_{\mathcal{H}}})$ and visible-related parameters $\ak{\thetaN}^{\mathcal{V}}$ and $\ak{\thetaN}^{\mathcal{VH}}$.
 \item Assuming $\sup_{N \geq 1} N_{\mathcal{H}}/N<\infty$ additionally, then the following properties 1.-7. hold:
 \begin{enumerate}[1.] \itemsep 0cm
-    \item an S-unstable visible model $P_{\boldsymbol \theta_N}$ is equivalent to the condition $\lim_{N\to \infty}  \elt/N  =\infty$; further, $P_{\boldsymbol \theta_N}$ is stable when $\elt/N$, $N \geq 1$, is bounded.
-    \item an S-unstable joint model $P_{\boldsymbol \theta_N}$ is equivalent to the condition $\lim_{N\to \infty} \max\{|\boldsymbol \theta_N^{\mathcal{H}}|_1, \Gam\}/N =\infty$; further,  $\tilde{P}_{\boldsymbol \theta_N}$ is stable when $[|\boldsymbol \theta_N^{\mathcal{H}}|_1+ \Gam]/N$, $N \geq 1$, is bounded.
-    \item if the visible model $P_{\boldsymbol \theta_N}$   is S-unstable, then  the joint model $\tilde{P}_{\boldsymbol \theta_N}$ is also S-unstable.
-    \item when $\lim_{N\to \infty}  (|\boldsymbol \theta_N^{\mathcal{V}}|_1-2|\boldsymbol \theta_N^{\mathcal{H}}|_1)/N =\infty$, both $P_{\boldsymbol \theta_N}$ and $\tilde{P}_{\boldsymbol \theta_N}$ are necessarily S-unstable.
-    \item when $\lim_{N\to \infty}|\boldsymbol \theta_N^{\mathcal{H}}|_1/N =\infty$,  the joint model $\tilde{P}_{\boldsymbol \theta_N}$ is necessarily S-unstable.
-    \item when $\sup_{N \geq 1} |\boldsymbol \theta_N^{\mathcal{H}}|_1 /N<\infty$, the visible model $P_{\boldsymbol \theta_N}$ being S-stable or S-unstable is equivalent to the joint model  $\tilde{P}_{\boldsymbol \theta_N}$ being stable or unstable.
-    \item  an S-stable visible model $P_{\boldsymbol \theta_N}$ results if
+    \item an S-unstable visible model $P_{\ak{\thetaN}}$ is equivalent to the condition $\lim_{N\to \infty}  \elt/N  =\infty$; further, $P_{\ak{\thetaN}}$ is stable when $\elt/N$, $N \geq 1$, is bounded.
+    \item an S-unstable joint model $P_{\ak{\thetaN}}$ is equivalent to the condition $\lim_{N\to \infty} \max\{|\ak{\thetaN}^{\mathcal{H}}|_1, \Gam\}/N =\infty$; further,  $\tilde{P}_{\ak{\thetaN}}$ is stable when $[|\ak{\thetaN}^{\mathcal{H}}|_1+ \Gam]/N$, $N \geq 1$, is bounded.
+    \item if the visible model $P_{\ak{\thetaN}}$   is S-unstable, then  the joint model $\tilde{P}_{\ak{\thetaN}}$ is also S-unstable.
+    \item when $\lim_{N\to \infty}  (|\ak{\thetaN}^{\mathcal{V}}|_1-2|\ak{\thetaN}^{\mathcal{H}}|_1)/N =\infty$, both $P_{\ak{\thetaN}}$ and $\tilde{P}_{\ak{\thetaN}}$ are necessarily S-unstable.
+    \item when $\lim_{N\to \infty}|\ak{\thetaN}^{\mathcal{H}}|_1/N =\infty$,  the joint model $\tilde{P}_{\ak{\thetaN}}$ is necessarily S-unstable.
+    \item when $\sup_{N \geq 1} |\ak{\thetaN}^{\mathcal{H}}|_1 /N<\infty$, the visible model $P_{\ak{\thetaN}}$ being S-stable or S-unstable is equivalent to the joint model  $\tilde{P}_{\ak{\thetaN}}$ being stable or unstable.
+    \item  an S-stable visible model $P_{\ak{\thetaN}}$ results if
         $$
-        |\boldsymbol \theta_N^{\mathcal{V}}|_1+ |\boldsymbol \theta_N^{\mathcal{VH}} |_1  \leq CN,\quad N \geq 1,
+        |\ak{\thetaN}^{\mathcal{V}}|_1+ |\ak{\thetaN}^{\mathcal{VH}} |_1  \leq CN,\quad N \geq 1,
         $$
-        for some $C>0$, while an S-stable joint model $\tilde{P}_{\boldsymbol \theta_N}$  results if
+        for some $C>0$, while an S-stable joint model $\tilde{P}_{\ak{\thetaN}}$  results if
         $$
-        |\boldsymbol \theta_N|_1 \equiv  |\boldsymbol \theta_N^{\mathcal{V}}|_1+|\boldsymbol \theta_N^{\mathcal{H}}|_1 +|\boldsymbol \theta_N^{\mathcal{VH}} |_1  \leq  C N ,\quad N \geq 1.
+        |\ak{\thetaN}|_1 \equiv  |\ak{\thetaN}^{\mathcal{V}}|_1+|\ak{\thetaN}^{\mathcal{H}}|_1 +|\ak{\thetaN}^{\mathcal{VH}} |_1  \leq  C N ,\quad N \geq 1.
         $$
@@
 In Proposition \@ref(prp:prop2)(iii), the relationships between RBM models with regard to instability, and the effects of different parameter types, follow from the bounds on model instability measures in Proposition \@ref(prp:prop2)(i)-(ii).
-Generally speaking, all instability in the marginal RBM model for the data $\boldsymbol X$ can be attributed to an excessively large model quantity $\elt$, which predominantly follows when main $\boldsymbol \theta_N^{\mathcal{V}}$ and interaction $\boldsymbol \theta_N^{\mathcal{VH}}$ parameters related to visible variables are too large in magnitude (e.g., upon accumulation in terms such as $|\boldsymbol \theta_N^{\mathcal{V}}|_1$, $\Gam$ or $\Gamc$).
-For example, for any bounded sequence $|\boldsymbol \theta_N^{\mathcal{H}}|/N$ of hidden parameters, if main visible parameters $\boldsymbol \theta_N^{\mathcal{V}}$ are too extreme ($|\boldsymbol \theta_N^{\mathcal{V}}|_1/N\to \infty$), this aspect  will guarantee instability in the visible model ($\elt/N\to \infty$).
-In fact, the instability measure $\elt \equiv \max_{ \boldsymbol x} \max_{ \boldsymbol h  } f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)-\min_{ \boldsymbol x } \max_{ \boldsymbol h }f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)$ for marginal/visible model represents a clearly smaller portion of the  instability measure $\REP_{(\boldsymbol X, \boldsymbol H)}(\boldsymbol \theta_N)\equiv \max_{ \boldsymbol x} \max_{ \boldsymbol h  } f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)-\min_{ \boldsymbol x } \min_{ \boldsymbol h } f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)$ in the joint RBM model, implying that an unstable marginal model (i.e., due to  $\boldsymbol \theta_N^{\mathcal{V}}$, $\boldsymbol \theta_N^{\mathcal{VH}}$) must always translate to an unstable joint model and that further potential causes of instability exist for the joint model, often due to the size $|\boldsymbol \theta_N^{\mathcal{H}}|_1$.
-For example, while the joint RBM model for $(\boldsymbol X,\boldsymbol H)$ must always be unstable due to a diverging combination of visible and/or interaction parameters ($|\boldsymbol \theta_N^{\mathcal{V}}|_1/N\to \infty$ or $\Gam/N\to \infty$) (Proposition \@ref(prp:prop2)(iii.2)), instability for the joint model can also result when the main hidden parameters $\boldsymbol \theta_N^{\mathcal{H}}$ become too large relative to sample size ($|\boldsymbol \theta_N^{\mathcal{H}}|_1/N\to \infty$ in Proposition \@ref(prp:prop2)(iii.5)).
-However, under Proposition \@ref(prp:prop2), the main hidden parameters $\boldsymbol \theta_N^{\mathcal{H}}$  do not necessarily entail a source of instability for the marginal visible model.
-To explain this distinction, consider a joint model where all parameters related to visibles are zero, $\boldsymbol \theta_N^{\mathcal{V}}= \boldsymbol \theta_N^{\mathcal{VH}}=\boldsymbol 0$, but the hidden-related parameters diverge in sum $|\boldsymbol \theta_N^{\mathcal{H}}|_1/N\to \infty$.
-Here, the explosive behavior among parameters $\boldsymbol \theta_N^{\mathcal{H}}$ induces instability in the joint model for $(\boldsymbol X, \boldsymbol H)$ but the marginal model for $\boldsymbol X$, however, has a perfectly stable (and in fact uniform) distribution in this case.
-When the hidden parameters are bounded relative to the sample size ($\sup_{N\geq 1} |\boldsymbol \theta_N^{\mathcal{H}}|_1/N<\infty$), then all instability in both the joint and marginal RBM models can be directly linked to excessively large visible $\boldsymbol \theta_N^{\mathcal{V}}$ and/or interaction parameters $\boldsymbol \theta_N^{\mathcal{VH}}$ so that features of stability/instability must be the same across both models (Proposition \@ref(prp:prop2)(iii.6)).
-Hence, to prevent instability in the joint model, the combined magnitudes of all parameters $\theta_N$ must be controlled (cf.
-Proposition \@ref(prp:prop2)(iii.7)), while a stable visible data model technically results in constraining only the sizes of visible-related parameters $\boldsymbol \theta_N^{\mathcal{V}}$, $\boldsymbol \theta_N^{\mathcal{VH}}$.
+Generally speaking, all instability in the marginal RBM model for the data $\boldsymbol X$ can be attributed to an excessively large model quantity $\elt$, which predominantly follows when main $\\ak{\thetaN}^{\mathcal{V}}$ and interaction $\ak{\thetaN}^{\mathcal{VH}}$ parameters related to visible variables are too large in magnitude (e.g., upon accumulation in terms such as $|\ak{\thetaN}^{\mathcal{V}}|_1$, $\Gam$ or $\Gamc$).
+For example, for any bounded sequence $|\ak{\thetaN}^{\mathcal{H}}|/N$ of hidden parameters, if main visible parameters $\ak{\thetaN}^{\mathcal{V}}$ are too extreme ($|\ak{\thetaN}^{\mathcal{V}}|_1/N\to \infty$), this aspect  will guarantee instability in the visible model ($\elt/N\to \infty$).
+In fact, the instability measure $\elt \equiv \max_{ \boldsymbol x} \max_{ \boldsymbol h  } f_{\ak{\thetaN}} (\boldsymbol x, \boldsymbol h)-\min_{ \boldsymbol x } \max_{ \boldsymbol h }f_{\ak{\thetaN}} (\boldsymbol x, \boldsymbol h)$ for marginal/visible model represents a clearly smaller portion of the  instability measure $\REP_{(\boldsymbol X, \boldsymbol H)}(\ak{\thetaN})\equiv \max_{ \boldsymbol x} \max_{ \boldsymbol h  } f_{\ak{\thetaN}} (\boldsymbol x, \boldsymbol h)-\min_{ \boldsymbol x } \min_{ \boldsymbol h } f_{\ak{\thetaN}} (\boldsymbol x, \boldsymbol h)$ in the joint RBM model, implying that an unstable marginal model (i.e., due to  $\ak{\thetaN}^{\mathcal{V}}$, $\ak{\thetaN}^{\mathcal{VH}}$) must always translate to an unstable joint model and that further potential causes of instability exist for the joint model, often due to the size $|\ak{\thetaN}^{\mathcal{H}}|_1$.
+For example, while the joint RBM model for $(\boldsymbol X,\boldsymbol H)$ must always be unstable due to a diverging combination of visible and/or interaction parameters ($|\ak{\thetaN}^{\mathcal{V}}|_1/N\to \infty$ or $\Gam/N\to \infty$) (Proposition \@ref(prp:prop2)(iii.2)), instability for the joint model can also result when the main hidden parameters $\ak{\thetaN}^{\mathcal{H}}$ become too large relative to sample size ($|\ak{\thetaN}^{\mathcal{H}}|_1/N\to \infty$ in Proposition \@ref(prp:prop2)(iii.5)).
+However, under Proposition \@ref(prp:prop2), the main hidden parameters $\ak{\thetaN}^{\mathcal{H}}$  do not necessarily entail a source of instability for the marginal visible model.
+To explain this distinction, consider a joint model where all parameters related to visibles are zero, $\ak{\thetaN}^{\mathcal{V}}= \ak{\thetaN}^{\mathcal{VH}}=\boldsymbol 0$, but the hidden-related parameters diverge in sum $|\ak{\thetaN}^{\mathcal{H}}|_1/N\to \infty$.
+Here, the explosive behavior among parameters $\ak{\thetaN}^{\mathcal{H}}$ induces instability in the joint model for $(\boldsymbol X, \boldsymbol H)$ but the marginal model for $\boldsymbol X$, however, has a perfectly stable (and in fact uniform) distribution in this case.
+When the hidden parameters are bounded relative to the sample size ($\sup_{N\geq 1} |\ak{\thetaN}^{\mathcal{H}}|_1/N<\infty$), then all instability in both the joint and marginal RBM models can be directly linked to excessively large visible $\ak{\thetaN}^{\mathcal{V}}$ and/or interaction parameters $\ak{\thetaN}^{\mathcal{VH}}$ so that features of stability/instability must be the same across both models (Proposition \@ref(prp:prop2)(iii.6)).
+Hence, to prevent instability in the joint model, the combined magnitudes of all parameters $\ak{\thetaN}$ must be controlled (cf.
+Proposition \@ref(prp:prop2)(iii.7)), while a stable visible data model technically results in constraining only the sizes of visible-related parameters $\ak{\thetaN}^{\mathcal{V}}$, $\ak{\thetaN}^{\mathcal{VH}}$.
 Nevertheless, because the joint model is often employed in practice for purposes of simulation and simulation-based inference, it is still reasonable to consider parameter choices for ensuring a stable joint model (and, consequently, a stable visible model as well).
@@
 ```{r rbm-plots, fig.show='hold', fig.cap="The sample mean value of $\\text{ELPR}(\\boldsymbol \\theta)/N_{\\mathcal{V}}$ (left) and $\\DN(\\boldsymbol \\theta)$ at each grid point for each combination of magnitude of $\\boldsymbol \\theta$.
-As the magnitude of $\\boldsymbol \\theta$ grows, so does the value of these metrics, indicating typical instability in the model.", fig.height=3.5, fig.align="ht!"}
+As the magnitude of $\\boldsymbol \\theta$ grows, so does the value of these metrics, indicating \\ak{instability} in the model.", fig.height=3.5, fig.align="ht!"}
 elpr_summary %>%
@@
 
-In our numerical experiment, we allow the two types of terms (main effects terms corresponding to visible and hidden parameters $\boldsymbol \theta_{main} = (\boldsymbol \theta_N^{\mathcal{V}}, \boldsymbol \theta_N^{\mathcal{H}})$ and interaction parameters $\boldsymbol \theta_N^{\mathcal{VH}}$) to have varying average magnitudes, $||\boldsymbol \theta_{main} || /(N_{\mathcal{H}}+N_{\mathcal{V}})$ and $||\boldsymbol \theta_{interaction} || /(N_{\mathcal{H}}*N_{\mathcal{V}})$ for a RBM with $N_\mathcal{V} = `r V`$ visibles and $N_\mathcal{H} = `r H`$ hiddens.
+In our numerical experiment, we allow the two types of terms (main effects terms corresponding to visible and hidden parameters $\boldsymbol \theta_{main} = (\ak{\thetaN}^{\mathcal{V}}, \ak{\thetaN}^{\mathcal{H}})$ and interaction parameters $\ak{\thetaN}^{\mathcal{VH}}$) to have varying average magnitudes, $||\boldsymbol \theta_{main} || /(N_{\mathcal{H}}+N_{\mathcal{V}})$ and $||\boldsymbol \theta_{interaction} || /(N_{\mathcal{H}}*N_{\mathcal{V}})$ for a RBM with $N_\mathcal{V} = `r V`$ visibles and $N_\mathcal{H} = `r H`$ hiddens.
 These average magnitudes vary on a grid between $`r min(elpr_summary$r1/(elpr_summary$H + elpr_summary$V))`$ and $`r max(elpr_summary$r1/(elpr_summary$H + elpr_summary$V))`$ with $`r length(unique(elpr_summary$r1/(elpr_summary$H + elpr_summary$V)))`$ breaks, yielding $`r length(unique(elpr_summary$r1/(elpr_summary$H + elpr_summary$V)))^2`$ grid points.
 At each point in the grid, $`r n`$ vectors ($\boldsymbol \theta_{main}$) are sampled uniformly on a sphere with radius corresponding to the first coordinate in the grid and $`r n`$ vectors ($\boldsymbol \theta_{interaction}$) are sampled uniformly on a sphere with radius corresponding to the second coordinate in the grid via sums of squared and scaled iid Normal$(0, 1)$ variables.
-These vectors are then paired to create $`r n`$ values of $\boldsymbol \theta_N$ with magnitudes at each point in the grid.
-The values $\REP(\boldsymbol \theta_N)/N_{\mathcal{V}}$ and $\DN(\boldsymbol \theta_N)$ are then calculated for each $\boldsymbol \theta_N$ and then summarized for each point in the grid using the sample mean.
+These vectors are then paired to create $`r n`$ values of $\ak{\thetaN}$ with magnitudes at each point in the grid.
+The values $\REP(\ak{\thetaN})/N_{\mathcal{V}}$ and $\DN(\ak{\thetaN})$ are then calculated for each $\ak{\thetaN}$ and then summarized for each point in the grid using the sample mean.
 The results of this numerical study are shown in Figure \@ref(fig:rbm-plots).
-From these two plots, it is clear that for larger magnitudes of the parameter vectors, there is evidence of S-instability in that the log-ratio of extremal probabilities scaled by $N_{\mathcal{V}}$ and the the biggest log-probability ratio for a one-component change in data outcomes are both increasing away from $\boldsymbol \theta_N = \boldsymbol 0$, further supporting \@ref(prp:prop2)(iii.2 and iii.5).
+From these two plots, it is clear that for larger magnitudes of the parameter vectors, there is evidence of S-instability in that the log-ratio of extremal probabilities scaled by $N_{\mathcal{V}}$ and the the biggest log-probability ratio for a one-component change in data outcomes are both increasing away from $\ak{\thetaN} = \boldsymbol 0$, further supporting \@ref(prp:prop2)(iii.2 and iii.5).
 
-In more complicated graphical models involving further or deeper hidden layers, the same issues and causes of instability similarly exist, but are compounded by a greater number of model parameters.
-S-unstable joint models will similarly follow if the combined sizes of all parameters are too great relative to the total number of variables, while instability in the data model for visible variables will depend only on the  main or interaction parameters directly related to visibles and how their accumulated magnitude compares to the observation sample size $N$.
+In more complicated graphical models involving further or deeper hidden layers, the same issues and causes of instability similarly exist, but are compounded by a greater number of model parameters \ak{and depend on the configuration of the particular models in consideration.
+Consider the following two models -- a RBM with $N$ visible nodes and one hidden layer with $N_H$ nodes and a deep RBM with $N$ visible nodes and $2$ layers, with $N_{H,1}$ and $N_{H,2}$ hidden nodes respectively such that $N_{H,1} + N_{H,2} = N_H$.
+The length of the parameter vector $\ak{\thetaN}$ in the single layer RBM is then equal to $N + N_H + N*N_H$, whereas the length of the parameter vector in the two layer RBM is $N + N_{H,1} + N_{H,2} + N*N_{H,1} + N_{H,1}*N_{H,2} = N + N_H + N*N_H + N_{H,2}(N_{H,1} - N)$.
+If $N_{H,1} > N$, then the deep model will have more parameters, if $N_{H,1} < N$, then the deep model will have less parameters, and if $N_{H,1} = N$, then the deep model will the same number of parameters as the single layer RBM.} S-unstable joint models will similarly follow if the combined sizes of all parameters are too great relative to the total number of variables, while instability in the data model for visible variables will depend only on the main or interaction parameters directly related to visibles \ak{(i.e.
+all the parameters for the single layer RBM and only the parameters for the visible layer, layer $H^{(1)}$, and the interactions between them for the two layer RBM)} and how their accumulated magnitude compares to the observation sample size $N$.
+\ak{This indicates that for the name number of hidden variables $N_H$, a deeper, rather than a wider, structure can be beneficial when considering instability in the data model for visible variables.
+However, as the size of a model considered grows, this point becomes moot.}
 
@@
 
-Due to their induced sensitivity in probability structure, S-instability in FOES models may often translate to numerical complications, and in fact obstructions, in both simulation and statistical inference based on likelihoods.
+Due to \ak{induced sensitivity in the probability structure of FOES models, S-instability} may often translate to numerical complications, and in fact obstructions, in both simulation and statistical inference based on likelihoods.
 We describe these aspects in Sections \@ref(mcmc)-\@ref(bayes) with regard to data simulation, maximum likelihood estimation and Bayes inference, respectively.
@@
 
-Suppose one aims to apply MCMC to simulate data $\boldsymbol X=(X_1,\ldots,X_n)$ from a FOES model $P_{\boldsymbol \theta_N}(\boldsymbol x)$, $\boldsymbol x \in\mathcal{X}^N$ whereby a chain is constructed with $P_{\boldsymbol \theta_N}$ as the stationary distribution.
+Suppose one aims to apply MCMC to simulate data $\boldsymbol X=(X_1,\ldots,X_n)$ from a FOES model $P_{\ak{\thetaN}}(\boldsymbol x)$, $\boldsymbol x \in\mathcal{X}^N$ whereby a chain is constructed with $P_{\ak{\thetaN}}$ as the stationary distribution.
 For an unstable FOES model, one-component changes in outcomes may produce
 radically different probabilities, which may then impose numerical barriers to MCMC.
-For example, consider implementation of a Gibbs sampler from the full conditional distributions $P_{\boldsymbol \theta_N} (X_i = x| \boldsymbol x_{-i})$, $x\in\mathcal{X}$, of each variable $X_i$ based on values $\boldsymbol x_{-i}\in \mathcal{X}^{N-1}$ for the remaining variables, say $\boldsymbol X_{-i}$, in $\boldsymbol X$.
-If a single change in $X_i$ from one value $x_1$ to another $x_2$ may produce two outcomes $\boldsymbol x^{(1)}$ and $\boldsymbol x^{(2)}$ for $\boldsymbol X$ with vastly different probabilities under the joint distribution $P_{\boldsymbol \theta_N}$, then the Gibbs sampler can have extreme log-ratios in its transition probabilities,
+For example, consider implementation of a Gibbs sampler from the full conditional distributions $P_{\ak{\thetaN}} (X_i = x| \boldsymbol x_{-i})$, $x\in\mathcal{X}$, of each variable $X_i$ based on values $\boldsymbol x_{-i}\in \mathcal{X}^{N-1}$ for the remaining variables, say $\boldsymbol X_{-i}$, in $\boldsymbol X$.
+If a single change in $X_i$ from one value $x_1$ to another $x_2$ may produce two outcomes $\boldsymbol x^{(1)}$ and $\boldsymbol x^{(2)}$ for $\boldsymbol X$ with vastly different probabilities under the joint distribution $P_{\ak{\thetaN}}$, then the Gibbs sampler can have extreme log-ratios in its transition probabilities,
 $$
-\log \left| \frac{P_{\boldsymbol \theta_N} (X_i =x_1 | \boldsymbol x^{(1)}_{-i})}{P_{\boldsymbol \theta_N} (X_i = x_2| \boldsymbol x^{(2)}_{-i}) } \right|=\log \left| \frac{P_{\boldsymbol \theta_N} ( \boldsymbol x^{(1)} )}{P_{\boldsymbol \theta_N} (  \boldsymbol x^{(2)} ) } \right|,
+\log \left| \frac{P_{\ak{\thetaN}} (X_i =x_1 | \boldsymbol x^{(1)}_{-i})}{P_{\ak{\thetaN}} (X_i = x_2| \boldsymbol x^{(2)}_{-i}) } \right|=\log \left| \frac{P_{\ak{\thetaN}} ( \boldsymbol x^{(1)} )}{P_{\ak{\thetaN}} (  \boldsymbol x^{(2)} ) } \right|,
 $$  
@@
 This mixing problem is due to the unstable stationary distribution (unbounded ratios of probabilities under the joint model), rather than in any particulars of the MCMC algorithm, and similar complications can also arise for Metropolis-Hastings algorithms for MCMC.
-Hence, while an unstable FOES model $P_{\boldsymbol \theta_N}$ may be valid and technically open to simulation by MCMC, the aspect of instability can render applications of MCMC as numerically infeasible for simulation purposes.
+Hence, while an unstable FOES model $P_{\ak{\thetaN}}$ may be valid and technically open to simulation by MCMC, the aspect of instability can render applications of MCMC as numerically infeasible for simulation purposes.
 This result is in line with conclusions of @handcock2003assessing and @schweinberger2011instability for other exponential models.
@@
 Volatility in the probability structure of an unstable model can also hamper efforts to maximize likelihood functions in statistical inference.
-When a FOES model is unstable along a parameter sequence \(\boldsymbol \theta_N\), the same model can further be unstable along parameters \(-\boldsymbol \theta_N\) in an opposite direction from the
-origin (model condition PSR and Corollary \@ref(cor:sign)). This can translate into
-potential sensitivity of the likelihood function around zero, and lead to numerical complications in maximizing the objective function.
+When a FOES model is unstable along a parameter sequence \(\ak{\thetaN}\), the same model can further be unstable along parameters \(-\ak{\thetaN}\) in an opposite direction from the origin (model condition PSR and Corollary \@ref(cor:sign)).
+This can translate into potential sensitivity of the likelihood function around zero, and lead to numerical complications in maximizing the objective function.
 We next provide a discussion of this issue in a way that builds upon and extends related findings by @schweinberger2011instability, who largely focused on the case of one-parameter exponential models.
 
-With many probability models, the modes and anti-modes in the probability structure under one parameter $\boldsymbol \theta_N$ are reversed in role when the parameter sign changes \(-\boldsymbol \theta_N\).
+With many probability models, the modes and anti-modes in the probability structure under one parameter $\ak{\thetaN}$ are reversed in role when the parameter sign changes \(-\ak{\thetaN}\).
 Because unstable models tend to degeneracy, the opposite signed parameters further push unstable models to assign nearly all probability to extremely opposite data configurations, given by modes/anti-modes.
@@
 ```{theorem, label="something", echo=TRUE}
-Let $P_{\boldsymbol \theta_N}$, $N \geq 1$, denote an S-unstable FOES model sequence, which additionally satisfies model condition PSR.
-Let $\boldsymbol{x}_{\max, \boldsymbol \theta_N},\boldsymbol{x}_{\min, \boldsymbol \theta_N}\in\mathcal{X}^N$ denote, respectively, a mode and anti-mode of the model $P_{\boldsymbol \theta_N}(\boldsymbol x)$, $\boldsymbol x\in\mathcal{X}^N$, for $N$ observations $\boldsymbol X = (X_1,\ldots,X_N)$, whereby $P_{\boldsymbol \theta_N }(\boldsymbol{x}_{\max, \boldsymbol \theta_N}) = \max_{\boldsymbol y\in\mathcal{X}^N} P_{\boldsymbol \theta_N }(\boldsymbol y)$ and $P_{\boldsymbol \theta_N }(\boldsymbol{x}_{\min, \boldsymbol \theta_N}) = \min_{\boldsymbol y\in\mathcal{X}^N} P_{\boldsymbol \theta_N }(\boldsymbol y)$.
+Let $P_{\ak{\thetaN}}$, $N \geq 1$, denote an S-unstable FOES model sequence, which additionally satisfies model condition PSR.
+Let $\boldsymbol{x}_{\max, \ak{\thetaN}},\boldsymbol{x}_{\min,  \ak{\thetaN}}\in\mathcal{X}^N$ denote, respectively, a mode and anti-mode of the model $P_{\ak{\thetaN}}(\boldsymbol x)$, $\boldsymbol x\in\mathcal{X}^N$, for $N$ observations $\boldsymbol X = (X_1,\ldots,X_N)$, whereby $P_{\ak{\thetaN} }(\boldsymbol{x}_{\max, \ak{\thetaN}}) = \max_{\boldsymbol y\in\mathcal{X}^N} P_{\ak{\thetaN} }(\boldsymbol y)$ and $P_{\ak{\thetaN} }(\boldsymbol{x}_{\min, \ak{\thetaN}}) = \min_{\boldsymbol y\in\mathcal{X}^N} P_{\ak{\thetaN} }(\boldsymbol y)$.
 
@@
 $$
-\frac{1}{\REP(\boldsymbol \theta_N)} \log\left[ \frac{P_{\boldsymbol \theta_N }(\boldsymbol X) }{ \displaystyle{\min_{\boldsymbol y\in\mathcal{X}^N}} P_{\boldsymbol \theta_N }(\boldsymbol y)}\right] = \frac{ \log P_{\boldsymbol \theta_N } ( \boldsymbol X)  - \log P_{\boldsymbol \theta_N }(\boldsymbol{x}_{\min, \boldsymbol \theta_N}) }
-{ \log P_{\boldsymbol \theta_N } (\boldsymbol{x}_{\max, \boldsymbol \theta_N})  - \log P_{\boldsymbol \theta_N }(\boldsymbol{x}_{\min, \boldsymbol \theta_N}) }
+\frac{1}{\REP(\ak{\thetaN})} \log\left[ \frac{P_{\ak{\thetaN} }(\boldsymbol X) }{ \displaystyle{\min_{\boldsymbol y\in\mathcal{X}^N}} P_{\ak{\thetaN} }(\boldsymbol y)}\right] = \frac{ \log P_{\ak{\thetaN} } ( \boldsymbol X)  - \log P_{\ak{\thetaN} }(\boldsymbol{x}_{\min, \ak{\thetaN}}) }
+{ \log P_{\ak{\thetaN} } (\boldsymbol{x}_{\max,\ak{\thetaN}})  - \log P_{\ak{\thetaN} }(\boldsymbol{x}_{\min, \ak{\thetaN}}) }
 \stackrel{p, \mathrm{E}}{\longrightarrow}
@@
 $$
-under $\boldsymbol\theta_N$ while
+under $\ak{\thetaN}$ while
 $$
-\frac{1}{\REP(-\boldsymbol \theta_N)} \log\left[ \frac{P_{-\boldsymbol \theta_N }(\boldsymbol X) }{  \displaystyle{\min_{\boldsymbol y\in\mathcal{X}^N}} P_{-\boldsymbol \theta_N }(\boldsymbol y)} \right]
-=    \frac{ \log P_{-\boldsymbol \theta_N } ( \boldsymbol X) - \log P_{ \boldsymbol \theta_N }(\boldsymbol{x}_{\max, \boldsymbol \theta_N})    }
-{ \log P_{-\boldsymbol \theta_N } (\boldsymbol{x}_{\min, \boldsymbol \theta_N})  - \log P_{-\boldsymbol \theta_N }(\boldsymbol{x}_{\max, \boldsymbol \theta_N}) }
+\frac{1}{\REP(-\ak{\thetaN})} \log\left[ \frac{P_{-\ak{\thetaN} }(\boldsymbol X) }{  \displaystyle{\min_{\boldsymbol y\in\mathcal{X}^N}} P_{-\ak{\thetaN} }(\boldsymbol y)} \right]
+=    \frac{ \log P_{-\ak{\thetaN} } ( \boldsymbol X) - \log P_{\ak{\thetaN} }(\boldsymbol{x}_{\max, \ak{\thetaN}})    }
+{ \log P_{-\ak{\thetaN} } (\boldsymbol{x}_{\min, \ak{\thetaN}})  - \log P_{-\ak{\thetaN} }(\boldsymbol{x}_{\max, \ak{\thetaN}}) }
 \stackrel{p,\mathrm{E}}{\longrightarrow}
@@
 $$
-under $-\boldsymbol\theta_N$, where
+under $-\ak{\thetaN}$, where
 $$
-\REP( \boldsymbol \theta_N) \equiv \log \frac{P_{\boldsymbol \theta_N } (\boldsymbol{x}_{\max, \boldsymbol \theta_N})}{P_{\boldsymbol \theta_N }(\boldsymbol{x}_{\min, \boldsymbol \theta_N})} = \log  \frac{P_{-\boldsymbol \theta_N } (\boldsymbol{x}_{\min, \boldsymbol \theta_N}) }{ P_{-\boldsymbol \theta_N }(\boldsymbol{x}_{\max, \boldsymbol \theta_N})}=   \REP( -\boldsymbol \theta_N), \quad N \geq 1.
+\REP( \ak{\thetaN}) \equiv \log \frac{P_{\ak{\thetaN} } (\boldsymbol{x}_{\max, \ak{\thetaN}})}{P_{\ak{\thetaN} }(\boldsymbol{x}_{\min, \ak{\thetaN}})} = \log  \frac{P_{-\ak{\thetaN} } (\boldsymbol{x}_{\min, \ak{\thetaN}}) }{ P_{-\ak{\thetaN} }(\boldsymbol{x}_{\max, \ak{\thetaN}})}=   \REP( -\ak{\thetaN}), \quad N \geq 1.
 $$
@@
 
-Theorem \@ref(thm:something) entails log-likelihood functions based on unstable models are both inversely related and degenerate at opposited signed parameters $\boldsymbol \theta_N$ or $-\boldsymbol \theta_N$, so that likelihoods are highest at different extremes in data configuration (e.g., $\boldsymbol{x}_{\max, \boldsymbol \theta_N}$ under $\boldsymbol \theta$-probabilities or $\boldsymbol{x}_{\min, \boldsymbol \theta_N}$ under $-\boldsymbol \theta$-probabilities).
-If the observed outcome $\boldsymbol x$ for data $\boldsymbol X$ is not a mode/anti-mode, then probabilities for the outcome may be small under both parameters $\boldsymbol \theta_N$ and $-\boldsymbol \theta_N$, in which case associated optimization steps may then shift around zero and struggle to converge.
+Theorem \@ref(thm:something) \ak{implies} log-likelihood functions based on unstable models are both inversely related and degenerate at opposited signed parameters $\ak{\thetaN}$ or $-\ak{\thetaN}$, so that likelihoods are highest at different extremes in data configuration (e.g., $\boldsymbol{x}_{\max, \ak{\thetaN}}$ under $\boldsymbol \theta$-probabilities or $\boldsymbol{x}_{\min, \ak{\thetaN}}$ under $-\boldsymbol \theta$-probabilities).
+If the observed outcome $\boldsymbol x$ for data $\boldsymbol X$ is not a mode/anti-mode, then probabilities for the outcome may be small under both parameters $\ak{\thetaN}$ and $-\ak{\thetaN}$, in which case associated optimization steps may then shift around zero and struggle to converge.
 
-In many model formulations, the zero parameter $\boldsymbol \theta_N=\boldsymbol 0$ is a "safe" position among parameters, representing a  guaranteed stable model (having uniform probabilities among outcomes), which can also  tether a broad parameter search attempted among unstable models.
+In many model formulations, the zero parameter $\ak{\thetaN}=\boldsymbol 0$ is a "safe" position among parameters, representing a  guaranteed stable model (having uniform probabilities among outcomes), which can also  tether a broad parameter search attempted among unstable models.
 @handcock2003assessing describes similar results for degenerate exponential models, and Theorem \@ref(thm:something) also supports an important finding of @schweinberger2011instability [Corollary 1] for one-parameter linear exponential models \@ref(eq:expo).
-In the latter case, the likelihood score function at $\boldsymbol \theta_N$ is the expected value $\mu(\boldsymbol \theta_N)\equiv \mathrm{E}_{\boldsymbol \theta_N} g(\boldsymbol X)$  of the sufficient statistic $g(\cdot)$, and optimization involves solving $\mu(\cdot)=g(\boldsymbol x)$ for an observed outcome $\boldsymbol x$.
+In the latter case, the likelihood score function at $\ak{\thetaN}$ is the expected value $\mu( \ak{\thetaN})\equiv \mathrm{E}_{\ak{\thetaN}} g(\boldsymbol X)$  of the sufficient statistic $g(\cdot)$, and optimization involves solving $\mu(\cdot)=g(\boldsymbol x)$ for an observed outcome $\boldsymbol x$.
 For unstable  models in this exponential class, @schweinberger2011instability [Corollary 1] shows
@@
 $$
-\lim_{N\to \infty}\frac{\mu(\boldsymbol \theta_N) -L_n}{U_n-L_N}= \left\{ \begin{array}{lcl}
-1 && \text{for } \boldsymbol \theta_N>0,\\
-0 && \text{for } \boldsymbol \theta_N<0, \end{array}\right.
+\lim_{N\to \infty}\frac{\mu(\ak{\thetaN}) -L_n}{U_n-L_N}= \left\{ \begin{array}{lcl}
+1 && \text{for } \ak{\thetaN}>0,\\
+0 && \text{for } \ak{\thetaN}<0, \end{array}\right.
 $$
@@
 As described by @schweinberger2011instability, the implication for maximum likelihood estimation is that, unless an observed outcome $\boldsymbol x$ falls at an extreme $U_N, L_N$ (i.e., modes/anti-modes), optimization steps in the parameter space can iterate in relatively small increments around zero and fail to converge.
-For unstable one-parameter exponential models, the maximum likelihood results of @schweinberger2011instability turn out to be a special case of Theorem \@ref(thm:something) and the LREP expansion \@ref(eq:UL) in this setting; namely, for an unstable model with $\boldsymbol \theta_N>0$,
+For unstable one-parameter exponential models, the maximum likelihood results of @schweinberger2011instability turn out to be a special case of Theorem \@ref(thm:something) and the LREP expansion \@ref(eq:UL) in this setting; namely, for an unstable model with $\ak{\thetaN}>0$,
 $$
-\frac{1}{\REP(\boldsymbol \theta_N)} \log\left[ \frac{P_{\boldsymbol \theta_N }(\boldsymbol X) }{\displaystyle{\min_{\boldsymbol y\in\mathcal{X}^N}} P_{\boldsymbol \theta_N }(\boldsymbol y)}\right] = \frac{\boldsymbol g(\boldsymbol X) - L_N}{U_N-L_N}\;\;\stackrel{p, \mathrm{E}}{\longrightarrow}\;\; 1
+\frac{1}{\REP(\ak{\thetaN})} \log\left[ \frac{P_{\ak{\thetaN} }(\boldsymbol X) }{\displaystyle{\min_{\boldsymbol y\in\mathcal{X}^N}} P_{ \ak{\thetaN} }(\boldsymbol y)}\right] = \frac{\boldsymbol g(\boldsymbol X) - L_N}{U_N-L_N}\;\;\stackrel{p, \mathrm{E}}{\longrightarrow}\;\; 1
 $$ 
-holds as $N\to \infty$ by Theorem \@ref(thm:something), while under $-\boldsymbol \theta_N<0$
+holds as $N\to \infty$ by Theorem \@ref(thm:something), while under $-\ak{\thetaN}<0$
 $$
-\frac{1}{\REP(-\boldsymbol \theta_N)} \log\left[ \frac{P_{-\boldsymbol \theta_N }(\boldsymbol X) }{\displaystyle{\min_{\boldsymbol y\in\mathcal{X}^N}} P_{-\boldsymbol \theta_N }(\boldsymbol y)}\right] = \frac{U_N -\boldsymbol g(\boldsymbol X)}{U_N-L_N} = 1 - \frac{\boldsymbol g(\boldsymbol X) - L_N}{U_N-L_N} \;\;\stackrel{p, \mathrm{E}}{\longrightarrow}\;\; 1.
+\frac{1}{\REP(-\ak{\thetaN})} \log\left[ \frac{P_{-\ak{\thetaN} }(\boldsymbol X) }{\displaystyle{\min_{\boldsymbol y\in\mathcal{X}^N}} P_{-\ak{\thetaN} }(\boldsymbol y)}\right] = \frac{U_N -\boldsymbol g(\boldsymbol X)}{U_N-L_N} = 1 - \frac{\boldsymbol g(\boldsymbol X) - L_N}{U_N-L_N} \;\;\stackrel{p, \mathrm{E}}{\longrightarrow}\;\; 1.
 $$
@@
 The potential challenges in chain mixing are similar to those presented in Section \@ref(mcmc), though in chain movements through the parameter space as opposed to the sample space in data generation.
-That is, in the Bayes setting for sampling a posterior distribution for $\boldsymbol \theta_N$,  a chain may  unstable to effectively explore the parameter space due partly to extreme and potentially unbounded probability ratios from parameter sign changes, which represents a parameter space analog to how one-component changes in the sample space may impact data simulation with unstable models.
-For example, if $\pi(\cdot)$ denotes a prior density for $\boldsymbol \theta_N$ and $q(\cdot | \cdot)$  denotes a proposal distribution  for use in a Metropolis-Hastings (MH) sampler, then MH acceptance probability becomes
+That is, in the Bayes setting for sampling a posterior distribution for $\ak{\thetaN}$,  a chain may \ak{be unable} to effectively explore the parameter space due partly to extreme and potentially unbounded probability ratios from parameter sign changes, which represents a parameter space analog to how one-component changes in the sample space may impact data simulation with unstable models.
+For example, if $\pi(\cdot)$ denotes a prior density for $\ak{\thetaN}$ and $q(\cdot | \cdot)$  denotes a proposal distribution  for use in a Metropolis-Hastings (MH) sampler, then MH acceptance probability becomes
 $$
 \alpha\left(\boldsymbol \theta^{(1)} \mid \boldsymbol \theta^{(2)}\right)= \min\left\{1,
-\frac{q(\boldsymbol \theta_N^{(2)} \mid \boldsymbol \theta_N^{(1)})}{q(\boldsymbol \theta_N^{(1)} \mid \boldsymbol \theta_N^{(2)})}
-\frac{P_{\boldsymbol \theta_N^{(1)}} ( \boldsymbol x ) \pi(\boldsymbol \theta_N^{(1)}) }{P_{\boldsymbol \theta^{(2)}_N} (  \boldsymbol x) \pi(\boldsymbol \theta_N^{(2)}) } \right\},
+\frac{q(\ak{\thetaN}^{(2)} \mid \ak{\thetaN}^{(1)})}{q(\ak{\thetaN}^{(1)} \mid  \ak{\thetaN}^{(2)})}
+\frac{P_{\ak{\thetaN}^{(1)}} ( \boldsymbol x ) \pi(\ak{\thetaN}^{(1)}) }{P_{\ak{\thetaN}^{(2)}} (  \boldsymbol x) \pi(\ak{\thetaN}^{(2)}) } \right\},
 $$
-which indicates how parameter sensitivity in the likelihood $P_{\boldsymbol \theta_N} (  \boldsymbol x)$ may complicate sampling of the posterior $P_{\boldsymbol \theta_N} (  \boldsymbol x) \pi(\boldsymbol \theta_N)$ (i.e., moving from $\boldsymbol \theta^{(1)}$ to $\boldsymbol \theta^{(2)}$ in the parameter space).
-Furthermore, the potential for model instability and the size of the parameter space can also become greater with the introduction of latent variables to existing data variables, as involved in some model formulations described in Sections \@ref(rbm)-\@ref(deep-learning).
+which indicates how parameter sensitivity in the likelihood $P_{\ak{\thetaN}} (  \boldsymbol x)$ may complicate sampling of the posterior $P_{ \ak{\thetaN}} (  \boldsymbol x) \pi(\ak{\thetaN})$ (i.e., moving from $\boldsymbol \theta^{(1)}$ to $\boldsymbol \theta^{(2)}$ in the parameter space).
+\ak{With this in mind, the instability results presented in Section \ref{instability-results} point to the potential use of regularization and penalization as a potential solution to avoid instability in FOES models.
+One example of a principled approach to penalization can be found in \cite{kaplan2019properties}, which approaches Bayesian fitting of a single layer RBM using priors to penalize the parameter values.} Furthermore, the potential for model instability and the size of the parameter space can also become greater with the introduction of latent variables to existing data variables, as involved in some model formulations described in Sections \@ref(rbm)-\@ref(deep-learning).
 As latent variables are often sampled with parameters in a Bayes MCMC approach, this aspect may further compound numerical problems in chain mixing.
@@
 We have shown for FOES models that instability manifests through small changes in data leading to potentially large changes in probability as well as the potential to place all probability on certain modal subsections of the sample space, which potentially could be small.
-Such instability is often due to a complex interaction between the model statistics used  (i.e., how numerous and large these may become) and the number and magnitudes of
-parameters in the model formulation.
-For many FOES models, the possibility exists, at least in principle, to constraint parameters in a way balances their potential contributions against those of model statistics in order to prevent probability instabilities.
+Such instability is often due to a complex interaction between the model statistics used  (i.e., how numerous and large these may become) and the number and magnitudes of parameters in the model formulation.
+For many FOES models, the possibility exists, at least in principle, to \ak{constrain} parameters in a way \ak{that} balances their potential contributions against those of model statistics in order to prevent probability instabilities.
 The FOES model class is quite broad and, in developing such models for large data sets, some caution should be used in parameter specification to control effects of model instability.

### 2019-06-11 writing/paper-note/docs/note.tex
started edits

@@
 
+\usepackage{color}
+\newcommand{\ak}[1]{{\color{blue} #1}}
+
 %imaiai starts here
@@
 
-\author{{\sc Andee Kaplan}\(^*\), \\[2pt] Department of Statistical Science, Duke University, P.O. Box 90251,
-Durham, NC 27708-0251, USA \\ \(^*\){\email{Corresponding author: \href{mailto:andrea.kaplan@duke.edu}{\nolinkurl{andrea.kaplan@duke.edu}}}}
-\\[6pt] {\sc Daniel Nordman} \\[2pt] Department of Statistics, Iowa State University, 2438 Osborn Dr., Ames,
-IA 50011-1090, USA \\ {\href{mailto:dnordman@iastate.edu}{\nolinkurl{dnordman@iastate.edu}}}  
-\\[6pt] {\sc Stephen Vardeman} \\[2pt] Departments of Statistics and Industrial and Manufacturing Systems
-Engineering, Iowa State University, 2438 Osborn Dr., Ames, IA
-50011-1090, USA \\ {\href{mailto:vardeman@iastate.edu}{\nolinkurl{vardeman@iastate.edu}}}  
+\author{{\sc Andee Kaplan}\(^*\), \\[2pt] Department of Statistical Science, Duke University, P.O.
+Box 90251, Durham, NC 27708-0251, USA \\ \(^*\){\email{Corresponding author: \href{mailto:andrea.kaplan@duke.edu}{\nolinkurl{andrea.kaplan@duke.edu}}}}
+\\[6pt] {\sc Daniel Nordman} \\[2pt] Department of Statistics, Iowa State University, 2438 Osborn Dr., Ames, IA 50011-1090, USA \\ {\href{mailto:dnordman@iastate.edu}{\nolinkurl{dnordman@iastate.edu}}}
+\\[6pt] {\sc Stephen Vardeman} \\[2pt] Departments of Statistics and Industrial and Manufacturing Systems Engineering, Iowa State University, 2438 Osborn Dr., Ames, IA 50011-1090, USA \\ {\href{mailto:vardeman@iastate.edu}{\nolinkurl{vardeman@iastate.edu}}}
 }
@@
 \begin{abstract}
-{A probability model exhibits instability if small changes in a data
-outcome result in large, and often unanticipated, changes in
-probability. This instability is a property of the probability model,
-rather than the fitted parameter vector. For correlated data structures
-found in several application areas, there is increasing interest in
-predicting/identifying such sensitivity in model probability structure.
-We consider the problem of quantifying instability for general
-probability models defined on sequences of observations, where each
-sequence of length \(N\) has a finite number of possible values. A
-sequence of probability models results, indexed by \(N\), that
-accommodates data of expanding dimension. Model instability is formally
-shown to occur when a certain log-probability ratio under such models
-grows faster than \(N\). In this case, a one component change in the
-data sequence can shift probability by orders of magnitude. Also, as
-instability becomes more extreme, the resulting probability models are
-shown to tend to degeneracy, placing all their probability on
-potentially small portions of the sample space. These results on
-instability apply to large classes of models commonly used in random
-graphs, network analysis, and machine learning contexts.}
+{A probability model exhibits instability if small changes in a data outcome result in large, and often unanticipated, changes in probability.
+This instability is a property of the probability \ak{model, prescribed by a probability distributional form evaluated at a given configuration of parameters}.
+For correlated data structures found in several application areas, there is increasing interest in \ak{identifying} such sensitivity in model probability structure.
+We consider the problem of quantifying instability for general probability models defined on sequences of observations, where each sequence of length \(N\) has a finite number of possible values \ak{that can be taken at each point}.
+A sequence of probability models results, indexed by \(N\), \ak{and an associated parameter sequence,} that accommodates data of expanding dimension.
+Model instability is formally shown to occur when a certain log-probability ratio under such models grows faster than \(N\).
+In this case, a one component change in the data sequence can shift probability by orders of magnitude.
+Also, as instability becomes more extreme, the resulting probability models are shown to tend to degeneracy, placing all their probability on potentially small portions of the sample space.
+These results on instability apply to large classes of models commonly used in random graphs, network analysis, and machine learning contexts.}
 {Degeneracy, Instability, Classification, Deep Learning, Graphical Models}
@@
 
-We consider the behavior, and the potential impropriety, of sequences of
-discrete probability models built to incorporate observations of
-increasing sample size \(N\). Interest is in identifying instability in
-such models, which is roughly characterized by probabilities with
-extreme sensitivity to small changes in data configuration. The concept
-of instability was introduced in the field of statistical physics (i.e.,
-point processes) by \citet{ruelle1999statistical} and then further
-extended by \citet{schweinberger2011instability} for a family of
-exponential models. At issue, models exhibiting instability are
-typically undesirable as these tend to provide poor representations of
-data or data-generation. As an example, such models can include
-near-degenerate distributions that assign essentially all probability
-mass to only a subset of an overall sample space. The latter issue in
-connection to degeneracy has been recognized as a concern in that
-dominant model outcomes may not resemble observed data
-\citep[cf.][]{handcock2003assessing}. As a compounding issue, model
-instability often has direct negative impacts for statistical inference
-and computations based on likelihood functions. Namely, volatilities in
-probability structure can potentially hamper the numerical evaluations
-required for maximum likelihood estimation as well as other model-based
-simulations via Markov Chain Monte Carlo (MCMC). These reasons motivate
-our general study of instability for a broad class of probability
-models, described next.
-
-In the model framework, let \(\boldsymbol X_N = (X_1, \dots, X_N)\)
-denote a collection of discrete random variables with a finite sample
-space, \(\mathcal{X}^N\), represented as some \(N\)-fold Cartesian
-product. That is, \(\mathcal{X}\) with \(|\mathcal{X}| < \infty\)
-denotes the set of potential outcomes for each single variable \(X_i\),
-so that the product space \(\mathcal{X}^N\) corresponds to values for
-the variables \(\boldsymbol X_N=(X_1,\ldots,X_N)\). For each \(N\), let
-\(P_{\boldsymbol \theta_N}\) denote a probability model on
-\(\mathcal{X}^N\), under which
-\(P_{\boldsymbol \theta_N}(x_1, \dots, x_N) > 0\) is the probability of
-the data outcome \((x_1, \dots, x_N) \in \mathcal{X}^N\). In this, we
-assume that the model support of \(P_{\boldsymbol \theta_N}\) is the
-sample space \(\mathcal{X}^N\). This framework produces probability
-models \(P_{\boldsymbol \theta_N}\), indexed by a generic sequence of
-parameters \(\boldsymbol \theta_N\), to describe data
-\(\boldsymbol X_N\) of any given sample size \(N \geq 1\). For
-simplicity, we will refer to this distributional class as \emph{Finite
-Outcome Everywhere Supported (FOES)} models in the following. The
-dimension and structure of such parameters are generic, without
-restriction, though natural cases will be seen to include those where
-\(\boldsymbol \theta_N \in \mathbb{R}^{q(N)}\) for some arbitrary
-integer-valued function \(q(\cdot) \geq 1\).
-
-Section \ref{examples} provides some examples of FOES models encountered
-in graph/network analysis and machine learning (i.e., deep learning
-models). These are used as references for later illustrations. Section
-\ref{instability-results} then establishes several formal results for
-FOES models with regard to instability.
-\citet{schweinberger2011instability} originally developed instability
-results specific to a certain class of discrete exponential models. For
-similar exponential models with random networks,
-\citet{handcock2003assessing} studied model degeneracy, where a
-probability model places near complete mass on modes and may thereby
-narrow the feasible model outcomes. As findings here and from
-\citet{schweinberger2011instability} suggest, model instability and
-degeneracy may also be related by viewing degeneracy as an extreme, or
-limiting form, of instability. Our main results establish a broad
-characterization of model instability, appropriate across the whole FOES
-model class, that incorporates results of
-\citet{schweinberger2011instability} as a special case. We prescribe a
-general and simple condition for identifying instability in a FOES model
-sequence, which quantifies whether certain maximal probabilities in a
-FOES model are too extreme relative to the sample size \(N\). When these
-conditions are met, the probability structure of a FOES model is shown
-to exhibit extreme sensitivity, with probability assignments possessing
-extreme peaks and troughs across nearly identical outcomes. As the
-measure of model instability increases, probabilities from an unstable
-FOES model additionally increase in volatility and provably slide into
-degeneracy. Section \ref{implications} then emphasizes the implications
-of such model instability, showing that such impropriety can be expected
-to numerically hinder maximum likelihood estimation and MCMC-based
-simulations. As one potential remedy, suggestions are given for
-constraining model parameterizations to avoid the most problematic
-regions of the parameter space. Proofs of the main results appear in
-Appendix \ref{appendix-instab}.
+We consider the behavior, and the potential impropriety, of sequences of discrete probability models built to incorporate observations of increasing sample size \(N\).
+Interest is in identifying instability in such models, which is roughly characterized by probabilities with extreme sensitivity to small changes in data configuration.
+The concept of instability was introduced in the field of statistical physics (i.e., point processes) by \citet{ruelle1999statistical} and then further extended by \citet{schweinberger2011instability} for a family of exponential models.
+At issue, models exhibiting instability are typically undesirable as these tend to provide poor representations of data or data-generation.
+As an example, such models can include near-degenerate distributions that assign essentially all probability mass to only a subset of an overall sample space.
+The latter issue in connection to degeneracy has been recognized as a concern in that dominant model outcomes may not resemble observed data \citep[cf.~][]{handcock2003assessing}.
+As a compounding issue, model instability often has direct negative impacts for statistical inference and computations based on likelihood functions.
+Namely, volatilities in probability structure can potentially hamper the numerical evaluations required for maximum likelihood estimation as well as other model-based simulations via Markov Chain Monte Carlo (MCMC).
+These reasons motivate our general study of instability for a broad class of probability models, described next.
+
+In the model framework, let \(\boldsymbol X_N = (X_1, \dots, X_N)\) denote a collection of discrete random variables with a finite sample space, \(\mathcal{X}^N\), represented as some \(N\)-fold Cartesian product.
+That is, \(\mathcal{X}\) with \(|\mathcal{X}| < \infty\) denotes the set of potential outcomes for each single variable \(X_i\), so that the product space \(\mathcal{X}^N\) corresponds to values for the variables \(\boldsymbol X_N=(X_1,\ldots,X_N)\).
+For each \(N\), let \(P_{\boldsymbol \theta_N}\) denote a probability model on \(\mathcal{X}^N\), under which \(P_{\boldsymbol \theta_N}(x_1, \dots, x_N) > 0\) is the probability of the data outcome \((x_1, \dots, x_N) \in \mathcal{X}^N\).
+In this, we assume that the model support of \(P_{\boldsymbol \theta_N}\) is the sample space \(\mathcal{X}^N\).
+This framework produces probability models \(P_{\boldsymbol \theta_N}\), indexed by a \ak{defining} sequence of parameters \(\boldsymbol \theta_N\), to describe data \(\boldsymbol X_N\) of any given sample size \(N \geq 1\).
+For simplicity, we will refer to this distributional class as \emph{Finite Outcome Everywhere Supported (FOES)} models in the following.
+The dimension and structure of such parameters are generic, without restriction, though natural cases will be seen to include those where \(\boldsymbol \theta_N \in \mathbb{R}^{q(N)}\) for some arbitrary integer-valued function \(q(\cdot) \geq 1\).
+
+Section \ref{examples} provides some examples of FOES models encountered in graph/network analysis and machine learning (i.e., deep learning models).
+These are used as references for later illustrations.
+Section \ref{instability-results} then establishes several formal results for FOES models with regard to instability.
+\citet{schweinberger2011instability} originally developed instability results specific to a certain class of discrete exponential models.
+For similar exponential models with random networks, \citet{handcock2003assessing} studied model degeneracy, where a probability model places near complete mass on modes and may thereby narrow the feasible model outcomes.
+As findings here and from \citet{schweinberger2011instability} suggest, model instability and degeneracy may also be related by viewing degeneracy as an extreme, or limiting form, of instability.
+Our main results establish a broad characterization of model instability, appropriate across the whole FOES model class, that incorporates results of \citet{schweinberger2011instability} as a special case.
+We prescribe a general and simple condition for identifying instability in a FOES model sequence, which quantifies whether certain maximal probabilities in a FOES model are too extreme relative to the sample size \(N\).
+When these conditions are met, the probability structure of a FOES model is shown to exhibit extreme sensitivity, with probability assignments possessing extreme peaks and troughs across nearly identical outcomes.
+As the measure of model instability increases, probabilities from an unstable FOES model additionally increase in volatility and provably slide into degeneracy.
+Section \ref{implications} then emphasizes the implications of such model instability, showing that such impropriety can be expected to numerically hinder maximum likelihood estimation and MCMC-based simulations.
+As one potential remedy, suggestions are given for constraining model parameterizations to avoid the most problematic regions of the parameter space.
+Proofs of the main results appear in Appendix \ref{appendix-instab}.
 
@@
 
-Many model families fall under the umbrella of FOES models. For
-illustration, this section presents three specific examples of FOES
-models, including models with deep architectures.
+Many model families \ak{lead to} FOES models.
+\ak{(For clarity, we use ``model family" to refer to generic class or functional form of parametric distributions, while "model" often refers here to a member of the class for an instantiation or selection of parameters.
+This distinction is not so important for the examples presented in Sections \ref{discrete-exponential-family-models}-\ref{deep-learning}, but becomes more so in Section \ref{instability-results} where model instability depends on both form and a given parameter configuration, rather than just model class form.)} For illustration, this section presents three specific examples of FOES model \ak{classes}, including models with deep architectures.
 
 \hypertarget{discrete-exponential-family-models}{%
-\subsection{Discrete Exponential Family
-Models}\label{discrete-exponential-family-models}}
-
-For random variables
-\(\boldsymbol X \equiv\boldsymbol X_N= (X_1, \dots, X_N)\) with sample
-space \(\mathcal{X}^N\), \(|\mathcal{X}| < \infty\), consider an
-exponential family model for \(\boldsymbol X\) with probability mass
-function given by \begin{equation}
+\subsection{Discrete Exponential Family Models}\label{discrete-exponential-family-models}}
+
+For random variables \(\boldsymbol X \equiv\boldsymbol X_N= (X_1, \dots, X_N)\) with sample space \(\mathcal{X}^N\), \(|\mathcal{X}| < \infty\), consider an exponential \ak{model family} for \(\boldsymbol X\) with probability mass function given by
+\begin{equation}
 \label{eq:expo}
 p_{N, \boldsymbol \theta}(\boldsymbol x) = \exp\left[\boldsymbol\eta^T(\boldsymbol \theta) \boldsymbol g_N(\boldsymbol x) - \psi(\boldsymbol \theta)\right], \qquad \boldsymbol x \in \mathcal{X}^N,
-\end{equation} depending on parameter vector
-\(\boldsymbol \theta \in \Theta_N \subset \mathbb{R}^{k}\) and natural
-parameter function
-\(\boldsymbol \eta : \mathbb{R}^k \mapsto \mathbb{R}^L\) with fixed
-positive integers \(k\) and \(L\) denoting their dimensions. Above,
-\(\boldsymbol g_N : \mathcal{X}^N \mapsto \mathbb{R}^L\) is a vector of
-sufficient statistics, while \[
+\end{equation}
+depending on parameter vector \(\boldsymbol \theta \in \Theta_N \subset \mathbb{R}^{k}\) and natural parameter function \(\boldsymbol \eta : \mathbb{R}^k \mapsto \mathbb{R}^L\) with fixed positive integers \(k\) and \(L\) denoting their dimensions.
+Above, \(\boldsymbol g_N : \mathcal{X}^N \mapsto \mathbb{R}^L\) is a vector of sufficient statistics, while
+\[
 \psi(\boldsymbol \theta) = \log \sum\limits_{\boldsymbol x \in \mathcal{X}^N}\exp\left[\boldsymbol \eta^T(\boldsymbol \theta) \boldsymbol g_N(\boldsymbol x) \right], \qquad \boldsymbol \theta \in \Theta_N\equiv \{\boldsymbol \theta \in \mathbb{R}^k : \psi(\boldsymbol \theta) < \infty \},
-\] denotes the normalizing function with parameter space \(\Theta_N\).
-The natural parameter function \(\eta (\boldsymbol \theta)\) has a
-linear form (i.e.,
-\(\eta (\boldsymbol \theta)= \bm{A} \boldsymbol \theta\) for a given
-\(L \times k\) matrix \(\bm{A}\)) in many common model formulations,
-though may also be nonlinear (e.g., curved exponential families). In the
-linear case, \(\eta (\boldsymbol \theta) = \boldsymbol \theta\) may be
-generally assumed in the exponential parameterization with a minor
-modification to the definition of sufficient statistics
-\(\boldsymbol g_N(\boldsymbol x)\).
-
-Such discrete exponential family models are special cases of the FOES
-models, as seen by defining
-\(P_{\boldsymbol \theta_N}(\boldsymbol x)\equiv p_{N,\boldsymbol \theta_N}(\boldsymbol x)> 0\),
-\(\boldsymbol x \in \mathcal{X}^N\), based on \eqref{eq:expo} and a
-parameter sequence
-\(\boldsymbol \theta_N \in \Theta_N \subset \mathbb{R}^k\). For example,
-if observations \(\boldsymbol X = (X_1,\ldots,X_N)\) correspond to \(N\)
-independent and identically distributed Bernoulli random variables, each
-indicating a binary \(0\)-\(1\) outcome, the resulting probabilities
-have exponential form \eqref{eq:expo} given by \begin{equation}
+\]
+denotes the normalizing function with parameter space \(\Theta_N\).
+The natural parameter function \(\eta (\boldsymbol \theta)\) has a linear form (i.e., \(\eta (\boldsymbol \theta)= \bm{A} \boldsymbol \theta\) for a given \(L \times k\) matrix \(\bm{A}\)) in many common model formulations, though may also be nonlinear (e.g., curved exponential families).
+In the linear case, \(\eta (\boldsymbol \theta) = \boldsymbol \theta\) may be generally assumed in the exponential parameterization with a minor modification to the definition of sufficient statistics \(\boldsymbol g_N(\boldsymbol x)\).
+
+Such discrete \ak{exponential models} are special cases of the FOES models, as seen by defining \(P_{\boldsymbol \theta_N}(\boldsymbol x)\equiv p_{N,\boldsymbol \theta_N}(\boldsymbol x)> 0\), \(\boldsymbol x \in \mathcal{X}^N\), based on \eqref{eq:expo} and a parameter sequence \(\boldsymbol \theta_N \in \Theta_N \subset \mathbb{R}^k\).
+For example, if observations \(\boldsymbol X = (X_1,\ldots,X_N)\) correspond to \(N\) independent and identically distributed Bernoulli random variables, each indicating a binary \(0\)-\(1\) outcome, the resulting probabilities have exponential form \eqref{eq:expo} given by
+\begin{equation}
 \label{eq:mod1}
@@
  \exp\left[\boldsymbol \theta_N \sum_{i=1}^N x_i\right], \qquad \boldsymbol x=(x_1,\ldots,x_N) \in\{0,1\}^N, 
- \end{equation} with sufficient statistic
-\(\boldsymbol g_N(\boldsymbol x)\equiv \sum_{i=1}^N x_i\) and ``log odds
-ratio'' parameter
-\(\boldsymbol \theta_N \equiv \log[ P_{\boldsymbol \theta_N}(X_i=1)/P_{\boldsymbol \theta_N}(X_i=0) ] \in \mathbb{R}\).
-More generally, supposing \(\boldsymbol X =(X_1,\ldots,X_N)\) represent
-\(N\) independent trials, each assuming an outcome \(\{1,\ldots,k\}\)
-among \(k\) possibilities (e.g., a die roll), a multinomial distribution
-is given by \begin{equation}
+ \end{equation}
+with sufficient statistic \(\boldsymbol g_N(\boldsymbol x)\equiv \sum_{i=1}^N x_i\) and ``log odds ratio'' parameter \(\boldsymbol \theta_N \equiv \log[ P_{\boldsymbol \theta_N}(X_i=1)/P_{\boldsymbol \theta_N}(X_i=0) ] \in \mathbb{R}\).
+More generally, supposing \(\boldsymbol X =(X_1,\ldots,X_N)\) represent \(N\) independent trials, each assuming an outcome \(\{1,\ldots,k\}\) among \(k\) possibilities (e.g., a die roll), a multinomial distribution is given by
+\begin{equation}
 \label{eq:mod11}
@@
 \exp\left[ \sum_{j=1}^k {\theta_{j,N}} \sum_{i=1}^N \mathbb{I}(x_i=j) \right], \qquad \boldsymbol x  \in\{1,\ldots,k\}^N, 
-\end{equation} with sufficient statistic
-\(\boldsymbol g_N(\boldsymbol x)\) involving a count
-\(\sum_{i=1}^N \mathbb{I}(x_i=j)\) for each outcome
-\(j \in \{1,\ldots,k\}\), where \(\mathbb{I}(\cdot)\) denotes the
-indicator function, and parameters
-\(\boldsymbol \theta_N=(\theta_{1,N},\ldots,\theta_{k,N})\in\mathbb{R}^k\)
-defining log-probability ratios
-\(\theta_{i,N}-\theta_{j,N} =\log [P_{\boldsymbol \theta_N}(X_1=i)/P_{\boldsymbol \theta_N}(X_1=j)]\).
-In addition to such standard models for discrete independent data,
-exponential models of FOES type commonly arise with dependent spatial
-data \citep{besag1974spatial} and network/relational data
-\citep{wasserman1994social, handcock2003assessing}. For a random graph
-or network with, say, \(n\) nodes, consider \(N={n \choose 2}\) random
-edges where the \(i\)th edge is associated with a pair of nodes
-\(s_i \equiv \{v_i,u_i\}\) and a binary variable \(X_i\in\{0,1\}\)
-indicating presence/absence of an edge among the node pair \(s_i\),
-\(i=1,\ldots,N\). Here the length \(N\) of the edge variable sequence
-\(\boldsymbol X = (X_1,\ldots,X_N)\) increases as a function of node
-number \(n\) and corresponding exponential models often incorporate
-graph topographical features derived from \(\boldsymbol X\). As an
-example, consider a graph model of exponential/FOES form prescribed by
+\end{equation}
+with sufficient statistic \(\boldsymbol g_N(\boldsymbol x)\) involving a count \(\sum_{i=1}^N \mathbb{I}(x_i=j)\) for each outcome \(j \in \{1,\ldots,k\}\), where \(\mathbb{I}(\cdot)\) denotes the indicator function, and parameters \(\boldsymbol \theta_N=(\theta_{1,N},\ldots,\theta_{k,N})\in\mathbb{R}^k\) defining log-probability ratios \(\theta_{i,N}-\theta_{j,N} =\log [P_{\boldsymbol \theta_N}(X_1=i)/P_{\boldsymbol \theta_N}(X_1=j)]\).
+In addition to such standard models for discrete independent data, exponential models of FOES type commonly arise with dependent spatial data \citep{besag1974spatial} and network/relational data \citep{wasserman1994social, handcock2003assessing}.
+For a random graph or network with, say, \(n\) nodes, consider \(N={n \choose 2}\) random edges where the \(i\)th edge is associated with a pair of nodes \(s_i \equiv \{v_i,u_i\}\) and a binary variable \(X_i\in\{0,1\}\) indicating presence/absence of an edge among the node pair \(s_i\), \(i=1,\ldots,N\).
+Here the length \(N\) of the edge variable sequence \(\boldsymbol X = (X_1,\ldots,X_N)\) increases as a function of node number \(n\) and corresponding exponential models often incorporate graph topographical features derived from \(\boldsymbol X\).
+As an example, consider a graph model of exponential/FOES form prescribed by
 \begin{equation}
@@
  \exp\left[\sum_{j=1}^3 \theta_{j,N} g_{j,N}(\boldsymbol x)\right], \quad\qquad \boldsymbol x=(x_1,\ldots,x_N)  \in\{0,1\}^N,
-\end{equation} \[
+\end{equation}
+\[
   g_{1,N}(\boldsymbol x) \equiv \sum_{i=1}^N  x_i, \qquad\quad g_{2,N}(\boldsymbol x) \equiv \sum_{1 \leq i<j \leq N,\atop s_i \cap s_j \neq \emptyset}\!\!\!   x_i x_j, \qquad
@@
 x_i x_j x_\ell,
-\] involving the numbers of edges, 2-stars and triangles among an
-outcome \(\boldsymbol x\) given by \(g_{1,N}(\boldsymbol x)\),
-\(g_{2,N}(\boldsymbol x)\) and \(g_{3,N}(\boldsymbol x)\), respectively,
-along with \(k=3\) real parameters
-\(\boldsymbol \theta_N \equiv (\theta_{1,N},\theta_{2,N},\theta_{3,N})\).
-For this network model \eqref{eq:mod2} in particular, as well as for more
-general models of form \eqref{eq:expo},
-\citet{schweinberger2011instability} considered instability in such
-exponential models with sequences of fixed parameters
-\(\boldsymbol \theta_N = (\theta_1,\ldots,\theta_k)\in\mathbb{R}^k\),
-\(N \geq 1\), of fixed dimension \(k\).
-
-For model sequences
-\(P_{\boldsymbol \theta_N}(\boldsymbol x)\equiv p_{N,\boldsymbol \theta_N}(\boldsymbol x)\)
-of the exponential type \eqref{eq:expo}, such as those in
-\eqref{eq:mod1}-\eqref{eq:mod2}, note that the dimension \(k\) of the
-parameter \(\boldsymbol \theta_N\in\Theta \subset \mathbb{R}^k\)
-necessarily remains the same for all sample sizes \(N \geq 1\) as the
-form of the natural parameter function \(\eta(\cdot)\) in \eqref{eq:expo}
-and the number of sufficient statistics
-\(\boldsymbol g_{N}(\boldsymbol x)\) do not depend on \(N\).
-Consequently, \(\boldsymbol \theta_N\) lies in a parameter space of
-fixed Euclidean dimension \(k\). However, this aspect need not be true
-for other types of FOES models considered in Sections \ref{rbm} -
-\ref{deep-learning}, where instead the numbers of parameters and
-sufficient statistics commonly increase with the sample size \(N\).
+\]
+involving the numbers of edges, 2-stars and triangles among an outcome \(\boldsymbol x\) given by \(g_{1,N}(\boldsymbol x)\), \(g_{2,N}(\boldsymbol x)\) and \(g_{3,N}(\boldsymbol x)\), respectively, along with \(k=3\) real parameters \(\boldsymbol \theta_N \equiv (\theta_{1,N},\theta_{2,N},\theta_{3,N})\).
+For this network model \eqref{eq:mod2} in particular, as well as for more general models of form \eqref{eq:expo}, \citet{schweinberger2011instability} considered instability in such exponential models with sequences of fixed parameters \(\boldsymbol \theta_N = (\theta_1,\ldots,\theta_k)\in\mathbb{R}^k\), \(N \geq 1\), of fixed dimension \(k\).
+
+For model sequences \(P_{\boldsymbol \theta_N}(\boldsymbol x)\equiv p_{N,\boldsymbol \theta_N}(\boldsymbol x)\) of the exponential type \eqref{eq:expo}, such as those in \eqref{eq:mod1}-\eqref{eq:mod2}, note that the dimension \(k\) of the parameter \(\boldsymbol \theta_N\in\Theta \subset \mathbb{R}^k\) necessarily remains the same for all sample sizes \(N \geq 1\) as the form of the natural parameter function \(\eta(\cdot)\) in \eqref{eq:expo} and the number of sufficient statistics \(\boldsymbol g_{N}(\boldsymbol x)\) do not depend on \(N\).
+Consequently, \(\boldsymbol \theta_N\) lies in a parameter space of fixed Euclidean dimension \(k\).
+However, this aspect need not be true for other types of FOES models considered in Sections \ref{rbm} - \ref{deep-learning}, where instead the numbers of parameters and sufficient statistics commonly increase with the sample size \(N\).
 
@@
 
-A restricted Boltzmann machine (RBM) is an undirected graphical model
-specified for discrete or continuous random variables, with binary
-variables being most common \citep[cf.][]{smolensky1986information}. A
-RBM architecture has two layers, hidden (\(\mathcal{H}\)) and visible
-(\(\mathcal{V}\)), with conditional independence within each layer. Let
-\(\boldsymbol X = (X_1,\ldots,X_N)\) denote the \(N\) random variables
-for visibles with support \(\mathcal{X}^N\) and
-\(\boldsymbol H = (H_1,\ldots,H_{N_\mathcal{H}})\) denote the
-\(N_\mathcal{H}\) random variables for hiddens with support
-\(\mathcal{X}^{N_\mathcal{H}}\) where \(\mathcal{X} = \{-1,1\}\). For
-parameters
-\(\boldsymbol \theta_N^{\mathcal{H}} \in \mathbb{R}^{N_\mathcal{H}}\),
-\(\boldsymbol \theta_N^{\mathcal{V}}\in \mathbb{R}^N\), and
-\(\boldsymbol \theta_N^{\mathcal{HV}}\) as a real matrix with dimension
-\(N_\mathcal{H} \times N\), the RBM model for
-\(\tilde{\boldsymbol X}=(\boldsymbol X,\boldsymbol H)\) has the joint
-probability mass function \begin{equation}
+A restricted Boltzmann machine (RBM) is an undirected graphical model specified for discrete or continuous random variables, with binary variables being most common \citep[cf.~][]{smolensky1986information}.
+A RBM architecture has two layers, hidden (\(\mathcal{H}\)) and visible (\(\mathcal{V}\)), with conditional independence within each layer.
+Let \(\boldsymbol X = (X_1,\ldots,X_N)\) denote the \(N\) random variables for visibles with support \(\mathcal{X}^N\) and \(\boldsymbol H = (H_1,\ldots,H_{N_\mathcal{H}})\) denote the \(N_\mathcal{H}\) random variables for hiddens with support \(\mathcal{X}^{N_\mathcal{H}}\) where \(\mathcal{X} = \{-1,1\}\).
+For \ak{specified} parameters \(\boldsymbol \theta_N^{\mathcal{H}} \in \mathbb{R}^{N_\mathcal{H}}\), \(\boldsymbol \theta_N^{\mathcal{V}}\in \mathbb{R}^N\), and \(\boldsymbol \theta_N^{\mathcal{HV}}\) as a real matrix with dimension \(N_\mathcal{H} \times N\), the RBM model for \(\tilde{\boldsymbol X}=(\boldsymbol X,\boldsymbol H)\) has the joint probability mass function
+\begin{equation}
 \label{eq:RBM1}
 \tilde{P}_{\boldsymbol \theta_N} (\tilde{\boldsymbol x}) = \exp\left[ (\boldsymbol \theta_N^{\mathcal{H}})^T \boldsymbol h + \boldsymbol (\boldsymbol \theta_N^{\mathcal{V}})^T \boldsymbol x + \boldsymbol h^T  \boldsymbol\theta_N^{\mathcal{HV}} \boldsymbol x - \psi(\boldsymbol \theta_N)\right], \quad \tilde{\boldsymbol x} = (\boldsymbol x, \boldsymbol h) \in \{\pm 1\}^{N+N_\mathcal{H}}
-\end{equation} with normalizing function \[
+\end{equation}
+with normalizing function
+\[
 \psi(\boldsymbol \theta_N) = \log \sum_{\tilde{\boldsymbol x} \in \{\pm 1\}^{N+N_H} } \exp\left[ (\boldsymbol \theta_N^{\mathcal{H}})^T \boldsymbol h + \boldsymbol (\boldsymbol \theta_N^{\mathcal{V}})^T \boldsymbol x + \boldsymbol h^T  \boldsymbol\theta_N^{\mathcal{HV}} \boldsymbol x\right].
-\] Let
-\(\boldsymbol \theta_N = (\boldsymbol \theta_N^{\mathcal{H}}, \boldsymbol \theta_N^{\mathcal{V}}, \boldsymbol\theta_N^{\mathcal{HV}} ) \in \Theta_N \equiv \mathbb{R}^{q(N)}\),
-with \(q(N) = N + N_\mathcal{H} + N*N_\mathcal{H}\), denote the
-parameter vector for the RBM, as indexed by the number \(N\) of visible
-random variables (which may differ from the actual lengths of these
-parameter vectors). The probability mass function for the visible
-variables \(\boldsymbol X = (X_1, \dots, X_N)\) follows from
-marginalizing the joint specification to yield \begin{equation}
+\]
+Let \(\boldsymbol \theta_N = (\boldsymbol \theta_N^{\mathcal{H}}, \boldsymbol \theta_N^{\mathcal{V}}, \boldsymbol\theta_N^{\mathcal{HV}} ) \in \Theta_N \equiv \mathbb{R}^{q(N)}\), with \(q(N) = N + N_\mathcal{H} + N*N_\mathcal{H}\), denote \ak{a given} parameter vector for the RBM, as indexed by the number \(N\) of visible random variables (which may differ from the actual lengths of these parameter vectors).
+The probability mass function for the visible variables \(\boldsymbol X = (X_1, \dots, X_N)\) follows from marginalizing the joint specification to yield
+\begin{equation}
 \label{eq:RBM2}
 P_{\boldsymbol \theta_N} (\boldsymbol x) = \sum\limits_{\boldsymbol h \in \{\pm 1\}^{N_{\mathcal{H}}}} \tilde{P}_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h), \qquad \boldsymbol x \in \{\pm 1\}^{N}\equiv \mathcal{X}^N.
-\end{equation} Here the baseline model \eqref{eq:RBM1} for hidden/visible
-variables is a linear exponential one in sufficient statistics
-\((\tilde{\boldsymbol X}, \boldsymbol X^T\boldsymbol H)\) using
-\(\tilde{\boldsymbol X}=(\boldsymbol X,\boldsymbol H)\) from
-\eqref{eq:RBM1}, but the form differs from the previous exponential models
-in \eqref{eq:expo} in that the lengths of parameters
-\(\boldsymbol \theta_N\) and statistics
-\((\tilde{\boldsymbol X}, \boldsymbol X^T\boldsymbol H)\) increase to
-incorporate more visible variables. That is, in contrast to
-\eqref{eq:expo}, the natural parameter function involved in the RBM model
-\eqref{eq:RBM1}, as the identity mapping of the parameters
-\(\boldsymbol \theta_N\in\mathbb{R}^{q(N)}\), naturally grows in
-dimension \(q(N)\to \infty\) to accommodate visible variables
-\(X_1, \dots, X_N\) of increasing sample size \(N\to \infty\).
-Additionally, one may further arbitrarily choose the number
-\(N_\mathcal{H}\) of hidden variables \(\boldsymbol H\) in the joint RBM
-model \eqref{eq:RBM1} to define a marginal model \eqref{eq:RBM2} for the
-\(N\) visible variables \(\boldsymbol X\), and the number
-\(N_\mathcal{H}\) of hiddens may also potentially increase with \(N\).
-Because \(|\mathcal{X}| = 2\) and
-\(P_{\boldsymbol \theta_N}(\boldsymbol x) > 0\) for all
-\(\boldsymbol x \in \mathcal{X}^N\), the RBM specification \eqref{eq:RBM2}
-for visibles \(\boldsymbol\) corresponds to a FOES model, while the
-joint distribution \eqref{eq:RBM1} for \((\boldsymbol X, \boldsymbol H)\)
-is also a FOES model. As this example also indicates, any model formed
-by marginalizing a base FOES model class, such as the RBM joint
-specification \eqref{eq:RBM1}, is again a FOES model.
+\end{equation}
+Here the baseline model \eqref{eq:RBM1} for hidden/visible variables is a linear exponential one in sufficient statistics \((\tilde{\boldsymbol X}, \boldsymbol X^T\boldsymbol H)\) using \(\tilde{\boldsymbol X}=(\boldsymbol X,\boldsymbol H)\) from \eqref{eq:RBM1}, but the form differs from the previous exponential models in \eqref{eq:expo} in that the lengths of parameters \(\boldsymbol \theta_N\) and statistics \((\tilde{\boldsymbol X}, \boldsymbol X^T\boldsymbol H)\) increase to incorporate more visible variables.
+That is, in contrast to \eqref{eq:expo}, the natural parameter function involved in the RBM model \eqref{eq:RBM1}, as the identity mapping of the parameters \(\boldsymbol \theta_N\in\mathbb{R}^{q(N)}\), naturally grows in dimension \(q(N)\to \infty\) to accommodate visible variables \(X_1, \dots, X_N\) of increasing sample size \(N\to \infty\).
+Additionally, one may further arbitrarily choose the number \(N_\mathcal{H}\) of hidden variables \(\boldsymbol H\) in the joint RBM model \eqref{eq:RBM1} to define a marginal model \eqref{eq:RBM2} for the \(N\) visible variables \(\boldsymbol X\), and the number \(N_\mathcal{H}\) of hiddens may also potentially increase with \(N\).
+Because \(|\mathcal{X}| = 2\) and \(P_{\boldsymbol \theta_N}(\boldsymbol x) > 0\) for all \(\boldsymbol x \in \mathcal{X}^N\), the RBM specification \eqref{eq:RBM2} for visibles \(\boldsymbol\) corresponds to a FOES model, while the joint distribution \eqref{eq:RBM1} for \((\boldsymbol X, \boldsymbol H)\) is also a FOES model.
+As this example also indicates, any model formed by marginalizing a base FOES model class, such as the RBM joint specification \eqref{eq:RBM1}, is again a FOES model.
 
@@
 
-Consider two models with ``deep architecture'' that contain multiple
-hidden (or latent) layers in addition to a visible layer of data, namely
-a deep Boltzmann machine \citep{salakhutdinov2009deep} and a deep belief
-network \citet{hinton2006fast}{]}. Let \(M\) denote the number of hidden
-layers included in the model and let \(N_{(H,1)}, \dots, N_{(H,M)}\)
-denote the numbers of hidden variables within each hidden layer. Then
-the random vector
-\(\tilde{\boldsymbol X} = \{H^{(1)}_1, \dots, H^{(1)}_{N_{(H,1)}}, \dots, H^{(M)}_1, \dots, H^{(M)}_{N_{(H,M)}}, \boldsymbol X\}\)
-collects both the hidden variables
-\(\{ H_{i}^{(j)} : i=1,\ldots, N_{(H,j)}, j=1,\ldots,M\}\) and visible
-variables \(\boldsymbol X =(X_1,\ldots,X_N)\) in a deep probabilistic
-model. Each variable outcome will again lie in
-\(\mathcal{X} = \{-1,1\}\).
-
-\textbf{Deep Boltzmann machine (DBM).} The DBM class of models maintains
-conditional independence within all layers in the model by stacking RBM
-models and only allowing conditional dependence between neighboring
-layers. The joint probability mass function for a DBM is \[
+Consider two models with ``deep architecture'' that contain multiple hidden (or latent) layers in addition to a visible layer of data, namely a deep Boltzmann machine \citep{salakhutdinov2009deep} and a deep belief network \citep{hinton2006fast}.
+Let \(M\) denote the number of hidden layers included in the model and let \(N_{(H,1)}, \dots, N_{(H,M)}\) denote the numbers of hidden variables within each hidden layer.
+Then the random vector \(\tilde{\boldsymbol X} = \{H^{(1)}_1, \dots, H^{(1)}_{N_{(H,1)}}, \dots, H^{(M)}_1, \dots, H^{(M)}_{N_{(H,M)}}, \boldsymbol X\}\) collects both the hidden variables \(\{ H_{i}^{(j)} : i=1,\ldots, N_{(H,j)}, j=1,\ldots,M\}\) and visible variables \(\boldsymbol X =(X_1,\ldots,X_N)\) in a deep probabilistic model.
+Each variable outcome will again lie in \(\mathcal{X} = \{-1,1\}\).
+
+\textbf{Deep Boltzmann machine (DBM).} The DBM class of models maintains conditional independence within all layers in the model by stacking RBM models and only allowing conditional dependence between neighboring layers.
+The joint probability mass function for a DBM is
+\[
 \tilde{P}_{\boldsymbol \theta_N} ( \tilde{\boldsymbol x} ) = \exp\left[ \sum\limits_{i = 1}^M\boldsymbol \alpha^{(i)T} \boldsymbol h^{(i)} + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^{(1)T} \Gamma^{(0)} \boldsymbol x + \sum\limits_{i = 1}^{M - 1} \boldsymbol h^{(i)T} \Gamma^{(i)} \boldsymbol h^{(i + 1)} - \psi(\boldsymbol \theta_N) \right],
-\] for
-\(\tilde{\boldsymbol x} = (\boldsymbol h^{(1)}, \dots, \boldsymbol h^{(M)}, \boldsymbol x) \in \mathcal{X}^{N_{(H,1)} + \cdots + N_{(H,M)} +N}\)
-where \[
+\]
+for \(\tilde{\boldsymbol x} = (\boldsymbol h^{(1)}, \dots, \boldsymbol h^{(M)}, \boldsymbol x) \in \mathcal{X}^{N_{(H,1)} + \cdots + N_{(H,M)} +N}\) where
+\[
 \psi(\boldsymbol \theta_N) = \log \sum\limits_{\tilde{\boldsymbol x} \in \mathcal{X}^{N_{(H,1)} + \cdots + N_{(H,M)} +N}} \exp\left[ \sum\limits_{i = 1}^M\boldsymbol \alpha^{(i)T} \boldsymbol h^{(i)} + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^{(1)T} \Gamma^{(0)} \boldsymbol x + \sum\limits_{i = 1}^{M - 1} \boldsymbol h^{(i)T} \Gamma^{(i)} \boldsymbol h^{(i + 1)}\right],
-\] is the normalizing function for
-\(\boldsymbol \theta_N = (\boldsymbol \alpha^{(1)}, \dots, \boldsymbol \alpha^{(M)}, \boldsymbol \beta,\Gamma^{(0)}, \dots, \Gamma^{(M - 1)}) \in \Theta_N \subset \mathbb{R}^{q(N)}\),
-consisting of model parameters \(\boldsymbol \beta \in \mathbb{R}^N\),
-\(\boldsymbol \alpha^{(i)} \in \mathbb{R}^{N_{(H,i)}}\),
-\(i = 1, \dots, M\), along with a matrix \(\Gamma^{(0)}\) of dimension
-\(N_{(H,1)} \times N\), and matrices \(\Gamma^{(i)}\) of dimension
-\(N_{(H,i)} \times N_{(H,i+1)}\) for \(i = 1, \dots, M-1\). The combined
-parameter vector \(\boldsymbol \theta_N\) has total length
-\(q(N)= N_{(H,1)}+\cdots N_{(H,M)} + N + N_{(H,1)}*N+N_{H,2}*H_{(H,1)}+\cdots +N_{(H,M)}*H_{(H,M)-1}\).
-The probability mass function for the visible random variables
-\(X_1, \dots, X_N\) follows from this joint specification as \[
+\]
+is the normalizing function for \(\boldsymbol \theta_N = (\boldsymbol \alpha^{(1)}, \dots, \boldsymbol \alpha^{(M)}, \boldsymbol \beta,\Gamma^{(0)}, \dots, \Gamma^{(M - 1)}) \in \Theta_N \subset \mathbb{R}^{q(N)}\), consisting of model parameters \(\boldsymbol \beta \in \mathbb{R}^N\), \(\boldsymbol \alpha^{(i)} \in \mathbb{R}^{N_{(H,i)}}\), \(i = 1, \dots, M\), along with a matrix \(\Gamma^{(0)}\) of dimension \(N_{(H,1)} \times N\), and matrices \(\Gamma^{(i)}\) of dimension
+\(N_{(H,i)} \times N_{(H,i+1)}\) for \(i = 1, \dots, M-1\).
+The combined parameter vector \(\boldsymbol \theta_N\) has total length \(q(N)= N_{(H,1)}+\cdots N_{(H,M)} + N + N_{(H,1)}*N+N_{H,2}*H_{(H,1)}+\cdots +N_{(H,M)}*H_{(H,M)-1}\).
+The probability mass function for the visible random variables \(X_1, \dots, X_N\) follows from this joint specification as
+\[
 P_{\boldsymbol \theta_N} (\boldsymbol x) = \sum\limits_{(\boldsymbol h^{(1)}, \dots, \boldsymbol h^{(M)}) \in \mathcal{X}^{N_{(H,1)} + \cdots + N_{(H,M)}}} \tilde{P}_{\boldsymbol \theta_N} (\boldsymbol h^{(1)}, \dots, \boldsymbol h^{(M)}, \boldsymbol x) , \qquad \boldsymbol x \in \mathcal{X}^N.
-\] Again like the RBM case, the DBM model specification is an example of
-a FOES model.
-
-\textbf{Deep belief network (DBN).} A DBN resembles a DBM in that there
-are multiple layers of latent random variables stacked in a deep
-architecture with no conditional dependence between layers. The
-difference between the DBM and DBN models is that all but the last
-stacked layer in a DBN are Bayesian networks
-\citep[see][]{pearl985bayesian}, rather than RBMs. A Bayesian network is
-a probabilistic graphical model that defines conditional dependence to
-be directed, rather than undirected (as with the RBM). Thus for visibles
-\(X_1, \dots, X_N\) with support
-\(\mathcal{X}^N, \mid \mathcal{X} \mid < \infty\), a DBN is also a FOES
-model with \(q(N)\) the length of parameter vector is dependent on the
-dimension of the visibles because
-\(P_{\boldsymbol \theta_N}(\boldsymbol x)>0\) for all
-\(\boldsymbol x \in\mathcal{X}^N\). Commonly, as in logistic belief nets
-\citep{neal1992connectionist}, a ``weight'' parameter is placed on each
-interaction between visibles, \(X_1, \dots, X_N\), and the first layer
-of latent variables, \(H^{(1)}_1, \dots, H^{(1)}_{N_{(H,1)}}\),
-satisfying the definition of a FOES model.
+\]
+Again like the RBM case, the DBM model specification is an example of a FOES model.
+
+\textbf{Deep belief network (DBN).} A DBN resembles a DBM in that there are multiple layers of latent random variables stacked in a deep architecture with no conditional dependence between layers.
+The difference between the DBM and DBN models is that all but the last stacked layer \ak{(layer $0$, which contains the visible random variables $X_1, \dots, X_N$)} in a DBN are Bayesian networks \citep[see][]{pearl985bayesian}, rather than RBMs.
+A Bayesian network is a probabilistic graphical model that defines conditional dependence to be directed, rather than undirected (as with the RBM).
+Thus for visibles \(X_1, \dots, X_N\) with support \(\mathcal{X}^N, \mid \mathcal{X} \mid < \infty\), a DBN is also a FOES model with \(q(N)\) the length of parameter vector is dependent on the dimension of the visibles because \(P_{\boldsymbol \theta_N}(\boldsymbol x)>0\) for all \(\boldsymbol x \in\mathcal{X}^N\).
+Commonly, as in logistic belief nets \citep{neal1992connectionist}, a ``weight'' parameter is placed on each interaction between visibles, \(X_1, \dots, X_N\), and the first layer of latent variables, \(H^{(1)}_1, \dots, H^{(1)}_{N_{(H,1)}}\), satisfying the definition of a FOES model.
 
@@
 
-We now present a formal definition for instability of FOES models as
-well as a simple condition for identifying instability in a FOES model
-sequence.
+We now present a formal definition for instability of FOES models as well as a simple condition for identifying instability in a FOES model sequence.
 
@@
 
-To define a measure of instability in FOES models, it is useful to
-consider the behavior of data models \(P_{\theta_N}\), again supported
-on a set \(\mathcal{X}^N\) of outcomes for
-\(\boldsymbol X\equiv \boldsymbol X_N =(X_1,\ldots,X_N)\), in connection
-to the sample size \(N\). A relevant quantity to this end is a log-ratio
-of extremal probabilities (LREP), defined as \begin{align}
+To define a measure of instability in FOES models, it is useful to consider the behavior of data models \(P_{\theta_N}\), again supported on a set \(\mathcal{X}^N\) of outcomes for \(\boldsymbol X\equiv \boldsymbol X_N =(X_1,\ldots,X_N)\) \textbackslash ak\{and for a prescribed configuration of parameters \(\boldsymbol \theta_N\), in connection to the sample size \(N\).
+A relevant quantity to this end is a log-ratio of extremal probabilities (LREP), defined as
+\begin{align}
 \label{eq:elpr}
  \REP (\boldsymbol \theta_N)  =  \log \left[\frac{\max\limits_{  \boldsymbol x\in \mathcal{X}^N}P_{\boldsymbol \theta_N}( \boldsymbol x)}{\min\limits_{ \boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}( \boldsymbol x)}\right],
-\end{align} based on maximum and minimal model probabilities. In what
-follows, the main idea is that instability, and other negative model
-features, can be associated with a FOES model formulation for \(N\)
-random variables where the LREP \eqref{eq:elpr} is overly large relative
-to the sample size \(N\). That is, a sequence of FOES probability models
-\(P_{\boldsymbol \theta_N}\) results in specifying the distribution of
-observations \(\boldsymbol X=(X_1,\ldots,X_N)\) for each sample size
-\(N \geq 1\) and instability will generally occur among these models
-whenever the corresponding LREP \eqref{eq:elpr} grows faster than \(N\).
+\end{align}
+based on maximum and minimal model probabilities.
+In what follows, the main idea is that instability, and other negative model features, can be associated with a FOES model formulation for \(N\) random variables where the LREP \eqref{eq:elpr} is overly large relative to the sample size \(N\).
+That is, a sequence of FOES probability models \(P_{\boldsymbol \theta_N}\) results in specifying \ak{a} distribution of observations \(\boldsymbol X=(X_1,\ldots,X_N)\) for each sample size \(N \geq 1\) \ak{(dependent on the prescribed values for $\boldsymbol \theta_N$, $N \geq 1$)} and instability will generally occur among these models whenever the corresponding LREP \eqref{eq:elpr} grows faster than \(N\).
 This leads to the following definition.
@@
 \BeginKnitrBlock{definition}[S-unstable FOES model]
-\protect\hypertarget{def:instabFSFS}{}{\label{def:instabFSFS}
-\iffalse (S-unstable FOES model) \fi{} }A FOES model formulation for
-\(\boldsymbol X_N=(X_1,\ldots,X_N)\) is \emph{Schweinberger-unstable} or
-\emph{S-unstable} if \begin{equation}
+\protect\hypertarget{def:instabFSFS}{}{\label{def:instabFSFS} \iffalse (S-unstable FOES model) \fi{} }A FOES model formulation \ak{$P_{\boldsymbol \theta_N}$} for \(\boldsymbol X_N=(X_1,\ldots,X_N)\), \ak{based on parameter specification $\boldsymbol \theta_N$, $N \geq 1$,} is \emph{Schweinberger-unstable} or \emph{S-unstable} if
+\begin{equation}
 \label{eq:Sun}
 \lim \limits_{N \rightarrow \infty} \frac{1}{N} \REP(\boldsymbol \theta_N) \equiv \lim \limits_{N \rightarrow \infty} \frac{1}{N}\log \left[\frac{\max\limits_{  \boldsymbol x\in \mathcal{X}^N}P_{\boldsymbol \theta_N}( \boldsymbol x)}{\min\limits_{ \boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}( \boldsymbol x)}\right] = \infty
-\end{equation} as the number of variables increases
-(\(N \rightarrow \infty\)).
+\end{equation}
+as the number of variables increases (\(N \rightarrow \infty\)).
 \EndKnitrBlock{definition}
 
-In other words, a model is S-unstable if
-\(\REP(\boldsymbol \theta_N)/N\) is an unbounded sequence of sample size
-\(N\); namely, given any \(C > 0\), there exists an integer \(N_C > 0\)
-so that \(\REP(\boldsymbol \theta_N)/N > C\) holds for all
-\(N \ge N_C\). A FOES model formulation may be termed S-stable if it
-fails to be S-unstable, i.e., if
-\(\sup_{N \geq 1}\REP(\boldsymbol \theta_N)/N\) is bounded.
-
-This definition of S-unstable is a generalization or reinterpretation of
-``unstable'' used in \citet{schweinberger2011instability} by allowing
-possibly non-exponential family models (e.g., RBM and DBM models in
-Sections \ref{rbm}-\ref{deep-learning} as well as a potentially
-increasing number \(q(N)\) of parameters through the parameter sequence
-\(\boldsymbol \theta_N\in \mathbb{R}^{q(N)}\). While this definition
-differs in form and scope from the original, it does match that in
-\citet{schweinberger2011instability} for the special case of exponential
-models (cf.~Section \ref{discrete-exponential-family-models} considered
-there. Section \ref{illustrations} provides several examples of unstable
-models as well as causes for model instability, where the latter may
-often be traced to issues in model form (i.e., data functions) and/or
-parameterization. We next describe several potentially undesirable
-features associated with S-unstable FOES models.
+In other words, a model is S-unstable if \(\REP(\boldsymbol \theta_N)/N\) is an unbounded sequence of sample size \(N\); namely, given any \(C > 0\), there exists an integer \(N_C > 0\) so that \(\REP(\boldsymbol \theta_N)/N > C\) holds for all \(N \ge N_C\).
+A FOES model formulation may be termed S-stable if it fails to be S-unstable, i.e., if \(\sup_{N \geq 1}\REP(\boldsymbol \theta_N)/N\) is bounded.
+
+This definition of S-unstable is a generalization or reinterpretation of ``unstable'' used in \citet{schweinberger2011instability} by allowing possibly non-exponential family models (e.g., RBM and DBM models in Sections \ref{rbm}-\ref{deep-learning} as well as a potentially increasing number \(q(N)\) of parameters through the parameter sequence \(\boldsymbol \theta_N\in \mathbb{R}^{q(N)}\).
+While this definition differs in form and scope from the original, it does match that in \citet{schweinberger2011instability} for the special case of exponential models (cf.~Section \ref{discrete-exponential-family-models} considered there \ak{which translates to $\boldsymbol{\theta}_N=\boldsymbol{\theta} \in \mathbb{R}^k$, $N \geq 1$, in our model notation).
+Note the definition (\ref{eq:Sun}) of model instability depends intricately on the prescribed parameter sequence $\boldsymbol{\theta}_N$ or the particular parameter values indexing the distribution $P_{\boldsymbol{\theta}_N}$.
+This also agrees with the notion of unstable distributions from
+\citet{schweinberger2011instability}, as well as other characterizations of distributional stability in networks (cf.
+\citet{handcock2003assessing}), that model instability is crucially tied to the parameter configuration (i.e., where $\boldsymbol{\theta}_N$ lies in the parameter space) in addition to distributional form.
+Hence, a given model family or form may may result in an unstable/stable model, depending on the parameters $\boldsymbol{\theta}_n$ chosen.} Section \ref{illustrations} provides several examples of unstable models as well as causes for model instability, where the latter may often be traced to issues in model form (i.e., data functions) and/or \ak{type of} parameterization.
+We next describe several potentially undesirable features associated with S-unstable FOES models.
 
 \BeginKnitrBlock{remark}
-\iffalse{} {Remark. } \fi{}In the definition \eqref{eq:Sun} of
-S-instability, we note that the numerical measure
-\(\REP(\boldsymbol \theta_N)/N\) of model instability is invariant to
-\emph{independent replications} of data. That is, let \(M \geq 1\)
-denote a possible number of replications and consider data
-\(\boldsymbol Y_{N,M} \equiv (\boldsymbol X^{(1)}_N, \dots, \boldsymbol X^{(M)}_N)\)
-formed by \(\{ \boldsymbol X^{(j)}_N\}_{j=1}^M\) as \(M\) iid
-replications of a random vector \(\boldsymbol X_N=(X_1,\ldots,X_N)\),
-where the latter follows a FOES model with probabilities
-\(P_{ \boldsymbol \theta_N}(\boldsymbol x)>0\),
-\(\boldsymbol x\in\mathcal{X}^N\). This leads to a joint model, say
-\(P_{ \boldsymbol \theta_N}(\boldsymbol y)\),
-\(\boldsymbol y\in\mathcal{X}^{NM}\), for \(\boldsymbol Y_{N,M}\)
-consisting of \(N*M\) random variables in total. Then, the LREP for
-\(\boldsymbol Y_{N,M}\), scaled by associated size, is given by
+\iffalse{} {Remark.
+} \fi{}In the definition \eqref{eq:Sun} of S-instability, we note that the numerical measure \(\REP(\boldsymbol \theta_N)/N\) of model instability is invariant to \emph{independent replications} of data.
+That is, let \(M \geq 1\) denote a possible number of replications and consider data \(\boldsymbol Y_{N,M} \equiv (\boldsymbol X^{(1)}_N, \dots, \boldsymbol X^{(M)}_N)\) formed by \(\{ \boldsymbol X^{(j)}_N\}_{j=1}^M\) as \(M\) iid replications of a random vector \(\boldsymbol X_N=(X_1,\ldots,X_N)\), where the latter follows a FOES model with probabilities \(P_{ \boldsymbol \theta_N}(\boldsymbol x)>0\), \(\boldsymbol x\in\mathcal{X}^N\).
+This leads to a joint model, say \(P_{ \boldsymbol \theta_N}(\boldsymbol y)\), \(\boldsymbol y\in\mathcal{X}^{NM}\), for \(\boldsymbol Y_{N,M}\) consisting of \(N*M\) random variables in total.
+Then, the LREP for \(\boldsymbol Y_{N,M}\), scaled by associated size, is given by
 \begin{align*}
@@
 &= \frac{1}{NM}\log\left[\frac{\max_{\boldsymbol x \in \mathcal{X}^{N}}P_{ \boldsymbol \theta_N}(\boldsymbol x)  }{\min_{\boldsymbol x \in \mathcal{X}^{N}}P_{ \boldsymbol \theta_N}(\boldsymbol x)} \right]^M \equiv \frac{1}{N}\REP_{\boldsymbol X_{N}}( \boldsymbol\theta_N ),
-\end{align*} where
-\(\REP_{\boldsymbol X_{N}}( \boldsymbol\theta_N ) \equiv \REP( \boldsymbol\theta_N )\)
-denotes the log-ratio of extremal probabilities for \(\boldsymbol X_N\)
-defined from \eqref{eq:elpr}. That is, due to iid properties, the
-sample-size corrected LREP for \(\boldsymbol Y_{N,M}\) equals the
-analog, \(\REP(\boldsymbol\theta_N )/N\), from the underlying common
-data model for \(\boldsymbol X_N\) alone, regardless of the level
-\(M \geq 1\) of independent replication. Consequently, the definition of
-an S-unstable model is unaffected by independent replication and all
-instability properties may be characterized by those of one observation
-from the common FOES model. For computational purposes, this aspect also
-implies that if the original data \(\boldsymbol X_N=(X_1,\ldots,X_N)\)
-in a FOES model consist of \(N\) iid random variables, then the
-size-scaled log-ratio \eqref{eq:elpr} may be calculated as \[
+\end{align*}
+where \(\REP_{\boldsymbol X_{N}}( \boldsymbol\theta_N ) \equiv \REP( \boldsymbol\theta_N )\) denotes the log-ratio of extremal probabilities for \(\boldsymbol X_N\) defined from \eqref{eq:elpr}.
+That is, due to iid properties, the sample-size corrected LREP for \(\boldsymbol Y_{N,M}\) equals the analog, \(\REP(\boldsymbol\theta_N )/N\), from the underlying common data model for \(\boldsymbol X_N\) alone, regardless of the level \(M \geq 1\) of independent replication.
+Consequently, the definition of an S-unstable model is unaffected by independent replication and all instability properties may be characterized by those of one observation from the common FOES model.
+For computational purposes, this aspect also implies that if the original data \(\boldsymbol X_N=(X_1,\ldots,X_N)\) in a FOES model consist of \(N\) iid random variables, then the size-scaled log-ratio \eqref{eq:elpr} may be calculated as
+\[
 \frac{1}{N}\REP( \boldsymbol\theta_N ) \equiv \frac{1}{N}\REP_{\boldsymbol X_{N}}( \boldsymbol\theta_N ) = \log\left[ \frac{\max_{ x \in \mathcal{X}}P_{\boldsymbol\theta_N}(X_1=x)} {\min_{ x \in \mathcal{X}}P_{\boldsymbol\theta_N}(X_1=x)} \right]
-\] based on the extremal probabilities of just one random variable
-\(X_1\).
+\]
+based on the extremal probabilities of just one random variable \(X_1\).
 \EndKnitrBlock{remark}
@@
 \hypertarget{characterizations-and-consequences-of-instability}{%
-\subsection{Characterizations and Consequences of
-Instability}\label{characterizations-and-consequences-of-instability}}
-
-As a basic characteristic, S-unstable FOES model sequences have
-extremely sensitive probability structures. One aspect is that small
-changes in data configuration can lead to very large changes in
-probability. Consider, for example, the quantity given by \[
+\subsection{Characterizations and Consequences of Instability}\label{characterizations-and-consequences-of-instability}}
+
+As a basic characteristic, S-unstable FOES model sequences have extremely sensitive probability structures.
+One aspect is that small changes in data configuration can lead to very large changes in probability.
+Consider, for example, the quantity given by
+\[
 \DN(\boldsymbol \theta_N) \equiv \max \left\{\log \frac{P_{\boldsymbol \theta_N}(\boldsymbol x)}{P_{\boldsymbol \theta_N}(\boldsymbol x^*)} : \boldsymbol x \text{ }\& \text{ } \boldsymbol x^* \in \mathcal{X}^N \text{ differ in exactly one component}\right\},
-\] which represents the biggest log-probability ratio for a
-one-component change in data outcomes in a FOES model with parameter
-\(\boldsymbol \theta_N\). We then have the following result prescribing
-the behavior of \(\DN(\boldsymbol \theta_N)\) for S-unstable FOES
-models.
+\]
+which represents the biggest log-probability ratio for a one-component change \ak{(i.e., the data vector differs in exactly one component value)} in data outcomes in a FOES model with parameter \(\boldsymbol \theta_N\).
+\ak{This is the smallest difference possible between two data vectors, making it a good notion of a small distance in the input space.} We then have the following result prescribing the behavior of \(\DN(\boldsymbol \theta_N)\) for S-unstable FOES models.
 
 \BeginKnitrBlock{theorem}
-\protect\hypertarget{thm:instab-elpr}{}{\label{thm:instab-elpr} }Let
-\(P_{\boldsymbol \theta_N}\), with support \(\mathcal{X}^N\),
-\(N\geq 1\), be a sequence of FOES models.
+\protect\hypertarget{thm:instab-elpr}{}{\label{thm:instab-elpr} }Let \(P_{\boldsymbol \theta_N}\), with support \(\mathcal{X}^N\), \(N\geq 1\), be a sequence of FOES models.
 \begin{enumerate}[(i)]
@@
 \item Suppose the FOES model sequence is S-unstable.
-Then, for all large $N$ and given any arbitrary $C>0$, there exist outcomes $\boldsymbol x,\boldsymbol x^*\in\mathcal{X}^N$, differing by one component, such that
+Then, \ak{for any $C>0$ there exists $N'$ such that for all $N>N'$}, there exist outcomes $\boldsymbol x,\boldsymbol x^*\in\mathcal{X}^N$, differing by one component, such that
     $$
@@
 
-Theorem \ref{thm:instab-elpr}(i) is a non-asymptotic result, which
-connects to the definition of instability in a FOES model through a
-log-ratio of extreme model probabilities \eqref{eq:elpr} being too large
-relative to the associated sample size \(N\). If so, Theorem
-\ref{thm:instab-elpr}(i) guarantees the FOES model must also exhibit
-correspondingly large changes in probability for very small differences
-among some data configurations, a property that intuitively captures a
-notion of instability. Furthermore, and perhaps more seriously under
-Theorem \ref{thm:instab-elpr}(ii), S-unstable models can never have
-universally bounded changes in probability among single component
-variations in data configurations. While not all one-component changes
-in data may produce massive changes in probability, unstable models must
-have some such data outcomes with this property. As a consequence,
-unstable probability structures may exhibit extreme sensitivity through
-large peaks and troughs over the sample space.
-
-Additionally, S-unstable FOES model sequences are also connected to
-degenerate models, where \emph{degeneracy} involves assigning
-essentially all probability to modes within the sample space, which
-could potentially represent a small subset among the totality of
-outcomes. For perspective, note that differing sizes of the scaled
-log-ratio \(\REP(\boldsymbol \theta_N)/N\) from \eqref{eq:elpr} induce a
-spectrum of levels of instability/stability and Theorem
-\ref{thm:instab-elpr} indicates increasing sensitivity of model
-probabilities as \eqref{eq:elpr} increases. Furthermore, as the
-instability measure grows and the log-ratio
-\(\REP(\boldsymbol \theta_N)/N\) diverges, as in the definition
-\eqref{eq:Sun} of S-unstable models, then a FOES model sequence will
-become degenerate. Theorem \ref{thm:degenFOES} provides a formal
-statement of such degeneracy due to S-instability. For a given
-\(0 < \epsilon < 1\), define a \(\epsilon\)-modal set of outcomes as
+Theorem \ref{thm:instab-elpr}(i) is a non-asymptotic result, which connects to the definition of instability in a FOES model through a log-ratio of extreme model probabilities \eqref{eq:elpr} being too large relative to the associated sample size \(N\).
+If so, Theorem \ref{thm:instab-elpr}(i) guarantees the FOES model must also exhibit correspondingly large changes in probability for very small differences among some data configurations, a property that intuitively captures a notion of instability.
+Furthermore, and perhaps more seriously under Theorem \ref{thm:instab-elpr}(ii),
+S-unstable models can never have universally bounded changes in probability among single component variations in data configurations.
+While not all one-component changes in data may produce massive changes in probability, unstable models must have some such data outcomes with this property.
+As a consequence, unstable probability structures may exhibit extreme sensitivity through large peaks and troughs over the sample space.
+
+Additionally, S-unstable FOES model sequences are also connected to degenerate models, where \emph{degeneracy} involves assigning essentially all probability to modes within the sample space, which could potentially represent a small subset among the totality of outcomes.
+For perspective, note that differing sizes of the scaled log-ratio \(\REP(\boldsymbol \theta_N)/N\) from \eqref{eq:elpr} induce a spectrum of levels of instability/stability and Theorem \ref{thm:instab-elpr} indicates increasing sensitivity of model probabilities as \eqref{eq:elpr} increases.
+Furthermore, as the instability measure grows and the log-ratio \(\REP(\boldsymbol \theta_N)/N\) diverges, as in the definition \eqref{eq:Sun} of S-unstable models, then a FOES model sequence will become degenerate.
+Theorem \ref{thm:degenFOES} provides a formal statement of such degeneracy due to S-instability.
+For a given \(0 < \epsilon < 1\), define a \(\epsilon\)-modal set of outcomes as
 \begin{equation}
@@
 \BeginKnitrBlock{theorem}
-\protect\hypertarget{thm:degenFOES}{}{\label{thm:degenFOES} }For any
-arbitrarily small \(0 < \epsilon < 1\), an S-unstable FOES model
-sequence \(P_{\boldsymbol \theta_N}\), \(N \geq 1\), for
-\(\boldsymbol X_N=(X_1, \dots, X_N)\) satisfies \[
+\protect\hypertarget{thm:degenFOES}{}{\label{thm:degenFOES} }For any arbitrarily small \(0 < \epsilon < 1\), an S-unstable FOES model sequence \(P_{\boldsymbol \theta_N}\), \(N \geq 1\), for \(\boldsymbol X_N=(X_1, \dots, X_N)\) satisfies
+\[
 P_{\boldsymbol \theta_N}\left( \boldsymbol X_N\in \mathcal{M}_{\epsilon, \boldsymbol \theta_N}\right) \rightarrow 1 \text{ as } N \rightarrow \infty.
@@
 
-In other words, as the sample size grows in S-unstable FOES models, all
-probability tends to concentrate mass on an \(\epsilon\)-modal set,
-where \(\epsilon\) can be made arbitrarily small. Intuitively, the
-occurrence of such degeneracy can be explained by a type of ``reverse''
-pigeonhole principle for unstable FOES models: if all outcomes should
-receive positive probability but the maximal probability far exceeds the
-minimal one in the model, then little probability remains for
-distribution among remaining model outcomes (i.e., if nearly all
-available pigeons are stuffed into one hole, the remaining pigeonholes
-must have few occupants). Degeneracy in unstable models can pose dangers
-in data modeling as well, particularly when a mode set represents a
-narrow collection of outcomes among those realistically possible for
-adequately describing data. In which case, model outcomes may fail to
-look like data of interest.
-
-Connected to degeneracy, S-unstable FOES models may also exhibit
-additional kinds of extreme and undesirable sensitivity in probabilities
-if model parameters \(\boldsymbol \theta_N\) can further be ``dialed''
-between positive and negative values. That is, some FOES models
-naturally involve parameter spaces covering a positive-negative spectrum
-of parameter possibilities, where the signs of parameters provide a
-standard device for increasing or decreasing probabilities of outcomes
-in the model formulation. In fact, for many models, the switch of a
-parameter sign serves to produce reciprocal probabilities, as outlined
-in the following model assumption about parameter sign reversal (PSR).
-
-\textbf{Model Condition PSR} \emph{(Reciprocal Probabilities from
-Parameter Sign Reversal)}: Let \(P_{\boldsymbol \theta_N}\), with
-support \(\mathcal{X}^N\), \(N\geq 1\), represent a sequence of FOES
-models. For each \(N \geq 1\) and any outcome
-\(\boldsymbol x \in \mathcal{X}^N\), suppose it holds that \[
+In other words, as the sample size grows in S-unstable FOES models, all probability tends to concentrate mass on an \(\epsilon\)-modal set, where \(\epsilon\) can be made arbitrarily small.
+Intuitively, the occurrence of such degeneracy can be explained by a type of ``reverse'' pigeonhole principle for unstable FOES models: if all outcomes should receive positive probability but the maximal probability far exceeds the minimal one in the model, then little probability remains for distribution among remaining model outcomes (i.e., if nearly all available pigeons are stuffed into one hole, the remaining pigeonholes must have few occupants).
+Degeneracy in unstable models can pose dangers in data modeling as well, particularly when a mode set represents a narrow collection of outcomes among those realistically possible for adequately describing data.
+In which case, model outcomes may fail to look like data of interest.
+
+Connected to degeneracy, S-unstable FOES models may also exhibit additional kinds of extreme and undesirable sensitivity in probabilities if model parameters \(\boldsymbol \theta_N\) can further be ``dialed'' between positive and negative values.
+That is, some FOES models naturally involve parameter spaces covering a positive-negative spectrum of parameter possibilities, where the signs of parameters provide a standard device for increasing or decreasing probabilities of outcomes in the model formulation.
+In fact, for many models, the switch of a parameter sign serves to produce reciprocal probabilities, as outlined in the following model assumption about parameter sign reversal (PSR).
+
+\textbf{Model Condition PSR} \emph{(Reciprocal Probabilities from Parameter Sign Reversal)}: Let \(P_{\boldsymbol \theta_N}\), with support \(\mathcal{X}^N\), \(N\geq 1\), represent a sequence of FOES models.
+For each \(N \geq 1\) and any outcome \(\boldsymbol x \in \mathcal{X}^N\), suppose it holds that
+\[
 P_{\boldsymbol \theta_N}(\boldsymbol x)  \cdot P_{-\boldsymbol \theta_N}(\boldsymbol x) \;=\;   \max\limits_{\boldsymbol y \in \mathcal{X}^N}P_{ \boldsymbol \theta_N}(\boldsymbol y)\cdot \min\limits_{\boldsymbol y \in \mathcal{X}^N}P_{-\boldsymbol \theta_N}(\boldsymbol y),
-\] where
-\(\max_{\boldsymbol y \in \mathcal{X}^N}P_{ \boldsymbol \theta_N}(\boldsymbol y)\)
-and
-\(\min_{\boldsymbol y \in \mathcal{X}^N}P_{-\boldsymbol \theta_N}(\boldsymbol y)\)
-denote the maximum and minimum probabilities under parameters
-\(\boldsymbol \theta_N\) and \(-\boldsymbol \theta_N\), respectively.
-
-The above model condition incorporates many standard parameterizations
-and follows, for instance, whenever
-\(P_{\boldsymbol \theta_N}(\boldsymbol x)/P_{\boldsymbol \theta_N}(\boldsymbol y) = [P_{-\boldsymbol \theta_N}(\boldsymbol x)/P_{-\boldsymbol \theta_N}(\boldsymbol y)]^{-1}\)
-holds for outcomes \(\boldsymbol x, \boldsymbol y \in\mathcal{X}^N\) in
-a FOES model. For instance, this latter condition is fulfilled for all
-linear exponential families from Section
-\ref{discrete-exponential-family-models} (e.g.,
-\eqref{eq:mod1}-\eqref{eq:mod2}) as well as all network models from Sections
-\ref{rbm}-\ref{deep-learning} (e.g., \eqref{eq:RBM1}-\eqref{eq:RBM2}). When
-parameters can be tuned in sign with effects prescribed in the model
-condition PSR, unstable FOES models will exhibit further probability
-sensitivities, as outlined in the following extension of Theorem
-\ref{thm:degenFOES}.
+\]
+where \(\max_{\boldsymbol y \in \mathcal{X}^N}P_{ \boldsymbol \theta_N}(\boldsymbol y)\) and \(\min_{\boldsymbol y \in \mathcal{X}^N}P_{-\boldsymbol \theta_N}(\boldsymbol y)\) denote the maximum and minimum probabilities under parameters \(\boldsymbol \theta_N\) and \(-\boldsymbol \theta_N\), respectively.
+
+The above model condition incorporates many standard parameterizations and follows, for instance, whenever \(P_{\boldsymbol \theta_N}(\boldsymbol x)/P_{\boldsymbol \theta_N}(\boldsymbol y) = [P_{-\boldsymbol \theta_N}(\boldsymbol x)/P_{-\boldsymbol \theta_N}(\boldsymbol y)]^{-1}\) holds for outcomes \(\boldsymbol x, \boldsymbol y \in\mathcal{X}^N\) in a FOES model.
+For instance, this latter condition is fulfilled for all linear exponential families from Section \ref{discrete-exponential-family-models} (e.g., \eqref{eq:mod1}-\eqref{eq:mod2}) as well as all network models from Sections \ref{rbm}-\ref{deep-learning} (e.g., \eqref{eq:RBM1}-\eqref{eq:RBM2}).
+When parameters can be tuned in sign with effects prescribed in the model condition PSR, unstable FOES models will exhibit further probability sensitivities, as outlined in the following extension of Theorem \ref{thm:degenFOES}.
 
 \BeginKnitrBlock{corollary}
-\protect\hypertarget{cor:sign}{}{\label{cor:sign} }Let
-\(P_{\boldsymbol \theta_N}\), with support \(\mathcal{X}^N\),
-\(N\geq 1\), be a sequence of FOES models satisfying model condition
-PSR. If the models \(P_{\boldsymbol \theta_N}\) are additionally
-S-unstable, then
+\protect\hypertarget{cor:sign}{}{\label{cor:sign} }Let \(P_{\boldsymbol \theta_N}\), with support \(\mathcal{X}^N\), \(N\geq 1\), be a sequence of FOES models satisfying model condition PSR.
+If the models \(P_{\boldsymbol \theta_N}\) are additionally S-unstable, then
 \begin{enumerate}[(i)]
@@
 
-For unstable models, Corollary \ref{cor:sign} shows that shifts in
-parameters around zero (i.e., from \(\boldsymbol \theta_N\) to
-\(-\boldsymbol \theta_N\)) can induce extreme changes in probability
-among subsets of the sample space, as another manifestation of
-instability and hyper-sensitivity in probability structure. For
-one-parameter exponential families, involving a fixed real-valued linear
-parameter \(\boldsymbol \theta_N = \theta \in \mathbb{R}\) and
-sufficient statistic \(\boldsymbol g_N(\boldsymbol x)\in \mathbb{R}\) in
-\eqref{eq:expo}, \citet[Theorem 3]{schweinberger2011instability} proved a
-result similar in spirit, though based on a characterization there in
-terms of maximum
-\(U_N \equiv \max_{\boldsymbol x\in\mathcal{X}^N}g_N(\boldsymbol x)\)
-and minimal
-\(L_N \equiv \min_{\boldsymbol x\in\mathcal{X}^N}g_N(\boldsymbol x)\)
-values of the sufficient statistic. For this case in particular, mode
-sets have specific, and essentially complementary, forms over positive
-and negative parameters, namely,
-\(\mathcal{M}_{\epsilon, \boldsymbol \theta_N} = \{\boldsymbol x \in\mathcal{X}^N: g_N(\boldsymbol x) > (1-\epsilon) U_N + \epsilon L_N \}\)
-and
-\(\mathcal{M}_{\epsilon, -\boldsymbol \theta_N} = \{\boldsymbol x \in\mathcal{X}^N: g_N(\boldsymbol x) < \epsilon U_N + (1-\epsilon) L_N \}\)
-for any \(\boldsymbol \theta_N>0\), and \citet[Theorem
-3]{schweinberger2011instability} showed each mode set collects all mass,
-under positive and negative parameters, respectively, with unstable
-models of this exponential type. However, for all unstable FOES models,
-Corollary \ref{cor:sign} generalizes the same principle that unstable
-models can push all probability to different, and in fact disjoint,
-parts of the sample space, depending on how parameters fall with respect
-to zero. This feature can numerically complicate likelihood
-manipulations, such as maximization or MCMC-based Bayes posterior
-sampling, as further discussed in Section \ref{implications}.
+For unstable models, Corollary \ref{cor:sign} shows that shifts in parameters around zero (i.e., from \(\boldsymbol \theta_N\) to \(-\boldsymbol \theta_N\)) can induce extreme changes in probability among subsets of the sample space, as another manifestation of instability and hyper-sensitivity in probability structure.
+For one-parameter exponential families, involving a fixed real-valued linear parameter \(\boldsymbol \theta_N = \theta \in \mathbb{R}\) and sufficient statistic \(\boldsymbol g_N(\boldsymbol x)\in \mathbb{R}\) in \eqref{eq:expo}, \citet[Theorem 3]{schweinberger2011instability} proved a result similar in spirit, though based on a characterization there in terms of maximum \(U_N \equiv \max_{\boldsymbol x\in\mathcal{X}^N}g_N(\boldsymbol x)\) and minimal \(L_N \equiv \min_{\boldsymbol x\in\mathcal{X}^N}g_N(\boldsymbol x)\) values of the sufficient statistic.
+For this case in particular, mode sets have specific, and essentially complementary, forms over positive and negative parameters, namely, \(\mathcal{M}_{\epsilon, \boldsymbol \theta_N} = \{\boldsymbol x \in\mathcal{X}^N: g_N(\boldsymbol x) > (1-\epsilon) U_N + \epsilon L_N \}\) and \(\mathcal{M}_{\epsilon, -\boldsymbol \theta_N} = \{\boldsymbol x \in\mathcal{X}^N: g_N(\boldsymbol x) < \epsilon U_N + (1-\epsilon) L_N \}\) for any \(\boldsymbol \theta_N>0\), and
+\citet[Theorem 3]{schweinberger2011instability} showed each mode set collects all mass, under positive and negative parameters, respectively, with unstable models of this exponential type.
+However, for all unstable FOES models, Corollary \ref{cor:sign} generalizes the same principle that unstable models can push all probability to different, and in fact disjoint, parts of the sample space, depending on how parameters fall with respect to zero.
+This feature can numerically complicate likelihood manipulations, such as maximization or MCMC-based Bayes posterior sampling, as further discussed in Section \ref{implications}.
 
 \BeginKnitrBlock{remark}
-\iffalse{} {Remark. } \fi{}Under the model condition PSR, Corollary
-\ref{cor:sign} can also be extended to cases where parameter components
-\(\boldsymbol \theta_N =(\boldsymbol \theta_{1,N}, \boldsymbol \theta_{N,2})\)
-(say) are not all changed in sign (e.g., \(-\boldsymbol \theta_N\)) but,
-more generally, are instead altered to another parameter configuration
-\(\boldsymbol \theta_{N}^A = ( \boldsymbol \theta_{1,N}^A, \boldsymbol \theta_{2,N}^A )\)
-involving a switch in sign only among some dominating model parameters
-\(\boldsymbol \theta_{N,2}^A=- \boldsymbol \theta_{N,2}\) with remaining
-parameters \(\boldsymbol \theta_{1,N}^A\) being arbitrarily chosen. If a
-change sign occurs among parameters (\(\pm \boldsymbol \theta_{N,2}\))
-which dominate the probability structure of the model, then the results
-of Corollary \ref{cor:sign} can still hold with
-\(\boldsymbol \theta_{N}^A\) replacing \(-\boldsymbol \theta_N\); as an
-example of one sufficient condition, if
-\(\lim_{N\to \infty} \max_{\boldsymbol x \in \mathcal{X}^N} |G_N(\boldsymbol x, -\boldsymbol \theta_{N}) - G_N(\boldsymbol x, \boldsymbol \theta_{N}^A)|=0\)
-holds in addition to Corollary \ref{cor:sign} assumptions, where \[
+\iffalse{} {Remark. } \fi{}Under the model condition PSR, Corollary \ref{cor:sign} can also be extended to
+cases where parameter components \(\boldsymbol \theta_N =(\boldsymbol \theta_{1,N}, \boldsymbol \theta_{N,2})\) (say) are not all changed in sign (e.g., \(-\boldsymbol \theta_N\)) but, more generally, are instead altered to another parameter configuration \(\boldsymbol \theta_{N}^A = ( \boldsymbol \theta_{1,N}^A, \boldsymbol \theta_{2,N}^A )\) involving a switch in sign only among some dominating model parameters \(\boldsymbol \theta_{N,2}^A=- \boldsymbol \theta_{N,2}\) with remaining parameters \(\boldsymbol \theta_{1,N}^A\) being arbitrarily chosen.
+If a change sign occurs among parameters (\(\pm \boldsymbol \theta_{N,2}\)) which dominate the probability structure of the model, then the results of Corollary \ref{cor:sign} can still hold with \(\boldsymbol \theta_{N}^A\) replacing \(-\boldsymbol \theta_N\); as an example of one sufficient condition, if \(\lim_{N\to \infty} \max_{\boldsymbol x \in \mathcal{X}^N} |G_N(\boldsymbol x, -\boldsymbol \theta_{N}) - G_N(\boldsymbol x, \boldsymbol \theta_{N}^A)|=0\) holds in addition to Corollary \ref{cor:sign} assumptions, where
+\[
 G_N(\boldsymbol x, \boldsymbol \theta) =   \frac{\log P_{\boldsymbol \theta}(\boldsymbol x) - \min_{\boldsymbol y \in\mathcal{X}^N}\log P_{\boldsymbol \theta}(\boldsymbol y) }{\max_{\boldsymbol y \in\mathcal{X}^N}\log P_{\boldsymbol \theta}(\boldsymbol y)  - \min_{\boldsymbol y \in\mathcal{X}^N}\log P_{\boldsymbol \theta}(\boldsymbol y) }, \quad \boldsymbol x \in\mathcal{X}^N,
-\] represents a standardized form of \(\boldsymbol \theta\)-model
-probabilities, then the results of Corollary \ref{cor:sign} apply to
-\(\boldsymbol \theta_{N}^A\) in addition to \(-\boldsymbol \theta_N\).
-As a consequence, an unstable model under \(\boldsymbol \theta_N\) can
-then imply that many more unstable models exist over a broader spectrum
-of possibilities for variations \(\boldsymbol \theta_{N}^A\) of
-\(\boldsymbol \theta_N\), which involves some amount of sign change
-among components of \(\boldsymbol \theta_N\).
+\]
+represents a standardized form of \(\boldsymbol \theta\)-model probabilities, then the results of Corollary \ref{cor:sign} apply to \(\boldsymbol \theta_{N}^A\) in addition to \(-\boldsymbol \theta_N\).
+As a consequence, an unstable model under \(\boldsymbol \theta_N\) can then imply that many more unstable models exist over a broader spectrum of possibilities for variations \(\boldsymbol \theta_{N}^A\) of \(\boldsymbol \theta_N\), which involves some amount of sign change among components of \(\boldsymbol \theta_N\).
 \EndKnitrBlock{remark}
@@
 
-Model instability can depend intricately on how functions of parameters
-and data \(\boldsymbol X_N=(X_1,\ldots,X_n)\) are combined in the
-formulation of the model probabilities, though some general causes may
-be identified. As one issue, a broad parameter space (or wide
-interpretation of this space) may admit some parameters as technically
-valid that have an undue and often undesirable impact on the model
-structure for a prescribed data size \(N\). In this case, both the size
-and dimension of model parameters can be problematic and induce
-instability. In combination to this last point, further causes of
-instability may also be traced to the magnitude of statistics in the
-model. Potentially massive, and thereby unstable, statistics were the
-primary focus of instability studies of
-\citet{schweinberger2011instability} for certain discrete exponential
-models having parameters/statistics of fixed dimension. However, as
-shown in the following, bounded statistics may still lead to instability
-if the parameter dimension is high. We next provides some examples to
-illustrate S-instability in FOES models, which also suggest some
-potential strategies for preventing unstable models.
+Model instability can depend intricately on how functions of parameters and data \(\boldsymbol X_N=(X_1,\ldots,X_n)\) are combined in the formulation of the model probabilities, though some general causes may be identified.
+As one issue, a broad parameter space (or wide interpretation of this space) may admit some parameters as technically valid that have an undue and often undesirable impact on the model structure for a prescribed data size \(N\).
+In this case, both the size and dimension of model parameters can be problematic and induce instability.
+In combination to this last point, further causes of instability may also be traced to the magnitude of statistics in the model.
+Potentially massive, and thereby unstable, statistics were the primary focus of instability studies of \citet{schweinberger2011instability} for certain discrete exponential models having parameters/statistics of fixed dimension.
+However, as shown in the following, bounded statistics may still lead to instability if the parameter dimension is high.
+We next \ak{provide} some examples to illustrate S-instability in FOES models, which also suggest some potential strategies for preventing unstable models.
 
@@
 
-As a baseline for comparisons, consider a simplistic model for
-\(\boldsymbol X_N=(X_1,\ldots,X_N)\) with uniform probabilities over the
-sample space, say
-\(P_{\boldsymbol \theta_N}(\boldsymbol x)= |\mathcal{X}|^{-N}\),
-\(\boldsymbol x \in \mathcal{X}^N\), where each random variable has
-\(|\mathcal{X}| \geq 1\) outcomes. In contrast to instability, model
-probabilities here are completely insensitive to changes in data
-outcomes across the sample space, and the associated log-ratio of
-extreme probabilities \eqref{eq:elpr} is \[
+As a baseline for comparisons, consider a simplistic model for \(\boldsymbol X_N=(X_1,\ldots,X_N)\) with uniform probabilities over the sample space, say \(P_{\boldsymbol \theta_N}(\boldsymbol x)= |\mathcal{X}|^{-N}\), \(\boldsymbol x \in \mathcal{X}^N\), where each random variable has \(|\mathcal{X}| \geq 1\) outcomes.
+In contrast to instability, model probabilities here are completely insensitive to changes in data outcomes across the sample space, and the associated log-ratio of extreme probabilities \eqref{eq:elpr} is
+\[
 \frac{1}{N} \REP(\boldsymbol \theta_N)=0\quad \;(\text{uniform probability model}),
-\] which is as small as possible. In fact, a LREP value of zero can only
-occur for a FOES model having uniform probabilities, and such
-equi-probability models are always S-stable.
+\]
+which is as small as possible. In fact, a LREP value of zero can only occur for a FOES model having uniform probabilities, and such equi-probability models are always S-stable.
 
@@
 
-A fundamental model considered in the instability work of
-\citet{schweinberger2011instability} involves a one-parameter
-exponential model corresponding to \eqref{eq:expo} with a real-valued
-parameter, say
-\(\boldsymbol \theta_N = \eta(\boldsymbol \theta_N)\in \mathbb{R}\), and
-sufficient statistic \(\boldsymbol g_N(\boldsymbol x)\in \mathbb{R}\).
-For such models, upon scaling by sample size \(N\), the log-ratio of
-extreme probabilities in \eqref{eq:elpr} for assessing instability becomes
+A fundamental model considered in the instability work of \citet{schweinberger2011instability} involves a one-parameter exponential model corresponding to \eqref{eq:expo} with a real-valued parameter, say \(\boldsymbol \theta_N = \eta(\boldsymbol \theta_N)\in \mathbb{R}\), and sufficient statistic \(\boldsymbol g_N(\boldsymbol x)\in \mathbb{R}\).
+For such models, upon scaling by sample size \(N\), the log-ratio of extreme probabilities in \eqref{eq:elpr} for assessing instability becomes
 \begin{equation}
@@
 \frac{1}{N}\REP(\boldsymbol \theta_N ) \equiv   |\boldsymbol \theta_N| \frac{(U_N-L_N)}{N} \;\quad \text{(one-parameter exponential model)},
-\end{equation} where
-\(U_N \equiv \max_{\boldsymbol x\in\mathcal{X}^N}g_N(\boldsymbol x)\)
-and
-\(L_N \equiv \min_{\boldsymbol x\in\mathcal{X}^N}g_N(\boldsymbol x)\)
-denote the maximal and minimal values of the single sufficient
-statistic. In this case, an S-unstable model results, by definition
-\eqref{eq:Sun}, whenever
-\(\lim_{N\to \infty} |\boldsymbol \theta_N| (U_N-L_N)/N= \infty\) holds
-or, in other words, if the combined magnitudes of parameter
-\(|\boldsymbol \theta_N|\) and maximal difference \(U_N-L_N\) in
-statistic values are overwhelmingly large relative to the sample size
-\(N\). If we further assume that
-\(\boldsymbol \theta_N =\theta\in\mathbb{R}\setminus \{0\}\) is a fixed
-(non-zero) parameter for all \(N \geq 1\), as considered in
-\citet{schweinberger2011instability}, then an S-unstable model results
-solely if the sufficient statistic admits a value \(U_N-L_N\) too large
-relative to number \(N\) of observations, i.e., if
-\((U_N-L_N)/N\to \infty\) as \(N\to \infty\). The latter aspect reflects
-the definition of \citet{schweinberger2011instability}, for this
-setting, that a real-valued \emph{statistic} \(g_N(\boldsymbol x)\) may
-be classified as \emph{unstable} when
-\(\lim_{N\to \infty}|(U_N-L_N)/N=\infty\) holds and as \emph{stable}
-otherwise (e.g., if \(\sup_{N \geq 1}(U_N-L_N)/N<\infty\)).
-
-For illustration, consider the iid Bernoulli model \eqref{eq:mod1} for
-\(\boldsymbol X_N=(X_1,\ldots,X_N)\) with log-odds ratio parameter
-\(\boldsymbol \theta_N = \log[ P_{\boldsymbol \theta_N}(X_1=1)/ P_{\boldsymbol \theta_N}(X_1=0)]\in\mathbb{R}\).
-Remark 1 (Section \ref{criterion}) then gives the model instability
-measure \eqref{eq:Sun} directly as \[
+\end{equation}
+where \(U_N \equiv \max_{\boldsymbol x\in\mathcal{X}^N}g_N(\boldsymbol x)\) and \(L_N \equiv \min_{\boldsymbol x\in\mathcal{X}^N}g_N(\boldsymbol x)\) denote the maximal and minimal values of the single sufficient statistic.
+In this case, an S-unstable model results, by definition \eqref{eq:Sun}, whenever \(\lim_{N\to \infty} |\boldsymbol \theta_N| (U_N-L_N)/N= \infty\) holds or, in other words, if the combined magnitudes of parameter \(|\boldsymbol \theta_N|\) and maximal difference \(U_N-L_N\) in statistic values are overwhelmingly large relative to the sample size \(N\).
+If we further assume that \(\boldsymbol \theta_N =\theta\in\mathbb{R}\setminus \{0\}\) is a fixed (non-zero) parameter for all \(N \geq 1\), as considered in \citet{schweinberger2011instability}, then an S-unstable model results solely if the sufficient statistic admits a value \(U_N-L_N\) too large relative to number \(N\) of observations, i.e., if \((U_N-L_N)/N\to \infty\) as \(N\to \infty\).
+The latter aspect reflects the definition of \citet{schweinberger2011instability}, for this setting, that a real-valued \emph{statistic} \(g_N(\boldsymbol x)\) may be classified as \emph{unstable} when \(\lim_{N\to \infty}|(U_N-L_N)/N=\infty\) holds and as \emph{stable} otherwise (e.g., if \(\sup_{N \geq 1}(U_N-L_N)/N<\infty\)).
+
+For illustration, consider the iid Bernoulli model \eqref{eq:mod1} for \(\boldsymbol X_N=(X_1,\ldots,X_N)\) with log-odds ratio parameter \(\boldsymbol \theta_N = \log[ P_{\boldsymbol \theta_N}(X_1=1)/ P_{\boldsymbol \theta_N}(X_1=0)]\in\mathbb{R}\).
+\ak{Remark 3.1} (Section \ref{criterion}) then gives the model instability measure \eqref{eq:Sun} directly as
+\[
 \frac{1}{N}\REP(\boldsymbol \theta_N ) = |\boldsymbol \theta_N|\quad\; \text{(iid Bernoulli model)},
-\] so that an unstable (or stable) model results for a divergent (or
-bounded) parameter sequence \(|\boldsymbol \theta_N|\). The above
-instability expression for the Bernoulli model follows as well from the
-\(N\)-scaled LREP value \eqref{eq:UL} for a one-parameter exponential
-distribution, using that the sufficient statistic involved
-\(g_N(\boldsymbol x)= \sum_{i=1}^N x_i\),
-\(\boldsymbol x =(x_1,\ldots,x_n)\in\{0,1\}^N\), has maximum and minimum
-values \(U_N=N\) and \(L_N=0\). In this case,
-\citet{schweinberger2011instability} has noted that the statistic is
-stable (i.e., bounded \((U_N-L_N)/N=1\)) and the Bernoulli model is as
-well when, in particular,
-\(\boldsymbol \theta_N=\bold \theta \in\mathbb{R}\) is fixed for
-\(N \geq 1\).
-
-Alternatively, considering a random graph with \(N={n \choose 2}\) edges
-among \(n\) nodes, the exponential graph model from \eqref{eq:mod2}, when
-based purely on the number of \(g_{2,N}(\boldsymbol x)\) of 2-stars or
-solely the number \(g_{3,N}(\boldsymbol x)\) of triangles,
-\(\boldsymbol x\in\{0,1\}^N\), has an measure of instability from
-\eqref{eq:Sun} as \[
+\]
+so that an unstable (or stable) model results for a divergent (or bounded) parameter sequence \(|\boldsymbol \theta_N|\).
+The above instability expression for the Bernoulli model follows as well from the \(N\)-scaled LREP value \eqref{eq:UL} for a one-parameter exponential distribution, using that the sufficient statistic involved \(g_N(\boldsymbol x)= \sum_{i=1}^N x_i\), \(\boldsymbol x =(x_1,\ldots,x_n)\in\{0,1\}^N\), has maximum and minimum values \(U_N=N\) and \(L_N=0\).
+In this case, \citet{schweinberger2011instability} has noted that the statistic is stable (i.e., bounded \((U_N-L_N)/N=1\)) and the Bernoulli model is as well when, in particular, \(\boldsymbol \theta_N=\boldsymbol \theta \in\mathbb{R}\) is fixed for \(N \geq 1\).
+
+Alternatively, considering a random graph with \(N={n \choose 2}\) edges among \(n\) nodes, the exponential graph model from \eqref{eq:mod2}, when based purely on the number of \(g_{2,N}(\boldsymbol x)\) of 2-stars or solely the number \(g_{3,N}(\boldsymbol x)\) of triangles, \(\boldsymbol x\in\{0,1\}^N\), has a measure of instability from \eqref{eq:Sun} as
+\[
 \frac{1}{N}\REP(\boldsymbol \theta_N )  = \left\{ \begin{array}{lcl} |\boldsymbol \theta_N| (n-2) && \text{(2-star graph model)}\\
 |\boldsymbol \theta_N|(n-2)/3 &&\text{(triangle graph model)},\end{array}\right.
-\] by using the (one-parameter exponential) LREP formula \eqref{eq:UL}
-with statistic maximums \(U_N= N(n-2)\) for 2-stars or \(U_N= N(n-2)/3\)
-for triangles and with minimums \(L_N=0\) in both cases. Because the
-variable number \(N\to \infty\) as the node number \(n\to \infty\), both
-counts of 2-stars and triangles are unstable statistics in the sense of
-\citet{schweinberger2011instability} (i.e.,
-\(\lim_{N\to \infty} (U_N-L_N)/N=\infty\)). Furthermore, both types of
-graph models are always S-unstable for all possible of parameter
-sequences \(\boldsymbol \theta_N \in\mathbb{R}\) that are bounded away
-from zero (i.e.,
-\(\lim_{N\to \infty}\REP(\boldsymbol \theta_N )/N=\infty\) then holds,
-including the fixed parameter case
-\(\boldsymbol \theta_N=\theta\in\mathbb{R}\setminus \{0\}\) from
-\citet{schweinberger2011instability}).
+\]
+by using the (one-parameter exponential) LREP formula \eqref{eq:UL} with statistic maximums \(U_N= N(n-2)\) for 2-stars or \(U_N= N(n-2)/3\) for triangles and with minimums \(L_N=0\) in both cases.
+Because the variable number \(N\to \infty\) as the node number \(n\to \infty\), both counts of 2-stars and triangles are unstable statistics in the sense of \citet{schweinberger2011instability} (i.e., \(\lim_{N\to \infty} (U_N-L_N)/N=\infty\)).
+Furthermore, both types of graph models are always S-unstable for all possible of parameter sequences \(\boldsymbol \theta_N \in\mathbb{R}\) that are bounded away from zero (i.e., \(\lim_{N\to \infty}\REP(\boldsymbol \theta_N )/N=\infty\) then holds, including the fixed parameter case \(\boldsymbol \theta_N=\theta\in\mathbb{R}\setminus \{0\}\) from \citet{schweinberger2011instability}).
 
 \hypertarget{fixed-dim-exp}{%
-\subsection{Fixed-dimensional Linear Exponential
-Models}\label{fixed-dim-exp}}
-
-As a generalization of the one-parameter exponential case, we next
-consider linear exponential families \eqref{eq:expo} with \(k\) parameters
-\(\boldsymbol \theta_N = (\theta_{1,N},\ldots,\theta_{k,N})^\prime\) and
-\(k\) sufficient statistics
-\(\boldsymbol g_N(\boldsymbol x) = (g_{1,N}(\boldsymbol x),\ldots, g_{k,N}(\boldsymbol x))^\prime\).
-Here the dimension \(k\) of model parameters/statistics is fixed, and we
-next prescribe a condition helpful to avoiding instability in such
-models. For this, define
-\(U_{i,N}=\max_{\boldsymbol x \in\mathcal{X}^N} g_{i,N}(\boldsymbol x)\)
-and
-\(L_{i,N}=\min_{\boldsymbol x \in\mathcal{X}^N} g_{i,N}(\boldsymbol x)\)
-as the maximal and minimal values of the \(i\)th statistic,
-\(i=1,\ldots,k\), based on observations
-\(\boldsymbol X_N=(X_1,\ldots,X_N)\).
+\subsection{Fixed-dimensional Linear Exponential Models}\label{fixed-dim-exp}}
+
+As a generalization of the one-parameter exponential case, we next consider linear exponential families \eqref{eq:expo} with \(k\) parameters \(\boldsymbol \theta_N = (\theta_{1,N},\ldots,\theta_{k,N})^\prime\) and \(k\) sufficient statistics \(\boldsymbol g_N(\boldsymbol x) = (g_{1,N}(\boldsymbol x),\ldots, g_{k,N}(\boldsymbol x))^\prime\).
+Here the dimension \(k\) of model parameters/statistics is fixed, and we next prescribe a condition helpful to avoiding instability in such models.
+For this, define \(U_{i,N}=\max_{\boldsymbol x \in\mathcal{X}^N} g_{i,N}(\boldsymbol x)\) and \(L_{i,N}=\min_{\boldsymbol x \in\mathcal{X}^N} g_{i,N}(\boldsymbol x)\) as the maximal and minimal values of the \(i\)th statistic, \(i=1,\ldots,k\), based on observations \(\boldsymbol X_N=(X_1,\ldots,X_N)\).
 
 \BeginKnitrBlock{proposition}
-\protect\hypertarget{prp:prop1}{}{\label{prp:prop1} }Let
-\(P_{\boldsymbol \theta_N}\), \(N \geq 1\), denote linear exponential
-models (\ref{eq:expo}) with parameters
-\(\boldsymbol \theta_N = (\theta_{1,N},\ldots,\theta_{k,N})^\prime \in \mathbb{R}^k\)
-and statistics
-\(\boldsymbol g_N(\boldsymbol x) = (g_{1,N}(\boldsymbol x),\ldots, g_{k,N}(\boldsymbol x))^\prime \in \mathbb{R}^k\),
-for fixed \(k \geq 1\). Then, the models \(P_{\boldsymbol \theta_N}\)
-are S-stable if \begin{equation}
+\protect\hypertarget{prp:prop1}{}{\label{prp:prop1} }Let \(P_{\boldsymbol \theta_N}\), \(N \geq 1\), denote linear exponential models (\ref{eq:expo}) with parameters \(\boldsymbol \theta_N = (\theta_{1,N},\ldots,\theta_{k,N})^\prime \in \mathbb{R}^k\) and statistics \(\boldsymbol g_N(\boldsymbol x) = (g_{1,N}(\boldsymbol x),\ldots, g_{k,N}(\boldsymbol x))^\prime \in \mathbb{R}^k\), for fixed \(k \geq 1\).
+Then, the models \(P_{\boldsymbol \theta_N}\) are S-stable if
+\begin{equation}
 \label{eq:prop1}
 \sup_{N \geq 1}\frac{1}{N} \max_{1 \leq i \leq k }|\theta_{i,N}|(U_{i,N}-L_{i,N})<\infty
-\end{equation} holds, i.e., if
-\(\max_{1 \leq i \leq k } |\theta_{i,N}|(U_{i,N}-L_{i,N})/N\) is bounded
-sequence of sample size \(N\).
+\end{equation}
+holds, i.e., if \(\max_{1 \leq i \leq k } |\theta_{i,N}|(U_{i,N}-L_{i,N})/N\) is bounded sequence of sample size \(N\).
 \EndKnitrBlock{proposition}
@@
 \BeginKnitrBlock{remark}
-\iffalse{} {Remark. } \fi{}In the one-parameter exponential case
-\(k=1\), recall the exponential model is stable/unstable depending on
-whether
-\(\REP(\boldsymbol \theta_N)/N = |\theta_{1,N}|(U_{1,N}-L_{1,N})/N \equiv |\boldsymbol \theta_{N}|(U_{N}-L_{N})/N\)
-in \eqref{eq:UL} is convergent/divergent. Hence, for \(k=1\), the
-condition \eqref{eq:prop1} of Proposition \ref{prp:prop1} captures the
-same notion of S-stability based on \eqref{eq:UL}.
+\iffalse{} {Remark.
+} \fi{}In the one-parameter exponential case \(k=1\), recall the exponential model is stable/unstable depending on whether \(\REP(\boldsymbol \theta_N)/N = |\theta_{1,N}|(U_{1,N}-L_{1,N})/N \equiv |\boldsymbol \theta_{N}|(U_{N}-L_{N})/N\) in \eqref{eq:UL} is convergent/divergent.
+Hence, for \(k=1\), the condition \eqref{eq:prop1} of Proposition \ref{prp:prop1} captures the same notion of S-stability based on \eqref{eq:UL}.
 \EndKnitrBlock{remark}
 
-Proposition \ref{prp:prop1} provides a sufficient condition for the
-stability of linear exponential models with fixed parameter dimension
-\(k\geq 1\), whereby an S-stable model is guaranteed if the compounded
-magnitude of each combination of parameter \(\theta_{i,N}\) and
-sufficient statistic value \((U_{i,N}-L_{i,N})\) is bounded by the
-sample size \(N\), \(i=1,\ldots,k\). This supports the findings of
-\citet{schweinberger2011instability}, who showed degeneracy follows in
-such models under one type of violation of the condition \eqref{eq:prop1}
-in Proposition \ref{prp:prop1} (namely, involving \(k>1\) non-zero
-parameters with \(k-1\) statistics being \(O(N)\) bounded while one
-statistic diverges in maximal size faster than the number \(N\) of
-observations). To further illustrate the result in Proposition
-\ref{prp:prop1}, consider the multinomial distribution \eqref{eq:mod11}
-for \(\boldsymbol X_N=(X_1,\ldots,X_N)\) having \(k\geq 2\) categories
-\(\{1,\ldots,k\}\) and \(k\) parameters
-\(\boldsymbol \theta_N = (\theta_{1,N},\ldots,\theta_{k,N})^\prime\).
-The variables are iid under this model so that Remark 1 (Section
-\ref{criterion} yields the corresponding \(N\)-scaled log-ratio of
-extreme probabilities \eqref{eq:elpr} as \begin{align*}
+Proposition \ref{prp:prop1} provides a sufficient condition for the stability of linear exponential models with fixed parameter dimension \(k\geq 1\), whereby an S-stable model is guaranteed if the \ak{product of magnitudes} of each combination of parameter \(\theta_{i,N}\) and sufficient statistic value \((U_{i,N}-L_{i,N})\) is bounded by the sample size \(N\), \(i=1,\ldots,k\).
+This supports the findings of \citet{schweinberger2011instability}, who showed degeneracy follows in such models under one type of violation of the condition \eqref{eq:prop1} in Proposition \ref{prp:prop1} (namely, involving \(k>1\) non-zero parameters with \(k-1\) statistics being \(O(N)\) bounded while one statistic diverges in maximal size faster than the number \(N\) of observations).
+To further illustrate the result in Proposition \ref{prp:prop1}, consider the multinomial distribution \eqref{eq:mod11} for \(\boldsymbol X_N=(X_1,\ldots,X_N)\) having \(k\geq 2\) categories \(\{1,\ldots,k\}\) and \(k\) parameters \(\boldsymbol \theta_N = (\theta_{1,N},\ldots,\theta_{k,N})^\prime\).
+The variables are iid under this model so that \ak{Remark 3.1} (Section \ref{criterion} yields the corresponding \(N\)-scaled log-ratio of extreme probabilities \eqref{eq:elpr} as
+\begin{align*}
   \frac{1}{N}\REP(\boldsymbol \theta_N) &= \frac{\max_{1 \leq i \leq k} P_{\boldsymbol \theta_N}(X_1=i)}{\min_{1 \leq i \leq k} P_{\boldsymbol \theta_N}(X_1=i)}\\
   &= \max_{1 \leq i \leq k} \theta_{i,N} - \min_{1 \leq i \leq k} \theta_{i,N} \qquad \text{(iid multinomial model)}.
-\end{align*} Hence, a multinomial model sequence is unstable (or stable)
-depending on whether (or not) the maximal parameter difference
-\(\max_{1 \leq i \leq k} \theta_{i,N} - \min_{1 \leq i \leq k} \theta_{i,N}\)
-diverges. Furthermore, using that each of the \(k\) sufficient (count)
-statistics from the multinomial model \eqref{eq:mod11} satisfies
-\((U_{i,N}-L_{i,N})/N=1\), we see that \eqref{eq:prop1} of Proposition
-\ref{prp:prop1} becomes purely a parameter condition,
-\(\sup_{N \geq 1}\max_{1 \leq i \leq k } |\theta_{i,N}| <\infty\), for
-ensuring that
-\(\REP(\boldsymbol \theta_N)/N =\max_{1 \leq i \leq k} \theta_{i,N} - \min_{1 \leq i \leq k} \theta_{i,N}\)
-is bounded and stability follows for the multinomial distribution.
-Additionally, a stable multinomial sequence (i.e., bounded
-\(\REP(\boldsymbol \theta_N)/N\)) turns out to be nearly equivalent to
-\eqref{eq:prop1} (e.g., these are the same if the smallest parameter
-\(\min_{1 \leq i \leq k } |\theta_{i,N}|\) remains bounded).
-
-When the condition \eqref{eq:prop1} of Proposition \ref{prp:prop1} is
-violated, this aspect suggests a potentially unstable model that may be
-investigated more closely. For example, consider the exponential graph
-model from \eqref{eq:mod2} involving counts of edges, 2-stars and
-triangles with fixed parameters
-\(\boldsymbol \theta_N = (\theta_{1},\theta_2,\theta_3)^\prime \in \mathbb{R}^3\)
-for \(N\geq 1\). If either the 2-star parameter \(\theta_2 \neq 0\) or
-triangle parameter \(\theta_3 \neq 0\) is non-zero, then
-\(\max_{1 \leq i \leq 3 } |\theta_{i}|(U_{i,N}-L_{i,N})/N \propto (n-2)\to \infty\)
-holds in \eqref{eq:prop1} by
-\((U_{2,N}-L_{2,N})/N = 3 (U_{3,N}-L_{3,N})/N=(n-2)\) for 2-star and
-triangle statistics (\(i=2,3\)), so that Proposition \ref{prp:prop1}
-hints that an unstable model may result when
-\(|\theta_2| + |\theta_3| \neq 0\). Relatedly, a result from
-\citet[Result 3]{schweinberger2011instability} states that this model is
-unstable for all fixed parameters excluding cases
-\(\theta_2 =\theta_3=0\) or \(\theta_2 = - \theta_3/3\). However, more
-is true in line with the instability suggested by Proposition
-\ref{prp:prop1} whenever \(|\theta_2| + |\theta_3| \neq 0\) (i.e.,
-excluding \(\theta_2 =\theta_3=0\)).
-
-To see this, consider an even number \(n>2\) of nodes and let
-\(\boldsymbol x_0\) denote the data outcome in
-\(\mathcal{X}^N \equiv \{0,1\}^N\) with all \(N = {n \choose 2}\) edges
-being zero, let \(\boldsymbol x_1\) denote the outcome with all edges
-being 1, and let \(\boldsymbol x_2\) denote the edge configuration from
-dividing the nodes into two equal groups, with no edges within a group
-and all edges between the groups (so that no triangles exist in
-\(\boldsymbol x_2\)). Then, the \(N\)-scaled log-ratio \eqref{eq:elpr} for
-the exponential graph model \eqref{eq:mod2} can, by definition, be bounded
-below by \begin{align*}
+\end{align*}
+Hence, a multinomial model sequence is unstable (or stable) depending on whether (or not) the maximal parameter difference \(\max_{1 \leq i \leq k} \theta_{i,N} - \min_{1 \leq i \leq k} \theta_{i,N}\) diverges.
+Furthermore, using that each of the \(k\) sufficient (count) statistics from the multinomial model \eqref{eq:mod11} satisfies \((U_{i,N}-L_{i,N})/N=1\), we see that \eqref{eq:prop1} of Proposition \ref{prp:prop1} becomes purely a parameter condition, \(\sup_{N \geq 1}\max_{1 \leq i \leq k } |\theta_{i,N}| <\infty\), for ensuring that \(\REP(\boldsymbol \theta_N)/N =\max_{1 \leq i \leq k} \theta_{i,N} - \min_{1 \leq i \leq k} \theta_{i,N}\) is bounded and stability follows for the multinomial distribution.
+Additionally, a stable multinomial sequence (i.e., bounded \(\REP(\boldsymbol \theta_N)/N\)) turns out to be nearly equivalent to \eqref{eq:prop1} (e.g., these are the same if the smallest parameter \(\min_{1 \leq i \leq k } |\theta_{i,N}|\) remains bounded).
+
+When the condition \eqref{eq:prop1} of Proposition \ref{prp:prop1} is violated, this aspect suggests a potentially unstable model that may be investigated more closely.
+For example, consider the exponential graph model from \eqref{eq:mod2} involving counts of edges, 2-stars and triangles with fixed parameters \(\boldsymbol \theta_N = (\theta_{1},\theta_2,\theta_3)^\prime \in \mathbb{R}^3\) for \(N\geq 1\).
+If either the 2-star parameter \(\theta_2 \neq 0\) or triangle parameter \(\theta_3 \neq 0\) is non-zero, then \(\max_{1 \leq i \leq 3 } |\theta_{i}|(U_{i,N}-L_{i,N})/N \propto (n-2)\to \infty\) holds in \eqref{eq:prop1} by \((U_{2,N}-L_{2,N})/N = 3 (U_{3,N}-L_{3,N})/N=(n-2)\) for 2-star and triangle statistics (\(i=2,3\)), so that Proposition \ref{prp:prop1} hints that an unstable model may result when \(|\theta_2| + |\theta_3| \neq 0\).
+Relatedly, a result from \citet[Result 3]{schweinberger2011instability} states that this model is unstable for all fixed parameters excluding cases \(\theta_2 =\theta_3=0\) or \(\theta_2 = - \theta_3/3\)\ak{, more in line} with the instability suggested by Proposition \ref{prp:prop1} whenever \(|\theta_2| + |\theta_3| \neq 0\) (i.e., excluding \(\theta_2 =\theta_3=0\)).
+
+To see this, consider an even number \(n>2\) of nodes and let \(\boldsymbol x_0\) denote the data outcome in \(\mathcal{X}^N \equiv \{0,1\}^N\) with all \(N = {n \choose 2}\) edges being zero, let \(\boldsymbol x_1\) denote the outcome with all edges being 1, and let \(\boldsymbol x_2\) denote the edge configuration from dividing the nodes into two equal groups, with no edges within a group and all edges between the groups (so that no triangles exist in \(\boldsymbol x_2\)).
+Then, the \(N\)-scaled log-ratio \eqref{eq:elpr} for the exponential graph model \eqref{eq:mod2} can, by definition, be bounded below by
+\begin{align*}
 \frac{1}{N}\REP_N(\boldsymbol \theta_N) &\geq \max_{i=1,2}\frac{1}{N}
@@
 &= (n-2) \max\left\{ \left| \theta_2 + \frac{\theta_3}{3}+\frac{\theta_1}{n-2} \right|, \frac{n}{4(n-1)} \left| \theta_2 + \frac{8\theta_1}{n-2} \right| \right\};
-\end{align*} a similar expression also holds for an odd node number
-\(n>2\). Consequently, for all fixed parameters excluding
-\(\theta_2=\theta_3=0\),
-\(\lim_{N\to \infty}\REP_N(\boldsymbol \theta_N)/N=\infty\) then follows
-and the graph model with 2-stars and triangles is S-unstable, as
-suggested by the breach of Proposition \ref{prp:prop1} for this model
-when \(|\theta_2|+|\theta_3|\neq 0\). That is, instability holds even
-under \(\theta_2 = - \theta_3/3\) case potentially allowed by
-Schweinberger's \citeyearpar{schweinberger2011instability} results.
+\end{align*}
+a similar expression also holds for an odd node number \(n>2\).
+Consequently, for all fixed parameters excluding \(\theta_2=\theta_3=0\), \(\lim_{N\to \infty}\REP_N(\boldsymbol \theta_N)/N=\infty\) then follows and the graph model with 2-stars and triangles is S-unstable, as suggested by the breach of Proposition \ref{prp:prop1} for this model when \(|\theta_2|+|\theta_3|\neq 0\).
+That is, instability holds even under \ak{the} \(\theta_2 = - \theta_3/3\) case potentially allowed by Schweinberger's \citeyearpar{schweinberger2011instability} results.
 
 \hypertarget{latent-variable-models-of-increasing-parameter-dimension}{%
-\subsection{Latent Variable Models of Increasing Parameter
-Dimension}\label{latent-variable-models-of-increasing-parameter-dimension}}
-
-We next consider instability of discrete data models based on
-exponential formulations involving hidden, or latent, variables, such as
-those probabilistic graphical models described in Sections
-\ref{rbm}-\ref{deep-learning}. We will focus on restricted Boltzmann
-machine (RBM) models (Section \ref{rbm}, having one layer of latent
-variables for simplicity, though the same instability concepts may be
-extended to other deep learning models (Section \ref{deep-learning}. For
-\(N\) visible variables
-\(\boldsymbol X \equiv \boldsymbol X_N = (X_1,\ldots,X_n)\) as data,
-each observation \(X_i\in\{\pm 1\}\) being binary, the RBM-based model
-\eqref{eq:RBM2} for \(\boldsymbol X\) is again of FOES-type, though not an
-exponential model. However, the distribution of visible variables is
-induced by an underlying joint exponential model \eqref{eq:RBM1} for both
-visible and latent variables \((\boldsymbol X, \boldsymbol H)\), where
-\(\boldsymbol H = (H_1,\ldots,H_{N_{\mathcal{H}}})\) denotes a vector of
-\(N_{\mathcal{H}}\) hidden variables (similarly binary). The joint model
-is of linear exponential form involving
-\(q(N)\equiv N + N_{\mathcal{H}} + N*N_{\mathcal{H}}\) sufficient
-statistics given by
-\((\boldsymbol X, \boldsymbol H, \boldsymbol X^T\boldsymbol H)\) and
-parameters
-\(\boldsymbol \theta_N = (\boldsymbol \theta_N^{\mathcal{V}},\boldsymbol \theta_N^{\mathcal{H}}, \boldsymbol \theta_N^{\mathcal{VH}} ) \in\mathbb{R}^{q(N)}\)
-corresponding to the \(N\) visible variables \(\boldsymbol X\) (i.e.,
-\(\boldsymbol \theta_N^{\mathcal{V}}\in\mathbb{R}^N\)), the
-\(N_{\mathcal{H}}\) hidden variables \(\boldsymbol H\) (i.e.,
-\(\boldsymbol \theta_N^{\mathcal{H}}\in\mathbb{R}^{N_{\mathcal{H}}}\)),
-and the \(N *N_{\mathcal{H}}\) cross-product variables
-\(\boldsymbol X^T\boldsymbol H\) (i.e.,
-\(\boldsymbol \theta_N^{\mathcal{VH}}\in\mathbb{R}^{N*N_{\mathcal{H}}}\)).
-However, unlike some previous exponential models considered in Sections
-\ref{one-param-exp}-\ref{fixed-dim-exp} (cf.~Proposition
-\ref{prp:prop1}, note that the RBM formulation always associates
-parameters with \emph{bounded} statistics (i.e., the components of
-\((\boldsymbol X, \boldsymbol H, \boldsymbol X^T\boldsymbol H)\)) so
-that model instability cannot arise here due to the magnitude of
-sufficient statistics exceeding the sample size \(N\). Instead, RBM
-instability may be linked solely to parameter configuration and the fact
-that the number \(q(N) \geq N\) of parameters necessarily increases with
-the number \(N\) of observations \(\boldsymbol X\), in contrast to
-previous exponential cases of fixed parameter dimension.
-
-To highlight the instability issues for the RBM model, consider a simple
-model for \(N\) visibles \(\boldsymbol X\) with no hidden variables
-(\(N_{\mathcal{H}}=0\)), for which model statements
-\eqref{eq:RBM1}-\eqref{eq:RBM2} coincide. An independence model then results
-for variables in \(\boldsymbol X\), which has \(q(N)=N\) parameters
-\(\boldsymbol \theta_N^{\mathcal{V}} = (\theta_{1,N}^{\mathcal{V}}, \ldots, \theta_{N,N}^{\mathcal{V}}) \in \mathbb{R}^{N}\),
-and the measure of model instability becomes \[
+\subsection{Latent Variable Models of Increasing Parameter Dimension}\label{latent-variable-models-of-increasing-parameter-dimension}}
+
+We next consider instability of discrete data models based on exponential formulations involving hidden, or latent, variables, such as those probabilistic graphical models described in Sections \ref{rbm}-\ref{deep-learning}.
+We will focus on restricted Boltzmann machine (RBM) models (Section \ref{rbm}, having one layer of latent variables for simplicity, though the same instability concepts may be extended to other deep learning models (Section \ref{deep-learning}.
+For \(N\) visible variables \(\boldsymbol X \equiv \boldsymbol X_N = (X_1,\ldots,X_n)\) as data, each observation \(X_i\in\{\pm 1\}\) being binary, the RBM-based model \eqref{eq:RBM2} for \(\boldsymbol X\) is again of FOES-type, though not an exponential model.
+However, the distribution of visible variables is induced by an underlying joint exponential model \eqref{eq:RBM1} for both visible and latent variables \((\boldsymbol X, \boldsymbol H)\), where \(\boldsymbol H = (H_1,\ldots,H_{N_{\mathcal{H}}})\) denotes a vector of \(N_{\mathcal{H}}\) hidden variables (similarly binary).
+The joint model is of linear exponential form involving \(q(N)\equiv N + N_{\mathcal{H}} + N*N_{\mathcal{H}}\) sufficient statistics given by \((\boldsymbol X, \boldsymbol H, \boldsymbol X^T\boldsymbol H)\) and parameters \(\boldsymbol \theta_N = (\boldsymbol \theta_N^{\mathcal{V}},\boldsymbol \theta_N^{\mathcal{H}}, \boldsymbol \theta_N^{\mathcal{VH}} ) \in\mathbb{R}^{q(N)}\) corresponding to the \(N\) visible variables \(\boldsymbol X\) (i.e., \(\boldsymbol \theta_N^{\mathcal{V}}\in\mathbb{R}^N\)), the \(N_{\mathcal{H}}\) hidden variables \(\boldsymbol H\) (i.e., \(\boldsymbol \theta_N^{\mathcal{H}}\in\mathbb{R}^{N_{\mathcal{H}}}\)), and the \(N *N_{\mathcal{H}}\) cross-product variables \(\boldsymbol X^T\boldsymbol H\) (i.e., \(\boldsymbol \theta_N^{\mathcal{VH}}\in\mathbb{R}^{N*N_{\mathcal{H}}}\)).
+However, unlike some previous exponential models considered in Sections \ref{one-param-exp}-\ref{fixed-dim-exp} (cf.~Proposition \ref{prp:prop1}, note that the RBM formulation always associates parameters with \emph{bounded} statistics (i.e., the components of \((\boldsymbol X, \boldsymbol H, \boldsymbol X^T\boldsymbol H)\)) so that model instability cannot arise here due to the magnitude of sufficient statistics exceeding the sample size \(N\).
+Instead, RBM instability may be linked solely to parameter configuration and the fact that the number \(q(N) \geq N\) of parameters necessarily increases with the number \(N\) of observations \(\boldsymbol X\), in contrast to previous exponential cases of fixed parameter dimension.
+
+To highlight the instability issues for the RBM model, consider a simple model for \(N\) visibles \(\boldsymbol X\) with no hidden variables (\(N_{\mathcal{H}}=0\)), for which model statements \eqref{eq:RBM1}-\eqref{eq:RBM2} coincide.
+An independence model then results for variables in \(\boldsymbol X\), which has \(q(N)=N\) parameters \(\boldsymbol \theta_N^{\mathcal{V}} = (\theta_{1,N}^{\mathcal{V}}, \ldots, \theta_{N,N}^{\mathcal{V}}) \in \mathbb{R}^{N}\), and the measure of model instability becomes
+\[
 \frac{1}{N}\REP(\boldsymbol \theta_N) = \frac{2}{N} \sum_{i=1}^N|\theta_{i,N}^{\mathcal{V}}| \quad \text{(RBM model, no hiddens)}.
-\] Hence, this model sequence for \(\boldsymbol X\) will be S-unstable
-model if the aggregation of absolute parameters grows faster than the
-number \(N\) of parameters/visible variables. Consequently, even for a
-simplest RBM model involving independence, preventing instability
-requires careful choice of parameters, particularly with regard to how a
-parameter configuration differs from zero. For more general RBM models,
-the number \(N_{\mathcal{H}}\) of hidden variables \(\boldsymbol H\) can
-also be chosen arbitrarily (i.e., as some function
-\(N_{\mathcal{H}}\equiv N_{N,\mathcal{H}}\) of \(N\)), which can
-substantially inflate the number \(q(N)\) of model parameters and
-further impact model instability through accumulated parameters. To
-better understand the effects of instability in the RBM structure,
-Proposition \ref{prp:prop2} next frames the general behavior of extreme
-probabilities in the joint RBM model \eqref{eq:RBM1} for
-\((\boldsymbol X, \boldsymbol H)\) and the implied RBM data model
-\eqref{eq:RBM2} for \(\boldsymbol X\) alone. Specifically, critical
-measures of instability may be closely connected in both models through
-tight bounds on their respective LREP values \eqref{eq:elpr}. As a result,
-Proposition \ref{prp:prop2} shows how an unstable distribution for
-observations \(\boldsymbol X\) may be traced to sources of instability
-in the original joint distribution for
-\((\boldsymbol X,\boldsymbol H)\). This also suggests a device for
-avoiding instability, as provided next.
-
-To state the result, let
-\(\REP_{\boldsymbol X}(\boldsymbol \theta_N) \equiv \REP(\boldsymbol \theta_N)\)
-denote the LREP value \eqref{eq:elpr} from the marginal distribution
-\(P_{\boldsymbol \theta_N}\) of visibles \(\boldsymbol X\) in
-\eqref{eq:RBM2} and write the LREP for the joint distribution
-\(\tilde{P}_{\boldsymbol \theta_N}\) of
-\((\boldsymbol X, \boldsymbol H)\) from \eqref{eq:RBM1} as \begin{align*}
+\]
+Hence, this model sequence for \(\boldsymbol X\) will be S-unstable model if the aggregation of absolute parameters grows faster than the number \(N\) of parameters/visible variables.
+Consequently, even for \ak{the} simplest RBM model involving independence, preventing instability requires careful choice of parameters, particularly with regard to how a parameter configuration differs from zero.
+For more general RBM models, the number \(N_{\mathcal{H}}\) of hidden variables \(\boldsymbol H\) can also be chosen arbitrarily (i.e., as some function \(N_{\mathcal{H}}\equiv N_{N,\mathcal{H}}\) of \(N\)), which can substantially inflate the number \(q(N)\) of model parameters and further impact model instability through accumulated parameters.
+To better understand the effects of instability in the RBM structure, Proposition \ref{prp:prop2} next frames the general behavior of extreme probabilities in the joint RBM model \eqref{eq:RBM1} for \((\boldsymbol X, \boldsymbol H)\) and the implied RBM data model \eqref{eq:RBM2} for \(\boldsymbol X\) alone.
+Specifically, critical measures of instability may be closely connected in both models through tight bounds on their respective LREP values \eqref{eq:elpr}.
+As a result, Proposition \ref{prp:prop2} shows how an unstable distribution for observations \(\boldsymbol X\) may be traced to sources of instability in the original joint distribution for \((\boldsymbol X,\boldsymbol H)\).
+This also suggests a device for avoiding instability, as provided next.
+
+To state the result, let \(\REP_{\boldsymbol X}(\boldsymbol \theta_N) \equiv \REP(\boldsymbol \theta_N)\) denote the LREP value \eqref{eq:elpr} from the marginal distribution \(P_{\boldsymbol \theta_N}\) of visibles \(\boldsymbol X\) in \eqref{eq:RBM2} and write the LREP for the joint distribution \(\tilde{P}_{\boldsymbol \theta_N}\) of \((\boldsymbol X, \boldsymbol H)\) from \eqref{eq:RBM1} as
+\begin{align*}
 \REP_{(\boldsymbol X, \boldsymbol H)}(\boldsymbol \theta_N) &=\log\left[  \frac{\max_{ (\boldsymbol x, \boldsymbol h) \in \{\pm 1\}^{N+N_{\mathcal{H}}}}
@@
 &= \left(\max_{ (\boldsymbol x, \boldsymbol h) \in \{\pm 1\}^{N+N_{\mathcal{H}}}} f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)-\min_{ (\boldsymbol x, \boldsymbol h) \in \{\pm 1\}^{N+N_{\mathcal{H}}}} f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)\right),
-\end{align*} written as a function \begin{equation}
+\end{align*}
+written as a function
+\begin{equation}
 \label{eq:f}
 f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)  \equiv  \sum_{i=1}^N x_i  \theta_{i,N}^{\mathcal{V}} +  \sum_{j=1}^{N_\mathcal{H}} h_j \theta_{j,N}^{\mathcal{H}} + \sum_{i=1}^N \sum_{j=1}^{N_\mathcal{H}} x_i h_j  \theta_{ij,N}^{\mathcal{VH}}
-\end{equation} of outcomes
-\(\boldsymbol x =(x_1,\ldots,x_N) \in\{\pm1\}^{N}\),
-\(\boldsymbol h =(h_1,\ldots,h_{N_{\mathcal{H}}}) \in\{\pm1\}^{N_{\mathcal{H}}}\)
-and parameters
-\(\boldsymbol \theta_N \equiv (\boldsymbol \theta_N^{\mathcal{V}},\boldsymbol \theta_N^{\mathcal{H}}, \boldsymbol \theta_N^{\mathcal{VH}})\),
-with \(\theta_{i,N}^{\mathcal{V}}\), \(\theta_{j,N}^{\mathcal{H}}\) and
-\(\theta_{ij,N}^{\mathcal{VH}}\) denoting respective parameter
-components, \(1 \leq i \leq N\), \(1 \leq j \leq N_{\mathcal{H}}\). Due
-to the marginalization steps in defining the distribution \eqref{eq:RBM2}
-of \(\boldsymbol X\), note that
-\(\REP_{\boldsymbol X}(\boldsymbol \theta_N)\) has no immediate
-analytical expression similar to that of
-\(\REP_{(\boldsymbol X, \boldsymbol H)}(\boldsymbol \theta_N)\). For
-clarity, recall also that S-instability \eqref{eq:Sun} in each model type
-refers to a respective divergence (i.e.,
-\(\lim_{N\to \infty} \REP_{(\boldsymbol X, \boldsymbol H)}(\boldsymbol \theta_N) /(N+N_{\mathcal{H}})=\infty\),
-\(\lim_{N\to \infty} \REP(\boldsymbol \theta_N) /N=\infty\)) upon
-scaling by the corresponding number of variables in a distribution. In
-the following, let \(|\boldsymbol y|_1 = \sum_{i=1}^d |y_i|\) denote the
-L1 norm of a generic vector \(\boldsymbol y =(y_1,\ldots,y_d)\),
-\(d \geq 1\). \BeginKnitrBlock{proposition}
-
-\protect\hypertarget{prp:prop2}{}{\label{prp:prop2} }Let
-\(P_{\boldsymbol \theta_N}\) denote a RBM-based data model \eqref{eq:RBM2}
-for \(N\geq 1\) visible variables
-\(\boldsymbol X \equiv \boldsymbol X_N\) derived from
-\(\tilde{P}_{\boldsymbol \theta_N}\) as the joint RBM distribution
-\eqref{eq:RBM1} of \((\boldsymbol X, \boldsymbol H)\) involving some
-number \(N_{\mathcal{H}} \equiv N_{N,\mathcal{H}}\geq 0\) of hidden
-variables \(\boldsymbol H \equiv \boldsymbol H_N\) and parameters
-\(\boldsymbol \theta_N \equiv (\boldsymbol \theta_N^{\mathcal{V}},\boldsymbol \theta_N^{\mathcal{H}}, \boldsymbol \theta_N^{\mathcal{VH}}) \in\mathbb{R}^{N}\times \mathbb{R}^{N_{\mathcal{H}}} \times \mathbb{R}^{N*N_{\mathcal{H}}}\).
+\end{equation}
+of outcomes \(\boldsymbol x =(x_1,\ldots,x_N) \in\{\pm1\}^{N}\), \(\boldsymbol h =(h_1,\ldots,h_{N_{\mathcal{H}}}) \in\{\pm1\}^{N_{\mathcal{H}}}\) and parameters \(\boldsymbol \theta_N \equiv (\boldsymbol \theta_N^{\mathcal{V}},\boldsymbol \theta_N^{\mathcal{H}}, \boldsymbol \theta_N^{\mathcal{VH}})\), with \(\theta_{i,N}^{\mathcal{V}}\), \(\theta_{j,N}^{\mathcal{H}}\) and \(\theta_{ij,N}^{\mathcal{VH}}\) denoting respective parameter components, \(1 \leq i \leq N\), \(1 \leq j \leq N_{\mathcal{H}}\).
+Due to the marginalization steps in defining the distribution \eqref{eq:RBM2} of \(\boldsymbol X\), note that \(\REP_{\boldsymbol X}(\boldsymbol \theta_N)\) has no immediate analytical expression similar to that of \(\REP_{(\boldsymbol X, \boldsymbol H)}(\boldsymbol \theta_N)\).
+For clarity, recall also that S-instability \eqref{eq:Sun} in each model type refers to a respective divergence (i.e., \(\lim_{N\to \infty} \REP_{(\boldsymbol X, \boldsymbol H)}(\boldsymbol \theta_N) /(N+N_{\mathcal{H}})=\infty\), \(\lim_{N\to \infty} \REP(\boldsymbol \theta_N) /N=\infty\)) upon scaling by the corresponding number of variables in a distribution.
+In the following, let \(|\boldsymbol y|_1 = \sum_{i=1}^d |y_i|\) denote the L1 norm of a generic vector \(\boldsymbol y =(y_1,\ldots,y_d)\), \(d \geq 1\).
+\BeginKnitrBlock{proposition}
+\protect\hypertarget{prp:prop2}{}{\label{prp:prop2} }Let \(P_{\boldsymbol \theta_N}\) denote a RBM-based data model \eqref{eq:RBM2} for \(N\geq 1\) visible variables \(\boldsymbol X \equiv \boldsymbol X_N\) derived from \(\tilde{P}_{\boldsymbol \theta_N}\) as the joint RBM distribution \eqref{eq:RBM1} of \((\boldsymbol X, \boldsymbol H)\) involving some number \(N_{\mathcal{H}} \equiv N_{N,\mathcal{H}}\geq 0\) of hidden variables \(\boldsymbol H \equiv \boldsymbol H_N\) and parameters \(\boldsymbol \theta_N \equiv (\boldsymbol \theta_N^{\mathcal{V}},\boldsymbol \theta_N^{\mathcal{H}}, \boldsymbol \theta_N^{\mathcal{VH}}) \in\mathbb{R}^{N}\times \mathbb{R}^{N_{\mathcal{H}}} \times \mathbb{R}^{N*N_{\mathcal{H}}}\).
 Then,
@@
 \BeginKnitrBlock{remark}
-\iffalse{} {Remark. } \fi{}The condition
-\(\sup_{N \geq 1} N_{\mathcal{H}}/N<\infty\) in Proposition
-\ref{prp:prop2}(iii) is often mild in practice (i.e., the number
-\(N_{\mathcal{H}}\) of hidden variables is typically not excessively
-larger than the number \(N\) of visible observations). This allows
-instability results for both marginal and joint RBM models to be more
-readily stated together, as the numbers \(N\) and \(N+N_{\mathcal{H}}\)
-of variables in these models become asymptotically equivalent.
+\iffalse{} {Remark.
+} \fi{}The condition \(\sup_{N \geq 1} N_{\mathcal{H}}/N<\infty\) in Proposition \ref{prp:prop2}(iii) is often mild in practice (i.e., the number \(N_{\mathcal{H}}\) of hidden variables is typically not excessively larger than the number \(N\) of visible observations).
+This allows instability results for both marginal and joint RBM models to be more readily stated together, as the numbers \(N\) and \(N+N_{\mathcal{H}}\) of variables in these models become asymptotically equivalent.
 \EndKnitrBlock{remark}
 
-In Proposition \ref{prp:prop2}(iii), the relationships between RBM
-models with regard to instability, and the effects of different
-parameter types, follow from the bounds on model instability measures in
-Proposition \ref{prp:prop2}(i)-(ii). Generally speaking, all instability
-in the marginal RBM model for the data \(\boldsymbol X\) can be
-attributed to an excessively large model quantity \(\elt\), which
-predominantly follows when main \(\boldsymbol \theta_N^{\mathcal{V}}\)
-and interaction \(\boldsymbol \theta_N^{\mathcal{VH}}\) parameters
-related to visible variables are too large in magnitude (e.g., upon
-accumulation in terms such as
-\(|\boldsymbol \theta_N^{\mathcal{V}}|_1\), \(\Gam\) or \(\Gamc\)). For
-example, for any bounded sequence
-\(|\boldsymbol \theta_N^{\mathcal{H}}|/N\) of hidden parameters, if main
-visible parameters \(\boldsymbol \theta_N^{\mathcal{V}}\) are too
-extreme (\(|\boldsymbol \theta_N^{\mathcal{V}}|_1/N\to \infty\)), this
-aspect will guarantee instability in the visible model
-(\(\elt/N\to \infty\)). In fact, the instability measure
-\(\elt \equiv \max_{ \boldsymbol x} \max_{ \boldsymbol h } f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)-\min_{ \boldsymbol x } \max_{ \boldsymbol h }f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)\)
-for marginal/visible model represents a clearly smaller portion of the
-instability measure
-\(\REP_{(\boldsymbol X, \boldsymbol H)}(\boldsymbol \theta_N)\equiv \max_{ \boldsymbol x} \max_{ \boldsymbol h } f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)-\min_{ \boldsymbol x } \min_{ \boldsymbol h } f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)\)
-in the joint RBM model, implying that an unstable marginal model (i.e.,
-due to \(\boldsymbol \theta_N^{\mathcal{V}}\),
-\(\boldsymbol \theta_N^{\mathcal{VH}}\)) must always translate to an
-unstable joint model and that further potential causes of instability
-exist for the joint model, often due to the size
-\(|\boldsymbol \theta_N^{\mathcal{H}}|_1\). For example, while the joint
-RBM model for \((\boldsymbol X,\boldsymbol H)\) must always be unstable
-due to a diverging combination of visible and/or interaction parameters
-(\(|\boldsymbol \theta_N^{\mathcal{V}}|_1/N\to \infty\) or
-\(\Gam/N\to \infty\)) (Proposition \ref{prp:prop2}(iii.2)), instability
-for the joint model can also result when the main hidden parameters
-\(\boldsymbol \theta_N^{\mathcal{H}}\) become too large relative to
-sample size (\(|\boldsymbol \theta_N^{\mathcal{H}}|_1/N\to \infty\) in
-Proposition \ref{prp:prop2}(iii.5)). However, under Proposition
-\ref{prp:prop2}, the main hidden parameters
-\(\boldsymbol \theta_N^{\mathcal{H}}\) do not necessarily entail a
-source of instability for the marginal visible model. To explain this
-distinction, consider a joint model where all parameters related to
-visibles are zero,
-\(\boldsymbol \theta_N^{\mathcal{V}}= \boldsymbol \theta_N^{\mathcal{VH}}=\boldsymbol 0\),
-but the hidden-related parameters diverge in sum
-\(|\boldsymbol \theta_N^{\mathcal{H}}|_1/N\to \infty\). Here, the
-explosive behavior among parameters
-\(\boldsymbol \theta_N^{\mathcal{H}}\) induces instability in the joint
-model for \((\boldsymbol X, \boldsymbol H)\) but the marginal model for
-\(\boldsymbol X\), however, has a perfectly stable (and in fact uniform)
-distribution in this case. When the hidden parameters are bounded
-relative to the sample size
-(\(\sup_{N\geq 1} |\boldsymbol \theta_N^{\mathcal{H}}|_1/N<\infty\)),
-then all instability in both the joint and marginal RBM models can be
-directly linked to excessively large visible
-\(\boldsymbol \theta_N^{\mathcal{V}}\) and/or interaction parameters
-\(\boldsymbol \theta_N^{\mathcal{VH}}\) so that features of
-stability/instability must be the same across both models (Proposition
-\ref{prp:prop2}(iii.6)). Hence, to prevent instability in the joint
-model, the combined magnitudes of all parameters \(\theta_N\) must be
-controlled (cf.~Proposition \ref{prp:prop2}(iii.7)), while a stable
-visible data model technically results in constraining only the sizes of
-visible-related parameters \(\boldsymbol \theta_N^{\mathcal{V}}\),
-\(\boldsymbol \theta_N^{\mathcal{VH}}\). Nevertheless, because the joint
-model is often employed in practice for purposes of simulation and
-simulation-based inference, it is still reasonable to consider parameter
-choices for ensuring a stable joint model (and, consequently, a stable
-visible model as well). Further evidence of this is seen in the
-following numerical example.
+In Proposition \ref{prp:prop2}(iii), the relationships between RBM models with regard to instability, and the effects of different parameter types, follow from the bounds on model instability measures in Proposition \ref{prp:prop2}(i)-(ii).
+Generally speaking, all instability in the marginal RBM model for the data \(\boldsymbol X\) can be attributed to an excessively large model quantity \(\elt\), which predominantly follows when main \(\boldsymbol \theta_N^{\mathcal{V}}\) and interaction \(\boldsymbol \theta_N^{\mathcal{VH}}\) parameters related to visible variables are too large in magnitude (e.g., upon accumulation in terms such as \(|\boldsymbol \theta_N^{\mathcal{V}}|_1\), \(\Gam\) or \(\Gamc\)).
+For example, for any bounded sequence \(|\boldsymbol \theta_N^{\mathcal{H}}|/N\) of hidden parameters, if main visible parameters \(\boldsymbol \theta_N^{\mathcal{V}}\) are too extreme (\(|\boldsymbol \theta_N^{\mathcal{V}}|_1/N\to \infty\)), this aspect will guarantee instability in the visible model (\(\elt/N\to \infty\)).
+In fact, the instability measure \(\elt \equiv \max_{ \boldsymbol x} \max_{ \boldsymbol h } f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)-\min_{ \boldsymbol x } \max_{ \boldsymbol h }f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)\) for marginal/visible model represents a clearly smaller portion of the instability measure \(\REP_{(\boldsymbol X, \boldsymbol H)}(\boldsymbol \theta_N)\equiv \max_{ \boldsymbol x} \max_{ \boldsymbol h } f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)-\min_{ \boldsymbol x } \min_{ \boldsymbol h } f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)\) in the joint RBM model, implying that an unstable marginal model (i.e., due to \(\boldsymbol \theta_N^{\mathcal{V}}\), \(\boldsymbol \theta_N^{\mathcal{VH}}\)) must always translate to an unstable joint model and that further potential causes of instability exist for the joint model, often due to the size \(|\boldsymbol \theta_N^{\mathcal{H}}|_1\).
+For example, while the joint RBM model for \((\boldsymbol X,\boldsymbol H)\) must always be unstable due to a diverging combination of visible and/or interaction parameters (\(|\boldsymbol \theta_N^{\mathcal{V}}|_1/N\to \infty\) or \(\Gam/N\to \infty\)) (Proposition \ref{prp:prop2}(iii.2)), instability for the joint model can also result when the main hidden parameters \(\boldsymbol \theta_N^{\mathcal{H}}\) become too large relative to sample size (\(|\boldsymbol \theta_N^{\mathcal{H}}|_1/N\to \infty\) in Proposition \ref{prp:prop2}(iii.5)).
+However, under Proposition \ref{prp:prop2}, the main hidden parameters \(\boldsymbol \theta_N^{\mathcal{H}}\) do not necessarily entail a source of instability for the marginal visible model.
+To explain this distinction, consider a joint model where all parameters related to visibles are zero, \(\boldsymbol \theta_N^{\mathcal{V}}= \boldsymbol \theta_N^{\mathcal{VH}}=\boldsymbol 0\), but the hidden-related parameters diverge in sum \(|\boldsymbol \theta_N^{\mathcal{H}}|_1/N\to \infty\).
+Here, the explosive behavior among parameters \(\boldsymbol \theta_N^{\mathcal{H}}\) induces instability in the joint model for \((\boldsymbol X, \boldsymbol H)\) but the marginal model for \(\boldsymbol X\), however, has a perfectly stable (and in fact uniform) distribution in this case.
+When the hidden parameters are bounded relative to the sample size (\(\sup_{N\geq 1} |\boldsymbol \theta_N^{\mathcal{H}}|_1/N<\infty\)), then all instability in both the joint and marginal RBM models can be directly linked to excessively large visible \(\boldsymbol \theta_N^{\mathcal{V}}\) and/or interaction parameters \(\boldsymbol \theta_N^{\mathcal{VH}}\) so that features of stability/instability must be the same across both models (Proposition \ref{prp:prop2}(iii.6)).
+Hence, to prevent instability in the joint model, the combined magnitudes of all parameters \(\theta_N\) must be controlled (cf.~Proposition \ref{prp:prop2}(iii.7)), while a stable visible data model technically results in constraining only the sizes of visible-related parameters \(\boldsymbol \theta_N^{\mathcal{V}}\), \(\boldsymbol \theta_N^{\mathcal{VH}}\).
+Nevertheless, because the joint model is often employed in practice for purposes of simulation and simulation-based inference, it is still reasonable to consider parameter choices for ensuring a stable joint model (and, consequently, a stable visible model as well).
+Further evidence of this is seen in the following numerical example.
 
@@
 \includegraphics{note_files/figure-latex/rbm-plots-1} \includegraphics{note_files/figure-latex/rbm-plots-2} \caption{The sample mean value of $\text{ELPR}(\boldsymbol \theta)/N_{\mathcal{V}}$ (left) and $\DN(\boldsymbol \theta)$ at each grid point for each combination of magnitude of $\boldsymbol \theta$.
-As the magnitude of $\boldsymbol \theta$ grows, so does the value of these metrics, indicating typical instability in the model.}\label{fig:rbm-plots}
+As the magnitude of $\boldsymbol \theta$ grows, so does the value of these metrics, indicating \ak{instability} in the model.}\label{fig:rbm-plots}
 \end{figure}
-In our numerical experiment, we allow the two types of terms (main
-effects terms corresponding to visible and hidden parameters
-\(\boldsymbol \theta_{main} = (\boldsymbol \theta_N^{\mathcal{V}}, \boldsymbol \theta_N^{\mathcal{H}})\)
-and interaction parameters \(\boldsymbol \theta_N^{\mathcal{VH}}\)) to
-have varying average magnitudes,
-\(||\boldsymbol \theta_{main} || /(N_{\mathcal{H}}+N_{\mathcal{V}})\)
-and
-\(||\boldsymbol \theta_{interaction} || /(N_{\mathcal{H}}*N_{\mathcal{V}})\)
-for a RBM with \(N_\mathcal{V} = 9\) visibles and \(N_\mathcal{H} = 5\)
-hiddens. These average magnitudes vary on a grid between \(0.001\) and
-\(3\) with \(20\) breaks, yielding \(400\) grid points. At each point in
-the grid, \(100\) vectors (\(\boldsymbol \theta_{main}\)) are sampled
-uniformly on a sphere with radius corresponding to the first coordinate
-in the grid and \(100\) vectors (\(\boldsymbol \theta_{interaction}\))
-are sampled uniformly on a sphere with radius corresponding to the
-second coordinate in the grid via sums of squared and scaled iid
-Normal\((0, 1)\) variables. These vectors are then paired to create
-\(100\) values of \(\boldsymbol \theta_N\) with magnitudes at each point
-in the grid. The values \(\REP(\boldsymbol \theta_N)/N_{\mathcal{V}}\)
-and \(\DN(\boldsymbol \theta_N)\) are then calculated for each
-\(\boldsymbol \theta_N\) and then summarized for each point in the grid
-using the sample mean. The results of this numerical study are shown in
-Figure \ref{fig:rbm-plots}. From these two plots, it is clear that for
-larger magnitudes of the parameter vectors, there is evidence of
-S-instability in that the log-ratio of extremal probabilities scaled by
-\(N_{\mathcal{V}}\) and the the biggest log-probability ratio for a
-one-component change in data outcomes are both increasing away from
-\(\boldsymbol \theta_N = \boldsymbol 0\), further supporting
-\ref{prp:prop2}(iii.2 and iii.5).
-
-In more complicated graphical models involving further or deeper hidden
-layers, the same issues and causes of instability similarly exist, but
-are compounded by a greater number of model parameters. S-unstable joint
-models will similarly follow if the combined sizes of all parameters are
-too great relative to the total number of variables, while instability
-in the data model for visible variables will depend only on the main or
-interaction parameters directly related to visibles and how their
-accumulated magnitude compares to the observation sample size \(N\).
+In our numerical experiment, we allow the two types of terms (main effects terms corresponding to visible and hidden parameters \(\boldsymbol \theta_{main} = (\boldsymbol \theta_N^{\mathcal{V}}, \boldsymbol \theta_N^{\mathcal{H}})\) and interaction parameters \(\boldsymbol \theta_N^{\mathcal{VH}}\)) to have varying average magnitudes, \(||\boldsymbol \theta_{main} || /(N_{\mathcal{H}}+N_{\mathcal{V}})\) and \(||\boldsymbol \theta_{interaction} || /(N_{\mathcal{H}}*N_{\mathcal{V}})\) for a RBM with \(N_\mathcal{V} = 9\) visibles and \(N_\mathcal{H} = 5\) hiddens.
+These average magnitudes vary on a grid between \(0.001\) and \(3\) with \(20\) breaks, yielding \(400\) grid points.
+At each point in the grid, \(100\) vectors (\(\boldsymbol \theta_{main}\)) are sampled uniformly on a sphere with radius corresponding to the first coordinate in the grid and \(100\) vectors (\(\boldsymbol \theta_{interaction}\)) are sampled uniformly on a sphere with radius corresponding to the second coordinate in the grid via sums of squared and scaled iid Normal\((0, 1)\) variables.
+These vectors are then paired to create \(100\) values of \(\boldsymbol \theta_N\) with magnitudes at each point in the grid.
+The values \(\REP(\boldsymbol \theta_N)/N_{\mathcal{V}}\) and \(\DN(\boldsymbol \theta_N)\) are then calculated for each \(\boldsymbol \theta_N\) and then summarized for each point in the grid using the sample mean.
+The results of this numerical study are shown in Figure \ref{fig:rbm-plots}.
+From these two plots, it is clear that for larger magnitudes of the parameter vectors, there is evidence of S-instability in that the log-ratio of extremal probabilities scaled by \(N_{\mathcal{V}}\) and the the biggest log-probability ratio for a one-component change in data outcomes are both increasing away from \(\boldsymbol \theta_N = \boldsymbol 0\), further supporting \ref{prp:prop2}(iii.2 and iii.5).
+
+In more complicated graphical models involving further or deeper hidden layers, the same issues and causes of instability similarly exist, but are compounded by a greater number of model parameters \ak{and depend on the configuration of the particular models in consideration.
+Consider the following two models -- a RBM with $N$ visible nodes and one hidden layer with $N_H$ nodes and a deep RBM with $N$ visible nodes and $2$ layers, with $N_{H,1}$ and $N_{H,2}$ hidden nodes respectively such that $N_{H,1} + N_{H,2} = N_H$.
+The length of the parameter vector $\boldsymbol \theta_N$ in the single layer RBM is then equal to $N + N_H + N*N_H$, whereas the length of the parameter vector in the two layer RBM is $N + N_{H,1} + N_{H,2} + N*N_{H,1} + N_{H,1}*N_{H,2} = N + N_H + N*N_H + N_{H,2}(N_{H,1} - N)$.
+If $N_{H,1} > N$, then the deep model will have more parameters, if $N_{H,1} < N$, then the deep model will have less parameters, and if $N_{H,1} = N$, then the deep model will the same number of parameters as the single layer RBM.} S-unstable joint models will similarly follow if the combined sizes of all parameters are too great relative to the total number of variables, while instability in the data model for visible variables will depend only on the main or interaction parameters directly related to visibles \ak{(i.e.
+all the parameters for the single layer RBM and only the parameters for the visible layer, layer $H^{(1)}$, and the interactions between them for the two layer RBM)} and how their accumulated magnitude compares to the observation sample size \(N\).
+\ak{This indicates that for the name number of hidden variables $N_H$, a deeper, rather than a wider, structure can be beneficial when considering instability in the data model for visible variables.
+However, as the size of a model considered grows, this point becomes moot.}
 
@@
 
-Due to their induced sensitivity in probability structure, S-instability
-in FOES models may often translate to numerical complications, and in
-fact obstructions, in both simulation and statistical inference based on
-likelihoods. We describe these aspects in Sections
-\ref{mcmc}-\ref{bayes} with regard to data simulation, maximum
-likelihood estimation and Bayes inference, respectively.
+Due to \ak{induced sensitivity in the probability structure of FOES models, S-instability} may often translate to numerical complications, and in fact obstructions, in both simulation and statistical inference based on likelihoods.
+We describe these aspects in Sections \ref{mcmc}-\ref{bayes} with regard to data simulation, maximum likelihood estimation and Bayes inference, respectively.
 
@@
 
-Suppose one aims to apply MCMC to simulate data
-\(\boldsymbol X=(X_1,\ldots,X_n)\) from a FOES model
-\(P_{\boldsymbol \theta_N}(\boldsymbol x)\),
-\(\boldsymbol x \in\mathcal{X}^N\) whereby a chain is constructed with
-\(P_{\boldsymbol \theta_N}\) as the stationary distribution. For an
-unstable FOES model, one-component changes in outcomes may produce
-radically different probabilities, which may then impose numerical
-barriers to MCMC. For example, consider implementation of a Gibbs
-sampler from the full conditional distributions
-\(P_{\boldsymbol \theta_N} (X_i = x| \boldsymbol x_{-i})\),
-\(x\in\mathcal{X}\), of each variable \(X_i\) based on values
-\(\boldsymbol x_{-i}\in \mathcal{X}^{N-1}\) for the remaining variables,
-say \(\boldsymbol X_{-i}\), in \(\boldsymbol X\). If a single change in
-\(X_i\) from one value \(x_1\) to another \(x_2\) may produce two
-outcomes \(\boldsymbol x^{(1)}\) and \(\boldsymbol x^{(2)}\) for
-\(\boldsymbol X\) with vastly different probabilities under the joint
-distribution \(P_{\boldsymbol \theta_N}\), then the Gibbs sampler can
-have extreme log-ratios in its transition probabilities, \[
+Suppose one aims to apply MCMC to simulate data \(\boldsymbol X=(X_1,\ldots,X_n)\) from a FOES model \(P_{\boldsymbol \theta_N}(\boldsymbol x)\), \(\boldsymbol x \in\mathcal{X}^N\) whereby a chain is constructed with \(P_{\boldsymbol \theta_N}\) as the stationary distribution.
+For an unstable FOES model, one-component changes in outcomes may produce
+radically different probabilities, which may then impose numerical barriers to MCMC.
+For example, consider implementation of a Gibbs sampler from the full conditional distributions \(P_{\boldsymbol \theta_N} (X_i = x| \boldsymbol x_{-i})\), \(x\in\mathcal{X}\), of each variable \(X_i\) based on values \(\boldsymbol x_{-i}\in \mathcal{X}^{N-1}\) for the remaining variables, say \(\boldsymbol X_{-i}\), in \(\boldsymbol X\).
+If a single change in \(X_i\) from one value \(x_1\) to another \(x_2\) may produce two outcomes \(\boldsymbol x^{(1)}\) and \(\boldsymbol x^{(2)}\) for \(\boldsymbol X\) with vastly different probabilities under the joint distribution \(P_{\boldsymbol \theta_N}\), then the Gibbs sampler can have extreme log-ratios in its transition probabilities,
+\[
 \log \left| \frac{P_{\boldsymbol \theta_N} (X_i =x_1 | \boldsymbol x^{(1)}_{-i})}{P_{\boldsymbol \theta_N} (X_i = x_2| \boldsymbol x^{(2)}_{-i}) } \right|=\log \left| \frac{P_{\boldsymbol \theta_N} ( \boldsymbol x^{(1)} )}{P_{\boldsymbol \theta_N} (  \boldsymbol x^{(2)} ) } \right|,
 \]\\
-as conditional probabilities are proportional to joint probabilities
-that, with unstable models, can have unbounded log-probability ratios in
-one-component changes (Theorem \ref{thm:instab-elpr}). This can hinder
-the ability of a chain to effectively explore the sample space of the
-observations \(\boldsymbol X\), as the chain may mix poorly by moving
-rapidly to, and slowly away from, sections of the sample space. In this
-case, for example, the Markov chain may become entrapped within a mode
-of the probability function, with rare chance of escaping to adequately
-mimic the occupation frequencies in the overall sample space. If modes
-of the unstable model are not unique, then important outcomes may be
-missed without multiple chains or impractically enormous numbers of MCMC
-samples. This mixing problem is due to the unstable stationary
-distribution (unbounded ratios of probabilities under the joint model),
-rather than in any particulars of the MCMC algorithm, and similar
-complications can also arise for Metropolis-Hastings algorithms for
-MCMC. Hence, while an unstable FOES model \(P_{\boldsymbol \theta_N}\)
-may be valid and technically open to simulation by MCMC, the aspect of
-instability can render applications of MCMC as numerically infeasible
-for simulation purposes. This result is in line with conclusions of
-\citet{handcock2003assessing} and \citet{schweinberger2011instability}
-for other exponential models.
+as conditional probabilities are proportional to joint probabilities that, with unstable models, can have unbounded log-probability ratios in one-component changes (Theorem
+\ref{thm:instab-elpr}).
+This can hinder the ability of a chain to effectively explore the sample space of the observations \(\boldsymbol X\), as the chain may mix poorly by moving rapidly to, and slowly away from, sections of the sample space.
+In this case, for example, the Markov chain may become entrapped within a mode of the probability function, with rare chance of escaping to adequately mimic the occupation frequencies in the overall sample space.
+If modes of the unstable model are not unique, then important outcomes may be missed without multiple chains or impractically enormous numbers of MCMC samples.
+This mixing problem is due to the unstable stationary distribution (unbounded ratios of probabilities under the joint model), rather than in any particulars of the MCMC algorithm, and similar complications can also arise for Metropolis-Hastings algorithms for MCMC.
+Hence, while an unstable FOES model \(P_{\boldsymbol \theta_N}\) may be valid and technically open to simulation by MCMC, the aspect of instability can render applications of MCMC as numerically infeasible for simulation purposes.
+This result is in line with conclusions of \citet{handcock2003assessing} and \citet{schweinberger2011instability} for other exponential models.
 
 \hypertarget{implications-for-maximum-likelihood-inference}{%
-\subsection{Implications for Maximum Likelihood
-Inference}\label{implications-for-maximum-likelihood-inference}}
-
-Volatility in the probability structure of an unstable model can also
-hamper efforts to maximize likelihood functions in statistical
-inference. When a FOES model is unstable along a parameter sequence
-\(\boldsymbol \theta_N\), the same model can further be unstable along
-parameters \(-\boldsymbol \theta_N\) in an opposite direction from the
-origin (model condition PSR and Corollary \ref{cor:sign}). This can
-translate into potential sensitivity of the likelihood function around
-zero, and lead to numerical complications in maximizing the objective
-function. We next provide a discussion of this issue in a way that
-builds upon and extends related findings by
-\citet{schweinberger2011instability}, who largely focused on the case of
-one-parameter exponential models.
-
-With many probability models, the modes and anti-modes in the
-probability structure under one parameter \(\boldsymbol \theta_N\) are
-reversed in role when the parameter sign changes
-\(-\boldsymbol \theta_N\). Because unstable models tend to degeneracy,
-the opposite signed parameters further push unstable models to assign
-nearly all probability to extremely opposite data configurations, given
-by modes/anti-modes. This is made concrete in Theorem
-\ref{thm:something}, relating the degeneracy from unstable models to the
-expected behavior of log-likelihood functions.
+\subsection{Implications for Maximum Likelihood Inference}\label{implications-for-maximum-likelihood-inference}}
+
+Volatility in the probability structure of an unstable model can also hamper efforts to maximize likelihood functions in statistical inference.
+When a FOES model is unstable along a parameter sequence \(\boldsymbol \theta_N\), the same model can further be unstable along parameters \(-\boldsymbol \theta_N\) in an opposite direction from the
+origin (model condition PSR and Corollary \ref{cor:sign}). This can translate into
+potential sensitivity of the likelihood function around zero, and lead to numerical complications in maximizing the objective function.
+We next provide a discussion of this issue in a way that builds upon and extends related findings by \citet{schweinberger2011instability}, who largely focused on the case of one-parameter exponential models.
+
+With many probability models, the modes and anti-modes in the probability structure under one parameter \(\boldsymbol \theta_N\) are reversed in role when the parameter sign changes \(-\boldsymbol \theta_N\).
+Because unstable models tend to degeneracy, the opposite signed parameters further push unstable models to assign nearly all probability to extremely opposite data configurations, given by modes/anti-modes.
+This is made concrete in Theorem \ref{thm:something}, relating the degeneracy from unstable models to the expected behavior of log-likelihood functions.
 
 \BeginKnitrBlock{theorem}
-\protect\hypertarget{thm:something}{}{\label{thm:something} }Let
-\(P_{\boldsymbol \theta_N}\), \(N \geq 1\), denote an S-unstable FOES
-model sequence, which additionally satisfies model condition PSR. Let
-\(\boldsymbol{x}_{\max, \boldsymbol \theta_N},\boldsymbol{x}_{\min, \boldsymbol \theta_N}\in\mathcal{X}^N\)
-denote, respectively, a mode and anti-mode of the model
-\(P_{\boldsymbol \theta_N}(\boldsymbol x)\),
-\(\boldsymbol x\in\mathcal{X}^N\), for \(N\) observations
-\(\boldsymbol X = (X_1,\ldots,X_N)\), whereby
-\(P_{\boldsymbol \theta_N }(\boldsymbol{x}_{\max, \boldsymbol \theta_N}) = \max_{\boldsymbol y\in\mathcal{X}^N} P_{\boldsymbol \theta_N }(\boldsymbol y)\)
-and
-\(P_{\boldsymbol \theta_N }(\boldsymbol{x}_{\min, \boldsymbol \theta_N}) = \min_{\boldsymbol y\in\mathcal{X}^N} P_{\boldsymbol \theta_N }(\boldsymbol y)\).
-
-Then, letting \(\stackrel{p, \mathrm{E}}{\longrightarrow}\) denote
-convergence in probability and expectation, as \(N\to \infty\), \[
+\protect\hypertarget{thm:something}{}{\label{thm:something} }Let \(P_{\boldsymbol \theta_N}\), \(N \geq 1\), denote an S-unstable FOES model sequence, which additionally satisfies model condition PSR.
+Let \(\boldsymbol{x}_{\max, \boldsymbol \theta_N},\boldsymbol{x}_{\min, \boldsymbol \theta_N}\in\mathcal{X}^N\) denote, respectively, a mode and anti-mode of the model \(P_{\boldsymbol \theta_N}(\boldsymbol x)\), \(\boldsymbol x\in\mathcal{X}^N\), for \(N\) observations \(\boldsymbol X = (X_1,\ldots,X_N)\), whereby \(P_{\boldsymbol \theta_N }(\boldsymbol{x}_{\max, \boldsymbol \theta_N}) = \max_{\boldsymbol y\in\mathcal{X}^N} P_{\boldsymbol \theta_N }(\boldsymbol y)\) and \(P_{\boldsymbol \theta_N }(\boldsymbol{x}_{\min, \boldsymbol \theta_N}) = \min_{\boldsymbol y\in\mathcal{X}^N} P_{\boldsymbol \theta_N }(\boldsymbol y)\).
+
+Then, letting \(\stackrel{p, \mathrm{E}}{\longrightarrow}\) denote convergence in probability and expectation, as \(N\to \infty\),
+\[
 \frac{1}{\REP(\boldsymbol \theta_N)} \log\left[ \frac{P_{\boldsymbol \theta_N }(\boldsymbol X) }{ \displaystyle{\min_{\boldsymbol y\in\mathcal{X}^N}} P_{\boldsymbol \theta_N }(\boldsymbol y)}\right] = \frac{ \log P_{\boldsymbol \theta_N } ( \boldsymbol X)  - \log P_{\boldsymbol \theta_N }(\boldsymbol{x}_{\min, \boldsymbol \theta_N}) }
@@
 \; 1
-\] under \(\boldsymbol\theta_N\) while \[
+\]
+under \(\boldsymbol\theta_N\) while
+\[
 \frac{1}{\REP(-\boldsymbol \theta_N)} \log\left[ \frac{P_{-\boldsymbol \theta_N }(\boldsymbol X) }{  \displaystyle{\min_{\boldsymbol y\in\mathcal{X}^N}} P_{-\boldsymbol \theta_N }(\boldsymbol y)} \right]
@@
 \; 1,
-\] under \(-\boldsymbol\theta_N\), where \[
+\]
+under \(-\boldsymbol\theta_N\), where
+\[
 \REP( \boldsymbol \theta_N) \equiv \log \frac{P_{\boldsymbol \theta_N } (\boldsymbol{x}_{\max, \boldsymbol \theta_N})}{P_{\boldsymbol \theta_N }(\boldsymbol{x}_{\min, \boldsymbol \theta_N})} = \log  \frac{P_{-\boldsymbol \theta_N } (\boldsymbol{x}_{\min, \boldsymbol \theta_N}) }{ P_{-\boldsymbol \theta_N }(\boldsymbol{x}_{\max, \boldsymbol \theta_N})}=   \REP( -\boldsymbol \theta_N), \quad N \geq 1.
@@
 
-Theorem \ref{thm:something} entails log-likelihood functions based on
-unstable models are both inversely related and degenerate at opposited
-signed parameters \(\boldsymbol \theta_N\) or \(-\boldsymbol \theta_N\),
-so that likelihoods are highest at different extremes in data
-configuration (e.g., \(\boldsymbol{x}_{\max, \boldsymbol \theta_N}\)
-under \(\boldsymbol \theta\)-probabilities or
-\(\boldsymbol{x}_{\min, \boldsymbol \theta_N}\) under
-\(-\boldsymbol \theta\)-probabilities). If the observed outcome
-\(\boldsymbol x\) for data \(\boldsymbol X\) is not a mode/anti-mode,
-then probabilities for the outcome may be small under both parameters
-\(\boldsymbol \theta_N\) and \(-\boldsymbol \theta_N\), in which case
-associated optimization steps may then shift around zero and struggle to
-converge.
-
-In many model formulations, the zero parameter
-\(\boldsymbol \theta_N=\boldsymbol 0\) is a ``safe'' position among
-parameters, representing a guaranteed stable model (having uniform
-probabilities among outcomes), which can also tether a broad parameter
-search attempted among unstable models. \citet{handcock2003assessing}
-describes similar results for degenerate exponential models, and Theorem
-\ref{thm:something} also supports an important finding of
-\citet[Corollary 1]{schweinberger2011instability} for one-parameter
-linear exponential models \eqref{eq:expo}. In the latter case, the
-likelihood score function at \(\boldsymbol \theta_N\) is the expected
-value
-\(\mu(\boldsymbol \theta_N)\equiv \mathrm{E}_{\boldsymbol \theta_N} g(\boldsymbol X)\)
-of the sufficient statistic \(g(\cdot)\), and optimization involves
-solving \(\mu(\cdot)=g(\boldsymbol x)\) for an observed outcome
-\(\boldsymbol x\). For unstable models in this exponential class,
-\citet[Corollary 1]{schweinberger2011instability} shows that \[
+Theorem \ref{thm:something} \ak{implies} log-likelihood functions based on unstable models are both inversely related and degenerate at opposited signed parameters \(\boldsymbol \theta_N\) or \(-\boldsymbol \theta_N\), so that likelihoods are highest at different extremes in data configuration (e.g., \(\boldsymbol{x}_{\max, \boldsymbol \theta_N}\) under \(\boldsymbol \theta\)-probabilities or \(\boldsymbol{x}_{\min, \boldsymbol \theta_N}\) under \(-\boldsymbol \theta\)-probabilities).
+If the observed outcome \(\boldsymbol x\) for data \(\boldsymbol X\) is not a mode/anti-mode, then probabilities for the outcome may be small under both parameters \(\boldsymbol \theta_N\) and \(-\boldsymbol \theta_N\), in which case associated optimization steps may then shift around zero and struggle to converge.
+
+In many model formulations, the zero parameter \(\boldsymbol \theta_N=\boldsymbol 0\) is a ``safe'' position among parameters, representing a guaranteed stable model (having uniform probabilities among outcomes), which can also tether a broad parameter search attempted among unstable models.
+\citet{handcock2003assessing} describes similar results for degenerate exponential models, and Theorem \ref{thm:something} also supports an important finding of \citet[Corollary 1]{schweinberger2011instability} for one-parameter linear exponential models \eqref{eq:expo}.
+In the latter case, the likelihood score function at \(\boldsymbol \theta_N\) is the expected value \(\mu(\boldsymbol \theta_N)\equiv \mathrm{E}_{\boldsymbol \theta_N} g(\boldsymbol X)\) of the sufficient statistic \(g(\cdot)\), and optimization involves solving \(\mu(\cdot)=g(\boldsymbol x)\) for an observed outcome \(\boldsymbol x\).
+For unstable models in this exponential class, \citet[Corollary 1]{schweinberger2011instability} shows
+that
+\[
 \lim_{N\to \infty}\frac{\mu(\boldsymbol \theta_N) -L_n}{U_n-L_N}= \left\{ \begin{array}{lcl}
@@
 0 && \text{for } \boldsymbol \theta_N<0, \end{array}\right.
-\] where again \(U_N\) and \(L_N\) denote the maximum and minimum values
-of the statistic \(g(\boldsymbol x)\),
-\(\boldsymbol x\in\mathcal{X}^N\). As described by
-\citet{schweinberger2011instability}, the implication for maximum
-likelihood estimation is that, unless an observed outcome
-\(\boldsymbol x\) falls at an extreme \(U_N, L_N\) (i.e.,
-modes/anti-modes), optimization steps in the parameter space can iterate
-in relatively small increments around zero and fail to converge. For
-unstable one-parameter exponential models, the maximum likelihood
-results of \citet{schweinberger2011instability} turn out to be a special
-case of Theorem \ref{thm:something} and the LREP expansion \eqref{eq:UL}
-in this setting; namely, for an unstable model with
-\(\boldsymbol \theta_N>0\), \[
+\]
+where again \(U_N\) and \(L_N\) denote the maximum and minimum values of the statistic \(g(\boldsymbol x)\), \(\boldsymbol x\in\mathcal{X}^N\).
+As described by \citet{schweinberger2011instability}, the implication for maximum likelihood estimation is that, unless an observed outcome \(\boldsymbol x\) falls at an extreme \(U_N, L_N\) (i.e., modes/anti-modes), optimization steps in the parameter space can iterate in relatively small increments around zero and fail to converge.
+For unstable one-parameter exponential models, the maximum likelihood results of \citet{schweinberger2011instability} turn out to be a special case of Theorem \ref{thm:something} and the LREP expansion \eqref{eq:UL} in this setting; namely, for an unstable model with \(\boldsymbol \theta_N>0\),
+\[
 \frac{1}{\REP(\boldsymbol \theta_N)} \log\left[ \frac{P_{\boldsymbol \theta_N }(\boldsymbol X) }{\displaystyle{\min_{\boldsymbol y\in\mathcal{X}^N}} P_{\boldsymbol \theta_N }(\boldsymbol y)}\right] = \frac{\boldsymbol g(\boldsymbol X) - L_N}{U_N-L_N}\;\;\stackrel{p, \mathrm{E}}{\longrightarrow}\;\; 1
-\] holds as \(N\to \infty\) by Theorem \ref{thm:something}, while under
-\(-\boldsymbol \theta_N<0\) \[
+\]
+holds as \(N\to \infty\) by Theorem \ref{thm:something}, while under \(-\boldsymbol \theta_N<0\)
+\[
 \frac{1}{\REP(-\boldsymbol \theta_N)} \log\left[ \frac{P_{-\boldsymbol \theta_N }(\boldsymbol X) }{\displaystyle{\min_{\boldsymbol y\in\mathcal{X}^N}} P_{-\boldsymbol \theta_N }(\boldsymbol y)}\right] = \frac{U_N -\boldsymbol g(\boldsymbol X)}{U_N-L_N} = 1 - \frac{\boldsymbol g(\boldsymbol X) - L_N}{U_N-L_N} \;\;\stackrel{p, \mathrm{E}}{\longrightarrow}\;\; 1.
-\] Again, when all probability in unstable models may be pushed to
-opposite extremes in the sample space, due to a combination of
-degeneracy and parameter sign, numerical complications in likelihood
-maximization may occur.
+\]
+Again, when all probability in unstable models may be pushed to opposite extremes in the sample space, due to a combination of degeneracy and parameter sign, numerical complications in likelihood maximization may occur.
 
@@
 
-The potential numerical difficulties with maximum likelihood with
-unstable models, as described in the previous section, can naturally
-carry over to Bayes inference. Considering that the degeneracy issues
-related to unstable models can cause likelihoods can be flat (e.g., near
-zero) for many parameters under a given data outcome and that sign
-changes in parameters can shift tremendous probability to extreme and
-opposite outcomes in the sample space (e.g., Corollary \ref{cor:sign},
-Theorem \ref{thm:something}), then numerical complications may arise
-with Bayes inference in sampling a posterior parameter space based on
-MCMC. The potential challenges in chain mixing are similar to those
-presented in Section \ref{mcmc}, though in chain movements through the
-parameter space as opposed to the sample space in data generation. That
-is, in the Bayes setting for sampling a posterior distribution for
-\(\boldsymbol \theta_N\), a chain may unstable to effectively explore
-the parameter space due partly to extreme and potentially unbounded
-probability ratios from parameter sign changes, which represents a
-parameter space analog to how one-component changes in the sample space
-may impact data simulation with unstable models. For example, if
-\(\pi(\cdot)\) denotes a prior density for \(\boldsymbol \theta_N\) and
-\(q(\cdot | \cdot)\) denotes a proposal distribution for use in a
-Metropolis-Hastings (MH) sampler, then MH acceptance probability becomes
+The potential numerical difficulties with maximum likelihood with unstable models, as described in the previous section, can naturally carry over to Bayes inference.
+Considering that the degeneracy issues related to unstable models can cause likelihoods can be flat (e.g., near zero) for many parameters under a given data outcome and that sign changes in parameters can shift tremendous probability to extreme and opposite outcomes in the sample space (e.g., Corollary \ref{cor:sign}, Theorem \ref{thm:something}), then numerical complications may arise with Bayes inference in sampling a posterior parameter space based on MCMC.
+The potential challenges in chain mixing are similar to those presented in Section \ref{mcmc}, though in chain movements through the parameter space as opposed to the sample space in data generation.
+That is, in the Bayes setting for sampling a posterior distribution for \(\boldsymbol \theta_N\), a chain may \ak{be unable} to effectively explore the parameter space due partly to extreme and potentially unbounded probability ratios from parameter sign changes, which represents a parameter space analog to how one-component changes in the sample space may impact data simulation with unstable models.
+For example, if \(\pi(\cdot)\) denotes a prior density for \(\boldsymbol \theta_N\) and \(q(\cdot | \cdot)\) denotes a proposal distribution for use in a Metropolis-Hastings (MH) sampler, then MH acceptance probability becomes
 \[
@@
 \frac{P_{\boldsymbol \theta_N^{(1)}} ( \boldsymbol x ) \pi(\boldsymbol \theta_N^{(1)}) }{P_{\boldsymbol \theta^{(2)}_N} (  \boldsymbol x) \pi(\boldsymbol \theta_N^{(2)}) } \right\},
-\] which indicates how parameter sensitivity in the likelihood
-\(P_{\boldsymbol \theta_N} ( \boldsymbol x)\) may complicate sampling of
-the posterior
-\(P_{\boldsymbol \theta_N} ( \boldsymbol x) \pi(\boldsymbol \theta_N)\)
-(i.e., moving from \(\boldsymbol \theta^{(1)}\) to
-\(\boldsymbol \theta^{(2)}\) in the parameter space). Furthermore, the
-potential for model instability and the size of the parameter space can
-also become greater with the introduction of latent variables to
-existing data variables, as involved in some model formulations
-described in Sections \ref{rbm}-\ref{deep-learning}. As latent variables
-are often sampled with parameters in a Bayes MCMC approach, this aspect
-may further compound numerical problems in chain mixing.
+\]
+which indicates how parameter sensitivity in the likelihood \(P_{\boldsymbol \theta_N} ( \boldsymbol x)\) may complicate sampling of the posterior \(P_{\boldsymbol \theta_N} ( \boldsymbol x) \pi(\boldsymbol \theta_N)\) (i.e., moving from \(\boldsymbol \theta^{(1)}\) to \(\boldsymbol \theta^{(2)}\) in the parameter space).
+Furthermore, the potential for model instability and the size of the parameter space can also become greater with the introduction of latent variables to existing data variables, as involved in some model formulations described in Sections \ref{rbm}-\ref{deep-learning}.
+As latent variables are often sampled with parameters in a Bayes MCMC approach, this aspect may further compound numerical problems in chain mixing.
 
@@
 
-For a large class of models that covers a broad range of applications
-(including ``deep learning''), we have developed a formal definition of
-instability in model probability structure and elucidated multiple
-consequences of instability. We have shown for FOES models that
-instability manifests through small changes in data leading to
-potentially large changes in probability as well as the potential to
-place all probability on certain modal subsections of the sample space,
-which potentially could be small. Such instability is often due to a
-complex interaction between the model statistics used (i.e., how
-numerous and large these may become) and the number and magnitudes of
-parameters in the model formulation. For many FOES models, the
-possibility exists, at least in principle, to constraint parameters in a
-way balances their potential contributions against those of model
-statistics in order to prevent probability instabilities. The FOES model
-class is quite broad and, in developing such models for large data sets,
-some caution should be used in parameter specification to control
-effects of model instability.
+For a large class of models that covers a broad range of applications (including ``deep learning''), we have developed a formal definition of instability in model probability structure and elucidated multiple consequences of instability.
+We have shown for FOES models that instability manifests through small changes in data leading to potentially large changes in probability as well as the potential to place all probability on certain modal subsections of the sample space, which potentially could be small.
+Such instability is often due to a complex interaction between the model statistics used (i.e., how numerous and large these may become) and the number and magnitudes of
+parameters in the model formulation.
+For many FOES models, the possibility exists, at least in principle, to \ak{constrain} parameters in a way \ak{that} balances their potential contributions against those of model statistics in order to prevent probability instabilities.
+The FOES model class is quite broad and, in developing such models for large data sets, some caution should be used in parameter specification to control effects of model instability.
 
@@
 
-\textbf{Proof of Proposition \ref{thm:instab-elpr}.} For part (i), we
-prove the contrapositive, supposing that
-\(\DN(\boldsymbol \theta_N) \le C\) holds for some \(C > 0\) and show
-\(\REP(\boldsymbol \theta_N) \leq NC\). Let
-\(\boldsymbol x_{min} \equiv \argmin\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)\)
-and
-\(\boldsymbol x_{max} \equiv \argmax\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)\).
-Note there exists a sequence
-\(\boldsymbol x_{min} \equiv \boldsymbol x_0, \boldsymbol x_1, \dots, \boldsymbol x_k \equiv \boldsymbol x_{max}\)
-in \(\mathcal{X}^N\) of component-wise switches to move from
-\(\boldsymbol x_{min}\) to \(\boldsymbol x_{max}\) in the sample space
-(i.e. \(\boldsymbol x_i, \boldsymbol x_{i + 1} \in \mathcal{X}^N\)
-differ in exactly \(1\) component, \(i = 0, \dots, k\)) for some integer
-\(k \in \{0, 1, \dots, N\}\). Under the FOES model, recall
-\(P_{\boldsymbol \theta_N}(\boldsymbol x) > 0\) holds so that
-\(\log P_{\boldsymbol \theta_N}(\boldsymbol x)\) is well-defined for
-each outcome \(\boldsymbol x \in \mathcal{X}^N\). Then, if \(k > 0\), it
-follows that \begin{align*}
+\textbf{Proof of Proposition \ref{thm:instab-elpr}.} For part (i), we prove the contrapositive, supposing that \(\DN(\boldsymbol \theta_N) \le C\) holds for some \(C > 0\) and show \(\REP(\boldsymbol \theta_N) \leq NC\).
+Let \(\boldsymbol x_{min} \equiv \argmin\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)\)
+and \(\boldsymbol x_{max} \equiv \argmax\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)\).
+Note there exists a sequence \(\boldsymbol x_{min} \equiv \boldsymbol x_0, \boldsymbol x_1, \dots, \boldsymbol x_k \equiv \boldsymbol x_{max}\) in \(\mathcal{X}^N\) of component-wise switches to move from \(\boldsymbol x_{min}\) to \(\boldsymbol x_{max}\) in the sample space (i.e.~\(\boldsymbol x_i, \boldsymbol x_{i + 1} \in \mathcal{X}^N\) differ in exactly \(1\) component, \(i = 0, \dots, k\)) for some integer \(k \in \{0, 1, \dots, N\}\).
+Under the FOES model, recall
+\(P_{\boldsymbol \theta_N}(\boldsymbol x) > 0\) holds so that \(\log P_{\boldsymbol \theta_N}(\boldsymbol x)\) is well-defined for each outcome \(\boldsymbol x \in \mathcal{X}^N\).
+Then, if \(k > 0\), it follows that
+\begin{align*}
 \REP(\boldsymbol \theta_N) = \log\left[\frac{P_{\boldsymbol \theta_N}(\boldsymbol x_{max})}{P_{\boldsymbol \theta_N}(\boldsymbol x_{min})}\right] &= \left|\sum\limits_{i = 1}^k\log\left(\frac{P_{\boldsymbol \theta_N}(\boldsymbol x_i)}{P_{\boldsymbol \theta_N}(\boldsymbol x_{i-1})}\right)\right| \\
 &\le \sum\limits_{i = 1}^k\left|\log\left(\frac{P_{\boldsymbol \theta_N}(\boldsymbol x_i)}{P_{\boldsymbol \theta}(\boldsymbol x_{i-1})}\right)\right| \le k \Delta_N(\boldsymbol \theta_N) \le NC,
-\end{align*} using \(k \le N\) and
-\(\Delta(\boldsymbol \theta_N) \le C\). If \(k = 0\), then
-\(\boldsymbol x_{max} = \boldsymbol x_{min}\) and the same bound above
-holds. This establishes part (i). To show part (ii), note the definition
-of S-instability (i.e.,
-\(\lim_{N\to \infty}\REP(\boldsymbol \theta_N)/N= \infty\)) combined
-with part (i) implies that
-\(\lim_{N\to \infty}\DN(\boldsymbol \theta_N)=\infty\). \hfill \(\Box\)
-
-\textbf{Proof of Proposition \ref{thm:degenFOES}.} As
-\(|\mathcal{X}|<\infty\) holds in the FOES model, we may suppose
-\(|\mathcal{X}|>1\); otherwise, \(\mathcal{X}^N\) has one outcome and
-the model is trivially degenerate for all \(N \geq 1\). Fix
-\(0 < \epsilon < 1\) and write
-\(\boldsymbol x_{min} \equiv \argmin\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)\)
-and
-\(\boldsymbol x_{max} \equiv \argmax\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)\).
-Then, \(\boldsymbol x_{max} \in M_{\epsilon, \boldsymbol \theta_N}\), so
-\(P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N}) \ge P_{\boldsymbol \theta_N}(\boldsymbol x_{max}) > 0\).
-If
-\(\boldsymbol x \in \mathcal{X}^N \setminus M_{\epsilon, \boldsymbol \theta_N}\),
-then by definition
-\(P_{\boldsymbol \theta_N}(\boldsymbol x) \le [P_{\boldsymbol \theta_N}(\boldsymbol x_{max})]^{1-\epsilon}[P_{\boldsymbol \theta_N}(\boldsymbol x_{min})]^{\epsilon}\)
-holds so that \[
+\end{align*}
+using \(k \le N\) and \(\Delta(\boldsymbol \theta_N) \le C\).
+If \(k = 0\), then \(\boldsymbol x_{max} = \boldsymbol x_{min}\) and the same bound above holds.
+This establishes part (i).
+To show part (ii), note the definition of S-instability (i.e., \(\lim_{N\to \infty}\REP(\boldsymbol \theta_N)/N= \infty\)) combined with part (i) implies that \(\lim_{N\to \infty}\DN(\boldsymbol \theta_N)=\infty\).
+\hfill \(\Box\)
+
+\textbf{Proof of Proposition \ref{thm:degenFOES}.} As \(|\mathcal{X}|<\infty\) holds in the FOES model, we may suppose
+\(|\mathcal{X}|>1\); otherwise, \(\mathcal{X}^N\) has one outcome and the model is trivially degenerate for all \(N \geq 1\).
+Fix \(0 < \epsilon < 1\) and write \(\boldsymbol x_{min} \equiv \argmin\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)\) and \(\boldsymbol x_{max} \equiv \argmax\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)\).
+Then, \(\boldsymbol x_{max} \in M_{\epsilon, \boldsymbol \theta_N}\), so \(P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N}) \ge P_{\boldsymbol \theta_N}(\boldsymbol x_{max}) > 0\).
+If \(\boldsymbol x \in \mathcal{X}^N \setminus M_{\epsilon, \boldsymbol \theta_N}\), then by definition
+\(P_{\boldsymbol \theta_N}(\boldsymbol x) \le [P_{\boldsymbol \theta_N}(\boldsymbol x_{max})]^{1-\epsilon}[P_{\boldsymbol \theta_N}(\boldsymbol x_{min})]^{\epsilon}\) holds so that
+\[
 1-P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N})
@@
   \le (|\mathcal{X}|^N)[P_{\boldsymbol \theta_N}(\boldsymbol x_{max})]^{1-\epsilon}[P_{\boldsymbol \theta_N}(\boldsymbol x_{min})]^{\epsilon}.
-\] From the lower bound on
-\(P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N})\) and the
-upper bound on
-\(1-P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N})\), it
-follows that \begin{align*}
+\]
+From the lower bound on \(P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N})\) and the
+upper bound on \(1-P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N})\), it follows that
+\begin{align*}
 \frac{1}{N}\log\left[\frac{P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N})}{1-P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N})}\right] & \ge \frac{1}{N} \log\left[\frac{P_{\boldsymbol\theta_N}(\boldsymbol x_{max})}{(|\mathcal{X}|^N)[P_{\boldsymbol \theta_N}(\boldsymbol x_{max})]^{1-\epsilon}[P_{\boldsymbol \theta_N}(\boldsymbol x_{min})]^{\epsilon}}\right] \\
 &= \frac{\epsilon}{N} \log\left[\frac{P_{\boldsymbol \theta_N}(\boldsymbol x_{max})}{P_{\boldsymbol \theta_N}(\boldsymbol x_{min})}\right] - \log |\mathcal{X}| \rightarrow \infty
-\end{align*} as \(N \rightarrow \infty\) by the definition of an
-S-unstable FOES model \eqref{eq:Sun}. Consequently,
-\(P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N}) \rightarrow 1\)
-as \(N \rightarrow \infty\) as claimed. \hfill \(\Box\)
-
-\textbf{Proof of Corollary \ref{cor:sign}.} The model condition PSR
-implies that \begin{equation}
+\end{align*}
+as \(N \rightarrow \infty\) by the definition of an S-unstable FOES model \eqref{eq:Sun}. Consequently,
+\(P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N}) \rightarrow 1\) as \(N \rightarrow \infty\) as claimed. \hfill \(\Box\)
+
+\textbf{Proof of Corollary \ref{cor:sign}.} The model condition PSR implies that
+\begin{equation}
 \label{eq:R}
 \frac{\max_{\boldsymbol y \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol y)  }{\min_{\boldsymbol y \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol y) } = \frac{\max_{\boldsymbol y \in \mathcal{X}^N}P_{-\boldsymbol \theta_N}(\boldsymbol y) }{\min_{\boldsymbol y \in \mathcal{X}^N}P_{-\boldsymbol \theta_N}(\boldsymbol y) }
-\end{equation} so that the log-ratio
-\(\REP(\boldsymbol \theta_N)=\REP(-\boldsymbol \theta_N)\) is the same
-for both \(\boldsymbol \theta_N\) and \(-\boldsymbol \theta_N\) in
-\eqref{eq:elpr}. Now part (i) of Corollary \ref{cor:sign} follows from
-\(\REP(\boldsymbol \theta_N)/N=\REP(-\boldsymbol \theta_N)/N\to \infty\)
-as \(N\to \infty\) in \eqref{eq:Sun}. To show part (ii), fix
-\(0 < \epsilon < 1\) and consider a \(\epsilon\)-mode set
-\(\mathcal{M}_{\epsilon, \boldsymbol \theta_N}\) under
-\(\boldsymbol \theta_N\) from \eqref{eq:mode}. If
-\(\boldsymbol x \in \mathcal{M}_{\epsilon, \boldsymbol \theta_N}^c \equiv \mathcal{X}^N \setminus \mathcal{M}_{\epsilon, \boldsymbol \theta_N}\),
-then, by definition, \[
+\end{equation}
+so that the log-ratio \(\REP(\boldsymbol \theta_N)=\REP(-\boldsymbol \theta_N)\) is the same for both \(\boldsymbol \theta_N\) and \(-\boldsymbol \theta_N\) in \eqref{eq:elpr}.
+Now part (i) of Corollary \ref{cor:sign} follows from \(\REP(\boldsymbol \theta_N)/N=\REP(-\boldsymbol \theta_N)/N\to \infty\) as \(N\to \infty\) in \eqref{eq:Sun}.
+To show part (ii), fix \(0 < \epsilon < 1\) and consider a \(\epsilon\)-mode set \(\mathcal{M}_{\epsilon, \boldsymbol \theta_N}\) under \(\boldsymbol \theta_N\) from \eqref{eq:mode}.
+If \(\boldsymbol x \in \mathcal{M}_{\epsilon, \boldsymbol \theta_N}^c \equiv \mathcal{X}^N \setminus \mathcal{M}_{\epsilon, \boldsymbol \theta_N}\), then, by definition,
+\[
 \frac{P_{\boldsymbol \theta_N}(\boldsymbol x)}{\min_{\boldsymbol y \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol y)} \leq \left[\frac{\max_{\boldsymbol y \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol y)  }{\min_{\boldsymbol y \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol y) } \right]^{1-\epsilon}
-\] holds, which is equivalent to \[
+\]
+holds, which is equivalent to
+\[
 \frac{\max_{\boldsymbol y \in \mathcal{X}^N}P_{-\boldsymbol \theta_N}(\boldsymbol y)}{P_{-\boldsymbol \theta_N}(\boldsymbol x)} \leq
 \left[\frac{\max_{\boldsymbol y \in \mathcal{X}^N}P_{-\boldsymbol \theta_N}(\boldsymbol y)  }{\min_{\boldsymbol y \in \mathcal{X}^N}P_{-\boldsymbol \theta_N}(\boldsymbol y) } \right]^{1-\epsilon}
-\] by model condition PSR and \eqref{eq:R}. The latter is in turn
-equivalent to \begin{equation}
+\]
+by model condition PSR and \eqref{eq:R}. The latter is in turn equivalent to
+\begin{equation}
 \label{eq:R2}
 \log P_{-\boldsymbol \theta_N}(\boldsymbol x) \geq \epsilon \max\limits_{\boldsymbol y \in \mathcal{X}^N} \log  P_{-\boldsymbol \theta_N}(\boldsymbol y) + (1-\epsilon)\min\limits_{\boldsymbol y \in \mathcal{X}^N} \log P_{-\boldsymbol \theta_N}(\boldsymbol y),
-\end{equation} so that
-\(\boldsymbol x \in \mathcal{M}_{\epsilon, \boldsymbol \theta_N}^c\) if
-and only if \eqref{eq:R2} holds. Next consider the \((1-\epsilon)\)-mode
-set \(\mathcal{M}_{1-\epsilon, -\boldsymbol \theta_N}\) under
-\(-\boldsymbol \theta_N\) from \eqref{eq:mode}. If
-\(\boldsymbol x \in\mathcal{M}_{1-\epsilon, -\boldsymbol \theta_N}\),
-then by definition \(\boldsymbol x\) fulfills \eqref{eq:R2} and so
-\(\boldsymbol x \in \mathcal{M}_{\epsilon, \boldsymbol \theta_N}^c\),
-showing that
-\(\mathcal{M}_{1-\epsilon, -\boldsymbol \theta_N} \subset \mathcal{M}_{\epsilon, \boldsymbol \theta_N}^c\).
-By this and the fact that that Theorem \ref{thm:degenFOES} and Corollary
-\ref{cor:sign}(i) entail that
-\(P_{-\boldsymbol \theta_N}(\boldsymbol X_N \in \mathcal{M}_{1-\epsilon, -\boldsymbol \theta_N})\to 1\)
-as \(N\to \infty\) (i.e., \(P_{-\boldsymbol \theta_N}\) is S-unstable),
-we have \[
+\end{equation}
+so that \(\boldsymbol x \in \mathcal{M}_{\epsilon, \boldsymbol \theta_N}^c\) if and only if \eqref{eq:R2} holds.
+Next consider the \((1-\epsilon)\)-mode set \(\mathcal{M}_{1-\epsilon, -\boldsymbol \theta_N}\) under \(-\boldsymbol \theta_N\) from \eqref{eq:mode}.
+If \(\boldsymbol x \in\mathcal{M}_{1-\epsilon, -\boldsymbol \theta_N}\), then by definition \(\boldsymbol x\) fulfills \eqref{eq:R2} and so \(\boldsymbol x \in \mathcal{M}_{\epsilon, \boldsymbol \theta_N}^c\), showing that \(\mathcal{M}_{1-\epsilon, -\boldsymbol \theta_N} \subset \mathcal{M}_{\epsilon, \boldsymbol \theta_N}^c\).
+By this and the fact that that Theorem \ref{thm:degenFOES} and Corollary \ref{cor:sign}(i) entail that \(P_{-\boldsymbol \theta_N}(\boldsymbol X_N \in \mathcal{M}_{1-\epsilon, -\boldsymbol \theta_N})\to 1\) as \(N\to \infty\) (i.e., \(P_{-\boldsymbol \theta_N}\) is S-unstable), we have
+\[
 1  = \lim_{N\to \infty} P_{-\boldsymbol \theta_N}(\boldsymbol X_N \in \mathcal{M}_{1-\epsilon, -\boldsymbol \theta_N}) \leq \lim_{N\to \infty} P_{-\boldsymbol \theta_N}(\boldsymbol X_N \in \mathcal{M}_{\epsilon, \boldsymbol \theta_N}^c ) \leq 1,
-\] proving Corollary \ref{cor:sign}(ii) \hfill \(\Box\)
-
-\textbf{Proof of Proposition \ref{prp:prop1}.} For any two outcomes
-\(\boldsymbol x_1, \boldsymbol x_2\in\mathcal{X}^N\), the log-ratio of
-probabilities from the linear exponential model \eqref{eq:expo} with \(k\)
-parameters/statistics satisfies \[
+\]
+proving Corollary \ref{cor:sign}(ii) \hfill \(\Box\)
+
+\textbf{Proof of Proposition \ref{prp:prop1}.} For any two outcomes \(\boldsymbol x_1, \boldsymbol x_2\in\mathcal{X}^N\), the log-ratio of probabilities from the linear exponential model \eqref{eq:expo} with \(k\) parameters/statistics satisfies
+\[
 \left|\log \left[ \frac{P_{\boldsymbol \theta_N}(\boldsymbol x_1)}{P_{\boldsymbol \theta_N}(\boldsymbol x_2)}  \right] \right| =
 \left|  \sum_{i=1}^k \theta_{i,N} [g_{i,k}(\boldsymbol x_1) - g_{i,k}(\boldsymbol x_2) ] \right|  \leq  \sum_{i=1}^k | \theta_{i,N}| (U_{i,N}-L_{i,N});
-\] consequently,
-\(\REP(\boldsymbol \theta_N ) \leq \sum_{i=1}^k | \theta_{i,N}| (U_{i,N}-L_{i,N})\)
-holds in \eqref{eq:elpr} and model stability in Proposition
-\ref{prp:prop1} follows from \eqref{eq:Sun}. \hfill \(\Box\)
-
-\textbf{Proof of Proposition \ref{prp:prop2}.} Writing
-\(\boldsymbol x=(x_1,\ldots,x_N)\) and
-\(\boldsymbol h = (h_1,\ldots,h_{N_{\mathcal{H}}})\) with all components
-\(x_i,h_j\in\{\pm 1\}\), probabilities in the joint RBM model
-\eqref{eq:RBM1} can be written as
-\(\tilde{P}_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h) = c(\boldsymbol \theta_N)\exp[ f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)]\)
-in terms of the function
-\(f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)\) from
-\eqref{eq:f} and the normalizing constant
-\(c(\boldsymbol \theta_N)= \exp [-\psi(\boldsymbol \theta_N)]\) from
-\eqref{eq:RBM1}. Let \(\boldsymbol x_M, \boldsymbol x_m\in\{\pm 1\}^N\) be
-such that
-\(P_{\boldsymbol \theta_N} (\boldsymbol x_M) = \max_{\boldsymbol x}P_{\boldsymbol \theta_N} (\boldsymbol x)\)
-and
-\(P_{\boldsymbol \theta_N} (\boldsymbol x_m) = \min_{\boldsymbol x}P_{\boldsymbol \theta_N} (\boldsymbol x)\)
-under the marginal RBM model
-\(P_{\boldsymbol \theta_N} (\boldsymbol x) = c(\boldsymbol \theta_N)\sum_{\boldsymbol h \in\{\pm 1\}^{\mathcal{N}_H}} \tilde{P}_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)= c(\boldsymbol \theta_N)\sum_{\boldsymbol h \in\{\pm 1\}^{\mathcal{N}_H}} \exp[ f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)]\)
-from \eqref{eq:RBM2}. Also, \(\boldsymbol x_0,x_1\in\{\pm 1\}^N\) be such
-that
-\(\max_{\boldsymbol h}f_{\boldsymbol \theta_N} (\boldsymbol x_0, \boldsymbol h)=\max_{\boldsymbol x}\max_{\boldsymbol h}f_{\boldsymbol \theta_N} (\boldsymbol x , \boldsymbol h)\)
-and
-\(\max_{\boldsymbol h}f_{\boldsymbol \theta_N} (\boldsymbol x_1, \boldsymbol h)=\min_{\boldsymbol x}\max_{\boldsymbol h}f_{\boldsymbol \theta_N} (\boldsymbol x , \boldsymbol h)\).
-Then, Proposition \ref{prp:prop2}(i) follows from
-\(\REP_{(\boldsymbol X)}(\boldsymbol \theta_N) = \log[P_{\boldsymbol \theta_N} (\boldsymbol x_M) /P_{\boldsymbol \theta_N} (\boldsymbol x_m) ]\)
-and the lower/upper bounds on
-\(P_{\boldsymbol \theta_N} (\boldsymbol x_M)\) and
-\(P_{\boldsymbol \theta_N} (\boldsymbol x_m)\) as \[
+\]
+consequently, \(\REP(\boldsymbol \theta_N ) \leq \sum_{i=1}^k | \theta_{i,N}| (U_{i,N}-L_{i,N})\) holds in \eqref{eq:elpr} and model stability in Proposition \ref{prp:prop1} follows from \eqref{eq:Sun}.
+\hfill \(\Box\)
+
+\textbf{Proof of Proposition \ref{prp:prop2}.}
+Writing \(\boldsymbol x=(x_1,\ldots,x_N)\) and \(\boldsymbol h = (h_1,\ldots,h_{N_{\mathcal{H}}})\) with all components \(x_i,h_j\in\{\pm 1\}\), probabilities in the joint RBM model \eqref{eq:RBM1} can be written as \(\tilde{P}_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h) = c(\boldsymbol \theta_N)\exp[ f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)]\) in terms of the function \(f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)\) from \eqref{eq:f} and the normalizing constant \(c(\boldsymbol \theta_N)= \exp [-\psi(\boldsymbol \theta_N)]\) from \eqref{eq:RBM1}.
+Let \(\boldsymbol x_M, \boldsymbol x_m\in\{\pm 1\}^N\) be such that \(P_{\boldsymbol \theta_N} (\boldsymbol x_M) = \max_{\boldsymbol x}P_{\boldsymbol \theta_N} (\boldsymbol x)\) and \(P_{\boldsymbol \theta_N} (\boldsymbol x_m) = \min_{\boldsymbol x}P_{\boldsymbol \theta_N} (\boldsymbol x)\) under the marginal RBM model \(P_{\boldsymbol \theta_N} (\boldsymbol x) = c(\boldsymbol \theta_N)\sum_{\boldsymbol h \in\{\pm 1\}^{\mathcal{N}_H}} \tilde{P}_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)= c(\boldsymbol \theta_N)\sum_{\boldsymbol h \in\{\pm 1\}^{\mathcal{N}_H}} \exp[ f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)]\) from \eqref{eq:RBM2}.
+Also, \(\boldsymbol x_0,x_1\in\{\pm 1\}^N\) be such that \(\max_{\boldsymbol h}f_{\boldsymbol \theta_N} (\boldsymbol x_0, \boldsymbol h)=\max_{\boldsymbol x}\max_{\boldsymbol h}f_{\boldsymbol \theta_N} (\boldsymbol x , \boldsymbol h)\) and \(\max_{\boldsymbol h}f_{\boldsymbol \theta_N} (\boldsymbol x_1, \boldsymbol h)=\min_{\boldsymbol x}\max_{\boldsymbol h}f_{\boldsymbol \theta_N} (\boldsymbol x , \boldsymbol h)\).
+Then, Proposition \ref{prp:prop2}(i) follows from \(\REP_{(\boldsymbol X)}(\boldsymbol \theta_N) = \log[P_{\boldsymbol \theta_N} (\boldsymbol x_M) /P_{\boldsymbol \theta_N} (\boldsymbol x_m) ]\) and the lower/upper bounds on \(P_{\boldsymbol \theta_N} (\boldsymbol x_M)\) and \(P_{\boldsymbol \theta_N} (\boldsymbol x_m)\) as
+\[
 c(\boldsymbol \theta_N) \exp[\max_{\boldsymbol h}f_{\boldsymbol \theta_N} (\boldsymbol x_0 , \boldsymbol h)]
 \leq P_{\boldsymbol \theta_N} (\boldsymbol x_0) \leq  P_{\boldsymbol \theta_N} (\boldsymbol x_M) \leq  2^{N_{\mathcal{H}}} c(\boldsymbol \theta_N) \exp[\max_{\boldsymbol x}\max_{\boldsymbol h}f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)]
-\] and \begin{align*}
+\]
+and
+\begin{align*}
 c(\boldsymbol \theta_N) \exp[\min_{\boldsymbol x}\max_{\boldsymbol h}f_{\boldsymbol \theta_N} (\boldsymbol x , \boldsymbol h)] &\leq c(\boldsymbol \theta_N) \exp[ \max_{\boldsymbol h}f_{\boldsymbol \theta_N} (\boldsymbol x_m , \boldsymbol h)] \\
@@
 \end{align*}
-To prove Proposition \ref{prp:prop2}, we next expand the function
-\(f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)\) from
-\eqref{eq:f} as \[
+To prove Proposition \ref{prp:prop2}, we next expand the function \(f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)\) from \eqref{eq:f} as
+\[
 f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)= \sum_{j=1}^{N_\mathcal{H}} h_j \theta_{j,N}^{\mathcal{H}}  +   \sum_{i=1}^N \left( \theta_{i,N}^{\mathcal{V}} + \sum_{j=1}^{N_\mathcal{H}}  h_j  \theta_{ij,N}^{\mathcal{VH}}\right)x_i=\sum_{i=1}^N x_i  \theta_{i,N}^{\mathcal{V}} +  \sum_{j=1}^{N_\mathcal{H}} \left( \theta_{j,N}^{\mathcal{H}}  + \sum_{i=1}^N  x_i  \theta_{ij,N}^{\mathcal{VH}}\right)h_j.
-\] By this and the fact that \(x_i,h_j\in\{\pm 1\}\), we then have
+\]
+By this and the fact that \(x_i,h_j\in\{\pm 1\}\), we then have
 \begin{align}
@@
 a_{\boldsymbol \theta_N, \mathcal{H}} (\boldsymbol h) \equiv \sum_{i=1}^N \left| \theta_{i,N}^{\mathcal{V}} + \sum_{j=1}^{N_\mathcal{H}}  h_j  \theta_{ij,N}^{\mathcal{VH}}\right|, &  b_{\boldsymbol \theta_N, \mathcal{V}} (\boldsymbol x) \equiv \sum_{j=1}^{N_\mathcal{H}} \left| \theta_{j,N}^{\mathcal{H}}  + \sum_{i=1}^N  x_i  \theta_{ij,N}^{\mathcal{VH}}\right|,
-\end{align} where
-\(\boldsymbol h^T \boldsymbol \theta_N^{\mathcal{H}}=\sum_{j=1}^{N_\mathcal{H}} h_j \theta_{j,N}^{\mathcal{H}}\),
-\(\boldsymbol x^T \boldsymbol \theta_N^{\mathcal{V}}= \sum_{i=1}^{N}x_i \theta_{i,N}^{\mathcal{V}}\)
-and
-\(\Gam\equiv \max_{\boldsymbol h} a_{\boldsymbol \theta_N, \mathcal{H}} (\boldsymbol h)\).
-From this, it follows that \begin{align*}
+\end{align}
+where \(\boldsymbol h^T \boldsymbol \theta_N^{\mathcal{H}}=\sum_{j=1}^{N_\mathcal{H}} h_j \theta_{j,N}^{\mathcal{H}}\), \(\boldsymbol x^T \boldsymbol \theta_N^{\mathcal{V}}= \sum_{i=1}^{N}x_i \theta_{i,N}^{\mathcal{V}}\) and \(\Gam\equiv \max_{\boldsymbol h} a_{\boldsymbol \theta_N, \mathcal{H}} (\boldsymbol h)\).
+From this, it follows that
+\begin{align*}
 \REP_{(\boldsymbol X, \boldsymbol H)}(\boldsymbol \theta_N) &=  \max_{\boldsymbol h}\max_{\boldsymbol x}f_{\boldsymbol \theta_N} (\boldsymbol x , \boldsymbol h) -  \min_{\boldsymbol h}\min_{\boldsymbol x}f_{\boldsymbol \theta_N} (\boldsymbol x , \boldsymbol h)\\
 &= \max_{\boldsymbol h_1 }  \max_{\boldsymbol h_2 }\left[ (\boldsymbol h_1 - \boldsymbol h_2)^T \boldsymbol \theta_N^{\mathcal{H}}  +   a_{\boldsymbol \theta_N, \mathcal{H}} (\boldsymbol h_1)  + a_{\boldsymbol \theta_N, \mathcal{H}} (\boldsymbol h_2)\right],
-\end{align*} which leads to the upper bound
-\(\REP_{(\boldsymbol X, \boldsymbol H)}(\boldsymbol \theta_N) \leq 2 \Gam + 2 |\boldsymbol \theta_N^{\mathcal{H}} |_1\).
-Then, taking \(\boldsymbol h_1=\boldsymbol h_2\) (i.e., before
-maximization) gives
-\(\REP_{(\boldsymbol X, \boldsymbol H)}(\boldsymbol \theta_N) \geq 2\Gam\)
-and taking \(\boldsymbol h_1=-\boldsymbol h_2\), such that
-\(\boldsymbol h_1^T \boldsymbol \theta_N^{\mathcal{H}} = |\theta_N^{\mathcal{H}}|_1\),
-gives
-\(\REP_{(\boldsymbol X, \boldsymbol H)}(\boldsymbol \theta_N) \geq 2|\theta_N^{\mathcal{H}}|_1\);
-this yields the lower bound
-\(\REP_{(\boldsymbol X, \boldsymbol H)}(\boldsymbol \theta_N)\geq 2\max\{\Gam, |\boldsymbol \theta_N^{\mathcal{H}} |_1\}\).
-
-We next consider \(\elt\) and, by \eqref{eq:max}, write \begin{align*}
+\end{align*}
+which leads to the upper bound \(\REP_{(\boldsymbol X, \boldsymbol H)}(\boldsymbol \theta_N) \leq 2 \Gam + 2 |\boldsymbol \theta_N^{\mathcal{H}} |_1\).
+Then, taking \(\boldsymbol h_1=\boldsymbol h_2\) (i.e., before maximization) gives \(\REP_{(\boldsymbol X, \boldsymbol H)}(\boldsymbol \theta_N) \geq 2\Gam\) and taking \(\boldsymbol h_1=-\boldsymbol h_2\), such that \(\boldsymbol h_1^T \boldsymbol \theta_N^{\mathcal{H}} = |\theta_N^{\mathcal{H}}|_1\), gives \(\REP_{(\boldsymbol X, \boldsymbol H)}(\boldsymbol \theta_N) \geq 2|\theta_N^{\mathcal{H}}|_1\); this yields the lower bound \(\REP_{(\boldsymbol X, \boldsymbol H)}(\boldsymbol \theta_N)\geq 2\max\{\Gam, |\boldsymbol \theta_N^{\mathcal{H}} |_1\}\).
+
+We next consider \(\elt\) and, by \eqref{eq:max}, write
+\begin{align*}
 \elt &=  \max_{\boldsymbol h}\max_{\boldsymbol x}f_{\boldsymbol \theta_N} (\boldsymbol x , \boldsymbol h) -  \max_{\boldsymbol h}\min_{\boldsymbol x}f_{\boldsymbol \theta_N} (\boldsymbol x , \boldsymbol h)\\
 &= \max_{\boldsymbol h_1} \min_{\boldsymbol h_2 }\left[ (\boldsymbol h_1 - \boldsymbol h_2)^T \boldsymbol \theta_N^{\mathcal{H}} + a_{\boldsymbol \theta_N, \mathcal{H}} (\boldsymbol h_1)  + a_{\boldsymbol \theta_N, \mathcal{H}} (\boldsymbol h_2)\right].
-\end{align*} Taking \(\boldsymbol h_1=\boldsymbol h_2\) and maximizing
-over both \(\boldsymbol h_1,\boldsymbol h_2\) produces the upper bound
-\(\elt \leq 2\Gam\). Then, using
-\((\boldsymbol h_1 - \boldsymbol h_2)^T \boldsymbol \theta_N^{\mathcal{H}} + a_{\boldsymbol \theta_N, \mathcal{H}} (\boldsymbol h_2) \geq - 2|\boldsymbol \theta_N^{\mathcal{H}} |_1\)
-and maximizing over \(\boldsymbol h_1\) gives
-\(\elt \geq \Gam- 2|\boldsymbol \theta_N^{\mathcal{H}} |_1\), while
-setting \(\boldsymbol h_1=\boldsymbol h_2^*\) for \(\boldsymbol h_2^*\)
-such that
-\(-(\boldsymbol h_2^*) ^T \boldsymbol \theta_N^{\mathcal{H}} + a_{\boldsymbol \theta_N, \mathcal{H}} (\boldsymbol h_2^*) = \min_{\boldsymbol h_2} [-\boldsymbol h_2^T \boldsymbol \theta_N^{\mathcal{H}} + a_{\boldsymbol \theta_N, \mathcal{H}} (\boldsymbol h_2)]\)
-gives
-\(\elt \geq 2 a_{\boldsymbol \theta_N, \mathcal{H}} (\boldsymbol h_2^*) \geq \Gamc \equiv \min_{\boldsymbol h} a_{\boldsymbol \theta_N, \mathcal{H}} (\boldsymbol h)\).
-Finally, note that for any \(\boldsymbol h\), the triangle inequality
-gives \begin{align*}
+\end{align*}
+Taking \(\boldsymbol h_1=\boldsymbol h_2\) and maximizing over both \(\boldsymbol h_1,\boldsymbol h_2\) produces the upper bound \(\elt \leq 2\Gam\).
+Then, using \((\boldsymbol h_1 - \boldsymbol h_2)^T \boldsymbol \theta_N^{\mathcal{H}} + a_{\boldsymbol \theta_N, \mathcal{H}} (\boldsymbol h_2) \geq - 2|\boldsymbol \theta_N^{\mathcal{H}} |_1\) and maximizing over \(\boldsymbol h_1\) gives \(\elt \geq \Gam- 2|\boldsymbol \theta_N^{\mathcal{H}} |_1\), while setting \(\boldsymbol h_1=\boldsymbol h_2^*\) for \(\boldsymbol h_2^*\) such that \(-(\boldsymbol h_2^*) ^T \boldsymbol \theta_N^{\mathcal{H}} + a_{\boldsymbol \theta_N, \mathcal{H}} (\boldsymbol h_2^*) = \min_{\boldsymbol h_2} [-\boldsymbol h_2^T \boldsymbol \theta_N^{\mathcal{H}} + a_{\boldsymbol \theta_N, \mathcal{H}} (\boldsymbol h_2)]\) gives \(\elt \geq 2 a_{\boldsymbol \theta_N, \mathcal{H}} (\boldsymbol h_2^*) \geq \Gamc \equiv \min_{\boldsymbol h} a_{\boldsymbol \theta_N, \mathcal{H}} (\boldsymbol h)\).
+Finally, note that for any \(\boldsymbol h\), the triangle inequality gives
+\begin{align*}
 \Gam  \equiv \max_{\boldsymbol h_1}  a_{\boldsymbol \theta_N, \mathcal{H}} (\boldsymbol h_1) &\geq
@@
 & \geq \sum_{i=1}^{N }  \left| \theta_{i,N}^{\mathcal{V}} \right| \equiv |\boldsymbol \theta_{N}^{\mathcal{V}}|_1.
-\end{align*} \hfill \(\Box\)
-\textbf{Proof of Theorem \ref{thm:something}.} Let
-\(L_{\boldsymbol \theta_{N}}(\boldsymbol X) = \log[ P_{\boldsymbol \theta_{N}}(\boldsymbol X)/ \min_{\boldsymbol y \in \mathcal{X}^N} P_{\boldsymbol \theta_{N}}(\boldsymbol y) ]/\REP(\boldsymbol \theta_{N})\),
-where again \(\boldsymbol X=(X_1, \dots, X_N)\) and
-\(\REP(\boldsymbol \theta_{N})= \log[\max_{\boldsymbol y \in \mathcal{X}^N} P_{\boldsymbol \theta_{N}}(\boldsymbol y)/ \min_{\boldsymbol y \in \mathcal{X}^N} P_{\boldsymbol \theta_{N}}(\boldsymbol y) ]\).
-As \(L_{\boldsymbol \theta_{N}}(\boldsymbol X)\in[0,1]\), convergence of
-\(L_{\boldsymbol \theta_{N}}(\boldsymbol X)\) to 1 in probability under
-\(P_{\boldsymbol \theta_{N}}\) is equivalent to convergence to \(1\) in
-expectation under \(P_{\boldsymbol \theta_{N}}\) (i.e., convergence in
-expectation implies probabilistic convergence by Markov's inequality
-while probabilistic convergence implies convergence in expectation by
-uniform integrability/boundedness).
-
-For \(\epsilon \in(0,1)\), let
-\(\mathcal{M}_{\epsilon, \boldsymbol \theta_N}\) denote a modal set as
-in \eqref{eq:mode}. By Theorem \ref{thm:degenFOES},
-\(P_{\boldsymbol \theta_N}\left( \boldsymbol X\in \mathcal{M}_{\epsilon, \boldsymbol \theta_N}\right) \rightarrow 1\)
-holds as \(N \rightarrow \infty\) and, by definition of \eqref{eq:mode},
-\(\boldsymbol X\in \mathcal{M}_{\epsilon, \boldsymbol \theta_N}\)
-follows if and only if
-\(1-L_{\boldsymbol \theta_{N}}(\boldsymbol X)<\epsilon\). Hence,
-\(L_{\boldsymbol \theta_{N}}(\boldsymbol X) \stackrel{p,\mathrm{E}}{\longrightarrow} 1\)
-holds under \(\boldsymbol \theta_{N}\) in Theorem \ref{thm:something}.
-The convergence
-\(L_{-\boldsymbol \theta_{N}}(\boldsymbol X) \stackrel{p,\mathrm{E}}{\longrightarrow} 1\)
-under \(-\boldsymbol \theta_{N}\) likewise follows from Corollary
-\ref{cor:sign}. \hfill \(\Box\)
+\end{align*}
+\hfill \(\Box\)
+
+\textbf{Proof of Theorem \ref{thm:something}.} Let \(L_{\boldsymbol \theta_{N}}(\boldsymbol X) = \log[ P_{\boldsymbol \theta_{N}}(\boldsymbol X)/ \min_{\boldsymbol y \in \mathcal{X}^N} P_{\boldsymbol \theta_{N}}(\boldsymbol y) ]/\REP(\boldsymbol \theta_{N})\), where again \(\boldsymbol X=(X_1, \dots, X_N)\) and \(\REP(\boldsymbol \theta_{N})= \log[\max_{\boldsymbol y \in \mathcal{X}^N} P_{\boldsymbol \theta_{N}}(\boldsymbol y)/ \min_{\boldsymbol y \in \mathcal{X}^N} P_{\boldsymbol \theta_{N}}(\boldsymbol y) ]\).
+As \(L_{\boldsymbol \theta_{N}}(\boldsymbol X)\in[0,1]\), convergence of \(L_{\boldsymbol \theta_{N}}(\boldsymbol X)\) to 1 in probability under \(P_{\boldsymbol \theta_{N}}\) is equivalent to convergence to \(1\) in expectation under \(P_{\boldsymbol \theta_{N}}\) (i.e., convergence in expectation implies probabilistic convergence by Markov's inequality while probabilistic convergence implies convergence in expectation by uniform integrability/boundedness).
+
+For \(\epsilon \in(0,1)\), let \(\mathcal{M}_{\epsilon, \boldsymbol \theta_N}\) denote a modal set as in \eqref{eq:mode}.
+By Theorem \ref{thm:degenFOES}, \(P_{\boldsymbol \theta_N}\left( \boldsymbol X\in \mathcal{M}_{\epsilon, \boldsymbol \theta_N}\right) \rightarrow 1\) holds as \(N \rightarrow \infty\) and, by definition of \eqref{eq:mode}, \(\boldsymbol X\in \mathcal{M}_{\epsilon, \boldsymbol \theta_N}\) follows if and only if
+\(1-L_{\boldsymbol \theta_{N}}(\boldsymbol X)<\epsilon\).
+Hence, \(L_{\boldsymbol \theta_{N}}(\boldsymbol X) \stackrel{p,\mathrm{E}}{\longrightarrow} 1\) holds under \(\boldsymbol \theta_{N}\) in Theorem \ref{thm:something}.
+The convergence \(L_{-\boldsymbol \theta_{N}}(\boldsymbol X) \stackrel{p,\mathrm{E}}{\longrightarrow} 1\) under \(-\boldsymbol \theta_{N}\) likewise follows from Corollary \ref{cor:sign}.
+\hfill \(\Box\)
 

### 2019-06-11 writing/paper-note/index.Rmd
started edits

@@
   A probability model exhibits instability if small changes in a data outcome result in large, and often unanticipated, changes in probability.
-This instability is a property of the probability model, rather than the fitted parameter vector.
-For correlated data structures found in several application areas, there is increasing interest in predicting/identifying such sensitivity in model probability structure.
-We consider the problem of quantifying instability for general probability models defined on sequences of observations, where each sequence of length $N$ has a finite number of possible values.
-A sequence of probability models results, indexed by $N$, that accommodates data of expanding dimension.
+This instability is a property of the probability \ak{model, prescribed by a probability distributional form evaluated at a given configuration of parameters}.
+For correlated data structures found in several application areas, there is increasing interest in \ak{identifying} such sensitivity in model probability structure.
+We consider the problem of quantifying instability for general probability models defined on sequences of observations, where each sequence of length $N$ has a finite number of possible values \ak{that can be taken at each point}.
+A sequence of probability models results, indexed by $N$, \ak{and an associated parameter sequence,} that accommodates data of expanding dimension.
 Model instability is formally shown to occur when a certain log-probability ratio under such models grows faster than $N$.

### 2019-06-18 writing/paper-note/appendices/appendix-note.Rmd
getting ready for another submission
revision of note

--- 2017-09-27 writing/paper-note/appendices/appendix-note.Rmd
@@
 
-**Proof of Proposition \@ref(prp:instab-elpr).** We prove the contrapositive, supposing that $\Delta(\boldsymbol \theta_N) \le C$ holds for some $C > 0$ and show $\text{ELPR}(\boldsymbol \theta_N) \leq NC$.
-Let $\boldsymbol x_{min} \equiv \argmin\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)$ and $\boldsymbol x_{max} \equiv \argmax\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)$.
+**Proof of Theorem \@ref(thm:instab-elpr).** For part (i), we prove the contrapositive, supposing that $\DN(\ak{\thetaN}) \le C$ holds for some $C > 0$ and show $\REP(\ak{\thetaN}) \leq NC$.
+Let $\boldsymbol x_{min} \equiv \argmin\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\ak{\thetaN}}(\boldsymbol x)$
+and $\boldsymbol x_{max} \equiv \argmax\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\ak{\thetaN}}(\boldsymbol x)$.
 Note there exists a sequence $\boldsymbol x_{min} \equiv \boldsymbol x_0, \boldsymbol x_1, \dots, \boldsymbol x_k \equiv \boldsymbol x_{max}$ in $\mathcal{X}^N$ of component-wise switches to move from $\boldsymbol x_{min}$ to $\boldsymbol x_{max}$ in the sample space (i.e.
-$\boldsymbol x_i, \boldsymbol x_{i + 1} \in \mathcal{X}^N$ differ in exactly $1$ component for $i = 0, \dots, k$) for some integer $k \in \{0, 1, \dots, N\}$.
-Under the FSFS model, recall $P_{\boldsymbol \theta_N}(\boldsymbol x) > 0$ holds so that $\log P_{\boldsymbol \theta_N}(\boldsymbol x)$ is well-defined for each outcome $\boldsymbol x \in \mathcal{X}^N$.
-Then, if $k > 0$, it follows that
+$\boldsymbol x_i, \boldsymbol x_{i + 1} \in \mathcal{X}^N$ differ in exactly $1$ component, $i = 0, \dots, k$) for some integer $k \in \{0, 1, \dots, N\}$.
+Under the FOES model, recall
+$P_{\ak{\thetaN}}(\boldsymbol x) > 0$ holds so that $\log P_{\ak{\thetaN}}(\boldsymbol x)$ is well-defined for each outcome $\boldsymbol x \in \mathcal{X}^N$. Then, if $k > 0$, it follows that
 \begin{align*}
-\text{ELPR}(\boldsymbol \theta_N) = \log\left[\frac{P_{\boldsymbol \theta_N}(\boldsymbol x_{max})}{P_{\boldsymbol \theta_N}(\boldsymbol x_{min})}\right] &= \left|\sum\limits_{i = 1}^k\log\left(\frac{P_{\boldsymbol \theta_N}(\boldsymbol x_i)}{P_{\boldsymbol \theta_N}(\boldsymbol x_{i-1})}\right)\right| \\
-&\le \sum\limits_{i = 1}^k\left|\log\left(\frac{P_{\boldsymbol \theta_N}(\boldsymbol x_i)}{P_{\boldsymbol \theta}(\boldsymbol x_{i-1})}\right)\right| \le k \Delta_N(\boldsymbol \theta_N) \le NC,
+\REP(\ak{\thetaN}) = \log\left[\frac{P_{\ak{\thetaN}}(\boldsymbol x_{max})}{P_{ \ak{\thetaN}}(\boldsymbol x_{min})}\right] &= \left|\sum\limits_{i = 1}^k\log\left(\frac{P_{\ak{\thetaN}}(\boldsymbol x_i)}{P_{\ak{\thetaN}}(\boldsymbol x_{i-1})}\right)\right| \\
+&\le \sum\limits_{i = 1}^k\left|\log\left(\frac{P_{\ak{\thetaN}}(\boldsymbol x_i)}{P_{\boldsymbol \theta}(\boldsymbol x_{i-1})}\right)\right| \le k \Delta_N(\ak{\thetaN}) \le NC,
 \end{align*}
-using $k \le N$ and $\Delta(\boldsymbol \theta_N) \le C$. If $k = 0$, then $\boldsymbol x_{max} = \boldsymbol x_{min}$ and the same bound above holds.\hfill $\Box$
-
-**Proof of Proposition \@ref(prp:degenFSFS).** Where $|\mathcal{X}|<\infty$ holds in the FSFS model, we may suppose $|\mathcal{X}|>1$ (i.e., $\mathcal{X}^N$ has more than one outcome) because otherwise the model is trivially degenerate for all $N \geq 1$.
-Fix $0 < \epsilon < 1$.
-Then, $\boldsymbol x_{max} \in M_{\epsilon, \boldsymbol \theta_N}$, so $P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N}) \ge P_{\boldsymbol \theta_N}(\boldsymbol x_{max}) > 0$.
-If $\boldsymbol x \in \mathcal{X}^N \setminus M_{\epsilon, \boldsymbol \theta_N}$, then by definition $P_{\boldsymbol \theta_N}(\boldsymbol x) \le [P_{\boldsymbol \theta_N}(\boldsymbol x_{max})]^{1-\epsilon}[P_{\boldsymbol \theta_N}(\boldsymbol x_{min})]^{\epsilon}$ holds so that
-\begin{align*}
-1-P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N})
-& = \sum\limits_{\boldsymbol x \in \mathcal{X}^N \setminus M_{\epsilon, \boldsymbol \theta_N}}P_{\boldsymbol \theta_N}(\boldsymbol x) \\
-& \le (|\mathcal{X}|^N)[P_{\boldsymbol \theta_N}(\boldsymbol x_{max})]^{1-\epsilon}[P_{\boldsymbol \theta_N}(\boldsymbol x_{min})]^{\epsilon}.
-\end{align*}
-From the lower bound on $P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N})$ and the upper bound on $1-P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N})$, it follows that
-\begin{align*}
-\frac{1}{N}\log\left[\frac{P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N})}{1-P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N})}\right] & \ge \frac{1}{N} \log\left[\frac{P_{\boldsymbol\theta_N}(\boldsymbol x_{max})}{(|\mathcal{X}|^N)[P_{\boldsymbol \theta_N}(\boldsymbol x_{max})]^{1-\epsilon}[P_{\boldsymbol \theta_N}(\boldsymbol x_{min})]^{\epsilon}}\right] \\
-&= \frac{\epsilon}{N} \log\left[\frac{P_{\boldsymbol \theta_N}(\boldsymbol x_{max})}{P_{\boldsymbol \theta_N}(\boldsymbol x_{min})}\right] - \log |\mathcal{X}| \rightarrow \infty
-\end{align*}
-as $N \rightarrow \infty$ by the definition of an S-unstable FSFS model (cf.
-Definition \@ref(def:instabFSFS)).
-Consequently, $P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N}) \rightarrow 1$ as $N \rightarrow \infty$ as claimed.
+using $k \le N$ and $\Delta(\ak{\thetaN}) \le C$.
+If $k = 0$, then $\boldsymbol x_{max} = \boldsymbol x_{min}$ and the same bound above holds.
+This establishes part (i).
+To show part (ii), note the definition of S-instability (i.e., $\lim_{N\to \infty}\REP(\ak{\thetaN})/N= \infty$) combined with part (i) implies that $\lim_{N\to \infty}\DN(\ak{\thetaN})=\infty$.
 \hfill $\Box$
 
+**Proof of Theorem \@ref(thm:degenFOES).** As $|\mathcal{X}|<\infty$ holds in the FOES model, we may suppose
+$|\mathcal{X}|>1$; otherwise, $\mathcal{X}^N$ has one outcome and the model is  trivially degenerate for all $N \geq 1$.
+Fix $0 < \epsilon < 1$ and write $\boldsymbol x_{min} \equiv \argmin\limits_{\boldsymbol x \in \mathcal{X}^N}P_{ \ak{\thetaN}}(\boldsymbol x)$ and $\boldsymbol x_{max} \equiv \argmax\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\ak{\thetaN}}(\boldsymbol x)$.
+Then, $\boldsymbol x_{max} \in M_{\epsilon, \ak{\thetaN}}$, so $P_{\ak{\thetaN}}(M_{\epsilon, \ak{\thetaN}}) \ge P_{ \ak{\thetaN}}(\boldsymbol x_{max}) > 0$.
+If $\boldsymbol x \in \mathcal{X}^N \setminus M_{\epsilon, \ak{\thetaN}}$, then by definition
+$P_{\ak{\thetaN}}(\boldsymbol x) \le [P_{\ak{\thetaN}}(\boldsymbol x_{max})]^{1-\epsilon}[P_{\ak{\thetaN}}(\boldsymbol x_{min})]^{\epsilon}$ holds so that
+$$
+1-P_{\ak{\thetaN}}(M_{\epsilon, \ak{\thetaN}})
+ = \sum\limits_{\boldsymbol x \in \mathcal{X}^N \setminus M_{\epsilon, \ak{\thetaN}}}P_{\ak{\thetaN}}(\boldsymbol x)
+  \le (|\mathcal{X}|^N)[P_{\ak{\thetaN}}(\boldsymbol x_{max})]^{1-\epsilon}[P_{ \ak{\thetaN}}(\boldsymbol x_{min})]^{\epsilon}.
+$$
+From the lower bound on $P_{\ak{\thetaN}}(M_{\epsilon, \ak{\thetaN}})$ and the
+upper bound on $1-P_{\ak{\thetaN}}(M_{\epsilon, \ak{\thetaN}})$, it follows that
+\begin{align*}
+\frac{1}{N}\log\left[\frac{P_{\ak{\thetaN}}(M_{\epsilon, \ak{\thetaN}})}{1-P_{ \ak{\thetaN}}(M_{\epsilon, \ak{\thetaN}})}\right] & \ge \frac{1}{N} \log\left[\frac{P_{\ak{\thetaN}}(\boldsymbol x_{max})}{(|\mathcal{X}|^N)[P_{ \ak{\thetaN}}(\boldsymbol x_{max})]^{1-\epsilon}[P_{\ak{\thetaN}}(\boldsymbol x_{min})]^{\epsilon}}\right] \\
+&= \frac{\epsilon}{N} \log\left[\frac{P_{\ak{\thetaN}}(\boldsymbol x_{max})}{P_{ \ak{\thetaN}}(\boldsymbol x_{min})}\right] - \log |\mathcal{X}| \rightarrow \infty
+\end{align*}
+as $N \rightarrow \infty$ by the definition of an S-unstable FOES model \@ref(eq:Sun).
+Consequently, $P_{\ak{\thetaN}}(M_{\epsilon, \ak{\thetaN}}) \rightarrow 1$ as $N \rightarrow \infty$ as claimed.
+\hfill $\Box$
+
+**Proof of Corollary \@ref(cor:sign).** The model condition PSR implies that
+\begin{equation}
+(\#eq:R)
+\frac{\max_{\boldsymbol y \in \mathcal{X}^N}P_{\ak{\thetaN}}(\boldsymbol y)  }{\min_{\boldsymbol y \in \mathcal{X}^N}P_{\ak{\thetaN}}(\boldsymbol y) } = \frac{\max_{\boldsymbol y \in \mathcal{X}^N}P_{-\ak{\thetaN}}(\boldsymbol y) }{\min_{\boldsymbol y \in \mathcal{X}^N}P_{-\ak{\thetaN}}(\boldsymbol y) }
+\end{equation}
+so that the log-ratio $\REP(\ak{\thetaN})=\REP(-\ak{\thetaN})$ is the same for both $\ak{\thetaN}$ and $-\ak{\thetaN}$ in \@ref(eq:elpr).
+Now part (i) of Corollary \@ref(cor:sign) follows from $\REP(\ak{\thetaN})/N=\REP(-\ak{\thetaN})/N\to \infty$ as $N\to \infty$ in \@ref(eq:Sun).
+To show part (ii), fix $0 < \epsilon < 1$ and consider a $\epsilon$-mode set $\mathcal{M}_{\epsilon, \ak{\thetaN}}$ under $\ak{\thetaN}$ from \@ref(eq:mode).
+If $\boldsymbol x \in \mathcal{M}_{\epsilon, \ak{\thetaN}}^c \equiv \mathcal{X}^N \setminus \mathcal{M}_{\epsilon, \ak{\thetaN}}$, then, by definition,
+$$
+\frac{P_{\ak{\thetaN}}(\boldsymbol x)}{\min_{\boldsymbol y \in \mathcal{X}^N}P_{ \ak{\thetaN}}(\boldsymbol y)} \leq \left[\frac{\max_{\boldsymbol y \in \mathcal{X}^N}P_{\ak{\thetaN}}(\boldsymbol y)  }{\min_{\boldsymbol y \in \mathcal{X}^N}P_{\ak{\thetaN}}(\boldsymbol y) } \right]^{1-\epsilon}
+$$
+holds, which is equivalent to
+$$
+\frac{\max_{\boldsymbol y \in \mathcal{X}^N}P_{-\ak{\thetaN}}(\boldsymbol y)}{P_{-\ak{\thetaN}}(\boldsymbol x)} \leq
+\left[\frac{\max_{\boldsymbol y \in \mathcal{X}^N}P_{-\ak{\thetaN}}(\boldsymbol y)  }{\min_{\boldsymbol y \in \mathcal{X}^N}P_{-\ak{\thetaN}}(\boldsymbol y) } \right]^{1-\epsilon}
+$$
+by model condition PSR and \@ref(eq:R). The latter is in turn equivalent to
+\begin{equation}
+(\#eq:R2)
+\log P_{-\ak{\thetaN}}(\boldsymbol x) \geq \epsilon \max\limits_{\boldsymbol y \in \mathcal{X}^N} \log  P_{-\ak{\thetaN}}(\boldsymbol y) + (1-\epsilon)\min\limits_{\boldsymbol y \in \mathcal{X}^N} \log P_{-\ak{\thetaN}}(\boldsymbol y),
+\end{equation}
+so that $\boldsymbol x \in \mathcal{M}_{\epsilon, \ak{\thetaN}}^c$ if and only if \@ref(eq:R2) holds.
+Next consider the $(1-\epsilon)$-mode set $\mathcal{M}_{1-\epsilon, -\ak{\thetaN}}$ under $-\ak{\thetaN}$ from \@ref(eq:mode).
+If $\boldsymbol x \in\mathcal{M}_{1-\epsilon, -\ak{\thetaN}}$, then by definition $\boldsymbol x$ fulfills \@ref(eq:R2) and so $\boldsymbol x \in \mathcal{M}_{\epsilon, \ak{\thetaN}}^c$, showing that $\mathcal{M}_{1-\epsilon, -\ak{\thetaN}} \subset \mathcal{M}_{\epsilon, \ak{\thetaN}}^c$.
+By this and the fact that that Theorem \@ref(thm:degenFOES) and Corollary \@ref(cor:sign)(i) entail that $P_{-\ak{\thetaN}}(\boldsymbol X_N \in \mathcal{M}_{1-\epsilon, -\ak{\thetaN}})\to 1$ as $N\to \infty$ (i.e., $P_{-\ak{\thetaN}}$ is S-unstable), we have
+$$
+1  = \lim_{N\to \infty} P_{-\ak{\thetaN}}(\boldsymbol X_N \in \mathcal{M}_{1-\epsilon, -\ak{\thetaN}}) \leq \lim_{N\to \infty} P_{-\ak{\thetaN}}(\boldsymbol X_N \in \mathcal{M}_{\epsilon, \ak{\thetaN}}^c ) \leq 1,
+$$
+proving Corollary \@ref(cor:sign)(ii) \hfill $\Box$
+
+**Proof of Proposition \@ref(prp:prop1).** For any two outcomes $\boldsymbol x_1, \boldsymbol x_2\in\mathcal{X}^N$, the log-ratio of probabilities from the linear exponential model \@ref(eq:expo) with $k$ parameters/statistics satisfies
+$$
+\left|\log \left[ \frac{P_{\ak{\thetaN}}(\boldsymbol x_1)}{P_{\ak{\thetaN}}(\boldsymbol x_2)}  \right] \right| =
+\left|  \sum_{i=1}^k \theta_{i,\ak{\thetaidx}} [g_{i,k}(\boldsymbol x_1) - g_{i,k}(\boldsymbol x_2) ] \right|  \leq  \sum_{i=1}^k | \theta_{i,\ak{\thetaidx}}| (U_{i,N}-L_{i,N});
+$$
+consequently, $\REP(\ak{\thetaN} )  \leq  \sum_{i=1}^k | \theta_{i,\ak{\thetaidx}}| (U_{i,N}-L_{i,N})$ holds in \@ref(eq:elpr) and model stability in Proposition \@ref(prp:prop1) follows from \@ref(eq:Sun).
+\hfill $\Box$
+
+**Proof of Proposition \@ref(prp:prop2).**
+Writing $\boldsymbol x=(x_1,\ldots,x_N)$ and $\boldsymbol h = (h_1,\ldots,h_{N_{\mathcal{H}}})$ with all components $x_i,h_j\in\{\pm 1\}$, probabilities in the joint RBM model \@ref(eq:RBM1) can be written as $\tilde{P}_{\ak{\thetaN}} (\boldsymbol x, \boldsymbol h) = c(\ak{\thetaN})\exp[ f_{\ak{\thetaN}} (\boldsymbol x, \boldsymbol h)]$ in terms of the function $f_{\ak{\thetaN}} (\boldsymbol x, \boldsymbol h)$ from \@ref(eq:f) and the normalizing constant $c(\ak{\thetaN})= \exp [-\psi(\ak{\thetaN})]$ from \@ref(eq:RBM1).
+Let $\boldsymbol x_M, \boldsymbol x_m\in\{\pm 1\}^N$ be such that $P_{\ak{\thetaN}} (\boldsymbol x_M) = \max_{\boldsymbol x}P_{\ak{\thetaN}} (\boldsymbol x)$ and $P_{\ak{\thetaN}} (\boldsymbol x_m) = \min_{\boldsymbol x}P_{\ak{\thetaN}} (\boldsymbol x)$ under the marginal RBM model $P_{\ak{\thetaN}} (\boldsymbol x) = c(\ak{\thetaN})\sum_{\boldsymbol h \in\{\pm 1\}^{\mathcal{N}_H}} \tilde{P}_{\ak{\thetaN}} (\boldsymbol x, \boldsymbol h)= c( \ak{\thetaN})\sum_{\boldsymbol h \in\{\pm 1\}^{\mathcal{N}_H}} \exp[ f_{\ak{\thetaN}} (\boldsymbol x, \boldsymbol h)]$ from \@ref(eq:RBM2).
+Also, $\boldsymbol x_0,x_1\in\{\pm 1\}^N$ be such that $\max_{\boldsymbol h}f_{\ak{\thetaN}} (\boldsymbol x_0, \boldsymbol h)=\max_{\boldsymbol x}\max_{\boldsymbol h}f_{\ak{\thetaN}} (\boldsymbol x , \boldsymbol h)$ and $\max_{\boldsymbol h}f_{\ak{\thetaN}} (\boldsymbol x_1, \boldsymbol h)=\min_{\boldsymbol x}\max_{\boldsymbol h}f_{\ak{\thetaN}} (\boldsymbol x , \boldsymbol h)$.
+Then, Proposition \@ref(prp:prop2)(i) follows from $\REP_{(\boldsymbol X)}(\ak{\thetaN}) = \log[P_{\ak{\thetaN}} (\boldsymbol x_M) /P_{\ak{\thetaN}} (\boldsymbol x_m) ]$ and the lower/upper bounds on $P_{\ak{\thetaN}} (\boldsymbol x_M)$ and $P_{\ak{\thetaN}} (\boldsymbol x_m)$ as
+$$
+c(\ak{\thetaN}) \exp[\max_{\boldsymbol h}f_{\ak{\thetaN}} (\boldsymbol x_0 , \boldsymbol h)]
+\leq P_{\ak{\thetaN}} (\boldsymbol x_0) \leq  P_{\ak{\thetaN}} (\boldsymbol x_M) \leq  2^{N_{\mathcal{H}}} c(\ak{\thetaN}) \exp[\max_{\boldsymbol x}\max_{\boldsymbol h}f_{\ak{\thetaN}} (\boldsymbol x, \boldsymbol h)]
+$$
+and
+\begin{align*}
+c(\ak{\thetaN}) \exp[\min_{\boldsymbol x}\max_{\boldsymbol h}f_{\ak{\thetaN}} (\boldsymbol x , \boldsymbol h)] &\leq c(\ak{\thetaN}) \exp[ \max_{\boldsymbol h}f_{\ak{\thetaN}} (\boldsymbol x_m , \boldsymbol h)] \\
+&\leq P_{\ak{\thetaN}} (\boldsymbol x_m) \\
+&\leq P_{\ak{\thetaN}} (\boldsymbol x_1) \\ 
+&\leq 2^{N_{\mathcal{H}}} c(\ak{\thetaN}) \exp[ \max_{\boldsymbol h}f_{\ak{\thetaN}} (\boldsymbol x_1 , \boldsymbol h)]\\
+&=2^{N_{\mathcal{H}}} c(\ak{\thetaN}) \exp[\min_{\boldsymbol x}\max_{\boldsymbol h}f_{\ak{\thetaN}} (\boldsymbol x , \boldsymbol h)]
+\end{align*}
+
+To prove Proposition \@ref(prp:prop2), we next expand the function $f_{\ak{\thetaN}} (\boldsymbol x, \boldsymbol h)$ from \@ref(eq:f) as
+$$
+f_{\ak{\thetaN}} (\boldsymbol x, \boldsymbol h)= \sum_{j=1}^{N_\mathcal{H}} h_j \theta_{j,\ak{\thetaidx}}^{\mathcal{H}}  +   \sum_{i=1}^N \left( \theta_{i,\ak{\thetaidx}}^{\mathcal{V}} + \sum_{j=1}^{N_\mathcal{H}}  h_j  \theta_{ij,\ak{\thetaidx}}^{\mathcal{VH}}\right)x_i=\sum_{i=1}^N x_i  \theta_{i,\ak{\thetaidx}}^{\mathcal{V}} +  \sum_{j=1}^{N_\mathcal{H}} \left( \theta_{j,\ak{\thetaidx}}^{\mathcal{H}}  + \sum_{i=1}^N  x_i  \theta_{ij,\ak{\thetaidx}}^{\mathcal{VH}}\right)h_j.
+$$
+By this and the fact that $x_i,h_j\in\{\pm 1\}$, we then have
+\begin{align}
+\nonumber \max_{\boldsymbol x}  f_{\ak{\thetaN}} (\boldsymbol x, \boldsymbol h) =  \sum_{j=1}^{N_\mathcal{H}} h_j \theta_{j,\ak{\thetaidx}}^{\mathcal{H}}  +   a_{\ak{\thetaN}, \mathcal{H}} (\boldsymbol h), & \min_{\boldsymbol x}  f_{\ak{\thetaN}} (\boldsymbol x, \boldsymbol h) =  \sum_{j=1}^{N_\mathcal{H}} h_j \theta_{j,\ak{\thetaidx}}^{\mathcal{H}}  - a_{\ak{\thetaN}, \mathcal{H}} (\boldsymbol h),\\
+\nonumber \max_{\boldsymbol h}  f_{\ak{\thetaN}} (\boldsymbol x, \boldsymbol h) = \sum_{i=1}^{N}x_i \theta_{i,\ak{\thetaidx}}^{\mathcal{V}} + b_{\ak{\thetaN}, \mathcal{V}} (\boldsymbol x), & \min_{\boldsymbol h}  f_{\ak{\thetaN}} (\boldsymbol x, \boldsymbol h) = \sum_{i=1}^{N}x_i \theta_{i,\ak{\thetaidx}}^{\mathcal{V}}  -   b_{\ak{\thetaN}, \mathcal{V}} (\boldsymbol x), \\
+(\#eq:max)
+a_{\ak{\thetaN}, \mathcal{H}} (\boldsymbol h) \equiv \sum_{i=1}^N \left| \theta_{i,\ak{\thetaidx}}^{\mathcal{V}} + \sum_{j=1}^{N_\mathcal{H}}  h_j  \theta_{ij,\ak{\thetaidx}}^{\mathcal{VH}}\right|, &  b_{\ak{\thetaN}, \mathcal{V}} (\boldsymbol x) \equiv \sum_{j=1}^{N_\mathcal{H}} \left| \theta_{j,\ak{\thetaidx}}^{\mathcal{H}}  + \sum_{i=1}^N  x_i  \theta_{ij,\ak{\thetaidx}}^{\mathcal{VH}}\right|,
+\end{align}
+where  $\boldsymbol h^T \ak{\thetaN}^{\mathcal{H}}=\sum_{j=1}^{N_\mathcal{H}} h_j \theta_{j,\ak{\thetaidx}}^{\mathcal{H}}$, $\boldsymbol x^T \ak{\thetaN}^{\mathcal{V}}=  \sum_{i=1}^{N}x_i \theta_{i,\ak{\thetaidx}}^{\mathcal{V}}$ and $\Gam\equiv \max_{\boldsymbol h} a_{\ak{\thetaN}, \mathcal{H}} (\boldsymbol h)$.
+From this, it follows that
+\begin{align*}
+\REP_{(\boldsymbol X, \boldsymbol H)}(\ak{\thetaN}) &=  \max_{\boldsymbol h}\max_{\boldsymbol x}f_{\ak{\thetaN}} (\boldsymbol x , \boldsymbol h) -  \min_{\boldsymbol h}\min_{\boldsymbol x}f_{\ak{\thetaN}} (\boldsymbol x , \boldsymbol h)\\
+&= \max_{\boldsymbol h_1 }  \max_{\boldsymbol h_2 }\left[ (\boldsymbol h_1 - \boldsymbol h_2)^T  \ak{\thetaN}^{\mathcal{H}}  +   a_{\ak{\thetaN}, \mathcal{H}} (\boldsymbol h_1)  + a_{\ak{\thetaN}, \mathcal{H}} (\boldsymbol h_2)\right],
+\end{align*}
+which leads to the upper bound $\REP_{(\boldsymbol X, \boldsymbol H)}(\ak{\thetaN}) \leq 2 \Gam + 2 | \ak{\thetaN}^{\mathcal{H}} |_1$.
+Then, taking $\boldsymbol h_1=\boldsymbol h_2$ (i.e., before maximization) gives $\REP_{(\boldsymbol X, \boldsymbol H)}(\ak{\thetaN}) \geq 2\Gam$ and taking $\boldsymbol h_1=-\boldsymbol h_2$, such that $\boldsymbol h_1^T \ak{\thetaN}^{\mathcal{H}} = |\ak{\thetaN}^{\mathcal{H}}|_1$, gives $\REP_{(\boldsymbol X, \boldsymbol H)}(\ak{\thetaN}) \geq 2|\ak{\thetaN}^{\mathcal{H}}|_1$; this yields the lower bound $\REP_{(\boldsymbol X, \boldsymbol H)}(\ak{\thetaN})\geq 2\max\{\Gam, |\ak{\thetaN}^{\mathcal{H}} |_1\}$.
+
+We next consider $\elt$ and, by \@ref(eq:max), write
+\begin{align*}
+\elt &=  \max_{\boldsymbol h}\max_{\boldsymbol x}f_{\ak{\thetaN}} (\boldsymbol x , \boldsymbol h) -  \max_{\boldsymbol h}\min_{\boldsymbol x}f_{\ak{\thetaN}} (\boldsymbol x , \boldsymbol h)\\
+&= \max_{\boldsymbol h_1} \min_{\boldsymbol h_2 }\left[ (\boldsymbol h_1 - \boldsymbol h_2)^T \ak{\thetaN}^{\mathcal{H}} + a_{\ak{\thetaN}, \mathcal{H}} (\boldsymbol h_1)  + a_{\ak{\thetaN}, \mathcal{H}} (\boldsymbol h_2)\right].
+\end{align*}
+Taking $\boldsymbol h_1=\boldsymbol h_2$ and maximizing over both $\boldsymbol h_1,\boldsymbol h_2$ produces the upper bound $\elt \leq 2\Gam$.
+Then, using $(\boldsymbol h_1 - \boldsymbol h_2)^T \ak{\thetaN}^{\mathcal{H}} + a_{\ak{\thetaN}, \mathcal{H}} (\boldsymbol h_2) \geq - 2|\ak{\thetaN}^{\mathcal{H}} |_1$ and maximizing over $\boldsymbol h_1$ gives $\elt \geq  \Gam- 2|\ak{\thetaN}^{\mathcal{H}} |_1$, while setting $\boldsymbol h_1=\boldsymbol h_2^*$ for $\boldsymbol h_2^*$ such that $-(\boldsymbol h_2^*) ^T  \ak{\thetaN}^{\mathcal{H}}  + a_{\ak{\thetaN}, \mathcal{H}} (\boldsymbol h_2^*) = \min_{\boldsymbol h_2} [-\boldsymbol h_2^T \ak{\thetaN}^{\mathcal{H}}  + a_{\ak{\thetaN}, \mathcal{H}} (\boldsymbol h_2)]$ gives $\elt \geq 2 a_{\ak{\thetaN}, \mathcal{H}} (\boldsymbol h_2^*) \geq \Gamc \equiv \min_{\boldsymbol h}  a_{ \ak{\thetaN}, \mathcal{H}} (\boldsymbol h)$.
+Finally, note that for any $\boldsymbol h$, the triangle inequality gives
+\begin{align*}
+\Gam  \equiv \max_{\boldsymbol h_1}  a_{\ak{\thetaN}, \mathcal{H}} (\boldsymbol h_1) &\geq
+[a_{\ak{\thetaN}, \mathcal{H}} (\boldsymbol h)  + a_{\ak{\thetaN}, \mathcal{H}} (-\boldsymbol h)]/2 \\
+&=
+2^{-1}\sum_{i=1}^{N } \left(\left| \theta_{i,\ak{\thetaidx}}^{\mathcal{V}}   + \sum_{j=1}^{N_{\mathcal{H}}} h_j \theta_{ij,\ak{\thetaidx}}^{\mathcal{VH}} \right|
++ \left| \theta_{i,\ak{\thetaidx}}^{\mathcal{V}}   - \sum_{j=1}^{N_{\mathcal{H}}} h_j \theta_{ij,\ak{\thetaidx}}^{\mathcal{VH}} \right|\right)\\
+& \geq \sum_{i=1}^{N }  \left| \theta_{i,\ak{\thetaidx}}^{\mathcal{V}} \right| \equiv | \ak{\thetaN}^{\mathcal{V}}|_1.
+\end{align*}
+\hfill $\Box$
+
+**Proof of Theorem \@ref(thm:something).**  Let $L_{\ak{\thetaN}}(\boldsymbol X) = \log[ P_{ \ak{\thetaN}}(\boldsymbol X)/ \min_{\boldsymbol y \in \mathcal{X}^N} P_{\ak{\thetaN}}(\boldsymbol y) ]/\REP(\ak{\thetaN})$, where again $\boldsymbol X=(X_1, \dots, X_N)$ and $\REP(\ak{\thetaN})= \log[\max_{\boldsymbol y \in \mathcal{X}^N} P_{\ak{\thetaN}}(\boldsymbol y)/ \min_{\boldsymbol y \in \mathcal{X}^N} P_{ \ak{\thetaN}}(\boldsymbol y) ]$.
+As $L_{\ak{\thetaN}}(\boldsymbol X)\in[0,1]$, convergence of $L_{ \ak{\thetaN}}(\boldsymbol X)$ to 1 in probability under $P_{\ak{\thetaN}}$ is equivalent to convergence to $1$ in expectation under $P_{\ak{\thetaN}}$ (i.e., convergence in expectation implies probabilistic convergence by Markov's inequality while probabilistic convergence implies convergence in expectation by uniform integrability/boundedness).
+
+For $\epsilon \in(0,1)$, let $\mathcal{M}_{\epsilon, \ak{\thetaN}}$ denote a modal set as in \@ref(eq:mode).
+By Theorem \@ref(thm:degenFOES), $P_{\ak{\thetaN}}\left( \boldsymbol X\in \mathcal{M}_{\epsilon, \ak{\thetaN}}\right) \rightarrow 1$ holds as $N \rightarrow \infty$ and, by definition of \@ref(eq:mode), $\boldsymbol X\in \mathcal{M}_{\epsilon, \ak{\thetaN}}$ follows if and only if
+$1-L_{\ak{\thetaN}}(\boldsymbol X)<\epsilon$.
+Hence, $L_{\ak{\thetaN}}(\boldsymbol X) \stackrel{p,\mathrm{E}}{\longrightarrow} 1$ holds under $\ak{\thetaN}$ in Theorem \@ref(thm:something).
+The convergence $L_{-\ak{\thetaN}}(\boldsymbol X) \stackrel{p,\mathrm{E}}{\longrightarrow} 1$  under $-\ak{\thetaN}$ likewise follows from Corollary \@ref(cor:sign).
+\hfill $\Box$
+
+```{remark, echo = TRUE}
+Consider the exponential graph model from \@ref(eq:mod2) involving counts of edges, 2-stars and triangles with fixed parameters $\thetaN = (\theta_{1},\theta_2,\theta_3)^\prime \in \mathbb{R}^3$ for $N\geq 1$.
+This model is unstable whenever $|\theta_2| + |\theta_3| \neq 0$.
+To see this, consider an even number $n>2$ of nodes and let $\boldsymbol x_0$ denote the data outcome in $\mathcal{X}^N \equiv \{0,1\}^N$ with all $N = {n \choose 2}$ edges being zero, let $\boldsymbol x_1$ denote the outcome with all edges being 1, and let $\boldsymbol x_2$ denote the edge configuration from dividing the nodes into two equal groups, with no edges within a group and all edges between the groups (so that no triangles exist in $\boldsymbol x_2$).
+Then, the $N$-scaled  log-ratio \@ref(eq:elpr) for the exponential graph model \@ref(eq:mod2) can, by definition, be bounded below by
+\begin{align*}
+\frac{1}{N}\REP_N(\ak{\thetaN}) &\geq \max_{i=1,2}\frac{1}{N}
+\left| \log\left[ \frac{P_{\ak{\thetaN}}(\boldsymbol x_i)}{P_{\ak{\thetaN}}(\boldsymbol x_0)}\right] \right| \\
+&= (n-2) \max\left\{ \left| \theta_2 + \frac{\theta_3}{3}+\frac{\theta_1}{n-2} \right|, \frac{n}{4(n-1)} \left| \theta_2 + \frac{8\theta_1}{n-2} \right| \right\};
+\end{align*}
+a similar expression also holds for an odd node number $n>2$.
+Consequently, for all fixed parameters excluding $\theta_2=\theta_3=0$, $\lim_{N\to \infty}\REP_N(\ak{\thetaN})/N=\infty$ then follows and the graph model with 2-stars and triangles is S-unstable, as suggested by the breach of Proposition \@ref(prp:prop1) for this model when $|\theta_2|+|\theta_3|\neq 0$.
+```
+
+```{remark, echo=TRUE}
+Let $M \geq 1$ denote a possible number of replications and consider data $\boldsymbol Y_{N,M} \equiv (\boldsymbol X^{(1)}_N, \dots, \boldsymbol X^{(M)}_N)$ formed by $\{ \boldsymbol X^{(j)}_N\}_{j=1}^M$ as $M$  iid replications of a random vector $\boldsymbol X_N=(X_1,\ldots,X_N)$, where the latter follows a FOES model with probabilities $P_{\ak{\thetaN}}(\boldsymbol x)>0$, $\boldsymbol x\in\mathcal{X}^N$.
+This leads to a joint model, say $P_{\ak{\thetaN}}(\boldsymbol y)$, $\boldsymbol y\in\mathcal{X}^{NM}$, for $\boldsymbol Y_{N,M}$ consisting of $N*M$ random variables in total.
+Then, the LREP for $\boldsymbol Y_{N,M}$, scaled by associated size, is given by
+\begin{align*}
+\frac{1}{NM}\REP_{\boldsymbol Y_{N,M}}(\ak{\thetaN} ) &\equiv \frac{1}{NM}\log\left[\frac{\max_{\boldsymbol y \in \mathcal{X}^{NM}}P_{\ak{\thetaN}}(\boldsymbol y)  }{\min_{\boldsymbol y \in \mathcal{X}^{NM}}P_{\ak{\thetaN}}(\boldsymbol y)} \right] \\
+&= \frac{1}{NM}\log\left[\frac{\max_{\boldsymbol x \in \mathcal{X}^{N}}P_{\ak{\thetaN}}(\boldsymbol x)  }{\min_{\boldsymbol x \in \mathcal{X}^{N}}P_{\ak{\thetaN}}(\boldsymbol x)} \right]^M \equiv \frac{1}{N}\REP_{\boldsymbol X_{N}}(\ak{\thetaN} ),
+\end{align*}
+where $\REP_{\boldsymbol X_{N}}(\ak{\thetaN} )  \equiv \REP(\ak{\thetaN} )$ denotes the log-ratio of extremal probabilities for $\boldsymbol X_N$ defined from \@ref(eq:elpr).
+That is, due to iid properties, the sample-size corrected  LREP for $\boldsymbol Y_{N,M}$ equals the analog, $\REP(\ak{\thetaN} )/N$, from the underlying common data model for $\boldsymbol X_N$ alone, regardless of the level $M \geq 1$ of independent replication.
+Hence, the definition of an S-unstable  model is unaffected by independent replication.
+For computational purposes, this aspect also implies that if the original data $\boldsymbol X_N=(X_1,\ldots,X_N)$ in a FOES model consist of $N$ iid random variables, then the size-scaled log-ratio \@ref(eq:elpr)  may be calculated as
+$$
+\frac{1}{N}\REP( \ak{\thetaN} ) \equiv \frac{1}{N}\REP_{\boldsymbol X_{N}}( \ak{\thetaN} ) = \log\left[ \frac{\max_{ x \in \mathcal{X}}P_{\ak{\thetaN}}(X_1=x)} {\min_{ x \in \mathcal{X}}P_{\ak{\thetaN}}(X_1=x)} \right]
+$$
+based on the extremal probabilities of just one random variable $X_1$.
+```
+
+# Details on Centered Graph Example
+\ak{To illustrate centered model parameterizations and n examination of stability in these, consider the two-star model (\ref{eq:mod2}) for the $N = {n \choose 2}$ edges in simple graph with $n$ nodes and binary edge-variables $(X_1,\ldots,X_N)\in\{0,1\}^N$.
+Here a common or standard parameterization in (\ref{eq:mod2}) leads to a conditional probability of ``$1$" for an edge $i$ as
+$$
+\mathrm{logit}[  P_{\thetaN}(X_{i} = 1 | X_j=x_j, j\neq i)=  \theta_1 + \theta_2 \sum_{j \in \mathcal{N}_i } x_j,$$ 
+based on summing other edge observations $x_j$ in a neighborhood $\mathcal{N}_i = \{{s}_j: {s}_i \cap {s}_j \neq \emptyset\}$ to edge $i$ (i.e., edges $j$,  marked by pairs of graph vertices ${s}_{j} = \{ v_{j_1}, v_{j_2} \}$, that share a common vertex with edge $i$ marked by the vertex pair ${s}_i =\{v_{i_1}, v_{i_2}\}$).
+In contrast, a centered conditional would yield
+$$
+\mathrm{logit}[  P_{\thetaN}(X_{i} = 1 | X_j=x_j, j\neq i)=  \theta_1 +  \frac{\theta_2}{2(n-1)}\sum_{j \in \mathcal{N}_i} (x_{j}-\kappa),
+$$
+involving a parameter $\theta_1 \equiv \mathrm{logit}(\kappa)$ for $\kappa \in (0,1)$ and $2(n-2)$ as the size of $\mathcal{N}_i$ (cf.
+\citep{kaiser2009exploring}); the corresponding joint model would involve parameters $\theta_{1,\thetaidx}= \mathrm{logit}(\kappa) - \kappa\theta_2$
+and $\theta_{2,\thetaidx} = \theta_2/(2(n-2))$ in (\ref{eq:mod2}).
+The purpose of the centerization is to have $\kappa \in(0,1)$ represent a model mean parameter (note $E X_i =\kappa$ is the edge proportion/probability under independence $\theta_2=0$), while a separate parameter $\theta_2$ (for dependence) modifies the conditional probability of ``$X_i=1$" up/down from $\kappa$, depending on neighbors $x_j=1$ or 0.
+A similar interpretation does not hold in the uncentered model; see \citep{kaiser2009exploring} and \citep{casleton2017local} for a discussion of the centered parameterization in spatial and network modeling, where the intent is to improve parameter interpretation and help detect degeneracy (e.g., intuitively given by large dependence parameters $\theta_2$, so that $\kappa$ fails to correspond to the mean or the marginal proportion of $1$'s in data generations from the model).
+However, from a different perspective using the model measures here, centered parameterizations also lead to models which are more stable over wider regions of the parameter space.
+To illustrate, in the standard/uncentered parameterization for a graph model (\ref{eq:mod2}) with only two-stars $\thetaN = (0,\theta_2,0)$, our measure $\REP(\thetaN)/N$ of model instability becomes $\REP(\thetaN)/N = |\theta_2| (n-2)$, which is unbounded as $N\to \infty$ for all non-zero parameters $\theta_2$ (i.e., all models with $\theta_2\neq 0$ are S-unstable).
+However, in the centered parameterization  with only two-stars, corresponding to $\kappa=1/2$ and $\thetaN= 0.5*(-\theta_2,\theta_2/(n-2),0)$, the measure becomes
+\begin{align*}
+   \REP(\thetaN)/N  &= 0.5*|\theta_2|  \max_{x_i, y_i \in\{0,1\}}\left|  \frac{1}{N}\sum_{i=1}^N x_i \left( 1 - \frac{1}{2(n-2)} \sum_{j \in \mathcal{N}_i} x_j\right) - \frac{1}{N}\sum_{i=1}^N y_i \left( 1 - \frac{1}{2(n-2)} \sum_{j \in \mathcal{N}_i} y_j\right)\right| \\
+   &\leq |\theta_2|,   
+\end{align*}
+which is bounded, for any fixed $\theta_2 \in\mathbb{R}$, as $N$ increases (i.e., note $\sum_{j \in \mathcal{N}_j} x_i/(2(n-2)) \in [0,1]$ is a conditional/neighborhood sample proportion while $\sum_{i=1}^N x_i/N \in[0,1]$ is a marginal proportion).
+This aspect owes to adjusting parameters by neighborhood sizes in centered conditional distributions, but centering also induces an additional effect of alternating signs in parameters (e.g., $\thetaN= 0.5*(-\theta_2,\theta_2/(n-2),0)$).
+The latter has been suggested in other contexts with exponential graph models for regulating degeneracy \citep[cf.][]{snijders2006new}.}
+
 \clearpage
-
-# References {-}

### 2019-06-18 writing/paper-note/chapters/instab.Rmd
revision of note

@@
 Section \@ref(implications) then emphasizes the implications of such model instability, showing that such impropriety can be expected to numerically hinder maximum likelihood estimation and MCMC-based simulations.
-As one potential remedy, suggestions are given for constraining model parameterizations to avoid the most problematic regions of the parameter space.
+As one potential remedy, suggestions are given for constraining model parameterizations to avoid the most problematic regions of the parameter space \ak{(see Section \ref{bayes} for further discussion)}.
 Proofs of the main results appear in Appendix \@ref(appendix-instab).
@@
 \ak{(For clarity, we use ``model family" to refer to generic class or functional form of parametric distributions, while "model" often refers here to a member of the class for an instantiation or selection of parameters.
-This distinction is not so important for the examples presented in Sections \ref{discrete-exponential-family-models}-\ref{deep-learning}, but becomes more so in Section \ref{instability-results} where model instability depends on both form and a given parameter configuration, rather than just model class form.)} For illustration, this section presents three specific examples of FOES model \ak{classes}, including models with deep architectures.
+This distinction is not so important for the examples presented in Sections \ref{discrete-exponential-family-models}-\ref{deep-learning}, but becomes more so in Section \ref{instability-results} where model instability depends on both form and a given parameter configuration)} For illustration, \ak{Sections \ref{discrete-exponential-family-models}-\ref{deep-learning} present} three specific examples of FOES model \ak{classes}, including models with deep architectures.
 
@@
 In addition to such standard models for discrete independent data, exponential models of FOES type commonly arise with dependent spatial data [@besag1974spatial] and network/relational data [@wasserman1994social; @handcock2003assessing].
-For a random graph  or network  with, say, $n$ nodes,  consider $N={n \choose 2}$ random edges where the $i$th edge is associated with a pair of nodes $s_i \equiv \{v_i,u_i\}$  and a binary variable $X_i\in\{0,1\}$ indicating   presence/absence of an edge among the node pair $s_i$, $i=1,\ldots,N$.
+For a random graph or network with, say, $n$ nodes \ak{$\{v_i\}_{i = 1}^n$},  consider $N={n \choose 2}$ random edges where the $i$th edge is associated with a pair of nodes $s_i \equiv \{v_{i_1},v_{i_2}\}$  and a binary variable $X_i\in\{0,1\}$ indicating   presence/absence of an edge among the node pair $s_i$, $i=1,\ldots,N$.
 Here the length $N$ of the edge variable sequence $\boldsymbol X = (X_1,\ldots,X_N)$ increases as a function of node number $n$ and corresponding exponential models often incorporate graph topographical features derived from $\boldsymbol X$.
@@
 \end{equation}
-with normalizing function
-$$
-\psi(\ak{\thetaN}) = \log \sum_{\tilde{\boldsymbol x} \in \{\pm 1\}^{N+N_H} } \exp\left[ (\ak{\thetaN}^{\mathcal{H}})^T \boldsymbol h + \boldsymbol (\ak{\thetaN}^{\mathcal{V}})^T \boldsymbol x + \boldsymbol h^T  \ak{\thetaN}^{\mathcal{HV}} \boldsymbol x\right].
-$$
+with \ak{normalizing function $\psi(\ak{\thetaN})$}.
+<!-- $$ -->
+<!-- \psi(\ak{\thetaN}) = \log \sum_{\tilde{\boldsymbol x} \in \{\pm 1\}^{N+N_H} } \exp\left[ (\ak{\thetaN}^{\mathcal{H}})^T \boldsymbol h + \boldsymbol (\ak{\thetaN}^{\mathcal{V}})^T \boldsymbol x + \boldsymbol h^T  \ak{\thetaN}^{\mathcal{HV}} \boldsymbol x\right].
+-->
+<!-- $$ -->
 Let $\ak{\thetaN} = (\ak{\thetaN}^{\mathcal{H}}, \ak{\thetaN}^{\mathcal{V}}, \ak{\thetaN}^{\mathcal{HV}} ) \in \Theta_\ak{\thetaidx} \equiv \mathbb{R}^{q(N)}$, with $q(N) = N + N_\mathcal{H} + N*N_\mathcal{H}$, denote \ak{a given} parameter vector for the RBM, as indexed by the number $N$ of visible random variables (which may differ from the actual lengths of these parameter vectors).
@@
 Here the baseline model \@ref(eq:RBM1) for hidden/visible variables is a linear exponential one in sufficient statistics $(\tilde{\boldsymbol X}, \boldsymbol X^T\boldsymbol H)$ using $\tilde{\boldsymbol X}=(\boldsymbol X,\boldsymbol H)$ from \@ref(eq:RBM1), but the form differs from the previous exponential models in \@ref(eq:expo) in that the lengths of parameters $\ak{\thetaN}$ and statistics  $(\tilde{\boldsymbol X}, \boldsymbol X^T\boldsymbol H)$ increase to incorporate more visible variables.
-That is, in contrast to \@ref(eq:expo), the natural parameter function involved in the RBM model \@ref(eq:RBM1), as the identity mapping of the parameters $\ak{\thetaN}\in\mathbb{R}^{q(N)}$, naturally grows in  dimension  $q(N)\to \infty$ to accommodate visible variables $X_1, \dots, X_N$ of increasing sample size $N\to \infty$.
+That is, in contrast to \@ref(eq:expo), the natural parameter function involved in the RBM model \@ref(eq:RBM1), as the identity mapping of the parameters $\ak{\thetaN}\in\mathbb{R}^{q(N)}$, naturally grows in  dimension  $q(N)\to \infty$ to accommodate visible variables $X_1, \dots, X_N$ of increasing data size $N\to \infty$.
 Additionally, one may further arbitrarily choose the number $N_\mathcal{H}$ of hidden variables $\boldsymbol H$ in the joint RBM model \@ref(eq:RBM1) to define a marginal model \@ref(eq:RBM2) for the $N$ visible variables $\boldsymbol X$, and the number $N_\mathcal{H}$ of hiddens may also potentially increase with $N$.
@@
 
-## Deep Learning
+\ak{\subsection{Binary Deep Learning Models} \label{deep-learning}}
 
@@
 
-**Deep Boltzmann machine (DBM).** The DBM class of models maintains conditional independence within all layers in the model by stacking RBM models and only allowing conditional dependence between neighboring layers.
+\noindent \emph{Deep Boltzmann machine (DBM).} The DBM class of models maintains conditional independence within all layers in the model by stacking RBM models and only allowing conditional dependence between neighboring layers.
 The joint probability mass function for a DBM is
@@
 $$ 
-for $\tilde{\boldsymbol x} = (\boldsymbol h^{(1)}, \dots, \boldsymbol h^{(M)}, \boldsymbol x) \in \mathcal{X}^{N_{(H,1)} + \cdots + N_{(H,M)} +N}$ where 
-$$
-\psi(\ak{\thetaN}) = \log \sum\limits_{\tilde{\boldsymbol x} \in \mathcal{X}^{N_{(H,1)} + \cdots + N_{(H,M)} +N}} \exp\left[ \sum\limits_{i = 1}^M\boldsymbol \alpha^{(i)T} \boldsymbol h^{(i)} + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^{(1)T} \Gamma^{(0)} \boldsymbol x + \sum\limits_{i = 1}^{M - 1} \boldsymbol h^{(i)T} \Gamma^{(i)} \boldsymbol h^{(i + 1)}\right],
-$$ 
+for $\tilde{\boldsymbol x} = (\boldsymbol h^{(1)}, \dots, \boldsymbol h^{(M)}, \boldsymbol x) \in \mathcal{X}^{N_{(H,1)} + \cdots + N_{(H,M)} +N}$ \ak{where again $\psi(\ak{\thetaN})$}
+<!-- $$ -->
+<!-- \psi(\ak{\thetaN}) = \log \sum\limits_{\tilde{\boldsymbol x} \in \mathcal{X}^{N_{(H,1)} + \cdots + N_{(H,M)} +N}} \exp\left[ \sum\limits_{i = 1}^M\boldsymbol \alpha^{(i)T} \boldsymbol h^{(i)} + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^{(1)T} \Gamma^{(0)} \boldsymbol x + \sum\limits_{i = 1}^{M - 1} \boldsymbol h^{(i)T} \Gamma^{(i)} \boldsymbol h^{(i + 1)}\right], -->
+<!-- $$  -->
 is the normalizing function for $\ak{\thetaN} = (\boldsymbol \alpha^{(1)}, \dots, \boldsymbol \alpha^{(M)}, \boldsymbol \beta,\Gamma^{(0)}, \dots, \Gamma^{(M - 1)}) \in \Theta_\ak{\thetaidx} \subset \mathbb{R}^{q(N)}$, consisting of model parameters $\boldsymbol \beta \in \mathbb{R}^N$, $\boldsymbol \alpha^{(i)} \in \mathbb{R}^{N_{(H,i)}}$, $i = 1, \dots, M$, along with a matrix $\Gamma^{(0)}$ of dimension $N_{(H,1)} \times N$, and matrices $\Gamma^{(i)}$ of dimension
@@
 
-**Deep belief network (DBN).** A DBN resembles a DBM in that there are multiple layers of latent random variables stacked in a deep architecture with no conditional dependence between layers.
-The difference between the DBM and DBN models is that all but the last stacked layer \ak{(layer $0$, which contains the visible random variables $X_1, \dots, X_N$)} in a DBN are Bayesian networks [see @pearl985bayesian], rather than RBMs.
-A Bayesian network is a probabilistic graphical model that defines conditional dependence to be directed, rather than undirected (as with the RBM).
+\noindent \emph{Deep belief network (DBN).} A DBN resembles a DBM in that there are multiple layers of latent random variables stacked in a deep architecture with no conditional dependence between layers.
+The difference between the DBM and DBN models is that all but the last stacked layer \ak{(e.g.
+"layer $0$" corresponding to the visible random variables $X_1, \dots, X_N$)} in a DBN are Bayesian networks [see @pearl985bayesian], rather than RBMs.
+A Bayesian network \ak{is a class of probabilistic graphical models that define} conditional dependence to be directed, rather than undirected (as with the RBM).
 Thus for visibles $X_1, \dots, X_N$ with support $\mathcal{X}^N, \mid \mathcal{X} \mid < \infty$, a DBN is also a FOES model with $q(N)$ the length of parameter vector is dependent on the dimension of the visibles because $P_{\ak{\thetaN}}(\boldsymbol x)>0$ for all $\boldsymbol x \in\mathcal{X}^N$.
@@
 We now present a formal definition for instability of FOES models as well as a simple condition for identifying instability in a FOES model sequence.
+\ak{The basic intuition is that, if the smallest and largest probabilities from a given model are ``too different," then instability may arise, which manifests as extreme sensitivity in other probability behavior: large shifts in probability may be associated with only small differences in the input space for FOES models, and is related to a particular model placing almost all probability on a subset of potential outcomes.
+We present both non-asymptotic (Theorem \ref{thm:instab-elpr}) and asymptotic (Theorem \ref{thm:degenFOES}) results to this effect.}
 
@@
 
-To define a measure of instability in FOES models, it is useful to consider the behavior of data  models $P_{\ak{\thetaN}}$, again supported on a set $\mathcal{X}^N$ of outcomes for $\boldsymbol X\equiv \boldsymbol X_N =(X_1,\ldots,X_N)$ \ak{and for a prescribed configuration of parameters $\ak{\thetaN}$,  in connection to the sample size $N$.
+To define a measure of instability in FOES models, it is useful to consider the behavior of data models $P_{\ak{\thetaN}}$, again supported on a set $\mathcal{X}^N$ of outcomes for $\boldsymbol X\equiv \boldsymbol X_N =(X_1,\ldots,X_N)$ and for a prescribed configuration of parameters $\ak{\thetaN}$, in connection to the sample size $N$.
 A relevant quantity to this end is a log-ratio of extremal probabilities (LREP), defined as
@@
 
-```{remark, label="rmk1", echo=TRUE}
+```{remark, echo = TRUE}
+\ak{Different notions of model stability/instability exist.
+For example, \citet{handcock2003assessing} refers to how small changes in parameters may dramatically change a probability mass function, while \citet{kaiser2007statistical} discusses how interpretation of model parameters and conditional expectations may break down in parts of the parameter space.
+However, these may be intuitively connected in the sense that the same (undesirable) parameter/model configurations may often be associated with varying conceptualizations of instability.
+While beyond the scope of the current paper, \citet{kaplan2019properties} investigates and numerically demonstrates some such connections for a class of RBM models.
+Section \ref{stable-models} also explains one relation between our notion of stable models and parameter interpretation in centered models \citep[][cf.]{kaiser2009exploring}.
+}
+```
+
+```{remark, echo=TRUE}
 In the definition \@ref(eq:Sun) of S-instability, we note that the numerical measure $\REP(\ak{\thetaN})/N$ of model instability is invariant to *independent replications* of data.
-That is, let $M \geq 1$ denote a possible number of replications and consider data $\boldsymbol Y_{N,M} \equiv (\boldsymbol X^{(1)}_N, \dots, \boldsymbol X^{(M)}_N)$ formed by $\{ \boldsymbol X^{(j)}_N\}_{j=1}^M$ as $M$  iid replications of a random vector $\boldsymbol X_N=(X_1,\ldots,X_N)$, where the latter follows a FOES model with probabilities $P_{\ak{\thetaN}}(\boldsymbol x)>0$, $\boldsymbol x\in\mathcal{X}^N$.
-This leads to a joint model, say $P_{\ak{\thetaN}}(\boldsymbol y)$, $\boldsymbol y\in\mathcal{X}^{NM}$, for $\boldsymbol Y_{N,M}$ consisting of $N*M$ random variables in total.
-Then, the LREP for $\boldsymbol Y_{N,M}$, scaled by associated size, is given by
-\begin{align*}
-\frac{1}{NM}\REP_{\boldsymbol Y_{N,M}}(\ak{\thetaN} ) &\equiv \frac{1}{NM}\log\left[\frac{\max_{\boldsymbol y \in \mathcal{X}^{NM}}P_{\ak{\thetaN}}(\boldsymbol y)  }{\min_{\boldsymbol y \in \mathcal{X}^{NM}}P_{\ak{\thetaN}}(\boldsymbol y)} \right] \\
-&= \frac{1}{NM}\log\left[\frac{\max_{\boldsymbol x \in \mathcal{X}^{N}}P_{\ak{\thetaN}}(\boldsymbol x)  }{\min_{\boldsymbol x \in \mathcal{X}^{N}}P_{\ak{\thetaN}}(\boldsymbol x)} \right]^M \equiv \frac{1}{N}\REP_{\boldsymbol X_{N}}(\ak{\thetaN} ),
-\end{align*}
-where $\REP_{\boldsymbol X_{N}}(\ak{\thetaN} )  \equiv \REP(\ak{\thetaN} )$ denotes the log-ratio of extremal probabilities for $\boldsymbol X_N$ defined from \@ref(eq:elpr).
-That is, due to iid properties, the sample-size corrected  LREP for $\boldsymbol Y_{N,M}$ equals the analog, $\REP(\ak{\thetaN} )/N$, from the underlying common data model for $\boldsymbol X_N$ alone, regardless of the level $M \geq 1$ of independent replication.
-Consequently, the definition of an S-unstable  model  is  unaffected by independent replication and all instability properties  may be characterized by those of one observation  from the common FOES model.
-For computational purposes, this aspect also implies that if the original data $\boldsymbol X_N=(X_1,\ldots,X_N)$ in a FOES model consist of $N$ iid random variables, then the size-scaled log-ratio \@ref(eq:elpr)  may be calculated as
-$$
-\frac{1}{N}\REP( \ak{\thetaN} ) \equiv \frac{1}{N}\REP_{\boldsymbol X_{N}}( \ak{\thetaN} ) = \log\left[ \frac{\max_{ x \in \mathcal{X}}P_{\ak{\thetaN}}(X_1=x)} {\min_{ x \in \mathcal{X}}P_{\ak{\thetaN}}(X_1=x)} \right]
-$$
-based on the extremal probabilities of just one random variable $X_1$.
+\ak{Consequently, the definition of an
+S-unstable model is unaffected by independent replication and all instability properties may be
+characterized by those of one observation from the common FOES model. Remark A.2 of the Appendix provides more details.}
 ```
@@
 which represents the biggest log-probability ratio for a one-component change \ak{(i.e., the data vector differs in exactly one component value)} in data outcomes in a FOES model with parameter $\ak{\thetaN}$.
-\ak{This is the smallest difference possible between two data vectors, making it a good notion of a small distance in the input space.} We then have the following result prescribing the behavior of $\DN(\ak{\thetaN})$ for S-unstable FOES models.
+\ak{This is the smallest difference possible between two data vectors, making it a good notion of a small distance in the input space and has been similarly used by \citet[][, Sec 2.2]{schweinberger2011instability}.} We then have the following result prescribing the behavior of $\DN(\ak{\thetaN})$ for S-unstable FOES models.
 
@@
 \begin{enumerate}[(i)]
-\item For any integer $N \geq 1$ and any given $C>0$, if $\REP(\ak{\thetaN})/N > C$ in (\ref{eq:elpr}), then
-    $$ \DN(\ak{\thetaN}) > C,$$
-    or probabilities from a one-component change in some outcome have log-ratio exceeding $C$.
+\item For any integer $N \geq 1$ and any given $C>0$, if $\REP(\ak{\thetaN})/N > C$ holds in (\ref{eq:elpr}), then
+    $$ \DN(\ak{\thetaN}) > C$$
+    follows.
 \item Suppose the FOES model sequence is S-unstable.
@@
 
-Theorem \@ref(thm:instab-elpr)(i) is a non-asymptotic result, which connects to the definition of instability in a FOES model through a log-ratio of extreme model probabilities \@ref(eq:elpr) being too large relative to the associated sample size $N$.
+\ak{Theorem \ref{thm:instab-elpr} aims to describe some model implications when a log-ratio of extreme model probabilities (\ref{eq:elpr}) is too large relative to the associated sample size $N$, which underlies the definition of a S-unstable FOES model (\ref{eq:Sun}).
+If so, Theorem \ref{thm:instab-elpr}(i) guarantees the FOES model must also exhibit correspondingly large changes in probability for small differences among some data configurations.
+Hence, as a further consequence in Theorem \ref{thm:instab-elpr}(ii), S-unstable models can never have universally bounded changes in probability among single component variations in data configurations.
+While not all one-component changes in data may produce massive changes in probability, unstable models must have some such data outcomes with this property.
+Hence,
+unstable probability structures may exhibit extreme sensitivity through large peaks and troughs over the sample space.
+Theorem \ref{thm:instab-elpr} supports and generalizes findings in \citet[Theorem 1]{schweinberger2011instability} for classes of discrete exponential models (i.e., the result there entails that, if $\REP(\thetaN)/N $ is unbounded as $N$ increases, then the nearest-point log-odds ratio  $\DN(\ak{\thetaN})$  is unbounded as well in these models).}
+
+<!-- Theorem \@ref(thm:instab-elpr)(i) is a non-asymptotic result, which connects to the definition of instability in a FOES model through a log-ratio of extreme model probabilities \@ref(eq:elpr) being too large relative to the associated sample size $N$.
 If so, Theorem \@ref(thm:instab-elpr)(i) guarantees the FOES model must also exhibit correspondingly large changes in probability for very small differences among some data configurations, a property that intuitively captures a notion of instability.
-Furthermore, and perhaps more seriously under Theorem \@ref(thm:instab-elpr)(ii),
-S-unstable models can never have universally bounded changes in probability among single component variations in data configurations.
+Furthermore, and perhaps more seriously under Theorem \@ref(thm:instab-elpr)(ii), S-unstable models can never have universally bounded changes in probability among single component variations in data configurations.
 While not all one-component changes in data may produce massive changes in probability, unstable models must have some such data outcomes with this property.
 As a consequence, unstable probability structures may exhibit extreme sensitivity through large peaks and troughs over the sample space.
+-->
 
@@
 However, for all unstable FOES models, Corollary \@ref(cor:sign) generalizes the same principle that unstable models can push all probability to different, and in fact disjoint, parts of the sample space, depending on how parameters fall with respect to zero.
-This feature can numerically complicate likelihood manipulations, such as maximization or MCMC-based Bayes posterior sampling, as further discussed in Section \@ref(implications).
+This feature can numerically complicate likelihood manipulations, such as maximization or MCMC-based Bayes posterior \ak{simulation}, as further discussed in Section \@ref(implications).
 
@@
 However, as shown in the following, bounded statistics may still lead to instability if the parameter dimension is high.
-We next \ak{provide} some examples to illustrate S-instability in FOES models, which also suggest some potential strategies for preventing unstable models.
+We next \ak{provide} some examples to illustrate S-instability in FOES models, which also suggest some potential strategies for preventing unstable models \ak{(e.g., for many model types in the following, a control of the combined magnitudes of parameters can ensure S-stability, which is also discussed further in Section \ref{stable-models})}.
 
@@
 
-A fundamental model considered in the instability work of @schweinberger2011instability involves a one-parameter exponential model corresponding to \@ref(eq:expo) with a real-valued parameter, say $\ak{\thetaN} =  \eta(\ak{\thetaN})\in \mathbb{R}$, and sufficient statistic $\boldsymbol g_N(\boldsymbol x)\in \mathbb{R}$.
+A fundamental model considered in the instability work of @schweinberger2011instability involves a one-parameter exponential model corresponding to \@ref(eq:expo) with a real-valued parameter, say $\ak{\thetaN} =  \eta(\ak{\thetaN})\in \mathbb{R}$, and sufficient statistic $g_N(\boldsymbol x)\in \mathbb{R}$.
 For such models, upon scaling by sample size $N$, the log-ratio of extreme probabilities in \@ref(eq:elpr) for assessing instability becomes
@@
 For illustration, consider the iid Bernoulli model \@ref(eq:mod1) for $\boldsymbol X_N=(X_1,\ldots,X_N)$ with log-odds ratio parameter $\ak{\thetaN} = \log[ P_{\ak{\thetaN}}(X_1=1)/ P_{\ak{\thetaN}}(X_1=0)]\in\mathbb{R}$.
-\ak{Remark 3.1} (Section \@ref(criterion)) then gives the model instability measure \@ref(eq:Sun) directly as
+\ak{Remark 3.2} (Section \@ref(criterion)) then gives the model instability measure \@ref(eq:Sun) directly as
 $$
@@
 so that an unstable (or stable) model results for a divergent (or bounded) parameter sequence  $|\ak{\thetaN}|$. 
-The above instability expression for the Bernoulli model follows as well from  the $N$-scaled LREP value \@ref(eq:UL) for a one-parameter exponential distribution, using that the sufficient statistic involved $g_N(\boldsymbol x)= \sum_{i=1}^N x_i$, $\boldsymbol x =(x_1,\ldots,x_n)\in\{0,1\}^N$, has maximum and minimum values $U_N=N$ and $L_N=0$.
-In this case, @schweinberger2011instability has noted that the statistic is stable (i.e., bounded $(U_N-L_N)/N=1$) and the Bernoulli model is as well when, in particular, $\ak{\thetaN}=\theta \in\mathbb{R}$ is fixed for $N \geq 1$.
+<!-- The above instability expression for the Bernoulli model follows as well from  the $N$-scaled LREP value \@ref(eq:UL) for a one-parameter exponential distribution, using that the sufficient statistic involved $g_N(\boldsymbol x)= \sum_{i=1}^N x_i$, $\boldsymbol x =(x_1,\ldots,x_n)\in\{0,1\}^N$, has maximum and minimum values $U_N=N$ and $L_N=0$.
+-->
+\ak{In this case}, @schweinberger2011instability has also noted that the statistic is stable (i.e., bounded $(U_N-L_N)/N=1$) and the Bernoulli model is as well when, in particular, $\ak{\thetaN}=\theta \in\mathbb{R}$ is fixed for $N \geq 1$.
 
@@
 Because the variable number $N\to \infty$ as the node number $n\to \infty$, both counts of 2-stars and triangles are unstable statistics in the sense of @schweinberger2011instability (i.e., $\lim_{N\to \infty} (U_N-L_N)/N=\infty$).
-Furthermore, both types of graph models are always S-unstable for all possible of parameter sequences $\ak{\thetaN} \in\mathbb{R}$ that are bounded away from zero (i.e.,  $\lim_{N\to \infty}\REP(\ak{\thetaN} )/N=\infty$ then  holds, including the fixed parameter case $ \ak{\thetaN}=\theta\in\mathbb{R}\setminus \{0\}$  from @schweinberger2011instability).
+Furthermore, both types of graph models are always S-unstable for all possible of parameter sequences $\ak{\thetaN} \in\mathbb{R}$ that are bounded away from zero (i.e.,  $\lim_{N\to \infty}\REP(\ak{\thetaN} )/N=\infty$ then  holds, including the fixed parameter case $\ak{\thetaN}=\theta\in\mathbb{R}\setminus \{0\}$  from @schweinberger2011instability).
 
@@
 To further illustrate the result in Proposition \@ref(prp:prop1), consider the multinomial distribution \@ref(eq:mod11) for $\boldsymbol X_N=(X_1,\ldots,X_N)$ having $k\geq 2$ categories $\{1,\ldots,k\}$ and $k$ parameters $\ak{\thetaN} = (\theta_{1,\ak{\thetaidx}},\ldots,\theta_{k,\ak{\thetaidx}})^\prime$.
-The variables are iid under this model so that \ak{Remark 3.1} (Section \@ref(criterion) yields the corresponding $N$-scaled  log-ratio of extreme probabilities \@ref(eq:elpr) as
+The variables are iid under this model so that \ak{Remark 3.2} (Section \@ref(criterion)) yields the corresponding $N$-scaled log-ratio of extreme probabilities \@ref(eq:elpr) as
 \begin{align*}
@@
 When the condition \@ref(eq:prop1) of Proposition \@ref(prp:prop1) is violated, this aspect suggests a potentially unstable model that may be investigated more closely.
-For example, consider the exponential graph model from \@ref(eq:mod2) involving counts of edges, 2-stars and triangles with fixed parameters $\thet_N = (\theta_{1},\theta_2,\theta_3)^\prime \in \mathbb{R}^3$ for $N\geq 1$.
+For example, consider the exponential graph model from \@ref(eq:mod2) involving counts of edges, 2-stars and triangles with fixed parameters $\thetaN = (\theta_{1},\theta_2,\theta_3)^\prime \in \mathbb{R}^3$ for $N\geq 1$.
 If either the 2-star parameter $\theta_2 \neq 0$ or triangle parameter $\theta_3 \neq 0$ is non-zero, then $\max_{1 \leq i \leq 3 } |\theta_{i}|(U_{i,N}-L_{i,N})/N \propto (n-2)\to \infty$ holds in \@ref(eq:prop1) by  $(U_{2,N}-L_{2,N})/N = 3 (U_{3,N}-L_{3,N})/N=(n-2)$ for 2-star and triangle statistics ($i=2,3$), so that Proposition \@ref(prp:prop1) hints that an unstable model may result when $|\theta_2| + |\theta_3| \neq 0$.
-Relatedly, a result from @schweinberger2011instability [Result 3] states that this model is unstable for all fixed parameters excluding cases $\theta_2 =\theta_3=0$ or $\theta_2 = - \theta_3/3$\ak{, more in line} with the instability suggested by Proposition \@ref(prp:prop1) whenever $|\theta_2| + |\theta_3| \neq 0$ (i.e., excluding $\theta_2 =\theta_3=0$).
-
-To see this, consider an even number $n>2$ of nodes and let $\boldsymbol x_0$ denote the data outcome in $\mathcal{X}^N \equiv \{0,1\}^N$ with all $N = {n \choose 2}$ edges being zero, let $\boldsymbol x_1$ denote the outcome with all edges being 1, and let $\boldsymbol x_2$ denote the edge configuration from dividing the nodes into two equal groups, with no edges within a group and all edges between the groups (so that no triangles exist in $\boldsymbol x_2$).
-Then, the $N$-scaled  log-ratio \@ref(eq:elpr) for the exponential graph model \@ref(eq:mod2) can, by definition, be bounded below by
-\begin{align*}
-\frac{1}{N}\REP_N(\ak{\thetaN}) &\geq \max_{i=1,2}\frac{1}{N}
-\left| \log\left[ \frac{P_{\ak{\thetaN}}(\boldsymbol x_i)}{P_{\ak{\thetaN}}(\boldsymbol x_0)}\right] \right| \\
-&= (n-2) \max\left\{ \left| \theta_2 + \frac{\theta_3}{3}+\frac{\theta_1}{n-2} \right|, \frac{n}{4(n-1)} \left| \theta_2 + \frac{8\theta_1}{n-2} \right| \right\};
-\end{align*}
-a similar expression also holds for an odd node number $n>2$.
-Consequently, for all fixed parameters excluding $\theta_2=\theta_3=0$, $\lim_{N\to \infty}\REP_N(\ak{\thetaN})/N=\infty$ then follows and the graph model with 2-stars and triangles is S-unstable, as suggested by the breach of Proposition \@ref(prp:prop1) for this model when $|\theta_2|+|\theta_3|\neq 0$.
-That is, instability holds even under \ak{the} $\theta_2 = - \theta_3/3$ case potentially allowed by Schweinberger's [-@schweinberger2011instability] results.
+Relatedly, a result from @schweinberger2011instability [Result 3] states that this model is unstable for all fixed parameters excluding cases $\theta_2 =\theta_3=0$ or $\theta_2 = - \theta_3/3$\ak{.
+However, more in line} with the instability suggested by Proposition \@ref(prp:prop1)\ak{, this model is unstable} whenever $|\theta_2| + |\theta_3| \neq 0$ (i.e., excluding $\theta_2 =\theta_3=0$)\ak{; Remark A.1 in the Appendix gives a proof.} That is, instability holds even under \ak{the} $\theta_2 = - \theta_3/3$ case potentially allowed by Schweinberger's [-@schweinberger2011instability] results.
+\ak{Thus, graph models of the form (\ref{eq:mod2}), with 2-stars and triangles are always S-unstable for all possible parameter sequences $\thetaN \in \mathbb{R}^3$ with $(\theta_2, \theta_3)$ bounded away from $\boldsymbol 0 \in \mathbb{R}^2$.}
 
@@
 $$
-Hence, this model sequence for $\boldsymbol X$ will be  S-unstable model if the aggregation of absolute parameters grows faster than the number $N$ of parameters/visible variables.
+Hence, this model sequence for $\boldsymbol X$ will be  S-unstable if the aggregation of absolute parameters grows faster than the number $N$ of parameters/visible variables.
 Consequently, even for \ak{the} simplest RBM model involving independence, preventing instability requires careful choice of parameters, particularly with regard to how a parameter configuration differs from zero.
@@
 \begin{enumerate}[(i)]
-\item the instability measure $\REP(\ak{\thetaN})$ for the marginal model $P_{\ak{\thetaN}}$ of  $\boldsymbol X$ satisfies
+\item the instability measure $\REP(\ak{\thetaN})$ for the marginal model $P_{\ak{\thetaN}}$ of  $\boldsymbol X$ satisfies \ak{
     $$
-    \left| \REP(\ak{\thetaN})  - \elt\right| \leq    N_{\mathcal{H}}  \log 2
-    $$
-    for
-    $$
-    \elt  \equiv   \max_{ \boldsymbol x} \max_{ \boldsymbol h  } f_{\ak{\thetaN}} (\boldsymbol x, \boldsymbol h)-\min_{ \boldsymbol x } \max_{ \boldsymbol h }f_{\ak{\thetaN}} (\boldsymbol x, \boldsymbol h)
-    $$
+    \left| \REP(\ak{\thetaN})  - \elt\right| \leq    N_{\mathcal{H}}  \log 2 \text{ for } \elt  \equiv   \max_{ \boldsymbol x} \max_{ \boldsymbol h  } f_{\ak{\thetaN}} (\boldsymbol x, \boldsymbol h)-\min_{ \boldsymbol x } \max_{ \boldsymbol h }f_{\ak{\thetaN}} (\boldsymbol x, \boldsymbol h)
+    $$}
     based on $f_{\ak{\thetaN}}$ from (\ref{eq:f}) with components $\boldsymbol x \in \{\pm 1\}^{N}, \boldsymbol h \in \{\pm 1\}^{N_{\mathcal{H}}}$.
@@
 
-In Proposition \@ref(prp:prop2)(iii), the relationships between RBM models with regard to instability, and the effects of different parameter types, follow from the bounds on model instability measures in Proposition \@ref(prp:prop2)(i)-(ii).
-Generally speaking, all instability in the marginal RBM model for the data $\boldsymbol X$ can be attributed to an excessively large model quantity $\elt$, which predominantly follows when main $\\ak{\thetaN}^{\mathcal{V}}$ and interaction $\ak{\thetaN}^{\mathcal{VH}}$ parameters related to visible variables are too large in magnitude (e.g., upon accumulation in terms such as $|\ak{\thetaN}^{\mathcal{V}}|_1$, $\Gam$ or $\Gamc$).
-For example, for any bounded sequence $|\ak{\thetaN}^{\mathcal{H}}|/N$ of hidden parameters, if main visible parameters $\ak{\thetaN}^{\mathcal{V}}$ are too extreme ($|\ak{\thetaN}^{\mathcal{V}}|_1/N\to \infty$), this aspect  will guarantee instability in the visible model ($\elt/N\to \infty$).
-In fact, the instability measure $\elt \equiv \max_{ \boldsymbol x} \max_{ \boldsymbol h  } f_{\ak{\thetaN}} (\boldsymbol x, \boldsymbol h)-\min_{ \boldsymbol x } \max_{ \boldsymbol h }f_{\ak{\thetaN}} (\boldsymbol x, \boldsymbol h)$ for marginal/visible model represents a clearly smaller portion of the  instability measure $\REP_{(\boldsymbol X, \boldsymbol H)}(\ak{\thetaN})\equiv \max_{ \boldsymbol x} \max_{ \boldsymbol h  } f_{\ak{\thetaN}} (\boldsymbol x, \boldsymbol h)-\min_{ \boldsymbol x } \min_{ \boldsymbol h } f_{\ak{\thetaN}} (\boldsymbol x, \boldsymbol h)$ in the joint RBM model, implying that an unstable marginal model (i.e., due to  $\ak{\thetaN}^{\mathcal{V}}$, $\ak{\thetaN}^{\mathcal{VH}}$) must always translate to an unstable joint model and that further potential causes of instability exist for the joint model, often due to the size $|\ak{\thetaN}^{\mathcal{H}}|_1$.
-For example, while the joint RBM model for $(\boldsymbol X,\boldsymbol H)$ must always be unstable due to a diverging combination of visible and/or interaction parameters ($|\ak{\thetaN}^{\mathcal{V}}|_1/N\to \infty$ or $\Gam/N\to \infty$) (Proposition \@ref(prp:prop2)(iii.2)), instability for the joint model can also result when the main hidden parameters $\ak{\thetaN}^{\mathcal{H}}$ become too large relative to sample size ($|\ak{\thetaN}^{\mathcal{H}}|_1/N\to \infty$ in Proposition \@ref(prp:prop2)(iii.5)).
-However, under Proposition \@ref(prp:prop2), the main hidden parameters $\ak{\thetaN}^{\mathcal{H}}$  do not necessarily entail a source of instability for the marginal visible model.
-To explain this distinction, consider a joint model where all parameters related to visibles are zero, $\ak{\thetaN}^{\mathcal{V}}= \ak{\thetaN}^{\mathcal{VH}}=\boldsymbol 0$, but the hidden-related parameters diverge in sum $|\ak{\thetaN}^{\mathcal{H}}|_1/N\to \infty$.
-Here, the explosive behavior among parameters $\ak{\thetaN}^{\mathcal{H}}$ induces instability in the joint model for $(\boldsymbol X, \boldsymbol H)$ but the marginal model for $\boldsymbol X$, however, has a perfectly stable (and in fact uniform) distribution in this case.
-When the hidden parameters are bounded relative to the sample size ($\sup_{N\geq 1} |\ak{\thetaN}^{\mathcal{H}}|_1/N<\infty$), then all instability in both the joint and marginal RBM models can be directly linked to excessively large visible $\ak{\thetaN}^{\mathcal{V}}$ and/or interaction parameters $\ak{\thetaN}^{\mathcal{VH}}$ so that features of stability/instability must be the same across both models (Proposition \@ref(prp:prop2)(iii.6)).
+\ak{With regard to instability and the effects of different parameter types, the relationships between RBM models in Proposition \ref{prp:prop2}(iii)} follow from the bounds on model instability measures in Proposition \@ref(prp:prop2)(i)-(ii).
+Generally speaking, all instability in the marginal RBM model for the data $\boldsymbol X$ can be attributed to an excessively large model quantity $\elt$, which predominantly follows when main $\ak{\thetaN}^{\mathcal{V}}$ and interaction $\ak{\thetaN}^{\mathcal{VH}}$ parameters related to visible variables are too large in magnitude (e.g., upon accumulation in terms such as $|\ak{\thetaN}^{\mathcal{V}}|_1$, $\Gam$ or $\Gamc$).
+For example, for any bounded sequence $|\ak{\thetaN}^{\mathcal{H}}|/N$ of hidden parameters, \ak{main visible parameters $\ak{\thetaN}^{\mathcal{V}}$ that are too extreme ($|\ak{\thetaN}^{\mathcal{V}}|_1/N\to \infty$) will guarantee} instability in the visible model ($\elt/N\to \infty$).
+In fact, the instability measure $\elt \equiv \max_{ \boldsymbol x} \max_{ \boldsymbol h  } f_{\ak{\thetaN}} (\boldsymbol x, \boldsymbol h)-\min_{ \boldsymbol x } \max_{ \boldsymbol h }f_{\ak{\thetaN}} (\boldsymbol x, \boldsymbol h)$ for marginal/visible model represents a clearly smaller portion of the  instability measure $\REP_{(\boldsymbol X, \boldsymbol H)}(\ak{\thetaN})\equiv \max_{ \boldsymbol x} \max_{ \boldsymbol h  } f_{\ak{\thetaN}} (\boldsymbol x, \boldsymbol h)-\min_{ \boldsymbol x } \min_{ \boldsymbol h } f_{\ak{\thetaN}} (\boldsymbol x, \boldsymbol h)$ in the joint RBM model\ak{.
+This implies} that an unstable marginal model (i.e., due to  $\ak{\thetaN}^{\mathcal{V}}$, $\ak{\thetaN}^{\mathcal{VH}}$) must always translate to an unstable joint model and that further potential causes of instability exist for the joint model, often due to the size $|\ak{\thetaN}^{\mathcal{H}}|_1$.
+\par
+
+\ak{While} the joint RBM model for $(\boldsymbol X,\boldsymbol H)$ must always be unstable due to a diverging combination of visible and/or interaction parameters ($|\ak{\thetaN}^{\mathcal{V}}|_1/N\to \infty$ or $\Gam/N\to \infty$) (Proposition \@ref(prp:prop2)(iii.2)), instability for the joint model can also result when the main hidden parameters $\ak{\thetaN}^{\mathcal{H}}$ become too large relative to sample size ($|\ak{\thetaN}^{\mathcal{H}}|_1/N\to \infty$ in Proposition \@ref(prp:prop2)(iii.5)).
+However, under Proposition \@ref(prp:prop2), the main hidden parameters $\ak{\thetaN}^{\mathcal{H}}$  do not necessarily entail a source of instability for the marginal visible model\ak{.
+%% To \ak{illustrate} this distinction, consider a joint model where all parameters related to visibles are zero, $\ak{\thetaN}^{\mathcal{V}}= \ak{\thetaN}^{\mathcal{VH}}=\boldsymbol 0$, but the hidden-related parameters diverge in sum $|\ak{\thetaN}^{\mathcal{H}}|_1/N\to \infty$.
+Here, the explosive behavior among parameters $\ak{\thetaN}^{\mathcal{H}}$ induces instability in the joint model for $(\boldsymbol X, \boldsymbol H)$\ak{.
+The} marginal model for $\boldsymbol X$, however, has a perfectly stable (and in fact uniform) distribution in this case.
+When} the hidden parameters are bounded relative to the sample size ($\sup_{N\geq 1} |\ak{\thetaN}^{\mathcal{H}}|_1/N<\infty$), then all instability in both the joint and marginal RBM models can be directly linked to excessively large visible $\ak{\thetaN}^{\mathcal{V}}$ and/or interaction parameters $\ak{\thetaN}^{\mathcal{VH}}$ so that features of stability/instability must be the same across both models (Proposition \@ref(prp:prop2)(iii.6)).
 Hence, to prevent instability in the joint model, the combined magnitudes of all parameters $\ak{\thetaN}$ must be controlled (cf.
-Proposition \@ref(prp:prop2)(iii.7)), while a stable visible data model technically results in constraining only the sizes of visible-related parameters $\ak{\thetaN}^{\mathcal{V}}$, $\ak{\thetaN}^{\mathcal{VH}}$.
-Nevertheless, because the joint model is often employed in practice for purposes of simulation and simulation-based inference, it is still reasonable to consider parameter choices for ensuring a stable joint model (and, consequently, a stable visible model as well).
+Proposition \@ref(prp:prop2)(iii.7)), while a stable visible data model technically results \ak{from} constraining only the sizes of visible-related parameters $\ak{\thetaN}^{\mathcal{V}}$, $\ak{\thetaN}^{\mathcal{VH}}$.
+Nevertheless, because the joint model \ak{often is} employed in practice for purposes of simulation and simulation-based inference, it is still reasonable to consider parameter choices for ensuring a stable joint model (and, consequently, a stable visible model as well).
 Further evidence of this is seen in the following numerical example.
-\par
 
@@
 
+\par
+
 In our numerical experiment, we allow the two types of terms (main effects terms corresponding to visible and hidden parameters $\boldsymbol \theta_{main} = (\ak{\thetaN}^{\mathcal{V}}, \ak{\thetaN}^{\mathcal{H}})$ and interaction parameters $\ak{\thetaN}^{\mathcal{VH}}$) to have varying average magnitudes, $||\boldsymbol \theta_{main} || /(N_{\mathcal{H}}+N_{\mathcal{V}})$ and $||\boldsymbol \theta_{interaction} || /(N_{\mathcal{H}}*N_{\mathcal{V}})$ for a RBM with $N_\mathcal{V} = `r V`$ visibles and $N_\mathcal{H} = `r H`$ hiddens.
@@
 From these two plots, it is clear that for larger magnitudes of the parameter vectors, there is evidence of S-instability in that the log-ratio of extremal probabilities scaled by $N_{\mathcal{V}}$ and the the biggest log-probability ratio for a one-component change in data outcomes are both increasing away from $\ak{\thetaN} = \boldsymbol 0$, further supporting \@ref(prp:prop2)(iii.2 and iii.5).
+\par
 
@@
 
-Due to \ak{induced sensitivity in the probability structure of FOES models, S-instability} may often translate to numerical complications, and in fact obstructions, in both simulation and statistical inference based on likelihoods.
-We describe these aspects in Sections \@ref(mcmc)-\@ref(bayes) with regard to data simulation, maximum likelihood estimation and Bayes inference, respectively.
+Due to \ak{sensitivity in the probability structure of an unstable FOES model, S-instability} may often translate to numerical complications, and in fact obstructions, in both simulation and statistical inference based on likelihoods.
+We describe these aspects in Sections \@ref(ml)-\@ref(bayes) with regard to data simulation, maximum likelihood estimation and Bayes inference.
 
-## Implications for Simulation {#mcmc}
+<!-- ## Implications for Simulation {#mcmc} -->
 
-Suppose one aims to apply MCMC to simulate data $\boldsymbol X=(X_1,\ldots,X_n)$ from a FOES model $P_{\ak{\thetaN}}(\boldsymbol x)$, $\boldsymbol x \in\mathcal{X}^N$ whereby a chain is constructed with $P_{\ak{\thetaN}}$ as the stationary distribution.
-For an unstable FOES model, one-component changes in outcomes may produce
-radically different probabilities, which may then impose numerical barriers to MCMC.
+<!-- Suppose one aims to apply MCMC to simulate data $\boldsymbol X=(X_1,\ldots,X_n)$ from a FOES model $P_{\ak{\thetaN}}(\boldsymbol x)$, $\boldsymbol x \in\mathcal{X}^N$ whereby a chain is constructed with $P_{\ak{\thetaN}}$ as the stationary distribution.
+For an unstable FOES model, one-component changes in outcomes may produce radically different probabilities, which may then impose numerical barriers to MCMC.
 For example, consider implementation of a Gibbs sampler from the full conditional distributions $P_{\ak{\thetaN}} (X_i = x| \boldsymbol x_{-i})$, $x\in\mathcal{X}$, of each variable $X_i$ based on values $\boldsymbol x_{-i}\in \mathcal{X}^{N-1}$ for the remaining variables, say $\boldsymbol X_{-i}$, in $\boldsymbol X$.
-If a single change in $X_i$ from one value $x_1$ to another $x_2$ may produce two outcomes $\boldsymbol x^{(1)}$ and $\boldsymbol x^{(2)}$ for $\boldsymbol X$ with vastly different probabilities under the joint distribution $P_{\ak{\thetaN}}$, then the Gibbs sampler can have extreme log-ratios in its transition probabilities,
-$$
-\log \left| \frac{P_{\ak{\thetaN}} (X_i =x_1 | \boldsymbol x^{(1)}_{-i})}{P_{\ak{\thetaN}} (X_i = x_2| \boldsymbol x^{(2)}_{-i}) } \right|=\log \left| \frac{P_{\ak{\thetaN}} ( \boldsymbol x^{(1)} )}{P_{\ak{\thetaN}} (  \boldsymbol x^{(2)} ) } \right|,
-$$  
-as conditional probabilities are proportional to joint probabilities that, with unstable models, can have unbounded log-probability ratios in one-component changes (Theorem
-\@ref(thm:instab-elpr)).
+If a single change in $X_i$ from one value $x_1$ to another $x_2$ may produce two outcomes $\boldsymbol x^{(1)}$ and $\boldsymbol x^{(2)}$ for $\boldsymbol X$ with vastly different probabilities under the joint distribution $P_{\ak{\thetaN}}$, then the Gibbs sampler can have extreme log-ratios in its transition probabilities,  -->
+<!-- $$ -->
+<!-- \log \left| \frac{P_{\ak{\thetaN}} (X_i =x_1 | \boldsymbol x^{(1)}_{-i})}{P_{\ak{\thetaN}} (X_i = x_2| \boldsymbol x^{(2)}_{-i}) } \right|=\log \left| \frac{P_{\ak{\thetaN}} ( \boldsymbol x^{(1)} )}{P_{\ak{\thetaN}} (  \boldsymbol x^{(2)} ) } \right|, -->
+<!-- $$   -->
+<!-- as conditional probabilities are proportional to joint probabilities that, with unstable models, can have unbounded log-probability ratios in one-component changes (Theorem -->
+<!-- \@ref(thm:instab-elpr)).
 This can hinder the ability of a chain to effectively explore the sample space of the observations $\boldsymbol X$, as the chain may mix poorly by moving rapidly to, and slowly away from, sections of the sample space.
@@
 This result is in line with conclusions of @handcock2003assessing and @schweinberger2011instability for other exponential models.
+-->
 
-## Implications for Maximum Likelihood Inference
+## Implications for Maximum Likelihood Inference {#ml}
 
@@
 Considering that the degeneracy issues related to unstable models can cause likelihoods can be flat (e.g., near zero) for many parameters under a given data outcome and that sign changes in parameters can shift tremendous probability to extreme and opposite outcomes in the sample space (e.g., Corollary \@ref(cor:sign), Theorem \@ref(thm:something)), then numerical complications may arise with Bayes inference in sampling a posterior parameter space based on MCMC.
-The potential challenges in chain mixing are similar to those presented in Section \@ref(mcmc), though in chain movements through the parameter space as opposed to the sample space in data generation.
-That is, in the Bayes setting for sampling a posterior distribution for $\ak{\thetaN}$,  a chain may \ak{be unable} to effectively explore the parameter space due partly to extreme and potentially unbounded probability ratios from parameter sign changes, which represents a parameter space analog to how one-component changes in the sample space may impact data simulation with unstable models.
+<!-- The potential challenges in chain mixing are similar to those presented in Section \@ref(mcmc), though in chain movements through the parameter space as opposed to the sample space in data generation.
+-->
+\ak{Instability may hinder effective chain mixing, related to the findings of \citet{handcock2003assessing} and \citet{schweinberger2011instability} with exponential graph models.
+For example, a Markov chain may become entrapped within a mode, and fail to adequately mimic the overall occupation frequencies required for a reasonable MCMC distributional approximaton.
+If modes of the unstable model are not unique, then important outcomes may be missed without multiple chains or impractically enormous numbers of MCMC samples.
+This mixing problem is due to the unstable stationary distribution (unbounded ratios of probabilities under the joint model), rather than in any particulars of the MCMC algorithm.
+Hence,} in the Bayes setting for sampling a posterior distribution for $\ak{\thetaN}$,  a chain may \ak{be unable} to effectively explore the parameter space due partly to extreme and potentially unbounded probability ratios from parameter sign changes.
+<!-- , which represents a parameter space analog to how one-component changes in the sample space may impact data simulation with unstable models.  -->
+
 For example, if $\pi(\cdot)$ denotes a prior density for $\ak{\thetaN}$ and $q(\cdot | \cdot)$ denotes a proposal distribution  for use in a Metropolis-Hastings (MH) sampler, then MH acceptance probability becomes
@@
 $$
-which indicates how parameter sensitivity in the likelihood $P_{\ak{\thetaN}} (  \boldsymbol x)$ may complicate sampling of the posterior $P_{ \ak{\thetaN}} (  \boldsymbol x) \pi(\ak{\thetaN})$ (i.e., moving from $\boldsymbol \theta^{(1)}$ to $\boldsymbol \theta^{(2)}$ in the parameter space).
-\ak{With this in mind, the instability results presented in Section \ref{instability-results} point to the potential use of regularization and penalization as a potential solution to avoid instability in FOES models.
-One example of a principled approach to penalization can be found in \cite{kaplan2019properties}, which approaches Bayesian fitting of a single layer RBM using priors to penalize the parameter values.} Furthermore, the potential for model instability and the size of the parameter space can also become greater with the introduction of latent variables to existing data variables, as involved in some model formulations described in Sections \@ref(rbm)-\@ref(deep-learning).
+which indicates how parameter sensitivity in the likelihood $P_{\ak{\thetaN}} (\boldsymbol x)$ may complicate sampling of the posterior $P_{ \ak{\thetaN}} (  \boldsymbol x) \pi(\ak{\thetaN})$ (i.e., moving from $\boldsymbol \theta^{(1)}$ to $\boldsymbol \theta^{(2)}$ in the parameter space).
+Furthermore, the potential for model instability and the size of the parameter space can also become greater with the introduction of latent variables to existing data variables, as involved in some model formulations described in Sections \@ref(rbm)-\@ref(deep-learning).
 As latent variables are often sampled with parameters in a Bayes MCMC approach, this aspect may further compound numerical problems in chain mixing.
+
+## Fitting Stable Models {#stable-models}
+
+\ak{The instability results presented in Section \ref{instability-results} also suggest the potential use of regularization and penalization as a solution to avoid instability when fitting FOES models.
+One example of a rigorous approach to penalization can be found in \citet{kaplan2019properties}, which employs Bayesian fitting for a single layer RBM using multivariate Gaussian priors with constrained covariance matrices to control the sizes of the parameter values.
+For example, considering a latent variable model described in Section \ref{fixed-dim-exp}, Proposition \ref{prp:prop2}(iii) suggests how parameter contributions from sources of ``main hidden," ``main visible," and ``interaction" parameters may each be constrained (relative to the visible variable sample size $N$) to ensure model stability.
+In a Bayes inference framework, priors can be developed so that parameters are then appropriately bounded (in a stochastic sense) to lie in a parameter subset compatible with stability.
+For example, for a latent model with numbers $N_{\mathcal{H}}$, $N_{\mathcal{V}}$ and $N_{\mathcal{VH}}$ for hidden, visible and interaction parameters, respectively, Proposition \ref{prp:prop2}(iii.7) suggests a potential prior specification where parameters have root second moments bounded by the parameter type: $C N/N_{\mathcal{V}}$, $C N/N_{\mathcal{H}}$, and $C N/N_{\mathcal{VH}}$ for visible, hidden and interactions parameters with a constant $C>0$ (i.e., so that the condition in Proposition \ref{prp:prop2}(iii.7) holds in a stochastic sense).
+The same general strategy can apply with other models with results from Section \ref{illustrations}, and the adequacy of the subsequent fitted model might then be assessed.
+
+As opposed to approaches for fitting (e.g., penalization as above),  the findings here on model stability/instability also support alternative strategies to model formulation and re-pameterization using ``centered conditional" distributions \citep[cf.][]{kaiser2007statistical}; see \citet{kaiser2009exploring} and \citet{kaiser2012centered} for applications with spatial data and \citet{casleton2017local} for network data.
+In a joint model $P_{\thetaN}$,  these authors consider associated (full) conditional distributions $P_{\thetaN}(X_{i} = x_i | X_j=x_j, j\neq i)$ and re-parameterize by centering/averaging sums in the expression of  $P_{\thetaN}(X_{i} = x_i | X_j=x_j, j\neq i)$ in order to separate model effects between ``mean" and ``dependence" parameters.
+The intent is to improve model interpretation and help detect degeneracy (characterized by the ``mean parameters" in the model failing to match averages observed in data generations from the model and attributable to large ``dependence parameters").
+The works above describe advantages of such centered parameterizations based on empirical studies, but findings here also support these parameterizations as useful for avoiding model instability, where the corresponding ``centered" models can be checked for $\REP(\thetaN)/N$ ratios that are bounded over wider possible parameters than ``uncentered" counterpart models.
+For illustration, the appendix provides details on a graph model example.}
 
@@
 For a large class of models that covers a broad range of applications (including "deep learning"), we have developed a formal definition of instability in model probability structure and elucidated multiple consequences of instability.
-We have shown for FOES models that instability manifests through small changes in data leading to potentially large changes in probability as well as the potential to place all probability on certain modal subsections of the sample space, which potentially could be small.
+We have shown for FOES models that instability manifests through \ak{a large amount of probability being placed on a subset of the theoretically possible realizations that may be narrower than intended.}
+<!-- small changes in data leading to potentially large changes in probability as well as the potential to place all probability on certain modal subsections of the sample space, which potentially could be small.
+-->
 Such instability is often due to a complex interaction between the model statistics used  (i.e., how numerous and large these may become) and the number and magnitudes of parameters in the model formulation.
 For many FOES models, the possibility exists, at least in principle, to \ak{constrain} parameters in a way \ak{that} balances their potential contributions against those of model statistics in order to prevent probability instabilities.
-The FOES model class is quite broad and, in developing such models for large data sets, some caution should be used in parameter specification to control effects of model instability.
+\ak{While the results on instability here only address model formulations with finite sample spaces, the same underlying principles might be extended to models for continuous data as well by discretizing the model (i.e., binning outcomes into intervals), as data typically have some given level of discretization.
+Such a generalization could help toward understanding potential model instability in a wider class of data applications.
+As it sits currently, t}he FOES model class is quite broad and \ak{findings here can help in identifying undesirable probability features in such models.}

### 2019-06-18 writing/paper-note/docs/note.tex
revision of note

@@
 \newcommand{\DN}{\Delta_N}
+\newcommand{\thetaidx}{q(N)}
+\newcommand{\thetaN}{\boldsymbol \theta_{\thetaidx}}
 
-\newcommand{\ma}{\mathrm{max}_{\boldsymbol \theta_N}}
-\newcommand{\mi}{\mathrm{min}_{\boldsymbol \theta_N}}
+\newcommand{\ma}{\mathrm{max}_{\thetaN}}
+\newcommand{\mi}{\mathrm{min}_{\thetaN}}
 
-\newcommand{\maa}{\mathrm{max}_{-\boldsymbol \theta_N}}
-\newcommand{\mii}{\mathrm{min}_{-\boldsymbol \theta_N}}
+\newcommand{\maa}{\mathrm{max}_{-\thetaN}}
+\newcommand{\mii}{\mathrm{min}_{-\thetaN}}
 
@@
 
-\newcommand{\elt}{A_{N}(\boldsymbol \theta_N) }
-\newcommand{\Gam}{B_{N}(\boldsymbol \theta_N) }
+\newcommand{\elt}{A_{N}(\thetaN) }
+\newcommand{\Gam}{B_{N}(\thetaN) }
 
-\newcommand{\Gamc}{C_{N}(\boldsymbol \theta_N) }
-\newcommand{\Gamt}{\Gamma_{N,2}(\boldsymbol \theta_N) }
+\newcommand{\Gamc}{C_{N}(\thetaN) }
+\newcommand{\Gamt}{\Gamma_{N,2}(\thetaN) }
 
@@
 {A probability model exhibits instability if small changes in a data outcome result in large, and often unanticipated, changes in probability.
-This instability is a property of the probability \ak{model, prescribed by a probability distributional form evaluated at a given configuration of parameters}.
+This instability is a property of the probability \ak{model, given by a distributional form and a given configuration of parameters}.
 For correlated data structures found in several application areas, there is increasing interest in \ak{identifying} such sensitivity in model probability structure.
@@
 That is, \(\mathcal{X}\) with \(|\mathcal{X}| < \infty\) denotes the set of potential outcomes for each single variable \(X_i\), so that the product space \(\mathcal{X}^N\) corresponds to values for the variables \(\boldsymbol X_N=(X_1,\ldots,X_N)\).
-For each \(N\), let \(P_{\boldsymbol \theta_N}\) denote a probability model on \(\mathcal{X}^N\), under which \(P_{\boldsymbol \theta_N}(x_1, \dots, x_N) > 0\) is the probability of the data outcome \((x_1, \dots, x_N) \in \mathcal{X}^N\).
-In this, we assume that the model support of \(P_{\boldsymbol \theta_N}\) is the sample space \(\mathcal{X}^N\).
-This framework produces probability models \(P_{\boldsymbol \theta_N}\), indexed by a \ak{defining} sequence of parameters \(\boldsymbol \theta_N\), to describe data \(\boldsymbol X_N\) of any given sample size \(N \geq 1\).
+For each \(N\), let \(P_{\ak{\thetaN}}\) denote a probability model on \(\mathcal{X}^N\), under which \(P_{\ak{\thetaN}}(x_1, \dots, x_N) > 0\) is the probability of the data outcome \((x_1, \dots, x_N) \in \mathcal{X}^N\).
+In this, we assume that the model support of \(P_{\ak{\thetaN}}\) is the sample space \(\mathcal{X}^N\).
+This framework produces probability models \(P_{\ak{\thetaN}}\), indexed by a \ak{defining} sequence of parameters \(\ak{\thetaN}\), to describe data \(\boldsymbol X_N\) of any given sample size \(N \geq 1\).
 For simplicity, we will refer to this distributional class as \emph{Finite Outcome Everywhere Supported (FOES)} models in the following.
-The dimension and structure of such parameters are generic, without restriction, though natural cases will be seen to include those where \(\boldsymbol \theta_N \in \mathbb{R}^{q(N)}\) for some arbitrary integer-valued function \(q(\cdot) \geq 1\).
+The dimension and structure of such parameters are generic, without restriction, though natural cases will be seen to include those where \(\ak{\thetaN} \in \mathbb{R}^{q(N)}\) for some arbitrary integer-valued function \(q(\cdot) \geq 1\).
 
@@
 Section \ref{implications} then emphasizes the implications of such model instability, showing that such impropriety can be expected to numerically hinder maximum likelihood estimation and MCMC-based simulations.
-As one potential remedy, suggestions are given for constraining model parameterizations to avoid the most problematic regions of the parameter space.
+As one potential remedy, suggestions are given for constraining model parameterizations to avoid the most problematic regions of the parameter space \ak{(see Section \ref{bayes} for further discussion)}.
 Proofs of the main results appear in Appendix \ref{appendix-instab}.
@@
 \end{equation}
-depending on parameter vector \(\boldsymbol \theta \in \Theta_N \subset \mathbb{R}^{k}\) and natural parameter function \(\boldsymbol \eta : \mathbb{R}^k \mapsto \mathbb{R}^L\) with fixed positive integers \(k\) and \(L\) denoting their dimensions.
+depending on parameter vector \(\boldsymbol \theta \in \Theta_{\ak{\thetaidx}} \subset \mathbb{R}^{k}\) and natural parameter function \(\boldsymbol \eta : \mathbb{R}^k \mapsto \mathbb{R}^L\) with fixed positive integers \(k\) and \(L\) denoting their dimensions.
 Above, \(\boldsymbol g_N : \mathcal{X}^N \mapsto \mathbb{R}^L\) is a vector of sufficient statistics, while
 \[
-\psi(\boldsymbol \theta) = \log \sum\limits_{\boldsymbol x \in \mathcal{X}^N}\exp\left[\boldsymbol \eta^T(\boldsymbol \theta) \boldsymbol g_N(\boldsymbol x) \right], \qquad \boldsymbol \theta \in \Theta_N\equiv \{\boldsymbol \theta \in \mathbb{R}^k : \psi(\boldsymbol \theta) < \infty \},
+\psi(\boldsymbol \theta) = \log \sum\limits_{\boldsymbol x \in \mathcal{X}^N}\exp\left[\boldsymbol \eta^T(\boldsymbol \theta) \boldsymbol g_N(\boldsymbol x) \right], \qquad \boldsymbol \theta \in \Theta_\ak{\thetaidx}\equiv \{\boldsymbol \theta \in \mathbb{R}^k : \psi(\boldsymbol \theta) < \infty \},
 \]
-denotes the normalizing function with parameter space \(\Theta_N\).
+denotes the normalizing function with parameter space \(\Theta_{\ak{\thetaidx}}\).
 The natural parameter function \(\eta (\boldsymbol \theta)\) has a linear form (i.e., \(\eta (\boldsymbol \theta)= \bm{A} \boldsymbol \theta\) for a given \(L \times k\) matrix \(\bm{A}\)) in many common model formulations, though may also be nonlinear (e.g., curved exponential families).
@@
 
-Such discrete \ak{exponential models} are special cases of the FOES models, as seen by defining \(P_{\boldsymbol \theta_N}(\boldsymbol x)\equiv p_{N,\boldsymbol \theta_N}(\boldsymbol x)> 0\), \(\boldsymbol x \in \mathcal{X}^N\), based on \eqref{eq:expo} and a parameter sequence \(\boldsymbol \theta_N \in \Theta_N \subset \mathbb{R}^k\).
+Such discrete \ak{exponential models} are special cases of the FOES models, as seen by defining \(P_{\ak{\thetaN}}(\boldsymbol x)\equiv p_{N, \ak{\thetaN}}(\boldsymbol x)> 0\), \(\boldsymbol x \in \mathcal{X}^N\), based on \eqref{eq:expo} and a parameter sequence \(\ak{\thetaN} \in \Theta_{\ak{\thetaidx}} \subset \mathbb{R}^k\).
 For example, if observations \(\boldsymbol X = (X_1,\ldots,X_N)\) correspond to \(N\) independent and identically distributed Bernoulli random variables, each indicating a binary \(0\)-\(1\) outcome, the resulting probabilities have exponential form \eqref{eq:expo} given by
@@
 \label{eq:mod1}
-P_{\boldsymbol \theta_N}(\boldsymbol x) \propto
- \exp\left[\boldsymbol \theta_N \sum_{i=1}^N x_i\right], \qquad \boldsymbol x=(x_1,\ldots,x_N) \in\{0,1\}^N, 
+P_{\ak{\thetaN}}(\boldsymbol x) \propto
+ \exp\left[\ak{\thetaN} \sum_{i=1}^N x_i\right], \qquad \boldsymbol x=(x_1,\ldots,x_N) \in\{0,1\}^N, 
  \end{equation}
-with sufficient statistic \(\boldsymbol g_N(\boldsymbol x)\equiv \sum_{i=1}^N x_i\) and ``log odds ratio'' parameter \(\boldsymbol \theta_N \equiv \log[ P_{\boldsymbol \theta_N}(X_i=1)/P_{\boldsymbol \theta_N}(X_i=0) ] \in \mathbb{R}\).
+with sufficient statistic \(\boldsymbol g_N(\boldsymbol x)\equiv \sum_{i=1}^N x_i\) and ``log odds ratio'' parameter \(\ak{\thetaN} \equiv \log[ P_{\ak{\thetaN}}(X_i=1)/P_{\ak{\thetaN}}(X_i=0) ] \in \mathbb{R}\).
 More generally, supposing \(\boldsymbol X =(X_1,\ldots,X_N)\) represent \(N\) independent trials, each assuming an outcome \(\{1,\ldots,k\}\) among \(k\) possibilities (e.g., a die roll), a multinomial distribution is given by
@@
 \label{eq:mod11}
-P_{\boldsymbol \theta_N}(\boldsymbol x) \propto  \exp\left[  \boldsymbol \theta_{N}^T g_N(\boldsymbol x)   \right] =
-\exp\left[ \sum_{j=1}^k {\theta_{j,N}} \sum_{i=1}^N \mathbb{I}(x_i=j) \right], \qquad \boldsymbol x  \in\{1,\ldots,k\}^N, 
+P_{\ak{\thetaN}}(\boldsymbol x) \propto  \exp\left[  \ak{\thetaN}^T g_N(\boldsymbol x)   \right] =
+\exp\left[ \sum_{j=1}^k {\theta_{j,\ak{\thetaidx}}} \sum_{i=1}^N \mathbb{I}(x_i=j) \right], \qquad \boldsymbol x  \in\{1,\ldots,k\}^N, 
 \end{equation}
-with sufficient statistic \(\boldsymbol g_N(\boldsymbol x)\) involving a count \(\sum_{i=1}^N \mathbb{I}(x_i=j)\) for each outcome \(j \in \{1,\ldots,k\}\), where \(\mathbb{I}(\cdot)\) denotes the indicator function, and parameters \(\boldsymbol \theta_N=(\theta_{1,N},\ldots,\theta_{k,N})\in\mathbb{R}^k\) defining log-probability ratios \(\theta_{i,N}-\theta_{j,N} =\log [P_{\boldsymbol \theta_N}(X_1=i)/P_{\boldsymbol \theta_N}(X_1=j)]\).
+with sufficient statistic \(\boldsymbol g_N(\boldsymbol x)\) involving a count \(\sum_{i=1}^N \mathbb{I}(x_i=j)\) for each outcome \(j \in \{1,\ldots,k\}\), where \(\mathbb{I}(\cdot)\) denotes the indicator function, and parameters \(\ak{\thetaN}=(\theta_{1,\ak{\thetaidx}},\ldots,\theta_{k,\ak{\thetaidx}})\in\mathbb{R}^k\) defining log-probability ratios \(\theta_{i,\ak{\thetaidx}}-\theta_{j,\ak{\thetaidx}} =\log [P_{\ak{\thetaN}}(X_1=i)/P_{\ak{\thetaN}}(X_1=j)]\).
 In addition to such standard models for discrete independent data, exponential models of FOES type commonly arise with dependent spatial data \citep{besag1974spatial} and network/relational data \citep{wasserman1994social, handcock2003assessing}.
-For a random graph or network with, say, \(n\) nodes, consider \(N={n \choose 2}\) random edges where the \(i\)th edge is associated with a pair of nodes \(s_i \equiv \{v_i,u_i\}\) and a binary variable \(X_i\in\{0,1\}\) indicating presence/absence of an edge among the node pair \(s_i\), \(i=1,\ldots,N\).
+For a random graph or network with, say, \(n\) nodes \ak{$\{v_i\}_{i = 1}^n$}, consider \(N={n \choose 2}\) random edges where the \(i\)th edge is associated with a pair of nodes \(s_i \equiv \{v_{i_1},v_{i_2}\}\) and a binary variable \(X_i\in\{0,1\}\) indicating presence/absence of an edge among the node pair \(s_i\), \(i=1,\ldots,N\).
 Here the length \(N\) of the edge variable sequence \(\boldsymbol X = (X_1,\ldots,X_N)\) increases as a function of node number \(n\) and corresponding exponential models often incorporate graph topographical features derived from \(\boldsymbol X\).
@@
 \label{eq:mod2}
-P_{\boldsymbol \theta_N}(\boldsymbol x) \propto
- \exp\left[\sum_{j=1}^3 \theta_{j,N} g_{j,N}(\boldsymbol x)\right], \quad\qquad \boldsymbol x=(x_1,\ldots,x_N)  \in\{0,1\}^N,
+P_{\ak{\thetaN}}(\boldsymbol x) \propto
+ \exp\left[\sum_{j=1}^3 \theta_{j,\ak{\thetaidx}} g_{j,N}(\boldsymbol x)\right], \quad\qquad \boldsymbol x=(x_1,\ldots,x_N)  \in\{0,1\}^N,
 \end{equation}
@@
 \]
-involving the numbers of edges, 2-stars and triangles among an outcome \(\boldsymbol x\) given by \(g_{1,N}(\boldsymbol x)\), \(g_{2,N}(\boldsymbol x)\) and \(g_{3,N}(\boldsymbol x)\), respectively, along with \(k=3\) real parameters \(\boldsymbol \theta_N \equiv (\theta_{1,N},\theta_{2,N},\theta_{3,N})\).
-For this network model \eqref{eq:mod2} in particular, as well as for more general models of form \eqref{eq:expo}, \citet{schweinberger2011instability} considered instability in such exponential models with sequences of fixed parameters \(\boldsymbol \theta_N = (\theta_1,\ldots,\theta_k)\in\mathbb{R}^k\), \(N \geq 1\), of fixed dimension \(k\).
+involving the numbers of edges, 2-stars and triangles among an outcome \(\boldsymbol x\) given by \(g_{1,N}(\boldsymbol x)\), \(g_{2,N}(\boldsymbol x)\) and \(g_{3,N}(\boldsymbol x)\), respectively, along with \(k=3\) real parameters \(\ak{\thetaN} \equiv (\theta_{1,\ak{\thetaidx}},\theta_{2,\ak{\thetaidx}},\theta_{3,\ak{\thetaidx}})\).
+For this network model \eqref{eq:mod2} in particular, as well as for more general models of form \eqref{eq:expo}, \citet{schweinberger2011instability} considered instability in such exponential models with sequences of fixed parameters \(\ak{\thetaN} = (\theta_1,\ldots,\theta_k)\in\mathbb{R}^k\), \(N \geq 1\), of fixed dimension \(k\).
 
-For model sequences \(P_{\boldsymbol \theta_N}(\boldsymbol x)\equiv p_{N,\boldsymbol \theta_N}(\boldsymbol x)\) of the exponential type \eqref{eq:expo}, such as those in \eqref{eq:mod1}-\eqref{eq:mod2}, note that the dimension \(k\) of the parameter \(\boldsymbol \theta_N\in\Theta \subset \mathbb{R}^k\) necessarily remains the same for all sample sizes \(N \geq 1\) as the form of the natural parameter function \(\eta(\cdot)\) in \eqref{eq:expo} and the number of sufficient statistics \(\boldsymbol g_{N}(\boldsymbol x)\) do not depend on \(N\).
-Consequently, \(\boldsymbol \theta_N\) lies in a parameter space of fixed Euclidean dimension \(k\).
+For model sequences \(P_{\ak{\thetaN}}(\boldsymbol x)\equiv p_{N,\ak{\thetaN}}(\boldsymbol x)\) of the exponential type \eqref{eq:expo}, such as those in \eqref{eq:mod1}-\eqref{eq:mod2}, note that the dimension \(k\) of the parameter \(\ak{\thetaN}\in\Theta \subset \mathbb{R}^k\) necessarily remains the same for all sample sizes \(N \geq 1\) as the form of the natural parameter function \(\eta(\cdot)\) in \eqref{eq:expo} and the number of sufficient statistics \(\boldsymbol g_{N}(\boldsymbol x)\) do not depend on \(N\).
+Consequently, \(\ak{\thetaN}\) lies in a parameter space of fixed Euclidean dimension \(k\).
 However, this aspect need not be true for other types of FOES models considered in Sections \ref{rbm} - \ref{deep-learning}, where instead the numbers of parameters and sufficient statistics commonly increase with the sample size \(N\).
@@
 Let \(\boldsymbol X = (X_1,\ldots,X_N)\) denote the \(N\) random variables for visibles with support \(\mathcal{X}^N\) and \(\boldsymbol H = (H_1,\ldots,H_{N_\mathcal{H}})\) denote the \(N_\mathcal{H}\) random variables for hiddens with support \(\mathcal{X}^{N_\mathcal{H}}\) where \(\mathcal{X} = \{-1,1\}\).
-For \ak{specified} parameters \(\boldsymbol \theta_N^{\mathcal{H}} \in \mathbb{R}^{N_\mathcal{H}}\), \(\boldsymbol \theta_N^{\mathcal{V}}\in \mathbb{R}^N\), and \(\boldsymbol \theta_N^{\mathcal{HV}}\) as a real matrix with dimension \(N_\mathcal{H} \times N\), the RBM model for \(\tilde{\boldsymbol X}=(\boldsymbol X,\boldsymbol H)\) has the joint probability mass function
+For \ak{specified} parameters \(\ak{\thetaN}^{\mathcal{H}} \in \mathbb{R}^{N_\mathcal{H}}\), \(\ak{\thetaN}^{\mathcal{V}}\in \mathbb{R}^N\), and \(\ak{\thetaN}^{\mathcal{HV}}\) as a real matrix with dimension \(N_\mathcal{H} \times N\), the RBM model for \(\tilde{\boldsymbol X}=(\boldsymbol X,\boldsymbol H)\) has the joint probability mass function
 \begin{equation}
 \label{eq:RBM1}
-\tilde{P}_{\boldsymbol \theta_N} (\tilde{\boldsymbol x}) = \exp\left[ (\boldsymbol \theta_N^{\mathcal{H}})^T \boldsymbol h + \boldsymbol (\boldsymbol \theta_N^{\mathcal{V}})^T \boldsymbol x + \boldsymbol h^T  \boldsymbol\theta_N^{\mathcal{HV}} \boldsymbol x - \psi(\boldsymbol \theta_N)\right], \quad \tilde{\boldsymbol x} = (\boldsymbol x, \boldsymbol h) \in \{\pm 1\}^{N+N_\mathcal{H}}
+\tilde{P}_{\ak{\thetaN}} (\tilde{\boldsymbol x}) = \exp\left[ (\ak{\thetaN}^{\mathcal{H}})^T \boldsymbol h + \boldsymbol (\ak{\thetaN}^{\mathcal{V}})^T \boldsymbol x + \boldsymbol h^T  \ak{\thetaN}^{\mathcal{HV}} \boldsymbol x - \psi(\ak{\thetaN})\right], \quad \tilde{\boldsymbol x} = (\boldsymbol x, \boldsymbol h) \in \{\pm 1\}^{N+N_\mathcal{H}}
 \end{equation}
-with normalizing function
-\[
-\psi(\boldsymbol \theta_N) = \log \sum_{\tilde{\boldsymbol x} \in \{\pm 1\}^{N+N_H} } \exp\left[ (\boldsymbol \theta_N^{\mathcal{H}})^T \boldsymbol h + \boldsymbol (\boldsymbol \theta_N^{\mathcal{V}})^T \boldsymbol x + \boldsymbol h^T  \boldsymbol\theta_N^{\mathcal{HV}} \boldsymbol x\right].
-\]
-Let \(\boldsymbol \theta_N = (\boldsymbol \theta_N^{\mathcal{H}}, \boldsymbol \theta_N^{\mathcal{V}}, \boldsymbol\theta_N^{\mathcal{HV}} ) \in \Theta_N \equiv \mathbb{R}^{q(N)}\), with \(q(N) = N + N_\mathcal{H} + N*N_\mathcal{H}\), denote \ak{a given} parameter vector for the RBM, as indexed by the number \(N\) of visible random variables (which may differ from the actual lengths of these parameter vectors).
+with \ak{normalizing function $\psi(\ak{\thetaN})$}.
+Let \(\ak{\thetaN} = (\ak{\thetaN}^{\mathcal{H}}, \ak{\thetaN}^{\mathcal{V}}, \ak{\thetaN}^{\mathcal{HV}} ) \in \Theta_\ak{\thetaidx} \equiv \mathbb{R}^{q(N)}\), with \(q(N) = N + N_\mathcal{H} + N*N_\mathcal{H}\), denote \ak{a given} parameter vector for the RBM, as indexed by the number \(N\) of visible random variables (which may differ from the actual lengths of these parameter vectors).
 The probability mass function for the visible variables \(\boldsymbol X = (X_1, \dots, X_N)\) follows from marginalizing the joint specification to yield
@@
 \label{eq:RBM2}
-P_{\boldsymbol \theta_N} (\boldsymbol x) = \sum\limits_{\boldsymbol h \in \{\pm 1\}^{N_{\mathcal{H}}}} \tilde{P}_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h), \qquad \boldsymbol x \in \{\pm 1\}^{N}\equiv \mathcal{X}^N.
+P_{\ak{\thetaN}} (\boldsymbol x) = \sum\limits_{\boldsymbol h \in \{\pm 1\}^{N_{\mathcal{H}}}} \tilde{P}_{\ak{\thetaN}} (\boldsymbol x, \boldsymbol h), \qquad \boldsymbol x \in \{\pm 1\}^{N}\equiv \mathcal{X}^N.
 \end{equation}
-Here the baseline model \eqref{eq:RBM1} for hidden/visible variables is a linear exponential one in sufficient statistics \((\tilde{\boldsymbol X}, \boldsymbol X^T\boldsymbol H)\) using \(\tilde{\boldsymbol X}=(\boldsymbol X,\boldsymbol H)\) from \eqref{eq:RBM1}, but the form differs from the previous exponential models in \eqref{eq:expo} in that the lengths of parameters \(\boldsymbol \theta_N\) and statistics \((\tilde{\boldsymbol X}, \boldsymbol X^T\boldsymbol H)\) increase to incorporate more visible variables.
-That is, in contrast to \eqref{eq:expo}, the natural parameter function involved in the RBM model \eqref{eq:RBM1}, as the identity mapping of the parameters \(\boldsymbol \theta_N\in\mathbb{R}^{q(N)}\), naturally grows in dimension \(q(N)\to \infty\) to accommodate visible variables \(X_1, \dots, X_N\) of increasing sample size \(N\to \infty\).
+Here the baseline model \eqref{eq:RBM1} for hidden/visible variables is a linear exponential one in sufficient statistics \((\tilde{\boldsymbol X}, \boldsymbol X^T\boldsymbol H)\) using \(\tilde{\boldsymbol X}=(\boldsymbol X,\boldsymbol H)\) from \eqref{eq:RBM1}, but the form differs from the previous exponential models in \eqref{eq:expo} in that the lengths of parameters \(\ak{\thetaN}\) and statistics \((\tilde{\boldsymbol X}, \boldsymbol X^T\boldsymbol H)\) increase to incorporate more visible variables.
+That is, in contrast to \eqref{eq:expo}, the natural parameter function involved in the RBM model \eqref{eq:RBM1}, as the identity mapping of the parameters \(\ak{\thetaN}\in\mathbb{R}^{q(N)}\), naturally grows in dimension \(q(N)\to \infty\) to accommodate visible variables \(X_1, \dots, X_N\) of increasing data size \(N\to \infty\).
 Additionally, one may further arbitrarily choose the number \(N_\mathcal{H}\) of hidden variables \(\boldsymbol H\) in the joint RBM model \eqref{eq:RBM1} to define a marginal model \eqref{eq:RBM2} for the \(N\) visible variables \(\boldsymbol X\), and the number \(N_\mathcal{H}\) of hiddens may also potentially increase with \(N\).
-Because \(|\mathcal{X}| = 2\) and \(P_{\boldsymbol \theta_N}(\boldsymbol x) > 0\) for all \(\boldsymbol x \in \mathcal{X}^N\), the RBM specification \eqref{eq:RBM2} for visibles \(\boldsymbol\) corresponds to a FOES model, while the joint distribution \eqref{eq:RBM1} for \((\boldsymbol X, \boldsymbol H)\) is also a FOES model.
+Because \(|\mathcal{X}| = 2\) and \(P_{\ak{\thetaN}}(\boldsymbol x) > 0\) for all \(\boldsymbol x \in \mathcal{X}^N\), the RBM specification \eqref{eq:RBM2} for visibles \(\boldsymbol\) corresponds to a FOES model, while the joint distribution \eqref{eq:RBM1} for \((\boldsymbol X, \boldsymbol H)\) is also a FOES model.
 As this example also indicates, any model formed by marginalizing a base FOES model class, such as the RBM joint specification \eqref{eq:RBM1}, is again a FOES model.
 
-\hypertarget{deep-learning}{%
-\subsection{Deep Learning}\label{deep-learning}}
+\ak{\subsection{Binary Deep Learning Models} \label{deep-learning}}
 
@@
 
-\textbf{Deep Boltzmann machine (DBM).} The DBM class of models maintains conditional independence within all layers in the model by stacking RBM models and only allowing conditional dependence between neighboring layers.
+\noindent \emph{Deep Boltzmann machine (DBM).} The DBM class of models maintains conditional independence within all layers in the model by stacking RBM models and only allowing conditional dependence between neighboring layers.
 The joint probability mass function for a DBM is
 \[
-\tilde{P}_{\boldsymbol \theta_N} ( \tilde{\boldsymbol x} ) = \exp\left[ \sum\limits_{i = 1}^M\boldsymbol \alpha^{(i)T} \boldsymbol h^{(i)} + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^{(1)T} \Gamma^{(0)} \boldsymbol x + \sum\limits_{i = 1}^{M - 1} \boldsymbol h^{(i)T} \Gamma^{(i)} \boldsymbol h^{(i + 1)} - \psi(\boldsymbol \theta_N) \right],
+\tilde{P}_{\ak{\thetaN}} ( \tilde{\boldsymbol x} ) = \exp\left[ \sum\limits_{i = 1}^M\boldsymbol \alpha^{(i)T} \boldsymbol h^{(i)} + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^{(1)T} \Gamma^{(0)} \boldsymbol x + \sum\limits_{i = 1}^{M - 1} \boldsymbol h^{(i)T} \Gamma^{(i)} \boldsymbol h^{(i + 1)} - \psi(\ak{\thetaN}) \right],
 \]
-for \(\tilde{\boldsymbol x} = (\boldsymbol h^{(1)}, \dots, \boldsymbol h^{(M)}, \boldsymbol x) \in \mathcal{X}^{N_{(H,1)} + \cdots + N_{(H,M)} +N}\) where
-\[
-\psi(\boldsymbol \theta_N) = \log \sum\limits_{\tilde{\boldsymbol x} \in \mathcal{X}^{N_{(H,1)} + \cdots + N_{(H,M)} +N}} \exp\left[ \sum\limits_{i = 1}^M\boldsymbol \alpha^{(i)T} \boldsymbol h^{(i)} + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^{(1)T} \Gamma^{(0)} \boldsymbol x + \sum\limits_{i = 1}^{M - 1} \boldsymbol h^{(i)T} \Gamma^{(i)} \boldsymbol h^{(i + 1)}\right],
-\]
-is the normalizing function for \(\boldsymbol \theta_N = (\boldsymbol \alpha^{(1)}, \dots, \boldsymbol \alpha^{(M)}, \boldsymbol \beta,\Gamma^{(0)}, \dots, \Gamma^{(M - 1)}) \in \Theta_N \subset \mathbb{R}^{q(N)}\), consisting of model parameters \(\boldsymbol \beta \in \mathbb{R}^N\), \(\boldsymbol \alpha^{(i)} \in \mathbb{R}^{N_{(H,i)}}\), \(i = 1, \dots, M\), along with a matrix \(\Gamma^{(0)}\) of dimension \(N_{(H,1)} \times N\), and matrices \(\Gamma^{(i)}\) of dimension
+for \(\tilde{\boldsymbol x} = (\boldsymbol h^{(1)}, \dots, \boldsymbol h^{(M)}, \boldsymbol x) \in \mathcal{X}^{N_{(H,1)} + \cdots + N_{(H,M)} +N}\) \ak{where again $\psi(\ak{\thetaN})$}
+is the normalizing function for \(\ak{\thetaN} = (\boldsymbol \alpha^{(1)}, \dots, \boldsymbol \alpha^{(M)}, \boldsymbol \beta,\Gamma^{(0)}, \dots, \Gamma^{(M - 1)}) \in \Theta_\ak{\thetaidx} \subset \mathbb{R}^{q(N)}\), consisting of model parameters \(\boldsymbol \beta \in \mathbb{R}^N\), \(\boldsymbol \alpha^{(i)} \in \mathbb{R}^{N_{(H,i)}}\), \(i = 1, \dots, M\), along with a matrix \(\Gamma^{(0)}\) of dimension \(N_{(H,1)} \times N\), and matrices \(\Gamma^{(i)}\) of dimension
 \(N_{(H,i)} \times N_{(H,i+1)}\) for \(i = 1, \dots, M-1\).
-The combined parameter vector \(\boldsymbol \theta_N\) has total length \(q(N)= N_{(H,1)}+\cdots N_{(H,M)} + N + N_{(H,1)}*N+N_{H,2}*H_{(H,1)}+\cdots +N_{(H,M)}*H_{(H,M)-1}\).
+The combined parameter vector \(\ak{\thetaN}\) has total length \(q(N)= N_{(H,1)}+\cdots N_{(H,M)} + N + N_{(H,1)}*N+N_{H,2}*H_{(H,1)}+\cdots +N_{(H,M)}*H_{(H,M)-1}\).
 The probability mass function for the visible random variables \(X_1, \dots, X_N\) follows from this joint specification as
 \[
-P_{\boldsymbol \theta_N} (\boldsymbol x) = \sum\limits_{(\boldsymbol h^{(1)}, \dots, \boldsymbol h^{(M)}) \in \mathcal{X}^{N_{(H,1)} + \cdots + N_{(H,M)}}} \tilde{P}_{\boldsymbol \theta_N} (\boldsymbol h^{(1)}, \dots, \boldsymbol h^{(M)}, \boldsymbol x) , \qquad \boldsymbol x \in \mathcal{X}^N.
+P_{\ak{\thetaN}} (\boldsymbol x) = \sum\limits_{(\boldsymbol h^{(1)}, \dots, \boldsymbol h^{(M)}) \in \mathcal{X}^{N_{(H,1)} + \cdots + N_{(H,M)}}} \tilde{P}_{\ak{\thetaN}} (\boldsymbol h^{(1)}, \dots, \boldsymbol h^{(M)}, \boldsymbol x) , \qquad \boldsymbol x \in \mathcal{X}^N.
 \]
@@
 
-\textbf{Deep belief network (DBN).} A DBN resembles a DBM in that there are multiple layers of latent random variables stacked in a deep architecture with no conditional dependence between layers.
-The difference between the DBM and DBN models is that all but the last stacked layer \ak{(layer $0$, which contains the visible random variables $X_1, \dots, X_N$)} in a DBN are Bayesian networks \citep[see][]{pearl985bayesian}, rather than RBMs.
-A Bayesian network is a probabilistic graphical model that defines conditional dependence to be directed, rather than undirected (as with the RBM).
-Thus for visibles \(X_1, \dots, X_N\) with support \(\mathcal{X}^N, \mid \mathcal{X} \mid < \infty\), a DBN is also a FOES model with \(q(N)\) the length of parameter vector is dependent on the dimension of the visibles because \(P_{\boldsymbol \theta_N}(\boldsymbol x)>0\) for all \(\boldsymbol x \in\mathcal{X}^N\).
+\noindent \emph{Deep belief network (DBN).} A DBN resembles a DBM in that there are multiple layers of latent random variables stacked in a deep architecture with no conditional dependence between layers.
+The difference between the DBM and DBN models is that all but the last stacked layer \ak{(e.g.
+"layer $0$" corresponding to the visible random variables $X_1, \dots, X_N$)} in a DBN are Bayesian networks \citep[see][]{pearl985bayesian}, rather than RBMs.
+A Bayesian network \ak{is a class of probabilistic graphical models that define} conditional dependence to be directed, rather than undirected (as with the RBM).
+Thus for visibles \(X_1, \dots, X_N\) with support \(\mathcal{X}^N, \mid \mathcal{X} \mid < \infty\), a DBN is also a FOES model with \(q(N)\) the length of parameter vector is dependent on the dimension of the visibles because \(P_{\ak{\thetaN}}(\boldsymbol x)>0\) for all \(\boldsymbol x \in\mathcal{X}^N\).
 Commonly, as in logistic belief nets \citep{neal1992connectionist}, a ``weight'' parameter is placed on each interaction between visibles, \(X_1, \dots, X_N\), and the first layer of latent variables, \(H^{(1)}_1, \dots, H^{(1)}_{N_{(H,1)}}\), satisfying the definition of a FOES model.
@@
 
-To define a measure of instability in FOES models, it is useful to consider the behavior of data models \(P_{\theta_N}\), again supported on a set \(\mathcal{X}^N\) of outcomes for \(\boldsymbol X\equiv \boldsymbol X_N =(X_1,\ldots,X_N)\) \textbackslash ak\{and for a prescribed configuration of parameters \(\boldsymbol \theta_N\), in connection to the sample size \(N\).
+To define a measure of instability in FOES models, it is useful to consider the behavior of data models \(P_{\ak{\thetaN}}\), again supported on a set \(\mathcal{X}^N\) of outcomes for \(\boldsymbol X\equiv \boldsymbol X_N =(X_1,\ldots,X_N)\) and for a prescribed configuration of parameters \(\ak{\thetaN}\), in connection to the sample size \(N\).
 A relevant quantity to this end is a log-ratio of extremal probabilities (LREP), defined as
@@
 \label{eq:elpr}
- \REP (\boldsymbol \theta_N)  =  \log \left[\frac{\max\limits_{  \boldsymbol x\in \mathcal{X}^N}P_{\boldsymbol \theta_N}( \boldsymbol x)}{\min\limits_{ \boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}( \boldsymbol x)}\right],
+ \REP (\ak{\thetaN})  =  \log \left[\frac{\max\limits_{  \boldsymbol x\in \mathcal{X}^N}P_{\ak{\thetaN}}( \boldsymbol x)}{\min\limits_{ \boldsymbol x \in \mathcal{X}^N}P_{\ak{\thetaN}}( \boldsymbol x)}\right],
 \end{align}
@@
 In what follows, the main idea is that instability, and other negative model features, can be associated with a FOES model formulation for \(N\) random variables where the LREP \eqref{eq:elpr} is overly large relative to the sample size \(N\).
-That is, a sequence of FOES probability models \(P_{\boldsymbol \theta_N}\) results in specifying \ak{a} distribution of observations \(\boldsymbol X=(X_1,\ldots,X_N)\) for each sample size \(N \geq 1\) \ak{(dependent on the prescribed values for $\boldsymbol \theta_N$, $N \geq 1$)} and instability will generally occur among these models whenever the corresponding LREP \eqref{eq:elpr} grows faster than \(N\).
+That is, a sequence of FOES probability models \(P_{\ak{\thetaN}}\) results in specifying \ak{a} distribution of observations \(\boldsymbol X=(X_1,\ldots,X_N)\) for each sample size \(N \geq 1\) \ak{(dependent on the prescribed values for $\ak{\thetaN}$, $N \geq 1$)} and instability will generally occur among these models whenever the corresponding LREP \eqref{eq:elpr} grows faster than \(N\).
 This leads to the following definition.
@@
 \BeginKnitrBlock{definition}[S-unstable FOES model]
-\protect\hypertarget{def:instabFSFS}{}{\label{def:instabFSFS} \iffalse (S-unstable FOES model) \fi{} }A FOES model formulation \ak{$P_{\boldsymbol \theta_N}$} for \(\boldsymbol X_N=(X_1,\ldots,X_N)\), \ak{based on parameter specification $\boldsymbol \theta_N$, $N \geq 1$,} is \emph{Schweinberger-unstable} or \emph{S-unstable} if
+\protect\hypertarget{def:instabFSFS}{}{\label{def:instabFSFS} \iffalse (S-unstable FOES model) \fi{} }A FOES model formulation \ak{$P_{\ak{\thetaN}}$} for \(\boldsymbol X_N=(X_1,\ldots,X_N)\), \ak{based on parameter specification $\ak{\thetaN}$, $N \geq 1$,} is \emph{Schweinberger-unstable} or \emph{S-unstable} if
 \begin{equation}
 \label{eq:Sun}
-\lim \limits_{N \rightarrow \infty} \frac{1}{N} \REP(\boldsymbol \theta_N) \equiv \lim \limits_{N \rightarrow \infty} \frac{1}{N}\log \left[\frac{\max\limits_{  \boldsymbol x\in \mathcal{X}^N}P_{\boldsymbol \theta_N}( \boldsymbol x)}{\min\limits_{ \boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}( \boldsymbol x)}\right] = \infty
+\lim \limits_{N \rightarrow \infty} \frac{1}{N} \REP(\ak{\thetaN}) \equiv \lim \limits_{N \rightarrow \infty} \frac{1}{N}\log \left[\frac{\max\limits_{  \boldsymbol x\in \mathcal{X}^N}P_{\ak{\thetaN}}( \boldsymbol x)}{\min\limits_{ \boldsymbol x \in \mathcal{X}^N}P_{\ak{\thetaN}}( \boldsymbol x)}\right] = \infty
 \end{equation}
@@
 
-In other words, a model is S-unstable if \(\REP(\boldsymbol \theta_N)/N\) is an unbounded sequence of sample size \(N\); namely, given any \(C > 0\), there exists an integer \(N_C > 0\) so that \(\REP(\boldsymbol \theta_N)/N > C\) holds for all \(N \ge N_C\).
-A FOES model formulation may be termed S-stable if it fails to be S-unstable, i.e., if \(\sup_{N \geq 1}\REP(\boldsymbol \theta_N)/N\) is bounded.
+In other words, a model is S-unstable if \(\REP(\ak{\thetaN})/N\) is an unbounded sequence of sample size \(N\); namely, given any \(C > 0\), there exists an integer \(N_C > 0\) so that \(\REP(\ak{\thetaN})/N > C\) holds for all \(N \ge N_C\).
+A FOES model formulation may be termed S-stable if it fails to be S-unstable, i.e., if \(\sup_{N \geq 1}\REP(\ak{\thetaN})/N\) is bounded.
 
-This definition of S-unstable is a generalization or reinterpretation of ``unstable'' used in \citet{schweinberger2011instability} by allowing possibly non-exponential family models (e.g., RBM and DBM models in Sections \ref{rbm}-\ref{deep-learning} as well as a potentially increasing number \(q(N)\) of parameters through the parameter sequence \(\boldsymbol \theta_N\in \mathbb{R}^{q(N)}\).
-While this definition differs in form and scope from the original, it does match that in \citet{schweinberger2011instability} for the special case of exponential models (cf.~Section \ref{discrete-exponential-family-models} considered there \ak{which translates to $\boldsymbol{\theta}_N=\boldsymbol{\theta} \in \mathbb{R}^k$, $N \geq 1$, in our model notation).
-Note the definition (\ref{eq:Sun}) of model instability depends intricately on the prescribed parameter sequence $\boldsymbol{\theta}_N$ or the particular parameter values indexing the distribution $P_{\boldsymbol{\theta}_N}$.
-This also agrees with the notion of unstable distributions from
-\citet{schweinberger2011instability}, as well as other characterizations of distributional stability in networks (cf.
-\citet{handcock2003assessing}), that model instability is crucially tied to the parameter configuration (i.e., where $\boldsymbol{\theta}_N$ lies in the parameter space) in addition to distributional form.
-Hence, a given model family or form may may result in an unstable/stable model, depending on the parameters $\boldsymbol{\theta}_n$ chosen.} Section \ref{illustrations} provides several examples of unstable models as well as causes for model instability, where the latter may often be traced to issues in model form (i.e., data functions) and/or \ak{type of} parameterization.
+This definition of S-unstable is a generalization or reinterpretation of ``unstable'' used in \citet{schweinberger2011instability} by allowing possibly non-exponential family models (e.g., RBM and DBM models in Sections \ref{rbm}-\ref{deep-learning} as well as a potentially increasing number \(q(N)\) of parameters through the parameter sequence \(\ak{\thetaN}\in \mathbb{R}^{q(N)}\).
+While this definition differs in form and scope from the original, it does match that in \citet{schweinberger2011instability} for the special case of exponential models (cf.~Section \ref{discrete-exponential-family-models} considered there \ak{which translates to $\ak{\thetaN}=\boldsymbol{\theta} \in \mathbb{R}^k$, $N \geq 1$, in our model notation).
+Note the definition (\ref{eq:Sun}) of model instability depends intricately on the prescribed parameter sequence $\ak{\thetaN}$ or the particular parameter values indexing the distribution $P_{\ak{\thetaN}}$.
+This also agrees with the notion of unstable distributions from \citet{schweinberger2011instability}, as well as other characterizations of distributional stability in networks (cf.
+\citet{handcock2003assessing}), that model instability is crucially tied to the parameter configuration (i.e., where $\ak{\thetaN}$ lies in the parameter space) in addition to distributional form.
+Hence, a given model family or form may may result in an unstable/stable model, depending on the parameters $\ak{\thetaN}$ chosen.} Section \ref{illustrations} provides several examples of unstable models as well as causes for model instability, where the latter may often be traced to issues in model form (i.e., data functions) and/or \ak{type of} parameterization.
 We next describe several potentially undesirable features associated with S-unstable FOES models.
@@
 \iffalse{} {Remark.
-} \fi{}In the definition \eqref{eq:Sun} of S-instability, we note that the numerical measure \(\REP(\boldsymbol \theta_N)/N\) of model instability is invariant to \emph{independent replications} of data.
-That is, let \(M \geq 1\) denote a possible number of replications and consider data \(\boldsymbol Y_{N,M} \equiv (\boldsymbol X^{(1)}_N, \dots, \boldsymbol X^{(M)}_N)\) formed by \(\{ \boldsymbol X^{(j)}_N\}_{j=1}^M\) as \(M\) iid replications of a random vector \(\boldsymbol X_N=(X_1,\ldots,X_N)\), where the latter follows a FOES model with probabilities \(P_{ \boldsymbol \theta_N}(\boldsymbol x)>0\), \(\boldsymbol x\in\mathcal{X}^N\).
-This leads to a joint model, say \(P_{ \boldsymbol \theta_N}(\boldsymbol y)\), \(\boldsymbol y\in\mathcal{X}^{NM}\), for \(\boldsymbol Y_{N,M}\) consisting of \(N*M\) random variables in total.
-Then, the LREP for \(\boldsymbol Y_{N,M}\), scaled by associated size, is given by
-\begin{align*}
-\frac{1}{NM}\REP_{\boldsymbol Y_{N,M}}( \boldsymbol\theta_N ) &\equiv \frac{1}{NM}\log\left[\frac{\max_{\boldsymbol y \in \mathcal{X}^{NM}}P_{ \boldsymbol \theta_N}(\boldsymbol y)  }{\min_{\boldsymbol y \in \mathcal{X}^{NM}}P_{ \boldsymbol \theta_N}(\boldsymbol y)} \right] \\
-&= \frac{1}{NM}\log\left[\frac{\max_{\boldsymbol x \in \mathcal{X}^{N}}P_{ \boldsymbol \theta_N}(\boldsymbol x)  }{\min_{\boldsymbol x \in \mathcal{X}^{N}}P_{ \boldsymbol \theta_N}(\boldsymbol x)} \right]^M \equiv \frac{1}{N}\REP_{\boldsymbol X_{N}}( \boldsymbol\theta_N ),
-\end{align*}
-where \(\REP_{\boldsymbol X_{N}}( \boldsymbol\theta_N ) \equiv \REP( \boldsymbol\theta_N )\) denotes the log-ratio of extremal probabilities for \(\boldsymbol X_N\) defined from \eqref{eq:elpr}.
-That is, due to iid properties, the sample-size corrected LREP for \(\boldsymbol Y_{N,M}\) equals the analog, \(\REP(\boldsymbol\theta_N )/N\), from the underlying common data model for \(\boldsymbol X_N\) alone, regardless of the level \(M \geq 1\) of independent replication.
-Consequently, the definition of an S-unstable model is unaffected by independent replication and all instability properties may be characterized by those of one observation from the common FOES model.
-For computational purposes, this aspect also implies that if the original data \(\boldsymbol X_N=(X_1,\ldots,X_N)\) in a FOES model consist of \(N\) iid random variables, then the size-scaled log-ratio \eqref{eq:elpr} may be calculated as
-\[
-\frac{1}{N}\REP( \boldsymbol\theta_N ) \equiv \frac{1}{N}\REP_{\boldsymbol X_{N}}( \boldsymbol\theta_N ) = \log\left[ \frac{\max_{ x \in \mathcal{X}}P_{\boldsymbol\theta_N}(X_1=x)} {\min_{ x \in \mathcal{X}}P_{\boldsymbol\theta_N}(X_1=x)} \right]
-\]
-based on the extremal probabilities of just one random variable \(X_1\).
+} \fi{}\ak{Different notions of model stability/instability exist.
+For example, \citet{handcock2003assessing} refers to how small changes in parameters may dramatically change a probability mass function, while \citet{kaiser2007statistical} discusses how interpretation of model parameters and conditional expectations may break down in parts of the parameter space.
+However, these may be intuitively connected in the sense that the same (undesirable) parameter/model configurations may often be associated with varying conceptualizations of instability.
+While beyond the scope of the current paper, \citet{kaplan2019properties} investigates and numerically demonstrates some such connections for a class of RBM models.
+Section \ref{stable-models} also explains one relation between our notion of stable models and parameter interpretation in centered models \citep[][cf.]{kaiser2009exploring}.
+}
+\EndKnitrBlock{remark}
+
+\BeginKnitrBlock{remark}
+\iffalse{} {Remark.
+} \fi{}In the definition \eqref{eq:Sun} of S-instability, we note that the numerical measure \(\REP(\ak{\thetaN})/N\) of model instability is invariant to \emph{independent replications} of data.
+\ak{Consequently, the definition of an
+S-unstable model is unaffected by independent replication and all instability properties may be
+characterized by those of one observation from the common FOES model. Remark A.2 of the Appendix provides more details.}
 \EndKnitrBlock{remark}
@@
 \[
-\DN(\boldsymbol \theta_N) \equiv \max \left\{\log \frac{P_{\boldsymbol \theta_N}(\boldsymbol x)}{P_{\boldsymbol \theta_N}(\boldsymbol x^*)} : \boldsymbol x \text{ }\& \text{ } \boldsymbol x^* \in \mathcal{X}^N \text{ differ in exactly one component}\right\},
+\DN(\ak{\thetaN}) \equiv \max \left\{\log \frac{P_{\ak{\thetaN}}(\boldsymbol x)}{P_{\ak{\thetaN}}(\boldsymbol x^*)} : \boldsymbol x \text{ }\& \text{ } \boldsymbol x^* \in \mathcal{X}^N \text{ differ in exactly one component}\right\},
 \]
-which represents the biggest log-probability ratio for a one-component change \ak{(i.e., the data vector differs in exactly one component value)} in data outcomes in a FOES model with parameter \(\boldsymbol \theta_N\).
-\ak{This is the smallest difference possible between two data vectors, making it a good notion of a small distance in the input space.} We then have the following result prescribing the behavior of \(\DN(\boldsymbol \theta_N)\) for S-unstable FOES models.
+which represents the biggest log-probability ratio for a one-component change \ak{(i.e., the data vector differs in exactly one component value)} in data outcomes in a FOES model with parameter \(\ak{\thetaN}\).
+\ak{This is the smallest difference possible between two data vectors, making it a good notion of a small distance in the input space and has been similarly used by \citet[][, Sec 2.2]{schweinberger2011instability}.} We then have the following result prescribing the behavior of \(\DN(\ak{\thetaN})\) for S-unstable FOES models.
 
 \BeginKnitrBlock{theorem}
-\protect\hypertarget{thm:instab-elpr}{}{\label{thm:instab-elpr} }Let \(P_{\boldsymbol \theta_N}\), with support \(\mathcal{X}^N\), \(N\geq 1\), be a sequence of FOES models.
+\protect\hypertarget{thm:instab-elpr}{}{\label{thm:instab-elpr} }Let \(P_{\ak{\thetaN}}\), with support \(\mathcal{X}^N\), \(N\geq 1\), be a sequence of FOES models.
 \begin{enumerate}[(i)]
-\item For any integer $N \geq 1$ and any given $C>0$, if $\REP(\boldsymbol \theta_N)/N > C$ in (\ref{eq:elpr}), then
-    $$ \DN(\boldsymbol \theta_N) > C,$$
-    or probabilities from a one-component change in some outcome have log-ratio exceeding $C$.
+\item For any integer $N \geq 1$ and any given $C>0$, if $\REP(\ak{\thetaN})/N > C$ holds in (\ref{eq:elpr}), then
+    $$ \DN(\ak{\thetaN}) > C$$
+    follows.
 \item Suppose the FOES model sequence is S-unstable.
@@
     $$
-    \frac{P_{\boldsymbol \theta_N}(\boldsymbol x)}{P_{\boldsymbol \theta_N}(\boldsymbol x^*)} > \exp[N C].
+    \frac{P_{\ak{\thetaN}}(\boldsymbol x)}{P_{\ak{\thetaN}}(\boldsymbol x^*)} > \exp[N C].
     $$
@@
 
-Theorem \ref{thm:instab-elpr}(i) is a non-asymptotic result, which connects to the definition of instability in a FOES model through a log-ratio of extreme model probabilities \eqref{eq:elpr} being too large relative to the associated sample size \(N\).
-If so, Theorem \ref{thm:instab-elpr}(i) guarantees the FOES model must also exhibit correspondingly large changes in probability for very small differences among some data configurations, a property that intuitively captures a notion of instability.
-Furthermore, and perhaps more seriously under Theorem \ref{thm:instab-elpr}(ii),
-S-unstable models can never have universally bounded changes in probability among single component variations in data configurations.
+\ak{Theorem \ref{thm:instab-elpr} aims to describe some model implications when a log-ratio of extreme model probabilities (\ref{eq:elpr}) is too large relative to the associated sample size $N$, which underlies the definition of a S-unstable FOES model (\ref{eq:Sun}).
+If so, Theorem \ref{thm:instab-elpr}(i) guarantees the FOES model must also exhibit correspondingly large changes in probability for small differences among some data configurations.
+Hence, as a further consequence in Theorem \ref{thm:instab-elpr}(ii), S-unstable models can never have universally bounded changes in probability among single component variations in data configurations.
 While not all one-component changes in data may produce massive changes in probability, unstable models must have some such data outcomes with this property.
-As a consequence, unstable probability structures may exhibit extreme sensitivity through large peaks and troughs over the sample space.
+Hence,
+unstable probability structures may exhibit extreme sensitivity through large peaks and troughs over the sample space.
+Theorem \ref{thm:instab-elpr} supports and generalizes findings in \citet[Theorem 1]{schweinberger2011instability} for classes of discrete exponential models (i.e., the result there entails that, if $\REP(\thetaN)/N $ is unbounded as $N$ increases, then the nearest-point log-odds ratio  $\DN(\ak{\thetaN})$  is unbounded as well in these models).}
 
 Additionally, S-unstable FOES model sequences are also connected to degenerate models, where \emph{degeneracy} involves assigning essentially all probability to modes within the sample space, which could potentially represent a small subset among the totality of outcomes.
-For perspective, note that differing sizes of the scaled log-ratio \(\REP(\boldsymbol \theta_N)/N\) from \eqref{eq:elpr} induce a spectrum of levels of instability/stability and Theorem \ref{thm:instab-elpr} indicates increasing sensitivity of model probabilities as \eqref{eq:elpr} increases.
-Furthermore, as the instability measure grows and the log-ratio \(\REP(\boldsymbol \theta_N)/N\) diverges, as in the definition \eqref{eq:Sun} of S-unstable models, then a FOES model sequence will become degenerate.
+For perspective, note that differing sizes of the scaled log-ratio \(\REP(\ak{\thetaN})/N\) from \eqref{eq:elpr} induce a spectrum of levels of instability/stability and Theorem \ref{thm:instab-elpr} indicates increasing sensitivity of model probabilities as \eqref{eq:elpr} increases.
+Furthermore, as the instability measure grows and the log-ratio \(\REP(\ak{\thetaN})/N\) diverges, as in the definition \eqref{eq:Sun} of S-unstable models, then a FOES model sequence will become degenerate.
 Theorem \ref{thm:degenFOES} provides a formal statement of such degeneracy due to S-instability.
@@
 \label{eq:mode}
-\mathcal{M}_{\epsilon, \boldsymbol \theta_N} \equiv \left\{\boldsymbol x \in \mathcal{X}^N: \log P_{\boldsymbol \theta_N}(\boldsymbol x) > (1-\epsilon)\max\limits_{\boldsymbol y \in \mathcal{X}^N} \log  P_{\boldsymbol \theta_N}(\boldsymbol y) + \epsilon\min\limits_{\boldsymbol y \in \mathcal{X}^N} \log P_{\boldsymbol \theta_N}(\boldsymbol y) \right\}.
+\mathcal{M}_{\epsilon, \ak{\thetaN}} \equiv \left\{\boldsymbol x \in \mathcal{X}^N: \log P_{\ak{\thetaN}}(\boldsymbol x) > (1-\epsilon)\max\limits_{\boldsymbol y \in \mathcal{X}^N} \log  P_{\ak{\thetaN}}(\boldsymbol y) + \epsilon\min\limits_{\boldsymbol y \in \mathcal{X}^N} \log P_{\ak{\thetaN}}(\boldsymbol y) \right\}.
 \end{equation}
 \BeginKnitrBlock{theorem}
-\protect\hypertarget{thm:degenFOES}{}{\label{thm:degenFOES} }For any arbitrarily small \(0 < \epsilon < 1\), an S-unstable FOES model sequence \(P_{\boldsymbol \theta_N}\), \(N \geq 1\), for \(\boldsymbol X_N=(X_1, \dots, X_N)\) satisfies
+\protect\hypertarget{thm:degenFOES}{}{\label{thm:degenFOES} }For any arbitrarily small \(0 < \epsilon < 1\), an S-unstable FOES model sequence \(P_{\ak{\thetaN}}\), \(N \geq 1\), for \(\boldsymbol X_N=(X_1, \dots, X_N)\) satisfies
 \[
-P_{\boldsymbol \theta_N}\left( \boldsymbol X_N\in \mathcal{M}_{\epsilon, \boldsymbol \theta_N}\right) \rightarrow 1 \text{ as } N \rightarrow \infty.
+P_{\ak{\thetaN}}\left( \boldsymbol X_N\in \mathcal{M}_{\epsilon, \ak{\thetaN}}\right) \rightarrow 1 \text{ as } N \rightarrow \infty.
 \]
@@
 
-Connected to degeneracy, S-unstable FOES models may also exhibit additional kinds of extreme and undesirable sensitivity in probabilities if model parameters \(\boldsymbol \theta_N\) can further be ``dialed'' between positive and negative values.
+Connected to degeneracy, S-unstable FOES models may also exhibit additional kinds of extreme and undesirable sensitivity in probabilities if model parameters \(\ak{\thetaN}\) can further be ``dialed'' between positive and negative values.
 That is, some FOES models naturally involve parameter spaces covering a positive-negative spectrum of parameter possibilities, where the signs of parameters provide a standard device for increasing or decreasing probabilities of outcomes in the model formulation.
@@
 
-\textbf{Model Condition PSR} \emph{(Reciprocal Probabilities from Parameter Sign Reversal)}: Let \(P_{\boldsymbol \theta_N}\), with support \(\mathcal{X}^N\), \(N\geq 1\), represent a sequence of FOES models.
+\textbf{Model Condition PSR} \emph{(Reciprocal Probabilities from Parameter Sign Reversal)}: Let \(P_{\ak{\thetaN}}\), with support \(\mathcal{X}^N\), \(N\geq 1\), represent a sequence of FOES models.
 For each \(N \geq 1\) and any outcome \(\boldsymbol x \in \mathcal{X}^N\), suppose it holds that
 \[
-P_{\boldsymbol \theta_N}(\boldsymbol x)  \cdot P_{-\boldsymbol \theta_N}(\boldsymbol x) \;=\;   \max\limits_{\boldsymbol y \in \mathcal{X}^N}P_{ \boldsymbol \theta_N}(\boldsymbol y)\cdot \min\limits_{\boldsymbol y \in \mathcal{X}^N}P_{-\boldsymbol \theta_N}(\boldsymbol y),
+P_{\ak{\thetaN}}(\boldsymbol x)  \cdot P_{-\ak{\thetaN}}(\boldsymbol x) \;=\;   \max\limits_{\boldsymbol y \in \mathcal{X}^N}P_{ \ak{\thetaN}}(\boldsymbol y)\cdot \min\limits_{\boldsymbol y \in \mathcal{X}^N}P_{-\ak{\thetaN}}(\boldsymbol y),
 \]
-where \(\max_{\boldsymbol y \in \mathcal{X}^N}P_{ \boldsymbol \theta_N}(\boldsymbol y)\) and \(\min_{\boldsymbol y \in \mathcal{X}^N}P_{-\boldsymbol \theta_N}(\boldsymbol y)\) denote the maximum and minimum probabilities under parameters \(\boldsymbol \theta_N\) and \(-\boldsymbol \theta_N\), respectively.
+where \(\max_{\boldsymbol y \in \mathcal{X}^N}P_{\ak{\thetaN}}(\boldsymbol y)\) and \(\min_{\boldsymbol y \in \mathcal{X}^N}P_{-\ak{\thetaN}}(\boldsymbol y)\) denote the maximum and minimum probabilities under parameters \(\ak{\thetaN}\) and \(-\ak{\thetaN}\), respectively.
 
-The above model condition incorporates many standard parameterizations and follows, for instance, whenever \(P_{\boldsymbol \theta_N}(\boldsymbol x)/P_{\boldsymbol \theta_N}(\boldsymbol y) = [P_{-\boldsymbol \theta_N}(\boldsymbol x)/P_{-\boldsymbol \theta_N}(\boldsymbol y)]^{-1}\) holds for outcomes \(\boldsymbol x, \boldsymbol y \in\mathcal{X}^N\) in a FOES model.
+The above model condition incorporates many standard parameterizations and follows, for instance, whenever \(P_{\ak{\thetaN}}(\boldsymbol x)/P_{\ak{\thetaN}}(\boldsymbol y) = [P_{-\ak{\thetaN}}(\boldsymbol x)/P_{-\ak{\thetaN}}(\boldsymbol y)]^{-1}\) holds for outcomes \(\boldsymbol x, \boldsymbol y \in\mathcal{X}^N\) in a FOES model.
 For instance, this latter condition is fulfilled for all linear exponential families from Section \ref{discrete-exponential-family-models} (e.g., \eqref{eq:mod1}-\eqref{eq:mod2}) as well as all network models from Sections \ref{rbm}-\ref{deep-learning} (e.g., \eqref{eq:RBM1}-\eqref{eq:RBM2}).
@@
 \BeginKnitrBlock{corollary}
-\protect\hypertarget{cor:sign}{}{\label{cor:sign} }Let \(P_{\boldsymbol \theta_N}\), with support \(\mathcal{X}^N\), \(N\geq 1\), be a sequence of FOES models satisfying model condition PSR.
-If the models \(P_{\boldsymbol \theta_N}\) are additionally S-unstable, then
+\protect\hypertarget{cor:sign}{}{\label{cor:sign} }Let \(P_{\ak{\thetaN}}\), with support \(\mathcal{X}^N\), \(N\geq 1\), be a sequence of FOES models satisfying model condition PSR.
+If the models \(P_{\ak{\thetaN}}\) are additionally S-unstable, then
 \begin{enumerate}[(i)]
-\item the models  $P_{-\boldsymbol \theta_N}$  defined by $-\boldsymbol \theta_N$  are also S-unstable;
-\item and for the complement $\mathcal{M}_{\epsilon, \boldsymbol \theta_N}^c \equiv \mathcal{X}^N \setminus \mathcal{M}_{\epsilon, \boldsymbol \theta_N}$ of any mode-set $\mathcal{M}_{\epsilon, \boldsymbol \theta_N}$ under $\boldsymbol \theta_N$ from (\ref{eq:mode}), with $0<\epsilon<1$, it holds under $-\boldsymbol \theta_N$ that
+\item the models  $P_{-\ak{\thetaN}}$  defined by $-\ak{\thetaN}$  are also S-unstable;
+\item and for the complement $\mathcal{M}_{\epsilon, \ak{\thetaN}}^c \equiv \mathcal{X}^N \setminus \mathcal{M}_{\epsilon, \ak{\thetaN}}$ of any mode-set $\mathcal{M}_{\epsilon, \ak{\thetaN}}$ under $\ak{\thetaN}$ from (\ref{eq:mode}), with $0<\epsilon<1$, it holds under $-\ak{\thetaN}$ that
     $$ 
-    P_{-\boldsymbol \theta_N}( \boldsymbol X_N \in \mathcal{M}_{\epsilon, \boldsymbol \theta_N}^c) \rightarrow 1\qquad \text{as } N\to \infty,
+    P_{-\ak{\thetaN}}( \boldsymbol X_N \in \mathcal{M}_{\epsilon, \ak{\thetaN}}^c) \rightarrow 1\qquad \text{as } N\to \infty,
     $$
-    while, by Theorem \ref{thm:degenFOES}, $P_{\boldsymbol \theta_N}( \boldsymbol X_N \in \mathcal{M}_{\epsilon, \boldsymbol \theta_N} ) \rightarrow 1$ holds for $\boldsymbol X_N = (X_1,\ldots,X_N)$ under $\boldsymbol \theta_N$.
+    while, by Theorem \ref{thm:degenFOES}, $P_{\ak{\thetaN}}( \boldsymbol X_N \in \mathcal{M}_{\epsilon, \ak{\thetaN}} ) \rightarrow 1$ holds for $\boldsymbol X_N = (X_1,\ldots,X_N)$ under $\ak{\thetaN}$.
 \end{enumerate}
@@
 
-For unstable models, Corollary \ref{cor:sign} shows that shifts in parameters around zero (i.e., from \(\boldsymbol \theta_N\) to \(-\boldsymbol \theta_N\)) can induce extreme changes in probability among subsets of the sample space, as another manifestation of instability and hyper-sensitivity in probability structure.
-For one-parameter exponential families, involving a fixed real-valued linear parameter \(\boldsymbol \theta_N = \theta \in \mathbb{R}\) and sufficient statistic \(\boldsymbol g_N(\boldsymbol x)\in \mathbb{R}\) in \eqref{eq:expo}, \citet[Theorem 3]{schweinberger2011instability} proved a result similar in spirit, though based on a characterization there in terms of maximum \(U_N \equiv \max_{\boldsymbol x\in\mathcal{X}^N}g_N(\boldsymbol x)\) and minimal \(L_N \equiv \min_{\boldsymbol x\in\mathcal{X}^N}g_N(\boldsymbol x)\) values of the sufficient statistic.
-For this case in particular, mode sets have specific, and essentially complementary, forms over positive and negative parameters, namely, \(\mathcal{M}_{\epsilon, \boldsymbol \theta_N} = \{\boldsymbol x \in\mathcal{X}^N: g_N(\boldsymbol x) > (1-\epsilon) U_N + \epsilon L_N \}\) and \(\mathcal{M}_{\epsilon, -\boldsymbol \theta_N} = \{\boldsymbol x \in\mathcal{X}^N: g_N(\boldsymbol x) < \epsilon U_N + (1-\epsilon) L_N \}\) for any \(\boldsymbol \theta_N>0\), and
+For unstable models, Corollary \ref{cor:sign} shows that shifts in parameters around zero (i.e., from \(\ak{\thetaN}\) to \(-\ak{\thetaN}\)) can induce extreme changes in probability among subsets of the sample space, as another manifestation of instability and hyper-sensitivity in probability structure.
+For one-parameter exponential families, involving a fixed real-valued linear parameter \(\ak{\thetaN} = \theta \in \mathbb{R}\) and sufficient statistic \(\boldsymbol g_N(\boldsymbol x)\in \mathbb{R}\) in \eqref{eq:expo}, \citet[Theorem 3]{schweinberger2011instability} proved a result similar in spirit, though based on a characterization there in terms of maximum \(U_N \equiv \max_{\boldsymbol x\in\mathcal{X}^N}g_N(\boldsymbol x)\) and minimal \(L_N \equiv \min_{\boldsymbol x\in\mathcal{X}^N}g_N(\boldsymbol x)\) values of the sufficient statistic.
+For this case in particular, mode sets have specific, and essentially complementary, forms over positive and negative parameters, namely, \(\mathcal{M}_{\epsilon, \ak{\thetaN}} = \{\boldsymbol x \in\mathcal{X}^N: g_N(\boldsymbol x) > (1-\epsilon) U_N + \epsilon L_N \}\) and \(\mathcal{M}_{\epsilon, -\ak{\thetaN}} = \{\boldsymbol x \in\mathcal{X}^N: g_N(\boldsymbol x) < \epsilon U_N + (1-\epsilon) L_N \}\) for any \(\ak{\thetaN}>0\), and
 \citet[Theorem 3]{schweinberger2011instability} showed each mode set collects all mass, under positive and negative parameters, respectively, with unstable models of this exponential type.
 However, for all unstable FOES models, Corollary \ref{cor:sign} generalizes the same principle that unstable models can push all probability to different, and in fact disjoint, parts of the sample space, depending on how parameters fall with respect to zero.
-This feature can numerically complicate likelihood manipulations, such as maximization or MCMC-based Bayes posterior sampling, as further discussed in Section \ref{implications}.
+This feature can numerically complicate likelihood manipulations, such as maximization or MCMC-based Bayes posterior \ak{simulation}, as further discussed in Section \ref{implications}.
 
@@
 \iffalse{} {Remark. } \fi{}Under the model condition PSR, Corollary \ref{cor:sign} can also be extended to
-cases where parameter components \(\boldsymbol \theta_N =(\boldsymbol \theta_{1,N}, \boldsymbol \theta_{N,2})\) (say) are not all changed in sign (e.g., \(-\boldsymbol \theta_N\)) but, more generally, are instead altered to another parameter configuration \(\boldsymbol \theta_{N}^A = ( \boldsymbol \theta_{1,N}^A, \boldsymbol \theta_{2,N}^A )\) involving a switch in sign only among some dominating model parameters \(\boldsymbol \theta_{N,2}^A=- \boldsymbol \theta_{N,2}\) with remaining parameters \(\boldsymbol \theta_{1,N}^A\) being arbitrarily chosen.
-If a change sign occurs among parameters (\(\pm \boldsymbol \theta_{N,2}\)) which dominate the probability structure of the model, then the results of Corollary \ref{cor:sign} can still hold with \(\boldsymbol \theta_{N}^A\) replacing \(-\boldsymbol \theta_N\); as an example of one sufficient condition, if \(\lim_{N\to \infty} \max_{\boldsymbol x \in \mathcal{X}^N} |G_N(\boldsymbol x, -\boldsymbol \theta_{N}) - G_N(\boldsymbol x, \boldsymbol \theta_{N}^A)|=0\) holds in addition to Corollary \ref{cor:sign} assumptions, where
+cases where parameter components \(\ak{\thetaN} =(\boldsymbol \theta_{1,\ak{\thetaidx}}, \boldsymbol \theta_{2, \ak{\thetaidx}})\) (say) are not all changed in sign (e.g., \(-\ak{\thetaN}\)) but, more generally, are instead altered to another parameter configuration \(\ak{\thetaN}^A = ( \boldsymbol \theta_{1,\ak{\thetaidx}}^A, \boldsymbol \theta_{2,\ak{\thetaidx}}^A )\) involving a switch in sign only among some dominating model parameters \(\boldsymbol \theta_{2,\ak{\thetaidx}}^A=- \boldsymbol \theta_{2,\ak{\thetaidx}}\) with remaining parameters \(\boldsymbol \theta_{1,\ak{\thetaidx}}^A\) being arbitrarily chosen.
+If a change sign occurs among parameters (\(\pm \boldsymbol \theta_{2,\ak{\thetaidx}}\)) which dominate the probability structure of the model, then the results of Corollary \ref{cor:sign} can still hold with \(\ak{\thetaN}^A\) replacing \(-\ak{\thetaN}\); as an example of one sufficient condition, if \(\lim_{N\to \infty} \max_{\boldsymbol x \in \mathcal{X}^N} |G_N(\boldsymbol x, -\ak{\thetaN}) - G_N(\boldsymbol x, \ak{\thetaN}^A)|=0\) holds in addition to Corollary \ref{cor:sign} assumptions, where
 \[
@@
 \]
-represents a standardized form of \(\boldsymbol \theta\)-model probabilities, then the results of Corollary \ref{cor:sign} apply to \(\boldsymbol \theta_{N}^A\) in addition to \(-\boldsymbol \theta_N\).
-As a consequence, an unstable model under \(\boldsymbol \theta_N\) can then imply that many more unstable models exist over a broader spectrum of possibilities for variations \(\boldsymbol \theta_{N}^A\) of \(\boldsymbol \theta_N\), which involves some amount of sign change among components of \(\boldsymbol \theta_N\).
+represents a standardized form of \(\boldsymbol \theta\)-model probabilities, then the results of Corollary \ref{cor:sign} apply to \(\ak{\thetaN}^A\) in addition to \(-\ak{\thetaN}\).
+As a consequence, an unstable model under \(\ak{\thetaN}\) can then imply that many more unstable models exist over a broader spectrum of possibilities for variations \(\ak{\thetaN}^A\) of \(\ak{\thetaN}\), which involves some amount of sign change among components of \(\ak{\thetaN}\).
 \EndKnitrBlock{remark}
@@
 
-As a baseline for comparisons, consider a simplistic model for \(\boldsymbol X_N=(X_1,\ldots,X_N)\) with uniform probabilities over the sample space, say \(P_{\boldsymbol \theta_N}(\boldsymbol x)= |\mathcal{X}|^{-N}\), \(\boldsymbol x \in \mathcal{X}^N\), where each random variable has \(|\mathcal{X}| \geq 1\) outcomes.
+As a baseline for comparisons, consider a simplistic model for \(\boldsymbol X_N=(X_1,\ldots,X_N)\) with uniform probabilities over the sample space, say \(P_{\ak{\thetaN}}(\boldsymbol x)= |\mathcal{X}|^{-N}\), \(\boldsymbol x \in \mathcal{X}^N\), where each random variable has \(|\mathcal{X}| \geq 1\) outcomes.
 In contrast to instability, model probabilities here are completely insensitive to changes in data outcomes across the sample space, and the associated log-ratio of extreme probabilities \eqref{eq:elpr} is
 \[
-\frac{1}{N} \REP(\boldsymbol \theta_N)=0\quad \;(\text{uniform probability model}),
+\frac{1}{N} \REP(\ak{\thetaN})=0\quad \;(\text{uniform probability model}),
 \]
@@
 
-A fundamental model considered in the instability work of \citet{schweinberger2011instability} involves a one-parameter exponential model corresponding to \eqref{eq:expo} with a real-valued parameter, say \(\boldsymbol \theta_N = \eta(\boldsymbol \theta_N)\in \mathbb{R}\), and sufficient statistic \(\boldsymbol g_N(\boldsymbol x)\in \mathbb{R}\).
+A fundamental model considered in the instability work of \citet{schweinberger2011instability} involves a one-parameter exponential model corresponding to \eqref{eq:expo} with a real-valued parameter, say \(\ak{\thetaN} = \eta(\ak{\thetaN})\in \mathbb{R}\), and sufficient statistic \(g_N(\boldsymbol x)\in \mathbb{R}\).
 For such models, upon scaling by sample size \(N\), the log-ratio of extreme probabilities in \eqref{eq:elpr} for assessing instability becomes
@@
 \label{eq:UL}
-\frac{1}{N}\REP(\boldsymbol \theta_N ) \equiv   |\boldsymbol \theta_N| \frac{(U_N-L_N)}{N} \;\quad \text{(one-parameter exponential model)},
+\frac{1}{N}\REP(\ak{\thetaN} ) \equiv   |\ak{\thetaN}| \frac{(U_N-L_N)}{N} \;\quad \text{(one-parameter exponential model)},
 \end{equation}
 where \(U_N \equiv \max_{\boldsymbol x\in\mathcal{X}^N}g_N(\boldsymbol x)\) and \(L_N \equiv \min_{\boldsymbol x\in\mathcal{X}^N}g_N(\boldsymbol x)\) denote the maximal and minimal values of the single sufficient statistic.
-In this case, an S-unstable model results, by definition \eqref{eq:Sun}, whenever \(\lim_{N\to \infty} |\boldsymbol \theta_N| (U_N-L_N)/N= \infty\) holds or, in other words, if the combined magnitudes of parameter \(|\boldsymbol \theta_N|\) and maximal difference \(U_N-L_N\) in statistic values are overwhelmingly large relative to the sample size \(N\).
-If we further assume that \(\boldsymbol \theta_N =\theta\in\mathbb{R}\setminus \{0\}\) is a fixed (non-zero) parameter for all \(N \geq 1\), as considered in \citet{schweinberger2011instability}, then an S-unstable model results solely if the sufficient statistic admits a value \(U_N-L_N\) too large relative to number \(N\) of observations, i.e., if \((U_N-L_N)/N\to \infty\) as \(N\to \infty\).
+In this case, an S-unstable model results, by definition \eqref{eq:Sun}, whenever \(\lim_{N\to \infty} |\ak{\thetaN}| (U_N-L_N)/N= \infty\) holds or, in other words, if the combined magnitudes of parameter \(|\ak{\thetaN}|\) and maximal difference \(U_N-L_N\) in statistic values are overwhelmingly large relative to the sample size \(N\).
+If we further assume that \(\ak{\thetaN} =\theta\in\mathbb{R}\setminus \{0\}\) is a fixed (non-zero) parameter for all \(N \geq 1\), as considered in \citet{schweinberger2011instability}, then an S-unstable model results solely if the sufficient statistic admits a value \(U_N-L_N\) too large relative to number \(N\) of observations, i.e., if \((U_N-L_N)/N\to \infty\) as \(N\to \infty\).
 The latter aspect reflects the definition of \citet{schweinberger2011instability}, for this setting, that a real-valued \emph{statistic} \(g_N(\boldsymbol x)\) may be classified as \emph{unstable} when \(\lim_{N\to \infty}|(U_N-L_N)/N=\infty\) holds and as \emph{stable} otherwise (e.g., if \(\sup_{N \geq 1}(U_N-L_N)/N<\infty\)).
 
-For illustration, consider the iid Bernoulli model \eqref{eq:mod1} for \(\boldsymbol X_N=(X_1,\ldots,X_N)\) with log-odds ratio parameter \(\boldsymbol \theta_N = \log[ P_{\boldsymbol \theta_N}(X_1=1)/ P_{\boldsymbol \theta_N}(X_1=0)]\in\mathbb{R}\).
-\ak{Remark 3.1} (Section \ref{criterion}) then gives the model instability measure \eqref{eq:Sun} directly as
+For illustration, consider the iid Bernoulli model \eqref{eq:mod1} for \(\boldsymbol X_N=(X_1,\ldots,X_N)\) with log-odds ratio parameter \(\ak{\thetaN} = \log[ P_{\ak{\thetaN}}(X_1=1)/ P_{\ak{\thetaN}}(X_1=0)]\in\mathbb{R}\).
+\ak{Remark 3.2} (Section \ref{criterion}) then gives the model instability measure \eqref{eq:Sun} directly as
 \[
-\frac{1}{N}\REP(\boldsymbol \theta_N ) = |\boldsymbol \theta_N|\quad\; \text{(iid Bernoulli model)},
+\frac{1}{N}\REP(\ak{\thetaN} ) = |\ak{\thetaN}|\quad\; \text{(iid Bernoulli model)},
 \]
-so that an unstable (or stable) model results for a divergent (or bounded) parameter sequence \(|\boldsymbol \theta_N|\).
-The above instability expression for the Bernoulli model follows as well from the \(N\)-scaled LREP value \eqref{eq:UL} for a one-parameter exponential distribution, using that the sufficient statistic involved \(g_N(\boldsymbol x)= \sum_{i=1}^N x_i\), \(\boldsymbol x =(x_1,\ldots,x_n)\in\{0,1\}^N\), has maximum and minimum values \(U_N=N\) and \(L_N=0\).
-In this case, \citet{schweinberger2011instability} has noted that the statistic is stable (i.e., bounded \((U_N-L_N)/N=1\)) and the Bernoulli model is as well when, in particular, \(\boldsymbol \theta_N=\boldsymbol \theta \in\mathbb{R}\) is fixed for \(N \geq 1\).
+so that an unstable (or stable) model results for a divergent (or bounded) parameter sequence \(|\ak{\thetaN}|\).
+\ak{In this case}, \citet{schweinberger2011instability} has also noted that the statistic is stable (i.e., bounded \((U_N-L_N)/N=1\)) and the Bernoulli model is as well when, in particular, \(\ak{\thetaN}=\theta \in\mathbb{R}\) is fixed for \(N \geq 1\).
 
@@
 \[
-\frac{1}{N}\REP(\boldsymbol \theta_N )  = \left\{ \begin{array}{lcl} |\boldsymbol \theta_N| (n-2) && \text{(2-star graph model)}\\
-|\boldsymbol \theta_N|(n-2)/3 &&\text{(triangle graph model)},\end{array}\right.
+\frac{1}{N}\REP(\ak{\thetaN} )  = \left\{ \begin{array}{lcl} |\ak{\thetaN}| (n-2) && \text{(2-star graph model)}\\
+|\ak{\thetaN}|(n-2)/3 &&\text{(triangle graph model)},\end{array}\right.
 \]
@@
 Because the variable number \(N\to \infty\) as the node number \(n\to \infty\), both counts of 2-stars and triangles are unstable statistics in the sense of \citet{schweinberger2011instability} (i.e., \(\lim_{N\to \infty} (U_N-L_N)/N=\infty\)).
-Furthermore, both types of graph models are always S-unstable for all possible of parameter sequences \(\boldsymbol \theta_N \in\mathbb{R}\) that are bounded away from zero (i.e., \(\lim_{N\to \infty}\REP(\boldsymbol \theta_N )/N=\infty\) then holds, including the fixed parameter case \(\boldsymbol \theta_N=\theta\in\mathbb{R}\setminus \{0\}\) from \citet{schweinberger2011instability}).
+Furthermore, both types of graph models are always S-unstable for all possible of parameter sequences \(\ak{\thetaN} \in\mathbb{R}\) that are bounded away from zero (i.e., \(\lim_{N\to \infty}\REP(\ak{\thetaN} )/N=\infty\) then holds, including the fixed parameter case \(\ak{\thetaN}=\theta\in\mathbb{R}\setminus \{0\}\) from \citet{schweinberger2011instability}).
 
@@
 
-As a generalization of the one-parameter exponential case, we next consider linear exponential families \eqref{eq:expo} with \(k\) parameters \(\boldsymbol \theta_N = (\theta_{1,N},\ldots,\theta_{k,N})^\prime\) and \(k\) sufficient statistics \(\boldsymbol g_N(\boldsymbol x) = (g_{1,N}(\boldsymbol x),\ldots, g_{k,N}(\boldsymbol x))^\prime\).
+As a generalization of the one-parameter exponential case, we next consider linear exponential families \eqref{eq:expo} with \(k\) parameters \(\ak{\thetaN} = (\theta_{1,\ak{\thetaidx}},\ldots,\theta_{k,\ak{\thetaidx}})^\prime\) and \(k\) sufficient statistics \(\boldsymbol g_N(\boldsymbol x) = (g_{1,N}(\boldsymbol x),\ldots, g_{k,N}(\boldsymbol x))^\prime\).
 Here the dimension \(k\) of model parameters/statistics is fixed, and we next prescribe a condition helpful to avoiding instability in such models.
@@
 \BeginKnitrBlock{proposition}
-\protect\hypertarget{prp:prop1}{}{\label{prp:prop1} }Let \(P_{\boldsymbol \theta_N}\), \(N \geq 1\), denote linear exponential models (\ref{eq:expo}) with parameters \(\boldsymbol \theta_N = (\theta_{1,N},\ldots,\theta_{k,N})^\prime \in \mathbb{R}^k\) and statistics \(\boldsymbol g_N(\boldsymbol x) = (g_{1,N}(\boldsymbol x),\ldots, g_{k,N}(\boldsymbol x))^\prime \in \mathbb{R}^k\), for fixed \(k \geq 1\).
-Then, the models \(P_{\boldsymbol \theta_N}\) are S-stable if
+\protect\hypertarget{prp:prop1}{}{\label{prp:prop1} }Let \(P_{\ak{\thetaN}}\), \(N \geq 1\), denote linear exponential models (\ref{eq:expo}) with parameters \(\ak{\thetaN} = (\theta_{1,\ak{\thetaidx}},\ldots,\theta_{k,\ak{\thetaidx}})^\prime \in \mathbb{R}^k\) and statistics \(\boldsymbol g_N(\boldsymbol x) = (g_{1,N}(\boldsymbol x),\ldots, g_{k,N}(\boldsymbol x))^\prime \in \mathbb{R}^k\), for fixed \(k \geq 1\).
+Then, the models \(P_{\ak{\thetaN}}\) are S-stable if
 \begin{equation}
 \label{eq:prop1}
-\sup_{N \geq 1}\frac{1}{N} \max_{1 \leq i \leq k }|\theta_{i,N}|(U_{i,N}-L_{i,N})<\infty
+\sup_{N \geq 1}\frac{1}{N} \max_{1 \leq i \leq k }|\theta_{i,\ak{\thetaidx}}|(U_{i,N}-L_{i,N})<\infty
 \end{equation}
-holds, i.e., if \(\max_{1 \leq i \leq k } |\theta_{i,N}|(U_{i,N}-L_{i,N})/N\) is bounded sequence of sample size \(N\).
+holds, i.e., if \(\max_{1 \leq i \leq k } |\theta_{i,\ak{\thetaidx}}|(U_{i,N}-L_{i,N})/N\) is bounded sequence of sample size \(N\).
 \EndKnitrBlock{proposition}
@@
 \iffalse{} {Remark.
-} \fi{}In the one-parameter exponential case \(k=1\), recall the exponential model is stable/unstable depending on whether \(\REP(\boldsymbol \theta_N)/N = |\theta_{1,N}|(U_{1,N}-L_{1,N})/N \equiv |\boldsymbol \theta_{N}|(U_{N}-L_{N})/N\) in \eqref{eq:UL} is convergent/divergent.
+} \fi{}In the one-parameter exponential case \(k=1\), recall the exponential model is stable/unstable depending on whether \(\REP(\ak{\thetaN})/N = |\theta_{1,\ak{\thetaidx}}|(U_{1,N}-L_{1,N})/N \equiv | \ak{\thetaN}|(U_{N}-L_{N})/N\) in \eqref{eq:UL} is convergent/divergent.
 Hence, for \(k=1\), the condition \eqref{eq:prop1} of Proposition \ref{prp:prop1} captures the same notion of S-stability based on \eqref{eq:UL}.
@@
 
-Proposition \ref{prp:prop1} provides a sufficient condition for the stability of linear exponential models with fixed parameter dimension \(k\geq 1\), whereby an S-stable model is guaranteed if the \ak{product of magnitudes} of each combination of parameter \(\theta_{i,N}\) and sufficient statistic value \((U_{i,N}-L_{i,N})\) is bounded by the sample size \(N\), \(i=1,\ldots,k\).
+Proposition \ref{prp:prop1} provides a sufficient condition for the stability of linear exponential models with fixed parameter dimension \(k\geq 1\), whereby an S-stable model is guaranteed if the \ak{product of magnitudes} of each combination of parameter \(\theta_{i,\ak{\thetaidx}}\) and sufficient statistic value \((U_{i,N}-L_{i,N})\) is bounded by the sample size \(N\), \(i=1,\ldots,k\).
 This supports the findings of \citet{schweinberger2011instability}, who showed degeneracy follows in such models under one type of violation of the condition \eqref{eq:prop1} in Proposition \ref{prp:prop1} (namely, involving \(k>1\) non-zero parameters with \(k-1\) statistics being \(O(N)\) bounded while one statistic diverges in maximal size faster than the number \(N\) of observations).
-To further illustrate the result in Proposition \ref{prp:prop1}, consider the multinomial distribution \eqref{eq:mod11} for \(\boldsymbol X_N=(X_1,\ldots,X_N)\) having \(k\geq 2\) categories \(\{1,\ldots,k\}\) and \(k\) parameters \(\boldsymbol \theta_N = (\theta_{1,N},\ldots,\theta_{k,N})^\prime\).
-The variables are iid under this model so that \ak{Remark 3.1} (Section \ref{criterion} yields the corresponding \(N\)-scaled log-ratio of extreme probabilities \eqref{eq:elpr} as
+To further illustrate the result in Proposition \ref{prp:prop1}, consider the multinomial distribution \eqref{eq:mod11} for \(\boldsymbol X_N=(X_1,\ldots,X_N)\) having \(k\geq 2\) categories \(\{1,\ldots,k\}\) and \(k\) parameters \(\ak{\thetaN} = (\theta_{1,\ak{\thetaidx}},\ldots,\theta_{k,\ak{\thetaidx}})^\prime\).
+The variables are iid under this model so that \ak{Remark 3.2} (Section \ref{criterion}) yields the corresponding \(N\)-scaled log-ratio of extreme probabilities \eqref{eq:elpr} as
 \begin{align*}
-  \frac{1}{N}\REP(\boldsymbol \theta_N) &= \frac{\max_{1 \leq i \leq k} P_{\boldsymbol \theta_N}(X_1=i)}{\min_{1 \leq i \leq k} P_{\boldsymbol \theta_N}(X_1=i)}\\
-  &= \max_{1 \leq i \leq k} \theta_{i,N} - \min_{1 \leq i \leq k} \theta_{i,N} \qquad \text{(iid multinomial model)}.
+  \frac{1}{N}\REP(\ak{\thetaN}) &= \frac{\max_{1 \leq i \leq k} P_{\ak{\thetaN}}(X_1=i)}{\min_{1 \leq i \leq k} P_{\ak{\thetaN}}(X_1=i)}\\
+  &= \max_{1 \leq i \leq k} \theta_{i,\ak{\thetaidx}} - \min_{1 \leq i \leq k} \theta_{i,\ak{\thetaidx}} \qquad \text{(iid multinomial model)}.
 \end{align*}
-Hence, a multinomial model sequence is unstable (or stable) depending on whether (or not) the maximal parameter difference \(\max_{1 \leq i \leq k} \theta_{i,N} - \min_{1 \leq i \leq k} \theta_{i,N}\) diverges.
-Furthermore, using that each of the \(k\) sufficient (count) statistics from the multinomial model \eqref{eq:mod11} satisfies \((U_{i,N}-L_{i,N})/N=1\), we see that \eqref{eq:prop1} of Proposition \ref{prp:prop1} becomes purely a parameter condition, \(\sup_{N \geq 1}\max_{1 \leq i \leq k } |\theta_{i,N}| <\infty\), for ensuring that \(\REP(\boldsymbol \theta_N)/N =\max_{1 \leq i \leq k} \theta_{i,N} - \min_{1 \leq i \leq k} \theta_{i,N}\) is bounded and stability follows for the multinomial distribution.
-Additionally, a stable multinomial sequence (i.e., bounded \(\REP(\boldsymbol \theta_N)/N\)) turns out to be nearly equivalent to \eqref{eq:prop1} (e.g., these are the same if the smallest parameter \(\min_{1 \leq i \leq k } |\theta_{i,N}|\) remains bounded).
+Hence, a multinomial model sequence is unstable (or stable) depending on whether (or not) the maximal parameter difference \(\max_{1 \leq i \leq k} \theta_{i,\ak{\thetaidx}} - \min_{1 \leq i \leq k} \theta_{i,\ak{\thetaidx}}\) diverges.
+Furthermore, using that each of the \(k\) sufficient (count) statistics from the multinomial model \eqref{eq:mod11} satisfies \((U_{i,N}-L_{i,N})/N=1\), we see that \eqref{eq:prop1} of Proposition \ref{prp:prop1} becomes purely a parameter condition, \(\sup_{N \geq 1}\max_{1 \leq i \leq k } |\theta_{i,\ak{\thetaidx}}| <\infty\), for ensuring that \(\REP(\ak{\thetaN})/N =\max_{1 \leq i \leq k} \theta_{i,\ak{\thetaidx}} - \min_{1 \leq i \leq k} \theta_{i,\ak{\thetaidx}}\) is bounded and stability follows for the multinomial distribution.
+Additionally, a stable multinomial sequence (i.e., bounded \(\REP(\ak{\thetaN})/N\)) turns out to be nearly equivalent to \eqref{eq:prop1} (e.g., these are the same if the smallest parameter \(\min_{1 \leq i \leq k } |\theta_{i,\ak{\thetaidx}}|\) remains bounded).
 
 When the condition \eqref{eq:prop1} of Proposition \ref{prp:prop1} is violated, this aspect suggests a potentially unstable model that may be investigated more closely.
-For example, consider the exponential graph model from \eqref{eq:mod2} involving counts of edges, 2-stars and triangles with fixed parameters \(\boldsymbol \theta_N = (\theta_{1},\theta_2,\theta_3)^\prime \in \mathbb{R}^3\) for \(N\geq 1\).
+For example, consider the exponential graph model from \eqref{eq:mod2} involving counts of edges, 2-stars and triangles with fixed parameters \(\thetaN = (\theta_{1},\theta_2,\theta_3)^\prime \in \mathbb{R}^3\) for \(N\geq 1\).
 If either the 2-star parameter \(\theta_2 \neq 0\) or triangle parameter \(\theta_3 \neq 0\) is non-zero, then \(\max_{1 \leq i \leq 3 } |\theta_{i}|(U_{i,N}-L_{i,N})/N \propto (n-2)\to \infty\) holds in \eqref{eq:prop1} by \((U_{2,N}-L_{2,N})/N = 3 (U_{3,N}-L_{3,N})/N=(n-2)\) for 2-star and triangle statistics (\(i=2,3\)), so that Proposition \ref{prp:prop1} hints that an unstable model may result when \(|\theta_2| + |\theta_3| \neq 0\).
-Relatedly, a result from \citet[Result 3]{schweinberger2011instability} states that this model is unstable for all fixed parameters excluding cases \(\theta_2 =\theta_3=0\) or \(\theta_2 = - \theta_3/3\)\ak{, more in line} with the instability suggested by Proposition \ref{prp:prop1} whenever \(|\theta_2| + |\theta_3| \neq 0\) (i.e., excluding \(\theta_2 =\theta_3=0\)).
-
-To see this, consider an even number \(n>2\) of nodes and let \(\boldsymbol x_0\) denote the data outcome in \(\mathcal{X}^N \equiv \{0,1\}^N\) with all \(N = {n \choose 2}\) edges being zero, let \(\boldsymbol x_1\) denote the outcome with all edges being 1, and let \(\boldsymbol x_2\) denote the edge configuration from dividing the nodes into two equal groups, with no edges within a group and all edges between the groups (so that no triangles exist in \(\boldsymbol x_2\)).
-Then, the \(N\)-scaled log-ratio \eqref{eq:elpr} for the exponential graph model \eqref{eq:mod2} can, by definition, be bounded below by
-\begin{align*}
-\frac{1}{N}\REP_N(\boldsymbol \theta_N) &\geq \max_{i=1,2}\frac{1}{N}
-\left| \log\left[ \frac{P_{\boldsymbol \theta_N}(\boldsymbol x_i)}{P_{\boldsymbol \theta_N}(\boldsymbol x_0)}\right] \right| \\
-&= (n-2) \max\left\{ \left| \theta_2 + \frac{\theta_3}{3}+\frac{\theta_1}{n-2} \right|, \frac{n}{4(n-1)} \left| \theta_2 + \frac{8\theta_1}{n-2} \right| \right\};
-\end{align*}
-a similar expression also holds for an odd node number \(n>2\).
-Consequently, for all fixed parameters excluding \(\theta_2=\theta_3=0\), \(\lim_{N\to \infty}\REP_N(\boldsymbol \theta_N)/N=\infty\) then follows and the graph model with 2-stars and triangles is S-unstable, as suggested by the breach of Proposition \ref{prp:prop1} for this model when \(|\theta_2|+|\theta_3|\neq 0\).
-That is, instability holds even under \ak{the} \(\theta_2 = - \theta_3/3\) case potentially allowed by Schweinberger's \citeyearpar{schweinberger2011instability} results.
+Relatedly, a result from \citet[Result 3]{schweinberger2011instability} states that this model is unstable for all fixed parameters excluding cases \(\theta_2 =\theta_3=0\) or \(\theta_2 = - \theta_3/3\)\ak{.
+However, more in line} with the instability suggested by Proposition \ref{prp:prop1}\ak{, this model is unstable} whenever \(|\theta_2| + |\theta_3| \neq 0\) (i.e., excluding \(\theta_2 =\theta_3=0\))\ak{; Remark A.1 in the Appendix gives a proof.} That is, instability holds even under \ak{the} \(\theta_2 = - \theta_3/3\) case potentially allowed by Schweinberger's \citeyearpar{schweinberger2011instability} results.
+\ak{Thus, graph models of the form (\ref{eq:mod2}), with 2-stars and triangles are always S-unstable for all possible parameter sequences $\thetaN \in \mathbb{R}^3$ with $(\theta_2, \theta_3)$ bounded away from $\boldsymbol 0 \in \mathbb{R}^2$.}
 
@@
 However, the distribution of visible variables is induced by an underlying joint exponential model \eqref{eq:RBM1} for both visible and latent variables \((\boldsymbol X, \boldsymbol H)\), where \(\boldsymbol H = (H_1,\ldots,H_{N_{\mathcal{H}}})\) denotes a vector of \(N_{\mathcal{H}}\) hidden variables (similarly binary).
-The joint model is of linear exponential form involving \(q(N)\equiv N + N_{\mathcal{H}} + N*N_{\mathcal{H}}\) sufficient statistics given by \((\boldsymbol X, \boldsymbol H, \boldsymbol X^T\boldsymbol H)\) and parameters \(\boldsymbol \theta_N = (\boldsymbol \theta_N^{\mathcal{V}},\boldsymbol \theta_N^{\mathcal{H}}, \boldsymbol \theta_N^{\mathcal{VH}} ) \in\mathbb{R}^{q(N)}\) corresponding to the \(N\) visible variables \(\boldsymbol X\) (i.e., \(\boldsymbol \theta_N^{\mathcal{V}}\in\mathbb{R}^N\)), the \(N_{\mathcal{H}}\) hidden variables \(\boldsymbol H\) (i.e., \(\boldsymbol \theta_N^{\mathcal{H}}\in\mathbb{R}^{N_{\mathcal{H}}}\)), and the \(N *N_{\mathcal{H}}\) cross-product variables \(\boldsymbol X^T\boldsymbol H\) (i.e., \(\boldsymbol \theta_N^{\mathcal{VH}}\in\mathbb{R}^{N*N_{\mathcal{H}}}\)).
+The joint model is of linear exponential form involving \(q(N)\equiv N + N_{\mathcal{H}} + N*N_{\mathcal{H}}\) sufficient statistics given by \((\boldsymbol X, \boldsymbol H, \boldsymbol X^T\boldsymbol H)\) and parameters \(\ak{\thetaN} = (\ak{\thetaN}^{\mathcal{V}}, \ak{\thetaN}^{\mathcal{H}}, \ak{\thetaN}^{\mathcal{VH}} ) \in\mathbb{R}^{q(N)}\) corresponding to the \(N\) visible variables \(\boldsymbol X\) (i.e., \(\ak{\thetaN}^{\mathcal{V}}\in\mathbb{R}^N\)), the \(N_{\mathcal{H}}\) hidden variables \(\boldsymbol H\) (i.e., \(\ak{\thetaN}^{\mathcal{H}}\in\mathbb{R}^{N_{\mathcal{H}}}\)), and the \(N *N_{\mathcal{H}}\) cross-product variables \(\boldsymbol X^T\boldsymbol H\) (i.e., \(\ak{\thetaN}^{\mathcal{VH}}\in\mathbb{R}^{N*N_{\mathcal{H}}}\)).
 However, unlike some previous exponential models considered in Sections \ref{one-param-exp}-\ref{fixed-dim-exp} (cf.~Proposition \ref{prp:prop1}, note that the RBM formulation always associates parameters with \emph{bounded} statistics (i.e., the components of \((\boldsymbol X, \boldsymbol H, \boldsymbol X^T\boldsymbol H)\)) so that model instability cannot arise here due to the magnitude of sufficient statistics exceeding the sample size \(N\).
@@
 To highlight the instability issues for the RBM model, consider a simple model for \(N\) visibles \(\boldsymbol X\) with no hidden variables (\(N_{\mathcal{H}}=0\)), for which model statements \eqref{eq:RBM1}-\eqref{eq:RBM2} coincide.
-An independence model then results for variables in \(\boldsymbol X\), which has \(q(N)=N\) parameters \(\boldsymbol \theta_N^{\mathcal{V}} = (\theta_{1,N}^{\mathcal{V}}, \ldots, \theta_{N,N}^{\mathcal{V}}) \in \mathbb{R}^{N}\), and the measure of model instability becomes
+An independence model then results for variables in \(\boldsymbol X\), which has \(q(N)=N\) parameters \(\ak{\thetaN}^{\mathcal{V}} = (\theta_{1,\ak{\thetaidx}}^{\mathcal{V}}, \ldots, \theta_{N,\ak{\thetaidx}}^{\mathcal{V}}) \in \mathbb{R}^{N}\), and the measure of model instability becomes
 \[
-\frac{1}{N}\REP(\boldsymbol \theta_N) = \frac{2}{N} \sum_{i=1}^N|\theta_{i,N}^{\mathcal{V}}| \quad \text{(RBM model, no hiddens)}.
+\frac{1}{N}\REP(\ak{\thetaN}) = \frac{2}{N} \sum_{i=1}^N|\theta_{i,\ak{\thetaidx}}^{\mathcal{V}}| \quad \text{(RBM model, no hiddens)}.
 \]
-Hence, this model sequence for \(\boldsymbol X\) will be S-unstable model if the aggregation of absolute parameters grows faster than the number \(N\) of parameters/visible variables.
+Hence, this model sequence for \(\boldsymbol X\) will be S-unstable if the aggregation of absolute parameters grows faster than the number \(N\) of parameters/visible variables.
 Consequently, even for \ak{the} simplest RBM model involving independence, preventing instability requires careful choice of parameters, particularly with regard to how a parameter configuration differs from zero.
@@
 
-To state the result, let \(\REP_{\boldsymbol X}(\boldsymbol \theta_N) \equiv \REP(\boldsymbol \theta_N)\) denote the LREP value \eqref{eq:elpr} from the marginal distribution \(P_{\boldsymbol \theta_N}\) of visibles \(\boldsymbol X\) in \eqref{eq:RBM2} and write the LREP for the joint distribution \(\tilde{P}_{\boldsymbol \theta_N}\) of \((\boldsymbol X, \boldsymbol H)\) from \eqref{eq:RBM1} as
+To state the result, let \(\REP_{\boldsymbol X}(\ak{\thetaN}) \equiv \REP(\ak{\thetaN})\) denote the LREP value \eqref{eq:elpr} from the marginal distribution \(P_{\ak{\thetaN}}\) of visibles \(\boldsymbol X\) in \eqref{eq:RBM2} and write the LREP for the joint distribution \(\tilde{P}_{\ak{\thetaN}}\) of \((\boldsymbol X, \boldsymbol H)\) from \eqref{eq:RBM1} as
 \begin{align*}
-\REP_{(\boldsymbol X, \boldsymbol H)}(\boldsymbol \theta_N) &=\log\left[  \frac{\max_{ (\boldsymbol x, \boldsymbol h) \in \{\pm 1\}^{N+N_{\mathcal{H}}}}
-\tilde{P}_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)}{\min_{ (\boldsymbol x, \boldsymbol h) \in \{\pm 1\}^{N+N_{\mathcal{H}}}`}
-\tilde{P}_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)} \right] \qquad \text{(joint RBM model)}    \\    
-&= \left(\max_{ (\boldsymbol x, \boldsymbol h) \in \{\pm 1\}^{N+N_{\mathcal{H}}}} f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)-\min_{ (\boldsymbol x, \boldsymbol h) \in \{\pm 1\}^{N+N_{\mathcal{H}}}} f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)\right),
+\REP_{(\boldsymbol X, \boldsymbol H)}(\ak{\thetaN}) &=\log\left[  \frac{\max_{ (\boldsymbol x, \boldsymbol h) \in \{\pm 1\}^{N+N_{\mathcal{H}}}}
+\tilde{P}_{\ak{\thetaN}} (\boldsymbol x, \boldsymbol h)}{\min_{ (\boldsymbol x, \boldsymbol h) \in \{\pm 1\}^{N+N_{\mathcal{H}}}`}
+\tilde{P}_{\ak{\thetaN}} (\boldsymbol x, \boldsymbol h)} \right] \qquad \text{(joint RBM model)}    \\    
+&= \left(\max_{ (\boldsymbol x, \boldsymbol h) \in \{\pm 1\}^{N+N_{\mathcal{H}}}} f_{\ak{\thetaN}} (\boldsymbol x, \boldsymbol h)-\min_{ (\boldsymbol x, \boldsymbol h) \in \{\pm 1\}^{N+N_{\mathcal{H}}}} f_{\ak{\thetaN}} (\boldsymbol x, \boldsymbol h)\right),
 \end{align*}
@@
 \label{eq:f}
-f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)  \equiv  \sum_{i=1}^N x_i  \theta_{i,N}^{\mathcal{V}} +  \sum_{j=1}^{N_\mathcal{H}} h_j \theta_{j,N}^{\mathcal{H}} + \sum_{i=1}^N \sum_{j=1}^{N_\mathcal{H}} x_i h_j  \theta_{ij,N}^{\mathcal{VH}}
+f_{\ak{\thetaN}} (\boldsymbol x, \boldsymbol h)  \equiv  \sum_{i=1}^N x_i  \theta_{i,\ak{\thetaidx}}^{\mathcal{V}} +  \sum_{j=1}^{N_\mathcal{H}} h_j \theta_{j,\ak{\thetaidx}}^{\mathcal{H}} + \sum_{i=1}^N \sum_{j=1}^{N_\mathcal{H}} x_i h_j  \theta_{ij,\ak{\thetaidx}}^{\mathcal{VH}}
 \end{equation}
-of outcomes \(\boldsymbol x =(x_1,\ldots,x_N) \in\{\pm1\}^{N}\), \(\boldsymbol h =(h_1,\ldots,h_{N_{\mathcal{H}}}) \in\{\pm1\}^{N_{\mathcal{H}}}\) and parameters \(\boldsymbol \theta_N \equiv (\boldsymbol \theta_N^{\mathcal{V}},\boldsymbol \theta_N^{\mathcal{H}}, \boldsymbol \theta_N^{\mathcal{VH}})\), with \(\theta_{i,N}^{\mathcal{V}}\), \(\theta_{j,N}^{\mathcal{H}}\) and \(\theta_{ij,N}^{\mathcal{VH}}\) denoting respective parameter components, \(1 \leq i \leq N\), \(1 \leq j \leq N_{\mathcal{H}}\).
-Due to the marginalization steps in defining the distribution \eqref{eq:RBM2} of \(\boldsymbol X\), note that \(\REP_{\boldsymbol X}(\boldsymbol \theta_N)\) has no immediate analytical expression similar to that of \(\REP_{(\boldsymbol X, \boldsymbol H)}(\boldsymbol \theta_N)\).
-For clarity, recall also that S-instability \eqref{eq:Sun} in each model type refers to a respective divergence (i.e., \(\lim_{N\to \infty} \REP_{(\boldsymbol X, \boldsymbol H)}(\boldsymbol \theta_N) /(N+N_{\mathcal{H}})=\infty\), \(\lim_{N\to \infty} \REP(\boldsymbol \theta_N) /N=\infty\)) upon scaling by the corresponding number of variables in a distribution.
+of outcomes \(\boldsymbol x =(x_1,\ldots,x_N) \in\{\pm1\}^{N}\), \(\boldsymbol h =(h_1,\ldots,h_{N_{\mathcal{H}}}) \in\{\pm1\}^{N_{\mathcal{H}}}\) and parameters \(\ak{\thetaN} \equiv (\ak{\thetaN}^{\mathcal{V}}, \ak{\thetaN}^{\mathcal{H}}, \ak{\thetaN}^{\mathcal{VH}})\), with \(\theta_{i,\ak{\thetaidx}}^{\mathcal{V}}\), \(\theta_{j,\ak{\thetaidx}}^{\mathcal{H}}\) and \(\theta_{ij,\ak{\thetaidx}}^{\mathcal{VH}}\) denoting respective parameter components, \(1 \leq i \leq N\), \(1 \leq j \leq N_{\mathcal{H}}\).
+Due to the marginalization steps in defining the distribution \eqref{eq:RBM2} of \(\boldsymbol X\), note that \(\REP_{\boldsymbol X}(\ak{\thetaN})\) has no immediate analytical expression similar to that of \(\REP_{(\boldsymbol X, \boldsymbol H)}(\ak{\thetaN})\).
+For clarity, recall also that S-instability \eqref{eq:Sun} in each model type refers to a respective divergence (i.e., \(\lim_{N\to \infty} \REP_{(\boldsymbol X, \boldsymbol H)}(\ak{\thetaN}) /(N+N_{\mathcal{H}})=\infty\), \(\lim_{N\to \infty} \REP(\ak{\thetaN}) /N=\infty\)) upon scaling by the corresponding number of variables in a distribution.
 In the following, let \(|\boldsymbol y|_1 = \sum_{i=1}^d |y_i|\) denote the L1 norm of a generic vector \(\boldsymbol y =(y_1,\ldots,y_d)\), \(d \geq 1\).
 \BeginKnitrBlock{proposition}
-\protect\hypertarget{prp:prop2}{}{\label{prp:prop2} }Let \(P_{\boldsymbol \theta_N}\) denote a RBM-based data model \eqref{eq:RBM2} for \(N\geq 1\) visible variables \(\boldsymbol X \equiv \boldsymbol X_N\) derived from \(\tilde{P}_{\boldsymbol \theta_N}\) as the joint RBM distribution \eqref{eq:RBM1} of \((\boldsymbol X, \boldsymbol H)\) involving some number \(N_{\mathcal{H}} \equiv N_{N,\mathcal{H}}\geq 0\) of hidden variables \(\boldsymbol H \equiv \boldsymbol H_N\) and parameters \(\boldsymbol \theta_N \equiv (\boldsymbol \theta_N^{\mathcal{V}},\boldsymbol \theta_N^{\mathcal{H}}, \boldsymbol \theta_N^{\mathcal{VH}}) \in\mathbb{R}^{N}\times \mathbb{R}^{N_{\mathcal{H}}} \times \mathbb{R}^{N*N_{\mathcal{H}}}\).
+\protect\hypertarget{prp:prop2}{}{\label{prp:prop2} }Let \(P_{\ak{\thetaN}}\) denote a RBM-based data model \eqref{eq:RBM2} for \(N\geq 1\) visible variables \(\boldsymbol X \equiv \boldsymbol X_N\) derived from \(\tilde{P}_{\ak{\thetaN}}\) as the joint RBM distribution \eqref{eq:RBM1} of \((\boldsymbol X, \boldsymbol H)\) involving some number \(N_{\mathcal{H}} \equiv N_{N,\mathcal{H}}\geq 0\) of hidden variables \(\boldsymbol H \equiv \boldsymbol H_N\) and parameters \(\ak{\thetaN} \equiv (\ak{\thetaN}^{\mathcal{V}},\ak{\thetaN}^{\mathcal{H}}, \ak{\thetaN}^{\mathcal{VH}}) \in\mathbb{R}^{N}\times \mathbb{R}^{N_{\mathcal{H}}} \times \mathbb{R}^{N*N_{\mathcal{H}}}\).
 Then,
 \begin{enumerate}[(i)]
-\item the instability measure $\REP(\boldsymbol \theta_N)$ for the marginal model $P_{\boldsymbol \theta_N}$ of  $\boldsymbol X$ satisfies
+\item the instability measure $\REP(\ak{\thetaN})$ for the marginal model $P_{\ak{\thetaN}}$ of  $\boldsymbol X$ satisfies \ak{
     $$
-    \left| \REP(\boldsymbol \theta_N)  - \elt\right| \leq    N_{\mathcal{H}}  \log 2
-    $$
-    for
-    $$
-    \elt  \equiv   \max_{ \boldsymbol x} \max_{ \boldsymbol h  } f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)-\min_{ \boldsymbol x } \max_{ \boldsymbol h }f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)
-    $$
-    based on $f_{\boldsymbol \theta_N}$ from (\ref{eq:f}) with components $\boldsymbol x \in \{\pm 1\}^{N}, \boldsymbol h \in \{\pm 1\}^{N_{\mathcal{H}}}$.
-\item The instability measure $\REP_{(\boldsymbol X, \boldsymbol H)}(\boldsymbol \theta_N)\equiv \left(\max_{ \boldsymbol x} \max_{ \boldsymbol h  } f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)-\min_{ \boldsymbol x } \min_{ \boldsymbol h }f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)\right)$ for the joint model $\tilde{P}_{\boldsymbol \theta_N}$ of $(\boldsymbol X, \boldsymbol H)$ satisfies
+    \left| \REP(\ak{\thetaN})  - \elt\right| \leq    N_{\mathcal{H}}  \log 2 \text{ for } \elt  \equiv   \max_{ \boldsymbol x} \max_{ \boldsymbol h  } f_{\ak{\thetaN}} (\boldsymbol x, \boldsymbol h)-\min_{ \boldsymbol x } \max_{ \boldsymbol h }f_{\ak{\thetaN}} (\boldsymbol x, \boldsymbol h)
+    $$}
+    based on $f_{\ak{\thetaN}}$ from (\ref{eq:f}) with components $\boldsymbol x \in \{\pm 1\}^{N}, \boldsymbol h \in \{\pm 1\}^{N_{\mathcal{H}}}$.
+\item The instability measure $\REP_{(\boldsymbol X, \boldsymbol H)}(\ak{\thetaN})\equiv \left(\max_{ \boldsymbol x} \max_{ \boldsymbol h  } f_{\ak{\thetaN}} (\boldsymbol x, \boldsymbol h)-\min_{ \boldsymbol x } \min_{ \boldsymbol h }f_{\ak{\thetaN}} (\boldsymbol x, \boldsymbol h)\right)$ for the joint model $\tilde{P}_{\ak{\thetaN}}$ of $(\boldsymbol X, \boldsymbol H)$ satisfies
     \begin{align*}
-    2\Gam +  2|\boldsymbol \theta_N^{\mathcal{H}} |_{1} \; \geq \; \REP_{(\boldsymbol X, \boldsymbol H)}(\boldsymbol \theta_N)    & \geq 
-    2\max\big\{\Gam,   |\boldsymbol \theta_N^{\mathcal{H}} |_{1}\big\} \\
+    2\Gam +  2|\ak{\thetaN}^{\mathcal{H}} |_{1} \; \geq \; \REP_{(\boldsymbol X, \boldsymbol H)}(\ak{\thetaN})    & \geq 
+    2\max\big\{\Gam,   |\ak{\thetaN}^{\mathcal{H}} |_{1}\big\} \\
     &\geq  2\Gam\\
     &\geq  \elt\\
-    &\geq   \max\big\{  \Gamc, \, \Gam  - 2|\boldsymbol \theta_N^{\mathcal{H}} |_{1}  \big\}
+    &\geq   \max\big\{  \Gamc, \, \Gam  - 2|\ak{\thetaN}^{\mathcal{H}} |_{1}  \big\}
     \end{align*}
@@
 
-In Proposition \ref{prp:prop2}(iii), the relationships between RBM models with regard to instability, and the effects of different parameter types, follow from the bounds on model instability measures in Proposition \ref{prp:prop2}(i)-(ii).
-Generally speaking, all instability in the marginal RBM model for the data \(\boldsymbol X\) can be attributed to an excessively large model quantity \(\elt\), which predominantly follows when main \(\boldsymbol \theta_N^{\mathcal{V}}\) and interaction \(\boldsymbol \theta_N^{\mathcal{VH}}\) parameters related to visible variables are too large in magnitude (e.g., upon accumulation in terms such as \(|\boldsymbol \theta_N^{\mathcal{V}}|_1\), \(\Gam\) or \(\Gamc\)).
-For example, for any bounded sequence \(|\boldsymbol \theta_N^{\mathcal{H}}|/N\) of hidden parameters, if main visible parameters \(\boldsymbol \theta_N^{\mathcal{V}}\) are too extreme (\(|\boldsymbol \theta_N^{\mathcal{V}}|_1/N\to \infty\)), this aspect will guarantee instability in the visible model (\(\elt/N\to \infty\)).
-In fact, the instability measure \(\elt \equiv \max_{ \boldsymbol x} \max_{ \boldsymbol h } f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)-\min_{ \boldsymbol x } \max_{ \boldsymbol h }f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)\) for marginal/visible model represents a clearly smaller portion of the instability measure \(\REP_{(\boldsymbol X, \boldsymbol H)}(\boldsymbol \theta_N)\equiv \max_{ \boldsymbol x} \max_{ \boldsymbol h } f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)-\min_{ \boldsymbol x } \min_{ \boldsymbol h } f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)\) in the joint RBM model, implying that an unstable marginal model (i.e., due to \(\boldsymbol \theta_N^{\mathcal{V}}\), \(\boldsymbol \theta_N^{\mathcal{VH}}\)) must always translate to an unstable joint model and that further potential causes of instability exist for the joint model, often due to the size \(|\boldsymbol \theta_N^{\mathcal{H}}|_1\).
-For example, while the joint RBM model for \((\boldsymbol X,\boldsymbol H)\) must always be unstable due to a diverging combination of visible and/or interaction parameters (\(|\boldsymbol \theta_N^{\mathcal{V}}|_1/N\to \infty\) or \(\Gam/N\to \infty\)) (Proposition \ref{prp:prop2}(iii.2)), instability for the joint model can also result when the main hidden parameters \(\boldsymbol \theta_N^{\mathcal{H}}\) become too large relative to sample size (\(|\boldsymbol \theta_N^{\mathcal{H}}|_1/N\to \infty\) in Proposition \ref{prp:prop2}(iii.5)).
-However, under Proposition \ref{prp:prop2}, the main hidden parameters \(\boldsymbol \theta_N^{\mathcal{H}}\) do not necessarily entail a source of instability for the marginal visible model.
-To explain this distinction, consider a joint model where all parameters related to visibles are zero, \(\boldsymbol \theta_N^{\mathcal{V}}= \boldsymbol \theta_N^{\mathcal{VH}}=\boldsymbol 0\), but the hidden-related parameters diverge in sum \(|\boldsymbol \theta_N^{\mathcal{H}}|_1/N\to \infty\).
-Here, the explosive behavior among parameters \(\boldsymbol \theta_N^{\mathcal{H}}\) induces instability in the joint model for \((\boldsymbol X, \boldsymbol H)\) but the marginal model for \(\boldsymbol X\), however, has a perfectly stable (and in fact uniform) distribution in this case.
-When the hidden parameters are bounded relative to the sample size (\(\sup_{N\geq 1} |\boldsymbol \theta_N^{\mathcal{H}}|_1/N<\infty\)), then all instability in both the joint and marginal RBM models can be directly linked to excessively large visible \(\boldsymbol \theta_N^{\mathcal{V}}\) and/or interaction parameters \(\boldsymbol \theta_N^{\mathcal{VH}}\) so that features of stability/instability must be the same across both models (Proposition \ref{prp:prop2}(iii.6)).
-Hence, to prevent instability in the joint model, the combined magnitudes of all parameters \(\theta_N\) must be controlled (cf.~Proposition \ref{prp:prop2}(iii.7)), while a stable visible data model technically results in constraining only the sizes of visible-related parameters \(\boldsymbol \theta_N^{\mathcal{V}}\), \(\boldsymbol \theta_N^{\mathcal{VH}}\).
-Nevertheless, because the joint model is often employed in practice for purposes of simulation and simulation-based inference, it is still reasonable to consider parameter choices for ensuring a stable joint model (and, consequently, a stable visible model as well).
-Further evidence of this is seen in the following numerical example.
+\ak{With regard to instability and the effects of different parameter types, the relationships between RBM models in Proposition \ref{prp:prop2}(iii)} follow from the bounds on model instability measures in Proposition \ref{prp:prop2}(i)-(ii).
+Generally speaking, all instability in the marginal RBM model for the data \(\boldsymbol X\) can be attributed to an excessively large model quantity \(\elt\), which predominantly follows when main \(\ak{\thetaN}^{\mathcal{V}}\) and interaction \(\ak{\thetaN}^{\mathcal{VH}}\) parameters related to visible variables are too large in magnitude (e.g., upon accumulation in terms such as \(|\ak{\thetaN}^{\mathcal{V}}|_1\), \(\Gam\) or \(\Gamc\)).
+For example, for any bounded sequence \(|\ak{\thetaN}^{\mathcal{H}}|/N\) of hidden parameters, \ak{main visible parameters $\ak{\thetaN}^{\mathcal{V}}$ that are too extreme ($|\ak{\thetaN}^{\mathcal{V}}|_1/N\to \infty$) will guarantee} instability in the visible model (\(\elt/N\to \infty\)).
+In fact, the instability measure \(\elt \equiv \max_{ \boldsymbol x} \max_{ \boldsymbol h } f_{\ak{\thetaN}} (\boldsymbol x, \boldsymbol h)-\min_{ \boldsymbol x } \max_{ \boldsymbol h }f_{\ak{\thetaN}} (\boldsymbol x, \boldsymbol h)\) for marginal/visible model represents a clearly smaller portion of the instability measure \(\REP_{(\boldsymbol X, \boldsymbol H)}(\ak{\thetaN})\equiv \max_{ \boldsymbol x} \max_{ \boldsymbol h } f_{\ak{\thetaN}} (\boldsymbol x, \boldsymbol h)-\min_{ \boldsymbol x } \min_{ \boldsymbol h } f_{\ak{\thetaN}} (\boldsymbol x, \boldsymbol h)\) in the joint RBM model\ak{.
+This implies} that an unstable marginal model (i.e., due to \(\ak{\thetaN}^{\mathcal{V}}\), \(\ak{\thetaN}^{\mathcal{VH}}\)) must always translate to an unstable joint model and that further potential causes of instability exist for the joint model, often due to the size \(|\ak{\thetaN}^{\mathcal{H}}|_1\).
 
 \par
+
+\ak{While} the joint RBM model for \((\boldsymbol X,\boldsymbol H)\) must always be unstable due to a diverging combination of visible and/or interaction parameters (\(|\ak{\thetaN}^{\mathcal{V}}|_1/N\to \infty\) or \(\Gam/N\to \infty\)) (Proposition \ref{prp:prop2}(iii.2)), instability for the joint model can also result when the main hidden parameters \(\ak{\thetaN}^{\mathcal{H}}\) become too large relative to sample size (\(|\ak{\thetaN}^{\mathcal{H}}|_1/N\to \infty\) in Proposition \ref{prp:prop2}(iii.5)).
+However, under Proposition \ref{prp:prop2}, the main hidden parameters \(\ak{\thetaN}^{\mathcal{H}}\) do not necessarily entail a source of instability for the marginal visible model\ak{.
+%% To \ak{illustrate} this distinction, consider a joint model where all parameters related to visibles are zero, $\ak{\thetaN}^{\mathcal{V}}= \ak{\thetaN}^{\mathcal{VH}}=\boldsymbol 0$, but the hidden-related parameters diverge in sum $|\ak{\thetaN}^{\mathcal{H}}|_1/N\to \infty$.
+Here, the explosive behavior among parameters $\ak{\thetaN}^{\mathcal{H}}$ induces instability in the joint model for $(\boldsymbol X, \boldsymbol H)$\ak{.
+The} marginal model for $\boldsymbol X$, however, has a perfectly stable (and in fact uniform) distribution in this case.
+When} the hidden parameters are bounded relative to the sample size (\(\sup_{N\geq 1} |\ak{\thetaN}^{\mathcal{H}}|_1/N<\infty\)), then all instability in both the joint and marginal RBM models can be directly linked to excessively large visible \(\ak{\thetaN}^{\mathcal{V}}\) and/or interaction parameters \(\ak{\thetaN}^{\mathcal{VH}}\) so that features of stability/instability must be the same across both models (Proposition \ref{prp:prop2}(iii.6)).
+Hence, to prevent instability in the joint model, the combined magnitudes of all parameters \(\ak{\thetaN}\) must be controlled (cf.~Proposition \ref{prp:prop2}(iii.7)), while a stable visible data model technically results \ak{from} constraining only the sizes of visible-related parameters \(\ak{\thetaN}^{\mathcal{V}}\), \(\ak{\thetaN}^{\mathcal{VH}}\).
+Nevertheless, because the joint model \ak{often is} employed in practice for purposes of simulation and simulation-based inference, it is still reasonable to consider parameter choices for ensuring a stable joint model (and, consequently, a stable visible model as well).
+Further evidence of this is seen in the following numerical example.
 \begin{figure}
@@
 \end{figure}
-In our numerical experiment, we allow the two types of terms (main effects terms corresponding to visible and hidden parameters \(\boldsymbol \theta_{main} = (\boldsymbol \theta_N^{\mathcal{V}}, \boldsymbol \theta_N^{\mathcal{H}})\) and interaction parameters \(\boldsymbol \theta_N^{\mathcal{VH}}\)) to have varying average magnitudes, \(||\boldsymbol \theta_{main} || /(N_{\mathcal{H}}+N_{\mathcal{V}})\) and \(||\boldsymbol \theta_{interaction} || /(N_{\mathcal{H}}*N_{\mathcal{V}})\) for a RBM with \(N_\mathcal{V} = 9\) visibles and \(N_\mathcal{H} = 5\) hiddens.
+\par
+
+In our numerical experiment, we allow the two types of terms (main effects terms corresponding to visible and hidden parameters \(\boldsymbol \theta_{main} = (\ak{\thetaN}^{\mathcal{V}}, \ak{\thetaN}^{\mathcal{H}})\) and interaction parameters \(\ak{\thetaN}^{\mathcal{VH}}\)) to have varying average magnitudes, \(||\boldsymbol \theta_{main} || /(N_{\mathcal{H}}+N_{\mathcal{V}})\) and \(||\boldsymbol \theta_{interaction} || /(N_{\mathcal{H}}*N_{\mathcal{V}})\) for a RBM with \(N_\mathcal{V} = 9\) visibles and \(N_\mathcal{H} = 5\) hiddens.
 These average magnitudes vary on a grid between \(0.001\) and \(3\) with \(20\) breaks, yielding \(400\) grid points.
 At each point in the grid, \(100\) vectors (\(\boldsymbol \theta_{main}\)) are sampled uniformly on a sphere with radius corresponding to the first coordinate in the grid and \(100\) vectors (\(\boldsymbol \theta_{interaction}\)) are sampled uniformly on a sphere with radius corresponding to the second coordinate in the grid via sums of squared and scaled iid Normal\((0, 1)\) variables.
-These vectors are then paired to create \(100\) values of \(\boldsymbol \theta_N\) with magnitudes at each point in the grid.
-The values \(\REP(\boldsymbol \theta_N)/N_{\mathcal{V}}\) and \(\DN(\boldsymbol \theta_N)\) are then calculated for each \(\boldsymbol \theta_N\) and then summarized for each point in the grid using the sample mean.
+These vectors are then paired to create \(100\) values of \(\ak{\thetaN}\) with magnitudes at each point in the grid.
+The values \(\REP(\ak{\thetaN})/N_{\mathcal{V}}\) and \(\DN(\ak{\thetaN})\) are then calculated for each \(\ak{\thetaN}\) and then summarized for each point in the grid using the sample mean.
 The results of this numerical study are shown in Figure \ref{fig:rbm-plots}.
-From these two plots, it is clear that for larger magnitudes of the parameter vectors, there is evidence of S-instability in that the log-ratio of extremal probabilities scaled by \(N_{\mathcal{V}}\) and the the biggest log-probability ratio for a one-component change in data outcomes are both increasing away from \(\boldsymbol \theta_N = \boldsymbol 0\), further supporting \ref{prp:prop2}(iii.2 and iii.5).
+From these two plots, it is clear that for larger magnitudes of the parameter vectors, there is evidence of S-instability in that the log-ratio of extremal probabilities scaled by \(N_{\mathcal{V}}\) and the the biggest log-probability ratio for a one-component change in data outcomes are both increasing away from \(\ak{\thetaN} = \boldsymbol 0\), further supporting \ref{prp:prop2}(iii.2 and iii.5).
+
+\par
 
@@
 Consider the following two models -- a RBM with $N$ visible nodes and one hidden layer with $N_H$ nodes and a deep RBM with $N$ visible nodes and $2$ layers, with $N_{H,1}$ and $N_{H,2}$ hidden nodes respectively such that $N_{H,1} + N_{H,2} = N_H$.
-The length of the parameter vector $\boldsymbol \theta_N$ in the single layer RBM is then equal to $N + N_H + N*N_H$, whereas the length of the parameter vector in the two layer RBM is $N + N_{H,1} + N_{H,2} + N*N_{H,1} + N_{H,1}*N_{H,2} = N + N_H + N*N_H + N_{H,2}(N_{H,1} - N)$.
+The length of the parameter vector $\ak{\thetaN}$ in the single layer RBM is then equal to $N + N_H + N*N_H$, whereas the length of the parameter vector in the two layer RBM is $N + N_{H,1} + N_{H,2} + N*N_{H,1} + N_{H,1}*N_{H,2} = N + N_H + N*N_H + N_{H,2}(N_{H,1} - N)$.
 If $N_{H,1} > N$, then the deep model will have more parameters, if $N_{H,1} < N$, then the deep model will have less parameters, and if $N_{H,1} = N$, then the deep model will the same number of parameters as the single layer RBM.} S-unstable joint models will similarly follow if the combined sizes of all parameters are too great relative to the total number of variables, while instability in the data model for visible variables will depend only on the main or interaction parameters directly related to visibles \ak{(i.e.
@@
 
-Due to \ak{induced sensitivity in the probability structure of FOES models, S-instability} may often translate to numerical complications, and in fact obstructions, in both simulation and statistical inference based on likelihoods.
-We describe these aspects in Sections \ref{mcmc}-\ref{bayes} with regard to data simulation, maximum likelihood estimation and Bayes inference, respectively.
+Due to \ak{sensitivity in the probability structure of an unstable FOES model, S-instability} may often translate to numerical complications, and in fact obstructions, in both simulation and statistical inference based on likelihoods.
+We describe these aspects in Sections \ref{ml}-\ref{bayes} with regard to data simulation, maximum likelihood estimation and Bayes inference.
 
-\hypertarget{mcmc}{%
-\subsection{Implications for Simulation}\label{mcmc}}
-
-Suppose one aims to apply MCMC to simulate data \(\boldsymbol X=(X_1,\ldots,X_n)\) from a FOES model \(P_{\boldsymbol \theta_N}(\boldsymbol x)\), \(\boldsymbol x \in\mathcal{X}^N\) whereby a chain is constructed with \(P_{\boldsymbol \theta_N}\) as the stationary distribution.
-For an unstable FOES model, one-component changes in outcomes may produce
-radically different probabilities, which may then impose numerical barriers to MCMC.
-For example, consider implementation of a Gibbs sampler from the full conditional distributions \(P_{\boldsymbol \theta_N} (X_i = x| \boldsymbol x_{-i})\), \(x\in\mathcal{X}\), of each variable \(X_i\) based on values \(\boldsymbol x_{-i}\in \mathcal{X}^{N-1}\) for the remaining variables, say \(\boldsymbol X_{-i}\), in \(\boldsymbol X\).
-If a single change in \(X_i\) from one value \(x_1\) to another \(x_2\) may produce two outcomes \(\boldsymbol x^{(1)}\) and \(\boldsymbol x^{(2)}\) for \(\boldsymbol X\) with vastly different probabilities under the joint distribution \(P_{\boldsymbol \theta_N}\), then the Gibbs sampler can have extreme log-ratios in its transition probabilities,
-\[
-\log \left| \frac{P_{\boldsymbol \theta_N} (X_i =x_1 | \boldsymbol x^{(1)}_{-i})}{P_{\boldsymbol \theta_N} (X_i = x_2| \boldsymbol x^{(2)}_{-i}) } \right|=\log \left| \frac{P_{\boldsymbol \theta_N} ( \boldsymbol x^{(1)} )}{P_{\boldsymbol \theta_N} (  \boldsymbol x^{(2)} ) } \right|,
-\]\\
-as conditional probabilities are proportional to joint probabilities that, with unstable models, can have unbounded log-probability ratios in one-component changes (Theorem
-\ref{thm:instab-elpr}).
-This can hinder the ability of a chain to effectively explore the sample space of the observations \(\boldsymbol X\), as the chain may mix poorly by moving rapidly to, and slowly away from, sections of the sample space.
-In this case, for example, the Markov chain may become entrapped within a mode of the probability function, with rare chance of escaping to adequately mimic the occupation frequencies in the overall sample space.
-If modes of the unstable model are not unique, then important outcomes may be missed without multiple chains or impractically enormous numbers of MCMC samples.
-This mixing problem is due to the unstable stationary distribution (unbounded ratios of probabilities under the joint model), rather than in any particulars of the MCMC algorithm, and similar complications can also arise for Metropolis-Hastings algorithms for MCMC.
-Hence, while an unstable FOES model \(P_{\boldsymbol \theta_N}\) may be valid and technically open to simulation by MCMC, the aspect of instability can render applications of MCMC as numerically infeasible for simulation purposes.
-This result is in line with conclusions of \citet{handcock2003assessing} and \citet{schweinberger2011instability} for other exponential models.
-
-\hypertarget{implications-for-maximum-likelihood-inference}{%
-\subsection{Implications for Maximum Likelihood Inference}\label{implications-for-maximum-likelihood-inference}}
+\hypertarget{ml}{%
+\subsection{Implications for Maximum Likelihood Inference}\label{ml}}
 
 Volatility in the probability structure of an unstable model can also hamper efforts to maximize likelihood functions in statistical inference.
-When a FOES model is unstable along a parameter sequence \(\boldsymbol \theta_N\), the same model can further be unstable along parameters \(-\boldsymbol \theta_N\) in an opposite direction from the
-origin (model condition PSR and Corollary \ref{cor:sign}). This can translate into
-potential sensitivity of the likelihood function around zero, and lead to numerical complications in maximizing the objective function.
+When a FOES model is unstable along a parameter sequence \(\ak{\thetaN}\), the same model can further be unstable along parameters \(-\ak{\thetaN}\) in an opposite direction from the origin (model condition PSR and Corollary \ref{cor:sign}).
+This can translate into potential sensitivity of the likelihood function around zero, and lead to numerical complications in maximizing the objective function.
 We next provide a discussion of this issue in a way that builds upon and extends related findings by \citet{schweinberger2011instability}, who largely focused on the case of one-parameter exponential models.
 
-With many probability models, the modes and anti-modes in the probability structure under one parameter \(\boldsymbol \theta_N\) are reversed in role when the parameter sign changes \(-\boldsymbol \theta_N\).
+With many probability models, the modes and anti-modes in the probability structure under one parameter \(\ak{\thetaN}\) are reversed in role when the parameter sign changes \(-\ak{\thetaN}\).
 Because unstable models tend to degeneracy, the opposite signed parameters further push unstable models to assign nearly all probability to extremely opposite data configurations, given by modes/anti-modes.
@@
 \BeginKnitrBlock{theorem}
-\protect\hypertarget{thm:something}{}{\label{thm:something} }Let \(P_{\boldsymbol \theta_N}\), \(N \geq 1\), denote an S-unstable FOES model sequence, which additionally satisfies model condition PSR.
-Let \(\boldsymbol{x}_{\max, \boldsymbol \theta_N},\boldsymbol{x}_{\min, \boldsymbol \theta_N}\in\mathcal{X}^N\) denote, respectively, a mode and anti-mode of the model \(P_{\boldsymbol \theta_N}(\boldsymbol x)\), \(\boldsymbol x\in\mathcal{X}^N\), for \(N\) observations \(\boldsymbol X = (X_1,\ldots,X_N)\), whereby \(P_{\boldsymbol \theta_N }(\boldsymbol{x}_{\max, \boldsymbol \theta_N}) = \max_{\boldsymbol y\in\mathcal{X}^N} P_{\boldsymbol \theta_N }(\boldsymbol y)\) and \(P_{\boldsymbol \theta_N }(\boldsymbol{x}_{\min, \boldsymbol \theta_N}) = \min_{\boldsymbol y\in\mathcal{X}^N} P_{\boldsymbol \theta_N }(\boldsymbol y)\).
+\protect\hypertarget{thm:something}{}{\label{thm:something} }Let \(P_{\ak{\thetaN}}\), \(N \geq 1\), denote an S-unstable FOES model sequence, which additionally satisfies model condition PSR.
+Let \(\boldsymbol{x}_{\max, \ak{\thetaN}},\boldsymbol{x}_{\min, \ak{\thetaN}}\in\mathcal{X}^N\) denote, respectively, a mode and anti-mode of the model \(P_{\ak{\thetaN}}(\boldsymbol x)\), \(\boldsymbol x\in\mathcal{X}^N\), for \(N\) observations \(\boldsymbol X = (X_1,\ldots,X_N)\), whereby \(P_{\ak{\thetaN} }(\boldsymbol{x}_{\max, \ak{\thetaN}}) = \max_{\boldsymbol y\in\mathcal{X}^N} P_{\ak{\thetaN} }(\boldsymbol y)\) and \(P_{\ak{\thetaN} }(\boldsymbol{x}_{\min, \ak{\thetaN}}) = \min_{\boldsymbol y\in\mathcal{X}^N} P_{\ak{\thetaN} }(\boldsymbol y)\).
 
@@
 \[
-\frac{1}{\REP(\boldsymbol \theta_N)} \log\left[ \frac{P_{\boldsymbol \theta_N }(\boldsymbol X) }{ \displaystyle{\min_{\boldsymbol y\in\mathcal{X}^N}} P_{\boldsymbol \theta_N }(\boldsymbol y)}\right] = \frac{ \log P_{\boldsymbol \theta_N } ( \boldsymbol X)  - \log P_{\boldsymbol \theta_N }(\boldsymbol{x}_{\min, \boldsymbol \theta_N}) }
-{ \log P_{\boldsymbol \theta_N } (\boldsymbol{x}_{\max, \boldsymbol \theta_N})  - \log P_{\boldsymbol \theta_N }(\boldsymbol{x}_{\min, \boldsymbol \theta_N}) }
+\frac{1}{\REP(\ak{\thetaN})} \log\left[ \frac{P_{\ak{\thetaN} }(\boldsymbol X) }{ \displaystyle{\min_{\boldsymbol y\in\mathcal{X}^N}} P_{\ak{\thetaN} }(\boldsymbol y)}\right] = \frac{ \log P_{\ak{\thetaN} } ( \boldsymbol X)  - \log P_{\ak{\thetaN} }(\boldsymbol{x}_{\min, \ak{\thetaN}}) }
+{ \log P_{\ak{\thetaN} } (\boldsymbol{x}_{\max,\ak{\thetaN}})  - \log P_{\ak{\thetaN} }(\boldsymbol{x}_{\min, \ak{\thetaN}}) }
 \stackrel{p, \mathrm{E}}{\longrightarrow}
@@
 \]
-under \(\boldsymbol\theta_N\) while
+under \(\ak{\thetaN}\) while
 \[
-\frac{1}{\REP(-\boldsymbol \theta_N)} \log\left[ \frac{P_{-\boldsymbol \theta_N }(\boldsymbol X) }{  \displaystyle{\min_{\boldsymbol y\in\mathcal{X}^N}} P_{-\boldsymbol \theta_N }(\boldsymbol y)} \right]
-=    \frac{ \log P_{-\boldsymbol \theta_N } ( \boldsymbol X) - \log P_{ \boldsymbol \theta_N }(\boldsymbol{x}_{\max, \boldsymbol \theta_N})    }
-{ \log P_{-\boldsymbol \theta_N } (\boldsymbol{x}_{\min, \boldsymbol \theta_N})  - \log P_{-\boldsymbol \theta_N }(\boldsymbol{x}_{\max, \boldsymbol \theta_N}) }
+\frac{1}{\REP(-\ak{\thetaN})} \log\left[ \frac{P_{-\ak{\thetaN} }(\boldsymbol X) }{  \displaystyle{\min_{\boldsymbol y\in\mathcal{X}^N}} P_{-\ak{\thetaN} }(\boldsymbol y)} \right]
+=    \frac{ \log P_{-\ak{\thetaN} } ( \boldsymbol X) - \log P_{\ak{\thetaN} }(\boldsymbol{x}_{\max, \ak{\thetaN}})    }
+{ \log P_{-\ak{\thetaN} } (\boldsymbol{x}_{\min, \ak{\thetaN}})  - \log P_{-\ak{\thetaN} }(\boldsymbol{x}_{\max, \ak{\thetaN}}) }
 \stackrel{p,\mathrm{E}}{\longrightarrow}
@@
 \]
-under \(-\boldsymbol\theta_N\), where
+under \(-\ak{\thetaN}\), where
 \[
-\REP( \boldsymbol \theta_N) \equiv \log \frac{P_{\boldsymbol \theta_N } (\boldsymbol{x}_{\max, \boldsymbol \theta_N})}{P_{\boldsymbol \theta_N }(\boldsymbol{x}_{\min, \boldsymbol \theta_N})} = \log  \frac{P_{-\boldsymbol \theta_N } (\boldsymbol{x}_{\min, \boldsymbol \theta_N}) }{ P_{-\boldsymbol \theta_N }(\boldsymbol{x}_{\max, \boldsymbol \theta_N})}=   \REP( -\boldsymbol \theta_N), \quad N \geq 1.
+\REP( \ak{\thetaN}) \equiv \log \frac{P_{\ak{\thetaN} } (\boldsymbol{x}_{\max, \ak{\thetaN}})}{P_{\ak{\thetaN} }(\boldsymbol{x}_{\min, \ak{\thetaN}})} = \log  \frac{P_{-\ak{\thetaN} } (\boldsymbol{x}_{\min, \ak{\thetaN}}) }{ P_{-\ak{\thetaN} }(\boldsymbol{x}_{\max, \ak{\thetaN}})}=   \REP( -\ak{\thetaN}), \quad N \geq 1.
 \]
@@
 
-Theorem \ref{thm:something} \ak{implies} log-likelihood functions based on unstable models are both inversely related and degenerate at opposited signed parameters \(\boldsymbol \theta_N\) or \(-\boldsymbol \theta_N\), so that likelihoods are highest at different extremes in data configuration (e.g., \(\boldsymbol{x}_{\max, \boldsymbol \theta_N}\) under \(\boldsymbol \theta\)-probabilities or \(\boldsymbol{x}_{\min, \boldsymbol \theta_N}\) under \(-\boldsymbol \theta\)-probabilities).
-If the observed outcome \(\boldsymbol x\) for data \(\boldsymbol X\) is not a mode/anti-mode, then probabilities for the outcome may be small under both parameters \(\boldsymbol \theta_N\) and \(-\boldsymbol \theta_N\), in which case associated optimization steps may then shift around zero and struggle to converge.
+Theorem \ref{thm:something} \ak{implies} log-likelihood functions based on unstable models are both inversely related and degenerate at opposited signed parameters \(\ak{\thetaN}\) or \(-\ak{\thetaN}\), so that likelihoods are highest at different extremes in data configuration (e.g., \(\boldsymbol{x}_{\max, \ak{\thetaN}}\) under \(\boldsymbol \theta\)-probabilities or \(\boldsymbol{x}_{\min, \ak{\thetaN}}\) under \(-\boldsymbol \theta\)-probabilities).
+If the observed outcome \(\boldsymbol x\) for data \(\boldsymbol X\) is not a mode/anti-mode, then probabilities for the outcome may be small under both parameters \(\ak{\thetaN}\) and \(-\ak{\thetaN}\), in which case associated optimization steps may then shift around zero and struggle to converge.
 
-In many model formulations, the zero parameter \(\boldsymbol \theta_N=\boldsymbol 0\) is a ``safe'' position among parameters, representing a guaranteed stable model (having uniform probabilities among outcomes), which can also tether a broad parameter search attempted among unstable models.
+In many model formulations, the zero parameter \(\ak{\thetaN}=\boldsymbol 0\) is a ``safe'' position among parameters, representing a guaranteed stable model (having uniform probabilities among outcomes), which can also tether a broad parameter search attempted among unstable models.
 \citet{handcock2003assessing} describes similar results for degenerate exponential models, and Theorem \ref{thm:something} also supports an important finding of \citet[Corollary 1]{schweinberger2011instability} for one-parameter linear exponential models \eqref{eq:expo}.
-In the latter case, the likelihood score function at \(\boldsymbol \theta_N\) is the expected value \(\mu(\boldsymbol \theta_N)\equiv \mathrm{E}_{\boldsymbol \theta_N} g(\boldsymbol X)\) of the sufficient statistic \(g(\cdot)\), and optimization involves solving \(\mu(\cdot)=g(\boldsymbol x)\) for an observed outcome \(\boldsymbol x\).
+In the latter case, the likelihood score function at \(\ak{\thetaN}\) is the expected value \(\mu( \ak{\thetaN})\equiv \mathrm{E}_{\ak{\thetaN}} g(\boldsymbol X)\) of the sufficient statistic \(g(\cdot)\), and optimization involves solving \(\mu(\cdot)=g(\boldsymbol x)\) for an observed outcome \(\boldsymbol x\).
 For unstable models in this exponential class, \citet[Corollary 1]{schweinberger2011instability} shows
@@
 \[
-\lim_{N\to \infty}\frac{\mu(\boldsymbol \theta_N) -L_n}{U_n-L_N}= \left\{ \begin{array}{lcl}
-1 && \text{for } \boldsymbol \theta_N>0,\\
-0 && \text{for } \boldsymbol \theta_N<0, \end{array}\right.
+\lim_{N\to \infty}\frac{\mu(\ak{\thetaN}) -L_n}{U_n-L_N}= \left\{ \begin{array}{lcl}
+1 && \text{for } \ak{\thetaN}>0,\\
+0 && \text{for } \ak{\thetaN}<0, \end{array}\right.
 \]
@@
 As described by \citet{schweinberger2011instability}, the implication for maximum likelihood estimation is that, unless an observed outcome \(\boldsymbol x\) falls at an extreme \(U_N, L_N\) (i.e., modes/anti-modes), optimization steps in the parameter space can iterate in relatively small increments around zero and fail to converge.
-For unstable one-parameter exponential models, the maximum likelihood results of \citet{schweinberger2011instability} turn out to be a special case of Theorem \ref{thm:something} and the LREP expansion \eqref{eq:UL} in this setting; namely, for an unstable model with \(\boldsymbol \theta_N>0\),
+For unstable one-parameter exponential models, the maximum likelihood results of \citet{schweinberger2011instability} turn out to be a special case of Theorem \ref{thm:something} and the LREP expansion \eqref{eq:UL} in this setting; namely, for an unstable model with \(\ak{\thetaN}>0\),
 \[
-\frac{1}{\REP(\boldsymbol \theta_N)} \log\left[ \frac{P_{\boldsymbol \theta_N }(\boldsymbol X) }{\displaystyle{\min_{\boldsymbol y\in\mathcal{X}^N}} P_{\boldsymbol \theta_N }(\boldsymbol y)}\right] = \frac{\boldsymbol g(\boldsymbol X) - L_N}{U_N-L_N}\;\;\stackrel{p, \mathrm{E}}{\longrightarrow}\;\; 1
+\frac{1}{\REP(\ak{\thetaN})} \log\left[ \frac{P_{\ak{\thetaN} }(\boldsymbol X) }{\displaystyle{\min_{\boldsymbol y\in\mathcal{X}^N}} P_{ \ak{\thetaN} }(\boldsymbol y)}\right] = \frac{\boldsymbol g(\boldsymbol X) - L_N}{U_N-L_N}\;\;\stackrel{p, \mathrm{E}}{\longrightarrow}\;\; 1
 \]
-holds as \(N\to \infty\) by Theorem \ref{thm:something}, while under \(-\boldsymbol \theta_N<0\)
+holds as \(N\to \infty\) by Theorem \ref{thm:something}, while under \(-\ak{\thetaN}<0\)
 \[
-\frac{1}{\REP(-\boldsymbol \theta_N)} \log\left[ \frac{P_{-\boldsymbol \theta_N }(\boldsymbol X) }{\displaystyle{\min_{\boldsymbol y\in\mathcal{X}^N}} P_{-\boldsymbol \theta_N }(\boldsymbol y)}\right] = \frac{U_N -\boldsymbol g(\boldsymbol X)}{U_N-L_N} = 1 - \frac{\boldsymbol g(\boldsymbol X) - L_N}{U_N-L_N} \;\;\stackrel{p, \mathrm{E}}{\longrightarrow}\;\; 1.
+\frac{1}{\REP(-\ak{\thetaN})} \log\left[ \frac{P_{-\ak{\thetaN} }(\boldsymbol X) }{\displaystyle{\min_{\boldsymbol y\in\mathcal{X}^N}} P_{-\ak{\thetaN} }(\boldsymbol y)}\right] = \frac{U_N -\boldsymbol g(\boldsymbol X)}{U_N-L_N} = 1 - \frac{\boldsymbol g(\boldsymbol X) - L_N}{U_N-L_N} \;\;\stackrel{p, \mathrm{E}}{\longrightarrow}\;\; 1.
 \]
@@
 Considering that the degeneracy issues related to unstable models can cause likelihoods can be flat (e.g., near zero) for many parameters under a given data outcome and that sign changes in parameters can shift tremendous probability to extreme and opposite outcomes in the sample space (e.g., Corollary \ref{cor:sign}, Theorem \ref{thm:something}), then numerical complications may arise with Bayes inference in sampling a posterior parameter space based on MCMC.
-The potential challenges in chain mixing are similar to those presented in Section \ref{mcmc}, though in chain movements through the parameter space as opposed to the sample space in data generation.
-That is, in the Bayes setting for sampling a posterior distribution for \(\boldsymbol \theta_N\), a chain may \ak{be unable} to effectively explore the parameter space due partly to extreme and potentially unbounded probability ratios from parameter sign changes, which represents a parameter space analog to how one-component changes in the sample space may impact data simulation with unstable models.
-For example, if \(\pi(\cdot)\) denotes a prior density for \(\boldsymbol \theta_N\) and \(q(\cdot | \cdot)\) denotes a proposal distribution for use in a Metropolis-Hastings (MH) sampler, then MH acceptance probability becomes
+\ak{Instability may hinder effective chain mixing, related to the findings of \citet{handcock2003assessing} and \citet{schweinberger2011instability} with exponential graph models.
+For example, a Markov chain may become entrapped within a mode, and fail to adequately mimic the overall occupation frequencies required for a reasonable MCMC distributional approximaton.
+If modes of the unstable model are not unique, then important outcomes may be missed without multiple chains or impractically enormous numbers of MCMC samples.
+This mixing problem is due to the unstable stationary distribution (unbounded ratios of probabilities under the joint model), rather than in any particulars of the MCMC algorithm.
+Hence,} in the Bayes setting for sampling a posterior distribution for \(\ak{\thetaN}\), a chain may \ak{be unable} to effectively explore the parameter space due partly to extreme and potentially unbounded probability ratios from parameter sign changes.
+
+For example, if \(\pi(\cdot)\) denotes a prior density for \(\ak{\thetaN}\) and \(q(\cdot | \cdot)\) denotes a proposal distribution for use in a Metropolis-Hastings (MH) sampler, then MH acceptance probability becomes
 \[
 \alpha\left(\boldsymbol \theta^{(1)} \mid \boldsymbol \theta^{(2)}\right)= \min\left\{1,
-\frac{q(\boldsymbol \theta_N^{(2)} \mid \boldsymbol \theta_N^{(1)})}{q(\boldsymbol \theta_N^{(1)} \mid \boldsymbol \theta_N^{(2)})}
-\frac{P_{\boldsymbol \theta_N^{(1)}} ( \boldsymbol x ) \pi(\boldsymbol \theta_N^{(1)}) }{P_{\boldsymbol \theta^{(2)}_N} (  \boldsymbol x) \pi(\boldsymbol \theta_N^{(2)}) } \right\},
+\frac{q(\ak{\thetaN}^{(2)} \mid \ak{\thetaN}^{(1)})}{q(\ak{\thetaN}^{(1)} \mid  \ak{\thetaN}^{(2)})}
+\frac{P_{\ak{\thetaN}^{(1)}} ( \boldsymbol x ) \pi(\ak{\thetaN}^{(1)}) }{P_{\ak{\thetaN}^{(2)}} (  \boldsymbol x) \pi(\ak{\thetaN}^{(2)}) } \right\},
 \]
-which indicates how parameter sensitivity in the likelihood \(P_{\boldsymbol \theta_N} ( \boldsymbol x)\) may complicate sampling of the posterior \(P_{\boldsymbol \theta_N} ( \boldsymbol x) \pi(\boldsymbol \theta_N)\) (i.e., moving from \(\boldsymbol \theta^{(1)}\) to \(\boldsymbol \theta^{(2)}\) in the parameter space).
+which indicates how parameter sensitivity in the likelihood \(P_{\ak{\thetaN}} (\boldsymbol x)\) may complicate sampling of the posterior \(P_{ \ak{\thetaN}} ( \boldsymbol x) \pi(\ak{\thetaN})\) (i.e., moving from \(\boldsymbol \theta^{(1)}\) to \(\boldsymbol \theta^{(2)}\) in the parameter space).
 Furthermore, the potential for model instability and the size of the parameter space can also become greater with the introduction of latent variables to existing data variables, as involved in some model formulations described in Sections \ref{rbm}-\ref{deep-learning}.
 As latent variables are often sampled with parameters in a Bayes MCMC approach, this aspect may further compound numerical problems in chain mixing.
+
+\hypertarget{stable-models}{%
+\subsection{Fitting Stable Models}\label{stable-models}}
+
+\ak{The instability results presented in Section \ref{instability-results} also suggest the potential use of regularization and penalization as a solution to avoid instability when fitting FOES models.
+One example of a rigorous approach to penalization can be found in \citet{kaplan2019properties}, which employs Bayesian fitting for a single layer RBM using multivariate Gaussian priors with constrained covariance matrices to control the sizes of the parameter values.
+For example, considering a latent variable model described in Section \ref{fixed-dim-exp}, Proposition \ref{prp:prop2}(iii) suggests how parameter contributions from sources of ``main hidden," ``main visible," and ``interaction" parameters may each be constrained (relative to the visible variable sample size $N$) to ensure model stability.
+In a Bayes inference framework, priors can be developed so that parameters are then appropriately bounded (in a stochastic sense) to lie in a parameter subset compatible with stability.
+For example, for a latent model with numbers $N_{\mathcal{H}}$, $N_{\mathcal{V}}$ and $N_{\mathcal{VH}}$ for hidden, visible and interaction parameters, respectively, Proposition \ref{prp:prop2}(iii.7) suggests a potential prior specification where parameters have root second moments bounded by the parameter type: $C N/N_{\mathcal{V}}$, $C N/N_{\mathcal{H}}$, and $C N/N_{\mathcal{VH}}$ for visible, hidden and interactions parameters with a constant $C>0$ (i.e., so that the condition in Proposition \ref{prp:prop2}(iii.7) holds in a stochastic sense).
+The same general strategy can apply with other models with results from Section \ref{illustrations}, and the adequacy of the subsequent fitted model might then be assessed.
+
+As opposed to approaches for fitting (e.g., penalization as above),  the findings here on model stability/instability also support alternative strategies to model formulation and re-pameterization using ``centered conditional" distributions \citep[cf.][]{kaiser2007statistical}; see \citet{kaiser2009exploring} and \citet{kaiser2012centered} for applications with spatial data and \citet{casleton2017local} for network data.
+In a joint model $P_{\thetaN}$,  these authors consider associated (full) conditional distributions $P_{\thetaN}(X_{i} = x_i | X_j=x_j, j\neq i)$ and re-parameterize by centering/averaging sums in the expression of  $P_{\thetaN}(X_{i} = x_i | X_j=x_j, j\neq i)$ in order to separate model effects between ``mean" and ``dependence" parameters.
+The intent is to improve model interpretation and help detect degeneracy (characterized by the ``mean parameters" in the model failing to match averages observed in data generations from the model and attributable to large ``dependence parameters").
+The works above describe advantages of such centered parameterizations based on empirical studies, but findings here also support these parameterizations as useful for avoiding model instability, where the corresponding ``centered" models can be checked for $\REP(\thetaN)/N$ ratios that are bounded over wider possible parameters than ``uncentered" counterpart models.
+For illustration, the appendix provides details on a graph model example.}
 
@@
 For a large class of models that covers a broad range of applications (including ``deep learning''), we have developed a formal definition of instability in model probability structure and elucidated multiple consequences of instability.
-We have shown for FOES models that instability manifests through small changes in data leading to potentially large changes in probability as well as the potential to place all probability on certain modal subsections of the sample space, which potentially could be small.
-Such instability is often due to a complex interaction between the model statistics used (i.e., how numerous and large these may become) and the number and magnitudes of
-parameters in the model formulation.
+We have shown for FOES models that instability manifests through \ak{a large amount of probability being placed on a subset of the theoretically possible realizations that may be narrower than intended.}
+Such instability is often due to a complex interaction between the model statistics used (i.e., how numerous and large these may become) and the number and magnitudes of parameters in the model formulation.
 For many FOES models, the possibility exists, at least in principle, to \ak{constrain} parameters in a way \ak{that} balances their potential contributions against those of model statistics in order to prevent probability instabilities.
-The FOES model class is quite broad and, in developing such models for large data sets, some caution should be used in parameter specification to control effects of model instability.
+\ak{While the results on instability here only address model formulations with finite sample spaces, the same underlying principles might be extended to models for continuous data as well by discretizing the model (i.e., binning outcomes into intervals), as data typically have some given level of discretization.
+Such a generalization could help toward understanding potential model instability in a wider class of data applications.
+As it sits currently, t}he FOES model class is quite broad and \ak{findings here can help in identifying undesirable probability features in such models.}
 
@@
 
-\textbf{Proof of Proposition \ref{thm:instab-elpr}.} For part (i), we prove the contrapositive, supposing that \(\DN(\boldsymbol \theta_N) \le C\) holds for some \(C > 0\) and show \(\REP(\boldsymbol \theta_N) \leq NC\).
-Let \(\boldsymbol x_{min} \equiv \argmin\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)\)
-and \(\boldsymbol x_{max} \equiv \argmax\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)\).
+\textbf{Proof of Theorem \ref{thm:instab-elpr}.} For part (i), we prove the contrapositive, supposing that \(\DN(\ak{\thetaN}) \le C\) holds for some \(C > 0\) and show \(\REP(\ak{\thetaN}) \leq NC\).
+Let \(\boldsymbol x_{min} \equiv \argmin\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\ak{\thetaN}}(\boldsymbol x)\)
+and \(\boldsymbol x_{max} \equiv \argmax\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\ak{\thetaN}}(\boldsymbol x)\).
 Note there exists a sequence \(\boldsymbol x_{min} \equiv \boldsymbol x_0, \boldsymbol x_1, \dots, \boldsymbol x_k \equiv \boldsymbol x_{max}\) in \(\mathcal{X}^N\) of component-wise switches to move from \(\boldsymbol x_{min}\) to \(\boldsymbol x_{max}\) in the sample space (i.e.~\(\boldsymbol x_i, \boldsymbol x_{i + 1} \in \mathcal{X}^N\) differ in exactly \(1\) component, \(i = 0, \dots, k\)) for some integer \(k \in \{0, 1, \dots, N\}\).
 Under the FOES model, recall
-\(P_{\boldsymbol \theta_N}(\boldsymbol x) > 0\) holds so that \(\log P_{\boldsymbol \theta_N}(\boldsymbol x)\) is well-defined for each outcome \(\boldsymbol x \in \mathcal{X}^N\).
-Then, if \(k > 0\), it follows that
+\(P_{\ak{\thetaN}}(\boldsymbol x) > 0\) holds so that \(\log P_{\ak{\thetaN}}(\boldsymbol x)\) is well-defined for each outcome \(\boldsymbol x \in \mathcal{X}^N\). Then, if \(k > 0\), it follows that
 \begin{align*}
-\REP(\boldsymbol \theta_N) = \log\left[\frac{P_{\boldsymbol \theta_N}(\boldsymbol x_{max})}{P_{\boldsymbol \theta_N}(\boldsymbol x_{min})}\right] &= \left|\sum\limits_{i = 1}^k\log\left(\frac{P_{\boldsymbol \theta_N}(\boldsymbol x_i)}{P_{\boldsymbol \theta_N}(\boldsymbol x_{i-1})}\right)\right| \\
-&\le \sum\limits_{i = 1}^k\left|\log\left(\frac{P_{\boldsymbol \theta_N}(\boldsymbol x_i)}{P_{\boldsymbol \theta}(\boldsymbol x_{i-1})}\right)\right| \le k \Delta_N(\boldsymbol \theta_N) \le NC,
+\REP(\ak{\thetaN}) = \log\left[\frac{P_{\ak{\thetaN}}(\boldsymbol x_{max})}{P_{ \ak{\thetaN}}(\boldsymbol x_{min})}\right] &= \left|\sum\limits_{i = 1}^k\log\left(\frac{P_{\ak{\thetaN}}(\boldsymbol x_i)}{P_{\ak{\thetaN}}(\boldsymbol x_{i-1})}\right)\right| \\
+&\le \sum\limits_{i = 1}^k\left|\log\left(\frac{P_{\ak{\thetaN}}(\boldsymbol x_i)}{P_{\boldsymbol \theta}(\boldsymbol x_{i-1})}\right)\right| \le k \Delta_N(\ak{\thetaN}) \le NC,
 \end{align*}
-using \(k \le N\) and \(\Delta(\boldsymbol \theta_N) \le C\).
+using \(k \le N\) and \(\Delta(\ak{\thetaN}) \le C\).
 If \(k = 0\), then \(\boldsymbol x_{max} = \boldsymbol x_{min}\) and the same bound above holds.
 This establishes part (i).
-To show part (ii), note the definition of S-instability (i.e., \(\lim_{N\to \infty}\REP(\boldsymbol \theta_N)/N= \infty\)) combined with part (i) implies that \(\lim_{N\to \infty}\DN(\boldsymbol \theta_N)=\infty\).
+To show part (ii), note the definition of S-instability (i.e., \(\lim_{N\to \infty}\REP(\ak{\thetaN})/N= \infty\)) combined with part (i) implies that \(\lim_{N\to \infty}\DN(\ak{\thetaN})=\infty\).
 \hfill \(\Box\)
 
-\textbf{Proof of Proposition \ref{thm:degenFOES}.} As \(|\mathcal{X}|<\infty\) holds in the FOES model, we may suppose
+\textbf{Proof of Theorem \ref{thm:degenFOES}.} As \(|\mathcal{X}|<\infty\) holds in the FOES model, we may suppose
 \(|\mathcal{X}|>1\); otherwise, \(\mathcal{X}^N\) has one outcome and the model is trivially degenerate for all \(N \geq 1\).
-Fix \(0 < \epsilon < 1\) and write \(\boldsymbol x_{min} \equiv \argmin\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)\) and \(\boldsymbol x_{max} \equiv \argmax\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)\).
-Then, \(\boldsymbol x_{max} \in M_{\epsilon, \boldsymbol \theta_N}\), so \(P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N}) \ge P_{\boldsymbol \theta_N}(\boldsymbol x_{max}) > 0\).
-If \(\boldsymbol x \in \mathcal{X}^N \setminus M_{\epsilon, \boldsymbol \theta_N}\), then by definition
-\(P_{\boldsymbol \theta_N}(\boldsymbol x) \le [P_{\boldsymbol \theta_N}(\boldsymbol x_{max})]^{1-\epsilon}[P_{\boldsymbol \theta_N}(\boldsymbol x_{min})]^{\epsilon}\) holds so that
+Fix \(0 < \epsilon < 1\) and write \(\boldsymbol x_{min} \equiv \argmin\limits_{\boldsymbol x \in \mathcal{X}^N}P_{ \ak{\thetaN}}(\boldsymbol x)\) and \(\boldsymbol x_{max} \equiv \argmax\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\ak{\thetaN}}(\boldsymbol x)\).
+Then, \(\boldsymbol x_{max} \in M_{\epsilon, \ak{\thetaN}}\), so \(P_{\ak{\thetaN}}(M_{\epsilon, \ak{\thetaN}}) \ge P_{ \ak{\thetaN}}(\boldsymbol x_{max}) > 0\).
+If \(\boldsymbol x \in \mathcal{X}^N \setminus M_{\epsilon, \ak{\thetaN}}\), then by definition
+\(P_{\ak{\thetaN}}(\boldsymbol x) \le [P_{\ak{\thetaN}}(\boldsymbol x_{max})]^{1-\epsilon}[P_{\ak{\thetaN}}(\boldsymbol x_{min})]^{\epsilon}\) holds so that
 \[
-1-P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N})
- = \sum\limits_{\boldsymbol x \in \mathcal{X}^N \setminus M_{\epsilon, \boldsymbol \theta_N}}P_{\boldsymbol \theta_N}(\boldsymbol x)
-  \le (|\mathcal{X}|^N)[P_{\boldsymbol \theta_N}(\boldsymbol x_{max})]^{1-\epsilon}[P_{\boldsymbol \theta_N}(\boldsymbol x_{min})]^{\epsilon}.
+1-P_{\ak{\thetaN}}(M_{\epsilon, \ak{\thetaN}})
+ = \sum\limits_{\boldsymbol x \in \mathcal{X}^N \setminus M_{\epsilon, \ak{\thetaN}}}P_{\ak{\thetaN}}(\boldsymbol x)
+  \le (|\mathcal{X}|^N)[P_{\ak{\thetaN}}(\boldsymbol x_{max})]^{1-\epsilon}[P_{ \ak{\thetaN}}(\boldsymbol x_{min})]^{\epsilon}.
 \]
-From the lower bound on \(P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N})\) and the
-upper bound on \(1-P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N})\), it follows that
+From the lower bound on \(P_{\ak{\thetaN}}(M_{\epsilon, \ak{\thetaN}})\) and the
+upper bound on \(1-P_{\ak{\thetaN}}(M_{\epsilon, \ak{\thetaN}})\), it follows that
 \begin{align*}
-\frac{1}{N}\log\left[\frac{P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N})}{1-P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N})}\right] & \ge \frac{1}{N} \log\left[\frac{P_{\boldsymbol\theta_N}(\boldsymbol x_{max})}{(|\mathcal{X}|^N)[P_{\boldsymbol \theta_N}(\boldsymbol x_{max})]^{1-\epsilon}[P_{\boldsymbol \theta_N}(\boldsymbol x_{min})]^{\epsilon}}\right] \\
-&= \frac{\epsilon}{N} \log\left[\frac{P_{\boldsymbol \theta_N}(\boldsymbol x_{max})}{P_{\boldsymbol \theta_N}(\boldsymbol x_{min})}\right] - \log |\mathcal{X}| \rightarrow \infty
+\frac{1}{N}\log\left[\frac{P_{\ak{\thetaN}}(M_{\epsilon, \ak{\thetaN}})}{1-P_{ \ak{\thetaN}}(M_{\epsilon, \ak{\thetaN}})}\right] & \ge \frac{1}{N} \log\left[\frac{P_{\ak{\thetaN}}(\boldsymbol x_{max})}{(|\mathcal{X}|^N)[P_{ \ak{\thetaN}}(\boldsymbol x_{max})]^{1-\epsilon}[P_{\ak{\thetaN}}(\boldsymbol x_{min})]^{\epsilon}}\right] \\
+&= \frac{\epsilon}{N} \log\left[\frac{P_{\ak{\thetaN}}(\boldsymbol x_{max})}{P_{ \ak{\thetaN}}(\boldsymbol x_{min})}\right] - \log |\mathcal{X}| \rightarrow \infty
 \end{align*}
-as \(N \rightarrow \infty\) by the definition of an S-unstable FOES model \eqref{eq:Sun}. Consequently,
-\(P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N}) \rightarrow 1\) as \(N \rightarrow \infty\) as claimed. \hfill \(\Box\)
+as \(N \rightarrow \infty\) by the definition of an S-unstable FOES model \eqref{eq:Sun}.
+Consequently, \(P_{\ak{\thetaN}}(M_{\epsilon, \ak{\thetaN}}) \rightarrow 1\) as \(N \rightarrow \infty\) as claimed.
+\hfill \(\Box\)
 
@@
 \label{eq:R}
-\frac{\max_{\boldsymbol y \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol y)  }{\min_{\boldsymbol y \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol y) } = \frac{\max_{\boldsymbol y \in \mathcal{X}^N}P_{-\boldsymbol \theta_N}(\boldsymbol y) }{\min_{\boldsymbol y \in \mathcal{X}^N}P_{-\boldsymbol \theta_N}(\boldsymbol y) }
+\frac{\max_{\boldsymbol y \in \mathcal{X}^N}P_{\ak{\thetaN}}(\boldsymbol y)  }{\min_{\boldsymbol y \in \mathcal{X}^N}P_{\ak{\thetaN}}(\boldsymbol y) } = \frac{\max_{\boldsymbol y \in \mathcal{X}^N}P_{-\ak{\thetaN}}(\boldsymbol y) }{\min_{\boldsymbol y \in \mathcal{X}^N}P_{-\ak{\thetaN}}(\boldsymbol y) }
 \end{equation}
-so that the log-ratio \(\REP(\boldsymbol \theta_N)=\REP(-\boldsymbol \theta_N)\) is the same for both \(\boldsymbol \theta_N\) and \(-\boldsymbol \theta_N\) in \eqref{eq:elpr}.
-Now part (i) of Corollary \ref{cor:sign} follows from \(\REP(\boldsymbol \theta_N)/N=\REP(-\boldsymbol \theta_N)/N\to \infty\) as \(N\to \infty\) in \eqref{eq:Sun}.
-To show part (ii), fix \(0 < \epsilon < 1\) and consider a \(\epsilon\)-mode set \(\mathcal{M}_{\epsilon, \boldsymbol \theta_N}\) under \(\boldsymbol \theta_N\) from \eqref{eq:mode}.
-If \(\boldsymbol x \in \mathcal{M}_{\epsilon, \boldsymbol \theta_N}^c \equiv \mathcal{X}^N \setminus \mathcal{M}_{\epsilon, \boldsymbol \theta_N}\), then, by definition,
+so that the log-ratio \(\REP(\ak{\thetaN})=\REP(-\ak{\thetaN})\) is the same for both \(\ak{\thetaN}\) and \(-\ak{\thetaN}\) in \eqref{eq:elpr}.
+Now part (i) of Corollary \ref{cor:sign} follows from \(\REP(\ak{\thetaN})/N=\REP(-\ak{\thetaN})/N\to \infty\) as \(N\to \infty\) in \eqref{eq:Sun}.
+To show part (ii), fix \(0 < \epsilon < 1\) and consider a \(\epsilon\)-mode set \(\mathcal{M}_{\epsilon, \ak{\thetaN}}\) under \(\ak{\thetaN}\) from \eqref{eq:mode}.
+If \(\boldsymbol x \in \mathcal{M}_{\epsilon, \ak{\thetaN}}^c \equiv \mathcal{X}^N \setminus \mathcal{M}_{\epsilon, \ak{\thetaN}}\), then, by definition,
 \[
-\frac{P_{\boldsymbol \theta_N}(\boldsymbol x)}{\min_{\boldsymbol y \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol y)} \leq \left[\frac{\max_{\boldsymbol y \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol y)  }{\min_{\boldsymbol y \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol y) } \right]^{1-\epsilon}
+\frac{P_{\ak{\thetaN}}(\boldsymbol x)}{\min_{\boldsymbol y \in \mathcal{X}^N}P_{ \ak{\thetaN}}(\boldsymbol y)} \leq \left[\frac{\max_{\boldsymbol y \in \mathcal{X}^N}P_{\ak{\thetaN}}(\boldsymbol y)  }{\min_{\boldsymbol y \in \mathcal{X}^N}P_{\ak{\thetaN}}(\boldsymbol y) } \right]^{1-\epsilon}
 \]
@@
 \[
-\frac{\max_{\boldsymbol y \in \mathcal{X}^N}P_{-\boldsymbol \theta_N}(\boldsymbol y)}{P_{-\boldsymbol \theta_N}(\boldsymbol x)} \leq
-\left[\frac{\max_{\boldsymbol y \in \mathcal{X}^N}P_{-\boldsymbol \theta_N}(\boldsymbol y)  }{\min_{\boldsymbol y \in \mathcal{X}^N}P_{-\boldsymbol \theta_N}(\boldsymbol y) } \right]^{1-\epsilon}
+\frac{\max_{\boldsymbol y \in \mathcal{X}^N}P_{-\ak{\thetaN}}(\boldsymbol y)}{P_{-\ak{\thetaN}}(\boldsymbol x)} \leq
+\left[\frac{\max_{\boldsymbol y \in \mathcal{X}^N}P_{-\ak{\thetaN}}(\boldsymbol y)  }{\min_{\boldsymbol y \in \mathcal{X}^N}P_{-\ak{\thetaN}}(\boldsymbol y) } \right]^{1-\epsilon}
 \]
@@
 \label{eq:R2}
-\log P_{-\boldsymbol \theta_N}(\boldsymbol x) \geq \epsilon \max\limits_{\boldsymbol y \in \mathcal{X}^N} \log  P_{-\boldsymbol \theta_N}(\boldsymbol y) + (1-\epsilon)\min\limits_{\boldsymbol y \in \mathcal{X}^N} \log P_{-\boldsymbol \theta_N}(\boldsymbol y),
+\log P_{-\ak{\thetaN}}(\boldsymbol x) \geq \epsilon \max\limits_{\boldsymbol y \in \mathcal{X}^N} \log  P_{-\ak{\thetaN}}(\boldsymbol y) + (1-\epsilon)\min\limits_{\boldsymbol y \in \mathcal{X}^N} \log P_{-\ak{\thetaN}}(\boldsymbol y),
 \end{equation}
-so that \(\boldsymbol x \in \mathcal{M}_{\epsilon, \boldsymbol \theta_N}^c\) if and only if \eqref{eq:R2} holds.
-Next consider the \((1-\epsilon)\)-mode set \(\mathcal{M}_{1-\epsilon, -\boldsymbol \theta_N}\) under \(-\boldsymbol \theta_N\) from \eqref{eq:mode}.
-If \(\boldsymbol x \in\mathcal{M}_{1-\epsilon, -\boldsymbol \theta_N}\), then by definition \(\boldsymbol x\) fulfills \eqref{eq:R2} and so \(\boldsymbol x \in \mathcal{M}_{\epsilon, \boldsymbol \theta_N}^c\), showing that \(\mathcal{M}_{1-\epsilon, -\boldsymbol \theta_N} \subset \mathcal{M}_{\epsilon, \boldsymbol \theta_N}^c\).
-By this and the fact that that Theorem \ref{thm:degenFOES} and Corollary \ref{cor:sign}(i) entail that \(P_{-\boldsymbol \theta_N}(\boldsymbol X_N \in \mathcal{M}_{1-\epsilon, -\boldsymbol \theta_N})\to 1\) as \(N\to \infty\) (i.e., \(P_{-\boldsymbol \theta_N}\) is S-unstable), we have
+so that \(\boldsymbol x \in \mathcal{M}_{\epsilon, \ak{\thetaN}}^c\) if and only if \eqref{eq:R2} holds.
+Next consider the \((1-\epsilon)\)-mode set \(\mathcal{M}_{1-\epsilon, -\ak{\thetaN}}\) under \(-\ak{\thetaN}\) from \eqref{eq:mode}.
+If \(\boldsymbol x \in\mathcal{M}_{1-\epsilon, -\ak{\thetaN}}\), then by definition \(\boldsymbol x\) fulfills \eqref{eq:R2} and so \(\boldsymbol x \in \mathcal{M}_{\epsilon, \ak{\thetaN}}^c\), showing that \(\mathcal{M}_{1-\epsilon, -\ak{\thetaN}} \subset \mathcal{M}_{\epsilon, \ak{\thetaN}}^c\).
+By this and the fact that that Theorem \ref{thm:degenFOES} and Corollary \ref{cor:sign}(i) entail that \(P_{-\ak{\thetaN}}(\boldsymbol X_N \in \mathcal{M}_{1-\epsilon, -\ak{\thetaN}})\to 1\) as \(N\to \infty\) (i.e., \(P_{-\ak{\thetaN}}\) is S-unstable), we have
 \[
-1  = \lim_{N\to \infty} P_{-\boldsymbol \theta_N}(\boldsymbol X_N \in \mathcal{M}_{1-\epsilon, -\boldsymbol \theta_N}) \leq \lim_{N\to \infty} P_{-\boldsymbol \theta_N}(\boldsymbol X_N \in \mathcal{M}_{\epsilon, \boldsymbol \theta_N}^c ) \leq 1,
+1  = \lim_{N\to \infty} P_{-\ak{\thetaN}}(\boldsymbol X_N \in \mathcal{M}_{1-\epsilon, -\ak{\thetaN}}) \leq \lim_{N\to \infty} P_{-\ak{\thetaN}}(\boldsymbol X_N \in \mathcal{M}_{\epsilon, \ak{\thetaN}}^c ) \leq 1,
 \]
@@
 \[
-\left|\log \left[ \frac{P_{\boldsymbol \theta_N}(\boldsymbol x_1)}{P_{\boldsymbol \theta_N}(\boldsymbol x_2)}  \right] \right| =
-\left|  \sum_{i=1}^k \theta_{i,N} [g_{i,k}(\boldsymbol x_1) - g_{i,k}(\boldsymbol x_2) ] \right|  \leq  \sum_{i=1}^k | \theta_{i,N}| (U_{i,N}-L_{i,N});
+\left|\log \left[ \frac{P_{\ak{\thetaN}}(\boldsymbol x_1)}{P_{\ak{\thetaN}}(\boldsymbol x_2)}  \right] \right| =
+\left|  \sum_{i=1}^k \theta_{i,\ak{\thetaidx}} [g_{i,k}(\boldsymbol x_1) - g_{i,k}(\boldsymbol x_2) ] \right|  \leq  \sum_{i=1}^k | \theta_{i,\ak{\thetaidx}}| (U_{i,N}-L_{i,N});
 \]
-consequently, \(\REP(\boldsymbol \theta_N ) \leq \sum_{i=1}^k | \theta_{i,N}| (U_{i,N}-L_{i,N})\) holds in \eqref{eq:elpr} and model stability in Proposition \ref{prp:prop1} follows from \eqref{eq:Sun}.
+consequently, \(\REP(\ak{\thetaN} ) \leq \sum_{i=1}^k | \theta_{i,\ak{\thetaidx}}| (U_{i,N}-L_{i,N})\) holds in \eqref{eq:elpr} and model stability in Proposition \ref{prp:prop1} follows from \eqref{eq:Sun}.
 \hfill \(\Box\)
@@
 \textbf{Proof of Proposition \ref{prp:prop2}.}
-Writing \(\boldsymbol x=(x_1,\ldots,x_N)\) and \(\boldsymbol h = (h_1,\ldots,h_{N_{\mathcal{H}}})\) with all components \(x_i,h_j\in\{\pm 1\}\), probabilities in the joint RBM model \eqref{eq:RBM1} can be written as \(\tilde{P}_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h) = c(\boldsymbol \theta_N)\exp[ f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)]\) in terms of the function \(f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)\) from \eqref{eq:f} and the normalizing constant \(c(\boldsymbol \theta_N)= \exp [-\psi(\boldsymbol \theta_N)]\) from \eqref{eq:RBM1}.
-Let \(\boldsymbol x_M, \boldsymbol x_m\in\{\pm 1\}^N\) be such that \(P_{\boldsymbol \theta_N} (\boldsymbol x_M) = \max_{\boldsymbol x}P_{\boldsymbol \theta_N} (\boldsymbol x)\) and \(P_{\boldsymbol \theta_N} (\boldsymbol x_m) = \min_{\boldsymbol x}P_{\boldsymbol \theta_N} (\boldsymbol x)\) under the marginal RBM model \(P_{\boldsymbol \theta_N} (\boldsymbol x) = c(\boldsymbol \theta_N)\sum_{\boldsymbol h \in\{\pm 1\}^{\mathcal{N}_H}} \tilde{P}_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)= c(\boldsymbol \theta_N)\sum_{\boldsymbol h \in\{\pm 1\}^{\mathcal{N}_H}} \exp[ f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)]\) from \eqref{eq:RBM2}.
-Also, \(\boldsymbol x_0,x_1\in\{\pm 1\}^N\) be such that \(\max_{\boldsymbol h}f_{\boldsymbol \theta_N} (\boldsymbol x_0, \boldsymbol h)=\max_{\boldsymbol x}\max_{\boldsymbol h}f_{\boldsymbol \theta_N} (\boldsymbol x , \boldsymbol h)\) and \(\max_{\boldsymbol h}f_{\boldsymbol \theta_N} (\boldsymbol x_1, \boldsymbol h)=\min_{\boldsymbol x}\max_{\boldsymbol h}f_{\boldsymbol \theta_N} (\boldsymbol x , \boldsymbol h)\).
-Then, Proposition \ref{prp:prop2}(i) follows from \(\REP_{(\boldsymbol X)}(\boldsymbol \theta_N) = \log[P_{\boldsymbol \theta_N} (\boldsymbol x_M) /P_{\boldsymbol \theta_N} (\boldsymbol x_m) ]\) and the lower/upper bounds on \(P_{\boldsymbol \theta_N} (\boldsymbol x_M)\) and \(P_{\boldsymbol \theta_N} (\boldsymbol x_m)\) as
+Writing \(\boldsymbol x=(x_1,\ldots,x_N)\) and \(\boldsymbol h = (h_1,\ldots,h_{N_{\mathcal{H}}})\) with all components \(x_i,h_j\in\{\pm 1\}\), probabilities in the joint RBM model \eqref{eq:RBM1} can be written as \(\tilde{P}_{\ak{\thetaN}} (\boldsymbol x, \boldsymbol h) = c(\ak{\thetaN})\exp[ f_{\ak{\thetaN}} (\boldsymbol x, \boldsymbol h)]\) in terms of the function \(f_{\ak{\thetaN}} (\boldsymbol x, \boldsymbol h)\) from \eqref{eq:f} and the normalizing constant \(c(\ak{\thetaN})= \exp [-\psi(\ak{\thetaN})]\) from \eqref{eq:RBM1}.
+Let \(\boldsymbol x_M, \boldsymbol x_m\in\{\pm 1\}^N\) be such that \(P_{\ak{\thetaN}} (\boldsymbol x_M) = \max_{\boldsymbol x}P_{\ak{\thetaN}} (\boldsymbol x)\) and \(P_{\ak{\thetaN}} (\boldsymbol x_m) = \min_{\boldsymbol x}P_{\ak{\thetaN}} (\boldsymbol x)\) under the marginal RBM model \(P_{\ak{\thetaN}} (\boldsymbol x) = c(\ak{\thetaN})\sum_{\boldsymbol h \in\{\pm 1\}^{\mathcal{N}_H}} \tilde{P}_{\ak{\thetaN}} (\boldsymbol x, \boldsymbol h)= c( \ak{\thetaN})\sum_{\boldsymbol h \in\{\pm 1\}^{\mathcal{N}_H}} \exp[ f_{\ak{\thetaN}} (\boldsymbol x, \boldsymbol h)]\) from \eqref{eq:RBM2}.
+Also, \(\boldsymbol x_0,x_1\in\{\pm 1\}^N\) be such that \(\max_{\boldsymbol h}f_{\ak{\thetaN}} (\boldsymbol x_0, \boldsymbol h)=\max_{\boldsymbol x}\max_{\boldsymbol h}f_{\ak{\thetaN}} (\boldsymbol x , \boldsymbol h)\) and \(\max_{\boldsymbol h}f_{\ak{\thetaN}} (\boldsymbol x_1, \boldsymbol h)=\min_{\boldsymbol x}\max_{\boldsymbol h}f_{\ak{\thetaN}} (\boldsymbol x , \boldsymbol h)\).
+Then, Proposition \ref{prp:prop2}(i) follows from \(\REP_{(\boldsymbol X)}(\ak{\thetaN}) = \log[P_{\ak{\thetaN}} (\boldsymbol x_M) /P_{\ak{\thetaN}} (\boldsymbol x_m) ]\) and the lower/upper bounds on \(P_{\ak{\thetaN}} (\boldsymbol x_M)\) and \(P_{\ak{\thetaN}} (\boldsymbol x_m)\) as
 \[
-c(\boldsymbol \theta_N) \exp[\max_{\boldsymbol h}f_{\boldsymbol \theta_N} (\boldsymbol x_0 , \boldsymbol h)]
-\leq P_{\boldsymbol \theta_N} (\boldsymbol x_0) \leq  P_{\boldsymbol \theta_N} (\boldsymbol x_M) \leq  2^{N_{\mathcal{H}}} c(\boldsymbol \theta_N) \exp[\max_{\boldsymbol x}\max_{\boldsymbol h}f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)]
+c(\ak{\thetaN}) \exp[\max_{\boldsymbol h}f_{\ak{\thetaN}} (\boldsymbol x_0 , \boldsymbol h)]
+\leq P_{\ak{\thetaN}} (\boldsymbol x_0) \leq  P_{\ak{\thetaN}} (\boldsymbol x_M) \leq  2^{N_{\mathcal{H}}} c(\ak{\thetaN}) \exp[\max_{\boldsymbol x}\max_{\boldsymbol h}f_{\ak{\thetaN}} (\boldsymbol x, \boldsymbol h)]
 \]
@@
 \begin{align*}
-c(\boldsymbol \theta_N) \exp[\min_{\boldsymbol x}\max_{\boldsymbol h}f_{\boldsymbol \theta_N} (\boldsymbol x , \boldsymbol h)] &\leq c(\boldsymbol \theta_N) \exp[ \max_{\boldsymbol h}f_{\boldsymbol \theta_N} (\boldsymbol x_m , \boldsymbol h)] \\
-&\leq P_{\boldsymbol \theta_N} (\boldsymbol x_m) \\
-&\leq P_{\boldsymbol \theta_N} (\boldsymbol x_1) \\ 
-&\leq 2^{N_{\mathcal{H}}} c(\boldsymbol \theta_N) \exp[ \max_{\boldsymbol h}f_{\boldsymbol \theta_N} (\boldsymbol x_1 , \boldsymbol h)]\\
-&=2^{N_{\mathcal{H}}} c(\boldsymbol \theta_N) \exp[\min_{\boldsymbol x}\max_{\boldsymbol h}f_{\boldsymbol \theta_N} (\boldsymbol x , \boldsymbol h)]
+c(\ak{\thetaN}) \exp[\min_{\boldsymbol x}\max_{\boldsymbol h}f_{\ak{\thetaN}} (\boldsymbol x , \boldsymbol h)] &\leq c(\ak{\thetaN}) \exp[ \max_{\boldsymbol h}f_{\ak{\thetaN}} (\boldsymbol x_m , \boldsymbol h)] \\
+&\leq P_{\ak{\thetaN}} (\boldsymbol x_m) \\
+&\leq P_{\ak{\thetaN}} (\boldsymbol x_1) \\ 
+&\leq 2^{N_{\mathcal{H}}} c(\ak{\thetaN}) \exp[ \max_{\boldsymbol h}f_{\ak{\thetaN}} (\boldsymbol x_1 , \boldsymbol h)]\\
+&=2^{N_{\mathcal{H}}} c(\ak{\thetaN}) \exp[\min_{\boldsymbol x}\max_{\boldsymbol h}f_{\ak{\thetaN}} (\boldsymbol x , \boldsymbol h)]
 \end{align*}
-To prove Proposition \ref{prp:prop2}, we next expand the function \(f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)\) from \eqref{eq:f} as
+To prove Proposition \ref{prp:prop2}, we next expand the function \(f_{\ak{\thetaN}} (\boldsymbol x, \boldsymbol h)\) from \eqref{eq:f} as
 \[
-f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h)= \sum_{j=1}^{N_\mathcal{H}} h_j \theta_{j,N}^{\mathcal{H}}  +   \sum_{i=1}^N \left( \theta_{i,N}^{\mathcal{V}} + \sum_{j=1}^{N_\mathcal{H}}  h_j  \theta_{ij,N}^{\mathcal{VH}}\right)x_i=\sum_{i=1}^N x_i  \theta_{i,N}^{\mathcal{V}} +  \sum_{j=1}^{N_\mathcal{H}} \left( \theta_{j,N}^{\mathcal{H}}  + \sum_{i=1}^N  x_i  \theta_{ij,N}^{\mathcal{VH}}\right)h_j.
+f_{\ak{\thetaN}} (\boldsymbol x, \boldsymbol h)= \sum_{j=1}^{N_\mathcal{H}} h_j \theta_{j,\ak{\thetaidx}}^{\mathcal{H}}  +   \sum_{i=1}^N \left( \theta_{i,\ak{\thetaidx}}^{\mathcal{V}} + \sum_{j=1}^{N_\mathcal{H}}  h_j  \theta_{ij,\ak{\thetaidx}}^{\mathcal{VH}}\right)x_i=\sum_{i=1}^N x_i  \theta_{i,\ak{\thetaidx}}^{\mathcal{V}} +  \sum_{j=1}^{N_\mathcal{H}} \left( \theta_{j,\ak{\thetaidx}}^{\mathcal{H}}  + \sum_{i=1}^N  x_i  \theta_{ij,\ak{\thetaidx}}^{\mathcal{VH}}\right)h_j.
 \]
@@
 \begin{align}
-\nonumber \max_{\boldsymbol x}  f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h) =  \sum_{j=1}^{N_\mathcal{H}} h_j \theta_{j,N}^{\mathcal{H}}  +   a_{\boldsymbol \theta_N, \mathcal{H}} (\boldsymbol h), & \min_{\boldsymbol x}  f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h) =  \sum_{j=1}^{N_\mathcal{H}} h_j \theta_{j,N}^{\mathcal{H}}  - a_{\boldsymbol \theta_N, \mathcal{H}} (\boldsymbol h),\\
-\nonumber \max_{\boldsymbol h}  f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h) = \sum_{i=1}^{N}x_i \theta_{i,N}^{\mathcal{V}} + b_{\boldsymbol \theta_N, \mathcal{V}} (\boldsymbol x), & \min_{\boldsymbol h}  f_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h) = \sum_{i=1}^{N}x_i \theta_{i,N}^{\mathcal{V}}  -   b_{\boldsymbol \theta_N, \mathcal{V}} (\boldsymbol x), \\
+\nonumber \max_{\boldsymbol x}  f_{\ak{\thetaN}} (\boldsymbol x, \boldsymbol h) =  \sum_{j=1}^{N_\mathcal{H}} h_j \theta_{j,\ak{\thetaidx}}^{\mathcal{H}}  +   a_{\ak{\thetaN}, \mathcal{H}} (\boldsymbol h), & \min_{\boldsymbol x}  f_{\ak{\thetaN}} (\boldsymbol x, \boldsymbol h) =  \sum_{j=1}^{N_\mathcal{H}} h_j \theta_{j,\ak{\thetaidx}}^{\mathcal{H}}  - a_{\ak{\thetaN}, \mathcal{H}} (\boldsymbol h),\\
+\nonumber \max_{\boldsymbol h}  f_{\ak{\thetaN}} (\boldsymbol x, \boldsymbol h) = \sum_{i=1}^{N}x_i \theta_{i,\ak{\thetaidx}}^{\mathcal{V}} + b_{\ak{\thetaN}, \mathcal{V}} (\boldsymbol x), & \min_{\boldsymbol h}  f_{\ak{\thetaN}} (\boldsymbol x, \boldsymbol h) = \sum_{i=1}^{N}x_i \theta_{i,\ak{\thetaidx}}^{\mathcal{V}}  -   b_{\ak{\thetaN}, \mathcal{V}} (\boldsymbol x), \\
 \label{eq:max}
-a_{\boldsymbol \theta_N, \mathcal{H}} (\boldsymbol h) \equiv \sum_{i=1}^N \left| \theta_{i,N}^{\mathcal{V}} + \sum_{j=1}^{N_\mathcal{H}}  h_j  \theta_{ij,N}^{\mathcal{VH}}\right|, &  b_{\boldsymbol \theta_N, \mathcal{V}} (\boldsymbol x) \equiv \sum_{j=1}^{N_\mathcal{H}} \left| \theta_{j,N}^{\mathcal{H}}  + \sum_{i=1}^N  x_i  \theta_{ij,N}^{\mathcal{VH}}\right|,
+a_{\ak{\thetaN}, \mathcal{H}} (\boldsymbol h) \equiv \sum_{i=1}^N \left| \theta_{i,\ak{\thetaidx}}^{\mathcal{V}} + \sum_{j=1}^{N_\mathcal{H}}  h_j  \theta_{ij,\ak{\thetaidx}}^{\mathcal{VH}}\right|, &  b_{\ak{\thetaN}, \mathcal{V}} (\boldsymbol x) \equiv \sum_{j=1}^{N_\mathcal{H}} \left| \theta_{j,\ak{\thetaidx}}^{\mathcal{H}}  + \sum_{i=1}^N  x_i  \theta_{ij,\ak{\thetaidx}}^{\mathcal{VH}}\right|,
 \end{align}
-where \(\boldsymbol h^T \boldsymbol \theta_N^{\mathcal{H}}=\sum_{j=1}^{N_\mathcal{H}} h_j \theta_{j,N}^{\mathcal{H}}\), \(\boldsymbol x^T \boldsymbol \theta_N^{\mathcal{V}}= \sum_{i=1}^{N}x_i \theta_{i,N}^{\mathcal{V}}\) and \(\Gam\equiv \max_{\boldsymbol h} a_{\boldsymbol \theta_N, \mathcal{H}} (\boldsymbol h)\).
+where \(\boldsymbol h^T \ak{\thetaN}^{\mathcal{H}}=\sum_{j=1}^{N_\mathcal{H}} h_j \theta_{j,\ak{\thetaidx}}^{\mathcal{H}}\), \(\boldsymbol x^T \ak{\thetaN}^{\mathcal{V}}= \sum_{i=1}^{N}x_i \theta_{i,\ak{\thetaidx}}^{\mathcal{V}}\) and \(\Gam\equiv \max_{\boldsymbol h} a_{\ak{\thetaN}, \mathcal{H}} (\boldsymbol h)\).
 From this, it follows that
 \begin{align*}
-\REP_{(\boldsymbol X, \boldsymbol H)}(\boldsymbol \theta_N) &=  \max_{\boldsymbol h}\max_{\boldsymbol x}f_{\boldsymbol \theta_N} (\boldsymbol x , \boldsymbol h) -  \min_{\boldsymbol h}\min_{\boldsymbol x}f_{\boldsymbol \theta_N} (\boldsymbol x , \boldsymbol h)\\
-&= \max_{\boldsymbol h_1 }  \max_{\boldsymbol h_2 }\left[ (\boldsymbol h_1 - \boldsymbol h_2)^T \boldsymbol \theta_N^{\mathcal{H}}  +   a_{\boldsymbol \theta_N, \mathcal{H}} (\boldsymbol h_1)  + a_{\boldsymbol \theta_N, \mathcal{H}} (\boldsymbol h_2)\right],
+\REP_{(\boldsymbol X, \boldsymbol H)}(\ak{\thetaN}) &=  \max_{\boldsymbol h}\max_{\boldsymbol x}f_{\ak{\thetaN}} (\boldsymbol x , \boldsymbol h) -  \min_{\boldsymbol h}\min_{\boldsymbol x}f_{\ak{\thetaN}} (\boldsymbol x , \boldsymbol h)\\
+&= \max_{\boldsymbol h_1 }  \max_{\boldsymbol h_2 }\left[ (\boldsymbol h_1 - \boldsymbol h_2)^T  \ak{\thetaN}^{\mathcal{H}}  +   a_{\ak{\thetaN}, \mathcal{H}} (\boldsymbol h_1)  + a_{\ak{\thetaN}, \mathcal{H}} (\boldsymbol h_2)\right],
 \end{align*}
-which leads to the upper bound \(\REP_{(\boldsymbol X, \boldsymbol H)}(\boldsymbol \theta_N) \leq 2 \Gam + 2 |\boldsymbol \theta_N^{\mathcal{H}} |_1\).
-Then, taking \(\boldsymbol h_1=\boldsymbol h_2\) (i.e., before maximization) gives \(\REP_{(\boldsymbol X, \boldsymbol H)}(\boldsymbol \theta_N) \geq 2\Gam\) and taking \(\boldsymbol h_1=-\boldsymbol h_2\), such that \(\boldsymbol h_1^T \boldsymbol \theta_N^{\mathcal{H}} = |\theta_N^{\mathcal{H}}|_1\), gives \(\REP_{(\boldsymbol X, \boldsymbol H)}(\boldsymbol \theta_N) \geq 2|\theta_N^{\mathcal{H}}|_1\); this yields the lower bound \(\REP_{(\boldsymbol X, \boldsymbol H)}(\boldsymbol \theta_N)\geq 2\max\{\Gam, |\boldsymbol \theta_N^{\mathcal{H}} |_1\}\).
+which leads to the upper bound \(\REP_{(\boldsymbol X, \boldsymbol H)}(\ak{\thetaN}) \leq 2 \Gam + 2 | \ak{\thetaN}^{\mathcal{H}} |_1\).
+Then, taking \(\boldsymbol h_1=\boldsymbol h_2\) (i.e., before maximization) gives \(\REP_{(\boldsymbol X, \boldsymbol H)}(\ak{\thetaN}) \geq 2\Gam\) and taking \(\boldsymbol h_1=-\boldsymbol h_2\), such that \(\boldsymbol h_1^T \ak{\thetaN}^{\mathcal{H}} = |\ak{\thetaN}^{\mathcal{H}}|_1\), gives \(\REP_{(\boldsymbol X, \boldsymbol H)}(\ak{\thetaN}) \geq 2|\ak{\thetaN}^{\mathcal{H}}|_1\); this yields the lower bound \(\REP_{(\boldsymbol X, \boldsymbol H)}(\ak{\thetaN})\geq 2\max\{\Gam, |\ak{\thetaN}^{\mathcal{H}} |_1\}\).
 
@@
 \begin{align*}
-\elt &=  \max_{\boldsymbol h}\max_{\boldsymbol x}f_{\boldsymbol \theta_N} (\boldsymbol x , \boldsymbol h) -  \max_{\boldsymbol h}\min_{\boldsymbol x}f_{\boldsymbol \theta_N} (\boldsymbol x , \boldsymbol h)\\
-&= \max_{\boldsymbol h_1} \min_{\boldsymbol h_2 }\left[ (\boldsymbol h_1 - \boldsymbol h_2)^T \boldsymbol \theta_N^{\mathcal{H}} + a_{\boldsymbol \theta_N, \mathcal{H}} (\boldsymbol h_1)  + a_{\boldsymbol \theta_N, \mathcal{H}} (\boldsymbol h_2)\right].
+\elt &=  \max_{\boldsymbol h}\max_{\boldsymbol x}f_{\ak{\thetaN}} (\boldsymbol x , \boldsymbol h) -  \max_{\boldsymbol h}\min_{\boldsymbol x}f_{\ak{\thetaN}} (\boldsymbol x , \boldsymbol h)\\
+&= \max_{\boldsymbol h_1} \min_{\boldsymbol h_2 }\left[ (\boldsymbol h_1 - \boldsymbol h_2)^T \ak{\thetaN}^{\mathcal{H}} + a_{\ak{\thetaN}, \mathcal{H}} (\boldsymbol h_1)  + a_{\ak{\thetaN}, \mathcal{H}} (\boldsymbol h_2)\right].
 \end{align*}
 Taking \(\boldsymbol h_1=\boldsymbol h_2\) and maximizing over both \(\boldsymbol h_1,\boldsymbol h_2\) produces the upper bound \(\elt \leq 2\Gam\).
-Then, using \((\boldsymbol h_1 - \boldsymbol h_2)^T \boldsymbol \theta_N^{\mathcal{H}} + a_{\boldsymbol \theta_N, \mathcal{H}} (\boldsymbol h_2) \geq - 2|\boldsymbol \theta_N^{\mathcal{H}} |_1\) and maximizing over \(\boldsymbol h_1\) gives \(\elt \geq \Gam- 2|\boldsymbol \theta_N^{\mathcal{H}} |_1\), while setting \(\boldsymbol h_1=\boldsymbol h_2^*\) for \(\boldsymbol h_2^*\) such that \(-(\boldsymbol h_2^*) ^T \boldsymbol \theta_N^{\mathcal{H}} + a_{\boldsymbol \theta_N, \mathcal{H}} (\boldsymbol h_2^*) = \min_{\boldsymbol h_2} [-\boldsymbol h_2^T \boldsymbol \theta_N^{\mathcal{H}} + a_{\boldsymbol \theta_N, \mathcal{H}} (\boldsymbol h_2)]\) gives \(\elt \geq 2 a_{\boldsymbol \theta_N, \mathcal{H}} (\boldsymbol h_2^*) \geq \Gamc \equiv \min_{\boldsymbol h} a_{\boldsymbol \theta_N, \mathcal{H}} (\boldsymbol h)\).
+Then, using \((\boldsymbol h_1 - \boldsymbol h_2)^T \ak{\thetaN}^{\mathcal{H}} + a_{\ak{\thetaN}, \mathcal{H}} (\boldsymbol h_2) \geq - 2|\ak{\thetaN}^{\mathcal{H}} |_1\) and maximizing over \(\boldsymbol h_1\) gives \(\elt \geq \Gam- 2|\ak{\thetaN}^{\mathcal{H}} |_1\), while setting \(\boldsymbol h_1=\boldsymbol h_2^*\) for \(\boldsymbol h_2^*\) such that \(-(\boldsymbol h_2^*) ^T \ak{\thetaN}^{\mathcal{H}} + a_{\ak{\thetaN}, \mathcal{H}} (\boldsymbol h_2^*) = \min_{\boldsymbol h_2} [-\boldsymbol h_2^T \ak{\thetaN}^{\mathcal{H}} + a_{\ak{\thetaN}, \mathcal{H}} (\boldsymbol h_2)]\) gives \(\elt \geq 2 a_{\ak{\thetaN}, \mathcal{H}} (\boldsymbol h_2^*) \geq \Gamc \equiv \min_{\boldsymbol h} a_{ \ak{\thetaN}, \mathcal{H}} (\boldsymbol h)\).
 Finally, note that for any \(\boldsymbol h\), the triangle inequality gives
 \begin{align*}
-\Gam  \equiv \max_{\boldsymbol h_1}  a_{\boldsymbol \theta_N, \mathcal{H}} (\boldsymbol h_1) &\geq
-[a_{\boldsymbol \theta_N, \mathcal{H}} (\boldsymbol h)  + a_{\boldsymbol \theta_N, \mathcal{H}} (-\boldsymbol h)]/2 \\
+\Gam  \equiv \max_{\boldsymbol h_1}  a_{\ak{\thetaN}, \mathcal{H}} (\boldsymbol h_1) &\geq
+[a_{\ak{\thetaN}, \mathcal{H}} (\boldsymbol h)  + a_{\ak{\thetaN}, \mathcal{H}} (-\boldsymbol h)]/2 \\
 &=
-2^{-1}\sum_{i=1}^{N } \left(\left| \theta_{i,N}^{\mathcal{V}}   + \sum_{j=1}^{N_{\mathcal{H}}} h_j \theta_{ij,N}^{\mathcal{VH}} \right|
-+ \left| \theta_{i,N}^{\mathcal{V}}   - \sum_{j=1}^{N_{\mathcal{H}}} h_j \theta_{ij,N}^{\mathcal{VH}} \right|\right)\\
-& \geq \sum_{i=1}^{N }  \left| \theta_{i,N}^{\mathcal{V}} \right| \equiv |\boldsymbol \theta_{N}^{\mathcal{V}}|_1.
+2^{-1}\sum_{i=1}^{N } \left(\left| \theta_{i,\ak{\thetaidx}}^{\mathcal{V}}   + \sum_{j=1}^{N_{\mathcal{H}}} h_j \theta_{ij,\ak{\thetaidx}}^{\mathcal{VH}} \right|
++ \left| \theta_{i,\ak{\thetaidx}}^{\mathcal{V}}   - \sum_{j=1}^{N_{\mathcal{H}}} h_j \theta_{ij,\ak{\thetaidx}}^{\mathcal{VH}} \right|\right)\\
+& \geq \sum_{i=1}^{N }  \left| \theta_{i,\ak{\thetaidx}}^{\mathcal{V}} \right| \equiv | \ak{\thetaN}^{\mathcal{V}}|_1.
 \end{align*}
@@
 
-\textbf{Proof of Theorem \ref{thm:something}.} Let \(L_{\boldsymbol \theta_{N}}(\boldsymbol X) = \log[ P_{\boldsymbol \theta_{N}}(\boldsymbol X)/ \min_{\boldsymbol y \in \mathcal{X}^N} P_{\boldsymbol \theta_{N}}(\boldsymbol y) ]/\REP(\boldsymbol \theta_{N})\), where again \(\boldsymbol X=(X_1, \dots, X_N)\) and \(\REP(\boldsymbol \theta_{N})= \log[\max_{\boldsymbol y \in \mathcal{X}^N} P_{\boldsymbol \theta_{N}}(\boldsymbol y)/ \min_{\boldsymbol y \in \mathcal{X}^N} P_{\boldsymbol \theta_{N}}(\boldsymbol y) ]\).
-As \(L_{\boldsymbol \theta_{N}}(\boldsymbol X)\in[0,1]\), convergence of \(L_{\boldsymbol \theta_{N}}(\boldsymbol X)\) to 1 in probability under \(P_{\boldsymbol \theta_{N}}\) is equivalent to convergence to \(1\) in expectation under \(P_{\boldsymbol \theta_{N}}\) (i.e., convergence in expectation implies probabilistic convergence by Markov's inequality while probabilistic convergence implies convergence in expectation by uniform integrability/boundedness).
+\textbf{Proof of Theorem \ref{thm:something}.} Let \(L_{\ak{\thetaN}}(\boldsymbol X) = \log[ P_{ \ak{\thetaN}}(\boldsymbol X)/ \min_{\boldsymbol y \in \mathcal{X}^N} P_{\ak{\thetaN}}(\boldsymbol y) ]/\REP(\ak{\thetaN})\), where again \(\boldsymbol X=(X_1, \dots, X_N)\) and \(\REP(\ak{\thetaN})= \log[\max_{\boldsymbol y \in \mathcal{X}^N} P_{\ak{\thetaN}}(\boldsymbol y)/ \min_{\boldsymbol y \in \mathcal{X}^N} P_{ \ak{\thetaN}}(\boldsymbol y) ]\).
+As \(L_{\ak{\thetaN}}(\boldsymbol X)\in[0,1]\), convergence of \(L_{ \ak{\thetaN}}(\boldsymbol X)\) to 1 in probability under \(P_{\ak{\thetaN}}\) is equivalent to convergence to \(1\) in expectation under \(P_{\ak{\thetaN}}\) (i.e., convergence in expectation implies probabilistic convergence by Markov's inequality while probabilistic convergence implies convergence in expectation by uniform integrability/boundedness).
 
-For \(\epsilon \in(0,1)\), let \(\mathcal{M}_{\epsilon, \boldsymbol \theta_N}\) denote a modal set as in \eqref{eq:mode}.
-By Theorem \ref{thm:degenFOES}, \(P_{\boldsymbol \theta_N}\left( \boldsymbol X\in \mathcal{M}_{\epsilon, \boldsymbol \theta_N}\right) \rightarrow 1\) holds as \(N \rightarrow \infty\) and, by definition of \eqref{eq:mode}, \(\boldsymbol X\in \mathcal{M}_{\epsilon, \boldsymbol \theta_N}\) follows if and only if
-\(1-L_{\boldsymbol \theta_{N}}(\boldsymbol X)<\epsilon\).
-Hence, \(L_{\boldsymbol \theta_{N}}(\boldsymbol X) \stackrel{p,\mathrm{E}}{\longrightarrow} 1\) holds under \(\boldsymbol \theta_{N}\) in Theorem \ref{thm:something}.
-The convergence \(L_{-\boldsymbol \theta_{N}}(\boldsymbol X) \stackrel{p,\mathrm{E}}{\longrightarrow} 1\) under \(-\boldsymbol \theta_{N}\) likewise follows from Corollary \ref{cor:sign}.
+For \(\epsilon \in(0,1)\), let \(\mathcal{M}_{\epsilon, \ak{\thetaN}}\) denote a modal set as in \eqref{eq:mode}.
+By Theorem \ref{thm:degenFOES}, \(P_{\ak{\thetaN}}\left( \boldsymbol X\in \mathcal{M}_{\epsilon, \ak{\thetaN}}\right) \rightarrow 1\) holds as \(N \rightarrow \infty\) and, by definition of \eqref{eq:mode}, \(\boldsymbol X\in \mathcal{M}_{\epsilon, \ak{\thetaN}}\) follows if and only if
+\(1-L_{\ak{\thetaN}}(\boldsymbol X)<\epsilon\).
+Hence, \(L_{\ak{\thetaN}}(\boldsymbol X) \stackrel{p,\mathrm{E}}{\longrightarrow} 1\) holds under \(\ak{\thetaN}\) in Theorem \ref{thm:something}.
+The convergence \(L_{-\ak{\thetaN}}(\boldsymbol X) \stackrel{p,\mathrm{E}}{\longrightarrow} 1\) under \(-\ak{\thetaN}\) likewise follows from Corollary \ref{cor:sign}.
 \hfill \(\Box\)
+
+\BeginKnitrBlock{remark}
+\iffalse{} {Remark.
+} \fi{}Consider the exponential graph model from \eqref{eq:mod2} involving counts of edges, 2-stars and triangles with fixed parameters \(\thetaN = (\theta_{1},\theta_2,\theta_3)^\prime \in \mathbb{R}^3\) for \(N\geq 1\).
+This model is unstable whenever \(|\theta_2| + |\theta_3| \neq 0\).
+To see this, consider an even number \(n>2\) of nodes and let \(\boldsymbol x_0\) denote the data outcome in \(\mathcal{X}^N \equiv \{0,1\}^N\) with all \(N = {n \choose 2}\) edges being zero, let \(\boldsymbol x_1\) denote the outcome with all edges being 1, and let \(\boldsymbol x_2\) denote the edge configuration from dividing the nodes into two equal groups, with no edges within a group and all edges between the groups (so that no triangles exist in \(\boldsymbol x_2\)).
+Then, the \(N\)-scaled log-ratio \eqref{eq:elpr} for the exponential graph model \eqref{eq:mod2} can, by definition, be bounded below by
+\begin{align*}
+\frac{1}{N}\REP_N(\ak{\thetaN}) &\geq \max_{i=1,2}\frac{1}{N}
+\left| \log\left[ \frac{P_{\ak{\thetaN}}(\boldsymbol x_i)}{P_{\ak{\thetaN}}(\boldsymbol x_0)}\right] \right| \\
+&= (n-2) \max\left\{ \left| \theta_2 + \frac{\theta_3}{3}+\frac{\theta_1}{n-2} \right|, \frac{n}{4(n-1)} \left| \theta_2 + \frac{8\theta_1}{n-2} \right| \right\};
+\end{align*}
+a similar expression also holds for an odd node number \(n>2\).
+Consequently, for all fixed parameters excluding \(\theta_2=\theta_3=0\), \(\lim_{N\to \infty}\REP_N(\ak{\thetaN})/N=\infty\) then follows and the graph model with 2-stars and triangles is S-unstable, as suggested by the breach of Proposition \ref{prp:prop1} for this model when \(|\theta_2|+|\theta_3|\neq 0\).
+\EndKnitrBlock{remark}
+
+\BeginKnitrBlock{remark}
+\iffalse{} {Remark.
+} \fi{}Let \(M \geq 1\) denote a possible number of replications and consider data \(\boldsymbol Y_{N,M} \equiv (\boldsymbol X^{(1)}_N, \dots, \boldsymbol X^{(M)}_N)\) formed by \(\{ \boldsymbol X^{(j)}_N\}_{j=1}^M\) as \(M\) iid replications of a random vector \(\boldsymbol X_N=(X_1,\ldots,X_N)\), where the latter follows a FOES model with probabilities \(P_{\ak{\thetaN}}(\boldsymbol x)>0\), \(\boldsymbol x\in\mathcal{X}^N\).
+This leads to a joint model, say \(P_{\ak{\thetaN}}(\boldsymbol y)\), \(\boldsymbol y\in\mathcal{X}^{NM}\), for \(\boldsymbol Y_{N,M}\) consisting of \(N*M\) random variables in total.
+Then, the LREP for \(\boldsymbol Y_{N,M}\), scaled by associated size, is given by
+\begin{align*}
+\frac{1}{NM}\REP_{\boldsymbol Y_{N,M}}(\ak{\thetaN} ) &\equiv \frac{1}{NM}\log\left[\frac{\max_{\boldsymbol y \in \mathcal{X}^{NM}}P_{\ak{\thetaN}}(\boldsymbol y)  }{\min_{\boldsymbol y \in \mathcal{X}^{NM}}P_{\ak{\thetaN}}(\boldsymbol y)} \right] \\
+&= \frac{1}{NM}\log\left[\frac{\max_{\boldsymbol x \in \mathcal{X}^{N}}P_{\ak{\thetaN}}(\boldsymbol x)  }{\min_{\boldsymbol x \in \mathcal{X}^{N}}P_{\ak{\thetaN}}(\boldsymbol x)} \right]^M \equiv \frac{1}{N}\REP_{\boldsymbol X_{N}}(\ak{\thetaN} ),
+\end{align*}
+where \(\REP_{\boldsymbol X_{N}}(\ak{\thetaN} ) \equiv \REP(\ak{\thetaN} )\) denotes the log-ratio of extremal probabilities for \(\boldsymbol X_N\) defined from \eqref{eq:elpr}.
+That is, due to iid properties, the sample-size corrected LREP for \(\boldsymbol Y_{N,M}\) equals the analog, \(\REP(\ak{\thetaN} )/N\), from the underlying common data model for \(\boldsymbol X_N\) alone, regardless of the level \(M \geq 1\) of independent replication.
+Hence, the definition of an S-unstable model is unaffected by independent replication.
+For computational purposes, this aspect also implies that if the original data \(\boldsymbol X_N=(X_1,\ldots,X_N)\) in a FOES model consist of \(N\) iid random variables, then the size-scaled log-ratio \eqref{eq:elpr} may be calculated as
+\[
+\frac{1}{N}\REP( \ak{\thetaN} ) \equiv \frac{1}{N}\REP_{\boldsymbol X_{N}}( \ak{\thetaN} ) = \log\left[ \frac{\max_{ x \in \mathcal{X}}P_{\ak{\thetaN}}(X_1=x)} {\min_{ x \in \mathcal{X}}P_{\ak{\thetaN}}(X_1=x)} \right]
+\]
+based on the extremal probabilities of just one random variable \(X_1\).
+\EndKnitrBlock{remark}
+
+\hypertarget{details-on-centered-graph-example}{%
+\section{Details on Centered Graph Example}\label{details-on-centered-graph-example}}
+
+\ak{To illustrate centered model parameterizations and n examination of stability in these, consider the two-star model (\ref{eq:mod2}) for the $N = {n \choose 2}$ edges in simple graph with $n$ nodes and binary edge-variables $(X_1,\ldots,X_N)\in\{0,1\}^N$.
+Here a common or standard parameterization in (\ref{eq:mod2}) leads to a conditional probability of ``$1$" for an edge $i$ as
+$$
+\mathrm{logit}[  P_{\thetaN}(X_{i} = 1 | X_j=x_j, j\neq i)=  \theta_1 + \theta_2 \sum_{j \in \mathcal{N}_i } x_j,$$ 
+based on summing other edge observations $x_j$ in a neighborhood $\mathcal{N}_i = \{{s}_j: {s}_i \cap {s}_j \neq \emptyset\}$ to edge $i$ (i.e., edges $j$,  marked by pairs of graph vertices ${s}_{j} = \{ v_{j_1}, v_{j_2} \}$, that share a common vertex with edge $i$ marked by the vertex pair ${s}_i =\{v_{i_1}, v_{i_2}\}$).
+In contrast, a centered conditional would yield
+$$
+\mathrm{logit}[  P_{\thetaN}(X_{i} = 1 | X_j=x_j, j\neq i)=  \theta_1 +  \frac{\theta_2}{2(n-1)}\sum_{j \in \mathcal{N}_i} (x_{j}-\kappa),
+$$
+involving a parameter $\theta_1 \equiv \mathrm{logit}(\kappa)$ for $\kappa \in (0,1)$ and $2(n-2)$ as the size of $\mathcal{N}_i$ (cf.
+\citep{kaiser2009exploring}); the corresponding joint model would involve parameters $\theta_{1,\thetaidx}= \mathrm{logit}(\kappa) - \kappa\theta_2$
+and $\theta_{2,\thetaidx} = \theta_2/(2(n-2))$ in (\ref{eq:mod2}).
+The purpose of the centerization is to have $\kappa \in(0,1)$ represent a model mean parameter (note $E X_i =\kappa$ is the edge proportion/probability under independence $\theta_2=0$), while a separate parameter $\theta_2$ (for dependence) modifies the conditional probability of ``$X_i=1$" up/down from $\kappa$, depending on neighbors $x_j=1$ or 0.
+A similar interpretation does not hold in the uncentered model; see \citep{kaiser2009exploring} and \citep{casleton2017local} for a discussion of the centered parameterization in spatial and network modeling, where the intent is to improve parameter interpretation and help detect degeneracy (e.g., intuitively given by large dependence parameters $\theta_2$, so that $\kappa$ fails to correspond to the mean or the marginal proportion of $1$'s in data generations from the model).
+However, from a different perspective using the model measures here, centered parameterizations also lead to models which are more stable over wider regions of the parameter space.
+To illustrate, in the standard/uncentered parameterization for a graph model (\ref{eq:mod2}) with only two-stars $\thetaN = (0,\theta_2,0)$, our measure $\REP(\thetaN)/N$ of model instability becomes $\REP(\thetaN)/N = |\theta_2| (n-2)$, which is unbounded as $N\to \infty$ for all non-zero parameters $\theta_2$ (i.e., all models with $\theta_2\neq 0$ are S-unstable).
+However, in the centered parameterization  with only two-stars, corresponding to $\kappa=1/2$ and $\thetaN= 0.5*(-\theta_2,\theta_2/(n-2),0)$, the measure becomes
+\begin{align*}
+   \REP(\thetaN)/N  &= 0.5*|\theta_2|  \max_{x_i, y_i \in\{0,1\}}\left|  \frac{1}{N}\sum_{i=1}^N x_i \left( 1 - \frac{1}{2(n-2)} \sum_{j \in \mathcal{N}_i} x_j\right) - \frac{1}{N}\sum_{i=1}^N y_i \left( 1 - \frac{1}{2(n-2)} \sum_{j \in \mathcal{N}_i} y_j\right)\right| \\
+   &\leq |\theta_2|,   
+\end{align*}
+which is bounded, for any fixed $\theta_2 \in\mathbb{R}$, as $N$ increases (i.e., note $\sum_{j \in \mathcal{N}_j} x_i/(2(n-2)) \in [0,1]$ is a conditional/neighborhood sample proportion while $\sum_{i=1}^N x_i/N \in[0,1]$ is a marginal proportion).
+This aspect owes to adjusting parameters by neighborhood sizes in centered conditional distributions, but centering also induces an additional effect of alternating signs in parameters (e.g., $\thetaN= 0.5*(-\theta_2,\theta_2/(n-2),0)$).
+The latter has been suggested in other contexts with exponential graph models for regulating degeneracy \citep[cf.][]{snijders2006new}.}
 

### 2019-11-16 presentations/silo_wisc/slides.Rmd
silo presentation

--- 2018-05-08 presentations/msmlc/slides.Rmd
@@
 author: Andee Kaplan
-shortname: Kaplan, et al.
+shortname: andee.kaplan@colostate.edu
 institute: |
-    | Duke University
-    | andrea.kaplan@duke.edu
-shortinstitute: andrea.kaplan@duke.edu
+    | Colorado State University
+    | andee.kaplan@colostate.edu
 date: |
-  | May 9, 2018
+  | November 20, 2019
   |
-  | Slides available at <http://bit.ly/kaplan-msmlc>
+  | Slides available at <http://bit.ly/kaplan-silo>
   | 
   | \footnotesize Joint work with D. Nordman and S. Vardeman
-shortdate: "May 9, 2018"
+shortdate: "November 20, 2019"
 output: 
   beamer_presentation:
+    keep_tex: true
     template: ../resources/latex/beamer.tex
@@
 nocite: |
-    @salakhutdinov2009deep, @kaiser2007statistical
+    @salakhutdinov2009deep, @kaiser2007statistical, @kaplan2019properties
 ---
@@
 1. The model properties are largely unexplored in the literature
-2. The commonly cited fitting methodology remains heuristic-based and abstruse [@hinton2006fast]
+2. The commonly cited fitting methodology remains heuristic-based [@hinton2006fast]
 
@@
 2. Explore the possibility of a rigorous fitting methodology
+
+#
+
+\begin{center}
+\Huge{Properties of RBM Model Class}
+\end{center}
 
@@
 - All probability will stack up on mode sets or potentially those few outcomes with the highest probability
-- Proofs found in [@kaplan2016note]
 
@@
 
+<!--
 # RBM quantities to compare
@@
 \end{align*}
+-->
 
-# Data coding to mitigate degeneracy
+<!-- # Data coding to mitigate degeneracy -->
 
-\vspace{-.1in}
-\begin{figure}
-\includegraphics[width=.55\linewidth]{../resources/images/toyhull_top.png}
-\includegraphics[width=.35\linewidth]{../resources/images/toyhull_top_2.png}
-\caption{Convex hull of the "statistic space" $\mathcal{T} = \{(v_1, h_1, v_1 h_1): v_1, h_1 \in \mathcal{C}\}$ for a toy RBM with one visible and one hidden node for $\mathcal{C} = \{0,1\}$ (left) and $\mathcal{C} = \{-1,1\}$ (right) data encoding.}
-\end{figure}
-\vspace{-.15in}
-- For the $\mathcal{C} = \{-1, 1 \}$ encoding of hiddens $(H_1, \dots, H_\nh)$ and visibles $(V_1, \dots, V_\nv)$, the origin is the center of the parameter space.
-- At $\boldsymbol \theta = \boldsymbol 0$, RBM is equivalent to elements of $X$ being distributed as iid Bernoulli$\left(\frac{1}{2}\right)$ $\Rightarrow$ No *near-degeneracy*, *instability*, or *uninterpretability*!
+<!-- \vspace{-.1in} -->
+<!-- \begin{figure} -->
+<!-- \includegraphics[width=.55\linewidth]{../resources/images/toyhull_top.png} -->
+<!-- \includegraphics[width=.35\linewidth]{../resources/images/toyhull_top_2.png} -->
+<!-- \caption{Convex hull of the "statistic space" $\mathcal{T} = \{(v_1, h_1, v_1 h_1): v_1, h_1 \in \mathcal{C}\}$ for a toy RBM with one visible and one hidden node for $\mathcal{C} = \{0,1\}$ (left) and $\mathcal{C} = \{-1,1\}$ (right) data encoding.} -->
+<!-- \end{figure} -->
+<!-- \vspace{-.15in} -->
+<!-- - For the $\mathcal{C} = \{-1, 1 \}$ encoding of hiddens $(H_1, \dots, H_\nh)$ and visibles $(V_1, \dots, V_\nv)$, the origin is the center of the parameter space. -->
+<!-- - At $\boldsymbol \theta = \boldsymbol 0$, RBM is equivalent to elements of $X$ being distributed as iid Bernoulli$\left(\frac{1}{2}\right)$ $\Rightarrow$ No *near-degeneracy*, *instability*, or *uninterpretability*!
+-->
 
@@
 
-# Manageable (a.k.a. small) examples
+# FOES models
+
+Statements about near-degeneracy, instability, and uninterpretability hold for a general class of models that includes the RBM.
+
+- $\boldsymbol X_N = (X_1, \dots, X_N)$ collection of discrete random variables with a finite sample space
+
+- $\mathcal{X}^N$ finite sample space, represented as some $N$-fold Cartesian  product
+
+- $P_{{\thetaN}}$ probability model on $\mathcal{X}^N$ for each $N$
+
+- Assume that the model support of $P_{{\thetaN}}$ is the sample space $\mathcal{X}^N$.
+
+#### Definition
+
+This framework produces *Finite Outcome Everywhere Supported (FOES)* models, $P_{{\thetaN}}$, indexed by a defining sequence of parameters ${\thetaN}$, to describe data $\boldsymbol X_N$ of any given sample size $N \geq 1$.
+
+[@kaplan2016note]
+
+# Examples
+
+- Discrete exponential family models
+
+- Discrete restricted Boltzmann machines
+
+- Binary deep learning models
+
+  \vspace{.1in}
+
+  1. **Deep Boltzmann machine (DBM).** Conditional independence within all layers, stacked RBM models with conditional dependence between neighboring layers
+
+  \vspace{.1in}
+
+  2.
+**Deep belief network (DBN).** Multiple layers of latent random variables stacked in a deep architecture with no conditional dependence between layers;  all but the last stacked layer are Bayesian networks (directed dependence), rather than RBMs
+
+# Visualizing properties in manageable (a.k.a. small) RBMs
 
@@
 ```
+
+#
+
+\begin{center}
+\Huge{Rigorous Fitting Methodology}
+\end{center}
 
@@
 
-- RBMs used for classification, but are concerning as statistical models due to *near-degeneracy*, *S-instability*, and *uninterpretability*
-- Rigorous fitting methodology is difficult due to the dimension of the parameter space & size of the latent variable space
-- Fitting a RBM model is also questionable as any distribution for the visibles can be approximated arbitrarily well
+- RBMs shown success for classification, but concerning as statistical models due to *near-degeneracy*, *S-instability*, and *uninterpretability*
+- Rigorous fitting methodology is difficult 
+    - Numerical complications in likelihood maximization may occur due to degeneracy and S-instability when all probability in pushed to opposite extremes in the sample space 
+    - For Bayesian inference, S-instability may hinder effective chain mixing
+- RBMs questionably useful as any distribution for the visibles can be approximated arbitrarily well
     - The empirical distribution of visibles is the best fitting model for observed cell data 
@@
 
-Skeptical that any model built using RBMs (i.e. deep Boltzmann machine) can achieve useful **prediction** or **inference** in a principled way without limiting the flexibility of the fitted model
+Skeptical that any model built using RBMs (e.g., DBM) can achieve useful **inference** in principled way without limiting flexibility of the fitted model
 
-# Future work
+# Relevant papers
 
-1. Generalization of instability results for other network models [ongoing, see @kaplan2016note]
 \vfill
-2. Image classification
-    - Ensemble methods (super learners) using AdaBoost [@freund1995desicion] 
-    - Decision theoretic based approach to approximating the likelihood ratio test for classification
+
+1.
+Kaplan, Andee, Daniel Nordman, and Stephen Vardeman.
+"Properties and Bayesian fitting of restricted Boltzmann machines." *Statistical Analysis and Data Mining: The ASA Data Science Journal* 12.1 (2019): 23-38.
+
+\vspace{0.25in}
+
+2. Kaplan, Andee, Daniel J. Nordman, and Stephen B. Vardeman. "On the S-instability and degeneracy of discrete deep learning models." *Information and Inference: A Journal of the IMA* (2019).
+
 \vfill
@@
 
-* Slides -- <http://bit.ly/kaplan-msmlc>
+* Slides -- <http://bit.ly/kaplan-silo>
 
 * Contact
-    * Email -- <andrea.kaplan@duke.edu>
+    * Email -- <andee.kaplan@colostate.edu>
     * Twitter -- <http://twitter.com/andeekaplan>

### 2019-11-16 presentations/silo_wisc/slides.tex
silo presentation

\documentclass[ignorenonframetext,]{beamer}
\setbeamertemplate{caption}[numbered]
\setbeamertemplate{caption label separator}{: }
\setbeamercolor{caption name}{fg=normal text.fg}
\beamertemplatenavigationsymbolsempty
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
\usetheme[]{CambridgeUS}
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\newif\ifbibliography
\usepackage{longtable,booktabs}
\usepackage{caption}
% These lines are needed to make table captions work with longtable:
\makeatletter
\def\fnum@table{\tablename~\thetable}
\makeatother
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight0.8\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}

% Prevent slide breaks in the middle of a paragraph:
\widowpenalties 1 10000
\raggedbottom

\AtBeginPart{
  \let\insertpartnumber\relax
  \let\partname\relax
  \frame{\partpage}
}
\AtBeginSection{
  \ifbibliography
  \else
    \let\insertsectionnumber\relax
    \let\sectionname\relax
    \frame{\sectionpage}
  \fi
}
\AtBeginSubsection{
  \let\insertsubsectionnumber\relax
  \let\subsectionname\relax
  \frame{\subsectionpage}
}

\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
\usepackage{multirow}

\newtheorem{result}{Result}

\DeclareMathOperator*{\argmin}{arg\,min}
\DeclareMathOperator*{\argmax}{arg\,max}

\newtheorem{prop}{Proposition}
\newtheorem{apptheorem}{Theorem}[section]
\newtheorem{applemma}{Lemma}[section]

\theoremstyle{definition}
\newtheorem{defn}{Definition}
\newtheorem{assumption}{Assumption}
\newtheorem{remark}{Remark}

\newcommand{\nv}{{n_{\scriptscriptstyle V}}}
\newcommand{\nh}{{n_{\scriptscriptstyle H}}}
\newcommand{\E}{E}
\newcommand{\thetaN}{\boldsymbol \theta_{\thetaidx}}
\newcommand{\thetaidx}{q(N)}

%beamer bug
\widowpenalties 1 150

\title[Model matters with RBMs]{Model matters with restricted Boltzmann machines}

\author[\href{mailto:andee.kaplan@colostate.edu}{\nolinkurl{andee.kaplan@colostate.edu}}]{Andee Kaplan}
\institute[]{Colorado State University\\
\href{mailto:andee.kaplan@colostate.edu}{\nolinkurl{andee.kaplan@colostate.edu}}}
\date[November 20, 2019]{November 20, 2019\\
~\\
Slides available at \url{http://bit.ly/kaplan-silo}\\
~\\
\footnotesize Joint work with D. Nordman and S. Vardeman}

\begin{document}
\frame{\titlepage}

\begin{frame}{What is this?}
\protect\hypertarget{what-is-this}{}

A restricted Boltzman machine (RBM) is an undirected probabilistic
graphical model with

\begin{enumerate}
\tightlist
\item
  two layers of random variables - one hidden and one visible
\item
  conditional independence within a layer (Smolensky 1986)
\end{enumerate}

\begin{figure}
\includegraphics[width=.6\linewidth]{../resources/images/rbm.png}
\caption{Hidden nodes are indicated by white circles and the visible nodes are indicated by blue circles.}
\end{figure}

\end{frame}

\begin{frame}{How is it used?}
\protect\hypertarget{how-is-it-used}{}

\begin{itemize}
\tightlist
\item
  Supervised learning, specifically image classification
\end{itemize}

\begin{figure}
\includegraphics[width=\linewidth]{../resources/images/visibles_image.pdf}
\caption{Image classification using a RBM: each image pixel comprises a node in the visible layer, $\mathcal{V}$ and the output of the RBM is used to create features passed to a supervised learning algorithm.}
\end{figure}

\end{frame}

\begin{frame}{Joint distribution}
\protect\hypertarget{joint-distribution}{}

\begin{itemize}
\item
  \(\boldsymbol x = (h_1, \dots, h_{\nh}, v_1,\dots,v_{\nv})\)
  represents visible and hidden nodes in a RBM
\item
  Each single ``binary'' random variable, visible \(v_i\) or hidden
  \(h_j\), takes values in a common coding set

  \begin{itemize}
  \tightlist
  \item
    \(\mathcal{C}=\{0,1\}\) or \(\mathcal{C}=\{-1,1\}\).
  \end{itemize}
\item
  A parametric form for probabilities

  \begin{align*} 
    \label{eqn:pmf} 
    f_{\boldsymbol \theta} (\boldsymbol x) = \frac{\exp\left(\sum\limits_{i = 1}^{\nv} \sum\limits_{j=1}^{\nh} \theta_{ij} v_i h_j + \sum\limits_{i = 1}^{\nv}\theta_{v_i} v_i + \sum\limits_{j = 1}^{\nh}\theta_{h_j} h_j\right)}{\gamma(\boldsymbol \theta)} \end{align*}

  where

  \[\gamma(\boldsymbol \theta) = \sum\limits_{\boldsymbol x \in \mathcal{C}^{\nh + \nv}}\exp\left(\sum\limits_{i = 1}^{\nv} \sum\limits_{j=1}^{\nh} \theta_{ij} v_i h_j + \sum\limits_{i = 1}^{\nv}\theta_{v_i} v_i + \sum\limits_{j = 1}^{\nh}\theta_{h_j} h_j\right)\]
\end{itemize}

\end{frame}

\begin{frame}{Deep learning}
\protect\hypertarget{deep-learning}{}

\begin{columns}[T] % align columns
\begin{column}{.48\textwidth}
\begin{itemize}
\item Stacking layers of RBMs in a deep architecture
\item Proponents claim the ability to learn "internal representations that become increasingly complex, which is considered to be a promising way of solving object and speech recognition problems" (Salakhutdinov and Hinton 2009, pp. 450).
\end{itemize}
\end{column}
\hfill
\begin{column}{.48\textwidth}
\begin{figure}
\includegraphics[width=\linewidth]{../resources/images/deep_rbm.png}
\caption{Three layer deep Boltzmann machine, with visible-to-hidden and hidden-to-hidden connections but no within-layer connections.}
\end{figure}
\end{column}
\end{columns}

\end{frame}

\begin{frame}{Why do I care?}
\protect\hypertarget{why-do-i-care}{}

\begin{enumerate}
\tightlist
\item
  The model properties are largely unexplored in the literature
\item
  The commonly cited fitting methodology remains heuristic-based
  (Hinton, Osindero, and Teh 2006)
\end{enumerate}

We want to

\begin{enumerate}
\tightlist
\item
  Provide steps toward understanding properties of the model class from
  the perspective of statistical theory
\item
  Explore the possibility of a rigorous fitting methodology
\end{enumerate}

\end{frame}

\begin{frame}{}
\protect\hypertarget{section}{}

\begin{center}
\Huge{Properties of RBM Model Class}
\end{center}

\end{frame}

\begin{frame}{Degeneracy, instability, and uninterpretability. Oh my!}
\protect\hypertarget{degeneracy-instability-and-uninterpretability.-oh-my}{}

The highly flexible nature of a RBM (\(\nh + \nv + \nh*\nv\) parameters)
makes at least three kinds of potential model impropriety of concern

\begin{enumerate}
\tightlist
\item
  \emph{degeneracy}
\item
  \emph{instability}, and
\item
  \emph{uninterpretability}
\end{enumerate}

\begin{quote}
A model should ``provide an explanation of the mechanism underlying the
observed phenomena'' (G. E. P. Box 1967).
\end{quote}

RBMs often

\begin{itemize}
\tightlist
\item
  fail to generate data with realistic variability and thus an
  unsatisfactory conceptualization of the data generation process (Li
  2014)
\item
  exhibit model instability (over-sensitivity) (Szegedy et al. 2013;
  Nguyen, Yosinski, and Clune 2014)
\end{itemize}

\end{frame}

\begin{frame}{Near-degeneracy}
\protect\hypertarget{near-degeneracy}{}

\begin{definition}[Model Degeneracy]
A disproportionate amount of probability is placed on only a few elements of the sample space, $\mathcal{C}^{\nh + \nv}$, by the model.
\end{definition}

RBM models exhibit \emph{near-degeneracy} when random variables in
\[Q_{\boldsymbol \theta}(\boldsymbol x) = \sum\limits_{i = 1}^{\nv} \sum\limits_{j=1}^{\nh} \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j,
\] have a mean vector \(\boldsymbol \mu(\boldsymbol \theta)\) close to
the boundary of the convex hull of
\(\mathcal{T} = \{\boldsymbol t(x): x \in \mathcal{C}^{\nh + \nv}\}\)
(Handcock 2003), where
\[\boldsymbol t(\boldsymbol x) = \{v_1, \dots, v_{\nv}, h_1, \dots, h_{\nh}, v_1 h_1, \dots, v_{\nv} h_{\nh} \}\]
and
\[\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta} \boldsymbol t(\boldsymbol X)\].

\end{frame}

\begin{frame}{Instability}
\protect\hypertarget{instability}{}

\begin{definition}[Instability]
Characterized by excessive sensitivity in the model, where small changes in the components of data outcomes, $\boldsymbol x$, lead to substantial changes in probability.
\end{definition}

\begin{itemize}
\tightlist
\item
  Concept of model deficiency related to \emph{instability} for a class
  of exponential families of distributions (Schweinberger 2011)
\item
  For the RBM, consider how model incorporates more visibles

  \begin{itemize}
  \tightlist
  \item
    Model parameters in a longer sequence
    \(\boldsymbol \theta_{\nv} \in \mathbb{R}^{\nv + \nh + \nv*\nh}, \nv \ge 1\)
  \item
    May also arbitrarily expand the number of hidden variables used
  \end{itemize}
\end{itemize}

\end{frame}

\begin{frame}{Unstable RBMs}
\protect\hypertarget{unstable-rbms}{}

\begin{definition}[S-unstable RBM]
A RBM model formulation is \emph{S-unstable} if
\begin{align*}
\lim\limits_{\nv \rightarrow \infty} \frac{1}{\nv} \text{LREP}(\boldsymbol \theta_{\nv}) = \infty.
\end{align*}
where 
\begin{align}
\label{eqn:elpr}
\text{LREP}(\boldsymbol \theta_{\nv}) = \log \left[\frac{\max\limits_{(v_1, \dots, v_{\nv}) \in \mathcal{C}^\nv}P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}{\min\limits_{(v_1, \dots, v_\nv) \in \mathcal{C}^\nv}P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}\right]
\end{align}
\end{definition}

S-unstable RBM models are undesirable for several reasons - small
changes in data outcomes can lead to overly-sensitive changes in
probability.

\end{frame}

\begin{frame}{One-pixel change}
\protect\hypertarget{one-pixel-change}{}

Consider the biggest log-probability ratio for a one-pixel (one
component) change in visibles (data outcomes)

\[
\Delta(\boldsymbol \theta_\nv) \equiv \max \left\{\log \frac{P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}{P_{\boldsymbol \theta_\nv}(v_1^*, \dots, v_\nv^*)} \right\},
\]

where
\((v_1, \dots, v_\nv) \& (v_1^*, \dots, v_\nv^*) \in \mathcal{C}^\nv\)
differ by exactly one component

\begin{result}
\label{prop:instab}
Let $c > 0$ and fix an integer $\nv \ge 1$. If $\frac{1}{\nv}\text{LREP}(\boldsymbol \theta_\nv) > c$, then $\Delta(\boldsymbol \theta_\nv) > c$.
\end{result}

If the \(\nv^{-1}LREP(\boldsymbol\theta_{\nv})\) is too large, then a
RBM model will exhibit large probability shifts for very small changes
in the data configuration.

\end{frame}

\begin{frame}{Tie to degeneracy}
\protect\hypertarget{tie-to-degeneracy}{}

Define an arbitrary modal set of possible outcomes (i.e.~set of highest
probability outcomes) for a given \(0 < \epsilon < 1\) as

\small

\begin{align*}
M_{\epsilon, \boldsymbol \theta_\nv} \equiv \left\{\boldsymbol v \in \mathcal{C}^\nv: \log P_{\boldsymbol \theta_\nv}(\boldsymbol v) > (1-\epsilon)\max\limits_{\boldsymbol v^*}P_{\boldsymbol \theta_\nv}(\boldsymbol v^*) + \epsilon\min\limits_{\boldsymbol v^*}P_{\boldsymbol \theta_\nv}(\boldsymbol v^*) \right\}
\end{align*} \normalsize

\begin{result}
\label{prop:degen}
For an S-unstable RBM model, and for any given $0 < \epsilon < 1$, $P_{\boldsymbol \theta_\nv}\left((v_1, \dots, v_\nv) \in M_{\epsilon, \boldsymbol \theta_\nv}\right) \rightarrow 1$ holds as $\nv \rightarrow \infty$.
\end{result}

\begin{itemize}
\tightlist
\item
  All probability will stack up on mode sets or potentially those few
  outcomes with the highest probability
\end{itemize}

\end{frame}

\begin{frame}{Uninterpretability}
\protect\hypertarget{uninterpretability}{}

\begin{definition}[Uninterpretability]
Characterized by marginal mean-structure (controlled by main effect parameters $\theta_{v_i}, \theta_{h_j}$) not being maintained in the model due to dependence (interaction parameters $\theta_{ij}$) (Kaiser 2007).
\end{definition}

\begin{itemize}
\tightlist
\item
  Model expectations,
  E\(\left[\boldsymbol X | \boldsymbol \theta\right]\)
\item
  Expectations given independence,
  E\(\left[\boldsymbol X | \boldsymbol \theta^* \right]\), where
  \(\boldsymbol \theta^*\) matches \(\boldsymbol \theta\) for all main
  effects but otherwise has \(\theta_{ij} = 0\) for
  \(i = 1, \dots, \nv, j = 1, \dots, \nh\)
\item
  If
  \(|\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] - \text{E}\left[\boldsymbol X | \boldsymbol \theta^*\right]|\)
  is large then the RBM with parameter vector \(\boldsymbol \theta\) is
  \emph{uninterpetable}
\end{itemize}

\end{frame}

\begin{frame}{FOES models}
\protect\hypertarget{foes-models}{}

Statements about near-degeneracy, instability, and uninterpretability
hold for a general class of models that includes the RBM.

\begin{itemize}
\item
  \(\boldsymbol X_N = (X_1, \dots, X_N)\) collection of discrete random
  variables with a finite sample space
\item
  \(\mathcal{X}^N\) finite sample space, represented as some \(N\)-fold
  Cartesian product
\item
  \(P_{{\thetaN}}\) probability model on \(\mathcal{X}^N\) for each
  \(N\)
\item
  Assume that the model support of \(P_{{\thetaN}}\) is the sample space
  \(\mathcal{X}^N\).
\end{itemize}

\begin{block}{Definition}

This framework produces \emph{Finite Outcome Everywhere Supported
(FOES)} models, \(P_{{\thetaN}}\), indexed by a defining sequence of
parameters \({\thetaN}\), to describe data \(\boldsymbol X_N\) of any
given sample size \(N \geq 1\).

(Kaplan, Nordman, and Vardeman 2019a)

\end{block}

\end{frame}

\begin{frame}{Examples}
\protect\hypertarget{examples}{}

\begin{itemize}
\item
  Discrete exponential family models
\item
  Discrete restricted Boltzmann machines
\item
  Binary deep learning models

  \vspace{.1in}

  \begin{enumerate}
  \tightlist
  \item
    \textbf{Deep Boltzmann machine (DBM).} Conditional independence
    within all layers, stacked RBM models with conditional dependence
    between neighboring layers
  \end{enumerate}

  \vspace{.1in}

  \begin{enumerate}
  \setcounter{enumi}{1}
  \tightlist
  \item
    \textbf{Deep belief network (DBN).} Multiple layers of latent random
    variables stacked in a deep architecture with no conditional
    dependence between layers; all but the last stacked layer are
    Bayesian networks (directed dependence), rather than RBMs
  \end{enumerate}
\end{itemize}

\end{frame}

\begin{frame}{Visualizing properties in manageable (a.k.a. small) RBMs}
\protect\hypertarget{visualizing-properties-in-manageable-a.k.a.-small-rbms}{}

\begin{itemize}
\tightlist
\item
  To explore the effects of RBM parameters \(\boldsymbol \theta\) on
  \emph{near-degeneracy}, \emph{instability}, and
  \emph{uninterpretability}, consider models of small size\\
\item
  For \(\nh, \nv \in \{1, \dots, 4\}\), sample 100 values of
  \(\boldsymbol \theta\)

  \begin{enumerate}
  \tightlist
  \item
    Split \(\boldsymbol \theta\) into
    \(\boldsymbol \theta_{interaction}\) and
    \(\boldsymbol \theta_{main}\)
  \item
    Allow the two types of terms to have varying average magnitudes,
    \(||\boldsymbol \theta_{main} || /(\nh+\nv)\) and
    \(||\boldsymbol \theta_{interaction} || /(\nh*\nv)\)
  \item
    Average magnitudes vary on a grid between 0.001 and 3 with 24
    breaks, yielding 576 grid points
  \end{enumerate}
\item
  Calculate metrics of model impropriety,
  \(\boldsymbol \mu(\boldsymbol \theta)\),
  \(\text{LREP}(\boldsymbol \theta)/\nv\), and the coordinates of
  \(\left|\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] - \text{E}\left[\boldsymbol X | \boldsymbol \theta^* \right] \right|\).
\item
  In the case of \emph{near-degeneracy}, classify each model as
  near-degenerate or ``viable'' based on the distance of
  \(\boldsymbol \mu(\boldsymbol \theta)\) from the boundary of the
  convex hull of \(\mathcal{T}\)
\end{itemize}

\end{frame}

\begin{frame}{Simulation results}
\protect\hypertarget{simulation-results}{}

\begin{figure}
\includegraphics[width=.32\linewidth]{slides_files/figure-beamer/unnamed-chunk-1-1} \includegraphics[width=.32\linewidth]{slides_files/figure-beamer/unnamed-chunk-1-2} \includegraphics[width=.32\linewidth]{slides_files/figure-beamer/unnamed-chunk-1-3} \caption{\label{fig:degen_plots}The fraction of models that were near-degenerate (left), the sample mean value of $\text{LREP}(\boldsymbol \theta)/\nv$ (middle), and the sample mean of the maximum component of the absolute difference between the model expectation vector, E$\left[\boldsymbol X | \boldsymbol \theta\right]$, and the expectation vector given independence, E$\left[\boldsymbol X | \boldsymbol \theta^* \right ]$ (right).}\label{fig:unnamed-chunk-1}
\end{figure}

\end{frame}

\begin{frame}{}
\protect\hypertarget{section-1}{}

\begin{center}
\Huge{Rigorous Fitting Methodology}
\end{center}

\end{frame}

\begin{frame}{Model fitting}
\protect\hypertarget{model-fitting}{}

\begin{enumerate}
\tightlist
\item
  Computational concerns: Fitting a RBM via maximum likelihood (ML)
  methods infeasible due to the intractibility of the normalizing term
  \(\gamma(\boldsymbol \theta)\)

  \begin{itemize}
  \tightlist
  \item
    Ad hoc methods used to avoid this problem with stochastic ML
  \item
    Employ a small number of MCMC draws to approximate
    \(\gamma(\boldsymbol \theta)\)
  \end{itemize}
\item
  Model parameterization concerns: With enough hiddens,

  \begin{itemize}
  \tightlist
  \item
    Potential to re-create any distribution for the data (Le Roux and
    Bengio 2008; Montufar and Ay 2011; and Montúfar, Rauh, and Ay 2011)

    \begin{itemize}
    \tightlist
    \item
      The model for the cell probabilities that has the highest
      likelihood over \emph{all possible model classes} is the empirical
      distribution
    \item
      The RBM model ensures that this empirical distribution can be
      arbitrarily well approximated
    \end{itemize}
  \item
    When empirical distribution contains empty cells, ML will chase
    parameters to \(\infty\) in order to zero out corresponding RBM cell
    probabilities
  \end{itemize}
\end{enumerate}

\end{frame}

\begin{frame}{Bayesian methods}
\protect\hypertarget{bayesian-methods}{}

\begin{itemize}
\tightlist
\item
  Consider what might be done in a principled manner, small test
\item
  To avoid model impropriety, avoid parts of the parameter space
  \(\mathbb{R}^{\nv + \nh + \nv*\nh}\) leading to
  \emph{near-degeneracy}, \emph{instability}, and
  \emph{uninterpretability}.

  \begin{itemize}
  \tightlist
  \item
    Shrink \(\boldsymbol \theta\) toward \(\boldsymbol 0\)

    \begin{enumerate}
    \tightlist
    \item
      Specify priors that place low probability on large values of
      \(||\boldsymbol \theta||\)
    \item
      Shrink \(\boldsymbol \theta_{interaction}\) more than
      \(\boldsymbol \theta_{main}\)
    \end{enumerate}
  \end{itemize}
\item
  Consider a test case with \(\nv = \nh = 4\)

  \begin{itemize}
  \tightlist
  \item
    \(\boldsymbol \theta\) chosen as a sampled value from a grid point
    in figure \ref{fig:degen_plots} with \(< 5\)\% near-degeneracy (not
    near the convex hull of the sufficient statistics)
  \item
    simulate \(n = 5,000\) as a training set and fit the RBM using three
    Bayes methodologies
  \end{itemize}
\end{itemize}

\end{frame}

\begin{frame}{Fitting methodologies}
\protect\hypertarget{fitting-methodologies}{}

\begin{enumerate}
\tightlist
\item
  \emph{A ``trick'' prior (BwTPLV)}

  \begin{itemize}
  \tightlist
  \item
    Cancel out normalizing term in the likelihood
  \item
    Resulting full conditionals of \(\boldsymbol \theta\) are
    multivariate Normal
  \item
    \(h_j\) are carried along as latent variables \begin{align*}
     \pi(\boldsymbol \theta) \propto \gamma(\boldsymbol \theta)^n \exp\left(-\frac{1}{2C_{1}}\boldsymbol \theta_{main}'\boldsymbol \theta_{main} -\frac{1}{2C_{2}}\boldsymbol \theta_{interaction}'\boldsymbol \theta_{interaction}\right), \vspace{-.75cm}
     \end{align*} where \(C_{2} < C_{1}\) (Li 2014)
  \end{itemize}
\end{enumerate}

\end{frame}

\begin{frame}{Fitting methodologies (cont'd)}
\protect\hypertarget{fitting-methodologies-contd}{}

\begin{enumerate}
\setcounter{enumi}{1}
\tightlist
\item
  \emph{A truncated Normal prior (BwTNLV)}

  \begin{itemize}
  \tightlist
  \item
    Independent spherical normal distributions as priors for
    \(\boldsymbol \theta_{main}\) and
    \(\boldsymbol \theta_{interaction}\)

    \begin{itemize}
    \tightlist
    \item
      \(\sigma_{interaction} < \sigma_{main}\)
    \item
      \emph{truncated} at \(3\sigma_{main}\) and
      \(3\sigma_{interaction}\), respectively
    \end{itemize}
  \item
    Simulation from the posterior using a geometric adaptive MH step
    (Zhou 2014)
  \item
    \(h_j\) are carried along in the MCMC implementation as latent
    variables
  \end{itemize}
\item
  \emph{A truncated Normal prior and marginalized likelihood (BwTNML)}

  \begin{itemize}
  \tightlist
  \item
    Marginalize out \(\boldsymbol h\) in
    \(f_{\boldsymbol \theta}(\boldsymbol x)\)
  \item
    Use the truncated Normal priors applied to the marginal
    probabilities for visible variables (recall visibles are the
    observed data, hiddens are not)
  \end{itemize}
\end{enumerate}

\end{frame}

\begin{frame}{Hyperparameters}
\protect\hypertarget{hyperparameters}{}

\begin{table}[ht]
\centering
\caption{The values used for the hyperparameters for all three fitting methods. A rule of thumb is imposed which decreases prior variances for the model parameters as the size of the model increases and also shrinks $\boldsymbol \theta_{interaction}$ more than $\boldsymbol \theta_{main}$. The common $C$ defining $C_1$ and $C_2$  in the BwTPLV method is chosen by tuning.}
\label{tab:hyperparam}
\begin{tabular}{|l|c|c|}
\hline 
Method & Hyperparameter & Value \\ 
\hline \hline
\multirow{2}{*}{BwTPLV} & $C_1$ & $\frac{C}{n}\frac{1}{\nh + \nv}$ \\
 & $C_2$ & $\frac{C}{n}\frac{1}{\nh*\nv}$ \\
\hline
\multirow{2}{*}{BwTNLV} & $\sigma^2_{main}$ & $\frac{1}{\nh + \nv}$ \\
 & $\sigma^2_{interaction}$ & $\frac{1}{\nh*\nv}$ \\
\hline
\multirow{2}{*}{BwTNML} & $\sigma^2_{main}$ & $\frac{1}{\nh + \nv}$ \\
 & $\sigma^2_{interaction}$ & $\frac{1}{\nh*\nv}$ \\
\hline
\end{tabular}
\end{table}

\end{frame}

\begin{frame}{Mixing}
\protect\hypertarget{mixing}{}

The BwTNLV (2) and the BwTNML method (3) are drawing from the same
stationary posterior distribution for images.

\begin{figure}
\includegraphics[width=.7\linewidth]{slides_files/figure-beamer/acf-1} \caption{\label{fig:acf}The autocorrelation functions (ACF) for the posterior probabilities of all $2^4 = 16$ possible outcomes for the vector of $4$ visibles assessed at multiple lags for each method with BwTNLV in black and BwTNML in red.}\label{fig:acf}
\end{figure}

\end{frame}

\begin{frame}{Effective sample size}
\protect\hypertarget{effective-sample-size}{}

\begin{itemize}
\tightlist
\item
  Overlapping blockmeans approach (Gelman, Shirley, and others 2011)

  \begin{itemize}
  \tightlist
  \item
    Crude estimate for the aysmptotic variance of the probability of
    each image
  \item
    Compare it to an estimate of the asymptotic variance assuming IID
    draws from the target distribution
  \end{itemize}
\end{itemize}

\tiny

\begin{longtable}[]{@{}rrrrrr@{}}
\caption{\label{tab:m_eff}The effective sample sizes for a chain of
length \(M = 1000\) regarding all \(16\) probabilities for possible
vector outcomes of visibles. BwTNLV would require at least \(4.7\) times
as many MCMC iterations to achieve the same amount of effective
information about the posterior distribution.}\tabularnewline
\toprule
Outcome & BwTNLV & BwTNML & Outcome & BwTNLV & BwTNML\tabularnewline
\midrule
\endfirsthead
\toprule
Outcome & BwTNLV & BwTNML & Outcome & BwTNLV & BwTNML\tabularnewline
\midrule
\endhead
1 & 73.00 & 509.43 & 9 & 83.47 & 394.90\tabularnewline
2 & 65.05 & 472.51 & 10 & 95.39 & 327.35\tabularnewline
3 & 87.10 & 1229.39 & 11 & 70.74 & 356.56\tabularnewline
4 & 72.64 & 577.73 & 12 & 81.40 & 338.30\tabularnewline
5 & 71.67 & 452.01 & 13 & 105.98 & 373.59\tabularnewline
6 & 66.49 & 389.78 & 14 & 132.61 & 306.91\tabularnewline
7 & 84.30 & 660.37 & 15 & 82.15 & 365.30\tabularnewline
8 & 75.46 & 515.09 & 16 & 98.05 & 304.57\tabularnewline
\bottomrule
\end{longtable}

\end{frame}

\begin{frame}{Posterior distributions of images}
\protect\hypertarget{posterior-distributions-of-images}{}

\begin{figure}
\centering
\includegraphics{slides_files/figure-beamer/fitting_plot-1.pdf}
\caption{\label{fig:fitting_plot}Posterior probabilities of \(16 = 2^4\)
possible realizations of \(4\) visibles using two of the three Bayesian
fitting techniques, BwTPLV and BwTNML. Black lines show true
probabilities of each vector of visibles based on the parameters used to
generate the training data while red lines show the empirical
distribution.}
\end{figure}

\end{frame}

\begin{frame}{Wrapping up}
\protect\hypertarget{wrapping-up}{}

\begin{itemize}
\tightlist
\item
  RBMs shown success for classification, but concerning as statistical
  models due to \emph{near-degeneracy}, \emph{S-instability}, and
  \emph{uninterpretability}
\item
  Rigorous fitting methodology is difficult

  \begin{itemize}
  \tightlist
  \item
    Numerical complications in likelihood maximization may occur due to
    degeneracy and S-instability when all probability in pushed to
    opposite extremes in the sample space
  \item
    For Bayesian inference, S-instability may hinder effective chain
    mixing
  \end{itemize}
\item
  RBMs questionably useful as any distribution for the visibles can be
  approximated arbitrarily well

  \begin{itemize}
  \tightlist
  \item
    The empirical distribution of visibles is the best fitting model for
    observed cell data
  \item
    There can be no ``smoothed distribution'' achieved in a RBM model of
    sufficient size with a rigorous likelihood-based method
  \end{itemize}
\end{itemize}

Skeptical that any model built using RBMs (e.g., DBM) can achieve useful
\textbf{inference} in principled way without limiting flexibility of the
fitted model

\end{frame}

\begin{frame}{Relevant papers}
\protect\hypertarget{relevant-papers}{}

\vfill

\begin{enumerate}
\tightlist
\item
  Kaplan, Andee, Daniel Nordman, and Stephen Vardeman. ``Properties and
  Bayesian fitting of restricted Boltzmann machines.'' \emph{Statistical
  Analysis and Data Mining: The ASA Data Science Journal} 12.1 (2019):
  23-38.
\end{enumerate}

\vspace{0.25in}

\begin{enumerate}
\setcounter{enumi}{1}
\tightlist
\item
  Kaplan, Andee, Daniel J. Nordman, and Stephen B. Vardeman. ``On the
  S-instability and degeneracy of discrete deep learning models.''
  \emph{Information and Inference: A Journal of the IMA} (2019).
\end{enumerate}

\vfill

\end{frame}

\begin{frame}{Thank you}
\protect\hypertarget{thank-you}{}

\begin{itemize}
\item
  Slides -- \url{http://bit.ly/kaplan-silo}
\item
  Contact

  \begin{itemize}
  \tightlist
  \item
    Email --
    \href{mailto:andee.kaplan@colostate.edu}{\nolinkurl{andee.kaplan@colostate.edu}}
  \item
    Twitter -- \url{http://twitter.com/andeekaplan}
  \item
    GitHub -- \url{http://github.com/andeek}
  \end{itemize}
\end{itemize}

\end{frame}

\begin{frame}{Appendix: Parameters used}
\protect\hypertarget{appendix-parameters-used}{}

\scriptsize

\begin{longtable}[]{@{}lrlrlr@{}}
\caption{\label{tab:theta}Parameters used to fit a test case with
\(\nv = \nh = 4\). This parameter vector was chosen as a sampled value
of \(\boldsymbol \theta\) that was not near the convex hull of the
sufficient statistics for a grid point in figure \ref{fig:degen_plots}
with \(< 5\)\% near-degeneracy.}\tabularnewline
\toprule
Parameter & Value & Parameter & Value & Parameter & Value\tabularnewline
\midrule
\endfirsthead
\toprule
Parameter & Value & Parameter & Value & Parameter & Value\tabularnewline
\midrule
\endhead
\(\theta_{v1}\) & -1.1043760 & \(\theta_{11}\) & -0.0006334 &
\(\theta_{31}\) & -0.0038301\tabularnewline
\(\theta_{v2}\) & -0.2630044 & \(\theta_{12}\) & -0.0021401 &
\(\theta_{32}\) & 0.0032237\tabularnewline
\(\theta_{v3}\) & 0.3411915 & \(\theta_{13}\) & 0.0047799 &
\(\theta_{33}\) & 0.0020681\tabularnewline
\(\theta_{v4}\) & -0.2583769 & \(\theta_{14}\) & 0.0025282 &
\(\theta_{34}\) & 0.0041429\tabularnewline
\(\theta_{h1}\) & -0.1939302 & \(\theta_{21}\) & 0.0012975 &
\(\theta_{41}\) & 0.0089533\tabularnewline
\(\theta_{h2}\) & -0.0572858 & \(\theta_{22}\) & 0.0000253 &
\(\theta_{42}\) & -0.0042403\tabularnewline
\(\theta_{h3}\) & -0.2101802 & \(\theta_{23}\) & -0.0004352 &
\(\theta_{43}\) & -0.0000480\tabularnewline
\(\theta_{h4}\) & 0.2402456 & \(\theta_{24}\) & -0.0086621 &
\(\theta_{44}\) & 0.0004767\tabularnewline
\bottomrule
\end{longtable}

\end{frame}

\begin{frame}[allowframebreaks]{References}
\protect\hypertarget{references}{}

\tiny

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-gelman2011inference}{}%
Gelman, Andrew, Kenneth Shirley, and others. 2011. ``Inference from
Simulations and Monitoring Convergence.'' \emph{Handbook of Markov Chain
Monte Carlo}, 163--74.

\leavevmode\hypertarget{ref-box1967discrimination}{}%
G. E. P. Box, W. J. Hill. 1967. ``Discrimination Among Mechanistic
Models.'' \emph{Technometrics} 9 (1). {[}Taylor \& Francis, Ltd.,
American Statistical Association, American Society for Quality{]}:
57--71.

\leavevmode\hypertarget{ref-handcock2003assessing}{}%
Handcock, Mark S. 2003. ``Assessing Degeneracy in Statistical Models of
Social Networks.'' Center for Statistics; the Social Sciences,
University of Washington. \url{http://www.csss.washington.edu/}.

\leavevmode\hypertarget{ref-hinton2006fast}{}%
Hinton, Geoffrey E, Simon Osindero, and Yee-Whye Teh. 2006. ``A Fast
Learning Algorithm for Deep Belief Nets.'' \emph{Neural Computation} 18
(7). MIT Press: 1527--54.

\leavevmode\hypertarget{ref-kaiser2007statistical}{}%
Kaiser, Mark S. 2007. ``Statistical Dependence in Markov Random Field
Models.'' \emph{Statistics Preprints} Paper 57. Digital Repository @
Iowa State University.
\url{http://lib.dr.iastate.edu/stat_las_preprints/57/}.

\leavevmode\hypertarget{ref-kaplan2016note}{}%
Kaplan, Andee, Daniel J Nordman, and Stephen B Vardeman. 2019a. ``On the
S-instability and degeneracy of discrete deep learning models.''
\emph{Information and Inference: A Journal of the IMA}, November.
\url{https://doi.org/10.1093/imaiai/iaz022}.

\leavevmode\hypertarget{ref-kaplan2019properties}{}%
Kaplan, Andee, Daniel Nordman, and Stephen Vardeman. 2019b. ``Properties
and Bayesian Fitting of Restricted Boltzmann Machines.''
\emph{Statistical Analysis and Data Mining: The ASA Data Science
Journal} 12 (1): 23--38. \url{https://doi.org/10.1002/sam.11396}.

\leavevmode\hypertarget{ref-le2008representational}{}%
Le Roux, Nicolas, and Yoshua Bengio. 2008. ``Representational Power of
Restricted Boltzmann Machines and Deep Belief Networks.'' \emph{Neural
Computation} 20 (6). MIT Press: 1631--49.

\leavevmode\hypertarget{ref-li2014biclustering}{}%
Li, Jing. 2014. ``Biclustering Methods and a Bayesian Approach to
Fitting Boltzmann Machines in Statistical Learning.'' PhD thesis, Iowa
State University; Graduate Theses; Dissertations.
\url{http://lib.dr.iastate.edu/etd/14173/}.

\leavevmode\hypertarget{ref-montufar2011refinements}{}%
Montufar, Guido, and Nihat Ay. 2011. ``Refinements of Universal
Approximation Results for Deep Belief Networks and Restricted Boltzmann
Machines.'' \emph{Neural Computation} 23 (5). MIT Press: 1306--19.

\leavevmode\hypertarget{ref-montufar2011expressive}{}%
Montúfar, Guido F, Johannes Rauh, and Nihat Ay. 2011. ``Expressive Power
and Approximation Errors of Restricted Boltzmann Machines.'' In
\emph{Advances in Neural Information Processing Systems}, 415--23. NIPS.

\leavevmode\hypertarget{ref-nguyen2014deep}{}%
Nguyen, Anh Mai, Jason Yosinski, and Jeff Clune. 2014. ``Deep Neural
Networks Are Easily Fooled: High Confidence Predictions for
Unrecognizable Images.'' \emph{arXiv Preprint arXiv:1412.1897}.
\url{http://arxiv.org/abs/1412.1897}.

\leavevmode\hypertarget{ref-salakhutdinov2009deep}{}%
Salakhutdinov, Ruslan, and Geoffrey E Hinton. 2009. ``Deep Boltzmann
Machines.'' In \emph{International Conference on Artificial Intelligence
and Statistics}, 448--55. AI \& Statistics.

\leavevmode\hypertarget{ref-schweinberger2011instability}{}%
Schweinberger, Michael. 2011. ``Instability, Sensitivity, and Degeneracy
of Discrete Exponential Families.'' \emph{Journal of the American
Statistical Association} 106 (496). Taylor \& Francis: 1361--70.

\leavevmode\hypertarget{ref-smolensky1986information}{}%
Smolensky, Paul. 1986. ``Information Processing in Dynamical Systems:
Foundations of Harmony Theory.'' DTIC Document.

\leavevmode\hypertarget{ref-szegedy2013intriguing}{}%
Szegedy, Christian, Wojciech Zaremba, Ilya Sutskever, Joan Bruna,
Dumitru Erhan, Ian J. Goodfellow, and Rob Fergus. 2013. ``Intriguing
Properties of Neural Networks.'' \emph{arXiv Preprint arXiv:1312.6199}.
\url{http://arxiv.org/abs/1312.6199}.

\leavevmode\hypertarget{ref-zhou2014some}{}%
Zhou, Wen. 2014. ``Some Bayesian and Multivariate Analysis Methods in
Statistical Machine Learning and Applications.'' PhD thesis, Iowa State
University; Graduate Theses; Dissertations.
\url{http://lib.dr.iastate.edu/etd/13816/}.

\end{frame}

\end{document}

### 2019-11-20 presentations/silo_wisc/slides.Rmd
update slides

@@
 **Deep belief network (DBN).** Multiple layers of latent random variables stacked in a deep architecture with no conditional dependence between layers;  all but the last stacked layer are Bayesian networks (directed dependence), rather than RBMs
+
+# Avoiding S-instability for the RBM
+
+Based on the definition of S-instability, we can show how an S-unstable distribution can arise (and potentially how to avoid this).
+
+- All instability in the marginal RBM model for the data $\boldsymbol X$ can be attributed to large magnitudes (or many terms) of $\boldsymbol \theta_{v}$ and/or $\boldsymbol \theta_{interaction}$
+- An S-unstable marginal model implies an S-unstable joint model
+- Further potential causes of S-instability exist for the joint model, often due to the size of $|\boldsymbol \theta_{v}|_1$
+
+#### Key Finding
+To prevent instability in the joint model of an RBM (and all models that embloy RBMs, e.g. a DBM), the combined magnitudes of all parameters $\boldsymbol \theta$ must be controlled.
 

### 2019-11-20 presentations/silo_wisc/slides.tex
update slides

@@
 
+\begin{frame}{Avoiding S-instability for the RBM}
+\protect\hypertarget{avoiding-s-instability-for-the-rbm}{}
+
+Based on the definition of S-instability, we can show how an S-unstable
+distribution can arise (and potentially how to avoid this).
+
+\begin{itemize}
+\tightlist
+\item
+  All instability in the marginal RBM model for the data
+  \(\boldsymbol X\) can be attributed to large magnitudes (or many
+  terms) of \(\boldsymbol \theta_{v}\) and/or
+  \(\boldsymbol \theta_{interaction}\)
+\item
+  An S-unstable marginal model implies an S-unstable joint model
+\item
+  Further potential causes of S-instability exist for the joint model,
+  often due to the size of \(|\boldsymbol \theta_{v}|_1\)
+\end{itemize}
+
+\begin{block}{Key Finding}
+
+To prevent instability in the joint model of an RBM (and all models that
+embloy RBMs, e.g.~a DBM), the combined magnitudes of all parameters
+\(\boldsymbol \theta\) must be controlled.
+
+\end{block}
+
+\end{frame}
+
 \begin{frame}{Visualizing properties in manageable (a.k.a. small) RBMs}
